---
title: "Dolutegravir (Punyawudho 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dolutegravir (Punyawudho 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Punyawudho_2025_dolutegravir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Punyawudho_2025_dolutegravir"))$description`
* Article: <https://doi.org/10.1002/psp4.13244>

Tuberculosis is the commonest opportunistic infection among people living with
HIV, and rifampicin is the cornerstone of its treatment. Rifampicin is a potent
inducer of the UGT1A1 and CYP3A4 pathways that clear dolutegravir, so
co-treated patients are told to take dolutegravir 50 mg **twice** daily instead
of the standard once-daily dose. In low- and middle-income countries
dolutegravir is dispensed as the fixed-dose TLD tablet (tenofovir disoproxil
fumarate 300 mg / lamivudine 300 mg / dolutegravir 50 mg), so the doubled dose
requires an extra single-agent pill, with the adherence and stock-out problems
that brings. This paper asks whether 100 mg **once** daily would do instead.

Every one of the 40 participants received rifampicin, so the parameter values in
this model are the *induced* values. The model cannot separate the rifampicin
effect from the drug-alone baseline, and the authors say so explicitly in their
limitations. It is a model of dolutegravir-with-rifampicin, not a model of
dolutegravir with a rifampicin covariate; for the latter see
`Kawuma_2023_dolutegravir` in this library.

Two features make the model unusual among the library's oral popPK entries and
they set up most of the checks below:

* **Bioavailability is fixed at 1 but carries a 90.6 %CV between-occasion
  variability.** Absorption is where nearly all of this drug's variability sits
  in this cohort; between-subject variability was retained on clearance only.
* **Total bilirubin is the sole retained covariate**, entering clearance
  through an exponential term. Dolutegravir and bilirubin compete for UGT1A1,
  so hyperbilirubinaemia raises dolutegravir exposure.

# Population

The model was fit to 332 dolutegravir plasma concentrations from 40 Thai adults
newly diagnosed with HIV/tuberculosis co-infection, enrolled at HIV-NAT (Thai
Red Cross AIDS Research Centre, Bangkok) inside NCT03731559 (Punyawudho 2025
Methods and Results). All were on rifampicin-based anti-tuberculosis therapy,
450 mg daily at 35-49 kg and 600 mg daily at 50 kg or more, and were randomised
to dolutegravir 50 mg once daily with food (n = 20) or 50 mg twice daily
without food (n = 20).

Intensive sampling was done at week 4: pre-dose and 1, 2, 4, 6, 8, 10 and 12 h
post-dose in both arms, with an extra 24 h sample in the once-daily arm.
Concentrations were measured by LC-MS/MS with an LLOQ of 0.1 mg/L; isolated
below-LLOQ values were imputed at LLOQ/2 and consecutive ones were dropped. One
participant whose concentrations were consistently below the LLOQ was removed
except for the pre-dose sample.

Baseline characteristics, from Punyawudho 2025 Table 1 (once-daily arm /
twice-daily arm): male 85% / 90%; age 37.5 / 35.6 years (range 25.0-60.5 and
21.6-53.2); weight 59.3 / 60.2 kg (range 41.1-78.4 and 47.1-86.0); serum
creatinine 0.891 / 0.895 mg/dL; total bilirubin 0.380 / 0.350 mg/dL (range
0.140-2.58 and 0.170-0.500).

The same information is available programmatically via
`readModelDb("Punyawudho_2025_dolutegravir")()$population`.

```{r population}
pop <- rxode2::rxode(readModelDb("Punyawudho_2025_dolutegravir"))$population
str(pop)
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Punyawudho_2025_dolutegravir.R`. They are collected
here for review. "Table 2" means Table 2 of Punyawudho 2025, column "NONMEM
Point estimate"; the parenthesised intervals in that table are the asymptotic
95% CIs, and a 1000-sample non-parametric bootstrap (97.3% successful
minimisations) agreed closely with all of them.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | `log(2.82)` | Table 2, `CL/F (L/h)` = 2.82, %RSE 7.2%, 95% CI 2.42-3.22; also Results text |
| `lvc` | `log(19.8)` | Table 2, `V/F (L)` = 19.8, %RSE 8.2%, 95% CI 16.6-22.9 |
| `lka` | `log(1.41)` | Table 2, `Ka (h-1)` = 1.41, %RSE 38.6%, 95% CI 0.344-2.48 |
| `ltlag` | `log(0.562)` | Table 2, `Lag time (h)` = 0.562, %RSE 33.6%, 95% CI 0.192-0.932 |
| `lfdepot` | `fixed(log(1))` | Table 2, `F` = "1 (fixed)"; Results, "The bioavailability (F) was fixed to 1" |
| `e_wt_cl` | `fixed(0.75)` | Methods, "The allometric exponents were fixed to the values of 0.75 and 1 for CL/F and V/F, respectively"; 60 kg reference from the Results CL/F equation |
| `e_wt_vc` | `fixed(1)` | Methods, as above |
| `e_tbili_cl` | `-0.297` | Table 2, `CL-Bilirubin` = -0.297, %RSE 9.7%, 95% CI -0.347 to -0.240; Results equation |
| `etalcl` | `0.0358313` | Table 2, `IIV-CL` = 19.1 %CV, 95% CI 12.8-23.7; `omega^2 = log(1 + 0.191^2)` |
| `etaiov_fdepot_*` | `0.5992957` | Table 2, `IOV-F1` = 90.6 %CV, 95% CI 86.9-94.1; `omega^2 = log(1 + 0.906^2)` |
| `etaiov_ka_*` | `0.2393318` | Table 2, `IOV-Ka` = 52.0 %CV, 95% CI 28.1-67.9; `omega^2 = log(1 + 0.520^2)` |
| `propSd` | `0.156` | Table 2, `RUV prop` = 15.6 %CV, 95% CI 9.27-20.0 |
| `CL/F = 2.82 * (WT/60)^0.75 * exp(-0.297 * (TBILI_mgdL - 0.38))` | n/a | Results, the paper's only displayed equation |
| One compartment, first-order absorption with lag time, first-order elimination | n/a | Results, "a one-compartment model with first-order absorption and elimination"; "The addition of lag time ... improved the model fit (dOFV = -15.9)"; "Adding a second compartment did not improve the fit" |
| Log-normal IIV and IOV; IIV on CL only; IOV on F and Ka | n/a | Methods ("assumed to be log-normally distributed"; "IOV was tested either following the inclusion of the IIV or substituting the IIV on absorption parameters (F and absorption rate constant; Ka)") and Results ("The IIV of V/F and lag time could not be precisely estimated"; "The addition of the IOV on the absorption parameters (F and Ka) significantly improved the fit") |
| Two occasions (pre-dose and post-dose) | n/a | Methods, "there were two occasions: the pre-dose occasion and the post-dose occasion" |
| Proportional residual error | n/a | Methods, "The residual unexplained variability (RUV) was characterized by proportional error model" |

```{r model-load}
mod <- readModelDb("Punyawudho_2025_dolutegravir")
ui <- rxode2::rxode(mod)
mod_typical <- rxode2::zeroRe(ui)

# The canonical TBILI column is SI umol/L; the paper works in US-convention
# mg/dL. 1 mg/dL = 17.1 umol/L (inst/references/covariate-columns.md).
MGDL_TO_UMOLL <- 17.1
```

# Deterministic checks on the printed equation

These are typical-value (`zeroRe()`) solves, so they carry no Monte-Carlo noise
at all and are the sharpest gates in the vignette: they pin the structure, the
units and the covariate arithmetic against numbers printed in the paper.

The first is the mass-balance identity for a linear one-compartment model with
complete input: the area under the plasma curve from a single dose out to
effective infinity is exactly `F * Dose / (CL/F)`, independently of the depot,
the absorption rate constant, the lag time and the ODE solver. With `F` fixed at
1 and a typical 60 kg individual at the cohort median total bilirubin of
0.38 mg/dL, the paper's printed clearance of 2.82 L/h therefore forces
`AUC = 50 / 2.82 = 17.73 mg*h/L`. A mis-transcribed clearance, a wrong reference
weight or a dose/volume unit slip all break it immediately.

```{r deterministic}
obs_grid <- c(seq(0, 6, by = 0.02), seq(6.1, 24, by = 0.1), seq(25, 480, by = 1))

typical_single_dose <- function(dose = 50, wt = 60, tbili_mgdl = 0.38) {
  d <- rxode2::et(amt = dose, cmt = "depot") |>
    rxode2::et(obs_grid) |>
    as.data.frame()
  d$WT <- wt
  d$TBILI <- tbili_mgdl * MGDL_TO_UMOLL
  d$OCC <- 1
  rxode2::rxSolve(mod_typical, d, returnType = "data.frame",
                  atol = 1e-10, rtol = 1e-8, addDosing = FALSE)
}

trap_auc <- function(df) {
  d <- dplyr::arrange(df, time)
  sum(diff(d$time) * (utils::head(d$Cc, -1) + utils::tail(d$Cc, -1)) / 2)
}

# Apparent clearance recovered from the solve, as Dose / AUC0-inf. Nothing here
# reads a model variable back out; it is inverted from the concentration curve.
cl_from_solve <- function(dose = 50, wt = 60, tbili_mgdl = 0.38) {
  dose / trap_auc(typical_single_dose(dose, wt, tbili_mgdl))
}

ref <- typical_single_dose()

deterministic <- tibble::tibble(
  Check = c(
    "AUC0-inf, 50 mg, 60 kg, TBILI 0.38 mg/dL (mg*h/L)",
    "CL/F recovered as Dose / AUC0-inf (L/h)",
    "CL/F ratio, TBILI 1.38 vs 0.38 mg/dL",
    "CL/F ratio, 80 vs 60 kg",
    "V/F recovered from the terminal slope (L)"
  ),
  Simulated = c(
    trap_auc(ref),
    cl_from_solve(),
    cl_from_solve(tbili_mgdl = 1.38) / cl_from_solve(),
    cl_from_solve(wt = 80) / cl_from_solve(),
    NA_real_
  ),
  `Punyawudho 2025` = c(50 / 2.82, 2.82, exp(-0.297), (80 / 60)^0.75, 19.8)
)

# Terminal-slope volume: kel = CL/V, so V = CL / kel, with kel read off the
# log-linear tail of the typical-value curve (far past the absorption phase).
tail_dat <- ref |> dplyr::filter(time >= 120, time <= 480, Cc > 0)
kel_hat <- -stats::coef(stats::lm(log(Cc) ~ time, data = tail_dat))[["time"]]
deterministic$Simulated[5] <- cl_from_solve() / kel_hat

deterministic <- deterministic |>
  dplyr::mutate(`Difference (%)` = 100 * (Simulated / `Punyawudho 2025` - 1))

knitr::kable(
  deterministic, digits = c(0, 5, 5, 4),
  caption = "Typical-value checks against values printed by Punyawudho 2025. Deterministic; the residuals are trapezoidal-integration and regression error only."
)

stopifnot(
  # Deterministic quantities. The realised residuals are all below 0.01%, so
  # 0.1% is a real gate: a 1 kg error in the reference weight moves the
  # clearance ratio by 1.2%, and a mg/dL-vs-umol/L slip on the bilirubin
  # centring moves the bilirubin ratio by a factor of 4.
  all(abs(deterministic$`Difference (%)`) < 0.1)
)

# The Results sentence "An elevation of 1 mg/dL in total bilirubin decreased the
# CL/F of DTG by 25.7%" is the same statement as row 3, read as a percentage.
pct_drop_per_mgdl <- 100 * (1 - cl_from_solve(tbili_mgdl = 1.38) / cl_from_solve())
stopifnot(abs(pct_drop_per_mgdl - 25.7) < 0.1)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the simulation design of Punyawudho 2025 Methods, "Simulations for evaluating
optimal dosage regimens": three regimens (50 mg once daily, 50 mg twice daily,
100 mg once daily) crossed with two body-weight strata (41-60 and 60.1-86 kg)
and the five DAIDS total-bilirubin grades (normal 0.1-1.29, grade 1 1.3-1.89,
grade 2 1.9-3.09, grade 3 3.1-6.09, grade 4 above 6.1 mg/dL), with weight and
bilirubin "allocated ... to each individual in each group ... at random". The
paper simulated 10,000 individuals per cell; the library caps vignette cohorts
at 200 per arm, so the percentages below carry a Monte-Carlo standard error of
up to about 3.5 percentage points and every assertion is written to tolerate
that.

Two encoding points matter.

* **Occasions alternate across doses.** The model carries two occasions, so
  `OCC` alternates 1, 2, 1, 2, ... across successive administrations, and each
  observation row takes the occasion of the dose that opened its interval. This
  is what gives consecutive doses independent bioavailability and absorption
  draws, which is the physical content of the paper's IOV.
* **The run-in is 720 h.** At grade 4 hyperbilirubinaemia clearance falls
  roughly ten-fold, so a subject in the low tail of the clearance distribution
  has a terminal half-life near 70 h; 720 h is more than ten of those.

```{r cohort}
n_arm <- 200L
t_end <- 720

# set.seed() seeds R's RNG, used here only for the covariate draws. It does NOT
# seed rxode2's simulation RNG, and rxode2's streams are partitioned per solver
# thread, so the etas are reproducible on this machine and different on a
# machine with a different thread count. Every assertion below is written to
# hold for any cohort the model can produce.
set.seed(20260907)

grades <- tibble::tribble(
  ~grade,    ~blo, ~bhi,
  "Normal",   0.1, 1.29,
  "Grade 1",  1.3, 1.89,
  "Grade 2",  1.9, 3.09,
  "Grade 3",  3.1, 6.09,
  "Grade 4",  6.1, 8.00
)
weights <- tibble::tribble(
  ~wtgrp,        ~wlo, ~whi,
  "41-60 kg",      41,   60,
  "60.1-86 kg",  60.1,   86
)
regimens <- tibble::tribble(
  ~regimen,        ~amt, ~ii,
  "50 mg OD",        50,  24,
  "50 mg b.i.d.",    50,  12,
  "100 mg OD",      100,  24
)

arms <- tidyr::expand_grid(regimens, weights, grades)
arms$arm_index <- seq_len(nrow(arms))

make_arm <- function(arm) {
  dose_times <- seq(0, t_end - arm$ii, by = arm$ii)
  n_dose <- length(dose_times)
  # id_offset keeps every arm's ids disjoint; duplicate ids across arms are
  # silently merged by rxSolve into one subject receiving the summed dose.
  ids <- (arm$arm_index - 1L) * n_arm + seq_len(n_arm)

  ev <- dplyr::bind_rows(
    tidyr::expand_grid(id = ids, time = dose_times) |>
      dplyr::mutate(amt = arm$amt, evid = 1, cmt = "depot"),
    tibble::tibble(id = ids, time = t_end, amt = 0, evid = 0, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))

  covs <- tibble::tibble(
    id    = ids,
    WT    = stats::runif(n_arm, arm$wlo, arm$whi),
    TBILI = stats::runif(n_arm, arm$blo, arm$bhi) * MGDL_TO_UMOLL
  )

  ev |>
    dplyr::left_join(covs, by = "id") |>
    dplyr::mutate(
      # Occasion index of the dosing interval this record falls in; the trough
      # at t_end belongs to the LAST interval, not to a notional (n+1)th dose.
      occ_index = pmin(floor(time / arm$ii), n_dose - 1),
      OCC       = ifelse(occ_index %% 2 == 0, 1, 2),
      regimen   = arm$regimen,
      wtgrp     = arm$wtgrp,
      grade     = arm$grade
    ) |>
    dplyr::select(-occ_index)
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), function(i) make_arm(arms[i, ])))

stopifnot(
  !anyDuplicated(unique(events[, c("id", "time", "evid")])),
  dplyr::n_distinct(events$id) == n_arm * nrow(arms)
)
```

```{r simulate}
rxode2::rxSetSeed(20260907)
sim <- rxode2::rxSolve(
  ui, events,
  keep = c("regimen", "wtgrp", "grade"),
  returnType = "data.frame", addDosing = FALSE
)

troughs <- sim |>
  dplyr::filter(time == t_end) |>
  dplyr::mutate(
    grade   = factor(grade, levels = grades$grade),
    regimen = factor(regimen, levels = regimens$regimen)
  )

stopifnot(
  nrow(troughs) == n_arm * nrow(arms),
  all(is.finite(troughs$Cc)),
  all(troughs$Cc > 0)
)
```

# Replicate published figures

```{r figure-3, fig.height = 5, fig.cap = "Replicates Figure 3 of Punyawudho 2025."}
ggplot(troughs, aes(grade, Cc, fill = regimen)) +
  geom_boxplot(outlier.size = 0.4, coef = 0, position = position_dodge(0.8)) +
  stat_summary(fun.min = function(x) stats::quantile(x, 0.05),
               fun.max = function(x) stats::quantile(x, 0.95),
               geom = "linerange", position = position_dodge(0.8)) +
  geom_hline(yintercept = 0.3, linetype = "dotted") +
  geom_hline(yintercept = 0.064, linetype = "dashed") +
  facet_wrap(~wtgrp) +
  scale_y_log10() +
  labs(
    x = "Total bilirubin grade", y = "Steady-state trough concentration (mg/L)",
    fill = NULL,
    title = "Figure 3 - simulated steady-state trough by bilirubin grade",
    caption = paste(
      "Boxes are the 25th-75th percentiles with the median; whiskers the 5th-95th.",
      "Dotted line: in vivo EC90 0.3 mg/L. Dashed line: PA-IC90 0.064 mg/L.",
      "Replicates Figure 3 of Punyawudho 2025."
    )
  ) +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 30, hjust = 1))
```

The qualitative shape of the paper's Figure 3 is reproduced: trough
concentrations rise monotonically with bilirubin grade in every regimen and both
weight strata, the twice-daily arm sits about two-fold above the 50 mg
once-daily arm at every grade, and 100 mg once daily lands between them.

# Replicate Table 3: target attainment

Table 3 of Punyawudho 2025 is the paper's headline quantitative output: the
percentage of simulated individuals whose steady-state trough exceeds the
in vitro protein-adjusted IC90 of 0.064 mg/L and the in vivo EC90 of 0.3 mg/L,
in each of the 30 regimen x weight x bilirubin-grade cells.

```{r table-3}
published <- tibble::tribble(
  ~wtgrp,       ~regimen,       ~grade,    ~pub_ic90, ~pub_ec90,
  "41-60 kg",   "50 mg OD",     "Normal",       79.2,      20.9,
  "41-60 kg",   "50 mg OD",     "Grade 1",      96.3,      58.2,
  "41-60 kg",   "50 mg OD",     "Grade 2",      99.4,      84.1,
  "41-60 kg",   "50 mg OD",     "Grade 3",     100.0,      98.8,
  "41-60 kg",   "50 mg OD",     "Grade 4",     100.0,     100.0,
  "41-60 kg",   "50 mg b.i.d.", "Normal",       99.9,      95.5,
  "41-60 kg",   "50 mg b.i.d.", "Grade 1",     100.0,      99.2,
  "41-60 kg",   "50 mg b.i.d.", "Grade 2",     100.0,      99.9,
  "41-60 kg",   "50 mg b.i.d.", "Grade 3",     100.0,     100.0,
  "41-60 kg",   "50 mg b.i.d.", "Grade 4",     100.0,     100.0,
  "41-60 kg",   "100 mg OD",    "Normal",       92.6,      49.7,
  "41-60 kg",   "100 mg OD",    "Grade 1",      98.9,      78.3,
  "41-60 kg",   "100 mg OD",    "Grade 2",      99.8,      93.8,
  "41-60 kg",   "100 mg OD",    "Grade 3",     100.0,      99.7,
  "41-60 kg",   "100 mg OD",    "Grade 4",     100.0,     100.0,
  "60.1-86 kg", "50 mg OD",     "Normal",       78.4,      16.2,
  "60.1-86 kg", "50 mg OD",     "Grade 1",      96.3,      53.4,
  "60.1-86 kg", "50 mg OD",     "Grade 2",      99.4,      80.7,
  "60.1-86 kg", "50 mg OD",     "Grade 3",     100.0,      98.3,
  "60.1-86 kg", "50 mg OD",     "Grade 4",     100.0,     100.0,
  "60.1-86 kg", "50 mg b.i.d.", "Normal",       99.9,      92.7,
  "60.1-86 kg", "50 mg b.i.d.", "Grade 1",      99.9,      94.8,
  "60.1-86 kg", "50 mg b.i.d.", "Grade 2",     100.0,      99.7,
  "60.1-86 kg", "50 mg b.i.d.", "Grade 3",     100.0,      99.9,
  "60.1-86 kg", "50 mg b.i.d.", "Grade 4",     100.0,     100.0,
  "60.1-86 kg", "100 mg OD",    "Normal",       92.9,      45.7,
  "60.1-86 kg", "100 mg OD",    "Grade 1",      98.9,      74.4,
  "60.1-86 kg", "100 mg OD",    "Grade 2",      99.8,      92.1,
  "60.1-86 kg", "100 mg OD",    "Grade 3",     100.0,      99.6,
  "60.1-86 kg", "100 mg OD",    "Grade 4",     100.0,     100.0
)

attainment <- function(df) {
  df |>
    dplyr::group_by(wtgrp, regimen, grade) |>
    dplyr::summarise(
      sim_ic90 = 100 * mean(Cc > 0.064),
      sim_ec90 = 100 * mean(Cc > 0.3),
      .groups  = "drop"
    )
}

tab3 <- attainment(troughs) |>
  dplyr::left_join(published, by = c("wtgrp", "regimen", "grade")) |>
  dplyr::mutate(
    d_ic90 = sim_ic90 - pub_ic90,
    d_ec90 = sim_ec90 - pub_ec90
  ) |>
  dplyr::arrange(wtgrp, regimen, grade)

tab3 |>
  dplyr::select(wtgrp, regimen, grade, sim_ic90, pub_ic90, d_ic90,
                sim_ec90, pub_ec90, d_ec90) |>
  dplyr::rename(
    "Weight"          = wtgrp,
    "Regimen"         = regimen,
    "Bilirubin"       = grade,
    "Sim > IC90 (%)"  = sim_ic90,
    "Pub > IC90 (%)"  = pub_ic90,
    "Diff (IC90)"     = d_ic90,
    "Sim > EC90 (%)"  = sim_ec90,
    "Pub > EC90 (%)"  = pub_ec90,
    "Diff (EC90)"     = d_ec90
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Replicates Table 3 of Punyawudho 2025. 'Pub' columns are transcribed from the paper (10,000 simulated individuals per cell); 'Sim' columns are this model at 200 per cell."
  )
```

```{r table-3-gates}
mad_ic90 <- mean(abs(tab3$d_ic90))
mad_ec90 <- mean(abs(tab3$d_ec90))

stopifnot(
  # Aggregate agreement, pooled over 6000 simulated subjects. Realised on the
  # authoring machine: 0.67 and 2.16 percentage points. The bounds are set well
  # outside that, because at 200 subjects per cell a single cell's Monte-Carlo
  # standard error reaches 3.5 percentage points; they are still gates -- a
  # 10% error in clearance moves the once-daily EC90 column by ~10 points and
  # a lost bilirubin effect moves the grade 2-4 columns by tens of points.
  mad_ic90 < 5,
  mad_ec90 < 8,
  # No individual cell may be wildly off. Realised maxima: 4.6 percentage
  # points on the IC90 column and 11.4 on the EC90 column (the 50 mg once-daily
  # grade 1 cells, where the attainment percentage sits near 50% and the
  # Monte-Carlo standard error at 200 subjects is therefore at its largest).
  max(abs(tab3$d_ic90)) < 15,
  max(abs(tab3$d_ec90)) < 20
)

# The paper's conclusions are orderings, and they are large effects that no
# plausible cohort draw can invert: twice-daily beats 100 mg once daily beats
# 50 mg once daily on EC90 attainment at normal bilirubin, and the twice-daily
# arm clears the IC90 essentially completely.
normal <- tab3 |> dplyr::filter(grade == "Normal")
ec90_by_regimen <- normal |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(ec90 = mean(sim_ec90), ic90 = mean(sim_ic90), .groups = "drop")

stopifnot(
  # Published gaps at normal bilirubin are ~74 and ~29 points; require half.
  ec90_by_regimen$ec90[ec90_by_regimen$regimen == "50 mg b.i.d."] -
    ec90_by_regimen$ec90[ec90_by_regimen$regimen == "50 mg OD"] > 35,
  ec90_by_regimen$ec90[ec90_by_regimen$regimen == "100 mg OD"] -
    ec90_by_regimen$ec90[ec90_by_regimen$regimen == "50 mg OD"] > 12,
  # "DTG 100 mg OD could serve as an alternative regimen" rests on this: more
  # than 90% of normal-bilirubin individuals clear the in vitro IC90. Published
  # values are 92.6 and 92.9; the gate allows for Monte-Carlo error but would
  # fail on the 79% that 50 mg once daily achieves.
  ec90_by_regimen$ic90[ec90_by_regimen$regimen == "100 mg OD"] > 85,
  ec90_by_regimen$ic90[ec90_by_regimen$regimen == "50 mg b.i.d."] > 97
)
```

# Which %CV convention?

Punyawudho 2025 Table 2 reports its random effects in a section headed
"Inter-individual/Inter-occasion variability (%CV)" but never states the formula
behind that column. Two readings are in common use for a log-normal random
effect, and they disagree materially at 90.6%:

* the **exact** log-normal coefficient of variation,
  `CV = sqrt(exp(omega^2) - 1)`, giving `omega^2 = log(1 + 0.906^2) = 0.599`;
* the **approximation** `CV = sqrt(omega^2)`, giving `omega^2 = 0.906^2 = 0.821`.

Nothing in the paper adjudicates directly, but Table 3 does so indirectly: the
attainment percentages are a functional of the whole simulated trough
distribution, so a wider bioavailability variance shifts them. The comparison
below re-solves the identical cohort under each reading with common random
numbers -- `rxSetSeed()` is re-issued before each solve and only the omega
matrix changes, so each subject keeps the same underlying standard-normal draws
and the two columns are paired.

```{r convention}
omega_sd <- ui$omega
om_diag  <- diag(omega_sd)
om_names <- rownames(omega_sd)
om_diag[om_names == "etalcl"] <- 0.191^2
om_diag[grepl("fdepot", om_names)] <- 0.906^2
om_diag[grepl("ka", om_names)] <- 0.520^2
diag(omega_sd) <- om_diag

rxode2::rxSetSeed(20260907)
sim_sd <- rxode2::rxSolve(
  ui, events, omega = omega_sd,
  keep = c("regimen", "wtgrp", "grade"),
  returnType = "data.frame", addDosing = FALSE
)

convention <- dplyr::bind_rows(
  attainment(troughs) |> dplyr::mutate(Convention = "CV = sqrt(exp(omega^2) - 1) (retained)"),
  attainment(dplyr::filter(sim_sd, time == t_end)) |>
    dplyr::mutate(Convention = "CV = sqrt(omega^2)")
) |>
  dplyr::left_join(published, by = c("wtgrp", "regimen", "grade")) |>
  dplyr::group_by(Convention) |>
  dplyr::summarise(
    `omega^2 for IOV-F1`         = ifelse(dplyr::first(Convention) == "CV = sqrt(omega^2)",
                                          0.906^2, log(1 + 0.906^2)),
    `Mean |diff|, IC90 column`   = mean(abs(sim_ic90 - pub_ic90)),
    `Mean |diff|, EC90 column`   = mean(abs(sim_ec90 - pub_ec90)),
    .groups = "drop"
  )

knitr::kable(
  convention, digits = 3,
  caption = "Mean absolute deviation from Table 3 of Punyawudho 2025 over all 30 cells, under the two readings of the paper's %CV column. Common random numbers; only the omega matrix differs."
)
```

The exact-CV reading is closer on both columns, and it is the one the packaged
model carries. The margin is small, and deliberately not asserted on: at 200
subjects per cell it is comparable to Monte-Carlo error, and the honest summary
is that **both readings reproduce Table 3 well** -- the choice moves the
bioavailability IOV variance from 0.599 to 0.821 and the target-attainment
percentages by one to two points. A reader who obtains the paper's NONMEM
control stream should treat this as the one convention question worth
re-checking; `omega` can be overridden at the `rxSolve()` call exactly as above
without editing the model file.

# Replicate Figure 2: the study's own regimens

Figure 2 of Punyawudho 2025 is a visual predictive check of the two randomised
arms at the week-4 intensive-sampling visit. The observed concentrations are not
public, so only the model-predicted percentile bands can be reproduced. The
cohort mirrors Table 1: weight and total bilirubin are drawn per arm from
truncated normals matching the reported means, standard deviations and ranges.

Note that food does **not** appear in the model. The paper screened it and found
no effect on either bioavailability or lag time (Results and Discussion), so the
"with food" and "without food" labels below distinguish the randomised arms but
carry no parameter.

```{r vpc-cohort}
n_vpc <- 200L
vpc_runin <- 14 * 24    # 14 days; the slowest VPC-cohort subject has a
                        # terminal half-life near 10 h at TBILI 2.58 mg/dL

rtnorm <- function(n, mean, sd, lo, hi) {
  x <- stats::rnorm(n, mean, sd)
  # Reflect rather than resample so the draw count is fixed and the tails stay
  # inside the reported range.
  pmin(pmax(x, lo), hi)
}

set.seed(20260908)

vpc_arms <- tibble::tribble(
  ~arm,                         ~amt, ~ii, ~tau, ~wt_m, ~wt_s, ~wt_lo, ~wt_hi, ~b_m,  ~b_s,  ~b_lo, ~b_hi, ~offset,
  "50 mg once daily, fed",        50,  24,   24,  59.3,  13.2,   41.1,   78.4, 0.380, 0.250, 0.140, 2.580,      0L,
  "50 mg twice daily, fasted",    50,  12,   12,  60.2,  18.5,   47.1,   86.0, 0.350, 0.150, 0.170, 0.500,   1000L
)

make_vpc_arm <- function(a) {
  # The dose train runs THROUGH the start of the observation window, so the
  # window is a complete dosing interval opened by the dose at vpc_runin.
  dose_times <- seq(0, vpc_runin, by = a$ii)
  n_dose <- length(dose_times)
  ids <- a$offset + seq_len(n_vpc)
  grid <- vpc_runin + c(seq(0, a$tau, by = 0.1))

  ev <- dplyr::bind_rows(
    tidyr::expand_grid(id = ids, time = dose_times) |>
      dplyr::mutate(amt = a$amt, evid = 1, cmt = "depot"),
    tidyr::expand_grid(id = ids, time = grid) |>
      dplyr::mutate(amt = 0, evid = 0, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))

  covs <- tibble::tibble(
    id    = ids,
    WT    = rtnorm(n_vpc, a$wt_m, a$wt_s, a$wt_lo, a$wt_hi),
    TBILI = rtnorm(n_vpc, a$b_m, a$b_s, a$b_lo, a$b_hi) * MGDL_TO_UMOLL
  )

  ev |>
    dplyr::left_join(covs, by = "id") |>
    dplyr::mutate(
      occ_index = pmin(floor(time / a$ii), n_dose - 1),
      OCC       = ifelse(occ_index %% 2 == 0, 1, 2),
      arm       = a$arm,
      amt_arm   = a$amt,
      tau       = a$tau
    ) |>
    dplyr::select(-occ_index)
}

vpc_events <- dplyr::bind_rows(
  lapply(seq_len(nrow(vpc_arms)), function(i) make_vpc_arm(vpc_arms[i, ]))
)
stopifnot(!anyDuplicated(unique(vpc_events[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20260908)
vpc_sim <- rxode2::rxSolve(
  ui, vpc_events, keep = c("arm", "amt_arm", "tau"),
  returnType = "data.frame", addDosing = FALSE
) |>
  dplyr::mutate(tad = time - vpc_runin)

stopifnot(nrow(vpc_sim) > 0, all(is.finite(vpc_sim$Cc)), all(vpc_sim$Cc > 0))
```

```{r figure-2, fig.cap = "Replicates Figure 2 of Punyawudho 2025."}
vpc_sim |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 0.9, colour = "steelblue") +
  geom_hline(yintercept = 0.3, linetype = "dotted") +
  geom_hline(yintercept = 0.064, linetype = "dashed") +
  facet_wrap(~arm, scales = "free_x") +
  scale_y_log10() +
  labs(
    x = "Time after dose at steady state (h)",
    y = "Dolutegravir concentration (mg/L)",
    title = "Figure 2 - predicted steady-state profiles by randomised arm",
    caption = paste(
      "Median with 5th-95th percentile band, 200 simulated subjects per arm.",
      "Dotted line: in vivo EC90 0.3 mg/L. Dashed line: PA-IC90 0.064 mg/L.",
      "Replicates the model-predicted bands of Figure 2 of Punyawudho 2025."
    )
  ) +
  theme_bw()
```

# PKNCA validation

Punyawudho 2025 publishes no non-compartmental analysis, so there is no
`ncaComparisonTable()` comparison to make. PKNCA is instead used to close the
loop on the exposure identity that underwrites the whole dosing argument: at
steady state, the area under the curve over one dosing interval is
`F * Dose / (CL/F)` for **any** interval length, so the once-daily and
twice-daily arms of a typical individual must return the same `AUC(0, tau)` --
50 / 2.82 = 17.73 mg*h/L -- even though their intervals differ two-fold. This
identity is what makes "100 mg once daily" and "50 mg twice daily" deliver equal
daily exposure while differing in trough, which is the paper's entire argument.

The gate runs on typical-value (`zeroRe()`) profiles, so it is deterministic and
the residual is pure trapezoidal error.

```{r pknca-typical}
typical_ss <- function(dose, ii, tau, wt = 60, tbili_mgdl = 0.38, label) {
  runin <- 480
  # As in the VPC block: dose through the start of the observation window so
  # the window is a complete dosing interval, not the tail after the last dose.
  dose_times <- seq(0, runin, by = ii)
  grid <- runin + seq(0, tau, by = 0.02)
  d <- dplyr::bind_rows(
    tibble::tibble(time = dose_times, amt = dose, evid = 1, cmt = "depot"),
    tibble::tibble(time = grid, amt = 0, evid = 0, cmt = "central")
  ) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    dplyr::mutate(
      id    = 1,
      WT    = wt,
      TBILI = tbili_mgdl * MGDL_TO_UMOLL,
      OCC   = 1
    )
  # A single-subject solve returns no `id` column, so re-attach it explicitly
  # rather than assuming rxSolve carries it through.
  rxode2::rxSolve(mod_typical, d, returnType = "data.frame",
                  atol = 1e-10, rtol = 1e-8, addDosing = FALSE) |>
    dplyr::mutate(id = 1L, time = time - runin, treatment = label,
                  amt_dose = dose, tau = tau)
}

ss_profiles <- dplyr::bind_rows(
  typical_ss(50, 24, 24, label = "50 mg OD"),
  typical_ss(50, 12, 12, label = "50 mg b.i.d."),
  typical_ss(100, 24, 24, label = "100 mg OD")
)

conc_df <- ss_profiles |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 record per (treatment, id). The steady-state grid already
# starts at the dosing time; this is a defensive no-op that would supply one if
# the grid ever changed, and keeps PKNCA from warning about an AUC range that
# starts before the first measurement.
conc_df <- dplyr::bind_rows(
  conc_df,
  conc_df |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(treatment, id, time, .keep_all = TRUE) |>
  dplyr::arrange(treatment, id, time)

dose_df <- ss_profiles |>
  dplyr::distinct(treatment, id, amt_dose) |>
  dplyr::mutate(time = 0) |>
  dplyr::rename(amt = amt_dose)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

tau_by_trt <- ss_profiles |> dplyr::distinct(treatment, tau)
intervals <- data.frame(
  start   = 0,
  end     = tau_by_trt$tau,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cmin    = TRUE
)
intervals$treatment <- tau_by_trt$treatment

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(
    dplyr::distinct(ss_profiles, treatment, amt_dose), by = "treatment"
  ) |>
  dplyr::mutate(
    `Closed form Dose / CL` = amt_dose / 2.82,
    `AUC difference (%)`    = 100 * (auclast / `Closed form Dose / CL` - 1)
  )

nca_wide |>
  dplyr::select(treatment, cmax, tmax, cmin, auclast,
                `Closed form Dose / CL`, `AUC difference (%)`) |>
  dplyr::rename(
    "Regimen"              = treatment,
    "Cmax (mg/L)"          = cmax,
    "Tmax (h)"             = tmax,
    "Ctrough (mg/L)"       = cmin,
    "AUCtau (mg*h/L)"      = auclast
  ) |>
  knitr::kable(
    digits = 4,
    caption = "PKNCA on typical-value steady-state profiles (60 kg, total bilirubin 0.38 mg/dL). AUCtau must equal Dose / (CL/F) exactly for every regimen and every interval length."
  )

stopifnot(
  # Deterministic. The realised residual is trapezoidal error on a 0.02 h grid.
  all(abs(nca_wide$`AUC difference (%)`) < 0.1),
  # The two 50 mg arms differ in interval length but not in AUCtau.
  abs(nca_wide$auclast[nca_wide$treatment == "50 mg b.i.d."] /
        nca_wide$auclast[nca_wide$treatment == "50 mg OD"] - 1) < 0.001,
  # 100 mg once daily doubles daily exposure relative to 50 mg once daily.
  abs(nca_wide$auclast[nca_wide$treatment == "100 mg OD"] /
        nca_wide$auclast[nca_wide$treatment == "50 mg OD"] - 2) < 0.001,
  # No typical-value profile peaks before the absorption lag time. Structural
  # and deterministic: a dropped or zeroed alag(depot) puts Tmax at 0.
  all(nca_wide$tmax > 0.562),
  # The typical individual on 50 mg once daily sits between the two targets --
  # above the 0.064 mg/L PA-IC90 but below the 0.3 mg/L EC90 -- which is the
  # structural reason the paper's 50 mg OD row of Table 3 reads ~79% / ~21%.
  nca_wide$cmin[nca_wide$treatment == "50 mg OD"] > 0.064,
  nca_wide$cmin[nca_wide$treatment == "50 mg OD"] < 0.3,
  # Twice-daily dosing puts the typical trough above the EC90.
  nca_wide$cmin[nca_wide$treatment == "50 mg b.i.d."] > 0.3
)
```

The typical-value troughs read off that table are the structural counterpart of
the paper's Table 3: 50 mg once daily leaves the typical individual below the
0.3 mg/L in vivo EC90, 100 mg once daily lifts it to roughly twice the
once-daily value, and twice-daily dosing puts it comfortably above. The paper's
percentages are the cohort version of the same statement.

```{r pknca-cohort}
# The same NCA parameters over the stochastic VPC cohort, for the two arms the
# study actually randomised. This is descriptive: the paper reports no NCA to
# compare against, so no assertion is made on the values, only on their
# structural sanity.
cohort_conc <- vpc_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, time = tad, Cc, treatment = arm)

cohort_conc <- dplyr::bind_rows(
  cohort_conc,
  cohort_conc |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(treatment, id, time, .keep_all = TRUE) |>
  dplyr::arrange(treatment, id, time)

cohort_dose <- vpc_sim |>
  dplyr::distinct(id, treatment = arm, amt_arm) |>
  dplyr::mutate(time = 0) |>
  dplyr::rename(amt = amt_arm)

cohort_tau <- vpc_sim |> dplyr::distinct(treatment = arm, tau)
cohort_intervals <- data.frame(
  start = 0, end = cohort_tau$tau,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, cmin = TRUE
)
cohort_intervals$treatment <- cohort_tau$treatment

cohort_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(cohort_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(cohort_dose, amt ~ time | treatment + id),
  intervals = cohort_intervals
))

cohort_wide <- as.data.frame(cohort_nca) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(cohort_wide) == 2 * n_vpc, !anyNA(cohort_wide$cmax))

cohort_wide |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cmax    = stats::median(cmax),
    tmax    = stats::median(tmax),
    cmin    = stats::median(cmin),
    auclast = stats::median(auclast),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Arm"                     = treatment,
    "Median Cmax (mg/L)"      = cmax,
    "Median Tmax (h)"         = tmax,
    "Median Ctrough (mg/L)"   = cmin,
    "Median AUCtau (mg*h/L)"  = auclast
  ) |>
  knitr::kable(
    digits = 3,
    caption = "PKNCA over the simulated week-4 cohort of each randomised arm. Descriptive only; Punyawudho 2025 reports no non-compartmental analysis."
  )

stopifnot(
  # Median Tmax sits in the window the paper's 1 and 2 h post-dose samples
  # bracket. Asserted on the median, NOT on `all(tmax > 0.562)`: at steady
  # state a subject whose current occasion drew a small bioavailability can
  # have its interval maximum at the interval-opening trough, carried over
  # from a previous dose, so Tmax = 0 is legitimate model behaviour and does
  # occur (about 0.5% of the twice-daily arm on the authoring machine). The
  # structural "no peak before the lag" statement is asserted deterministically
  # on the typical-value profiles above, where it is exactly true.
  dplyr::between(stats::median(cohort_wide$tmax), 1, 6),
  # Median AUCtau against the typical-value closed form Dose / (CL/F). The
  # median sits slightly BELOW it: bioavailability is log-normal with median 1,
  # and the cohort's weight distribution is centred near the 60 kg reference.
  # Realised on the authoring machine: 0.96 and 1.13 for the two arms. The band
  # is wide enough for any cohort draw and still fails on a factor-of-two dose,
  # volume or clearance error.
  all(dplyr::between(
    cohort_wide |>
      dplyr::group_by(treatment) |>
      dplyr::summarise(r = stats::median(auclast) / (50 / 2.82),
                       .groups = "drop") |>
      dplyr::pull(r),
    0.6, 1.5
  ))
)
```

# Assumptions and deviations

* **The `%CV` convention for the random effects is not stated by the paper.**
  The exact log-normal form `CV = sqrt(exp(omega^2) - 1)` is used, giving
  `omega^2 = log(1 + (CV/100)^2)`. The arbitration is the "Which %CV
  convention?" section above: it reproduces Table 3 slightly better than the
  `CV = sqrt(omega^2)` approximation, but the margin is comparable to
  Monte-Carlo error at this cohort size and both readings reproduce the table
  well. This is the single most consequential undocumented choice in the
  extraction; the residual-error row is unaffected, since for a proportional
  error model the tabulated %CV is the standard deviation directly.
* **The reference weight for `V/F` is assumed to be 60 kg**, the same as for
  `CL/F`. The paper prints the allometric reference only inside its `CL/F`
  equation, while stating in Methods that both parameters were scaled
  allometrically with fixed exponents. A shared reference is the standard
  construction and is consistent with the cohort medians of 59.3 and 60.2 kg.
* **Total bilirubin is carried in SI umol/L** because that is the canonical unit
  in `inst/references/covariate-columns.md`, and converted to the paper's mg/dL
  inline in `model()` at 1 mg/dL = 17.1 umol/L. The centring value 0.38 mg/dL
  is 6.50 umol/L. Pass `TBILI` in umol/L.
* **Two occasions are encoded**, exactly as the paper defines them ("the
  pre-dose occasion and the post-dose occasion"). For the multi-dose steady-
  state simulations here, `OCC` is alternated across successive doses so that
  consecutive administrations draw independent bioavailability and absorption
  behaviour. The paper does not state how it mapped its two occasions onto the
  10,000-subject steady-state simulation; alternating is the reading that
  preserves the physical meaning of between-occasion variability.
* **Grade 4 hyperbilirubinaemia is bounded at 8 mg/dL** for simulation. The
  DAIDS grade is open-ended ("> 6.1 mg/dL") and the paper does not say what
  upper bound it sampled to. Every grade 3 and grade 4 cell of Table 3 is
  100.0% in both targets, so the choice cannot affect the comparison.
* **Covariate distributions within a simulation cell are uniform**, following
  the paper's "The allocation of weight and total bilirubin value to each
  individual in each group was performed at random". The distributional family
  is not stated.
* **Cohort size is 200 per arm, not the paper's 10,000.** This is a library-wide
  vignette cap. It puts a Monte-Carlo standard error of up to 3.5 percentage
  points on each cell of the Table 3 comparison, which is why the assertions
  there are on the pooled mean absolute deviation and on effect orderings rather
  than on individual cells.
* **The VPC cohort's covariate draws are truncated normals** matching the Table 1
  summaries, clipped to the reported ranges. Punyawudho 2025 does not publish
  the observed concentrations, so Figure 2 can be reproduced only as
  model-predicted percentile bands, not as a true visual predictive check
  against data.
* **Food carries no parameter.** The paper randomised one arm to dosing with
  food and screened food as a covariate, finding no effect on either
  bioavailability or lag time. The arm labels in the Figure 2 reproduction are
  therefore descriptive only. The authors flag this as a limitation and
  attribute the residual 0-5 h over-prediction in their goodness-of-fit plots to
  fed subjects whose absorption was delayed beyond the estimated lag.
* **This model is valid only with rifampicin.** Every participant received it,
  so `CL/F` = 2.82 L/h is the induced clearance. There is no covariate that can
  be switched off to obtain dolutegravir alone.

## Errata and source-reporting notes

* **Table 1 row labels say "median (range)" but the values are formatted as
  "mean +/- SD (range)".** For example age in the once-daily arm is
  "37.5 +/- 17.7 (25.0-60.5)". A median cannot carry a standard deviation, and
  17.7 years is implausible as a spread for a range of 25.0-60.5 in n = 20, so
  the "+/-" figure is read here as a standard deviation and the leading figure
  as a mean. The population metadata records both readings' consequences by
  quoting the ranges, which are unambiguous. No model parameter depends on this.
* **Table 1's arm concentration counts sum to 300, not the 332 reported in the
  Results.** The table gives 160 and 140 plasma concentrations for the two arms
  while the Results text and Abstract both state "A total of 332 DTG
  concentrations from 40 PLWH were analyzed". No model parameter depends on the
  count; `population$n_observations` records the Results figure of 332.
* **The Table 2 asymptotic 95% CI for `IOV-F1` (86.9-94.1) is far narrower than
  its own %RSE of 20.3% implies** and than the bootstrap interval of 70.0-134.
  The point estimate of 90.6 is used, which is unaffected.
* **The Discussion quotes the trough targets as "0.064 and 0.3 mg/mL"** in two
  places where the Methods, Abstract, Table 3 and Figure 3 all say mg/L. mg/L is
  correct; mg/mL would be a thousand-fold error.
* **No supplement was needed.** The Supporting Information is cited only for
  concentration-time profiles stratified by treatment group; every parameter
  value, the covariate equation and the full simulation design are in the main
  text.
