---
title: "Vestronidase alfa (Qi 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vestronidase alfa (Qi 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("Qi_2019_vestronidaseAlfa")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1007/s40262-018-0721-y>
* Erratum: <https://doi.org/10.1007/s40262-018-0726-6>
* Full text (open access): <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6451706/>

Vestronidase alfa (UX003) is a recombinant human beta-glucuronidase enzyme
replacement therapy for mucopolysaccharidosis type VII (MPS VII, Sly syndrome).
Qi 2019 pooled three trials to build a two-compartment population PK model and a
sequential inhibitory-Imax exposure-response model for urinary glycosaminoglycan
(uGAG) reduction, and used them to justify the approved 4 mg/kg every-other-week
(QOW) regimen.

# Population

The analysis pooled 23 subjects with MPS VII from three trials (Qi 2019 Table 1):
3 subjects from a phase I/II dose-titration study (NCT01856218, 1-4 mg/kg QOW),
8 subjects under 5 years of age from a phase II study (NCT02418455, 4 mg/kg QOW),
and 12 subjects from a phase III blind-start placebo-controlled study
(NCT02230566, 4 mg/kg QOW). Ages ranged from 1.7 to 25.3 years (median 10.1) and
baseline body weights from 9.41 to 104.0 kg (median 24.8, mean 37.7). The cohort
was 52% female and 61% White / 22% Other / 13% Asian / 4% Black. Fifteen subjects
contributed intensively sampled PK profiles and eight contributed sparse
profiles. All doses were given as approximately 4-hour intravenous infusions in
which the first 2.5% of the total volume was infused over the first hour.

The same metadata is available programmatically:

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Qi_2019_vestronidaseAlfa.R`. They are collected here
for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 1.97 L/h | Table 2, point estimate (RSE 9.61%) |
| `lvc` (Vc) | 1.52 L | Table 2, point estimate (RSE 9.64%) |
| `lq` (Q) | 0.931 L/h | Table 2, point estimate (RSE 16.3%) |
| `lvp` (Vp) | 3.11 L | Table 2, point estimate (RSE 4.23%) |
| `e_wt_cl_q` | 0.587 | Table 2, "BWT on CL and Q" (RSE 12.4%) |
| `e_wt_vc_vp` | 0.483 | Table 2, "BWT on Vc and Vp" (RSE 17.7%) |
| Reference weight 20 kg | n/a | Table 2 footnote b ("Centered on a body weight of 20 kg") |
| Exponents estimated, not fixed | n/a | Table 2 footnote c (CIs exclude 0.75 / 1) |
| `etalcl`, `etalvc`, `etalq`, `etalvp` | 0.170, 0.0944, 0.563, 0.0236 | Table 2, omega^2 column (reported directly) |
| `propSd` | sqrt(0.109) = 0.330 | Table 2, sigma^2 proportional = 0.109 (CV 33.0%) |
| IIV model `theta_i = theta_TV * exp(eta_i)` | n/a | Methods, Population PK Modeling, first displayed equation |
| Residual model `DV = IPRED * (1 + eps)` | n/a | Methods, Population PK Modeling, second displayed equation |
| `rbase_ucs`, `rbase_uds` (I0) | fixed at 0 | Methods, Exposure-Response Analysis ("I0 is the no-drug response and fixed to 0") |
| `limax_ucs` / `limax_uds` | 82.0% / 76.9% | Table 4, Imax row |
| `lic50_ucs` / `lic50_uds` | 8.6 / 7.3 ug*h/mL | Table 4, IC50 row |
| `etalimax_ucs` / `etalimax_uds` | 7.9% / 8.7% CV | Table 4, IIV for Imax row |
| `addSd_uCSchange` / `addSd_uDSchange` | 6.1 / 6.4 | Table 4, additive residual error row |
| `I = I0 - Imax * AUC / (AUC + IC50)` | n/a | Methods, Exposure-Response Analysis, displayed equation |

Two transcription points are worth stating explicitly because both are easy to
get wrong from a secondary source.

**The omega values are variances, transcribed directly.** Table 2 reports
`omega^2` in its own column, so no back-transformation from the CV% column is
performed. That matters here: Table 2 footnote *d* declares that the CV% column
uses `CV = sqrt(exp(omega^2) - 1)` **only when `omega^2 > 0.15`**, and a plain
`sqrt(omega^2)` below that threshold. The column therefore switches formula row
by row (log-normal for CL and Q, plain SD for Vc and Vp), and inverting it with a
single formula would misstate the Vc and Vp variances by a few percent.

**The exposure-response equation carries a leading minus sign.** The displayed
equation is `I = I0 - Imax * AUC / (AUC + IC50)`, verified against the JATS
MathML of the source XML. With `I0` fixed at 0, `I` is the percent *change from
baseline* and is therefore negative, which is what the Figure 4 axis
("% Change from baseline") shows.

# Structural identities

These checks are deterministic: they compare the packaged model against
quantities the paper derives from the very same parameters, so any disagreement
is a transcription error rather than simulation noise. They are gated tightly.

```{r identities}
# Typical values at the 20 kg reference weight, read back out of the model.
th <- setNames(ui$theta, names(ui$theta))
CL <- exp(th[["lcl"]]); Vc <- exp(th[["lvc"]])
Q  <- exp(th[["lq"]]);  Vp <- exp(th[["lvp"]])

# Terminal disposition half-life of a two-compartment model.
k10 <- CL / Vc; k12 <- Q / Vc; k21 <- Q / Vp
a <- k10 + k12 + k21
lambda_z <- (a - sqrt(a^2 - 4 * k10 * k21)) / 2
thalf <- log(2) / lambda_z

# Weight-normalised CL and Vss at the cohort median weight (24.8 kg). The paper
# normalises the 20 kg typical values by the median weight WITHOUT re-applying
# the allometric exponents -- see the note below.
wt_median <- 24.8
cl_per_kg  <- CL / wt_median
vss_per_kg <- (Vc + Vp) / wt_median

ident <- tibble::tibble(
  Quantity = c("Terminal half-life (h)",
               "CL normalised by median weight (L/h/kg)",
               "Vss = Vc + Vp normalised by median weight (L/kg)"),
  Model    = c(thalf, cl_per_kg, vss_per_kg),
  Published = c(3.6, 0.079, 0.19),
  Source   = c("Results, Population PK", "Results, Population PK", "Results, Population PK")
) |>
  mutate(`% diff` = 100 * (Model - Published) / Published)

knitr::kable(ident, digits = 3,
             caption = "Model-internal identities against the values Qi 2019 derives from the same parameters.")

stopifnot(
  abs(thalf - 3.6) / 3.6 < 0.01,          # paper rounds to 3.6 h; model gives 3.601
  abs(cl_per_kg - 0.079) / 0.079 < 0.01,
  abs(vss_per_kg - 0.19) / 0.19 < 0.02
)
```

Note that the paper's per-kilogram values are the 20 kg typical values divided by
the median weight, *not* allometrically rescaled to 24.8 kg. Rescaling would give
CL = 1.97 * (24.8/20)^0.587 = `r round(1.97 * (24.8/20)^0.587, 3)` L/h, i.e.
`r round(1.97 * (24.8/20)^0.587 / 24.8, 4)` L/h/kg. The identity above reproduces
the paper's own arithmetic, and the allometric value is used everywhere a profile
is actually simulated.

# Event tables

Vestronidase alfa is infused over approximately 4 hours with the first 2.5% of
the volume delivered in the first hour, so every dose is a two-rate infusion.

```{r helpers}
# Two-rate 4-hour infusion: 2.5% of the dose over hour 1, 97.5% over hours 1-4.
qi_dose_rows <- function(id, dose_mg, start_times, wt) {
  n <- length(start_times)
  tibble::tibble(
    # Interleave the two rate segments of each infusion: t0, t0 + 1 h.
    time = as.vector(rbind(start_times, start_times + 1)),
    amt  = rep(c(0.025 * dose_mg, 0.975 * dose_mg), times = n),
    rate = rep(c(0.025 * dose_mg / 1, 0.975 * dose_mg / 3), times = n),
    id = id, evid = 1L, cmt = "central", dvid = NA_integer_, WT = wt
  ) |>
    dplyr::select(id, time, amt, rate, evid, cmt, dvid, WT)
}

# Observation rows. The model declares three endpoints (Cc, uCSchange,
# uDSchange), so rxode2 requires each observation record to identify its
# endpoint. `dvid = 1L` selects Cc; every algebraic observable is still returned
# as a column, so one dvid = 1L grid yields the PK and both PD outputs.
qi_obs_rows <- function(id, times, wt) {
  tidyr::expand_grid(id = id, time = times) |>
    dplyr::transmute(id, time, amt = NA_real_, rate = NA_real_, evid = 0L,
                     cmt = NA_character_, dvid = 1L, WT = wt)
}

qi_solve <- function(events, ..., typical = TRUE) {
  args <- list(mod, as.data.frame(events), returnType = "data.frame",
               # rxode2's ODE->linCmt auto-conversion corrupts the dvid->cmt
               # mapping for multi-endpoint models (known failure pattern 5b).
               useLinCmt = FALSE, ...)
  if (typical) { args$omega <- NA; args$sigma <- NA }
  do.call(rxode2::rxSolve, args)
}
```

# Replicating Table 3: alternative dosing regimens

Qi 2019 simulated three regimens at the cohort median weight of 24.8 kg and
reported, for each, the maximum concentration and the time spent above
`Kuptake` = 0.3 ug/mL -- the in vitro concentration at which receptor-mediated
cellular uptake reaches half its maximum. The comparison of those times is the
paper's central dosing argument.

```{r table3}
wt_sim <- 24.8
regimens <- tibble::tribble(
  ~regimen,        ~dose_mgkg, ~dose_times,
  "4 mg/kg QOW",   4,          c(0, 336),   # two doses per 4-week month
  "4 mg/kg Q4W",   4,          c(0),
  "8 mg/kg Q4W",   8,          c(0)
)

# One "subject" per regimen, solved over a full 4-week (672 h) month. The grid is
# dense around each infusion and coarse in the washout tail.
grid_fine <- c(seq(0, 48, by = 0.02), seq(336, 384, by = 0.02))
grid_tail <- c(seq(48, 336, by = 1), seq(384, 672, by = 1))
obs_times <- sort(unique(c(grid_fine, grid_tail)))

ev_t3 <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  dplyr::bind_rows(
    qi_dose_rows(i, r$dose_mgkg * wt_sim, r$dose_times[[1]], wt_sim),
    qi_obs_rows(i, obs_times, wt_sim)
  ) |>
    dplyr::mutate(regimen = r$regimen, dose_mgkg = r$dose_mgkg)
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# Test the frame directly: wrapping it in unique() would strip the very
# duplicates this asserts the absence of, making the check unfalsifiable.
stopifnot(!anyDuplicated(ev_t3[, c("id", "time", "evid")]))

sim_t3 <- qi_solve(ev_t3, keep = c("regimen", "dose_mgkg"))

kuptake <- 0.3
t3 <- sim_t3 |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    cmax = max(Cc),
    # Total time above Kuptake within the 4-week month, summed over the
    # trapezoids between consecutive observation times.
    t_above = sum(diff(time) * ((head(Cc, -1) > kuptake) + (tail(Cc, -1) > kuptake)) / 2),
    auc_month = max(auc_tau),
    .groups = "drop"
  )

published_t3 <- tibble::tibble(
  regimen  = c("4 mg/kg QOW", "4 mg/kg Q4W", "8 mg/kg Q4W"),
  cmax_pub = c(11.3, 11.4, 21.5),
  t_pub    = c(30.1, 14.9, 18.1)
)

t3_cmp <- t3 |>
  dplyr::left_join(published_t3, by = "regimen") |>
  dplyr::mutate(cmax_diff = 100 * (cmax - cmax_pub) / cmax_pub,
                t_diff    = 100 * (t_above - t_pub) / t_pub)

t3_cmp |>
  dplyr::select(regimen, cmax, cmax_pub, cmax_diff, t_above, t_pub, t_diff) |>
  dplyr::rename("Regimen" = regimen,
                "Cmax simulated (ug/mL)" = cmax, "Cmax published" = cmax_pub,
                "Cmax % diff" = cmax_diff,
                "Time > Kuptake simulated (h)" = t_above,
                "Time > Kuptake published (h)" = t_pub,
                "Time % diff" = t_diff) |>
  knitr::kable(digits = 2, caption = "Replication of Qi 2019 Table 3 at 24.8 kg.")
```

```{r table3-gate}
# Deterministic typical-value solve, so these are tight bounds. Cmax for the
# 8 mg/kg arm is the loosest at about +5%: the model is strictly linear, so its
# Cmax is exactly twice the 4 mg/kg value, whereas the published Table 3 is the
# median of 500 stochastic profiles simulated per regimen and its 21.5 is 1.90x
# rather than 2.00x its own 11.3.
stopifnot(
  max(abs(t3_cmp$cmax_diff)) < 8,
  max(abs(t3_cmp$t_diff))    < 5
)

# The paper's own two summary claims about these numbers.
qow <- t3_cmp$t_above[t3_cmp$regimen == "4 mg/kg QOW"]
stopifnot(
  # "4 mg/kg Q4W ... 50% decrease" relative to 4 mg/kg QOW.
  abs((1 - t3_cmp$t_above[t3_cmp$regimen == "4 mg/kg Q4W"] / qow) - 0.50) < 0.05,
  # "8 mg/kg Q4W ... 40% decrease" relative to 4 mg/kg QOW.
  abs((1 - t3_cmp$t_above[t3_cmp$regimen == "8 mg/kg Q4W"] / qow) - 0.40) < 0.05
)
```

The `Kuptake` column of Table 3 is a **per-month** quantity, not a per-dosing-interval
one, despite the table footnote reading "time (h) > 0.3 ug/mL in one dosing
interval". This is unambiguous from the numbers themselves: vestronidase alfa has
a 3.6 h terminal half-life against a 336 h dosing interval, so there is no
accumulation whatsoever and a 4 mg/kg dose produces an identical profile whether
it is repeated every 2 or every 4 weeks. A per-dosing-interval reading would
therefore force the 4 mg/kg QOW and 4 mg/kg Q4W rows to be equal, whereas the
published values differ exactly two-fold (30.1 vs 14.9 h). Summing over the two
QOW doses in a 4-week month, as done above, reproduces all three published values
to within 3% and recovers both of the paper's stated comparisons (a 50% and a 40%
reduction against QOW). The footnote is a labelling slip; the modelling argument
it supports is sound.

```{r figure-conc, fig.cap = "Simulated typical-value profiles at 24.8 kg over one 4-week month. The dashed line is Kuptake = 0.3 ug/mL. Replicates the simulation underlying Qi 2019 Table 3."}
sim_t3 |>
  dplyr::filter(!is.na(Cc), time <= 384) |>
  ggplot(aes(time, pmax(Cc, 1e-4), colour = regimen)) +
  geom_line() +
  geom_hline(yintercept = kuptake, linetype = "dashed") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Serum vestronidase alfa (ug/mL)", colour = NULL,
       caption = "Kuptake = 0.3 ug/mL (dashed).") +
  theme(legend.position = "bottom")
```

# Exposure over a dosing interval

The exposure metric driving the PD layer is the individual AUC over a dosing
interval. Because elimination is linear, that is exactly `Dose / CL`, and the
`auc_tau` state must reproduce it.

```{r auc-identity}
auc_1dose <- t3_cmp$auc_month[t3_cmp$regimen == "4 mg/kg Q4W"]
cl_at_wt  <- CL * (wt_sim / 20)^ui$theta[["e_wt_cl_q"]]
auc_closed <- 4 * wt_sim / cl_at_wt

cat(sprintf("AUC from the auc_tau state: %.4f ug*h/mL\n", auc_1dose))
cat(sprintf("Dose / CL closed form:      %.4f ug*h/mL\n", auc_closed))

stopifnot(abs(auc_1dose - auc_closed) / auc_closed < 0.001)
```

# Replicating Figure 4: uGAG exposure-response

Figure 4 plots the percent change from baseline in urinary chondroitin sulfate
(uCS) and dermatan sulfate (uDS) against individual AUC, with the population
prediction overlaid. The virtual cohort below spans the published dose levels
(1, 2 and 4 mg/kg QOW) and the published weight range.

Because `auc_tau` accumulates from the start of the solve, the uGAG observables
must be read at the **end of a single dosing interval**, where `auc_tau` holds
the AUC over that interval. That is the discipline the model file documents; a
mid-interval reading reports the effect of the partial AUC accrued so far.

```{r cohort}
# set.seed() seeds R's RNG, not rxode2's; rxode2 partitions its streams per
# solver thread, so the cohort differs between a 16-thread workstation and a
# 2-thread CI runner. Every assertion below is written to hold for any cohort
# the model can produce (see known-vignette-failure-patterns.md pattern 12).
set.seed(20190501)
n_per_arm <- 60L

make_arm <- function(dose_mgkg, id_offset) {
  # Log-normal weights with the published median of 24.8 kg, truncated to the
  # published range of 9.41-104.0 kg (Table 1, overall column).
  wt <- pmin(pmax(rlnorm(n_per_arm, meanlog = log(24.8), sdlog = 0.6), 9.41), 104.0)
  tibble::tibble(id = id_offset + seq_len(n_per_arm), WT = wt,
                 dose_mgkg = dose_mgkg,
                 arm = paste0(dose_mgkg, " mg/kg QOW"))
}

subjects <- dplyr::bind_rows(
  make_arm(1, 0L), make_arm(2, 100L), make_arm(4, 200L)
)

# One dosing interval per subject; observe at the end of it (336 h) plus a dense
# early grid so the PK profile is also usable.
pd_times <- sort(unique(c(seq(0, 48, by = 0.1), seq(48, 336, by = 4), 336)))

ev_pd <- dplyr::bind_rows(lapply(seq_len(nrow(subjects)), function(i) {
  s <- subjects[i, ]
  dplyr::bind_rows(
    qi_dose_rows(s$id, s$dose_mgkg * s$WT, 0, s$WT),
    qi_obs_rows(s$id, pd_times, s$WT)
  ) |>
    dplyr::mutate(arm = s$arm, dose_mgkg = s$dose_mgkg)
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(ev_pd[, c("id", "time", "evid")]))

# Between-subject variability on, residual error off (Figure 4 plots individual
# post hoc exposures against the population prediction, not replicate assays).
sim_pd <- qi_solve(ev_pd, keep = c("arm", "dose_mgkg"), typical = FALSE, sigma = NA)
```

```{r figure-4, fig.cap = "Replicates Figure 4 of Qi 2019: percent change from baseline in uCS (a) and uDS (b) against vestronidase alfa AUC. Points are simulated subjects at the end of one dosing interval; the line is the population prediction."}
endpoints <- sim_pd |>
  dplyr::filter(time == 336) |>
  dplyr::select(id, arm, auc_tau, uCSchange, uDSchange)

curve_df <- tibble::tibble(auc = seq(1, 120, length.out = 400)) |>
  dplyr::mutate(
    uCS = -exp(ui$theta[["limax_ucs"]]) * auc / (auc + exp(ui$theta[["lic50_ucs"]])),
    uDS = -exp(ui$theta[["limax_uds"]]) * auc / (auc + exp(ui$theta[["lic50_uds"]]))
  ) |>
  tidyr::pivot_longer(c(uCS, uDS), names_to = "analyte", values_to = "pred")

endpoints |>
  tidyr::pivot_longer(c(uCSchange, uDSchange), names_to = "analyte",
                      values_to = "change") |>
  dplyr::mutate(analyte = dplyr::recode(analyte, uCSchange = "uCS", uDSchange = "uDS")) |>
  ggplot(aes(auc_tau, change)) +
  geom_point(aes(shape = arm), alpha = 0.6) +
  geom_line(data = curve_df, aes(auc, pred), linewidth = 0.9) +
  facet_wrap(~analyte) +
  coord_cartesian(xlim = c(0, 125), ylim = c(-100, 0)) +
  labs(x = "AUC (ug*h/mL)", y = "% change from baseline", shape = NULL) +
  theme(legend.position = "bottom")
```

```{r figure-4-gate}
# 1. The model's algebraic output must equal the published closed form exactly
#    when the etas are zeroed -- this is a transcription check, not a
#    simulation check, so it is gated at solver tolerance.
ev_typ <- dplyr::bind_rows(
  qi_dose_rows(1L, 4 * wt_sim, 0, wt_sim),
  qi_obs_rows(1L, c(0, 336), wt_sim)
) |>
  dplyr::arrange(time, dplyr::desc(evid))
typ <- qi_solve(ev_typ) |> dplyr::filter(time == 336)

closed_ucs <- -82.0 * typ$auc_tau / (typ$auc_tau + 8.6)
closed_uds <- -76.9 * typ$auc_tau / (typ$auc_tau + 7.3)
stopifnot(
  abs(typ$uCSchange - closed_ucs) < 1e-6,
  abs(typ$uDSchange - closed_uds) < 1e-6
)

# 2. Cohort-level checks, written as robust central statistics rather than
#    extremes (the cohort is not reproducible across solver thread counts).
by_arm <- endpoints |>
  dplyr::group_by(arm) |>
  dplyr::summarise(median_auc = median(auc_tau),
                   median_ucs = median(uCSchange),
                   median_uds = median(uDSchange),
                   .groups = "drop")
knitr::kable(by_arm, digits = 1,
             caption = "Simulated cohort medians at the end of one dosing interval.")

stopifnot(
  # Qi 2019 Results: the uGAG reduction has reached its plateau at 4 mg/kg, and
  # Figure 4 shows the 4 mg/kg cloud sitting around -75% for uCS. Bounds carry
  # deliberate headroom: the arm median realised -68.5% (uCS) / -64.3% (uDS)
  # here, and the cohort is redrawn on every machine because rxode2 partitions
  # its RNG per solver thread. The bounds still go red on a dropped exp(), a
  # sign flip, or a mis-transcribed Imax or IC50, all of which move these by
  # tens of percentage points. Do not tighten them back to the realised values.
  dplyr::between(by_arm$median_ucs[by_arm$arm == "4 mg/kg QOW"], -80, -60),
  dplyr::between(by_arm$median_uds[by_arm$arm == "4 mg/kg QOW"], -78, -57),
  # Exposure must increase monotonically with dose across the three arms; each
  # step is a factor of two in dose against a factor of ~2 in AUC, far outside
  # cohort noise.
  by_arm$median_auc[by_arm$arm == "4 mg/kg QOW"] >
    by_arm$median_auc[by_arm$arm == "1 mg/kg QOW"] * 2
)
```

The paper states that the exposure-response relationship flattens above roughly
40 ug*h/mL and that the typical mean steady-state AUC at 4 mg/kg QOW is
57.9 ug*h/mL. The typical-value AUC at the *median* weight is
`r round(auc_closed, 1)` ug*h/mL; the cohort *mean* is larger because AUC rises
with body weight (as `WT^(1 - 0.587)`) over a right-skewed weight distribution,
and because the mean of a log-normal clearance inflates mean exposure by
`exp(omega^2 / 2)` = `r round(exp(0.170 / 2), 3)`. The simulated cohort mean is
reported below for reference; it is not gated, because it depends on the assumed
weight distribution rather than on the model.

```{r mean-auc}
mean_auc_4 <- mean(endpoints$auc_tau[endpoints$arm == "4 mg/kg QOW"])
cat(sprintf("Simulated cohort mean AUC at 4 mg/kg QOW: %.1f ug*h/mL (paper: 57.9)\n",
            mean_auc_4))
```

# PKNCA validation

Qi 2019 compares its population model against an independent noncompartmental
analysis of the same trials, reported as "data on file": CL 0.07 L/h/kg,
Vss 0.23 L/kg and terminal half-life 2.5 h. To make that comparison meaningful
the NCA below is run on the study's **actual sampling schedule** (Methods, PK
Sampling and Assay) rather than on the dense simulation grid: predose, 60 and
120 min after the start of infusion, the end of infusion, and 5, 15, 30, 60,
120, 180, 360 and 480 min after the end of infusion.

The NCA is run on the typical-value profile rather than the full-IIV cohort. That
is deliberate: a half-life estimate over a cohort with 43% CV on clearance is
both `NA`-poisoned (subjects whose terminal points fall below solver resolution
fail `lambda.z`) and `tmax`-selected, so it would not test the transcription.

```{r pknca}
# The published clinical sampling schedule, in hours. Infusion ends at 4 h.
sample_times <- c(0, 1, 2, 4,
                  4 + c(5, 15, 30, 60, 120, 180, 360, 480) / 60)

ev_nca <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  dplyr::bind_rows(
    qi_dose_rows(i, r$dose_mgkg * wt_sim, 0, wt_sim),  # first dose only
    qi_obs_rows(i, sample_times, wt_sim)
  ) |>
    dplyr::mutate(regimen = r$regimen)
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_nca_raw <- qi_solve(ev_nca, keep = "regimen")

sim_nca <- sim_nca_raw |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

# Guarantee a time-zero row per (id, regimen); pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, regimen) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, regimen, time)

stopifnot(all(sim_nca$Cc >= 0), nrow(sim_nca) > 0)

dose_df <- ev_nca |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id, regimen) |>
  dplyr::summarise(time = 0, amt = sum(amt), .groups = "drop") |>
  dplyr::select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
summary(nca_res)
```

## Comparison against published NCA

```{r nca-compare}
# Qi 2019 reports Cmax per regimen (Table 3) and a single pooled NCA half-life
# of 2.5 h (Results, "data on file"). AUC0-inf is not published per regimen, so
# the Dose/CL identity established above is used as the reference.
published_nca <- tibble::tibble(
  regimen    = c("4 mg/kg QOW", "4 mg/kg Q4W", "8 mg/kg Q4W"),
  cmax       = c(11.3, 11.4, 21.5),
  half.life  = c(2.5, 2.5, 2.5),
  aucinf.obs = c(auc_closed, auc_closed, 2 * auc_closed)
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_nca,
  by        = "regimen",
  units     = c(cmax = "ug/mL", aucinf.obs = "ug*h/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = "Simulated vs. published NCA. * differs from reference by >20%.")
```

```{r nca-gate}
# Locate the percent-difference column by name rather than by position.
diff_col <- grep("diff", names(cmp), ignore.case = TRUE, value = TRUE)[1]
stopifnot(!is.na(diff_col))
pct <- suppressWarnings(as.numeric(gsub("[^0-9.eE+-]", "", as.character(cmp[[diff_col]]))))
param_col <- names(cmp)[1]

# Cmax and AUC are deterministic consequences of the transcribed parameters and
# must agree closely. Half-life is a genuine model-vs-NCA difference (the paper
# itself reports 3.6 h from the model against 2.5 h from the NCA) and is
# reported rather than gated -- see the note below.
# ncaParamLabel() renders half.life as "t\u00bd"; the escape keeps this file ASCII.
is_hl <- grepl("half|t\u00bd", as.character(cmp[[param_col]]), ignore.case = TRUE)
stopifnot(all(abs(pct[!is_hl & !is.na(pct)]) < 20))
```

Cmax reproduces to within 5% on all three regimens. AUC0-inf comes out about 4%
low across the board, which is the sampling schedule rather than the model: the
clinical schedule stops 8 h after the end of infusion, so `aucinf.obs`
extrapolates from a truncated profile. The exact `Dose / CL` identity is checked
separately above on the dense grid.

The half-life row is starred and is deliberately excluded from the gate, but it
deserves a precise statement rather than a hand-wave. There are three numbers,
not two:

| quantity | value |
|---|---|
| Terminal half-life implied by the transcribed parameters | 3.60 h (reproduced exactly, above) |
| NCA on the model, using the published sampling schedule | 3.47 h |
| Qi 2019's own NCA on the observed data ("data on file") | 2.5 h |

Truncating the sampling window at 12 h therefore accounts for only a small part
of the gap (3.60 h to 3.47 h). The remaining difference between 3.47 h and the
paper's 2.5 h is **not** reproducible from the published model, and this vignette
does not claim otherwise. Qi 2019 reports both figures side by side without
reconciling them, and the 2.5 h value also appears in the Discussion cited to the
US prescribing information rather than to this analysis. Plausible contributors
are residual error and per-subject `lambda.z` window selection on real data,
and the eight subjects who contributed only sparse profiles -- none of which a
typical-value solve reproduces. The same pattern appears in the paper's CL and
Vss comparison (model 0.079 L/h/kg and 0.19 L/kg against NCA 0.07 L/h/kg and
0.23 L/kg), where the NCA implies both slower clearance and a larger volume.

# Assumptions and deviations

- **Weight distribution.** Qi 2019 reports only the median (24.8 kg), mean
  (37.7 kg) and range (9.41-104.0 kg) of baseline body weight. The virtual cohort
  uses a log-normal distribution with the published median and `sdlog = 0.6`,
  truncated to the published range. No individual weights are published.
- **Dose-arm sizes.** The published trials are highly unbalanced (3, 8 and 12
  subjects) and 20 of the 23 subjects received 4 mg/kg. The Figure 4 cohort uses
  60 subjects per dose arm so all three arms are visible; it is illustrative of
  the exposure-response shape, not a reproduction of the trial's allocation.
- **PD residual error is not applied in the Figure 4 replication.** Figure 4
  plots individual post hoc exposures against the population prediction, so the
  simulation carries between-subject variability on `Imax` but no residual error.
- **`auc_tau` accumulates from the start of the solve.** The paper's exposure
  metric is the AUC over a dosing interval, so the uGAG observables are only
  interpretable at the end of a single dosing interval. This vignette solves one
  interval per subject for the PD sections. There is no accumulation to worry
  about (3.6 h half-life against a 336 h interval), so the first interval and the
  steady-state interval are identical.
- **`I0` is encoded as an untransformed `fixed(0)`.** The canonical baseline
  parameter is normally carried on the log scale (`lrbase`), which is impossible
  for a value of exactly zero; the quantity is a percent change and is not sign
  constrained. It is kept in the model rather than dropped so that the published
  equation form remains auditable.
- **Table 4's additive residual error is in percentage points**, not ug*h/mL. The
  dependent variable of the exposure-response model is a percent change from
  baseline, and Table 4 gives no units on that row.
- **Table 3's `Kuptake` footnote is inconsistent with its own numbers.** It reads
  "in one dosing interval" but the QOW row is a per-4-week-month total; see the
  Table 3 section above for the argument and the numerical demonstration.
- **Cmax for the 8 mg/kg Q4W regimen differs by about 5%.** The model is strictly
  linear, so its Cmax is exactly twice the 4 mg/kg value; the published Table 3
  is the median of 500 stochastic profiles simulated separately per regimen, and
  its own 21.5 / 11.3 ratio is 1.90 rather than 2.00. This is Monte-Carlo noise
  in the published summary, not a transcription difference.
- **The paper's NCA half-life of 2.5 h is not reproducible from the published
  model.** Running NCA on the model over the study's own sampling schedule gives
  3.47 h, against the 3.60 h implied by the parameters and the 2.5 h Qi 2019
  reports from its own noncompartmental analysis. This is recorded as a known
  deviation and excluded from the NCA gate rather than being tuned away; see the
  PKNCA section for the full argument.
- **Erratum.** The erratum to this article (Clin Pharmacokinet. 2019;58(5):685,
  doi:10.1007/s40262-018-0726-6) corrects a ClinicalTrials.gov identifier in the
  Introduction and changes no parameter value.
- **Supplement.** The electronic supplementary material (DOCX, 207 kB) contains
  assay methodology (Supplementary Methods S1.1-S1.4) and two supporting figures
  (dose proportionality; exposure by age group). It contains no model parameter
  values; all parameters come from Tables 2 and 4 of the main article.
