---
title: "Tacrolimus and mycophenolic acid in kidney transplantation (Quintairos 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tacrolimus and mycophenolic acid in kidney transplantation (Quintairos 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Quintairos 2021 built two population PK models, one for tacrolimus and one for
mycophenolic acid (MPA), in the same cohort of adult de novo kidney transplant
recipients. The two models were fitted separately, so they are packaged as two
model files:

* `Quintairos_2021_tacrolimus`: `r rxode2::rxode(readModelDb("Quintairos_2021_tacrolimus"))$description`
* `Quintairos_2021_mycophenolic_acid`: `r rxode2::rxode(readModelDb("Quintairos_2021_mycophenolic_acid"))$description`
* Citation: `r rxode2::rxode(readModelDb("Quintairos_2021_tacrolimus"))$reference`
* Article: <https://doi.org/10.1371/journal.pone.0245880> (open access). The
  supporting information includes the NONMEM control streams (S1 Appendix) and
  the analysis datasets (S1-S3 Tables).

The paper then used the individual exposures predicted by these models, together
with two urinary biomarkers, as candidate predictors in a logistic regression of
acute rejection.

# Population

The PK sub-study included 58 of the 80 adult de novo kidney recipients enrolled
at three European centres: Charite Berlin and Heidelberg in Germany, and
Fundacio Puigvert in Barcelona, Spain (EudraCT 2013-001817-33). The recipients
had deceased (28) or living (30) donors. Median age was 48 years (IQR 38-58),
and 20 of 58 were female. Median weight was 73 kg (IQR 62.9-86.8) and median
GFR 44 mL/min (IQR 15-55) (Table 1). All patients were Caucasian. Eight
patients (14%) had biopsy-proven acute rejection.

All patients received twice-daily oral tacrolimus (Prograf) and mycophenolate
mofetil (Myfenax), methylprednisolone, and basiliximab induction. Doses were
adjusted by therapeutic drug monitoring. Sampling was intensive at week 1 (0 to
12 h post-dose) and at 0, 1.5, 2 and 4 h post-dose at months 1, 2, 3 and 6.
The tacrolimus model used 1102 whole-blood concentrations and the MPA model
1071 plasma concentrations.

The same information is available programmatically through the `population`
element of each model, for example
`readModelDb("Quintairos_2021_tacrolimus")()$population`.

# Source trace

Parameter values come from Table 4. The model structure comes from the S1
Appendix NONMEM control streams. These are ADVAN4/TRANS4 two-compartment models
with first-order absorption and `ALAG1`, MU-referenced exponential random
effects, and the error models transcribed below. The stream `$THETA` records
are initial estimates (for example, the MPA V2 initial value is 46 L against the
final 106 L), so no value is taken from them.

| Model | Parameter / equation | Value | Source |
|-------|----------------------|-------|--------|
| Tacrolimus | CL/F (`lcl`) | 16.5 L/h/70 kg | Table 4 |
| Tacrolimus | Vc/F (`lvc`) | 311 L/70 kg | Table 4 |
| Tacrolimus | Q/F (`lq`) | 20.5 L/h/70 kg | Table 4 |
| Tacrolimus | Vp/F (`lvp`) | 56300 L/70 kg | Table 4 |
| Tacrolimus | ka (`lka`) | 3.08 1/h | Table 4 |
| Tacrolimus | tlag (`ltlag`) | 0.295 h | Table 4 |
| Tacrolimus | BSV CL, Q, Vc | 57.6%, 68.9%, 55.6% -> omega 0.3318, 0.4747, 0.3091 | Table 4 |
| Tacrolimus | Residual error (`expSd`, log-normal) | 0.366 | Table 4; control stream `Y = LOG(F) + EPS(1)` |
| Tacrolimus | `Cc = central / vc * 1000` (ng/mL from mg and L) | n/a | control stream `S2 = V2/1000` |
| MPA | CL/F (`lcl`) | 11.8 L/h/70 kg | Table 4 |
| MPA | Vc/F (`lvc`) | 106 L/70 kg | Table 4 |
| MPA | Q/F (`lq`) | 37.1 L/h/70 kg | Table 4 |
| MPA | Vp/F (`lvp`) | 800 L/70 kg, fixed | Table 4; control stream `800 FIX` |
| MPA | ka (`lka`) | 1.79 1/h | Table 4 |
| MPA | tlag (`ltlag`) | 0.243 h | Table 4 |
| MPA | BSV CL, Vc, Vp | 34.9%, 133.8%, 164.6% -> omega 0.1218, 1.790, 2.709 | Table 4 |
| MPA | Residual error (`propSd`) | 0.553 | Table 4; control stream `Y = F + F*EPS(1)` |
| MPA | ka multiplied by `WT/70` | exponent 1, fixed | control stream `KA = EXP(MU_5+ETA(5)) * DCOV` |
| Both | Allometry `(WT/70)^0.75` on CL/F and Q/F, `(WT/70)^1` on Vc/F and Vp/F | fixed | Methods, Pharmacokinetic models; MPA control stream `ECOV`, `DCOV` |
| Both | Fixed random effects `omega = 0.01` (tacrolimus ka, Vp, tlag; MPA Q, ka, tlag) | fixed | control stream `$OMEGA 0.01 FIX` |

## Reading of the BSV column

Table 4 prints BSV as a percentage without stating the transformation. The
packaged models read it as `sqrt(omega) x 100`. The MPA control stream supports
this. Its initial `$OMEGA` values (0.123, 1.85, 2.41) are close to the squares
of the printed percentages (0.1218, 1.790, 2.709). They are far from the
log-normal CV reading `log(1 + CV^2)` (0.115, 1.03, 1.31), which would halve
the Vc and Vp variances.

```{r omega-reading}
mpa_bsv <- c(CL = 34.9, VC = 133.8, VP = 164.6) / 100
stream_init <- c(CL = 0.123, VC = 1.85, VP = 2.41)
omega_check <- data.frame(
  parameter = names(mpa_bsv),
  stream_initial = stream_init,
  sqrt_reading = mpa_bsv^2,
  cv_reading = log(1 + mpa_bsv^2)
)
knitr::kable(omega_check, digits = 3, row.names = FALSE)
# The square-root reading is within 13% of every stream value; the CV reading
# is off by more than 40% for Vc and Vp.
stopifnot(
  all(abs(omega_check$sqrt_reading / omega_check$stream_initial - 1) < 0.15),
  all(abs(omega_check$cv_reading / omega_check$stream_initial - 1)[2:3] > 0.4)
)
```

# Virtual cohort

The simulation follows the study's visit schedule: week 1 (day 7) and months
1, 2, 3 and 6 (days 30, 60, 90, 180). Doses are given every 12 h. Table 2
pairs each visit's mean trough with that visit's mean dose (it reports
Ctrough/Dose per visit), so the dose leading up to each visit is the Table 2
mean dose for that visit, and the month-6 dose continues after the last visit.
Table 2 reports doses per day. They are halved to twice-daily amounts here,
matching the per-administration amounts in the S1 and S2 Table datasets (for
example, tacrolimus 7.5 mg per dose at week 1). Mycophenolate mofetil doses
are converted to MPA molar equivalents by the molecular-weight ratio
320.34 / 433.49; 1000 mg of MMF gives the 738.96 mg of MPA carried in the S2
Table dataset.

Body weight is log-normal with a median of 73 kg and a log-scale SD of 0.24,
which reproduces the Table 1 IQR of 62.9-86.8 kg. Each drug has one arm of 200
subjects.

```{r cohort}
set.seed(20210122)
rxode2::rxSetSeed(20210122)
n_sub <- 200
visit_day <- c(7, 30, 60, 90, 180)
visit_lab <- c("Week 1", "Month 1", "Month 2", "Month 3", "Month 6")
end_day <- 184

mmf_to_mpa <- 320.34 / 433.49
daily_dose <- list(
  tacrolimus = c(14.60, 10.63, 7.78, 6.79, 5.29), # Table 2, mg/day
  mpa = c(1875.91, 1655.41, 1552.02, 1402.64, 1238.67) * mmf_to_mpa # Table 2 MMF mg/day -> MPA
)

# Dense sampling over the week-1 and month-6 dosing intervals (for the profile
# figure and NCA), plus a pre-dose trough at every visit.
profile_hours <- c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 2, 2.5, 3, 4, 5, 6, 8, 10, 12)
obs_times <- sort(unique(c(
  visit_day * 24 - 0.05,
  visit_day[1] * 24 + profile_hours,
  visit_day[5] * 24 + profile_hours
)))

make_cohort <- function(daily, drug, id_offset) {
  wt <- exp(rnorm(n_sub, log(73), 0.24))
  # Dose k is given from the previous visit up to visit k, so the trough
  # sampled just before visit k reflects dose k; the month-6 dose continues
  # over the month-6 profile interval.
  starts <- c(0, visit_day) * 24
  ends <- c(visit_day, end_day) * 24
  daily <- c(daily, daily[5])
  doses <- data.frame(
    time = starts,
    amt = daily / 2,
    ii = 12,
    addl = (ends - starts) / 12 - 1
  )
  bind_rows(lapply(seq_len(n_sub), function(i) {
    bind_rows(
      doses |> mutate(evid = 1L, cmt = "depot"),
      data.frame(time = obs_times, amt = 0, ii = 0, addl = 0, evid = 0L, cmt = "central")
    ) |>
      mutate(id = id_offset + i, WT = wt[i], drug = drug)
  })) |>
    arrange(id, time, desc(evid))
}

events_tac <- make_cohort(daily_dose$tacrolimus, "Tacrolimus", 0L)
events_mpa <- make_cohort(daily_dose$mpa, "MPA", n_sub)
stopifnot(
  length(unique(events_tac$id)) == n_sub,
  length(unique(events_mpa$id)) == n_sub,
  !any(unique(events_tac$id) %in% unique(events_mpa$id))
)
```

# Simulation

```{r simulate}
mod_tac <- readModelDb("Quintairos_2021_tacrolimus")
mod_mpa <- readModelDb("Quintairos_2021_mycophenolic_acid")

sim_tac <- rxSolve(mod_tac, events_tac, keep = c("drug", "WT"), returnType = "data.frame")
sim_mpa <- rxSolve(mod_mpa, events_mpa, keep = c("drug", "WT"), returnType = "data.frame")

typ_tac <- rxSolve(zeroRe(mod_tac), events_tac |> filter(id == 1) |> mutate(WT = 70),
  returnType = "data.frame"
)
typ_mpa <- rxSolve(zeroRe(mod_mpa), events_mpa |> filter(id == n_sub + 1) |> mutate(WT = 70),
  returnType = "data.frame"
)
```

## Trough concentrations by visit (Table 2)

Table 2 gives the mean observed trough at each visit. The table below compares
it with the simulated mean pre-dose concentration under the Table 2 dose
schedule. The comparison uses the individual prediction `Cc`, without residual
error.

```{r troughs}
trough_times <- visit_day * 24 - 0.05
obs_trough <- data.frame(
  drug = rep(c("Tacrolimus", "MPA"), each = 5),
  visit = rep(visit_lab, 2),
  observed = c(8.85, 11.15, 8.92, 9.11, 8.35, 2.37, 2.93, 2.70, 2.90, 2.37) # Table 2
)
sim_trough <- bind_rows(sim_tac, sim_mpa) |>
  filter(time %in% trough_times) |>
  mutate(visit = visit_lab[match(time, trough_times)]) |>
  group_by(drug, visit) |>
  summarise(simulated_mean = mean(Cc), simulated_median = median(Cc), .groups = "drop")
trough_cmp <- obs_trough |>
  left_join(sim_trough, by = c("drug", "visit")) |>
  mutate(pct_diff = 100 * (simulated_mean - observed) / observed)

trough_cmp |>
  dplyr::rename(
    Drug = drug, Visit = visit, `Observed mean (Table 2)` = observed,
    `Simulated mean` = simulated_mean, `Simulated median` = simulated_median,
    `Difference (%)` = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Tacrolimus in ng/mL, MPA in mg/L.")
```

The models have no time-varying clearance. The paper reports that
dose-normalised troughs rose steadily over the 6 months, and Table 2 shows
tacrolimus C/D rising from 0.61 at week 1 to 1.63 at month 6. The per-visit
comparison is therefore expected to drift by some tens of percent. The gate
below is on the centre of the comparison across visits. A mis-transcribed
clearance, volume or unit conversion would move every visit together by far
more.

```{r trough-gate}
chk <- trough_cmp |>
  group_by(drug) |>
  summarise(mean_pct = mean(pct_diff), max_abs_pct = max(abs(pct_diff)))
knitr::kable(chk, digits = 1)
# Realised centres: tacrolimus -10.9% / -13.8% and MPA +5.8% / +7.8% on a
# 16-thread / 1-2-thread solve (largest single visit 27-29%). Moving each dose
# to the period AFTER its visit instead gives about +9% and +15%, so the
# dose-schedule assumption alone moves the centre by about 20 points. 25%
# admits that plus the cohort noise. A mis-transcribed CL/F, a volume or the
# ng/mL scaling moves every visit by 50% or more and still fails here.
stopifnot(
  all(abs(chk$mean_pct) < 25),
  all(chk$max_abs_pct < 45)
)
```

## Week-1 and month-6 dosing-interval profiles

The figure shows the simulated concentrations over one 12-h dosing interval at
week 1 and at month 6: the median, the 5th-95th percentile band, and the
typical 70 kg patient. The shape (tmax near 1-1.5 h after the lag time) is the
one the paper's prediction-corrected VPCs show (Figures 1 and 2).

```{r profiles, fig.width = 7, fig.height = 5}
prof <- bind_rows(sim_tac, sim_mpa) |>
  mutate(
    visit = case_when(
      time >= visit_day[1] * 24 & time <= visit_day[1] * 24 + 12 ~ "Week 1",
      time >= visit_day[5] * 24 & time <= visit_day[5] * 24 + 12 ~ "Month 6"
    ),
    tad = time - ifelse(visit == "Week 1", visit_day[1], visit_day[5]) * 24
  ) |>
  filter(!is.na(visit)) |>
  group_by(drug, visit, tad) |>
  summarise(
    med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
    .groups = "drop"
  )
typ_prof <- bind_rows(
  typ_tac |> mutate(drug = "Tacrolimus"),
  typ_mpa |> mutate(drug = "MPA")
) |>
  mutate(
    visit = case_when(
      time >= visit_day[1] * 24 & time <= visit_day[1] * 24 + 12 ~ "Week 1",
      time >= visit_day[5] * 24 & time <= visit_day[5] * 24 + 12 ~ "Month 6"
    ),
    tad = time - ifelse(visit == "Week 1", visit_day[1], visit_day[5]) * 24
  ) |>
  filter(!is.na(visit))

ggplot(prof, aes(tad, med, colour = visit, fill = visit)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_line(data = typ_prof, aes(tad, Cc, colour = visit), linetype = "dashed", inherit.aes = FALSE) +
  facet_wrap(~drug, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Concentration (tacrolimus ng/mL; MPA mg/L)",
    colour = NULL, fill = NULL,
    caption = "Solid: simulated median; band: 5th-95th percentile; dashed: typical 70 kg patient."
  ) +
  theme_bw()
```

# NCA of the week-1 and month-6 dosing intervals

The paper reports no NCA. PKNCA is run over the week-1 and month-6 dosing
intervals of the simulated cohort, one drug at a time because the concentration
units differ. The results summarise the exposures the models imply.

```{r nca}
run_nca <- function(sim, events, conc_unit) {
  conc <- sim |>
    filter(!is.na(Cc)) |>
    mutate(treatment = drug) |>
    select(id, time, Cc, treatment)
  dose <- events |>
    filter(evid == 1) |>
    mutate(treatment = drug) |>
    select(id, time, amt, treatment)
  starts <- c(visit_day[1], visit_day[5]) * 24
  # Dose records carry ADDL; add explicit dose rows at the start of each NCA
  # interval so PKNCA sees the dose that opens it.
  dose_int <- dose |>
    group_by(id, treatment) |>
    reframe(time = starts, amt = amt[findInterval(starts, time)])
  intervals <- data.frame(
    start = starts, end = starts + 12,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
  )
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id, concu = conc_unit, timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_int, amt ~ time | treatment + id, doseu = "mg")
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
nca_tac <- run_nca(sim_tac, events_tac, "ng/mL")
nca_mpa <- run_nca(sim_mpa, events_mpa, "mg/L")
summary(nca_tac)
summary(nca_mpa)
```

As a consistency check, the median MPA AUC over the week-1 dosing interval is
compared with the MPA target range of 30-60 mg*h/L that is used for
therapeutic drug monitoring. The simulated median should fall near that range,
because the doses were adjusted to reach it.

```{r nca-check}
auc_mpa <- as.data.frame(nca_mpa) |>
  filter(PPTESTCD == "auclast", start == visit_day[1] * 24)
med_auc_mpa <- median(auc_mpa$PPORRES)
med_auc_mpa
stopifnot(med_auc_mpa > 20, med_auc_mpa < 80)
```

# Assumptions and deviations

* **Weight scaling on tacrolimus.** The S1 Appendix tacrolimus control stream
  has no weight scaling. The article states that all disposition parameters
  were scaled a priori with fixed exponents of 0.75 (flows) and 1 (volumes).
  Table 4 gives the tacrolimus estimates per 70 kg, and the Discussion says the
  fixed-exponent scaling improved the model predictions. The packaged
  tacrolimus model therefore includes the scaling. The deposited dataset
  supports this choice. With every Table 4 value fixed and only the individual
  random effects estimated (FOCEi, diagonal omega), the objective function was
  7.4 points lower with the scaling than without it. The stream's `$THETA`
  records are initial estimates, so it is not the run that produced Table 4.
* **Weight scaling on MPA ka.** The MPA control stream multiplies ka by
  `WT/70`. The article describes scaling of the disposition parameters only,
  and Table 4 prints ka in 1/h. The packaged model keeps the executed form.
  The same fixed-estimate check on the S2 Table dataset favoured ka *without*
  weight scaling by 5.7 objective-function points. Both forms give the same
  ka at 70 kg and differ by a factor of about 0.7 to 1.5 across the 49-108 kg
  cohort weight range.
* **Omega block off-diagonals.** Both control streams estimate a 3 x 3 omega
  block: tacrolimus CL, Q and Vc; MPA CL, Vc and Vp. The paper reports only the
  variances, so the covariances are set to zero. Simulated marginal variability
  matches the paper. Joint behaviour, such as a CL-Vc correlation, does not.
* **Fixed random effects.** Both control streams carry `$OMEGA 0.01 FIX` on
  the parameters without estimated BSV. Table 4 lists those parameters as
  "not estimated". They are kept as `fixed(0.01)`, which is part of the
  executed model and adds about 10% variability to those parameters.
* **Tacrolimus residual error** is additive on the log scale in the control
  stream and printed as "Proportional 36.6%" in Table 4. It is encoded as
  log-normal (`lnorm`) with SD 0.366.
* **MMF dosing** is entered as MPA molar equivalents, as in the paper. A user
  simulating an MMF dose must multiply it by 320.34 / 433.49 = 0.739.
* **Virtual dose schedule.** Real doses were individualised by therapeutic
  drug monitoring. The simulation gives the Table 2 mean dose for each visit
  over the period leading up to that visit, at the nominal visit days.
