---
title: "Pyrazinamide and ethambutol in HIV-associated critical illness (Rao 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pyrazinamide and ethambutol in HIV-associated critical illness (Rao 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Rao 2021 fitted four separate population PK models, one per first-line
anti-tuberculosis drug, to steady-state serum concentrations from Ugandan adults
living with HIV who were hospitalised with sepsis. Two of the four are packaged
here:

* `Rao_2021_pyrazinamide`:
  `r rxode2::rxode(readModelDb("Rao_2021_pyrazinamide"))$description`
* `Rao_2021_ethambutol`:
  `r rxode2::rxode(readModelDb("Rao_2021_ethambutol"))$description`
* Citation: `r rxode2::rxode(readModelDb("Rao_2021_pyrazinamide"))$reference`
* Article (open access): <https://doi.org/10.3390/antibiotics10060739>

The rifampin and isoniazid models from the same paper are **not** packaged. The
section [Rifampin and isoniazid](#rifampin-and-isoniazid) explains why.

# Population

Rao 2021 (Methods 4.1, Results, Table 1) enrolled 81 consecutive people living
with HIV at Mbarara Regional Referral Hospital, Uganda. Each met the sepsis
definition (two or more SIRS criteria plus suspected infection), with or without
suspected tuberculous meningitis, and was starting first-line therapy for
presumed drug-susceptible tuberculosis. Of the 81, 36.3% were women. The median
(IQR) age was 36 (30-43.5) years, weight 52.5 (45.3-58) kg and mid-upper arm
circumference 22 (20.4-24) cm. The median CD4 count was 169 cells/mm3, 76% were
on antiretroviral therapy and 33% had concurrent meningitis. By the two-week PK
visit 18 had died and 14 had been lost to follow-up, withdrawn or had incomplete
sampling, which left **49 patients in the PK analysis** (26 with microbiologically
confirmed TB and 23 with clinical TB; Table 2). Table 1 describes the 81 enrolled,
not the 49 analysed.

Drugs were given once daily as weight-based fixed-dose combinations. The median
(IQR) dose was 25.4 (23-28) mg/kg for pyrazinamide and 18.4 (16.5-19.6) mg/kg for
ethambutol. Serum was sampled at 1, 2, 4 and 6 h post-dose after two weeks of
treatment, and each model was fitted to 173 concentrations (Results 2.3, 2.4).

The same information is available programmatically:

```{r population}
str(readModelDb("Rao_2021_pyrazinamide")()$population, give.attr = FALSE)
```

# Source trace

Both packaged models are one-compartment models with first-order absorption, an
absorption lag time, linear elimination and an additive residual error
(Results 2.3 and 2.4). Neither retains a covariate. All parameters are apparent
oral values.

| Equation / parameter | Pyrazinamide | Ethambutol | Source location |
|----------------------|--------------|------------|-----------------|
| `lka` (ka, 1/h) | log(0.08) | log(0.15) | Table 3, 'Estimate (%CV)' |
| `lvc` (V/F, L) | log(1.5) | log(75.17) | Table 3 |
| `lcl` (CL/F, L/h) | log(2.6) | log(51.6) | Table 3 |
| `ltlag` (Tlag, h) | log(0.2) | log(0.4) | Table 3 |
| `etalka` (variance) | 0.1 | 0.6 | Table 3, 'IIV (Shrinkage)'; pyrazinamide printed '0.1-0.11' |
| `etalvc` (variance) | 0.9 | 0.4 | Table 3 |
| `etalcl` (variance) | 0.4 | 0.3 | Table 3 |
| `etaltlag` (variance) | 1.3 | 0.6 | Table 3 |
| `addSd` (mg/L) | fixed(0) | fixed(0) | Additive error named in Results 2.3 / 2.4; magnitude not reported |
| `d/dt(depot)`, `d/dt(central)`, `alag(depot)` | -- | -- | Results 2.3 / 2.4: 'one-compartment distribution model with a lag-time for oral absorption' |
| exponential IIV, `exp(theta + eta)` | -- | -- | Methods 4.4: 'Interindividual variability was included using exponential function' |

# Rifampin and isoniazid

Table 3 of Rao 2021 also lists final estimates for rifampin (one-compartment,
lag, additive error) and isoniazid (two-compartment, lag, proportional error).
Neither can be packaged faithfully from the published text:

* **Rifampin.** Results 2.1 states that 'weight was a significant covariate for
  clearance', but the weight coefficient is not reported. The printed clearance,
  0.1 L/h, is therefore the intercept of an unstated covariate equation, not a
  typical value. Read on its own, it gives a steady-state AUC0-24 of 6000 mg*h/L
  for 600 mg, where the paper's observed median is 21.7 mg*h/L. The Methods say
  only that covariates enter 'using an exponential function in multiplicative
  fashion', and the centring is not given.
* **Isoniazid.** Results 2.2 retains a mid-upper arm circumference (MUAC) effect
  on the central volume, which is printed as 2.9 L, but the coefficient is not
  reported. Separately, the printed clearance of 9.2 L/h cannot reproduce the
  paper's own Table 4 simulation. AUC0-24 after a fully absorbed dose is
  dose / CL whatever the volume, so it would be about 32.6 mg*h/L at 300 mg,
  against the Table 4 median of 10.2 mg*h/L.

The paper has no supplement, and its data are available only on request from the
corresponding author. These are gaps in what was reported, not missing files.
Both models can be added if the coefficients become available.

# Virtual cohort and simulation

Table 4 of Rao 2021 summarises Cmax and AUC0-24 simulated from the final models
at five dose levels per drug. Solving the typical-value models (below) shows that
those simulations are **single-dose** profiles. A single pyrazinamide dose
reproduces the Table 4 medians to within about 7%. Steady state overshoots them by
more than 20%, because pyrazinamide absorption is slow (ka = 0.08 1/h) and
accumulates.

Neither model has covariates, so the virtual cohort is 100 subjects per dose
level. Each gets one oral dose into `depot` with observations on `central` every
0.25 h to 24 h.

```{r cohort}
doses <- list(
  pyrazinamide = c(1000, 1500, 2000, 2500, 3000),
  ethambutol = c(800, 1200, 1600, 2000, 2400)
)
n_per_arm <- 100L
obs_times <- seq(0, 24, by = 0.25)

make_arm <- function(drug, dose, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  dose_rows <- tibble(
    id = ids, time = 0, amt = dose, evid = 1L, cmt = "depot"
  )
  obs_rows <- tidyr::expand_grid(id = ids, time = obs_times) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
  bind_rows(dose_rows, obs_rows) |>
    mutate(drug = drug, treatment = paste(dose, "mg")) |>
    arrange(id, time, desc(evid))
}

events <- list()
for (drug in names(doses)) {
  arms <- lapply(seq_along(doses[[drug]]), function(i) {
    make_arm(drug, doses[[drug]][i], id_offset = (i - 1L) * n_per_arm)
  })
  events[[drug]] <- bind_rows(arms)
  stopifnot(!anyDuplicated(unique(events[[drug]][, c("id", "time", "evid")])))
}
```

```{r simulate}
rxode2::rxSetSeed(20210618)
mods <- list(
  pyrazinamide = readModelDb("Rao_2021_pyrazinamide"),
  ethambutol = readModelDb("Rao_2021_ethambutol")
)
sim <- bind_rows(lapply(names(mods), function(drug) {
  as.data.frame(rxode2::rxSolve(
    mods[[drug]],
    events = events[[drug]],
    keep = c("drug", "treatment")
  ))
}))
```

# Typical-value replication of Table 4

The Table 4 medians summarise individual profiles, which the paper simulated from
the 49 patients' own parameter estimates. The typical-value solve is the
deterministic anchor for those medians.

```{r typical}
published_t4 <- tibble::tribble(
  ~drug, ~dose, ~cmax, ~auc24, ~cmax_target, ~auc_target, ~pct_cmax, ~pct_auc,
  "pyrazinamide", 1000, 25.7, 303.8, 20, 363, 69.4, 42.9,
  "pyrazinamide", 1500, 38.0, 466.0, 20, 363, 89.8, 71.4,
  "pyrazinamide", 2000, 50.4, 624.7, 20, 363, 91.8, 85.1,
  "pyrazinamide", 2500, 63.1, 778.5, 20, 363, 96.0, 89.6,
  "pyrazinamide", 3000, 75.8, 933.0, 20, 363, 98.0, 91.7,
  "ethambutol", 800, 1.6, 13.0, 2, NA, 32.6, NA,
  "ethambutol", 1200, 2.4, 19.5, 2, NA, 63.2, NA,
  "ethambutol", 1600, 3.2, 26.0, 2, NA, 89.8, NA,
  "ethambutol", 2000, 3.9, 32.5, 2, NA, 93.8, NA,
  "ethambutol", 2400, 4.7, 39.0, 2, NA, 93.8, NA
)

trap_auc <- function(time, conc) {
  sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2)
}

typical_profile <- function(drug, dose, steady_state = FALSE) {
  ev <- if (steady_state) {
    rxode2::et(amt = dose, cmt = "depot", ii = 24, addl = 29) |>
      rxode2::et(seq(696, 720, by = 0.05), cmt = "central")
  } else {
    rxode2::et(amt = dose, cmt = "depot") |>
      rxode2::et(seq(0, 24, by = 0.05), cmt = "central")
  }
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mods[[drug]]), events = ev))
  tibble(cmax = max(s$Cc), auc24 = trap_auc(s$time, s$Cc))
}

typical <- published_t4 |>
  rowwise() |>
  mutate(
    sd = list(typical_profile(drug, dose)),
    ss = list(typical_profile(drug, dose, steady_state = TRUE))
  ) |>
  ungroup() |>
  mutate(
    cmax_sd = sapply(sd, `[[`, "cmax"),
    auc_sd = sapply(sd, `[[`, "auc24"),
    cmax_ss = sapply(ss, `[[`, "cmax"),
    auc_ss = sapply(ss, `[[`, "auc24"),
    pct_cmax_sd = 100 * (cmax_sd - cmax) / cmax,
    pct_auc_sd = 100 * (auc_sd - auc24) / auc24,
    pct_auc_ss = 100 * (auc_ss - auc24) / auc24
  )

typical |>
  select(drug, dose, cmax, cmax_sd, auc24, auc_sd, auc_ss, pct_cmax_sd, pct_auc_sd, pct_auc_ss) |>
  dplyr::rename(
    "Drug" = drug,
    "Dose (mg)" = dose,
    "Table 4 Cmax" = cmax,
    "Typical single-dose Cmax" = cmax_sd,
    "Table 4 AUC0-24" = auc24,
    "Typical single-dose AUC0-24" = auc_sd,
    "Typical steady-state AUC0-24" = auc_ss,
    "Cmax % diff (single dose)" = pct_cmax_sd,
    "AUC % diff (single dose)" = pct_auc_sd,
    "AUC % diff (steady state)" = pct_auc_ss
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Typical-value Cmax (mg/L) and AUC0-24 (mg*h/L) against the Table 4 simulated medians of Rao 2021."
  )
```

```{r typical-gate}
pza <- typical[typical$drug == "pyrazinamide", ]
emb <- typical[typical$drug == "ethambutol", ]
stopifnot(
  # Pyrazinamide: single dose reproduces both Table 4 medians at every dose
  # (realised 3-7%). A mis-transcribed ka, V or CL moves these by tens of
  # percent or more.
  all(abs(pza$pct_cmax_sd) < 10),
  all(abs(pza$pct_auc_sd) < 10),
  # ...and steady state does not (realised +23% to +27%), which is what
  # identifies Table 4 as single-dose simulations.
  all(pza$pct_auc_ss > 15),
  # Ethambutol: Cmax within 10% (realised -3% to -5%). The AUC0-24 medians
  # run about 15% below the typical value at every dose; see Assumptions.
  all(abs(emb$pct_cmax_sd) < 10),
  all(abs(emb$pct_auc_sd) < 20)
)
```

These are deterministic solves with no random draw, so the bounds above are tight
on purpose.

# Replicate Figures 6 and 8

```{r nca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(group = paste(drug, treatment)) |>
  dplyr::select(id, time, Cc, drug, treatment, group)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |>
    dplyr::distinct(id, drug, treatment, group) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, group, time, .keep_all = TRUE) |>
  dplyr::arrange(group, id, time)

dose_df <- bind_rows(events) |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(group = paste(drug, treatment)) |>
  dplyr::select(id, time, amt, drug, treatment, group)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | group + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | group + id,
  doseu = "mg"
)
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

per_subject <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::select(group, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(
    dplyr::distinct(dose_df, group, drug, treatment, amt),
    by = "group"
  )
```

```{r figures-6-8, fig.width = 7, fig.height = 6}
# Replicates Figures 6 (pyrazinamide) and 8 (ethambutol) of Rao 2021:
# simulated Cmax and AUC0-24 by single oral dose, with the Table 4 targets.
targets <- tibble::tribble(
  ~drug, ~metric, ~target,
  "pyrazinamide", "Cmax (mg/L)", 20,
  "pyrazinamide", "AUC0-24 (mg*h/L)", 363,
  "ethambutol", "Cmax (mg/L)", 2
)

per_subject |>
  tidyr::pivot_longer(c(cmax, auclast), names_to = "metric", values_to = "value") |>
  dplyr::mutate(
    metric = ifelse(metric == "cmax", "Cmax (mg/L)", "AUC0-24 (mg*h/L)"),
    dose = factor(amt)
  ) |>
  ggplot(aes(dose, value)) +
  geom_boxplot(outlier.size = 0.6) +
  geom_hline(data = targets, aes(yintercept = target), linetype = "dashed") +
  facet_wrap(drug ~ metric, scales = "free", ncol = 2) +
  labs(
    x = "Single oral dose (mg)", y = NULL,
    caption = "Replicates Figures 6 and 8 of Rao 2021; dashed lines are the Table 4 targets."
  ) +
  theme_bw()
```

# PKNCA comparison against Table 4

```{r nca-compare}
reference <- published_t4 |>
  dplyr::mutate(group = paste(drug, paste(dose, "mg"))) |>
  dplyr::select(group, cmax, auclast = auc24)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by = "group",
  params = c("cmax", "auclast"),
  units = c(cmax = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Simulated (single dose, 100 subjects per dose) vs. Rao 2021 Table 4 medians. * differs from the reference by >20%."
)
```

```{r nca-gate}
chk <- per_subject |>
  dplyr::group_by(drug, amt) |>
  dplyr::summarise(cmax = median(cmax), auclast = median(auclast), .groups = "drop") |>
  dplyr::inner_join(published_t4, by = c("drug", "amt" = "dose")) |>
  dplyr::mutate(
    pct_cmax = 100 * (cmax.x - cmax.y) / cmax.y,
    pct_auc = 100 * (auclast - auc24) / auc24
  )
stopifnot(
  # Centre, not extremes: the cohort median across the five doses. A wrong
  # dose, unit or clearance moves every dose level together.
  abs(median(chk$pct_cmax[chk$drug == "pyrazinamide"])) < 15,
  abs(median(chk$pct_auc[chk$drug == "pyrazinamide"])) < 15,
  # Ethambutol Cmax: the expected offset is about -9% (a 2000-subject run
  # gave a median of 1.45 against 1.6 mg/L at 800 mg), so the bound leaves
  # room for cohort noise while still failing on a mis-transcribed V or ka.
  abs(median(chk$pct_cmax[chk$drug == "ethambutol"])) < 20,
  abs(median(chk$pct_auc[chk$drug == "ethambutol"])) < 25
)
```

For pyrazinamide, the simulated medians fall within about 15% of Table 4 at every
dose. For ethambutol, the AUC0-24 medians agree within about 10%. The Cmax medians
run 5-25% low, and the gap is widest at 800 mg, where Table 4 rounds to one
decimal place. The ka variance of 0.6 is wide, and it pulls the median Cmax below
the typical-value Cmax: a subject with slow absorption has a lower peak, while a
subject with fast absorption gains comparatively little.

## Target attainment

```{r attainment}
per_subject |>
  dplyr::inner_join(published_t4, by = c("drug", "amt" = "dose")) |>
  dplyr::group_by(drug, amt) |>
  dplyr::summarise(
    sim_cmax = 100 * mean(cmax.x >= cmax_target),
    pub_cmax = dplyr::first(pct_cmax),
    sim_auc = if (all(is.na(auc_target))) NA_real_ else 100 * mean(auclast >= auc_target),
    pub_auc = dplyr::first(pct_auc),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Drug" = drug,
    "Dose (mg)" = amt,
    "Simulated % Cmax >= target" = sim_cmax,
    "Table 4 % Cmax >= target" = pub_cmax,
    "Simulated % AUC0-24 >= target" = sim_auc,
    "Table 4 % AUC0-24 >= target" = pub_auc
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Percentage attaining the Table 4 targets (pyrazinamide Cmax 20 mg/L and AUC0-24 363 mg*h/L; ethambutol Cmax 2 mg/L)."
  )
```

This table is descriptive. The Table 4 percentages come from the 49 patients' own
(shrunken) parameter estimates, so their spread is narrower than a simulation
from the full between-subject variances. For pyrazinamide the simulated
attainment is close to Table 4 at every dose. For ethambutol it is lower, because
the simulated median Cmax sits below Table 4 (see above) and the wider spread
puts more subjects under the 2 mg/L target.

# Scale of the IIV column

Table 3 heads the variability column 'IIV (Shrinkage)' without saying whether the
number is a variance or a standard deviation. Phoenix NLME reports the Omega
diagonal as variances, and Table 4 supports that reading. The paper's simulations
use individual estimates, which shrinkage pulls toward the typical value, so the
spread in Table 4 can be *no wider* than a simulation from the true between-subject
variances. Its AUC0-24 interquartile ratio (Q3/Q1) is 2.01 for pyrazinamide at
1000 mg and 1.86 for ethambutol at 800 mg. Under the standard-deviation reading, a
full-variability simulation gives a spread narrower than the shrunken one Table 4
shows, which is impossible. Under the variance reading it gives a wider one, as it
should.

```{r iiv-scale}
iqr_ratio <- function(x) unname(quantile(x, 0.75) / quantile(x, 0.25))
per_subject |>
  dplyr::filter(amt %in% c(1000, 800)) |>
  dplyr::group_by(drug, amt) |>
  dplyr::summarise(sim_iqr_ratio = iqr_ratio(auclast), .groups = "drop") |>
  dplyr::mutate(table4_iqr_ratio = ifelse(drug == "pyrazinamide", 466.1 / 231.6, 18.6 / 10)) |>
  dplyr::rename(
    "Drug" = drug,
    "Dose (mg)" = amt,
    "Simulated AUC0-24 Q3/Q1 (variance reading)" = sim_iqr_ratio,
    "Table 4 AUC0-24 Q3/Q1" = table4_iqr_ratio
  ) |>
  knitr::kable(digits = 2)
```

When the maintainers repeated this with a large cohort under the standard-deviation
reading, it gave AUC0-24 Q3/Q1 ratios of about 1.7 (pyrazinamide) and 1.5
(ethambutol). Both are below the Table 4 values, so that reading was rejected.

# Steady state against the observed NCA

Results 2.3 and 2.4 report observed non-compartmental values at the two-week
visit. For pyrazinamide these are a median Cmax of 34 mg/L and AUC0-24 of 351
mg*h/L at 25.4 mg/kg; for ethambutol, a Cmax of 1.8 mg/L and AUC0-24 of 14.3
mg*h/L at 18.4 mg/kg. For a 52.5 kg patient those doses are about 1334 mg and
966 mg.

```{r steady-state}
bind_rows(
  typical_profile("pyrazinamide", 25.4 * 52.5, steady_state = TRUE) |>
    mutate(drug = "pyrazinamide", obs_cmax = 34, obs_auc = 351),
  typical_profile("ethambutol", 18.4 * 52.5, steady_state = TRUE) |>
    mutate(drug = "ethambutol", obs_cmax = 1.8, obs_auc = 14.3)
) |>
  dplyr::select(drug, obs_cmax, cmax, obs_auc, auc24) |>
  dplyr::rename(
    "Drug" = drug,
    "Observed NCA Cmax" = obs_cmax,
    "Typical steady-state Cmax" = cmax,
    "Observed NCA AUC0-24" = obs_auc,
    "Typical steady-state AUC0-24" = auc24
  ) |>
  knitr::kable(digits = 1, caption = "Typical steady-state values at the median mg/kg dose for a 52.5 kg patient.")
```

This comparison is not a validation gate. The observed AUC0-24 values were
extrapolated to 24 h from four samples between 1 and 6 h post-dose, and only in
patients whose 6-h sample lay in the elimination phase (Methods 4.4). For
pyrazinamide, the terminal phase is absorption-limited (half-life about
ln 2 / 0.08 = 8.7 h), and a 6-h window captures little of it. The model's
steady-state pyrazinamide AUC0-24 is therefore well above the observed NCA value.
For ethambutol the steady-state Cmax agrees with the observed median, and the
AUC0-24 is about 30% higher.
The paper's own Table 4 simulations also sit below steady state, because they are
single-dose.

# Assumptions and deviations

* **Rifampin and isoniazid are not packaged.** The weight coefficient on rifampin
  clearance and the MUAC coefficient on isoniazid central volume are not reported,
  and the printed isoniazid clearance does not reproduce Table 4. See
  [Rifampin and isoniazid](#rifampin-and-isoniazid).
* **Residual error magnitude.** Results 2.3 and 2.4 name an additive residual
  error for both drugs but give no value, and the paper has no supplement. `addSd`
  is `fixed(0)`, so simulations carry between-subject variability only.
* **IIV scale.** The Table 3 'IIV (Shrinkage)' entries are read as Phoenix NLME
  Omega variances of exponential random effects, with the parenthesised number
  taken as the shrinkage fraction. See [Scale of the IIV column](#scale-of-the-iiv-column).
* **Pyrazinamide ka IIV.** Table 3 prints '0.1-0.11' in the IIV (Shrinkage)
  column, while every other row reads 'value (shrinkage)'. It is read as a
  variance of 0.1 with shrinkage 0.11.
* **Pyrazinamide V/F of 1.5 L.** This is far below the roughly 40-50 L usually
  reported for pyrazinamide. With ka = 0.08 1/h, well below CL/V = 1.7 1/h, the
  fitted model is flip-flop: its terminal slope is set by ka, and it reproduces
  the paper's own Table 4 medians. It is kept as published and not reinterpreted.
* **Ethambutol AUC0-24.** The typical-value AUC0-24 is about 15% above the Table 4
  median at every dose, while Cmax agrees within 5%. The Table 4 medians summarise
  49 individual clearance estimates with 30% shrinkage, and their median need not
  equal the typical value. The parameters were not adjusted.
* **Dosing.** Table 4 is reproduced with single oral doses, the reading
  the typical-value comparison supports. The paper does not state it.
* **Covariates screened but not retained.** Age, sex, weight, MUAC and TB
  diagnostic group (Methods 4.4) are recorded in `covariatesDataExcluded`.
* No erratum or correction notice for this article was found on the publisher
  page or Europe PMC as of 2026-09-28.
