---
title: "Trametinib (Ravix 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Trametinib (Ravix 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Ravix_2024_trametinib"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.3390/cancers16122193>

Trametinib is an allosteric MEK1/MEK2 inhibitor used for BRAF V600-mutant solid
tumours, given orally at 0.5-2 mg once daily. Ravix and colleagues fit a
population PK model to real-life therapeutic-drug-monitoring data from the OpTAT
study, then used it to ask two questions: whether the licensed 2 mg daily dose
actually reaches the 10.6 ng/mL efficacy threshold, and how much exposure is lost
when patients miss doses.

The structural model is a two-compartment disposition with first-order absorption
and linear elimination. Because no intravenous data were collected, every
disposition parameter is an *apparent* (F-scaled) value and no bioavailability
term is estimated. The absorption rate constant was not identifiable from the
sparse opportunistic sampling and was fixed to a literature value.

# Population

The model was fit to 113 plasma concentrations from 33 adults treated for a solid
tumour at Lausanne University Hospital (Ravix 2024 Table 1). Two thirds had
melanoma (22 patients, 67%); the remainder had ovarian cancer (3), breast cancer
(2), cholangiocarcinoma (2), and one each of thyroid carcinoma, gastrointestinal
stromal tumour, hepatocellular carcinoma and ileocecal carcinoma. Fifteen
patients (45%) were women. Median (min, max) age was 63 (30, 85) years, body
weight 70 (45, 96) kg, BMI 25.3 (17.3, 33.8) kg/m^2 and fat-free mass 46.35 (32,
68) kg.

Sampling was sparse and opportunistic: a median of 3 (1, 11) samples per patient,
drawn 5 h (0.13, 202 h) after the reported intake. Twenty-two patients (67%) took
trametinib together with dabrafenib. Eleven patients had their dose history
captured electronically by a MEMS pillbox; for the rest it was reconstructed from
consultation notes, assuming steady state where no information existed.

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Ravix_2024_trametinib.R`. They are collected here for
review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (ka) | 0.913 /h, fixed | Table 2, row "ka (h-1)", printed as "0.913 fixed"; attributed to reference [38] (Balakirouchenane et al.) in Results 3.2 and the Discussion |
| `lcl` (CL/F) | 3.96 L/h | Table 2, row "thetaCL (L.h-1)", RSE 6% |
| `lvc` (Vc/F) | 108 L | Table 2, row "V2 (L)", RSE 16% |
| `lq` (Q/F) | 29.4 L/h | Table 2, row "Q (L.h-1)", RSE 30% |
| `lvp` (Vp/F) | 286 L | Table 2, row "V3 (L)", RSE 25% |
| `e_age_cl` | -0.69 | Table 2, row "thetaAGE", RSE 32% |
| `e_ffm_cl` | 1.41 | Table 2, row "thetaFFM", RSE 17% |
| `etalcl` | 23% CV -> omega^2 = 0.0515459 | Table 2, row "IIV_CL (%)", RSE 14% |
| `propSd` | 20% | Table 2, row "sigma_prop (%)", RSE 10% |
| CL covariate equation | `CL = thetaCL * (1 + thetaAGE * (AGE - 63)/63) * (1 + thetaFFM * (FFM - 46.35)/46.35)` | Table 2 footnote ("Final equation") |
| AGE reference (63 y), FFM reference (46.35 kg) | medians | Table 2 footnote; Table 1 cohort medians |
| Two-compartment structure, first-order absorption, linear elimination | n/a | Results Section 3.2, paragraph 1 |
| IIV on CL only | n/a | Results Section 3.2 ("IIV to the other PK parameters ... estimated to be close to 0") |
| Proportional residual error | n/a | Results Section 3.2 ("The proportional error model was retained") |
| FFM derivation (Deurenberg) | `FFM = BW * (1 - BFPbmi/100)`; `BFPbmi = 1.20*BMI + 0.23*AGE - 10.8*SEX - 5.4` | Methods Section 2.1, Equations (1) and (2) |
| Therapeutic target Cmin >= 10.6 ng/mL | n/a | Methods Section 2.2.5; Table 3 caption |

Note that `ka` is the only fixed parameter. Everything else in Table 2 is an
estimate with a reported RSE and a bootstrap interval.

# Reproducing the values the paper prints about its own model

These checks are deterministic -- no random effects, no simulated cohort -- so
they are gated tightly. Each one independently exercises a different part of the
transcription.

## The Deurenberg fat-free-mass derivation

The paper derives FFM from body weight, BMI, age and sex rather than measuring
it. Evaluating its Equations (1) and (2) at the Table 1 median demographics
(70 kg, BMI 25.3 kg/m^2, 63 years) and mixing the sexes at the reported 45%
female should return the reported median FFM of 46.35 kg. It is a sharp check on
both the equation and the sex coding: the paper codes `SEX = 1` for men, whereas
the register's canonical `SEXF` column codes 1 for women.

```{r deurenberg}
# Ravix 2024 Methods Eq (1) and (2). SEXF is the canonical column (1 = female);
# the paper's SEX is 1 = male, hence the (1 - SEXF).
ffm_deurenberg <- function(WT, BMI, AGE, SEXF) {
  bfp <- 1.20 * BMI + 0.23 * AGE - 10.8 * (1 - SEXF) - 5.4
  WT * (1 - bfp / 100)
}

ffm_female <- ffm_deurenberg(WT = 70, BMI = 25.3, AGE = 63, SEXF = 1)
ffm_male   <- ffm_deurenberg(WT = 70, BMI = 25.3, AGE = 63, SEXF = 0)
ffm_mix    <- 0.45 * ffm_female + 0.55 * ffm_male   # Table 1: 15/33 = 45% women

tibble(
  Quantity = c("FFM, median woman", "FFM, median man", "FFM, 45% female mix",
               "FFM, Ravix 2024 Table 1 median"),
  `kg` = round(c(ffm_female, ffm_male, ffm_mix, 46.35), 2)
) |>
  knitr::kable(caption = "Deurenberg FFM at the Table 1 median demographics.")

# Deterministic: a wrong sex code or a dropped /100 moves this by kilograms.
stopifnot(abs(ffm_mix - 46.35) / 46.35 < 0.02)
```

The mix returns `r round(ffm_mix, 2)` kg against the reported 46.35 kg, a
`r sprintf("%.1f", 100 * abs(ffm_mix - 46.35) / 46.35)`% difference.

## The clearance covariate equation

Results Section 3.2 prints two worked consequences of the covariate model, and
both are exact identities that the packaged `ini()` values must reproduce:

* "a 66% increase in CL for an FFM of 68 kg ... compared with the population
  median of 46.35 kg";
* "a CL of 5.39 mL/h for a 30-year-old patient versus 3.22 mL/h for an
  80-year-old patient (40% decrease)". (The unit is a typo in the source: these
  are L/h, consistent with the 3.96 L/h typical value in Table 2.)

```{r covariate-equation}
th <- ui$theta
cl_typ <- function(AGE, FFM) {
  exp(th[["lcl"]]) *
    (1 + th[["e_age_cl"]] * (AGE - 63) / 63) *
    (1 + th[["e_ffm_cl"]] * (FFM - 46.35) / 46.35)
}

cl_age30 <- cl_typ(AGE = 30, FFM = 46.35)
cl_age80 <- cl_typ(AGE = 80, FFM = 46.35)
ffm_gain <- 100 * (cl_typ(AGE = 63, FFM = 68) / cl_typ(AGE = 63, FFM = 46.35) - 1)

tibble(
  Claim = c("CL at age 30 y (L/h)", "CL at age 80 y (L/h)",
            "Decrease from age 30 to 80 (%)", "CL increase at FFM 68 kg (%)"),
  Paper = c(5.39, 3.22, 40, 66),
  Model = round(c(cl_age30, cl_age80, 100 * (1 - cl_age80 / cl_age30), ffm_gain), 2)
) |>
  knitr::kable(caption = "Ravix 2024 Results 3.2 worked values vs the packaged model.")

stopifnot(
  abs(cl_age30 - 5.39) < 0.01,
  abs(cl_age80 - 3.22) < 0.01,
  abs(ffm_gain - 66) < 0.5
)
```

Both reproduce to the precision the paper prints them at. This is what confirms
the reading of the Table 2 footnote equation as a product of two linear
median-centred deviation terms, and it pins the sign of `e_age_cl` -- which the
PDF text layer renders ambiguously -- because only a negative coefficient makes
clearance fall with age.

## Absorption and disposition of the typical patient

`ka` was fixed rather than estimated, and the paper justifies the choice by the
tmax it produces: "ka was set to a literature value of 0.913 h-1 and provided a
median (min, max) Tmax of 1.75 h (1.51 h, 1.84 h)". Reproducing that tmax
exercises the depot-to-central wiring end to end.

```{r typical-profile}
mod     <- readModelDb("Ravix_2024_trametinib")
mod_typ <- rxode2::zeroRe(mod)

# Steady state on 2 mg once daily, fine grid over the last dosing interval.
ev_typ <- rxode2::et(amt = 2, cmt = "depot", ii = 24, until = 24 * 40)
ev_typ <- rxode2::et(ev_typ, seq(0, 24 * 41, by = 0.02))
ev_typ <- as.data.frame(ev_typ) |> mutate(id = 1L, AGE = 63, FFM = 46.35)

sim_typ <- rxode2::rxSolve(mod_typ, ev_typ) |> as.data.frame()
last_interval <- sim_typ |>
  filter(!is.na(Cc), time > 24 * 39, time <= 24 * 40) |>
  mutate(tad = time - 24 * 39)

tmax_ss <- last_interval$tad[which.max(last_interval$Cc)]

# Terminal half-life of the typical patient, from the two-compartment
# eigenvalue rather than from a fitted NCA slope (deterministic, no cohort).
kel <- th[["lcl"]] |> exp() / exp(th[["lvc"]])
k12 <- exp(th[["lq"]]) / exp(th[["lvc"]])
k21 <- exp(th[["lq"]]) / exp(th[["lvp"]])
s   <- kel + k12 + k21
beta <- 0.5 * (s - sqrt(s^2 - 4 * kel * k21))
thalf_terminal <- log(2) / beta

tibble(
  Quantity = c("Steady-state tmax (h)", "Steady-state Cmax (ng/mL)",
               "Steady-state Cmin (ng/mL)", "Terminal half-life (h)"),
  Value = round(c(tmax_ss, max(last_interval$Cc), min(last_interval$Cc),
                  thalf_terminal), 2)
) |>
  knitr::kable(caption = "Typical patient (age 63 y, FFM 46.35 kg) on 2 mg once daily.")

# The paper reports a median tmax of 1.75 h across its patients. A mis-wired
# depot or a mis-transcribed ka moves tmax by hours, not by tenths of an hour.
stopifnot(abs(tmax_ss - 1.75) < 0.25)
# Deterministic given the four disposition thetas; guards against a swapped
# Q/Vp or a transposed volume.
stopifnot(abs(thalf_terminal - 74.0) < 3)
```

```{r figure-typical, fig.cap = "Typical-value steady-state profile on 2 mg once daily. The dashed line is the 10.6 ng/mL efficacy threshold used throughout Ravix 2024."}
ggplot(last_interval, aes(tad, Cc)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 10.6, linetype = "dashed") +
  labs(x = "Time after dose (h)", y = "Trametinib (ng/mL)",
       title = "Typical patient at steady state, 2 mg once daily") +
  theme_bw()
```

The typical patient sits above the threshold throughout the interval
(Cmin `r round(min(last_interval$Cc), 1)` ng/mL), which is consistent with the
paper's finding that roughly two thirds of patients are on target at 2 mg.

Note the terminal half-life of `r round(thalf_terminal)` h. The trametinib label
quotes 127 h, and the authors flag the discrepancy themselves: their sampling was
concentrated within 24 h of the dose, with few samples beyond 15 h, which biases
the peripheral volume downwards relative to the two prior published models
(286 L here versus 417 L and 568 L). This is a property of the published model,
not of the transcription, and it is not gated below.

# Virtual cohort

The individual data are not public (Ravix 2024 Data Availability Statement), so
the cohort below is simulated. It follows the paper's own simulation design
(Methods Section 2.2.5): four strata defined by sex and the 65-year age cut-off,
with age drawn from a uniform distribution over each stratum's range.

The paper drew fat-free mass from a sex-specific uniform distribution spanning
the 10th to 90th percentiles reported by Larsson et al., a reference that is not
part of this paper. Rather than invent those percentiles, the cohort here derives
FFM from the paper's own Deurenberg equation applied to demographics sampled
across the Table 1 ranges. The consequences of that substitution are discussed
under Assumptions and deviations.

```{r cohort}
# set.seed() seeds R's RNG (used for the covariate draws below). It does NOT
# seed rxode2's simulation RNG, whose streams are partitioned per solver
# thread, so the eta draws differ between machines with different thread
# counts. Every assertion below is written to hold for any cohort the model
# can produce.
set.seed(20240611)

N_PER_ARM <- 200L   # skill cap: never more than 200 participants per arm

# Height, BMI and age are drawn uniformly across the Table 1 ranges; body
# weight follows from height and BMI, and draws whose weight falls outside the
# observed 45-96 kg range are rejected. This keeps all three marginals inside
# the observed ranges while keeping them mutually consistent.
make_arm <- function(n, age_lo, age_hi, sexf, arm, id_offset) {
  out <- tibble(
    AGE  = runif(4000, age_lo, age_hi),
    HT   = runif(4000, 150, 188),
    BMI  = runif(4000, 17.3, 33.8),
    SEXF = sexf
  ) |>
    mutate(WT = BMI * (HT / 100)^2) |>
    filter(WT >= 45, WT <= 96) |>
    slice_head(n = n) |>
    mutate(FFM = ffm_deurenberg(WT, BMI, AGE, SEXF),
           arm = arm,
           id  = id_offset + row_number())
  stopifnot(nrow(out) == n)   # fail loudly if rejection sampling came up short
  out
}

cohort <- bind_rows(
  make_arm(N_PER_ARM, 30, 65, 1, "Women 30-65 y",   0L),
  make_arm(N_PER_ARM, 30, 65, 0, "Men 30-65 y",   200L),
  make_arm(N_PER_ARM, 65, 85, 1, "Women 65-85 y", 400L),
  make_arm(N_PER_ARM, 65, 85, 0, "Men 65-85 y",   600L)
) |>
  mutate(arm = factor(arm, levels = c("Men 30-65 y", "Women 30-65 y",
                                      "Men 65-85 y", "Women 65-85 y")))

cohort |>
  group_by(arm) |>
  summarise(n = n(),
            `Age (median)` = round(median(AGE), 1),
            `WT (median)`  = round(median(WT), 1),
            `FFM (median)` = round(median(FFM), 1),
            `FFM (p10)`    = round(quantile(FFM, 0.10), 1),
            `FFM (p90)`    = round(quantile(FFM, 0.90), 1),
            .groups = "drop") |>
  knitr::kable(caption = "Simulated cohort, 200 participants per arm.")

# The pooled cohort should land near the Table 1 medians it was built from.
# Wide enough to admit the sampling noise of 800 draws; still red if the
# Deurenberg derivation or the weight range were wrong.
stopifnot(
  abs(median(cohort$FFM) - 46.35) / 46.35 < 0.10,
  abs(median(cohort$WT)  - 70)    / 70    < 0.10
)
```

Pooled across arms the cohort has a median FFM of
`r round(median(cohort$FFM), 2)` kg and a median weight of
`r round(median(cohort$WT), 1)` kg, against the reported medians of 46.35 kg and
70 kg.

# Simulation

Each participant receives 2 mg once daily for 60 days, matching the paper's
"steady state (assumed here after 60 days)" in the Figure 4 caption.

```{r simulate}
dosing <- cohort |>
  tidyr::crossing(time = seq(0, 24 * 59, by = 24)) |>
  mutate(amt = 2, evid = 1L, cmt = "depot")

# Observations go on the ODE state `central`, never on the observable `Cc`.
obs <- cohort |>
  tidyr::crossing(time = c(seq(24, 24 * 60, by = 24),         # daily troughs
                           seq(24 * 59, 24 * 60, by = 1))) |> # last interval
  mutate(amt = NA_real_, evid = 0L, cmt = "central")

events <- bind_rows(dosing, obs) |>
  distinct(id, time, evid, .keep_all = TRUE) |>
  arrange(id, time, -evid)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("arm", "AGE", "FFM", "SEXF")) |>
  as.data.frame()
stopifnot(all(sim$Cc >= 0, na.rm = TRUE))
```

```{r figure-2, fig.cap = "Simulated steady-state profiles over the last dosing interval, by sex and age stratum. Solid line is the median, ribbon the 5th-95th percentile. Compare Figure 3 of Ravix 2024, which plots the same strata as Cmin distributions."}
sim |>
  filter(!is.na(Cc), time >= 24 * 59, time <= 24 * 60) |>
  mutate(tad = time - 24 * 59) |>
  group_by(arm, tad) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 10.6, linetype = "dashed") +
  facet_wrap(~arm) +
  labs(x = "Time after dose (h)", y = "Trametinib (ng/mL)",
       title = "Steady-state profiles on 2 mg once daily",
       caption = "Dashed line: 10.6 ng/mL efficacy threshold (Ravix 2024).") +
  theme_bw()
```

# Replicating Table 3: target attainment by dose

Table 3 of Ravix 2024 reports the percentage of patients reaching
Cmin >= 10.6 ng/mL at each of four daily doses, in each of the four strata.

The model is linear in dose, so trough concentrations at 0.5, 1 and 1.5 mg are
the 2 mg troughs scaled by the dose ratio. That is verified explicitly before it
is used, because it is a structural claim about the model rather than an
assumption.

```{r linearity}
# Verify dose-proportionality on a subset rather than assuming it.
check_ids <- cohort |> group_by(arm) |> slice_head(n = 10) |> ungroup()
ev_1mg <- bind_rows(
  check_ids |> tidyr::crossing(time = seq(0, 24 * 59, by = 24)) |>
    mutate(amt = 1, evid = 1L, cmt = "depot"),
  check_ids |> mutate(time = 24 * 60, amt = NA_real_, evid = 0L, cmt = "central")
) |> arrange(id, time, -evid)

cmin_1mg <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_1mg) |>
  as.data.frame() |> filter(!is.na(Cc), time == 24 * 60) |> select(id, Cc_1mg = Cc)

ev_2mg <- ev_1mg |> mutate(amt = ifelse(evid == 1L, 2, amt))
cmin_2mg <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_2mg) |>
  as.data.frame() |> filter(!is.na(Cc), time == 24 * 60) |> select(id, Cc_2mg = Cc)

lin <- inner_join(cmin_1mg, cmin_2mg, by = "id") |>
  mutate(pct_diff = 100 * (Cc_2mg / 2 - Cc_1mg) / Cc_1mg)
stopifnot(nrow(lin) == nrow(check_ids), max(abs(lin$pct_diff)) < 0.01)
```

Halving the dose halves the trough to within
`r sprintf("%.4f", max(abs(lin$pct_diff)))`%, so the scaling below is exact.

```{r table-3}
cmin_ss <- sim |>
  filter(!is.na(Cc), time == 24 * 60) |>
  select(id, arm, Cmin_2mg = Cc)

attainment <- lapply(c(0.5, 1, 1.5, 2), function(d) {
  cmin_ss |>
    group_by(arm) |>
    summarise(dose = d, pct = 100 * mean(Cmin_2mg * d / 2 >= 10.6), .groups = "drop")
}) |>
  bind_rows() |>
  tidyr::pivot_wider(names_from = dose, values_from = pct,
                     names_glue = "{dose} mg")

published <- tibble::tribble(
  ~arm,             ~`0.5 mg`, ~`1 mg`, ~`1.5 mg`, ~`2 mg`,
  "Men 30-65 y",             0,       1,        13,      37,
  "Women 30-65 y",           0,      16,        47,      71,
  "Men 65-85 y",             0,      10,        46,      75,
  "Women 65-85 y",           2,      45,        79,      93
)

bind_rows(
  attainment |> mutate(Source = "Simulated"),
  published  |> mutate(Source = "Ravix 2024 Table 3")
) |>
  relocate(Source) |>
  arrange(arm, Source) |>
  mutate(across(ends_with("mg"), \(x) round(x, 1))) |>
  knitr::kable(caption = paste(
    "Percentage of patients with steady-state Cmin >= 10.6 ng/mL.",
    "Simulated values use a fat-free-mass distribution derived from this",
    "paper's Table 1; Ravix 2024 used sex-specific percentiles from an",
    "external reference (see Assumptions and deviations)."
  ))
```

The simulated attainment is systematically **higher** than the published values,
by roughly 6 to 25 percentage points at 2 mg. The mechanism is identified under
Assumptions and deviations and is a property of the covariate distribution rather
than of the model transcription.

What does reproduce is the **ordering** of the four strata, which is the paper's
substantive conclusion: men are less likely to reach the target than women of the
same age, and older patients are more likely to reach it than younger patients of
the same sex. The ordering assertions below are stated on median trough
concentration rather than on a thresholded percentage, because a median is far
less sensitive to which cohort the RNG happens to draw.

```{r ordering}
med <- cmin_ss |>
  group_by(arm) |>
  summarise(median_Cmin = median(Cmin_2mg), .groups = "drop")

knitr::kable(
  med |> dplyr::rename("Stratum" = arm, "Median Cmin at 2 mg (ng/mL)" = median_Cmin) |>
    mutate(across(where(is.numeric), \(x) round(x, 1))),
  caption = "Median steady-state trough by stratum."
)

g <- setNames(med$median_Cmin, as.character(med$arm))

# Effect sizes, not orderings of near-zero differences. Realised ratios were
# 1.32 (sex, under 65), 1.45 (sex, over 65), 1.59 and 1.74 (age) on a 16-thread
# render; the bounds below sit well inside those with headroom, and each still
# goes red if a covariate sign or reference value were wrong.
stopifnot(
  g[["Women 30-65 y"]] / g[["Men 30-65 y"]]   > 1.15,  # sex effect, via FFM
  g[["Women 65-85 y"]] / g[["Men 65-85 y"]]   > 1.15,
  g[["Men 65-85 y"]]   / g[["Men 30-65 y"]]   > 1.25,  # age effect, direct
  g[["Women 65-85 y"]] / g[["Women 30-65 y"]] > 1.25
)

# Attainment must rise with dose in every stratum (exact under linear scaling).
stopifnot(all(apply(as.matrix(attainment[, -1]), 1, \(r) all(diff(r) >= 0))))
```

# PKNCA validation

A separate single-dose arm supports non-compartmental analysis. Two hundred
participants drawn across the full Table 1 demographic range receive a single
2 mg dose, sampled densely through absorption and out to 336 h (about 4.5
terminal half-lives).

```{r pknca-cohort}
set.seed(7)
sd_cohort <- tibble(
  id   = 1:200,
  AGE  = runif(200, 30, 85),
  HT   = runif(200, 150, 188),
  BMI  = runif(200, 17.3, 33.8),
  SEXF = rbinom(200, 1, 0.45)
) |>
  mutate(WT  = pmin(pmax(BMI * (HT / 100)^2, 45), 96),
         FFM = ffm_deurenberg(WT, BMI, AGE, SEXF),
         treatment = "2 mg single dose")

sd_grid <- sort(unique(c(seq(0, 6, by = 0.05), seq(6, 24, by = 0.25),
                         seq(24, 336, by = 3))))

sd_events <- bind_rows(
  sd_cohort |> mutate(time = 0, amt = 2, evid = 1L, cmt = "depot"),
  sd_cohort |> tidyr::crossing(time = sd_grid) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  arrange(id, time, -evid)

sd_sim <- rxode2::rxSolve(mod, events = sd_events,
                          keep = c("treatment", "FFM", "AGE")) |>
  as.data.frame()
# Guard against solver noise going negative in the far tail, which would make
# PKNCA take log() of a negative number and return NaN for aucinf.
stopifnot(all(sd_sim$Cc >= 0, na.rm = TRUE))
```

```{r pknca}
# Filter on !is.na(Cc) ONLY -- a `time > 0` or `Cc > 0` filter would drop the
# time-zero row that PKNCA needs to anchor AUC0-*.
sim_nca <- sd_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- sd_events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Internal identity: AUC(0-inf) must equal Dose / CL

For a linear model the extrapolated AUC after a single dose is exactly
`Dose / CL`, evaluated with each subject's own clearance. Both sides of this
comparison use the same drawn parameters, so the only difference is trapezoidal
and extrapolation error and a tight bound is the correct gate.

```{r auc-identity}
per_subject <- as.data.frame(nca_res) |>
  tidyr::pivot_wider(id_cols = id, names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(sd_sim |> dplyr::distinct(id, cl), by = "id") |>
  # dose 2 mg, cl in L/h -> mg/L; x1000 for ng/mL, matching the model's output
  dplyr::mutate(auc_theoretical = 2 / cl * 1000,
                auc_pct_diff    = 100 * (aucinf.obs - auc_theoretical) / auc_theoretical)

stopifnot(nrow(per_subject) == 200, !anyNA(per_subject$auc_pct_diff))

tibble(
  Statistic = c("Median % difference", "Max absolute % difference"),
  Value = round(c(median(per_subject$auc_pct_diff),
                  max(abs(per_subject$auc_pct_diff))), 4)
) |>
  knitr::kable(caption = "PKNCA AUC(0-inf) against the analytic Dose/CL.")

# Pure numerical error on identical parameters; realised max was 0.17%.
stopifnot(max(abs(per_subject$auc_pct_diff)) < 1)
```

## Comparison against published values

Ravix 2024 does not report an NCA table of its own, but it does report the tmax
its fixed `ka` produces. The trametinib label values quoted in the paper's
Introduction are included for context.

```{r nca-compare}
sim_summary <- per_subject |>
  dplyr::mutate(treatment = "2 mg single dose") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(cmax = median(cmax), tmax = median(tmax),
                   aucinf.obs = median(aucinf.obs), half.life = median(half.life),
                   .groups = "drop")

published_nca <- tibble::tribble(
  ~treatment,          ~tmax, ~half.life,
  "2 mg single dose",   1.75,        127
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = published_nca,
  by        = "treatment",
  params    = c("tmax", "half.life"),
  units     = c(tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated (median of 200 subjects) vs published. * differs by >20%.",
  "tmax is the value Ravix 2024 reports for its own model; the 127 h",
  "half-life is the product label's, not an output of this paper's model."
))

# tmax is a genuine gate on this paper's own reported value.
stopifnot(abs(sim_summary$tmax - 1.75) < 0.35)
```

The tmax row agrees. The half-life row is starred and is a **known deviation,
deliberately excluded from the gate**: 127 h is the label's terminal half-life
from richly sampled studies, whereas this model was fit to data concentrated
within 24 h of dosing. The authors say so themselves in the Discussion, noting
that their peripheral volume (286 L) is well below the 417 L and 568 L of the two
prior published trametinib models and attributing it to the sampling window. A
smaller peripheral volume gives a shorter terminal phase. Reproducing the label's
half-life would require changing the published parameters, which is exactly what
must not be done.

Per-subject NCA half-life is additionally unreliable here: with 23% CV on
clearance and a fixed absorption rate, the fitted terminal slope varies from
`r round(min(per_subject$half.life))` to
`r round(max(per_subject$half.life))` h across subjects. The deterministic
eigenvalue-based half-life computed earlier
(`r round(thalf_terminal)` h) is the meaningful summary and is gated there.

# Assumptions and deviations

* **Fat-free-mass distribution.** Ravix 2024 drew FFM from a sex-specific uniform
  distribution spanning the 10th-90th percentiles of Larsson et al. (their
  reference 49), which is not reproduced in this paper. Rather than invent those
  percentiles, this vignette derives FFM from the paper's own Deurenberg
  equations applied to age, height, BMI and sex sampled across the Table 1
  ranges. The resulting pooled median FFM (`r round(median(cohort$FFM), 1)` kg)
  matches the reported 46.35 kg closely, but the per-sex spread is narrower and
  lower than a general-population reference would give -- the simulated men have
  a median FFM of `r round(median(cohort$FFM[cohort$SEXF == 0]), 1)` kg. Lower
  FFM means lower clearance and higher troughs, which is why the simulated Table
  3 attainment percentages run above the published ones while their ordering is
  preserved. Obtaining the Larsson percentiles would be required to reproduce the
  percentages themselves.
* **Height distribution not sex-specific.** Table 1 reports height for the pooled
  cohort only (170 cm; 150, 188), so the same height range is used for both
  sexes. The simulated male-female FFM difference therefore arises entirely from
  the `-10.8 * SEX` term of the Deurenberg body-fat equation and understates the
  real difference, which also has a height component.
* **Body weight by rejection sampling.** Weight is computed from the sampled
  height and BMI, and draws outside the observed 45-96 kg range are rejected.
  This keeps the three marginals inside their reported ranges while keeping them
  mutually consistent; the paper does not report their joint distribution.
* **No bioavailability parameter.** The study had no intravenous arm and did not
  estimate F, so CL, Vc, Q and Vp are apparent oral values. The 72% bioavailability
  quoted in the paper's Introduction is a label value for the drug and is not part
  of the fitted model; applying it would rescale every disposition parameter.
* **Concomitant dabrafenib is not encoded.** Two thirds of the cohort took
  dabrafenib, and it was significant on clearance in the covariate analysis
  (+40%). The authors removed it after a sensitivity analysis in the 11 patients
  with electronically monitored dosing showed the effect was a confound of
  unrecorded missed doses, so it is absent from the final model and from this
  encoding. It is recorded in the model file's `covariatesDataExcluded` metadata
  along with the other eleven screened-but-dropped covariates.
* **IIV on clearance only.** Adding between-subject variability to any other
  parameter gave estimates near zero, so the simulated variability in volumes and
  intercompartmental clearance is nil. Simulated profiles are consequently
  narrower in the distribution phase than real data would be.
* **Terminal half-life.** The packaged model gives about
  `r round(thalf_terminal)` h against the label's 127 h; see the PKNCA comparison
  section. This is a property of the published parameter estimates, documented by
  the authors, and is not corrected here.
* **The "5.39 mL/h" unit typo.** Results Section 3.2 prints the age-effect worked
  example in mL/h. The values are L/h -- they are reproduced exactly by the L/h
  parameters of Table 2 -- and are treated as L/h throughout.
* **Non-paper-derived parameter values.** None. Every `ini()` value comes from
  Ravix 2024 Table 2, and both covariate reference values come from its footnote.
* **Residual error is excluded from the simulated figures.** `Cc` from `rxSolve`
  is the individual prediction; the 20% proportional residual is a measurement
  model and is deliberately not added to exposure-based target-attainment
  calculations, matching the paper's approach.
