---
title: "Ethinylestradiol + drospirenone extended-cycle contraception (Reif 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ethinylestradiol + drospirenone extended-cycle contraception (Reif 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Reif 2013 developed **two independent population PK models** from a single
Phase III trial, one per analyte of the combined oral contraceptive (COC).
Following the library's replicate-the-author's-structure policy they are shipped
as two model files with this one shared vignette.

```{r models}
ee <- rxode2::rxode(readModelDb("Reif_2013_ethinylestradiol"))
drsp <- rxode2::rxode(readModelDb("Reif_2013_drospirenone"))
```

* Citation: `r ee$reference`
* Article: <https://doi.org/10.1136/jfprhc-2012-100397> (PMC3632974, open access)
* `Reif_2013_ethinylestradiol` -- `r ee$description`
* `Reif_2013_drospirenone` -- `r drsp$description`

# Population

Approximately 1100 healthy young women in a Phase III, multicentre, randomised,
open-label, parallel-group trial (protocol 308683 / NCT00266032) received
ethinylestradiol (EE) 20 ug / drospirenone (DRSP) 3 mg once daily for one year
under one of three regimens: a flexible extended regimen with management of
intracyclic bleeding (flexibleMIB, 24-120 days' active intake followed by a
4-day tablet-free interval), a conventional 28-day cyclic regimen (24 days'
active intake followed by 4 days of placebo), or a fixed extended regimen
(120 days' uninterrupted active intake followed by a 4-day tablet-free
interval).

Four serum samples were taken per subject under an optimised sparse-sampling
scheme: two samples 45-120 minutes apart during the first cycle (Week 3, days
15-21) and two more after about 6 months (Week 27). For 99% of samples the
sampling time fell 0-36 h after the preceding dose. The EE dataset held 4218
concentrations from 1109 subjects and the DRSP dataset 4042 concentrations from
1096 subjects (Table 1).

Baseline demographics are Table 2 of the paper (EE dataset, n = 1109): median
age 24 years (5th-95th percentile 19-34), median body weight 62 kg (51-79.8),
median BMI 22 kg/m^2 (18.8-28). More than 98% of subjects were Caucasian, so
ethnic group was not tested as a covariate; smoking and concomitant medication
occurred in fewer than 10% of subjects and were likewise not tested.

The same information is available programmatically from either model's
`population` metadata:

```{r population}
str(ee$population)
```

# Source trace

Every `ini()` entry in both model files carries an in-file comment naming its
source location. They are collected here for review.

| Model | Equation / parameter | Value | Source location |
|---|---|---|---|
| EE | `lka` | 0.295 1/h | Appendix 1, `ka` (RSE 6.98%) |
| EE | `lcl` | 25.3 L/h | Appendix 1, `TVCL/F` (RSE 1.24%) |
| EE | `lvc` | 23.9 L | Appendix 1, `V2/F` (RSE 13.6%) |
| EE | `lvp` | 1330 L | Appendix 1, `V3/F` deep peripheral (RSE 3.62%) |
| EE | `vp2 <- vc` | 23.9 L | Appendix 1 footnote: 'For the model it was assumed that V2F = V4F' |
| EE | `lq` | 52.9 L/h | Appendix 1, `Q3/F` (RSE 7.01%) |
| EE | `lq2` | 8.49 L/h | Appendix 1, `Q4/F` (RSE 34.3%) |
| EE | `ltlag` | 0.353 h | Appendix 1, `ALAG` (RSE 2.78%) |
| EE | `lfdepot` | 1 (fixed) | Appendix 1, `F_week3` = 1 |
| EE | `e_occ2_fdepot` | 0.0815 | Appendix 1, `Diff_F_week27` = 8.15% (RSE 11.0%) |
| EE | `e_wt_cl` | 0.00591 1/kg | Appendix 1, `CL_BW` = 0.591 %/kg (RSE 20.1%) |
| EE | `e_age_cl` | 0.208 | Appendix 1, `CL_AGE` = 20.8 %/LN(year) (RSE 29.1%) |
| EE | `etalcl` | 0.105761 | Appendix 1, `IIV_CL` = 33.4 CV% -> log(0.334^2 + 1) |
| EE | `propSd` | 0.240481 | Appendix 1, proportional error 24.4 CV% -> sqrt(log(0.244^2 + 1)) |
| EE | `F = F_week3 * (1 + F_week27 * OCA)` | n/a | Appendix 1 footnote (equation) |
| EE | `CL = TVCL*EXP(ETA_CL)*CO2*CO1` | n/a | Appendix 1 footnote (equation) |
| EE | `CO1 = 1 + CL_AGE * (LOG(AGE) - LOG(24))` | n/a | Appendix 1 footnote (equation) |
| EE | `CO2 = 1 + CL_BW * (BW - 62)` | n/a | Appendix 1 footnote (equation) |
| DRSP | `lka` | 2.18 1/h | Appendix 2, `ka` (RSE 8.26%) |
| DRSP | `lcl` | 3.52 L/h | Appendix 2, `TVCL_week3/F` (RSE 0.98%) |
| DRSP | `e_occ2_cl` | -0.0655 | Appendix 2, `Diff_CL_week27/F` = -6.55% (RSE 13.0%) |
| DRSP | `lvc` | 51.6 L | Appendix 2, `V2/F` (RSE 3.64%) |
| DRSP | `lvp` | 204 L | Appendix 2, `V3/F` (RSE 7.99%) |
| DRSP | `lq` | 17.5 L/h | Appendix 2, `Q3/F` (RSE 4.62%) |
| DRSP | `ltlag` | 0.372 h | Appendix 2, `ALAG` (RSE 2.54%) |
| DRSP | `lfdepot` | 1 (fixed) | Appendix 2, `F` = 1 |
| DRSP | `e_wt_cl` | 0.00672 1/kg | Appendix 2, `CL_BW` = 0.672 %/kg (RSE 14.5%) |
| DRSP | `etalcl` | 0.267670 | Appendix 2, `IIV_CL/F` = 55.4 CV% -> log(0.554^2 + 1) |
| DRSP | `etalfdepot` | 0.204227 | Appendix 2, `IIV_F` = 47.6 CV% -> log(0.476^2 + 1) |
| DRSP | `cov(etalcl, etalfdepot)` | 0.212764 | Appendix 2, `rho_CL/F,F` = 0.91 |
| DRSP | `etaiov_fdepot_1/2` | 0.037320 | Appendix 2, `IOV_F` = 19.5 CV% -> log(0.195^2 + 1) |
| DRSP | `propSd` | 0.095879 | Appendix 2, proportional error 9.61 CV% -> sqrt(log(0.0961^2 + 1)) |
| DRSP | `addSd` | 0.95 ng/mL | Appendix 2, additive error 0.95 ug/l (RSE 26.0%) |
| DRSP | `CL = TVCL_week3*(1 + TVCL_visit5*OCA)*EXP(ETA(1))*CO2` | n/a | Appendix 2 footnote (equation) |
| DRSP | `CO2 = 1 + CL_BW * (BW - 62)` | n/a | Appendix 2 footnote (equation) |

Both appendices define their reported `CV%` as
`sqrt(exp(OMEGA) - 1) * 100`, so every variance above is recovered as
`log(CV^2 + 1)` and the proportional residual SD as `sqrt(log(CV^2 + 1))`
rather than the CV read off directly. The additive DRSP residual carries no
such footnote and is used as printed.

# Regimen calendars and the sampling occasions

`OCC = 1` is the Week 3 visit (the paper's `OCA = 0`) and `OCC = 2` is the
Week 27 visit. Figure 1B of the paper states that the Week 27 samples were
taken on days 15-21 of the seventh cycle for the conventional regimen and on
days 59-65 of the second cycle for the fixed extended regimen. Study **day 189**
satisfies both simultaneously, so it is used as the Week 27 sampling day for
every regimen.

```{r calendars}
# Study days on which an active tablet is taken, for a repeating
# <active>/<hormone-free> cycle.
dose_days <- function(active, hormone_free, until) {
  day <- seq.int(0, until)
  day[(day %% (active + hormone_free)) < active]
}

regimens <- list(
  "Flexible MIB"   = c(active = 72,  hfi = 4),  # 72 days = the trial's average
  "Conventional"   = c(active = 24,  hfi = 4),
  "Fixed extended" = c(active = 120, hfi = 4)
)

week27_day <- 189
week3_day <- 20

# Every regimen is mid-active-treatment across the whole Week 27 window.
stopifnot(vapply(
  regimens,
  function(r) {
    d <- dose_days(r[["active"]], r[["hfi"]], week27_day + 1)
    all(c(week27_day, week27_day + 1) %in% d)
  },
  logical(1)
))
```

# Deterministic replication of Table 3

Table 3 tabulates typical-subject steady-state exposure (AUC0-24h,ss) at the
5th, 50th and 95th percentiles of the weight and age distributions. These are
typical-value predictions, so they are reproduced with the random effects zeroed
and compared exactly -- there is no cohort noise to absorb.

```{r typical-helper}
# Dense early grid so NCA resolves Tmax (about 1.2-1.3 h for both analytes).
obs_grid <- c(seq(0, 4, by = 1 / 6), seq(4.5, 12, by = 0.5), seq(13, 24, by = 1))

# Trapezoidal AUC over one 24 h steady-state window from a typical-value solve.
typical_auc <- function(mod, dose_amt, days, window_day, params) {
  ev <-
    rxode2::et(amt = dose_amt, time = days * 24, cmt = "depot") |>
    rxode2::et(window_day * 24 + obs_grid, cmt = "central")
  s <- rxode2::rxSolve(rxode2::zeroRe(mod), ev, params, returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  stopifnot(nrow(s) == length(obs_grid), all(s$Cc >= 0))
  sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2)
}

daily <- 0:week3_day
```

```{r table3}
table3 <- dplyr::bind_rows(
  tidyr::expand_grid(analyte = "EE", WT = c(51, 62, 79.8), AGE = 24),
  tidyr::expand_grid(analyte = "EE", WT = 62, AGE = c(19, 34)),
  tidyr::expand_grid(analyte = "DRSP", WT = c(51, 62, 79.8), AGE = NA_real_)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    Simulated = if (analyte == "EE") {
      typical_auc(ee, 20000, daily, week3_day, c(WT = WT, AGE = AGE, OCC = 1))
    } else {
      typical_auc(drsp, 3000, daily, week3_day, c(WT = WT, OCC = 1))
    }
  ) |>
  dplyr::ungroup() |>
  # Reif 2013 Table 3, read row-wise.
  dplyr::mutate(Published = c(844.3, 789.8, 715.0, 829.9, 736.7, 917.6, 850.6, 760.5)) |>
  dplyr::mutate(`% diff` = 100 * (Simulated - Published) / Published)

table3 |>
  dplyr::mutate(
    Unit = ifelse(analyte == "EE", "pg*h/mL", "ng*h/mL"),
    dplyr::across(c(Simulated, Published, `% diff`), \(x) round(x, 2))
  ) |>
  dplyr::rename(
    "Analyte" = analyte, "Body weight (kg)" = WT, "Age (years)" = AGE,
    "AUC0-24h,ss unit" = Unit
  ) |>
  knitr::kable(caption = "Replicates Table 3 of Reif 2013: typical-subject AUC0-24h,ss by body weight and age.")

# Deterministic: no random effects are drawn, so this bound is a true
# transcription gate on the clearance, dose, unit and covariate-model form.
stopifnot(max(abs(table3$`% diff`)) < 1)
```

The paper's narrative statements fall out of the same numbers: EE exposure at
79.8 kg is 15.3% below that at 51 kg, EE exposure at 34 years is 11.2% below
that at 19 years, and DRSP exposure at 51 kg is 17.2% above that at 79.8 kg
(each expressed, as the paper does, relative to the larger of the pair).

```{r table3-narrative}
pct_drop <- function(hi, lo) 100 * (hi - lo) / hi

# Address a Table 3 cell by its covariate values rather than by position:
# the frame binds three sub-grids, so a positional index silently picks the
# wrong anchor row (e.g. the age comparison's 19-year row vs the 24-year one).
cell <- function(a, wt, age) {
  i <- which(
    table3$analyte == a & table3$WT == wt &
      (if (is.na(age)) is.na(table3$AGE) else !is.na(table3$AGE) & table3$AGE == age)
  )
  stopifnot(length(i) == 1L)
  table3$Simulated[i]
}

narrative <- data.frame(
  Claim = c(
    "EE: 79.8 kg vs 51 kg exposure reduction",
    "EE: 34 years vs 19 years exposure reduction",
    "DRSP: 51 kg vs 79.8 kg exposure increase"
  ),
  Published = c(15.3, 11.2, 17.2),
  Simulated = c(
    pct_drop(cell("EE", 51, 24), cell("EE", 79.8, 24)),
    pct_drop(cell("EE", 62, 19), cell("EE", 62, 34)),
    pct_drop(cell("DRSP", 51, NA), cell("DRSP", 79.8, NA))
  )
)
knitr::kable(narrative, digits = 2, caption = "Percentage exposure differences quoted in the Results text.")
stopifnot(max(abs(narrative$Simulated - narrative$Published)) < 0.5)
```

# Deterministic replication of the occasion effects

EE relative bioavailability is 8.15% higher at Week 27; DRSP apparent clearance
is 6.55% lower (3.52 -> 3.29 L/h). Both propagate to exposure exactly.

```{r occasion}
ee_w3 <- typical_auc(ee, 20000, daily, week3_day, c(WT = 62, AGE = 24, OCC = 1))
ee_w27 <- typical_auc(ee, 20000, daily, week3_day, c(WT = 62, AGE = 24, OCC = 2))
dr_w3 <- typical_auc(drsp, 3000, daily, week3_day, c(WT = 62, OCC = 1))
dr_w27 <- typical_auc(drsp, 3000, daily, week3_day, c(WT = 62, OCC = 2))

occ <- data.frame(
  Analyte = c("EE", "DRSP"),
  Quantity = c("F higher at Week 27", "CL/F lower at Week 27"),
  `Expected ratio` = c(1.0815, 1 / (1 - 0.0655)),
  `Simulated ratio` = c(ee_w27 / ee_w3, dr_w27 / dr_w3),
  check.names = FALSE
)
knitr::kable(occ, digits = 4, caption = "Week 27 / Week 3 steady-state exposure ratio.")
stopifnot(max(abs(occ$`Simulated ratio` - occ$`Expected ratio`)) < 0.005)
```

# Deterministic replication of the paper's central claim

> "Extension of the established 24-day treatment period does not change the
> steady-state pharmacokinetics of EE or DRSP."

At the Week 27 sampling day the three regimens are all mid-active-treatment, and
the typical-value exposure is identical to within a small fraction of a percent.

```{r regimen-equivalence}
reg_typ <- lapply(names(regimens), function(nm) {
  r <- regimens[[nm]]
  d <- dose_days(r[["active"]], r[["hfi"]], week27_day + 1)
  data.frame(
    Regimen = nm,
    `EE AUC0-24h,ss (pg*h/mL)` =
      typical_auc(ee, 20000, d, week27_day, c(WT = 62, AGE = 24, OCC = 2)),
    `DRSP AUC0-24h,ss (ng*h/mL)` =
      typical_auc(drsp, 3000, d, week27_day, c(WT = 62, OCC = 2)),
    check.names = FALSE
  )
}) |> dplyr::bind_rows()

knitr::kable(reg_typ, digits = 2, caption = "Typical-value steady-state exposure at Week 27 by regimen.")

spread <- function(x) 100 * (max(x) - min(x)) / stats::median(x)
# Deterministic (zeroRe), so the realised spread is a fixed property of the
# model: 0.06% for EE and 0.14% for DRSP. 1% still goes red if a regimen
# calendar is mis-built and a sampling window straddles a hormone-free interval.
stopifnot(
  spread(reg_typ$`EE AUC0-24h,ss (pg*h/mL)`) < 1,
  spread(reg_typ$`DRSP AUC0-24h,ss (ng*h/mL)`) < 1
)
```

# Virtual cohort

Original observed data are not public. The cohort below reproduces the Table 2
covariate distributions: body weight normally distributed (mean 63.0 kg,
SD 8.6 kg) and age log-normally distributed with median 24 years, as the paper
states.

```{r cohort}
# `set.seed()` seeds R's RNG for the covariate draws only. It does NOT seed
# rxode2's simulation RNG, whose streams are partitioned per solver thread, so
# the etas below differ between a 2-core CI runner and a 16-thread workstation.
# Every cohort-derived assertion in this vignette is written to hold for any
# cohort the model can produce.
set.seed(20130108)
n_per_arm <- 200L

make_cohort <- function(n, arm, id_offset = 0L) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    arm = arm,
    WT = pmin(pmax(stats::rnorm(n, 63.0, 8.6), 40), 110),
    AGE = pmin(pmax(stats::rlnorm(n, log(24), 0.175), 18), 45)
  )
}

make_events <- function(cohort, dose_amt, days, window_day, occ) {
  doses <- tidyr::expand_grid(cohort, .day = days) |>
    dplyr::mutate(time = .day * 24, amt = dose_amt, evid = 1L, cmt = "depot") |>
    dplyr::select(-.day)
  obs <- tidyr::expand_grid(cohort, .t = obs_grid) |>
    dplyr::mutate(
      time = window_day * 24 + .t, amt = NA_real_, evid = 0L, cmt = "central"
    ) |>
    dplyr::select(-.t)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(OCC = occ) |>
    dplyr::arrange(id, time, evid)
}

# Four arms: the Week 3 occasion (identical for all regimens, since every
# subject is still inside the mandatory 24-day phase) plus one Week 27 arm per
# regimen. IDs are offset so the arms never collide inside rxSolve.
arms <- dplyr::bind_rows(
  tibble::tibble(
    arm = "Week 3", occ = 1, window_day = week3_day,
    days = list(daily), offset = 0L
  ),
  tibble::tibble(
    arm = paste("Week 27 -", names(regimens)), occ = 2, window_day = week27_day,
    days = lapply(regimens, function(r) dose_days(r[["active"]], r[["hfi"]], week27_day + 1)),
    offset = c(1L, 2L, 3L) * 1000L
  )
)

cohorts <- Map(make_cohort, n_per_arm, arms$arm, arms$offset)
events_ee <- Map(make_events, cohorts, 20000, arms$days, arms$window_day, arms$occ) |>
  dplyr::bind_rows()
events_drsp <- Map(make_events, cohorts, 3000, arms$days, arms$window_day, arms$occ) |>
  dplyr::bind_rows()

stopifnot(
  !anyDuplicated(unique(events_ee[, c("id", "time", "evid")])),
  nrow(dplyr::distinct(events_ee, id)) == 4L * n_per_arm
)
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20130108)
sim_ee <- rxode2::rxSolve(ee, events = events_ee, keep = c("arm", "WT", "AGE")) |>
  as.data.frame()
sim_drsp <- rxode2::rxSolve(drsp, events = events_drsp, keep = c("arm", "WT", "AGE")) |>
  as.data.frame()

stopifnot(
  all(sim_ee$Cc[!is.na(sim_ee$Cc)] >= 0),
  all(sim_drsp$Cc[!is.na(sim_drsp$Cc)] >= 0)
)
```

```{r figure-1b, fig.cap = "Replicates Figure 1B of Reif 2013: median and 90% prediction interval of EE and DRSP serum concentration over one steady-state dosing interval."}
dplyr::bind_rows(
  sim_ee |> dplyr::mutate(Analyte = "EE (pg/mL)"),
  sim_drsp |> dplyr::mutate(Analyte = "DRSP (ng/mL)")
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad = time - ifelse(arm == "Week 3", week3_day, week27_day) * 24) |>
  dplyr::group_by(Analyte, arm, tad) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_wrap(~Analyte, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Serum concentration",
    colour = NULL, fill = NULL,
    caption = "Replicates Figure 1B of Reif 2013."
  ) +
  theme(legend.position = "bottom")
```

# PKNCA validation

Steady-state NCA over the 24 h dosing interval that starts at the sampling day,
computed per subject and grouped by arm (PKNCA recipe 3).

```{r pknca}
nca_arm <- function(sim, arm_label, window_day, dose_amt) {
  conc <- sim |>
    dplyr::filter(arm == arm_label, !is.na(Cc)) |>
    dplyr::transmute(id, arm, time = time - window_day * 24, Cc)
  # Time-zero guarantee: the grid already starts at the window open, so this
  # only fires if a future edit drops it. Real rows win the distinct().
  conc <- dplyr::bind_rows(
    conc,
    conc |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)
  stopifnot(nrow(conc) > 0)

  dose <- conc |>
    dplyr::distinct(id, arm) |>
    dplyr::mutate(time = 0, amt = dose_amt)

  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(as.data.frame(conc), Cc ~ time | arm + id),
    PKNCA::PKNCAdose(as.data.frame(dose), amt ~ time | arm + id),
    intervals = data.frame(
      start = 0, end = 24,
      cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
    )
  ))
  as.data.frame(res$result)
}

nca_ee <- Map(nca_arm, list(sim_ee), arms$arm, arms$window_day, 20000) |>
  dplyr::bind_rows()
nca_drsp <- Map(nca_arm, list(sim_drsp), arms$arm, arms$window_day, 3000) |>
  dplyr::bind_rows()

stopifnot(
  nrow(nca_ee) > 0, nrow(nca_drsp) > 0,
  !any(is.na(nca_ee$PPORRES[nca_ee$PPTESTCD == "auclast"]))
)
```

## Comparison against the published exposure distribution

Table 4 of the paper reports the distribution of *individual* AUC0-24h,ss by
regimen, pooling the exposures estimated for each subject. The simulated
counterpart pools each subject's Week 3 and Week 27 exposure, then summarises by
regimen. `ncaComparisonTable()` aggregates the per-subject simulated values with
the median, which is what Table 4's "Median" column reports.

```{r nca-pooled}
pool_by_regimen <- function(nca) {
  auc <- nca |>
    dplyr::filter(PPTESTCD == "auclast") |>
    dplyr::select(arm, id, PPORRES)
  wk3 <- auc |> dplyr::filter(arm == "Week 3")
  dplyr::bind_rows(lapply(names(regimens), function(nm) {
    dplyr::bind_rows(
      wk3,
      auc |> dplyr::filter(arm == paste("Week 27 -", nm))
    ) |>
      dplyr::mutate(Regimen = nm)
  })) |>
    dplyr::mutate(PPTESTCD = "auclast")
}

pooled_ee <- pool_by_regimen(nca_ee)
pooled_drsp <- pool_by_regimen(nca_drsp)

geomean <- function(x) exp(mean(log(x)))
geocv <- function(x) 100 * sqrt(exp(stats::var(log(x))) - 1)

summ <- dplyr::bind_rows(
  pooled_ee |> dplyr::mutate(Analyte = "EE (pg*h/mL)"),
  pooled_drsp |> dplyr::mutate(Analyte = "DRSP (ng*h/mL)")
) |>
  dplyr::group_by(Analyte, Regimen) |>
  dplyr::summarise(
    `5th pct` = stats::quantile(PPORRES, 0.05),
    Median = stats::median(PPORRES),
    `95th pct` = stats::quantile(PPORRES, 0.95),
    `Geometric mean` = geomean(PPORRES),
    `Geometric CV (%)` = geocv(PPORRES),
    .groups = "drop"
  )

knitr::kable(summ, digits = 1, caption = "Simulated counterpart of Table 4 of Reif 2013: distribution of individual AUC0-24h,ss by regimen.")
```

```{r nca-compare-ee}
# Reif 2013 Table 4, EE block (medians; the column ncaComparisonTable compares).
published_ee <- tibble::tribble(
  ~Regimen,         ~auclast,
  "Flexible MIB",   811,
  "Conventional",   818,
  "Fixed extended", 782
)

cmp_ee <- nlmixr2lib::ncaComparisonTable(
  simulated = pooled_ee |> dplyr::select(Regimen, id, PPTESTCD, PPORRES),
  reference = published_ee,
  by = "Regimen",
  units = c(auclast = "pg*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_ee, caption = "EE: simulated vs published median AUC0-24h,ss. * differs from reference by >20%.")
```

```{r nca-compare-drsp}
# Reif 2013 Table 4, DRSP block, plus the Methods statement that DRSP Cmax was
# assumed to be reached 1.4 h after administration.
published_drsp <- tibble::tribble(
  ~Regimen,         ~auclast, ~tmax,
  "Flexible MIB",   879,      1.4,
  "Conventional",   876,      1.4,
  "Fixed extended", 862,      1.4
)

tmax_drsp <- nca_drsp |>
  dplyr::filter(PPTESTCD == "tmax") |>
  dplyr::mutate(
    Regimen = dplyr::case_when(
      arm == "Week 3" ~ NA_character_,
      TRUE ~ sub("^Week 27 - ", "", arm)
    )
  ) |>
  dplyr::filter(!is.na(Regimen)) |>
  dplyr::select(Regimen, id, PPTESTCD, PPORRES)

cmp_drsp <- nlmixr2lib::ncaComparisonTable(
  simulated = dplyr::bind_rows(pooled_drsp |> dplyr::select(Regimen, id, PPTESTCD, PPORRES), tmax_drsp),
  reference = published_drsp,
  by = "Regimen",
  units = c(auclast = "ng*h/mL", tmax = "h"),
  tolerance_pct = 20
)
knitr::kable(cmp_drsp, caption = "DRSP: simulated vs published median AUC0-24h,ss and assumed Tmax. * differs from reference by >20%.")
```

```{r nca-assertions}
pct <- function(tbl) {
  v <- suppressWarnings(abs(as.numeric(gsub("[^0-9.eE+-]", "", tbl$`% diff`))))
  # A table whose every row is NA would make max(..., na.rm = TRUE) return -Inf
  # and turn the gate below into one that cannot go red (pattern 10).
  stopifnot(nrow(tbl) > 0, sum(!is.na(v)) == nrow(tbl))
  v
}
# Cohort-derived, so the bound has to survive a different eta draw on a
# different thread count. Realised |% diff| stayed under 6% for both analytes
# across repeated renders; 20% is the tolerance ncaComparisonTable already
# flags at, and a mis-transcribed clearance, dose or unit moves these medians
# by tens of percent.
stopifnot(max(pct(cmp_ee), na.rm = TRUE) < 20, max(pct(cmp_drsp), na.rm = TRUE) < 20)

# Table 4 reports total geoCV of 31.9% (EE) and 27.9% (DRSP). The simulated
# spread is driven by the published IIV (EE 33.4 CV% on CL alone; DRSP 55.4 and
# 47.6 CV% on a strongly correlated CL/F and F pair, so the exposure CV is far
# below either), plus the covariate distributions.
geo <- summ$`Geometric CV (%)`
stopifnot(all(geo > 15), all(geo < 55))
```

```{r figure-4, fig.cap = "Replicates Figure 4 of Reif 2013: distribution of individual steady-state exposure by regimen."}
dplyr::bind_rows(
  pooled_ee |> dplyr::mutate(Analyte = "EE (pg*h/mL)"),
  pooled_drsp |> dplyr::mutate(Analyte = "DRSP (ng*h/mL)")
) |>
  ggplot(aes(Regimen, PPORRES)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~Analyte, scales = "free_y") +
  labs(
    x = NULL, y = "AUC0-24h,ss",
    caption = "Replicates Figure 4 of Reif 2013."
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# Replicate Figure 5

Figure 5 shows the minimum and maximum DRSP concentrations expected under the
conventional 24/4-day regimen and an extended 72/4-day flexibleMIB regimen. The
maxima are taken 1.4 h after each dose, as the paper's Methods specify; the
minima are the pre-dose troughs, extended through the hormone-free interval.

```{r figure-5, fig.cap = "Replicates Figure 5 of Reif 2013: minimum and maximum DRSP concentrations under a conventional 24/4-day and an extended 72/4-day flexibleMIB regimen."}
fig5_days <- 220
fig5 <- lapply(c("Conventional (24/4)" = 24, "Flexible MIB (72/4)" = 72), function(active) {
  d <- dose_days(active, 4, fig5_days)
  ev <-
    rxode2::et(amt = 3000, time = d * 24, cmt = "depot") |>
    rxode2::et(sort(c(seq(0, fig5_days) * 24 - 1e-6, d * 24 + 1.4)), cmt = "central")
  s <- rxode2::rxSolve(rxode2::zeroRe(drsp), ev, c(WT = 62, OCC = 1),
                       returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  s$Metric <- ifelse(
    abs(s$time / 24 - round(s$time / 24)) < 1e-6, "Minimum", "Maximum"
  )
  s
})
fig5 <- dplyr::bind_rows(fig5, .id = "Regimen")

ggplot(fig5, aes(time / 24, Cc, colour = Metric)) +
  geom_line() +
  facet_wrap(~Regimen, ncol = 1) +
  labs(
    x = "Study day", y = "DRSP serum concentration (ng/mL)", colour = NULL,
    caption = "Replicates Figure 5 of Reif 2013."
  )
```

The three conclusions the paper draws from Figure 5 are checked directly. All
are typical-value quantities, so the bounds are tight.

```{r figure-5-checks}
ss_max <- fig5 |>
  dplyr::filter(Metric == "Maximum", time / 24 > 150) |>
  dplyr::group_by(Regimen) |>
  dplyr::summarise(Cmax = stats::median(Cc), .groups = "drop")

# (iii) DRSP level at the end of the 4-day hormone-free interval, after
#       preceding active-treatment runs of very different length.
hfi_end <- vapply(c(24, 72, 120), function(active) {
  d <- dose_days(active, 4, active + 3)
  ev <-
    rxode2::et(amt = 3000, time = d * 24, cmt = "depot") |>
    rxode2::et((active + 4) * 24, cmt = "central")
  s <- rxode2::rxSolve(rxode2::zeroRe(drsp), ev, c(WT = 62, OCC = 1),
                       returnType = "data.frame")
  s$Cc[!is.na(s$Cc)]
}, numeric(1))

fig5_tab <- data.frame(
  Claim = c(
    "(i) steady-state Cmax is the same under both regimens",
    "(iii) concentration at the end of the 4-day hormone-free interval is independent of the preceding treatment length (24 / 72 / 120 days)"
  ),
  `Simulated values` = c(
    paste(round(ss_max$Cmax, 2), collapse = " / "),
    paste(round(hfi_end, 3), collapse = " / ")
  ),
  `Spread (%)` = c(spread(ss_max$Cmax), spread(hfi_end)),
  check.names = FALSE
)
knitr::kable(fig5_tab, digits = 3, caption = "Figure 5 conclusions, checked against the model.")

# Deterministic. Realised 0.00% and 0.09%; 1% still goes red if the model's
# distribution parameters or the hormone-free interval length are wrong.
stopifnot(fig5_tab$`Spread (%)` < 1)
```

Conclusion (ii) -- that the time to DRSP steady state does not depend on the
length of the preceding treatment period -- follows from the regimen-equivalence
check above, where the Week 27 typical-value exposures of the 24/4, 72/4 and
120/4 regimens agree to within 0.14%.

# Assumptions and deviations

- **Covariate distributions.** Table 2 gives the mean, SD and percentiles of
  body weight and age and states that body weight is normally distributed and
  age log-normally distributed, but not the joint distribution. The virtual
  cohort draws them independently: `WT ~ N(63.0, 8.6)` truncated to 40-110 kg,
  and `AGE ~ lognormal(log(24), 0.175)` truncated to 18-45 years, the `sdlog`
  chosen to reproduce the reported mean/SD pair (24.8 / 4.4 years). The paper
  reports no weight-age correlation (only weight with BMI, which the final
  models do not use).
- **Sampling day for the Week 27 occasion.** The paper gives the Week 27 window
  per regimen in the Figure 1B caption rather than as a study day. Study day 189
  is the single day that satisfies the caption for both the conventional
  (days 15-21 of cycle 7) and the fixed extended (days 59-65 of cycle 2)
  regimens, and it falls mid-active-treatment for the flexibleMIB regimen at its
  reported average cycle length; it is used for all three.
- **Flexible MIB cycle length.** The flexible regimen has no fixed cycle. The
  paper reports an average of 72 days' active treatment in the trial, and uses a
  72/4-day regimen for its own Figure 5 simulations; the same 72/4 calendar is
  used here.
- **Week 3 arm is not split by regimen.** At Week 3 (days 15-21 of the first
  cycle) every subject is still inside the mandatory 24-day active phase, so all
  three regimens are identical by construction and one arm is simulated.
- **CV%-to-variance conversion.** Both appendices define their reported CV% as
  `sqrt(exp(OMEGA) - 1) * 100`, including for the proportional residual error.
  Every variance is therefore recovered as `log(CV^2 + 1)`, and `propSd` as
  `sqrt(log(CV^2 + 1))` -- 0.240481 rather than 0.244 for EE and 0.095879 rather
  than 0.0961 for DRSP. The difference is under 1.5% either way. The additive
  DRSP residual (0.95 ug/l) carries no such footnote and is used as printed.
- **EE second peripheral volume.** `V4/F` is not an independently estimated
  parameter: the Appendix 1 footnote states 'For the model it was assumed that
  V2F = V4F', so the model file carries a single `lvc` and sets `vp2 <- vc`
  inside `model()`. Consistent with this, Appendix 1 reports no RSE for `V4/F`.
  The paper's `V3/F` (deep) maps to `peripheral1` and `V4/F` to `peripheral2`,
  preserving the paper's numbering rather than ordering by depth.
- **Dose units.** EE doses are entered as 20000 ng and DRSP doses as 3000 ug so
  that `central / vc` lands directly in the paper's reporting units (pg/mL for
  EE, ng/mL for DRSP). No value is converted; only the amount unit is chosen to
  match the reported concentration unit.
- **Inter-occasion variability on DRSP bioavailability.** Appendix 2 reports one
  shared IOV magnitude (19.5 CV%) without a per-occasion breakdown, so the two
  occasions carry the same variance with the second fixed, which is the NONMEM
  `$OMEGA BLOCK(1) SAME` idiom. `rxode2` cannot simulate the `eta ~ var | OCC`
  multi-level syntax, so the occasion-indicator expansion is used; it emits a
  benign mu-referencing warning that affects estimation only, not simulation.
- **No regimen covariate.** The paper tested 'treatment group' as a covariate
  and rejected it (drop in objective function of 3.032 for EE and 1.256 for
  DRSP), so the shipped models carry no regimen term. Regimen enters only
  through the dosing calendar.
- **Nothing was tuned.** No parameter was adjusted to improve agreement with any
  published value.
