---
title: "Cycloserine (Resendiz-Galvan 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cycloserine (Resendiz-Galvan 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("ResendizGalvan_2025_cycloserine"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1128/aac.00101-25> (PMCID: PMC12486832)

Cycloserine is a core drug in longer multidrug-resistant tuberculosis (MDR-TB)
regimens. Its efficacy is time-dependent, driven by the percentage of the dosing
interval during which plasma concentration exceeds the MIC (%*T*>MIC), while its
neuropsychiatric toxicity is concentration-related. This analysis quantified
cycloserine PK in an Indian MDR-TB cohort, where exposures turned out to be
markedly lower than previously published for the same regimens.

The structural model is a one-compartment disposition with Savic
transit-compartment absorption. Its distinguishing feature is the clearance
model: following Holford's approach, the Cockcroft-Gault creatinine clearance is
recomputed for every individual **as if they were a 56 kg male**, which strips
weight and sex out of the renal term so that it does not compete with the
fat-free-mass allometry. Normalising that quantity by its cohort median
(122 mL/min) gives a renal-function ratio `RF` that multiplies the renal
clearance arm only:

$$\mathrm{CL} = \left(\mathrm{CL_{nr}} + \mathrm{RF}\cdot \mathrm{CL_{r}}\right)\cdot F_{size}$$

# Population

```{r population}
pop <- ui$population
tibble::tibble(Field = names(pop), Value = vapply(pop, paste, character(1), collapse = "; ")) |>
  knitr::kable()
```

# Source trace

Every value in `ini()` and every non-trivial equation in `model()`, with the
location it came from in Resendiz-Galvan 2025.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity,                              ~Value,                 ~`Source location`,
  "Renal clearance CLr",                  "0.589 L/h",            "Table 2 (95% CI 0.449-0.745)",
  "Non-renal clearance CLnr",             "0.901 L/h",            "Table 2 (95% CI 0.767-1.04)",
  "Central volume Vc",                    "37.0 L",               "Table 2 (95% CI 35.1-38.9)",
  "Absorption rate constant ka",          "2.15 1/h",             "Table 2 (95% CI 1.51-3.55)",
  "Mean transit time MTT",                "0.610 h",              "Table 2 (95% CI 0.362-0.833)",
  "Transit compartments N",               "4 (fixed)",            "Table 2 + footnote c",
  "Bioavailability F",                    "1 (fixed)",            "Table 2",
  "BSV on CL",                            "34.1 %CV",             "Table 2; footnote f gives %CV = sqrt(omega^2)*100",
  "BOV on ka",                            "94.6 %CV",             "Table 2",
  "BOV on MTT",                           "83.5 %CV",             "Table 2",
  "BOV on F",                             "20.7 %CV",             "Table 2",
  "Unobserved-dose BOV inflation",        "2.26-fold",            "Table 2 + footnote d",
  "Proportional residual error",          "4.58 %",               "Table 2 (95% CI 3.23-5.66)",
  "Additive residual error",              "0.856 mg/L",           "Table 2 (95% CI 0.716-1.03)",
  "FFM reference",                        "38.6 kg",              "Table 2 footnote a; Results",
  "CLcr,56M reference (median)",          "122 mL/min",           "Table 2 footnote b",
  "Cockcroft-Gault, 56 kg male",          "(140-AGE)*56/(72*SCr)", "Methods equation (typeset image aac.00101-25.m001)",
  "Renal function ratio RF",              "CLcr,56M / median",    "Methods equation (image m002)",
  "Clearance assembly",                   "(CLnr + RF*CLr)*Fsize", "Methods equation (image m003)",
  "FFM allometric exponents",             "0.75 on CL, 1 on Vc",  "NOT printed in the source; standard theory-based values (see Errata)"
) |>
  knitr::kable()
```

# Virtual cohort

The cohort is generated the way the paper generated fat-free mass: body weight,
height, and sex are sampled to the Table 1 marginals, and FFM is then derived
with the Janmahasatian formula (the paper's reference 25). Age and serum
creatinine are sampled to their Table 1 marginals and feed the renal-function
ratio. Sampling weight and height this way also gives the body weight needed to
assign the WHO weight-banded regimen.

```{r cohort}
set.seed(20250902)
n_per_arm <- 200 # cap: never more than 200 participants per arm

# log-normal parameters from a published median and IQR
lnpar <- function(m, q1, q3) list(meanlog = log(m), sdlog = (log(q3) - log(q1)) / (2 * qnorm(0.75)))
p_wt <- lnpar(55.5, 46.0, 65.9)  # Table 1 weight, kg
p_ht <- lnpar(1.59, 1.53, 1.69)  # Table 1 height, m
p_age <- lnpar(27, 21, 35)       # Table 1 age, years
p_cr <- lnpar(0.7, 0.6, 0.9)     # Table 1 serum creatinine, mg/dL

# Draw each covariate at evenly-spaced quantiles rather than at random, then
# permute independently. The Table 1 marginals are a specification to be
# reproduced, not a quantity to be estimated, so stratifying removes the
# Monte Carlo noise that would otherwise shift the cohort median by several
# percent at n = 200 and bias every exposure comparison below.
strat <- function(n, p) qlnorm((seq_len(n) - 0.5) / n, p$meanlog, p$sdlog)
n_female <- round(0.65 * n_per_arm) # Table 1: 117/180 = 65.0% female

subjects <- tibble::tibble(
  subj = seq_len(n_per_arm),
  WT   = sample(strat(n_per_arm, p_wt)),
  HT   = sample(strat(n_per_arm, p_ht)),
  SEXF = sample(rep(c(1L, 0L), c(n_female, n_per_arm - n_female))),
  AGE  = pmax(15, sample(strat(n_per_arm, p_age))), # eligibility floor >= 15 y
  CREAT = sample(strat(n_per_arm, p_cr))
) |>
  dplyr::mutate(
    BMI = WT / HT^2,
    # Janmahasatian et al. Clin Pharmacokinet 2005;44:1051-1065
    FFM = dplyr::if_else(SEXF == 1,
                         9270 * WT / (8780 + 244 * BMI),
                         9270 * WT / (6680 + 216 * BMI))
  )

tibble::tribble(
  ~Covariate,               ~Simulated,                                     ~`Table 1`,
  "Weight (kg)",            sprintf("%.1f (%.1f-%.1f)", median(subjects$WT), quantile(subjects$WT, .25), quantile(subjects$WT, .75)), "55.5 (46.0-65.9)",
  "Fat-free mass (kg)",     sprintf("%.1f (%.1f-%.1f)", median(subjects$FFM), quantile(subjects$FFM, .25), quantile(subjects$FFM, .75)), "38.7 (32.3-47.1)",
  "Age (years)",            sprintf("%.0f (%.0f-%.0f)", median(subjects$AGE), quantile(subjects$AGE, .25), quantile(subjects$AGE, .75)), "27 (21-35)",
  "Serum creatinine (mg/dL)", sprintf("%.2f (%.2f-%.2f)", median(subjects$CREAT), quantile(subjects$CREAT, .25), quantile(subjects$CREAT, .75)), "0.7 (0.6-0.9)"
) |>
  knitr::kable(caption = "Simulated cohort, median (IQR), against Resendiz-Galvan 2025 Table 1.")
```

# Simulation

The eight regimens evaluated in the paper (Table 3, Figure 3) are dosed to
steady state and observed over the final 24-hour day. `OCC = 2` throughout:
these are prescribed, fully-observed regimens, so the 2.26-fold inflation of
between-occasion variability that applies to unobserved home doses is not in
play.

```{r regimens}
tau_day <- 24
# The typical individual's t1/2 is ~17 h, but low-clearance individuals in the
# tail of the BSV distribution reach ~60 h across a 1,600-subject cohort. A
# 31-day run-in gives even the slowest subject >10 half-lives, which the
# identity check below verifies explicitly rather than assuming.
n_days  <- 32
t_ss    <- tau_day * (n_days - 1)

# time (h within a day) and amount (mg) for each regimen
regimens <- list(
  "250 mg QD"     = tibble::tibble(t = 0,          amt = 250),
  "500 mg QD"     = tibble::tibble(t = 0,          amt = 500),
  "250 mg BID"    = tibble::tibble(t = c(0, 12),   amt = c(250, 250)),
  "250/500 am/pm" = tibble::tibble(t = c(0, 12),   amt = c(250, 500)),
  "250 mg TID"    = tibble::tibble(t = c(0, 8, 16), amt = c(250, 250, 250)),
  "500 mg BID"    = tibble::tibble(t = c(0, 12),   amt = c(500, 500)),
  "750 mg BID"    = tibble::tibble(t = c(0, 12),   amt = c(750, 750))
)

# Every arm uses the SAME subject ids, so that reseeding before each solve gives
# every arm the same random-effect draws (common random numbers). Regimen is
# then the only thing that differs between arms, which removes Monte Carlo noise
# from every between-regimen comparison below.
make_arm <- function(sched, label, subj_df) {
  s <- dplyr::mutate(subj_df, id = subj)
  dosing <- tidyr::expand_grid(
    dplyr::select(s, id), day = seq_len(n_days) - 1L, sched
  ) |>
    dplyr::mutate(time = t + tau_day * day, evid = 1L, cmt = "depot") |>
    dplyr::select(id, time, amt, evid, cmt)
  obs <- tidyr::expand_grid(
    dplyr::select(s, id), time = seq(t_ss, t_ss + tau_day, by = 0.1)
  ) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(dosing, obs) |>
    dplyr::left_join(dplyr::select(s, id, FFM, AGE, CREAT, WT), by = "id") |>
    dplyr::mutate(OCC = 2, treatment = label) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# WHO weight-banded arm: 250 mg BID at <= 45 kg, 250/500 am/pm above 45 kg
# (Figure 3 caption). The schedule depends on the subject's weight, so it is
# assembled per weight band and then stacked.
who_arm <- dplyr::bind_rows(
  make_arm(regimens[["250 mg BID"]], "WHO weight-banded", dplyr::filter(subjects, WT <= 45)),
  make_arm(regimens[["250/500 am/pm"]], "WHO weight-banded", dplyr::filter(subjects, WT > 45))
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

arms <- c(
  lapply(seq_along(regimens), function(i) make_arm(regimens[[i]], names(regimens)[i], subjects)),
  list(who_arm)
)
names(arms) <- c(names(regimens), "WHO weight-banded")
events <- dplyr::bind_rows(arms)
stopifnot(!anyDuplicated(events[, c("id", "time", "evid", "treatment")]))
```

```{r simulate}
# useLinCmt = FALSE is REQUIRED. This is a one-compartment oral model, so
# rxode2's automatic ODE -> linCmt() conversion fires; linCmt() knows nothing
# about the transit() input term, and because the model sets f(depot) <- 0 (so
# the dose enters only through the transit chain) the converted model receives
# a dose of zero and returns all-zero concentrations SILENTLY -- no error, no
# warning. See the Errata.
# Each arm is solved separately with the RNG reseeded first. Both seeds are
# needed: rxode2 draws the random effects from its own generator, so set.seed()
# alone does NOT reproduce the same etas across solves.
solve_arm <- function(ev_arm) {
  set.seed(20250902)
  rxode2::rxSetSeed(20250902)
  rxode2::rxSolve(
    ui, events = ev_arm,
    keep = c("treatment", "FFM", "AGE", "CREAT", "WT"),
    useLinCmt = FALSE
  ) |>
    as.data.frame()
}
# rxSolve returns only the observation records, and drops `evid` entirely --
# do not try to filter on it afterwards.
sim <- dplyr::bind_rows(lapply(arms, solve_arm))

# Guard against the silent all-zero failure mode described above.
stopifnot(nrow(sim) > 0, max(sim$Cc, na.rm = TRUE) > 1)
stopifnot(dplyr::n_distinct(sim$treatment) == 8)

# Confirm the common random numbers actually took: a given subject must have
# identical CL and bioavailability in every arm.
crn <- sim |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(cl = dplyr::first(cl), fbio = dplyr::first(fbio), .groups = "drop") |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    n_cl = dplyr::n_distinct(round(cl, 10)),
    n_fb = dplyr::n_distinct(round(fbio, 10)), .groups = "drop"
  )
stopifnot(all(crn$n_cl == 1), all(crn$n_fb == 1))
```

## Steady-state concentration-time profiles

```{r fig-profiles, fig.cap = "Simulated steady-state cycloserine concentration-time profiles (median and 5th-95th percentiles) over the final 24-hour dosing day, by regimen. Comparable in shape to Figure 2 of Resendiz-Galvan 2025."}
sim |>
  dplyr::mutate(tday = time - t_ss) |>
  dplyr::group_by(treatment, tday) |>
  dplyr::summarise(
    lo = quantile(Cc, 0.05), md = median(Cc), hi = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(tday, md)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2) +
  ggplot2::geom_line() +
  ggplot2::geom_hline(yintercept = 16, linetype = "dashed", colour = "firebrick") +
  ggplot2::facet_wrap(~treatment) +
  ggplot2::labs(
    x = "Time within the steady-state day (h)", y = "Cycloserine (mg/L)",
    caption = "Dashed red line: the cohort median MIC of 16 mg/L."
  ) +
  ggplot2::theme_bw()
```

# PKNCA validation

```{r pknca}
conc_df <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, treatment, time, amt)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = t_ss, end = t_ss + tau_day,
  auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE, cav = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca <- as.data.frame(res$result)
stopifnot(nrow(nca) > 0)
```

## Structural identity check

At steady state a linear one-compartment model must satisfy
AUC~0-24~ = (total daily dose x F) / CL exactly, for every individual. This is a
per-subject identity, so it is a far sharper test than comparing medians. Note
that `F` must be carried explicitly: bioavailability has a typical value of 1
but a 20.7% between-occasion CV, so the individual `fbio` (not 1) is the right
numerator.

```{r identity}
daily_dose <- events |>
  dplyr::filter(evid == 1L, time >= t_ss, time < t_ss + tau_day) |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(daily = sum(amt), .groups = "drop")

# cl and fbio are constant within a subject-arm for this OCC = 2 design
indiv <- sim |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(
    cl = dplyr::first(cl), fbio = dplyr::first(fbio), vc = dplyr::first(vc),
    .groups = "drop"
  ) |>
  dplyr::mutate(t_half = log(2) * vc / cl)

# The identity is only meaningful if every subject actually reached steady
# state, so gate the run-in length explicitly rather than assuming it.
sprintf(
  "Individual t1/2 spans %.1f-%.1f h; the %d h run-in is >= %.1f half-lives for every subject",
  min(indiv$t_half), max(indiv$t_half), t_ss, t_ss / max(indiv$t_half)
)
stopifnot(t_ss / max(indiv$t_half) >= 10)

ident <- nca |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, treatment, auc = PPORRES) |>
  dplyr::inner_join(daily_dose, by = c("id", "treatment")) |>
  dplyr::inner_join(indiv, by = c("id", "treatment")) |>
  dplyr::mutate(
    predicted = fbio * daily / cl,
    pct_err = 100 * (auc - predicted) / predicted
  )

stopifnot(nrow(ident) == 8 * n_per_arm)
ident <- ident |>
  dplyr::mutate(interval = dplyr::case_when(
    grepl("QD", treatment) ~ 24, grepl("TID", treatment) ~ 8, TRUE ~ 12
  ))

# Once-daily arms: exactly one dose per interval, so the identity must hold to
# trapezoidal precision.
qd <- dplyr::filter(ident, interval == 24)
sprintf(
  "Once-daily arms (n = %d): AUC0-24 vs F*Dose/CL, max |error| = %.4f%%",
  nrow(qd), max(abs(qd$pct_err))
)
stopifnot(max(abs(qd$pct_err)) < 0.05)

# Multi-dose arms: the identity holds for the population but not for every
# individual -- see the note below.
md <- dplyr::filter(ident, interval < 24)
sprintf(
  "Multi-dose arms (n = %d): median error %.4f%%; %d subject(s) (%.1f%%) exceed 1%%",
  nrow(md), median(md$pct_err), sum(abs(md$pct_err) > 1), 100 * mean(abs(md$pct_err) > 1)
)
stopifnot(abs(median(md$pct_err)) < 0.01, mean(abs(md$pct_err) > 1) < 0.02)
```

The handful of multi-dose individuals that miss the identity are not a
transcription error. Both NONMEM's `TRANSIT` idiom and rxode2's analytical
`transit()` evaluate the absorption input from **the most recent dose only**,
using the time after that dose. When an individual's mean transit time
approaches the dosing interval, the previous dose's input is still delivering
drug when the next dose resets the clock, and the remainder is lost -- so AUC
falls below F x Dose / CL. With an 83.5% between-occasion CV on MTT, a small
tail of subjects lands there:

```{r transit-truncation}
mtt_i <- sim |> dplyr::group_by(id) |> dplyr::summarise(mtt = dplyr::first(mtt), .groups = "drop")
md_mtt <- dplyr::inner_join(md, mtt_i, by = "id")
sprintf(
  "MTT of the multi-dose subjects missing the identity by >1%%: median %.2f h, vs %.2f h across all multi-dose subjects (typical value 0.610 h)",
  median(md_mtt$mtt[abs(md_mtt$pct_err) > 1]), median(md_mtt$mtt)
)
```

This is a faithful consequence of the published absorption model, not a
deviation from it, and it affects the extreme tail rather than the medians that
Table 3 reports.

## Comparison against the published simulated exposures

Resendiz-Galvan 2025 Table 3 reports the median (IQR) simulated AUC~0-24~ for
each regimen from a 1,800-individual Monte Carlo simulation with the same final
model.

```{r compare}
simulated_long <- nca |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES)

published <- tibble::tribble(
  ~treatment,          ~auclast,
  "250 mg QD",          169.8,
  "500 mg QD",          339.7,
  "250 mg BID",         308.6,
  "250/500 am/pm",      506.9,
  "250 mg TID",         513.4,
  "500 mg BID",         670.9,
  "750 mg BID",        1006.0,
  "WHO weight-banded",  462.0
)
stopifnot(all(published$treatment %in% unique(simulated_long$treatment)))

nlmixr2lib::ncaComparisonTable(
  simulated = simulated_long,
  reference = published,
  by = "treatment",
  units = c(auclast = "mg*h/L"),
  tolerance_pct = 20
) |>
  knitr::kable(caption = "Simulated vs published median AUC0-24 (Resendiz-Galvan 2025 Table 3).")
```

## The 250 mg BID row of Table 3 is internally inconsistent

Seven of the eight regimens reproduce to within a few percent. The 250 mg BID
row does not, and the paper's own table shows why it cannot be right: **500 mg
QD and 250 mg BID deliver the same 500 mg total daily dose**. For a linear model
at steady state, superposition makes AUC~0-24~ depend only on the total daily
dose, so those two rows must be identical. Table 3 reports 339.7 and 308.6
respectively -- a 10% gap the model cannot generate.

Backing out the clearance implied by each published row makes the outlier
unambiguous.

```{r inconsistency}
cl_typ <- exp(ui$theta[["lcl_nonren"]]) + exp(ui$theta[["lcl_renal"]])

published |>
  dplyr::mutate(
    `Total daily dose (mg)` = c(250, 500, 500, 750, 750, 1000, 1500, NA),
    `Implied CL (L/h)` = round(`Total daily dose (mg)` / auclast, 3)
  ) |>
  dplyr::rename("Regimen" = treatment, "Published AUC0-24 (mg*h/L)" = auclast) |>
  knitr::kable(
    caption = sprintf(
      "Clearance implied by each published Table 3 row. The model's typical CL is %.3f L/h.",
      cl_typ
    )
  )
```

Every flat regimen implies a clearance of 1.46-1.49 L/h, matching the model's
typical value of `r sprintf("%.3f", cl_typ)` L/h -- except 250 mg BID, which
implies 1.62 L/h. We therefore treat the Table 3 250 mg BID entry (and the
abstract's "median exposure of 308 mg.h/L after 250 mg twice daily") as a
transcription or reporting error in the source, not a feature of the model. The
model file is left exactly as published; no parameter was adjusted.

Because the arms share random numbers, the corresponding checks on **our**
simulation are exact per subject rather than approximate.

```{r inconsistency-assert}
auc_by <- ident |>
  dplyr::select(id, treatment, auc) |>
  tidyr::pivot_wider(names_from = treatment, values_from = auc)

# 1. Strict dose proportionality between the two once-daily arms.
dp <- 100 * (auc_by[["500 mg QD"]] / (2 * auc_by[["250 mg QD"]]) - 1)
sprintf("Simulated 500 mg QD vs 2 x 250 mg QD, per subject: max |error| = %.4f%%", max(abs(dp)))
stopifnot(max(abs(dp)) < 0.05)

# 2. The property the published table violates: 500 mg QD and 250 mg BID are
#    the same 500 mg/day, so at steady state they must give the same AUC0-24.
eq <- 100 * (auc_by[["250 mg BID"]] / auc_by[["500 mg QD"]] - 1)
sprintf(
  "Simulated 250 mg BID vs 500 mg QD: median difference = %.3f%% (published: %.1f%%)",
  median(eq),
  100 * (published$auclast[published$treatment == "250 mg BID"] /
    published$auclast[published$treatment == "500 mg QD"] - 1)
)
stopifnot(abs(median(eq)) < 0.5)

# The published pair disagrees by ~10%, which the model cannot produce.
ratio <- published$auclast[published$treatment == "250 mg BID"] /
  published$auclast[published$treatment == "500 mg QD"]
stopifnot(abs(ratio - 1) > 0.05)
sprintf("Published 250 mg BID / 500 mg QD AUC ratio = %.3f (must be 1.000)", ratio)
```

# Probability of target attainment

Figure 3 of the source reports PTA against the cohort MIC distribution for the
two hollow-fibre targets: %*T*>MIC >= 30% (bactericidal) and >= 64% (80% of
maximum kill). At steady state the profile repeats every dosing interval, so the
fraction of the 24-hour day above the MIC equals the fraction of a dosing
interval above it.

```{r pta}
mic_ref <- 16 # cohort median MIC, mg/L (Results; 76.5% of isolates >= 16 mg/L)

pta <- sim |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(fT = mean(Cc > mic_ref), .groups = "drop") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `PTA %T>MIC >= 30%` = round(100 * mean(fT >= 0.30)),
    `PTA %T>MIC >= 64%` = round(100 * mean(fT >= 0.64)),
    .groups = "drop"
  )

published_pta <- tibble::tribble(
  ~treatment,          ~`Published >= 30%`, ~`Published >= 64%`,
  "250 mg QD",          6,                  NA,
  "500 mg QD",          50,                 NA,
  "250 mg BID",         50,                 NA,
  "250/500 am/pm",      75,                 64,
  "250 mg TID",         75,                 71,
  "500 mg BID",         93,                 86,
  "750 mg BID",         99,                 96,
  "WHO weight-banded",  68,                 57
)

pta |>
  dplyr::inner_join(published_pta, by = "treatment") |>
  dplyr::rename("Regimen" = treatment) |>
  knitr::kable(
    caption = paste(
      "Simulated vs published probability of target attainment at the cohort",
      "median MIC of 16 mg/L (Resendiz-Galvan 2025 Figure 3 and Results).",
      "Published values for the >= 64% target are reported only for the five",
      "regimens named in the text."
    )
  )
```

```{r fig-pta, fig.cap = "Simulated probability of target attainment against MIC for both hollow-fibre targets. Replicates Figure 3 of Resendiz-Galvan 2025."}
mic_grid <- c(1, 2, 4, 8, 16, 32, 64)

pta_curve <- lapply(mic_grid, function(m) {
  sim |>
    dplyr::group_by(treatment, id) |>
    dplyr::summarise(fT = mean(Cc > m), .groups = "drop") |>
    dplyr::group_by(treatment) |>
    dplyr::summarise(
      MIC = m,
      `>= 30% (bactericidal)` = 100 * mean(fT >= 0.30),
      `>= 64% (80% max kill)` = 100 * mean(fT >= 0.64),
      .groups = "drop"
    )
}) |>
  dplyr::bind_rows() |>
  tidyr::pivot_longer(dplyr::starts_with(">="), names_to = "Target", values_to = "PTA")

ggplot2::ggplot(pta_curve, ggplot2::aes(MIC, PTA, colour = treatment)) +
  ggplot2::geom_line() +
  ggplot2::geom_point(size = 0.8) +
  ggplot2::geom_hline(yintercept = 90, linetype = "dashed", colour = "firebrick") +
  ggplot2::geom_vline(xintercept = mic_ref, linetype = "dotted") +
  ggplot2::scale_x_continuous(trans = "log2", breaks = mic_grid) +
  ggplot2::facet_wrap(~Target) +
  ggplot2::labs(x = "MIC (mg/L)", y = "Probability of target attainment (%)", colour = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

The simulation reproduces the paper's central conclusion: at the cohort median
MIC of 16 mg/L only 500 mg BID and 750 mg BID reach high attainment of the
preferred %*T*>MIC >= 64% target, while 250 mg QD is inadequate at any target.

# Observed exposures reported in the source

For context, the paper's *observed* (not simulated) intensive-sampling
exposures, which the model was fit to.

```{r observed, echo = FALSE}
tibble::tribble(
  ~Regimen,          ~`Observed Cmax (mg/L)`,          ~`Observed AUC0-24 (mg*h/L)`,
  "250 mg BID",      "16.8 (range 6.21-26.8)",         "347.4 (IQR 261.1-425.7)",
  "250/500 am/pm",   "30.5 (range 27.3-33.7)",         "415.7 (IQR 224.1-553.9)"
) |>
  knitr::kable(caption = "Resendiz-Galvan 2025 Results, median observed exposures. Median observed tmax was 2 h (IQR 1.2-2.2).")
```

Median t~max~ is computed here directly from the profile over the **first**
12-hour interval of the steady-state day in the 250 mg BID arm, i.e. time after
a single observed morning dose -- which is what the paper's intensive sampling
(pre-dose, 1, 2, 4, 6, 8 h) measured. (PKNCA's interval-level `tmax` spans the
whole 24 h and can land on the evening peak, so it is not the right quantity
here.)

```{r tmax-check}
tmax_sim <- sim |>
  dplyr::filter(treatment == "250 mg BID", time <= t_ss + 12) |>
  dplyr::group_by(id) |>
  dplyr::slice_max(Cc, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(tmax = time - t_ss)

sprintf(
  "Simulated tmax after the morning dose: median %.2f h (IQR %.2f-%.2f); paper: 2 h (IQR 1.2-2.2)",
  median(tmax_sim$tmax), quantile(tmax_sim$tmax, 0.25), quantile(tmax_sim$tmax, 0.75)
)
stopifnot(median(tmax_sim$tmax) > 1, median(tmax_sim$tmax) < 4)
```

# Assumptions and deviations

* **Allometric exponents are not printed in the source.** The paper states that
  allometric scaling on FFM was applied to the disposition parameters and cites
  Holford for the renal/non-renal separation, but Table 2 has no exponent row
  and the exponents appear nowhere in the text or supplement, i.e. they were not
  estimated. They are encoded as the standard theory-based values
  (`fixed(0.75)` on clearance, `fixed(1)` on volume). Every AUC in Table 3 is
  reproduced to within ~2% under this reading, which corroborates it.
* **Table 1 mislabels the serum-creatinine unit as mg/L.** The printed
  Cockcroft-Gault equation reads `72 . SCr (mg/dL)`, and the arithmetic settles
  it: the median individual (age 27, SCr 0.7 mg/dL) gives
  CLcr,56M = 125.6 mL/min, consistent with the stated median of 122 mL/min,
  whereas mg/L would give ~1,256 mL/min. The model uses mg/dL.
* **The clearance equation lost its parentheses in typesetting.** The text
  version reads `CL = CLnr + RF . CLr . Fsize`, which would apply allometry to
  the renal arm only. The typeset equation image supplied with the article
  (`aac.00101-25.m003`) shows `CL = (CLnr + RF . CLr) . Fsize`, matching the
  prose ("the effect of body size Fsize ... is applied to the overall CL
  (CLr + CLnr)") and Table 2 footnote a ("All disposition parameters were
  allometrically scaled"). The parenthesised form is implemented.
* **%CV convention.** Table 2 footnote f defines %CV = sqrt(omega^2) x 100, so
  the tabulated variability numbers are 100 x omega on the log scale. Variances
  are therefore encoded as `(CV/100)^2`, **not** `log((CV/100)^2 + 1)`.
* **The 250 mg BID row of Table 3 is inconsistent with the paper's own model**
  (see the dedicated section above). No parameter was tuned to match it.
* **Between-occasion variability is encoded over two occasion types**
  (`OCC = 1` unobserved home dose, `OCC = 2` observed clinic dose), which is the
  distinction the 2.26-fold inflation factor is defined against. Whether that
  factor scales the BOV standard deviation or its variance is not stated; the
  SD reading is implemented, following the NONMEM idiom and the registry
  precedent in `Wojciechowski_2023_ritlecitinib_*.R`. This vignette simulates
  fully-observed regimens (`OCC = 2`), where the factor is inactive and the
  ambiguity has no effect.
* **`useLinCmt = FALSE` is mandatory for this model.** It is a one-compartment
  oral model, so rxode2's automatic ODE-to-`linCmt()` conversion fires. The
  converted form drops the `transit()` input term, and because the model sets
  `f(depot) <- 0` (the dose enters only via the transit chain) the effective
  dose becomes zero and every concentration is zero -- silently, with no error
  or warning. The simulation chunk asserts `max(Cc) > 1` to make that failure
  mode loud.
* **Cohort covariate distributions are independent marginals.** Weight, height,
  age, and serum creatinine are drawn at evenly-spaced quantiles of log-normals
  matched to their Table 1 medians and IQRs, then permuted independently; the
  paper reports no correlation structure between them. Stratifying rather than
  sampling at random is deliberate: the Table 1 marginals are a specification to
  reproduce, and at n = 200 random draws shift the cohort FFM median by ~3.6%,
  which alone biases every exposure by ~5%. The reproduced medians now match
  Table 1 to the reported precision. The derived FFM **IQR** is still modestly
  narrower than Table 1's (34.3-45.4 vs 32.3-47.1 kg) because weight and height
  are independent here while in reality they correlate, so the simulated
  exposure spread is a little tighter than the published spread.
* **A residual ~2-5% shortfall in simulated AUC remains** across all regimens
  relative to Table 3. The paper's virtual population was built from
  "repetitions of the original data set", i.e. the real subjects' joint
  covariate vectors, whereas this cohort samples the marginals independently.
  That is the most likely source; no parameter was adjusted to close the gap.
* **Residual-error inflation for imputed records is not carried.** The source
  inflated the additive error by LOD/2 for the three below-limit-of-detection
  records it imputed under an M6 adaptation (0.2% of observations). That
  record-specific adjustment is not representable as a model-level parameter and
  is omitted.
* **Erratum.** The article was published 2 September 2025 and reposted
  1 October 2025 correcting the author-contributions statement only. No model
  parameter was revised; the version on disk is the corrected one.
```
