---
title: "Emicizumab (Retout 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Emicizumab (Retout 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Retout_2020_emicizumab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Retout_2020_emicizumab"))$description`
* Article: [doi:10.1007/s40262-020-00904-z](https://doi.org/10.1007/s40262-020-00904-z)
  (open access, CC BY-NC 4.0)
* Trial registries: HAVEN 1
  [NCT02622321](https://clinicaltrials.gov/study/NCT02622321), HAVEN 2
  [NCT02795767](https://clinicaltrials.gov/study/NCT02795767), HAVEN 3
  [NCT02847637](https://clinicaltrials.gov/study/NCT02847637), HAVEN 4
  [NCT03020160](https://clinicaltrials.gov/study/NCT03020160), Japanese
  phase I/II JapicCTI-121934 / JapicCTI-132195.

Emicizumab is a humanized bispecific IgG4 antibody that bridges activated
factor IX and factor X, replacing the cofactor function of the missing
activated factor VIII in persons with hemophilia A (PwHA). Retout 2020 pooled
the phase I/II study and the four phase III HAVEN studies into one
population PK analysis. The final model is a linear one-compartment model with
first-order subcutaneous absorption and elimination and no lag time. Its
covariates are body weight and albumin on CL/F, body weight and Black race on
V/F, and an age-dependent apparent bioavailability F that declines linearly
above 30 years.

The earlier analysis of the phase I/II data alone is packaged separately as
`Yoneyama_2017_emicizumab`. Retout 2020's exposure-bleeding analysis
(Section 2.4, Figure 6) is a descriptive comparison of estimated average
concentrations across annualized-bleeding-rate categories. It fits no model,
so nothing from it is packaged here.

# Population

The analysis population was 389 PwHA with or without factor VIII (FVIII)
inhibitors from five studies (Retout 2020 Tables 1-3): the Japanese phase
I/II study (n = 18), HAVEN 1 (n = 112), HAVEN 2 (pediatric, n = 63), HAVEN 3
(n = 148) and HAVEN 4 (n = 48). Age ranged from 1.22 to 77.0 years (median
30.0) and body weight from 9.50 to 156 kg (median 69.1). Median albumin was
45.0 g/L. The population was 62.7% White, 8.0% Black, 22.9% Asian (including
Japanese) and 6.4% other or unknown, and 49.9% had FVIII inhibitors. Most
PwHA received 3 mg/kg QW for 4 weeks followed by 1.5 mg/kg QW (75.1%); 12.6%
received the 3 mg/kg Q2W maintenance and 12.3% the 6 mg/kg Q4W maintenance.
The paper does not tabulate sex. Hemophilia A is X-linked, so the cohort is
expected to be male.

The same information is available programmatically via
`readModelDb("Retout_2020_emicizumab")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
table below collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | log(0.536) 1/day | Table 4, KA |
| `lcl` | log(0.272) L/day | Table 4, CL/F (BW 70 kg, ALB 45 g/L, age <= 30 years) |
| `lvc` | log(10.4) L | Table 4, V/F |
| `e_wt_cl` | 0.911 | Table 4, 'Effect of BW on CL/F' |
| `e_alb_cl` | 0.0157 | Table 4, 'Effect of ALB on CL/F'; sign from the Section 3.1 CL/F equation |
| `e_wt_vc` | 1.00 | Table 4, 'Effect of BW on V/F' |
| `e_race_black_vc` | -0.215 | Table 4, 'Effect of Black on V/F' |
| `e_age_fdepot` | 0.00651 | Table 4, 'Effect of AGE>30 years on F'; sign from the Section 3.1 F equation |
| `etalcl`, `etalvc`, `etalka` | 0.079152, 0.064927, 0.422400 | Table 4, BPV 28.7%, 25.9%, 72.5% CV; `omega^2 = log(1 + CV^2)` |
| cov(`etalcl`, `etalvc`) | 0.015556 | Table 4, correlation CL/F-V/F 0.217 |
| cov(`etalcl`, `etalka`) | -0.062352 | Table 4, correlation CL/F-KA -0.341 |
| cov(`etalvc`, `etalka`) | 0 | Not reported in Table 4 |
| `addSd` | 0.025 ug/mL (fixed) | Table 4, sigma1 additive; Section 3.1 (half the phase I/II LLOQ) |
| `propSd` | 0.146 | Table 4, sigma2 proportional 14.6% |
| `cl`, `vc`, `fdepot` equations | n/a | Section 3.1 covariate equations |
| One-compartment ODE, first-order absorption, no lag | n/a | Section 3.1, first paragraph |

# Closed-form checks against the paper's reference patient

Retout 2020 reports steady-state predictions for a reference PwHA who is White
or Asian, 30 years old, 70 kg and has albumin 45 g/L, dosed at 1.5 mg/kg QW
(Section 3.2). For a linear model, `Cav,SS = F * Dose / (CL/F * tau)`
exactly, so each of the paper's statements is a direct test of the packaged
parameters and equation signs.

```{r closed-form}
mod <- readModelDb("Retout_2020_emicizumab")
ui <- rxode2::rxode(mod)
th <- setNames(ui$iniDf$est, ui$iniDf$name)

typ_cl <- function(WT, ALB) {
  exp(th[["lcl"]]) * (WT / 70)^th[["e_wt_cl"]] * (1 - th[["e_alb_cl"]] * (ALB - 45))
}
typ_f <- function(AGE) {
  1 - th[["e_age_fdepot"]] * (AGE > 30) * (AGE - 30)
}
typ_cav <- function(WT, ALB, AGE, mgkg = 1.5, tau = 7) {
  typ_f(AGE) * mgkg * WT / (typ_cl(WT, ALB) * tau)
}

cav_ref <- typ_cav(70, 45, 30)
closed <- tibble::tribble(
  ~Claim, ~Paper, ~Packaged,
  "Cav,SS reference PwHA (ug/mL)", 55.1, cav_ref,
  "Cav,SS at ALB 33 g/L (ug/mL)", 46.4, typ_cav(70, 33, 30),
  "Cav,SS at ALB 57 g/L (ug/mL)", 68.0, typ_cav(70, 57, 30),
  "Exposure change at 77 years (%)", -31, 100 * (typ_cav(70, 45, 77) / cav_ref - 1),
  "CL/F change at 9 kg (%)", -85, 100 * ((9 / 70)^th[["e_wt_cl"]] - 1),
  "CL/F change at 156 kg (%)", 108, 100 * ((156 / 70)^th[["e_wt_cl"]] - 1),
  "V/F change at 9 kg (%)", -87, 100 * ((9 / 70)^th[["e_wt_vc"]] - 1),
  "V/F change at 156 kg (%)", 123, 100 * ((156 / 70)^th[["e_wt_vc"]] - 1),
  "V/F change, Black race (%)", -22, 100 * th[["e_race_black_vc"]],
  "Typical absorption half-life (day)", 1.27, log(2) / exp(th[["lka"]])
)
knitr::kable(closed, digits = 2, caption = "Reference-patient statements in Retout 2020 (Sections 3.1, 3.2 and 4, Figure 2 caption) against the packaged model. The half-life row compares the typical value with the Table 5 median.")

# All ten are deterministic functions of the packaged parameters. Each
# published value is printed to 2-3 significant figures, so agreement to
# within the printed rounding is the test. A sign error on the albumin or
# age term moves the corresponding row by 20-40%.
stopifnot(
  abs(cav_ref - 55.1) < 0.1,
  abs(typ_cav(70, 33, 30) - 46.4) < 0.1,
  abs(typ_cav(70, 57, 30) - 68.0) < 0.1,
  abs(100 * (typ_cav(70, 45, 77) / cav_ref - 1) - (-31)) < 0.5,
  abs(100 * ((9 / 70)^th[["e_wt_cl"]] - 1) - (-85)) < 0.5,
  abs(100 * ((156 / 70)^th[["e_wt_cl"]] - 1) - 108) < 0.5,
  abs(100 * ((9 / 70)^th[["e_wt_vc"]] - 1) - (-87)) < 0.5,
  abs(100 * ((156 / 70)^th[["e_wt_vc"]] - 1) - 123) < 0.5,
  abs(log(2) / exp(th[["lka"]]) - 1.27) < 0.05
)
```

The albumin rows show which sign is right. Table 4 prints the albumin
coefficient as `+1.57e-2`. Under the Methods' generic form
`(1 + theta * (ALB - 45))`, that would give 68.0 ug/mL at 33 g/L and
46.4 ug/mL at 57 g/L, the reverse of what the paper reports. The Section 3.1
equation `(1 - 0.0157 * (ALB - 45))` reproduces both published values, so
clearance falls as albumin rises. The age coefficient works the same way:
Table 4 prints `+6.51e-3`, and the equation's minus sign gives the published
31% exposure loss at 77 years.

# Replicate Figure 4: covariate impact at steady state

Figure 4 of Retout 2020 shows typical-value steady-state profiles over one
1.5 mg/kg QW dosing interval for 12 covariate combinations. The paper
simulated steady state directly. Here the regimen is run for two years
(3 mg/kg QW loading for 4 weeks, then 1.5 mg/kg QW) and the final interval
is plotted. At the typical 26.5-day half-life, the remaining gap from true
steady state is below 1e-8.

```{r fig4-events}
# One subject per row of `subj`. The regimen is 3 mg/kg QW x 4 loading, then
# maintenance `mgkg` every `tau` days up to day 756; observations every 6 h
# over the final interval [756 - tau, 756] (6 h is the paper's step, Section 2.3).
t_end <- 28 + 728

make_events <- function(subj, mgkg, tau, obs_times) {
  maint <- seq(28, t_end - tau, by = tau)
  dose <- tidyr::crossing(
    subj,
    tibble::tibble(time = c(0, 7, 14, 21, maint), mgkg_dose = c(rep(3, 4), rep(mgkg, length(maint))))
  ) |>
    dplyr::mutate(amt = mgkg_dose * WT, evid = 1L, cmt = "depot") |>
    dplyr::select(-mgkg_dose)
  obs <- tidyr::crossing(subj, tibble::tibble(time = obs_times)) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = "central")
  dplyr::bind_rows(dose, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

fig4_subj <- tibble::tribble(
  ~panel, ~label, ~WT, ~AGE, ~ALB, ~RACE_BLACK,
  "Body weight", "31 kg - 8 y", 31, 8, 45, 0,
  "Body weight", "70 kg - 30 y", 70, 30, 45, 0,
  "Body weight", "156 kg - 30 y", 156, 30, 45, 0,
  "Age", "6 y - 20 kg", 20, 6, 45, 0,
  "Age", "12 y - 40 kg", 40, 12, 45, 0,
  "Age", "30 y - 70 kg", 70, 30, 45, 0,
  "Age", "50 y - 70 kg", 70, 50, 45, 0,
  "Age", "77 y - 70 kg", 70, 77, 45, 0,
  "Race", "White or Asian", 70, 30, 45, 0,
  "Race", "Black", 70, 30, 45, 1,
  "Albumin", "ALB = 33 g/L", 70, 30, 33, 0,
  "Albumin", "ALB = 45 g/L", 70, 30, 45, 0,
  "Albumin", "ALB = 57 g/L", 70, 30, 57, 0
) |>
  dplyr::mutate(id = dplyr::row_number())

ev_fig4 <- make_events(
  dplyr::select(fig4_subj, id, WT, AGE, ALB, RACE_BLACK),
  mgkg = 1.5, tau = 7, obs_times = seq(t_end - 7, t_end, by = 0.25)
)
```

```{r fig4-simulate}
mod_typ <- rxode2::zeroRe(mod)

# Every solve below is on the zero-variance model. Where per-subject etas are
# supplied as event-table columns, rxode2 warns that a multi-subject solve
# has no omega, which is intended here. Muffle that one warning only.
muffle_no_omega <- function(w) {
  if (grepl("without 'omega'", conditionMessage(w), fixed = TRUE)) {
    invokeRestart("muffleWarning")
  }
}
solve_typ <- function(events, ...) {
  withCallingHandlers(
    rxode2::rxSolve(mod_typ, events = events, ...),
    warning = muffle_no_omega
  )
}

sim_fig4 <- solve_typ(ev_fig4, rtol = 1e-10, atol = 1e-12) |>
  as.data.frame() |>
  dplyr::left_join(dplyr::select(fig4_subj, id, panel, label), by = "id") |>
  dplyr::mutate(
    tad = time - (t_end - 7),
    label = factor(label, levels = unique(fig4_subj$label))
  )
```

```{r fig4-plot, fig.height = 6}
ggplot(sim_fig4, aes(tad, Cc)) +
  geom_line(colour = "red", linewidth = 1) +
  facet_wrap(~ panel + label, ncol = 5) +
  coord_cartesian(ylim = c(0, 75)) +
  labs(
    x = "Time over a dosing interval at steady state (day)",
    y = "Predicted emicizumab concentration (ug/mL)",
    caption = "Replicates Figure 4 of Retout 2020 (1.5 mg/kg QW, typical values)."
  ) +
  theme_bw()
```

```{r fig4-check}
fig4_sum <- sim_fig4 |>
  dplyr::group_by(id, label) |>
  dplyr::summarise(Cmax = max(Cc), Ctrough = min(Cc), .groups = "drop")
knitr::kable(fig4_sum, digits = 1, caption = "Typical steady-state Cmax and Ctrough per Figure 4 scenario.")

race <- fig4_sum |> dplyr::filter(label %in% c("White or Asian", "Black"))
race_trough <- 100 * (race$Ctrough[race$label == "Black"] / race$Ctrough[race$label == "White or Asian"] - 1)
race_peak <- 100 * (race$Cmax[race$label == "Black"] / race$Cmax[race$label == "White or Asian"] - 1)
c(race_trough_pct = race_trough, race_peak_pct = race_peak)

# Section 3.2: in Black PwHA, Ctrough,SS falls by less than 2% and Cmax,SS
# rises by less than 1% relative to the White/Asian reference. Both are
# typical-value solves (no random draws); measured -1.32% and +0.70%. The
# lower bounds catch a race effect that is too small: -0.0215 gives -0.13%
# and +0.07%, which the paper's "< 2%" / "< 1%" alone would accept.
stopifnot(race_trough < -1, race_trough > -2, race_peak > 0.5, race_peak < 1)
```

The simulated profiles match Figure 4 as read by eye. The weight panels peak
at 52.7, 56.6 and 60.7 ug/mL (Figure 4: about 52.5, 56.5 and 60.5); the
77-year panel peaks at 39.3 ug/mL (about 39); and the 33 / 57 g/L albumin
panels peak at 47.8 and 69.4 ug/mL (about 48 and 69.5).

# Virtual cohort for Table 5

Table 5 of Retout 2020 summarises steady-state exposure from the individual
PK parameters of 381 PwHA under each maintenance regimen. The individual data
are not public. The virtual cohort below approximates the analysis population
from Tables 2-3: 16% pediatric (HAVEN 2: age 1.2-15.7 years, median weight
22.6 kg) and 84% adolescent/adult (median age about 33 years, median weight
about 74 kg). Albumin is centred on 45 g/L and 8% of the cohort is Black.

Between-subject random effects are drawn in R from the packaged `omega`
matrix and passed to a `zeroRe()` model as per-subject columns. The same 200
individuals (same covariates and random effects) therefore receive each of
the three regimens, as in Table 5. This reproduces the paper's result that
`Cav,SS` is identical across regimens. The cohort is also identical on every
machine, because no draw depends on rxode2's thread-partitioned RNG.

```{r cohort}
set.seed(20200605)

# Draw from a lognormal, rejecting and redrawing outside [lo, hi] so the band
# limits are definitions, not clamps that pile subjects on the boundary.
rtrunc_lnorm <- function(n, median, sdlog, lo, hi) {
  x <- stats::rlnorm(n, log(median), sdlog)
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- stats::rlnorm(sum(bad), log(median), sdlog)
    bad <- x < lo | x > hi
  }
  x
}
rtrunc_norm <- function(n, mean, sd, lo, hi) {
  x <- stats::rnorm(n, mean, sd)
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- stats::rnorm(sum(bad), mean, sd)
    bad <- x < lo | x > hi
  }
  x
}

# Approximate median weight-for-age (10 kg at 1 year, 23 kg at 7 years,
# 46 kg at 12 years) with 15% lognormal spread, redrawn into the HAVEN 2
# weight range 9.5-63 kg.
rped_wt <- function(age) {
  med <- 10 * exp(0.139 * (age - 1))
  x <- med * exp(stats::rnorm(length(age), 0, 0.15))
  bad <- x < 9.5 | x > 63
  while (any(bad)) {
    x[bad] <- med[bad] * exp(stats::rnorm(sum(bad), 0, 0.15))
    bad <- x < 9.5 | x > 63
  }
  x
}

make_cohort <- function(n_ped, n_adult, id_offset = 0L) {
  age_ped <- stats::runif(n_ped, 1, 12)
  wt_ped <- rped_wt(age_ped)
  age_ad <- rtrunc_lnorm(n_adult, 33, 0.40, 12, 77)
  wt_ad <- rtrunc_lnorm(n_adult, 74, 0.20, 40, 156)
  n <- n_ped + n_adult
  tibble::tibble(
    id = id_offset + seq_len(n),
    age_group = rep(c("1 to <12 years", ">=12 years"), c(n_ped, n_adult)),
    AGE = c(age_ped, age_ad),
    WT = c(wt_ped, wt_ad),
    ALB = rtrunc_norm(n, 45, 3.5, 33, 57),
    RACE_BLACK = stats::rbinom(n, 1, 0.08)
  )
}

# Random effects from the packaged omega, via its Cholesky factor.
draw_etas <- function(n, omega) {
  z <- matrix(stats::rnorm(n * ncol(omega)), n, ncol(omega))
  eta <- z %*% chol(omega)
  colnames(eta) <- colnames(omega)
  tibble::as_tibble(eta)
}

cohort_t5 <- make_cohort(n_ped = 32, n_adult = 168)
cohort_t5 <- dplyr::bind_cols(cohort_t5, draw_etas(nrow(cohort_t5), ui$omega))
stopifnot(!anyDuplicated(cohort_t5$id))
summary(cohort_t5[, c("AGE", "WT", "ALB", "RACE_BLACK")])
```

```{r table5-simulate}
regimens <- tibble::tribble(
  ~regimen, ~mgkg, ~tau,
  "1.5 mg/kg QW", 1.5, 7,
  "3 mg/kg Q2W", 3, 14,
  "6 mg/kg Q4W", 6, 28
)
subj_cols <- c("id", "WT", "AGE", "ALB", "RACE_BLACK", "etalcl", "etalvc", "etalka")

sim_t5 <- lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  ev <- make_events(
    cohort_t5[, subj_cols],
    mgkg = r$mgkg, tau = r$tau,
    obs_times = seq(t_end - r$tau, t_end, by = 0.25)
  )
  solve_typ(ev, rtol = 1e-8, atol = 1e-10, maxsteps = 1e6) |>
    as.data.frame() |>
    dplyr::mutate(regimen = r$regimen, tau = r$tau)
}) |>
  dplyr::bind_rows()

# The explicit etas must have reached the solve.
chk_eta <- sim_t5 |>
  dplyr::distinct(id, cl) |>
  dplyr::left_join(cohort_t5, by = "id")
stopifnot(
  nrow(chk_eta) == nrow(cohort_t5),
  max(abs(chk_eta$cl / (typ_cl(chk_eta$WT, chk_eta$ALB) * exp(chk_eta$etalcl)) - 1)) < 1e-8
)
```

# PKNCA validation against Table 5

The steady-state interval is the final dosing interval of each regimen.
`cmin` over the interval is the paper's `Ctrough,SS` (defined in Section 2.3
as the minimum simulated value).

```{r pknca}
conc_t5 <- sim_t5 |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, regimen, time, Cc)

dose_t5 <- lapply(seq_len(nrow(regimens)), function(i) {
  make_events(
    cohort_t5[, subj_cols],
    mgkg = regimens$mgkg[i], tau = regimens$tau[i], obs_times = numeric(0)
  ) |>
    dplyr::mutate(regimen = regimens$regimen[i])
}) |>
  dplyr::bind_rows() |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, regimen, time, amt)

conc_obj <- PKNCA::PKNCAconc(conc_t5, Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(dose_t5, amt ~ time | regimen + id)
intervals <- regimens |>
  dplyr::transmute(
    regimen,
    start = t_end - tau, end = t_end,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
  ) |>
  as.data.frame()
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
published_t5 <- tibble::tribble(
  ~regimen, ~cmax, ~tmax, ~cmin, ~auclast, ~cav,
  "1.5 mg/kg QW", 53.9, 2.50, 49.9, 366, 52.3,
  "3 mg/kg Q2W", 57.0, 3.50, 45.6, 733, 52.3,
  "6 mg/kg Q4W", 65.9, 4.50, 36.5, 1465, 52.3
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_t5,
  by = "regimen",
  units = c(cmax = "ug/mL", tmax = "day", cmin = "ug/mL", auclast = "ug*day/mL", cav = "ug/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Median steady-state exposure: virtual cohort (PKNCA) vs. Retout 2020 Table 5 medians. AUC is over one dosing interval tau. * differs from the published value by >20%."
)
```

```{r nca-assert}
res_t5 <- as.data.frame(nca_res$result) |>
  dplyr::select(id, regimen, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

med_t5 <- res_t5 |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(dplyr::across(c(cmax, cmin, auclast, cav), stats::median), .groups = "drop") |>
  dplyr::left_join(published_t5, by = "regimen", suffix = c("_sim", "_pub"))
pct <- with(med_t5, 100 * cbind(
  cmax = cmax_sim / cmax_pub - 1, cmin = cmin_sim / cmin_pub - 1,
  auclast = auclast_sim / auclast_pub - 1, cav = cav_sim / cav_pub - 1
))
pct

# Structural gate on the cohort median. Measured +3.3% to +4.3% on every
# metric. The cohort only approximates the 381 analysis subjects, so the
# bound is 10%; a mis-transcribed CL/F, V/F, dose or unit moves every median
# by tens of percent. The cohort is drawn with R's RNG only, so these values
# are the same on every machine.
stopifnot(nrow(med_t5) == 3L, all(abs(pct) < 10))

# Exact identity, as in Table 5: for a linear model with the same individuals,
# Cav,SS = F * (mg/kg * WT) / (CL/F * tau) does not depend on the regimen.
# The mg/kg-per-day dose rate is the same for all three regimens, so each
# subject's Cav must agree with the closed form to within trapezoid error
# and the small remaining approach to steady state (measured 4e-4).
cav_check <- res_t5 |>
  dplyr::left_join(regimens, by = "regimen") |>
  dplyr::left_join(chk_eta[, c("id", "cl", "AGE", "WT")], by = "id") |>
  dplyr::mutate(cav_closed = typ_f(AGE) * mgkg * WT / (cl * tau))
cav_err <- max(abs(cav_check$cav / cav_check$cav_closed - 1))
cav_err
stopifnot(nrow(cav_check) == 3L * nrow(cohort_t5), cav_err < 5e-3)
```

Every simulated median is 3-4% above Table 5, and the offset is the same for
every metric and regimen. Tmax and the regimen-dependent peak and trough
pattern match. A uniform offset points to the covariate distribution rather
than the model structure. The published Cav,SS median (52.3 ug/mL) sits 5%
below the 55.1 ug/mL reference patient, while the virtual cohort's (54.2)
sits only 2% below. The real population therefore had more exposure-lowering
covariates (age over 30, low albumin, low weight) than the virtual one.

Table 5 also reports elimination and absorption half-lives. These are
derived directly from each subject's parameters.

```{r secondary}
sec <- chk_eta |>
  dplyr::left_join(
    sim_t5 |> dplyr::distinct(id, vc, ka),
    by = "id"
  ) |>
  dplyr::summarise(
    t_half = stats::median(log(2) * vc / cl),
    t_half_abs = stats::median(log(2) / ka)
  )
tibble::tibble(
  Parameter = c("t1/2 (day)", "t1/2,abs (day)"),
  Published_median = c(25.1, 1.27),
  Simulated_median = c(sec$t_half, sec$t_half_abs)
) |>
  knitr::kable(digits = 2, caption = "Secondary PK parameters (Table 5 medians).")
```

The Table 5 medians come from empirical Bayes estimates, which are shrunk
toward the typical value; Table 4 reports 40.6% shrinkage for KA. The spread
of these secondary parameters in the published table is therefore narrower
than in a simulation from the full `omega`. The medians are unaffected.

# Replicate Figure 5: pediatric vs. adolescent/adult profiles

Figure 5 of Retout 2020 shows the median and 90% prediction interval over
the first 6 months for PwHA aged 1 to < 12 years and >= 12 years under each
maintenance regimen (all after a 3 mg/kg QW loading for 4 weeks). The paper
reports negligible differences between the age groups.

```{r fig5-simulate}
set.seed(5)
cohort_f5 <- make_cohort(n_ped = 100, n_adult = 100)
cohort_f5 <- dplyr::bind_cols(cohort_f5, draw_etas(nrow(cohort_f5), ui$omega))

sim_f5 <- lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  ev <- make_events(
    cohort_f5[, subj_cols],
    mgkg = r$mgkg, tau = r$tau, obs_times = seq(0, 182, by = 1)
  ) |>
    dplyr::filter(time <= 182)
  solve_typ(ev, maxsteps = 1e6) |>
    as.data.frame() |>
    dplyr::mutate(regimen = r$regimen)
}) |>
  dplyr::bind_rows() |>
  dplyr::left_join(cohort_f5[, c("id", "age_group")], by = "id")

f5_sum <- sim_f5 |>
  dplyr::group_by(regimen, age_group, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  )
```

```{r fig5-plot, fig.height = 5.5}
ggplot(f5_sum, aes(time / 7, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "blue") +
  facet_grid(age_group ~ regimen) +
  labs(
    x = "Time (weeks)",
    y = "Emicizumab concentration (ug/mL)",
    caption = "Replicates Figure 5 of Retout 2020: median and 90% prediction interval."
  ) +
  theme_bw()
```

```{r fig5-check}
# Negligible age-group difference: median concentration over weeks 5-26,
# pediatric relative to >= 12 years, per regimen. Measured -2.7% to -3.5%:
# weight-based dosing with a CL/F exponent of 0.911 leaves children slightly
# lower, and the F decline above 30 years pulls the adult group partly back.
# The same cohort is drawn on every machine (R's RNG only). With the
# canonical 0.75 exponent instead, children sit 17-18% lower and this
# check fails.
age_diff <- sim_f5 |>
  dplyr::filter(time >= 28) |>
  dplyr::group_by(regimen, age_group) |>
  dplyr::summarise(med = stats::median(Cc), .groups = "drop") |>
  tidyr::pivot_wider(names_from = age_group, values_from = med) |>
  dplyr::mutate(pct = 100 * (`1 to <12 years` / `>=12 years` - 1))
knitr::kable(age_diff, digits = 1, caption = "Median concentration after loading, by age group.")
stopifnot(nrow(age_diff) == 3L, all(abs(age_diff$pct) < 10))
```

# Assumptions and deviations

- **Correlation between V/F and KA.** Table 4 reports correlations for
  CL/F-V/F (0.217) and CL/F-KA (-0.341) only. The V/F-KA covariance is set to
  0; the resulting 3 x 3 `omega` is positive definite.
- **CV to variance.** Table 4 reports between-person variability as CV% under
  an exponential model. It is converted with `omega^2 = log(1 + CV^2)`. The
  choice matters only for KA (72.5% CV: 0.422 vs 0.526 with the
  `omega^2 = CV^2` approximation). CL/F and V/F change by less than 5%.
- **Signs of the albumin and age effects.** Table 4 prints both coefficients
  as positive magnitudes. The model follows the Section 3.1 equations, which
  carry a minus sign. The paper's own Cav,SS predictions (46.4 and
  68.0 ug/mL at 33 and 57 g/L albumin, -31% at 77 years) confirm this; see
  the closed-form checks above. The Discussion's phrase "a positive
  correlation between ALB and CL/F" contradicts both the equation and those
  predictions, and is not followed.
- **Residual error.** Table 4 footnote d specifies a combined
  additive-plus-proportional model without the NONMEM coding. It is encoded
  as nlmixr2's `add(addSd) + prop(propSd)`, with the additive part fixed at
  0.025 ug/mL as published.
- **Age as a baseline covariate.** The paper tests covariates at baseline.
  AGE, WT and ALB are therefore treated as time-fixed per subject.
- **Virtual cohort.** Individual covariates are not public. Age, weight,
  albumin and race are drawn from distributions chosen to match the medians
  and ranges in Tables 2-3; the pediatric weight-for-age curve is an
  approximation. The paper resampled the observed covariates instead.
- **Exposure-bleeding analysis.** The annualized-bleeding-rate comparison
  (Section 2.4, Figure 6) is descriptive and has no model to package.
- **Errata.** No correction notice for Retout 2020 was found in Crossref or
  Europe PMC as of 2026-09-27.
