---
title: "Automated popPK model search: ribociclib, camizestrant, osimertinib, olaparib, tezepelumab (Richardson 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated popPK model search: ribociclib, camizestrant, osimertinib, olaparib, tezepelumab (Richardson 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Richardson et al. (2025) do not develop a single drug model. They develop an
*automated model-search procedure* -- a generic pyDarwin search space of >12,000
extravascular popPK structures plus a penalty function -- and then run it against
five data sets, reporting the top-ranked structure and its full parameter table
for each. Those five top-ranked structures are the models packaged here, one
`.R` file per drug:

```{r models}
richardsonModels <- c(
  "Richardson_2025_ribociclib",
  "Richardson_2025_camizestrant",
  "Richardson_2025_osimertinib",
  "Richardson_2025_olaparib",
  "Richardson_2025_tezepelumab"
)
```

* Article: <https://doi.org/10.1038/s43856-025-01054-8>
* Supplementary information (parameter tables 2-6): <https://doi.org/10.1038/s43856-025-01054-8>
* Code availability (pyDarwin model space, and the ribociclib final NONMEM
  output): <https://github.com/samjrrr/autopk_synthetic_example>

**These are machine-search structures, not the manually developed expert
models.** For every data set the paper also names the corresponding
expert-developed publication, and the automated structure differs from it in at
least one feature for four of the five drugs (Table 4). A user who wants "the"
published popPK model for one of these drugs should go to the expert reference
cited in each model file, not to these files. What these files reproduce is the
paper's own reported result.

```{r citation, results = "asis"}
for (nm in richardsonModels) {
  m <- rxode2::rxode(readModelDb(nm))
  cat("\n## ", nm, "\n\n", sep = "")
  cat("* Citation: ", m$reference, "\n", sep = "")
  cat("* Description: ", m$description, "\n", sep = "")
}
```

# Population

The five data sets are summarised in Table 1 and Supplementary Table 7 of the
source. One is synthetic; the other four are pooled Phase 1 clinical data sets
held by AstraZeneca (available via Vivli, not distributed with the paper).

```{r population}
popTable <- do.call(rbind, lapply(richardsonModels, function(nm) {
  p <- readModelDb(nm)()$population
  data.frame(
    Model      = nm,
    Species    = p$species,
    N          = p$n_subjects,
    Studies    = p$n_studies,
    `Dose range` = if (is.null(p$dose_range)) NA_character_ else p$dose_range,
    check.names = FALSE
  )
}))
knitr::kable(popTable, row.names = FALSE)
```

Observation counts (Supplementary Table 7): ribociclib 1,344; camizestrant
2,709; osimertinib 3,766; olaparib 7,397; tezepelumab 1,477.

The ribociclib data set is **synthetic** -- 96 individuals simulated from the
published ribociclib popPK model of Lu et al. (2021) with the covariate effects
excluded, at 100, 200, 400 and 600 mg with Phase 1-like sampling (Methods,
"Data"). The paper reports it as the one case where the automated search
recovered the data-generating structure exactly, so it functions as a positive
control for the method rather than as a clinical model.

No demographic covariates appear in any of the five models: the paper's search
space contains no demographic covariate features at all (Supplementary Table 1),
only structural features, IIV switches, a dose-dependency switch and the
residual-error form. Accordingly the paper reports no baseline demographics for
the four clinical cohorts, and the packaged models declare no demographic
covariates. The full metadata is available programmatically:

```{r population-meta}
str(readModelDb("Richardson_2025_tezepelumab")()$population)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them. "Suppl Table N" refers to the Supplementary
Information PDF of the source; "`.lst`" refers to `FinalResultFile.lst` in the
Code Availability repository, which is the authors' own NONMEM output for the
top-ranked ribociclib model.

## Structure

| Model | Structure | Source location |
|-------|-----------|-----------------|
| ribociclib | 2-cmt; zero-order input into central over duration D with lag; KA fixed to 0 so the depot is inactive; dose effects on CL, Vp, Q; combined error | Table 4 row 1; Suppl Data 1 `ribociclib` top row; `tokens.json` `ZERO_ORDER_ABS` option 2; `FinalControlFile.mod` |
| camizestrant | 2-cmt; first-order absorption with 3 transit compartments, all transfers = KA; dose effects on CL, Vc; combined error (log scale) | Table 4 row 2; Suppl Data 1 `camizestrant` top row; `tokens.json` `TRANSIT_COMP` option 2 |
| osimertinib | 1-cmt; first-order absorption with 4 transit compartments, all transfers = KA; no dose effect; combined error | Table 4 row 3; Suppl Data 1 `osimertinib` top row; `tokens.json` `TRANSIT_COMP` option 3 |
| olaparib | 2-cmt; first-order absorption with 4 transit compartments; dose effect on Vc; proportional error | Table 4 row 4; Suppl Data 1 `olaparib` top row; `tokens.json` `TRANSIT_COMP` option 3, `ERRORFIX` option 1 |
| tezepelumab | 2-cmt; sequential zero-order input into the depot over duration D1 then first-order KA to central; no dose effect; proportional error | Table 4 row 5; Suppl Data 1 `tezepelumab` top row; `tokens.json` `ZERO_ORDER_ABS` option 1, `ERRORFIX` option 1 |

Two structural forms come from the published model-space files rather than from
the paper's prose, because the prose does not state them:

* **Dose dependency** is a power model on dose normalised to a reference dose:
  `MU(P) = LOG(TVP) + LOG(DOSE/DOSEREF) * THETA(cov_P)`, i.e.
  `P = TVP * (DOSE/DOSEREF)^theta` (`template.txt` and `tokens.json`
  `DOSE_COV_*`; confirmed in `FinalControlFile.mod`).
* **Transit absorption** places N transit compartments in series between depot
  and central with *every* transfer rate constant set to KA -- which is why the
  parameter tables label the estimate `KA = KTR` (`tokens.json`
  `TRANSIT_COMP`).

## Parameters

| Model | Parameter | Value | Source location |
|-------|-----------|-------|-----------------|
| ribociclib | `lcl` | 23.40 L/h | Suppl Table 2 THETA1 |
| ribociclib | `lvc` | 244.38 L | Suppl Table 2 THETA2 |
| ribociclib | `lvp` | 714.62 L | Suppl Table 2 THETA7 |
| ribociclib | `lq` | 68.67 L/h | Suppl Table 2 THETA8 |
| ribociclib | `ltlag` | 0.36 h | Suppl Table 2 THETA11 |
| ribociclib | `ld1` | 3.47 h | Suppl Table 2 THETA12 |
| ribociclib | `e_dose_cl` | -0.493 | `.lst` THETA3 (Suppl Table 2 -0.49) |
| ribociclib | `e_dose_vp` | -0.606 | `.lst` THETA9 (Suppl Table 2 -0.61) |
| ribociclib | `e_dose_q` | -0.747 | `.lst` THETA10 (Suppl Table 2 -0.75) |
| ribociclib | IIV CL, Vc, Vp, Q | 0.243, 0.903, 0.244, 0.315 | `.lst` OMEGA(1,1), (2,2), (4,4), (5,5) |
| ribociclib | IIV tlag, D | 0.01 FIX, 0.01 FIX | `.lst` OMEGA(6,6), (7,7) |
| ribociclib | `addSd`, `propSd` | 3.42 ng/mL, 0.217 | Suppl Table 2 THETA5; `.lst` THETA6 |
| camizestrant | `lcl` | 62.40 L/h | Suppl Table 3 THETA1 |
| camizestrant | `lvc` | 857.55 L | Suppl Table 3 THETA2 |
| camizestrant | `lvp` | 439.30 L | Suppl Table 3 THETA8 |
| camizestrant | `lq` | 36.48 L/h | Suppl Table 3 THETA9 |
| camizestrant | `lktr` | 2.76 1/h | Suppl Table 3 THETA5 |
| camizestrant | `e_dose_cl`, `e_dose_vc` | -0.17, -0.34 | Suppl Table 3 THETA3, THETA4 |
| camizestrant | IIV CL, Vc, ktr | 40%, 41%, 56% CV | Suppl Table 3 OMEGA(1,1)-(3,3) |
| camizestrant | IIV Vp, Q | 15% CV FIX | Suppl Table 3 OMEGA(4,4), (5,5) |
| camizestrant | `addSd`, `propSd` | 1.01 ng/mL, 0.32 | Suppl Table 3 THETA6, THETA7 |
| osimertinib | `lcl` | 13.84 L/h | Suppl Table 4 THETA1 |
| osimertinib | `lvc` | 1101.20 L | Suppl Table 4 THETA2 |
| osimertinib | `lktr` | 1.70 1/h | Suppl Table 4 THETA3 |
| osimertinib | IIV CL, Vc, ktr | 53%, 52%, 48% CV | Suppl Table 4 OMEGA(1,1)-(3,3) |
| osimertinib | `addSd`, `propSd` | 2.61 ng/mL, 0.17 | Suppl Table 4 THETA4, THETA5 |
| olaparib | `lcl` | 5.21 L/h | Suppl Table 5 THETA1 |
| olaparib | `lvc` | 35.06 L | Suppl Table 5 THETA2 |
| olaparib | `lvp` | 18.77 L | Suppl Table 5 THETA7 |
| olaparib | `lq` | 0.41 L/h | Suppl Table 5 THETA8 |
| olaparib | `lktr` | 9.22 1/h | Suppl Table 5 THETA4 |
| olaparib | `e_dose_vc` | 0.17 | Suppl Table 5 THETA3 |
| olaparib | IIV CL, Vc, ktr, Q | 55%, 32%, 60%, 190% CV | Suppl Table 5 OMEGA(1,1)-(3,3), (5,5) |
| olaparib | IIV Vp | 15% CV FIX | Suppl Table 5 OMEGA(4,4) |
| olaparib | `addSd`, `propSd` | 0 FIX, 0.36 | Suppl Table 5 THETA5, THETA6 |
| tezepelumab | `lcl` | 0.21 L/day | Suppl Table 6 THETA1 |
| tezepelumab | `lvc` | 2.84 L | Suppl Table 6 THETA2 |
| tezepelumab | `lvp` | 3.6 L | Suppl Table 6 THETA6 |
| tezepelumab | `lq` | 6.04 L/day | Suppl Table 6 THETA7 |
| tezepelumab | `lka` | 0.44 1/day | Suppl Table 6 THETA3 |
| tezepelumab | `ld1` | 0.11 day | Suppl Table 6 THETA8 |
| tezepelumab | IIV CL, Vc, ka, Vp, D1 | 39%, 69%, 56%, 34%, 82% CV | Suppl Table 6 OMEGA(1,1)-(4,4), (6,6) |
| tezepelumab | IIV Q | 15% CV FIX | Suppl Table 6 OMEGA(5,5) |
| tezepelumab | `addSd`, `propSd` | 0 FIX, 0.1 | Suppl Table 6 THETA4, THETA5 |
| all | `lfdepot` / `lfcentral` | 1 FIX | `template.txt` `F1 = 1`; Suppl Table 6 footnote for tezepelumab SC |

## The omega scale

Supplementary Tables 3-6 report IIV only as CV%, without saying which
convention converts a NONMEM OMEGA variance to that percentage. The ribociclib
model settles it, because the authors published the actual NONMEM output
alongside the tabulated CV%: the `.lst` OMEGA diagonal is
(0.243, 0.903, 0.244, 0.315, 0.01, 0.01) and Supplementary Table 2 reports
(52%, 121%, 53%, 61%, 10%, 10%). Only the log-normal form reproduces that
row by row -- the naive `sqrt(omega^2)` reading gives 49% and 95% for the first
two, which does not match.

```{r omega-scale}
lstOmega  <- c(CL = 0.243, Vc = 0.903, Vp = 0.244, Q = 0.315, tlag = 0.01, D = 0.01)
tabulated <- c(52, 121, 53, 61, 10, 10)
data.frame(
  Parameter          = names(lstOmega),
  `omega^2 (.lst)`   = lstOmega,
  `sqrt(exp(w2)-1)`  = round(sqrt(exp(lstOmega) - 1) * 100),
  `sqrt(w2)`         = round(sqrt(lstOmega) * 100),
  `Suppl Table 2`    = tabulated,
  check.names = FALSE, row.names = NULL
) |>
  knitr::kable()
```

Every other model therefore uses `omega^2 = log(1 + CV^2)` to invert the
tabulated CV%.

# Simulation set-up

Each model needs its own dosing route, time unit and observation grid. The
`useModelledDuration` flag marks the two models whose dose record must carry
`rate = -2` so that rxode2 takes the input duration from the model's `dur()`
statement.

```{r spec}
spec <- list(
  Richardson_2025_ribociclib = list(
    dose = 600, cmt = "central", dur = TRUE,  tmax = 336, tdense = 24, tunit = "h",
    cunit = "ng/mL", refDose = 600, nTransfer = NA, extra = 0.36 + 3.47),
  Richardson_2025_camizestrant = list(
    dose = 100, cmt = "depot",   dur = FALSE, tmax = 240, tdense = 24, tunit = "h",
    cunit = "ng/mL", refDose = 100, nTransfer = 4,  extra = NULL),
  Richardson_2025_osimertinib = list(
    dose = 80,  cmt = "depot",   dur = FALSE, tmax = 504, tdense = 36, tunit = "h",
    cunit = "ng/mL", refDose = NA, nTransfer = 5,  extra = NULL),
  Richardson_2025_olaparib = list(
    dose = 100, cmt = "depot",   dur = FALSE, tmax = 336, tdense = 12, tunit = "h",
    cunit = "ng/mL", refDose = 100, nTransfer = 5,  extra = NULL),
  Richardson_2025_tezepelumab = list(
    dose = 210, cmt = "depot",   dur = TRUE,  tmax = 224, tdense = 21, tunit = "day",
    cunit = "ug/mL", refDose = NA, nTransfer = NA, extra = NULL)
)
```

The observation grid is deliberately **non-uniform**: dense over the absorption
and distribution phase, sparse through the terminal phase. A uniform grid over
the whole window under-samples the peak, and trapezoidal AUC then reads several
percent low for the fast-absorbing models (olaparib's `ktr` of 9.22/h puts Tmax
below 2 h in a 336 h window). This is a property of the numerical integration,
not of the models, but it would otherwise be indistinguishable from a real
discrepancy in the validation below.

```{r events}
buildEvents <- function(s, nSub = 1) {
  ev <- if (s$dur) {
    rxode2::et(amt = s$dose, cmt = s$cmt, rate = -2)
  } else {
    rxode2::et(amt = s$dose, cmt = s$cmt)
  }
  grid <- sort(unique(c(
    seq(0, s$tdense, length.out = 150),
    seq(s$tdense, s$tmax, length.out = 150),
    s$extra
  )))
  ev <- rxode2::et(ev, grid, cmt = "central")
  ev <- as.data.frame(ev)
  ev$DOSE <- s$dose
  if (nSub > 1) {
    ev <- do.call(rbind, lapply(seq_len(nSub), function(i) {
      x <- ev; x$id <- i; x
    }))
  }
  ev
}
```

For ribociclib the exact expected peak time `tlag + D = 0.36 + 3.47 = 3.83 h` is
added to the grid explicitly, so the Tmax check below is not limited by grid
resolution.

Observation records target `cmt = "central"`, the ODE state. rxode2 returns the
algebraic observable `Cc` as a column on those rows; naming `Cc` as a
compartment would inject an extra compartment slot and renumber the ODE states.

# Typical-value profiles

`rxode2::zeroRe()` strips the random effects, giving the typical-value profile
implied by each parameter table.

```{r typical}
typical <- do.call(rbind, lapply(names(spec), function(nm) {
  s <- spec[[nm]]
  mod <- rxode2::zeroRe(readModelDb(nm)())
  out <- rxode2::rxSolve(mod, buildEvents(s), returnType = "data.frame")
  data.frame(Model = nm, time = out$time, Cc = out$Cc,
             tunit = s$tunit, cunit = s$cunit)
}))

ggplot(typical, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ Model, scales = "free", ncol = 2) +
  scale_y_log10() +
  labs(x = "Time after dose (h, or day for tezepelumab)",
       y = "Concentration (ng/mL, or ug/mL for tezepelumab)",
       title = "Typical-value single-dose profiles") +
  theme_bw()
```

# Validation

The source paper reports **no** NCA summary -- it ranks structures by objective
function value, penalty and fitness (Table 5), which are properties of the
NONMEM fit to data that is not distributed. There is therefore no published
Cmax / AUC / half-life table to compare against. Instead each model is validated
against quantities that follow *analytically* from its own parameter table, which
is a strict test of the encoded structure: if a compartment is mis-wired, an
absorption chain is mis-specified, a unit conversion is wrong, or a rate constant
is transposed, these identities break.

## Mass balance: AUC(0-inf) must equal Dose / CL

For any linear model with complete input (F = 1) the dose eventually clears
through CL alone, so `AUC(0-inf) = Dose / CL` exactly, independent of the number
of compartments, the absorption model or the transit-chain length.

## Terminal slope

For the four two-compartment models the terminal rate constant is the smaller
root of `lambda^2 - (kel + k12 + k21) * lambda + kel * k21 = 0`; for the
one-compartment osimertinib model it is simply `kel = CL / Vc`.

```{r analytic}
analyticTargets <- function(nm, s) {
  mod  <- readModelDb(nm)()
  ini  <- mod$theta
  cl   <- exp(ini[["lcl"]])
  vc   <- exp(ini[["lvc"]])
  kel  <- cl / vc
  if ("lvp" %in% names(ini)) {
    vp  <- exp(ini[["lvp"]])
    q   <- exp(ini[["lq"]])
    k12 <- q / vc
    k21 <- q / vp
    b   <- kel + k12 + k21
    lambda <- (b - sqrt(b^2 - 4 * kel * k21)) / 2
  } else {
    lambda <- kel
  }
  data.frame(
    Model     = nm,
    aucinf.obs = 1000^(s$cunit == "ng/mL") * s$dose / cl,
    half.life  = log(2) / lambda
  )
}
refNca <- do.call(rbind, Map(analyticTargets, names(spec), spec))
knitr::kable(refNca, row.names = FALSE, digits = 3,
             caption = "Analytic targets derived from each model's own parameter table")
```

The `1000^(...)` factor converts `Dose (mg) / CL (L/h)` from mg*h/L to the
ng*h/mL scale used by the four small-molecule models; tezepelumab is already on
the mg/L = ug/mL scale.

## PKNCA

NCA is run with PKNCA on the typical-value profiles, grouped by model. The
concentration frame is filtered on `!is.na(Cc)` only -- dropping the `time = 0`
record would make PKNCA warn that the AUC interval starts before the first
measurement.

```{r pknca}
ncaConc <- typical |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = Model, treatment = Model)

ncaDose <- do.call(rbind, lapply(names(spec), function(nm) {
  data.frame(id = nm, treatment = nm, time = 0, dose = spec[[nm]]$dose)
}))

oConc <- PKNCA::PKNCAconc(ncaConc, Cc ~ time | id + treatment)
oDose <- PKNCA::PKNCAdose(ncaDose, dose ~ time | id + treatment)

intervals <- do.call(rbind, lapply(names(spec), function(nm) {
  data.frame(
    start = 0, end = Inf, id = nm, treatment = nm,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
}))

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(oConc, oDose, intervals = intervals))
simNca <- as.data.frame(res) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
knitr::kable(simNca, digits = 3, caption = "PKNCA results on the typical-value profiles")
```

## Simulated versus analytic

```{r comparison}
simForCompare <- simNca |>
  dplyr::rename(Model = treatment) |>
  dplyr::select(Model, aucinf.obs, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(simForCompare),
  reference = refNca,
  by        = "Model",
  units     = c(aucinf.obs = "conc*time", half.life = "time")
)
knitr::kable(cmp, caption = "Simulated NCA versus analytic targets")
attr(cmp, "footnote")
```

`AUC(0-inf)` agrees with `Dose / CL` to within a fraction of a percent for all
five models, and the terminal half-life agrees with the exact eigenvalue to
better than 1%. The residuals are trapezoidal integration and terminal
extrapolation on a finite grid, not model discrepancies -- the mass-balance
identity is exact for these structures, so any real error in the compartment
wiring, the unit conversion, or the placement of the absorption input would show
up here as a gross mismatch rather than a fraction of a percent.

# Absorption-model check

The absorption models differ between the five structures, and Tmax is the
observable that distinguishes them. A pure first-order chain of `n` equal-rate
steps into a compartment whose elimination is much slower than `ktr` peaks at
roughly `n / ktr`; a zero-order input peaks at the end of the input duration.

**Ribociclib gives an exact check.** Its dose enters the central compartment
directly as a zero-order input, and for a zero-order input into a linear system
the concentration rises for as long as the input runs. The peak must therefore
fall exactly at the end of the input window, `tlag + D = 0.36 + 3.47 = 3.83 h`.
If the lag or the duration had been attached to the wrong compartment, or the two
transposed, this would not hold.

```{r tmax-ribociclib}
riboTmax <- typical |>
  dplyr::filter(Model == "Richardson_2025_ribociclib") |>
  dplyr::slice_max(Cc, n = 1)
c(simulated = riboTmax$time, expected = 0.36 + 3.47)
```

**The transit models give an ordering check.** For a chain of `n` equal-rate
first-order transfers the mean transit time from dose to central is `n / ktr`,
and because elimination is far slower than absorption in all three models, Tmax
must land somewhat after that mean transit time and must be ordered inversely
with `ktr`. The exact peak has no closed form once distribution is included, so
this is a consistency check rather than an identity.

```{r tmax-transit}
transitCheck <- do.call(rbind, lapply(
  c("Richardson_2025_camizestrant", "Richardson_2025_osimertinib",
    "Richardson_2025_olaparib"),
  function(nm) {
    s   <- spec[[nm]]
    ktr <- exp(readModelDb(nm)()$theta[["lktr"]])
    tm  <- typical$time[typical$Model == nm][which.max(typical$Cc[typical$Model == nm])]
    data.frame(Model = nm, `ktr (1/h)` = ktr, Transfers = s$nTransfer,
               `MTT = n/ktr (h)` = s$nTransfer / ktr, `Tmax (simulated, h)` = tm,
               check.names = FALSE)
  }
))
knitr::kable(transitCheck, digits = 3)
```

Tmax exceeds the mean transit time in each case and falls in the same order as
`1 / ktr` (olaparib fastest, camizestrant next, osimertinib slowest), as the
chain lengths and rate constants require.

**Tezepelumab** combines a zero-order input into the depot over
`D1 = 0.11 day` with a slow first-order transfer (`ka = 0.44 /day`), so the
0.11-day input is negligible against absorption and the peak is governed by
`ka` relative to elimination -- it should fall several days after dosing, which
it does (`r round(typical$time[typical$Model == "Richardson_2025_tezepelumab"][which.max(typical$Cc[typical$Model == "Richardson_2025_tezepelumab"])], 1)`
days).

# Dose dependency

Three of the five structures carry a dose effect. Because it is a power model on
`DOSE / refDose`, the typical value of the affected parameter is exactly the
tabulated value when `DOSE = refDose`, and departs from it as a straight line on
log-log axes elsewhere. The exponents are negative for ribociclib and
camizestrant (CL falls as dose rises -- more than proportional exposure) and
positive for olaparib (Vc rises with dose).

```{r dose-dep}
doseModels <- c("Richardson_2025_ribociclib", "Richardson_2025_camizestrant",
                "Richardson_2025_olaparib")
doseGrid <- do.call(rbind, lapply(doseModels, function(nm) {
  s <- spec[[nm]]
  mod <- rxode2::zeroRe(readModelDb(nm)())
  do.call(rbind, lapply(s$refDose * c(0.25, 0.5, 1, 2, 4), function(d) {
    s2 <- s; s2$dose <- d
    out <- rxode2::rxSolve(mod, buildEvents(s2), returnType = "data.frame")
    auc <- sum(diff(out$time) *
                 (head(out$Cc, -1) + tail(out$Cc, -1)) / 2)
    data.frame(Model = nm, dose = d, auclast = auc)
  }))
}))

ggplot(doseGrid, aes(dose, auclast)) +
  geom_line() + geom_point() +
  facet_wrap(~ Model, scales = "free", ncol = 3) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Dose (mg)", y = "AUC(0-tlast) (ng*h/mL)",
       title = "Dose-exposure relationship for the three dose-dependent structures") +
  theme_bw()
```

```{r dose-dep-table}
doseGrid |>
  dplyr::group_by(Model) |>
  dplyr::mutate(`Dose / refDose` = dose / dose[3L],
                `AUC / AUC at refDose` = auclast / auclast[3L]) |>
  dplyr::ungroup() |>
  dplyr::select(Model, dose, `Dose / refDose`, `AUC / AUC at refDose`) |>
  knitr::kable(digits = 3,
               caption = "Dose proportionality. A value above the dose ratio indicates more than proportional exposure.")
```

For ribociclib, CL scales as `(DOSE/600)^-0.493`, so a 4-fold dose increase
raises AUC by `4^(1 + 0.493) = ` `r round(4^1.493, 2)`-fold rather than 4-fold.
That is the nonlinearity the search detected and it is reproduced here.

# Between-subject variability

A 100-subject cohort per model exercises the IIV structure, including the two
large variances the paper itself flags: 121% CV on ribociclib Vc and 190% CV on
olaparib Q (the latter incurs a penalty of 81.53, the single largest estimated
value penalty across all five top models -- Supplementary Table 5).

```{r vpc, fig.height = 6}
set.seed(20250101)
nSub <- 200
vpc <- do.call(rbind, lapply(names(spec), function(nm) {
  s <- spec[[nm]]
  out <- rxode2::rxSolve(readModelDb(nm)(), buildEvents(s, nSub = nSub),
                         returnType = "data.frame")
  data.frame(Model = nm, id = out$id, time = out$time, Cc = out$Cc)
}))

vpcSummary <- vpc |>
  dplyr::filter(Cc > 0) |>
  dplyr::group_by(Model, time) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(vpcSummary, aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line(aes(y = p50), linewidth = 0.7) +
  facet_wrap(~ Model, scales = "free", ncol = 2) +
  scale_y_log10() +
  labs(x = "Time after dose (h, or day for tezepelumab)",
       y = "Concentration (ng/mL, or ug/mL for tezepelumab)",
       title = paste0("Median and 5th-95th percentile, n = ", nSub, " per model")) +
  theme_bw()
```

For the two models with **no dose effect** (osimertinib and tezepelumab) the
cohort gives an exact gate on the IIV encoding. Individual exposure is
`AUC(0-inf)_i = Dose / CL_i` with `CL_i = exp(lcl + eta_i)` log-normal, and
`1 / CL_i` is then log-normal with the same variance, so the coefficient of
variation of individual AUC(0-inf) must equal the tabulated CL CV%: 53% for
osimertinib and 39% for tezepelumab. Recovering that number from the simulated
cohort confirms both the omega scale and that the eta is attached to CL.

Because the exposures are log-normal, the variance is estimated on the log scale
(`omega^2 = var(log(AUC))`, then `CV% = sqrt(exp(omega^2) - 1) * 100`), which is
the consistent estimator for this distribution. A cohort of
`r nSub` is a finite sample, so the estimate carries Monte Carlo uncertainty;
the gate is whether the tabulated CV% falls inside the 95% confidence interval
of the simulated estimate, not whether the point estimates coincide.

```{r vpc-cv}
aucInfBySubject <- function(d) {
  d <- d[d$Cc > 0, ]
  n  <- nrow(d)
  tail_idx <- seq(max(1, n - 40), n)
  lz  <- -stats::coef(stats::lm(log(d$Cc[tail_idx]) ~ d$time[tail_idx]))[2]
  auc <- sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  auc + utils::tail(d$Cc, 1) / lz
}

cvFromLogVar <- function(v) sqrt(exp(v) - 1) * 100

cvTable <- vpc |>
  dplyr::group_by(Model, id) |>
  dplyr::group_modify(~ data.frame(auc = aucInfBySubject(.x))) |>
  dplyr::group_by(Model) |>
  dplyr::summarise(v = stats::var(log(auc)), n = dplyr::n(), .groups = "drop") |>
  dplyr::mutate(
    `Simulated CV%` = round(cvFromLogVar(v)),
    `95% CI`        = paste0(
      round(cvFromLogVar(v * (n - 1) / stats::qchisq(0.975, n - 1))), "-",
      round(cvFromLogVar(v * (n - 1) / stats::qchisq(0.025, n - 1))), "%"),
    `Tabulated CL CV%` = c(40, 55, 53, 52, 39)[match(
      Model, c("Richardson_2025_camizestrant", "Richardson_2025_olaparib",
               "Richardson_2025_osimertinib", "Richardson_2025_ribociclib",
               "Richardson_2025_tezepelumab"))],
    `Exact gate` = Model %in% c("Richardson_2025_osimertinib",
                                "Richardson_2025_tezepelumab")) |>
  dplyr::select(-v, -n)
knitr::kable(cvTable,
             caption = "Simulated exposure variability against the tabulated CL CV%")
```

For the two models where the identity holds exactly, the tabulated CL CV% falls
inside the simulated confidence interval, confirming both the omega scale and
that the eta is attached to CL. The other three models are not expected to
match, and should not: ribociclib and camizestrant carry a dose effect plus IIV
on volumes and absorption, and olaparib's 190% CV on Q feeds into the terminal
phase, so exposure variability there is a mixture of several etas rather than CL
alone.

# Assumptions and deviations

These are the points where the packaged models required a decision that the
source does not settle. Each is a deliberate, documented choice, not a fitted
value.

1. **The reference dose `DOSEREF` is not reported for any data set.** The
   dose-dependency term is `(DOSE / DOSEREF)^theta`, so the tabulated typical
   values of the affected parameters are the values *at the reference dose*. The
   authors' data sets carried a `DOSEREF` column (`$INPUT` of
   `FinalControlFile.mod`) but its numeric value appears nowhere in the paper,
   the supplement or the code repository. The packaged models use:
   * ribociclib: `DOSEREF = 600 mg`, the highest of the four simulated dose
     levels and the therapeutic dose. The dose range *is* reported for this data
     set (Methods, "Data").
   * camizestrant and olaparib: `DOSEREF = 100 mg`, a deliberately round,
     arbitrary anchor. Neither the reference dose nor the dose range is reported
     for these cohorts, so no better-founded value exists on the record. This is
     a declared normalisation constant, not a value read from the source.

   A different reference dose `D` rescales each affected typical value by
   `(refDose / D)^theta` and leaves the *shape* of the dose dependence -- the
   exponent, and therefore every dose-ratio prediction -- unchanged. The
   validation above is run at `DOSE = refDose`, where the dose term equals 1, so
   none of the gates depend on this choice.

2. **Concentration units are inferred, not stated.** The parameter tables give
   CL in L/h (or L/day) and volumes in L but never name the concentration unit,
   while reporting additive residual errors of 3.42, 1.01 and 2.61. Those
   magnitudes are only self-consistent on a ng/mL scale: at 600 mg with
   Vc = 244 L, ribociclib's central concentration is ~2.5 mg/L, so an additive
   error of 3.42 mg/L would exceed the peak concentration, whereas 3.42 ng/mL is
   a plausible assay floor. The four small-molecule models therefore compute
   `Cc = 1000 * central / vc` (mg and L in, ng/mL out) and the mAb model uses
   `Cc = central / vc` (mg/L = ug/mL, the conventional scale for a monoclonal
   antibody, and its additive error is fixed to zero so the scale is
   unconstrained).

3. **Camizestrant's log-scale error is encoded on the linear scale.** The
   camizestrant data were fitted on the log scale (Table 1) with a "combined
   error model in log scale" (Table 4). The search space's linear-scale form is
   `W = SQRT(add^2 + (prop * IPRED)^2)`; its log-scale counterpart is that same
   `W` divided by `IPRED`, so the two are the same error model to first order.
   The packaged model uses `Cc ~ add(addSd) + prop(propSd)` on the linear scale
   with the tabulated 1.01 and 0.32.

4. **The ribociclib depot compartment is omitted.** For the zero-order
   absorption option the search space still declares a depot but fixes its
   transfer rate constant to zero (`tokens.json` `ZERO_ORDER_ABS` option 2 sets
   `0 FIX ; KA`, and Supplementary Table 2 records THETA4 as `0 (fix)`; the
   footnote states KA is fixed "to avoid modification of THETA numbering"). The
   dose goes directly into the central compartment, so the depot is inert and
   carrying it would add an unreachable state.

5. **No demographic covariates.** The paper's search space contains no
   demographic covariate features (Supplementary Table 1) -- covariate selection
   was explicitly out of scope, since automated tools for it already exist
   (Discussion). The only covariate in any of these models is the dose itself.
   Nothing was screened and discarded, so there is no
   `covariatesDataExcluded` list either.

6. **No published diagnostics are reproduced.** Supplementary Figures 1-4 are
   VPC and goodness-of-fit plots against the clinical observations. Those data
   are AstraZeneca-controlled (available through Vivli, per the Data
   Availability statement) and are not distributed with the paper, so the
   observed layer cannot be redrawn here. The VPC-style figure above shows the
   simulated distribution only.

7. **Half-life is compared loosely.** PKNCA's automatic terminal-slope selection
   reads an oral terminal half-life slightly long on a finite observation grid;
   the analytic eigenvalue is exact. The comparison table flags differences over
   20%, which this does not approach.

8. **Precision mixing for ribociclib.** Where the authors' published NONMEM
   output carries more significant figures than the rounded supplementary table
   (the three dose exponents, and the OMEGA diagonal which the table gives only
   as whole-percent CV), the `.lst` value is used and the table value is quoted
   in the in-file comment. The two agree to the supplementary table's precision.

9. **These are not the expert models.** Four of the five automated structures
   differ from the corresponding manually developed publication (Table 4), and
   the olaparib expert model additionally contains a CL autoinhibition mechanism
   that the search space cannot express at all. The paper's own framing is that
   these structures are "comparable to" the expert models, which is a claim
   about the search procedure, not an assertion that these are the definitive
   popPK models for these five drugs.
