---
title: "Axitinib (Rini 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Axitinib (Rini 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Rini_2013_axitinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Rini_2013_axitinib"))$description`
* Article: <https://doi.org/10.1002/jcph.73>

Rini 2013 pooled plasma axitinib concentrations from 590 subjects across
17 trials -- 383 healthy volunteers, 181 patients with metastatic renal
cell carcinoma (mRCC), and 26 patients with other solid tumours (Table 1)
-- and fit a single linear two-compartment population PK model with
first-order absorption and an estimated absorption lag time. Twelve
covariates were screened. Age > 60 years, Japanese ethnicity and active
smoking were retained on systemic clearance `CL`; body weight was
retained as a power function on the central volume `Vc`. The food and
crystal-polymorph effects of clinical interest -- fasting on the
absorption rate constant `ka` and on absolute oral bioavailability `F`,
and the `F` reduction for the marketed Form XLI polymorph relative to
the earlier Form IV -- were carried in the model throughout.

Rini 2013 is the pooled healthy-volunteer-plus-patient analysis. This
library also ships `Garrett_2014_axitinib`, the companion
healthy-volunteer-only analysis by an overlapping author group; the two
models share a structure but not their parameter values or their
covariate sets, and neither supersedes the other.

The paper's second half is an exposure-response analysis in a 168-patient
mRCC subset: multivariate Cox proportional-hazards regression for
progression-free and overall survival, and logistic regression for
RECIST partial response. Those are not encoded in the model file -- see
the Errata section for why -- but the exposure metric they are built on,
`AUC = total daily dose * F / CL`, is fully specified and is used as a
validation target below.

# Population

The pooled PK dataset (Rini 2013 Results "Subject characteristics";
reproduced programmatically via
`readModelDb("Rini_2013_axitinib")$population`) covers 590 subjects with
median (range) body weight 74 kg (37-136), creatinine clearance
103 mL/min (8-214), AST 22 U/L (9-154), ALT 21 U/L (5-188) and bilirubin
0.7 mg/dL (0.1-3). Overall median age is 42 years (18-85), but the two
sub-populations differ sharply: patients have a median age of 60 years
(32-85) and healthy volunteers 32 years (18-69). Most subjects are male
(85%) and Caucasian (61%). Only 19 of 590 subjects (3%) were active
smokers. Normal renal and hepatic function were study-entry
requirements, which is why creatinine clearance, AST, ALT and bilirubin
all screened out.

Healthy volunteers received a single 5 mg oral dose (one study
additionally gave 1 mg intravenously to anchor absolute bioavailability,
and one Chinese study gave single 5, 7 and 10 mg doses); patients
received 5 mg orally twice daily with titration up to 10 mg twice daily
permitted. Sixteen hepatically impaired subjects (study 13) and sixteen
study-1 patients started above the 5 mg bid maximum tolerated dose were
excluded from the analysis dataset.

The exposure-response subset is 168 of the 178 patients enrolled in the
three mRCC studies (Table 3): median age 60 years (34-85), 71% male, 68%
cytokine-refractory and 32% sorafenib-refractory, 65% with baseline
ECOG PS 0, median haemoglobin 12.4 g/dL and median corrected serum
calcium 9.4 mg/dL.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Rini_2013_axitinib.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model -- linear two-compartment, first-order absorption, absorption lag time | n/a | Methods "Model development"; Results "PK model" |
| Reference body weight for the `Vc` power function | 74.1 kg | Results "PK model" typical-value equation (screening median body weight 74 kg) |
| `lka` (fed reference) | log(0.482) | Table 2 "k a (hour-1): fed" = 0.482 |
| `lcl` | log(14.6) | Table 2 "CL (L/h)" = 14.6 |
| `lvc` (at 74.1 kg) | log(47.3) | Table 2 "V c (L)" = 47.3 |
| `lq` | log(4.00) | Table 2 "Q (L/h)" = 4.00 |
| `lvp` | log(393) | Table 2 "V p (L)" = 393 |
| `lfdepot` (Form IV, fed reference) | log(0.457) | Table 2 "F: fed/Form IV" = 0.457 |
| `ltlag` | log(0.454) | Table 2 "t lag (hour)" = 0.454 |
| `e_age_cl` (age > 60 y on CL) | -0.213 | Table 2 "Age >60-yr effect on CL"; Results "PK model" equation `CL = 14.6 * (1 - 0.213 * Age>60) * ...` |
| `e_jpn_cl` (Japanese on CL) | -0.249 | Table 2 "Japanese ethnicity effect on CL"; Results "PK model" equation `... * (1 - 0.249 * RaceJapanese) * ...` |
| `e_smoke_cl` (active smoking on CL) | 1.02 | Table 2 "Smoking status on CL"; Results "PK model" equation `... * (1 + 1.02 * Smokeractive)` |
| `e_wt_vc` (power exponent of WT on Vc) | 0.778 | Table 2 "Weight effect on V c"; Results "PK model" equation `Vc = 47.3 * (weight / 74.1 kg)^0.778` |
| `e_fast_ka` (fasting on ka) | 1.97 | Table 2 "Fasting effect on k a"; footnote e: `ka_fasted = 0.482 * (1 + 1.97) = 1.43 /h` |
| `e_fast_f` (fasting on F, Form IV only) | 0.330 | Table 2 "Fasting on F, Form IV"; footnote f: `F = 0.457 * (1 + 0.33) = 0.608` |
| `e_xli_f` (Form XLI on F) | -0.121 | Table 2 "Form XLI on F"; footnote g: `F = 0.457 * (1 - 0.121) = 0.402`, and no food effect with Form XLI |
| `omega^2` CL | 0.599^2 = 0.358801 | Table 2 "CL (L/h) 14.6 (59.9)" with footnote a (IIV given as %CV) |
| `omega^2` Vc | 0.397^2 = 0.157609 | Table 2 "V c (L) 47.3 (39.7)" |
| `omega^2` shared by Q and Vp | 0.868^2 = 0.753424 | Table 2 "Q (L/h) 4.00 (86.8)" with footnote d (Q and Vp 100% correlated, same IIV) |
| `omega^2` ka | 0.77^2 = 0.5929 | Table 2 "k a (hour-1): fed 0.482 (77)" |
| `expSd` (oral log-scale residual SD) | 0.582 | Table 2 "Residual error (%), oral" = 58.2 |
| IV log-scale residual SD (not encoded) | 0.335 | Table 2 "Residual error (%), intravenous" = 33.5 (see Errata) |
| Exposure metric for the PK/PD analysis | `AUC = total daily dose * F / CL` | Methods "PK/PD Modeling" |
| Published median mean-daily AUC, 168 mRCC patients | 375 h*ng/mL (32.8-1,728) | Table 3 "AUC, hxng/mL" |
| Cox regression coefficients for AUC and dBP | see Errata | Table 5 |

# Typical-value covariate identities

Every covariate effect in Rini 2013 has a closed-form typical value that
the paper either prints as an equation (Results "PK model") or works out
in a Table 2 footnote. Because both sides of these checks use the same
parameters, the comparison is exact arithmetic and is asserted tightly.

```{r typical-values}
mod     <- readModelDb("Rini_2013_axitinib")
mod_typ <- rxode2::zeroRe(mod)

# One subject per covariate combination, per the zero-random-effect
# convention: a typical-value cohort has no within-arm variability.
combos <- tibble::tribble(
  ~arm,                                   ~WT,  ~AGE_GT60, ~RACE_JAPANESE, ~SMOKE, ~FED, ~FORM_AXI_XLI,
  "Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", 74.1, 0L, 0L, 0L, 1L, 0L,
  "Age > 60 y",                           74.1, 1L, 0L, 0L, 1L, 0L,
  "Japanese",                             74.1, 0L, 1L, 0L, 1L, 0L,
  "Active smoker",                        74.1, 0L, 0L, 1L, 1L, 0L,
  "Age > 60 y and Japanese",              74.1, 1L, 1L, 0L, 1L, 0L,
  "58 kg (10th percentile)",              58,   0L, 0L, 0L, 1L, 0L,
  "94 kg (90th percentile)",              94,   0L, 0L, 0L, 1L, 0L,
  "Fasted, Form IV",                      74.1, 0L, 0L, 0L, 0L, 0L,
  "Fed, Form XLI",                        74.1, 0L, 0L, 0L, 1L, 1L,
  "Fasted, Form XLI",                     74.1, 0L, 0L, 0L, 0L, 1L
) |>
  dplyr::mutate(id = dplyr::row_number())

typ_events <- dplyr::bind_rows(
  combos |> dplyr::mutate(time = 0, evid = 1L, amt = 5, cmt = "depot"),
  combos |> dplyr::mutate(time = 1, evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# Only non-model columns are passed through `keep`; every covariate here is
# also a model variable, and keeping one of those risks shadowing it.
typ <- rxode2::rxSolve(
  mod_typ,
  events = as.data.frame(typ_events),
  keep   = "arm"
) |>
  as.data.frame() |>
  dplyr::group_by(arm) |>
  dplyr::slice_head(n = 1) |>
  dplyr::ungroup() |>
  dplyr::arrange(match(arm, combos$arm)) |>
  dplyr::select(arm, ka, cl, vc, q, vp, fdepot, tlag)

knitr::kable(
  typ |>
    dplyr::rename(
      "Covariate combination" = arm,
      "ka (1/h)"  = ka,
      "CL (L/h)"  = cl,
      "Vc (L)"    = vc,
      "Q (L/h)"   = q,
      "Vp (L)"    = vp,
      "F"         = fdepot,
      "tlag (h)"  = tlag
    ),
  digits  = 4,
  caption = "Typical-value PK parameters by covariate combination (random effects zeroed)."
)

get_par <- function(arm_label, par) typ[[par]][typ$arm == arm_label]

ref_cl  <- get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "cl")
ref_vc  <- get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "vc")
ref_ka  <- get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "ka")
ref_f   <- get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "fdepot")

identities <- tibble::tibble(
  claim = c(
    "CL, Vc, Q, Vp, ka, F, tlag reproduce Table 2 typical values",
    "Age > 60 y lowers CL by 21.3% (Results 'PK model' equation)",
    "Japanese ethnicity lowers CL by 24.9% (Results 'PK model' equation)",
    "Active smoking raises CL by 102% (Results 'PK model' equation)",
    "Age > 60 y and Japanese combine multiplicatively",
    "Vc scales as (WT / 74.1)^0.778 at 58 kg and 94 kg",
    "Fasted ka = 1.43 /h (Table 2 footnote e)",
    "Fasted F, Form IV = 0.608 (Table 2 footnote f)",
    "Form XLI F = 0.402 in the fed state (Table 2 footnote g)",
    "Form XLI F is unchanged by fasting (Table 2 footnote g)"
  ),
  holds = c(
    isTRUE(all.equal(
      c(ref_ka, ref_cl, ref_vc,
        get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "q"),
        get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "vp"),
        ref_f,
        get_par("Reference (74.1 kg, <=60 y, non-Japanese, non-smoker, fed, Form IV)", "tlag")),
      c(0.482, 14.6, 47.3, 4.00, 393, 0.457, 0.454),
      tolerance = 1e-6
    )),
    isTRUE(all.equal(get_par("Age > 60 y", "cl") / ref_cl, 1 - 0.213, tolerance = 1e-8)),
    isTRUE(all.equal(get_par("Japanese", "cl") / ref_cl, 1 - 0.249, tolerance = 1e-8)),
    isTRUE(all.equal(get_par("Active smoker", "cl") / ref_cl, 1 + 1.02, tolerance = 1e-8)),
    isTRUE(all.equal(get_par("Age > 60 y and Japanese", "cl") / ref_cl,
                     (1 - 0.213) * (1 - 0.249), tolerance = 1e-8)),
    isTRUE(all.equal(
      c(get_par("58 kg (10th percentile)", "vc"), get_par("94 kg (90th percentile)", "vc")) / ref_vc,
      c((58 / 74.1)^0.778, (94 / 74.1)^0.778), tolerance = 1e-8
    )),
    isTRUE(all.equal(get_par("Fasted, Form IV", "ka"), 1.43, tolerance = 2e-3)),
    isTRUE(all.equal(get_par("Fasted, Form IV", "fdepot"), 0.608, tolerance = 2e-3)),
    isTRUE(all.equal(get_par("Fed, Form XLI", "fdepot"), 0.402, tolerance = 2e-3)),
    isTRUE(all.equal(get_par("Fasted, Form XLI", "fdepot"),
                     get_par("Fed, Form XLI", "fdepot"), tolerance = 1e-10))
  )
)

knitr::kable(
  identities |> dplyr::rename("Published claim" = claim, "Reproduced" = holds),
  caption = "Closed-form typical-value identities from Rini 2013 Table 2 and the Results 'PK model' equations."
)

stopifnot(all(identities$holds))
```

# Virtual cohort: single-dose healthy volunteers

Original observed data are not publicly available. The single-dose
cohort below approximates the healthy-volunteer Phase I studies of
Table 1 (5 mg oral axitinib, dense sampling to 48 h) across the three
food / polymorph conditions that Rini 2013 Figure 1 and Table 2
emphasise. Healthy volunteers had a median age of 32 years, so
`AGE_GT60 = 0`; `RACE_JAPANESE = 0` and `SMOKE = 0` set the reference
subgroup.

```{r cohort-single-dose}
set.seed(20130501)

n_per_arm <- 100L

make_sd_arm <- function(n, fed_value, form_xli_value, label, id_offset) {
  cov <- tibble::tibble(
    id           = id_offset + seq_len(n),
    WT           = pmax(40, rnorm(n, mean = 74, sd = 15)),   # Results: median 74 kg (37-136)
    AGE_GT60     = 0L,
    RACE_JAPANESE = 0L,
    SMOKE        = 0L,
    FED          = fed_value,
    FORM_AXI_XLI = form_xli_value,
    treatment    = label
  )
  obs_times <- c(0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 6, 8, 12, 16, 24, 32, 36, 48)
  dplyr::bind_rows(
    cov |> dplyr::mutate(time = 0, evid = 1L, amt = 5, cmt = "depot"),
    tidyr::expand_grid(id = cov$id, time = obs_times) |>
      dplyr::left_join(cov, by = "id") |>
      dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

sd_events <- dplyr::bind_rows(
  make_sd_arm(n_per_arm, 1L, 0L, "Form IV fed",    0L),
  make_sd_arm(n_per_arm, 0L, 0L, "Form IV fasted", n_per_arm),
  make_sd_arm(n_per_arm, 1L, 1L, "Form XLI fed",   2L * n_per_arm)
)

stopifnot(!anyDuplicated(sd_events[, c("id", "time", "evid")]))
```

```{r simulate-single-dose}
sd_sim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(sd_events),
  keep   = "treatment"
) |>
  as.data.frame() |>
  dplyr::mutate(
    # rxSolve can return `id` as a factor; PKNCA needs the concentration and
    # dose frames to agree on its type.
    id        = as.integer(as.character(id)),
    treatment = factor(treatment, levels = c("Form IV fed", "Form IV fasted", "Form XLI fed"))
  )
```

## Replicating Rini 2013 Figure 1

Figure 1 shows visual predictive checks of Form IV in fed (panel A) and
fasted (panel B) subjects, plotting the 2.5th, 50th and 97.5th
percentiles down to the lower limit of quantification of 0.1 ng/mL
(Methods "PK Assessment": the assay linear range was 0.1-25 ng/mL in the
studies contributing the bulk of the single-dose data). Observed data are
not publicly distributed, so the panel below reproduces the simulated
percentile bands from the packaged model. The bands are computed from the
`sim` column, which carries the 58.2% oral residual error on top of the
interindividual variability, so they are comparable to percentiles of
observed concentrations rather than of individual predictions.

```{r figure-1}
lloq <- 0.1

sd_sim |>
  dplyr::filter(time > 0, treatment != "Form XLI fed") |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q025 = stats::quantile(sim, 0.025, na.rm = TRUE),
    Q50  = stats::quantile(sim, 0.500, na.rm = TRUE),
    Q975 = stats::quantile(sim, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = lloq, linetype = "dashed") +
  facet_wrap(~ treatment) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Plasma axitinib (ng/mL)",
    title = "Simulated VPC: single 5 mg oral dose of axitinib, Form IV",
    caption = paste(
      "Replicates the fed (A) and fasted (B) panels of Rini 2013 Figure 1.",
      "Dashed line is the 0.1 ng/mL assay LLOQ."
    )
  )
```

## Single-dose NCA

```{r pknca-single-dose}
sd_nca_conc <- sd_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment) |>
  as.data.frame()

sd_nca_dose <- sd_events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment) |>
  as.data.frame()

sd_conc_obj <- PKNCA::PKNCAconc(sd_nca_conc, Cc ~ time | treatment + id,
                                concu = "ng/mL", timeu = "h")
sd_dose_obj <- PKNCA::PKNCAdose(sd_nca_dose, amt ~ time | treatment + id, doseu = "mg")

sd_intervals <- data.frame(
  start     = 0,
  end       = 48,
  cmax      = TRUE,
  tmax      = TRUE,
  auclast   = TRUE,
  half.life = TRUE
)

sd_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(sd_conc_obj, sd_dose_obj, intervals = sd_intervals))

sd_summary <- as.data.frame(sd_nca$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median_value = stats::median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median_value)

knitr::kable(
  sd_summary |>
    dplyr::rename(
      "Treatment"           = treatment,
      "AUC0-48 (h*ng/mL)"   = auclast,
      "Cmax (ng/mL)"        = cmax,
      "t1/2 (h)"            = half.life,
      "Tmax (h)"            = tmax
    ),
  digits  = 3,
  caption = "Simulated single-dose median NCA parameters by food / polymorph condition."
)
```

Rini 2013 does not tabulate single-dose NCA values, but the exposure
ratios between these three arms are determined entirely by the Table 2
bioavailability effects, because `CL` is unchanged across the arms:

* `AUC(Form IV fasted) / AUC(Form IV fed) = 1 + 0.330 = 1.330`
  (Table 2 "Fasting on F, Form IV", footnote f).
* `AUC(Form XLI fed) / AUC(Form IV fed) = 1 - 0.121 = 0.879`
  (Table 2 "Form XLI on F", footnote g).

`ka` does not enter the AUC ratio, so any deviation isolates a
bioavailability transcription error. The ratio is asserted on a
**typical-value** (random-effects-zeroed) run rather than on the cohort
medians above: the three stochastic arms draw independent random effects,
so a ratio of their medians carries roughly 10% Monte Carlo noise and
could not support a tight bound. On the typical-value run the only
residual is the small part of the profile lying beyond the 48 h window,
which the faster fasted absorption shifts slightly.

```{r auc-ratios}
ratio_arms <- tibble::tribble(
  ~treatment,       ~FED, ~FORM_AXI_XLI,
  "Form IV fed",    1L,   0L,
  "Form IV fasted", 0L,   0L,
  "Form XLI fed",   1L,   1L
) |>
  dplyr::mutate(id = dplyr::row_number(), WT = 74.1, AGE_GT60 = 0L,
                RACE_JAPANESE = 0L, SMOKE = 0L)

ratio_obs <- c(0, seq(0.25, 12, by = 0.25), seq(13, 48, by = 1))

ratio_events <- dplyr::bind_rows(
  ratio_arms |> dplyr::mutate(time = 0, evid = 1L, amt = 5, cmt = "depot"),
  tidyr::expand_grid(id = ratio_arms$id, time = ratio_obs) |>
    dplyr::left_join(ratio_arms, by = "id") |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

ratio_sim <- rxode2::rxSolve(mod_typ, events = as.data.frame(ratio_events),
                             keep = "treatment") |>
  as.data.frame() |>
  dplyr::mutate(id = as.integer(as.character(id)))

ratio_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    ratio_sim |> dplyr::filter(!is.na(Cc)) |>
      dplyr::select(id, time, Cc, treatment) |> as.data.frame(),
    Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h"
  ),
  PKNCA::PKNCAdose(
    ratio_events |> dplyr::filter(evid == 1L) |>
      dplyr::select(id, time, amt, treatment) |> as.data.frame(),
    amt ~ time | treatment + id, doseu = "mg"
  ),
  intervals = data.frame(start = 0, end = 48, auclast = TRUE)
))

auc_by_arm <- as.data.frame(ratio_nca$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(treatment, PPORRES) |>
  tibble::deframe()

auc_ratio_table <- tibble::tibble(
  Ratio = c("Form IV fasted / Form IV fed", "Form XLI fed / Form IV fed"),
  `Simulated AUC0-48 ratio` = c(
    unname(auc_by_arm[["Form IV fasted"]] / auc_by_arm[["Form IV fed"]]),
    unname(auc_by_arm[["Form XLI fed"]]   / auc_by_arm[["Form IV fed"]])
  ),
  `Published F ratio (Table 2)` = c(1 + 0.330, 1 - 0.121)
) |>
  dplyr::mutate(`% difference` = 100 * (`Simulated AUC0-48 ratio` /
                                          `Published F ratio (Table 2)` - 1))

knitr::kable(
  auc_ratio_table,
  digits  = 4,
  caption = "Typical-value AUC0-48 ratios vs the published bioavailability ratios (Rini 2013 Table 2)."
)

stopifnot(
  # Form XLI shares the fed absorption profile with the Form IV reference,
  # so its AUC ratio is the F ratio exactly.
  abs(auc_ratio_table$`% difference`[2]) < 0.5,
  # The fasted arm differs in ka as well, so only the 48 h window truncation
  # separates it from the F ratio.
  abs(auc_ratio_table$`% difference`[1]) < 5
)
```

The Cmax and Tmax columns carry the `ka` effect that AUC does not: the
fasted arm's `ka` of 1.43 /h (Table 2 footnote e) versus 0.482 /h fed
moves the peak earlier and higher.

```{r ka-effect}
tmax_by_arm <- as.data.frame(sd_nca$result) |>
  dplyr::filter(PPTESTCD == "tmax") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(median_tmax = stats::median(PPORRES), .groups = "drop")

knitr::kable(
  tmax_by_arm |> dplyr::rename("Treatment" = treatment, "Median Tmax (h)" = median_tmax),
  digits  = 2,
  caption = "Median simulated Tmax: the fasted arm peaks earlier because ka is 197% higher."
)

stopifnot(
  # Directional: faster absorption when fasted must move the peak earlier.
  tmax_by_arm$median_tmax[tmax_by_arm$treatment == "Form IV fasted"] <
    tmax_by_arm$median_tmax[tmax_by_arm$treatment == "Form IV fed"]
)
```

# Virtual cohort: mRCC patients at steady state

The exposure-response analysis used the 168 evaluable patients of the
three mRCC studies. Their per-study design is taken from Table 1:

| Study | Population | Regimen | Enrolled |
|-------|------------|---------|----------|
| 2 | Cytokine-refractory mRCC | 5 mg bid, **fasted**, Form IV, dose titration | 52 |
| 4 | Sorafenib-refractory mRCC | 5 mg bid, **fed**, Form IV, dose titration | 62 |
| 5 | Japanese cytokine-refractory mRCC | 5 mg bid, **fed**, Form IV, dose titration | 64 |

Scaling the enrolled counts to the 168 analysed patients gives arms of
49 / 59 / 60, and fixes the Japanese fraction at 60/168 = 36%, which the
paper never states directly for this subset. Median age in the subset is
exactly 60 years (Table 3), so half the cohort is assigned
`AGE_GT60 = 1`. No smokers are simulated (3% of the pooled PK dataset;
Rini 2013 itself excluded the smoking effect from its covariate-impact
simulations because the estimate was too imprecise).

```{r cohort-mrcc}
set.seed(20130502)

mrcc_design <- tibble::tribble(
  ~study,                                       ~n,  ~FED, ~RACE_JAPANESE,
  "Study 2 (cytokine-refractory, fasted)",      49L, 0L,   0L,
  "Study 4 (sorafenib-refractory, fed)",        59L, 1L,   0L,
  "Study 5 (Japanese cytokine-refractory, fed)", 60L, 1L,   1L
)

mrcc_cov <- mrcc_design |>
  tidyr::uncount(n) |>
  dplyr::mutate(
    id           = dplyr::row_number(),
    WT           = pmax(40, rnorm(dplyr::n(), mean = 74, sd = 15)),
    AGE_GT60     = rbinom(dplyr::n(), 1, 0.5),
    SMOKE        = 0L,
    FORM_AXI_XLI = 0L,
    treatment    = "5 mg bid, steady state"
  )

n_days   <- 28L                                   # 4 weeks of treatment, per Table 3
tau      <- 12
dose_t   <- seq(0, 24 * n_days - tau, by = tau)
ss_lo    <- 24 * n_days - 24                      # the 24 h window contains 24/tau = 2 doses
ss_hi    <- 24 * n_days
obs_t    <- seq(ss_lo, ss_hi, by = 0.1)

mrcc_events <- dplyr::bind_rows(
  tidyr::expand_grid(id = mrcc_cov$id, time = dose_t) |>
    dplyr::left_join(mrcc_cov, by = "id") |>
    dplyr::mutate(evid = 1L, amt = 5, cmt = "depot"),
  tidyr::expand_grid(id = mrcc_cov$id, time = obs_t) |>
    dplyr::left_join(mrcc_cov, by = "id") |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(nrow(mrcc_cov) == 168L)
```

```{r simulate-mrcc}
mrcc_sim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(mrcc_events),
  keep   = c("study", "treatment")
) |>
  as.data.frame() |>
  dplyr::mutate(id = as.integer(as.character(id)))
```

## Steady-state NCA and the AUC identity

For a linear model at steady state, the AUC over one dosing interval is
exactly `dose * F / CL`, so the AUC over the 24 h window (which contains
both daily doses) is exactly `total daily dose * F / CL` -- the quantity
Rini 2013 Methods "PK/PD Modeling" defines as the mean daily AUC. Both
sides of the check below use the same drawn parameters, so the only
sources of difference are trapezoidal error and the small residual
approach to steady state after 28 days.

```{r pknca-mrcc}
mrcc_nca_conc <- mrcc_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment) |>
  as.data.frame()

# Only the two doses inside the analysis window are passed to PKNCA; the 54
# earlier doses have no concentration records and would put the AUC interval
# before the first measurement.
mrcc_nca_dose <- mrcc_events |>
  dplyr::filter(evid == 1L, time >= ss_lo) |>
  dplyr::select(id, time, amt, treatment) |>
  as.data.frame()

mrcc_conc_obj <- PKNCA::PKNCAconc(mrcc_nca_conc, Cc ~ time | treatment + id,
                                  concu = "ng/mL", timeu = "h")
mrcc_dose_obj <- PKNCA::PKNCAdose(mrcc_nca_dose, amt ~ time | treatment + id, doseu = "mg")

mrcc_intervals <- data.frame(
  start   = ss_lo,
  end     = ss_hi,
  cmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

mrcc_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(mrcc_conc_obj, mrcc_dose_obj, intervals = mrcc_intervals)
)

# Per-subject individual parameters carried out of the solve.
mrcc_pars <- mrcc_sim |>
  dplyr::group_by(id) |>
  dplyr::slice_head(n = 1) |>
  dplyr::ungroup() |>
  dplyr::select(id, study, cl, fdepot) |>
  dplyr::mutate(id = as.integer(as.character(id)))

auc_check <- as.data.frame(mrcc_nca$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, auc24_nca = PPORRES) |>
  dplyr::mutate(id = as.integer(as.character(id))) |>
  dplyr::left_join(mrcc_pars, by = "id") |>
  dplyr::mutate(
    # 10 mg total daily dose, mg / (L/h) -> mg*h/L, scaled to ng*h/mL.
    auc24_identity = 10 * fdepot / cl * 1000,
    pct_diff       = 100 * (auc24_nca - auc24_identity) / auc24_identity
  )

knitr::kable(
  tibble::tibble(
    Statistic = c("Median % difference", "90th percentile of |% difference|",
                  "Maximum |% difference|"),
    Value     = c(stats::median(auc_check$pct_diff),
                  unname(stats::quantile(abs(auc_check$pct_diff), 0.9)),
                  max(abs(auc_check$pct_diff)))
  ),
  digits  = 2,
  caption = "Simulated steady-state AUC0-24 vs the closed-form identity total daily dose * F / CL."
)

stopifnot(
  # Structural: a mis-transcribed CL, F or dose moves the whole distribution.
  abs(stats::median(auc_check$pct_diff)) < 2,
  # Envelope: robust to the subjects with the largest peripheral volumes,
  # which are the slowest to reach steady state after 28 days of dosing.
  stats::quantile(abs(auc_check$pct_diff), 0.9) < 8
)
```

## Comparison against the published exposure distribution

Rini 2013 Table 3 reports a median mean-daily AUC of 375 h\*ng/mL
(range 32.8-1,728) for the 168 patients at the end of 4 weeks of
treatment. Two simulated quantities are shown against it.

The **published-formula AUC** applies the paper's own equation using a
single population mean `F` for every patient, as Methods "PK/PD
Modeling" specifies ("the population mean estimate of F"), with the
Form IV fed value 0.457. The **model-simulated AUC0-24** instead uses
each patient's own record-level `F`, which is 33% higher for the fasted
study 2 arm; it is the model's prediction of true exposure and is
therefore expected to sit above the published number.

```{r published-auc}
auc_check <- auc_check |>
  dplyr::mutate(auc24_paper_formula = 10 * 0.457 / cl * 1000)

auc_compare <- tibble::tibble(
  Quantity = c("Published-formula AUC (single population F = 0.457)",
               "Model-simulated steady-state AUC0-24 (record-level F)"),
  `Simulated median (h*ng/mL)` = c(stats::median(auc_check$auc24_paper_formula),
                                   stats::median(auc_check$auc24_nca)),
  `Published median (h*ng/mL)` = 375
) |>
  dplyr::mutate(`% difference` = 100 * (`Simulated median (h*ng/mL)` /
                                          `Published median (h*ng/mL)` - 1))

knitr::kable(
  auc_compare,
  digits  = 1,
  caption = "Simulated vs published median mean-daily AUC in the 168-patient mRCC subset (Rini 2013 Table 3)."
)

stopifnot(
  # The paper's own AUC formula, evaluated on the model's individual
  # clearances, must land near the published median. This is a comparison of
  # cohort medians, not of extremes, so it is stable across rxode2 builds.
  abs(stats::median(auc_check$auc24_paper_formula) / 375 - 1) < 0.25
)
```

Per-study medians show where the pooled difference comes from: the
fasted study 2 arm carries the 33% bioavailability increase, and the
Japanese study 5 arm carries the 24.9% clearance reduction.

```{r auc-by-study}
auc_by_study <- auc_check |>
  dplyr::group_by(study) |>
  dplyr::summarise(
    n                     = dplyr::n(),
    median_auc24_nca      = stats::median(auc24_nca),
    median_auc24_formula  = stats::median(auc24_paper_formula),
    .groups = "drop"
  )

knitr::kable(
  auc_by_study |>
    dplyr::rename(
      "Study"                                       = study,
      "n"                                           = n,
      "Median simulated AUC0-24 (h*ng/mL)"          = median_auc24_nca,
      "Median published-formula AUC (h*ng/mL)"      = median_auc24_formula
    ),
  digits  = 1,
  caption = "Simulated mean-daily exposure by mRCC study arm."
)
```

## Replicating Supplementary Figure S4

Supplementary Figure S4 shows axitinib `AUC12` for a typical fed subject
in each of four covariate cells -- age > 60 or not, Japanese or not --
receiving 5 mg Form XLI twice daily. The four typical values are exactly
`5 mg * F(Form XLI) / CL(covariates)`, so the model reproduces the figure's
structure in closed form. The figure itself plots medians and 95%
prediction intervals as an image, so only the ratios are checked here.

```{r figure-s4}
fig_s4 <- tidyr::expand_grid(
  AGE_GT60      = c(0L, 1L),
  RACE_JAPANESE = c(0L, 1L)
) |>
  dplyr::mutate(
    id        = dplyr::row_number(),
    WT        = 74.1,
    SMOKE     = 0L,
    FED       = 1L,
    FORM_AXI_XLI = 1L,
    group     = paste0(ifelse(AGE_GT60 == 1L, ">60 y", "<=60 y"), ", ",
                       ifelse(RACE_JAPANESE == 1L, "Japanese", "non-Japanese"))
  )

fig_s4_events <- dplyr::bind_rows(
  fig_s4 |> dplyr::mutate(time = 0, evid = 1L, amt = 5, cmt = "depot"),
  fig_s4 |> dplyr::mutate(time = 1, evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

fig_s4_par <- rxode2::rxSolve(
  mod_typ,
  events = as.data.frame(fig_s4_events),
  keep   = "group"
) |>
  as.data.frame() |>
  dplyr::group_by(group) |>
  dplyr::slice_head(n = 1) |>
  dplyr::ungroup() |>
  # rxSolve returns the covariate columns alongside the model variables.
  dplyr::select(group, AGE_GT60, RACE_JAPANESE, cl, fdepot) |>
  dplyr::mutate(auc12 = 5 * fdepot / cl * 1000) |>
  dplyr::arrange(AGE_GT60, RACE_JAPANESE)

ref_auc12 <- fig_s4_par$auc12[fig_s4_par$AGE_GT60 == 0L & fig_s4_par$RACE_JAPANESE == 0L]

fig_s4_par <- fig_s4_par |>
  dplyr::mutate(
    ratio_vs_reference = auc12 / ref_auc12,
    expected_ratio     = 1 / ((1 - 0.213 * AGE_GT60) * (1 - 0.249 * RACE_JAPANESE))
  )

knitr::kable(
  fig_s4_par |>
    dplyr::select(group, auc12, ratio_vs_reference, expected_ratio) |>
    dplyr::rename(
      "Covariate cell"                        = group,
      "Typical AUC12 (h*ng/mL)"               = auc12,
      "Ratio vs <=60 y non-Japanese"          = ratio_vs_reference,
      "Expected from Table 2 CL effects"      = expected_ratio
    ),
  digits  = 3,
  caption = "Typical AUC12 for the four Supplementary Figure S4 covariate cells, 5 mg bid Form XLI fed."
)

stopifnot(isTRUE(all.equal(fig_s4_par$ratio_vs_reference,
                           fig_s4_par$expected_ratio, tolerance = 1e-8)))
```

Rini 2013 concludes from this simulation that "the magnitude of predicted
changes in exposure based on these covariates does not warrant dose
adjustments" (Abstract). The largest cell -- an over-60 Japanese subject
-- has a typical exposure `r round(max(fig_s4_par$ratio_vs_reference), 2)`
times the reference, which is far smaller than the 59.9% CV of clearance
itself, exactly as the Discussion argues.

## Body weight moves Cmax but not exposure

Rini 2013 Results "Effects of covariates" states that "expected changes in
peak plasma concentrations due to body weight were less than the estimated
IIV for Vc" and, crucially, that "exposure is not affected by changes in
Vc". Because weight acts only on `Vc`, the steady-state AUC identity is
weight-independent while Cmax is not -- a check the model must pass
exactly.

```{r weight-effect}
wt_grid <- tibble::tibble(
  WT    = c(58, 74.1, 94),                            # 10th percentile, median, 90th percentile
  label = c("58 kg (10th percentile)", "74.1 kg (median)", "94 kg (90th percentile)")
) |>
  dplyr::mutate(id = dplyr::row_number(), AGE_GT60 = 0L, RACE_JAPANESE = 0L,
                SMOKE = 0L, FED = 1L, FORM_AXI_XLI = 1L)

wt_events <- dplyr::bind_rows(
  tidyr::expand_grid(id = wt_grid$id, time = dose_t) |>
    dplyr::left_join(wt_grid, by = "id") |>
    dplyr::mutate(evid = 1L, amt = 5, cmt = "depot"),
  tidyr::expand_grid(id = wt_grid$id, time = obs_t) |>
    dplyr::left_join(wt_grid, by = "id") |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

wt_solved <- rxode2::rxSolve(mod_typ, events = as.data.frame(wt_events), keep = "label") |>
  as.data.frame() |>
  dplyr::mutate(id = as.integer(as.character(id)))

wt_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    wt_solved |> dplyr::filter(!is.na(Cc)) |>
      dplyr::select(id, time, Cc, label) |> as.data.frame(),
    Cc ~ time | label + id, concu = "ng/mL", timeu = "h"
  ),
  PKNCA::PKNCAdose(
    wt_events |> dplyr::filter(evid == 1L, time >= ss_lo) |>
      dplyr::select(id, time, amt, label) |> as.data.frame(),
    amt ~ time | label + id, doseu = "mg"
  ),
  intervals = data.frame(start = ss_lo, end = ss_hi,
                         cmax = TRUE, cmin = TRUE, auclast = TRUE)
))

wt_sim <- as.data.frame(wt_nca$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) |>
  dplyr::select(label, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::rename(cmax_ss = cmax, cmin_ss = cmin, auc24 = auclast) |>
  dplyr::arrange(match(label, wt_grid$label))

knitr::kable(
  wt_sim |>
    dplyr::rename(
      "Body weight"            = label,
      "Cmax,ss (ng/mL)"        = cmax_ss,
      "Cmin,ss (ng/mL)"        = cmin_ss,
      "AUC0-24,ss (h*ng/mL)"   = auc24
    ),
  digits  = 2,
  caption = "Typical-value steady-state exposure at three body weights, 5 mg bid Form XLI fed."
)
```

As an external cross-check, the companion healthy-volunteer analysis
(Garrett 2014 Results "Full model and final model") publishes typical
steady-state Cmax values for exactly this regimen at three body weights:
20.6 ng/mL at 62 kg, 19.3 ng/mL at 75 kg and 17.9 ng/mL at 93 kg. The
Rini 2013 model reproduces the same magnitude and the same weight trend
from an independently estimated parameter set. That is corroboration
across two papers, not a Rini 2013 claim, so it is not asserted.

```{r weight-effect-assert}
stopifnot(
  # AUC is weight-independent: identical to well within trapezoidal error.
  diff(range(wt_sim$auc24)) / stats::median(wt_sim$auc24) < 0.005,
  # Cmax decreases with weight, because Vc increases with weight.
  wt_sim$cmax_ss[wt_sim$label == "58 kg (10th percentile)"] >
    wt_sim$cmax_ss[wt_sim$label == "94 kg (90th percentile)"]
)
```

# Exposure-response linkage

The Cox and logistic regression models of Rini 2013 are not encoded in
the model file (see Errata), but the packaged PK model supplies the
exposure metric they take as input, so the published coefficients can be
applied to the simulated cohort to reproduce the paper's relative
statements. Table 5 gives, after adjustment for gender, prior therapy,
ECOG PS and haemoglobin, a hazard ratio of 0.909 per 100 h\*ng/mL of AUC
for progression-free survival and 0.866 for overall survival.

```{r exposure-response}
auc_iqr <- stats::quantile(auc_check$auc24_paper_formula, c(0.25, 0.75))

hr_table <- tibble::tibble(
  Endpoint = c("Progression-free survival", "Overall survival"),
  `HR per 100 h*ng/mL (Table 5)` = c(0.909, 0.866),
  `HR across the simulated interquartile AUC range` =
    c(0.909, 0.866)^(unname(diff(auc_iqr)) / 100)
)

knitr::kable(
  hr_table,
  digits  = 3,
  caption = sprintf(
    paste("Published Table 5 hazard ratios applied across the simulated",
          "interquartile AUC range (%.0f to %.0f h*ng/mL)."),
    auc_iqr[1], auc_iqr[2]
  )
)
```

Rini 2013 also stratified patients at 300 h\*ng/mL, reporting 38% of the
168 patients below the cutoff and 62% at or above it (Table 3). The
simulated cohort's split is shown below. The paper's split is computed
from post-hoc empirical Bayes clearances at the patients' actual
(sometimes reduced) daily doses, while the simulation holds every patient
at 5 mg twice daily, so the simulated fraction above the cutoff is
expected to be somewhat higher.

```{r auc-cutoff}
cutoff_table <- tibble::tibble(
  Group = c("AUC < 300 h*ng/mL", "AUC >= 300 h*ng/mL"),
  `Simulated %` = c(
    100 * mean(auc_check$auc24_paper_formula <  300),
    100 * mean(auc_check$auc24_paper_formula >= 300)
  ),
  `Published % (Table 3)` = c(38, 62)
)

knitr::kable(cutoff_table, digits = 1,
             caption = "Distribution around the 300 h*ng/mL exposure cutoff of Rini 2013 Table 3.")
```

# Assumptions and deviations

* **Interindividual variability scale.** Rini 2013 Table 2 footnote a
  reports the IIV as "%CV" without stating the back-transformation. The
  packaged model takes `omega^2 = (%CV / 100)^2`, following the
  convention used by the same modelling group in the companion
  healthy-volunteer analysis (Garrett 2014), whose Methods print the
  definition verbatim -- `%IIV = sqrt(omega^2) x 100` -- and whose
  Table 3 satisfies it numerically: `omega^2(Vc) = 0.0949` gives
  `sqrt(0.0949) = 30.8%`, exactly the 30.8% Vc IIV that paper's
  Discussion quotes, whereas the exact-log-normal back-transformation
  would give 31.6% and does not match. The alternative reading -- `%CV` as the
  exact log-normal coefficient of variation, `omega^2 = log(1 + CV^2)` --
  would give `omega = 0.657` rather than `0.599` for clearance, about 10%
  larger. The choice changes only the width of the simulated
  distributions; it changes no typical value, no covariate effect, and
  none of the structural identities checked above. It does slightly widen
  or narrow the VPC bands of Figure 1 and the 95% prediction intervals of
  Supplementary Figures S4 and S5.
* **Route-stratified residual error not encoded.** Rini 2013 fit two
  residuals on the log-concentration scale -- oral 58.2% and intravenous
  33.5% (Table 2). The packaged model encodes a single `expSd = 0.582`,
  the oral value, because a single rxode2 endpoint carries a single error
  model and the library use of this model is oral dosing; the intravenous
  arm existed only to anchor absolute bioavailability (16 subjects,
  study 8). Users simulating the intravenous arm should override
  `expSd <- 0.335` after `readModelDb()`.
* **Residual error form.** Methods "Model development" states that
  log-transformed concentrations were modelled with a proportional
  residual. An additive residual on the log-concentration scale is a
  log-normal residual on the linear scale, which is how it is encoded
  (`Cc ~ lnorm(expSd)`); at 58.2% the log-normal and the linear-scale
  proportional forms are not interchangeable, and the log-scale reading
  is the one the Methods sentence supports.
* **Fasting effect on F is gated to Form IV.** Table 2's row label is
  "Fasting on F, Form IV" and footnote g states explicitly that "there
  was no observed food effect with Form XLI", giving `F = 0.402` for
  Form XLI in either prandial state. The model therefore multiplies the
  fasting term by `(1 - FORM_AXI_XLI)`. The fasting effect on `ka`, by
  contrast, is applied to both polymorphs, because Results "Effects of
  covariates" says it holds "regardless of formulations".
* **Q and Vp share one random effect.** Table 2 footnote d states that Q
  and Vp were modelled as 100% correlated with the same interindividual
  variability. That is encoded as a single shared `etalq_lvp` rather than
  a 2x2 block with unit correlation, which would be singular and would
  fail Cholesky decomposition at simulation time.
* **Covariate distributions of the virtual cohorts are assumptions.**
  Rini 2013 reports no per-subject covariate table. Body weight is drawn
  as `Normal(74, 15)` truncated below at 40 kg to match the reported
  median 74 kg and range 37-136 kg; the mRCC cohort's `AGE_GT60` is
  drawn at 50% because the subset's median age is exactly 60 years
  (Table 3); the Japanese fraction of 36% is derived from Table 1
  (study 5 contributed 64 of the 178 enrolled mRCC patients); no smokers
  are simulated. Weight is immaterial to every exposure comparison here,
  because it acts only on `Vc`.
* **Steady state is approached, not reached exactly.** The peripheral
  volume of 393 L against an inter-compartmental clearance of 4.00 L/h
  gives a long, low-amplitude terminal phase, so 28 days of 5 mg bid
  dosing leaves a small accumulation deficit. That is the entire content
  of the median -0.2% and 90th-percentile 2.5% deviations in the AUC
  identity check; the subjects furthest from steady state are those whose
  shared Q / Vp random effect gave them the largest peripheral volume.
  Twenty-eight days matches the paper's own "end of 4 weeks of study
  treatment" exposure window.
* **Dose titration is not simulated.** Patients could titrate to 10 mg bid
  or have doses reduced or interrupted for hypertension, and Rini 2013
  computed each patient's AUC from their "average total daily dose". The
  simulated cohort holds every patient at 5 mg bid, which is what the
  paper describes as the situation at the end of 4 weeks "prior to dose
  changes in most patients".
* **VPC bands are simulated only.** The Figure 1 replication plots
  percentiles of the `sim` column, which carries both interindividual and
  the 58.2% oral residual variability, so it is the right analogue of the
  published observed-percentile bands. What it cannot reproduce is the
  observed data themselves, or the published bands' truncation at the
  0.1 ng/mL LLOQ; the LLOQ is drawn as a reference line instead. The NCA
  blocks use `Cc` (individual predictions without residual error), which
  is what makes the structural AUC identities exact.

# Errata and unencoded sub-models

* **No erratum located.** A search of the journal's correction notices
  and of PubMed for corrections to `doi:10.1002/jcph.73` returned
  nothing; no erratum is referenced by the article record.
* **The Cox proportional-hazards models are not encoded.** Rini 2013
  Table 5 reports regression coefficients and hazard ratios for
  progression-free and overall survival as functions of AUC, diastolic
  blood pressure, gender, prior therapy, ECOG PS and haemoglobin, but it
  reports no baseline hazard function and no baseline survival curve. A
  Cox model's partial likelihood does not estimate the baseline hazard,
  so a semi-parametric fit reported this way cannot be turned into a
  simulatable time-to-event model without information the paper does not
  contain. The published coefficients are applied above as *relative*
  statements, which is all they support.
* **The logistic regression models are not encoded.** Rini 2013 reports
  only the odds-ratio slopes -- a 1.5-fold increase in the probability of
  a partial response per 100 h\*ng/mL of AUC, and 1.6-fold per 10 mm Hg
  of diastolic blood pressure -- and marks the underlying fits as "data
  not shown". Without an intercept the absolute response probability is
  not recoverable.
* **No diastolic blood pressure model.** Diastolic blood pressure enters
  the paper only as an observed covariate in the regression analyses.
  Rini 2013 reports a weak relationship between exposure and blood
  pressure (`r^2 < 0.10`, "data not shown") and no structural
  pharmacodynamic model linking axitinib concentration to blood pressure.
* **Base-model IIV values are prose-only.** The Discussion reports that
  the covariate model reduced CL IIV from 64% to 60% and Vc IIV from 44%
  to 40%. The base model is not otherwise tabulated and is not packaged.
