---
title: "Robbie_2012_palivizumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robbie_2012_palivizumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Palivizumab population PK simulations

Replicates key figures and parameter reference values from Robbie et al. (2012),
"Population Pharmacokinetics of Palivizumab, a Humanized Anti-Respiratory
Syncytial Virus Monoclonal Antibody, in Adults and Children"
(Antimicrob Agents Chemother 56:4927-4936; erratum AAC 56:5431 adds
Anderson/Allegaert/Holford 2006 as the maturation-model reference).

- **Parameter reference values**: reproduce Table 2 (adult, 70 kg, 40-week PAGE)
  and Table 3 (pediatric reference: 4.5 kg, 12.3-month PAGE).
- **Figure 3 (corrected per erratum)**: simulated concentration-time profiles
  for 5 monthly 15 mg/kg IM doses vs. an abbreviated 3 monthly dose regimen.
- **Figure 5 (corrected per erratum)**: group median concentration-time
  profiles stratified by gestational age + postnatal age cohorts.
- **PKNCA exposure summary**: steady-state Cmax / Cmin / AUC0-tau over the
  label 5-monthly regimen.

Because the original study data are not publicly available, these figures show
simulated prediction intervals from virtual infant populations whose covariate
distributions approximate the published trial demographics (Robbie 2012 Table 1).
Body weight trajectories are generated using WHO weight-for-age growth standards
(combined sex, 0-24 months).

## Population

Baseline demographics reproduced from Robbie 2012 Table 1:

| Cohort | N | Key covariates | PK samples |
|---|---|---|---|
| Adults | 116 | age 19.3-58.9 y (mean 32.8); weight 48.9-104.3 kg (mean 72.5); 65.5% female | 1,661 |
| Pediatric | 1,767 | PAGE 6.89-34.7 mo (median 12.6); weight 0.92-16.3 kg (median 5.64); gestational age 22-41 wk (median 30; 62% preterm); 36% CLD of prematurity; 54% White, 20% Black, 18% Hispanic, 2% Asian, 6% Other | 4,095 |

The pooled dataset spans 22 clinical studies (9 adult: 7 healthy-volunteer + 2
transplant; 13 pediatric). Adult cohort was dosed IV or IM across studies;
pediatric cohort was dosed IM. Dose range 3-30 mg/kg; label regimen is 5
monthly doses of 15 mg/kg IM.

## Source trace

| Element | Source location |
|---|---|
| 2-compartment structural model with first-order IM absorption | Robbie 2012, Methods, "Population pharmacokinetic model" subsection (ADVAN4 TRANS4). |
| Allometric weight scaling (CL/Q exp 0.75; Vc/Vp exp 1.0; 70 kg reference) | Robbie 2012, eq. (3) and Table 2 footnote. |
| PAGE definition PAGE = AGE(months) + GA(weeks)/4.35 | Robbie 2012, eq. (1). |
| Maturation on CL (asymptotic exponential centered at 40-week PAGE) | Robbie 2012, eq. (1a) / eq. (4); Anderson/Allegaert/Holford 2006 parameterization added as reference 1a by the 2012 erratum (AAC 56:5431 / PMC3457364). |
| CL, Vc, Vp, Q, ka, F1 final estimates (adult reference, 70 kg) | Robbie 2012, Table 2. |
| beta, TCL maturation parameters | Robbie 2012, Table 2 ("maturation parameters"). |
| Race effects on CL (Black, Hispanic, Asian, Other) and on Vc (Hispanic) | Robbie 2012, Table 2. |
| Chronic lung disease of prematurity effect on CL (+20%) | Robbie 2012, Table 2 ("theta_CLD" row). |
| ADA titer step-function effect on CL (4 bins: 10, 20, 40, >=80) | Robbie 2012, Table 2 ("theta_titer" rows). |
| IIV: CL CV 48.7%, Vc CV 61.7%, correlation rho = 0.62 | Robbie 2012, Table 2. |
| Residual proportional error, sigma^2 = 0.0639 | Robbie 2012, Table 2. |
| Pediatric reference scaling (4.5 kg, 12.3-month PAGE) | Robbie 2012, Table 3. |
| Figure 3 (corrected per erratum): 3- vs 5-dose profiles | Robbie 2012, Figure 3 (PDF page 7); erratum replaces the original panel. |
| Figure 5 (corrected per erratum): profiles stratified by postnatal age cohort | Robbie 2012, Figure 5 (PDF page 9); erratum updates legend. |
| Equation 10 on page 4931, left column: 16,900 | Robbie 2012 erratum; corrects original "6,900" value. |

## WHO weight-for-age growth curve helper

WHO weight-for-age LMS parameters for combined sex, 0-24 months
(WHO Multicentre Growth Reference Study Group, 2006). Formula:
weight (kg) = M * (1 + L * S * z)^(1/L).

```{r who-lms}
who_lms <- data.frame(
  age_mo = 0:24,
  L = c(0.3487, 0.2297, 0.1970, 0.1738, 0.1553, 0.1395, 0.1257,
    0.1125, 0.0998, 0.0875, 0.0756, 0.0640, 0.0527, 0.0418,
    0.0313, 0.0211, 0.0113, 0.0018, -0.0073, -0.0161, -0.0245,
    -0.0326, -0.0404, -0.0479, -0.0551),
  M = c(3.3464,  4.4709,  5.5675,  6.3762,  7.0023,  7.5105,  7.9340,
    8.2970,  8.6151,  8.9014,  9.1649,  9.4122,  9.6479,  9.8749,
    10.0953, 10.3108, 10.5228, 10.7319, 10.9385, 11.1430, 11.3462,
    11.5480, 11.7478, 11.9459, 12.1424),
  S = c(0.14602, 0.13395, 0.12385, 0.11876, 0.11535, 0.11254, 0.11056,
    0.10947, 0.10868, 0.10814, 0.10765, 0.10722, 0.10706, 0.10695,
    0.10695, 0.10700, 0.10710, 0.10723, 0.10737, 0.10754, 0.10773,
    0.10793, 0.10814, 0.10835, 0.10858)
)
who_weight <- function(pna_mo, z) {
  pna_mo <- pmax(0, pmin(pna_mo, 24))
  L <- approx(who_lms$age_mo, who_lms$L, xout = pna_mo)$y
  M <- approx(who_lms$age_mo, who_lms$M, xout = pna_mo)$y
  S <- approx(who_lms$age_mo, who_lms$S, xout = pna_mo)$y
  M * (1 + L * S * z)^(1 / L)
}
```

## Load model and confirm reference parameter values

```{r load-model}
mod <- readModelDb("Robbie_2012_palivizumab")
```

Sanity-check reference values against the paper's Tables 2 and 3: a
single typical subject (no between-subject variability, no dose)
should yield the published per-subject CL, Vc, Vp, Q at each reference.

```{r ref-values}
mod0 <- rxode2::zeroRe(mod)

ref_ev <- function(WT, PAGE) {
  data.frame(
    ID = 1, TIME = c(0, 1e-6),
    AMT = c(1, 0),
    EVID = c(1, 0),
    CMT = c("depot", "central"),
    DV = NA,
    WT = WT, PAGE = PAGE,
    RACE_BLACK = 0, RACE_HISPANIC = 0, RACE_ASIAN = 0, RACE_OTHER = 0,
    CLD_PREM = 0, ADA_TITER = 0
  )
}

adult <- as.data.frame(rxode2::rxSolve(mod0, events = ref_ev(70, 12 * 30 + 9)))
ped   <- as.data.frame(rxode2::rxSolve(mod0, events = ref_ev(4.5, 12.3)))

ref_table <- data.frame(
  parameter = c("CL (mL/day)", "Vc (mL)", "Vp (mL)", "Q (mL/day)"),
  paper_adult  = c(198, 4090, 2230, 879),
  model_adult  = round(c(adult$cl[1] * 1000, adult$vc[1] * 1000,
                          adult$vp[1] * 1000, adult$q[1] * 1000), 1),
  paper_pediatric = c(11.0, 263, 143, 112),
  model_pediatric = round(c(ped$cl[1] * 1000, ped$vc[1] * 1000,
                             ped$vp[1] * 1000, ped$q[1] * 1000), 1)
)
knitr::kable(ref_table,
  caption = "Reference parameter reproduction against Robbie 2012 Tables 2 and 3.")
```

Small deviations at the adult reference (~1%) arise because the adult
reference in Table 2 is reported without the maturation multiplier; the
model output here evaluates the full CL expression at a typical-adult PAGE.

## Helper: build a virtual pediatric cohort

```{r make-cohort}
make_cohort <- function(n, ga_range, pna0_range, max_day,
                        cld_prob = 0, ada_titer_fn = function(n) rep(0, n),
                        obs_days = seq(0, max_day, by = 7),
                        race_probs = c(White = 0.54, Black = 0.20,
                                       Hispanic = 0.18, Asian = 0.02, Other = 0.06),
                        n_doses = 5, dose_mg_per_kg = 15) {
  GA <- runif(n, ga_range[1], ga_range[2])
  PNA_0 <- runif(n, pna0_range[1], pna0_range[2])
  wt_z <- pmax(-2, pmin(2, rnorm(n, 0, 1)))
  WT_0 <- who_weight(PNA_0, wt_z)
  CLD_PREM <- as.integer(runif(n) < cld_prob)
  ADA_T <- ada_titer_fn(n)
  race <- sample(names(race_probs), n, replace = TRUE, prob = race_probs)

  pop <- data.frame(
    ID = seq_len(n), GA, PNA_0, wt_z, WT_0,
    RACE_BLACK    = as.integer(race == "Black"),
    RACE_HISPANIC = as.integer(race == "Hispanic"),
    RACE_ASIAN    = as.integer(race == "Asian"),
    RACE_OTHER    = as.integer(race == "Other"),
    CLD_PREM = CLD_PREM, ADA_TITER = ADA_T
  )

  # Dose records: n_doses at 30-day intervals, 15 mg/kg of current weight
  dose_times <- (0:(n_doses - 1)) * 30
  d_dose <- pop[rep(seq_len(n), each = length(dose_times)), ] |>
    mutate(
      TIME = rep(dose_times, times = n),
      EVID = 1, CMT = "depot", DV = NA,
      PNA = PNA_0 + TIME / 30.44,
      PAGE = GA / 4.35 + PNA,
      WT = who_weight(pmax(0, PNA), wt_z),
      AMT = dose_mg_per_kg * WT
    )

  # Observation records: weekly sampling
  d_obs <- pop[rep(seq_len(n), each = length(obs_days)), ] |>
    mutate(
      TIME = rep(obs_days, times = n),
      EVID = 0, CMT = "central", DV = NA, AMT = 0,
      PNA = PNA_0 + TIME / 30.44,
      PAGE = GA / 4.35 + PNA,
      WT = who_weight(pmax(0, PNA), wt_z)
    )

  bind_rows(d_dose, d_obs) |>
    arrange(ID, TIME, desc(EVID)) |>
    select(ID, TIME, AMT, EVID, CMT, DV, WT, PAGE,
      RACE_BLACK, RACE_HISPANIC, RACE_ASIAN, RACE_OTHER,
      CLD_PREM, ADA_TITER, WT_0, GA, PNA_0)
}
```

## Figure 3 (erratum-corrected) - 3 vs 5 monthly doses

Replicates the erratum-corrected Figure 3: simulated concentration-time
profiles from the abbreviated 3 monthly 15 mg/kg IM dose regimen compared
with the label 5 monthly doses, in a typical pediatric cohort.

```{r fig3-sim}
set.seed(22802243) # PMID seed
n_f3 <- 300
obs_days_f3 <- seq(0, 240, by = 2)

ev_5 <- make_cohort(n_f3, ga_range = c(26, 41), pna0_range = c(0, 6),
                    max_day = 240, cld_prob = 0.2,
                    obs_days = obs_days_f3, n_doses = 5) |>
  mutate(regimen = "5 monthly doses (label)")
ev_3 <- make_cohort(n_f3, ga_range = c(26, 41), pna0_range = c(0, 6),
                    max_day = 240, cld_prob = 0.2,
                    obs_days = obs_days_f3, n_doses = 3) |>
  mutate(regimen = "3 monthly doses (abbreviated)")

sim_f3 <- bind_rows(ev_5, ev_3)

out_f3 <- rxode2::rxSolve(mod, events = sim_f3) |>
  as.data.frame() |>
  left_join(sim_f3 |> select(ID, regimen) |> distinct(),
            by = c("id" = "ID"))

d_f3 <- out_f3 |>
  group_by(regimen, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r fig3-plot, fig.width = 7, fig.height = 4.5}
ggplot(d_f3, aes(x = time, y = Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 40, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 235, y = 42, hjust = 1, vjust = 0,
           label = "40 ug/mL target", size = 3, colour = "grey40") +
  scale_colour_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(0, 240, by = 30)) +
  labs(
    x = "Time after first dose (days)",
    y = "Palivizumab serum concentration (ug/mL)",
    title = "Figure 3 (erratum-corrected) - 3 vs 5 monthly 15 mg/kg IM doses",
    caption = paste0(
      "Simulated median (line) and 90% prediction interval (shaded) in ",
      n_f3, " virtual infants per regimen.\n",
      "Dashed line: 40 ug/mL nominal target."
    ),
    colour = NULL, fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Figure 5 (erratum-corrected) - profiles stratified by postnatal age

The erratum-corrected legend specifies "Group median concentration-time
profiles of palivizumab at 15 mg/kg stratified by gestational age and
postnatal age cohorts, 0 to 6 months (A), 7 to 12 months (B), and 13 to
24 months (C)." Replicate with three cohorts differing in postnatal age
at first dose.

```{r fig5-sim}
set.seed(4927) # first journal page
cohorts <- list(
  list(label = "A. Postnatal 0 to 6 mo",   pna = c(0,  6)),
  list(label = "B. Postnatal 7 to 12 mo",  pna = c(7, 12)),
  list(label = "C. Postnatal 13 to 24 mo", pna = c(13, 24))
)

obs_days_f5 <- seq(0, 180, by = 2)

sim_f5 <- bind_rows(lapply(cohorts, function(co) {
  make_cohort(200, ga_range = c(26, 41), pna0_range = co$pna,
              max_day = 180, cld_prob = 0.2,
              obs_days = obs_days_f5, n_doses = 5) |>
    mutate(cohort = co$label)
}))

out_f5 <- rxode2::rxSolve(mod, events = sim_f5) |>
  as.data.frame() |>
  left_join(sim_f5 |> select(ID, cohort) |> distinct(),
            by = c("id" = "ID"))

d_f5 <- out_f5 |>
  group_by(cohort, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r fig5-plot, fig.width = 8, fig.height = 4.5}
ggplot(d_f5, aes(x = time, y = Q50, colour = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_continuous(breaks = seq(0, 180, by = 30)) +
  labs(
    x = "Time after first dose (days)",
    y = "Palivizumab serum concentration (ug/mL)",
    title = "Figure 5 (erratum-corrected) - Profiles by postnatal age cohort at first dose",
    caption = "Simulated median (line) and 90% prediction interval (shaded). N = 200 per cohort, 5 monthly 15 mg/kg IM doses.",
    colour = NULL, fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## PKNCA exposure summary for the 5-monthly regimen

Compute steady-state Cmax, Cmin, and AUC0-tau over the last dosing interval
for the Figure 3 5-dose cohort, using PKNCA (not inline trapezoidal
integration) as the convention for validation vignettes. The grouping
variable keeps adult vs pediatric separation possible for future multi-group
comparisons; here we carry a single `treatment` level.

```{r pknca}
tau <- 30
last_dose_time <- 4 * tau

conc_df <- out_f3 |>
  as.data.frame() |>
  filter(regimen == "5 monthly doses (label)", !is.na(Cc)) |>
  mutate(treatment = "pediatric_5dose_15mpk") |>
  select(id, time, Cc, treatment)

dose_df <- ev_5 |>
  filter(EVID == 1) |>
  mutate(treatment = "pediatric_5dose_15mpk") |>
  select(id = ID, time = TIME, amt = AMT, treatment)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start   = last_dose_time,
  end     = last_dose_time + tau,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-summary}
nca_df <- as.data.frame(res$result)

nca_summary <- nca_df |>
  group_by(PPTESTCD) |>
  summarise(
    median = round(median(PPORRES, na.rm = TRUE), 2),
    q05    = round(quantile(PPORRES, 0.05, na.rm = TRUE), 2),
    q95    = round(quantile(PPORRES, 0.95, na.rm = TRUE), 2),
    .groups = "drop"
  )
knitr::kable(nca_summary,
  caption = "Steady-state NCA (final dosing interval) of the 5-monthly 15 mg/kg IM regimen in virtual pediatric infants.")
```

The paper does not tabulate individual pediatric NCA results, so this table
serves as a forward-prediction of steady-state exposure under the label
regimen; reviewers can compare qualitatively against the 40 ug/mL target
cited in the paper's discussion.

## Assumptions and deviations

- **Maturation parameterization.** The paper's eq. 4 / 1a reads
  `TVCL = theta_CL * (1 - beta * exp(...))` but the reported beta = 0.411
  and TCL = 62.3 months only reproduce the pediatric Table 3 reference
  (CL = 11.0 mL/day at 4.5 kg, 12.3-mo PAGE) and the abstract's 10.2 to
  11.9 mL/day PAGE 7-18 month range when the equation is taken as
  `(1 - (1-beta) * exp(...))`, i.e., beta = fraction of adult CL at
  term = 0.411. This is the Anderson / Allegaert / Holford (2006)
  parameterization added by the 2012 erratum as reference 1a. The model
  file uses this form and documents the choice in the ini() block and
  model() comments.
- **ADA titer as a step function.** Robbie 2012 modelled ADA titer as
  four separate multiplicative bins (10, 20, 40, >=80) rather than as a
  continuous trend. The implementation derives the bin indicators inside
  model() from the continuous canonical `ADA_TITER` column so that users
  supply a single column. Values 0 (ADA-negative) map to the reference
  theta = 1.
- **Hispanic race on Vc.** A Hispanic effect on Vc (1.06, CI included
  unity) is reported alongside the CL race effects; it is included for
  structural fidelity.
- **Residual variance.** The paper reports sigma^2_prop = 0.0639 and a
  proportional-error CV of 23.4%. The model uses `propSd = sqrt(0.0639) =
  0.2528` (25.3%); the 23.4% figure appears to be a derived summary rather
  than the fundamental estimate.
- **Race and weight for the virtual cohort.** Race distribution matches
  Robbie 2012 Table 1 pediatric proportions (54/20/18/2/6 White/Black/
  Hispanic/Asian/Other); weight uses WHO combined-sex LMS with each
  infant's z-score held constant.
- **Original observed data not available.** All figures show simulated
  prediction intervals only.

## Reference

* `r rxode2::rxode(readModelDb("Robbie_2012_palivizumab"))$reference`
