---
title: "Meropenem (Romano-Aguilar 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meropenem (Romano-Aguilar 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
mod <- readModelDb("RomanoAguilar_2022_meropenem")
ui <- rxode2::rxode(mod)
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1093/ofid/ofac492.663>

Romano-Aguilar and colleagues developed a population PK model of meropenem in
critically ill Mexican adults and used Monte Carlo simulation to propose initial
dosing regimens by creatinine clearance for time-above-MIC targets of 50% and
100% of the dosing interval. The work is published as an IDWeek 2022 poster
abstract. Its parameter table (Table 2), demographics (Table 1), dosing
recommendations (Table 3) and visual predictive check (Figure 1) are printed as
image panels in the abstract. A one-compartment model described the data, with
clearance proportional to Cockcroft-Gault creatinine clearance.

This vignette checks the packaged encoding against closed-form results, then
compares a virtual cohort with the observed medians of Figure 1. It also
re-simulates the probability of target attainment (PTA) of every Table 3
regimen.

# Population

Seventy-eight critically ill adults (45 female, 33 male) were enrolled in a
prospective observational study at Hospital Central "Dr. Ignacio Morones
Prieto", San Luis Potosi, Mexico. Median age was 47 years (range 17-92), weight
67 kg (40-98), height 1.62 m (1-1.9), serum creatinine 0.93 mg/dL (0.1-5.11) and
Cockcroft-Gault creatinine clearance 103.14 mL/min/1.73 m^2 (13.5-241.4). Daily
meropenem doses ranged from 500 to 6000 mg (3000 mg in 52 patients), mostly
every 8 h (60 patients), as intravenous infusions over 0.5 h (33), 1 h (16),
2 h (6) or 3 h (23). Blood was sampled pre-dose and 1, 3 and 6 h post-dose, and
plasma meropenem was measured by HPLC (abstract Methods and Table 1).

The same information is available programmatically via the model's `population`
metadata (`readModelDb("RomanoAguilar_2022_meropenem")()$population`).

# Source trace

The per-parameter origin is also recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/RomanoAguilar_2022_meropenem.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (typical CL at CRCL = 102.23 mL/min/1.73 m^2) | 11.9 L/h | Table 2 theta1 (RSE 8%, bootstrap median 11.93); Results text |
| `lvc` (volume of distribution) | 25.2 L | Table 2 theta2 (RSE 11%, bootstrap median 24.92); Results text |
| `var(etalcl)` | log(1 + 0.562^2) = 0.2745 | Table 2 IIV on CL 56.2 CV% (see Assumptions) |
| `var(etalvc)` | log(1 + 0.47^2) = 0.1996 | Table 2 IIV on V 47 CV% (see Assumptions) |
| `addSd` | 3.53 mg/L | Table 2 residual variability sigma, ug/mL |
| `CL = 11.9 * (CRCL / 102.23) * exp(eta_CL)` | n/a | Results 'CL (L/h) = 11.9 * (CLCr/102.23)'; Table 2 'CL = theta1 x (CLCr/102)' |
| `V = 25.2 * exp(eta_V)` | n/a | Results 'V (L) = 25.2'; Table 2 |
| one-compartment model, linear elimination | n/a | Results 'best described by a one compartment model' |
| `Cc ~ add(addSd)` | n/a | Table 2 residual variability in ug/mL |

# Structural verification

These checks do not depend on a random draw, so tight bounds are correct.

## Variance scale against Table 2

Table 2 reports both interindividual variabilities as CV%. Back-transforming
the packaged variances with `CV = sqrt(exp(omega^2) - 1)` must return the
printed values.

```{r check-variance}
om <- ui$omega
var_chk <- data.frame(
  term = c("IIV on CL", "IIV on V"),
  packaged_variance = c(om["etalcl", "etalcl"], om["etalvc", "etalvc"]),
  table2_cv_pct = c(56.2, 47)
) |>
  dplyr::mutate(cv_pct_from_variance = 100 * sqrt(exp(packaged_variance) - 1))

knitr::kable(
  var_chk |>
    dplyr::rename("Term" = term,
                  "Packaged variance" = packaged_variance,
                  "Table 2 CV (%)" = table2_cv_pct,
                  "sqrt(exp(v) - 1) x 100" = cv_pct_from_variance),
  digits = c(0, 5, 1, 3),
  caption = "Packaged variances back-transformed to the Table 2 CV scale."
)

# The variances are stored to five significant figures.
stopifnot(max(abs(var_chk$cv_pct_from_variance - var_chk$table2_cv_pct)) < 0.01)
```

## Covariate relation

The solved individual clearance must equal the Results equation.

```{r check-covariate}
typ <- rxode2::zeroRe(ui)
crcl_grid <- c(13.5, 50, 102.23, 150, 241.4)
ev_cov <- do.call(rbind, lapply(seq_along(crcl_grid), function(i) {
  data.frame(id = i, time = c(0, 1), amt = c(1000, 0), rate = c(2000, 0),
             evid = c(1, 0), cmt = "central", CRCL = crcl_grid[i])
}))
cov_chk <- rxode2::rxSolve(typ, ev_cov, returnType = "data.frame") |>
  dplyr::group_by(id) |>
  dplyr::summarise(CRCL = CRCL[1], cl_solved = cl[1], vc_solved = vc[1],
                   .groups = "drop") |>
  dplyr::mutate(cl_equation = 11.9 * CRCL / 102.23,
                rel_err = abs(cl_solved / cl_equation - 1),
                thalf = log(2) * vc_solved / cl_solved)

knitr::kable(
  cov_chk |>
    dplyr::select(CRCL, cl_solved, cl_equation, rel_err, thalf) |>
    dplyr::rename("CRCL (mL/min/1.73 m^2)" = CRCL,
                  "CL solved (L/h)" = cl_solved,
                  "CL = 11.9 x CRCL/102.23 (L/h)" = cl_equation,
                  "Relative error" = rel_err,
                  "Typical t1/2 (h)" = thalf),
  digits = c(2, 3, 3, 12, 2),
  caption = "Typical clearance and half-life across creatinine clearance."
)

stopifnot(max(cov_chk$rel_err) < 1e-10)
stopifnot(all(abs(cov_chk$vc_solved - 25.2) < 1e-10))
```

## Closed-form steady state

For a one-compartment model given a zero-order infusion of rate `R` over `T`
every `tau`, the steady-state concentration has an exact closed form. The ODE
solve and the closed form use the same individual parameters, so the
difference is integrator error only.

```{r check-closed-form}
css_1cmt <- function(t, cl, vc, dose, tinf, tau) {
  k <- cl / vc
  r <- dose / tinf
  cur <- ifelse(t <= tinf, 1 - exp(-k * t),
                (1 - exp(-k * tinf)) * exp(-k * (t - tinf)))
  past <- (1 - exp(-k * tinf)) * exp(-k * (t - tinf)) * exp(-k * tau) /
    (1 - exp(-k * tau))
  r / cl * (cur + past)
}

set.seed(2022)
n_cf <- 20
cf_par <- data.frame(id = seq_len(n_cf), CRCL = exp(rnorm(n_cf, log(103.14), 0.5)))
obs_t <- c(seq(0, 3, by = 0.1), seq(3.25, 8, by = 0.25))
ev_cf <- do.call(rbind, lapply(seq_len(n_cf), function(i) {
  rbind(
    data.frame(id = i, time = 0, amt = 1000, rate = 1000 / 3, ii = 8, ss = 1,
               evid = 1, cmt = "central", CRCL = cf_par$CRCL[i]),
    data.frame(id = i, time = obs_t, amt = 0, rate = 0, ii = 0, ss = 0,
               evid = 0, cmt = "central", CRCL = cf_par$CRCL[i])
  )
}))

rxode2::rxSetSeed(2022)
cf <- rxode2::rxSolve(ui, ev_cf, returnType = "data.frame",
                      rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12,
                      maxsteps = 1e6) |>
  dplyr::mutate(Cc_closed = css_1cmt(time, cl, vc, dose = 1000, tinf = 3, tau = 8),
                rel_err = abs(Cc / Cc_closed - 1))

cat(sprintf("max relative error vs closed form: %.3g\n", max(cf$rel_err)))
stopifnot(max(cf$rel_err) < 1e-5)
```

# Virtual cohort

The observed data are not public. Each virtual patient draws creatinine
clearance log-normally around the Table 1 median of 103.14 mL/min/1.73 m^2, kept
inside the observed range 13.5-241.4. The log SD of 0.5 is an assumption that
places the observed maximum near the 95th percentile of the draw. The regimen is
drawn from the three Table 1 marginal distributions (daily dose, dosing interval
and infusion time) independently. The dose per administration is the daily dose
divided by the number of doses per day. Concentrations are simulated at steady
state.

```{r cohort}
# set.seed() fixes the covariate and regimen draw. rxSetSeed() fixes rxode2's
# eta and residual draws within one rxode2 build and thread count only, so the
# assertions below are on medians, never on extremes.
set.seed(611)
n_sub <- 200
draw_crcl <- function(n) {
  x <- exp(rnorm(n, log(103.14), 0.5))
  bad <- x < 13.5 | x > 241.4
  while (any(bad)) {
    x[bad] <- exp(rnorm(sum(bad), log(103.14), 0.5))
    bad <- x < 13.5 | x > 241.4
  }
  x
}
cohort <- data.frame(
  id = seq_len(n_sub),
  CRCL = draw_crcl(n_sub),
  daily = sample(c(500, 1000, 1500, 2000, 3000, 6000), n_sub, replace = TRUE,
                 prob = c(1, 6, 2, 11, 52, 6)),
  tau = sample(c(6, 8, 12, 24), n_sub, replace = TRUE, prob = c(4, 60, 12, 2)),
  tinf = sample(c(0.5, 1, 2, 3), n_sub, replace = TRUE, prob = c(33, 16, 6, 23))
) |>
  dplyr::mutate(dose = daily * tau / 24)

# Observations over one steady-state interval, up to the next pre-dose sample.
events <- do.call(rbind, lapply(seq_len(n_sub), function(i) {
  p <- cohort[i, ]
  tt <- sort(unique(c(seq(0, p$tau, by = 0.25), 1, 3, 6)))
  rbind(
    data.frame(id = p$id, time = 0, amt = p$dose, rate = p$dose / p$tinf,
               ii = p$tau, ss = 1, evid = 1, cmt = "central", CRCL = p$CRCL),
    data.frame(id = p$id, time = tt, amt = 0, rate = 0, ii = 0, ss = 0,
               evid = 0, cmt = "central", CRCL = p$CRCL)
  )
}))

summary(cohort[, c("CRCL", "dose", "tau", "tinf")])
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(611)
sim <- rxode2::rxSolve(ui, events, returnType = "data.frame", maxsteps = 1e6)
stopifnot(!anyNA(sim$Cc))
```

## Visual predictive check

Figure 1 of the abstract plots the observed concentrations against time after
dose, with the observed median and the 5th and 95th percentiles. The figure
below shows the simulated median and the 5th-95th percentile band (with residual
error) over the virtual cohort. The points are the observed medians read off
Figure 1 by the maintainers.

```{r fig-vpc}
# Observed median line of Figure 1, digitised at its vertices.
fig1_median <- data.frame(
  time = c(1.45, 2.24, 3.14, 5.24, 6.54, 7.65),
  conc = c(27.1, 14.9, 12.0, 4.6, 5.5, 1.2)
)

vpc <- sim |>
  dplyr::filter(time <= 16) |>
  dplyr::group_by(time) |>
  dplyr::filter(dplyr::n() >= 20) |>
  dplyr::summarise(p05 = quantile(sim, 0.05), p50 = median(sim),
                   p95 = quantile(sim, 0.95), .groups = "drop")

ggplot(vpc, aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 1) +
  geom_point(data = fig1_median, aes(time, conc), size = 2) +
  labs(x = "Time after dose (h)", y = "Meropenem (mg/L)",
       title = "Simulated median and 90% interval, steady state",
       caption = "Replicates Figure 1 of Romano-Aguilar 2022; points are the digitised observed medians.")
```

The observed median line in Figure 1 joins bin medians over sparse, mixed
regimens. It is therefore compared with the simulated median only at the three
nominal sampling times, 1, 3 and 6 h, within a factor of two.

```{r vpc-check}
sim_med <- sim |>
  dplyr::filter(time %in% c(1, 3, 6)) |>
  dplyr::group_by(time) |>
  dplyr::summarise(sim_median = median(sim), .groups = "drop") |>
  dplyr::mutate(fig1_median = stats::approx(fig1_median$time, fig1_median$conc,
                                            xout = time, rule = 2)$y,
                ratio = sim_median / fig1_median)

knitr::kable(
  sim_med |>
    dplyr::rename("Time after dose (h)" = time,
                  "Simulated median (mg/L)" = sim_median,
                  "Figure 1 observed median (mg/L)" = fig1_median,
                  "Ratio" = ratio),
  digits = 2,
  caption = "Simulated vs digitised observed median concentration."
)

# A mis-transcribed clearance, volume or covariate reference moves the whole
# profile by far more than two-fold at one or more of these times.
stopifnot(all(sim_med$ratio > 0.5 & sim_med$ratio < 2))
```

# Probability of target attainment (Table 3)

Table 3 lists the initial regimens that achieve PTA > 90% for 50% and 100%
fT>MIC at each creatinine clearance. The abstract does not state the MIC used.
Each listed regimen is re-simulated below with 200 virtual patients at the
stated creatinine clearance and steady state. The table reports the share of
patients whose concentration (without residual error) exceeds the MIC for at
least the target fraction of the interval, over a grid of MIC values.

```{r pta}
tab3 <- read.csv(text = "
crcl,target,dose,tau,tinf
20,50,1000,24,0.5
20,50,500,12,0.5
20,50,500,8,0.5
40,50,500,12,2
40,50,1000,12,0.5
40,50,500,8,0.5
60,50,1000,12,3
60,50,1500,12,2
60,50,500,8,0.5
80,50,2000,12,3
80,50,500,8,3
80,50,1000,8,1
100,50,500,8,3
100,50,1000,8,2
120,50,1000,8,3
120,50,1500,8,2
120,50,500,6,2
140,50,1000,8,3
140,50,500,6,3
160,50,1500,8,3
160,50,500,6,3
180,50,500,6,3
180,50,1500,6,2
20,100,500,12,0.5
20,100,500,6,0.5
40,100,1000,8,2
40,100,1500,8,0.5
40,100,500,6,0.5
60,100,1000,6,3
60,100,1500,6,2")
tab3$row <- seq_len(nrow(tab3))
mics <- c(1, 2, 4)
n_pta <- 200
grid_frac <- seq(0, 1, by = 0.02)

ev_pta <- do.call(rbind, lapply(tab3$row, function(r) {
  p <- tab3[r, ]
  ids <- (r - 1) * n_pta + seq_len(n_pta)
  rbind(
    data.frame(id = ids, time = 0, amt = p$dose, rate = p$dose / p$tinf,
               ii = p$tau, ss = 1, evid = 1, cmt = "central", CRCL = p$crcl,
               row = r),
    data.frame(id = rep(ids, each = length(grid_frac)),
               time = rep(grid_frac * p$tau, n_pta), amt = 0, rate = 0, ii = 0,
               ss = 0, evid = 0, cmt = "central", CRCL = p$crcl, row = r)
  )
}))

rxode2::rxSetSeed(2022611)
sim_pta <- rxode2::rxSolve(ui, ev_pta, returnType = "data.frame", keep = "row",
                           maxsteps = 1e6)

pta <- do.call(rbind, lapply(mics, function(m) {
  sim_pta |>
    dplyr::group_by(row, id) |>
    dplyr::summarise(ft = mean(Cc > m), .groups = "drop") |>
    dplyr::left_join(tab3, by = "row") |>
    dplyr::group_by(row) |>
    dplyr::summarise(pta = mean(ft >= target / 100 - 1e-9), .groups = "drop") |>
    dplyr::mutate(mic = m)
})) |>
  dplyr::left_join(tab3, by = "row") |>
  dplyr::mutate(regimen = sprintf("%g mg q%gh over %g h", dose, tau, tinf),
                mic = paste0("PTA at MIC ", mic, " mg/L (%)"),
                pta = 100 * pta) |>
  dplyr::select(crcl, target, regimen, mic, pta) |>
  tidyr::pivot_wider(names_from = mic, values_from = pta)

knitr::kable(
  pta |>
    dplyr::rename("CLCr (mL/min/1.73 m^2)" = crcl,
                  "fT>MIC target (%)" = target,
                  "Table 3 regimen" = regimen),
  digits = 0,
  caption = "Simulated PTA of every Table 3 regimen (Table 3 lists them as PTA > 90%)."
)

pta2 <- pta[["PTA at MIC 2 mg/L (%)"]]
pta4 <- pta[["PTA at MIC 4 mg/L (%)"]]
cat(sprintf("median PTA across Table 3 rows: %.0f%% at MIC 2 mg/L, %.0f%% at MIC 4 mg/L\n",
            median(pta2), median(pta4)))
# Table 3's regimens sit at the 90% boundary for an MIC of about 2 mg/L: the
# median row is near 90% there and well below it at 4 mg/L. A two-fold error
# in clearance or volume moves the median row PTA at 2 mg/L away from 90% by
# far more than the +/- 7 points allowed here.
stopifnot(abs(median(pta2) - 91) < 7, median(pta4) < 85)
```

The listed regimens cluster at the 90% boundary for an MIC of 2 mg/L, which is
consistent with the abstract's simulations having used that MIC (the EUCAST
clinical breakpoint for Enterobacterales and *Pseudomonas aeruginosa*).

For augmented renal clearance and the 100% fT>MIC target the abstract recommends
continuous infusion, because no intermittent regimen reached PTA > 90%. The
typical steady-state concentration of a continuous infusion is
`daily dose / 24 / CL`:

```{r continuous}
ci <- expand.grid(daily = c(3000, 6000), CRCL = c(80, 130, 180, 240)) |>
  dplyr::mutate(cl = 11.9 * CRCL / 102.23, css = daily / 24 / cl)
knitr::kable(
  ci |>
    dplyr::select(daily, CRCL, cl, css) |>
    dplyr::rename("Daily dose (mg)" = daily,
                  "CLCr (mL/min/1.73 m^2)" = CRCL,
                  "Typical CL (L/h)" = cl,
                  "Typical Css (mg/L)" = css),
  digits = c(0, 0, 2, 2),
  caption = "Typical continuous-infusion steady-state concentrations."
)
```

# PKNCA validation

Non-compartmental analysis over one steady-state interval of the two most
common regimens, 1000 mg q8h over 0.5 h and over 3 h, in 200 patients each
drawn from the same creatinine clearance distribution. At steady state the
interval AUC must equal `Dose / CL` for every subject. The abstract reports no
NCA of its own.

```{r pknca}
set.seed(8)
nca_regimens <- data.frame(treatment = c("1000 mg q8h, 0.5 h", "1000 mg q8h, 3 h"),
                           tinf = c(0.5, 3))
nca_cohort <- do.call(rbind, lapply(seq_len(nrow(nca_regimens)), function(j) {
  data.frame(id = (j - 1) * n_sub + seq_len(n_sub),
             treatment = nca_regimens$treatment[j],
             tinf = nca_regimens$tinf[j],
             CRCL = draw_crcl(n_sub))
}))
nca_t <- sort(unique(c(seq(0, 3, by = 0.1), seq(3.25, 8, by = 0.25))))
ev_nca <- do.call(rbind, lapply(seq_len(nrow(nca_cohort)), function(i) {
  p <- nca_cohort[i, ]
  rbind(
    data.frame(id = p$id, time = 0, amt = 1000, rate = 1000 / p$tinf, ii = 8,
               ss = 1, evid = 1, cmt = "central", CRCL = p$CRCL,
               treatment = p$treatment),
    data.frame(id = p$id, time = nca_t, amt = 0, rate = 0, ii = 0, ss = 0,
               evid = 0, cmt = "central", CRCL = p$CRCL, treatment = p$treatment)
  )
}))

# Tight tolerances: at the highest creatinine clearances the trough falls to
# a level where default absolute tolerance can return slightly negative
# concentrations, which PKNCA's log-down trapezoid cannot integrate.
rxode2::rxSetSeed(8)
sim_nca <- rxode2::rxSolve(ui, ev_nca, returnType = "data.frame",
                           keep = "treatment", maxsteps = 1e6,
                           rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12)
stopifnot(min(sim_nca$Cc) >= 0)

conc <- sim_nca |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
dose_df <- ev_nca |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

o_conc <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 0, end = 8, cmax = TRUE, tmax = TRUE, cmin = TRUE,
                        auclast = TRUE, cav = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
nca_res <- as.data.frame(nca$result)

nca_summary <- nca_res |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

knitr::kable(
  nca_summary |>
    dplyr::select(treatment, cmax, tmax, cmin, auclast, cav) |>
    dplyr::rename("Regimen" = treatment, "Cmax (mg/L)" = cmax, "Tmax (h)" = tmax,
                  "Cmin (mg/L)" = cmin, "AUC0-8 (mg*h/L)" = auclast,
                  "Cavg (mg/L)" = cav),
  digits = 2,
  caption = "Median steady-state NCA parameters (simulated, individual predictions)."
)

auc_chk <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::left_join(sim_nca |> dplyr::distinct(id, cl), by = "id") |>
  dplyr::mutate(rel_err = PPORRES * cl / 1000 - 1)

cat(sprintf("AUC0-8 x CL / Dose - 1: median %.2g, 90th pct |err| %.2g\n",
            median(auc_chk$rel_err), quantile(abs(auc_chk$rel_err), 0.9)))
# Trapezoidal error on a 0.1-0.25 h grid is small for typical half-lives; a
# wrong clearance, dose or infusion rate moves this by tens of percent.
stopifnot(abs(median(auc_chk$rel_err)) < 0.01,
          quantile(abs(auc_chk$rel_err), 0.9) < 0.05)
```

# Assumptions and deviations

* **Variance scale of the interindividual variability.** Table 2 prints both
  terms as CV% under an omega-squared symbol and does not say how CV% was
  computed. The log-normal conversion `omega^2 = log(1 + CV^2)` is used, giving
  variances of 0.2745 (CL) and 0.1996 (V). The alternative reading
  `omega^2 = CV^2` (0.316 and 0.221) cannot be ruled out from the abstract:
  the bootstrap columns are percentile intervals, which do not discriminate
  the two, and the Table 3 PTA values change by at most a few percentage
  points between the readings.
* **Residual error.** Table 2 reports one residual term, sigma = 3.53 in
  ug/mL, with no description of the error model. A value in concentration units
  is read as an additive standard deviation.
* **Covariate reference.** The Results print the reference as 102.23 and Table
  2 as 102; 102.23 is used. The abstract does not say which cohort statistic it
  is (the Table 1 median is 103.14).
* **Cockcroft-Gault details.** The body weight entering the Cockcroft-Gault
  formula and the normalisation to 1.73 m^2 are not described. `CRCL` should be
  supplied in mL/min/1.73 m^2 as in Table 1. Clearance is linear in `CRCL`
  through the origin with no cap, so it approaches zero for anuric patients;
  the model was fitted to 13.5-241.4 mL/min/1.73 m^2.
* **Virtual cohort.** Creatinine clearance is drawn log-normally around the
  Table 1 median within the observed range, and the regimen from the three
  Table 1 marginal frequencies independently. The real pairing of doses,
  intervals and infusion times is not reported. Steady state is assumed at
  sampling.
* **Figure 1 digitisation.** The observed median line was read at its vertices
  from the image; precision is about +/- 1 mg/L.
* **MIC in Table 3.** The abstract does not state the MIC used for its PTA
  simulations. The re-simulation reports a grid of MICs; the regimens
  sit at the 90% boundary for 2 mg/L. Whether the abstract's simulations
  included residual error is not stated; the re-simulation excludes it.
* **Protein binding.** The abstract states its targets as %t > MIC on total
  concentration; meropenem binding is low (about 2%), and total concentrations
  are used.
* **Interindividual variability correlation.** No covariance between the CL
  and V random effects is reported; they are independent.
