---
title: "Nadroparin (Romano 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nadroparin (Romano 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("Romano_2023_nadroparin")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1111/bcp.15634>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC9878197/>

Romano 2023 is a population pharmacokinetic analysis of nadroparin, a
low-molecular-weight heparin, given as thromboprophylaxis (5700 IU
subcutaneously twice daily) to critically ill COVID-19 patients in the
intensive care unit. The model is fitted to plasma **anti-Xa activity**, so
clearance and volume are apparent anti-Xa parameters (CL/F in mL/h, Vd/F in
mL) and the observation is in IU/mL. The clinical question is which regimen
best reaches the 0.3-0.7 IU/mL anti-Xa target measured 4 h after a dose; the
paper's answer is 5700 IU twice daily.

The final model is one-compartment with first-order absorption. Apparent
clearance carries all of the inter-individual variability and five covariate
effects -- higher C-reactive protein, D-dimer and CKD-EPI eGFR raise it, and
vasopressor and corticosteroid use lower it by 25.1% and 22.5%:

$$
CL_i = 2230 \left(\frac{CRP}{100}\right)^{0.182}
\left(\frac{\text{D-dimer}}{10}\right)^{0.117}
\left(\frac{GFR_{CKD\text{-}EPI}}{80}\right)^{0.368}
0.749^{\text{Vasopressors}} \, 0.775^{\text{Corticosteroids}} \, e^{\eta_i}
$$

with CRP in mg/L, D-dimer in mg/L and eGFR in mL/min/1.73 m^2. The control
stream in the paper's Appendix SA computes the eGFR inside the model from
serum creatinine, age and sex with the 2009 CKD-EPI equation; the packaged
model does the same, so it takes `CREAT` (umol/L), `AGE` and `SEXF` rather
than a precomputed eGFR column.

# Population

The modelling cohort is 65 critically ill adults with PCR-confirmed SARS-CoV-2
infection admitted to the Erasmus MC intensive care unit (Rotterdam, the
Netherlands) between March 2020 and January 2021, who contributed 280 anti-Xa
samples (21 below the 0.1 IU/mL limit of quantification). A further 25
patients (147 samples) formed an internal validation cohort. Modelling-cohort
medians (Table 1): age 62 years, body weight 89 kg, CRP 76 mg/L, D-dimer
1.29 mg/L, CKD-EPI eGFR 94.4 mL/min/1.73 m^2; 87% were mechanically
ventilated, 89% received vasopressors and 65% corticosteroids at the start of
sampling. Sex and race are not tabulated.

# Source trace

| Model element | Value | Source |
|---|---|---|
| Structure: depot -> central, first-order absorption and elimination | one compartment | Results 3.2; Table 2 (ka, CL, Vd only); Appendix SA ADVAN5, K12 = KA, K20 = CL/V2 |
| `lka` | log(0.276 1/h) | Table 2 final model; Appendix SA THETA(3) |
| `lcl` | log(2230 mL/h) | Table 2 final model and CL equation; Appendix SA THETA(4) |
| `lvc` | log(11 000 mL) | Table 2 final model; Appendix SA THETA(5) |
| `e_crp_cl` | 0.182, (CRP/100)^theta | Table 2 'CL-CRP'; Appendix SA THETA(6), COV_CL |
| `e_ddimer_cl` | 0.117, (D-dimer/10 mg/L)^theta | Table 2 'CL-D-dimer'; Appendix SA THETA(7), COV_DDMR; unit from Table 1 and Figure 5 axis |
| `e_inotrope_cl` | 0.749^Vasopressors | Table 2 'CL-use of vasopressors'; Appendix SA THETA(8), COV_VASO |
| `e_crcl_cl` | 0.368, (eGFR/80)^theta | Table 2 'CL-GFR CKD-EPI'; Appendix SA THETA(9), COV_CL2 |
| `e_steroid_cl` | 0.775^Corticosteroids | Table 2 'CL-use of corticosteroids'; Appendix SA THETA(10), COV_CORT |
| CKD-EPI 2009 eGFR | 141, kappa 61.9 / 79.6 umol/L, alpha -0.329 / -0.411, -1.209, 0.993^AGE, x 1.018 female | Appendix SA $PK |
| `etalcl` | variance 0.115 (34.9% CV) | Appendix SA $OMEGA; Table 2 IIV on CL |
| `addSd` | 0.0859 IU/mL | Appendix SA THETA(1); Table 2 0.086 |
| `propSd` | 0.2 | Appendix SA THETA(2); Table 2 0.2 |
| Error form | sqrt(add^2 + prop^2 IPRED^2) | Appendix SA $ERROR, W |

# Closed-form checks

## The in-model eGFR reproduces the CKD-EPI 2009 equation

The `egfr` variable computed inside the model is compared with an independent
evaluation of the 2009 CKD-EPI equation (creatinine converted from umol/L to
mg/dL) for creatinine values on both sides of the sex-specific knot.

```{r egfr-check}
ckd_epi_2009 <- function(scr_umol, age, female) {
  scr <- scr_umol / 88.4
  kappa <- ifelse(female == 1, 0.7, 0.9)
  alpha <- ifelse(female == 1, -0.329, -0.411)
  141 * pmin(scr / kappa, 1)^alpha * pmax(scr / kappa, 1)^-1.209 *
    0.993^age * ifelse(female == 1, 1.018, 1)
}

egfr_grid <- tidyr::expand_grid(
  CREAT = c(40, 61.9, 70, 79.6, 120, 250),
  SEXF = c(0, 1)
) |>
  mutate(id = row_number(), AGE = 62, CRP = 100, DDIMER = 10000,
         CONMED_INOTROPE = 0, CONMED_STEROID = 0)

egfr_ev <- egfr_grid |>
  mutate(time = 0, amt = 0, evid = 0, cmt = "central") |>
  select(id, time, amt, evid, cmt, everything())

egfr_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), egfr_ev, returnType = "data.frame")
egfr_chk <- egfr_grid |>
  mutate(egfr_model = egfr_sim$egfr,
         egfr_closed = ckd_epi_2009(CREAT, AGE, SEXF))

knitr::kable(egfr_chk |> select(CREAT, SEXF, egfr_model, egfr_closed), digits = 2)
# Same equation evaluated twice: the only difference is the umol/L kappa
# rounding (61.9 and 79.6 vs 0.7 * 88.4 and 0.9 * 88.4).
stopifnot(all(abs(egfr_chk$egfr_model / egfr_chk$egfr_closed - 1) < 0.002))
```

## Typical clearance at the reference patient

Romano 2023 Table 2 states the typical values hold for a patient with CRP
100 mg/L, D-dimer 10 mg/L and eGFR 80 mL/min/1.73 m^2 on neither
corticosteroids nor vasopressors. With the creatinine back-calculated to give
an eGFR of exactly 80, the model's typical clearance must be 2230 mL/h.

```{r reference-cl}
# Inverts the CKD-EPI 2009 equation for creatinine (umol/L).
ckd_epi_2009_creat <- function(egfr, age, female) {
  kappa <- ifelse(female == 1, 61.9, 79.6)
  alpha <- ifelse(female == 1, -0.329, -0.411)
  r <- egfr / (141 * 0.993^age * ifelse(female == 1, 1.018, 1))
  kappa * ifelse(r >= 1, r^(1 / alpha), r^(-1 / 1.209))
}

ref_ev <- data.frame(
  id = 1, time = 0, amt = 0, evid = 0, cmt = "central",
  CRP = 100, DDIMER = 10000, AGE = 62, SEXF = 0,
  CREAT = ckd_epi_2009_creat(80, 62, 0),
  CONMED_INOTROPE = 0, CONMED_STEROID = 0
)
ref_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), ref_ev, returnType = "data.frame")
c(egfr = ref_sim$egfr, cl_mL_h = ref_sim$cl)
stopifnot(abs(ref_sim$egfr - 80) < 1e-6, abs(ref_sim$cl - 2230) < 1e-6)
```

## The D-dimer reference is 10 mg/L

The D-dimer reference of 10 sits at the top of the observed range (Table 1
median 1.24 mg/L), which is unusual enough to check. Figure 5 plots D-dimer
on a 0-10 mg/L axis, and a second, independent check comes from the
structural model: before covariates were added, Romano 2023 estimated a
population clearance of 1140 mL/h (Table 2, structural model). Evaluating the
final model at the modelling-cohort median covariates, with each binary
effect weighted by its prevalence (the geometric mean of the factor across
the cohort), should land near that value if D-dimer is read in mg/L. Reading
the reference as the cohort median instead would move clearance by a factor
of (10 / 1.29)^0.117 = 1.27.

```{r structural-consistency}
theta <- ui$theta
cl_median_patient <- unname(
  exp(theta["lcl"]) *
    (76 / 100)^theta["e_crp_cl"] *
    (1.29 / 10)^theta["e_ddimer_cl"] *
    (94.4 / 80)^theta["e_crcl_cl"] *
    theta["e_inotrope_cl"]^0.89 *
    theta["e_steroid_cl"]^0.65
)
c(final_model_at_median = cl_median_patient, structural_model = 1140)
stopifnot(abs(cl_median_patient / 1140 - 1) < 0.05)
```

## The simulated steady state matches the closed form

```{r closed-form-ss}
ss_closed <- function(dose, tau, t, cl, v, ka) {
  k <- cl / v
  dose * ka / (v * (ka - k)) *
    (exp(-k * t) / (1 - exp(-k * tau)) - exp(-ka * t) / (1 - exp(-ka * tau)))
}

cf_ev <- data.frame(
  id = 1, time = c(0, 240 + 4), amt = c(5700, 0), evid = c(1, 0),
  cmt = c("depot", "central"), ii = c(12, 0), addl = c(20, 0),
  CRP = 100, DDIMER = 10000, AGE = 62, SEXF = 0,
  CREAT = ckd_epi_2009_creat(80, 62, 0),
  CONMED_INOTROPE = 0, CONMED_STEROID = 0
)
cf_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), cf_ev, returnType = "data.frame")
cf_ref <- ss_closed(5700, 12, 4, 2230, 11000, 0.276)
c(simulated = cf_sim$Cc[cf_sim$time == 244], closed_form = cf_ref)
# Same parameters on both sides: the difference is solver tolerance and the
# residual distance from steady state after 20 doses (k * 240 h = 48.7).
stopifnot(abs(cf_sim$Cc[cf_sim$time == 244] / cf_ref - 1) < 1e-4)
```

# Virtual cohort

The cohort reproduces the modelling-cohort covariate distributions in Table 1.
CRP, D-dimer and eGFR are drawn log-normally from their medians and
interquartile ranges, age normally, and the two medication indicators as
Bernoulli draws at the Table 1 prevalences. The paper does not tabulate sex,
so 70% of patients are assumed male; serum creatinine is then back-calculated
from the drawn eGFR, age and sex, so the eGFR the model computes reproduces
the drawn value exactly. Covariates are held constant within a patient.

```{r cohort}
rxode2::rxSetSeed(20230501)
n_per_arm <- 200

lnorm_from_iqr <- function(n, median, q1, q3) {
  exp(rnorm(n, log(median), (log(q3) - log(q1)) / (2 * qnorm(0.75))))
}

cohort <- tibble(
  pid = seq_len(n_per_arm),
  AGE = pmin(pmax(rnorm(n_per_arm, 62, (70 - 53) / (2 * qnorm(0.75))), 25), 85),
  SEXF = rbinom(n_per_arm, 1, 0.30),
  CRP = lnorm_from_iqr(n_per_arm, 76.0, 35.5, 124),
  DDIMER = 1000 * lnorm_from_iqr(n_per_arm, 1.29, 0.87, 2.19),
  egfr_target = pmin(lnorm_from_iqr(n_per_arm, 94.4, 78.6, 111), 160),
  CONMED_INOTROPE = rbinom(n_per_arm, 1, 0.89),
  CONMED_STEROID = rbinom(n_per_arm, 1, 0.65)
) |>
  mutate(CREAT = ckd_epi_2009_creat(egfr_target, AGE, SEXF))

cohort |>
  summarise(across(c(AGE, CRP, DDIMER, egfr_target, CREAT), median),
            vasopressor = mean(CONMED_INOTROPE), corticosteroid = mean(CONMED_STEROID),
            female = mean(SEXF)) |>
  knitr::kable(digits = 2)
```

# Simulation of the dosing regimens in Figure 4

Each patient receives each of the four regimens Romano 2023 compares
(3800 IU, 5700 IU and 7600 IU twice daily, 5700 IU once daily) for ten days,
long enough to be at steady state (the elimination half-life is about 7 h
for a patient at the cohort-median covariates and 3.4 h at the reference
patient). The last dosing interval is sampled densely.

```{r events}
regimens <- tibble(
  regimen = c("3800 IU BID", "5700 IU BID", "7600 IU BID", "5700 IU QD"),
  dose = c(3800, 5700, 7600, 5700),
  tau = c(12, 12, 12, 24)
) |>
  mutate(n_dose = 240 / tau, t_last = 240 - tau)

arms <- tidyr::crossing(regimens, cohort) |>
  mutate(id = row_number())

dose_rows <- arms |>
  transmute(id, time = 0, amt = dose, evid = 1, cmt = "depot",
            ii = tau, addl = n_dose - 1)
obs_rows <- arms |>
  select(id, t_last, tau) |>
  tidyr::crossing(tad = seq(0, 24, by = 0.25)) |>
  filter(tad <= tau) |>
  transmute(id, time = t_last + tad, amt = 0, evid = 0, cmt = "central",
            ii = 0, addl = 0)

events <- bind_rows(dose_rows, obs_rows) |>
  left_join(arms |> select(id, regimen, AGE, SEXF, CRP, DDIMER, CREAT,
                           CONMED_INOTROPE, CONMED_STEROID), by = "id") |>
  arrange(id, time, desc(evid)) |>
  select(id, time, amt, evid, cmt, ii, addl, everything())
```

```{r simulate}
sim <- rxode2::rxSolve(mod, events |> select(-regimen), returnType = "data.frame") |>
  left_join(arms |> select(id, regimen, t_last), by = "id") |>
  mutate(tad = time - t_last)
```

## Steady-state anti-Xa 4 h and 12 h after a dose (Figure 4)

Figure 4 of Romano 2023 shows, for the modelling cohort, the steady-state
anti-Xa activity 4 h after a dose computed from each patient's empirical
Bayes clearance. The digitised medians are compared with the simulated
individual predictions (`Cc`, which carries the clearance random effect but
not residual error) at the same times.

```{r fig4}
at_times <- sim |>
  filter(tad %in% c(4, 12)) |>
  mutate(regimen = factor(regimen, levels = regimens$regimen))

ggplot(at_times |> filter(tad == 4), aes(regimen, Cc)) +
  geom_boxplot(outlier.size = 0.8) +
  geom_hline(yintercept = c(0.3, 0.7), linetype = "dashed", colour = "darkgreen") +
  labs(x = NULL, y = "Steady-state anti-Xa at t = 4 h (IU/mL)",
       caption = "Replicates Figure 4 (modelling cohort, t = 4 h) of Romano 2023.")
```

```{r fig4-compare}
fig4 <- tribble(
  ~regimen, ~tad, ~published_median, ~published_pct_in_range,
  "3800 IU BID", 4, 0.30, 46.2,
  "5700 IU BID", 4, 0.45, 55.4,
  "7600 IU BID", 4, 0.60, 52.3,
  "5700 IU QD", 4, 0.31, 52.3,
  "5700 IU BID", 12, 0.27, NA,
  "7600 IU BID", 12, 0.36, NA
)

fig4_cmp <- at_times |>
  group_by(regimen, tad) |>
  summarise(simulated_median = median(Cc),
            simulated_pct_in_range = 100 * mean(Cc >= 0.3 & Cc <= 0.7),
            .groups = "drop") |>
  mutate(regimen = as.character(regimen)) |>
  inner_join(fig4, by = c("regimen", "tad")) |>
  mutate(pct_diff_median = 100 * (simulated_median / published_median - 1))

fig4_cmp |>
  select(regimen, tad, published_median, simulated_median, pct_diff_median,
         published_pct_in_range, simulated_pct_in_range) |>
  dplyr::rename(
    "Regimen" = regimen, "Time after dose (h)" = tad,
    "Figure 4 median (IU/mL)" = published_median,
    "Simulated median (IU/mL)" = simulated_median,
    "Median difference (%)" = pct_diff_median,
    "Figure 4 % in 0.3-0.7" = published_pct_in_range,
    "Simulated % in 0.3-0.7" = simulated_pct_in_range
  ) |>
  knitr::kable(digits = c(0, 0, 2, 2, 1, 1, 1))

# Robust centre check: a mis-transcribed clearance, volume or covariate
# reference moves every regimen's median by tens of percent.
stopifnot(
  abs(median(fig4_cmp$pct_diff_median)) < 10,
  all(abs(fig4_cmp$pct_diff_median) < 25)
)
```

The simulated medians agree with the digitised Figure 4 medians. The
fraction of patients inside the target range is reported but not asserted on:
Figure 4 is built from empirical Bayes estimates, which shrink toward the
typical value (14% shrinkage on clearance), and the observed cohort's
covariate joint distribution is not reproduced exactly by the independent
draws above, so the tails -- and hence the in-range fraction -- differ more
than the centre does.

## Profile over a steady-state dosing interval (cf. Figure 2)

Figure 2 is a prediction-corrected visual predictive check against time after
dose for the observed data, which mix first and later doses. The plot below
shows the simulated 5700 IU twice-daily steady-state interval with residual
error (`sim`), the observable that the visual predictive check compares with
the data. The observed median in Figure 2 peaks near 0.43 IU/mL around 4 h
after the dose.

```{r vpc}
sim |>
  filter(regimen == "5700 IU BID") |>
  group_by(tad) |>
  summarise(p05 = quantile(sim, 0.05), p50 = quantile(sim, 0.5),
            p95 = quantile(sim, 0.95), .groups = "drop") |>
  ggplot(aes(tad)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = p50), colour = "steelblue4") +
  geom_hline(yintercept = c(0.3, 0.7), linetype = "dashed", colour = "darkgreen") +
  labs(x = "Time after dose (h)", y = "Anti-Xa (IU/mL)",
       caption = "Median and 90% interval, 5700 IU BID at steady state; compare Figure 2 of Romano 2023.")
```

# PKNCA validation

Non-compartmental analysis of the steady-state dosing interval, grouped by
regimen. Romano 2023 reports no NCA, so the check is internal: at steady state
the area over one dosing interval equals dose / CL for each patient.

```{r pknca}
nca_conc <- sim |>
  filter(!is.na(Cc)) |>
  select(id, regimen, tad, Cc)

nca_dose <- arms |>
  transmute(id, regimen, tad = 0, dose)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ tad | regimen + id)
dose_obj <- PKNCA::PKNCAdose(nca_dose, dose ~ tad | regimen + id)

intervals <- regimens |>
  transmute(regimen, start = 0, end = tau, cmax = TRUE, tmax = TRUE,
            auclast = TRUE, cmin = TRUE, cav = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_summary <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cmin", "cav")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::rename(
    "Regimen" = regimen, "Cmax (IU/mL)" = cmax, "Tmax (h)" = tmax,
    "AUCtau (IU*h/mL)" = auclast, "Cmin (IU/mL)" = cmin, "Cav (IU/mL)" = cav
  ) |>
  knitr::kable(digits = 3)
```

```{r auc-vs-dose-over-cl}
auc_chk <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  select(id, regimen, auclast = PPORRES) |>
  left_join(sim |> distinct(id, cl), by = "id") |>
  left_join(arms |> select(id, dose), by = "id") |>
  mutate(pct_diff = 100 * (auclast / (dose / cl) - 1))
summary(auc_chk$pct_diff)
# Same drawn clearance on both sides; the residual is trapezoidal error on a
# 15-min grid plus, for the slowest-clearing patients, a small shortfall from
# full steady state after ten days.
stopifnot(
  abs(median(auc_chk$pct_diff)) < 0.5,
  quantile(abs(auc_chk$pct_diff), 0.9) < 2
)
```

# Assumptions and deviations

* **Sex distribution.** Table 1 does not report sex. The virtual cohort
  assumes 70% male; sex enters the model only through the CKD-EPI eGFR, and
  because creatinine is back-calculated from the drawn eGFR the assumption
  does not change any simulated clearance.
* **Creatinine.** Table 1 reports eGFR but not serum creatinine. The cohort
  draws eGFR from its Table 1 median and interquartile range and inverts the
  CKD-EPI equation for creatinine.
* **Covariate distributions.** CRP, D-dimer and eGFR are drawn independently
  (log-normal from Table 1 medians and IQRs); the paper reports no
  collinearity among the retained covariates (Figure S2). Covariates are held
  constant within a patient, whereas the source data were time-varying.
* **Missing-covariate imputation.** The Appendix SA control stream replaces a
  CRP of 0 (missing) by 100 mg/L and a missing eGFR by 80 mL/min/1.73 m^2. Its
  eGFR branch tests a variable `GFR` that is neither a data column nor
  defined in the code shown; because the eGFR exponent was estimated with a
  33% RSE, the published run must have computed eGFR from creatinine, which
  is what the packaged model does. Neither imputation rule is part of the
  packaged model; supply observed values.
* **Baseline term.** The control stream writes `IPRED = F + BL` with
  `BL = TVBL`, but `TVBL` is never defined and Results 3.2 states that a
  baseline parameter was not evaluated. The packaged model has no baseline.
* **Unused peripheral compartment.** Appendix SA declares a `PERIPH`
  compartment under ADVAN5 but defines no rate constants for it, and Table 2
  reports no intercompartmental clearance or peripheral volume; the model is
  one-compartment.
* **Below-quantification data.** The paper fitted samples below 0.1 IU/mL with
  the M3 likelihood. That is an estimation device; simulations from the
  packaged model return uncensored values.
* **D-dimer units.** The canonical `DDIMER` column is in ng/mL; the source
  recorded mg/L, so the model divides by 10000 ng/mL (the paper's 10 mg/L
  reference). The structural-model check above supports the mg/L reading.
* **Figures 5 and 6 (Monte Carlo covariate scenarios)** are not replicated:
  the paper draws covariates uniformly between unreported minima and maxima
  and does not state how the binary indicators were sampled.
* No erratum was found for Romano 2023 (EuropePMC search, 2026-10-11).
