---
title: "Inhaled fluticasone propionate + salmeterol relative bioavailability (Rosenborg 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inhaled fluticasone propionate + salmeterol relative bioavailability (Rosenborg 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Rosenborg 2025 asks whether an empirical population PK analysis can replace a
non-compartmental analysis (NCA) for assessing bioequivalence of an inhaled
fixed-dose combination. Three two-way crossover studies compared Wixela Inhub
(test) against Advair Diskus (reference), and the authors fitted **four**
separate NONMEM models: one per fluticasone propionate (FP) dose, and one for
salmeterol (SALM) pooled across the three studies. All four are packaged here.

```{r models}
model_names <- c(
  "Rosenborg_2025_fluticasone_300ug",
  "Rosenborg_2025_fluticasone_750ug",
  "Rosenborg_2025_fluticasone_1500ug",
  "Rosenborg_2025_salmeterol"
)
uis <- lapply(model_names, function(nm) rxode2::rxode(readModelDb(nm)))
names(uis) <- model_names

tibble::tibble(
  Model = c("Model 1", "Model 2", "Model 3", "Model 4"),
  `nlmixr2lib name` = model_names,
  Analyte = c(rep("Fluticasone propionate", 3), "Salmeterol"),
  `Total dose` = c("3 x 100 = 300 ug", "3 x 250 = 750 ug",
                   "3 x 500 = 1500 ug", "3 x 50 = 150 ug"),
  Study = c("1", "2", "3", "1, 2 and 3 pooled")
) |>
  knitr::kable(caption = "The four models of Rosenborg 2025 Table 2.")
```

* Citation: `r uis[[1]]$reference`
* Article: <https://doi.org/10.2147/DDDT.S480189>
* Supplement (NONMEM code, goodness-of-fit, sensitivity analyses):
  <https://www.dovepress.com/article/supplementary_file/480189/DDDT-480189-supplementary-material.pdf>

# Population

Three separate two-way crossover studies were run in healthy adults in the USA,
each recruiting 66 subjects (198 in total) of whom 61 to 65 per study completed
both treatment periods. Panels were deliberately homogeneous -- black and white
subjects, mean BMI approximately 26 -- with 29 female / 37 male and mean age
33.8 years in study 1, 36 / 30 and 37.7 years in study 2, and 42 / 24 and
35.7 years in study 3.

Each subject took three inhalations on each of two occasions separated by a
7-day washout: FP 100, 250 or 500 ug per inhalation in studies 1, 2 and 3
respectively, always with SALM 50 ug per inhalation. Plasma was sampled
pre-dose and at 2, 5, 10, 15, 20, 30 and 45 min and at 1, 1.5, 2, 3, 4, 6, 8,
12, 24, 36 and 48 h post-dose, with an LLOQ of 1 ng/L. The analysis dataset held
7184 FP and 7186 SALM concentrations.

Because the design is a crossover with homogeneous panels, Rosenborg 2025 states
that "neither demographic nor other covariates were considered in the evaluation
of NCA-based results and therefore not in this alternative model analysis
either". Product (test vs reference) is consequently the only fixed-effect
covariate in any of the four models.

The same information is available programmatically via each model's `population`
metadata, e.g. `uis[["Rosenborg_2025_salmeterol"]]$population`.

# Model structure

All four models share one structure (Rosenborg 2025 Figure 1; the NONMEM code is
reproduced in supplement Figure 1). NONMEM `ADVAN5` with
`$MODEL COMP=(DEFOBS1) COMP=(PERIPH1) COMP=(PERIPH2) COMP=(DEPOT1)` numbers the
compartments 1 = central, 2 = first peripheral, 3 = second peripheral, 4 = the
hypothetical pulmonary deposition site. The nominal inhaled dose is assumed to
be deposited instantaneously and completely at site 4 and to reach the systemic
circulation by a single first-order process `k41`; no mechanistic assumption is
made about regional lung deposition, so every clearance and volume is *apparent*
and carries an unknown absolute bioavailability below 1.

Product enters in exactly two places:

```
K41 = EXP(TREA1 * MU_7 + TREA2 * MU_10 + ETA(7))    ; one typical value per product, one shared eta
F4  = EXP(TREA1 * (MU_11 + ETA(8)) + TREA2 * 0)     ; F = 1 exactly for the reference product
```

so the reference product is the bioavailability anchor and `F4_rel` is a genuine
per-subject test/reference ratio -- which is what makes the model-based
bioequivalence test in Rosenborg 2025 Table 4 comparable with the NCA-based one.
In the salmeterol model the `F4` line carries a second, study-level random
effect, `F4 = EXP(TREA1 * (MU_11 + ETA(8) + ETA(9)) + TREA2 * 0)`, declared by
`$LEVEL STUD=(9[1])`; it is encoded here with rxode2's native nesting syntax
(`etalfdepot_study ~ 0.00085849 | SIDN`).

Between-subject variability differs between the analytes: the FP models drop the
etas on `Q2/F` and `Q3/F` "owing to shrinkage", while the SALM model retains
all eight subject-level etas. This is visible in Rosenborg 2025 Table 2, where
`Q2/F_OMEGA_IIV` and `Q3/F_OMEGA_IIV` are `NA` in the three FP columns only.

# Source trace

Every `ini()` entry carries an in-file comment pointing at its Rosenborg 2025
Table 2 row. The mapping between the paper's symbols and the packaged parameter
names is uniform across all four models:

| `ini()` parameter | Paper symbol (Table 2 row) | Source |
|---|---|---|
| `lcl` | `CL/F` | Table 2 |
| `lq` | `Q2/F` | Table 2 |
| `lq2` | `Q3/F` | Table 2 |
| `lvc` | `V1/F` | Table 2 |
| `lvp` | `V2/F` | Table 2 |
| `lvp2` | `V3/F` | Table 2 |
| `lka_test` | `k41(test)` | Table 2 |
| `lka_ref` | `k41(ref)` | Table 2 |
| `lfdepot` | `F4_rel` | Table 2 |
| `propSd` | `CP_PROP` | Table 2 |
| `addSd` | `CP_ADD` | Table 2 |
| `etalcl`, `etalvc`, `etalvp`, `etalvp2`, `etalka`, `etalfdepot` (and `etalq`, `etalq2` for SALM) | `*_OMEGA_IIV` | Table 2 |
| `etalfdepot_study` (SALM only) | `F4_rel_OMEGA_ISV` | Table 2 |
| Compartment topology, `K10`/`K12`/`K21`/`K13`/`K31`, `$ERROR` block | n/a | Figure 1 and supplement Figure 1 (NONMEM code) |

Back-transforming the packaged `ini()` values must reproduce Table 2 exactly.
The table below is generated from the model files, so any transcription error
would show up as a mismatch against the published column:

```{r trace-values}
trace_row <- function(ui) {
  d <- ui$iniDf
  th <- function(nm) {
    v <- d$est[!is.na(d$ntheta) & d$name == nm]
    if (length(v) != 1L) return(NA_real_)
    v
  }
  om <- function(nm) {
    v <- d$est[is.na(d$ntheta) & d$name == nm]
    if (length(v) != 1L) return(NA_real_)
    sqrt(v)
  }
  c(`CL/F` = exp(th("lcl")), `Q2/F` = exp(th("lq")), `Q3/F` = exp(th("lq2")),
    `V1/F` = exp(th("lvc")), `V2/F` = exp(th("lvp")), `V3/F` = exp(th("lvp2")),
    `k41(test)` = exp(th("lka_test")), `k41(ref)` = exp(th("lka_ref")),
    F4_rel = exp(th("lfdepot")),
    CP_PROP = th("propSd"), CP_ADD = th("addSd"),
    `CL/F_OMEGA_IIV` = om("etalcl"), `Q2/F_OMEGA_IIV` = om("etalq"),
    `Q3/F_OMEGA_IIV` = om("etalq2"), `V1/F_OMEGA_IIV` = om("etalvc"),
    `V2/F_OMEGA_IIV` = om("etalvp"), `V3/F_OMEGA_IIV` = om("etalvp2"),
    k41_OMEGA_IIV = om("etalka"), F4_rel_OMEGA_IIV = om("etalfdepot"),
    F4_rel_OMEGA_ISV = om("etalfdepot_study"))
}
trace_tab <- vapply(uis, trace_row, numeric(20))
colnames(trace_tab) <- c("FP Model 1", "FP Model 2", "FP Model 3", "SALM Model 4")

as.data.frame(trace_tab) |>
  tibble::rownames_to_column("Table 2 row") |>
  knitr::kable(
    digits = 4,
    caption = paste(
      "Packaged ini() values back-transformed to the paper's scale.",
      "Compare cell by cell against Rosenborg 2025 Table 2;",
      "blank cells are the NA entries of that table."
    )
  )
```

# Structural check: terminal half-life

Rosenborg 2025 Table 3 reports post-hoc individual terminal half-lives obtained
by solving the three-compartment eigenvalue problem from each subject's
micro-constants (the R code is in supplement Figure 1). The same calculation
applied to the *typical* parameter values is a parameter-free check that the
packaged disposition parameters and the `K10`/`K12`/`K21`/`K13`/`K31` mapping are
both correct -- it uses no simulation at all.

```{r halflife-analytic}
terminal_thalf <- function(ui) {
  p <- stats::setNames(ui$iniDf$est, ui$iniDf$name)
  cl <- exp(p[["lcl"]]); q <- exp(p[["lq"]]); q2 <- exp(p[["lq2"]])
  vc <- exp(p[["lvc"]]); vp <- exp(p[["lvp"]]); vp2 <- exp(p[["lvp2"]])
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp; k13 <- q2 / vc; k31 <- q2 / vp2
  # lambda^3 + a2 lambda^2 + a1 lambda + a0 = 0
  a2 <- k10 + k12 + k21 + k13 + k31
  a1 <- k10 * k21 + k10 * k31 + k21 * k31 + k12 * k31 + k13 * k21
  a0 <- k10 * k21 * k31
  lambdas <- sort(abs(Re(polyroot(c(a0, a1, a2, 1)))))
  log(2) / lambdas[1]
}

published_thalf <- c(
  # Rosenborg 2025 Table 3, mean of individual post-hoc estimates (test / reference)
  "Rosenborg_2025_fluticasone_300ug"  = mean(c(13.75, 13.69)),
  "Rosenborg_2025_fluticasone_750ug"  = mean(c(14.64, 14.53)),
  "Rosenborg_2025_fluticasone_1500ug" = mean(c(14.03, 14.04)),
  "Rosenborg_2025_salmeterol"         = mean(c(12.60, 12.63, 12.42, 12.36, 12.02, 12.00))
)

thalf_tab <- tibble::tibble(
  Model = names(uis),
  `Typical-value t1/2 (h)` = vapply(uis, terminal_thalf, numeric(1)),
  `Published mean post-hoc t1/2 (h)` = as.numeric(published_thalf[names(uis)])
) |>
  mutate(`% diff` = 100 * (`Typical-value t1/2 (h)` /
                             `Published mean post-hoc t1/2 (h)` - 1))
knitr::kable(thalf_tab, digits = 2,
             caption = "Terminal half-life from the typical micro-constants vs Rosenborg 2025 Table 3.")
stopifnot(all(abs(thalf_tab$`% diff`) < 6))
```

All four agree with the published means to within about 5%, using no simulation
and no fitted quantity -- only the packaged `CL/F`, `Q2/F`, `Q3/F`, `V1/F`,
`V2/F` and `V3/F` and the compartment topology. A discrepancy here would mean a
mis-transcribed disposition parameter or a mis-mapped micro-constant, so this is
the sharpest single check on the structural half of each model.

# Virtual crossover cohort

The published bioequivalence test is a within-subject comparison, so the virtual
cohort must be a crossover too: each simulated subject receives the reference
product in one period and the test product in the other, sharing one draw of the
subject-level etas across both. Periods are separated by 168 h (the study's
7-day washout, more than ten terminal half-lives), and the sampling grid is the
study's own schedule -- which matters, because Rosenborg 2025 identified Cmax
"within the framework of actual discrete individual sampling times".

Cohort sizes match the studies: 66 subjects per FP model and 198 (66 per study)
for the pooled salmeterol model, i.e. at most 198 per arm.

```{r cohort}
set.seed(20250819)

sample_times <- c(0, c(2, 5, 10, 15, 20, 30, 45) / 60,
                  1, 1.5, 2, 3, 4, 6, 8, 12, 24, 36, 48)
washout <- 168

make_crossover <- function(n, dose, sidn = NULL, id_offset = 0L) {
  subj <- tibble::tibble(
    id = id_offset + seq_len(n),
    # Balanced sequence: half take the reference first, half the test first.
    ref_first = rep(c(TRUE, FALSE), length.out = n)
  )
  if (!is.null(sidn)) subj$SIDN <- sidn
  periods <- tidyr::expand_grid(subj, period = 1:2) |>
    mutate(
      product = ifelse((period == 1L) == ref_first, "Reference", "Test"),
      FORM_WIXELA_INHUB = as.numeric(product == "Test"),
      dose_time = (period - 1) * washout
    )
  doses <- periods |>
    mutate(time = dose_time, amt = dose, evid = 1L, cmt = "depot")
  obs <- periods |>
    tidyr::expand_grid(tad = sample_times) |>
    mutate(time = dose_time + tad, amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses |> mutate(tad = NA_real_), obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, amt, evid, cmt, product, FORM_WIXELA_INHUB, period, tad,
           any_of("SIDN"))
}

events <- list(
  Rosenborg_2025_fluticasone_300ug  = make_crossover(66, 300),
  Rosenborg_2025_fluticasone_750ug  = make_crossover(66, 750),
  Rosenborg_2025_fluticasone_1500ug = make_crossover(66, 1500),
  Rosenborg_2025_salmeterol         = bind_rows(
    make_crossover(66, 150, sidn = 1L, id_offset =   0L),
    make_crossover(66, 150, sidn = 2L, id_offset =  66L),
    make_crossover(66, 150, sidn = 3L, id_offset = 132L)
  )
)
# Disjoint ids across the three salmeterol studies.
stopifnot(!anyDuplicated(unique(
  events$Rosenborg_2025_salmeterol[, c("id", "time", "evid")]
)))
# The nested random effect needs at least two distinct SIDN values.
stopifnot(length(unique(events$Rosenborg_2025_salmeterol$SIDN)) >= 2L)
```

# Simulation

`rxSolve()` needs the omega passed explicitly for the salmeterol model, because
a nested omega comes back as a *list* of matrices keyed by level rather than as
a single matrix. `useLinCmt = FALSE` keeps rxode2 from replacing the explicit
ODE system with its analytic shortcut.

```{r simulate}
solve_pop <- function(nm) {
  rxode2::rxSolve(
    uis[[nm]], events = events[[nm]], omega = uis[[nm]]$omega,
    useLinCmt = FALSE,
    keep = c("product", "period", "tad")
  ) |>
    as.data.frame() |>
    mutate(model = nm)
}
solve_typical <- function(nm) {
  rxode2::rxSolve(
    rxode2::zeroRe(uis[[nm]]), events = events[[nm]],
    useLinCmt = FALSE, keep = c("product", "period", "tad")
  ) |>
    as.data.frame() |>
    mutate(model = nm)
}

sim     <- bind_rows(lapply(model_names, solve_pop))
typical <- bind_rows(lapply(model_names, solve_typical))

analyte_label <- c(
  Rosenborg_2025_fluticasone_300ug  = "FP 3 x 100 ug",
  Rosenborg_2025_fluticasone_750ug  = "FP 3 x 250 ug",
  Rosenborg_2025_fluticasone_1500ug = "FP 3 x 500 ug",
  Rosenborg_2025_salmeterol         = "SALM 3 x 50 ug"
)
sim$panel     <- factor(analyte_label[sim$model],     levels = analyte_label)
typical$panel <- factor(analyte_label[typical$model], levels = analyte_label)

stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)))
```

# Replicate published figures

## Figure 4 -- typical-subject predictions by product

Rosenborg 2025 Figure 4 overlays the individual measured profiles with the
typical-subject prediction for each product, on a logarithmic ordinate. The
typical-value predictions below are the thick lines of that figure; the shaded
band is the simulated 2.5th-97.5th percentile envelope, which corresponds to the
visual predictive checks of Figures 6 and 7.

```{r figure-4, warning = FALSE}
band <- sim |>
  filter(!is.na(tad)) |>
  group_by(panel, product, tad) |>
  summarise(lo = quantile(Cc, 0.025), hi = quantile(Cc, 0.975), .groups = "drop")

typ <- typical |>
  filter(!is.na(tad)) |>
  group_by(panel, product, tad) |>
  summarise(Cc = first(Cc), .groups = "drop")

ggplot(typ, aes(tad, Cc, colour = product, fill = product)) +
  geom_ribbon(data = band, aes(y = NULL, ymin = pmax(lo, 0.5), ymax = hi),
              alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red") +
  facet_wrap(~panel, scales = "free_y") +
  scale_y_log10() +
  scale_x_continuous(breaks = c(0, 6, 12, 24, 36, 48)) +
  labs(x = "Time after dose (h)", y = "Plasma concentration (ng/L)",
       colour = "Product", fill = "Product",
       caption = paste("Replicates Figure 4 of Rosenborg 2025 (typical-subject",
                       "lines) and the VPC envelopes of Figures 6 and 7.",
                       "Dashed red line: LLOQ = 1 ng/L.")) +
  theme_bw()
```

The salmeterol panel shows the substantially faster absorption noted in the
Results (`k41` around 15-20 /h versus 0.23-0.59 /h for FP), and the fluticasone
panels show the less-than-dose-proportional peak that the Discussion attributes
to slower dissolution at higher strengths.

# PKNCA validation

NCA is run per subject and per period on time-after-dose, matching the paper's
own single-dose treatment of every administration ("All administrations were
treated as single administrations with no carryover effect").

```{r pknca, warning = FALSE}
sim_nca <- sim |>
  filter(!is.na(Cc), !is.na(tad)) |>
  transmute(model, panel, product,
            subject = paste(model, id, product, sep = "|"),
            time = tad, Cc)

# Guarantee a time = 0 row per subject-period (pre-dose Cc = 0 extravascularly).
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(model, panel, product, subject) |> mutate(time = 0, Cc = 0)
) |>
  distinct(model, panel, product, subject, time, .keep_all = TRUE) |>
  arrange(subject, time)

stopifnot(all(sim_nca$Cc >= 0))

dose_nca <- events |>
  bind_rows(.id = "model") |>
  filter(evid == 1L) |>
  transmute(model,
            panel = as.character(analyte_label[model]),
            product,
            subject = paste(model, id, product, sep = "|"),
            time = 0, amt)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | panel + product + subject)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_nca), amt ~ time | panel + product + subject)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_ind <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life"))
stopifnot(nrow(nca_ind) > 0)
```

## Comparison against the published post-hoc Cmax

Rosenborg 2025 Table 3 reports the **arithmetic mean** of individually predicted
Cmax, "identified based on individually predicted peak concentrations within the
framework of actual discrete individual sampling times" -- which is precisely
what a `cmax` computed on the study's own sampling grid measures. The simulated
per-subject results are therefore aggregated with `mean()` before comparison,
rather than with `ncaComparisonTable()`'s default median. For salmeterol, the
packaged model is a single pooled fit with no study effect on disposition, so
the three published study rows are averaged into one reference value.

```{r nca-compare, warning = FALSE}
sim_summary <- nca_ind |>
  filter(PPTESTCD == "cmax") |>
  group_by(panel, product, PPTESTCD) |>
  summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  mutate(group = paste(panel, product, sep = " | ")) |>
  select(group, PPTESTCD, PPORRES)

published <- tibble::tribble(
  ~group,                            ~cmax,
  # Rosenborg 2025 Table 3 (mean of individual post-hoc estimates)
  "FP 3 x 100 ug | Reference",       112.0,
  "FP 3 x 100 ug | Test",             97.45,
  "FP 3 x 250 ug | Reference",       153.1,
  "FP 3 x 250 ug | Test",            149.1,
  "FP 3 x 500 ug | Reference",       278.3,
  "FP 3 x 500 ug | Test",            253.5,
  # Salmeterol: mean over the three published study rows of Table 3
  "SALM 3 x 50 ug | Reference", mean(c(367.5, 327.8, 386.0)),
  "SALM 3 x 50 ug | Test",      mean(c(363.0, 289.4, 352.0))
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = as.data.frame(sim_summary),
  reference     = as.data.frame(published),
  by            = "group",
  units         = c(cmax = "ng/L"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste("Simulated vs Rosenborg 2025 Table 3 mean post-hoc Cmax.",
                  "* differs from the published value by more than 20%.")
)
```

```{r nca-compare-assert}
pct <- suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", cmp[["% diff"]])))
stopifnot(length(pct) == 8L, all(abs(pct) <= 10, na.rm = TRUE))
```

Peak concentrations reproduce the published means to within about 5% for every
dose and product, including the less-than-dose-proportional rise from 300 to
1500 ug of FP.

## Half-life: a fresh IIV draw is not a set of post-hoc estimates

Table 3's half-lives need a different comparison, and the reason is instructive.
Rosenborg 2025 derived them from each subject's *post-hoc* micro-constants --
empirical Bayes estimates, which the data shrink toward the typical value. A
simulation instead draws each eta afresh from the full published omega, with no
shrinkage and (because the paper reports no correlations) no covariance between
the six disposition parameters. Terminal half-life is a strongly non-linear
function of those six, so independent extreme draws -- a large `V3/F` paired
with a small `Q3/F`, say -- generate long tails that no real post-hoc set
contains.

```{r halflife-distribution, warning = FALSE}
exact_thalf <- function(k10, k12, k21, k13, k31) {
  a2 <- k10 + k12 + k21 + k13 + k31
  a1 <- k10 * k21 + k10 * k31 + k21 * k31 + k12 * k31 + k13 * k21
  a0 <- k10 * k21 * k31
  log(2) / min(abs(Re(polyroot(c(a0, a1, a2, 1)))))
}

subject_pars <- sim |>
  filter(!is.na(kel)) |>
  group_by(panel, product, id) |>
  slice(1) |>
  ungroup() |>
  mutate(t_exact = mapply(exact_thalf, kel, k12, k21, k13, k31))

nca_hl <- nca_ind |>
  filter(PPTESTCD == "half.life") |>
  group_by(panel, product) |>
  summarise(`NCA t1/2 mean` = mean(PPORRES, na.rm = TRUE),
            `NCA t1/2 median` = median(PPORRES, na.rm = TRUE), .groups = "drop")

published_hl <- tibble::tribble(
  ~panel,           ~product,     ~pub_mean, ~pub_sd,
  "FP 3 x 100 ug",  "Reference",  13.69,     2.506,
  "FP 3 x 100 ug",  "Test",       13.75,     2.511,
  "FP 3 x 250 ug",  "Reference",  14.53,     2.859,
  "FP 3 x 250 ug",  "Test",       14.64,     2.875,
  "FP 3 x 500 ug",  "Reference",  14.04,     2.157,
  "FP 3 x 500 ug",  "Test",       14.03,     2.172,
  "SALM 3 x 50 ug", "Reference",  mean(c(12.63, 12.36, 12.00)), mean(c(1.761, 2.023, 2.289)),
  "SALM 3 x 50 ug", "Test",       mean(c(12.60, 12.42, 12.02)), mean(c(1.780, 2.153, 2.301))
)

hl_tab <- subject_pars |>
  group_by(panel, product) |>
  summarise(`Exact t1/2 mean` = mean(t_exact),
            `Exact t1/2 median` = median(t_exact),
            `Exact t1/2 SD` = sd(t_exact), .groups = "drop") |>
  left_join(nca_hl, by = c("panel", "product")) |>
  mutate(panel = as.character(panel)) |>
  left_join(published_hl, by = c("panel", "product")) |>
  mutate(`Median vs published mean, % diff` = 100 * (`NCA t1/2 median` / pub_mean - 1))

hl_tab |>
  dplyr::rename("Substance | dose" = panel, "Product" = product,
                "Published mean (h)" = pub_mean, "Published SD (h)" = pub_sd) |>
  knitr::kable(digits = 2,
               caption = paste("Terminal half-life: exact eigenvalue solution and",
                               "PKNCA lambda-z on the simulated cohort, against",
                               "Rosenborg 2025 Table 3."))
```

```{r halflife-assert}
# The MEDIAN simulated half-life tracks the published mean; the simulated
# arithmetic mean and SD do not, and are not expected to.
stopifnot(all(abs(hl_tab$`Median vs published mean, % diff`) < 20))
# The published SDs span 1.76-2.88 h. Every simulated SD exceeds the largest of
# them, which is the fresh-draw-versus-shrunken-EBE signature described above.
stopifnot(all(hl_tab$`Exact t1/2 SD` > 2.9))
```

The medians land within 20% of the published means -- and the typical-value
calculation in the structural check above lands within 5% -- while the simulated
SDs are 3 to 17 times the published ones. The inflation is largest exactly where
the published omegas are largest: FP 3 x 500 ug carries `V2/F_OMEGA_IIV` = 1.739
(the eta distribution the Results section calls "remarkably skewed") and the
salmeterol model is the only one retaining etas on both distribution clearances.
Neither is a defect in the packaged models -- the same omegas reproduce the
concentration profiles and the bioequivalence assessment below -- but it does
mean a *typical-value* or *median* statistic is the right thing to compare
against a table of post-hoc half-lives.

# Replicate Table 4: bioequivalence assessment

Table 4 is the paper's headline result. Because `F4_rel` is anchored on a
reference product with `F = 1`, the model-predicted relative extent of
bioavailability for a subject is simply that subject's test/reference AUC ratio,
and the relative rate marker is the test/reference Cmax ratio. Both are computed
here as geometric means with 90% confidence intervals, on the paired crossover
data, exactly as a bioequivalence test would be run.

```{r table-4}
paired <- nca_ind |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs")) |>
  mutate(id = sub("\\|[^|]*$", "", subject)) |>
  select(panel, product, id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = product, values_from = PPORRES) |>
  filter(is.finite(Test), is.finite(Reference), Test > 0, Reference > 0) |>
  mutate(lr = log(Test / Reference))

be <- paired |>
  group_by(panel, PPTESTCD) |>
  summarise(
    n = dplyr::n(),
    Estimate = exp(mean(lr)),
    Lower.ci = exp(mean(lr) - stats::qt(0.95, dplyr::n() - 1) * stats::sd(lr) / sqrt(dplyr::n())),
    Upper.ci = exp(mean(lr) + stats::qt(0.95, dplyr::n() - 1) * stats::sd(lr) / sqrt(dplyr::n())),
    .groups = "drop"
  ) |>
  mutate(Parameter = ifelse(PPTESTCD == "aucinf.obs", "F4_rel", "Cmax_ratio"))

published_be <- tibble::tribble(
  # Rosenborg 2025 Table 4, model-based columns
  ~panel,           ~Parameter,   ~pub_est, ~pub_lo, ~pub_hi,
  "FP 3 x 100 ug",  "F4_rel",     1.023,    0.991,   1.056,
  "FP 3 x 250 ug",  "F4_rel",     1.048,    0.992,   1.107,
  "FP 3 x 500 ug",  "F4_rel",     0.935,    0.898,   0.974,
  "FP 3 x 100 ug",  "Cmax_ratio", 0.854,    0.827,   0.883,
  "FP 3 x 250 ug",  "Cmax_ratio", 0.980,    0.927,   1.037,
  "FP 3 x 500 ug",  "Cmax_ratio", 0.910,    0.873,   0.948,
  # Salmeterol Table 4 rows are per study; the pooled model has one value,
  # so the three published studies are averaged.
  "SALM 3 x 50 ug", "F4_rel",     mean(c(1.075, 1.013, 0.998)),
                                  mean(c(1.045, 0.974, 0.960)),
                                  mean(c(1.107, 1.054, 1.037)),
  "SALM 3 x 50 ug", "Cmax_ratio", mean(c(0.966, 0.890, 0.887)),
                                  mean(c(0.938, 0.855, 0.854)),
                                  mean(c(0.995, 0.927, 0.922))
)

be_tab <- be |>
  mutate(panel = as.character(panel)) |>
  left_join(published_be, by = c("panel", "Parameter")) |>
  mutate(`% diff` = 100 * (Estimate / pub_est - 1)) |>
  select(panel, Parameter, n, Estimate, Lower.ci, Upper.ci,
         pub_est, pub_lo, pub_hi, `% diff`) |>
  arrange(Parameter, panel)

be_tab |>
  dplyr::rename(
    "Substance | dose"   = panel,
    "Parameter"          = Parameter,
    "N"                  = n,
    "Simulated estimate" = Estimate,
    "Simulated lower"    = Lower.ci,
    "Simulated upper"    = Upper.ci,
    "Published estimate" = pub_est,
    "Published lower"    = pub_lo,
    "Published upper"    = pub_hi
  ) |>
  knitr::kable(digits = 3,
               caption = paste("Replicates Rosenborg 2025 Table 4: model-based",
                               "relative extent (F4_rel) and relative rate",
                               "(Cmax_ratio) of bioavailability, test vs",
                               "reference, with 90% confidence intervals."))
```

## A structural identity behind Table 4

Because the reference product is anchored at `F = 1` and the two periods of a
crossover share every disposition parameter, the test/reference AUC ratio for
the typical subject is *exactly* `exp(lfdepot)` -- `AUC = Dose * F / CL`, and
`CL` cancels. That makes the relative-extent column of Table 4 a parameter-free
identity check on `lfdepot` and on the `f(depot)` encoding, independent of any
sampling noise. Integrating the typical-value profile over 400 h (more than
25 terminal half-lives) recovers it to better than 0.1%:

```{r frel-identity}
identity_events <- function(nm, dose) {
  base <- tidyr::expand_grid(product = c("Reference", "Test"),
                             time = c(0, seq(0.005, 400, length.out = 4000))) |>
    mutate(id = as.integer(factor(product)),
           FORM_WIXELA_INHUB = as.numeric(product == "Test"),
           amt = NA_real_, evid = 0L, cmt = "central")
  dosing <- base |> distinct(id, product, FORM_WIXELA_INHUB) |>
    mutate(time = 0, amt = dose, evid = 1L, cmt = "depot")
  out <- bind_rows(dosing, base) |> arrange(id, time, desc(evid))
  if (nm == "Rosenborg_2025_salmeterol") out$SIDN <- out$id  # >= 2 levels required
  out
}
doses <- c(300, 750, 1500, 150)

frel_identity <- do.call(rbind, Map(function(nm, dose) {
  ev <- identity_events(nm, dose)
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(uis[[nm]]), ev,
                                     useLinCmt = FALSE, keep = "product"))
  auc <- vapply(split(s, s$product), function(d) {
    d <- d[order(d$time), ]
    sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  }, numeric(1))
  data.frame(
    Model = nm,
    `Simulated AUC ratio` = unname(auc[["Test"]] / auc[["Reference"]]),
    `exp(lfdepot)` = exp(uis[[nm]]$iniDf$est[!is.na(uis[[nm]]$iniDf$ntheta) &
                                               uis[[nm]]$iniDf$name == "lfdepot"]),
    check.names = FALSE
  )
}, model_names, doses))
frel_identity$`% diff` <-
  100 * (frel_identity$`Simulated AUC ratio` / frel_identity$`exp(lfdepot)` - 1)

knitr::kable(frel_identity, digits = 4, row.names = FALSE,
             caption = "Typical-value test/reference AUC ratio vs the packaged F4_rel.")
stopifnot(all(abs(frel_identity$`% diff`) < 0.5))
```

## Regression assertions

```{r table-4-assert}
# Every 90% CI lies inside the conventional 0.80-1.25 bioequivalence window,
# which is what Rosenborg 2025 reports for all twelve rows of Table 4.
stopifnot(all(be_tab$Lower.ci > 0.80), all(be_tab$Upper.ci < 1.25))

# Simulated point estimates agree with the published model-based ones.
stopifnot(nrow(be_tab) == 8L, all(abs(be_tab$`% diff`) < 6))
```

Both bioequivalence conclusions of the paper are reproduced: every 90% confidence
interval falls inside 0.80-1.25, and the simulated point estimates match the
published model-based estimates to within about 4%. The residual gap between the
population estimate and `exp(lfdepot)` is finite-sample noise in the drawn
`etalfdepot` (with `n = 66` and an omega of 0.16-0.27, one standard error on the
geometric mean is already 2-3%), which is why the identity above is checked on
the typical-value solve rather than on the cohort. The simulated intervals are a
little narrower than the published ones because the simulation carries no
residual error into the post-hoc estimates and has no dropouts.

# Assumptions and deviations

- **Dose units.** The paper reports `DOSE` in ug and `DV` in ng/L while `CL/F`
  is in L/h and `V1/F` in L, so `central / vc` is in ug/L. The observation is
  written `Cc <- 1000 * central / vc` to land on the ng/L scale on which the
  concentrations, the LLOQ of 1 ng/L and `CP_ADD` are reported. Doses are
  supplied in ug. The conversion is verified by the reproduction of Table 3
  above.
- **Log-scale `Q2/F` and `Q3/F` in the FP models.** The supplement's FP code
  writes `Q2 = THETA(2) + ETA(2)` on the linear scale, but the accompanying
  OMEGA is fixed to zero (Table 2 reports `NA` for both IIV rows) and
  Rosenborg 2025 states that parameters carrying random effects are log-normal
  "except for pharmacokinetic parameters without consideration of random
  effects". With no eta, the linear THETA and the library's canonical
  log-transformed `lq` / `lq2` describe the same typical value, so the canonical
  form is used. The salmeterol code writes `Q2 = EXP(MU_2 + ETA(2))` and is
  encoded directly.
- **Errata: which parameter carries the salmeterol interstudy random effect.**
  Rosenborg 2025 Results describes "the jointly estimated random effect of study
  on apparent elimination and inter compartmental clearances of SALM". Three
  other places in the paper disagree and are followed here: the supplement's
  NONMEM code writes `F4 = EXP(TREA1*(MU_11 + ETA(8) + ETA(9)) + TREA2*0)` with
  `$LEVEL STUD=(9[1])`, Figure 1 labels the salmeterol deposition compartment
  "F4_rel, ETA8/9, TREA", and Table 2 names the row `F4_rel_OMEGA_ISV`. The
  sentence in Results appears to be a drafting slip.
- **No demographic covariates.** The paper deliberately fitted none, so the
  virtual cohort carries no weight, age, sex or race columns. Generalisation
  beyond a homogeneous healthy-adult panel is not supported by this model; the
  Discussion states this explicitly as a limitation.
- **Crossover sequence.** Half the simulated subjects take the reference product
  first and half the test product first. The model has no period or sequence
  effect, so this is cosmetic; it is included so the event tables look like the
  study they represent.
- **Absolute bioavailability is not identifiable.** With no parenteral reference
  treatment, all clearances and volumes are apparent and biased upward relative
  to intravenous estimates. The Discussion infers an absolute bioavailability
  just under 15% by comparison with Thorsson 2001, but that value is not part of
  this model.
- **Half-life comparison uses the typical value and the median, not the
  simulated mean.** Table 3's half-lives are means of *post-hoc* (empirical
  Bayes) individual estimates, which the data shrink toward the typical value;
  the published SDs are 1.76-2.88 h. Simulation draws each eta afresh from the
  full published omega with no shrinkage and, because Rosenborg 2025 reports no
  correlations between etas, no covariance between the six disposition
  parameters. Terminal half-life is strongly non-linear in those six, so the
  simulated distribution is right-skewed with an SD 3 to 17 times the published
  one and an inflated arithmetic mean. The typical-value and median statistics
  are the like-for-like comparisons and both agree with the paper; see the
  half-life section for the full accounting. No parameter was adjusted.
- **Below-LLOQ handling.** The published fit used the M3-style likelihood of
  Bauer 2019 for the 1 ng/L LLOQ. Simulation from the packaged model does not
  censor, so the simulated tails extend below 1 ng/L; the LLOQ is drawn on the
  figure for reference.
- **Excluded from the packaged models.** The supplement's Sections 7.1 and 7.2
  document a biphasic-absorption model fitted across studies with `F4_rel` fixed
  to 1, and an alternative parameterisation placing the product effect on `CL/F`
  and `V1/F` instead of on `F4_rel`. Both were reported as rejected exploratory
  alternatives rather than final models, so neither is packaged.
