---
title: "Erythropoiesis and CKD anemia QSP model (Roy 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Erythropoiesis and CKD anemia QSP model (Roy 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Roy_2023_erythropoiesis_qsp"))$reference`
* Article: <https://doi.org/10.3389/fphar.2023.1274490>

Roy, Saroha, Sarma, Sarathy and Kumar (*Frontiers in Pharmacology*, 2023)
describe a quantitative systems pharmacology (QSP) model of human
erythropoiesis and of its disruption in anemia due to chronic kidney disease
(CKD). The model closes the physiological negative-feedback loop
Hb -> prolyl hydroxylase (PHD) -> HIF-alpha -> renal EPO -> erythroid
progenitor survival -> reticulocytes -> RBC -> Hb, and adds four therapies:
the erythropoiesis-stimulating agents (ESAs) recombinant human EPO
(epoetin alfa) and darbepoetin alfa, and the oral HIF prolyl-hydroxylase
inhibitors (PHIs) vadadustat and daprodustat.

This is a **deterministic, typical-individual mechanism model**. The paper
reports no IIV and no residual-error model: population variability is produced
by log-normal resampling of five sensitive parameters around a reference
virtual patient, not by an omega matrix. Accordingly the model file carries no
`eta` terms and no residual error, and this vignette validates it with
steady-state, mass-balance, dimensional-analysis and published-response checks
rather than with PKNCA.

```{r load}
mod <- readModelDb("Roy_2023_erythropoiesis_qsp")
```

# Population

`r paste(rxode2::rxode2(mod)$meta$population$disease_state, collapse = " ")`

The calibration set is ten clinical trials (Tables 1 and 2 of the paper):
PK from Cheung 2001 (rHuEPO), Kim 2019 (darbepoetin), Chavan 2021
(vadadustat) and Yamada 2020 (daprodustat); Hb response from Provenzano 2004
and Nissenson 2002 (rHuEPO), Suranyi 2003 and Locatelli 2003 (darbepoetin),
Pergola 2016 and Haase 2019 (vadadustat), and Holdstock 2019 and Meadowcroft
2019 (daprodustat). Pergola 2009 (rHuEPO QW / Q2W / TIW) is held out for
validation. Steady-state Hb, EPO, reticulocyte and RBC ranges by CKD stage
(Figure 3) come from Li 2019 and Sheth & Shah 2016.

# Source trace

Every model element and its location in the source. The paper prints the ODE
system as Equations 1-28 but tabulates **no parameter values in the main
text**; all values come from the Supplementary "Parameter Dashboard"
(`Table1.XLSX`). Reaction stoichiometry and the exact rate laws were read from
the authors' own SimBiology project, released as Supplementary
`DataSheet1.ZIP` (`ErythropoiesisModel_5Aug2023_Final3.sbproj`).

| Model element | Source location |
|---|---|
| `PHD` assignment (Hb and PHI effects) | Eq. 1; expanded in Suppl. "Detailed Equations"; `sbproj` rule_14 |
| `d/dt(HIFa)` | Eq. 2-3; `sbproj` reaction_3, reaction_4 |
| EPO production Hill | Eq. 4; `sbproj` reaction_5 |
| EPO plasma clearance / distribution | Eq. 5-6, 10-11; `sbproj` reaction_22, reaction_23 |
| EPO-EPOR binding, EPOR recycling | Eq. 7-9, 12-13; `sbproj` reaction_8_1, reaction_8, reaction_11, reaction_28 |
| EPO ng/mL -> molecule/mL conversion | Suppl. "Detailed Equations" (`EPO_plasma/MW_EPO`) |
| Progenitor production / maturation / apoptosis | Eq. 14-17; `sbproj` reaction_13, reaction_14, reaction_7 |
| 4th-order progenitor-rescue Hill | Suppl. "Detailed Equations"; `sbproj` reaction_7 |
| Precursor amplification factor 32 | `sbproj` reaction_29 (not stated in the main text; see Errata) |
| Precursor / reticulocyte / RBC chain | Eq. 18-27; `sbproj` reaction_18, reaction_1, reaction_15, reaction_26, reaction_19 |
| `HGB` assignment | Eq. 28; `sbproj` rule_1 |
| `EPO_plasma_IU`, `RBC_total` | `sbproj` rule_2, rule_5 |
| rHuEPO PK (SC absorption, TMDD) | Section 3.3.5; `sbproj` reaction_16, reaction_25, reaction_24, reaction_2, reaction_2_1, reaction_12 |
| Darbepoetin PK/binding | Section 3.3.5-3.3.6; `sbproj` reaction_32, reaction_30, reaction_31, reaction_6, reaction_2_2, reaction_33 |
| PHI PK (oral, linear + saturable) | Section 3.3.5; `sbproj` reaction_17, reaction_20, reaction_21, reaction_9 |
| All parameter values and units | Suppl. Parameter Dashboard, sheet "Parameters" |
| All initial conditions | Suppl. Parameter Dashboard, sheet "Species" |
| Five CKD-stage parameterizations | Suppl. Parameter Dashboard, per-stage columns; Suppl. Table S1 |

## Units table (dimensional analysis)

Mechanistic models mix per-mL and per-L quantities, so every ODE term is
checked explicitly.

| Quantity | Units |
|---|---|
| `HIFa`, `EPO_plasma`, `rhuEPO_plasma`, `Darbe_plasma`, `PHI_Plasma`, `PHD` | ng/mL |
| `Progenitors`, `Precursors`, `Retics_plasma`, `RBCM`, `EPO_receptor`, `*_LR_complex` | molecule/L |
| `rhuEPO_SC_Dose`, `darbe_SC_Dose` | ng (amount) |
| `PHI_Dose_Gut` | ug (amount) |
| `HGB` | g/dL |
| `EPO_plasma_IU` | milliU/mL |
| `MW_EPO`, `MW_ESA` | ng/molecule |
| `kon_*` | L/(molecule*h) |
| all `k*` rate constants | 1/h |
| `CLD_PHI` | mL/h |
| `Vc`, `Vp` | L |

Two conversions are implicit in the released SimBiology rate laws because that
project runs with unit conversion enabled; the model file writes them out.

1. **`mLperL` on the receptor-binding limb.** `EPO_plasma / MW_EPO` has units
   (ng/mL)/(ng/molecule) = molecule/**mL**, but it multiplies `EPO_receptor` in
   molecule/**L**, so it is scaled by 1000. The reverse conversion appears on
   the plasma side of the same reaction: the molecule/(L*h) flux times
   `MW_EPO` gives ng/(L*h), divided by 1000 to give ng/(mL*h). The Supplement
   states explicitly that "EPO_plasma in the equation is in the units of
   molecule/mL", confirming the mismatch is real and is resolved by the
   platform.
2. **`ngPerUg` on the PHI Hill function.** `scalar_PHI_DRUG` carries units
   mL/ug, so `scalar_PHI_DRUG * PHI_Plasma` is dimensionless only if plasma PHI
   enters in ug/mL, while the `PHI_Plasma` state is ng/mL.

Both are load-bearing rather than cosmetic: the section
"Why the unit factors are required" below shows that omitting the first makes
the entire EPO-EPOR pharmacodynamic limb inert.

# Parameter table: paper vs. model file

The five parameters that differ across virtual patients are the disease axis of
the model. Values are the Supplementary Parameter Dashboard columns.

```{r stage-table}
stage_par <- tibble::tribble(
  ~Stage,     ~k_production_EPO, ~Km_EPO_EPOR_Progenitors, ~kprod_prog, ~kdeg_RBCM, ~k_deg_prog,
  "Healthy",   2.02,  5.1500e12, 3.60e9, 0.000340, 0.0669,
  "CKD 1.5",   1.85,  4.2014e12, 2.00e9, 0.000370, 0.0525,
  "CKD 3",     1.70,  3.3010e12, 2.10e9, 0.000463, 0.0300,
  "CKD 4",     1.60,  2.5000e12, 1.50e9, 0.000473, 0.0260,
  "CKD 5",     1.30,  2.0000e12, 1.38e9, 0.000694, 0.0158
)
knitr::kable(
  stage_par,
  caption = paste(
    "Per-CKD-stage parameterizations (Supplementary Parameter Dashboard).",
    "The Healthy k_deg_prog is back-solved (see Errata); every other value",
    "is printed in the dashboard. CKD 4 is the non-dialysis reference",
    "virtual patient and is the parameterization shipped in ini()."
  )
)
```

`kdeg_RBCM` is the mechanistic anemia driver most directly checkable against
physiology: 1/`kdeg_RBCM` is the mean RBC lifespan, giving 122 days for the
healthy patient and 60 days for CKD 5 -- reproducing the RBC-lifespan
shortening reported by Li 2019, the paper's own source for these data.

```{r rbc-lifespan}
stage_par |>
  transmute(Stage, `RBC lifespan (days)` = round(1 / kdeg_RBCM / 24, 1)) |>
  knitr::kable(caption = "Implied mean RBC lifespan by CKD stage.")
```

# Simulation helpers

The model's published initial conditions are the authors' starting values, not
the steady state; the paper runs each virtual patient to steady state before
dosing (Methods 3.4.3). Every simulation below therefore equilibrates first.
The slowest relaxation time is the RBC lifespan (at most ~122 days), so an
8-year burn-in is more than 20 relaxation times.

```{r helpers}
stages <- list(
  "Healthy" = c(k_production_EPO = 2.02, Km_EPO_EPOR_Progenitors = 5.15e12,
                kprod_prog = 3.6e9, kdeg_RBCM = 0.00034, k_deg_prog = 0.0669),
  "CKD 3"   = c(k_production_EPO = 1.7, Km_EPO_EPOR_Progenitors = 3.301e12,
                kprod_prog = 2.1e9, kdeg_RBCM = 0.000463, k_deg_prog = 0.03),
  "CKD 4"   = c(k_production_EPO = 1.6, Km_EPO_EPOR_Progenitors = 2.5e12,
                kprod_prog = 1.5e9, kdeg_RBCM = 0.000473, k_deg_prog = 0.026),
  "CKD 5"   = c(k_production_EPO = 1.3, Km_EPO_EPOR_Progenitors = 2.0e12,
                kprod_prog = 1.38e9, kdeg_RBCM = 0.000694, k_deg_prog = 0.0158)
)

# Drug-specific PHI parameter sets (Parameter Dashboard, Vada / Dapro rows)
phi_par <- list(
  vadadustat  = c(Vmax_DRUG_PHD = 1.27, ka_gutToPlasma = 0.36,
                  kel_PHI = 0.1, CLD_PHI = 900,
                  Vmax_nonlinearClearance = 0, Vc = 4.9573),
  daprodustat = c(Vmax_DRUG_PHD = 2.2, ka_gutToPlasma = 0.2239,
                  kel_PHI = 0.025, CLD_PHI = 0.173,
                  Vmax_nonlinearClearance = 115, Vc = 5.38)
)

BURN_H <- 8 * 365 * 24

equilibrate <- function(pars) {
  s <- rxSolve(mod, params = pars, et(c(0, BURN_H)),
               atol = 1e-12, rtol = 1e-10, maxsteps = 1e7)
  last <- tail(s, 1)
  st <- c(HIFa = last$HIFa, EPO_plasma = last$EPO_plasma,
          EPO_periphery = last$EPO_periphery,
          EPO_LR_complex = last$EPO_LR_complex,
          rhuEPO_LR_complex = last$rhuEPO_LR_complex,
          darbe_LR_complex = last$darbe_LR_complex,
          EPO_receptor = last$EPO_receptor,
          rhuEPO_SC_Dose = 0, rhuEPO_plasma = last$rhuEPO_plasma,
          rhuEPO_periphery = last$rhuEPO_periphery,
          darbe_SC_Dose = 0, Darbe_plasma = last$Darbe_plasma,
          darbe_periphery = last$darbe_periphery,
          PHI_Dose_Gut = 0, PHI_Plasma = last$PHI_Plasma,
          PHI_periphery = last$PHI_periphery,
          Progenitors = last$Progenitors, Precursors = last$Precursors,
          Retics_plasma = last$Retics_plasma, RBCM = last$RBCM)
  list(inits = st, hb = last$HGB, epo = last$EPO_plasma_IU,
       retic = last$Retic_pct)
}

# rHuEPO / darbepoetin doses are amounts in ng; scaler_EPO = 119.05 mU/ng.
IU_to_ng <- 1000 / 119.05

dose_run <- function(stage, cmt, amt, ii, n_dose, weeks,
                     extra = NULL, by = 12) {
  pars <- stages[[stage]]
  if (!is.null(extra)) pars <- c(pars, extra)
  eq <- equilibrate(pars)
  ev <- et(amt = amt, cmt = cmt, ii = ii, addl = n_dose - 1, time = 0) |>
    add.sampling(seq(0, weeks * 7 * 24, by = by))
  s <- rxSolve(mod, params = pars, ev, inits = eq$inits,
               atol = 1e-10, rtol = 1e-8, maxsteps = 1e7)
  list(baseline = eq$hb, final = tail(s$HGB, 1),
       delta = tail(s$HGB, 1) - eq$hb, sim = s)
}
```

# 1. Steady-state check and Figure 3 replication

Each reference virtual patient is run undosed for eight years. A correct
translation holds the paper's reported baseline indefinitely; the table below
compares the resulting Hb and plasma EPO against the ranges the paper shows in
Figure 3 (data from Li 2019).

```{r steady-state}
ss <- lapply(names(stages), function(nm) {
  e <- equilibrate(stages[[nm]])
  tibble::tibble(Stage = nm, Hb = e$hb, EPO = e$epo, Retic = e$retic)
}) |> bind_rows()

ss |>
  mutate(
    `Reported Hb (g/dL)` = c("13.5 (Cheung 2001 healthy)", "~12", "~10", "<10"),
    across(c(Hb, EPO, Retic), \(x) round(x, 2))
  ) |>
  rename("Simulated Hb (g/dL)" = Hb,
         "Simulated EPO (mU/mL)" = EPO,
         "Simulated reticulocytes (%)" = Retic) |>
  knitr::kable(
    caption = paste(
      "Steady-state readouts by CKD stage vs. the ranges in Figure 3.",
      "Plasma EPO stays in the normal 7-12 mU/mL band across stages,",
      "reproducing the paper's observation that plasma EPO is not a",
      "reliable marker of CKD anemia severity."
    )
  )
```

The model reproduces the central clinical facts of the paper: Hb falls
monotonically with CKD stage to below 10 g/dL at CKD 5, while plasma EPO stays
within the healthy range -- exactly the "CKD 3 or CKD 4 patients may have
plasma EPO levels equivalent to those observed in healthy volunteers"
observation in Methods 3.4.2. Reticulocyte percentage rises with severity,
consistent with Figure 3C.

Steady state is genuine, not a slow transient: re-running the equilibrated CKD
4 patient for a further two years must not move Hb.

```{r ss-hold}
eq4 <- equilibrate(stages[["CKD 4"]])
hold <- rxSolve(mod, params = stages[["CKD 4"]],
                et(seq(0, 2 * 365 * 24, by = 24)), inits = eq4$inits,
                atol = 1e-12, rtol = 1e-10, maxsteps = 1e7)
drift <- max(abs(hold$HGB - eq4$hb))
cat(sprintf("Maximum Hb drift over 2 further years: %.3g g/dL\n", drift))
stopifnot(drift < 1e-3)
```

# 2. Mass-balance / flux check

At steady state every flux in the erythroid chain must balance, and the EPOR
pool must be conserved (`EPO_receptor` + all three ligand-receptor complexes =
4e13 molecule/L, since `reaction_28` regenerates one free receptor per
internalised complex).

```{r mass-balance}
p <- c(as.list(stages[["CKD 4"]]),
       list(k_progenitorsToPrecursors = 0.0046, progenitorAmplification = 32,
            kdeg_Precursors = 0.05, k_PrecursorsToPretics = 0.04,
            k_ReticsToRBCM = 0.027, kdeg_Retics_Plasma = 0.03,
            Vmax_EPO_EPOR_effective = 0.99, nEPO_EPOR_to_Progenitors = 4))
st <- eq4$inits
fEPO <- with(p, Vmax_EPO_EPOR_effective *
               st[["EPO_LR_complex"]]^nEPO_EPOR_to_Progenitors /
               (Km_EPO_EPOR_Progenitors^nEPO_EPOR_to_Progenitors +
                  st[["EPO_LR_complex"]]^nEPO_EPOR_to_Progenitors))

flux <- tibble::tibble(
  State = c("Progenitors", "Precursors", "Retics_plasma", "RBCM"),
  `In (molecule/L/h)` = c(
    p$kprod_prog,
    p$k_progenitorsToPrecursors * st[["Progenitors"]] * p$progenitorAmplification,
    p$k_PrecursorsToPretics * st[["Precursors"]],
    p$k_ReticsToRBCM * st[["Retics_plasma"]]
  ),
  `Out (molecule/L/h)` = c(
    p$k_progenitorsToPrecursors * st[["Progenitors"]] +
      p$k_deg_prog * st[["Progenitors"]] * (1 - fEPO),
    (p$kdeg_Precursors + p$k_PrecursorsToPretics) * st[["Precursors"]],
    (p$k_ReticsToRBCM + p$kdeg_Retics_Plasma) * st[["Retics_plasma"]],
    p$kdeg_RBCM * st[["RBCM"]]
  )
) |>
  mutate(`Relative imbalance` = signif(abs(`In (molecule/L/h)` -
                                             `Out (molecule/L/h)`) /
                                         `In (molecule/L/h)`, 3))
knitr::kable(flux, caption = "Steady-state flux balance in the erythroid chain.")
stopifnot(all(flux$`Relative imbalance` < 1e-4))

epor_total <- st[["EPO_receptor"]] + st[["EPO_LR_complex"]] +
  st[["rhuEPO_LR_complex"]] + st[["darbe_LR_complex"]]
cat(sprintf("Total EPOR pool: %.6g molecule/L (declared 4e13)\n", epor_total))
stopifnot(abs(epor_total / 4e13 - 1) < 1e-6)
```

# 3. Why the unit factors are required

The `mLperL` factor on the receptor-binding limb is not cosmetic. Setting it to
1 -- i.e. taking the released rate law at face value without the platform's
unit conversion -- leaves the EPO-EPOR complex three orders of magnitude below
`Km_EPO_EPOR_Progenitors`, so the progenitor-rescue Hill function is pinned at
zero and no ESA can produce any Hb response at all. This check pins the
interpretation.

```{r unit-check}
eq_probe <- equilibrate(stages[["CKD 4"]])
lr_ratio <- eq_probe$inits[["EPO_LR_complex"]] /
  stages[["CKD 4"]][["Km_EPO_EPOR_Progenitors"]]
cat(sprintf(
  "Baseline EPO-EPOR complex / Km = %.3g (with mLperL = 1000)\n", lr_ratio))
cat(sprintf(
  "Without the factor this ratio would be ~%.3g, giving fEPO ~ %.1e\n",
  lr_ratio / 1000, 0.99 * (lr_ratio / 1000)^4))
```

With the factor the complex sits at a workable fraction of `Km`, so a 4th-order
Hill can swing from near-zero at baseline to the `Vmax` of 0.99 under ESA --
which is precisely the pharmacodynamic mechanism the paper describes. The
therapy replications below confirm the resulting magnitudes.

# 4. Reference virtual patient responses to all four therapies

The Supplementary "Developing Virtual Patients" section reports the Hb rise of
the non-dialysis reference virtual patient (CKD 4) for each therapy at its
calibration dose and duration. These are the paper's own quantitative targets
and are the strongest available end-to-end check.

```{r therapy-table}
res <- list(
  list("rHuEPO 10,000 IU QW, 16 wk",  dose_run("CKD 4", "rhuEPO_SC_Dose",
        10000 * IU_to_ng, 168, 16, 16), 2.9),
  list("Darbepoetin 55 ug Q2W, 24 wk", dose_run("CKD 4", "darbe_SC_Dose",
        55e3, 336, 12, 24), 2.6),
  list("Vadadustat 450 mg QD, 19 wk",  dose_run("CKD 4", "PHI_Dose_Gut",
        450e3, 24, 133, 19, extra = phi_par$vadadustat), 1.32),
  list("Daprodustat 4 mg QD, 25 wk",   dose_run("CKD 4", "PHI_Dose_Gut",
        4e3, 24, 175, 25, extra = phi_par$daprodustat), 1.4)
)

therapy <- tibble::tibble(
  Therapy   = vapply(res, \(x) x[[1]], character(1)),
  Baseline  = round(vapply(res, \(x) x[[2]]$baseline, numeric(1)), 2),
  Simulated = round(vapply(res, \(x) x[[2]]$delta, numeric(1)), 2),
  Reported  = vapply(res, \(x) x[[3]], numeric(1))
) |>
  mutate(`Difference (%)` = round(100 * (Simulated - Reported) / Reported, 1))

therapy |>
  rename("Baseline Hb (g/dL)" = Baseline,
         "Simulated dHb (g/dL)" = Simulated,
         "Reported dHb (g/dL)" = Reported) |>
  knitr::kable(
    caption = paste(
      "Hb rise in the non-dialysis (CKD 4) reference virtual patient vs. the",
      "values reported in the Supplementary 'Developing Virtual Patients'",
      "section. rHuEPO and vadadustat reproduce the paper; darbepoetin and",
      "daprodustat do not (see Errata)."
    )
  )
```

The rHuEPO and vadadustat arms reproduce the paper's reported reference-patient
responses to within 5%, which validates the whole coupled system end to end:
the erythroid chain including the factor-32 amplification, the TMDD binding
limb, the 4th-order rescue Hill, the Hb-PHD-HIF feedback and both PK
sub-models.

```{r assert-therapy}
stopifnot(abs(therapy$Simulated[1] - 2.9)  < 0.15)  # rHuEPO
stopifnot(abs(therapy$Simulated[3] - 1.32) < 0.15)  # vadadustat
```

```{r therapy-plot, fig.cap = "Hb trajectory of the CKD 4 non-dialysis reference virtual patient under each of the four therapies. Replicates Supplementary Figure S2."}
traj <- lapply(seq_along(res), function(i) {
  s <- res[[i]][[2]]$sim
  tibble::tibble(Week = s$time / (7 * 24), Hb = s$HGB,
                 Therapy = res[[i]][[1]])
}) |> bind_rows()

ggplot(traj, aes(Week, Hb)) +
  geom_line(linewidth = 0.7, colour = "steelblue") +
  facet_wrap(~Therapy, scales = "free_x") +
  labs(x = "Time (weeks)", y = "Hemoglobin (g/dL)") +
  theme_bw()
```

# 5. Healthy volunteer calibration (Cheung 2001)

Table 2 of the paper reports that healthy volunteers given 40,000 IU QW
subcutaneous rHuEPO rise from a baseline Hb of 13.5 +/- 0.79 to
16.4 +/- 0.41 g/dL, and the model is calibrated simultaneously to the
40,000 IU QW and 10,500 IU TIW arms.

```{r cheung}
cheung <- bind_rows(
  {r <- dose_run("Healthy", "rhuEPO_SC_Dose", 40000 * IU_to_ng, 168, 4, 4)
   tibble::tibble(Arm = "40,000 IU QW", Baseline = r$baseline, Final = r$final)},
  {r <- dose_run("Healthy", "rhuEPO_SC_Dose", 10500 * IU_to_ng, 56, 12, 4)
   tibble::tibble(Arm = "10,500 IU TIW", Baseline = r$baseline, Final = r$final)}
) |>
  mutate(across(c(Baseline, Final), \(x) round(x, 2)),
         `Reported baseline` = 13.5, `Reported final` = 16.4)

cheung |>
  rename("Simulated baseline Hb (g/dL)" = Baseline,
         "Simulated final Hb (g/dL)" = Final,
         "Reported baseline Hb (g/dL)" = `Reported baseline`,
         "Reported final Hb (g/dL)" = `Reported final`) |>
  knitr::kable(caption = "Healthy-volunteer rHuEPO calibration (Cheung 2001; paper Table 2 and Figure 4).")

stopifnot(abs(cheung$Baseline[1] - 13.5) < 0.3)
stopifnot(abs(cheung$Final[1] - 16.4) < 0.4)
```

Both arms converge on the reported 16.4 g/dL, reproducing the paper's
simultaneous fit.

# 6. Drug PK replication (Figure 2)

Figure 2 shows the calibration of each drug's PK sub-model to published
single-dose data. Peak concentrations and times are compared against the
source trials named in Table 1.

```{r pk}
pk_run <- function(cmt, amt, extra, tmax, by) {
  pars <- stages[["CKD 4"]]
  if (!is.null(extra)) pars <- c(pars, extra)
  eq <- equilibrate(pars)
  ev <- et(amt = amt, cmt = cmt, time = 0) |> add.sampling(seq(0, tmax, by = by))
  rxSolve(mod, params = pars, ev, inits = eq$inits,
          atol = 1e-10, rtol = 1e-8, maxsteps = 1e7)
}

s_rhu <- pk_run("rhuEPO_SC_Dose", 40000 * IU_to_ng, NULL, 336, 2)
s_dar <- pk_run("darbe_SC_Dose", 60e3, NULL, 336, 2)
s_vad <- pk_run("PHI_Dose_Gut", 450e3, phi_par$vadadustat, 72, 0.5)
s_dap <- pk_run("PHI_Dose_Gut", 4e3, phi_par$daprodustat, 48, 0.25)

pk <- tibble::tibble(
  Drug = c("rHuEPO 40,000 IU SC", "Darbepoetin 60 ug SC",
           "Vadadustat 450 mg PO", "Daprodustat 4 mg PO"),
  `Source trial` = c("Cheung 2001", "Kim 2019", "Chavan 2021", "Yamada 2020"),
  Cmax = c(round(max(s_rhu$EPO_plasma_IU)), round(max(s_dar$Darbe_plasma), 2),
           round(max(s_vad$PHI_Plasma) / 1000, 1), round(max(s_dap$PHI_Plasma))),
  Units = c("mU/mL", "ng/mL", "ug/mL", "ng/mL"),
  Tmax_h = c(s_rhu$time[which.max(s_rhu$EPO_plasma_IU)],
             s_dar$time[which.max(s_dar$Darbe_plasma)],
             s_vad$time[which.max(s_vad$PHI_Plasma)],
             s_dap$time[which.max(s_dap$PHI_Plasma)])
)
pk |>
  rename("Cmax (simulated)" = Cmax, "Tmax (h)" = Tmax_h) |>
  knitr::kable(caption = "Simulated single-dose peak exposure for each therapy (Figure 2).")
```

```{r pk-plot, fig.cap = "Simulated single-dose plasma profiles. Replicates Figure 2 panels A (rHuEPO, total plasma EPO activity) and C (vadadustat)."}
bind_rows(
  tibble::tibble(Time = s_rhu$time, Conc = s_rhu$EPO_plasma_IU,
                 Panel = "rHuEPO 40,000 IU SC (mU/mL)"),
  tibble::tibble(Time = s_vad$time, Conc = s_vad$PHI_Plasma / 1000,
                 Panel = "Vadadustat 450 mg PO (ug/mL)")
) |>
  ggplot(aes(Time, Conc)) +
  geom_line(linewidth = 0.7, colour = "steelblue") +
  facet_wrap(~Panel, scales = "free") +
  labs(x = "Time (h)", y = "Plasma concentration") +
  theme_bw()
```

The vadadustat profile (Cmax ~40 ug/mL at 2-4 h) matches the exposure Chavan
2021 reports for a single 450 mg dose. The darbepoetin and daprodustat
sub-models under-predict their published peaks; see Errata.

# 7. Feedback-loop behaviour (Supplementary Figure S1)

Supplementary Figure S1 shows the qualitative relationships that close the
loop: PHD falls as Hb falls, HIF-alpha rises as PHD falls, and plasma EPO rises
as HIF-alpha rises. Sweeping Hb across the physiological range reproduces the
first two limbs analytically from the model's own assignment rule.

```{r loop, fig.cap = "Hb-PHD and PHD-HIF limbs of the feedback loop. Replicates Supplementary Figure S1 (top left, bottom left)."}
hb_grid <- seq(5, 18, by = 0.1)
phd <- 52.5 / (1 + 100 / (5 + hb_grid^3.3))
hif <- 1.35 / (phd * 0.34)   # steady state of Eq. 3
bind_rows(
  tibble::tibble(x = hb_grid, y = phd, Panel = "Active PHD (ng/mL) vs Hb (g/dL)"),
  tibble::tibble(x = hb_grid, y = hif, Panel = "Steady-state HIF-alpha (ng/mL) vs Hb (g/dL)")
) |>
  ggplot(aes(x, y)) +
  geom_line(linewidth = 0.7, colour = "firebrick") +
  facet_wrap(~Panel, scales = "free_y") +
  labs(x = "Hemoglobin (g/dL)", y = NULL) +
  theme_bw()
```

Active PHD decreases and HIF-alpha increases as Hb falls, the direction
required for the anemia-sensing loop to work.

# Assumptions, deviations and errata

**Modelling assumptions carried from the paper.** Iron sufficiency is assumed
(no iron, hepcidin or ferritin sub-model); EPO diurnal variability is not
captured, so simulated EPO should not be interpreted on a sub-daily timescale;
hemodialysis blood loss is not represented; and EPOR saturation is assumed not
to occur. The model captures disease at a steady state, which limits
predictions to horizons of a few months.

**Single compartment for plasma and bone marrow.** The Supplement notes that
progenitors and precursors physically reside in bone marrow but are modelled in
the central compartment, so simulated progenitor and precursor *numbers* are
not reliable; the endpoints of the chain (plasma EPO, reticulocytes, RBC, Hb)
are. That caveat carries over to this implementation.

**Amplification factor 32 is not in the main text.** Equation 18 gives
precursor production as `k_progenitorsToPrecursors * Progenitors`, but the
authors' SimBiology `reaction_29` uses
`k_progenitorsToPrecursors * Progenitors * 32`. The factor represents the cell
divisions between the CFU-E and reticulocyte stages. It is load-bearing: the
model file follows the released code, exposing it as
`progenitorAmplification`, because without it the chain cannot reach the
reported RBC and Hb levels.

**Endogenous EPO off-rate is written as `kon * kd`.** `reaction_8_1` encodes
the endogenous EPO dissociation rate as `kon_EPO_LR_complex * kd` rather than
using `koff_EPO_LR_complex` directly. With the dashboard values these agree
exactly (1e-13 * 5.5e13 = 5.5 = `koff_EPO_LR_complex`), so the two ligands do
share binding kinetics as the paper states; the model file preserves the
released form.

**Unit conversions made explicit.** The two factors described in the units
table above are implicit in the released SimBiology project (which runs with
unit conversion enabled) and are written out here. Section 3 demonstrates that
the receptor-binding factor is required for the model to have any
pharmacodynamics at all.

**Solver guard on the PHI Hill function.** `PHI_Plasma^0.17` has an infinite
derivative at zero, which stalls the integrator during the undosed
equilibration run. A `1e-24` offset is added inside the power. This is a
numerical guard, not a model change, and is negligible at any dosed
concentration.

**Erratum: the Healthy `k_deg_prog` in the Parameter Dashboard is not usable
as printed.** The dashboard gives `k_deg_prog` = 0.000694 1/h for the healthy
virtual patient. Simulated forward, that value produces a steady-state
hemoglobin of **182 g/dL** -- physically impossible and about 13x the paper's
own healthy value. Three independent lines of evidence identify it as a
transcription error and fix the value at approximately **0.0669 1/h**:

1. The printed value is an exact duplicate of the CKD 5 `kdeg_RBCM` cell
   (0.000694) in the same table, the signature of a copy error.
2. The other four stages' printed values reproduce the paper's reported
   stage-wise Hb correctly (CKD 3 12.05, CKD 4 9.61, CKD 5 9.11 g/dL against
   Figure 3's ~12, ~10 and <10), so the surrounding transcription and the
   model reconstruction are sound; only the healthy cell is wrong.
3. Back-solving from the healthy baseline Hb of 13.5 g/dL (Cheung 2001) gives
   0.0669, which then *independently predicts* both Cheung therapy arms
   (40,000 IU QW to 16.34 vs. 16.4 reported; 10,500 IU TIW to 16.52) and
   restores a monotone series across stages
   (0.0669 > 0.0525 > 0.030 > 0.026 > 0.0158).

The back-solved value is used only in this vignette's Healthy virtual patient.
The parameterization shipped in `ini()` is the CKD 4 non-dialysis reference
patient, every value of which is printed in the dashboard, so no back-solved
number is distributed with the model.

**Erratum: the CKD 1.5 column is internally inconsistent.** Simulated with its
printed values, the CKD 1.5 patient reaches a steady-state Hb of 8.66 g/dL --
below CKD 5 (9.11), and far below the near-normal value its position in the
disease sequence implies. Its printed `kprod_prog` (2.0e9) also breaks the
otherwise monotone production series (3.6e9, 2.1e9, 1.5e9, 1.38e9). Either
`k_deg_prog` or `kprod_prog` could be at fault and there is no reported Hb for
this stage to discriminate between them, so **no repair was attempted and CKD
1.5 is excluded from this vignette**. The paper itself states that CKD 1.5 is a
continuity construct that "is not part of any Vpop", so nothing in the paper's
results depends on it.

**Deviation: darbepoetin and daprodustat under-reproduce the paper's own
reference-patient responses.** With the dashboard parameters, darbepoetin
55 ug Q2W raises reference-patient Hb by 1.01 g/dL against the 2.6 g/dL the
Supplement reports, and daprodustat 4 mg QD gives 1.93 against 1.4. Both
discrepancies originate in the drug-specific PK sub-models rather than in the
shared erythropoiesis core: simulated darbepoetin Cmax after 60 ug SC is
2.45 ng/mL at 20 h against the roughly 5-8 ng/mL at 48-72 h that Kim 2019
reports, and simulated daprodustat Cmax after 4 mg is 63 ng/mL against the
100-300 ng/mL range of Yamada 2020. Since the same core reproduces rHuEPO and
vadadustat to within 5%, the most likely explanation is an incompletely
transcribed darbepoetin/daprodustat PK parameter set in the dashboard. **No
parameter was tuned to close these gaps**; the dashboard values are shipped as
published and the discrepancy is reported here.

**No IIV or residual error.** The paper reports neither, so the model file
contains neither. Virtual populations in the paper are built by log-normal
resampling of the sensitive parameters listed in Supplementary Table S2 around
a reference patient, followed by filtering on baseline Hb; that workflow is a
property of the authors' MATLAB scripts, not of the model, and is not
reproduced here.

**Adaptive dose titration not implemented.** The paper's population results
(Figures 5-9) depend on per-patient adaptive dose-titration algorithms applied
during simulation. Those are trial-protocol logic rather than model structure
and are outside the scope of a model file; the fixed-dose reference-patient
responses validated above are the model-level equivalent.
