---
title: "Letermovir (Royston 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Letermovir (Royston 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Royston_2025_letermovir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Royston_2025_letermovir"))$description`
* Article: <https://doi.org/10.1128/aac.00697-25>
* Upstream phase III model (source of the fixed Ka, Tlag and their IIV):
  Prohn et al. (2021) <https://doi.org/10.1002/psp4.12593>

Royston 2025 is a short-form post hoc pharmacometric analysis of a prospective
therapeutic drug monitoring study in allogeneic hematopoietic cell transplant
(HCT) recipients receiving oral letermovir as primary cytomegalovirus
prophylaxis. The paper makes two claims: (1) the industry-sponsored phase III
model of Prohn 2021 substantially over-predicts real-world letermovir
concentrations, and (2) a simplified one-compartment model refitted to the
real-world data describes them much better (AIC 4,830 -> 3,739).

This vignette packages and validates the **adjusted real-world model** -- the
model the authors themselves fitted, reported in the right-hand column of
Table 1. The phase III comparator model is *not* packaged here; see
[Assumptions and deviations](#assumptions-and-deviations) for why.

# Population

```{r population}
pop <- rxode2::rxode(readModelDb("Royston_2025_letermovir"))$population
str(pop)
```

Forty adult CMV-seropositive allogeneic HCT recipients received primary CMV
prophylaxis with oral letermovir between 1 March 2020 and 20 April 2021 at the
University Hospitals of Geneva, dosed once daily at 480 mg, or at 240 mg when
coadministered with cyclosporine (6 of the 40 patients). The analysis pooled
296 plasma concentrations: 217 trough samples drawn 24 +/- 2 h post-dose and 79
non-trough samples.

This short-form publication does not tabulate baseline demographics; age,
weight and sex distributions are reported in the parent study (Royston 2022,
<https://doi.org/10.1128/aac.00657-22>). Because the final model in Table 1
carries no demographic covariates, the absence of those distributions does not
affect the simulations below.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Royston_2025_letermovir.R` carries an in-file
comment naming its origin. They are collected here for review. All values come
from the **"Adjusted model estimate from one-compartmental model fitted to
real-world data"** column of Royston 2025 Table 1 unless noted.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (fixed) | `log(0.15)` 1/h | Table 1, Ka row: "0.15 (fixed)" -- fixed from the Prohn 2021 phase III model |
| `ltlag` (fixed) | `log(0.674)` h | Table 1, Tlag row: "0.674 (fixed)" -- fixed from Prohn 2021 |
| `lcl` | `log(26.5)` L/h | Table 1, CL/F row: 26.5 (RSE 12.6%) |
| `lvc` | `log(115.2)` L | Table 1, V/F row: 115.2 (RSE 12%) |
| `e_csa_cl` | `-0.38` | Table 1, fCSA row: -0.38 (RSE 69.5%) |
| `etalka` (fixed) | `0.72^2` | Table 1, Ka row `[0.72 (fixed)]`; footnote a: SD of log-transformed parameter |
| `etalvc` (fixed) | `0.23^2` | Table 1, V/F row `[0.23 (fixed)]`; footnote a |
| `etalcl` | `0.58^2` | Table 1, CL/F row `[IIV 0.58 (RSE 17%)]`; footnote a |
| `etaiov_cl_1..8` | `0.62^2` | Table 1, CL/F row `[IOV 0.62 (RSE 7%)]`; footnote a |
| `propSd` | `0.29` | Table 1, "Residual error (%/ug/L)" row: "29%/-" (proportional only, no additive term) |
| `d/dt(depot)`, `d/dt(central)`, `alag(depot)` | n/a | Results: "simplified one-compartment with fixed absorption rate and lag time"; Discussion: "Simplifications to an oral one-compartment model appear justified given the lack of bi-phasic decline during 24-h dosing intervals (Fig. 1)" |
| `cl <- ... * exp(e_csa_cl * CONMED_CSA)` | n/a | Results: CL/F "reduced by 32% with cyclosporin"; see the falsifier note below |

## Why the cyclosporine effect is exponential, not linear

Table 1 reports `fCSA = -0.38`, but the Results text states that CL/F is
*"reduced by 32% with cyclosporin"*. Those two numbers pin the functional form
exactly:

```{r csa-form}
c(exponential = 1 - exp(-0.38), linear = 0.38)
```

`1 - exp(-0.38)` is 31.6%, which rounds to the quoted 32%; a linear
`1 - 0.38` form would have been quoted as 38%. The multiplicative-on-log-scale
form is therefore the one the authors fitted, and it is also Monolix's standard
parameterization for a categorical covariate on a log-normally distributed
parameter. (The Discussion separately restates the raw coefficient as
"-38%", which is the coefficient value, not the back-transformed effect.)

# Virtual cohort

Original observed data are not publicly available. Two arms of 200 virtual
subjects each reproduce the study's two regimens. The final model carries no
demographic covariates, so the only subject-level inputs are the cyclosporine
indicator and the occasion index.

Each subject receives 8 consecutive once-daily doses, one per occasion
(`OCC = 1..8`), matching the paper's treatment of each patient visit as a
separate occasion. A trough sample is drawn at the end of every dosing
interval, and the final interval is additionally sampled densely for NCA.

Note that the sample at `time = 168 h` serves twice over: it is occasion 7's
end-of-interval trough *and*, because the 0.674 h absorption lag leaves the
concentration continuous across a dose, it is also occasion 8's time-zero
concentration. Everything downstream is therefore keyed off `time` rather than
off a recomputed occasion index, which would have to classify that row as one
or the other.

```{r cohort}
set.seed(20250822)

n_per_arm <- 200L
n_occ <- 8L
tau <- 24
t_last <- tau * (n_occ - 1L)  # dose time opening the final interval

make_arm <- function(n, dose, csa, label, id_offset = 0L) {
  ids <- id_offset + seq_len(n)

  doses <- tidyr::expand_grid(id = ids, time = tau * (seq_len(n_occ) - 1L)) |>
    dplyr::mutate(
      amt = dose, evid = 1L, cmt = "depot",
      OCC = as.integer(time / tau) + 1L
    )

  # Trough samples: end of every dosing interval. The sample ending occasion k
  # coincides with the dose opening occasion k+1; sorting evid = 0 ahead of
  # evid = 1 makes it a pre-dose sample, which is what the paper drew. The
  # covariate OCC on that row is k, so it is eliminated under occasion k's CL/F.
  troughs <- tidyr::expand_grid(id = ids, time = tau * seq_len(n_occ)) |>
    dplyr::mutate(OCC = as.integer(time / tau))

  # Dense grid over the final dosing interval only, for NCA and the profile.
  dense <- tidyr::expand_grid(id = ids, time = seq(t_last, tau * n_occ, by = 0.5)) |>
    dplyr::mutate(OCC = n_occ)

  obs <- dplyr::bind_rows(troughs, dense) |>
    dplyr::distinct(id, time, .keep_all = TRUE) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")

  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(CONMED_CSA = csa, treatment = label, dose_mg = dose) |>
    dplyr::arrange(id, time, evid)
}

events <- dplyr::bind_rows(
  make_arm(n_per_arm, 480, 0L, "480 mg QD", id_offset = 0L),
  make_arm(n_per_arm, 240, 1L, "240 mg QD + CSA", id_offset = n_per_arm)
)

stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
stopifnot(length(unique(events$id)) == 2L * n_per_arm)
```

# Simulation

```{r simulate}
mod <- readModelDb("Royston_2025_letermovir")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("treatment", "CONMED_CSA", "OCC", "dose_mg"),
  addDosing = FALSE
) |>
  as.data.frame() |>
  dplyr::mutate(
    is_trough = time > 0 & abs(time %% tau) < 1e-8,
    in_last = time >= t_last,
    tad_last = time - t_last
  )

stopifnot(nrow(sim) > 0, !anyNA(sim$Cc), all(sim$Cc >= 0))
# Every subject must contribute one trough per occasion.
stopifnot(all(table(sim$id[sim$is_trough]) == n_occ))
```

`Cc` is the individual prediction and carries no residual error; the `sim`
column adds the 29% proportional residual error. Comparisons against
*observed* concentrations below use `sim`, because the published troughs are
measured values that include assay error. Structural checks use `Cc`.

```{r sim-column-check}
stopifnot("sim" %in% names(sim))
```

A typical-value companion simulation, with every random effect zeroed, is used
below for the deterministic checks.

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(rxode2::rxode(mod))

sim_typ <- rxode2::rxSolve(
  mod_typical,
  events |> dplyr::filter(id %in% c(1L, 1L + n_per_arm)),
  keep = c("treatment", "dose_mg"), addDosing = FALSE
) |>
  as.data.frame() |>
  dplyr::mutate(is_trough = time > 0 & abs(time %% tau) < 1e-8,
                in_last = time >= t_last,
                tad_last = time - t_last)
```

# Steady state

At the typical value the model is flip-flop: `ka` (0.15 1/h, half-life 4.6 h)
is slower than `kel` (CL/F divided by V/F, 0.23 1/h), so the terminal phase is
absorption-limited with a 4.6 h half-life against a 24 h dosing interval.
Accumulation is correspondingly slight, which is what makes the paper's
treatment of "each patient visit as an occasion at steady state" reasonable.

```{r accumulation-typical}
typ_accum <- sim_typ |>
  dplyr::filter(is_trough) |>
  dplyr::mutate(occ = as.integer(time / tau)) |>
  dplyr::select(treatment, occ, Cc) |>
  dplyr::mutate(trough_ng_mL = 1000 * Cc) |>
  dplyr::select(-Cc) |>
  tidyr::pivot_wider(names_from = occ, values_from = trough_ng_mL)

knitr::kable(typ_accum, digits = 1,
             caption = "Typical-value trough by occasion (ng/mL).")

# Accumulation, and confirmation that the plateau is genuinely reached.
typ_ratio <- sim_typ |>
  dplyr::filter(is_trough) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    accumulation_occ8_over_occ1 = Cc[time == tau * n_occ] / Cc[time == tau],
    plateau_occ8_over_occ4 = Cc[time == tau * n_occ] / Cc[time == tau * 4],
    .groups = "drop"
  )
knitr::kable(typ_ratio, digits = 5,
             caption = "Typical-value accumulation and plateau ratios.")

# A single-exponential approximation using the absorption-limited terminal rate.
# It under-states the 240 mg + CSA arm because there kel (0.157) sits very close
# to ka (0.15), so the second exponential still contributes at trough.
c(single_exponential_approximation = 1 / (1 - exp(-0.15 * tau)))

# Total accumulation over eight occasions is under 6% in both arms ...
stopifnot(all(typ_ratio$accumulation_occ8_over_occ1 > 1.02),
          all(typ_ratio$accumulation_occ8_over_occ1 < 1.06))
# ... and the trough is flat to within 0.1% from occasion 4 onward, so the
# final interval is unambiguously at steady state.
stopifnot(all(abs(typ_ratio$plateau_occ8_over_occ4 - 1) < 1e-3))
```

That tidy picture holds only near the centre of the population. IIV on CL/F and
V/F, plus an inter-occasion SD of 0.62 on CL/F, spread `kel` widely, and
subjects in the low-clearance tail have `kel < ka` and accumulate materially
across occasions. The paper's per-occasion steady-state approximation is
therefore good for a typical patient and progressively weaker in that tail --
worth knowing before reusing this model for an accumulation question.

```{r halflife-spread}
kel_dist <- sim |>
  dplyr::filter(in_last, OCC == n_occ) |>
  dplyr::group_by(id) |>
  dplyr::summarise(kel = dplyr::first(kel), .groups = "drop")

round(quantile(log(2) / kel_dist$kel, c(0, 0.05, 0.5, 0.95, 1)), 2)
```

# Replicate published figures

```{r figure-1, fig.width = 7, fig.height = 4}
# Replicates Figure 1 of Royston 2025: letermovir concentrations on a
# time-after-dose scale, by regimen. Royston 2025 plots the real-world
# observations against the phase III model's expected profile; here the ribbon
# is the adjusted (real-world) model's own prediction interval, and the dashed
# line marks the published observed median trough for each regimen.
published_trough <- tibble::tribble(
  ~treatment,         ~median, ~q1,  ~q3,
  "480 mg QD",         260,     123,  518,
  "240 mg QD + CSA",   418,     235,  945
)

sim |>
  dplyr::filter(in_last) |>
  dplyr::group_by(treatment, tad_last) |>
  dplyr::summarise(
    Q05 = 1000 * quantile(sim, 0.05),
    Q50 = 1000 * quantile(sim, 0.50),
    Q95 = 1000 * quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad_last, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  geom_hline(
    data = published_trough,
    aes(yintercept = median), linetype = "dashed", colour = "firebrick"
  ) +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Letermovir concentration (ng/mL)",
    title = "Steady-state profile by regimen",
    caption = paste(
      "Adjusted-model prediction interval (5th-50th-95th percentile, with",
      "residual error). Dashed line: published observed median trough",
      "(Royston 2025 Results). Compare Figure 1 of Royston 2025."
    )
  )
```

# PKNCA validation

NCA runs on the final dosing interval, rebased so time 0 is the dose opening
that interval. The `time = 168 h` sample supplies the time-zero concentration
directly, so no zero is imputed.

```{r pknca}
nca_conc <- sim |>
  dplyr::filter(in_last, !is.na(Cc)) |>
  dplyr::select(id, time = tad_last, Cc, treatment) |>
  dplyr::arrange(id, treatment, time)

# A real time-zero record must already be present (steady-state pre-dose
# concentration), not an imputed zero.
stopifnot(all(tapply(nca_conc$time, nca_conc$id, min) == 0))
stopifnot(all(nca_conc$Cc[nca_conc$time == 0] > 0))

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id)

dose_df <- events |>
  dplyr::filter(evid == 1, time == t_last) |>
  dplyr::mutate(time = 0) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

`ctrough` is read directly off the solved profile rather than requested from
PKNCA. The quantity the paper reports is the measured 24 h post-dose
concentration, which includes residual error, and PKNCA's `cmin` is not that
value here -- the absorption lag means the within-interval minimum sits at
`tad = 0.674 h`, appreciably below the end-of-interval concentration.

```{r ctrough}
# Pool troughs across all eight occasions, matching the paper's pooling of 217
# trough samples across visits. Uses `sim` (with residual error) because the
# published values are measured concentrations.
ctrough_obs <- sim |>
  dplyr::filter(is_trough) |>
  dplyr::transmute(id, treatment, PPTESTCD = "ctrough", PPORRES = 1000 * sim)

# Confirm the interior-minimum trap is real for this model, per subject.
cmin_vs_ctrough <- sim |>
  dplyr::filter(in_last) |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(
    cmin = min(Cc),
    ctrough = Cc[tad_last == tau],
    tmin_h = tad_last[which.min(Cc)],
    .groups = "drop"
  ) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cmin_ng_mL = 1000 * median(cmin),
    ctrough_ng_mL = 1000 * median(ctrough),
    pct_below = 100 * median((ctrough - cmin) / ctrough),
    median_tmin_h = median(tmin_h),
    .groups = "drop"
  )

knitr::kable(
  cmin_vs_ctrough, digits = c(0, 1, 1, 1, 3),
  caption = "Interior minimum vs end-of-interval concentration, per subject."
)

# The minimum sits at the end of the absorption lag, not at the end of the
# interval, so substituting cmin would have understated the published trough.
stopifnot(all(cmin_vs_ctrough$pct_below > 5))
stopifnot(all(abs(cmin_vs_ctrough$median_tmin_h - 0.674) < 0.3))
```

## Comparison against published NCA

Royston 2025 reports observed median (IQR) trough concentrations for each
regimen. It reports no observed Cmax, AUC or half-life for the real-world
data, so `ctrough` is the only directly comparable quantity.

```{r nca_compare}
sim_long <- dplyr::bind_rows(
  as.data.frame(nca_res$result) |>
    dplyr::select(treatment, PPTESTCD, PPORRES),
  ctrough_obs |> dplyr::select(treatment, PPTESTCD, PPORRES)
)

published <- tibble::tribble(
  ~treatment,        ~ctrough,
  "480 mg QD",        260,
  "240 mg QD + CSA",  418
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_long,
  reference = published,
  by = "treatment",
  params = "ctrough",
  units = c(ctrough = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs published observed median trough concentration.",
    "* differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r ctrough-in-iqr}
med <- sim |>
  dplyr::filter(is_trough) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    median_ng_mL = 1000 * median(sim),
    q1_ng_mL = 1000 * quantile(sim, 0.25),
    q3_ng_mL = 1000 * quantile(sim, 0.75),
    .groups = "drop"
  )
knitr::kable(med, digits = 1,
             caption = "Simulated pooled trough distribution by regimen (ng/mL).")

# The 480 mg arm's simulated median must fall inside the published IQR
# (123-518 ng/mL).
med_480 <- med$median_ng_mL[med$treatment == "480 mg QD"]
stopifnot(med_480 > 123, med_480 < 518)
```

Both arms simulate below their published median: the 480 mg arm by about a
third (though comfortably inside the published IQR of 123-518 ng/mL), and the
240 mg + cyclosporine arm by roughly 2.3-fold. Neither is tuned away. Two
distinct things are going on, and they are worth separating.

**The level.** The typical-value trough is 206 ng/mL at 480 mg, but the *median
of the simulated population* is lower still, near 178. That gap is not a bug:
the trough is a non-linear function of three random effects, and `etalka` in
particular carries a fixed IIV of 0.72 on the log scale -- a roughly four-fold
spread in absorption rate either side of the median. Since a slower `ka`
flattens the profile and a faster one deepens the trough, the median of the
resulting distribution sits below the trough of the median subject. The same
mechanism widens the simulated IQR well beyond the published one.

**The separation between arms.** The model predicts almost identical troughs
for the two regimens, because halving the dose and cutting CL/F by 32% nearly
cancel. The published data instead show the cyclosporine arm *higher* than the
480 mg arm (418 vs 260 ng/mL). Recovering that ordering requires a much larger
cyclosporine effect than the one fitted:

```{r implied-csa}
# Coefficient implied by the published trough ratio, if trough scales as
# Dose / (CL/F): (240 / (26.5 * exp(b))) / (480 / 26.5) = 418 / 260.
implied_b <- log(0.5 * 260 / 418)
se_b <- 0.695 * 0.38  # RSE 69.5% of the point estimate
c(
  implied = implied_b,
  fitted = -0.38,
  ci_low = -0.38 - 1.96 * se_b,
  ci_high = -0.38 + 1.96 * se_b
)
```

The implied coefficient sits just outside the lower end of the fitted effect's
approximate 95% confidence interval -- which is precisely the imprecision the
authors flag, calling for work to "enhance confidence in the magnitude of
decreased oral clearance with cyclosporin (estimated in this analysis to -38%
with however large RSE)". That arm holds only 6 of the 40 patients, its
published IQR spans a four-fold range (235-945 ng/mL), and the authors
explicitly describe their covariate findings as "hypothesis-generating rather
than as confirmed predictors". The discrepancy is a property of the published
model, faithfully reproduced, not of this implementation.

Note also that the exponential-versus-linear question settled above does not
explain the gap: a linear `1 - 0.38` reading moves the cyclosporine arm up by
only about a fifth, still far short of 418 ng/mL.

# Structural identity checks

## Exact identity on the typical-value ladder

With the random effects zeroed, every occasion shares one CL/F, the profile
reaches true steady state, and `AUC over tau = Dose / (CL/F)` holds exactly.

```{r identity-typical}
typ_check <- sim_typ |>
  dplyr::filter(in_last) |>
  dplyr::group_by(treatment, dose_mg) |>
  dplyr::summarise(
    cl = dplyr::last(cl),
    auc_trapz = sum(diff(tad_last) *
                      (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    auc_expected = dose_mg / cl,
    pct_diff = 100 * (auc_trapz - auc_expected) / auc_expected
  )

knitr::kable(typ_check, digits = 4,
             caption = "Typical-value AUC over tau vs Dose/(CL/F).")

# Only the 0.5 h trapezoidal grid separates the two; the identity is exact.
stopifnot(max(abs(typ_check$pct_diff)) < 0.5)

# The typical CL/F values are exactly the published ones.
stopifnot(
  abs(typ_check$cl[typ_check$treatment == "480 mg QD"] - 26.5) < 1e-6,
  abs(typ_check$cl[typ_check$treatment == "240 mg QD + CSA"] -
        26.5 * exp(-0.38)) < 1e-6
)
```

## Exact per-subject mass balance

`AUC = Dose / (CL/F)` is a *steady-state* identity, and with an inter-occasion
SD of 0.62 on CL/F no individual profile is at a single steady state: the drug
carried into the final interval was eliminated under occasion 7's clearance
while the interval itself runs on occasion 8's. Asserting the steady-state form
per subject would be testing an identity this model does not satisfy -- across
these 400 subjects the carried-over amount ranges from -104% to +67% of the
dose, so the naive form is off by a median of 5.9% and by as much as 104%.

The identity that *does* hold per subject, exactly and without any steady-state
assumption, is mass balance over the interval: everything that goes in either
leaves via clearance or is still in the body at the end.

```
(CL/F) * AUC_tau = Dose - [ (depot + central)_end - (depot + central)_start ]
```

```{r identity-massbalance}
mass_balance <- sim |>
  dplyr::filter(in_last) |>
  dplyr::arrange(id, tad_last) |>
  dplyr::group_by(id, treatment, dose_mg) |>
  dplyr::summarise(
    # The tad_last == 0 row carries OCC = 7 (it is occasion 7's trough), so
    # take the clearance governing this interval from the end of it.
    cl = dplyr::last(cl),
    auc = sum(diff(tad_last) *
                (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    d_state = (dplyr::last(depot) + dplyr::last(central)) -
      (dplyr::first(depot) + dplyr::first(central)),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    auc_expected = (dose_mg - d_state) / cl,
    pct_diff = 100 * (auc - auc_expected) / auc_expected
  )

stopifnot(nrow(mass_balance) == 2L * n_per_arm)
summary(mass_balance$pct_diff)

# Exact up to trapezoidal error on the 0.5 h grid.
stopifnot(median(abs(mass_balance$pct_diff)) < 0.1)
stopifnot(max(abs(mass_balance$pct_diff)) < 2)
```

For contrast, the same subjects scored against the steady-state form the
carry-over invalidates:

```{r identity-naive-contrast}
naive <- mass_balance |>
  dplyr::mutate(pct_naive = 100 * (auc - dose_mg / cl) / (dose_mg / cl))

c(
  massbalance_median_abs_pct = median(abs(naive$pct_diff)),
  naive_median_abs_pct = median(abs(naive$pct_naive)),
  naive_max_abs_pct = max(abs(naive$pct_naive))
)
```

```{r identity-cl-constant}
cl_by_id <- sim |>
  dplyr::filter(in_last, OCC == n_occ) |>
  dplyr::group_by(id, treatment, dose_mg) |>
  dplyr::summarise(cl = dplyr::first(cl), n_cl = dplyr::n_distinct(cl),
                   .groups = "drop")

# CL/F must be constant within the final occasion.
stopifnot(all(cl_by_id$n_cl == 1L))
```

```{r identity-csa}
# The cyclosporine effect multiplies CL/F by exp(-0.38).
cl_ratio <- cl_by_id |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(median_cl = median(cl), .groups = "drop")

observed_ratio <-
  cl_ratio$median_cl[cl_ratio$treatment == "240 mg QD + CSA"] /
  cl_ratio$median_cl[cl_ratio$treatment == "480 mg QD"]

c(observed = observed_ratio, expected = exp(-0.38))

# Both arms draw from the same eta distribution, so the median CL/F ratio
# converges on exp(-0.38); 200 subjects per arm leaves some sampling noise.
stopifnot(abs(observed_ratio - exp(-0.38)) < 0.08)
```

```{r identity-iov}
# IOV is the dominant variance component: the pooled SD of log CL/F must exceed
# the IIV SD of 0.58 and approach sqrt(0.58^2 + 0.62^2) = 0.849.
sd_pooled <- sim |>
  dplyr::filter(is_trough, treatment == "480 mg QD") |>
  dplyr::summarise(sd_log_cl = sd(log(cl))) |>
  dplyr::pull(sd_log_cl)

c(pooled_sd = sd_pooled, expected = sqrt(0.58^2 + 0.62^2), iiv_only = 0.58)
stopifnot(sd_pooled > 0.58, abs(sd_pooled - sqrt(0.58^2 + 0.62^2)) < 0.1)
```

# Assumptions and deviations

* **The phase III comparator model is not packaged.** Royston 2025 Table 1
  reproduces the Prohn 2021 phase III model in its left-hand column, but that
  column is internally inconsistent as printed: it gives `CL = 4.8 L/h`,
  `F = 0.85` and `CL/F = 31.9 L/h (calculated)`, yet `4.8 / 0.85 = 5.6`, not
  31.9. The tabulated `fCSA = -0.258 (calculated)` likewise cannot be recovered
  from the tabulated `CL`, `CL_CSA`, `F` and `F_CSA`: those give a 72% *increase*
  in CL/F with cyclosporine, not the 26% decrease reported. A third, independent
  check disagrees with both: the paper's own phase III simulations predict a
  median trough of 480 ng/mL at 480 mg and 1,104 ng/mL at 240 mg with
  cyclosporine, i.e. a 4.6-fold *higher* dose-normalised exposure with
  cyclosporine, which implies a log coefficient near -1.5 rather than -0.258.
  The three readings of the same column cannot be reconciled, so no single
  self-consistent parameter set can be recovered from it. Table 1's heading
  further restricts that column to "non-asian subjects", so the phase III
  model carries at least one covariate that the table does not tabulate. A
  runnable phase III model must therefore be extracted from the primary,
  Prohn et al. (2021) <https://doi.org/10.1002/psp4.12593>, not from this
  table. No phase III parameters were used here except `Ka`, `Tlag` and their
  IIV, which Royston 2025 fixed into its own model, which the adjusted-model
  column reports independently, and which are unaffected by the inconsistency.
* **Number of occasions.** The paper treats "each patient visit as an occasion
  at steady state" but does not report how many visits each patient had. The
  model file implements eight occasions (`OCC = 1..8`), chosen to cover the
  observed sampling density of 296 observations across 40 patients (~7.4 per
  patient). This is an implementation cap, not a paper value; extending it
  means adding further `etaiov_cl_<n>` entries. All occasions share the single
  published IOV variance, so the cap affects only how many distinct occasions
  can be simulated, never the variance itself.
* **Screened covariates are documented, not implemented.** Royston 2025 reports
  univariable associations between CL/F and body weight, age, serum albumin,
  vomiting, acute GvHD, prednisone and posaconazole use, but tabulates no
  regression coefficient for any of them, and none entered the final model in
  Table 1. Those with an existing canonical column name are recorded in the
  model file's `covariatesDataExcluded` list so the provenance of the covariate
  screen is preserved without implying an implementable effect; the remainder
  (vomiting, nausea, diarrhea, acute GvHD, posaconazole and pantoprazole use,
  days since letermovir start, infectious complications, and measured
  cyclosporin concentration) are named in a comment above that list rather than
  given entries, so that no new canonical column is minted for an effect that
  cannot be implemented. The paper itself describes these associations as
  "hypothesis-generating rather than as confirmed predictors".
* **The cyclosporine effect applies to CL/F only.** No cyclosporine effect on
  V/F is reported, so V/F is shared between arms.
* **Baseline demographics are not reproduced.** The short-form paper does not
  tabulate age, weight or sex. Since the final model carries no demographic
  covariates, the virtual cohort needs none.
* **Supplementary material was not available.** Fig. S1
  (`AAC00697-25-S0001.docx`) holds residual diagnostics for both models. Its
  caption confirms it contains goodness-of-fit plots only, no parameter values,
  so its absence does not affect the extraction.
* **Concentration units.** The model works in mg and L, so `Cc` is in ug/mL.
  The paper reports concentrations in ng/mL; this vignette multiplies by 1000
  wherever it compares against published values.
