---
title: "L-glutamine (Sadaf 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{L-glutamine (Sadaf 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Sadaf_2024_glutamine"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Sadaf_2024_glutamine"))$description`
* Article: <https://doi.org/10.1007/s40262-024-01349-4>
* Trial registration: [NCT04684381](https://clinicaltrials.gov/study/NCT04684381)

L-glutamine (Endari) was approved by the FDA in 2017 to reduce acute
complications of sickle cell disease, but the approved 0.3 g/kg twice-daily
regimen was chosen empirically. Sadaf 2024 is the first population PK analysis
of L-glutamine in sickle cell disease and the first PK study of multiple-dose,
long-term oral therapy.

The distinguishing feature of this model is that L-glutamine is **endogenous**:
it is the most abundant free amino acid in plasma and every participant already
carries several hundred umol/L of it before the first dose. The authors handled
this by subtracting each participant's own pre-dose baseline from every
measurement and fitting the *increment* attributable to the dose. The packaged
model therefore predicts the exogenous increment, and `Cc` is zero before the
first dose. To recover a measured plasma concentration, add the subject's
`GLN_BL` back on.

# Population

Twelve participants completed the analysis population of a single-centre,
open-label, dose-ascending phase IV trial run at Cincinnati Children's Hospital
Medical Center between January and June 2021, contributing 400 plasma
L-glutamine concentrations. They were allocated to three groups of four:
paediatric sickle cell disease (age < 18 years), adult sickle cell disease, and
healthy adult volunteers. Overall the cohort spanned 6.8-43.8 years and
30.3-119.1 kg, and two thirds were female. Among the eight participants with
sickle cell disease, genotypes were evenly split (four homozygous HbSS, four
HbSC) as was hydroxyurea use. Healthy volunteers had normal haemoglobin
profiles and blood counts. Fifteen participants were enrolled; the three who
did not complete at least the third study visit were withdrawn and excluded.

Each participant received three ascending dose levels over three consecutive
weeks: 0.1 g/kg twice daily, 0.3 g/kg twice daily, and 0.6 g/kg once daily,
dissolved in 120 mL of water. Sampling was pre-dose and at 30, 60, 120, 180,
and 240 minutes after the morning dose at each of four study visits. Doses were
given either fasting or after a standardised meal so that a food effect could
be tested.

Baseline demographics are Sadaf 2024 Table 1, reported per allocation group:

```{r population-table}
tibble::tribble(
  ~Group,                ~N, ~`Age (years)`,      ~`Weight (kg)`,        ~`Baseline L-glutamine (umol/L)`,
  "Paediatric SCD",      4L, "12.0 (6.8-15.1)",   "64.3 (30.3-101.5)",   "571 (406-720)",
  "Adult SCD",           4L, "23.5 (20.4-29.3)",  "63.45 (49.3-72.8)",   "625 (491-758)",
  "Healthy volunteers",  4L, "37.3 (29.6-43.8)",  "105.1 (67.5-119.1)",  "531 (364-756)"
) |>
  knitr::kable(caption = "Sadaf 2024 Table 1: mean (min-max) by allocation group.")
```

The same information is available programmatically via
`readModelDb("Sadaf_2024_glutamine")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Sadaf_2024_glutamine.R`. The table below
collects them in one place for review. Every value comes from Sadaf 2024
Table 2 or from the numbered equations in Methods 2.6; nothing was digitised
from a figure, inferred from a related publication, or supplied by
correspondence.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | `log(0.91)` 1/h | Table 2, `KA` = 0.91 (RSE 6.4%; bootstrap 0.78-0.99) |
| `lcl` | `log(78.5)` L/h per 70 kg | Table 2, `theta1` = 78.5 (RSE 7.6%; bootstrap 67.1-91.2) |
| `lvc` | `log(52.3)` L per 70 kg | Table 2, `theta3` = 52.3 (RSE 19.5%; bootstrap 30.9-69.9) |
| `e_wt_cl` | `fixed(0.75)` | Eq. 3 ("power is the coefficient set at 0.75 for CL/F"); printed as a literal exponent in the Table 2 `CLi` equation |
| `e_wt_vc` | `fixed(1)` | Eq. 3 ("... and 1 for V/F"); printed as a literal exponent in the Table 2 `Vi` equation |
| `e_gln_bl_cl` | `-0.96` | Table 2, `theta2` (RSE 19.8%; bootstrap -1.3 to -0.4) |
| `e_dose_gln_gkg_vc` | `0.27` | Table 2, `theta4` (RSE 45.0%; bootstrap 0.09-0.60) |
| `etalcl` | `fixed(0)` | Table 2, "Inter-patient variability (variance)", IIV CL = "Fixed to 0" |
| `etalvc` | `0.078` (variance) | Table 2, IIV V (RSE 19.7%; bootstrap 0.004-0.226) |
| `etalka` | `fixed(0)` | Table 2, IIV Ka = "Fixed to 0" |
| `etaiov_cl_1..4` | `0.081` (variance) | Table 2, "Inter-occasion variability (variance)", IOV CL (RSE 36.4%; bootstrap 0.028-0.144); four occasions per Methods 2.6.1 |
| `propSd` | `0.7` = `sqrt(0.49)` | Table 2, "Residual variability (variance)", Prop.Err. = 0.49 (RSE 10.9%; bootstrap 0.39-0.59) |
| additive residual | omitted | Table 2, Add.Err. = "Fixed to 0"; Eq. 2 as printed is `Y = Cpred * (1 + eps_prop)` |
| `GLN_BL` reference 683 umol/L | n/a | Table 2 footnote: "Glu_BSLstardard standard glutamine baseline concentration (683 umol/L)" |
| `DOSE_GLN_GKG` reference 0.1 g/kg | n/a | Table 2 `Vi` equation, `(DOSE per kg/0.1)` |
| Exponential IIV, `Pi = Ppop * exp(eta_i)` | n/a | Eq. 1 |
| Power covariate form, `Pi = Ppop * (COVcon/COVmedian)^theta` | n/a | Eq. 5 |
| One-compartment, first-order absorption and disposition | n/a | Results 3.3, "a one-compartment model with first-order input and disposition was found to outperform other base models examined" |
| Baseline subtraction of endogenous L-glutamine | n/a | Methods 2.6.1, "the individual baseline l-glutamine concentration was subtracted from each l-glutamine measurement" |

Three of the reported quantities are **variances**, not standard deviations or
CV%: Table 2 heads the three random-effect blocks "Inter-patient variability
(variance)", "Inter-occasion variability (variance)", and "Residual variability
(variance)". The IIV and IOV entries are therefore carried into `ini()`
unchanged (rxode2 `~` declarations take variances), while `propSd` is a
standard deviation and is entered as `sqrt(0.49) = 0.7`. Reading 0.49 as an SD
instead would understate the residual error by 30%, and reading the IIV
variances as CV% would overstate them by an order of magnitude, so this is the
single most consequential transcription decision in the file.

# Virtual cohort

Original observed data are not publicly available. The cohort below draws
weights and pre-dose L-glutamine baselines from the three allocation groups of
Table 1, in the 4:4:4 proportion of the study, using truncated normals whose
means match the reported group means and whose spread is set so that the
reported min-max spans roughly four standard deviations.

Every virtual participant receives all three dose levels, mirroring the
crossover-like ascending-dose design. Arms are given disjoint ID ranges because
`rxSolve` treats `id` as the subject key.

```{r cohort}
set.seed(20240224)

n_per_arm <- 100L
mw_gln    <- 146.14  # g/mol, L-glutamine

rtnorm <- function(n, mean, lo, hi) {
  pmin(pmax(stats::rnorm(n, mean, (hi - lo) / 4), lo), hi)
}

# Per-subject covariates, drawn once and reused across the three dose arms so
# the design is paired, as in the trial.
group_levels <- c("Paediatric SCD", "Adult SCD", "Healthy volunteer")
subjects <- tibble::tibble(
  subject = seq_len(n_per_arm),
  group   = sample(group_levels, n_per_arm, replace = TRUE)
) |>
  mutate(
    WT = case_when(
      group == "Paediatric SCD"    ~ rtnorm(n(),  64.30,  30.3, 101.5),
      group == "Adult SCD"         ~ rtnorm(n(),  63.45,  49.3,  72.8),
      group == "Healthy volunteer" ~ rtnorm(n(), 105.10,  67.5, 119.1)
    ),
    GLN_BL = case_when(
      group == "Paediatric SCD"    ~ rtnorm(n(), 571, 406, 720),
      group == "Adult SCD"         ~ rtnorm(n(), 625, 491, 758),
      group == "Healthy volunteer" ~ rtnorm(n(), 531, 364, 756)
    )
  )

# Observation grid: the paper's nominal sampling times (0, 30, 60, 120, 180,
# 240 min) plus a dense grid so the concentration-time figure is smooth.
obs_dense    <- sort(unique(c(seq(0, 6, by = 0.05))))
obs_nominal  <- c(0, 0.5, 1, 2, 3, 4)

make_arm <- function(gkg, occ, id_offset) {
  s <- subjects |>
    mutate(
      id           = id_offset + subject,
      DOSE_GLN_GKG = gkg,
      OCC          = occ,
      treatment    = paste0(format(gkg, nsmall = 1), " g/kg"),
      # Dose amount in umol so that central (umol) / vc (L) is umol/L.
      amt_umol     = gkg * WT / mw_gln * 1e6
    )
  doses <- s |>
    mutate(time = 0, amt = amt_umol, evid = 1L, cmt = "depot")
  obs <- s |>
    tidyr::crossing(time = obs_dense) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |>
    select(id, subject, group, time, amt, evid, cmt,
           WT, GLN_BL, DOSE_GLN_GKG, OCC, treatment) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(0.1, 1L, id_offset =              0L),
  make_arm(0.3, 2L, id_offset =     n_per_arm),
  make_arm(0.6, 3L, id_offset = 2L * n_per_arm)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(dplyr::n_distinct(events$id) == 3L * n_per_arm)
```

The `OCC` column multiplexes the four inter-occasion etas on log-CL/F. Because
Sadaf 2024 reports a single IOV variance shared by all four occasions, the
occasion indices are exchangeable: assigning arms to occasions 1, 2, 3 rather
than to some other permutation changes nothing about the predicted
distribution.

# Simulation

```{r simulate}
mod <- readModelDb("Sadaf_2024_glutamine")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "group", "WT", "GLN_BL")
) |>
  as.data.frame()

# rxSolve has been observed to silently drop subjects; assert the count.
stopifnot(dplyr::n_distinct(sim$id) == 3L * n_per_arm)
```

For the typical-value predictions used to replicate Figure 3, the random
effects are zeroed:

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(mod)

events_typical <- events |>
  filter(subject == 1L) |>
  mutate(WT = 70, GLN_BL = 683, amt = ifelse(evid == 1L, DOSE_GLN_GKG * 70 / mw_gln * 1e6, amt))

sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events_typical,
  keep   = c("treatment"),
  omega  = NA
) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2 - concentration-time profiles

Sadaf 2024 Figure 2 plots mean (+/- SD) *absolute* plasma L-glutamine against
time after dose, with each participant's visit-1 baseline drawn as a dotted red
reference line, to show that concentrations return to baseline between doses.
Because the packaged model predicts the baseline-subtracted increment, the
absolute concentration is recovered by adding `GLN_BL` back on.

```{r figure-2}
# Replicates Figure 2 of Sadaf 2024: absolute plasma L-glutamine vs. time.
sim |>
  filter(!is.na(Cc)) |>
  mutate(total = Cc + GLN_BL) |>
  group_by(treatment, time) |>
  summarise(
    mean_total = mean(total),
    lo         = mean(total) - stats::sd(total),
    hi         = mean(total) + stats::sd(total),
    .groups    = "drop"
  ) |>
  ggplot(aes(time, mean_total)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = mean(subjects$GLN_BL), linetype = "dotted",
             colour = "red", linewidth = 0.6) +
  facet_wrap(~treatment) +
  labs(
    x = "Time after morning dose (h)",
    y = "Plasma L-glutamine (umol/L)",
    title = "Figure 2 - absolute plasma L-glutamine by dose level",
    caption = paste(
      "Replicates Figure 2 of Sadaf 2024. Mean +/- SD of the simulated cohort.",
      "Dotted red line is the cohort mean pre-dose baseline."
    )
  )
```

Sadaf 2024 Results 3.2 makes a specific claim about these profiles: "At the
lowest dose (0.1 g/kg), l-glutamine concentrations returned to baseline
concentrations within 4 hours. However, with 0.3 g/kg or 0.6 g/kg dosing, the
plasma l-glutamine concentration remained higher than baseline at 4 h." That is
checkable rather than merely visual, so it is checked:

```{r return-to-baseline}
ret <- sim_typical |>
  filter(!is.na(Cc)) |>
  group_by(treatment) |>
  summarise(
    `Cmax (umol/L)`            = max(Cc),
    `Cc at 4 h (umol/L)`       = Cc[which.min(abs(time - 4))],
    .groups = "drop"
  ) |>
  mutate(`Percent of baseline remaining at 4 h` =
           round(100 * `Cc at 4 h (umol/L)` / 683, 1))

pct <- setNames(ret$`Percent of baseline remaining at 4 h`, ret$treatment)

# Pin the typical-value predictions exactly, so a regression in the model file
# or in rxode2 fails here rather than silently changing the figures.
stopifnot(
  isTRUE(all.equal(unname(pct[c("0.1 g/kg", "0.3 g/kg", "0.6 g/kg")]),
                   c(4.9, 19.5, 46.0), tolerance = 0.01))
)

# The paper's qualitative claim, stated separately from the exact values: at
# 0.1 g/kg the increment at 4 h is a few percent of the endogenous baseline
# (indistinguishable from it on a plot with SD bands), while at 0.3 and
# 0.6 g/kg it is still a substantial and monotonically growing fraction of it.
stopifnot(
  pct[["0.1 g/kg"]] < 10,
  pct[["0.3 g/kg"]] > 15,
  pct[["0.6 g/kg"]] > 40,
  !is.unsorted(pct[c("0.1 g/kg", "0.3 g/kg", "0.6 g/kg")], strictly = TRUE)
)

ret |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  dplyr::rename("Dose level" = treatment) |>
  knitr::kable(
    caption = paste(
      "Typical-value return to baseline at 4 h. At 0.1 g/kg the remaining",
      "increment is a few percent of the 683 umol/L endogenous baseline;",
      "at 0.3 and 0.6 g/kg it is still a large fraction of it, as",
      "Sadaf 2024 Results 3.2 describes."
    )
  )
```

## Figure 3 - dose-normalised Cmax falls with dose escalation

This is the paper's central non-linearity finding, and the reason the model
carries dose as a covariate on apparent volume at all.

```{r figure-3}
# Replicates Figure 3 of Sadaf 2024: dose-normalised Cmax vs. dose level.
cmax_norm <- sim |>
  filter(!is.na(Cc)) |>
  group_by(id, treatment, DOSE_GLN_GKG) |>
  summarise(cmax = max(Cc), .groups = "drop") |>
  mutate(cmax_norm = cmax / (DOSE_GLN_GKG / 0.1))

cmax_norm |>
  ggplot(aes(treatment, cmax_norm)) +
  geom_boxplot(outlier.alpha = 0.3) +
  labs(
    x = "Dose level",
    y = "Cmax normalised to the 0.1 g/kg dose (umol/L)",
    title = "Figure 3 - dose-normalised Cmax decreases with dose escalation",
    caption = "Replicates Figure 3 of Sadaf 2024."
  )
```

```{r figure-3-typical}
sim_typical |>
  filter(!is.na(Cc)) |>
  group_by(treatment) |>
  summarise(
    `Tmax (h)`                 = time[which.max(Cc)],
    `Cmax (umol/L)`            = max(Cc),
    .groups = "drop"
  ) |>
  mutate(
    `Dose multiple of 0.1 g/kg`   = c(1, 3, 6),
    `Dose-normalised Cmax`        = round(`Cmax (umol/L)` / `Dose multiple of 0.1 g/kg`, 1),
    `Cmax (umol/L)`               = round(`Cmax (umol/L)`, 1)
  ) |>
  dplyr::rename("Dose level" = treatment) |>
  knitr::kable(
    caption = paste(
      "Typical-value (70 kg, baseline 683 umol/L) predictions.",
      "Dose-normalised Cmax falls and Tmax rises across the dose range,",
      "reproducing both non-linearity findings of Sadaf 2024."
    )
  )
```

Two published findings are reproduced here at once. Dose-normalised Cmax falls
monotonically across the three dose levels (Figure 3 and Results 3.2), and Tmax
rises with dose - Sadaf 2024 reports "the Tmax was higher at the dose of
0.6 g/kg (p < 0.05), indicating slower absorption at higher doses". Both fall
out of the single `(DOSE per kg / 0.1)^0.27` term on apparent volume: inflating
V/F lowers kel = CL/V, which pushes the model further into flip-flop kinetics
(ka = 0.91 1/h against kel = 1.50, 1.12, and 0.93 1/h at the three dose levels)
and so both depresses and delays the peak. No separate absorption-rate covariate
is needed.

# PKNCA validation

The NCA below uses the paper's own nominal sampling schedule (pre-dose and 30,
60, 120, 180, 240 minutes) rather than the dense grid, so that the comparison
against the published NCA is made on comparable data. Like the paper, the NCA
is run on the baseline-subtracted concentrations: Sadaf 2024 Methods 2.5 states
that "Cmax values were obtained by subtracting baseline concentrations from
peak values".

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc), time %in% obs_nominal) |>
  select(id, time, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment); pre-dose Cc = 0 is correct for
# this baseline-subtracted, extravascular model.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

dose_df <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "umol/L", timeu = "hr")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "umol")

intervals <- data.frame(
  start      = 0,
  end        = 4,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-summary}
summary(nca_res)
```

## Comparison against published NCA

Sadaf 2024 reports only two NCA parameters numerically, and reports both pooled
across dose levels and visits rather than per dose group: Tmax averaged
1.2 +/- 0.9 h (Results 3.2) and the apparent terminal half-life averaged
1.0 +/- 0.7 h, computed only "from visits when samples were rich enough".
Cmax appears in Figure 3 in dose-normalised form only, with no axis values
transcribable as a point estimate, so it cannot enter a numeric comparison.
The published values are therefore repeated across the three simulated arms
below; the per-arm spread of the simulated values is the informative part.

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,  ~tmax, ~half.life,
  "0.1 g/kg",  1.2,   1.0,
  "0.3 g/kg",  1.2,   1.0,
  "0.6 g/kg",  1.2,   1.0
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. published NCA. Published values are the pooled",
    "cohort means of Sadaf 2024 Results 3.2, repeated for each arm.",
    "* differs from reference by >20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_footnote, results = "asis", echo = FALSE}
fn <- attr(cmp, "footnote")
if (!is.null(fn)) cat("\n", fn, "\n", sep = "")
```

The simulated half-lives (0.84, 0.93, and 1.07 h) bracket the published pooled
mean of 1.0 h, and none of the three differs from it by more than 20%. This is
worth a note, because a naive reading of the model would not predict it: with
CL/F = 78.5 L/h and V/F = 52.3 L the elimination rate constant at the reference
dose is 1.50 1/h, implying a *disposition* half-life of only 0.46 h. But
ka = 0.91 1/h is smaller than kel, so the model sits in flip-flop kinetics and
the observable terminal slope is governed by absorption rather than
elimination, with an asymptotic value of log(2)/0.91 = 0.76 h. The values above
run a little longer than that because the paper's 0-4 h window ends before the
profile is purely terminal, and they lengthen with dose because inflating V/F
lowers kel toward ka, which flattens the observed decline. That a
1-hour-half-life observation falls out of a model whose disposition half-life
is 28 minutes is a meaningful check that absorption and disposition were
transcribed into the right roles, and not swapped.

Simulated Tmax runs slightly below the published pooled mean of 1.2 h. Two
reasons, both structural rather than a transcription problem. First, the
published mean pools all four visits including the 0.6 g/kg visit, whose Tmax
the paper explicitly reports as higher; the simulated per-arm values bracket
the pooled mean in the same direction. Second, Tmax on the paper's sparse
nominal grid can only take the values 0.5, 1, 2, 3, or 4 h, so a published
mean of 1.2 h with an SD of 0.9 h reflects a handful of subjects whose observed
peak landed at 2 h or later. No parameter was adjusted to close the gap.

# Assumptions and deviations

- **The model predicts the baseline-subtracted increment, not the measured
  concentration.** Sadaf 2024 Methods 2.6.1 subtracted each participant's
  pre-dose L-glutamine from every measurement before fitting, assuming the
  endogenous level is constant. `Cc` is therefore zero before the first dose,
  and a measured plasma concentration is `Cc + GLN_BL`. Users simulating this
  model to predict laboratory values must add the baseline back on, as the
  Figure 2 chunk above does.
- **`GLN_BL` does double duty.** The same per-subject baseline both defines the
  zero of the concentration scale and enters the CL/F covariate equation. When
  simulating, one value per subject must be supplied and used consistently for
  both purposes.
- **Virtual cohort distributions are assumed.** Sadaf 2024 Table 1 reports
  group means and min-max but no distributional form or correlation structure.
  Weight and baseline L-glutamine are drawn here as independent truncated
  normals per allocation group with the spread set so the reported range spans
  about four SD. In the real cohort the two are almost certainly correlated
  with age and group, and n = 4 per group makes the reported ranges unstable.
- **Occasion-to-dose-level mapping is a presentation choice.** The paper defines
  four occasions (Methods 2.6.1) corresponding to study visits, and Figure 1
  carries the visit-by-visit dose and meal schema, but the running text does
  not state which visit carried which dose level unambiguously ("the dose was
  increased (visit 2) or continued unchanged (visit 3)"). This vignette assigns
  the three simulated arms to occasions 1, 2, and 3. Because Table 2 reports a
  single IOV variance shared across all four occasions, the assignment is
  exchangeable and does not affect the predicted distribution.
- **Cmax could not be compared numerically.** Sadaf 2024 presents Cmax only in
  Figure 3, dose-normalised, with no transcribable axis values, and reports no
  Cmax table. Only Tmax and half-life entered the comparison table. Consistent
  with the standing policy, no value was digitised off the figure.
- **The published Tmax and half-life are pooled, not per dose level.** They are
  repeated across the three arms in the comparison table above. The paper's
  per-dose statement about Tmax (higher at 0.6 g/kg, p < 0.05) is checked
  separately in the typical-value table under Figure 3.
- **Screened but unretained covariates are documented, not encoded.** Age,
  height, body surface area, sex, serum creatinine, haemoglobin genotype,
  baseline L-glutamate, and meal intake were all tested by stepwise selection
  and none was retained (Results 3.3). No point estimates are published for
  any of them, so they appear in the model file's `covariatesDataExcluded`
  metadata rather than in `covariateData`. Note in particular that the meal
  covariate was *not* coded as a plain binary: Methods 2.6.2 used 0 for
  fasting, 0.5 for a snack, 1 for a full meal, and 1.5 for a snack plus a full
  meal between consecutive measurements.
- **Allometric exponents are treated as fixed.** Eq. 3 states the exponents
  were "set at" 0.75 and 1, and Table 2 prints them as literal exponents with
  no estimate, RSE, or bootstrap interval, unlike every estimated theta in the
  same table. Results 3.3 separately reports that free estimation gave 0.73 and
  0.78; those exploratory values are not the final model and are not used here.
- **A Michaelis-Menten alternative was tested and rejected by the authors.**
  Results 3.3 reports that a Michaelis-Menten base model was unstable, which is
  why the observed capacity-limited behaviour is carried by a dose effect on
  apparent volume instead. The packaged model replicates the authors' final
  structure, not the rejected one.
- **The parent-metabolite extension was abandoned by the authors.** Methods
  2.6.1 reports that a simultaneous L-glutamine / L-glutamate model was
  attempted but the conversion parameter sat near the zero boundary, so
  L-glutamate is not part of the final model and is not extracted.

# Errata and source-quality notes

- Sadaf 2024 Table 1 prints the paediatric age range as "6.8-5.1", which is not
  a range. The correct upper bound is 15.1 years: the paediatric group is
  defined as age < 18 years with a mean of 12.0, and the Abstract and Results
  give the overall cohort range as 6.8-43.8 years with the healthy-volunteer
  group topping out at 43.8. The vignette population table above uses 15.1.
- The Table 2 footnote spells the standard-baseline symbol "Glu_BSLstardard",
  a typographical variant of "standard". The value it defines, 683 umol/L, is
  unambiguous and is the normalisation constant used in the model.
- Table 2's footnote also glosses "BIO relative bioavailability", but no BIO
  parameter appears anywhere in the table or in the model equations. It appears
  to be a leftover abbreviation; no bioavailability term is extracted.
- Methods 2.6.1 describes the residual error as "a proportional and additive
  error model", but Eq. 2 as typeset is purely proportional,
  `Y = Cpred * (1 + eps_prop)`, and Table 2 reports the additive component as
  "Fixed to 0". The equation and the table agree with each other, so the model
  carries a proportional error only; the prose is describing the error model
  that was *tested* rather than the one that was retained.
- No erratum or corrigendum for this article was found. The article is open
  access under CC BY-NC 4.0.
- The Electronic Supplementary Material (participant flow diagram, adverse-event
  table, and goodness-of-fit plots) contains no parameter values needed for the
  model, so the extraction does not depend on it.
