---
title: "Ciprofloxacin whole-body PBPK (Sadiq 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ciprofloxacin whole-body PBPK (Sadiq 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Sadiq_2017_ciprofloxacin_pbpk"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Sadiq_2017_ciprofloxacin_pbpk"))$description`
* Article: <https://doi.org/10.1007/s10928-016-9486-9> (open access, CC-BY 4.0)
* Supplement: Electronic Supplementary Material `10928_2016_9486_MOESM1_ESM.pdf`,
  "NONMEM code WB-PBPK-PD ciprofloxacin.mod", retrieved from the EuropePMC
  `supplementaryFiles` endpoint for PMC5376394.

Sadiq et al. built a whole-body PBPK model for ciprofloxacin in intensive-care
patients and fitted it **to plasma concentrations alone**. What makes that
possible is the use of literature tissue-to-plasma partition coefficients as
informative `$PRIOR` (NWPRI) distributions with 25 % uncertainty: the plasma
data cannot identify eleven partition coefficients on their own, but they can
move each one away from its prior where the data demand it.

This vignette validates the PBPK half of the paper. The bacterial-kill PKPD
module that the paper couples downstream of it is not part of this model file;
see [Assumptions and deviations](#assumptions-and-deviations).

```{r load_model}
mod <- readModelDb("Sadiq_2017_ciprofloxacin_pbpk")
ui <- rxode2::rxode2(mod)

# The model must solve its ODEs rather than collapsing to an analytic
# solution: a `cl`/`vc` pair can make rxode2 silently auto-solve and discard
# the explicit d/dt() system.
stopifnot(length(ui$linCmt) == 0L, length(ui$state) == 15L)
ui$state
```

# Population

`r rxode2::rxode2(mod)$population$disease_state`

102 adult ICU patients (27 women, 75 men), mean total body weight 77 +/- 16 kg,
mean age 60 +/- 17 years, measured creatinine clearance 82 +/- 51 mL/min. All
were mechanically ventilated. 588 plasma concentrations were available, on
average 5.8 per patient across an average of 3.1 dosing-interval occasions.
86 of the 102 received the 400 mg twice-daily 1 h infusion that is simulated
below.

No tissue was sampled in this study. Every tissue profile in this vignette is
therefore a *prediction* informed by the literature Kp priors, not a fit to
tissue data, which is exactly the claim the paper makes for it.

# Source trace

Every value in `ini()` and every equation in `model()` comes from one of the
three locations below. "ESM" is the deposited NONMEM control stream.

| Model element | Source location |
|---|---|
| `lcl_nonren`, `fsec`, the eleven `lkp_*` | Table 1, columns "Model estimate log-scale (+/- SE)" and "Model estimate normal scale (RSE)" |
| `etalcl`, `etalkp` | ESM `$OMEGA` 0.316 and 0.306; cross-checked against Table 1 rows "IIV CL (CV %)" = 56 and "IIV Kp (CV %)" = 55 |
| `expSd` | ESM `$SIGMA` 0.112; `sqrt(0.112)` = 0.3347, Table 1 "Proportional residual error (%)" = 33 |
| `fu` = 0.65 | Methods ("fu = 0.65"); ESM `$PK` `FUP = 0.65` |
| `co_coef`, `co_exp` | ESM `$PK` `CO = (15*(WT)**(0.74))` |
| `crcl_cap` | ESM `$PK` `IF (CRCL.GT.150) CRCL2=150` |
| `dens_skin`, `dens_adipose` | ESM `$PK` divisors in `VSKN = ...*(WT/1.18)` and `VADI = ...*(WT/0.916)` |
| `fvol_*_f` / `fvol_*_m` (26 values) | ESM `$PK`, the `IF (SEX.EQ.0)` and `IF (SEX.EQ.1)` volume blocks |
| `fq_*_f` / `fq_*_m` (20 values) | ESM `$PK`, the `IF (SEX.EQ.0)` and `IF (SEX.EQ.1)` flow blocks |
| Total clearance split `cl <- cl_renal + cl_nonren` | Article Equation 2 |
| `cl_renal <- CRCL * fu * (1 + fsec)` | Article Equation 3; ESM `$PK` `CLR = (((CRCL2*60/1000)*FUP)*(1+RSEC))*EXP(ETA(1))` |
| The 13 mass-balance ODEs | ESM `$DES` `DADT(1)`-`DADT(13)` |
| `a_urine` + `a_metabolized` | ESM `$DES` `DADT(14)`, split into its two printed terms |
| Observation `Cc <- c_venous`, `lnorm` residual | ESM `$ERROR` `IPRD2 = A(2)/VVEN`, `IPRED = LOG(IPRD2)`, `Y = IPRED + EPS(1)` |
| Unbound extracellular multipliers (Figure 3 below) | ESM `$ERROR` `CC1`-`CC13`; the underlying definition is article Equation 4 |

## Structural check: the published fractions balance

The deposited code prints two full sets of fractional blood flows. Both sets
must exhaust cardiac output exactly, which is a free transcription check on
twenty numbers.

```{r flow_balance}
fq_female <- c(brain = 0.12, heart = 0.05, skin = 0.05, muscle = 0.12,
               adipose = 0.085, spleen = 0.03, gut = 0.16, kidney = 0.17,
               hepatic_artery = 0.065, other = 0.15)
fq_male <- c(brain = 0.12, heart = 0.04, skin = 0.05, muscle = 0.17,
             adipose = 0.05, spleen = 0.03, gut = 0.15, kidney = 0.19,
             hepatic_artery = 0.065, other = 0.135)

c(female = sum(fq_female), male = sum(fq_male))

# Both arterial-side sets sum to exactly 1. A single mistyped digit anywhere
# in the twenty transcribed flow fractions breaks this.
stopifnot(
  abs(sum(fq_female) - 1) < 1e-12,
  abs(sum(fq_male) - 1) < 1e-12
)
```

# Virtual cohort

The cohort reproduces the demographics the paper reports. Creatinine clearance
is drawn from the reported mean and SD and then floored at 10 mL/min, because a
normal draw with mean 82 and SD 51 would otherwise produce non-physiological
negative values; the paper reports only the mean and SD, so the shape of the
distribution is an assumption (see Errata).

```{r cohort}
rxode2::rxSetSeed(20260916)
set.seed(20260916)

n_sub <- 102L

make_cohort <- function(n, dose_mg, inf_dur_h, label, id_offset = 0L) {
  # 27 of 102 patients were women.
  covs <- data.frame(
    id = seq_len(n) + id_offset,
    WT = pmax(40, rnorm(n, mean = 77, sd = 16)),
    SEXF = as.numeric(seq_len(n) <= round(n * 27 / 102)),
    CRCL = pmax(10, rnorm(n, mean = 82, sd = 51))
  )
  ev <- rxode2::et(
    amt = dose_mg, rate = dose_mg / inf_dur_h,
    ii = 12, addl = 9, cmt = "venous", id = covs$id
  ) |>
    rxode2::et(seq(0, 120, by = 0.25), cmt = "venous", id = covs$id) |>
    as.data.frame()
  ev <- dplyr::left_join(ev, covs, by = "id")
  ev$treatment <- label
  ev
}

events <- dplyr::bind_rows(
  make_cohort(n_sub, 400, 1.0, "400 mg q12h (1 h infusion)", id_offset = 0L),
  make_cohort(n_sub, 200, 0.5, "200 mg q12h (30 min infusion)", id_offset = 1000L)
)

# Duplicate (id, time, evid) rows would silently collapse subjects.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
dplyr::count(events, treatment)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  ui, events,
  keep = c("treatment", "WT", "SEXF", "CRCL"),
  returnType = "data.frame"
)
nrow(sim)
```

## Mass balance

The thirteen drug-holding tissues plus the two elimination integrators must
account for every milligram administered. Both sides of this comparison use the
same drawn parameters for the same subject, so the residual is pure solver
error and a tight bound is the correct assertion.

```{r mass_balance}
states <- c("arterial", "venous", "lung", "brain", "heart", "skin", "muscle",
            "adipose", "spleen", "gut", "liver", "kidney", "other")

mb <- sim |>
  dplyr::group_by(id, treatment) |>
  dplyr::slice_max(time, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    in_body = rowSums(dplyr::across(dplyr::all_of(states))),
    eliminated = a_urine + a_metabolized,
    dosed = ifelse(grepl("^400", treatment), 400, 200) * 10,
    rel_err = abs(in_body + eliminated - dosed) / dosed
  )

max(mb$rel_err)
stopifnot(max(mb$rel_err) < 1e-6)
```

# Replicate published figures

## Figure 2: plasma concentration-time profile

Figure 2 of Sadiq 2017 is a prediction-corrected VPC of plasma concentration
over one dosing interval. Reproduced here as the simulated median and 5th/95th
percentiles over the final steady-state interval.

```{r fig2, fig.cap = "Replicates Figure 2 of Sadiq 2017: simulated plasma ciprofloxacin over one steady-state dosing interval."}
last_int <- sim |>
  dplyr::filter(time >= 108, time <= 120) |>
  dplyr::mutate(tad = time - 108)

pc <- last_int |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05), med = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(pc, aes(tad)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = med), colour = "steelblue", linewidth = 1) +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Plasma ciprofloxacin (mg/L)") +
  theme_bw()
```

## Figure 3: predicted unbound extracellular tissue concentrations

Figure 3 shows unbound ciprofloxacin in the extracellular space of each tissue
at 400 mg b.i.d. Total tissue concentrations are converted to unbound
extracellular concentrations with the per-organ factors the deposited `$ERROR`
block applies (article Equation 4,
`fue = 1 / (1 + (E/P) * (1 - fu) / fu)`, with the rat E/P ratios of
reference 14). Those factors are output scalings, not model dynamics, which is
why they live here rather than in `ini()`.

```{r fig3, fig.cap = "Replicates Figure 3 of Sadiq 2017: predicted unbound extracellular ciprofloxacin by tissue, 400 mg b.i.d."}
# ESM $ERROR: CC3..CC13 multiply the total tissue concentration by these.
fue <- c(lung = 0.79, brain = 0.79, heart = 0.79, skin = 0.65, muscle = 0.79,
         adipose = 0.79, spleen = 0.79, gut = 0.67, liver = 0.79,
         kidney = 0.79, other = 0.79)

# Typical-value patient (median cohort covariates), deterministic.
typ_cov <- data.frame(WT = 77, SEXF = 0, CRCL = 110)
typ_ev <- rxode2::et(amt = 400, rate = 400, ii = 12, addl = 9, cmt = "venous") |>
  rxode2::et(seq(0, 120, by = 0.1), cmt = "venous")
typ <- rxode2::rxSolve(rxode2::zeroRe(ui), typ_ev, typ_cov,
                       returnType = "data.frame")

typ_ss <- typ |> dplyr::filter(time >= 108, time <= 120)

tissue_long <- lapply(names(fue), function(tis) {
  data.frame(
    tissue = tis,
    tad = typ_ss$time - 108,
    cu = typ_ss[[tis]] / typ_ss[[paste0("v_", tis)]] * fue[[tis]]
  )
}) |> dplyr::bind_rows()

ggplot(tissue_long, aes(tad, cu, colour = tissue)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Unbound extracellular ciprofloxacin (mg/L)",
       colour = "Tissue") +
  theme_bw()
```

# Reproducing the paper's own quantitative claims

The paper reports no NCA table, so the quantitative statements in its Results
and Discussion are the available reference values. Each is asserted here rather
than merely narrated: a prose claim in a vignette is read by no gate, so a
false one would render green.

All three checks below run on the deterministic typical-value solve
(`zeroRe()`), so they carry no cohort-draw dependence.

```{r claims}
fu <- 0.65
cl_nonren <- exp(2.60)
cl_renal <- 110 * 60 / 1000 * fu * (1 + 0.674)

# --- Claim 1 (Results): "The renal clearance was for a typical patient with a
# CRCL of 110 ml min-1, estimated to be 49 % of the total clearance."
# Elimination is driven by the unbound liver concentration on the non-renal
# arm, so the effective plasma-referenced non-renal clearance is fu * CL_NR.
last_typ <- typ |> dplyr::slice_max(time, n = 1, with_ties = FALSE)
renal_pct <- 100 * last_typ$a_urine / (last_typ$a_urine + last_typ$a_metabolized)
renal_pct

# --- Claim 2 (Results): "Cmax was achieved at the end of the 1 h constant rate
# infusion in plasma as well as in all other tissues and organs."
tmax_plasma <- typ_ss$time[which.max(typ_ss$Cc)] - 108
tmax_tissue <- vapply(names(fue), function(tis) {
  typ_ss$time[which.max(typ_ss[[tis]])] - 108
}, numeric(1))
c(plasma = tmax_plasma, tmax_tissue)

# --- Claim 3 (Results): "Kidney and lung were predicted to have higher
# exposures as compared to the other organs while muscle, brain and adipose
# were predicted to have relatively low exposures."
auc_tissue <- tapply(tissue_long$cu, tissue_long$tissue, function(cu) {
  tad <- sort(unique(tissue_long$tad))
  sum(diff(tad) * (head(cu, -1) + tail(cu, -1)) / 2)
})
round(sort(auc_tissue, decreasing = TRUE), 2)
```

```{r claim_gates}
ranked <- names(sort(auc_tissue, decreasing = TRUE))
low_three <- c("muscle", "brain", "adipose")

# Claim 2 holds for plasma and for every NAMED organ, but not for the lumped
# rest-of-body compartment, which peaks at 1.9 h. `other` stands in for every
# tissue the model does not resolve and is not one of the organs Figure 3
# plots, so it is checked separately rather than folded into the claim.
named_tissue <- setdiff(names(tmax_tissue), "other")

stopifnot(
  # Claim 1: the paper says 49 %. Bounded rather than pinned because the
  # simulated share integrates a real hepatic-extraction gradient that the
  # closed form ignores.
  renal_pct > 44, renal_pct < 52,
  # Claim 2: Cmax at the end of the 1 h infusion in plasma (exact on the
  # 0.1 h grid) and within one grid step of it in every named organ.
  abs(tmax_plasma - 1) < 1e-8,
  all(tmax_tissue[named_tissue] >= 1 - 1e-8),
  all(tmax_tissue[named_tissue] <= 1.1 + 1e-8),
  # ... and the lumped remainder does lag, which the paper's blanket
  # "all other tissues and organs" wording does not admit.
  tmax_tissue[["other"]] > 1.5,
  # Claim 3: kidney is the highest-exposure organ, and all three of the
  # organs the paper calls low sit in the bottom half of the ranking.
  ranked[1] == "kidney",
  all(match(low_three, ranked) > length(ranked) / 2)
)
```

Claim 2 is reproduced with one qualification. Plasma peaks exactly at the end
of the 1 h infusion, and every named organ peaks there or one 0.1 h grid step
later (muscle and adipose, the two large slowly-equilibrating tissues, at
1.1 h). The lumped `other` compartment -- 23 % of body weight receiving 13.5 %
of cardiac output, standing in for every tissue the model does not resolve --
peaks at 1.9 h. The paper's wording ("in plasma as well as in all other tissues
and organs") does not distinguish it, but `other` is not an organ and is not
among the tissues Figure 3 plots.

The renal share the model produces (`r sprintf("%.1f %%", renal_pct)`)
reproduces the paper's stated 49 %. Note that this only works once the `fu`
factor on the non-renal arm is carried: the naive `CL_R / (CL_R + CL_NR)`
gives 35 %, which is what makes this a real check on the clearance
parameterisation rather than a restatement of it.

The one place the model does **not** follow the prose is the paper's companion
statement about a poorly-functioning kidney; see Errata.

# PKNCA validation

NCA over the final steady-state dosing interval, by treatment arm.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

# The cohort event tables carry one dose record per subject with `addl = 9`,
# so there is no materialised row at t = 108. Build the steady-state dose
# frame explicitly: one row per subject at the start of the interval being
# analysed, carrying that subject's dose amount.
dose_df <- events |>
  dplyr::filter(!is.na(amt), amt > 0) |>
  dplyr::distinct(id, treatment, amt) |>
  dplyr::mutate(time = 108)

stopifnot(nrow(dose_df) == dplyr::n_distinct(events$id))

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 108, end = 120,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, cmin = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res <- PKNCA::pk.nca(nca_data)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_summary <- nca_wide |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    n = dplyr::n(),
    Cmax_median = median(cmax),
    Tmax_median = median(tmax),
    AUCtau_median = median(auclast),
    Ctrough_median = median(cmin),
    .groups = "drop"
  )

nca_summary |>
  dplyr::rename(
    "Treatment" = treatment,
    "N" = n,
    "Cmax (mg/L)" = Cmax_median,
    "Tmax (h)" = Tmax_median,
    "AUC0-tau (mg*h/L)" = AUCtau_median,
    "Ctrough (mg/L)" = Ctrough_median
  ) |>
  knitr::kable(digits = 3, caption = "Simulated steady-state NCA by treatment arm (medians).")
```

## Dose proportionality

The model is linear in dose: nothing in the mass balance saturates. Halving the
dose must halve every exposure metric exactly, which is a check on the dosing
and infusion-rate encoding rather than on the parameter values.

Run on the deterministic typical-value patient, so the ratio carries no
cohort-draw dependence. The two arms differ in infusion duration as well as in
dose, which makes this a check on the `rate` encoding too: AUC over a complete
dosing interval is invariant to infusion duration, so any leakage of duration
into exposure would show up here.

```{r dose_prop}
auc_tau <- function(dose_mg, inf_dur_h) {
  ev <- rxode2::et(amt = dose_mg, rate = dose_mg / inf_dur_h,
                   ii = 12, addl = 9, cmt = "venous") |>
    rxode2::et(seq(0, 120, by = 0.1), cmt = "venous")
  s <- rxode2::rxSolve(rxode2::zeroRe(ui), ev, typ_cov, returnType = "data.frame")
  s <- s[s$time >= 108 & s$time <= 120, ]
  tad <- s$time - 108
  sum(diff(tad) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
}

ratio <- auc_tau(400, 1.0) / auc_tau(200, 0.5)
ratio

stopifnot(abs(ratio - 2) < 1e-6)
```

## Clearance recovery

At steady state, dose divided by AUC over the dosing interval recovers the
effective plasma-referenced clearance.

```{r cl_recovery}
auc_typ <- local({
  tad <- typ_ss$time - 108
  sum(diff(tad) * (head(typ_ss$Cc, -1) + tail(typ_ss$Cc, -1)) / 2)
})
cl_obs <- 400 / auc_typ
cl_pred <- cl_renal + fu * cl_nonren
c(observed = cl_obs, closed_form = cl_pred,
  pct_diff = 100 * (cl_obs - cl_pred) / cl_pred)

# The closed form assumes the liver's emergent plasma-equivalent concentration
# equals the venous concentration. It does not exactly, because the liver
# extracts; a few percent of disagreement is the hepatic gradient, not a
# transcription error.
stopifnot(abs(100 * (cl_obs - cl_pred) / cl_pred) < 10)
```

# Assumptions and deviations {#assumptions-and-deviations}

## Errata: two disagreements between Table 1 and the deposited control stream

1. **`Kp,muscle` sign.** Table 1 prints `-0.0229` in the log-scale column and
   `0.977` in the normal-scale column; `exp(-0.0229) = 0.9774`, so the two
   printed columns agree with each other. The deposited `$THETA 7` prints
   `0.0229` with no minus sign. This model uses Table 1's `-0.0229`, because
   Table 1 is self-consistent across two independent columns while the control
   stream carries a single number. The control stream does not drop minus signs
   generally: `-0.257`, `-0.335`, `-0.885`, `-0.26`, `-0.33` and `-0.80` all
   survive elsewhere in the same block, so the discrepancy is in the source and
   not an artefact of text extraction.

2. **`f_secretion` scale.** The deposited `$PK` block computes
   `TVRSEC = EXP(THETA(2))`, which with `THETA(2) = 0.674` would make the
   secretion factor `exp(0.674) = 1.96`. This model uses `0.674` on the natural
   scale, for three independent reasons: Table 1 places `0.674` in the
   "Model estimate **normal scale**" column and leaves that row's log-scale cell
   empty (unlike `CL_NR` and every `Kp`, which have both); article Equation 3 is
   written `CL_R = CRCL x fu,plasma x (1 + f_Secretion)` with a bare term; and
   the Methods call it "the ... fraction of the ciprofloxacin renal clearance
   that is dependent on secretion", which cannot exceed 1. The prior block is
   the same story (`$THETA 0.57 FIX` against Table 1's prior of 0.57 on the
   normal scale, while the `CL_NR` prior `1.97` *is* logged, `exp(1.97) = 7.17`).
   The reading adopted here is the one that reproduces the paper's own 49 %
   renal-share claim; see the claims section above.

## Deviations from the source

* **Eliminated-drug integrator split.** The deposited `$DES` carries one
  cumulative sink, `DADT(14) = C1*CLR + (C11*CLH*FUP/KHEP)`. It is split here
  into `a_urine` and `a_metabolized`, one per printed term, so the two routes
  can be read separately. Their sum reproduces `DADT(14)` exactly and the
  dynamics are unchanged.
* **`SEXF` inverted relative to the source.** The control stream codes
  `SEX = 0` for female; the library canonical is `SEXF` with 1 = female, so
  `SEXF = 1 - SEX`. The direction is fixed independently by the volume block
  (the `SEX = 0` branch carries the higher adipose and lower muscle fraction)
  and by the flow block (higher adipose and lower muscle flow), so this is not
  an inference from the label alone.
* **Per-organ `fue` multipliers are not `ini()` parameters.** The deposited code
  applies them in `$ERROR`, i.e. as output scalings that do not feed back into
  the mass balance, so they are applied in this vignette rather than in the
  model file. There is at present no canonical library name for a per-organ
  unbound extracellular fraction; minting one was left to a naming decision
  rather than taken silently.

## Assumptions

* **Covariate distributions.** The paper reports only means and SDs for weight
  and creatinine clearance. The virtual cohort draws both from normal
  distributions with those moments, flooring weight at 40 kg and creatinine
  clearance at 10 mL/min to keep draws physiological. The reported SD for
  creatinine clearance (51 on a mean of 82) is large enough that the true
  distribution is certainly right-skewed rather than normal, so the cohort's
  tails should not be read as the study's tails. Nothing asserted in this
  vignette depends on them.
* **Dosing compartment.** The deposited data file is not published, so the
  compartment the dose record targets is not directly observable. The dose is
  administered into `venous` here: the drug was given as an intravenous
  infusion into a peripheral vein, `$ERROR` observes compartment 2 (venous
  blood) as the plasma prediction, and venous blood is the conventional input
  compartment for a whole-body PBPK IV dose.
* **Sex assignment in the cohort** is deterministic (the first 27 of every 102
  subjects are female) rather than random, so the female fraction is exactly
  the study's 26.5 % in every draw.

## Claims in the paper that this model does not reproduce

The paper's renal-share statements at reduced renal function are mutually
inconsistent and neither matches the model. Results says that at
`CRCL = 50 mL/min` "renal clearance was predicted to constitute 13 % of the
total clearance"; the Discussion says "with decreased CRCL of 50 ml min-1 renal
clearance was reduced to 16 % of total clearance". The model gives 27 % at that
creatinine clearance. The `CRCL = 110` statement (49 %) *is* reproduced, and
Equation 3 is linear in `CRCL`, so no single reading of the clearance model can
satisfy both the 110 and the 50 statements. This is recorded as a
paper-internal inconsistency rather than resolved.

Similarly, the Results and Discussion both describe `Kp,muscle` as "60 % higher"
than its literature prior, while Table 1 gives an estimate of 0.977 against a
prior of 1.6 -- i.e. lower, by a ratio whose magnitude (1.6 / 0.977 = 1.64) is
plausibly where the "60 %" came from. The tabulated values are used.

## Scope

This file covers the WB-PBPK model only. The paper goes on to couple it to an
in-vitro-derived bacterial-kill PKPD model for *E. coli* (Figures 4-6),
including a neutrophil-mediated immune term. All of that module's parameters
are present in the deposited control stream, so it is extractable, but it
introduces six bacterial states per strain -- growing, resting and
non-plateable (filamentous), each with a pre-existing-resistant counterpart --
for which the library has no canonical compartment names, and its "resting"
state collides with the register's existing `R`, which means *resistant*. That
naming decision is deliberately left to the maintainers rather than guessed.
