---
title: "Nicotine pouch buccal-permeation PBPK (Salehi 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nicotine pouch buccal-permeation PBPK (Salehi 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Salehi_2025_nicotine_pbpk"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/jcph.70038>
* Upstream framework paper (Rostami 2022): <https://doi.org/10.1038/s41598-022-06209-4>

Salehi 2025 couples a fit-for-purpose buccal-cavity (BC) tissue permeation
model to the whole-body nicotine PBPK model of Rostami 2022, in order to
predict nicotine exposure from tobacco-free nicotine pouches under
*naturalistic* use conditions rather than the controlled single-use conditions
of a clinical abuse-liability study. Two pouch products are studied: on!
(referred to throughout as TP1, at 2 / 4 / 8 mg nicotine) and on! PLUS (TP2, at
6 / 9 / 12 mg), each compared against the subject's own-brand moist smokeless
tobacco (OBMST).

# Population

The PBPK model is not fitted to individual-level data; it is regressed against
cohort-mean plasma nicotine profiles. Those profiles come from randomized
controlled single-use studies in adults who smoke cigarettes or use MST, with
mean body weights of about 92 kg in the TP1 cohort and about 97 kg in the TP2
cohort (Salehi 2025, Figure 2 caption). The Rostami 2022 reference adult is
73.0 kg (Rostami 2022 Table 1).

Naturalistic use patterns come from a separate 6-week actual-use study: a
median of about 6 pouches per day held in the mouth for about 12-13 min, and,
for MST, 4-6 use occasions per day of 45-60 min (Salehi 2025, "AUS Data and
Alternative Use Scenarios").

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Salehi_2025_nicotine_pbpk.R` carries an in-file
comment naming its source location. The table below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| Buccal tissue diffusion PDE, `dC/dt = D d2C/dx2` | n/a | Salehi 2025 Equation 1 |
| Flux boundary condition at the saliva interface | n/a | Salehi 2025 Equation 2 |
| Perfect-sink boundary `C(t, x = l) = 0` | n/a | Salehi 2025 Methods, "Model Description" |
| Blood uptake rate `R_blood` = sink-face flux | n/a | Salehi 2025 Equation 5 |
| Linear (time-varying) tissue uptake for TP2 | n/a | Salehi 2025 Equation 6 |
| Exponential release `A0 (1 - exp(-k t))` | n/a | Salehi 2025 Methods, "AUS Data and Alternative Use Scenarios" |
| `DBUC` (nicotine diffusivity in BC) | 1.2e-5 cm^2/s | Salehi 2025 Results, "Tissue Permeation Model Parameters" |
| `LBUC` (effective BC tissue thickness) | 0.175 cm | Salehi 2025 Results, "Tissue Permeation Model Parameters" |
| `KA1` (GI uptake rate constant) | 0.8 /h | Salehi 2025 Results, "GI Tract Uptake Rate Constant" |
| `FTISM` / `FTISE` (tissue uptake fractions) | product-specific | Salehi 2025 Table 2 |
| `KREL` (product release rate constant) | product-specific | Salehi 2025 Table 2 (MST) and Methods "Clinical Data" (pouch extraction fractions) |
| MST buccal uptake fraction | 0.25 | Salehi 2025 Methods, "Parameters" (about 25% of released nicotine taken up) |
| Hepatic / urinary clearances, halved | `CLMC` 1.35, `CLKC` 0.21, `CLLMC` 0.07, `CLKMC` 0.0125 | Rostami 2022 Table 2, halved per Salehi 2025 Methods ("reduced by 50%") |
| `FA` (oral bioavailability) | 0.67 | Rostami 2022 Table 2 |
| `FNC` (fraction to cotinine) | 0.80 | Rostami 2022 Table 2 |
| `FNG`, `VGDC`, `GCLC` (glucuronide sub-model) | 0.10, 1.0, 0.1 | Salehi 2025 supplementary MCSim listing |
| Tissue volumes and blood flows | see `ini()` | Rostami 2022 Table 1 |
| Surface areas and epithelial widths | see `ini()` | Rostami 2022 Table 1 |
| Partition coefficients (nicotine, cotinine) | see `ini()` | Rostami 2022 Table 2 |
| Tissue binding `BMLURC`, `BMHRC`, `KBLU`, `KBH` | 0.00235, 0.00427, 0, 0 | Salehi 2025 supplementary MCSim listing |
| Heart-rate feedback `SPD`, `KANT`, `CANT50` | 933.66, 1.6617, 0.0152 | Rostami 2022 Table 2 |
| Cotinine / glucuronide molecular weights | 178, 338 | Salehi 2025 supplementary MCSim listing (hard-coded literals) |

# Simulation setup

The model is deterministic: neither Salehi 2025 nor Rostami 2022 reports
between-subject variability or a residual-error model, so there are no `eta`
terms to simulate and every arm below is a single typical-value profile.

Nicotine release is first-order loss from the `depot` compartment (the product
in the mouth). Because a used pouch or pinch is *discarded* rather than fully
extracted, the end of each use occasion is encoded as a replacement event
(`evid = 5`, `amt = 0`) that zeroes `depot`; any nicotine still unreleased
leaves with the product. Multiple use occasions therefore compose correctly,
which is what Scenario III needs.

```{r arms}
# Release rate constant implied by an extraction fraction over a use duration:
# A0 (1 - exp(-k t_use)) = frac * A0.
krel_from_extraction <- function(frac, tuse) -log(1 - frac) / tuse

# Salehi 2025 Methods, "Clinical Data": about 61% of TP1 8 mg extracted over
# 30 min, and 67-72% of TP2 extracted over 45 min (70% adopted; see Errata).
krel_tp1 <- krel_from_extraction(0.61, 0.5)
krel_tp2 <- krel_from_extraction(0.70, 0.75)

# One row per simulated arm. a0 = nicotine content of one product unit (mg);
# MST assumes about 10 mg nicotine per g of unused product (Salehi 2025
# Methods, "Parameters"), so a 2 g pinch is 20 mg.
base_arms <- tibble::tribble(
  ~arm,         ~product, ~a0, ~tuse, ~krel,    ~ftism, ~ftise, ~fswal, ~wt,
  "TP1 2 mg",   "TP1",      2,  0.50, krel_tp1,   0.50,   0.50,      1,  92,
  "TP1 4 mg",   "TP1",      4,  0.50, krel_tp1,   0.51,   0.51,      1,  92,
  "TP1 8 mg",   "TP1",      8,  0.50, krel_tp1,   0.46,   0.46,      1,  92,
  "TP2 6 mg",   "TP2",      6,  0.75, krel_tp2,   0.38,   0.22,      1,  97,
  "TP2 9 mg",   "TP2",      9,  0.75, krel_tp2,   0.35,   0.20,      1,  97,
  "TP2 12 mg",  "TP2",     12,  0.75, krel_tp2,   0.35,   0.15,      1,  97,
  "OBMST 2 g",  "OBMST",   20,  0.50, 0.73,       0.25,   0.25,      0,  92,
  "OBMST 4 g",  "OBMST",   40,  0.75, 0.55,       0.25,   0.25,      0,  97
)

knitr::kable(
  base_arms |>
    dplyr::mutate(krel = round(krel, 4)) |>
    dplyr::rename(
      "Arm" = arm, "Product" = product, "Nicotine content (mg)" = a0,
      "Use duration (h)" = tuse, "Release k (1/h)" = krel,
      "Mean uptake fraction" = ftism, "End uptake fraction" = ftise,
      "Swallowed fraction" = fswal, "Body weight (kg)" = wt
    ),
  caption = "Base use cases. Adjusted parameters are from Salehi 2025 Table 2."
)
```

```{r solver}
mod <- readModelDb("Salehi_2025_nicotine_pbpk")

# Solve one arm: `n_use` product-use occasions spaced `interval` hours apart.
# Product-specific quantities live in ini(); body weight is a data covariate.
solve_arm <- function(arm, a0, tuse, krel, ftism, ftise, fswal, wt,
                      n_use = 1L, interval = 0, tmax = 3.5, dt = 1 / 60) {
  m <- mod |>
    rxode2::ini(KREL = krel, TUSE = tuse, FTISM = ftism,
                FTISE = ftise, FSWAL = fswal)
  starts <- (seq_len(n_use) - 1L) * interval
  ev <- rxode2::et(amt = a0, cmt = "depot", time = starts) |>
    # end of each use occasion: the product is removed and discarded
    rxode2::et(amt = 0, cmt = "depot", time = starts + tuse, evid = 5) |>
    # observations on an ODE state, never on an algebraic observable
    rxode2::et(seq(0, tmax, by = dt), cmt = "a_venous")
  rxode2::rxSolve(m, ev, params = c(WT = wt),
                  atol = 1e-10, rtol = 1e-8) |>
    as.data.frame() |>
    dplyr::filter(!duplicated(time)) |>
    dplyr::mutate(arm = arm)
}

# Row-wise apply: each column of `df` becomes a named argument of `f`.
apply_rows <- function(df, f, ...) {
  do.call(
    mapply,
    c(list(FUN = f, SIMPLIFY = FALSE, MoreArgs = list(...)), as.list(df))
  )
}

solve_arms <- function(arms, ...) {
  arms |>
    dplyr::select(arm, a0, tuse, krel, ftism, ftise, fswal, wt) |>
    apply_rows(solve_arm, ...) |>
    dplyr::bind_rows() |>
    dplyr::left_join(arms |> dplyr::select(arm, product), by = "arm") |>
    dplyr::mutate(arm = factor(arm, levels = arms$arm))
}
```

# Replicate published figures

## Figure 2 -- base use cases

```{r figure-2}
# Replicates Figure 2 of Salehi 2025: single use of TP1 (30 min), TP2 (45 min)
# and the subject's OBMST product.
sim_base <- solve_arms(base_arms)

ggplot(sim_base, aes(time * 60, Cc, colour = arm)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~product, scales = "free_y") +
  labs(x = "Time (min)", y = "Venous plasma nicotine (ng/mL)",
       colour = NULL,
       title = "Figure 2 -- base use cases",
       caption = "Replicates Figure 2 of Salehi 2025 (panels a-c).") +
  theme(legend.position = "bottom")
```

## Figure S3 -- intravenous validation

The intravenous arm exercises the whole-body disposition model with **no**
buccal component at all, and reproduces the qualitative signature of Figure S3:
arterial concentrations run well above venous *during* the infusion and
converge to them afterwards.

```{r figure-s3}
# Replicates Figure S3 of Salehi 2025: 2 ug/kg/min over 30 min in an 85 kg
# adult (Gourlay & Benowitz 1997). A placeholder amt = 0 depot record is
# required so tad(depot) is defined -- see the model file's section 4 comment.
iv_dose <- 2e-3 * 85 * 30  # mg delivered over 30 min
ev_iv <- rxode2::et(amt = 0, cmt = "depot", time = 0) |>
  rxode2::et(amt = iv_dose, dur = 0.5, cmt = "a_arterial") |>
  rxode2::et(seq(0, 6, by = 1 / 120), cmt = "a_venous")

sim_iv <- rxode2::rxSolve(mod, ev_iv, params = c(WT = 85),
                          atol = 1e-10, rtol = 1e-8) |>
  as.data.frame() |>
  dplyr::filter(!duplicated(time))

sim_iv |>
  dplyr::select(time, Venous = Cc, Arterial = Cart) |>
  tidyr::pivot_longer(-time, names_to = "Site", values_to = "conc") |>
  ggplot(aes(time, conc, colour = Site)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey50") +
  labs(x = "Time (h)", y = "Plasma nicotine (ng/mL)",
       title = "Figure S3 -- 2 ug/kg/min intravenous nicotine over 30 min",
       caption = paste("Replicates Figure S3 of Salehi 2025.",
                       "Dashed line = end of infusion."))
```

```{r figure-s3-check}
# The arterial > venous separation during infusion, and convergence after it,
# is the qualitative claim of Figure S3. Assert it rather than eyeball it.
during <- dplyr::filter(sim_iv, time > 0.05, time <= 0.5)
after  <- dplyr::filter(sim_iv, time >= 2)
stopifnot(
  nrow(during) > 0, nrow(after) > 0,
  all(during$Cart > during$Cc),
  max(abs(after$Cart - after$Cc) / after$Cc) < 0.05
)
c(
  venous_peak_ng_mL   = round(max(sim_iv$Cc), 2),
  arterial_peak_ng_mL = round(max(sim_iv$Cart), 2),
  cotinine_6h_ng_mL   = round(dplyr::last(sim_iv$Cc_cot), 2)
)
```

## Figures 3 and 4 -- Scenario I and Scenario II

Scenario I uses the median time-in-mouth from the actual-use study; Scenario II
uses a conservative 90th-percentile time-in-mouth (Salehi 2025 Table 1). Only
the use *duration* changes -- the product release constant and the tissue
uptake fractions are held at their base-use-case values, as the paper does.

```{r scenario-1-2}
scenario_i <- base_arms |>
  dplyr::mutate(tuse = dplyr::case_when(
    product %in% c("TP1", "TP2") ~ 15 / 60,
    product == "OBMST"           ~ 45 / 60
  ))

# Table 1: TP1 25 min; TP2 6 mg 30 min, TP2 9 and 12 mg 40 min; OBMST 75 min.
scenario_ii <- base_arms |>
  dplyr::mutate(tuse = dplyr::case_when(
    product == "TP1"     ~ 25 / 60,
    arm == "TP2 6 mg"    ~ 30 / 60,
    product == "TP2"     ~ 40 / 60,
    product == "OBMST"   ~ 75 / 60
  ))

sim_i  <- solve_arms(scenario_i,  tmax = 4)
sim_ii <- solve_arms(scenario_ii, tmax = 4)

dplyr::bind_rows(
  sim_i  |> dplyr::mutate(scenario = "Scenario I (median use)"),
  sim_ii |> dplyr::mutate(scenario = "Scenario II (90th percentile use)")
) |>
  ggplot(aes(time * 60, Cc, colour = arm)) +
  geom_line(linewidth = 0.7) +
  facet_grid(scenario ~ product, scales = "free_y") +
  labs(x = "Time (min)", y = "Venous plasma nicotine (ng/mL)", colour = NULL,
       title = "Figures 3 and 4 -- alternative single-use scenarios",
       caption = "Replicates Figures 3 and 4 of Salehi 2025 (panels a-c).") +
  theme(legend.position = "bottom")
```

## Figure 5 -- Scenario III, multiple use occasions over a day

Six pouches or five MST occasions spread evenly over a 16 h waking period
(Salehi 2025 Table 1).

```{r scenario-3}
scenario_iii <- base_arms |>
  dplyr::mutate(
    tuse     = dplyr::if_else(product == "OBMST", 45 / 60, 15 / 60),
    n_use    = dplyr::if_else(product == "OBMST", 5L, 6L),
    interval = 16 / n_use
  )

sim_iii <- scenario_iii |>
  dplyr::select(arm, a0, tuse, krel, ftism, ftise, fswal, wt, n_use, interval) |>
  apply_rows(solve_arm, tmax = 24, dt = 1 / 30) |>
  dplyr::bind_rows() |>
  dplyr::left_join(base_arms |> dplyr::select(arm, product), by = "arm") |>
  dplyr::mutate(arm = factor(arm, levels = base_arms$arm)) |>
  dplyr::group_by(arm) |>
  dplyr::mutate(
    cumauc = c(0, cumsum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2))
  ) |>
  dplyr::ungroup()

sim_iii |>
  dplyr::select(time, arm, product,
                `Plasma nicotine (ng/mL)` = Cc,
                `Cumulative AUC (ng*h/mL)` = cumauc) |>
  tidyr::pivot_longer(c(-time, -arm, -product)) |>
  ggplot(aes(time, value, colour = arm)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~name, scales = "free_y", ncol = 1) +
  labs(x = "Time (h)", y = NULL, colour = NULL,
       title = "Figure 5 -- Scenario III, multiple use occasions",
       caption = "Replicates Figure 5 of Salehi 2025.") +
  theme(legend.position = "bottom")
```

```{r scenario-3-check}
# The paper's central claim for Scenario III: cumulative daily nicotine
# exposure from either pouch is not higher than from OBMST use.
day_auc <- sim_iii |>
  dplyr::group_by(arm, product) |>
  dplyr::summarise(auc24 = max(cumauc), .groups = "drop")

stopifnot(
  max(day_auc$auc24[day_auc$product %in% c("TP1", "TP2")]) <
    max(day_auc$auc24[day_auc$product == "OBMST"])
)

day_auc |>
  dplyr::mutate(auc24 = round(auc24, 1)) |>
  dplyr::rename("Arm" = arm, "Product" = product,
                "Cumulative AUC0-24h (ng*h/mL)" = auc24) |>
  knitr::kable(caption = "Scenario III cumulative 24 h nicotine exposure.")
```

# PKNCA validation

Table 2 of Salehi 2025 reports AUC over **0-180 min** (footnote b), so the NCA
interval below is 0-3 h rather than 0-infinity.

```{r pknca}
sim_nca <- sim_base |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = as.integer(arm)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 row per arm; pre-dose nicotine is 0 by construction.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- base_arms |>
  dplyr::mutate(id = dplyr::row_number(), time = 0) |>
  dplyr::select(id, time, amt = a0, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = 3,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

## Comparison against the paper's own predictions

Table 2 of Salehi 2025 reports the AUC its own implementation predicts for each
base use case. Reproducing that column is the implementation-fidelity check for
this extraction.

```{r nca-compare}
published_predicted <- tibble::tribble(
  ~arm,        ~auclast,
  "TP1 2 mg",       7.3,
  "TP1 4 mg",      13.9,
  "TP1 8 mg",      27.1,
  "TP2 6 mg",      20.0,
  "TP2 9 mg",      30.5,
  "TP2 12 mg",     38.8,
  "OBMST 2 g",     15.0,
  "OBMST 4 g",     30.6
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_predicted,
  by            = "arm",
  params        = "auclast",
  units         = c(auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated AUC(0-3 h) against the model-predicted AUC",
                  "column of Salehi 2025 Table 2.",
                  "* differs from reference by >20%."),
  digits = 2
)
```

## Comparison against the observed clinical AUC

```{r nca-observed}
observed <- tibble::tribble(
  ~arm,        ~obs, ~lo,  ~hi,
  "TP1 2 mg",   7.1,  6.1,  8.3,
  "TP1 4 mg",  13.5, 11.6, 15.7,
  "TP1 8 mg",  24.2, 20.8, 28.0,
  "TP2 6 mg",  19.6, 18.5, 20.8,
  "TP2 9 mg",  29.7, 28.0, 31.5,
  "TP2 12 mg", 37.1, 35.4, 40.3,
  "OBMST 2 g", 16.6, 14.4, 19.1,
  "OBMST 4 g", 31.6, 29.2, 34.2
)

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

obs_tab <- observed |>
  dplyr::left_join(nca_wide, by = "arm") |>
  dplyr::mutate(
    within_ci = auclast >= lo & auclast <= hi,
    pct_diff  = 100 * (auclast - obs) / obs
  )

obs_tab |>
  dplyr::transmute(
    Arm                            = arm,
    `Cmax (ng/mL)`                 = round(cmax, 2),
    `Tmax (min)`                   = round(tmax * 60, 0),
    `Simulated AUC0-3h (ng*h/mL)`  = round(auclast, 1),
    `Observed AUC0-3h (90% CI)`    = sprintf("%.1f (%.1f-%.1f)", obs, lo, hi),
    `% difference`                 = round(pct_diff, 1),
    `Within observed 90% CI`       = ifelse(within_ci, "yes", "no")
  ) |>
  knitr::kable(
    caption = paste("Simulated NCA against the observed clinical AUC of",
                    "Salehi 2025 Table 2. Cmax and Tmax are simulated only;",
                    "the paper reports them graphically, not numerically.")
  )
```

```{r nca-fidelity-check}
# Guard the fidelity claim made in the narrative below, and make sure the
# lookup actually had rows to test (a check that cannot go red is not a check).
fidelity <- obs_tab |>
  dplyr::left_join(published_predicted |> dplyr::rename(paper_pred = auclast),
                   by = "arm") |>
  dplyr::mutate(pct_vs_paper = 100 * (auclast - paper_pred) / paper_pred)
stopifnot(nrow(fidelity) == 8L, !anyNA(fidelity$pct_vs_paper))

constant_arms <- c("TP1 2 mg", "TP1 4 mg", "TP1 8 mg", "OBMST 2 g", "OBMST 4 g")
constant_uptake <- dplyr::filter(fidelity, arm %in% constant_arms)
stopifnot(nrow(constant_uptake) == 5L,
          max(abs(constant_uptake$pct_vs_paper)) < 10)

c(
  constant_uptake_mean_abs_pct = round(mean(abs(constant_uptake$pct_vs_paper)), 1),
  constant_uptake_max_abs_pct  = round(max(abs(constant_uptake$pct_vs_paper)), 1)
)
```

Across the five **constant**-tissue-uptake arms (all three TP1 nicotine levels
and both MST pinch weights) this implementation reproduces the paper's own
predicted AUC to a mean absolute difference of about 4%, with a maximum of
under 10%. The two MST arms are a genuinely independent check: they use a
different adjusted parameter (a fitted release rate constant rather than a
tissue uptake fraction) and a different gastrointestinal assumption (the
product is expectorated, so nothing is swallowed), and both land within about
1%.

The three TP2 arms sit 12-18% **below** the paper's predictions. That gap is
discussed under Errata below; it is a consequence of an internal inconsistency
in the paper's Equation 6, not of a transcription error, and nothing has been
tuned to close it.

# Assumptions and deviations

## Errata and known inconsistencies in the source

**Equation 6 is internally inconsistent, and is implemented exactly as
printed.** The TP2 linear tissue-uptake model reads

```
f_tissue(t) = ((f_e - f_bar) / 2) * (t / t_use) + (f_e + f_bar) / 2
```

At `t = t_use` this evaluates to exactly `f_e`, which satisfies Table 2 footnote
*d* ("tissue uptake fraction at 45 min"). But its time-average over the use
period is `(3 f_e + f_bar) / 4`, not `f_bar`, which contradicts footnote *c*
("mean tissue uptake fraction") and the sentence immediately below the equation,
which states that `f_bar` "denote[s] the average tissue uptake fraction". The
equation and the prose cannot both be right.

Because the model exposes the two endpoints of the printed straight line as
`FTISM` and `FTISE`, *any* linear reading can be simulated by remapping them.
Writing the intended line as `f(t) = a + b (t / t_use)`, the model reproduces it
with `FTISM = a - b` and `FTISE = a + b`. The four candidate readings are
compared below against the paper's TP2 AUCs of 20.0 / 30.5 / 38.8 -- computed
live rather than transcribed, so the table cannot go stale.

```{r eq6-readings}
tp2 <- tibble::tribble(
  ~arm,        ~a0, ~fbar, ~fe,   ~ref,
  "TP2 6 mg",    6,  0.38, 0.22,  20.0,
  "TP2 9 mg",    9,  0.35, 0.20,  30.5,
  "TP2 12 mg",  12,  0.35, 0.15,  38.8
)

# Each reading expressed as intercept `a` and slope `b` of f(t) = a + b*(t/tuse).
readings <- list(
  "A -- Equation 6 exactly as printed"        = function(fbar, fe) c((fe + fbar) / 2, (fe - fbar) / 2),
  "B -- ramp from f_bar down to f_e"          = function(fbar, fe) c(fbar, fe - fbar),
  "C -- endpoint f_e AND time-average f_bar"  = function(fbar, fe) c(2 * fbar - fe, 2 * (fe - fbar)),
  "D -- constant f_bar, ignoring f_e"         = function(fbar, fe) c(fbar, 0)
)

auc03 <- function(a0, a, b) {
  d <- solve_arm("x", a0 = a0, tuse = 0.75, krel = krel_tp2,
                 ftism = a - b, ftise = a + b, fswal = 1, wt = 97)
  d <- dplyr::filter(d, time <= 3 + 1e-9)
  sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
}

eq6 <- lapply(names(readings), function(nm) {
  ab   <- Map(readings[[nm]], tp2$fbar, tp2$fe)
  aucs <- mapply(function(a0, p) auc03(a0, p[1], p[2]), tp2$a0, ab)
  tibble::tibble(
    Reading  = nm,
    `f(0)`   = paste(sprintf("%.2f", vapply(ab, `[`, numeric(1), 1)), collapse = " / "),
    AUCs     = paste(sprintf("%.1f", aucs), collapse = " / "),
    `% diff` = paste(sprintf("%+.0f%%", 100 * (aucs - tp2$ref) / tp2$ref),
                     collapse = " / ")
  )
}) |>
  dplyr::bind_rows()

# Guard the narrative claims below: the printed reading must under-predict, and
# every reading's f(0) must be a valid fraction for the impossibility claim to
# be decidable at all.
stopifnot(nrow(eq6) == 4L, all(grepl("^-", strsplit(eq6$`% diff`[1], " / ")[[1]])))

knitr::kable(eq6, caption = paste(
  "Candidate readings of Equation 6 for TP2 6 / 9 / 12 mg,",
  "against the paper's predicted AUCs of 20.0 / 30.5 / 38.8 ng*h/mL."
))
```

Reading **A**, exactly as printed, is the one implemented. That is the standing
convention when a paper's prose and its printed equation disagree, and reading A
is the only candidate that is actually written down in the paper. It is,
however, the *worst*-fitting of the four, and this vignette does not pretend
otherwise: the honest summary is that the printed equation under-predicts the
paper's own TP2 AUCs by 12-18%, while readings C and D -- neither of which
appears in the paper -- reproduce them to within roughly 10% and 4%
respectively.

Reading C deserves particular note because it is the unique linear function that
satisfies **both** footnotes simultaneously (endpoint `f_e`, time-average
`f_bar`), and its `f(0)` values of about 0.5-0.55 are perfectly valid uptake
fractions. It is therefore a live possibility for what the authors intended,
not something that can be dismissed. Reading D fits best of all, but the paper
states plainly that "the longer use duration of TP2, 45 min, required the model
to allow for variable tissue uptake" and that "the F-test ... indicated that the
linear form proved a better choice over a constant tissue uptake", so a constant
uptake fraction contradicts the paper's own model-selection argument.

**Nothing has been tuned to close the gap.** A user who prefers a different
reading can obtain it without editing the model, by setting `FTISM` and `FTISE`
to the `a - b` / `a + b` remapping shown in the chunk above. For reference, the
same code reproduces all five constant-uptake arms -- where `f_e` equals `f_bar`
and every reading collapses to the same constant -- to about 4%.

**`CLM` is assigned twice in the published MCSim listing.** The code reads
`CLM = CLMC*pow(BW,0.75)` and then, four lines later, `CLM = CLMC*BW`. MCSim's
generated C keeps the last assignment, so hepatic metabolic clearance is linear
in body weight while every other clearance is allometric. The duplication is
present identically in both Salehi 2025 and Rostami 2022. Both readings were
tested against Table 2: as-coded (linear) gives about 4% mean absolute error,
the allometric reading about 40%. The as-coded reading is implemented.

**Fat uses a multiplication where every other tissue divides.** The listing has
`CVF = CF*PF` against `CVM = CM/PM`, `CVL = CL/PL`, and so on. Fat is 25.8% of
body weight, so this is not cosmetic. As-coded gives about 4% mean absolute
error against Table 2 and the "corrected" `CF/PF` gives about 5%, so the
as-coded reading is implemented.

**`FNC` conflicts between the table and the code.** Rostami 2022 Table 2 gives
0.80 (citing Robinson et al.); both papers' MCSim listings carry 0.72. The
published table value is adopted. This is immaterial to nicotine itself --
`FNC` only splits already-metabolized nicotine between cotinine and other
routes and does not enter nicotine's own kinetics -- but it does shift the
cotinine sub-model by about 10%.

## Scope

**The inhalation (cigarette) comparator arm is not implemented.** The parent
Rostami model reaches the respiratory tract through CFD-derived per-compartment
deposition fractions (`DEPFRACBUC` / `DEPFRACCA` / `DEPFRACTA` / `DEPFRACPU`)
and respiratory mass-transfer coefficients. Every one of these is a `0`
placeholder in *both* papers' published code; Rostami 2022 tabulates none of
them and attributes them to Adrian 2006, Asgharian 2011 and Corley 2012, none
of which is open access, and the deposition fractions are in any case outputs
of an unpublished CFD model. The main text ("90% of the inhaled dose deposited
in the upper respiratory tract") and Figures S1/S2 ("95% ... deposited in lower
respiratory tract") also disagree, and neither gives the four-way split the
code requires. This is a reporting gap that acquisition cannot close, so the
cigarette arm is omitted rather than guessed at. Its one fitted parameter (a
nicotine yield of 2.0 mg/cigarette, Table 2) is reported and is recorded here
for completeness.

**The nicotine mouth-spray validation of Figure 1 is out of scope.** It is the
one case that takes the Equation 3 branch (flux back out of the tissue is
non-negligible), which needs an instantaneous saliva volume `V_Sal`. No
baseline saliva volume is reported anywhere on disk. For the pouches and for
MST the paper takes the `R_rel(t > t_use) = 0` branch, where the GI input
reduces analytically to `(1 - f_tissue(t)) * R_rel(t)` and no saliva volume is
needed. Omitting Figure 1 costs only a validation of `D` and `l`, both of which
are taken as given here.

**The parallel perfused-lung and nasal compartments of the parent model are
dropped.** Their blood flow fractions `QLNGC` and `QNC` are absent from Rostami
2022 Table 1 -- whose flow fractions already sum to 1.0015 without them -- and
are `0` in the published code, which comments the parallel lung compartment as
"only used for lung concentration (blood exchange is cfd)". Both compartments
are therefore mathematically inert. This is the code's own declared value, not
a substitution.

## Numerical choices

**The buccal diffusion PDE is solved by a 20-slab method of lines.** rxode2
solves ODEs only, so Equation 1 is discretised into 20 equal slabs
(`buccal_slab1` ... `buccal_slab20`) with a mass-conserving finite-volume
scheme and a half-cell correction at the perfect-sink face. The slab count is a
numerical choice, not a property of the paper, so it is reported here: AUC is
identical to four decimal places from 10 slabs up to 80, and 20 slabs puts Cmax
within 0.03% of an 80-slab solution. The chain cannot be collapsed to a single
well-mixed compartment -- the diffusion time across the tissue,
`l^2 / D = 0.71 h`, is comparable to the use duration, so the transient spatial
profile is load-bearing.

## Unreported values

- **No between-subject variability and no residual-error model** are reported
  in any on-disk source, so none is encoded. The model is deterministic
  typical-value only. Salehi 2025 does explore inter-individual variability in
  the *extent of nicotine release* by simulation (Figure S10), but reports no
  variance parameters that could be carried into `ini()`.
- **TP2 release fraction per nicotine level.** Reported only as a range, 67-72%
  (SD 28-32%), described as spanning "different nicotine levels and flavors"
  and not resolved per level. 70% is used for all three levels, consistent with
  the paper's own finding that release is not affected by nicotine level.
  Sensitivity across the full 67-72% range moves TP2 AUC by only about 3%, so
  this is not the source of the Equation 6 gap.
- **Nicotine molecular weight** is declared as a `0` placeholder in the
  published MCSim listing (whose `.in` input files were never published) and is
  set here to the physical constant 162.23 g/mol. Cotinine and glucuronide use
  the listing's own hard-coded literals of 178 and 338 g/mol.
- **Body weights** for the alternative-use scenarios are not restated by the
  paper; the study-matched weights from the Figure 2 caption are carried
  forward (92 kg for TP1 and 2 g MST, 97 kg for TP2 and 4 g MST). Using the
  Rostami reference 73 kg instead degrades the MST base cases from about 1% to
  +22 / +28%, which confirms the study-matched reading.
