---
title: "Antibody tumor-to-lymph-node PBPK (Salgado 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Antibody tumor-to-lymph-node PBPK (Salgado 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 6)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Salgado_2022_atezolizumab_bladder_pbpk"))$reference`
* Article: [Antibodies (Basel) 2022;11(2):28](https://doi.org/10.3390/antib11020028) (open access)

Salgado and Cao built a proof-of-concept minimal physiologically based
pharmacokinetic (PBPK) model of how a therapeutic antibody distributes from
plasma into an organ-specific primary tumor (PT) and on to its tumor-draining
lymph node (TDLN). They calibrated the same structural model separately to
eleven antibody / tumor pairs, using immuno-PET standardized uptake values
(SUV) digitized from eight published imaging studies, and then used the
calibrated model to explore metastasis, surgical resection, and
tumor-induced inflammation.

Each calibration is packaged as its own model; all eleven share the same
`model()` block and differ only in the Table 1 / Table 2 parameter values.

```{r models}
models <- tibble::tribble(
  ~panel, ~model,                                          ~label,
  "a",    "Salgado_2022_fresolimumab_glioblastoma_pbpk",   "a) Brain: 89Zr-fresolimumab",
  "b",    "Salgado_2022_bevacizumab_nsclc_pbpk",           "b) Lung: 89Zr-bevacizumab",
  "c",    "Salgado_2022_bevacizumab_breast_pbpk",          "c) Breast: 89Zr-bevacizumab",
  "d",    "Salgado_2022_bevacizumab_renal_pbpk",           "d) Renal: 89Zr-bevacizumab",
  "e",    "Salgado_2022_trastuzumab_breast_pbpk",          "e) Breast: 64Cu-DOTA-trastuzumab",
  "f",    "Salgado_2022_trastuzumab_esophagogastric_pbpk", "f) Esophagogastric: 89Zr-trastuzumab",
  "g",    "Salgado_2022_MMOT0530A_pancreatic_pbpk",        "g) Pancreatic: 89Zr-MMOT0530A",
  "h",    "Salgado_2022_MMOT0530A_ovarian_pbpk",           "h) Ovarian: 89Zr-MMOT0530A",
  "i",    "Salgado_2022_atezolizumab_nsclc_pbpk",          "i) Lung: 89Zr-atezolizumab",
  "j",    "Salgado_2022_atezolizumab_breast_pbpk",         "j) Breast: 89Zr-atezolizumab",
  "k",    "Salgado_2022_atezolizumab_bladder_pbpk",        "k) Bladder: 89Zr-atezolizumab"
)
mods <- lapply(models$model, readModelDb)
names(mods) <- models$panel
knitr::kable(models |> dplyr::rename("Figure 2 panel" = panel, "Model" = model, "Tumor: tracer" = label))
```

# Population

The model was calibrated to published group-mean (+/- SD) SUV time courses,
not to individual patients, so no subject count or demographic table applies.
The eleven calibration data sets (Salgado 2022 Section 2.2 and Figure 2) are
89Zr-fresolimumab in recurrent high-grade glioma (den Hollander 2015);
89Zr-bevacizumab in non-small-cell lung cancer (Bahce 2014), primary breast
cancer (Gaykema 2013) and renal cell carcinoma (Oosting 2015);
64Cu-DOTA-trastuzumab in HER2-positive breast cancer (Tamura 2013);
89Zr-trastuzumab in esophagogastric cancer (O'Donoghue 2018); 89Zr-MMOT0530A
(anti-mesothelin) in pancreatic and ovarian cancer (Lamberts 2016); and
89Zr-atezolizumab (anti-PD-L1) in non-small-cell lung, breast and bladder cancer
(Bensch 2018). Each model's `population` metadata names its source study
(`readModelDb("<model>")()$population`).

# Model structure

The ODEs are Salgado 2022 Appendix A, Eqs. A1-A5. With `C` denoting total
antibody concentration, `Cf` the free antibody concentration, `R_T` the total
target concentration and `AR` the antibody-target complex:

* Plasma (A1): `Vp dCp/dt = Input - (1 - sigma_V) Lorgan Cp - CLp Cp + Leff Cf2 FcRnb2`
* PT interstitium (A2): `VISF,PT dC_PT/dt = (1 - sigma_V) Lorgan Cp - (1 - sigma_L) Laff Cf1 FcRnb1 - kint AR_PT VISF,PT`
* PT target (A3): `dR_T,PT/dt = ksyn,PT - kdeg (R_T,PT - AR_PT) - kint AR_PT`, initial value R01
* TDLN interstitium (A4): `VISF,TDLN dC_TDLN/dt = (1 - sigma_L) Laff Cf1 FcRnb1 - Leff Cf2 FcRnb2 - kint AR_TDLN VISF,TDLN`
* TDLN target (A5): `dR_T,TDLN/dt = ksyn,TDLN - kdeg (R_T,TDLN - AR_TDLN) - kint AR_TDLN`, initial value R02
* Free antibody (A6, A7): `Cf = 0.5 ((C - R_T - Kd) + sqrt((C - R_T - Kd)^2 + 4 C Kd))`
* Complex (A8, A9): `AR = R_T Cf / (Kd + Cf)`
* Lymphatic FcRn salvage (A10): `FcRnb = 0.418 + ([FcRn] / (Cf + Kd,FcRn + [FcRn]))^dln`

The packaged models use `central` (plasma, nmol), `is_tumor` (PT
interstitium, nmol), `lnode` (TDLN interstitium, nmol), and
`total_target_tumor` / `total_target_lnode` (nM). Outputs are the total
antibody concentrations `Cc`, `Ctumor` and `Clnode` (nM). Doses are in nmol.

## Target synthesis: the printed term versus the steady-state term

Eqs. A3 and A5 as printed use the baseline concentration R01 (R02) itself as
the zero-order synthesis term, with the same R01 (R02) as initial condition.
With `kdeg = 0.01 1/h` that is not a steady state: the total target would
climb toward `R01 / kdeg`, 100-fold above baseline. The Figure 1 caption names
the synthesis rate `ksyn`, and the steady-state form `ksyn = kdeg * R01`
reproduces the Figure 2 calibration curves, whereas the literal form does
not. The packaged models therefore use `ksyn = kdeg * R0`. The chunk below
re-solves the pancreatic calibration (Figure 2g) both ways at a 1 nmol dose.

SUV is computed as `C (nM) * 70 L / dose (nmol)` (Appendix F with a 70 L
whole-body volume; see Assumptions).

```{r helpers}
suv_volume <- 70 # L; whole-body volume used to convert nM to SUV (see Assumptions)

# One-subject event table: an IV bolus into central, then observation rows
# keyed by dvid (the model has three declared endpoints, none of them a state).
make_events <- function(dose, times) {
  dplyr::bind_rows(
    data.frame(id = 1L, time = 0, evid = 1L, amt = dose, cmt = "central", dvid = NA_integer_),
    data.frame(id = 1L, time = times, evid = 0L, amt = NA_real_, cmt = NA_character_, dvid = 1L)
  )
}

simulate_suv <- function(mod, dose = 1, times = seq(0, 200, by = 2), params = NULL) {
  sim <- rxode2::rxSolve(mod, events = make_events(dose, times), params = params, useLinCmt = FALSE) |>
    as.data.frame()
  data.frame(
    time = sim$time,
    Plasma = sim$Cc * suv_volume / dose,
    PT = sim$Ctumor * suv_volume / dose,
    TDLN = sim$Clnode * suv_volume / dose
  )
}
```

```{r synthesis-check}
ui_panc <- rxode2::rxode(mods[["g"]])
ui_panc_literal <- ui_panc |>
  rxode2::model(ksyn_tumor <- r0_tumor) |>
  rxode2::model(ksyn_lnode <- r0_lnode)
syn <- dplyr::bind_rows(
  simulate_suv(ui_panc) |> dplyr::mutate(form = "ksyn = kdeg * R0 (packaged)"),
  simulate_suv(ui_panc_literal) |> dplyr::mutate(form = "ksyn = R0 (as printed)")
)
syn_tab <- syn |>
  dplyr::filter(time %in% c(50, 100, 150, 200)) |>
  dplyr::select(form, time, PT, TDLN)
knitr::kable(syn_tab, digits = 2, caption = "Pancreatic calibration (Figure 2g): PT and TDLN SUV under the two synthesis forms. Figure 2g shows the TDLN curve rising to about 6 by 150-200 h.")

tdln_150 <- syn |> dplyr::filter(time == 150)
stopifnot(
  # Figure 2g TDLN reads about 6 at 150 h. The packaged form lands within 20%;
  # the printed form is about 25-fold lower.
  abs(tdln_150$TDLN[tdln_150$form == "ksyn = kdeg * R0 (packaged)"] / 6 - 1) < 0.2,
  tdln_150$TDLN[tdln_150$form == "ksyn = R0 (as printed)"] < 1
)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. All
eleven models share the parameters in the first table; the second table lists
the organ- and antibody-specific values (Salgado 2022 Tables 1 and 2).

| Parameter | Value | Source |
|-----------|-------|--------|
| `sigma_l` (sigma_L) | 0.2 (fixed) | Table 1 footnote (ref 27) |
| `l_aff` (Laff) | 0.004 L/h (fixed) | Table 1 footnote (ref 28) |
| `l_eff` (Leff) | 0.004 L/h (fixed) | Table 1 footnote (ref 28) |
| `lv_lnode` (VISF,TDLN) | log(0.0000584 L) (fixed) | Table 1 footnote (ref 28) |
| `fcrn` ([FcRn]) | 40000 nM = 40 uM (fixed) | Table 1 footnote (ref 29) |
| `f_fcrn_base` (Base) | 0.418 (fixed) | Appendix B (ref 47) |
| `kdeg` | 0.01 1/h (fixed) | Table 2 footnote |
| `kint` | 0.01 1/h (fixed) | Table 2 footnote |
| `propSd`, `propSd_Ctumor`, `propSd_Clnode` | 0 (fixed) | not reported; deterministic model |
| ODEs `central`, `is_tumor`, `total_target_tumor`, `lnode`, `total_target_lnode` | n/a | Appendix A Eqs. A1-A5 (synthesis term: see above) |
| `cf_tumor`, `cf_lnode` | n/a | Appendix A Eqs. A6, A7 |
| `complex_tumor`, `complex_lnode` | n/a | Appendix A Eqs. A8, A9 |
| `fcrnb_tumor`, `fcrnb_lnode` | n/a | Appendix A Eq. A10; Appendix B Eqs. A11-A15 |

```{r source-trace}
param_names <- c(
  "sigma_v", "l_organ", "v_isf_tumor", "lvc", "r0_tumor", "r0_lnode",
  "kd", "kd_fcrn", "dln", "lcl"
)
ptab <- lapply(seq_len(nrow(models)), function(i) {
  ini <- rxode2::rxode(mods[[i]])$iniDf
  est <- stats::setNames(ini$est, ini$name)[param_names]
  data.frame(
    panel = models$panel[i],
    sigma_V = est[["sigma_v"]],
    Lorgan = est[["l_organ"]],
    VISF_PT = est[["v_isf_tumor"]],
    Vp = exp(est[["lvc"]]),
    R01 = est[["r0_tumor"]],
    R02 = est[["r0_lnode"]],
    Kd = est[["kd"]],
    Kd_FcRn = est[["kd_fcrn"]],
    dln = est[["dln"]],
    CLp = exp(est[["lcl"]])
  )
}) |>
  dplyr::bind_rows()
ptab |>
  dplyr::rename(
    "Panel" = panel, "sigma_V" = sigma_V, "Lorgan (L/h)" = Lorgan,
    "VISF,PT (L)" = VISF_PT, "Vp (L)" = Vp, "R01 (nM)" = R01, "R02 (nM)" = R02,
    "Kd (nM)" = Kd, "Kd,FcRn (nM)" = Kd_FcRn, "dln" = dln, "CLp (L/h)" = CLp
  ) |>
  knitr::kable(caption = "Organ- and antibody-specific parameters as packaged (Salgado 2022 Table 1: sigma_V, Lorgan, VISF,PT, Vp; Table 2: R01, R02, Kd, Kd,FcRn, dln, CLp).")
```

# Replicate Figure 2: calibration in eleven tumor / TDLN pairs

The doses given in the source imaging studies are not restated in Salgado
2022. The simulations below use a 1 nmol tracer dose; the model is nearly
dose-proportional in SUV terms at tracer doses (see the dose-sensitivity
table further down), so the SUV curves are insensitive to this choice except
for the high-affinity, small-volume ovarian and renal calibrations.

```{r figure-2, fig.height = 8}
fig2 <- lapply(seq_len(nrow(models)), function(i) {
  simulate_suv(mods[[i]]) |> dplyr::mutate(label = models$label[i], panel = models$panel[i])
}) |>
  dplyr::bind_rows() |>
  tidyr::pivot_longer(c(Plasma, PT, TDLN), names_to = "site", values_to = "SUV")

ggplot(fig2, aes(time, SUV, colour = site)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~label, scales = "free_y", ncol = 4, labeller = label_wrap_gen(24)) +
  scale_colour_manual(values = c(Plasma = "#7f0000", PT = "#f39c12", TDLN = "#0b1a3a")) +
  labs(x = "Time (h)", y = "SUV", colour = NULL, caption = "Replicates Figure 2 of Salgado 2022 (model curves; 1 nmol dose).") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The table compares the simulated maximum SUV over 0-200 h with the maximum of
the model curves in Figure 2, read off the published figure by the
maintainers (readings are approximate, about +/- 0.5 SUV).

```{r figure-2-compare}
fig2_published <- tibble::tribble(
  ~panel, ~PT_pub, ~TDLN_pub,
  "a",     5.0,     4.2,
  "b",     4.6,     3.9,
  "c",     1.7,     2.7,
  "d",    21.0,     6.0,
  "e",    11.2,     8.9,
  "f",    13.8,    11.3,
  "g",    11.2,     6.0,
  "h",    18.0,    16.5,
  "i",    13.4,    11.8,
  "j",     7.0,    11.3,
  "k",    12.8,    10.4
)
fig2_cmp <- fig2 |>
  dplyr::group_by(panel, site) |>
  dplyr::summarise(SUVmax = max(SUV), .groups = "drop") |>
  tidyr::pivot_wider(names_from = site, values_from = SUVmax) |>
  dplyr::left_join(fig2_published, by = "panel") |>
  dplyr::mutate(
    PT_pct = 100 * (PT / PT_pub - 1),
    TDLN_pct = 100 * (TDLN / TDLN_pub - 1)
  )
stopifnot(nrow(fig2_cmp) == 11, !anyNA(fig2_cmp$PT_pct))
fig2_cmp |>
  dplyr::select(panel, PT, PT_pub, PT_pct, TDLN, TDLN_pub, TDLN_pct) |>
  dplyr::rename(
    "Panel" = panel,
    "PT SUVmax (sim)" = PT, "PT SUVmax (Fig. 2)" = PT_pub, "PT diff (%)" = PT_pct,
    "TDLN SUVmax (sim)" = TDLN, "TDLN SUVmax (Fig. 2)" = TDLN_pub, "TDLN diff (%)" = TDLN_pct
  ) |>
  knitr::kable(digits = 1, caption = "Simulated versus published (Figure 2 model curve) maximum SUV over 0-200 h.")

# Deterministic model: these are solver-exact numbers, not cohort statistics.
# Seven calibrations reproduce the published PT curve maximum within 10%
# (measured: -8% to +6%); the gate is set at 15%. Panels f, h, j and k are
# known deviations discussed below and are not gated.
good_pt <- c("a", "b", "c", "d", "e", "g", "i")
stopifnot(all(abs(fig2_cmp$PT_pct[fig2_cmp$panel %in% good_pt]) < 15))
# Plasma SUV at time zero equals the whole-body-to-plasma volume ratio.
p0 <- fig2 |> dplyr::filter(time == 0, site == "Plasma")
stopifnot(all(abs(p0$SUV - suv_volume / ptab$Vp[match(p0$panel, ptab$panel)]) < 1e-6))
```

Agreement is close for the PT in seven of the eleven calibrations and for the
plasma curves throughout (the plasma SUV starts at 70/Vp = 14, or 10 for the
esophagogastric calibration with Vp = 7 L, as in Figure 2). The known
deviations are:

* **Esophagogastric (panel f)**: the simulated PT and TDLN peaks are about
  40% below the published curves at any tracer dose, although the plasma
  curve matches. With VISF,PT = 0.005 L the simulated PT concentration is
  close to the plasma concentration, so a PT SUV of 13-14 is not reachable
  with the printed Table 1 / Table 2 values for this row.
* **Ovarian (panel h)**: the simulated PT and TDLN SUVs depend strongly on
  dose, because the tiny interstitial volume (0.001 L) holds a saturable target
  pool. The published curves (rapid rise to a peak near 15 h, then a decline
  paralleling plasma) resemble the simulation at doses of roughly 50-100 nmol,
  where the target is saturated; the dose used for the calibration is not
  reported.
* **Atezolizumab breast and bladder (panels j, k)** and the TDLN of the
  bevacizumab breast and renal calibrations (panels c, d) run 20-30% below
  the published curves; the PT:TDLN gradient (the quantity the paper studies)
  is preserved.

```{r dose-sensitivity}
dose_sens <- lapply(seq_len(nrow(models)), function(i) {
  lapply(c(1, 10, 100), function(d) {
    s <- simulate_suv(mods[[i]], dose = d)
    data.frame(panel = models$panel[i], dose = d, PT = max(s$PT), TDLN = max(s$TDLN))
  }) |>
    dplyr::bind_rows()
}) |>
  dplyr::bind_rows() |>
  tidyr::pivot_wider(names_from = dose, values_from = c(PT, TDLN), names_sep = " SUVmax, dose (nmol) ")
dose_sens |>
  dplyr::rename("Panel" = panel) |>
  knitr::kable(digits = 1, caption = "Maximum PT and TDLN SUV over 0-200 h at 1, 10 and 100 nmol.")
```

# Replicate Figure 4: surgical resection

Figure 4 contrasts the TDLN SUV before surgery (neoadjuvant setting) with the
residual TDLN after resection of the primary tumor (adjuvant setting).
Appendix D (Table A1) parameterizes the resected case on the atezolizumab
breast calibration with R01 = 0.001 nM, R02 = 2.5 nM and dln = 5, and the text
reduces the local lymph flows (Lorgan and Laff) by 50% and 80%.

```{r figure-4}
brA <- mods[["j"]]
brA_ini <- stats::setNames(rxode2::rxode(brA)$iniDf$est, rxode2::rxode(brA)$iniDf$name)
resected <- c(r0_tumor = 0.001, r0_lnode = 2.5, dln = 5)
fig4 <- dplyr::bind_rows(
  simulate_suv(brA) |> dplyr::mutate(scenario = "Neoadjuvant (Table 2 breast, atezolizumab)"),
  simulate_suv(brA, params = c(resected, l_organ = 0.5 * brA_ini[["l_organ"]], l_aff = 0.5 * brA_ini[["l_aff"]])) |>
    dplyr::mutate(scenario = "Adjuvant, lymph flows -50%"),
  simulate_suv(brA, params = c(resected, l_organ = 0.2 * brA_ini[["l_organ"]], l_aff = 0.2 * brA_ini[["l_aff"]])) |>
    dplyr::mutate(scenario = "Adjuvant, lymph flows -80%")
)
ggplot(fig4, aes(time, TDLN, linetype = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (h)", y = "TDLN SUV", linetype = NULL, caption = "Replicates Figure 4 of Salgado 2022.") +
  theme_bw() +
  theme(legend.position = "bottom", legend.direction = "vertical")

fig4_200 <- fig4 |>
  dplyr::filter(time == 200) |>
  dplyr::select(scenario, TDLN) |>
  dplyr::mutate(Fig4 = c(10.5, 8.8, 4.0))
fig4_200 |>
  dplyr::rename("Scenario" = scenario, "TDLN SUV at 200 h (sim)" = TDLN, "TDLN SUV at 200 h (Fig. 4, read)" = Fig4) |>
  knitr::kable(digits = 1)
stopifnot(
  fig4_200$TDLN[1] > fig4_200$TDLN[2],
  fig4_200$TDLN[2] > fig4_200$TDLN[3]
)
```

The simulation reproduces the direction and ordering of Figure 4 (surgery
lowers TDLN exposure, more so for the larger lymph-flow loss) but predicts a
larger reduction than the figure shows (about 30% and 80% versus about 15% and
60% read from the figure). The paper does not state which parameter set the
neoadjuvant curve used or whether Leff was also reduced, so the published
magnitudes could not be pinned down.

# Replicate Figure 5: peri- and intra-tumoral TDLNs

Appendix E (Tables A2 and A3) defines three TDLN networks on the 64Cu-DOTA-
trastuzumab breast calibration: a peritumoral node with ten-fold increased
afferent lymph flow, a tumor-positive (+) intratumoral node with collapsed
lymphatics, and a tumor-negative (-) intratumoral node with no target (R02 = 0).
The values are used exactly as printed, including the CLp column.

```{r figure-5}
brT <- mods[["e"]]
appE_common <- c(sigma_v = 0.87, l_organ = 0.008, v_isf_tumor = 0.1122, lvc = log(5),
                 r0_tumor = 500, kd = 5, kd_fcrn = 2400, dln = 5)
fig5 <- dplyr::bind_rows(
  simulate_suv(brT, params = c(appE_common, l_aff = 0.04, l_eff = 0.004, r0_lnode = 100, lcl = log(0.004))) |>
    dplyr::mutate(scenario = "Peritumoral"),
  simulate_suv(brT, params = c(appE_common, l_aff = 0.00001, l_eff = 0.00001, r0_lnode = 100, lcl = log(0.00001))) |>
    dplyr::mutate(scenario = "(+)-Intratumoral"),
  simulate_suv(brT, params = c(appE_common, l_aff = 0.004, l_eff = 0.004, r0_lnode = 0, lcl = log(0.004))) |>
    dplyr::mutate(scenario = "(-)-Intratumoral")
)
ggplot(fig5[fig5$TDLN > 0, ], aes(time, TDLN, linetype = scenario)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "TDLN SUV", linetype = NULL, caption = "Replicates Figure 5 of Salgado 2022.") +
  theme_bw() +
  theme(legend.position = "bottom")

fig5_200 <- fig5 |> dplyr::filter(time == 200) |> dplyr::select(scenario, TDLN)
knitr::kable(fig5_200 |> dplyr::rename("Scenario" = scenario, "TDLN SUV at 200 h" = TDLN), digits = 3)
peri <- fig5_200$TDLN[fig5_200$scenario == "Peritumoral"]
pos <- fig5_200$TDLN[fig5_200$scenario == "(+)-Intratumoral"]
neg <- fig5_200$TDLN[fig5_200$scenario == "(-)-Intratumoral"]
# Figure 5 (log scale) separates the three networks by roughly an order of
# magnitude each: peritumoral about 8-15, (+) about 0.6, (-) about 0.04-0.08.
stopifnot(pos / peri < 0.1, neg / pos < 0.2, pos > 0.3, pos < 1.5)
```

Figure 5 is reproduced in ordering and order of magnitude: the (+)
intratumoral node settles near SUV 0.7 and the (-) intratumoral node below
0.1, as in the figure. The published peritumoral curve rises faster (to about
15 within a day) than the simulation does.

# Figure 6: sensitivity of the PT profile to key parameters

Figure 6 varies R01/R02, dln, Kd, kint and kdeg from 0.1- to 100-fold. The
paper does not state which calibration underlies the figure, so the
atezolizumab bladder calibration (panel k) is used here to check the direction
of each effect stated in Section 3.5: the PT profile rises with R01/R02 and
dln, falls with Kd and kint, and is insensitive to kdeg.

```{r figure-6, fig.height = 7}
bl <- mods[["k"]]
bl_ini <- stats::setNames(rxode2::rxode(bl)$iniDf$est, rxode2::rxode(bl)$iniDf$name)
folds <- c(0.1, 0.3, 0.5, 0.7, 1, 3, 5, 7, 10, 50, 100)
sens_par <- c("r0_tumor", "dln", "kd", "kint", "kdeg")
fig6 <- lapply(sens_par, function(p) {
  lapply(folds, function(f) {
    pp <- stats::setNames(f * bl_ini[[p]], p)
    simulate_suv(bl, params = pp) |> dplyr::mutate(parameter = p, fold = f)
  }) |>
    dplyr::bind_rows()
}) |>
  dplyr::bind_rows()
ggplot(fig6, aes(time, PT, group = fold, colour = log10(fold))) +
  geom_line() +
  facet_wrap(~parameter, scales = "free_y") +
  scale_colour_viridis_c(name = "log10(fold)") +
  labs(x = "Time (h)", y = "PT SUV", caption = "Analogue of Figure 6 of Salgado 2022 on the bladder calibration.") +
  theme_bw()

pt200 <- fig6 |> dplyr::filter(time == 200)
at <- function(p, f) pt200$PT[pt200$parameter == p & pt200$fold == f]
sens_tab <- data.frame(
  parameter = sens_par,
  fold_0.1 = vapply(sens_par, at, numeric(1), f = 0.1),
  fold_1 = vapply(sens_par, at, numeric(1), f = 1),
  fold_10 = vapply(sens_par, at, numeric(1), f = 10)
)
sens_tab |>
  dplyr::rename("Parameter" = parameter, "PT SUV 200 h, 0.1x" = fold_0.1, "1x" = fold_1, "10x" = fold_10) |>
  knitr::kable(digits = 2, row.names = FALSE)
stopifnot(
  at("r0_tumor", 10) > at("r0_tumor", 1), at("r0_tumor", 1) > at("r0_tumor", 0.1),
  at("dln", 10) > at("dln", 1),
  at("kd", 10) < at("kd", 1), at("kd", 1) < at("kd", 0.1),
  at("kint", 10) < at("kint", 1), at("kint", 1) < at("kint", 0.1),
  # kdeg: Section 3.5 calls it 'not a sensitive model parameter'
  abs(at("kdeg", 10) / at("kdeg", 0.1) - 1) < 0.25
)
```

# PKNCA: plasma disposition

Plasma is a single pool cleared by CLp, with a fraction of the dose diverted
into the tumor and TDLN, where some of it is internalized with its target
rather than returning to plasma. The plasma AUC therefore sits at or below
Dose / CLp; the shortfall is largest (about 20%) for the atezolizumab lung
calibration, whose low CLp lets the tumor influx compete with systemic
clearance. For most calibrations the terminal half-life is close to
ln(2) Vp / CLp. Two exceptions follow from the structure: in the brain
calibration (panel a) the large tumor interstitium (0.265 L) drains back to
plasma slowly through the lymph and sets a longer terminal phase, and in the
atezolizumab lung calibration (panel i) internalization in the tumor adds to
CLp and shortens it.

```{r pknca}
nca_times <- c(0, 1, 2, 4, 8, 12, 24, 48, 72, 96, 120, 168, 240, 336, 504, 672, 1008, 1344, 2016)
pk <- lapply(seq_len(nrow(models)), function(i) {
  ev <- make_events(1, nca_times)
  sim <- rxode2::rxSolve(mods[[i]], events = ev, useLinCmt = FALSE) |> as.data.frame()
  data.frame(id = i, panel = models$panel[i], time = sim$time, Cc = sim$Cc)
}) |>
  dplyr::bind_rows()
stopifnot(!anyNA(pk$Cc), all(pk$Cc > 0))
dose_df <- data.frame(id = seq_len(nrow(models)), panel = models$panel, time = 0, amt = 1)

conc_obj <- PKNCA::PKNCAconc(pk |> dplyr::filter(!is.na(Cc)), Cc ~ time | panel + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | panel + id, route = "intravascular")
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, aucinf.obs = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tab <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  dplyr::select(panel, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(ptab |> dplyr::select(panel, Vp, CLp), by = "panel") |>
  dplyr::mutate(
    auc_ratio = aucinf.obs * CLp / 1,
    thalf_closed = log(2) * Vp / CLp
  )
nca_tab |>
  dplyr::select(panel, cmax, aucinf.obs, auc_ratio, half.life, thalf_closed) |>
  dplyr::rename(
    "Panel" = panel, "Cmax (nM)" = cmax, "AUC0-inf (nM*h)" = aucinf.obs,
    "AUC x CLp / Dose" = auc_ratio, "t1/2 (h, PKNCA)" = half.life,
    "ln(2) Vp / CLp (h)" = thalf_closed
  ) |>
  knitr::kable(digits = 3, caption = "PKNCA plasma NCA after a 1 nmol IV dose.")
stopifnot(
  # Cmax at time zero is Dose / Vp exactly.
  all(abs(nca_tab$cmax - 1 / nca_tab$Vp) < 1e-6),
  # AUC x CLp / Dose is the fraction of the dose eliminated by CLp; the rest
  # is internalized in the tumor and TDLN, so the ratio cannot exceed 1. A
  # 1-h-step trapezoid to 20000 h gives 0.79 (atezolizumab NSCLC: low CLp and
  # a large tumor influx) to 1.00 (ovarian, esophagogastric).
  all(nca_tab$auc_ratio < 1.01), all(nca_tab$auc_ratio > 0.7)
)
```

No NCA results are published for these immuno-PET calibrations, so there is
no published-versus-simulated NCA table.

# Assumptions and deviations

* **Target synthesis term.** Eqs. A3 and A5 print the baseline target
  concentration (R01, R02) as the synthesis rate. The packaged models use
  `ksyn = kdeg * R0`, which keeps the target at its baseline before dosing and
  reproduces Figure 2; the printed form drives total target up to 100-fold over
  baseline and misses Figure 2 (shown above for the pancreatic calibration).
* **SUV conversion.** Appendix F defines SUV as concentration at the target
  site times whole-body volume divided by the injected dose, without stating
  the whole-body volume. A volume of 70 L reproduces the published plasma SUV
  at time zero (about 14 for Vp = 5 L and about 10 for Vp = 7 L). The tumor and
  TDLN SUVs use the interstitial-fluid concentration of total (free plus
  bound) antibody, which reproduces the published PT curves.
* **Dose.** The calibration doses of the source imaging studies are not given
  in Salgado 2022. The figures use a 1 nmol tracer dose; the dose-sensitivity
  table shows where this matters.
* **Free-antibody root.** Eqs. A6 and A7 are coded in the algebraically
  identical rationalized form when `C - R_T - Kd < 0`, to avoid round-off
  near time zero; the values are unchanged.
* **Fixed versus calibrated parameters.** The paper lists sigma_V, R01, R02,
  Kd, dln and CLp as the organ- and antibody-specific calibrated parameters
  but cites literature sources for Kd and Kd,FcRn. sigma_V, R01, R02, dln and
  CLp are therefore left unfixed and every literature or physiological value
  (including Kd) is wrapped in `fixed()`. No standard errors were reported.
* **No variability.** The model was calibrated to group means; there is no
  between-subject or residual variability, and the residual-error parameters
  are fixed to zero.
* **Single channel.** Eq. A1 sums over tumor / TDLN channels when metastases
  are added. The lung-metastasis channels of Figure 3a-b and the target
  densities of Figure 3c-d are not reported, so only the single-channel
  (primary tumor) models are packaged and Figure 3 is not reproduced.
* **Appendix E CLp column.** Table A3 prints CLp = 0.004, 0.00001 and 0.004
  L/h for the three TDLN networks, which equal the Laff values of Table A2 and
  differ from the 0.063 L/h of the trastuzumab breast calibration. They are
  used as printed.
* **Literature check.** No correction notice for Salgado 2022 was found in
  Europe PMC or in the Crossref record of the DOI as of 2026-10-01.
