---
title: "Rat neuro-PBPK for six drugs from three BBB cell lines (Sanchez-Dengra 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rat neuro-PBPK for six drugs from three BBB cell lines (Sanchez-Dengra 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The paper

Sanchez-Dengra and colleagues coupled a single in vitro blood-brain barrier
(BBB) assay to a small semi-physiological PBPK model to predict drug
concentrations in the rat brain. Three cell lines were used for the in vitro
assay -- MDCK, MDCK-MDR1 (MDCK transfected with human P-glycoprotein) and the
human brain endothelial line hCMEC/D3 -- and six model drugs were studied:
amitriptyline, caffeine, carbamazepine, fleroxacin, pefloxacin and zolpidem.

> Sanchez-Dengra B, Gonzalez-Alvarez I, Bermejo M, Gonzalez-Alvarez M.
> Physiologically Based Pharmacokinetic (PBPK) Modeling for Predicting Brain
> Levels of Drug in Rat. *Pharmaceutics.* 2021;13(9):1402.
> [doi:10.3390/pharmaceutics13091402](https://doi.org/10.3390/pharmaceutics13091402)

* Article (open access): <https://doi.org/10.3390/pharmaceutics13091402>
* Supplement (Tables S1-S2): <https://www.mdpi.com/article/10.3390/pharmaceutics13091402/s1>

For every drug and every cell line the authors fitted the model to published
mean plasma and brain profiles in Berkeley Madonna, estimating the plasma
volume of distribution `Vd`, the elimination rate constant `kel`, the
absorption rate constant `ka` (extravascular doses only) and three scaling
factors that carry the in vitro inputs over to the rat (Table 3). The paper
therefore reports **18 fitted parameterisations** (six drugs times three cell
lines) of one shared structure. Each is a separate model in this package:

```{r models}
drugs <- c("amitriptyline", "caffeine", "carbamazepine", "fleroxacin", "pefloxacin", "zolpidem")
cells <- c(mdck = "MDCK", mdckmdr1 = "MDCK-MDR1", hcmec = "hCMEC/D3")
grid <- expand.grid(cell = names(cells), drug = drugs, stringsAsFactors = FALSE)
grid$model <- sprintf("SanchezDengra_2021_%s_rat_pbpk_%s", grid$drug, grid$cell)
grid$cell_label <- factor(unname(cells[grid$cell]), levels = cells)
stopifnot(nrow(grid) == 18L, all(grid$model %in% modeldb$name))
mods <- lapply(setNames(grid$model, grid$model), function(m) rxode2::rxode(readModelDb(m)))
grid[, c("model", "drug", "cell_label")]
```

The models are deterministic: the paper fitted mean profiles and estimated no
between-animal variability and no residual error, so there are no `eta` or
residual-error terms, and every simulation below is a single typical-value
solve.

## Model structure

Figure 1 of the paper has one plasma compartment (`central`, volume `Vd`) and
two CNS compartments: brain tissue (`brain`, volume `Vb`) and cerebrospinal
fluid (`csf`, volume `VCSF`). Only unbound drug crosses a barrier. For total
concentrations `Cp` (plasma) and `Cb` (brain) and the CSF concentration
`CCSF`, Equations 4-6 are

```
Vd   dCp/dt   = - PS_BBB,in Cu,p + PS_BBB,out Cu,b - PS_BCSFB,in Cu,p
                + PS_BCSFB,out CCSF + Qsink CCSF - kel Cp Vd
Vb   dCb/dt   =   PS_BBB,in Cu,p - PS_BBB,out Cu,b - Qbulk Cu,b
VCSF dCCSF/dt =   PS_BCSFB,in Cu,p - PS_BCSFB,out CCSF - Qsink CCSF + Qbulk Cu,b
```

with the unbound concentrations `Cu,p = fu,plasma Cp` (Figure 1) and
`Cu,b = SC3 fu,brain Cb` (Equation 14), all CSF drug unbound, and the barrier
permeability-surface-area products built from the in vitro apparent
permeabilities (Equations 10-13):

```
PS_BBB,in    = SC1 Papp,A-B S_BBB      PS_BCSFB,in  = SC1 Papp,A-B S_BCSFB
PS_BBB,out   = SC2 Papp,B-A S_BBB      PS_BCSFB,out = SC2 Papp,B-A S_BCSFB
```

An extravascular dose adds a depot `dA/dt = -ka A` feeding plasma at `ka A`
(Equations 7-8); an intravenous infusion adds the zero-order input `k0`
(Equation 9), which the package models supply as an infusion event rather
than as a term in the ODE. Nothing is eliminated from the brain or the CSF:
every molecule that enters the CNS returns to plasma, and `kel` is the only
exit from the system.

## Population

The in vivo data are **literature mean profiles** (Methods 2.4, references
16-20), one rat study per drug, so the paper reports no number of animals. The
rat weights of the source studies were 190-300 g (Table 1). The brain data are
total brain concentrations for amitriptyline and zolpidem and unbound brain
(or brain extracellular fluid) concentrations for the other four drugs; the
plasma data are total plasma concentrations except for carbamazepine, whose
plasma profile is unbound (Figure 2 legends). The same information is in each
model's `population` metadata:

```{r population}
str(readModelDb("SanchezDengra_2021_caffeine_rat_pbpk_mdck")()$population)
```

## Source trace

Every `ini()` value carries an in-file comment naming its source. They are
collected here.

| Quantity | Symbol in model | Source |
|---|---|---|
| Model structure | `d/dt(central)`, `d/dt(brain)`, `d/dt(csf)` | Figure 1; Equations 4-6 |
| Extravascular input | `d/dt(depot)`, `ka` | Equations 7-8 |
| Infusion input | infusion event into `central` | Equation 9 |
| Barrier PS products | `ps_bbb_in`, `ps_bbb_out`, `ps_bcsfb_in`, `ps_bcsfb_out` | Equations 10-13 |
| Unbound brain concentration | `Cu_brain <- sc3 * fu_brain * Cbrain` | Equation 14 |
| Unbound plasma concentration | `Cu <- fu * Cc` | Figure 1 |
| `Vd`, `kel`, `ka` (fitted) | `lvc`, `lkel`, `lka` | Table 3 (Table 1 holds the initial estimates) |
| `SC1`, `SC2`, `SC3` per cell line (fitted) | `sc1`, `sc2`, `sc3` | Table 3 |
| `Papp,A-B`, `Papp,B-A`, `fu,brain` per cell line (fixed) | `papp_ab`, `papp_ba`, `fu_brain` | Table 2 |
| `fu,plasma` (fixed) | `fu` | Table 1 |
| `Vb` = 1.28 cm^3, `VCSF` = 0.25 cm^3 | `lvbrain`, `lvcsf` | Methods 2.4 (Ball et al., ref 14) |
| `Qbulk` = 0.012 cm^3/s, `Qsink` = 0.132 cm^3/s | `qbulk`, `qsink` | Methods 2.4 (Ball et al., ref 14) |
| `S_BBB` = 187.5 cm^2, `S_BCSFB` = 0.0375 cm^2 | `s_bbb`, `s_bcsfb` | Methods 2.4 (ref 14; Engelhard et al., ref 15) |
| Doses `D`, infusion rates `k0`, rat weights | event tables below | Table 1 |
| QSPR polynomials `lnSC = f(logP)` | vignette only | Figure 3 (printed equations) |
| logP of each drug | vignette only | Supplementary Table S2 |

The eighteen files differ only in the Table 1-3 values. The parameter values
of all eighteen, read back out of the packaged models:

```{r param-table}
param_row <- function(m) {
  ini <- mods[[m]]$iniDf
  v <- setNames(ini$est, ini$name)
  data.frame(
    model = m,
    Vd_mL = exp(v[["lvc"]]),
    kel_per_s = signif(exp(v[["lkel"]]), 3),
    ka_per_s = if ("lka" %in% names(v)) signif(exp(v[["lka"]]), 3) else NA_real_,
    fu_plasma = v[["fu"]],
    Papp_AB = v[["papp_ab"]],
    Papp_BA = v[["papp_ba"]],
    fu_brain = v[["fu_brain"]],
    SC1 = v[["sc1"]],
    SC2 = v[["sc2"]],
    SC3 = v[["sc3"]]
  )
}
params <- do.call(rbind, lapply(grid$model, param_row))
knitr::kable(params, row.names = FALSE)
```

## Dosing designs (Table 1)

Table 1 gives, per drug, a dose `D` and/or an infusion rate `k0`. Read
together with Figure 2 the designs are:

* **Amitriptyline** and **carbamazepine** -- a single extravascular dose `D`
  (5,000,000 ng and 3,600,000 ng) with first-order absorption `ka`.
* **Zolpidem** -- a single intravenous bolus `D` = 499,700 ng; `D / Vd` =
  2688 ng/mL is the fitted plasma line's value at time zero in Figure 2F.
* **Fleroxacin** and **pefloxacin** -- an intravenous loading bolus `D`
  followed by a constant infusion `k0` for the whole 14,400 s record. Both
  `D` and `k0` are printed; that `D` is a bolus at time zero is confirmed by
  `D / Vd` = 3405 and 7012 ng/mL, the fitted plasma lines' starting values in
  Figure 2D and 2E.
* **Caffeine** -- a constant infusion `k0` = 833.333 ng/s with no bolus. The
  paper does not print the infusion duration; it is taken as 14,400 s (4 h),
  where the fitted plasma and brain lines of Figure 2B peak.

```{r designs}
design <- data.frame(
  drug = drugs,
  bolus_amt = c(5e6, 0, 3.6e6, 1114350, 3676500, 499700),
  bolus_cmt = c("depot", "central", "depot", "central", "central", "central"),
  inf_rate = c(0, 833.333, 0, 83.125, 214.542, 0),
  inf_dur = c(0, 14400, 0, 14400, 14400, 0),
  t_end = c(86400, 28800, 50400, 14400, 14400, 21600),
  stringsAsFactors = FALSE
)
# Which concentrations Figure 2 plots for each drug (legends of Figure 2)
design$plasma_var <- c("Cc", "Cc", "Cu", "Cc", "Cc", "Cc")
design$brain_var <- c("Cbrain", "Cu_brain", "Cu_brain", "Cu_brain", "Cu_brain", "Cbrain")
knitr::kable(design)

make_events <- function(drug, obs_times) {
  d <- design[design$drug == drug, ]
  ev <- rxode2::et(obs_times, cmt = "central")
  if (d$bolus_amt > 0) {
    ev <- rxode2::et(ev, amt = d$bolus_amt, cmt = d$bolus_cmt, time = 0)
  }
  if (d$inf_rate > 0) {
    ev <- rxode2::et(ev, amt = d$inf_rate * d$inf_dur, rate = d$inf_rate, cmt = "central", time = 0)
  }
  ev
}

solve_one <- function(m, drug, obs_times, ...) {
  s <- as.data.frame(rxode2::rxSolve(mods[[m]], make_events(drug, obs_times), ...))
  s$model <- m
  s
}
```

## Replicating Figure 2: the fitted profiles

Each fitted parameterisation is solved on the design of its source study. The
panels plot the same quantities as Figure 2 of the paper (total or unbound, as
in its legends).

```{r fig2-sim}
fig2 <- lapply(seq_len(nrow(grid)), function(i) {
  d <- design[design$drug == grid$drug[i], ]
  obs <- sort(unique(c(seq(0, d$t_end, length.out = 401), 1:120)))
  obs <- obs[obs <= d$t_end]
  s <- solve_one(grid$model[i], grid$drug[i], obs)
  data.frame(
    drug = grid$drug[i], cell = grid$cell_label[i], time = s$time,
    plasma = s[[d$plasma_var]], brain = s[[d$brain_var]]
  )
})
fig2 <- do.call(rbind, fig2)
fig2_long <- fig2 |>
  pivot_longer(c(plasma, brain), names_to = "matrix", values_to = "conc") |>
  mutate(matrix = factor(matrix, levels = c("plasma", "brain")))
```

```{r fig2-plot, fig.height = 9, fig.cap = "Replicates Figure 2 of Sanchez-Dengra 2021: fitted plasma and brain profiles for each drug and cell line (the three cell-line fits overlap, as in the paper)."}
ggplot(fig2_long, aes(time, conc, colour = cell)) +
  geom_line() +
  facet_wrap(drug ~ matrix, scales = "free", ncol = 4) +
  labs(x = "t (s)", y = "C (ng/mL)", colour = "Cell line") +
  theme_bw() +
  theme(legend.position = "bottom")
```

The fitted lines of Figure 2 were read by eye at the points below (fitted-line
peaks, or the start and end of the constant-infusion records). They are
visual reads of a printed figure, so the check allows 15 percent -- enough for
the reading error, and far too little for a wrong dose, volume or unit, which
move these values several-fold. The check uses the MDCK fit, the cell line the
paper recommends; the next check covers the other two.

```{r fig2-anchors}
anchor <- data.frame(
  drug = c(
    "amitriptyline", "amitriptyline", "caffeine", "caffeine", "carbamazepine",
    "carbamazepine", "fleroxacin", "fleroxacin", "pefloxacin", "pefloxacin",
    "zolpidem", "zolpidem"
  ),
  matrix = c(
    "plasma", "brain", "plasma", "brain", "plasma", "brain",
    "plasma", "brain", "plasma", "brain", "plasma", "brain"
  ),
  what = c(
    "peak", "peak", "peak", "peak", "peak", "peak",
    "end", "end", "end", "end", "start", "peak"
  ),
  figure2 = c(310, 6000, 34000, 30500, 1050, 490, 5400, 1380, 6400, 2000, 2700, 880)
)
sim_at <- function(drug, matrix, what) {
  x <- fig2_long[fig2_long$drug == drug & fig2_long$matrix == matrix & fig2_long$cell == "MDCK", ]
  if (nrow(x) == 0L) stop("no simulated rows for ", drug, " ", matrix)
  switch(what,
    peak = max(x$conc),
    start = x$conc[which.min(x$time)],
    end = x$conc[which.max(x$time)]
  )
}
anchor$simulated <- mapply(sim_at, anchor$drug, anchor$matrix, anchor$what)
anchor$pct_diff <- round(100 * (anchor$simulated / anchor$figure2 - 1), 1)
anchor$simulated <- signif(anchor$simulated, 4)
knitr::kable(anchor)
stopifnot(nrow(anchor) == 12L, all(abs(anchor$pct_diff) < 15))
```

The paper states that the three cell-line fits overlap in both plasma and brain
(Figure 2 and Discussion). Brain exposure of the MDCK-MDR1 and hCMEC/D3 fits,
relative to the MDCK fit of the same drug:

```{r fig2-overlap}
overlap <- fig2 |>
  group_by(drug, cell) |>
  summarise(
    brain_auc = sum(diff(time) * (head(brain, -1) + tail(brain, -1)) / 2),
    brain_cmax = max(brain),
    .groups = "drop"
  ) |>
  group_by(drug) |>
  mutate(
    auc_vs_mdck_pct = round(100 * (brain_auc / brain_auc[cell == "MDCK"] - 1), 1),
    cmax_vs_mdck_pct = round(100 * (brain_cmax / brain_cmax[cell == "MDCK"] - 1), 1)
  ) |>
  ungroup()
overlap$deviation <- overlap$drug == "amitriptyline" & overlap$cell != "MDCK"
knitr::kable(overlap |> select(drug, cell, auc_vs_mdck_pct, cmax_vs_mdck_pct, deviation))
# Measured within 2.0 percent for every row not flagged as a deviation.
stopifnot(nrow(overlap) == 18L, sum(overlap$deviation) == 2L, all(abs(overlap$auc_vs_mdck_pct[!overlap$deviation]) < 5))
```

The fits overlap for five of the six drugs. For **amitriptyline** the
MDCK-MDR1 and hCMEC/D3 total brain profiles lie 11 and 27 percent below the
MDCK one, whereas all three overlap in Figure 2A. This is a known deviation,
and it comes from rounding in Table 3. Amitriptyline is extensively bound in
brain, and the fitted `SC3` values are printed to two decimals as 0.05, 0.02
and 0.01. The unbound brain concentration is fixed by the barrier
permeabilities, so the three fits agree on it (checked below). The total
concentration is the unbound one divided by `SC3 * fu,brain`, which as printed
is 0.00185, 0.00208 and 0.00252 for the three cell lines. For the total
profiles to overlap as they do in the paper, the unrounded hCMEC/D3 `SC3` must
be about 0.0073, not 0.01. The Figure 3C point for amitriptyline sits at
`ln SC3` of about -4.9 (`SC3` of about 0.0075), which agrees. The packaged
models keep the printed 0.02 and 0.01. The check below confirms that this is
the whole explanation: the unbound profiles overlap, and the ratio of the
total-brain AUCs equals the ratio of the printed `SC3 * fu,brain` products.

```{r fig2-amitriptyline}
ami <- lapply(grid$model[grid$drug == "amitriptyline"], function(m) {
  s <- solve_one(m, "amitriptyline", seq(0, 86400, by = 60))
  v <- params[params$model == m, ]
  data.frame(
    model = m,
    auc_unbound = sum(diff(s$time) * (head(s$Cu_brain, -1) + tail(s$Cu_brain, -1)) / 2),
    auc_total = sum(diff(s$time) * (head(s$Cbrain, -1) + tail(s$Cbrain, -1)) / 2),
    binding = v$SC3 * v$fu_brain
  )
})
ami <- do.call(rbind, ami)
ami$unbound_vs_mdck <- ami$auc_unbound / ami$auc_unbound[1]
ami$total_vs_mdck <- ami$auc_total / ami$auc_total[1]
ami$binding_ratio <- ami$binding[1] / ami$binding
knitr::kable(ami, digits = 4)
stopifnot(
  nrow(ami) == 3L, grepl("_mdck$", ami$model[1]),
  all(abs(ami$unbound_vs_mdck - 1) < 0.02),
  all(abs(ami$total_vs_mdck / ami$binding_ratio - 1) < 0.02)
)
```

## Structural identities

Two properties of the structure hold exactly for every parameterisation, and
test the ODEs and the unit handling independently of any figure.

### Plasma AUC and the dose

Because the CNS returns everything to plasma, the total dose leaves the system
only through `kel`, so `kel * Vd * AUC(0-inf, total plasma) = total dose`
whatever the CNS parameters are. A mis-signed exchange term or a
mis-scaled permeability would break it. The AUC is computed with PKNCA on a
long, dense grid of each drug's Table 1 design, grouped by drug and cell line.

```{r auc-identity}
ident <- lapply(seq_len(nrow(grid)), function(i) {
  ini <- mods[[grid$model[i]]]$iniDf
  kel <- exp(ini$est[ini$name == "lkel"])
  obs <- sort(unique(c(0, 10^seq(0, log10(40 * log(2) / kel), length.out = 800))))
  s <- solve_one(grid$model[i], grid$drug[i], obs, rtol = 1e-10, atol = 1e-10, maxsteps = 1e6)
  data.frame(treatment = paste(grid$drug[i], grid$cell[i], sep = "_"), id = 1L, time = s$time, Cc = s$Cc)
})
ident <- do.call(rbind, ident)
stopifnot(all(ident$Cc >= -1e-6 * max(ident$Cc)))
conc_ident <- ident |>
  filter(!is.na(Cc)) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(treatment, id, time, Cc)
dose_ident <- design |>
  mutate(total_dose = bolus_amt + inf_rate * inf_dur) |>
  select(drug, total_dose) |>
  right_join(grid |> select(drug, cell), by = "drug") |>
  mutate(treatment = paste(drug, cell, sep = "_"), id = 1L, time = 0)
o_conc <- PKNCAconc(conc_ident, Cc ~ time | treatment + id)
o_dose <- PKNCAdose(dose_ident, total_dose ~ time | treatment + id)
o_data <- PKNCAdata(o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE),
  options = list(auc.method = "lin up/log down")
)
nca_ident <- as.data.frame(pk.nca(o_data))
auc_check <- nca_ident |>
  filter(PPTESTCD == "aucinf.obs") |>
  left_join(dose_ident |> select(treatment, total_dose, drug, cell), by = "treatment") |>
  left_join(params |> mutate(treatment = sub("^SanchezDengra_2021_(.*)_rat_pbpk_(.*)$", "\\1_\\2", model)),
    by = "treatment"
  ) |>
  mutate(ratio = kel_per_s * Vd_mL * PPORRES / total_dose)
knitr::kable(auc_check |> select(treatment, total_dose, aucinf = PPORRES, ratio), digits = 4)
# The ratio is 1 up to the trapezoid and extrapolation error of the NCA
# (measured at most 1.1e-5 on this grid); a sign error in any exchange term
# or a 1e-6 unit slip moves it by far more than 0.5 percent.
stopifnot(nrow(auc_check) == 18L, all(abs(auc_check$ratio - 1) < 0.005))
```

`kel_per_s` above comes from `signif(..., 3)`, which equals the printed Table 3
value exactly.

### Steady-state brain partitioning

At steady state Equation 5 gives the unbound brain-to-plasma ratio directly:
`Kp,uu,brain = Cu,b / Cu,p = PS_BBB,in / (PS_BBB,out + Qbulk)`, and Equation 6
then gives `CCSF / Cu,p = (PS_BCSFB,in + Qbulk Kp,uu) / (PS_BCSFB,out + Qsink)`.
A long constant infusion drives each parameterisation to steady state, and the
simulated ratios are compared with the closed forms.

```{r kpuu}
kp <- lapply(seq_len(nrow(grid)), function(i) {
  v <- params[params$model == grid$model[i], ]
  kel <- v$kel_per_s
  t_ss <- 60 * log(2) / kel
  ev <- rxode2::et(amt = t_ss, rate = 1, cmt = "central", time = 0)
  ev <- rxode2::et(ev, t_ss * (1 - 1e-6), cmt = "central")
  s <- as.data.frame(rxode2::rxSolve(mods[[grid$model[i]]], ev, rtol = 1e-10, atol = 1e-12, maxsteps = 1e6))
  s <- s[nrow(s), ]
  ps_in <- v$SC1 * v$Papp_AB * 1e-6 * 187.5
  ps_out <- v$SC2 * v$Papp_BA * 1e-6 * 187.5
  kpuu <- ps_in / (ps_out + 0.012)
  csf_ratio <- (v$SC1 * v$Papp_AB * 1e-6 * 0.0375 + 0.012 * kpuu) /
    (v$SC2 * v$Papp_BA * 1e-6 * 0.0375 + 0.132)
  data.frame(
    model = grid$model[i],
    kpuu_closed = kpuu, kpuu_sim = s$Cu_brain / s$Cu,
    csf_closed = csf_ratio, csf_sim = s$Ccsf / s$Cu
  )
})
kp <- do.call(rbind, kp)
kp$kpuu_relerr <- kp$kpuu_sim / kp$kpuu_closed - 1
kp$csf_relerr <- kp$csf_sim / kp$csf_closed - 1
knitr::kable(kp, digits = 5)
# Measured at most 4e-13 with rtol = 1e-10; 1e-5 leaves headroom for the
# numeric integrator while still failing on any structural or unit error.
stopifnot(nrow(kp) == 18L, all(abs(kp$kpuu_relerr) < 1e-5), all(abs(kp$csf_relerr) < 1e-5))
```

The steady-state `Kp,uu,brain` values sit near or below 1 for every drug. This
is a property of the fitted parameters, not a structural constraint; `Qbulk`
enters the denominator, so the packaged structure can in principle produce
`Kp,uu` above 1 when `PS_BBB,in` is large.

## Replicating Figure 3: the QSPR sub-model

To predict brain levels for a new drug, the authors regressed the natural
logarithm of each fitted scaling factor on the drug's lipophilicity (logP):
a cubic for `SC1` and `SC2` and a parabola for `SC3`, separately for each cell
line (Figure 3). The polynomial coefficients are printed under each panel of
Figure 3 and the logP values are in Supplementary Table S2.

```{r qspr}
logp <- c(
  amitriptyline = 4.81, caffeine = -0.55, carbamazepine = 2.77,
  fleroxacin = 0.98, pefloxacin = 0.75, zolpidem = 3.02
)
# Coefficients (x^3, x^2, x, intercept) of Figure 3, one row per cell line and
# scaling factor, with the printed R^2.
qspr <- data.frame(
  cell = rep(c("mdck", "mdckmdr1", "hcmec"), each = 3),
  sc = rep(c("sc1", "sc2", "sc3"), times = 3),
  a3 = c(-0.019, 0.080, 0, -0.044, 0.083, 0, -0.055, 0.020, 0),
  a2 = c(0.284, -0.63, -0.291, 0.470, -0.488, -0.308, 0.380, -0.275, -0.247),
  a1 = c(-0.048, 2.016, 0.967, -0.020, 1.794, 0.844, 0.184, 1.946, 0.180),
  a0 = c(1.237, 1.300, -0.888, 0.903, 1.116, -0.779, 1.358, 1.125, 0.394),
  r2_printed = c(0.969, 0.955, 0.854, 0.894, 0.831, 0.959, 0.643, 0.855, 0.895),
  stringsAsFactors = FALSE
)
qspr_predict <- function(row, x) exp(row$a3 * x^3 + row$a2 * x^2 + row$a1 * x + row$a0)

qspr_check <- lapply(seq_len(nrow(qspr)), function(k) {
  q <- qspr[k, ]
  fitted_sc <- vapply(drugs, function(dr) {
    params[[toupper(q$sc)]][params$model == sprintf("SanchezDengra_2021_%s_rat_pbpk_%s", dr, q$cell)]
  }, numeric(1))
  y <- log(fitted_sc)
  yhat <- log(qspr_predict(q, logp[drugs]))
  refit <- lm(y ~ poly(logp[drugs], if (q$sc == "sc3") 2 else 3, raw = TRUE))
  data.frame(
    cell = q$cell, sc = q$sc, r2_printed = q$r2_printed,
    r2_printed_coefs = 1 - sum((y - yhat)^2) / sum((y - mean(y))^2),
    r2_refit = summary(refit)$r.squared,
    max_coef_diff = max(abs(rev(coef(refit)) - c(q$a3, q$a2, q$a1, q$a0)[if (q$sc == "sc3") 2:4 else 1:4]))
  )
})
qspr_check <- do.call(rbind, qspr_check)
knitr::kable(qspr_check, digits = 3)
stopifnot(
  nrow(qspr_check) == 9L,
  # The printed polynomials, evaluated at the Table S2 logP values, reproduce
  # the printed R^2 of every panel against the Table 3 scaling factors.
  all(abs(qspr_check$r2_printed_coefs - qspr_check$r2_printed) < 0.01),
  # A least-squares refit recovers the printed coefficients (rounded to
  # 3 decimals) of every SC1 and SC2 panel (measured max 0.001); the SC3
  # panels carry the rounding of amitriptyline's SC3, see below.
  all(qspr_check$max_coef_diff[qspr_check$sc != "sc3"] < 0.01)
)
```

The three data sets that enter each regression -- the Table 3 scaling
factors, the Table S2 logP values and the Figure 3 coefficients -- are
therefore mutually consistent. The refit coefficients of the three `SC3`
panels differ slightly from the printed ones (by 0.013, 0.006 and 0.044 for
MDCK, MDCK-MDR1 and hCMEC/D3), although all three still reproduce the printed
R^2 to within 0.006. The cause is the same Table 3 rounding found above:
amitriptyline's `SC3` is printed as 0.05, 0.02 and 0.01, and at the extreme
logP of 4.81 its point has the most leverage in a six-point parabola. The
effect is largest for hCMEC/D3, where the Figure 3C point sits near
`ln SC3` = -4.9 (`SC3` of about 0.0075) rather than at ln 0.01 = -4.6. The
printed Figure 3 coefficients are used below.

```{r fig3-plot, fig.height = 6, fig.cap = "Replicates Figure 3 of Sanchez-Dengra 2021: ln(scaling factor) against logP with the printed QSPR polynomials."}
curve_x <- seq(-0.8, 5, by = 0.05)
qspr_curves <- do.call(rbind, lapply(seq_len(nrow(qspr)), function(k) {
  data.frame(cell = qspr$cell[k], sc = qspr$sc[k], logp = curve_x, lnsc = log(qspr_predict(qspr[k, ], curve_x)))
}))
qspr_points <- params |>
  mutate(
    drug = sub("^SanchezDengra_2021_(.*)_rat_pbpk_.*$", "\\1", model),
    cell = sub("^.*_rat_pbpk_", "", model)
  ) |>
  pivot_longer(c(SC1, SC2, SC3), names_to = "sc", values_to = "value") |>
  mutate(sc = tolower(sc), logp = logp[drug], lnsc = log(value))
lab <- function(x) unname(c(cells, sc1 = "ln SC1", sc2 = "ln SC2", sc3 = "ln SC3")[x])
ggplot(mapping = aes(logp, lnsc)) +
  geom_line(data = qspr_curves) +
  geom_point(data = qspr_points, aes(colour = drug)) +
  facet_grid(sc ~ factor(cell, levels = names(cells)), scales = "free_y", labeller = as_labeller(lab)) +
  labs(x = "logP", y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Replicating Figure 4: brain profiles predicted from logP

In the paper's internal validation, the fitted scaling factors of each model
were replaced by those predicted from logP by the QSPR, and the brain profiles
re-simulated (Figure 4). The packaged models do this with `ini()`:

```{r fig4-sim}
fig4 <- lapply(seq_len(nrow(grid)), function(i) {
  d <- design[design$drug == grid$drug[i], ]
  rows <- qspr[qspr$cell == grid$cell[i], ]
  sc_pred <- vapply(split(rows, rows$sc), function(r) qspr_predict(r, logp[[grid$drug[i]]]), numeric(1))
  m_qspr <- rxode2::ini(mods[[grid$model[i]]], sc1 = sc_pred[["sc1"]], sc2 = sc_pred[["sc2"]], sc3 = sc_pred[["sc3"]])
  obs <- sort(unique(c(seq(0, d$t_end, length.out = 401), 1:120)))
  obs <- obs[obs <= d$t_end]
  s_fit <- solve_one(grid$model[i], grid$drug[i], obs)
  s_qspr <- as.data.frame(rxode2::rxSolve(m_qspr, make_events(grid$drug[i], obs)))
  rbind(
    data.frame(drug = grid$drug[i], cell = grid$cell_label[i], scaling = "fitted (Table 3)", time = s_fit$time, brain = s_fit[[d$brain_var]]),
    data.frame(drug = grid$drug[i], cell = grid$cell_label[i], scaling = "QSPR (Figure 3)", time = s_qspr$time, brain = s_qspr[[d$brain_var]])
  )
})
fig4 <- do.call(rbind, fig4)
```

```{r fig4-plot, fig.height = 8, fig.cap = "Replicates Figure 4 of Sanchez-Dengra 2021: brain profiles simulated with QSPR-predicted scaling factors (solid) against the fitted profiles (dashed)."}
ggplot(fig4, aes(time, brain, colour = cell, linetype = scaling)) +
  geom_line() +
  scale_linetype_manual(values = c("fitted (Table 3)" = "dashed", "QSPR (Figure 3)" = "solid")) +
  facet_wrap(~drug, scales = "free", ncol = 2) +
  labs(x = "t (s)", y = "Brain C (ng/mL)", colour = "Cell line", linetype = NULL) +
  theme_bw() +
  theme(legend.position = "bottom", legend.box = "vertical")
```

The QSPR-predicted lines of Figure 4 were read by eye at their peaks (at the
end of the record for fleroxacin, whose brain concentration rises throughout):

```{r fig4-anchors}
fig4_anchor <- data.frame(
  drug = rep(drugs, each = 3),
  cell = rep(unname(cells), times = 6),
  what = rep(c("peak", "peak", "peak", "end", "peak", "peak"), each = 3),
  figure4 = c(6300, 5500, 4000, 31000, 31000, 31000, 610, 1090, 410, 1500, 1800, 1680, 1850, 1480, 1840, 400, 380, 2550)
)
fig4_at <- function(drug, cell, what) {
  x <- fig4[fig4$drug == drug & fig4$cell == cell & fig4$scaling == "QSPR (Figure 3)", ]
  if (nrow(x) == 0L) stop("no simulated rows for ", drug, " ", cell)
  if (what == "peak") max(x$brain) else x$brain[which.max(x$time)]
}
fig4_anchor$simulated <- signif(mapply(fig4_at, fig4_anchor$drug, fig4_anchor$cell, fig4_anchor$what), 4)
fig4_anchor$pct_diff <- round(100 * (fig4_anchor$simulated / fig4_anchor$figure4 - 1), 1)
knitr::kable(fig4_anchor)
# Measured within 8 percent everywhere except amitriptyline/MDCK (+13
# percent, discussed below). A wrong QSPR coefficient or logP moves these
# predictions by tens of percent or more (compare the cell lines).
stopifnot(nrow(fig4_anchor) == 18L, all(abs(fig4_anchor$pct_diff) < 15))
```

The QSPR predictions reproduce Figure 4, including the two failures the paper
discusses: carbamazepine is over-predicted with MDCK-MDR1 inputs and zolpidem
with hCMEC/D3 inputs. The largest difference is the amitriptyline MDCK peak,
13 percent above the paper's line. Two things plausibly contribute. The
paper's simulated lines are drawn through a few time points (the kinks in
Figure 2A), which cuts off the top of a peak that lasts only minutes. And at
logP = 4.81 the cubic term multiplies coefficient rounding by more than 100.

## Brain exposure with PKNCA

Brain `Cmax` and `AUC(0-tlast)` of the fitted and QSPR-predicted profiles, by
drug, cell line and scaling source:

```{r nca-brain}
conc_brain <- fig4 |>
  filter(!is.na(brain)) |>
  mutate(
    treatment = paste(drug, cell, scaling, sep = " | "),
    id = 1L,
    brain = pmax(brain, 0)
  )
dose_brain <- conc_brain |>
  distinct(treatment, id, drug) |>
  left_join(design |> mutate(total_dose = bolus_amt + inf_rate * inf_dur) |> select(drug, total_dose), by = "drug") |>
  mutate(time = 0)
o_conc_b <- PKNCAconc(conc_brain, brain ~ time | treatment + id)
o_dose_b <- PKNCAdose(dose_brain, total_dose ~ time | treatment + id)
o_data_b <- PKNCAdata(o_conc_b, o_dose_b,
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, auclast = TRUE)
)
nca_brain <- as.data.frame(pk.nca(o_data_b)) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  tidyr::separate(treatment, into = c("drug", "cell", "scaling"), sep = " \\| ")
stopifnot(nrow(nca_brain) == 36L, all(is.finite(nca_brain$cmax)), all(is.finite(nca_brain$auclast)))
knitr::kable(
  nca_brain |>
    dplyr::rename("Brain Cmax (ng/mL)" = cmax, "Brain AUC0-tlast (ng*s/mL)" = auclast),
  digits = 1
)
```

The paper does not tabulate brain `Cmax` or AUC. Its Table 4 gives the mean
prediction error of the QSPR-simulated brain `Cmax` and AUC against the
**experimental** data; those data are not published in tabular form, so Table 4
cannot be recomputed here. As a proxy, the prediction error of the
QSPR-simulated profile against the fitted profile (which tracked the data
with a mean AUC error near 5 percent, Table 4 first row) is:

```{r pe-proxy}
pe <- nca_brain |>
  pivot_wider(names_from = scaling, values_from = c(cmax, auclast)) |>
  mutate(
    pe_cmax = 100 * abs(`cmax_fitted (Table 3)` - `cmax_QSPR (Figure 3)`) / `cmax_fitted (Table 3)`,
    pe_auc = 100 * abs(`auclast_fitted (Table 3)` - `auclast_QSPR (Figure 3)`) / `auclast_fitted (Table 3)`
  )
pe_summary <- pe |>
  group_by(cell) |>
  summarise(mean_pe_cmax = mean(pe_cmax), mean_pe_auc = mean(pe_auc), .groups = "drop") |>
  mutate(cell = factor(cell, levels = cells)) |>
  arrange(cell) |>
  mutate(
    paper_pe_cmax = c(19.23, 35.71, 49.77),
    paper_pe_auc = c(22.34, 48.21, 46.69)
  )
knitr::kable(
  pe_summary |>
    dplyr::rename(
      "Cell line" = cell,
      "Mean PE% Cmax (QSPR vs fitted)" = mean_pe_cmax,
      "Mean PE% AUC (QSPR vs fitted)" = mean_pe_auc,
      "Table 4 PE% Cmax (vs data)" = paper_pe_cmax,
      "Table 4 PE% AUC (vs data)" = paper_pe_auc
    ),
  digits = 1
)
```

```{r pe-proxy-check}
# The paper's conclusion (MDCK predicts best) holds for the proxy too:
# measured 21.5 percent AUC error for MDCK against 42.5 and 41.1.
stopifnot(nrow(pe_summary) == 3L, which.min(pe_summary$mean_pe_auc) == 1L, which.min(pe_summary$mean_pe_cmax) == 1L)
```

The proxy and Table 4 measure different things -- the proxy excludes the
fitting error and the paper's PE% is taken as a signed per-drug value before
averaging (Equation 15) -- so the numbers are shown side by side rather than
compared. The paper's conclusion that the MDCK-based QSPR gives the best
predictions is visible in both.

## Assumptions and deviations

* **Eighteen files for one structure.** The paper fitted each drug separately
  for each of the three cell lines (Table 3), so each drug-cell-line
  combination is its own model, named
  `SanchezDengra_2021_<drug>_rat_pbpk_<cellline>`. The in vivo parameters
  `Vd`, `kel` and `ka` are the same in the three files of a drug because
  Table 3 reports a single value per drug.
* **Deterministic model.** No between-animal variability or residual error
  was estimated (mean profiles fitted in Berkeley Madonna), so the models have
  no `eta` or residual-error terms.
* **Time in seconds.** All rate constants, flows and infusion rates are
  printed per second; the models keep the paper's units (time s, dose ng,
  concentration ng/mL, volume cm^3 = mL).
* **Physiological flows used as printed.** `Qbulk` = 0.012 cm^3/s and `Qsink`
  = 0.132 cm^3/s (Methods 2.4, citing Ball et al.) are large compared with
  the rat brain interstitial and CSF flows used elsewhere in this package
  (0.2 and 2.2 uL/min in `Yamamoto_2017_quinidine_rat_pbpk`, Table 3 of that
  paper). They are used exactly as printed, because the
  fitted scaling factors were estimated with them and the Figure 2 fits are
  reproduced with them (see the anchor check above); any unit inconsistency
  in the physiological constants is absorbed by the fitted `SC1`-`SC3`.
* **Rounded amitriptyline `SC3`.** Table 3 prints amitriptyline's `SC3` as
  0.05 (MDCK), 0.02 (MDCK-MDR1) and 0.01 (hCMEC/D3). The models use these
  printed values. As a result the MDCK-MDR1 and hCMEC/D3 total brain
  concentrations of amitriptyline are 11 and 27 percent lower than the MDCK
  ones, whereas the paper's fits overlap. Unbound brain concentrations are not
  affected. The unrounded values implied by the overlap and by Figure 3 are
  about 0.018 and 0.0073.
* **Caffeine infusion duration.** Table 1 gives the caffeine infusion rate
  (833.333 ng/s) but not its duration; 14,400 s (4 h) was taken from the
  peak of the fitted lines in Figure 2B.
* **Fleroxacin and pefloxacin dosing.** Table 1 gives both `D` and `k0`; `D`
  is modelled as an intravenous bolus at time zero followed by the infusion
  for the whole record, which reproduces the non-zero starting plasma
  concentrations of Figure 2D-E.
* **Extravascular route and bioavailability.** Equations 7-8 feed the whole
  dose through `ka` into plasma; bioavailability is not a parameter, so any
  incomplete absorption is absorbed into the fitted `Vd`.
* **`kel` and `ka` source trace.** The values are copied from Table 3, where
  they are printed in `x 10^-n` notation (for example amitriptyline
  `kel` = 1.17 x 10^-4 s^-1).
* **Errata.** No erratum or correction was found for this article (Crossref
  update metadata, checked 2026-09-29).
