---
title: "Doxorubicin cardiotoxicity QSP, rat and human (Sang 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Doxorubicin cardiotoxicity QSP, rat and human (Sang 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5, message = FALSE)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Sang_2021_doxorubicin_rat_qsp")()$reference`
* Article: <https://doi.org/10.1002/psp4.12719>

Sang et al. (2021) built a quantitative systems pharmacology (QSP) model of
doxorubicin-induced cardiac dysfunction in rats and translated it to humans to
estimate a "safe-equivalent" cumulative dose for patients with pre-existing
cardiovascular disease. The paper contributes two model files:

* `Sang_2021_doxorubicin_rat_qsp` -- the fitted rat QSP-PK-PD model: a
  five-compartment doxorubicin PK model (intraperitoneal depot, plasma, heart,
  two peripheral compartments) feeding a cumulative heart-tissue AUC, which
  impairs bioenergy production and myocardial compliance inside a
  cardiovascular turnover system (stroke volume SV, left ventricular
  end-diastolic volume LVEDV, heart rate HR, total peripheral resistance TPR;
  MAP = HR x TPR x SV; LVESV = LVEDV - SV; LVEF = SV / LVEDV).
* `Sang_2021_doxorubicin_human_qsp` -- the human translation (Table 1): the
  dissipation rate constants are allometrically scaled, FB_LVESV is rescaled by
  the baseline LVESV, and only the bioenergy-production (systolic) arm is
  retained. The paper obtained human heart exposure from the separately
  published He 2018 multiscale doxorubicin PBPK model, which is not part of
  nlmixr2lib, so this file takes the heart-tissue doxorubicin concentration as
  the time-varying covariate `CEFFECT` (ug/mL).

The rat structure and every constant come from the Mlxtran code the authors
deposited with the Supporting Information; the parameter values match Tables
S4 and S5 except where noted under "Assumptions and deviations".

# Population

**Rat.** The PK model was fitted first to ten literature PK studies (2-6 mg/kg
single i.v. or i.p. doses in Wistar-Kyoto, Sprague-Dawley, Donryu and BDIX
rats; Table S1) and then fixed. The QSP and PD parameters were estimated
jointly (Monolix 2018R1, SAEM) on three literature echocardiography studies in
Sprague-Dawley rats (Chang 2015, Kim 2012, Lee 2014; 90 rats, 1.25-3 mg/kg i.p.
repeated doses) and an in-house study of 3.75 mg/kg i.p. every 3 days for 4
doses in healthy Wistar-Kyoto (WKY) rats, isoproterenol-hypertrophied WKY rats
and spontaneously hypertensive rats (SHR). Body weights were 250-329 g (Table S1).

**Human.** No human data were fitted. 13,994 virtual adult patients (8599
cardiovascular-healthy, 5395 with enlarged LVEDV and/or hypertension) were
generated by Monte Carlo sampling of the Table 1 baselines with 30% variation
and simulated for one year of 3-weekly doxorubicin (cumulative 120-900 mg/m^2).

```{r population}
str(readModelDb("Sang_2021_doxorubicin_rat_qsp")()$population)
str(readModelDb("Sang_2021_doxorubicin_human_qsp")()$population)
```

# Source trace

| Equation / parameter | Rat value | Human value | Source location |
|---|---|---|---|
| LVEDV turnover (`d/dt(edv)`) | -- | -- | Eq. 1; Mlxtran `ddt_LVEDV` |
| LVESV = LVEDV - SV; LVEF = SV/LVEDV | -- | -- | Eqs. 2-3; Mlxtran |
| HR, TPR, SV turnover | -- | -- | Eqs. 4-6; Mlxtran `ddt_HR`, `ddt_TPR`, `ddt_SV` |
| MAP = HR x TPR x SV | -- | -- | Eq. 7 |
| FB_MAP = FB_MAP_0 (MAP_base/MAP_0)^-1.98 | MAP_0 = 106.596 | 106.596 | Eq. 8; Mlxtran `FB_MAP` |
| E_drug_EP = exp(-AUC^h/(AUC^h + AUC50_EP^h)) | -- | -- | Eq. 9; Mlxtran `EdrugEP`, `EEP_SV` |
| MC transit chain and E_MC_LVEDV = 1 - MC_T3 | -- | not used | Eqs. 10-12; Mlxtran `MCtrans1-3` |
| E_MC_SV = 1 - AUC/(AUC + AUC50_MC) | -- | not used | Eq. 13; Mlxtran `EMC_SV` |
| Allometric scaling of kout, exponent -0.25 | -- | -- | Eq. 14 |
| FB_LVESV scaled by baseline LVESV ratio | -- | -- | Eq. 15 |
| `lka` | 4.532 1/h | -- | Mlxtran `k`; Table S4 ka 4.53 |
| `lkel` | 1.07 1/h | -- | Mlxtran `ke`; Table S4 |
| `lvc` | 442 mL | -- | Mlxtran `V1`; Table S4 |
| `lv_heart` | 60.5 mL | -- | Mlxtran `V2`; Table S4 |
| `lkin_heart`, `lkout_heart` | 6.75, 0.605 1/h | -- | Mlxtran `K12`, `K21`; Table S4 k_in_heart, k_out_heart |
| `lk12`, `lk21` | 6.74, 13.4 1/h | -- | Mlxtran `K13`, `K31`; Table S4 k12, k21 |
| `lk13`, `lk31` | 3.07, 0.0568 1/h | -- | Mlxtran `K14`, `K41`; Table S4 k13, k31 (0.0586) |
| `lrbase_hr` / `HR_BL` | 423 beats/min (fixed) | covariate (70) | Mlxtran `HR0`; Table S5 / Table 1 |
| `lkout_hr` | 11.58 1/h | 2.83 1/h | Mlxtran; Table 1 |
| `lkout_tpr` | 3.58 1/h | 0.875 1/h | Mlxtran; Table 1 |
| `lkout_edv` | 0.126 1/h | 0.0308 1/h | Mlxtran; Table 1 |
| `lkout_sv` | 0.126 1/h | 0.0308 1/h | Table S5; Table 1 |
| `lfb0` | 0.0029 1/mmHg | 0.0029 1/mmHg | Mlxtran; Table 1 FB_MAP_0 |
| `e_bslmap_fb` | -1.98 | -1.98 | Methods (Snelder 2014) |
| `lfb_lvesv` | 1.43 1/mL | 2.532e-3 1/mL | Table S5; Table 1 |
| `lktr` | 0.021 1/h | -- | Table S5 kt |
| `lauc50_ep` | 1390 ug*h/mL | 1390 ug*h/mL | Table S5; Table 1 |
| `lauc50_mc` | 1704 ug*h/mL | -- | Table S5 |
| `hill_ep`, `hill_mc` | 3, 1 | 3, -- | Results; Mlxtran |
| `etalfb_lvesv`, `etalauc50_ep` | -- | CV 12.4%, 7.06% | Table 1 CV column |
| Residual errors | not reported (fixed 0) | not reported (fixed 0) | Supplementary Methods Eqs. 1-4 |

# Rat model

```{r rat-model}
mod_rat <- readModelDb("Sang_2021_doxorubicin_rat_qsp")
rat_ui <- rxode2::rxode2(mod_rat)
rat_par <- setNames(rat_ui$iniDf$est, rat_ui$iniDf$name)
```

## Doxorubicin PK (Figure S2)

Figure S2 shows plasma and heart concentrations after a 5 mg/kg dose. The
deposited code works in per-animal amounts, so a 5 mg/kg dose to a 250 g rat is
1250 ug; the resulting heart peak of about 11 ug/mL and plasma concentration of
about 20 ng/mL at 24 h match the Figure S2 axes (in ng/mL).

```{r rat-pk}
dose_ug <- 5 * 0.250 * 1000
t_pk <- sort(unique(c(0, 10^seq(-3, log10(2000), length.out = 300), 24, 48, 72)))
pk_events <- dplyr::bind_rows(
  data.frame(id = 1L, time = 0, evid = 1L, amt = dose_ug, cmt = "central", treatment = "5 mg/kg i.v."),
  data.frame(id = 1L, time = t_pk, evid = 0L, amt = 0, cmt = NA_character_, treatment = "5 mg/kg i.v."),
  data.frame(id = 2L, time = 0, evid = 1L, amt = dose_ug, cmt = "depot", treatment = "5 mg/kg i.p."),
  data.frame(id = 2L, time = t_pk, evid = 0L, amt = 0, cmt = NA_character_, treatment = "5 mg/kg i.p.")
) |>
  dplyr::mutate(dvid = ifelse(evid == 0L, 1L, NA_integer_), LVEDV_BL = 0.385, LVESV_BL = 0.085, MAP_BL = 106.596, STRAIN_SD = 1)

pk_sim <- rxode2::rxSolve(mod_rat, pk_events, keep = "treatment", returnType = "data.frame",
                          atol = 1e-10, rtol = 1e-10, maxsteps = 1e6)

# The time-zero row (concentration 0) is left out of this log-scale plot only;
# the PKNCA input below keeps it.
pk_long <- pk_sim[pk_sim$time > 0, ] |>
  dplyr::select(treatment, time, Plasma = Cc, Heart = Cheart) |>
  tidyr::pivot_longer(c(Plasma, Heart), names_to = "matrix", values_to = "conc")

ggplot(pk_long, aes(time, conc * 1000, colour = treatment)) +
  geom_line() +
  facet_wrap(~matrix) +
  scale_y_log10() +
  coord_cartesian(xlim = c(0, 72)) +
  labs(x = "Time (h)", y = "Doxorubicin (ng/mL)", colour = NULL,
       caption = "Replicates Figure S2 of Sang 2021 (predicted median).")
```

```{r rat-pk-check}
pk_iv <- pk_sim[pk_sim$treatment == "5 mg/kg i.v.", ]
heart_peak <- max(pk_iv$Cheart)
plasma_24 <- pk_iv$Cc[pk_iv$time == 24] * 1000
knitr::kable(data.frame(
  Quantity = c("Heart Cmax (ug/mL)", "Plasma at 24 h (ng/mL)"),
  Simulated = signif(c(heart_peak, plasma_24), 3),
  "Figure S2" = c("about 10 (1e4 ng/mL axis)", "about 15-20"),
  check.names = FALSE
))
stopifnot(heart_peak > 5, heart_peak < 20, plasma_24 > 10, plasma_24 < 40)
```

### NCA (PKNCA)

The paper reports no NCA table for the rat PK, so the NCA here checks the
implementation against its own mass balance: plasma AUC(0-inf) must equal
Dose / (kel * V1) for both routes (i.p. absorption is complete), and the heart
AUC follows from the heart-to-plasma exchange.

```{r rat-nca}
nca_conc <- pk_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)
nca_dose <- pk_events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, treatment, time, amt)
conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id)
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                        aucinf.obs = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
nca_tab |>
  dplyr::rename("Route" = treatment, "Cmax (ug/mL)" = cmax, "Tmax (h)" = tmax,
                "AUC0-inf (ug*h/mL)" = aucinf.obs, "t1/2 (h)" = half.life) |>
  knitr::kable(digits = 4)

auc_expected <- dose_ug / (exp(rat_par[["lkel"]]) * exp(rat_par[["lvc"]]))
auc_pct <- 100 * (nca_tab$aucinf.obs / auc_expected - 1)
heart_expected <- auc_expected * exp(rat_par[["lvc"]]) * exp(rat_par[["lkin_heart"]]) /
  exp(rat_par[["lkout_heart"]]) / exp(rat_par[["lv_heart"]])
heart_pct <- 100 * (tail(pk_iv$auc_heart, 1) / heart_expected - 1)
c(expected_plasma_auc = auc_expected, pct_diff = auc_pct,
  expected_heart_auc = heart_expected, heart_state_pct_diff = heart_pct)
stopifnot(all(abs(auc_pct) < 2), abs(heart_pct) < 1)
```

## Pre-dose steady state

With no dose, every turnover state must stay at its baseline: the production
rates are derived from the pre-dose steady state.

```{r rat-steady}
ss_events <- data.frame(id = 1L, time = c(0, 24, 240, 2400), evid = 0L, amt = 0, dvid = 1L,
                        LVEDV_BL = 0.385, LVESV_BL = 0.085, MAP_BL = 106.596, STRAIN_SD = 0)
ss <- rxode2::rxSolve(mod_rat, ss_events, returnType = "data.frame")
ss[, c("time", "LVEF", "LVEDV", "LVESV", "MAP", "hr", "tpr")]
stopifnot(
  max(abs(ss$LVEDV / 0.385 - 1)) < 1e-6,
  max(abs(ss$LVESV / 0.085 - 1)) < 1e-6,
  max(abs(ss$MAP / 106.596 - 1)) < 1e-6
)
```

## Figure 2 and Table S2

The panels of Figure 2 are typical-value predictions using each study's own
baselines. Doses are per-animal (mg/kg x study body weight from Table S1).
Baselines for studies 11-13 are read from the starting point of the Figure 2
curves; for the in-house groups 14A-C they are the Table S2 day-0 means. Dosing
intervals for the literature studies are read from the Figure 2 arrows; the
in-house schedule is every 3 days for 4 doses (Supplementary Methods).

```{r rat-studies, warning = FALSE}
studies <- data.frame(
  study = c("11", "12", "13", "14A", "14B", "14C"),
  STRAIN_SD = c(1, 1, 1, 0, 0, 0),
  dose_ug = c(3 * 275, 1.25 * 329.3, 2.5 * 250, 3.75 * 250, 3.75 * 250, 3.75 * 250),
  n_dose = c(6, 16, 6, 4, 4, 4),
  ii = c(168, 55, 46, 72, 72, 72),
  lvef_bl = c(78, 80, 85, (1 - 75.6 / 264.2) * 100,
              (1 - 96.8 / 292.3) * 100, (1 - 107.3 / 300.3) * 100),
  lvedv_bl = c(0.37, 0.39, 0.34, 0.2642, 0.2923, 0.3003),
  map_bl = c(106.596, 106.596, 106.596, 112.7, 120.3, 156.7),
  tend = c(1344, 1344, 1008, 384, 384, 384)
)
studies$lvesv_bl <- studies$lvedv_bl * (1 - studies$lvef_bl / 100)

make_study <- function(i) {
  s <- studies[i, ]
  obs <- sort(unique(c(seq(0, s$tend, by = 4), 96, 192, 360)))
  dplyr::bind_rows(
    data.frame(time = (seq_len(s$n_dose) - 1) * s$ii, evid = 1L, amt = s$dose_ug, cmt = "depot", dvid = NA_integer_),
    data.frame(time = obs, evid = 0L, amt = 0, cmt = NA_character_, dvid = 1L)
  ) |>
    dplyr::mutate(id = i, study = s$study, STRAIN_SD = s$STRAIN_SD,
                  LVEDV_BL = s$lvedv_bl, LVESV_BL = s$lvesv_bl, MAP_BL = s$map_bl)
}
rat_events <- dplyr::bind_rows(lapply(seq_len(nrow(studies)), make_study)) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
rat_sim <- rxode2::rxSolve(mod_rat, rat_events, keep = "study", returnType = "data.frame",
                           maxsteps = 1e6)

rat_sim |>
  dplyr::select(study, time, "LVEF (%)" = LVEF, "LVEDV (mL)" = LVEDV) |>
  tidyr::pivot_longer(-c(study, time)) |>
  ggplot(aes(time, value)) +
  geom_line(colour = "red") +
  facet_grid(name ~ study, scales = "free") +
  labs(x = "Time since first dose (h)", y = NULL,
       caption = "Replicates Figure 2 of Sang 2021 (a: studies 11-13; b: study 14A-C).")
```

Study 11 is the panel with the largest change. The end-of-study values of its
Figure 2 curve were read from the figure by the maintainers.

```{r rat-fig2-check}
s11 <- rat_sim[rat_sim$study == "11" & rat_sim$time == 1344, ]
fig2_tab <- data.frame(
  Quantity = c("LVEF at 1344 h (%)", "LVEDV at 1344 h (mL)"),
  Simulated = round(c(s11$LVEF, s11$LVEDV), 3),
  "Figure 2a (read from curve)" = c(54.6, 0.458),
  check.names = FALSE
)
knitr::kable(fig2_tab)
stopifnot(abs(s11$LVEF - 54.6) < 3, abs(s11$LVEDV - 0.458) < 0.025)
```

The in-house groups can be compared with the observed Table S2 means (n = 9 per
group) at days 4, 8 and 15.

```{r rat-tableS2}
obs_s2 <- data.frame(
  study = rep(c("14A", "14B", "14C"), each = 3),
  day = rep(c(4, 8, 15), 3),
  lvef_obs = c(63.8, 65.2, 77.1, 62.2, 56.6, 68.3, 59.8, 58.3, 61.7),
  lvef_sd = c(4.3, 3.3, 7.4, 3.9, 4.4, 8.5, 3.4, 4.5, 6.3),
  lvedv_obs = c(261.0, 242.0, 242.4, 288.2, 271.2, 240.2, 307.8, 307.1, 253.4) / 1000
)
pred_s2 <- rat_sim |>
  dplyr::filter(study %in% obs_s2$study, time %in% c(96, 192, 360)) |>
  dplyr::mutate(day = time / 24) |>
  dplyr::select(study, day, lvef_pred = LVEF, lvedv_pred = LVEDV)
s2_cmp <- dplyr::inner_join(obs_s2, pred_s2, by = c("study", "day"))
stopifnot(nrow(s2_cmp) == 9L)
s2_cmp |>
  dplyr::mutate(lvef_pred = round(lvef_pred, 1), lvedv_pred = round(lvedv_pred, 3)) |>
  dplyr::rename("Group" = study, "Day" = day, "LVEF obs (%)" = lvef_obs, "LVEF SD" = lvef_sd,
                "LVEF pred (%)" = lvef_pred, "LVEDV obs (mL)" = lvedv_obs,
                "LVEDV pred (mL)" = lvedv_pred) |>
  knitr::kable()
lvef_z <- (s2_cmp$lvef_pred - s2_cmp$lvef_obs) / s2_cmp$lvef_sd
stopifnot(median(abs(lvef_z)) < 1, all(abs(lvef_z) < 2.5))
```

The typical curve tracks the day-4 and day-8 nadir; the day-15 observations
recover further than the model in the healthy group 14A, as the Figure 2b
curve also shows.

# Human model

## Rat-to-human scaling (Table 1)

```{r scaling}
scale_f <- (0.25 / 70)^0.25
scaling <- data.frame(
  Parameter = c("kout_SV", "kout_LVEDV", "kout_HR", "kout_TPR", "FB_LVESV"),
  Rat = c(0.126, 0.126, 11.58, 3.58, 1.43),
  Recomputed = c(0.126 * scale_f, 0.126 * scale_f, 11.58 * scale_f, 3.58 * scale_f,
                 1.43 * 0.085 / (113 - 65)),
  "Table 1 human" = c(0.0308, 0.0308, 2.83, 0.875, 2.532e-3),
  check.names = FALSE
)
knitr::kable(scaling, digits = 5)
stopifnot(all(abs(scaling$Recomputed / scaling$`Table 1 human` - 1) < 0.005))
```

## Pre-treatment steady state

```{r human-model, warning = FALSE}
mod_hum <- readModelDb("Sang_2021_doxorubicin_human_qsp")
mod_hum_typ <- rxode2::zeroRe(rxode2::rxode2(mod_hum))

hum_ss <- data.frame(
  id = rep(1:2, each = 3), time = rep(c(0, 240, 2400), 2), evid = 0L, amt = 0, dvid = 1L,
  CEFFECT = 0, LVEDV_BL = rep(c(113, 141), each = 3), LVESV_BL = rep(c(48, 76), each = 3),
  HR_BL = 70, MAP_BL = rep(c(70 * 0.02 * 65, 70 * 0.025 * 65), each = 3)
)
hs <- rxode2::rxSolve(mod_hum_typ, hum_ss, returnType = "data.frame")
hs[, c("id", "time", "LVEF", "LVEDV", "LVESV", "MAP")]
stopifnot(max(abs(hs$LVEDV / rep(c(113, 141), each = 3) - 1)) < 1e-6,
          max(abs(hs$MAP / rep(c(91, 113.75), each = 3) - 1)) < 1e-6)
```

## Figure 4: sensitivity to baseline LVEDV and MAP

Figure 4 plots the relative change of LVEF from baseline over about 11 months
of 3-weekly doxorubicin for patients insensitive (AUC50_EP = 1390 ug*h/mL,
panels a-c) and sensitive (AUC50_EP = 463.3 ug*h/mL, one third, panels d-f)
to doxorubicin, varying baseline LVEDV and MAP. The floor of -65.2% in panels d
and e (LVEF frozen at 20% by the model's numerical guard) shows that baseline
LVEF was held at 65/113 = 57.5% as LVEDV was varied.

The heart-concentration profile came from the He 2018 PBPK model and is not
reproduced here. Because the turnover states equilibrate within days, the late
plateau of Figure 4 depends on the final cumulative heart AUC and not on the
shape of the profile, so an illustrative driver is used: nine 24-hour
constant-concentration pulses, 21 days apart. Its total AUC is calibrated so
that the reference curve of panel d (LVEDV 113 mL, MAP 91 mmHg) ends at the
digitised -31.9%; every other curve is then a prediction.

```{r fig4}
fig4_events <- function(auc_total, lvedv, map, id = 1L) {
  starts <- (0:8) * 21 * 24
  tt <- sort(unique(c(seq(0, 330 * 24, by = 24), starts, starts + 24)))
  on <- vapply(tt, function(t) any(t >= starts & t < starts + 24), logical(1))
  sv <- lvedv * 65 / 113
  data.frame(id = id, time = tt, evid = 0L, amt = 0, dvid = 1L,
             CEFFECT = ifelse(on, auc_total / (9 * 24), 0),
             LVEDV_BL = lvedv, LVESV_BL = lvedv - sv, HR_BL = 70, MAP_BL = map)
}
lvef_change <- function(auc_total, auc50, lvedv, map) {
  m <- mod_hum_typ |> rxode2::ini(lauc50_ep = log(auc50))
  s <- rxode2::rxSolve(m, fig4_events(auc_total, lvedv, map), covsInterpolation = "locf",
                       returnType = "data.frame", maxsteps = 1e6)
  s$change <- 100 * (s$LVEF / (65 / 113 * 100) - 1)
  s
}
auc_cal <- uniroot(function(a) tail(lvef_change(a, 463.3, 113, 91)$change, 1) + 31.9,
                   c(200, 1000), tol = 1e-3)$root
auc_cal

fig4_cases <- dplyr::bind_rows(
  data.frame(panel = "a", auc50 = 1390, lvedv = c(113, 130, 146, 163, 180), map = c(91, 104, 118, 131, 145),
             digitised = c(-2.47, -2.68, -2.93, -3.20, -3.60)),
  data.frame(panel = "b", auc50 = 1390, lvedv = c(113, 130, 146, 163, 180), map = 91,
             digitised = c(-2.47, -2.65, -2.84, -3.02, -3.29)),
  data.frame(panel = "c", auc50 = 1390, lvedv = 113, map = c(91, 106, 120, 135, 150),
             digitised = c(-2.47, NA, NA, NA, -2.68)),
  data.frame(panel = "d", auc50 = 463.3, lvedv = c(113, 130, 146, 163, 180), map = c(91, 104, 118, 131, 145),
             digitised = c(-31.9, -34.1, -36.8, -41.3, -65.2)),
  data.frame(panel = "e", auc50 = 463.3, lvedv = c(113, 130, 146, 163, 180), map = 91,
             digitised = c(-31.9, -33.3, -35.4, -38.6, -65.2)),
  data.frame(panel = "f", auc50 = 463.3, lvedv = 113, map = c(91, 106, 120, 135, 150),
             digitised = c(-31.9, NA, NA, NA, -34.6))
)
fig4_sim <- dplyr::bind_rows(lapply(seq_len(nrow(fig4_cases)), function(i) {
  cs <- fig4_cases[i, ]
  s <- lvef_change(auc_cal, cs$auc50, cs$lvedv, cs$map)
  data.frame(panel = cs$panel, curve = sprintf("LVEDV = %g, MAP = %g", cs$lvedv, cs$map),
             day = s$time / 24, change = s$change, auc_heart = s$auc_heart)
}))

ggplot(fig4_sim, aes(day, change, colour = curve)) +
  geom_line() +
  facet_wrap(~panel, scales = "free_y") +
  labs(x = "Time since enrollment (day)", y = "LVEF change from baseline (%)", colour = NULL,
       caption = "Replicates Figure 4 of Sang 2021 with an illustrative heart-exposure driver.") +
  theme(legend.position = "bottom")
```

```{r fig4-check}
plateau <- fig4_sim |>
  dplyr::group_by(panel, curve) |>
  dplyr::summarise(simulated = dplyr::last(change), auc_end = dplyr::last(auc_heart), .groups = "drop")
# Every curve must have integrated the whole driver: a covariate supplied on too
# sparse a time grid is stepped over by the solver and silently lost.
stopifnot(all(abs(plateau$auc_end / auc_cal - 1) < 1e-3))
fig4_cmp <- fig4_cases |>
  dplyr::mutate(curve = sprintf("LVEDV = %g, MAP = %g", lvedv, map)) |>
  dplyr::inner_join(plateau, by = c("panel", "curve"))
stopifnot(nrow(fig4_cmp) == nrow(fig4_cases))
fig4_cmp |>
  dplyr::filter(!is.na(digitised)) |>
  dplyr::mutate(simulated = round(simulated, 2)) |>
  dplyr::select("Panel" = panel, "Baselines" = curve, "Figure 4 plateau (%)" = digitised,
                "Simulated (%)" = simulated) |>
  knitr::kable()
chk <- fig4_cmp[!is.na(fig4_cmp$digitised), ]
stopifnot(nrow(chk) == 24L, all(abs(chk$simulated - chk$digitised) < 1.5))
```

A single calibrated exposure reproduces every digitised plateau within about
one percentage point, including the ordering across baselines, the small MAP
effect (panels c and f) and the 20% LVEF floor.

## Figure S5: sensitivity to AUC50_EP

Figure S5 varies AUC50_EP over a nine-fold range for the typical healthy
patient (LVEDV 113 mL, MAP 91 mmHg). With the same calibrated exposure:

```{r figS5}
s5 <- data.frame(auc50 = c(463.3, 802.5, 1390, 2407),
                 digitised = c(-32.2, -10.8, -2.65, -0.6))
s5$simulated <- vapply(s5$auc50, function(a) tail(lvef_change(auc_cal, a, 113, 91)$change, 1),
                       numeric(1))
s5 |>
  dplyr::mutate(simulated = round(simulated, 2)) |>
  dplyr::rename("AUC50_EP (ug*h/mL)" = auc50, "Figure S5 plateau (%)" = digitised,
                "Simulated (%)" = simulated) |>
  knitr::kable()
stopifnot(all(abs(s5$simulated - s5$digitised) < 1.5))
```

The Results text says LVEF reduction "increased from 5% to 32%" as AUC50_EP
fell from 1390 to 463.3; the Figure S5 and Figure 4 curves show about 2.5% at
1390, which the model reproduces.

## Virtual cohort (illustrative)

The published incidence curves (Figure 3) need the He 2018 heart exposure and
cannot be reproduced here. The chunk below shows how a virtual cohort is set
up: baselines vary log-normally with 30% CV around the Table 1 typical values
(healthy: SV 65 mL, LVEDV 113 mL, HR 70 beats/min, TPR 0.02 mmHg*min/mL;
diseased: LVEDV 141 mL and TPR 0.025), with the Table 1 variability on
FB_LVESV and AUC50_EP. Draws whose baseline LVEF falls outside 35-80% are
redrawn. The driver is the same illustrative pulse train, at the calibrated
total heart AUC, supplied on a daily grid: rxode2 does not restart the
integrator where a covariate changes, so a `CEFFECT` profile given only at its
change points can be stepped over and lost. The chunk checks that every
patient accumulated the full heart AUC.

```{r cohort}
rxode2::rxSetSeed(20210512)
set.seed(20210512)
draw_arm <- function(n, lvedv0, tpr0, arm, id0) {
  out <- data.frame()
  while (nrow(out) < n) {
    d <- data.frame(sv = 65 * exp(rnorm(n, 0, sqrt(log(1.09)))),
                    lvedv = lvedv0 * exp(rnorm(n, 0, sqrt(log(1.09)))),
                    hr = 70 * exp(rnorm(n, 0, sqrt(log(1.09)))),
                    tpr = tpr0 * exp(rnorm(n, 0, sqrt(log(1.09)))))
    d <- d[d$sv / d$lvedv > 0.35 & d$sv / d$lvedv < 0.80, ]
    out <- rbind(out, d)
  }
  out <- out[seq_len(n), ]
  out$id <- id0 + seq_len(n)
  out$arm <- arm
  out
}
pts <- rbind(draw_arm(100, 113, 0.020, "Cardiovascular healthy", 0L),
             draw_arm(100, 141, 0.025, "Cardiovascular disease", 100L))
cohort_events <- dplyr::bind_rows(lapply(seq_len(nrow(pts)), function(i) {
  p <- pts[i, ]
  e <- fig4_events(auc_cal, p$lvedv, p$hr * p$tpr * p$sv, id = p$id)
  e$LVESV_BL <- p$lvedv - p$sv
  e$HR_BL <- p$hr
  e$arm <- p$arm
  e
}))
cohort <- rxode2::rxSolve(mod_hum, cohort_events, keep = "arm", covsInterpolation = "locf",
                          returnType = "data.frame", maxsteps = 1e6)
cohort_end <- cohort |>
  dplyr::group_by(id, arm) |>
  dplyr::summarise(change = 100 * (dplyr::last(LVEF) / dplyr::first(LVEF) - 1),
                   auc_end = dplyr::last(auc_heart), .groups = "drop")
cohort_end |>
  dplyr::group_by(arm) |>
  dplyr::summarise(N = dplyr::n(), median_change = median(change),
                   pct_ge_10pct_decline = 100 * mean(change <= -10)) |>
  dplyr::rename("Arm" = arm, "Median LVEF change (%)" = median_change,
                "Patients with >= 10% decline (%)" = pct_ge_10pct_decline) |>
  knitr::kable(digits = 1)
stopifnot(nrow(cohort_end) == 200L, all(is.finite(cohort_end$change)),
          all(abs(cohort_end$auc_end / auc_cal - 1) < 1e-3))
```

# Assumptions and deviations

* **Units.** Table S4 prints the PK volumes as L/kg and Tables 1 and S5 print
  AUC50 in mg*h/mL. The deposited code only reproduces Figure S2 (ng/mL axes)
  when the dose is a per-animal amount and the volumes are per-animal mL, and it
  only reproduces the gradual Figure 2 decline when that amount is in ug (in ng
  the bioenergy effect saturates at the first dose and LVEF collapses to the 20%
  floor). The models therefore use per-animal ug doses, mL volumes, ug/mL
  concentrations and AUC50 in ug*h/mL (numerically mg*h/L). Heart rate is in
  beats/min although the tables print beats/h; the model is unaffected because
  TPR is derived from MAP / (HR * SV).
* **PK values.** The model uses the as-run Mlxtran constants. Table S4 labels the
  plasma-heart exchange `k_in_heart` / `k_out_heart` and the two peripheral
  pairs `k12/k21` and `k13/k31`, matching the code's `K12/K21`, `K13/K31` and
  `K14/K41`. Table S4 prints the slow return rate as 0.0586 1/h; the code uses
  0.0568 1/h, which is kept because it is the value the QSP parameters were
  estimated against. ka is 4.532 (code) versus 4.53 (Table S4).
* **Strain switch.** The code's `Species` regressor removes the
  myocardial-compliance effect (`EdrugMC * (1 - Species)`, commented "Wistar").
  It is encoded as `STRAIN_SD` (1 = Sprague-Dawley), because studies 11-13 were
  Sprague-Dawley rats with the systolic-dysfunction phenotype of Figure 2a and
  study 14A-C used WKY / SHR rats with the diastolic phenotype of Figure 2b.
* **Numerical guard and feedback clamp.** The deposited code freezes all four
  turnover states once SV or LVESV is non-positive or LVEF is at or below 20%,
  and clamps the MAP feedback term at 1 for HR, TPR and LVEDV but not for SV.
  Both are reproduced as coded, in the human model too.
* **Reference MAP of Eq. 8.** The code uses MAP_0 = 106.596 mmHg (the typical
  rat MAP). Table 1 does not list MAP_0 among the rescaled parameters, so the
  human model keeps 106.596. Replacing it with the human typical 91 mmHg moves
  the Figure 4 plateaus by at most 0.5 percentage points, so the figure cannot
  distinguish the two readings.
* **Human model scope.** Only the bioenergy-production (systolic) effect is
  translated, as stated in the Methods and reflected in Table 1, which lists
  AUC50_EP but not AUC50_MC or kt. The human heart-tissue concentration must be
  supplied as `CEFFECT` from an external PK model; Sang 2021 used He 2018
  (Pharm Res 35:174), which is not in nlmixr2lib.
* **Human variability.** The Table 1 CV column gives 12.4% for FB_LVESV and
  7.06% for AUC50_EP, the rat relative standard errors of those estimates. They
  are encoded as log-normal between-patient variability (variance
  log(1 + CV^2)) and held fixed. The 30% variability on the baselines is not in
  the model because the baselines are covariates; the virtual cohort above
  samples them log-normally, and the redraw window for baseline LVEF (35-80%)
  is the maintainers' choice, as the paper does not state how implausible draws
  were handled.
* **Table 1 "65/115".** The SV0 row reads "65/115" for healthy/diseased. A
  diseased SV of 115 mL with LVEDV 141 mL would give LVEF 82% and MAP 201 mmHg;
  with SV 65 mL, TPR 0.025 and HR 70 the diseased MAP is 113.75 mmHg, close to
  the 115 mmHg hypertension threshold. The cohort above keeps SV at 65 mL for
  both arms.
* **Residual error.** Supplementary Methods name additive error for LVEF and
  MAP and proportional error for LVEDV and LVESV but report no estimates;
  they are fixed to 0. No between-animal variability was reported for the rat
  model.
* **Study schedules.** Figure 2b shows six dose arrows for study 14, whereas
  Table S1 and the Supplementary Methods give four doses every 3 days; the
  latter is used. Schedules for studies 11-13 are read from the Figure 2a
  arrows.
