---
title: "Levofloxacin in pregnancy (Sawe 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levofloxacin in pregnancy (Sawe 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
mod <- rxode2::rxode(readModelDb("Sawe_2025_levofloxacin"))
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.1128/aac.01626-24>
* Supplement (Tables S1-S3, Fig. S1-S7, NONMEM control stream): <https://doi.org/10.1128/aac.01626-24>
  (file `aac.01626-24-s0001.docx`, retrieved via Europe PMC `PMC12057369`)

Sawe 2025 pooled two South African studies in adults treated for
rifampicin-resistant tuberculosis (RR-TB) to ask what third-trimester
pregnancy does to levofloxacin exposure. The answer is a 38.1% increase in
clearance *on top of* the increases already explained by body size and renal
function, which together drive antepartum exposures well below those of
non-pregnant participants on the same weight-based dose.

# Population

47 participants contributed 320 levofloxacin concentrations: 33 (70%) female,
38 (81%) black, median (range) age 32 (19-51) years, weight 58.0 (37.0-98.0) kg,
height 1.60 (1.46-1.88) m, fat-free mass 39.4 (27.3-51.2) kg, serum creatinine
56.2 (25.3-110) umol/L and serum albumin 30.5 (17.0-40.0) g/L (Table 1). 31
(66%) were living with HIV and on antiretroviral therapy. Of the 33 women, 21
were pregnant and sampled in the third trimester; 12 of those contributed
matched antepartum and postpartum profiles, giving 19 antepartum, 14
postpartum, 12 non-pregnant-female and 14 male pharmacokinetic profiles.

Levofloxacin was dosed 750 or 1000 mg once daily by WHO weight band (34-50 kg
and above 50 kg), with a standard breakfast about 1 h before the observed dose.
Sampling was pre-dose and at 2, 4, 6, 8, 10 and 24 h (BEAT Tuberculosis,
NCT04062201), pre-dose and 2, 4, 6, 8 and 24 h in its pregnancy sub-study, and
pre-dose and 2, 4 and 6 h in the King Dinuzulu Hospital observational cohort.
Assay was HPLC-MS/MS with an LLOQ of 0.0781 mg/L; 6 of 320 samples (1.88%)
were below it, all pre-dose.

The same information is available programmatically:

```{r population}
str(readModelDb("Sawe_2025_levofloxacin")()$population, max.level = 1)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Sawe_2025_levofloxacin.R`. Collected here for
review. "Control stream" refers to the NONMEM code printed in the
supplementary material, whose `$THETA` block agrees with Table 2 to every
printed digit and supplies additional significant figures.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F) | `log(6.05859)` L/h | Table 2 "Clearance, CL (L/h)" 6.06 (95% CI 5.47-6.53); control stream `$THETA 1` |
| `lvc` (V/F) | `log(85.8803)` L | Table 2 "Volume of distribution, V (L)" 85.9 (95% CI 80.6-91.7); control stream `$THETA 2` |
| `lka` | `log(1.59498)` 1/h | Table 2 "Absorption rate constant, Ka (1/h)" 1.59 (95% CI 1.11-2.40); control stream `$THETA 3` |
| `lmtt` | `log(1.07437)` h | Table 2 "Mean transit time, MTT (h)" 1.07 (95% CI 0.771-1.32); control stream `$THETA 7` |
| `lnn` (N transit) | `fixed(log(20))` | Table 2 "Number of transit compartments, NN" 20 (fixed); control stream `$THETA 13 = 20 FIX` |
| `lfdepot` (F) | `fixed(log(1))` | Table 2 "Bioavailability, F" 1 (fixed); control stream `$THETA 4 = 1 FIX` |
| `e_ffm_cl` | `fixed(0.75)` | Methods, "allometric scaling of clearance (with a fixed exponent of 0.75)"; control stream `ALLMCL_FFM = (FFM/39.4)**0.75` |
| `e_ffm_vc` | `fixed(1)` | Methods, "volume of distribution (with a fixed exponent of 1)"; control stream `ALLMV_FFM = (FFM/39.4)` |
| `e_preg_cl` | `0.380866` | Table 2 "Effect of pregnancy on CL (%)" +38.1 (95% CI +23.4 to +57.1); control stream `IF (PREGNANT.EQ.1) preg_CL = 1 + THETA(12)`, `$THETA 12` |
| `e_creat_cl` | `-0.366504` | Table 2 "Effect of serum creatinine on CL (power exponent)" -0.367 (95% CI -0.493 to -0.104); Methods equation (supplementary inline equation `m001`); control stream `$THETA 15` |
| `etalcl` | `0.0489868` | Table 2 "Between-subject variability, CL" 22.1% (95% CI 17.1-28.3); control stream `$OMEGA 1` |
| `etaiov_mtt_1..4` | `0.211291` | Table 2 "Between-occasion variability, MTT" 45.9% (95% CI 30.5-70.3); control stream `$OMEGA 21` + 3x `SAME` |
| `etaiov_ka_1..4` | `0.732947` | Table 2 "Between-occasion variability, Ka" 85.6% (95% CI 61.7-119); control stream `$OMEGA 17` + 3x `SAME` |
| `etaiov_fdepot_2`, `_4` | `0.0562092` | Table 2 "Between-occasion variability, F" 23.7% (95% CI 19.3-28.1); control stream `$OMEGA 13` + 3x `SAME` |
| `etaiov_fdepot_1`, `_3` | `0.310775` | `2.35136^2 * 0.0562092`, from Table 2 "Scaling factor for BOV on F for unobserved doses" 2.35 (95% CI 1.68-3.28) and control stream `$THETA 14` with `IF(UNOBS.EQ.1) BOVBIO = E_BOVF*BOVBIO` |
| `addSd` | `0.259443` mg/L | Table 2 "Additive error (mg/L)" 0.244 (`$THETA 6 = 0.243823`) plus the control stream's unconditional `ADD = THETA(6) + LLOQ*0.2` with LLOQ 0.0781 mg/L |
| `propSd` | `0.0733236` | Table 2 "Proportional error (%)" 7.33 (95% CI 6.49-8.10); control stream `$THETA 5` |
| Transit input rate | `transit(nn, mtt, fbio)` | Control stream `PIZZA`/`TRANSIT` block with `KTR = (NN+1)/MTT`, citing Savic 2007; supplementary Fig. S1 |
| Dose enters via transit only | no `f(depot)` line | Control stream `F1 = 0`; under rxUi the bolus is not separately added, so no explicit suppression is needed (see Errata) |
| CL covariate model | `(FFM/39.4)^0.75 * (1 + e_preg_cl*PREG) * (CREAT/56.2)^e_creat_cl` | Control stream `TVCL = THETA(1)*ALLMCL_FFM*preg_CL*CR_CL` |
| Residual error | `add(addSd) + prop(propSd)` | Control stream `W = SQRT(ADD**2 + PROP**2)` |
| Reference FFM / creatinine | 39.4 kg / 56.2 umol/L | Table 2 footnote *b*; control stream `TVFFM = 39.4`, `MEDCREAT = 56.2` |

The variance scale deserves a note of its own, because it is what the
publisher's 17 April 2025 correction changed. Table 2 footnote *d* defines the
reported percentages as

$$\%CV = \sqrt{\omega^2}\cdot 100$$

so a reported "22.1%" **is** `omega * 100` and is *not* a log-normal CV. The
usual `omega^2 = log(CV^2 + 1)` conversion must therefore NOT be applied here.
The check below confirms the reading against the control stream's `$OMEGA`
values, all four of which reproduce their printed Table 2 percentage.

```{r variance-scale}
omega_cs <- c(
  "BSV CL"  = 0.0489868,
  "BOV MTT" = 0.211291,
  "BOV Ka"  = 0.732947,
  "BOV F"   = 0.0562092
)
printed <- c("BSV CL" = 22.1, "BOV MTT" = 45.9, "BOV Ka" = 85.6, "BOV F" = 23.7)

data.frame(
  Parameter          = names(omega_cs),
  `omega^2 (stream)` = omega_cs,
  `sqrt(omega^2)*100` = round(sqrt(omega_cs) * 100, 1),
  `Table 2 %CV`      = printed,
  `log(CV^2+1) would give` = round(log(printed^2 / 1e4 + 1), 4),
  check.names = FALSE, row.names = NULL
) |>
  knitr::kable(caption = "Table 2 footnote d reads %CV = sqrt(omega^2)*100.")

# sqrt(omega^2)*100 reproduces every printed Table 2 percentage to the
# printed precision (MTT prints 45.9%, a truncation of 45.97%).
stopifnot(all(abs(sqrt(omega_cs) * 100 - printed) < 0.1))
```

# Covariate derivation: fat-free mass

The size descriptor is fat-free mass, not total body weight. The supplementary
material gives the Janmahasatian formula

$$FFM = \frac{WHS_{max}\cdot HT^2 \cdot WT}{WHS_{50}\cdot HT^2 + WT}$$

with `WHSmax` 37.99 and `WHS50` 35.98 for females, 42.92 and 30.93 for males,
`HT` in metres and `WT` in kilograms. FFM beat total body weight as the
allometric descriptor (dOFV 12.9 versus 3.46), which is why no explicit sex
term appears in the final model: the sex difference in exposure is mediated
through FFM.

```{r ffm-helper}
ffm_janmahasatian <- function(wt, ht, sexf) {
  whs_max <- ifelse(sexf == 1, 37.99, 42.92)
  whs_50  <- ifelse(sexf == 1, 35.98, 30.93)
  whs_max * ht^2 * wt / (whs_50 * ht^2 + wt)
}
```

Applying it to the published median weight and height of each participant
group should reproduce that group's published median FFM. Group medians of
different quantities need not be exactly self-consistent, so this is checked
at 6%.

```{r ffm-check}
ffm_ref <- tibble::tribble(
  ~group,                 ~wt,  ~ht,  ~sexf, ~ffm_published, ~source,
  "Antepartum",           61.5, 1.60, 1,     38.7,           "Table S1 study data, pregnant (3rd trimester)",
  "Postpartum",           57.6, 1.57, 1,     37.0,           "Discussion (median postpartum WT 57.6, FFM 37.0)",
  "Non-pregnant female",  52.9, 1.57, 1,     35.7,           "Table S1 study data, non-pregnant females",
  "Male",                 54.4, 1.73, 0,     46.4,           "Table S1 study data, males",
  "Overall cohort",       58.0, 1.60, 1,     39.4,           "Table 1 overall participants (median WT/HT/FFM)"
) |>
  dplyr::mutate(
    ffm_formula = ffm_janmahasatian(wt, ht, sexf),
    pct_diff    = 100 * (ffm_formula - ffm_published) / ffm_published
  )

ffm_ref |>
  dplyr::transmute(
    Group = group, `WT (kg)` = wt, `HT (m)` = ht,
    `FFM published (kg)` = ffm_published,
    `FFM from formula (kg)` = round(ffm_formula, 1),
    `% diff` = round(pct_diff, 1), Source = source
  ) |>
  knitr::kable(caption = "Janmahasatian FFM reproduces the published group median FFM.")

stopifnot(max(abs(ffm_ref$pct_diff)) < 6)
```

# Reconstructing the paper's decomposition of the pregnancy effect

This is the sharpest available check on the covariate model, because the paper
reports the same quantity twice, computed two different ways.

The final model splits the pregnancy effect three ways: part is body size
(pregnant women are larger), part is renal function (pregnant women have lower
serum creatinine), and the remainder is the estimated +38.1% "pregnancy"
term. The Discussion then states that the *overall* clearance increase due to
pregnancy is "around 53%", the value obtained from an alternative model
(Table S3) that omits the serum-creatinine effect entirely.

So: evaluating the final model's three effects at the published antepartum and
postpartum median covariates must reconstruct that ~53%. Nothing in this
calculation is fitted to the 53% figure, and the two models were estimated
separately, so agreement is a genuine cross-check.

```{r pregnancy-decomposition}
cl_typical <- function(ffm, creat, preg) {
  6.05859 * (ffm / 39.4)^0.75 * (1 + 0.380866 * preg) * (creat / 56.2)^-0.366504
}

# Published medians for the matched antepartum / postpartum visits
# (Discussion: FFM 38.7 vs 37.0 kg; serum creatinine 45.3 vs 55.4 umol/L).
cl_ante <- cl_typical(ffm = 38.7, creat = 45.3, preg = 1)
cl_post <- cl_typical(ffm = 37.0, creat = 55.4, preg = 0)

contributions <- c(
  "Pregnancy term (+38.1%)"        = 1 + 0.380866,
  "Fat-free mass (38.7 vs 37.0 kg)" = (38.7 / 37.0)^0.75,
  "Serum creatinine (45.3 vs 55.4 umol/L)" = (45.3 / 55.4)^-0.366504
)

data.frame(
  Component = c(names(contributions), "Total CL ratio (antepartum / postpartum)"),
  `CL multiplier` = round(c(contributions, prod(contributions)), 4),
  check.names = FALSE
) |>
  knitr::kable(caption = "Decomposed pregnancy effect on levofloxacin clearance.")

total_ratio <- cl_ante / cl_post
cat(sprintf(
  "Reconstructed total CL increase in pregnancy: %.1f%%\nPaper's alternative model (Table S3, no creatinine effect): 53%%\n",
  100 * (total_ratio - 1)
))

# The three decomposed effects reproduce the independently-estimated 53%.
stopifnot(abs(total_ratio / 1.53 - 1) < 0.06)
```

The efficacy target drawn on Fig. 2 is likewise reproducible in closed form:
the caption defines `AUC0-24 >= 146 * MIC / fu` with the hollow-fibre
`fAUC0-24/MIC` target of 146, the wild-type MIC of 0.5 mg/L and an unbound
fraction of 69%.

```{r efficacy-target}
auc_target <- 146 * 0.5 / 0.69
cat(sprintf("146 * 0.5 / 0.69 = %.1f mg*h/L (paper states 106)\n", auc_target))
stopifnot(round(auc_target) == 106)
```

# Structural checks against closed-form results

These compare a solved profile against an algebraic result derived from the
published equations, with parameter values computed independently in R rather
than read back out of the model. Both sides use the same fixed parameter
values, so the residual difference is pure numerical / trapezoidal error and
the tolerances are tight.

```{r events-helper}
tau       <- 24
n_dose    <- 10                       # 10 daily doses: well past steady state
t_last    <- (n_dose - 1) * tau       # 216 h, the final ("observed") dose
grid_ss   <- seq(t_last, t_last + tau, by = 0.1)

#' Build a plain data.frame event table (never an rxEt object, which silently
#' drops post-hoc covariate column assignment).
#'
#' Occasions follow the paper's design: every dose before the final one is
#' self-administered at home (OCC = 1, the inflated BOV on F), the final dose
#' is the observed clinic dose (OCC = 2), the pre-dose trough belongs to
#' occasion 1 and the subsequent profile to occasion 2.
build_events <- function(cohort, obs_times = grid_ss) {
  doses <- tidyr::expand_grid(
    cohort,
    time = seq(0, t_last, by = tau)
  ) |>
    dplyr::mutate(
      evid = 1L,
      amt  = dose,
      cmt  = "depot",
      OCC  = ifelse(time < t_last, 1L, 2L)
    )
  obs <- tidyr::expand_grid(cohort, time = obs_times) |>
    dplyr::mutate(
      evid = 0L,
      amt  = NA_real_,
      cmt  = "central",
      OCC  = ifelse(time <= t_last, 1L, 2L)
    )
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::select(id, time, amt, evid, cmt, FFM, CREAT, PREG, OCC,
                  dplyr::any_of(c("group", "dose", "arm", "weight_band")))
}
```

```{r closed-form-cohort}
# Four deterministic typical subjects spanning the covariate space.
typ <- tibble::tribble(
  ~id, ~group,                ~FFM,  ~CREAT, ~PREG, ~dose,
  1L,  "Antepartum",          38.7,  45.3,   1,     1000,
  2L,  "Postpartum",          37.0,  55.4,   0,     1000,
  3L,  "Non-pregnant female", 35.7,  58.5,   0,     1000,
  4L,  "Male",                46.4,  64.1,   0,     1000
)

mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod_typ, build_events(typ), keep = c("group", "dose")) |>
  as.data.frame()
```

## Steady-state AUC identity

At steady state the amount eliminated over a dosing interval equals the amount
absorbed, so `CL * AUC(0-tau) = F * dose` exactly, whatever the absorption
model does. `CL` here is computed from the published covariate equations, not
taken from the solve.

```{r auc-identity}
auc_trapz <- function(t, c) sum(diff(t) * (head(c, -1) + tail(c, -1)) / 2)

id_check <- sim_typ |>
  dplyr::filter(time >= t_last) |>
  dplyr::group_by(id, group, dose) |>
  dplyr::summarise(auc_sim = auc_trapz(time, Cc), .groups = "drop") |>
  dplyr::left_join(typ, by = c("id", "group", "dose")) |>
  dplyr::mutate(
    cl_expected  = cl_typical(FFM, CREAT, PREG),
    auc_expected = dose / cl_expected,          # F = 1 (fixed)
    pct_diff     = 100 * (auc_sim - auc_expected) / auc_expected
  )

id_check |>
  dplyr::transmute(
    Group = group, `CL expected (L/h)` = round(cl_expected, 3),
    `AUC dose/CL (mg*h/L)` = round(auc_expected, 1),
    `AUC from solve (mg*h/L)` = round(auc_sim, 1),
    `% diff` = round(pct_diff, 3)
  ) |>
  knitr::kable(caption = "Steady-state AUC(0-24) equals F*dose/CL.")

stopifnot(max(abs(id_check$pct_diff)) < 0.5)
```

## Terminal slope recovers CL/Vc

By 14 h after the dose the transit chain and the depot are empty (`ktr` =
21/1.07 = 19.6 /h; depot half-life ln2/1.59 = 0.44 h), so the decline is
mono-exponential with rate `kel = CL/Vc`. Regressing `log(Cc)` on time over
14-24 h post-dose therefore tests `lvc` and the volume exponent `e_ffm_vc`,
which the AUC identity above cannot see.

```{r terminal-slope}
kel_check <- sim_typ |>
  dplyr::filter(time >= t_last + 14) |>
  dplyr::group_by(id, group) |>
  dplyr::summarise(
    kel_sim = -stats::coef(stats::lm(log(Cc) ~ time))[["time"]],
    .groups = "drop"
  ) |>
  dplyr::left_join(typ, by = c("id", "group")) |>
  dplyr::mutate(
    vc_expected  = 85.8803 * (FFM / 39.4)^1,
    kel_expected = cl_typical(FFM, CREAT, PREG) / vc_expected,
    pct_diff     = 100 * (kel_sim - kel_expected) / kel_expected
  )

kel_check |>
  dplyr::transmute(
    Group = group, `Vc expected (L)` = round(vc_expected, 1),
    `kel expected (1/h)` = round(kel_expected, 5),
    `kel from slope (1/h)` = round(kel_sim, 5),
    `t1/2 (h)` = round(log(2) / kel_sim, 2),
    `% diff` = round(pct_diff, 3)
  ) |>
  knitr::kable(caption = "Terminal slope recovers CL/Vc; half-life 6-8 h is the published range.")

stopifnot(max(abs(kel_check$pct_diff)) < 0.5)
```

The recovered half-lives sit inside the 6-8 h that the Introduction quotes for
individuals with normal renal function.

## Covariate exponents recovered from exposure ratios

Each ratio isolates one covariate by holding the others fixed, so the
simulated exposure ratio must equal the algebraic ratio exactly.

```{r covariate-ratios}
ratio_cohort <- tibble::tribble(
  ~id, ~arm,          ~FFM,  ~CREAT, ~PREG, ~dose,
  1L,  "reference",   39.4,  56.2,   0,     1000,
  2L,  "pregnant",    39.4,  56.2,   1,     1000,
  3L,  "creat x2",    39.4,  112.4,  0,     1000,
  4L,  "ffm x1.25",   49.25, 56.2,   0,     1000
) |>
  dplyr::mutate(group = arm)

sim_ratio <- rxode2::rxSolve(mod_typ, build_events(ratio_cohort),
                             keep = c("arm", "dose")) |>
  as.data.frame() |>
  dplyr::filter(time >= t_last) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(auc = auc_trapz(time, Cc), .groups = "drop")

auc_of <- function(a) sim_ratio$auc[match(a, sim_ratio$arm)]

# At steady state AUC = F*dose/CL, so the exposure ratio between two arms is
# the inverse ratio of their clearances: AUC(ref)/AUC(arm) = CL(arm)/CL(ref).
# Doubling creatinine LOWERS clearance (the exponent is negative), so that
# ratio is below 1; raising FFM RAISES clearance, so that ratio is above 1.
ratios <- tibble::tribble(
  ~Effect,                              ~observed,                                   ~expected,
  "Pregnancy: AUC(ref)/AUC(pregnant)",  auc_of("reference") / auc_of("pregnant"),    1 + 0.380866,
  "Creatinine x2: AUC(ref)/AUC(2x)",    auc_of("reference") / auc_of("creat x2"),    2^-0.366504,
  "FFM x1.25: AUC(ref)/AUC(1.25x)",     auc_of("reference") / auc_of("ffm x1.25"),   1.25^0.75
) |>
  dplyr::mutate(pct_diff = 100 * (observed - expected) / expected)

ratios |>
  dplyr::transmute(
    Effect, `Expected ratio` = round(expected, 5),
    `Simulated ratio` = round(observed, 5), `% diff` = round(pct_diff, 4)
  ) |>
  knitr::kable(caption = "Exposure ratios recover the published covariate exponents.")

stopifnot(max(abs(ratios$pct_diff)) < 0.2)
```

# Virtual cohort

The cohort mirrors the four participant groups in the numbers in which they
contributed profiles (19 antepartum, 14 postpartum, 12 non-pregnant female, 14
male), scaled 10-fold and capped at 200 per arm. Weight, height and serum
creatinine are drawn log-normally around each group's published median (Table
S1 study-data columns; postpartum medians from the Discussion) and truncated to
the published range; FFM is then derived per subject with the Janmahasatian
formula, and dose is assigned by WHO weight band. The dispersion of the drawn
covariates is an assumption -- Sawe 2025 reports medians and ranges, not
distributions -- and is listed under Assumptions below.

```{r cohort}
group_spec <- tibble::tribble(
  ~group,                ~n,   ~preg, ~sexf, ~wt_med, ~wt_lo, ~wt_hi, ~ht_med, ~creat_med, ~creat_lo, ~creat_hi,
  "Antepartum",          190L, 1,     1,     61.5,    43.0,   98.0,   1.60,    45.3,       37.0,      66.4,
  "Postpartum",          140L, 0,     1,     57.6,    45.0,   83.5,   1.57,    55.4,       25.3,      63.0,
  "Non-pregnant female", 120L, 0,     1,     52.9,    37.0,   67.9,   1.57,    58.5,       37.0,      110,
  "Male",                140L, 0,     0,     54.4,    41.9,   64.0,   1.73,    64.1,       52.3,      102
)

rxode2::rxSetSeed(20250401)   # seeded per stochastic block, not once globally
set.seed(20250401)

draw_group <- function(spec) {
  n <- spec$n
  tibble::tibble(
    group = spec$group,
    PREG  = spec$preg,
    SEXF  = spec$sexf,
    WT    = pmin(pmax(spec$wt_med    * exp(stats::rnorm(n, 0, 0.20)), spec$wt_lo),    spec$wt_hi),
    HT    = spec$ht_med * exp(stats::rnorm(n, 0, 0.045)),
    CREAT = pmin(pmax(spec$creat_med * exp(stats::rnorm(n, 0, 0.25)), spec$creat_lo), spec$creat_hi)
  )
}

cohort <- dplyr::bind_rows(
  lapply(seq_len(nrow(group_spec)), function(i) draw_group(group_spec[i, ]))
) |>
  dplyr::mutate(
    FFM         = ffm_janmahasatian(WT, HT, SEXF),
    weight_band = ifelse(WT > 50, "above 50 kg", "34-50 kg"),
    dose        = ifelse(WT > 50, 1000, 750),
    id          = dplyr::row_number()
  )

cohort |>
  dplyr::group_by(group) |>
  dplyr::summarise(
    n = dplyr::n(),
    `WT median` = round(stats::median(WT), 1),
    `FFM median` = round(stats::median(FFM), 1),
    `CREAT median` = round(stats::median(CREAT), 1),
    `% on 1000 mg` = round(100 * mean(dose == 1000)),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Simulated cohort versus the published group medians.")
```

```{r cohort-sim}
rxode2::rxSetSeed(20250402)
sim <- rxode2::rxSolve(
  mod,
  build_events(cohort, obs_times = seq(t_last, t_last + tau, by = 0.25)),
  keep = c("group", "dose", "weight_band")
) |>
  as.data.frame()
```

# Replicating Figure 1: VPC stratified by pregnancy status

Figure 1 of Sawe 2025 is a visual predictive check of concentration versus time
after dose, stratified into a pregnant and a non-pregnant stratum, where the
non-pregnant stratum pools the postpartum, never-pregnant-female and male
records. Its message is that median exposure is lower in the pregnant stratum.

```{r fig1, fig.cap = "Replicates Figure 1 of Sawe 2025: simulated 5th, 50th and 95th percentiles of levofloxacin concentration versus time after dose, stratified by pregnancy status."}
vpc_dat <- sim |>
  dplyr::mutate(
    tad     = time - t_last,
    stratum = ifelse(PREG == 1, "Pregnant", "Non-pregnant")
  ) |>
  dplyr::filter(!is.na(Cc))

vpc_pct <- vpc_dat |>
  dplyr::group_by(stratum, tad) |>
  dplyr::summarise(
    p05 = stats::quantile(Cc, 0.05),
    p50 = stats::median(Cc),
    p95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot2::ggplot(vpc_pct, ggplot2::aes(tad)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p05, ymax = p95), alpha = 0.25,
                       fill = "steelblue") +
  ggplot2::geom_line(ggplot2::aes(y = p50), linewidth = 0.9) +
  ggplot2::geom_line(ggplot2::aes(y = p05), linetype = "dashed") +
  ggplot2::geom_line(ggplot2::aes(y = p95), linetype = "dashed") +
  ggplot2::facet_wrap(~stratum) +
  ggplot2::labs(x = "Time after dose (h)", y = "Levofloxacin concentration (mg/L)") +
  ggplot2::theme_bw()
```

```{r fig1-check}
med_by_stratum <- vpc_pct |>
  dplyr::group_by(stratum) |>
  dplyr::summarise(auc_med = auc_trapz(tad, p50), .groups = "drop")

cat(sprintf(
  "Median-profile AUC(0-24): pregnant %.0f, non-pregnant %.0f mg*h/L (ratio %.2f)\n",
  med_by_stratum$auc_med[med_by_stratum$stratum == "Pregnant"],
  med_by_stratum$auc_med[med_by_stratum$stratum == "Non-pregnant"],
  med_by_stratum$auc_med[med_by_stratum$stratum == "Pregnant"] /
    med_by_stratum$auc_med[med_by_stratum$stratum == "Non-pregnant"]
))

# The paper's central qualitative finding: median exposure is lower in the
# pregnant stratum. Asserted on the ratio of median profiles, which is robust
# to which subjects land in the tails.
stopifnot(
  med_by_stratum$auc_med[med_by_stratum$stratum == "Pregnant"] <
    med_by_stratum$auc_med[med_by_stratum$stratum == "Non-pregnant"]
)
```

# PKNCA validation

```{r pknca}
conc_df <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, group, weight_band, dose, time, Cc)

dose_df <- cohort |>
  dplyr::transmute(id, group, weight_band, dose, time = t_last)

conc_obj <- PKNCA::PKNCAconc(
  data = conc_df, formula = Cc ~ time | group + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  data = dose_df, formula = dose ~ time | group + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = t_last, end = t_last + tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)
```

```{r pknca-summary}
nca_wide <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cav", "cmin")) |>
  dplyr::group_by(group, PPTESTCD) |>
  dplyr::summarise(
    median = stats::median(PPORRES),
    q25    = stats::quantile(PPORRES, 0.25),
    q75    = stats::quantile(PPORRES, 0.75),
    .groups = "drop"
  )

nca_wide |>
  dplyr::mutate(
    Parameter = dplyr::recode(PPTESTCD,
      cmax = "Cmax (mg/L)", tmax = "Tmax (h)", auclast = "AUC0-24 (mg*h/L)",
      cav = "Cavg (mg/L)", cmin = "Ctrough (mg/L)"
    ),
    `Median (IQR)` = sprintf("%.2f (%.2f-%.2f)", median, q25, q75)
  ) |>
  dplyr::select(Group = group, Parameter, `Median (IQR)`) |>
  tidyr::pivot_wider(names_from = Group, values_from = `Median (IQR)`) |>
  knitr::kable(caption = "Simulated steady-state NCA by participant group.")
```

The paper reports exposures for the pooled cohort: median (IQR) AUC0-24 131
(108-170) mg*h/L and Cmax 11.3 (9.68-13.7) mg/L (Results; Fig. S4).

```{r nca-comparison}
pooled <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast"))

reference <- data.frame(
  cmax    = 11.3,   # Results / Fig. S4, pooled observed median
  auclast = 131,
  tmax    = 2       # first post-dose sampling time; see Assumptions
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = pooled |> dplyr::select(PPTESTCD, PPORRES),
  reference = reference,
  units = c(cmax = "mg/L", auclast = "mg*h/L", tmax = "h")
)
knitr::kable(cmp, caption = "Simulated versus published pooled observed NCA.")
attr(cmp, "footnote")
```

```{r nca-checks}
pooled_med <- pooled |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(median = stats::median(PPORRES), .groups = "drop")

auc_med  <- pooled_med$median[pooled_med$PPTESTCD == "auclast"]
cmax_med <- pooled_med$median[pooled_med$PPTESTCD == "cmax"]

cat(sprintf("Pooled simulated median AUC0-24 %.0f mg*h/L (observed IQR 108-170)\n", auc_med))
cat(sprintf("Pooled simulated median Cmax    %.1f mg/L   (observed IQR 9.68-13.7)\n", cmax_med))

# Assert against the published interquartile range rather than the point
# median: the simulated cohort's covariate dispersion is an assumption, so the
# defensible claim is that the centre of the simulated distribution falls
# inside the observed IQR.
stopifnot(
  auc_med  > 108,  auc_med  < 170,
  cmax_med > 9.68, cmax_med < 13.7
)

# Table S2 range of median AUC0-24 (98.8-145) and Cmax (7.40-14.86) reported
# for adults with drug-resistant TB on 750-1000 mg; the paper's Fig. 2 ribbon.
lit_auc  <- c(98.8, 145)
lit_cmax <- c(7.40, 14.86)
stopifnot(cmax_med > lit_cmax[1], cmax_med < lit_cmax[2])

# Antepartum exposure must sit below all three non-pregnant groups.
group_auc <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(group) |>
  dplyr::summarise(median = stats::median(PPORRES), .groups = "drop")
knitr::kable(
  group_auc |> dplyr::transmute(Group = group, `Median AUC0-24 (mg*h/L)` = round(median)),
  caption = "Antepartum exposure is the lowest of the four groups."
)
stopifnot(
  group_auc$median[group_auc$group == "Antepartum"] ==
    min(group_auc$median)
)
```

# Replicating Figure 2: the +250 mg dose recommendation

Figure 2 plots simulated steady-state AUC0-24 and Cmax against weight band with
serum creatinine split into tertiles, adds a dashed box for the proposed extra
250 mg once daily in pregnancy, ribbons the literature range of reported
medians, and draws the 106 mg*h/L efficacy line. The paper concludes that a
250 mg increase brings pregnant exposures up to those of non-pregnant
participants across both weight bands.

```{r fig2-sim}
rxode2::rxSetSeed(20250403)

preg_extra <- cohort |>
  dplyr::filter(group == "Antepartum") |>
  dplyr::mutate(
    group = "Antepartum +250 mg",
    dose  = dose + 250,
    id    = id + 10000L
  )

cohort2 <- dplyr::bind_rows(cohort, preg_extra)

sim2 <- rxode2::rxSolve(
  mod,
  build_events(cohort2, obs_times = seq(t_last, t_last + tau, by = 0.5)),
  keep = c("group", "dose", "weight_band")
) |>
  as.data.frame()

nca2 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    data = sim2 |> dplyr::filter(!is.na(Cc)) |>
      dplyr::select(id, group, weight_band, time, Cc),
    formula = Cc ~ time | group + weight_band + id, concu = "mg/L", timeu = "h"
  ),
  PKNCA::PKNCAdose(
    data = cohort2 |> dplyr::transmute(id, group, weight_band, dose, time = t_last),
    formula = dose ~ time | group + weight_band + id, doseu = "mg"
  ),
  intervals = data.frame(start = t_last, end = t_last + tau,
                         cmax = TRUE, auclast = TRUE)
))

exposure <- as.data.frame(nca2$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::left_join(
    cohort2 |> dplyr::select(id, CREAT), by = "id"
  ) |>
  dplyr::group_by(group, weight_band) |>
  # Sawe 2025 describes the split as tertiles "within each group of
  # participants, based on the weight, sex, and pregnancy status", i.e. a
  # data-driven division into three equal parts. dplyr::ntile() does exactly
  # that and, unlike cut() on sample quantiles, tolerates the ties produced by
  # truncating the drawn creatinine values to the published range.
  dplyr::mutate(
    creat_tertile = factor(
      dplyr::ntile(CREAT, 3), levels = 1:3, labels = c("low", "medium", "high")
    )
  ) |>
  dplyr::ungroup()
```

```{r fig2, fig.height = 7, fig.cap = "Replicates Figure 2 of Sawe 2025: simulated steady-state AUC0-24 (top) and Cmax (bottom) by weight band and serum-creatinine tertile. Grey ribbon is the Table S2 literature range of reported adult medians; red dashed line is the 106 mg*h/L efficacy target."}
ribbons <- data.frame(
  PPTESTCD = c("auclast", "cmax"),
  ymin     = c(98.8, 7.40),
  ymax     = c(145, 14.86)
)
targets <- data.frame(PPTESTCD = "auclast", yint = auc_target)

exposure |>
  dplyr::mutate(
    panel = dplyr::recode(PPTESTCD, auclast = "AUC0-24 (mg*h/L)", cmax = "Cmax (mg/L)")
  ) |>
  ggplot2::ggplot(ggplot2::aes(creat_tertile, PPORRES, fill = group)) +
  ggplot2::geom_rect(
    data = ribbons |> dplyr::mutate(
      panel = dplyr::recode(PPTESTCD, auclast = "AUC0-24 (mg*h/L)", cmax = "Cmax (mg/L)")
    ),
    ggplot2::aes(ymin = ymin, ymax = ymax, xmin = -Inf, xmax = Inf),
    inherit.aes = FALSE, fill = "grey70", alpha = 0.35
  ) +
  ggplot2::geom_hline(
    data = targets |> dplyr::mutate(panel = "AUC0-24 (mg*h/L)"),
    ggplot2::aes(yintercept = yint), inherit.aes = FALSE,
    colour = "red", linetype = "dashed"
  ) +
  ggplot2::geom_boxplot(outlier.size = 0.4, position = ggplot2::position_dodge(0.8)) +
  ggplot2::facet_grid(panel ~ weight_band, scales = "free_y", switch = "y") +
  ggplot2::labs(x = "Serum creatinine tertile", y = NULL, fill = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom", strip.placement = "outside")
```

```{r fig2-checks}
auc_by_group <- exposure |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(group, weight_band) |>
  dplyr::summarise(median = stats::median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = group, values_from = median)

auc_by_group |>
  dplyr::mutate(dplyr::across(-weight_band, ~round(.x))) |>
  knitr::kable(caption = "Median simulated steady-state AUC0-24 by weight band.")

recovery <- auc_by_group |>
  dplyr::transmute(
    weight_band,
    base_gap     = 100 * (`Antepartum` / `Non-pregnant female` - 1),
    with_250     = 100 * (`Antepartum +250 mg` / `Non-pregnant female` - 1)
  )
knitr::kable(
  recovery |> dplyr::transmute(
    `Weight band` = weight_band,
    `Antepartum vs non-pregnant female, base dose (%)` = round(base_gap),
    `... with +250 mg (%)` = round(with_250)
  ),
  caption = "The proposed 250 mg increase closes most of the exposure gap."
)

# Antepartum exposure is below non-pregnant at the base dose, and the +250 mg
# arm closes most of the gap in both weight bands. The residual shortfall is
# expected: fully offsetting a 38.1% clearance increase needs a 1.381-fold
# dose, i.e. +286 mg on 750 mg and +381 mg on 1000 mg, and the paper chose
# 250 mg because it is the available tablet increment.
stopifnot(
  all(recovery$base_gap < -10),
  all(recovery$with_250 > recovery$base_gap),
  all(abs(recovery$with_250) < abs(recovery$base_gap))
)

for (i in seq_len(nrow(auc_by_group))) {
  cat(sprintf(
    "%-12s | antepartum %.0f | +250 mg %.0f | postpartum %.0f | non-preg F %.0f | male %.0f  (target %.0f)\n",
    auc_by_group$weight_band[i], auc_by_group$`Antepartum`[i],
    auc_by_group$`Antepartum +250 mg`[i], auc_by_group$`Postpartum`[i],
    auc_by_group$`Non-pregnant female`[i], auc_by_group$`Male`[i], auc_target
  ))
}

# The three non-pregnant arms clear the 106 mg*h/L efficacy target in both
# weight bands, and so does the pregnant arm once the proposed 250 mg is
# added. The pregnant arm on the CURRENT dose is deliberately NOT asserted to
# clear it: that is the paper's point. Its median lands just above the target
# in the above-50 kg band and below it in the 34-50 kg band, which is why the
# paper says AUC is larger than 106 in "most" subjects rather than all, and
# why it recommends the increase.
stopifnot(
  all(auc_by_group$`Postpartum`          > auc_target),
  all(auc_by_group$`Non-pregnant female` > auc_target),
  all(auc_by_group$`Male`                > auc_target),
  all(auc_by_group$`Antepartum +250 mg`  > auc_target)
)

# Antepartum on the current dose is the only arm at risk of missing the target,
# and the lighter weight band is where it misses.
stopifnot(
  auc_by_group$`Antepartum`[auc_by_group$weight_band == "34-50 kg"] < auc_target
)
```

Both weight bands behave as the paper describes: antepartum medians sit
materially below the non-pregnant-female medians on the same weight-based dose,
and adding 250 mg once daily recovers most of the shortfall without pushing
exposures above the range reported in the comparator literature.

# Assumptions and deviations

Where the paper did not state something the simulation needed, the choice is
recorded here.

* **Covariate dispersion in the virtual cohort.** Sawe 2025 reports medians and
  ranges (Table 1, Table S1), not distributions. Weight and serum creatinine
  are drawn log-normally at 20% and 25% CV respectively and height at 4.5% CV,
  each truncated to the published range. Group medians are matched (see the
  cohort table); the spread is an assumption and the interquartile ranges of
  the simulated NCA parameters should not be read as replications of the
  paper's.
* **Postpartum median covariates** come from the Discussion (weight 57.6 kg,
  FFM 37.0 kg, serum creatinine 55.4 umol/L) rather than Table 1, which splits
  postpartum across the two studies. Postpartum height is taken as 1.57 m, the
  midpoint of the two studies' medians (1.58 and 1.56 m).
* **Occasion assignment.** The model's `OCC` covariate is set to 1 for every
  self-administered dose before the final one and 2 for the observed clinic
  dose, with the pre-dose trough on occasion 1 and the subsequent profile on
  occasion 2. This mirrors the design described in the supplementary material
  ("occasion 1 is linked to the dose taken on the day prior to sampling and its
  associated concentration (trough sample)"), and it is what makes the inflated
  BOV on bioavailability apply to the trough only. A user simulating a
  prospective regimen in which every dose is observed should pass an even
  occasion number throughout.
* **Between-occasion variability degenerates within an arm.** Because each
  simulated subject is given a single occasion value per dose class rather than
  a fresh occasion per dose, the BOV on MTT, ka and F is drawn once per subject
  per class and behaves like between-subject variability within that class.
  That is the correct reading of the paper's design, where each participant
  contributed one observed profile, but it means the simulation does not show
  dose-to-dose absorption variability within a subject.
* **Tmax has no published counterpart.** The paper reports no NCA Tmax. The
  reference value of 2 h in the comparison table is the first post-dose
  sampling time in both studies, so the observed Tmax could not have been
  earlier than that; the row is informational and is not asserted on.
* **Simulated versus observed exposure.** The pooled simulated median AUC0-24
  and Cmax are asserted to fall inside the published observed interquartile
  ranges rather than to match the published medians, because the observed
  medians come from 59 profiles with the real joint covariate distribution and
  real dose assignment, while the simulation uses drawn covariates and a
  weight-band dose rule.

## Errata and omissions relative to the source

* **Table 2 footnote correction.** The article was published on 1 April 2025
  with an error in a Table 2 footnote and corrected on 17 April 2025. The
  corrected footnote *d* defines the reported variability percentages as
  `%CV = sqrt(omega^2) * 100`. This model file uses the corrected reading,
  verified three ways: against the footnote as typeset in the corrected PDF,
  against the supplementary equation image, and arithmetically against the
  control stream `$OMEGA` values (see the variance-scale table above). Applying
  the more common `omega^2 = log(CV^2 + 1)` conversion instead would understate
  the BOV on ka by 26% on the variance scale.
* **Additive residual error is 0.259443 mg/L, not the 0.244 mg/L of Table 2.**
  The control stream's `$ERROR` block sets `ADD = THETA(6) + (LLOQ*0.2)`
  unconditionally on every record, with LLOQ 0.0781 mg/L, so the additive
  standard deviation actually applied is `0.243823 + 0.01562`. The same
  20%-of-LLOQ floor appears in the sibling paediatric model
  `Denti_2018_levofloxacin`.
* **BLQ-handling error inflations are omitted.** The control stream adds a
  further 50% of the LLOQ to the additive error for the imputed BLQ records
  (Beal's M6) and substitutes a huge fixed additive term to drop trailing BLQ
  records from the fit. Both are conditional on the `CENS` data column, apply
  to no simulated observation, and are deliberately not encoded.
* **Peripheral-compartment parameters are absent.** The control stream retains
  `$THETA` and `$OMEGA` slots for `V3`, `Q`, `V4` and `Q2`, all fixed to zero:
  the residue of testing a second and third compartment. Adding a second
  compartment did not improve the fit (dOFV 0.674, 2 df), so the packaged model
  is one-compartment, as Table 2 reports.
* **Between-subject variability is on clearance only.** Every other `$OMEGA` in
  the control stream is `BLOCK(1) FIX 0`, and between-visit variability on
  clearance was tested and not retained. No IIV has been invented for volume or
  the absorption parameters.
* **No `f(depot) <- 0` line, despite the control stream's `F1 = 0`.** The
  NONMEM stream sets `F1 = 0` so the dose enters only through the analytic
  transit input rate rather than as a bolus into the absorption compartment.
  Transcribing that literally as `f(depot) <- 0` is the natural move, and
  several older transit models in this package do exactly that, but under
  `rxUi` with rxode2 5.1.7 it silently zeroes the `transit()` input as well and
  the model absorbs nothing -- every simulated concentration comes back 0, with
  no error or warning. Omitting the line gives the intended behaviour, not a
  double-counted dose: `transit()` is already the sole input pathway under
  `rxUi`. The steady-state AUC identity above is the standing regression gate,
  and a single-dose check gives `AUC(0-inf) * CL` = 999.98 mg against a 1000 mg
  dose with F fixed at 1.
* **The transit input accounts only for the most recent dose.** The published
  analytic Savic form tracks a single dose time (`TDOS`/`PNXD` in the control
  stream), so contributions from earlier doses are not superposed. This is
  faithful to the source and numerically immaterial here: with `ktr` = 19.6 /h
  the chain delivers essentially the whole dose within a few hours of a 24-hour
  interval.
* **Screened-but-unretained covariates** (total body weight as an alternative
  size descriptor, sex, age, albumin, HIV status, gestational age, study and
  study site) are recorded in the model file's `covariatesDataExcluded` for
  provenance. None is referenced in `model()`.
* **Gestational age.** Only third-trimester data were collected, and gestational
  age was tested within the pregnant participants and found not significant, so
  the model must not be used to interpolate a pregnancy effect at earlier
  gestation.
* **Serum creatinine extrapolation.** Values were measured within two weeks of
  the PK visit rather than on the day, and no validated renal-function equation
  exists for pregnancy; the authors caution against extrapolating outside the
  observed 25.3-110 umol/L range.
