---
title: "Balovaptan (Schaedeli Stark 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Balovaptan (Schaedeli Stark 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("SchaedeliStark_2024_balovaptan"))$reference`
* Description: `r rxode2::rxode(readModelDb("SchaedeliStark_2024_balovaptan"))$description`
* Article: <https://doi.org/10.1007/s10928-023-09898-0> (open access; PMCID PMC11136808)
* Supplement: online supplementary material to the same DOI (goodness-of-fit
  figures plus the development-stage explicit-binding and empirical-binding
  two-compartment models; it carries no final-model parameters)

```{r model}
mod <- rxode2::rxode(readModelDb("SchaedeliStark_2024_balovaptan"))
mod
```

# Population

The model was fitted to 3985 balovaptan plasma concentrations from 370 adult
and pediatric individuals across five studies (Table 1 of the source). Three
phase I studies contributed richly sampled neurotypical adults -- NCT01418963
(n = 24, multiple-ascending dose 12 to 52 mg once daily for 14 days),
NCT03579719 (n = 15, 5 mg once daily fed) and NCT03586726 (n = 16, 10 mg once
daily fasted). Two phase II studies in autism spectrum disorder (ASD) with
IQ >= 70 contributed the bulk of the subjects but sparse, mainly steady-state
sampling: VANILLA (NCT01793441, n = 146 adult men, 1.5 / 4 / 10 mg once daily
with food) and aV1ation (NCT02901431, n = 169 children and adolescents aged 5
to 17 years, age-adjusted doses of 1.5 to 10 mg once daily).

Overall the cohort was 91.9% male, with a median age of 19 years (range 5 to
64) and a median body weight of 72.8 kg (range 18.8 to 152.0). Only 4.3% of the
participants with PK data were dosed fasted, which is why the food effect on
mean transit time had to be fixed rather than estimated. Doses spanned 1.5 to
52 mg once daily; all fitted data were oral. The intravenous and receptor
occupancy work described below is *simulation* from this model for a planned
phase II malignant cerebral edema trial (NCT05399550) -- no IV data were fitted.

```{r population}
str(readModelDb("SchaedeliStark_2024_balovaptan")()$population)
```

# Model structure

One-compartment disposition with transit-compartment absorption, an empirical
saturable-binding model that makes the central volume a decreasing function of
the central amount, and an empirical gut-extraction process gated by a slow
turnover pool (Fig. 3 and p. 234 of the source):

```
d/dt(depot)      = Input - ktr*depot - kgutex*moderator1*depot
d/dt(transit_i)  = ktr*transit_{i-1} - ktr*transit_i          (i = 1..6)
d/dt(central)    = ktr*transit6 - ke*central
d/dt(moderator1) = kin - kout*moderator1*(1 + smod*depot),   moderator1(0) = 1

ktr = (ntr + 1)/mtt          mtt = MTT_fasted * 3.39^FED
ke  = cl/vc                  cl  = CL / (1 + exp(AGEslope*(AGE50 - AGE)))
vc  = V0 * (WT/76)^1.00 * (1 - vbmax*central/(central + vba50))
Cc  = 1000 * central/vc      RO  = 100*Cc/(kd_fu + Cc)
```

Two features drive all of the non-linearity. The **dynamic central volume**
shrinks by up to 80.5% as the central amount rises, which makes both Cmax and
Cmin more than dose-proportional. The **gut extraction** term competes with
absorption out of the depot; the turnover pool `moderator1` starts at 1, is
suppressed by the amount in the depot (via `smod`), and recovers only slowly
(`kout` = 0.00205 1/h, a 338 h half-life). Extraction is therefore large on the
first dose and small at steady state, and -- because a *small* dose suppresses
the pool less -- it removes proportionally more drug at low doses. That is the
mechanism behind the source's statement of "lower oral bioavailability at low
balovaptan doses".

# Source trace

Every `ini()` value and every non-obvious `model()` equation, with its location
in the source.

| Model quantity | Value | Source location |
| --- | --- | --- |
| `lcl` (CL/F) | 8.52 L/h | Table 2, row "CL (L/h)" |
| `lvc` (V0) | 565 L | Table 2, row "V0 (L)" |
| `lmtt` (MTT, fasted) | 0.367 h | Table 2, row "MTT (h)" |
| `ntr` | 5.81 | Table 2, row "Ntr" |
| `lvbmax` (VLmax) | 0.805 | Table 2, row "VLmax (fraction)" |
| `lvba50` (VLA50) | 3.36 mg | Table 2, row "VLA50"; unit corrected, see Errata |
| `lkgutex` (Kgut) | 1.89 1/h | Table 2, row "Kgut (1/h)" |
| `lkout` (Kout) | 0.00205 1/h, fixed | Table 2, row "Kout (1/h)" |
| `lsmod` (S) | 22.0 1/mg | Table 2, row "S" |
| `e_wt_vc` | 1.00, fixed | Table 2, row "WT V0"; Results p. 234 |
| `e_fed_mtt` | 3.39, fixed | Table 2, row "FOOD MTT"; see Errata |
| `e_age_cl` (AGEslope) | 0.267 1/year | Table 2, row "AGEslope" |
| `age50_cl` (AGE50) | 5.34 years | Table 2, row "AGE50 (years)" |
| `lkd_fu` (Kb/fu_plasma) | 3.16 ng/mL | Back-solved from Table 3; see Errata |
| `etalcl`, `etalvc`, `etalmtt` | 0.150, 0.0944, 0.0423 | Table 2, rows "IIV CL/V0/MTT" and footnote b |
| `propSd_early`, `propSd_late`, `propSd_exp_kdes` | 1.20, 0.216, 1.32 1/h | Table 2, rows "RUV early/late/rate" and footnote |
| `addSd` | 0.025 ng/mL, fixed | Table 2, row "RUV add" and footnote a |
| `addSd_RO` | 0, fixed | Not reported; RO was simulated, not fitted |
| Depot / transit / central ODEs | -- | p. 234 equations; transit chain per Fig. 3 |
| Turnover ODE, `moderator1(0) = 1` | -- | p. 234 equation; Results p. 233 |
| `ktr = (ntr + 1)/mtt`, `ke = cl/vc` | -- | p. 234 "micro-constants of mass transfer" |
| `vc` empirical binding | -- | p. 234 and Fig. 3 caption |
| `cl` age maturation | -- | p. 234; sign corrected, see Errata |
| `RO` | -- | Methods p. 230 |
| `Cc = 1000 * central/vc` | -- | Unit conversion, mg/L to ng/mL |

# Assumptions and deviations (Errata)

The source contains three typesetting defects and one reporting gap. Each was
resolved against the paper's own numbers rather than assumed; the discriminating
evidence is reproduced in the validation sections below.

**1. The printed age-clearance equation is missing a `1 +`.** As printed
(p. 234) it reads `CLi = CL * (1 - 1/exp(-AGEslope*(AGE50 - AGE)))`, which
returns **-143.9%** of adult clearance at age 2 and **exactly 0%** at AGE50 --
directly contradicting Table 2's own definition of AGE50 as the "age where 50%
of adult CL/F is reached". Restoring a dropped `1 +` in the denominator gives
the logistic `CL/(1 + exp(AGEslope*(AGE50 - AGE)))`, which reproduces four
independent printed anchors (checked below).

**2. Table 2 prints VLA50 in `ug`, but the value is in mg.** Read as 3.36 ug,
the binding term is saturated at every clinical dose, the central volume becomes
a constant, and the model loses the entire non-linearity it was built to
describe. Read as 3.36 mg it reproduces Table 3 to a median of 1.3%; the ug
reading is off by up to 69% (checked below). The mg reading is also the only one
dimensionally consistent with the `smod * depot` term, which needs 1/mg.

**3. The food effect on MTT is multiplicative, not an absolute 3.39 h.** Table 2
lists "FOOD MTT -- Effect of food on MTT -- 3.39 (Fixed)", exactly parallel to
its sibling row "WT V0 -- Effect of body weight on V0 -- 1.00 (Fixed)", which is
unambiguously a multiplicative exponent. The Results text instead calls it "an
estimate of 3.39 h". The multiplicative reading gives a fed Tmax of 2.4 h and an
AUCss span of -9.6% to +14.6%, matching Fig. 6A and the paper's own printed
"-10% and +15%" claim; the additive and absolute readings give Tmax of 6.2 h and
5.7 h and AUCss spans of -32%/+23% and -29%/+22% (checked below).

**4. Neither `Kb` nor `fu_plasma` is reported anywhere.** The receptor occupancy
equation is printed but both of its constants are absent from the paper, the
supplement and the Fig. 8 panel. Only their **ratio** is identifiable, and
Table 3's seven paired (concentration, occupancy) medians over-determine it: the
intersection of the rounding intervals implied by all seven rows is
[3.076, 3.446] ng/mL. The model ships `kd_fu` = 3.16 ng/mL, the median of the
seven row-wise point estimates. **This value is not from the paper text or
tables** -- it is back-solved arithmetic on printed medians.

Two further implementation notes:

* **Transit chain length.** `Ntr` = 5.81 is not an integer, so the chain is
  realised with 6 transit compartments (7 transfers). The printed
  `ktr = (Ntr + 1)/MTT` relation is kept intact, so both published values are
  used as printed and the realised mean transit time is 2.8% longer than the
  nominal MTT. Realising 5 transit compartments instead moves Tmax to 2.16 h,
  further from Fig. 6A.
* **Where gut extraction acts.** The printed ODE set elides the transit chain
  entirely (it shows only `dAa/dt` and `dAc/dt`), which would leave `Ntr` and
  `MTT` jointly unidentifiable even though Table 2 reports both with tight
  bootstrap intervals. Fig. 3 settles the structure: the "TCAM" label sits on
  the absorption-to-central arrow, and the `Kgut x At` arrow leaves the
  absorption box only. Applying extraction to the whole chain instead blows the
  AUCss span out to -52%/+163% (checked below).
* **Time-varying residual error.** The printed RUV expression omits the minus
  sign in its exponent, which would send the SD to infinity rather than decay
  it; the Results text ("mono-exponential decay from a high value immediately
  after dosing to a lower level ... half-life of 0.53 h", and ln(2)/1.32 = 0.53)
  fixes the sign unambiguously.

# Simulation helper

Two endpoints (`Cc` and `RO`) means observation records are tagged with `dvid`
rather than `cmt`, so that every solve returns both endpoint columns.

```{r helper}
baloEvents <- function(doseTimes, doseAmts, doseCmt, obsTimes,
                       WT = 76, AGE = 45, FED = 0, dur = NA_real_) {
  dosing <- data.frame(
    time = doseTimes, amt = doseAmts, evid = 1L, dur = dur,
    cmt = doseCmt, dvid = NA_integer_
  )
  obs <- data.frame(
    time = obsTimes, amt = NA_real_, evid = 0L, dur = NA_real_,
    cmt = NA_character_, dvid = 1L
  )
  out <- rbind(dosing, obs)
  out <- out[order(out$time, -out$evid), ]
  out$id <- 1L
  out$WT <- WT
  out$AGE <- AGE
  out$FED <- FED
  out
}

# Typical-value (all etas zero) version of the model: the source's Figs. 5-7 and
# Table 3 are all "typical individual" simulations, so these gates are
# deterministic and can be asserted tightly.
modTypical <- rxode2::zeroRe(mod)
```

# Validation 1: Table 3, intravenous stepped dosing and receptor occupancy

Table 3 of the source reports simulated median plasma concentrations and brain
V1a receptor occupancies at seven time points following three 60-minute IV
infusions of 50, 25 and 15 mg on days 1, 2 and 3, for a typical 76 kg adult.
Because the IV route bypasses absorption entirely, this gate isolates the
disposition model, the empirical binding term and the occupancy relationship.

```{r table3}
ivEvents <- baloEvents(
  doseTimes = c(0, 24, 48),
  doseAmts  = c(50, 25, 15),
  doseCmt   = "central",
  dur       = 1,
  obsTimes  = sort(unique(c(seq(0, 120, by = 0.05), c(1, 24, 25, 48, 49, 72, 96))))
)
ivSim <- rxode2::rxSolve(modTypical, ivEvents, returnType = "data.frame")

table3 <- data.frame(
  time    = c(1, 24, 25, 48, 49, 72, 96),
  Cc_pub  = c(352, 74, 261, 59, 166, 42, 15),
  RO_pub  = c(99, 96, 99, 95, 98, 93, 82)
) %>%
  dplyr::left_join(ivSim[, c("time", "Cc", "RO")], by = "time") %>%
  dplyr::mutate(
    Cc_pct_diff = 100 * (Cc - Cc_pub) / Cc_pub,
    RO_diff_pp  = RO - RO_pub
  )

table3 %>%
  dplyr::mutate(
    dplyr::across(c(Cc, Cc_pct_diff, RO_diff_pp), ~ round(.x, 2)),
    RO = round(RO, 1)
  ) %>%
  dplyr::rename(
    "Time (h)"                = time,
    "Published Cc (ng/mL)"    = Cc_pub,
    "Simulated Cc (ng/mL)"    = Cc,
    "Cc difference (%)"       = Cc_pct_diff,
    "Published RO (%)"        = RO_pub,
    "Simulated RO (%)"        = RO,
    "RO difference (points)"  = RO_diff_pp
  ) %>%
  knitr::kable(caption = "Replicates Table 3 of Schaedeli Stark et al. (2024), 76 kg reference column.")
```

The reconstruction reproduces all seven published concentrations to a median of
1.3% and all seven occupancies to within 0.2 percentage points.

```{r table3-assert}
stopifnot(
  median(abs(table3$Cc_pct_diff)) < 3,
  max(abs(table3$Cc_pct_diff)) < 6,
  max(abs(table3$RO_diff_pp)) < 0.5
)
```

## The VLA50 unit reading, falsified

Re-running the identical gate with VLA50 read as 3.36 **ug** rather than mg:

```{r vla50}
modUg <- modTypical %>% rxode2::ini(lvba50 = log(3.36e-3))
ugSim <- rxode2::rxSolve(modUg, ivEvents, returnType = "data.frame")
ugDiff <- 100 * (ugSim$Cc[match(table3$time, ugSim$time)] - table3$Cc_pub) / table3$Cc_pub

data.frame(
  Reading = c("VLA50 = 3.36 mg", "VLA50 = 3.36 ug"),
  Median  = round(c(median(abs(table3$Cc_pct_diff)), median(abs(ugDiff))), 1),
  Worst   = round(c(max(abs(table3$Cc_pct_diff)), max(abs(ugDiff))), 1)
) %>%
  dplyr::rename(
    "Median absolute difference (%)" = Median,
    "Worst absolute difference (%)"  = Worst
  ) %>%
  knitr::kable(caption = "Table 3 reproduction under the two VLA50 unit readings.")

stopifnot(median(abs(ugDiff)) > 5 * median(abs(table3$Cc_pct_diff)))
```

# Validation 2: Fig. 6A, dose-normalized steady-state profiles

Fig. 6A shows dose-normalized concentration-time profiles at steady state for a
typical adult participant at 1.5, 4, 10, 20 and 50 mg once daily. Both phase II
ASD studies dosed with food, so these are fed profiles (`FED = 1`).

```{r fig6a-sim}
oralDoses <- c(1.5, 4, 10, 20, 50)

simSteadyState <- function(dose, model, ...) {
  ev <- baloEvents(
    doseTimes = seq(0, 13 * 24, by = 24),
    doseAmts  = rep(dose, 14),
    doseCmt   = "depot",
    obsTimes  = seq(13 * 24, 14 * 24, by = 0.05),
    FED       = 1,
    ...
  )
  out <- rxode2::rxSolve(model, ev, returnType = "data.frame")
  out <- out[out$time >= 13 * 24, ]
  out$time <- out$time - 13 * 24
  out$dose_mg <- dose
  out
}

fig6a <- dplyr::bind_rows(lapply(oralDoses, simSteadyState, model = modTypical))

ggplot2::ggplot(
  fig6a,
  ggplot2::aes(time, Cc / dose_mg, colour = factor(dose_mg))
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  ggplot2::labs(
    x = "Time (h)",
    y = "Dose-normalized concentration (ng/mL/mg)",
    colour = "Dose (mg)",
    title = "Replicates Figure 6A of Schaedeli Stark et al. (2024)"
  ) +
  ggplot2::theme_bw()
```

The published panel shows the characteristic signature of the empirical binding
model: dose-normalized peaks that *rise* with dose, curves that cross at about
16 to 17 hours, and dose-normalized troughs that *fall* with dose.

## NCA comparison against Fig. 6A

NCA is computed with `PKNCA` over the final 24-hour dosing interval. The
reference values are read from Fig. 6A (the source publishes no NCA table, so
these are figure-derived and are labelled as such).

```{r pknca-typical}
ncaConc <- fig6a %>%
  dplyr::filter(!is.na(Cc)) %>%
  dplyr::mutate(
    id       = paste0("typical_", dose_mg),
    dose_lbl = paste0(dose_mg, " mg")
  ) %>%
  dplyr::select(id, dose_lbl, dose_mg, time, Cc)

ncaDose <- ncaConc %>%
  dplyr::group_by(id, dose_lbl, dose_mg) %>%
  dplyr::summarise(time = 0, .groups = "drop")

oConc <- PKNCA::PKNCAconc(ncaConc, Cc ~ time | id / dose_lbl)
oDose <- PKNCA::PKNCAdose(ncaDose, dose_mg ~ time | id + dose_lbl)
oData <- PKNCA::PKNCAdata(
  oConc, oDose,
  intervals = data.frame(
    start = 0, end = 24,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
  )
)
ncaRes <- PKNCA::pk.nca(oData)

# Fig. 6A dose-normalized peak concentrations (ng/mL/mg), read from the panel,
# converted to absolute Cmax; Tmax is read as ~2.5 h at every dose.
fig6aRead <- data.frame(
  dose_lbl = paste0(oralDoses, " mg"),
  cmax     = c(4.9, 5.6, 6.65, 7.75, 8.85) * oralDoses,
  tmax     = rep(2.5, length(oralDoses))
)

ncaTable <- nlmixr2lib::ncaComparisonTable(
  simulated = ncaRes,
  reference = fig6aRead,
  by        = "dose_lbl",
  params    = c("cmax", "tmax"),
  units     = c(cmax = "ng/mL", tmax = "h")
)
ncaTable %>%
  dplyr::rename("Dose" = dose_lbl) %>%
  knitr::kable(
    caption = "Simulated steady-state NCA versus values read from Figure 6A."
  )
attr(ncaTable, "footnote")
```

# Validation 3: Fig. 6B and the printed AUCss span

The source states in its Discussion that "median systemic availability across
the dosing range of 1.5 mg to 50.0 mg QD is expected to remain within -10% and
+15% of the reference adult dose of 10 mg". This is a printed claim, not a
figure read, and it is a sharp test of the gut-extraction component: the
non-linearity has to be present, but only weakly.

```{r fig6b}
aucss <- fig6a %>%
  dplyr::group_by(dose_mg) %>%
  dplyr::summarise(
    auc = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    auc_dn  = auc / dose_mg,
    pct_ref = 100 * (auc_dn / auc_dn[dose_mg == 10] - 1),
    # At steady state AUCss = F * Dose / CL exactly, so this recovers the
    # fraction of each dose that escapes gut extraction.
    f_escape = auc_dn * 8.52 / 1000
  )

aucss %>%
  dplyr::mutate(dplyr::across(c(auc_dn, pct_ref, f_escape), ~ round(.x, 3))) %>%
  dplyr::rename(
    "Dose (mg)"                        = dose_mg,
    "AUCss (h*ng/mL)"                  = auc,
    "Dose-normalized AUCss (h*ng/mL/mg)" = auc_dn,
    "Difference from 10 mg (%)"        = pct_ref,
    "Fraction escaping gut extraction" = f_escape
  ) %>%
  knitr::kable(caption = "Replicates Figure 6B of Schaedeli Stark et al. (2024); the source's own Fig. 6B medians run from about 88 to 112 h*ng/mL/mg over this dose range.")
```

The simulated span is -9.6% to +14.6% against the source's printed -10% and
+15%. The final column is an independent mass-balance check: at steady state
`AUCss = F * Dose / CL`, so dividing the dose-normalized AUCss by 1/CL recovers
the fraction of each dose escaping gut extraction. It rises from 0.76 at 1.5 mg
to 0.96 at 50 mg, which is exactly the source's "lower oral bioavailability at
low balovaptan doses", and it stays inside (0, 1) as any bioavailability must.

```{r fig6b-assert}
stopifnot(
  # The printed claim, reproduced to better than one percentage point at both ends.
  abs(aucss$pct_ref[aucss$dose_mg == 1.5] - (-10)) < 1,
  abs(aucss$pct_ref[aucss$dose_mg == 50] - (+15)) < 1,
  # Mass balance: a bioavailable fraction, increasing with dose.
  all(aucss$f_escape > 0), all(aucss$f_escape < 1),
  !is.unsorted(aucss$f_escape)
)
```

## The food-effect and gut-extraction readings, falsified

```{r alternatives}
spanFor <- function(model) {
  a <- dplyr::bind_rows(lapply(c(1.5, 10, 50), simSteadyState, model = model)) %>%
    dplyr::group_by(dose_mg) %>%
    dplyr::summarise(
      auc_dn = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2) / dose_mg[1],
      .groups = "drop"
    )
  round(100 * (a$auc_dn / a$auc_dn[a$dose_mg == 10] - 1), 1)[c(1, 3)]
}
tmaxFor <- function(model) {
  s <- simSteadyState(10, model)
  s$time[which.max(s$Cc)]
}

# The food effect enters as mtt * 3.39^FED. An additive or absolute reading
# cannot be expressed by changing a parameter value, so each is built as its own
# rxode2 model sharing every other line with the packaged one.
modAdditive <- rxode2::rxode({
  cl <- 8.52 / (1 + exp(0.267 * (5.34 - AGE)))
  vc <- 565 * (WT / 76) * (1 - 0.805 * central / (central + 3.36))
  mtt <- 0.367 + 3.39 * FED
  ktr <- (5.81 + 1) / mtt
  ke <- cl / vc
  moderator1(0) <- 1
  d/dt(depot) <- -ktr * depot - 1.89 * moderator1 * depot
  d/dt(transit1) <- ktr * depot - ktr * transit1
  d/dt(transit2) <- ktr * transit1 - ktr * transit2
  d/dt(transit3) <- ktr * transit2 - ktr * transit3
  d/dt(transit4) <- ktr * transit3 - ktr * transit4
  d/dt(transit5) <- ktr * transit4 - ktr * transit5
  d/dt(transit6) <- ktr * transit5 - ktr * transit6
  d/dt(central) <- ktr * transit6 - ke * central
  d/dt(moderator1) <- 0.00205 - 0.00205 * moderator1 * (1 + 22 * depot)
  Cc <- 1000 * central / vc
})

modAllChain <- rxode2::rxode({
  cl <- 8.52 / (1 + exp(0.267 * (5.34 - AGE)))
  vc <- 565 * (WT / 76) * (1 - 0.805 * central / (central + 3.36))
  mtt <- 0.367 * 3.39^FED
  ktr <- (5.81 + 1) / mtt
  ke <- cl / vc
  moderator1(0) <- 1
  d/dt(depot) <- -ktr * depot - 1.89 * moderator1 * depot
  d/dt(transit1) <- ktr * depot - ktr * transit1 - 1.89 * moderator1 * transit1
  d/dt(transit2) <- ktr * transit1 - ktr * transit2 - 1.89 * moderator1 * transit2
  d/dt(transit3) <- ktr * transit2 - ktr * transit3 - 1.89 * moderator1 * transit3
  d/dt(transit4) <- ktr * transit3 - ktr * transit4 - 1.89 * moderator1 * transit4
  d/dt(transit5) <- ktr * transit4 - ktr * transit5 - 1.89 * moderator1 * transit5
  d/dt(transit6) <- ktr * transit5 - ktr * transit6 - 1.89 * moderator1 * transit6
  d/dt(central) <- ktr * transit6 - ke * central
  d/dt(moderator1) <- 0.00205 - 0.00205 * moderator1 * (1 + 22 * depot)
  Cc <- 1000 * central / vc
})

alt <- data.frame(
  Reading = c(
    "As packaged: food x3.39, extraction on depot only",
    "Food ADDITIVE (MTT + 3.39 h)",
    "Gut extraction on depot AND all six transits"
  ),
  Low  = c(spanFor(modTypical)[1], spanFor(modAdditive)[1], spanFor(modAllChain)[1]),
  High = c(spanFor(modTypical)[2], spanFor(modAdditive)[2], spanFor(modAllChain)[2]),
  Tmax = round(c(tmaxFor(modTypical), tmaxFor(modAdditive), tmaxFor(modAllChain)), 2)
)
alt %>%
  dplyr::rename(
    "AUCss at 1.5 mg vs 10 mg (%)" = Low,
    "AUCss at 50 mg vs 10 mg (%)"  = High,
    "Tmax at 10 mg (h)"            = Tmax
  ) %>%
  knitr::kable(caption = "Only the packaged reading reproduces the printed -10% / +15% span and the Fig. 6A Tmax of about 2.5 h.")

stopifnot(
  abs(alt$Low[2]) > 20, abs(alt$High[3]) > 100,
  alt$Tmax[2] > 5, alt$Tmax[1] > 2 && alt$Tmax[1] < 3
)
```

# Validation 4: Fig. 7, the age effect on clearance

Fig. 7 plots the proportion of adult CL/F against age. The source states that
about 90% of adult CL/F is reached at 14 years and that the curve plateaus at
approximately 20 years; Table 2 defines AGE50 as the age at 50%; and the plotted
curve begins at about 0.29 at age 2.

```{r fig7}
ageGrid <- seq(2, 65, by = 0.1)
logistic <- 1 / (1 + exp(0.267 * (5.34 - ageGrid)))
asPrinted <- 1 - 1 / exp(-0.267 * (5.34 - ageGrid))

ggplot2::ggplot(
  data.frame(age = ageGrid, frac = logistic),
  ggplot2::aes(age, frac)
) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = c(0.5, 0.9), linetype = "dashed") +
  ggplot2::scale_x_continuous(trans = "log10", breaks = c(2, 4, 7, 10, 14, 20, 30, 50)) +
  ggplot2::labs(
    x = "Age (years)", y = "Proportion of adult CL/F",
    title = "Replicates Figure 7 of Schaedeli Stark et al. (2024)"
  ) +
  ggplot2::theme_bw()

anchorAges <- c(2, 5.34, 14, 20)
ageCheck <- data.frame(
  age       = anchorAges,
  anchor    = c("Fig. 7 curve starts at about 0.29",
                "Table 2: AGE50 is the age at 50% of adult CL/F",
                "Results: about 90% of adult CL/F at 14 years",
                "Results: plateau at approximately 20 years"),
  logistic  = round(1 / (1 + exp(0.267 * (5.34 - anchorAges))), 3),
  as_printed = round(1 - 1 / exp(-0.267 * (5.34 - anchorAges)), 3)
)
ageCheck %>%
  dplyr::rename(
    "Age (years)"                = age,
    "Published anchor"           = anchor,
    "Logistic (as implemented)"  = logistic,
    "Equation exactly as printed" = as_printed
  ) %>%
  knitr::kable(caption = "The dropped `1 +` in the printed age-clearance equation.")
```

The logistic reading hits every anchor. The equation exactly as printed returns
a negative clearance at age 2 and exactly zero at AGE50, so it cannot be what
was fitted.

```{r fig7-assert}
stopifnot(
  abs(ageCheck$logistic[1] - 0.29) < 0.01,
  abs(ageCheck$logistic[2] - 0.50) < 1e-8,
  abs(ageCheck$logistic[3] - 0.90) < 0.02,
  ageCheck$logistic[4] > 0.97,
  # The printed form is not merely different, it is impossible.
  ageCheck$as_printed[1] < 0,
  abs(ageCheck$as_printed[2]) < 1e-8
)
```

The source's dosing recommendation follows directly: adult doses from age 10,
70% of the adult dose for ages 5 to 9, and 40% for ages 2 to 4.

```{r pediatric}
pedGroups <- data.frame(
  group = c("2-4 years", "5-9 years", "10-17 years"),
  lower = c(2, 5, 10), upper = c(4, 9, 17),
  recommended_pct = c(40, 70, 100)
)
pedGroups$mean_cl_pct <- round(vapply(seq_len(nrow(pedGroups)), function(i) {
  a <- seq(pedGroups$lower[i], pedGroups$upper[i], by = 0.1)
  100 * mean(1 / (1 + exp(0.267 * (5.34 - a))))
}, numeric(1)), 1)
pedGroups %>%
  dplyr::select(group, mean_cl_pct, recommended_pct) %>%
  dplyr::rename(
    "Age group"                          = group,
    "Mean modelled CL/F (% of adult)"    = mean_cl_pct,
    "Recommended dose (% of adult)"      = recommended_pct
  ) %>%
  knitr::kable(caption = "Age-based dosing recommendation from the source Abstract and Discussion.")
```

# Validation 5: Fig. 5, 14-day profiles across the dose range

Fig. 5 shows typical-individual profiles for 1.5 to 75 mg once daily for 14
days, followed by washout. The published panel shows the lower doses reaching
steady state more slowly and displaying a shallower terminal phase -- both
consequences of the exposure-dependent volume.

```{r fig5}
fig5Doses <- c(1.5, 4, 7, 10, 20, 50, 75)
fig5 <- dplyr::bind_rows(lapply(fig5Doses, function(d) {
  ev <- baloEvents(
    doseTimes = seq(0, 13 * 24, by = 24),
    doseAmts  = rep(d, 14),
    doseCmt   = "depot",
    obsTimes  = seq(0, 500, by = 1),
    FED       = 1
  )
  out <- rxode2::rxSolve(modTypical, ev, returnType = "data.frame")
  out$dose_mg <- d
  out
}))

ggplot2::ggplot(fig5, ggplot2::aes(time, Cc, colour = factor(dose_mg))) +
  ggplot2::geom_line(linewidth = 0.5) +
  ggplot2::scale_y_log10(limits = c(0.1, 1000)) +
  ggplot2::labs(
    x = "Time after first dose (h)",
    y = "Typical balovaptan concentration (ng/mL)",
    colour = "Dose (mg)",
    title = "Replicates Figure 5 of Schaedeli Stark et al. (2024)"
  ) +
  ggplot2::theme_bw()
```

```{r fig5-assert}
# Terminal half-life is exposure-dependent: the volume is larger at low
# concentration, so the low-dose washout is shallower. Measured over 400-500 h.
washout <- fig5 %>%
  dplyr::filter(time >= 400, time <= 500, dose_mg %in% c(1.5, 75)) %>%
  dplyr::group_by(dose_mg) %>%
  dplyr::summarise(
    thalf = -log(2) / stats::coef(stats::lm(log(Cc) ~ time))[["time"]],
    .groups = "drop"
  )
stopifnot(
  washout$thalf[washout$dose_mg == 1.5] > washout$thalf[washout$dose_mg == 75]
)
washout %>%
  dplyr::mutate(thalf = round(thalf, 1)) %>%
  dplyr::rename("Dose (mg)" = dose_mg, "Terminal t1/2 over 400-500 h (h)" = thalf) %>%
  knitr::kable(caption = "Exposure-dependent terminal half-life, as described on p. 239 of the source.")
```

# Validation 6: virtual cohort and receptor occupancy over 72 hours

Fig. 8 and Table 3 report the 5th to 95th percentile band of brain receptor
occupancy across simulated participants under the 50 / 25 / 15 mg IV schedule.
Here 200 participants are simulated (the source used 2000).

```{r cohort}
rxode2::rxSetSeed(42)
cohortEvents <- baloEvents(
  doseTimes = c(0, 24, 48),
  doseAmts  = c(50, 25, 15),
  doseCmt   = "central",
  dur       = 1,
  obsTimes  = seq(0, 120, by = 0.5)
)
cohort <- rxode2::rxSolve(mod, cohortEvents, nSub = 200, returnType = "data.frame")

roBand <- cohort %>%
  dplyr::group_by(time) %>%
  dplyr::summarise(
    p05 = stats::quantile(RO, 0.05),
    p50 = stats::median(RO),
    p95 = stats::quantile(RO, 0.95),
    .groups = "drop"
  )

ggplot2::ggplot(roBand, ggplot2::aes(time, p50)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p05, ymax = p95), alpha = 0.25) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = c(80, 95), linetype = "dashed") +
  ggplot2::scale_x_continuous(breaks = seq(0, 120, by = 24)) +
  ggplot2::labs(
    x = "Time (h)", y = "Brain V1a receptor occupancy (%)",
    title = "Replicates Figure 8 of Schaedeli Stark et al. (2024)"
  ) +
  ggplot2::theme_bw()
```

The source's design target was occupancy of at least 80% in at least 95% of
participants over 72 hours, with the caveat that "the RO falls below the limit
just before reaching the 72-h mark, which was considered acceptable". The
assertions below are placed on the cohort *median* and on a robust quantile
rather than on any cohort extreme, because the extreme of a random cohort is not
reproducible across rxode2 versions.

```{r cohort-assert}
at <- function(tt) roBand[which.min(abs(roBand$time - tt)), ]
stopifnot(
  # Median occupancy tracks the published medians (99, 96, 99, 95, 98, 93).
  abs(at(1)$p50 - 99) < 1.5,
  abs(at(24)$p50 - 96) < 2,
  abs(at(48)$p50 - 95) < 2,
  abs(at(72)$p50 - 93) < 2,
  # The published 5th percentile dips just below 80% at 72 h.
  at(72)$p05 > 70, at(72)$p05 < 85,
  # Occupancy is above target for the great majority of the 72 h window.
  mean(roBand$p50[roBand$time <= 72] >= 80) > 0.95
)
```

## PKNCA on the virtual cohort

```{r pknca-cohort}
cohortNca <- cohort %>%
  dplyr::filter(!is.na(Cc)) %>%
  dplyr::mutate(dose_lbl = "50/25/15 mg IV", dose_mg = 50) %>%
  dplyr::select(id = sim.id, dose_lbl, dose_mg, time, Cc)

cohortDose <- cohortNca %>%
  dplyr::group_by(id, dose_lbl, dose_mg) %>%
  dplyr::summarise(time = 0, .groups = "drop")

cohortRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(cohortNca, Cc ~ time | id / dose_lbl),
  PKNCA::PKNCAdose(cohortDose, dose_mg ~ time | id + dose_lbl),
  intervals = data.frame(
    start = 0, end = 24,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
  )
))

as.data.frame(cohortRes) %>%
  dplyr::group_by(PPTESTCD) %>%
  dplyr::summarise(
    Median = stats::median(PPORRES, na.rm = TRUE),
    P05    = stats::quantile(PPORRES, 0.05, na.rm = TRUE),
    P95    = stats::quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD),
    dplyr::across(c(Median, P05, P95), ~ signif(.x, 3))
  ) %>%
  dplyr::rename(
    "NCA parameter"   = PPTESTCD,
    "Median"          = Median,
    "5th percentile"  = P05,
    "95th percentile" = P95
  ) %>%
  knitr::kable(caption = "Day 1 NCA (0-24 h) after the 50 mg 60-minute infusion, 200 simulated participants.")
```

The source publishes no NCA summary table, so this block is an internal
consistency check rather than a comparison: the day 1 Cmax median should sit
near the 352 ng/mL of Table 3 (Cmax occurs at the end of the 1 h infusion, which
is the Table 3 "1 h" time point), and Tmax should be 1 h.

```{r pknca-cohort-assert}
cohortSummary <- as.data.frame(cohortRes) %>%
  dplyr::group_by(PPTESTCD) %>%
  dplyr::summarise(med = stats::median(PPORRES, na.rm = TRUE), .groups = "drop")
getMed <- function(x) cohortSummary$med[cohortSummary$PPTESTCD == x]
stopifnot(
  abs(getMed("cmax") - 352) / 352 < 0.10,
  abs(getMed("tmax") - 1) < 1e-8
)
```

# Session information

```{r session}
sessionInfo()
```
