---
title: "Atogepant (Schlachter 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Atogepant (Schlachter 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Schlachter_2026_atogepant"))$reference`
* Description: `r rxode2::rxode(readModelDb("Schlachter_2026_atogepant"))$description`
* Article: <https://doi.org/10.1007/s40262-025-01566-5>
* Supplement: <https://doi.org/10.1007/s40262-025-01566-5> (Supplementary Information, open access)

Atogepant (Qulipta / Aquipta) is an oral calcitonin gene-related peptide (CGRP)
receptor antagonist approved for the preventive treatment of episodic and
chronic migraine. Schlachter 2026 reports the population pharmacokinetic model
that supported dose selection and the regulatory filings, developed in three
stages across the programme. This vignette packages the **Phase 3 Model**, the
final model of the paper; the Phase 1 and Phase 2 Models are earlier
development iterations of the same analysis and are superseded by it, so per
the library's replicate-the-author's-structure policy only the final model is
extracted.

# Population

The Phase 3 Model was fitted to 11,763 atogepant concentrations from 1356
participants: 351 healthy participants from 12 phase 1 studies, 463 patients
with episodic migraine from the phase 2b/3 study CGP-MD-01, and 542 patients
with episodic migraine from the pivotal phase 3 ADVANCE study (Table 1 and
supplementary Tables S1 and S2). Median age was 39 years (range 18 to 78),
median body weight 77.6 kg (40.7 to 196), and 74.7% were female. The population
was 76.0% Caucasian, 17.9% Black or African American and 3.5% Asian. Hepatic
function was normal in 98.2%, with 8 subjects in each of the mild, moderate and
severe strata contributed by the dedicated hepatic-impairment study CGP-PK-01.
Renal function was normal in 87.2%, mildly impaired in 11.9% and moderately
impaired in 1.0%; no subject had severe renal impairment. Concomitant
medications of interest were itraconazole (40 subjects), rifampicin (31) and
quinidine (25), each from a dedicated drug-drug-interaction study, plus
concomitant statins in 67 subjects.

The model was then externally validated **without re-estimation** against 350
patients with chronic migraine (PROGRESS) and 127 patients with episodic
migraine for whom two to four classes of conventional oral preventive
treatments had failed (ELEVATE). Those 477 subjects are not part of the 1356
above.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Schlachter_2026_atogepant"))$population
str(pop, max.level = 1)
```

# Model structure

Three disposition compartments with linear elimination from the central
compartment, and a **sequential zero-order then first-order absorption process
with a lag time**. The bioavailable dose is released into the depot at a
zero-order rate over a duration `Tk0`, and the depot empties into the central
compartment at first-order rate `ka`, with the whole input delayed by `ALAG`.

`ka` is not estimated independently. Section 3.1 states it "was linked to the
zero-order input parameters through `ka = Fk0 / [Tk0 * (1 - Fk0)]`", which
rearranges exactly to `Fk0 = Tk0 / (Tk0 + 1/ka)`: `Fk0` is the fraction of the
**total absorption time constant** contributed by the zero-order step, not (as
the surrounding prose loosely glosses it) a fraction of the dose. The
consequence matters for the covariate model: because dose and formulation are
applied to `Tk0`, they propagate into `ka`, which is why Table 2 labels those
two estimates as effects "on ka" while the display equations apply them to
`Tk0`.

```{r structure}
ui <- rxode2::rxode(readModelDb("Schlachter_2026_atogepant"))
cat("ODE states :", paste(ui$state, collapse = ", "), "\n")
cat("Covariates :", paste(sort(ui$allCovs), collapse = ", "), "\n")
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Schlachter_2026_atogepant.R` carries an in-file
comment naming its source location. They are collected here for review. "Eq."
refers to the unnumbered display equations printed in Section 3.1 of the paper.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F, patients) | 17.4 L/h | Table 2, "Apparent clearance patients" (RSE 2%) |
| `lvc` (V1/F) | 86.1 L | Table 2, "Apparent central volume of distribution" (RSE 2.3%) |
| `lq` (Q/F) | 1.43 L/h | Table 2, "Apparent first intercompartmental clearance" (RSE 10%) |
| `lvp` (V2/F) | 40.5 L | Table 2, "Apparent first peripheral volume of distribution" (RSE 7.3%) |
| `lq2` (Q2/F) | 1.68 L/h | Table 2, "Apparent second intercompartmental clearance" (RSE 7.4%) |
| `lvp2` (V3/F) | 13.0 L | Table 2, "Apparent second peripheral volume of distribution" (RSE 11.9%) |
| `ld1` (Tk0) | 0.908 h | Table 2, "Duration zero-order absorption" (RSE 4.7%) |
| `ltlag` (ALAG) | 0.276 h | Table 2, "Lag time" (RSE 0.5%) |
| `fk0` (Fk0) | 0.693 | Table 2, "Fraction zero-order absorption" (RSE 2.9%) |
| `lfdepot` (Frel at 60 mg) | 1 (fixed) | Sect. 3.1 Frel Eq.; Frel is purely relative, anchored at the 60 mg reference |
| `bpratio` | 0.573 | Table 2, "Blood-plasma ratio" (RSE 2%) |
| `e_dis_healthy_cl` | 0.3161 | **Derived** from the two Table 2 typical values: 22.9 / 17.4 - 1 |
| `e_hepimp_sev_cl` | -0.366 | Table 2, "Effect of severe hepatic impairment on CL/F"; Sect. 3.1 CL/F Eq. |
| `e_conmed_itraconazole_cl` | -0.662 | Table 2, "Itraconazole effect on CL/F"; Sect. 3.1 CL/F Eq. |
| `e_conmed_quinidine_cl` | -0.285 | Table 2, "Quinidine effect on CL/F"; Sect. 3.1 CL/F Eq. |
| `e_conmed_rifampicin_sd_cl` | -0.128 | Table 2, "Rifampin effect on CL/F after first dose" |
| `e_conmed_rifampicin_md_cl` | 0.818 | Table 2, "Rifampin effect on CL/F following multiple doses" |
| `e_wt_vc` | 0.411 | Table 2, "Exponential weight effect on V1/F"; Sect. 3.1 V1/F Eq. (76.8 kg reference) |
| `e_dose_atogepant_mg_fdepot` | 0.119 | Table 2, "Exponential dose effect on Frel"; Sect. 3.1 Frel Eq. (60 mg reference) |
| `e_conmed_itraconazole_fdepot` | 0.949 | Table 2, "Itraconazole effect on Frel" |
| `e_conmed_rifampicin_sd_fdepot` | 1.42 | Table 2, "Rifampin effect on Frel after first dose" |
| `e_conmed_rifampicin_md_fdepot` | -0.248 | Table 2, "Rifampin effect on Frel following multiple doses" |
| `e_fed_highfat_tlag` | 0.672 | Table 2, "Food effect on ALAG"; Sect. 3.1 ALAG Eq. |
| `e_dose_atogepant_mg_d1` | 0.199 | Table 2 (labelled "on ka"); Sect. 3.1 Tk0 Eq. applies it to Tk0 |
| `e_form_atogepant_earlytab_d1` | -0.353 | Table 2 (labelled "on ka"); Sect. 3.1 Tk0 Eq. applies it to Tk0 |
| `etalcl` | 0.0243 | Table 2, "omega2 CL/F"; sqrt = 15.6% CV, matching Sect. 3.1 |
| `etald1` | 0.163 | Table 2, "omega2 tk0"; sqrt = 40.4% CV |
| `etalfdepot` | 0.234 | Table 2, "omega2 Frel"; sqrt = 48.4% CV |
| `etalq` / `etalvp` block | 0.471 / 0.422 / 0.601 | Table 2, "omega2 Q/F", "Cov Q/F,V2/F", "omega2 V2/F"; sqrt of diagonals = 68.6% and 77.5% CV |
| `etaiov_fdepot_*` | 0.0373 | Table 2, "omega IOV Frel"; sqrt = 19.3% CV |
| `propSdPhase1` | 0.307 | Table 2, "sigma prop"; 30.7% CV |
| `propSdCgpPk02` | 0.228 | Table 2, "sigma prop (study CGP-PK-02)"; 22.8% CV |
| `propSdCgpMd01` | 0.585 | Table 2, "sigma prop (study CGP-MD-01)"; 58.5% CV |
| `propSdPhase3` | 0.491 | Table 2, "sigma prop (ADVANCE and PROGRESS studies)"; 49.1% CV |
| ODE system (3 compartments, linear elimination) | n/a | Sect. 3.2.1 and the Discussion |
| `ka = Fk0 / (Tk0 * (1 - Fk0))` | n/a | Sect. 3.1, verbatim |
| Concentration scaling (`* 1000`) | n/a | Units: mg dose / L volume gives mg/L; the paper reports ng/mL |

# Deterministic structural checks

These compare quantities the paper states in prose against the packaged model's
own arithmetic. They are **deterministic** -- no simulated cohort is involved --
so the tolerances are tight and are expected to stay tight.

```{r structural-checks}
th <- ui$theta

# Fk0 / Tk0 / ka relationship, Sect. 3.1.
tk0_ref <- exp(th[["ld1"]])
fk0_ref <- th[["fk0"]]
ka_derived <- fk0_ref / (tk0_ref * (1 - fk0_ref))

# Frel ratio between the 60 mg and 10 mg doses, Sect. 3.1 / Discussion.
frel_ratio_60_10 <- (60 / 60)^th[["e_dose_atogepant_mg_fdepot"]] /
  (10 / 60)^th[["e_dose_atogepant_mg_fdepot"]]

# Lag time after a high-fat meal, Sect. 3.1.
alag_fed <- exp(th[["ltlag"]]) * (1 + th[["e_fed_highfat_tlag"]])

# Tk0 shortening for the early phase 1 tablet, Sect. 3.1.
tk0_pct_shorter <- -100 * th[["e_form_atogepant_earlytab_d1"]]

# Healthy-participant clearance recovered from the patient typical value.
cl_healthy <- exp(th[["lcl"]]) * (1 + th[["e_dis_healthy_cl"]])

structural <- tibble::tribble(
  ~Quantity,                                  ~Published, ~Model,
  "Derived ka (1/h)",                          2.48,       ka_derived,
  "Frel ratio, 60 mg vs 10 mg",                1.24,       frel_ratio_60_10,
  "Lag time after a high-fat meal (h)",        0.46,       alag_fed,
  "Tk0 shortening, early phase 1 tablet (%)", 35.0,        tk0_pct_shorter,
  "CL/F in healthy participants (L/h)",       22.9,        cl_healthy,
  "Blood-to-plasma ratio",                     0.573,      th[["bpratio"]]
) |>
  mutate(`Diff (%)` = 100 * (Model - Published) / Published)

knitr::kable(structural, digits = 3,
             caption = "Model arithmetic vs values stated in Schlachter 2026 Section 3.1.")

# Each published figure is rounded to 2-3 significant figures, so 1% is ample
# headroom for rounding while still catching a mis-transcribed estimate (which
# would move these by tens of percent).
stopifnot(max(abs(structural$`Diff (%)`)) < 1)
```

## Interindividual variability on the reported CV scale

Table 2 reports variances; the Table 1 footnote states that IIV is derived as
`sqrt(omega^2) * 100`. Recomputing the percent CVs quoted in Section 3.1 from
the packaged variances confirms the scale (variance, not SD) was read
correctly.

```{r iiv-scale}
om <- ui$omega
iiv <- tibble::tribble(
  ~Parameter,        ~Variance,               ~`Published CV (%)`,
  "CL/F",             om["etalcl", "etalcl"],         15.6,
  "Tk0",              om["etald1", "etald1"],         40.4,
  "Frel",             om["etalfdepot", "etalfdepot"], 48.4,
  "Q/F",              om["etalq", "etalq"],           68.6,
  "V2/F",             om["etalvp", "etalvp"],         77.5,
  "Frel (IOV)",       om["etaiov_fdepot_1", "etaiov_fdepot_1"], 19.3
) |>
  mutate(`Model CV (%)` = 100 * sqrt(Variance),
         `Diff (%)` = 100 * (`Model CV (%)` - `Published CV (%)`) / `Published CV (%)`)

knitr::kable(iiv, digits = 3,
             caption = "IIV variances back-transformed to the percent CVs quoted in Section 3.1.")

stopifnot(max(abs(iiv$`Diff (%)`)) < 0.5)

# The Q/F - V2/F covariance implies a strong positive correlation.
cor_q_vp <- om["etalq", "etalvp"] / sqrt(om["etalq", "etalq"] * om["etalvp", "etalvp"])
cat(sprintf("Implied Q/F - V2/F correlation: %.3f\n", cor_q_vp))
stopifnot(cor_q_vp > 0.7, cor_q_vp < 0.9)
```

# Mass-balance check on the ODE system

For a linear model the total exposure after a single dose is exactly
`Dose * Frel / (CL/F)`, independent of the absorption and distribution
structure. Reproducing that identity from the solved ODE system confirms the
compartment wiring, the bioavailability target, and the mg-to-ng/mL scaling
simultaneously. Because both sides use the same drawn parameters, the only
difference is numerical-integration error and the bound is correspondingly
tight.

```{r mass-balance}
mod  <- readModelDb("Schlachter_2026_atogepant")
modT <- rxode2::zeroRe(mod)

# Reference covariate set: a typical migraine patient at the population median
# weight, Formulation 5, fasted, no interacting comedication, phase 3 study.
base_covs <- list(
  OCC = 1, DIS_HEALTHY = 0, HEPIMP_SEV = 0, CONMED_ITRACONAZOLE = 0,
  CONMED_QUINIDINE = 0, CONMED_RIFAMPICIN_SD = 0, CONMED_RIFAMPICIN_MD = 0,
  WT = 77.6, DOSE_ATOGEPANT_MG = 60, FORM_ATOGEPANT_EARLYTAB = 0,
  FED_HIGHFAT = 0, STUDY_CGP_PK_02 = 0, STUDY_CGP_MD_01 = 0,
  STUDY_ATOGEPANT_PHASE3 = 1
)

# `rate = -2` tells rxode2 to use the model's own dur(depot) for the
# zero-order input rather than treating the record as a bolus.
solve_typical <- function(dose, times, covs = list(), ii = NULL, until = NULL) {
  cv <- modifyList(base_covs, c(list(DOSE_ATOGEPANT_MG = dose), covs))
  ev <- if (is.null(ii)) {
    rxode2::et(amt = dose, cmt = "depot", rate = -2)
  } else {
    rxode2::et(amt = dose, cmt = "depot", rate = -2, ii = ii, until = until)
  }
  # Observation rows point at the ODE state `central`; rxode2 returns the
  # algebraic observables Cc and Cb as columns at those rows.
  ev <- rxode2::et(ev, times, cmt = "central")
  d <- as.data.frame(ev)
  for (nm in names(cv)) d[[nm]] <- cv[[nm]]
  rxode2::rxSolve(modT, d, returnType = "data.frame") |>
    dplyr::filter(!is.na(Cc))
}

trap_auc <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}

# 21 days is many multiples of the ~11 h terminal half-life, so the truncated
# AUC is within rounding of AUC(0, Inf).
sd_grid <- sort(unique(c(seq(0, 24, by = 0.02), seq(24, 504, by = 0.5))))
mb <- lapply(c(10, 30, 60), function(d) {
  s <- solve_typical(d, sd_grid)
  frel <- (d / 60)^ui$theta[["e_dose_atogepant_mg_fdepot"]]
  tibble::tibble(
    Dose = paste(d, "mg"),
    `AUC from ODE (ng*h/mL)` = trap_auc(s$time, s$Cc),
    `Dose * Frel / (CL/F)`   = d * frel / exp(ui$theta[["lcl"]]) * 1000
  )
}) |> bind_rows() |>
  mutate(`Diff (%)` = 100 * (`AUC from ODE (ng*h/mL)` - `Dose * Frel / (CL/F)`) /
           `Dose * Frel / (CL/F)`)

knitr::kable(mb, digits = 2,
             caption = "Single-dose mass balance: solved AUC vs the closed-form identity.")

stopifnot(max(abs(mb$`Diff (%)`)) < 0.5)
```

# Covariate-effect replication

Every intrinsic and extrinsic factor the paper retained is exercised here by
solving the packaged model with and without the covariate set and comparing the
realised exposure ratio against the ratio implied by the published
coefficients. This is the check that the covariate wiring -- which parameter
each effect lands on, and in which direction -- is correct. It is deterministic,
so the tolerance is tight.

```{r covariate-effects}
# AUC ratio implied by the published coefficients: a factor on Frel raises
# exposure, a factor on CL/F lowers it.
scenarios <- tibble::tribble(
  ~Scenario,                       ~covs,                                  ~cl_fac, ~frel_fac,
  "Healthy vs migraine patient",   list(DIS_HEALTHY = 1),                   1.3161,   1,
  "Severe hepatic impairment",     list(HEPIMP_SEV = 1),                    1 - 0.366, 1,
  "Itraconazole",                  list(CONMED_ITRACONAZOLE = 1),           1 - 0.662, 1 + 0.949,
  "Quinidine",                     list(CONMED_QUINIDINE = 1),              1 - 0.285, 1,
  "Rifampicin, single dose",       list(CONMED_RIFAMPICIN_SD = 1),          1 - 0.128, 1 + 1.42,
  "Rifampicin, multiple doses",    list(CONMED_RIFAMPICIN_MD = 1),          1 + 0.818, 1 - 0.248
)

ref <- solve_typical(60, sd_grid)
auc_ref <- trap_auc(ref$time, ref$Cc)

cov_tab <- lapply(seq_len(nrow(scenarios)), function(i) {
  s <- solve_typical(60, sd_grid, covs = scenarios$covs[[i]])
  tibble::tibble(
    Scenario = scenarios$Scenario[i],
    `Expected AUC ratio` = scenarios$frel_fac[i] / scenarios$cl_fac[i],
    `Simulated AUC ratio` = trap_auc(s$time, s$Cc) / auc_ref
  )
}) |> bind_rows() |>
  mutate(`Diff (%)` = 100 * (`Simulated AUC ratio` - `Expected AUC ratio`) /
           `Expected AUC ratio`)

knitr::kable(cov_tab, digits = 3,
             caption = paste(
               "Covariate effects on single-dose AUC, relative to a typical",
               "migraine patient with no interacting comedication."
             ))

stopifnot(max(abs(cov_tab$`Diff (%)`)) < 0.5)
```

The paper describes these same effects in words in Section 3.1 and the
Discussion: itraconazole and quinidine lower CL/F by 66.2% and 28.5%;
itraconazole raises Frel 1.95-fold; a single rifampicin dose raises Frel
2.4-fold and lowers CL/F by 12.8%, while multiple rifampicin doses raise CL/F
1.82-fold and lower Frel by 24.8%; severe hepatic impairment lowers CL/F by
36.6%. The table above confirms each lands on the right parameter with the
right sign, and shows the net exposure consequence -- most strikingly the
5.8-fold AUC increase with itraconazole, which combines the clearance and
bioavailability effects in the same direction.

Body weight acts on V1/F alone, so it shifts Cmax without changing AUC:

```{r weight-effect}
wt_tab <- lapply(c(50, 77.6, 120), function(w) {
  s <- solve_typical(60, sd_grid, covs = list(WT = w))
  tibble::tibble(`Body weight (kg)` = w,
                 `Cmax (ng/mL)` = max(s$Cc),
                 `AUC (ng*h/mL)` = trap_auc(s$time, s$Cc))
}) |> bind_rows()

knitr::kable(wt_tab, digits = 1,
             caption = "Body weight acts on V1/F only: Cmax moves, AUC does not.")

# AUC must be analytically identical across weights (weight touches no
# clearance term); the residual spread is trapezoidal discretisation error,
# which grows slightly as the peak sharpens at low weight. Realised 5.0e-6
# (3448.296 / 3448.304 / 3448.313, i.e. agreement to 7 significant figures);
# 1e-4 leaves two orders of magnitude of headroom while still catching a
# weight term leaking into a clearance parameter, which would be percent-level.
# Both checks are deterministic -- no cohort is drawn -- so the bounds are tight.
stopifnot(diff(range(wt_tab$`AUC (ng*h/mL)`)) / mean(wt_tab$`AUC (ng*h/mL)`) < 1e-4)
stopifnot(wt_tab$`Cmax (ng/mL)`[1] > wt_tab$`Cmax (ng/mL)`[3])
```

# Virtual cohort

Original observed data are not public. The cohort below approximates the three
once-daily arms of the phase 3 ADVANCE study, whose model-derived steady-state
exposures the paper tabulates in Table 3. All ADVANCE participants were
patients with episodic migraine dosed with the Formulation 5 tablet and had no
interacting comedication, so only body weight varies across the cohort.

```{r cohort}
# `set.seed()` seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD -- so this cohort is
# reproducible on this machine and different on a machine with a different
# thread count. Every assertion downstream is written to hold for ANY cohort
# the model can produce (see pattern 12 of known-vignette-failure-patterns.md).
set.seed(20260911)

n_per_arm <- 200
tau       <- 24
n_days    <- 14                # >> 5 terminal half-lives (~11 h)
ss_start  <- (n_days - 1) * tau
obs_grid  <- seq(ss_start, ss_start + tau, by = 0.1)

# ADVANCE body weight: Table 1 gives median 77.6 kg, range 40.7-196 for the
# whole Phase 3 Model population. A log-normal centred on the median with a
# 0.25 log-SD reproduces that spread; weight is truncated to the observed range.
make_arm <- function(dose, label, id_offset) {
  subj <- tibble::tibble(
    id    = id_offset + seq_len(n_per_arm),
    arm   = label,
    WT    = pmin(pmax(stats::rlnorm(n_per_arm, log(77.6), 0.25), 40.7), 196),
    DOSE_ATOGEPANT_MG = dose
  )
  doses <- subj |>
    tidyr::crossing(time = seq(0, (n_days - 1) * tau, by = tau)) |>
    mutate(amt = dose, evid = 1L, cmt = "depot", rate = -2,
           # One occasion per dosing week, cycling through the four IOV slots.
           OCC = pmin(4L, as.integer(time %/% (7 * tau)) + 1L))
  obs <- subj |>
    tidyr::crossing(time = obs_grid) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = NA_real_,
           OCC = pmin(4L, as.integer(time %/% (7 * tau)) + 1L))
  bind_rows(doses, obs) |>
    mutate(
      DIS_HEALTHY = 0, HEPIMP_SEV = 0, CONMED_ITRACONAZOLE = 0,
      CONMED_QUINIDINE = 0, CONMED_RIFAMPICIN_SD = 0, CONMED_RIFAMPICIN_MD = 0,
      FORM_ATOGEPANT_EARLYTAB = 0, FED_HIGHFAT = 0,
      STUDY_CGP_PK_02 = 0, STUDY_CGP_MD_01 = 0, STUDY_ATOGEPANT_PHASE3 = 1
    ) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(10, "10 mg QD", 0L),
  make_arm(30, "30 mg QD", 1000L),
  make_arm(60, "60 mg QD", 2000L)
)

# Disjoint IDs across arms are mandatory: rxSolve treats id as the subject key
# and silently merges duplicates into a single over-dosed subject.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(length(unique(events$id)) == 3 * n_per_arm)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("arm", "WT")) |>
  as.data.frame()
stopifnot(nrow(sim) > 0, any(!is.na(sim$Cc)))
```

```{r figure-3}
# Companion to Figure 3 of Schlachter 2026: the simulated steady-state
# concentration-time envelope over one dosing interval, by dose arm.
sim |>
  filter(!is.na(Cc)) |>
  mutate(tad = time - ss_start) |>
  group_by(arm, tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 8.4, linetype = "dashed", colour = "firebrick") +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Atogepant plasma concentration (ng/mL)",
    title = "Simulated steady-state profiles by ADVANCE dose arm",
    caption = paste(
      "Median with 5th-95th percentile ribbon, 200 subjects per arm.",
      "Dashed line: the EC90 of 8.4 ng/mL (Sect. 2.3.4).",
      "Companion to Figure 3 of Schlachter 2026."
    )
  )
```

# PKNCA validation

Steady-state NCA over the final dosing interval, stratified by dose arm.

```{r pknca}
# Only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the interval
# anchor row and trigger PKNCA's "AUC range starting before the first
# measurement" warning on every subject.
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "ng/mL", timeu = "h")

dose_df <- events |>
  filter(evid == 1L, time == ss_start) |>
  select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start   = ss_start,
  end     = ss_start + tau,
  cmax    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

### Why Ctrough is computed outside PKNCA here

`Ctrough` is deliberately **not** requested from PKNCA above, because neither
PKNCA parameter means what Table 3's "Ctrough" means for this dataset:

* `ctrough` is the concentration at the end of the interval **anchored on a dose
  record**. The simulated cohort's last dose is at `ss_start`, so nothing
  anchors the interval end at `ss_start + tau` and PKNCA returns `NA` for every
  subject.
* `cmin` is the minimum **within** the interval, which for this model is not the
  trough. The 0.276 h absorption lag means concentrations keep falling for a
  little after each dose, so the within-interval minimum occurs just after
  `ss_start`, not at `ss_start + tau`.

Since the last dose is exactly one `tau` before the end of the observation grid,
the trough is simply the simulated concentration at `ss_start + tau`, taken
directly:

```{r ctrough}
trough <- sim |>
  filter(!is.na(Cc), abs(time - (ss_start + tau)) < 1e-8) |>
  group_by(arm) |>
  summarise(sim = median(Cc), .groups = "drop") |>
  mutate(PPTESTCD = "ctrough")

stopifnot(nrow(trough) == 3, all(is.finite(trough$sim)))
```

## Comparison against published Table 3

Table 3 of Schlachter 2026 reports individual model-derived steady-state
exposures for each treatment arm as mean (SD) and median (5th to 95th
percentile). The medians for the three ADVANCE once-daily arms are transcribed
below.

```{r nca-compare}
published <- tibble::tribble(
  ~arm,        ~auclast, ~cmax, ~cav,  ~ctrough,
  "10 mg QD",     464,    73.3,  19.3,   2.62,
  "30 mg QD",    1560,   244,    65.1,   8.85,
  "60 mg QD",    3220,   504,   134,    18.4
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published |> select(-ctrough),
  by        = "arm",
  units     = c(auclast = "ng*h/mL", cmax = "ng/mL", cav = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated steady-state NCA (median across 200 subjects per arm) vs the",
    "published Table 3 medians for the ADVANCE once-daily arms.",
    "* differs from the reference by more than 20%."
  )
)
```

```{r nca-gate}
# Recompute the simulated-vs-published percent differences numerically: the
# `% diff` column of ncaComparisonTable() is formatted TEXT, not a number.
# Ctrough is folded back in from the direct calculation above.
sim_all <- bind_rows(
  as.data.frame(nca_res) |>
    filter(PPTESTCD %in% c("auclast", "cmax", "cav")) |>
    group_by(arm, PPTESTCD) |>
    summarise(sim = median(PPORRES), .groups = "drop"),
  trough |> select(arm, PPTESTCD, sim)
)

gate <- published |>
  tidyr::pivot_longer(-arm, names_to = "PPTESTCD", values_to = "ref") |>
  inner_join(sim_all, by = c("arm", "PPTESTCD")) |>
  mutate(pct = 100 * (sim - ref) / ref)

# A gate with missing or zero rows cannot go red.
stopifnot(nrow(gate) == 12, !anyNA(gate$pct))

knitr::kable(gate |> arrange(PPTESTCD, arm), digits = 2,
             caption = "Percent difference, simulated median vs published median.")

# Structural gate on the CENTRE and on a robust quantile -- never on the
# extreme, which is not reproducible across rxode2 versions or solver thread
# counts (the cohort differs). A mis-transcribed clearance, dose, exponent or
# unit moves the WHOLE distribution by tens of percent and blows the median
# bound immediately.
#
# Realised on this cohort: median +3.3%, q90 |7.0|%, max |17.9|% (the 60 mg
# trough). The per-arm pattern is systematic, not noise: 10 and 30 mg sit
# within a few percent while 60 mg runs consistently high, because the
# published 60 mg medians imply a slightly weaker dose effect on Frel than the
# printed 0.119 exponent reproduces. See the Errata below.
stopifnot(abs(median(gate$pct)) < 15)
stopifnot(quantile(abs(gate$pct), 0.9) < 25)
```

## Dose proportionality

Atogepant exposure is close to dose-proportional, with a modest
greater-than-proportional trend contributed by the `(dose/60)^0.119`
bioavailability term. Section 3.1 quantifies it as a 1.24-fold higher Frel at
60 mg than at 10 mg.

```{r dose-proportionality}
dp <- gate |>
  filter(PPTESTCD == "auclast") |>
  mutate(dose = as.numeric(sub(" mg QD", "", arm))) |>
  arrange(dose) |>
  mutate(`Dose-normalised AUC24 (simulated)` = sim / dose,
         `Dose-normalised AUC24 (published)` = ref / dose)

knitr::kable(
  dp |> select(arm, `Dose-normalised AUC24 (simulated)`,
               `Dose-normalised AUC24 (published)`),
  digits = 2,
  caption = "Dose-normalised steady-state AUC24 rises modestly with dose."
)

# Trend, not step-by-step monotonicity on a noisy cohort statistic.
stopifnot(
  dp$`Dose-normalised AUC24 (simulated)`[3] >
    dp$`Dose-normalised AUC24 (simulated)`[1]
)
```

# Assumptions and deviations

## Errata and internal inconsistencies in the source

* **The Frel display equation swaps the two rifampicin labels.** The Section
  3.1 `Frel` display equation prints `x (1 - 0.248) with rifampin (after first
  dose)` and `x (1 + 1.42) with rifampin (after multiple doses)`. Table 2, the
  Section 3.1 prose ("Frel and CL/F were 2.4-fold higher and 12.8% lower
  ... following coadministration of single-dose atogepant and single-dose
  rifampin ... Following the coadministration of multiple rifampin doses ...
  Frel decreased by 24.8%") and the Abstract all assign `+1.42` to the single
  dose and `-0.248` to multiple doses. The model follows the three-to-one
  majority. The mechanism corroborates it: the acute OATP1B1-inhibition phase
  raises exposure and the induction phase lowers it, which is also how the
  paper's own Discussion reads the result.

* **Table 2 labels two absorption covariates as effects "on ka" while the
  display equation applies them to Tk0.** The dose effect (0.199) and the
  formulation effect (-0.353) appear in Table 2 under "Exponential dose effect
  on ka" and "Formulation 2 tablet effect on ka/formulation 4 tablet effect on
  ka", but the Section 3.1 display equation applies both to `Tk0`. These are not
  in conflict: the model derives `ka = Fk0 / (Tk0 * (1 - Fk0))`, so an effect
  applied to `Tk0` propagates into `ka`. The model applies them to `Tk0` as the
  equation does. The Table 2 row label for the formulation effect is
  additionally a carry-over from the Phase 1 Model column, where two separate
  estimates (-0.44 and -0.42) sat directly on `ka`.

* **The source does not agree with itself about which formulations carry the
  Tk0 effect.** The display equation says "phase 1 / formulations 3 and 4", the
  prose says "the formulation 1 tablet used in early phase 1 studies", and the
  Table 2 alias says "Formulation 2 tablet / formulation 4 tablet". All three
  agree the effect belongs to the early phase 1 tablets and that the reference
  is the Formulation 5 tablet, so a single aggregate indicator
  (`FORM_ATOGEPANT_EARLYTAB`) is used. Formulations 1 to 4 together account for
  only 82 of the 1356 subjects and none of the phase 2b/3, phase 3 or
  validation data.

* **"Approximately 69% of the orally bioavailable dose was absorbed through a
  zero-order process" is a loose gloss of Fk0.** The printed relation
  `ka = Fk0 / [Tk0 * (1 - Fk0)]` rearranges to `Fk0 = Tk0 / (Tk0 + 1/ka)`, so
  `Fk0` is the fraction of the total absorption *time constant* that is
  zero-order, not a fraction of the dose. Reading it as a dose fraction would
  imply a parallel zero-order plus first-order input, which contradicts both
  the printed relation and the paper's description of the process as
  "sequential". The arithmetic confirms the time-constant reading:
  `0.908 / (0.908 + 1/2.486) = 0.693` exactly reproduces the tabulated Fk0, and
  the relation returns the paper's derived `ka` of 2.48/h.

* **Table 3, ADVANCE 30 mg QD, Cmax row** prints a median of 244 with a "5th to
  95th percentile" range of "431-1420" whose lower bound exceeds the median.
  Only the medians are used in the comparison above.

## Implementation choices

* **`e_dis_healthy_cl = 0.3161` is derived, not printed.** Table 2 reports the
  two apparent-clearance typical values separately (17.4 L/h in patients, 22.9
  L/h in healthy participants) rather than a ratio. The patient value is the
  model's typical value because Section 2.3.1 states the typical value is
  parameterised for the level constituting the largest proportion of the
  population, and patients supply 1005 of 1356 subjects. The healthy effect is
  therefore `22.9 / 17.4 - 1 = 0.3161`, which reproduces both printed values
  exactly. Encoding the paper's alternative phrasing instead (patients are
  "23.7% lower" than healthy) would have anchored on the healthy value and
  returned 17.47 L/h for patients rather than the printed 17.4.

* **Four IOV occasions.** The source reports the inter-occasion variance on
  `Frel` (0.0373, i.e. 19.3% CV) but never states how many occasions its
  analysis dataset defined, nor how an occasion was delimited. Four occasions
  are encoded, sharing the single reported variance via the registered
  `fixed()` idiom for NONMEM's `$OMEGA BLOCK(1) SAME`. The cohort above maps one
  occasion to each dosing week. Pass `OCC = 1` for single-occasion data.

* **The blood-to-plasma ratio is a derived output, not a second endpoint.**
  Studies CGP-PK-02 and CGP-MD-01 contributed dried-blood-sample concentrations,
  which the model relates to plasma through the estimated ratio of 0.573. The
  packaged model computes `Cb = Cc * bpratio` as an output column rather than
  declaring a second estimated endpoint, because the source stratifies residual
  error by **study**, not by matrix, so a separate endpoint would carry the same
  error magnitude and add no information -- while forcing every observation row
  in every downstream simulation to carry a `dvid`.

* **Residual error requires the study indicators.** Four proportional residual
  magnitudes are packaged, selected by `STUDY_CGP_PK_02`, `STUDY_CGP_MD_01` and
  `STUDY_ATOGEPANT_PHASE3`. With all three set to 0 a record takes the
  reference magnitude of 0.307 ("all phase 1 studies except CGP-PK-02"). The
  cohort above sets `STUDY_ATOGEPANT_PHASE3 = 1`, giving the 49.1% CV that
  applies to the ADVANCE data the comparison targets.

* **Dosing requires `rate = -2`.** The zero-order input duration is modelled
  (`dur(depot) <- d1`), so dose records must carry `rate = -2` for rxode2 to
  apply it; a plain bolus silently bypasses the zero-order absorption step.

* **`DOSE_ATOGEPANT_MG` must be supplied as its own column.** A covariate
  column literally named `DOSE` (any casing) is consumed by rxode2's
  `etTrans()` and never reaches `model()`, so the dose amount is carried both as
  the event-table `amt` and as this separate covariate column.

* **Covariates screened but not retained** (age, sex, race, renal function,
  concomitant statins, BCRP inhibitors and BCRP substrates) are recorded in the
  model file's `covariatesDataExcluded` metadata rather than `covariateData`,
  preserving the provenance of the Figure 4 covariate screen. Every one changed
  model-predicted steady-state exposure by less than 20% and was judged not
  clinically significant.

## Agreement with the published Table 3, by arm

The steady-state comparison above is close overall (median +3.3% across the 12
simulated-vs-published medians) but the residual is **systematic by dose rather
than random**: the 10 mg and 30 mg arms agree within a few percent on every
metric, while the 60 mg arm runs consistently high (+6% on AUC24 and Cav, +7% on
Cmax, +18% on Ctrough).

The arithmetic locates this in the dose effect on relative bioavailability, not
in the disposition parameters. At steady state a linear model gives
`AUC24 = Dose * Frel / (CL/F)` exactly, so the published medians imply
`Frel(60)/Frel(10) = (3220/464) / 6 = 1.157`, whereas the printed exponent of
0.119 gives `(60/10)^0.119 = 1.236`. The packaged model reproduces the printed
exponent; the published Table 3 medians imply a slightly flatter one (about
0.081). Both the exponent and the "approximately 1.24-fold higher at the 60 mg
dose compared with the 10 mg dose" statement in Section 3.1 are reproduced
exactly by the model (see the deterministic checks above), so this is a
disagreement *within* the source between its parameter table and its derived
exposure table, not a transcription error in the extraction. The most likely
mechanism is that Table 3 summarises individual Bayesian post hoc estimates,
which carry 44% shrinkage on CL/F and 43.9% on Frel and are therefore pulled
toward the population mean by an amount that differs across arms.

The deviation is recorded here rather than tuned away, and the gate is set on
the centre and a robust quantile so that it still goes red for any
transcription-scale error while tolerating this known offset.

## Known model limitation carried from the source

The paper reports that the Phase 3 Model **under-predicts Cmax** in both
external-validation cohorts (Section 3.3 for PROGRESS and Section 3.5 for
ELEVATE), and states this "could not be explained by demographic differences in
the study populations" and that neither phase 3 sampling scheme specifically
targeted Cmax. Relatedly, the model-derived Cmax values are lower than the
non-compartmental Cmax of roughly 740 ng/mL quoted in the Introduction for a
60 mg dose from rich phase 1 sampling. This is a property of the published
model, not of the extraction: the comparison against the paper's own
model-derived Table 3 values above agrees within the gate, and the deterministic
mass-balance and covariate checks are exact.

# Session information

```{r session}
sessionInfo()
```
