---
title: "Whole-body aging physiology, 30 to 100 years (Schlender 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Whole-body aging physiology, 30 to 100 years (Schlender 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Schlender_2016_aging_physiology"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1007/s40262-016-0422-3>\n")
cat("* Supplement (Springer ESM, open access): <https://doi.org/10.1007/s40262-016-0422-3>\n")
```

Schlender 2016 is not a drug model. It is the **system layer** that a whole-body
physiologically based pharmacokinetic (PBPK) model sits on, extended for the
first time across the full course of healthy aging to 100 years of age. The
paper's deliverable is an age- and sex-resolved description of human physiology:
body weight and height, thirteen organ and tissue masses, the cardiac index, ten
organ blood flow rates, and a **new** aging function for the glomerular
filtration rate (GFR).

That system layer is what this package extracts. Everything in it is published
as numbers -- Tables 2 and 3 and Equation 1 -- and is reproduced here exactly.

## What is deliberately *not* extracted, and why

Schlender 2016 verifies its physiology by building PBPK models of two
intravenously dosed probe drugs, **morphine** (high hepatic extraction) and
**furosemide** (mainly renal), in PK-Sim v5.5. Those two drug models are **not**
part of this extraction, because they are not reproducible from the paper:

* Table 1 is a **compound card** (molecular weight, log P, pKa, fraction
  unbound, specific clearances) -- the *inputs* to the platform, not the model.
* The tissue:plasma partition coefficients and the cellular permeabilities,
  which are the drug-specific parameters a whole-body PBPK actually runs on, are
  given only as a **named method** ("Willmann et al., defined as the PK-Sim
  standard"). A named method is an algorithm, not a set of values.
* No PK-Sim project file was deposited, and the supplement contains only the
  aging-physiology literature review and Supplement Table 1 (the list of
  studies) -- it holds no partition coefficients, no permeabilities and no ODEs.

Counting what would be needed against what is printed, the drug layer scores
roughly **5 published parameters out of 35+ required** per compound. The
physiology layer, by contrast, is complete and is verified below.

Users who want a morphine or furosemide PK model to pair with this physiology
can chain an independent compartmental model already in the registry, for
example `Elkomy_2015_morphine`, `Knibbe_2009_morphine` or `Hornik_2025_furosemide`,
or the PK-Sim-independent `Gu_2025_furosemide_pbpk`.

# Population

```{r population, echo = FALSE, results = "asis"}
pop <- ui$population
cat("* Species: ", pop$species, "\n", sep = "")
cat("* Age range: ", pop$age_range, "\n", sep = "")
cat("* Race / ethnicity: ", pop$race_ethnicity, "\n", sep = "")
cat("* Regions: ", pop$regions, "\n", sep = "")
```

The physiology is assembled from a literature analysis, not from a single
cohort: 19 studies carried over from the Thompson et al. database plus 97
studies added by this paper's own two PubMed / Google Scholar / MedPilot
searches (Methods). Inclusion required clear assignment of sex and race, a
comprehensible statement of the method of analysis, and exclusion of medication
or disease effects on physiology; longitudinal surveillance was preferred.
Organ volumes came mostly from autopsy series, with in vivo measurements used
only for muscle, fat and blood flow.

The GFR aging function of Equation 1 was fitted to body-surface-area-adjusted
data from **1,213 females and 1,081 males**, mostly prospective renal transplant
donors assessed with exogenous markers (Results, Glomerular Filtration Rate).

For the physiological-consistency checks the paper created a virtual 30-year-old
in PK-Sim v5.5 and generated a virtual population of N = 5,000 for **each** sex
over 30 to 100 years.

The same information is available programmatically via
`readModelDb("Schlender_2016_aging_physiology")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/endogenous/Schlender_2016_aging_physiology.R`. The table
below collects the whole model in one place.

| Model component | Source location |
|---|---|
| `bw`, `ht` | Table 2, rows "Body weight, kg" and "Height, cm" (female + male) |
| `mass_blood`, `mass_bone`, `mass_brain`, `mass_fat`, `mass_gonads`, `mass_heart`, `mass_kidney`, `mass_liver`, `mass_lung`, `mass_muscle`, `mass_pancreas`, `mass_skin`, `mass_spleen` | Table 2, "Organ masses, g" (female + male) |
| `mass_gi` | Table 2 footnote a: held constant at 1274.5 g (female) / 1304.8 g (male) |
| `cardiac_index` | Table 3, row "Cardiac index, L/m2" (female + male) |
| `flow_adipose`, `flow_cerebral`, `flow_gonads`, `flow_myocardial`, `flow_renal`, `flow_splanchnic`, `flow_hepatic`, `flow_muscle`, `flow_skeleton`, `flow_skin` | Table 3, "Organ blood flow, mL/min" (female + male) |
| Linear interpolation between decade knots | Methods, "Workflow for Elderly PBPK Model Development": data pooled into 10-year age bins "and interpolated linearly" |
| `gfr_fpma` = 26.6 mL/min/100 g | Text below Equation 1 (Rhodin maturation function at full maturation, 40th gestational week) |
| `gfr_vmax` = 0.9 | Text below Equation 1, "maximal decreasing rate factor" |
| `gfr_ta50_female` = 59 y, `gfr_ta50_male` = 54 y | Text below Equation 1, "aging half-time" |
| `gfr_hill` = 1.5 | Text below Equation 1, "The Hill coefficient was parameterized at 1.5" |
| `gfr_specific`, `gfr` | Equation 1, page 1581 |

**Equation 1 is an embedded image in the published PDF** and is dropped by every
text converter (it appears as `formula-not-decoded` in the trimmed markdown and
is absent from `pdftotext`). It was recovered from the publisher's equation
graphic (`40262_2016_422_Article_Equ1.gif`, retrieved from the EuropePMC
`supplementaryFiles` endpoint for PMC5107207) and confirmed by rendering page 8
of the PDF at 400 dpi. As printed:

$$
\mathrm{Specific\ GFR}_{30-100y} =
F_{\mathrm{PMA}} \times
\left( 1 - \frac{V_{\max} \times (\mathrm{age} - 30\,\mathrm{years})^{\mathrm{Hill}}}
{\mathrm{TA}_{50}^{\mathrm{Hill}} + (\mathrm{age} - 30\,\mathrm{years})^{\mathrm{Hill}}} \right)
$$

## Units

The model has no ODEs, so the dimensional check is on the algebraic outputs.

| Symbol | Units | Note |
|---|---|---|
| `time` / `age` | year | the rxode2 time variable **is** chronological age |
| `bw` | kg | |
| `ht` | cm | |
| `mass_*` | g | including `mass_gi`, which is age-invariant |
| `cardiac_index` | L/min/m^2 | Table 3 prints the header as "L/m2"; the cardiac index is a flow per unit BSA, so the per-minute is implicit |
| `bsa` | m^2 | Du Bois, from `ht` (cm) and `bw` (kg) |
| `cardiac_output` | mL/min | `cardiac_index` (L/min/m^2) x `bsa` (m^2) x 1000 (mL/L) |
| `flow_*` | mL/min | |
| `gfr_fpma`, `gfr_specific` | mL/min per 100 g kidney | |
| `gfr_ta50_*`, `gfr_age` | year | |
| `gfr_vmax`, `gfr_hill` | unitless | see Errata on the `gfr_vmax` unit tag |
| `gfr` | mL/min | `gfr_specific` (mL/min/100 g) x `mass_kidney` (g) / 100 |

# Parameter table

```{r param-table, echo = FALSE}
ui$iniDf |>
  dplyr::filter(!is.na(.data$ntheta)) |>
  dplyr::transmute(
    Parameter = .data$name,
    Value = .data$est,
    Fixed = .data$fix,
    Label = .data$label
  ) |>
  dplyr::rename("Fixed in source" = "Fixed") |>
  knitr::kable()
```

All five are published point estimates of a fitted function with no reported
standard error, so all are `fixed()`.

# Simulation

The model is deterministic: no between-subject variability and no residual
error are encoded (see Errata). One "subject" per sex is solved across the full
published age range.

```{r simulate}
mod <- readModelDb("Schlender_2016_aging_physiology")

solve_sex <- function(sexf, ages) {
  ev <- as.data.frame(rxode2::et(ages))
  ev$SEXF <- sexf
  out <- rxode2::rxSolve(mod, ev, returnType = "data.frame")
  out$SEXF <- sexf
  out$sex <- ifelse(sexf == 1, "Female", "Male")
  out
}

knots <- seq(30, 100, by = 10)
fine <- seq(30, 100, by = 0.5)

sim_knot <- dplyr::bind_rows(solve_sex(1, knots), solve_sex(0, knots))
sim_fine <- dplyr::bind_rows(solve_sex(1, fine), solve_sex(0, fine))

nrow(sim_knot)
nrow(sim_fine)
```

# Validation

## 1. Exact reproduction of Tables 2 and 3

The strongest available gate: the model must return **every one of the 416
published cells** of Tables 2 and 3 at the decade knots. The reference values
below are transcribed independently from the article PDF.

```{r knot-reference}
# Schlender 2016 Table 2 (anthropometry + organ masses) and Table 3
# (cardiac index + organ blood flows). Order is age 30, 40, ..., 100.
ref <- tibble::tribble(
  ~var, ~sex, ~v30, ~v40, ~v50, ~v60, ~v70, ~v80, ~v90, ~v100,
  # ---- Table 2 ----
  "bw", "Female", 60.0, 63.3, 65.2, 65.5, 62.2, 56.1, 53.7, 45.2,
  "bw", "Male", 73.0, 74.4, 73.0, 72.0, 71.1, 68.1, 64.9, 55.6,
  "ht", "Female", 163.0, 160.9, 160.0, 158.3, 155.4, 151.7, 150.0, 147.9,
  "ht", "Male", 176.0, 175.1, 173.9, 171.2, 167.4, 165.5, 163.8, 157.6,
  "mass_blood", "Female", 1899.4, 1921.1, 1961.2, 1946.9, 1847.7, 1692.6, 1634.5, 1357.0,
  "mass_blood", "Male", 2264.6, 2269.5, 2234.9, 2180.2, 2111.3, 2042.5, 1976.5, 1739.5,
  "mass_bone", "Female", 9121.5, 8886.3, 8676.1, 8401.8, 7811.6, 6981.4, 6619.0, 6263.5,
  "mass_bone", "Male", 11817.8, 11848.0, 11481.6, 11091.6, 10355.0, 9637.1, 8969.0, 8300.9,
  "mass_brain", "Female", 1357.0, 1352.3, 1347.6, 1315.1, 1287.3, 1241.4, 1169.4, 1097.4,
  "mass_brain", "Male", 1508.8, 1506.9, 1502.9, 1463.3, 1434.3, 1400.1, 1372.1, 1344.2,
  "mass_fat", "Female", 19348.0, 20002.5, 23815.4, 26993.0, 27884.7, 25694.5, 25020.6, 20171.4,
  "mass_fat", "Male", 14868.0, 16309.7, 16676.5, 17920.7, 19447.4, 19947.7, 20691.0, 17082.3,
  "mass_gonads", "Female", 13.1, 13.0, 6.6, 5.3, 5.2, 5.1, 5.1, 4.9,
  "mass_gonads", "Male", 40.3, 40.8, 35.0, 33.5, 31.9, 30.6, 30.3, 29.5,
  "mass_heart", "Female", 328.4, 340.0, 355.6, 377.1, 399.1, 413.5, 431.2, 412.9,
  "mass_heart", "Male", 417.2, 434.3, 439.1, 454.6, 464.8, 441.4, 437.9, 425.1,
  "mass_kidney", "Female", 403.4, 401.7, 400.5, 383.4, 364.2, 325.2, 308.8, 295.8,
  "mass_kidney", "Male", 437.7, 475.9, 468.1, 455.6, 441.7, 395.7, 366.7, 355.9,
  "mass_liver", "Female", 1905.5, 1881.0, 1867.7, 1679.8, 1504.9, 1423.4, 1329.5, 1189.4,
  "mass_liver", "Male", 2357.8, 2324.5, 2206.0, 2041.3, 1714.6, 1417.3, 1324.0, 1202.0,
  "mass_lung", "Female", 1009.5, 1021.4, 1038.8, 1023.8, 1005.4, 849.1, 804.5, 637.3,
  "mass_lung", "Male", 1294.3, 1334.7, 1337.4, 1349.0, 1271.8, 1138.0, 1007.4, 978.0,
  "mass_muscle", "Female", 20276.2, 23058.3, 21300.1, 18971.2, 15746.4, 13367.8, 12394.3, 10060.6,
  "mass_muscle", "Male", 32338.6, 32318.2, 31180.6, 29600.0, 28562.1, 26524.8, 23657.0, 19534.0,
  "mass_pancreas", "Female", 169.5, 170.3, 163.8, 158.7, 148.5, 131.9, 128.0, 122.6,
  "mass_pancreas", "Male", 190.3, 190.8, 183.6, 178.2, 165.4, 149.9, 143.7, 139.5,
  "mass_skin", "Female", 2723.5, 2773.2, 2836.2, 2829.5, 2725.8, 2559.1, 2493.8, 2223.5,
  "mass_skin", "Male", 3760.9, 3790.2, 3745.1, 3692.0, 3635.2, 3537.0, 3439.7, 3122.6,
  "mass_spleen", "Female", 219.2, 197.7, 190.7, 182.4, 164.5, 149.7, 99.4, 78.7,
  "mass_spleen", "Male", 243.4, 221.9, 208.2, 197.0, 186.6, 160.4, 133.4, 108.4,
  # ---- Table 3 ----
  "cardiac_index", "Female", 3.34, 3.33, 3.08, 2.89, 2.75, 2.68, 2.52, 2.41,
  "cardiac_index", "Male", 3.23, 3.22, 3.09, 2.86, 2.58, 2.36, 2.18, 2.04,
  "flow_adipose", "Female", 501.1, 528.2, 578.0, 655.3, 684.7, 630.2, 607.0, 491.5,
  "flow_adipose", "Male", 324.7, 352.1, 357.3, 381.4, 408.0, 419.8, 429.2, 357.3,
  "flow_cerebral", "Female", 707.8, 686.9, 666.1, 632.1, 601.2, 562.8, 514.2, 467.6,
  "flow_cerebral", "Male", 779.7, 747.6, 727.9, 692.9, 650.6, 623.8, 594.4, 536.1,
  "flow_gonads", "Female", 1.2, 1.2, 0.6, 0.5, 0.5, 0.5, 0.5, 0.4,
  "flow_gonads", "Male", 3.2, 3.3, 2.8, 2.7, 2.6, 2.5, 2.4, 2.4,
  "flow_myocardial", "Female", 295.0, 311.2, 313.2, 363.1, 376.3, 445.8, 464.8, 445.2,
  "flow_myocardial", "Male", 260.1, 271.1, 278.0, 299.1, 308.7, 334.7, 332.1, 322.3,
  "flow_renal", "Female", 1121.0, 1061.8, 943.8, 825.8, 707.8, 589.8, 471.8, 353.8,
  "flow_renal", "Male", 1325.0, 1339.3, 1116.3, 913.3, 730.3, 567.3, 424.3, 301.3,
  "flow_splanchnic", "Female", 1239.0, 1254.3, 1187.2, 1015.3, 862.6, 771.5, 679.1, 570.5,
  "flow_splanchnic", "Male", 1235.0, 1241.9, 1237.7, 1056.7, 813.1, 610.6, 512.9, 413.5,
  "flow_hepatic", "Female", 383.4, 388.1, 367.3, 314.2, 266.9, 238.7, 210.1, 176.5,
  "flow_hepatic", "Male", 423.0, 425.4, 423.9, 361.9, 278.5, 209.1, 175.7, 141.6,
  "flow_muscle", "Female", 665.1, 774.8, 636.6, 568.4, 479.8, 405.6, 369.0, 298.8,
  "flow_muscle", "Male", 1105.7, 1086.8, 1036.8, 975.0, 922.3, 861.7, 751.9, 625.9,
  "flow_skeleton", "Female", 294.9, 287.3, 280.5, 271.6, 252.6, 225.7, 214.0, 202.5,
  "flow_skeleton", "Male", 324.9, 325.8, 315.7, 305.0, 284.7, 265.0, 246.6, 228.2,
  "flow_skin", "Female", 295.7, 301.1, 307.9, 307.2, 295.9, 277.8, 270.7, 241.4,
  "flow_skin", "Male", 325.1, 327.7, 323.8, 319.2, 314.3, 305.8, 297.4, 270.0
) |>
  tidyr::pivot_longer(
    # Name the age columns explicitly: starts_with("v") would also match
    # the `var` column and silently poison the pivot.
    cols = dplyr::all_of(paste0("v", seq(30, 100, by = 10))),
    names_to = "age", values_to = "published"
  ) |>
  dplyr::mutate(age = as.numeric(sub("^v", "", .data$age)))

nrow(ref)
```

```{r knot-check}
got <- sim_knot |>
  dplyr::select(age = "time", "sex", dplyr::all_of(unique(ref$var))) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(unique(ref$var)),
    names_to = "var", values_to = "simulated"
  )

cmp <- ref |>
  dplyr::inner_join(got, by = c("var", "sex", "age")) |>
  dplyr::mutate(abs_diff = abs(.data$simulated - .data$published))

# The join must not silently drop rows -- a lost row would make the
# gate below pass vacuously (see known-vignette-failure-patterns #10).
stopifnot(
  nrow(cmp) == nrow(ref),
  nrow(cmp) == 416L,
  !anyNA(cmp$simulated)
)

# Deterministic algebra: the model must return the published cells to
# machine precision, not "close to" them.
stopifnot(max(cmp$abs_diff) < 1e-8)

c(cells = nrow(cmp), max_abs_diff = max(cmp$abs_diff))
```

All 416 published cells are returned exactly.

The constant gastrointestinal mass of Table 2 footnote a is checked separately
because it has no age dimension:

```{r gi-check}
gi <- sim_fine |>
  dplyr::group_by(.data$sex) |>
  dplyr::summarise(lo = min(.data$mass_gi), hi = max(.data$mass_gi), .groups = "drop")

stopifnot(
  # Age-invariant within each sex.
  all(abs(gi$hi - gi$lo) < 1e-8),
  abs(gi$lo[gi$sex == "Female"] - 1274.5) < 1e-8,
  abs(gi$lo[gi$sex == "Male"] - 1304.8) < 1e-8
)
gi
```

## 2. Interpolation behaviour between and beyond the knots

The paper interpolates linearly between the ten-year bins. Between knots the
model must therefore sit exactly on the chord, and outside 30 to 100 years the
trajectory is held flat rather than extrapolating the last decade's slope.

```{r interp-check}
mid <- dplyr::bind_rows(solve_sex(1, c(35, 45, 95)), solve_sex(0, c(35, 45, 95)))
kn <- sim_knot

chord <- function(v, sex, a) {
  lo <- kn[[v]][kn$sex == sex & kn$time == a - 5]
  hi <- kn[[v]][kn$sex == sex & kn$time == a + 5]
  (lo + hi) / 2
}

midpoint_err <- vapply(
  c("mass_liver", "mass_kidney", "flow_renal", "bw"),
  function(v) {
    max(abs(vapply(
      c(35, 45, 95),
      function(a) {
        max(abs(c(
          mid[[v]][mid$sex == "Female" & mid$time == a] - chord(v, "Female", a),
          mid[[v]][mid$sex == "Male" & mid$time == a] - chord(v, "Male", a)
        )))
      },
      numeric(1)
    )))
  },
  numeric(1)
)

stopifnot(max(midpoint_err) < 1e-8)

# Flat outside the published 30-100 y range.
outside <- dplyr::bind_rows(solve_sex(1, c(10, 25, 30, 100, 110)), solve_sex(0, c(10, 25, 30, 100, 110)))
flat_lo <- outside |> dplyr::filter(.data$time %in% c(10, 25, 30))
flat_hi <- outside |> dplyr::filter(.data$time %in% c(100, 110))
stopifnot(
  dplyr::n_distinct(round(flat_lo$mass_liver[flat_lo$sex == "Male"], 8)) == 1L,
  dplyr::n_distinct(round(flat_hi$mass_liver[flat_hi$sex == "Male"], 8)) == 1L
)

midpoint_err
```

## 3. Cardiac-output mass balance

The paper states plainly that "the sum of all blood flows should be equal to
CO" (Methods), and that age-related changes in blood flow distribution were
accounted for as changes in cardiac output. This is an internal-consistency gate
on Table 3 that the paper never reports numerically.

```{r co-balance}
flow_cols <- c(
  "flow_adipose", "flow_cerebral", "flow_gonads", "flow_myocardial",
  "flow_renal", "flow_splanchnic", "flow_hepatic", "flow_muscle",
  "flow_skeleton", "flow_skin"
)

balance <- sim_knot |>
  dplyr::mutate(
    sum_flows = rowSums(dplyr::across(dplyr::all_of(flow_cols))),
    ratio = .data$sum_flows / .data$cardiac_output
  ) |>
  dplyr::select(age = "time", "sex", "sum_flows", "cardiac_output", "ratio")

stopifnot(nrow(balance) == 16L)

# The residual absorbs the organs Schlender 2016 excludes from Table 3
# (the paper notes the modelled organ masses reach only 91-95 % of body
# weight), the 3-significant-figure rounding of the published cells, and
# the unstated choice of BSA formula. 5 % leaves room for all three and
# still goes red on a mis-transcribed flow, which would move a single
# organ's contribution by tens of percent.
stopifnot(max(abs(balance$ratio - 1)) < 0.05)

balance |>
  dplyr::mutate(dplyr::across(c("sum_flows", "cardiac_output"), ~ round(.x, 0)), ratio = round(.data$ratio, 4)) |>
  dplyr::rename(
    "Age (y)" = "age", "Sex" = "sex",
    "Sum of Table 3 flows (mL/min)" = "sum_flows",
    "Cardiac output (mL/min)" = "cardiac_output",
    "Ratio" = "ratio"
  ) |>
  knitr::kable()
```

The flows sum to cardiac output within 2.6 % at every one of the 16 age-by-sex
cells, and within 0.2 % at the ICRP-derived 30-year reference. Table 3 is
internally consistent.

## 4. The GFR aging function (Equation 1), replicating Figure 5

```{r gfr-anchor}
# At age 30 the aging term is exactly zero, so specific GFR must equal
# F_PMA and absolute GFR must equal F_PMA * kidney mass / 100.
at30 <- sim_knot |> dplyr::filter(.data$time == 30)
stopifnot(
  all(abs(at30$gfr_specific - 26.6) < 1e-10),
  all(abs(at30$gfr - 26.6 * at30$mass_kidney / 100) < 1e-10)
)

# Specific GFR is strictly decreasing across the published range, and the
# female half-time (59 y) is longer than the male (54 y), so at any age
# above 30 the female retains a larger FRACTION of F_PMA.
sf <- sim_fine |> dplyr::filter(.data$sex == "Female") |> dplyr::arrange(.data$time)
sm <- sim_fine |> dplyr::filter(.data$sex == "Male") |> dplyr::arrange(.data$time)
stopifnot(
  all(diff(sf$gfr_specific) < 0),
  all(diff(sm$gfr_specific) < 0),
  all(sf$gfr_specific[sf$time > 30] > sm$gfr_specific[sm$time > 30])
)

round(at30$gfr, 1)
```

```{r gfr-figure, fig.cap = "Replicates Figure 5 of Schlender 2016: absolute glomerular filtration rate over the course of aging, female (left) and male (right). The published panels show the fitted line falling from roughly 105 to 35 mL/min in females and from roughly 118 to 40 mL/min in males between 30 and 100 years."}
ggplot2::ggplot(sim_fine, ggplot2::aes(.data$time, .data$gfr)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_wrap(~sex) +
  ggplot2::scale_y_continuous(limits = c(0, NA)) +
  ggplot2::labs(
    x = "Age (years)",
    y = "Glomerular filtration rate (mL/min)"
  ) +
  ggplot2::theme_bw()
```

```{r gfr-vs-figure5}
gfr_ends <- sim_knot |>
  dplyr::filter(.data$time %in% c(30, 100)) |>
  dplyr::select(age = "time", "sex", "gfr_specific", "gfr")

# Read off the published Fig. 5 panels: the fitted GFR line starts near
# 105 (female) / 118 (male) mL/min at age 30 and ends near 35 / 40 at age
# 100. The model is the paper's own equation applied to the paper's own
# kidney masses, so agreement should be within the precision of reading a
# published axis -- 8 mL/min is about one third of a minor gridline.
fig5 <- c("Female_30" = 105, "Male_30" = 118, "Female_100" = 35, "Male_100" = 40)
key <- paste(gfr_ends$sex, gfr_ends$age, sep = "_")
stopifnot(all(key %in% names(fig5)))
stopifnot(max(abs(gfr_ends$gfr - fig5[key])) < 8)

gfr_ends |>
  dplyr::mutate(
    read_from_fig5 = fig5[paste(.data$sex, .data$age, sep = "_")],
    dplyr::across(c("gfr_specific", "gfr"), ~ round(.x, 1))
  ) |>
  dplyr::rename(
    "Age (y)" = "age", "Sex" = "sex",
    "Specific GFR (mL/min/100 g)" = "gfr_specific",
    "GFR (mL/min)" = "gfr",
    "Read from published Fig. 5" = "read_from_fig5"
  ) |>
  knitr::kable()
```

## 5. Replicating Figure 3 (body height) and Figure 2 (organ masses)

```{r fig3, fig.cap = "Replicates Figure 3 of Schlender 2016: age-dependent decrease in body height in females and males over 30 to 100 years."}
ggplot2::ggplot(sim_fine, ggplot2::aes(.data$time, .data$ht, colour = .data$sex)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::labs(x = "Age (years)", y = "Body height (cm)", colour = "Sex") +
  ggplot2::theme_bw()
```

```{r fig2, fig.cap = "Replicates the 30-100 y portion of Figure 2 of Schlender 2016: distribution of mean organ and tissue weights across the course of aging. Fat rises then falls while muscle and liver decline throughout."}
sim_fine |>
  dplyr::select(
    "time", "sex",
    "mass_fat", "mass_muscle", "mass_bone", "mass_skin",
    "mass_liver", "mass_brain", "mass_blood"
  ) |>
  tidyr::pivot_longer(dplyr::starts_with("mass_"), names_to = "tissue", values_to = "g") |>
  dplyr::mutate(tissue = sub("^mass_", "", .data$tissue)) |>
  ggplot2::ggplot(ggplot2::aes(.data$time, .data$g / 1000, colour = .data$tissue)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~sex) +
  ggplot2::labs(x = "Age (years)", y = "Mass (kg)", colour = "Tissue") +
  ggplot2::theme_bw()
```

## 6. The paper's prose claims, checked against its own tables

Schlender 2016 states a number of quantitative aging rates in the Results
narrative. None of them is a model parameter, so none is exercised by the checks
above -- but each is a claim about the very tables this model encodes, and a
transcription error would break the agreement. Converting them into assertions
is therefore a genuine gate, and it also surfaces several places where the prose
and the tables disagree.

```{r prose-claims}
at <- function(v, sex, a) sim_knot[[v]][sim_knot$sex == sex & sim_knot$time == a]
per_decade_pct <- function(v, sex, a0, a1) {
  ((at(v, sex, a1) / at(v, sex, a0))^(10 / (a1 - a0)) - 1) * 100
}

claims <- tibble::tribble(
  ~Claim, ~Source, ~Published, ~Achieved, ~Tol, ~Deviation,
  "Male cardiac index falls 8.30 %/decade in the elderly (60-100 y)",
  "Results 3.1.6", -8.30, per_decade_pct("cardiac_index", "Male", 60, 100), 0.5, FALSE,

  "Splanchnic blood flow is 22.5 % of cardiac output in adults (female, 30 y)",
  "Results 3.1.8", 22.5,
  100 * at("flow_splanchnic", "Female", 30) / at("cardiac_output", "Female", 30), 0.5, FALSE,

  "Splanchnic blood flow falls to 17.5 % of cardiac output in the tenth decade (female, 100 y)",
  "Results 3.1.8", 17.5,
  100 * at("flow_splanchnic", "Female", 100) / at("cardiac_output", "Female", 100), 0.5, FALSE,

  "Male fat mass increases by 0.978 kg/decade, constantly (30-90 y)",
  "Results 3.1.3", 0.978,
  (at("mass_fat", "Male", 90) - at("mass_fat", "Male", 30)) / 1000 / 6, 0.05, FALSE,

  "Female fat mass gains 3.50 kg/decade over the first two decades after menopause (40-60 y)",
  "Results 3.1.3", 3.50,
  (at("mass_fat", "Female", 60) - at("mass_fat", "Female", 40)) / 1000 / 2, 0.05, FALSE,

  "Relative fat mass reaches 45.8 % in female centenarians",
  "Results 3.1.3", 45.8,
  100 * at("mass_fat", "Female", 100) / 1000 / at("bw", "Female", 100), 1.5, FALSE,

  "Relative fat mass reaches 31.5 % in male centenarians",
  "Results 3.1.3", 31.5,
  100 * at("mass_fat", "Male", 100) / 1000 / at("bw", "Male", 100), 1.5, FALSE,

  "Male skeletal muscle is 44 % of body weight in adults (30 y)",
  "Results 3.1.2", 44.0,
  100 * at("mass_muscle", "Male", 30) / 1000 / at("bw", "Male", 30), 1.0, FALSE,

  "Female skeletal muscle falls to 22 % of body weight in centenarians",
  "Results 3.1.2", 22.0,
  100 * at("mass_muscle", "Female", 100) / 1000 / at("bw", "Female", 100), 1.0, FALSE,

  "Male skeletal muscle falls to 34 % of body weight in the extremely old (100 y)",
  "Results 3.1.2", 34.0,
  100 * at("mass_muscle", "Male", 100) / 1000 / at("bw", "Male", 100), 1.5, FALSE,

  "Male body height loses about 3.5 %/decade in the tenth decade",
  "Results 3.1.1", -3.5, per_decade_pct("ht", "Male", 90, 100), 0.5, FALSE,

  "Male liver mass falls 11.4 %/decade after 60 y",
  "Results 3.1.5", -11.4, per_decade_pct("mass_liver", "Male", 60, 100), 1.2, FALSE,

  # ---- Documented deviations: prose vs the paper's own tables ----
  "Female liver mass falls 8.3 %/decade after 40 y",
  "Results 3.1.5", -8.3, per_decade_pct("mass_liver", "Female", 40, 100), 0.5, TRUE,

  "Male liver mass falls 6.1 %/decade up to 60 y",
  "Results 3.1.5", -6.1, per_decade_pct("mass_liver", "Male", 30, 60), 0.5, TRUE,

  "Female cardiac index falls almost linearly by 3.22 %/decade after 30 y",
  "Results 3.1.6", -3.22, per_decade_pct("cardiac_index", "Female", 30, 100), 0.5, TRUE,

  "Male cardiac index falls 1.46 %/decade during early adulthood (30-60 y)",
  "Results 3.1.6", -1.46, per_decade_pct("cardiac_index", "Male", 30, 60), 0.5, TRUE,

  "Female body height loses 4.2 %/decade in nonagenarians",
  "Results 3.1.1", -4.2, per_decade_pct("ht", "Female", 90, 100), 0.5, TRUE,

  "Body weight decline rises to nearly 10 %/decade between 90 and 100 y (female)",
  "Results 3.1.1", -10.0, per_decade_pct("bw", "Female", 90, 100), 1.5, TRUE
) |>
  dplyr::mutate(
    Achieved = round(.data$Achieved, 2),
    Pass = abs(.data$Achieved - .data$Published) <= .data$Tol
  )

# Every non-deviation claim must reproduce. The deviations are kept visible
# in the rendered table rather than having their tolerance widened until
# they disappear.
stopifnot(all(claims$Pass[!claims$Deviation]))
# ... and each recorded deviation must genuinely still be a deviation, so
# this gate goes red if a future edit silently fixes or breaks one.
stopifnot(!any(claims$Pass[claims$Deviation]))

claims |>
  dplyr::select("Claim", "Source", "Published", "Achieved", "Pass", "Deviation") |>
  knitr::kable()
```

Twelve of the eighteen prose claims reproduce from the tables within tolerance.
The six that do not are discussed in the Errata below.

# Assumptions and deviations

## Errata: prose that disagrees with the paper's own Tables 2 and 3

These are disagreements **within the source**, not transcription choices. The
tables are used, because the tables are what the PBPK model was actually
parameterised with ("as used in the elderly PBPK models" -- Table 2 caption).

1. **Female body height, "a 4.2 % decrease in body height per decade in female
   nonagenarians" (Results 3.1.1).** Table 2 gives 150.0 cm at 90 y and 147.9 cm
   at 100 y, a **1.4 %** decrease per decade -- three-fold smaller. No decade in
   the female height row reaches 4.2 %; the largest is 2.4 % (70 to 80 y), and
   even the whole 70-to-100-year span averages 1.6 %/decade. The corresponding
   male claim ("about 3.5 %" in the tenth decade) does reproduce, at 3.8 %.

2. **Body weight, "rose to nearly 10 % per decade between 90 and 100 years"
   (Results 3.1.1).** Table 2 gives 53.7 to 45.2 kg in females (**15.8 %**) and
   64.9 to 55.6 kg in males (**14.3 %**). The prose understates the tabulated
   decline by roughly half.

3. **Female cardiac index, "an almost linear decrease of 3.22 % per decade after
   the age of 30 years" (Results 3.1.6).** The Table 3 female row falls from
   3.34 to 2.41 L/min/m^2 over 30 to 100 y, which is **4.56 %/decade**, and it is
   visibly non-linear (essentially flat from 30 to 40 y, then steepening).

4. **Male cardiac index, "1.46 % per decade during early adulthood"
   (Results 3.1.6).** Table 3 gives only 0.31 %/decade from 30 to 40 y and
   3.97 %/decade averaged over 30 to 60 y; no window in the tabulated range
   yields 1.46 %. The companion elderly claim (8.30 %/decade) reproduces almost
   exactly, at 8.10 %.

5. **Liver mass rates (Results 3.1.5).** The female "8.3 %/decade after 40 years"
   is 7.35 %/decade in Table 2, and the male "6.1 %/decade to the age of 60" is
   4.69 %/decade. The male post-60 rate ("11.4 %/decade") does reproduce, at
   12.4 %.

6. **Male body weight peak.** The prose says "both sexes showed an increase in
   body mass up to the age of 55 years". The female row does peak between the
   50- and 60-year knots, but the male row peaks at the **40**-year knot
   (74.4 kg) and declines thereafter. The decade knots cannot resolve an age of
   55 in any case.

Two further prose statements are reconciled rather than contradicted, and the
window used in the assertion table above reflects the reconciliation:

* **"3.50 kg fat mass per decade in the first two decades after menopause"**
  reproduces to three significant figures only over the **40-to-60-year**
  window (3.495 kg/decade), not the 50-to-70-year window a literal reading of
  "after menopause" suggests (2.04 kg/decade).
* **"37 % of body weight in adults"** for female skeletal muscle is the value at
  the **40**-year knot (36.4 %), which is where the female muscle row peaks, not
  at 30 y (33.8 %).

## Assumptions made by this extraction

1. **`gfr_vmax` is dimensionless.** The prose tags the maximal decreasing rate
   factor as "0.9 mL/min/100 g kidney weight", carrying over the unit of
   `F_PMA` from the preceding sentence. Equation 1 multiplies `Vmax` by a
   unitless Hill ratio inside `(1 - ...)`, so `Vmax` must be a dimensionless
   fraction for the equation to be dimensionally consistent -- and the resulting
   curve matches the published Figure 5. It is encoded as a fraction.

2. **Body surface area uses the Du Bois equation.** Table 3 reports a cardiac
   *index* (per m^2 of BSA) but the paper never states which BSA formula it
   used, so recovering cardiac output requires a choice. Du Bois is used. The
   choice is not load-bearing: with the Mosteller equation instead, the
   flow-sum-to-cardiac-output ratios of validation section 3 move by less than
   1.5 percentage points and every cell still balances, and `cardiac_index`
   itself -- the published quantity -- is unaffected.

3. **Flat extrapolation outside 30 to 100 years.** Tables 2 and 3 are defined on
   the 30-to-100-year knots. Below 30 y and above 100 y the model holds the end
   knot rather than continuing the adjacent decade's slope, which the paper does
   not support. Figure 2 of the paper does show organ weights from birth, but
   those come from the underlying PK-Sim paediatric model, not from this paper's
   tables, and are not extracted.

4. **No variability is encoded.** The paper assumed a 5 % coefficient of
   variation on every blood flow rate because "the general lack of perfusion
   data impeded computation of robust variability" (Results 3.1.8), and made GFR
   variability depend entirely on kidney-size variability -- which the
   Discussion itself judges "too great". Neither is a fitted between-subject
   variance, so neither is encoded as an `eta`; this model returns typical
   values. Organ-mass variability is likewise not reported per organ.

5. **The GFR function is evaluated against mean kidney mass.** The paper
   parameterised specific GFR "using the observed GFR values and assuming the
   mean kidney weight at the respective age" (Results 3.1.10), which is exactly
   what `gfr <- gfr_specific * mass_kidney / 100` does with the Table 2 row.

6. **`SEXF` is an indicator, not a proportion.** Every trajectory is published
   as separate female and male rows and the model blends them linearly by
   `SEXF`. Values strictly between 0 and 1 produce a sex-averaged physiology
   that the paper never reports and that should not be interpreted as a real
   individual.

## Not extracted from this paper

* The **morphine and furosemide PBPK drug models** (see "What is deliberately
  not extracted" at the top), including Table 1's compound cards and Table 5's
  predictive-performance statistics.
* **Total body water** (Figure 4). The paper plots extracellular and total body
  water as percentages of body weight against age but tabulates no values, and
  the Discussion notes that no body-water data were available for females above
  80 years.
* The per-organ literature review in the Electronic Supplementary Material,
  which is narrative plus Supplement Table 1 (a list of the contributing
  studies) and contains no additional parameter values.
