---
title: "NLG207 conjugated and free camptothecin (Schmidt 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NLG207 conjugated and free camptothecin (Schmidt 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20200908)
set.seed(20200908)
```

# Model and source

```{r model}
mod <- rxode2::rxode2(readModelDb("Schmidt_2020_camptothecin"))
mod_typ <- rxode2::zeroRe(mod)
```

* Citation: `r mod$reference`
* Article: <https://doi.org/10.1007/s00280-020-04134-9>
* PubMed Central (open access): <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7515962/>

NLG207 (formerly CRLX101) is a nanoparticle-drug conjugate in which the
topoisomerase-I inhibitor camptothecin (CPT) is covalently linked to a
cyclodextrin-containing polymer that self-assembles into nanoparticles. CPT is
released from the polymer by hydrolysis. Schmidt 2020 measured nanoparticle-bound
(conjugated) and free CPT in the same plasma samples and fit both together in
NONMEM (FOCE-I, ADVAN13):

* **Conjugated CPT** (`central_np`, `peripheral1_np`; output `Cc_np`) follows a
  linear two-compartment model. Its only route out of the body is conversion to
  free CPT in the central compartment.
* **The release clearance** is the sum of two parts (Figure 1). One is a
  steady-state part, `CL_B`. The other is a fast part, `CL_F`, that decays
  mono-exponentially with time after the start of each infusion:
  `CL1 = CL_B + CL_F * exp(-(ln 2 / t1/2) * t)`.
* **Free CPT** (`central`, `peripheral1`; output `Cc`) follows a second linear
  two-compartment model with first-order elimination (`CL3`). The fraction of
  conjugated CPT that ends up as free CPT is not identifiable, so the free-CPT
  volumes and clearances are apparent values relative to that fraction.

All clearances and volumes except `CL_F` and `t1/2` are scaled allometrically
on body weight, with fixed exponents (1 for volumes, 0.75 for clearances) and a
reference weight of 70 kg.

The `_np` suffix follows the library's nanoparticle convention: the bare
compartment and output names hold the free (released) drug, and the `_np`
suffix holds the drug that is still conjugated.

# Population

```{r population, echo = FALSE}
pop <- mod$population
data.frame(
  Field = c(
    "Subjects", "Studies", "Observations", "Age", "Weight", "Female",
    "Race", "Disease", "Dosing"
  ),
  Value = c(
    pop$n_subjects, pop$n_studies, pop$n_observations,
    paste0(pop$age_median, " (", pop$age_range, ")"),
    paste0(pop$weight_median, " (", pop$weight_range, ")"),
    paste0(pop$sex_female_pct, "%"),
    paste(names(pop$race_ethnicity), paste0(pop$race_ethnicity, "%"), collapse = ", "),
    pop$disease_state, pop$dose_range
  )
) |>
  knitr::kable(caption = "Study population (Schmidt 2020 Table 1).")
```

There were 27 adults with advanced solid tumours in two phase II trials at the
US National Cancer Institute: NCT02769962 (NLG207 plus olaparib, n = 24) and
NCT03531827 (NLG207 plus enzalutamide, n = 3). All received NLG207 12 mg/m^2 IV
over 1 or 2 h at cycle 1. Dense samples were taken up to 24 or 48 h after the
end of infusion, plus a sample at about 14 to 16 days in NCT02769962. Five
patients were sampled again at cycle 6. Cycle-1 samples were collected before
the combination partner was given.

# Source trace

| Model element | Value | Source |
|---|---|---|
| Structure: 2-cmt conjugated + 2-cmt free CPT, release only from conjugated central | -- | Methods 'Pharmacokinetic model-building procedure'; Fig. 1 |
| `CL1 = CL_B + CL_F * exp(-(ln 2 / t1/2) * t)` | -- | Fig. 1 equation |
| Time origin of `t` = start of each infusion | -- | Discussion (release half-life 0.38 h at start of infusion, 22 h at 4 h) |
| `lvc_np` (V1, 70 kg) | 3.16 L | Table 2 |
| `lvp_np` (V2, 70 kg) | 2.09 L | Table 2 |
| `lq_np` (Q1, 70 kg) | 0.0482 L/h | Table 2 |
| `lcl_exp_inf_np` (CL_B, 70 kg) | 0.0988 L/h | Table 2 |
| `lcl_exp_component_np` (CL_F) | 5.71 L/h | Table 2 |
| `lcl_exp_kdes_np` | ln(2)/0.307 = 2.258 1/h | Table 2 (t1/2 = 0.307 h) |
| `lvc` (V3, 70 kg) | 21.1 L | Table 2 |
| `lvp` (V4, 70 kg) | 19.4 L | Table 2 |
| `lq` (Q3, 70 kg) | 25.6 L/h | Table 2 |
| `lcl` (CL3, 70 kg) | 0.874 L/h | Table 2 |
| `e_wt_vc_vp`, `e_wt_cl_q` | 1, 0.75 (fixed) | Table 2 formulas; Results 'Covariate model' |
| `etalvc_np`, `etalcl_exp_inf_np` (block) | CV 18.1%, 33.5%; r = 0.918 | Table 2 |
| `etalcl_exp_component_np` | CV 62.6% | Table 2 |
| `etalvc`, `etalcl` (block) | CV 79.8%, 42.2%; r = 0.884 | Table 2 |
| `propSd_np`, `addSd_np` | 12.3%, 5.07 ng/mL | Table 2 'BoundRE' |
| `propSd`, `addSd` | 24.8%, 0.396 ng/mL | Table 2 'FreeRE' |
| Residual form `C * (1 + eps_prop) + eps_add` | -- | Methods Eq. 2 |

# Deterministic checks against the paper's derived quantities

The paper does not report an NCA table. Its Discussion does quote several
quantities that follow directly from the typical parameters, and each one can be
recomputed without any random draw. The first three check the release clearance,
and the last two check the free-CPT disposition parameters.

```{r derived}
th <- mod$theta
v1 <- exp(th[["lvc_np"]])
clb <- exp(th[["lcl_exp_inf_np"]])
clf <- exp(th[["lcl_exp_component_np"]])
kdes <- exp(th[["lcl_exp_kdes_np"]])
cl1 <- function(t) clb + clf * exp(-kdes * t)
release_hl <- function(t) log(2) * v1 / cl1(t)

biexp_hl <- function(cl, v1, q, v2) {
  k10 <- cl / v1
  k12 <- q / v1
  k21 <- q / v2
  s <- k10 + k12 + k21
  d <- sqrt(s^2 - 4 * k10 * k21)
  c(alpha = log(2) / ((s + d) / 2), beta = log(2) / ((s - d) / 2))
}
hl_free <- biexp_hl(
  exp(th[["lcl"]]), exp(th[["lvc"]]), exp(th[["lq"]]), exp(th[["lvp"]])
)

derived <- tibble::tribble(
  ~Quantity, ~Model, ~Published,
  "Release half-life at start of infusion (h)", release_hl(0), 0.38,
  "Release half-life at 4 h after start (h)", release_hl(4), 22,
  "CL_F contribution / CL_B at 1.8 h", clf * exp(-kdes * 1.8) / clb, 1,
  "Free CPT distribution half-life (min)", 60 * hl_free[["alpha"]], 16,
  "Free CPT terminal half-life (h)", hl_free[["beta"]], 32.4
) |>
  dplyr::mutate(`% diff` = 100 * (.data$Model - .data$Published) / .data$Published)

derived |>
  dplyr::mutate(Model = signif(.data$Model, 3), `% diff` = round(.data$`% diff`, 1)) |>
  knitr::kable(caption = "Typical-value quantities quoted in the Schmidt 2020 Discussion.")

# Deterministic: no random effects are involved, so a tight bound is correct.
# The published values are rounded to 2-3 significant figures ("~16 min",
# "near equal contribution").
stopifnot(all(abs(derived$`% diff`) < 5))
```

All five agree to within the rounding of the published text. The 0.38 h and
22 h release half-lives pin both the release equation and its time origin at the
start of the infusion. If `t` were measured from the end of a 1 h infusion
instead, the half-life at 4 h after the start would be about 21 h, not 22 h.

# Typical-subject profile and mass balance

The typical 70 kg subject has a BSA of about 1.8 m^2, so a 12 mg/m^2 dose over
1 h is about 21.6 mg of CPT.

```{r typical}
make_events <- function(subjects, obs_times, n_doses = 1L) {
  dose_times <- (seq_len(n_doses) - 1) * 336
  doses <- tidyr::crossing(subjects, time = dose_times) |>
    dplyr::mutate(
      amt = .data$dose_mg, evid = 1L, cmt = "central_np",
      dur = .data$tinf, dvid = NA_integer_
    )
  obs <- tidyr::crossing(subjects, time = obs_times) |>
    dplyr::mutate(amt = 0, evid = 0L, cmt = NA_character_, dur = NA_real_, dvid = 1L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(.data$id, .data$time, dplyr::desc(.data$evid)) |>
    dplyr::select("id", "time", "amt", "evid", "cmt", "dur", "dvid", dplyr::everything()) |>
    as.data.frame()
}

typ_subject <- data.frame(id = 1L, WT = 70, dose_mg = 12 * 1.8, tinf = 1)
obs_fine <- sort(unique(c(seq(0, 4, by = 0.05), seq(4.25, 48, by = 0.25), seq(49, 2000, by = 1))))
sim_typ <- rxode2::rxSolve(
  mod_typ, make_events(typ_subject, obs_fine),
  returnType = "data.frame"
)
```

```{r typical-plot, fig.cap = "Typical-subject conjugated and free CPT after 12 mg/m^2 over 1 h (70 kg, BSA 1.8 m^2)."}
sim_typ |>
  dplyr::filter(.data$time <= 96) |>
  dplyr::select("time", "Conjugated CPT" = "Cc_np", "Free CPT" = "Cc") |>
  tidyr::pivot_longer(-"time", names_to = "Analyte", values_to = "conc") |>
  dplyr::filter(.data$conc > 0) |>
  ggplot(aes(.data$time, .data$conc, colour = .data$Analyte)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time after start of infusion (h)", y = "Concentration (ng/mL)", colour = NULL)
```

Conjugated CPT has no elimination route other than release, so every milligram
dosed eventually leaves the body as free CPT through `CL3`. The dose must
therefore equal `CL3 * AUC(0-inf)` of free CPT. This check is exact, and it
would fail if the release flux were lost or double-counted between the two
sub-models.

```{r mass-balance}
cl3_typ <- exp(th[["lcl"]])
auc_free <- with(
  dplyr::filter(sim_typ, !is.na(.data$Cc)),
  sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2)
)
# Remaining tail beyond the last point, using the conjugated terminal slope that
# governs the free-CPT decline at late times.
tail_end <- tail(sim_typ, 2)
lz <- log(tail_end$Cc[1] / tail_end$Cc[2]) / diff(tail_end$time)
auc_free_inf <- auc_free + tail(sim_typ$Cc, 1) / lz
recovered_mg <- cl3_typ * auc_free_inf / 1000 # L/h * ng.h/mL = ug -> mg
c(dose_mg = typ_subject$dose_mg, recovered_mg = recovered_mg)
stopifnot(abs(recovered_mg / typ_subject$dose_mg - 1) < 0.005)
```

The model also has to reset the release clock at each dose, since each
infusion delivers fresh nanoparticles. The release clearance should return to
`CL_B + CL_F` at the start of cycle 2 (336 h).

```{r reset}
sim_2 <- rxode2::rxSolve(
  mod_typ, make_events(typ_subject, c(0, 1, 4, 336, 337, 340), n_doses = 2L),
  returnType = "data.frame"
)
sim_2 |> dplyr::select("time", "cl_release") |> knitr::kable(digits = 4)
stopifnot(
  isTRUE(all.equal(sim_2$cl_release[sim_2$time == 336], clb + clf)),
  isTRUE(all.equal(sim_2$cl_release[sim_2$time == 340], sim_2$cl_release[sim_2$time == 4]))
)
```

# Percent free CPT

The paper reports observed mean (SEM) percent free CPT (free / total). It was
3.13 +/- 0.21% in samples taken within 3 h of the start of infusion, and it
rose to 20.42 +/- 0.44% by about 50 h after the dose. These are summaries of
observed data rather than model outputs, so they serve as a loose external
check.

```{r pct-free}
pct_free <- sim_typ |>
  dplyr::filter(.data$time > 0) |>
  dplyr::mutate(pct = 100 * .data$Cc / (.data$Cc + .data$Cc_np))
early <- mean(pct_free$pct[pct_free$time <= 3])
at50 <- pct_free$pct[pct_free$time == 50]
c(early_0_3h = early, at_50h = at50)
stopifnot(abs(early - 3.13) < 1, abs(at50 - 20.42) / 20.42 < 0.25)
```

The typical subject gives about `r round(early, 1)`% free CPT over the first
3 h and `r round(at50, 1)`% at 50 h. The observed values are 3.13% and 20.42%.
The 50 h value is about 15% lower in the model. Two things contribute to that
gap: the observed figure is a mean of per-sample ratios over a skewed
population, and at 50 h free CPT carries the largest between-subject
variability in the model (V3 CV 79.8%).

# Virtual cohort and VPC

The cohort has 100 subjects per infusion duration. Body weight is log-normal
around the cohort median of 70.4 kg and is redrawn until it falls inside the
observed 46.4-105 kg range. Sex is 55.6% female. Height is drawn by sex and is
used only to compute a Mosteller BSA for the per-m^2 dose.

```{r cohort}
n_per_arm <- 100L
draw_wt <- function(n) {
  out <- numeric(0)
  while (length(out) < n) {
    w <- exp(rnorm(n, log(70.4), 0.22))
    out <- c(out, w[w >= 46.4 & w <= 105])
  }
  out[seq_len(n)]
}
cohort <- tibble::tibble(
  id = seq_len(2L * n_per_arm),
  arm = rep(c("12 mg/m^2 over 1 h", "12 mg/m^2 over 2 h"), each = n_per_arm),
  tinf = rep(c(1, 2), each = n_per_arm),
  WT = draw_wt(2L * n_per_arm),
  female = rbinom(2L * n_per_arm, 1, 0.556),
  HT = ifelse(female == 1, rnorm(2L * n_per_arm, 162, 7), rnorm(2L * n_per_arm, 176, 7)),
  BSA = sqrt(HT * WT / 3600),
  dose_mg = 12 * BSA
)
summary(cohort[, c("WT", "BSA", "dose_mg")])
```

```{r cohort-sim}
obs_vpc <- sort(unique(c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 2.5, 3, 4, 5, 6, 8, 10, 12, 16, 24, 36, 48, 50)))
sim_cohort <- rxode2::rxSolve(
  mod, make_events(dplyr::select(cohort, "id", "arm", "WT", "dose_mg", "tinf"), obs_vpc),
  keep = "arm", returnType = "data.frame"
)
```

```{r vpc, fig.height = 6, fig.cap = "Replicates the layout of Figure 3 of Schmidt 2020: simulated 2.5th, 50th and 97.5th percentiles of conjugated (a) and free (b) CPT up to 50 h. The shading is the 95% prediction interval of individual predictions, without residual error."}
vpc_df <- sim_cohort |>
  dplyr::select("id", "arm", "time", "Conjugated CPT" = "Cc_np", "Free CPT" = "Cc") |>
  tidyr::pivot_longer(c("Conjugated CPT", "Free CPT"), names_to = "Analyte", values_to = "conc") |>
  dplyr::filter(.data$time > 0) |>
  dplyr::group_by(.data$arm, .data$Analyte, .data$time) |>
  dplyr::summarise(
    lo = quantile(.data$conc, 0.025), med = median(.data$conc),
    hi = quantile(.data$conc, 0.975), .groups = "drop"
  )
ggplot(vpc_df, aes(.data$time, .data$med, colour = .data$arm, fill = .data$arm)) +
  geom_ribbon(aes(ymin = .data$lo, ymax = .data$hi), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_y_log10() +
  facet_wrap(~Analyte, ncol = 1, scales = "free_y") +
  labs(x = "Time after start of infusion (h)", y = "Concentration (ng/mL)", colour = NULL, fill = NULL)
```

Over the whole sampling window, conjugated CPT stays well above free CPT. For
the typical subject it is about 35-fold higher at the end of a 1 h infusion,
and the gap narrows to about 5-fold by 48 h. Free CPT peaks late, well after the end of infusion.
Both features match the paper's description of Figure 3 and its conclusion that
the formulation holds on to most of the CPT during the first 48 h.

# PKNCA

The NCA below covers cycle 1 (0 to 336 h) for both analytes, grouped by
infusion-duration arm.

```{r nca}
obs_nca <- sort(unique(c(obs_vpc, 72, 96, 120, 168, 240, 336)))
sim_nca <- rxode2::rxSolve(
  mod, make_events(dplyr::select(cohort, "id", "arm", "WT", "dose_mg", "tinf"), obs_nca),
  keep = "arm", returnType = "data.frame"
)
dose_df <- cohort |>
  dplyr::transmute(id = .data$id, treatment = .data$arm, time = 0, amt = .data$dose_mg) |>
  as.data.frame()

run_nca <- function(column) {
  conc_df <- sim_nca |>
    dplyr::transmute(id = .data$id, treatment = .data$arm, time = .data$time, conc = .data[[column]]) |>
    dplyr::filter(!is.na(.data$conc)) |>
    as.data.frame()
  conc_obj <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id, concu = "ng/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")
  intervals <- data.frame(start = 0, end = 336, cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res) |>
    dplyr::filter(.data$PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
    dplyr::group_by(.data$treatment, .data$PPTESTCD) |>
    dplyr::summarise(median = median(.data$PPORRES, na.rm = TRUE), .groups = "drop") |>
    tidyr::pivot_wider(names_from = "PPTESTCD", values_from = "median")
}
nca_np <- run_nca("Cc_np") |> dplyr::mutate(Analyte = "Conjugated CPT", .before = 1)
nca_free <- run_nca("Cc") |> dplyr::mutate(Analyte = "Free CPT", .before = 1)
nca_all <- dplyr::bind_rows(nca_np, nca_free)
nca_all |>
  dplyr::rename(
    "Arm" = "treatment", "Cmax (ng/mL)" = "cmax", "Tmax (h)" = "tmax",
    "AUC0-336 (ng*h/mL)" = "auclast", "t1/2 (h)" = "half.life"
  ) |>
  knitr::kable(digits = 3, caption = "Median simulated cycle-1 NCA by analyte and infusion duration.")
```

```{r nca-gate}
free_tmax <- nca_free$tmax
np_hl <- nca_np$half.life
stopifnot(
  # Free CPT peaks well after the end of infusion (delayed Cmax; paper Results).
  all(free_tmax > 6),
  # The free-CPT terminal decline follows the release-limited conjugated
  # terminal phase (~55 h for the typical subject), not the 32.4 h disposition
  # half-life of free CPT itself.
  all(abs(nca_free$half.life / np_hl - 1) < 0.15),
  all(np_hl > 40 & np_hl < 75)
)
```

The effect of infusion duration is checked on the typical subject rather than
on the two random cohorts, so that it does not depend on the draw.

```{r tinf-typical}
auc_typ <- function(infusion_h) {
  s <- rxode2::rxSolve(
    mod_typ, make_events(dplyr::mutate(typ_subject, tinf = infusion_h), obs_fine[obs_fine <= 336]),
    returnType = "data.frame"
  )
  sum(diff(s$time) * (head(s$Cc_np, -1) + tail(s$Cc_np, -1)) / 2)
}
auc_np_typ <- c(`1 h` = auc_typ(1), `2 h` = auc_typ(2))
auc_np_typ
# A longer infusion delivers more of the dose after the fast release arm has
# decayed, so conjugated AUC is higher for the 2 h infusion.
stopifnot(auc_np_typ[["2 h"]] > auc_np_typ[["1 h"]])
```

The paper reports no NCA of its own, so there is no published table to put
beside these values, and the gates above are structural. One consequence of the
model is worth noting. Once the fast release has died away, the terminal decline
of free CPT is set by release from conjugated CPT, at about
`r round(stats::median(nca_free$half.life))` h. That is longer than the 32.4 h
terminal half-life that free CPT would show on its own (Discussion), so the
32.4 h figure is a property of the free-CPT parameters and not the slope seen in
plasma.

For the typical subject, the 2 h infusion gives a conjugated AUC0-336 about
`r round(100 * (auc_np_typ[["2 h"]] / auc_np_typ[["1 h"]] - 1), 1)`% higher than
the 1 h infusion. That is expected, because more of the dose arrives after the
fast release arm has decayed.

# Assumptions and deviations

* **Dose basis.** NLG207 doses are given as mg/m^2 of CPT equivalents, so the
  dose in the event table is mg of CPT and concentrations are ng/mL of CPT. The
  paper does not restate this. It is consistent with the assay ranges and with
  the prior NCA steady-state volume of 4.63 L at 12 mg/m^2 quoted in the
  Discussion, against V1 + V2 = 5.25 L here.
* **IIV scale.** Table 2 reports BSV as CV%. It is converted with
  `omega^2 = log(CV^2 + 1)`, the exact log-normal relation, because the paper
  states log-normal BSV (Eq. 1). Covariances come from the reported
  correlations, `cov = r * sqrt(omega1^2 * omega2^2)`.
* **Residual error.** Methods Eq. 2 is `C * (1 + eps_prop) + eps_add` with
  separate epsilons, which is encoded as `add() + prop()` for each analyte.
* **Infusion profile.** In NCT02769962 the first 10 to 15 minutes of infusion
  ran at a slower rate. The vignette uses a constant-rate infusion over 1 h or
  2 h.
* **Release clock.** The time in the release equation is the time since the
  start of the most recent infusion (`tad()`). It resets at each dose, as the
  Discussion's release half-lives imply.
* **Free-CPT parameters.** These are relative to the unknown fraction of
  conjugated CPT that is converted. The model assumes the whole release flux
  enters free-CPT central, so free-CPT volumes and clearance are apparent
  values.
* **Virtual cohort.** Height, the body-weight distribution shape and Mosteller
  BSA are assumptions, since the paper reports only the median and range of
  body weight. The VPC shows individual-prediction percentiles without residual
  error and without observed data, so it reproduces the layout of Figure 3 but
  not its content.
* **Errata.** A Europe PMC search on 2026-09-27 found no erratum or correction
  for this article.
