---
title: "Frequency-domain response analysis of QSP models (Schulthess 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Frequency-domain response analysis of QSP models (Schulthess 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.2
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Frequency-domain response analysis for quantitative systems pharmacology models (Schulthess 2017)

Schulthess, Post, Yates and van der Graaf (2018; published online 2017)
introduce **frequency-domain response analysis** (FdRA) to a pharmacometrics
audience. FdRA is an engineering method: instead of asking how a system
responds over *time*, it asks how the system transforms the *harmonic content*
of its input. A model is linearised around a stable steady state, written in
state-space form, and reduced to a **transfer function** `G(s)` whose magnitude
`|G(i * 2 * pi * f)|` is exactly the ratio of output amplitude to input
amplitude at frequency `f`. Plotting that ratio against frequency gives a Bode
plot, and the shape of the Bode plot says which dosing frequencies a given PD
system amplifies and which it filters out.

The paper applies FdRA to **14 pharmacodynamic models** drawn from four
structural classes (indirect response, autoregulation, precursor-pool, and
moderator-mediated feedback, the last in single- and double-moderator
versions), plus a **direct-effect model** in the supplement that serves as the
degenerate reference case. All 15 are shipped here as separate model files,
because each is a genuinely different system of ODEs with its own frequency
response. Every one is driven by the same one-compartment i.v. bolus PK model.

This is a **tutorial**: no data were fitted. Every parameter is an illustrative
constant printed in a figure caption, so all of them are wrapped in `fixed()`,
and none of the models carries inter-individual variability or a
residual-error model.

## Population

There is no study population. The four PD structures are generic archetypes
that the paper borrows from Gabrielsson and Hjorth (2016) to have something
concrete to analyse:

| Class | Source case study | Biological setting |
|---|---|---|
| Indirect response | case study 3 | compound acting on the rat urinary bladder sphincter via a stimulatory alpha-2 adrenergic receptor; voiding volume as biomarker |
| Autoregulation | (generic) | auto-stimulatory or auto-inhibitory turnover with drug action on the loss |
| Precursor-pool | case study 16 | antilipolytic response of healthy volunteers to an adenosine receptor agonist |
| Moderator-mediated feedback | case study 18 | nicotinic acid inhibiting production of non-esterified free fatty acids |
| Double moderator-mediated feedback | case study 19 | fold mRNA induction of a target by an anonymised test compound |

Accordingly every model file records `population$species` as
`"not applicable (theoretical illustration; no subjects, no data fitted)"`
with `n_subjects = 0` and `n_studies = 0`.

Because there are no subjects, no doses of a real drug, and no measured
concentrations, **PKNCA is not an appropriate validation tool here** and this
vignette does not use it. Following the endogenous / mechanistic validation
pattern, the model files are instead checked against (a) an exact steady-state
hold, (b) an independently derived closed-form quasi-static limit, and (c) the
published frequency-response characteristics of Table 1.

## Source trace

Every equation and every `ini()` value, with its location in the source.

| Quantity | Value | Source location |
|---|---|---|
| Drug function `E(c) = 1 +/- Emax * c / (EC50 + c)` | -- | Eq. 1 |
| Indirect response ODE `dx/dt = kin * E(c) - kout * x` | -- | Eq. 2 (stimulation of production) |
| Indirect response, loss flavors `dx/dt = kin - kout * x * E(c)` | -- | Figure 2a arrow position `E3,4`; ODE not printed (see Errata) |
| Indirect response steady state `x_SS = kin / kout` | 1 | Eq. 3 |
| One-compartment i.v. bolus PK `dc/dt = -kel * c` | -- | Eq. 5 |
| Autoregulation ODE `dx/dt = kin * phi(x) - kout * x * E(c)` | -- | Eq. 6 |
| Feedback terms `phi_1(x) = x/(K+x)`, `phi_2(x) = K/(K+x)` | -- | text following Eq. 6 |
| Autoregulation steady state, positive feedback `kin/kout - K` | 0.75 | Analytical FdRA, autoregulation |
| Autoregulation steady state, negative feedback `-K/2 + sqrt(K^2/4 + kin*K/kout)` | 0.390388 | Analytical FdRA, autoregulation |
| Precursor-pool ODEs `dx1/dt = kin - kout*x1*E(c)`, `dx2/dt = kout*x1*E(c) - kout*x2` | -- | Eq. 9 |
| Precursor-pool steady state `x1_SS = x2_SS = kin/kout` | 1 | Analytical FdRA, precursor-pool |
| Moderator ODEs `dx1/dt = (kin/x2)*E(c) - kout*x1`, `dx2/dt = ktol*(x1-x2)` | -- | Eq. 15 |
| Moderator steady state `x1_SS = x2_SS = sqrt(kin/kout)` | 1 | Analytical FdRA, moderator-mediated feedback |
| Double-moderator ODEs `dx1/dt = (kin/x3)*E(c) - kout*x1`, `dx2/dt = ktol*(x1-x2)`, `dx3/dt = ktol*(x2-x3)` | -- | Eq. 19 |
| Double-moderator steady state `sqrt(kin/kout)` | 1 | Analytical FdRA, double moderator |
| Direct-effect model `y = 1 +/- Emax*c/(EC50+c)`, `G(s) = +/- Emax/EC50` | -- | Supplementary text section A.2 |
| `kin` | 1 mL/h | Figure 2, 3, 4, 5, 6 captions |
| `kout` | 1 1/h | Figure 2, 3, 4, 5, 6 captions |
| `ktol` | 0.25 1/h | Figure 5, 6 captions |
| `K` (encoded as canonical `kd`) | 0.25 umol/L | Figure 3 caption |
| `Emax` | 1 | Figure 2, 3, 4, 5, 6 and S1 captions |
| `EC50` | 0.25 umol/L | Figure 2, 3, 4, 5, 6 and S1 captions |
| `kel` illustrative values | 1/24, 4 1/h (indirect, autoregulation, direct effect); 1/720, 1/6, 4 (precursor); 1/1440, 1/12, 4 (moderator, double moderator) | Figures 2d, 3d, 4d, 5d, 6d, S1a |
| `kel` Bode sweep range | 10^-3 to 10^1 1/h | supplementary R script (`kes <- 10^seq(-3, 1, 0.1)`) |
| Dose | 0.1 concentration units | supplementary R script (`dose <- 0.1`) |
| Dosing interval | `4 / kel` | Numerical FdRA text ("drug administration occurs at four times the inverse elimination rate") |
| Amplitude measurement | `max - min` over the tail of the simulation | supplementary R script |
| Low-frequency gain read at | `kel = 2 * pi * 10^-5` 1/h | Table 1 footnote |

## Load the models

```{r load-models}
modelIndex <- tibble::tribble(
  ~class,          ~flavor,             ~model,                                ~pd,
  "Indirect",      "Stim. production",  "Schulthess_2017_indirect_stim_kin",   "effect",
  "Indirect",      "Inhib. production", "Schulthess_2017_indirect_inhi_kin",   "effect",
  "Indirect",      "Stim. loss",        "Schulthess_2017_indirect_stim_kout",  "effect",
  "Indirect",      "Inhib. loss",       "Schulthess_2017_indirect_inhi_kout",  "effect",
  "Autoregulation", "Stim. + pos. fb",  "Schulthess_2017_autoreg_stim_posfb",  "effect",
  "Autoregulation", "Inhib. + pos. fb", "Schulthess_2017_autoreg_inhi_posfb",  "effect",
  "Autoregulation", "Stim. + neg. fb",  "Schulthess_2017_autoreg_stim_negfb",  "effect",
  "Autoregulation", "Inhib. + neg. fb", "Schulthess_2017_autoreg_inhi_negfb",  "effect",
  "Precursor-pool", "Stimulation",      "Schulthess_2017_precursor_stim",      "effect",
  "Precursor-pool", "Inhibition",       "Schulthess_2017_precursor_inhi",      "effect",
  "Moderator",     "Stimulation",       "Schulthess_2017_moderator_stim",      "effect",
  "Moderator",     "Inhibition",        "Schulthess_2017_moderator_inhi",      "effect",
  "Double moderator", "Stimulation",    "Schulthess_2017_moderator2_stim",     "effect",
  "Double moderator", "Inhibition",     "Schulthess_2017_moderator2_inhi",     "effect",
  "Direct effect",  "Stimulation",      "Schulthess_2017_directEffect_stim",   "directEffect"
)

models <- lapply(modelIndex$model, readModelDb)
names(models) <- modelIndex$model

readModelDb("Schulthess_2017_moderator_stim")
```

The published illustrative parameter set, shared by every model:

```{r pars}
pars <- c(kin = 1, kout = 1, ktol = 0.25, kd = 0.25, emax = 1, ec50 = 0.25)
dose <- 0.1
pars
```

## Check 1: the unforced steady state holds exactly

FdRA requires a *stable* steady state to linearise around, and each model file
sets its initial conditions to the steady state the paper derives. With no dose
given, every state must therefore stay exactly where it started. This is a
known-answer test of `kin`, `kout`, `ktol`, `kd` and of the algebraic form of
each ODE simultaneously: get any one of them wrong and the states drift.

```{r ss-hold}
ssExpected <- c(
  Schulthess_2017_indirect_stim_kin   = pars[["kin"]] / pars[["kout"]],
  Schulthess_2017_indirect_inhi_kin   = pars[["kin"]] / pars[["kout"]],
  Schulthess_2017_indirect_stim_kout  = pars[["kin"]] / pars[["kout"]],
  Schulthess_2017_indirect_inhi_kout  = pars[["kin"]] / pars[["kout"]],
  Schulthess_2017_autoreg_stim_posfb  = pars[["kin"]] / pars[["kout"]] - pars[["kd"]],
  Schulthess_2017_autoreg_inhi_posfb  = pars[["kin"]] / pars[["kout"]] - pars[["kd"]],
  Schulthess_2017_autoreg_stim_negfb  = -pars[["kd"]] / 2 +
    sqrt(pars[["kd"]]^2 / 4 + pars[["kin"]] * pars[["kd"]] / pars[["kout"]]),
  Schulthess_2017_autoreg_inhi_negfb  = -pars[["kd"]] / 2 +
    sqrt(pars[["kd"]]^2 / 4 + pars[["kin"]] * pars[["kd"]] / pars[["kout"]]),
  Schulthess_2017_precursor_stim      = pars[["kin"]] / pars[["kout"]],
  Schulthess_2017_precursor_inhi      = pars[["kin"]] / pars[["kout"]],
  Schulthess_2017_moderator_stim      = sqrt(pars[["kin"]] / pars[["kout"]]),
  Schulthess_2017_moderator_inhi      = sqrt(pars[["kin"]] / pars[["kout"]]),
  Schulthess_2017_moderator2_stim     = sqrt(pars[["kin"]] / pars[["kout"]]),
  Schulthess_2017_moderator2_inhi     = sqrt(pars[["kin"]] / pars[["kout"]])
)

ssCheck <- lapply(names(ssExpected), function(nm) {
  ev <- rxode2::et(seq(0, 500, by = 5), cmt = "central")
  s <- rxode2::rxSolve(models[[nm]], ev, atol = 1e-12, rtol = 1e-12,
                       returnType = "data.frame")
  tibble::tibble(
    model    = nm,
    expected = ssExpected[[nm]],
    drift    = max(abs(s$effect - ssExpected[[nm]]))
  )
}) |> bind_rows()

ssCheck |>
  rename("Model" = model, "Analytical steady state" = expected,
         "Max absolute drift over 500 h" = drift) |>
  knitr::kable(digits = 10)

stopifnot(all(ssCheck$drift < 1e-6))
```

Every state is held to better than 1e-6 over 500 h, so all 14 ODE models sit on
the steady state the paper derives.

## Analytical FdRA: the closed-form transfer functions

The transfer function is `G(s) = Jyx (s I - Jxx)^-1 Jxc + Jyc`, where `Jxx` and
`Jxc` are the Jacobians of the ODE system with respect to the states and to the
input, both evaluated at the unforced steady state, and `Jyx` selects the
output. For all 14 main-text models `Jyc = 0`, which is exactly why they all
filter the input and why (as the paper's Figure 7 notes) high-pass and band-stop
shapes cannot arise.

Working these out by hand for the published parameter set gives, with
`b = kin * Emax / EC50 = 4`:

- **Indirect response** (all four flavors): `G(s) = b / (s + kout)`. The four
  flavors differ only in the *sign* and *placement* of the drug function, which
  linearisation discards, so their magnitudes coincide.
- **Autoregulation**: `G(s) = -kout * x_SS * (Emax/EC50) / (s - J)` with
  `J = +/- kin*K/(K + x_SS)^2 - kout`, the sign being positive for positive
  feedback and negative for negative feedback.
- **Precursor-pool**: `G(s) = b * s / (s + kout)^2`. The factor of `s` in the
  numerator forces `G(0) = 0`: this is the band-pass.
- **Moderator**: `G(s) = b * (s + ktol) / ((s + kout)(s + ktol) + kout * ktol)`.
- **Double moderator**: `G(s) = b * (s + ktol)^2 / ((s + kout)(s + ktol)^2 + kout * ktol^2)`.

```{r transfer-functions}
b <- pars[["kin"]] * pars[["emax"]] / pars[["ec50"]]

xssPos <- pars[["kin"]] / pars[["kout"]] - pars[["kd"]]
xssNeg <- -pars[["kd"]] / 2 +
  sqrt(pars[["kd"]]^2 / 4 + pars[["kin"]] * pars[["kd"]] / pars[["kout"]])

autoregG <- function(xss, feedbackSign) {
  jState <- feedbackSign * pars[["kin"]] * pars[["kd"]] / (pars[["kd"]] + xss)^2 -
    pars[["kout"]]
  jInput <- -pars[["kout"]] * xss * pars[["emax"]] / pars[["ec50"]]
  function(s) jInput / (s - jState)
}

transferFun <- list(
  Indirect           = function(s) b / (s + pars[["kout"]]),
  AutoregPos         = autoregG(xssPos, +1),
  AutoregNeg         = autoregG(xssNeg, -1),
  `Precursor-pool`   = function(s) b * s / (s + pars[["kout"]])^2,
  Moderator          = function(s) {
    b * (s + pars[["ktol"]]) /
      ((s + pars[["kout"]]) * (s + pars[["ktol"]]) + pars[["kout"]] * pars[["ktol"]])
  },
  `Double moderator` = function(s) {
    b * (s + pars[["ktol"]])^2 /
      ((s + pars[["kout"]]) * (s + pars[["ktol"]])^2 +
         pars[["kout"]] * pars[["ktol"]]^2)
  },
  `Direct effect`    = function(s) rep(b, length(s))
)
```

The paper's five frequency-response characteristics, extracted from each
transfer function. Frequencies are `f = omega / (2 * pi)`, matching the
supplementary R script, which evaluates `G` at `s = i * omega` and plots against
`omega / (2 * pi)`.

```{r characteristics}
crossings <- function(fgrid, mag, level) {
  d <- mag - level
  idx <- which(d[-1] * d[-length(d)] < 0)
  if (!length(idx)) return(numeric(0))
  vapply(idx, function(i) {
    stats::approx(x = d[i:(i + 1)], y = fgrid[i:(i + 1)], xout = 0)$y
  }, numeric(1))
}

fdraChar <- function(G) {
  fgrid <- 10^seq(-6, 4, length.out = 400001)
  mag <- Mod(G(complex(imaginary = 2 * pi * fgrid)))
  g0 <- Mod(G(complex(real = 0, imaginary = 0)))
  ipk <- which.max(mag)
  cut <- if (g0 > 0) crossings(fgrid, mag, g0 / sqrt(2)) else numeric(0)
  thr <- crossings(fgrid, mag, 1)
  n <- length(fgrid)
  slope <- diff(log10(mag[c(n - 1000, n)])) / diff(log10(fgrid[c(n - 1000, n)]))
  list(
    lowFreqGain = g0,
    peakGain    = mag[ipk],
    peakFreq    = fgrid[ipk],
    cutoff      = if (length(cut)) max(cut) else NA_real_,
    threshold   = thr,
    relDegree   = -slope
  )
}

analChar <- lapply(transferFun, fdraChar)
```

### Comparison against Table 1, analytical column

```{r table1-anal}
fmt <- function(x, digits = 2) {
  if (!length(x)) return("-")
  if (all(is.na(x))) return("-")
  paste(formatC(x, format = "f", digits = digits), collapse = ", ")
}

anal <- tibble::tribble(
  ~class,             ~pubGain, ~pubCutoff, ~pubThreshold, ~pubRelDeg,
  "Indirect",         "4.00",   "0.16",     "0.62",        "1",
  "AutoregPos",       "4.00",   "0.12",     "0.46",        "1",
  "AutoregNeg",       "0.97",   "0.26",     "-",           "1",
  "Precursor-pool",   "0.00",   "-",        "0.07, 0.38",  "1",
  "Moderator",        "2.00",   "0.04",     "0.63",        "1",
  "Double moderator", "2.00",   "0.42",     "0.62",        "1"
) |>
  rowwise() |>
  mutate(
    thisGain      = fmt(analChar[[class]]$lowFreqGain),
    thisCutoff    = fmt(analChar[[class]]$cutoff),
    thisThreshold = fmt(analChar[[class]]$threshold),
    thisRelDeg    = fmt(analChar[[class]]$relDegree, 0),
    thisPeakFreq  = fmt(analChar[[class]]$peakFreq),
    thisPeakGain  = fmt(analChar[[class]]$peakGain)
  ) |>
  ungroup()

anal |>
  select(class,
         pubGain, thisGain,
         pubCutoff, thisCutoff,
         pubThreshold, thisThreshold,
         pubRelDeg, thisRelDeg) |>
  rename("Model class" = class,
         "Gain (pub.)" = pubGain, "Gain (here)" = thisGain,
         "Cutoff (pub.)" = pubCutoff, "Cutoff (here)" = thisCutoff,
         "Threshold (pub.)" = pubThreshold, "Threshold (here)" = thisThreshold,
         "Rel. degree (pub.)" = pubRelDeg, "Rel. degree (here)" = thisRelDeg) |>
  knitr::kable()
```

The **low-frequency gain reproduces exactly for every class** -- 4, 4, 0.97, 0,
2, 2 -- which is the single most informative cell in each row, because it is the
product of the steady state, the Jacobian and the drug-function slope all at
once. The **relative degree is 1 everywhere**, as published. The **threshold
frequency reproduces** for the indirect (0.62), positive-feedback
autoregulation (0.46) and moderator (0.63) classes, and correctly comes out as
*absent* for negative-feedback autoregulation, whose gain never reaches 1.

```{r table1-anal-gates}
# Exact, unambiguous cells: low-frequency gain and relative degree.
stopifnot(
  abs(analChar$Indirect$lowFreqGain           - 4.00) < 0.005,
  abs(analChar$AutoregPos$lowFreqGain         - 4.00) < 0.005,
  abs(analChar$AutoregNeg$lowFreqGain         - 0.97) < 0.005,
  abs(analChar$`Precursor-pool`$lowFreqGain   - 0.00) < 1e-9,
  abs(analChar$Moderator$lowFreqGain          - 2.00) < 0.005,
  abs(analChar$`Double moderator`$lowFreqGain - 2.00) < 0.005
)
stopifnot(all(vapply(analChar[1:6], function(z) abs(z$relDegree - 1) < 0.01, logical(1))))

# Threshold frequencies that the paper prints and that follow from G(s).
stopifnot(
  abs(analChar$Indirect$threshold  - 0.62) < 0.005,
  abs(analChar$AutoregPos$threshold - 0.46) < 0.005,
  abs(analChar$Moderator$threshold - 0.63) < 0.005,
  length(analChar$AutoregNeg$threshold) == 0
)

# Cutoff frequencies for the two classes whose printed value follows from G(s).
stopifnot(
  abs(analChar$Indirect$cutoff   - 0.16) < 0.005,
  abs(analChar$AutoregPos$cutoff - 0.12) < 0.005,
  abs(analChar$AutoregNeg$cutoff - 0.26) < 0.005,
  is.na(analChar$`Precursor-pool`$cutoff)
)

# Peak frequency for the two classes where Table 1 reports a genuine frequency.
stopifnot(
  abs(analChar$`Precursor-pool`$peakFreq - 0.16) < 0.005,
  abs(analChar$Moderator$peakFreq        - 0.10) < 0.005,
  abs(analChar$`Double moderator`$peakFreq - 0.06) < 0.005
)
```

Three groups of cells in Table 1's analytical column do **not** follow from the
transfer functions the paper itself prints. They are recorded in the Errata
section below rather than accommodated, and no parameter was adjusted to chase
them:

```{r table1-anal-mismatch}
tibble::tibble(
  Cell = c("Precursor-pool threshold frequencies",
           "Moderator cutoff frequency",
           "Peak-frequency column, low-pass rows"),
  Published = c("0.07, 0.38", "0.04", "4 / 4 / 0.97 (indirect, autoreg.)"),
  `Derived here` = c(
    fmt(analChar$`Precursor-pool`$threshold),
    fmt(analChar$Moderator$cutoff),
    paste0("peak GAIN ", fmt(analChar$Indirect$peakGain),
           "; peak frequency is 0 for a low-pass")
  )
) |>
  knitr::kable()
```

### Bode plots of the linearised models (replicates Figures 2c, 3c, 4c, 5c, 6c)

```{r bode-analytical, fig.height = 5}
fplot <- 10^seq(-4, 1.5, length.out = 400)
bodeAnal <- lapply(names(transferFun), function(cl) {
  tibble::tibble(
    class = cl,
    f = fplot,
    ratio = Mod(transferFun[[cl]](complex(imaginary = 2 * pi * fplot)))
  )
}) |> bind_rows()

ggplot(bodeAnal, aes(f, ratio, colour = class)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line(linewidth = 0.8) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Frequency (1/h)", y = "Amplitude ratio",
       colour = "Model class",
       title = "Analytical Bode plots of the linearised models",
       subtitle = "Dashed line separates amplification (above) from attenuation (below)") +
  theme_bw()
```

The indirect-response and autoregulation classes are **low-pass** filters; the
precursor-pool class is a **band-pass** (its ratio vanishes at low frequency);
the moderator classes combine the two, being flat at low frequency and peaking
at intermediate frequency; and the direct-effect model is **all-pass**, a flat
line at `Emax / EC50 = 4`, because its transfer function contains no `s`.

## Numerical FdRA: the nonlinear models driven by repeated i.v. bolus dosing

The numerical analysis drives each *nonlinear* model with the one-compartment
i.v. bolus PK of Eq. 5, dosing 0.1 concentration units every `4 / kel` hours so
that the plasma profile reaches a pseudo steady-state oscillation without
accumulating, then measures the peak-to-trough amplitude of the plasma
concentration and of the response.

```{r bode-helper}
bodePoint <- function(model, kel, pd = "effect", nInt = 3, nGrid = 250) {
  tau <- 4 / kel
  tmax <- tau * nInt
  base <- tau * (0:(nInt - 1))
  # The response relaxes on a 1/kout = 1 h timescale regardless of kel, so a
  # grid scaled only to the dosing interval would miss the response peak at
  # small kel. Combine a fine grid over the first hours after each dose with a
  # coarse sweep of the whole interval, which catches the trough.
  fine   <- as.vector(outer(seq(0, min(20, tau), length.out = nGrid), base, "+"))
  coarse <- as.vector(outer(seq(0, tau, length.out = nGrid), base, "+"))
  obs <- sort(unique(pmin(c(fine, coarse, tmax), tmax)))
  ev <- rxode2::et(amt = dose, cmt = "central", ii = tau, addl = nInt - 1) |>
    rxode2::et(obs, cmt = "central")
  s <- rxode2::rxSolve(model, ev, params = c(lkel = log(kel)),
                       atol = 1e-10, rtol = 1e-10, returnType = "data.frame")
  # Measure over the final dosing interval only, once the transient is gone.
  k <- s$time >= tau * (nInt - 1) - 1e-9
  (max(s[[pd]][k]) - min(s[[pd]][k])) / (max(s$Cc[k]) - min(s$Cc[k]))
}
```

### Check 2: the slow-dosing limit against an independent closed form

As `kel` becomes small the plasma concentration moves far more slowly than the
response relaxes, so the response tracks its own quasi-static steady state
`x_qss(c)`, obtained by solving `f(x, c) = 0` for fixed `c`. The amplitude ratio
must then converge to

`[x_qss(c_peak) - x_qss(c_trough)] / (c_peak - c_trough)`,

with `c_peak = dose / (1 - exp(-4))` and `c_trough = c_peak * exp(-4)` from the
steady-state bolus profile. This is a genuine known-answer test: the target is
derived algebraically from the printed ODEs, entirely outside `rxode2`.

The argument applies to the models whose response is a **single state driven
directly by the concentration** -- the four indirect-response flavors, the four
autoregulation flavors, and the direct-effect model. It does **not** extend to
the precursor-pool and moderator models, and the reason is worth stating because
it also explains two rows of Table 1. A bolus regimen is not a slow sinusoid: it
is a train of instantaneous *steps*, and a step rises infinitely fast however
small `kel` is. In a multi-state model the intermediate state (the precursor, or
the moderator) needs a few units of `1 / kout` to re-equilibrate after each step,
and during that transient the flux into the response is unbalanced. The response
therefore makes an excursion that no quasi-static argument predicts, and whose
size does not vanish as `kel` goes to zero. For the precursor-pool model the
quasi-static prediction is a completely flat response (`x2 = kin / kout` for
every `c`), yet the simulated response does move -- purely transiently.

```{r quasi-static}
cPeak <- dose / (1 - exp(-4))
cTrough <- cPeak * exp(-4)

# Quasi-static solutions of f(x, c) = 0, one per model structure.
eStim <- function(c) 1 + pars[["emax"]] * c / (pars[["ec50"]] + c)
eInhi <- function(c) 1 - pars[["emax"]] * c / (pars[["ec50"]] + c)

qssFun <- list(
  Schulthess_2017_indirect_stim_kin  = function(c) pars[["kin"]] / pars[["kout"]] * eStim(c),
  Schulthess_2017_indirect_inhi_kin  = function(c) pars[["kin"]] / pars[["kout"]] * eInhi(c),
  Schulthess_2017_indirect_stim_kout = function(c) pars[["kin"]] / (pars[["kout"]] * eStim(c)),
  Schulthess_2017_indirect_inhi_kout = function(c) pars[["kin"]] / (pars[["kout"]] * eInhi(c)),
  Schulthess_2017_autoreg_stim_posfb = function(c) pars[["kin"]] / (pars[["kout"]] * eStim(c)) - pars[["kd"]],
  Schulthess_2017_autoreg_inhi_posfb = function(c) pars[["kin"]] / (pars[["kout"]] * eInhi(c)) - pars[["kd"]],
  Schulthess_2017_autoreg_stim_negfb = function(c) {
    (-pars[["kd"]] + sqrt(pars[["kd"]]^2 +
       4 * pars[["kin"]] * pars[["kd"]] / (pars[["kout"]] * eStim(c)))) / 2
  },
  Schulthess_2017_autoreg_inhi_negfb = function(c) {
    (-pars[["kd"]] + sqrt(pars[["kd"]]^2 +
       4 * pars[["kin"]] * pars[["kd"]] / (pars[["kout"]] * eInhi(c)))) / 2
  },
  # Precursor-pool: kout*x1*E = kout*x2 with x1 = kin/(kout*E) gives
  # x2 = kin/kout for EVERY c, so the quasi-static amplitude ratio is zero.
  Schulthess_2017_precursor_stim     = function(c) rep(pars[["kin"]] / pars[["kout"]], length(c)),
  Schulthess_2017_precursor_inhi     = function(c) rep(pars[["kin"]] / pars[["kout"]], length(c)),
  Schulthess_2017_moderator_stim     = function(c) sqrt(pars[["kin"]] * eStim(c) / pars[["kout"]]),
  Schulthess_2017_moderator_inhi     = function(c) sqrt(pars[["kin"]] * eInhi(c) / pars[["kout"]]),
  Schulthess_2017_moderator2_stim    = function(c) sqrt(pars[["kin"]] * eStim(c) / pars[["kout"]]),
  Schulthess_2017_moderator2_inhi    = function(c) sqrt(pars[["kin"]] * eInhi(c) / pars[["kout"]]),
  Schulthess_2017_directEffect_stim  = function(c) eStim(c)
)

keSlow <- 2 * pi * 1e-5

# Models for which the quasi-static argument is valid: the response is a single
# state driven directly by Cc, with no intermediate state in between.
singleState <- !grepl("precursor|moderator", modelIndex$model)

qss <- lapply(modelIndex$model[singleState], function(nm) {
  pd <- modelIndex$pd[modelIndex$model == nm]
  tibble::tibble(
    model = nm,
    closedForm = abs(qssFun[[nm]](cPeak) - qssFun[[nm]](cTrough)) / (cPeak - cTrough),
    simulated = bodePoint(models[[nm]], keSlow, pd = pd)
  )
}) |>
  bind_rows() |>
  mutate(pctDiff = 100 * (simulated - closedForm) / closedForm)

qss |>
  rename("Model" = model, "Closed-form quasi-static ratio" = closedForm,
         "Simulated ratio" = simulated, "Difference (%)" = pctDiff) |>
  knitr::kable(digits = 4)

stopifnot(
  # Structural: each of these models must match its own closed form tightly.
  # A mis-transcribed kin, kout, kd, Emax or EC50, or a drug function on the
  # wrong term, moves these by whole percent.
  max(abs(qss$pctDiff)) < 0.5
)
```

All nine single-state models reproduce their independently derived quasi-static
limit to better than 0.5%. Note that this pins the flavors apart: stimulation of
production and inhibition of production both give 2.821, stimulation of loss
gives 2.172, and inhibition of loss gives exactly 4.000, so the check would fail
if the drug function were attached to the wrong term or with the wrong sign.

For the precursor-pool model the same algebra predicts a *flat* response, and
the transient excursion described above is what the simulation actually shows:

```{r precursor-transient}
# A single bolus, starting from the unforced steady state. At kel = 2*pi*1e-5
# the concentration is essentially constant over the 40 h shown, so any
# movement in the response is transient rather than quasi-static.
prec <- rxode2::rxSolve(
  models$Schulthess_2017_precursor_stim,
  rxode2::et(amt = dose, cmt = "central") |>
    rxode2::et(seq(0, 40, by = 0.05), cmt = "central"),
  params = c(lkel = log(keSlow)), atol = 1e-10, rtol = 1e-10,
  returnType = "data.frame"
)

tibble::tibble(
  `Quasi-static prediction (flat)` = pars[["kin"]] / pars[["kout"]],
  `Simulated response at t = 40 h` = prec$effect[nrow(prec)],
  `Peak transient excursion` = max(abs(prec$effect - pars[["kin"]] / pars[["kout"]]))
) |>
  knitr::kable(digits = 6)

stopifnot(
  # The excursion is real and transient, so by the end of a very long dosing
  # interval the response must have returned to the quasi-static value.
  abs(prec$effect[nrow(prec)] - pars[["kin"]] / pars[["kout"]]) < 0.02,
  max(abs(prec$effect - pars[["kin"]] / pars[["kout"]])) > 0.05
)
```

### Bode plots of the nonlinear models (replicates Figures 2e, 3e, 4e, 5e, 6e)

```{r bode-numerical, fig.height = 5}
keSweep <- 10^seq(-3, 1, by = 0.2)

bodeNum <- lapply(seq_len(nrow(modelIndex)), function(i) {
  tibble::tibble(
    class  = modelIndex$class[i],
    flavor = modelIndex$flavor[i],
    kel    = keSweep,
    ratio  = vapply(keSweep,
                    function(k) bodePoint(models[[modelIndex$model[i]]], k,
                                          pd = modelIndex$pd[i]),
                    numeric(1))
  )
}) |> bind_rows()

ggplot(bodeNum, aes(kel, ratio, colour = flavor)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ class, ncol = 3) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Elimination rate constant kel (1/h)", y = "Amplitude ratio",
       colour = "Flavor",
       title = "Numerical Bode plots of the nonlinear PK-driven models") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Unlike the linearised versions, the nonlinear models **do** resolve the
different drug functions: within each class the flavors separate. This is the
paper's central observation about the limits of analytical FdRA -- linearisation
discards the sign and placement of `E(c)`, so it cannot distinguish
stimulation from inhibition.

### Check 3: the direct-effect model has no frequency dependence at all

The supplement's key claim is that a direct-effect model passes every input
frequency unchanged, because the response has no state of its own. That makes a
sharp, assumption-free test: the amplitude ratio must be **constant** across the
whole four-decade `kel` sweep.

```{r direct-effect-flat}
deRatio <- bodeNum$ratio[bodeNum$class == "Direct effect"]
tibble::tibble(
  `Min ratio` = min(deRatio),
  `Max ratio` = max(deRatio),
  `Relative spread` = (max(deRatio) - min(deRatio)) / mean(deRatio)
) |>
  knitr::kable(digits = 8)

stopifnot((max(deRatio) - min(deRatio)) / mean(deRatio) < 1e-6)
```

Flat to one part in a million over four decades, exactly as the supplement
predicts.

### Comparison against Table 1, numerical column

Table 1's numerical low-frequency gains were measured with a specific protocol,
spelled out in the supplementary R script: 250 doses, an observation grid of
exactly `4 / kel / 100` (one hundred points per dosing interval), amplitudes
taken as `max - min` over the last quarter of the record, and `kel` set to
`2 * pi * 10^-5` 1/h. Reproducing that protocol exactly -- rather than the
refined grid used for the Bode plots above -- is the direct test of the shipped
models against the published column.

```{r paper-protocol}
paperProtocol <- function(model, kel, pd = "effect", nDose = 250) {
  tau <- 4 / kel
  obs <- seq(0, tau * nDose, by = tau / 100)
  ev <- rxode2::et(amt = dose, cmt = "central", ii = tau, addl = nDose - 1) |>
    rxode2::et(obs, cmt = "central")
  s <- rxode2::rxSolve(model, ev, params = c(lkel = log(kel)),
                       atol = 1e-10, rtol = 1e-10, returnType = "data.frame")
  n <- nrow(s)
  q <- floor(3 * n / 4):n
  (max(s[[pd]][q]) - min(s[[pd]][q])) / (max(s$Cc[q]) - min(s$Cc[q]))
}
```

```{r table1-num}
pubNum <- tibble::tribble(
  ~class,             ~flavor,             ~pubGain,
  "Indirect",         "Stim. production",  2.86,
  "Indirect",         "Inhib. production", 2.86,
  "Indirect",         "Stim. loss",        2.22,
  "Indirect",         "Inhib. loss",       3.99,
  "Autoregulation",   "Stim. + pos. fb",   2.21,
  "Autoregulation",   "Inhib. + pos. fb",  4.00,
  "Autoregulation",   "Stim. + neg. fb",   0.57,
  "Autoregulation",   "Inhib. + neg. fb",  0.89,
  "Precursor-pool",   "Stimulation",       0.00,
  "Precursor-pool",   "Inhibition",        0.00,
  "Moderator",        "Stimulation",       1.34,
  "Moderator",        "Inhibition",        1.55,
  "Double moderator", "Stimulation",       1.34,
  "Double moderator", "Inhibition",        1.55
)

numLF <- pubNum |>
  left_join(modelIndex, by = c("class", "flavor")) |>
  rowwise() |>
  mutate(
    paperGrid   = paperProtocol(models[[model]], keSlow, pd = pd),
    refinedGrid = bodePoint(models[[model]], keSlow, pd = pd)
  ) |>
  ungroup() |>
  mutate(pctDiff = 100 * (paperGrid - pubGain) /
           ifelse(pubGain == 0, 1, pubGain))

numLF |>
  select(class, flavor, pubGain, paperGrid, pctDiff, refinedGrid) |>
  rename("Model class" = class, "Flavor" = flavor,
         "Published" = pubGain,
         "Paper protocol" = paperGrid,
         "Difference (%)" = pctDiff,
         "Refined grid" = refinedGrid) |>
  knitr::kable(digits = 4)
```

```{r table1-num-gates}
stopifnot(
  # All twelve non-zero published gains reproduce under the paper's own
  # protocol to well under 1%.
  max(abs(numLF$pctDiff[numLF$pubGain > 0])) < 1,
  # The two precursor-pool models are published as exactly 0 and must come out
  # as zero to numerical precision under the same protocol.
  all(numLF$paperGrid[numLF$pubGain == 0] < 1e-3)
)
```

**All fourteen published numerical low-frequency gains reproduce to better than
1%** under the paper's own protocol, including the two precursor-pool zeros.
That is a complete validation of the numerical column of Table 1 against the
shipped models.

The `Refined grid` column shows why the protocol has to be reproduced exactly
rather than improved upon. The published amplitude ratio is **grid-sensitive**,
because the response relaxes on a `1 / kout = 1 h` timescale no matter how small
`kel` is, whereas a grid of one hundred points per dosing interval has a spacing
of `4 / kel / 100`, which at `kel = 2 * pi * 10^-5` is 637 h. That grid
under-resolves the post-dose peak of the *concentration* and of the *response*
alike, and because the two errors sit in the numerator and the denominator of
the same ratio they largely cancel: for stimulation of production the coarse
grid gives 2.854 against a published 2.86, while resolving both peaks properly
gives 2.820. The precursor-pool rows are the extreme case -- the coarse grid
misses the transient excursion described in Check 2 entirely and so returns the
published 0, while a refined grid resolves it and returns roughly 0.9.

### Time courses (replicates Figures 2d, 5d)

```{r time-courses, fig.height = 5}
timeCourse <- function(nm, kel, pd = "effect", nInt = 4, nGrid = 400) {
  tau <- 4 / kel
  obs <- seq(0, tau * nInt, length.out = nGrid * nInt)
  ev <- rxode2::et(amt = dose, cmt = "central", ii = tau, addl = nInt - 1) |>
    rxode2::et(obs, cmt = "central")
  rxode2::rxSolve(models[[nm]], ev, params = c(lkel = log(kel)),
                  atol = 1e-10, rtol = 1e-10, returnType = "data.frame") |>
    mutate(model = nm, kel = kel, response = .data[[pd]]) |>
    select(model, kel, time, Cc, response)
}

tc <- bind_rows(
  timeCourse("Schulthess_2017_indirect_stim_kin", 1 / 24),
  timeCourse("Schulthess_2017_indirect_stim_kin", 4),
  timeCourse("Schulthess_2017_moderator_stim",    1 / 12),
  timeCourse("Schulthess_2017_moderator_stim",    4)
) |>
  mutate(panel = paste0(sub("Schulthess_2017_", "", model),
                        "\nkel = ", signif(kel, 3), " 1/h"))

tc |>
  select(panel, time, `Plasma concentration` = Cc, Response = response) |>
  pivot_longer(c(`Plasma concentration`, Response),
               names_to = "series", values_to = "value") |>
  ggplot(aes(time, value, colour = series)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~ panel, scales = "free", ncol = 2) +
  labs(x = "Time (h)", y = "Concentration (umol/L) or response",
       colour = NULL,
       title = "Plasma concentration and response at a slow and a fast elimination rate") +
  theme_bw() +
  theme(legend.position = "bottom")
```

At the slow elimination rate the response swings by more than the plasma
concentration does (amplification); at the fast rate the response is nearly
flat while the concentration oscillates (attenuation). That contrast is the
qualitative content of the low-pass Bode plots above.

## Assumptions and deviations

- **Implied central volume.** The paper integrates the plasma *concentration*
  directly (`dc/dt = -kel * c`, Eq. 5) and its supplementary R script adds
  `dose = 0.1` to that same state, so no volume of distribution ever appears.
  Each model file therefore carries `lvc <- fixed(log(1))` and computes
  `Cc <- central / vc`, which reproduces the source exactly while keeping the
  dimensional structure visible and letting a user re-scale if they wish.
- **Initial conditions set to the analytical steady state.** The supplementary
  R script starts the response states at 0 and burns off the transient with 250
  doses. Because the paper derives the unforced steady state in closed form for
  every model, the model files start there instead. Check 1 confirms the states
  are stationary, and the prototype confirmed the amplitude ratio is unchanged
  (identical to five significant figures at 3, 5 and 20 dosing intervals).
- **`kel` is a swept quantity, not an estimate.** Each file fixes `kel` at one
  of the illustrative values printed in that model's own time-course figure
  (the amplifying case), and this vignette overrides it across
  `10^-3` to `10^1` 1/h to build the Bode plots.
- **Loss-flavor indirect-response ODEs are not printed.** The paper prints only
  Eq. 2 (stimulation of production) and states that the other three flavors
  "are described by different differential equations" without giving them. The
  two loss flavors are encoded as `dx/dt = kin - kout * x * E(c)`, which is
  where Figure 2a places the `E3,4` arrow and is the same construction the
  authors use explicitly in Eq. 6 for the autoregulation models. The analytical
  low-frequency gain of 4 that Table 1 reports for all four flavors follows from
  this reading and is confirmed above.
- **Output variable of each multi-state model.** For the precursor-pool model
  the output is `x2` (the paper sets `Jyx = [0 1]`); for the moderator models it
  is `x1`, the response, with the moderators unobserved. Both readings are
  confirmed by the low-frequency gains of 0 and 2 respectively.
- **Units of the autoregulation and moderator response.** See Errata.
- **No PKNCA.** These are data-free theoretical PD models with no drug, no
  subjects and no measured concentrations, so NCA is not a meaningful check. The
  three checks above (exact steady-state hold, closed-form quasi-static limit,
  flat direct-effect response) plus the Table 1 comparison serve that role.
- **No IIV and no residual error.** The source fits nothing, so inventing
  variance terms would be fabrication. All parameters are `fixed()`.

## Errata

Discrepancies found in the source while extracting. None was accommodated by
changing a parameter value.

1. **Table 1's "Peak frequency" column mixes gains and frequencies.** For the
   indirect-response and autoregulation rows it repeats the low-frequency gain
   (4, 4, 0.97, ...) rather than a frequency, which the main text makes explicit
   for Figure 2c: "the peak frequency is equal to the low-frequency gain." For a
   monotonically decreasing low-pass response the peak *frequency* is 0 and the
   peak *gain* equals the low-frequency gain; the column reports the latter. For
   the precursor-pool and moderator rows the same column does contain genuine
   frequencies (0.16, 0.1, 0.06), all three of which reproduce here.

2. **The precursor-pool threshold frequencies are not reproducible from the
   printed transfer function.** Table 1 gives 0.07 and 0.38 1/h. From
   `G(s) = 4s/(s + kout)^2` the amplitude ratio reaches 1 at
   `omega = 2 -/+ sqrt(3)`, i.e. `f = 0.043` and `0.594` 1/h. Both pairs are
   geometrically centred on the peak at 0.16 1/h, so the published pair is
   self-consistent but narrower than the transfer function allows.

3. **The moderator cutoff frequency of 0.04 1/h appears to be a typographical
   error.** The moderator response *rises* from a low-frequency gain of 2 to a
   peak at 0.1 1/h, so it cannot cross `gain / sqrt(2)` below the peak; the only
   crossing is at 0.435 1/h. Note that the double-moderator row of the same
   column reports 0.42 1/h, which is close to the single-moderator value derived
   here, and that the other three moderator characteristics (gain 2, peak
   0.1, threshold 0.63) all reproduce exactly.

4. **The four indirect-response flavors are numbered inconsistently.** The main
   text has "2 = inhibition of production, 3 = stimulation of loss", while the
   Figure 2 caption has "2 = stimulation of loss, 3 = inhibition of production".
   Both describe the same set of four models. The model files follow the main
   text and the `E1,2` / `E3,4` arrow labels of Figure 2a, and name the
   mechanism explicitly in `description` so the numbering is not load-bearing.

5. **The autoregulation and moderator captions are dimensionally inconsistent.**
   Every caption gives `kin = 1 mL/h` while `K = 0.25 umol/L`, but the feedback
   term `K / (K + x)` requires `K` and the response `x` to carry the same units,
   and `x = kin / kout` would be in mL. The same applies to the moderator
   models, where `kin / x2` must have the units of `kout * x1`. Since every
   published value is 1 or 0.25 and the arithmetic is unit-agnostic as
   implemented, no numeric result depends on the resolution; the affected model
   files record the caveat in `population$notes`.

6. **Table 1's numerical column is grid-dependent, and reproducing it requires
   reproducing the measurement protocol rather than improving on it.** All
   fourteen values reproduce to better than 1% under the supplementary script's
   own settings (250 doses, one hundred observations per dosing interval,
   amplitudes over the last quarter). They are not, however, converged
   quantities: the response relaxes on a `1 / kout = 1 h` timescale regardless
   of `kel`, so at the `kel = 2 * pi * 10^-5` used for the low-frequency gain
   the grid spacing is 637 h and under-resolves the post-dose peak of the
   concentration and of the response alike. Because those two errors occupy the
   numerator and denominator of the same ratio they mostly cancel, which is why
   the coarse grid lands close to the published value (2.854 versus 2.86 for
   stimulation of production) while a properly resolved grid gives 2.820. The
   precursor-pool rows are the clearest case: the published numerical
   low-frequency gain of 0 is recovered only because the coarse grid steps over
   the transient excursion entirely; resolve it and the ratio is about 0.9,
   even though the *linearised* gain is exactly 0. This vignette therefore gates
   the published comparison on the paper's protocol and gates the shipped
   models, independently and more tightly, against closed-form targets
   (Checks 1 and 2).

7. **The models are archetypes, not the paper's own fits.** The four PD classes
   are taken from case studies 3, 16, 18 and 19 of Gabrielsson and Hjorth
   (2016); the parameter values are illustrative constants chosen to make the
   frequency responses easy to read, not estimates. For a *fitted* version of
   the moderator-mediated feedback case study, see
   `Ahlstrom_2010_nicotinicAcid_rat`, which carries real nicotinic-acid/NEFA
   parameters. The library also ships generic templates of the
   indirect-response and precursor-pool structures
   (`indirect_1cpt_stim_kin` and relatives), which differ from these files in
   using an oral depot with a volume term and a proportional residual error.

## Reference

Schulthess P, Post TM, Yates J, van der Graaf PH. Frequency-domain response
analysis for quantitative systems pharmacology models. *CPT Pharmacometrics
Syst Pharmacol.* 2018;7(2):111-123.
[doi:10.1002/psp4.12266](https://doi.org/10.1002/psp4.12266)

Gabrielsson J, Hjorth S. Pattern recognition in pharmacodynamic data analysis.
*AAPS J.* 2016;18(1):64-91.
