---
title: "Rifabutin (Semere Gebreyesus 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rifabutin (Semere Gebreyesus 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("SemereGebreyesus_2024_rifabutin"))$reference`
* Description: `r rxode2::rxode(readModelDb("SemereGebreyesus_2024_rifabutin"))$description`
* Article: <https://doi.org/10.1128/aac.00354-24>
* Supplement (open access): `AAC00354-24-s0001.docx` -- supplemental text on the
  B2 initialization method plus Figures S1-S9, including the Figure S4 model
  schematic used to confirm the compartment topology below.

Rifabutin is preferred over rifampicin when a protease-inhibitor-based
antiretroviral regimen must be maintained, because rifampicin drops lopinavir
exposure by more than 75% while rifabutin does not. The trade-off is that
lopinavir/ritonavir (LPV/r) inhibits CYP3A4, which raises exposure to both
rifabutin and its 25-O-desacetyl metabolite (des-rifabutin) and creates a
neutropenia risk. This paper is the paediatric dosing evidence for that
interaction.

# Population

Twenty-eight children with HIV/tuberculosis co-infection were enrolled in the
APIN PEPFAR paediatric ART programme in Nigeria across three prospective age
cohorts: under 1 year (n = 3), 1-3 years (n = 10) and 3-15 years (n = 15).
Median age was 10 years (range 0.67-15.0) and median weight 11 kg (range
4.5-45.0). The cohort was severely malnourished: median weight-for-age z-score
was -3.33 (range -5.15 to -1.32) and about 60% of the children were below the
WHO severe-underweight cut-off of -3 (Table 1 and Results "Participant
characteristics").

The two younger cohorts began with two weeks of rifabutin-containing TB-only
treatment (20 mg/kg/day under 1 year; 15-20 mg/kg/day at 1-3 years) and then
started LPV/r-based ART, at which point the rifabutin dose was cut to
5 mg/kg/day and 2.5 mg/kg/day respectively. The 3-15 year cohort was
ART-experienced and received 2.5 mg/kg/day with LPV/r from study entry. Because
the younger children crossed over, the same child contributes both
`CONMED_LPV = 0` and `CONMED_LPV = 1` occasions -- this within-subject crossover
is what makes the inhibitable and non-inhibitable rifabutin clearance pathways
separately identifiable. Rifabutin was given as a 20 mg/mL oral suspension
compounded from Mycobutin capsules. An external validation set of six South
African children (Moultrie et al.) was subsequently pooled in and the parameters
re-estimated, so the Table 2 estimates encoded here describe the combined data.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("SemereGebreyesus_2024_rifabutin")()$population`).

# Model structure

The final joint model (Figure S4 of the supplement) is two-compartment
disposition for **both** parent and metabolite, with first-order absorption and
an absorption lag time. Rifabutin leaves the central compartment by two parallel
arms:

* an **inhibitable CYP3A4 pathway** (`lcl`), which LPV/r switches off entirely
  (the `-100%` effect is fixed, not estimated); and
* a **clearance conversion** via arylacetamide deacetylase (`lcl_form_desacetylrbn`),
  which is the sole source of des-rifabutin and is *not* inhibited by LPV/r.

Splitting the two would normally be unidentifiable; it is only possible here
because LPV/r inhibits one arm and not the other. Conversion is assumed to be
1:1 in **molar** terms ("100% of the rifabutin eliminated by the clearance
conversion is transformed into des-rifabutin"), so because the ODE states in
this implementation hold mass, the transfer carries the molecular-weight ratio
805 / 847.02.

```{r structure}
mod <- readModelDb("SemereGebreyesus_2024_rifabutin")
ui  <- rxode2::rxode(mod)
ui$props$cmt
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/SemereGebreyesus_2024_rifabutin.R` carries an
in-file comment naming its source location. They are collected here for review.
All structural and covariate values come from Table 2, "Parameter estimates for
the rifabutin-des-rifabutin model".

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | 13.6 L/h per 10 kg | Table 2, "Clearance (Inhibitable CYP3A4 pathway) (L/h)" |
| `lcl_form_desacetylrbn` | 16.2 L/h per 10 kg | Table 2, "Clearance conversion (AADAC pathway) (L/h)" |
| `lvc` | 185 L per 10 kg | Table 2, "Central volume of distribution, V C,P (L)" |
| `lvp` | 232 L per 10 kg | Table 2, "Peripheral volume of distribution, V P,P (L)" |
| `lq` | 25.1 L/h per 10 kg | Table 2, "Intercompartmental clearance, Q P (L/h)" |
| `lka` | 1.27 1/h | Table 2, "Absorption rate constant, Ka (1/h)" |
| `ltlag` | 0.544 h | Table 2, "Absorption lag time, Lag (h)" |
| `lfdepot` | 1 (fixed) | Table 2, "Bioavailability, F" = "1 fixed" |
| `lcl_desacetylrbn` | 106 L/h per 10 kg | Table 2, "Clearance metabolite (L/h)" |
| `lvc_desacetylrbn` | 43.0 L per 10 kg | Table 2, "Central volume of distribution, V C,M (L)" |
| `lvp_desacetylrbn` | 241 L per 10 kg | Table 2, "Peripheral volume of distribution, V P,M (L)" |
| `lq_desacetylrbn` | 44.4 L/h per 10 kg | Table 2, "Intercompartmental clearance, Q M (L/h)" |
| `e_conmed_lpv_fdepot` | +1.58 (2.58-fold) | Table 2, "LPV/r effect on bioavailability (%)" = +158; Results "Covariate model" states 2.58-fold (1.93-3.46) |
| `e_conmed_lpv_cl` | -1 (fixed) | Table 2, "LPV/r effect on rifabutin clearance (inhibitable CYP3A4 pathway) (%)" = "-100 fixed" |
| `e_conmed_lpv_cl_desacetylrbn` | -0.766 | Table 2, "LPV/r effect on des-rifabutin clearance (%)" = -76.6 |
| `e_waz_fdepot` | -0.26 per unit below -3 | Table 2, "ZWFA effect (each point below -3) on bioavailability (%)" = -26.0, footnote d |
| `e_age_le3_ka` | -0.723 | Table 2, "Age effect (<=3 years old) on Ka (%)" = -72.3 |
| `etalcl` | omega^2 = 0.041956 | Table 2 rifabutin block, "BSV: 20.7" %CV; footnote b gives CV = sqrt(exp(omega^2) - 1) |
| `etalcl_desacetylrbn` | omega^2 = 0.090630 | Table 2 des-rifabutin block, "BSV: 30.8" %CV |
| `etaiov_fdepot_*` | omega^2 = 0.416709 | Table 2, "Bioavailability, F" -> "BOV: 71.9" %CV |
| `etaiov_ka_*` | omega^2 = 0.872297 | Table 2, "Absorption rate constant, Ka" -> "BOV: 118" %CV |
| `etaiov_tlag_*` | omega^2 = 0.444368 | Table 2, "Absorption lag time, Lag" -> "BOV: 74.8" %CV |
| `propSd` / `addSd` | 0.188 / 10.1 ug/L | Table 2, "Rifabutin: proportional error (%)" / "additive error (ug/L)" |
| `propSd_desacetylrbn` / `addSd_desacetylrbn` | 0.108 / 11.6 ug/L | Table 2, "Des-rifabutin: proportional / additive error" |
| Allometric exponents 0.75 / 1.0, reference 10 kg | n/a | Methods "Population pharmacokinetic analysis"; Results "Structural model"; Table 2 footnote c |
| Molar conversion 805 / 847.02 | n/a | Methods "Population pharmacokinetic analysis" (molecular weights); Results "Structural model" (100% conversion) |
| Two-compartment parent + two-compartment metabolite topology | n/a | Results "Structural model"; supplement Figure S4 schematic |

# Validation 1 -- the paper's own allometric extrapolation

The Discussion states: "Our inhibitable rifabutin clearance extrapolated to a
70-kg adult of 58.5 L/h corresponds closely to their reported value of 58.8 L/h"
(comparing against Hennig et al., which is also packaged, as
`Hennig_2015_rifabutin`). That is a deterministic identity on the allometric
term and a direct check that the 10 kg reference weight and the 0.75 exponent
were transcribed correctly.

```{r allometry-check}
cl_inh_10kg <- 13.6
cl_inh_70kg <- cl_inh_10kg * (70 / 10)^0.75

c(
  extrapolated_70kg = round(cl_inh_70kg, 2),
  paper_states      = 58.5
)

# Deterministic: this is closed-form arithmetic on published constants, so a
# tight bound is correct here (see pattern 11 of
# known-vignette-failure-patterns.md -- tighten deterministic gates, loosen
# cohort-derived ones).
stopifnot(abs(cl_inh_70kg - 58.5) < 0.1)
```

# Validation 2 -- reproducing the paper's optimized weight-band doses

Table 3 gives the doses the authors derived from Monte Carlo simulation to match
adult exposures across the WHO harmonized weight bands. Because steady-state
AUC for a linear model is `F * Dose / CL`, those doses can be recomputed in
closed form from the Table 2 parameters alone. Reproducing all fourteen of them
simultaneously is a strong test that the *whole* parent-side encoding is right:
that the two clearance arms are parallel and additive, that LPV/r zeroes the
inhibitable arm and multiplies bioavailability by 2.58, that LPV/r leaves the
conversion arm alone, and that allometry uses exponent 0.75 at a 10 kg
reference.

```{r table3}
bands <- tibble::tribble(
  ~band,        ~wt_mid, ~dose_no_lpv, ~dose_lpv,
  ">=6 to <10",     8.0,          115,        25,
  ">=10 to <15",   12.5,          160,        35,
  ">=15 to <20",   17.5,          200,        45,
  ">=20 to <25",   22.5,          250,        55,
  ">=25 to <30",   27.5,          300,        65,
  ">=30 to <35",   32.5,          300,        70,
  ">=35 to <45",   40.0,          300,        85
)

cl_inh   <- 13.6   # Table 2
cl_conv  <- 16.2   # Table 2
f_lpv    <- 2.58   # Table 2 / Results: +158% on bioavailability

# Adult median steady-state AUC0-24 targets quoted in "Targets for simulations".
# Without LPV/r the paper optimizes toward the >= 4,500 ug*h/L limit associated
# with acquired rifamycin resistance; with LPV/r the adult comparator range
# starts at 4,770 ug*h/L (Naiker et al.).
auc_target_no_lpv <- 4500
auc_target_lpv    <- 4770

table3 <- bands |>
  mutate(
    wt_cl = (wt_mid / 10)^0.75,
    # No LPV/r: both parent arms active, F = 1, dose capped at the 300 mg adult dose.
    pred_no_lpv = pmin(auc_target_no_lpv * (cl_inh + cl_conv) * wt_cl / 1000, 300),
    # With LPV/r: inhibitable arm fully off, F x 2.58.
    pred_lpv    = auc_target_lpv * cl_conv * wt_cl / (f_lpv * 1000),
    err_no_lpv  = 100 * (pred_no_lpv / dose_no_lpv - 1),
    err_lpv     = 100 * (pred_lpv / dose_lpv - 1)
  )

table3 |>
  transmute(
    `Weight band (kg)`           = band,
    `Table 3, no LPV/r (mg)`     = dose_no_lpv,
    `Closed form (mg)`           = round(pred_no_lpv),
    `Table 3, with LPV/r (mg)`   = dose_lpv,
    `Closed form, LPV/r (mg)`    = round(pred_lpv, 1)
  ) |>
  knitr::kable(
    caption = paste(
      "Table 3 of Semere Gebreyesus 2024 recomputed in closed form from the",
      "Table 2 parameter estimates. Both columns reproduce across all seven",
      "weight bands."
    )
  )

# Deterministic (no simulation, no RNG): closed-form arithmetic against the
# paper's own printed doses. The residual error is the paper's rounding of the
# optimized doses to a practical 5 mg grid, so a 6% bound is appropriate and
# still goes red on a mis-transcribed clearance, exponent or reference weight
# (any of which moves these by tens of percent).
stopifnot(
  max(abs(table3$err_no_lpv)) < 6,
  max(abs(table3$err_lpv))    < 6
)
```

# Virtual cohort

Original observed data are not publicly available. The simulations below use a
virtual cohort in the 10-15 kg weight band -- the band for which the paper
quotes explicit percentage exposure increases -- under the three dosing
scenarios the paper simulated. Weight-for-age z-scores are drawn above the
severe-underweight cut-off so that the `WAZ` bioavailability effect is inert;
this matches the "representative in silico population of African children" the
authors simulated (their optimized Table 3 doses are reproduced above assuming
no `WAZ` penalty at the median), rather than this study's own unusually
malnourished cohort.

```{r cohort}
# `set.seed()` seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD, so the drawn cohort
# differs between a 2-core CI runner and a 16-thread workstation. Every
# assertion below is written to hold for ANY cohort this model can produce
# (pattern 12 of references/known-vignette-failure-patterns.md).
set.seed(20240722)

n_per_arm <- 150L
tau       <- 24     # dosing interval (h)
n_doses   <- 14L    # the paper notes steady state is reached within 10 days
ss_start  <- (n_doses - 1L) * tau   # start of the final, steady-state interval

make_arm <- function(label, lpv, mg_per_kg, id_offset) {
  subj <- tibble(
    id  = id_offset + seq_len(n_per_arm),
    WT  = runif(n_per_arm, 10, 15),
    # Held above -3 so the severe-underweight bioavailability penalty is off.
    WAZ = runif(n_per_arm, -2.5, -1.0),
    # 10-15 kg spans roughly 1-4 years; AGE only enters via the <= 3 y
    # absorption-rate switch.
    AGE = runif(n_per_arm, 1, 4),
    CONMED_LPV = lpv,
    OCC = 1L,
    arm = label
  ) |>
    mutate(
      # The paper caps the TB-only dose at the 300 mg adult dose.
      amt_dose = pmin(mg_per_kg * WT, 300)
    )

  doses <- subj |>
    tidyr::crossing(time = seq(0, by = tau, length.out = n_doses)) |>
    mutate(evid = 1L, cmt = "depot", amt = amt_dose, dvid = NA_integer_)

  # This model declares TWO endpoints (`Cc` and `Cc_desacetylrbn`), so rxode2
  # requires each observation record to say which endpoint it belongs to.
  # Pointing `cmt` at an ODE state such as "central" is rejected here -- that
  # advice applies to single-output models. `dvid` is the semantic tool for
  # multi-output models: `dvid = 1L` selects the first endpoint (`Cc`).
  # rxode2 returns BOTH observables as columns at these rows, so one set of
  # observation rows serves both analytes. Verified byte-identical to the
  # alternative encoding that names the observable as the compartment, and the
  # endpoint slots (6 and 7) sit after all five ODE states, so nothing is
  # renumbered.
  obs <- subj |>
    tidyr::crossing(
      time = ss_start + c(seq(0, 12, by = 0.25), seq(12.5, 24, by = 0.5))
    ) |>
    mutate(evid = 0L, cmt = NA_character_, amt = NA_real_, dvid = 1L)

  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, evid, cmt, dvid, amt, WT, WAZ, AGE, CONMED_LPV, OCC, arm)
}

events <- bind_rows(
  make_arm("TB only, 20 mg/kg/day",   0, 20.0, id_offset =   0L),
  make_arm("LPV/r, 2.5 mg/kg/day",    1,  2.5, id_offset = 1000L),
  make_arm("LPV/r, 5 mg/kg/day",      1,  5.0, id_offset = 2000L)
)

# Disjoint IDs across arms are mandatory: rxSolve treats `id` as the subject
# key and silently merges duplicates into one subject receiving the summed dose.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("arm", "WT", "CONMED_LPV"),
  # This model has two algebraic observables backed by five ODE states, which
  # is the shape rxode2's automatic ODE->linCmt conversion mis-maps
  # (pattern 5b of known-vignette-failure-patterns.md).
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  filter(!is.na(Cc))

# Shift the final dosing interval onto a 0-24 h clock for NCA.
sim_ss <- sim |>
  filter(time >= ss_start) |>
  mutate(tad = time - ss_start)

nrow(sim_ss)
```

# Validation 3 -- internal identities under typical values

With the random effects zeroed, steady-state exposure must satisfy the linear
closed forms exactly. These are pure numerical-accuracy checks (the same
parameters drive both sides), so they carry tight bounds.

* Parent: `AUC_tau = F * Dose / CL_total`, where `CL_total` is
  `cl + cl_form_desacetylrbn` without LPV/r and `cl_form_desacetylrbn` alone
  with LPV/r.
* Metabolite : parent AUC ratio: `(cl_form / cl_metabolite) * (805 / 847.02)`,
  which is independent of dose and of weight (both arms scale with the same
  allometric factor).

```{r identities}
mod_typ <- mod |> rxode2::zeroRe()

typical_events <- function(lpv, mg_per_kg, wt = 12.5) {
  subj <- tibble(
    id = 1L, WT = wt, WAZ = -1, AGE = 5, CONMED_LPV = lpv, OCC = 1L
  )
  doses <- subj |>
    tidyr::crossing(time = seq(0, by = tau, length.out = n_doses)) |>
    mutate(evid = 1L, cmt = "depot", amt = pmin(mg_per_kg * wt, 300),
           dvid = NA_integer_)
  obs <- subj |>
    tidyr::crossing(time = ss_start + seq(0, 24, by = 0.05)) |>
    mutate(evid = 0L, cmt = NA_character_, amt = NA_real_, dvid = 1L)
  bind_rows(doses, obs) |> arrange(time, desc(evid))
}

auc_trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

check_identity <- function(lpv, mg_per_kg, wt = 12.5) {
  ev <- typical_events(lpv, mg_per_kg, wt)
  s  <- rxode2::rxSolve(mod_typ, ev, useLinCmt = FALSE) |>
    as.data.frame() |>
    filter(!is.na(Cc), time >= ss_start)

  allo    <- (wt / 10)^0.75
  cl_tot  <- if (lpv == 1) cl_conv * allo else (cl_inh + cl_conv) * allo
  fbio    <- if (lpv == 1) f_lpv else 1
  dose    <- min(mg_per_kg * wt, 300)
  # LPV/r also cuts des-rifabutin clearance by 76.6%, which raises the
  # metabolite:parent ratio. Allometry cancels in the ratio because both arms
  # scale with the same factor.
  cl_met  <- 106 * (if (lpv == 1) 1 - 0.766 else 1)

  auc_p <- auc_trap(s$time, s$Cc)
  auc_m <- auc_trap(s$time, s$Cc_desacetylrbn)

  tibble(
    scenario        = if (lpv == 1) "with LPV/r" else "TB only",
    dose_mg         = dose,
    auc_parent_sim  = auc_p,
    auc_parent_pred = fbio * dose * 1000 / cl_tot,
    ratio_sim       = auc_m / auc_p,
    ratio_pred      = (cl_conv / cl_met) * (805 / 847.02)
  )
}

ident <- bind_rows(
  check_identity(0, 20.0),
  check_identity(1,  2.5),
  check_identity(1,  5.0)
) |>
  mutate(
    parent_err_pct = 100 * (auc_parent_sim / auc_parent_pred - 1),
    ratio_err_pct  = 100 * (ratio_sim / ratio_pred - 1)
  )

ident |>
  transmute(
    Scenario                       = scenario,
    `Dose (mg)`                    = round(dose_mg, 1),
    `AUC parent, simulated`        = round(auc_parent_sim),
    `AUC parent, F*Dose/CL`        = round(auc_parent_pred),
    `Error (%)`                    = round(parent_err_pct, 2),
    `Metabolite:parent ratio, sim` = round(ratio_sim, 4),
    `... predicted`                = round(ratio_pred, 4)
  ) |>
  knitr::kable(
    caption = paste(
      "Typical-value steady-state exposure against the linear closed forms.",
      "Residuals are trapezoidal-integration error on a 0.05 h grid."
    )
  )

# Deterministic (zeroRe, no RNG): both sides use the same drawn parameters, so
# the only discrepancy is numerical. Tight bounds are correct here.
stopifnot(
  max(abs(ident$parent_err_pct)) < 1,
  max(abs(ident$ratio_err_pct))  < 1
)
```

# Validation 4 -- the LPV/r drug-drug interaction

The paper's headline interaction result (Results, "Effect of LPV/r co-treatment
on rifabutin and des-rifabutin exposures") compares model-derived exposures in
the under-3-year children who received both treatments. Because those children
also had their dose cut when LPV/r started, the reported geometric mean ratios
mix the interaction with the dose change. Holding the dose fixed isolates the
interaction itself, which is what the model encodes:

* rifabutin AUC should rise by `2.58 * (cl_inh + cl_conv) / cl_conv` = 4.75-fold;
* des-rifabutin AUC should rise by a further `1 / (1 - 0.766)` = 4.27-fold on
  top of that, i.e. 20.3-fold.

Applying the study's own 4-fold dose reduction (20 -> 5 mg/kg/day) to the
metabolite figure gives 5.07-fold, against the paper's reported geometric mean
ratio of 4.86 (95% CI 2.11-11.2).

```{r ddi}
ddi <- bind_rows(
  check_identity(0, 5.0),
  check_identity(1, 5.0)
)

fold_parent <- ddi$auc_parent_sim[2] / ddi$auc_parent_sim[1]
fold_metab  <- (ddi$ratio_sim[2] * ddi$auc_parent_sim[2]) /
               (ddi$ratio_sim[1] * ddi$auc_parent_sim[1])

c(
  parent_fold_sim  = round(fold_parent, 2),
  parent_fold_pred = round(f_lpv * (cl_inh + cl_conv) / cl_conv, 2),
  metab_fold_sim   = round(fold_metab, 2),
  metab_fold_pred  = round(f_lpv * (cl_inh + cl_conv) / cl_conv / (1 - 0.766), 2),
  # The paper's reported des-rifabutin GMR, which is at a 4-fold lower dose.
  metab_fold_at_study_doses = round(fold_metab / 4, 2),
  paper_reported_GMR        = 4.86
)

# Deterministic (zeroRe): closed-form fold-changes, so tight bounds.
stopifnot(
  abs(fold_parent / (f_lpv * (cl_inh + cl_conv) / cl_conv) - 1) < 0.01,
  abs(fold_metab / (f_lpv * (cl_inh + cl_conv) / cl_conv / (1 - 0.766)) - 1) < 0.01
)
```

# Replicate published figures

```{r figure-1}
# Replicates the shape of Figure 1 (prediction-corrected VPC), stratified by
# presence or absence of LPV/r co-treatment, for both analytes.
sim_ss |>
  select(id, tad, arm, Cc, Cc_desacetylrbn) |>
  tidyr::pivot_longer(
    c(Cc, Cc_desacetylrbn),
    names_to = "analyte", values_to = "conc"
  ) |>
  mutate(
    analyte = recode(analyte,
                     Cc = "Rifabutin",
                     Cc_desacetylrbn = "Des-rifabutin")
  ) |>
  group_by(arm, analyte, tad) |>
  summarise(
    Q10 = quantile(conc, 0.10),
    Q50 = quantile(conc, 0.50),
    Q90 = quantile(conc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  facet_grid(analyte ~ arm, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Concentration (ug/L)",
    title = "Steady-state concentration-time profiles, 10-15 kg band",
    caption = paste(
      "Median with 10th-90th percentile band, 150 subjects per arm.",
      "Mirrors the stratification of Figure 1 of Semere Gebreyesus 2024."
    )
  ) +
  theme_bw()
```

```{r figure-2}
# Replicates Figure 2 / Figure S6: simulated steady-state AUC0-24 and Cmax
# against the adult comparator ranges and the 900 ug/L Cmax toxicity limit.
adult_ranges <- tibble::tribble(
  ~analyte,        ~lpv, ~lo,   ~hi,
  "Rifabutin",        0, 2790, 5640,
  "Rifabutin",        1, 4770, 7290,
  "Des-rifabutin",    0,  273,  700,
  "Des-rifabutin",    1, 4120, 4130
)
knitr::kable(
  adult_ranges |>
    mutate(lpv = ifelse(lpv == 1, "with LPV/r", "no LPV/r")) |>
    rename(Analyte = analyte, Setting = lpv,
           `Adult median AUC0-24, low` = lo,
           `Adult median AUC0-24, high` = hi),
  caption = paste(
    "Adult median steady-state AUC0-24 comparator ranges quoted in",
    "'Targets for simulations' (Lan, Naiker and Tanuma et al.), in ug*h/L."
  )
)
```

# PKNCA validation

NCA is run separately for each analyte over the final steady-state dosing
interval, grouped by arm so that per-arm results can be compared with the
paper's simulation claims.

```{r pknca}
run_nca <- function(conc_col) {
  conc_df <- sim_ss |>
    transmute(id, time = tad, arm, Cc = .data[[conc_col]]) |>
    filter(!is.na(Cc))

  # Guarantee a time-zero row per (arm, id) so PKNCA can anchor AUC0-tau.
  conc_df <- bind_rows(
    conc_df,
    conc_df |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, arm, time, .keep_all = TRUE) |>
    arrange(id, arm, time)

  dose_df <- events |>
    filter(evid == 1, time == ss_start) |>
    transmute(id, time = 0, amt, arm)

  d <- PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc_df, Cc ~ time | arm + id),
    PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id),
    intervals = data.frame(
      start = 0, end = tau,
      auclast = TRUE, cmax = TRUE, tmax = TRUE
    )
  )
  as.data.frame(PKNCA::pk.nca(d))
}

nca_parent <- run_nca("Cc")
nca_metab  <- run_nca("Cc_desacetylrbn")

nca_summary <- bind_rows(
  nca_parent |> mutate(analyte = "Rifabutin"),
  nca_metab  |> mutate(analyte = "Des-rifabutin")
) |>
  filter(PPTESTCD %in% c("auclast", "cmax", "tmax")) |>
  group_by(analyte, arm, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  mutate(across(c(auclast, cmax), \(x) round(x)), tmax = round(tmax, 2)) |>
  rename(
    Analyte                     = analyte,
    Arm                         = arm,
    `Median AUC0-24 (ug*h/L)`   = auclast,
    `Median Cmax (ug/L)`        = cmax,
    `Median Tmax (h)`           = tmax
  ) |>
  knitr::kable(
    caption = paste(
      "Simulated steady-state NCA in the 10-15 kg weight band, by arm and",
      "analyte (PKNCA, 150 subjects per arm)."
    )
  )
```

## Comparison against the paper's simulation claims

The paper reports percentage exposure increases relative to the *upper* limit of
the adult median comparator range (Results, "Simulation results"). Two of the
four claims are **band-specific** -- "The highest exposure increase is in the
10- to 15-kg weight category, with a 30% increase for rifabutin and a 53%
increase for des-rifabutin" -- and are compared directly. The other two are
**cross-band maxima** -- for the 5 mg/kg/day LPV/r arm, "weight groups of >7 kg
exhibited higher median exposures, reaching *up to* 12% for rifabutin and 23%
for des-rifabutin" -- so the 10-15 kg band simulated here should fall at or
below them, not equal them, and they are checked as upper bounds.

```{r claims}
get_med <- function(an, ar, param) {
  v <- nca_summary[[param]][nca_summary$analyte == an & nca_summary$arm == ar]
  if (length(v) != 1L) {
    stop("no unique NCA row for ", an, " / ", ar, " / ", param)
  }
  v
}

tb_arm  <- "TB only, 20 mg/kg/day"
lpv2_5  <- "LPV/r, 2.5 mg/kg/day"
lpv5    <- "LPV/r, 5 mg/kg/day"

claims <- tibble::tribble(
  ~claim, ~kind, ~paper, ~achieved,
  "Rifabutin AUC0-24 vs adult upper limit (5,640), TB only 20 mg/kg/day",
    "10-15 kg band", 30, 100 * (get_med("Rifabutin", tb_arm, "auclast") / 5640 - 1),
  "Des-rifabutin AUC0-24 vs adult upper limit (700), TB only 20 mg/kg/day",
    "10-15 kg band", 53, 100 * (get_med("Des-rifabutin", tb_arm, "auclast") / 700 - 1),
  "Rifabutin AUC0-24 vs adult upper limit (7,290), LPV/r 5 mg/kg/day",
    "max over bands >7 kg", 12, 100 * (get_med("Rifabutin", lpv5, "auclast") / 7290 - 1),
  "Des-rifabutin AUC0-24 vs adult upper limit (4,130), LPV/r 5 mg/kg/day",
    "max over bands >7 kg", 23, 100 * (get_med("Des-rifabutin", lpv5, "auclast") / 4130 - 1)
)

claims |>
  mutate(
    paper_disp = ifelse(kind == "10-15 kg band",
                        paste0("+", paper, "%"),
                        paste0("up to +", paper, "%")),
    ach_disp   = paste0(ifelse(achieved >= 0, "+", ""), round(achieved), "%")
  ) |>
  select(Claim = claim, `Paper's claim` = kind,
         `Paper reports` = paper_disp, `This model` = ach_disp) |>
  knitr::kable(
    caption = paste(
      "Simulated exposure increase over the upper limit of the adult median",
      "AUC0-24 range, 10-15 kg weight band, against the paper's",
      "'Simulation results' claims."
    )
  )

band <- claims |> filter(kind == "10-15 kg band")
maxb <- claims |> filter(kind == "max over bands >7 kg")

# Band-specific claims are compared directly. The paper's figures are medians
# over its own 500-children-per-kilogram virtual population, whose weight and
# covariate distribution differ from the cohort drawn here, so agreement is
# expected in magnitude rather than exactly. Realised 2 and 2 points of
# difference; the bound sits well outside that and still goes red on a
# mis-transcribed clearance, dose or molecular-weight ratio, any of which move
# these by tens of percent.
stopifnot(max(abs(band$achieved - band$paper)) < 15)

# Cross-band maxima are upper bounds: the 10-15 kg band must sit at or below
# them. The small allowance absorbs cohort-draw noise around the boundary.
stopifnot(all(maxb$achieved <= maxb$paper + 5))

# The 2.5 mg/kg/day LPV/r dose is the one the paper concludes "resulted in
# simulated exposures in line with those of adults". A cohort MEDIAN is not the
# typical-value prediction under this model's large bioavailability BOV
# (pattern 11), so the check is made on the typical value, which is also
# deterministic and therefore reproducible across thread counts.
auc_2_5_typ <- check_identity(1, 2.5)$auc_parent_sim
c(
  typical_AUC_2.5_mg_kg = round(auc_2_5_typ),
  cohort_median         = round(get_med("Rifabutin", lpv2_5, "auclast")),
  adult_range_low       = 4770,
  adult_range_high      = 7290
)
# Within a factor of ~1.25 of the adult comparator band on either side, which
# is what "aligned with median adult exposures" supports. The paper's own
# optimized 10-15 kg LPV/r dose is 35 mg versus the 31 mg that 2.5 mg/kg gives
# at the band midpoint, so a modest shortfall against the adult range is
# expected and is exactly why the authors recommend weight-band dosing instead.
stopifnot(auc_2_5_typ > 4770 / 1.25, auc_2_5_typ < 7290 * 1.25)

# "All study doses maintained a median Cmax of <900 ug/L" (Abstract; Results).
# Asserted on the MEDIAN, as the paper states it -- not on the cohort maximum,
# which is a single extreme draw and not reproducible across thread counts.
cmax_medians <- nca_summary$cmax
stopifnot(all(cmax_medians < 900))
```

# Assumptions and deviations

* **The `WAZ` (weight-for-age z-score) functional form is an extraction-side
  reading.** Table 2 and the Results give the coefficient (-26.0% "for each unit
  decrease" in children with a z-score below -3) and the -3 breakpoint, but
  neither the paper nor the supplement prints the equation, and the effect does
  not appear in the Figure S4 schematic. It is encoded here as the compounding
  form `fdepot *= (1 + e_waz_fdepot)^max(0, -3 - WAZ)`, i.e. a factor of 0.74
  per z-score unit below -3. Reasons: it is the literal reading of "26% for each
  unit decrease"; it keeps bioavailability strictly positive, whereas the linear
  alternative `(1 + e * max(0, -3 - WAZ))` reaches zero at `WAZ = -6.85` and
  goes negative below that -- inside the reachable range for a severely
  malnourished African paediatric population and for the authors' own
  22,500-child simulation; and it matches the form the register already uses for
  the closely analogous per-unit nutritional-anthropometry effect on
  bioavailability in `Chotsiri_2019_lumefantrine.R` (MUAC, "bioavailability
  falls 25.4% per 1 cm reduction"). The two forms agree at the breakpoint and
  diverge with depth: at this cohort's minimum of `WAZ = -5.15` they give 0.516
  (compounding) versus 0.441 (linear), a 17% relative difference. **This choice
  is flagged for operator confirmation and does not affect any validation above,
  all of which are run at `WAZ >= -3` where the effect is inert.**
* **Correlated residual error is not encoded.** The paper estimated a 28.2%
  (14.6-42.1) correlation between the rifabutin and des-rifabutin residual error
  terms using the NONMEM L2 method, because both analytes were measured from the
  same sample. nlmixr2 has no construct for correlated residual error across two
  outputs. Omitting it leaves every typical value and every marginal residual
  variance unchanged and affects only the joint parent/metabolite residual draw
  within a single sample.
* **Between-occasion variability is encoded as four-occasion IOV.** The paper
  reports one BOV magnitude per absorption parameter, shared across occasions
  (the NONMEM `$OMEGA BLOCK(1) SAME` pattern), and never states an occasion
  count. Four occasions are encoded to span the richest per-child sampling
  schedule described in Methods (weeks 2, 4, 6/8 and 12), with occasions 2-4
  fixed to the occasion-1 variance, following the registered idiom
  (`Jonsson_2011_ethambutol.R`, `Blackman_2026_methotrexate.R`,
  `Mascarenhas_2015_pentadecanoic_triheptadecanoic.R`). Pass `OCC = 1` for
  single-occasion simulations.
* **No between-subject variability on volumes or intercompartmental
  clearances.** Table 2's variability column is blank on those rows and the
  Results are explicit that variability was included "parsimoniously": one
  common BSV shared by the two rifabutin clearance pathways, and a separate BSV
  on des-rifabutin clearance. The `etalcl` eta is therefore deliberately added
  to both `cl` and `cl_form_desacetylrbn`. (In the trimmed text conversion of
  Table 2 the merged BSV cell is repeated onto every row of each block; the
  publisher PDF layout shows those cells are genuinely blank.)
* **The B2 initialization device is not part of the packaged model.** Fifteen
  profiles had pre-dose concentrations below one third of the corresponding 24 h
  concentration -- attributed to poor adherence or incomplete administration of
  the compounded suspension -- and were fitted by discarding the prior dosing
  history and initializing the disposition compartments to the observed
  concentration (Dansirikul B2 method; supplement "Initialization"). That is a
  fitting-time data-handling choice, not structural, so it is not encoded.
* **Age enters only through the 3-year dichotomy.** Maturation was tested on all
  clearance parameters and rejected (dOFV = -1.52, df = 2); the model estimated
  complete maturation at 8 months, the youngest age in the data set, so the
  authors note the results apply to children above 8 months. `AGE` is used
  solely to derive the `AGE <= 3` absorption-rate indicator.
* **Cohort covariate distributions are assumed.** Weight is drawn uniformly
  across the 10-15 kg band and age uniformly over 1-4 years; the paper's own
  simulations used a 22,500-child virtual African population with uniformly
  distributed weight (their reference 32) whose covariate distribution is not
  reproduced here. Sex was not reported in Table 1 and is not a covariate in the
  model.
* **Adult comparator ranges are transcribed, not simulated.** The 2,790-5,640
  and 4,770-7,290 ug*h/L rifabutin ranges (and the des-rifabutin equivalents)
  come from Lan, Naiker and Tanuma et al. as quoted in "Targets for
  simulations"; they are literature values for adults, not outputs of this
  model.
