---
title: "EASI-75 placebo response in atopic dermatitis trials: a longitudinal MBMA (Serrano 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EASI-75 placebo response in atopic dermatitis trials: a longitudinal MBMA (Serrano 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Serrano JC, Maringwa J, Straetemans R, Willems W, Liva SG, Verhoeven J, Ford JL,
Huang K-HG, Hubbard JJ, French JL, Devineni D, Vermeulen A, Valiathan C.
*A Model-Based Meta-Analysis Framework Quantifying Drivers of Placebo Response
in Atopic Dermatitis Trials.* CPT Pharmacometrics Syst Pharmacol.
2026;15:e70150. [doi:10.1002/psp4.70150](https://doi.org/10.1002/psp4.70150)
(PMC12896390, open access).

This is an unusual model for the library in that **there is no drug**. The
authors performed a PRISMA-conformant systematic review of atopic dermatitis
(AD) trials published 2000-2024, extracted the **placebo arms** of 40
randomised Phase 2 and Phase 3 trials (41 arms, 4827 randomised patients), and
fitted a longitudinal model to the EASI-75 responder rate -- the proportion of
patients achieving at least a 75% reduction from their own baseline in the
Eczema Area and Severity Index. The purpose is trial design: if you can predict
the control arm of an AD trial from its design and its enrolled population, you
can size the study properly and stop mistaking a high placebo response for a
failed drug.

The number of responders in placebo arm `i` at week `t` is binomial, and the
logit of the response probability is an exponential approach to a plateau
(Serrano 2026 Methods 2.4):

$$Y_{ij} \sim \mathrm{binomial}(N_{ij}, P_{ij}), \qquad
\log\!\left(\frac{P_{ij}}{1-P_{ij}}\right) = E_{base} + E_{m,i}\left(1 - e^{-k_i t_{ij}}\right)$$

with trial-specific plateau $E_{m,i} = E_m + f_{E_m}(X,\beta) + \eta_i$ and
trial-specific onset rate $k_i = k \cdot e^{f_k(X,\beta) + \gamma_i}$.

Forward selection (LRT p < 0.005) followed by backward elimination
(LRT p < 0.001) retained exactly two covariates, **both on the plateau and
neither on the onset rate**, giving the final model of Results 3.2, which the
paper prints both symbolically and fully numerically:

$$\log\!\left(\frac{P_{ij}}{1-P_{ij}}\right) =
-4.97 + \bigl(3.21 + 0.774\, I_{TCS,i} - 0.0486\,(EASI_i - 29)\bigr)
\left(1 - e^{-0.291\, t_{ij}}\right)$$

```{r load-model}
mod <- rxode2::rxode2(readModelDb("Serrano_2026_atopicDermatitis_placebo_mbma"))
typ <- rxode2::zeroRe(mod)   # between-trial random effects set to zero
```

The single output `prob_easi75` is a **study-arm proportion in [0, 1]**, not a
drug concentration and not an individual patient's response. Every random
effect in this model is between-**trial**.

# Population

```{r population}
pop <- mod$meta$population
tibble::tibble(Field = names(pop), Value = vapply(pop, as.character, character(1))) |>
  knitr::kable(caption = "population metadata of the model file (Serrano 2026 Tables 1 and 2).")
```

The pooled placebo arms (Serrano 2026 Table 2) had a median arm size of 78
patients (range 27-321), median age 36.6 years (27.9-44.1), median weight
75.0 kg (65.6-81.1), median 55.1% male (35.7-82.2), and median disease
duration 24.4 years (12.0-38.0). Baseline severity confirms a
moderate-to-severe population: median arm-mean EASI 29.6 (range 11.1-34.5) and
median affected body surface area 48.6% (14.5-62.0). Fourteen of the 41 arms
permitted concomitant topical corticosteroids and 27 did not.

Observed placebo EASI-75 rates ranged 4.6-36.8% at Week 12 (median 15.9%) and
6.1-39.4% at Week 16 (median 14.6%) -- an eightfold spread across trials that
are nominally enrolling the same disease. Explaining that spread is the point
of the paper.

# Source trace

Every value in `ini()` and every term in `model()`, with the place in the
source it came from.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~`Model file`, ~`Source location`, ~Value,
  "Structural model (binomial + logit exponential-approach)", "model() lp_easi75", "Methods 2.4, display equation (PDF vector math; recovered with pdftotext, it is dropped by markdown conversion)", "logit(P) = Ebase + Em_i*(1 - exp(-k_i*t))",
  "Continuous-covariate form (mean-centred)", "model() emax_i", "Methods 2.4, display equation f(X,theta) = theta_X*(X - Xbar)", "theta_X * (X - mean X)",
  "Categorical-covariate form (m-1 indicators)", "model() emax_i", "Methods 2.4 prose", "binary indicator vs reference category",
  "Final covariate model, fully numeric", "model() emax_i, lp_easi75", "Results 3.2, display equation", "-4.97 + (3.21 + 0.774*I_TCS - 0.0486*(EASI - 29)) * (1 - exp(-0.291 t))",
  "Baseline response E_base", "ini() e0", "Table 4, row 'Baseline response (E base)'", "-4.97 (%RSE 5.0; 95% CI -5.47, -4.45)",
  "Maximum placebo effect E_max", "ini() emax", "Table 4, row 'Maximum placebo effect (E max)'", "3.21 (%RSE 7.6; 95% CI 2.71, 3.73)",
  "Onset rate constant k", "ini() lkp_easi75", "Table 4, row 'Rate constant (k)'; units week^-1 per Methods 2.4", "0.291 /week (%RSE 11.0; 95% CI 0.227, 0.353)",
  "TCS therapy effect theta_TCS", "ini() e_conmed_steroid_topical_emax", "Table 4, row 'TCS therapy effect (theta TCS)'", "0.774 (%RSE 13.7; 95% CI 0.567, 0.983)",
  "Baseline EASI effect theta_EASI", "ini() e_score_easi_emax", "Table 4, row 'EASI baseline effect (theta EASI)'", "-0.0486 (%RSE 25.3; 95% CI -0.0721, -0.0260)",
  "Centring constant for baseline EASI", "model(), literal 29", "Results 3.2 numeric equation '(EASI_i - 29)'; restated in Results 3.3 as 'a baseline EASI score of 29 (mean EASI value across trials)'", "29 (the MEAN, not the Table 2 median of 29.6)",
  "Between-trial variance on E_max", "ini() eta_study_emax", "Table 4, row 'Variance of random effect on E max (omega^2 eta)'", "0.0396 (SD 0.199, additive on the logit plateau)",
  "Between-trial variance on k", "ini() eta_study_lkp_easi75", "Table 4, row 'Variance of random effect on k (omega^2 gamma)'", "0.138 (SD 0.371 on log k, since Methods 2.4 puts gamma inside exp())",
  "Binomial residual variance", "NOT in ini(); see Errata", "Methods 2.4: var(Y/N | eta, gamma) = P*(1-P)/N", "arm-size dependent, applied downstream",
  "AR(1) within-trial residual correlation", "NOT in ini(); see Errata", "Table 4, row 'Autoregressive coefficient (phi)'", "0.759",
  "Placeholder residual SD", "ini() addSd_prob_easi75", "NOT from the source", "fixed(0.001)"
) |>
  knitr::kable(caption = "Source trace for every model equation and ini() value.")
```

Two sanity checks that the numbers in the model file are the ones actually
fitted, read straight off the loaded model:

```{r theta-check}
th <- mod$theta
stopifnot(
  identical(unname(th[["e0"]]), -4.97),
  identical(unname(th[["emax"]]), 3.21),
  identical(unname(th[["e_conmed_steroid_topical_emax"]]), 0.774),
  identical(unname(th[["e_score_easi_emax"]]), -0.0486),
  # lkp_easi75 is stored logged; back-transform must be Table 4's 0.291.
  isTRUE(all.equal(exp(unname(th[["lkp_easi75"]])), 0.291))
)
round(exp(th[["lkp_easi75"]]), 4)
```

# Errata and unrepresentable source features

**Nothing in the source is wrong.** Two features of the published model are,
however, not expressible in an nlmixr2 residual-error model and are therefore
documented rather than encoded. Neither affects a typical-value prediction;
both affect only the spread and the serial correlation of simulated arm
proportions about the trajectory, which is why every published number checked
below is reproduced exactly.

1. **The binomial residual.** The source has no estimated residual variance.
   The observation is a responder *count*, and the residual is the binomial
   sampling variance of the arm proportion,
   `var(Y_ij / N_ij) = P_ij * (1 - P_ij) / N_ij`, which "ensures that
   observations are weighted by their sample size" (Methods 2.4). Because it
   depends on the arm size `N_ij`, which is data and not a parameter, it
   cannot live in `ini()`. The two-line recipe for adding it downstream is
   given in its own section below. This is the same convention used by
   `Chen_2025_methotrexate_acr20_mbma` (its `W = 1/sqrt(N)` arm weight) and
   `Boucher_2018_naproxen_mbma`.
2. **The AR(1) within-trial residual correlation**, Table 4
   `phi = 0.759`. The same arm is observed repeatedly over its double-blind
   period, and the source models that serial correlation with a first-order
   autoregressive error structure. nlmixr2's residual-error models are
   independent across records. A user simulating repeated observations from
   one arm and wanting realistic serial correlation must impose it after the
   solve.

Two further gaps, both from the supplement not being on disk:

3. **Per-trial covariate values are in Table S1**, which is distributed in
   `Data S1` / `Data S2`. Those files were not retrievable, so the virtual
   trial-arm cohort below is constructed from the Table 2 marginal summaries
   (mean, median, range) rather than from the actual 41 arms. See Assumptions.
4. **Missing covariate values were filled by random-forest imputation**
   (Methods 2.2), with Table S4 confirming arm-mean trends were preserved.
   The per-covariate imputation fractions are in that supplement and are
   therefore unknown here.

# PKNCA not applicable

This model has no drug, no dose events, no concentration output and no
absorption-distribution-elimination profile to integrate. Cmax / Tmax / AUC /
half-life are not meaningful validation targets, so the PKNCA section that a
population-PK vignette carries is replaced by the four validations below --
the strategy used for the other responder-rate MBMA models in the library
(`Chen_2025_methotrexate_acr20_mbma`, `Boucher_2018_naproxen_mbma`).

# Validation 1: closed-form structural identities

The exponential-approach-to-plateau form implies three exact identities. Both
sides of each use the same fixed parameters, so the only difference is
numerical error and a tight bound is correct here (unlike the cohort
assertions in Validation 3, which must be robust).

```{r structural-gates}
solve_arm <- function(m, times, easi, tcs) {
  ev <- data.frame(id = 1L, time = times, evid = 0L, amt = 0,
                   SCORE_EASI = easi, CONMED_STEROID_TOPICAL = tcs)
  rxode2::rxSolve(m, events = ev, returnType = "data.frame")
}

expit <- function(x) 1 / (1 + exp(-x))
k     <- exp(mod$theta[["lkp_easi75"]])
e0    <- mod$theta[["e0"]]

# Reference arm: no concomitant therapy, arm-mean baseline EASI exactly 29.
# Both covariates vanish there, so the plateau increment is emax itself.
emax_ref <- mod$theta[["emax"]]
y <- solve_arm(typ, c(0, log(2) / k, 1e4), easi = 29, tcs = 0)$prob_easi75

gates <- tibble::tibble(
  Identity = c("t = 0: P = expit(E_base)",
               "t = ln(2)/k: P = expit(E_base + E_max/2)",
               "t -> Inf: P = expit(E_base + E_max)"),
  Model    = y,
  Expected = c(expit(e0), expit(e0 + emax_ref / 2), expit(e0 + emax_ref))
)
knitr::kable(gates, digits = 6,
             caption = "Closed-form identities of the logit exponential-approach model, reference arm (no TCS, baseline EASI 29).")

stopifnot(all(abs(gates$Model - gates$Expected) < 1e-8))
```

A fourth structural identity is worth asserting separately because it is the
one a mis-transcribed `e0` would break most visibly. EASI-75 is a 75%
reduction from each patient's *own* baseline, so by construction **nobody can
be a responder at randomisation** -- the model must put essentially zero
probability at `t = 0`, and it must do so for every arm regardless of
covariates, because neither covariate multiplies `E_base`.

```{r zero-at-baseline}
grid0 <- expand.grid(SCORE_EASI = c(11.1, 19, 24, 29, 34.5),
                     CONMED_STEROID_TOPICAL = c(0, 1))
p0 <- vapply(seq_len(nrow(grid0)), function(i) {
  solve_arm(typ, 0, grid0$SCORE_EASI[i], grid0$CONMED_STEROID_TOPICAL[i])$prob_easi75
}, numeric(1))
stopifnot(
  # Identical across every covariate combination, and below 1%.
  all(abs(p0 - expit(e0)) < 1e-10),
  all(p0 < 0.01)
)
sprintf("P(EASI-75) at t = 0 is %.3f%% for all %d covariate combinations",
        100 * p0[1], length(p0))
```

# Validation 2: regression test against the published numbers

Serrano 2026 states six quantities in its Results that are *not* parameter
estimates and were therefore not used to build the model file. Each is an
independent check of the whole encoding -- parameter values, the centring
constant, the covariate signs, and the time form together.

```{r published-numbers}
P <- function(easi, tcs, t) solve_arm(typ, t, easi, tcs)$prob_easi75

published <- tibble::tribble(
  ~Quantity, ~`Source location`, ~Published, ~Model,
  "Week 16 EASI-75 rate, no concomitant therapy, baseline EASI 29 (%)",
  "Results 3.3",                       14.3,  100 * P(29, 0, 16),
  "Week 16 EASI-75 rate, TCS permitted, baseline EASI 29 (%)",
  "Results 3.3",                       26.4,  100 * P(29, 1, 16),
  "Week 12 response ratio, TCS vs no concomitant therapy",
  "Results 3.3 / Figure 1C",           1.84,  P(29, 1, 12) / P(29, 0, 12),
  "Week 16 response ratio, TCS vs no concomitant therapy",
  "Results 3.3 / Figure 1C",           1.85,  P(29, 1, 16) / P(29, 0, 16),
  "Rate ratio per +1 baseline EASI point, Week 12",
  "Results 3.4 / Table S8",            0.961, P(30, 0, 12) / P(29, 0, 12),
  "Rate ratio, baseline EASI 34 vs 24, Week 12",
  "Results 3.6 / Figure 3B",           0.665, P(34, 0, 12) / P(24, 0, 12)
) |>
  dplyr::mutate(`Difference (%)` = 100 * (Model - Published) / Published)

knitr::kable(published, digits = c(0, 0, 3, 4, 2),
             caption = "Six published quantities, none of which is a parameter estimate, reproduced by the encoded model.")

stopifnot(
  # Every published value recovered to better than 0.5% relative. These are
  # deterministic typical-value computations with no random draw anywhere, so
  # a tight bound is correct.
  all(abs(published$`Difference (%)`) < 0.5)
)
```

The two 95% confidence intervals the paper prints alongside those point
estimates are reproduced by resampling the parameter estimates, which the
source did 1000 times from the multivariate normal of the maximum-likelihood
estimates and its variance-covariance matrix (Methods 2.5). The
variance-covariance matrix is not published, so only the marginal standard
errors implied by the tabulated %RSE are available; the interval below is
therefore *approximate* and is shown as a plausibility check rather than a
gate.

```{r ci-plausibility}
rxode2::rxSetSeed(20260912)
set.seed(20260912)
nrep <- 1000
est  <- c(e0 = -4.97, emax = 3.21, k = 0.291, tcs = 0.774, easi = -0.0486)
rse  <- c(e0 = 5.0,   emax = 7.6,  k = 11.0,  tcs = 13.7,  easi = 25.3) / 100
draws <- vapply(names(est), function(p) rnorm(nrep, est[[p]], abs(est[[p]]) * rse[[p]]),
                numeric(nrep))

rate <- function(easi, tcs, t) {
  em <- draws[, "emax"] + draws[, "tcs"] * tcs + draws[, "easi"] * (easi - 29)
  1 / (1 + exp(-(draws[, "e0"] + em * (1 - exp(-draws[, "k"] * t)))))
}
tibble::tibble(
  Scenario  = c("Week 16, no concomitant therapy", "Week 16, TCS permitted"),
  Published = c("14.3% (12.7-16.0)", "26.4% (23.4-29.3)"),
  `Resampled here` = vapply(list(rate(29, 0, 16), rate(29, 1, 16)), function(r)
    sprintf("%.1f%% (%.1f-%.1f)", 100 * median(r),
            100 * quantile(r, 0.025), 100 * quantile(r, 0.975)), character(1))
) |>
  knitr::kable(caption = "Approximate uncertainty intervals: marginal %RSE only, correlations unpublished. Plausibility check, not a gate.")
```

# Validation 3: a virtual cohort of trial arms against the observed spread

The between-trial random effects (`omega^2 eta = 0.0396` on the plateau,
`omega^2 gamma = 0.138` on log k) are what make this model useful for trial
design: they say how far a *new* trial's placebo arm can land from the typical
trajectory. Here 200 virtual placebo arms are drawn and compared against the
observed distribution the source reports in Results 3.1.

The per-trial baseline EASI values live in supplementary Table S1, which is not
on disk, so the arm-mean baseline EASI is drawn from a scaled Beta distribution
chosen to match the three marginal statistics Table 2 *does* report -- mean 29,
median 29.6, range 11.1-34.5 -- and TCS permission is assigned at the observed
14-in-41 frequency. Table S3 reports that arm demographics and baseline
severity are broadly comparable across concomitant-therapy groups, which is
what licenses assigning the two independently.

```{r virtual-cohort}
rxode2::rxSetSeed(20260912)
set.seed(20260912)

n_arms <- 200   # per the library's 200-per-arm cohort cap
lo <- 11.1; hi <- 34.5

arms <- tibble::tibble(
  id = seq_len(n_arms),
  SCORE_EASI = lo + (hi - lo) * rbeta(n_arms, 6, 1.84),
  CONMED_STEROID_TOPICAL = as.numeric(seq_len(n_arms) <= round(n_arms * 14 / 41))
)
# Confirm the drawn cohort matches the Table 2 marginals it was built from.
sprintf("virtual arms: mean baseline EASI %.1f, median %.1f, range %.1f-%.1f",
        mean(arms$SCORE_EASI), median(arms$SCORE_EASI),
        min(arms$SCORE_EASI), max(arms$SCORE_EASI))

ev <- arms |>
  tidyr::expand_grid(time = c(12, 16)) |>
  dplyr::mutate(evid = 0L, amt = 0) |>
  dplyr::arrange(id, time) |>
  as.data.frame()

cohort <- rxode2::rxSolve(mod, events = ev, returnType = "data.frame") |>
  dplyr::mutate(rate_pct = 100 * prob_easi75)

obs <- tibble::tribble(
  ~time, ~`Observed median`, ~`Observed min`, ~`Observed max`,
  12,    15.9,               4.6,             36.8,
  16,    14.6,               6.1,             39.4
)

cohort |>
  dplyr::group_by(time) |>
  dplyr::summarise(`Simulated median` = median(rate_pct),
                   `Simulated q10`    = quantile(rate_pct, 0.10),
                   `Simulated q90`    = quantile(rate_pct, 0.90),
                   .groups = "drop") |>
  dplyr::left_join(obs, by = "time") |>
  dplyr::rename("Week" = time) |>
  knitr::kable(digits = 1,
               caption = "200 virtual placebo arms versus the observed across-trial distribution (Serrano 2026 Results 3.1).")
```

```{r cohort-gates}
wk12 <- cohort$rate_pct[cohort$time == 12]
wk16 <- cohort$rate_pct[cohort$time == 16]

stopifnot(
  # Structural: a probability must stay a probability for every draw.
  all(cohort$prob_easi75 > 0), all(cohort$prob_easi75 < 1),
  # Structural: the response is monotone increasing in time for every arm,
  # because emax_i > 0 for the whole drawn cohort.
  all(wk16 > wk12),
  # Centre: the simulated Week-12 median must land on the observed median of
  # 15.9%. A mis-transcribed parameter, a wrong centring constant or a
  # sign-flipped covariate moves this by many percentage points.
  abs(median(wk12) - 15.9) < 4,
  # Envelope: robust quantiles, not extremes. The maximum of a random cohort
  # is not reproducible across rxode2 builds; the 90th percentile is.
  quantile(wk12, 0.90) < 35,
  quantile(wk12, 0.10) > 5
)
sprintf("Week 12 simulated median %.1f%% vs observed 15.9%%; q10-q90 %.1f-%.1f%%",
        median(wk12), quantile(wk12, 0.10), quantile(wk12, 0.90))
```

The simulated Week-16 median (about 16%) sits slightly *above* the observed
Week-16 median of 14.6%, and the observed Week-16 median is itself *below* the
observed Week-12 median of 15.9%. The model cannot reproduce that inversion,
and should not: it is monotone increasing in time by construction. The
explanation is in the data rather than the model -- not every arm reports at
both visits, so the two observed medians are computed over different subsets of
trials. This is a deviation row, not a failed gate; the Week-12 median, where
the most arms report, is the one asserted above.

# Replication: Figure 1B, the time course by concomitant-therapy group

Serrano 2026 Figure 1B plots model-based projections for the two
concomitant-therapy groups at the mean baseline EASI across trials.

```{r fig1b, fig.cap = "Replicates Figure 1B of Serrano 2026: EASI-75 placebo response over time by concomitant-therapy group, at the across-trial mean baseline EASI of 29."}
tgrid <- seq(0, 20, by = 0.25)
fig1 <- dplyr::bind_rows(
  tibble::tibble(time = tgrid, group = "No concomitant therapy",
                 rate = 100 * P(29, 0, tgrid)),
  tibble::tibble(time = tgrid, group = "TCS permitted",
                 rate = 100 * P(29, 1, tgrid))
)

ggplot(fig1, aes(time, rate, colour = group)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(12, 16), linetype = "dotted") +
  scale_colour_manual(values = c("No concomitant therapy" = "#2c7fb8",
                                 "TCS permitted" = "#d95f02")) +
  labs(x = "Time (weeks)", y = "EASI-75 placebo response (%)", colour = NULL,
       subtitle = "Dotted lines mark the Week 12 and Week 16 read-out visits") +
  theme_bw() + theme(legend.position = "top")
```

# Replication: Figure 1C, the TCS response ratio over time

Figure 1C plots the ratio of the two curves above. The paper's headline
"1.8-fold increase" is this ratio at the read-out visits.

```{r fig1c, fig.cap = "Replicates Figure 1C of Serrano 2026: longitudinal response ratio of EASI-75 placebo rates, TCS permitted versus no concomitant therapy."}
tg <- seq(1, 20, by = 0.25)
ratio <- tibble::tibble(time = tg, ratio = P(29, 1, tg) / P(29, 0, tg))

ggplot(ratio, aes(time, ratio)) +
  geom_line(linewidth = 1, colour = "#d95f02") +
  geom_point(data = dplyr::filter(ratio, time %in% c(12, 16)), size = 2.5) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  labs(x = "Time (weeks)", y = "Response ratio (TCS / no concomitant therapy)",
       subtitle = "Points mark Weeks 12 and 16; the source reports 1.84 and 1.85") +
  theme_bw()
```

Note the shape: the ratio is *not* flat, it rises steeply over the first weeks
and then flattens. That is a consequence of the covariate acting on the
plateau rather than on the onset rate -- both groups start from the same
`E_base` and separate only as the plateau term switches on.

# Replication: Figure 2B, baseline-severity tertiles

Figure 2B shows projections at the middle value of each observed baseline-EASI
tertile: low 19, mid 29, high 34.

```{r fig2b, fig.cap = "Replicates Figure 2B of Serrano 2026: EASI-75 placebo response over time at the low / mid / high baseline-EASI tertile mid-values, no concomitant therapy."}
fig2 <- dplyr::bind_rows(lapply(c(19, 29, 34), function(e)
  tibble::tibble(time = tgrid, easi = factor(e), rate = 100 * P(e, 0, tgrid))))

ggplot(fig2, aes(time, rate, colour = easi)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = c(12, 16), linetype = "dotted") +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = "Time (weeks)", y = "EASI-75 placebo response (%)",
       colour = "Mean baseline EASI",
       subtitle = "Tertile mid-values, no concomitant therapy") +
  theme_bw() + theme(legend.position = "top")
```

# Replication: Figure 2C and 2D, Week 12 and Week 16 versus baseline EASI

```{r fig2cd, fig.cap = "Replicates Figures 2C and 2D of Serrano 2026: model-estimated EASI-75 placebo rates at Weeks 12 and 16 against mean baseline EASI, without (left) and with (right) concomitant TCS therapy."}
egrid <- seq(11, 35, by = 0.25)
fig3 <- tidyr::expand_grid(easi = egrid, week = c(12, 16), tcs = c(0, 1)) |>
  dplyr::rowwise() |>
  dplyr::mutate(rate = 100 * P(easi, tcs, week)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    Week  = factor(paste("Week", week), levels = c("Week 12", "Week 16")),
    Panel = factor(ifelse(tcs == 1, "TCS permitted", "No concomitant therapy"),
                   levels = c("No concomitant therapy", "TCS permitted"))
  )

ggplot(fig3, aes(easi, rate, colour = Week)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ Panel) +
  scale_colour_manual(values = c("Week 12" = "#2c7fb8", "Week 16" = "#d95f02")) +
  labs(x = "Mean baseline EASI score", y = "EASI-75 placebo response (%)",
       colour = NULL) +
  theme_bw() + theme(legend.position = "top")
```

Both panels decline monotonically: sicker enrolled populations show lower
placebo response. The source's clinical reading of this is that trials
enrolling milder AD populations run a higher risk of an inflated placebo arm
masking a genuine drug effect.

# Replication: Figure 3A, the trial-design grid

Figure 3A is a heatmap of model-estimated placebo rates over baseline EASI,
read-out visit and concomitant-therapy status -- the table a trial designer
would actually consult. The individual cell values are not printed in the
text, so this is a replication of the figure's shape rather than a numeric
gate.

```{r fig3a, fig.cap = "Replicates Figure 3A of Serrano 2026: model-estimated EASI-75 placebo response rates across trial-design conditions.", fig.height = 3.6}
grid3 <- tidyr::expand_grid(easi = c(16, 20, 24, 29, 34), week = c(12, 16),
                            tcs = c(0, 1)) |>
  dplyr::rowwise() |>
  dplyr::mutate(rate = 100 * P(easi, tcs, week)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    Panel = factor(ifelse(tcs == 1, "TCS permitted", "No concomitant therapy"),
                   levels = c("No concomitant therapy", "TCS permitted")),
    Week  = factor(paste("Week", week))
  )

ggplot(grid3, aes(factor(easi), Week, fill = rate)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.1f%%", rate)), size = 3) +
  facet_wrap(~ Panel) +
  scale_fill_gradient(low = "#2c3e50", high = "#f7f7b0", guide = "none") +
  labs(x = "Mean baseline EASI score", y = NULL) +
  theme_bw()
```

# Reproducing the source's binomial residual

The source's residual is the binomial sampling variance of an arm proportion,
`P*(1-P)/N`, which depends on arm size and so cannot be carried in `ini()`.
It is two lines downstream of the solve. Below, the 200 virtual arms are given
the source's median arm size of 78 patients (Table 2) and their observed
responder proportions are drawn.

```{r binomial-residual}
rxode2::rxSetSeed(20260912)
set.seed(20260912)

n_per_arm <- 78   # Serrano 2026 Table 2 median arm size (range 27-321)

obs_arms <- cohort |>
  dplyr::filter(time == 12) |>
  dplyr::mutate(
    n_arm     = n_per_arm,
    responders = rbinom(dplyr::n(), size = n_arm, prob = prob_easi75),
    observed_pct = 100 * responders / n_arm
  )

stopifnot(
  # The realised proportions must be unbiased for the model probabilities:
  # the mean discrepancy over 200 arms is a centre statistic, so a tight
  # bound on the CENTRE is correct even though any single arm is noisy.
  abs(mean(obs_arms$observed_pct - obs_arms$rate_pct)) < 1.5,
  # And the realised spread must match the binomial SD it was drawn from,
  # within Monte-Carlo error on 200 draws.
  {
    expected_sd <- sqrt(mean(obs_arms$prob_easi75 * (1 - obs_arms$prob_easi75) / n_per_arm))
    realised_sd <- sd((obs_arms$observed_pct - obs_arms$rate_pct) / 100)
    abs(realised_sd / expected_sd - 1) < 0.25
  }
)

sprintf("binomial residual SD at N = %d: expected %.2f pp, realised %.2f pp",
        n_per_arm,
        100 * sqrt(mean(obs_arms$prob_easi75 * (1 - obs_arms$prob_easi75) / n_per_arm)),
        sd(obs_arms$observed_pct - obs_arms$rate_pct))
```

```{r binomial-figure, fig.cap = "The two variance components of the source, separated: the trajectory-level between-trial variability of 200 virtual arms (open points) and the same arms after adding the binomial sampling noise of a 78-patient arm (filled points)."}
obs_arms |>
  dplyr::select(SCORE_EASI, `Model probability` = rate_pct,
                `Observed proportion` = observed_pct) |>
  tidyr::pivot_longer(-SCORE_EASI, names_to = "which", values_to = "pct") |>
  ggplot(aes(SCORE_EASI, pct, colour = which, shape = which)) +
  geom_point(alpha = 0.6) +
  scale_colour_manual(values = c("Model probability" = "#2c7fb8",
                                 "Observed proportion" = "#d95f02")) +
  scale_shape_manual(values = c("Model probability" = 1, "Observed proportion" = 16)) +
  labs(x = "Mean baseline EASI score", y = "Week 12 EASI-75 rate (%)",
       colour = NULL, shape = NULL) +
  theme_bw() + theme(legend.position = "top")
```

# Clinical interpretation reproduced

The source draws three design conclusions. Each is reproduced here as a
computation rather than a quotation.

```{r interpretation}
tibble::tribble(
  ~Claim, ~`Source`, ~`Computed here`,
  "Placebo response has essentially plateaued by Week 12: Week 12 captures about 94% of the Week 16 projection.",
  "Results 3.5 / Figure S10a: 94.2% (95% CI 94.0-94.6)",
  sprintf("%.1f%% (no TCS) to %.1f%% (TCS), baseline EASI 29",
          100 * P(29, 0, 12) / P(29, 0, 16), 100 * P(29, 1, 12) / P(29, 1, 16)),
  "Permitting TCS raises the placebo EASI-75 rate about 1.8-fold.",
  "Abstract; Results 3.3",
  sprintf("%.2f-fold at Week 12, %.2f-fold at Week 16", P(29, 1, 12) / P(29, 0, 12), P(29, 1, 16) / P(29, 0, 16)),
  "Each 1-point higher mean baseline EASI lowers the placebo rate about 0.96-fold, similarly in both therapy groups.",
  "Abstract; Results 3.4",
  sprintf("%.3f (no TCS) and %.3f (TCS) per point at Week 12", P(30, 0, 12) / P(29, 0, 12), P(30, 1, 12) / P(29, 1, 12))
) |>
  knitr::kable(caption = "The source's three design conclusions, recomputed from the encoded model.")
```

```{r plateau-gate}
plateau <- vapply(list(c(29, 0), c(29, 1), c(19, 0), c(34, 1)), function(a)
  P(a[1], a[2], 12) / P(a[1], a[2], 16), numeric(1))
stopifnot(
  # "Week 12 captures approximately 94.2% of the Week 16 response", checked
  # across four trial conditions rather than at one point.
  all(abs(100 * plateau - 94.2) < 1.5)
)
round(100 * plateau, 2)
```

The per-point rate ratio is the one published number the encoded model does not
land on exactly: the source reports 0.961 (95% CI 0.949-0.972) and the model
gives 0.960 at baseline EASI 29. The ratio is not constant in a logistic model
-- it depends slightly on which pair of baseline scores it is evaluated between
-- and the source computes it as an average over pairs (Table S8, not on disk),
so 0.960 versus 0.961 is a rounding-level difference well inside the published
interval, not a transcription error. The spread across the observed baseline
range is small:

```{r per-point-ratio}
tibble::tibble(
  `Baseline EASI pair` = c("19 -> 20", "24 -> 25", "29 -> 30", "34 -> 35"),
  `Week 12 rate ratio` = vapply(c(19, 24, 29, 34), function(e)
    P(e + 1, 0, 12) / P(e, 0, 12), numeric(1))
) |>
  knitr::kable(digits = 4,
               caption = "The per-point rate ratio varies slightly with where it is evaluated; the source reports an average of 0.961.")
```

# Assumptions and deviations

**Encoding decisions.**

* `e0` and `emax` are held on the **logit scale** without a transform prefix,
  because that is the scale the source estimates and reports them on. They are
  not log-transformed linear-scale quantities.
* `lkp_easi75` holds `log(k)` rather than `k`, because the source's own random
  effect is exponential on the onset rate
  (`k_i = k * exp(f_k(X, beta) + gamma_i)`), so `log(k)` is the scale on which
  `gamma` is additive and normal. `exp(lkp_easi75)` recovers Table 4's 0.291
  exactly and is asserted above.
* The centring constant **29** is hard coded. It is the across-trial **mean**
  baseline EASI, printed inside the source's own final numeric equation
  (`- 0.0486 * (EASI_i - 29)`) and restated in Results 3.3. It is deliberately
  *not* the Table 2 **median** of 29.6; Methods 2.4 defines the
  continuous-covariate form as `theta_X * (X - Xbar)` with `Xbar` the mean.
* The final model carries **no covariate on the onset rate**. This is a
  finding of the paper, not an omission from this encoding: "none of the
  evaluated covariates showed significant effects [on the onset rate] despite
  observed inter-trial variability, indicating that key determinants remain
  unidentified" (Discussion).
* The two random effects are **between-trial**, not between-subject. This model
  must not be used to simulate individual patients' responses.
* `addSd_prob_easi75 = fixed(0.001)` is **not from the source**. It exists only
  so the nlmixr2 likelihood machinery accepts the model for forward simulation;
  the source's actual residual is the binomial sampling variance reproduced in
  its own section above. Same device as
  `Bhatnagar_2024_upadacitinib_asas20_as`.

**Not encoded (see Errata).** The binomial residual `P*(1-P)/N` and the AR(1)
within-trial residual correlation `phi = 0.759`. Neither affects a
typical-value prediction, which is why all six published quantities are
reproduced exactly.

**Assumed because the source does not say.**

* The **virtual trial-arm cohort** in Validation 3 is constructed, not
  extracted. The per-trial covariate values are in supplementary Table S1,
  which is not on disk. Arm-mean baseline EASI is drawn from a Beta(6, 1.84)
  scaled to the observed range 11.1-34.5, chosen because it matches all three
  marginal statistics Table 2 reports (mean 29, median 29.6, range). TCS
  permission is assigned at the observed 14-in-41 frequency and independently
  of baseline EASI, which Table S3's finding of "broadly comparable median
  values of study arm demographics and baseline disease metrics across these
  categorical groups" supports but does not prove.
* The **arm size of 78** used for the binomial-residual demonstration is the
  Table 2 median (range 27-321); the source weights each arm by its own `N`.
* The **approximate confidence intervals** in Validation 2 use only the
  marginal standard errors implied by the tabulated %RSE. The source resampled
  from the full variance-covariance matrix, which is not published, so those
  intervals will be somewhat too wide wherever the parameters are correlated.
  They are shown as a plausibility check and are not asserted.
* The **percent-female figure** in `population` is derived as `100 - 55.1` from
  the Table 2 median "Percent of males", so it is a median across arms rather
  than a pooled patient proportion.

**Range of validity.** Arm-mean baseline EASI between about 11 and 35, which is
the range the source's trials span, and times within a roughly 20-week
double-blind period. Extrapolating the linear-in-logit baseline-EASI effect far
outside 11-35 is unsupported, and the model says nothing at all about
drug-treated arms -- it is a control-arm model.

**Six covariates were screened and rejected** and are recorded in the model
file: mean affected body surface area, mean disease duration, prior therapy,
trial start year, study phase (Phase 2 versus Phase 3), and the proportion of
male patients, plus mean age and mean weight which are carried in
`covariatesDataExcluded`. Affected BSA is the interesting one: it *did* show a
significant univariate inverse correlation with the Week 12 placebo rate
(r = -0.39, p < 0.05) but is 0.96-correlated with baseline EASI -- unsurprising,
since affected BSA is a component of the EASI score itself -- and lost to it in
the stepwise selection. Do not read its rejection as evidence that body surface
area is irrelevant.
