---
title: "Acetaminophen + diclofenac opioid-sparing effect (Sethi 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acetaminophen + diclofenac opioid-sparing effect (Sethi 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7.5, fig.height = 3.6)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sethi_2024_acetaminophen_diclofenac_mbma"))$reference`
* Article: <https://doi.org/10.1007/s40122-023-00569-z>
* Supplementary material (Tables S1-S5, Figures S1-S4): included with the
  article at the DOI above.

Sethi 2024 is a model-based meta-analysis (MBMA) of the *opioid-sparing effect*
(OSE) of combining acetaminophen and diclofenac in acute pain. The authors set
out to answer a question about mild-to-moderate osteoarthritis pain, for which
no combination trial exists, by borrowing evidence from the acute-pain setting
and extrapolating.

Two endpoints were considered. Pain-score reduction was examined only
descriptively -- the eleven identified trials used too many different pain
scales, and the trials that permitted opioid patient-controlled analgesia (PCA)
confound a pain score with the rescue opioid the patient self-administered. The
authors therefore fit their MBMA to the second endpoint, cumulative opioid PCA
consumption, which is *less* susceptible to that confounding. **Only the
opioid-sparing model is a fitted model, and it is the only thing this vignette
reproduces**; the pain-score analysis (Supplementary Table S3, Figures S1-S2)
is a tabulation of observed ratios with no parameters.

The model is a static algebraic MBMA operating on trial-arm means. It has no
time course, no dose events and no pharmacokinetics.

```{r load-model}
mod_full <- rxode2::rxode(readModelDb("Sethi_2024_acetaminophen_diclofenac_mbma"))
mod_typ  <- rxode2::zeroRe(mod_full)
mod_full
```

# Population

```{r population}
pop <- mod_full$population
knitr::kable(
  data.frame(Field = names(pop), Value = unlist(lapply(pop, as.character))),
  row.names = FALSE,
  caption = "Population metadata recorded with the model."
)
```

The five trials that entered the final MBMA are the subset of Sethi 2024
Table 2 marked "Included in the final analysis = Yes". Their sample sizes sum
to exactly the 353 patients reported in the Abstract:

```{r included-trials}
trials <- data.frame(
  study      = c("Montgomery 1996", "Beck 2000", "Siddik 2001",
                 "Hiller 2004", "Munishankar 2008"),
  population = c("Women", "Women", "Women", "Adults", "Women"),
  indication = c("Elective gynecological surgery", "Hysterectomy",
                 "Cesarean section", "Tonsillectomy", "Cesarean section"),
  n          = c(59L, 65L, 80L, 71L, 78L),
  opioid_pca = c("Morphine", "Morphine", "Morphine", "Oxycodone", "Morphine"),
  route      = c("Rectal", "Rectal", "Intravenous; rectal",
                 "Intravenous", "Oral")
)
stopifnot(sum(trials$n) == 353L)

knitr::kable(
  trials |>
    dplyr::rename(
      "Study"       = study,
      "Population"  = population,
      "Indication"  = indication,
      "N"           = n,
      "Opioid PCA"  = opioid_pca,
      "Route"       = route
    ),
  caption = "The five trials in the final opioid-sparing MBMA (Sethi 2024 Tables 1 and 2)."
)
```

Two further PCA-reporting trials from Table 2 were excluded by the authors:
Breivik 1999 (n = 72; only a limited percentage of subjects used PCA in each
arm, and PCA was reported as a percentage rather than in mg) and Riad 2007
(n = 108; a pediatric population).

# Source trace

Every structural equation and every `ini()` value, with its location in the
source.

```{r source-trace}
source_trace <- tibble::tribble(
  ~quantity, ~value, ~source,
  "Eq 1: dY_ij,ose = eo_i,ose + f(Drug_ij, theta) + e_ij,ose",
  "structure",
  "Methods, Model Development, Eq (1) (p148)",

  "Eq 2: f(Drug)_ose = f(acet) + f(diclof) + gamma * f(acet) * f(diclof)",
  "structure",
  "Methods, Model Development, Eq (2) (p148)",

  "Residual: var(e_ij,ose) = sigma_ij^2 / N_ij (observed per-arm squared SE)",
  "data-supplied",
  "Methods, Model Development (p148); no sigma is reported in Table 3",

  "Between-trial placebo response eo_i,ose: unstructured / non-parametric per trial",
  "not reported per trial",
  "Methods, Model Development (p148); Discussion notes only one placebo-controlled trial informs it",

  "e0 (typical placebo response, mg)",
  "64.7 (fixed)",
  "Figure 2 caption: 'assuming a typical placebo response (64.7 mg)'",

  "e_acetaminophen_e0 (mg)",
  "-18.93",
  "Table 3 e.ace = -18.9 [-31.4, -6.43], RSE 29%; Supplementary Table S5 prints the same quantity as -18.93 (-29.47, -8.33)",

  "e_diclofenac_e0 (mg)",
  "-28.41",
  "Table 3 e.dic = -28.4 [-40.7, -16.2], RSE 19%; Supplementary Table S5 prints the same quantity as -28.41 (-39.15, -17.68)",

  "gamma_ad (1/mg)",
  "0.025",
  "Table 3 c = 0.025 [0.0148, 0.0353], RSE 18%; Supplementary Table S4 gives p = 0.028, AIC 115 -> 113",

  "addSd (mg)",
  "0 (fixed)",
  "Not estimated by the source; the residual variance is the observed per-arm squared SE (Methods)",

  "ACETAMINOPHEN / DICLOFENAC arm indicators",
  "0 or 1 per arm",
  "Table 1 Treatments and ROA columns; Supplementary Table S5 row labels ('Systemic acetaminophen', 'Systemic diclofenac')"
)

knitr::kable(
  source_trace |>
    dplyr::rename("Quantity" = quantity, "Value" = value, "Source location" = source),
  caption = "Source trace for every model equation and parameter."
)
```

The `-18.93` / `-28.41` choice deserves a note. Table 3 rounds the two drug
effects to three significant figures (`-18.9`, `-28.4`). For a monotherapy arm
the "model predicted mean opioid PCA difference from placebo" in Supplementary
Table S5 *is* that same parameter, and S5 prints it to four significant figures.
This vignette therefore uses the higher-precision S5 printing; the two agree to
their common precision. (The two tables report different 95% CIs for the same
quantity because Table 3's are asymptotic parameter intervals and Table S5's
come from resampling the final variance-covariance matrix 1000 times.)

# Study-arm design

The model consumes two binary study-arm indicators. Both are 1 on a combination
arm and both are 0 on a placebo arm, so the four treatment conditions the paper
reports are covered by four rows. There is no dose column: the drug effects are
dose-independent (see "Assumptions and deviations").

```{r arms}
arms <- data.frame(
  id            = 1:4,
  time          = 0,
  amt           = 0,
  evid          = 0L,
  arm           = c("Placebo", "Systemic acetaminophen",
                    "Systemic diclofenac",
                    "Systemic acetaminophen + diclofenac"),
  ACETAMINOPHEN = c(0, 1, 0, 1),
  DICLOFENAC    = c(0, 0, 1, 1),
  stringsAsFactors = FALSE
)

knitr::kable(
  arms |>
    dplyr::select(arm, ACETAMINOPHEN, DICLOFENAC) |>
    dplyr::rename("Treatment arm" = arm),
  caption = "Study-arm design matrix for the four treatment conditions."
)
```

# Simulation

```{r simulate}
# NOTE: rxSolve() returns columns literally named `sim` and `ipredSim`. Drop
# them, and never name a data frame `sim` here -- inside a dplyr data mask the
# column would shadow the data frame and `sim$col` fails with
# "$ operator is invalid for atomic vectors".
armSim <- rxode2::rxSolve(mod_typ, events = arms, keep = c("arm")) |>
  as.data.frame() |>
  dplyr::select(-dplyr::any_of(c("sim", "ipredSim"))) |>
  dplyr::mutate(
    arm            = as.character(arm),
    opioid_mg      = Cc,
    placebo_adj_mg = f_drug
  )

placebo_mg <- armSim$opioid_mg[armSim$arm == "Placebo"]
ace_ref_mg <- armSim$opioid_mg[armSim$arm == "Systemic acetaminophen"]

knitr::kable(
  armSim |>
    dplyr::select(arm, opioid_mg, placebo_adj_mg) |>
    dplyr::rename(
      "Treatment arm"                          = arm,
      "Predicted opioid PCA use (mg)"          = opioid_mg,
      "Placebo-adjusted opioid PCA use (mg)"   = placebo_adj_mg
    ),
  digits  = 2,
  caption = "Typical-placebo predictions for the four treatment conditions."
)
```

The placebo arm must return the fixed typical placebo response exactly, and
`f_drug` must be exactly the difference from it:

```{r placebo-identity}
stopifnot(
  abs(placebo_mg - 64.7) < 1e-8,
  all(abs((armSim$opioid_mg - placebo_mg) - armSim$placebo_adj_mg) < 1e-8)
)
```

# Replication of Figure 2 / Supplementary Table S5

Sethi 2024 Figure 2 plots the *placebo-adjusted* opioid PCA use for the two
monotherapies and the combination; Supplementary Table S5 tabulates the same
three numbers with resampling 95% confidence intervals and adds a percent
difference from acetaminophen monotherapy.

Because the paper does not publish the final variance-covariance matrix, the
confidence intervals cannot be resampled here. The intervals drawn below are
**the paper's published intervals**, shown as reference around the model's point
predictions; only the points are simulated.

```{r figure2, fig.height = 3.2}
published <- tibble::tribble(
  ~arm,                                  ~pub_mg,  ~pub_lo,  ~pub_hi, ~pub_pct_vs_ace,
  "Systemic acetaminophen",               -18.93,  -29.47,   -8.33,   NA_real_,
  "Systemic diclofenac",                  -28.41,  -39.15,  -17.68,   -20.70,
  "Systemic acetaminophen + diclofenac",  -33.87,  -44.43,  -22.26,   -32.62
)

fig2 <- armSim |>
  dplyr::filter(arm != "Placebo") |>
  dplyr::inner_join(published, by = "arm") |>
  dplyr::mutate(
    arm = factor(arm, levels = c("Systemic acetaminophen + diclofenac",
                                 "Systemic diclofenac",
                                 "Systemic acetaminophen"))
  )

ggplot(fig2, aes(x = placebo_adj_mg, y = arm)) +
  geom_segment(aes(x = pub_lo, xend = pub_hi, y = arm, yend = arm), linewidth = 0.7) +
  geom_point(size = 4) +
  scale_x_continuous(limits = c(-50, 0)) +
  labs(
    x = "Simulated placebo-adjusted opioid PCA use (mg)",
    y = NULL,
    caption = "Points: this model. Bars: Sethi 2024 Supplementary Table S5 resampling 95% CI."
  ) +
  theme_bw()
```

Replicates Figure 2 of Sethi 2024.

## Side-by-side comparison against the published values

```{r comparison-table}
cmp <- fig2 |>
  dplyr::mutate(
    pct_vs_ace = 100 * (opioid_mg / ace_ref_mg - 1),
    pct_diff_mg = 100 * (placebo_adj_mg - pub_mg) / abs(pub_mg)
  ) |>
  dplyr::arrange(dplyr::desc(placebo_adj_mg)) |>
  dplyr::select(arm, placebo_adj_mg, pub_mg, pct_diff_mg, pct_vs_ace, pub_pct_vs_ace)

knitr::kable(
  cmp |>
    dplyr::rename(
      "Treatment arm"                         = arm,
      "Simulated (mg)"                        = placebo_adj_mg,
      "Sethi 2024 Table S5 (mg)"              = pub_mg,
      "Difference (%)"                        = pct_diff_mg,
      "Simulated vs acetaminophen (%)"        = pct_vs_ace,
      "Table S5 vs acetaminophen (%)"         = pub_pct_vs_ace
    ),
  digits  = 2,
  caption = "Placebo-adjusted opioid PCA use and percent difference from acetaminophen monotherapy, simulated versus published."
)
```

Both quantities are deterministic functions of the printed parameters -- there
is no inter-subject variability and no residual error in this model -- so the
only discrepancy possible is the rounding of the printed parameters. The bounds
below are tight for that reason (see the repository note on vignette
assertions: a tight bound is correct when the two sides differ only by numerical
or rounding error, not by a per-subject physical mechanism).

```{r comparison-gate}
stopifnot(
  # Each monotherapy arm must reproduce its own parameter exactly.
  abs(cmp$pct_diff_mg[cmp$arm == "Systemic acetaminophen"]) < 1e-6,
  abs(cmp$pct_diff_mg[cmp$arm == "Systemic diclofenac"])    < 1e-6,
  # The combination arm is the only place the rounded gamma bites: gamma is
  # printed to two significant figures, which moves the prediction by <0.1%.
  abs(cmp$pct_diff_mg[cmp$arm == "Systemic acetaminophen + diclofenac"]) < 0.5,
  # The percent-difference-from-acetaminophen column is the check that makes
  # the Figure-2-caption placebo response (64.7 mg) load-bearing: it cancels
  # out of the mg column entirely but not out of this one.
  all(abs(cmp$pct_vs_ace - cmp$pub_pct_vs_ace) < 0.5, na.rm = TRUE)
)
```

## The paper's headline claim

> "The final model predicted about 32% less opioid use with the combination than
> acetaminophen monotherapy based on the mean point estimate." (Results)

```{r headline-claim}
ace_mg   <- ace_ref_mg
combo_mg <- armSim$opioid_mg[armSim$arm == "Systemic acetaminophen + diclofenac"]
pct_less <- 100 * (1 - combo_mg / ace_mg)
pct_less

stopifnot(abs(pct_less - 32) < 1.5)
```

# Sub-additivity of the combination

The paper's central structural finding is the *sign* of the interaction
coefficient. A positive `gamma` makes the combination sub-additive: its
opioid-sparing effect is smaller in magnitude than the sum of the two
monotherapy effects, and the shortfall is exactly `gamma * f(ace) * f(diclof)`.

```{r subadditivity}
f_ace   <- armSim$placebo_adj_mg[armSim$arm == "Systemic acetaminophen"]
f_dic   <- armSim$placebo_adj_mg[armSim$arm == "Systemic diclofenac"]
f_combo <- armSim$placebo_adj_mg[armSim$arm == "Systemic acetaminophen + diclofenac"]

subadd <- data.frame(
  quantity = c("f(acetaminophen)", "f(diclofenac)",
               "Additive expectation f(ace) + f(dic)",
               "Modelled combination effect",
               "Sub-additivity shortfall (gamma * f(ace) * f(dic))"),
  mg = c(f_ace, f_dic, f_ace + f_dic, f_combo, f_combo - (f_ace + f_dic))
)

knitr::kable(
  subadd |> dplyr::rename("Quantity" = quantity, "Placebo-adjusted (mg)" = mg),
  digits  = 2,
  caption = "Decomposition of the combination effect (Sethi 2024 Eq 2)."
)

stopifnot(
  # Sub-additive: the combination spares LESS opioid than the sum of the parts.
  f_combo > f_ace + f_dic,
  # ...and the shortfall is exactly the interaction product.
  abs((f_combo - (f_ace + f_dic)) - 0.025 * f_ace * f_dic) < 1e-8,
  # ...but the combination is still better than either monotherapy alone.
  f_combo < f_ace,
  f_combo < f_dic
)
```

The combination is nonetheless predicted to be the best of the three
treatments, and the paper's comparison against diclofenac monotherapy is the
weaker of the two claims: the combination improves on diclofenac by only

```{r combo-vs-dic}
dic_mg <- armSim$opioid_mg[armSim$arm == "Systemic diclofenac"]
100 * (1 - combo_mg / dic_mg)
```

percent, versus ~32% against acetaminophen -- consistent with the paper's
"Differences in the effect size of the combination were less conclusive versus
diclofenac monotherapy."

# Dose independence is a real scope limit

Sethi 2024 could not identify a dose-response from five trials, so the drug
effects are per-arm indicators. Beck 2000 makes this concrete: it contributed
both a 1200 mg and a 2400 mg acetaminophen arm, and the model returns the same
prediction for both.

```{r dose-independence}
beck <- data.frame(
  id            = 1:2,
  time          = 0,
  amt           = 0,
  evid          = 0L,
  arm           = c("Acetaminophen 1200 mg", "Acetaminophen 2400 mg"),
  ACETAMINOPHEN = 1,
  DICLOFENAC    = 0,
  stringsAsFactors = FALSE
)

beck_sim <- rxode2::rxSolve(mod_typ, events = beck, keep = "arm") |>
  as.data.frame()

stopifnot(length(unique(beck_sim$Cc)) == 1L)
beck_sim$Cc
```

Do not use this model to compare acetaminophen or diclofenac doses.

# Predictions for the sixteen fitted trial arms

For completeness, the typical-placebo prediction for each of the sixteen arms
of the five included trials (Sethi 2024 Table 1, Treatments column). These are
*not* comparable to each trial's observed arm means, because the fitted model
gave each trial its own unstructured placebo response and those per-trial
estimates are not published; every row below uses the single typical 64.7 mg
placebo instead.

```{r trial-arms}
trial_arms <- tibble::tribble(
  ~study,             ~treatment,                    ~ACETAMINOPHEN, ~DICLOFENAC,
  "Montgomery 1996",  "ace 1500 mg",                  1, 0,
  "Montgomery 1996",  "ace 1500 mg + dic 100 mg",     1, 1,
  "Montgomery 1996",  "dic 100 mg",                   0, 1,
  "Beck 2000",        "ace 1200 mg",                  1, 0,
  "Beck 2000",        "ace 1200 mg + dic 100 mg",     1, 1,
  "Beck 2000",        "ace 2400 mg",                  1, 0,
  "Siddik 2001",      "ace 2000 mg",                  1, 0,
  "Siddik 2001",      "ace 2000 mg + dic 100 mg",     1, 1,
  "Siddik 2001",      "dic 100 mg",                   0, 1,
  "Siddik 2001",      "placebo",                      0, 0,
  "Hiller 2004",      "ace 2000 mg",                  1, 0,
  "Hiller 2004",      "ace 2000 mg + dic 75 mg",      1, 1,
  "Hiller 2004",      "dic 75 mg",                    0, 1,
  "Munishankar 2008", "ace 1000 mg",                  1, 0,
  "Munishankar 2008", "ace 1000 mg + dic 100 mg",     1, 1,
  "Munishankar 2008", "dic 100 mg",                   0, 1
) |>
  dplyr::mutate(id = dplyr::row_number(), time = 0, amt = 0, evid = 0L)

stopifnot(nrow(trial_arms) == 16L)

arm_sim <- rxode2::rxSolve(mod_typ, events = as.data.frame(trial_arms),
                           keep = c("study", "treatment")) |>
  as.data.frame()

knitr::kable(
  arm_sim |>
    dplyr::select(study, treatment, Cc) |>
    dplyr::rename(
      "Study"                                = study,
      "Treatment arm"                        = treatment,
      "Predicted opioid PCA use (mg)"        = Cc
    ),
  digits  = 2,
  caption = "Typical-placebo predictions for the sixteen arms of the five included trials."
)
```

Only four distinct predicted values appear across the sixteen arms -- one per
treatment condition (placebo, acetaminophen, diclofenac, combination). That is
the dose-independence limitation restated: the model distinguishes treatment
*presence*, not treatment *amount*.

```{r three-values}
stopifnot(length(unique(round(arm_sim$Cc, 6))) == 4L)
sort(unique(round(arm_sim$Cc, 2)))
```

# Assumptions and deviations

* **No PKNCA validation.** This model has no time course, no dose events and no
  concentrations, so non-compartmental analysis does not apply. It is validated
  instead against the paper's own published model predictions (Supplementary
  Table S5 and Figure 2), its headline 32% claim, and internal algebraic
  identities (placebo recovery, sub-additivity decomposition).

* **`e0` is fixed, not estimated.** The source modelled the trial-specific
  placebo response with an unstructured (non-parametric) per-trial model and
  does not report the individual per-trial estimates. The single value used here
  is the typical placebo response of 64.7 mg that the paper itself assumed for
  its Figure 2 and Table S5 simulations, taken from the Figure 2 caption. This
  is a **figure-caption-sourced value**, not a table-sourced one; it is a
  printed number, not a digitized one. It is load-bearing for the
  percent-difference-from-acetaminophen column, and reproduces that published
  column to within 0.1 percentage points, which is strong corroboration that
  64.7 mg is the value the authors used. The paper's own Discussion flags that
  only one of the five trials (Siddik 2001) carried a placebo arm, so this
  placebo response rests on a single trial. Supplementary Figure S3 (a boxplot
  of the reported mean opioid use per study) shows that Siddik 2001 was also the
  *highest*-consumption trial of the five, with arm means spanning roughly
  30-67 mg, while Hiller 2004 sat lowest at roughly 24-32 mg. The typical-placebo
  predictions in this vignette (30.8-64.7 mg) are therefore anchored near the top
  of the observed range across trials, not at its centre. Those bounds are read
  off a figure and are approximate; they are quoted only as context and no
  assertion in this vignette depends on them.

* **Residual error is fixed to zero.** The source defines the within-trial
  residual as `var(e_ij,ose) = sigma_ij^2 / N_ij`, where `sigma_ij` is the
  *observed* standard deviation in arm `j` of trial `i` -- i.e. the residual
  variance is the squared standard error of each arm mean, supplied by the
  data. No residual magnitude is estimated and Table 3 reports no sigma.
  `addSd` is therefore `fixed(0)`; downstream code that wants arm-level noise
  must supply the per-arm standard error explicitly. The per-arm `sigma_ij` and
  `N_ij` values are not tabulated in the paper or its supplement (only the total
  N per trial is, in Table 2), so the weights cannot be reconstructed here.

* **No between-trial random effect.** The between-trial component is the
  unstructured per-trial placebo response, not a Gaussian study-level eta. The
  paper is explicit that it chose an unstructured model because the variability
  "is determined by a substantial number of unexplained factors and thus likely
  to be highly non-Gaussian". There is consequently no omega to encode, and this
  model is *not* suitable for simulating a distribution of trial outcomes.

* **Confidence intervals are not reproducible.** The paper's intervals come from
  resampling 1000 parameter sets from the final variance-covariance matrix,
  which is not published. The intervals shown in the Figure 2 replication are
  the paper's own, overlaid for reference.

* **Drug effects are dose-independent.** Acetaminophen arms spanning
  1000-2400 mg and diclofenac arms spanning 75-100 mg all receive their drug's
  single per-arm effect. This is the source model's own structure, not a
  simplification made here, and it is demonstrated explicitly in the "Dose
  independence" section above.

* **Diclofenac is SYSTEMIC in every fitted arm.** The paper's title concerns
  *topical* diclofenac, but every diclofenac arm in the five included trials
  used oral, rectal or intravenous diclofenac (Table 1, ROA column), and
  Supplementary Table S5 labels its rows "Systemic diclofenac". The move from
  systemic acute-pain evidence to topical diclofenac in chronic osteoarthritis
  is a qualitative extrapolation argued in the paper's Discussion on the grounds
  of shared pain pathways; it is **not** a fitted effect and is not encoded in
  this model.

* **Opioid PCA is pooled across opioids.** Four trials used morphine PCA and one
  (Hiller 2004) used oxycodone PCA, and the endpoint is raw mg with no potency
  normalization. The two opioids are not equipotent, so the pooled effect sizes
  in mg mix scales.

* **Surgical indication is not a covariate.** The paper observed that opioid PCA
  use differs markedly by procedure (cesarean pain requiring more than
  tonsillectomy) but had too few trials to estimate it. This unmodelled
  heterogeneity sits inside the per-trial placebo responses in the original fit
  and is absent altogether from the typical-placebo predictions here.

* **The pain-score endpoint is not modelled.** Sethi 2024 fit no model to pain
  scores; the pain-score material (Supplementary Table S3, Figures S1-S2) is an
  exploratory tabulation of observed between-arm ratios. There is nothing to
  extract.

* **"Corrected publication 2024".** The article's copyright line reads
  "The Author(s) 2024, corrected publication 2024". No correction or erratum is
  indexed for this DOI in Crossref (the record carries no `relation` or
  `update-to`) or in EuropePMC, and no content in the article is flagged as
  corrected. The on-disk PDF is the corrected version of record and its values
  are used as final.
