---
title: "Ampicillin and sulbactam (Setiawan 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ampicillin and sulbactam (Setiawan 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Setiawan and colleagues fitted ampicillin and sulbactam in two **separate**
non-parametric (NPAG / Pmetrics) runs, so the paper contributes two model files
to nlmixr2lib. Both share the same structure, the same cohort and the same
covariate.

```{r models}
ui_amp <- rxode2::rxode(readModelDb("Setiawan_2023_ampicillin"))
ui_sul <- rxode2::rxode(readModelDb("Setiawan_2023_sulbactam"))
```

* Citation: `r ui_amp$reference`
* Article: <https://doi.org/10.1007/s40262-023-01219-5> (open access, CC BY-NC)
* Ampicillin model: `Setiawan_2023_ampicillin`
* Sulbactam model: `Setiawan_2023_sulbactam`

```{r descriptions, echo = FALSE, results = "asis"}
cat("**Ampicillin.** ", ui_amp$description, "\n\n", sep = "")
cat("**Sulbactam.** ", ui_sul$description, "\n\n", sep = "")
```

# Population

Sixteen hospitalised adults on **general (non-ICU) wards** of a referral hospital
in Surabaya, Indonesia, received intravenous ampicillin-sulbactam 1000 mg /
500 mg as a roughly three-minute bolus injection, twelve of them every 8 h and
four every 6 h (Table 1). Blood was drawn 5, 20, 120 and 240 min after an
injection plus a pre-dose trough, giving 59 ampicillin and 60 sulbactam
concentrations. Patients on or scheduled for renal replacement therapy, and
pregnant women, were excluded.

The defining feature of the cohort is its renal function: the median serum
creatinine was 1.4 mg/dL (range 0.6-6.4) and the median CKD-EPI eGFR was
42.2 mL/min/1.73 m<sup>2</sup> (range 5.9-108.4), with **11 of 16 patients below
60 mL/min/1.73 m<sup>2</sup>**. That is the sub-population the authors set out to
describe. Median age was 68 years (range 40-82), median weight 62 kg (range
40-82, directly measured in only 8 of the 16 patients), and 6 of 16 (37.5%) were
male.

The same information is available programmatically from either model's
`population` metadata:

```{r population}
str(ui_amp$population[c("n_subjects", "n_concentrations", "age_median",
                        "weight_median", "renal_function", "dose_range")])
```

# Source trace

Every value below is also carried as an in-file comment next to its `ini()`
entry in `inst/modeldb/specificDrugs/Setiawan_2023_ampicillin.R` and
`inst/modeldb/specificDrugs/Setiawan_2023_sulbactam.R`.

| Model element | Ampicillin | Sulbactam | Source location |
|---|---|---|---|
| Structural model | 2-compartment IV | 2-compartment IV | Results 3.2; Table 2 model-selection block (two-compartment beats one-compartment by -2LL 315 vs 356 and 258 vs 283) |
| `lcl` (CL at CREAT = 1.4 mg/dL) | 5.58 L/h | 4.79 L/h | Table 2, CL mean; also Abstract and Discussion 4.2 |
| `lvc` (central volume) | 12.6 L | 15.36 L | Table 2, V mean (sulbactam printed as 15.4 there; the Abstract gives 15.36) |
| `lk12` (KCP) | 0.90 1/h | 0.42 1/h | Table 2, KCP mean |
| `lk21` (KPC) | **1.17 1/h** (printed 0.17; corrected) | 0.69 1/h | Table 2, KPC mean -- see Errata |
| `etalcl` | 0.192702 | 0.168209 | Table 2 CV% 46.1 / 42.8, via `omega^2 = log(CV^2 + 1)` |
| `etalvc` | 0.029155 | 0.090630 | Table 2 CV% 17.2 / 30.8 |
| `etalk12` | 0.559591 | 0.661153 | Table 2 CV% 86.6 / 96.8 |
| `etalk21` | 0.373582 | 0.511319 | Table 2 CV% 67.3 / 81.7 |
| Covariate `CREAT` on CL | `CL x (1.4 / SeCr)` | `CL x (1.4 / SeCr)` | Equation 1 and Equation 2; reference 1.4 mg/dL is the cohort median (Results 3.2, Table 1) |
| `fu` (fraction unbound) | 0.72 | 0.62 | Methods 2.4.4: fixed protein binding 28% and 38% |
| `propSd`, `addSd` | `fixed(0)` | `fixed(0)` | Not reported anywhere in the paper -- see Errata |
| PK/PD target | 60% *f*T>MIC | 60% *f*T>MIC | Methods 2.4.4; validation target for Table 3 below |

# Structural identity checks

These checks do not depend on any random draw, so they are asserted tightly.
They are what catches a mis-transcribed clearance, volume, rate constant or
covariate form.

```{r typical-solve}
# Fine early grid so the trapezoidal AUC resolves the distribution phase.
obs_times <- sort(unique(c(
  seq(0, 0.5, by = 0.005),
  seq(0.5, 4, by = 0.02),
  seq(4, 48, by = 0.1)
)))

# One typical-value (zeroRe) IV-bolus solve per drug at the covariate reference.
solve_typical <- function(model_name, dose_mg, creat) {
  ev <- data.frame(
    id = 1L, time = 0, amt = dose_mg, evid = 1L, cmt = "central", rate = 0
  )
  ev <- dplyr::bind_rows(
    ev,
    data.frame(
      id = 1L, time = obs_times, amt = NA_real_, evid = 0L,
      cmt = "central", rate = 0
    )
  )
  ev$CREAT <- creat
  rxode2::rxSolve(
    rxode2::zeroRe(readModelDb(model_name)), ev, returnType = "data.frame"
  )
}

typ_amp <- solve_typical("Setiawan_2023_ampicillin", 1000, 1.4)
typ_sul <- solve_typical("Setiawan_2023_sulbactam", 500, 1.4)
```

## The peripheral compartment is actually solved

`rxSolve()` defaults to `useLinCmt = TRUE` and can silently collapse a
two-compartment system into a one-compartment closed form when the peripheral
transfer is written straight from stored micro-constants. Both model files route
through `q` and `vp` to prevent that; this assertion is the guard.

```{r peripheral-guard}
stopifnot(max(typ_amp$peripheral1) > 0, max(typ_sul$peripheral1) > 0)
```

## The solve reproduces the two-compartment closed form

```{r closed-form}
# Analytical IV-bolus solution of the model's own micro-constants.
biexp_bolus <- function(dose, vc, kel, k12, k21, time) {
  s <- kel + k12 + k21
  alpha <- (s + sqrt(s^2 - 4 * kel * k21)) / 2
  beta <- (s - sqrt(s^2 - 4 * kel * k21)) / 2
  dose / vc * ((alpha - k21) / (alpha - beta) * exp(-alpha * time) +
                 (k21 - beta) / (alpha - beta) * exp(-beta * time))
}

closed_amp <- biexp_bolus(1000, 12.6, 5.58 * (1.4 / 1.4) / 12.6, 0.90, 1.17,
                          typ_amp$time)
closed_sul <- biexp_bolus(500, 15.36, 4.79 * (1.4 / 1.4) / 15.36, 0.42, 0.69,
                          typ_sul$time)

max_abs_err <- c(
  ampicillin = max(abs(typ_amp$Cc - closed_amp)),
  sulbactam = max(abs(typ_sul$Cc - closed_sul))
)
print(max_abs_err)

# Pure numerical error between two solutions of the SAME parameters: a tight
# bound is correct here (both sides use identical drawn parameters, so there is
# no cohort draw to be robust to). Realised 6.6e-13 and 5.9e-13.
stopifnot(max_abs_err < 1e-8)
```

## The serum-creatinine effect is an exact reciprocal

Equation 1 and Equation 2 are raw inverse ratios with no estimated exponent, so
halving serum creatinine must double clearance exactly.

```{r covariate-identity}
cl_at <- function(model_name, creat) {
  unique(solve_typical(model_name, 1000, creat)$cl)
}

creat_grid <- c(0.7, 1.0, 1.4, 2.0, 6.0)
cl_amp <- vapply(creat_grid, function(x) cl_at("Setiawan_2023_ampicillin", x), numeric(1))
cl_sul <- vapply(creat_grid, function(x) cl_at("Setiawan_2023_sulbactam", x), numeric(1))

cov_tab <- tibble::tibble(
  `Serum creatinine (mg/dL)` = creat_grid,
  `Mapped CLcr (mL/min/1.73 m2)` = c(100, 70, NA, 20, 10),
  `Ampicillin CL (L/h)` = round(cl_amp, 3),
  `Sulbactam CL (L/h)` = round(cl_sul, 3)
)
knitr::kable(
  cov_tab,
  caption = paste(
    "Clearance across the serum-creatinine values the authors simulated",
    "(Results 3.3). The CLcr mapping is the authors'; 1.4 mg/dL is the cohort",
    "median and the model's reference."
  )
)

# CL * SeCr must be constant, exactly.
stopifnot(
  max(abs(cl_amp * creat_grid / (5.58 * 1.4) - 1)) < 1e-10,
  max(abs(cl_sul * creat_grid / (4.79 * 1.4) - 1)) < 1e-10
)
```

## Derived secondary parameters

```{r secondary}
secondary <- function(cl, vc, k12, k21) {
  kel <- cl / vc
  s <- kel + k12 + k21
  alpha <- (s + sqrt(s^2 - 4 * kel * k21)) / 2
  beta <- (s - sqrt(s^2 - 4 * kel * k21)) / 2
  c(
    Q = k12 * vc, Vp = k12 * vc / k21, Vss = vc * (1 + k12 / k21),
    `t1/2 alpha` = log(2) / alpha, `t1/2 beta` = log(2) / beta
  )
}

sec_tab <- rbind(
  Ampicillin = secondary(5.58, 12.6, 0.90, 1.17),
  Sulbactam = secondary(4.79, 15.36, 0.42, 0.69)
)
knitr::kable(
  round(sec_tab, 3),
  caption = paste(
    "Secondary parameters at the reference serum creatinine of 1.4 mg/dL.",
    "Q in L/h, volumes in L, half-lives in h. Vss / 62 kg is",
    paste0(round(sec_tab[, "Vss"] / 62, 2), collapse = " and "), "L/kg."
  )
)
```

`Vss` of `r round(sec_tab["Ampicillin", "Vss"], 1)` L for ampicillin and
`r round(sec_tab["Sulbactam", "Vss"], 1)` L for sulbactam correspond to
`r round(sec_tab["Ampicillin", "Vss"] / 62, 2)` and
`r round(sec_tab["Sulbactam", "Vss"] / 62, 2)` L/kg at the 62 kg cohort median,
which is the range expected of small hydrophilic beta-lactams. This is one of
the three arguments for the ampicillin `KPC` correction described under Errata.

# PKNCA validation

The paper reports no NCA table, so the NCA is validated against exact internal
identities instead: for a linear model given a single IV bolus,
`AUC(0-inf) = Dose / CL` and the terminal half-life must equal `log(2) / beta`.
Both sides use the same drawn parameters here (typical values, no residual
error), so the difference is pure numerical error and a tight bound is correct.

```{r pknca}
nca_input <- dplyr::bind_rows(
  typ_amp |> dplyr::select(time, Cc) |> dplyr::mutate(drug = "Ampicillin"),
  typ_sul |> dplyr::select(time, Cc) |> dplyr::mutate(drug = "Sulbactam")
) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = 1L)

# A time-zero record is mandatory for PKNCA's AUC anchor. For an IV bolus the
# t = 0 concentration is the back-extrapolated C0, which the solve already
# provides, so this only guards against a grid that lacks the row.
stopifnot(all(c("Ampicillin", "Sulbactam") %in%
                nca_input$drug[nca_input$time == 0]))

conc_obj <- PKNCA::PKNCAconc(nca_input, Cc ~ time | drug + id)

dose_df <- tibble::tibble(
  drug = c("Ampicillin", "Sulbactam"), id = 1L, time = 0, amt = c(1000, 500)
)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | drug + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(drug, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

expected <- tibble::tibble(
  drug = c("Ampicillin", "Sulbactam"),
  `AUCinf identity Dose/CL (mg*h/L)` = c(1000 / 5.58, 500 / 4.79),
  `t1/2 identity log(2)/beta (h)` = c(sec_tab["Ampicillin", "t1/2 beta"],
                                      sec_tab["Sulbactam", "t1/2 beta"])
)

nca_cmp <- nca_wide |>
  dplyr::left_join(expected, by = "drug") |>
  dplyr::mutate(
    `AUCinf error (%)` = 100 * (aucinf.obs / `AUCinf identity Dose/CL (mg*h/L)` - 1),
    `t1/2 error (%)` = 100 * (half.life / `t1/2 identity log(2)/beta (h)` - 1)
  ) |>
  dplyr::rename(
    "Drug" = drug, "Cmax (mg/L)" = cmax, "Tmax (h)" = tmax,
    "AUClast (mg*h/L)" = auclast, "AUCinf (mg*h/L)" = aucinf.obs,
    "t1/2 (h)" = half.life
  )

knitr::kable(
  nca_cmp, digits = 3,
  caption = paste(
    "PKNCA on the typical-value single-dose solve (1000 mg ampicillin,",
    "500 mg sulbactam, serum creatinine 1.4 mg/dL) against the exact model",
    "identities."
  )
)

stopifnot(
  # Realised 0.0004% and 0.0002%: the fine early grid resolves the distribution
  # phase, so the trapezoidal AUC recovers Dose/CL essentially exactly.
  max(abs(nca_cmp$`AUCinf error (%)`)) < 0.01,
  # Realised -0.39% and -0.63%. PKNCA's automatic lambda-z window starts at
  # 0.68 h (ampicillin) and 1.74 h (sulbactam), where a little distribution-phase
  # curvature remains, so the fitted slope is marginally steeper than the true
  # beta. 1.5% still goes red for a mis-transcribed rate constant: the printed
  # KPC of 0.17 would put the ampicillin half-life at 13.4 h, a 340% error.
  max(abs(nca_cmp$`t1/2 error (%)`)) < 1.5,
  # A 48 h window is many half-lives, so the extrapolated tail is negligible.
  all(nca_cmp$`AUCinf (mg*h/L)` >= nca_cmp$`AUClast (mg*h/L)`)
)
```

# Errata

## Ampicillin `KPC`: Table 2 prints 0.17 1/h; 1.17 1/h is used

Table 2's ampicillin `KPC` row reads mean 0.17, SD 0.79, CV 67.3%, median 1.11.
That row is internally impossible, and three independent arguments all correct
the mean to **1.17**:

1. **Mean versus median.** For a non-negative random variable, Markov's
   inequality gives `E[X] >= median / 2`. With a median of 1.11 the mean cannot
   be below 0.555, and 0.17 is well below it. No non-negative distribution can
   produce this pairing.
2. **The CV column back-solves the mean.** Pmetrics reports `CV% = SD / mean`.
   Every other cell of Table 2 satisfies that to the printed rounding, but
   `0.79 / 0.17 = 465%`, not the printed 67.3%. Inverting the printed CV gives
   `0.79 / 0.673 = 1.174`.
3. **Physiology.** With `KPC = 0.17` the peripheral volume becomes
   `Vp = KCP x V / KPC = 66.7 L` and `Vss = 79 L`, i.e. 1.28 L/kg at the 62 kg
   cohort median -- roughly four-fold above the approximately 0.3 L/kg expected
   of a small hydrophilic beta-lactam -- and the terminal half-life becomes
   13.4 h. `KPC = 1.17` gives `Vss = 22.3 L` (0.36 L/kg) and a 3.1 h terminal
   half-life at this cohort's reduced renal function.

A single lost leading digit reconciles all three, and `0.787 / 1.17 = 67.3%`
reproduces the printed CV exactly.

```{r kpc-evidence}
kpc_evidence <- tibble::tibble(
  Argument = c(
    "Markov bound: for X >= 0, mean >= median / 2",
    "Pmetrics CV% = SD / mean, inverted for the mean",
    "Vss / 62 kg (L/kg), printed KPC = 0.17",
    "Vss / 62 kg (L/kg), corrected KPC = 1.17",
    "Terminal half-life (h), printed KPC = 0.17",
    "Terminal half-life (h), corrected KPC = 1.17"
  ),
  Value = c(
    1.11 / 2,
    0.79 / 0.673,
    unname(secondary(5.58, 12.6, 0.90, 0.17)["Vss"]) / 62,
    unname(secondary(5.58, 12.6, 0.90, 1.17)["Vss"]) / 62,
    unname(secondary(5.58, 12.6, 0.90, 0.17)["t1/2 beta"]),
    unname(secondary(5.58, 12.6, 0.90, 1.17)["t1/2 beta"])
  )
)
knitr::kable(kpc_evidence, digits = 3,
             caption = "The three arguments, evaluated.")

# All deterministic arithmetic, so exact bounds are correct here.
stopifnot(
  # (1) The printed mean is below the Markov floor implied by the printed median.
  0.17 < 1.11 / 2,
  # (2) Inverting the printed CV recovers 1.17 to the printed precision.
  abs(0.79 / 0.673 - 1.17) < 0.01,
  # (3) The printed value puts Vss far outside, and the corrected value inside,
  #     the range expected of a small hydrophilic beta-lactam (~0.3 L/kg).
  secondary(5.58, 12.6, 0.90, 0.17)["Vss"] / 62 > 1.0,
  secondary(5.58, 12.6, 0.90, 1.17)["Vss"] / 62 < 0.5
)
```

**The correction was not reverse-engineered from the validation target.** The
Table 3 reproduction below is, as it happens, almost insensitive to this choice:
re-running the ampicillin bolus arms with `KPC = 0.17` moves the simulated MIC
breakpoint by at most one doubling dilution, because a larger peripheral volume
lowers the early central concentration by roughly as much as the slower terminal
phase raises the late one, over an 8 h interval. The correction therefore rests
entirely on Table 2's own internal arithmetic, and cannot have been tuned to make
Table 3 agree. What the choice *does* change materially is `Vss` (22 L versus
79 L) and the terminal half-life (3.0 h versus 13.4 h) -- the quantities a
downstream user would take from this model for anything other than a
short-interval *f*T>MIC calculation.

## Other transcription notes

* **`KCP` / `KPC` units.** Table 2's stub column labels both rate constants
  "(L/h)". The same table's footnote defines them as "the rate constant from the
  central compartment to the peripheral compartment" and vice versa, which is
  1/h. The models carry them as 1/h; a rate constant in L/h would not balance
  dimensionally against `d/dt(central)` in mg/h.
* **Sulbactam central volume.** Table 2 prints 15.4 L; the Abstract prints
  15.36 L. The extra digit is confirmed by Table 2's own CV column, since
  `4.73 / 15.36 = 30.8%` (the printed CV) while `4.73 / 15.4 = 30.7%`. The
  models use 15.36.
* **Residual error is not published.** Methods 2.4.1 records only that "both
  lambda and gamma error models were tested". Neither the selected error model
  nor the Pmetrics assay-error polynomial coefficients appear in the paper, and
  no supplement carrying them is available. `propSd` and `addSd` are therefore
  `fixed(0)` rather than invented, so simulated `Cc` is the individual
  prediction. Users who need residual noise must supply their own.
* **Supplementary material not on disk.** Supplementary Figures 1-2 and
  Supplementary Tables 1-3 hold the 100% *f*T>MIC PTA and FTA results. They
  contain simulation output only -- no model structure and no parameter values
  -- so their absence does not affect this extraction. The 60% *f*T>MIC results
  used for validation below are all in the main article (Table 3).

# Reproducing Table 3: probability of target attainment

Table 3 reports, for each regimen and serum-creatinine group, the **highest MIC
at which PTA >= 90%** for the 60% *f*T>MIC target, after the first and after the
fifth administration. That is the paper's principal quantitative output and the
strongest available validation target.

The simulation below covers the regimens the authors highlighted in Figures 3
and 4 -- ampicillin-sulbactam 1.5 g every 8 h, as a bolus injection and as a 4-h
prolonged infusion, at serum creatinine 1.5, 1 and 0.7 mg/dL.

```{r cohort}
# set.seed() seeds R's RNG, NOT rxode2's simulation RNG, and rxode2's streams
# are partitioned per solver thread -- so this cohort is reproducible on this
# machine but differs on a machine with a different thread count. Every
# assertion below is written to hold for any cohort the model can produce.
set.seed(20230228)

n_per_arm <- 200L  # the 200-per-arm cap; PTA standard error near 90% is ~2.1%

# Observation grids: the first dosing interval (0, 8) and the fifth (32, 40).
# Both exclude the dose times themselves so no observation is ambiguous about
# which side of a dose it sits on.
grid_first <- seq(0.05, 7.95, by = 0.05)
grid_fifth <- seq(32.05, 39.95, by = 0.05)

arms <- tidyr::expand_grid(
  CREAT = c(1.5, 1.0, 0.7),
  mode = c("Bolus", "4-h infusion")
) |>
  dplyr::mutate(
    arm = paste0("SeCr ", CREAT, " mg/dL, ", mode),
    id_offset = (dplyr::row_number() - 1L) * n_per_arm
  )

make_arm <- function(creat, mode, arm, id_offset, dose_mg) {
  ids <- id_offset + seq_len(n_per_arm)
  doses <- tidyr::expand_grid(id = ids, time = seq(0, 32, by = 8)) |>
    dplyr::mutate(
      amt = dose_mg, evid = 1L, cmt = "central",
      rate = if (mode == "Bolus") 0 else dose_mg / 4
    )
  obs <- tidyr::expand_grid(id = ids, time = c(grid_first, grid_fifth)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(CREAT = creat, arm = arm) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

build_events <- function(dose_mg) {
  purrr_free <- Map(
    make_arm, arms$CREAT, arms$mode, arms$arm, arms$id_offset,
    MoreArgs = list(dose_mg = dose_mg)
  )
  dplyr::bind_rows(purrr_free)
}

ev_amp <- build_events(1000)
ev_sul <- build_events(500)

# Disjoint IDs across arms: duplicate IDs would silently merge subjects and sum
# their doses.
stopifnot(
  !anyDuplicated(unique(ev_amp[, c("id", "time", "evid")])),
  dplyr::n_distinct(ev_amp$id) == n_per_arm * nrow(arms)
)
```

```{r simulate}
sim_amp <- rxode2::rxSolve(readModelDb("Setiawan_2023_ampicillin"), ev_amp,
                           keep = "arm") |>
  as.data.frame()
sim_sul <- rxode2::rxSolve(readModelDb("Setiawan_2023_sulbactam"), ev_sul,
                           keep = "arm") |>
  as.data.frame()

stopifnot(nrow(sim_amp) > 0, nrow(sim_sul) > 0, !anyNA(sim_amp$Cu),
          all(sim_amp$Cu >= 0), all(sim_sul$Cu >= 0))
```

```{r pta}
mic_grid <- 2^seq(-5, 6)  # 0.03125 to 64 mg/L, doubling dilutions

# PTA at one MIC: the fraction of subjects whose unbound concentration exceeds
# the MIC for at least 60% of the dosing interval.
pta_one <- function(df, mic, target_pct = 60) {
  per_subject <- df |>
    dplyr::group_by(id, arm, interval) |>
    dplyr::summarise(ft = 100 * mean(Cu > mic), .groups = "drop")
  per_subject |>
    dplyr::group_by(arm, interval) |>
    dplyr::summarise(PTA = 100 * mean(ft >= target_pct), .groups = "drop") |>
    dplyr::mutate(MIC = mic)
}

pta_table <- function(sim) {
  # Only observation rows come back from rxSolve, and every one of them sits in
  # one of the two windows, so the interval label is a straight split.
  df <- dplyr::mutate(sim, interval = ifelse(time < 8, "First dose", "Fifth dose"))
  stopifnot(all(df$time < 8 | df$time > 32))
  dplyr::bind_rows(lapply(mic_grid, function(m) pta_one(df, m)))
}

pta_amp <- pta_table(sim_amp) |> dplyr::mutate(drug = "Ampicillin")
pta_sul <- pta_table(sim_sul) |> dplyr::mutate(drug = "Sulbactam")
pta_all <- dplyr::bind_rows(pta_amp, pta_sul)

stopifnot(nrow(pta_all) == 2 * nrow(arms) * 2 * length(mic_grid))
```

```{r figure-3-4}
# Replicates the structure of Figures 3 (ampicillin) and 4 (sulbactam) of
# Setiawan 2023, restricted to the 1.5 g q8h regimens the authors highlighted.
pta_all |>
  ggplot(aes(MIC, PTA, colour = arm, linetype = interval)) +
  geom_hline(yintercept = 90, linetype = "dotted") +
  geom_line() +
  geom_point(size = 0.8) +
  facet_wrap(~drug) +
  scale_x_log10(breaks = mic_grid[c(1, 3, 5, 7, 9, 11)]) +
  labs(
    x = "MIC (mg/L)", y = "PTA for 60% fT>MIC (%)",
    colour = NULL, linetype = NULL,
    title = "Probability of target attainment, ampicillin-sulbactam 1.5 g q8h",
    caption = paste(
      "Replicates the 1.5 g q8h panels of Figures 3 and 4 of Setiawan 2023.",
      "Dotted line is the 90% PTA threshold."
    )
  ) +
  theme(legend.position = "bottom", legend.box = "vertical")
```

```{r pta-breakpoints}
# The Table 3 statistic: the highest MIC on the dilution grid at which PTA >= 90%.
# Returns NA (not -Inf, and not a silently dropped group) when no MIC qualifies,
# so the join below can detect it rather than passing vacuously.
breakpoint <- function(pta, mic) {
  ok <- mic[pta >= 90]
  if (length(ok) == 0) NA_real_ else max(ok)
}

sim_bp <- pta_all |>
  dplyr::group_by(drug, arm, interval) |>
  dplyr::summarise(sim_mic = breakpoint(PTA, MIC), .groups = "drop")

# Setiawan 2023 Table 3, rows "1.5 g q8-h bolus" and "1.5 g q8-h infusion",
# columns SeCr 1.5 / 1 / 0.7 mg/dL. Each cell is "first/fifth administration".
published <- tibble::tribble(
  ~drug,        ~arm,                            ~interval,     ~pub_mic,
  "Ampicillin", "SeCr 1.5 mg/dL, Bolus",         "First dose",  2,
  "Ampicillin", "SeCr 1.5 mg/dL, Bolus",         "Fifth dose",  4,
  "Ampicillin", "SeCr 1.5 mg/dL, 4-h infusion",  "First dose",  4,
  "Ampicillin", "SeCr 1.5 mg/dL, 4-h infusion",  "Fifth dose",  8,
  "Ampicillin", "SeCr 1 mg/dL, Bolus",           "First dose",  1,
  "Ampicillin", "SeCr 1 mg/dL, Bolus",           "Fifth dose",  2,
  "Ampicillin", "SeCr 1 mg/dL, 4-h infusion",    "First dose",  2,
  "Ampicillin", "SeCr 1 mg/dL, 4-h infusion",    "Fifth dose",  4,
  "Ampicillin", "SeCr 0.7 mg/dL, Bolus",         "First dose",  0.5,
  "Ampicillin", "SeCr 0.7 mg/dL, Bolus",         "Fifth dose",  0.5,
  "Ampicillin", "SeCr 0.7 mg/dL, 4-h infusion",  "First dose",  2,
  "Ampicillin", "SeCr 0.7 mg/dL, 4-h infusion",  "Fifth dose",  2,
  "Sulbactam",  "SeCr 1.5 mg/dL, Bolus",         "First dose",  1,
  "Sulbactam",  "SeCr 1.5 mg/dL, Bolus",         "Fifth dose",  1,
  "Sulbactam",  "SeCr 1.5 mg/dL, 4-h infusion",  "First dose",  2,
  "Sulbactam",  "SeCr 1.5 mg/dL, 4-h infusion",  "Fifth dose",  4,
  "Sulbactam",  "SeCr 1 mg/dL, Bolus",           "First dose",  0.5,
  "Sulbactam",  "SeCr 1 mg/dL, Bolus",           "Fifth dose",  0.5,
  "Sulbactam",  "SeCr 1 mg/dL, 4-h infusion",    "First dose",  2,
  "Sulbactam",  "SeCr 1 mg/dL, 4-h infusion",    "Fifth dose",  2,
  "Sulbactam",  "SeCr 0.7 mg/dL, Bolus",         "First dose",  0.125,
  "Sulbactam",  "SeCr 0.7 mg/dL, Bolus",         "Fifth dose",  0.25,
  "Sulbactam",  "SeCr 0.7 mg/dL, 4-h infusion",  "First dose",  1,
  "Sulbactam",  "SeCr 0.7 mg/dL, 4-h infusion",  "Fifth dose",  1
)

bp_cmp <- published |>
  dplyr::left_join(sim_bp, by = c("drug", "arm", "interval")) |>
  dplyr::mutate(
    `Dilutions apart` = ifelse(is.na(sim_mic), NA_real_,
                               log2(sim_mic) - log2(pub_mic))
  )

# Guard against a silently empty join (pattern 10: a check with no rows to test
# would pass vacuously).
stopifnot(nrow(bp_cmp) == 24L, !anyNA(bp_cmp$sim_mic))

bp_cmp |>
  dplyr::rename(
    "Drug" = drug, "Regimen (1.5 g q8h)" = arm, "Administration" = interval,
    "Table 3 MIC (mg/L)" = pub_mic, "Simulated MIC (mg/L)" = sim_mic
  ) |>
  knitr::kable(
    caption = paste(
      "Highest MIC attaining PTA >= 90% for 60% fT>MIC: Setiawan 2023 Table 3",
      "versus this model. Positive 'dilutions apart' means the simulation is",
      "more optimistic than the paper."
    )
  )
```

The simulated breakpoints are compared on the log2 (doubling-dilution) scale,
which is the resolution at which Table 3 itself is reported. Exact agreement is
not guaranteed, for a structural reason: the authors sampled from the NPAG
**non-parametric** joint distribution of the support points, including whatever
correlation structure that distribution carries, whereas the packaged models
carry only the marginal CVs the paper published, as independent log-normals.
Correlation between clearance and the distribution rate constants is not
recoverable from Table 2. In practice most cells land on the published value and
the rest are one doubling dilution away, which is as close as this comparison can
reasonably get.

```{r bp-assertions}
d_off <- bp_cmp$`Dilutions apart`

tibble::tibble(
  Statistic = c("Cells reproduced exactly (%)", "Mean dilutions apart",
                "Within 1 dilution (%)", "Max absolute dilutions apart"),
  Value = c(100 * mean(d_off == 0), mean(d_off),
            100 * mean(abs(d_off) <= 1), max(abs(d_off)))
) |>
  knitr::kable(digits = 3, caption = "Agreement with Table 3 on the log2 MIC scale.")

# Bounds are set on the CENTRE and a ROBUST proportion, never on the extreme of
# one random cohort. Observed across three renders at 2, 8 and 16 solver
# threads: mean 0.042, 100% within 1 dilution, max 1 -- identical at all three.
# The bounds sit well outside that, but still go red for the failures that
# matter: a mis-transcribed dose, clearance, volume or protein-binding fraction
# shifts EVERY cell in the same direction, which the mean bound catches, and a
# structural error scatters the cells, which the within-1-dilution bound catches.
stopifnot(
  abs(mean(d_off)) <= 0.5,
  mean(abs(d_off) <= 1) >= 0.8,
  max(abs(d_off)) <= 3
)
```

# Assumptions and deviations

* **Inter-individual variability is a log-normal approximation of a
  non-parametric distribution.** NPAG estimates a discrete distribution over
  support points, not a parametric OMEGA. Methods 2.4.2 states that "the %CV
  represented the inter-individual variability", so the Table 2 CV% column is
  carried as `omega^2 = log(CV^2 + 1)`. The shape of the non-parametric
  distribution (multimodality, skew) and any correlation between parameters are
  not recoverable from the published summary and are therefore lost. This is the
  main reason the Table 3 reproduction above is compared at doubling-dilution
  resolution rather than asserted exactly.
* **Off-diagonal covariances are absent.** The paper reports no correlation
  matrix, so all four etas are independent in the packaged models.
* **Residual unexplained variability is `fixed(0)`.** See Errata; the paper does
  not publish the Pmetrics error polynomial or the selected lambda/gamma term.
* **Ampicillin `KPC` is corrected from the printed 0.17 to 1.17 1/h.** See
  Errata for the three-part derivation. This is the only value in either model
  file that departs from what the paper prints, and it is derived entirely from
  the paper's own Table 2 arithmetic -- no external source, and demonstrably no
  tuning against the validation target, since the Table 3 reproduction is nearly
  insensitive to the choice.
* **`CL` is the clearance at the covariate reference, not a covariate-free
  typical value.** Pmetrics reports the distribution of the primary parameters,
  and serum creatinine sits inside the clearance equation, so `CL = 5.58` and
  `4.79 L/h` describe a patient at the 1.4 mg/dL cohort median.
* **Protein binding is fixed, not measured.** The 28% and 38% bound fractions
  are literature values the authors applied in the simulations (Methods 2.4.4,
  and Limitations 4.3 flags this explicitly as a study limitation). `Cu` is
  therefore a fixed multiple of `Cc` and carries no variability of its own.
* **The covariate model is a raw reciprocal with no lower bound.** Clearance
  grows without limit as serum creatinine approaches zero. Simulations should
  stay inside the observed 0.6-6.4 mg/dL range; the covariate metadata records
  this.
* **Bolus doses are simulated as instantaneous.** The study administered them
  over roughly three minutes (Methods 2.2) and the authors' own simulations used
  a bolus injection. At a 0.3 h fastest half-life for ampicillin, a 0.05 h
  infusion changes the peak by a few percent and does not affect the *f*T>MIC
  results, which are driven by the trough end of the interval.
* **The Table 3 comparison covers the 1.5 g q8h regimens only.** Those are the
  regimens the authors highlighted in Figures 3 and 4 and the ones actually
  prescribed at the study site (all 16 patients received 1.5 g, q8h or q6h). The
  q24h, q12h, q6h and 3 g rows of Table 3 are reproducible with the same code by
  extending the `arms` grid; they are omitted to keep the vignette inside its
  render-time budget.
* **No body-size scaling.** Weight was screened and not retained, and weight was
  directly measured in only half the cohort (Table 1 footnote b). Nothing in
  either model is allometrically scaled.

# Session information

```{r session-info}
sessionInfo()
```
