---
title: "Shikimic acid PK and antioxidant PD in broilers (Shen 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Shikimic acid PK and antioxidant PD in broilers (Shen 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Shikimic acid (SA) is a highly polar polyhydroxy cyclic organic acid, best known
as the synthetic precursor of oseltamivir but of independent interest as an
"indirect" antioxidant that acts by activating the Nrf2/Keap1 pathway rather
than by scavenging radicals directly. Shen 2026 is the first report of its
pharmacokinetics in poultry.

Sixteen healthy 5-week-old yellow-feathered broilers were randomised to a single
50 mg/kg dose of SA given either intravenously (wing vein) or by oral gavage,
n = 8 per group. Plasma SA was measured by HPLC-MS/MS. In the **oral group
only**, serum total antioxidant capacity (T-AOC), superoxide dismutase (SOD)
activity and malondialdehyde (MDA) were measured at 0, 1, 2, 4, 6, 8 and 12 h.

T-AOC and SOD rose and MDA fell, but all three peaked well after the plasma
peak, producing hysteresis loops (counter-clockwise for T-AOC and SOD, clockwise
for MDA). The authors therefore linked plasma to an **effect compartment** and
drove an **indirect-response** turnover model from the effect-site
concentration, fitting each bird individually in Phoenix WinNonlin. Each
biomarker was fitted separately, with its own `ke0`, `S`, `kin` and `kout`
(Table 5), so the paper contributes **three** model files that share one
plasma description:

```{r models}
sa_models <- c(
  "Shen_2026_shikimicAcid_broiler_taoc",  # T-AOC: SA stimulates production (kin)
  "Shen_2026_shikimicAcid_broiler_sod",   # SOD:   SA stimulates production (kin)
  "Shen_2026_shikimicAcid_broiler_mda"    # MDA:   SA stimulates loss       (kout)
)
# Parse each model once; the parsed object exposes the file-level metadata
# (reference, description, population) as well as being solvable.
mods <- lapply(sa_models, function(x) rxode2::rxode(readModelDb(x)))
names(mods) <- sa_models
vapply(mods, function(m) paste(m$state, collapse = ", "), character(1))
```

* Citation: `r mods[[1]]$reference`
* Article: [Poultry Science 2026;105:106959](https://doi.org/10.1016/j.psj.2026.106959)

## Population

```{r population}
pop <- mods[[1]]$population
data.frame(
  Field = c("Species", "Subjects (oral arm)", "Age", "Body weight", "Sex",
            "Health status", "Dose", "Region"),
  Value = c(pop$species, as.character(pop$n_subjects), pop$age_median,
            pop$weight_range, pop$sex, pop$disease_state, pop$dose_range,
            pop$regions)
) |>
  knitr::kable()
```

Every parameter is fixed at the published mean of the eight individual fits.
Shen 2026 reports only means and SDs of individual estimates, never an omega
matrix, so there is **no between-subject variability to simulate** and the model
is deterministic. A single typical-value profile is therefore the complete
prediction; no virtual cohort is drawn.

# Source trace

```{r source-trace}
tribble(
  ~Quantity, ~`Source location`, ~Value,
  "Dose, route, species, n",        "Methods, 'Animals' / 'Dosing and sample collection'", "50 mg/kg PO gavage, broiler, n = 8",
  "dCe/dt = ke0 * (Cp - Ce)",       "Methods, 'Statistical analysis and PK-PD modeling'",  "effect-compartment link",
  "dE/dt = kin*(1+S*Ce) - kout*E",  "Methods, same section (T-AOC, SOD)",                  "stimulation of production",
  "dE/dt = kin - kout*(1+S*Ce)*E",  "Methods, same section (MDA)",                         "stimulation of loss",
  "E0 = kin/kout",                  "Methods, same section",                               "stated explicitly",
  "T-AOC ke0, S, kin, kout",        "Table 5, T-AOC row",                                  "0.44, 0.01, 0.37, 0.14",
  "SOD ke0, S, kin, kout",          "Table 5, SOD row",                                    "0.89, 0.04, 8.98, 0.31",
  "MDA ke0, S, kin, kout",          "Table 5, MDA row",                                    "1.06, 0.01, 0.84, 0.24",
  "CL/F = Dose/AUC_po",             "Table 4 (AUC 132.04 h*ug/mL)",                        "378.67 mL/h/kg",
  "V/F = Vz/F",                     "Table 3 (Vz 0.35 L/kg), Table 4 (F 40.37%)",          "866.98 mL/kg",
  "ka",                             "RECONSTRUCTED - not published (see Errata)",          "1.2378 /h",
  "Residual error structures",      "Methods, 'Statistical analysis and PK-PD modeling'",  "multiplicative PK, additive PD",
  "Residual error magnitudes",      "NOT reported (only initial SDs 0.1 / 1.0)",           "fixed at 0"
) |>
  knitr::kable()
```

# Reference data digitised from the published figures

Shen 2026 tabulates the non-compartmental results (Tables 3 and 4) but plots the
observed profiles only. The observed values below were digitised from the
published figures by locating the marker centroids and bar edges in pixel space
and mapping them through the axis-tick calibration; recovered sample times
matched the nominal sampling schedule to within 0.02 h, and the digitised
trapezoidal AUC(0-12) of 137.3 h*ug/mL sits within 4% of the published
AUC(0-inf) of 132.04 h*ug/mL.

```{r digitised-data}
# Figure 2B, main panel: mean (n = 8) plasma SA after 50 mg/kg PO.
obs_pk <- data.frame(
  time = c(0, 0.25, 0.5, 0.75, 1, 2, 3, 4, 6, 8, 12),
  conc = c(0, 1.68, 3.03, 4.30, 8.25, 23.03, 23.54, 27.85, 12.86, 4.65, 0.92)
)

# Figure 3A/B/C: mean serum biomarker bar heights.
obs_pd <- data.frame(
  time = c(0, 1, 2, 4, 6, 8, 12),
  taoc = c(2.675, 2.633, 2.414, 2.768, 2.928, 3.089, 2.819),  # umol/mL
  sod  = c(30.127, 31.392, 34.051, 37.722, 44.430, 45.316, 37.215),  # U/mL
  mda  = c(3.534, 3.449, 3.196, 2.975, 2.985, 2.985, 3.207)   # nmol/mL
)

# Tables 3 and 4, transcribed (mean of the individual NCA fits).
nca_published <- data.frame(
  treatment  = "SA 50 mg/kg PO",
  cmax       = 32.68,
  tmax       = 3.38,
  aucinf.obs = 132.04,
  half.life  = 1.80
)
```

## The baseline identity `E0 = kin/kout` certifies Table 5

Shen 2026 states `E0 = kin/kout` but never prints E0, and Table 5 carries no
units at all. The 0 h bar of each Figure 3 panel *is* that E0, and the two
numbers are published independently, so the ratio is a free consistency check on
both the transcription and the units (`kin` carries the biomarker's units per
hour, so it can only agree if the units are right).

```{r baseline-check}
e0 <- vapply(mods, function(m) {
  p <- m$theta
  unname(exp(p[["lkin"]]) / exp(p[["lkout"]]))
}, numeric(1))

baseline_check <- data.frame(
  Biomarker  = c("T-AOC (umol/mL)", "SOD (U/mL)", "MDA (nmol/mL)"),
  `kin/kout` = round(unname(e0), 3),
  `Figure 3 0 h bar` = unname(unlist(obs_pd[obs_pd$time == 0, c("taoc", "sod", "mda")])),
  check.names = FALSE
) |>
  mutate(`% diff` = round(100 * (`kin/kout` - `Figure 3 0 h bar`) / `Figure 3 0 h bar`, 1))
knitr::kable(baseline_check)

# Every baseline must land within the digitisation error of the observed bar.
stopifnot(all(abs(baseline_check$`% diff`) < 5))
```

# Simulation

The model is deterministic (no IIV, residual SDs fixed at zero), so a single
profile is simulated per biomarker model. Amounts are per kg body weight, so the
50 mg/kg dose enters as 50,000 ug/kg and concentrations come out in ug/mL.

```{r simulate}
dose_ug_per_kg <- 50000  # 50 mg/kg

# Each model carries TWO endpoints (Cc and its biomarker), so observation rows
# must name an endpoint via dvid rather than sitting on a bare compartment;
# rxode2 then returns every state and algebraic observable as a column. The
# grid is refined around the absorption peak so the NCA Tmax reflects the model
# rather than the spacing of the observation grid.
grid <- sort(unique(c(obs_pk$time, obs_pd$time,
                      seq(0, 24, by = 0.05), seq(0.8, 2.0, by = 0.01))))

ev <- as.data.frame(
  rxode2::et(amt = dose_ug_per_kg, cmt = "depot") |>
    rxode2::et(grid)
)
ev$dvid <- ifelse(ev$evid == 0, 1L, NA_integer_)
ev$id <- 1L

sims <- lapply(mods, function(m) {
  rxode2::rxSolve(m, ev, returnType = "data.frame", addDosing = FALSE)
})

stopifnot(
  all(vapply(sims, nrow, integer(1)) > 0),
  all(vapply(sims, function(s) !anyNA(s$Cc), logical(1)))
)

# The plasma layer is shared by all three files; confirm it really is identical
# before using the T-AOC solve as the single PK reference.
stopifnot(
  max(abs(sims[[1]]$Cc - sims[[2]]$Cc)) < 1e-8,
  max(abs(sims[[1]]$Cc - sims[[3]]$Cc)) < 1e-8
)

sim_pk <- sims[[1]][, c("time", "Cc")]
sim_pd <- data.frame(
  time = sims[[1]]$time,
  taoc = sims[["Shen_2026_shikimicAcid_broiler_taoc"]]$taoc,
  sod  = sims[["Shen_2026_shikimicAcid_broiler_sod"]]$sod,
  mda  = sims[["Shen_2026_shikimicAcid_broiler_mda"]]$mda
)
```

## Replicating Figure 2B: plasma SA after 50 mg/kg PO

```{r fig2b, fig.cap = "Replicates Figure 2B of Shen 2026. Line = model, points = digitised observed mean (n = 8)."}
ggplot(sim_pk, aes(time, Cc)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = obs_pk, aes(time, conc), size = 2.2, colour = "firebrick") +
  coord_cartesian(xlim = c(0, 12)) +
  labs(x = "Time (h)", y = "Plasma shikimic acid (ug/mL)",
       title = "Shen 2026 Figure 2B: oral 50 mg/kg",
       subtitle = "Model reproduces Cmax and AUC; the peak is reached too early (see Errata)") +
  theme_bw()
```

The reconstructed plasma layer hits the published Cmax and AUC exactly but peaks
at 1.3 h rather than the reported 3.38 h. That gap is **structural, not a fitting
failure**: it is proven unavoidable in the Errata below.

## Replicating Figure 3: serum antioxidant biomarkers

```{r fig3, fig.width = 8, fig.height = 3.2, fig.cap = "Replicates Figure 3 of Shen 2026. Lines = model, points = digitised observed means (n = 8)."}
lbl <- c(taoc = "T-AOC (umol/mL)", sod = "SOD (U/mL)", mda = "MDA (nmol/mL)")

sim_long <- sim_pd |>
  pivot_longer(-time, names_to = "biomarker", values_to = "value") |>
  mutate(biomarker = factor(lbl[biomarker], levels = lbl))
obs_long <- obs_pd |>
  pivot_longer(-time, names_to = "biomarker", values_to = "value") |>
  mutate(biomarker = factor(lbl[biomarker], levels = lbl))

ggplot(sim_long, aes(time, value)) +
  geom_line(linewidth = 0.8) +
  geom_point(data = obs_long, size = 2.2, colour = "firebrick") +
  facet_wrap(~biomarker, scales = "free_y") +
  coord_cartesian(xlim = c(0, 12)) +
  labs(x = "Time (h)", y = NULL, title = "Shen 2026 Figure 3") +
  theme_bw()
```

```{r fig3-numeric}
# Interpolate the simulation onto the observation times rather than matching
# grid values: seq(0, 24, by = 0.05) does not land exactly on integers in
# floating point, so a `%in%` join would silently drop rows.
pd_cmp <- bind_rows(lapply(c("taoc", "sod", "mda"), function(b) {
  data.frame(
    biomarker = b,
    time      = obs_pd$time,
    observed  = obs_pd[[b]],
    simulated = stats::approx(sim_pd$time, sim_pd[[b]], xout = obs_pd$time)$y
  )
})) |>
  mutate(`% diff` = round(100 * (simulated - observed) / observed, 1),
         across(c(observed, simulated), ~ round(.x, 3)))
stopifnot(!anyNA(pd_cmp$simulated))

pd_cmp |>
  rename("Biomarker" = biomarker, "Time (h)" = time,
         "Observed (Fig 3)" = observed, "Simulated" = simulated) |>
  knitr::kable()

# The model must reproduce the DIRECTION of every biomarker's response and stay
# close to the observed course. Both sides are deterministic here -- there is no
# random draw anywhere in this vignette -- so these bounds are exact and stable.
peak <- pd_cmp |>
  group_by(biomarker) |>
  summarise(base = observed[time == 0],
            obs_extreme = observed[which.max(abs(observed - observed[time == 0]))],
            sim_extreme = simulated[which.max(abs(simulated - simulated[time == 0]))],
            .groups = "drop") |>
  mutate(obs_dir = sign(obs_extreme - base), sim_dir = sign(sim_extreme - base))

stopifnot(
  # T-AOC and SOD rise, MDA falls -- in the model exactly as in the data.
  all(peak$obs_dir == peak$sim_dir),
  peak$sim_dir[peak$biomarker == "taoc"] > 0,
  peak$sim_dir[peak$biomarker == "sod"] > 0,
  peak$sim_dir[peak$biomarker == "mda"] < 0,
  # Every simulated point is within 25% of the digitised observation.
  max(abs(pd_cmp$`% diff`)) < 25,
  # and the typical agreement is much tighter than that.
  median(abs(pd_cmp$`% diff`)) < 6
)
```

## Replicating Figure 4: concentration-effect hysteresis

The paper's headline qualitative finding is that the biomarker response lags
plasma concentration, giving counter-clockwise loops for T-AOC and SOD and a
clockwise loop for MDA.

```{r fig4, fig.width = 8, fig.height = 3.2, fig.cap = "Replicates Figure 4 of Shen 2026: concentration-effect loops over 0-12 h, traced chronologically."}
hyst <- sim_pd |>
  mutate(Cc = sim_pk$Cc) |>
  filter(time <= 12) |>
  pivot_longer(c(taoc, sod, mda), names_to = "biomarker", values_to = "value") |>
  mutate(biomarker = factor(lbl[biomarker], levels = lbl))

ggplot(hyst, aes(Cc, value, colour = time)) +
  geom_path(linewidth = 0.8) +
  facet_wrap(~biomarker, scales = "free_y") +
  scale_colour_viridis_c(name = "Time (h)") +
  labs(x = "Plasma shikimic acid (ug/mL)", y = NULL,
       title = "Shen 2026 Figure 4: hysteresis loops") +
  theme_bw()
```

```{r fig4-check}
# A hysteresis loop encloses area; a direct (non-delayed) effect would collapse
# onto a single curve. Use the shoelace formula on the closed 0-12 h path and
# check the SIGN: positive = counter-clockwise, negative = clockwise, both
# computed on the biomarker scaled to its own baseline so the three are
# comparable.
loop_area <- function(x, y) {
  y <- y / y[1]
  x <- c(x, x[1]); y <- c(y, y[1])
  n <- length(x)
  sum(x[-n] * y[-1] - x[-1] * y[-n]) / 2
}
areas <- vapply(c("taoc", "sod", "mda"), function(b) {
  d <- sim_pd[sim_pd$time <= 12, ]
  loop_area(sim_pk$Cc[sim_pk$time <= 12], d[[b]])
}, numeric(1))
round(areas, 4)

stopifnot(
  # T-AOC and SOD: counter-clockwise (as Shen 2026 Fig 4A, 4B).
  areas[["taoc"]] > 0, areas[["sod"]] > 0,
  # MDA: clockwise (as Shen 2026 Fig 4C).
  areas[["mda"]] < 0
)
```

# PKNCA validation against the published non-compartmental analysis

```{r pknca}
conc_df <- sim_pk |>
  filter(!is.na(Cc)) |>
  mutate(id = 1L, treatment = "SA 50 mg/kg PO") |>
  select(id, time, Cc, treatment)

dose_df <- data.frame(id = 1L, time = 0, amt = dose_ug_per_kg,
                      treatment = "SA 50 mg/kg PO")

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "ug")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
sim_nca <- as.data.frame(res)
```

```{r nca-table}
tbl <- ncaComparisonTable(
  simulated = sim_nca,
  reference = nca_published,
  by = "treatment",
  units = c(cmax = "ug/mL", tmax = "h", aucinf.obs = "h*ug/mL", half.life = "h")
)
knitr::kable(tbl)
attr(tbl, "footnote")
```

`Cmax` and `AUC(0-inf)` match the published values essentially exactly, because
the plasma layer was constructed to reproduce them. The terminal half-life,
which was **not** used in the construction, comes out at 1.60 h against the
1.80 h reported for the oral arm and the 1.61 h reported for the intravenous arm
(Table 3) -- an independent check that the disposition is right. `Tmax` is the
one parameter the structure cannot reach; see the Errata.

```{r nca-assert}
got <- setNames(sim_nca$PPORRES, sim_nca$PPTESTCD)

# Cmax and AUC are matched by construction, so require them to be tight.
stopifnot(
  abs(got[["cmax"]] / 32.68 - 1) < 0.01,
  abs(got[["aucinf.obs"]] / 132.04 - 1) < 0.01
)
# Half-life was NOT used to build the model: check it against the independently
# reported intravenous t1/2 of 1.61 h (Table 3).
stopifnot(abs(got[["half.life"]] / 1.61 - 1) < 0.10)
# Tmax is knowingly and provably wrong -- pin the known value so that a change
# in the reconstruction is caught rather than silently accepted.
stopifnot(abs(got[["tmax"]] - 1.30) < 0.10)
```

# Assumptions and deviations (Errata)

## The plasma layer is reconstructed, not published

This is the single most important caveat for anyone using these models.

Shen 2026 states that "the pharmacokinetic component was described by a
one-compartment model with extravascular input and clearance parameterization",
but **reports no parameter of that fit anywhere**. Table 5 is explicitly and
exclusively the PD parameter table; Tables 3 and 4 are Phoenix WinNonlin
non-compartmental analysis, not the compartmental fit. There is no `ka`, no
`V/F`, no `CL/F` and no lag time in the paper, and the supplement contains only
the analytical stability tables S1 and S2. The plasma layer was therefore
reconstructed here, under an operator decision, from the paper's own published
results:

| Parameter | Value | Provenance |
|---|---|---|
| `CL/F` | 378.6731 mL/h/kg | `Dose / AUC_po` = 50,000 / 132.04 (Table 4) -- an exact NCA identity |
| `V/F`  | 866.9804 mL/kg   | `Vz / F` = 350 / 0.4037 (Tables 3 and 4) |
| `ka`   | 1.237772 /h      | **reconstructed**: the value that makes the model reproduce the Table 4 Cmax of 32.68 ug/mL |

The implied `kel` of 0.43677 /h agrees independently with the intravenous
`CL/Vz` = 0.42857 /h of Table 3 and with the terminal slope of the digitised
Figure 2B curve (0.43-0.44 /h), so only `ka` is genuinely unconstrained by
published numbers.

## Why Tmax is not reproduced, and why no other choice would do better

The published oral triplet (Cmax 32.68 ug/mL, Tmax 3.38 h, AUC 132.04 h*ug/mL)
is **infeasible for any lag-free one-compartment first-order-absorption model**,
so this is not a fitting failure and no parameterisation can repair it:

- For that structure `Tmax = ln(ka/kel)/(ka - kel)`, which is bounded above by
  `1/kel`. Honouring Tmax = 3.38 h therefore forces `kel <= 0.2959 /h`.
- With `CL/F` pinned by the published AUC, that forces `V/F >= 1279.9 mL/kg`.
- Cmax is then bounded by `(Dose/V)/e = 14.37 ug/mL` -- a factor of **2.3 below**
  the observed 32.68 ug/mL.

An absorption lag does not rescue it either: Figure 2B shows a non-zero mean
concentration already at 0.25 h (1.68 ug/mL), so no delay is admissible. Direct
least-squares fits to the digitised Figure 2B curve confirm the diagnosis rather
than fixing it -- the optimiser drives `ka` onto the `ka = kel` boundary and
still underpredicts the peak by roughly half.

The reason is visible in the Figure 2B inset: individual birds peak at 2, 3 or
4 h (Table 4), and several (inset birds 5 and 7) show frankly double-peaked
absorption, consistent with the crop reservoir the authors invoke in the
Discussion. Averaging those profiles produces a broad 2-4 h plateau that no
single Bateman function has.

Given that the triplet cannot all be honoured, this extraction keeps **Cmax and
AUC** -- the two exposure quantities the downstream indirect-response models
actually depend on -- and accepts a too-early Tmax. The Figure 3 replication
above is the evidence that this choice works: the effect-compartment link and
the turnover model together low-pass filter the plasma profile heavily, so the
biomarker time courses are reproduced despite the peak arriving early.

**Users who need a quantitatively faithful plasma profile should not use this
layer.** The PD layer (Table 5) is published in full and is unaffected.

## Other assumptions

- **No between-subject variability.** Shen 2026 fitted each bird individually
  and reported only the mean and SD across individuals, never an omega matrix.
  All parameters are fixed at the published means and the models are
  deterministic. The SDs in Table 5 are large (`ke0` for T-AOC is
  0.44 +/- 0.42), so the typical-value profile should not be read as
  representing a typical bird tightly.
- **Residual error magnitudes are fixed at zero.** The paper specifies the
  structures (multiplicative for plasma, additive for the biomarkers) but gives
  only the *initial* standard deviations handed to the estimator (0.1 and 1.0),
  not final estimates. The structures are encoded; the magnitudes are fixed at
  zero.
- **Mean-of-individual-fit parameters applied to a single profile.** Table 5
  reports the mean of eight individual fits. Simulating one profile with those
  means is not identical to averaging eight individual simulations, which
  accounts for part of the residual disagreement in the Figure 3 comparison.
- **Observed values are digitised from figures.** Shen 2026 plots but does not
  tabulate the observed plasma and biomarker time courses, so the reference
  points in this vignette carry digitisation error of roughly a few percent.
  The published Tables 3 and 4 values used in the NCA comparison are
  transcribed, not digitised.
- **Oral route only.** Antioxidant markers were measured only after oral
  dosing, so there is no intravenous PD model. The intravenous arm was analysed
  non-compartmentally only, and Table 3 describes a multi-exponential decay that
  the one-compartment oral model does not attempt to represent.
- **No covariates.** Shen 2026 fitted no covariate model of any kind; both sexes
  were studied but sex was not tested as a covariate.
- **New canonical compartment names.** `taoc`, `sod` and `mda` were registered
  in `inst/references/compartment-names.md` as standard clinical PD-output
  biomarkers, alongside existing entries such as `ldh`, `hb` and `anc`.
