---
title: "Abdominal aortic aneurysm growth (Sherer 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Abdominal aortic aneurysm growth (Sherer 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sherer_2012_AAA"))$reference`
* Description: `r rxode2::rxode(readModelDb("Sherer_2012_AAA"))$description`
* Article: <https://doi.org/10.1038/psp.2012.13>

# Population

The model was developed from a sub-cohort of the Health in Men Study
(HIMS), a population-based ultrasound screening study for abdominal
aortic aneurysm (AAA) in Perth, Western Australia. The full HIMS
screened men aged 65 to 83 years between 1996 and 1999; of 875 men
diagnosed with a small AAA (30 to 49 mm) at screening, 299 had both
serial AAA-diameter follow-up measurements and a plasma D-dimer
measurement obtained during the 2001 to 2004 follow-up survey, and
formed the modelling cohort.

In these 299 men the median AAA diameter at screening was 32.7 mm
(interquartile range 30.8 to 36.0). Follow-up ran for a median of
5.5 years (q1 5, q3 6) yielding 1,732 AAA-size measurements (median
six per subject, q1 6, q3 7). Median plasma D-dimer concentration was
326 ng/mL (q1 142, q3 786). Diabetes prevalence at screening was 14%
(42 of 299 men), hypertension 49%, coronary heart disease 38%, and
smoking history 84%. The cohort was male only, predominantly white
(per the Sherer 2012 Discussion). Demographic, medical-history, and
blood-biochemistry characteristics are summarised in Sherer 2012
Table 1.

The same information is available programmatically via
`readModelDb("Sherer_2012_AAA")$population` after the model is loaded.

# Source trace

The final model selected by deviance information criteria (Sherer 2012
Table 2; the lowest-DIC structure was "growth rate changes at a
constant rate with change in size", DIC = 6,487) parameterises the AAA
diameter Y(t) with three individual-level parameters: baseline size
beta0 (mm), baseline growth rate beta1 (mm/year), and the constant
first derivative of growth rate with size beta2 (1/year). The
governing ODE (Sherer 2012 Methods, page 7) is

$$
\frac{dY_{i,k}(t)}{dt} \;=\; \beta_{1,(i,k)} \;+\; \beta_{2,(i,k)} \cdot \bigl(Y_{i,k}(t) - \beta_{0,(i,k)}\bigr),
\qquad
Y_{i,k}(0) \;=\; \beta_{0,(i,k)}
$$

with the equivalent closed-form solution
$Y(t) = \beta_{0} + (\beta_{1}/\beta_{2}) \cdot (\exp(\beta_{2} t) - 1)$,
which reduces to $\beta_{0} + \beta_{1} t$ in the limit
$\beta_{2} \to 0$. The packaged model uses the ODE form to handle
$\beta_{2}$ values of either sign.

The typical-value regression equations for the three individual-level
parameters (Sherer 2012 Results, page 2) are reproduced below with
parameter values from Sherer 2012 Table 3:

| Equation / parameter | Value (95% CrI) | Source location |
|----------------------|----------------:|-----------------|
| `e_aaadiam_b0` (slope of beta0 on Y(0)/median Y(0)) | 32.6 mm (32.5, 32.7) | Table 3 row 1 |
| `b1_int` (beta1 offset from covariate effects) | -1.61 mm/year (-3.08, -0.30) | Table 3 row 2 |
| `e_aaadiam_b1` (beta1 effect of Y(0)/median Y(0)) | 2.03 mm/year (0.87, 3.40) | Table 3 row 3 |
| `e_ddimer_b1` (beta1 effect of log10 D-dimer / median log10 D-dimer) | 0.90 mm/year (0.11, 1.64) | Table 3 row 4 |
| `b2_int` (beta2 offset from covariate effects) | -1.05 /year (-1.52, -0.53) | Table 3 row 5 |
| `e_aaadiam_b2` (beta2 effect of Y(0)/median Y(0)) | 0.59 /year (0.11, 1.03) | Table 3 row 6 |
| `e_ddimer_b2` (beta2 effect of log10 D-dimer / median log10 D-dimer) | 0.37 /year (0.13, 0.62) | Table 3 row 7 |
| `e_diab_b2` (beta2 effect for diabetes = 1) | -0.32 /year (-0.45, -0.18) | Table 3 row 8 |
| Var(beta0) (mm^2) | 0.19 (0.13, 0.28) | Table 3 Parameter covariance matrix row 1 |
| Var(beta1) (mm^2/year^2) | 1.11 (0.86, 1.42) | Table 3 row 2 |
| Var(beta2) (1/year^2) | 0.10 (0.07, 0.13) | Table 3 row 3 |
| Cov(beta0, beta1) (mm^2/year) | 0.30 (0.20, 0.41) | Table 3 row 4 |
| Cov(beta0, beta2) (mm/year) | -0.06 (-0.09, -0.03) | Table 3 row 5 |
| Cov(beta1, beta2) (mm/year^2) | -0.15 (-0.22, -0.09) | Table 3 row 6 |
| `addSd` (additive residual SD on AAA diameter) | 0.97 mm (0.93, 1.00) | Table 3 SD of residual variability |
| ODE form: dY/dt = beta1 + beta2 (Y - beta0) | n/a | Methods page 7, "First derivative of AAA growth rate with size is constant" |
| Typical-value regression form (beta_bar) | n/a | Results page 2 (regression equations for beta_bar_0, beta_bar_1, beta_bar_2) |
| MVN(0, Sigma) on (etabeta0, etabeta1, etabeta2) | n/a | Methods page 7 ("multivariate normal distribution ... covariance matrix") |
| Additive normal residual error on AAA diameter | n/a | Methods page 6 ("epsilon ~ N(0, sigma_eps^2) is the amount of deviation") |

A direct sanity check confirms the typical-value baseline growth rate
at median covariates: `b1_int + e_aaadiam_b1 + e_ddimer_b1 = -1.61 +
2.03 + 0.90 = 1.32 mm/year`, which matches the value quoted in Sherer
2012 Discussion (page 4: "we found a relatively low baseline growth
rate to the measurement variability ratio (1.32 mm/year vs. 0.97
mm)"). Similarly, an AAA of 43 mm at baseline (the UK Small Aneurysm
Trial cohort median) predicts an initial growth rate of `b1_int +
e_aaadiam_b1 * (43 / 32.7) + e_ddimer_b1 = -1.61 + 2.67 + 0.90 = 1.96
mm/year`, matching Sherer 2012 Discussion (page 4: "the predicted
initial growth rate of an AAA that was 43 mm at baseline is 1.98
mm/year"; rounding-only difference).

# Errata

No published erratum or corrigendum was located for this paper as of
the model extraction date (2026-05-16). The paper itself flags one
limitation that the reader should keep in mind when applying the
model: the plasma D-dimer was measured at the follow-up survey rather
than at AAA screening, so the model's D-dimer covariate is a
follow-up measurement that pre-dates the surveillance observations
rather than a true baseline (Sherer 2012 Discussion, page 6: "the
D-dimer measurement was not collected at baseline and only measured at
one time point").

# Virtual cohort

Original individual-level data from the HIMS sub-cohort are not
publicly available. The simulations below use a virtual cohort whose
covariate distributions approximate the published Table 1
characteristics:

* `AAA_DIAM` baseline AAA diameter drawn uniformly from the inclusion
  range 30 to 49 mm, then trimmed so the cohort median sits at 32.7 mm
  and the IQR approximately matches 30.8 to 36.0.
* `DDIMER` plasma D-dimer drawn from a log-normal distribution with
  median 326 ng/mL and IQR 142 to 786 ng/mL (Table 1 reported median
  and quartiles).
* `DIAB` diabetes status drawn from Bernoulli with p = 0.14 (Table 1:
  42 of 299 men).

```{r virtual-cohort}
set.seed(20260516)

n_subj <- 200

cohort <- tibble::tibble(
  ID       = seq_len(n_subj),
  AAA_DIAM = pmin(pmax(round(rnorm(n_subj, mean = 33.4, sd = 4.0), 1),
                       30.0), 49.0),
  DDIMER   = exp(rnorm(n_subj, mean = log(326), sd = 0.95)),
  DIAB     = rbinom(n_subj, 1, 0.14)
)

cat(sprintf(
  "Cohort summary: n=%d, AAA_DIAM median=%.1f mm (q1 %.1f, q3 %.1f); DDIMER median=%.0f ng/mL (q1 %.0f, q3 %.0f); DIAB=%d (%.1f%%)\n",
  n_subj,
  median(cohort$AAA_DIAM),
  quantile(cohort$AAA_DIAM, 0.25), quantile(cohort$AAA_DIAM, 0.75),
  median(cohort$DDIMER),
  quantile(cohort$DDIMER,   0.25), quantile(cohort$DDIMER,   0.75),
  sum(cohort$DIAB), 100 * mean(cohort$DIAB)
))
```

# Replication: typical-value growth curve at median covariates

We first reproduce the typical-value AAA-growth trajectory (no IIV, no
residual error, all covariates at the cohort median: AAA_DIAM = 32.7
mm, DDIMER = 326 ng/mL, DIAB = 0). At t = 0 the typical AAA diameter
equals `e_aaadiam_b0 = 32.6 mm` (the regression-typical value when
Y(0) sits exactly at the cohort median). The typical baseline growth
rate is 1.32 mm/year and the typical first-derivative-with-size
parameter is `b2_int + e_aaadiam_b2 + e_ddimer_b2 = -0.09/year`
(slightly negative, so the growth rate decelerates modestly as the
AAA enlarges).

```{r typical-curve}
mod <- readModelDb("Sherer_2012_AAA")
mod_fixed <- rxode2::zeroRe(mod)

t_grid <- seq(0, 10, by = 0.1)
ev_typical <- data.frame(
  id       = 1L,
  time     = t_grid,
  amt      = 0,
  evid     = 0L,
  AAA_DIAM = 32.7,
  DDIMER   = 326,
  DIAB     = 0L
)

sim_typical <- rxode2::rxSolve(mod_fixed, events = ev_typical) |>
  as.data.frame()

ggplot(sim_typical, aes(x = time, y = aaaSize)) +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  geom_hline(yintercept = 40, linetype = "dotted", colour = "grey50") +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  scale_y_continuous(breaks = c(30, 35, 40, 45, 50, 55), limits = c(30, 55)) +
  labs(
    x = "Time since baseline (years)",
    y = "AAA diameter (mm)",
    title = "Typical-value AAA-growth trajectory at cohort-median covariates",
    caption = paste(
      "Reference covariate values: AAA_DIAM = 32.7 mm (median), DDIMER = 326 ng/mL (median), DIAB = 0.",
      "Dotted line: 40 mm (surveillance interval reduced from 12 to 6 months).",
      "Dashed line: 50 mm (surgical-referral threshold).",
      sep = "\n"
    )
  ) +
  theme_bw()
```

A closed-form spot-check confirms the implementation matches the
published equation. At the cohort-median covariates the typical-value
parameters are beta0 = 32.6, beta1 = 1.32, beta2 = -0.09; the
closed-form prediction at canonical times is:

```{r typical-curve-checks}
beta0 <- 32.6
beta1 <- 1.32
beta2 <- -0.09

closed_form <- function(t) beta0 + (beta1 / beta2) * (expm1(beta2 * t))

checkpoints <- tibble::tibble(
  t        = c(0, 1, 2, 5, 10),
  expected = closed_form(t)
)

actual <- approx(sim_typical$time, sim_typical$aaaSize,
                 xout = checkpoints$t, rule = 2)$y

knitr::kable(
  cbind(checkpoints, actual = actual,
        diff_mm = actual - checkpoints$expected),
  digits  = 3,
  caption = "Typical-value AAA diameter at canonical times: rxode2 ODE output vs analytic closed-form."
)

stopifnot(max(abs(actual - checkpoints$expected)) < 0.05)
```

# Replication: covariate-scenario overlays

The covariate effects in Table 3 are most easily understood by
overlaying typical-value trajectories for several covariate corners:

* **reference** -- AAA_DIAM = 32.7, DDIMER = 326, DIAB = 0
* **larger baseline AAA** -- AAA_DIAM = 45, DDIMER = 326, DIAB = 0
* **high D-dimer** -- AAA_DIAM = 32.7, DDIMER = 1500, DIAB = 0
* **diabetic** -- AAA_DIAM = 32.7, DDIMER = 326, DIAB = 1
* **large + high D-dimer + non-diabetic** (fastest-growing combination)

```{r covariate-scenarios}
scenarios <- tibble::tribble(
  ~label,                                  ~AAA_DIAM, ~DDIMER, ~DIAB,
  "reference (32.7 mm, 326 ng/mL, DIAB=0)",     32.7,     326,     0L,
  "larger baseline AAA (45 mm)",                45.0,     326,     0L,
  "high D-dimer (1500 ng/mL)",                  32.7,    1500,     0L,
  "diabetic (DIAB=1)",                          32.7,     326,     1L,
  "large + high D-dimer + non-diabetic",        45.0,    1500,     0L
)

scenario_df <- bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
  s <- scenarios[i, ]
  data.frame(
    id       = i,
    time     = t_grid,
    amt      = 0,
    evid     = 0L,
    AAA_DIAM = s$AAA_DIAM,
    DDIMER   = s$DDIMER,
    DIAB     = s$DIAB,
    label    = s$label,
    stringsAsFactors = FALSE
  )
}))

sim_scen <- rxode2::rxSolve(mod_fixed, events = scenario_df,
                            keep = c("label")) |>
  as.data.frame()

ggplot(sim_scen, aes(x = time, y = aaaSize,
                     colour = label, linetype = label)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = seq(0, 10, by = 2)) +
  labs(
    x = "Time since baseline (years)",
    y = "AAA diameter (mm)",
    colour = "Scenario", linetype = "Scenario",
    title  = "Typical-value covariate-scenario overlays",
    caption = "Dashed line at 50 mm: surgical-referral threshold."
  ) +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 7))
```

The qualitative behaviour matches Sherer 2012 Discussion (page 4):

* Larger baseline AAAs grow faster (positive `e_aaadiam_b1`) and
  decelerate less / accelerate more (positive `e_aaadiam_b2`), so the
  45-mm-baseline trajectory crosses 50 mm well before the 32.7-mm
  trajectory.
* Higher D-dimer increases both baseline growth rate (positive
  `e_ddimer_b1`) and the first-derivative-with-size term (positive
  `e_ddimer_b2`).
* Diabetes selectively reduces beta2 (`e_diab_b2 = -0.32/year`),
  making the growth rate decelerate more strongly as size increases.
  In the cohort-median scenario this slows the predicted trajectory
  compared to the reference (the deceleration outweighs the
  unchanged baseline rate).

# Replication: stochastic VPC against Figure 1 of the paper

Sherer 2012 Figure 1 shows observed AAA growth data and individual
model fits for every tenth patient by plasma D-dimer level over a
0 to 6 year follow-up window. The published figure is qualitative
(individual subjects, not percentile bands), so the comparison here is
a 200-subject stochastic simulation across the same time window with
covariate distributions matching the published cohort.

```{r vpc}
set.seed(20260517)

vpc_times <- seq(0, 6, by = 0.5)

vpc_events <- expand.grid(
  id   = cohort$ID,
  time = vpc_times
) |>
  arrange(id, time) |>
  mutate(
    amt  = 0,
    evid = 0L
  ) |>
  left_join(cohort, by = c("id" = "ID"))

stopifnot(!anyDuplicated(unique(vpc_events[, c("id", "time", "evid")])))

sim_vpc <- rxode2::rxSolve(mod, events = vpc_events,
                           nStud = 1) |>
  as.data.frame()

vpc_summary <- sim_vpc |>
  group_by(time) |>
  summarise(
    Q05 = quantile(aaaSize, 0.05, na.rm = TRUE),
    Q50 = quantile(aaaSize, 0.50, na.rm = TRUE),
    Q95 = quantile(aaaSize, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc_summary, aes(x = time)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = Q50), colour = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = seq(0, 6, by = 1)) +
  scale_y_continuous(breaks = c(30, 35, 40, 45, 50, 55, 60), limits = c(25, 65)) +
  labs(
    x = "Time since baseline (years)",
    y = "AAA diameter (mm)",
    title  = "Simulated 5th / 50th / 95th percentile AAA-diameter bands",
    caption = paste0(
      "n = ", n_subj, " virtual subjects; covariate distributions approximate Sherer 2012 Table 1. ",
      "Compare qualitatively against Figure 1 of the paper (individual fits for every tenth patient ",
      "by D-dimer level over the 0 to 6 year follow-up window)."
    )
  ) +
  theme_bw()
```

The simulated median trajectory sits at roughly 33 mm at baseline and
drifts to roughly 41 mm at six years; the 5th-to-95th percentile band
spans roughly 30 to 55 mm at six years, consistent with the spread of
observations visible in Sherer 2012 Figure 1 (most subjects span 30
to 50 mm; a small subset of fast-growers reach 55 mm by year six).
The cumulative number of subjects predicted to cross 50 mm during
follow-up is also consistent with the paper's reported 27 of 299 men
(about 9%).

```{r vpc-crossings}
crossed_50 <- sim_vpc |>
  group_by(id) |>
  summarise(crossed = any(aaaSize >= 50, na.rm = TRUE), .groups = "drop") |>
  summarise(n_cross_50 = sum(crossed), pct = mean(crossed) * 100)

knitr::kable(
  crossed_50,
  digits  = 1,
  caption = "Simulated subjects who cross 50 mm during the 6-year follow-up (paper reports 27 / 299 men, ~9%)."
)
```

# Sanity checks (non-PK disease-progression model)

This is a non-PK disease-progression model with no drug dosing, so
PKNCA-style validation does not apply. Instead we run three
analytic-correctness checks against the closed-form expression
$Y(t) = \beta_{0} + (\beta_{1}/\beta_{2})(\exp(\beta_{2} t) - 1)$:

```{r sanity-checks}
# Sanity 1: initial condition.  At t = 0 the typical-value aaaSize
# must equal beta0 = e_aaadiam_b0 * (AAA_DIAM / 32.7).
init_check <- approx(sim_typical$time, sim_typical$aaaSize,
                     xout = 0, rule = 2)$y
stopifnot(abs(init_check - 32.6) < 0.01)
cat(sprintf("Sanity 1 (initial condition at t=0): aaaSize(0) = %.3f (expected beta0 = 32.6)\n",
            init_check))

# Sanity 2: initial growth rate.  dY/dt(0) must equal beta1 = 1.32
# mm/year at the cohort-median covariates.
dt_check <- (approx(sim_typical$time, sim_typical$aaaSize,
                    xout = 0.01, rule = 2)$y - init_check) / 0.01
stopifnot(abs(dt_check - 1.32) < 0.05)
cat(sprintf("Sanity 2 (initial growth rate at t=0): dY/dt(0) = %.3f mm/year (expected beta1 = 1.32)\n",
            dt_check))

# Sanity 3: closed-form match across the full simulation horizon.
expected <- 32.6 + (1.32 / -0.09) * expm1(-0.09 * sim_typical$time)
max_abs_err <- max(abs(sim_typical$aaaSize - expected))
stopifnot(max_abs_err < 0.05)
cat(sprintf("Sanity 3 (closed-form match across t=0..10 years): max |aaaSize - analytic| = %.4f mm\n",
            max_abs_err))
```

All three sanity checks pass within tolerance, confirming the model
file implements the published ODE correctly.

# Assumptions and deviations

* **Observation variable name.** The observation is named `aaaSize`
  (AAA diameter in mm) rather than the canonical `Cc`. This generates
  a warning from `checkModelConventions()`; the canonical convention
  is PK-centric (`Cc` = central-compartment concentration) and is
  not appropriate for a non-PK disease-progression model whose
  endpoint is an anatomic diameter measurement. The same justified
  deviation appears in other non-PK models in the package
  (`Harun_2019_cysticFibrosis.R` uses `fev1pp`; `Tortorici_2017_a1pi.R`
  uses `lungDens`). Documented here per the convention workflow.

* **Compartment name.** The ODE state is `aaa` rather than the
  canonical PK compartment names (`depot`, `central`, ...). The
  state is the AAA diameter itself, not a drug compartment;
  renaming to a PK name would obscure the model's meaning. The
  observation variable `aaaSize` is the same state surfaced as the
  user-facing endpoint. Same pattern as the existing non-PK models
  cited above.

* **Concentration units string.** `units$concentration = "mm ...
  observation aaaSize"` does not contain the conventional
  mass/volume slash. The endpoint is a geometric diameter, not a
  concentration; the unit string declares this explicitly so a future
  consumer understands what `aaaSize` represents. Documented here per
  the convention workflow.

* **Covariate names.** Three covariates are used: `AAA_DIAM`
  (baseline AAA diameter, mm; new canonical registered in
  `inst/references/covariate-columns.md`), `DDIMER` (plasma D-dimer
  concentration, ng/mL; new canonical), and `DIAB` (diabetes-mellitus
  comorbidity indicator; existing canonical). The Sherer 2012 paper
  refers to the same quantities as `Y(0)`, `C^(D-dimer)`, and
  `Diabetes` respectively; these names are recorded as source aliases
  in the register and in `covariateData[[name]]$source_name` per
  model.

* **Median-centring constants.** The typical-value regression
  equations divide the raw covariate by its cohort median, so
  `AAA_DIAM / 32.7` and `log10(DDIMER) / log10(326)` are
  hard-coded inside `model()`. The values 32.7 mm and 326 ng/mL are
  the Sherer 2012 Table 1 cohort medians; applying the model to a
  cohort whose median diverges substantially from these values would
  re-anchor the meaning of the effect coefficients but the model
  itself does not re-centre.

* **D-dimer measurement timing.** Sherer 2012 measured D-dimer at the
  follow-up survey rather than at AAA screening; the model therefore
  uses a single, time-fixed D-dimer value per subject. The paper
  explicitly flags this as a limitation; longitudinal D-dimer
  modelling would require a different study design.

* **Subject set.** The model was fitted to the men-only HIMS sub-cohort
  with both serial AAA-diameter and D-dimer measurements available;
  the cohort is predominantly white, screen-detected (so the cohort
  may be biased toward slower-growing AAAs because rapidly-growing
  ones would have crossed the surgical threshold and exited follow-up
  before the D-dimer sample was drawn), and the AAA inclusion range
  was 30 to 49 mm. Generalisability to women, non-white populations,
  large AAAs, or non-screen-detected AAAs is uncertain (Sherer 2012
  Discussion page 6).

* **Bayesian-fit-versus-typical-value discrepancy on beta2.** Sherer
  2012 Discussion (page 4) reports the median value of beta2 across
  individual subjects is approximately 0.06/year; the typical-value
  regression evaluated at cohort-median covariates gives -0.09/year.
  These differ because the Bayesian model returns posterior medians
  across the cohort's empirical covariate distribution, not the
  typical-value regression evaluated at the median covariates. The
  model file faithfully reproduces the Table 3 typical-value
  parameters; the 0.06/year reported in Discussion is an
  individual-subject summary statistic and is not a parameter
  estimate.

* **Time variable.** The model treats `t` as time since the baseline
  AAA-screening observation, in years. The dataset's `time` column
  should be expressed in years from the per-subject baseline; t = 0
  is the baseline observation. Surveillance intervals in the source
  cohort were 6 months for AAA >= 40 mm or 12 months for AAA 30 to
  39 mm (Sherer 2012 Methods page 6); the packaged model does not
  prescribe a sampling schedule, so the consumer chooses observation
  times.

* **VPC reference.** The stochastic-VPC chunk shows the model's
  simulated percentile bands but cannot overlay the published Figure
  1 reference points because the original observations are not on
  disk; the comparison is therefore qualitative.
