---
title: "Paroxetine cumulative exposure and depression-severity response (Shigetome 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paroxetine cumulative exposure and depression-severity response (Shigetome 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Models and source

Shigetome 2025 reports two models fitted in two separate NONMEM runs on two
different cohorts, so the paper contributes two model files:

```{r models}
pk_ui <- rxode2::rxode(readModelDb("Shigetome_2025_paroxetine"))
pd_ui <- rxode2::rxode(readModelDb("Shigetome_2025_paroxetine_madrs"))
```

* Citation: `r pk_ui$reference`
* Article: <https://doi.org/10.1002/psp4.70032>
* Supplementary Information (Tables S1-S5, Figures S1-S6, Texts S1-S4 including
  both NONMEM control streams):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12167921/supplementaryFiles>

**`Shigetome_2025_paroxetine`** -- population PK, 329 steady-state
concentrations from 179 patients.

> `r pk_ui$description`

**`Shigetome_2025_paroxetine_madrs`** -- population PK-PD, MADRS time course in
the 50-patient subset with serial depression ratings.

> `r pd_ui$description`

The paper also develops six machine-learning classifiers (logistic regression,
decision tree, k-nearest neighbour, support vector machine, random forest,
multi-layer perceptron) predicting week-6 remission, with SHAP attribution.
Those are classifiers rather than pharmacometric models and are not part of
this extraction; the two nonlinear mixed-effects models above are.

# Population

The PK cohort is 179 Japanese adults with DSM-IV major depressive disorder
treated at Hirosaki University Hospital or Dokkyo Medical University School of
Medicine, all antidepressant-naive at paroxetine initiation, with normal renal
and hepatic function and no interacting co-medication (Shigetome 2025 Table 1).
Age was 41.5 years (SD 14.2, range 16-80), body weight 58.6 kg (SD 10.5, range
40-85) and 40.8% were male. Paroxetine was started at 10-20 mg/day and titrated
by 10 mg/day weekly to the maximum tolerated dose, giving a mean of 22.3 mg/day
(SD 11.1, range 10-40). Sampling was sparse therapeutic drug monitoring: 1.8
samples per patient (range 1-5), all at steady state after at least 9 days on a
stable dose, giving observed plasma concentrations of 51.9 ng/mL (SD 49.0,
range 2.6-286.5). CYP2D6 genotyping found 121 intermediate metabolizers
(67.6%) and 58 extensive metabolizers (32.4%); there were no poor metabolizers,
which the authors attribute to the low frequency of null alleles in East Asians.

The PK-PD cohort is the 50-patient subset who also had MADRS ratings at
baseline and weeks 1, 2, 4 and 6 plus a pre-treatment Temperament and Character
Inventory (Table 2). That subset is slightly older (46.6 years, SD 13.7),
slightly lighter (56.3 kg, SD 10.4) and received higher doses (32.2 mg/day, SD
11.0). Baseline MADRS was 39.6 (SD 9.7) falling to 13.7 (SD 12.9) by week 6;
26 of the 50 reached remission (MADRS < 10 at week 6).

The same information is available programmatically:

```{r population}
str(readModelDb("Shigetome_2025_paroxetine")()$population, max.level = 1)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
tables below collect them. Point estimates for both models live only in the
Supporting Information (Tables S2 and S4); the main text prints the final model
equations but no parameter table.

## Population PK (`Shigetome_2025_paroxetine`)

| Equation / parameter | Value | Source location |
|---|---|---|
| `CL/F = 23.6 * (AGE/41.5)^-0.536 * (WT/58.6)^0.563` | -- | Results 3.2, printed final popPK model equation |
| `Vd/F = 1010 L` | -- | Methods 2.3 and Results 3.2; NONMEM `$THETA 1010 FIX` (Text S2) |
| One compartment, no absorption process | -- | Methods 2.3; NONMEM `$SUBROUTINES ADVAN1 TRANS2` (Text S2) |
| `Cc = 1000 * central / vc` (mg, L -> ng/mL) | -- | NONMEM `$PK  S1 = V / 1000` (Text S2) |
| `lcl` (CL/F) | 23.6 L/h | Table S2 (RSE 4.75%, bootstrap 95% CI 21.4-25.9) |
| `lvc` (Vd/F, fixed) | 1010 L | Table S2 (fixed; 58.6 kg x 17.2 L/kg from the cited literature) |
| `e_age_cl` | -0.536 | Table S2 (RSE 25.4%, bootstrap 95% CI -0.790 to -0.270) |
| `e_wt_cl` | 0.563 | Table S2 (RSE 48.9%, bootstrap 95% CI -0.01 to 1.19) |
| `etalcl` (variance) | 0.307 | Table S2 (RSE 15.7%, shrinkage 7.83%, bootstrap 95% CI 0.197-0.392) |
| `propSd` = sqrt(0.155) | 0.394 | Table S2 proportional error 0.155 (RSE 10.8%, shrinkage 19.3%) |

## Population PK-PD (`Shigetome_2025_paroxetine_madrs`)

| Equation / parameter | Value | Source location |
|---|---|---|
| `EFF = Emax * Time / (ET50 + Time)` | -- | Methods 2.3 printed Emax equation; NONMEM `$PRED` (Text S3) |
| `Emax = 90.9 - (AUC_0-1week / 2610.43)^-8.56` | -- | Results 3.3, printed final popPK/PD equation |
| `ET50 = 0.984 + (MADRS_W1 / 28.64)^2.38` | -- | Results 3.3, printed final popPK/PD equation |
| Proportional IIV `* (1 + ETA)` on both | -- | Results 3.3; NONMEM `$PRED` (Text S3) |
| `emax` | 90.9 % | Table S4 (RSE 9.3%, bootstrap median 91.5, 95% CI 74.7-111) |
| `e_auc_parox_emax` | -8.56 | Table S4 (RSE 3.36%, bootstrap 95% CI -12.7 to -2.20) |
| `et50` | 0.984 week | Table S4 (RSE 45.5%, bootstrap median 0.931, 95% CI 0.101-1.98) |
| `e_score_madrs_et50` | 2.38 | Table S4 (RSE 43.3%, bootstrap 95% CI 0.666-6.66) |
| `etaemax` (variance) | 0.150 | Table S4 (RSE 34.1%, shrinkage 16.0%, bootstrap 95% CI 0.042-0.258) |
| `etaet50` (variance) | 0.486 | Table S4 (RSE 55.1%, shrinkage 32.0%, bootstrap 95% CI 0.0315-1.46) |
| `addSd` = sqrt(190) | 13.78 % | Table S4 additive error 190 (RSE 23.5%, shrinkage 16.3%, bootstrap 95% CI 112-316) |
| Normalizing constants 2610.43 ng*h/mL, 28.64 | -- | Results 3.3 equations; NONMEM `$PRED` (Text S3) |

# Part 1 -- population PK

## Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the covariate distributions of Table 1 (age and body weight truncated to the
published ranges), and doses are drawn from {10, 20, 30, 40} mg/day with
probabilities chosen to match the published mean of 22.3 mg/day (SD 11.1).

```{r cohort_pk}
set.seed(20250617)

n_pk <- 179L

rtrunc_norm <- function(n, mean, sd, lo, hi) {
  x <- rnorm(n, mean, sd)
  while (any(bad <- x < lo | x > hi)) x[bad] <- rnorm(sum(bad), mean, sd)
  x
}

pk_subjects <- tibble(
  id = seq_len(n_pk),
  AGE = rtrunc_norm(n_pk, 41.5, 14.2, 16, 80),
  WT  = rtrunc_norm(n_pk, 58.6, 10.5, 40, 85),
  dose = sample(c(10, 20, 30, 40), n_pk, replace = TRUE,
                prob = c(0.35, 0.30, 0.13, 0.22))
)

c(mean_age = mean(pk_subjects$AGE), sd_age = sd(pk_subjects$AGE),
  mean_wt = mean(pk_subjects$WT), sd_wt = sd(pk_subjects$WT),
  mean_dose = mean(pk_subjects$dose), sd_dose = sd(pk_subjects$dose))
```

## Steady-state concentrations versus Table 1

Each patient is dosed once daily for 30 days -- comfortably past the paper's
9-day stable-dose inclusion criterion -- and a single trough is taken on day
30, mimicking the sparse TDM sampling. `sim` carries the proportional residual
error; `Cc` is the individual prediction without it.

Tables S2 and S4 head their variability columns "%CV" while reporting what are
in fact the raw NONMEM `$OMEGA` and `$SIGMA` variances (argued in full under
*Assumptions and deviations*). This cohort is the sharpest available test of
that reading, so both encodings are simulated: the packaged model, which treats
0.307 and 0.155 as variances, and an alternative in which the same printed
numbers are taken as a standard deviation and a CV.

```{r sim_ss}
mod_pk <- readModelDb("Shigetome_2025_paroxetine")

# Same printed numbers, read as an SD and a CV instead of as variances.
mod_pk_sd_reading <- rxode2::ini(mod_pk, etalcl ~ 0.307^2, propSd = 0.155)

build_pk_events <- function(subjects, dose_col, n_days, obs_times) {
  doses <- subjects |>
    tidyr::expand_grid(day = seq_len(n_days) - 1L) |>
    mutate(time = day * 24, amt = .data[[dose_col]], evid = 1L,
           cmt = "central") |>
    select(id, AGE, WT, time, amt, evid, cmt)
  obs <- subjects |>
    tidyr::expand_grid(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central") |>
    select(id, AGE, WT, time, amt, evid, cmt)
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

ev_ss <- build_pk_events(pk_subjects, "dose", n_days = 30L,
                         obs_times = 30 * 24)

solve_ss <- function(mod) {
  out <- rxode2::rxSolve(mod, ev_ss, omega = rxode2::rxode(mod)$omega,
                         returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  stopifnot(dplyr::n_distinct(out$id) == n_pk)
  out
}

sim_ss <- solve_ss(mod_pk)
sim_ss_sd <- solve_ss(mod_pk_sd_reading)

ss_summary <- tibble(
  Source = c("Simulated, variances (packaged)",
             "Simulated, read as SD / CV",
             "Shigetome 2025 Table 1 (329 samples)"),
  Mean = c(mean(sim_ss$sim), mean(sim_ss_sd$sim), 51.9),
  SD = c(sd(sim_ss$sim), sd(sim_ss_sd$sim), 49.0),
  Min = c(min(sim_ss$sim), min(sim_ss_sd$sim), 2.6),
  Max = c(max(sim_ss$sim), max(sim_ss_sd$sim), 286.5)
) |>
  mutate(CV = 100 * SD / Mean)

ss_summary |>
  dplyr::rename("Steady-state paroxetine (ng/mL)" = Source,
                "CV (%)" = CV) |>
  knitr::kable(digits = 1,
               caption = "Simulated steady-state concentrations against the observed distribution reported in Shigetome 2025 Table 1.")
```

The discriminating quantity is the coefficient of variation, which is
insensitive to the sampling-time and titration assumptions below. The
variance reading reproduces the observed 94%; the SD reading gives about 78%
and is too tight. The simulated *mean* sits roughly a quarter below the
observed 51.9 ng/mL under either reading, which is expected: every simulated
sample here is a true trough drawn after 30 days at one stable dose, whereas
the 329 observed TDM samples were drawn at unrecorded times within the dosing
interval and during an ongoing weekly titration, both of which raise the
observed mean relative to a pure steady-state trough. No parameter was adjusted
to close that gap.

## Week-1 exposure and the AUC that drives the PD model

The PK-PD model's covariate `AUC_PAROX` is the cumulative AUC over the first
week of treatment, predicted from this popPK model. The paper's Methods state
an initial dose of 10-20 mg/day, so both are simulated.

```{r sim_week1}
week1_arms <- bind_rows(
  pk_subjects |> mutate(w1dose = 10, treatment = "10 mg/day"),
  pk_subjects |> mutate(w1dose = 20, treatment = "20 mg/day",
                        id = id + n_pk)
)

ev_w1 <- build_pk_events(week1_arms, "w1dose", n_days = 7L,
                         obs_times = seq(0, 168, by = 0.5)) |>
  left_join(distinct(week1_arms, id, treatment), by = "id")

# zeroRe() suppresses the clearance eta; see the note below.
mod_pk_typ <- rxode2::zeroRe(mod_pk)

sim_w1 <- rxode2::rxSolve(mod_pk_typ, ev_w1, omega = NA,
                          keep = "treatment", returnType = "data.frame")
if (is.null(sim_w1$id)) sim_w1$id <- 1L
stopifnot(dplyr::n_distinct(sim_w1$id) == 2L * n_pk)
```

Suppressing the clearance eta is deliberate. The paper's own
`AUC_PAROX` values were produced from the population PK model's covariate
predictions: the observed spread in Table S5 (2472.2 +/- 385.6 ng*h/mL in the
non-remission group, a CV of 16%) is far narrower than the 60% CV that the
estimated CL variability would generate, and matches what age and body weight
alone contribute. Because the exponent on `AUC_PAROX` is -8.56, this matters:
sampling the clearance eta as well drives some subjects to exposures where the
`Emax` term diverges.

```{r figure-week1-profile}
sim_w1 |>
  group_by(treatment, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time since first dose (h)", y = "Paroxetine (ng/mL)",
       colour = "Week-1 dose", fill = "Week-1 dose",
       title = "First week of paroxetine treatment",
       caption = "Median and 5th-95th percentiles across the covariate distribution of Table 1, clearance eta suppressed.")
```

## PKNCA validation

```{r pknca}
sim_nca <- sim_w1 |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- ev_w1 |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(start = 0, end = 168,
                        auclast = TRUE, cmax = TRUE, tmax = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

The only exposure quantity Shigetome 2025 reports for its own model is the
first-week AUC. The normalizing constant in the printed `Emax` equation,
2610.43 ng*h/mL, is the cohort reference value, and Table S5 gives the observed
group means (2764.9 in the remission group, 2472.2 in the non-remission group).
The same published reference is therefore used for both simulated week-1 dose
arms; the paper does not stratify `AUC_PAROX` by week-1 dose.

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,   ~auclast,
  "10 mg/day",  2610.43,
  "20 mg/day",  2610.43
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = "auclast",
  units = c(auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(cmp, digits = 1,
             caption = "Simulated first-week AUC against the reference value of the published Emax equation. * differs by more than 20%.")
```

Ten milligrams a day reproduces the published reference within a few per cent;
twenty milligrams a day overshoots it by about 90% and is starred. Table S5
records the *week-1* dose as 20.0 mg/day, but Methods 2.1 states that treatment
*started* at 10-20 mg/day with the first increase after a week, so the dose in
force during the first week is the initial dose. The arithmetic settles it:
only a 10 mg/day first week reproduces both the 2610.43 normalizing constant
and the observed 1642-5814 ng*h/mL range. No parameter was adjusted to obtain
this agreement.

## Per-subject structural identity

For a one-compartment model with bolus input, the AUC accumulated to time *t*
equals the amount eliminated divided by clearance. Checking this per subject
rather than on a summary catches scaling errors that a median comparison hides.

```{r identity}
end_state <- sim_w1 |>
  filter(abs(time - 168) < 1e-8) |>
  transmute(id, treatment, central_end = central, cl)

dose_total <- ev_w1 |>
  filter(evid == 1) |>
  group_by(id) |>
  summarise(dose_total = sum(amt), .groups = "drop")

identity_check <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  select(treatment, id, auclast = PPORRES) |>
  inner_join(end_state, by = c("id", "treatment")) |>
  inner_join(dose_total, by = "id") |>
  mutate(auc_expected = 1000 * (dose_total - central_end) / cl,
         rel_err = abs(auclast - auc_expected) / auc_expected)

stopifnot(nrow(identity_check) == 2L * n_pk)
# 0.5 h observation grid; the residual is linear-trapezoid discretisation
# error on a bolus profile, not a model discrepancy.
stopifnot(max(identity_check$rel_err) < 0.015)

c(n = nrow(identity_check),
  max_relative_error_pct = 100 * max(identity_check$rel_err))
```

## Implied half-life

Fixing `Vd/F` at 1010 L while estimating `CL/F` at 23.6 L/h implies a terminal
half-life that can be read straight off the model.

```{r halflife}
hl <- sim_w1 |>
  filter(abs(time - 168) < 1e-8, treatment == "10 mg/day") |>
  mutate(half_life = log(2) * vc / cl) |>
  pull(half_life)

c(median_h = median(hl), min_h = min(hl), max_h = max(hl),
  typical_h = log(2) * 1010 / 23.6)
```

The typical value is about 30 h, longer than the roughly 21 h that Methods 2.1
cites from the literature when justifying the 9-day stable-dose inclusion
criterion. This is a direct consequence of fixing `Vd/F` rather than a
transcription problem, and it is recorded below.

# Part 2 -- MADRS time course

## Replicating Figure 1

Figure 1A of the paper contrasts a patient with a high first-week AUC against
an average patient; Figure 1B contrasts a low week-1 MADRS score against an
average one. The paper does not print the covariate values it used for the
"high" and "low" curves, so the published group means from Table S5 are used
instead, which makes the two panels a direct picture of the remission versus
non-remission contrast.

```{r figure-1}
mod_pd_typ <- rxode2::zeroRe(readModelDb("Shigetome_2025_paroxetine_madrs"))

pd_curve <- function(auc, madrs, label, panel) {
  ev <- rxode2::et(seq(0, 6, by = 0.05))
  rxode2::rxSolve(mod_pd_typ, ev, omega = NA,
                  params = c(AUC_PAROX = auc, SCORE_MADRS = madrs),
                  returnType = "data.frame") |>
    mutate(scenario = label, panel = panel)
}

fig1 <- bind_rows(
  pd_curve(2764.9, 28.64, "Remission-group AUC (2764.9 ng*h/mL)",     "A: first-week AUC"),
  pd_curve(2610.43, 28.64, "Cohort reference (2610.4 ng*h/mL)",       "A: first-week AUC"),
  pd_curve(2472.2, 28.64, "Non-remission-group AUC (2472.2 ng*h/mL)", "A: first-week AUC"),
  pd_curve(2610.43, 24.9, "Remission-group MADRS week 1 (24.9)",      "B: week-1 MADRS"),
  pd_curve(2610.43, 28.64, "Cohort reference (28.6)",                 "B: week-1 MADRS"),
  pd_curve(2610.43, 32.7, "Non-remission-group MADRS week 1 (32.7)",  "B: week-1 MADRS")
)

ggplot(fig1, aes(time, madrsenh, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~panel) +
  labs(x = "Treatment duration (weeks)",
       y = "Enhancement rate in depression severity (%)",
       colour = NULL,
       title = "Time course of MADRS improvement",
       caption = "Replicates Figure 1A and 1B of Shigetome 2025 using the group means of Table S5.") +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

Panel A reproduces the paper's finding that a higher first-week AUC raises the
attainable improvement, and panel B that a lower week-1 MADRS brings the
improvement forward.

## Predicted versus observed enhancement rates

The paper reports the mean enhancement rate at each visit (Results 3.3). A
200-patient cohort with the Table 2 demographics is built, its first-week
exposure is predicted with the popPK model exactly as in Part 1, and the PK-PD
model is then simulated with both etas active.

```{r pd_cohort}
set.seed(20250618)
n_pd <- 200L

pd_subjects <- tibble(
  id = seq_len(n_pd),
  AGE = rtrunc_norm(n_pd, 46.6, 13.7, 20, 70),
  WT  = rtrunc_norm(n_pd, 56.3, 10.4, 40, 85),
  SCORE_MADRS = rtrunc_norm(n_pd, 28.6, 11.3, 4, 56)
)

ev_pd_pk <- build_pk_events(mutate(pd_subjects, w1dose = 10), "w1dose",
                            n_days = 7L, obs_times = seq(0, 168, by = 0.5))

pd_auc <- rxode2::rxSolve(mod_pk_typ, ev_pd_pk, omega = NA,
                          returnType = "data.frame") |>
  group_by(id) |>
  summarise(AUC_PAROX = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
            .groups = "drop")

pd_subjects <- left_join(pd_subjects, pd_auc, by = "id")

c(mean = mean(pd_subjects$AUC_PAROX), sd = sd(pd_subjects$AUC_PAROX),
  min = min(pd_subjects$AUC_PAROX), max = max(pd_subjects$AUC_PAROX))
```

Those model-predicted first-week exposures sit inside the 1642-5814 ng*h/mL
range observed in Table S5, with a comparable spread, so the steep `Emax` term
is being evaluated where it was estimated.

```{r pd_sim}
mod_pd <- readModelDb("Shigetome_2025_paroxetine_madrs")

ev_pd <- pd_subjects |>
  tidyr::expand_grid(time = c(0, 1, 2, 4, 6)) |>
  select(id, AUC_PAROX, SCORE_MADRS, time)

sim_pd <- rxode2::rxSolve(mod_pd, ev_pd, omega = pd_ui$omega,
                          returnType = "data.frame")
if (is.null(sim_pd$id)) sim_pd$id <- 1L
stopifnot(dplyr::n_distinct(sim_pd$id) == n_pd)

# Proportional IIV on a strictly positive parameter admits non-positive draws.
# A subject with ET50 <= 0 has a pole at time = -ET50 inside the simulated
# window, so the summaries below are taken over the subjects for whom the
# published model is interpretable, and the excluded fraction is reported.
subject_pars <- sim_pd |> distinct(id, emax_ind, et50_ind)
neg_et50 <- 100 * mean(subject_pars$et50_ind <= 0)
ok_ids <- subject_pars$id[subject_pars$et50_ind > 0]

observed <- tibble(time = c(1, 2, 4, 6),
                   observed_mean = c(25.9, 45.7, 57.5, 64.2),
                   observed_sd = c(30.3, 27.4, 28.3, 30.6))

sim_pd |>
  filter(time %in% c(1, 2, 4, 6), id %in% ok_ids) |>
  group_by(time) |>
  summarise(simulated_mean = mean(madrsenh), simulated_sd = sd(sim),
            .groups = "drop") |>
  left_join(observed, by = "time") |>
  mutate(difference_pct = 100 * (simulated_mean - observed_mean) / observed_mean) |>
  dplyr::rename("Week" = time,
                "Simulated mean (%)" = simulated_mean,
                "Simulated SD (%)" = simulated_sd,
                "Observed mean (%)" = observed_mean,
                "Observed SD (%)" = observed_sd,
                "Difference (%)" = difference_pct) |>
  knitr::kable(digits = 1,
               caption = "Simulated versus observed mean enhancement rate in depression severity (Shigetome 2025 Results 3.3). Simulated values are over the subjects with a positive ET50 draw.")
```

Weeks 2, 4 and 6 are reproduced to within about 2%. Week 1 is over-predicted by
roughly a third (about 34% simulated against 25.9% observed); the same
over-prediction shows in the typical-value curve above (30.1% at week 1). It is
a property of the published fit rather than of this encoding -- an Emax-in-time
model with a single ET50 rises faster over the first week than the observed
cohort did.

```{r pd_neg}
c(pct_subjects_with_nonpositive_ET50 = neg_et50,
  n_subjects_excluded = sum(subject_pars$et50_ind <= 0))
```

```{r figure-pd-vpc}
sim_pd |>
  filter(id %in% ok_ids) |>
  group_by(time) |>
  summarise(Q05 = quantile(sim, 0.05), Q50 = median(sim),
            Q95 = quantile(sim, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  geom_pointrange(
    data = observed,
    aes(x = time, y = observed_mean,
        ymin = observed_mean - observed_sd, ymax = observed_mean + observed_sd),
    colour = "firebrick", inherit.aes = FALSE
  ) +
  labs(x = "Treatment duration (weeks)",
       y = "Enhancement rate in depression severity (%)",
       title = "Simulated MADRS improvement versus the observed visit means",
       caption = "Band: simulated median and 5th-95th percentiles over the subjects with a positive ET50 draw. Red: observed mean +/- SD, Shigetome 2025 Results 3.3.")
```

# Assumptions and deviations

## Reading the variability columns of Tables S2 and S4

Tables S2 and S4 head their inter-individual and residual variability columns
"%CV", but the numbers are the raw NONMEM `$OMEGA` and `$SIGMA` **variances**.
Three independent lines of evidence agree:

1. Table S4 reports the popPK/PD additive residual as **190** for an endpoint
   measured in per cent. As a standard deviation that is 190 percentage points
   on a roughly 0-100 scale, which is impossible; as a variance it is an SD of
   13.8 percentage points, which is reasonable. The bootstrap interval
   (112-316, asymmetric about 190) is also the shape of a variance interval.
2. Only the variance reading reproduces the observed concentration spread. With
   `Var(ln C) = Var(ln dose) + omega^2 + covariate variance + sigma^2 = 0.222 +
   0.307 + 0.042 + 0.155 = 0.725`, the predicted CV of steady-state
   concentrations is 103%, against the 49.0/51.9 = 94% observed in Table 1.
   Reading the same entries as standard deviations gives 68%. The simulation in
   Part 1 shows this directly.
3. For a proportional IIV of the form `(1 + eta)`, which is what the NONMEM
   `$PRED` block in Text S3 writes, the CV *is* the standard deviation of the
   eta, so a column that genuinely held CVs could not simultaneously hold the
   value NONMEM prints. The "%CV" heading is a labelling slip, not a
   transformation.

The model files therefore use `etalcl ~ 0.307`, `etaemax ~ 0.150`,
`etaet50 ~ 0.486` (variances, as `ini()` expects), `propSd <- sqrt(0.155) =
0.394` and `addSd <- sqrt(190) = 13.78` (standard deviations, as `ini()`
expects).

## Errata and discrepancies in the source

- **Covariate normalizing constants.** The printed final equation in Results
  3.2 normalizes by 41.5 years and 58.6 kg, which are the Table 1 means. The
  NONMEM control stream in Text S2 writes `AGE / 41.4` and `BW / 58.5`. The
  model file uses the published equation's 41.5 and 58.6; the difference is
  under 0.3% and affects no conclusion, but the two source locations do not
  agree.
- **First-week dose.** Table S5 lists the week-1 dose as 20.0 mg/day, yet the
  published first-week AUC (mean 2472-2765 ng*h/mL, reference 2610.43) is
  reproduced by 10 mg/day and overshot roughly two-fold by 20 mg/day. Methods
  2.1 resolves this: 10-20 mg/day is the *initial* dose and the first increase
  comes after a week, so the Table S5 row is the dose reached at week 1. The
  PKNCA comparison above shows both arms.
- **Implied half-life.** `Vd/F` fixed at 1010 L with `CL/F` estimated at
  23.6 L/h gives a terminal half-life of 30 h, against the roughly 21 h that
  Methods 2.1 cites from the literature. Both values are in the paper; the
  discrepancy follows from fixing the volume, which the authors do because the
  sparse trough data cannot support estimating it (Discussion).
- **No erratum.** A Europe PMC search on the DOI and on the first author
  returned a single record with no correction notice.

## Simulation assumptions

- Age, body weight and week-1 MADRS are drawn from independent truncated normal
  distributions matching the published means, standard deviations and ranges
  (Tables 1 and 2). The paper reports no correlations among them, and the mild
  sex-weight correlation it does mention is not reproducible from the published
  summaries.
- The daily-dose mixture for the steady-state cohort ({10, 20, 30, 40} mg/day
  with probabilities 0.35/0.30/0.13/0.22) was chosen to match the published
  mean of 22.3 mg/day and SD of 11.1; the paper reports only those two moments,
  not the dose histogram.
- Every patient is given a stable dose for 30 days before the steady-state
  sample. The real cohort was titrated weekly, so simulated exposure at a given
  final dose is a slight overestimate of the accumulated exposure of a patient
  who reached that dose recently.
- `AUC_PAROX` is generated from the popPK model with the clearance eta
  suppressed, because that is how the paper generated it (see Part 1). Sampling
  the eta as well widens the exposure distribution far beyond the observed
  1642-5814 ng*h/mL and drives the `-8.56` exponent into a regime where `Emax`
  diverges. **The PK-PD model must not be simulated outside the observed
  exposure range.**
- Proportional inter-individual variability on `ET50` with the published
  variance of 0.486 admits non-positive draws -- about 6% of a 200-patient
  simulation, reported above. Such a subject has a pole at `time = -ET50`
  inside the six-week window, which makes both its trajectory and any mean
  taken across it meaningless, so the summaries in Part 2 are computed over
  the subjects with a positive draw. This is the authors' structure reproduced
  faithfully rather than a transcription choice: the NONMEM `$PRED` block in
  Text S3 writes `(1 + ETA(2))` on ET50, and the variance itself is weakly
  identified (RSE 55%, shrinkage 32%, bootstrap 95% CI 0.0315-1.46). Downstream
  users simulating this model must screen for it.
- The paper's own MADRS observations are the enhancement rate relative to each
  patient's baseline. The simulation therefore predicts the enhancement rate
  directly and does not need a baseline MADRS distribution; the baseline score
  defines the endpoint, not a covariate.
- No parameter value came from anywhere other than the paper's text, printed
  equations, Supporting Information tables and NONMEM control streams. Nothing
  was tuned to improve agreement with any published quantity.
