---
title: "Vancomycin in pediatric liver transplant recipients (Shoji 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin in pediatric liver transplant recipients (Shoji 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Shoji_2021_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Shoji_2021_vancomycin"))$description`
* Article: <https://doi.org/10.1128/Spectrum.00460-21> (open access,
  [PMC8510181](https://europepmc.org/article/PMC/PMC8510181))
* Supplement (Fig. S1 goodness-of-fit plots, Table S1 dosing simulations by
  creatinine clearance): `spectrum00460-21_supp_1_seq2.pdf`, published
  with the open-access article (PMC8510181).

Shoji 2021 is a single-center retrospective population-PK analysis of
intravenous vancomycin in children after liver transplantation, followed by
Monte Carlo dosing simulations. The final model is a one-compartment model with
allometric body-weight scaling, and with serum creatinine and the number of days
from liver transplantation to the start of vancomycin (DFLT) as power-function
covariates on clearance.

# Population

161 pediatric liver transplant recipients (270 vancomycin treatment episodes,
1,158 serum concentrations) treated at the National Center for Child Health and
Development, Tokyo, between 2006 and 2014 (Shoji 2021 Table 1). Median age 13.3
months (IQR 7.6-53.5, range 1-186 months); median weight 9.1 kg (IQR 6.8-16.2,
range 3.1-61.0); 55.3% female. Underlying diseases: biliary atresia 52.2%,
metabolic disease 18.6%, fulminant hepatitis 16.1%, liver cirrhosis and liver
fibrosis 4.3% each, vascular abnormalities 2.5%, liver tumor 1.9%. Median serum
creatinine 0.16 mg/dL (IQR 0.12-0.23, range 0.06-5.43); median DFLT 17 days
(IQR 6-31, range 0-357). All patients received tacrolimus (median trough 8.2
ug/mL). Patients on renal replacement therapy were excluded. The median dose
was 15.0 mg/kg (IQR 14.0-15.0), usually every 6 h (ages 1 month to 12 years) or
every 8 h (13-17 years).

The same information is available programmatically via
`readModelDb("Shoji_2021_vancomycin")()$population`.

# Source trace

Each `ini()` entry in `inst/modeldb/specificDrugs/Shoji_2021_vancomycin.R`
carries an in-file comment giving its origin; the table collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (theta_CL) | 0.29 L/h/kg^0.75 | Table 3 final model, RSE 4.13%; bootstrap median 0.29 (95% CI 0.26-0.32) |
| `lvc` (theta_V) | 1.00 L/kg | Table 3 final model, RSE 5.97%; bootstrap median 1.00 (95% CI 0.89-1.14) |
| `e_creat_cl` (theta_sCr) | -0.70 | Table 3 final model, RSE 8.23%; bootstrap median -0.69 (95% CI -0.79 to -0.57) |
| `e_pod_cl` (theta_DFLT) | -0.09 | Table 3 final model, RSE 34.8%; bootstrap median -0.09 (95% CI -0.15 to -0.03) |
| `e_wt_cl` | 0.75, fixed | Methods ("scaled allometrically by subject weight (weight^0.75)"); Table 3 footnote b |
| `e_wt_vc` | 1, fixed | Methods ("scaled by subject weight (weight^1.0)"); Table 3 footnote c |
| `etalcl`, `etalvc` | 0, fixed | Exponential IIV declared in the Results equations; final-model magnitudes not printed (see Errata) |
| `propSd`, `addSd` | 0, fixed | Results: "an additive/proportional error model was selected"; final-model magnitudes not printed (see Errata) |
| `d/dt(central)` | n/a | Results: "A one-compartment model was selected as the structural model" |
| `cl <- ...` | n/a | Results: `TVCL(l/h) = CLpop x (sCr/0.16)^theta_sCr x (DFLT/17)^theta_DFLT x wt^0.75 x exp(eta_CL)`; Table 3 footnote b |
| `vc <- ...` | n/a | Results: `TVVi = Vpop x wt x exp(eta_V)`; Table 3 footnote c |
| Reference sCr 0.16 mg/dL, DFLT 17 days | n/a | Table 1 cohort medians |

The final-model equations are typeset as images in the article; the forms above
were read from the publisher's equation image `spectrum.00460-21-m001.jpg`,
deposited with the article, and match Table 3 footnote b.

# Check 1 -- typical values against the published post hoc averages

Shoji 2021 Results report post hoc average clearance and volume of 0.18 +/-
0.08 L/h/kg and 1.01 +/- 0.35 L/kg. At the cohort medians (9.1 kg, sCr 0.16
mg/dL, DFLT 17 days), the covariate terms for sCr and DFLT are exactly 1, so the
typical clearance per kg is `0.29 * 9.1^-0.25` and the typical volume per kg is
exactly theta_V.

```{r typical-values}
mod <- readModelDb("Shoji_2021_vancomycin") |> rxode2::zeroRe()

typical_cl_per_kg <- 0.29 * 9.1^0.75 / 9.1
typical_v_per_kg <- 1.00

# Solve the packaged model at the medians and read back its own cl and vc.
ev_med <- data.frame(
  id = 1L, time = c(0, 1), amt = c(15 * 9.1, 0), rate = c(15 * 9.1, 0),
  evid = c(1L, 0L), cmt = "central", WT = 9.1, CREAT = 0.16, POD = 17
)
sim_med <- rxode2::rxSolve(mod, events = ev_med) |> as.data.frame()

stopifnot(
  abs(sim_med$cl[1] / 9.1 - typical_cl_per_kg) < 1e-10,
  abs(sim_med$vc[1] / 9.1 - typical_v_per_kg) < 1e-10,
  # The typical clearance of a median child sits within one post hoc SD of the
  # published post hoc average (0.18 +/- 0.08 L/h/kg) ...
  abs(typical_cl_per_kg - 0.18) < 0.08,
  # ... and the typical volume is within 0.01 L/kg of it (1.01 +/- 0.35 L/kg).
  abs(typical_v_per_kg - 1.01) < 0.02
)

tibble(
  Quantity = c("CL (L/h/kg)", "V (L/kg)"),
  `Model, median child` = signif(c(typical_cl_per_kg, typical_v_per_kg), 3),
  `Shoji 2021 post hoc mean (SD)` = c("0.18 (0.08)", "1.01 (0.35)")
) |>
  knitr::kable()
```

# Replicating Figure 1 -- clearance against serum creatinine by DFLT

Figure 1 of Shoji 2021 plots post hoc clearance (L/h/kg) against serum
creatinine on a log axis, separately for episodes with DFLT < 14 days and >= 14
days. Without the individual random effects the model gives the typical curves
below, drawn for a 9.1 kg child at DFLT 7 days and 30 days (roughly the centres
of the two groups). The published scatter spans about 0.02-0.42 L/h/kg and falls
steeply between sCr 0.1 and 0.5 mg/dL, the shape of a power function with
exponent -0.70.

```{r figure-1, fig.cap = "Typical clearance against serum creatinine for a 9.1 kg child. Replicates the shape of Figure 1 of Shoji 2021 (which plots post hoc estimates)."}
fig1 <- expand.grid(
  CREAT = exp(seq(log(0.05), log(5), length.out = 200)),
  POD = c(7, 30)
) |>
  mutate(
    cl_per_kg = 0.29 * 9.1^0.75 * (CREAT / 0.16)^-0.70 * (POD / 17)^-0.09 / 9.1,
    group = ifelse(POD < 14, "DFLT < 14 days (7 days shown)", "DFLT >= 14 days (30 days shown)")
  )

# The DFLT effect: going from day 7 to day 30 lowers clearance by a fixed factor
# at every creatinine value.
pod_ratio <- (7 / 17)^-0.09 / (30 / 17)^-0.09
stopifnot(abs(pod_ratio - (30 / 7)^0.09) < 1e-12)
pod_ratio

ggplot(fig1, aes(CREAT, cl_per_kg, linetype = group)) +
  geom_line() +
  scale_x_log10(breaks = c(0.05, 0.5, 5)) +
  coord_cartesian(ylim = c(0, 0.5)) +
  labs(x = "Serum creatinine (mg/dL)", y = "Vancomycin clearance (L/h/kg)", linetype = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The two curves differ by only `r signif(100 * (pod_ratio - 1), 2)`%, which is
all a DFLT exponent of -0.09 can produce between day 7 and day 30. The
published figure separates the groups more than that, because the post hoc
estimates also carry the random effects and the creatinine values differ between
the groups.

# Virtual cohort

The article does not publish its individual data. The cohort below is built from
the Table 1 medians and interquartile ranges, each covariate taken as log-normal
(median = the published median, log-scale SD from the published IQR) and
truncated to the published range. Because the packaged model has no
between-subject variability, a simulation is a deterministic function of the
covariates, so the cohort is a fixed quantile grid rather than a random draw and
every number below is reproducible on any machine. DFLT is truncated below at 1
day, because the covariate term is infinite at 0 (see Errata). The covariates
are taken as independent; the article does not report their correlations.

```{r cohort}
n_per_arm <- 100L
u <- ppoints(n_per_arm)
i <- seq_len(n_per_arm)
shift <- function(k) ((i + k - 1L) %% n_per_arm) + 1L

# Truncated log-normal quantile: median m, IQR (q1, q3), range (lo, hi).
qtrunc_lnorm <- function(p, m, q1, q3, lo, hi) {
  sdlog <- log(q3 / q1) / (2 * qnorm(0.75))
  plo <- plnorm(lo, log(m), sdlog)
  phi <- plnorm(hi, log(m), sdlog)
  qlnorm(plo + p * (phi - plo), log(m), sdlog)
}

covariates <- tibble(
  cohort_id = i,
  WT = qtrunc_lnorm(u, 9.1, 6.8, 16.2, 3.1, 61.0), # Table 1 wt, kg
  CREAT = qtrunc_lnorm(u[shift(33L)], 0.16, 0.12, 0.23, 0.06, 5.43), # Table 1 sCr, mg/dL
  POD = qtrunc_lnorm(u[shift(66L)], 17, 6, 31, 1, 357) # Table 1 DFLT, days
)

stopifnot(
  abs(median(covariates$WT) - 9.1) / 9.1 < 0.1,
  abs(median(covariates$CREAT) - 0.16) / 0.16 < 0.1,
  abs(median(covariates$POD) - 17) / 17 < 0.1,
  min(covariates$POD) >= 1
)

summary(covariates[, c("WT", "CREAT", "POD")])
```

# Simulation

The three regimens of Shoji 2021 Table 4 are simulated for 7 days as 1-hour
infusions, which is long enough to reach steady state in every cohort member.
The article does not state the infusion duration; 1 hour is an assumption.

```{r simulate}
regimens <- tibble(
  regimen = c("15 mg/kg q8h", "15 mg/kg q6h", "20 mg/kg q6h"),
  dose_mgkg = c(15, 15, 20),
  tau = c(8, 6, 6)
)

make_arm <- function(k) {
  r <- regimens[k, ]
  dose_times <- seq(0, 168 - r$tau, by = r$tau)
  obs_times <- sort(unique(c(seq(144, 168, by = 0.05))))
  cov_k <- covariates |> mutate(id = cohort_id + (k - 1L) * n_per_arm)
  doses <- tidyr::crossing(cov_k, time = dose_times) |>
    mutate(amt = r$dose_mgkg * WT, rate = amt / 1, evid = 1L)
  obs <- tidyr::crossing(cov_k, time = obs_times) |>
    mutate(amt = 0, rate = 0, evid = 0L)
  bind_rows(doses, obs) |>
    mutate(cmt = "central", regimen = r$regimen, tau = r$tau) |>
    arrange(id, time, desc(evid))
}
events <- bind_rows(lapply(seq_len(nrow(regimens)), make_arm))

sim <- rxode2::rxSolve(
  mod,
  events = as.data.frame(events),
  keep = c("regimen", "tau", "WT", "CREAT", "POD")
) |>
  as.data.frame()

stopifnot(
  length(unique(sim$id)) == 3L * n_per_arm,
  all(is.finite(sim$Cc)),
  all(sim$Cc > 0)
)
```

# Typical steady-state profiles

The median child's steady-state profile over the last dosing interval of each
regimen. Figure 2 of Shoji 2021 (the VPC, which pools every dose level and
dosing interval against time after dose) has a simulated median of about 18
ug/mL just after the dose, falling to about 10 ug/mL by 5 h and about 5 ug/mL
by 10 h. The typical peaks and troughs, tabulated below the figure, lie inside
that band.

```{r typical-profile, fig.cap = "Typical steady-state concentration-time profiles for a 9.1 kg child with sCr 0.16 mg/dL and DFLT 17 days."}
ev_typ <- bind_rows(lapply(seq_len(nrow(regimens)), function(k) {
  r <- regimens[k, ]
  bind_rows(
    tibble(time = seq(0, 168 - r$tau, by = r$tau), amt = r$dose_mgkg * 9.1, evid = 1L),
    tibble(time = seq(168 - r$tau, 168, by = 0.05), amt = 0, evid = 0L)
  ) |>
    mutate(id = k, rate = amt, cmt = "central", WT = 9.1, CREAT = 0.16, POD = 17, regimen = r$regimen)
}))
sim_typ <- rxode2::rxSolve(mod, events = as.data.frame(ev_typ), keep = "regimen") |>
  as.data.frame() |>
  group_by(regimen) |>
  mutate(tad = time - min(time)) |>
  ungroup()

ggplot(sim_typ, aes(tad, Cc, colour = regimen)) +
  geom_line() +
  labs(x = "Time after dose (h)", y = "Vancomycin (ug/mL)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")

typ_peak_trough <- sim_typ |>
  group_by(regimen) |>
  summarise(peak = max(Cc), trough = Cc[which.max(tad)], .groups = "drop")

# Typical peaks and troughs within the 5-35 ug/mL span of the Figure 2 median.
stopifnot(
  all(typ_peak_trough$peak > 15 & typ_peak_trough$peak < 35),
  all(typ_peak_trough$trough > 4 & typ_peak_trough$trough < 15)
)

typ_peak_trough |>
  dplyr::rename(Regimen = regimen, `Peak (ug/mL)` = peak, `Trough (ug/mL)` = trough) |>
  knitr::kable(digits = 1)
```

# PKNCA -- steady-state AUC over 24 h

PKNCA computes each subject's AUC over the final 24 h (144-168 h) and the
trough at 168 h. For a linear one-compartment model at steady state, AUC over a
24-h window is exactly the daily dose divided by clearance. Both sides use the
same parameters, so the difference is only the numerical error of the
trapezoidal rule and a tight bound is the right gate.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | regimen + id,
  concu = "ug/mL", timeu = "h"
)
dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, regimen)
dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_df), amt ~ time | regimen + id,
  doseu = "mg"
)
intervals <- data.frame(start = 144, end = 168, auclast = TRUE, cmax = TRUE, clast.obs = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(regimen, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

per_subject <- sim |>
  group_by(id) |>
  slice(1) |>
  ungroup() |>
  select(id, cl, WT, CREAT, POD) |>
  left_join(events |> distinct(id, regimen, tau), by = "id") |>
  left_join(regimens |> select(regimen, dose_mgkg), by = "regimen")

auc_chk <- nca_wide |>
  left_join(per_subject, by = c("id", "regimen")) |>
  mutate(
    auc24_closed = dose_mgkg * WT * 24 / tau / cl,
    auc_relerr = abs(auclast - auc24_closed) / auc24_closed
  )

stopifnot(
  nrow(auc_chk) == 3L * n_per_arm,
  all(is.finite(auc_chk$auc_relerr)),
  max(auc_chk$auc_relerr) < 0.005
)
max(auc_chk$auc_relerr)
```

# Comparison with Table 4 -- target attainment by regimen

Shoji 2021 Table 4 reports the percentage of 1,000 simulated patients reaching a
trough above 10 ug/mL and an AUC24/MIC of at least 400. Those percentages depend
on the between-subject variability and on the covariate joint distribution of
the authors' simulation, neither of which is reproducible here. What the
packaged model does determine is the typical (median-covariate) child, and a
regimen should reach a target for the typical child exactly when more than half
of the published population reaches it. That is checked below for the AUC24/MIC
target, which the typical child clears or misses by at least 10% in every row.

```{r table-4}
typ_auc24 <- regimens |>
  mutate(
    cl_typ = 0.29 * 9.1^0.75,
    auc24 = dose_mgkg * 9.1 * 24 / tau / cl_typ,
    # Shoji 2021 Table 4, 'Total' column.
    pub_auc_mic05 = c(100.0, 100.0, 100.0),
    pub_auc_mic1 = c(7.4, 16.3, 88.1)
  )

stopifnot(
  # MIC 0.5: every regimen reaches AUC24 >= 200 for the typical child (Table 4: 100%).
  all(typ_auc24$auc24 >= 200),
  # MIC 1: the typical child reaches AUC24 >= 400 exactly where Table 4 exceeds 50%.
  identical(typ_auc24$auc24 >= 400, typ_auc24$pub_auc_mic1 > 50)
)

# Cohort attainment (covariate spread only, no between-subject variability).
cohort_pta <- auc_chk |>
  group_by(regimen) |>
  summarise(
    auc_mic05 = 100 * mean(auclast / 0.5 >= 400),
    auc_mic1 = 100 * mean(auclast / 1 >= 400),
    trough10 = 100 * mean(clast.obs > 10),
    .groups = "drop"
  )

typ_auc24 |>
  left_join(cohort_pta, by = "regimen") |>
  mutate(
    pub_trough10 = c(31.2, 54.0, 71.6),
    auc24 = round(auc24)
  ) |>
  select(regimen, auc24, pub_auc_mic1, auc_mic1, pub_auc_mic05, auc_mic05, pub_trough10, trough10) |>
  dplyr::rename(
    Regimen = regimen,
    `Typical AUC24 (ug*h/mL)` = auc24,
    `AUC24/MIC>=400, MIC 1: Table 4 (%)` = pub_auc_mic1,
    `AUC24/MIC>=400, MIC 1: cohort (%)` = auc_mic1,
    `AUC24/MIC>=400, MIC 0.5: Table 4 (%)` = pub_auc_mic05,
    `AUC24/MIC>=400, MIC 0.5: cohort (%)` = auc_mic05,
    `Trough>10: Table 4 (%)` = pub_trough10,
    `Trough>10: cohort (%)` = trough10
  ) |>
  knitr::kable(digits = 1)
```

The cohort percentages come from covariate spread alone and are shown for
orientation only; they are not expected to match the published values.

# Assumptions and deviations

* **Between-subject and residual variability are zero.** Shoji 2021 Table 3
  prints `% omega CL` = 46.47 (RSE 2.88%), `% omega Vc` = 48.14 (RSE 6.55%) and a
  proportional residual error of 56.5% (RSE 4.02%) only in the *base-model*
  column; the final-model and bootstrap columns are blank for these rows. The
  base model has no serum-creatinine term, and serum creatinine alone lowered the
  objective function by 435 points (Table 2), so the base-model clearance
  variability overstates the final model's. Combining final-model fixed effects
  with base-model variances would give a model the authors never fitted, so the
  final-model variances are encoded as `fixed(0)`. Users who need stochastic
  simulations can supply their own values; the base-model values above are
  upper-bound guides for clearance and plausible guides for volume. Final-model
  eta shrinkage was 22.0% (CL) and 27.0% (V), and epsilon shrinkage 4.55%.
* **Residual error form.** The Results state that "an additive/proportional
  error model was selected", but Table 3 prints only a proportional term (for
  the base model). The model carries both `propSd` and `addSd`, each fixed at 0.
* **DFLT = 0.** Table 1 reports DFLT from 0 days, but with a negative exponent
  the printed term `(DFLT/17)^-0.09` is infinite at 0. The article does not say
  how episodes starting on the day of transplant were coded. The model is
  encoded as printed and needs `POD > 0`; this vignette uses `POD >= 1`.
* **DFLT is a per-episode constant.** It is defined as the days from
  transplantation to the *first* day of vancomycin, and it is stored in the
  canonical `POD` column. Supply the value at the start of the treatment
  episode on every row, not a day count that advances with each observation.
* **Allometric weight enters unnormalised.** theta_CL is per kg^0.75 and theta_V
  per kg, as in the published equations, not relative to a 70 kg reference.
* **Infusion duration.** Not stated in the article; 1 hour is used here.
* **Virtual cohort.** Log-normal covariates matched to the Table 1 medians and
  IQRs, truncated to the published ranges, and taken as independent.

## Errata noted in the source

* Table 3 footnote c prints `V (l) = theta_V x wt x exp(eta_CL)`; the Results
  equation has `exp(eta_V)`, which is used here.
* The Results text says "only 16% of the patients with a DFLT of <14 days
  achieved an AUC24/MIC of >=400 for an MIC of 1 ug/ml with ... 15 mg/kg q6h",
  but Table 4 gives 3.3% for DFLT < 14 days and 16.3% for the total population
  (the Abstract's "16%" refers to all patients).
* Table 4, 20 mg/kg q6h, trough > 10 ug/mL: the total (71.6%) is lower than both
  subgroups (76.3% and 100.0%), which cannot happen for a weighted average of
  them. At least one of the three cells is misprinted.
