---
title: "Efavirenz (Siccardi 2012)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efavirenz (Siccardi 2012)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Siccardi_2012_efavirenz"))$reference`
* Description: `r rxode2::rxode(readModelDb("Siccardi_2012_efavirenz"))$description`
* Article: <https://doi.org/10.1038/clpt.2012.61>

Siccardi 2012 is a two-arm paper. It builds a whole-body in-vitro-in-vivo
extrapolation (IVIVE) model of efavirenz in the Simcyp Population-Based
Simulator (version 10.1), and it independently fits a NONMEM population PK
model plus two binary logistic exposure-response models to clinical data. The
IVIVE arm is then used to simulate 600, 400 and 200 mg once-daily regimens in
virtual subjects carrying each CYP2B6 c.516G>T genotype.

**Only the second arm is packaged here.** The IVIVE arm is a platform
whole-body PBPK model whose system parameters (organ volumes, blood flows,
enzyme abundances, ISEF and fu-mic corrections) come from the Simcyp North
European Caucasian population library rather than from the paper, and no ODE
system is written out; it is therefore not reproducible from any on-disk
source. What *is* fully specified is the population PK model (Table 2 and the
Results section) and the two logistic PK/PD relationships (Equation 1 and the
"PD of dose reduction" section), and those are what
`modellib("Siccardi_2012_efavirenz")` contains. Table 1 and Table 3 of the
paper are simulation *outputs* of the IVIVE arm, so they are used below as
reference values with that provenance stated explicitly, not as targets the
population PK arm is expected to hit exactly.

# Population

The population PK model was estimated from 202 plasma efavirenz samples in 157
subjects: nine healthy volunteers from the Royal Free NHS Trust, London (six
samples each) and 148 HIV-1-positive patients from the German KompNet cohort
(one random sample each). All patients were at steady state on an
efavirenz-based regimen with two nucleoside reverse-transcriptase inhibitors;
patients on a boosted protease inhibitor or on any other interacting drug were
excluded. 110 of 157 subjects (70%) were men, the median age was 41 years
(IQR 22-67) and the median body weight was 70.5 kg (IQR 49-98 kg). Observed
concentrations spanned 530-26,020 ng/mL. By genotype, 41 subjects were 516GT
and 10 were 516TT, leaving 106 as 516GG wild-type (Siccardi 2012 Results,
"Population PK model").

The two logistic exposure-response models were fitted to **different** cohorts,
drawn from Csajka 2003 and Marzolini 2001 (Siccardi 2012 Results, "PD of dose
reduction"). Viral suppression: 93 patients, 14 with therapeutic failure, 65
men (68%), median age 38 years (IQR 33-44), median weight 66 kg (IQR 60-75).
CNS side effects: 121 patients, 30 with CNS side effects, 88 men (73%), median
age 39 years (IQR 35-45), median weight 66 kg (IQR 57-75). Both cohorts were
co-treated with the older nucleoside reverse-transcriptase inhibitors
zidovudine, stavudine and didanosine, which the paper's Discussion flags as a
limitation for extrapolation to modern tenofovir/emtricitabine backbones.

The same information is available programmatically from the model's
`population` metadata (`readModelDb("Siccardi_2012_efavirenz")()$population`).

```{r population}
pop <- readModelDb("Siccardi_2012_efavirenz")()$population
str(pop[c("species", "n_subjects", "weight_median", "genotype_distribution")])
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Siccardi_2012_efavirenz.R`. The table
below collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | 0.36 1/h | Table 2, Population PK column, ka (RSE 76%) |
| `lcl` | 13.3 L/h | Table 2, Population PK column, CL/F 516GG (RSE 7%); also Results, "Population PK model" |
| `lvc` | 4.3 L/kg x 70.5 kg = 303 L | Table 2, Population PK column, Vd = 4.3 (RSE 10%); cohort median weight from Results, "Population PK model" |
| `lfdepot` | 1 (fixed) | Structural anchor: the paper reports apparent CL/F and V/F only |
| `e_516gt_cl` | 9.7 - 13.3 = -3.6 L/h | Table 2, Population PK column, CL/F 516GT = 9.7 L/h (RSE 12%); form from Methods, "Population PK model at standard regimens" |
| `e_516tt_cl` | 2.6 - 13.3 = -10.7 L/h | Table 2, Population PK column, CL/F 516TT = 2.6 L/h (RSE 41%) |
| `logite0_supp` | -8.38 | Results, "PD of dose reduction", below Equation 1 (SE 100%) |
| `e_c816_supp` | 3.12 | Results, "PD of dose reduction" (SE 42%); corroborated by OR = 22.6 for log10 C8-16h |
| `logite0_cns` | -6.65 | Results, "PD of dose reduction" (SE 43%) |
| `e_c816_cns` | 1.68 | Results, "PD of dose reduction" (SE 51%); corroborated by OR = 5.4 for log10 C8-16h |
| `etalcl` | log(0.574^2 + 1) = 0.28478 | Results, "Population PK model": 57.4% CV on CL/F (base model), exponential IIV model |
| `propSd` | 0 (fixed) | Results, "Population PK model": proportional structure stated, magnitude never reported |
| One-compartment first-order absorption | n/a | Results, "Population PK model": "A one-compartment model described the data better than a two-compartment model" |
| `tvcl = CL0 + theta1 * GT + theta2 * TT` | n/a | Methods, "Population PK model at standard regimens", final paragraph |
| `p = 1 / (1 + e^-(A + B log10 C8-16h))` | n/a | Equation 1, page 497 |
| `C8-16h` = mean concentration 8-16 h post-dose | n/a | Methods, "PK/PD analysis", final paragraph |

# The logistic link is natural-log, not base-10

Equation 1 of the paper is written with a natural exponential,

$$p = \frac{1}{1 + e^{-(A + B \log_{10} C_{8-16h})}},$$

but the sentence immediately after it says the relationship fed into the Simcyp
PD module was `log10(p / (1 - p)) = A + B x`. Those two statements are not the
same model, and the choice matters enormously: at 2,000 ng/mL the natural-logit
form gives a suppression probability of 0.87 and the base-10 form gives 0.994.

The paper's own reported odds ratios settle it. An odds ratio per unit of
`log10 C8-16h` is `exp(B)` under a natural logit and `10^B` under a base-10
logit, and the paper reports OR = 22.6 for viral suppression and OR = 5.4 for
CNS side effects. This is an exact, simulation-free gate on the encoding.

```{r gate-odds-ratio}
mod <- readModelDb("Siccardi_2012_efavirenz")
th  <- rxode2::rxode(mod)$theta

or_check <- tibble::tibble(
  endpoint  = c("Viral suppression", "CNS side effects"),
  B         = c(th[["e_c816_supp"]], th[["e_c816_cns"]]),
  or_natural = exp(B),
  or_base10  = 10^B,
  or_published = c(22.6, 5.4)
) |>
  dplyr::mutate(
    pct_diff_natural = 100 * (or_natural - or_published) / or_published,
    pct_diff_base10  = 100 * (or_base10  - or_published) / or_published
  )

or_check |>
  dplyr::rename(
    "Endpoint"                = endpoint,
    "B"                       = B,
    "exp(B)"                  = or_natural,
    "10^B"                    = or_base10,
    "Published OR"            = or_published,
    "% diff, natural logit"   = pct_diff_natural,
    "% diff, base-10 logit"   = pct_diff_base10
  ) |>
  knitr::kable(digits = 3, caption =
    "Equation 1 read as a natural logit reproduces both published odds ratios.")

# B is printed to three significant figures, so the rounding interval on
# exp(B) is exp(B +/- 0.005), i.e. +/- 0.5%. Require the natural-logit form to
# land inside that and the base-10 form to fail it.
stopifnot(
  all(abs(or_check$pct_diff_natural) < 1),
  all(abs(or_check$pct_diff_base10)  > 100)
)
```

The natural-logit reading is inside the printed rounding interval for both
endpoints; the base-10 reading is wrong by a factor of 58 and 7 respectively.
The model therefore encodes Equation 1 as printed, and the following sentence
is treated as an erratum.

# Virtual cohort

Original observed data are not publicly available. The simulations below use
nine arms (three CYP2B6 516 genotypes x three once-daily doses), matching the
paper's simulation design.

```{r cohort}
NSUB  <- 200L                # per arm; the skill cap
NDAY  <- 90L                 # days of once-daily dosing before the readout
TSS   <- 24 * (NDAY - 1)     # time of the final dose, i.e. the steady-state interval
SEED  <- 20120701L

arms <- tidyr::expand_grid(
  genotype = c("516GG", "516GT", "516TT"),
  dose_mg  = c(600, 400, 200)
) |>
  dplyr::mutate(
    tcount    = rep(0:2, each = 3),   # CYP2B6 516 T-allele count
    treatment = paste(genotype, dose_mg, "mg")
  )

arms |>
  dplyr::rename(
    "Genotype"                = genotype,
    "Once-daily dose (mg)"    = dose_mg,
    "516 T-allele count"      = tcount,
    "Arm label"               = treatment
  ) |>
  knitr::kable(caption = "Simulated arms (Siccardi 2012 Methods, 'Simulation design').")
```

90 days of dosing is not extravagance. The 516TT typical clearance is 2.6 L/h,
so with a 303 L volume the typical terminal half-life is 81 h, and a subject in
the lower tail of the 57.4% CV clearance distribution reaches 200 h. Steady
state is set by the slowest subjects, not the typical one, so the duration is
verified rather than assumed below.

# Typical-value steady state

The model carries two bookkeeping states that reconstruct the paper's PD
driver: `auc_8_16` integrates concentration only while `tad()` lies between 8
and 16 h, and `t_8_16` accumulates the elapsed time inside that window, so
their ratio (`c816`) is the window mean. Because both run from the start of the
simulation, `c816` is the running mean over *every* elapsed window. To read the
value for one steady-state interval exactly, the two states are restarted at
the top of that interval: solve to steady state, then re-solve a single
interval with the disposition states carried over and the two accumulators set
back to zero.

```{r typical, message = FALSE}
mod_typ <- mod |> rxode2::zeroRe()

typical <- lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]

  # Stage 1 -- run to steady state and capture the pre-dose disposition states.
  ev1 <- rxode2::et(amt = a$dose_mg, ii = 24, until = TSS) |>
    rxode2::et(TSS + 24) |>
    as.data.frame() |>
    dplyr::mutate(SNP_CYP2B6_RS3745274_T_COUNT = a$tcount)
  s1 <- rxode2::rxSolve(mod_typ, ev1, returnType = "data.frame")
  trough <- s1[nrow(s1), ]

  # Stage 2 -- one interval with the accumulators restarted at zero.
  ev2 <- rxode2::et(amt = a$dose_mg) |>
    rxode2::et(seq(0, 24, by = 0.05)) |>
    as.data.frame() |>
    dplyr::mutate(SNP_CYP2B6_RS3745274_T_COUNT = a$tcount)
  s2 <- rxode2::rxSolve(
    mod_typ, ev2,
    inits = c(depot = trough$depot, central = trough$central,
              auc_8_16 = 0, t_8_16 = 0),
    returnType = "data.frame"
  )

  at16 <- s2[which.min(abs(s2$time - 16)), ]
  data.frame(
    treatment  = a$treatment,
    genotype   = a$genotype,
    dose_mg    = a$dose_mg,
    cl_paper   = c(13.3, 9.7, 2.6)[a$tcount + 1],
    window_h   = at16$t_8_16,
    c816       = at16$c816,
    c816_grid  = mean(s2$Cc[s2$time >= 8 & s2$time <= 16]),
    psupp      = at16$psupp,
    pcns       = at16$pcns,
    cmax       = max(s2$Cc),
    ctrough    = min(s2$Cc),
    auc_tau    = sum(diff(s2$time) *
                       (utils::head(s2$Cc, -1) + utils::tail(s2$Cc, -1)) / 2),
    ss_rel     = abs(utils::tail(s2$Cc, 1) - s2$Cc[1]) / s2$Cc[1]
  )
}) |>
  dplyr::bind_rows()
```

Four exact structural checks on the typical-value arms.

```{r gate-structural}
gates <- typical |>
  dplyr::mutate(
    # (1) steady state: the interval must open and close at the same value
    ss_rel        = ss_rel,
    # (2) the accumulator window really is 8 h wide
    window_err    = abs(window_h - 8),
    # (3) the accumulator ratio equals a direct grid mean over the window
    c816_rel      = abs(c816 - c816_grid) / c816_grid,
    # (4) steady-state average concentration equals the closed form
    #     Cavg = Dose / (CL/F * tau); dose in mg, CL in L/h, Cc in ng/mL
    cavg_closed   = dose_mg * 1000 / (cl_paper * 24),
    cavg_solved   = auc_tau / 24,
    cavg_rel      = abs(cavg_solved - cavg_closed) / cavg_closed
  )

gates |>
  dplyr::select(treatment, ss_rel, window_err, c816_rel,
                cavg_closed, cavg_solved, cavg_rel) |>
  dplyr::rename(
    "Arm"                          = treatment,
    "Steady state, rel. Cc(0) vs Cc(24)" = ss_rel,
    "Window width error (h)"       = window_err,
    "c816 vs grid mean, rel."      = c816_rel,
    "Cavg closed form (ng/mL)"     = cavg_closed,
    "Cavg from AUC (ng/mL)"        = cavg_solved,
    "Cavg rel. error"              = cavg_rel
  ) |>
  knitr::kable(digits = c(0, 10, 6, 8, 1, 1, 8), caption =
    "Exact structural gates on the typical-value steady-state interval.")

stopifnot(
  max(gates$ss_rel)     < 1e-5,   # steady state reached by day 89
  max(gates$window_err) < 1e-2,   # 8 h window, integrator-step resolution
  max(gates$c816_rel)   < 1e-4,   # accumulator ratio == direct grid mean
  max(gates$cavg_rel)   < 1e-4    # AUC over tau == Dose / (CL/F * tau)
)
```

The fourth gate is the one that pins the whole PK layer: it reproduces
`Dose / (CL/F * tau)` to better than one part in ten thousand for all nine
arms, which simultaneously confirms the three genotype clearances, the mg-to-ng
unit conversion in the observation equation, and that the volume was read as
303 L rather than 4.3 L (see the Errata below).

Next, the model's own `psupp` and `pcns` outputs must equal Equation 1
evaluated at the model's own `c816`, using the coefficients read back out of
the model object rather than retyped.

```{r gate-identity}
identity_chk <- typical |>
  dplyr::mutate(
    psupp_eq1 = 1 / (1 + exp(-(th[["logite0_supp"]] +
                                 th[["e_c816_supp"]] * log10(c816)))),
    pcns_eq1  = 1 / (1 + exp(-(th[["logite0_cns"]] +
                                 th[["e_c816_cns"]]  * log10(c816)))),
    d_supp    = abs(psupp - psupp_eq1),
    d_cns     = abs(pcns  - pcns_eq1)
  )
stopifnot(max(identity_chk$d_supp) < 1e-10, max(identity_chk$d_cns) < 1e-10)

typical |>
  dplyr::select(treatment, c816, psupp, pcns, cmax, ctrough) |>
  dplyr::rename(
    "Arm"                            = treatment,
    "C8-16h (ng/mL)"                 = c816,
    "P(viral suppression)"           = psupp,
    "P(CNS side effects)"            = pcns,
    "Cmax,ss (ng/mL)"                = cmax,
    "Ctrough,ss (ng/mL)"             = ctrough
  ) |>
  knitr::kable(digits = c(0, 1, 4, 4, 1, 1), caption =
    "Typical-value steady state and the two exposure-response probabilities.")
```

The typical-value probabilities are also a check against the clinical
thresholds the paper's Introduction quotes: a 1,000 ng/mL minimum effective
concentration and a 1,000-4,000 ng/mL therapeutic range.

```{r gate-thresholds}
thresh <- tibble::tibble(c816 = c(1000, 4000)) |>
  dplyr::mutate(
    psupp = 1 / (1 + exp(-(th[["logite0_supp"]] + th[["e_c816_supp"]] * log10(c816)))),
    pcns  = 1 / (1 + exp(-(th[["logite0_cns"]]  + th[["e_c816_cns"]]  * log10(c816))))
  )
thresh |>
  dplyr::rename(
    "C8-16h (ng/mL)"       = c816,
    "P(viral suppression)" = psupp,
    "P(CNS side effects)"  = pcns
  ) |>
  knitr::kable(digits = 3, caption =
    "Equation 1 across the therapeutic range quoted in the Introduction.")

# Monotonic and bracketing: suppression rises across the therapeutic window,
# CNS risk rises with it, and the concentration at which suppression is
# even-odds sits below the quoted minimum effective concentration.
c50_supp <- 10^(-th[["logite0_supp"]] / th[["e_c816_supp"]])
stopifnot(
  diff(thresh$psupp) > 0, diff(thresh$pcns) > 0,
  c50_supp < 1000
)
cat(sprintf("C8-16h at 50%% suppression probability: %.0f ng/mL\n", c50_supp))
```

# Figure 1 (right-hand panels): steady-state profiles by genotype

```{r figure-1}
prof <- lapply(which(arms$dose_mg == 600), function(i) {
  a <- arms[i, ]
  rxode2::rxSetSeed(SEED)
  ev <- rxode2::et(amt = a$dose_mg, ii = 24, until = TSS) |>
    rxode2::et(TSS + seq(0, 24, by = 0.5)) |>
    rxode2::et(id = seq_len(NSUB)) |>
    as.data.frame() |>
    dplyr::mutate(
      SNP_CYP2B6_RS3745274_T_COUNT = a$tcount,
      genotype                     = a$genotype
    )
  rxode2::rxSolve(mod, ev, keep = "genotype", returnType = "data.frame") |>
    dplyr::filter(time >= TSS) |>
    dplyr::mutate(tad_h = time - TSS)
}) |>
  dplyr::bind_rows()

prof |>
  dplyr::group_by(genotype, tad_h) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(tad_h, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~genotype) +
  scale_y_log10(limits = c(100, 30000)) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Time after dose (h)", y = "Efavirenz concentration (ng/mL)",
       title = "Steady-state efavirenz profiles, 600 mg once daily",
       caption = paste("Replicates the population-PK (right-hand) panels of",
                       "Figure 1 of Siccardi 2012. Line = median,",
                       "band = 5th to 95th percentile."))
```

The published panels use the same 100-30,000 ng/mL log axis and the same 0-24 h
window. The 516GG and 516GT panels overlay the observed 530-26,020 ng/mL data
range; the 516TT panel sits substantially higher than the published one, which
is the CL/F disagreement the paper itself flags and which is quantified below.

# Cohort exposure-response

For a cohort the accumulator states cannot be restarted per subject, so the
steady-state window mean is recovered by **differencing** them across the final
interval: `c816 = (auc_8_16(t+16) - auc_8_16(t)) / (t_8_16(t+16) - t_8_16(t))`,
where `t` is the time of the final dose. All nine arms share one seed, so each
subject carries the same clearance random effect across doses and genotypes
(common random numbers).

```{r cohort-sim}
cohort <- lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  rxode2::rxSetSeed(SEED)
  ev <- rxode2::et(amt = a$dose_mg, ii = 24, until = TSS) |>
    rxode2::et(c(TSS - 24, TSS, TSS + 16, TSS + 24)) |>
    rxode2::et(id = seq_len(NSUB)) |>
    as.data.frame() |>
    dplyr::mutate(SNP_CYP2B6_RS3745274_T_COUNT = a$tcount)

  rxode2::rxSolve(mod, ev, returnType = "data.frame") |>
    dplyr::filter(time %in% c(TSS - 24, TSS, TSS + 16, TSS + 24)) |>
    dplyr::select(id, time, Cc, auc_8_16, t_8_16) |>
    dplyr::group_by(id) |>
    dplyr::summarise(
      c816        = (auc_8_16[time == TSS + 16] - auc_8_16[time == TSS]) /
                    (t_8_16[time == TSS + 16]   - t_8_16[time == TSS]),
      window_h    = t_8_16[time == TSS + 16] - t_8_16[time == TSS],
      trough_prev = Cc[time == TSS - 24],
      trough_this = Cc[time == TSS],
      trough_next = Cc[time == TSS + 24],
      .groups     = "drop"
    ) |>
    dplyr::mutate(treatment = a$treatment, genotype = a$genotype,
                  dose_mg = a$dose_mg)
}) |>
  dplyr::bind_rows() |>
  dplyr::mutate(
    psupp = 1 / (1 + exp(-(th[["logite0_supp"]] + th[["e_c816_supp"]] * log10(c816)))),
    pcns  = 1 / (1 + exp(-(th[["logite0_cns"]]  + th[["e_c816_cns"]]  * log10(c816))))
  )

cat("cohort rows:", nrow(cohort), "\n")
```

Two gates on the cohort itself, before any comparison against the paper.

```{r gate-cohort}
# (1) Steady state by day 89 -- consecutive troughs must agree. Assert on the
#     centre and a robust quantile rather than the extreme: the slowest
#     subject is whichever one happened to draw the lowest clearance, and that
#     is not reproducible across rxode2 builds.
ss_rel <- with(cohort, abs(trough_next - trough_this) / trough_this)
cat(sprintf("trough-to-trough relative change: median %.2e, q90 %.2e, q99 %.2e\n",
            median(ss_rel), quantile(ss_rel, 0.90), quantile(ss_rel, 0.99)))
stopifnot(median(ss_rel) < 1e-4, quantile(ss_rel, 0.90) < 1e-3,
          quantile(ss_rel, 0.99) < 1e-2)

# (2) Dose linearity under common random numbers. Elimination is linear, so at
#     a fixed clearance C8-16h must scale exactly with dose. With the same
#     random effects drawn in every arm this becomes an exact per-subject
#     identity rather than a noisy comparison of medians.
lin <- cohort |>
  dplyr::select(genotype, id, dose_mg, c816) |>
  tidyr::pivot_wider(names_from = dose_mg, values_from = c816, names_prefix = "d") |>
  dplyr::mutate(r400 = d400 / d600 - 2 / 3, r200 = d200 / d600 - 1 / 3)
cat(sprintf("dose-linearity max deviation: 400/600 %.2e, 200/600 %.2e\n",
            max(abs(lin$r400)), max(abs(lin$r200))))
stopifnot(max(abs(lin$r400)) < 1e-3, max(abs(lin$r200)) < 1e-3,
          max(abs(cohort$window_h - 8)) < 1e-2)
```

## Comparison against Table 3 of the paper

Table 3 reports mean probabilities with 5th-95th percentiles, computed from the
**IVIVE** exposures. The population PK arm packaged here is a different model
of the same drug, so the comparison is a cross-arm one; the paper's own
acceptance criterion for that comparison (Results, "Population PK model") was a
less-than-twofold difference in the PK parameters.

```{r table-3-compare}
published_t3 <- tibble::tribble(
  ~genotype, ~dose_mg, ~psupp_pub, ~pcns_pub,
  "516GG",   600,      0.770,      0.253,
  "516GG",   400,      0.692,      0.207,
  "516GG",   200,      0.538,      0.131,
  "516GT",   600,      0.816,      0.302,
  "516GT",   400,      0.751,      0.241,
  "516GT",   200,      0.622,      0.176,
  "516TT",   600,      0.888,      0.361,
  "516TT",   400,      0.817,      0.291,
  "516TT",   200,      0.716,      0.202
)

t3 <- cohort |>
  dplyr::group_by(genotype, dose_mg) |>
  dplyr::summarise(
    psupp_sim = mean(psupp), psupp_p5 = quantile(psupp, 0.05),
    psupp_p95 = quantile(psupp, 0.95),
    pcns_sim  = mean(pcns),  pcns_p5  = quantile(pcns, 0.05),
    pcns_p95  = quantile(pcns, 0.95), .groups = "drop"
  ) |>
  dplyr::left_join(published_t3, by = c("genotype", "dose_mg")) |>
  dplyr::mutate(d_supp = psupp_sim - psupp_pub, d_cns = pcns_sim - pcns_pub) |>
  dplyr::arrange(genotype, dplyr::desc(dose_mg))

t3 |>
  dplyr::select(genotype, dose_mg, psupp_sim, psupp_pub, d_supp,
                pcns_sim, pcns_pub, d_cns) |>
  dplyr::rename(
    "Genotype"                = genotype,
    "Dose (mg)"               = dose_mg,
    "P(suppression), model"   = psupp_sim,
    "P(suppression), Table 3" = psupp_pub,
    "Difference"              = d_supp,
    "P(CNS), model"           = pcns_sim,
    "P(CNS), Table 3"         = pcns_pub,
    "Difference "             = d_cns
  ) |>
  knitr::kable(digits = 3, caption =
    "Cohort mean probabilities vs Table 3 of Siccardi 2012 (Table 3 was generated from the IVIVE arm).")
```

Three separable findings, in order of how well they land.

**1. CNS side effects, 516GG and 516GT: reproduced.** Across those six arms the
mean probability of CNS side effects agrees with Table 3 to within 0.01
absolute. The mean of a nonlinear function over a cohort depends on the *spread*
of exposures as well as its centre, so this also exercises the 57.4% CV
clearance variability and not only the typical value.

```{r gate-cns}
cns_ggt <- t3 |> dplyr::filter(genotype != "516TT")
cat(sprintf("CNS, 516GG + 516GT: mean |diff| %.4f, max |diff| %.4f\n",
            mean(abs(cns_ggt$d_cns)), max(abs(cns_ggt$d_cns))))
# Tolerance from the Monte-Carlo standard error: the per-arm SD of pcns is
# about 0.07 over 200 subjects, so the SE of each arm mean is about 0.005.
# Gate on the mean absolute difference (stable across seeds) with a max
# envelope at roughly four standard errors.
stopifnot(mean(abs(cns_ggt$d_cns)) < 0.015, max(abs(cns_ggt$d_cns)) < 0.030)
```

**2. Viral suppression: a systematic offset of about +0.08.** Every one of the
six 516GG/516GT arms overshoots Table 3 in the same direction and by a similar
amount, which is the signature of a wrong intercept rather than a wrong
exposure. Note that an exposure error would move both endpoints; the CNS
endpoint is on target.

```{r diagnose-suppression}
supp_ggt <- t3 |> dplyr::filter(genotype != "516TT")
cat(sprintf("Suppression, 516GG + 516GT: differences %s\n",
            paste(sprintf("%+.3f", supp_ggt$d_supp), collapse = " ")))
stopifnot(all(supp_ggt$d_supp > 0))   # systematic, one-directional

# What intercept would reproduce Table 3? Refit A on the nine arms holding
# B = 3.12 (which the odds-ratio gate above already confirmed).
obj <- function(a) {
  p <- cohort |>
    dplyr::mutate(p = 1 / (1 + exp(-(a + th[["e_c816_supp"]] * log10(c816))))) |>
    dplyr::group_by(genotype, dose_mg) |>
    dplyr::summarise(p = mean(p), .groups = "drop") |>
    dplyr::left_join(published_t3, by = c("genotype", "dose_mg"))
  sum((p$p - p$psupp_pub)^2)
}
a_fit <- optimize(obj, c(-14, -5))$minimum
cat(sprintf("printed intercept %.2f; intercept implied by Table 3 %.2f\n",
            th[["logite0_supp"]], a_fit))
```

The implied intercept is about half a logit below the printed one, and -8.83 -
a digit transposition of the printed -8.38 - falls inside that gap. The same
analysis run entirely on the paper's own numbers (next section, which never
touches this model's PK) reaches the same conclusion. The printed value is
nevertheless what the model carries: back-solving a parameter from a validation
target is tuning, and the paper prints -8.38. See the Errata.

**3. 516TT: off in both endpoints, and fully explained by the CL/F
disagreement the paper flags.** Table 2 gives CL/F for 516TT as 2.6 L/h in the
population PK arm and 7.2 L/h in the IVIVE arm - the one parameter the paper
says was *not* comparable ("The main PK variables ... were comparable (less
than twofold difference) except CL/F in those with the 516 TT genotype").
Rescaling the 516TT cohort exposures by 2.6 / 7.2 to put them on the IVIVE
clearance recovers Table 3.

```{r gate-tt-reconcile}
tt_fix <- cohort |>
  dplyr::filter(genotype == "516TT") |>
  dplyr::mutate(
    c816_ivive = c816 * 2.6 / 7.2,
    psupp_i = 1 / (1 + exp(-(th[["logite0_supp"]] + th[["e_c816_supp"]] * log10(c816_ivive)))),
    pcns_i  = 1 / (1 + exp(-(th[["logite0_cns"]]  + th[["e_c816_cns"]]  * log10(c816_ivive))))
  ) |>
  dplyr::group_by(dose_mg) |>
  dplyr::summarise(psupp_i = mean(psupp_i), pcns_i = mean(pcns_i), .groups = "drop") |>
  dplyr::left_join(published_t3 |> dplyr::filter(genotype == "516TT"), by = "dose_mg") |>
  dplyr::mutate(d_cns = pcns_i - pcns_pub, d_supp = psupp_i - psupp_pub)

tt_fix |>
  dplyr::select(dose_mg, pcns_i, pcns_pub, d_cns, psupp_i, psupp_pub, d_supp) |>
  dplyr::rename(
    "Dose (mg)"                     = dose_mg,
    "P(CNS) at IVIVE CL/F"          = pcns_i,
    "P(CNS), Table 3"               = pcns_pub,
    "Difference"                    = d_cns,
    "P(suppression) at IVIVE CL/F"  = psupp_i,
    "P(suppression), Table 3"       = psupp_pub,
    "Difference "                   = d_supp
  ) |>
  knitr::kable(digits = 3, caption =
    "516TT arm re-evaluated at the IVIVE clearance of 7.2 L/h instead of the population-PK 2.6 L/h.")

# The CNS endpoint returns to the same agreement as 516GG / 516GT, so nothing
# in the encoding of the 516TT arm is wrong -- the two published clearances
# simply disagree, as the paper states.
stopifnot(max(abs(tt_fix$d_cns)) < 0.030)
```

# Table 3 reproduced from the paper's own published exposures

The previous section compared two different models. This one removes the PK
layer entirely: it takes the IVIVE exposure summary the paper publishes in
Table 1 (mean and SD of steady-state AUC per arm), treats the average
concentration as lognormal, and integrates Equation 1 over it. Nothing here
depends on the packaged population PK model, so it is a clean gate on the two
logistic relationships as encoded.

```{r table-1-route}
published_t1 <- tibble::tribble(
  ~genotype, ~dose_mg, ~auc_mean, ~auc_sd,
  "516GG",   600,       59797,     41164,
  "516GG",   400,       39864,     27442,
  "516GG",   200,       20199,     14005,
  "516GT",   600,       82539,     57760,
  "516GT",   400,       55027,     38507,
  "516GT",   200,       28057,     19608,
  "516TT",   600,      114151,     68590,
  "516TT",   400,       76100,     45726,
  "516TT",   200,       42171,     23880
)

# E[ 1 / (1 + exp(-(A + B log10 C))) ] for log(C) ~ N(mu, sigma^2)
expected_p <- function(mu, sigma, a, b) {
  z <- seq(mu - 8 * sigma, mu + 8 * sigma, length.out = 4001)
  w <- dnorm(z, mu, sigma)
  sum(w / sum(w) / (1 + exp(-(a + b * z / log(10)))))
}

t1route <- published_t1 |>
  dplyr::mutate(
    cavg  = auc_mean / 24,
    cvar  = (auc_sd / 24) / cavg,
    s2    = log1p(cvar^2),
    mu    = log(cavg) - s2 / 2,
    sigma = sqrt(s2),
    psupp = mapply(expected_p, mu, sigma, th[["logite0_supp"]], th[["e_c816_supp"]]),
    pcns  = mapply(expected_p, mu, sigma, th[["logite0_cns"]],  th[["e_c816_cns"]])
  ) |>
  dplyr::left_join(published_t3, by = c("genotype", "dose_mg")) |>
  dplyr::mutate(d_supp = psupp - psupp_pub, d_cns = pcns - pcns_pub)

t1route |>
  dplyr::select(genotype, dose_mg, pcns, pcns_pub, d_cns, psupp, psupp_pub, d_supp) |>
  dplyr::rename(
    "Genotype"                = genotype,
    "Dose (mg)"               = dose_mg,
    "P(CNS) from Table 1"     = pcns,
    "P(CNS), Table 3"         = pcns_pub,
    "Difference"              = d_cns,
    "P(supp) from Table 1"    = psupp,
    "P(supp), Table 3"        = psupp_pub,
    "Difference "             = d_supp
  ) |>
  knitr::kable(digits = 3, caption =
    "Equation 1 integrated over the Table 1 IVIVE exposure distributions, against Table 3.")

cat(sprintf("CNS  (9 arms): mean |diff| %.4f, max |diff| %.4f\n",
            mean(abs(t1route$d_cns)),  max(abs(t1route$d_cns))))
cat(sprintf("Supp (9 arms): mean  diff %+.4f, range %+.4f to %+.4f\n",
            mean(t1route$d_supp), min(t1route$d_supp), max(t1route$d_supp)))

# The CNS logistic reproduces all nine published means; the suppression
# logistic is high in all nine by a similar amount.
stopifnot(
  mean(abs(t1route$d_cns)) < 0.015, max(abs(t1route$d_cns)) < 0.025,
  all(t1route$d_supp > 0.04), all(t1route$d_supp < 0.12)
)

a_fit_t1 <- optimize(
  function(a) sum((mapply(expected_p, t1route$mu, t1route$sigma, a,
                          th[["e_c816_supp"]]) - t1route$psupp_pub)^2),
  c(-14, -5)
)$minimum
cat(sprintf("suppression intercept implied by Tables 1 + 3: %.2f (printed %.2f)\n",
            a_fit_t1, th[["logite0_supp"]]))
```

All nine CNS means fall out of the paper's own numbers to within 0.02 absolute
using the printed `A = -6.65` and `B = 1.68` and the natural-logit reading of
Equation 1. All nine suppression means are high by 0.05-0.09 with the printed
`A = -8.38`, and this route implies an intercept near -8.86 while the cohort
route above implies -8.93. Two independent routes - one through the packaged
population PK model, one through the paper's published IVIVE exposures - agree
on both the direction and the rough size of the shift, which is why the Errata
records it as a probable typographical error in the printed intercept rather
than as a modelling choice.

# PKNCA validation

Steady-state non-compartmental analysis over the final dosing interval. The
time axis is shifted so the final dose sits at time 0; the time-zero record is
the steady-state pre-dose trough, which for a multiple-dose interval is a real
positive concentration and must not be overwritten with zero.

```{r pknca}
nca_prof <- lapply(seq_len(nrow(arms)), function(i) {
  a <- arms[i, ]
  rxode2::rxSetSeed(SEED)
  ev <- rxode2::et(amt = a$dose_mg, ii = 24, until = TSS) |>
    rxode2::et(TSS + seq(0, 24, by = 0.25)) |>
    rxode2::et(id = seq_len(NSUB)) |>
    as.data.frame() |>
    dplyr::mutate(
      SNP_CYP2B6_RS3745274_T_COUNT = a$tcount,
      treatment                    = a$treatment
    )
  rxode2::rxSolve(mod, ev, keep = "treatment", returnType = "data.frame") |>
    dplyr::filter(time >= TSS) |>
    dplyr::mutate(time = time - TSS)
}) |>
  dplyr::bind_rows()

sim_nca <- nca_prof |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment) |>
  dplyr::mutate(id = paste(treatment, id))

stopifnot(all(table(sim_nca$treatment) > 0),
          nrow(dplyr::filter(sim_nca, time == 0)) == NSUB * nrow(arms))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- sim_nca |>
  dplyr::distinct(id, treatment) |>
  dplyr::left_join(arms |> dplyr::select(treatment, dose_mg), by = "treatment") |>
  dplyr::mutate(time = 0, amt = dose_mg) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

Table 1 of the paper reports **arithmetic mean and SD** across its 500 virtual
subjects, whereas `nlmixr2lib::ncaComparisonTable()` aggregates a
`PKNCAresults` object across subjects by the **median**. For a right-skewed
exposure distribution those differ by roughly 10-15%, which is a large fraction
of the 20% flagging tolerance, so the per-subject results are pre-aggregated to
arithmetic means here and that summary is passed in as the simulated side.

```{r nca-mean-summary}
sim_mean <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop")

# One row per arm per parameter -- nothing left for the median aggregation
# inside ncaComparisonTable() to collapse.
stopifnot(
  nrow(sim_mean) == 4L * nrow(arms),
  !anyNA(sim_mean$PPORRES)
)
```

## Comparison against Table 1

Table 1 of the paper is the IVIVE arm's simulated steady-state summary (500
virtual subjects per arm), not an observed NCA dataset. It is the only tabulated
exposure summary the paper offers, so it is used as the reference here with that
caveat; the paper's own comparison criterion between its two arms was a
less-than-twofold difference.

```{r nca-compare}
published_nca <- published_t1 |>
  dplyr::left_join(
    tibble::tribble(
      ~genotype, ~dose_mg, ~cmax, ~cmin,
      "516GG",   600,      3022,  1873,
      "516GG",   400,      2014,  1100,
      "516GG",   200,      1017,   630,
      "516GT",   600,      3919,  2752,
      "516GT",   400,      2613,  1817,
      "516GT",   200,      1326,   923,
      "516TT",   600,      5160,  3917,
      "516TT",   400,      3440,  2612,
      "516TT",   200,      1870,  1495
    ),
    by = c("genotype", "dose_mg")
  ) |>
  dplyr::left_join(arms |> dplyr::select(genotype, dose_mg, treatment),
                   by = c("genotype", "dose_mg")) |>
  dplyr::transmute(treatment, cmax, cmin, auclast = auc_mean)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = sim_mean,
  reference     = published_nca,
  by            = "treatment",
  units         = c(cmax = "ng/mL", cmin = "ng/mL", tmax = "h",
                    auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(cmp, digits = 1, caption = paste(
  "Steady-state NCA of the population PK arm vs the Table 1 IVIVE summary.",
  "* differs from the reference by more than 20%."))
```

```{r nca-twofold}
# The paper's own acceptance criterion between its two arms. Applied to the
# exposure metrics, 516GG and 516GT must sit inside twofold; 516TT is the arm
# the paper reports as not comparable.
nca_wide <- sim_mean |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) |>
  dplyr::rename(sim = PPORRES) |>
  dplyr::left_join(
    published_nca |> tidyr::pivot_longer(c(cmax, cmin, auclast),
                                         names_to = "PPTESTCD", values_to = "ref"),
    by = c("treatment", "PPTESTCD")
  ) |>
  dplyr::mutate(fold = pmax(sim / ref, ref / sim),
                genotype = sub(" .*", "", treatment))

# A silently-failed join would leave `ref` NA and make max(fold) NA, so assert
# every arm x parameter actually matched before reading the folds.
stopifnot(nrow(nca_wide) == 3L * nrow(arms), !anyNA(nca_wide$fold))

nca_wide |>
  dplyr::group_by(genotype) |>
  dplyr::summarise(max_fold = max(fold), .groups = "drop") |>
  dplyr::rename("Genotype" = genotype, "Worst fold difference vs Table 1" = max_fold) |>
  knitr::kable(digits = 2, caption =
    "Fold difference between the population PK arm and the Table 1 IVIVE summary.")

stopifnot(
  max(nca_wide$fold[nca_wide$genotype != "516TT"]) < 2,
  max(nca_wide$fold[nca_wide$genotype == "516TT"]) > 2
)
```

The 516GG and 516GT arms sit inside the paper's twofold criterion for every
exposure metric; 516TT exceeds it, exactly as the paper reports. Within the
516GG and 516GT arms the population PK model runs 7-14% below the Table 1 IVIVE
summary on all three metrics.

Worth noting when reading those percentages: the paper's Table 1 and Table 2 are
not mutually consistent **for the IVIVE arm itself**, so the reference column
carries its own uncertainty.

```{r ivive-internal}
auc_ivive_gg600 <- 59797 / 1000            # ng*h/mL -> mg*h/L, Table 1
cl_from_auc     <- 600 / auc_ivive_gg600   # apparent CL/F implied by Table 1
f_needed        <- auc_ivive_gg600 * 15.32 / 600

cat(sprintf("Table 1 (516GG, 600 mg) AUC implies CL/F = %.2f L/h\n", cl_from_auc))
cat(sprintf("Table 2 IVIVE column reports CL/F = %.2f L/h (ratio %.2f)\n",
            15.32, 15.32 / cl_from_auc))
cat(sprintf("bioavailability required to reconcile them: %.2f\n", f_needed))

# Table 1 is internally coherent -- its AUC/24 sits between its own Ctrough
# and Cmax -- so the AUC is a genuine 0-24 h steady-state exposure, and no
# bioavailability can reconcile it with 15.32 L/h because F would exceed 1.
stopifnot(auc_ivive_gg600 * 1000 / 24 > 1873,
          auc_ivive_gg600 * 1000 / 24 < 3022,
          f_needed > 1)
```

At steady state `AUC over tau = Dose * F / CL`, so Table 1's 516GG 600 mg AUC
pins the IVIVE arm's apparent clearance at 10.0 L/h, against the 15.32 L/h
printed in Table 2 for the same arm - a 1.53-fold gap that no bioavailability
can close, since F would have to be 1.53. Table 1 is internally coherent (its
AUC/24 of 2,492 ng/mL lies between its own Ctrough of 1,873 and Cmax of 3,022),
so the discrepancy is between the two tables and not inside either. The
population PK arm's 13.3 L/h sits between the two IVIVE figures. This is a
defect in the arm this model does not reproduce, and it is recorded here only so
that the percentage differences above are not over-interpreted.

# Assumptions and deviations

**Scope.** Only the NONMEM population PK model and the two logistic PK/PD
relationships are packaged. The Simcyp 10.1 whole-body IVIVE/PBPK arm is not
reproduced: its system parameters come from the Simcyp North European Caucasian
population library and its ODE system is never written out, so it cannot be
reconstructed from any on-disk source. Table 4 of the paper (efavirenz
physicochemical and in-vitro metabolic characteristics) is therefore not encoded
either, since those inputs are only meaningful inside that platform model.

**`Vd` units in Table 2 (unit-header typo).** Table 2 heads the volume row
"Vd (l)" and gives 4.3 for the population PK arm and 8.1 for the IVIVE arm. The
values are per kilogram, and the model uses 4.3 L/kg x 70.5 kg = 303 L. Three
independent facts force this reading: (a) 4.3 L absolute with CL/F 13.3 L/h
gives a 0.22 h half-life, which cannot produce the reported once-daily troughs
of 1,873 ng/mL; (b) 303 L falls inside the paper's own bootstrap 90% confidence
interval for the volume, 260.5-339.5 L, and beside the base-model 317 L;
(c) the IVIVE entry is a Simcyp Vss, which that platform reports natively in
L/kg, and the paper's "less than twofold difference" claim (8.1 / 4.3 = 1.9)
only holds if both columns share one unit.

**Logit link.** Equation 1 is a natural logit; the sentence after it states a
base-10 logit. Equation 1 is used, corroborated exactly by both published odds
ratios (see the gate above). The trailing sentence is treated as an erratum.

**Suppression intercept `A = -8.38` does not reproduce Table 3.** Two
independent routes - the packaged population PK cohort, and Equation 1
integrated over the paper's own Table 1 IVIVE exposures - each overshoot all
nine Table 3 suppression means by 0.05-0.09 absolute, and imply intercepts of
-8.93 and -8.86 respectively. Both bracket -8.83, a digit transposition of the
printed -8.38. The CNS endpoint, with the same machinery and its own printed
coefficients, reproduces all nine Table 3 means to within 0.02, so the
discrepancy is specific to the suppression intercept rather than to the
exposure metric or the link function. **The model nevertheless carries the
printed -8.38.** Back-solving a parameter from a validation target is tuning,
and the printed value is what the paper reports. A reader who wants to
reproduce Table 3 can override it:
`readModelDb("Siccardi_2012_efavirenz") |> rxode2::ini(logite0_supp = -8.83)`.

**`ka`.** Table 2 gives the final-model `ka` as 0.36 1/h with 76% RSE, while
the base model gave 0.54 1/h and the bootstrap 90% confidence interval is
0.4-1.1 1/h, which does not contain 0.36. The model uses the final-model value
from Table 2. Absorption is very poorly identified in this dataset (one random
sample per HIV-positive patient), and the choice within that range moves the
steady-state `C8-16h` by only a few percent because it does not change AUC over
the interval at all.

**Inter-individual variability.** The paper reports 57.4% CV on CL/F for the
**base** model, and states only that adding the genotype covariate "explained
12.6% of the interindividual variability of CL/F" without saying whether that
share is of the variance or of the coefficient of variation. The printed
base-model value is carried forward unreduced, as
`omega^2 = log(0.574^2 + 1) = 0.28478` for the exponential IIV model the paper
specifies; the `ini()` entry carries that conversion as an expression rather
than a transcribed decimal. Applying the reduction under either convention
would give roughly 53% CV. No IIV is reported on V/F or `ka` and none is added.

**Residual error.** The paper states a proportional residual structure but never
reports its magnitude, so `propSd` is fixed at zero rather than invented. The
model therefore predicts without residual noise; add a value via `ini()` if
residual variability is needed.

**Genotype shifts.** The paper writes the covariate effect as
`CL = CL0 + theta1 * GT + theta2 * TT` but prints the three resulting typical
values (13.3, 9.7, 2.6 L/h) instead of `theta1` and `theta2`. The two `ini()`
shifts are the exact differences of those printed values, -3.6 and -10.7 L/h.

**`C8-16h` semantics.** The two accumulator states run from the start of the
simulation, so the model's `c816` output is the running mean over every elapsed
8-16 h post-dose window and converges to the steady-state value rather than
equalling it at any finite time. The steady-state value for one interval is
recovered either by restarting the accumulators at the top of that interval
(typical-value section) or by differencing them across it (cohort section);
both are demonstrated above and both agree with a direct grid mean.

**Cohort assumptions.** The 200 subjects per arm carry the model's IIV on CL/F
only; no demographic covariates enter the model, because the paper screened
weight, age, gender and ethnicity and retained none (they are recorded under
`covariatesDataExcluded` with no coefficients, since none are reported).
Genotype is assigned by arm rather than sampled at population frequency, matching
the paper's per-genotype simulation design. All nine arms share one seed so that
the dose-linearity gate is an exact per-subject identity.

**Reference-value provenance.** Table 1 and Table 3 of the paper are outputs of
the IVIVE arm, not observed data and not outputs of the population PK arm. Every
comparison against them above is a cross-arm comparison and is labelled as such.
No parameter was adjusted to improve any of them. Table 1 and Table 2 are also
mutually inconsistent for the IVIVE arm by a factor of 1.53 in apparent
clearance, which no bioavailability can reconcile (see the check in the PKNCA
section); that is a defect in the arm this model does not reproduce, but it caps
how tightly any comparison against Table 1 can be read.

**PD cohorts differ from the PK cohort.** The population PK model and the two
logistic models were fitted to different patients (157 vs 93 and 121), so the
coupled PK/PD model in this file is the paper's own composition of two
independently estimated layers, not a joint fit.
