---
title: "Levofloxacin and moxifloxacin in MDR-TB (Sidamo 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levofloxacin and moxifloxacin in MDR-TB (Sidamo 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Sidamo 2022 fitted two separate one-compartment population PK models, one for
levofloxacin and one for moxifloxacin, to steady-state plasma concentrations from
Ethiopian adults treated for multidrug- or rifampicin-resistant tuberculosis.
Both are packaged:

* `Sidamo_2022_levofloxacin`:
  `r rxode2::rxode(readModelDb("Sidamo_2022_levofloxacin"))$description`
* `Sidamo_2022_moxifloxacin`:
  `r rxode2::rxode(readModelDb("Sidamo_2022_moxifloxacin"))$description`
* Citation: `r rxode2::rxode(readModelDb("Sidamo_2022_levofloxacin"))$reference`
* Article (open access): <https://doi.org/10.2147/IDR.S389442>

# Population

The study (Methods, Results and Table 1) followed outpatients with MDR/RR-TB at
four hospitals in Southern Ethiopia. Patients were 18 years or older, with or
without HIV co-infection, and had received a levofloxacin-based regimen (750 or
1000 mg/day) or a moxifloxacin-based regimen (600 mg/day) for at least 8 days.
Severely ill and anaemic patients (haematocrit below 25%) were excluded. Plasma
was sampled pre-dose and at 2, 4, 6, 9, 12 and 24 h after a directly observed
dose taken after breakfast. Of 55 sampled patients, 39 had acceptable profiles:
**21 on levofloxacin and 18 on moxifloxacin**.

| Characteristic (Table 1) | Levofloxacin (n = 21) | Moxifloxacin (n = 18) |
|---|---|---|
| Age, years | 26 (20-30) | 26 (20-37) |
| Male, n (%) | 12 (57.1) | 8 (44.4) |
| Weight, kg | 45 (42.5-52.3) | 51 (46.6-65.0) |
| BMI, kg/m^2 | 16.7 (15.6-18) | 17.3 (16.2-19.2) |
| Malnourished (BMI < 16), n (%) | 9 (42.9) | 10 (55.6) |
| Serum creatinine, mg/dL | 1 (0.8-1.3) | 1 (0.8-1.4) |

Values are median (IQR) unless stated. The same information is available
programmatically:

```{r population}
str(readModelDb("Sidamo_2022_levofloxacin")()$population, give.attr = FALSE)
```

# Source trace

Both models are one-compartment with first-order absorption and linear
elimination, fitted in Phoenix NLME with first-order conditional estimation. The
levofloxacin model has an absorption lag time and a combined additive and
proportional residual error. The moxifloxacin model has no lag time and an
additive residual error (Results, 'The Population PK Model'). All parameters are
apparent oral values.

| Equation / parameter | Levofloxacin | Moxifloxacin | Source location |
|----------------------|--------------|--------------|-----------------|
| `lka` (ka, 1/h) | log(0.2) | log(0.5) | Table 2, 'GM (%RSE)' |
| `lvc` (V/F, L) | log(122) | log(102.2) | Table 2 |
| `lcl` (CL/F, L/h) | log(10.4) | log(20) | Table 2 |
| `ltlag` (Tlag, h) | log(0.2) | -- | Table 2 |
| Covariate centring value | CREAT 1.1 mg/dL on CL/F | BMI 17.8 kg/m^2 on V/F | Table 2, 'Cr (mg/dL)' and 'BMI (Kg.m2)' rows; see [Reading Table 2](#reading-table-2) |
| `e_creat_cl`, `e_bmi_vc` (power exponents) | fixed(0) | fixed(0) | Not reported; Methods Equation 1 gives the power form |
| `etalka` (variance) | 0.01941 (14 %CV) | 0.2987 (59 %CV) | Table 2, '%CV (Shrinkage, %)', as log(1 + CV^2) |
| `etalvc` (variance) | 0.065413 (26 %CV) | 0.0099503 (10 %CV) | Table 2 |
| `etalcl` (variance) | 0.36158 (66 %CV) | 0.00072873 (2.7 %CV) | Table 2 |
| `etaltlag` (variance) | 0.0011553 (3.4 %CV) | -- | Table 2 |
| `propSd`, `addSd` | fixed(0), fixed(0) | --, fixed(0) | Error models named in Results; magnitudes not reported |
| `(COV / COVmean)^theta2` | CL/F | V/F | Methods Equation 1 |
| `exp(theta + eta)` | -- | -- | Methods Equation 3, exponential IIV |
| `d/dt(depot)`, `d/dt(central)`, `alag(depot)` | -- | -- | Results: one-compartment, first-order absorption, lag for levofloxacin only |

# Reading Table 2

Results state that serum creatinine was retained on levofloxacin clearance and
BMI on moxifloxacin volume. Methods Equation 1 gives the form,
`TVP = theta1 * (COV / COVmean)^theta2`. Table 2 prints a single number for each
covariate in the 'GM (%RSE)' column: 'Cr (mg/dL) 1.1 (10)' and 'BMI (Kg.m2) 17.8
(1.1)'. It does not say whether that number is the exponent theta2 or the
centring value COVmean.

The packaged models read it as **COVmean**, with the exponent unreported:

* Both rows carry the covariate's units, which a dimensionless exponent would not.
* Both values match the cohort's mean covariate. Levofloxacin creatinine has a
  median of 1 mg/dL and a right-skewed IQR of 0.8-1.3, so its mean is near 1.1.
  Moxifloxacin BMI has a median of 17.3 kg/m^2 and an IQR of 16.2-19.2, so its
  mean is near 17.8.
* An exponent of 17.8 on BMI is not physically meaningful. It would cut V/F by
  60% between the median BMI and the centring BMI, a difference of only 0.5
  kg/m^2. A single reading must hold for both rows, and only the COVmean reading
  does.
* The same group's earlier Phoenix analysis (Rao 2021, packaged as
  `Rao_2021_pyrazinamide` and `Rao_2021_ethambutol`) also left its covariate
  coefficients unprinted.

The exponents are therefore `fixed(0)` placeholders. The covariate columns stay
in the model, so the structure is the published one, but `CREAT` and `BMI` have
no effect until the coefficients are known. The '%CV' column also prints values
for both covariate rows, 14.7 and 4.5, each with shrinkage 'NA'. As exponents,
14.7 on creatinine would be as implausible as 17.8 on BMI. A residual-error
reading does not fit either. Neither value is used.

# Virtual cohort and simulation

The paper simulated each patient's own profile at 750, 1000 and 1500 mg of
levofloxacin and 600, 800 and 1000 mg of moxifloxacin (Methods, 'Drug Exposure').
The virtual cohort here is 100 subjects per dose. Each subject gets 8 once-daily
doses into `depot`, and `central` is observed every 0.5 h over the last dosing
interval (168-192 h), which is at steady state for both drugs. Creatinine and BMI
are drawn log-normally to match the Table 1 median and IQR of each drug's cohort.
With the exponents at zero they do not change the simulation, but they keep the
covariate path exercised.

```{r cohort}
doses <- list(
  levofloxacin = c(750, 1000, 1500),
  moxifloxacin = c(600, 800, 1000)
)
n_per_arm <- 100L
obs_times <- seq(168, 192, by = 0.5)

# Log-normal draw matching a published median and IQR.
draw_lognormal <- function(n, median, q1, q3) {
  exp(rnorm(n, log(median), log(q3 / q1) / (2 * qnorm(0.75))))
}

make_arm <- function(drug, dose, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  covs <- tibble(
    id = ids,
    CREAT = draw_lognormal(n_per_arm, 1, 0.8, if (drug == "levofloxacin") 1.3 else 1.4),
    BMI = if (drug == "levofloxacin") {
      draw_lognormal(n_per_arm, 16.7, 15.6, 18)
    } else {
      draw_lognormal(n_per_arm, 17.3, 16.2, 19.2)
    }
  )
  dose_rows <- tidyr::expand_grid(id = ids, time = seq(0, 168, by = 24)) |>
    mutate(amt = dose, evid = 1L, cmt = "depot")
  obs_rows <- tidyr::expand_grid(id = ids, time = obs_times) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
  bind_rows(dose_rows, obs_rows) |>
    left_join(covs, by = "id") |>
    mutate(drug = drug, treatment = paste(dose, "mg")) |>
    arrange(id, time, desc(evid))
}

rxode2::rxSetSeed(20221201)
events <- list()
for (drug in names(doses)) {
  arms <- lapply(seq_along(doses[[drug]]), function(i) {
    make_arm(drug, doses[[drug]][i], id_offset = (i - 1L) * n_per_arm)
  })
  events[[drug]] <- bind_rows(arms)
  stopifnot(!anyDuplicated(events[[drug]][, c("id", "time", "evid")]))
}
```

```{r simulate}
mods <- list(
  levofloxacin = readModelDb("Sidamo_2022_levofloxacin"),
  moxifloxacin = readModelDb("Sidamo_2022_moxifloxacin")
)
sim <- bind_rows(lapply(names(mods), function(drug) {
  as.data.frame(rxode2::rxSolve(
    mods[[drug]],
    events = events[[drug]],
    keep = c("drug", "treatment")
  ))
}))
stopifnot(!anyNA(sim$Cc))
```

```{r profiles, fig.width = 7, fig.height = 4}
sim |>
  mutate(tad = time - 168) |>
  group_by(drug, treatment, tad) |>
  summarise(
    q05 = quantile(Cc, 0.05), q50 = median(Cc), q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, colour = NA) +
  geom_line() +
  facet_wrap(~drug, scales = "free_y") +
  labs(
    x = "Time after dose at steady state (h)", y = "Plasma concentration (mg/L)",
    colour = "Dose", fill = "Dose",
    caption = "Median and 90% interval of 100 simulated subjects per dose (between-subject variability only)."
  ) +
  theme_bw()
```

# Typical-value comparison with Table 3

Table 3 gives the median Cmax and AUC0-24 at each dose, simulated from the
patients' individual estimates, alongside the observed non-compartmental values
at the doses actually given. The paper does not say whether its simulations are
single-dose or steady-state, so the typical-value models are solved both ways.

```{r typical}
published_t3 <- tibble::tribble(
  ~drug, ~dose, ~source, ~cmax, ~auc24,
  "levofloxacin", 750, "simulated", 7.5, 60.9,
  "levofloxacin", 1000, "simulated", 10.0, 81.5,
  "levofloxacin", 1500, "simulated", 15.1, 122.7,
  "levofloxacin", 750, "observed", 5.3, 58.1,
  "levofloxacin", 1000, "observed", 7.4, 103.4,
  "moxifloxacin", 600, "simulated", 2.6, 25.1,
  "moxifloxacin", 800, "simulated", 3.5, 33.4,
  "moxifloxacin", 1000, "simulated", 4.4, 41.8,
  "moxifloxacin", 600, "observed", 3.2, 28.2
)

trap_auc <- function(time, conc) {
  sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2)
}

typical_profile <- function(drug, dose, steady_state = FALSE) {
  ev <- if (steady_state) {
    rxode2::et(amt = dose, cmt = "depot", ii = 24, addl = 29) |>
      rxode2::et(seq(696, 720, by = 0.05), cmt = "central")
  } else {
    rxode2::et(amt = dose, cmt = "depot") |>
      rxode2::et(seq(0, 24, by = 0.05), cmt = "central")
  }
  ev <- as.data.frame(ev)
  ev$CREAT <- 1.1
  ev$BMI <- 17.8
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mods[[drug]]), events = ev))
  tibble(cmax = max(s$Cc), auc24 = trap_auc(s$time, s$Cc))
}

typical <- published_t3 |>
  rowwise() |>
  mutate(
    sd = list(typical_profile(drug, dose)),
    ss = list(typical_profile(drug, dose, steady_state = TRUE))
  ) |>
  ungroup() |>
  mutate(
    cmax_sd = sapply(sd, `[[`, "cmax"),
    auc_sd = sapply(sd, `[[`, "auc24"),
    cmax_ss = sapply(ss, `[[`, "cmax"),
    auc_ss = sapply(ss, `[[`, "auc24"),
    pct_cmax_ss = 100 * (cmax_ss - cmax) / cmax,
    pct_auc_sd = 100 * (auc_sd - auc24) / auc24,
    pct_auc_ss = 100 * (auc_ss - auc24) / auc24
  )

typical |>
  select(drug, dose, source, cmax, cmax_sd, cmax_ss, auc24, auc_sd, auc_ss, pct_cmax_ss, pct_auc_ss) |>
  dplyr::rename(
    "Drug" = drug,
    "Dose (mg)" = dose,
    "Table 3 row" = source,
    "Table 3 Cmax" = cmax,
    "Typical single-dose Cmax" = cmax_sd,
    "Typical steady-state Cmax" = cmax_ss,
    "Table 3 AUC0-24" = auc24,
    "Typical single-dose AUC0-24" = auc_sd,
    "Typical steady-state AUC0-24" = auc_ss,
    "Cmax % diff (steady state)" = pct_cmax_ss,
    "AUC % diff (steady state)" = pct_auc_ss
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Typical-value Cmax (mg/L) and AUC0-24 (mg*h/L) at the covariate centring values, against the Table 3 medians of Sidamo 2022."
  )
```

```{r typical-gate}
lfx <- typical[typical$drug == "levofloxacin", ]
mxf <- typical[typical$drug == "moxifloxacin", ]
mxf_obs <- mxf[mxf$source == "observed", ]
lfx_sim <- lfx[lfx$source == "simulated", ]
stopifnot(
  # Moxifloxacin: the steady-state typical value reproduces the observed
  # median Cmax and AUC0-24 at 600 mg (realised +2% and +6%). A
  # mis-transcribed CL or V moves these by tens of percent.
  abs(mxf_obs$pct_cmax_ss) < 15,
  abs(mxf_obs$pct_auc_ss) < 15,
  # Levofloxacin: the Table 3 simulated AUC0-24 medians lie between the
  # single-dose and steady-state typical values at every dose, and stay
  # dose-proportional, as the linear model requires.
  all(lfx_sim$auc_sd < lfx_sim$auc24 & lfx_sim$auc24 < lfx_sim$auc_ss),
  abs(lfx_sim$auc24[3] / lfx_sim$auc24[1] - 2) < 0.05
)
```

These are deterministic solves with no random draw, so the bounds above are tight
on purpose.

The moxifloxacin model reproduces the observed steady-state exposure at 600 mg.
Its simulated rows in Table 3 run about 15-20% below the typical value, and the
Table 3 means (31.5, 40.6 and 50.8 mg*h/L) are within 5% of it.

For levofloxacin, the AUC0-24 medians lie between the single-dose and
steady-state solves. The Table 3 Cmax values do not. The simulated medians (7.5,
10.0 and 15.1 mg/L) are about 1.8 times the steady-state typical value, and even
the observed medians (5.3 and 7.4 mg/L) are about 30% above it. With ka =
0.2 1/h and V/F = 122 L, the model's ratio of Cmax to AUC0-24 at steady state is
about 0.057 1/h. The simulated rows of Table 3 imply 0.12 1/h and the observed
rows 0.07-0.09 1/h. Rounding ka to one decimal place cannot account for this,
because even instant absorption limits the ratio to about 0.098 1/h at
V/F = 122 L. The model is kept as published. The [Assumptions](#assumptions-and-deviations)
section records the gap.

# PKNCA comparison against Table 3

```{r nca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(group = paste(drug, treatment)) |>
  dplyr::select(id, time, Cc, drug, treatment, group)

dose_df <- bind_rows(events) |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(group = paste(drug, treatment)) |>
  dplyr::select(id, time, amt, drug, treatment, group)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | group + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | group + id,
  doseu = "mg"
)
intervals <- data.frame(start = 168, end = 192, cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

per_subject <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::select(group, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(
    dplyr::distinct(dose_df, group, drug, treatment, amt),
    by = "group"
  )
```

```{r nca-compare}
reference <- published_t3 |>
  dplyr::filter(source == "simulated") |>
  dplyr::mutate(group = paste(drug, paste(dose, "mg"))) |>
  dplyr::select(group, cmax, auclast = auc24)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by = "group",
  params = c("cmax", "auclast"),
  units = c(cmax = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Simulated steady state (100 subjects per dose) vs. the Sidamo 2022 Table 3 simulated medians. * differs from the reference by >20%."
)
```

```{r nca-gate}
chk <- per_subject |>
  dplyr::group_by(drug, amt) |>
  dplyr::summarise(auclast = median(auclast), .groups = "drop") |>
  dplyr::inner_join(
    dplyr::filter(published_t3, source == "simulated"),
    by = c("drug", "amt" = "dose")
  ) |>
  dplyr::mutate(pct_auc = 100 * (auclast - auc24) / auc24)
stopifnot(
  # Centre, not extremes: the median over the three doses of each drug. The
  # expected offsets are about +18% (levofloxacin) and +20% (moxifloxacin),
  # where the Table 3 medians summarise shrunken individual estimates. A wrong
  # dose, unit or clearance moves every dose together, by far more.
  abs(median(chk$pct_auc[chk$drug == "levofloxacin"])) < 35,
  abs(median(chk$pct_auc[chk$drug == "moxifloxacin"])) < 35
)
```

The starred Cmax rows for levofloxacin repeat the typical-value gap above. The
AUC0-24 medians are about 15-20% above Table 3 for both drugs. This is the size of
offset expected when one median comes from a full-variability simulation and the
other from 18-21 shrunken individual estimates.

## Spread of AUC0-24

AUC0-24 at steady state is dose / CL, whatever the volume and absorption, so its
spread across subjects reflects between-subject variability in CL alone. The
Table 3 values come from the patients' own estimates, which shrinkage pulls
toward the typical value. Their spread should therefore be *no wider* than that
of a simulation with the full between-subject variances.

```{r spread}
iqr_ratio <- function(x) unname(quantile(x, 0.75) / quantile(x, 0.25))
spread <- per_subject |>
  dplyr::filter(amt %in% c(750, 600)) |>
  dplyr::group_by(drug, amt) |>
  dplyr::summarise(sim_iqr_ratio = iqr_ratio(auclast), .groups = "drop") |>
  dplyr::mutate(
    table3_iqr_ratio = ifelse(drug == "levofloxacin", 81.8 / 40.0, 47.0 / 19.0)
  )
spread |>
  dplyr::rename(
    "Drug" = drug,
    "Dose (mg)" = amt,
    "Simulated AUC0-24 Q3/Q1" = sim_iqr_ratio,
    "Table 3 simulated AUC0-24 Q3/Q1" = table3_iqr_ratio
  ) |>
  knitr::kable(digits = 2)
```

For levofloxacin the full-variability spread (66% CV on CL) is at least as wide
as Table 3's, as it should be. For moxifloxacin it is not. The printed 2.7% CV on
CL gives an AUC0-24 interquartile ratio near 1.0, while Table 3 shows about 2.5
for the simulated rows and 2.6 (19.3-50.8 mg*h/L) for the observed NCA at 600 mg.
The Discussion also reports individual 600 mg AUC0-24 values from 3.5 to
72.2 mg*h/L. None of the other variability entries in the moxifloxacin rows of
Table 2 would be wide enough to explain this if moved to CL. The moxifloxacin
model therefore reproduces the typical exposure but understates its
between-subject spread; see [Assumptions](#assumptions-and-deviations).

# Probability of target attainment

Figures 4 and 5 report the percentage of patients whose AUC0-24/MIC reaches the
kill target (146 for levofloxacin, 53 for moxifloxacin) or the
resistance-suppression target (360 and 100). The values below are those quoted in
the Results text. A dash means the text gives no number.

```{r pta}
pta_pub <- tibble::tribble(
  ~drug, ~amt, ~mic, ~target, ~published,
  "levofloxacin", 750, 0.5, 146, 29,
  "levofloxacin", 1000, 0.5, 146, 62,
  "levofloxacin", 1500, 0.5, 146, 95,
  "levofloxacin", 750, 0.5, 360, 0,
  "levofloxacin", 1000, 0.5, 360, 0,
  "levofloxacin", 1500, 0.5, 360, 4.8,
  "levofloxacin", 1500, 1, 146, 24,
  "moxifloxacin", 600, 0.25, 53, 94.4,
  "moxifloxacin", 800, 0.25, 53, 94.4,
  "moxifloxacin", 1000, 0.25, 53, 94.4,
  "moxifloxacin", 600, 0.25, 100, 50,
  "moxifloxacin", 800, 0.25, 100, 77.8,
  "moxifloxacin", 1000, 0.25, 100, 94.4,
  "moxifloxacin", 600, 1, 53, 11.1,
  "moxifloxacin", 800, 1, 53, 27.8,
  "moxifloxacin", 1000, 1, 53, 38.9
)

pta <- pta_pub |>
  dplyr::left_join(per_subject, by = c("drug", "amt"), relationship = "many-to-many") |>
  dplyr::group_by(drug, amt, mic, target, published) |>
  dplyr::summarise(simulated = 100 * mean(auclast / mic >= target), .groups = "drop")

pta |>
  dplyr::rename(
    "Drug" = drug,
    "Dose (mg)" = amt,
    "MIC (mg/L)" = mic,
    "AUC0-24/MIC target" = target,
    "Published PTA (%)" = published,
    "Simulated PTA (%)" = simulated
  ) |>
  knitr::kable(digits = c(0, 0, 2, 0, 1, 1))
```

This table is descriptive, and the two columns are not expected to agree closely.
The published values count 18-21 patients from their own estimates. For
levofloxacin, the simulation's AUC0-24 runs about 20% higher at steady state.
Its 66% CV on CL also spreads the distribution more widely, which moves
attainment in both directions around the threshold. For moxifloxacin, the
near-zero CL variability makes simulated attainment almost all-or-nothing: AUC0-24
is close to 30, 40 and 50 mg*h/L for every subject. Those values sit above
13.25 mg*h/L (target 53 at MIC 0.25 mg/L) and above 25 mg*h/L (target 100 at
MIC 0.25 mg/L). They sit below 53 mg*h/L (target 53 at MIC 1 mg/L). The
published values are graded instead, which again reflects the wider individual
spread discussed above.

# Assumptions and deviations

* **Covariate exponents.** Serum creatinine on levofloxacin CL/F and BMI on
  moxifloxacin V/F are retained in the published models. Their exponents are
  not reported and are `fixed(0)`. The single value in each covariate row of
  Table 2 (1.1 mg/dL and 17.8 kg/m^2) is read as the centring value of Equation 1.
  See [Reading Table 2](#reading-table-2). If the exponents become available,
  set `e_creat_cl` and `e_bmi_vc` through `ini()`.
* **Unused '%CV' entries.** The covariate rows of Table 2 also carry '%CV' values
  of 14.7 (creatinine) and 4.5 (BMI), with shrinkage 'NA'. The BIC rows carry
  '(9)' and '(3)'. Their meaning is not stated, and none of them is used.
* **Residual error magnitude.** Results name a combined additive and proportional
  error for levofloxacin and an additive error for moxifloxacin, but give no
  values, and the paper has no supplement. `propSd` and `addSd` are `fixed(0)`,
  so simulations carry between-subject variability only.
* **IIV scale.** Methods say IIV was 'reported in terms of percent coefficient of
  variability'. The '%CV' column is converted to log-normal variances with
  omega^2 = log(1 + CV^2). The Table 2 footnote defines %CV as
  '100 x IIV/parameter estimate'. Read literally, that would give a levofloxacin
  V/F variance of 31.7, so it is not used.
* **Moxifloxacin CL variability.** The printed 2.7% CV on CL is kept. It cannot
  reproduce the paper's own spread of AUC0-24 (see
  [Spread of AUC0-24](#spread-of-auc0-24)), so stochastic simulations from this
  model understate between-subject variability in moxifloxacin exposure. The
  value was not adjusted.
* **Levofloxacin Cmax.** The model's steady-state Cmax is about 45% below the
  Table 3 simulated medians and about 25% below the observed medians, while
  AUC0-24 agrees. The parameters were not adjusted. See
  [Typical-value comparison with Table 3](#typical-value-comparison-with-table-3).
* **Dosing.** The paper's sampling was at steady state, after at least 8 days of
  treatment, and the simulations here are steady-state. The Table 3 simulated
  levofloxacin AUC0-24 medians lie between the single-dose and steady-state
  typical values, so neither reading is excluded.
* **Covariates screened but not retained.** Nutritional status, sex, adverse drug
  effects, comorbidities, ALT, AST, bilirubin, and the other drug's covariate
  (Methods 'Covariates') are recorded in `covariatesDataExcluded` where a
  canonical column exists.
* No erratum or correction notice for this article was found on Europe PMC as
  of 2026-10-10.
