---
title: "PEGylated asparaginase and anti-PEG IgM (Siebel 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PEGylated asparaginase and anti-PEG IgM (Siebel 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20220301)
```

# Model and source

Siebel 2022 asks whether antibodies against polyethylene glycol (PEG) change
the pharmacokinetics of PEGylated asparaginase (PEG-ASNase) in children with
acute lymphoblastic leukaemia treated in the AIEOP-BFM ALL 2009 trial. The
authors took their group's previously published population PK model (the
14-compartment de-PEGylation transit chain of Wurthwein 2021) as a *reference
model*, refitted it to the antibody data set, and screened anti-PEG IgG and IgM
measured before and after dosing as covariates on the initial clearance
`CLinitial`. The final model adds a single effect: a **hockey stick on the
natural-log pre-existing anti-PEG IgM level**, active only above an estimated
cut point and only at the first induction dose.

* Citation: `r rxode2::rxode(readModelDb("Siebel_2022_pegasparaginase"))$reference`
* Article: <https://doi.org/10.1007/s13318-021-00741-w>
* Supplement (NONMEM control stream of the final model in Section 3):
  Electronic Supplementary Material 1 at the same DOI.

The package also ships the later models of the same group,
`Wurthwein_2025_pegasparaginase_*`, fitted to a larger AIEOP-BFM ALL 2009 data
set. Two of them carry the same IgM hockey stick with the cut point estimated
here (1.30 on the log scale) held fixed.

```{r load-model}
mod <- rxode2::rxode(readModelDb("Siebel_2022_pegasparaginase"))
mod0 <- rxode2::zeroRe(mod)
```

# Population

1444 children with acute lymphoblastic leukaemia from the German and Czech part
of AIEOP-BFM ALL 2009 (EudraCT 2007-004270-43), who received 3403 PEG-ASNase
administrations: two induction doses on protocol IA days 12 and 26 (all
patients) and one re-induction dose on protocol II day 8 (non-high-risk
patients only), each 2500 U/m^2 as a 2-hour infusion with a 3750 U cap. Median
age 5.13 years (range 1.06-18.32), median body surface area (BSA) 0.78 m^2
(0.41-2.58), 842 male / 602 female (Table 1). 6261 activity samples were
analysed, drawn mostly 7 and 14 days after each dose (ESM Table S1), together
with 2082 pre-dose and 6412 post-dose anti-PEG antibody measurements (Table 2).
Median pre-existing anti-PEG IgM before induction was 1.37 mean fluorescence
intensity (MFI), range 0.19-18.1; 146 of the 1444 patients were above the
finally estimated cut point of 3.67 MFI.

Samples at or after a hypersensitivity reaction or silent inactivation were
excluded before fitting, so **the model describes standard elimination only**.

The same information is available programmatically via
`readModelDb("Siebel_2022_pegasparaginase")()$population`.

# Structural model

Asparaginase activity is carried by a chain of 14 serial compartments that share
one serum volume `V`. Drug steps from compartment *i* to *i+1* at `Qtr/V`
(standing in for progressive hydrolysis of the PEG moiety) and every compartment
is eliminated at `CLinitial/V`; the terminal compartment passes its `Qtr/V` out of
the system. Apparent clearance therefore rises within a dosing interval from
`CLinitial` towards `CLinitial + Qtr` as mass migrates down the chain. The assay
measures total catalytic activity, so the observation is
`Cc = (central + transit1 + ... + transit13) / V`. All of this is read directly
off the `$DES` and `$ERROR` blocks in ESM Section 3.

Covariates act on `V`, `CLinitial` and `Qtr` exactly as in the control stream:

* BSA enters **linearly, centred on 0.79 m^2** (`THETA(1) * (1 + SCV1 * (BSA - 0.79))`),
  with one slope for `V` and one shared by `CLinitial` and `Qtr`.
* Age above 8 years increases `CLinitial` linearly; females have 7% lower
  `CLinitial`.
* `V` and `CLinitial` change stepwise at the second induction dose and at
  re-induction, relative to the first induction dose.
* Anti-PEG IgM: `CLinitial` is multiplied by
  `1 + 0.414 * max(log(IgM) - 1.30, 0)` **at the first induction dose only**.

## The anti-PEG IgM hockey stick

Section 3.3 states the effect in one sentence: "each additional log unit of
anti-PEG IgMprior above the estimated cut point (i.e. an increase in anti-PEG
IgMprior from 1.30 to 2.30 on the log scale or, equivalently, from 3.67 to 9.97
on the linear scale) was related to an increase in CLinitial of 41.4%". The chunk
below reads the individual `cl` back out of solved model output across a range
of IgM levels, so it checks the encoding as it runs, not a re-derivation of it.

```{r igm-effect, fig.cap = "Initial clearance relative to a patient below the cut point, as a function of the pre-existing anti-PEG IgM level, at the first induction dose (solid) and at the second induction dose (dashed). Model-derived from Siebel 2022 Table 3 and ESM Section 3."}
igm_grid <- c(0.19, 0.5, 1, 1.37, 2, 3, 3.67, 5, 7, 9.97, 12, 15, 18.1)

cl_at <- function(igm, occ) {
  ev <- rxode2::et(amt = 1950, dur = 2 / 24, cmt = "central") |>
    rxode2::et(1, cmt = "central") |>
    as.data.frame() |>
    mutate(BSA = 0.78, AGE = 5.13, SEXF = 0, OCC = occ, ABPEG_IGM = igm)
  s <- rxode2::rxSolve(mod0, ev, returnType = "data.frame")
  s$cl[s$time == 1]
}

cl_ref <- cl_at(1, 1)
igm_tab <- tidyr::expand_grid(igm = igm_grid, occ = c(1, 2)) |>
  rowwise() |>
  mutate(cl_rel = cl_at(igm, occ) / (cl_ref * ifelse(occ == 2, 1 - 0.107, 1))) |>
  ungroup()

ggplot(igm_tab, aes(igm, cl_rel, linetype = factor(occ))) +
  geom_line() +
  geom_point(data = filter(igm_tab, occ == 1)) +
  geom_vline(xintercept = 3.67, linetype = "dotted") +
  scale_x_log10() +
  scale_linetype_manual(values = c("1" = "solid", "2" = "dashed"),
                        labels = c("1" = "1st induction dose", "2" = "2nd induction dose")) +
  labs(x = "Pre-existing anti-PEG IgM (MFI, log axis)",
       y = "CLinitial relative to below-cut-point patient", linetype = NULL) +
  theme_bw()
```

```{r igm-effect-gate}
f <- function(igm, occ = 1) igm_tab$cl_rel[igm_tab$igm == igm & igm_tab$occ == occ]
c(at_median_1.37 = f(1.37), at_cut_3.67 = f(3.67), one_log_above_9.97 = f(9.97),
  max_18.1 = f(18.1), second_dose_at_18.1 = f(18.1, 2))

stopifnot(
  # Below the cut point there is no effect. 3.67 is the paper's rounding of
  # exp(1.30) = 3.6693, so it sits 0.0002 log units above the cut point and
  # its factor is 1.00008; hence the bound.
  all(abs(igm_tab$cl_rel[igm_tab$igm < 3.67] - 1) < 1e-9),
  abs(f(3.67) - 1) < 5e-4,
  # 9.97 MFI is one natural-log unit above 3.67, so the factor is 1 + 0.414
  # exactly (Section 3.3, '41.4%'); the tolerance only absorbs the rounding of
  # 3.67 and 9.97 in the text.
  abs(f(9.97) - 1.414) < 0.002,
  # The effect is restricted to the first induction dose.
  all(abs(igm_tab$cl_rel[igm_tab$occ == 2] - 1) < 1e-6)
)
```

## Mass balance

A linear chain has a closed-form total AUC. With `ke = CLinitial/V`,
`kt = Qtr/V` and `ratio = kt/(ke + kt)`, a dose `D` gives
`AUC = D / (V * (ke + kt)) * sum(ratio^j, j = 0..13)`. Reproducing it from the solved
ODEs confirms that the chain is wired correctly and conserves mass.

```{r mass-balance}
ev_mb <- rxode2::et(amt = 1950, cmt = "central") |>
  rxode2::et(seq(0, 150, by = 0.01), cmt = "central") |>
  as.data.frame() |>
  mutate(BSA = 0.78, AGE = 5.13, SEXF = 0, OCC = 1, ABPEG_IGM = 9.97)
mb <- rxode2::rxSolve(mod0, ev_mb, returnType = "data.frame")

auc_num <- sum(diff(mb$time) * (head(mb$Cc, -1) + tail(mb$Cc, -1)) / 2)
V <- mb$vc[1]; ke <- mb$cl[1] / V; kt <- mb$q[1] / V; r <- kt / (ke + kt)
auc_cf <- 1950 / (V * (ke + kt)) * sum(r^(0:13))

c(numeric = auc_num, closed_form = auc_cf, pct_diff = 100 * (auc_num - auc_cf) / auc_cf)

# A solve against its own closed form is pure numerical error, so the bound is
# tight: no per-subject random mechanism enters either side.
stopifnot(abs(auc_num - auc_cf) / auc_cf < 0.005)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location.

| Item | Value(s) | Source location |
|---|---|---|
| Chain of 14 serial compartments; dose into compartment 1; shared `V` | topology | ESM Section 3 `$MODEL`, `$DES`; ESM Fig. S1 |
| Observation = total activity over all 14 states divided by `V` | structural | ESM Section 3 `$ERROR` (`TOT = A(1)+...+A(14); IPRED = TOT/V1`) |
| `V`, `CLinitial`, `Qtr` | 1.71 L, 0.128 L/day, 0.954 L/day (at BSA 1 m^2) | Table 3, final covariate model |
| BSA linear, centred on 0.79 m^2; slopes on `V` and on `CLinitial` + `Qtr` | 1.60, 1.49 | ESM Section 3 `$PK`; Table 3 and footnote (a) |
| Quoting convention: values reported for BSA = 1 m^2 | convention | Table 3, note under the table |
| Phase changes in `V` (2nd induction dose, re-induction) | -0.158, -0.288 | Table 3 and footnote (b); ESM Section 3 `V1KOV` |
| Phase changes in `CLinitial` (2nd induction dose, re-induction) | -0.107, -0.438 | Table 3 and footnote (b); ESM Section 3 `CLKOV` |
| Age hockey stick, break point fixed at 8 years | 0.011 per year | Table 3 footnote (c); ESM Section 3 (`THETA(13)` `0 FIX`) |
| Sex on `CLinitial`, males the reference (`SEX` 1/2) | -0.070 | Table 3 footnote (d); ESM Section 3 `CLSEX` |
| Anti-PEG IgM slope above the cut point, first induction dose only | 0.414 per log unit | Table 3 footnote (e); ESM Section 3 (`THETA(17)`, `THETA(18)` `0 FIX`) |
| Anti-PEG IgM cut point, natural-log scale | 1.30 (3.67 MFI) | Table 3 footnote (f); Section 3.2.1 |
| IIV on `CLinitial`; no IIV on `V` or `Qtr` | 25.4% | Table 3; ESM Section 3 `$OMEGA` (`0 FIX`) |
| IOV on `V` and `CLinitial`, one occasion per administration | 11.9%, 22.8% | Table 3; ESM Section 3 `$OMEGA BLOCK(1) SAME` |
| Combined proportional + additive residual error | 0.191, 5.36 U/L | Table 3; ESM Section 3 `$ERROR`, `$SIGMA 1 FIX` |
| Published simulated day-14 troughs used as gates below | 517, 339, 305, 194 U/L | Section 3.3; ESM Table S5; Fig. 4 |

# Reproducing the published simulations

Section 3.3, ESM Table S5 and Figure 4 report Monte Carlo simulations of
1000 patients per arm, with BSA and age fixed at their medians (0.78 m^2,
5.13 years) and sex at the most frequent value (male), for two induction doses
14 days apart. Two regimens (2500 U/m^2 over 2 h; 1500 U/m^2 over 1 h) are
crossed with two IgM levels: at the cut point (3.67 MFI, no effect) and one log
unit above it (9.97 MFI). The published summary is the median (IQR) trough
14 days after the first dose.

```{r published}
published <- tibble::tribble(
  ~regimen,          ~igm,  ~median, ~q25, ~q75, ~pct_below_100,
  "2500 U/m^2, 2 h", 3.67,  517,     391,  632,  0.5,
  "2500 U/m^2, 2 h", 9.97,  339,     222,  457,  4.9,
  "1500 U/m^2, 1 h", 3.67,  305,     241,  367,  1.7,
  "1500 U/m^2, 1 h", 9.97,  194,     135,  262,  14.9
) |>
  mutate(dose_m2 = ifelse(grepl("^2500", regimen), 2500, 1500),
         dur_h   = ifelse(grepl("2 h$", regimen), 2, 1))
```

## Typical-value troughs

```{r typical}
bsa_med <- 0.78
trough_typical <- function(dose_m2, dur_h, igm) {
  ev <- rxode2::et(amt = dose_m2 * bsa_med, dur = dur_h / 24, cmt = "central") |>
    rxode2::et(c(7, 14), cmt = "central") |>
    as.data.frame() |>
    mutate(BSA = bsa_med, AGE = 5.13, SEXF = 0, OCC = 1, ABPEG_IGM = igm)
  s <- rxode2::rxSolve(mod0, ev, returnType = "data.frame")
  s$Cc[s$time == 14]
}

typ <- published |>
  rowwise() |>
  mutate(typical = trough_typical(dose_m2, dur_h, igm)) |>
  ungroup() |>
  mutate(pct_diff = 100 * (typical - median) / median)

typ |>
  select(regimen, igm, median, typical, pct_diff) |>
  dplyr::rename(
    "Regimen" = regimen,
    "Anti-PEG IgM (MFI)" = igm,
    "Published median (U/L)" = median,
    "Typical-value solve (U/L)" = typical,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = c(0, 2, 0, 0, 1),
               caption = "Day-14 trough after the first induction dose. Published values from Section 3.3 and ESM Table S5.")
```

```{r typical-gate}
stopifnot(
  # A mis-transcribed V, CLinitial, Qtr, BSA slope or antibody term moves these
  # troughs by tens of percent. The remaining gap compares a typical-value solve
  # with a 1000-patient Monte Carlo median and is under 3%. The solve is
  # deterministic, so the bound does not depend on the random-number stream.
  abs(median(typ$pct_diff)) < 3,
  max(abs(typ$pct_diff)) < 5
)
```

The typical-value solve lands within 3% of every published median, which is
about the Monte Carlo noise of a 1000-patient median.

## Which reading of the BSA quoting convention?

Table 3 reports `V`, `CLinitial` and `Qtr` "for a child with BSA = 1 m^2 for
better comparison". Because BSA enters linearly and centred on 0.79 m^2, that
means the NONMEM `THETA` is the printed value divided by `1 + slope * 0.21`, and
that is how the model file is encoded. The predecessor paper (Wurthwein 2021)
describes the same convention differently -- values "converted from L/0.79m2 to
L/m2" -- which read literally gives `THETA = printed * 0.79`. The two readings
differ by 3-5% in the typical parameters. Siebel 2022's own simulations can
arbitrate, because they are typical-covariate simulations of exactly these
parameters. The chain has a Poisson closed form, so both readings can be
evaluated without re-encoding the model:

```{r bsa-reading}
trough_cf <- function(V, CL, Q, D, t = 14) {
  ke <- CL / V; kt <- Q / V
  D * exp(-(ke + kt) * t) * sum((kt * t)^(0:13) / factorial(0:13)) / V
}
readings <- list(
  "BSA = 1 m^2 (model file)" = c(V = 1.71 / (1 + 1.60 * 0.21),
                                 CL = 0.128 / (1 + 1.49 * 0.21),
                                 Q = 0.954 / (1 + 1.49 * 0.21)),
  "printed x 0.79"           = c(V = 1.71 * 0.79, CL = 0.128 * 0.79, Q = 0.954 * 0.79)
)
bsa_cmp <- lapply(names(readings), function(nm) {
  th <- readings[[nm]]
  V  <- th[["V"]]  * (1 + 1.60 * (bsa_med - 0.79))
  CL <- th[["CL"]] * (1 + 1.49 * (bsa_med - 0.79))
  Q  <- th[["Q"]]  * (1 + 1.49 * (bsa_med - 0.79))
  published |>
    rowwise() |>
    mutate(reading = nm,
           closed_form = trough_cf(V, CL * ifelse(igm > 3.67, 1 + 0.414 * (log(igm) - 1.30), 1),
                                   Q, dose_m2 * bsa_med)) |>
    ungroup()
}) |>
  bind_rows() |>
  mutate(pct_diff = 100 * (closed_form - median) / median)

bsa_cmp |>
  group_by(reading) |>
  summarise(mean_abs_pct_diff = mean(abs(pct_diff)), .groups = "drop") |>
  dplyr::rename("Reading" = reading, "Mean |difference| vs published (%)" = mean_abs_pct_diff) |>
  knitr::kable(digits = 1, caption = "Typical-value closed-form trough (bolus) under each reading of the quoting convention, against the four published medians.")
```

```{r bsa-reading-gate}
err <- bsa_cmp |> group_by(reading) |> summarise(e = mean(abs(pct_diff)))
# The model file's reading must also agree with its own closed form, which
# confirms that the solve and the closed form describe the same parameters.
stopifnot(
  err$e[err$reading == "BSA = 1 m^2 (model file)"] < err$e[err$reading == "printed x 0.79"],
  max(abs(bsa_cmp$closed_form[bsa_cmp$reading == "BSA = 1 m^2 (model file)"] /
            typ$typical - 1)) < 0.02
)
```

The BSA = 1 m^2 reading is closer to the paper's own numbers, which supports
the encoding used here. The evidence is moderate rather than decisive: the
published values are Monte Carlo medians of 1000 patients, whose own sampling
noise is around 1-2%.

## Stochastic replication of Figure 4

200 virtual patients per arm (four arms), with IIV and IOV, two induction doses
on days 0 and 14 and observation to day 60. `OCC` switches from 1 to 2 at the
second dose; it is a time-varying covariate, so it is attached to a
materialised data frame.

```{r fig4-sim}
n_per_arm <- 200
grid <- sort(unique(c(seq(0, 60, by = 0.25), 13.999)))
arms <- published |> mutate(arm = row_number())

ev_fig4 <- lapply(seq_len(nrow(arms)), function(a) {
  ar <- arms[a, ]
  base <- rxode2::et(amt = ar$dose_m2 * bsa_med, time = c(0, 14),
                     dur = ar$dur_h / 24, cmt = "central") |>
    rxode2::et(grid, cmt = "central") |>
    as.data.frame()
  lapply(seq_len(n_per_arm), function(i) {
    b <- base
    b$id <- (a - 1) * n_per_arm + i
    b
  }) |>
    bind_rows() |>
    mutate(arm = a, regimen = ar$regimen, igm = ar$igm)
}) |>
  bind_rows() |>
  mutate(BSA = bsa_med, AGE = 5.13, SEXF = 0, ABPEG_IGM = igm,
         OCC = ifelse(time < 14, 1, 2))

fig4 <- rxode2::rxSolve(mod, ev_fig4, returnType = "data.frame",
                        keep = c("arm", "regimen", "igm"))
stopifnot(dplyr::n_distinct(fig4$id) == 4 * n_per_arm, !anyNA(fig4$Cc))
```

```{r fig4-plot, fig.cap = "Simulated PEG-ASNase activity in induction: median (line) and 95% prediction interval (band), 200 patients per arm. Replicates Figure 4 of Siebel 2022 (a: 2500 U/m^2 over 2 h; b: 1500 U/m^2 over 1 h). Dotted line: 100 U/L."}
fig4 |>
  filter(time != 13.999) |>
  group_by(regimen, igm, time) |>
  summarise(lo = quantile(Cc, 0.025), md = median(Cc), hi = quantile(Cc, 0.975),
            .groups = "drop") |>
  mutate(`Anti-PEG IgM` = paste(igm, "MFI"),
         regimen = factor(regimen, levels = c("2500 U/m^2, 2 h", "1500 U/m^2, 1 h"))) |>
  ggplot(aes(time, md, colour = `Anti-PEG IgM`, fill = `Anti-PEG IgM`)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 100, linetype = "dotted") +
  facet_wrap(~regimen, scales = "free_y") +
  scale_colour_manual(values = c("3.67 MFI" = "firebrick", "9.97 MFI" = "steelblue")) +
  scale_fill_manual(values = c("3.67 MFI" = "firebrick", "9.97 MFI" = "steelblue")) +
  labs(x = "Time after first dose (days)", y = "PEG-ASNase activity (U/L)") +
  theme_bw()
```

```{r fig4-table}
fig4_trough <- fig4 |>
  filter(time == 13.999) |>
  group_by(arm) |>
  summarise(sim_median = median(Cc), sim_q25 = quantile(Cc, 0.25),
            sim_q75 = quantile(Cc, 0.75), sim_below_100 = 100 * mean(Cc < 100),
            .groups = "drop") |>
  left_join(arms, by = "arm") |>
  mutate(pct_diff_median = 100 * (sim_median - median) / median,
         pct_diff_q25 = 100 * (sim_q25 - q25) / q25,
         pct_diff_q75 = 100 * (sim_q75 - q75) / q75)

fig4_trough |>
  transmute(regimen, igm = sprintf("%.2f", igm),
            published = sprintf("%.0f (%.0f-%.0f)", median, q25, q75),
            simulated = sprintf("%.0f (%.0f-%.0f)", sim_median, sim_q25, sim_q75),
            pct_diff_median, pct_below_100, sim_below_100) |>
  dplyr::rename(
    "Regimen" = regimen,
    "Anti-PEG IgM (MFI)" = igm,
    "Published median (IQR), U/L" = published,
    "Simulated median (IQR), U/L" = simulated,
    "Median difference (%)" = pct_diff_median,
    "Published < 100 U/L (%)" = pct_below_100,
    "Simulated < 100 U/L (%)" = sim_below_100
  ) |>
  knitr::kable(digits = 1, caption = "Day-14 trough after the first induction dose. Published values from Section 3.3 and ESM Table S5.")
```

```{r fig4-gate}
# Assert on the centre and on the quartiles, never on the tails: the share of
# patients below 100 U/L lives in the tail of a 200-patient cohort and is
# reported above without a gate.
stopifnot(
  abs(median(fig4_trough$pct_diff_median)) < 7,
  max(abs(fig4_trough$pct_diff_median)) < 15,
  max(abs(c(fig4_trough$pct_diff_q25, fig4_trough$pct_diff_q75))) < 20,
  # Structural ordering: at each regimen the high-IgM arm has the lower trough.
  all(fig4_trough$sim_median[fig4_trough$igm == 9.97] <
        fig4_trough$sim_median[fig4_trough$igm == 3.67])
)
```

# PKNCA validation

PKNCA recomputes exposure over the first induction dosing interval (0-14 days)
for a small covariate lattice, grouped by IgM arm. Siebel 2022 reports no NCA
table, so the reference is the closed-form `AUC(0-14)` of each virtual
patient's own solved parameters. For a dose into the head of the chain the
amount in compartment *j* is `D * exp(-(ke + kt) t) (kt t)^j / j!`, which
integrates to `D * kt^j / (ke + kt)^(j + 1) * pgamma((ke + kt) T, j + 1)`. The
closed form is for a bolus; the simulated 2-hour infusion shifts the profile by
about an hour, which changes `AUC(0-14)` by well under 1%.

```{r pknca}
lattice <- tidyr::expand_grid(
  BSA = c(0.5, 0.65, 0.78, 1.0, 1.4),
  AGE = c(3, 5.13, 12),
  SEXF = c(0, 1)
) |>
  # Keep BSA and age plausibly paired (small children are young).
  filter(!(BSA <= 0.65 & AGE == 12), !(BSA >= 1.4 & AGE == 3))

nca_subj <- tidyr::expand_grid(lattice, ABPEG_IGM = c(1.37, 9.97)) |>
  mutate(id = row_number(),
         treatment = ifelse(ABPEG_IGM > 3.67, "IgM 9.97 MFI", "IgM 1.37 MFI (median)"),
         amt = 2500 * BSA)

nca_grid <- sort(unique(c(seq(0, 1, by = 0.02), seq(1, 14, by = 0.25))))
ev_nca <- lapply(seq_len(nrow(nca_subj)), function(i) {
  s <- nca_subj[i, ]
  e <- rxode2::et(amt = s$amt, dur = 2 / 24, cmt = "central") |>
    rxode2::et(nca_grid, cmt = "central") |>
    as.data.frame()
  e$id <- s$id
  e
}) |>
  bind_rows() |>
  left_join(select(nca_subj, id, BSA, AGE, SEXF, ABPEG_IGM, treatment), by = "id") |>
  mutate(OCC = 1)

sim_nca <- rxode2::rxSolve(mod0, ev_nca, returnType = "data.frame",
                           keep = "treatment")

nca_conc <- sim_nca |>
  filter(!is.na(Cc)) |>
  transmute(id, time, Cc, treatment)
nca_dose <- nca_subj |>
  transmute(id, time = 0, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                             concu = "U/L", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id, doseu = "U")
intervals <- data.frame(start = 0, end = 14, cmax = TRUE, tmax = TRUE,
                        auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

as.data.frame(nca_res) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "IgM arm" = treatment,
    "AUC0-14d (U*day/L)" = auclast,
    "Cmax (U/L)" = cmax,
    "Tmax (day)" = tmax
  ) |>
  knitr::kable(digits = 2, caption = "PKNCA over the first induction dosing interval, median per IgM arm.")
```

```{r pknca-comparison}
auc_cf_0T <- function(D, V, CL, Q, T = 14) {
  ke <- CL / V; kt <- Q / V; j <- 0:13
  D / V * sum(kt^j / (ke + kt)^(j + 1) * pgamma((ke + kt) * T, j + 1))
}
indiv <- sim_nca |>
  filter(time == 1) |>
  select(id, vc, cl, q) |>
  left_join(select(nca_subj, id, amt, treatment), by = "id") |>
  rowwise() |>
  mutate(auc_ref = auc_cf_0T(amt, vc, cl, q)) |>
  ungroup()

reference <- indiv |>
  group_by(treatment) |>
  summarise(PPORRES = median(auc_ref), .groups = "drop") |>
  mutate(PPTESTCD = "auclast")

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = reference,
  by            = "treatment",
  params        = "auclast",
  units         = c(auclast = "U*day/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "PKNCA AUC(0-14 days) against the closed-form AUC of each virtual patient's solved parameters, median per IgM arm. * marks rows differing by >20%.")
```

```{r pknca-gate}
per_subj <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  select(id, auclast = PPORRES) |>
  left_join(indiv, by = "id") |>
  mutate(pct = 100 * (auclast - auc_ref) / auc_ref)
summary(per_subj$pct)

stopifnot(
  # Same parameters on both sides; the gap is the 2-hour infusion versus the
  # bolus closed form plus trapezoid error, so a tight bound is correct.
  max(abs(per_subj$pct)) < 1.5,
  # The arm above the cut point has the lower median exposure.
  median(per_subj$auclast[per_subj$treatment == "IgM 9.97 MFI"]) <
    median(per_subj$auclast[per_subj$treatment != "IgM 9.97 MFI"])
)
```

# Assumptions and deviations

* **Reference model versus final model.** Table 3 prints both the reference
  model (no antibody covariate, refitted to this data set) and the final
  covariate model. Only the final model is packaged, as for any base-versus-final
  model-development paper. The reference model is a refit of the Wurthwein 2021
  structure with no new covariate; its largest departure from the final model is
  the `CLinitial` change at the second induction dose (-0.125 against -0.107),
  which the authors attribute partly to the antibody effect.
* **`V`, `CLinitial` and `Qtr` are absolute, not per square metre.** The
  control stream multiplies nothing by BSA beyond the linear centred term;
  the `L/m^2` tags record the convention of quoting values "for a child with
  BSA = 1 m^2". `model()` renormalises each BSA factor to 1 at BSA = 1 m^2 so
  the `ini()` values are the printed numbers. The alternative reading suggested
  by the wording of Wurthwein 2021 fits the paper's own simulations less well
  (see above).
* **Log base.** The antibody level is log-transformed with the natural
  logarithm: the paper pairs 1.30 with 3.67 MFI and 2.30 with 9.97 MFI, and
  `exp(1.30) = 3.67`, `exp(2.30) = 9.97`.
* **Antibody level supplied per record.** The control stream's `IGMPLOG` is the
  level before the first dose of the current treatment phase. Because the
  coefficient is non-zero only at the first induction dose, only the
  pre-induction value affects predictions; users can supply it on every record.
* **Omega scale.** Table 3 gives IIV and IOV as percentages. They are taken as
  log-normal CVs, `omega^2 = log(1 + CV^2)`, the same convention used for the
  Wurthwein 2025 models of the same group; with `omega^2 = CV^2` the variances
  would be 3-4% larger.
* **NONMEM `$OMEGA BLOCK(1) SAME`** has no nlmixr2 equivalent, so occasions 2
  and 3 carry their own etas with the variance fixed equal to the occasion-1
  estimate.
* **Compartment naming.** The control stream names the states `CENTRAL` and
  `PERI2`-`PERI14`. These are not classical peripheral compartments (no
  back-flow, one shared volume), so the `transit<n>` chain prefix is used.
* **`OCC` carries the phase effects, the IOV and the antibody gate.** This is
  the authors' own encoding. Occasion codes are renumbered from `2/3/5` to
  `1..3`.
* **Simulation settings.** The published simulations do not state whether
  residual error was included; the replication above uses the individual
  prediction `Cc` without residual error. It uses 200 rather than 1000 patients
  per arm to keep the article fast to build.
* **Silent inactivation and hypersensitivity are out of scope.** Records at or
  after either were excluded before fitting.
* **Screened but not retained:** anti-PEG IgG (weaker than IgM and uninformative
  once IgM is in the model) and all post-dose antibody levels. They are listed
  in `covariatesDataExcluded`.

# Reference

Siebel C, Lanvers-Kaminsky C, Alten J, Smisek P, Nath CE, Rizzari C, Boos J,
Wurthwein G. Impact of Antibodies Against Polyethylene Glycol on the
Pharmacokinetics of PEGylated Asparaginase in Children with Acute Lymphoblastic
Leukaemia: A Population Pharmacokinetic Approach. *Eur J Drug Metab
Pharmacokinet*. 2022;47(2):187-198. <https://doi.org/10.1007/s13318-021-00741-w>
