---
title: "Inhaled fluticasone furoate and vilanterol in COPD (Siederer 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inhaled fluticasone furoate and vilanterol in COPD (Siederer 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Siederer_2016_fluticasoneFuroate"))$reference`
* Article (open access): <https://doi.org/10.1007/s13318-015-0303-4>
* Online Resource (Tables S1-S2, Figs. S1-S2): supplementary PDF
  `13318_2015_303_MOESM1_ESM.pdf` attached to the article record
  (PMC5119845).

Fluticasone furoate (FF, an inhaled corticosteroid) and vilanterol (VI, a
long-acting beta2-agonist) are co-formulated in a single dry-powder inhaler
(ELLIPTA) for maintenance treatment of COPD. Earlier characterisations of both
molecules rested on small non-compartmental studies whose profiles were heavily
censored, because inhaled systemic exposure sits close to the assay limit of
quantification. Siederer 2016 pools concentration-time data from three Phase III
COPD trials, one Phase II COPD trial and one intensively-sampled Phase I
healthy-volunteer study, and fits it with the NONMEM M3 likelihood so that the
39% (FF) and 30% (VI) of records below the limit of quantification contribute as
censored observations rather than being discarded.

**The paper contains two independent models**, stated explicitly in Sect. 2.3:
"The FF and VI concentration-time data were modeled independently." They share
neither parameters nor a likelihood, so they are extracted as two model files:

```{r load-models}
ff <- rxode2::rxode(readModelDb("Siederer_2016_fluticasoneFuroate"))
vi <- rxode2::rxode(readModelDb("Siederer_2016_vilanterol"))
```

| | Fluticasone furoate | Vilanterol |
|---|---|---|
| Structure | 2-compartment (ADVAN4 TRANS4) | 3-compartment (ADVAN11 TRANS4) |
| Absorption | first-order, `ka` | zero-order, duration `D1` |
| Analysis dataset | 1307 subjects, 11,789 obs, 39% BLQ | 1167 subjects, 10,807 obs, 30% BLQ |
| Studies | 4 (Phase II study excluded) | 5 |
| Covariates retained | race on CL/F | age + weight + study on CL/F; age + sex + smoking + study on V1/F |
| Residual error | additive on the **linear** concentration scale | additive on the **log** concentration scale |

# Units

Both model files declare `dosing = "ug"` and volumes in L, so `Cc` is in
ug/L, i.e. **ng/mL**. Siederer 2016 reports every concentration and exposure in
**pg/mL** and **pg\*h/mL**. The conversion factor is therefore 1000, applied
explicitly wherever simulated output is compared with the paper:

```{r units}
NG_PER_ML_TO_PG_PER_ML <- 1000
```

No scale factor is buried inside either model, so dose, volume and clearance stay
mutually consistent and the mass-balance identity below can be used as a gate.

# Population

Demographics are Online Resource Table S2; study designs are Table S1.

```{r population}
pop <- rxode2::rxode(readModelDb("Siederer_2016_vilanterol"))$meta$population
tibble::tibble(
  Field = c(
    "Species", "N (FF analysis)", "N (VI analysis)", "Median age",
    "Age range", "Mean weight", "Weight range", "Female", "White/Caucasian",
    "COPD", "Mean % predicted FEV1 (COPD)"
  ),
  Value = c(
    "human", "1307", "1167", "61.0 years", "18-85 years", "75.3-75.7 kg",
    "34.6-174.6 kg", "67-68%", "82%", "94%", "48.1-48.8%"
  )
) |>
  knitr::kable(caption = "Online Resource Table S2, both analysis datasets.")
```

The five contributing studies (Online Resource Table S1):

```{r studies}
tibble::tibble(
  Study = c("1", "2", "3", "4", "5"),
  Protocol = c("HZC112206", "HZC112207", "HZC110946", "HZC111348", "HZA102936"),
  Phase = c("III", "III", "III", "II", "I"),
  Population = c("COPD", "COPD", "COPD", "COPD", "Healthy"),
  `FF/VI doses (ug)` = c(
    "50/25, 100/25", "100/25, 200/25", "50/25, 100/25, 200/25",
    "400/25", "200/25, 800/100"
  ),
  `In FF analysis` = c("yes", "yes", "yes", "no (0-4 h data only)", "yes"),
  `In VI analysis` = c("yes", "yes", "yes", "yes", "yes")
) |>
  knitr::kable(caption = "Online Resource Table S1.")
```

# Source trace

Every `ini()` value, with the exact location it was read from. Both tables in the
paper print each THETA twice, as a log-scale "Ln estimate" and as its exponential
"Estimate"; the model files encode `log(<untransformed estimate>)` because that
column carries more significant figures.

```{r source-trace-ff}
tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "lka", "log(0.0523)", "Table 1 row 'ka (h-1)' (Ln estimate -2.95)",
  "lcl", "log(230)", "Table 1 row 'CL/F (L/h)'; Sect. 3.1.2 typical value",
  "lvc", "fixed(log(1.36))", "Table 1 row 'V2/F (L)' = 1.36 FIXED; Sect. 3.1.2",
  "lq", "log(268)", "Table 1 row 'Q/F (L/h)'",
  "lvp", "log(111)", "Table 1 row 'V3/F (L)'",
  "e_race_asian_east_se_cl", "-0.211", "Table 1 row 'RACE1 = 2 on CL/F' Ln estimate",
  "e_race_black_cl", "0.0602", "Table 1 row 'RACE1 = 3 on CL/F' Ln estimate",
  "e_race_asian_central_arabic_amind_oth_cl", "-0.265", "Table 1 row 'RACE1 = 4 on CL/F' Ln estimate",
  "race covariate equation", "Ln(theta) = theta_1 + COV", "Sect. 3.1.2 Eq. (a)",
  "etalcl", "fixed(0)", "Sect. 3.1.2 (IIV declared, magnitude not reported)",
  "addSd", "fixed(0)", "Sect. 3.1.2 additive error; magnitude not reported"
) |>
  knitr::kable(caption = "Fluticasone furoate model (Siederer 2016 Table 1).")
```

```{r source-trace-vi}
tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "lcl_copd", "log(94.6)", "Table 2 row 'CL/F, COPD (L/h)'; Sect. 3.2.2 typical value",
  "lcl_hvt", "log(135.6)", "Table 2 row 'CL/F, HVT (L/h)'",
  "lvc_copd", "log(639.0)", "Table 2 row 'V1/F, COPD (L)'; Sect. 3.2.2 typical value",
  "lvc_hvt", "log(159.2)", "Table 2 row 'V1/F, HVT (L)'",
  "lvp_copd", "log(177.7)", "Table 2 row 'V2/F, COPD (L)'",
  "lvp_hvt", "log(507.8)", "Table 2 row 'V2/F, HVT (L)'",
  "lq", "log(242.3)", "Table 2 row 'Q2/F (L/h)'",
  "lq2", "log(141.2)", "Table 2 row 'Q3/F (L/h)'",
  "lvp2", "log(2100.6)", "Table 2 row 'V3/F' (header unit misprinted as L/h)",
  "ld1", "log(0.098)", "Table 2 row 'D1 (h)'",
  "e_age_cl", "-0.433", "Table 2 row 'Age on CL/F, COPD' Ln estimate",
  "e_wt_cl", "0.421", "Table 2 row 'Wt on CL/F, COPD' Ln estimate",
  "e_age_vc", "-0.499", "Table 2 row 'Age on V1/F, COPD' Ln estimate",
  "e_sexf_vc", "-0.128", "Table 2 row 'Sex on V1/F, COPD' Ln estimate",
  "e_smoke_vc", "0.295", "Table 2 row 'Smoking on V1/F, COPD' Ln estimate",
  "e_study_hzc111348_cl", "-0.465", "Table 2 row 'Study 4 on CL/F, COPD' Ln estimate",
  "e_study_hzc110946_vc", "-0.358", "Table 2 row 'Study 3 on V1/F, COPD' Ln estimate",
  "e_study_hzc111348_vc", "-1.24", "Table 2 row 'Study 4 on V1/F, COPD' Ln estimate",
  "categorical covariate equation", "Ln CL = theta_1 + theta_COV x (cov - 1)", "Sect. 3.2.2 Eq. (b)",
  "continuous covariate equation", "power model, see Errata", "Sect. 3.2.2 Eq. (c), as corrected",
  "etalcl, etalvp2", "fixed(0)", "Sect. 2.5 / Sect. 3.2.2 (IIV declared, magnitudes not reported)",
  "expSd", "fixed(0)", "Sect. 3.2.2 additive error on the log scale; magnitude not reported"
) |>
  knitr::kable(caption = "Vilanterol model (Siederer 2016 Table 2).")
```

# Fluticasone furoate

## Typical clearance by race category

Sect. 3.1.2 quotes the typical CL/F for each of the four race groupings. Because
FF carries no continuous covariate and no estimable IIV, these are exact
predictions of the model, so they can be checked to the paper's printed
precision.

```{r ff-cl-by-race}
ff_race <- tibble::tibble(
  race_label = c(
    "1: White/Caucasian",
    "2: East Asian / Japanese / South East Asian",
    "3: African American",
    "4: Central Asian / White-Arabic / Am. Indian / other"
  ),
  RACE_ASIAN_EAST_SE = c(0, 1, 0, 0),
  RACE_BLACK = c(0, 0, 1, 0),
  RACE_ASIAN_CENTRAL_ARABIC_AMIND_OTH = c(0, 0, 0, 1),
  paper_cl = c(230, 186, 244, 176)
) |>
  mutate(
    model_cl = exp(
      log(230) - 0.211 * RACE_ASIAN_EAST_SE + 0.0602 * RACE_BLACK -
        0.265 * RACE_ASIAN_CENTRAL_ARABIC_AMIND_OTH
    )
  )

ff_race |>
  select(race_label, paper_cl, model_cl) |>
  rename(
    "Race grouping (RACE1)" = race_label,
    "Sect. 3.1.2 CL/F (L/h)" = paper_cl,
    "Model CL/F (L/h)" = model_cl
  ) |>
  knitr::kable(digits = 1)

# Strict: the model must round to each printed value.
stopifnot(all(round(ff_race$model_cl) == ff_race$paper_cl))
```

## Steady-state simulation

Once-daily inhalation. `ss = 1` puts the system at steady state directly rather
than integrating a 30-dose train.

```{r ff-sim}
ff_doses <- c(50, 100, 200)

ff_events <-
  tidyr::expand_grid(
    dose = ff_doses,
    race_idx = seq_len(nrow(ff_race))
  ) |>
  mutate(id = dplyr::row_number()) |>
  rowwise() |>
  group_split() |>
  lapply(function(g) {
    dose_row <- data.frame(
      id = g$id, time = 0, amt = g$dose, evid = 1, cmt = "depot",
      ii = 24, ss = 1
    )
    obs_row <- data.frame(
      id = g$id, time = seq(0, 24, by = 0.02), amt = NA_real_, evid = 0,
      cmt = "central", ii = 0, ss = 0
    )
    out <- dplyr::bind_rows(dose_row, obs_row)
    out$dose <- g$dose
    out$race_idx <- g$race_idx
    out
  }) |>
  dplyr::bind_rows() |>
  left_join(
    ff_race |> mutate(race_idx = dplyr::row_number()) |>
      select(
        race_idx, race_label, RACE_ASIAN_EAST_SE, RACE_BLACK,
        RACE_ASIAN_CENTRAL_ARABIC_AMIND_OTH
      ),
    by = "race_idx"
  ) |>
  arrange(id, time, dplyr::desc(evid))

ff_sim <- rxode2::rxSolve(
  ff, ff_events,
  keep = c("dose", "race_label"),
  returnType = "data.frame"
)
```

```{r ff-profile-plot, fig.cap = "Steady-state fluticasone furoate plasma concentration over one 24 h dosing interval, by nominal dose and race grouping. Compare the shape with Fig. 2 of Siederer 2016 (VPC by dose regimen)."}
ff_sim |>
  filter(!is.na(Cc)) |>
  mutate(Cc_pg = Cc * NG_PER_ML_TO_PG_PER_ML) |>
  ggplot(aes(time, Cc_pg, colour = race_label)) +
  geom_line() +
  facet_wrap(~ paste0(dose, " ug once daily")) +
  labs(
    x = "Time after dose (h)", y = "FF concentration (pg/mL)",
    colour = "RACE1 grouping"
  ) +
  theme_bw() +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

The profile is strikingly flat: the estimated `ka` of 0.0523 /h corresponds to an
absorption half-life of 13.3 h against a disposition half-life well under an
hour, so inhaled FF is deeply flip-flop and the terminal slope is the absorption
rate. That is the pharmacology the model encodes, and it is why trough
concentrations in the Phase III trials were frequently quantifiable despite a
very low peak.

## Mass-balance gate

At steady state, AUC over the dosing interval must equal dose / CL exactly. This
gate uses no published number, so it cannot be satisfied by a transcription
error in either direction -- it tests the ODE system, the compartment mapping and
the unit declaration together.

```{r ff-mass-balance}
ff_auc <- ff_sim |>
  filter(!is.na(Cc)) |>
  group_by(id, dose, race_label) |>
  summarise(
    auc_tau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  left_join(
    ff_race |> select(race_label, model_cl), by = "race_label"
  ) |>
  mutate(
    dose_over_cl = dose / model_cl,
    rel_err = (auc_tau - dose_over_cl) / dose_over_cl
  )

# The residual is pure trapezoidal quadrature error on the observation grid,
# and converges as O(h^2): the relative error is 1.16e-3 at a 0.1 h grid,
# 5.41e-4 at 0.05 h, 1.76e-4 at 0.02 h (used here) and 6.12e-5 at 0.01 h.
# A structural fault -- wrong compartment mapping, wrong unit declaration,
# a mis-transcribed clearance -- moves this by tens of percent, so 1e-3
# is a strict gate at this grid density and not a tuned tolerance.
stopifnot(max(abs(ff_auc$rel_err)) < 1e-3)
sprintf(
  "AUC(0-tau) matches dose/CL for all %d profiles; max relative error %.2e",
  nrow(ff_auc), max(abs(ff_auc$rel_err))
)
```

## Non-compartmental analysis

The race grouping is carried into the PKNCA formula alongside dose, so each of
the twelve dose x race profiles gets its own NCA rather than being collapsed.
That matters for the comparison below: Table 3's "FF dose" rows pool the whole
dataset, which is 82% White/Caucasian, so the correct comparator for them is the
White/Caucasian profile -- not a median over four equally-weighted race arms,
which would sit about 12% high.

```{r ff-nca}
ff_conc <- ff_sim |>
  filter(!is.na(Cc)) |>
  transmute(
    id = factor(id), time, treatment = paste0(dose, " ug"),
    race_label,
    Cc = Cc * NG_PER_ML_TO_PG_PER_ML
  )

ff_dose_df <- ff_events |>
  filter(evid == 1) |>
  transmute(
    id = factor(id), time, amt = dose, treatment = paste0(dose, " ug"),
    race_label
  )

ff_conc_obj <- PKNCA::PKNCAconc(
  ff_conc, Cc ~ time | treatment + race_label + id,
  concu = "pg/mL", timeu = "h"
)
ff_dose_obj <- PKNCA::PKNCAdose(
  ff_dose_df, amt ~ time | treatment + race_label + id,
  doseu = "ug"
)

ff_intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

ff_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(ff_conc_obj, ff_dose_obj, intervals = ff_intervals)
)

# White/Caucasian reference profiles only, for comparison with Table 3.
ff_white <- as.data.frame(ff_nca$result) |>
  filter(race_label == ff_race$race_label[1]) |>
  select(treatment, PPTESTCD, PPORRES)
```

## Comparison against the published race effect

Siederer 2016 Table 3 reports model-predicted steady-state AUC0-24 stratified by
dose and race grouping. The **ratio** of each race group's AUC to the
White/Caucasian AUC is a scale-free test of the covariate coefficients, and it is
the comparison the model can be held to: AUC is inversely proportional to CL/F,
so the predicted ratio is simply `exp(-coefficient)`.

```{r ff-race-ratio}
table3_auc <- tibble::tribble(
  ~dose, ~race1, ~race2, ~race3, ~race4, ~n2, ~n3, ~n4,
  50, 79.05, 102.77, 81.19, 99.25, 37L, 16L, 4L,
  100, 176.04, 223.04, 128.25, 202.34, 111L, 13L, 9L,
  200, 319.68, 394.61, 245.42, 352.68, 20L, 17L, 3L
)

ratio_tbl <- table3_auc |>
  transmute(
    Dose = paste0(dose, " ug"),
    `RACE1 = 2` = race2 / race1,
    `RACE1 = 3` = race3 / race1,
    `RACE1 = 4` = race4 / race1
  ) |>
  tidyr::pivot_longer(-Dose, names_to = "Grouping", values_to = "Table 3 ratio") |>
  mutate(
    `Model ratio` = dplyr::case_when(
      Grouping == "RACE1 = 2" ~ exp(0.211),
      Grouping == "RACE1 = 3" ~ exp(-0.0602),
      Grouping == "RACE1 = 4" ~ exp(0.265)
    ),
    `% diff` = 100 * (`Table 3 ratio` - `Model ratio`) / `Model ratio`
  )

ratio_tbl |>
  knitr::kable(
    digits = c(0, 0, 3, 3, 1),
    caption = paste(
      "AUC0-24 ratio versus White/Caucasian. The model ratio is exp(-coefficient)",
      "and does not depend on dose."
    )
  )
```

`RACE1 = 2` is the only race coefficient the paper estimates well (%RSE 28.5,
N = 111 at the 100 ug dose, and the 95% CI excludes 1). Its Table 3 ratio must
therefore match the model closely at every dose:

```{r ff-race2-gate}
race2 <- ratio_tbl |> filter(Grouping == "RACE1 = 2")
stopifnot(
  # Structural: a mis-signed or mis-scaled race coefficient blows this instantly.
  max(abs(race2$`% diff`)) < 6
)
sprintf(
  "RACE1 = 2 AUC ratio reproduced at all three doses; max |%% diff| = %.1f%%",
  max(abs(race2$`% diff`))
)
```

`RACE1 = 3` (%RSE 199, 95% CI spans 1, N = 13-17) and `RACE1 = 4` (%RSE 50.6,
N = 3-9) scatter widely, which is exactly what Sect. 3.1.2 warns about: "the
fixed-effect parameter RACE1 = 3 was estimated with less precision (%RSE 199 %)
and should thus be interpreted with caution". Table 3's values are summaries of
*post hoc* individual estimates in cells holding as few as three subjects, so
they are not held to a tolerance here.

## Comparison against published absolute exposures

```{r ff-nca-comparison}
table3_abs <- tibble::tribble(
  ~treatment, ~cmax, ~auclast,
  "50 ug", 7.52, 82.92,
  "100 ug", 11.73, 181.82,
  "200 ug", 21.62, 319.69
)

ff_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = ff_white,
  reference = table3_abs,
  by = "treatment",
  params = c("cmax", "auclast"),
  units = c(cmax = "pg/mL", auclast = "pg*h/mL")
)
knitr::kable(ff_cmp, digits = 1)
attr(ff_cmp, "footnote")
```

**Every row is flagged, all in the same direction and by a similar factor.**
This is a self-consistency problem inside the paper, not a transcription error in
the model file, and it is worked through in the Errata below. The key observation
is that in the best-determined dose stratum the discrepancy is a pure scale
factor, leaving the profile shape untouched:

```{r ff-scale-factor}
# Work from the PKNCA result directly rather than from the formatted
# comparison table, so the arithmetic does not depend on display formatting.
ff_wide <- ff_white |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

ff_scale <- ff_wide |>
  left_join(table3_abs, by = "treatment", suffix = c("_model", "_table3")) |>
  mutate(
    cmax_ratio = cmax_model / cmax_table3,
    auc_ratio = auclast_model / auclast_table3,
    auc_over_cmax_model = auclast_model / cmax_model,
    auc_over_cmax_table3 = auclast_table3 / cmax_table3
  )

ff_scale |>
  select(
    treatment, cmax_ratio, auc_ratio,
    auc_over_cmax_model, auc_over_cmax_table3
  ) |>
  rename(
    "Dose" = treatment,
    "Cmax model/Table 3" = cmax_ratio,
    "AUC model/Table 3" = auc_ratio,
    "AUC/Cmax model (h)" = auc_over_cmax_model,
    "AUC/Cmax Table 3 (h)" = auc_over_cmax_table3
  ) |>
  knitr::kable(digits = 2)

ff_100 <- ff_scale |> filter(treatment == "100 ug")

sprintf(
  paste(
    "AUC offset present at every dose (%.2fx to %.2fx).",
    "At the 100 ug stratum (N = 724, the best-determined) the offset is",
    "%.3fx on Cmax and %.3fx on AUC -- the same number, so the profile shape",
    "is unaffected: AUC/Cmax is %.2f h in the model and %.2f h in Table 3."
  ),
  min(ff_scale$auc_ratio), max(ff_scale$auc_ratio),
  ff_100$cmax_ratio, ff_100$auc_ratio,
  ff_100$auc_over_cmax_model, ff_100$auc_over_cmax_table3
)

stopifnot(
  # The AUC offset is large and present at every dose.
  all(ff_scale$auc_ratio > 2.2), all(ff_scale$auc_ratio < 3.0),
  # At the best-determined stratum the offset is a pure scale factor: it is
  # the same for Cmax and for AUC, so the shape statistic is preserved.
  abs(ff_100$cmax_ratio - ff_100$auc_ratio) < 0.05,
  abs(ff_100$auc_over_cmax_model - ff_100$auc_over_cmax_table3) /
    ff_100$auc_over_cmax_table3 < 0.02
)
```

The 100 ug stratum is the one to read: it holds 724 of the 1357 subject-dose
records in Table 3, and there the offset is a clean scale factor -- 2.39x on
both Cmax and AUC, leaving the shape statistic AUC/Cmax at 15.5 h in both. The
50 ug stratum departs furthest (1.86x on Cmax, and an AUC/Cmax of 11.0 h against
the model's 15.5 h), which is expected: Sect. 3.1.1 reports that 72% of samples
at FF/VI 50/25 ug were below the limit of quantification, and Sect. 4 notes that
average exposure at that dose "is predicted to be BLQ of the analytical assay".
Table 3's own dose rows are not dose-proportional either (AUC rises 2.19-fold
from 50 to 100 ug and 1.76-fold from 100 to 200 ug), so no single constant can
reconcile Table 3 with *any* linear model; the dose groups contain different
subjects with different race mixes and different degrees of censoring.

The shape of the profile -- and therefore `ka`, `Q/F` and `V3/F` -- is reproduced
exactly; only the dose-to-clearance scale differs.

# Vilanterol

## Typical parameter values

Sect. 3.2.2 quotes typical CL/F and V1/F for several covariate combinations.
These are exact model predictions and are checked to printed precision.

```{r vi-typical}
vi_typical <- tibble::tribble(
  ~Quantity, ~Condition, ~Paper, ~Model,
  "CL/F (L/h)", "COPD, 60 y, 70 kg, Studies 1-3", 94.6, exp(log(94.6)),
  "CL/F (L/h)", "COPD, 60 y, 70 kg, Study 4", 59.4, exp(log(94.6) - 0.465),
  "V1/F (L)", "COPD, 60 y, non-smoking male, Studies 1-2", 639, exp(log(639.0)),
  "V1/F (L)", "COPD, 60 y, non-smoking male, Study 3", 447, exp(log(639.0) - 0.358),
  "V1/F (L)", "COPD, 60 y, non-smoking male, Study 4", 185, exp(log(639.0) - 1.24)
) |>
  mutate(`% diff` = 100 * (Model - Paper) / Paper)

knitr::kable(vi_typical, digits = c(0, 0, 1, 1, 2))
stopifnot(max(abs(vi_typical$`% diff`)) < 0.3)
```

The covariate *magnitudes* quoted in Sect. 3.2.2 are equally exact:

```{r vi-covariate-magnitudes}
vi_cov <- tibble::tribble(
  ~Statement, ~`Paper`, ~`Model`,
  "CL/F decrease over age 41-84 y", 0.27, 1 - exp(-0.433 * log(84 / 41)),
  "CL/F reduction, 160 kg down to 35 kg", 0.47, 1 - exp(0.421 * log(35 / 160)),
  "V1/F decrease over age 41-84 y", 0.30, 1 - exp(-0.499 * log(84 / 41)),
  "V1/F lower in females", 0.12, 1 - exp(-0.128),
  "V1/F increase with current smoking", 0.34, exp(0.295) - 1
)
vi_cov |>
  mutate(across(c(Paper, Model), ~ scales::percent(.x, accuracy = 0.1))) |>
  knitr::kable()

# Strict: each quoted percentage must be reproduced to within half a
# percentage point. This is the gate that falsifies the printed form of
# Sect. 3.2.2 Eq. (c) in favour of the power model -- see Errata.
stopifnot(max(abs(vi_cov$Paper - vi_cov$Model)) < 0.005)
```

## Steady-state simulation by study stratum

```{r vi-sim}
vi_arms <- tibble::tribble(
  ~arm, ~DIS_COPD, ~AGE, ~WT, ~SEXF, ~SMOKE, ~STUDY_HZC110946, ~STUDY_HZC111348,
  "Study 1/2 (HZC112206/7)", 1, 60, 70, 0, 0, 0, 0,
  "Study 3 (HZC110946)", 1, 60, 70, 0, 0, 1, 0,
  "Study 4 (HZC111348)", 1, 60, 70, 0, 0, 0, 1,
  "Study 1/2, female", 1, 60, 70, 1, 0, 0, 0,
  "Study 1/2, current smoker", 1, 60, 70, 0, 1, 0, 0,
  "Healthy volunteers (HZA102936)", 0, 40, 75, 0, 0, 0, 0
) |>
  mutate(id = dplyr::row_number())

vi_events <- vi_arms |>
  group_split(id) |>
  lapply(function(g) {
    dose_row <- data.frame(
      id = g$id, time = 0, amt = 25, evid = 1, cmt = "central",
      rate = -2, ii = 24, ss = 1
    )
    obs_row <- data.frame(
      id = g$id, time = seq(0, 24, by = 0.02), amt = NA_real_, evid = 0,
      cmt = "central", rate = 0, ii = 0, ss = 0
    )
    dplyr::bind_rows(dose_row, obs_row)
  }) |>
  dplyr::bind_rows() |>
  left_join(vi_arms, by = "id") |>
  arrange(id, time, dplyr::desc(evid))

vi_sim <- rxode2::rxSolve(
  vi, vi_events, keep = "arm", returnType = "data.frame"
)
```

```{r vi-profile-plot, fig.cap = "Steady-state vilanterol plasma concentration over one 24 h dosing interval after 25 ug once daily, by study stratum and covariate contrast. Compare with Fig. 4 of Siederer 2016 (VPC stratified by study)."}
vi_sim |>
  filter(!is.na(Cc), time <= 8) |>
  mutate(Cc_pg = Cc * NG_PER_ML_TO_PG_PER_ML) |>
  ggplot(aes(time, Cc_pg, colour = arm)) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "VI concentration (pg/mL, log scale)",
    colour = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

## Non-compartmental analysis

```{r vi-nca}
vi_conc <- vi_sim |>
  filter(!is.na(Cc)) |>
  transmute(
    id = factor(id), time, treatment = arm,
    Cc = Cc * NG_PER_ML_TO_PG_PER_ML
  )

vi_dose_df <- vi_events |>
  filter(evid == 1) |>
  transmute(id = factor(id), time, amt = 25, treatment = arm)

vi_conc_obj <- PKNCA::PKNCAconc(
  vi_conc, Cc ~ time | treatment + id,
  concu = "pg/mL", timeu = "h"
)
vi_dose_obj <- PKNCA::PKNCAdose(
  vi_dose_df, amt ~ time | treatment + id,
  doseu = "ug"
)

vi_nca <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(
    vi_conc_obj, vi_dose_obj,
    intervals = data.frame(
      start = 0, end = 24,
      cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
    )
  )
)
```

## Comparison against published exposures

Siederer 2016 Table 4 reports model-predicted steady-state Cmax and AUC0-24 by
study. The three COPD study strata are compared directly.

```{r vi-nca-comparison}
table4 <- tibble::tribble(
  ~treatment, ~cmax, ~auclast,
  "Study 1/2 (HZC112206/7)", 43.2, 273.7,
  "Study 3 (HZC110946)", 49.7, 249.2,
  "Study 4 (HZC111348)", 120.5, 408.2
)

vi_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = vi_nca,
  reference = table4,
  by = "treatment",
  params = c("cmax", "auclast"),
  units = c(cmax = "pg/mL", auclast = "pg*h/mL")
)
knitr::kable(vi_cmp, digits = 1)
attr(vi_cmp, "footnote")
```

The Study 1/2 stratum -- 1002 of the 1091 COPD subjects, and the reference
category that carries no study effect -- reproduces to well under 5% on both
parameters. That is the strict gate:

```{r vi-gate}
vi_pct <- vi_cmp |>
  mutate(pct = suppressWarnings(as.numeric(gsub("[^0-9.eE+-]", "", `% diff`))))

ref_rows <- vi_pct |> filter(treatment == "Study 1/2 (HZC112206/7)")
stopifnot(
  # Structural: a mis-transcribed CL/F, V1/F, D1 or dose, or a wrong
  # compartment mapping, moves these by tens of percent.
  max(abs(ref_rows$pct)) < 5
)

# The two small study strata (N = 50 and N = 39 subjects) are held to a
# looser envelope. Each simulated arm here is a single typical subject --
# a 60-year-old, 70 kg, non-smoking male -- whereas the matching Table 4
# row is a geometric mean over a cohort that is 68% female and 53% current
# smokers (Online Resource Table S2). Current smoking alone raises V1/F by
# 34% and so lowers Cmax by about a fifth, which is why the volume-driven
# Cmax rows sit high while the clearance-driven AUC rows do not.
stopifnot(max(abs(vi_pct$pct), na.rm = TRUE) < 20)

sprintf(
  "Study 1/2 reference stratum: max |%% diff| = %.1f%%; all strata: max = %.1f%%",
  max(abs(ref_rows$pct)), max(abs(vi_pct$pct), na.rm = TRUE)
)
```

A sharper structural check is available for Study 3, because Siederer 2016
retains a study effect on V1/F for HZC110946 but **none on CL/F**. Steady-state
AUC over a dosing interval depends only on clearance, so the model's Study 3
AUC0-24 must be *identical* to its Studies 1/2 value. Mis-assigning the Study 3
coefficient to clearance instead of volume -- an easy transcription slip, since
the Study 4 coefficient acts on both -- would break this immediately:

```{r vi-study3-auc-identity}
vi_summary <- as.data.frame(vi_nca$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

auc_s12 <- vi_summary$auclast[vi_summary$treatment == "Study 1/2 (HZC112206/7)"]
auc_s3 <- vi_summary$auclast[vi_summary$treatment == "Study 3 (HZC110946)"]

sprintf(
  "AUC0-24: Studies 1/2 = %.3f, Study 3 = %.3f pg*h/mL (relative difference %.2e)",
  auc_s12, auc_s3, abs(auc_s3 - auc_s12) / auc_s12
)

# The residual is trapezoidal quadrature error: Study 3's smaller V1/F gives a
# sharper peak and so a slightly different quadrature bias. It converges O(h^2)
# -- 1.18e-5 at the 0.02 h grid used here, 5.21e-6 at 0.01 h, 1.95e-6 at
# 0.005 h. Assigning the Study 3 coefficient to clearance instead of volume
# would instead give exp(-0.358), a 43% difference, so this is a strict gate.
stopifnot(abs(auc_s3 - auc_s12) / auc_s12 < 1e-4)

# Both also sit within 3e-5 of dose / CL = 25 ug / 94.6 L/h, so vilanterol
# satisfies the same steady-state mass-balance identity checked for
# fluticasone furoate above.
stopifnot(abs(auc_s12 - 1000 * 25 / 94.6) / (1000 * 25 / 94.6) < 1e-4)
```

Table 4 nevertheless reports 273.7 and 249.2 pg\*h/mL for those two strata, a 9%
spread the final model attributes to nothing -- a reminder that Table 4 rows are
summaries of *post hoc* individual estimates in cohorts of very different size
(N = 496 and N = 50), not model predictions at a common covariate vector.

The Study 4 contrast that Sect. 4 highlights is also reproduced:

```{r vi-study4-contrast}
vi_summary <- as.data.frame(vi_nca$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

s12 <- vi_summary |> filter(treatment == "Study 1/2 (HZC112206/7)")
s4 <- vi_summary |> filter(treatment == "Study 4 (HZC111348)")

tibble::tibble(
  Contrast = c("Cmax, Study 4 / Study 1-2", "AUC0-24, Study 4 / Study 1-2"),
  `Sect. 4` = c("~2.7-fold", "~1.5-fold"),
  Model = c(s4$cmax / s12$cmax, s4$auclast / s12$auclast)
) |>
  knitr::kable(digits = 2)

stopifnot(
  s4$cmax / s12$cmax > 2.5, s4$cmax / s12$cmax < 3.2,
  s4$auclast / s12$auclast > 1.4, s4$auclast / s12$auclast < 1.8
)
```

## Smoking and sex contrasts

Table 4 also summarises exposure by sex and smoking status. These rows pool all
studies and all covariate values, so they are *not* clean isolations of the
model coefficient -- they are marginal summaries over a cohort in which age,
weight and study membership also differ between the strata.

```{r vi-sex-smoke}
contrast_tbl <- vi_summary |>
  filter(treatment %in% c(
    "Study 1/2 (HZC112206/7)", "Study 1/2, female", "Study 1/2, current smoker"
  ))

base <- contrast_tbl |> filter(treatment == "Study 1/2 (HZC112206/7)")
fem <- contrast_tbl |> filter(treatment == "Study 1/2, female")
smk <- contrast_tbl |> filter(treatment == "Study 1/2, current smoker")

tibble::tibble(
  Contrast = c("Cmax, female / male", "Cmax, current / former smoker"),
  `Table 4 (marginal)` = c(42.1 / 45.7, 38.3 / 49.4),
  `Model (all else equal)` = c(fem$cmax / base$cmax, smk$cmax / base$cmax)
) |>
  knitr::kable(digits = 3)
```

The smoking contrast survives marginalisation almost intact -- Table 4 gives
0.775 and Sect. 4 quotes "22 %" lower Cmax in current smokers, against 0.746
all-else-equal from the model. The sex contrast does not: holding everything else
fixed the model predicts females have a *higher* Cmax (12% lower V1/F), yet Table
4's marginal female/male ratio is 0.92. The paper reaches the same conclusion
from its own numbers -- Sect. 4: "there appeared to be no notable difference in
VI systemic exposure for males compared with females (Table 4)". A marginal
summary over a cohort whose strata differ in several correlated covariates can
mask and even invert a single covariate's effect, so only the isolated
coefficient (checked above against Table 2) is gated.

## Covariate-driven cohort

With no estimable IIV, all between-subject variability in this model comes from
the covariates. A virtual COPD cohort sampled over the Table S2 demographic
ranges shows the spread the covariate model alone produces, which is the
quantity Online Resource Fig. S2 plots.

```{r vi-cohort}
set.seed(20260916)
n_cohort <- 200

cohort <- tibble::tibble(
  id = seq_len(n_cohort),
  AGE = round(runif(n_cohort, 41, 84)),
  WT = round(rlnorm(n_cohort, meanlog = log(73), sdlog = 0.22)),
  SEXF = rbinom(n_cohort, 1, 0.68),
  SMOKE = rbinom(n_cohort, 1, 0.53),
  DIS_COPD = 1,
  STUDY_HZC110946 = 0,
  STUDY_HZC111348 = 0
) |>
  mutate(WT = pmin(pmax(WT, 35), 160))

cohort_events <- cohort |>
  group_split(id) |>
  lapply(function(g) {
    dplyr::bind_rows(
      data.frame(
        id = g$id, time = 0, amt = 25, evid = 1, cmt = "central",
        rate = -2, ii = 24, ss = 1
      ),
      data.frame(
        id = g$id, time = seq(0, 24, by = 0.1), amt = NA_real_, evid = 0,
        cmt = "central", rate = 0, ii = 0, ss = 0
      )
    )
  }) |>
  dplyr::bind_rows() |>
  left_join(cohort, by = "id") |>
  arrange(id, time, dplyr::desc(evid))

cohort_sim <- rxode2::rxSolve(
  vi, cohort_events, keep = c("AGE", "WT", "SEXF", "SMOKE"),
  returnType = "data.frame"
)

cohort_exposure <- cohort_sim |>
  filter(!is.na(Cc)) |>
  group_by(id, AGE, WT) |>
  summarise(
    cmax = max(Cc) * NG_PER_ML_TO_PG_PER_ML,
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) *
      NG_PER_ML_TO_PG_PER_ML,
    .groups = "drop"
  )
```

```{r vi-cohort-plot, fig.cap = "Steady-state vilanterol AUC0-24 and Cmax against age and bodyweight in a 200-subject virtual COPD cohort. Replicates Online Resource Fig. S2 of Siederer 2016."}
cohort_exposure |>
  tidyr::pivot_longer(c(cmax, auc), names_to = "param", values_to = "value") |>
  tidyr::pivot_longer(c(AGE, WT), names_to = "cov", values_to = "cov_value") |>
  mutate(
    param = dplyr::recode(param, cmax = "Cmax (pg/mL)", auc = "AUC0-24 (pg*h/mL)"),
    cov = dplyr::recode(cov, AGE = "Age (years)", WT = "Weight (kg)")
  ) |>
  ggplot(aes(cov_value, value)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  facet_grid(param ~ cov, scales = "free") +
  labs(x = NULL, y = NULL) +
  theme_bw()
```

```{r vi-cohort-summary}
tibble::tibble(
  Statistic = c("Geometric mean", "5th percentile", "95th percentile"),
  `Cmax (pg/mL)` = c(
    exp(mean(log(cohort_exposure$cmax))),
    quantile(cohort_exposure$cmax, 0.05),
    quantile(cohort_exposure$cmax, 0.95)
  ),
  `AUC0-24 (pg*h/mL)` = c(
    exp(mean(log(cohort_exposure$auc))),
    quantile(cohort_exposure$auc, 0.05),
    quantile(cohort_exposure$auc, 0.95)
  )
) |>
  knitr::kable(digits = 1)

# Centre of the covariate-driven cohort against Table 4 'All studies'
# (Cmax 43.2 pg/mL, AUC0-24 265.7 pg*h/mL). Robust to the sampled cohort,
# because the assertion is on the geometric mean, not on an extreme.
gm_auc <- exp(mean(log(cohort_exposure$auc)))
stopifnot(abs(gm_auc - 265.7) / 265.7 < 0.15)
sprintf("Cohort geometric-mean AUC0-24 %.1f vs Table 4 'All studies' 265.7 pg*h/mL", gm_auc)
```

Sect. 4 notes "there was no notable trend in predicted systemic exposure with
increasing age or decreasing bodyweight", and the loess curves above show why:
age lowers CL/F (raising AUC) and V1/F (raising Cmax) simultaneously, while
weight raises CL/F, so the net covariate-driven spread in exposure is modest
relative to the between-study difference.

# Assumptions and deviations

## Errata and corrections applied

**1. The continuous-covariate equation is printed without its inner logarithm.**
Siederer 2016 Sect. 3.2.2 prints

> Ln CL = theta_1 + theta_COV x (covariate / median)

but the model files implement the power form

> Ln P = theta_1 + theta_COV x Ln(covariate / reference)

Three independent statements in the paper falsify the printed form and confirm
the power form:

* Sect. 3.2.2 quotes "a reduction (47 %) in inhaled clearance ... with
  decreasing bodyweight (range of 160-35 kg)". The power form gives
  `exp(0.421 * log(35/160)) = 0.527`, a 47.3% reduction. The printed
  linear-ratio form gives `exp(0.421 * (35 - 160)/70) = 0.471`, a 52.9%
  reduction.
* `exp(4.55) = 94.6 L/h` is quoted as the typical CL/F at *exactly* 60 years and
  70 kg, and `exp(6.46) = 639 L` as the typical V1/F at *exactly* 60 years. Both
  require the covariate term to vanish at the reference, which only the power
  form does; the printed form would leave `theta_COV x 1` behind and give 93.5
  L/h and 388 L instead.
* Sect. 3.2.2 quotes a 30% decrease in V1/F over 41-84 years; the power form
  gives `exp(-0.499 * log(84/41)) = 0.699`.

The gate in "Typical parameter values" above re-runs all of these on every
render, so the correction cannot silently regress.

**2. Reference values for the power model are 60 years and 70 kg**, taken from
the Sect. 3.2.2 typical-value statements ("a subject with COPD (aged 60 years and
weighing 70 kg)"; "non-smoking male, aged 60 years"). The word used in the
printed equation is "median", and Online Resource Table S2 gives a median age of
61.0 years and a *mean* weight of 75.3 kg for the VI dataset, so the model's
reference values are the rounded standards the results text uses rather than the
tabulated dataset statistics. Using 61 y / 75.3 kg would shift `exp(theta_1)`
away from the paper's own quoted typical values.

**3. Table 2's `V3/F` row header misprints the unit as `L/h`.** The table
footnote ("V2/F, V3/F volumes of peripheral compartment") and dimensional
analysis both confirm it is a volume (L). Encoded as a volume.

**4. Table 2's `D1` footnote reads "input duration as FF/VI or FF alone"** in a
vilanterol parameter table. D1 is the vilanterol zero-order input duration; the
footnote text appears to have been copied from the fluticasone furoate table.

**5. Fluticasone furoate Table 1 and Table 3 are mutually inconsistent by a
factor of roughly 2.4-2.7 on AUC.** This is the largest deviation in the
extraction and it is *not* resolvable from the sources on disk, so nothing was
tuned.

* Table 1 is internally consistent four ways: its "Ln estimate" and "Estimate"
  columns are exact exponentials of each other, its confidence intervals agree,
  and it reproduces all four typical CL/F values quoted in Sect. 3.1.2 (230,
  186, 244, 176 L/h) exactly -- see the gate in "Typical clearance by race
  category".
* The model built from Table 1 satisfies the steady-state mass-balance identity
  `AUC(0-tau) = dose / CL` to machine precision, so the ODE system, compartment
  mapping and unit declaration are correct.
* Nonetheless the model predicts steady-state AUC0-24 that is 2.6x, 2.4x and
  2.7x *higher* than Table 3's model-predicted exposures at 50, 100 and 200 ug
  respectively. At the 100 ug stratum -- which holds 724 of Table 3's
  subject-dose records and is by far the best determined -- the offset is a
  clean scale factor: 2.388x on Cmax and 2.391x on AUC, leaving Table 3's own
  AUC/Cmax ratio of 15.50 h essentially identical to the model's 15.52 h. The
  profile *shape* therefore agrees and only the dose-to-clearance scale differs.
  Reproducing Table 3's absolute numbers at 100 ug would require CL/F = 549 L/h,
  or an effective FF dose of 41.9 ug in place of 100 ug; neither appears
  anywhere in the paper. The emitted-dose footnote in Online Resource Table S1
  ("Emitted doses 46/22, 92/22, 184/22 and 368/22 ug") accounts for only a
  factor of 1.09.
* The offset is not perfectly constant across doses, because Table 3 is itself
  not dose-proportional: its AUC rises 2.19-fold from 50 to 100 ug and
  1.76-fold from 100 to 200 ug, whereas a linear model is exactly
  dose-proportional. Some departure is expected -- Sect. 2.5 defines AUC0-24 as
  nominal dose divided by each subject's *post hoc* CL/F, and the dose groups
  contain different subjects with different race mixes -- but it means no single
  constant reconciles Table 3 with any linear model. The 50 ug stratum deviates
  most (AUC/Cmax 11.0 h against the model's 15.5 h), which is consistent with
  Sect. 3.1.1's report that 72% of its samples were below the limit of
  quantification.
* The equivalent check for vilanterol passes cleanly: Table 2's `exp(4.55) =
  94.6 L/h` and a 25 ug dose give 264.3 pg\*h/mL against Table 4's "All studies"
  geometric mean of 265.7 pg\*h/mL, a 0.5% agreement. So the inconsistency is
  specific to the fluticasone furoate analysis, not a systematic units error
  across the paper.
* **The paper's own visual predictive check sides with Table 1.** Fig. 2 panel
  (b) is the 100 ug stratum, plotted on a linear pg/mL axis. Its observed
  concentration cloud spans roughly 0-90 pg/mL with the observed median running
  from about 45 pg/mL early in the interval down to about 25 pg/mL at 24 h, and
  the model prediction intervals bracket it. A geometric-mean Cmax of 11.73
  pg/mL and Cavg of 7.6 pg/mL -- what Table 3 reports for that same 100 ug
  stratum -- cannot be reconciled with that figure, whereas the Table 1 model
  simulated here (Cmax 28.0 pg/mL, Cavg 18.1 pg/mL) sits inside it. Table 3,
  not Table 1, is the outlier.
* The model file therefore encodes **Table 1**, which is the paper's stated final
  parameter table and the only self-consistent description of the model. Users
  reproducing Table 3's absolute exposures should be aware that Table 3 cannot be
  regenerated from Table 1. Sect. 3.1.2 notes that the 50/25 ug treatment was
  excluded during model building and that "this model and the base model were
  re-run including all treatments" afterwards; if Table 1 and Table 3 come from
  the two different runs, that would explain a systematic offset, but the paper
  does not say so.

## Parameters not reported by the source

**Inter-individual variability magnitudes are not reported anywhere.** Both
Sect. 3.1.2 and Sect. 3.2.2 state that "Inter-individual variances (exponential
model) were estimated with reasonable precision" and give only %RSE bounds
(<= 36% for FF, <= 25% for VI except V3/F at 64%). Neither Table 1, Table 2 nor
the Online Resource lists an OMEGA row. Per project policy, no variance was
invented. Instead, etas are declared at `fixed(0)` on only the parameters for
which the paper gives positive evidence of an estimated IIV:

* FF: `etalcl`, because Sect. 2.5 derives "individual post hoc estimates of CL/F".
* VI: `etalcl` for the same reason, and `etalvp2`, named explicitly in Sect. 3.2.2
  ("ETA on volume of the peripheral compartment (V3/F) where %RSE was 64 %").

Consequently **both models are deterministic**: simulated between-subject
variability comes only from covariates. Every gate in this vignette is an exact
comparison rather than a distributional one, which is why tight tolerances are
used throughout. Anyone needing a stochastic simulation must supply their own
omega.

**Residual-error magnitudes are not reported either**, so `addSd` (FF) and
`expSd` (VI) are also `fixed(0)`.

**The residual-error scale differs between the two models**, and the paper does
not say so in prose -- both simply say "An additive error model described the
residual variability." The figures settle it: Fig. 1 (FF goodness-of-fit) has
axes "Observed FF Concentration (pg/mL)" against "Population Prediction
(pg/mL)", untransformed, so FF's additive error is additive in pg/mL and maps to
`add()`. Fig. 3 (VI goodness-of-fit) has axes "Observed log-Value (pg/mL)"
against "Population log-Prediction (pg/mL)", and the Fig. 4 VPC axis is "LN
concentration (pg/mL)", so VI was fitted on the natural-log concentration scale
and its additive error is log-normal in linear space, mapping to `lnorm()`.

**VI residual error was separated by population** (Sect. 3.2.2: "CL/F, V1/F,
volume of the peripheral compartment (V2/F), and residual error were separated by
population"). Neither stratum's magnitude is reported, so a single `expSd` is
declared.

**Which VI parameters carried fixed IIV in the base model is stated but not
quantified.** Sect. 3.2.2 notes "inter-subject variability on D1, Q2/F, and V2/F
was fixed", without giving the fixed values, and describes the *structural base
model* rather than the final model.

## Other assumptions

* **Race grouping.** Siederer 2016's `RACE1` is a paper-defined collapse of a
  nine-level race column driven by small per-category N (Sect. 2.3.1). Its
  groupings do not correspond to any race indicator already in the nlmixr2lib
  covariate register, so two new composite covariates were registered:
  `RACE_ASIAN_EAST_SE` (RACE1 = 2) and `RACE_ASIAN_CENTRAL_ARABIC_AMIND_OTH`
  (RACE1 = 4). `RACE1 = 3` maps onto the existing `RACE_BLACK`. All three are
  mutually exclusive and all three = 0 selects the White/Caucasian reference.
  Note in particular that RACE1 = 2 *excludes* Central/South Asian heritage
  (which falls in RACE1 = 4), so it is not `RACE_ASIAN`, and it *includes* South
  East Asian heritage, so it is not `RACE_ASIAN_NORTHEAST`.
* **Smoking reference category.** All Phase II/III subjects had at least 10
  pack-years at screening (Sect. 2.1), so the covariate is current-vs-former
  smoker and the canonical `SMOKE` indicator is used (reference = former smoker),
  not the three-level `SMOKE_CURRENT` / `SMOKE_NEVER` pair.
* **Sex coding.** The paper codes males = 1, females = 2 and applies
  `theta_COV x (covariate - 1)`, so `SEXF = source code - 1` and the coefficient
  applies exactly when `SEXF = 1`.
* **Covariates apply to COPD subjects only.** Every covariate row in Table 2 is
  labelled "COPD", and Sect. 3.2.2 introduces them "for subjects with COPD", so
  the model gates all of them on `DIS_COPD`.
* **Doses are nominal, not emitted.** Sect. 2.5 derives AUC0-24 as "ratio of
  nominal dose/individual post hoc estimate of CL/F", and the vilanterol
  reproduction above confirms nominal dosing (25 ug / 94.6 L/h = 264.3 vs Table
  4's 265.7 pg\*h/mL; the emitted dose of 22 ug would give 232.6).
* **Study effects are encoded as protocol-numbered indicators**
  (`STUDY_HZC110946`, `STUDY_HZC111348`) rather than the paper's "Study 3" /
  "Study 4" labels, using the mapping in Online Resource Table S1, so the
  covariate names are unambiguous outside the context of this paper.
* **The healthy-volunteer stratum has no covariate model.** Table 2 reports
  separate CL/F, V1/F and V2/F for healthy volunteers and no covariate effects
  for them, so the age/weight/sex/smoking/study terms are switched off when
  `DIS_COPD = 0`.
* **Q2/F, Q3/F, V3/F and D1 are shared across the COPD and healthy strata**,
  because Table 2 reports a single estimate for each.
