---
title: "Unbound ceftolozane and tazobactam during CVVHDF (Sime 2019b)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unbound ceftolozane and tazobactam during CVVHDF (Sime 2019b)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Sime et al. fitted the two analytes of the fixed 2:1 ceftolozane-tazobactam
combination as two separate nonparametric (Pmetrics NPAG) models sharing one
four-compartment structure, so the paper contributes **two** model files. The
`b` in the file names distinguishes this paper from the same group's 2019 study
of the same drug in patients *without* renal dysfunction
(`Sime_2019_ceftolozane` / `Sime_2019_tazobactam`, doi:10.1128/AAC.01265-19).

```{r models}
cef <- rxode2::rxode(readModelDb("Sime_2019b_ceftolozane"))
taz <- rxode2::rxode(readModelDb("Sime_2019b_tazobactam"))
```

* Citation: `r cef$reference`
* Article: <https://doi.org/10.1128/AAC.01655-19> (PMCID: PMC7187594, open access)
* Ceftolozane: `r cef$description`
* Tazobactam: `r taz$description`

The structure is the paper's Figure 1. The dose is infused into the prefilter
compartment (`central`, volume V_pre), which loses drug by a residual
non-CVVHDF clearance (`cl`) and by the CVVHDF clearance (`cl_crrt`) into an
effluent compartment (`circuit_dialysate`, volume V_effluent) that drains at
the first-order rate Kd. The prefilter compartment also exchanges with a
postfilter compartment (`postfilter`, V_post) through the rate constants K12 /
K21 and with a peripheral compartment (`peripheral1`, V_peripheral) through
Q_pre; the postfilter compartment exchanges with the same peripheral
compartment through Q_post. The three fitted outputs are the unbound prefilter
plasma concentration `Cc`, the unbound postfilter plasma concentration
`Cpostfilter`, and the effluent concentration `Ceffluent`.

**All concentrations are UNBOUND.** The assay isolated the unbound fraction by
ultracentrifugation before measurement, so no protein-binding conversion
appears in either file.

# Population

Six critically ill adults in the intensive care unit of the Royal Brisbane and
Women's Hospital, Australia, all on continuous venovenous hemodiafiltration
(CVVHDF) with a Prismaflex machine and an AN69 ST100 or ST150 filter (Tables 1
and 2). Five were male. Median age was 61.5 years (Q1-Q3 58-65; range 23-66),
median weight 72.5 kg (65-95), median serum creatinine 138 umol/L, median
albumin 26.5 g/L, median APACHE II 32 and median SOFA 11.5. Blood flow rates
were 100-200 mL/min and dialysate flow rates 1,000-1,500 mL/h. Every patient
received 1.5 g ceftolozane-tazobactam (1000 mg ceftolozane + 500 mg
tazobactam) every 8 h as a 1-h infusion, and was sampled over one dosing
interval in prefilter plasma, postfilter plasma and effluent.

# Source trace

| Quantity | Ceftolozane | Tazobactam | Source |
|---|---|---|---|
| Structure (4 compartments, clearances, transfers) | -- | -- | Figure 1 and its legend |
| CL_CVVHDF (L/h), `lcl_crrt` | 2.659 (SD 0.783) | 2.973 (SD 0.603) | Table 3 |
| CL_residual (L/h), `lcl` | 0.596 (SD 0.504) | 3.254 (SD 0.867) | Table 3 |
| V_pre (L), `lvc` | 25.184 (SD 7.499) | 28.206 (SD 6.603) | Table 3 |
| V_post (L), `lvpostfilter` | 17.578 (SD 10.871) | 19.685 (SD 14.382) | Table 3 |
| K12 (1/h), `lk12` | 0.43 (SD 0.718) | 0.561 (SD 0.638) | Table 3 |
| K21 (1/h), `lk21` | 0.676 (SD 0.908) | 1.01 (SD 1.01) | Table 3 |
| Kd (1/h), `lkd` | 1.596 (SD 0.495) | 1.584 (SD 0.483) | Table 3 |
| V_effluent (L), `lvdialysate` | 2.178 (SD 0.801) | 2.609 (SD 1.561) | Table 3 |
| Q_pre (L/h), `lq` | 0.834 (SD 1.863) | 2.547 (SD 2.455) | Table 3 |
| Q_post (L/h), `lqpostfilter` | 2.42 (SD 1.451) | 3.612 (SD 1.204) | Table 3 |
| V_peripheral (L), `lvp` | 73.379 (SD 39.042) | 77.196 (SD 32.267) | Table 3 |
| IIV, `eta*` | log((SD/mean)^2 + 1) | same | Table 3 SD column |
| Residual error | fixed(0) | fixed(0) | not reported (Methods) |
| Dose split 2:1 | 1000 mg per 1.5 g | 500 mg per 1.5 g | Methods, 'Ceftolozane-tazobactam dosing' |

The IIV variances are derived from the transcribed values. This check is pure
arithmetic on the model files, so it is asserted tightly.

```{r cv-check}
tab3 <- data.frame(
  eta = c(
    "etalcl_crrt", "etalcl", "etalvc", "etalvpostfilter", "etalk12", "etalk21",
    "etalkd", "etalvdialysate", "etalq", "etalqpostfilter", "etalvp"
  ),
  cef_mean = c(2.659, 0.596, 25.184, 17.578, 0.43, 0.676, 1.596, 2.178, 0.834, 2.42, 73.379),
  cef_sd = c(0.783, 0.504, 7.499, 10.871, 0.718, 0.908, 0.495, 0.801, 1.863, 1.451, 39.042),
  taz_mean = c(2.973, 3.254, 28.206, 19.685, 0.561, 1.01, 1.584, 2.609, 2.547, 3.612, 77.196),
  taz_sd = c(0.603, 0.867, 6.603, 14.382, 0.638, 1.01, 0.483, 1.561, 2.455, 1.204, 32.267)
)
om_cef <- diag(cef$omega)[tab3$eta]
om_taz <- diag(taz$omega)[tab3$eta]
stopifnot(
  all(abs(om_cef - log((tab3$cef_sd / tab3$cef_mean)^2 + 1)) < 1e-4),
  all(abs(om_taz - log((tab3$taz_sd / tab3$taz_mean)^2 + 1)) < 1e-4)
)
tab3 |>
  mutate(cef_cv = round(100 * cef_sd / cef_mean), taz_cv = round(100 * taz_sd / taz_mean)) |>
  select(eta, cef_cv, taz_cv) |>
  rename("Random effect" = eta, "Ceftolozane CV (%)" = cef_cv, "Tazobactam CV (%)" = taz_cv) |>
  knitr::kable()
```

# Virtual cohorts

The paper's simulations drew 1,000 virtual patients from the NPAG support
points. Here each regimen gets 150 virtual patients drawn from the independent
log-normal marginals encoded in the model files. There are no covariates to
sample. The regimens are a subset of Tables 4, 5 and S1; the combination dose
is split 2:1 between the two analytes.

```{r regimens}
n_per_arm <- 150
regimens <- tibble::tribble(
  ~regimen, ~combo_g, ~type, ~dur,
  "0.375 g q8h", 0.375, "intermittent", 1,
  "0.75 g q8h", 0.75, "intermittent", 1,
  "1.5 g q8h", 1.5, "intermittent", 1,
  "3.0 g q8h", 3.0, "intermittent", 1,
  "1.5 g 4 h EI q8h", 1.5, "intermittent", 4,
  "1.5 g LD + 4.5 g CI", 1.5, "continuous", 1
)
n_days <- 10
tau <- 8

# Dose rows for one regimen, in mg of one analyte. The continuous-infusion
# arm is a 1-h loading dose followed, from the end of the loading dose, by the
# daily dose (3 x the q8h dose) infused over each 24 h.
dose_rows <- function(reg, frac, ids) {
  amt1 <- reg$combo_g * 1000 * frac
  if (reg$type == "intermittent") {
    t_dose <- seq(0, n_days * 24 - tau, by = tau)
    d <- expand.grid(id = ids, time = t_dose)
    d$amt <- amt1
    d$rate <- amt1 / reg$dur
  } else {
    ld <- data.frame(id = ids, time = 0, amt = amt1, rate = amt1)
    t_ci <- 1 + seq(0, (n_days - 1) * 24, by = 24)
    ci <- expand.grid(id = ids, time = t_ci)
    ci$amt <- 3 * amt1
    ci$rate <- 3 * amt1 / 24
    d <- rbind(ld, ci)
  }
  d$evid <- 1L
  d$cmt <- "central"
  d$dvid <- NA_integer_
  d
}

obs_times <- sort(unique(c(seq(0, 24, by = 0.1), 96, seq(216, 224, by = 0.1))))

build_events <- function(frac) {
  out <- lapply(seq_len(nrow(regimens)), function(i) {
    ids <- (i - 1) * n_per_arm + seq_len(n_per_arm)
    obs <- expand.grid(id = ids, time = obs_times)
    obs$amt <- NA_real_
    obs$rate <- NA_real_
    obs$evid <- 0L
    # Three endpoints are declared, so observation rows are keyed by dvid
    # with no cmt; every output (Cc, Cpostfilter, Ceffluent) is returned on
    # each row.
    obs$cmt <- NA_character_
    obs$dvid <- 1L
    ev <- rbind(dose_rows(regimens[i, ], frac, ids), obs)
    ev$regimen <- regimens$regimen[i]
    ev
  })
  ev <- do.call(rbind, out)
  ev[order(ev$id, ev$time, -ev$evid), ]
}
ev_cef <- build_events(2 / 3)
ev_taz <- build_events(1 / 3)
```

# Simulation

```{r simulate}
# rxSetSeed() fixes the draw within one rxode2 build and thread count only;
# every assertion below is written on medians and robust proportions so that
# it holds for any cohort the model can produce.
rxode2::rxSetSeed(20191220)
# sigma = NA: residual error is fixed(0) and is not simulated.
sim_cef <- rxode2::rxSolve(cef, ev_cef, sigma = NA, keep = "regimen", returnType = "data.frame")
sim_taz <- rxode2::rxSolve(taz, ev_taz, sigma = NA, keep = "regimen", returnType = "data.frame")
sim <- bind_rows(
  mutate(sim_cef, drug = "ceftolozane"),
  mutate(sim_taz, drug = "tazobactam")
)
# Solver sanity: all four states survive and nothing goes negative.
stopifnot(
  all(sim$Cc >= -1e-8), all(sim$Cpostfilter >= -1e-8), all(sim$Ceffluent >= -1e-8),
  max(sim$peripheral1) > 0, max(sim$postfilter) > 0
)
```

## Typical-value profiles in the three fitted matrices

Replicates the matrices of Figures 2 and 3 (prefilter, postfilter and effluent)
as typical-value curves over the first 24 h of 1.5 g q8h. The paper's own
figures are observed-versus-predicted scatter plots, which need the
individual data.

```{r typical}
typ_events <- function(frac) {
  d <- dose_rows(regimens[regimens$regimen == "1.5 g q8h", ], frac, 1L)
  o <- data.frame(
    id = 1L, time = obs_times, amt = NA_real_, rate = NA_real_, evid = 0L,
    cmt = NA_character_, dvid = 1L
  )
  r <- rbind(d, o)
  r[order(r$time, -r$evid), ]
}
typ <- bind_rows(
  mutate(rxode2::rxSolve(cef, typ_events(2 / 3), omega = NA, sigma = NA, returnType = "data.frame"), drug = "ceftolozane"),
  mutate(rxode2::rxSolve(taz, typ_events(1 / 3), omega = NA, sigma = NA, returnType = "data.frame"), drug = "tazobactam")
)
typ |>
  filter(time <= 24) |>
  select(time, drug, Prefilter = Cc, Postfilter = Cpostfilter, Effluent = Ceffluent) |>
  pivot_longer(c(Prefilter, Postfilter, Effluent), names_to = "matrix", values_to = "conc") |>
  ggplot(aes(time, conc, colour = matrix)) +
  geom_line() +
  facet_wrap(~drug, scales = "free_y") +
  labs(x = "Time (h)", y = "Unbound concentration (mg/L)", colour = NULL) +
  theme_bw()
```

## Prefilter percentiles over the first dosing interval

Supplemental Figure S1 is a visual predictive check of prefilter plasma during
the sampled dosing interval. The sampled interval's position in each patient's
course is not stated, so the check is not replicated quantitatively; the
first-interval percentiles below show the same 5th-95th percentile layout.

```{r vpc}
sim |>
  filter(regimen == "1.5 g q8h", time <= 8) |>
  group_by(drug, time) |>
  summarise(
    p05 = quantile(Cc, 0.05), p25 = quantile(Cc, 0.25), p50 = median(Cc),
    p75 = quantile(Cc, 0.75), p95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey85") +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "grey65") +
  geom_line(aes(y = p50)) +
  facet_wrap(~drug, scales = "free_y") +
  labs(x = "Time after first dose (h)", y = "Unbound prefilter concentration (mg/L)") +
  theme_bw()
```

# PKNCA validation

Steady-state NCA over the last dosing interval (216-224 h) of every regimen,
for each analyte, compared against the medians of Table 5. For intermittent
regimens Cmax is the end-of-infusion concentration of Table 5 and Cmin its
trough; for the continuous infusion Table 5 reports only the steady-state
concentration, compared against Cmin.

```{r pknca}
run_nca <- function(s, ev) {
  conc <- s |>
    filter(!is.na(Cc)) |>
    select(id, time, Cc, regimen)
  dose <- ev |>
    filter(evid == 1) |>
    select(id, time, amt, regimen)
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | regimen + id, concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | regimen + id, doseu = "mg")
  intervals <- data.frame(start = 216, end = 224, cmax = TRUE, cmin = TRUE, auclast = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
nca_cef <- run_nca(sim_cef, ev_cef)
nca_taz <- run_nca(sim_taz, ev_taz)

# Table 5, median steady-state concentrations (mg/L).
ref_cef <- data.frame(
  regimen = regimens$regimen,
  cmax = c(13, 27, 54, 107, 44, NA),
  cmin = c(7, 14, 28, 56, 31, 36)
)
ref_taz <- data.frame(
  regimen = regimens$regimen,
  cmax = c(4.4, 8.8, 17.5, 35.0, 12.3, NA),
  cmin = c(1.5, 3.0, 6.1, 12.1, 7.0, 9.7)
)
cmp_cef <- ncaComparisonTable(nca_cef, ref_cef, by = "regimen", params = c("cmax", "cmin"), units = c(cmax = "mg/L", cmin = "mg/L"))
cmp_taz <- ncaComparisonTable(nca_taz, ref_taz, by = "regimen", params = c("cmax", "cmin"), units = c(cmax = "mg/L", cmin = "mg/L"))
```

```{r pknca-cef}
knitr::kable(cmp_cef, caption = "Ceftolozane, steady state: simulated median vs Table 5 median.")
```

```{r pknca-taz}
knitr::kable(cmp_taz, caption = "Tazobactam, steady state: simulated median vs Table 5 median.")
```

```{r pknca-gate}
# Recompute the percentage differences numerically for the gate.
pct <- function(nca, ref) {
  as.data.frame(nca$result) |>
    filter(PPTESTCD %in% c("cmax", "cmin")) |>
    group_by(regimen, PPTESTCD) |>
    summarise(sim = median(PPORRES), .groups = "drop") |>
    inner_join(pivot_longer(ref, c(cmax, cmin), names_to = "PPTESTCD", values_to = "ref"), by = c("regimen", "PPTESTCD")) |>
    filter(!is.na(ref)) |>
    mutate(pct_diff = 100 * (sim - ref) / ref)
}
chk <- bind_rows(pct(nca_cef, ref_cef), pct(nca_taz, ref_taz))
stopifnot(
  # Structural: a wrong clearance, dose split or unit moves every row by tens
  # of percent.
  abs(median(chk$pct_diff)) < 10,
  # Envelope on per-regimen MEDIANS (not extremes); Table 5 rounds the small
  # tazobactam values to one or two significant figures.
  quantile(abs(chk$pct_diff), 0.9) < 25
)
```

## Closed-form checks

The system is linear, so for the typical patient the steady-state
AUC over one interval must equal dose / (CL_residual + CL_CVVHDF), and the
drug removed into the effluent over that interval must be the fraction
CL_CVVHDF / (CL_residual + CL_CVVHDF) of the dose. Both are deterministic
(no cohort draw), so the bounds are tight.

```{r closed-form}
# The slowly equilibrating peripheral compartment leaves the typical patient
# about 0.7% short of true steady state at day 10, so the identity is checked
# on the last interval of a 40-day regimen on a 0.02 h grid.
long_typ <- function(mod, amt) {
  d <- data.frame(
    id = 1L, time = seq(0, 40 * 24 - tau, by = tau), amt = amt, rate = amt,
    evid = 1L, cmt = "central", dvid = NA_integer_
  )
  o <- data.frame(
    id = 1L, time = seq(40 * 24 - tau, 40 * 24, by = 0.02), amt = NA_real_,
    rate = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L
  )
  e <- rbind(d, o)
  rxode2::rxSolve(mod, e[order(e$time, -e$evid), ], omega = NA, sigma = NA, returnType = "data.frame")
}
cf <- bind_rows(
  mutate(long_typ(cef, 1000), drug = "ceftolozane"),
  mutate(long_typ(taz, 500), drug = "tazobactam")
) |>
  group_by(drug) |>
  summarise(
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  mutate(
    dose = c(1000, 500),
    cl_tot = c(0.596 + 2.659, 3.254 + 2.973),
    auc_expected = dose / cl_tot,
    pct_diff = 100 * (auc - auc_expected) / auc_expected,
    frac_cvvhdf = c(2.659, 2.973) / cl_tot
  )
knitr::kable(cf, digits = 3)
stopifnot(all(abs(cf$pct_diff) < 0.5))
```

The paper gives total ceftolozane clearance during CVVHDF as 3.3 L/h
(Discussion), i.e. 0.596 + 2.659 = 3.255 L/h. From Table 3, about 82% of
ceftolozane clearance but only 48% of tazobactam clearance is extracorporeal
(the `frac_cvvhdf` column).

The Discussion also states that "steady state was achieved only after 4 days
for ceftolozane". Below, the typical-patient pre-dose trough on each of days
1-6 of 1.5 g q8h is expressed as a fraction of the day-40 trough (true steady
state). Doubling the dose must double the concentration exactly (Results,
"doubling the dose resulted in doubling the steady-state concentration").

```{r steady-state}
daily_trough <- function(mod, amt) {
  d <- data.frame(
    id = 1L, time = seq(0, 40 * 24 - tau, by = tau), amt = amt, rate = amt,
    evid = 1L, cmt = "central", dvid = NA_integer_
  )
  o <- data.frame(
    id = 1L, time = seq(24, 40 * 24, by = 24), amt = NA_real_,
    rate = NA_real_, evid = 0L, cmt = NA_character_, dvid = 1L
  )
  e <- rbind(d, o)
  s <- rxode2::rxSolve(mod, e[order(e$time, -e$evid), ], omega = NA, sigma = NA, returnType = "data.frame")
  s$Cc / s$Cc[s$time == 40 * 24]
}
ss_frac <- data.frame(
  day = 1:6,
  ceftolozane = daily_trough(cef, 1000)[1:6],
  tazobactam = daily_trough(taz, 500)[1:6]
)
knitr::kable(ss_frac, digits = 3, caption = "Typical pre-dose trough as a fraction of the steady-state trough.")
# Deterministic (typical values, no cohort draw). Ceftolozane is well short of
# steady state at day 2 and within ~15% of it by day 4; tazobactam, with twice
# the clearance, gets there sooner.
stopifnot(
  ss_frac$ceftolozane[2] < 0.8,
  ss_frac$ceftolozane[4] > 0.85,
  all(ss_frac$tazobactam > ss_frac$ceftolozane)
)

trough <- function(d, t) d$Cc[which.min(abs(d$time - t))]
typ_double <- rxode2::rxSolve(cef, transform(typ_events(2 / 3), amt = 2 * amt, rate = 2 * rate), omega = NA, sigma = NA, returnType = "data.frame")
stopifnot(abs(trough(typ_double, 224) / trough(filter(typ, drug == "ceftolozane"), 224) - 2) < 1e-6)
```

# Probability of target attainment

%fT above a threshold is computed from the prefilter concentration (Methods:
"Prefilter patient plasma exposure was used for all PTA assessments") over the
first 24 h and over the steady-state interval 216-224 h.

```{r pta}
ft_above <- function(d, thr, t0, t1) {
  x <- d[d$time >= t0 & d$time <= t1, ]
  mean(x$Cc > thr)
}
pta <- function(drug_name, thr, target_pct, t0, t1) {
  sim |>
    filter(drug == drug_name) |>
    group_by(regimen, id) |>
    group_modify(~ data.frame(ft = ft_above(.x, thr, t0, t1))) |>
    group_by(regimen) |>
    summarise(pta = mean(100 * ft >= target_pct), .groups = "drop")
}
pta_tab <- bind_rows(
  mutate(pta("ceftolozane", 4, 40, 0, 24), target = "Ceftolozane 40% fT>4 mg/L, first 24 h"),
  mutate(pta("tazobactam", 1, 20, 0, 24), target = "Tazobactam 20% fT>1 mg/L, first 24 h"),
  mutate(pta("tazobactam", 2, 50, 0, 24), target = "Tazobactam 50% fT>2 mg/L, first 24 h"),
  mutate(pta("tazobactam", 4, 100, 216, 224), target = "Tazobactam 100% fT>4 mg/L, steady state")
)
pta_tab |>
  pivot_wider(names_from = regimen, values_from = pta) |>
  knitr::kable(digits = 2)
```

```{r pta-gate}
get_pta <- function(tgt, reg) pta_tab$pta[pta_tab$target == tgt & pta_tab$regimen == reg]
stopifnot(
  # Results: PTA >= 0.9 for 40% fT>MIC at MIC 4 mg/L in the first 24 h. The
  # floor sits below 0.9 to absorb the cohort draw; 0.375 g q8h is excluded
  # (see the deviation noted below the table).
  all(sapply(setdiff(regimens$regimen, "0.375 g q8h"), function(r) get_pta("Ceftolozane 40% fT>4 mg/L, first 24 h", r)) > 0.8),
  # Table S1: 20% fT>1 mg/L was 1.0 for every regimen.
  all(sapply(regimens$regimen, function(r) get_pta("Tazobactam 20% fT>1 mg/L, first 24 h", r)) > 0.9),
  # Table S1: 50% fT>2 mg/L in the first 24 h 0.1 at 0.375 g q8h, 1.0 at
  # 0.75 g q8h and above.
  get_pta("Tazobactam 50% fT>2 mg/L, first 24 h", "0.375 g q8h") < 0.5,
  get_pta("Tazobactam 50% fT>2 mg/L, first 24 h", "1.5 g q8h") > 0.9,
  # Table S1: 100% fT>4 mg/L at steady state 0.0 at 0.75 g q8h, 1.0 at 1.5 g q8h.
  get_pta("Tazobactam 100% fT>4 mg/L, steady state", "0.75 g q8h") < 0.4,
  get_pta("Tazobactam 100% fT>4 mg/L, steady state", "1.5 g q8h") > 0.6
)
```

Two published statements are not reproduced:

* **0.375 g q8h, ceftolozane 40% fT>MIC.** The Results state that all
  simulated regimens reached PTA >= 0.9 for 40% fT>MIC at MIC <= 4 mg/L in the
  first 24 h; here 0.375 g q8h falls short (see the table). The lowest dose is
  also the one with the lowest cumulative fractional response in Table 4
  (0.82), and the gap is consistent with the log-normal approximation to the
  NPAG distribution (see Assumptions and deviations).
* **100% fT above a threshold in the first 24 h** (Table 4 for ceftolozane,
  Table S1 for tazobactam) is not computed. Over a window that starts at the
  first dose the concentration begins at zero, so 100% cannot be reached as
  defined here; the paper does not state the time grid over which its
  first-24 h percentages were evaluated.

# Assumptions and deviations

* **Nonparametric to parametric.** Table 3 summarises each NPAG marginal as a
  mean and SD. Each mean is encoded as the median of a log-normal
  distribution with CV = SD / mean, following the group's sibling models
  (`Sime_2019_ceftolozane`, `Sime_2019_tazobactam`) and other Pmetrics models
  in the library. Several CVs exceed 100% (ceftolozane K12 167%, K21 134%,
  Q_pre 223%; tazobactam K12 114%, K21 100%), so the simulated population
  means of those parameters exceed the tabulated means. The simulated
  per-regimen medians nonetheless reproduce Table 5, and exposure (AUC,
  average concentration) depends only on the two clearances, whose CVs are
  20-85%.
* **Parameter correlations** are not recoverable from the publication, so all
  random effects are independent. The simulated interquartile ranges are
  narrower than those of Table 5 (e.g. ceftolozane 1.5 g q8h trough Q3 42
  mg/L in the paper), which a discrete, correlated NPAG distribution can
  produce and an independent log-normal approximation does not.
* **Residual error** is not reported: the Methods list the Pmetrics error
  models tested but neither the selected form nor its coefficients. All six
  residual parameters are `fixed(0)`.
* **Effluent compartment name.** The effluent compartment gains drug by the
  CVVHDF clearance and drains first-order, which is the role of the
  library's `circuit_dialysate` state; its fitted volume (about 2 L) is an
  empirical quantity, not the physical dialysate-side volume of the filter.
  The postfilter compartment is likewise an empirical compartment and is
  carried under the paper-specific name `postfilter`.
* **CVVHDF settings are fixed into the parameters.** No CVVHDF setting was
  retained as a covariate, so the model applies only to settings similar to
  Table 2 (blood flow 100-200 mL/min, dialysate 1,000-1,500 mL/h).
* **Continuous-infusion timing.** The continuous infusion is assumed to start
  at the end of the 1-h loading dose.
* **Supplemental Figure S1** (prefilter VPC) is shown only in layout, because
  the position of the sampled dosing interval in each patient's course is not
  stated.
