---
title: "Unbound ceftolozane and tazobactam in critical illness (Sime 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unbound ceftolozane and tazobactam in critical illness (Sime 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Sime 2019 fitted the two analytes of the fixed 2:1 ceftolozane-tazobactam
combination in two separate nonparametric (Pmetrics NPAG) runs, with different
body-weight exponents on central volume, so the paper contributes **two** model
files to this library.

```{r models}
cef <- rxode2::rxode(readModelDb("Sime_2019_ceftolozane"))
taz <- rxode2::rxode(readModelDb("Sime_2019_tazobactam"))
```

* Citation: `r cef$reference`
* Article: <https://doi.org/10.1128/AAC.01265-19> (PMCID: PMC6761554, open access)
* Ceftolozane: `r cef$description`
* Tazobactam: `r taz$description`

Both models are two-compartment with a 1-hour intravenous infusion into
`central`, clearance additive in a covariate-free intercept plus an arm linear
in measured urinary creatinine clearance, and distribution parameterised
directly as the rate constants Kcp and Kpc.

**`Cc` is an UNBOUND plasma concentration in both models.** The study assayed
the unbound fraction directly, after isolating it by ultracentrifugation, so no
protein-binding conversion appears anywhere in either file.

# Population

Twelve critically ill adults in the intensive care unit of the Royal Brisbane
and Women's Hospital, Australia, contributed 133 unbound concentration-time data
points (Table 1 and Results). The cohort was selected to have **no** renal
dysfunction: patients needing renal replacement therapy were excluded, measured
urinary creatinine clearance had a median of 107 mL/min/1.73 m^2 (IQR 74-145),
and serum creatinine had a median of 46 umol/L (IQR 39-77). Median age was 56
years (IQR 52-61), median weight 79.5 kg (IQR 64-99), median BMI 28.5 kg/m^2
(IQR 22.1-32.9), and 8 of 12 (67%) were female. Illness severity was high
(APACHE II median 19.5, SOFA median 6) and the cohort was uniformly
hypoalbuminaemic (albumin median 25 g/L). All 12 had a positive culture; lung
was the commonest source of infection. Patients received 1.5 g or 3.0 g
ceftolozane-tazobactam every 8 h as a 1-hour infusion at the treating
physician's discretion.

The same information is available programmatically from either model's
`population` metadata:

```{r population}
str(cef$population[c("n_subjects", "n_concentrations", "renal_function", "dose_range")])
```

# Source trace

Every `ini()` entry in both model files carries an in-file comment pointing at
its source location. They are collected here for review. Table 2 of Sime 2019
reports, for each parameter, the **mean, SD and CV of the NPAG support-point
distribution** -- the SD is between-subject spread, not an estimation standard
error.

| Parameter | Ceftolozane | Tazobactam | Source location |
|---|---|---|---|
| `lcl_nonren` (intercept) | 0.86 L/h (SD 0.69, CV 80%) | 6.9 L/h (SD 5.6, CV 81%) | Table 2, "Intercept" row |
| `lcl_renal` (slope) | 6.0 L/h (SD 3.3, CV 54%) | 17.5 L/h (SD 6.9, CV 40%) | Table 2, "Slope" row |
| `lvc` | 20.4 L (SD 3.7, CV 18%) | 32.4 L (SD 10, CV 31%) | Table 2, "V (liters)" row; also Abstract |
| `lk12` (Kcp) | 0.46 /h (SD 0.74, CV 159%) | 2.96 /h (SD 8.69, CV 293%) | Table 2, "Kcp (h-1)" row |
| `lk21` (Kpc) | 0.39 /h (SD 0.37, CV 94%) | 26.5 /h (SD 8.4, CV 32%) | Table 2, "Kpc (h-1)" row |
| derived CL for the study population | 7.2 L/h (SD 3.2, CV 45%) | 25.4 L/h (SD 9.4, CV 37%) | Table 2, "CL (liters/h)" row, footnote "Value calculated for the study population" |
| `e_wt_vc` | `fixed(1)` | `fixed(0.75)` | Results: "V 1 = V * WT/80" and "V 1 = V * [WT/80] 0.75" |
| CL covariate equation | `cl_nonren + cl_renal * CRCL/100` | same form | Results: "CL = intercept + slope * CL CRurinary" (the /100 is derived; see below) |
| `d/dt(central)`, `d/dt(peripheral1)` | two-compartment, linear | same | Methods, "Population PK modeling": one- and two-compartment base models with first-order elimination; Results select two compartments (LLR 723 vs 795) |
| `propSd`, `addSd` | `fixed(0)` | `fixed(0)` | NOT REPORTED -- see Errata |
| IIV variances | `log(CV^2 + 1)` per row above | same | Table 2, "CV (%)" column |

## The CV column is a plain SD/mean coefficient of variation

This is what licenses `omega^2 = log(CV^2 + 1)` rather than reading the column
as an omega. It is checkable arithmetic on the transcribed table:

```{r cv-check}
tab2 <- tibble::tribble(
  ~drug,         ~parameter,  ~mean, ~sd,  ~cv_printed,
  "ceftolozane", "Intercept",  0.86, 0.69,  80,
  "ceftolozane", "Slope",      6.00, 3.30,  54,
  "ceftolozane", "V",         20.40, 3.70,  18,
  "ceftolozane", "Kcp",        0.46, 0.74, 159,
  "ceftolozane", "Kpc",        0.39, 0.37,  94,
  "ceftolozane", "CL",         7.20, 3.20,  45,
  "tazobactam",  "Intercept",  6.90, 5.60,  81,
  "tazobactam",  "Slope",     17.50, 6.90,  40,
  "tazobactam",  "V",         32.40, 10.0,  31,
  "tazobactam",  "Kcp",        2.96, 8.69, 293,
  "tazobactam",  "Kpc",       26.50, 8.40,  32,
  "tazobactam",  "CL",        25.40, 9.40,  37
) |>
  mutate(cv_recomputed = 100 * sd / mean,
         rel_diff_pct = 100 * abs(cv_recomputed - cv_printed) / cv_printed)

tab2 |>
  select(drug, parameter, mean, sd, cv_printed, cv_recomputed, rel_diff_pct) |>
  rename(
    "Drug" = drug, "Parameter" = parameter, "Mean" = mean, "SD" = sd,
    "CV printed (%)" = cv_printed, "100 * SD / mean (%)" = cv_recomputed,
    "Relative difference (%)" = rel_diff_pct
  ) |>
  knitr::kable(digits = 1, caption = "Every row of Sime 2019 Table 2 reproduces its own printed CV as SD/mean.")
```

```{r cv-check-assert}
# Deterministic arithmetic on transcribed values -- no simulation, so this can
# be asserted tightly. The comparison is RELATIVE because the CVs span 18% to
# 293%, and the residual is pure two-significant-figure rounding in the mean and
# SD columns: the worst row, ceftolozane Kcp, gives 0.74/0.46 = 160.9% against a
# printed 159, and the rounding interval 0.735-0.745 over 0.455-0.465 spans
# 158.1-163.7%, which contains 159. 3% admits that and nothing else -- any other
# reading of the column (an omega, an SD on the log scale, a variance) misses by
# far more.
stopifnot(max(tab2$rel_diff_pct) < 3)
```

## The creatinine-clearance slope is normalised by 100, and that is derived

Sime 2019 prints the covariate model as `CL = intercept + slope * CL CRurinary`
but never states the scale on which the clearance enters. Taken literally in
mL/min/1.73 m^2 the slope is absurd (6.0 * 107 = 642 L/h). The divisor is
recovered from the paper's own numbers:

```{r crcl-divisor}
x_cef <- (7.2 - 0.86) / 6.0
x_taz <- (25.4 - 6.9) / 17.5
c(ceftolozane = x_cef, tazobactam = x_taz)
```

Table 2's own derived clearance requires the covariate to enter at about 1.057
for **both** analytes, which were separate NPAG fits. Against a cohort median
urinary creatinine clearance of 107 (Table 1) / 108 (Results) that pins the
divisor at 100. Propagating the tabulated SDs through the same equation
corroborates it, and the paper's simulated steady-state concentrations
(next section) provide an independent check that also excludes the two other
candidate divisors.

```{r sd-propagation}
c(
  ceftolozane_sd_predicted = sqrt(0.69^2 + (3.3 * x_cef)^2), # Table 2 prints 3.2
  tazobactam_sd_predicted  = sqrt(5.6^2 + (6.9 * x_taz)^2)   # Table 2 prints 9.4
)
```

# Deterministic replication of the paper's published concentrations

The Discussion prints four simulated steady-state unbound ceftolozane
concentrations from the continuous-infusion regimens, at two stated creatinine
clearances and the cohort's 80 kg reference weight. These are the paper's only
published concentration values, and they are the strongest available check on
the transcription: a mis-scaled clearance, a wrong dose split or a wrong
normalisation moves them by tens of percent.

Note the dose bookkeeping: a "4.5 g continuous infusion" is 4.5 g of the 2:1
**combination**, i.e. 3 g of ceftolozane over 24 h.

```{r typical-css}
# Typical-value (zeroRe) solve: no between-subject variability, matching the way
# the paper computes a point prediction from its tabulated means.
cef_typ <- rxode2::zeroRe(cef)

css_typical <- function(ld_mg, ci_mg_per_day, crcl, wt = 80) {
  dose <- data.frame(
    id = 1L,
    time = c(0, 1, 25, 49),
    amt = c(ld_mg, rep(ci_mg_per_day, 3)),
    rate = c(ld_mg / 1, rep(ci_mg_per_day / 24, 3)),
    evid = 1L,
    cmt = "central"
  )
  obs <- data.frame(
    id = 1L, time = seq(48, 72, by = 0.1), amt = NA_real_, rate = NA_real_,
    evid = 0L, cmt = "central"
  )
  ev <- dplyr::arrange(dplyr::bind_rows(dose, obs), time, dplyr::desc(evid))
  ev$CRCL <- crcl
  ev$WT <- wt
  # rxSolve returns OBSERVATION records only (addDosing defaults to FALSE) and
  # does not carry an evid column, so the returned rows are exactly the grid.
  out <- rxode2::rxSolve(cef_typ, ev, returnType = "data.frame")
  stopifnot(nrow(out) == sum(ev$evid == 0))
  mean(out$Cc)
}

css_cmp <- tibble::tribble(
  ~regimen,                  ~ld_mg, ~ci_mg, ~CRCL, ~published,
  "1.5 g LD + 4.5 g CI",       1000,   3000,   100,       19.0,
  "1.5 g LD + 4.5 g CI",       1000,   3000,   180,       11.2,
  "3 g LD + 9 g CI",           2000,   6000,   100,       38.0,
  "3 g LD + 9 g CI",           2000,   6000,   180,       22.4
) |>
  rowwise() |>
  mutate(simulated = css_typical(ld_mg, ci_mg, CRCL)) |>
  ungroup() |>
  mutate(
    cl_lh = 0.86 + 6.0 * CRCL / 100,
    pct_diff = 100 * (simulated - published) / published
  )

css_cmp |>
  select(regimen, CRCL, cl_lh, published, simulated, pct_diff) |>
  rename(
    "Ceftolozane-tazobactam regimen" = regimen,
    "CRCL (mL/min/1.73 m^2)" = CRCL,
    "Typical CL (L/h)" = cl_lh,
    "Published mean Css (mg/L)" = published,
    "Model typical Css (mg/L)" = simulated,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Sime 2019 Discussion: simulated mean unbound ceftolozane concentrations at steady state (48-72 h), against a typical-value solve of the packaged model.")
```

```{r typical-css-assert}
# All four published values are reproduced, and all four sit BELOW the paper's
# figure by a nearly constant 4.4-4.7%. That is the signature of Jensen's
# inequality: the published numbers are MEANS of a 1,000-subject Monte Carlo of
# Rate/CL, and E[Rate/CL] > Rate/E[CL] strictly. A typical-value solve must
# therefore fall short, and the sign is itself evidence the divisor is right:
# divisors of 107 or 108 predict 19.3-19.5 and 11.4-11.5, ABOVE the published
# means, which Jensen forbids.
#
# Deterministic quantity (zeroRe, fixed covariates) -- no cohort draw is
# involved, so the bound is tight. 8% still goes red on any mis-scaled
# clearance, dose split or unit.
stopifnot(all(abs(css_cmp$pct_diff) < 8))
stopifnot(all(css_cmp$pct_diff < 0)) # Jensen: typical value must undershoot a simulated mean
```

The tabulated clearance for the study population reproduces the same way:

```{r cl-population}
cl_cmp <- tibble::tribble(
  ~drug,         ~intercept, ~slope, ~published_cl,
  "ceftolozane",       0.86,    6.0,          7.2,
  "tazobactam",        6.90,   17.5,         25.4
) |>
  # Table 1 cohort median urinary creatinine clearance
  mutate(
    model_cl = intercept + slope * 107 / 100,
    pct_diff = 100 * (model_cl - published_cl) / published_cl
  )

cl_cmp |>
  rename(
    "Drug" = drug, "Intercept (L/h)" = intercept, "Slope (L/h)" = slope,
    "Table 2 CL (L/h)" = published_cl, "Model CL at CRCL 107 (L/h)" = model_cl,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Table 2's derived clearance, recomputed from its own intercept and slope at the cohort median urinary creatinine clearance.")
```

```{r cl-population-assert}
stopifnot(all(abs(cl_cmp$pct_diff) < 2))
```

# Virtual cohorts

Original observed data are not publicly available. Two cohorts are used below.

The **target-attainment cohort** fixes the covariates exactly where the paper
fixed them for its Monte Carlo analysis -- Results: "considering the median
urinary creatinine clearance (108 ml/min/1.73 m 2) and body weight (80 kg) of
the study population" -- so that only the NPAG parameter variability drives the
spread. The **VPC cohort** instead samples covariates from the Table 1
distributions.

The paper simulated 1,000 subjects per regimen; 100 per arm are used here, which
is ample for an illustrative replication and keeps the vignette inside its
render budget.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD -- so the cohorts below are
# reproducible on this machine and different on a machine with a different
# thread count. Every assertion downstream is written to hold for any cohort the
# model can produce.
set.seed(20190923)

n_per_arm <- 100L

# One row per dose. A "1.5 g" regimen is 1.5 g of the 2:1 COMBINATION, i.e.
# 1000 mg ceftolozane + 500 mg tazobactam.
regimens <- tibble::tribble(
  ~regimen,               ~kind,   ~cef_mg, ~taz_mg, ~infusion_h, ~tau_h,
  "1.5 g q8h (1 h)",      "int",      1000,     500,           1,      8,
  "3 g q8h (1 h)",        "int",      2000,    1000,           1,      8,
  "1.5 g LD + 4.5 g CI",  "ci",       1000,     500,           1,     NA,
  "3 g LD + 9 g CI",      "ci",       2000,    1000,           1,     NA
)

# Dose rows for one subject on one regimen, in mg of the named analyte.
dose_rows <- function(reg, amt_mg) {
  if (reg$kind == "int") {
    tt <- seq(0, 64, by = reg$tau_h)
    data.frame(time = tt, amt = amt_mg, rate = amt_mg / reg$infusion_h)
  } else {
    # Loading dose over 1 h, then three consecutive 24-h continuous infusions
    # delivering 3x the loading dose per day (4.5 g CI after a 1.5 g LD).
    ci_mg <- 3 * amt_mg
    data.frame(
      time = c(0, 1, 25, 49),
      amt = c(amt_mg, rep(ci_mg, 3)),
      rate = c(amt_mg / reg$infusion_h, rep(ci_mg / 24, 3))
    )
  }
}

# Observation grid: coarse early, 0.1 h over the 48-72 h steady-state window so
# the %fT>MIC calculation and the AUC0-tau trapezoid both resolve properly.
obs_times <- sort(unique(c(seq(0, 48, by = 0.5), seq(48, 72, by = 0.1))))

make_arm <- function(reg, amt_mg, crcl, wt, n, id_offset) {
  subj <- tibble(id = id_offset + seq_len(n), CRCL = crcl, WT = wt)
  dos <- dose_rows(reg, amt_mg) |> mutate(evid = 1L, cmt = "central")
  obs <- data.frame(
    time = obs_times, amt = NA_real_, rate = NA_real_, evid = 0L,
    cmt = "central"
  )
  subj |>
    tidyr::crossing(bind_rows(dos, obs)) |>
    mutate(regimen = reg$regimen) |>
    arrange(id, time, desc(evid))
}

build_events <- function(amt_col, crcl, wt, n) {
  out <- list()
  for (i in seq_len(nrow(regimens))) {
    reg <- regimens[i, ]
    out[[i]] <- make_arm(reg, reg[[amt_col]], crcl, wt, n, id_offset = (i - 1L) * n)
  }
  bind_rows(out)
}

ev_cef <- build_events("cef_mg", crcl = 108, wt = 80, n = n_per_arm)
ev_taz <- build_events("taz_mg", crcl = 108, wt = 80, n = n_per_arm)

stopifnot(!anyDuplicated(unique(ev_cef[, c("id", "time", "evid")])))
stopifnot(!anyDuplicated(unique(ev_taz[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim_cef <- rxode2::rxSolve(
  cef, ev_cef,
  keep = c("regimen", "CRCL", "WT"), returnType = "data.frame"
)
sim_taz <- rxode2::rxSolve(
  taz, ev_taz,
  keep = c("regimen", "CRCL", "WT"), returnType = "data.frame"
)

# Solver sanity: a two-compartment solve must retain the peripheral state, and
# concentrations must never go negative (a negative tail would silently poison
# both the NCA and the %fT>MIC calculation).
stopifnot(is.null(cef$linCmt), is.null(taz$linCmt))
stopifnot(all(c("central", "peripheral1") %in% cef$state))
stopifnot(all(sim_cef$Cc >= 0), all(sim_taz$Cc >= 0))
```

# Replicate Figure 2 -- visual predictive check for unbound ceftolozane

Figure 2 of Sime 2019 is a VPC of unbound ceftolozane over a single 8-hour
dosing interval, with the observed concentrations overlaid. The observed data
are not available, so only the simulated envelope is reproduced here. This
cohort samples body weight and urinary creatinine clearance from the Table 1
distributions rather than fixing them.

```{r figure-2}
# Log-normal covariate draws matched to the Table 1 medians and IQRs
# (WT 79.5, IQR 64-99; CRCL 107, IQR 74-145). The IQR of a log-normal spans
# 1.349 sd on the log scale.
lnorm_from_iqr <- function(n, med, q1, q3) {
  stats::rlnorm(n, meanlog = log(med), sdlog = (log(q3) - log(q1)) / 1.349)
}

vpc_reg <- regimens[1, ] # 1.5 g q8h, the licensed cUTI / cIAI regimen
ev_vpc <- {
  subj <- tibble(
    id = seq_len(n_per_arm),
    WT = lnorm_from_iqr(n_per_arm, 79.5, 64, 99),
    CRCL = lnorm_from_iqr(n_per_arm, 107, 74, 145)
  )
  dos <- dose_rows(vpc_reg, vpc_reg$cef_mg) |> mutate(evid = 1L, cmt = "central")
  obs <- data.frame(
    time = seq(0, 72, by = 0.25), amt = NA_real_, rate = NA_real_,
    evid = 0L, cmt = "central"
  )
  subj |>
    tidyr::crossing(bind_rows(dos, obs)) |>
    arrange(id, time, desc(evid))
}

sim_vpc <- rxode2::rxSolve(cef, ev_vpc, keep = c("WT", "CRCL"), returnType = "data.frame")

sim_vpc |>
  filter(time >= 64) |>
  mutate(tad = time - 64) |>
  group_by(tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 4, linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Unbound ceftolozane (mg/L)",
    title = "Figure 2 - simulated 5th, 50th and 95th percentiles at steady state",
    caption = paste(
      "Replicates the structure of Figure 2 of Sime 2019 (1.5 g q8h, 1-h infusion).",
      "Dashed line: EUCAST P. aeruginosa clinical breakpoint, 4 mg/L.",
      "Observed concentrations are not publicly available and are not overlaid."
    )
  )
```

# PKNCA validation

Steady-state noncompartmental analysis over the final complete 8-hour dosing
interval (64-72 h) of the two intermittent regimens. Sime 2019 reports no NCA
table of its own, so the reference for comparison is the closed-form
steady-state identity `AUC0-tau = Dose / CL`, which the model must satisfy
exactly for a linear system.

```{r pknca}
nca_for <- function(sim, ev, mod, analyte) {
  sim_nca <- sim |>
    filter(regimen %in% c("1.5 g q8h (1 h)", "3 g q8h (1 h)")) |>
    # Only !is.na(Cc): a `time > 0` or `Cc > 0` filter here would drop the
    # time-zero anchor and trigger PKNCA's "AUC range starting before the first
    # measurement" warning once per subject.
    filter(!is.na(Cc)) |>
    select(id, time, Cc, regimen)

  sim_nca <- bind_rows(
    sim_nca,
    sim_nca |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, regimen, time, .keep_all = TRUE) |>
    arrange(id, regimen, time)

  conc_obj <- PKNCA::PKNCAconc(
    as.data.frame(sim_nca), Cc ~ time | regimen + id,
    concu = "mg/L", timeu = "h"
  )

  dose_df <- ev |>
    filter(evid == 1, regimen %in% c("1.5 g q8h (1 h)", "3 g q8h (1 h)")) |>
    select(id, time, amt, regimen) |>
    as.data.frame()
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id, doseu = "mg")

  intervals <- data.frame(
    start = 64, end = 72,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE, auclast = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res) |> mutate(analyte = analyte)
}

nca_all <- bind_rows(
  nca_for(sim_cef, ev_cef, cef, "ceftolozane"),
  nca_for(sim_taz, ev_taz, taz, "tazobactam")
)

nca_summary <- nca_all |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "cav", "auclast")) |>
  group_by(analyte, regimen, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  rename(
    "Analyte" = analyte, "Regimen" = regimen,
    "AUC0-tau (mg*h/L)" = auclast, "Cavg (mg/L)" = cav,
    "Cmax (mg/L)" = cmax, "Cmin (mg/L)" = cmin, "Tmax (h)" = tmax
  ) |>
  knitr::kable(digits = 2, caption = "Median steady-state NCA over 64-72 h, unbound concentrations, 100 subjects per arm at CRCL 108 mL/min/1.73 m^2 and 80 kg.")
```

## Closed-form check: AUC0-tau equals Dose / CL

For a linear model at steady state the area under one dosing interval is exactly
`Dose / CL`, independent of the compartment structure and of the infusion
duration. Running this on the **typical-value** model turns it into a tight
deterministic gate; the identity is checked per analyte and per dose level.

```{r auc-identity}
auc_typical <- function(mod, amt_mg, crcl = 108, wt = 80) {
  mt <- rxode2::zeroRe(mod)
  dos <- data.frame(
    time = seq(0, 64, by = 8), amt = amt_mg, rate = amt_mg / 1,
    evid = 1L, cmt = "central"
  )
  obs <- data.frame(
    time = seq(64, 72, by = 0.02), amt = NA_real_, rate = NA_real_,
    evid = 0L, cmt = "central"
  )
  ev <- bind_rows(dos, obs) |> arrange(time, desc(evid))
  ev$id <- 1L
  ev$CRCL <- crcl
  ev$WT <- wt
  out <- rxode2::rxSolve(mt, ev, returnType = "data.frame")
  stopifnot(nrow(out) == sum(ev$evid == 0))
  # Trapezoid over the closed 64-72 h interval
  sum(diff(out$time) * (head(out$Cc, -1) + tail(out$Cc, -1)) / 2)
}

auc_check <- tibble::tribble(
  ~analyte,      ~regimen,          ~amt_mg, ~intercept, ~slope,
  "ceftolozane", "1.5 g q8h (1 h)",    1000,       0.86,    6.0,
  "ceftolozane", "3 g q8h (1 h)",      2000,       0.86,    6.0,
  "tazobactam",  "1.5 g q8h (1 h)",     500,       6.90,   17.5,
  "tazobactam",  "3 g q8h (1 h)",      1000,       6.90,   17.5
) |>
  mutate(
    cl_lh = intercept + slope * 108 / 100,
    auc_closed_form = amt_mg / cl_lh
  ) |>
  rowwise() |>
  mutate(
    auc_solved = auc_typical(if (analyte == "ceftolozane") cef else taz, amt_mg)
  ) |>
  ungroup() |>
  mutate(pct_diff = 100 * (auc_solved - auc_closed_form) / auc_closed_form)

auc_check |>
  select(analyte, regimen, cl_lh, auc_closed_form, auc_solved, pct_diff) |>
  rename(
    "Analyte" = analyte, "Regimen" = regimen, "Typical CL (L/h)" = cl_lh,
    "Dose / CL (mg*h/L)" = auc_closed_form,
    "Solved AUC0-tau (mg*h/L)" = auc_solved,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Steady-state mass balance: the solved AUC over one dosing interval against the closed-form Dose / CL.")
```

```{r auc-identity-assert}
# Deterministic: zeroRe, fixed covariates, so the only error is the trapezoid
# on a 0.02 h grid. 0.2% is far tighter than any structural mistake could
# survive, and the identity fails outright if the dose is not delivered into
# central, if the infusion rate is wrong, or if the peripheral compartment
# leaks mass.
stopifnot(all(abs(auc_check$pct_diff) < 0.2))

# Dose proportionality: the system is linear, so doubling the dose must double
# the exposure exactly.
prop_ratio <- auc_check |>
  select(analyte, regimen, auc_solved) |>
  tidyr::pivot_wider(names_from = regimen, values_from = auc_solved) |>
  mutate(ratio = `3 g q8h (1 h)` / `1.5 g q8h (1 h)`)
stopifnot(all(abs(prop_ratio$ratio - 2) < 0.01))

# The 1-hour infusion is actually being delivered as an infusion, not as a
# bolus. PKNCA reports tmax RELATIVE to the interval start, so an infusion
# peaking at its end gives tmax = 1 h; a bolus would give tmax = 0 and a Cmax
# roughly 1.5-fold higher. Deterministic in structure -- every subject shares
# the same infusion duration regardless of the parameter draw -- so the window
# only has to admit the 0.1 h observation grid.
tmax_med <- nca_all |> filter(PPTESTCD == "tmax") |> pull(PPORRES) |> median()
stopifnot(tmax_med >= 0.9, tmax_med <= 1.2)
```

# Probability of target attainment

Sime 2019's principal output is a probability-of-target-attainment analysis for
ceftolozane against `%fT>MIC` targets of 40%, 60% and 100%, and for tazobactam
against `20% fT>1 mg/L`. Both are evaluated at steady state over 48-72 h, at the
cohort median covariates, which is exactly the target-attainment cohort
simulated above.

```{r pta}
pct_time_above <- function(sim, threshold) {
  sim |>
    filter(time >= 48, time <= 72) |>
    group_by(regimen, id) |>
    summarise(pct = 100 * mean(Cc > threshold), .groups = "drop")
}

mics <- c(0.25, 0.5, 1, 2, 4, 8, 16, 32)

pta_cef <- lapply(mics, function(m) {
  pct_time_above(sim_cef, m) |>
    group_by(regimen) |>
    summarise(
      `40%` = 100 * mean(pct >= 40),
      `60%` = 100 * mean(pct >= 60),
      `100%` = 100 * mean(pct >= 100),
      .groups = "drop"
    ) |>
    mutate(MIC = m)
}) |>
  bind_rows() |>
  tidyr::pivot_longer(c(`40%`, `60%`, `100%`), names_to = "target", values_to = "PTA")

pta_cef |>
  tidyr::pivot_wider(names_from = MIC, values_from = PTA, names_prefix = "MIC ") |>
  arrange(regimen, target) |>
  rename("Regimen" = regimen, "fT>MIC target" = target) |>
  knitr::kable(digits = 0, caption = "Ceftolozane PTA (%) at steady state (48-72 h), by MIC (mg/L). Replicates the structure of Sime 2019 Table 3, steady-state block. 100 subjects per arm at CRCL 108 mL/min/1.73 m^2, 80 kg.")
```

```{r pta-taz}
# Tazobactam target: 20% fT > 1 mg/L (Methods, "Dosing simulations").
pta_taz <- pct_time_above(sim_taz, 1) |>
  group_by(regimen) |>
  summarise(PTA = 100 * mean(pct >= 20), .groups = "drop")

pta_taz |>
  rename("Regimen" = regimen, "PTA for 20% fT>1 mg/L (%)" = PTA) |>
  knitr::kable(digits = 0, caption = "Tazobactam PTA at steady state. Sime 2019 Results: 'for tazobactam, all simulated dosing regimens had a 100% probability of achieving the recommended target'.")
```

## Published claims checked against the simulation

```{r pta-claims}
cef_pta <- function(reg, target, mic) {
  v <- pta_cef$PTA[pta_cef$regimen == reg & pta_cef$target == target & pta_cef$MIC == mic]
  if (length(v) != 1L) stop("no unique PTA cell for '", reg, "' / ", target, " / MIC ", mic)
  v
}

# `floor` is the gate threshold, NOT the paper's number. Every one of these is a
# cohort quantity: the paper drew 1,000 subjects, this vignette draws 100, and
# the parameter distribution here is an over-dispersed independent-marginal
# approximation to a nonparametric joint density (see Assumptions). The floors
# therefore sit well below the paper's own statements, far enough that a
# different cohort draw cannot flip them, and still go red on any error that
# moves exposure by a factor of two. `deviation = TRUE` marks a claim this model
# reproducibly does NOT reach; those are reported but excluded from the gate
# rather than having the floor widened until they pass.
claims <- tibble::tribble(
  ~claim,                                                                    ~achieved,                                    ~floor, ~deviation,
  "1.5 g q8h: 100% PTA at 40% fT>MIC up to the 4 mg/L breakpoint",
  cef_pta("1.5 g q8h (1 h)", "40%", 4), 85, FALSE,

  "3 g q8h: 100% PTA at 40% fT>MIC up to the 4 mg/L breakpoint",
  cef_pta("3 g q8h (1 h)", "40%", 4), 85, FALSE,

  "1.5 g q8h: >=90% PTA at 100% fT>MIC up to MIC 2 mg/L",
  cef_pta("1.5 g q8h (1 h)", "100%", 2), 70, FALSE,

  "1.5 g LD + 4.5 g CI: >=90% PTA at 100% fT>MIC up to MIC 8 mg/L",
  cef_pta("1.5 g LD + 4.5 g CI", "100%", 8), 75, FALSE,

  "3 g LD + 9 g CI: >=90% PTA at 100% fT>MIC up to MIC 16 mg/L",
  cef_pta("3 g LD + 9 g CI", "100%", 16), 80, FALSE,

  "All tazobactam regimens: 100% PTA for 20% fT>1 mg/L",
  min(pta_taz$PTA), 90, FALSE,

  "3 g q8h: 100% PTA at 60% fT>MIC up to the 4 mg/L breakpoint",
  cef_pta("3 g q8h (1 h)", "60%", 4), 95, TRUE,

  "3 g q8h: >=90% PTA at 100% fT>MIC up to MIC 4 mg/L",
  cef_pta("3 g q8h (1 h)", "100%", 4), 90, TRUE
) |>
  mutate(pass = achieved >= floor)

claims |>
  select(claim, achieved, floor, pass, deviation) |>
  rename(
    "Published claim (Sime 2019 Results)" = claim,
    "Simulated PTA (%)" = achieved,
    "Gate floor (%)" = floor,
    "Pass" = pass,
    "Known deviation" = deviation
  ) |>
  knitr::kable(digits = 0, caption = "Sime 2019's target-attainment conclusions, checked against this model. The last two rows are reproducible under-attainment, not noise; see Assumptions and deviations.")
```

```{r pta-claims-assert}
stopifnot(all(claims$pass[!claims$deviation]))
```

The two flagged rows are both at the most demanding end of the analysis -- the
`100% fT>MIC` target, and the `60% fT>MIC` target at the breakpoint MIC, for the
intermittent 3 g q8h regimen. They fail in one direction only: this model
attains **less** than the paper reports (roughly 80% against 100%, and 93%
against 100%). That is the expected signature of the over-dispersed
independent-marginal IIV documented below -- attaining a target for *all* of a
dosing interval is decided entirely by the fast-clearance tail of the cohort,
which is exactly the part of the distribution this approximation inflates. The
central tendency is unaffected, which is why every deterministic gate above is
tight. Nothing here suggests a transcription error, and the model was not tuned
to close the gap.

# Assumptions and deviations

- **Nonparametric distribution approximated by independent log-normal
  marginals.** NPAG estimates a discrete joint density over support points.
  Sime 2019 publishes only the marginal mean, SD and CV of each parameter, so
  the joint density -- and every parameter correlation in it -- is not
  recoverable. Each marginal is encoded here as a log-normal with
  `omega^2 = log(CV^2 + 1)`, independent of the others. This is the same
  treatment as `Setiawan_2023_sulbactam` and
  `Hughes_2024_vancomycin_nonparametric`, and it is an approximation, not a
  transcription.

  The cost is measurable. The intercept and slope of a linear clearance model
  trade off strongly against one another, and treating them as independent
  over-disperses total clearance: independent marginals give a clearance CV
  near 49% at fixed CRCL, whereas the paper's own simulated steady-state
  concentrations have a CV near 29% (5.5/19 and 3.4/11.2). Central tendency is
  unaffected -- which is why every deterministic gate above is tight -- but the
  simulated spread from this file is wider than the paper's, and PTA in the
  tails is correspondingly pessimistic. Do not read a per-subject percentile
  from this model as the paper's percentile.

  This is measurable in the target-attainment table above and is recorded there
  as two known deviations: at the `100% fT>MIC` target and MIC 4 mg/L the
  3 g q8h regimen attains about 80% here against the paper's `>=90%`, and at
  `60% fT>MIC` and MIC 4 it attains about 93% against the paper's 100%. Both
  miss in the same direction, both sit at the end of the analysis most sensitive
  to the fast-clearance tail, and no other claim in the paper is affected.

- **The tabulated mean is encoded as the log-normal MEDIAN.** Table 2 labels
  its central column "Mean". Carrying it as `log(mean)` makes the encoded
  log-normal's own mean `mean * sqrt(1 + CV^2)`, which exceeds the tabulated
  value -- by 28% on the ceftolozane intercept and by a factor of 3.1 on the
  tazobactam Kcp, whose CV is 293%. The alternative (shifting the log-scale
  location down so the log-normal mean matches the table) was rejected because
  every published number in the paper that can be checked -- Table 2's derived
  clearance and the four steady-state concentrations reproduced above -- is
  computed from the tabulated values as point estimates, and only the median
  encoding reproduces them under `zeroRe()`. The consequence for a user is
  that `zeroRe()` gives the paper's typical patient, while a stochastic cohort
  mean sits above it.

- **The creatinine-clearance slope is normalised by 100 mL/min/1.73 m^2, and
  that normalisation is derived rather than printed.** The five-step derivation
  is above and is repeated in `covariateData$CRCL` of both model files. Every
  published number in the paper that depends on the scale is reproduced by it,
  and the two alternative candidates (the cohort median, 107 or 108) are
  excluded by the sign of the residual under Jensen's inequality.

- **Residual unexplained variability is not reported and is carried as
  `fixed(0)`.** The Methods list the candidate error forms -- an additive
  `Error = (SD^2 + lambda^2)^0.5`, a multiplicative `Error = SD * gamma`, and
  an assay-error polynomial `Error = C0 + C1 * obs` whose coefficients "were
  optimized interactively" -- but do not say which was selected, and print none
  of lambda, gamma, C0 or C1. No supplement was deposited with the article.
  Inventing a residual magnitude would be fabrication, so both `propSd` and
  `addSd` are fixed at zero and the model is a structural-plus-IIV model only.
  It must not be used as-is for residual-error-aware work (for example
  simulating assay-realistic observations, or a VPC against observed data).

- **Ceftolozane and tazobactam are separate model files.** The paper fitted
  them in two independent NPAG runs, with genuinely different body-weight
  exponents on volume (1 for ceftolozane, 0.75 for tazobactam). They are not
  harmonised here.

- **The intercept is an extrapolation.** Every patient in the study had
  preserved-to-augmented renal function; patients needing renal replacement
  therapy were excluded. The clearance intercept is therefore the value of an
  additive term in a region the data never visited, and neither model carries
  information about renal impairment. This matters more for tazobactam, whose
  intercept is 27% of total clearance at the cohort median against 12% for
  ceftolozane.

- **Covariate distributions for the VPC cohort are assumed log-normal**, fitted
  to the Table 1 medians and interquartile ranges. The paper reports no
  distributional form. The target-attainment cohort does not depend on this
  assumption: it fixes CRCL and weight at the values the paper itself fixed
  them at (108 mL/min/1.73 m^2 and 80 kg).

- **Observed concentrations are not overlaid on the Figure 2 replication**
  because the individual data are not publicly available; only the simulated
  envelope is shown.

- **No published NCA table exists to compare against.** Sime 2019 reports no
  Cmax, Tmax, AUC or half-life, so the PKNCA section is validated against the
  closed-form steady-state identity `AUC0-tau = Dose/CL` and against exact dose
  proportionality rather than against transcribed reference values. The four
  steady-state concentrations in the Discussion are the paper's only published
  concentration numbers and are checked separately above.

# Errata

No erratum, corrigendum or author correction to Sime 2019 was located. Nothing
in this vignette supersedes the published article.
