---
title: "Oxcarbazepine and its 10-monohydroxy derivative in children with obesity (Sinha 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oxcarbazepine and its 10-monohydroxy derivative in children with obesity (Sinha 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE}
.mod_meta <- rxode2::rxode(readModelDb("Sinha_2026_oxcarbazepine"))
```

* Citation: `r .mod_meta$meta$reference`
* Description: `r .mod_meta$meta$description`
* Article: <https://doi.org/10.1007/s40262-025-01579-0>

Oxcarbazepine (OXZ) is an antiepileptic whose pharmacological activity is
carried almost entirely by its active metabolite, the 10-monohydroxy
derivative (MHD). OXZ is cleared extremely fast (~175 L/h in healthy adults)
by ubiquitously expressed cytosolic aldo-keto reductases, essentially all of
it via conversion to MHD; MHD is then cleared slowly (~3 L/h). The result is
a roughly 45-fold higher plasma exposure of MHD than of parent OXZ after an
oral OXZ dose.

The packaged model is Sinha 2026's final model **M6**: a one-compartment OXZ
compartment with first-order absorption feeding a one-compartment MHD, with

* **complete conversion** -- OXZ's elimination clearance *is* MHD's formation
  clearance, so the entire parent elimination flux appears in the metabolite
  compartment;
* **molar-mass correction** on both transfer directions, using the salt
  factors computed from the molecular weights of OXZ (252.27 Da) and MHD
  (254.28 Da);
* a first-order **back-transformation** `Kbt` returning MHD to OXZ, supported
  by the observation that OXZ is quantifiable in plasma after an intravenous
  MHD dose; and
* allometric scaling of every clearance and volume to **pharmacokinetic
  weight (PKWT)**, a fat-free-mass-based body-size descriptor, referenced to
  50 kg.

Body size was the only covariate retained. `V_MHD` was fixed to the published
healthy-adult non-compartmental estimate of 50 L because it is not
identifiable in a joint parent-metabolite framework fitted to sparse data.

# Population

Sinha 2026 pooled two multicentre, prospective, standard-of-care paediatric
PK studies run by the Pediatric Trials Network: **POP01**
(NICHD-2011-POP01, NCT01431326; n = 73, any patient under 21 years, with or
without obesity) and **AED01** (NICHD-2015-AED01, NCT02993861; n = 27, ages
2-18 years, all with obesity). All 100 participants were receiving
oxcarbazepine per local standard of care -- doses were not protocol-assigned.

Participants spanned postnatal ages of 44 days to 20.90 years (median 9
years) and body weights of 3.825 to 156.9 kg (median 27.2 kg). Fifty-two
percent had obesity, defined as a BMI-for-age at or above the 95th
percentile of the CDC growth charts (applicable only to participants at
least 2 years old). A total of 425 plasma concentrations were available (212
OXZ, 213 MHD), a median of 3 (range 1-7) per participant per analyte, none
below the 1 ng/mL lower limit of quantification.

Co-medications recorded on the dosing day or the prior day were
phenobarbital (n = 6), valproic acid (n = 7), topiramate (n = 8) and
levetiracetam (n = 10). No participant received carbamazepine, so
phenobarbital was the only enzyme-inducing antiepileptic assessable, and at
n = 6 it was underpowered. Renal function was normal throughout (serum
creatinine IQR 0.23-0.63 mg/dL; estimated GFR IQR 95-156 mL/min/1.73 m^2).
Gestational-age records existed for only four participants, so postmenstrual
age rests on an imputed 40-week gestation for 96% of the dataset.

The same metadata is available programmatically via
`readModelDb("Sinha_2026_oxcarbazepine")$meta$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location in
`inst/modeldb/specificDrugs/Sinha_2026_oxcarbazepine.R`. The table below
collects them. Point estimates are from Sinha 2026 **Table 4** (final model),
cross-checked against **ESM1 Table S3 row M6** and the **ESM2 NONMEM control
stream** for the final model; all three agree on every value.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` (Ka) | log(0.269) -> 0.269 1/h | Table 4, RSE 28%; ESM2 `$THETA(1)` |
| `lcl` (CL_OXZ at 50 kg PKWT) | log(220) -> 220 L/h | Table 4, RSE 13%; ESM2 `$THETA(4)`; ESM1 Table S3 M6 |
| `lvc` (V_OXZ at 50 kg PKWT) | log(33.1) -> 33.1 L | Table 4, RSE 34%; ESM2 `$THETA(2)`; ESM1 Table S3 M6 |
| `lcl_mhd` (CL_MHD at 50 kg PKWT) | log(3.05) -> 3.05 L/h | Table 4, RSE 7%; ESM2 `$THETA(5)`; ESM1 Table S3 M6 |
| `lvc_mhd` (V_MHD) | fixed(log(50)) -> 50 L | Table 4 "50 (fixed)"; ESM2 `$THETA(3)` `50 FIX`; fixed to the healthy-adult NCA estimate per Results 3.2.1 (M5) |
| `lkbt` (Kbt) | log(0.0433) -> 0.0433 1/h | Table 4, RSE 22%; ESM2 `$THETA(9)` |
| `e_ffm_cl` | fixed(1) | Table 4 "1 (fixed)"; ESM2 `$THETA(7)` `1 FIX`; estimated at 0.98 in M3b then fixed |
| `e_ffm_cl_mhd` | 0.671 | Table 4, RSE 10%; ESM2 `$THETA(8)` |
| `e_ffm_vc` | 0.752 | Table 4, RSE 10%; ESM2 `$THETA(6)`, shared by `TVV2` and `TVV3` |
| `etalcl` variance | 0.51^2 = 0.2601 | Table 4: BSV CL_OXZ 51% (RSE 19%, shrinkage 26%); ESM2 `$OMEGA` ETA(4) |
| `etalcl_mhd` variance | 0.36^2 = 0.1296 | Table 4: BSV CL_MHD 36% (RSE 12%, shrinkage 16%); ESM2 `$OMEGA` ETA(5) |
| `etalvc_mhd` variance | 0.80^2 = 0.64 | Table 4: BSV V_MHD 80% (RSE 20%, shrinkage 33%); ESM2 `$OMEGA` ETA(3) |
| BSV on Ka, Kbt, V_OXZ | none (0 fixed) | Table 4 "0 (fixed)"; ESM2 `$OMEGA` ETA(1), ETA(2), ETA(9) `0 FIX`; Results 3.2.1 model M2 |
| `propSd` (OXZ) | 0.52 | Table 4: sigma_prop,OXZ 52% CV (RSE 13%, shrinkage 14%) |
| `propSd_mhd` (MHD) | 0.24 | Table 4: sigma_prop,MHD 24% CV (RSE 28%, shrinkage 20%) |
| `sf_ft` (forward salt factor) | 254.28 / 252.27 = 1.00797 | Methods 2.3.1; ESM2 `$DES` hardcodes 1.00796 |
| `sf_bt` (backward salt factor) | 252.27 / 254.28 = 0.992094 | Methods 2.3.1; ESM2 `$DES` hardcodes 0.992095 |
| ODE `d/dt(depot)` | `-ka * depot` | ESM2 `$DES` `DADT(1)`; Fig. 1 Eq. 1 |
| ODE `d/dt(central)` | `ka*depot - kel*central + kbt*central_mhd*sf_bt` | ESM2 `$DES` `DADT(2)`; Fig. 1 Eq. 2 |
| ODE `d/dt(central_mhd)` | `kel*central*sf_ft - kelm*central_mhd - kbt*central_mhd` | ESM2 `$DES` `DADT(3)`; Fig. 1 Eq. 3 |
| Scaling form | `(PKWT / 50)^theta` on every CL and V | ESM2 `$PK` `TVV2`, `TVV3`, `TVCL`, `TVCLM`; Eqs. 7-10, with Eq. 10 as restated by the 2026 Correction |

The article carries a published Correction (Clin Pharmacokinet 2026;65:345,
doi:10.1007/s40262-025-01613-1) restating Eq. 10 as
`TV_V,OXZ = 33.1 * (PKWT/50)^0.752`; the original printed it as a duplicate of
the V_MHD equation. No parameter value changed, and the corrected form matches
the ESM2 control stream this extraction was built from. See the closing
section for detail.

# Fat-free mass (PKWT) derivation

The model consumes `FFM` (the paper's PKWT) directly as a data column. PKWT
is a composite: fat-free mass for subjects at least 3 years old, and total
body weight below 3 years, because the Al-Sallami paediatric equation was
developed in children 3 years and older (Methods 2.3.2).

ESM1 Eqs. S1-S4 give the derivation. The adult (Janmahasatian) forms are
used for 18 <= PNA < 21 years; the paediatric (Al-Sallami) forms multiply
those by an age-correction factor.

**Correction applied.** ESM1 Eq. S4 prints the female paediatric numerator as
`+0.11`. That cannot be right: the age-correction multiplier must converge to
1 at adolescence so the paediatric equation reduces to the adult Eq. S2, and
the male numerator is exactly `(1 - 0.88) = 0.12`. The female numerator is
therefore `(1 - 1.11) = -0.11`. The chunk below implements the corrected form
and asserts the convergence numerically, so the correction is a gate rather
than a claim.

```{r ffm}
ffm_adult <- function(WT, BMI, female) {
  ifelse(female, 9270 * WT / (8780 + 244 * BMI),
                 9270 * WT / (6680 + 216 * BMI))
}

# Al-Sallami paediatric age-correction multiplier, a + (1 - a)/(1 + (PNA/b)^-c).
ffm_child_factor <- function(PNA, female) {
  ifelse(female,
         1.11 - 0.11 / (1 + (PNA / 7.1)^-1.1),
         0.88 + 0.12 / (1 + (PNA / 13.4)^-12.7))
}

ffm <- function(WT, BMI, PNA, female) {
  ifelse(PNA >= 18, ffm_adult(WT, BMI, female),
         ffm_child_factor(PNA, female) * ffm_adult(WT, BMI, female))
}

# PKWT: FFM for >= 3 years, total body weight below 3 years (Methods 2.3.2).
pkwt <- function(WT, BMI, PNA, female) {
  ifelse(PNA < 3, WT, ffm(WT, BMI, PNA, female))
}

# GATE: with the corrected sign the multiplier converges to 1 in the age
# limit, so the paediatric equation reduces to the adult one. With the sign as
# printed in ESM1 Eq. S4 it converges to 1.22 instead -- a 22% error in every
# female fat-free mass. Asserting both directions is what makes this a gate
# rather than a claim.
as_printed_female <- function(PNA) 1.11 + 0.11 / (1 + (PNA / 7.1)^-1.1)

limit <- c(
  `male (corrected)`   = ffm_child_factor(1e6, FALSE),
  `female (corrected)` = ffm_child_factor(1e6, TRUE),
  `female (as printed in Eq. S4)` = as_printed_female(1e6)
)
stopifnot(abs(limit[["male (corrected)"]]   - 1) < 1e-6)
stopifnot(abs(limit[["female (corrected)"]] - 1) < 1e-3)
stopifnot(abs(limit[["female (as printed in Eq. S4)"]] - 1.22) < 1e-3)

knitr::kable(
  data.frame(
    Form = names(limit),
    `Limit as PNA -> Inf` = round(limit, 6),
    `At PNA = 18 y` = round(c(ffm_child_factor(18, FALSE),
                              ffm_child_factor(18, TRUE),
                              as_printed_female(18)), 4),
    check.names = FALSE
  ),
  row.names = FALSE,
  caption = "The Al-Sallami age-correction multiplier must converge to 1 so the paediatric equation reduces to the adult Eq. S2. Only the corrected female numerator of -0.11 does so; the +0.11 printed in ESM1 Eq. S4 converges to 1.22."
)
```

Note that the female multiplier converges slowly -- its Hill exponent is 1.1
against 12.7 for males -- so it is still 2.9% above 1 at 18 years, where the
implementation switches to the adult form. That small discontinuity is a
property of the published equations, not of this implementation.

# Structural verification: exact steady-state identities

The distinctive feature of this model is the bidirectional OXZ <-> MHD
transfer. Two exact identities fall out of it at steady state, and they are
the sharpest available test that the ODEs, the molar-mass factors and the
allometric scaling are all implemented correctly.

Setting `d/dt = 0` in the ODE system with a mean input rate `R` (mg OXZ/h),
and using that `sf_ft * sf_bt = 1` exactly:

* **Metabolite.** The back-transformation cancels completely:
  `CL_MHD * Cavg_MHD = R * sf_ft`, so
  `AUC(0-tau),MHD = Dose * sf_ft / CL_MHD`. The recycling loop moves MHD out
  and straight back in, so it cannot change metabolite exposure.
* **Parent.** Recycling *does* inflate parent exposure:
  `AUC(0-tau),OXZ = Dose * (1 + Kbt * V_MHD / CL_MHD) / CL_OXZ`. At the
  reference 50 kg PKWT the inflation factor is
  `1 + 0.0433 * 50 / 3.05 = 1.71`, so ignoring back-transformation would
  understate parent exposure by 41%.

Both identities are asserted per subject across a body-size ladder.

```{r identities}
mod <- rxode2::rxode(readModelDb("Sinha_2026_oxcarbazepine"))

SF_FT   <- 254.28 / 252.27
REF_FFM <- 50
tv <- function(FFM) list(
  cl     = 220  * (FFM / REF_FFM)^1,
  vc     = 33.1 * (FFM / REF_FFM)^0.752,
  cl_mhd = 3.05 * (FFM / REF_FFM)^0.671,
  vc_mhd = 50   * (FFM / REF_FFM)^0.752,
  kbt    = 0.0433
)

ffm_ladder <- c(8, 15, 25, 35, 50, 70)
tau <- 12
n_dose <- 30
dose_mg <- 450

ev_id <- function(i, FFM) {
  d <- data.frame(id = i, time = (seq_len(n_dose) - 1) * tau, amt = dose_mg,
                  evid = 1L, cmt = "depot", dvid = NA_integer_, FFM = FFM)
  o <- data.frame(id = i,
                  time = seq((n_dose - 1) * tau, n_dose * tau, length.out = 1201),
                  amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L, FFM = FFM)
  rbind(d, o)
}
ev_ident <- do.call(rbind, Map(ev_id, seq_along(ffm_ladder), ffm_ladder))
ev_ident <- ev_ident[order(ev_ident$id, ev_ident$time, -ev_ident$evid), ]

sim_ident <- rxode2::rxSolve(
  rxode2::zeroRe(mod)$simulationModel, ev_ident,
  keep = "FFM", returnType = "data.frame"
)

trapz <- function(t, y) {
  o <- order(t); t <- t[o]; y <- y[o]
  sum(0.5 * (y[-1] + y[-length(y)]) * diff(t))
}

ident <- sim_ident |>
  dplyr::filter(time >= (n_dose - 1) * tau) |>
  dplyr::group_by(FFM) |>
  dplyr::summarise(auc_mhd = trapz(time, Cc_mhd),
                   auc_oxz = trapz(time, Cc), .groups = "drop") |>
  dplyr::mutate(
    pred_mhd = dose_mg * SF_FT / vapply(FFM, \(f) tv(f)$cl_mhd, numeric(1)),
    pred_oxz = dose_mg *
      (1 + vapply(FFM, \(f) tv(f)$kbt * tv(f)$vc_mhd / tv(f)$cl_mhd, numeric(1))) /
      vapply(FFM, \(f) tv(f)$cl, numeric(1)),
    pct_mhd = 100 * (auc_mhd - pred_mhd) / pred_mhd,
    pct_oxz = 100 * (auc_oxz - pred_oxz) / pred_oxz,
    ratio   = auc_mhd / auc_oxz
  )

stopifnot(nrow(ident) == length(ffm_ladder))
# Tolerance is set to the accuracy actually achieved (both identities agree to
# better than 0.001%), so this stays a regression gate rather than a formality.
stopifnot(all(abs(ident$pct_mhd) < 0.01), all(abs(ident$pct_oxz) < 0.01))

ident |>
  dplyr::transmute(
    `PKWT (kg)` = FFM,
    `MHD AUC simulated` = auc_mhd, `MHD AUC identity` = pred_mhd,
    `MHD % diff` = pct_mhd,
    `OXZ AUC simulated` = auc_oxz, `OXZ AUC identity` = pred_oxz,
    `OXZ % diff` = pct_oxz,
    `MHD:OXZ ratio` = ratio
  ) |>
  knitr::kable(
    digits = c(0, 2, 2, 4, 4, 4, 4, 1),
    caption = "Steady-state AUC(0-12 h) from the ODE solution versus the exact analytical identities, 450 mg twice daily. Agreement is within 0.5% at every body size."
  )
```

The **MHD:OXZ exposure ratio** column is an independent check against a
published number: Sinha 2026's Introduction states that rapid formation and
slow elimination "results in a ~45-fold higher plasma exposure of MHD than
OXZ following an oral dose of OXZ [5]", citing a study in **healthy adults**.
The ratio is strongly body-size dependent in this model -- it rises from ~25
at 8 kg PKWT to ~47 at 70 kg, because the exponents on parent clearance (1,
fixed), metabolite clearance (0.671) and the volumes (0.752) all differ. The
comparison against the published value is therefore made at adult-like body
size, where the model reproduces it closely.

```{r ratio-check}
PUBLISHED_RATIO <- 45   # Sinha 2026 Introduction, citing reference [5] (healthy adults)
ratio_ref <- ident$ratio[ident$FFM == 50]
ratio_pct <- 100 * abs(ratio_ref - PUBLISHED_RATIO) / PUBLISHED_RATIO
cat(sprintf("MHD:OXZ steady-state exposure ratio at the 50 kg PKWT reference: %.1f-fold vs the published ~%d-fold (%.1f%% difference)\n",
            ratio_ref, PUBLISHED_RATIO, ratio_pct))
stopifnot(ratio_pct < 15)
```

# Reference-subject clearances versus published non-compartmental estimates

Sinha 2026's Discussion validates the model's clearances against published
healthy-adult non-compartmental estimates: "The estimated CL of OXZ and MHD
in a reference adult with 50 kg FFM were 220 L/h and 3.05 L/h, respectively,
which align well with the reported NCA estimates of ~175 L/h and ~3 L/h."

```{r cl-vs-nca}
cl_cmp <- data.frame(
  Parameter = c("CL_OXZ at 50 kg PKWT (L/h)", "CL_MHD at 50 kg PKWT (L/h)"),
  `Model` = c(tv(50)$cl, tv(50)$cl_mhd),
  `Published healthy-adult NCA` = c(175, 3),
  check.names = FALSE
)
cl_cmp$`% diff` <- round(100 * (cl_cmp$Model - cl_cmp$`Published healthy-adult NCA`) /
                           cl_cmp$`Published healthy-adult NCA`, 1)
stopifnot(abs(cl_cmp$Model[1] - 220) < 1e-6, abs(cl_cmp$Model[2] - 3.05) < 1e-6)
knitr::kable(
  cl_cmp, digits = 2,
  caption = "Model reference-subject clearances versus the healthy-adult NCA estimates the paper compares against. The parent difference is expected and discussed by the authors as evidence of substantial extra-hepatic metabolism."
)
```

# Virtual cohort: label dosing tiers with and without obesity

The paper's central question is whether the label's age- and weight-tiered
maintenance doses (Table 1) give children with obesity the same MHD exposure
as children without obesity. The dosing scheme simulated is:

| Age (years) | Body weight (kg) | Daily dose | Frequency |
|---|---|---|---|
| 2-4 | -- | 60 mg/kg/day | Twice daily |
| > 4 | 20-29 | 900 mg/day | Twice daily |
| > 4 | > 29-39 | 1200 mg/day | Twice daily |
| > 4 | > 39 | 1800 mg/day | Twice daily |

The paper drew 2000 virtual children from the NHANES and Pediatric Trials
Network demographic repositories. Those repositories are not on disk, so the
cohort below is instead constructed from quantities the paper itself reports,
and the construction is stated explicitly rather than sampled from an
external growth reference:

* **BMI strata.** Table 2 gives a median BMI of 18.5 kg/m^2 for POP01 (mixed
  obesity status) and 24.6 kg/m^2 for AED01 (every participant obese, BMI
  percentile 95.22-100). These are used as the non-obese and obese strata, so
  a child with obesity weighs `24.6 / 18.5 = 1.33` times as much as a
  non-obese child of the same height.
* **Age-matched pairs.** As in the paper, each child with obesity is compared
  against a non-obese child of the *same age and height*. The pair therefore
  differs only in body composition -- and the heavier child is re-tiered by
  the label rules, which is precisely the compensation mechanism the paper
  describes.
* **Heights** are *derived*, not assumed: a non-obese weight is sampled across
  the span of the label's weight bands, and the height follows as
  `sqrt(WT / 18.5)`.
* **Ages** are interpolated across the label's tier structure and are flagged
  as an assumption in the closing section.

```{r cohort}
set.seed(20260117)
n_pairs <- 200

BMI_LEAN  <- 18.5
BMI_OBESE <- 24.6

# Age anchors spanning the label's tier structure; ages between anchors are
# linearly interpolated so the cohort covers a continuum rather than four
# discrete cells.
age_anchor_wt  <- c(12, 24.5, 34, 50, 60)
age_anchor_pna <- c(3,   7,   11, 15, 18)
pna_for_lean_wt <- function(wt) {
  approx(age_anchor_wt, age_anchor_pna, xout = wt, rule = 2)$y
}

# Label daily dose (mg) from Table 1, by age and total body weight. The
# Figure 3 caption supplies the fallback: "Patients whose characteristics did
# not align with the defined tiers were assigned the same dosing and covariate
# values as patients aged 2-4 years", which covers children over 4 years who
# weigh less than the 20 kg floor of the weight bands.
label_daily_dose <- function(PNA, WT) {
  dplyr::case_when(
    PNA <= 4  ~ 60 * WT,
    WT  <  20 ~ 60 * WT,
    WT  <= 29 ~ 900,
    WT  <= 39 ~ 1200,
    TRUE      ~ 1800
  )
}

lean_wt <- exp(runif(n_pairs, log(12), log(60)))
pna     <- pna_for_lean_wt(lean_wt)
female  <- rbinom(n_pairs, 1, 0.5) == 1
ht      <- sqrt(lean_wt / BMI_LEAN)

make_arm <- function(bmi, label, id_offset) {
  wt <- bmi * ht^2
  tibble::tibble(
    id       = id_offset + seq_len(n_pairs),
    pair     = seq_len(n_pairs),
    PNA      = pna,
    HT       = ht,
    BMI      = bmi,
    WT       = wt,
    female   = female,
    obesity  = label,
    FFM      = pkwt(wt, bmi, pna, female),
    daily_mg = label_daily_dose(pna, wt)
  )
}

cohort <- dplyr::bind_rows(
  make_arm(BMI_LEAN,  "Without obesity", 0L),
  make_arm(BMI_OBESE, "With obesity",    n_pairs)
) |>
  dplyr::mutate(
    tier = dplyr::case_when(
      PNA <= 4 | WT < 20 ~ "60 mg/kg/day",
      WT  <= 29          ~ "900 mg/day",
      WT  <= 39          ~ "1200 mg/day",
      TRUE               ~ "1800 mg/day"
    ),
    tier = factor(tier, levels = c("60 mg/kg/day", "900 mg/day",
                                   "1200 mg/day", "1800 mg/day"))
  )

knitr::kable(
  cohort |>
    dplyr::group_by(obesity) |>
    dplyr::summarise(n = dplyr::n(),
                     `median age (y)` = median(PNA),
                     `median WT (kg)` = median(WT),
                     `median PKWT (kg)` = median(FFM),
                     `median daily dose (mg)` = median(daily_mg),
                     .groups = "drop"),
  digits = 1,
  caption = "Age-matched virtual cohort. Children with obesity are heavier at the same age and height, so many move up a label dose tier."
)
```

# Simulation

Dosing is twice daily, and steady state is taken as the exposure after the
sixth dose, following the paper.

```{r simulate}
TAU <- 12
N_DOSE_SS <- 6

build_events <- function(n_dose) {
  dplyr::bind_rows(
    cohort |>
      tidyr::crossing(dose_idx = seq_len(n_dose)) |>
      dplyr::mutate(time = (dose_idx - 1) * TAU, amt = daily_mg / 2,
                    evid = 1L, cmt = "depot", dvid = NA_integer_) |>
      dplyr::select(-dose_idx),
    cohort |>
      tidyr::crossing(time = seq((n_dose - 1) * TAU, n_dose * TAU, by = 0.25)) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

sim_cohort <- rxode2::rxSolve(
  mod$simulationModel, build_events(N_DOSE_SS),
  keep = c("FFM", "WT", "PNA", "tier", "obesity", "daily_mg"),
  returnType = "data.frame"
)
stopifnot("cl_mhd" %in% names(sim_cohort))
```

`Cc` and `Cc_mhd` are individual predictions and carry no residual error, so
the residual component is added explicitly where the comparison calls for it
(the paper's proportional MHD error is 24% CV).

```{r exposure}
ss <- sim_cohort |>
  dplyr::filter(time >= (N_DOSE_SS - 1) * TAU) |>
  dplyr::group_by(id, tier, obesity, WT, FFM, PNA, daily_mg) |>
  dplyr::summarise(
    cl_mhd_i = dplyr::first(cl_mhd),
    ctrough  = Cc_mhd[which.max(time)],
    cmax     = max(Cc_mhd),
    cavg     = trapz(time, Cc_mhd) / TAU,
    .groups  = "drop"
  ) |>
  dplyr::mutate(ctrough_ruv = ctrough * (1 + rnorm(dplyr::n(), 0, 0.24)))

stopifnot(nrow(ss) == nrow(cohort), all(is.finite(ss$ctrough)))
```

# Replicate published figures

## Fig. 4a: steady-state MHD concentration-time profiles

```{r fig4a, fig.width = 7.5, fig.height = 4.5}
sim_cohort |>
  dplyr::filter(time >= (N_DOSE_SS - 1) * TAU) |>
  dplyr::mutate(tad = time - (N_DOSE_SS - 1) * TAU) |>
  dplyr::group_by(obesity, tad) |>
  dplyr::summarise(Q05 = quantile(Cc_mhd, 0.05), Q50 = median(Cc_mhd),
                   Q95 = quantile(Cc_mhd, 0.95), .groups = "drop") |>
  ggplot(aes(tad, Q50, colour = obesity, fill = obesity)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = c(3, 35), linetype = "dashed", colour = "grey30") +
  labs(x = "Time after the last dose (h)", y = "MHD plasma concentration (mg/L)",
       colour = NULL, fill = NULL,
       title = "Steady-state MHD exposure at label dosing, with and without obesity",
       caption = paste("Replicates Figure 4a of Sinha 2026. Dashed lines mark the",
                       "3-35 mg/L ILAE reference trough range.")) +
  theme(legend.position = "bottom")
```

## Fig. 4b and Fig. 5: steady-state trough concentrations

```{r fig4b, fig.width = 7.5, fig.height = 4.5}
ggplot(ss, aes(obesity, ctrough, fill = obesity)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.6) +
  geom_hline(yintercept = c(3, 35), linetype = "dashed", colour = "grey30") +
  facet_wrap(~ tier, nrow = 1) +
  labs(x = NULL, y = expression(C[trough*","*ss]~of~MHD~(mg/L)), fill = NULL,
       title = "Steady-state MHD trough by dosing tier and obesity status",
       caption = "Replicates Figures 4b and 5 of Sinha 2026.") +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

## Fig. 7: dosing rate to clearance ratio

Figure 7 plots the individual dosing rate (mg/h) divided by MHD clearance
(L/h) -- which, by the steady-state identity derived above, *is* the average
steady-state MHD concentration (up to the `sf_ft` molar factor). The paper's
claim is that this ratio is comparable across the age and obesity ranges.

The identity gate for this figure is run on a typical-value (`zeroRe`) version
of the same cohort taken out to 20 doses. It has to be: at the paper's
six-dose horizon the 80% between-subject variability on `V_MHD` leaves a
handful of subjects with MHD half-lives beyond 100 h, who are genuinely not at
steady state yet and for whom the steady-state identity therefore should not
hold. Removing the random effects removes that confound and leaves the
structural claim exposed.

```{r fig7-gate}
ss_tv <- rxode2::rxSolve(
  rxode2::zeroRe(mod)$simulationModel, build_events(20),
  keep = c("FFM", "WT", "tier", "obesity", "daily_mg"),
  returnType = "data.frame"
) |>
  dplyr::filter(time >= 19 * TAU) |>
  dplyr::group_by(id, daily_mg) |>
  dplyr::summarise(cl_mhd_i = dplyr::first(cl_mhd),
                   cavg = trapz(time, Cc_mhd) / TAU, .groups = "drop") |>
  dplyr::mutate(ratio_analytic = (daily_mg / 24) * SF_FT / cl_mhd_i,
                pct = 100 * (cavg - ratio_analytic) / ratio_analytic)

stopifnot(nrow(ss_tv) == nrow(cohort))
stopifnot(max(abs(ss_tv$pct)) < 0.01)
cat(sprintf("Analytic dose-rate/CL ratio reproduces simulated Cavg,ss for all %d subjects; worst deviation %.4f%%.\n",
            nrow(ss_tv), max(abs(ss_tv$pct))))
```

```{r fig7, fig.width = 7.5, fig.height = 4.5}
ss <- ss |>
  dplyr::mutate(dose_rate = daily_mg / 24,
                ratio_analytic = dose_rate * SF_FT / cl_mhd_i)

ggplot(ss, aes(WT, ratio_analytic, colour = obesity)) +
  geom_point(alpha = 0.5, size = 1.1) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linewidth = 0.9) +
  labs(x = "Total body weight (kg)",
       y = expression(Dosing~rate/CL[MHD]==C[avg*","*ss]~(mg/L)),
       colour = NULL,
       title = "Dosing rate to MHD clearance ratio across body size and obesity status",
       caption = "Replicates Figure 7 of Sinha 2026.") +
  theme(legend.position = "bottom")
```

The `stopifnot` in the preceding chunk is the load-bearing part: it confirms
per subject that the simulated `Cavg,ss` equals the analytic
`dose rate * sf_ft / CL_MHD`, which is what makes Figure 7 a legitimate
read-out of average exposure rather than a proxy for one.

# PKNCA validation

A single-dose simulation at the reference body size supplies the standard NCA
parameters for both analytes.

```{r pknca}
ev_sd <- dplyr::bind_rows(
  tibble::tibble(id = 1L, time = 0, amt = 450, evid = 1L, cmt = "depot",
                 dvid = NA_integer_, FFM = 50),
  tibble::tibble(id = 1L, time = seq(0, 168, length.out = 673),
                 amt = NA_real_, evid = 0L, cmt = "central",
                 dvid = 1L, FFM = 50)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_sd <- rxode2::rxSolve(
  rxode2::zeroRe(mod)$simulationModel, ev_sd,
  keep = "FFM", returnType = "data.frame"
)
if (is.null(sim_sd$id)) sim_sd$id <- 1L

dose_df <- data.frame(id = 1L, time = 0, amt = 450)

nca_for <- function(analyte) {
  cdat <- sim_sd |>
    dplyr::select(id, time, conc = dplyr::all_of(analyte)) |>
    dplyr::filter(!is.na(conc))
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(cdat, conc ~ time | id, concu = "mg/L", timeu = "hr"),
    PKNCA::PKNCAdose(dose_df, amt ~ time | id, doseu = "mg"),
    intervals = data.frame(start = 0, end = 168, cmax = TRUE, tmax = TRUE,
                           auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE)
  ))
  res$result |> dplyr::mutate(analyte = analyte)
}

nca_all <- dplyr::bind_rows(nca_for("Cc"), nca_for("Cc_mhd"))
stopifnot(nrow(nca_all) > 0)

nca_all |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life")) |>
  dplyr::transmute(
    Analyte = ifelse(analyte == "Cc", "Oxcarbazepine (parent)",
                     "MHD (active metabolite)"),
    `NCA parameter` = PPTESTCD,
    Value = PPORRES
  ) |>
  knitr::kable(
    digits = 3,
    caption = "PKNCA results from a typical-value 450 mg single oral dose at the 50 kg PKWT reference."
  )
```

## Comparison against published exposure values

The paper's directly comparable published numbers are the interquartile
ranges of simulated steady-state MHD trough concentration across the whole
2-20 year virtual population: **13.81-27.13 mg/L** with obesity and
**13.79-28.16 mg/L** without.

```{r published-comparison}
iqr_tbl <- ss |>
  dplyr::group_by(obesity) |>
  dplyr::summarise(
    `Simulated Q1` = quantile(ctrough, 0.25),
    `Simulated Q3` = quantile(ctrough, 0.75),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Published Q1` = ifelse(obesity == "With obesity", 13.81, 13.79),
    `Published Q3` = ifelse(obesity == "With obesity", 27.13, 28.16)
  ) |>
  dplyr::rename(`Obesity status` = obesity) |>
  dplyr::relocate(`Obesity status`, `Published Q1`, `Simulated Q1`,
                  `Published Q3`, `Simulated Q3`)

knitr::kable(
  iqr_tbl, digits = 2,
  caption = "Interquartile range of steady-state MHD trough concentration at label dosing: this vignette's virtual cohort versus Sinha 2026 Results 3.3."
)

# The paper's headline conclusion is that the two groups' exposures are
# "largely overlapping". Score that on the same quantity and the same
# statistic the paper reports -- the IQR of C_trough,ss -- using the fraction
# of the two IQRs' union that they share. Applying the metric to the paper's
# own published quartiles gives the benchmark to compare against.
iqr_overlap <- function(q1a, q3a, q1b, q3b) {
  inter <- max(0, min(q3a, q3b) - max(q1a, q1b))
  union <- max(q3a, q3b) - min(q1a, q1b)
  inter / union
}
published_overlap <- iqr_overlap(13.81, 27.13, 13.79, 28.16)
simulated_overlap <- iqr_overlap(
  iqr_tbl$`Simulated Q1`[iqr_tbl$`Obesity status` == "With obesity"],
  iqr_tbl$`Simulated Q3`[iqr_tbl$`Obesity status` == "With obesity"],
  iqr_tbl$`Simulated Q1`[iqr_tbl$`Obesity status` == "Without obesity"],
  iqr_tbl$`Simulated Q3`[iqr_tbl$`Obesity status` == "Without obesity"]
)
cat(sprintf("IQR overlap of C_trough,ss between obesity groups: %.0f%% simulated vs %.0f%% published.\n",
            100 * simulated_overlap, 100 * published_overlap))
stopifnot(published_overlap > 0.9, simulated_overlap > 0.5)

med <- tapply(ss$ctrough, ss$obesity, median)
cat(sprintf("Median C_trough,ss: %.1f mg/L with obesity vs %.1f mg/L without (%.0f%% apart).\n",
            med[["With obesity"]], med[["Without obesity"]],
            100 * abs(diff(med)) / mean(med)))

# Every tier should sit inside the ILAE 3-35 mg/L reference trough range.
tier_med <- ss |>
  dplyr::group_by(tier, obesity) |>
  dplyr::summarise(n = dplyr::n(), med = median(ctrough), .groups = "drop")
stopifnot(nrow(tier_med) >= 6, all(tier_med$n > 0),
          all(tier_med$med > 3), all(tier_med$med < 35))
knitr::kable(tier_med |> dplyr::rename(`Dosing tier` = tier,
                                       `Obesity status` = obesity,
                                       `n` = n,
                                       `Median C_trough,ss (mg/L)` = med),
             digits = 2,
             caption = "Median steady-state MHD trough per label dose tier; all fall inside the 3-35 mg/L ILAE reference range cited by the paper.")
```

# Assumptions and deviations

* **Between-subject-variability scale.** Sinha 2026 Methods 2.2 defines BSV
  as "the standard deviation of log-normally distributed random effects with
  a zero mean", and Table 4 reports those quantities under a column headed
  "BSV estimates (CV)". The packaged `ini()` therefore treats each reported
  percentage as `omega` itself and squares it (e.g. `etalcl ~ 0.51^2`), rather
  than inverting the exact log-normal relation
  `omega^2 = log(CV^2 + 1)`. Two independent checks support this reading.
  First, the same table reports the *proportional residual* errors as "CV%",
  and for a NONMEM proportional error the CV is exactly `sqrt(sigma^2)` with
  no exponential transform -- so the paper's "CV%" label denotes
  `sqrt(variance) * 100`. Second, the reported relative standard error on the
  BSV of CL_MHD is 12%, below the `sqrt(2/N) = 14.1%` asymptotic RSE expected
  for a *variance* estimate at N = 100 subjects. That second check is
  corroborating rather than decisive -- the asymptotic relation is only
  approximate in the presence of shrinkage -- but it points the same way as
  the Methods definition, which is the primary evidence here.
  Under the alternative reading the variances would be 12%, 9% and 24% smaller
  for CL_OXZ, CL_MHD and V_MHD respectively; the structural parameters and all
  typical-value predictions in this vignette are unaffected either way.
* **ESM1 Eq. S4 sign error.** The supplement prints the female paediatric
  fat-free-mass numerator as `+0.11`. It must be `-0.11`: the multiplier has
  to converge to 1 at adolescence so Eq. S4 reduces to the adult Eq. S2, and
  the male numerator is exactly `(1 - 0.88) = 0.12`. The vignette uses the
  corrected value and asserts the convergence numerically. This affects only
  the vignette's virtual-cohort construction, not any model parameter.
* **Molar-mass factors written as exact ratios.** The ESM2 control stream
  hardcodes the salt factors as 1.00796 and 0.992095. The model file writes
  them as `254.28 / 252.27` and `252.27 / 254.28`, matching Methods 2.3.1's
  stated derivation. The two forms agree to six significant figures, and the
  ratio form is exactly reciprocal, which the steady-state identities above
  require.
* **Concentration units.** The model works in mg/L (amounts in mg, volumes in
  L). The source control stream instead sets `S2 = V2/1000` and `S3 = V3/1000`
  so that NONMEM predictions come out in ng/mL, matching the 1 ng/mL assay
  LLOQ. This is a pure scale factor of 1000; mg/L is used here because it is
  the unit the paper reports every exposure in (trough IQRs, the 3-35 mg/L
  ILAE range).
* **No bioavailability parameter.** The source control stream has no `F1`, so
  `F` is implicitly 1 -- consistent with the Introduction's statement that OXZ
  is fully absorbed. Clearances and volumes are therefore absolute, not
  apparent.
* **Parameters with BSV fixed to zero carry no eta.** Table 4 reports BSV of
  "0 (fixed)" for Ka, Kbt and V_OXZ, matching `0 FIX` in the ESM2 `$OMEGA`
  block. These are encoded as parameters without an eta rather than as
  `~ fixed(0)`, which is numerically equivalent and avoids a singular OMEGA.
* **Virtual cohort is a paper-anchored reconstruction, not the paper's.** The
  paper sampled 2000 children from the NHANES and Pediatric Trials Network
  demographic repositories, neither of which is on disk. The cohort here is
  built from quantities the paper reports: the BMI strata are the Table 2
  medians for the mixed-obesity POP01 cohort (18.5 kg/m^2) and the all-obese
  AED01 cohort (24.6 kg/m^2), and heights follow from a non-obese weight
  sampled across the span of the label's weight bands. The **weight-to-age
  mapping** -- anchors of 12 kg at 3 y, 24.5 kg at 7 y, 34 kg at 11 y, 50 kg at
  15 y and 60 kg at 18 y, linearly interpolated -- **is an assumption of this
  vignette**, not a paper value; the paper sampled uniformly across ten age
  bins drawn from the two repositories. The trough IQR comparison is therefore
  a consistency check on the exposure range, not a reproduction of the paper's
  Monte-Carlo result, and the simulated IQR overlap between obesity groups is
  expected to be lower than the paper's 93% because a deterministic
  weight-to-age mapping concentrates subjects at the label's discrete tier
  boundaries where a real cohort spreads across them. The exact structural
  identities, the MHD:OXZ exposure ratio and the reference-subject clearances
  are the checks that do not depend on the cohort construction at all.
* **Residual error in the trough comparison.** `Cc_mhd` from `rxSolve` is an
  individual prediction without residual error. The published trough IQRs may
  or may not include the 24% proportional residual; the vignette reports the
  no-residual values (a `ctrough_ruv` column carrying the residual is computed
  alongside for inspection).
* **Steady-state convention.** The paper states that "the simulated exposure
  post-sixth dose was considered at steady state", and the cohort simulation
  follows that convention so the trough comparison is scored on the same
  footing as the published numbers. For a typical subject the MHD half-life is
  roughly 10-12 h, so six 12-hourly doses reach about 99% of true steady state.
  The 80% between-subject variability on `V_MHD`, however, gives a minority of
  subjects half-lives beyond 100 h, and those are genuinely still accumulating
  at six doses -- a property of the published model that the paper's own
  simulations share. Because of that, the two exact-identity checks are run
  without random effects (the body-size ladder at 30 doses, the cohort gate at
  20), where every subject is fully converged and the structural claim is
  testable at the 0.01% level.
* **Covariates screened but not retained.** PNA, PMA (as a sigmoidal
  maturation function), serum creatinine, eGFR, phenobarbital, valproic acid,
  levetiracetam and topiramate were all tested and none improved the fit
  (ESM1 Table S5). They are recorded in the model file's
  `covariatesDataExcluded` field rather than `covariateData`, since the model
  does not reference them. The Discussion notes the dataset was underpowered
  for the enzyme-inducing-antiepileptic effect (n = 6 on phenobarbital, none
  on carbamazepine) that other oxcarbazepine population PK analyses -- including
  `Rodrigues_2017_oxcarbazepine` in this package -- did detect.
* **Race and ethnicity.** Race, ethnicity and a race-white/black indicator
  were carried in the analysis dataset (ESM2 `$INPUT`) but no summary is
  tabulated in the publication, so `population$race_ethnicity` records that
  they were not reported and the virtual cohort does not stratify on them.
* **Erratum -- checked and applied.** A Correction was published on 6 January
  2026 (Clin Pharmacokinet 2026;65:345, doi:10.1007/s40262-025-01613-1,
  PMID 41493701). It is a labelling fix: Eq. 10 of the original was printed as
  `TV_V,MHD (L) = 50 * (PKWT/50)^0.752` -- a duplicate of the preceding V_MHD
  equation -- and should read `TV_V,OXZ (L) = 33.1 * (PKWT/50)^0.752`. **No
  parameter value changes**, and the corrected form is exactly what the ESM2
  control stream (`TVV2 = THETA(2)*(PKWT/50)**THETA(6)` with
  `THETA(2) = 33.1`) and Table 4 already specify, so the packaged model was
  unaffected. The correction is recorded in the model file's `reference`
  metadata and independently confirms the V_OXZ / V_MHD assignment and the
  shared 0.752 volume exponent used here.
