---
title: "Malnutrition physiological scaling parameters for PBPK (Sjogren 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Malnutrition physiological scaling parameters for PBPK (Sjogren 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Sjogren_2021_malnutrition_physiology"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article (open access): <https://doi.org/10.3390/pharmaceutics13020204>\n")
```

Sjogren 2021 is not a drug model. It builds a physiologically based bridge from
a healthy to a **malnourished** population for whole-body physiologically based
pharmacokinetic (PBPK) modelling. Its own contribution is a set of
**physiological scaling parameters (PSPs)**: multiplicative factors, one per
system parameter, that turn the physiology of a non-malnourished individual into
that of a malnourished one at three levels of malnutrition (mild, intermediate
and severe; Table 2). The authors derived the PSPs from two adult studies
(Table 1), then multiplied them onto virtual *pediatric* populations generated
by PK-Sim, and evaluated the result against six drugs studied in malnourished
children.

The PSP layer is published in full as numbers and is what this package extracts.
Each output of the model is a PSP; multiply it onto the matching quantity of an
age-appropriate non-malnourished physiology.

## What is deliberately *not* extracted, and why

The six drug models (caffeine, cefoxitin, ciprofloxacin, lumefantrine,
pyrimethamine and sulfadoxine) were built in PK-Sim 8.0 on its generic 15-organ
whole-body model, and the paper describes that structure only in prose
(Methods 2.2). Table 4 gives compound-card inputs -- lipophilicity, fraction
unbound, pKa, solubility, intestinal permeability and elimination constants --
but tissue:plasma partitioning is given only as the *name* of a PK-Sim method
("PK-Sim Standard" or "Rodgers and Rowland"), several inputs are "as per PK-Sim
library value", the pediatric virtual populations come from PK-Sim's internal
population algorithm, and no project file was deposited (Data Availability
Statement: "Data sharing not applicable"). None of the drug layer can therefore
be reproduced outside that platform. For the same reason the virtual-population
z-score comparisons of Figures 2-4 and the drug profiles of Figures 5-8 are not
replicated here.

# Population

The PSPs were computed from two adult studies summarised in Table 1:

* **Barac-Nieto et al.** (ref 18): 49 Colombian men graded into mild,
  intermediate and severe nutritional impairment (body weight / height 89.5 %,
  82.7 % and 73.9 % of standard). Source of the serum albumin, hematocrit, fat
  mass and total blood volume changes.
* **Bosy-Westphal et al.** (ref 19): 39 Dutch men and women, underweight vs
  intermediate weight. Source of the brain, heart, liver, spleen and kidney
  masses, bone mineral content, lean soft-tissue trunk mass and skeletal muscle
  mass.

No suitable data set for malnourished children was found, so the paper assumes
that the *relative* physiological change caused by malnutrition in adults also
applies to children. The ICRP 30-year-old man (176 cm, 73 kg) is the
non-malnourished reference. The paper checked the scaling against 131 severely
malnourished and 266 non-malnourished Malian and Nigerien children aged 6-59
months (ref 23), and chose **severe** malnutrition for every PK evaluation
because it moved a virtual 46-48-month-old population to a mean weight-for-height
z-score of about -3.

The same information is available programmatically as
`readModelDb("Sjogren_2021_malnutrition_physiology")()$population`.

# Source trace

| Model element | Source location | Value / form |
|---|---|---|
| Every `psp_<component>_<level>` in `ini()` (48 values) | Table 2 (page 9) | One fixed value per component and level; the 'Not Malnourished' column is 1 throughout |
| `psp_<component> = 1 + sum(level indicator * (psp - 1))` | Table 2 layout; Methods 2.3.1 ("scaling the system parameters of each virtual individual ... to the developed set of PSPs") | Mutually exclusive level indicators, 1 when not malnourished |
| `psp_gonads`, `psp_intestine`, `psp_lung`, `psp_stomach` share one factor | Table 2 row 'Gonads, intestines, lung, stomach'; Methods 2.3.2 (lean body mass as surrogate) | `psp_leanorg` |
| `psp_albumin`, `psp_aag` share the plasma-protein factor | Table 2 row 'Plasma proteins'; Methods 2.3.2 ("Equal effect of nutrition status on alpha-1-acid glycoprotein levels as for albumin was assumed") | `psp_plasmaprot` |
| Organ PSP derivation | Equation 1 (page 4) | `PSP = (Xr,M / Xt,M) / (Xr,N / Xt,N)`; recovered from the page image, the text converters drop it |
| Plasma-protein PSP derivation | Methods 2.3.2 and Table 1 | Barac-Nieto albumin over the 4.25 g/100 mL reference |
| Skin PSP derivation | Methods 2.3.2 | Ratio of Du Bois body surface area, `BSA = 0.007184 * BWT^0.425 * HT^0.725` |
| Mild / intermediate values where only severe was measured | Methods 2.3.2 ("Linear extrapolations based on adjusted BWT/HT ratios") | Checked below |
| Consequences for a 30-year-old man | Table 3 (page 9) | Reproduced below |

# Parameter table

```{r param-table, echo = FALSE}
ini_df <- as.data.frame(ui$iniDf) |>
  dplyr::filter(!is.na(ntheta)) |>
  dplyr::select(name, est, label)
knitr::kable(ini_df, digits = 3)
```

# Simulation

The model has no states and no time dependence. One row per malnutrition level
is enough to return every PSP.

```{r simulate}
mod <- readModelDb("Sjogren_2021_malnutrition_physiology")

levels_df <- data.frame(
  id = 1:4,
  level = c("Not malnourished", "Mild", "Intermediate", "Severe"),
  MAL_NOURISH_MILD = c(0, 1, 0, 0),
  MAL_NOURISH_MOD = c(0, 0, 1, 0),
  MAL_NOURISH_SEV = c(0, 0, 0, 1)
)
ev <- levels_df |>
  dplyr::mutate(time = 0, evid = 0, amt = 0) |>
  dplyr::select(id, time, evid, amt, MAL_NOURISH_MILD, MAL_NOURISH_MOD, MAL_NOURISH_SEV)

sim <- rxode2::rxSolve(mod, ev, returnType = "data.frame", keep = "id") |>
  dplyr::left_join(levels_df[, c("id", "level")], by = "id")
stopifnot(nrow(sim) == 4)
```

# Validation

## 1. Exact reproduction of Table 2

```{r table2}
components <- c(
  "bone", "brain", "fat", "leanorg", "heart", "kidney", "liver", "muscle",
  "pancreas", "skin", "spleen", "blood_art", "blood_ven", "blood_portal",
  "plasmaprot", "hct"
)
# Sjogren 2021 Table 2, columns Not Malnourished / Mild / Intermediate / Severe.
table2 <- tibble::tribble(
  ~component,     ~nm, ~mild, ~mod,  ~sev,
  "bone",         1,   0.947, 0.913, 0.869,
  "brain",        1,   0.918, 0.865, 0.797,
  "fat",          1,   0.817, 0.822, 0.624,
  "leanorg",      1,   0.936, 0.894, 0.84,
  "heart",        1,   0.902, 0.839, 0.758,
  "kidney",       1,   0.874, 0.792, 0.686,
  "liver",        1,   0.872, 0.789, 0.682,
  "muscle",       1,   0.893, 0.771, 0.715,
  "pancreas",     1,   0.936, 0.894, 0.84,
  "skin",         1,   0.954, 0.922, 0.879,
  "spleen",       1,   0.844, 0.743, 0.612,
  "blood_art",    1,   1.03,  0.992, 0.833,
  "blood_ven",    1,   1.02,  0.979, 0.822,
  "blood_portal", 1,   1.02,  0.982, 0.825,
  "plasmaprot",   1,   0.894, 0.706, 0.494,
  "hct",          1,   0.945, 0.791, 0.681
) |>
  tidyr::pivot_longer(-component, names_to = "lvl", values_to = "published") |>
  dplyr::mutate(level = c(
    nm = "Not malnourished", mild = "Mild", mod = "Intermediate", sev = "Severe"
  )[lvl])

model_psp <- sim |>
  dplyr::select(level, dplyr::all_of(paste0("psp_", components))) |>
  tidyr::pivot_longer(-level, names_to = "component", values_to = "model") |>
  dplyr::mutate(component = sub("^psp_", "", component))

chk2 <- dplyr::inner_join(table2, model_psp, by = c("component", "level"))
# 16 components x 4 levels; a lost row would make the gate pass vacuously.
stopifnot(nrow(chk2) == 64)
# Deterministic algebra on the published cells: machine precision.
stopifnot(all(abs(chk2$model - chk2$published) < 1e-12))

# The shared factors are exactly the shared rows.
stopifnot(
  all(sim$psp_gonads == sim$psp_leanorg),
  all(sim$psp_intestine == sim$psp_leanorg),
  all(sim$psp_lung == sim$psp_leanorg),
  all(sim$psp_stomach == sim$psp_leanorg),
  all(sim$psp_albumin == sim$psp_plasmaprot),
  all(sim$psp_aag == sim$psp_plasmaprot)
)
```

```{r fig-table2, fig.height = 6, fig.cap = "Replicates Table 2 of Sjogren 2021: physiological scaling parameters by component and malnutrition level (not malnourished = 1)."}
chk2 |>
  dplyr::filter(level != "Not malnourished") |>
  dplyr::mutate(
    level = factor(level, levels = c("Mild", "Intermediate", "Severe")),
    component = factor(component, levels = rev(components))
  ) |>
  ggplot(aes(x = model, y = component, colour = level)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.5) +
  labs(x = "Physiological scaling parameter (malnourished / non-malnourished)", y = NULL, colour = "Malnutrition") +
  theme_bw()
```

## 2. The PSPs agree with the derivations the paper describes

Three of the derivation rules in Methods 2.3.2 can be re-run from numbers the
paper prints. They check the transcription of Table 2 against independent
printed values (Tables 1 and 3), not against itself.

```{r derivations}
sev <- sim[sim$level == "Severe", ]
mild <- sim[sim$level == "Mild", ]
int <- sim[sim$level == "Intermediate", ]

# (a) Plasma proteins: Barac-Nieto serum albumin (Table 1: 3.8 / 3 / 2.1 g/100 mL)
#     over the 4.25 g/100 mL non-malnourished reference.
albumin_psp <- c(3.8, 3, 2.1) / 4.25
model_pp <- c(mild$psp_plasmaprot, int$psp_plasmaprot, sev$psp_plasmaprot)
stopifnot(all(abs(round(albumin_psp, 3) - model_pp) < 1e-12))

# (b) Skin: ratio of Du Bois BSA at constant height, i.e. (BWT_M / BWT_N)^0.425,
#     with the Table 3 body weights (73.0 -> 65.3 / 60.3 / 53.9 kg).
skin_psp <- (c(65.3, 60.3, 53.9) / 73)^0.425
model_skin <- c(mild$psp_skin, int$psp_skin, sev$psp_skin)
stopifnot(all(abs(round(skin_psp, 3) - model_skin) < 1e-12))

# (c) Linear extrapolation in the adjusted BWT/HT ratio for the organs measured
#     only at severe malnutrition: PSP = 1 - slope * (1 - r), with r the
#     Barac-Nieto percent of standard BWT/HT (Table 1: 89.5 / 82.7 / 73.9 %).
r <- c(mild = 0.895, int = 0.827, sev = 0.739)
lin_organs <- c("bone", "brain", "leanorg", "heart", "kidney", "liver", "pancreas", "spleen")
lin <- do.call(rbind, lapply(lin_organs, function(o) {
  p_sev <- sev[[paste0("psp_", o)]]
  slope <- (1 - p_sev) / (1 - r[["sev"]])
  data.frame(
    component = o,
    mild_pred = 1 - slope * (1 - r[["mild"]]),
    mild_pub = mild[[paste0("psp_", o)]],
    int_pred = 1 - slope * (1 - r[["int"]]),
    int_pub = int[[paste0("psp_", o)]]
  )
}))
stopifnot(nrow(lin) == length(lin_organs))
# Rounding of the 3-decimal published PSPs allows at most ~1 unit in the
# last place after propagation through the slope.
stopifnot(
  all(abs(lin$mild_pred - lin$mild_pub) <= 0.0015),
  all(abs(lin$int_pred - lin$int_pub) <= 0.0015)
)
lin |>
  dplyr::rename(
    "Component" = component,
    "Mild, extrapolated" = mild_pred, "Mild, Table 2" = mild_pub,
    "Intermediate, extrapolated" = int_pred, "Intermediate, Table 2" = int_pub
  ) |>
  knitr::kable(digits = 3)
```

Muscle and fat are *not* on the line (muscle mild 0.893 vs 0.885 extrapolated,
intermediate 0.771 vs 0.811): the paper adjusted those two PSPs so that the
scaled organ weights add up to the target body weight (Methods 2.3.2). The
hematocrit PSPs imply a constant non-malnourished reference hematocrit, which
the paper does not print:

```{r hct}
hct_ref <- c(44.4, 37.2, 32) / c(mild$psp_hct, int$psp_hct, sev$psp_hct)
round(hct_ref, 2)
# One reference for all three levels, to within the 3-decimal rounding.
stopifnot(diff(range(hct_ref)) < 0.1)
```

## 3. Reproducing Table 3 by chaining a registry physiology

Table 3 applies the PSPs to the ICRP 30-year-old man of PK-Sim (176 cm, 73 kg).
The organ masses of that same individual are the 30-year male knot of
`Schlender_2016_aging_physiology`, the PK-Sim adult physiology already in this
package. Multiplying each Schlender organ mass by its Sjogren PSP, and summing,
should return the Table 3 body weights -- which the paper states were the design
target (mild, intermediate and severe losses of 11 %, 17 % and 26 %).

```{r table3}
phys <- readModelDb("Schlender_2016_aging_physiology")
man30 <- rxode2::rxSolve(
  phys,
  data.frame(id = 1, time = 30, evid = 0, amt = 0, SEXF = 0),
  returnType = "data.frame"
)
stopifnot(nrow(man30) == 1, abs(man30$bw - 73) < 1e-9, abs(man30$ht - 176) < 1e-9)

# Schlender organ -> Sjogren PSP. Schlender pools arterial, venous and
# portal blood into one mass; the venous PSP (the largest pool) is used.
# The constant GI mass is stomach + intestines.
organ_map <- c(
  mass_blood = "psp_blood_ven", mass_bone = "psp_bone", mass_brain = "psp_brain",
  mass_fat = "psp_fat", mass_gonads = "psp_gonads", mass_heart = "psp_heart",
  mass_kidney = "psp_kidney", mass_liver = "psp_liver", mass_lung = "psp_lung",
  mass_muscle = "psp_muscle", mass_pancreas = "psp_pancreas", mass_skin = "psp_skin",
  mass_spleen = "psp_spleen", mass_gi = "psp_intestine"
)
masses <- unlist(man30[names(organ_map)])
remainder_g <- 1000 * man30$bw - sum(masses)
remainder_g

table3 <- sim |>
  dplyr::rowwise() |>
  dplyr::mutate(
    bwt = (sum(masses * unlist(dplyr::pick(dplyr::all_of(unname(organ_map))))) + remainder_g) / 1000
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    ht = man30$ht,
    bmi = bwt / (ht / 100)^2,
    bwt_ht = bwt / (ht / 100),
    bsa = 0.007184 * bwt^0.425 * ht^0.725
  ) |>
  dplyr::select(level, bwt, ht, bmi, bwt_ht, bsa)

published3 <- data.frame(
  level = c("Not malnourished", "Mild", "Intermediate", "Severe"),
  bwt_pub = c(73.0, 65.3, 60.3, 53.9),
  bmi_pub = c(23.6, 21.1, 19.5, 17.4),
  bwt_ht_pub = c(41.4, 37.1, 34.3, 30.6),
  bsa_pub = c(1.9, 1.8, 1.7, 1.7)
)
chk3 <- dplyr::inner_join(table3, published3, by = "level")
stopifnot(nrow(chk3) == 4)

chk3 |>
  dplyr::mutate(bwt_pct_diff = 100 * (bwt - bwt_pub) / bwt_pub) |>
  dplyr::select(level, bwt, bwt_pub, bwt_pct_diff, bmi, bmi_pub, bwt_ht, bwt_ht_pub, bsa, bsa_pub) |>
  dplyr::rename(
    "Level" = level, "BWT (kg)" = bwt, "BWT, Table 3" = bwt_pub, "BWT % diff" = bwt_pct_diff,
    "BMI" = bmi, "BMI, Table 3" = bmi_pub, "BWT/HT (kg/m)" = bwt_ht,
    "BWT/HT, Table 3" = bwt_ht_pub, "BSA (m2)" = bsa, "BSA, Table 3" = bsa_pub
  ) |>
  knitr::kable(digits = 2)

# A mis-transcribed muscle or fat PSP (together ~65 % of body weight) moves the
# total by kilograms; the chained reconstruction must land within 0.5 %.
mal3 <- chk3[chk3$level != "Not malnourished", ]
stopifnot(
  nrow(mal3) == 3,
  all(abs(chk3$bwt - chk3$bwt_pub) / chk3$bwt_pub < 0.005),
  all(abs(round(chk3$bmi, 1) - chk3$bmi_pub) <= 0.1),
  all(abs(round(mal3$bwt_ht, 1) - mal3$bwt_ht_pub) <= 0.1),
  all(round(chk3$bsa, 1) == chk3$bsa_pub)
)
# The printed non-malnourished BWT/HT (41.4) is the paper's own rounding
# slip: 73.0 kg / 1.76 m = 41.48, which rounds to 41.5, and the same row's
# BMI (23.6) confirms 73.0 kg. Gate the discrepancy so a later edit cannot
# silently hide or change it.
stopifnot(
  round(73.0 / 1.76, 1) == 41.5,
  chk3$bwt_ht_pub[chk3$level == "Not malnourished"] == 41.4
)
```

The non-malnourished row closes by construction (the unscaled remainder is
defined from it). The three malnourished rows are genuine predictions: the
Schlender organ masses come from a different paper, and only the PSPs come from
Sjogren 2021.

# Using the PSPs

Each PSP multiplies the corresponding non-malnourished quantity. For a PBPK
model with explicit organ volumes, scale each volume by its PSP; scale albumin
and alpha-1-acid glycoprotein concentrations by `psp_albumin` / `psp_aag` and
the hematocrit by `psp_hct`. The paper keeps tissue composition (the volume
fractions of water, lipids and proteins used to compute partition coefficients)
out of the PSP set (Methods 2.3.2), and Table 2 has no blood-flow rows.

```{r use-example}
# Severely malnourished organ masses of the 30-year-old reference man (g).
sev_masses <- masses * unlist(sim[sim$level == "Severe", unname(organ_map)])
data.frame(
  organ = sub("^mass_", "", names(masses)),
  not_malnourished_g = masses,
  severe_g = sev_masses
) |>
  dplyr::rename("Organ" = organ, "Not malnourished (g)" = not_malnourished_g, "Severe (g)" = severe_g) |>
  knitr::kable(digits = 1, row.names = FALSE)
```

# Assumptions and deviations

* **Only the physiology layer is extracted.** The six drug models are PK-Sim
  models whose partition coefficients, pediatric populations and several inputs
  are platform-internal, with no project file deposited; see "What is
  deliberately not extracted" above.
* **Malnutrition level is encoded as three mutually exclusive indicators**
  (`MAL_NOURISH_MILD`, `MAL_NOURISH_MOD`, `MAL_NOURISH_SEV`; all 0 = not
  malnourished). The paper's "intermediate" level is carried on the `_MOD`
  member of the severity-suffix family. Setting more than one indicator to 1 is
  not meaningful and is not trapped by the model.
* **Fat PSP is not monotone in severity as printed** (mild 0.817, intermediate
  0.822). The paper adjusted fat and muscle to make the scaled organ weights add
  up to the target body weight, so the printed values are kept.
* **The severe organ PSPs cannot be re-derived here.** Equation 1 corrects the
  Bosy-Westphal organ ratios for the height difference between study groups
  with ICRP height-typical organ values, which the paper does not print. The
  mild and intermediate values of those organs *are* checked, as linear
  extrapolations of the severe value (section 2).
* **The reference hematocrit is not printed**; the three hematocrit PSPs imply
  one reference of about 47 % (section 2).
* **Table 3 reconstruction.** The organ masses come from
  `Schlender_2016_aging_physiology` (the PK-Sim ICRP 30-year-old man). That
  model pools arterial, venous and portal blood, which is scaled here with the
  venous PSP; the arterial (0.833) or portal (0.825) value changes the severe
  body weight by at most 0.03 kg. The 0.16 kg by which the Schlender organ
  masses fall short of 73 kg is left unscaled. The intermediate body weight
  reconstructs to 60.4 kg against a printed 60.3 kg (0.1 %).
* **Table 3 erratum (non-malnourished BWT/HT).** The paper prints 41.4 kg/m for
  the non-malnourished man, but its own 73.0 kg and 176 cm give 41.48 kg/m
  (41.5), and the BMI of 23.6 on the same row confirms 73.0 kg. The three
  malnourished BWT/HT values reproduce; the 41.4 is treated as a rounding slip
  in the paper.
* **Adult data, applied to children.** The PSPs are derived from adults; their
  application to pediatric populations is the paper's own assumption (Methods
  2.3.2), not something this model adds.
* **No variability.** The paper applies the PSPs as fixed factors to every
  virtual individual; this model returns them as typical values.
