---
title: "Vancomycin (Smit 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Smit 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Smit_2020_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Smit_2020_vancomycin"))$description`
* Article: [Br J Clin Pharmacol 2020;86(2):303-317](https://doi.org/10.1111/bcp.14144)
  (open access, PMC7015748)

# Population

Smit 2020 enrolled 20 morbidly obese adults scheduled for laparoscopic
sleeve gastrectomy or gastric bypass (median weight 139.0 kg, range
110.6-234.6; median BMI 45.5 kg/m^2, range 40.8-65.7; median age 38.0
years, range 23-54) and 8 nonobese healthy volunteers (median weight 69.5
kg, range 60.0-84.7; median BMI 21.2 kg/m^2; median age 25.5 years, range
20-55) at St. Antonius Hospital, Nieuwegein, The Netherlands (Table 1).
Renal impairment (eGFR < 60 mL/min/1.73 m^2) was an exclusion criterion;
median measured 24-h creatinine clearance was 141.4 mL/min in the obese
and 117.9 mL/min in the nonobese group. The sex distribution is not
reported.

Obese patients received a single 12.5 mg/kg infusion of vancomycin
(maximum 2500 mg) during or immediately after surgery; nonobese
volunteers received 1000 mg. All infusions ran at 10 mg/min. 326 plasma
samples were taken over 48 h (11-13 per subject), and the 24 samples
(7%) below the 1.5 mg/L limit of detection were handled with the M3
method.

The same information is available programmatically via
`readModelDb("Smit_2020_vancomycin")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to the source.
All values are from the final-model column of Smit 2020 Table 2.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at 70 kg) | log(5.72) L/h | Table 2, CL70kg |
| `e_wt_cl` | 0.535 | Table 2, theta1 in CL70kg x (TBW/70)^theta1 |
| `lvc` (V1 at 36.5 y) | log(16.7) L | Table 2, V1_36.5yr |
| `e_age_vc_vp` | 0.0136 /year | Table 2, theta2 (same estimate listed under both V1 and V2) |
| `lq` (Q V1-V2) | log(15.8) L/h | Table 2 |
| `lvp` (V2 at 70 kg, 36.5 y) | log(6.98) L | Table 2, V2_70kg;36.5yr |
| `lq2` (Q V1-V3) | log(5.21) L/h | Table 2 |
| `lvp2` (V3) | log(19.5) L | Table 2 |
| `etalcl` (nonobese) | fixed(0.002784) | Table 2, IIV CL nonobese 5.28% FIX; footnote b |
| `etalcl_obese` | 0.05920 | Table 2, IIV CL obese 24.7%; footnote b |
| `etalvc` | 0.18664 | Table 2, IIV V1 45.3%; footnote b |
| `propSd` | 0.0392 | Table 2, proportional error; footnote c ('shown as sigma') |
| `addSd` | 1.07 mg/L | Table 2, additive error |
| `cl = CL70kg * (WT/70)^theta1` | n/a | Table 2 row header |
| `vc = V1_36.5yr * (1 + theta2 * (AGE - 36.5))` | n/a | Table 2 row header |
| `vp = V2 * (WT/70) * (1 + theta2 * (AGE - 36.5))` | n/a | Table 2 row header |
| Three-compartment, first-order elimination | n/a | Results 3.1 |
| Combined additive + proportional error | n/a | Results 3.1 |

IIV was reported as CV% computed as sqrt(exp(omega^2) - 1) (Table 2
footnote b), so each variance is `omega^2 = log(1 + CV^2)`.

```{r model}
mod <- readModelDb("Smit_2020_vancomycin")
mod_typical <- rxode2::zeroRe(mod)
```

# Single-dose simulation of the study design

A virtual cohort of 200 obese and 200 nonobese subjects reproduces the
study's dosing: 12.5 mg/kg (maximum 2500 mg) or 1000 mg, infused at 10
mg/min. Obese weights are drawn log-normally around the Table 1 median
(139 kg) and truncated to the observed range; nonobese weights around 69.5
kg. Ages are drawn uniformly across each group's Table 1 range.

```{r cohort}
set.seed(2020)
n_arm <- 200

make_cohort <- function(n, obese, id_offset) {
  if (obese) {
    wt <- pmin(pmax(exp(rnorm(n, log(139), 0.2)), 110.6), 234.6)
    age <- runif(n, 23, 54)
    dose <- pmin(12.5 * wt, 2500)
  } else {
    wt <- pmin(pmax(exp(rnorm(n, log(69.5), 0.1)), 60), 84.7)
    age <- runif(n, 20, 55)
    dose <- rep(1000, n)
  }
  subj <- tibble(
    id = id_offset + seq_len(n),
    WT = wt,
    AGE = age,
    DIS_OBESE_MORBID = as.integer(obese),
    group = if (obese) "Morbidly obese" else "Nonobese",
    amt = dose,
    rate = 600 # 10 mg/min = 600 mg/h
  )
  obs_times <- sort(unique(c(seq(0, 12, by = 0.25), seq(13, 96, by = 1))))
  doses <- subj |> mutate(time = 0, evid = 1L, cmt = "central")
  obs <- subj |>
    select(id, WT, AGE, DIS_OBESE_MORBID, group) |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0L, amt = 0, rate = 0, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_cohort(n_arm, obese = TRUE, id_offset = 0L),
  make_cohort(n_arm, obese = FALSE, id_offset = n_arm)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("group", "WT", "DIS_OBESE_MORBID")
) |>
  as.data.frame()
```

## Concentration-time profiles (cf. Figure 1)

```{r figure-1}
sim |>
  filter(time <= 48) |>
  group_by(group, time) |>
  summarise(
    Q025 = quantile(sim, 0.025, na.rm = TRUE),
    Q50 = quantile(sim, 0.50, na.rm = TRUE),
    Q975 = quantile(sim, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = pmax(Q025, 0.1), ymax = Q975), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = 1.5, linetype = "dotted", colour = "grey40") +
  facet_wrap(~group) +
  scale_y_log10() +
  labs(
    x = "Time after start of infusion (h)",
    y = "Vancomycin concentration (mg/L)",
    title = "Simulated median and 95% interval, single dose",
    caption = paste(
      "Compare with the pcVPCs in Figure 1 of Smit 2020 (observed data are not",
      "public). Dotted line: 1.5 mg/L limit of detection."
    )
  )
```

## PKNCA validation

The paper reports no NCA table, so the NCA is checked against the
model itself. For each subject, the single-dose AUC0-inf must equal the
dose divided by that subject's individual clearance. Both sides use the
same drawn parameters, so the only differences are the trapezoidal
error and the lambda-z extrapolation.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, group)
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, group) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, group, time, .keep_all = TRUE) |>
  arrange(id, group, time)

dose_df <- events |>
  filter(evid == 1) |>
  mutate(duration = amt / rate) |>
  select(id, time, amt, duration, group)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | group + id)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | group + id,
  route = "intravascular", duration = "duration"
)
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) |>
  select(id, group, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

indiv_cl <- sim |>
  group_by(id) |>
  summarise(cl = first(cl), .groups = "drop")

chk <- nca_wide |>
  left_join(dose_df |> select(id, amt), by = "id") |>
  left_join(indiv_cl, by = "id") |>
  mutate(auc_expected = amt / cl, pct_diff = 100 * (aucinf.obs - auc_expected) / auc_expected)

chk |>
  group_by(group) |>
  summarise(
    "Median Cmax (mg/L)" = median(cmax),
    "Median AUC0-inf (mg*h/L)" = median(aucinf.obs),
    "Median t1/2 (h)" = median(half.life),
    "Median AUC % diff vs dose/CL" = median(pct_diff),
    "90th pct |% diff|" = quantile(abs(pct_diff), 0.9),
    .groups = "drop"
  ) |>
  rename("Group" = group) |>
  knitr::kable(digits = 2, caption = "PKNCA single-dose summary and AUC0-inf check against dose / individual CL.")

stopifnot(
  abs(median(chk$pct_diff)) < 2,
  quantile(abs(chk$pct_diff), 0.9) < 5
)
```

# Dosing simulations

The paper's dosing simulations (Methods 2.7) used obese individuals with
a uniform weight distribution between 90 and 230 kg receiving continuous
infusions, with day-3 AUC24h (48-72 h) compared against 400 mg*h/L (PTA)
and 700 mg*h/L (PTOX). The simulation age distribution is not stated;
the reference age of 36.5 years is used here, and Figure A4 of the paper
reports no clinically significant age effect.

## Figure 3C: 35 mg/kg/day continuous infusion, maximum 5500 mg/day

Because only clearance determines the day-3 AUC once the infusion is near
steady state, and CL is log-normal in the obese group with variance
`etalcl_obese`, PTA and PTOX follow in closed form from the typical-value
day-3 AUC. The typical-value AUC comes from a deterministic solve.

```{r fig3-typical}
wt_grid <- sort(c(seq(90, 230, by = 10), 155))
ci_events <- function(wt, id, mg_per_kg = 35, cap = 5500, ld = 0) {
  rate_ci <- min(mg_per_kg * wt, cap) / 24
  t_ci <- if (ld > 0) 2 else 0
  obs <- tibble(time = seq(0, 72, by = 0.1), evid = 0L, amt = 0, rate = 0)
  ci <- tibble(time = t_ci, evid = 1L, amt = rate_ci * (72 - t_ci), rate = rate_ci)
  doses <- if (ld > 0) bind_rows(tibble(time = 0, evid = 1L, amt = ld, rate = 600), ci) else ci
  bind_rows(doses, obs) |>
    mutate(id = id, WT = wt, AGE = 36.5, DIS_OBESE_MORBID = 1L, cmt = "central") |>
    arrange(time, desc(evid))
}
auc_window <- function(time, conc, from, to) {
  keep <- time >= from & time <= to
  t <- time[keep]
  y <- conc[keep]
  sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
}

ev_typ <- bind_rows(lapply(seq_along(wt_grid), function(i) ci_events(wt_grid[i], i)))
sim_typ <- rxode2::rxSolve(mod_typical, events = ev_typ, keep = "WT") |> as.data.frame()

omega_cl <- sqrt(0.05920) # etalcl_obese
fig3 <- sim_typ |>
  group_by(id, WT) |>
  summarise(auc_d3 = auc_window(time, Cc, 48, 72), cl = first(cl), .groups = "drop") |>
  mutate(
    dose = pmin(35 * WT, 5500),
    auc_ss = dose / cl,
    pct_of_ss = 100 * auc_d3 / auc_ss,
    PTA = pnorm(log(auc_d3 / 400) / omega_cl),
    PTOX = 1 - pnorm(log(700 / auc_d3) / omega_cl),
    P_below_900 = pnorm(log(900 / auc_d3) / omega_cl)
  )

fig3 |>
  select(WT, dose, auc_d3, pct_of_ss, PTA, PTOX, P_below_900) |>
  rename(
    "Weight (kg)" = WT, "Daily dose (mg)" = dose,
    "Typical day-3 AUC24h (mg*h/L)" = auc_d3, "Day-3 AUC as % of steady state" = pct_of_ss,
    "PTA (AUC > 400)" = PTA, "PTOX (AUC > 700)" = PTOX, "P(AUC < 900)" = P_below_900
  ) |>
  knitr::kable(digits = c(0, 0, 0, 1, 3, 3, 3), caption = "35 mg/kg/day (max 5500 mg/day) continuous infusion, day 3.")
```

The paper states for this regimen (Results 3.4) that PTA was below 90%
but still above 80% for individuals below 110 kg and above 210 kg, that
PTOX peaked around 150-160 kg, and that 94% of individuals at 150-160 kg
had an AUC24h below 900 mg*h/L. The model reproduces all four claims.
The one exception is the 90 kg edge of the simulated weight range, where
the closed-form PTA is `r round(100 * fig3$PTA[fig3$WT == 90])`%. The
paper summarises PTA per weight category; averaged over the 90, 100 and
110 kg grid points the closed-form PTA is
`r round(100 * mean(fig3$PTA[fig3$WT %in% c(90, 100, 110)]))`%.

```{r fig3-gates}
# These are deterministic (a typical-value solve plus the log-normal CDF),
# so the bounds do not depend on the random-number stream.
pta_in <- function(w) fig3$PTA[fig3$WT %in% w]
stopifnot(
  # 100-110 kg and 210-230 kg: PTA below 90% but above 80%
  all(pta_in(c(100, 110, 210, 220, 230)) > 0.80),
  all(pta_in(c(100, 110, 210, 220, 230)) < 0.90),
  # 130-200 kg: at least 90%
  all(pta_in(seq(130, 200, by = 10)) >= 0.90),
  # PTOX peaks where the 5500 mg/day cap starts to bind (about 157 kg)
  fig3$WT[which.max(fig3$PTOX)] %in% c(150, 155, 160),
  # about 94% of individuals at 150-160 kg below 900 mg*h/L
  all(abs(fig3$P_below_900[fig3$WT %in% c(150, 155, 160)] - 0.94) < 0.02),
  # the day-3 AUC is essentially at steady state, so Figure 3 does not
  # depend on whether a loading dose was given
  all(abs(fig3$pct_of_ss - 100) < 1)
)
```

A stochastic cohort (200 per weight band) shows the same picture with
Monte Carlo noise.

```{r fig3-mc}
set.seed(3)
bands <- c(100, 130, 160, 190, 220)
n_band <- 200
ev_mc <- bind_rows(lapply(seq_along(bands), function(b) {
  wts <- runif(n_band, bands[b] - 15, bands[b] + 15)
  bind_rows(lapply(seq_len(n_band), function(j) {
    ci_events(wts[j], (b - 1) * n_band + j) |>
      filter(evid == 1 | time >= 48) |>
      mutate(band = paste0(bands[b] - 15, "-", bands[b] + 15, " kg"))
  }))
}))
sim_mc <- rxode2::rxSolve(mod, events = ev_mc, keep = c("band", "WT")) |>
  as.data.frame() |>
  mutate(band = factor(band, levels = paste0(bands - 15, "-", bands + 15, " kg")))
mc <- sim_mc |>
  group_by(id, band) |>
  summarise(auc_d3 = auc_window(time, Cc, 48, 72), .groups = "drop")

mc |>
  group_by(band) |>
  summarise(
    "Mean AUC24h" = mean(auc_d3),
    "2.5th pct" = quantile(auc_d3, 0.025),
    "97.5th pct" = quantile(auc_d3, 0.975),
    "PTA" = mean(auc_d3 > 400),
    "PTOX" = mean(auc_d3 > 700),
    .groups = "drop"
  ) |>
  rename("Weight band" = band) |>
  knitr::kable(digits = 2, caption = "Stochastic replicate of Figure 3C (n = 200 per band).")

ggplot(mc, aes(band, auc_d3)) +
  geom_boxplot() +
  geom_hline(yintercept = c(400, 700), linetype = "dashed", colour = "grey50") +
  labs(
    x = "Weight band", y = "Day-3 AUC24h (mg*h/L)",
    caption = "Replicates the left panel of Figure 3C of Smit 2020 (35 mg/kg/day, maximum 5500 mg/day)."
  )
```

## Figure 4: loading dose before a continuous infusion

The paper recommends a 1500 mg loading dose (10 mg/min) followed 2 h
after its start by the 35 mg/kg/day continuous infusion, reporting a
day-1/day-3 AUC24h ratio close to 1 at every weight. The typical-value
ratios are:

```{r figure-4}
lds <- c(500, 1000, 1500, 2000, 2500)
ev_ld <- bind_rows(lapply(seq_along(lds), function(k) {
  bind_rows(lapply(seq_along(wt_grid), function(i) {
    ci_events(wt_grid[i], (k - 1) * length(wt_grid) + i, ld = lds[k]) |> mutate(ld = lds[k])
  }))
}))
sim_ld <- rxode2::rxSolve(mod_typical, events = ev_ld, keep = c("WT", "ld")) |> as.data.frame()
ratios <- sim_ld |>
  group_by(id, WT, ld) |>
  summarise(ratio = auc_window(time, Cc, 0, 24) / auc_window(time, Cc, 48, 72), .groups = "drop")

ggplot(ratios, aes(WT, ratio, colour = factor(ld))) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  labs(
    x = "Weight (kg)", y = "AUC24h day 1 / AUC24h day 3", colour = "Loading dose (mg)",
    caption = "Typical-value replicate of Figure 4 of Smit 2020."
  )

r1500 <- ratios$ratio[ratios$ld == 1500]
stopifnot(all(abs(r1500 - 1) < 0.15))
```

With 1500 mg, the typical ratio stays between `r round(min(r1500), 2)`
and `r round(max(r1500), 2)` across 90-230 kg.

## Figure 5: trough concentration versus AUC24h, twice-daily dosing

For 35 mg/kg/day (maximum 5500 mg/day) given as two infusions per day at
10 mg/min, the paper reports that the trough concentrations matching
the 400-700 mg*h/L AUC24h window span 5.7-14.6 mg/L. The paper samples
the trough 0.5 h before the second dose of day 3.

```{r figure-5}
set.seed(5)
n_bid <- 200
wts_bid <- runif(n_bid, 90, 230)
ev_bid <- bind_rows(lapply(seq_len(n_bid), function(j) {
  dose <- min(35 * wts_bid[j], 5500) / 2
  bind_rows(
    tibble(time = seq(0, 60, by = 12), evid = 1L, amt = dose, rate = 600),
    tibble(time = c(seq(48, 72, by = 0.25), 59.5), evid = 0L, amt = 0, rate = 0)
  ) |>
    distinct(time, evid, .keep_all = TRUE) |>
    mutate(id = j, WT = wts_bid[j], AGE = 36.5, DIS_OBESE_MORBID = 1L, cmt = "central") |>
    arrange(time, desc(evid))
}))
sim_bid <- rxode2::rxSolve(mod, events = ev_bid, keep = "WT") |> as.data.frame()
bid <- sim_bid |>
  group_by(id, WT) |>
  summarise(
    auc_d3 = auc_window(time, Cc, 48, 72),
    trough = Cc[which.min(abs(time - 59.5))],
    .groups = "drop"
  )
in_window <- bid |> filter(auc_d3 >= 400, auc_d3 <= 700)
trough_range <- quantile(in_window$trough, c(0.025, 0.975))

ggplot(bid, aes(trough, auc_d3)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, colour = "black") +
  geom_hline(yintercept = c(400, 700), linetype = "dashed") +
  geom_vline(xintercept = trough_range, colour = "red") +
  labs(
    x = "Day-3 trough concentration (mg/L)", y = "Day-3 AUC24h (mg*h/L)",
    caption = "Replicates Figure 5A of Smit 2020 (n = 200 here, 10 000 in the paper)."
  )

summary(lm(auc_d3 ~ trough, data = bid))$r.squared
```

The 2.5th-97.5th percentiles of the trough among simulated subjects in the
AUC window are `r round(trough_range[1], 1)`-`r round(trough_range[2], 1)`
mg/L, against 5.7-14.6 mg/L in the paper.

```{r figure-5-gate}
# Robust centre check: the median in-window trough lies inside the paper's
# window, and each simulated percentile is within 3 mg/L of the paper's value.
stopifnot(
  median(in_window$trough) > 5.7, median(in_window$trough) < 14.6,
  abs(trough_range[[1]] - 5.7) < 3,
  abs(trough_range[[2]] - 14.6) < 3
)
```

## Figure A4: age

```{r figure-a4}
ages <- c(20, 30, 40, 50)
ev_age <- bind_rows(lapply(seq_along(ages), function(i) {
  ci_events(130, i, ld = 1500) |> filter(time <= 24 | evid == 1) |> mutate(AGE = ages[i])
}))
rxode2::rxSolve(mod_typical, events = ev_age, keep = "AGE") |>
  as.data.frame() |>
  ggplot(aes(time, Cc, colour = factor(AGE))) +
  geom_line() +
  labs(
    x = "Time (h)", y = "Vancomycin concentration (mg/L)", colour = "Age (years)",
    caption = paste(
      "Replicates Figure A4 of Smit 2020: 130 kg, 1500 mg loading dose then",
      "35 mg/kg/day continuous infusion from 2 h."
    )
  )
```

# Assumptions and deviations

- **Cohort-specific CL variability.** Table 2 gives separate IIV on CL
  for the nonobese (5.28%, fixed) and obese (24.7%) groups with a single
  typical value. The model draws both etas and selects one with
  `DIS_OBESE_MORBID`. Each group's variance is therefore exact, but
  `etalcl_obese` is not mu-referenced. This affects re-estimation
  efficiency, not simulation; the load-time note about non-mu-referenced
  etas is expected.
- **Shared age slope.** Table 2 lists theta2 = 0.0136 under both V1 and
  V2 with identical precision, so it is encoded as one parameter
  (`e_age_vc_vp`) used in both equations.
- **Residual error scale.** Table 2 footnote c states the proportional
  error is shown as sigma (a standard deviation). The additive error is
  taken as an SD in mg/L by the same convention.
- **Simulation cohorts.** The paper's dosing simulations used 10 000
  individuals and did not state the age distribution. Here the closed-form
  PTA/PTOX use the reference age of 36.5 years, and the stochastic
  replicates use 200 subjects per band. The single-dose cohort weights are
  log-normal around the Table 1 medians, since individual weights are not
  published.
- **Figure 3 without a loading dose.** The Methods do not say whether the
  Figure 3 regimens included a loading dose. The day-3 AUC is simulated
  without one; the typical day-3 AUC is already within a few percent of
  steady state (table above), so a loading dose barely changes it.
- **PTOX level.** The closed-form PTOX at 150-160 kg is about 30%. The
  paper's Discussion describes PTOX as below 20% "for most individuals"
  across the whole 90-230 kg range, and gives no per-weight value to
  compare against. The 94% below 900 mg*h/L at 150-160 kg, which it does
  give, is reproduced.
- **BLQ handling.** The M3 method was used for estimation only and has no
  effect on the model structure.
- No erratum or correction notice linked to this article (PMID
  31661553) was found in Europe PMC as of 2026-09-25.
