---
title: "Finerenone PK, UACR, serum potassium and eGFR (Snelder 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Finerenone PK, UACR, serum potassium and eGFR (Snelder 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: Snelder N, Heinig R, Drenth HJ, Joseph A, Kolkhof P, Lippert J,
  Garmann D, Ploeger B, Eissing T. Population Pharmacokinetic and
  Exposure-Response Analysis of Finerenone: Insights Based on Phase IIb Data
  and Simulations to Support Dose Selection for Pivotal Trials in Type 2
  Diabetes with Chronic Kidney Disease. Clin Pharmacokinet.
  2020;59(3):359-370.
* Article: <https://doi.org/10.1007/s40262-019-00820-x>

The paper contributes four models, all fitted to the pooled ARTS-DN and
ARTS-DN Japan phase IIb data (the "final" ARTS-DN+JP models):

| Model | Endpoint | Source table |
|---|---|---|
| `Snelder_2020_finerenone` | finerenone plasma concentration | ESM Table S1 |
| `Snelder_2020_finerenone_uacr` | urinary albumin:creatinine ratio | ESM Table S2 |
| `Snelder_2020_finerenone_potassium` | serum potassium | ESM Table S3 |
| `Snelder_2020_finerenone_egfr` | eGFR (CKD-EPI) | ESM Table S4 |

The three PK/PD models were fitted sequentially, with individual post hoc PK
parameters from the final PK model as the input (ESM "Phase IIb PK/PD Model
Development"). Each PK/PD file therefore carries the PK model with every PK
parameter and PK variance wrapped in `fixed()`; the PK residual error is not
carried because concentrations are not an output of those models.

```{r load-models}
mod_pk <- readModelDb("Snelder_2020_finerenone")
mod_uacr <- readModelDb("Snelder_2020_finerenone_uacr")
mod_k <- readModelDb("Snelder_2020_finerenone_potassium")
mod_egfr <- readModelDb("Snelder_2020_finerenone_egfr")
```

# Population

ARTS-DN (NCT01874431, global, no sites in Japan) and ARTS-DN Japan
(NCT01968668) randomised adults with type 2 diabetes and persistent
albuminuria (UACR >= 30 mg/g) on a renin-angiotensin system blocker to
placebo or finerenone 1.25, 2.5, 5, 7.5, 10, 15 or 20 mg once daily for 90
days. The PK/PD dataset held 893 subjects (799 ARTS-DN, 94 ARTS-DN Japan) and
the PK dataset 787 subjects with 4597 quantifiable concentrations (paper
Table 2). Median (5th-95th percentile) baseline body weight was 90.6
(64.1-126.3) kg in ARTS-DN and 71.6 (54-100) kg in ARTS-DN Japan; eGFR-MDRD
was 63.9 (33.5-102.5) and 61.5 (41.0-87.2) mL/min/1.73 m^2; median age was
65 and 64 years (paper Table 3). The PK/PD dataset was 75.7% Caucasian,
9.1% non-Japanese Asian, 10.5% Japanese and 2.9% African American.

```{r population}
rxode2::rxode2(mod_pk)$population
```

# Source trace

| Quantity | Value | Source |
|---|---|---|
| Ka (depot and all three transits) | 10.7 1/h | ESM Table S1 (footnote d for the shared rate) |
| CL/F | 37.3 L/h | ESM Table S1 |
| Vc/F (= Vp/F) | 123 L | ESM Table S1; equal volumes per ESM "Phase IIa Models" and paper Discussion 4.1 |
| Q/F | 0.433 L/h | ESM Table S1 |
| Lag time | 0.215 h (fixed) | ESM Table S1 |
| F | 1 (fixed) | ESM Table S1 |
| BW slope on Vc/F | 0.449 | ESM Table S1 |
| eGFR-MDRD slope on CL/F and 1/F | 0.101 | ESM Table S1 |
| Covariate formulae and medians (88.5 kg, 63.53 mL/min/1.73 m^2) | - | ESM Table S1 footnotes |
| omega^2 Ka, CL/F, V/F, cov(CL/F, V/F) | 0.585, 0.2, 0.0927, 0.0928 | ESM Table S1 |
| PK proportional sigma^2 | 0.179 | ESM Table S1 |
| UACR baseline, screening UACR <= 300 / > 300 mg/g | 96.8 / 633 mg/g | ESM Table S2 (cat2 / cat3); paper Results 3.2.1 |
| UACR kout | 0.0014 1/h | ESM Table S2 |
| UACR Imax, IC50 | 1 (fixed), 12.9 ug/L | ESM Table S2; ESM Eq. S2 and S7 |
| UACR omega^2 baseline | 0.512 | ESM Table S2 |
| UACR additive sigma^2 (log scale); Japanese factor | 0.181; 0.692 | ESM Table S2 and footnote a |
| Potassium baseline; Japanese effect | 4.31 mmol/L; -3.13% | ESM Table S3 and footnote b |
| Potassium kin | 0.0815 mmol/L/h | ESM Table S3 |
| Potassium log-linear slope | 0.0204 | ESM Table S3 (footnote a); ESM Eq. S3 and S5 |
| Potassium omega^2 baseline (additive) | 0.102 | ESM Table S3 |
| Potassium proportional sigma^2; Japanese factor | 0.00432; 0.752 | ESM Table S3 and footnote c |
| eGFR-EPI baseline | 66.6 mL/min/1.73 m^2 | ESM Table S4 |
| eGFR kout | 0.0023 1/h | ESM Table S4 |
| eGFR power-model slope, exponent | 0.0359, 0.231 | ESM Table S4 (footnote a); ESM Eq. S2 and S6 |
| eGFR omega^2 baseline (additive); Japanese factor | 451; 0.647 | ESM Table S4 and footnote b |
| eGFR proportional sigma^2; Japanese factor | 0.00944; 0.675 | ESM Table S4 and footnote c |

# Pharmacokinetics

## Virtual cohorts

Body weight and eGFR-MDRD are drawn log-normally around the Table 3 medians,
with log-scale SDs chosen so the 5th-95th percentiles approximately match
Table 3. Covariates are drawn independently; their correlation is not
reported.

```{r pk-cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)
n_arm <- 200

make_cohort <- function(n, wt_med, wt_p5, wt_p95, gfr_med, gfr_p5, gfr_p95,
                        japanese, id0) {
  data.frame(
    id = id0 + seq_len(n),
    WT = exp(rnorm(n, log(wt_med), log(wt_p95 / wt_p5) / (2 * 1.645))),
    CRCL = exp(rnorm(n, log(gfr_med), log(gfr_p95 / gfr_p5) / (2 * 1.645))),
    RACE_JAPANESE = japanese
  )
}

cohort_pk <- dplyr::bind_rows(
  make_cohort(n_arm, 90.6, 64.1, 126.3, 63.9, 33.5, 102.5, 0, 0) |>
    dplyr::mutate(study = "ARTS-DN"),
  make_cohort(n_arm, 71.6, 54, 100, 61.5, 41.0, 87.2, 1, 1000) |>
    dplyr::mutate(study = "ARTS-DN Japan")
)
```

Each subject receives finerenone 20 mg once daily for 30 days, with dense
sampling over the last dosing interval (day 30, the Cmax,ss day of paper
Figure 1).

```{r pk-sim}
dose_mg <- 20
obs_times <- 29 * 24 + seq(0, 24, by = 0.25)
ev_pk <- dplyr::bind_rows(
  cohort_pk |>
    dplyr::mutate(time = 0, evid = 1, amt = dose_mg, cmt = "depot", ii = 24, addl = 29),
  tidyr::crossing(cohort_pk, time = obs_times) |>
    dplyr::mutate(evid = 0, amt = 0, cmt = "central", ii = 0, addl = 0)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_pk <- rxode2::rxSolve(mod_pk, ev_pk, keep = "study", returnType = "data.frame")
```

```{r pk-profile, fig.cap = "Simulated finerenone concentration over the day-30 dosing interval after 20 mg once daily (median and 90% prediction interval, including residual error). Compare with the ARTS-DN VPC of ESM Figure S1."}
sim_pk |>
  dplyr::mutate(tad = time - 29 * 24) |>
  dplyr::group_by(study, tad) |>
  dplyr::summarise(
    med = median(sim), lo = quantile(sim, 0.05), hi = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80") +
  geom_line() +
  scale_y_log10() +
  facet_wrap(~study) +
  labs(x = "Time after dose (h)", y = "Finerenone (ug/L)")
```

## PKNCA: dose-normalised steady-state exposure

Paper Figure 1 reports the median dose-normalised AUCss (26.6 and 26.7
ug*h/L/mg) and Cmax,ss at day 30 (6.8 and 7.3 ug/L/mg) for ARTS-DN and
ARTS-DN Japan, calculated from empirical Bayes estimates. The simulated
concentrations are divided by the dose and analysed against a 1 mg dose.

```{r pknca}
conc_nca <- sim_pk |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(time = time - 29 * 24, Cdn = Cc / dose_mg) |>
  dplyr::select(id, study, time, Cdn)
dose_nca <- cohort_pk |>
  dplyr::distinct(id, study) |>
  dplyr::mutate(time = 0, amt = 1)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_nca, Cdn ~ time | study + id),
  PKNCA::PKNCAdose(dose_nca, amt ~ time | study + id),
  intervals = data.frame(start = 0, end = 24, auclast = TRUE, cmax = TRUE)
))

published_nca <- data.frame(
  study = c("ARTS-DN", "ARTS-DN Japan"),
  auclast = c(26.6, 26.7),
  cmax = c(6.8, 7.3)
)
cmp_nca <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_nca,
  by = "study",
  units = c(auclast = "ug*h/L per mg", cmax = "ug/L per mg"),
  tolerance_pct = 10
)
knitr::kable(cmp_nca, caption = "Median dose-normalised AUC0-24,ss and Cmax,ss: simulation versus paper Figure 1.")
```

```{r pknca-assert}
nca_med <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::group_by(study, PPTESTCD) |>
  dplyr::summarise(med = median(PPORRES), .groups = "drop") |>
  dplyr::left_join(
    tidyr::pivot_longer(published_nca, -study, names_to = "PPTESTCD", values_to = "ref"),
    by = c("study", "PPTESTCD")
  )
nca_med
# Cohort medians: a mis-transcribed CL/F, volume or unit would move these by
# tens of percent.
stopifnot(all(abs(nca_med$med / nca_med$ref - 1) < 0.10))
```

## Covariate effects on exposure

The paper states that a subject with eGFR-MDRD 30 mL/min/1.73 m^2 has a
32.9% higher AUCss and a 17.5% higher Cmax than a subject with eGFR-MDRD 90,
and that a 50 kg subject has a 43.1% higher Cmax than a 100 kg subject
(Results 3.1). These are typical-value contrasts. The eGFR contrast is
reproduced exactly by the ARTS-DN-only slope (0.126, ESM Table S1, second
column) and not by the final ARTS-DN+JP slope (0.101), so the quoted
percentages were computed with the intermediate ARTS-DN-only model; both are
shown. The ARTS-DN-only model differs from the final model in Ka, CL/F,
Vc/F, Q/F and both covariate slopes (ESM Table S1); its covariate medians
(90.4 kg, 63.9 mL/min/1.73 m^2) are not overridable here, so the final-model
medians are used for both.

```{r covariate-contrasts}
mod_pk_typ <- rxode2::zeroRe(mod_pk)
mod_pk_artsdn <- mod_pk_typ |>
  rxode2::ini(
    lka = log(11.4), lcl = log(36.9), lvc = log(125), lq = log(0.415),
    e_wt_vc = 0.516, e_crcl_cl = 0.126
  )

ss_metrics <- function(model, wt, crcl) {
  ev <- rxode2::et(amt = 1, ii = 24, addl = 29, cmt = "depot") |>
    rxode2::et(29 * 24 + seq(0, 24, by = 0.05), cmt = "central") |>
    as.data.frame() |>
    dplyr::mutate(WT = wt, CRCL = crcl)
  s <- as.data.frame(rxode2::rxSolve(model, ev))
  c(auc = sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2), cmax = max(s$Cc))
}

contrast <- function(model) {
  g30 <- ss_metrics(model, 88.5, 30)
  g90 <- ss_metrics(model, 88.5, 90)
  w50 <- ss_metrics(model, 50, 63.53)
  w100 <- ss_metrics(model, 100, 63.53)
  100 * c(
    auc_egfr30_vs_90 = g30[["auc"]] / g90[["auc"]] - 1,
    cmax_egfr30_vs_90 = g30[["cmax"]] / g90[["cmax"]] - 1,
    cmax_wt50_vs_100 = w50[["cmax"]] / w100[["cmax"]] - 1
  )
}

tab_contrast <- data.frame(
  contrast = c("AUCss, eGFR 30 vs 90", "Cmax,ss, eGFR 30 vs 90", "Cmax,ss, BW 50 vs 100 kg"),
  paper = c(32.9, 17.5, 43.1),
  artsdn_only = contrast(mod_pk_artsdn),
  final = contrast(mod_pk_typ)
)
tab_contrast |>
  dplyr::rename(
    "Contrast (% higher)" = contrast,
    "Paper Results 3.1" = paper,
    "ARTS-DN-only parameters" = artsdn_only,
    "Final ARTS-DN+JP parameters" = final
  ) |>
  knitr::kable(digits = 1, row.names = FALSE)

# Closed-form typical-value contrasts: no random component, so a tight bound.
stopifnot(
  abs(tab_contrast$artsdn_only[1] - 32.9) < 0.3,
  abs(tab_contrast$artsdn_only[2] - 17.5) < 0.5,
  abs(tab_contrast$artsdn_only[3] - 43.1) < 3
)
```

The steady-state AUC for 1 mg is F / CL = 1 / 37.3 L/h = 26.8 ug*h/L in the
typical subject, matching the 26.6 ug*h/L/mg median of Figure 1.

# Pharmacodynamics

## Typical time courses

```{r pd-typical, fig.cap = "Typical-value response over 180 days of once-daily finerenone (median covariates, screening UACR <= 300 mg/g, non-Japanese)."}
typ_ev <- function(dose) {
  ev <- rxode2::et(seq(0, 180 * 24, by = 12), cmt = "central")
  if (dose > 0) ev <- rxode2::et(ev, amt = dose, ii = 24, addl = 179, cmt = "depot")
  as.data.frame(ev) |>
    dplyr::mutate(WT = 88.5, CRCL = 63.53, UACR = 200, RACE_JAPANESE = 0)
}
typ_run <- function(model, state, label) {
  dplyr::bind_rows(lapply(c(0, 10, 20), function(d) {
    s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(model), typ_ev(d)))
    data.frame(time = s$time, value = s[[state]], dose = paste(d, "mg"), endpoint = label)
  }))
}
typ_pd <- dplyr::bind_rows(
  typ_run(mod_uacr, "uacr", "UACR (mg/g)"),
  typ_run(mod_k, "serum_k", "Serum potassium (mmol/L)"),
  typ_run(mod_egfr, "egfr", "eGFR-EPI (mL/min/1.73 m^2)")
)
ggplot(typ_pd, aes(time / 24, value, colour = dose)) +
  geom_line() +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(x = "Day", y = NULL, colour = "Dose")
```

## Time to 99% of the steady-state drug effect

The abstract gives 138, 20 and 85 days for UACR, serum potassium and eGFR.
The typical 20 mg response is evaluated once a day at trough, relative to
its value after 400 days.

```{r t99}
# A 400-day solve, so the last value is the steady state even for UACR
ev_t99 <- rxode2::et(amt = 20, ii = 24, addl = 399, cmt = "depot") |>
  rxode2::et(seq(0, 400 * 24, by = 24), cmt = "central") |>
  as.data.frame() |>
  dplyr::mutate(WT = 88.5, CRCL = 63.53, UACR = 200, RACE_JAPANESE = 0)
t99_days <- function(model, state) {
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(model), ev_t99))
  frac <- (s[[state]] - s[[state]][1]) / (s[[state]][nrow(s)] - s[[state]][1])
  s$time[which(frac >= 0.99)[1]] / 24
}
t99 <- data.frame(
  endpoint = c("UACR", "Serum potassium", "eGFR-EPI"),
  days = c(t99_days(mod_uacr, "uacr"), t99_days(mod_k, "serum_k"), t99_days(mod_egfr, "egfr")),
  paper = c(138, 20, 85)
)
knitr::kable(t99 |> dplyr::rename("Endpoint" = endpoint, "Simulated (days)" = days, "Paper (days)" = paper))
# Deterministic typical-value solve: tight bound.
stopifnot(all(abs(t99$days / t99$paper - 1) < 0.05))
```

## Scale of the baseline inter-individual variability

ESM Tables S3 and S4 print the baseline omega^2 for serum potassium (0.102)
and eGFR-EPI (451) without stating the scale. Paper Table 3 gives the
observed baseline percentiles, which include residual error. For eGFR a
variance of 451 is only meaningful on the additive scale. For potassium the
two readings differ enough for Table 3 to decide: an exponential IIV of
variance 0.102 would put the 95th percentile of baseline potassium above
7 mmol/L. The quantiles below are computed from the model's distributions
(no ODE solve), so they are not subject to cohort sampling noise.

```{r baseline-scale}
q <- c(0.05, 0.5, 0.95)
k_add <- qnorm(q, 4.31, sqrt(0.102 + 4.31^2 * 0.00432))
k_exp <- 4.31 * exp(qnorm(q, 0, sqrt(0.102 + 0.00432)))
g_add <- qnorm(q, 66.6, sqrt(451 + 66.6^2 * 0.00944))
tab_bl <- data.frame(
  quantity = rep(c("Serum potassium (mmol/L)", "eGFR-EPI (mL/min/1.73 m^2)"), each = 3),
  percentile = rep(c("5th", "50th", "95th"), 2),
  table3_artsdn = c(3.6, 4.3, 5.0, 33.3, 66.3, 101.3),
  additive_iiv = c(k_add, g_add),
  exponential_iiv = c(k_exp, rep(NA, 3))
)
tab_bl |>
  dplyr::rename(
    "Quantity" = quantity, "Percentile" = percentile,
    "Paper Table 3 (ARTS-DN)" = table3_artsdn,
    "Additive IIV (as coded)" = additive_iiv,
    "Exponential IIV (rejected)" = exponential_iiv
  ) |>
  knitr::kable(digits = 2)
stopifnot(
  all(abs(k_add - c(3.6, 4.3, 5.0)) < 0.1),
  k_exp[3] > 7,
  # eGFR: the median and upper tail are compared tightly. The observed 5th
  # percentile is trimmed by the ARTS-DN eGFR >= 30 inclusion criterion,
  # which an untruncated normal IIV does not reproduce, so it may only sit
  # below the observed value, and by a few units.
  abs(g_add[2] - 66.3) < 1,
  abs(g_add[3] - 101.3) < 3,
  g_add[1] < 33.3,
  g_add[1] > 33.3 - 5
)
```

The approximation of the observed baseline as normal (IIV plus a
proportional error evaluated at the typical baseline) is adequate here
because both coefficients of variation are small. The simulated eGFR 5th
percentile falls a few units below the observed one because ARTS-DN
enrolled only patients with eGFR of at least 30 mL/min/1.73 m^2, which
trims the lower tail of the observed baseline distribution.

## Phase III dose-response scenario (paper Figure 3)

Paper Table 4 defines the scenario: 180 days of once-daily dosing with
measurements on days 1, 30, 60, 90 and 180; eGFR 25-90 mL/min/1.73 m^2;
serum potassium below 4.8 mmol/L at run-in and at screening. The cohort
below uses the ARTS-DN body-weight and eGFR distributions with eGFR
truncated to 25-90. The same event table drives all three PK/PD models.

Because UACR and eGFR are linear turnover models, the ratio of a
subject's response to its own baseline does not depend on the baseline.
The percentage of subjects crossing a threshold is therefore computed from
each simulated subject's model-predicted ratio, with the residual error of
the baseline and the follow-up measurement integrated analytically. This
keeps Monte-Carlo noise from 200 subjects per arm out of percentages that
are only a few percent.

```{r ph3-sim}
cohort_ph3 <- function(n, id0) {
  crcl <- exp(rnorm(5 * n, log(63.9), log(102.5 / 33.5) / (2 * 1.645)))
  data.frame(
    id = id0 + seq_len(n),
    WT = exp(rnorm(n, log(90.6), log(126.3 / 64.1) / (2 * 1.645))),
    CRCL = head(crcl[crcl >= 25 & crcl <= 90], n),
    UACR = 200,
    RACE_JAPANESE = 0
  )
}
arms <- c(0, 10, 20, 30)
visit_days <- c(1, 30, 60, 90, 180)
ev_ph3 <- dplyr::bind_rows(lapply(seq_along(arms), function(i) {
  cov <- cohort_ph3(n_arm, 1000 * i) |> dplyr::mutate(arm = arms[i])
  obs <- tidyr::crossing(cov, time = (visit_days - 1) * 24) |>
    dplyr::mutate(evid = 0, amt = 0, ii = 0, addl = 0)
  if (arms[i] == 0) {
    return(obs)
  }
  dplyr::bind_rows(
    cov |> dplyr::mutate(time = 0, evid = 1, amt = arms[i], ii = 24, addl = 179, cmt = "depot"),
    obs
  )
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

run_ph3 <- function(model, state) {
  ev <- ev_ph3 |> dplyr::mutate(cmt = ifelse(evid == 0, state, cmt))
  s <- rxode2::rxSolve(model, ev, keep = "arm", returnType = "data.frame")
  s |>
    dplyr::mutate(day = time / 24 + 1, value = .data[[state]]) |>
    dplyr::group_by(id) |>
    dplyr::mutate(ratio = value / value[day == 1]) |>
    dplyr::ungroup() |>
    dplyr::select(id, arm, day, ratio)
}
ph3_uacr <- run_ph3(mod_uacr, "uacr")
ph3_egfr <- run_ph3(mod_egfr, "egfr")
ph3_k <- run_ph3(mod_k, "serum_k")
```

### UACR targets

UACR is modelled on the log scale with residual SD `sqrt(0.181)`, so the log
of an observed ratio has SD `sqrt(2 * 0.181)` around the log of the
predicted ratio.

```{r ph3-uacr}
sd_uacr <- sqrt(0.181)
targets <- c(0.8, 0.7, 0.6, 0.5)
pct_uacr <- ph3_uacr |>
  dplyr::filter(day == 180, arm > 0) |>
  dplyr::group_by(arm) |>
  dplyr::reframe(
    target = targets,
    pct = sapply(targets, function(x) 100 * mean(pnorm((log(x) - log(ratio)) / (sqrt(2) * sd_uacr))))
  )
ref_uacr <- data.frame(
  arm = rep(c(20, 30), each = 4),
  target = rep(targets, 2),
  paper = c(66.4, 58.6, 49.0, 39.5, 66.4 + 5.5, 58.6 + 5.2, 49.0 + 6.5, 39.5 + 5.7)
)
chk_uacr <- dplyr::inner_join(pct_uacr, ref_uacr, by = c("arm", "target"))
chk_uacr |>
  dplyr::mutate(target = paste("<=", target)) |>
  dplyr::rename(
    "Dose (mg QD)" = arm, "UACR day 180 / baseline" = target,
    "Simulated (%)" = pct, "Paper Results 3.3.1 (%)" = paper
  ) |>
  knitr::kable(digits = 1)
# Semi-analytic percentages over 200 PK-variable subjects: the centre and
# the envelope, not the extremes, are asserted.
stopifnot(
  abs(median(chk_uacr$pct - chk_uacr$paper)) < 4,
  max(abs(chk_uacr$pct - chk_uacr$paper)) < 6
)
```

```{r ph3-uacr-plot, fig.cap = "Percentage of subjects reaching each UACR target at day 180 by dose. Replicates Figure 3a of Snelder 2020 (points: values quoted in Results 3.3.1)."}
ggplot(pct_uacr, aes(arm, pct, colour = factor(target))) +
  geom_line() +
  geom_point(data = ref_uacr, aes(y = paper), shape = 1, size = 3) +
  labs(x = "Finerenone dose (mg once daily)", y = "Subjects reaching target (%)", colour = "UACR ratio <=")
```

### eGFR decreases

The eGFR residual error is proportional (SD `sqrt(0.00944)`) on both the
baseline and the day-180 measurement; the baseline error is integrated
numerically.

```{r ph3-egfr}
sd_egfr <- sqrt(0.00944)
grid_e <- seq(-6, 6, length.out = 241)
w_e <- dnorm(grid_e) / sum(dnorm(grid_e))
p_drop <- function(r, x) {
  sum(w_e * pnorm(((1 - x) * (1 + sd_egfr * grid_e) / r - 1) / sd_egfr))
}
drops <- c(0.25, 0.30, 0.40)
pct_egfr <- ph3_egfr |>
  dplyr::filter(day == 180) |>
  dplyr::group_by(arm) |>
  dplyr::reframe(
    drop = drops,
    pct = sapply(drops, function(x) 100 * mean(sapply(ratio, p_drop, x = x)))
  )
ref_egfr <- data.frame(
  arm = c(20, 20, 30, 30),
  drop = c(0.25, 0.30, 0.25, 0.30),
  paper = c(5.0, 1.8, 5.5, 2.0)
)
chk_egfr <- dplyr::inner_join(pct_egfr, ref_egfr, by = c("arm", "drop"))
chk_egfr |>
  dplyr::mutate(drop = paste0(">= ", 100 * drop, "%")) |>
  dplyr::rename(
    "Dose (mg QD)" = arm, "eGFR decrease at day 180" = drop,
    "Simulated (%)" = pct, "Paper Results 3.3.3 (%)" = paper
  ) |>
  knitr::kable(digits = 2)
stopifnot(
  all(abs(chk_egfr$pct - chk_egfr$paper) < 1),
  all(pct_egfr$pct[pct_egfr$drop == 0.40 & pct_egfr$arm %in% c(20, 30)] < 0.5)
)
```

The paper also states fewer than 0.1% of subjects lose 40% or more; the
simulated values at 20 and 30 mg are
`r paste(signif(pct_egfr$pct[pct_egfr$drop == 0.40 & pct_egfr$arm %in% c(20, 30)], 2), collapse = " and ")`%.

### Serum potassium above 5.5 mmol/L

Unlike UACR and eGFR, this percentage depends on each subject's baseline and
on the screening rule, so the true baseline is integrated over its additive
IIV distribution, N(4.31, 0.102). Inclusion requires the observed
pre-treatment potassium to be below 4.8 mmol/L. Table 4 names two such
measurements (run-in and screening); whether the day-1 measurement was also
conditioned on is not stated, and the paper stresses that the result is
highly sensitive to this rule (Discussion 4.3). Both readings are shown.
The measurement ratios are taken from each subject's own solve; after day
30 they are at steady state (20 days to 99%), where the ratio does not
depend on the baseline.

```{r ph3-k}
sd_k <- sqrt(0.00432)
grid_k <- seq(-6, 6, length.out = 241)
w_k <- dnorm(grid_k) / sum(dnorm(grid_k))
r0 <- 4.31 + sqrt(0.102) * grid_k
p_exceed <- function(ratios, n_screen) {
  incl <- pnorm((4.8 / r0 - 1) / sd_k)^n_screen
  none <- sapply(r0, function(b) prod(pnorm((5.5 / (b * ratios) - 1) / sd_k)))
  sum(w_k * incl * (1 - none)) / sum(w_k * incl)
}
pct_k <- ph3_k |>
  dplyr::group_by(arm, id) |>
  dplyr::summarise(
    two = p_exceed(ratio, 2),
    three = p_exceed(ratio, 3),
    .groups = "drop"
  ) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(two = 100 * mean(two), three = 100 * mean(three), .groups = "drop") |>
  dplyr::left_join(data.frame(arm = c(20, 30), paper = c(1.3, 1.7)), by = "arm")
pct_k |>
  dplyr::rename(
    "Dose (mg QD)" = arm,
    "Simulated, 2 screening values < 4.8 (%)" = two,
    "Simulated, 3 screening values < 4.8 (%)" = three,
    "Paper Results 3.3.2 (%)" = paper
  ) |>
  knitr::kable(digits = 2)
chk_k <- dplyr::filter(pct_k, !is.na(paper))
# The published value sits between the two readings of the screening rule.
stopifnot(
  all(chk_k$three < chk_k$paper + 0.2),
  all(chk_k$two > chk_k$paper - 0.2),
  all(diff(pct_k$two) > 0)
)
```

# Assumptions and deviations

* **Additive baseline IIV for serum potassium and eGFR.** ESM Tables S3 and
  S4 print the baseline omega^2 without its scale, and the Methods state
  that random effects were exponential. The paper Table 3 baseline
  percentiles are reproduced only by an additive (normal) IIV for both
  endpoints (see "Scale of the baseline inter-individual variability"), so
  both are coded additively. The UACR baseline IIV is exponential, the
  natural scale for data the Methods say were log-transformed.
* **"Ethnic effect on sigma^2" in ESM Table S4 is the baseline IIV.** The
  footnote gives 291 for the Japanese population, which is 0.647 times the
  baseline omega^2 of 451; no global additive residual variance appears in
  the table. The factor is therefore applied to the baseline IIV variance.
* **Residual-variance factors for Japanese subjects** multiply variances,
  so the coded SDs are multiplied by the square root of each factor.
  `RACE_JAPANESE` is equivalent to the ARTS-DN Japan study indicator
  because ARTS-DN enrolled no Japanese subjects.
* **UACR baseline categories.** ESM Table S2 labels the baselines "cat2"
  and "cat3"; the paper text defines the split as screening UACR above
  300 mg/g versus at or below it, which is how `UACR` is used.
* **Log-linear potassium effect.** ESM Eq. S5 is read as the natural
  logarithm, `log(Cc + 1)` with Cc in ug/L.
* **Covariate time dependence.** Whether body weight and eGFR-MDRD entered
  as baseline or time-varying values is not stated; the simulations use
  baseline values.
* **PK in the PK/PD files.** The PK/PD models were fitted to individual
  post hoc PK; the files instead carry the population PK model with its
  variability fixed, which reproduces the PK distribution rather than any
  individual.
* **Covariate contrasts in Results 3.1** were evidently computed with the
  ARTS-DN-only PK model (ESM Table S1, second column): its eGFR slope
  reproduces 32.9% and 17.5% exactly, while the final slope gives the
  smaller values in the table above. The packaged model carries the final
  estimates.
* **ESM Table S4 "ARTS-DN only" slope of 0.371** is ten times the final
  estimate (0.0359) and would suppress eGFR almost completely at clinical
  concentrations; it is presumably a misprint and is not used.
* **Phase III threshold checks** use typical PD parameters and no parameter
  uncertainty, whereas the paper simulated 5000 subjects per arm including
  NONMEM parameter uncertainty (ESM "Simulations"). Only the median
  percentages are compared.
* **rxode2 warns that `etarbase` in the eGFR model is not mu-referenced**
  because the Japanese factor scales the eta; the warning does not affect
  simulation.
