---
title: "Dapagliflozin, empagliflozin and canagliflozin renal SGLT1/2 QSP (Sokolov 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dapagliflozin, empagliflozin and canagliflozin renal SGLT1/2 QSP (Sokolov 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Sokolov_2020_sglt_qsp"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Sokolov_2020_sglt_qsp"))$description`
* Article: <https://doi.org/10.1002/psp4.12498>

Sokolov et al. (2020) used a quantitative systems pharmacology (QSP) model of
renal glucose filtration, SGLT-mediated reabsorption and urinary excretion,
first reported by Yakovleva et al. (2019), to compare how strongly
dapagliflozin, empagliflozin and canagliflozin inhibit SGLT2 (S1/S2 segments of
the proximal tubule) and SGLT1 (S3 segment) in type 2 diabetes mellitus (T2DM).
The supplement deposits the complete IQRtools model code
(`SGLT_model_code.txt`) and the fitting / simulation script
(`Script_Fitting_Simulation.R`); the packaged model is a line-by-line
translation of that code.

Each gliflozin is dosed orally in mg into its own `depot_<drug>` state and
passes through a transit chain into a plasma compartment of volume 2.75 L
shared by all three drugs (dapagliflozin also has a peripheral compartment).
The unbound drug is filtered at the glomerular filtration rate into the S1/S2
lumen (`pct_<drug>`), flows on to the S3 lumen (`pst_<drug>`), the bladder and
the urine. Glucose is filtered at `GFR x GLU` and reabsorbed by
Michaelis-Menten SGLT2 kinetics in S1/S2 and SGLT1 kinetics in S3. The drug
free in the tubular lumen inhibits each transporter competitively. The model
is a typical-value model fit to study-level mean data. It has no
between-subject variability and no residual error.

```{r model}
mod <- rxode2::rxode2(readModelDb("Sokolov_2020_sglt_qsp"))
mw <- c(dapagliflozin = 408.87, empagliflozin = 450.91, canagliflozin = 444.5)
drugs <- names(mw)
drug_cols <- c(dapagliflozin = "#E41A1C", empagliflozin = "#377EB8", canagliflozin = "#4DAF4A")
```

# Population

The model was fit to published, study-level mean data from phase I and II
trials in healthy volunteers and adults with T2DM (Sokolov 2020 Table S1):
plasma concentration-time profiles, 24-hour urinary drug excretion and 24-hour
urinary glucose excretion (UGE). Each trial arm's mean daily plasma glucose
(MPG) and eGFR were supplied as regressors. The pharmacodynamic T2DM data set
comprised 7 dapagliflozin arms (2.5-100 mg), 11 canagliflozin arms (25-400 mg)
and 14 empagliflozin arms (1-100 mg). Median MPG was 7.8, 9.24 and 10.43 mM,
and median eGFR 98.1, 100 and 100 mL/min/1.73 m^2, in the dapagliflozin,
empagliflozin and canagliflozin trials respectively; the pooled T2DM medians
were 9.33 mM and 100 mL/min/1.73 m^2 (Table S2). Subject counts and
demographics are not reported because only arm means were modelled.

The same information is available programmatically via
`readModelDb("Sokolov_2020_sglt_qsp")()$population`.

# Source trace

Every `ini()` value comes from the `MODEL PARAMETERS` section of the deposited
`SGLT_model_code.txt` and carries an in-file comment. The ODEs and rate laws
are Sokolov 2020 Table S3. Where Table S3 and the code differ, the code is
followed; see *Assumptions and deviations*.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lvc` (plasma volume, all drugs) | log(2.75 L), fixed | Code `Vpl` |
| `v_pct`, `v_pst`, `v_bladder` | 0.045, 0.01944, 0.2 L, fixed | Code `Vlumen1`, `Vlumen2`, `Vbladder` |
| `q_pct`, `q_pst`, `q_urine` | 2.7, 0.72, 0.055 L/h, fixed | Code `Qlumen`, `Qbladder`, `Qurine` |
| `lfdepot_*` | 0.78 / 0.78 / 0.65, fixed | Code `Fdapa`, `Fempa`, `Fcana` |
| `fu_*` | 0.086 / 0.22 / 0.01, fixed | Code `fupdapa`, `fupempa`, `fupcana` |
| `mw_*`, `mw_glucose` | 408.87 / 450.91 / 444.5 / 180.156 g/mol, fixed | Code `MW*` |
| `lktr_*` | 10.38 / 19.8 / 11.93 1/h | Code `k_tr_d`, `k_tr_e`, `k_tr_c` |
| `lka_*` | 0.5071 / 0.09984 / 0.1358 1/h | Code `kabs*` |
| `lcl_*` | 13.07 / 5.78 / 8.242 L/h | Code `CL*pls` |
| `lq_dapagliflozin`, `lvp_dapagliflozin` | 13.65 L/h, 98.57 L | Code `Qdapa`, `Vprfdapa` |
| `km_sglt2`, `km_sglt1` | 4, 0.5 mM, fixed | Code `KmreabsSGLT2`, `KmreabsSGLT1` |
| `vmax_total_healthy`, `vmax_total_t2dm` | 105.6, 140 mmol/h, fixed | Code `Vmax_hs`, `Vmax_t2d` |
| `lvmax_sglt2_healthy`, `lvmax_sglt2_t2dm` | 87.07, 111.4 mmol/h | Code `VmaxreabsSGLT2hs`, `VmaxreabsSGLT2t2d` |
| `lki_sglt2_*` (healthy) | 103.1 / 638.6 / 364.3 pM | Code `Ki*SGLT2ex` |
| `ratio_sglt1_*` | 1157 / 1249 / 158, fixed | Code `Ki*SGLT1ex` expressions (in vitro SGLT1:SGLT2 ratio) |
| `f_ki_t2dm` | 0.3047 | Code `coef` |
| `reabs_base` | 39.174 mmol/h, fixed | Code `reabsbase` |
| Transit chains (5 / 5 / 4 states) | n/a | Code `MODEL REACTIONS` `Vtr*_d`, `Vtr*_e`, `Vtr*_c` |
| Drug disposition ODEs | n/a | Table S3 (Dapagliflozin / Empagliflozin / Canagliflozin blocks) |
| SGLT2 / SGLT1 reabsorption rate laws | n/a | Table S3 `V_reabs1`, `V_reabs2`; code `reabsSGLT2dr`, `reabsSGLT1dr` |
| T2DM Ki factor and Vmax switch | n/a | Code `khs` / `kt2d` terms in `reabsSGLT*dr` |
| GFR from eGFR (`CRCL * 60 / 1000`) | n/a | Script, Figure 1 block (`GFR * 1000 / 60`) |

The six SGLT parameters (`lvmax_sglt2_*`, `lki_sglt2_*`, `f_ki_t2dm`) are the
ones the deposited script estimates; the PK parameters were estimated in the
companion analysis. Together these are the paper's 17 estimated parameters.
The 27 literature parameters are `fixed()`.

# Exact gates on derived quantities

The T2DM Ki values in Table 1 are not model parameters. They are derived from
the healthy SGLT2 Ki, the T2DM factor and the in vitro ratio. The glucose
filtration fluxes in Table S2 are `GFR x MPG`. Both must reproduce exactly.

```{r ki-gate}
ini_df <- mod$iniDf
th <- setNames(ini_df$est, ini_df$name)
ki2_t2dm <- exp(th[paste0("lki_sglt2_", drugs)]) * th[["f_ki_t2dm"]] / 1000 # nM
ki1_t2dm <- ki2_t2dm * th[paste0("ratio_sglt1_", drugs)]
ki_tbl <- data.frame(
  drug = drugs,
  ki2_model = unname(ki2_t2dm), ki2_table1 = c(0.031, 0.195, 0.111),
  ki1_model = unname(ki1_t2dm), ki1_table1 = c(36.35, 243.03, 17.55)
)
knitr::kable(
  ki_tbl |>
    dplyr::rename(
      "Drug" = drug,
      "Ki SGLT2 model (nM)" = ki2_model, "Ki SGLT2 Table 1 (nM)" = ki2_table1,
      "Ki SGLT1 model (nM)" = ki1_model, "Ki SGLT1 Table 1 (nM)" = ki1_table1
    ),
  digits = 3,
  caption = "T2DM inhibitory constants derived from the packaged parameters versus Sokolov 2020 Table 1."
)
# Table 1 prints the SGLT2 Ki to 3 decimals. The SGLT1 Ki agree to 0.1%: the
# canagliflozin value is 17.538 nM here and in the deposited script's own
# Figure 4 block (57.559 * 0.3047), against 17.55 printed in Table 1.
stopifnot(
  all(abs(ki_tbl$ki2_model - ki_tbl$ki2_table1) <= 0.0005 + 1e-9),
  all(abs(ki_tbl$ki1_model / ki_tbl$ki1_table1 - 1) < 1e-3)
)

flux <- data.frame(
  scenario = c("dapagliflozin", "empagliflozin", "canagliflozin", "common"),
  mpg = c(7.8, 9.24, 10.43, 9.33),
  egfr = c(98.1, 100, 100, 100),
  flux_tableS2 = c(45.89, 55.48, 62.61, 55.98)
) |>
  dplyr::mutate(flux_model = egfr * 60 / 1000 * mpg)
# Table S2 prints the medians rounded (e.g. MPG 7.8 mM) while its flux column
# is the product of the unrounded medians, so agreement is to 0.1%, not exact.
stopifnot(all(abs(flux$flux_model / flux$flux_tableS2 - 1) < 2e-3))
```

# Steady-state initial condition

The deposited code starts the tubular glucose states at zero and runs a 72-h
drug-free burn-in before dosing. The packaged model instead starts each glucose
state at its drug-free steady state, so a dose can be given at time 0. Without
a dose, the glucose states and the UGE rate must stay constant over the whole
simulation.

```{r ss-gate}
make_subject <- function(id, drug, dose, glu = 9.33, egfr = 100, t2dm = 1,
                         times = seq(0, 24, by = 0.1)) {
  dose_row <- data.frame(
    id = id, time = 0, evid = 1, amt = dose, cmt = paste0("depot_", drug)
  )
  obs_rows <- data.frame(
    id = id, time = times, evid = 0, amt = 0, cmt = paste0("central_", drug)
  )
  dplyr::bind_rows(dose_row, obs_rows) |>
    dplyr::mutate(GLU = glu, CRCL = egfr, DIS_DIAB = t2dm, drug = drug, dose = dose)
}
solve_df <- function(d, m = mod) {
  as.data.frame(rxode2::rxSolve(m, d, keep = c("drug", "dose"), returnType = "data.frame"))
}

pbo <- solve_df(make_subject(1, "dapagliflozin", 0, times = seq(0, 200, by = 1)))
drift <- pbo |>
  dplyr::summarise(
    dplyr::across(c(glu_pct, glu_pst, glu_bladder), ~ max(abs(.x / .x[1] - 1)))
  )
uge_rate <- diff(pbo$uge) / diff(pbo$time)
stopifnot(
  all(unlist(drift) < 1e-6),
  max(abs(uge_rate / uge_rate[1] - 1)) < 1e-6
)
drift
```

# Table 1: drug concentrations in plasma, kidney and bladder

Single oral doses at the highest approved doses (dapagliflozin 10 mg,
empagliflozin 25 mg, canagliflozin 300 mg) in a typical T2DM subject with the
common median MPG (9.33 mM) and eGFR (100 mL/min/1.73 m^2). As in the deposited
script, the 0-24 h average is the mean over a 0.1-h output grid.

```{r table1}
approved <- c(dapagliflozin = 10, empagliflozin = 25, canagliflozin = 300)
typ <- dplyr::bind_rows(lapply(seq_along(drugs), function(i) {
  make_subject(i, drugs[i], approved[[drugs[i]]], times = seq(0, 120, by = 0.1))
})) |>
  solve_df()

conc_long <- dplyr::bind_rows(lapply(drugs, function(d) {
  s <- typ[typ$drug == d, ]
  data.frame(
    drug = d, time = s$time,
    Plasma = s[[paste0("Cc_", d)]] / mw[[d]] * 1000,
    `S1-S2` = s[[paste0("pct_", d)]] / 0.045 * 1e6,
    S3 = s[[paste0("pst_", d)]] / 0.01944 * 1e6,
    Bladder = s[[paste0("bladder_", d)]] / 0.2 * 1e6,
    check.names = FALSE
  )
})) |>
  tidyr::pivot_longer(c(Plasma, `S1-S2`, S3, Bladder), names_to = "site", values_to = "conc_nM")

t1_sim <- conc_long |>
  dplyr::filter(time <= 24) |>
  dplyr::group_by(drug, site) |>
  dplyr::summarise(cmax = max(conc_nM), cavg = mean(conc_nM), .groups = "drop")

t1_ref <- data.frame(
  drug = rep(drugs, each = 4),
  site = rep(c("Plasma", "S1-S2", "S3", "Bladder"), 3),
  cmax_ref = c(275, 52.5, 196, 987, 533, 260, 976, 8172, 6156, 137, 513, 3922),
  cavg_ref = c(51.7, 9.87, 37.0, 462, 229, 112, 420, 5164, 2109, 46.9, 176, 2205)
)
t1 <- dplyr::left_join(t1_ref, t1_sim, by = c("drug", "site")) |>
  dplyr::mutate(
    cmax_pct = 100 * (cmax - cmax_ref) / cmax_ref,
    cavg_pct = 100 * (cavg - cavg_ref) / cavg_ref
  )
knitr::kable(
  t1 |>
    dplyr::select(drug, site, cmax_ref, cmax, cmax_pct, cavg_ref, cavg, cavg_pct) |>
    dplyr::rename(
      "Drug" = drug, "Site" = site,
      "Cmax Table 1 (nM)" = cmax_ref, "Cmax model (nM)" = cmax, "Cmax % diff" = cmax_pct,
      "Cavg0-24 Table 1 (nM)" = cavg_ref, "Cavg0-24 model (nM)" = cavg, "Cavg % diff" = cavg_pct
    ),
  digits = 2,
  caption = "Replicates Table 1 of Sokolov 2020 (model-predicted Cmax and Cavg0-24)."
)
# Same parameters and same deterministic model as the paper's simulation; the
# residual difference is rounding of the printed values and the output grid.
stopifnot(
  nrow(t1) == 12L, !anyNA(t1$cmax), !anyNA(t1$cavg),
  max(abs(t1$cmax_pct)) < 1,
  max(abs(t1$cavg_pct)) < 2
)
```

```{r fig3-conc, fig.height = 5}
ggplot(conc_long |> dplyr::filter(time <= 120), aes(time / 24, conc_nM, colour = drug)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~site, nrow = 1) +
  scale_y_log10(limits = c(1e-5, 1e4)) +
  scale_colour_manual(values = drug_cols) +
  labs(x = "Time (days)", y = "Free drug concentration (nM)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Replicates Figure 3 (top row) of Sokolov 2020: drug concentrations in plasma,
the S1/S2 and S3 tubular lumen and the bladder after a single approved dose.

# Figure 3 (bottom row): reabsorption and cumulative UGE

```{r fig3-reabs, fig.height = 5}
pbo5 <- solve_df(make_subject(99, "dapagliflozin", 0, times = seq(0, 120, by = 0.1)))
reabs <- typ |>
  dplyr::left_join(pbo5 |> dplyr::select(time, uge_pbo = uge), by = "time") |>
  dplyr::group_by(drug) |>
  dplyr::mutate(
    `Total reabsorption (% of max)` = 100 * reabs_total / max(reabs_total),
    `SGLT1 contribution (%)` = 100 * reabs_sglt1 / reabs_total,
    `SGLT2 contribution (%)` = 100 * reabs_sglt2 / reabs_total,
    `Placebo-normalized cumulative UGE (g)` = uge - uge_pbo
  ) |>
  dplyr::ungroup()
reabs |>
  dplyr::select(drug, time, dplyr::ends_with("(%)"), dplyr::ends_with("of max)"), dplyr::ends_with("(g)")) |>
  tidyr::pivot_longer(-c(drug, time)) |>
  ggplot(aes(time / 24, value, colour = drug)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~name, nrow = 1, scales = "free_y") +
  scale_colour_manual(values = drug_cols) +
  labs(x = "Time (days)", y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The Results report a 24-hour UGE of 92 g (dapagliflozin), 98 g (empagliflozin)
and 104 g (canagliflozin), and a placebo-normalized cumulative UGE over 0-5
days of 184, 200 and 157 g.

```{r uge-gate}
uge_tbl <- reabs |>
  dplyr::group_by(drug) |>
  dplyr::summarise(
    uge24 = uge[time == 24],
    uge_cum5 = `Placebo-normalized cumulative UGE (g)`[time == 120]
  ) |>
  dplyr::left_join(
    data.frame(drug = drugs, uge24_ref = c(92, 98, 104), uge_cum5_ref = c(184, 200, 157)),
    by = "drug"
  )
knitr::kable(
  uge_tbl |>
    dplyr::select(drug, uge24_ref, uge24, uge_cum5_ref, uge_cum5) |>
    dplyr::rename(
      "Drug" = drug, "24-h UGE paper (g)" = uge24_ref, "24-h UGE model (g)" = uge24,
      "0-5 d placebo-normalized UGE paper (g)" = uge_cum5_ref,
      "0-5 d placebo-normalized UGE model (g)" = uge_cum5
    ),
  digits = 1,
  caption = "24-hour and cumulative UGE after a single approved dose (Sokolov 2020 Results)."
)
stopifnot(
  all(abs(uge_tbl$uge24 - uge_tbl$uge24_ref) / uge_tbl$uge24_ref < 0.02),
  all(abs(uge_tbl$uge_cum5 - uge_tbl$uge_cum5_ref) / uge_tbl$uge_cum5_ref < 0.02)
)
```

# Figure 2: UGE dose response in T2DM

The deposited script simulates the 24-h UGE after a single dose. It does this
under two scenarios: the "in situ" scenario uses drug-specific median MPG and
eGFR (Figure 2a), and the "normalized" scenario uses the common medians
(Figure 2b).

```{r fig2}
doses <- c(0.1, 0.25, 0.5, 1, 2.5, 5, 10, 25, 50, 100, 300, 1000)
scen <- data.frame(
  scenario = c(rep("in situ", 3), rep("normalized", 3)),
  drug = rep(drugs, 2),
  glu = c(7.8, 9.24, 10.43, rep(9.33, 3)),
  egfr = c(98.1, 100, 100, rep(100, 3))
)
dr_events <- list()
id <- 0L
for (k in seq_len(nrow(scen))) {
  for (dz in doses) {
    id <- id + 1L
    dr_events[[id]] <- make_subject(id, scen$drug[k], dz,
      glu = scen$glu[k], egfr = scen$egfr[k], times = 24
    ) |>
      dplyr::mutate(scenario = scen$scenario[k])
  }
}
dr_events <- dplyr::bind_rows(dr_events)
dr <- as.data.frame(rxode2::rxSolve(mod, dr_events,
  keep = c("drug", "dose", "scenario"), returnType = "data.frame"
))

ggplot(dr, aes(dose, uge, colour = drug)) +
  geom_line(linewidth = 0.8) +
  geom_point() +
  facet_wrap(~scenario) +
  scale_x_log10() +
  scale_colour_manual(values = drug_cols) +
  coord_cartesian(ylim = c(0, 180)) +
  labs(x = "Dose (mg)", y = "24-h UGE (g/day)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Replicates Figure 2 of Sokolov 2020 (model curves only; the observed arm means
are in the unpublished analysis data set).

In the normalized scenario the Results state that the UGE is 82.3 and 91.4 g/day
for 5 and 10 mg dapagliflozin, 91.1 and 98.4 g/day for 10 and 25 mg
empagliflozin, and 86.4 and 104.3 g for 100 and 300 mg canagliflozin.

```{r fig2-gate}
uge24_pbo <- pbo$uge[pbo$time == 24]
dr_ref <- data.frame(
  drug = rep(drugs, each = 2),
  dose = c(5, 10, 10, 25, 100, 300),
  uge_ref = c(82.3, 91.4, 91.1, 98.4, 86.4, 104.3)
)
dr_cmp <- dr |>
  dplyr::filter(scenario == "normalized") |>
  dplyr::select(drug, dose, uge) |>
  dplyr::inner_join(dr_ref, by = c("drug", "dose")) |>
  dplyr::mutate(
    pct = 100 * (uge - uge_ref) / uge_ref,
    uge_net = uge - uge24_pbo,
    pct_net = 100 * (uge_net - uge_ref) / uge_ref
  )
knitr::kable(
  dr_cmp |>
    dplyr::select(drug, dose, uge_ref, uge, pct, uge_net, pct_net) |>
    dplyr::rename(
      "Drug" = drug, "Dose (mg)" = dose, "UGE paper (g/day)" = uge_ref,
      "Total UGE model (g/day)" = uge, "% diff (total)" = pct,
      "UGE minus drug-free baseline (g/day)" = uge_net, "% diff (net)" = pct_net
    ),
  digits = 2,
  caption = "Normalized-scenario 24-h UGE (Sokolov 2020 Results, Figure 2b)."
)
stopifnot(
  nrow(dr_cmp) == 6L,
  max(abs(dr_cmp$pct)) < 2,
  max(abs(dr_cmp$pct_net)) < 0.1
)
```

The total 24-h UGE is 0.6-0.7% above every value in the Results paragraph.
The offset equals the drug-free 24-h UGE (`r round(uge24_pbo, 2)` g). With
that baseline subtracted, all six values agree to within 0.1%. The paragraph
therefore appears to report UGE net of the drug-free baseline, although the
paper does not say so. The separate statement that the 24-h UGE after the
approved doses is 92, 98 and 104 g matches the **total** UGE (previous
section).

# Figure 4: SGLT1 exposure and contribution to UGE

Figure 4a compares the average S3 concentration over the 24 h after dosing
with the T2DM SGLT1 Ki. Figure 4b gives the share of 24-h UGE that is due to
SGLT1 inhibition. The script computes it by re-running the model with SGLT1
inhibition switched off (the code's `k_sglt1 = 1e8`, which multiplies every
SGLT1 Ki), and the packaged model does the same through the `ratio_sglt1_*`
parameters. The paper reports S3 exposure ratios of 0.5 (dapagliflozin 5 mg),
1.0 (dapagliflozin 10 mg), 1.7 (empagliflozin 25 mg) and 10 (canagliflozin
300 mg), and SGLT1 contributions of 1.59%, 2.41% and about 10%.

```{r fig4}
mod_no_sglt1 <- rxode2::ini(
  mod,
  ratio_sglt1_dapagliflozin = fixed(1157e8),
  ratio_sglt1_empagliflozin = fixed(1249e8),
  ratio_sglt1_canagliflozin = fixed(158e8)
)
f4_events <- dplyr::bind_rows(
  make_subject(1, "dapagliflozin", 5),
  make_subject(2, "dapagliflozin", 10),
  make_subject(3, "empagliflozin", 25),
  make_subject(4, "canagliflozin", 300)
)
f4_on <- solve_df(f4_events)
f4_off <- solve_df(f4_events, mod_no_sglt1)
ki1_named <- setNames(ki1_t2dm, drugs)
f4 <- f4_on |>
  dplyr::group_by(id, drug, dose) |>
  dplyr::summarise(
    cavg_s3 = mean(dplyr::case_when(
      drug == "dapagliflozin" ~ pst_dapagliflozin,
      drug == "empagliflozin" ~ pst_empagliflozin,
      TRUE ~ pst_canagliflozin
    ) / 0.01944 * 1e6),
    uge_on = uge[time == 24],
    .groups = "drop"
  ) |>
  dplyr::left_join(
    f4_off |> dplyr::filter(time == 24) |> dplyr::select(id, uge_off = uge),
    by = "id"
  ) |>
  dplyr::mutate(
    ratio_s3 = cavg_s3 / ki1_named[drug],
    sglt1_share = 100 * (uge_on - uge_off) / uge_on,
    ratio_ref = c(0.5, 1.0, 1.7, 10),
    share_ref = c(NA, 1.59, 2.41, 10)
  )
knitr::kable(
  f4 |>
    dplyr::select(drug, dose, ratio_ref, ratio_s3, share_ref, sglt1_share) |>
    dplyr::rename(
      "Drug" = drug, "Dose (mg)" = dose,
      "Cavg S3 / Ki SGLT1 paper" = ratio_ref, "Cavg S3 / Ki SGLT1 model" = ratio_s3,
      "SGLT1 share of UGE paper (%)" = share_ref, "SGLT1 share of UGE model (%)" = sglt1_share
    ),
  digits = 2,
  caption = "Replicates Figure 4 of Sokolov 2020."
)
stopifnot(
  # Exposure ratios are printed to 1-2 significant figures.
  all(abs(f4$ratio_s3 - f4$ratio_ref) / f4$ratio_ref < 0.06),
  # The two exact SGLT1 shares; the canagliflozin value is printed as ~10 %.
  all(abs(f4$sglt1_share[2:3] - f4$share_ref[2:3]) < 0.1),
  abs(f4$sglt1_share[4] - 10) < 1
)

ggplot(f4 |> dplyr::filter(dose != 5), aes(drug, sglt1_share, fill = drug)) +
  geom_col(colour = "black", show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f%%", sglt1_share)), vjust = -0.5) +
  scale_fill_manual(values = drug_cols) +
  labs(x = NULL, y = "SGLT1 contribution to 24-h UGE (%)") +
  theme_bw()
```

The S3 exposure ratios reproduce Figure 4a. The SGLT1 shares of 24-h UGE are
1.62% and 2.45% for dapagliflozin and empagliflozin, against 1.59% and 2.41%
printed. They are small differences between two nearly equal UGE values
(about 0.03 g of about 92 g), which the rounding of the deposited parameter
values can explain. Canagliflozin gives 10.8%, against "approximately 10%" in
the text and a bar labelled 10 in Figure 4b.

# PKNCA: plasma exposure after a single approved dose

```{r pknca}
nca_conc <- conc_long |>
  dplyr::filter(site == "Plasma", time <= 24) |>
  dplyr::mutate(id = match(drug, drugs), treatment = drug) |>
  dplyr::filter(!is.na(conc_nM))
nca_dose <- data.frame(
  id = seq_along(drugs), time = 0, treatment = drugs,
  dose_nmol = unname(approved[drugs] / mw[drugs] * 1e6)
)
conc_obj <- PKNCA::PKNCAconc(nca_conc, conc_nM ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(nca_dose, dose_nmol ~ time | treatment + id)
intervals <- data.frame(start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_ref <- data.frame(
  treatment = drugs,
  cmax = c(275, 533, 6156),
  cav = c(51.7, 229, 2109)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = nca_ref,
  by = "treatment",
  units = c(cmax = "nM", cav = "nM"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "PKNCA plasma Cmax and 0-24 h average concentration after a single",
    "approved dose versus Sokolov 2020 Table 1 (plasma rows).",
    "* marks a difference above 20%."
  )
)
nca_num <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cav")) |>
  dplyr::select(treatment, PPTESTCD, sim = PPORRES) |>
  dplyr::left_join(
    tidyr::pivot_longer(nca_ref, -treatment, names_to = "PPTESTCD", values_to = "ref"),
    by = c("treatment", "PPTESTCD")
  )
# One deterministic typical subject per drug; the only difference from the
# paper is the linear-trapezoid AUC versus the grid mean used for Table 1.
stopifnot(nrow(nca_num) == 6L, max(abs(nca_num$sim / nca_num$ref - 1)) < 0.02)
```

All three drugs reproduce Table 1 closely, and no row is flagged.

# Assumptions and deviations

* **Code over Table S3.** The deposited code has 5 (dapagliflozin),
  5 (empagliflozin) and 4 (canagliflozin) first-order transit states between
  the dosing state and plasma. Table S3 writes absorption directly from the
  dosing state and omits them. Table S3 also shows no T2DM factor on Ki and
  prints `Cana_lumen1` in the glucose `V_lumen` rate law. The packaged model
  follows the code throughout. With the code's structure, Table 1, the Figure 2
  UGE values and the Figure 4 SGLT1 shares all reproduce (see the gates
  above). The T2DM Ki in Table 1 equal the healthy Ki times `coef`, which
  confirms the factor.
* **Steady-state glucose start.** The code starts the tubular glucose states at
  zero and burns in for 72 h. The packaged model starts them at the analytic
  drug-free steady state, so a dose can be given at time 0. The steady-state
  gate above confirms the two are equivalent.
* **Plasma glucose is not a state.** Mean daily plasma glucose (`GLU`) is a
  per-subject input and is not lowered by the drug. This is the paper's own
  stated limitation: it considered only 24-h mean glucose and cumulative UGE.
* **eGFR used as absolute GFR.** GFR in L/h is `CRCL * 60 / 1000`, as in the
  deposited script. The BSA-normalized eGFR is not de-normalized.
* **Omitted bookkeeping.** The code's AUC accumulator states and its
  urine-reset events at fixed times are not carried. They are output
  bookkeeping for particular simulations. AUCs and interval UGE are computed
  from the output instead (e.g. `uge(t2) - uge(t1)`). The code's `k_sglt1`
  switch is reproduced by scaling the `ratio_sglt1_*` parameters (Figure 4
  section).
* **No variability.** The fit used mean data and the error-model estimates are
  not reported. The model therefore has no IIV and no residual error, and is
  intended for typical-value simulation.
* **Tubular fluid specimen.** The tubular-lumen states are recorded with
  specimen `urine` (glomerular filtrate on its way to the bladder), the
  closest term in the controlled vocabulary.
