---
title: "Andrographolide (Songvut 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Andrographolide (Songvut 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r helpers, include = FALSE}
# rxode2 emits "multi-subject simulation without without 'omega'" whenever an
# event table carries more than one ID and the model declares no random
# effects. This model has no random effects BY DESIGN (the source fitted a
# single typical-value curve to the pooled 12-patient cohort by nonlinear
# regression), so that warning is expected and not actionable. Muffle exactly
# that message and let every other warning through -- a blanket
# suppressWarnings() around the solve would also hide real problems.
solve_er <- function(ui, events) {
  withCallingHandlers(
    rxode2::rxSolve(ui, events = events, returnType = "data.frame"),
    warning = function(w) {
      if (grepl("multi-subject simulation without", conditionMessage(w))) {
        invokeRestart("muffleWarning")
      }
    }
  )
}

# Closed form of the encoded model, used to cross-check the rxode2 solve.
er_closed_form <- function(auc, e0, emax, auc50, hill) {
  e0 + emax * auc^hill / (auc50^hill + auc^hill)
}
```

# Model and source

```{r model}
mod <- readModelDb("Songvut_2026_andrographolide")
ui <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.3389/fphar.2026.1781740>
* Supplement: <https://www.frontiersin.org/articles/10.3389/fphar.2026.1781740/full#supplementary-material>
  (Figures S1-S3 only: HPLC chromatograms, diterpenoid chemical structures, MS
  chromatograms. The supplement contains no model parameters.)

Description: `r ui$description`

# What this paper contributes, and what it does not

Songvut 2026 is an open-label, single-centre, single-arm pharmacokinetic /
pharmacodynamic and safety study of standardized *Andrographis paniculata*
aqueous extract capsules in 12 Thai adults with mild COVID-19. It reports two
distinct analyses, and only one of them is a fitted model:

* **The pharmacokinetics are non-compartmental only.** Table 2 reports Cmax,
  Tmax, AUC(0-4 h), AUC(0-inf), MRT, Vd/F, Cl/F and half-life for each of the
  four major diterpenoids, computed in PK Solutions 2.0 by the linear
  trapezoidal rule (paper Section 2.3.6). `Cl/F`, `Vd/F` and `t1/2` there are
  NCA-derived summary statistics, not a fitted structural model. There is no
  population PK model in the paper, and therefore none in this package: nothing
  in `Songvut_2026_andrographolide` produces a concentration-time profile.
* **The exposure-response relationship is a fitted model**, and it is what this
  package reproduces: an exploratory sigmoidal Emax curve relating each
  patient's day-5 andrographolide AUC(0-4 h) to their log10 RdRp viral-load
  reduction between day 1 and day 5, fitted by unweighted nonlinear regression
  in GraphPad Prism 9.3.0 (paper Sections 2.4 and 3.3.2, Figure 5).

Because there is no PK layer, this vignette has no PKNCA section: there is no
concentration-time profile to integrate and no simulated NCA parameter to
compare against Table 2. The validation instead follows the pattern used for
mechanistic and algebraic models in `references/endogenous-validation.md` --
closed-form agreement, anchor reproduction, monotonicity and boundedness, and a
goodness-of-fit round trip against the paper's own reported `r2`. The
published NCA table is transcribed below for context, clearly marked as
reference-only.

The paper is emphatic, and this vignette repeats it, that the exposure-response
analysis is **exploratory and hypothesis-generating, not confirmatory**. There
was no placebo arm, so the observed viral-load decline overlaps the expected
window of natural clearance in vaccinated patients with mild COVID-19 and
cannot be attributed to the extract. With n = 12 at a single dose level and a
narrow exposure range, the paper could not estimate a confidence interval for
any parameter.

# Population

Twelve adults (10 female, 2 male; 83.33% female) of Thai nationality, aged
18-60 years by inclusion criterion with a mean age of 34.08 +/- 9.29 years and
a mean body mass index of 21.07 +/- 2.19 kg/m^2, all with RT-PCR-confirmed
SARS-CoV-2 infection within 72 h of symptom onset and classified as mild
COVID-19 under the WHO 2020 clinical-management guidance -- symptomatic without
viral pneumonia or hypoxia, resting SpO2 at least 95%, normal chest radiograph
(paper Tables 3-5 and Section 2.3.1). All were vaccinated, with the most recent
dose more than one year before enrollment, and none had significant
comorbidities. Body weight is not reported; only BMI is, so no weight-based
scaling is possible or needed.

Each participant received capsules equivalent to a labelled 30 mg of
andrographolide (three capsules of 10 mg labelled content; measured content
10.28 +/- 0.03 mg per capsule, paper Table 1) every 8 h for 5 consecutive days,
i.e. 90 mg/day, alongside as-needed symptomatic standard of care. Intensive
0-4 h plasma sampling was performed after the first dose on day 1 and again
after the morning dose on day 5. All 12 completed the protocol with 100%
compliance, no dropouts and no protocol deviations, so all 12 contribute to the
exposure-response fit.

The same information is available programmatically:

```{r population}
str(readModelDb("Songvut_2026_andrographolide")()$population)
```

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry
in `inst/modeldb/specificDrugs/Songvut_2026_andrographolide.R`. Collected here:

| Equation / parameter | Value | Source location |
|---|---|---|
| Sigmoidal Emax equation | `Effect = Emax * AUC^gamma / (EAUC50^gamma + AUC^gamma)` | Section 2.4 "Statistical analysis", displayed equation |
| Baseline term `e0` added to the printed equation | 1.81 | **Not printed.** Recovered from Figure 5; see "The printed equation versus Figure 5" below |
| `lemax` = log(2.24) | 2.24 -log10 copies/uL | Section 3.3.2: "Emax = 2.24 log10 copies/mL" |
| `lauc50` = log(29.80) | 29.80 ug*h/L | Section 3.3.2: "EAUC50 = 29.80 ug h/L"; corroborated in Section 4.3 |
| `lhill` = log(8) | 8 (unitless) | Section 3.3.2: "The Hill coefficient was estimated to be approximately 8 and approached the upper boundary of the constrained parameter range" |
| `addSd` = 1.245 | 1.245 -log10 copies/uL | **Derived**, not printed: from `r2` = 0.4 (Section 3.3.2) and the 12 log10 reductions in Table 3, using ordinary least squares (Section 2.4: "Nonlinear regression was performed without weighting") |
| Reported goodness of fit | `r2` = 0.4 | Section 3.3.2 |
| `AUC_ANDRO` covariate | day-5 AUC(0-4 h), mean 30.12 +/- 15.83 ug*h/L | Table 2, andrographolide (AP1) block; Figure 5 x-axis |
| Per-subject response values | 12 log10 reductions, 0.20 to 5.10 | Table 3, "Log10 reduction (between day 1 and 5)" column |
| No between-subject variability | n/a | Section 2.4 -- a single curve fitted by nonlinear regression to the pooled cohort; no hierarchical model |
| No covariate effects | n/a | The paper screened no covariates on the exposure-response model |

# The printed equation versus Figure 5

This is the one place where the packaged model deliberately departs from the
paper's printed equation, so the reasoning is set out in full.

The equation printed in Section 2.4 is a **three-parameter** sigmoid with no
baseline term:

```
Effect = Emax * AUC^gamma / (EAUC50^gamma + AUC^gamma)
```

Under that form, with the printed `Emax` = 2.24, `EAUC50` = 29.80 and
`gamma` = 8, the curve would run from 0 at zero exposure to 2.24 at saturation,
and the half-maximal point would sit at y = 1.12. Figure 5 shows something
else: a lower asymptote near 1.85, a plateau near 4.05 carrying the "Emax"
annotation, and the dashed `EAUC50` guide line drawn at y near 3.0.

Three independent pieces of the paper's own evidence adjudicate this, and all
three point the same way -- the fitted curve carries a nonzero baseline:

1. **The plateau.** `e0 + emax` = 1.81 + 2.24 = 4.05 reproduces the figure's
   plateau, the point the figure labels "Emax".
2. **The guide line.** `e0 + emax/2` = 2.93 reproduces where the dashed
   `EAUC50` line is drawn (y near 3.0), rather than the 1.12 the printed form
   requires.
3. **The reported goodness of fit.** Scored against the paper's own 12
   response values (Table 3) and the digitised Figure 5 exposures, the printed
   three-parameter form gives `r2` = -0.90 -- worse than predicting the mean --
   while the four-parameter form gives `r2` = +0.37, which is the paper's
   reported 0.4. The printed form cannot reproduce the paper's own stated fit.

The most likely explanation is typesetting: GraphPad Prism's standard sigmoidal
dose-response models are parameterised with `Bottom` and `Top`, so the reported
"Emax = 2.24" is the `Top - Bottom` **increment** rather than the absolute
plateau, and the displayed equation dropped the `Bottom` term.

The packaged model therefore encodes the four-parameter form

```
reduction = e0 + emax * AUC^hill / (auc50^hill + AUC^hill)
```

with `e0` = 1.81 recovered from Figure 5. This resolution was ratified by the
operator (see Assumptions and deviations). The `e0` term is not a fudge factor:
in an uncontrolled single-arm study it is exactly the quantity the design cannot
separate from drug effect -- spontaneous viral clearance -- which is why the
paper's own Discussion stresses that the decline cannot be attributed to the
extract.

# Digitised Figure 5 data

The 12 response values are printed exactly in Table 3. The 12 matching
exposures are not tabulated per subject anywhere in the paper, only plotted, so
they were digitised from Figure 5.

```{r digitised}
fig5 <- tibble::tibble(
  # x: per-subject day-5 AUC(0-4 h), digitised from Figure 5 (ug*h/L).
  AUC_ANDRO = c(7.5, 11.5, 16.0, 14.0, 23.0, 30.0, 36.0, 39.5, 40.0, 46.0, 43.5, 57.0),
  # y: per-subject log10 viral-load reduction, EXACT from paper Table 3.
  observed  = c(2.12, 4.83, 1.31, 0.40, 0.20, 3.09, 4.01, 5.10, 4.31, 4.05, 2.31, 3.84)
)
stopifnot(nrow(fig5) == 12L)
```

The digitisation is validated independently of anything it is later used for:
the paper reports the cohort's day-5 AUC(0-4 h) as 30.12 +/- 15.83 ug*h/L in
Table 2, a number that played no part in reading the figure.

```{r gate-digitisation}
auc_mean <- mean(fig5$AUC_ANDRO)
auc_sd <- sd(fig5$AUC_ANDRO)
c(mean = auc_mean, sd = auc_sd)

# Table 2, andrographolide (AP1), 90 mg/day column: 30.12 +/- 15.83 ug*h/L.
stopifnot(
  abs(auc_mean - 30.12) / 30.12 < 0.02,
  abs(auc_sd - 15.83) / 15.83 < 0.05
)

# The response values must be Table 3's, not a re-reading of the figure: their
# arithmetic mean is the paper's reported day-5 reduction of 2.96 log10.
stopifnot(abs(mean(fig5$observed) - 2.96) < 0.01)
```

Both the centre and the spread of the digitised exposures land on the
independently published values, which is a strong check that the x-axis was read
correctly. Note the caveats: Table 3 lists two nearly identical responses
(4.01 and 4.05) whose assignment between the x = 36 and x = 46 points cannot be
resolved from the figure, an ambiguity of 0.04 log10 that moves nothing; and
individual digitised exposures carry roughly +/- 1 ug*h/L of reading error.

# Simulation

The model is algebraic and cross-sectional -- no ODE states, no dosing events,
no time dependence -- so an event table is observation rows only, each carrying
that subject's `AUC_ANDRO`.

```{r simulate-cohort}
events_cohort <- fig5 |>
  mutate(id = row_number(), time = 0, evid = 0L) |>
  select(id, time, evid, AUC_ANDRO)

# Join on `id`, not on AUC_ANDRO: joining on a floating-point covariate that
# has round-tripped through the solver is a needless footgun.
sim_cohort <- solve_er(ui, events_cohort) |>
  select(id, predicted = viralLoadReduction) |>
  left_join(mutate(fig5, id = row_number()), by = "id")

stopifnot(nrow(sim_cohort) == 12L, !anyNA(sim_cohort$predicted))
```

A dense exposure grid gives the fitted curve. It stays inside the 200-per-arm
cohort cap.

```{r simulate-curve}
grid_auc <- seq(0, 60, length.out = 121)
stopifnot(length(grid_auc) <= 200L)

curve <- solve_er(
  ui,
  data.frame(id = seq_along(grid_auc), time = 0, evid = 0L, AUC_ANDRO = grid_auc)
) |>
  select(AUC_ANDRO, predicted = viralLoadReduction)
```

# Validation

## Gate 1 -- rxode2 solve equals the closed form exactly

Both sides use the same parameter draw, so the only difference is numerical.
A tight bound is correct here.

```{r gate-closed-form}
pars <- ui$theta
e0 <- exp(pars[["le0"]])
emax <- exp(pars[["lemax"]])
auc50 <- exp(pars[["lauc50"]])
hill <- exp(pars[["lhill"]])
c(e0 = e0, emax = emax, auc50 = auc50, hill = hill)

analytic <- er_closed_form(curve$AUC_ANDRO, e0, emax, auc50, hill)
max_abs_err <- max(abs(curve$predicted - analytic))
max_abs_err

stopifnot(length(analytic) == 121L, max_abs_err < 1e-8)
```

## Gate 2 -- the three Figure 5 anchors

The lower asymptote, the dashed `EAUC50` guide line and the labelled plateau
are three separately readable features of Figure 5. All three must fall out of
the encoded parameters.

```{r gate-anchors}
anchors <- solve_er(
  ui,
  data.frame(
    id = 1:3, time = 0, evid = 0L,
    AUC_ANDRO = c(0, auc50, 1e6)  # zero exposure, half-maximal, saturating
  )
) |>
  transmute(
    anchor = c("lower asymptote (AUC = 0)", "at EAUC50", "plateau (saturating)"),
    predicted = viralLoadReduction,
    figure5 = c(1.81, 2.93, 4.05)
  )

anchors |>
  rename("Figure 5 feature" = anchor,
         "Model" = predicted,
         "Read from Figure 5" = figure5) |>
  knitr::kable(digits = 4, caption = "Gate 2: the encoded curve reproduces all three readable features of Figure 5.")

stopifnot(max(abs(anchors$predicted - anchors$figure5)) < 1e-6)
```

## Gate 3 -- monotone and bounded over the observed exposure range

A sigmoidal Emax curve with a positive Hill coefficient must increase
monotonically and stay within `[e0, e0 + emax]`. This catches a sign error or a
mistyped exponent that the anchor gate alone would not.

```{r gate-shape}
stopifnot(
  all(diff(curve$predicted) > 0),
  all(curve$predicted >= e0 - 1e-9),
  all(curve$predicted <= e0 + emax + 1e-9),
  # Hill = 8 makes the transition sharp: essentially all of the increment is
  # spent between half and double the EAUC50.
  er_closed_form(auc50 / 2, e0, emax, auc50, hill) < e0 + 0.01 * emax,
  er_closed_form(auc50 * 2, e0, emax, auc50, hill) > e0 + 0.99 * emax
)
```

## Gate 4 -- the encoded model reproduces the paper's reported r2

The paper reports `r2` = 0.4 (Section 3.3.2), to one significant figure. The
encoded model, scored against the paper's own 12 response values, must land
there -- and the printed three-parameter alternative must not.

```{r gate-r2}
sst <- sum((sim_cohort$observed - mean(sim_cohort$observed))^2)
r2_of <- function(pred) 1 - sum((sim_cohort$observed - pred)^2) / sst

r2_encoded <- r2_of(sim_cohort$predicted)
r2_printed <- r2_of(er_closed_form(sim_cohort$AUC_ANDRO, 0, emax, auc50, hill))

tibble::tibble(
  form = c("Encoded: 4-parameter, e0 = 1.81",
           "Printed Section 2.4: 3-parameter, no baseline"),
  r2 = c(r2_encoded, r2_printed)
) |>
  rename("Model form" = form, "r2 vs the paper's 12 responses" = r2) |>
  knitr::kable(digits = 3, caption = "Gate 4: only the encoded four-parameter form reproduces the paper's reported r2 of 0.4.")

stopifnot(
  # Rounds to the reported 0.4.
  abs(r2_encoded - 0.4) < 0.05,
  # The printed form is worse than predicting the mean.
  r2_printed < 0
)
```

## Gate 5 -- the encoded residual SD round-trips

`addSd` was derived from the reported `r2` = 0.4 and Table 3's responses, so
this gate is not circular: it asks whether the SD implied by the *encoded
model's actual residuals* agrees with the value derived from the *reported*
fit. A structural transcription error would separate the two.

```{r gate-residual}
rmse_actual <- sqrt(mean((sim_cohort$observed - sim_cohort$predicted)^2))
addSd_encoded <- pars[["addSd"]]
c(rmse_actual = rmse_actual, addSd_encoded = addSd_encoded,
  pct_diff = 100 * (rmse_actual - addSd_encoded) / addSd_encoded)

stopifnot(abs(rmse_actual - addSd_encoded) / addSd_encoded < 0.05)
```

The residual error is large relative to the effect it sits on -- an SD of about
1.25 log10 against a total modelled range of 2.24 log10. That is the honest
picture of an `r2` = 0.4 fit to 12 patients, and it is why the paper declines
to treat these parameters as confirmatory.

The stochastic layer is exercised once to confirm the error model is wired up
and that the residual band brackets the observations.

```{r gate-residual-sim}
set.seed(20260901)
rxode2::rxSetSeed(20260901)

rep_events <- tidyr::expand_grid(replicate = 1:50, fig5) |>
  mutate(id = row_number(), time = 0, evid = 0L) |>
  select(id, time, evid, AUC_ANDRO)

rep_sim <- solve_er(ui, rep_events)
sd_sim <- sd(rep_sim$sim - rep_sim$viralLoadReduction)
c(addSd_encoded = addSd_encoded, sd_of_simulated_residuals = sd_sim)

stopifnot(
  nrow(rep_sim) == 600L,
  # 600 draws recover the encoded SD to well within Monte Carlo error.
  abs(sd_sim - addSd_encoded) / addSd_encoded < 0.10
)
```

## Gate 6 -- the paper's "AUC approached EAUC50" claim

Sections 3.3.2 and 4.3 both state that the day-5 mean AUC(0-4 h) of
30.12 ug*h/L approached the estimated `EAUC50` of 29.80 ug*h/L, i.e. that the
regimen sat near the steep middle of the curve. That is a checkable claim about
the encoded model, not just about two numbers.

```{r gate-eauc50-claim}
at_cohort_mean <- er_closed_form(30.12, e0, emax, auc50, hill)
half_maximal <- e0 + emax / 2
frac_of_increment <- (at_cohort_mean - e0) / emax
c(exposure_excess_over_auc50 = 30.12 / auc50 - 1,
  at_cohort_mean = at_cohort_mean, half_maximal = half_maximal,
  fraction_of_max_increment = frac_of_increment)

stopifnot(
  # The cohort mean exposure sits 1.07% above auc50.
  abs(30.12 / auc50 - 1) < 0.02,
  # Which puts the response within 2.2 percentage points of half-maximal.
  abs(frac_of_increment - 0.5) < 0.025,
  abs(at_cohort_mean - half_maximal) / half_maximal < 0.02
)
```

Note how the steep Hill coefficient amplifies: the cohort mean exposure is only
1.07% above `auc50`, but that becomes a 2.1% excess in the effect increment.
With `hill` = 8 the curve is so sharp near `auc50` that small exposure
differences translate into disproportionate predicted-response differences --
another reason to treat the fitted steepness as poorly identified rather than
mechanistic.

## Gate 7 -- dimensional analysis

`AUC_ANDRO` and `auc50` must share units so the sigmoid is dimensionless, and
`e0`, `emax` and `addSd` must all be on the log10-reduction scale. Scaling both
exposure terms by any common factor must leave the prediction unchanged.

```{r gate-dimensions}
str(ui$units)

# Exposure enters only as AUC/auc50 ratios, so a common rescaling is a no-op.
scaled <- er_closed_form(fig5$AUC_ANDRO * 1000, e0, emax, auc50 * 1000, hill)
stopifnot(max(abs(scaled - sim_cohort$predicted)) < 1e-9)
```

# Replicate Figure 5

```{r figure-5, fig.width = 6.5, fig.height = 4.5}
# Replicates Figure 5 of Songvut 2026: exploratory exposure-effect relationship
# between day-5 andrographolide AUC(0-4 h) and log10 RdRp viral-load reduction.
ggplot(curve, aes(AUC_ANDRO, predicted)) +
  geom_hline(yintercept = e0 + emax, linetype = "dotted", colour = "grey40") +
  geom_segment(
    aes(x = 0, xend = auc50, y = half_maximal, yend = half_maximal),
    linetype = "dashed", colour = "grey40"
  ) +
  geom_segment(
    aes(x = auc50, xend = auc50, y = 0, yend = half_maximal),
    linetype = "dashed", colour = "grey40"
  ) +
  geom_line(linewidth = 1.2, colour = "#2C7FB8") +
  geom_point(data = fig5, aes(AUC_ANDRO, observed), shape = 4, size = 2.5) +
  annotate("text", x = 58, y = e0 + emax + 0.18, label = "Emax",
           colour = "#2C7FB8", hjust = 1) +
  annotate("text", x = auc50, y = -0.25, label = "EAUC50", size = 3) +
  coord_cartesian(xlim = c(0, 60), ylim = c(0, 6), clip = "off") +
  labs(
    x = "AUC(0-4 h, day 5) (ug*h/L)",
    y = "Decrease in log10 viral load\n(-log10 copies/uL)",
    title = "Figure 5 - exposure-effect relationship",
    caption = paste(
      "Replicates Figure 5 of Songvut 2026. Crosses are the 12 patients:",
      "responses exact from Table 3, exposures digitised from the figure."
    )
  ) +
  theme_bw()
```

The reproduced panel matches the published figure feature for feature: the
nonzero lower asymptote, the plateau annotated `Emax`, and the dashed guide
lines meeting the curve at `EAUC50` on the x-axis and near y = 3 on the y-axis.

# Observed versus predicted

```{r obs-pred}
sim_cohort |>
  arrange(AUC_ANDRO) |>
  mutate(residual = observed - predicted) |>
  select(AUC_ANDRO, observed, predicted, residual) |>
  rename(
    "AUC(0-4 h, day 5) (ug*h/L)" = AUC_ANDRO,
    "Observed reduction (log10)" = observed,
    "Model-predicted reduction (log10)" = predicted,
    "Residual (log10)" = residual
  ) |>
  knitr::kable(digits = 3, caption = "Per-patient observed versus predicted log10 viral-load reduction. Observed values are exact from Table 3; exposures are digitised from Figure 5.")
```

The two largest residuals are the patient at 11.5 ug*h/L who cleared 4.83 log10
and the patient at 23.0 ug*h/L who cleared only 0.20 log10 -- both at the low
end of the exposure range. This is the pattern the paper itself flags in
Section 4.3: "greater variability in viral load reduction was observed at lower
AUC levels, indicating limitations of the model in lower exposure range."

# Published non-compartmental parameters (reference only)

Transcribed from Table 2, andrographolide (AP1) block. **These are not gated**
against anything: the paper reports no PK model, so this package cannot
simulate a concentration-time profile to compare them with. They are here so a
downstream user assembling a PK layer for `AUC_ANDRO` has the published targets
in one place.

```{r published-nca}
tibble::tribble(
  ~parameter,             ~single_dose,     ~multiple_dose,
  "Cmax (ug/L)",          "18.63 +/- 11.81", "17.82 +/- 10.56",
  "Tmax (h)",             "0.75 [0.31]",     "0.75 [0.5]",
  "AUC(0-4 h) (ug*h/L)",  "28.54 +/- 8.72",  "30.12 +/- 15.83",
  "AUC(0-inf) (ug*h/L)",  "32.81 +/- 8.91",  "not reported",
  "MRT (h)",              "2.15 +/- 0.69",   "4.19 +/- 3.88",
  "Vd/F (L/kg)",          "27.47 +/- 13.55", "41.02 +/- 32.22",
  "Cl/F (L/h/kg)",        "14.48 +/- 3.66",  "15.05 +/- 11.67",
  "Half-life (h)",        "1.21 [0.19]",     "1.48 [1.99]"
) |>
  rename(
    "NCA parameter" = parameter,
    "Single dose, 30 mg (n = 12)" = single_dose,
    "Multiple dose, 90 mg/day (n = 12)" = multiple_dose
  ) |>
  knitr::kable(caption = "Songvut 2026 Table 2, andrographolide (AP1). Mean +/- SD, except Tmax and half-life as median [IQR]. Reference only - not reproducible from this package, which contains no PK model.")
```

# Assumptions and deviations

* **The encoded model departs from the printed equation by adding a baseline
  term `e0`.** The paper's Section 2.4 prints a three-parameter sigmoid with no
  baseline. The packaged model is the four-parameter form
  `e0 + emax * AUC^hill / (auc50^hill + AUC^hill)`. This was escalated to the
  operator rather than decided silently, because recovering a value from a
  figure to *fill a reporting gap* is routine while using one to *override a
  printed structural claim* is not, and this act does both. It was ratified on
  2026-08-31 (task `oare_PMC13022710`, sidecar request-001, option B) on the
  evidence set out in "The printed equation versus Figure 5" above: the printed
  form cannot reproduce the paper's own reported `r2`, and the four-parameter
  form additionally reproduces two further independent Figure 5 features
  (plateau, guide line) that the printed form cannot explain. Gate 4 above is
  the load-bearing check.
* **Non-paper-derived parameter value: `e0` = 1.81** is not printed anywhere in
  the paper. It was recovered as the least-squares baseline with `emax`,
  `auc50` and `hill` held at their printed values, fitted to Table 3's
  responses against the digitised Figure 5 exposures (recovered 1.809), and it
  is also the value read directly off the Figure 5 lower asymptote. Treat it as
  carrying the uncertainty of a figure reading, roughly +/- 0.05 log10.
* **Derived parameter value: `addSd` = 1.245** is not printed either, but unlike
  `e0` it needs no figure: it follows from two printed quantities, since
  unweighted nonlinear regression (Section 2.4) means ordinary least squares and
  therefore `r2 = 1 - SSE/SST`, with `SST` computable exactly from Table 3's 12
  responses. The `n` denominator matches nlmixr2's maximum-likelihood SD
  parameterisation; the degrees-of-freedom-corrected alternative with four
  estimated parameters would be 1.52. Because `r2` is printed to one
  significant figure, this SD is only determined to about +/- 0.05.
* **Per-subject exposures are digitised from Figure 5**, since the paper
  tabulates the responses (Table 3) but not the matching per-subject AUC values.
  The digitisation is validated against Table 2's independently reported
  30.12 +/- 15.83 ug*h/L (Gate above). Individual values carry roughly
  +/- 1 ug*h/L of reading error, and the assignment of Table 3's 4.01 versus
  4.05 between the two right-hand points cannot be resolved from the figure --
  a 0.04 log10 ambiguity that changes no conclusion. The digitised exposures are
  used only for validation in this vignette; no packaged parameter depends on an
  individual digitised value.
* **The Hill coefficient is at a boundary.** The paper states it "was estimated
  to be approximately 8 and approached the upper boundary of the constrained
  parameter range", but does not report the bound. It is encoded as 8 and is
  *not* wrapped in `fixed()`, because it was estimated rather than held; but the
  steepness it implies is poorly identified and should not be read as a
  mechanistic finding. Gate 3 shows how sharp it makes the transition: with
  `hill` = 8, essentially the whole effect increment is spent between half and
  double the `EAUC50`.
* **No between-subject variability and no covariate effects are encoded**,
  because the source has none. The paper fitted a single typical-value curve to
  the pooled 12-patient cohort by nonlinear regression, with no hierarchical
  structure, and screened no covariates on the exposure-response model. No
  variance components were invented to fill the gap.
* **Units of `Emax`.** Section 3.3.2 gives `Emax` in "log10 copies/mL" while
  Table 3 tabulates viral loads and log10 reductions in copies/uL. A log10 ratio
  is dimensionless, so the value is unaffected; the model file and this vignette
  use the Table 3 convention (`-log10 copies/uL`).
* **Convention-lint deviation: the observation variable is
  `viralLoadReduction`.** `checkModelConventions()` warns that this is not a
  registered single-output canonical. The registered `viralLoad` name is not
  appropriate -- the endpoint is a *change from baseline* between two days, not
  a viral load, and the model has no baseline load from which one could be
  reconstructed (per-subject baselines in Table 3 span 541 to 4.4e6 copies/uL).
  The warning is accepted rather than papered over, following the same shape as
  other change-from-baseline single-output models already in the library
  (`Morganroth_2015_moxifloxacin.R` with `DDQTcF`,
  `Gong_2023_pemigatinib_creatinine.R` with `creatPctChange`). Promoting a
  `viralLoadReduction` canonical is left to the operator.
* **No PK layer, and therefore no PKNCA validation.** See "What this paper
  contributes, and what it does not". Downstream users must supply
  `AUC_ANDRO` from observed concentrations or from an external andrographolide
  PK model; none exists for this formulation in the published literature.
* **No erratum applies.** No correction notice was found for this article.
* **The supplement contains no model parameters** (Figures S1-S3 only), so
  nothing in the model file derives from it.
* **Exploratory status.** Every parameter here comes from an uncontrolled,
  single-arm, single-dose-level fit to 12 patients for which the paper could not
  estimate confidence intervals. The model reproduces what was published; it
  does not establish that andrographolide exposure causes viral-load reduction,
  and the paper explicitly declines that inference.
