---
title: "Linezolid (Soraluce 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linezolid (Soraluce 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Soraluce_2020_linezolid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Soraluce_2020_linezolid"))$description`
* Article: [Pharmaceutics. 2020;12(1):54](https://doi.org/10.3390/pharmaceutics12010054) (open access)

# Population

Forty critically ill adults treated with linezolid for a suspected Gram-positive
infection in the intensive care units of three Spanish university hospitals
(Araba, Doce de Octubre and Joan XXIII). Twenty-three received continuous renal
replacement therapy (CRRT; 18 continuous venovenous hemodiafiltration, 5
continuous venovenous hemodialysis) and 17 did not. Infection sources were
pulmonary (14), abdominal (10), neurological (9), biliary (2) and other (5).
Of the 40 patients, 29 were men. Median age was 72 years (range 22-85) without
CRRT and 68 (37-79) with CRRT, and median weight 71 kg (60-95) and 74 kg
(55-110) respectively (Table 1).

Renal function was measured, not estimated: creatinine clearance came from a
10-hour urine collection, `Clcr = (Cru * Vu) / (Crp * 600 min)`. Median Clcr was
71.2 mL/min (11.0-179.5) without CRRT and 6.0 mL/min (0.0-45.6) with CRRT. For
CRRT patients the extracorporeal clearance `CLEC = Sc * Qef` was computed per
patient from the measured sieving coefficient and the effluent flow: median
2.51 L/h (0.79-3.09).

All patients received 600 mg every 12 h as a 30-min intravenous infusion (one
patient 60 min) and were sampled at steady state after a mean of 8 doses: 311
plasma concentrations in total. An external validation cohort of 11 further
patients, none on CRRT (median Clcr 111 mL/min, range 45-240; Table 2),
received a 600 mg loading dose followed by 50 mg/h by continuous infusion.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Soraluce_2020_linezolid"))$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source in
`inst/modeldb/specificDrugs/Soraluce_2020_linezolid.R`; the table collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `CL = (CLNR + CLR * (Clcr/44) + Sc * Qef) * exp(eta1)` | n/a | Table 3 footnote b (final model) |
| `V1 = 16.2 * exp(eta2)` | n/a | Table 3 footnote d |
| Two-compartment, linear elimination | n/a | Results 3.1.1 |
| `CLEC = Sc * Qef`, per-patient data column (`QEFF`), CRRT only | n/a | Section 2.1; Section 2.3.2; Table 3 footnote c |
| `Clcr` from 10-h urine collection | n/a | Section 2.1 |
| `lcl_nonren` (CLNR) | 2.62 L/h (RSE 18%) | Table 3, final model |
| `lcl_renal` (theta in CLR) | 4.35 L/h at Clcr = 44 mL/min (RSE 19%) | Table 3, final model |
| `lvc` (V1) | 16.2 L (RSE 14%) | Table 3, final model |
| `lq` (Q) | 71.7 L/h (RSE 14%) | Table 3, final model |
| `lvp` (V2) | 29.0 L (RSE 7%) | Table 3, final model |
| `etalcl` | 61.5% -> omega^2 = log(0.615^2 + 1) = 0.3205 | Table 3, final model |
| `etalvc` | 65.9% -> omega^2 = log(0.659^2 + 1) = 0.3606 | Table 3, final model |
| No CL-V1 covariance | n/a | Results 3.1.1 |
| `addSd` | 0.266 mg/L (RSE 24%) | Table 3, final model |
| `propSd` | 0.159 (RSE 19%) | Table 3, final model |

# Virtual cohort

The observed data are not public. Two arms of 200 subjects each reproduce the
two strata of Table 1 at the standard 600 mg every 12 h, 30-min infusion
regimen. Covariates are drawn from log-normal distributions whose medians are
the Table 1 medians, truncated to the Table 1 ranges (the paper gives only
median and range):

* no CRRT: `CRCL` median 71.2 mL/min, truncated to 11.0-179.5; `QEFF = 0`;
* CRRT: `CRCL` median 6.0 mL/min, truncated to 0-45.6; `QEFF` median 2.51 L/h,
  truncated to 0.79-3.09.

```{r cohort}
# rxSetSeed() fixes the draw within an rxode2 build but not across builds or
# solver-thread counts; every assertion below is on a closed form or on
# quantities computed from the same drawn parameters, never on a cohort
# extreme.
rxode2::rxSetSeed(20200109)
set.seed(20200109)

N_PER_ARM <- 200L
DOSE <- 600 # mg
TAU <- 12 # h
TINF <- 0.5 # h

rtlnorm <- function(n, median, gsd, lo, hi) {
  pmin(pmax(exp(rnorm(n, log(median), log(gsd))), lo), hi)
}

cohort <- dplyr::bind_rows(
  tibble::tibble(
    id = seq_len(N_PER_ARM),
    arm = "No CRRT",
    CRCL = rtlnorm(N_PER_ARM, 71.2, 1.8, 11.0, 179.5),
    QEFF = 0
  ),
  tibble::tibble(
    id = N_PER_ARM + seq_len(N_PER_ARM),
    arm = "CRRT",
    CRCL = rtlnorm(N_PER_ARM, 6.0, 2.5, 0, 45.6),
    QEFF = rtlnorm(N_PER_ARM, 2.51, 1.4, 0.79, 3.09)
  )
)

cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    CRCL_median = median(CRCL),
    CRCL_min = min(CRCL),
    CRCL_max = max(CRCL),
    QEFF_median = median(QEFF),
    .groups = "drop"
  )

# One steady-state dosing interval per subject, observed on the central state.
# The grid is dense over the first 2 h: subjects in the upper tail of the
# clearance distribution (CL up to ~100 L/h) have a steep post-infusion
# decline, and a coarse grid there leaves a few percent of trapezoidal error
# in the AUC check below.
obs_times <- sort(unique(round(c(seq(0, 2, by = 0.02), seq(2, TAU, by = 0.25)), 6)))
events <- rxode2::et(amt = DOSE, dur = TINF, ii = TAU, ss = 1, cmt = "central") |>
  rxode2::et(obs_times, cmt = "central") |>
  rxode2::et(id = cohort$id) |>
  as.data.frame() |>
  dplyr::left_join(cohort, by = "id")
```

# Simulation

```{r simulate}
mod <- readModelDb("Soraluce_2020_linezolid")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("arm", "CRCL", "QEFF"),
  rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12
) |>
  as.data.frame()
```

## Steady-state profiles (compare Figure 2)

Figure 2 of the paper is a VPC of the observed steady-state interval, 0-12 h
after dose, with the observed median peaking near 20 mg/L at the end of the
infusion and falling to about 3-5 mg/L at 12 h. The plot below shows the
simulated median and 2.5th-97.5th percentile band of the individual
predictions for each arm.

```{r figure-2}
vpc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    p025 = quantile(Cc, 0.025),
    p50 = median(Cc),
    p975 = quantile(Cc, 0.975),
    .groups = "drop"
  )

ggplot(vpc, aes(time, p50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = p025, ymax = p975), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(
    x = "Time after dose (h)", y = "Linezolid (mg/L)", colour = NULL, fill = NULL,
    title = "Steady state, 600 mg q12h as a 30-min infusion",
    caption = paste(
      "Compare Figure 2 of Soraluce 2020.", N_PER_ARM,
      "simulated subjects per arm; median and 2.5th-97.5th percentiles."
    )
  )
```

# PKNCA validation

NCA over the steady-state interval, grouped by arm. At steady state the
area under the curve over one dosing interval equals `Dose / CL` exactly, and
the simulation reports each subject's drawn `cl`, so the NCA AUC can be checked
against the closed form subject by subject. Both sides use the same drawn
parameters, so the only difference is trapezoidal error on the observation grid and
a tight bound is appropriate.

```{r pknca}
conc_df <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

dose_df <- cohort |>
  dplyr::transmute(id, time = 0, amt = DOSE, arm)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | arm + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = TAU,
  auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)

nca_res |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(median = signif(median(PPORRES), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Arm" = arm,
    "AUC0-12 (mg*h/L)" = auclast,
    "Cmax (mg/L)" = cmax,
    "Cmin (mg/L)" = cmin,
    "Tmax (h)" = tmax
  ) |>
  knitr::kable(caption = "Median steady-state NCA parameters by arm (simulated).")

auc_chk <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::left_join(
    sim |> dplyr::distinct(id, cl),
    by = "id"
  ) |>
  dplyr::mutate(pct_diff = 100 * (PPORRES - DOSE / cl) / (DOSE / cl))

summary(auc_chk$pct_diff)

# Same drawn parameters on both sides: pure numerical error, so all() is right.
# Dropping the renal or the CRRT arm from the solved clearance would move the
# AUC by tens of percent in one arm or the other.
stopifnot(
  nrow(auc_chk) == 2L * N_PER_ARM,
  all(abs(auc_chk$pct_diff) < 1)
)
```

The paper reports no NCA parameters for the model-building cohort, so there is
no published NCA table to compare against; the comparison against published
results is the Figure 3 check below.

# Replicate published figures

## Figure 3 - continuous infusion, simulated steady-state concentration

The external validation (Section 2.3.4) simulated 5000 subjects at each of six
creatinine clearances (40, 80, 120, 160, 200 and 240 mL/min, no CRRT) receiving
50 mg/h by continuous infusion, and the inset of Figure 3 prints the pooled
simulated steady-state concentration as median 3.30 mg/L (2.5th-97.5th
percentile 0.85-14.73). At steady state `Css = 50 / CL`, independent of the
volumes, and with the single log-normal random effect on the total clearance
the pooled distribution is an equal-weight mixture of six log-normals, whose
quantiles can be computed exactly by root-finding. This makes the comparison
deterministic and therefore able to hold on every machine.

```{r figure-3-analytic}
omega_cl <- 0.320546 # log(0.615^2 + 1)
CRCL_GRID <- c(40, 80, 120, 160, 200, 240)
tv_cl <- 2.62 + 4.35 * CRCL_GRID / 44 # Table 3 footnote b, QEFF = 0

# P(Css <= x) for the pooled six-stratum mixture, Css = 50 / (tv_cl * exp(eta))
p_css <- function(x) {
  mean(stats::pnorm(log(50 / (tv_cl * x)), 0, sqrt(omega_cl), lower.tail = FALSE))
}
q_css <- function(p) {
  stats::uniroot(function(x) p_css(x) - p, c(1e-3, 1e3), tol = 1e-10)$root
}

fig3 <- tibble::tibble(
  statistic = c("median", "2.5th percentile", "97.5th percentile"),
  p = c(0.5, 0.025, 0.975),
  published = c(3.30, 0.85, 14.73)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(model = q_css(p)) |>
  dplyr::ungroup() |>
  dplyr::mutate(pct_diff = 100 * (model - published) / published)

fig3 |>
  dplyr::mutate(model = round(model, 2), pct_diff = round(pct_diff, 1)) |>
  dplyr::select(statistic, model, published, pct_diff) |>
  dplyr::rename(
    "Statistic" = statistic,
    "This model (mg/L)" = model,
    "Figure 3 inset (mg/L)" = published,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(caption = "Pooled simulated Css at 50 mg/h. Replicates the Figure 3 inset of Soraluce 2020.")

# Deterministic: a mis-read renal reference (44 mL/min), a dropped renal arm
# or the omega = CV^2 reading moves at least one quantile by far more than 3%.
stopifnot(all(abs(fig3$pct_diff) < 3))
```

The same calculation on the alternative reading of the IIV column, with the
printed 61.5% taken as the standard deviation of eta (omega^2 = 0.615^2), gives
a 97.5th percentile of about 15.9 mg/L, 8% above the published 14.73, and a
2.5th percentile of about 0.78 mg/L, 8% below the published 0.85; only the
median is unchanged. The log-normal CV reading
omega^2 = log(CV^2 + 1) used in the model file reproduces all three printed
values, and it also shows that the paper's percentiles exclude residual error:
adding the Table 3 residual error widens the lower tail to about 0.72-0.77 mg/L.

```{r figure-3-omega-scale}
q_alt <- local({
  omega_cl <- 0.615^2
  p_alt <- function(x) {
    mean(stats::pnorm(log(50 / (tv_cl * x)), 0, sqrt(omega_cl), lower.tail = FALSE))
  }
  stats::uniroot(function(x) p_alt(x) - 0.975, c(1e-3, 1e3), tol = 1e-10)$root
})
round(q_alt, 2)
stopifnot(abs(100 * (q_alt - 14.73) / 14.73) > 5)
```

A stochastic simulation of the same design checks that the ODE model itself
reaches the analytic steady state: 600 mg over 30 min, then 50 mg/h, with
concentrations over the 12-108 h window that Figure 3 plots. It uses 100
subjects per clearance stratum.

```{r figure-3-sim, fig.height = 4}
N_STRATUM <- 100L
ci_cohort <- tidyr::expand_grid(CRCL = CRCL_GRID, k = seq_len(N_STRATUM)) |>
  dplyr::mutate(id = dplyr::row_number(), QEFF = 0) |>
  dplyr::select(-k)

ci_events <- rxode2::et(amt = 600, dur = 0.5, cmt = "central") |>
  rxode2::et(amt = 50 * 107.5, dur = 107.5, time = 0.5, cmt = "central") |>
  rxode2::et(c(0, seq(12, 108, by = 12)), cmt = "central") |>
  rxode2::et(id = ci_cohort$id) |>
  as.data.frame() |>
  dplyr::left_join(ci_cohort, by = "id")

ci_sim <- rxode2::rxSolve(mod, events = ci_events, keep = "CRCL") |>
  as.data.frame() |>
  dplyr::filter(time >= 12)

# Late in the window every subject is close to 50 / cl; compare there with the
# subject's own drawn clearance. Subjects with very low clearance are still
# approaching steady state, so this is gated on the median, not on all().
late <- ci_sim |>
  dplyr::filter(time == 108) |>
  dplyr::mutate(pct_diff = 100 * (Cc - 50 / cl) / (50 / cl))
stopifnot(abs(median(late$pct_diff)) < 2)

ci_q <- c(
  p025 = unname(fig3$model[fig3$p == 0.025]),
  p50 = unname(fig3$model[fig3$p == 0.5]),
  p975 = unname(fig3$model[fig3$p == 0.975])
)

ggplot(ci_sim, aes(time, Cc)) +
  geom_point(alpha = 0.15, size = 0.8) +
  geom_hline(yintercept = ci_q[["p50"]]) +
  geom_hline(yintercept = ci_q[c("p025", "p975")], linetype = "dashed") +
  scale_x_continuous(breaks = seq(12, 108, 12)) +
  coord_cartesian(ylim = c(0, 25)) +
  labs(
    x = "Time (h)", y = "Css (mg/L)",
    title = "600 mg loading dose then 50 mg/h, Clcr 40-240 mL/min",
    caption = paste(
      "Replicates Figure 3 of Soraluce 2020. Points: simulated individual",
      "predictions; lines: exact pooled median and 2.5th/97.5th percentiles."
    )
  )
```

## Table 4 - probability of target attainment at 600 mg every 12 h

Table 4 reports the share of the *observed* patients whose measured
concentrations met `AUC24/MIC >= 80` and `100% T>MIC`, which the model can only
approximate: the simulated cohort draws covariates from the Table 1 medians and
ranges, not from the 40 real patients, and the observed percentages come from
17 and 23 patients (one patient is 4-6 percentage points). The table is
therefore shown for context and not gated.

```{r table-4}
auc24 <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::transmute(id, arm, auc24 = 2 * PPORRES)
cmin <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id, arm) |>
  dplyr::summarise(cmin = min(Cc), .groups = "drop")

pta <- tidyr::expand_grid(MIC = c(1, 2, 4), arm = c("CRRT", "No CRRT")) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    auc_sim = 100 * mean(auc24$auc24[auc24$arm == arm] / MIC >= 80),
    t_sim = 100 * mean(cmin$cmin[cmin$arm == arm] > MIC)
  ) |>
  dplyr::ungroup() |>
  dplyr::left_join(
    tibble::tribble(
      ~MIC, ~arm, ~auc_obs, ~t_obs,
      1, "CRRT", 96, 83,
      2, "CRRT", 52, 65,
      4, "CRRT", 0, 30,
      1, "No CRRT", 76, 76,
      2, "No CRRT", 65, 71,
      4, "No CRRT", 6, 29
    ),
    by = c("MIC", "arm")
  )

pta |>
  dplyr::mutate(dplyr::across(c(auc_sim, t_sim), round)) |>
  dplyr::select(arm, MIC, auc_sim, auc_obs, t_sim, t_obs) |>
  dplyr::rename(
    "Arm" = arm,
    "MIC (mg/L)" = MIC,
    "AUC24/MIC >= 80, simulated (%)" = auc_sim,
    "AUC24/MIC >= 80, Table 4 (%)" = auc_obs,
    "100% T>MIC, simulated (%)" = t_sim,
    "100% T>MIC, Table 4 (%)" = t_obs
  ) |>
  knitr::kable(caption = "Target attainment at 600 mg q12h. Observed values from Table 4 of Soraluce 2020.")
```

The model agrees with the observed attainment at an MIC of 1 mg/L and with the
paper's main conclusion that 600 mg every 12 h leaves many patients below
target at 2 and 4 mg/L. It does not reproduce the paper's finding of no
difference between the arms. In the virtual cohort the CRRT arm reaches the
targets more often than the no-CRRT arm, because its typical total clearance
(`2.62 + 4.35 * 6/44 + 2.51`, about 5.7 L/h) is well below that of the no-CRRT
arm (`2.62 + 4.35 * 71.2/44`, about 9.7 L/h): the extracorporeal clearance
replaces only part of the renal clearance lost with renal failure. The observed
no-CRRT patients reached the targets more often than the virtual ones (65% vs
about 35% at an MIC of 2 mg/L). The Table 4 percentages are computed from the
measured concentrations of 17 and 23 real patients rather than from the model,
so a mismatch of this size does not by itself point to an error in the model
file. The paper's own simulated checks (the Figure 3 inset above) are
reproduced to within 1%.

# Assumptions and deviations

* **IIV scale.** Table 3 prints the IIV as percentages without a formula. The
  model uses the log-normal CV reading, omega^2 = log(CV^2 + 1). The pooled
  simulated steady-state concentrations printed in the Figure 3 inset are
  reproduced to within 1% on this reading, while omega^2 = CV^2 misses the
  97.5th percentile by 8% (and the 2.5th percentile by 8%).
* **Residual error on the log scale.** NONMEM was fitted to log-transformed
  concentrations with a combined error model. It is encoded here as additive
  (0.266 mg/L) plus proportional (0.159) on the linear concentration scale,
  the usual nlmixr2 equivalent. The two forms agree closely for concentrations
  well above the additive term, but they are not identical at low
  concentrations.
* **Proportional error units.** The Table 3 row reads 'Residual
  error_proportional (%)' with the value 0.159; it is taken as a fraction
  (15.9%), since 0.159% would be implausibly small for ICU data.
* **QEFF for non-CRRT subjects.** The paper adds CLEC "only for those undergoing
  CRRT"; the model has no separate indicator, so `QEFF` must be set to 0 for
  subjects not on CRRT.
* **CRCL normalisation.** `CRCL` is the raw urine-measured clearance in mL/min,
  not BSA-normalised. Supplying a Cockcroft-Gault estimate would be a change of
  assay; the Discussion attributes the strength of the covariate relationship
  in this cohort to the measured value.
* **Base-model value discrepancy.** Results 3.1.2 gives the base-model IIV on CL
  as 98.7% and the Discussion as 97.8%; only the final model is packaged, so
  this has no effect.
* **Virtual cohort covariates.** Only the median and range of `CRCL` and CLEC
  are published. The virtual cohort uses truncated log-normal distributions
  with the published medians, geometric SDs chosen by the maintainers so the
  draws span the published ranges; the Table 4 comparison depends on this
  choice and is not gated.
* **Unmodelled features.** The Discussion reports a secondary peak 2-4 h after
  dose in about a third of patients, which the authors attributed to
  enterohepatic circulation but could not model; the packaged model, like the
  published one, does not reproduce it.
