---
title: "Primaquine (Sridharan 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Primaquine (Sridharan 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Sridharan_2019_primaquine"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Sridharan_2019_primaquine"))$description`
* Article: [Indian J Pharmacol. 2019;51(1):17-23](https://doi.org/10.4103/ijp.ijp_230_16)
  (open access; PMC6444836)

```{r model}
mod <- rxode2::rxode2(readModelDb("Sridharan_2019_primaquine"))
mod
```

# Population

The model was fitted to 458 primaquine concentration records pooled from 53
adults enrolled at a single Indian centre (Seth GS Medical College and KEM
Hospital, Mumbai) across three separately registered single-dose studies:
healthy volunteers (CTRI/2011/06/001803), patients with hepatic dysfunction
(CTRI/2011/06/001794) and patients with renal dysfunction
(CTRI/2010/091/000356). The pooled cohort comprised 13 normal healthy
individuals, 12 patients with mild and 6 with moderate hepatic dysfunction
graded by Child-Pugh criteria, and 22 patients with renal dysfunction diagnosed
by National Kidney Foundation KDOQI criteria from serum creatinine (Table 1).

Every participant received a single oral 15-mg primaquine phosphate tablet
after a standardized breakfast following an overnight fast, with liquids
restricted for 2 h and food for 4 h post-dose. Blood was drawn pre-dose and at
0.5, 1.0, 1.5, 2, 3, 4, 6, 8, 12 and 24 h and assayed by reversed-phase HPLC.
Group median (range) body weights were 66 kg (62-95) in healthy volunteers,
65.5 kg (49-75) in mild hepatic dysfunction, 56 kg (47-67) in moderate hepatic
dysfunction and 55 kg (40-74) in renal dysfunction; median ages ranged from
25.5 years in healthy volunteers to 50.5 years in moderate hepatic dysfunction
(Table 1). Sex is reported only as per-group male:female ratios, so exact
female counts are not recoverable.

The same information is available programmatically via the model's `population`
metadata. `readModelDb()` returns the model *function*, whose metadata is
reachable once it has been parsed by `rxode2::rxode2()` -- see the `mod` object
built above.

```{r population}
str(mod$meta$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Sridharan_2019_primaquine.R`. The table
below collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment, first-order absorption and elimination | n/a | Results, "Model development and evaluation": "One compartment model with first-order absorption best described the observed data." |
| `lka` | `log(0.9462)` | Table 2, "Ka theta" = 0.9462 (RSE 18.08%) |
| `lcl` | `log(39.14)` | Table 2, "CL theta" = 39.14 L/h (RSE 10.34%) |
| `lvc` | `log(438)` | Table 2, "V theta" = 438 L (RSE 11.9%) |
| `e_wt_vc` | `fixed(1)` | Methods, "Population pharmacokinetic modeling": "Volume of distribution was modeled, which was normalized to 70-kg person ... in all models tested." No exponent is printed. |
| `e_hepimp_mod_vc` | `2.859` | Table 2, "Moderate HD on V theta" = 2.859 (RSE 30.17%) |
| Covariate function `vc * (1 + theta_mod * HEPIMP_MOD)` | n/a | Methods: `TVP = P x (1 + theta_mild x FLAG) x (1 + theta_mod x FLAG1)` |
| `etalka` | `0.599695` | Table 2, "BSV on Ka (% CV)" = 77.44; 0.7744^2 |
| `etalcl` | `0.423541` | Table 2, "BSV on CL (% CV)" = 65.08; 0.6508^2 |
| `etalvc` | `0.443290` | Table 2, "BSV on V (% CV)" = 66.58; 0.6658^2 |
| IIV form `P_i = TVP * exp(eta_i)` | n/a | Methods, Equation 1 |
| `propSd` | `0.3206` | Table 2, "Proportional (% CV)" = 32.06 (RSE 12.526) |
| `addSd` | `1.5` | Table 2, "Additive (ng/ml)" = 1.5 (RSE 102.5) |
| Residual form `C * (1 + eps_prop) + eps_add` | n/a | Methods, Equation 2 (subscripts transposed as printed; see Errata) |

# Virtual cohort

The original observed data are not publicly available. The cohorts below
reproduce the published design -- a single oral 15-mg dose sampled at the
paper's 11 nominal times -- in the four strata of Table 1, drawing body weights
from each stratum's published median and range. 200 subjects per stratum are
simulated rather than the 13 / 12 / 6 / 22 actually enrolled, so that the
summary statistics compared against Table 1 are stable; the covariate
distributions are unchanged.

```{r cohort}
# `set.seed()` seeds R's RNG (used below for the covariate draws and for the
# residual error applied explicitly). It does NOT seed rxode2's between-subject
# RNG, whose streams are partitioned per solver thread -- so the eta draws
# differ between a 2-core CI runner and a many-core workstation. Every
# assertion below is therefore written either on typical values (no RNG at all)
# or on the centre of a 200-subject stratum with headroom.
set.seed(20190117)

n_per_arm <- 200L

# Table 1 body weights: median (range) per stratum.
strata <- tibble::tribble(
  ~group,                          ~hepimp_mod, ~wt_med, ~wt_lo, ~wt_hi,
  "Healthy",                                 0,      66,     62,     95,
  "Mild hepatic dysfunction",                0,    65.5,     49,     75,
  "Moderate hepatic dysfunction",            1,      56,     47,     67,
  "Renal dysfunction",                       0,      55,     40,     74
)

# Draw weights from a log-normal centred on the published median, with the
# dispersion set so the published range spans roughly the central 98%.
draw_wt <- function(n, med, lo, hi) {
  sdlog <- (log(hi) - log(lo)) / (2 * stats::qnorm(0.99))
  pmin(pmax(stats::rlnorm(n, log(med), sdlog), lo), hi)
}

obs_times <- c(0, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12, 24)
dose_ug <- 15 * 1000 # 15 mg expressed in ug, the model's dosing unit

subjects <- dplyr::bind_rows(lapply(seq_len(nrow(strata)), function(i) {
  s <- strata[i, ]
  tibble::tibble(
    id = (i - 1L) * n_per_arm + seq_len(n_per_arm),
    group = s$group,
    HEPIMP_MOD = s$hepimp_mod,
    WT = draw_wt(n_per_arm, s$wt_med, s$wt_lo, s$wt_hi)
  )
}))

events <- dplyr::bind_rows(
  subjects |>
    dplyr::mutate(time = 0, amt = dose_ug, evid = 1L, cmt = "depot"),
  tidyr::expand_grid(subjects, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

stopifnot(
  nrow(subjects) == 4L * n_per_arm,
  !anyDuplicated(unique(events[, c("id", "time", "evid")]))
)
knitr::kable(
  subjects |>
    dplyr::group_by(group) |>
    dplyr::summarise(
      n = dplyr::n(),
      `Median weight (kg)` = round(stats::median(WT), 1),
      `Weight range (kg)` = paste0(round(min(WT)), "-", round(max(WT))),
      .groups = "drop"
    ) |>
    dplyr::rename(Stratum = group, N = n),
  caption = "Simulated cohort, compared against Table 1 of the source."
)
```

# Simulation of the published study

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("group", "WT", "HEPIMP_MOD"),
  returnType = "data.frame",
  useLinCmt = FALSE
)

# Apply the combined residual error explicitly rather than reading rxode2's
# `sim` column, so the error model in force is visible in the vignette. The
# model declares add() + prop() with variances adding, which is NONMEM's
# C = C' * (1 + eps_prop) + eps_add.
propSd <- 0.3206
addSd <- 1.5
sim <- sim |>
  dplyr::mutate(
    Cobs = Cc + stats::rnorm(dplyr::n(), 0, sqrt(addSd^2 + (propSd * Cc)^2)),
    # The pre-dose sample carries no drug; the assay cannot return a negative
    # concentration, so non-positive simulated values are treated as below the
    # limit of quantification and dropped before NCA.
    Cobs = dplyr::if_else(time == 0, 0, Cobs),
    Cobs = dplyr::if_else(Cobs < 0 & time > 0, NA_real_, Cobs)
  )

head(sim[, c("id", "time", "group", "WT", "HEPIMP_MOD", "Cc", "Cobs")], 12)
```

## Replicating Figure 2 (visual predictive check strata)

Figure 2 of the source stratifies the VPC into "All Other subjects" and
"Moderate HD subjects", and its y-axis ceilings differ by roughly an order of
magnitude (about 170 ng/mL versus about 30 ng/mL). The panels below reproduce
the simulated 5th / 50th / 95th percentile envelope on the same stratification.

```{r figure2, fig.cap = "Replicates Figure 2 of Sridharan 2019: simulated concentration envelope after a single oral 15-mg dose, stratified on moderate hepatic dysfunction."}
vpc <- sim |>
  dplyr::filter(!is.na(Cobs)) |>
  dplyr::mutate(
    stratum = dplyr::if_else(
      HEPIMP_MOD == 1, "Moderate HD subjects", "All Other subjects"
    )
  ) |>
  dplyr::group_by(stratum, time) |>
  dplyr::summarise(
    lo = stats::quantile(Cobs, 0.05),
    mid = stats::median(Cobs),
    hi = stats::quantile(Cobs, 0.95),
    .groups = "drop"
  )

ggplot2::ggplot(vpc, ggplot2::aes(time)) +
  ggplot2::geom_ribbon(
    ggplot2::aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.25
  ) +
  ggplot2::geom_line(ggplot2::aes(y = mid), colour = "firebrick", linewidth = 1) +
  ggplot2::facet_wrap(~stratum, scales = "free_y") +
  ggplot2::labs(x = "Time (h)", y = "Primaquine concentration (ng/mL)") +
  ggplot2::theme_bw()

knitr::kable(
  vpc |>
    dplyr::group_by(stratum) |>
    dplyr::summarise(`Peak of median profile (ng/mL)` = round(max(mid), 1), .groups = "drop") |>
    dplyr::rename(Stratum = stratum),
  caption = "Peak of the simulated median profile by VPC stratum."
)
```

The moderate-hepatic-dysfunction panel peaks roughly 2.5-fold lower than the
pooled remainder, which is the qualitative separation Figure 2 shows. The
separation is smaller here than the 3.3-fold seen in the typical-value profiles
below because these are medians of simulated *observations*, where the additive
residual error lifts the lower concentrations.

## Replicating Figure 3 (national-guideline dosing regimens)

Figure 3 simulates typical profiles for the dosing regimens of the Indian
national guidelines, side by side for "All Other Subjects" and "Moderate HD
Subjects". Because the model scales the apparent volume linearly with body
weight and these regimens are specified per kilogram, the peak concentration is
almost independent of the weight assumed; 70 kg is used below.

```{r figure3, fig.cap = "Replicates Figure 3 of Sridharan 2019: typical-value profiles for the national-guideline regimens, with and without moderate hepatic dysfunction."}
typ_profile <- function(dose_mg_per_kg, n_days, hepimp, wt = 70) {
  amt <- dose_mg_per_kg * wt * 1000 # mg/kg -> ug
  tmax_h <- 24 * n_days + 24
  ev <- dplyr::bind_rows(
    tibble::tibble(
      time = 24 * (seq_len(n_days) - 1),
      amt = amt, evid = 1L, cmt = "depot"
    ),
    tibble::tibble(
      time = seq(0, tmax_h, by = 0.25),
      amt = NA_real_, evid = 0L, cmt = "central"
    )
  ) |>
    dplyr::mutate(id = 1L, WT = wt, HEPIMP_MOD = hepimp) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    as.data.frame()
  rxode2::rxSolve(
    mod, ev,
    omega = NA, sigma = NA, returnType = "data.frame", useLinCmt = FALSE
  ) |>
    dplyr::filter(!is.na(Cc))
}

regimens <- tibble::tribble(
  ~label,                      ~dose, ~days,
  "0.75 mg/kg single dose",     0.75,     1,
  "0.75 mg/kg/day, 14 days",    0.75,    14,
  "0.50 mg/kg/day, 14 days",    0.50,    14,
  "0.25 mg/kg/day, 14 days",    0.25,    14
)

fig3 <- dplyr::bind_rows(lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  dplyr::bind_rows(
    typ_profile(r$dose, r$days, 0) |> dplyr::mutate(stratum = "All Other Subjects"),
    typ_profile(r$dose, r$days, 1) |> dplyr::mutate(stratum = "Moderate HD Subjects")
  ) |>
    dplyr::mutate(label = factor(r$label, levels = regimens$label))
}))

ggplot2::ggplot(fig3, ggplot2::aes(time, Cc, colour = stratum)) +
  ggplot2::geom_line(linewidth = 0.6) +
  ggplot2::facet_wrap(~label, scales = "free_x", ncol = 2) +
  ggplot2::labs(
    x = "Time (h)", y = "Predicted concentration (ng/mL)", colour = NULL
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "top")
```

Figure 3 of the source is a raster plot without printed numbers, but its
single-dose panel (top row) can be read off the axes: the "All Other Subjects"
peak sits just under 100 ng/mL and the "Moderate HD Subjects" peak near
30 ng/mL. Those two readings are what distinguish the two possible readings of
the covariate coefficient (see Errata), so they are asserted here.

```{r figure3-gate}
peak <- fig3 |>
  dplyr::filter(label == "0.75 mg/kg single dose") |>
  dplyr::group_by(stratum) |>
  dplyr::summarise(cmax = max(Cc), .groups = "drop")
knitr::kable(
  peak |> dplyr::rename(Stratum = stratum, `Peak (ng/mL)` = cmax),
  digits = 1,
  caption = "Typical peak after a single 0.75 mg/kg dose, against Figure 3."
)

cmax_other <- peak$cmax[peak$stratum == "All Other Subjects"]
cmax_mod <- peak$cmax[peak$stratum == "Moderate HD Subjects"]
stopifnot(
  # Read off the Figure 3 single-dose panel: a little under 100, and about 30.
  cmax_other > 85, cmax_other < 105,
  cmax_mod > 24, cmax_mod < 34,
  # The alternative reading of the covariate coefficient (a bare 2.859-fold
  # volume rather than 1 + 2.859) would put this peak at 37.3 ng/mL.
  cmax_mod < 34
)
```

# Closed-form verification

For a one-compartment model with first-order absorption, the typical-value
solution is analytic. This check compares the packaged model, solved with
between-subject and residual variability switched off, against that closed
form. Both sides use the same parameter values, so the only difference is
numerical integration error and the bound is tight.

```{r closed-form}
ka <- 0.9462
cl <- 39.14
v70 <- 438

closed_form <- function(dose_ug, wt, hepimp) {
  v <- v70 * (wt / 70) * (1 + 2.859 * hepimp)
  kel <- cl / v
  tmax <- log(ka / kel) / (ka - kel)
  cmax <- (dose_ug / v) * (ka / (ka - kel)) *
    (exp(-kel * tmax) - exp(-ka * tmax))
  c(cmax = cmax, tmax = tmax, auc = dose_ug / cl, half_life = log(2) / kel)
}

solved <- function(dose_ug, wt, hepimp) {
  ev <- dplyr::bind_rows(
    tibble::tibble(time = 0, amt = dose_ug, evid = 1L, cmt = "depot"),
    tibble::tibble(
      time = seq(0, 600, by = 0.01), amt = NA_real_, evid = 0L, cmt = "central"
    )
  ) |>
    dplyr::mutate(id = 1L, WT = wt, HEPIMP_MOD = hepimp) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    as.data.frame()
  s <- rxode2::rxSolve(
    mod, ev,
    omega = NA, sigma = NA, returnType = "data.frame", useLinCmt = FALSE
  ) |>
    dplyr::filter(!is.na(Cc))
  i <- which.max(s$Cc)
  c(
    cmax = s$Cc[i], tmax = s$time[i],
    auc = sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2),
    half_life = log(2) / (cl / (v70 * (wt / 70) * (1 + 2.859 * hepimp)))
  )
}

cf_chk <- dplyr::bind_rows(lapply(
  list(
    list(lbl = "15 mg, 70 kg, no hepatic dysfunction", wt = 70, hep = 0),
    list(lbl = "15 mg, 56 kg, moderate hepatic dysfunction", wt = 56, hep = 1)
  ),
  function(s) {
    a <- closed_form(dose_ug, s$wt, s$hep)
    b <- solved(dose_ug, s$wt, s$hep)
    tibble::tibble(
      Scenario = s$lbl,
      Quantity = c("Cmax (ng/mL)", "Tmax (h)", "AUC0-inf (ng*h/mL)", "t1/2 (h)"),
      `Closed form` = as.numeric(a),
      `rxode2` = as.numeric(b)
    )
  }
)) |>
  dplyr::mutate(`% diff` = 100 * (rxode2 - `Closed form`) / `Closed form`)

knitr::kable(cf_chk, digits = c(0, 0, 3, 3, 3), caption = "Packaged model against the analytic one-compartment oral solution.")

stopifnot(
  # Same parameters on both sides: this is pure integration error.
  max(abs(cf_chk$`% diff`[cf_chk$Quantity != "Tmax (h)"])) < 0.5,
  # Tmax is resolved only to the 0.01 h observation grid.
  max(abs(cf_chk$rxode2[cf_chk$Quantity == "Tmax (h)"] -
    cf_chk$`Closed form`[cf_chk$Quantity == "Tmax (h)"])) < 0.02
)
```

The moderate-hepatic-dysfunction row also pins the covariate arithmetic: the
apparent volume is multiplied by `1 + 2.859 = 3.859`, which lengthens the
terminal half-life from 7.8 h to 29.9 h at a fixed clearance.

# PKNCA validation

Non-compartmental analysis is run on the simulated study using the same design
the paper analysed -- a single 15-mg oral dose sampled at 11 nominal times --
so that the trapezoidal and extrapolation behaviour of NCA applies equally to
both sides of the comparison.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cobs)) |>
  dplyr::select(id, time, Cobs, group)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, group) |>
  dplyr::mutate(amt = amt / 1000) # ug -> mg, to match the published dose unit

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cobs ~ time | group + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | group + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = 0,
  end = Inf,
  cmax = TRUE,
  tmax = TRUE,
  aucinf.obs = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

## Comparison against the published NCA

Table 1 of the source reports median (range) NCA parameters per stratum. The
comparison below uses those medians as the reference.

```{r nca-comparison}
published <- tibble::tribble(
  ~group,                          ~cmax, ~tmax, ~aucinf.obs, ~half.life,
  "Healthy",                        29.3,   3.0,       314.1,        4.1,
  "Mild hepatic dysfunction",       45.6,   3.0,       303.0,        6.6,
  "Moderate hepatic dysfunction",   14.4,   3.5,       117.2,        4.3,
  "Renal dysfunction",              45.6,   2.0,       261.9,        3.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "group",
  units = c(
    cmax = "ng/mL", tmax = "h",
    aucinf.obs = "ng*h/mL", half.life = "h"
  ),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated versus published NCA (Table 1 medians).",
    "* differs from the reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

```{r nca-gate}
chk <- cmp |>
  dplyr::mutate(
    parameter = cmp[[1]],
    pct = suppressWarnings(as.numeric(gsub("[^0-9.eE+-]", "", `% diff`)))
  )
cmax_pct <- chk$pct[grepl("^Cmax", chk$parameter)]
tmax_pct <- chk$pct[grepl("^Tmax", chk$parameter)]

stopifnot(
  # Structural gate on the exposure the model does reproduce. A mis-transcribed
  # dose, volume or covariate multiplier moves peak concentration by a factor,
  # not by tens of percent, so a centre statistic over the four strata catches
  # it with room to spare for the sampling noise of a 200-subject median.
  length(cmax_pct) == 4L,
  stats::median(abs(cmax_pct), na.rm = TRUE) < 30,
  max(abs(cmax_pct), na.rm = TRUE) < 45,
  # Tmax depends on ka and on kel, so it also moves if either is wrong, but it
  # is resolved only to the paper's coarse nominal sampling grid.
  stats::median(abs(tmax_pct), na.rm = TRUE) < 30
)
```

**Peak exposure is reproduced in all four strata.** Median simulated Cmax lands
within 27% of the published median everywhere and within 10% in the healthy and
moderate-hepatic-dysfunction strata, and Tmax matches the published median
exactly in three of the four. That is the comparison most sensitive to a
mis-transcribed dose, apparent volume, or hepatic covariate multiplier, and it
is the one asserted above.

**AUC and terminal half-life are systematically over-predicted, and this is a
property of the published model rather than of the packaged implementation.**
Two distinct effects are visible in the table:

* In the healthy, mild-hepatic-dysfunction and renal-dysfunction strata,
  simulated `AUC0-inf` runs 20-30% above the published medians and half-life
  runs up to 95% above them. The model's typical clearance, 39.14 L/h, is lower
  than the median apparent clearance Table 1 reports for every one of those
  strata (45.6, 50.95 and 60.4 L/h), and a lower clearance at a fixed volume
  produces both a larger AUC and a longer terminal slope. The two sides of the
  comparison are not independent estimates of the same quantity: Table 1's
  values are medians of individually-derived NCA on 11 samples per subject,
  while the model is a population fit to the same records, and the paper itself
  publishes both without reconciling them.
* The moderate-hepatic-dysfunction stratum diverges much further, at +163% on
  AUC and +201% on half-life. Table 1's own NCA shows that group with both an
  approximately threefold higher apparent volume (886.71 versus 300.7 L) *and*
  an approximately fourfold higher apparent clearance (183.71 versus 45.6 L/h)
  -- the signature of reduced bioavailability, since `CL/F` and `V/F` both
  scale as `1/F`. The final population model attributes the entire effect to
  volume and leaves clearance untouched, so it necessarily predicts an
  unchanged AUC and a terminal half-life stretched from 7.8 h to 29.9 h, where
  the observed group had a median half-life of 4.3 h and roughly a third of the
  healthy group's AUC. The authors reach the same diagnosis in the Discussion:
  "We hypothesize that presence of a gut wall edema may hamper the absorption
  of the drug in these patients leading to reduced bioavailability."

No parameter has been tuned to close either gap. Users simulating total
exposure in moderate hepatic dysfunction with this model should be aware that
its AUC and half-life predictions for that stratum are not supported by the
source paper's own non-compartmental analysis.

# Assumptions and deviations

* **Dosing unit.** The model's dosing unit is micrograms so that amounts over
  volumes in litres give ng/mL directly, the unit in which Table 1, Table 2 and
  the additive residual error are all reported. A 15-mg dose is therefore
  entered as `amt = 15000`.
* **Dose is 15 mg as administered.** The paper describes both a "single-dose
  (15 mg) PQ" and a "Tablet PQ phosphate 15-mg" and does not state whether the
  15 mg is base or phosphate salt. Simulating 15 mg reproduces the paper's own
  Table 1 peak concentrations and Figure 3 panels; simulating the 8.6-mg base
  equivalent of a 15-mg phosphate tablet would not. The parameters are apparent
  (`CL/F`, `V/F`) and are consistent only with the dose amount the authors
  themselves entered.
* **Weight exponent on volume.** The paper states that volume was "normalized
  to 70-kg person" in all models but never prints an exponent. `e_wt_vc` is
  held at 1, the linear scaling that a plain normalization denotes, rather than
  estimated. No weight scaling is applied to clearance, which the paper's
  covariate search did not retain.
* **Between-subject variability convention.** Table 2 publishes the etas as
  "% CV". The same column header is used for the proportional residual error,
  where it can only mean `100 x sqrt(variance)`, so the etas are read on that
  convention and each variance is `(%CV/100)^2`. Reading them instead as
  log-normal CVs, `log(1 + CV^2)`, would give 0.353 / 0.367 / 0.470 in place of
  0.424 / 0.443 / 0.600. The paper gives no footnote to settle it.
* **Bioavailability.** `F` is never estimated, so every clearance and volume in
  the model is apparent. No `lfdepot` is carried.
* **Flip-flop constraint.** The estimation constrained `Ka` above the
  elimination rate constant. At the published estimates the constraint is
  inactive (`Ka` = 0.946/h against `kel` = 0.089/h at 70 kg), so it is not
  encoded in the model; it would only matter when refitting.
* **Sex.** `sex_female_pct` is left `NA` because the paper reports sex only as
  per-group male:female ratios (5:1, 5:1, all male, 2:1) that do not resolve to
  integer counts for the 12- and 22-subject groups.
* **Simulated cohort size.** 200 subjects per stratum are simulated rather than
  the 13 / 12 / 6 / 22 enrolled, so that the medians compared against Table 1
  are stable across machines. Covariate distributions match Table 1.
* **BLQ handling.** Simulated concentrations that fall below zero once the
  additive residual error is applied are dropped before NCA; the paper does not
  report an assay lower limit of quantification.

# Errata and internal inconsistencies in the source

The source is internally inconsistent in four places. In each case the value
used in the model is the one supported by Table 2, the Results text and the
paper's own figures, over the Abstract.

1. **Volume of distribution.** The Abstract reports the final `V` as 498 L; the
   Results text ("volume of distribution (Vd) was 438 L") and Table 2
   (`V theta` = 438, bootstrap median 439.6, 95% CI 345.37-550.53) both give
   438 L. 438 L is used, and it is the value that reproduces Figure 3.
2. **Proportional residual error.** The Abstract reports "proportion error 12%
   CV"; Table 2 reports 32.06% CV with a bootstrap 95% CI of 27.98-35.74, which
   excludes 12%. 32.06% is used.
3. **Which hepatic stratum is significant.** The Abstract says "Mild hepatic
   dysfunction was a significant covariate on volume of distribution"; the
   Results ("Only the moderate hepatic dysfunction showed to be significant on
   the volume of distribution"), the Table 2 row label ("Moderate HD on V"),
   the Conclusion and both figures all say moderate. Moderate is used, and mild
   hepatic dysfunction is recorded in `covariatesDataExcluded`.
4. **Residual-error equation subscripts.** Methods Equation 2 is printed as
   `Cij = C'ij (1 + eps_add,j) + eps_prop,j`, which transposes the two error
   terms: the multiplicative term is labelled additive and the standalone term
   proportional. Table 2's units settle the orientation, since the additive
   term carries ng/mL and the proportional term % CV. The model uses the
   conventional orientation, `C * (1 + eps_prop) + eps_add`.

Two further points of reading:

* **Magnitude of the hepatic covariate effect.** The Methods covariate function
  is proportional, `TVP = P x (1 + theta_mild x FLAG) x (1 + theta_mod x FLAG1)`,
  so `theta_mod = 2.859` is a *fractional* increase and the multiplier on volume
  is `1 + 2.859 = 3.859`. The prose ("the parameter increased approximately
  three folds") reads naturally either as that fractional increase or as a bare
  2.859-fold multiplier. Figure 3 settles it: at 0.75 mg/kg as a single dose it
  shows peaks near 95 and 30 ng/mL for the two strata, which the 3.859-fold
  volume reproduces (93.7 and 28.3 ng/mL) and a bare 2.859-fold volume does not
  (93.7 and 37.3 ng/mL). This is asserted in the Figure 3 section above.
* **Additive-error RSE.** The Results text gives the additive residual variability
  RSE as 116%; Table 2 gives 102.5%. Neither enters the model.

# Session information

```{r session}
sessionInfo()
```
