---
title: "Ixazomib efficacy and safety in multiple myeloma (Srimani 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ixazomib efficacy and safety in multiple myeloma (Srimani 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Srimani 2022 built a joint exposure-response framework for ixazomib in the
phase III TOURMALINE-MM1 trial (ixazomib or placebo, each with lenalidomide
and dexamethasone, "LenDex"). Individual ixazomib exposure came from the
Gupta 2017 population PK model. Four packaged models carry the paper:

| Model | Endpoints |
|-------|-----------|
| `Srimani_2022_ixazomib_mprotein` | Serum M-protein (two-population indirect response), dropout, relapse and progression-free-survival (PFS) hazards |
| `Srimani_2022_ixazomib_platelets` | Platelet count (Friberg chain with ixazomib and lenalidomide K-PD effects) |
| `Srimani_2022_ixazomib_diarrhea` | Weekly diarrhea-grade transition probabilities (discrete-time Markov model) |
| `Srimani_2022_ixazomib_rash` | Weekly rash-grade transition probabilities (discrete-time Markov model) |

The efficacy endpoints are one file because the dropout, relapse and PFS
hazards use the individual M-protein parameters. The three safety models were
fitted separately and are separate files. Each file carries the fixed Gupta
2017 PK model (also available on its own as `modellib("Gupta_2017_ixazomib")`).

* Citation: `r rxode2::rxode(readModelDb("Srimani_2022_ixazomib_mprotein"))$reference`
* Article: <https://doi.org/10.1002/psp4.12815> (open access; Appendix S1 holds
  the rash model, the transition matrices and the safety-population demographics)

All four models run on an hour time base, the NONMEM unit of the source:
the paper's per-week values are the hourly estimates times 168.

# Population

The exposure-efficacy dataset holds 467 of the 720 treated patients (240
ixazomib, 227 placebo) with a baseline serum M-protein of at least 10 g/L and at
least three M-protein observations: median age 66 years (40-91), 43.7% female,
87.6% White and 7.3% Asian, median baseline M-protein 23 g/L, 20.8% with
high-risk cytogenetics and 55.9% previously exposed to an immunomodulatory drug
(IMiD; Table 1). The safety models use all 720 treated patients (361 ixazomib,
359 placebo; Supplementary Table 4): median age 66 years (30-91), 8.9% Asian,
55.0% IMiD-exposed, median baseline platelet count 197 x 10^9/L. Patients
received ixazomib 4 mg or placebo on days 1, 8 and 15 and lenalidomide 25 mg on
days 1-21 of 28-day cycles, with dexamethasone 40 mg weekly.

The same information is available programmatically via each model's
`population` metadata, e.g.
`readModelDb("Srimani_2022_ixazomib_mprotein")()$population`.

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
table collects the equations and the parameters.

| Model / element | Value | Source location |
|-----------------|-------|-----------------|
| Ixazomib PK (all files) | Gupta 2017 Table 3, fixed | Methods "Pharmacokinetic/pharmacodynamic modeling" (individual PK by Bayesian estimation from the published model) |
| M-protein `mprotein <- MCPROT * (mprotein_rel + rplus)` | n/a | Equations 1-2 |
| Resistant growth `rplus <- exp(kl * (t - T_MPROTEIN_NADIR)) - 1` after the nadir | n/a | Equation 3 |
| `d/dt(mprotein_rel) <- kr * (rss - mprotein_rel)`, `rss <- yss * (1 - imax*C/(ic50 + C))` | n/a | Equations 4-5 |
| `lkr`, `lyss`, `lkl` | -6.70, -1.95, -9.78 | Table 2 |
| `imax`, `lic50` | 0.758, 1.19 | Table 2 |
| `e_cytohr_kr`, `e_imidnaive_yss` | 0.590, -0.427 | Table 2; Equations 6-7 |
| `etalkr`, `etalyss`, `etalkl` (variances) | 0.655, 2.39, 1.34 | Table 2 |
| `propSd_mprotein`, `addSd_mprotein` | 0.218, 0.5 (fixed) | Table 2 |
| Dropout hazard | n/a | Equations 8-9 |
| `llambda0_drop`, `lambda_rss_drop`, `let50_drop`, `lt0_drop`, `llambda_mprot_drop`, `lalpha_bl_drop` | -10.1, 1.25e-5, 8.14, 6.86, -6.60, 0.148 | Table 2 |
| Relapse hazard (log-logistic AFT) | n/a | Equations 10-12 |
| `llambda0_relapse`, `lalpha_relapse`, `lbeta_relapse`, `probit_tnadir0` | -1.23, 10.9, 0.938, -1.67 | Table 2 |
| `alpha_ixa_relapse`, `alpha_kr_relapse`, `alpha_kr0_relapse` | 0.0509, -2960, 0.0509 | Table 2 |
| PFS hazard | n/a | Equation 13 |
| `llambda0_pfs`, `lambda_ixa_pfs`, `let50_pfs` | -9.21, -0.00108, 7.82 | Table 2 |
| `e_kr_pfs`, `e_yss_pfs`, `e_mcprot_pfs` | 0.798, 0.680, 0.784 | Table 2 |
| Platelet chain `prol -> transit1 -> transit2 -> circ` | n/a | Equations 15-18; Figure 1d |
| Individual baseline `bl_i` | n/a | Equation 19 |
| `e_ixa <- slp_ixa * (Cc + k_ixa * AUC)`, `e_len <- slp_len * ceff_len` | n/a | Equations 20-21 |
| `lkprp`, `lkin`, `bl`, `e_plt_base_bl` | -3.20, -3.21, 203, 0.837 | Table 3 |
| `slp_ixa`, `k_ixa`, `lkel`, `slp_len` | 0.000859, 8.18e-5, log(0.0483), 0.0134 | Table 3 |
| `propSd_PLT`, `addSd_PLT` | 0.160, 28.1 | Table 3 |
| Diarrhea and rash cumulative logits | n/a | Equation 14; Supplementary Equations S1-S2 |
| Diarrhea `b01` ... `b33` | -5.28 ... -1.78 | Table 3 |
| Diarrhea `lemax_time_g10`, `lk_time_g10`, `e_week1_g01`, `slp_aucwk_g0`, `e_imidnaive_b3` | 2.25, -10.4, 0.933, 0.000715, -2.87 | Table 3 |
| Diarrhea `etab01`, `etab11`, `etab21`, `etab31` (variances) | 1.86, 3.72, 3.24, 2.74 | Table 3 |
| Rash `b01` ... `b33` | -7.37 ... -2.19 | Supplementary Table 5 |
| Rash `lemax_time_g10`, `lk_time_g10`, `e_transient_g0`, `lk_transient_g0`, `slp_aucwk_g0`, `e_asian_b0` | 0.921, -7.83, 3.81, -7.13, 0.000929, 1.19 | Supplementary Table 5 |
| Rash `etab01`, `etab11`, `etab21` (variances) | 1.80, 0.685, 0.523 | Supplementary Table 5 |

# Helpers

The event tables use the named ODE states. Ixazomib doses go into `depot`, and
lenalidomide doses for the platelet model into `depot_kpd`. Random effects are
drawn in R from each model's omega matrix and passed to a `zeroRe()` copy of the
model, so the cohorts below are identical on every machine and the same
individual can be solved twice. The `delay()` used for the weekly AUC needs a
dense solver, so every solve uses `atol = rtol = 1e-6`.

```{r helpers}
hr_per_week <- 168

ixa_dose_times <- function(ncycle) {
  as.vector(outer(c(0, 7, 14), 28 * (seq_len(ncycle) - 1), "+")) * 24
}

len_dose_times <- function(ncycle) {
  as.vector(outer(0:20, 28 * (seq_len(ncycle) - 1), "+")) * 24
}

draw_etas <- function(mod, n, seed) {
  om <- rxode2::rxode(mod)$omega
  set.seed(seed)
  z <- matrix(stats::rnorm(n * nrow(om)), nrow = n) %*% chol(om)
  colnames(z) <- rownames(om)
  data.frame(id = seq_len(n), z)
}

# Doses for the ixazomib arm plus observation rows on `obs_cmt` at `obs_times`.
build_events <- function(cohort, obs_times, obs_cmt, ncycle, len_doses = FALSE) {
  ixa <- cohort |>
    filter(arm == "Ixazomib") |>
    select(id) |>
    tidyr::crossing(time = ixa_dose_times(ncycle)) |>
    mutate(amt = 4, evid = 1L, cmt = "depot")
  len <- if (len_doses) {
    cohort |>
      select(id) |>
      tidyr::crossing(time = len_dose_times(ncycle)) |>
      mutate(amt = 25, evid = 1L, cmt = "depot_kpd")
  } else {
    NULL
  }
  obs <- cohort |>
    select(id) |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = 0, evid = 0L, cmt = obs_cmt)
  bind_rows(ixa, len, obs) |>
    left_join(cohort, by = "id") |>
    arrange(id, time, desc(evid))
}

solve_ind <- function(mod, events, etas, extra = NULL) {
  p <- etas
  if (!is.null(extra)) {
    p <- dplyr::left_join(p, extra, by = "id")
  }
  as.data.frame(rxode2::rxSolve(
    rxode2::zeroRe(mod), events,
    params = p, atol = 1e-6, rtol = 1e-6,
    returnType = "data.frame"
  ))
}

# Advance a 4-state Markov chain one week at a time. `pr` holds one row per
# subject-week with columns p00 ... p33 ordered by id and week; `u` holds one
# uniform draw per subject-week.
markov_walk <- function(pr, u) {
  grade <- integer(nrow(pr))
  current <- 0L
  for (k in seq_len(nrow(pr))) {
    if (k == 1L || pr$id[k] != pr$id[k - 1L]) current <- 0L
    probs <- c(
      pr[[paste0("p", current, "0")]][k], pr[[paste0("p", current, "1")]][k],
      pr[[paste0("p", current, "2")]][k], pr[[paste0("p", current, "3")]][k]
    )
    current <- findInterval(u[k], cumsum(probs)[1:3])
    grade[k] <- current
  }
  grade
}

# Typical-value grade distribution after each week, from a transition-matrix
# sequence (rows of `pr` ordered by week for one subject).
markov_expected <- function(pr) {
  x <- c(1, 0, 0, 0)
  out <- matrix(NA_real_, nrow(pr), 4)
  for (k in seq_len(nrow(pr))) {
    P <- matrix(unlist(pr[k, paste0("p", rep(0:3, each = 4), rep(0:3, 4))]), 4, 4, byrow = TRUE)
    x <- as.numeric(x %*% P)
    out[k, ] <- x
  }
  out
}
```

# Ixazomib exposure (PKNCA)

The PK layer is the fixed Gupta 2017 model. A single 4 mg dose for a
typical patient (BSA 1.87 m^2) must give AUC0-inf = F x Dose / CL =
0.58 x 4000 / 1.86 = 1247 ng*h/mL, and PKNCA run on the solved profile checks
the unit handling of every file.

```{r pknca}
mod_mp <- readModelDb("Srimani_2022_ixazomib_mprotein")
pk_times <- sort(unique(c(seq(0, 24, by = 0.25), seq(25, 24 * 70, by = 6))))
pk_ev <- rbind(
  data.frame(id = 1L, time = 0, amt = 4, evid = 1L, cmt = "depot"),
  data.frame(id = 1L, time = pk_times, amt = 0, evid = 0L, cmt = "central")
)
pk_ev$BSA <- 1.87
pk_ev$MCPROT <- 23
pk_ev$TUM_CYTOGENETIC_HIGH_RISK <- 0
pk_ev$PRIOR_IMMUNOMODULATORY_DRUG <- 1
pk_ev$T_MPROTEIN_NADIR <- 1e9
pk_ev$treatment <- "Ixazomib 4 mg single dose"
pk_sim <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod_mp), pk_ev,
  atol = 1e-6, rtol = 1e-6, returnType = "data.frame"
))
# A one-subject solve returns no id column.
pk_sim$id <- 1L
pk_sim$treatment <- "Ixazomib 4 mg single dose"

conc <- pk_sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment) |>
  distinct(id, time, .keep_all = TRUE)
dose <- pk_ev |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
o_conc <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose, amt ~ time | treatment + id)
o_data <- PKNCA::PKNCAdata(o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE)
)
nca <- as.data.frame(PKNCA::pk.nca(o_data))
nca_wide <- nca |>
  select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
auc_expected <- 0.58 * 4000 / 1.86
nca_wide |>
  transmute(cmax, tmax, half.life, aucinf.obs, auc_expected = auc_expected) |>
  dplyr::rename(
    "Cmax (ng/mL)" = cmax, "Tmax (h)" = tmax, "t1/2 (h)" = half.life,
    "AUC0-inf (ng*h/mL)" = aucinf.obs, "F*Dose/CL (ng*h/mL)" = auc_expected
  ) |>
  knitr::kable(digits = 2)
# A deterministic typical-value solve; 3% leaves room for trapezoidal error
# over the 70-day grid and the extrapolated tail.
stopifnot(abs(nca_wide$aucinf.obs / auc_expected - 1) < 0.03)
```

# Virtual cohorts

Two arms of 200 virtual patients each, with covariates resembling Table 1 and
Supplementary Table 4. BSA is not reported by Srimani 2022 and is drawn around
the Gupta 2017 reference of 1.87 m^2.

```{r cohort}
n_arm <- 200
set.seed(20221)
cohort <- tibble(
  id = seq_len(2 * n_arm),
  arm = rep(c("Ixazomib", "Placebo"), each = n_arm),
  BSA = pmin(pmax(stats::rnorm(2 * n_arm, 1.87, 0.2), 1.4), 2.4),
  MCPROT = pmin(pmax(exp(stats::rnorm(2 * n_arm, log(23), 0.55)), 10), 102),
  TUM_CYTOGENETIC_HIGH_RISK = stats::rbinom(2 * n_arm, 1, 0.208),
  PRIOR_IMMUNOMODULATORY_DRUG = stats::rbinom(2 * n_arm, 1, 0.56),
  RACE_ASIAN = stats::rbinom(2 * n_arm, 1, 0.089),
  PLT_BASE = pmin(pmax(exp(stats::rnorm(2 * n_arm, log(197), 0.33)), 35), 666)
)
cohort |>
  group_by(arm) |>
  summarise(
    n = n(), `median BSA` = median(BSA), `median M-protein (g/L)` = median(MCPROT),
    `high-risk cytogenetics (%)` = 100 * mean(TUM_CYTOGENETIC_HIGH_RISK),
    `IMiD-exposed (%)` = 100 * mean(PRIOR_IMMUNOMODULATORY_DRUG), `Asian (%)` = 100 * mean(RACE_ASIAN),
    `median platelets (10^9/L)` = median(PLT_BASE)
  ) |>
  knitr::kable(digits = 1)
```

# Efficacy: M-protein, dropout, relapse and PFS

## Drawing the nadir time from the relapse model

The M-protein model needs each patient's time of nadir, `T_MPROTEIN_NADIR`, after which
the resistant population grows. The paper derived it from the observed data;
its relapse time-to-event model stands in for that time ("time to relapse ...
was not included directly as a random parameter in the M-protein PK/PD model.
Instead, a log-logistic accelerated failure time TTE model described relapse").
The simulation therefore runs twice with the same random effects: the first
pass gives each patient's relapse survival curve, a relapse time is drawn from
it, and the second pass uses that time as `T_MPROTEIN_NADIR`. Dropout times are drawn
from the dropout survival in the same way and end each patient's on-study
M-protein record.

```{r efficacy}
ncycle_eff <- 40
eff_times <- seq(0, 160, by = 2) * hr_per_week
eff_ev <- build_events(cohort, eff_times, "mprotein_rel", ncycle_eff) |>
  mutate(T_MPROTEIN_NADIR = 1e9)
eta_mp <- draw_etas(mod_mp, nrow(cohort), seed = 101)
pass1 <- solve_ind(mod_mp, eff_ev, eta_mp)

set.seed(102)
draws <- tibble(id = cohort$id, u_rel = stats::runif(nrow(cohort)), u_drop = stats::runif(nrow(cohort)))
event_time <- function(time, surv, u) {
  k <- which(surv < u)
  if (length(k) == 0L) 1e9 else time[k[1]]
}
nadir_drawn <- pass1 |>
  left_join(draws, by = "id") |>
  group_by(id) |>
  summarise(T_MPROTEIN_NADIR = event_time(time, surv_relapse, u_rel[1]))

eff_ev2 <- eff_ev |>
  select(-T_MPROTEIN_NADIR) |>
  left_join(nadir_drawn, by = "id")
pass2 <- solve_ind(mod_mp, eff_ev2, eta_mp)
# Dropout is drawn from the second pass, whose dropout hazard sees the
# M-protein regrowth after the nadir.
drop_drawn <- pass2 |>
  left_join(draws, by = "id") |>
  group_by(id) |>
  summarise(t_drop = event_time(time, surv_drop, u_drop[1]))
pass2 <- pass2 |>
  left_join(cohort |> select(id, arm), by = "id") |>
  left_join(drop_drawn, by = "id")
stopifnot(!anyNA(pass2$mprotein), !anyNA(pass2$surv_pfs))
```

## M-protein (Figure 2a)

```{r mprotein-vpc}
mp_onstudy <- pass2 |>
  filter(time <= t_drop) |>
  mutate(week = time / hr_per_week, mprot_int = round(mprotein))
mp_sum <- mp_onstudy |>
  group_by(arm, week) |>
  summarise(
    n = n(), median = median(mprot_int),
    p05 = quantile(mprot_int, 0.05), p95 = quantile(mprot_int, 0.95), .groups = "drop"
  ) |>
  filter(n >= 20, week <= 90)
ggplot(mp_sum, aes(week)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), linewidth = 1) +
  facet_wrap(~arm) +
  labs(
    x = "Time (weeks)", y = "M-protein (g/L)",
    title = "Replicates Figure 2a of Srimani 2022",
    caption = "Median and 5th-95th percentiles of on-study model predictions, rounded to integers as in the paper."
  )
mp_wk40 <- mp_sum |> filter(week == 40)
knitr::kable(mp_wk40, digits = 1, caption = "On-study M-protein at week 40.")
```

Figure 2a shows an observed median M-protein falling from the 23 g/L baseline
to about 3-5 g/L by week 20 and staying there, slightly lower with ixazomib.
The model reproduces the level and the arm ordering of the median. Its upper
tail is wider than the paper's: the simulated 95th percentile stays near
20-30 g/L, against a simulated 95% band of roughly 9-20 g/L in Figure 2a. The
tail is set by the large inter-individual variance of `Yss` (2.39, i.e. 155%;
a patient at the 95th percentile has a drug-free nadir above baseline), and
the paper's VPC kept only the observation times of patients still on study,
which this virtual cohort can approximate only through the simulated dropout.
The 95th percentile is therefore not gated.

```{r mprotein-check}
# Typical (median) response depth: Yss = exp(-1.95) = 14.3% of a 23 g/L
# baseline is 3.3 g/L without ixazomib; a transcription error in kR, Yss, Imax
# or IC50 moves the week-40 median by several g/L.
stopifnot(
  all(mp_wk40$median >= 1 & mp_wk40$median <= 6),
  mp_wk40$median[mp_wk40$arm == "Ixazomib"] <= mp_wk40$median[mp_wk40$arm == "Placebo"]
)
```

## Dropout, relapse and PFS (Figure 2b-d)

The curves below are the population-averaged survival from the cumulative
hazards (the mean of the individual survival functions), which is what the
Kaplan-Meier estimate of a large trial converges to.

```{r survival}
surv_long <- pass2 |>
  group_by(arm, time) |>
  summarise(
    Dropout = mean(surv_drop), Relapse = mean(surv_relapse), PFS = mean(surv_pfs),
    .groups = "drop"
  ) |>
  pivot_longer(c(Dropout, Relapse, PFS), names_to = "endpoint", values_to = "surv") |>
  mutate(week = time / hr_per_week, endpoint = factor(endpoint, c("Dropout", "Relapse", "PFS")))
ggplot(surv_long, aes(week, 100 * surv, colour = arm)) +
  geom_line(linewidth = 1) +
  facet_wrap(~endpoint) +
  coord_cartesian(xlim = c(0, 140), ylim = c(0, 100)) +
  labs(
    x = "Time (weeks)", y = "Patients without event (%)", colour = NULL,
    title = "Replicates Figure 2b-d of Srimani 2022"
  )

median_week <- function(week, surv) {
  k <- which(surv < 0.5)
  if (length(k) == 0L) NA_real_ else week[k[1]]
}
pfs_tab <- surv_long |>
  filter(endpoint == "PFS") |>
  group_by(arm) |>
  summarise(sim_median_wk = median_week(week, surv), surv_120 = surv[week == 120]) |>
  mutate(
    obs_median_wk = ifelse(arm == "Ixazomib", 20.6, 14.7) * 365.25 / 12 / 7,
    pct_diff = 100 * (sim_median_wk / obs_median_wk - 1)
  )
pfs_tab |>
  dplyr::rename(
    "Arm" = arm, "Simulated median PFS (weeks)" = sim_median_wk,
    "Simulated PFS at 120 weeks" = surv_120, "Observed median PFS (weeks)" = obs_median_wk,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Median PFS: model versus the trial (20.6 vs 14.7 months).")
```

The trial's observed medians are a poor target here: Figure 2d shows the
paper's own simulations over-predicting ixazomib-arm PFS (the observed curve
leaves the 95% band after week 100), and the Results call this "a slight
overprediction of PFS". The gate therefore compares the model with the paper's
simulated 95% bands, read off Figure 2b-d by the maintainers at weeks 80 and
120.

```{r survival-check}
bands <- tribble(
  ~endpoint, ~arm, ~week, ~lo, ~hi,
  "Dropout", "Ixazomib", 80, 48, 63, "Dropout", "Ixazomib", 120, 22, 46,
  "Dropout", "Placebo", 80, 48, 62, "Dropout", "Placebo", 120, 22, 46,
  "Relapse", "Ixazomib", 80, 47, 58, "Relapse", "Ixazomib", 120, 33, 52,
  "Relapse", "Placebo", 80, 48, 60, "Relapse", "Placebo", 120, 33, 52,
  "PFS", "Ixazomib", 80, 60, 74, "PFS", "Ixazomib", 120, 45, 66,
  "PFS", "Placebo", 80, 50, 66, "PFS", "Placebo", 120, 38, 56
)
band_cmp <- bands |>
  left_join(surv_long |> mutate(endpoint = as.character(endpoint)) |> select(endpoint, arm, week, surv),
    by = c("endpoint", "arm", "week")
  ) |>
  mutate(sim = 100 * surv)
stopifnot(nrow(band_cmp) == 12, !anyNA(band_cmp$sim))
band_cmp |>
  select(endpoint, arm, week, sim, lo, hi) |>
  dplyr::rename(
    "Endpoint" = endpoint, "Arm" = arm, "Week" = week, "Simulated (%)" = sim,
    "Paper band low (%)" = lo, "Paper band high (%)" = hi
  ) |>
  knitr::kable(digits = 1, caption = "Population-averaged survival against the simulated 95% bands of Figure 2b-d.")
# The digitised band edges carry about 2-3 points of reading error; 4 points
# of slack keeps the gate from tripping on that while a wrong hazard
# parameter (time unit, sign, centring) moves these values by 10-40 points.
stopifnot(all(band_cmp$sim > band_cmp$lo - 4 & band_cmp$sim < band_cmp$hi + 4))
pfs_ratio <- pfs_tab$sim_median_wk[pfs_tab$arm == "Ixazomib"] /
  pfs_tab$sim_median_wk[pfs_tab$arm == "Placebo"]
# Ixazomib must lengthen PFS: a weekly-AUC coefficient read per ug*h/mL would
# remove the drug effect and give a ratio near 1.
stopifnot(pfs_ratio > 1.3)
```

Dropout, relapse and PFS all fall inside the paper's simulated bands. The
simulated median PFS is about 30% longer than observed on placebo and longer
still with ixazomib, the same over-prediction the paper's Figure 2d shows.

# Platelet count (Figure 4 and Supplementary Figure S14)

```{r platelets}
mod_plt <- readModelDb("Srimani_2022_ixazomib_platelets")
plt_times <- seq(0, 24 * 7 * 24, by = 24 * 3.5)
plt_ev <- build_events(cohort, plt_times, "circ", ncycle = 6, len_doses = TRUE)
eta_plt <- draw_etas(mod_plt, nrow(cohort), seed = 201)
plt_sim <- solve_ind(mod_plt, plt_ev, eta_plt) |>
  left_join(cohort |> select(id, arm), by = "id") |>
  mutate(week = time / hr_per_week)
plt_sum <- plt_sim |>
  group_by(arm, week) |>
  summarise(median = median(PLT), p10 = quantile(PLT, 0.1), p90 = quantile(PLT, 0.9), .groups = "drop")
ggplot(plt_sum, aes(week)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = median), linewidth = 1) +
  geom_hline(yintercept = c(50, 75), linetype = "dashed") +
  facet_wrap(~arm) +
  labs(
    x = "Time (weeks)", y = expression("Platelet count (" * 10^9 * "/L)"),
    title = "Replicates Figure 4 and Supplementary Figure S14 of Srimani 2022 (first 6 cycles)",
    caption = "Model-predicted median and 10th-90th percentiles without residual error."
  )
```

In the placebo arm of Supplementary Figure S14 the observed median falls from
about 195 to about 150-165 x 10^9/L in weeks 2-3 of each cycle and recovers to
about 200 in the rest week; Figure 4 shows the same pattern with ixazomib, a few
percent lower. The typical-value check below pins that cycle.

```{r platelet-check}
typ_ev <- data.frame(id = 1L, time = c(len_dose_times(1), seq(0, 28 * 24, by = 12)))
typ_ev$amt <- c(rep(25, 21), rep(0, nrow(typ_ev) - 21))
typ_ev$evid <- c(rep(1L, 21), rep(0L, nrow(typ_ev) - 21))
typ_ev$cmt <- c(rep("depot_kpd", 21), rep("circ", nrow(typ_ev) - 21))
typ_ev$BSA <- 1.87
typ_ev$PLT_BASE <- 197
typ <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod_plt), typ_ev[order(typ_ev$time, -typ_ev$evid), ],
  returnType = "data.frame"
))
typ_nadir <- min(typ$PLT)
typ_day28 <- typ$PLT[typ$time == 28 * 24]
c(baseline = typ$PLT[1], nadir = typ_nadir, day28 = typ_day28)
# Deterministic typical placebo cycle: 203 -> ~151 -> ~196 x 10^9/L. Driving
# the LenDex effect by the K-PD amount instead of its rate gives a nadir near
# 26, and a slope transcribed per week instead of per hour gives ~0.
stopifnot(typ_nadir > 140, typ_nadir < 165, typ_day28 > 185)
```

# Diarrhea and rash (Discrete-time Markov models)

## Transition matrices without drug, time or covariate effects

With no ixazomib, no explicit time effect and the reference covariates, the
model's transition probabilities must equal the matrices printed in Appendix S1
(Supplementary Equation S3 for diarrhea, IMiD-exposed; Supplementary Equation 14
for rash, non-Asian). The printed matrices list the destination grade by row
(3, 2, 1, 0) and the current grade by column.

```{r matrices}
mod_dia <- readModelDb("Srimani_2022_ixazomib_diarrhea")
mod_rash <- readModelDb("Srimani_2022_ixazomib_rash")
base_ev <- data.frame(id = 1L, time = 1000, amt = 0, evid = 0L, cmt = "central", BSA = 1.87, PRIOR_IMMUNOMODULATORY_DRUG = 1, RACE_ASIAN = 0)
pnames <- paste0("p", rep(0:3, each = 4), rep(0:3, 4))
to_matrix <- function(s) matrix(unlist(s[1, pnames]), 4, 4, byrow = TRUE)
dia_base <- to_matrix(rxode2::rxSolve(rxode2::zeroRe(mod_dia), base_ev,
  params = c(lemax_time_g10 = -Inf, e_week1_g01 = 0), returnType = "data.frame"
))
rash_base <- to_matrix(rxode2::rxSolve(rxode2::zeroRe(mod_rash), base_ev,
  params = c(lemax_time_g10 = -Inf, e_transient_g0 = 0), returnType = "data.frame"
))
# Supplementary Equations S3 and 14, rows = destination grade 3, 2, 1, 0;
# columns = current grade 3, 2, 1, 0.
s3 <- matrix(c(
  0.966, 0.000637, 0.000236, 0.000215,
  0.00515, 0.856, 0.00108, 0.00126,
  0.00308, 0.0171, 0.662, 0.00357,
  0.0258, 0.126, 0.337, 0.995
), 4, 4, byrow = TRUE)
s14 <- matrix(c(
  0.843, 0.00260, 0.00128, 0.0000361,
  0.0143, 0.799, 0.00103, 0.000134,
  0.00714, 0.0124, 0.597, 0.000459,
  0.136, 0.186, 0.401, 0.999
), 4, 4, byrow = TRUE)
# Reorder the published matrices to current grade (row) 0-3 by destination
# grade (column) 0-3.
from_pub <- function(m) t(m)[4:1, 4:1]
round(dia_base, 5)
round(rash_base, 5)
rel_err_dia <- abs(dia_base / from_pub(s3) - 1)
rel_err_rash <- abs(rash_base / from_pub(s14) - 1)
# Same parameters, same formula: only the 3-significant-figure rounding of
# the printed matrices and of the tables separates the two sides. The largest
# gaps are both grade-2-to-3 cells (diarrhea 4%, rash 3%), where the rounded
# B23 (2.21, 1.99) enters through exp(). A mis-transcribed B parameter moves
# its cells by tens of percent.
c(diarrhea = max(rel_err_dia), rash = max(rel_err_rash))
stopifnot(max(rel_err_dia) < 0.05, max(rel_err_rash) < 0.05)
```

## Transitions out of grade 0 versus weekly AUC (Supplementary Figures S10 and S12)

The ixazomib effect is linear in the weekly AUC on the logits out of grade 0.
Shifting `b01` by `slp_aucwk_g0 * AUC` for a drug-free patient traces the curves
of Supplementary Figures S10 and S12.

```{r auc-curves}
auc_grid <- seq(0, 2500, by = 250)
grid_ev <- data.frame(id = seq_along(auc_grid), time = 1000, amt = 0, evid = 0L, cmt = "central", BSA = 1.87, PRIOR_IMMUNOMODULATORY_DRUG = 1, RACE_ASIAN = 0)
curve_one <- function(mod, b01, slp, label, asian = 0) {
  ev <- grid_ev
  ev$RACE_ASIAN <- asian
  p <- data.frame(id = ev$id, b01 = b01 + slp * auc_grid, lemax_time_g10 = -Inf)
  if ("e_week1_g01" %in% rxode2::rxode(mod)$iniDf$name) p$e_week1_g01 <- 0
  if ("e_transient_g0" %in% rxode2::rxode(mod)$iniDf$name) p$e_transient_g0 <- 0
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, params = p, returnType = "data.frame"))
  tibble(model = label, AUC = auc_grid, `0 to 0` = s$p00, `0 to 1` = s$p01, `0 to 2` = s$p02, `0 to 3` = s$p03)
}
curves <- bind_rows(
  curve_one(mod_dia, -5.28, 0.000715, "Diarrhea"),
  curve_one(mod_rash, -7.37, 0.000929, "Rash, non-Asian"),
  curve_one(mod_rash, -7.37, 0.000929, "Rash, Asian", asian = 1)
) |>
  pivot_longer(-c(model, AUC), names_to = "transition", values_to = "p")
ggplot(curves, aes(AUC, 100 * p, colour = model)) +
  geom_line(linewidth = 1) +
  facet_wrap(~transition, scales = "free_y") +
  labs(
    x = "Weekly AUC (ng*h/mL)", y = "Transition probability (%)", colour = NULL,
    title = "Replicates Supplementary Figures S10 and S12 of Srimani 2022"
  )
end_vals <- curves |>
  filter(AUC == 2500, transition != "0 to 0") |>
  pivot_wider(names_from = transition, values_from = p)
knitr::kable(end_vals |> mutate(across(-model, ~ 100 * .x)), digits = 3, caption = "Transition probabilities (%) at a weekly AUC of 2500 ng*h/mL.")
# Read off the figures by the maintainers at AUC = 2500: diarrhea 0 to 1
# ~2.05%, 0 to 2 ~0.73%; rash Asian 0 to 1 ~1.5%, non-Asian ~0.45%.
stopifnot(
  abs(100 * end_vals$`0 to 1`[end_vals$model == "Diarrhea"] - 2.05) < 0.2,
  abs(100 * end_vals$`0 to 2`[end_vals$model == "Diarrhea"] - 0.73) < 0.1,
  abs(100 * end_vals$`0 to 1`[end_vals$model == "Rash, Asian"] - 1.5) < 0.15,
  abs(100 * end_vals$`0 to 1`[end_vals$model == "Rash, non-Asian"] - 0.45) < 0.08
)
```

## One-year grade prevalence for typical patients

The paper reports model-predicted prevalence after one year: for diarrhea,
grade 3 in 1.2% of ixazomib-treated IMiD-exposed patients, 0.1% of IMiD-naive
patients, and 0.5% versus 0.1% on placebo (Results); for rash, grade 0/1/2/3 in
97.3/2.4/0.2/0.1% of non-Asian and 91.5/7.5/0.7/0.3% of Asian ixazomib-treated
patients, and grade 0 in 99.1% versus 97.3% of non-Asian and Asian placebo
patients (Supplement 2.2.2). The supplement calls the one-year values steady
state, and they are reproduced when ixazomib is given every week without the
rest week, so that the weekly AUC settles at F x Dose / CL = 1247 ng*h/mL; on
the trial schedule (3 of 4 weeks) the ixazomib-arm grade 1-3 values come out
about a quarter lower. The typical-value Markov chains below use the weekly
transition probabilities of a typical patient on that continuous schedule.

```{r one-year}
week_times <- seq(1, 52) * hr_per_week
typ_cohort <- tidyr::expand_grid(arm = c("Ixazomib", "Placebo"), PRIOR_IMMUNOMODULATORY_DRUG = c(1, 0), RACE_ASIAN = c(0, 1)) |>
  mutate(id = row_number(), BSA = 1.87, MCPROT = 23, TUM_CYTOGENETIC_HIGH_RISK = 0, PLT_BASE = 197)
# 4 mg every week (no rest week) for 53 weeks.
typ_ev <- bind_rows(
  typ_cohort |> filter(arm == "Ixazomib") |> select(id) |>
    tidyr::crossing(time = seq(0, 52) * hr_per_week) |> mutate(amt = 4, evid = 1L, cmt = "depot"),
  typ_cohort |> select(id) |> tidyr::crossing(time = week_times) |> mutate(amt = 0, evid = 0L, cmt = "central")
) |>
  left_join(typ_cohort, by = "id") |>
  arrange(id, time, desc(evid))
one_year <- function(mod, label) {
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), typ_ev, atol = 1e-6, rtol = 1e-6, returnType = "data.frame"))
  s |>
    group_by(id) |>
    group_modify(~ {
      x <- markov_expected(.x)
      tibble(G0 = x[52, 1], G1 = x[52, 2], G2 = x[52, 3], G3 = x[52, 4])
    }) |>
    ungroup() |>
    left_join(typ_cohort |> select(id, arm, PRIOR_IMMUNOMODULATORY_DRUG, RACE_ASIAN), by = "id") |>
    mutate(model = label)
}
dia_1y <- one_year(mod_dia, "Diarrhea") |> filter(RACE_ASIAN == 0)
rash_1y <- one_year(mod_rash, "Rash") |> filter(PRIOR_IMMUNOMODULATORY_DRUG == 1)

dia_tab <- dia_1y |>
  transmute(arm, IMiD = ifelse(PRIOR_IMMUNOMODULATORY_DRUG == 1, "exposed", "naive"), sim_G3 = 100 * G3) |>
  mutate(paper_G3 = c(1.2, 0.1, 0.5, 0.1)[match(paste(arm, IMiD), c("Ixazomib exposed", "Ixazomib naive", "Placebo exposed", "Placebo naive"))])
dia_tab |>
  dplyr::rename("Arm" = arm, "Prior IMiD" = IMiD, "Simulated grade 3 (%)" = sim_G3, "Paper grade 3 (%)" = paper_G3) |>
  knitr::kable(digits = 2, caption = "Diarrhea grade 3 prevalence after one year.")

rash_tab <- rash_1y |>
  transmute(arm, race = ifelse(RACE_ASIAN == 1, "Asian", "non-Asian"), G0 = 100 * G0, G1 = 100 * G1, G2 = 100 * G2, G3 = 100 * G3)
rash_paper <- tibble(
  arm = c("Ixazomib", "Ixazomib", "Placebo", "Placebo"), race = c("non-Asian", "Asian", "non-Asian", "Asian"),
  paper_G0 = c(97.3, 91.5, 99.1, 97.3), paper_G1 = c(2.4, 7.5, NA, NA)
)
rash_cmp <- left_join(rash_tab, rash_paper, by = c("arm", "race"))
rash_cmp |>
  dplyr::rename("Arm" = arm, "Race" = race, "Sim G0 (%)" = G0, "Sim G1 (%)" = G1, "Sim G2 (%)" = G2, "Sim G3 (%)" = G3, "Paper G0 (%)" = paper_G0, "Paper G1 (%)" = paper_G1) |>
  knitr::kable(digits = 2, caption = "Rash grade prevalence after one year.")
```

```{r one-year-check}
# Deterministic typical-value chains; the gaps left are the rounding of the
# printed percentages. With the slow time effect on the grade 0 -> 1 logit
# instead of on recovery from grade 1, the AUC-free placebo rash values are
# 98.9% / 96.4% grade 0 (paper 99.1% / 97.3%) and this gate fails.
stopifnot(
  all(abs(dia_tab$sim_G3 - dia_tab$paper_G3) < 0.15),
  all(abs(rash_cmp$G0 - rash_cmp$paper_G0) < 0.5),
  all(abs(rash_cmp$G1 - rash_cmp$paper_G1) < 0.5, na.rm = TRUE)
)
```

## Diarrhea and rash prevalence over time (Figure 3 and Supplementary Figure S11)

Individual grade sequences are drawn week by week from each simulated
patient's transition probabilities, with the per-grade random intercepts.

```{r ae-vpc}
ae_weeks <- seq(1, 78)
ae_ev <- build_events(cohort, ae_weeks * hr_per_week, "central", ncycle = 20)
ae_prev <- function(mod, label, seed) {
  etas <- draw_etas(mod, nrow(cohort), seed = seed)
  s <- solve_ind(mod, ae_ev, etas) |> arrange(id, time)
  set.seed(seed + 1)
  s$grade <- markov_walk(s, stats::runif(nrow(s)))
  s |>
    left_join(cohort |> select(id, arm), by = "id") |>
    mutate(week = time / hr_per_week) |>
    count(arm, week, grade) |>
    group_by(arm, week) |>
    mutate(pct = 100 * n / sum(n)) |>
    ungroup() |>
    tidyr::complete(arm, week, grade = 0:3, fill = list(n = 0, pct = 0)) |>
    mutate(model = label)
}
ae <- bind_rows(ae_prev(mod_dia, "Diarrhea", 301), ae_prev(mod_rash, "Rash", 401))
ggplot(ae |> filter(arm == "Ixazomib"), aes(week, pct, colour = model)) +
  geom_line() +
  facet_wrap(~grade, scales = "free_y", labeller = label_both) +
  labs(
    x = "Time (weeks)", y = "Patients with grade (%)", colour = NULL,
    title = "Replicates Figure 3 and Supplementary Figure S11 (ixazomib arm)"
  )
ae_pts <- ae |>
  filter(arm == "Ixazomib", week %in% c(4, 40, 76)) |>
  select(model, week, grade, pct) |>
  pivot_wider(names_from = grade, values_from = pct, names_prefix = "grade ")
knitr::kable(ae_pts, digits = 1, caption = "Simulated prevalence (%) in the ixazomib arm.")
```

Figure 3 shows observed grade-0 diarrhea in the ixazomib arm falling from
about 90% to 80% over 75 weeks while grade 1 rises from about 5% to 13%;
Supplementary Figure S11 shows rash concentrated in the first weeks (grade 0
about 87% at week 4, about 95% from week 40). The simulated sequences follow
both shapes.

```{r ae-check}
g <- function(model, week, grade) ae_pts[[paste("grade", grade)]][ae_pts$model == model & ae_pts$week == week]
# 200 patients per arm drawn with a fixed R seed: the margins admit the
# Monte-Carlo spread of a 200-patient prevalence (a few percentage points).
stopifnot(
  g("Diarrhea", 76, 0) < g("Diarrhea", 4, 0),
  g("Diarrhea", 76, 0) > 65, g("Diarrhea", 76, 0) < 92,
  g("Rash", 4, 0) < g("Rash", 40, 0),
  g("Rash", 40, 0) > 88
)
```

# Assumptions and deviations

* **Weekly AUC.** The paper integrates the PK model to a weekly AUC without
  stating the window. The models use the AUC over the 168 h ending at the
  evaluation time, computed with `delay()`. This needs a dense solver, and an
  occasional patient fails at the default tolerances; `atol = rtol = 1e-6`
  avoids it.
* **Weekly-AUC units.** Table 2 labels the PFS coefficient `ml/(ug.h)`, but its
  "untransformed" value exp(-0.00108) = 0.999 uses the estimate as printed, and
  Table 3 converts the diarrhea slope from 0.000715 to 0.715 per ug*h/mL while
  Supplementary Figures S10 and S12 plot AUC in ng*h/mL. All AUC coefficients
  are therefore read per ng*h/mL. On a per-ug reading the PFS model would carry
  no ixazomib effect, against the trial's PFS difference.
* **Nadir time.** `T_MPROTEIN_NADIR` is a covariate, because the paper derived it from
  each patient's observed profile. The vignette draws it from the relapse
  hazard. The 4.71% of patients without an estimable nadir (`p_tnadir0`) were
  excluded from the paper's PK/PD fit; the vignette does not simulate them
  separately.
* **Relapse parameters.** Table 2 prints the same estimate, 0.0509, for
  `alpha_IXA` and `alpha_kR,0`, with different standard errors and confidence
  intervals. Both are used as printed. The relapse multiplier `lambda0`
  (exp(-1.23) = 0.292) multiplies a log-logistic hazard and is dimensionless on
  the model's hour time base, although Table 2 converts it to 49.0/week.
* **Platelet loss rate.** The paper gives no separate `kout`. Because the
  system is normalised to baseline with production `kIn`, the circulating loss
  rate equals `kIn`, and the precursor states start at `kIn / kprp`.
* **Lenalidomide K-PD driver.** "Len conc." is read as the K-PD effect rate
  `kLEN * A` (mg/h; Jacqmin 2007). Driving the effect by the amount `A` itself
  drops the placebo platelet count to about 13% of baseline within a cycle,
  against the 20-25% dip of Supplementary Figure S14. Lenalidomide doses are
  given as 25 mg; patients with reduced creatinine clearance received 10 mg.
* **Markov time effects.** The functional form and target of the slow time
  effect (`P1|0T`, `K1|0T`) are not printed, and the main text describes it as a
  "slow transition from grade 0 to grade 1". Appendix S1 writes `p_i|j` for the
  move from grade i to grade j, which makes `P1|0` the grade-1-to-0 recovery and
  `P0|1` the grade-0-to-1 onset (the diarrhea `P0|1I`, "rapid diarrhea onset in
  week 1"). The models therefore add `exp(P1|0T) * (1 - exp(-K * t))` to the
  grade >= 1 logit from grade 1, which slows recovery and raises grade 1
  prevalence over time, the effect both models are said to describe. This
  placement and form reproduce every one-year prevalence the paper prints;
  putting the same term on the grade-0-to-1 logit misses the AUC-free placebo
  rash values and roughly doubles the late grade-1 diarrhea of Figure 3. The
  diarrhea `P0|1I` term is added to the grade-0-to-1 logit during the first
  168 h. The rash transient `ETIME * exp(-KTIME * t)` shifts every logit out of
  grade 0, which gives the early grade 2-3 rash of Supplementary Figure S11.
* **Typical exposure behind the one-year values.** The paper's one-year
  ("steady state") predictions are reproduced with uninterrupted weekly dosing
  (weekly AUC 1247 ng*h/mL), not with the 3-of-4-week trial schedule.
* **Covariate reference groups.** The printed diarrhea matrix is the
  IMiD-exposed one, so the IMiD shift applies to naive patients
  (`1 - PRIOR_IMMUNOMODULATORY_DRUG`); the M-protein `Yss` effect is coded the same way.
* **Likelihood.** The Markov models emit transition probabilities only; their
  categorical likelihood conditioned on the previous grade cannot be written as
  an nlmixr2 observation, so these files are for simulation.
* **M-protein upper tail.** The simulated on-study 95th percentile of
  M-protein is about 1.5 times the paper's simulated band (see the Figure 2a
  section); the median is reproduced.
* **Virtual cohort.** BSA, baseline M-protein and baseline platelet count are
  drawn from log-normal or normal distributions matched to the reported
  medians; the paper reports no BSA distribution.
