---
title: "Bictegravir drug-drug interactions with CYP3A and UGT1A1 perpetrators (Stader 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bictegravir drug-drug interactions with CYP3A and UGT1A1 perpetrators (Stader 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Stader_2021_bictegravir_ddi_pbpk"))$reference`
* Description: `r rxode2::rxode(readModelDb("Stader_2021_bictegravir_ddi_pbpk"))$description`
* Article: <https://doi.org/10.1002/cpt.2221> (open access, PMC8597021)
* Supplement: Supporting Information s001 (Tables S1-S7, Figures S1-S9) and
  s002 (`CPT_Matlab_Code`, the Matlab source of the PBPK framework with its
  drug library `Drug/DrugLibrary/*.m`).

Stader, Battegay and Marzolini used their in-house whole-body PBPK framework
(Matlab 2017a) to predict how inhibitors and inducers of CYP3A and UGT1A1
change the exposure of the integrase inhibitor bictegravir. They first built
two new perpetrator models (voriconazole and cobicistat, Table S1), verified
the framework against five clinical DDI studies of bictegravir (Table 1,
Figure 1), and then simulated 12 single-perpetrator and 8 two-perpetrator
scenarios (Figures 2 and 3, Table S7).

The paper describes the framework in a paragraph; the equations, the
virtual-population generator and every drug's inputs are in the deposited
Matlab code, which is the source of this implementation. The same framework
with bictegravir alone is packaged as `Stader_2021_bictegravir_pbpk` (Stader
et al. 2021, doi:10.1002/cpt.2178, whose supplement deposits the identical
code). This model adds what a DDI run needs: two co-administered drugs, and
enzyme-turnover states through which the drugs interact.

# How a framework DDI run maps onto the model

A framework run with N drugs solves 2N copies of the whole-body model. The
first N "alone" copies simulate each drug with only its own interaction
parameters active. In the last N "DDI" copies every drug sees all the others,
but the inhibitor and inducer concentrations it sees are read from the other
drugs' *alone* copies, not from their DDI copies (`PBPK_PreProcessing.m` and
`PBPK_ODE_solution.m`). For a victim that does not itself inhibit or induce,
such as bictegravir, the published DDI magnitude therefore needs three
things: the perpetrators' alone copies, the victim's DDI copy and the
victim's alone copy (the control arm). The model carries them as:

* the **victim** block (bare state names, output `Cc`): its DDI copy when
  `ddi_on = 1`, its alone copy (the control arm) when `ddi_on = 0`;
* the **perpetrator** and **second perpetrator** blocks (suffixes
  `_perpetrator`, `_perpetrator2`, outputs `Cc_perpetrator`,
  `Cc_perpetrator2`): their alone copies, always.

`on_perpetrator` and `on_perpetrator2` say whether a slot is part of the run
at all. The flags matter even for undosed drugs because of the first as-run
behaviour below. Every drug's inputs are `ini()` values, and the model's
`drugLibrary` metadata holds the as-run inputs of all 13 drugs in the
deposited library, so any scenario is a matter of loading those values into
the slots (as data columns, below). This follows the parameterised-perpetrator
design of `CherkaouiRbati_2017_midazolam_qsp`.

Four behaviours of the deposited code matter for reproducing its output, and
the model keeps all of them.

1. **Hepatic CYP synthesis is suppressed by every drug without an induction
   parameter.** Hepatic CYP induction is coded as
   `(IndMax - 1) * C / (IC50 + C)`. A drug that sets no `IndMax` for an enzyme
   leaves it at 0, and an `IC50` of 0 is replaced by 1 uM, so each such drug
   contributes `-C / (1 + C)` to the synthesis of every hepatic CYP. In an
   alone copy these terms use the copy's own drug concentration but each other
   drug's unbound fraction. (Intestinal and UGT induction use `IndMax * C /
   (IC50 + C)` and are unaffected.)
2. **Drug 1's unbound fraction enters every drug's albumin-binding term.**
   `PBPK_Drug_distribution.m` computes the Rodgers and Rowland `KaPR` of
   drug `d` from `DRUG.fup(d)`, a linear index into the subject-by-drug array
   of albumin-adjusted unbound fractions, which reads drug 1's unbound
   fraction of virtual subject `d` (strong bases, which use `KaAP`, are
   unaffected). Drug 1 is the run's drug with the lowest index in the
   framework's drug library, because the drug-parameter arrays are rebuilt in
   library order (`PBPK_Drug_PostProcessing.m`). `fupref_perpetrator` and
   `fupref_perpetrator2` say whether a perpetrator is drug 1. The model uses
   each subject's own drug-1 unbound fraction, which is exact when all
   subjects of a run are identical (as in the checks below); in a drawn
   cohort the framework would use another subject's value.
3. **Microsomal protein variability collapses when ritonavir is the last
   drug of the run.** `PBPK_Population_Liver.m` sets the MPPGL CV to 2.3%
   instead of 46% when the last drug of the run is ritonavir. This only
   affects the random draws (scale `etamppgl` and `etamppgl_redraw` by
   2.3/46 when drawing a cohort for such a run); the typical-subject
   simulations below are unaffected.
4. **Dosing lags.** Ritonavir and efavirenz carry `LagTime = 1` h, which
   `PBPK_StudyDesign.m` uses to place every oral dose 1 h after its nominal
   time. The model applies it as `alag()` on the stomach states.

```{r model}
mod <- readModelDb("Stader_2021_bictegravir_ddi_pbpk")
ui <- rxode2::rxode(mod)
mod_tv <- rxode2::zeroRe(ui)
lib <- ui$meta$drugLibrary

# Framework drug-library order (PBPK_DefineParameters.m). Clarithromycin's
# drug file is deposited but not registered in the library; it is appended.
lib_index <- c(
  midazolam = 1, ketoconazole = 2, voriconazole = 3, nilotinib = 4,
  rifampicin = 5, bictegravir = 6, atazanavir = 7, darunavir = 8,
  ritonavir = 9, cobicistat = 10, efavirenz = 11, etravirine = 12,
  clarithromycin = 13
)

slot_values <- function(drug, suffix) {
  stats::setNames(lib[[drug]], paste0(lib$parameter, suffix))
}

# Parameters of one framework run, as a one-row data frame. Absent slots are
# filled with the victim's values and switched off.
scenario_params <- function(victim, perpetrator = NA, perpetrator2 = NA, ddi_on = 1) {
  run <- c(victim, perpetrator, perpetrator2)
  run <- run[!is.na(run)]
  first <- run[which.min(lib_index[run])]
  p <- c(
    slot_values(victim, ""),
    slot_values(if (is.na(perpetrator)) victim else perpetrator, "_perpetrator"),
    slot_values(if (is.na(perpetrator2)) victim else perpetrator2, "_perpetrator2"),
    ddi_on = ddi_on,
    on_perpetrator = as.numeric(!is.na(perpetrator)),
    on_perpetrator2 = as.numeric(!is.na(perpetrator2)),
    fupref_perpetrator = as.numeric(!is.na(perpetrator) && first == perpetrator),
    fupref_perpetrator2 = as.numeric(!is.na(perpetrator2) && first == perpetrator2)
  )
  as.data.frame(as.list(p))
}

# Oral doses into a slot's stomach, `amt` mg every `ii` h from `start` to `end`.
oral_doses <- function(cmt, amt, ii, start, end) {
  if (is.na(amt)) {
    return(NULL)
  }
  data.frame(time = seq(start, end, by = ii), amt = amt, evid = 1L, cmt = cmt, rate = 0)
}

# One id per row of `runs`: its dose records, observation records on the
# victim's venous state (Cc and the perpetrator outputs are returned at every
# observation row), covariates, and the run's drug parameters as data columns
# (rxode2 reads a data column named like a parameter in preference to the
# ini() value).
build_data <- function(runs, obs_times) {
  bind_rows(lapply(seq_len(nrow(runs)), function(i) {
    r <- runs[i, ]
    ev <- bind_rows(
      oral_doses("stomach", r$amt, r$ii, r$start, r$end),
      oral_doses("stomach_perpetrator", r$amt_p, r$ii_p, r$start_p, r$end_p),
      oral_doses("stomach_perpetrator2", r$amt_p2, r$ii_p2, r$start_p2, r$end_p2),
      data.frame(time = obs_times, amt = 0, evid = 0L, cmt = "venous", rate = 0)
    )
    cbind(
      id = i, ev, AGE = r$AGE, SEXF = r$SEXF, HT = r$HT, WT = r$WT,
      etagastric = r$etagastric,
      scenario_params(r$victim, r$perpetrator, r$perpetrator2, r$ddi_on)
    )
  })) |>
    arrange(id, time, desc(evid))
}

# A typical subject of the paper's population: the generator's mean height
# and weight for the given age and sex, every random draw at its median.
typical_subject <- function(sexf, age = 35) {
  ht <- -0.0039 * age^2 + 0.238 * age - 12.5 * sexf + 176
  data.frame(
    AGE = age, SEXF = sexf, HT = ht,
    WT = -0.0039 * age^2 + 1.12 * ht + 0.611 * age - 0.424 * sexf - 137
  )
}
```

# Population

A PBPK model has no fitted population; its population is a virtual one drawn
by the framework's generator from age- and sex-dependent regressions. Each DDI
scenario was simulated in 100 virtual individuals (10 trials of 10, 50% women)
aged 20 to 50 years (Methods). The model was verified against clinical DDI
studies of bictegravir with voriconazole, darunavir/cobicistat, atazanavir,
atazanavir/cobicistat and rifampicin (data provided by Gilead Sciences;
Table 1), and the new voriconazole and cobicistat models against published
studies in healthy volunteers (Table S2). The same information is available
from `readModelDb("Stader_2021_bictegravir_ddi_pbpk")()$population`.

The model takes age, sex, height and weight as covariates. The generator's
remaining per-subject random draws (organ weights, blood flows, haematocrit,
albumin, microsomal protein, enzyme abundances, gut transit times) are fixed
etas whose variances are the squared CVs of the deposited code, so
`rxode2::rxSolve(mod, ...)` with random effects draws a virtual cohort and
`rxode2::zeroRe(mod)` gives the typical subject.

# Source trace

Every `ini()` value carries a comment giving the drug file and line it comes
from (`Drug/DrugLibrary/<drug>.m`); the voriconazole and cobicistat values
also match Table S1, which reprints them. The `drugLibrary` metadata holds the
same as-run values for all 13 drugs. As-run means the values the framework
actually uses: a Km, Kapp, IC50 or tissue scalar of 0 is replaced by 1, and a
Ki of 0 means no inhibition.

```{r source-trace-drugs}
lib |>
  select(
    parameter, bictegravir, voriconazole, cobicistat, ketoconazole, clarithromycin,
    ritonavir, darunavir, atazanavir, nilotinib, rifampicin, efavirenz, midazolam
  ) |>
  filter(if_any(-parameter, ~ .x != 0)) |>
  knitr::kable(
    caption = paste(
      "As-run drug inputs of the drugs used in the paper (drugLibrary metadata).",
      "Units are those of the corresponding ini() labels."
    ),
    digits = 4
  )
```

```{r source-trace-equations}
tibble::tribble(
  ~Component, ~Source,
  "Virtual individual: height, weight, BSA, organ weights, blood and lymph flows, GFR", "PBPK_Population_Demographics.m, PBPK_Population_Tissue.m",
  "Hepatic microsomal protein and enzyme abundances (CYPs, UGTs)", "PBPK_Population_Liver.m",
  "Gut segment lengths, volumes, transit times and intestinal CYP3A4", "PBPK_Population_GIT.m",
  "Ionisation, plasma binding, Rodgers and Rowland tissue partitioning", "PBPK_Drug_distribution.m",
  "Compartmental absorption and transit (effective permeability)", "PBPK_Drug_absorption.m",
  "Renal and biliary clearance, hepatic metabolism", "PBPK_Drug_elimination.m",
  "Whole-body ODEs, enzyme turnover, inhibition, inactivation, induction", "PBPK_ODE_solution.m (rhs_function)",
  "Oral dosing times and lag", "PBPK_StudyDesign.m",
  "Reported concentration (venous blood state times molecular weight)", "PBPK_ExtractConcentration.m",
  "Voriconazole and cobicistat drug inputs", "Table S1; Drug/DrugLibrary/voriconazole.m, cobicistat.m",
  "All other drug inputs", "Drug/DrugLibrary/<drug>.m",
  "Perpetrator doses of the prospective scenarios", "Table S3",
  "Scenario design (14 days of perpetrator, bictegravir 50 mg daily from day 7)", "Methods, 'Simulations of DDI scenarios'"
) |>
  knitr::kable(caption = "Source of each model component (files of the deposited s002 code).")
```

# Check against the deposited code

The maintainers ran the deposited Matlab code unmodified in GNU Octave 9.2
(with `unique()` called with its `'first'` option, which is Matlab's default
and not Octave's) for ten framework runs covering every perpetrator of the
paper: one deterministic 35-year-old man (the generator's mean height and
weight, variability switched off, which fixes the gastric emptying time at
0.25 h), victim dosed at 72 and 96 h, perpetrators dosed from time 0 for five
days at the Table S3 doses. The table below holds the framework's venous
concentrations (ng/mL) at three times of each run, for the victim's DDI and
alone (control) copies and the perpetrators' alone copies. The model is
solved for the same runs and must reproduce every value; the two sides use
identical inputs, so any difference is numerical error and the bound is
tight.

```{r deposited-code-values}
deposited <- read.csv(text = "
run,output,time,conc
atvcob,perpetrator,78.02510,790.4512
atvcob,perpetrator,89.97490,110.3191
atvcob,perpetrator,107.94979,343.1772
atvcob,victim_control,78.02510,1693.568
atvcob,victim_control,89.97490,491.0577
atvcob,victim_control,107.94979,1096.204
atvcob,victim_ddi,78.02510,1889.047
atvcob,victim_ddi,89.97490,549.0376
atvcob,victim_ddi,107.94979,1097.599
bcla,perpetrator,77.94958,973.6385
bcla,perpetrator,89.94958,973.9266
bcla,perpetrator,108.00000,457.9473
bcla,victim_control,77.94958,1846.255
bcla,victim_control,89.94958,1097.162
bcla,victim_control,108.00000,1924.851
bcla,victim_ddi,77.94958,2039.221
bcla,victim_ddi,89.94958,1464.838
bcla,victim_ddi,108.00000,2616.4
befv,perpetrator,78.02183,4140.864
befv,perpetrator,89.97380,2461.382
befv,perpetrator,108.04803,3279.968
befv,victim_control,78.02183,1840.467
befv,victim_control,89.97380,1095.984
befv,victim_control,108.04803,1920.747
befv,victim_ddi,78.02183,1582.706
befv,victim_ddi,89.97380,680.649
befv,victim_ddi,108.04803,1187.832
bnil,perpetrator,78.02510,4869.894
bnil,perpetrator,89.97490,4200.544
bnil,perpetrator,107.94979,5714.753
bnil,victim_control,78.02510,1800.336
bnil,victim_control,89.97490,1030.198
bnil,victim_control,107.94979,1819.59
bnil,victim_ddi,78.02510,2241.765
bnil,victim_ddi,89.97490,1928.512
bnil,victim_ddi,107.94979,3669.464
bnilrif,perpetrator,78.02510,5146.927
bnilrif,perpetrator,89.97490,4598.529
bnilrif,perpetrator,107.94979,6373.78
bnilrif,perpetrator2,78.02510,8689.898
bnilrif,perpetrator2,89.97490,689.4053
bnilrif,perpetrator2,107.94979,2401.016
bnilrif,victim_control,78.02510,1800.534
bnilrif,victim_control,89.97490,1030.982
bnilrif,victim_control,107.94979,1822.767
bnilrif,victim_ddi,78.02510,2075.085
bnilrif,victim_ddi,89.97490,1484.777
bnilrif,victim_ddi,107.94979,2748.524
brtv,perpetrator,78.02183,178.805
brtv,perpetrator,89.97380,88.49902
brtv,perpetrator,108.04803,137.638
brtv,victim_control,78.02183,1841.003
brtv,victim_control,89.97380,1098.278
brtv,victim_control,108.04803,1930.288
brtv,victim_ddi,78.02183,1886.13
brtv,victim_ddi,89.97380,1141.984
brtv,victim_ddi,108.04803,1992.338
bvor,perpetrator,78.02510,7138.102
bvor,perpetrator,89.97490,7976.37
bvor,perpetrator,107.94979,13125.04
bvor,victim_control,78.02510,1791.775
bvor,victim_control,89.97490,1014.575
bvor,victim_control,107.94979,1786.448
bvor,victim_ddi,78.02510,2014.2
bvor,victim_ddi,89.97490,1421.396
bvor,victim_ddi,107.94979,2535.905
bvorefv,perpetrator,78.02183,19747.98
bvorefv,perpetrator,89.97380,24119.54
bvorefv,perpetrator,108.04803,34277.59
bvorefv,perpetrator2,78.02183,3452.792
bvorefv,perpetrator2,89.97380,1858.085
bvorefv,perpetrator2,108.04803,2549.92
bvorefv,victim_control,78.02183,1792.257
bvorefv,victim_control,89.97380,1015.416
bvorefv,victim_control,108.04803,1781.32
bvorefv,victim_ddi,78.02183,1751.378
bvorefv,victim_ddi,89.97380,910.3617
bvorefv,victim_ddi,108.04803,1590.754
drvcob,perpetrator,78.02510,798.1882
drvcob,perpetrator,89.97490,113.077
drvcob,perpetrator,107.94979,348.108
drvcob,victim_control,78.02510,626.9302
drvcob,victim_control,89.97490,15.00125
drvcob,victim_control,107.94979,104.1511
drvcob,victim_ddi,78.02510,2684.229
drvcob,victim_ddi,89.97490,347.536
drvcob,victim_ddi,107.94979,1034.989
mdzvor,perpetrator,78.02510,7250.892
mdzvor,perpetrator,89.97490,8006.023
mdzvor,perpetrator,107.94979,13107.23
mdzvor,victim_control,78.02510,3.273652
mdzvor,victim_control,89.97490,0.1213418
mdzvor,victim_control,107.94979,0.624096
mdzvor,victim_ddi,78.02510,23.52838
mdzvor,victim_ddi,89.97490,2.562598
mdzvor,victim_ddi,107.94979,7.742433
")

deposited_runs <- tibble::tribble(
  ~run, ~victim, ~amt, ~perpetrator, ~amt_p, ~ii_p, ~perpetrator2, ~amt_p2,
  "atvcob", "atazanavir", 400, "cobicistat", 150, 24, NA, NA,
  "bcla", "bictegravir", 50, "clarithromycin", 500, 12, NA, NA,
  "befv", "bictegravir", 50, "efavirenz", 600, 24, NA, NA,
  "bnil", "bictegravir", 50, "nilotinib", 400, 24, NA, NA,
  "bnilrif", "bictegravir", 50, "nilotinib", 400, 24, "rifampicin", 600,
  "brtv", "bictegravir", 50, "ritonavir", 100, 24, NA, NA,
  "bvor", "bictegravir", 50, "voriconazole", 300, 24, NA, NA,
  "bvorefv", "bictegravir", 50, "voriconazole", 300, 24, "efavirenz", 600,
  "drvcob", "darunavir", 800, "cobicistat", 150, 24, NA, NA,
  "mdzvor", "midazolam", 7.5, "voriconazole", 300, 24, NA, NA
) |>
  mutate(
    ii = 24, start = 72, end = 96,
    start_p = 0, end_p = 120 - ii_p,
    ii_p2 = 24, start_p2 = 0, end_p2 = 96
  ) |>
  tidyr::crossing(ddi_on = c(1, 0)) |>
  bind_cols(typical_subject(0)) |>
  # Variability off: gastric emptying 0.25 h, the lower end of U(0.25, 1).
  mutate(etagastric = -40)

deposited_sim <- rxode2::rxSolve(
  mod_tv, build_data(deposited_runs, sort(unique(deposited$time))),
  returnType = "data.frame", cores = 2L
) |>
  mutate(run = deposited_runs$run[id], ddi_on = deposited_runs$ddi_on[id])

port <- bind_rows(
  deposited_sim |>
    filter(ddi_on == 1) |>
    select(run, time, victim_ddi = Cc, perpetrator = Cc_perpetrator, perpetrator2 = Cc_perpetrator2) |>
    pivot_longer(-c(run, time), names_to = "output", values_to = "model"),
  deposited_sim |>
    filter(ddi_on == 0) |>
    transmute(run, time, output = "victim_control", model = Cc)
)
deposited_check <- inner_join(deposited, port, by = c("run", "output", "time")) |>
  mutate(rel_diff = abs(model - conc) / conc)

deposited_check |>
  group_by(run) |>
  summarise(values = n(), max_rel_diff = max(rel_diff), .groups = "drop") |>
  rename("Run" = run, "Values compared" = values, "Max relative difference" = max_rel_diff) |>
  knitr::kable(digits = 8, caption = "Model versus the deposited code executed in Octave.")

stopifnot(
  nrow(deposited_check) == nrow(deposited),
  max(deposited_check$rel_diff) < 1e-4
)
```

The model reproduces all `r nrow(deposited_check)` values to a relative
difference of at most `r sprintf("%.1e", max(deposited_check$rel_diff))`, so it is
the deposited code, including the four behaviours listed above.

# Prospective DDI scenarios (Table S7, Figures 2 and 3)

The paper's 20 scenarios give each perpetrator (or pair) for 14 days at its
Table S3 dose and bictegravir 50 mg once daily from day 7, and report the AUC
ratio (DDI over control) on the first and the seventh day of bictegravir.
Here each scenario is solved for a typical man and a typical woman aged 35
years; the ratio reported is the mean of the two, compared with the paper's
day-1 values (Results text) and day-7 medians (Table S7). A typical subject's
ratio is not a cohort median, but the ratios are driven by the drug inputs
far more than by the physiology, so the comparison locates the model on the
published distribution.

```{r s7-scenarios}
s7 <- tibble::tribble(
  ~scenario, ~perpetrator, ~perpetrator2, ~amt_p, ~amt_p2, ~day1_paper, ~median, ~ci_lo, ~ci_hi,
  "Voriconazole", "voriconazole", NA, 300, NA, 1.21, 1.68, 1.19, 2.53,
  "Ketoconazole", "ketoconazole", NA, 400, NA, 1.21, 1.75, 1.17, 2.63,
  "Clarithromycin", "clarithromycin", NA, 500, NA, 1.21, 1.75, 1.19, 2.61,
  "Cobicistat", "cobicistat", NA, 150, NA, 1.21, 1.55, 1.14, 2.11,
  "Ritonavir", "ritonavir", NA, 100, NA, 1.02, 1.08, 0.62, 1.84,
  "Darunavir+Ritonavir", "darunavir", "ritonavir", 800, 100, NA, 1.19, 0.59, 1.89,
  "Darunavir+Cobicistat", "darunavir", "cobicistat", 800, 150, NA, 1.44, 1.13, 1.96,
  "Atazanavir", "atazanavir", NA, 400, NA, 1.37, 2.53, 1.92, 3.36,
  "Atazanavir+Cobicistat", "atazanavir", "cobicistat", 300, 150, NA, 2.43, 1.71, 3.21,
  "Nilotinib", "nilotinib", NA, 400, NA, 1.39, 2.85, 1.77, 3.82,
  "Rifampicin", "rifampicin", NA, 600, NA, 0.50, 0.30, 0.20, 0.42,
  "Efavirenz", "efavirenz", NA, 600, NA, 0.77, 0.56, 0.37, 0.74,
  "Voriconazole+Nilotinib", "voriconazole", "nilotinib", 300, 400, 1.52, 3.44, 2.00, 4.97,
  "Voriconazole+Rifampicin", "voriconazole", "rifampicin", 300, 600, 1.05, 1.10, 0.73, 1.68,
  "Voriconazole+Efavirenz", "voriconazole", "efavirenz", 300, 600, 1.05, 1.10, 0.62, 1.78,
  "Ritonavir+Nilotinib", "ritonavir", "nilotinib", 100, 400, 1.52, 3.62, 2.11, 5.04,
  "Ritonavir+Rifampicin", "ritonavir", "rifampicin", 100, 600, 1.05, 1.22, 0.71, 1.92,
  "Ritonavir+Efavirenz", "ritonavir", "efavirenz", 100, 600, 1.05, 1.14, 0.67, 1.75,
  "Nilotinib+Rifampicin", "nilotinib", "rifampicin", 400, 600, 1.06, 1.31, 0.55, 2.27,
  "Nilotinib+Efavirenz", "nilotinib", "efavirenz", 400, 600, 1.29, 2.13, 0.96, 3.23
) |>
  mutate(
    ii_p = ifelse(perpetrator == "clarithromycin", 12, 24),
    ii_p2 = 24,
    group = ifelse(is.na(perpetrator2) | perpetrator %in% c("darunavir", "atazanavir"),
      "One perpetrator (Figure 2)", "Two perpetrators (Figure 3)"
    )
  )

s7_runs <- s7 |>
  mutate(
    victim = "bictegravir", amt = 50, ii = 24, start = 144, end = 312,
    start_p = 0, end_p = 336 - ii_p, start_p2 = 0, end_p2 = 312,
    etagastric = 0
  ) |>
  tidyr::crossing(ddi_on = c(1, 0), SEXF = c(0, 1))
s7_runs <- bind_cols(select(s7_runs, -SEXF), bind_rows(lapply(s7_runs$SEXF, typical_subject)))
s7_runs$id <- seq_len(nrow(s7_runs))
s7_runs$arm <- ifelse(s7_runs$ddi_on == 1, "DDI", "control")
s7_runs$sex <- ifelse(s7_runs$SEXF == 1, "woman", "man")

obs_times <- sort(unique(c(seq(144, 168, by = 0.25), seq(288, 312, by = 0.25))))
s7_data <- build_data(s7_runs, obs_times)
s7_sim <- rxode2::rxSolve(mod_tv, s7_data, returnType = "data.frame", cores = 2L) |>
  left_join(select(s7_runs, id, scenario, arm, sex), by = "id")
```

The day-1 and day-7 AUCs of bictegravir are computed with PKNCA over the
dosing intervals 144-168 h and 288-312 h.

```{r s7-nca}
s7_conc <- s7_sim |>
  filter(!is.na(Cc)) |>
  select(id, scenario, arm, sex, time, Cc)
s7_dose <- s7_data |>
  filter(evid == 1, cmt == "stomach") |>
  select(id, time, amt) |>
  left_join(select(s7_runs, id, scenario, arm, sex), by = "id")

s7_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(s7_conc, Cc ~ time | scenario + sex + arm + id),
  PKNCA::PKNCAdose(s7_dose, amt ~ time | scenario + sex + arm + id),
  intervals = data.frame(start = c(144, 288), end = c(168, 312), cmax = TRUE, auclast = TRUE)
))

s7_ratio <- as.data.frame(s7_nca) |>
  filter(PPTESTCD == "auclast") |>
  mutate(day = ifelse(start == 144, "day1", "day7")) |>
  select(scenario, sex, arm, day, PPORRES) |>
  pivot_wider(names_from = arm, values_from = PPORRES) |>
  mutate(ratio = DDI / control) |>
  group_by(scenario, day) |>
  summarise(ratio = mean(ratio), .groups = "drop") |>
  pivot_wider(names_from = day, values_from = ratio) |>
  right_join(s7, by = "scenario") |>
  arrange(match(scenario, s7$scenario)) |>
  mutate(
    day1_pct = 100 * (day1 / day1_paper - 1),
    day7_pct = 100 * (day7 / median - 1)
  )

s7_ratio |>
  transmute(
    scenario,
    day1_paper, day1 = round(day1, 2), day1_pct = round(day1_pct),
    paper = sprintf("%.2f (%.2f-%.2f)", median, ci_lo, ci_hi),
    day7 = round(day7, 2), day7_pct = round(day7_pct)
  ) |>
  rename(
    "Scenario" = scenario,
    "Day 1, paper" = day1_paper, "Day 1, model" = day1, "Day 1, % diff" = day1_pct,
    "Day 7, paper median (95% CI)" = paper, "Day 7, model" = day7, "Day 7, % diff" = day7_pct
  ) |>
  knitr::kable(caption = paste(
    "Bictegravir AUC ratios (DDI / control): typical subjects versus the paper.",
    "Day-1 values are from the Results text (one value per group of scenarios);",
    "day-7 values are the Table S7 medians and 95% confidence intervals."
  ))
```

```{r s7-figure, fig.height = 6, fig.cap = "Replicates Figures 2 and 3 of Stader 2021: bictegravir day-7 AUC ratios. Bars: the paper's median and 95% confidence interval (Table S7); points: the model's typical subjects. Dashed lines: the efficacy-safety margin of 0.5- and 2.4-fold."}
s7_ratio |>
  mutate(scenario = factor(scenario, levels = rev(s7$scenario))) |>
  ggplot(aes(y = scenario)) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.3, colour = "grey50") +
  geom_point(aes(x = median, shape = "Paper median"), colour = "grey30", size = 2.5) +
  geom_point(aes(x = day7, shape = "Model, typical subjects"), colour = "firebrick", size = 2.5) +
  geom_vline(xintercept = c(0.5, 2.4), linetype = "dashed", colour = "red") +
  scale_x_log10() +
  scale_shape_manual(values = c("Paper median" = 16, "Model, typical subjects" = 4)) +
  facet_grid(group ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Day-7 AUC ratio (DDI / control)", y = NULL, shape = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r s7-profiles, fig.height = 5, fig.cap = "Bictegravir concentrations on the seventh day of dosing (typical man) with and without selected perpetrators."}
s7_sim |>
  filter(
    sex == "man", time >= 288,
    scenario %in% c("Voriconazole", "Atazanavir", "Nilotinib", "Rifampicin", "Efavirenz", "Ritonavir")
  ) |>
  ggplot(aes(time - 288, Cc, colour = arm)) +
  geom_line() +
  facet_wrap(~scenario) +
  labs(x = "Time after the seventh bictegravir dose (h)", y = "Bictegravir (ng/mL)", colour = NULL) +
  theme_bw()
```

Of the 20 scenarios, `r sum(abs(s7_ratio$day7_pct) < 15)` reproduce the
paper's day-7 median to within 15% (the strong CYP3A inhibitors,
darunavir/cobicistat, atazanavir with and without cobicistat, rifampicin,
efavirenz, voriconazole with rifampicin and ritonavir with nilotinib), and
all but `r sum(abs(s7_ratio$day1_pct) >= 15, na.rm = TRUE)` of the day-1
ratios agree to within 15%. The other nine deviate in two consistent
directions. Every other nilotinib scenario (nilotinib alone and with
voriconazole, rifampicin or efavirenz) gives a larger ratio than the paper:
the model's nilotinib inhibits bictegravir clearance more strongly.
Ritonavir alone, darunavir with ritonavir, ritonavir with rifampicin or
efavirenz, and voriconazole with efavirenz give a smaller ratio: more net
induction. In every one of them the gap is larger on day 7 than on day 1.
The deposited code executed in Octave behaves the same way (previous
section), so the gap lies between the deposited code and the runs behind the
published figures, not in this implementation; it is not resolvable from the
paper or its supplements (see Assumptions and deviations).

```{r s7-checks}
reproduced <- c(
  "Voriconazole", "Ketoconazole", "Clarithromycin", "Cobicistat", "Darunavir+Cobicistat",
  "Atazanavir", "Atazanavir+Cobicistat", "Rifampicin", "Efavirenz",
  "Voriconazole+Rifampicin", "Ritonavir+Nilotinib"
)
larger <- c("Nilotinib", "Voriconazole+Nilotinib", "Nilotinib+Rifampicin", "Nilotinib+Efavirenz")
smaller <- c("Ritonavir", "Darunavir+Ritonavir", "Ritonavir+Rifampicin", "Ritonavir+Efavirenz", "Voriconazole+Efavirenz")
day1_checked <- c(reproduced[!grepl("\\+", reproduced)], "Ritonavir", "Voriconazole+Nilotinib", "Ritonavir+Nilotinib")
stopifnot(
  setequal(s7_ratio$scenario[abs(s7_ratio$day7_pct) < 15], reproduced),
  all(abs(s7_ratio$day1_pct[s7_ratio$scenario %in% day1_checked]) < 15),
  sum(abs(s7_ratio$day1_pct) >= 15, na.rm = TRUE) <= 3,
  # The deviating scenarios deviate in the directions described above, more
  # on day 7 than on day 1.
  all(s7_ratio$day7_pct[s7_ratio$scenario %in% larger] > 15),
  all(s7_ratio$day7_pct[s7_ratio$scenario %in% smaller] < -15),
  with(
    filter(s7_ratio, scenario %in% c(larger, smaller), !is.na(day1_pct)),
    all(abs(day7_pct) > abs(day1_pct))
  )
)
```

# Voriconazole on its own (Table S4)

Voriconazole is one of the two drug models new in this paper, and Table S4
compares its predictions with published healthy-volunteer studies. Here the
voriconazole drug file is loaded into the victim slot, without perpetrators,
and single doses are simulated for the typical man; PKNCA gives Cmax,
AUC0-inf and the terminal half-life, compared with the paper's *predicted*
means (Table S4 reports AUC to tau; for a single dose that is compared with
AUC0-inf).

```{r voriconazole-sim}
vori_regimens <- tibble::tribble(
  ~regimen, ~amt, ~dur, ~cmax, ~aucinf.obs, ~half.life,
  "single iv; 50 mg", 50, 2, 309, 1236, 3.9,
  "single iv; 100 mg", 100, 2, 546, 3484, 4.6,
  "single iv; 200 mg", 200, 1, 2176, 7174, 4.6,
  "single iv; 400 mg", 400, 2, 3738, 20291, 5.4,
  "single oral; 400 mg", 400, NA, 1597, 12806, 5.3
)
vori_obs <- sort(unique(c(seq(0, 4, by = 0.1), seq(4.25, 72, by = 0.25))))
vori_data <- bind_rows(lapply(seq_len(nrow(vori_regimens)), function(i) {
  r <- vori_regimens[i, ]
  dose <- if (is.na(r$dur)) {
    data.frame(time = 0, amt = r$amt, evid = 1L, cmt = "stomach", rate = 0)
  } else {
    data.frame(time = 0, amt = r$amt, evid = 1L, cmt = "venous", rate = r$amt / r$dur)
  }
  cbind(
    id = i,
    bind_rows(dose, data.frame(time = vori_obs, amt = 0, evid = 0L, cmt = "venous", rate = 0)),
    typical_subject(0), etagastric = 0,
    scenario_params("voriconazole")
  )
}))
vori_sim <- rxode2::rxSolve(mod_tv, vori_data, returnType = "data.frame", cores = 2L) |>
  mutate(regimen = vori_regimens$regimen[id])

vori_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(filter(vori_sim, !is.na(Cc)), Cc ~ time | regimen + id),
  PKNCA::PKNCAdose(
    mutate(filter(vori_data, evid == 1), regimen = vori_regimens$regimen[id]),
    amt ~ time | regimen + id
  ),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, aucinf.obs = TRUE, half.life = TRUE)
))

vori_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(vori_nca) |> filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")),
  reference = select(vori_regimens, regimen, cmax, aucinf.obs, half.life),
  by = "regimen",
  units = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL", half.life = "h"),
  tolerance_pct = 20
)
vori_cmp |>
  rename("Regimen" = regimen) |>
  knitr::kable(caption = paste(
    "Voriconazole single doses, typical man, versus the Table S4 predicted means.",
    "* differs by more than 20%."
  ))
```

```{r voriconazole-figure, fig.cap = "Voriconazole concentrations after single doses (typical man)."}
ggplot(filter(vori_sim, time <= 24), aes(time, Cc, colour = regimen)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Voriconazole (ng/mL)", colour = NULL) +
  theme_bw()
```

```{r voriconazole-checks}
vori_auc_ratio <- as.data.frame(vori_nca) |>
  filter(PPTESTCD == "aucinf.obs") |>
  left_join(vori_regimens, by = "regimen") |>
  mutate(ratio = PPORRES / aucinf.obs)
stopifnot(all(vori_auc_ratio$ratio > 1.2))
```

The deposited voriconazole model gives an AUC `r sprintf("%.1f", min(vori_auc_ratio$ratio))`
to `r sprintf("%.1f", max(vori_auc_ratio$ratio))` times the paper's predicted
means, although its inputs are exactly those printed in Table S1 (plus the
tissue scalar below). The same happens when the deposited code itself is run
(previous section, where the voriconazole concentrations are those of the
Octave runs). Because voriconazole inactivates CYP3A4 almost completely at
either exposure, its effect on bictegravir (Table S7) is reproduced
regardless.

# Assumptions and deviations

* **Source.** The model implements the Matlab code deposited as
  Supplementary Material s002. Its header reads "Publication Date:
  12/12/2020, Publication: CPT - bictegravir", and the same code is deposited
  with Stader et al. 2021 (doi:10.1002/cpt.2178). The implementation
  reproduces that code exactly (section "Check against the deposited code").
* **Deposited code versus published results.** The deposited code does not
  reproduce every published prediction: the day-7 AUC ratios of the
  nilotinib and ritonavir-containing scenarios differ from Table S7 (section
  "Prospective DDI scenarios"), and the voriconazole exposure is higher than
  the Table S4 predictions (its clearance is lower; the half-life is about
  twice as long). These differences are those of the deposited code itself,
  possibly because the published runs used a later revision of the code or of
  the drug files; neither the paper nor the supplements allow them to be
  reconciled, and the model keeps the deposited values rather than adjusting
  any of them. Two single-input explanations were ruled out: neutralising the
  first as-run behaviour (setting `indmax_cyp3a4` to 1 for the drugs that
  leave it at 0) does not close the nilotinib gap, and switching off
  voriconazole's own CYP3A4 inactivation does not close the voriconazole
  gap.
* **As-run behaviours kept.** The four behaviours listed at the top (hepatic
  CYP synthesis suppression by non-inducers, drug 1's unbound fraction in
  every albumin-binding term, the MPPGL CV of 2.3% when ritonavir is the last
  drug, and the 1-h dosing lag of ritonavir and efavirenz) are part of the
  deposited code and change its output, so they are kept.
* **Voriconazole tissue scalar.** The voriconazole drug file sets a global
  tissue-partition scalar of 2.75 (`kpscalar`, citing Li et al.), which
  Table S1 does not list; the model uses it as deposited.
* **Clarithromycin library index.** Clarithromycin has a drug file but is not
  registered in the framework's drug list (`PBPK_DefineParameters.m`); it is
  placed after the registered drugs, which only matters for deciding which
  drug of a run is drug 1 (bictegravir precedes it either way).
* **Doses as printed.** Table S3 gives voriconazole 300 mg and nilotinib
  400 mg once daily (both are usually given twice daily); the scenarios use
  the printed regimens.
* **Typical subjects.** The paper simulated 100 virtual individuals per
  scenario; to keep the article's run time short, each scenario is solved for
  a typical man and a typical woman (the generator's mean height and weight at
  35 years, every random draw at its median, gastric emptying 0.625 h).
  Cohorts can be drawn from the model's random effects; see the MPPGL note
  above for ritonavir-last runs.
* **Clinical DDI studies (Table 1, Figure 1).** The bictegravir doses and
  sampling designs of the five clinical DDI studies (data from Gilead
  Sciences) are not reported, so Table 1 is not replicated.
* **Reported concentration.** As in the framework, the reported "plasma"
  concentration is the venous-blood state times the molecular weight,
  without a blood-to-plasma correction.
