---
title: "Levonorgestrel and ethinyl estradiol transdermal system (Stanczyk 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levonorgestrel and ethinyl estradiol transdermal system (Stanczyk 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Stanczyk_2022_levonorgestrel"))$reference`
* Levonorgestrel model: `r rxode2::rxode(readModelDb("Stanczyk_2022_levonorgestrel"))$description`
* Ethinyl estradiol model: `r rxode2::rxode(readModelDb("Stanczyk_2022_ethinylestradiol"))$description`
* Article: <https://doi.org/10.1371/journal.pone.0279640>
* Supporting information (S1-S4 Tables, S1-S2 Figs): <https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0279640#sec014>

Stanczyk 2022 fitted two separate one-compartment population PK models in
Phoenix NLME 8.1, one for levonorgestrel (LNG) and one for ethinyl estradiol
(EE), to cycle-2 data from women wearing the once-weekly LNG/EE transdermal
delivery system (TDS). The models were then used to simulate a 12-week extended
regimen and to show that weeks 3 and 12 give the same exposure. The two models
are packaged as `Stanczyk_2022_levonorgestrel` and
`Stanczyk_2022_ethinylestradiol`. The final parameter estimates are printed
only in the supporting information (S1 and S2 Tables); the simulated
concentrations at each nominal time point are in S3 and S4 Tables.

# Population

The source study was ATI-CL14 (NCT01243580), a phase 1, open-label, randomized,
single-centre trial run in 2009 in 36 healthy, normotensive, nonsmoking women
aged 18 to 45 years with a BMI of 18 to 32 kg/m^2 and 24- to 35-day menstrual
intervals. The popPK models used only the cycle-2 data of the 18 women who wore
the TDS in cycles 1 and 2. Samples were taken before patch application and at 6,
12, 24, 48, 72, 120, 144 and 168 h in weeks 1 and 3, plus 174 to 240 h after the
third patch (23 samples per woman). Each patch contains 2.6 mg LNG and 2.3 mg EE
and nominally delivers 120 ug/day LNG and 30 ug/day EE. No covariate analysis was
performed and no demographic table for the 18 women is given (Methods, "Study
design" and "PopPK models").

# Source trace

| Quantity | Value | Source |
|---|---|---|
| EE structure: zero-order input into central over 168 h (fraction Frel) plus first-order depot (1 - Frel) | -- | Methods, "EE popPK models" |
| EE V | 2490 L | S1 Table |
| EE CL | 107 L/h | S1 Table |
| EE Ka | 0.0204 1/h | S1 Table |
| EE Frel (zero-order route) | 0.732 | S1 Table |
| EE F | 0.30 (fixed) | S1 Table; Methods |
| EE IIV V / CL / Ka | 43.4 / 34.6 / 53.1 % CV | S1 Table |
| EE residual error | 26.9 % proportional | S1 Table and footnote |
| LNG structure: depot infused over D1 (fraction Frel) with first-order absorption, plus zero-order input into central over 168 h (1 - Frel) | -- | Methods, "LNG popPK model" |
| LNG V, CL (patches 1-2) | 163 L, 2.46 L/h | S2 Table |
| LNG week-3 effect on V, CL | 85.0 %, 87.0 % (multipliers 0.85, 0.87) | S2 Table; reading checked below |
| LNG Ka | 0.0281 1/h | S2 Table |
| LNG Frel (first-order route) | 0.378 | S2 Table |
| LNG F | 0.27 (fixed) | S2 Table; Methods |
| LNG D1 (depot infusion duration) | 0.5 h, **back-solved** | not printed; S4 Table week 1 (see below) |
| LNG IIV V / CL / Frel | 38.9 / 44.4 / 23.5 % CV | S2 Table |
| LNG residual error | proportional SD 0.206, additive SD 0.0860 ug/L | S2 Table and footnote |
| Dose per patch | 2600 ug LNG, 2300 ug EE | Methods, "Extended regimen simulations" |

# Virtual cohort and extended-regimen simulation

The paper simulated 1000 women wearing 12 consecutive weekly patches and summarised
weeks 1, 3 and 12. Here 200 women are simulated per drug with the full
between-subject variability and without residual error, as in the paper. Each
patch is two dose records of the patch content, one to `depot` and one to
`central`.

The central record is given an explicit `dur = 168` rather than `rate = -2`.
The model fixes the central input duration at 168 h, so the two are the same
model, but rxode2 5.1.8 mis-handles a modelled-duration (`rate = -2`) infusion
that ends at the instant the next one starts when a covariate changes on the
new dose row, as `CYCLE` does here: the weekly central input is then delivered
almost at once. An explicit duration is unaffected (checked below against the
closed-form steady-state AUC). The LNG depot record keeps `rate = -2` so the
model's 0.5 h `dur(depot)` applies; the EE depot record is a bolus.

`CYCLE` is the number of the patch being worn. The observation that closes week
*w* (and is time 0 of week *w + 1*) is given the CYCLE of week *w* and placed
before the next patch, so it is the pre-dose value the paper tabulates as
"time 0".

```{r events}
n_sub <- 200
n_patch <- 12
dose_lng <- 2600 # ug LNG per patch
dose_ee <- 2300 # ug EE per patch
weeks_obs <- c(1, 3, 12)

dose_rows <- tidyr::expand_grid(patch = seq_len(n_patch), cmt = c("depot", "central")) |>
  mutate(time = 168 * (patch - 1), evid = 1L, CYCLE = patch)

obs_rows <- tidyr::expand_grid(week = weeks_obs, rt = 0:168) |>
  mutate(
    time = 168 * (week - 1) + rt,
    evid = 0L,
    cmt = "central",
    rate = 0,
    dur = 0,
    # Time 0 of a week is the pre-dose value, still under the previous patch.
    CYCLE = ifelse(rt == 0, pmax(week - 1, 1), week)
  )

make_events <- function(amt, depot_rate) {
  doses <- dose_rows |>
    mutate(
      amt = amt,
      rate = ifelse(cmt == "depot", depot_rate, 0),
      dur = ifelse(cmt == "central", 168, 0)
    )
  one <- bind_rows(doses, mutate(obs_rows, amt = 0)) |>
    # observation before dose at a shared time
    arrange(time, evid) |>
    select(time, evid, amt, cmt, rate, dur, CYCLE)
  tidyr::expand_grid(id = seq_len(n_sub), one) |>
    select(id, time, evid, amt, cmt, rate, dur, CYCLE)
}
ev_lng <- make_events(dose_lng, depot_rate = -2)
ev_ee <- make_events(dose_ee, depot_rate = 0)
```

```{r simulate}
mod_lng <- readModelDb("Stanczyk_2022_levonorgestrel")
mod_ee <- readModelDb("Stanczyk_2022_ethinylestradiol")

rxode2::rxSetSeed(20221227)
sim_lng <- rxode2::rxSolve(mod_lng, ev_lng, returnType = "data.frame")
rxode2::rxSetSeed(20221228)
sim_ee <- rxode2::rxSolve(mod_ee, ev_ee, returnType = "data.frame")

week_of <- function(time) {
  # Week 1 is [0, 168]; week 3 is [336, 504]; week 12 is [1848, 2016].
  w <- rep(NA_real_, length(time))
  for (wk in weeks_obs) {
    w[time >= 168 * (wk - 1) & time <= 168 * wk & is.na(w)] <- wk
  }
  w
}

conc <- bind_rows(
  sim_lng |> mutate(drug = "LNG"),
  sim_ee |> mutate(drug = "EE")
) |>
  select(id, time, drug, Cc) |>
  mutate(week = week_of(time), rt = time - 168 * (week - 1)) |>
  filter(!is.na(week)) |>
  # each solve returns one row per observation record
  distinct(id, drug, week, rt, .keep_all = TRUE) |>
  mutate(conc_pg = 1000 * Cc) # ng/mL -> pg/mL
stopifnot(!anyNA(conc$Cc))
```

`Cc` is the individual prediction, i.e. without residual error, matching the
paper.

# Back-solving the LNG depot infusion duration

The Methods state that the duration of the zero-order infusion into the LNG
dosing compartment "was estimated by the model", but S2 Table does not print it.
Every other LNG parameter is printed, so the duration was back-solved from the
paper's own week-1 simulation (S4 Table). Only the earliest time points depend
on it: by 24 h the depot has received its whole dose for any duration of a few
hours. The table below shows the typical-value concentrations for a range of
durations; the shortest durations fit the S4 Table 6-h and 12-h values best.
The maintainers fitted the duration by least squares to the S4 Table week-1
means and medians at 6 and 12 h with a 4000-subject simulation, scanning 0.1 to
2 h; the optimum was 0.5 h, which is what the model ships. Any duration up to
about 2 h stays within 5% of those four values, so the exact value hardly
matters, but durations of 8 h or more do not fit.

```{r d1-scan}
ev_wk1 <- rxode2::et(amt = dose_lng, cmt = "depot", rate = -2) |>
  rxode2::et(amt = dose_lng, cmt = "central", rate = -2) |>
  rxode2::et(c(6, 12, 24, 48)) |>
  as.data.frame() |>
  mutate(CYCLE = 1)
d1_scan <- lapply(c(0.05, 0.5, 2, 8, 24), function(d1) {
  s <- rxode2::rxSolve(rxode2::zeroRe(mod_lng), ev_wk1,
    params = c(ld1 = log(d1)), returnType = "data.frame")
  data.frame(d1 = d1, time = s$time, conc = round(1000 * s$Cc))
}) |>
  bind_rows() |>
  tidyr::pivot_wider(names_from = time, values_from = conc, names_prefix = "t")
d1_scan |>
  dplyr::rename(
    "Depot infusion duration (h)" = d1,
    "6 h" = t6, "12 h" = t12, "24 h" = t24, "48 h" = t48
  ) |>
  knitr::kable(caption = "Typical-value LNG concentration (pg/mL) in week 1 by depot infusion duration. S4 Table week-1 medians: 330, 591, 949 and 1260 pg/mL.")
```

```{r d1-check}
wk1_early <- conc |>
  filter(drug == "LNG", week == 1, rt %in% c(6, 12)) |>
  group_by(rt) |>
  summarise(mean = mean(conc_pg), median = median(conc_pg), .groups = "drop")
paper_wk1 <- data.frame(rt = c(6, 12), mean = c(357, 632), median = c(330, 591))
wk1_early
stopifnot(
  # The 0.5 h duration reproduces the S4 Table early week-1 values; an 8 h
  # duration would put the 6 h median about 40% low.
  all(abs(wk1_early$median / paper_wk1$median - 1) < 0.15),
  all(abs(wk1_early$mean / paper_wk1$mean - 1) < 0.15)
)
```

# Reading of the LNG week-3 effect

S2 Table prints "Week 3 effect on volume (%) 85.0" and "Week 3 effect on clearance
(%) 87.0". These are read as multipliers: from week 3, V = 0.85 x 163 L and
CL = 0.87 x 2.46 L/h. Two checks support this reading:

* At steady state the weekly AUC equals F x dose / CL. With CL = 0.87 x 2.46 L/h
  this is 0.27 x 2600 / 2.14 = 328 h*ng/mL, against the paper's week-12
  geometric mean AUC(0-168) of 326 h*ng/mL (Table 3). The other readings,
  1.87 x CL or 0.13 x CL, give 153 or 2190 h*ng/mL.
* The effect has to start at the third patch, not the second. The S4 Table week-3
  pre-dose mean (1410 pg/mL) is the end of week 2 under the reference volume; if
  the lower volume applied already in week 2, the pre-dose mean would be about
  1640 pg/mL. The paper's extended-regimen simulation applies the week-3 values
  to every later patch, as the week-12 rows show.

```{r week3-checks}
typ_lng <- rxode2::zeroRe(mod_lng)
ev_typ <- ev_lng |> filter(id == 1)
s_typ <- rxode2::rxSolve(typ_lng, ev_typ, returnType = "data.frame") |>
  mutate(week = week_of(time), rt = time - 168 * (week - 1)) |>
  filter(week == 12) |>
  distinct(rt, .keep_all = TRUE)
auc_wk12_typ <- sum(diff(s_typ$rt) * (head(s_typ$Cc, -1) + tail(s_typ$Cc, -1)) / 2)
auc_wk12_ss <- 0.27 * dose_lng / (0.87 * 2.46)
c(simulated = auc_wk12_typ, closed_form = auc_wk12_ss)
stopifnot(
  # Steady-state identity on the typical subject: the same parameters on both
  # sides, so only trapezoid and carry-over error separate them.
  abs(auc_wk12_typ / auc_wk12_ss - 1) < 0.02
)

predose_wk3 <- conc |> filter(drug == "LNG", week == 3, rt == 0)
c(mean = mean(predose_wk3$conc_pg), median = median(predose_wk3$conc_pg))
stopifnot(
  # S4 Table week-3 time-0 median 1270 pg/mL; the week-2 onset reading gives
  # about 1490.
  abs(median(predose_wk3$conc_pg) / 1270 - 1) < 0.10
)
```

# Replicating Figure 1 and S3/S4 Tables

```{r figure1, fig.width = 7, fig.height = 4.5}
mean_prof <- conc |>
  group_by(drug, week, rt) |>
  summarise(mean_conc = mean(conc_pg), .groups = "drop") |>
  mutate(
    week = factor(paste("Week", week), levels = paste("Week", weeks_obs)),
    drug = factor(drug, levels = c("EE", "LNG"))
  )
ggplot(mean_prof, aes(rt, mean_conc, colour = week)) +
  geom_line() +
  facet_wrap(~drug, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 168, 24)) +
  labs(
    x = "Time since patch application (h)",
    y = "Mean concentration (pg/mL)",
    colour = NULL,
    caption = "Replicates Figure 1 of Stanczyk 2022 (A: EE, B: LNG)."
  ) +
  theme_bw()
```

```{r tables-s3-s4}
nominal <- c(0, 6, 12, 24, 48, 72, 120, 144, 168)
paper_s3s4 <- bind_rows(
  data.frame(
    drug = "EE", week = rep(weeks_obs, each = 9), rt = nominal,
    paper_mean = c(
      0, 15.4, 25.2, 35.2, 39.4, 38.1, 34.3, 33.0, 32.1,
      32.8, 41.4, 46.1, 49.4, 46.8, 42.4, 36.0, 34.2, 32.9,
      33.0, 41.5, 46.2, 49.5, 46.9, 42.4, 36.1, 34.2, 33.0
    ),
    paper_median = c(
      0, 14.1, 23.5, 33.5, 37.7, 36.0, 31.9, 30.7, 29.8,
      30.3, 38.6, 43.5, 46.9, 43.7, 39.0, 33.3, 31.6, 30.3,
      30.4, 38.6, 43.5, 47.0, 43.7, 39.1, 33.4, 31.6, 30.4
    )
  ),
  data.frame(
    drug = "LNG", week = rep(weeks_obs, each = 9), rt = nominal,
    paper_mean = c(
      0, 357, 632, 1000, 1320, 1380, 1310, 1260, 1220,
      1410, 1940, 2140, 2370, 2430, 2280, 1930, 1780, 1680,
      1700, 1980, 2190, 2410, 2470, 2320, 1950, 1810, 1700
    ),
    paper_median = c(
      0, 330, 591, 949, 1260, 1330, 1250, 1190, 1140,
      1270, 1770, 1970, 2200, 2250, 2100, 1730, 1590, 1480,
      1480, 1760, 1980, 2200, 2250, 2110, 1730, 1590, 1480
    )
  )
)

cmp_s3s4 <- conc |>
  filter(rt %in% nominal) |>
  group_by(drug, week, rt) |>
  summarise(sim_mean = mean(conc_pg), sim_median = median(conc_pg), .groups = "drop") |>
  inner_join(paper_s3s4, by = c("drug", "week", "rt")) |>
  mutate(pct_diff_mean = ifelse(paper_mean > 0, 100 * (sim_mean / paper_mean - 1), NA_real_))

cmp_s3s4 |>
  mutate(across(c(sim_mean, sim_median), ~ signif(.x, 3)), pct_diff_mean = round(pct_diff_mean, 1)) |>
  dplyr::rename(
    "Drug" = drug, "Week" = week, "Time (h)" = rt,
    "Simulated mean (pg/mL)" = sim_mean, "Paper mean (pg/mL)" = paper_mean,
    "Simulated median (pg/mL)" = sim_median, "Paper median (pg/mL)" = paper_median,
    "Mean % diff" = pct_diff_mean
  ) |>
  knitr::kable(caption = "Simulated concentrations against S3 Table (EE) and S4 Table (LNG) of Stanczyk 2022.")

stopifnot(
  # Centre and robust envelope across all 52 nonzero time points
  # (200 simulated women against the paper's 1000).
  abs(median(cmp_s3s4$pct_diff_mean, na.rm = TRUE)) < 5,
  quantile(abs(cmp_s3s4$pct_diff_mean), 0.9, na.rm = TRUE) < 12
)
```

# NCA with PKNCA

AUC(0-168), Cmax and Css(0-168) = AUC(0-168)/168 (PKNCA `cav`) are computed per
woman and week with PKNCA, using the paper's linear-up/log-down rule. Css(48-168)
is the average over 48 to 168 h.

```{r pknca}
nca_conc <- conc |>
  mutate(treatment = paste0(drug, " week ", week)) |>
  select(id, treatment, drug, week, rt, conc_ng = Cc)

conc_obj <- PKNCA::PKNCAconc(nca_conc, conc_ng ~ rt | treatment + id)
intervals <- data.frame(
  start = c(0, 48),
  end = c(168, 168),
  cmax = c(TRUE, FALSE),
  auclast = c(TRUE, FALSE),
  cav = c(TRUE, TRUE)
)
nca_data <- PKNCA::PKNCAdata(conc_obj, intervals = intervals,
  options = list(auc.method = "lin up/log down"))
nca_res <- PKNCA::pk.nca(nca_data)

nca_wide <- as.data.frame(nca_res) |>
  # PKNCA also reports the dependencies it computed (e.g. AUC for 48-168 h)
  filter(
    (start == 0 & PPTESTCD %in% c("cmax", "auclast", "cav")) |
      (start == 48 & PPTESTCD == "cav")
  ) |>
  mutate(PPTESTCD = ifelse(PPTESTCD == "cav" & start == 48, "cav48", PPTESTCD)) |>
  select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

geo <- function(x) exp(mean(log(x)))
nca_geo <- nca_wide |>
  group_by(treatment) |>
  summarise(
    cmax = geo(cmax) * 1000, auclast = geo(auclast),
    cav = geo(cav) * 1000, cav48 = geo(cav48) * 1000, .groups = "drop"
  )
```

Accumulation ratios are per woman, week 3 or 12 over week 1, as in Table 1.

```{r accumulation}
acc <- nca_wide |>
  tidyr::separate(treatment, c("drug", "week"), sep = " week ") |>
  select(drug, week, id, cmax, auclast) |>
  tidyr::pivot_wider(names_from = week, values_from = c(cmax, auclast)) |>
  mutate(
    ar_cmax_3 = cmax_3 / cmax_1, ar_cmax_12 = cmax_12 / cmax_1,
    ar_auc_3 = auclast_3 / auclast_1, ar_auc_12 = auclast_12 / auclast_1
  ) |>
  group_by(drug) |>
  summarise(across(starts_with("ar_"), geo), .groups = "drop")
knitr::kable(acc, digits = 2, caption = "Simulated geometric-mean accumulation ratios. Stanczyk 2022 Tables 2 and 3: EE AR_Cmax 1.17 / 1.17 and AR_AUC 1.17 / 1.18 (weeks 3 / 12); LNG AR_Cmax 1.63 / 1.64 and AR_AUC 1.68 / 1.70.")
paper_ar <- data.frame(
  drug = c("EE", "LNG"),
  ar_cmax_3 = c(1.17, 1.63), ar_cmax_12 = c(1.17, 1.64),
  ar_auc_3 = c(1.17, 1.68), ar_auc_12 = c(1.18, 1.70)
)
ar_ratio <- as.matrix(acc[, -1]) / as.matrix(paper_ar[match(acc$drug, paper_ar$drug), -1])
stopifnot(
  # Ratios of exposures within the same simulated women, so the cohort draw
  # largely cancels; a wrong week-3 effect or onset moves the LNG ratios by
  # 15% or more.
  all(abs(ar_ratio - 1) < 0.05)
)
```

```{r nca-compare}
published <- data.frame(
  treatment = c("EE week 1", "EE week 3", "EE week 12", "LNG week 1", "LNG week 3", "LNG week 12"),
  cmax = c(40.8, 47.9, 48.0, 1420, 2310, 2330),
  auclast = c(5.47, 6.42, 6.43, 192, 323, 326),
  cav = c(32.6, 38.2, 38.3, 1140, 1920, 1940)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_geo |> select(treatment, cmax, auclast, cav),
  reference = published,
  by = "treatment",
  params = c("cmax", "auclast", "cav"),
  units = c(cmax = "pg/mL", auclast = "h*ng/mL", cav = "pg/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  digits = 2,
  caption = paste(
    "Simulated against published geometric means (Stanczyk 2022 Tables 2 and 3).",
    attr(cmp, "footnote")
  )
)

pct <- cmp[["% diff"]]
pct <- as.numeric(gsub("[^0-9.-]", "", as.character(pct))) # "+4.3%" -> 4.3
stopifnot(!anyNA(pct))
stopifnot(
  abs(median(pct)) < 5,
  max(abs(pct)) < 15
)
```

The published Css(48-168) geometric means (EE 33.5 / 35.7 / 35.7 pg/mL; LNG
1250 / 1840 / 1860 pg/mL for weeks 1 / 3 / 12) compare with the simulated
values below.

```{r css48}
nca_geo |>
  select(treatment, cav48) |>
  dplyr::rename("Drug and week" = treatment, "Simulated Css(48-168) (pg/mL)" = cav48) |>
  knitr::kable(digits = 1)
```

# Assumptions and deviations

* **LNG depot infusion duration not printed.** The Methods say it was estimated
  but S2 Table omits it. The packaged 0.5 h was back-solved from the S4 Table
  week-1 6-h and 12-h values (see above). The other S4 Table entries, S3 Table,
  and Tables 2 and 3 were not used in the fit and serve as the validation.
* **Week-3 effect.** Read as multipliers 0.85 on V and 0.87 on CL, applied from
  the third consecutive patch onwards (`CYCLE >= 3`); see the two checks above.
  The source data contained only weeks 1 and 3 of a three-patch cycle, so
  applying the week-3 values to weeks 4 to 12 is the paper's extrapolation,
  which this model reproduces. Changing V at the start of week 3 makes the
  concentration step up by 1/0.85, as it does in the paper's simulation.
* **Dose amount.** The model dose is the patch content (2.6 mg LNG, 2.3 mg EE)
  and F (27% LNG, 30% EE) applies to it. The nominal delivery rates (120 and
  30 ug/day) are not the model dose: with them the EE AUC would be about a third
  of Table 2. The content-based dose reproduces the steady-state AUC as
  F x dose / CL (EE 6.45 h*ng/mL against 6.42 to 6.43 in Table 2).
* **IIV scale.** The supplement reports IIV as CV%. It is converted with
  omega^2 = log(1 + CV^2), the Phoenix exponential random-effect form. No
  correlations are reported, so the IIV is diagonal. IIV on the LNG Frel is
  exponential, as for the other parameters; with an SD of 0.23 the chance that
  Frel exceeds 1 is about 1e-5.
* **Residual error.** EE: proportional, 26.9%. LNG: Phoenix mixed error with
  variance addSd^2 + (propSd x C)^2, which is the nlmixr2 `add() + prop()`
  combined form. The multiplicative SD is printed as -0.206; only its square
  enters the model.
* **Css(0-168).** Taken as PKNCA `cav` over 0 to 168 h, which matches the
  paper's definition AUC(0-168)/168 (Table 1).
* **Concentrations** are in ng/mL (ug/L) in the model and converted to pg/mL
  for comparison with the paper.
