---
title: "Amikacin (Steffens 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amikacin (Steffens 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Steffens_2025_amikacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Steffens_2025_amikacin"))$description`
* Article: <https://doi.org/10.3390/antibiotics14060531>

Steffens and colleagues built a one-compartment intravenous population PK model
for amikacin from non-routine therapeutic drug monitoring (TDM) at a Brazilian
teaching hospital, then used it to ask whether standard once-daily regimens
reach the aminoglycoside efficacy target Cmax/MIC >= 8.

# Population

Thirty-nine patients (43 enrolled, 4 excluded for inappropriate collection time
or missing data) admitted to the Hospital Universitario de Santa Maria between
May 2018 and February 2020 contributed 113 amikacin concentrations, 2-6 samples
per subject (53 peak, 60 trough). The cohort was 84.6% male, median age 51 years
(range 4-75, including two pediatric patients), median weight 69.40 kg (range
15.60-143.80), and median Cockcroft-Gault creatinine clearance 79.01 mL/min
(range 12.97-517.97) -- seven patients below 30 mL/min and ten above
120 mL/min. Doses ranged 225-1500 mg given every 12, 24, 48 or 72 h, with
1000 mg q24h the modal regimen (35.7%). *Klebsiella pneumoniae* (n = 20) and
*Pseudomonas aeruginosa* (n = 7) were the commonest isolates. Baseline
demographics are Steffens 2025 Table 1.

Samples were drawn at steady state after at least three days of therapy: the
trough 30 min before a dose and the peak 30 min after the end of the infusion.
Mean +/- SD concentrations were 41.96 +/- 20.20 ug/mL (peak) and
8.75 +/- 15.38 ug/mL (trough).

The same information is available programmatically via
`readModelDb("Steffens_2025_amikacin")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at CRCL = 0) | `log(1.49)` | Table 2, "Cl (L/h)" final estimate 1.49 (RSE 12.96%) |
| `lvc` | `log(23.18)` | Table 2, "Vd (L)" final estimate 23.18 (RSE 23.73%) |
| `e_crcl_cl` | `0.004` | Table 2, "beta CrCl on Cl" final estimate 0.004 (RSE 57.15%) |
| `etalcl` | `0.4489` = 0.67^2 | Table 2, "IIV (%CV) omega Cl" = 0.67 (74.8%), RSE 17.01% |
| `etalvc` | `0.2209` = 0.47^2 | Table 2, "omega Vd" = 0.47 (49.89%), RSE 37.69% |
| `propSd` | `0.38` | Table 2, "Proportional error model" 0.38 (RSE 14.86%); Abstract confirms a proportional error model |
| Structure: 1 compartment, linear elimination | n/a | Section 2.2 ("the one-compartment model, with linear elimination ... presented a good description of the data") |
| `d/dt(central) <- -kel * central`, `kel <- cl/vc` | n/a | Section 2.2; standard one-compartment IV parameterisation |
| `log(cl) = log(1.49) + 0.004 * CRCL` (uncentered) | n/a | Reconstructed -- see "Establishing the covariate form" below |
| CRCL by Cockcroft-Gault, raw mL/min | n/a | Section 4.3 |
| Simulated regimens (500/750 mg q12h; 1000/1500/2000 mg q24h, 1 h infusion) | n/a | Section 4.4.2 |

## Establishing the covariate form

The paper prints the covariate coefficient (`beta CrCl on Cl` = 0.004) but never
writes the covariate equation, so the centering had to be established from the
paper's own numbers rather than read off the page. Two facts settle it:

1. Monolix -- the software used here -- enters a continuous covariate
   **uncentered** by default and names the coefficient exactly as Table 2 prints
   it (`beta_Cl_CrCl`).
2. Under the uncentered reading, the final model reproduces the paper's own
   covariate-free base model at the population median creatinine clearance:
   `1.49 * exp(0.004 * 79.01) = 2.044 L/h` against the base-model CL of
   **2.04 L/h** reported in Section 2.2. A median-centered reading would instead
   place CL = 1.49 L/h at the median -- a 27% drop from the base model with no
   explanation in the text.

The model therefore encodes `log(CL) = log(1.49) + 0.004 * CRCL`, so `exp(lcl)`
is clearance extrapolated to CRCL = 0 rather than clearance in a typical
patient. This is checked mechanically below.

# Structural verification

These checks are deterministic -- one typical subject, no random effects, no
cohort -- so they are reproducible on any machine and are asserted tightly.

```{r structural}
mod <- readModelDb("Steffens_2025_amikacin")
mod_typ <- rxode2::zeroRe(mod)

# Build a single-subject IV-infusion event table.
iv_events <- function(crcl, dose, tau, tinf = 1, n_dose = 1, obs) {
  dose_times <- (seq_len(n_dose) - 1) * tau
  dplyr::bind_rows(
    data.frame(id = 1L, time = dose_times, amt = dose, evid = 1L,
               cmt = "central", rate = dose / tinf, CRCL = crcl),
    data.frame(id = 1L, time = obs, amt = NA_real_, evid = 0L,
               cmt = "central", rate = 0, CRCL = crcl)
  ) |>
    dplyr::arrange(time, dplyr::desc(evid))
}

# --- Check 1: the packaged solve matches the closed-form 1-compartment
# IV-infusion solution. Both sides use the same parameters, so the only
# difference is numerical and the tolerance is tight.
crcl_med <- 79.01
tinf <- 1
dose1 <- 1000
obs1 <- sort(unique(c(seq(0, 24, by = 0.05), tinf)))

sol1 <- rxode2::rxSolve(mod_typ, iv_events(crcl_med, dose1, 24, tinf, 1, obs1),
                        omega = NA, returnType = "data.frame")

cl_typ <- sol1$cl[1]
vc_typ <- sol1$vc[1]
kel_typ <- cl_typ / vc_typ
r0 <- dose1 / tinf

closed_form <- function(t) {
  plateau <- r0 / (kel_typ * vc_typ)
  ifelse(t <= tinf,
         plateau * (1 - exp(-kel_typ * t)),
         plateau * (1 - exp(-kel_typ * tinf)) * exp(-kel_typ * (t - tinf)))
}

chk1 <- sol1 |>
  dplyr::filter(time > 0) |>
  dplyr::mutate(analytic = closed_form(time),
                rel = abs(Cc - analytic) / analytic)

max_rel <- max(chk1$rel)
# Realised ~1e-13 (analytic linear-compartment solve). 1e-6 leaves ample
# headroom for a solver-backed path while still catching any structural error.
stopifnot(max_rel < 1e-6)

# --- Check 2 (load-bearing): the covariate form. exp(lcl) must be CL at
# CRCL = 0, and CL at the population median CRCL must recover the paper's
# own covariate-free base-model clearance of 2.04 L/h.
cl_at <- function(crcl) {
  rxode2::rxSolve(mod_typ, iv_events(crcl, dose1, 24, tinf, 1, c(1, 2)),
                  omega = NA, returnType = "data.frame")$cl[1]
}
cl_zero <- cl_at(0)
cl_median <- cl_at(crcl_med)

stopifnot(abs(cl_zero - 1.49) < 1e-6)            # Table 2 intercept
stopifnot(abs(cl_median - 2.04) < 0.01)          # Section 2.2 base-model CL

# --- Check 3: the covariate actually reaches the solved profile. A covariate
# silently dropped from the solve would leave the concentration-time profile
# unchanged; assert it moves by an amount the covariate implies.
sol_lo <- rxode2::rxSolve(mod_typ, iv_events(30, dose1, 24, tinf, 1, obs1),
                          omega = NA, returnType = "data.frame")
sol_hi <- rxode2::rxSolve(mod_typ, iv_events(200, dose1, 24, tinf, 1, obs1),
                          omega = NA, returnType = "data.frame")
trough_ratio <- sol_lo$Cc[sol_lo$time == 24] / sol_hi$Cc[sol_hi$time == 24]
stopifnot(trough_ratio > 2)   # realised ~3.6; a dropped covariate gives 1.0
```

```{r structural-table, echo = FALSE}
tibble::tibble(
  Check = c("Solve vs closed-form 1-cmt IV infusion (max relative error)",
            "CL at CRCL = 0 (L/h)",
            "CL at median CRCL 79.01 mL/min (L/h)",
            "24 h trough ratio, CRCL 30 vs 200 mL/min"),
  Expected = c("< 1e-6", "1.49 (Table 2)", "2.04 (Section 2.2 base model)", "> 1 (covariate active)"),
  Achieved = c(format(max_rel, digits = 3),
               sprintf("%.4f", cl_zero),
               sprintf("%.4f", cl_median),
               sprintf("%.2f", trough_ratio))
) |>
  knitr::kable(caption = "Deterministic structural checks.")
```

## Steady-state mass balance

At steady state the amount cleared over one dosing interval equals the dose, so
`CL * AUCtau = Dose`. AUC comes from PKNCA, not an inline trapezoid.

```{r mass-balance}
tau_mb <- 24
# 20 q24h doses is many multiples of the ~8 h half-life, so the final interval
# is steady state to well below the assertion tolerance.
obs_mb <- sort(unique(c(seq(19 * tau_mb, 20 * tau_mb, by = 0.02),
                        19 * tau_mb + tinf)))
sol_mb <- rxode2::rxSolve(mod_typ,
                          iv_events(crcl_med, dose1, tau_mb, tinf, 20, obs_mb),
                          omega = NA, returnType = "data.frame")

mb_conc <- sol_mb |>
  dplyr::filter(time >= 19 * tau_mb) |>
  dplyr::transmute(id = 1L, regimen = "mass balance",
                   time = time - 19 * tau_mb, Cc = Cc)

mb_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(mb_conc, Cc ~ time | regimen + id,
                   concu = "ug/mL", timeu = "h"),
  PKNCA::PKNCAdose(data.frame(id = 1L, regimen = "mass balance",
                              time = 0, amt = dose1),
                   amt ~ time | regimen + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = tau_mb, auclast = TRUE)
))

auc_tau <- as.data.frame(mb_nca$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::pull(PPORRES)

cleared <- cl_typ * auc_tau          # L/h * ug/mL*h == mg
mb_pct <- 100 * (cleared - dose1) / dose1
# Realised ~0.0x%; the residual is trapezoidal discretisation of the infusion
# ramp. 1.5% still goes red on a mis-scaled volume, dose or clearance.
stopifnot(abs(mb_pct) < 1.5)
```

Amount cleared per 24 h interval: **`r sprintf("%.1f", cleared)` mg** against a
1000 mg dose (`r sprintf("%+.3f", mb_pct)`%).

# Virtual cohort

Original data are not public. The cohort below draws creatinine clearance from a
lognormal matched to the published median (79.01 mL/min) and mean
(100.3 mL/min), truncated to the observed range 12.97-517.97 mL/min. Random
effects are drawn in R and passed to the solver as columns so that **all five
regimens use the identical set of 200 virtual patients** -- this makes the
dose-ranking comparisons below exact rather than a race between noisy arms.

```{r cohort}
set.seed(20250522)

n_sub <- 200   # per arm; at or below the 200/arm cap

# Lognormal on CRCL: median 79.01 (Table 1) fixes mu; the published mean of
# 100.3 fixes sigma via mean = median * exp(sigma^2/2).
crcl_mu <- log(79.01)
crcl_sigma <- sqrt(2 * log(100.3 / 79.01))

subjects <- tibble::tibble(
  sid = seq_len(n_sub),
  CRCL = pmin(pmax(exp(rnorm(n_sub, crcl_mu, crcl_sigma)), 12.97), 517.97),
  # omega values are the log-scale SDs from Table 2 (0.67 on CL, 0.47 on Vd).
  etalcl = rnorm(n_sub, 0, 0.67),
  etalvc = rnorm(n_sub, 0, 0.47)
)

regimens <- tibble::tribble(
  ~regimen,          ~dose, ~tau,
  "500 mg q12h",       500,   12,
  "750 mg q12h",       750,   12,
  "1000 mg q24h",     1000,   24,
  "1500 mg q24h",     1500,   24,
  "2000 mg q24h",     2000,   24
)

# Observe the final dosing interval only, after dosing to steady state.
make_arm <- function(regimen, dose, tau, id_offset) {
  n_dose <- ceiling(20 * 24 / tau)
  last <- (n_dose - 1) * tau
  obs_t <- sort(unique(c(
    seq(0, tau, by = 0.25),
    seq(0, 2, by = 0.05),            # resolve the infusion peak
    tinf, tinf + 0.5, tau - 0.5, tau # paper's sampling times and interval ends
  )))
  subj <- subjects |>
    dplyr::mutate(id = id_offset + sid, regimen = regimen, tau = tau)
  dplyr::bind_rows(
    subj |> tidyr::crossing(time = (seq_len(n_dose) - 1) * tau) |>
      dplyr::mutate(amt = dose, evid = 1L, cmt = "central", rate = dose / tinf),
    subj |> tidyr::crossing(time = last + obs_t) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
  ) |>
    dplyr::mutate(last_dose = last) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  lapply(seq_len(nrow(regimens)), function(i) {
    make_arm(regimens$regimen[i], regimens$dose[i], regimens$tau[i],
             id_offset = (i - 1L) * n_sub)
  })
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
# Random effects are supplied per subject as data columns, so `omega = NA`
# suppresses rxode2's own draw. rxode2 warns that a multi-subject simulation
# has no omega; that is expected here and is suppressed rather than silenced
# elsewhere.
sim <- suppressWarnings(
  rxode2::rxSolve(mod, events = events, omega = NA,
                  keep = c("regimen", "tau", "last_dose"),
                  returnType = "data.frame")
)

sim_ss <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad = time - last_dose) |>
  dplyr::filter(tad >= 0)

stopifnot(nrow(sim_ss) > 0, all(sim_ss$Cc >= 0))
```

# PKNCA validation

Steady-state NCA over the final dosing interval, stratified by regimen. The q12h
and q24h arms have different intervals, so they are run separately and combined.

```{r pknca}
run_nca <- function(dat, tau) {
  # Filter with !is.na() only. The observation grid already puts a record at
  # both interval ends: at tad = 0 (the steady-state pre-dose trough, which is
  # the correct anchor value here -- NOT zero, since this is a steady-state
  # interval) and at tad = tau, which PKNCA needs for ctrough.
  conc <- dat |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time = tad, Cc, regimen) |>
    dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
    dplyr::arrange(id, regimen, time)
  # Fail loudly rather than let PKNCA warn once per subject.
  ends <- conc |>
    dplyr::group_by(id, regimen) |>
    dplyr::summarise(has0 = any(time == 0), hasTau = any(time == tau),
                     .groups = "drop")
  stopifnot(all(ends$has0), all(ends$hasTau))

  dose_df <- dat |>
    dplyr::distinct(id, regimen) |>
    dplyr::left_join(regimens |> dplyr::select(regimen, amt = dose), by = "regimen") |>
    dplyr::mutate(time = 0)

  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(conc, Cc ~ time | regimen + id,
                     concu = "ug/mL", timeu = "h"),
    PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id, doseu = "mg"),
    intervals = data.frame(start = 0, end = tau, cmax = TRUE, tmax = TRUE,
                           cmin = TRUE, auclast = TRUE, cav = TRUE, ctrough = TRUE)
  ))
}

nca_res <- dplyr::bind_rows(
  lapply(c(12, 24), function(tt) {
    as.data.frame(run_nca(dplyr::filter(sim_ss, tau == tt), tt)$result)
  })
)

nca_wide <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav", "ctrough")) |>
  dplyr::select(regimen, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(nca_wide) == nrow(regimens) * n_sub, !anyNA(nca_wide$cmax))
```

```{r nca-summary, echo = FALSE}
nca_wide |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    cmax_med = median(cmax), cmax_lo = quantile(cmax, 0.05), cmax_hi = quantile(cmax, 0.95),
    cmin_med = median(cmin), auc_med = median(auclast),
    .groups = "drop"
  ) |>
  dplyr::left_join(regimens |> dplyr::select(regimen, dose, tau), by = "regimen") |>
  dplyr::arrange(tau, dose) |>
  dplyr::transmute(
    regimen,
    `Cmax,ss (ug/mL)` = sprintf("%.1f [%.1f-%.1f]", cmax_med, cmax_lo, cmax_hi),
    `Cmin,ss (ug/mL)` = sprintf("%.2f", cmin_med),
    `AUCtau (ug*h/mL)` = sprintf("%.0f", auc_med)
  ) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(caption = "Simulated steady-state NCA by regimen (median [5th-95th percentile], n = 200 per arm).")
```

## Comparison against published concentrations

Steffens 2025 reports no formal NCA parameters. It does report the observed TDM
peak and trough concentrations, which map onto the steady-state `cmax` and
`ctrough` of the modal 1000 mg q24h regimen. The published values pool **all**
regimens (225-1500 mg, q12-q72h), so this is an indicative rather than a
like-for-like comparison; the trough in particular has a published SD
(15.38 ug/mL) nearly twice its mean.

```{r nca-compare}
published <- tibble::tribble(
  ~regimen,        ~cmax,  ~ctrough,
  "1000 mg q24h",  41.96,  8.75      # Steffens 2025 Section 2.1, observed means
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res |> dplyr::filter(regimen == "1000 mg q24h"),
  reference = published,
  by = "regimen",
  params = c("cmax", "ctrough"),
  units = c(cmax = "ug/mL", ctrough = "ug/mL"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = "Simulated vs. published amikacin concentrations. * differs by >20%.")
```

```{r nca-compare-detail, echo = FALSE}
# The published "peak" was drawn 30 min after the end of a 1 h infusion, not at
# the true Cmax; compare on the paper's own sampling clock as well.
peak_at_sample <- sim_ss |>
  dplyr::filter(regimen == "1000 mg q24h", abs(tad - (tinf + 0.5)) < 1e-9) |>
  dplyr::pull(Cc)
trough_q12 <- nca_wide |>
  dplyr::filter(regimen %in% c("500 mg q12h", "750 mg q12h")) |>
  dplyr::pull(ctrough)
cmax_q24 <- nca_wide$cmax[nca_wide$regimen == "1000 mg q24h"]
ctrough_q24 <- nca_wide$ctrough[nca_wide$regimen == "1000 mg q24h"]
cmax_pct <- 100 * (median(cmax_q24) - 41.96) / 41.96
ctrough_pct <- 100 * (median(ctrough_q24) - 8.75) / 8.75
```

**Both rows exceed the 20% tolerance, in the directions the pooling predicts.**

* *Cmax is high by `r sprintf("%.0f%%", cmax_pct)`.*
  Two effects push the same way. The published mean pools doses from 225 mg
  upward, so it sits below a pure 1000 mg arm; and the published peak was drawn
  **30 min after the end of the infusion**, not at the true Cmax. Sampled on the
  paper's own clock (1.5 h after the start of the infusion), this model gives a
  median of `r sprintf("%.1f ug/mL", median(peak_at_sample))` rather than the
  `r sprintf("%.1f", median(cmax_q24))` ug/mL reported as `cmax` above --
  closer to the published 41.96 ug/mL, though the dose-pooling gap remains.
* *Ctrough is low by `r sprintf("%.0f%%", abs(ctrough_pct))`.* The published
  trough pools the q12h and q48-72h patients as well. In this simulation the
  q12h arms trough at a median of
  `r sprintf("%.1f ug/mL", median(trough_q12))` against
  `r sprintf("%.1f", median(ctrough_q24))` ug/mL
  for 1000 mg q24h, so pooling them in raises the observed mean well above the
  q24h-only value. The published trough SD (15.38 ug/mL) is nearly twice its
  mean, which is itself the signature of a heavily mixed set of intervals.

Neither discrepancy is evidence of a transcription error, and no parameter was
adjusted in response to them.

# Replicate Figure 4 -- probability of target attainment

Figure 4 of Steffens 2025 plots the probability of reaching Cmax/MIC >= 8 for
each simulated regimen against the MIC of the strain, with a 90% attainment
line.

```{r figure-4}
mics <- c(1, 2, 4, 8, 16, 32)

pta <- tidyr::crossing(nca_wide |> dplyr::select(regimen, id, cmax), MIC = mics) |>
  dplyr::group_by(regimen, MIC) |>
  dplyr::summarise(PTA = 100 * mean(cmax >= 8 * MIC), .groups = "drop") |>
  dplyr::left_join(regimens |> dplyr::select(regimen, dose, tau), by = "regimen") |>
  dplyr::arrange(tau, dose, MIC)

ggplot(pta, aes(MIC, PTA, colour = regimen)) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  geom_line() + geom_point() +
  scale_x_log10(breaks = mics) +
  labs(x = "MIC (mg/L)", y = "Probability of target attainment (%)",
       colour = "Regimen",
       title = "Cmax/MIC >= 8 attainment by regimen",
       caption = "Replicates Figure 4 of Steffens 2025 (n = 200 simulated patients per arm).")
```

```{r pta-table, echo = FALSE}
pta |>
  dplyr::select(regimen, MIC, PTA) |>
  tidyr::pivot_wider(names_from = MIC, values_from = PTA,
                     names_prefix = "MIC ") |>
  dplyr::mutate(dplyr::across(dplyr::starts_with("MIC "), ~ sprintf("%.1f", .x))) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(caption = "Simulated PTA (%) for Cmax/MIC >= 8.")
```

```{r pta-checks}
pta_at <- function(reg, mic) {
  v <- pta$PTA[pta$regimen == reg & pta$MIC == mic]
  if (length(v) != 1L) stop("no unique PTA row for '", reg, "' at MIC ", mic)
  v
}

# Structural claims that hold for ANY cohort this model can produce.
# (1) PTA is non-increasing in MIC -- true by construction of P(Cmax >= 8*MIC).
stopifnot(
  pta |>
    dplyr::group_by(regimen) |>
    dplyr::summarise(ok = all(diff(PTA) <= 0), .groups = "drop") |>
    dplyr::pull(ok) |>
    all()
)
# (2) Within a dosing interval, a higher dose never attains less. Exact here
# because every arm uses the same 200 virtual patients and Cmax is linear in
# dose (verified to 10 significant figures during development).
stopifnot(
  pta |>
    dplyr::group_by(tau, MIC) |>
    dplyr::arrange(dose, .by_group = TRUE) |>
    dplyr::summarise(ok = all(diff(PTA) >= 0), .groups = "drop") |>
    dplyr::pull(ok) |>
    all()
)
```

# Published-claim checklist

```{r claims}
ctau_over_target <- nca_wide |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(pct = 100 * mean(ctrough > 2.5), .groups = "drop")

claims <- tibble::tribble(
  ~Claim, ~Published, ~Simulated, ~Deviation,
  "CL at CRCL = 0 (L/h)", "1.49", sprintf("%.3f", cl_zero), FALSE,
  "CL at median CRCL (L/h), base model", "2.04", sprintf("%.3f", cl_median), FALSE,
  "Vc (L)", "23.18", sprintf("%.3f", vc_typ), FALSE,
  "1000 mg q24h PTA at MIC 4 (%)", ">= 90", sprintf("%.1f", pta_at("1000 mg q24h", 4)), TRUE,
  "1500 mg q24h PTA at MIC 8 (%)", "92.2", sprintf("%.1f", pta_at("1500 mg q24h", 8)), TRUE,
  "2000 mg q24h PTA at MIC 16 (%)", "< 90", sprintf("%.1f", pta_at("2000 mg q24h", 16)), FALSE,
  "Patients above the Ctrough 2.5 ug/mL toxicity target, 1000 mg q24h (%)",
    ">= 90", sprintf("%.1f", ctau_over_target$pct[ctau_over_target$regimen == "1000 mg q24h"]),
    FALSE
)

knitr::kable(claims, caption = "Published claims against this implementation. Rows marked Deviation are discussed below and are excluded from the gate.")
```

```{r claim-gates}
# Gated (non-deviation) claims.
stopifnot(abs(cl_zero - 1.49) < 0.01)
stopifnot(abs(cl_median - 2.04) < 0.01)
stopifnot(abs(vc_typ - 23.18) < 0.01)
# 2000 mg q24h falls short of 90% attainment above MIC 9 (Section 2.4 / Discussion).
stopifnot(pta_at("2000 mg q24h", 16) < 90)
# "none of the simulated doses were effective" against Ctrough <= 2.5 ug/mL
# (Section 2.4): a large majority of patients exceed the toxicity target. The
# bound is a majority claim, not the published 90%, so it survives cohort noise.
stopifnot(all(ctau_over_target$pct > 60))
```

# Assumptions and deviations

## Covariate centering (reconstructed, not printed)

Steffens 2025 never writes the CL-CRCL covariate equation. The uncentered
exponential form `log(CL) = log(1.49) + 0.004 * CRCL` was established from
Monolix's default handling of a continuous covariate plus the paper's own
internal consistency (`1.49 * exp(0.004 * 79.01) = 2.044` reproduces the
base-model CL of 2.04 L/h to three significant figures). The alternative -- a
median-centered form making 1.49 L/h the typical clearance -- cannot be excluded
from the text alone, but it leaves the 27% gap to the base model unexplained.
Readers refitting this model should confirm the centering against the authors if
it is load-bearing for their application.

## Variability scale

Table 2 prints `omega Cl` as `0.67 (74.8)` and `omega Vd` as `0.47 (49.89)`. The
parenthesised figures are lognormal %CV computed from omega as a **log-scale
standard deviation**: `sqrt(exp(0.67^2) - 1) = 75.3%` and
`sqrt(exp(0.47^2) - 1) = 49.7%`, matching the printed 74.8% and 49.89% to within
the two-decimal rounding of omega itself. Reading the same numbers as variances
would give 97.7% and 77.5%, which the paper does not print. `ini()` therefore
carries the squared values 0.4489 and 0.2209.

## Residual error

Section 2.2 records that a combined additive-plus-proportional error model was
*tested* during model building, but Table 2 and the Abstract both report the
final model with a **proportional** term only (0.38), and no additive estimate
is given anywhere. Only the proportional term is encoded.

## Probability of target attainment (known deviation)

Two PTA claims are not reproduced:

* Steffens 2025 reports **92.2%** attainment of Cmax/MIC >= 8 for 1500 mg q24h
  at MIC 8 mg/L; this implementation gives
  `r sprintf("%.1f%%", pta_at("1500 mg q24h", 8))`.
* Steffens 2025 states 1000 mg q24h reaches the target "for all simulated
  patients" up to MIC 4 mg/L; this implementation gives
  `r sprintf("%.1f%%", pta_at("1000 mg q24h", 4))`.

The shortfall is not a centering artefact -- it is present under *both* readings
of the covariate equation, because Cmax after a 1 h infusion is governed almost
entirely by volume, and the published `omega Vd` of 0.47 (49.9% CV) spreads
Cmax far too widely for 92% of patients to clear a threshold that sits close to
the typical-patient value.

```{r pta-sensitivity, echo = FALSE}
cmax_1500 <- nca_wide$cmax[nca_wide$regimen == "1500 mg q24h"]
# What log-scale SD on Cmax would put 92.2% of patients above the 64 ug/mL
# target, given the median this model produces?
sd_needed <- log(median(cmax_1500) / 64) / qnorm(0.922)
cv_needed <- 100 * sqrt(exp(sd_needed^2) - 1)
cv_actual <- 100 * sqrt(exp(sd(log(cmax_1500))^2) - 1)
```

The median steady-state Cmax at 1500 mg q24h is
`r sprintf("%.1f ug/mL", median(cmax_1500))` against the 64 ug/mL target, with a
realised spread of `r sprintf("%.0f%%", cv_actual)` CV. Putting 92.2% of
patients above the target from that median would require the spread to be about
`r sprintf("%.0f%%", cv_needed)` CV -- under a third of what the published
`omega Vd` of 0.47 implies. No plausible centering of the creatinine-clearance
covariate closes a gap of that size.

The paper does not state what creatinine-clearance distribution its Simulx
Monte Carlo used, nor whether Cmax was taken at the end of infusion, at the
paper's 30-min post-infusion sampling time, or from the first dose rather than
at steady state. Any of these could account for part of the gap. No parameter
was tuned to close it; the deviation is recorded and excluded from the assertion
gate, and the structural PTA claims (monotone decline with MIC, monotone rise
with dose) are gated instead.

## Simulation assumptions

* **Creatinine clearance distribution** -- drawn lognormal with median
  79.01 mL/min and mean 100.3 mL/min (Table 1), truncated to the observed range
  12.97-517.97. The paper reports an SD of 102.2 mL/min, heavier-tailed than
  this lognormal (SD ~78), driven by the single 517.97 mL/min subject.
* **Infusion duration** -- 1 h for every simulated regimen, per Section 4.4.2.
  The observed TDM data came from infusions whose duration is not reported.
* **Steady state** -- 20 days of dosing before the observed interval, many
  multiples of the ~8 h half-life.
* **Random effects** -- drawn in R and passed as data columns so all five
  regimens share the same 200 virtual patients. This makes the dose-ranking
  comparison exact; it does not change any per-arm marginal distribution.
* **Comparison caveat** -- the published mean peak and trough pool all observed
  regimens (225-1500 mg, q12-q72h) and all sampling days, while the simulated
  values are for 1000 mg q24h at steady state.
* No covariate other than creatinine clearance is in the final model. Body
  weight, age, BMI, sex and dialysis status were screened by the authors and not
  retained; they are recorded in the model file's `covariatesDataExcluded`.
