---
title: "Stevens_2012_remoxipride"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Stevens_2012_remoxipride}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Stevens_2012_remoxipride"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Stevens_2012_remoxipride"))$description`
* Article: <https://doi.org/10.1007/s10928-012-9262-4>
* DDMORE Foundation Model Repository entry: [DDMODEL00000268](https://repository.ddmore.eu/model/DDMODEL00000268)

This model was extracted from the DDMORE Foundation Model Repository
bundle for `DDMODEL00000268` (scraped to
`dpastoor/ddmore_scraping/268/`). The bundle contains:

- `Executable_PK_rats.txt` -- NMTRAN control stream (NONMEM 7.3,
  ADVAN9 + NWPRI prior on the prolactin elimination and EC50
  THETAs).
- `Output_real_PK_rats.lst` -- NONMEM listing on the original real
  rat dataset; the `FINAL PARAMETER ESTIMATE` block (lines
  1283-1313) supplies the parameter values used here.
- `Output_simulated_PK_rats.lst` and `Simulated_PK_rats.csv` -- a
  shipped simulated dataset and re-fit listing, used as the
  self-consistency reference in this vignette.
- `DDMODEL00000268.rdf` -- model classification (PK/PD purpose,
  research-stage metadata).
- `20120910- DDMORE-WP1.3. Specificationdocument_remoxipride
  ECMdL.doc` -- DDMORE-internal model-encoding specification
  document; reproduces the publication's abstract and parameter
  labels (`BSLsd`, `BSLdd`, `kr,pr`, `kel,pr`, `EC50,pr`,
  `EC50,rem`) but not the parameter values themselves (those are
  embedded as MS Equation 3.0 OLE objects that are not
  text-extractable). Confirms the publication match: Stevens et
  al. 2012, J Pharmacokinet Pharmacodyn, doi:10.1007/s10928-012-9262-4.

The Stevens 2012 publication itself is not on disk in this
worktree, so the standard publication-figure replication and
PKNCA-vs-published-NCA checks are out of scope. The validation in
this vignette therefore follows the F.2 self-consistency and F.3
mechanistic-sanity substitutes from the extraction skill.

# Population

Stevens 2012 fits the prolactin pool model to plasma-prolactin time
courses from male Wistar rats with chronic intracerebral
microdialysis cannulae and femoral-vein cannulae. The shipped
simulated dataset contains 43 rats across two studies (`STUD = 1`
and `STUD = 3`) and four dose levels (3.8, 4.0, 8.0, 16.0 mg/kg
intravenous remoxipride), all at body weight 0.252 kg. Three
experimental cohorts feed the joint fit:

1. A baseline-variation study without drug administration (used to
   pin the typical baseline plasma prolactin TBSL).
2. A single-dose dose-response study at 4 / 8 / 16 mg/kg IV
   (`STUDY_DD = 0`, the reference category).
3. A double-dose study at 3.8 mg/kg IV with varying inter-dose
   intervals (`STUDY_DD = 1`, encoded via `STUD == 3` in the
   source data; the typical baseline is shifted by
   `1 + e_studydd_bsl = 1 - 0.290 = 0.710` for this group).

The PK structural parameters are fixed from a previously published
rat remoxipride PK study (Westerhout et al. 2011,
doi:10.1007/s11095-011-0395-8) and reproduced here verbatim from
`THETA(3)..THETA(13)` `FIX` in the DDMORE control stream. The PD
parameters (TH1, TH2, TH14-TH19) are estimated under an NWPRI
prior on TH1 (prolactin elimination rate) and TH2 (remoxipride
brain-ECF EC50).

The model is preclinical (rat-only). Stevens et al. 2012 separately
extrapolated the model to humans by allometric scaling of the
prolactin release and elimination rates plus literature-derived
human baseline prolactin and remoxipride PK; that human-scaled
projection is **not** encoded in this implementation -- only the
rat-fitted typical-value parameters are.

The same metadata is available programmatically via
`readModelDb("Stevens_2012_remoxipride")$population`.

# Source trace

Per-parameter and per-equation origin (also recorded as in-file
comments in `inst/modeldb/ddmore/Stevens_2012_remoxipride.R`):

| Equation / parameter   | Value                | Source location |
|------------------------|----------------------|-----------------|
| `lcl`                  | `fixed(log(1.12))`   | `Output_real_PK_rats.lst` line 1283, `TH3 FIX` (.mod L107: CL3 L/h) |
| `lvc`                  | `fixed(log(0.0881))` | `.lst` L1283, `TH4 FIX` (.mod L108: V3 L/kg) |
| `lq_brain`             | `fixed(log(0.700))`  | `.lst` L1283, `TH5 FIX` (.mod L109: Q4 plasma<->brain ECF) |
| `lvbrain`              | `fixed(log(0.873))`  | `.lst` L1283, `TH6 FIX` (.mod L110: V4 brain ECF, L/kg) |
| `lq`                   | `fixed(log(1.20))`   | `.lst` L1283, `TH7 FIX` (.mod L111: Q5 plasma<->peripheral) |
| `lvp`                  | `fixed(log(0.417))`  | `.lst` L1283, `TH8 FIX` (.mod L112: V5 peripheral, L/kg) |
| `lfk40`                | `fixed(log(0.302))`  | `.lst` L1283, `TH9 FIX` (.mod L113: K40 = FK40 * K30) |
| `lka`                  | `fixed(log(1.54))`   | `.lst` L1283, `TH10 FIX` (.mod L114: intranasal absorption rate KA) |
| `lftot`                | `fixed(log(0.892))`  | `.lst` L1283, `TH11 FIX` (.mod L115: total intranasal bioavailability) |
| `f_systemic`           | `fixed(0.249)`       | `.lst` L1283, `TH12 FIX` (.mod L116: F1 = FTOT * THETA(12)) |
| `lk24`                 | `fixed(log(0.0331))` | `.lst` L1283, `TH13 FIX` (.mod L117: K24 * 1000 = 33.1, divided by 1000 inside $PK) |
| `lkel_prl` (= K70)     | `log(5.20)`          | `.lst` L1283, `TH1` (estimated under NWPRI; .mod L60) |
| `lec50_prl`            | `log(0.0510)`        | `.lst` L1283, `TH2` (estimated under NWPRI; .mod L62) |
| `lbsl`                 | `log(6.64)`          | `.lst` L1284, `TH14` (.mod L56: typical baseline plasma prolactin, single-dose study) |
| `lemax_prl`            | `log(17.5)`          | `.lst` L1284, `TH15` (.mod L61) |
| `lstdm`                | `log(0.125)`         | `.lst` L1284, `TH16` (.mod L63: positive-feedback synthesis-stimulation slope) |
| `e_studydd_bsl`        | `-0.290`             | `.lst` L1284, `TH17` (.mod L56: TBSL = THETA(14) * (1 + THETA(17) * STDY)) |
| `lk_release` (= K67)   | `log(0.740)`         | `.lst` L1284, `TH18` (.mod L59: lactotroph->plasma release rate) |
| `gamma_prl`            | `fixed(1.00)`        | `.lst` L1284, `TH19 FIX` (.mod L64: Hill coefficient on the brain-ECF remoxipride stimulation term) |
| `etalbsl`              | `~ 0.0465`           | `.lst` line 1294, `OMEGA(1,1)` (log-scale variance on baseline prolactin; the .mod also fixes OMEGA(2,2) and OMEGA(3,3) to zero) |
| `propSd`               | `sqrt(0.0780)`       | `.lst` L1310, `SIGMA(1,1)` (proportional residual; .mod L143) |
| `addSd`                | `sqrt(6.62)`         | `.lst` L1313, `SIGMA(2,2)` (additive residual on prolactin in ug/L; .mod L144) |
| `d/dt(depot) = -ka * depot` | n/a              | `.mod` `$DES` L75 (DADT(1)) |
| `d/dt(depot_brain) = -k24 * depot_brain` | n/a   | `.mod` `$DES` L76 (DADT(2)) |
| `d/dt(central) = ka*depot - kel_plasma*central - k34*central - k35*central + k43*brain_ecf + k53*peripheral1` | n/a | `.mod` `$DES` L77 (DADT(3)) |
| `d/dt(brain_ecf) = k24*depot_brain + k34*central - k43*brain_ecf - k40*brain_ecf` | n/a | `.mod` `$DES` L78 (DADT(4)) |
| `d/dt(peripheral1) = k35*central - k53*peripheral1` | n/a | `.mod` `$DES` L88 (DADT(5)) |
| `eff = 1 + emax_prl * cp_brain^gamma_prl / (ec50_prl^gamma_prl + cp_brain^gamma_prl)` | n/a | `.mod` `$DES` L80-86 (drug effect on prolactin release) |
| `lna7 = max(prolactin, bsl_i)`              | n/a | `.mod` `$DES` L85-93 (positive-feedback floor) |
| `std = stdm * (lna7 - bsl_i)`               | n/a | `.mod` `$DES` L95 |
| `kf = bsl_i * kel_prl`                      | n/a | `.mod` `$PK` L68 (KF = BSL * K70) |
| `kft = kf * (1 + std)`                      | n/a | `.mod` `$DES` L96 |
| `d/dt(lactotroph) = kft - lactotroph * k_release * eff` | n/a | `.mod` `$DES` L97 (DADT(6)) |
| `d/dt(prolactin) = lactotroph * k_release * eff - prolactin * kel_prl` | n/a | `.mod` `$DES` L98 (DADT(7)) |
| `lactotroph(0) = bsl * kel_prl / k_release` | n/a | `.mod` `$PK` L66 (`A_0(6) = (BSL * K70) / K67`) |
| `prolactin(0) = bsl`                        | n/a | `.mod` `$PK` L67 (`A_0(7) = BSL`) |
| `prolactin ~ prop(propSd) + add(addSd)`     | n/a | `.mod` `$ERROR` L102-103 (`Y = IPRED * (1 + EPS(1)) + EPS(2)`) |
| `f(depot) <- ftot * f_systemic`             | n/a | `.mod` `$PK` L37 (`F1 = FTOT * THETA(12)`) |
| `f(depot_brain) <- ftot * (1 - f_systemic)` | n/a | `.mod` `$PK` L38 (`F2 = FTOT - F1`) |

The .mod also includes two NONMEM `EXIT 1 ...` early-exit guards
(`IF(K30.LT.K40) EXIT 1 200`, `IF(KA.LT.K24) EXIT 1 400`, `.mod`
L46-47). With the published parameter values both inequalities hold
(`K30 = 12.71 > K40 = 3.84`; `KA = 1.54 > K24 = 0.033`), so the
guards never fire; they are not reproduced in the nlmixr2
implementation.

# Validation strategy

This bundle ships a simulated dataset (`Simulated_PK_rats.csv`) and
a re-fit listing (`Output_simulated_PK_rats.lst`) but the linked
publication is not on disk. Following
`references/ddmore-source.md` Section  "Validation strategy by model type"
(decision tree -> no PKNCA -> mechanistic / endogenous -> F.2 + F.3
substitutes), this vignette validates by:

1. **Steady-state hold (F.1 / endogenous-validation).** With no
   drug administration, the lactotroph and plasma-prolactin states
   must stay at their analytic baselines. The .mod sets
   `A_0(6) = BSL * K70 / K67` and `A_0(7) = BSL` so the system is
   at steady state at `t = 0`; integrating forward without dose
   inputs must keep the states at those values within numerical
   tolerance.
2. **Drug-response sanity.** A single intravenous remoxipride dose
   must produce a transient brain-ECF concentration that drives an
   Emax-shaped stimulation of plasma prolactin, with prolactin
   rising above baseline, peaking, and decaying back as the brain
   ECF clears. A larger dose must produce a larger (and longer-
   lasting) response.
3. **F.2 self-consistency against the bundle's simulated dataset.**
   Re-simulate the bundle's typical-value trajectory at the
   conditions present in `Simulated_PK_rats.csv` (rat 304:
   3.8 mg/kg IV at `t = 0` and `t = 4` h, `STUD = 3`,
   `WT = 0.252` kg) and confirm the simulated prolactin profile
   visually tracks the observed `DV` values from the CSV. The
   bundle's `Output_simulated_PK_rats.lst` records the same
   trajectory under the .mod's NONMEM solver; an exact
   numerical match is not the goal (rxode2 and NONMEM
   `ADVAN9 TOL=3` use different ODE solvers and tolerances), but
   the typical-value shape and magnitude must be consistent.

# Setup

```{r model-load}
mod <- rxode2::rxode2(readModelDb("Stevens_2012_remoxipride"))
mod_typical <- rxode2::zeroRe(mod)

state_names <- mod$state
state_names
```

# 1. Steady-state hold (no drug)

With no remoxipride dose, the analytic baselines
`prolactin(0) = bsl` and `lactotroph(0) = bsl * kel_prl / k_release`
must hold across the integration horizon. For the single-dose
study reference (`STUDY_DD = 0`) this means `prolactin = 6.64`
ug/L and `lactotroph = 6.64 * 5.20 / 0.74 ~= 46.66` "amount-units"
(the prolactin-pool compartment has no scale factor in the source,
so the lactotroph state is in the same units as plasma prolactin
times time-dimension; what matters is that it stays constant).

```{r baseline-no-drug}
ev_baseline <- rxode2::et(seq(0, 24, by = 1))
ev_baseline_df <- as.data.frame(ev_baseline)
ev_baseline_df$WT <- 0.252
ev_baseline_df$STUDY_DD <- 0L

sim_baseline <- rxode2::rxSolve(mod_typical, ev_baseline_df,
                                returnType = "data.frame")

baseline_summary <- sim_baseline |>
  dplyr::filter(time %in% c(0, 1, 6, 12, 24)) |>
  dplyr::transmute(
    time,
    prolactin  = round(prolactin, 6),
    lactotroph = round(lactotroph, 6),
    central    = round(central, 8),
    brain_ecf  = round(brain_ecf, 8)
  )
knitr::kable(
  baseline_summary,
  caption = paste(
    "Steady-state hold: with no dose, plasma prolactin stays at",
    "6.64 ug/L and the lactotroph store stays at 46.66 across the",
    "24-hour horizon."
  )
)
```

The same hold for the double-dosing-study cohort
(`STUDY_DD = 1`) shifts the baseline by `1 + e_studydd_bsl = 0.71`
to `prolactin = 6.64 * 0.71 ~= 4.71` ug/L.

```{r baseline-no-drug-dd}
ev_baseline_dd_df <- ev_baseline_df
ev_baseline_dd_df$STUDY_DD <- 1L
sim_baseline_dd <- rxode2::rxSolve(mod_typical, ev_baseline_dd_df,
                                   returnType = "data.frame")
cat(sprintf(
  "STUDY_DD = 1 baseline: prolactin = %.4f ug/L (expected %.4f).\n",
  sim_baseline_dd$prolactin[1],
  6.64 * (1 + (-0.290))
))
```

# 2. Drug-response sanity (3.8, 8, and 16 mg/kg IV)

A typical-value simulation at three rat-equivalent IV dose levels.
Each dose is a 30-minute IV infusion with infusion rate matching
the simulated dataset's `RATE = 1.95` mg/h pattern (`RATE = AMT *
2`, equivalent to a 0.5-h infusion duration).

```{r drug-response, fig.height = 4.0}
make_iv_events <- function(dose_mgkg, wt_kg = 0.252, study_dd = 0L) {
  amt <- dose_mgkg * wt_kg
  ev  <- rxode2::et(amt = amt, rate = amt * 2, time = 0,
                    cmt = "central") |>
    rxode2::et(seq(0, 12, by = 0.05))
  ev_df <- as.data.frame(ev)
  ev_df$WT <- wt_kg
  ev_df$STUDY_DD <- study_dd
  ev_df
}

dose_labels <- c("3.8 mg/kg", "8 mg/kg", "16 mg/kg")
sim_doses <- dplyr::bind_rows(lapply(seq_along(dose_labels), function(i) {
  d <- c(3.8, 8.0, 16.0)[i]
  rxode2::rxSolve(mod_typical, make_iv_events(d),
                  returnType = "data.frame") |>
    dplyr::mutate(dose = factor(dose_labels[i], levels = dose_labels))
}))

ggplot(sim_doses, aes(time, prolactin, colour = dose)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = 6.64, linetype = "dashed",
             colour = "grey50") +
  labs(
    x = "Time (h)",
    y = "Plasma prolactin (ug/L)",
    title = paste(
      "Typical-value plasma prolactin vs time, single 30-min IV",
      "remoxipride infusion (rat, STUDY_DD = 0)"
    ),
    caption = paste(
      "Dashed line = baseline plasma prolactin (6.64 ug/L).",
      "Larger doses produce larger and longer-lasting peaks,",
      "consistent with an Emax-shaped stimulation of prolactin",
      "release driven by transient brain-ECF remoxipride."
    ),
    colour = "IV dose"
  )

dose_peaks <- sim_doses |>
  dplyr::group_by(dose) |>
  dplyr::summarise(
    peak_prolactin_ug_L = round(max(prolactin), 2),
    time_to_peak_h      = round(time[which.max(prolactin)], 2),
    return_to_baseline_h = round(
      time[which(time > time[which.max(prolactin)] &
                 prolactin < 6.64 * 1.01)[1]],
      2
    ),
    .groups = "drop"
  )
knitr::kable(
  dose_peaks,
  caption = paste(
    "Peak plasma prolactin scales with IV dose level. Time to peak",
    "(~0.2 h) tracks the rapid brain-ECF kinetics; return to within",
    "1% of baseline takes longer at higher doses because the",
    "lactotroph pool depletes and refills on a slower time scale."
  )
)
```

# 3. F.2 self-consistency vs. the bundle's simulated dataset

Re-simulate the bundle's typical-value trajectory at the same
dosing conditions used by the first rat in
`Simulated_PK_rats.csv` (ID 304, double-dose study, 3.8 mg/kg IV
at `t = 0` and `t = 4` h, body weight 0.252 kg) and overlay the
observed `DV` values from the CSV.

```{r f2-self-consistency, fig.height = 4.5}
# The Simulated_PK_rats.csv shipped with the DDMORE bundle is not
# distributed with this package; for the self-consistency check we
# reconstruct rat 304's event schedule from its hard-coded
# definition and compare against an inline copy of the
# observed-DV vector at the documented sample times.
rat304_obs <- tibble::tribble(
  ~time, ~DV,
   0.000,  7.79,
   0.167,  8.81,
   0.367, 12.50,
   0.450, 10.70,
   0.583, 10.20,
   0.667,  3.92,
   1.000,  6.00,
   2.000,  0.546,
   4.100,  1.33,
   4.230, 19.40,
   4.300,  0.940,
   4.430, -1.62,
   4.570,  7.39,
   4.650,  7.80,
   4.770,  8.95,
   5.100,  1.01,
   5.770, -0.239,
   7.100,  2.85,
   8.100,  2.78
)

ev_304 <- rxode2::et(amt = 0.97, rate = 1.95, time = 0,
                     cmt = "central") |>
  rxode2::et(amt = 0.97, rate = 1.95, time = 4,
             cmt = "central") |>
  rxode2::et(seq(0, 9, by = 0.02))
ev_304_df <- as.data.frame(ev_304)
ev_304_df$WT <- 0.252
ev_304_df$STUDY_DD <- 1L

sim_304 <- rxode2::rxSolve(mod_typical, ev_304_df,
                           returnType = "data.frame")

ggplot() +
  geom_line(data = sim_304, aes(time, prolactin),
            colour = "steelblue", linewidth = 0.6) +
  geom_point(data = rat304_obs, aes(time, DV),
             colour = "tomato", size = 1.6) +
  geom_hline(yintercept = 6.64 * 0.71, linetype = "dashed",
             colour = "grey50") +
  labs(
    x = "Time (h)",
    y = "Plasma prolactin (ug/L)",
    title = paste(
      "F.2 self-consistency: simulated typical-value trajectory",
      "(blue line) vs observed DV (red dots) for rat 304 from",
      "Simulated_PK_rats.csv"
    ),
    caption = paste(
      "Two 30-min IV infusions of 3.8 mg/kg (cumulative AMT 0.97",
      "mg per dose) at t = 0 and t = 4 h, double-dosing study",
      "(STUDY_DD = 1). Dashed line = STUDY_DD = 1 baseline (~4.71",
      "ug/L). The observed DV vector is the simulated noise-added",
      "data shipped with the bundle; the typical-value trajectory",
      "(no IIV, no residual error) traces the centre of the",
      "scatter, consistent with the bundle's own typical-value",
      "trajectory in Output_simulated_PK_rats.lst."
    )
  )
```

The blue typical-value trajectory recovers the observed pattern:
- the first IV pulse drives plasma prolactin up to a peak around
  10-20 ug/L within ~0.3 h,
- prolactin then dips below baseline (lactotroph depletion) and
  starts recovering,
- the second IV pulse at t = 4 h re-stimulates the system,
  producing a second peak,
- the late-time tail returns toward the STUDY_DD = 1 baseline
  (~4.71 ug/L).

A small fraction of the observed DV values are negative (e.g.
-1.62 at t = 4.43 h) -- these are the result of the additive
residual `addSd ~= 2.57` ug/L applied on top of an IPRED that drops
toward zero during the lactotroph-depletion phase. The model
reproduces the depletion phase (the typical-value trajectory dips
below baseline near t = 1 h and t = 5 h) but cannot produce
negative concentrations on its own; the simulated negatives are
purely a function of the additive-error noise model.

# Assumptions and deviations

- **Preclinical (rat-only) implementation.** Stevens et al. 2012
  also developed an allometrically scaled human projection of the
  prolactin pool model. Only the rat-fitted parameters (with PK
  fixed from Westerhout 2011) are encoded here; the human-scaled
  parameters are not. Users wanting to simulate the human
  projection should not load this model.
- **Non-canonical compartment names.** `depot_brain`, `brain_ecf`,
  `lactotroph`, and `prolactin` are paper-meaningful names (the
  source biology distinguishes a systemic-absorption depot from a
  direct nose-to-brain depot, a microdialysed brain-ECF
  compartment from the plasma central, a lactotroph pool from the
  plasma prolactin pool). They do not appear in
  `R/conventions.R::compartments`; `checkModelConventions()`
  flags them as warnings, which is the same convention adopted by
  other paper-specific mechanistic models in
  `inst/modeldb/ddmore/` (e.g. `Mohamed_2016_colistin_meropenem`,
  `Bizzotto_2016_glucose`).
- **Non-canonical observation name.** The observation variable is
  `prolactin`, not `Cc`, because the observed quantity is plasma
  prolactin (an endogenous hormone) rather than a drug
  concentration. `checkModelConventions()` flags this as a
  warning; the rationale is the same as for `tumorSize`,
  `freeIgE`, and other paper-named non-PK output variables in
  `inst/references/covariate-columns.md`.
- **Paper-specific covariate `STUDY_DD`.** The .mod's `STUD` data
  column is a paper-specific integer (1 = baseline-variation
  study, 2 = single-dose study, 3 = double-dose study); the .mod
  uses only `STUD == 3`, so this implementation collapses the
  covariate into a binary `STUDY_DD` (1 = double-dose-study
  cohort, 0 = the reference categories). `STUDY_DD` is not in
  `inst/references/covariate-columns.md`;
  `checkModelConventions()` flags it as a warning. The source-
  data column is `STUD` and the mapping (`STUDY_DD =
  as.integer(STUD == 3)`) is documented in `covariateData`.
- **Brain-ECF concentration units.** The .mod's `S3 = V3` and
  `S4 = V4` with `V3` / `V4` in L/kg means the central and
  brain-ECF "concentrations" are body-weight-normalized:
  `central / vc` carries an extra factor of kg (mg*kg/L). The
  `ec50_prl` parameter is on the same scale (mg*kg/L) so the
  Emax driver is dimensionally consistent inside the model. Users
  comparing brain-ECF concentrations against absolute mg/L values
  in the literature should multiply by body weight (in kg).
- **No publication on disk.** The Stevens 2012 paper itself is
  paywalled (J Pharmacokinet Pharmacodyn) and is not in the
  worktree. The DDMORE specification document confirms the
  publication match (title, authors, journal, DOI in the
  document's reference block) but stores parameter tables as
  embedded MS Equation 3.0 OLE objects that are not text-
  extractable. The cross-check against published parameter
  tables is therefore not performed; the `.lst` final estimates
  are the sole source of values.
- **Solver tolerance.** The .mod uses `ADVAN9 TOL=3`. rxode2's
  default LSODA tolerances (atol=1e-8, rtol=1e-6) are tighter,
  so trajectory differences at the third significant figure
  between the rxode2 simulation and the bundle's
  `Output_simulated_PK_rats.lst` are expected; differences at
  the first or second significant figure would indicate a
  translation bug.
