---
title: "Veliparib (Stodtmann 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Veliparib (Stodtmann 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Stodtmann_2021_veliparib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Stodtmann_2021_veliparib"))$description`
* Article: <https://doi.org/10.1002/jcph.1875> (open access, PMC8453554)

Stodtmann et al. pooled individual veliparib concentrations from nine AbbVie
studies (six phase 1, one phase 2 and two phase 3) into one population PK
analysis. The final model is one-compartment with first-order absorption and
linear clearance. The time base is **days** and concentrations are in
**ug/mL**, exactly as the paper reports them (Table 4: `CL/F (L/day)`,
`ka (1/day)`; additive errors in ug/mL).

The library also contains `Niu_2017_veliparib`, an earlier single-study
parent-plus-M8 model from a different group (hours, ng/mL, two compartments).
The two models are independent fits to different data.

# Population

The model was fit to 9160 plasma concentrations from 1470 adults (Stodtmann
2021 Table 3, All Subjects): median age 55 years (22-86), median body weight
66.0 kg (35.7-182), 97% female, 75% White, 12% Asian, 4% Black. The subjects
had ovarian, fallopian tube or primary peritoneal cancer (58%), breast cancer
(38%) or other solid tumours (5%), and received oral veliparib 10-400 mg twice
daily, alone or with temozolomide, carboplatin/gemcitabine or
carboplatin/paclitaxel (Table 1). Median Cockcroft-Gault creatinine clearance
was 96.5 mL/min (28.2-289); 32% had mild and 9% moderate renal impairment.
Median serum albumin was 40 g/L (20-52).

The same information is available programmatically via
`readModelDb("Stodtmann_2021_veliparib")()$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its row in Table 4. The
typical-value equations are printed in the Results section *Significant
Covariates*:

* `CL/F = 479 * (min(CrCL,120)/120)^0.513 * (ALB/40)^0.427 * 1.20^Male * 0.885^CYP2D6` L/day
* `Vc/F = 152 * (WTKG/70)^0.505 * (ALB/40)^0.260 * 1.25^Male` L
* `ka = 59.4 * 1.11^Fasting * 0.356^Fed` 1/day

| Parameter | Value | Source location |
|-----------|-------|-----------------|
| `lka` | log(59.4) 1/day | Table 4 'ka (1/day)' |
| `lcl` | log(479) L/day | Table 4 'CL/F (L/day)' |
| `lvc` | log(152) L | Table 4 'Vc/F (L)' |
| `e_fed_ka` | 0.356 | Table 4 'Fed on ka' |
| `e_fast_ka` | 1.11 | Table 4 'Fasting on ka' |
| `e_crcl_cl` | 0.513 | Table 4 'Creatinine clearance on CL/F'; cap and reference 120 mL/min from the CL/F equation and Table 2 |
| `e_alb_cl` | 0.427 | Table 4 'Albumin on CL/F'; reference 40 g/L from the CL/F equation |
| `e_male_cl` | 1.20 | Table 4 'Male on CL/F' |
| `e_cyp2d6inh_cl` | 0.885 | Table 4 'Strong inhibitors of CYP2D6 on CL/F' |
| `e_wt_vc` | 0.505 | Table 4 'Body weight on Vc/F'; reference 70 kg from Table 2 and the Vc/F equation |
| `e_alb_vc` | 0.260 | Table 4 'Albumin on Vc/F' |
| `e_male_vc` | 1.25 | Table 4 'Male on Vc/F' |
| `etalcl` | 0.085 (variance) | Table 4 'BSV on CL/F' (29.8% CV) |
| `etalvc` | 0.064 (variance) | Table 4 'BSV on Vc/F' (25.7% CV) |
| `addSd_abs`, `propSd_abs` | 0.004 ug/mL, 0.208 | Table 4 absorption-phase error rows |
| `addSd_elim`, `propSd_elim` | 2.96e-7 ug/mL, 0.078 | Table 4 elimination-phase error rows |
| Phase switch at 2.5 h after dose | 2.5/24 day | Results: 'absorption phase (before Tmax at 2.5 hours)' |
| `d/dt(depot)`, `d/dt(central)` | n/a | Results: '1-compartment model with linear clearance and first-order absorption' |

# Covariate effects on steady-state exposure

The paper reports its covariate effects as the change in steady-state AUC,
computed as dose divided by clearance, relative to a reference subject (female,
no strong CYP2D6 inhibitor, CrCL >= 120 mL/min, albumin 40 g/L; Results and
Figure 3). Because `AUCss = Dose / (CL/F)`, each effect is a pure function of
the typical-value clearance, so the packaged model must reproduce the printed
numbers from its own parameters. The check below reads the typical CL/F from a
typical-value solve of the packaged model (not from a re-typed formula).

The paper's values are medians of its bootstrap distribution; the model uses the
point estimates. The two agree to within a few tenths of a percentage point for
every row, so the gate is 0.5 percentage points.

```{r cov-effects}
mod <- readModelDb("Stodtmann_2021_veliparib")
mod_typ <- rxode2::zeroRe(mod)

scenarios <- tibble::tribble(
  ~scenario, ~CRCL, ~ALB, ~SEXF, ~CONMED_CYP2D6_INH, ~published_pct,
  "Reference", 120, 40, 1, 0, 0,
  "Mild renal impairment (CrCL 75 mL/min)", 75, 40, 1, 0, 27.3,
  "Moderate renal impairment (CrCL 45 mL/min)", 45, 40, 1, 0, 65.4,
  "Male", 120, 40, 0, 0, -16.5,
  "Strong CYP2D6 inhibitor", 120, 40, 1, 1, 13.0,
  "Albumin 45 g/L", 120, 45, 1, 0, -4.91,
  "Albumin 35 g/L", 120, 35, 1, 0, 5.87
) |>
  mutate(id = seq_len(n()), WT = 70, FED = 0, FED_MISSING = 1)

ev_typ <- scenarios |>
  select(id, CRCL, ALB, SEXF, CONMED_CYP2D6_INH, WT, FED, FED_MISSING) |>
  tidyr::crossing(time = c(0, 1)) |>
  mutate(evid = 0L, amt = 0, cmt = "central")

sim_typ <- rxode2::rxSolve(mod_typ, events = ev_typ, keep = "id") |>
  as.data.frame()
cl_typ <- sim_typ |>
  group_by(id) |>
  summarise(cl = first(cl), vc = first(vc), .groups = "drop")

cov_tab <- scenarios |>
  left_join(cl_typ, by = "id") |>
  mutate(
    model_pct = 100 * (cl[scenario == "Reference"] / cl - 1),
    diff_pts = model_pct - published_pct
  )

cov_tab |>
  select(scenario, cl, published_pct, model_pct, diff_pts) |>
  dplyr::rename(
    "Scenario" = scenario,
    "Typical CL/F (L/day)" = cl,
    "Published AUCss change (%)" = published_pct,
    "Model AUCss change (%)" = model_pct,
    "Difference (points)" = diff_pts
  ) |>
  knitr::kable(digits = 2, caption = "Replicates the covariate effects of Stodtmann 2021 Results and Figure 3.")

stopifnot(
  nrow(cov_tab) == 7L,
  abs(cov_tab$cl[cov_tab$scenario == "Reference"] - 479) < 1e-6,
  all(abs(cov_tab$diff_pts) < 0.5)
)
```

The Discussion also states that a 10 kg change in body weight changes Vc/F by
only about 7%. Around the 66 kg median, `(76/66)^0.505 = 1.074`:

```{r wt-vc}
ev_wt <- tibble(id = 1:2, WT = c(66, 76)) |>
  mutate(CRCL = 120, ALB = 40, SEXF = 1, CONMED_CYP2D6_INH = 0, FED = 0, FED_MISSING = 1) |>
  tidyr::crossing(time = c(0, 1)) |>
  mutate(evid = 0L, amt = 0, cmt = "central")
vc_wt <- rxode2::rxSolve(mod_typ, events = ev_wt, keep = "id") |>
  as.data.frame() |>
  group_by(id) |>
  summarise(vc = first(vc), .groups = "drop")
vc_change_pct <- 100 * (vc_wt$vc[2] / vc_wt$vc[1] - 1)
vc_change_pct
stopifnot(abs(vc_change_pct - 7) < 1)
```

# Virtual cohort

The observed data are not public. The virtual cohort draws covariates
independently from distributions matched to the Table 3 summaries (All
Subjects), redrawing any value outside the observed range rather than clamping
it. The prandial state is left unknown (`FED_MISSING = 1`), as it was for 95%
of the fitted subjects. Three twice-daily regimens used in the phase 2/3
studies are simulated, 150 subjects each.

```{r cohort}
set.seed(20210901)
rxode2::rxSetSeed(20210901)

# Draw from a sampler, redrawing values outside [lo, hi].
draw_in_range <- function(n, sampler, lo, hi) {
  x <- sampler(n)
  bad <- x < lo | x > hi
  while (any(bad)) {
    x[bad] <- sampler(sum(bad))
    bad <- x < lo | x > hi
  }
  x
}

make_cohort <- function(n, dose_mg, id_offset) {
  tibble(
    id = id_offset + seq_len(n),
    regimen = paste0(dose_mg, " mg BID"),
    dose = dose_mg,
    WT = draw_in_range(n, function(k) rlnorm(k, log(66), 0.25), 35.7, 182),
    ALB = draw_in_range(n, function(k) rnorm(k, 39.6, 5.05), 20, 52),
    CRCL = draw_in_range(n, function(k) rlnorm(k, log(96.5), 0.34), 28.2, 289),
    SEXF = rbinom(n, 1, 0.97),
    CONMED_CYP2D6_INH = rbinom(n, 1, 0.05),
    FED = 0,
    FED_MISSING = 1
  )
}

n_per_arm <- 150L
cohort <- bind_rows(
  make_cohort(n_per_arm, 120, 0L),
  make_cohort(n_per_arm, 150, n_per_arm),
  make_cohort(n_per_arm, 300, 2L * n_per_arm)
)

# Twice-daily dosing for 7 days (tau = 0.5 day); steady state is reached well
# before the last interval (typical half-life log(2) * 152 / 479 = 0.22 day).
tau <- 0.5
n_dose <- 14L
t_last <- tau * (n_dose - 1)
obs_times <- sort(unique(c(
  seq(0, t_last, by = 0.25),
  t_last + c(0, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5, 6, 8, 10, 12) / 24
)))

doses <- cohort |>
  tidyr::crossing(time = tau * (seq_len(n_dose) - 1)) |>
  mutate(evid = 1L, amt = dose, cmt = "depot")
obs <- cohort |>
  tidyr::crossing(time = obs_times) |>
  mutate(evid = 0L, amt = 0, cmt = "central")
events <- bind_rows(doses, obs) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, evid, amt, cmt, regimen, dose, WT, ALB, CRCL, SEXF,
         CONMED_CYP2D6_INH, FED, FED_MISSING)
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("regimen", "dose")) |>
  as.data.frame()
stopifnot(!anyNA(sim$Cc))
```

## Steady-state profile

The paper's Figure 2 is a prediction-corrected VPC over all nine studies, whose
dose, food and sampling mix cannot be reconstructed, so it is not overlaid
numerically. The plot below shows the simulated steady-state dosing interval of
each regimen (individual predictions, 5th/50th/95th percentiles).

```{r ss-profile, fig.width = 7, fig.height = 4}
sim |>
  filter(time >= t_last) |>
  mutate(tad_h = (time - t_last) * 24) |>
  group_by(regimen, tad_h) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad_h, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~regimen) +
  scale_y_log10() +
  labs(
    x = "Time after dose at steady state (h)", y = "Veliparib Cc (ug/mL)",
    caption = "Simulated steady-state interval; compare the shape with Figure 2 of Stodtmann 2021."
  )
```

# PKNCA validation

Non-compartmental analysis of the last (steady-state) dosing interval, grouped
by regimen.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(id, time, Cc, regimen)

dose_nca <- events |>
  filter(evid == 1, time == t_last) |>
  select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | regimen + id)
intervals <- data.frame(
  start = t_last, end = t_last + tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res$result)

nca_tab |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Regimen" = regimen,
    "AUCtau (ug*day/mL)" = auclast,
    "Cavg (ug/mL)" = cav,
    "Cmax (ug/mL)" = cmax,
    "Cmin (ug/mL)" = cmin,
    "Tmax (day)" = tmax
  ) |>
  knitr::kable(digits = 3, caption = "Median simulated steady-state NCA by regimen.")
```

The paper reports no NCA table. Its exposure metric is `AUCss = Dose / (CL/F)`,
so the simulated per-subject steady-state AUC over one interval must equal the
dose divided by that subject's own clearance. This is a mass-balance identity on
the solved ODE; the tolerance covers only the trapezoidal error of the sampling
grid around the absorption peak.

```{r auc-identity}
auc_check <- nca_tab |>
  filter(PPTESTCD == "auclast") |>
  left_join(
    sim |> group_by(id) |> summarise(cl = first(cl), dose = first(dose), .groups = "drop"),
    by = "id"
  ) |>
  mutate(pct_diff = 100 * (PPORRES / (dose / cl) - 1))

summary(auc_check$pct_diff)
stopifnot(
  nrow(auc_check) == 3L * n_per_arm,
  abs(median(auc_check$pct_diff)) < 3,
  quantile(abs(auc_check$pct_diff), 0.9) < 5
)
```

# Phase-specific residual error

The residual error switches from the absorption-phase terms to the
elimination-phase terms 2.5 h after each dose. Each simulated observation is
standardised by the combined SD that Table 4 implies for its phase,
`sqrt(add^2 + (prop * Cc)^2)`. If the model applies the right pair of terms in
each phase, the standardised residuals have SD 1 in both phases. Applying the
absorption terms late (or the elimination terms early) moves the SD by a factor
of about 2.7.

```{r residual-phase}
res <- sim |>
  mutate(
    tad_h = 24 * (time - tau * pmin(floor(time / tau + 1e-9), n_dose - 1)),
    phase = ifelse(tad_h <= 2.5, "absorption (<= 2.5 h)", "elimination (> 2.5 h)"),
    sd_expected = ifelse(
      tad_h <= 2.5,
      sqrt(0.004^2 + (0.208 * Cc)^2),
      sqrt(2.96e-7^2 + (0.078 * Cc)^2)
    ),
    z = (sim - Cc) / sd_expected
  ) |>
  group_by(phase) |>
  summarise(n = n(), sd_z = sd(z), .groups = "drop")
knitr::kable(res, digits = 3, caption = "SD of standardised residuals by phase (expected 1).")
stopifnot(
  nrow(res) == 2L,
  all(abs(res$sd_z - 1) < 0.05)
)
```

# Assumptions and deviations

* **Categorical covariates are multiplicative factors (`theta^I`).** The
  Methods give a generic categorical equation `TVP = theta * (1 + I * theta_cat)`,
  but the final equations print `1.20^Male`, `0.885^CYP2D6`, `1.25^Male`,
  `1.11^Fasting` and `0.356^Fed`, and the Results' effect sizes (males -16.5%,
  strong CYP2D6 inhibitors +13.0% AUCss) are reproduced only by the power
  form. Under the `(1 + theta)` form males would have 55% lower AUCss, and
  food would speed absorption rather than delay it as the Discussion describes.
  The model follows the printed final equations.
* **Residual-error rows are read as standard deviations.** Table 4 does not
  say whether its RUV estimates are SDs or variances. The additive rows carry
  concentration units (ug/mL), not squared units, and the elimination-phase
  proportional RSE of 1.44% is below `sqrt(2/9160) = 1.48%`, the smallest RSE
  a variance estimated from 9160 observations can have. The bootstrap CIs of
  the RUV rows are wider than their %RSE implies, so the RSEs are not a strong
  argument by themselves.
* **Combined error form.** The paper says only "combined (additive and
  proportional)"; the nlmixr2 default (`combined2`, variances added) is used.
* **Absorption-phase boundary.** The Results place the switch "before Tmax at
  2.5 hours", and the outlier rule applies to "time since last dose of more
  than 2.5 hours", so an observation exactly 2.5 h after the dose is assigned to
  the absorption phase.
* **Food effect reference level.** The published ka (59.4 1/day) is for an
  unknown prandial state, which covers 95% of the subjects. It is encoded with
  `FED_MISSING = 1`; set `FED_MISSING = 0` and `FED = 0/1` to simulate a
  defined fasting or fed dose.
* **Creatinine clearance** is raw Cockcroft-Gault in mL/min, not
  BSA-normalised, and is capped at 120 mL/min inside the model.
* **Discussion vs Results.** The Discussion quotes a 25% AUCss increase for
  mild renal impairment where the Results (and the model) give 27.3%. The
  Figure 3 legend calls CrCL 75 and 45 mL/min "moderate" and "severe", where
  the Results call them mild and moderate. Neither affects the model.
* **Virtual cohort.** Covariates are drawn independently; correlations between
  body weight, albumin, creatinine clearance and sex were not reported. Sex and
  strong-CYP2D6-inhibitor frequencies follow Table 3 (3% male, 5% with a
  strong CYP2D6 inhibitor).
* No erratum or correction notice for this article was found on Europe PMC as
  of 2026-09-29.
