---
title: "Phenobarbital (Stoschus 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phenobarbital (Stoschus 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Stoschus_2025_phenobarbital"))$reference`
* Description: `r rxode2::rxode(readModelDb("Stoschus_2025_phenobarbital"))$description`
* Article: <https://doi.org/10.1111/epi.18517>

Stoschus 2025 developed a population pharmacokinetic model for phenobarbital in
critically ill adults treated for refractory (RSE) and superrefractory (SRSE)
status epilepticus, and used it to derive ideal-body-weight-based loading and
maintenance dosing recommendations for a 12-h dosing schedule.

The structural model is one compartment with first-order absorption and
first-order elimination. The absorption rate constant could not be estimated
because most therapeutic drug monitoring (TDM) samples were drawn after the
absorption phase, so it was fixed to a literature value. Ideal body weight (IBW)
was the only covariate retained, entering as classical allometric scaling on
both volume of distribution and clearance.

```{r model}
mod <- readModelDb("Stoschus_2025_phenobarbital")
ui <- rxode2::rxode(mod)
```

# Population

The model was fit to 301 TDM samples from 37 adults treated between 2015 and
2024 at a single German neurointensive care unit (Stoschus 2025 Table 1).
Thirty-five patients (95%) had superrefractory status epilepticus and two (5%)
refractory status epilepticus; the median Status Epilepticus Severity Score was
5. Thirteen patients (35%) were female and the median age was 63 years
(5th-95th percentile 24-83). Median total body weight was 78.5 kg and median
ideal body weight 68.8 kg -- the latter is the reference for the allometric
terms. Renal function was largely preserved (median estimated glomerular
filtration rate 102 mL/min, median serum creatinine 0.7 mg/dL), with three
patients (8%) on renal replacement therapy.

Patients received guideline-directed loading doses of 10-20 mg/kg and
maintenance doses of 1-4 mg/kg over a median treatment duration of 16 days
(median administered dose 200 mg). Of 1242 doses, 824 (66%) were oral -- tablets
dispersed in water and given via gastric tube -- and 418 (34%) were intravenous
5-min infusions. The median observed phenobarbital concentration was 32.8 mg/L,
with a median of 6 TDM samples per patient.

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its origin in
`inst/modeldb/specificDrugs/Stoschus_2025_phenobarbital.R`. The table collects
them for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lfdepot` (F) | 0.96 | Table 2, row `F` (RSE 3.95%; 95% CI 0.85-0.99) |
| `lka` (ka, fixed) | 1.9 /h | Table 2, row `k a, h-1` with footnote "ka was fixed according to literature values"; Methods 2.2 |
| `lvc` (V) | 34.3 L | Table 2, row `V, L` (RSE 12.7%; 95% CI 26.81-43.89) |
| `lcl` (CL) | 0.38 L/h | Table 2, row `CL, L/h` (RSE 7.81%; 95% CI 0.32-0.44) |
| `e_ibw_vc` (fixed) | 1 | Results 3.2, Equation 1: `V_i = V * (IBW_i / medIBW)^1` |
| `e_ibw_cl` (fixed) | 0.75 | Results 3.2, Equation 2: `CL_i = CL * (IBW_i / medIBW)^0.75` |
| reference `medIBW` | 68.8 kg | Table 1, row "Ideal body weight, kg" (cohort median) |
| `etalvc` | 0.50799 | Table 2 IIV row `V`: SD 0.71, CV 81.36%; `omega^2 = log(1 + 0.8136^2)` |
| `etalcl` | 0.157914 | Table 2 IIV row `CL`: SD 0.4, CV 41.36%; `omega^2 = log(1 + 0.4136^2)` |
| `etaiov_cl_1..8` | 0.127331 | Table 2 IOV row `CL`: SD 0.36, CV 36.85%; `omega^2 = log(1 + 0.3685^2)` |
| `propSd` | 0.082 | Table 2, row `Proportional` (RSE 7.76%; 95% CI 0.071-0.096) |
| one-compartment, first-order absorption and elimination | n/a | Methods 2.2; Results 3.2 |
| IIV on V and CL, IOV on CL, proportional RUV | n/a | Results 3.2, first paragraph |

The two structural equations are printed in the article as images; the exponents
were read from the PDF text layer as `^1` on V (Equation 1) and `^.75` on CL
(Equation 2).

## Variance scale

Table 2 reports interindividual and interoccasion variability as an SD (rounded
to two significant figures) alongside a CV% (four significant figures). Methods
2.2 states that a lognormal distribution was assumed. Round-tripping each row
through the lognormal relation confirms the SD column is the omega on the log
scale, so the variances used here are back-solved from the more precise CV%.

```{r variance-scale}
tibble::tibble(
  parameter = c("IIV V", "IIV CL", "IOV CL"),
  cv_pct    = c(81.36, 41.36, 36.85),
  sd_printed = c(0.71, 0.40, 0.36)
) |>
  mutate(
    omega2       = log(1 + (cv_pct / 100)^2),
    sd_backsolved = sqrt(omega2),
    cv_roundtrip = 100 * sqrt(exp(sd_printed^2) - 1)
  ) |>
  knitr::kable(digits = 4, caption =
    "The printed SD reproduces the printed CV% and vice versa, confirming the lognormal log-scale reading.")
```

# Typical-value checks

Two quantities are fixed by the structural parameters and can be checked in
closed form before any simulation.

The paper states an elimination half-life of 62.6 h, "derived from our estimated
population-specific parameters" (Methods 2.3), which it used to justify
assessing steady-state trough concentrations at t = 336 h.

```{r half-life}
V  <- 34.3
CL <- 0.38
c(published_h = 62.6, closed_form_h = log(2) * V / CL)
```

Results 3.2 also reports the weight-normalised typical values obtained when
dividing by the cohort median *total* body weight of 78.5 kg.

```{r normalised}
c(V_L_per_kg_paper = 0.44, V_L_per_kg_model = V / 78.5,
  CL_L_h_kg_paper = 0.0048, CL_L_h_kg_model = CL / 78.5)
```

# Virtual cohort and simulation

Original patient data are not public. Simulations below use virtual cohorts
whose ideal body weights are the discrete values the paper simulated (55-80 kg
in 5-kg steps), all inside the observed cohort range of 52.4-82.4 kg.

Occasions follow the paper's definition -- the interval between consecutive TDM
samples, cohort median 48 h -- so the `OCC` column advances every 48 h and is
capped at the eight occasions the model file encodes.

```{r helpers}
occ_of <- function(time) pmin(floor(time / 48) + 1, 8)

# Build one 12-h-schedule regimen: loading dose at t = 0, maintenance doses
# every 12 h thereafter, observed on `obs_times`.
regimen <- function(ld, md, route, ibw, obs_times, last_dose = 324) {
  cmt <- if (identical(route, "IV")) "central" else "depot"
  ev <- rxode2::et(amt = ld, cmt = cmt, time = 0)
  if (md > 0) {
    for (tt in seq(12, last_dose, by = 12)) {
      ev <- rxode2::et(ev, amt = md, cmt = cmt, time = tt)
    }
  }
  ev <- rxode2::et(ev, obs_times, cmt = "central")
  d <- as.data.frame(ev)
  d$IBW <- ibw
  d$OCC <- occ_of(d$time)
  d
}

# Solve a cohort. The seed is set immediately before every rxSolve so that
# runs sharing a seed and an event-table shape draw the *same* subjects --
# common random numbers, which is what makes dose grids comparable.
solve_cohort <- function(events, n, seed) {
  set.seed(seed)
  as.data.frame(rxode2::rxSolve(ui, events, nSub = n, returnType = "data.frame"))
}
```

## Instrument check: are the published omegas actually realised?

Before using the cohort to judge the model, confirm the simulated cohort
reproduces the parameters that were put into it. A validation built on a
mis-drawn cohort measures nothing.

```{r instrument-check}
chk <- solve_cohort(regimen(1000, 0, "IV", 68.8, 12), n = 2000L, seed = 11L) |>
  filter(time == 12)

tibble::tibble(
  quantity = c("median V (L)", "median CL (L/h)", "sd(log V)", "sd(log CL)"),
  simulated = c(median(chk$vc), median(chk$cl), sd(log(chk$vc)), sd(log(chk$cl))),
  expected  = c(34.3, 0.38, sqrt(0.50799), sqrt(0.157914 + 0.127331))
) |>
  mutate(pct_diff = 100 * (simulated - expected) / expected) |>
  knitr::kable(digits = 4, caption =
    "Simulated cohort reproduces the published typical values and the IIV (V) and IIV+IOV (CL) log-scale spreads.")
```

The clearance spread combines the interindividual and interoccasion terms,
because at any single time point a subject carries one IIV draw and one
occasion's IOV draw.

# Replicating Figure 2

Figure 2 of Stoschus 2025 shows concentration-time profiles at IBW 70 kg for a
12-h schedule with the optimal loading dose of 1100 mg and maintenance dose of
150 mg, across the four combinations of loading and maintenance route, with the
median, the 50% prediction interval, and the 18-40 mg/L target band.

```{r figure-2}
obs_grid <- sort(unique(c(seq(0, 336, by = 2), 12, 336)))

fig2 <- bind_rows(lapply(
  list(c("IV", "IV"), c("IV", "PO"), c("PO", "IV"), c("PO", "PO")),
  function(rr) {
    ld_route <- rr[1]; md_route <- rr[2]
    # Loading and maintenance routes can differ, so build the two dose blocks
    # separately and bind them.
    cmt_ld <- if (ld_route == "IV") "central" else "depot"
    cmt_md <- if (md_route == "IV") "central" else "depot"
    ev <- rxode2::et(amt = 1100, cmt = cmt_ld, time = 0)
    for (tt in seq(12, 324, by = 12)) ev <- rxode2::et(ev, amt = 150, cmt = cmt_md, time = tt)
    ev <- rxode2::et(ev, obs_grid, cmt = "central")
    d <- as.data.frame(ev); d$IBW <- 70; d$OCC <- occ_of(d$time)
    solve_cohort(d, n = 200L, seed = 303L) |>
      mutate(arm = paste0(ld_route, " load / ", md_route, " maint."))
  }
))

fig2 |>
  group_by(arm, time) |>
  summarise(Q25 = quantile(Cc, 0.25), Q50 = median(Cc), Q75 = quantile(Cc, 0.75),
            .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q25, ymax = Q75), alpha = 0.3, fill = "steelblue") +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = c(18, 40), linetype = "dotted", colour = "red") +
  facet_wrap(~arm) +
  labs(x = "Time (h)", y = "Phenobarbital concentration (mg/L)",
       caption = paste("Replicates Figure 2 of Stoschus 2025: IBW 70 kg, loading dose 1100 mg,",
                       "maintenance dose 150 mg every 12 h. Line = median,",
                       "band = 50% prediction interval, dotted = 18-40 mg/L target range."))
```

As the paper reports, the intravenous and oral profiles are nearly
indistinguishable -- a direct consequence of the 96% oral bioavailability.

```{r figure-2-route-difference}
fig2 |>
  filter(time == 336) |>
  group_by(arm) |>
  summarise(median_Cmin = median(Cc), .groups = "drop") |>
  knitr::kable(digits = 2, caption =
    "Trough at t = 336 h by route combination; oral and intravenous agree closely.")
```

# PKNCA validation

NCA is run on a typical-value single-dose profile, where each parameter has an
exact closed-form counterpart. Using `zeroRe()` matters here: with an 81% CV on
volume the population *mean* profile is not the typical-value profile, so a
cohort median would not be comparable to a closed-form identity.

```{r pknca}
ui_typ <- rxode2::zeroRe(ui)

nca_events <- regimen(1000, 0, "IV", 68.8, obs_times = sort(unique(
  c(seq(0, 24, by = 0.25), seq(24, 720, by = 2))
)))
set.seed(404L)
sim_nca_raw <- as.data.frame(
  rxode2::rxSolve(ui_typ, nca_events, returnType = "data.frame")
)
if (is.null(sim_nca_raw$id)) sim_nca_raw$id <- 1L

sim_nca <- sim_nca_raw |>
  filter(!is.na(Cc)) |>
  mutate(treatment = "1000 mg IV, typical value") |>
  select(id, time, Cc, treatment)

# Guarantee a time-zero anchor for AUC.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- data.frame(id = 1L, time = 0, amt = 1000,
                      treatment = "1000 mg IV, typical value")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))
```

## Comparison against published and closed-form values

The paper reports one NCA-comparable quantity directly, the 62.6 h elimination
half-life. The remaining rows are exact closed-form identities for a
one-compartment intravenous bolus at the reference IBW, which gate the NCA
machinery as well as the model: an error in the dosing route, the observation
grid, or the PKNCA setup breaks the AUC identity immediately.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,                  ~cmax,   ~tmax, ~aucinf.obs, ~half.life,
  "1000 mg IV, typical value", 1000 / V, 0,     1000 / CL,   62.6
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "mg/L", aucinf.obs = "mg*h/L", tmax = "h", half.life = "h"),
  tolerance_pct = 20
) |>
  knitr::kable(caption = paste(
    "Simulated vs reference NCA. half.life is the published 62.6 h;",
    "cmax = dose/V, aucinf.obs = dose/CL are closed-form identities.",
    "* marks a >20% difference."))
```

The AUC identity `AUC(0-inf) = dose / CL` is the strongest single check in this
vignette: it simultaneously gates the ODE, the dose encoding, the observation
window, and PKNCA's settings.

# Replicating Table 3: optimal 12-h dosing

Table 3 of Stoschus 2025 gives, for each IBW from 55 to 80 kg, the loading and
maintenance doses maximising the probability that the trough concentration falls
in 18-40 mg/L. Loading doses were selected on the trough at t = 12 h and
maintenance doses on the trough at t = 336 h, with the optimal loading dose
applied first.

Because the model is linear in dose and the loading-dose trough involves a
single dose, the whole 100-2500 mg loading grid can be evaluated exactly from
one unit-dose solve per IBW and route -- the same subjects are used at every
dose, which removes the Monte Carlo noise that would otherwise dominate the
comparison between adjacent doses.

```{r linearity-check}
u <- solve_cohort(regimen(1, 0, "IV", 70, 12), n = 200L, seed = 505L) |> filter(time == 12)
m <- solve_cohort(regimen(1100, 0, "IV", 70, 12), n = 200L, seed = 505L) |> filter(time == 12)
c(max_relative_error = max(abs(1100 * u$Cc - m$Cc) / m$Cc))
```

Dose-proportionality holds to machine precision, so the scaling is exact rather
than an approximation.

```{r table3-loading}
IBWS <- seq(55, 80, by = 5)
in_target <- function(x) mean(x >= 18 & x <= 40)
load_grid <- seq(100, 2500, by = 100)

loading_published <- tibble::tibble(
  IBW = rep(IBWS, each = 2),
  route = rep(c("IV", "PO"), times = 6),
  dose_paper = c(1000, 900, 1000, 1000, 1000, 1000, 1100, 1100, 1300, 1300, 1300, 1300)
)

loading <- bind_rows(lapply(IBWS, function(ib) {
  bind_rows(lapply(c("IV", "PO"), function(rt) {
    unit <- solve_cohort(regimen(1, 0, rt, ib, 12), n = 200L, seed = 606L) |>
      filter(time == 12)
    p <- vapply(load_grid, function(D) in_target(D * unit$Cc), numeric(1))
    dp <- loading_published$dose_paper[loading_published$IBW == ib &
                                         loading_published$route == rt]
    tibble::tibble(IBW = ib, route = rt,
                   dose_model = load_grid[which.max(p)], pta = max(p),
                   dose_paper = dp, pta_at_paper = in_target(dp * unit$Cc))
  }))
}))

loading |>
  mutate(`PTA at model optimum (%)` = round(100 * pta, 1),
         `PTA at Table 3 dose (%)`  = round(100 * pta_at_paper, 1),
         `PTA loss (pp)` = round(100 * (pta - pta_at_paper), 1)) |>
  select(IBW, route, `Model optimum (mg)` = dose_model,
         `Table 3 (mg)` = dose_paper,
         `PTA at model optimum (%)`, `PTA at Table 3 dose (%)`, `PTA loss (pp)`) |>
  knitr::kable(caption = paste(
    "Loading doses: model optimum vs Stoschus 2025 Table 3. The last column is a",
    "paired comparison on identical subjects, so it isolates the dose choice from",
    "Monte Carlo noise."))
```

Maintenance doses need the full multiple-dose regimen, so each dose is simulated
directly. The seed is reset before every solve and every event table has the
same shape, so all doses within an IBW-route combination are evaluated on the
same virtual subjects.

```{r table3-maintenance}
maint_grid <- seq(50, 300, by = 50)
ld_paper <- list(IV = c(1000, 1000, 1000, 1100, 1300, 1300),
                 PO = c(900, 1000, 1000, 1100, 1300, 1300))

maintenance <- bind_rows(lapply(seq_along(IBWS), function(i) {
  bind_rows(lapply(c("IV", "PO"), function(rt) {
    p <- vapply(maint_grid, function(MD) {
      s <- solve_cohort(regimen(ld_paper[[rt]][i], MD, rt, IBWS[i], 336),
                        n = 200L, seed = 707L)
      in_target(s$Cc[s$time == 336])
    }, numeric(1))
    dp <- if (IBWS[i] == 55) 100 else 150
    tibble::tibble(IBW = IBWS[i], route = rt,
                   dose_model = maint_grid[which.max(p)], pta = max(p),
                   dose_paper = dp, pta_at_paper = p[match(dp, maint_grid)])
  }))
}))

maintenance |>
  mutate(`PTA at model optimum (%)` = round(100 * pta, 1),
         `PTA at Table 3 dose (%)`  = round(100 * pta_at_paper, 1),
         `PTA loss (pp)` = round(100 * (pta - pta_at_paper), 1)) |>
  select(IBW, route, `Model optimum (mg)` = dose_model,
         `Table 3 (mg)` = dose_paper,
         `PTA at model optimum (%)`, `PTA at Table 3 dose (%)`, `PTA loss (pp)`) |>
  knitr::kable(caption = paste(
    "Maintenance doses: model optimum vs Stoschus 2025 Table 3, with the same",
    "paired comparison on identical subjects."))
```

## Probability of target attainment

```{r pta-summary}
tibble::tibble(
  quantity = c("Loading, IV", "Loading, oral", "Maintenance, IV", "Maintenance, oral"),
  model_pct = 100 * c(
    mean(loading$pta[loading$route == "IV"]),
    mean(loading$pta[loading$route == "PO"]),
    mean(maintenance$pta[maintenance$route == "IV"]),
    mean(maintenance$pta[maintenance$route == "PO"])
  ),
  published_pct = c(42.7, 42.5, 44.3, 44.5)
) |>
  mutate(difference_pp = model_pct - published_pct) |>
  knitr::kable(digits = 1, caption =
    "Target attainment at each optimal dose, averaged over IBW 55-80 kg (Stoschus 2025 Results 3.3).")
```

The maintenance optimum reproduces Table 3 exactly at the five ideal body weights
where the paper recommends 150 mg, and selects 150 mg rather than 100 mg at
55 kg. The loading optima come out systematically one to two 100-mg grid steps
*below* Table 3.

That last difference should be read against the `PTA loss` column rather than
against the dose itself. Because the attainment surface is very flat near its
peak, giving the paper's Table 3 dose instead of this model's optimum costs only
a few percentage points of target attainment on identical simulated subjects,
and the argmax moves between adjacent grid points when the random seed changes.
The grid point that wins is therefore not a stable quantity at 200 subjects per
arm, whereas the paired attainment loss is. The practical conclusion is that the
published doses and the doses this model prefers are close to equivalent, not
that one is right and the other wrong.

The absolute attainment percentages, however, come out higher in this
reproduction than the published values, and the gap is larger for the
maintenance doses than the loading doses. This is discussed under Assumptions
and deviations below; no parameter was adjusted to close it.

# Assumptions and deviations

* **Number of occasions.** The paper defines an occasion as the interval between
  consecutive TDM samples (cohort median 48 h) and reports a single IOV
  magnitude, but never states an occasion count. The model file encodes eight
  48-h occasions spanning 0-384 h, chosen to cover the 336 h at which the paper
  assessed steady-state troughs. Records with `OCC` outside 1-8 carry no IOV.
  Because a single IOV variance is shared, occasions 2-8 are fixed equal to
  occasion 1. The marginal clearance variability at any one time point does not
  depend on this choice; only the correlation of clearance over time does.

* **Target attainment is reproduced higher than published.** Averaged over IBW
  55-80 kg the model gives about 50% attainment for loading doses (paper: 42.7%
  intravenous, 42.5% oral) and about 65% for maintenance doses (paper: 44.3% and
  44.5%). The instrument check above confirms the simulated cohort
  carries exactly the published typical values and log-scale spreads, and the
  closed-form half-life, weight-normalised parameters, and NCA identities all
  agree, so the discrepancy is not a mis-encoded parameter. The most likely
  explanation is a difference in the simulation setup rather than in the model:
  the paper simulated in C++ via `mrgsolve` with 100 000 (loading) and 10 000
  (maintenance) replicates per covariate-dose combination, and Appendix S1 --
  which documents that simulation code and is not available on disk for this
  extraction -- may specify additional variability, a different occasion
  structure, or a different trough definition. Notably, the published attainment
  of about 42-43% is what a lognormal trough distribution with a log-scale SD of
  0.71 would give, which is the interindividual SD on volume alone; the trough
  spread realised here is smaller because volume partly cancels between the
  `dose/V` and `exp(-CL/V * t)` terms. No parameter was tuned to reconcile the
  numbers.

* **Ideal body weight formula.** Methods 2.1 states IBW was computed with the
  formula of Brower et al. (paper reference 25) but does not reproduce the
  formula, and that reference is not on disk. A user deriving IBW from height
  and sex must consult that reference; the model takes `IBW` as a supplied
  covariate column and does not derive it.

* **Appendix S1 not available.** The supplement holds the CV definition
  (Equation 1), the covariate screening list (Table S1), the OFV table
  (Table S2), and the diagnostic and 24-h-schedule figures (Figures S1-S3). No
  final parameter value depends on it: all fixed effects, IIV, IOV, and residual
  error are in main-text Table 2, and both covariate equations are in the main
  text. The lognormal CV relation was confirmed by round-tripping Table 2's own
  SD and CV% columns against each other rather than by reading Equation 1.

* **Covariates screened but not retained.** Age, sex, total body weight, height,
  renal replacement therapy, serum creatinine, bilirubin, albumin, ALT, AST, and
  concomitant valproate, phenytoin, metamizole, and cenobamate were tested and
  excluded (Methods 2.2; Results 3.2). They are recorded in the model file's
  `covariatesDataExcluded` metadata rather than `covariateData`, since none is
  referenced in `model()`. The Discussion cautions that clearance effects of
  concomitant valproate and phenytoin reported elsewhere "cannot be ruled out
  and might be included in the estimated CL parameter".

* **Dosing routes.** Bioavailability applies to the `depot` compartment only.
  Intravenous doses are given directly to `central`, which is what makes F
  identifiable in this cohort; the paper's 5-min infusions are modelled as
  boluses, negligible against a 62.6 h half-life.

* **Extrapolation limits.** The paper restricts its dosing recommendations to
  IBW 55-80 kg within an observed range of 52.4-82.4 kg, and states the model is
  for adults only and "not intended for pediatric use". Simulations here stay
  inside that range.
