---
title: "Fluconazole (Stott 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fluconazole (Stott 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod      <- readModelDb("Stott_2018_fluconazole")
mod_meta <- rxode2::rxode(mod)$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (DOI): <https://doi.org/10.1128/AAC.00885-18>

This vignette validates the packaged `Stott_2018_fluconazole` model -- the
four-compartment (gut, central, CSF/CNS, peripheral) population PK model that
Stott and colleagues fitted to 508 plasma and 167 CSF fluconazole
concentrations from 43 Vietnamese and Ugandan adults with HIV-associated
cryptococcal meningitis, using the non-parametric adaptive grid algorithm in
Pmetrics 1.5.0.

It is the earlier companion of `modellib("Stott_2023_flucytosine")`: same group,
same disease, same four-compartment gut / central / CSF / peripheral structure
and the same asymmetric first-order transfer parameterisation.

Three quantities are exact algebraic consequences of the Table 3 parameter set
and are the primary gates below:

* dose-proportionality of exposure, the signature of a linear model;
* a steady-state CSF:plasma concentration ratio of `Kcs/Ksc * Vc/Vcns`;
* a terminal disposition rate of `SCL / Vss` with
  `Vss = Vc * (1 + Kcs/Ksc + Kcp/Kpc)`.

A fourth section compares the model against the exposures the paper itself
reports. **The plasma exposures the paper derived from its 43 patients
reproduce closely; the exposures and target-attainment percentages it derived
from its 5000-draw Monte Carlo do not, and cannot be recovered from Table 3
under any reading.** That is analysed, not tuned, in *Assumptions and
deviations*.

# Population

43 patients were recruited over 11 months between January and November 2016 at
the Hospital for Tropical Diseases, Ho Chi Minh City, Vietnam (n = 23) and
Masaka General Hospital, Uganda (n = 20): 3 from a multicentre randomised
controlled trial of adjuvant dexamethasone (ISRCTN 59144167) and, after that
trial stopped early, 40 from a prospective descriptive study at the same sites.

Median age was 33 years (range 20-73), median weight 48 kg (range 32-68) and
median BMI 18 kg/m^2 (range 12-25) -- a severely wasted cohort. Median baseline
creatinine was 70 umol/L (range 37-167) and median Cockcroft-Gault eGFR
84.8 mL/min/1.73 m^2 (range 35.4-146.7).

Fluconazole was given orally, or by nasogastric tube where conscious level
precluded swallowing. The majority received 800 mg q24h; two patients received
one-off 400 mg doses, two received one-off 600 mg doses, and one patient's
800 mg q24h regimen was escalated to 1200 mg q24h for six days from day 8. All
patients also received amphotericin B deoxycholate 1 mg/kg q24h.

```{r population}
str(mod_meta$population)
```

No covariate is retained in the final model. Weight, BMI, sex, ethnicity
(equivalently, enrollment site) and baseline eGFR were screened by bidirectional
stepwise multivariate regression against the Bayesian posterior estimates of
volume and clearance. Only weight reached the screening threshold against
volume, and only marginally (slope 0.22, 95% CI -0.06 to 0.51, P = 0.05); the
weight-scaled variant (Table 2, "Model 2") gave comparable log likelihood, AIC,
bias and imprecision, so the covariate-free base model was retained. Those
screened-but-dropped covariates are recorded in the model file's
`covariatesDataExcluded` metadata so the provenance of the screen survives
without creating unused `covariateData` entries.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Stott_2018_fluconazole.R`. The table below
collects them in one place. Every structural value is the **Mean** column of
Stott 2018 Table 3; every IIV variance is `log(1 + (SD/Mean)^2)` from the
**Mean** and **SD** columns of the same table.

| Parameter / equation                           | Value      | Source location                                                                 |
|------------------------------------------------|------------|---------------------------------------------------------------------------------|
| `lka` (Ka, gut to central)                     | log(8.78)  | Table 3 row "Ka (h-1)", Mean = 8.78                                             |
| `lcl` (SCL/F)                                  | log(0.72)  | Table 3 row "SCL/F (liters/h)", Mean = 0.72                                     |
| `lvc` (Vc/F)                                   | log(18.07) | Table 3 row "Vc/F (liters)", Mean = 18.07                                       |
| `lk_central_csf` (paper's Kcs)                 | log(35.43) | Table 3 row "Kcs (h-1)", Mean = 35.43                                           |
| `lk_csf_central` (paper's Ksc)                 | log(28.63) | Table 3 row "Ksc (h-1)", Mean = 28.63                                           |
| `lk12` (paper's Kcp)                           | log(12.20) | Table 3 row "Kcp (h-1)", Mean = 12.20                                           |
| `lk21` (paper's Kpc)                           | log(18.10) | Table 3 row "Kpc (h-1)", Mean = 18.10                                           |
| `lvcsf` (Vcns/F)                               | log(32.07) | Table 3 row "Vcns/F (liters)", Mean = 32.07                                     |
| `etalka`                                       | 1.05144    | Table 3: SD 11.98 / Mean 8.78 -> CV 1.36446 -> log(1 + CV^2)                    |
| `etalcl`                                       | 0.10536    | Table 3: SD 0.24 / Mean 0.72 -> CV 0.33333 -> log(1 + CV^2)                     |
| `etalvc`                                       | 0.11506    | Table 3: SD 6.31 / Mean 18.07 -> CV 0.34920 -> log(1 + CV^2)                    |
| `etalk_central_csf`                            | 0.14010    | Table 3: SD 13.74 / Mean 35.43 -> CV 0.38781 -> log(1 + CV^2)                   |
| `etalk_csf_central`                            | 0.11577    | Table 3: SD 10.03 / Mean 28.63 -> CV 0.35033 -> log(1 + CV^2)                   |
| `etalk12`                                      | 0.60883    | Table 3: SD 11.17 / Mean 12.20 -> CV 0.91557 -> log(1 + CV^2)                   |
| `etalk21`                                      | 0.18876    | Table 3: SD 8.25 / Mean 18.10 -> CV 0.45580 -> log(1 + CV^2)                    |
| `etalvcsf`                                     | 0.26327    | Table 3: SD 17.60 / Mean 32.07 -> CV 0.54880 -> log(1 + CV^2)                   |
| `d/dt(depot)`                                  | n/a        | Equation 1: dX(1)/dt = -Ka * X(1)                                               |
| `d/dt(central)`                                | n/a        | Equation 2: dX(2)/dt = Ka*X(1) - (Kcp + Kcs + SCL/V)*X(2) + Ksc*X(3) + Kpc*X(4) |
| `d/dt(csf)`                                    | n/a        | Equation 3: dX(3)/dt = Kcs*X(2) - Ksc*X(3)                                      |
| `d/dt(peripheral1)`                            | n/a        | Equation 4: dX(4)/dt = Kcp*X(2) - Kpc*X(4)                                      |
| `Cc <- central / vc`                           | n/a        | Equation 5: Y(1) = X(2)/V                                                       |
| `Ccsf <- csf / vcsf`                           | n/a        | Equation 6: Y(2) = X(3)/Vcns                                                    |
| `propSd`, `addSd`, `propSd_Ccsf`, `addSd_Ccsf` | fixed(0)   | Methods: additive lambda plus assay-variance weighting declared, no estimate published |
| initial conditions ICgut / ICcentral / ICCNS / ICperipheral | not encoded | Table 3 and Methods; the source zeroes them for every simulation it reports (see *Assumptions and deviations*) |

# Virtual cohort

The individual concentrations are not public, so the vignette uses virtual
cohorts at the four dosages of the paper's own Monte Carlo analysis: 400, 800,
1200 and 2000 mg q24h. No model parameter scales with body weight, so weight
enters nowhere and no weight distribution is needed.

Dosing runs q24h from 0 to 144 h and NCA is taken over the paper's window of
144-168 h after treatment initiation, on the 0.1 h grid used for its own
trapezoidal AUC. Initial conditions are zero, matching the source's simulation
setup ("for the simulations, the initial conditions of all compartments were
defaulted to zero").

```{r cohort}
rxode2::rxSetSeed(20260919)

n_per_arm <- 200L # per-arm cap for validation vignettes
t_end     <- 168  # h
ss_start  <- 144  # h; the paper's AUC window is 144-168 h
grid_h    <- 0.1  # paper: trapezoidal approximation

doses <- c(400, 800, 1200, 2000)

# Observation rows are anchored on the ODE state `central`, never on the
# algebraic observable names Cc / Ccsf: referencing an observable in `cmt`
# makes rxUi inject a compartment slot after the ODE states and renumber them.
# This model has two endpoints, so observation rows additionally carry
# `dvid = 1`. rxSolve returns both Cc and Ccsf as columns on every observation
# row, so one anchor per time point covers both matrices.
make_arm <- function(dose_mg, n, id_offset, obs_times) {
  ev  <- rxode2::et(amt = dose_mg, time = seq(0, ss_start, by = 24), cmt = "depot")
  ev  <- rxode2::et(ev, obs_times, cmt = "central")
  one <- as.data.frame(ev)
  one$dvid <- ifelse(one$evid == 0L, 1L, NA_integer_)

  out <- do.call(rbind, lapply(seq_len(n), function(i) {
    tmp    <- one
    tmp$id <- id_offset + i
    tmp
  }))
  out$dose_mg    <- dose_mg
  out$dose_label <- paste0(dose_mg, " mg q24h")
  out[
    order(out$id, out$time, -out$evid),
    c("id", setdiff(names(out), "id"))
  ]
}

obs_ss <- seq(ss_start, t_end, by = grid_h)
events_ss <- dplyr::bind_rows(lapply(seq_along(doses), function(i) {
  make_arm(doses[i], n_per_arm, (i - 1L) * n_per_arm, obs_ss)
}))

# Full 0-168 h profile on a coarser grid, 800 mg arm only, for the Figure 1
# replication.
events_vpc <- make_arm(800, n_per_arm, 10000L, seq(0, t_end, by = 0.5))

stopifnot(
  !anyDuplicated(events_ss[, c("id", "time", "evid")]),
  !anyDuplicated(events_vpc[, c("id", "time", "evid")])
)
```

# Simulation

Two solves are used. The **typical-value** solve (`rxode2::zeroRe()`) carries the
Table 3 mean parameter set with IIV switched off and is what the deterministic
gates are asserted against. The **cohort** solve carries the diagonal log-normal
IIV and is what the published exposure distributions are compared against.

`useLinCmt = FALSE` is required: rxode2's automatic ODE-to-`linCmt()` conversion
corrupts the observable-to-state mapping for models like this one that carry two
observables reading from two different ODE states.

```{r simulate}
mod_typical <- rxode2::zeroRe(mod)

sim_typ_ss <- rxode2::rxSolve(
  mod_typical, events_ss,
  keep = c("dose_mg", "dose_label"), useLinCmt = FALSE
) |>
  as.data.frame()

sim_cohort_ss <- rxode2::rxSolve(
  mod, events_ss,
  keep = c("dose_mg", "dose_label"), useLinCmt = FALSE
) |>
  as.data.frame()

sim_cohort_vpc <- rxode2::rxSolve(
  mod, events_vpc,
  keep = c("dose_mg", "dose_label"), useLinCmt = FALSE
) |>
  as.data.frame()

stopifnot(
  all(sim_typ_ss$Cc[!is.na(sim_typ_ss$Cc)] >= 0),
  all(sim_cohort_ss$Ccsf[!is.na(sim_cohort_ss$Ccsf)] >= 0)
)
```

# Structural identity checks

These are deterministic functions of the transcribed Table 3 values. Both sides
use the same parameters, so the only difference is numerical integration error
and a tight bound is the correct gate.

```{r identities}
trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

auc_typ <- sim_typ_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::arrange(dose_mg, time) |>
  dplyr::group_by(dose_mg, dose_label) |>
  dplyr::summarise(
    auc_plasma = trapz(time, Cc),
    auc_csf    = trapz(time, Ccsf),
    .groups    = "drop"
  ) |>
  dplyr::mutate(
    ratio            = auc_csf / auc_plasma,
    auc_per_100mg    = 100 * auc_plasma / dose_mg
  )

# (1) Dose proportionality: AUC / dose is constant for a linear model.
dose_prop_spread <-
  diff(range(auc_typ$auc_per_100mg)) / mean(auc_typ$auc_per_100mg)

# (2) Steady-state CSF:plasma concentration ratio is Kcs/Ksc * Vc/Vcns.
ratio_closed_form <- (35.43 / 28.63) * (18.07 / 32.07)

knitr::kable(
  auc_typ |>
    dplyr::select(dose_label, auc_plasma, auc_csf, ratio, auc_per_100mg) |>
    dplyr::rename(
      "Regimen"                    = dose_label,
      "Plasma AUC144-168 (mg*h/L)" = auc_plasma,
      "CSF AUC144-168 (mg*h/L)"    = auc_csf,
      "CSF:plasma"                 = ratio,
      "Plasma AUC per 100 mg"      = auc_per_100mg
    ),
  digits  = 3,
  caption = paste(
    "Typical-value exposure at the four dosages of Stott 2018 Figure 3.",
    "Exposure is exactly dose-proportional and the CSF:plasma ratio is",
    "dose-invariant, as a linear model requires."
  )
)

cat(sprintf(
  "Dose-proportionality spread across 400 / 800 / 1200 / 2000 mg: %.4f%%\n",
  100 * dose_prop_spread
))
cat(sprintf(
  "CSF:plasma AUC ratio (typical): %.5f; Kcs/Ksc * Vc/Vcns = %.5f\n",
  auc_typ$ratio[1], ratio_closed_form
))

stopifnot(
  # Linear PK: identical exposure per mg across a 5-fold dose range.
  dose_prop_spread < 1e-6,
  # Exact algebraic identity of the ODE system, at every dose.
  all(abs(auc_typ$ratio - ratio_closed_form) < 1e-4)
)
```

The terminal disposition rate is a second exact identity. With no elimination
from either the CSF or the peripheral compartment, the pseudo-equilibrium amount
ratios are `Kcs/Ksc` and `Kcp/Kpc`, so the apparent steady-state volume is
`Vc * (1 + Kcs/Ksc + Kcp/Kpc)` and the terminal rate is `SCL / Vss`.

```{r half-life-identity}
vss_closed_form <- 18.07 * (1 + 35.43 / 28.63 + 12.20 / 18.10)
thalf_closed_form <- log(2) * vss_closed_form / 0.72

ev_single <- rxode2::et(amt = 800, time = 0, cmt = "depot")
ev_single <- rxode2::et(ev_single, seq(0, 504, by = 0.5), cmt = "central")
ev_single <- as.data.frame(ev_single)
ev_single$dvid       <- ifelse(ev_single$evid == 0L, 1L, NA_integer_)
ev_single$id         <- 1L
ev_single$dose_label <- "800 mg single dose"

sim_single <- rxode2::rxSolve(
  mod_typical, ev_single, keep = "dose_label", useLinCmt = FALSE
) |>
  as.data.frame()

# rxSolve() drops the `id` column entirely for a single-subject event table.
if (is.null(sim_single$id)) sim_single$id <- 1L

tail_phase <- sim_single |>
  dplyr::filter(time >= 336, !is.na(Cc), Cc > 0)
slope <- stats::coef(stats::lm(log(Cc) ~ time, data = tail_phase))[["time"]]
thalf_solved <- log(2) / -slope

cat(sprintf("Apparent Vss = Vc * (1 + Kcs/Ksc + Kcp/Kpc) = %.3f L\n", vss_closed_form))
cat(sprintf(
  "Terminal half-life: solved %.3f h vs closed form log(2) * Vss / SCL = %.3f h\n",
  thalf_solved, thalf_closed_form
))

stopifnot(abs(thalf_solved - thalf_closed_form) / thalf_closed_form < 0.01)
```

Stott 2018 does not report a half-life, so this is an internal-consistency gate
rather than a comparison against the publication: a mis-transcribed clearance,
volume or transfer constant moves it by hours. The roughly 51 h terminal
half-life is longer than fluconazole's commonly quoted ~30 h because this model
places 66% of the steady-state drug mass outside the central compartment.

# Replicate published figures

## Figure 1 -- plasma and CSF concentration-time profiles

Figure 1 of the paper plots the raw plasma and CSF concentrations from all 43
patients on 800 mg q24h. The individual observations are not public; the
simulated envelope at the same regimen is shown below.

```{r figure-1-profiles}
vpc <- sim_cohort_vpc |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, Ccsf) |>
  tidyr::pivot_longer(c(Cc, Ccsf), names_to = "matrix", values_to = "conc") |>
  dplyr::mutate(matrix = dplyr::recode(matrix, Cc = "Plasma", Ccsf = "CSF")) |>
  dplyr::group_by(matrix, time) |>
  dplyr::summarise(
    p05 = stats::quantile(conc, 0.05, na.rm = TRUE),
    p50 = stats::median(conc, na.rm = TRUE),
    p95 = stats::quantile(conc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc, aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), fill = "grey75", alpha = 0.6) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~matrix) +
  labs(
    x        = "Time (hours)",
    y        = "Fluconazole concentration (mg/L)",
    title    = "Replicates Figure 1 of Stott 2018: simulated concentration envelope",
    subtitle = "800 mg q24h; median with 5th-95th percentile band, 200 subjects"
  ) +
  theme_minimal()
```

Accumulation is still visibly incomplete at 168 h, which is the direct
consequence of the ~51 h terminal half-life derived above and matters for the
comparisons that follow: the paper's 144-168 h window sits at about 90% of
steady state, not at steady state.

## Typical-value profile over the final dosing interval

```{r figure-profile}
prof <- sim_typ_ss |>
  dplyr::filter(dose_mg == 800, !is.na(Cc)) |>
  dplyr::mutate(time_post_dose = time - ss_start) |>
  dplyr::select(time_post_dose, Cc, Ccsf) |>
  tidyr::pivot_longer(c(Cc, Ccsf), names_to = "matrix", values_to = "conc") |>
  dplyr::mutate(matrix = dplyr::recode(matrix, Cc = "Plasma", Ccsf = "CSF"))

ggplot(prof, aes(time_post_dose, conc, colour = matrix)) +
  geom_line(linewidth = 0.7) +
  labs(
    x        = "Time after the dose at 144 h (hours)",
    y        = "Concentration (mg/L)",
    colour   = "Matrix",
    title    = "Typical-value profile over the paper's 144-168 h AUC window",
    subtitle = "CSF tracks plasma at the model's 0.70 ratio, damped by the CSF leg"
  ) +
  theme_minimal()
```

## Figure 3 -- simulated AUC144-168 distributions by dosage

```{r figure-3-auc}
auc_cohort <- sim_cohort_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::arrange(id, time) |>
  dplyr::group_by(id, dose_mg, dose_label) |>
  dplyr::summarise(
    auc_plasma  = trapz(time, Cc),
    auc_csf     = trapz(time, Ccsf),
    cmax_plasma = max(Cc),
    cmin_plasma = min(Cc),
    cmax_csf    = max(Ccsf),
    cmin_csf    = min(Ccsf),
    .groups     = "drop"
  ) |>
  dplyr::mutate(ratio = auc_csf / auc_plasma)

auc_long <- auc_cohort |>
  dplyr::select(id, dose_label, auc_plasma, auc_csf) |>
  tidyr::pivot_longer(c(auc_plasma, auc_csf),
    names_to = "matrix", values_to = "auc"
  ) |>
  dplyr::mutate(matrix = dplyr::recode(matrix,
    auc_plasma = "Plasma",
    auc_csf    = "CSF"
  ))

ggplot(auc_long, aes(stats::reorder(dose_label, auc), auc, fill = matrix)) +
  geom_boxplot(outlier.size = 0.6) +
  scale_y_log10() +
  labs(
    x        = NULL,
    y        = "AUC144-168 (mg*h/L, log scale)",
    fill     = "Matrix",
    title    = "Replicates Figure 3 of Stott 2018",
    subtitle = "Simulated plasma and CSF AUC distributions at escalating dosages"
  ) +
  theme_minimal()
```

## Figure 4 -- probability of pharmacodynamic target attainment

Figure 4 plots the proportion of simulated patients reaching a plasma
AUC/MIC of at least 389.3, against the MIC of the infecting isolate. The shape
is reproduced below from the simulated cohort.

```{r figure-4-pta}
pd_target <- 389.3 # Stott 2018: murine stasis target, plasma AUC/MIC
mic_grid  <- c(0.25, 0.5, 1, 2, 4, 8, 16, 32)

pta <- tidyr::crossing(
  auc_cohort |> dplyr::select(dose_label, dose_mg, auc_plasma),
  mic = mic_grid
) |>
  dplyr::group_by(dose_label, dose_mg, mic) |>
  dplyr::summarise(
    pta = 100 * mean(auc_plasma / mic >= pd_target),
    .groups = "drop"
  )

ggplot(pta, aes(mic, pta, colour = stats::reorder(dose_label, dose_mg))) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.4) +
  scale_x_log10(breaks = mic_grid) +
  labs(
    x        = "MIC (mg/L)",
    y        = "Probability of target attainment (%)",
    colour   = "Regimen",
    title    = "Replicates Figure 4 of Stott 2018 in shape",
    subtitle = "Plasma AUC/MIC >= 389.3; see Assumptions for why the levels differ"
  ) +
  theme_minimal()

pta_1200_mic2 <- pta$pta[pta$dose_mg == 1200 & pta$mic == 2]
cat(sprintf(
  "PTA at 1200 mg q24h, MIC 2.0 mg/L: %.1f%% simulated vs 61%% published\n",
  pta_1200_mic2
))

# PTA must fall monotonically as the MIC rises and rise with dose. Those are
# structural, unlike the absolute level (see Assumptions and deviations).
pta_wide <- pta |>
  dplyr::arrange(dose_mg, mic) |>
  dplyr::group_by(dose_mg) |>
  dplyr::summarise(monotone = all(diff(pta) <= 1e-9), .groups = "drop")
stopifnot(
  all(pta_wide$monotone),
  pta$pta[pta$dose_mg == 2000 & pta$mic == 4] >
    pta$pta[pta$dose_mg == 400 & pta$mic == 4]
)
```

# PKNCA validation

NCA is run on the typical-value solve over the paper's 144-168 h window. A
full-IIV cohort makes Cmax / Cmin tmax-selected and poisons `lambda.z`, so the
per-parameter table uses the typical patient; the cohort is used for the
exposure medians in the next section.

```{r pknca-window}
nca_conc <- sim_typ_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(matrix = "Plasma") |>
  dplyr::select(id, time, dose_label, matrix, conc = Cc) |>
  dplyr::bind_rows(
    sim_typ_ss |>
      dplyr::filter(!is.na(Ccsf)) |>
      dplyr::mutate(matrix = "CSF") |>
      dplyr::select(id, time, dose_label, matrix, conc = Ccsf)
  )

nca_dose <- events_ss |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, dose_label) |>
  tidyr::crossing(matrix = c("Plasma", "CSF"))

nca_win <- suppressWarnings(PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_conc, conc ~ time | dose_label + matrix + id,
    concu = "mg/L", timeu = "hr"
  ),
  PKNCA::PKNCAdose(nca_dose, amt ~ time | dose_label + matrix + id,
    doseu = "mg"
  ),
  intervals = data.frame(
    start = ss_start, end = t_end,
    cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
  )
)))

knitr::kable(
  summary(nca_win),
  caption = paste(
    "Typical-value NCA over the paper's 144-168 h window, by matrix and",
    "dosage."
  )
)
```

The terminal half-life needs a single-dose profile with a long washout, so it
comes from the separate typical-value solve built above.

```{r pknca-half-life}
nca_conc_single <- sim_single |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(matrix = "Plasma") |>
  dplyr::select(id, time, dose_label, matrix, conc = Cc) |>
  dplyr::bind_rows(
    sim_single |>
      dplyr::filter(!is.na(Ccsf)) |>
      dplyr::mutate(matrix = "CSF") |>
      dplyr::select(id, time, dose_label, matrix, conc = Ccsf)
  )
stopifnot(all(nca_conc_single$conc >= 0), any(nca_conc_single$time == 0))

nca_dose_single <- ev_single |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, dose_label) |>
  tidyr::crossing(matrix = c("Plasma", "CSF"))

nca_single <- suppressWarnings(PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_conc_single, conc ~ time | dose_label + matrix + id,
    concu = "mg/L", timeu = "hr"
  ),
  PKNCA::PKNCAdose(nca_dose_single, amt ~ time | dose_label + matrix + id,
    doseu = "mg"
  ),
  intervals = data.frame(
    start = 0, end = 504,
    cmax = TRUE, tmax = TRUE, half.life = TRUE, aucinf.obs = TRUE
  )
)))

knitr::kable(
  summary(nca_single),
  caption = "Typical-value single-dose NCA (800 mg), used for the terminal half-life."
)
```

```{r half-life-nca-gate}
nca_single_res <- as.data.frame(nca_single$result)

half_life_plasma <- nca_single_res |>
  dplyr::filter(matrix == "Plasma", PPTESTCD == "half.life") |>
  dplyr::pull(PPORRES)
aucinf_plasma <- nca_single_res |>
  dplyr::filter(matrix == "Plasma", PPTESTCD == "aucinf.obs") |>
  dplyr::pull(PPORRES)
stopifnot(length(half_life_plasma) == 1L, length(aucinf_plasma) == 1L)

cat(sprintf(
  "PKNCA terminal half-life: %.2f h (closed form log(2) * Vss / SCL = %.2f h)\n",
  half_life_plasma, thalf_closed_form
))
cat(sprintf(
  "PKNCA single-dose AUCinf: %.1f mg*h/L (closed form dose / SCL = %.1f)\n",
  aucinf_plasma, 800 / 0.72
))

stopifnot(
  # Same parameters on both sides; the gap is lambda.z window selection.
  abs(half_life_plasma - thalf_closed_form) / thalf_closed_form < 0.05,
  # Exact mass-balance identity: all drug eventually clears through SCL.
  abs(aucinf_plasma - 800 / 0.72) / (800 / 0.72) < 0.02
)
```

# Comparison against published NCA

Stott 2018 reports exposure two ways, and they behave very differently against
this model.

1. **From the 43 study patients' Bayesian posteriors.** The 38 patients on
   800 mg q24h had a median plasma AUC144-168 of 945.4 mg*h/L (IQR 799.2-1139.8)
   and a median CSF AUC144-168 of 784.2 mg*h/L (IQR 615.9-879.4), with a mean
   CSF:plasma ratio of 0.82 (SD 0.22).
2. **From a 5000-draw Monte Carlo of the final joint density.** At 1200 mg q24h
   the median plasma AUC144-168 was 1143.2 mg*h/L (IQR 988.4-1378.0) and the
   median CSF AUC144-168 982.9 mg*h/L (IQR 781.0-1185.9), with a mean CSF:plasma
   ratio of 0.89 (SD 0.44).

```{r nca-compare}
sim_summary <- auc_cohort |>
  dplyr::filter(dose_mg %in% c(800, 1200)) |>
  dplyr::mutate(group = paste0(dose_mg, " mg q24h"))

simulated_long <- dplyr::bind_rows(
  sim_summary |>
    dplyr::transmute(id, group, matrix = "Plasma",
      PPTESTCD = "auclast", PPORRES = auc_plasma),
  sim_summary |>
    dplyr::transmute(id, group, matrix = "CSF",
      PPTESTCD = "auclast", PPORRES = auc_csf)
)

published <- tibble::tribble(
  ~group,          ~matrix,  ~auclast,
  "800 mg q24h",   "Plasma", 945.4,
  "800 mg q24h",   "CSF",    784.2,
  "1200 mg q24h",  "Plasma", 1143.2,
  "1200 mg q24h",  "CSF",    982.9
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = simulated_long,
  reference     = published,
  by            = c("group", "matrix"),
  units         = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)

cmp |>
  dplyr::rename("Regimen" = group, "Matrix" = matrix) |>
  knitr::kable(
    digits  = 2,
    caption = paste(
      "Simulated cohort medians versus Stott 2018.",
      "The 800 mg references are Bayesian posterior medians across the 38",
      "patients who received that dosage; the 1200 mg references are medians",
      "of the paper's 5000-draw Monte Carlo.",
      "* differs from the published value by more than 20%."
    )
  )
```

```{r nca-compare-gates}
med <- function(g, m) {
  v <- simulated_long$PPORRES[simulated_long$group == g &
    simulated_long$matrix == m]
  if (length(v) < 1L) stop("no rows for ", g, " / ", m)
  stats::median(v)
}
pct <- function(sim, ref) 100 * (sim - ref) / ref

ratio_mean_1200 <- mean(auc_cohort$ratio[auc_cohort$dose_mg == 1200])
ratio_mean_800  <- mean(auc_cohort$ratio[auc_cohort$dose_mg == 800])

cat(sprintf("800 mg  plasma AUC144-168 median: %7.1f vs  945.4 (%+.1f%%)\n",
            med("800 mg q24h", "Plasma"), pct(med("800 mg q24h", "Plasma"), 945.4)))
cat(sprintf("800 mg  CSF    AUC144-168 median: %7.1f vs  784.2 (%+.1f%%)\n",
            med("800 mg q24h", "CSF"), pct(med("800 mg q24h", "CSF"), 784.2)))
cat(sprintf("1200 mg plasma AUC144-168 median: %7.1f vs 1143.2 (%+.1f%%)\n",
            med("1200 mg q24h", "Plasma"), pct(med("1200 mg q24h", "Plasma"), 1143.2)))
cat(sprintf("1200 mg CSF    AUC144-168 median: %7.1f vs  982.9 (%+.1f%%)\n",
            med("1200 mg q24h", "CSF"), pct(med("1200 mg q24h", "CSF"), 982.9)))
cat(sprintf("Mean CSF:plasma ratio, 800 mg: %.3f (paper posteriors 0.82)\n",
            ratio_mean_800))
cat(sprintf("Mean CSF:plasma ratio, 1200 mg: %.3f (paper Monte Carlo 0.89)\n",
            ratio_mean_1200))

# The 800 mg plasma median is the paper's best-supported exposure number -- it
# comes from 38 real patients' posteriors rather than from the Monte Carlo -- and
# it is the gate here. Realised across authoring runs at 2 and 16 solver threads:
# -0.1% and -2.8%. rxSetSeed() fixes the draw per solver-thread count and not
# across them, so the bound is set well outside that spread while still going red
# on a mis-transcribed clearance or dose, which move exposure by a factor.
stopifnot(
  abs(pct(med("800 mg q24h", "Plasma"), 945.4)) < 25,
  # Mean of a ratio of four log-normals; realised 0.90 and 1.04 against the
  # paper's 0.89. Centre only -- the spread is over-dispersed, see below.
  abs(pct(ratio_mean_1200, 0.89)) < 40
)
```

The 800 mg plasma row -- the only exposure the paper derives directly from its
own patients -- reproduces within a few percent. The CSF rows sit 10-25% low and
both 1200 mg rows sit about 20% high. The reasons are structural and are set out
below; no parameter was adjusted.

# Assumptions and deviations

- **Non-parametric support points approximated by a diagonal log-normal.**
  Stott 2018 fitted the model with Pmetrics' non-parametric adaptive grid
  algorithm, which yields a discrete, correlated support-point distribution
  rather than an OMEGA matrix; Table 3 publishes only a per-parameter Mean,
  Median and SD. The packaged model takes the **Mean** column as the typical
  value -- the set the paper endorses, since "the mean parameter estimates
  better fitted the data than medians and were used to calculate Bayesian
  estimates of drug exposure for each individual patient" -- and sets each IIV
  variance to `log(1 + (SD/Mean)^2)`, with no off-diagonal covariances. This is
  the same encoding used for the companion `Stott_2023_flucytosine` extraction.
  Two consequences follow, and neither is a transcription error:
  - The model's population **median** for each parameter equals the paper's
    reported **mean**, not its reported median. For Ka the two differ by a
    factor of five (mean 8.78 /h, median 1.73 /h), which is the signature of a
    strongly skewed support-point distribution that no single log-normal can
    carry faithfully. Ka is not identifiable from these data in any case -- the
    absorption rate is at least two orders of magnitude faster than elimination,
    so it has essentially no influence on exposure.
  - Dropping the off-diagonals **over-disperses derived quantities**. The
    CSF:plasma ratio is a function of four parameters (Kcs, Ksc, Vc, Vcns) whose
    support-point values are co-determined in the real fit; with independent etas
    the simulated ratio reproduces the published **mean** (0.89) but carries an
    SD roughly twice the published 0.44. Any analysis that depends on the spread
    of the ratio rather than its centre should not use this encoding.

- **The paper's Monte Carlo results cannot be recovered from its own Table 3,
  and this is a property of the source rather than of the encoding.** Two
  independent checks show it:
  - *Level.* Under a linear model the 144-168 h exposure is
    `accumulation x dose / SCL`. At 1200 mg q24h with the Table 3 **mean**
    clearance this is 1499 mg*h/L, and with the Table 3 **median** clearance it
    is about 1552 mg*h/L. The paper reports a simulated median of
    1143.2 mg*h/L, which implies a clearance near 0.94 L/h -- 31% above the
    reported mean and 45% above the reported median. No reading of Table 3
    produces it. By contrast the same arithmetic applied at 800 mg reproduces
    the paper's patient-posterior median almost exactly, which is why that row
    is the one gated above.
  - *Spread.* The paper reports 61% target attainment at 1200 mg q24h against an
    MIC of 2.0 mg/L, i.e. 39% of its 5000 draws fall below an AUC of 778.6.
    Relative to its own reported median of 1143.2 that places the 39th
    percentile at 0.68 of the median, which for a log-normal implies a CV of
    roughly 240%. The clearance SD in Table 3 is 0.24 on a mean of 0.72, a CV of
    33%. The Monte Carlo distribution behind Figures 3 and 4 is therefore far
    more dispersed than the variability Table 3 reports, and the simulated PTA
    in this vignette is correspondingly much higher than the published curve.
    Figure 4 is reproduced for its **shape** -- monotone decreasing in MIC,
    increasing in dose, which is what the gate asserts -- not its level.

  Both observations point the same way: Table 3 is internally consistent with
  the patient-derived exposures and with the model equations, but not with the
  simulation outputs. The packaged model encodes Table 3 faithfully and the gap
  is documented rather than closed by tuning.

- **Estimated initial conditions are deliberately not encoded.** Some patients
  had taken fluconazole at an undocumented time before enrolment, leaving
  detectable drug in the first PK sample. The source absorbed this by estimating
  a non-zero initial condition for each of the four compartments (Table 3:
  ICgut 34.67 mg, ICcentral 35.86 mg, ICCNS 31.06 mg, ICperipheral 34.29 mg),
  each multiplied by a binary switch set to 1 when fluconazole was detected in
  the first PK sample. These are a fitting device for this specific 43-patient
  data set rather than a generalisable patient characteristic, and they are
  close to unidentified -- all four medians sit at 49.96-49.99 mg, essentially
  the midpoint of the search grid. The source itself states that "for the
  simulations, the initial conditions of all compartments were defaulted to
  zero", so every simulated result it reports uses zero initial conditions, as
  does this vignette. Users wishing to reproduce the baseline-positive patients
  should pre-load the compartments with the Table 3 amounts. This follows the
  established handling of the same artefact in
  `modellib("Debord_2001_cyclosporin")` (residual concentration C0) and
  `modellib("Mosha_2014_lumefantrine")` (residual previous-treatment dose F0).

- **Residual error fixed to zero.** The Methods state that model error was split
  between process noise, "modeled using lambda, an additive error term", and
  assay variance used to weight the data. The *form* is therefore given but no
  numeric value for lambda or for the assay error polynomial appears anywhere in
  the paper, and there is no supplement. Both an additive and a proportional
  term are carried for each output so the structure is available, and both are
  fixed to zero rather than invented. The only precision figures the paper gives
  are LC-MS/MS quality-control CVs (plasma intraday < 3.4%, interday < 6.7%; CSF
  intraday < 5.2%, interday < 5.3%), which characterise the assay and not the
  model's residual variability. Simulations therefore carry IIV but no residual
  error.

- **The 144-168 h window is not steady state.** The terminal half-life implied
  by Table 3 is about 51 h, so after six q24h doses the window sits at roughly
  90% of steady-state exposure. The vignette dose-loads to 144 h exactly as the
  paper describes rather than solving to steady state, so the comparison is
  like-for-like; the 10% shortfall against `dose / SCL` is real and expected,
  not a numerical artefact.

- **Table 1's combined sex counts are internally inconsistent.** The "Combined"
  column reports 23 males and 20 females, which contradicts both its own
  per-site columns (13 + 8 = 21 males; 10 + 12 = 22 females) and the Results
  text, "Twenty-two patients (52%) were female". The per-site sums and the
  Results text agree, so the model metadata records 22 female of 43 (51%) and
  treats the Combined column as a transcription error in the source. Nothing in
  the model depends on it -- sex was screened and rejected.

- **Weight enters nowhere.** No model parameter scales with body weight, and
  fluconazole was dosed as a flat mg amount rather than per kg, so unlike the
  flucytosine companion this vignette needs no weight assumption at all.

- **The paper's Kcs / Ksc and Kcp / Kpc become `lk_central_csf` /
  `lk_csf_central` and `lk12` / `lk21`.** Those canonical names were ratified in
  `inst/references/parameter-names.md` alongside the `Stott_2023_flucytosine`
  extraction, which fits the identical structure; this model reuses them and
  introduces no new parameter, compartment or covariate name. `csf` is already a
  canonical compartment, and the enrollment-site covariate the paper screens
  under the name "ethnicity" maps onto the existing `REGION_VIETNAM` canonical.

- **Observations are anchored on `central`, not on `Cc` / `Ccsf`.** Writing an
  algebraic observable into an event table's `cmt` column makes rxUi inject a
  compartment slot for it after the ODE states and silently renumber them.
  Because this model declares two endpoints, observation rows additionally carry
  `dvid = 1`; rxSolve returns both observables as columns on every observation
  row, so one anchor per time point covers both matrices.
