---
title: "Amphotericin B deoxycholate (Stott 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Amphotericin B deoxycholate (Stott 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
ui <- rxode2::rxode(readModelDb("Stott_2018_amphotericinB_deoxycholate"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1128/AAC.02526-17>

Stott 2018 fitted four two-compartment models to the same data set with the
nonparametric adaptive grid (NPAG) algorithm of Pmetrics 1.5.0. Model 2, which
carries body weight on both clearance and central volume as an additive linear
intercept-plus-slope term, was selected as the final model (highest log
likelihood -42.8 and lowest AIC 101.9 of the four; Table 3). Models 1, 3 and 4
are base / rejected development steps and are not packaged separately.

**Solve this model with `useLinCmt = FALSE`.** Its `model()` block exposes only
the micro-constants `k12` and `k21` and no `q` / `vp` pair, so `rxSolve`'s
default ODE-to-`linCmt()` auto-conversion has no well-formed two-compartment
macro-parameterisation to build and silently discards `peripheral1`. The
collapse preserves `Dose / CL`, so an AUC check alone cannot detect it; the
terminal-half-life gate below is what makes it visible.

# Population

The model was fitted to 42 HIV-positive adults with cryptococcal meningitis, 22
recruited at the Hospital for Tropical Diseases in Ho Chi Minh City, Vietnam and
20 at Masaka General Hospital, Uganda (Stott 2018 Table 1). Twenty-two (52%)
were female. The combined cohort had a median age of 33 years (range 20 to 73),
a median weight of 48 kg (32 to 68), a median body mass index of 18 kg/m^2 (12
to 25), a median serum creatinine of 69 umol/L (37 to 167) and a median
Cockcroft-Gault eGFR of 76.7 mL/min/1.73 m^2 (35.4 to 146.7). No demographic
variable differed significantly between the two sites.

All patients received amphotericin B deoxycholate 1 mg/kg once daily as an
intravenous infusion over 5 to 6 h, together with fluconazole 800 mg/day.
Sampling was predose then at 1, 2, 4, 8, 12 and 24 h after the start of the
infusion (the first five patients additionally at 16 and 20 h), on treatment
days 1 or 2 and day 7, with additional sparse samples taken alongside clinically
indicated lumbar punctures out to day 17. Of 553 plasma samples, 479 were
analysable (mean 11.4 per patient, range 6 to 18); 74 were discarded because the
draw time had not been recorded.

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Stott_2018_amphotericinB_deoxycholate.R` carries an
in-file comment pointing at its source location. They are collected here.

| Equation / parameter | Value | Source location |
|---|---|---|
| `d/dt(central)` | n/a | Equation a.2 (Materials and Methods): `dX(1)/dt = R(1) - [ (Int_c + Wt*Sl_c) / (Int_v + Wt*Sl_v) + K12 ] * X(1) + K21 * X(2)` |
| `d/dt(peripheral1)` | n/a | Equation b: `dX(2)/dt = K12 * X(1) - K21 * X(2)` |
| `Cc` | n/a | Equation c: `Y(1) = X(1) / V` |
| `cl` | n/a | Equation e: `SCL = Int_c + (Wt * Sl_c)` |
| `vc` | n/a | Equation f: `V = Int_v + (Wt * Sl_v)` |
| `lcl` | `log(0.67)` | Table 2, model 2, `SCL_intercept (liters/h)` mean |
| `e_wt_cl` | `0.03` | Table 2, model 2, `SCL_slope (liters/h/kg)` mean |
| `lvc` | `log(1.76)` | Table 2, model 2, `Vc_intercept (liters)` mean |
| `e_wt_vc` | `0.82` | Table 2, model 2, `Vc_slope (liters/kg)` mean |
| `lk12` | `log(5.36)` | Table 2, model 2, `K12 (h-1)` mean |
| `lk21` | `log(9.92)` | Table 2, model 2, `K21 (h-1)` mean |
| `etalcl` | `0.000223` | Table 2, model 2, `SCL_intercept` SD 0.01; `log(1 + (0.01/0.67)^2)` |
| `etae_wt_cl` | `0.105361` | Table 2, model 2, `SCL_slope` SD 0.01; `log(1 + (0.01/0.03)^2)` |
| `etalvc` | `0.429977` | Table 2, model 2, `Vc_intercept` SD 1.29; `log(1 + (1.29/1.76)^2)` |
| `etae_wt_vc` | `0.668759` | Table 2, model 2, `Vc_slope` SD 0.80; `log(1 + (0.80/0.82)^2)` |
| `etalk12` | `0.951894` | Table 2, model 2, `K12` SD 6.76; `log(1 + (6.76/5.36)^2)` |
| `etalk21` | `0.928184` | Table 2, model 2, `K21` SD 12.27; `log(1 + (12.27/9.92)^2)` |
| `propSd` | `0.093` | Materials and Methods, "Measurement of amphotericin B concentrations": assay coefficient of variation below 9.3 percent over 0.05 to 8 mg/L. Not a fitted residual - see Assumptions and deviations |
| `WT` | covariate | Table 1, `Weight (kg)` row, Combined column: median 48, range 32 to 68 |

The paper's printed constants are restated once here and used from this list
throughout, so that every check below is driven by numbers transcribed from the
paper rather than by the model's own variables.

```{r paper-constants}
paper <- list(
  cl_int = 0.67, # Table 2 model 2 SCL intercept (L/h)
  cl_slope = 0.03, # Table 2 model 2 SCL slope (L/h/kg)
  vc_int = 1.76, # Table 2 model 2 Vc intercept (L)
  vc_slope = 0.82, # Table 2 model 2 Vc slope (L/kg)
  k12 = 5.36, # Table 2 model 2 K12 (1/h)
  k21 = 9.92, # Table 2 model 2 K21 (1/h)
  wt_median = 48, # Table 1 combined median weight (kg)
  wt_min = 32, # Table 1 combined weight range
  wt_max = 68,
  tinf = 5.5, # Monte Carlo simulation: "infused over 5.5 h"
  tau = 24 # q24h
)

paper_cl <- function(wt) paper$cl_int + paper$cl_slope * wt
paper_vc <- function(wt) paper$vc_int + paper$vc_slope * wt
```

# Deterministic structural checks

These checks use the typical-value model (`zeroRe()`) and compare it against
arithmetic carried out independently from the printed constants above. They are
deterministic, so they are asserted tightly.

```{r typical-model}
ui_typ <- rxode2::zeroRe(ui)
```

## Clearance and central volume versus body weight

Equations e and f are additive linear functions of weight with a separately
estimated intercept and slope. Note that the intercept is the value
extrapolated to zero body weight, **not** a typical-patient value, and that
nothing is centred on a reference weight.

```{r check-covariate}
wt_grid <- seq(paper$wt_min, paper$wt_max, by = 2)
ev_cov <- tidyr::crossing(tibble(id = seq_along(wt_grid), WT = wt_grid), time = c(0, 1)) |>
  mutate(evid = 0L, amt = NA_real_, cmt = "central")

sim_cov <- rxode2::rxSolve(
  ui_typ,
  events = as.data.frame(ev_cov),
  keep = "WT",
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  distinct(id, WT, cl, vc)

chk_cov <- sim_cov |>
  mutate(
    cl_paper = paper_cl(WT),
    vc_paper = paper_vc(WT),
    cl_pct = 100 * (cl - cl_paper) / cl_paper,
    vc_pct = 100 * (vc - vc_paper) / vc_paper
  )

# Deterministic algebra: the model must reproduce equations e and f exactly.
stopifnot(
  max(abs(chk_cov$cl_pct)) < 1e-8,
  max(abs(chk_cov$vc_pct)) < 1e-8
)

chk_cov |>
  filter(WT %in% c(32, 48, 68)) |>
  transmute(
    "Weight (kg)" = WT,
    "CL, paper equation e (L/h)" = cl_paper,
    "CL, model (L/h)" = cl,
    "Vc, paper equation f (L)" = vc_paper,
    "Vc, model (L)" = vc
  ) |>
  knitr::kable(digits = 4, caption = "Equations e and f reproduced at the minimum, median and maximum cohort weights.")
```

```{r figure-2a}
# Companion to Figure 2a of Stott 2018. NOTE: the regression lines printed in
# that figure (CL = 0.05*weight - 0.2, V = 1.08*weight - 24.8) are ordinary
# least-squares fits through the MODEL 1 Bayesian posteriors, not the model 2
# population equations plotted here, so the two are not expected to coincide.
chk_cov |>
  select(WT, Clearance = cl, `Central volume` = vc) |>
  tidyr::pivot_longer(-WT) |>
  ggplot(aes(WT, value)) +
  geom_line() +
  facet_wrap(~name, scales = "free_y") +
  labs(
    x = "Body weight (kg)", y = "Typical value (L/h or L)",
    title = "Model 2 weight relations",
    caption = "Equations e and f of Stott 2018 over the observed weight range (32-68 kg)."
  ) +
  theme_bw()
```

## Whole-ODE check against an independent biexponential

The two-compartment infusion system has a closed-form solution. The function
below is written from scratch out of the paper's printed micro-constants and
weight relations and is independent of the packaged model, so agreement
exercises the full ODE block, the dosing route and the observation equation at
once. It is also the gate that detects the silent `useLinCmt` compartment
collapse, which leaves `Dose / CL` -- and therefore every AUC check -- intact
while changing the shape of the curve.

```{r closed-form}
biexp_conc <- function(times, wt, mgkg, ndose, tau, tinf, p = paper) {
  cl <- p$cl_int + p$cl_slope * wt
  vc <- p$vc_int + p$vc_slope * wt
  k10 <- cl / vc
  b <- k10 + p$k12 + p$k21
  cprod <- k10 * p$k21
  alpha <- (b + sqrt(b^2 - 4 * cprod)) / 2
  beta <- (b - sqrt(b^2 - 4 * cprod)) / 2
  rate <- mgkg * wt / tinf
  # Unit-impulse amount response in the central compartment, integrated over a
  # zero-order input of duration tinf and superposed over the dosing history.
  ca <- (alpha - p$k21) / ((alpha - beta) * alpha)
  cb <- (p$k21 - beta) / ((alpha - beta) * beta)
  # Central amount under an infusion of `rate` started u time units ago and
  # never stopped. A finite infusion is that, minus the same thing started
  # tinf later.
  step <- function(u) {
    ifelse(u <= 0, 0, rate * (ca * (1 - exp(-alpha * pmax(u, 0))) + cb * (1 - exp(-beta * pmax(u, 0)))))
  }
  amt <- rep(0, length(times))
  for (d in seq_len(ndose)) {
    t0 <- (d - 1) * tau
    amt <- amt + step(times - t0) - step(times - t0 - tinf)
  }
  list(conc = amt / vc, alpha = alpha, beta = beta, cl = cl, vc = vc)
}

t_grid <- seq(0, 192, by = 0.5)
ref <- biexp_conc(t_grid, paper$wt_median, 1.0, ndose = 7, tau = paper$tau, tinf = paper$tinf)

ev_typ <- bind_rows(
  tibble(
    id = 1L, time = seq(0, paper$tau * 6, by = paper$tau), evid = 1L,
    amt = 1.0 * paper$wt_median, rate = 1.0 * paper$wt_median / paper$tinf, cmt = "central"
  ),
  tibble(id = 1L, time = t_grid, evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")
) |>
  mutate(WT = paper$wt_median) |>
  arrange(time, desc(evid))

sim_typ <- rxode2::rxSolve(
  ui_typ,
  events = as.data.frame(ev_typ),
  useLinCmt = FALSE,
  returnType = "data.frame"
)

cmp_typ <- tibble(time = t_grid, model = sim_typ$Cc[match(t_grid, sim_typ$time)], closed = ref$conc) |>
  filter(time > 0)
rel_err <- max(abs(cmp_typ$model - cmp_typ$closed) / cmp_typ$closed)

# Deterministic: two solutions of the same linear ODE. 1e-4 is ~100x the
# realised solver error and still goes red on a dropped peripheral
# compartment, a wrong micro-constant or a wrong infusion duration.
stopifnot(rel_err < 1e-4)
rel_err
```

```{r figure-closed-form}
cmp_typ |>
  tidyr::pivot_longer(-time) |>
  mutate(name = recode(name, model = "Packaged model", closed = "Closed-form biexponential")) |>
  ggplot(aes(time, value, colour = name, linetype = name)) +
  geom_line(linewidth = 0.7) +
  labs(
    x = "Time (h)", y = "Amphotericin B (mg/L)", colour = NULL, linetype = NULL,
    title = "Typical-value profile, 1 mg/kg q24h at 48 kg",
    caption = "Packaged model versus a closed-form solution built from the printed constants."
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

## Terminal half-life

```{r check-halflife}
# Closed-form terminal half-life from the paper's printed constants at the
# median cohort weight.
t_half_paper <- log(2) / ref$beta

# Empirical terminal slope of the packaged model after a single dose, fitted
# well clear of the distribution phase.
ev_single <- bind_rows(
  tibble(
    id = 1L, time = 0, evid = 1L, amt = paper$wt_median,
    rate = paper$wt_median / paper$tinf, cmt = "central"
  ),
  tibble(id = 1L, time = seq(0, 240, by = 0.5), evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")
) |>
  mutate(WT = paper$wt_median) |>
  arrange(time, desc(evid))

sim_single <- rxode2::rxSolve(
  ui_typ,
  events = as.data.frame(ev_single),
  useLinCmt = FALSE,
  returnType = "data.frame"
)

# The peripheral compartment must be present; its absence is the useLinCmt
# collapse and would halve the fitted half-life below.
stopifnot("peripheral1" %in% names(sim_single))

late <- sim_single |> filter(time >= 150, time <= 240, Cc > 0)
t_half_model <- log(2) / -coef(stats::lm(log(Cc) ~ time, data = late))[["time"]]

# Deterministic. Realised agreement is <0.1%; 2% still goes red on the
# useLinCmt collapse, which returns log(2)/kel = 13.5 h instead of 20.8 h.
stopifnot(abs(t_half_model - t_half_paper) / t_half_paper < 0.02)

tibble(
  Quantity = c(
    "alpha (1/h)", "beta (1/h)",
    "Terminal half-life, closed form (h)", "Terminal half-life, packaged model (h)"
  ),
  Value = c(ref$alpha, ref$beta, t_half_paper, t_half_model)
) |>
  knitr::kable(digits = 4, caption = "Hybrid rate constants at the median cohort weight of 48 kg.")
```

## Steady-state mass balance

At steady state the AUC over one dosing interval equals `Dose / CL` exactly.
Stott 2018 reports its simulated exposures as AUC over 144 to 168 h, that is the
seventh dosing interval, so the same interval is used here.

```{r check-massbalance}
w7 <- sim_typ |> filter(time >= 144, time <= 168)
auc_typ <- sum(diff(w7$time) * (head(w7$Cc, -1) + tail(w7$Cc, -1)) / 2)
auc_closed <- paper$wt_median / paper_cl(paper$wt_median)

# Deterministic; the small shortfall is the residual approach to steady state
# after seven doses (terminal half-life 20.8 h), not solver error.
stopifnot(abs(auc_typ - auc_closed) / auc_closed < 0.02)

tibble(
  Quantity = c("Dose / CL at 48 kg (mg*h/L)", "Simulated AUC 144-168 h (mg*h/L)"),
  Value = c(auc_closed, auc_typ)
) |>
  knitr::kable(digits = 3, caption = "Steady-state mass balance for the typical 48 kg patient at 1 mg/kg q24h.")
```

# Virtual cohort

Original observed data are not publicly available. The Monte Carlo section of
Stott 2018 simulated the three dosages for which clinical outcome data were
available (0.4, 0.7 and 1.0 mg/kg q24h), converted each weight-based dosage to
an absolute dose by multiplying by the simulated patient's weight, infused over
5.5 h, and limited simulated weights to the range observed in the clinical
cohort. The cohort below follows that description, at 200 subjects per arm.

```{r cohort}
# set.seed() seeds R's RNG, which is what draws the weights below. It does NOT
# seed rxode2's eta sampler, whose streams are partitioned per solver thread, so
# the drawn cohort differs between a 2-core CI runner and a 16-thread
# workstation. Every assertion below is written to hold for any cohort this
# model can produce.
set.seed(20180626)

n_arm <- 200L
n_dose <- 8L

# One virtual population, dosed three ways, as in the paper's Monte Carlo
# section. Sharing the weight vector across arms removes weight sampling as a
# source of between-arm difference; the eta draws still differ per subject id.
cohort_wt <- stats::runif(n_arm, paper$wt_min, paper$wt_max)

make_cohort <- function(n, mgkg, label, id_offset = 0L) {
  subj <- tibble(
    id = id_offset + seq_len(n),
    WT = cohort_wt,
    treatment = label
  ) |>
    mutate(dose_mg = mgkg * WT)

  dosing <- subj |>
    tidyr::crossing(time = paper$tau * seq(0, n_dose - 1L)) |>
    mutate(evid = 1L, amt = dose_mg, rate = dose_mg / paper$tinf, cmt = "central")

  obs <- subj |>
    tidyr::crossing(time = seq(0, paper$tau * n_dose, by = 0.5)) |>
    mutate(evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")

  bind_rows(dosing, obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, evid, amt, rate, cmt, WT, treatment)
}

arms <- tibble(
  mgkg = c(0.4, 0.7, 1.0),
  label = c("0.4 mg/kg q24h", "0.7 mg/kg q24h", "1.0 mg/kg q24h")
)

events <- bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  make_cohort(n_arm, arms$mgkg[i], arms$label[i], id_offset = (i - 1L) * n_arm)
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
sim <- rxode2::rxSolve(
  ui,
  events = as.data.frame(events),
  keep = c("WT", "treatment"),
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  mutate(treatment = factor(treatment, levels = arms$label))

stopifnot(
  dplyr::n_distinct(sim$id) == 3L * n_arm,
  all(sim$Cc >= 0)
)
```

# Replicate published figures

```{r figure-4}
# Replicates Figure 4 of Stott 2018: 5th, 50th and 95th percentiles of
# amphotericin B concentration for the 1 mg/kg q24h regimen. Stott 2018 carries
# its VPC to 420 h; 192 h is shown here, which covers the PK sampling window and
# the AUC 144-168 h interval used for the exposure comparison.
sim |>
  filter(treatment == "1.0 mg/kg q24h") |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2) +
  geom_line(aes(y = Q50), linewidth = 0.6) +
  geom_line(aes(y = Q05), linewidth = 0.3) +
  geom_line(aes(y = Q95), linewidth = 0.3) +
  labs(
    x = "Time (h)", y = "Amphotericin B (mg/L)",
    title = "Simulated 5th, 50th and 95th percentiles, 1 mg/kg q24h",
    caption = "Companion to Figure 4 of Stott 2018 (observed data not available)."
  ) +
  theme_bw()
```

```{r figure-6}
# Replicates Figure 6 of Stott 2018: AUC 144-168 h distributions by regimen.
auc_subject <- sim |>
  filter(time >= 144, time <= 168) |>
  arrange(treatment, id, time) |>
  group_by(treatment, id) |>
  summarise(
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

ggplot(auc_subject, aes(auc)) +
  geom_histogram(bins = 30, colour = "white") +
  facet_wrap(~treatment, scales = "free_y") +
  labs(
    x = "AUC 144-168 h (mg*h/L)", y = "Simulated patients",
    title = "Simulated AUC distributions by dosage",
    caption = "Companion to Figure 6 of Stott 2018."
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject; pre-dose concentration is zero for this
# intravenous model. Existing time = 0 rows win.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca),
  Cc ~ time | treatment + id,
  concu = "mg/L", timeu = "h"
)

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | treatment + id, doseu = "mg")

# The seventh dosing interval, matching the AUC 144-168 h window that Stott 2018
# reports for its Monte Carlo simulations.
intervals <- data.frame(
  start = 144,
  end = 168,
  auclast = TRUE,
  cmax = TRUE,
  cmin = TRUE,
  tmax = TRUE,
  cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_res$result |>
  filter(PPTESTCD %in% c("auclast", "cmax", "cmin", "cav")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(
    median = median(PPORRES),
    q25 = quantile(PPORRES, 0.25),
    q75 = quantile(PPORRES, 0.75),
    .groups = "drop"
  ) |>
  mutate(PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  rename(
    "Regimen" = treatment, "NCA parameter" = PPTESTCD,
    "Median" = median, "P25" = q25, "P75" = q75
  ) |>
  knitr::kable(digits = 2, caption = "Simulated steady-state NCA over 144 to 168 h.")
```

## Comparison against the published Monte Carlo exposures

```{r nca_compare}
published <- tibble::tribble(
  ~treatment, ~auclast,
  "0.4 mg/kg q24h", 5.83,
  "0.7 mg/kg q24h", 10.16,
  "1.0 mg/kg q24h", 14.51
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = "auclast",
  units = c(auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated versus published median AUC 144-168 h. * differs from the reference by more than 20 percent.",
  align = c("l", "l", "r", "r", "r")
)
```

All three rows are starred: the simulated median exposures run about 45 to 50
percent above the medians Stott 2018 obtained from its own Monte Carlo runs
(reported medians 5.83, 10.16 and 14.51 mg\*h/L for 0.4, 0.7 and 1.0 mg/kg
q24h). **This deviation is documented, not tuned away**, and the ratio is
essentially constant across the three dosages, which locates it in the
clearance scale of the simulated population rather than in the dose, the
infusion or the ODE structure.

The likely mechanism is the distribution shape, not the point estimates. Two
lines of evidence:

* The typical-value exposure is not in question. The mass-balance check above
  gives `Dose / CL = ` `r round(auc_closed, 2)` mg\*h/L for the median 48 kg
  patient at 1 mg/kg from the paper's own printed clearance equation, and the
  packaged model reproduces it. Any simulated median near 14.5 mg\*h/L requires
  a simulated clearance around 3.3 L/h, well above the 2.11 L/h that equations
  e and f give at the cohort median weight.
* Pmetrics NPAG estimates a discrete, non-Gaussian joint distribution over
  support points, and Table 2 shows how far from log-normal it is: the reported
  median `K21` is 0.46 /h against a mean of 9.92 /h, and the median `Vc` slope
  is 0.36 L/kg against a mean of 0.82 L/kg. A subject drawn near the `K21`
  median has a terminal half-life of roughly 63 h rather than the 20.8 h of the
  mean-parameter patient, and so is still far from steady state at 144 to 168 h.
  Stott 2018's own Figure 4 shows exactly this: its simulated percentiles are
  still visibly accumulating at 400 h. A log-normal approximation with median
  9.92 /h places almost no mass near 0.46 /h and therefore reaches steady state
  within the interval, which raises the interval AUC.

Independent non-compartmental literature that Stott 2018 itself cites sits on
the side of the values simulated here. Bekersky et al. measured an AUC0-24 of
13.9 +/- 2 mg\*h/L after 0.6 mg/kg in healthy volunteers, which scales linearly
to about 16 mg\*h/L at 0.7 mg/kg, against 10.16 mg\*h/L from the paper's
simulation; and Ayestaran et al. measured 28.98 +/- 15.46 mg\*h/L at 1 mg/kg in
neutropenic patients, which the Discussion of Stott 2018 acknowledges its own
simulated 14.51 mg\*h/L is "considerably lower" than. The values simulated here
(15.0 and 21.7 mg\*h/L) sit between the two published NCA results. The paper's
Figure 1 day-7 observations, clustered around 0.5 to 1.5 mg/L at 1 mg/kg, also
imply an interval average nearer 1 mg/L (AUC about 24 mg\*h/L) than the 0.60
mg/L (AUC 14.5 mg\*h/L) that the reported simulation median implies.

The gate below therefore asserts only the internally verifiable properties: dose
proportionality of the simulated exposures, and agreement of the simulated
median with the published ratio *between* dose levels.

```{r nca_gate}
# Dose proportionality is structural in a linear model, so it is checked
# deterministically on the typical-value model first: exact, no cohort noise.
auc_typ_dose <- vapply(arms$mgkg, function(mgkg) {
  ev <- bind_rows(
    tibble(
      id = 1L, time = paper$tau * seq(0, n_dose - 1L), evid = 1L,
      amt = mgkg * paper$wt_median, rate = mgkg * paper$wt_median / paper$tinf, cmt = "central"
    ),
    tibble(id = 1L, time = seq(144, 168, by = 0.25), evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")
  ) |>
    mutate(WT = paper$wt_median) |>
    arrange(time, desc(evid))
  s <- rxode2::rxSolve(ui_typ, events = as.data.frame(ev), useLinCmt = FALSE, returnType = "data.frame")
  s <- s[s$time >= 144 & s$time <= 168, ]
  sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
}, numeric(1))

dose_norm_typ <- auc_typ_dose / arms$mgkg
stopifnot(max(dose_norm_typ) / min(dose_norm_typ) - 1 < 1e-8)

med <- auc_subject |>
  group_by(treatment) |>
  summarise(auc = median(auc), .groups = "drop") |>
  arrange(treatment)

# The same claim on the simulated cohort. Each arm draws its own etas, so the
# arm medians carry sampling noise: the spread ran 1.013 to 1.041 over three
# independently seeded 200-per-arm cohorts. 1.15 admits that and still goes red
# on a non-linear encoding, which would move it by tens of percent.
dose_norm <- med$auc / arms$mgkg
stopifnot(max(dose_norm) / min(dose_norm) < 1.15)

# Ratio to the 1 mg/kg arm, compared against the published ratios. This is
# insensitive to the clearance-scale deviation discussed above; the realised
# deviation ran 0.007 to 0.016 over the same three cohorts.
ratio_sim <- med$auc / med$auc[3]
ratio_pub <- published$auclast / published$auclast[3]
stopifnot(max(abs(ratio_sim - ratio_pub)) < 0.06)

tibble(
  Regimen = med$treatment,
  "Typical-value AUC / dose (mg*h/L per mg/kg)" = dose_norm_typ,
  "Cohort median AUC / dose (mg*h/L per mg/kg)" = dose_norm,
  "Simulated ratio to 1.0 mg/kg" = ratio_sim,
  "Published ratio to 1.0 mg/kg" = ratio_pub
) |>
  knitr::kable(digits = 3, caption = "Dose proportionality and between-regimen exposure ratios.")
```

# Assumptions and deviations

* **Model 2 only.** Stott 2018 reports four two-compartment models. Models 1, 3
  and 4 are base and rejected covariate-model development steps (Table 3: model
  2 has the highest log likelihood and lowest AIC), so only model 2 is packaged,
  per the library's replicate-the-author's-structure policy for
  base-plus-final development papers.

* **Nonparametric distributions approximated as log-normal.** Pmetrics NPAG
  estimates a discrete joint distribution over support points and Stott 2018
  publishes only its marginal mean, median and SD (Table 2). Each parameter is
  therefore encoded as a log-normal whose median is the reported mean and whose
  variance is `log(1 + (SD / mean)^2)`. The reported means are used as the
  typical values because Results states that "the mean parameter values
  predicted the observed values better than the medians", and because the
  Abstract quotes the means. The nonparametric shape, which Table 2 shows is
  strongly right-skewed (median `K21` 0.46 /h against a mean of 9.92 /h), cannot
  be reproduced; this is the documented cause of the AUC deviation above.

* **IIV on the covariate slopes.** In a nonparametric fit every one of the four
  quantities in equations e and f is a subject-level parameter, so both the
  intercept and the slope of the clearance and volume relations carry their own
  eta (`etalcl` / `etae_wt_cl` and `etalvc` / `etae_wt_vc`). Correlations among
  the six parameters are not published, so the etas are encoded as independent.

* **Residual error is an assay-precision stand-in, not a fitted estimate.**
  Pmetrics weighted observations by the inverse of the estimated assay variance
  and the paper publishes neither the resulting error polynomial nor its gamma
  or lambda scale factor, so no fitted residual magnitude exists to transcribe.
  `propSd` is set to 0.093 from the reported HPLC assay coefficient of variation
  (Materials and Methods: "The coefficient of variation was < 9.3% over the
  concentration range of 0.05 to 8 mg/liter"). It is an analytical-precision
  figure, excludes model misspecification, and will understate the observed
  scatter in Figure 3. All checks in this vignette are run on `Cc`, the
  individual prediction, so none of them depends on it.

* **Estimated initial condition not encoded.** Materials and Methods describes a
  parameterised non-zero initial condition, switched on by a binary indicator,
  to accommodate patients whose first sample followed an undocumented prior
  dose. That parameter is not reported in Table 2, and the paper's own Monte
  Carlo simulations "set the initial conditions of the central and peripheral
  compartments at a default value of zero". It is a data-handling nuisance
  parameter and is omitted here.

* **Abstract / Table 2 discrepancy on the `K12` SD.** The Abstract prints the
  `K12` standard deviation as 6.67 /h and Table 2 prints 6.76 /h. The parameter
  table is taken as authoritative and `etalk12` is derived from 6.76; the
  difference is a digit transposition and moves the variance by under 3 percent.

* **Screened but rejected covariates.** Cockcroft-Gault eGFR (model 4) and
  ethnicity were tested and not retained. They are recorded in the model file's
  `covariatesDataExcluded` metadata rather than in `covariateData`, since
  `model()` does not reference them.

* **Simulated weights.** Drawn uniformly over the observed 32 to 68 kg range,
  following "simulated weight values were limited to the range observed in our
  clinical cohort". The paper does not state the distribution used within that
  range.

* **Infusion duration.** Patients received the drug over 5 to 6 h; the paper's
  Monte Carlo section specifies 5.5 h, which is used here.

* **Meta-analysis layer not extracted.** The paper's second component is a
  dose-adjusted random-effects meta-regression of CSF sterility and mortality
  across six study arms, of the form `theta_i = beta_0 + beta_1 * dose_i + u_i`.
  Only the residual heterogeneity `tau` values and the P values for the dose
  moderator are reported (Figure 5); `beta_0` and `beta_1` are not published for
  any of the three outcomes, so no model can be encoded from the printed values.
