---
title: "Liposomal amphotericin B (Stott 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Liposomal amphotericin B (Stott 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
ui <- rxode2::rxode(readModelDb("Stott_2022_amphotericinB_liposomal"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1093/jac/dkac389>

Stott 2022 pooled the pharmacokinetic substudies of the Phase II and Phase III
AMBITION-cm trials of high-dose, short-course liposomal amphotericin B
(AmBisome) for HIV-associated cryptococcal meningoencephalitis and fitted the
data with the nonparametric adaptive grid (NPAG) algorithm of Pmetrics 1.9.7. A
two-compartment model with a zero-order infusion into the central compartment
and first-order elimination was retained: a third compartment and
Michaelis-Menten elimination did not improve the fit. No covariate was
retained. The same paper also reports an exposure-toxicity logistic regression
and a meta-analysis of clinical outcomes after AmBisome monotherapy; neither is
encodable from the printed values (see Assumptions and deviations).

**Solve this model with `useLinCmt = FALSE`.** Its `model()` block exposes the
micro-constants `k12` and `k21` (the paper's `KCP` and `KPC`) rather than a
`q` / `vp` pair, so a `linCmt()` conversion has no well-formed two-compartment
macro-parameterisation to build.

# Population

The model was fitted to 87 adults with HIV-associated cryptococcal
meningoencephalitis: 56 from all four arms of the Phase II trial (Princess
Marina Hospital, Gaborone, Botswana and Bugando Medical Centre and Sekou Toure
Hospital, Mwanza, Tanzania; January 2015 to August 2016) and 31 from the
single-dose intervention arm of the Phase III trial (Queen Elizabeth Central
Hospital, Blantyre, Malawi; November 2018 to October 2019). Table 1 gives 55
men and 32 women, a median age of 37 years (IQR 32-43), a median weight of 52.0
kg (IQR 46.5-58.7), a median serum creatinine of 64.0 umol/L (IQR 58.0-87.2)
and a median CD4+ count of 30 cells/mm^3 (IQR 12-60).

AmBisome was infused over 2 h after pre-hydration. The Phase II arms were 3
mg/kg/day for 14 days; 10 mg/kg on day 1 only; 10 mg/kg on day 1 plus 5 mg/kg
on day 3; and 10 mg/kg on day 1 plus 5 mg/kg on days 3 and 7, each with
fluconazole. The Phase III intervention arm was a single 10 mg/kg dose with
flucytosine and fluconazole. Phase II samples were drawn at the end of the
infusion, at 6 h and at 24 h; Phase III samples at 0, 2, 4, 7, 12 and 23 h after
the start of the day-1 infusion and at 2, 4, 7, 12 and 23 h on day 7. The data
set held 565 plasma concentrations (mean 6.5, range 2-12 per patient) of total
amphotericin B (liposome-associated plus free), assayed by UPLC-MS/MS with a
lower limit of quantitation of 0.25 mg/L.

```{r population}
str(ui$population)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Stott_2022_amphotericinB_liposomal.R` carries an
in-file comment pointing at its source location. They are collected here.

| Equation / parameter | Value | Source location |
|---|---|---|
| `d/dt(central)` | n/a | Equation 1: `dX(1)/dt = R(1) - (SCL/V + KCP) * X(1) + KPC * X(2)` |
| `d/dt(peripheral1)` | n/a | Equation 2: `dX(2)/dt = KCP * X(1) - KPC * X(2)` |
| `Cc` | n/a | Equation 3: `Y(1) = X(1) / V` |
| `lcl` | `log(0.416)` | Table 2, `Clearance (L/h)` mean (Abstract: "clearance 0.416 (0.363) L/h") |
| `lvc` | `log(4.566)` | Table 2, `Volume (L)` mean |
| `lk12` | `log(2.222)` | Table 2, `KCP (h-1)` mean |
| `lk21` | `log(2.951)` | Table 2, `KPC (h-1)` mean |
| `etalcl` | `0.566123` | Table 2, `Clearance` SD 0.363; `log(1 + (0.363/0.416)^2)` |
| `etalvc` | `0.682635` | Table 2, `Volume` SD 4.518; `log(1 + (4.518/4.566)^2)` |
| `etalk12` | `1.18612` | Table 2, `KCP` SD 3.351; `log(1 + (3.351/2.222)^2)` |
| `etalk21` | `1.06546` | Table 2, `KPC` SD 4.070; `log(1 + (4.070/2.951)^2)` |
| `addSd` | `0` (fixed) | Results: "additive error was selected for the final model"; the fitted lambda is not printed |
| `propSd` | `0.09` (fixed) | Materials and methods, "Bioanalysis of PK samples": assay CV "<9.0% over the concentration range 0.25-50.0 mg/L" |

The paper's printed constants are restated once here and used from this list
throughout, so that every check below is driven by numbers transcribed from the
paper rather than by the model's own variables.

```{r paper-constants}
paper <- list(
  # Table 2, means (the encoded typical values)
  mean = c(cl = 0.416, vc = 4.566, k12 = 2.222, k21 = 2.951),
  # Table 2, medians (shown for comparison only)
  median = c(cl = 0.345, vc = 3.698, k12 = 0.218, k21 = 0.484),
  # Table 2, SDs
  sd = c(cl = 0.363, vc = 4.518, k12 = 3.351, k21 = 4.070),
  wt_median = 52.0, # Table 1 median weight (kg)
  wt_q1 = 46.5, # Table 1 weight IQR
  wt_q3 = 58.7,
  tinf = 2, # 'administered in a 2 h IV infusion'
  lloq = 0.25 # mg/L
)
```

# Deterministic structural checks

These checks use the typical-value model (`zeroRe()`) and compare it against
arithmetic carried out independently from the printed constants above. They are
deterministic, so they are asserted tightly.

```{r typical-model}
ui_typ <- rxode2::zeroRe(ui)
```

## Whole-ODE check against an independent biexponential

The two-compartment infusion system has a closed-form solution. The function
below is written from the paper's printed micro-constants and is independent of
the packaged model, so agreement exercises the ODE block, the infusion and the
observation equation at once. It also detects a dropped peripheral compartment,
which leaves `Dose / CL` -- and therefore any AUC check -- intact while
changing the shape of the curve.

```{r closed-form}
biexp_conc <- function(times, dose_times, doses, tinf, p) {
  k10 <- p[["cl"]] / p[["vc"]]
  b <- k10 + p[["k12"]] + p[["k21"]]
  alpha <- (b + sqrt(b^2 - 4 * k10 * p[["k21"]])) / 2
  beta <- (b - sqrt(b^2 - 4 * k10 * p[["k21"]])) / 2
  ca <- (alpha - p[["k21"]]) / ((alpha - beta) * alpha)
  cb <- (p[["k21"]] - beta) / ((alpha - beta) * beta)
  # Central amount under an infusion of unit rate started u time units ago and
  # never stopped; a finite infusion is that minus the same thing started tinf
  # later, superposed over the dosing history.
  step <- function(u) {
    ifelse(u <= 0, 0, ca * (1 - exp(-alpha * pmax(u, 0))) + cb * (1 - exp(-beta * pmax(u, 0))))
  }
  amt <- rep(0, length(times))
  for (d in seq_along(dose_times)) {
    rate <- doses[d] / tinf
    amt <- amt + rate * (step(times - dose_times[d]) - step(times - dose_times[d] - tinf))
  }
  list(conc = amt / p[["vc"]], alpha = alpha, beta = beta)
}

# Phase II arm 4 at the median weight: 10 mg/kg day 1, 5 mg/kg days 3 and 7.
dose_times <- c(0, 48, 144)
doses <- c(10, 5, 5) * paper$wt_median
t_grid <- seq(0, 240, by = 0.5)
ref <- biexp_conc(t_grid, dose_times, doses, paper$tinf, paper$mean)

ev_typ <- bind_rows(
  tibble(
    id = 1L, time = dose_times, evid = 1L, amt = doses,
    rate = doses / paper$tinf, cmt = "central"
  ),
  tibble(id = 1L, time = t_grid, evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")
) |>
  arrange(time, desc(evid))

sim_typ <- rxode2::rxSolve(
  ui_typ,
  events = as.data.frame(ev_typ),
  useLinCmt = FALSE,
  rtol = 1e-10, atol = 1e-12,
  returnType = "data.frame"
)

cmp_typ <- tibble(time = t_grid, model = sim_typ$Cc[match(t_grid, sim_typ$time)], closed = ref$conc) |>
  filter(time > 0)
rel_err <- max(abs(cmp_typ$model - cmp_typ$closed) / cmp_typ$closed)

# Two solutions of the same linear ODE at tight solver tolerances. 1e-6 is far
# above the realised error and still goes red on a dropped peripheral
# compartment, a swapped micro-constant or a wrong infusion duration.
stopifnot(rel_err < 1e-6)
rel_err
```

```{r figure-closed-form}
cmp_typ |>
  tidyr::pivot_longer(-time) |>
  mutate(name = recode(name, model = "Packaged model", closed = "Closed-form biexponential")) |>
  ggplot(aes(time, value, colour = name, linetype = name)) +
  geom_line(linewidth = 0.7) +
  labs(
    x = "Time (h)", y = "Amphotericin B (mg/L)", colour = NULL, linetype = NULL,
    title = "Typical-value profile, 10 mg/kg day 1 + 5 mg/kg days 3 and 7 at 52 kg",
    caption = "Packaged model versus a closed-form solution built from the Table 2 means."
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

## Mass balance and terminal half-life after a single 10 mg/kg dose

After a single dose, the area under the curve to infinity equals `Dose / CL`
exactly, and the terminal half-life is `log(2) / beta`.

```{r check-single}
dose_single <- 10 * paper$wt_median
ev_single <- bind_rows(
  tibble(id = 1L, time = 0, evid = 1L, amt = dose_single, rate = dose_single / paper$tinf, cmt = "central"),
  tibble(id = 1L, time = seq(0, 336, by = 0.25), evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")
) |>
  arrange(time, desc(evid))

sim_single <- rxode2::rxSolve(
  ui_typ,
  events = as.data.frame(ev_single),
  useLinCmt = FALSE,
  rtol = 1e-10, atol = 1e-12,
  returnType = "data.frame"
)
stopifnot("peripheral1" %in% names(sim_single))

ref_single <- biexp_conc(0, 0, dose_single, paper$tinf, paper$mean)
t_half_paper <- log(2) / ref_single$beta

auc_0_336 <- sum(diff(sim_single$time) * (head(sim_single$Cc, -1) + tail(sim_single$Cc, -1)) / 2)
auc_inf <- auc_0_336 + tail(sim_single$Cc, 1) / ref_single$beta
auc_closed <- dose_single / paper$mean[["cl"]]

late <- sim_single |> filter(time >= 120, time <= 240)
t_half_model <- log(2) / -coef(stats::lm(log(Cc) ~ time, data = late))[["time"]]

# Deterministic. The trapezoid on a 0.25 h grid is the only approximation in
# the AUC (realised error well under 0.1%); 1% still goes red on a wrong
# clearance or dose unit. The half-life fit is on a pure monoexponential tail.
stopifnot(
  abs(auc_inf - auc_closed) / auc_closed < 0.01,
  abs(t_half_model - t_half_paper) / t_half_paper < 0.01
)

tibble(
  Quantity = c(
    "Dose (mg)", "Dose / CL (mg*h/L)", "Simulated AUC0-inf (mg*h/L)",
    "alpha (1/h)", "beta (1/h)",
    "Terminal half-life, closed form (h)", "Terminal half-life, packaged model (h)"
  ),
  Value = c(dose_single, auc_closed, auc_inf, ref_single$alpha, ref_single$beta, t_half_paper, t_half_model)
) |>
  knitr::kable(digits = 3, caption = "Typical patient (52 kg) after a single 10 mg/kg 2 h infusion.")
```

## Table 2 means versus medians

Table 2 reports both the mean and the median of every parameter's
nonparametric marginal distribution, and Methods says that "both mean and
median parameter values were examined" without stating which set the final
model's population predictions used. The Abstract presents the means as the
population estimates, so the means are encoded. The two sets differ by a factor
of about ten on `KCP` and six on `KPC`, which is a direct sign that the NPAG
distribution of the distribution rate constants is far from log-normal.

The Discussion states that "plasma concentrations fall to negligible levels
48-72 h after a single dose of 10 mg/kg". Both parameter sets satisfy that
statement, so it does not discriminate between them; the visible difference is
in the first hours, where the medians' slow distribution gives a peak almost
twice as high.

```{r mean-vs-median}
mean_median <- bind_rows(
  tibble(set = "Table 2 means (encoded)", time = t_grid, Cc = biexp_conc(t_grid, 0, dose_single, paper$tinf, paper$mean)$conc),
  tibble(set = "Table 2 medians", time = t_grid, Cc = biexp_conc(t_grid, 0, dose_single, paper$tinf, paper$median)$conc)
)

neg <- mean_median |>
  group_by(set) |>
  summarise(
    cmax = max(Cc),
    c48 = Cc[time == 48],
    c72 = Cc[time == 72],
    .groups = "drop"
  ) |>
  mutate(pct72 = 100 * c72 / cmax)

# The Discussion's 'negligible by 48-72 h' claim: under 5% of the peak by 72 h
# for both parameter sets (realised 2.5% and 1.0%). Deterministic.
stopifnot(all(neg$pct72 < 5))

neg |>
  rename(
    "Parameter set" = set, "Cmax (mg/L)" = cmax, "C at 48 h (mg/L)" = c48,
    "C at 72 h (mg/L)" = c72, "C72 as % of Cmax" = pct72
  ) |>
  knitr::kable(digits = 2, caption = "Typical 52 kg patient after a single 10 mg/kg dose, closed form.")

ggplot(mean_median |> filter(time > 0, time <= 168), aes(time, Cc, colour = set)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = paper$lloq, linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Amphotericin B (mg/L)", colour = NULL,
    title = "Single 10 mg/kg dose at 52 kg: Table 2 means versus medians",
    caption = "Dashed line: assay LLOQ 0.25 mg/L."
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the five AmBisome regimens of the two trials at 200 virtual patients per
regimen. Weight drives only the mg/kg dose, since no covariate was retained;
it is drawn from a log-normal matched to the Table 1 median and IQR.

```{r cohort}
# set.seed() seeds R's RNG, which draws the weights below. It does NOT seed
# rxode2's eta sampler, whose streams are partitioned per solver thread, so the
# drawn cohort differs between machines. Every assertion below holds for any
# cohort this model can produce.
set.seed(20221122)

n_arm <- 200L
wt_sdlog <- log(paper$wt_q3 / paper$wt_q1) / (2 * stats::qnorm(0.75))
cohort_wt <- pmin(pmax(stats::rlnorm(n_arm, log(paper$wt_median), wt_sdlog), 30), 100)

regimens <- list(
  "Phase II: 3 mg/kg/day x 14 days" = tibble(day = 1:14, mgkg = 3),
  "Phase II: 10 mg/kg day 1" = tibble(day = 1, mgkg = 10),
  "Phase II: 10 mg/kg day 1 + 5 mg/kg day 3" = tibble(day = c(1, 3), mgkg = c(10, 5)),
  "Phase II: 10 mg/kg day 1 + 5 mg/kg days 3, 7" = tibble(day = c(1, 3, 7), mgkg = c(10, 5, 5)),
  "Phase III: 10 mg/kg day 1" = tibble(day = 1, mgkg = 10)
)

make_arm <- function(i) {
  label <- names(regimens)[i]
  subj <- tibble(id = (i - 1L) * n_arm + seq_len(n_arm), WT = cohort_wt, treatment = label)
  dosing <- subj |>
    tidyr::crossing(regimens[[i]]) |>
    mutate(
      time = 24 * (day - 1), evid = 1L, amt = mgkg * WT,
      rate = amt / paper$tinf, cmt = "central"
    ) |>
    select(-day, -mgkg)
  obs <- subj |>
    tidyr::crossing(time = seq(0, 168, by = 1)) |>
    mutate(evid = 0L, amt = NA_real_, rate = NA_real_, cmt = "central")
  bind_rows(dosing, obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, evid, amt, rate, cmt, WT, treatment)
}

events <- bind_rows(lapply(seq_along(regimens), make_arm))
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

```{r simulate}
sim <- rxode2::rxSolve(
  ui,
  events = as.data.frame(events),
  keep = c("WT", "treatment"),
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  mutate(treatment = factor(treatment, levels = names(regimens)))

stopifnot(
  dplyr::n_distinct(sim$id) == length(regimens) * n_arm,
  !anyNA(sim$Cc),
  all(sim$Cc >= -1e-6 * max(sim$Cc))
)
```

# Replicate published figures

Stott 2022 publishes only goodness-of-fit plots (Figure 1: observed versus
population and individual predicted concentrations, regression r^2 0.52 and
0.90), which cannot be reproduced without the observed data. The simulated
concentration-time profiles below are therefore companions rather than
replications.

```{r figure-profiles, fig.height = 6}
sim |>
  filter(time > 0) |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2) +
  geom_line(aes(y = Q50), linewidth = 0.6) +
  geom_hline(yintercept = paper$lloq, linetype = "dashed") +
  facet_wrap(~treatment, ncol = 2) +
  scale_y_log10() +
  coord_cartesian(ylim = c(0.01, NA)) +
  labs(
    x = "Time (h)", y = "Amphotericin B (mg/L)",
    title = "Simulated median and 90% interval by regimen",
    caption = "Dashed line: assay LLOQ 0.25 mg/L. Values below 0.01 mg/L are not drawn."
  ) +
  theme_bw()
```

# PKNCA validation

Stott 2022 computed AUC0-24, AUC0-168 and the first-week Cmax from Bayesian
posterior predictions for its exposure-toxicity analysis, but does not print
their values, so there is no published NCA to compare against. The block below
reports the same metrics for the simulated cohort and checks the single-dose
arms against the per-subject identity `AUC0-inf = Dose / CL`.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject; pre-dose concentration is zero.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca),
  Cc ~ time | treatment + id,
  concu = "mg/L", timeu = "h"
)

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = 0,
  end = c(24, 168),
  auclast = TRUE,
  cmax = c(FALSE, TRUE),
  tmax = c(FALSE, TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_res$result |>
  filter(PPTESTCD %in% c("auclast", "cmax", "tmax")) |>
  mutate(PPTESTCD = ifelse(PPTESTCD == "auclast", paste0("auc0-", end), PPTESTCD)) |>
  group_by(treatment, PPTESTCD) |>
  summarise(
    median = median(PPORRES),
    q25 = quantile(PPORRES, 0.25),
    q75 = quantile(PPORRES, 0.75),
    .groups = "drop"
  ) |>
  rename(
    "Regimen" = treatment, "NCA parameter" = PPTESTCD,
    "Median" = median, "P25" = q25, "P75" = q75
  ) |>
  knitr::kable(digits = 2, caption = "Simulated NCA by regimen: AUC0-24 and AUC0-168 (mg*h/L), first-week Cmax (mg/L) and tmax (h).")
```

```{r pknca-single}
single_arms <- c("Phase II: 10 mg/kg day 1", "Phase III: 10 mg/kg day 1")

intervals_single <- data.frame(start = 0, end = Inf, aucinf.obs = TRUE, half.life = TRUE)
conc_single <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca |> filter(treatment %in% single_arms)),
  Cc ~ time | treatment + id,
  concu = "mg/L", timeu = "h"
)
dose_single_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_df |> filter(treatment %in% single_arms)),
  amt ~ time | treatment + id,
  doseu = "mg"
)
nca_single <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_single, dose_single_obj, intervals = intervals_single))

chk <- nca_single$result |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(id, aucinf = PPORRES) |>
  inner_join(
    sim |> distinct(id, cl, treatment),
    by = "id"
  ) |>
  inner_join(dose_df |> select(id, amt), by = "id") |>
  mutate(pct_diff = 100 * (aucinf - amt / cl) / (amt / cl)) |>
  filter(!is.na(pct_diff))

# AUC0-inf from PKNCA against each subject's own Dose / CL. The two sides use
# the same drawn parameters, so the centre is numerical error only; the tail
# carries the extrapolation error of subjects whose sampled k21 is small enough
# that the 168 h window does not reach the terminal phase.
stopifnot(
  nrow(chk) > 0.9 * length(single_arms) * n_arm,
  abs(median(chk$pct_diff)) < 2,
  quantile(abs(chk$pct_diff), 0.9) < 10
)

chk |>
  group_by(treatment) |>
  summarise(
    "Subjects" = n(),
    "Median AUC0-inf (mg*h/L)" = median(aucinf),
    "Median Dose/CL (mg*h/L)" = median(amt / cl),
    "Median % difference" = median(pct_diff),
    "90th percentile |% difference|" = quantile(abs(pct_diff), 0.9),
    .groups = "drop"
  ) |>
  rename("Regimen" = treatment) |>
  knitr::kable(digits = 2, caption = "Single-dose arms: PKNCA AUC0-inf against each subject's Dose / CL.")
```

# Assumptions and deviations

* **Nonparametric distributions approximated as log-normal.** Pmetrics NPAG
  estimates a discrete joint distribution over support points and Stott 2022
  publishes only each marginal's mean, median and SD (Table 2). Each parameter
  is encoded as a log-normal whose median is the reported mean and whose
  variance is `log(1 + (SD / mean)^2)` (CVs of 87, 99, 151 and 138% for
  clearance, volume, `KCP` and `KPC`). No correlations are published, so the
  etas are independent. The nonparametric shape cannot be reproduced: for a
  log-normal the mean-to-median ratio is `exp(omega^2 / 2)`, which is 1.33 and
  1.41 for clearance and volume against the reported 1.21 and 1.23, and 1.81
  and 1.70 for `KCP` and `KPC` against the reported 10.2 and 6.1. Simulated
  distribution-phase behaviour is therefore the least reliable part of the
  stochastic model.

* **Means used as the typical values.** Methods says both mean and median
  parameter values were examined, without stating which the final model's
  population predictions used; the Abstract reports the means as the
  "population PK parameter estimates". The means are encoded, as in the same
  group's amphotericin B deoxycholate model (`Stott_2018_amphotericinB_deoxycholate`),
  and the medians are compared above. The Discussion's single-dose washout
  statement holds for either set.

* **Residual error is an assay-precision stand-in, not a fitted estimate.**
  Results states that additive error (the Pmetrics `lambda` term, combined in
  quadrature with the assay error polynomial) was selected, but neither the
  assay polynomial nor the fitted lambda is printed. `addSd` (lambda) is
  therefore fixed to 0 and `propSd` to the reported assay coefficient of
  variation of 9% (Materials and methods: "The coefficient of variation was
  <9.0% over the concentration range 0.25-50.0 mg/L"). The encoded residual
  understates the observation scatter. Every check in this vignette is run on
  `Cc`, the individual prediction, so none depends on it.

* **No covariates.** Age, sex, weight, baseline serum creatinine and CD4+ count
  were screened by stepwise multivariate linear regression on the Bayesian
  posterior parameters and none was retained. They are recorded in the model
  file's `covariatesDataExcluded` metadata. Weight enters the simulation only
  through the mg/kg dose.

* **Typographical errors in the source.** Table 1 gives serum creatinine in
  "mmol/L"; a median of 64.0 is only possible in umol/L, the unit the Toxicity
  section uses. Results describes the Figure 1 regression as being of
  "observed-versus-predicted concentrations of DAmB", where the analyte is
  liposomal amphotericin B. Neither affects a model value.

* **Simulated weights.** Table 1 reports the weight median and IQR only; the
  virtual cohort draws a log-normal with median 52.0 kg and the IQR-matched
  log-SD, truncated to 30-100 kg.

* **Exposure-toxicity and meta-analysis layers not extracted.** The
  exposure-toxicity logistic regressions report only odds ratios of 1.0 per unit
  AUC (Table 3) with no intercepts, and the random-effects meta-analysis of
  AmBisome monotherapy outcomes reports only residual heterogeneity (`tau^2`
  0.0023 falling to 0.0003 with moderators) and P values, not the moderator
  coefficients. Neither can be encoded from the printed values.

* **Errata.** No erratum or correction was found for this article (Europe PMC
  search, October 2026).
