---
title: "Sipoglitazar (Stringer 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sipoglitazar (Stringer 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Stringer_2014_sipoglitazar"))$reference`
* Description: `r rxode2::rxode(readModelDb("Stringer_2014_sipoglitazar"))$description`
* Article: <https://doi.org/10.1002/jcph.227>

Sipoglitazar is an oral PPAR alpha/delta/gamma agonist that is cleared by
UDP-glucuronosyltransferase 2B15. The `UGT2B15*2` allele reduces
glucuronidation, so `UGT2B15*2/*2` subjects clear the drug about 3.3-fold more
slowly than `UGT2B15*1/*1` subjects and reach correspondingly higher exposure.
The question this paper asks is whether that pharmacokinetic difference is
large enough to matter clinically -- i.e. whether it propagates into the
glycemic response -- and whether genotype-guided dose selection can equalise it.

The model is **pharmacodynamic only**. No drug is dosed through the event
table and there is no plasma concentration state: exposure enters as a
steady-state `AUC0-24h` computed from the subject's total daily dose and
genotype-specific clearance. Consequently there is nothing for a
non-compartmental analysis to integrate, and this vignette uses the
turnover-model validation strategy (steady-state identities, closed-form
checks, perturbation recovery, and replication of the paper's own published
simulations) rather than a PKNCA section.

```{r load-model}
# Build the two model objects from INDEPENDENT rxode() calls. rxode2's
# zeroRe() modifies the object it is handed, so `mod_typ <- zeroRe(mod_iiv)`
# would silently zero the omega of `mod_iiv` as well and every "with IIV"
# simulation below would quietly run at the typical value.
mod_iiv <- rxode2::rxode(readModelDb("Stringer_2014_sipoglitazar"))
mod_typ <- rxode2::zeroRe(rxode2::rxode(readModelDb("Stringer_2014_sipoglitazar")))
```

# Population

Two 13-week Phase II randomized, double-blind trials in **780 drug-naive
adults with type 2 diabetes mellitus** were pooled (sipoglitazar n = 572,
rosiglitazone n = 72, placebo n = 136). Entry required a diagnosis of type 2
diabetes with no prior anti-diabetic medication, a screening HbA1c above 7.0%
and below 10.0%, and age above 35 and below 75 years.

Baseline characteristics (supplement Table S1): age median 56 years (range
34-75); 388 male / 392 female; body weight median 88.8 kg (range 55-160);
duration of disease median 1.0 year (range 0-30.9); FPG baseline median
9.3 mmol/L (range 2.9-20.8); HbA1c baseline median 7.9% (range 6.9-9.9).
UGT2B15 genotype was `*1/*1` : `*1/*2` : `*2/*2` = 149 : 357 : 194 among the
700 genotyped subjects (21% : 51% : 28%); genotype was not collected in the
remaining 80 subjects (10%), who were carried with a population-average
clearance.

FPG and HbA1c were sampled at -1, 0, 2, 4, 6, 8, 10 and 12 weeks. All subjects
received dietary advice for the whole trial, which is what the model's LEFPG
and LEHB "lifestyle effect" parameters absorb.

The same information is available programmatically via
`readModelDb("Stringer_2014_sipoglitazar")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Stringer_2014_sipoglitazar.R`. The tables
below collect them in one place for review.

## Equations

| Model element | Source location |
|---------------|-----------------|
| `DEF = Emax * AUC / (AUC50 + AUC)` | Equation (1), p. 454 |
| `LEHB_active = LEHB_placebo - LEHB_factor` | Equation (2), p. 455 |
| `dFPG/dt = KinG * (1 +/- LEFPG) - KoutG * (1 + DEF + ROTE) * FPG` | Equation (3), p. 455 (the paper really does typeset a plus-minus sign here; see Assumptions) |
| `dHbA1c/dt = KinH * (1 - LEHB) * FPG^gamma - KoutH * HbA1c` | Equation (4), p. 455 |
| `AUC = total daily dose / CL` at steady state | Methods "Population Data Analysis", p. 454 |
| `KinG = KoutG * BSL_FPG` (baseline pinning) | Implied by the indirect-response parameterisation; `KinG` is not tabulated |
| `KinH = KoutH * BSL_HbA1c / BSL_FPG^gamma` (baseline pinning) | Implied by the indirect-response parameterisation; `KinH` is not tabulated |
| Box-Cox eta transform `(exp(lambda * eta) - 1) / lambda` | Methods "Intra-Individual Variability", citing Petersson 2009 (reference 18) |

## Parameters

| Parameter (file) | Paper symbol | Value | Source location |
|------------------|--------------|-------|-----------------|
| `lrbase_fpg` | BSL FPG | 9.41 mmol/L | Table 2 |
| `lrbase_fpg_bid32` | BSL FPG (32 mg BID) | 9.02 mmol/L | Table 2, footnote a |
| `lkout_fpg` | KoutG | 0.027 1/day | Table 2 |
| `lrbase_hba1c` | BSL HbA1c | 7.96 % | Table 2 |
| `lkout_hba1c` | KoutH | 0.031 1/day | Table 2 |
| `lgamma` | Gamma (printed as lambda in Eq. 4) | 0.71 | Table 2 |
| `lemax` | Emax | 48.7 % -> 0.487 fraction | Table 2 |
| `lauc50` | AUC50 | 1.15 mg*day/L | Table 2 |
| `e_rosi_kout_fpg` | ROTE | 0.28 | Table 2 |
| `e_lifestyle_kin_fpg` | LEFPG | 0 (fixed) | Table 2, footnote b |
| `e_placebo_kin_hba1c` | LEHB placebo | 0.037 | Table 2 |
| `e_active_kin_hba1c` | LEHB factor | 0.017 | Table 2, footnote c / Eq. (2) |
| `boxcox_rbase_hba1c` | Box-Cox | 2.39 | Table 2 |
| `etalrbase_fpg` | omega^2 BSL FPG | 0.05 | Table 2 |
| `etalrbase_hba1c` | omega^2 BSL HbA1c | 0.01 | Table 2 |
| `etae_lifestyle_kin_fpg` | omega^2 LEFPG | 0.04 | Table 2 |
| `etae_placebo_kin_hba1c` | omega^2 LEHB | 0.01 | Table 2 |
| `propSd_glucose` | Residual error FPG | 10.6 % -> 0.106 | Table 2 |
| `propSd_hba1c` | Residual error HbA1c | 2.0 % -> 0.020 | Table 2 |
| `lcl_ugt2b15_s1s1` | CL, `UGT2B15*1/*1` | 5.04 L/h (fixed) | Table 1 |
| `lcl_ugt2b15_s1s2` | CL, `UGT2B15*1/*2` | 3.35 L/h (fixed) | Table 1 |
| `lcl_ugt2b15_s2s2` | CL, `UGT2B15*2/*2` | 1.53 L/h (fixed) | Table 1 |
| `lcl_ugt2b15_missing` | population-average CL | 3.21 L/h (**derived**, see Errata) | Not printed; derived from Table 1 + Table S1 |

## Dimensional analysis

Time is **days** throughout, because `KoutG` and `KoutH` are reported in
`days^-1` and `AUC50` in `mg*day/L`. Both ODE states are carried on the
measured scale (a concentration and a percentage), not as amounts, which is the
standard indirect-response parameterisation.

| Term | Units | Product |
|------|-------|---------|
| `KinG = KoutG * BSL_FPG` | `(1/day) * (mmol/L)` | `mmol/(L*day)` |
| `KoutG * (1 + DEF + ROTE) * FPG` | `(1/day) * (unitless) * (mmol/L)` | `mmol/(L*day)` |
| `AUC = dose / (CL * 24)` | `mg / ((L/h) * (h/day))` | `mg*day/L`, matching AUC50 |
| `DEF = Emax * AUC / (AUC50 + AUC)` | `(unitless) * (mg*day/L)/(mg*day/L)` | unitless |
| `KinH = KoutH * BSL_HbA1c / BSL_FPG^gamma` | `(1/day) * % / (mmol/L)^0.71` | `% * (L/mmol)^0.71 / day` |
| `KinH * (1 - LEHB) * FPG^gamma` | `% * (L/mmol)^0.71/day * (mmol/L)^0.71` | `%/day` |
| `KoutH * HbA1c` | `(1/day) * %` | `%/day` |

Both ODE right-hand sides reduce to `[state]/day`, and the `mmol/L` raised to
the fractional power `gamma` cancels exactly between `KinH` and the `FPG^gamma`
driver, so the fractional-power term introduces no residual unit. The one
non-obvious conversion is the `* 24` that turns the Table 1 clearances (L/h)
into L/day; without it, `AUC` would be 24-fold too large and every drug effect
would saturate.

# Structural checks

## Placebo FPG holds exactly at baseline

The paper states that "the FPG data in the placebo group on average showed no
change over time" -- LEFPG has a population value of exactly zero. With the
random effects zeroed, the placebo FPG trajectory must therefore be a flat line
at 9.41 mmol/L to solver precision. This check also confirms that `zeroRe()`
actually took effect (any leaked IIV would move the trajectory).

```{r placebo-hold}
ev_placebo <- data.frame(
  id = 1L, time = seq(0, 395, by = 5), amt = NA_real_, evid = 0L,
  cmt = "glucose", dvid = 1L,
  TRT = 0, DOSE_SIPOGLITAZAR_MGD = 0,
  UGT2B15_STAR2_HET = 0, UGT2B15_STAR2_HOM = 0, UGT2B15_MISSING = 1
)
s_placebo <- rxode2::rxSolve(mod_typ, ev_placebo, useLinCmt = FALSE) |>
  as.data.frame()

stopifnot(max(abs(s_placebo$glucose - 9.41)) < 1e-6)
```

Placebo HbA1c, by contrast, must decline: LEHB inhibits `KinH` by 3.7% in the
placebo arm while FPG stays at its baseline, so the closed-form steady state is
`BSL_HbA1c * (1 - LEHB_placebo)` exactly.

```{r placebo-hba1c}
hba1c_placebo_ss <- 7.96 * (1 - 0.037)
c(closed_form = hba1c_placebo_ss, simulated = tail(s_placebo$hba1c, 1))
stopifnot(abs(tail(s_placebo$hba1c, 1) - hba1c_placebo_ss) < 1e-3)
```

## Closed-form steady state across a dose / genotype ladder

Setting both derivatives to zero gives

```
FPG_ss   = BSL_FPG / (1 + DEF + ROTE)
HbA1c_ss = BSL_HbA1c * (1 - LEHB) * (FPG_ss / BSL_FPG)^gamma
```

Because both sides of this comparison use the same parameter draw, the only
difference is numerical integration error, so a tight bound is the correct
assertion here.

```{r closed-form}
genotypes <- tibble::tibble(
  genotype = c("*1/*1", "*1/*2", "*2/*2"),
  cl_lh    = c(5.04, 3.35, 1.53),
  UGT2B15_STAR2_HET = c(0, 1, 0),
  UGT2B15_STAR2_HOM = c(0, 0, 1)
)

ladder <- tidyr::crossing(genotypes, dose_mgd = c(8, 16, 32, 64, 96)) |>
  dplyr::mutate(
    id = dplyr::row_number(),
    TRT = 1, DOSE_SIPOGLITAZAR_MGD = dose_mgd, UGT2B15_MISSING = 0,
    auc  = dose_mgd / (cl_lh * 24),
    def  = 0.487 * auc / (1.15 + auc),
    lehb = 0.037 - 0.017,
    fpg_cf   = 9.41 / (1 + def),
    hba1c_cf = 7.96 * (1 - lehb) * (fpg_cf / 9.41)^0.71
  )

ev_ladder <- ladder |>
  dplyr::select(id, TRT, DOSE_SIPOGLITAZAR_MGD,
                UGT2B15_STAR2_HET, UGT2B15_STAR2_HOM, UGT2B15_MISSING) |>
  tidyr::crossing(time = c(0, 2000)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "glucose", dvid = 1L)

ss <- rxode2::rxSolve(mod_typ, ev_ladder, useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(time == 2000) |>
  dplyr::select(id, fpg_sim = glucose, hba1c_sim = hba1c) |>
  dplyr::left_join(dplyr::select(ladder, id, genotype, dose_mgd, fpg_cf, hba1c_cf),
                   by = "id") |>
  dplyr::mutate(
    fpg_relerr   = abs(fpg_sim   - fpg_cf)   / fpg_cf,
    hba1c_relerr = abs(hba1c_sim - hba1c_cf) / hba1c_cf
  )

stopifnot(max(ss$fpg_relerr) < 1e-5, max(ss$hba1c_relerr) < 1e-5)

ss |>
  dplyr::transmute(
    Genotype = genotype, `Dose (mg/day)` = dose_mgd,
    `FPG closed form` = round(fpg_cf, 3), `FPG solved` = round(fpg_sim, 3),
    `HbA1c closed form` = round(hba1c_cf, 3), `HbA1c solved` = round(hba1c_sim, 3)
  ) |>
  knitr::kable(caption = "Solved steady state against the closed form implied by Equations (3) and (4).")
```

## Perturbation recovery

Displacing both states away from baseline and running forward with no treatment
effect must return them to the same attractor, which confirms that the
production terms are pinned to the baselines rather than to the initial
conditions.

```{r perturbation}
recover <- lapply(c(0.5, 1.0, 2.0), function(f) {
  s <- rxode2::rxSolve(
    mod_typ, ev_placebo, useLinCmt = FALSE,
    inits = c(glucose = f * 9.41, hba1c = f * 7.96)
  ) |> as.data.frame()
  data.frame(start = f, fpg_end = tail(s$glucose, 1), hba1c_end = tail(s$hba1c, 1))
}) |> dplyr::bind_rows()

recover
stopifnot(
  max(abs(recover$fpg_end - 9.41)) < 1e-3,
  max(abs(recover$hba1c_end - hba1c_placebo_ss)) < 1e-3
)
```

# Replication of the published simulations

## Change from baseline at 6 months (Results, p. 457)

The paper reports simulated median changes from baseline at 6 months for
sipoglitazar 64 mg/day in each genotype and for rosiglitazone 8 mg QD. Because
the population median of every random effect in this model is its typical value
(the Box-Cox transform maps `eta = 0` to `eta_bc = 0`, so the HbA1c baseline
median is preserved at 7.96%), the typical-value trajectory is the right
comparator for a published median.

```{r six-month}
arms <- tibble::tibble(
  arm = c("Sipoglitazar 64 mg, *1/*1", "Sipoglitazar 64 mg, *1/*2",
          "Sipoglitazar 64 mg, *2/*2", "Rosiglitazone 8 mg QD"),
  id  = 1:4,
  TRT = c(1, 1, 1, 3),
  DOSE_SIPOGLITAZAR_MGD = c(64, 64, 64, 0),
  UGT2B15_STAR2_HET = c(0, 1, 0, 0),
  UGT2B15_STAR2_HOM = c(0, 0, 1, 0),
  UGT2B15_MISSING   = c(0, 0, 0, 0),
  # Results p. 457 and Discussion p. 460.
  paper_dfpg   = c(-1.2, -1.6, -2.1, -2.0),
  paper_dhba1c = c(-0.9, -1.1, -1.4, -1.2)
)

ev_arms <- arms |>
  dplyr::select(id, TRT, DOSE_SIPOGLITAZAR_MGD,
                UGT2B15_STAR2_HET, UGT2B15_STAR2_HOM, UGT2B15_MISSING) |>
  tidyr::crossing(time = c(0, 182.6)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "glucose", dvid = 1L)

six <- rxode2::rxSolve(mod_typ, ev_arms, useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(time == 182.6) |>
  dplyr::select(id, glucose, hba1c) |>
  dplyr::left_join(arms, by = "id") |>
  dplyr::mutate(
    model_dfpg   = glucose - 9.41,
    model_dhba1c = hba1c   - 7.96
  )

six |>
  dplyr::transmute(
    Arm = arm,
    `Model dFPG (mmol/L)` = round(model_dfpg, 2),
    `Paper dFPG (mmol/L)` = paper_dfpg,
    `Model dHbA1c (%)`    = round(model_dhba1c, 2),
    `Paper dHbA1c (%)`    = paper_dhba1c
  ) |>
  knitr::kable(caption = "Simulated change from baseline at 6 months vs Stringer 2014 Results (p. 457) and Discussion (p. 460).")

# The paper reports to one decimal, so agreement to within half a reporting
# unit is the tightest bound the answer key can support.
stopifnot(
  max(abs(six$model_dfpg   - six$paper_dfpg))   < 0.06,
  max(abs(six$model_dhba1c - six$paper_dhba1c)) < 0.06
)
```

All eight published values are reproduced to within 0.06 units. That is a
strong joint test: it simultaneously pins the `Emax` percent-to-fraction
rescaling, the L/h to L/day clearance conversion, the total-daily-dose basis of
`AUC`, the value of `gamma`, the LEHB reading of Equation (2), and the absence
of a lifestyle effect on the rosiglitazone arm. Getting any one of those wrong
moves at least one row well outside the bound.

## Figure 2: simulated FPG and HbA1c over one year

```{r figure2, fig.cap = "Replicates Figure 2a and 2b of Stringer 2014: simulated FPG and HbA1c over time for sipoglitazar 64 mg by genotype and for rosiglitazone 8 mg QD."}
ev_traj <- arms |>
  dplyr::select(id, TRT, DOSE_SIPOGLITAZAR_MGD,
                UGT2B15_STAR2_HET, UGT2B15_STAR2_HOM, UGT2B15_MISSING) |>
  tidyr::crossing(time = seq(0, 395, by = 5)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "glucose", dvid = 1L)

traj <- rxode2::rxSolve(mod_typ, ev_traj, useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::left_join(dplyr::select(arms, id, arm), by = "id") |>
  dplyr::mutate(months = time / 30.44) |>
  tidyr::pivot_longer(c(glucose, hba1c), names_to = "endpoint", values_to = "value") |>
  dplyr::mutate(endpoint = dplyr::recode(endpoint,
                                         glucose = "FPG (mmol/L)",
                                         hba1c   = "HbA1c (%)"))

ggplot(traj, aes(months, value, colour = arm, linetype = arm)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(x = "Time (months)", y = NULL, colour = NULL, linetype = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The one-year asymptotes are FPG 8.16 / 7.85 / 7.28 mmol/L for `*1/*1` /
`*1/*2` / `*2/*2` at 64 mg and 7.35 mmol/L for rosiglitazone, matching the
plateaus drawn in Figure 2a.

## Time to 90% of steady state

The paper reports that 90% of steady state is reached at approximately
2.3 months for FPG and 3.7 months for HbA1c.

```{r t90}
t90_of <- function(x, times) {
  target <- x[1] + 0.9 * (tail(x, 1) - x[1])
  approx(x, times, xout = target)$y / 30.44
}

fine <- rxode2::rxSolve(
  mod_typ,
  arms |>
    dplyr::select(id, TRT, DOSE_SIPOGLITAZAR_MGD,
                  UGT2B15_STAR2_HET, UGT2B15_STAR2_HOM, UGT2B15_MISSING) |>
    tidyr::crossing(time = seq(0, 600, by = 1)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "glucose", dvid = 1L),
  useLinCmt = FALSE
) |> as.data.frame()

t90 <- fine |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    t90_fpg   = t90_of(glucose, time),
    t90_hba1c = t90_of(hba1c, time),
    .groups = "drop"
  ) |>
  dplyr::left_join(dplyr::select(arms, id, arm), by = "id")

t90 |>
  dplyr::transmute(Arm = arm,
                   `t90 FPG (months)`   = round(t90_fpg, 2),
                   `t90 HbA1c (months)` = round(t90_hba1c, 2)) |>
  knitr::kable(caption = "Time to 90% of steady state; Stringer 2014 reports approximately 2.3 months (FPG) and 3.7 months (HbA1c).")

# Across the four simulated arms the paper quotes a single rounded value for
# each endpoint, so assert on the arm-wise mean rather than any one arm.
stopifnot(
  abs(mean(t90$t90_fpg)   - 2.3) < 0.4,
  abs(mean(t90$t90_hba1c) - 3.7) < 0.4
)
```

## Figure 4: genotype-based versus titration-based dosing

Under titration-based dosing every subject starts at 32 mg/day; FPG is assessed
every two months and the dose is stepped up. The paper's simulated schedule is
32 -> 64 mg at 2 months -> 96 mg at 4 months for `UGT2B15*1/*1`, and a single
step 32 -> 64 mg at 2 months for `UGT2B15*1/*2`. Genotype-based dosing (Design
C) starts those subjects at 96 and 64 mg/day respectively. The daily dose is a
time-varying covariate, so the titration schedule is carried on the event
table.

```{r titration}
titration_dose <- function(genotype, time) {
  if (genotype == "*1/*1") {
    ifelse(time < 60.9, 32, ifelse(time < 121.8, 64, 96))
  } else {
    ifelse(time < 60.9, 32, 64)
  }
}

design_c <- c("*1/*1" = 96, "*1/*2" = 64)

ev_titr <- lapply(seq_along(design_c), function(k) {
  g <- names(design_c)[k]
  het <- as.integer(g == "*1/*2")
  tt <- sort(unique(c(seq(0, 395, by = 5), 60.9, 121.8, 182.6)))
  dplyr::bind_rows(
    data.frame(id = 10L + k, strategy = "Titration", genotype = g, time = tt,
               DOSE_SIPOGLITAZAR_MGD = titration_dose(g, tt)),
    data.frame(id = 20L + k, strategy = "Genotype (Design C)", genotype = g, time = tt,
               DOSE_SIPOGLITAZAR_MGD = design_c[[k]])
  ) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "glucose", dvid = 1L,
                  TRT = 1, UGT2B15_STAR2_HET = het,
                  UGT2B15_STAR2_HOM = 0, UGT2B15_MISSING = 0)
}) |> dplyr::bind_rows()

titr <- rxode2::rxSolve(
  mod_typ,
  dplyr::select(ev_titr, id, time, amt, evid, cmt, dvid, TRT,
                DOSE_SIPOGLITAZAR_MGD, UGT2B15_STAR2_HET,
                UGT2B15_STAR2_HOM, UGT2B15_MISSING),
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::left_join(
    dplyr::distinct(dplyr::select(ev_titr, id, strategy, genotype)),
    by = "id"
  )

titr_2mo <- titr |>
  dplyr::filter(abs(time - 60.9) < 1e-6) |>
  dplyr::transmute(Genotype = genotype, Strategy = strategy,
                   `Model FPG at 2 months (mmol/L)` = round(glucose, 2)) |>
  dplyr::arrange(Genotype, Strategy)

# Stringer 2014 Results p. 457 / p. 458: for *1/*1, FPG at 2 months was 8.8
# (titration) and 8.1 (genotype); for *1/*2, 8.6 and 8.1.
titr_2mo$`Paper FPG at 2 months (mmol/L)` <- c(8.1, 8.8, 8.1, 8.6)
knitr::kable(titr_2mo, caption = "FPG at 2 months under genotype-based (Design C) and titration-based dosing, against Stringer 2014.")

stopifnot(
  max(abs(titr_2mo$`Model FPG at 2 months (mmol/L)` -
          titr_2mo$`Paper FPG at 2 months (mmol/L)`)) < 0.08
)
```

```{r figure4, fig.cap = "Replicates Figure 4a and 4b of Stringer 2014: FPG and HbA1c under genotype-based (Design C) and titration-based dosing for UGT2B15*1/*1 and *1/*2."}
titr |>
  tidyr::pivot_longer(c(glucose, hba1c), names_to = "endpoint", values_to = "value") |>
  dplyr::mutate(
    months = time / 30.44,
    endpoint = dplyr::recode(endpoint, glucose = "FPG (mmol/L)", hba1c = "HbA1c (%)")
  ) |>
  ggplot(aes(months, value, colour = strategy)) +
  geom_line(linewidth = 0.7) +
  facet_grid(endpoint ~ paste("UGT2B15", genotype), scales = "free_y") +
  labs(x = "Time (months)", y = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The genotype-based curves fall faster over the first six months and reach their
plateau earlier, which is the paper's central operational claim: genotype-guided
starting doses buy time to glycemic control, while the eventual magnitude of the
FPG reduction is the same under both strategies.

```{r titration-plateau}
plateau <- titr |>
  dplyr::filter(time == 395) |>
  dplyr::select(genotype, strategy, glucose) |>
  tidyr::pivot_wider(names_from = strategy, values_from = glucose)

plateau
# "ultimately the magnitude of FPG reduction achieved for the two approaches
# would be expected to be the same" (Discussion, p. 460) -- for *1/*1 both
# strategies converge on 96 mg/day, so the plateaus must coincide exactly.
stopifnot(
  abs(plateau$Titration[plateau$genotype == "*1/*1"] -
      plateau$`Genotype (Design C)`[plateau$genotype == "*1/*1"]) < 1e-3
)
```

# Genotype-guided dose selection and responder rates

The paper's clinical-trial simulation defines a responder as a subject whose
HbA1c falls by more than 0.7% at 6 months, and compares three genotype-based
dose designs against a single dose given to everyone.

Each arm is simulated with **common random numbers**: the seed is reset before
every arm, so the same 100 virtual subjects (the same draws of the baseline and
lifestyle random effects) are carried through every dose and genotype cell.
Comparisons across cells are therefore paired, and the dose- and
exposure-ordering of the responder rate becomes an exact property of the model
rather than something that has to survive Monte Carlo noise.

```{r responders}
n_per_arm <- 100L
dose_grid <- c(8, 16, 32, 64, 96)

sim_arm <- function(het, hom, dose) {
  rxode2::rxSetSeed(20140401)   # common random numbers across every arm
  ev <- data.frame(
    id   = rep(seq_len(n_per_arm), each = 2L),
    time = rep(c(0, 182.6), times = n_per_arm),
    amt = NA_real_, evid = 0L, cmt = "glucose", dvid = 1L,
    TRT = 1, DOSE_SIPOGLITAZAR_MGD = dose,
    UGT2B15_STAR2_HET = het, UGT2B15_STAR2_HOM = hom, UGT2B15_MISSING = 0
  )
  s <- rxode2::rxSolve(mod_iiv, ev, useLinCmt = FALSE) |> as.data.frame()
  base <- s$hba1c[s$time == 0]
  post <- s$hba1c[s$time == 182.6]
  100 * mean(post - base < -0.7)
}

resp <- tidyr::crossing(
  genotypes |> dplyr::select(genotype, UGT2B15_STAR2_HET, UGT2B15_STAR2_HOM),
  dose_mgd = dose_grid
) |>
  dplyr::rowwise() |>
  dplyr::mutate(pct_responder = sim_arm(UGT2B15_STAR2_HET,
                                        UGT2B15_STAR2_HOM, dose_mgd)) |>
  dplyr::ungroup() |>
  dplyr::select(genotype, dose_mgd, pct_responder)

resp |>
  tidyr::pivot_wider(names_from = genotype, values_from = pct_responder) |>
  dplyr::rename(`Dose (mg/day)` = dose_mgd) |>
  knitr::kable(caption = "Percent of simulated subjects achieving an HbA1c reduction greater than 0.7% at 6 months, by dose and genotype (replicates Figure 3 of Stringer 2014).")
```

Reading the three published designs out of that grid:

```{r designs}
designs <- tibble::tribble(
  ~design,    ~`*1/*1`, ~`*1/*2`, ~`*2/*2`,
  "Design A",       32,       16,        8,
  "Design B",       64,       32,       16,
  "Design C",       96,       64,       32
) |>
  tidyr::pivot_longer(-design, names_to = "genotype", values_to = "dose_mgd") |>
  dplyr::left_join(resp, by = c("genotype", "dose_mgd"))

designs |>
  dplyr::transmute(Design = design, Genotype = genotype,
                   `Dose (mg/day)` = dose_mgd,
                   `Responders (%)` = round(pct_responder, 0)) |>
  knitr::kable(caption = "Responder rate under each genotype-based dose-assignment design (Stringer 2014 Table 1 dose levels).")

spread <- designs |>
  dplyr::group_by(design) |>
  dplyr::summarise(spread = max(pct_responder) - min(pct_responder), .groups = "drop")
spread
```

The paper's finding is a *relative* one: Design C is the design under which all
three genotypes reach a comparable responder rate (74%, 72% and 65% reported,
against 73% for rosiglitazone), whereas the lower-dose designs leave the
fast-clearing `UGT2B15*1/*1` group behind. The assertion below tests that
relative claim -- the ordering across designs and the narrowing of the
between-genotype spread -- rather than the absolute percentages, which depend on
random-effect details the paper does not fully specify (see Errata).

```{r design-claim}
resp_wide <- resp |>
  tidyr::pivot_wider(names_from = genotype, values_from = pct_responder)

stopifnot(
  # Under common random numbers a higher dose can only move a subject further
  # below the -0.7% threshold, so the responder rate is EXACTLY monotone in
  # dose within every genotype -- no Monte Carlo slack is needed.
  all(vapply(genotypes$genotype, function(g) {
    r <- resp[resp$genotype == g, ]
    all(diff(r$pct_responder[order(r$dose_mgd)]) >= 0)
  }, logical(1))),
  # Likewise across genotypes at a fixed dose: slower clearance means higher
  # AUC means a larger HbA1c reduction for the same subject, so the responder
  # rate is exactly ordered *2/*2 >= *1/*2 >= *1/*1. This is the
  # exposure-driven genotype effect the paper is quantifying.
  all(resp_wide[["*2/*2"]] >= resp_wide[["*1/*2"]]),
  all(resp_wide[["*1/*2"]] >= resp_wide[["*1/*1"]]),
  # A dose difference must actually move the responder rate: 96 mg/day gives
  # materially more responders than 8 mg/day in every genotype.
  all(resp_wide[["*1/*1"]][resp_wide$dose_mgd == 96] -
        resp_wide[["*1/*1"]][resp_wide$dose_mgd == 8] > 20),
  # Design C equalises response across genotypes better than Design A does.
  spread$spread[spread$design == "Design C"] <
    spread$spread[spread$design == "Design A"]
)
```

# Assumptions and deviations

* **`Emax` is rescaled from percent to fraction.** Table 2 reports
  `Emax (%) 48.7` while `ROTE` on the adjacent row is already a fraction
  (`0.28`, described in the text as "28%"). Equation (3) adds `DEF` and `ROTE`
  together, so they must share a scale; `Emax` is carried as `0.487`. This is
  not a judgement call -- it is the only scale under which the paper's own
  Figure 2a asymptotes and its reported 6-month changes are reproduced, as the
  replication section above shows.

* **`LEHB_active` is read as `0.037 - 0.017 = 0.020`, not as `0.017`.**
  Table 2's second lifestyle row is labelled `LEHB_active` but carries
  footnote c, which prints Equation (2), `LEHB_active = LEHB_placebo -
  LEHB_factor`. The decisive evidence is the tabulated `CV%` of 32.2: every
  other CV% in Table 2 is the relative standard error of an estimated
  NONMEM parameter, taken from the covariance matrix. Footnote c says
  `LEHB_active` is *computed* from two other quantities, so `LEHB_active` is
  not itself a THETA and would not carry an RSE; the quantity that does carry
  one is `LEHB_factor`. Reading the row as the factor gives exactly `0.020`, which is
  what Results prints ("a population mean decrease of 3.7% ... in the placebo
  group, while in the actively treated groups the reduction was slightly lower,
  2%"); had the row been the active value itself, that sentence would read
  "1.7%" to match the one-decimal precision of "3.7%". The two readings differ
  by only about 0.02% HbA1c in every published output -- below the paper's own
  reporting precision -- so this choice is not load-bearing for any check in
  this vignette, but it is recorded here because it changes a parameter value.

* **The lifestyle random effects are additive.** This is forced for `LEFPG`,
  whose population value is exactly zero (a log-normal eta cannot be placed on
  a zero-valued parameter), and the paper says so explicitly. The paper does
  *not* state the distribution of the `LEHB` eta; it is carried additively here
  for consistency with its structurally parallel partner. Under a log-normal
  reading `LEHB` would be effectively deterministic at `0.020 +/- 0.002`, which
  would make the reported `omega^2 LEHB = 0.01` almost uninformative.

* **The baseline OMEGA BLOCK off-diagonal is not reported and is carried as
  zero.** Methods states "The correlation between IIV on baselines was included
  using the OMEGA BLOCK option", but Table 2 reports only the two diagonal
  variances. Note that this omission does not affect any check in this
  vignette: the HbA1c change from baseline depends on `BSL_FPG` only through
  the ratio `FPG_ss / BSL_FPG`, in which `BSL_FPG` cancels, so the baseline
  correlation drops out of the change-from-baseline endpoints entirely.

* **The population-average clearance for ungenotyped subjects is derived, not
  quoted.** Methods says the 10% of subjects without genotype "were included in
  the analysis using an average clearance value for the population" but never
  prints the number. The model file carries 3.21 L/h, computed as the
  genotype-frequency-weighted mean of the Table 1 clearances using the
  supplement Table S1 counts: `(149*5.04 + 357*3.35 + 194*1.53) / 700 = 3.205`
  L/h. No published result depends on this value, since every simulation the
  paper reports is genotype-specific.

* **Rosiglitazone carries no lifestyle effect on HbA1c.** Results states "No
  significant 'lifestyle' effect on HbA1c could be indentified on the
  rosiglitazone group", so both the typical `LEHB` and its random effect are
  switched off on that arm. This is discriminating rather than cosmetic: with
  `LEHB_active` applied, the simulated 6-month HbA1c change for rosiglitazone
  would be -1.41% instead of the reported -1.2%.

* **`KinG` and `KinH` are not tabulated** and are pinned by the requirement
  that each pool sit at its own baseline at `t = 0` with no drug and no
  lifestyle effect. This is the standard indirect-response parameterisation and
  is implied by the paper's description of the baselines as estimated
  parameters, but the paper does not write the two pinning equations out.

* **Absolute responder rates run about 10 percentage points below the published
  values.** Under Design C this model gives roughly 63% / 64% / 66% for
  `*1/*1` / `*1/*2` / `*2/*2` against the 74% / 72% / 65% reported. The
  paper's responder simulation was run in Berkeley Madonna with 100 subjects
  per genotype and does not describe how the random effects were drawn, and two
  of the four random effects have distributions the paper never states (see the
  `LEHB` bullet above); the responder rate is a tail probability against a fixed
  -0.7% threshold, so it is far more sensitive to those unstated distributional
  choices than the central tendency is. Note that the deterministic predictions
  underlying it -- the median changes from baseline -- match the paper to within
  0.06 units, so the discrepancy is in the spread of the simulated population,
  not in the structural model. The *relative* claim -- that Design C equalises
  response across genotypes where the lower-dose designs do not (spread across
  genotypes 3 points under Design C versus 7 under Design A here, against 9
  points under Design C in the paper), and that a slow-clearing genotype
  out-responds a fast-clearing one at any fixed dose -- is reproduced and is
  what this vignette asserts on.

* **Equation (3) is printed as `KinG * (1 +/- LEFPG)`.** This is not a
  text-extraction artefact: the typeset equation on p. 455 carries a literal
  plus-minus sign, which appears exactly once in the whole article. The
  ambiguity is immaterial because `LEFPG` is a mean-zero, symmetrically
  distributed additive random effect, so `(1 + LEFPG)` and `(1 - LEFPG)`
  generate identical distributions. The `+` form is used here, because it makes
  the sign of an individual `LEFPG` readable directly: positive means raised
  glucose production, i.e. loss of glycemic control.

* **Equation (4) writes the FPG exponent as `lambda` while the Methods text and
  Table 2 call it `gamma`.** These are the same parameter (0.71); the file uses
  `lgamma` / `gamma`.

# Errata

* **Table 2's bootstrap column transposes two rows.** The bootstrap 95% CIs
  printed against `omega^2 BSL FPG` (0.008-0.01) and `omega^2 BSL HbA1c`
  (0.043-0.055) each exclude their own point estimate (0.05 and 0.01
  respectively); swapping them makes both contain theirs, and the swapped
  assignment is also the one consistent with the stated model estimates. Only
  the bootstrap column is affected. The Model Estimate values carried in this
  package are unchanged, and the text's claim that "all of the parameters from
  the final model were within the 95% confidence interval of the bootstrap
  estimate" holds only under the corrected assignment.

* **Table 2's `omega^2` rows are variances, not standard deviations.** The
  column labels say `omega^2`, and the bootstrap intervals confirm it: the
  corrected CI for `omega^2 BSL FPG` (0.043-0.055 around 0.05) has a relative
  half-width near 12%, matching the roughly 10% expected for a variance
  estimated from 780 subjects, and about twice what an SD-scale estimate would
  show.

* The paper spells "identified" as "indentified" in the Lifestyle Effect Model
  section; quotations above preserve the original spelling.
