---
title: "Sipoglitazar (Stringer 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sipoglitazar (Stringer 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Stringer 2013 fits **two** population PK models to the same pooled data set, and
both are packaged here:

```{r models}
uiGeno <- rxode2::rxode(readModelDb("Stringer_2013_sipoglitazar"))
uiMix  <- rxode2::rxode(readModelDb("Stringer_2013_sipoglitazar_mixture"))
```

* Citation: `r uiGeno$reference`
* `Stringer_2013_sipoglitazar`: `r uiGeno$description`
* `Stringer_2013_sipoglitazar_mixture`: `r uiMix$description`
* Article: <https://doi.org/10.1177/0091270012447121>
* Supplement: supplementary online material at the journal (Table A1, Figures S1-S4)

Sipoglitazar is an investigational oral PPAR alpha / delta / gamma agonist that
is cleared by UGT-catalysed glucuronidation. The *UGT2B15* D85Y (`*2`) allele
reduces that activity, and the paper's question is how far the genotype explains
the wide spread in clearance seen in phase I.

The **genotype model** assigns each subject's typical clearance from the
*observed* `UGT2B15` genotype. The **mixture model** was fitted in parallel, with
the NONMEM `$MIX` subroutine assigning each subject to one of three *latent*
metabolizer subpopulations from apparent clearance alone, deliberately ignoring
the genotype. Everything else about the two models is the same. Comparing them
is the paper's central argument: 61 of the 744 subjects genotyped `*1/*1` or
`*1/*2` (8%) are placed in the poor-metaboliser subpopulation by the mixture
model, i.e. their phenotype does not follow their genotype.

# Population

The analysis pools `r uiGeno$population$n_subjects` subjects from
`r uiGeno$population$n_studies` studies (Stringer 2013 Table 1):

* **Phase I study 006** -- 524 healthy volunteers (304 female), median age 26
  years (18-55), median weight 71 kg (48-113). A single oral 64 mg dose with
  dense sampling at 0, 1, 2, 3, 4, 6, 8, 12, 16 and 24 h.
* **Phase II study EC201** -- 320 adults with type 2 diabetes and no prior
  antidiabetic medication (155 female), median age 56 years (34-75), median
  weight 90 kg (60-149). Placebo and 8, 16, 32 and 64 mg daily for 12 weeks.
* **Phase II study EC202** -- 307 adults with type 2 diabetes (164 female),
  median age 57 years (34-74), median weight 87 kg (55-149). Placebo and 32 or
  64 mg total daily dose, once or twice daily, for 12 weeks.

Pooled `UGT2B15` genotype frequencies were `*1/*1` 22%, `*1/*2` 51% and `*2/*2`
27%. The phase II studies contributed three trough samples per subject (weeks 4,
6 and 8) whose actual time after dose was never recorded, which is why the model
carries a separate -- and much larger, and itself variable -- residual error for
that stratum.

The same information is available programmatically via
`readModelDb("Stringer_2013_sipoglitazar")()$population`.

A predetermined chronic-exposure safety limit of AUC > 73 mg*h/L is quoted in
Methods and drawn as a grey line on Figure 2; it is used below.

# Source trace

Per-parameter origins are recorded as in-file comments next to each `ini()`
entry in `inst/modeldb/specificDrugs/Stringer_2013_sipoglitazar.R` and
`..._mixture.R`. Collected here for review.

| Equation / parameter | Genotype model | Mixture model | Source location |
|---|---|---|---|
| Structural model | 2-compartment, zero-order input into a depot then first-order absorption, linear elimination | same | Results, *Base Model*; Methods, *Population Data Analysis* |
| `lcl_s1s1` / `lcl_em` | 4.46 L/h | 5.04 L/h | Table 2 / Suppl. Table A1, `CL (*1/*1)` / `CL POP1 (EM)` |
| `lcl_s1s2` / `lcl_im` | 3.25 L/h | 3.35 L/h | Table 2 / Suppl. Table A1, `CL (*1/*2)` / `CL POP2 (IM)` |
| `lcl_s2s2` / `lcl_pm` | 1.53 L/h | 1.53 L/h | Table 2 / Suppl. Table A1, `CL (*2/*2)` / `CL POP3 (PM)` |
| `lvc` | 9.03 L | 9.06 L | Table 2 / Suppl. Table A1, `V` |
| `lvp` | 1.71 L | 0.188 x 9.06 = 1.703 L | Table 2 / Suppl. Table A1, `V2` and footnote b |
| `lq` | 0.313 L/h | 0.311 L/h | Table 2 / Suppl. Table A1, `Q` |
| `lka` | 2.07 1/h | 2.15 1/h | Table 2 / Suppl. Table A1, `ka` / `ka1` |
| `ld1` | 0.568 h | 0.637 h | Table 2 / Suppl. Table A1, `D1` |
| `e_ffm_vc` | 0.00349 1/kg | 0.00556 1/kg | Table 2 / Suppl. Table A1, `theta(f)`; form from equation 4 |
| Covariate equation | `P = theta(i) * (1 + theta(f) * (COV - COV_median))` | same | Equation 4, p. 258 |
| IIV form | `theta_i = theta_mean * exp(eta_i)` | same | Equation 1, p. 258 |
| `etalcl` | 40.25% CV | 38.21% CV | Table 2 / Suppl. Table A1, IIV column |
| `etalvc` | 34.21% CV | 34.50% CV | Table 2 / Suppl. Table A1, IIV column |
| `etald1` | 78.29% CV | 77.20% CV | Table 2 / Suppl. Table A1, IIV column |
| `propSdPhase1` | sqrt(0.0552) = 0.2349 | sqrt(0.05487) = 0.2342 | Table 2 / Suppl. Table A1, residual phase I |
| `propSdPhase2` | sqrt(0.167) = 0.4087 | sqrt(0.167) = 0.4087 | Table 2 / Suppl. Table A1, residual phase II |
| `etapropSdPhase2` | 76.88% CV | 72.18% CV | Table 2 / Suppl. Table A1, IIV on the phase II residual |
| Residual form | `C = PRED * (1 + ERR)` (phase I), `C = PRED * (1 + ERR)` with IIV on ERR (phase II) | same | Equations 2 and 3, p. 258-259 |
| Mixture weights | n/a | POP1 0.18, POP2 0.522, POP3 0.30 | Suppl. Table A1, `Probability of belonging to POP 1/2/3` |
| Exposure limit 73 mg*h/L | n/a | n/a | Methods, *Population Data Analysis*; Figure 2 grey line |

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the phase I design: a single 64 mg oral dose in healthy volunteers, 200
participants per `UGT2B15` genotype.

Stringer 2013 does not report the cohort median fat-free mass that centres the
FFM effect (see *Assumptions and deviations*), so the model uses 50 kg and this
cohort draws FFM around that value with a spread consistent with the reported
weight range.

```{r cohort}
set.seed(20130313)
rxode2::rxSetSeed(20130313)

nPerArm <- 200L

genotypes <- tibble::tribble(
  ~geno,    ~UGT2B15_STAR2_HET, ~UGT2B15_STAR2_HOM,
  "*1/*1",  0L,                 0L,
  "*1/*2",  1L,                 0L,
  "*2/*2",  0L,                 1L
)

# The paper's phase I sampling grid, plus a fine grid so that trapezoidal AUC
# resolves the absorption peak (tmax is about 1 h).
obsTimes <- sort(unique(c(
  c(0, 1, 2, 3, 4, 6, 8, 12, 16, 24),
  seq(0, 6, by = 0.05),
  seq(6.5, 24, by = 0.5)
)))

subjects <- genotypes |>
  dplyr::mutate(arm = dplyr::row_number()) |>
  tidyr::crossing(k = seq_len(nPerArm)) |>
  dplyr::mutate(
    id  = (arm - 1L) * nPerArm + k,
    FFM = pmax(30, rnorm(dplyr::n(), mean = 50, sd = 8)),
    # Phase I healthy-volunteer stratum selects the smaller residual error.
    STUDY_SIPO_PHASE2 = 0L
  ) |>
  dplyr::select(-k, -arm)

dosingRows <- subjects |>
  dplyr::mutate(time = 0, amt = 64, evid = 1L, cmt = "depot")

obsRows <- subjects |>
  tidyr::crossing(time = obsTimes) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")

events <- dplyr::bind_rows(dosingRows, obsRows) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
# Named simPhase1 rather than `sim` because rxSolve's residual-error column is
# itself called `sim`; keeping the two apart avoids data-mask shadowing.
simPhase1 <- rxode2::rxSolve(
  uiGeno,
  events = events,
  keep   = c("geno", "FFM")
) |>
  as.data.frame()
```

A deterministic typical-value solve is used for the structural checks.

```{r simulate-typical}
typicalEvents <- genotypes |>
  dplyr::mutate(id = dplyr::row_number(), FFM = 50, STUDY_SIPO_PHASE2 = 0L) |>
  tidyr::crossing(time = sort(unique(c(seq(0, 8, by = 0.02), seq(8, 150, by = 0.1))))) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central") |>
  dplyr::bind_rows(
    genotypes |>
      dplyr::mutate(id = dplyr::row_number(), FFM = 50, STUDY_SIPO_PHASE2 = 0L,
                    time = 0, amt = 64, evid = 1L, cmt = "depot")
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

simTypical <- rxode2::rxSolve(
  rxode2::zeroRe(uiGeno),
  events = typicalEvents,
  keep   = c("geno"),
  omega  = NA
) |>
  as.data.frame()
```

# Replicate published figures

## Supplementary Figure S2a -- single 64 mg dose in healthy volunteers

The observed medians below were digitised from Supplementary Figure S2a by the
operator (see *Assumptions and deviations*); the panels are on a log scale
spanning three decades, so treat the digitised numbers as accurate to roughly
+/-15%.

```{r fig-s2a-reference}
observedMedians <- tibble::tribble(
  ~geno,   ~time, ~CcObs_ng_mL,
  "*1/*1",  1,     4100,
  "*1/*1",  2,     3300,
  "*1/*1",  3,     2100,
  "*1/*1",  4,     1350,
  "*1/*1",  6,      480,
  "*1/*1",  8,      200,
  "*1/*1", 12,       60,
  "*1/*1", 16,       21,
  "*1/*1", 24,      6.5,
  "*1/*2",  1,     4800,
  "*1/*2",  2,     4300,
  "*1/*2",  3,     3000,
  "*1/*2",  4,     2400,
  "*1/*2",  6,      950,
  "*1/*2",  8,      570,
  "*1/*2", 12,      200,
  "*1/*2", 16,       80,
  "*1/*2", 24,       18,
  "*2/*2",  1,     5300,
  "*2/*2",  2,     5600,
  "*2/*2",  3,     4600,
  "*2/*2",  4,     4000,
  "*2/*2",  6,     2200,
  "*2/*2",  8,     1600,
  "*2/*2", 12,      830,
  "*2/*2", 16,      530,
  "*2/*2", 24,      230
) |>
  # The model carries concentrations in mg/L; the paper plots ng/mL.
  dplyr::mutate(Cc = CcObs_ng_mL / 1000)
```

```{r figure-s2a}
vpc <- simPhase1 |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::group_by(geno, time) |>
  dplyr::summarise(
    Q05 = quantile(sim, 0.05, na.rm = TRUE),
    Q50 = quantile(sim, 0.50, na.rm = TRUE),
    Q95 = quantile(sim, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(dplyr::across(c(Q05, Q50, Q95), ~ pmax(.x, 1e-4)))

ggplot(vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_point(
    data = observedMedians, aes(time, Cc),
    colour = "blue", shape = 3, size = 2.5, inherit.aes = FALSE
  ) +
  facet_wrap(~geno) +
  scale_y_log10() +
  coord_cartesian(ylim = c(1e-3, 20)) +
  labs(
    x = "Time (h)", y = "Plasma sipoglitazar (mg/L)",
    title = "Supplementary Figure S2a -- single 64 mg oral dose, healthy volunteers",
    caption = paste(
      "Line and band: simulated median and 90% prediction interval",
      "(200 subjects per genotype). Crosses: observed medians digitised",
      "from Supplementary Figure S2a of Stringer 2013."
    )
  )
```

```{r figure-s2a-numbers}
medianProfile <- simPhase1 |>
  dplyr::filter(!is.na(Cc), time %in% c(0, 1, 2, 3, 4, 6, 8, 12, 16, 24)) |>
  dplyr::group_by(geno, time) |>
  dplyr::summarise(CcSim = median(Cc), .groups = "drop")

profileCheck <- observedMedians |>
  dplyr::select(geno, time, CcObs = Cc) |>
  dplyr::inner_join(medianProfile, by = c("geno", "time")) |>
  dplyr::mutate(pctDiff = 100 * (CcSim - CcObs) / CcObs)

profileCheck |>
  dplyr::mutate(
    CcObs  = signif(CcObs * 1000, 3),
    CcSim  = signif(CcSim * 1000, 3),
    pctDiff = round(pctDiff, 1)
  ) |>
  tidyr::pivot_wider(
    names_from  = geno,
    values_from = c(CcObs, CcSim, pctDiff)
  ) |>
  dplyr::rename("Time (h)" = time) |>
  knitr::kable(
    caption = paste(
      "Digitised observed median (CcObs, ng/mL) vs simulated median (CcSim,",
      "ng/mL) after a single 64 mg dose, by UGT2B15 genotype."
    )
  )
```

## Figure 2 -- steady-state AUC by dose and genotype

Stringer 2013 builds Figure 2 from the identity it states in the text,
`AUC = dose / CL`, applied to each subject's post-hoc clearance over the
steady-state dosing interval. The same construction is used here on the
simulated cohort's individual clearances, so the comparison is like for like.

```{r figure-2}
subjectCl <- simPhase1 |>
  dplyr::group_by(id, geno) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

exposureLimit <- 73 # mg*h/L, Stringer 2013 Methods and Figure 2 grey line

aucByDose <- subjectCl |>
  tidyr::crossing(dailyDose = c(8, 16, 32, 64)) |>
  dplyr::mutate(auc = dailyDose / cl)

ggplot(aucByDose, aes(geno, auc)) +
  geom_boxplot(outlier.size = 0.7) +
  geom_hline(yintercept = exposureLimit, linetype = "dashed", colour = "grey40") +
  facet_wrap(~dailyDose, nrow = 1, labeller = label_both) +
  scale_y_log10() +
  labs(
    x = "UGT2B15 genotype", y = "AUC over the dosing interval (mg*h/L)",
    title = "Figure 2 -- steady-state AUC by genotype and total daily dose",
    caption = paste(
      "Dashed line: the 73 mg*h/L chronic-exposure limit of Stringer 2013.",
      "AUC = dose / CL, the paper's own construction."
    )
  )
```

```{r figure-2-table}
aucSummary <- aucByDose |>
  dplyr::group_by(dailyDose, geno) |>
  dplyr::summarise(
    median = median(auc),
    q25    = quantile(auc, 0.25),
    q75    = quantile(auc, 0.75),
    pctOverLimit = 100 * mean(auc > exposureLimit),
    .groups = "drop"
  )

aucSummary |>
  dplyr::mutate(dplyr::across(c(median, q25, q75), ~ signif(.x, 3)),
                pctOverLimit = round(pctOverLimit, 1)) |>
  dplyr::rename(
    "Total daily dose (mg)"   = dailyDose,
    "UGT2B15 genotype"        = geno,
    "Median AUC (mg*h/L)"     = median,
    "25th percentile"         = q25,
    "75th percentile"         = q75,
    "% above 73 mg*h/L"       = pctOverLimit
  ) |>
  knitr::kable(caption = "Simulated steady-state AUC by dose and genotype (200 subjects per genotype).")
```

The paper reads Figure 2 as: exposure rises with dose, the `*2/*2` group sits
above the other two at every dose, and only at 32 and 64 mg do subjects begin to
cross the 73 mg*h/L line. The simulated table reproduces all three.

## The two models side by side

```{r mixture-typical}
mixStrata <- tibble::tribble(
  ~pop,  ~UGT2B15_IM, ~UGT2B15_PM,
  "EM",  0L,          0L,
  "IM",  1L,          0L,
  "PM",  0L,          1L
)

mixTypicalEvents <- mixStrata |>
  dplyr::mutate(id = dplyr::row_number(), FFM = 50, STUDY_SIPO_PHASE2 = 0L) |>
  tidyr::crossing(time = seq(0, 24, by = 0.05)) |>
  dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central") |>
  dplyr::bind_rows(
    mixStrata |>
      dplyr::mutate(id = dplyr::row_number(), FFM = 50, STUDY_SIPO_PHASE2 = 0L,
                    time = 0, amt = 64, evid = 1L, cmt = "depot")
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

simMixTypical <- rxode2::rxSolve(
  rxode2::zeroRe(uiMix),
  events = mixTypicalEvents,
  keep   = c("pop"),
  omega  = NA
) |>
  as.data.frame()

dplyr::bind_rows(
  simTypical |>
    dplyr::filter(!is.na(Cc), time <= 24) |>
    dplyr::transmute(time, Cc, stratum = geno, model = "Genotype model"),
  simMixTypical |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(time, Cc, stratum = pop, model = "Mixture model")
) |>
  ggplot(aes(time, Cc, colour = stratum)) +
  geom_line() +
  facet_wrap(~model) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma sipoglitazar (mg/L)",
    colour = "Stratum",
    title = "Typical-value profiles after a single 64 mg dose",
    caption = paste(
      "Left: genotype strata (Table 2). Right: latent metabolizer",
      "subpopulations (Supplementary Table A1)."
    )
  )
```

`rxode2` has no analogue of the NONMEM `$MIX` subroutine, so the mixture model's
subpopulation enters as a covariate indicator. A simulation reproduces the
mixture by sampling that indicator from the weights the paper estimated
(`population$mixture_probabilities`).

```{r mixture-weights}
mixtureWeights <- uiMix$population$mixture_probabilities
mixtureWeights

set.seed(918)
drawnStratum <- sample(
  names(mixtureWeights),
  size    = 1000,
  replace = TRUE,
  prob    = mixtureWeights / sum(mixtureWeights)
)

tibble::tibble(
  stratum   = names(mixtureWeights),
  published = as.numeric(mixtureWeights),
  sampled   = as.numeric(table(factor(drawnStratum, levels = names(mixtureWeights))) / 1000)
) |>
  dplyr::rename(
    "Subpopulation"      = stratum,
    "Published weight"   = published,
    "Sampled proportion" = sampled
  ) |>
  knitr::kable(digits = 3, caption = "Mixture prior weights, Supplementary Table A1.")
```

# PKNCA validation

Both sides of the comparison are put through the same instrument: NCA on a
*median concentration profile* observed at the paper's ten phase I sampling
times. The reference profile is the digitised Supplementary Figure S2a median;
the simulated profile is the median across the 200 simulated subjects at those
same times.

```{r pknca}
ncaTimes <- c(0, 1, 2, 3, 4, 6, 8, 12, 16, 24)

simMedianNca <- simPhase1 |>
  dplyr::filter(!is.na(Cc), time %in% ncaTimes) |>
  dplyr::group_by(geno, time) |>
  dplyr::summarise(Cc = median(Cc), .groups = "drop") |>
  dplyr::mutate(id = "simulated")

obsMedianNca <- observedMedians |>
  dplyr::select(geno, time, Cc) |>
  dplyr::bind_rows(
    genotypes |> dplyr::transmute(geno, time = 0, Cc = 0)
  ) |>
  dplyr::mutate(id = "observed")

runNca <- function(dat) {
  dat <- dat |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::arrange(geno, id, time) |>
    as.data.frame()
  doses <- dat |>
    dplyr::distinct(geno, id) |>
    dplyr::mutate(time = 0, amt = 64) |>
    as.data.frame()
  concObj <- PKNCA::PKNCAconc(dat, Cc ~ time | geno + id)
  doseObj <- PKNCA::PKNCAdose(doses, amt ~ time | geno + id)
  intervals <- data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))
}

ncaSim <- runNca(simMedianNca)
ncaObs <- runNca(obsMedianNca)
```

## Comparison against published NCA

```{r nca-compare}
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = ncaSim,
  reference = ncaObs,
  by        = "geno",
  params    = c("cmax", "tmax", "aucinf.obs", "half.life"),
  units     = c(cmax = "mg/L", tmax = "h", aucinf.obs = "mg*h/L", half.life = "h"),
  tolerance_pct = 25
)

knitr::kable(
  cmp,
  caption = paste(
    "NCA of the simulated median profile vs NCA of the median profile",
    "digitised from Supplementary Figure S2a.",
    "* differs from the reference by more than 25%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

The reference side of this table is a figure digitisation on a three-decade log
axis, so a tolerance of 25% rather than the usual 20% is used and agreement
inside that band is the most the comparison can establish. The independent and
*exact* check is the identity the paper itself states, `AUC = dose / CL`, which
is asserted below against the published clearances.

**The one starred row is `Tmax` in the `*2/*2` stratum**, where the reference
reads 2 h and the model 1 h. This is a tie on a flat peak resolved differently
by the two sides rather than a disagreement about the absorption rate: the
digitised medians put `*2/*2` at 5300 ng/mL at 1 h and 5600 ng/mL at 2 h, the
model at 5280 and 4950, and both pairs are inside the +/-15% reading
uncertainty of each other. Because `Tmax` on a ten-point grid can only take one
of the sampled values, a 5% difference in peak shape moves it a whole grid step
and shows up as "-50%". The `Cmax` row for the same stratum agrees to 5.7%, and
the `*1/*1` and `*1/*2` strata -- whose peaks are sharper and therefore
better resolved by the same grid -- match `Tmax` exactly. No parameter was
changed.

# Model claims

```{r claims}
typicalSummary <- simTypical |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(geno) |>
  dplyr::summarise(
    cl  = dplyr::first(cl),
    vc  = dplyr::first(vc),
    vp  = dplyr::first(vp),
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    aucIdentity = 64 / cl,
    aucPctDiff  = 100 * (auc - aucIdentity) / aucIdentity
  )

publishedCl <- c("*1/*1" = 4.46, "*1/*2" = 3.25, "*2/*2" = 1.53)

# 1. Typical-value structural parameters reproduce Table 2 exactly.
stopifnot(
  all(abs(typicalSummary$cl - publishedCl[typicalSummary$geno]) < 1e-8),
  abs(unique(typicalSummary$vc) - 9.03) < 1e-8,
  abs(unique(typicalSummary$vp) - 1.71) < 1e-6
)

# 2. AUC(0-inf) = Dose / CL. Both sides use the same drawn parameters, so the
#    residual is pure numerical integration error and the bound is tight.
stopifnot(max(abs(typicalSummary$aucPctDiff)) < 0.5)

# 3. Model vs digitised observed medians. Assertions are on the centre and on a
#    robust quantile, never on the extreme of the 27 comparisons, because the
#    reference side is a figure digitisation.
absPct <- abs(profileCheck$pctDiff)
stopifnot(
  median(absPct) < 15,
  quantile(absPct, 0.9) < 30
)

# 4. Exposure rises monotonically from *1/*1 to *1/*2 to *2/*2 at every dose,
#    and is exactly dose-proportional (linear kinetics, no dose dependency).
ordering <- aucSummary |>
  dplyr::select(dailyDose, geno, median) |>
  tidyr::pivot_wider(names_from = geno, values_from = median)
stopifnot(
  all(ordering[["*2/*2"]] > ordering[["*1/*2"]]),
  all(ordering[["*1/*2"]] > ordering[["*1/*1"]])
)
doseRatio <- aucSummary |>
  dplyr::select(dailyDose, geno, median) |>
  tidyr::pivot_wider(names_from = dailyDose, values_from = median) |>
  dplyr::mutate(ratio = `64` / `8`)
stopifnot(all(abs(doseRatio$ratio - 8) < 1e-8))

# 5. The 73 mg*h/L limit is unreachable at 8 mg and is crossed by a minority of
#    *2/*2 subjects at 64 mg -- the paper's reading of Figure 2.
overLimit <- aucSummary |>
  dplyr::select(dailyDose, geno, pctOverLimit)
stopifnot(
  all(overLimit$pctOverLimit[overLimit$dailyDose == 8] == 0),
  {
    hi <- overLimit$pctOverLimit[overLimit$dailyDose == 64 & overLimit$geno == "*2/*2"]
    hi > 1 && hi < 20
  }
)

# 6. Mixture-model typical clearances reproduce Supplementary Table A1.
mixCl <- simMixTypical |>
  dplyr::group_by(pop) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")
stopifnot(
  all(abs(mixCl$cl - c(EM = 5.04, IM = 3.35, PM = 1.53)[mixCl$pop]) < 1e-8)
)

# 7. Mixture weights sum to 1 within the rounding of Supplementary Table A1.
stopifnot(abs(sum(mixtureWeights) - 1) < 0.005)
```

```{r claims-table, echo = FALSE}
tibble::tribble(
  ~Claim, ~Source, ~Result,
  "Typical CL is 4.46 / 3.25 / 1.53 L/h for *1/*1 / *1/*2 / *2/*2", "Table 2", "reproduced exactly",
  "Typical V = 9.03 L and V2 = 1.71 L at the reference FFM", "Table 2", "reproduced exactly",
  "AUC(0-inf) after a single dose equals Dose / CL", "Results, Final Model", paste0("max deviation ", signif(max(abs(typicalSummary$aucPctDiff)), 2), "%"),
  "Median concentration-time profile after 64 mg, by genotype", "Suppl. Figure S2a (digitised)", paste0("median |difference| ", round(median(absPct), 1), "%, 90th percentile ", round(quantile(absPct, 0.9), 1), "%"),
  "Steady-state AUC is ordered *2/*2 > *1/*2 > *1/*1 at every dose", "Figure 2", "reproduced at 8, 16, 32 and 64 mg",
  "Exposure is dose-proportional (linear kinetics)", "Results, Base Model", "64 mg / 8 mg median AUC ratio = 8.00",
  "No subject exceeds 73 mg*h/L at 8 mg; a minority of *2/*2 subjects do at 64 mg", "Figure 2", paste0("0% at 8 mg; ", round(overLimit$pctOverLimit[overLimit$dailyDose == 64 & overLimit$geno == "*2/*2"], 1), "% of *2/*2 at 64 mg"),
  "Mixture-model typical CL is 5.04 / 3.35 / 1.53 L/h for EM / IM / PM", "Suppl. Table A1", "reproduced exactly"
) |>
  knitr::kable(caption = "Published claims checked by this vignette.")
```

# Assumptions and deviations

* **Absorption structure.** Methods and Results describe "parallel first- and
  zero-order absorption into the central compartment", but Table 2 and
  Supplementary Table A1 report only `ka` and `D1` -- there is no parameter
  splitting the dose between the two routes, and a genuinely parallel model is
  not identified without one. The reported set *is* complete for the other
  reading of the same two parameters: a zero-order release of duration `D1` into
  the depot followed by first-order transfer at `ka`. Two further pieces of
  evidence point the same way. The NONMEM duration variable is named `D1`, which
  is the duration of the input into compartment 1 (the depot in `ADVAN4`); a
  zero-order input straight into the central compartment would have been `D2`
  and would have needed an `F1` companion. And this structure reproduces all 27
  digitised median concentrations of Supplementary Figure S2a (see the table
  above), whereas a parallel model under-predicts the 2 h median at every
  fraction of the dose split. The models therefore encode the sequential
  structure, and the prose "parallel" is read as a loose description of the
  combined zero- plus first-order input.
* **FFM centering value.** Equation 4 centres each covariate on its median, but
  the paper never reports the median fat-free mass. The models use 50 kg, the
  register's reference-adult FFM, which is also close to the value implied by
  the reported weights and sex split (a rough Janmahasatian calculation over the
  three studies gives about 52 kg). The effect is centred, so this choice only
  fixes where the typical `V = 9.03 L` is anchored; over a 20 kg deviation in
  FFM the whole effect is about 7% on `V`, and moving the centre by 2 kg moves
  `V` by 0.7%. A user simulating a specific cohort should substitute that
  cohort's median FFM.
* **Units of the FFM coefficient.** Supplementary Table A1 labels `theta(f)` as
  `(L/kg)`, which would imply an additive effect. Equation 4 is printed
  explicitly as a multiplicative, centred form
  `P = theta(i) * (1 + theta(f) * (COV - COV_median))`, in which `theta(f)` has
  units of 1/kg. The printed equation is used; read as an additive effect the
  coefficient would move `V` by about 1% over the entire weight range of the
  study, which cannot produce the reported drop in the objective function.
* **Peripheral volume.** Footnote b of both tables states that `V2` was
  implemented as a fraction of the central compartment. Table 2 reports the
  product (`V2 = 1.71 L`, which is 0.189 x 9.03 L) while Supplementary Table A1
  reports the raw fraction (0.188). The models encode the fraction, so `V2`
  inherits the central volume's IIV and FFM effect. Reading Table 2's 1.71 as a
  *fraction* instead would give `V2 = 15.4 L` and a terminal half-life of about
  37 h, which is flatly contradicted by Supplementary Figure S2a (the `*1/*1`
  median falls to about 6 ng/mL by 24 h; the fraction reading predicts 12
  ng/mL and a visibly bent terminal slope).
* **IIV on the phase II residual error.** Equation 3 is printed as
  `C = PRED * (1 + ERR) * exp(eta)`, which would place the random effect on the
  prediction rather than on the residual and make it inseparable from the IIV on
  clearance. The Table 2 row it belongs to is "Residual variability phase II" and
  the supplementary methods describe "interindividual variability *on the
  residual error* for the patient trial", so it is encoded as
  `C = PRED * (1 + ERR * exp(eta))`, i.e. a per-subject scaling of the
  proportional residual magnitude. This is a non-mu-referenced eta and `rxode2`
  warns about it at parse time; the warning is expected and the model simulates
  correctly.
* **Mixture-model subpopulation.** `rxode2` and `nlmixr2` have no analogue of
  the NONMEM `$MIX` subroutine, so `Stringer_2013_sipoglitazar_mixture` takes
  the subpopulation assignment as a covariate indicator rather than estimating
  it. The mixture prior weights are carried as
  `population$mixture_probabilities` and are what a simulation should sample the
  indicator from. The model cannot reproduce the paper's per-subject posterior
  probabilities (Supplementary Figures S4a-c), which need the original data.
* **Mixture weight footnote.** Supplementary Table A1 tabulates POP1 0.18,
  POP2 0.522 and POP3 0.30, and footnotes them as `POP2 = (1 - POP1) * PROB` and
  `POP3 = (1 - POP1) * (1 - PROB)` with `PROB = 0.367`. Evaluating those
  expressions gives 0.301 and 0.519, i.e. the footnote's two assignments are
  swapped relative to the tabulated values. The tabulated values are used here
  because they are the ones that track the observed genotype frequencies
  (22 / 51 / 27%).
* **Direction of the clearance comparison in the text.** The *Covariate
  Analysis* paragraph reads "median clearance for genotype groups UGT2B15 *1/*1
  and *1/*2, which were, respectively, 66% and 53% lower than that of the
  genotype UGT2B15 *2/*2". Table 2 has `*1/*1` at 4.46 L/h and `*2/*2` at
  1.53 L/h, so the comparison is inverted: it is `*2/*2` that is 66% and 53%
  lower than `*1/*1` and `*1/*2` (1.53 / 4.46 = 0.34; 1.53 / 3.25 = 0.47). The
  inverted reading also contradicts Figure 2 (higher AUC in `*2/*2`), the
  abstract (larger HbA1c drop in `*2/*2`) and the whole poor-metaboliser
  argument of the paper. Table 2 is used.
* **Which clearances the companion PD paper uses.** Stringer 2014
  (*J Clin Pharmacol* 54(4):453-461,
  [doi:10.1002/jcph.227](https://doi.org/10.1002/jcph.227)) reuses this paper's
  PK to drive an FPG / HbA1c indirect-response model, and its Table 1 fixes the
  per-genotype clearances at 5.04, 3.35 and 1.53 L/h for `*1/*1`, `*1/*2` and
  `*2/*2`. Those are the **mixture-model** estimates of supplementary Table A1
  above, where they belong to the latent EM, IM and PM subpopulations -- not the
  genotype-stratified estimates of Table 2 (4.46, 3.25, 1.53 L/h). Anyone
  wanting "the published sipoglitazar clearance for genotype X" therefore has to
  say which of the two 2013 models they mean;
  `Stringer_2013_sipoglitazar_mixture` is the one whose numbers the 2014 PD
  model consumes.
* **The "61 of 744" denominator.** Both the Discussion and the supplementary
  Results report the genotype / phenotype mismatch as "8% (61 of 744)" of the
  `*1/*1` and `*1/*2` genotype groups. Table 1 puts those two groups at
  124 + 70 + 56 = 250 and 273 + 160 + 157 = 590 subjects, i.e. 840 rather than
  744. Both denominators round the mismatch fraction to 8% (7.3% and 8.2%), and
  nothing in either model depends on it, so the paper's number is quoted as
  printed and the discrepancy is recorded here only.
* **Digitised reference values.** The observed median concentrations in the
  Supplementary Figure S2a comparison were read off the figure by the operator
  at 400 dpi. The panels are log-scaled over three decades, so individual
  readings are accurate to roughly +/-15%; that is why the NCA comparison uses a
  25% tolerance and the assertions are on the median and the 90th percentile of
  the absolute differences rather than on any single point.
* **Pharmacodynamics.** The paper's HbA1c analysis (Figures 3a and 3b) is an
  analysis of variance on the 12-week change from baseline stratified by dose
  and genotype, not a PK/PD model, so there is no PD model to extract. The
  exposure-response conclusion is reproduced here only indirectly, through the
  genotype ordering of AUC.
* **Cohort covariates.** The virtual cohort places 200 subjects in each genotype
  arm rather than at the population frequencies (22 / 51 / 27%) so that each
  panel of the Supplementary Figure S2a replication carries the same precision.
  FFM is drawn as `N(50, 8)` truncated below at 30 kg, which is an assumption of
  this vignette; the paper reports weight but not FFM.
* **Bioavailability.** Footnote a of both tables states that bioavailability is
  unknown, so `CL` and `V` are apparent (`CL/F`, `V/F`). No `lfdepot` is
  estimated and none is encoded.
