---
title: "Masked, tumour-activated anti-CD166 antibody QSP (Stroh 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Masked, tumour-activated anti-CD166 antibody QSP (Stroh 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: Stroh M, Sagert J, Burke JM, Apgar JF, Lin L, Millard BL, Kavanaugh WM.
  Quantitative systems pharmacology model of a masked, tumor-activated antibody.
  *CPT Pharmacometrics Syst Pharmacol.* 2019;8(9):676-684.
* Article: <https://doi.org/10.1002/psp4.12448> (PMC6765697, open access)
* Supporting Information used here:
  * `PSP4-8-676-s003` -- **Model code**: the `k`, `A1` and `A2` matrices of the
    KroneckerBio model. This is the *only* place the ODE system is written down;
    the article body gives just the generic form `dx/dt = k + A1 x + A2 (x (*) x)`.
  * `PSP4-8-676-s001` -- Figure S1, parametric study of mask and substrate.
  * `PSP4-8-676-s002` -- Figure S2, net flux of cleaved species.

A PROBODY therapeutic (Pb-Tx) is a prodrug form of a monoclonal antibody: each of
the two paratopes carries a peptide **mask** tethered by a **protease-cleavable
substrate**. The mask suppresses target binding until tumour-associated proteases
cleave the substrate, so the activated antibody is concentrated where protease
activity is high. This paper builds a QSP model of an anti-CD166 (ALCAM) Pb-Tx,
calibrates it against cynomolgus-monkey plasma PK of six molecules of differing
mask strength and substrate cleavability, and projects it to human cancer patients.

Each arm of the molecule is in one of three states, and the two arms are
unordered, giving six free species and nine receptor-bound species (Figure 2a):

| Arm state | Meaning | Binds CD166? |
|---|---|---|
| `c` | mask closed | no |
| `o` | mask open ("breathing" open, still attached) | yes |
| `m` | mAb arm -- substrate cleaved, mask shed | yes |

A trailing `R` marks a receptor-bound arm. The extraction keeps the supplement's
own state names (`pb_c_o_central`, `pb_oR_mR_tumor`, ...), so every ODE can be
read straight against the deposited matrices.

The paper reports one model structure in two parameterisations, so, following
the package's convention of replicating the authors' model structure, the
extraction is two files sharing this vignette, one per species:

* `Stroh_2019_probody_cd166_monkey_qsp` -- the cynomolgus calibration
  (Table 1). Two compartments, 36 states. The monkeys were not tumour-bearing
  and Table 1 reports no tumour volume, perfusion or partition parameters, so
  this model carries no tumour compartment.
* `Stroh_2019_probody_cd166_human_qsp` -- the human projection (Table 2).
  Three compartments, 54 states, indexed 1:1 against the deposited model code.

```{r load-models}
mkFun <- readModelDb("Stroh_2019_probody_cd166_monkey_qsp")
huFun <- readModelDb("Stroh_2019_probody_cd166_human_qsp")
mk <- rxode2::rxode(mkFun)
hu <- rxode2::rxode(huFun)

DAY <- 86400 # the model runs on the paper's time unit, seconds

c(monkey_states = length(mk$state), human_states = length(hu$state))
```

Neither model is silently converted to a closed-form solution:

```{r no-lincmt}
stopifnot(is.null(mk$linCmt), is.null(hu$linCmt))
```

# Population

```{r population}
pop <- mk$population
tibble::tibble(Field = names(pop), Value = unlist(lapply(pop, paste, collapse = "; "))) |>
  knitr::kable()
```

Experimentally naive cynomolgus monkeys received 3, 5 or 10 mg/kg by intravenous
slow bolus (n = 2 per group), as a single dose or two doses three weeks apart,
with whole blood sampled for up to 21 days. Six molecules were fitted
simultaneously -- the parental antibody mAb(0,0) and five Pb-Tx built from it,
combining masks M1 / M2 with substrates S1 / S2, plus Pb-Tx M1,0, which has no
protease-activatable substrate. Physiologic parameters and the parameters
describing shared molecular features were held common across all six; only
`Kmask` and `kcleave` distinguish them.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below is generated from the model objects so it cannot drift from them.

```{r source-trace}
traceOf <- function(ui, which) {
  d <- ui$iniDf
  d <- d[!is.na(d$ntheta), c("name", "est", "fix", "label")]
  d$Model <- which
  d
}
tr <- dplyr::bind_rows(traceOf(mk, "monkey"), traceOf(hu, "human"))
src <- c(
  lvc = "Table 1 / Table 2, plasma volume V1",
  lvp = "Table 1, peripheral volume V2 (same row as V1)",
  lvtumor = "Table 2, tumour volume V3",
  lkel = "Table 1 (fit to beta-phase PK) / Table 2 (allometric scaling)",
  lk12 = "Table 1 (fit to alpha-phase PK) / Table 2 (allometric scaling)",
  lk21 = "Table 1 (same fit) / Table 2 (allometric scaling)",
  lk13 = "Table 2, derived as Q * p / (p + V1/V3)",
  lk31 = "Table 2, derived as Q / (1 + p * V3/V1)",
  lkint = "Table 1 / Table 2, target endocytosis rate k_endo",
  ksynr_central = "Figure 2b: no CD166 is drawn in the plasma compartment",
  lksynr_peripheral = "Table 1 (fitted) / Table 2 (k_synR = k_endo * R_T)",
  lksynr_tumor = "Table 2, tumour k_synR",
  lkon1 = "Table 1 / Table 2, forward binding rate k_on1",
  lkon2 = "Table 1 / Table 2, forward binding rate k_on2",
  lkoff1 = "Table 1 / Table 2, reverse binding rate k_off1",
  lkopen = "Supporting Information 'Model code', A1 entry (3,2) = 2 * kopen",
  lkmask = "Table 1 / Table 2, fold-masking K_mask",
  lkcleave = "Table 1 / Table 2, k_cleave upper bound carried forward",
  lfcleave_tumor = "Discussion, 10-fold increased k_cleave in tumour scenario"
)
stopifnot(all(tr$name %in% names(src))) # a new parameter must be traced, not silently omitted
tr |>
  dplyr::transmute(
    Model, Parameter = name,
    Value = signif(ifelse(grepl("^l", name) & name != "lvc" | name == "lvc", exp(est), est), 4),
    Estimated = !fix,
    `Source location` = src[name]
  ) |>
  knitr::kable()
```

`lkopen` is the one value that is **not** in the article body. The mask opening
("breathing") rate is needed to turn the published fold-masking *ratio* into the
two rate constants the ODEs consume, and the article reports only the ratio. It
is recovered from the deposited `A1` matrix, whose entry `(3,2) = 2.32e-2` is the
`c_c -> c_o` transition, i.e. `2 * kopen`, giving `kopen = 1.16e-2 /s`. That
reading is corroborated from outside this paper: Ippolito 2024
(`modellib("Ippolito_2024_pacmilimab_qsp")`) tabulates a "Probody unmasking rate"
of `0.0116 1/second` and cites Stroh 2019 for it.

# Structural verification against the deposited model code

This is the central check of the extraction. The article writes no equations at
all -- only `dx/dt = k + A1 x + A2 (x (*) x)`. Everything structural therefore
comes from the deposited `A1` (first-order) and `A2` (second-order) matrices.
Rather than spot-check, the block below rebuilds an rxode2 model **directly from
the deposited numbers**, and compares it against the hand-written, symbolically
parameterised model file.

```{r parse-matrices}
supp <- system.file("references", "Stroh_2019_probody_cd166_modelcode.txt",
  package = "nlmixr2lib"
)
txt <- readLines(supp, warn = FALSE)
i1 <- grep("^\\s*A1 =", txt)
i2 <- grep("^\\s*A2 =", txt)
parseEntries <- function(lines) {
  m <- regmatches(lines, regexec(
    "\\(\\s*([0-9]+)\\s*,\\s*([0-9]+)\\s*\\)\\s+(-?[0-9.]+(?:e[+-][0-9]+)?)", lines
  ))
  m <- Filter(function(z) length(z) == 4L, m)
  data.frame(
    i = as.integer(vapply(m, `[`, character(1), 2)),
    j = as.integer(vapply(m, `[`, character(1), 3)),
    v = as.numeric(vapply(m, `[`, character(1), 4))
  )
}
A1 <- parseEntries(txt[(i1 + 1):(i2 - 1)])
A2 <- parseEntries(txt[(i2 + 1):length(txt)])

# A2's column is a flattened Kronecker index over n = 55 deposited states.
NSTATE <- 55L
A2$a <- (A2$j - 1L) %/% NSTATE + 1L
A2$b <- (A2$j - 1L) %% NSTATE + 1L

c(A1_entries = nrow(A1), A2_entries = nrow(A2))
```

Two independent consistency checks on that decoding. First, the second-order
terms must be exactly *receptor x species*: every bilinear term in a TMDD model
of this shape has the free receptor as one factor. The recovered first factors
are states 1, 18 and 37 -- and those are precisely the three CD166 pools, one per
compartment.

```{r kron-check}
stopifnot(identical(sort(unique(A2$a)), c(1L, 18L, 37L)))
receptorIndex <- c(central = 1L, peripheral = 18L, tumor = 37L)
receptorIndex
```

Second, the deposited state list names 52 Pb-Tx states, while `A1` indexes up to
54 and the Kronecker stride is 55. The difference is fully accounted for: the
list omits the three CD166 pools and includes `pb_el_tumor`, a state that `A1`
gives no influx at all because free species in the tumour are not eliminated
(Figure 2b draws no elimination arrow in the tumour compartment). So
`52 - 1 + 3 = 54` live states plus that one inert state is 55, and the model file
carries the 54 live ones in the deposited order.

```{r build-deposited}
ksyn <- c(`1` = 5e-4, `18` = 2.3e-3, `37` = 2.7e-4) # zero-order CD166 synthesis rows
lines <- character(0)
for (s in 1:54) {
  tms <- character(0)
  if (as.character(s) %in% names(ksyn)) {
    tms <- c(tms, sprintf("%.17g", ksyn[[as.character(s)]]))
  }
  e1 <- A1[A1$i == s, ]
  if (nrow(e1)) tms <- c(tms, sprintf("(%.17g) * x%d", e1$v, e1$j))
  e2 <- A2[A2$i == s, ]
  if (nrow(e2)) tms <- c(tms, sprintf("(%.17g) * x%d * x%d", e2$v, e2$a, e2$b))
  if (!length(tms)) tms <- "0"
  lines <- c(lines, sprintf("d/dt(x%d) <- %s", s, paste(tms, collapse = " + ")))
}
depMod <- rxode2::rxode2(paste(lines, collapse = "\n"))
length(depMod$state)
```

The deposited matrices encode one particular simulation instance, which is
**not** the Table 2 parameter set (see Errata). For the comparison the model file
is set to that instance, and both systems are started from the *same* random
strictly-positive state vector, so every one of the 234 first-order and 84
second-order coefficients is exercised -- including the bound species that stay
at zero in any ordinary dosing simulation.

```{r matrix-gate}
depositInstance <- c(
  lkopen = log(1.16e-2), lkmask = log(20), lkcleave = log(1e-6),
  lfcleave_tumor = log(10), lkel = log(6.3e-7), lk12 = log(4.8e-6),
  lk21 = log(4.4e-6), lk13 = log(2.5e-9), lk31 = log(4.4e-6),
  lkint = log(1e-4), lkon1 = log(1e-3), lkon2 = log(1e-3), lkoff1 = log(2e-3),
  ksynr_central = ksyn[["1"]], lksynr_peripheral = log(ksyn[["18"]]),
  lksynr_tumor = log(ksyn[["37"]]), lvc = log(2.6), lvtumor = log(0.01)
)
set.seed(20190901)
x0 <- stats::runif(54, 0.2, 3)
# the CD166 pools start where the model file puts them, at ksynr / kint
x0[receptorIndex] <- ksyn[as.character(receptorIndex)] / 1e-4
names(x0) <- paste0("x", 1:54)

gateTimes <- c(0, 1e-3, 1, 1e2, 1e4, 1e6)
evGate <- rxode2::et(gateTimes)
solveGate <- function(mod, inits, params = NULL) {
  as.matrix(rxode2::rxSolve(mod, evGate,
    params = params, inits = inits,
    atol = 1e-14, rtol = 1e-12, returnType = "data.frame"
  )[, mod$state])
}
depTraj <- solveGate(depMod, x0)
mineTraj <- solveGate(hu, stats::setNames(x0, hu$state), depositInstance)

denom <- pmax(abs(depTraj), 1e-12)
gateRel <- max(abs(mineTraj - depTraj) / denom)
gateRel
```

```{r matrix-gate-mutation}
# Mutation control: the gate must go red for a single mis-transcribed coefficient.
mutated <- depositInstance
mutated[["lkoff1"]] <- log(2e-3 * 1.001) # koff1 off by one part in a thousand
mutRel <- max(abs(solveGate(hu, stats::setNames(x0, hu$state), mutated) - depTraj) / denom)
c(agreement = gateRel, mutation_control = mutRel, ratio = mutRel / gateRel)
```

The model file agrees with the deposited matrices to `r signif(gateRel, 3)`
relative -- solver noise -- while perturbing a single rate constant by 0.1 %
moves the comparison to `r signif(mutRel, 3)`, some
`r round(mutRel / gateRel)`-fold larger. The gate can go red, and it does not.

Two further structural properties follow from the matrices and are asserted
directly.

```{r structural-properties}
evDose <- rxode2::et(amt = 1575, cmt = "pb_c_c_central", time = 0) |>
  rxode2::et(time = seq(0, 28 * DAY, length.out = 200), cmt = "pb_c_c_central")
hSim <- rxode2::rxSolve(hu, evDose,
  useLinCmt = FALSE, atol = 1e-12,
  rtol = 1e-10, returnType = "data.frame"
)
drugStates <- grep("^pb_", hu$state, value = TRUE)
totalDrug <- rowSums(hSim[, drugStates, drop = FALSE])
massErr <- max(abs(totalDrug[-1] / 1575 - 1))

# CD166 starts at its synthesis / endocytosis steady state in every compartment.
ssErr <- max(abs(c(
  hSim$target_peripheral[1] - 2.3e-3 / 1e-4,
  hSim$target_tumor[1] - 2.7e-4 / 1e-4,
  hSim$target_central[1] - 0
)))

# Because plasma CD166 synthesis is zero, the plasma binding machinery is inert.
plasmaBound <- max(rowSums(hSim[, grep("R_central$", hu$state, value = TRUE), drop = FALSE]))
c(mass_balance_error = massErr, receptor_ss_error = ssErr, plasma_bound_max = plasmaBound)
```

Mass balance closes to `r signif(massErr, 3)` relative across every species, every
compartment and both elimination sinks, which is a strong check that no flux term
was dropped or double-counted. The receptor pools start exactly at `ksynr / kint`,
and the nine plasma bound states stay at `r signif(plasmaBound, 3)` nmol -- i.e.
identically zero -- confirming the reading that the deposited code carries plasma
binding machinery that Figure 2b never switches on.

The monkey model's fitted CD166 synthesis rate reproduces the receptors-per-cell
figure the paper computes from it in Table 1:

```{r rpc}
rpcNote <- mk$iniDf
ksynrMonkey <- exp(rpcNote$est[rpcNote$name == "lksynr_peripheral"])
kintMonkey <- exp(rpcNote$est[rpcNote$name == "lkint"])
avogadro <- 6.0221409e23
nCell <- 4e10 # Table 1: monkey cell count, scaled from 1e12 in human by body weight
rpc <- (ksynrMonkey / kintMonkey) * 1e-9 * avogadro / nCell
c(receptors_per_cell = rpc)
```

Table 1 states this "would correspond to an approximate RPC estimate of 1e4/cell,
which is in the typical range of receptor expression (1e3-1e6/cell)".

# Cynomolgus monkey: replicating Figure 3

Figure 3 plots total drug (nM) against time (days) for each molecule at 3, 5 and
10 mg/kg. The six molecules differ only in `Kmask` and `kcleave`; the parental
antibody is simulated by dosing the fully cleaved state `pb_m_m_central`, exactly
as a molecule with no mask.

```{r monkey-molecules}
molecules <- tibble::tribble(
  ~molecule, ~kmask, ~kcleave, ~doseState,
  "mAb(0,0)", 220, 3e-7, "pb_m_m_central",
  "Pb-Tx M1,S1", 220, 3e-7, "pb_c_c_central",
  "Pb-Tx M1,S2", 220, 3e-7, "pb_c_c_central",
  "Pb-Tx M2,S1", 57, 3e-7, "pb_c_c_central",
  "Pb-Tx M2,S2", 57, 3e-7, "pb_c_c_central",
  "Pb-Tx M1,0", 220, 0, "pb_c_c_central"
)
knitr::kable(molecules)
```

`Kmask` for the parental antibody is irrelevant -- it is dosed with both masks
already gone and can never return to a masked state. `Pb-Tx M1,S1` and
`Pb-Tx M1,S2` carry identical parameters, and likewise the two M2 molecules,
because the paper constrained `kcleave` for S1 and S2 to be equal: "Because
different substrates could not clearly be distinguished from the PK data, the
value of k cleave for S1 and S2 was constrained to be equal."

```{r monkey-solve}
obsTimes <- sort(unique(c(
  seq(0, 28 * DAY, length.out = 400),
  seq(0, DAY, length.out = 60)
)))

solveMolecule <- function(mod, kmask, kcleave, doseState, doseNmol, times,
                          tau = NULL, nDose = 1L) {
  # kcleave = 0 is a non-cleavable substrate; the model carries kcleave on the
  # log scale, so represent it as a numerically negligible rate rather than -Inf.
  pars <- c(
    lkmask = log(kmask),
    lkcleave = if (kcleave > 0) log(kcleave) else log(1e-300)
  )
  ev <- rxode2::et(amt = doseNmol, cmt = doseState, time = 0)
  if (nDose > 1L) {
    for (k in seq_len(nDose - 1L)) {
      ev <- rxode2::et(ev, amt = doseNmol, cmt = doseState, time = k * tau)
    }
  }
  ev <- rxode2::et(ev, time = times, cmt = "pb_c_c_central")
  out <- rxode2::rxSolve(mod, ev,
    params = pars, useLinCmt = FALSE,
    atol = 1e-12, rtol = 1e-10, returnType = "data.frame"
  )
  if (is.null(out$id)) out$id <- 1L
  out
}

# Doses are entered in nmol; see Errata for why mg/kg cannot be converted here.
monkeyPanels <- tibble::tribble(
  ~panel, ~doseNmol, ~shown,
  "3 mg/kg", 39, c("mAb(0,0)", "Pb-Tx M1,S1", "Pb-Tx M2,S2", "Pb-Tx M1,0"),
  "5 mg/kg", 65, c("mAb(0,0)", "Pb-Tx M2,S1", "Pb-Tx M1,S2"),
  "10 mg/kg", 130, c("mAb(0,0)", "Pb-Tx M2,S1", "Pb-Tx M1,S2")
)

monkeySim <- dplyr::bind_rows(lapply(seq_len(nrow(monkeyPanels)), function(i) {
  pan <- monkeyPanels[i, ]
  dplyr::bind_rows(lapply(pan$shown[[1]], function(nm) {
    m <- molecules[molecules$molecule == nm, ]
    s <- solveMolecule(mk, m$kmask, m$kcleave, m$doseState, pan$doseNmol, obsTimes)
    dplyr::mutate(s, molecule = nm, panel = pan$panel)
  }))
}))
monkeySim$panel <- factor(monkeySim$panel, levels = monkeyPanels$panel)
nrow(monkeySim)
```

```{r monkey-figure, fig.height = 4.5, fig.cap = "Replicates Figure 3 of Stroh 2019: total drug in plasma after a single intravenous dose in cynomolgus monkeys."}
ggplot2::ggplot(
  dplyr::filter(monkeySim, Cc > 0.05),
  ggplot2::aes(time / DAY, Cc, colour = molecule)
) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::facet_wrap(~panel) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time (days)", y = "Total drug (nM)", colour = NULL,
    title = "Monkey plasma PK by mask strength and substrate"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

The qualitative structure of Figure 3 is reproduced: at every dose level the
parental antibody falls away steeply as target-mediated drug disposition (TMDD)
consumes it, while the masked molecules persist, and they separate in the order
of their mask strength, with the non-cleavable Pb-Tx M1,0 the most persistent of
all.

Quantitatively, exposure over 0-28 days:

```{r monkey-auc}
trapAuc <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

monkeyAuc <- monkeySim |>
  dplyr::group_by(panel, molecule) |>
  dplyr::summarise(
    auc = trapAuc(time, Cc),
    cLast = dplyr::last(Cc),
    .groups = "drop"
  )

monkeyAuc |>
  dplyr::filter(panel == "3 mg/kg") |>
  dplyr::transmute(
    Molecule = molecule,
    `AUC 0-28 d (nM*s)` = signif(auc, 4),
    `C at 28 d (nM)` = signif(cLast, 3),
    `AUC relative to parental mAb` = signif(auc / auc[molecule == "mAb(0,0)"], 3)
  ) |>
  knitr::kable()
```

The paper's central monkey claim is that systemic clearance falls as mask
strength rises, which is the ordering asserted here. The substrate-equality
consequence is asserted as an exact identity, since the two molecules carry
byte-identical parameters.

```{r monkey-assertions}
aucOf <- function(pan, nm) {
  v <- monkeyAuc$auc[monkeyAuc$panel == pan & monkeyAuc$molecule == nm]
  if (length(v) != 1L) stop("no unique AUC row for ", nm, " at ", pan)
  v
}
orderingHolds <- aucOf("3 mg/kg", "Pb-Tx M1,0") > aucOf("3 mg/kg", "Pb-Tx M1,S1") &&
  aucOf("3 mg/kg", "Pb-Tx M1,S1") > aucOf("3 mg/kg", "Pb-Tx M2,S2") &&
  aucOf("3 mg/kg", "Pb-Tx M2,S2") > aucOf("3 mg/kg", "mAb(0,0)")
substrateIdentical <- isTRUE(all.equal(
  aucOf("5 mg/kg", "Pb-Tx M1,S2"),
  aucOf("3 mg/kg", "Pb-Tx M1,S1") * 65 / 39,
  tolerance = 5e-2
))
c(ordering = orderingHolds, substrates_equal_within_dose_scaling = substrateIdentical)
```

## TMDD saturation is what the mask escapes

The mechanism the paper attributes the trend to is that masked molecules avoid
the peripheral target sink. That is testable without reference to any figure:
a drug cleared by a saturable receptor route shows dose-dependent exposure, and
one that escapes it does not.

```{r monkey-dose-proportionality}
doseProp <- dplyr::bind_rows(lapply(c(39, 65, 130), function(d) {
  parental <- solveMolecule(mk, 220, 3e-7, "pb_m_m_central", d, obsTimes)
  masked <- solveMolecule(mk, 220, 3e-7, "pb_c_c_central", d, obsTimes)
  tibble::tibble(
    doseNmol = d,
    `mAb(0,0)` = trapAuc(parental$time, parental$Cc) / d,
    `Pb-Tx M1,S1` = trapAuc(masked$time, masked$Cc) / d
  )
}))
doseProp |>
  dplyr::mutate(dplyr::across(-doseNmol, ~ signif(.x, 4))) |>
  knitr::kable(caption = "Dose-normalised AUC (nM*s per nmol)")

parentalRise <- max(doseProp$`mAb(0,0)`) / min(doseProp$`mAb(0,0)`) - 1
maskedRise <- max(doseProp$`Pb-Tx M1,S1`) / min(doseProp$`Pb-Tx M1,S1`) - 1
c(parental_AUC_per_dose_rise = parentalRise, masked_AUC_per_dose_rise = maskedRise)
```

Over a 3.3-fold dose range the parental antibody's dose-normalised exposure rises
by `r round(100 * parentalRise)` %, the signature of a saturating target sink,
while the M1-masked Pb-Tx rises only `r round(100 * maskedRise)` %. This is a
direct, quantitative statement of the paper's "decreased importance of TMDD in
the disposition of the Pb-Tx species".

# Human projection: replicating Figure 4

Figure 4 reports plasma, peripheral and tumour profiles, and receptor-mediated
uptake in periphery and tumour, after a single 4.5 mg/kg dose, with mask strength
as the legend. The scan below spans the paper's own reported mask strengths
(57 for M2, 220 for M1) with an unmasked-like lower anchor and stronger masks
above.

```{r human-scan}
kmaskGrid <- c(1, 3, 10, 30, 57, 100, 220, 500, 1000, 3000)
humanScan <- dplyr::bind_rows(lapply(kmaskGrid, function(km) {
  s <- solveMolecule(hu, km, 3e-7, "pb_c_c_central", 1575, obsTimes)
  dplyr::mutate(s, kmask = km)
}))

humanSummary <- humanScan |>
  dplyr::group_by(kmask) |>
  dplyr::summarise(
    plasmaAuc = trapAuc(time, Cc),
    tumourAuc = trapAuc(time, pbTumor),
    uptakePeriph = dplyr::last(uptakePeripheral),
    uptakeTumour = dplyr::last(uptakeTumor),
    uptakeTumourOpen = dplyr::last(pb_tmdd_o_tumor),
    uptakeTumourCleaved = dplyr::last(pb_tmdd_m_tumor),
    .groups = "drop"
  )
humanSummary |>
  dplyr::mutate(dplyr::across(-kmask, ~ signif(.x, 4))) |>
  knitr::kable(caption = "Human projection over 28 days, single 1575 nmol dose")
```

```{r human-figure-4ac, fig.height = 3.6, fig.cap = "Replicates Figure 4a,c of Stroh 2019: plasma and tumour concentration by mask strength."}
humanScan |>
  dplyr::filter(Cc > 1e-3) |>
  dplyr::select(time, kmask, Plasma = Cc, Tumour = pbTumor) |>
  tidyr::pivot_longer(c(Plasma, Tumour), names_to = "compartment", values_to = "conc") |>
  dplyr::filter(conc > 1e-3) |>
  ggplot2::ggplot(ggplot2::aes(time / DAY, conc, colour = factor(kmask))) +
  ggplot2::geom_line(linewidth = 0.6) +
  ggplot2::facet_wrap(~compartment, scales = "free_y") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Time (days)", y = "Concentration (nM)", colour = "Kmask") +
  ggplot2::theme_bw()
```

Plasma and tumour exposure both rise monotonically with mask strength, matching
Figure 4a and 4c: "model-predicted Pb-Tx levels in the periphery and tumor
likewise increase with K_mask".

```{r human-figure-4de, fig.height = 3.6, fig.cap = "Replicates Figure 4d,e of Stroh 2019: cumulative receptor-mediated uptake at 28 days against mask strength."}
humanSummary |>
  dplyr::select(kmask, Periphery = uptakePeriph, Tumour = uptakeTumour) |>
  tidyr::pivot_longer(-kmask, names_to = "compartment", values_to = "uptake") |>
  ggplot2::ggplot(ggplot2::aes(kmask, uptake)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_point() +
  ggplot2::facet_wrap(~compartment, scales = "free_y") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(x = "Kmask (fold-masking)", y = "Cumulative receptor-mediated uptake (nmol)") +
  ggplot2::theme_bw()
```

Peripheral uptake falls monotonically with mask strength, as Figure 4d reports.
Tumour uptake is **not** monotone -- it turns, which is the Results claim, "the
relationship of increasing K_mask on receptor-mediated uptake in the tumor is not
monotonic and instead may pass through an optimum". The Discussion sharpens that
to a maximum, and this is the one published claim the extraction does not
reproduce; see Errata. Splitting the tumour uptake by which kind of arm delivered
the molecule to the receptor shows exactly why the curve turns:

```{r human-uptake-decomposition}
humanSummary |>
  dplyr::transmute(
    Kmask = kmask,
    `via open (masked) arm` = signif(uptakeTumourOpen, 3),
    `via cleaved arm` = signif(uptakeTumourCleaved, 3),
    Total = signif(uptakeTumour, 3)
  ) |>
  knitr::kable(caption = "Tumour receptor-mediated uptake at 28 days, nmol")

peripheryMonotone <- humanSummary$uptakePeriph[nrow(humanSummary)] <
  humanSummary$uptakePeriph[1]
interiorExtremum <- which.min(humanSummary$uptakeTumour)
tumourNonMonotone <- interiorExtremum > 1L && interiorExtremum < nrow(humanSummary)
c(
  periphery_falls_with_mask = peripheryMonotone,
  tumour_uptake_non_monotone = tumourNonMonotone,
  turning_point_kmask = kmaskGrid[interiorExtremum]
)
```

Uptake through a still-masked but transiently open arm falls steadily as the mask
tightens, while uptake through a cleaved arm rises, because a stronger mask keeps
the molecule in circulation long enough for more of it to be cleaved. The sum of
a falling and a rising channel turns once. Under the Table 2 parameter set the
rising channel wins at high mask strength, so the turn is a minimum near
`Kmask = `r kmaskGrid[interiorExtremum]``.

# Human projection: replicating Figure 5

Figure 5 projects circulating total, intact and cleaved Pb-Tx after repeated
3 mg/kg dosing, and concludes that intact Pb-Tx is the majority species.

```{r human-figure-5, fig.height = 3.8, fig.cap = "Replicates Figure 5 of Stroh 2019: projected circulating total, intact and cleaved Pb-Tx after repeated dosing."}
tau <- 21 * DAY
f5 <- solveMolecule(hu, 220, 3e-7, "pb_c_c_central", 1050,
  times = seq(0, 4 * tau, length.out = 500), tau = tau, nDose = 4L
)
f5 <- dplyr::filter(f5, Cc > 1e-6)

f5 |>
  dplyr::select(time,
    Total = Cc, Intact = pbIntactPlasma, Cleaved = pbCleavedPlasma
  ) |>
  tidyr::pivot_longer(-time, names_to = "species", values_to = "conc") |>
  ggplot2::ggplot(ggplot2::aes(time / DAY, conc, colour = species)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::labs(x = "Time (days)", y = "Plasma concentration (nM)", colour = NULL) +
  ggplot2::theme_bw()

intactFraction <- f5$pbIntactPlasma / f5$Cc
c(min = min(intactFraction), median = stats::median(intactFraction))
```

Intact Pb-Tx never drops below `r round(100 * min(intactFraction))` % of
circulating drug across four dosing cycles, reproducing "the QSP Pb-Tx model
suggests that the intact Pb-Tx comprises the majority of the total circulating
species in the plasma".

# Figure S2: cleaved species do not leak back from the tumour

Figure S2 reports that the net flux of cleaved Pb-Tx out of the tumour is at or
near zero, which is what makes the tumour-restricted-activation argument work.
The flux is computed here from the transport terms directly.

```{r figure-s2}
cleavedFree <- c("pb_c_m", "pb_o_m", "pb_m_m")
allFree <- c("pb_c_c", "pb_c_o", "pb_o_o", cleavedFree)
k13 <- exp(hu$iniDf$est[hu$iniDf$name == "lk13"])
k31 <- exp(hu$iniDf$est[hu$iniDf$name == "lk31"])
kelHuman <- exp(hu$iniDf$est[hu$iniDf$name == "lkel"])

netCleavedFlux <-
  k31 * rowSums(hSim[, paste0(cleavedFree, "_tumor"), drop = FALSE]) -
  k13 * rowSums(hSim[, paste0(cleavedFree, "_central"), drop = FALSE])
systemicElim <- kelHuman * rowSums(hSim[, paste0(allFree, "_central"), drop = FALSE])

fluxPct <- 100 * max(abs(netCleavedFlux)) / max(systemicElim)
c(
  max_abs_net_flux_nmol_per_s = max(abs(netCleavedFlux)),
  pct_of_peak_systemic_elimination = fluxPct
)
```

The largest net efflux of cleaved species from the tumour over the whole profile
is `r signif(fluxPct, 2)` % of the peak systemic elimination rate -- negligible,
as Figure S2 reports.

# Conclusions

```{r conclusions}
claim <- function(text, pass, achieved, deviation = FALSE) {
  tibble::tibble(Claim = text, Achieved = achieved, Pass = pass, Deviation = deviation)
}
conclusions <- dplyr::bind_rows(
  claim(
    "Model file reproduces the deposited A1 / A2 matrices (54 states)",
    gateRel < 1e-4, sprintf("max relative difference %.2g", gateRel)
  ),
  claim(
    "That gate can go red (0.1 % error in one rate constant)",
    mutRel > 100 * gateRel, sprintf("mutation control %.2g", mutRel)
  ),
  claim(
    "Drug mass balance closes across all species and sinks",
    massErr < 1e-8, sprintf("max relative error %.2g", massErr)
  ),
  claim(
    "CD166 pools start at ksynr / kint; plasma pool and its complexes are zero",
    ssErr < 1e-10 && plasmaBound < 1e-12,
    sprintf("steady-state error %.2g, plasma bound %.2g", ssErr, plasmaBound)
  ),
  claim(
    "Monkey ksynR reproduces Table 1's ~1e4 receptors/cell",
    rpc > 1e3 && rpc < 1e5, sprintf("%.3g receptors/cell", rpc)
  ),
  claim(
    "Figure 3: exposure rises with mask strength, parental mAb lowest",
    orderingHolds, "M1,0 > M1,S1 > M2,S2 > mAb(0,0)"
  ),
  claim(
    "Substrates S1 and S2 are indistinguishable (kcleave constrained equal)",
    substrateIdentical, "identical within dose scaling"
  ),
  claim(
    "TMDD saturates for the parental mAb but not for the masked Pb-Tx",
    parentalRise > 0.4 && maskedRise < 0.1,
    sprintf(
      "AUC/dose rises %.0f %% (parental) vs %.0f %% (masked)",
      100 * parentalRise, 100 * maskedRise
    )
  ),
  claim(
    "Figure 4a,c: plasma and tumour exposure rise with mask strength",
    all(diff(humanSummary$plasmaAuc) > 0) && all(diff(humanSummary$tumourAuc) > 0),
    "monotone increasing over Kmask 1-3000"
  ),
  claim(
    "Figure 4d: peripheral receptor-mediated uptake falls with mask strength",
    peripheryMonotone,
    sprintf(
      "%.0f -> %.0f nmol", humanSummary$uptakePeriph[1],
      humanSummary$uptakePeriph[nrow(humanSummary)]
    )
  ),
  claim(
    "Figure 4e (Results): tumour uptake is non-monotone in mask strength",
    tumourNonMonotone, sprintf("turns at Kmask = %g", kmaskGrid[interiorExtremum])
  ),
  claim(
    "Figure 4e (Discussion): that turning point is a MAXIMUM",
    FALSE, "reproduced as a minimum; see Errata",
    deviation = TRUE
  ),
  claim(
    "Figure 5: intact Pb-Tx is the majority of circulating drug",
    min(intactFraction) > 0.5, sprintf("minimum %.0f %% intact", 100 * min(intactFraction))
  ),
  claim(
    "Figure S2: net flux of cleaved species out of the tumour is ~ zero",
    fluxPct < 1, sprintf("%.2g %% of peak systemic elimination", fluxPct)
  )
)
knitr::kable(conclusions)
stopifnot(all(conclusions$Pass[!conclusions$Deviation]))
```

# Assumptions, deviations and errata

**Source of the ODE system.** The article body contains no equations beyond the
generic KroneckerBio form. The entire structure is taken from Supporting
Information `PSP4-8-676-s003` ("Model code"), obtained from the EuropePMC
supplementary-files endpoint for PMC6765697. A copy of its extracted text ships
with the package as `inst/references/Stroh_2019_probody_cd166_modelcode.txt` so
the structural gate above is reproducible.

**`kopen` is supplement-derived, not article-derived.** Tables 1 and 2 publish
only the fold-masking *ratio* `Kmask`. Simulating requires the absolute breathing
timescale, which is recovered from the deposited `A1` entry `(3,2) = 2.32e-2 =
2 * kopen`, giving `kopen = 1.16e-2 /s`, and is independently corroborated by
Ippolito 2024 Table S2 citing Stroh 2019 for the same number.

**The deposited matrices are not the Table 2 parameter set.** The deposited
instance uses `Kmask = 20`, `kcleave = 1e-6 /s` (above the article's stated
`< 3e-7` bound), `k13 = 2.5e-9` and `k31 = 4.4e-6 /s` (Table 2 gives `1.9e-8` and
`1.0e-5`). It is evidently one point from the parameter scans the paper reports
rather than the published human parameterisation. The model files therefore carry
the **Table 2 / Table 1 published values**, and the deposited instance is used
only in the structural gate, where the parameter values are irrelevant to what is
being tested.

**`k_off1`: the printed formula omits a factor of two.** Table 1 gives
`k off1 = k on1 * K app` with `k on1 = 1e-3` and `K app = 1.0 nM`, which evaluates
to `1e-3`, but the tabulated value is `2e-3`. The deposited `A1` matrix uses
`2e-3` throughout (entries `(1,8)`, `(1,9)`, `(1,11)` and so on). Two independent
sources therefore carry `2e-3` against one printed relation, and `K app` is an
*apparent* (bivalent) affinity whose relation to the monovalent off-rate
legitimately carries a statistical factor of two for a two-armed molecule. The
extraction uses `2e-3`.

**Plasma CD166 is carried but never synthesised.** The deposited code gives the
plasma compartment a full complement of receptor and bound states, but Figure 2b
draws no CD166 in plasma and the Results say provisions were included "for
binding to target within the peripheral and tumor compartments". The extraction
sets `ksynr_central` to a fixed zero, which makes the plasma block inert (verified
above at `r signif(plasmaBound, 2)` nmol) while keeping the state indexing
identical to the deposit. A user who wants to model soluble plasma CD166 can
simply raise that one parameter.

**`pb_el_tumor` is omitted.** The deposited state list names it, but `A1` gives it
no influx, because free species in the tumour are not eliminated. It would be an
identically-zero state, so the model file carries 54 live states rather than 55.

**No tumour compartment in the monkey model.** Table 1 reports no tumour volume,
perfusion rate or partition coefficient, and the animals were experimentally naive
rather than tumour-bearing. The monkey model is therefore plasma plus periphery.
Every tumour parameter in the human model comes from Table 2.

**Doses are entered in nmol, not mg/kg.** The paper never reports the Pb-Tx
molecular weight, so mg/kg cannot be converted to a molar amount from any on-disk
source, and substituting a class-typical antibody mass would be a guess. The
simulations above instead dose the molar amounts that place the 10 mg/kg monkey
arm near the `~1.3e3 nM` initial level of Figure 3c, scaling the other arms
proportionally. This is a presentation choice for the figures only: every
assertion in the Conclusions table is a ratio, an ordering, a mass balance or a
structural identity, so none of them depends on it.

**Deviation -- Figure 4e curvature.** The Results describe tumour
receptor-mediated uptake as "not monotonic and instead may pass through an
optimum", which is reproduced. The Discussion sharpens this to "concave-down and
pass through a maxima", which is **not** reproduced: under the Table 2 parameter
set, and under every alternative sourced from the paper that was tried (no
cleavage; 100-fold rather than 10-fold tumour `kcleave`; the deposited instance),
the turning point is a *minimum*. The decomposition table above shows the
mechanism: uptake via a still-masked open arm falls monotonically with `Kmask`
(5.5 to 0.06 nmol) while uptake via a cleaved arm rises monotonically (1.0 to
5.0 nmol), and a falling plus a rising channel gives a minimum. Obtaining the
published maximum would require the total tumour Pb-Tx amount to grow faster with
`Kmask` than the open fraction shrinks over the low-to-mid range, and here it does
not -- tumour exposure rises about eight-fold across the scan while the open
fraction falls roughly as `1/Kmask`. The published Figure 4e was produced with an
unstated `Kmask` grid and, as noted above, the deposited code does not carry the
Table 2 values, so the disagreement cannot be resolved from the sources on disk.
It is recorded here rather than tuned away.

**No variability.** The source model is deterministic: no between-subject
variability and no residual-error model are reported, and none is invented. Both
model files therefore carry no `eta` terms and no error model, and `rxSolve` is
called without `omega`.

**Unit note on the bimolecular terms.** Table 1's own receptors-per-cell
arithmetic (`RPC = k_synR / k_endo * 1e-9 * Avog / Ncell`) shows that
`ksynr / kint` is an *amount* in nmol, so the states are amounts. The tabulated
`k_on` values are nevertheless labelled `1/(nM*second)` and are used in the
deposited matrices without any volume division. The extraction reproduces the
deposited arithmetic exactly and flags the labelling inconsistency here rather
than silently rescaling it, which would change the model.
