---
title: "Tigecycline (Su 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tigecycline (Su 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Su_2024_tigecycline"))$reference`
* Description: `r rxode2::rxode(readModelDb("Su_2024_tigecycline"))$description`
* Article: <https://doi.org/10.3389/fphar.2024.1342947>
* Supplement (Tables S1-S3, Figures S1-S3): <https://www.frontiersin.org/articles/10.3389/fphar.2024.1342947/full#supplementary-material>

# Population

Su and colleagues ran a single-centre prospective study at Sir Run Run Shaw
Hospital (Hangzhou, China) between July 2019 and July 2023, enrolling 98
critically ill adults who received intermittent intravenous tigecycline as a
30-minute infusion every 12 h. Sampling was intensive rather than opportunistic:
eight samples per patient, drawn immediately before the seventh dose and at 0.5,
1, 2, 3, 4, 6 and 12 h afterwards, so every subject contributed a complete
steady-state dosing interval. That yielded 751 concentrations for model
building.

The cohort was 66.3% male with a mean age of 63.4 years (SD 18.0, range 20-92)
and a mean weight of 62.3 kg (SD 12.4, range 38-92.5). Renal function spanned
almost two orders of magnitude  -  Cockcroft-Gault creatinine clearance median
77.0 mL/min, IQR 51.1-142, range 6.40-338  -  while patients on haemodialysis,
peritoneal dialysis or CRRT were excluded. Liver chemistry was likewise
dispersed: albumin median 27.9 g/L (IQR 25.8-30.3), GGT median 53 U/L (IQR
30-103) and total bilirubin median 15.7 umol/L (IQR 10.0-30.4). Infections were
predominantly intra-abdominal (37.8%) and pulmonary (31.6%), most often caused
by carbapenem-resistant *A. baumannii* (46.9%) or *K. pneumoniae* (24.5%).
Almost all patients (92.9%) received 100 mg/day; 7.14% received 200 mg/day.
Baseline demographics are Table 1 of the paper.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Su_2024_tigecycline")()$population`).

# Source trace

Every `ini()` entry carries an in-file comment pointing at its source location
in `inst/modeldb/specificDrugs/Su_2024_tigecycline.R`. The table below collects
them for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment structure, first-order elimination | n/a | Results, "PPK model development" (AIC 8242.914 two-compartment vs 9325.681 one-compartment) |
| 30-minute IV infusion, q12h | n/a | Methods, "Patient inclusion, drug administration and sample collection" |
| `lcl` (CL intercept) | 3.09 L/h | Table 2, "CL (L/h)"; final-model equation `CL (L/h) = 3.09+(CCr/77) x 3.28` |
| `e_crcl_cl` | 3.28 L/h per (CCr/77) | Table 2, `theta CLCR-CL` |
| `lvc` (V1 at 61 kg) | 32.1 L | Table 2, "V1 (L)"; final-model equation `V1 (L) = 32.1x(BW/61)^1.95` |
| `e_wt_vc` | 1.95 | Table 2, `theta BW-V1` |
| `lq` (Q at reference liver chemistry) | 39.7 L/h | Table 2, "Q (L/h)"; final-model equation `Q (L/h) = 39.7x(logGGT/1.7)^0.956 x (logTBIL/1.2)^-0.912` |
| `e_ggt_q` | 0.956 | Table 2, `theta GGT-Q` |
| `e_tbili_q` | -0.912 | Table 2, `theta TBIL-Q` |
| `lvp` (V2 at 61 kg, reference albumin) | 113 L | Table 2, "V2 (L)"; final-model equation `V2 (L) = 113x(BW/61)^1.61 x (logALB/1.4)^4.52` |
| `e_wt_vp` | 1.61 | Table 2, `theta BW-V2` |
| `e_alb_vp` | 4.52 | Table 2, `theta ALB-V2` |
| Linear covariate form `theta_pop + theta_cov * cov/cov_median` | n/a | Methods Eq. 1 (used for CCr on CL) |
| Power covariate form `theta_pop * (cov/cov_median)^theta_cov` | n/a | Methods Eq. 3 (used for the four remaining effects) |
| `etalcl` | 27.0% CV | Table 2, `omega CL (%)` |
| `etalvc` | 72.5% CV | Table 2, `omega V1 (%)` |
| `etalq` | 18.8% CV | Table 2, `omega Q (%)` |
| No IIV on V2 | n/a | Results, "it was not possible to estimate the interindividual variability for V2" |
| `propSd` | 0.142 | Table 2, `sigma (%) 2.02`, read as a variance (see Assumptions) |
| Covariate reference values 77 mL/min, 61 kg, 1.7, 1.2, 1.4 | n/a | Divisors printed inside the four final-model equations |
| PK/PD targets AUC/MIC >= 6.96 / 12.8 / 17.9 | n/a | Methods, "Simulation and dosing regimen optimization" |
| PTA strata CCr 30 / 80 / 130 mL/min | n/a | Methods, "Simulation and dosing regimen optimization" |

# Virtual cohort

The original concentrations are not public, so the checks below use virtual
cohorts whose covariate distributions reproduce the Table 1 marginals. Skewed
laboratory covariates are drawn log-normally with the log-scale SD implied by
the published inter-quartile range (`sd = log(IQR ratio) / 1.349`); body weight
is drawn normally from its published mean and SD. Every draw is truncated to
the observed range in Table 1.

Five cohorts of 200 subjects each are built. Four are dosed to *exact* steady
state (rxode2's `ss = 1`): one with the full covariate distribution and three
with creatinine clearance pinned at 30, 80 and 130 mL/min (the strata of
Figure 4). These carry the NCA and PTA work, where steady state is the intended
condition. The fifth replicates the paper's actual sampling occasion - the
seventh q12h dose, i.e. 72 h of therapy, which is *not* steady state for the
slower subjects - and is used only for the Figure 3 comparison, so that
simulated and observed profiles are compared at the same accumulation state.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned PER SOLVER THREAD, so the cohort below is
# reproducible on this machine and different on a machine with a different
# thread count. Every assertion downstream is written to hold for ANY cohort
# the model can produce; see pattern 12 of the skill's
# known-vignette-failure-patterns reference.
set.seed(20240129)
rxode2::rxSetSeed(20240129)

n_per_arm <- 200L

# Table 1 marginals. sd_log is recovered from the published IQR.
sd_log <- function(q1, q3) log(q3 / q1) / 1.349

draw_covariates <- function(n, crcl = NULL) {
  tibble::tibble(
    CRCL  = if (is.null(crcl)) {
      pmax(6.4, pmin(338, exp(rnorm(n, log(77), sd_log(51.1, 142)))))
    } else {
      rep(crcl, n)
    },
    WT    = pmax(38, pmin(92.5, rnorm(n, 62.3, 12.4))),
    ALB   = pmax(17.3, pmin(38.3, exp(rnorm(n, log(27.9), sd_log(25.8, 30.3))))),
    GGT   = pmax(8,    pmin(1088, exp(rnorm(n, log(53),   sd_log(30, 103))))),
    TBILI = pmax(4,    pmin(144,  exp(rnorm(n, log(15.7), sd_log(10.0, 30.4)))))
  )
}

# Steady state is imposed EXACTLY, with rxode2's `ss = 1` dosing flag, rather
# than approached by stacking a finite number of doses. That matters here: the
# peripheral volume is large (113 L at the reference covariates) and IIV on V1
# is 72.5%, so a subject in the tail of the eta distribution has a terminal
# half-life of several days and is still ~50% short of steady state after nine
# q12h doses. Accumulation shortfall is a PHYSICAL, per-subject difference, so
# it would silently corrupt the AUC identity below and vary with whichever
# cohort the machine happens to draw. `ss = 1` removes it entirely.
tad_obs <- c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)

# `n_doses = NULL` gives exact steady state; an integer instead stacks that many
# q12h doses and samples around the last one, which is what the paper did.
make_cohort <- function(n, stratum, crcl = NULL, id_offset = 0L, n_doses = NULL) {
  cov <- draw_covariates(n, crcl) |>
    dplyr::mutate(id = id_offset + seq_len(n), stratum = stratum)
  t_ref <- if (is.null(n_doses)) 0 else (n_doses - 1) * 12
  doses <- cov |>
    dplyr::mutate(time = 0, amt = 50, evid = 1L, dur = 0.5, ii = 12,
                  addl = if (is.null(n_doses)) NA_integer_ else n_doses - 1L,
                  ss   = if (is.null(n_doses)) 1L else 0L,
                  cmt  = "central")
  obs <- cov |>
    tidyr::crossing(time = t_ref + tad_obs) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_,
                  ii = NA_real_, addl = NA_integer_, ss = NA_integer_,
                  cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(t_ref = t_ref) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

ss_strata   <- c("Full covariate distribution", "CCr 30 mL/min",
                 "CCr 80 mL/min", "CCr 130 mL/min")
crcl_strata <- c("CCr 30 mL/min", "CCr 80 mL/min", "CCr 130 mL/min")
dose7_stratum <- "Seventh dose (paper sampling)"

events <- dplyr::bind_rows(
  make_cohort(n_per_arm, "Full covariate distribution", NULL, id_offset =   0L),
  make_cohort(n_per_arm, "CCr 30 mL/min",                30, id_offset = 200L),
  make_cohort(n_per_arm, "CCr 80 mL/min",                80, id_offset = 400L),
  make_cohort(n_per_arm, "CCr 130 mL/min",              130, id_offset = 600L),
  # Replicates the paper's own sampling occasion: the seventh q12h dose, i.e.
  # 72 h of therapy, NOT steady state. Used only for the Figure 3 comparison so
  # that simulated and observed profiles carry the same accumulation state.
  make_cohort(n_per_arm, dose7_stratum, NULL, id_offset = 800L, n_doses = 7L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Su_2024_tigecycline")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("stratum", "WT", "CRCL", "ALB", "GGT", "TBILI", "t_ref")
) |>
  as.data.frame() |>
  dplyr::mutate(tad = time - t_ref,
                Cc_ugL = Cc * 1000)   # model works in mg/L; the paper plots ug/L

# At steady state the trough must repeat: C(0) == C(12) for every subject.
ss_check <- sim |>
  dplyr::filter(stratum %in% ss_strata, tad %in% c(0, 12)) |>
  tidyr::pivot_wider(id_cols = c(stratum, id), names_from = tad,
                     values_from = Cc, names_prefix = "t") |>
  dplyr::mutate(pct = 100 * (t12 - t0) / t0)
stopifnot(max(abs(ss_check$pct)) < 1e-6)
```

## Structural identity check

Before any cohort-level comparison, confirm that the typical-value parameters at
the paper's reference covariates reproduce the Table 2 estimates exactly. These
are deterministic, so the tolerance is tight.

```{r structural-identity}
ref <- tibble::tibble(CRCL = 77, WT = 61, ALB = 10^1.4, GGT = 10^1.7, TBILI = 10^1.2)

typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = tibble::tibble(id = 1L, time = 0, amt = 50, evid = 1L, dur = 0.5,
                          cmt = "central") |>
    dplyr::bind_rows(tibble::tibble(id = 1L, time = c(0, 1), amt = NA_real_,
                                    evid = 0L, dur = NA_real_, cmt = "central")) |>
    dplyr::bind_cols(ref[rep(1, 3), ]) |>
    dplyr::arrange(time, dplyr::desc(evid))
) |> as.data.frame()

structural <- tibble::tibble(
  Parameter = c("CL at CCr 77 (L/h)", "V1 at 61 kg (L)",
                "Q at reference liver chemistry (L/h)", "V2 at 61 kg, reference ALB (L)"),
  Published = c(3.09 + 3.28, 32.1, 39.7, 113),
  Model     = c(typ$cl[1], typ$vc[1], typ$q[1], typ$vp[1])
) |>
  dplyr::mutate(`% difference` = 100 * (Model - Published) / Published)

knitr::kable(structural, digits = 3,
             caption = "Typical-value parameters at the paper's reference covariates.")

# Deterministic - a mis-transcribed value or exponent breaks this immediately.
stopifnot(max(abs(structural$`% difference`)) < 0.5)
```

Note the first row: the paper's Abstract and Results call 3.09 L/h "the typical
value of CL", but the printed final-model equation is
`CL = 3.09 + (CCr/77) x 3.28`, so 3.09 L/h is the *intercept* and the typical
patient at the median CCr of 77 mL/min clears 6.37 L/h. The equation governs;
see Assumptions below.

# Replicate published figures

## Figure 3  -  steady-state concentration-time profile

Figure 3 of Su 2024 is a visual predictive check over one dosing interval, with
the observed median and 5th / 95th percentiles drawn on a log concentration
axis in ug/L. Because the underlying observations are not public, the published
percentiles below were read off the rendered figure and are therefore accurate
to roughly +/-10%. They are used only for a wide order-of-magnitude gate (a
factor-of-two error in dose or volume breaks it); nothing is tuned to them.

This comparison uses the seventh-dose cohort, not the exact-steady-state one,
because Su 2024 sampled after 72 h of therapy and with 72.5% IIV on V1 a
sizeable minority of subjects have not accumulated fully by then. Comparing an
exact-steady-state simulation against those observations would bias the model
side upwards.

```{r figure-3}
vpc <- sim |>
  dplyr::filter(stratum == dose7_stratum) |>
  dplyr::group_by(tad) |>
  dplyr::summarise(Q05 = quantile(Cc_ugL, 0.05),
                   Q50 = quantile(Cc_ugL, 0.50),
                   Q95 = quantile(Cc_ugL, 0.95), .groups = "drop")

ggplot(vpc, aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "firebrick", linewidth = 1) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Tigecycline concentration (ug/L)",
       title = "Steady-state dosing interval, 50 mg q12h",
       caption = "Replicates Figure 3 of Su 2024 (median with 5th-95th percentile band).")
```

```{r figure-3-table}
digitised <- tibble::tribble(
  ~tad, ~pub_Q05, ~pub_Q50, ~pub_Q95,
  0.5,       680,     1600,     3300,
  2,         255,      570,     1400,
  4,         210,      440,     1180,
  6,         195,      420,     1120,
  12,        100,      320,      750
)

fig3 <- digitised |>
  dplyr::left_join(vpc, by = "tad") |>
  dplyr::mutate(`Median ratio (model / paper)` = Q50 / pub_Q50,
                `Spread sd(log C), model` = log(Q95 / Q05) / 3.29,
                `Spread sd(log C), paper` = log(pub_Q95 / pub_Q05) / 3.29)

fig3 |>
  dplyr::select(tad, pub_Q50, Q50, `Median ratio (model / paper)`,
                `Spread sd(log C), paper`, `Spread sd(log C), model`) |>
  dplyr::rename("TAD (h)" = tad,
                "Paper median (ug/L)" = pub_Q50,
                "Model median (ug/L)" = Q50) |>
  knitr::kable(digits = c(1, 0, 0, 2, 3, 3),
               caption = "Digitised Figure 3 percentiles vs the simulated cohort. Published values were read off the rendered figure (+/-10%).")
```

The medians agree to within about a third across the interval, and the model's
dispersion is of the right order. Two systematic differences are visible and are
recorded as deviations rather than tuned away. First, the simulated profile is
**flatter** than the observed one: the model matches the peak (ratio 0.87) but
over-predicts the 12 h trough by roughly 35%. Second, the simulated 5th-95th
spread is narrower than the observed spread at every time point - `sd(log C)` of
roughly 0.35-0.52 against 0.48-0.61 - with the gap largest early in the
interval. The second of these bears directly on the Table 2 variability block,
whose reported scale is ambiguous; note that the *wider* (variance) reading of
the `omega` rows would close this particular gap while over-shooting elsewhere.
See Assumptions.

```{r figure-3-gate}
# Loose, but still red-capable: a mis-transcribed volume (32.1 vs 113 L) or a
# dose error (50 vs 100 mg) moves these medians by a factor of 2 or more.
stopifnot(all(fig3$`Median ratio (model / paper)` > 0.6),
          all(fig3$`Median ratio (model / paper)` < 1.7))
```

## Figure 4  -  probability of target attainment

Figure 4 reports PTA against the three CART-derived PK/PD targets (AUC/MIC
>= 6.96 for complicated intra-abdominal infection, >= 12.8 for
community-acquired pneumonia and >= 17.9 for complicated skin and skin structure
infection), stratified by creatinine clearance, for the 50 mg q12h and
100 mg q12h regimens.

Steady-state exposure in a linear model is exactly dose-proportional, so the
100 mg q12h arm is obtained by doubling the simulated 50 mg AUC rather than by
simulating a second cohort.

```{r pta}
auc_ss <- sim |>
  dplyr::filter(stratum %in% crcl_strata) |>
  dplyr::group_by(stratum, id) |>
  dplyr::arrange(tad, .by_group = TRUE) |>
  dplyr::summarise(
    # Linear-up / linear-down trapezoid over the steady-state interval, then
    # scaled to 24 h; the paper's targets are AUC0-24 / MIC ratios.
    auc24_50mg = 2 * sum(diff(tad) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(auc24_100mg = 2 * auc24_50mg)

mic_grid <- c(0.125, 0.25, 0.5, 1, 2, 4, 8)
targets  <- c(`AUC/MIC >= 6.96 (cIAI)` = 6.96,
              `AUC/MIC >= 12.8 (CAP)`  = 12.8,
              `AUC/MIC >= 17.9 (cSSSI)` = 17.9)

pta <- tidyr::expand_grid(
  stratum = unique(auc_ss$stratum),
  regimen = c("50 mg q12h", "100 mg q12h"),
  MIC     = mic_grid,
  target_label = names(targets)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    PTA = {
      a <- auc_ss$auc24_50mg[auc_ss$stratum == stratum]
      if (regimen == "100 mg q12h") a <- 2 * a
      100 * mean(a / MIC >= targets[[target_label]])
    }
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(stratum = factor(stratum, levels = c("CCr 30 mL/min", "CCr 80 mL/min",
                                                     "CCr 130 mL/min")))

ggplot(pta, aes(MIC, PTA, colour = stratum, linetype = regimen)) +
  geom_line() + geom_point(size = 1) +
  geom_hline(yintercept = 90, linetype = "dotted") +
  scale_x_log10(breaks = mic_grid) +
  facet_wrap(~target_label) +
  labs(x = "MIC (mg/L)", y = "PTA (%)", colour = NULL, linetype = NULL,
       caption = "Replicates Figure 4 of Su 2024 (panels A-C, stratified by creatinine clearance).") +
  theme(legend.position = "bottom")
```

```{r pta-monotonic}
# Deterministic consequence of the CL equation: exposure falls as CCr rises.
med_auc <- auc_ss |>
  dplyr::group_by(stratum) |>
  dplyr::summarise(median_auc24 = median(auc24_50mg), .groups = "drop")

knitr::kable(med_auc, digits = 2,
             caption = "Median steady-state AUC0-24 (mg*h/L) at 50 mg q12h by CCr stratum.")

ratio_30_130 <- med_auc$median_auc24[med_auc$stratum == "CCr 30 mL/min"] /
  med_auc$median_auc24[med_auc$stratum == "CCr 130 mL/min"]
# CL(30)/CL(130) = (3.09 + 3.28*30/77) / (3.09 + 3.28*130/77) = 4.368 / 8.627,
# so the exposure ratio is ~1.98 by construction and only cohort noise moves it.
stopifnot(ratio_30_130 > 1.7, ratio_30_130 < 2.3)
```

# PKNCA validation

PKNCA is used for the steady-state interval NCA. The strongest available gate
here is an **internal identity**: for a linear model at steady state, the AUC
over one dosing interval must equal dose / CL exactly. The paper reports no NCA
table of its own, so this identity  -  plus the structural check above  -  is what
pins the transcription.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(stratum %in% ss_strata, !is.na(Cc)) |>
  dplyr::select(id, stratum, time, Cc)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | stratum + id)

dose_df <- events |>
  dplyr::filter(evid == 1, stratum %in% ss_strata) |>
  dplyr::select(id, stratum, time, amt)

dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | stratum + id)

intervals <- data.frame(
  start   = 0,
  end     = 12,
  auclast = TRUE,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tab <- as.data.frame(nca_res)

nca_summary <- nca_tab |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "tmax", "cmin")) |>
  tidyr::pivot_wider(id_cols = c(stratum, id), names_from = PPTESTCD,
                     values_from = PPORRES)

nca_summary |>
  dplyr::group_by(stratum) |>
  dplyr::summarise(dplyr::across(c(auclast, cmax, tmax, cmin), median), .groups = "drop") |>
  dplyr::rename("Stratum" = stratum,
                "AUC0-12,ss (mg*h/L)" = auclast,
                "Cmax,ss (mg/L)" = cmax,
                "Tmax (h)" = tmax,
                "Cmin,ss (mg/L)" = cmin) |>
  knitr::kable(digits = 3,
               caption = "Median steady-state NCA parameters at 50 mg q12h, by CCr stratum.")
```

```{r pknca-identity}
# AUC0-12,ss must equal dose / CL for every individual. Because steady state is
# imposed exactly (`ss = 1`), the only residual difference is trapezoidal
# discretisation of a convex curve, which is one-sided: every subject should
# come out slightly HIGH, by well under a percent.
cl_by_id <- sim |>
  dplyr::filter(stratum %in% crcl_strata) |>
  dplyr::group_by(stratum, id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

identity_chk <- nca_summary |>
  dplyr::inner_join(cl_by_id, by = c("stratum", "id")) |>
  dplyr::mutate(expected = 50 / cl,
                pct_diff = 100 * (auclast - expected) / expected)

knitr::kable(
  identity_chk |>
    dplyr::group_by(stratum) |>
    dplyr::summarise(`Median % difference` = median(pct_diff),
                     `Min % difference`    = min(pct_diff),
                     `Max % difference`    = max(pct_diff), .groups = "drop") |>
    dplyr::rename("Stratum" = stratum),
  digits = 3,
  caption = "PKNCA AUC0-12,ss against the closed-form dose / CL. Pure trapezoidal error."
)

# Both sides use the same drawn parameters and steady state is exact, so this is
# numerical error only and a tight two-sided bound is correct (see the repository
# note on cohort-extreme assertions). It is not quite one-sided: the trapezoid
# over-estimates the convex post-peak decline but under-estimates the concave
# rise during the 0.5 h infusion, and which term wins depends on the subject.
stopifnot(max(abs(identity_chk$pct_diff)) < 2)
```

# Published dosing claims

Su 2024 states three quantitative conclusions from the Monte Carlo simulation.
Each is checked below against the packaged model at the median-CCr stratum,
which is the closest match to the paper's typical patient.

```{r claims}
pta_at <- function(stratum_, regimen_, mic, target_label_) {
  v <- pta$PTA[pta$stratum == stratum_ & pta$regimen == regimen_ &
                 pta$MIC == mic & pta$target_label == target_label_]
  if (length(v) != 1L) stop("no unique PTA row for ", stratum_, " / ", regimen_,
                            " / MIC ", mic, " / ", target_label_)
  v
}

claims <- tibble::tribble(
  ~Claim, ~Source, ~Achieved, ~Pass, ~Deviation,
  "50 mg q12h attains the cIAI target (6.96) with PTA >= 90% at MIC 1",
  "Results, 'sufficient to achieve an AUC/MIC ratio of 6.96 at MICs <= 1 mg/L'",
  pta_at("CCr 80 mL/min", "50 mg q12h", 1, "AUC/MIC >= 6.96 (cIAI)"), NA, FALSE,

  "50 mg q12h does NOT attain the cIAI target at MIC 2",
  "Results, same sentence (attainment limited to MIC <= 1 mg/L)",
  pta_at("CCr 80 mL/min", "50 mg q12h", 2, "AUC/MIC >= 6.96 (cIAI)"), NA, FALSE,

  "50 mg q12h does NOT attain the CAP target (12.8) at MIC 1",
  "Results, 'may attain the suboptimal target at MICs > 0.5' for AUC/MIC 12.8",
  pta_at("CCr 80 mL/min", "50 mg q12h", 1, "AUC/MIC >= 12.8 (CAP)"), NA, FALSE,

  "100 mg q12h restores the CAP target at MIC 1 (PTA >= 90%)",
  "Conclusion, '100 mg every 12 h was needed for community-acquired pneumonia'",
  pta_at("CCr 80 mL/min", "100 mg q12h", 1, "AUC/MIC >= 12.8 (CAP)"), NA, FALSE,

  "50 mg q12h does NOT attain the cSSSI target (17.9) at MIC 0.5",
  "Results, 'may attain the suboptimal target at ... 0.25 mg/L' for AUC/MIC 17.9",
  pta_at("CCr 80 mL/min", "50 mg q12h", 0.5, "AUC/MIC >= 17.9 (cSSSI)"), NA, TRUE
)

claims$Pass <- c(
  claims$Achieved[1] >= 90,
  claims$Achieved[2] <  90,
  claims$Achieved[3] <  90,
  claims$Achieved[4] >= 90,
  claims$Achieved[5] <  90
)

claims |>
  dplyr::mutate(Achieved = sprintf("%.1f%% PTA", Achieved)) |>
  dplyr::rename("Achieved (model)" = Achieved, "Known deviation" = Deviation) |>
  knitr::kable(caption = "Published dosing conclusions checked against the packaged model.")

# Gate on the claims that reproduce; the flagged deviation stays visible in the
# table but is excluded rather than having the bound widened until it passes.
stopifnot(all(claims$Pass[!claims$Deviation]))
```

The one deviation is the cSSSI boundary. Su 2024 places the 50 mg q12h
attainment ceiling for the 17.9 target at MIC 0.25 mg/L; the packaged model puts
it at 0.5 mg/L, i.e. one two-fold dilution more permissive. The direction is
consistent with the discrepancies noted for Figure 3: the transcribed model
carries less exposure variability than the published analysis did, which lifts
PTA in the tail of the MIC grid. The paper's headline recommendation  -  50 mg
q12h for complicated intra-abdominal infection, 100 mg q12h for pneumonia and
skin and skin structure infections  -  is reproduced.

# Assumptions and deviations

- **`CL = 3.09 L/h` is an intercept, not a typical value.** The Abstract, the
  Results and the Conclusion all describe 3.09 L/h as "the typical value of CL",
  but Methods Eq. 1 defines the linear covariate model as
  `theta_i = theta_pop + theta_cov * cov_i / cov_median` and the printed
  final-model equation is `CL = 3.09 + (CCr/77) x 3.28`. Evaluated at the
  cohort's median CCr of 77 mL/min this gives 6.37 L/h, and the equation is
  encoded as printed. Three independent lines of evidence support the equation
  over the prose. First, the other three parameters are internally consistent
  with their equations (V1, Q and V2 all reduce to their Table 2 values at the
  reference covariates), so CL is the only row where "Table 2 value" and
  "typical value" would differ  -  exactly what an intercept does. Second,
  Supplementary Table S1 tabulates ten previously published tigecycline models
  with CL between 4.81 and 23.1 L/h; 6.37 L/h sits at the low end, consistent
  with the paper's own claim that its CL was "lower than those estimated in
  previously published studies", whereas 3.09 L/h would be far below every one
  of them. Third, the paper's PTA conclusions (Figure 4) are attainable only
  with the higher clearance: at 3.09 L/h the 50 mg q12h regimen would deliver an
  AUC0-24 of 32 mg*h/L and would comfortably clear the 6.96 target at MIC 2,
  contradicting the stated MIC <= 1 mg/L ceiling. The same reading  -  Table 2
  reporting the intercept of an additive linear renal-clearance model  -  is used
  in `Delattre_2010_amikacin`, which has the identical covariate structure.
- **`log` in the Q and V2 equations is base 10.** The paper writes `logGGT`,
  `logTBIL` and `logALB` without stating the base. The divisors printed in the
  equations settle it: 1.7, 1.2 and 1.4 against the Table 1 medians of 53 U/L,
  15.7 umol/L and 27.9 g/L give `log10` values of 1.72, 1.20 and 1.45, while the
  natural logarithms are 3.97, 2.75 and 3.33. All three agree on base 10, so it
  is not a coincidence of rounding on any single covariate. The normalizer is
  the median of the log-transformed covariate, not the log of the median
  covariate  -  the equations divide `logGGT` by 1.7, not by `log10(53)` written
  out  -  but for these values the two are the same number to the printed
  precision.
- **The residual-error scale in Table 2 is ambiguous, and the literal reading is
  falsified.** The variability block prints `omega CL (%) 27.0`,
  `omega V1 (%) 72.5`, `omega Q (%) 18.8` and `sigma (%) 2.02`. Read uniformly
  as coefficients of variation, the first three are ordinary popPK values but
  the last is a 2% proportional residual error, which is not attainable for
  clinical LC-MS/MS concentration data. The paper's own Figure 2 rules it out
  directly: a 2.02% proportional CV puts the 95% band of DV about IPRED at
  +/-4%, whereas digitising the DV-vs-IPRED panel gives a band of roughly
  +/-20-35% in the densely sampled region  -  an order of magnitude wider, and far
  beyond the plotting-symbol size. The value is therefore encoded as the
  proportional-error **variance** expressed in percent, `sigma^2 = 0.0202`, i.e.
  `propSd = sqrt(0.0202) = 0.142`, which reproduces the observed scatter. The
  three `omega` rows are encoded as coefficients of variation via the repository
  convention `omega^2 = log(CV^2 + 1)`, matching how the same style of table is
  handled in `Delattre_2010_amikacin`. This mixed reading is the best-supported
  one, but it is a reading: applying the variance convention to the `omega` rows
  as well would give SDs of 0.52 / 0.85 / 0.43 instead of 0.27 / 0.72 / 0.19.
  Digitising the Figure 3 VPC does not cleanly separate the two  -  the
  coefficient-of-variation reading is slightly too narrow at late times and the
  variance reading is clearly too wide  -  and digitising the Figure 4 PTA slopes
  gives an intermediate value. Any downstream use that depends on the magnitude
  of the between-subject variability (rather than on typical-value exposure)
  should treat this as uncertain.
- **Steady state is imposed exactly, not approached.** Su 2024 sampled around
  the seventh dose, i.e. after 72 h of q12h therapy. Reproducing that literally
  by stacking doses is not reproducible across cohorts: IIV on V1 is 72.5% and
  V2 is large, so a subject in the tail of the eta distribution has a terminal
  half-life of several days and is still tens of percent short of steady state
  at 72 h - or even at 480 h. The vignette therefore uses rxode2's `ss = 1`
  dosing flag, which solves the exact periodic steady state for each subject's
  own parameters. This makes the AUC identity a statement about numerical
  quadrature rather than about accumulation, which is what it is meant to test,
  and it matches the steady-state condition the paper's own Monte Carlo
  simulation assumes. The one comparison against published *observations* -
  Figure 3 - deliberately does **not** use it, and instead replicates the
  seventh-dose occasion the paper actually sampled, so that both sides carry the
  same accumulation state.
- **No IIV on V2.** The paper states plainly that "it was not possible to
  estimate the interindividual variability for V2", so V2 carries no eta. This
  is a reported structural fact, not an omission.
- **Covariate reference values.** 77 mL/min, 61 kg, 1.7, 1.2 and 1.4 are taken
  from the divisors inside the printed final-model equations. Table 1 reports
  body weight as a mean (62.3 kg) rather than a median, so 61 kg is known only
  from the equation; the two are close enough that the distinction does not
  matter numerically.
- **Screened but not retained.** Supplementary Table S3 records that daily dose
  on V1 and total bilirubin on CL both entered forward inclusion and were then
  removed in backward elimination, and that age, BMI, sex, BUN, ALT, AST and ALP
  were screened without entering. Only the five covariates of the final model
  are encoded.
- **Virtual cohort.** Covariate distributions are reconstructed from the Table 1
  marginals (mean and SD for weight, median and IQR for the laboratory values)
  under independence. The real cohort's covariates are certainly correlated  - 
  albumin, bilirubin and GGT co-vary with illness severity  -  so the simulated
  spread of V2 and Q is wider than the study's would have been.
- **Concentration units.** The model works in mg/L (mg dose over L volume);
  Su 2024 plots concentrations in ug/L, i.e. 1000 x the model's `Cc`. The
  vignette carries an explicit `Cc_ugL` column for figure comparisons.
- **No published NCA table.** Su 2024 reports no observed Cmax / AUC / half-life
  summary, so there is no external NCA comparison to make. The PKNCA section
  gates on the internal steady-state identity AUC0-12,ss = dose / CL instead.
