---
title: "Voriconazole (Suetsugu 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Voriconazole (Suetsugu 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Suetsugu_2021_voriconazole"))$reference`
* Description: `r rxode2::rxode(readModelDb("Suetsugu_2021_voriconazole"))$description`
* Article: <https://doi.org/10.1007/s40268-021-00365-0> (open access, PMC8602551)

Suetsugu et al. analysed routine voriconazole therapeutic-drug-monitoring
troughs from adult allogeneic haematopoietic stem cell transplant (allo-HSCT)
recipients to quantify the drug-drug interaction with letermovir, a
cytomegalovirus prophylactic that induces CYP2C19 and CYP2C9. Because every
observation was a steady-state trough, the authors did not fit a
compartmental model. They fitted the steady-state Michaelis-Menten relation
between the daily maintenance dose and the trough concentration directly
(Eq. 1):

$$C_{ss,\,trough} = \frac{K_m \times F \times \text{Daily dose}}{V_{max} - F \times \text{Daily dose}}$$

with $F$ fixed to 1 and $V_{max,i} = 670 \times 1.72^{LMV_i} \times 1.30^{mPSL_i}$
(Eq. 4), where `LMV` and `mPSL` flag concomitant letermovir and
methylprednisolone.

The packaged model is therefore **algebraic**: it has no time dimension, no
compartments and no dosing events. The daily dose enters as the covariate
`DOSE_VORI_MGD` (mg/day), and each observation row returns the steady-state
trough `Cc` (mg/L) for that dose level and co-medication state.

# Population

The model was fitted to 216 steady-state voriconazole troughs from 47 Japanese
adults (23 male, 24 female; median age 51 years, range 22-69; median weight
55.0 kg, range 31.3-90.3) who underwent allo-HSCT at Kyushu University Hospital
between April 2016 and March 2020 with tacrolimus graft-versus-host disease
prophylaxis (Suetsugu 2021 Section 2.1 and Table 1). Voriconazole was given for
treatment (n = 33) or prophylaxis (n = 14) of invasive fungal infection, orally
(38), intravenously (1) or both (8). Nineteen patients received letermovir
(71 of the 216 samples; Fig. 1), 18 methylprednisolone, 19 prednisolone and 2
dexamethasone. Patients starting voriconazole more than 100 days after HSCT,
under 20 years old, or with liver dysfunction were excluded, as were
concentrations within 5 days of starting voriconazole.

The same information is available programmatically:

```{r population}
str(rxode2::rxode(readModelDb("Suetsugu_2021_voriconazole"))$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Suetsugu_2021_voriconazole.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Steady-state trough equation | $K_m F D / (V_{max} - F D)$ | Eq. 1, Section 2.5 |
| Categorical covariate form | $P = P_{pop} \times \theta^X$ | Eq. 3, Section 2.6 |
| Final $V_{max}$ equation | $670 \times 1.72^{LMV} \times 1.30^{mPSL}$ | Eq. 4, Section 3.3 |
| `lkm` | log(1.97) mg/L | Table 2, Km 1.97 ug/mL |
| `lvmax` | log(670) mg/day | Table 2, Vmax 670 mg/day |
| `lfdepot` | fixed(log(1)) | Section 2.5, 'the F-value was set to 1' |
| `e_conmed_letermovir_vmax` | 1.72 | Table 2, 'Effect of LMV on Vmax' |
| `e_conmed_methylprednisolone_vmax` | 1.30 | Table 2, 'Effect of mPSL on Vmax' |
| `etalvmax` | 0.05421 = log(1 + 0.236^2) | Table 2, IIV Vmax 23.6 CV%; exponential model (Section 2.5) |
| `addSd` | 0.77 mg/L | Table 2, 'Additive error (ug/mL)' 0.77 |

# Typical-value check against the closed form

With the random effects zeroed, the model must return exactly Eq. 1 with the
Table 2 typical values. The event table puts one observation row per dose
level and co-medication state on a single subject; `time` is only a row
index, since the model has no time dimension.

```{r typical}
mod <- readModelDb("Suetsugu_2021_voriconazole")

grid <- tidyr::expand_grid(
  group = c("No DDI", "with LMV", "with mPSL"),
  dose = c(300, 400, 500, 600)
) |>
  mutate(
    CONMED_LETERMOVIR = as.integer(group == "with LMV"),
    CONMED_METHYLPREDNISOLONE = as.integer(group == "with mPSL")
  )

ev_typ <- grid |>
  transmute(
    id = 1L,
    time = seq_len(n()),
    evid = 0L,
    amt = 0,
    DOSE_VORI_MGD = dose,
    CONMED_LETERMOVIR,
    CONMED_METHYLPREDNISOLONE
  )

typ <- as.data.frame(rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = ev_typ,
  returnType = "data.frame"
))
stopifnot(nrow(typ) == nrow(grid))

vmax_typ <- 670 * 1.72^grid$CONMED_LETERMOVIR * 1.30^grid$CONMED_METHYLPREDNISOLONE
grid$closed_form <- 1.97 * grid$dose / (vmax_typ - grid$dose)
grid$model <- typ$Cc

# Same parameters on both sides, so the only difference is floating point.
stopifnot(all(abs(grid$model - grid$closed_form) < 1e-8))

grid |>
  select(group, dose, model) |>
  tidyr::pivot_wider(names_from = group, values_from = model) |>
  dplyr::rename("Daily dose (mg)" = dose) |>
  knitr::kable(digits = 2, caption = "Typical steady-state trough (mg/L).")
```

The typical values already show the paper's conclusions: without an
interacting co-medication 300-400 mg/day lands in the 1-5 mg/L target, while
letermovir (Vmax x 1.72) pushes the typical trough at 300 mg/day below 1 mg/L.
At 600 mg/day without letermovir or methylprednisolone the typical trough is
`r round(grid$model[grid$group == "No DDI" & grid$dose == 600], 1)` mg/L because
the dose is about 10% below Vmax; the curve is steep there, and for patients
whose individual Vmax is below 600 mg/day no steady state exists at all.

```{r typical-curve, fig.width = 7, fig.height = 4.5, fig.cap = "Typical steady-state trough versus daily dose (Eq. 1). The grey band is the 1-5 mg/L target. Each curve rises without bound as the dose approaches its Vmax (670, 871 and 1152 mg/day)."}
curve_grid <- tidyr::expand_grid(
  group = c("No DDI", "with LMV", "with mPSL"),
  dose = seq(100, 640, by = 10)
) |>
  mutate(
    CONMED_LETERMOVIR = as.integer(group == "with LMV"),
    CONMED_METHYLPREDNISOLONE = as.integer(group == "with mPSL")
  )
ev_curve <- curve_grid |>
  transmute(
    id = 1L,
    time = seq_len(n()),
    evid = 0L,
    amt = 0,
    DOSE_VORI_MGD = dose,
    CONMED_LETERMOVIR,
    CONMED_METHYLPREDNISOLONE
  )
curve_sim <- as.data.frame(rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = ev_curve,
  returnType = "data.frame"
))
stopifnot(nrow(curve_sim) == nrow(curve_grid))
curve_grid$Cc <- curve_sim$Cc

ggplot(curve_grid, aes(dose, Cc, colour = group)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1, ymax = 5, alpha = 0.15) +
  geom_line(linewidth = 0.8) +
  coord_cartesian(ylim = c(0, 15)) +
  labs(
    x = "Daily voriconazole dose (mg/day)",
    y = "Typical steady-state trough (mg/L)",
    colour = NULL
  ) +
  theme_bw()
```

# Virtual cohort and replication of Figure 4

Figure 4 of the paper shows Monte Carlo steady-state troughs (1000 per
scenario) for 300-600 mg/day in three groups: no interacting co-medication,
with letermovir, and with methylprednisolone. Each simulated patient needs only
an individual Vmax (drawn from the IIV) and one trough with additive residual
error. Here each scenario is simulated with 200 virtual patients.

Two properties of the equation shape the figure. When a patient's Vmax is at or
below the dose, Eq. 1 is negative or infinite: no steady state exists. With a
0.77 mg/L additive error, low troughs are also often simulated below zero. The
paper does not say how it handled either case. Its box plots have no
non-positive values and omit the no-DDI 500 and 600 mg/day scenarios. The
no-DDI group has no steady state for about 10% of patients at 500 mg/day and
about 32% at 600 mg/day. We reproduce the figure by keeping simulated troughs
that have a steady state and are positive, and by showing only the scenarios
the paper shows. Without that filter the medians do not match (see the
Assumptions section).

```{r cohort}
n_per_arm <- 200L

scen <- tidyr::expand_grid(
  group = c("No DDI", "with LMV", "with mPSL"),
  dose = c(300, 400, 500, 600)
) |>
  filter(!(group == "No DDI" & dose > 400)) |>
  mutate(scenario = seq_len(n()))

ev_mc <- scen |>
  tidyr::uncount(n_per_arm) |>
  mutate(
    id = seq_len(n()),
    time = 0,
    evid = 0L,
    amt = 0,
    DOSE_VORI_MGD = dose,
    CONMED_LETERMOVIR = as.integer(group == "with LMV"),
    CONMED_METHYLPREDNISOLONE = as.integer(group == "with mPSL")
  )

rxode2::rxSetSeed(20211015)
mc <- as.data.frame(rxode2::rxSolve(mod, events = ev_mc, returnType = "data.frame"))
stopifnot(nrow(mc) == nrow(ev_mc))

mc <- ev_mc |>
  select(id, group, dose) |>
  left_join(mc |> select(id, vmax, sim), by = "id") |>
  mutate(retained = vmax > dose & sim > 0)

mc |>
  group_by(group, dose) |>
  summarise(
    `No steady state (%)` = 100 * mean(vmax <= dose),
    `Retained (%)` = 100 * mean(retained),
    .groups = "drop"
  ) |>
  dplyr::rename("Group" = group, "Daily dose (mg)" = dose) |>
  knitr::kable(digits = 1, caption = "Share of simulated troughs dropped before plotting.")
```

The published box-plot statistics come from the vector graphics of Figure 4
in the article PDF. They are the exact end points of each box (25th and 75th
percentiles), its median line, and its whiskers (10th and 90th percentiles),
converted to mg/L using the 1 and 5 mg/L target lines as the axis
calibration.

```{r fig4-published}
fig4 <- read.table(header = TRUE, text = "
group      dose  q10    q25    q50    q75    q90
No_DDI     300   0.664  1.068  1.774  2.456  3.364
No_DDI     400   1.373  2.026  3.112  4.751  7.607
with_LMV   300   0.228  0.478  0.910  1.405  1.809
with_LMV   400   0.322  0.695  1.214  1.809  2.363
with_LMV   500   0.540  1.006  1.650  2.424  3.301
with_LMV   600   0.851  1.560  2.334  3.513  4.875
with_mPSL  300   0.353  0.664  1.183  1.746  2.276
with_mPSL  400   0.664  1.155  1.774  2.580  3.575
with_mPSL  500   1.125  1.871  2.764  3.949  6.369
with_mPSL  600   1.622  2.704  4.163  7.296  14.307
") |>
  mutate(group = sub("_", " ", group))
```

```{r fig4, fig.width = 7, fig.height = 4.5, fig.cap = "Replicates Figure 4 of Suetsugu 2021: simulated steady-state voriconazole troughs. Boxes are the 25th-75th percentiles with the median; whiskers are the 10th-90th percentiles. Red diamonds are the published medians. The grey band is the 1-5 mg/L target."}
sim_q <- mc |>
  filter(retained) |>
  group_by(group, dose) |>
  summarise(
    q10 = quantile(sim, 0.10),
    q25 = quantile(sim, 0.25),
    q50 = quantile(sim, 0.50),
    q75 = quantile(sim, 0.75),
    q90 = quantile(sim, 0.90),
    .groups = "drop"
  )

ggplot(sim_q, aes(x = factor(dose))) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1, ymax = 5, alpha = 0.15) +
  geom_boxplot(
    aes(ymin = q10, lower = q25, middle = q50, upper = q75, ymax = q90),
    stat = "identity",
    fill = "grey80",
    width = 0.6
  ) +
  geom_point(data = fig4, aes(y = q50), colour = "red", shape = 18, size = 3) +
  facet_grid(~group, scales = "free_x", space = "free_x") +
  coord_cartesian(ylim = c(0, 15)) +
  labs(x = "Daily dose (mg)", y = "Voriconazole concentration (mg/L)") +
  theme_bw()
```

```{r fig4-compare}
cmp <- sim_q |>
  tidyr::pivot_longer(q10:q90, names_to = "stat", values_to = "simulated") |>
  inner_join(
    fig4 |> tidyr::pivot_longer(q10:q90, names_to = "stat", values_to = "published"),
    by = c("group", "dose", "stat")
  ) |>
  mutate(pct_diff = 100 * (simulated - published) / published)
stopifnot(nrow(cmp) == 50L)

med_cmp <- cmp |> filter(stat == "q50")

med_cmp |>
  select(group, dose, published, simulated, pct_diff) |>
  dplyr::rename(
    "Group" = group,
    "Daily dose (mg)" = dose,
    "Published median (mg/L)" = published,
    "Simulated median (mg/L)" = simulated,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Median steady-state trough: simulation against Figure 4.")

# Bounds calibrated on 2000 independent 200-per-scenario cohorts drawn from
# the same model: |median of the ten median differences| reached at most
# 10.1% (99.9th percentile 7.1%), and the 80th percentile of the 50 absolute
# differences at most 19.2%. A Vmax of 607 instead of 670, a letermovir or
# methylprednisolone factor of 1.27 / 1.03, a residual SD of 0.077, or an IIV
# variance of 0.236 each failed the pair of bounds in every cohort tried.
stopifnot(
  # Structural: centre of the ten median differences.
  abs(median(med_cmp$pct_diff)) < 12,
  # Envelope over all 50 box-plot statistics, robust to the Monte Carlo noise
  # of 200 patients per scenario in the tails.
  quantile(abs(cmp$pct_diff), 0.8) < 25
)
```

With 200 000 patients per scenario the same filter reproduces all ten
published medians within 3.2% and 49 of the 50 box-plot statistics within 10%;
the exception is the with-methylprednisolone 600 mg/day 10th percentile (+13%)
(checked while preparing this vignette).
The remaining differences above come from Monte Carlo noise with 200 patients
per scenario, which is largest at the 10th and 90th percentiles.

# Target attainment

The paper's dosing conclusion (Section 3.5) is that 300 or 400 mg/day reaches
the 1-5 mg/L target without an interacting co-medication, 500 or 600 mg/day
with letermovir, and 400 or 500 mg/day with methylprednisolone. The same
retained simulations give:

```{r pta}
pta <- mc |>
  filter(retained) |>
  group_by(group, dose) |>
  summarise(`In 1-5 mg/L (%)` = 100 * mean(sim >= 1 & sim <= 5), .groups = "drop") |>
  tidyr::pivot_wider(names_from = group, values_from = `In 1-5 mg/L (%)`) |>
  dplyr::rename("Daily dose (mg)" = dose)
options(knitr.kable.NA = "not shown")
knitr::kable(pta, digits = 0, caption = "Percentage of retained simulated troughs within 1-5 mg/L (no-DDI 500 and 600 mg/day are not shown in the paper).")
```

# Why no PKNCA section

The model has no time dimension. It returns a single steady-state trough for
each dose level, and there is no concentration-time profile to put through a
non-compartmental analysis. The paper reports no NCA parameters. The model is
validated against its closed form and against the paper's own simulation
(Figure 4).

# Assumptions and deviations

* **Algebraic encoding.** The paper's model is Eq. 1 itself; it estimates no
  volume of distribution or absorption rate, and none is invented here. The
  model therefore cannot produce concentration-time profiles, only
  steady-state troughs for a given daily dose.
* **Daily dose as a covariate.** Eq. 1 takes the daily maintenance dose as an
  input, so the packaged model reads it from the `DOSE_VORI_MGD` column
  (mg/day, summed over the day) on each observation row instead of from dosing
  events. Oral and intravenous doses are treated alike because the paper fixed
  F to 1 (Section 2.5). A sensitivity analysis with F = 0.459 (ESM-4) is not
  part of the final model and is not included.
* **No steady state when dose >= Vmax.** For patients whose individual Vmax is
  at or below the daily dose, Eq. 1 returns a negative or infinite value. The
  model returns that value unchanged, as the published equation does; users
  should treat such rows as "no steady state" (unbounded accumulation), not
  as concentrations.
* **Handling of non-positive simulated troughs in Figure 4.** The paper does
  not describe it. Keeping only troughs with a steady state and a positive
  value (after additive residual error) reproduces the published box plots.
  Keeping all values, or plotting the individual predictions without residual
  error, does not: for example, the with-letermovir 300 mg/day median is then
  0.74 or 0.69 mg/L instead of the published 0.91 mg/L. The simulated values
  in this article use the filter that reproduces the paper.
* **IIV scale.** Table 2 reports the IIV of Vmax as 23.6 CV% for an exponential
  model. It is converted with omega^2 = log(1 + CV^2) = 0.05421; the
  alternative omega^2 = CV^2 = 0.0557 differs by 3% and cannot be told apart
  from Figure 4.
* **Additive residual error.** Table 2 gives the additive error in ug/mL
  (0.77), i.e. a standard deviation in concentration units, used as `addSd`.
* **Screened covariates.** Sex, age, body weight, albumin, CRP, PPIs,
  prednisolone and dexamethasone were screened and not retained. They are
  listed in `covariatesDataExcluded` for provenance and do not enter the model.
* **Errata.** No correction notice was found for this article in Europe PMC
  (checked 2026-09-29).
