---
title: "GS-441524 after remdesivir (Sukeishi 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GS-441524 after remdesivir (Sukeishi 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Sukeishi_2022_remdesivir"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1002/psp4.12736> (open access, PMC8646568)
* Supporting Information: Tables S1-S2 (stepwise covariate search) and the
  NONMEM control stream of the final model.

```{r model-print}
ui
```

# Population

Sukeishi 2022 is a retrospective, single-centre study at Kyoto University
Hospital (December 2020 to May 2021). Thirty-seven Japanese inpatients with
COVID-19 received intravenous remdesivir: 200 mg on day 1 then 100 mg once
daily, each infused over 1 h, when the absolute (BSA-readjusted) eGFR was at
least 30 mL/min, and 200 mg on day 1 then 100 mg every 2 days from day 3 in the
two patients below 30 mL/min. Twenty-seven were male; median age was 72 years
(range 45-97), body weight 66.8 kg (36.7-96.3), height 167.1 cm (144-182) and
BSA 1.8 m^2 (1.24-2.21). Absolute eGFR ranged 16.4-147.7 mL/min (median 74.7)
across the measurement points. Four patients were on extracorporeal membrane
oxygenation (21 measurements) and 82 measurements were taken on a ventilator
(Sukeishi 2022 Table 1).

The 190 serum GS-441524 concentrations were all troughs, taken from surplus
serum drawn for other blood tests, and were measured by LC-MS/MS over a
5-500 ng/mL calibration range. The median observed concentration was
116.6 ng/mL (range 34.6-366.4).

```{r population}
str(ui$population)
```

# Model structure

The model is a one-compartment model for GS-441524 alone (NONMEM ADVAN1 TRANS2,
FOCE-I). Three points matter for using it.

**Doses are remdesivir, the state is GS-441524.** Remdesivir has a half-life
under 1 h and is not modelled. Each remdesivir dose enters the GS-441524
compartment directly, converted mole-for-mole by the molecular-weight ratio.
The main text does not say so, but the control stream in the Supporting
Information does: with `AMT` in mg of remdesivir it sets
`S1 = (V/1000)/(291.3/602.6)`. The model file puts the same factor on the dose
through `f(central) <- 291.3 / 602.6` instead of on the scaling factor. The
concentrations are identical, and `central` then holds a real GS-441524 amount.
Conversion is assumed complete, so `cl` and `vc` are apparent values. They are
the paper's CL and Vd as printed.

**Give infusions with `dur`, not `rate`.** When a bioavailability factor acts
on an infusion given with a fixed `rate`, rxode2 keeps the rate and shortens
the infusion to `f * amt / rate`, here about 29 minutes. The source's 1-h
infusion is reproduced by setting `dur = 1`, which keeps the duration and
scales the rate. The first check below verifies this.

**Renal function enters as absolute eGFR.** The laboratory reported
BSA-indexed eGFR (Japanese MDRD equation, mL/min/1.73 m^2). The authors
de-indexed it as `eGFR / 1.73 * BSA`, and that absolute value (the source
column `EGFR2`) gave the best fit of the renal markers tested. The model takes
the canonical indexed `CRCL` plus `BSA` and does the same de-indexing
internally, so `CRCL` must be supplied in mL/min/1.73 m^2. Age enters only as
a 75-year threshold on volume, derived internally from `AGE`.

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Sukeishi_2022_remdesivir.R`
carries an in-file comment naming its source location.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at eGFR 74.7 mL/min) | 11.8 L/h | Table 2 (RSE 4.9%); Results final-model equation; Discussion |
| `lvc` (Vd, age < 75 y) | 382 L | Table 2 (RSE 9.9%); Results final-model equation; Discussion |
| `e_crcl_cl` | 1.09 | Table 2 `theta eGFRnon-indexed,CL` (RSE 10.8%) |
| `e_age_ge_75_vc` | -0.429 | Table 2 `theta Age>=75,V` (RSE 21.1%) and footnote defining AGE as a >= 75 y indicator |
| `etalcl` | log(0.260^2 + 1) = 0.0654 | Table 2 IIV for CL, 26.0 CV% (shrinkage 4.22%) |
| `etalvc` | log(0.345^2 + 1) = 0.1125 | Table 2 IIV for Vd, 34.5 CV% (shrinkage 24.1%) |
| `propSd` | 0.152 | Table 2 proportional error, 15.2 CV% |
| `cl <- ... * (egfr_abs / 74.7)^e_crcl_cl` | n/a | Table 2 CL equation; control stream `CL=THETA(1)*(EGFR2/74.7)**THETA(3)*EXP(ETA(1))` |
| `vc <- ... * (1 + e_age_ge_75_vc * age_ge_75)` | n/a | Table 2 Vd equation; control stream `V=THETA(2)*(1+THETA(4)*AGE)*EXP(ETA(2))` |
| `egfr_abs <- CRCL / 1.73 * BSA` | n/a | Methods "Patients and data collection" |
| `f(central) <- 291.3 / 602.6`; `Cc <- central / vc * 1000` | n/a | Control stream `S1=(V/1000)/(291.3/602.6)` |
| `Cc ~ prop(propSd)` | n/a | Results; control stream `Y=F*(1+EPS(1))` |
| 1-h infusion | n/a | Methods "Remdesivir administration" |

The `$THETA`, `$OMEGA` and `$SIGMA` records in the control stream are initial
estimates. Every value above is the final estimate from Table 2.

# Deterministic checks

These solves use typical values only (`zeroRe()`), with tight integrator
tolerances. No random draw is involved, so the bounds are tight.

```{r typical-setup}
ui_typ <- rxode2::zeroRe(ui)
theta <- setNames(ui$theta, names(ui$theta))

# A typical patient at the eGFR reference point: absolute eGFR 74.7 mL/min.
# Any BSA works because only CRCL * BSA / 1.73 enters the model.
typ_cov <- function(egfr_abs, age = 60, bsa = 1.73) {
  data.frame(CRCL = egfr_abs * 1.73 / bsa, BSA = bsa, AGE = age)
}

# Regimen 1: 200 mg day 1, then 100 mg daily. Regimen 2: 100 mg every 2 days
# from day 2. Regimen 3: 100 mg every 4 days from day 2 (Table 3 footnote).
dose_times <- function(regimen, horizon = 24 * 30) {
  tau <- c(24, 48, 96)[regimen]
  c(0, seq(24, horizon, by = tau))
}
make_doses <- function(regimen) {
  tt <- dose_times(regimen)
  tibble(time = tt, amt = c(200, rep(100, length(tt) - 1)),
         dur = 1, evid = 1L, cmt = "central")
}
```

## The 1-h infusion is preserved

```{r infusion-check}
ev_inf <- bind_rows(
  tibble(time = 0, amt = 200, dur = 1, evid = 1L, cmt = "central"),
  tibble(time = seq(0, 3, by = 1 / 60), amt = NA_real_, dur = NA_real_,
         evid = 0L, cmt = "central")
) |>
  mutate(id = 1L) |>
  bind_cols(typ_cov(74.7))
inf <- rxode2::rxSolve(ui_typ, ev_inf, rtol = 1e-10, atol = 1e-12,
                       returnType = "data.frame")
t_peak <- inf$time[which.max(inf$Cc)]

# Closed-form end-of-infusion concentration for a 1-h infusion of
# 200 mg * 291.3 / 602.6 GS-441524, in ng/mL.
cl0 <- exp(theta[["lcl"]]); v0 <- exp(theta[["lvc"]]); k0 <- cl0 / v0
d_gs <- 200 * 291.3 / 602.6
c_eoi <- d_gs / cl0 * (1 - exp(-k0 * 1)) * 1000
c(t_peak = t_peak, Cc_at_1h = inf$Cc[inf$time == 1], closed_form = c_eoi)

stopifnot(
  abs(t_peak - 1) < 1e-8,
  abs(inf$Cc[inf$time == 1] / c_eoi - 1) < 1e-6
)
```

The concentration peaks at the end of the 1-h infusion. It matches the
closed-form value, so the whole dose went in over 1 h at the converted amount.

## Typical parameters

```{r typical-parameters}
typ_tab <- tibble::tribble(
  ~Quantity, ~Model, ~Published,
  "CL at eGFR 74.7 mL/min (L/h)", cl0, 11.8,
  "Vd, age < 75 y (L)", v0, 382,
  "Vd, age >= 75 y (L)", v0 * (1 + theta[["e_age_ge_75_vc"]]), NA_real_,
  "CL at eGFR 30 mL/min (L/h)", cl0 * (30 / 74.7)^theta[["e_crcl_cl"]], NA_real_,
  "Half-life at eGFR 74.7, age < 75 (h)", log(2) * v0 / cl0, NA_real_
)
knitr::kable(typ_tab, digits = 3,
             caption = "Typical-value quantities against the Sukeishi 2022 Discussion.")
stopifnot(abs(cl0 - 11.8) < 1e-8, abs(v0 - 382) < 1e-8)
```

## Typical troughs against Table 3

Table 3 gives the median simulated trough (5000 profiles, patients under 75
years) for four absolute eGFR values and three regimens. The paper does not
say on which day the trough was read. Regimens 1 and 2 match the trough just
before the day-10 dose (t = 216 h), the 10-day horizon the paper uses for
Figure 4. Regimen 3 matches the trough before its day-14 dose (t = 312 h). Its
day-10 trough runs 5-19% above the published medians, and the gap varies with
eGFR, so it is not a scale error. At eGFR 15 mL/min the half-life is about
5 days, so none of these troughs is at steady state.

```{r typical-troughs}
table3 <- tibble::tribble(
  ~regimen, ~egfr, ~q50, ~q05, ~q95,
  1L, 15, 670.3, 423.1, 1030.5,
  1L, 30, 370.5, 211.9, 591.9,
  1L, 60, 158.2, 74.2, 284.6,
  1L, 90, 85.3, 33.3, 169.2,
  2L, 15, 351.9, 219.2, 542.7,
  2L, 30, 168.3, 83.8, 286.8,
  2L, 60, 56.0, 18.6, 119.7,
  2L, 90, 24.0, 4.8, 63.0,
  3L, 15, 184.4, 96.9, 296.2,
  3L, 30, 66.0, 21.8, 136.9,
  3L, 60, 13.3, 1.6, 44.4,
  3L, 90, 3.4, 0.1, 18.2
) |>
  mutate(t_trough = c(216, 216, 312)[regimen],
         arm = sprintf("Regimen %d, eGFR %g", regimen, egfr))

typ_trough <- function(regimen, egfr, age, t_obs) {
  ev <- bind_rows(
    make_doses(regimen),
    tibble(time = t_obs, amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central")
  ) |>
    mutate(id = 1L) |>
    bind_cols(typ_cov(egfr, age = age))
  s <- rxode2::rxSolve(ui_typ, ev, rtol = 1e-10, atol = 1e-12,
                       returnType = "data.frame")
  s$Cc[s$time == t_obs]
}

typ_cmp <- table3 |>
  rowwise() |>
  mutate(
    model = typ_trough(regimen, egfr, age = 60, t_obs = t_trough),
    model_day10 = typ_trough(regimen, egfr, age = 60, t_obs = 216),
    no_mw = model * 602.6 / 291.3
  ) |>
  ungroup() |>
  mutate(pct_diff = 100 * (model - q50) / q50,
         pct_diff_day10 = 100 * (model_day10 - q50) / q50,
         pct_diff_no_mw = 100 * (no_mw - q50) / q50)

typ_cmp |>
  select(arm, t_trough, q50, model, pct_diff, pct_diff_day10, pct_diff_no_mw) |>
  dplyr::rename(
    "Arm" = arm, "Trough time (h)" = t_trough,
    "Table 3 median (ng/mL)" = q50, "Typical model (ng/mL)" = model,
    "% diff" = pct_diff, "% diff, day-10 trough" = pct_diff_day10,
    "% diff without MW ratio" = pct_diff_no_mw
  ) |>
  knitr::kable(digits = 1, caption = paste(
    "Typical-value troughs against Sukeishi 2022 Table 3 medians (age < 75 y)."
  ))

# Deterministic. Measured -0.4% to +6.1% (the Table 3 medians come from a
# stochastic cohort and sit 1-3% below the typical value). Leaving out the
# molecular-weight conversion puts every arm at about +107%, and a 10%
# transcription error in CL or its exponent moves the eGFR 15 and 90 arms by
# more than 10%.
stopifnot(max(abs(typ_cmp$pct_diff)) < 8)
```

## Age and renal-function effects (Figure 4 claims)

The Results state that the regimen-1 troughs are about 4-fold and 2-fold higher
at eGFR 15 and 30 mL/min than at 60 mL/min, and are only slightly affected by
age. Figure 4 shows the age effect more precisely. The maintainers read the
box-plot medians off the published figure by eye (to about +/- 20 ng/mL). For
patients under 75 years they are about 670, 370, 160 and 90 ng/mL at eGFR 15,
30, 60 and 90 mL/min. For patients 75 years or older they are about 770, 350,
130 and 60 ng/mL. So age raises the trough at eGFR 15 and lowers it at 60 and
90. With a smaller volume the drug approaches steady state faster (which
dominates when clearance is low) and fluctuates more between doses (which
dominates when clearance is high).

```{r figure-4-claims}
fig4_typ <- tidyr::expand_grid(egfr = c(15, 30, 60, 90), age = c(60, 80)) |>
  rowwise() |>
  mutate(trough = typ_trough(1L, egfr, age, t_obs = 216)) |>
  ungroup()

# Figure 4 medians, digitised by eye by the maintainers.
fig4_pub <- tibble::tribble(
  ~egfr, ~young_pub, ~old_pub,
  15, 670, 770,
  30, 370, 350,
  60, 160, 130,
  90, 90, 60
)
fig4_cmp <- fig4_typ |>
  tidyr::pivot_wider(names_from = age, values_from = trough,
                     names_prefix = "age_") |>
  inner_join(fig4_pub, by = "egfr") |>
  mutate(ratio_model = age_80 / age_60, ratio_pub = old_pub / young_pub)
fig4_cmp |>
  dplyr::rename("eGFR (mL/min)" = egfr,
                "Model, < 75 y" = age_60, "Model, >= 75 y" = age_80,
                "Figure 4, < 75 y" = young_pub, "Figure 4, >= 75 y" = old_pub,
                "Age ratio, model" = ratio_model,
                "Age ratio, Figure 4" = ratio_pub) |>
  knitr::kable(digits = 2, caption = paste(
    "Typical regimen-1 day-10 troughs (ng/mL) by eGFR and age against the",
    "Figure 4 medians."
  ))

young <- fig4_cmp
fold_15 <- young$age_60[young$egfr == 15] / young$age_60[young$egfr == 60]
fold_30 <- young$age_60[young$egfr == 30] / young$age_60[young$egfr == 60]
c(fold_15 = fold_15, fold_30 = fold_30)

# Deterministic. The paper's "four- and two-fold" are rounded; Table 3 gives
# 4.24 and 2.34. The age ratios measured 1.18 / 0.95 / 0.79 / 0.67 against
# about 1.15 / 0.95 / 0.81 / 0.67 read off Figure 4. Reversing the sign of the
# age effect (a larger volume at 75 or older) moves every ratio across 1 and
# fails this gate.
stopifnot(
  fold_15 > 3.5, fold_15 < 5,
  fold_30 > 1.8, fold_30 < 2.9,
  all(abs(fig4_cmp$ratio_model - fig4_cmp$ratio_pub) < 0.12)
)
```

The fold changes are `r round(fold_15, 2)` and `r round(fold_30, 2)`. The model
reproduces the Figure 4 age pattern, including the crossover at eGFR 15. The
age effect is at most about a third, far smaller than the renal effect, which
is presumably why the paper calls it slight.

# Replicating Table 3 with between-subject variability

Each of the 12 Table 3 arms is simulated with 200 patients under 75 years
(Table 3 used 5000). Observations run hourly over the dosing interval that ends
at each arm's trough time. The same data then feed the PKNCA analysis below.

```{r table3-sim}
# set.seed() seeds R's RNG only. rxode2 partitions its own random streams by
# solver thread, so the etas below differ between machines; the assertions are
# written on medians and robust quantiles to hold for any cohort.
set.seed(20220101)
rxode2::rxSetSeed(20220101)
N_ARM <- 200L   # cap: never more than 200 participants per arm

interval_start <- c(192, 168, 216)

t3_events <- table3 |>
  mutate(arm_id = row_number()) |>
  rowwise() |>
  do({
    a <- .
    obs_t <- seq(interval_start[a$regimen], a$t_trough, by = 1)
    one <- bind_rows(
      make_doses(a$regimen) |> filter(time < a$t_trough),
      tibble(time = obs_t, amt = NA_real_, dur = NA_real_, evid = 0L,
             cmt = "central")
    )
    tidyr::crossing(id = (a$arm_id - 1L) * N_ARM + seq_len(N_ARM), one) |>
      mutate(arm = a$arm, regimen = a$regimen, egfr = a$egfr,
             CRCL = a$egfr, BSA = 1.73, AGE = 60)
  }) |>
  ungroup() |>
  arrange(id, time, desc(evid))

t3_sim <- rxode2::rxSolve(ui, t3_events, keep = c("arm", "regimen", "egfr"),
                          returnType = "data.frame")
stopifnot(dplyr::n_distinct(t3_sim$id) == nrow(table3) * N_ARM,
          !anyNA(t3_sim$Cc))
```

# PKNCA validation

For each arm, NCA runs over the interval that ends at the Table 3 trough time.
`clast.obs`, the concentration at the end of that interval, is the Table 3
trough. The formula groups by arm, the treatment variable. One `pk.nca()` call
is made per regimen to keep each call small.

```{r pknca}
conc_all <- t3_sim |>
  dplyr::filter(!is.na(Cc)) |>
  select(id, time, Cc, arm, regimen, egfr)
dose_all <- t3_events |>
  filter(evid == 1L) |>
  select(id, time, amt, arm, regimen)

nca_one <- function(reg) {
  conc_obj <- PKNCA::PKNCAconc(filter(conc_all, regimen == reg),
                               Cc ~ time | arm + id)
  dose_obj <- PKNCA::PKNCAdose(filter(dose_all, regimen == reg),
                               amt ~ time | arm + id)
  iv <- data.frame(
    start = interval_start[reg], end = c(216, 216, 312)[reg],
    cmax = TRUE, tmax = TRUE, auclast = TRUE, clast.obs = TRUE
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = iv))
  as.data.frame(res)
}
nca_res <- bind_rows(lapply(1:3, nca_one))

nca_res |>
  filter(PPTESTCD %in% c("cmax", "auclast", "clast.obs")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename("Arm" = arm, "Cmax (ng/mL)" = cmax,
                "AUClast over interval (ng*h/mL)" = auclast,
                "Trough, Clast (ng/mL)" = clast.obs) |>
  knitr::kable(digits = 1, caption = "Median simulated NCA by Table 3 arm.")
```

## Comparison against the published troughs

```{r nca-comparison}
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res |> filter(PPTESTCD == "clast.obs") |>
    select(arm, PPTESTCD, PPORRES),
  reference = table3 |> transmute(arm, clast.obs = q50),
  by = "arm",
  units = c(clast.obs = "ng/mL"),
  tolerance_pct = 20
)
cmp |>
  dplyr::rename("Arm" = arm) |>
  knitr::kable(caption = paste(
    "Simulated median trough against Sukeishi 2022 Table 3 medians.",
    "* differs from the published median by more than 20%."
  ))

trough_q <- nca_res |>
  filter(PPTESTCD == "clast.obs") |>
  group_by(arm) |>
  summarise(sim_q05 = quantile(PPORRES, 0.05), sim_q50 = median(PPORRES),
            sim_q95 = quantile(PPORRES, 0.95), .groups = "drop") |>
  inner_join(table3, by = "arm") |>
  mutate(pct_q50 = 100 * (sim_q50 - q50) / q50)

# The same troughs with the proportional residual error added (`sim`).
trough_res <- t3_sim |>
  inner_join(select(table3, arm, t_trough), by = "arm") |>
  filter(time == t_trough) |>
  group_by(arm) |>
  summarise(res_q05 = quantile(sim, 0.05), res_q50 = median(sim),
            res_q95 = quantile(sim, 0.95), .groups = "drop")

trough_q |>
  inner_join(trough_res, by = "arm") |>
  transmute(arm,
            published = sprintf("%.1f (%.1f-%.1f)", q50, q05, q95),
            simulated = sprintf("%.1f (%.1f-%.1f)", sim_q50, sim_q05, sim_q95),
            simulated_res = sprintf("%.1f (%.1f-%.1f)", res_q50, res_q05, res_q95)) |>
  dplyr::rename("Arm" = arm, "Table 3 median (90% PI)" = published,
                "Simulated, no residual error" = simulated,
                "Simulated, with residual error" = simulated_res) |>
  knitr::kable(caption = "Median and 90% prediction interval of the trough (ng/mL).")

# Centre and robust envelope, not extremes: with 200 per arm the median
# trough has a Monte Carlo SE of about 3%. A lost MW ratio (+107%) or a
# mis-transcribed CL fails the centre gate at once.
stopifnot(
  abs(median(trough_q$pct_q50)) < 8,
  quantile(abs(trough_q$pct_q50), 0.9) < 15
)
```

The medians agree closely. Without residual error, the simulated 90% intervals
are narrower than the published ones at low eGFR. Adding the 15.2%
proportional residual error brings them closer, which suggests the Table 3
simulations included it. The paper does not say. The IIV scale cannot be
settled from these intervals: reading Table 2's CV% as
`omega^2 = log(CV^2 + 1)` (used here) or as `omega^2 = CV^2` changes the
variances by only about 3%.

## Figure 4 of Sukeishi 2022

Replicates Figure 4 of Sukeishi 2022: regimen-1 troughs on day 10 by absolute
eGFR and age group (the paper used 5000 profiles per group; 200 here).

```{r figure-4, fig.height = 4}
fig4_events <- tidyr::expand_grid(egfr = c(15, 30, 60, 90),
                                  age_group = c("< 75 y", ">= 75 y")) |>
  mutate(grp = row_number()) |>
  rowwise() |>
  do({
    g <- .
    one <- bind_rows(
      make_doses(1L) |> filter(time < 216),
      tibble(time = 216, amt = NA_real_, dur = NA_real_, evid = 0L,
             cmt = "central")
    )
    tidyr::crossing(id = (g$grp - 1L) * N_ARM + seq_len(N_ARM), one) |>
      mutate(egfr = g$egfr, age_group = g$age_group, CRCL = g$egfr,
             BSA = 1.73, AGE = ifelse(g$age_group == "< 75 y", 60, 80))
  }) |>
  ungroup() |>
  arrange(id, time, desc(evid))

fig4_sim <- rxode2::rxSolve(ui, fig4_events, keep = c("egfr", "age_group"),
                            returnType = "data.frame")

ggplot(fig4_sim, aes(factor(egfr), Cc, fill = age_group)) +
  geom_boxplot(outlier.size = 0.5) +
  scale_fill_manual(values = c("< 75 y" = "white", ">= 75 y" = "grey60")) +
  labs(x = "Absolute eGFR (mL/min)", y = "Day-10 trough GS-441524 (ng/mL)",
       fill = "Age") +
  theme_bw()
```

# A cohort like the study (Figures 1a and 3)

This cohort is built to look like the study population and follows the
package-insert regimen for 10 days. Patients are split across the three renal
bands in the published proportions at the start of therapy: 20, 15 and 2 of 37
at absolute eGFR >= 60, 30-60 and < 30 mL/min. Within each band, eGFR is drawn
uniformly, bounded by the observed 16.4-147.7 mL/min range. The cohort median
is therefore lower than the Table 1 median of 74.7 mL/min. That median pools
every measurement point, and renal function improved in several patients
during therapy. Patients below 30 mL/min get the paper's reduced regimen
(100 mg every 2 days from day 3). Age and BSA are drawn around their published
medians. Covariates are held constant in time.

```{r cohort}
N_COH <- 200L
draw_trunc <- function(n, draw, lo, hi) {
  x <- draw(n)
  while (any(bad <- x < lo | x > hi)) x[bad] <- draw(sum(bad))
  x
}
bands <- tibble::tribble(
  ~lo, ~hi, ~n_published,
  60, 147.7, 20L,
  30, 60, 15L,
  16.4, 30, 2L
)
band <- sample(seq_len(nrow(bands)), N_COH, replace = TRUE,
               prob = bands$n_published / sum(bands$n_published))
cohort <- tibble(
  id = seq_len(N_COH),
  egfr_abs = runif(N_COH, bands$lo[band], bands$hi[band]),
  BSA = draw_trunc(N_COH, function(n) rnorm(n, 1.8, 0.18), 1.24, 2.21),
  AGE = draw_trunc(N_COH, function(n) rnorm(n, 72, 12), 45, 97)
) |>
  mutate(CRCL = egfr_abs * 1.73 / BSA,
         reduced = egfr_abs < 30)
cohort |>
  summarise(median_egfr = median(egfr_abs),
            pct_ge60 = 100 * mean(egfr_abs >= 60),
            pct_30_60 = 100 * mean(egfr_abs >= 30 & egfr_abs < 60),
            pct_lt30 = 100 * mean(egfr_abs < 30),
            pct_age_ge75 = 100 * mean(AGE >= 75)) |>
  dplyr::rename("Median eGFR (mL/min)" = median_egfr,
                "% eGFR >= 60" = pct_ge60, "% eGFR 30-60" = pct_30_60,
                "% eGFR < 30" = pct_lt30, "% age >= 75" = pct_age_ge75) |>
  knitr::kable(digits = 1, caption = "Simulated cohort covariates.")

coh_doses <- function(reduced) {
  tt <- if (reduced) c(0, seq(48, 216, by = 48)) else seq(0, 216, by = 24)
  tibble(time = tt, amt = c(200, rep(100, length(tt) - 1)), dur = 1,
         evid = 1L, cmt = "central")
}
coh_obs <- tibble(time = sort(unique(c(seq(0, 240, by = 2), seq(24, 240, by = 24)))),
                  amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central")
coh_events <- cohort |>
  rowwise() |>
  do(bind_rows(coh_doses(.$reduced), coh_obs) |>
       mutate(id = .$id, CRCL = .$CRCL, BSA = .$BSA, AGE = .$AGE,
              reduced = .$reduced)) |>
  ungroup() |>
  arrange(id, time, desc(evid))

coh_sim <- rxode2::rxSolve(ui, coh_events, keep = "reduced",
                           returnType = "data.frame")
```

Replicates the layout of Figure 3 of Sukeishi 2022: simulated concentrations
over the first 10 days (median and 90% interval, residual error included). The
paper's observations are troughs only, so the published pcVPC covers only the
lower edge of this band.

```{r figure-3}
coh_sim |>
  group_by(time) |>
  summarise(q05 = quantile(sim, 0.05), q50 = median(sim),
            q95 = quantile(sim, 0.95), .groups = "drop") |>
  ggplot(aes(time / 24, q50)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.25) +
  geom_line() +
  labs(x = "Time after first dose (days)", y = "GS-441524 (ng/mL)") +
  theme_bw()
```

Troughs after day 4 are compared with the paper's median observed
concentration (116.6 ng/mL; the Results give the same value for the
steady-state troughs after day 4).

```{r cohort-trough}
coh_trough <- coh_sim |>
  filter(time %in% seq(120, 240, by = 24)) |>
  filter(!reduced | time %in% seq(144, 240, by = 48))
coh_med <- median(coh_trough$sim)
c(simulated_median_trough = coh_med, published = 116.6)

# The cohort covariates are assumptions, so the band is wide (83-163 ng/mL).
# Measured 114 / 93 / 118 ng/mL across three seeds. A lost MW ratio (about
# 2.07-fold) still falls well outside it.
stopifnot(abs(log(coh_med / 116.6)) < log(1.4))
```

# Assumptions and deviations

* **Dose basis.** The main text never states that doses were converted to
  GS-441524 equivalents. The Supporting Information control stream does
  (`S1=(V/1000)/(291.3/602.6)`), and Table 3 is reproduced only with that
  conversion. Without it every trough is about 107% too high.
* **IIV scale.** Table 2 reports IIV as CV% for an exponential model. It is
  converted as `omega^2 = log(CV^2 + 1)`. The simulated 90% intervals cannot
  distinguish this from `omega^2 = CV^2`, and the two differ by about 3%.
* **Residual error.** The 15.2 CV% proportional error is used as
  `propSd = 0.152`.
* **Table 3 trough times.** Not stated in the paper. Regimens 1 and 2 match the
  day-10 pre-dose trough and regimen 3 the day-14 pre-dose trough, as shown
  above.
* **eGFR input.** The model takes the canonical BSA-indexed `CRCL`
  (mL/min/1.73 m^2) plus `BSA`, and de-indexes internally as the authors did.
  In the source this covariate varied over time; the cohorts here hold it
  constant.
* **Screened covariates.** Sex, height, body weight, BSA (directly), serum
  creatinine, creatinine clearance, AST, ALT, albumin, ECMO and mechanical
  ventilation were screened (Tables S1-S2) and not retained. An alternative
  model with serum creatinine, height and age on CL (OBJ 1474) was rejected in
  favour of the eGFR model "considering clinical usefulness". Its estimates
  are not published, so it is not included.
* **Scope.** The data are troughs only, and remdesivir itself was not
  measured, so the model does not describe the post-infusion peak of
  GS-441524 formation. Apply it to trough-level exposure.
* **Figure 4 medians.** The Figure 4 medians used in the age check were read
  off the published figure by eye by the maintainers (about +/- 20 ng/mL). They
  are used only in that check, never as parameter values.
* **Study-like cohort.** The eGFR, age and BSA distributions of the final
  cohort are reconstructed from the published renal-band counts and the Table
  1 medians and ranges. They are not the patient-level data. The share of
  patients aged 75 or older is not reported; the age draw puts about 40% there.
