---
title: "Piperacillin/tazobactam in sepsis and septic shock (Sulaiman 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piperacillin/tazobactam in sepsis and septic shock (Sulaiman 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sulaiman_2026_piperacillin_tazobactam"))$reference`
* Description: `r rxode2::rxode(readModelDb("Sulaiman_2026_piperacillin_tazobactam"))$description`
* Article: <https://doi.org/10.1093/jac/dkag199>

```{r model}
mod <- readModelDb("Sulaiman_2026_piperacillin_tazobactam")
ui  <- rxode2::rxode(mod)
th  <- ui$theta
```

Piperacillin and tazobactam are co-formulated in a fixed 8:1 ratio, so a 4.5 g
dose delivers 4 g of piperacillin and 0.5 g of tazobactam. The model gives each
drug its own one-compartment, first-order-elimination disposition; the two are
fitted together but do not interconvert. Piperacillin uses the unsuffixed
canonical names and tazobactam carries the `_taz` sibling-drug suffix.

```{r constants}
# Fixed clinical facts used throughout, all from the source paper.
PIP_MG  <- 4000   # piperacillin per 4.5 g dose (g)  -- clinical 8:1 combination
TAZ_MG  <-  500   # tazobactam   per 4.5 g dose
TAU     <-    6   # dosing interval (h) of the 82%-majority regimen, Table 1
FU      <- 0.70   # unbound fraction; Methods assume 30% protein binding
MIC_PIP <-   16   # EUCAST ECOFF MIC for P. aeruginosa (mg/L)
CT_TAZ  <-    2   # tazobactam critical threshold (mg/L)
TOX_PIP <-  160   # piperacillin Cmin / Css toxicity threshold (mg/L)
```

# Population

Forty-five adults with sepsis or septic shock by the Sepsis-3 criteria were
recruited between March 2017 and March 2018 in three Malaysian intensive care
units (Sultan Ahmad Shah Medical Centre, Hospital Universiti Sains Malaysia,
University of Malaya Medical Centre). Median age was 61 years (range 18-88),
median weight 65 kg (range 31-114) and 20/45 (44%) were female. Median
Cockcroft-Gault creatinine clearance on admission was 70 mL/min (range 30-161)
and median serum albumin 24 g/L, with hypoalbuminaemia in 53%. Median APACHE II
was 14 and median SOFA 6; 73% were mechanically ventilated and the lung was the
primary infection site in 71%. Patients on renal replacement therapy, or with a
creatinine clearance below 30 mL/min or a plasma creatinine above 200 umol/L,
were excluded, so the cohort spans preserved to augmented renal function only.

Most patients (82%) received 4.5 g every 6 h; administration was by intermittent
infusion in 31%, extended (3-4 h) infusion in 29% and continuous infusion in 40%,
the last preceded by a 4.5 g loading dose over 30-60 min. A total of 381 plasma
samples were drawn across two occasions -- day 1 and day 3 of therapy -- with 26
patients contributing both occasions and 19 a single occasion (Table 1, Results).

The same information is available programmatically from the model's `population`
metadata (`readModelDb("Sulaiman_2026_piperacillin_tazobactam")()$population`).

# Source trace

Every `ini()` entry carries an in-file comment naming its origin in
`inst/modeldb/specificDrugs/Sulaiman_2026_piperacillin_tazobactam.R`. They are
collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| One-compartment, first-order elimination, both drugs | n/a | Results, "Pharmacokinetic model building" |
| `lcl` (piperacillin CL) | 8.78 L/h | Table 2, `CLpip`, %RSE 5.69 |
| `lvc` (piperacillin V) | 25.53 L | Table 2, `Vpip`, %RSE 9.52 |
| `lcl_taz` (tazobactam CL) | 8.74 L/h | Table 2, `CLtaz`, %RSE 6.39 |
| `lvc_taz` (tazobactam V) | 27.15 L | Table 2, `Vtaz`, %RSE 9.77 |
| `CL = CLpop * (CLcr/71)^0.87` | n/a | Results, first display equation |
| `CL = CLpop * (CLcr/71)^0.90` | n/a | Results, second display equation |
| `e_crcl_cl` | 0.87 | Table 2, "CLcr effect on CLpip", %RSE 10.2 |
| `e_crcl_cl_taz` | 0.90 | Table 2, "CLcr effect on CLtaz", %RSE 10.8 |
| `crcl_ref_cl` | 71 mL/min | Results, "normalized to the population median value of 71 mL/min" |
| `etalcl` | 0.0576 | Table 2, BSV `CLpip` 24.35% |
| `etalvc` | 0.1681 | Table 2, BSV `Vpip` 42.78% |
| `etalcl_taz` | 0.0900 | Table 2, BSV `CLtaz` 30.69% |
| `etalvc_taz` | 0.1681 | Table 2, BSV `Vtaz` 42.78% |
| `etaiov_cl_*` | 0.0961 | Table 2, BOV `CLpip` 31.76% |
| `etaiov_vc_*` | 0.0729 | Table 2, BOV `Vpip` 27.50% |
| `etaiov_cl_taz_*` | 0.1024 | Table 2, BOV `CLtaz` 32.84% |
| `etaiov_vc_taz_*` | 0.0576 | Table 2, BOV `Vtaz` 24.35% |
| `addSd` / `propSd` | 2.43 mg/L / 17% | Table 2, "Random error", piperacillin |
| `addSd_taz` / `propSd_taz` | 0.40 mg/L / 16% | Table 2, "Random error", tazobactam |

## The variability table reports CV%, not omega

Table 2 heads its between-subject and between-occasion blocks with "(%)" but
never states the transformation. The two readings -- a raw log-scale standard
deviation reported as a percentage, or a coefficient of variation computed from
one -- differ by a factor that matters, so the choice is a transcription
decision, not a formatting one.

It is settled by inverting the published percentages. Under the log-normal
identity `CV% = sqrt(exp(omega^2) - 1) * 100`, **all eight** published values
invert to an `omega` that is round to exactly two decimal places. A raw-SD
reading would leave eight arbitrary four-figure numbers.

```{r iiv-scale}
published_cv <- tibble::tribble(
  ~block,                          ~parameter,           ~cv_pct,
  "Between-subject variability",   "CL piperacillin",      24.35,
  "Between-subject variability",   "V piperacillin",       42.78,
  "Between-subject variability",   "CL tazobactam",        30.69,
  "Between-subject variability",   "V tazobactam",         42.78,
  "Between-occasion variability",  "CL piperacillin",      31.76,
  "Between-occasion variability",  "V piperacillin",       27.50,
  "Between-occasion variability",  "CL tazobactam",        32.84,
  "Between-occasion variability",  "V tazobactam",         24.35
) |>
  mutate(
    omega_implied  = sqrt(log(1 + (cv_pct / 100)^2)),
    omega_rounded  = round(omega_implied, 2),
    cv_recomputed  = sqrt(exp(omega_rounded^2) - 1) * 100,
    variance       = omega_rounded^2,
    residual_pct   = abs(cv_recomputed - cv_pct)
  )

# Deterministic (no simulation): every published CV% must be reproduced by a
# two-decimal omega to better than 0.01 percentage points.
stopifnot(max(published_cv$residual_pct) < 0.01)

published_cv |>
  select(block, parameter, cv_pct, omega_rounded, variance, cv_recomputed) |>
  rename(
    "Block"              = block,
    "Parameter"          = parameter,
    "Published CV (%)"   = cv_pct,
    "Implied omega"      = omega_rounded,
    "Variance in ini()"  = variance,
    "Recomputed CV (%)"  = cv_recomputed
  ) |>
  knitr::kable(digits = 4,
               caption = "Table 2's '%' columns invert exactly to two-decimal omegas.")
```

The `ini()` variances therefore reproduce Table 2 to the last printed digit, and
the model file records the inversion next to every eta.

# The published continuous-infusion steady state validates both clearances

The Figure 1 caption reports observed median concentrations at 50% and 100% of
the dosing interval by administration method. For the continuous-infusion arm
these two are the same number, which identifies them as a steady state, and at
steady state `Css = infusion rate / CL` -- a closed form with no volume term and
no simulation noise.

Inverting each published continuous-infusion concentration through the model's
own covariate equation therefore recovers the creatinine clearance of the
patients who produced it. Four cells (two drugs x two occasions) are available
and they must agree.

```{r css-inversion}
rate_pip <- PIP_MG / TAU   # mg/h delivered continuously
rate_taz <- TAZ_MG / TAU

crcl_from_css <- function(css, rate, cl_pop, expo) {
  71 * ((rate / css) / cl_pop)^(1 / expo)
}

css_obs <- tibble::tribble(
  ~occasion,             ~drug,          ~css_mgL,
  "Occasion 1 (day 1)",  "piperacillin",   134.22,
  "Occasion 1 (day 1)",  "tazobactam",      15.08,
  "Occasion 2 (day 3)",  "piperacillin",   124.42,
  "Occasion 2 (day 3)",  "tazobactam",      15.39
) |>
  mutate(
    rate_mgh   = if_else(drug == "piperacillin", rate_pip, rate_taz),
    cl_implied = rate_mgh / css_mgL,
    crcl_implied = if_else(
      drug == "piperacillin",
      crcl_from_css(css_mgL, rate_mgh, exp(th[["lcl"]]),     th[["e_crcl_cl"]]),
      crcl_from_css(css_mgL, rate_mgh, exp(th[["lcl_taz"]]), th[["e_crcl_cl_taz"]])
    )
  )

css_obs |>
  rename(
    "Occasion"                    = occasion,
    "Drug"                        = drug,
    "Published Css (mg/L)"        = css_mgL,
    "Infusion rate (mg/h)"        = rate_mgh,
    "Implied CL (L/h)"            = cl_implied,
    "Implied CLcr (mL/min)"       = crcl_implied
  ) |>
  knitr::kable(digits = 2,
               caption = paste("Continuous-infusion steady state inverted through the",
                               "model's covariate equation."))
```

All four cells land between `r sprintf("%.0f", min(css_obs$crcl_implied))` and
`r sprintf("%.0f", max(css_obs$crcl_implied))` mL/min -- a tight cluster well
inside the cohort's 30-161 mL/min range, and squarely consistent with Table 3,
where the two groups that the continuous-infusion patients dominate had median
creatinine clearances of 46 mL/min (therapeutic, 10/20 on continuous infusion)
and 30 mL/min (toxic, 4/4 on continuous infusion). Two chemically distinct drugs
with independently estimated clearances and independently estimated exponents
agree on the same answer.

The check is made non-circular by deriving one creatinine clearance from a single
cell -- occasion 1, piperacillin -- and using it to *predict* the other three.

```{r css-crossprediction}
crcl_star <- css_obs$crcl_implied[1]   # from occasion-1 piperacillin only

predict_css <- function(drug, crcl) {
  if (drug == "piperacillin") {
    rate_pip / (exp(th[["lcl"]])     * (crcl / 71)^th[["e_crcl_cl"]])
  } else {
    rate_taz / (exp(th[["lcl_taz"]]) * (crcl / 71)^th[["e_crcl_cl_taz"]])
  }
}

cross <- css_obs |>
  slice(2:4) |>
  mutate(
    css_predicted = vapply(drug, predict_css, numeric(1), crcl = crcl_star),
    pct_diff      = 100 * (css_predicted - css_mgL) / css_mgL
  )

# Deterministic: no cohort is drawn, so these bounds are reproducible anywhere.
stopifnot(
  # The four independently inverted CLcr values agree with each other.
  max(css_obs$crcl_implied) / min(css_obs$crcl_implied) < 1.35,
  all(css_obs$crcl_implied > 30, css_obs$crcl_implied < 60),
  # One cell predicts the other three within 20% (realised max 14.0%).
  max(abs(cross$pct_diff)) < 20
)

cross |>
  select(occasion, drug, css_mgL, css_predicted, pct_diff) |>
  rename(
    "Occasion"                = occasion,
    "Drug"                    = drug,
    "Published Css (mg/L)"    = css_mgL,
    "Predicted Css (mg/L)"    = css_predicted,
    "% difference"            = pct_diff
  ) |>
  knitr::kable(digits = 2,
               caption = paste0("Occasion-1 piperacillin implies CLcr = ",
                                sprintf("%.1f", crcl_star),
                                " mL/min; that value predicts the other three cells."))
```

A mis-transcribed clearance, exponent, dose or unit in either drug breaks this
immediately, because the two drugs would then disagree about the creatinine
clearance of the same patients.

## The intermittent-infusion decay slope validates both volumes

The steady-state check above is blind to the volume of distribution. The
intermittent-infusion arm is not: between 50% and 100% of a 6 h interval the
concentration is in pure post-infusion decay, so the ratio of the two published
medians gives an apparent elimination rate constant, and `kel = CL / V` converts
it back to a creatinine clearance.

```{r decay-slope}
decay_obs <- tibble::tribble(
  ~occasion,            ~drug,           ~c_50pct, ~c_100pct,
  "Occasion 1 (day 1)", "piperacillin",     53.36,     24.84,
  "Occasion 1 (day 1)", "tazobactam",        6.70,      2.79,
  "Occasion 2 (day 3)", "piperacillin",     55.63,     16.52,
  "Occasion 2 (day 3)", "tazobactam",        6.12,      2.18
) |>
  mutate(
    kel_apparent = log(c_50pct / c_100pct) / (TAU / 2),
    cl_implied   = if_else(drug == "piperacillin",
                           exp(th[["lvc"]]), exp(th[["lvc_taz"]])) * kel_apparent,
    crcl_implied = if_else(
      drug == "piperacillin",
      71 * (cl_implied / exp(th[["lcl"]]))^(1 / th[["e_crcl_cl"]]),
      71 * (cl_implied / exp(th[["lcl_taz"]]))^(1 / th[["e_crcl_cl_taz"]])
    )
  )

drug_ratio <- decay_obs |>
  group_by(occasion) |>
  summarise(ratio = max(crcl_implied) / min(crcl_implied), .groups = "drop")

# Deterministic. The medians at 50% and 100% of the interval are medians of
# possibly different subjects, so this is a coarser check than the steady-state
# one: it asks only that both volumes place the arm inside the cohort's own
# renal-function range, and that the two drugs agree to within 1.5-fold.
# Realised: CLcr 50-86 mL/min, drug ratios 1.27 and 1.12.
stopifnot(
  all(decay_obs$crcl_implied > 30, decay_obs$crcl_implied < 161),
  all(drug_ratio$ratio < 1.5)
)

decay_obs |>
  select(occasion, drug, c_50pct, c_100pct, kel_apparent, crcl_implied) |>
  rename(
    "Occasion"                    = occasion,
    "Drug"                        = drug,
    "Published C at 50% (mg/L)"   = c_50pct,
    "Published C at 100% (mg/L)"  = c_100pct,
    "Apparent kel (1/h)"          = kel_apparent,
    "Implied CLcr (mL/min)"       = crcl_implied
  ) |>
  knitr::kable(digits = 3,
               caption = paste("Intermittent-infusion decay between 50% and 100% of the",
                               "interval, inverted through kel = CL / V."))
```

# Virtual cohort

Original observed data are not publicly available, so the cohort below draws
creatinine clearance from a log-normal matched to the published median of
70 mL/min and mean of 75 mL/min and truncated to the observed 30-161 mL/min
range. Weight, albumin and age are not covariates in the final model and are not
simulated.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, whose
# streams are partitioned per solver thread, so this cohort differs between a
# 2-thread CI runner and a 16-thread workstation. Every assertion below is
# written to hold for any cohort the model can produce.
set.seed(20260901)

DOSE_ROWS <- function(regimen) {
  spec <- switch(
    regimen,
    "Continuous infusion"           = list(t = c(0, 0.5, 6.5, 12.5, 18.5),
                                           dur = c(0.5, rep(TAU, 4))),
    "Extended infusion (4 h)"       = list(t = seq(0, 18, by = TAU), dur = rep(4, 4)),
    "Intermittent infusion (0.5 h)" = list(t = seq(0, 18, by = TAU), dur = rep(0.5, 4)),
    stop("unknown regimen: ", regimen)
  )
  n <- length(spec$t)
  data.frame(
    time = rep(spec$t, each = 2),
    amt  = rep(c(PIP_MG, TAZ_MG), n),
    rate = rep(c(PIP_MG, TAZ_MG), n) / rep(spec$dur, each = 2),
    evid = 1L,
    # Dose rows address the real ODE state; observation rows below use dvid
    # alone, which is the form that works for a two-endpoint rxUi.
    cmt  = rep(c("central", "central_taz"), n),
    dvid = NA_integer_
  )
}

OBS_ROWS <- function(grid) {
  data.frame(time = grid, amt = NA_real_, rate = NA_real_, evid = 0L,
             cmt = NA_character_, dvid = 1L)
}

# Expand a per-subject frame (id, CRCL, OCC, regimen, ...) into an event table.
make_events <- function(subjects, grid) {
  bind_rows(lapply(split(subjects, subjects$regimen), function(g) {
    tpl <- bind_rows(DOSE_ROWS(g$regimen[1]), OBS_ROWS(grid))
    g[rep(seq_len(nrow(g)), each = nrow(tpl)), , drop = FALSE] |>
      bind_cols(tpl[rep(seq_len(nrow(tpl)), nrow(g)), , drop = FALSE])
  })) |>
    arrange(id, time, desc(evid))
}

# Log-normal with median 70 and mean 75 mL/min, rejection-truncated to 30-161.
draw_crcl <- function(n) {
  sigma <- sqrt(2 * log(75 / 70))
  out <- numeric(0)
  while (length(out) < n) {
    y <- rlnorm(3 * n, meanlog = log(70), sdlog = sigma)
    out <- c(out, y[y >= 30 & y <= 161])
  }
  out[seq_len(n)]
}

N_PER_ARM <- 100   # well under the 200-per-arm cap
regimens  <- c("Intermittent infusion (0.5 h)", "Extended infusion (4 h)",
               "Continuous infusion")

subjects <- bind_rows(lapply(seq_along(regimens), function(k) {
  data.frame(
    id      = (k - 1L) * N_PER_ARM + seq_len(N_PER_ARM),   # disjoint id ranges
    CRCL    = draw_crcl(N_PER_ARM),
    OCC     = 1L,                                          # day 1 = occasion 1
    regimen = regimens[k]
  )
}))

events <- make_events(subjects, seq(0, 24, by = 0.25))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events = events,
  keep      = "regimen",
  # rxode2's automatic ODE -> linCmt conversion corrupts the dvid mapping for
  # multi-endpoint models; see known-vignette-failure-patterns.md pattern 5b.
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  filter(!is.na(Cc))

stopifnot(nrow(sim) > 0, all(sim$Cc >= 0), all(sim$Cc_taz >= 0))
```

```{r figure-1-profiles}
# Replicates the concentration-time picture of Figure 1 of Sulaiman 2026:
# observed piperacillin (a, b) and tazobactam (c, d) by method of administration.
sim |>
  select(time, regimen, Piperacillin = Cc, Tazobactam = Cc_taz) |>
  pivot_longer(c(Piperacillin, Tazobactam), names_to = "drug", values_to = "conc") |>
  group_by(time, regimen, drug) |>
  summarise(Q05 = quantile(conc, 0.05), Q50 = median(conc),
            Q95 = quantile(conc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_hline(data = data.frame(drug = c("Piperacillin", "Tazobactam"),
                               y = c(MIC_PIP, CT_TAZ)),
             aes(yintercept = y), linetype = "dashed") +
  facet_grid(drug ~ regimen, scales = "free_y") +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time (h)", y = "Total plasma concentration (mg/L)",
       caption = paste("Replicates Figure 1 of Sulaiman 2026. Median with 5th-95th",
                       "percentile band; dashed lines are the EUCAST ECOFF MIC of",
                       "16 mg/L (piperacillin) and the 2 mg/L tazobactam threshold."))
```

## Concentrations at 50% and 100% of the dosing interval

The Figure 1 caption tabulates observed median (IQR) concentrations at 50% and
100% of the dosing interval, by drug and administration method. Those are
reproduced below against the simulated cohort, taking the 18-24 h interval of
day 1.

An important confound has to be stated first. The published arms are **not**
renal-function-matched: Table 3 shows continuous infusion concentrated in the
low-creatinine-clearance groups (median 46 and 30 mL/min) and intermittent
infusion in the high group (median 112 mL/min), and the steady-state inversion
above put the continuous-infusion arm at roughly
`r sprintf("%.0f", crcl_star)` mL/min. The virtual cohort here draws all three
arms from the *whole-cohort* distribution (median 70 mL/min), so the
continuous-infusion arm is expected to sit low. It is carried as a known,
quantified deviation rather than tuned away.

```{r figure-1-table}
published_f1 <- tibble::tribble(
  ~regimen,                        ~drug,           ~timepoint,          ~med,   ~q25,   ~q75,
  "Intermittent infusion (0.5 h)", "Piperacillin",  "50% of interval",   53.36,  28.48, 122.70,
  "Intermittent infusion (0.5 h)", "Piperacillin",  "100% of interval",  24.84,   6.28,  50.93,
  "Intermittent infusion (0.5 h)", "Tazobactam",    "50% of interval",    6.70,   3.59,  14.49,
  "Intermittent infusion (0.5 h)", "Tazobactam",    "100% of interval",   2.79,   1.13,   7.16,
  "Extended infusion (4 h)",       "Piperacillin",  "50% of interval",   76.87,  34.28, 181.69,
  "Extended infusion (4 h)",       "Piperacillin",  "100% of interval",  22.99,   4.10,  77.09,
  "Extended infusion (4 h)",       "Tazobactam",    "50% of interval",    8.49,   3.87,  22.34,
  "Extended infusion (4 h)",       "Tazobactam",    "100% of interval",   2.49,   0.72,   7.60,
  "Continuous infusion",           "Piperacillin",  "50% of interval",  142.30, 101.61, 202.07,
  "Continuous infusion",           "Piperacillin",  "100% of interval", 134.22, 106.68, 177.97,
  "Continuous infusion",           "Tazobactam",    "50% of interval",   18.29,  13.25,  30.67,
  "Continuous infusion",           "Tazobactam",    "100% of interval",  15.08,  13.10,  31.50
)

simulated_f1 <- sim |>
  filter(abs(time - 21) < 1e-8 | abs(time - 24) < 1e-8) |>
  mutate(timepoint = if_else(abs(time - 21) < 1e-8,
                             "50% of interval", "100% of interval")) |>
  select(regimen, timepoint, Piperacillin = Cc, Tazobactam = Cc_taz) |>
  pivot_longer(c(Piperacillin, Tazobactam), names_to = "drug", values_to = "conc") |>
  group_by(regimen, drug, timepoint) |>
  summarise(sim_med = median(conc), .groups = "drop")

f1 <- published_f1 |>
  left_join(simulated_f1, by = c("regimen", "drug", "timepoint")) |>
  mutate(
    in_iqr   = sim_med >= q25 & sim_med <= q75,
    ratio    = sim_med / med
  )
stopifnot(nrow(f1) == 12L, !anyNA(f1$sim_med))

# The renal-function-matched arms (intermittent, extended) must land inside the
# published interquartile range in every one of their eight cells. Those IQRs
# are wide, but a mis-transcribed clearance, volume, dose or unit moves the
# whole distribution far enough to break several at once. The continuous arm is
# excluded and reported separately (see the narrative above).
matched <- f1 |> filter(regimen != "Continuous infusion")
stopifnot(nrow(matched) == 8L, all(matched$in_iqr))

# Continuous infusion: the model must be LOW here, by the amount the
# creatinine-clearance confound predicts, not high and not in agreement.
ci_arm <- f1 |> filter(regimen == "Continuous infusion")
stopifnot(all(ci_arm$ratio < 1), all(ci_arm$ratio > 0.3))

f1 |>
  transmute(
    regimen, drug, timepoint,
    published = sprintf("%.2f (%.2f-%.2f)", med, q25, q75),
    simulated = sprintf("%.2f", sim_med),
    ratio,
    flag = if_else(in_iqr, "", "outside published IQR")
  ) |>
  rename(
    "Administration"            = regimen,
    "Drug"                      = drug,
    "Timepoint"                 = timepoint,
    "Published median (IQR)"    = published,
    "Simulated median"          = simulated,
    "Simulated / published"     = ratio,
    "Note"                      = flag
  ) |>
  knitr::kable(digits = 2,
               caption = paste("Occasion 1 concentrations at 50% and 100% of the",
                               "dosing interval (Figure 1 caption of Sulaiman 2026)."))
```

Every intermittent- and extended-infusion cell falls inside the published
interquartile range. The continuous-infusion cells come out low by a factor of
about `r sprintf("%.1f", 1 / mean(ci_arm$ratio))`, which is what the
`r sprintf("%.0f", crcl_star)` mL/min inferred for that arm predicts against the
70 mL/min the virtual cohort was given -- the deviation is a property of the
cohort, not of the model.

# PKNCA validation

The source paper reports no non-compartmental parameters, so PKNCA is used here
as an independent implementation check: NCA run on a simulated single 4.5 g dose
must recover the clearance and half-life the model was solved with. Because both
sides of the comparison use the same drawn parameters, the difference is pure
numerical error and the tolerances are correspondingly tight.

```{r nca-simulate}
nca_strata <- c(50, 71, 110)
nca_subjects <- bind_rows(lapply(seq_along(nca_strata), function(k) {
  data.frame(
    id      = (k - 1L) * 60L + seq_len(60L),
    CRCL    = nca_strata[k],
    OCC     = 1L,
    regimen = sprintf("CLcr %d mL/min", nca_strata[k])
  )
}))

# Single 4.5 g dose over 30 min, sampled to 12 h (about six half-lives).
nca_events <- nca_subjects[rep(seq_len(nrow(nca_subjects)), each = 2), ] |>
  mutate(time = 0, amt = rep(c(PIP_MG, TAZ_MG), nrow(nca_subjects)),
         rate = rep(c(PIP_MG, TAZ_MG), nrow(nca_subjects)) / 0.5,
         evid = 1L, cmt = rep(c("central", "central_taz"), nrow(nca_subjects)),
         dvid = NA_integer_) |>
  bind_rows(
    nca_subjects[rep(seq_len(nrow(nca_subjects)), each = 49L), ] |>
      mutate(time = rep(seq(0, 12, by = 0.25), nrow(nca_subjects)),
             amt = NA_real_, rate = NA_real_, evid = 0L,
             cmt = NA_character_, dvid = 1L)
  ) |>
  arrange(id, time, desc(evid))

nca_sim <- rxode2::rxSolve(mod, events = nca_events, keep = "regimen",
                           useLinCmt = FALSE, returnType = "data.frame") |>
  filter(!is.na(Cc))
stopifnot(all(nca_sim$Cc >= 0), all(nca_sim$Cc_taz >= 0))

# One row per subject holding the parameters the ODE was actually solved with.
subject_pars <- nca_sim |>
  group_by(regimen, id) |>
  summarise(cl = first(cl), vc = first(vc),
            cl_taz = first(cl_taz), vc_taz = first(vc_taz), .groups = "drop")
```

```{r nca-run}
run_nca <- function(conc_col, dose_mg) {
  conc <- nca_sim |>
    transmute(id, time, regimen, Cc = .data[[conc_col]])
  # Guarantee a time-zero record; for an IV infusion the pre-dose value is 0.
  conc <- bind_rows(
    conc,
    conc |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, regimen, time, .keep_all = TRUE) |>
    arrange(id, regimen, time)

  dose <- nca_events |>
    filter(evid == 1L, amt == dose_mg) |>
    select(id, time, amt, regimen)

  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | regimen + id,
                               concu = "ug/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | regimen + id, doseu = "mg")

  intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                          aucinf.obs = TRUE, half.life = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_pip <- run_nca("Cc",     PIP_MG)
nca_taz <- run_nca("Cc_taz", TAZ_MG)

tidy_nca <- function(res, dose_mg, label) {
  as.data.frame(res$result) |>
    filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
    select(regimen, id, PPTESTCD, PPORRES) |>
    pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
    mutate(drug = label, cl_nca = dose_mg / aucinf.obs)
}

nca_tbl <- bind_rows(tidy_nca(nca_pip, PIP_MG, "Piperacillin"),
                     tidy_nca(nca_taz, TAZ_MG, "Tazobactam")) |>
  left_join(subject_pars, by = c("regimen", "id")) |>
  mutate(
    cl_model  = if_else(drug == "Piperacillin", cl, cl_taz),
    vc_model  = if_else(drug == "Piperacillin", vc, vc_taz),
    thalf_model = log(2) * vc_model / cl_model,
    cl_pct    = 100 * (cl_nca - cl_model) / cl_model,
    thalf_pct = 100 * (half.life - thalf_model) / thalf_model
  )
stopifnot(nrow(nca_tbl) == 360L, !anyNA(nca_tbl$cl_pct), !anyNA(nca_tbl$thalf_pct))

# Both sides use the SAME drawn parameters, so this is numerical error only and
# a tight all() bound is the correct form (not a quantile envelope).
stopifnot(
  max(abs(nca_tbl$cl_pct))    < 2,
  max(abs(nca_tbl$thalf_pct)) < 2,
  all(abs(nca_tbl$tmax - 0.5) < 1e-8)   # Cmax at end of the 30-min infusion
)
```

```{r nca-table}
nca_tbl |>
  group_by(drug, regimen) |>
  summarise(
    cmax       = median(cmax),
    aucinf.obs = median(aucinf.obs),
    half.life  = median(half.life),
    thalf_ref  = median(thalf_model),
    cl_nca     = median(cl_nca),
    cl_ref     = median(cl_model),
    .groups    = "drop"
  ) |>
  rename(
    "Drug"                        = drug,
    "Stratum"                     = regimen,
    "Cmax (mg/L)"                 = cmax,
    "AUC0-inf (mg*h/L)"           = aucinf.obs,
    "t1/2 by PKNCA (h)"           = half.life,
    "t1/2 = ln2*V/CL (h)"         = thalf_ref,
    "CL from NCA (L/h)"           = cl_nca,
    "CL in the model (L/h)"       = cl_ref
  ) |>
  knitr::kable(digits = 3,
               caption = paste("PKNCA on a simulated single 4.5 g dose recovers the",
                               "model's own clearance and half-life at every",
                               "creatinine-clearance stratum."))
```

The typical-value half-lives -- `r sprintf("%.2f", log(2) * exp(th[["lvc"]]) / exp(th[["lcl"]]))` h
for piperacillin and `r sprintf("%.2f", log(2) * exp(th[["lvc_taz"]]) / exp(th[["lcl_taz"]]))` h
for tazobactam at the reference creatinine clearance -- match the 0.7-1.5 h
usually quoted for these agents in patients with normal renal function once the
reduced clearance of this cohort is accounted for, and are consistent with the
Discussion's statement that clearance here "was at the lower end of the reported
range".

# Probability of target attainment (Figure 5)

The paper's headline analysis simulates 15 regimens and scores, per creatinine
clearance stratum, the joint efficacy target -- 100% *f*T > MIC for piperacillin
combined with 85% *f*T > 2 mg/L for tazobactam -- and the piperacillin toxicity
threshold of a total trough or steady-state concentration above 160 mg/L, with
free concentrations obtained by assuming 30% protein binding. Three
representative regimens are reproduced here across the paper's creatinine
clearance range.

```{r pta-simulate}
pta_regimens <- c("Intermittent infusion (0.5 h)", "Extended infusion (4 h)",
                  "Continuous infusion")
pta_strata   <- c(30, 50, 70, 90, 110, 130)
N_PTA        <- 50

pta_subjects <- expand_grid(regimen = pta_regimens, CRCL = pta_strata,
                            k = seq_len(N_PTA)) |>
  mutate(id = row_number(), OCC = 1L, stratum = CRCL) |>
  select(-k) |>
  as.data.frame()

# Observations only where the metrics are computed: the final dosing interval of
# day 1, plus a time-zero anchor.
pta_events <- make_events(pta_subjects, c(0, seq(18, 24, by = 0.1)))
stopifnot(!anyDuplicated(unique(pta_events[, c("id", "time", "evid")])))

pta_sim <- rxode2::rxSolve(mod, events = pta_events,
                           keep = c("regimen", "stratum"),
                           useLinCmt = FALSE, returnType = "data.frame") |>
  filter(!is.na(Cc), time > 18)
stopifnot(nrow(pta_sim) > 0, all(pta_sim$Cc >= 0))

pta <- pta_sim |>
  group_by(regimen, stratum, id) |>
  summarise(
    # 100% fT > MIC for piperacillin: the free trough must clear the MIC.
    eff_pip = min(FU * Cc) > MIC_PIP,
    # 85% fT > CT for tazobactam: fraction of the interval above threshold.
    eff_taz = mean(FU * Cc_taz > CT_TAZ) >= 0.85,
    # Toxicity on TOTAL piperacillin Cmin / Css.
    tox     = min(Cc) > TOX_PIP,
    .groups = "drop"
  ) |>
  group_by(regimen, stratum) |>
  summarise(PTA_eff = mean(eff_pip & eff_taz), PTA_tox = mean(tox), .groups = "drop")
stopifnot(nrow(pta) == length(pta_regimens) * length(pta_strata))
```

```{r figure-5}
# Replicates Figure 5 of Sulaiman 2026: PTA for efficacy and for piperacillin
# toxicity by renal function on day 1 of therapy.
pta |>
  pivot_longer(c(PTA_eff, PTA_tox), names_to = "endpoint", values_to = "pta") |>
  mutate(endpoint = recode(endpoint,
                           PTA_eff = "Efficacy (joint target)",
                           PTA_tox = "Piperacillin toxicity")) |>
  ggplot(aes(stratum, 100 * pta, colour = regimen)) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  geom_line() +
  geom_point() +
  facet_wrap(~endpoint) +
  scale_x_continuous(breaks = pta_strata) +
  labs(x = "Cockcroft-Gault creatinine clearance (mL/min)",
       y = "Probability of target attainment (%)", colour = NULL,
       caption = paste("Replicates Figure 5 of Sulaiman 2026. Dashed line is the",
                       "90% PTA criterion the paper uses to call a regimen optimal.")) +
  theme(legend.position = "bottom")

pta |>
  mutate(across(c(PTA_eff, PTA_tox), ~ 100 * .x)) |>
  pivot_wider(names_from = stratum, values_from = c(PTA_eff, PTA_tox)) |>
  knitr::kable(digits = 0,
               caption = "PTA (%) for efficacy and piperacillin toxicity by CLcr stratum.")
```

```{r pta-claims}
pooled <- pta |> group_by(regimen) |> summarise(m = mean(PTA_eff), .groups = "drop")
get_pool <- function(r) {
  v <- pooled$m[pooled$regimen == r]
  if (length(v) != 1L) stop("no unique pooled PTA for '", r, "'")
  v
}
ci_mid <- pta |> filter(regimen == "Continuous infusion", stratum %in% c(50, 70, 90, 110))
ii_hi  <- pta |> filter(regimen == "Intermittent infusion (0.5 h)", stratum >= 70)
stopifnot(nrow(ci_mid) == 4L, nrow(ii_hi) == 4L)

by_stratum <- pta |> group_by(stratum) |> summarise(m = mean(PTA_eff), .groups = "drop")
low_crcl  <- by_stratum$m[by_stratum$stratum ==  30]
high_crcl <- by_stratum$m[by_stratum$stratum == 130]

tox_high <- pta |> filter(stratum >= 90)
tox_ci30 <- pta$PTA_tox[pta$regimen == "Continuous infusion" & pta$stratum == 30]
stopifnot(length(tox_ci30) == 1L)

# Bounds were set from three independent draws (two at 16 solver threads, one at
# 2) and placed OUTSIDE the observed range, because rxode2 partitions its RNG
# streams per thread and CI draws a different cohort. Realised ranges are in the
# comment on each line; do not tighten them back to a single run.
stopifnot(
  # "PTA improved progressively from intermittent to extended to continuous"
  # (Results, "Dosing simulations and PTA"). Realised gaps 0.32-0.36 and
  # 0.27-0.31.
  get_pool("Continuous infusion") - get_pool("Extended infusion (4 h)") > 0.15,
  get_pool("Extended infusion (4 h)") -
    get_pool("Intermittent infusion (0.5 h)") > 0.15,
  # "PTA markedly declined with increasing CLcr". Trend across the whole range,
  # not step by step -- the extended-infusion arm is not monotone between the
  # 110 and 130 strata in every draw. Realised decline 0.53-0.60.
  low_crcl - high_crcl > 0.30,
  # "optimal PTA for efficacy was achieved only with continuous infusion
  # regimens in patients with CLcr <= 110". Realised minimum 0.94-0.96 for
  # continuous infusion over 50-110 mL/min.
  min(ci_mid$PTA_eff) >= 0.85,
  # ... and intermittent infusion is nowhere near the 90% criterion once renal
  # function is preserved. Realised maximum 0.18-0.32.
  max(ii_hi$PTA_eff) < 0.60,
  # "Toxic concentrations were generally uncommon (< 5% for most regimens), but
  # the probability increased substantially with ... reduced CLcr (<= 50)".
  # Realised: 0-0.007 at CLcr >= 90, and 0.42-0.54 for continuous infusion at
  # CLcr 30.
  mean(tox_high$PTA_tox) < 0.10,
  tox_ci30 > 0.10
)
```

Every qualitative claim the paper makes from Figure 5 is reproduced: attainment
improves from intermittent through extended to continuous infusion, falls with
rising creatinine clearance, reaches the 90% criterion only under continuous
infusion across the preserved-renal-function range, and carries a toxicity risk
that is negligible above 90 mL/min but substantial at 30 mL/min. That last point
is the paper's reason for recommending a 4.5 g loading dose followed by 4.5 g
every 6 h as a continuous infusion specifically for creatinine clearances of
50-130 mL/min, rather than for everyone.

# Assumptions and deviations

* **Between-subject and between-occasion variability scale.** Table 2 labels
  these columns "(%)" without saying what the percentage is. They are read as
  log-normal CV%, because all eight values invert to `omega` values that are
  round to two decimal places under `CV% = sqrt(exp(omega^2) - 1) * 100`
  (maximum residual 0.005 percentage points); a raw-standard-deviation reading
  would require eight arbitrary four-figure numbers. The inversion is shown
  above and recorded against every eta in the model file.
* **Residual-error combination rule.** The paper states only that "residual
  unexplained variability was described by a combined error model". Monolix
  offers two combined forms -- `combined1`, where the residual SD is
  `a + b * f`, and `combined2`, where it is `sqrt(a^2 + (b * f)^2)` -- and
  Table 2 reports `a` and `b` without distinguishing them. Text S4, which might
  resolve it, is not among the supplementary items released with the
  open-access article. The model uses nlmixr2's default `add() + prop()`, which
  is the `combined2` form. The two differ most where the additive and
  proportional contributions are comparable, i.e. in the low-concentration tail;
  they do not affect any of the median-based comparisons in this vignette.
* **Centring constant 71 vs 70 mL/min.** The Results text and both printed
  display equations normalise creatinine clearance to 71 mL/min, described as
  "the population median value", while Table 1 reports the cohort median as
  70 mL/min (mean 75, SD 40). The implemented value 71 is used. The 1 mL/min
  difference changes a typical clearance by 1.2%.
* **Between-occasion variability encoding.** rxode2 parses but cannot simulate
  the multi-level `eta ~ var | occ` syntax, so the two occasions are expanded
  into explicit indicator-selected eta slots (`etaiov_*_1`, `etaiov_*_2`) with
  the second occasion's variance `fixed()` to the first. This is the same
  shared-magnitude structure NONMEM writes as `$OMEGA BLOCK(1) SAME`, and it
  reproduces the single per-parameter magnitude Table 2 reports. Loading the
  model emits a `some etas defaulted to non-mu referenced` warning; that is the
  accepted cost of the idiom and affects estimation only, not simulation.
* **Supplementary material not on disk.** Tables S2, S6, S7, S8, S9 and Texts
  S1, S3, S4, S5 are referenced by the paper but were not retrievable
  (EuropePMC returns no supplementary bundle for PMC13280637 and the publisher
  landing page is access-controlled). Nothing in the model depends on them: all
  fixed effects, variances and residual-error terms are in Table 2 and both
  covariate equations are printed in the Results. What is lost is the exact
  combined-error form (above), the model-building trace of Table S6, and the
  full list of 15 simulated regimens in Text S5 -- of which three
  representative ones are reproduced here.
* **Covariates screened but not retained.** Body weight, serum albumin and age
  are recorded in Table 1 and discussed, but creatinine clearance on clearance
  was the only effect retained in the final model. They are documented in the
  model file's `covariatesDataExcluded` rather than `covariateData`, so no
  allometric or albumin term is implied. In particular the volumes of
  distribution carry no weight scaling, which the Discussion notes explicitly.
* **Virtual-cohort creatinine clearance distribution.** The paper reports the
  median (70), mean (75), standard deviation (40) and range (30-161) but not the
  shape. A log-normal matched to the median and mean and rejection-truncated to
  the observed range is used; its standard deviation (about 29 mL/min) is
  narrower than the reported 40. This affects the width of the simulated
  percentile bands, not the central tendency.
* **The published administration arms are not renal-function-matched.**
  Table 3 shows continuous-infusion patients concentrated in the low-creatinine-
  clearance groups and intermittent-infusion patients in the high group. The
  virtual cohort draws all three arms from the whole-cohort distribution, so the
  continuous-infusion arm simulates low against the Figure 1 caption values.
  This is reported as a quantified deviation, and the steady-state inversion
  above independently recovers roughly
  `r sprintf("%.0f", crcl_star)` mL/min for that arm, which accounts for it.
* **Protein binding.** The 30% figure (unbound fraction 0.70) used for the
  target-attainment analysis is an assumption the paper imports from the
  literature for critically ill patients, not a quantity it estimated; the
  popPK model itself was fitted to total plasma concentrations. It is therefore
  applied in this vignette rather than encoded in the model file.
* **Simulated regimens.** Three of the paper's 15 regimens are reproduced
  (intermittent 4.5 g q6h over 30 min, extended 4.5 g q6h over 4 h, and 4.5 g
  loading dose followed by 4.5 g q6h as a continuous infusion). The remaining
  twelve are enumerated only in Text S5, which is not on disk.
* **No non-paper-derived parameter values.** Every `ini()` entry comes from
  Table 2 or from one of the two printed covariate equations in the Results; no
  value was digitised from a figure, supplied by correspondence, or carried from
  an upstream model.
