---
title: "Maribavir (Sun 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maribavir (Sun 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sun_2025_maribavir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Sun_2025_maribavir"))$description`
* Article: <https://doi.org/10.1002/psp4.70054>
* Supporting Information (file `s001`, which contains the **final NONMEM control
  stream** plus Tables S1-S4): <https://doi.org/10.1002/psp4.70054>

Maribavir is an orally bioavailable benzimidazole riboside with a selective
multimodal mechanism of action against human cytomegalovirus (CMV). Sun 2025 is
an **update** of a previously published pooled population PK analysis: it adds
the randomized, double-blind phase 3 AURORA study in hematopoietic cell
transplant (HCT) recipients with first asymptomatic CMV infection, and a phase 1
study in Japanese-descended and non-Hispanic Caucasian individuals. Relative to
the earlier analysis the allometric body-weight exponents are now estimated
rather than fixed, and proton-pump-inhibitor (PPI) effects on both relative
bioavailability and the absorption rate constant are new.

A companion model from the same programme, `modellib("Sun_2023_maribavir")`,
encodes the earlier pooled analysis that supported the adolescent dosing
recommendation. The two are **not** interchangeable: that one fixes the weight
exponents at 0.75 / 1 and carries no PPI effect.

```{r model}
mod <- rxode2::rxode(readModelDb("Sun_2025_maribavir"))
mod
```

# Population

The parameter-estimation population is 930 individuals contributing 7431
maribavir plasma concentration records, pooled across phase 1, 2 and 3 studies
(Sun 2025 Table 1 and Section 3.1):

* **Phase 1 (n = 206, 4231 records)** -- 157 healthy volunteers, 19 with renal
  impairment, 20 stable renal transplant recipients and 10 with hepatic
  impairment. None have CMV infection.
* **Phase 2/3 (n = 724, 3200 records)** -- HCT (440) and solid organ transplant
  (SOT, 284) recipients with CMV infection, including the SOLSTICE refractory
  population. CMV category: asymptomatic 644, symptomatic 44, organ disease 36.
* **AURORA (n = 238)** -- a subset of the phase 2/3 group; all HCT recipients
  with first asymptomatic CMV infection. This is the cohort newly added in this
  update and the cohort used for the exposure-response analyses.

Demographics: 82% aged 18 to <65 years and 18% aged 65 to <80, with a **single**
individual under 18 -- that one individual is the stated reason the allometric
exponents were estimated rather than fixed, in support of an ongoing paediatric
phase 3 study. Weight median 73.0 kg (range 36.1-141), mean 74.2 (SD 17.4). 41%
female. Race: Caucasian 77%, Black 13%, Asian 7%, Other 3%.

Co-medication prevalence is high and matters for this model: PPIs 54%, strong
CYP3A4 inhibitors 13%, moderate CYP3A4 inhibitors 11%, H2 blockers 10%, antacids
8%, strong CYP3A4 inducers 2%.

Below-limit-of-quantification records (297; 3.5% of post-dose values) were
excluded under method M1. Estimation used NONMEM 7.5.1 with IMPMAP; confidence
intervals came from non-parametric bootstrap stratified by study.

# Source trace

Every `ini()` value comes from the published Table 2. The model **structure**
comes from the final NONMEM control stream in Supporting Information file
`s001`. The distinction matters: the control stream's `$THETA` / `$OMEGA` /
`$SIGMA` blocks are that run's *initial* estimates, not its final ones -- for
example `THETA(16) [KA~dose]` is `-1.17` there against a published final
`-1.02`, and `-1.17` is in fact the lower bound of the published 95% CI. Values
therefore come from the table, structure from the code.

| Model element | Source location | Value |
|---|---|---|
| CL/F reference | Table 2 `CL/F (L/h)` | 3.94 (95% CI 3.69-4.20) |
| Vc/F reference | Table 2 `Vc/F (L)` | 17.8 (16.9-18.9) |
| Q/F reference | Table 2 `Q/F (L/h)` | 1.24 (0.962-1.60) |
| Vp/F reference | Table 2 `Vp/F (L)` | 7.10 (6.04-8.35) |
| Ka at 800 mg | Table 2 `Ka (1/h)` | 0.697 (0.594-0.819) |
| Absorption lag | Table 2 `Lag (h)` | 0.212 (0.192-0.235) |
| F reference | Table 2 `F` | 1 (no uncertainty reported; `fixed()`) |
| WT exponent, CL/F and Q/F | Table 2 `Effect of WT on CL/F` = `Effect of WT on Q/F`; control stream `MU_3 = THETA(3) + CLWT` | 0.301 (0.159-0.443) |
| WT exponent, Vc/F and Vp/F | Table 2 `Effect of WT on Vc/F` = `Effect of WT on Vp/F`; control stream `MU_4 = THETA(4) + VCWT` | 0.536 (0.378-0.694) |
| Strong CYP3A4 inhibitor on CL/F | Table 2 `Effect of CYP3AINH on CL/F` | x0.709 (0.681-0.737) |
| CYP3A4 inducer on CL/F | Table 2 `Effect of CYP3AIND on CL/F` | x2.27 (2.15-2.38) |
| CMV status on CL/F | Table 2 `Effect of CMV on CL/F` | x0.678 (0.626-0.733) |
| Dose on Ka | Table 2 `Effect of dose on Ka` | x(DOSE/800)^-1.02 (-1.17 to -0.875) |
| PPI on Ka | Table 2 `Effect of PPI on Ka` | x0.457 (0.367-0.568) |
| PPI on F | Table 2 `Effect of PPI on F` | x0.905 (0.849-0.964) |
| IIV diagonals (6) | Table 2 `IIV (%)` column | 49.5, 25.3, 145, 110, 72.0, 48.8% |
| IIV block structure | Control stream `3 OMEGA BLOCK(2)` over `ETA(1..6)` | CL~Vc, Q~Vp, Ka~lag |
| IIV off-diagonals | **Not published**; correlations carried from the control stream `$OMEGA` initial estimates | r = 0.673, 0.820, -0.686 |
| Additive residual | Table 2 `sigma^2 add` | 0.000112 -> SD 0.010583 |
| Proportional residual, phase 1 | Table 2 `sigma^2 prop Phase 1` | 0.0705 -> SD 0.265518 |
| Proportional residual, phase 2/3 | Table 2 `sigma^2 prop Phase 2 & 3` | 0.158 -> SD 0.397492 |
| Structural model, `$ERROR`, `F1 = PPIF`, `ALAG1` | Control stream (`ADVAN4 TRANS4`) | -- |

Two structural readings deserve emphasis, because Table 2 alone is ambiguous
about both and only the control stream settles them.

**Two weight exponents, not four.** Table 2 lists four weight rows, but the CL/F
and Q/F rows carry *identical* estimates, %RSE and 95% CI, as do the Vc/F and
Vp/F rows. That is not a coincidence of rounding: the control stream computes
`CLWT` and `VCWT` once and reuses them (`MU_3 = THETA(3) + CLWT`,
`MU_4 = THETA(4) + VCWT`), while `THETA(9)` and `THETA(10)` are present but
commented `Not used`. The model file therefore has two exponent parameters.

**The `IIV (%)` back-transform.** Table 2 reports IIV as a percentage without
stating the convention. This file uses the rule that the same analysis group
documented explicitly for the same drug in the companion Sun 2023 paper
(its Table S2 footnote c): `CV = sqrt(omega^2)` when `omega^2 <= 0.15` and
`CV = sqrt(exp(omega^2) - 1)` otherwise. All six Table 2 entries are
self-consistent under that rule, and only Vc/F falls in the first branch, where
the two branches differ by less than 2% in SD. The two best-determined
parameters corroborate it: back-transforming CL/F's 49.5% gives `omega^2` =
0.2192 against the control stream's initial 0.214, and the lag time's 48.8%
gives 0.2136 against an initial 0.203.

# Virtual cohort

Two arms of 200 participants each, matching the two health-status strata that
Sun 2025 Table 3 reports separately. Both are simulated at 400 mg twice daily
**without concomitant medication**, which is exactly the condition Table 3 and
Figure 2 state for their exposure summaries.

Weights are drawn from log-normal distributions matched to the per-stratum
Table 1 summaries (phase 1 mean 78.1, SD 15.1; phase 2/3 mean 73.1, SD 17.8) and
truncated to the reported overall range 36.1-141 kg.

```{r cohort}
set.seed(20250704)
rxode2::rxSetSeed(20250704)

n_arm <- 200L

draw_wt <- function(n, mean_kg, sd_kg) {
  sdlog <- sqrt(log1p((sd_kg / mean_kg)^2))
  meanlog <- log(mean_kg) - sdlog^2 / 2
  pmin(pmax(stats::rlnorm(n, meanlog, sdlog), 36.1), 141)
}

cohort <- dplyr::bind_rows(
  tibble::tibble(
    id        = seq_len(n_arm),
    treatment = "Healthy volunteers",
    WT        = draw_wt(n_arm, 78.1, 15.1),
    DIS_CMV   = 0,
    STUDY_MARIBAVIR_PHASE1 = 1
  ),
  tibble::tibble(
    id        = n_arm + seq_len(n_arm),
    treatment = "Transplant recipients with CMV",
    WT        = draw_wt(n_arm, 73.1, 17.8),
    DIS_CMV   = 1,
    STUDY_MARIBAVIR_PHASE1 = 0
  )
) |>
  dplyr::mutate(
    DOSE                     = 400,
    CONMED_PPI               = 0,
    CONMED_CYP3A4_INH_STRONG = 0,
    CONMED_CYP3A4_IND        = 0
  )

cohort |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    n = dplyr::n(),
    `Weight mean (kg)`   = round(mean(WT), 1),
    `Weight median (kg)` = round(stats::median(WT), 1),
    `Weight SD (kg)`     = round(stats::sd(WT), 1),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Simulated cohort weight distributions (Sun 2025 Table 1 targets: phase 1 mean 78.1 SD 15.1, median 75.6; phase 2/3 mean 73.1 SD 17.8, median 71.3).")
```

# Simulation

400 mg twice daily for 10 days (20 doses, `ii = 12`, `addl = 19`), which is far
beyond the 2 days to steady state the paper reports for this regimen. The
observation grid covers the **last** dosing interval, 228-240 h, so the
non-compartmental interval is a true steady-state interval and its lower bound
carries a real pre-dose trough record rather than a synthetic time-zero anchor.

```{r simulate}
events <- rxode2::et(amt = 400, ii = 12, addl = 19, cmt = "depot") |>
  rxode2::et(seq(228, 240, by = 0.1), cmt = "central") |>
  rxode2::et(id = cohort$id)

ev_df <- as.data.frame(events) |>
  dplyr::left_join(cohort, by = "id")

sim <- rxode2::rxSolve(mod, ev_df, returnType = "data.frame") |>
  dplyr::left_join(dplyr::select(cohort, id, treatment), by = "id")

str(dplyr::select(sim, id, time, Cc), max.level = 1)
```

## Steady-state concentration-time profiles

```{r fig-profiles, fig.cap = "Simulated steady-state maribavir profiles over the last 12 h dosing interval of 400 mg twice daily, by health status. Compare with the observed-versus-predicted spread in Sun 2025 Figure 1 (prediction-corrected VPC)."}
sim |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    med = stats::median(Cc),
    lo  = stats::quantile(Cc, 0.05),
    hi  = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time - 228, med, colour = treatment, fill = treatment)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time after dose at steady state (h)",
    y = "Maribavir concentration (ug/mL)",
    colour = NULL, fill = NULL
  ) +
  ggplot2::theme(legend.position = "bottom")
```

# Structural checks (typical value, no random effects)

These are deterministic: `zeroRe()` removes both the between-subject and the
residual variability, so the numbers below depend only on the `ini()` values and
the `model()` algebra. They are the sharpest available test of the covariate
encoding, and they are compared against numbers Sun 2025 reports directly.

```{r structural}
mod_typ <- rxode2::zeroRe(mod)

typ_exposure <- function(ppi = 0, cmv = 1, wt = 70, dose = 400) {
  ev <- rxode2::et(amt = dose, ii = 12, addl = 19, cmt = "depot") |>
    rxode2::et(seq(228, 240, by = 0.01), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt
  d$DOSE <- dose
  d$DIS_CMV <- cmv
  d$CONMED_PPI <- ppi
  d$CONMED_CYP3A4_INH_STRONG <- 0
  d$CONMED_CYP3A4_IND <- 0
  d$STUDY_MARIBAVIR_PHASE1 <- 0
  s <- rxode2::rxSolve(mod_typ, d, returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  c(
    auc  = sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2),
    cmax = max(s$Cc),
    cmin = min(s$Cc)
  )
}

cmv_noppi <- typ_exposure(ppi = 0, cmv = 1)
cmv_ppi   <- typ_exposure(ppi = 1, cmv = 1)
healthy   <- typ_exposure(ppi = 0, cmv = 0)
```

## Steady-state mass balance

At steady state the amount cleared over one dosing interval must equal the
delivered dose, so `AUCss x CL/F = F x Dose` exactly. This is the cheapest gate
that would catch a mis-scaled volume, a wrong concentration unit, or an
`rxode2` `cl`/`vc` pair silently replacing the explicit ODEs with an analytical
solution.

```{r mass-balance}
cl_cmv_70kg <- 3.94 * 0.678
mass_balance <- unname(cmv_noppi["auc"] * cl_cmv_70kg / 400)
mass_balance

stopifnot(abs(mass_balance - 1) < 1e-3)
```

## Covariate effect ratios against the published values

Sun 2025 quantifies three covariate contrasts numerically. Each is reproduced
below from the model algebra alone. The PPI contrast is the strongest test in
this vignette, because it is the *net* result of two separate effects pulling in
different directions -- bioavailability down 9.5% and absorption rate down 54.3%
-- and the paper reports its AUC, Cmax and Cmin consequences separately.

```{r covariate-ratios}
ratios <- tibble::tibble(
  Contrast = c(
    "PPI vs no PPI, AUCss", "PPI vs no PPI, Cmax,ss", "PPI vs no PPI, Cmin,ss",
    "CMV vs healthy, AUCss", "CMV vs healthy, Cmax,ss", "CMV vs healthy, Cmin,ss"
  ),
  Simulated = c(
    cmv_ppi["auc"] / cmv_noppi["auc"],
    cmv_ppi["cmax"] / cmv_noppi["cmax"],
    cmv_ppi["cmin"] / cmv_noppi["cmin"],
    cmv_noppi["auc"] / healthy["auc"],
    cmv_noppi["cmax"] / healthy["cmax"],
    cmv_noppi["cmin"] / healthy["cmin"]
  ),
  Published = c(0.905, 0.775, 1.03, 1.46, 1.20, 2.11),
  Source = c(
    "Figure 2b / Section 3.1.3 (-9.5%)", "Figure 2b / Section 3.1.3 (-22.5%)", "Figure 2b",
    "Section 3.1.3", "Section 3.1.3", "Section 3.1.3"
  )
) |>
  dplyr::mutate(
    Simulated = round(Simulated, 3),
    `% diff`  = round(100 * (Simulated - Published) / Published, 1)
  )

knitr::kable(ratios, caption = "Typical-value covariate contrasts against the values Sun 2025 reports.")

# The PPI AUCss ratio is a pure function of the F effect and must reproduce
# 0.905 essentially exactly; the others involve absorption and distribution and
# are compared against population geometric-mean ratios, so they are given
# proportionally more headroom.
stopifnot(
  abs(ratios$`% diff`[1]) < 0.5,
  max(abs(ratios$`% diff`)) < 5
)
```

# PKNCA validation

Non-compartmental analysis over the steady-state interval 228-240 h, grouped by
treatment arm so each arm can be compared against its own Table 3 row.

```{r pknca}
conc_data <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)

dose_data <- cohort |>
  dplyr::transmute(id, treatment, time = 228, dose = 400)

o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | id / treatment)
# PKNCAdose does not accept slash (nested) grouping, only PKNCAconc does.
o_dose <- PKNCA::PKNCAdose(dose_data, dose ~ time | id + treatment)

intervals <- data.frame(
  start = 228, end = 240,
  auclast = TRUE, cmax = TRUE, cmin = TRUE, half.life = TRUE
)

o_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))

nca_res <- as.data.frame(o_nca) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin", "half.life"))

stopifnot(nrow(nca_res) > 0L, !anyNA(nca_res$PPORRES))
```

```{r pknca-summary}
nca_gm <- nca_res |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    gm  = exp(mean(log(PPORRES))),
    cv  = 100 * sqrt(exp(stats::var(log(PPORRES))) - 1),
    .groups = "drop"
  )

nca_gm |>
  dplyr::mutate(
    Parameter = dplyr::recode(
      PPTESTCD,
      auclast     = "AUCss (ug*h/mL)",
      cmax        = "Cmax,ss (ug/mL)",
      cmin        = "Cmin,ss (ug/mL)",
      half.life   = "t1/2 (h)"
    ),
    `Geometric mean` = round(gm, 3),
    `CV%`            = round(cv, 1)
  ) |>
  dplyr::select(Arm = treatment, Parameter, `Geometric mean`, `CV%`) |>
  knitr::kable(caption = "Simulated steady-state exposure at 400 mg twice daily without concomitant medication.")
```

# Comparison against published NCA

Sun 2025 Table 3 reports geometric mean (CV%) steady-state exposure for the same
regimen and the same two strata, derived from post-hoc Bayesian individual
parameter estimates.

```{r nca-comparison}
reference <- tibble::tribble(
  ~treatment,                       ~PPTESTCD,   ~PPORRES,
  "Healthy volunteers",             "auclast",     97.0,
  "Healthy volunteers",             "cmax",        16.7,
  "Healthy volunteers",             "cmin",         2.57,
  "Healthy volunteers",             "half.life",    4.84,
  "Transplant recipients with CMV", "auclast",    142.0,
  "Transplant recipients with CMV", "cmax",        20.1,
  "Transplant recipients with CMV", "cmin",         5.43,
  "Transplant recipients with CMV", "half.life",    6.68
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by        = "treatment",
  params    = c(
    "AUCss (ug*h/mL)" = "auclast",
    "Cmax,ss (ug/mL)" = "cmax",
    "Cmin,ss (ug/mL)" = "cmin",
    "t1/2 (h)"        = "half.life"
  ),
  label_first_column = "NCA parameter"
) |>
  knitr::kable(caption = "Simulated versus Sun 2025 Table 3 steady-state exposure. Values marked with a star differ by more than 20%.")
```

AUCss and Cmax,ss reproduce the published values closely in both arms (-1.3% to
+10.8%). Note that the two tables in this section aggregate differently:
`ncaComparisonTable()` compares **medians**, whereas the percent-difference
table below compares **geometric means**, which is what Sun 2025 Table 3
reports. The geometric-mean comparison is the like-for-like one.

Two comparisons deserve comment, including the two rows
`ncaComparisonTable()` stars as exceeding 20%.

* **Cmin,ss is the starred metric in both arms** (+21.1% and +29.9% on medians;
  +12.3% and +18.4% on geometric means). This is expected and is not evidence of
  a transcription error, for a specific reason: Table 3's values summarise
  *post-hoc Bayesian* individual estimates, and Table 2 reports shrinkage of
  46.8% for Ka and 53.6% for the absorption lag time -- the two parameters that
  determine the trough. Post-hoc estimates of a heavily shrunk parameter are
  pulled toward the typical value, which raises the trough relative to a
  prospective draw from the full prior such as this one. Consistent with that
  reading, the *typical-value* trough ratios in the Structural checks section
  above -- which involve no sampling at all -- match the published values to
  within 2.5%, and AUCss, whose driving parameter CL/F has only 6.9% shrinkage,
  matches to within a few percent. Cmin,ss is also the most variable metric in
  the source (CV 76-92%).
* **t1/2** is not comparable on equal terms. Sun 2025 derived it from each
  individual's full model-predicted profile, whereas PKNCA estimates lambda-z
  from the terminal points of a single 12 h steady-state interval of a
  two-compartment drug. A 12 h window cannot resolve a terminal phase that the
  model places beyond it, so the NCA half-life is expected to be the *effective*
  rather than the true terminal value. It is reported for completeness, not used
  as a gate.

The assertions below follow the repository convention of gating on the centre
and on robust quantiles rather than on the extreme of a random cohort, because
the per-subject extreme of a simulated cohort is not reproducible across rxode2
versions.

```{r nca-assertions}
check <- nca_gm |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin")) |>
  dplyr::inner_join(reference, by = c("treatment", "PPTESTCD")) |>
  dplyr::mutate(pct_diff = 100 * (gm - PPORRES) / PPORRES)

check |>
  dplyr::select(treatment, PPTESTCD, simulated_gm = gm, published = PPORRES, pct_diff) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Percent difference between simulated and published geometric mean steady-state exposure.")

stopifnot(
  # AUCss and Cmax,ss are driven by clearance and volume, which are the two
  # best-determined parameters in Table 2; a mis-transcribed value moves these
  # by tens of percent.
  max(abs(check$pct_diff[check$PPTESTCD %in% c("auclast", "cmax")])) < 15,
  # Cmin,ss is absorption- and lag-driven with very high published variability.
  max(abs(check$pct_diff[check$PPTESTCD == "cmin"])) < 30
)
```

# Dose-dependent absorption

The dose effect on Ka is unusual enough to be worth showing on its own: the
exponent is negative, so absorption slows as the dose increases. At 1200 mg the
typical Ka is `(1200/800)^-1.02` = `r round((1200 / 800)^-1.02, 3)` times its
800 mg value, while at 400 mg it is `r round((400 / 800)^-1.02, 3)` times.
Because the model is written on apparent (`/F`) parameters and clearance is
dose-independent, AUCss stays exactly dose-proportional -- consistent with the
paper's statement that maribavir PK is dose-proportional after single doses of
50-1600 mg -- while the peak-to-trough shape changes.

```{r fig-dose, fig.cap = "Typical-value steady-state profiles for a 70 kg transplant recipient with CMV at three dose levels, showing the flattening of the profile at higher doses that the negative dose-on-Ka exponent produces."}
dose_profiles <- lapply(c(400, 800, 1200), function(dd) {
  ev <- rxode2::et(amt = dd, ii = 12, addl = 19, cmt = "depot") |>
    rxode2::et(seq(228, 240, by = 0.05), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- 70
  d$DOSE <- dd
  d$DIS_CMV <- 1
  d$CONMED_PPI <- 0
  d$CONMED_CYP3A4_INH_STRONG <- 0
  d$CONMED_CYP3A4_IND <- 0
  d$STUDY_MARIBAVIR_PHASE1 <- 0
  s <- rxode2::rxSolve(mod_typ, d, returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  data.frame(time = s$time - 228, Cc = s$Cc, Dose = paste(dd, "mg BID"))
}) |>
  dplyr::bind_rows()

ggplot2::ggplot(dose_profiles, ggplot2::aes(time, Cc, colour = Dose)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    x = "Time after dose at steady state (h)",
    y = "Maribavir concentration (ug/mL)",
    colour = NULL
  ) +
  ggplot2::theme(legend.position = "bottom")
```

# Assumptions and deviations

## Structural readings taken from the control stream

* **Two weight exponents rather than four.** Table 2 presents four weight-effect
  rows, but the control stream reuses `CLWT` for Q/F and `VCWT` for Vp/F, with
  `THETA(9)` and `THETA(10)` present but commented `Not used`. Encoding four
  independent exponents would have been a silent structural error that the
  identical Table 2 estimates would never have exposed.
* **Values from the table, structure from the code.** The control stream's
  `$THETA` / `$OMEGA` / `$SIGMA` blocks are initial estimates, not final ones.
  They are close to the published values but not equal to them (`THETA(16)` is
  `-1.17` initially versus a final `-1.02`), so none of them was used as a value.

## Not published, and therefore approximated

* **The three IIV correlations.** Table 2 reports only the six IIV magnitudes; it
  has no correlation column and no off-diagonal rows. The control stream shows
  the *structure* -- three `$OMEGA BLOCK(2)` blocks pairing CL with Vc, Q with
  Vp, and Ka with the lag time -- and the header line
  `;; 2. Description: 3 OMEGA BLOCKs` confirms that structure is final. This
  model file keeps that block structure, puts the published Table 2 variances on
  the diagonals, and carries the correlation **coefficients** (0.673, 0.820 and
  -0.686) from the control stream's own `$OMEGA` initial estimates, rescaled onto
  the published variances. These three covariances are the least well-sourced
  numbers in the file. They were carried rather than invented, and rather than
  set to zero, because the block structure is a published feature of the final
  model and zeroing the off-diagonals would misrepresent it; but a user who
  needs exact published uncertainty should treat the correlations as
  provisional. The magnitudes and the diagonals are unaffected.
* **The `IIV (%)` back-transform convention** is not stated in this paper. See
  the Source trace section for the rule used, its provenance in the same
  analysis group's companion Sun 2023 paper, and the internal-consistency check
  that supports it. Only Vc/F is sensitive to the choice, and only by <2% in SD.

## Exposure-response models: reported but not extracted

Sun 2025 is a two-part paper. The population PK model above is fully extracted.
The exposure-response (E-R) analyses are **not** extracted here, and the reason
differs by endpoint:

* **Efficacy (Table S3).** Two logistic regressions -- confirmed CMV clearance at
  week 8, and clearance with no tissue-invasive disease maintained to week 16 --
  are reported in full, with intercepts, an AUCss slope and every risk-factor
  coefficient. They are extractable as written. They are not extracted in this
  pass because doing so requires roughly eight new canonical covariate columns
  that do not yet exist in `inst/references/covariate-columns.md` (baseline
  CD8+CD69+pp65 stimulation category, treatment-emergent maribavir-resistance
  mutation, high baseline CMV DNA, post-HCT T-cell infusion, an Asia Pacific
  enrolling-region indicator, and days from HCT onset to treatment), and a new
  canonical name needs operator ratification rather than a librarian's guess.
  The paper's headline result is that **neither** relationship is statistically
  significant: the AUCss coefficients are 0.00920 (SE 0.0137, p = 0.503) and
  -0.0131 (SE 0.0113, p = 0.247) per 10 ug*h/mL, i.e. essentially flat.
* **Safety versus AUCss (Figure 4).** Nausea and vomiting show statistically
  significant positive relationships (p = 0.0394 and 0.00296), but the figure
  prints **no coefficients** -- only the p-value, the fitted curve and the
  observed quartile proportions. There is nothing to transcribe, and recovering
  an intercept and slope by digitising a raster figure would manufacture
  precision the paper never published.
* **Safety versus AUCday (Figure S3, Table S4).** Fourteen logistic regressions
  across dysgeusia, nausea, vomiting, diarrhea, neutropenia, raised
  immunosuppressant concentrations, invasive infection, acute GvHD, renal
  disorder, anemia, pyrexia, headache, thrombocytopenia and serious adverse
  events. These *are* fully parameterised in the supplement and are extractable;
  they carry the same new-canonical requirement as the efficacy models, plus
  several more (T-cell depletion modality, prior HCT, prior CMV prophylaxis,
  total baseline white blood cell category, transplanted nucleated cell number).

One discrepancy in the source is worth recording for whoever extracts these
later. For **anemia**, Table S4 assigns odds ratio 3.12 to "Europe vs. North
America" and 4.62 to "Asia Pacific vs. North America", while Figure S3 panel j
assigns 3.12 to Asia Pacific and 4.62 to Europe. Figure S3 is internally
consistent and Table S4 is not: the figure's estimates exponentiate to its own
odds ratios (`exp(1.14) = 3.13` for Asia Pacific, `exp(1.53) = 4.62` for
Europe), whereas Table S4 reports odds ratios only and cannot be checked against
itself. The figure should be preferred.

## Simulation assumptions

* **Weight distributions** are log-normal draws matched to the Table 1 per-stratum
  mean and SD and truncated to the reported 36.1-141 kg range. Sun 2025 does not
  publish the empirical weight distribution, and Table 1 reports "NR" for the
  AURORA weights specifically.
* **Cohort size** is 200 per arm, the repository cap, against published strata of
  157 and 724 individuals. The comparison is therefore of geometric means and
  CV%, not of individual predictions.
* **Post-hoc versus prospective simulation.** Table 3's values are summaries of
  *post-hoc Bayesian* individual estimates, which are shrunk toward the typical
  value. This simulation draws from the full prior. Shrinkage is low for CL/F
  (6.9%) and moderate for Vc/F (30.8%) but high for Q/F, Vp/F and the lag time
  (54-56%), so the simulated CV% for trough-sensitive metrics is expected to
  exceed the published CV%.
* **No concomitant medication** in any simulated arm, matching the stated
  conditions for Table 3 and Figure 2.
```
