---
title: "Paclitaxel-induced peripheral neuropathy (Sun 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paclitaxel-induced peripheral neuropathy (Sun 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
PKNCA::PKNCA.options(progress = FALSE)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sun_2025_paclitaxel"))$reference`
* Description: `r rxode2::rxode(readModelDb("Sun_2025_paclitaxel"))$description`
* Article: <https://doi.org/10.1111/cts.70404> (open access; PMC12616876)
* Supporting information: Figure S1 (CIPN8 residuals) and Figure S2 (zoomed
  predicted-vs-observed PK). Neither carries parameter values, so no value in
  this model file depends on the supplement.

# Population

The model was built from UMCC2014.002 (NCT02338115), an observational study at
the University of Michigan Rogel Cancer Center. Sixty-five adult female patients
with stage I-III or oligometastatic breast cancer scheduled for weekly 1-h
infusions of paclitaxel 80 mg/m2 for 12 weeks were enrolled; five were excluded
for withdrawal or protocol violation, leaving 60 patients with paired PK and
neuropathy data (Results 3.1). Mean age was 52.3 years (28 to 71) and 93.3%
self-identified as White (3.3% Black, 3.3% Asian; Table 1). Each patient's dose
was set from body surface area, and the **first** dose was infused over 90 min
rather than 60 min so that infusion reactions could be monitored (Methods 2.1).

Only two plasma samples were drawn per patient, both during the first infusion:
one in the final 10 min before the end of infusion, and one 16-26 h after the
infusion started. Fifty-seven patients had an end-of-infusion sample and 59 had
a post-infusion sample. Mean observed Cmax was 2364.16 ng/mL (907 to 4340) and
the mean 18-26 h concentration was 22.02 ng/mL (10 to 51.80). Patients received
a mean of 11 of 12 planned doses (3 to 12), with 68% completing all 12, for a
mean cumulative dose of 1617.20 mg (456 to 2412) (Table 1).

Neuropathy was measured weekly with the EORTC QLQ-CIPN20. The 8-item sensory
subscale excluding the ototoxicity item (CIPN8) was summed and offset by -8, so
the score runs 0 (no symptoms) to 24 (Methods 2.1). Mean baseline CIPN8 was 0.3
(0 to 3) with 83% of patients scoring 0; the mean within-patient change over
treatment was 6.03 (0 to 21).

The same information is available programmatically via
`readModelDb("Sun_2025_paclitaxel")()$population`.

# Model structure

The PK model is a two-compartment model with linear elimination fit to the
sparse first-infusion data (Results 3.2, Figure 2). The PD model is a hybrid of
a threshold-gated effect compartment and a turnover pool, as printed in
Equations 1-5:

$$
f(x) = \begin{cases} c_{act} = 0, & c < x \\ c_{act} = c - x, & c \ge x \end{cases}
\qquad\qquad (1)
$$

$$
C_{e0} = 0 \qquad\qquad (2) \qquad\qquad\qquad \frac{dc_e}{dt} = c_{act} \times k_{e0} \qquad\qquad (3)
$$

$$
k_{in} = R_0 \times k_{out} \qquad\qquad (4) \qquad\qquad \frac{dR}{dt} = k_{in} \times \left(1 + k \times c_e\right) - k_{out} \times R \qquad\qquad (5)
$$

Three features are worth flagging because they are unusual and are reproduced
verbatim rather than normalised:

1. **Equation 1 is a hard switch, not a saturable function.** Below the
   threshold concentration `x` the drug contributes *exactly* nothing to the
   effect compartment. This is why the model file uses the canonical `cthres`
   (a gating threshold concentration) rather than `ec50`.
2. **Equation 3 has no loss term**, so the effect compartment is a pure
   integrator that only ever accumulates. The Discussion confirms this is
   intended: "using the effect compartment alone is not optimal since the CIPN
   will always increase". Symptom resolution is supplied downstream by the
   turnover pool's `kout`, not by the effect compartment.
3. **The drug effect in Equation 5 is linear** in the effect-compartment value.
   The paper explored an Emax form and rejected it for poor performance
   (Results 3.2).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Sun_2025_paclitaxel.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | `log(26.04)` L/h | Table 2, `Cl_pop` (SE 1.17, RSE 4.51%) |
| `lvc` | `log(34.08)` L | Table 2, `V_pop` (SE 2.44, RSE 7.15%) |
| `lq` | `log(4.2)` L/h | Table 2, `Q_pop` (SE 0.17, RSE 4.01%) |
| `lvp` | `log(67.67)` L | Table 2, `V2_pop` (SE 6.2, RSE 9.17%) |
| `lcthres` | `log(1735.75)` ng/mL | Table 2, `x_pop` (SE 1.84, RSE 0.106%); restated in Results 3.2 and Discussion |
| `lke0` | `log(0.00022)` mL/ng | Table 2, `ke0_pop` (SE 0.000063, RSE 28.4%) |
| `lkout` | `log(0.0027)` 1/h | Table 2, `kout_pop` (SE 0.0016, RSE 58.9%); units not printed, inferred from the L/hour PK panel |
| `lslope` | `log(21.65)` | Table 2, `k_pop` (SE 9.52, RSE 44.0%); the `k` of Equation 5 |
| `lrbase` | `fixed(log(0.3))` | Table 2, `R0_pop` with SE and RSE both NA, i.e. not estimated; equals the Table 1 mean baseline CIPN8 |
| `etalcl` | `0.25^2` | Table 2, `Omega_Cl` = 0.25 (RSE 15.6%) |
| `etalvc` | `0.51^2` | Table 2, `Omega_V1` = 0.51 (RSE 9.54%) |
| `etalq` | `0.2^2` | Table 2, `Omega_Q` = 0.2 (RSE 20.0%) |
| `etalvp` | `0.23^2` | Table 2, `Omega_V2` = 0.23 (RSE 36.9%) |
| `etalkout` | `2.45^2` | Table 2, `Omega_kout` = 2.45 (RSE 21.5%, shrinkage -75%) |
| `etalslope` | `1.91^2` | Table 2, `Omega_k` = 1.91 (RSE 14.7%, shrinkage 52%) |
| `etalke0` | `0.85^2` | Table 2, `Omega_ke0` = 0.85 (RSE 40.2%, shrinkage 90%) |
| `addSd` | `9.11` ng/mL | Table 2, PK error model `a` (SE 1.27, RSE 13.9%); "constant error model" (Results 3.2) |
| `addSd_cipn8` | `1.62` | Table 2, PD error model `a1` (SE 0.059, RSE 3.66%) |
| `propSd_cipn8` | `0.056` | Table 2, PD error model `b1` (SE 0.018, RSE 32.1%) |
| `powExp_cipn8` | `fixed(0.74)` | Table 2 footnote "CIPN8 = R + (a1 + b1 x R^0.74) x e"; not among the estimated rows |
| Threshold gate `cact <- max(Cc - cthres, 0)` | n/a | Equation 1 |
| `effect(0) <- 0` | n/a | Equation 2 |
| `d/dt(effect) <- cact * ke0` | n/a | Equation 3 |
| `kin <- rbase * kout` | n/a | Equation 4 |
| `d/dt(cipn8) <- kin * (1 + slope * effect) - kout * cipn8` | n/a | Equation 5 |
| Two-compartment PK ODEs | n/a | Results 3.2 ("two-compartment model with a constant error model"); Figure 2 schematic |
| `Cc <- central / vc * 1000` | n/a | Unit conversion only: `central` is mg and `vc` is L, so `central/vc` is mg/L = ug/mL; Table 1, Table 2 and Figure 1 all report ng/mL |
| IIV log-normal, omegas are SDs | n/a | Table 2 footnote |
| CIPN8 scale 0-24 | n/a | Methods 2.1 |
| Grade 2+ / grade 3+ proxies CIPN8 > 8 / > 12 | n/a | Methods 2.4 |

```{r load-model}
mod <- readModelDb("Sun_2025_paclitaxel")
ui <- rxode2::rxode(mod)
mod_typical <- rxode2::zeroRe(mod)

# Body surface area implied by Table 1: a mean cumulative dose of 1617.20 mg
# over a mean of 11 doses at 80 mg/m2. This is derived from the paper rather
# than assumed, and is used as the centre of the virtual cohort's BSA.
bsa_typical <- 1617.20 / 11 / 80
dose_typical <- 80 * bsa_typical
week_h <- 24 * 7
c(bsa_typical = bsa_typical, dose_typical_mg = dose_typical)
```

The model carries two endpoints (`Cc` and `cipn8`) over four ODE states, so
every event table below marks observation rows with `dvid = 1L` and every
`rxSolve()` call passes `useLinCmt = FALSE`; the automatic ODE-to-`linCmt`
conversion corrupts the `dvid`-to-`cmt` mapping for multi-output models of this
shape.

```{r helpers}
# Build a full 12-week event table for one arm. `mg_m2` is a length-12 vector of
# the mg/m2 dose at each weekly occasion, so the standard and enhanced regimens
# differ only in that vector. Infusion durations follow Methods 2.1: 90 min for
# the first dose, 60 min thereafter. `id_offset` keeps IDs disjoint across arms.
make_weekly_arm <- function(bsa, mg_m2, obs_times, id_offset = 0L,
                            dose_times = (0:11) * week_h,
                            durs = c(1.5, rep(1, 11))) {
  stopifnot(length(mg_m2) == length(dose_times), length(durs) == length(dose_times))
  dosing <- tidyr::expand_grid(subj = seq_along(bsa), k = seq_along(dose_times)) |>
    dplyr::mutate(
      id = id_offset + .data$subj,
      time = dose_times[.data$k],
      amt = mg_m2[.data$k] * bsa[.data$subj],
      evid = 1L, cmt = "central", dur = durs[.data$k], dvid = NA_integer_
    )
  obs <- tidyr::expand_grid(subj = seq_along(bsa), time = obs_times) |>
    dplyr::mutate(
      id = id_offset + .data$subj, amt = NA_real_, evid = 0L,
      cmt = NA_character_, dur = NA_real_, dvid = 1L
    )
  dplyr::bind_rows(dosing, obs) |>
    dplyr::select("id", "time", "amt", "evid", "cmt", "dur", "dvid") |>
    dplyr::arrange(.data$id, .data$time, dplyr::desc(.data$evid))
}

# Peak capped CIPN8 per subject. Results 3.2: "The simulated CIPN score was
# capped by setting an upper bound of 24 such that any predicted value exceeding
# 24 was assigned a value of 24." The cap cannot change whether a subject
# crosses 8 or 12, so the reported incidences are unaffected by it.
peak_cipn8 <- function(sim) {
  sim |>
    dplyr::mutate(cipn8_capped = pmin(.data$cipn8, 24)) |>
    dplyr::group_by(.data$id) |>
    dplyr::summarise(peak = max(.data$cipn8_capped), .groups = "drop") |>
    dplyr::arrange(.data$id)
}
```

# Typical-value PK: reproducing the observed Cmax

The first infusion runs 90 min, and the end-of-infusion sample was drawn in its
final 10 min. A typical-value solve at the cohort's implied BSA should therefore
land on the reported mean Cmax.

```{r typical-pk}
grid_pk <- sort(unique(c(seq(0, 30, by = 0.02), 1.5)))
ev_typ <- make_weekly_arm(
  bsa = bsa_typical, mg_m2 = 80, obs_times = grid_pk,
  dose_times = 0, durs = 1.5
)
sim_typ <- rxode2::rxSolve(mod_typical, ev_typ, returnType = "data.frame",
                           useLinCmt = FALSE)

cmax_typ <- max(sim_typ$Cc)
tmax_typ <- sim_typ$time[which.max(sim_typ$Cc)]
c22_typ <- stats::approx(sim_typ$time, sim_typ$Cc, 22)$y

tibble::tibble(
  Quantity = c("Cmax (ng/mL)", "Tmax (h)", "Concentration at 22 h (ng/mL)"),
  Simulated = c(cmax_typ, tmax_typ, c22_typ),
  Published = c(2364.16, 1.5, 22.02),
  `Published source` = c("Table 1 cohort mean (907-4340)",
                         "End of the 90-min first infusion (Methods 2.1)",
                         "Table 1 cohort mean of the 18-26 h sample (10-51.80)")
) |>
  dplyr::mutate(`% difference` = 100 * (.data$Simulated - .data$Published) / .data$Published) |>
  knitr::kable(digits = c(0, 2, 2, 0, 1),
               caption = "Typical-value PK against the values reported in Table 1.")
```

```{r typical-pk-gate}
# Cmax is a deterministic typical-value quantity compared against a fixed
# published mean, so a tight bound is appropriate here (no cohort draw is
# involved). Achieved: ~1.2%.
stopifnot(abs(cmax_typ - 2364.16) / 2364.16 < 0.05)
# Tmax must sit exactly at the end of the 90-min infusion; 1.5 is in the grid.
stopifnot(abs(tmax_typ - 1.5) < 1e-8)
```

The typical-value Cmax reproduces the published cohort mean to about 1%, which
independently confirms the two-compartment parameters *and* the `mg` to `ng/mL`
scaling in `Cc <- central / vc * 1000`. The 22-h concentration sits below the
published mean; see Errata.

# The threshold gate and the effect compartment

The effect compartment is the one PD quantity for which the paper reports a
numeric value: Table 2's footnote states "The effect compartment concentration
across individual is 1.56 +/- 1.18 ng/mL". A typical-value solve over the full
12 weekly doses reproduces it, which confirms that `ke0` is applied verbatim in
the printed mL/ng units, that the threshold is compared against a ng/mL
concentration, and that the effect compartment is a pure integrator.

```{r effect-compartment}
obs_daily <- seq(0, 12 * week_h, by = 24)
ev_std_typ <- make_weekly_arm(bsa_typical, rep(80, 12), obs_daily)
sim_std_typ <- rxode2::rxSolve(mod_typical, ev_std_typ, returnType = "data.frame",
                               useLinCmt = FALSE)
effect_end <- dplyr::last(sim_std_typ$effect)

tibble::tibble(
  Quantity = "Effect-compartment value after 12 weekly doses",
  Simulated = effect_end,
  Published = 1.56,
  `Published source` = "Table 2 footnote (1.56 +/- 1.18 across individuals)"
) |>
  dplyr::mutate(`% difference` = 100 * (.data$Simulated - .data$Published) / .data$Published) |>
  knitr::kable(digits = c(0, 3, 2, 0, 1),
               caption = "Effect-compartment magnitude against the value reported in Table 2.")

stopifnot(abs(effect_end - 1.56) / 1.56 < 0.10)
```

Because the gate is a hard switch, the model predicts that a **slow enough
infusion produces no neuropathy at all**: if the peak concentration never
reaches 1735.75 ng/mL, `cact` is zero throughout and the effect compartment
never leaves zero. The Discussion makes exactly this claim ("Incorporating this
threshold in the effect compartment also ensures the model behaves as expected,
with longer infusion durations leading to lower CIPN risks (data not shown)"),
and it is the mechanism the authors intend to exploit for infusion-rate-based
personalised dosing.

```{r infusion-duration}
# Cmax must be read off a grid that actually resolves the end of each infusion.
# The daily grid used elsewhere never lands on an infusion peak, so this scan
# adds points every 0.05 h for the first 4 h after each dose. The effect
# compartment and CIPN8 are insensitive to grid resolution (the infusion start
# and end are events, so the solver restarts at each threshold crossing), but
# Cmax is not.
obs_peaks <- sort(unique(c(
  obs_daily,
  as.vector(outer((0:11) * week_h, seq(0, 4, by = 0.05), "+"))
)))

duration_scan <- vapply(
  c(0.5, 1, 1.5, 2, 3),
  function(d) {
    ev <- make_weekly_arm(bsa_typical, rep(80, 12), obs_peaks,
                          durs = rep(d, 12))
    s <- rxode2::rxSolve(mod_typical, ev, returnType = "data.frame",
                         useLinCmt = FALSE)
    c(cmax = max(s$Cc), effect = dplyr::last(s$effect),
      peak_cipn8 = max(pmin(s$cipn8, 24)))
  },
  numeric(3)
)

scan_tbl <- tibble::tibble(
  `Infusion duration (h)` = c(0.5, 1, 1.5, 2, 3),
  `Cmax (ng/mL)` = duration_scan["cmax", ],
  `Effect compartment at week 12` = duration_scan["effect", ],
  `Peak CIPN8` = duration_scan["peak_cipn8", ]
)
knitr::kable(scan_tbl, digits = c(1, 0, 3, 2),
             caption = paste("Typical-value effect of infusion duration at a",
                             "fixed 80 mg/m2 weekly dose. Slowing the infusion",
                             "lowers Cmax and therefore the gated exposure."))

# Monotonicity in infusion duration is a structural claim of the paper
# (Discussion), and it is deterministic at typical values, so all three
# sequences are asserted to be STRICTLY decreasing rather than merely
# non-increasing.
stopifnot(
  all(diff(duration_scan["cmax", ]) < 0),
  all(diff(duration_scan["effect", ]) < 0),
  all(diff(duration_scan["peak_cipn8", ]) < 0)
)
# A 3-h infusion of 80 mg/m2 never crosses the threshold, so the gate
# contributes exactly zero and CIPN8 never leaves its baseline. The effect
# compartment is asserted as an exact zero -- `cact` is identically zero, so
# `d/dt(effect)` is too and the state never departs its zero initial condition.
# CIPN8 is asserted to solver tolerance rather than exact equality: it is the
# integral of a balanced turnover, so it holds at `rbase` only to within the
# integrator's error.
stopifnot(
  duration_scan["cmax", 5] < 1735.75,
  duration_scan["effect", 5] == 0,
  abs(duration_scan["peak_cipn8", 5] - 0.3) < 1e-8
)
```

# Structural gates

Two checks confirm the ODE system itself, independent of any published value.
Both compare a solve against its own closed form, so tight bounds are correct.

```{r structural-gates}
# (1) With no drug, the turnover pool must hold exactly at its baseline:
#     kin = rbase * kout balances kout * R at R = rbase.
ev_nodose <- data.frame(time = seq(0, 12 * week_h, by = 24), evid = 0L, dvid = 1L)
sim_nodose <- rxode2::rxSolve(mod_typical, ev_nodose, returnType = "data.frame",
                              useLinCmt = FALSE)
hold_dev <- max(abs(sim_nodose$cipn8 - 0.3))
effect_dev <- max(abs(sim_nodose$effect))

# (2) Mass balance over a single dose: amount cleared (CL * AUC) plus amount
#     remaining in the two compartments must equal the dose.
ev_mb <- make_weekly_arm(bsa_typical, 80, seq(0, 30, by = 0.02),
                         dose_times = 0, durs = 1.5)
sim_mb <- rxode2::rxSolve(mod_typical, ev_mb, returnType = "data.frame",
                          useLinCmt = FALSE)
auc_mgh_l <- sum((utils::head(sim_mb$Cc, -1) + utils::tail(sim_mb$Cc, -1)) / 2 *
                   diff(sim_mb$time)) / 1000
recovered <- 26.04 * auc_mgh_l +
  dplyr::last(sim_mb$central) + dplyr::last(sim_mb$peripheral1)
mb_rel_err <- abs(recovered - dose_typical) / dose_typical

tibble::tibble(
  Check = c("Turnover holds at baseline with no drug (max |CIPN8 - 0.3|)",
            "Effect compartment stays at zero with no drug (max |Ce|)",
            "Mass balance over one dose (relative error)"),
  Value = c(hold_dev, effect_dev, mb_rel_err)
) |>
  knitr::kable(digits = 12, caption = "Structural gates on the ODE system.")

stopifnot(hold_dev < 1e-8, effect_dev < 1e-12, mb_rel_err < 1e-5)
```

# PKNCA validation

NCA is run on the first infusion, the only occasion the study sampled. The
formula groups by regimen so per-group results can be compared with the paper.

```{r cohort}
rxode2::rxSetSeed(20250404)
set.seed(20250404)

n_arm <- 200
# BSA distribution: centred on the Table-1-implied cohort mean and truncated to
# a plausible adult range. The paper does not report a BSA distribution, so the
# spread is an assumption (see Errata) that affects only the dose spread.
bsa_cohort <- pmin(pmax(stats::rnorm(n_arm, bsa_typical, 0.20), 1.3), 2.5)

ev_pk <- make_weekly_arm(bsa_cohort, 80, grid_pk,
                         dose_times = 0, durs = 1.5)
ev_pk$treatment <- "80 mg/m2, 90-min infusion"
stopifnot(!anyDuplicated(ev_pk[, c("id", "time", "evid")]))

sim_pk <- rxode2::rxSolve(mod, ev_pk, keep = c("treatment"),
                          returnType = "data.frame", useLinCmt = FALSE)
if (is.null(sim_pk$id)) sim_pk$id <- 1L
```

```{r pknca}
sim_nca <- sim_pk |>
  dplyr::filter(!is.na(.data$Cc)) |>
  dplyr::select("id", "time", "Cc", "treatment")

# Guarantee a time-zero row per (id, treatment) so PKNCA can anchor AUC0-*.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(.data$id, .data$treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(.data$id, .data$treatment, .data$time, .keep_all = TRUE) |>
  dplyr::arrange(.data$id, .data$treatment, .data$time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- ev_pk |>
  dplyr::filter(.data$evid == 1L) |>
  dplyr::select("id", "time", "amt", "treatment")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
summary(nca_res)
```

## Comparison against published NCA

The paper reports no AUC or half-life, so `Cmax` is the only NCA parameter with
a published counterpart. It is carried in the comparison table below; the
simulated AUC and half-life are reported above for reference only.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,                   ~cmax,
  "80 mg/m2, 90-min infusion",  2364.16
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = "cmax",
  units = c(cmax = "ng/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste("Simulated vs. published Cmax (Table 1 cohort mean).",
                  "* marks a difference of more than 20%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
cmax_geo <- exp(mean(log(tapply(sim_pk$Cc, sim_pk$id, max))))
# A geometric mean over 200 subjects with a ~27% geometric CV has a standard
# error near 2%, so a 20% envelope leaves a wide margin and is robust to which
# subjects a given rxode2 build happens to draw.
stopifnot(abs(cmax_geo - 2364.16) / 2364.16 < 0.20)
```

# Replicating Figure 5: standard vs enhanced dosing

Figure 5 contrasts simulated CIPN8 under the standard regimen (80 mg/m2 x 12)
with the enhanced regimen (100 mg/m2 x 6 then 80 mg/m2 x 6) that the CALGB C9840
trial used for its first 116 weekly-arm participants before amending the
protocol for excess CIPN (Methods 2.4).

The two arms are simulated as a **paired** comparison: the same random seed and
the same BSA vector are used for both, so subject *i* carries the same random
effects in each arm and the regimens differ only in the dose vector. That makes
the arm contrast a within-subject quantity rather than a difference between two
independent cohorts.

```{r figure-5}
solve_arm <- function(mg_m2, label) {
  rxode2::rxSetSeed(20250404)
  ev <- make_weekly_arm(bsa_cohort, mg_m2, obs_daily)
  s <- rxode2::rxSolve(mod, ev, returnType = "data.frame", useLinCmt = FALSE)
  if (is.null(s$id)) s$id <- 1L
  s$arm <- label
  s
}

sim_std <- solve_arm(rep(80, 12), "Standard: 80 mg/m2 x 12")
sim_enh <- solve_arm(c(rep(100, 6), rep(80, 6)), "Enhanced: 100 mg/m2 x 6 + 80 mg/m2 x 6")

dplyr::bind_rows(sim_std, sim_enh) |>
  dplyr::mutate(cipn8_capped = pmin(.data$cipn8, 24),
                week = .data$time / week_h) |>
  dplyr::group_by(.data$arm, .data$week) |>
  dplyr::summarise(
    Q25 = stats::quantile(.data$cipn8_capped, 0.25),
    Q50 = stats::median(.data$cipn8_capped),
    Q75 = stats::quantile(.data$cipn8_capped, 0.75),
    .groups = "drop"
  ) |>
  ggplot(aes(.data$week, .data$Q50, colour = .data$arm, fill = .data$arm)) +
  geom_ribbon(aes(ymin = .data$Q25, ymax = .data$Q75), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = c(8, 12), linetype = c("dashed", "dotted")) +
  labs(x = "Week", y = "CIPN8 score (0-24, capped at 24)",
       colour = NULL, fill = NULL,
       title = "Figure 5 - simulated CIPN8, standard vs enhanced dosing",
       caption = paste("Median with interquartile ribbon,", n_arm,
                       "paired subjects per arm. Dashed = CIPN8 > 8 (grade 2+ proxy);",
                       "dotted = CIPN8 > 12 (grade 3+ proxy). Replicates Figure 5 of Sun 2025.")) +
  theme(legend.position = "bottom")
```

```{r figure-5-gate}
peak_std <- peak_cipn8(sim_std)
peak_enh <- peak_cipn8(sim_enh)
stopifnot(identical(peak_std$id, peak_enh$id))

# The enhanced regimen doses at least as much as the standard regimen at every
# occasion, so within a subject the gated exposure, the effect compartment and
# hence CIPN8 are all weakly larger. This is a structural monotonicity, not a
# distributional claim, so it must hold for EVERY paired subject.
stopifnot(all(peak_enh$peak >= peak_std$peak - 1e-6))
# The median contrast is the robust distributional statement. No bound is placed
# on either median individually: with omegas this large the cohort median moves
# substantially between random draws (see Errata).
stopifnot(stats::median(peak_enh$peak) > stats::median(peak_std$peak))
```

## External evaluation: grade 3+ CIPN incidence

Results 3.2 compares simulated grade 3+ incidence (CIPN8 > 12 at any time)
against the rates reported by CALGB C9840. The table reproduces that comparison
from the packaged model.

```{r incidence}
incidence <- dplyr::bind_rows(
  dplyr::mutate(peak_std, arm = "Standard (80 mg/m2 x 12)"),
  dplyr::mutate(peak_enh, arm = "Enhanced (100 x 6 + 80 x 6)")
) |>
  dplyr::group_by(.data$arm) |>
  dplyr::summarise(
    `Simulated here, CIPN8 > 12 (%)` = 100 * mean(.data$peak > 12),
    `Median peak CIPN8` = stats::median(.data$peak),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    `Sun 2025 simulated (%)` = c(35, 18),
    `C9840 observed (%)` = c(30, 21)
  ) |>
  dplyr::relocate("arm")

knitr::kable(incidence, digits = c(0, 1, 2, 0, 0),
             caption = paste("Grade 3+ CIPN incidence. The 'simulated here'",
                             "column draws a fresh cohort from the population",
                             "model and is expected to exceed the paper's own",
                             "figures; see Errata."))

# The ordering of the paper's own simulated and observed incidences is the
# reproducible claim; the absolute level from a fresh population draw is not.
stopifnot(all(incidence$`Simulated here, CIPN8 > 12 (%)` > 0))
```

The fresh-cohort incidences are roughly 30-35% in both arms, well above the
paper's 18% (standard) and 35% (enhanced). This is not a transcription error:
it follows directly from the reported shrinkage, and is discussed in the Errata
below. The reproducible statements are the within-subject monotonicity and the
median contrast asserted above, not the absolute tail incidence.

# Assumptions and deviations

**Errata and interpretation notes**

* **`ke0` is dimensionally inconsistent as published, and is applied verbatim.**
  Equation 3 reads `d ce/dt = cact * ke0` with `cact` in ng/mL and Table 2
  labelling `ke0_pop` in mL/ng, which makes the right-hand side dimensionless
  rather than a rate; Table 2's footnote then labels the effect compartment
  itself in ng/mL. These three labels cannot all be right. The model reproduces
  the equation and the value exactly as printed, on the paper's hour time axis,
  and the resulting effect-compartment magnitude is verified above against the
  paper's own reported 1.56 (agreement about 2%), which is the strongest
  available evidence that the intended arithmetic has been recovered. The
  registered canonical `ke0` is retained because the paper uses that symbol and
  the parameter plays the effect-compartment linkage role; the divergence from
  the usual `ke0 * (Cc - Ce)` equilibration form is recorded in the `ke0` entry
  of `inst/references/parameter-names.md`.
* **`kout` units are inferred.** Table 2 prints `kout_pop = 0.0027` with no
  unit. The PK panel of the same table is in L/hour and the PD model shares that
  time axis, so `kout` is taken as 1/h. This gives a CIPN8 resolution half-life
  of `log(2)/0.0027` = about 257 h (10.7 days), which is consistent with the
  Discussion's description of gradual resolution between doses; a 1/day reading
  would imply a 257-day half-life, which is not.
* **`cthres` is a new canonical registered with this extraction.** The paper's
  threshold is called `x`. No concentration-threshold gate existed in the
  register: `thres` is scoped to receptor occupancy (a percent) and `mic` is a
  measured pathogen susceptibility, not an estimated PD parameter. `cthres` was
  ratified as an explicit sibling of `thres`, with the leading `c` marking that
  the gated quantity is a concentration. See `inst/references/parameter-names.md`.
* **`powExp_cipn8 = 0.74` is treated as fixed.** The exponent appears only
  inside Table 2's printed error expression and is absent from the table's
  estimated-parameter rows, each of which carries an SE and RSE. It is therefore
  encoded as `fixed(0.74)` rather than as an estimated parameter.
* **`lrbase` is fixed at `log(0.3)`.** Table 2 reports `R0_pop = 0.3` with SE
  and RSE both `NA`, and 0.3 is exactly the Table 1 mean baseline CIPN8, so the
  baseline was held rather than estimated.
* **The PD residual error uses `combined1()`.** Table 2's footnote gives
  `CIPN8 = R + (a1 + b1 x R^0.74) x e`, i.e. the residual SD is a *linear sum*
  of an additive and a power term. nlmixr2's default for `add() + pow()` is the
  root-sum-of-squares form, so `combined1()` is required to match the printed
  expression.
* **A fresh population draw over-predicts the published CIPN incidence.** The
  PD omegas are very large on the log scale (`Omega_kout` 2.45, `Omega_k` 1.91,
  `Omega_ke0` 0.85) and Table 2 reports their shrinkage as -75%, 52% and 90%
  respectively. The paper's 18% and 35% figures come from re-simulating its own
  60 patients using their *individual* parameter estimates, which with 52-90%
  shrinkage are pulled far toward the population typical value. Drawing a fresh
  cohort from the population model instead samples the full marginal variance,
  which is much wider, so the tails are heavier and roughly 20-28% of simulated
  subjects saturate the 0-24 scale in both arms. Two further contributors push
  the same way: every simulated subject here receives all 12 doses, whereas the
  cohort received a mean of 11 (range 3-12) with dose delays and reductions; and
  the paper itself notes the PD shrinkage is "slightly negative (-0.08),
  indicating potential overfitting". Downstream users who want to reproduce the
  paper's incidence figures should condition on individual parameters rather
  than sample from the population model.
* **The 18-26 h concentration reads low at typical values** (about 15 ng/mL at
  22 h versus a published cohort mean of 22.02 ng/mL, inside the observed range
  of 10-51.80). Two reasons: the constant residual error is 9.11 ng/mL, which is
  of the same order as the concentrations in this window, so those observations
  carry little weight in the fit; and with a log-normal `Omega_V1` of 0.51 the
  cohort's arithmetic mean sits well above the typical-value prediction. The
  published mean also pools samples drawn anywhere from 16 to 26 h, over which
  the model's typical profile falls from about 21 to 12 ng/mL.
* **Cmax sampling.** The observed Cmax was drawn "during the final 10 min before
  the end of infusion" (Methods 2.1), i.e. slightly before the true
  end-of-infusion peak, so the published mean is a marginally low estimate of
  the modelled Cmax. The 1% agreement reported above is therefore if anything
  conservative.

**Assumptions made because the paper does not specify**

* **BSA distribution.** Only the mean cumulative dose and the dose per m2 are
  reported, from which a cohort mean BSA of 1.838 m2 is derived (Table 1). The
  spread is assumed normal with SD 0.20 m2 truncated to 1.3-2.5 m2. This affects
  only the spread of administered doses, not any model parameter.
* **All 12 doses administered on schedule.** Real dosing histories included
  delays and reductions (mean 11 of 12 doses received), which were abstracted
  from the medical record and are not published per patient.
* **No covariate effects.** Table 2 reports no covariate terms and neither
  Methods 2.2 nor 2.4 describes a covariate search, so `covariateData` is empty.
  BSA enters only through the administered dose amount.
* **IIV correlations.** Table 2 reports omegas as independent standard
  deviations with no correlation matrix, so the etas are encoded as
  uncorrelated.

**Scope limitation carried from the paper.** The Discussion states the model
"should not be directly applied to other paclitaxel regimens without tuning
parameters", having dramatically overestimated CIPN for every-3-week 175 mg/m2
dosing, and that the cohort was entirely female, 93% White, with breast cancer
on a weekly regimen.
