---
title: "Tilmicosin PBPK-PD against Pasteurella multocida in swine (Sun 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tilmicosin PBPK-PD against Pasteurella multocida in swine (Sun 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", error = FALSE,
  fig.width = 7, fig.height = 4.5, dpi = 96
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sun_2026_tilmicosin_pbpk"))$reference`
* Article: <https://doi.org/10.1021/acs.jafc.5c11368>
* Supporting Information (Tables S1-S10 and the complete Berkeley Madonna
  source code): distributed as `jf5c11368_si_001.pdf` with the article.

Tilmicosin (TIL) is a macrolide, the semi-synthetic hydrolysis product of
tylosin, licensed for swine respiratory disease. Sun and colleagues built a
whole-body PBPK model for oral tilmicosin in pigs, coupled it to a hollow-fiber
infection model (HFIM) of *Pasteurella multocida*, and used the pair to
recommend a dosing regimen and project a withdrawal interval for edible tissue.

Two things make the paper unusual. First, it reports the **first collection of
pulmonary interstitial fluid (PIF) from conscious pigs by microdialysis**; PIF
is where *P. multocida* actually lives, so the free drug concentration there,
not in plasma, is the exposure that matters. Second, the PK/PD analysis is run
**twice**, once on plasma exposure and once on PIF exposure, and the two give
materially different targets.

This paper therefore contributes **three** model files, all sharing this
vignette:

| Model file | What it is |
|---|---|
| `Sun_2026_tilmicosin_pbpk` | 17-state whole-body oral PBPK in swine |
| `Sun_2026_tilmicosin_pkpd_plasma` | sigmoid Imax PK/PD integration, plasma arm |
| `Sun_2026_tilmicosin_pkpd_pif` | sigmoid Imax PK/PD integration, PIF arm |

The model equations are transcribed from the Supporting Information's Berkeley
Madonna source code and from main-text equations 3-5; parameter values come from
Table 2 (chemical-specific), Table 3 (PK/PD), and Supporting Information
Tables S3 (physiology) and S4 (Monte Carlo distributions).

## Structure

Four perfusion-limited, well-stirred compartments -- liver, kidney, muscle and a
lumped rest-of-body -- are connected by venous and arterial blood pools. Three
features distinguish this from a generic whole-body PBPK:

1. **A lumped four-segment oral absorption chain.** Drug moves
   `stomach` -> `duodenum` -> `a_small_intestine` -> `a_large_intestine` ->
   `a_feces` by first-order transit, and is absorbed out of the duodenum
   (`Ka1` = 0.8 /h) and, roughly 100-fold more slowly, out of the remaining
   small intestine (`Ka2` = 0.007 /h). The paper resolves the duodenum but
   lumps everything distal to it, which is why the extraction uses
   `a_small_intestine` / `a_large_intestine` rather than the full-resolution
   `jejunum` / `ileum` / `cecum` / `colon` chain.
2. **Absorbed drug enters the liver, not venous blood**, so oral tilmicosin
   undergoes first pass.
3. **The lung is permeability-limited.** It resolves into pulmonary blood
   (`vp_lung`), interstitial fluid (`is_lung`) and tissue (`int_lung`), with
   only unbound drug crossing out of blood and out of tissue.

Elimination is hepatobiliary (major) plus renal (minor). Critically, the
**biliary output is returned to the duodenum**, so the model carries
enterohepatic recirculation; `a_bile` is a cumulative record of biliary output
rather than a closed sink.

```{r load-models}
pbpk <- rxode2::rxode2(readModelDb("Sun_2026_tilmicosin_pbpk"))
pd_plasma <- rxode2::rxode2(readModelDb("Sun_2026_tilmicosin_pkpd_plasma"))
pd_pif <- rxode2::rxode2(readModelDb("Sun_2026_tilmicosin_pkpd_pif"))

# Typical-value (deterministic) form of the PBPK for the calibration and
# structural checks; the Monte Carlo section below uses the full model.
pbpk_tv <- rxode2::zeroRe(pbpk)

WT_REF <- 40 # kg -- Sun 2026 Table S3 and SI code both fix BW = 40
```

# Population

```{r population, echo = FALSE}
pop <- pbpk$population
```

The PBPK model's own microdialysis experiment used
`r pop$n_subjects` crossbred (Landrace x Large White x Duroc) pigs of
`r pop$weight_range`. The structural model was calibrated and validated against
four further published swine datasets spanning 10-40 mg/kg oral doses
(Sun 2026 Table 1). All animals were healthy.

The challenge organism is `r pop$organism`.

Note the mismatch flagged in the model file: the **model's reference body weight
is 40 kg while the microdialysis pigs weighed 20 +/- 1 kg**. Because organ
volumes, blood flows, per-kg clearances and the mg/kg dose all scale linearly in
body weight, predicted *concentrations* in this model are essentially invariant
to weight and only amounts change, so this affects the residue/withdrawal
arithmetic rather than the concentration profiles validated below.

# Source trace

Every value in the three model files, and where it came from.

```{r source-trace, echo = FALSE}
trace <- tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "QCC, cardiac output", "4.944 L/h/kg", "Table S3",
  "QLC / QKC / QMC", "0.2725 / 0.1200 / 0.2510", "Table S3",
  "QRC (rest of body)", "1 - QLC - QKC - QMC = 0.3565", "SI code; matches Table S3",
  "VLC / VKC / VMC", "0.0247 / 0.0040 / 0.4000", "Table S3",
  "VartC / VvenC / VLUC", "0.0440 / 0.0160 / 0.0100", "Table S3",
  "VRC (rest of body)", "complement = 0.5013", "SI code; matches Table S3",
  "FVBLU / FVILU", "0.0200 / 0.4850", "Table S3 (SI code declares 0.262 / 0.188 -- Errata E2)",
  "VLUT (pulmonary tissue)", "VLU - VLUE - VLUB", "SI code (Errata E2)",
  "Kst, gastric emptying", "0.8 /h", "Table 2",
  "Kd, duodenal transit", "2.2 /h", "Table 2",
  "Kint, ileocolic transit", "2.0 /h", "Table 2",
  "Ka1, duodenal absorption", "0.8 /h", "Table 2",
  "Ka2, small-intestinal absorption", "0.007 /h", "Table 2",
  "Kf, fecal excretion", "0.05 /h", "Table 2",
  "PL / PK / PM / PLu / PR", "20 / 23 / 1.8 / 2.2 / 6.24", "Table 2",
  "Pb, pulmonary plasma binding", "0.188 bound (fu = 0.812)", "Table 2",
  "Pt, pulmonary tissue binding", "0.200", "Table 2",
  "K2, blood -> PIF", "2.6355773 /h", "SI code value in eq 3 slot (Errata E1)",
  "K1, PIF -> blood", "0.03 /h", "SI code value in eq 3 slot (Errata E1)",
  "K4, PIF -> tissue", "0.3 /h", "SI code value in eq 4 slot (Errata E1)",
  "K3, tissue -> PIF", "9.645148 /h", "SI code value in eq 4 slot (Errata E1)",
  "KbileC, hepatobiliary clearance", "0.126 L/h/kg", "Table 2 + Table S4 (SI code swaps -- Errata E3)",
  "KurineC, renal clearance", "0.002 L/h/kg", "Table 2 + Table S4 (SI code swaps -- Errata E3)",
  "IIV on PL/PK/PM/PLU", "lognormal CV 20%, omega^2 = log(1.04)", "Table S4",
  "IIV on Kd/Ka1/KurineC/KbileC", "lognormal CV 30%, omega^2 = log(1.09)", "Table S4",
  "Lung ODEs", "eqs 3-5 + SI code RLUB / RLUE / RLUT", "main text + SI",
  "All other ODEs", "SI code (Berkeley Madonna listing)", "SI",
  "Imax / IC50 / E0 / gamma, plasma", "17.11 / 130.05 / 3.57 / 0.46", "Table 3, plasma column",
  "Imax / IC50 / E0 / gamma, PIF", "6.82 / 16.06 / 2.36 / 2.27", "Table 3, PIF column",
  "Sigmoid Imax equation", "E = E0 - Imax*C^g/(C^g + IC50^g)", "eq 8",
  "Starting inoculum", "~1 x 10^6 CFU/mL", "SI, HFIM establishment",
  "propSd, addSd", "placeholders, FIXED", "not reported in Sun 2026"
)
knitr::kable(trace, caption = "Source trace for every model value.")
```

# Calibration: replicating Figure 3B

Sun 2026 Figure 3B compares the PBPK prediction of PIF concentration against the
authors' microdialysis observations (Supporting Information Table S7) after a
single 40 mg/kg oral dose, and the text makes a precise, falsifiable claim:

> all predicted values except one at the 48 h time point fell within the 2-fold
> error range of the observed data

That sentence is the answer key used below to settle Errata E1 and E2. It is
reproduced here without tuning anything.

```{r figure-3b}
obs_pif <- tibble::tibble(
  time = c(0.25, 0.5, 1, 2, 3, 4, 6, 8, 12, 24, 48),
  obs  = c(0.86, 2.84, 6.14, 9.16, 11.08, 9.82, 9.00, 8.21, 6.53, 3.83, 0.95),
  sd   = c(0.09, 0.24, 0.25, 0.59, 0.28, 0.57, 0.30, 0.25, 0.38, 0.11, 0.12)
)

ev_single <- rxode2::et(amt = 40 * WT_REF, cmt = "stomach") |>
  rxode2::et(seq(0, 48, by = 0.1)) |>
  rxode2::et(obs_pif$time)
ev_single <- as.data.frame(ev_single)
ev_single$WT <- WT_REF

sim_single <- as.data.frame(
  rxode2::rxSolve(pbpk_tv, ev_single, returnType = "data.frame")
)

chk <- obs_pif |>
  mutate(
    pred  = approx(sim_single$time, sim_single$Cpif, xout = time)$y,
    ratio = pmax(pred / obs, obs / pred)
  )

n_within <- sum(chk$ratio < 2)
worst_t <- chk$time[which.max(chk$ratio)]
```

```{r figure-3b-plot, echo = FALSE, fig.cap = "Replicates Figure 3B of Sun 2026: PBPK prediction of tilmicosin in pulmonary interstitial fluid after a single 40 mg/kg oral dose, against the microdialysis observations of Table S7 (mean +/- SD, n = 3). Shaded band is the 2-fold error range around the prediction."}
ggplot() +
  geom_ribbon(
    data = sim_single,
    aes(time, ymin = Cpif / 2, ymax = Cpif * 2), alpha = 0.15, fill = "steelblue"
  ) +
  geom_line(data = sim_single, aes(time, Cpif), colour = "steelblue", linewidth = 0.9) +
  geom_pointrange(
    data = obs_pif,
    aes(time, obs, ymin = obs - sd, ymax = obs + sd)
  ) +
  labs(
    x = "Time (h)", y = "Tilmicosin in PIF (ug/mL)",
    title = "PIF calibration, single 40 mg/kg oral dose"
  ) +
  theme_bw()
```

```{r figure-3b-assert}
knitr::kable(
  chk |> mutate(across(c(pred, ratio), ~ round(.x, 3))),
  caption = "Observed (Table S7) vs predicted PIF concentration, with the fold-difference."
)

stopifnot(
  # The paper's own claim, verbatim: exactly one point outside 2-fold, and it
  # is the 48 h point. This is a whole-chain answer key -- it simultaneously
  # confirms the Errata E1 lung-constant assignment and the Errata E2 lung
  # sub-compartment volumes, because no other reading reproduces it.
  n_within == 10L,
  identical(worst_t, 48),
  # Peak height and timing.
  abs(max(sim_single$Cpif) - 11.08) / 11.08 < 0.15,
  # The paper reports MAPE between 32.7% and 54.4% in PIF, muscle, liver and
  # kidney (SI, "PBPK Model Validation").
  mean(abs(chk$pred - chk$obs) / chk$obs) < 0.544
)
```

`r n_within` of `r nrow(chk)` predictions fall within 2-fold of the observation,
and the single failure is at `r worst_t` h, exactly as the paper states. The
simulated peak is `r round(max(sim_single$Cpif), 2)` ug/mL against an observed
`r max(obs_pif$obs)` ug/mL, and the mean absolute percentage error is
`r round(100 * mean(abs(chk$pred - chk$obs) / chk$obs), 1)`%, inside the
32.7-54.4% band the Supporting Information reports for this matrix.

**Why this matters.** Table 2 and the SI code disagree about which of the four
lung transfer constants belongs to which direction, and about the lung
sub-compartment volumes. Sweeping all sixteen combinations of the two
transposable pairs against these eleven observations, only the assignment
shipped in the model file reproduces the sentence above; the literal Table 2
reading gives a PIF Cmax of 1.06 ug/mL against an observed 11.08. See Errata E1
and E2.

# Structural check: mass balance

The Supporting Information closes a mass balance over the model. The physical
states (everything except the two bookkeeping integrators `a_bile` and
`a_oral_absorbed`) must always sum to the administered dose.

```{r mass-balance}
final <- tail(sim_single, 1)
physical <- with(final,
  stomach + duodenum + a_small_intestine + a_large_intestine + liver +
    venous + arterial + kidney + muscle + other +
    vp_lung + is_lung + int_lung + urine + a_feces
)
rel_err <- abs(physical - 40 * WT_REF) / (40 * WT_REF)

stopifnot(rel_err < 1e-8)
```

Mass balance closes to a relative error of `r signif(rel_err, 3)`.

The route split is itself informative: of the `r 40 * WT_REF` mg administered,
`r round(final$a_oral_absorbed)` mg
(`r round(100 * final$a_oral_absorbed / (40 * WT_REF), 1)`%) is absorbed across
the gut wall over 48 h, `r round(final$a_feces)` mg leaves in faeces, and only
`r round(final$urine, 1)` mg appears in urine. That ordering matches the paper's
prose -- *"following excretion in bile, the majority are excreted in feces,
while a minor portion is eliminated via renal excretion in urine"* -- and is one
of the three documents used to settle Errata E3.

# Structural check: replicating the Table S9 sensitivity analysis

Sun 2026 perturbed each parameter by +10% and reported the normalised
sensitivity coefficient of the AUC in each tissue (Supporting Information
Table S9). Reproducing that table is a strong test of the whole flow-limited
backbone, because each partition coefficient should move its own tissue's AUC
and nothing else.

```{r sensitivity}
base_pars <- c(
  lkp_liver = log(20), lkp_kidney = log(23), lkp_lung = log(2.2),
  lkp_muscle = log(1.8), lkt_duodenum = log(2.2), lka_duodenum = log(0.8),
  lcl_renal = log(0.002), lcl_nonren = log(0.126)
)

# The SI code's own scenario: 40 mg/kg once daily for three days.
ev_sens <- rxode2::et(amt = 40 * WT_REF, cmt = "stomach", ii = 24, addl = 2) |>
  rxode2::et(seq(0, 720, by = 0.5))
ev_sens <- as.data.frame(ev_sens)
ev_sens$WT <- WT_REF

auc_vec <- function(delta = NULL) {
  p <- base_pars
  if (!is.null(delta)) p[delta] <- log(exp(p[delta]) * 1.1)
  ss <- as.data.frame(
    rxode2::rxSolve(pbpk_tv, ev_sens, params = p, returnType = "data.frame")
  )
  trap <- function(x) sum(diff(ss$time) * (head(x, -1) + tail(x, -1)) / 2)
  c(AUCCV = trap(ss$Cc), AUCCLUE = trap(ss$Cpif), AUCCK = trap(ss$Ckidney),
    AUCCL = trap(ss$Cliver), AUCCM = trap(ss$Cmuscle))
}

base_auc <- auc_vec()
nsc <- vapply(
  names(base_pars),
  function(nm) (auc_vec(nm) - base_auc) / base_auc / 0.1,
  numeric(5)
)
nsc <- round(t(nsc), 2)
```

```{r sensitivity-table, echo = FALSE}
published_s9 <- tibble::tribble(
  ~Parameter,     ~AUCCV, ~AUCCLUE, ~AUCCK, ~AUCCL, ~AUCCM,
  "lkp_liver",      0.00,  0.00,  0.00,  1.00,  0.00,
  "lkp_kidney",     0.00,  0.00,  1.00,  0.00,  0.00,
  "lkp_lung",       0.00,  1.00,  0.00,  0.00,  0.00,
  "lkp_muscle",     0.00,  0.00,  0.00,  0.00,  0.99,
  "lkt_duodenum",  -0.92, -0.92, -0.92, -0.92, -0.92,
  "lka_duodenum",   0.93,  0.93,  0.93,  0.93,  0.93,
  "lcl_renal",     -0.57, -0.57, -0.61, -0.55, -0.57,
  "lcl_nonren",    -0.39, -0.39, -0.39, -0.39, -0.39
)

cmp_s9 <- as.data.frame(nsc) |>
  tibble::rownames_to_column("Parameter") |>
  tidyr::pivot_longer(-Parameter, names_to = "Tissue", values_to = "Replicated") |>
  left_join(
    published_s9 |>
      tidyr::pivot_longer(-Parameter, names_to = "Tissue", values_to = "Published"),
    by = c("Parameter", "Tissue")
  ) |>
  mutate(Agrees = ifelse(abs(Replicated - Published) <= 0.10, "yes", "NO"))

knitr::kable(
  cmp_s9 |>
    tidyr::pivot_wider(
      id_cols = Parameter,
      names_from = Tissue,
      values_from = c(Replicated, Published)
    ),
  caption = paste(
    "Replicated vs published (Table S9) normalised sensitivity coefficients,",
    "+10% perturbation, 40 mg/kg once daily x 3."
  )
)
```

```{r sensitivity-assert}
agree_by_param <- cmp_s9 |>
  group_by(Parameter) |>
  summarise(all_agree = all(Agrees == "yes"), .groups = "drop")

stopifnot(
  # All four partition coefficients must reproduce exactly and be perfectly
  # selective: each moves its own tissue's AUC with NSC 1 and no other.
  all(agree_by_param$all_agree[
    agree_by_param$Parameter %in%
      c("lkp_liver", "lkp_kidney", "lkp_lung", "lkp_muscle")
  ]),
  # Both absorption-chain parameters reproduce, and with the published sign.
  all(nsc["lkt_duodenum", ] < 0),
  all(nsc["lka_duodenum", ] > 0),
  # The two elimination rows do NOT reproduce; see Errata E3. Assert the
  # ordering the paper's own prose requires (biliary >> renal), which is the
  # opposite of what Table S9 prints.
  all(abs(nsc["lcl_nonren", ]) > abs(nsc["lcl_renal", ]))
)
```

Six of the eight rows reproduce. All four partition coefficients are exactly
selective -- each moves only its own tissue's AUC, with a normalised
sensitivity of 1.00 -- and the two absorption-chain parameters reproduce with
the right sign and close magnitude (`r nsc["lkt_duodenum", "AUCCV"]` vs -0.92
for duodenal transit; `r nsc["lka_duodenum", "AUCCV"]` vs 0.93 for duodenal
absorption).

The two **elimination** rows do not reproduce, and this is diagnostic rather
than a transcription failure. Under the shipped assignment, hepatobiliary
clearance dominates (NSC `r nsc["lcl_nonren", "AUCCV"]`) and renal clearance is
negligible (`r nsc["lcl_renal", "AUCCV"]`) -- exactly the ordering the paper's
own prose describes. Table S9 prints the opposite ordering (-0.39 biliary,
-0.57 renal), and swapping the two values back to the SI code's assignment does
not reproduce Table S9 either. Four matching rows plus two failing rows
localises the defect to those two rows of Table S9, not to the model backbone.
See Errata E3.

# PKNCA validation

Supporting Information Table S8 reports non-compartmental parameters for
tilmicosin in PIF, computed from the Table S7 observations after a single
40 mg/kg oral dose. The comparison below runs PKNCA over **the paper's own
sampling times** for both the observed means and the model prediction, so the
two sides are scored on the same quantity.

```{r pknca}
# Filter on !is.na() ONLY -- a `time > 0` or `Cc > 0` filter would drop the
# time-zero anchor PKNCA needs for AUC.
sim_at_obs <- as.data.frame(rxode2::rxSolve(
  pbpk_tv,
  local({
    e <- as.data.frame(
      rxode2::et(amt = 40 * WT_REF, cmt = "stomach") |>
        rxode2::et(c(0, obs_pif$time))
    )
    e$WT <- WT_REF
    e
  }),
  returnType = "data.frame"
))

nca_conc <- bind_rows(
  tibble::tibble(
    time = c(0, obs_pif$time), Cc = c(0, obs_pif$obs),
    treatment = "Observed (Table S7)"
  ),
  tibble::tibble(
    time = sim_at_obs$time, Cc = sim_at_obs$Cpif,
    treatment = "Simulated"
  )
) |>
  mutate(id = 1L) |>
  filter(!is.na(Cc))

nca_dose <- nca_conc |>
  distinct(id, treatment) |>
  mutate(time = 0, amt = 40 * WT_REF)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id),
  intervals = intervals
))
```

## Comparison against published NCA

```{r nca-compare}
published_nca <- tibble::tribble(
  ~treatment,            ~cmax, ~tmax, ~auclast, ~half.life,
  "Observed (Table S7)", 11.08,  3.00,   216.03,      13.14,
  "Simulated",           11.08,  3.00,   216.03,      13.14
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_nca,
  by            = "treatment",
  params        = c("cmax", "tmax", "auclast", "half.life"),
  units         = c(cmax = "ug/mL", auclast = "ug*h/mL",
                    tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated and observed NCA against the published Table S8 values.",
    "Both rows are scored against the same published reference, so the",
    "'Observed' row measures the NCA setup and the 'Simulated' row measures",
    "the model. * differs from the reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

```{r nca-assert}
wide <- as.data.frame(nca_res) |>
  tidyr::pivot_wider(
    id_cols = treatment, names_from = PPTESTCD, values_from = PPORRES
  )
obs_row <- wide |> filter(treatment == "Observed (Table S7)")
sim_row <- wide |> filter(treatment == "Simulated")

stopifnot(
  # The NCA setup itself is correct: re-running PKNCA on the paper's own
  # observed means reproduces the published Cmax, Tmax and half-life.
  abs(obs_row$cmax - 11.08) < 0.01,
  abs(obs_row$tmax - 3.00) < 0.01,
  abs(obs_row$half.life - 13.14) / 13.14 < 0.05,
  # The published AUC of 216.03 lies between the observed AUClast and AUCinf,
  # so the paper's integration convention is bracketed rather than pinned.
  obs_row$auclast < 216.03, obs_row$aucinf.obs > 216.03,
  # The model reproduces peak height and timing well ...
  abs(sim_row$cmax - 11.08) / 11.08 < 0.15,
  abs(sim_row$tmax - 3.00) < 0.01,
  # ... but over-predicts the terminal phase, which is the paper's own
  # acknowledged limitation. Pin the direction and rough size so a future
  # change to the model has to explain itself.
  sim_row$half.life > 2 * obs_row$half.life,
  sim_row$auclast > obs_row$auclast
)
```

Running PKNCA on the paper's own observed means reproduces Table S8's Cmax
(`r round(obs_row$cmax, 2)` vs 11.08), Tmax (`r obs_row$tmax` vs 3.00) and
half-life (`r round(obs_row$half.life, 2)` vs 13.14) essentially exactly, which
validates the NCA setup. The published AUC of 216.03 `ug*h/mL` sits between the
observed AUClast (`r round(obs_row$auclast, 1)`) and AUCinf
(`r round(obs_row$aucinf.obs, 1)`), so the paper does not state its integration
convention precisely enough to pin.

The **simulated** row matches on peak height (`r round(sim_row$cmax, 2)` ug/mL)
and peak time exactly, but over-predicts the terminal phase: half-life
`r round(sim_row$half.life, 1)` h against an observed
`r round(obs_row$half.life, 1)` h, and AUClast `r round(sim_row$auclast)` against
`r round(obs_row$auclast)`. This is the same deficiency the paper reports itself
-- it is the sole 2-fold failure at 48 h in Figure 3B, and the reason the
paper's PIF MAPE is 47% rather than single-digit. It is *not* a transcription
error: the model is faithful and the model is imperfect. Because the PK/PD index
below uses AUC over the first 24 h, where the fit is good, the consequence for
the dosing conclusions is smaller than the terminal-phase numbers suggest.

# PK/PD: the sigmoid Imax models

Sun 2026 equation 8 relates the PK/PD index `C` = AUC0-24/MIC to the signed
change in bacterial count over a 24 h window:

$$E = E_0 - \frac{I_{max} \cdot C^{\gamma}}{C^{\gamma} + IC_{50}^{\gamma}}$$

The published PDF drops the minus sign (a symbol-font casualty), but the sign
convention is fixed independently by Table 3's own target rows, reproduced
below.

```{r pd-targets}
# Read the sigmoid parameters straight off the compiled model object, so the
# targets below are derived from the shipped model rather than re-typed.
sigmoid_pars <- function(ui) {
  th <- ui$theta
  list(e0 = exp(th[["le0"]]), imax = exp(th[["limax"]]),
       ic50 = exp(th[["lic50"]]), g = exp(th[["lhill"]]))
}

sigmoid_effect <- function(p, C) {
  p$e0 - p$imax * C^p$g / (C^p$g + p$ic50^p$g)
}

solve_index <- function(ui, target_E) {
  p <- sigmoid_pars(ui)
  uniroot(
    function(C) sigmoid_effect(p, C) - target_E,
    c(1e-8, 1e8), tol = 1e-12
  )$root
}
```

```{r pd-targets-table}
targets <- tibble::tibble(
  Endpoint = c("Bacteriostatic (E = 0)", "Bactericidal (E = -3)",
               "Eradication (E = -4)"),
  E = c(0, -3, -4),
  `Published plasma` = c(7.17, 46.54, 78.66),
  `Published PIF` = c(12.13, 28.48, 51.08)
) |>
  rowwise() |>
  mutate(
    `Replicated plasma` = round(solve_index(pd_plasma, E), 3),
    `Replicated PIF` = round(solve_index(pd_pif, E), 3)
  ) |>
  ungroup()

knitr::kable(
  targets |> select(-E),
  caption = paste(
    "Replicates Table 3 of Sun 2026: AUC0-24/MIC (h) required for each",
    "antibacterial endpoint, in plasma and in pulmonary interstitial fluid."
  )
)

stopifnot(
  max(abs(targets$`Replicated plasma` - targets$`Published plasma`) /
        targets$`Published plasma`) < 0.001,
  max(abs(targets$`Replicated PIF` - targets$`Published PIF`) /
        targets$`Published PIF`) < 0.001
)
```

All six published targets are reproduced to better than 0.1%, which is what
confirms the equation-8 sign convention and the Table 3 transcription.

```{r figure-6, echo = FALSE, fig.cap = "Replicates Figure 6 of Sun 2026: fitted sigmoid Imax curves relating the PK/PD index AUC0-24/MIC to the change in bacterial count, for plasma (A) and pulmonary interstitial fluid (B). Points are the observed HFIM data of Supporting Information Table S10."}
curve_for <- function(ui, label) {
  p <- sigmoid_pars(ui)
  C <- 10^seq(0, 3, length.out = 200)
  tibble::tibble(C = C, E = sigmoid_effect(p, C), matrix = label)
}

curves <- bind_rows(
  curve_for(pd_plasma, "A. Plasma"),
  curve_for(pd_pif, "B. PIF")
)

s10 <- tibble::tribble(
  ~dose, ~time, ~pd_plasma, ~auc_plasma, ~pd_pif, ~auc_pif,
  20, 24,  1.27,  27.93,  1.03,  86.30,
  20, 48,  0.90,  63.82, -1.40, 197.29,
  20, 72,  0.31, 101.79, -1.73, 314.64,
  40, 24, -0.96,  55.87, -1.35, 172.68,
  40, 48, -1.23, 127.65, -1.76, 394.58,
  40, 72, -1.50, 203.58, -2.12, 629.28,
  50, 24, -1.10,  69.83, -2.27, 215.85,
  50, 48, -1.40, 159.57, -2.83, 493.22,
  50, 72, -1.74, 254.48, -2.83, 786.60,
  60, 24, -1.43,  83.80, -2.76, 259.02,
  60, 48, -1.67, 191.48, -3.75, 591.87,
  60, 72, -3.54, 305.38, -3.75, 943.92
)

s10_long <- bind_rows(
  s10 |> transmute(C = auc_plasma, E = pd_plasma, matrix = "A. Plasma"),
  s10 |> transmute(C = auc_pif, E = pd_pif, matrix = "B. PIF")
)

ggplot(curves, aes(C, E)) +
  geom_line(colour = "steelblue", linewidth = 0.9) +
  geom_point(data = s10_long, alpha = 0.8) +
  geom_hline(yintercept = c(0, -3, -4), linetype = "dashed", alpha = 0.4) +
  scale_x_log10() +
  facet_wrap(~matrix) +
  labs(
    x = "AUC0-24/MIC (h)",
    y = "Change in bacterial count (log10 CFU/mL)"
  ) +
  theme_bw()
```

The fitted curves pass through the published target values by construction, but
the Table S10 observations scatter widely around them -- particularly the
plasma arm, whose Hill coefficient of 0.46 makes the curve very shallow. The
paper reports no goodness-of-fit statistic for these fits, so the scatter is
carried forward as published (Errata E7).

## Simulating a treatment course

The two PD models integrate the bacterial density so that log10(CFU/mL) moves by
exactly `E` over each 24 h window. Driving them with the Table S10 index for the
recommended 40 mg/kg once-daily regimen:

```{r pd-sim}
simulate_kill <- function(model, index_value, hours = 72) {
  ev <- as.data.frame(rxode2::et(seq(0, hours, by = 1)))
  ev$AUCMIC_TILM <- index_value
  as.data.frame(rxode2::rxSolve(model, ev, returnType = "data.frame"))
}

kill <- bind_rows(
  simulate_kill(pd_plasma, 55.87) |> mutate(matrix = "Plasma (index 55.87 h)"),
  simulate_kill(pd_pif, 172.68) |> mutate(matrix = "PIF (index 172.68 h)")
)
```

```{r pd-sim-plot, echo = FALSE, fig.cap = "Predicted P. multocida time-kill over a 72 h course at the recommended 40 mg/kg once-daily regimen, driven by the 24 h PK/PD index Sun 2026 Table S10 reports for that regimen in each matrix. Dashed lines mark the bactericidal (3-log) and eradication (4-log) reductions from the 1e6 CFU/mL inoculum."}
ggplot(kill, aes(time, Cc, colour = matrix)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = c(6, 3, 2), linetype = "dashed", alpha = 0.4) +
  labs(x = "Time (h)", y = "log10 CFU/mL", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r pd-sim-assert}
pif_24 <- kill |> filter(matrix == "PIF (index 172.68 h)", time == 24) |> pull(Cc)
pl_24 <- kill |> filter(matrix == "Plasma (index 55.87 h)", time == 24) |> pull(Cc)

stopifnot(
  # At the recommended regimen the PIF index (172.68 h) is far above the PIF
  # eradication target of 51.08 h, so the PIF arm must clear more than 4 log10
  # in the first 24 h -- the paper's headline conclusion.
  6 - pif_24 > 4,
  # The plasma index (55.87 h) sits between the plasma bactericidal (46.54 h)
  # and eradication (78.66 h) targets, so the plasma arm must clear more than
  # 3 but less than 4 log10.
  6 - pl_24 > 3, 6 - pl_24 < 4
)
```

At the recommended regimen the PIF arm clears
`r round(6 - pif_24, 2)` log10 CFU/mL in the first 24 h, above the 4-log
eradication threshold, while the plasma arm clears `r round(6 - pl_24, 2)`
log10 -- bactericidal but short of eradication. That ordering is precisely the
paper's conclusion: *"the 40 mg/kg regimen achieving the eradication of
P. multocida at the infection site"* while plasma exposure alone would be scored
as merely bactericidal.

## Why the index is carried as a ratio, not as AUC divided by an MIC

The register's rule (see `AUCMIC_TYLO` in `inst/references/covariate-columns.md`)
is to carry an absolute `AUC_<DRUG>` plus a model `mic` parameter whenever the
paper reports the challenge strain's MIC. Sun 2026 *does* report one -- ZJWZ-A,
8 ug/mL. But that MIC does not reconstruct the paper's own index, and this is
worth showing because it is the reason the models ship with `AUCMIC_TILM`.

```{r mic-check}
ev_mc <- rxode2::et(amt = 40 * WT_REF, cmt = "stomach", ii = 24, addl = 2) |>
  rxode2::et(seq(0, 72, by = 0.05))
ev_mc <- as.data.frame(ev_mc)
ev_mc$WT <- WT_REF
s_mc <- as.data.frame(
  rxode2::rxSolve(pbpk_tv, ev_mc, returnType = "data.frame")
)

cum_auc <- function(tt, cc, upto) {
  keep <- tt <= upto
  sum(diff(tt[keep]) * (head(cc[keep], -1) + tail(cc[keep], -1)) / 2)
}

implied <- tibble::tibble(
  Matrix = c("Plasma", "PIF"),
  `Model AUC0-24 (ug*h/mL)` = c(
    cum_auc(s_mc$time, s_mc$Cc, 24),
    cum_auc(s_mc$time, s_mc$Cpif, 24)
  ),
  `Table S10 index at 24 h (h)` = c(55.87, 172.68)
) |>
  mutate(`Implied MIC (ug/mL)` =
           `Model AUC0-24 (ug*h/mL)` / `Table S10 index at 24 h (h)`)

knitr::kable(
  implied |> mutate(across(where(is.numeric), ~ round(.x, 3))),
  caption = paste(
    "Back-solving MIC = AUC / index against the verified PBPK, for the",
    "40 mg/kg once-daily regimen. The reported isolate MIC is 8 ug/mL."
  )
)

stopifnot(
  # The two matrices imply mutually inconsistent MICs ...
  max(implied$`Implied MIC (ug/mL)`) / min(implied$`Implied MIC (ug/mL)`) > 1.3,
  # ... and both are far below the reported 8 ug/mL.
  all(implied$`Implied MIC (ug/mL)` < 8 / 3)
)
```

The plasma and PIF columns of Table S10 imply MICs of
`r round(implied$"Implied MIC (ug/mL)"[1], 3)` and
`r round(implied$"Implied MIC (ug/mL)"[2], 3)` ug/mL respectively -- mutually
inconsistent, and both roughly an order of magnitude below the reported
8 ug/mL. Adopting the absolute-AUC form with `mic = 8` would put the PIF index
at roughly `r round(implied$"Model AUC0-24 (ug*h/mL)"[2] / 8, 1)` h at the
recommended dose, *below* the 28.48 h bactericidal target, and would therefore
ship a model contradicting the paper's own headline conclusion. Carrying the
ratio directly reproduces the paper exactly and introduces no inferred value.

The general lesson, recorded in the register entry for `AUCMIC_TILM`: **a
reported MIC is not automatically a splittable MIC.** Test whether it actually
reconstructs the paper's own index column before applying the
`AUC_<DRUG>`-versus-`AUCMIC_<DRUG>` rule.

# Population PBPK (Monte Carlo)

Sun 2026 propagated eight chemical-specific parameters through a 1000-animal
Monte Carlo analysis (Table S4). The cohort below is reduced to 100 animals for
build time; the distributions are unchanged.

```{r monte-carlo}
set.seed(20260829)
rxode2::rxSetSeed(20260829)

N_ANIMALS <- 100

ev_pop <- rxode2::et(amt = 40 * WT_REF, cmt = "stomach", ii = 24, addl = 2) |>
  rxode2::et(seq(0, 96, by = 0.5))
ev_pop <- as.data.frame(ev_pop)
ev_pop$WT <- WT_REF

pop_sim <- as.data.frame(rxode2::rxSolve(
  pbpk, ev_pop, nSub = N_ANIMALS, returnType = "data.frame"
))
```

```{r monte-carlo-plot, echo = FALSE, fig.cap = "Population PBPK prediction of tilmicosin in pulmonary interstitial fluid, 40 mg/kg once daily for 3 days, 100 virtual animals sampling the Table S4 distributions. Line is the median, band the 5th-95th percentiles."}
pop_q <- pop_sim |>
  group_by(time) |>
  summarise(
    lo = quantile(Cpif, 0.05), md = median(Cpif), hi = quantile(Cpif, 0.95),
    .groups = "drop"
  )

ggplot(pop_q, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = md), colour = "steelblue", linewidth = 0.9) +
  geom_pointrange(
    data = obs_pif,
    aes(time, y = obs, ymin = obs - sd, ymax = obs + sd), inherit.aes = FALSE
  ) +
  labs(x = "Time (h)", y = "Tilmicosin in PIF (ug/mL)") +
  theme_bw()
```

```{r monte-carlo-assert}
# Per-animal AUC0-24 in PIF; the quantity the PK/PD index is built on.
auc24 <- pop_sim |>
  filter(time <= 24) |>
  group_by(sim.id) |>
  summarise(
    auc = sum(diff(time) * (head(Cpif, -1) + tail(Cpif, -1)) / 2),
    .groups = "drop"
  )

stopifnot(
  # Assert on the CENTRE and on ROBUST QUANTILES, never on the extremes of a
  # random cohort (see CLAUDE.md). The typical-value AUC0-24 must sit close to
  # the cohort median, because all eight sampled parameters are lognormal with
  # the Table 2 value as the median.
  abs(median(auc24$auc) / cum_auc(s_mc$time, s_mc$Cpif, 24) - 1) < 0.10,
  # Spread is bounded: the Table S4 CVs are 20-30% on parameters that mostly
  # act sub-proportionally on PIF AUC.
  quantile(auc24$auc, 0.95) / quantile(auc24$auc, 0.05) < 5
)
```

The cohort median AUC0-24 in PIF is
`r round(median(auc24$auc), 1)` `ug*h/mL` against a typical-value
`r round(cum_auc(s_mc$time, s_mc$Cpif, 24), 1)` `ug*h/mL`, confirming that the
Table 2 point estimates are being carried as medians rather than as arithmetic
means (Errata E6).

# Assumptions and deviations

## Errata

**E1 -- Table 2 transposes all four lung transfer constants within each
bidirectional pair.** The Supporting Information code declares `K1..K4`, but
those declarations are *dead code*: the code's own ODEs reference
`KBE`/`KEB`/`KET`/`KTE`, which are never declared anywhere in the deposited
listing. The only surviving name-to-slot map is therefore main-text equations
3-5, which place `K2` on pulmonary blood -> PIF, `K1` on PIF -> pulmonary blood,
`K4` on PIF -> tissue and `K3` on tissue -> PIF. Feeding the SI's declared
*values* through the main text's *slots* gives the assignment shipped here:

| ODE slot | Shipped (1/h) | Table 2 prints for that direction |
|---|---|---|
| pulmonary blood -> PIF | 2.6355773 | 0.03 |
| PIF -> pulmonary blood | 0.03 | 2.63 |
| PIF -> lung tissue | 0.3 | 9.65 |
| lung tissue -> PIF | 9.645148 | 0.30 |

This reading is not a preference: of the sixteen possible combinations of the
two transposable pairs and the two lung-volume readings, it is the only one that
reproduces the paper's own Figure 3B claim, as demonstrated above. The literal
Table 2 reading gives a PIF Cmax of 1.06 ug/mL against an observed 11.08.

*General lesson:* a deposited control stream can be internally dead. Check that
its declarations are actually referenced by its own ODEs before treating them as
authoritative.

**E2 -- the Supporting Information code's lung sub-compartment volumes are a
copy-paste leftover from the same laboratory's previous paper.** The code
declares `FVBLU = 0.262` and `FVILU = 0.188`; Table S3 gives 0.0200 and 0.4850.
The code's two numbers are *exactly* the cefquinome values in
`Mi_2023_cefquinome_pbpk` (`v_vp_lung <- v_lung * 0.262`,
`v_is_lung <- v_lung * 0.188`), which shares the Berkeley Madonna template and a
co-author. Table S3's values are shipped and are required for the Figure 3B
match. Separately, Table S3's own `FVTLU = 0.0100` for pulmonary tissue is
inconsistent with its siblings: the code computes pulmonary tissue as the
complement, `VLUT = VLU - VLUE - VLUB`, which is 0.495 of lung volume. The
complement is shipped.

**E3 -- `KbileC` and `KurineC` are swapped in the Supporting Information code,
and Table S9's two elimination rows do not reproduce under either assignment.**
The code declares `KbileC = 0.002` and `KurineC = 0.126`, with their unit
comments swapped alongside. Three documents agree against it: Table 2 (hepatic
0.126, renal 0.002), Table S4 (Monte Carlo means, same), and the prose
*"following excretion in bile, the majority are excreted in feces, while a minor
portion is eliminated via renal excretion in urine."* Table 2's assignment is
shipped, and the mass-balance section above confirms it produces the stated
faecal-dominant route split. Table S9's elimination sensitivities reproduce
under neither assignment; since the other six rows of that table reproduce, the
defect is localised to those two rows.

**E4 -- the lung transfer fluxes are dimensionally asymmetric, as published.**
The inbound terms multiply a *concentration*
(`K2 * ALUB * (1 - Pb) / VLUB`) while the outbound terms multiply an *amount*
(`K1 * ALUE`). The four constants therefore cannot all carry the "/h" that
Table 2 assigns them. This is printed identically in the main text and in the
deposited code, so it is a genuine feature of the published model rather than a
transcription slip, and it is reproduced as published. (The sibling
`Mi_2023_cefquinome_pbpk` divides by the sub-compartment volume on both sides;
Sun 2026 does not.) A practical consequence, noted in the model file's `WT`
covariate: this is the only part of the model whose behaviour is not invariant
to body weight.

**E5 -- five variables are used but never declared in the deposited code.**
`Kdm` (in the "other small intestine" equation) is `Kd`, the duodenal transit
constant -- the only rate constant that can drain the duodenum into that
segment, and what Figure 1 and Table 2 describe. `PBtissue` (in the lung
equations) is `Pt`. `CVK`, `CVM` and `CVR` take the standard flow-limited form
`A / (V * P)` shown explicitly for `CVL`. The same defect class is documented
for `Mi_2023_cefquinome_pbpk` (its Errata E3).

**E6 -- Table S4 parameterises each lognormal by its arithmetic mean; the model
files use the Table 2 value as the median.** This puts the median about 1-2%
below the tabulated mean, and is done so that a typical-value simulation
reproduces the deterministic Table 2 model and Figures 3-4 exactly. The Monte
Carlo section above confirms the choice.

**E7 -- the Table 3 sigmoid parameters reproduce Table 3's own target rows
exactly but not the Table S10 observations.** All six published AUC/MIC targets
are recovered to better than 0.1%, so the transcription is certainly right, but
the (index, effect) pairs tabulated in Table S10 scatter substantially around
the fitted curves, as the Figure 6 replication shows. Sun 2026 reports no
goodness-of-fit statistic for these fits, so the scatter is carried forward as
published.

**E8 -- the PIF AUC disagrees between the paper's own sections.** The Results
text gives 235.70 `ug*h/mL` with a half-life of 13.10 h; Table S8 gives 216.03 and
13.14. Table S8 is used here as the more specific record.

## Assumptions

* **Body weight.** The model's 40 kg reference (Table S3 and the SI code) does
  not match the 20 +/- 1 kg pigs of the microdialysis experiment. All
  simulations here use the model's own 40 kg reference. As noted above,
  predicted concentrations are essentially weight-invariant in this model.
* **Residual error is not modelled.** Sun 2026 hand-calibrated in Berkeley
  Madonna by minimising the sum of squared residuals and reports only MAPE.
  `propSd` (PBPK) and `addSd` (both PD models) are FIXED placeholders for
  syntactic completeness and must not be read as estimates.
* **Between-animal variability is assumed, not estimated.** The Table S4
  coefficients of variation are the authors' literature defaults (20% for
  partition coefficients and transport rate constants, 30% otherwise), not
  variances fitted to data. All eta terms are therefore `fixed()`.
* **No physiological parameter carries variability.** Table S4 lists only
  chemical-specific parameters, so unlike `Mi_2023_cefquinome_pbpk` this model
  gives cardiac output no between-animal variability.
* **The dose equation (eq 9) is not reproduced.** Sun 2026 Table 4 derives
  per-endpoint doses from
  `dose = (AUC24h/MIC) * MIC90 * CL / (fu * F)`, but the paper does not report
  the `MIC90`, `CL`, `fu` and `F` values it substituted, so the arithmetic
  cannot be checked without inventing three of the four inputs. The Table 4
  doses are quoted in the model descriptions but not asserted here.
* **The withdrawal-interval analysis is not reproduced.** It depends on the
  residue Monte Carlo against Chinese and EMA maximum residue limits, which
  requires the tissue-residue datasets rather than the model alone.
