---
title: "Olanzapine and samidorphan (Sun 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Olanzapine and samidorphan (Sun 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Sun 2021 developed two separate population PK models from the same pooled
clinical program of the olanzapine/samidorphan combination tablet (OLZ/SAM):
one for olanzapine and one for samidorphan. Each is packaged as its own model.

* Citation: `r rxode2::rxode(readModelDb("Sun_2021_olanzapine"))$reference`
* Olanzapine model: `r rxode2::rxode(readModelDb("Sun_2021_olanzapine"))$description`
* Samidorphan model: `r rxode2::rxode(readModelDb("Sun_2021_samidorphan"))$description`
* Article: <https://doi.org/10.1002/jcph.1911> (open access; supplement Table S1
  lists the pooled studies and Figure S1 the model schematic)

# Population

The olanzapine model was fit to 601 subjects (9905 concentrations) from 10
OLZ/SAM studies: nine intensively sampled phase 1 studies and the sparsely
sampled phase 3 study ALK3831-A305. The samidorphan model was fit to 521
subjects (9321 concentrations) from the same 10 studies plus the
samidorphan-alone phase 1 study ALK33-B109 (Sun 2021 Table S1). Subjects were
healthy adults or adults with schizophrenia (41% / 59% in the olanzapine data
set, 57% / 43% in the samidorphan data set), with median age 36 and 34 years
(range 18-73), median body weight 76.9 and 76.4 kg, 32% and 29% women, and
42% and 50% Black (Sun 2021 Table 1). Olanzapine and samidorphan doses ranged
from 5 to 30 mg orally. Ten subjects with moderate hepatic impairment
(Child-Pugh B) and subjects with severe renal impairment came from dedicated
organ-impairment studies, and rifampin coadministration from a dedicated
drug-drug interaction study.

The same information is available programmatically via each model's
`population` metadata (e.g. `readModelDb("Sun_2021_olanzapine")()$population`).

# Source trace

Both models share the structure of Sun 2021 Figure S1: two-compartment
disposition with first-order absorption from a depot, a lag time on
absorption, and first-order elimination. Continuous covariates enter as
`TVP = theta_P * (COV / TVCOV)^theta_COV` and categorical covariates as
`TVP = theta_P * theta_CAT^CAT` (Methods, "Model Development"). Random
effects are exponential and the residual error is proportional,
`Yobs = Ypred * (1 + eps)` (Methods).

## Olanzapine (Table 3)

| Parameter | Value | Source location |
|-----------|-------|-----------------|
| `lcl` (CL/F) | log(15.5 L/h) | Table 3 |
| `lvc` (Vc/F) | log(656 L) | Table 3 |
| `lka` (ka) | log(0.861 1/h) | Table 3 |
| `ltlag` (ALAG) | fixed(log(0.782 h)) | Table 3, footnote a; Results |
| `lvp` (Vp/F) | log(225 L) | Table 3 |
| `lq` (Q/F) | log(6.15 L/h) | Table 3 |
| `e_wt_cl` | fixed(0.75), reference 70 kg | Table 3, footnote b; Figure 3 caption |
| `e_wt_vc` | fixed(1.0), reference 70 kg | Table 3, footnote b; Figure 3 caption |
| `e_age_vc` | 0.356, reference 36 years | Table 3; Figure 3 caption |
| `e_conmed_rifampicin_cl` | 1.80 | Table 3 |
| `e_smoke_cl` | 1.30 | Table 3; Table 2 (reference pools not-recorded status) |
| `e_fed_fdepot` | 0.943 | Table 3 |
| `e_hepimp_mod_cl` | fixed(0.875) | Table 3, footnote a |
| `e_renalimp_sev_cl` | 0.801 | Table 3 |
| `e_race_black_cl` | 1.10 | Table 3 |
| `e_sexf_cl` | 0.862 | Table 3 |
| `etalcl`, `etalvc`, `etalka`, `etaltlag` | 0.171, 0.127, 0.209, 0.319 | Table 3 (omega^2) |
| `etalvp`, `etalq` | fixed(0.223) each | Table 3; Results ("fixed at 50%") |
| `etaiov_ka_1..3` | 0.630 (occasions 2-3 fixed equal) | Table 3 'Interoccasion variability in Ka' |
| `propSd` | sqrt(0.0462) = 0.215 | Table 3 |

## Samidorphan (Table 4)

| Parameter | Value | Source location |
|-----------|-------|-----------------|
| `lcl` (CL/F) | log(35.4 L/h) | Table 4 |
| `lvc` (Vc/F) | log(297 L) | Table 4 |
| `lvp` (Vp/F) | log(124 L) | Table 4 |
| `lka` (ka) | log(6.61 1/h) | Table 4 |
| `ltlag` (ALAG) | log(0.323 h) | Table 4 |
| `lq` (Q/F) | log(12.1 L/h) | Table 4 |
| `e_wt_cl` | fixed(0.75), reference 70 kg | Table 4, footnote a; Figure 3 caption |
| `e_wt_vc` | fixed(1.0), reference 70 kg | Table 4, footnote a; Figure 3 caption |
| `e_conmed_rifampicin_cl` | 2.70 | Table 4 |
| `e_hepimp_mod_cl` | 0.810 | Table 4 |
| `e_renalimp_sev_cl` | 0.570 | Table 4 |
| `e_fed_ka` | 0.107 | Table 4 |
| `e_study_alk3831a305_tlag` | 10.1 | Table 4, footnotes b and c; Table 2 |
| `e_form_sam_tab_tlag` | 1.41 | Table 4 |
| `etalcl`, `etalvc`, `etalka`, `etaltlag`, `etalvp` | 0.087, 0.054, 1.76, 0.131, 0.681 | Table 4 (omega^2) |
| `etalq` | fixed(0.223) | Table 4; Results ("fixed at 50%") |
| `propSd` | sqrt(0.061) = 0.247 | Table 4 |

# Virtual cohort: the Figure 3 covariate individuals

Sun 2021 Figure 3 compares steady-state exposure of a set of "covariate
individuals" with a reference subject: a 36-year-old, 70 kg, non-Black,
nonsmoking man with normal hepatic and renal function taking OLZ/SAM
10 mg/10 mg once daily, fasted. Each covariate individual differs from the
reference in one covariate only. The paper simulated 500 profiles per
individual; here 200 are simulated per individual.

The between-subject random effects are drawn once in R from each model's
`omega` matrix and passed to the solver as data columns of a
`zeroRe()` copy of the model, so every covariate individual shares the same
200 sets of random effects (common random numbers). This makes each
exposure ratio a within-subject comparison and makes the whole simulation
reproducible across machines. For olanzapine a single occasion is simulated
(`OCC = 1`), so each subject carries one draw of the ka inter-occasion effect.

```{r cohort}
n_per_individual <- 200
tau <- 24
n_doses <- 28 # 4 weeks of once-daily dosing: steady state for the slow tail of the olanzapine CL/F distribution
t_ss <- (n_doses - 1) * tau
obs_times <- t_ss + c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 2, 2.5, 3, 3.5, 4, 5, 6, 8, 10, 12, 16, 20, 24)

reference_subject <- tibble(
  WT = 70, AGE = 36, SEXF = 0, RACE_BLACK = 0, SMOKE = 0, FED = 0,
  CONMED_RIFAMPICIN = 0, HEPIMP_MOD = 0, RENALIMP_SEV = 0, OCC = 1,
  FORM_SAM_TAB = 0, STUDY_ALK3831A305 = 0
)

# Figure 3 covariate individuals: label, changed covariate(s)
individuals <- list(
  "Reference" = list(),
  "Weight: low (44 kg)" = list(WT = 44),
  "Weight: high (141 kg)" = list(WT = 141),
  "Age: low (18 y)" = list(AGE = 18),
  "Age: high (73 y)" = list(AGE = 73),
  "Rifampin: presence" = list(CONMED_RIFAMPICIN = 1),
  "Smoking: smoker" = list(SMOKE = 1),
  "Hepatic impairment: moderate" = list(HEPIMP_MOD = 1),
  "Renal impairment: severe" = list(RENALIMP_SEV = 1),
  "Food: fed" = list(FED = 1),
  "Race: Black" = list(RACE_BLACK = 1),
  "Sex: female" = list(SEXF = 1),
  "Tablet: samidorphan alone" = list(FORM_SAM_TAB = 1)
)

make_individual <- function(label, changes, eta, id_offset) {
  covs <- reference_subject
  for (nm in names(changes)) covs[[nm]] <- changes[[nm]]
  tibble(
    id = id_offset + seq_len(nrow(eta)),
    sid = seq_len(nrow(eta)),
    individual = label
  ) |>
    bind_cols(covs[rep(1, nrow(eta)), ], as_tibble(eta))
}

make_events <- function(model_name, labels, seed) {
  ui <- rxode2::rxode(readModelDb(model_name))
  omega <- ui$omega
  set.seed(seed)
  eta <- MASS::mvrnorm(n_per_individual, rep(0, nrow(omega)), omega)
  colnames(eta) <- rownames(omega)
  subjects <- bind_rows(lapply(seq_along(labels), function(i) {
    make_individual(labels[i], individuals[[labels[i]]], eta,
      id_offset = (i - 1L) * n_per_individual
    )
  }))
  bind_rows(
    subjects |> tidyr::crossing(time = (seq_len(n_doses) - 1) * tau) |>
      mutate(evid = 1L, amt = 10),
    subjects |> tidyr::crossing(time = obs_times) |>
      mutate(evid = 0L, amt = 0)
  ) |>
    mutate(cmt = "depot") |>
    arrange(id, time, desc(evid))
}

olz_labels <- setdiff(names(individuals), "Tablet: samidorphan alone")
sam_labels <- c(
  "Reference", "Weight: low (44 kg)", "Weight: high (141 kg)",
  "Rifampin: presence", "Hepatic impairment: moderate",
  "Renal impairment: severe", "Food: fed", "Tablet: samidorphan alone"
)
events_olz <- make_events("Sun_2021_olanzapine", olz_labels, seed = 1430)
events_sam <- make_events("Sun_2021_samidorphan", sam_labels, seed = 1441)
stopifnot(
  !anyDuplicated(unique(events_olz[, c("id", "time", "evid")])),
  !anyDuplicated(unique(events_sam[, c("id", "time", "evid")]))
)
```

# Simulation

```{r simulate}
simulate_model <- function(model_name, events) {
  mod <- rxode2::zeroRe(rxode2::rxode(readModelDb(model_name)))
  rxode2::rxSolve(mod,
    events = events, keep = c("individual", "sid"),
    returnType = "data.frame"
  ) |>
    filter(time >= t_ss) |>
    mutate(tad = time - t_ss)
}
sim_olz <- simulate_model("Sun_2021_olanzapine", events_olz)
sim_sam <- simulate_model("Sun_2021_samidorphan", events_sam)
```

## Steady-state profiles of the reference subject (compare Figure 2)

Sun 2021 Figure 2 shows prediction-corrected VPCs over a 24-hour dosing
interval against the observed data, which are not public. The simulated
steady-state interval for the reference subject is shown below for orientation.

```{r figure-2, fig.width = 7, fig.height = 3.5}
bind_rows(
  sim_olz |> filter(individual == "Reference") |> mutate(drug = "Olanzapine"),
  sim_sam |> filter(individual == "Reference") |> mutate(drug = "Samidorphan")
) |>
  group_by(drug, tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~drug, scales = "free_y") +
  labs(
    x = "Time after dose at steady state (h)", y = "Concentration (ng/mL)",
    title = "Reference subject, OLZ/SAM 10 mg/10 mg once daily",
    caption = "Median and 5th-95th percentiles of 200 simulated subjects; compare Figure 2 of Sun 2021."
  )
```

# PKNCA validation

Steady-state Cmax,ss and AUCtau are computed with PKNCA over the 28th dosing
interval for every covariate individual.

```{r pknca}
run_nca <- function(sim, events) {
  conc <- sim |>
    filter(!is.na(Cc)) |>
    select(id, time, Cc, individual)
  doses <- events |>
    filter(evid == 1) |>
    select(id, time, amt, individual)
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | individual + id)
  dose_obj <- PKNCA::PKNCAdose(doses, amt ~ time | individual + id)
  intervals <- data.frame(start = t_ss, end = t_ss + tau, cmax = TRUE, auclast = TRUE)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}
nca_olz <- run_nca(sim_olz, events_olz)
nca_sam <- run_nca(sim_sam, events_sam)
```

## Reference-subject exposure against the Figure 3 footnotes

The Figure 3 footnotes give the reference-subject steady-state exposure:
olanzapine Cmax,ss 31.7 ng/mL and AUCtau 635 ng*h/mL; samidorphan Cmax,ss
33.4 ng/mL and AUCtau 284 ng*h/mL. The table compares the median of the 200
simulated reference subjects.

```{r nca-compare}
published_ref <- tibble::tribble(
  ~individual, ~cmax, ~auclast,
  "Reference", 31.7, 635
)
cmp_olz <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(nca_olz) |> filter(individual == "Reference"),
  reference = published_ref,
  by = "individual",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_olz, caption = "Olanzapine reference subject: simulated vs published (Figure 3 footnotes a, b). * differs by >20%.")
```

```{r nca-compare-sam}
published_ref_sam <- tibble::tribble(
  ~individual, ~cmax, ~auclast,
  "Reference", 33.4, 284
)
cmp_sam <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(nca_sam) |> filter(individual == "Reference"),
  reference = published_ref_sam,
  by = "individual",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_sam, caption = "Samidorphan reference subject: simulated vs published (Figure 3 footnotes c, d). * differs by >20%.")
```

```{r reference-gate}
ref_median <- function(nca, param) {
  as.data.frame(nca) |>
    filter(individual == "Reference", PPTESTCD == param) |>
    pull(PPORRES) |>
    median()
}
# The typical-value AUCtau is Dose / CL exactly: 10 mg / 15.5 L/h = 645 and
# 10 mg / 35.4 L/h = 282 ng*h/mL, 1.6% above and 0.5% below the published
# values. A mis-transcribed CL/F, dose unit or the ng/mL scaling moves these
# by tens of percent.
stopifnot(
  abs(ref_median(nca_olz, "auclast") / 635 - 1) < 0.05,
  abs(ref_median(nca_olz, "cmax") / 31.7 - 1) < 0.08,
  abs(ref_median(nca_sam, "auclast") / 284 - 1) < 0.05,
  abs(ref_median(nca_sam, "cmax") / 33.4 - 1) < 0.08
)
```

# Replicate Figure 3: covariate effects on steady-state exposure

Each covariate individual's geometric-mean exposure is divided by that of the
reference subject (a within-subject ratio, since all individuals share the
same random effects). The published ratios are read from the value labels
printed in Sun 2021 Figure 3.

```{r figure-3-table}
published_ratios <- tibble::tribble(
  ~drug, ~individual, ~cmax, ~auclast,
  "Olanzapine", "Weight: low (44 kg)", 1.46, 1.43,
  "Olanzapine", "Weight: high (141 kg)", 0.59, 0.61,
  "Olanzapine", "Age: low (18 y)", 1.04, 0.99,
  "Olanzapine", "Age: high (73 y)", 0.97, 1.01,
  "Olanzapine", "Rifampin: presence", 0.65, 0.58,
  "Olanzapine", "Smoking: smoker", 0.83, 0.78,
  "Olanzapine", "Hepatic impairment: moderate", 1.12, 1.16,
  "Olanzapine", "Renal impairment: severe", 1.21, 1.25,
  "Olanzapine", "Food: fed", 0.98, 0.99,
  "Olanzapine", "Race: Black", 0.96, 0.94,
  "Olanzapine", "Sex: female", 1.14, 1.18,
  "Samidorphan", "Weight: low (44 kg)", 1.49, 1.39,
  "Samidorphan", "Weight: high (141 kg)", 0.52, 0.60,
  "Samidorphan", "Rifampin: presence", 0.77, 0.36,
  "Samidorphan", "Hepatic impairment: moderate", 1.05, 1.23,
  "Samidorphan", "Renal impairment: severe", 1.25, 1.79,
  "Samidorphan", "Food: fed", 0.71, 1.01,
  "Samidorphan", "Tablet: samidorphan alone", 0.98, 1.01
) |>
  pivot_longer(c(cmax, auclast), names_to = "PPTESTCD", values_to = "published")

gm_ratio <- function(nca, drug) {
  res <- as.data.frame(nca) |>
    filter(PPTESTCD %in% c("cmax", "auclast")) |>
    select(individual, id, PPTESTCD, PPORRES)
  sid_map <- res |>
    distinct(individual, id) |>
    group_by(individual) |>
    mutate(sid = rank(id)) |>
    ungroup()
  res <- res |> left_join(sid_map, by = c("individual", "id"))
  ref <- res |>
    filter(individual == "Reference") |>
    select(sid, PPTESTCD, ref = PPORRES)
  res |>
    filter(individual != "Reference") |>
    left_join(ref, by = c("sid", "PPTESTCD")) |>
    group_by(individual, PPTESTCD) |>
    summarise(simulated = exp(mean(log(PPORRES / ref))), .groups = "drop") |>
    mutate(drug = drug)
}

ratios <- bind_rows(
  gm_ratio(nca_olz, "Olanzapine"),
  gm_ratio(nca_sam, "Samidorphan")
) |>
  inner_join(published_ratios, by = c("drug", "individual", "PPTESTCD")) |>
  mutate(
    parameter = ifelse(PPTESTCD == "cmax", "Cmax,ss", "AUCtau"),
    pct_diff = 100 * (simulated / published - 1)
  )

ratios |>
  select(drug, individual, parameter, simulated, published, pct_diff) |>
  arrange(drug, parameter, individual) |>
  dplyr::rename(
    "Drug" = drug, "Covariate individual" = individual,
    "Exposure" = parameter, "Simulated ratio" = simulated,
    "Published ratio (Figure 3)" = published, "% difference" = pct_diff
  ) |>
  knitr::kable(digits = c(0, 0, 0, 3, 2, 1), caption = "Fold change in steady-state exposure relative to the reference subject.")
```

```{r figure-3, fig.width = 7, fig.height = 7}
ratios |>
  ggplot(aes(y = individual)) +
  geom_vline(xintercept = 1) +
  geom_vline(xintercept = c(0.8, 1.25), linetype = "dashed") +
  geom_point(aes(x = published, shape = "Published (Figure 3)"), size = 2.5) +
  geom_point(aes(x = simulated, shape = "Simulated"), size = 2) +
  scale_shape_manual(values = c("Published (Figure 3)" = 1, "Simulated" = 16)) +
  facet_grid(drug ~ parameter, scales = "free_y", space = "free_y") +
  labs(
    x = "Fold change relative to the reference subject", y = NULL, shape = NULL,
    caption = "Replicates Figure 3 of Sun 2021."
  ) +
  theme(legend.position = "bottom")
```

```{r figure-3-gate}
# The simulation uses the same R-drawn random effects on every machine, so
# these ratios are reproducible and the bounds can sit just outside the
# observed differences: at most 4.8% for AUCtau (olanzapine food) and 5.7%
# for Cmax,ss (samidorphan rifampin). The paper's own ratios come from
# independent 500-subject draws per individual and scatter by a few percent
# around the model-implied values (e.g. 1 / 1.80 = 0.556 against the printed
# 0.58 for olanzapine rifampin AUCtau). A mis-transcribed covariate
# coefficient moves its ratio by 10% or more.
stopifnot(
  nrow(ratios) == nrow(published_ratios),
  all(abs(ratios$pct_diff[ratios$PPTESTCD == "auclast"]) < 6),
  all(abs(ratios$pct_diff[ratios$PPTESTCD == "cmax"]) < 8)
)
```

The simulated ratios reproduce every Figure 3 value within 6%. The two largest
differences are the olanzapine food effect on AUCtau (model-implied 0.943,
exactly the Table 3 relative bioavailability, against the printed 0.99) and
the samidorphan rifampin effect on Cmax,ss (0.81 against the printed 0.77).
Both are consistent with the sampling noise of the paper's independent
500-subject simulations; the published 90% prediction interval for the
olanzapine food effect, 0.95-1.03, sits just above the value its own Table 3
coefficient implies.

# Assumptions and deviations

- **Covariate centering.** The Methods state that continuous covariates were
  centered at "typical values" without giving them. Body weight is centered
  at 70 kg and age at 36 years, the reference subject of Figure 3. The
  reference-subject AUCtau printed in Figure 3 confirms the 70 kg centering:
  Dose / CL/F = 645 (olanzapine) and 282 (samidorphan) ng*h/mL against the
  printed 635 and 284, whereas centering at the data-set median weights would
  give 694 and 301 ng*h/mL.
- **Olanzapine inter-occasion variability.** Table 3 reports one IOV variance
  on ka (0.630) but neither the occasion definition nor the number of
  occasions. Three occasions are encoded through the `OCC` column, the largest
  number of separate PK sampling occasions per subject among the pooled
  studies (Table S1); occasions 2 and 3 carry the same variance. Records with
  `OCC` outside 1-3 receive no IOV.
- **Samidorphan study ALK3831-A305 lag-time factor.** Table 4 marks
  'Change in ALAG' (10.1) with footnote b, 'Fixed at estimate from previous
  stable model', but also prints an RSE (11.0%) and a 95% CI; Table 2 does not
  mark it as fixed and the Results name only the Q/F variability as fixed in
  the samidorphan model. It is therefore encoded as estimated. The factor
  compensates for imputed dose times in that study (Table 4 footnote c) and
  should be left at `STUDY_ALK3831A305 = 0` for simulation.
- **Smoking reference level.** The olanzapine smoking effect contrasts smokers
  with nonsmokers pooled with subjects whose smoking status was not recorded
  (Table 2; Discussion), so `SMOKE = 0` covers both.
- **Renal and hepatic impairment.** Both indicators were informed by the
  dedicated organ-impairment studies (ALK3831-A105 and ALK3831-A106); mild and
  moderate renal impairment and other hepatic categories were not retained.
- **No IIV covariances** are reported in Tables 3 and 4, so all random effects
  are independent.
- **Units.** Table 3 and Table 4 print the ka unit as 'h'; ka is a first-order
  rate constant and is encoded in 1/h.
- No erratum or correction notice for Sun 2021 was found (Crossref, checked
  2026-09-29).
