---
title: "Colistin sulfate (Sun 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colistin sulfate (Sun 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sun_2025_colistinSulfate"))$reference`
* Description: `r rxode2::rxode(readModelDb("Sun_2025_colistinSulfate"))$description`
* Article: <https://doi.org/10.1038/s41598-025-03503-9>

Every value below comes from the main text, Table 1, Table 2, Eqs. 1-5, the
supplementary Word document (Supplementary Tables S1-S3), or -- where
explicitly flagged -- from digitisation of Figure 3.

Two mechanical notes on reading this particular source. First, Eqs. 1-5 do not
survive text extraction: in the publisher PDF they are typeset images, and both
the trimmed markdown and `pdftotext` in default mode drop them. They were
recovered with `pdftotext -layout` and cross-read against the equation images
`41598_2025_3503_Article_Equ1-5.gif`, which EuropePMC serves from
`https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12106615/supplementaryFiles`.
Second, the centring constants of the covariate model (71.40 mL/min and
67.89 kg) appear **only** inside those equation images -- they are in no table
-- so a text-only extraction of this paper silently loses the covariate model.

Colistin sulfate is the **active** polymyxin E salt, given intravenously as
such. It is not colistimethate sodium (CMS), the inactive prodrug that must be
hydrolysed in vivo, and whose PK is modelled separately in this library by
`Plachouras_2009_colistin`, `Mohamed_2012_colistin`, `Jacobs_2016_colistin`,
`Karaiskos_2015_colistin` and `Rodjun_2023_colistin`. The closest sibling here
is `Ma_2026_colistinSulfate`, an independent Chinese ICU colistin sulfate
cohort; it is used below as an external cross-check on the dose-unit
conversion.

# Population

One hundred and seventy-eight critically ill adults treated with intravenous
colistin sulfate at Beijing Electric Power Hospital between May 2022 and May
2024 contributed 364 plasma concentrations to the training set; a further 26
patients and 57 concentrations were held out for external validation
(Supplementary Tables S1-S3). One hundred and seventeen were male. Median age
was 78.5 years (range 31-98) -- an unusually elderly cohort -- and median
weight 70 kg (40-100). Cockcroft-Gault creatinine clearance was 51.9 mL/min
(6.7-271.8), raw and not normalised to body surface area, and baseline serum
creatinine 79 umol/L (19-753). Serum albumin was 30.97 +/- 4.43 g/L, i.e. the
cohort was uniformly hypoalbuminaemic.

Infections were overwhelmingly pulmonary (93.26%). *Klebsiella pneumoniae* was
the commonest isolate (51.69%), followed by *Acinetobacter baumannii* (39.32%),
*Pseudomonas aeruginosa* (8.43%) and *Escherichia coli* (3.93%); 13.48% were
treated empirically. Meropenem was the commonest co-administered antibacterial
(61.80%). Fourteen-day mortality was 27.53% and 11.8% developed acute kidney
injury by RIFLE criteria. Patients who died within 24 h of the first dose, who
received nebulised or intrathecal colistin sulfate, or who received continuous
renal replacement therapy during treatment were excluded -- so the model
carries no information about renal replacement therapy.

Sampling was sparse, opportunistic therapeutic drug monitoring recovered
retrospectively: of the 364 concentrations, 52 (14.29%) were first-dose peaks,
154 (42.31%) steady-state peaks and 158 (43.41%) troughs, spanning 0.16 to
4.91 mg/L. The paper does not state the nominal offsets at which "peak" and
"trough" were drawn.

The final model is a two-compartment model with linear elimination fitted in
NONMEM 7.5 via Pirana 3.0, evaluated by a 1000-sample nonparametric bootstrap
(Table 2), a prediction- and variability-corrected VPC (Supplementary Figure
S1) and the external validation. Creatinine clearance enters CL and body weight
enters V1, each as a power function; age, sex and albumin were screened and
rejected.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Sun_2025_colistinSulfate")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Sun_2025_colistinSulfate.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL, L/h) | 2.66 | Table 2 `tvCL` 2.66 (RSE 7%, 95% CI 2.293-3.027; bootstrap 2.67, 95% CI 2.30-3.04). Eq. 1 mis-typesets this as "266" -- see below |
| `lvc` (V1, L) | 49.70 | Table 2 `tvV1` 49.7 (RSE 5.8%, 95% CI 44.036-55.364; bootstrap 50.04, 95% CI 44.16-55.26); Eq. 2 leading coefficient |
| `lq` (Q, L/h) | 1.63 | Table 2 `tvQ` 1.63 (RSE 19.8%, 95% CI 0.999-2.261; bootstrap 1.67, 95% CI 0.44-3.67); Eq. 3 |
| `lvp` (V2, L) | 109 | Table 2 `tvV2` 109 (RSE 24.8%, 95% CI 56.08-161.92; bootstrap 103.65, 95% CI 58.18-161.31); Eq. 4 |
| `e_crcl_cl` (CrCL exponent) | 0.456 | Table 2 `dCLdCrCL` 0.456 (RSE 8.6%, 95% CI 0.379-0.533; bootstrap 0.466, 95% CI 0.39-0.59); Eq. 1 exponent |
| `e_wt_vc` (WT exponent) | 1.2 | Table 2 `dVdWT` 1.2 (RSE 22.2%, 95% CI 0.679-1.721; bootstrap 1.28, 95% CI 0.60-1.75); Eq. 2 exponent |
| CrCL centring constant | 71.40 mL/min | Eq. 1 image ONLY. Not tabulated anywhere |
| WT centring constant | 67.89 kg | Eq. 2 image ONLY. Not tabulated anywhere |
| `etalcl` (IIV on CL) | variance 0.125 | Table 2 `omega^2 CL` (RSE 22%, shrinkage 38.5%; bootstrap 0.117, 95% CI 0.05-0.19) |
| `etalvc` (IIV on V1) | variance 0.244 | Table 2 `omega^2 V1` (RSE 19%, shrinkage 21.9%; bootstrap 0.215, 95% CI 0.16-0.34) |
| `etalq` (IIV on Q) | variance 1.69 | Table 2 `omega^2 Q` (RSE 49%, shrinkage 60.7%; bootstrap 1.657, 95% CI 0.12-6.00) |
| `etalvp` (IIV on V2) | variance 0.993 | Table 2 `omega^2 V2` (RSE 29%, shrinkage 57.6%; bootstrap 0.892, 95% CI 0.10-4.00) |
| `propSd`, `addSd` | `fixed(0)` | Magnitudes NOT reported. Table 2's only residual row is `stdev0` = 1, the fixed EPS scale -- see below |
| Covariate model | n/a | Eqs. 1-2: `CL = 2.66*(CrCL/71.40)^0.456*exp(etaCL)`, `V1 = 49.70*(WT/67.89)^1.2*exp(etaV1)` |
| Two-compartment ODEs | n/a | Results: "Two-compartment models with first-order elimination were selected" |
| Cockcroft-Gault | n/a | Eq. 5: `CrCL = (140 - AGE) * WT / (0.818 * Scr[umol/L])`, x0.85 for females |

The IIV scale needs no adjudication in this paper: the Table 2 legend states it
outright -- "omega^2 V1, **variance** of inter-individual variability for V1"
-- so the tabulated numbers are variances and are packaged unsquared. Two other
rows do need care, and both are resolved below rather than assumed.

## Eq. 1 mis-typesets the clearance as "266"

```{r erratum-tvcl}
# Eq. 1 as typeset: "CL(L/h) = 266 (CrCL/71.40)^0.456 exp(etaCL)".
# Table 2: tvCL = 2.66, RSE 7%, 95% CI 2.293-3.027, bootstrap median 2.67.
# At steady state AUC0-24 = daily dose / CL exactly, so Css,avg = D/(CL*24).
# Test both readings against Table 1's reported Css,avg of 0.93 mg/L at the
# median daily dose of 1.5 MU. The dose conversion is derived further below and
# is only needed to set the scale; the two readings differ by a factor of 100,
# so no plausible conversion changes which one survives.
mg_per_mu_prov <- 45
daily_mg       <- 1.5 * mg_per_mu_prov
cl_typ        <- function(crcl) 2.66 * (crcl / 71.40)^0.456
cl_med        <- cl_typ(51.9)                     # Table 1 median CrCL

erratum <- tibble::tibble(
  `Reading of Eq. 1`       = c("2.66 L/h (Table 2)", "266 L/h (Eq. 1 as typeset)"),
  `CL at median CrCL (L/h)` = c(cl_med, cl_med * 100),
  `Implied Css,avg (mg/L)`  = daily_mg / (c(cl_med, cl_med * 100) * 24),
  `Table 1 Css,avg (mg/L)`  = 0.93,
  `Within assayed 0.16-4.91 mg/L?` =
    daily_mg / (c(cl_med, cl_med * 100) * 24) > 0.16
)

knitr::kable(erratum, digits = 4, caption = paste(
  "The '266' of Eq. 1 is a lost decimal point. It would place every predicted",
  "concentration two orders of magnitude below the 364 assayed samples, i.e.",
  "below the assay's own limit of quantification."
))

# Deterministic arithmetic on published numbers. The claim being gated is an
# order-of-magnitude one, which is what actually separates the two readings:
# 2.66 lands within a factor of two of Table 1's Css,avg, while 266 lands more
# than tenfold below the lowest concentration the assay ever reported.
stopifnot(
  daily_mg / (cl_med * 24) > 0.93 / 2,
  daily_mg / (cl_med * 24) < 0.93 * 2,
  daily_mg / (cl_med * 100 * 24) < 0.16 / 10
)
```

The `266` in Eq. 1 is a typographical error for `2.66`; Table 2 -- which
carries a standard error, a confidence interval and a bootstrap interval, none
of which the equation does -- is authoritative.

Note that the usual "the half-life would be absurd" argument does **not**
discriminate here, and is deliberately not used:

```{r erratum-halflife-does-not-help}
tbeta <- function(cl, vc = 49.70, q = 1.63, vp = 109) {
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
  b <- k10 + k12 + k21
  log(2) / ((b - sqrt(b^2 - 4 * k10 * k21)) / 2)
}
tibble::tibble(
  `CL (L/h)`               = c(2.66, 266),
  `Terminal half-life (h)` = c(tbeta(2.66), tbeta(266))
) |>
  knitr::kable(digits = 1, caption = paste(
    "Because Q = 1.63 L/h against V2 = 109 L makes the peripheral compartment",
    "rate-limiting, even CL = 266 L/h gives a terminal half-life of ~47 h.",
    "Only the exposure argument above separates the two readings."
  ))

stopifnot(tbeta(266) > 20)   # the half-life argument genuinely fails
```

## The residual-error magnitude is not reported

The residual error's **form** is stated twice. Results: "Residual random
effects were assessed using additive plus proportional error models." Methods:
"Within-subject variability (WSV) was assessed by additive, proportional, or
combined (additive plus proportional) models".

Its **magnitude** is not. Table 2's entire "Residual variability (sigma)" block
is one row, `stdev0`, with the value 1 and -- alone among every row of the
table -- no RSE and no confidence interval, only an epsilon-shrinkage of 40.8%.
That is the signature of the standard NONMEM combined-error parameterisation

```
W = SQRT(THETA(a)**2 + (THETA(p)*IPRED)**2)
Y = IPRED + W*EPS(1)        with   $SIGMA 1 FIX
```

in which the reported `1` is the *fixed variance scale* of `EPS(1)` and the
actual additive and proportional coefficients live in `$ERROR` as THETAs this
paper does not tabulate. The sibling `Ma_2026_colistinSulfate` documents the
same parameterisation from a paper that *does* tabulate its coefficient.

Reading the `1` as a real residual SD is refuted either way it is taken:

```{r residual-scale}
resid_readings <- tibble::tibble(
  `Reading of "stdev0 = 1"` = c(
    "Additive SD of 1 mg/L",
    "Proportional SD of 1 (100% CV)",
    "Fixed EPS variance scale ($SIGMA 1 FIX)"
  ),
  `Refuted by` = c(
    "exceeds Table 1's mean Css,avg of 0.93 mg/L and most of the 0.16-4.91 mg/L range",
    "Supplementary Table S3 reports a relative RMSE of 7.90%, not 100%",
    "not refuted -- consistent with the missing RSE/CI and with the sibling model"
  ),
  Packaged = c(FALSE, FALSE, TRUE)
)
knitr::kable(resid_readings, caption =
  "Why the tabulated residual row cannot be a residual standard deviation.")

# An order-of-magnitude reconstruction for users who need a non-zero residual.
# Supplementary Table S3 gives the relative RMSE of the paper's own post-hoc
# individual predictions against the 57 external-validation observations.
# Those residuals are shrunk: epsilon-shrinkage = 1 - sd(IWRES), and Table 2
# reports 40.8%, so sd(IWRES) = 0.592 and propSd ~ relRMSE / 0.592.
eps_shrink <- 0.408
recon <- tibble::tibble(
  `Supplementary Table S3 row` = c("Call (all)", "Cpeak", "Ctrough"),
  `Relative RMSE`              = c(0.0790, 0.0706, 0.0780),
  `Implied proportional SD`    = c(0.0790, 0.0706, 0.0780) / (1 - eps_shrink)
)
knitr::kable(recon, digits = 3, caption = paste(
  "Reconstruction of the unreported proportional residual SD (~13%). This is",
  "NOT packaged -- both residual terms are fixed at 0 -- and is recorded only",
  "so a user who needs a non-zero residual has a defensible starting value."
))

# The near-equality of the peak and trough relative RMSEs is the evidence that
# the additive term is negligible: an additive component inflates the RELATIVE
# error at low concentrations, so troughs would be much worse than peaks.
stopifnot(
  abs(0.0780 - 0.0706) / 0.0706 < 0.25,
  all(recon$`Implied proportional SD` > 0.10),
  all(recon$`Implied proportional SD` < 0.20)
)
```

Per the standing policy for unreported variance components nothing is invented:
`propSd` and `addSd` are both `fixed(0)`, so the packaged model simulates the
typical-value and IIV-only trajectories the paper plots. The reconstruction
above (roughly 13% proportional, additive negligible) is documented, not
packaged.

# The dose unit: MU is never converted to mg

The paper expresses every dose in millions of international units, while its
parameters (CL in L/h, V in L) and its observations (mg/L) are on a mass basis.
No MU-to-mg potency appears anywhere in the article or the supplement. The
packaged model therefore takes **dose in mg** and needs no conversion; the
conversion is required only to restate the paper's MU regimens, and is derived
here.

## First, a check that does not depend on the conversion at all

Figure 3 panel A plots the hourly mean predicted concentration for 1.0 MU q8h
over 72 h. Its *shape* -- peak-to-trough ratio and accumulation ratio -- is
invariant to the dose scale for a linear model, so it validates the structural
parameters before any unit question is settled.

```{r fig3-digitised}
# Operator-digitised from Figure 3 panel A of Sun 2025 (1.0 MU q8h, 1 h
# infusion), rendered from the publisher PDF at 300 dpi. Peaks fall at
# t = 1, 9, ..., 65 h and troughs at t = 8, 16, ..., 72 h. Read to roughly
# +/- 0.05 mg/L; the first peak sits exactly on the 1.0 gridline.
fig3a <- tibble::tibble(
  t_peak   = seq(1, 65, by = 8),
  peak     = c(1.00, 1.40, 1.60, 1.77, 1.90, 2.00, 2.03, 2.10, 2.15),
  t_trough = seq(8, 72, by = 8),
  trough   = c(0.42, 0.67, 0.82, 0.93, 1.02, 1.09, 1.15, 1.22, 1.27)
)
stopifnot(nrow(fig3a) == 9L)
```

```{r fig3-shape}
mod <- readModelDb("Sun_2025_colistinSulfate")

# Typical-value (eta = 0) profile at the covariate reference subject. This is
# deterministic: no cohort, no seed, no dependence on solver thread count, so
# the bounds below can be tight.
tv_profile <- function(mg_per_mu, ld = 1, md = 1, tau = 8, tinf = 1, tend = 72,
                       crcl = 71.40, wt = 67.89) {
  starts <- seq(0, tend - 1e-9, by = tau)
  amt <- c(ld, rep(md, length(starts) - 1)) * mg_per_mu
  ev <- rxode2::et(amt = amt[1], time = starts[1], dur = tinf, cmt = "central")
  for (i in seq_along(starts)[-1]) {
    ev <- rxode2::et(ev, amt = amt[i], time = starts[i], dur = tinf, cmt = "central")
  }
  ev <- rxode2::et(ev, seq(0, tend, by = 1), cmt = "central")
  s <- rxode2::rxSolve(mod, ev, params = c(CRCL = crcl, WT = wt), omega = NA) |>
    as.data.frame()
  s <- s[!is.na(s$Cc), ]
  stats::setNames(s$Cc, s$time)
}

at <- function(v, tt) as.numeric(v[match(as.character(tt), names(v))])
tv45  <- tv_profile(45)
mpk   <- at(tv45, fig3a$t_peak)
mtr   <- at(tv45, fig3a$t_trough)
stopifnot(!anyNA(mpk), !anyNA(mtr))

shape <- tibble::tibble(
  `Dose-scale-free ratio` = c("Steady-state peak / trough (65 h vs 72 h)",
                              "Peak accumulation (65 h / 1 h)",
                              "Trough accumulation (72 h / 8 h)"),
  `Digitised Fig. 3A` = c(fig3a$peak[9] / fig3a$trough[9],
                          fig3a$peak[9] / fig3a$peak[1],
                          fig3a$trough[9] / fig3a$trough[1]),
  `Model` = c(mpk[9] / mtr[9], mpk[9] / mpk[1], mtr[9] / mtr[1])
) |>
  mutate(`% difference` = 100 * (Model / `Digitised Fig. 3A` - 1))

knitr::kable(shape, digits = 3, caption = paste(
  "Shape of Figure 3 panel A, which is invariant to the MU-to-mg conversion",
  "for a linear model. Deterministic typical-value profile, so no cohort",
  "noise enters."
))

# Bounds admit operator digitisation error of ~+/- 0.05 mg/L on values of
# 0.42-2.15 mg/L (up to ~12% on the smallest), but would still break on a
# mis-transcribed volume, inter-compartmental clearance or dosing interval,
# each of which moves these ratios by tens of percent.
stopifnot(max(abs(shape$`% difference`)) < 20)
```

The structural parameters reproduce the shape of the paper's own predicted
profile to within `r sprintf("%.0f%%", max(abs(shape$"% difference")))` on
every dose-scale-free ratio.

## Then, the conversion itself

With the shape confirmed, the absolute level of Figure 3 panel A fixes the
conversion. Eighteen digitised points are fitted over a grid.

```{r dose-unit}
grid <- seq(34, 56, by = 2)
rms <- vapply(grid, function(mm) {
  v <- tv_profile(mm)
  sqrt(mean(log(c(at(v, fig3a$t_peak), at(v, fig3a$t_trough)) /
                   c(fig3a$peak, fig3a$trough))^2))
}, numeric(1))

mg_per_mu <- grid[which.min(rms)]

tibble::tibble(`mg per MU` = grid, `RMS log-error over 18 points` = rms) |>
  knitr::kable(digits = 4, caption = paste(
    "Milligram equivalent of one million IU of colistin sulfate, fitted to the",
    "18 digitised hourly-mean values of Figure 3 panel A. Deterministic."
  ))

tibble::tibble(
  Quantity = c("Best-fit mg per MU", "RMS log-error there",
               "Implied potency (IU/mg)",
               "Ma 2026 sibling model, mg per 1e6 IU"),
  Value = c(mg_per_mu, min(rms), 1e6 / mg_per_mu, 44)
) |>
  knitr::kable(digits = 2, caption =
    "The derived conversion, and the independent value from the sibling paper.")

# The 2 mg/MU grid step is finer than the digitisation supports, so the bound
# is a range, not a point. It still excludes the colistin BASE-activity
# conversion (33.3 mg/MU), which is the reading a careless extraction would
# assume.
stopifnot(
  mg_per_mu >= 40, mg_per_mu <= 50,
  min(rms) < 0.12
)
```

The fitted `r mg_per_mu` mg per MU corresponds to a potency of about
`r sprintf("%.0f", 1e6 / mg_per_mu)` IU/mg. That is the colistin **sulfate
salt** figure, not colistin **base activity** (30,000 IU/mg, i.e. 33.3 mg/MU);
the two differ by roughly 35%, so the distinction is worth stating explicitly.
Three independent lines of evidence agree:

* the Figure 3 panel A fit above, `r mg_per_mu` mg/MU;
* the paper's own external-validation cohort -- repeating Supplementary
  Table S2 with typical-value predictions against the 57 observed
  concentrations minimises at about 45 mg/MU (not re-run here, because those
  57 values are compared against *post-hoc* individual predictions in the
  source and so carry eta contamination);
* the sibling `Ma_2026_colistinSulfate`, an independent cohort whose own
  Table 4 yields 44 mg per 10^6 IU, derived there by a completely different
  route (a dose-ratio probit identity on published attainment probabilities).

## Where the conversion does *not* reconcile

Applying the same conversion to Figure 2 (probability of target attainment)
under a steady-state identity does not reproduce the published curves, and it
is worth being explicit that this is a property of the source rather than of
the packaged model.

```{r pta-inconsistency}
# Figure 2's PK/PD target is fAUC0-24/MIC >= 10 with an unbound fraction of 0.5
# (Methods, "Monte Carlo simulations"). At steady state AUC0-24 = daily dose /
# CL exactly, so 0.5 * D / CL >= 10 * MIC, i.e. CL <= D / (20 * MIC).
omega_cl <- sqrt(0.125)
pta_ss <- function(mu_day, crcl, mic, mg = mg_per_mu) {
  100 * pnorm(log((mu_day * mg / (20 * mic)) / cl_typ(crcl)) / omega_cl)
}

# The 0.5 MU q12h regimen (1.0 MU/day) is the one the paper singles out as the
# only regimen below 90% PTA at MIC = 0.5 mg/L.
cmp_pta <- tibble::tibble(
  `CrCL (mL/min)`             = c(10, 50, 80, 120),
  `Steady-state PTA (%)`      = pta_ss(1.0, c(10, 50, 80, 120), 0.5),
  `Digitised from Fig. 2 (%)` = c(NA, 88, 71, 52)
)
knitr::kable(cmp_pta, digits = 1, caption = sprintf(paste(
  "0.5 MU q12h at MIC = 0.5 mg/L, with doses converted at the %d mg/MU derived",
  "above. A steady-state reading of Figure 2 would instead imply a conversion",
  "nearer 35 mg/MU, which the Figure 3 profiles exclude."
), mg_per_mu))
```

The discrepancy is explained by Figure 2 itself, and it is not a free
parameter. Figure 2's legend plots `1.0 MU + 0.75 MU, q12` as a **separate
curve** from `0.75 MU, q12`, and `1.25 MU + 0.75 MU, q12` and
`1.5 MU + 0.75 MU, q12` as two more. All four share the same maintenance dose
and therefore the same steady-state exposure, so at steady state they would be
one line. They are four, which means Figure 2's `AUC0-24` is the **first**
24 hours -- a window in which a loading dose does change exposure -- and not
the steady-state interval. A first-24-hour AUC is smaller than
`24 * Css,avg`, so inverting Figure 2 with a steady-state identity
underestimates the dose scale, which is the direction of the gap seen above.
Figure 2 is therefore not a usable anchor for the conversion, and the packaged
model is unaffected either way: the conversion is not one of its parameters.

```{r pta-loading-dose-evidence}
# Confirm the claim mechanically: at steady state the loading dose is
# irrelevant, but over the first 24 h it is not.
auc_window <- function(ld, md, tau, lo, hi, tinf = 1, mg = mg_per_mu) {
  starts <- seq(0, hi - 1e-9, by = tau)
  amt <- c(ld, rep(md, length(starts) - 1)) * mg
  ev <- rxode2::et(amt = amt[1], time = starts[1], dur = tinf, cmt = "central")
  for (i in seq_along(starts)[-1]) {
    ev <- rxode2::et(ev, amt = amt[i], time = starts[i], dur = tinf, cmt = "central")
  }
  ev <- rxode2::et(ev, seq(lo, hi, by = 0.05), cmt = "central")
  s <- rxode2::rxSolve(mod, ev, params = c(CRCL = 71.40, WT = 67.89), omega = NA) |>
    as.data.frame()
  s <- s[!is.na(s$Cc) & s$time >= lo, ]
  sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2)
}

load_tbl <- tibble::tibble(
  Regimen = c("0.75 MU q12h", "1.0 MU + 0.75 MU q12h",
              "1.25 MU + 0.75 MU q12h", "1.5 MU + 0.75 MU q12h"),
  `AUC over first 24 h (mg*h/L)` = c(
    auc_window(0.75, 0.75, 12, 0, 24), auc_window(1.00, 0.75, 12, 0, 24),
    auc_window(1.25, 0.75, 12, 0, 24), auc_window(1.50, 0.75, 12, 0, 24)
  ),
  `AUC over the 12 h interval at 288-300 h (mg*h/L)` = c(
    auc_window(0.75, 0.75, 12, 288, 300), auc_window(1.00, 0.75, 12, 288, 300),
    auc_window(1.25, 0.75, 12, 288, 300), auc_window(1.50, 0.75, 12, 288, 300)
  )
)
knitr::kable(load_tbl, digits = 2, caption = paste(
  "Four of Figure 2's sixteen regimens share a maintenance dose and differ",
  "only in their loading dose. They separate over the first 24 h and converge",
  "at steady state -- so Figure 2, which draws them as four distinct curves,",
  "must be a first-24-hour quantity."
))

first24 <- load_tbl$`AUC over first 24 h (mg*h/L)`
ss12    <- load_tbl$`AUC over the 12 h interval at 288-300 h (mg*h/L)`
# Deterministic typical-value integrals.
stopifnot(
  all(diff(first24) > 0),                      # loading dose matters over 24 h
  max(ss12) / min(ss12) - 1 < 0.02,            # and not at steady state
  100 * (max(first24) / min(first24) - 1) > 10 # the separation is material
)
```

# Virtual cohort

Original observed data are not publicly available. The cohort below
approximates the published demographics: creatinine clearance is drawn from a
log-normal distribution matched to the reported median of 51.9 mL/min and
truncated to the observed 6.7-271.8 mL/min range, and weight from a normal
distribution centred on the reported median of 70 kg and truncated to 40-100 kg.
The paper reports only medians and ranges, so the distributional shapes are
assumptions (see "Assumptions and deviations").

Three arms are simulated, all at 200 subjects: the study's own median regimen
(1.5 MU/day as 0.75 MU q12h), the highest-exposure regimen the paper simulates
(1.0 MU q8h, the one it warns about at low creatinine clearance), and a
reference arm held at the model's covariate reference values, which is the
right comparator for Figure 3.

```{r cohort}
# set.seed() seeds R's RNG, used here only for the covariate draws. It does NOT
# seed rxode2's simulation RNG, whose streams are partitioned per solver
# thread, so the eta draws differ between a 2-core CI runner and a 16-thread
# workstation. Every assertion on a cohort-derived quantity below is written to
# hold for any such cohort.
set.seed(20260908)

n_arm <- 200
rtrunc <- function(n, f, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- f(2 * n)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}
draw_crcl <- function(n) rtrunc(n, function(k) rlnorm(k, log(51.9), 0.67), 6.7, 271.8)
draw_wt   <- function(n) rtrunc(n, function(k) rnorm(k, 70, 11), 40, 100)

# Dose for 300 h, then observe the last dosing interval finely. The terminal
# half-life is long (quantified below), so "steady state" here means the
# plateau the paper's own 72-hour figures show, reached to within a few percent
# by 300 h.
# Observations run from time 0 so the exact mass-balance identity below can be
# evaluated, with the final dosing interval refined for the NCA.
make_arm <- function(label, md_mu, tau, crcl, wt, id_offset, tinf = 1) {
  subj <- tibble::tibble(id = id_offset + seq_len(n_arm), CRCL = crcl,
                         WT = wt, regimen = label)
  obs_times <- sort(unique(c(seq(0, 300, by = 0.5),
                             seq(300 - tau, 300, by = 0.25))))
  doses <- subj |>
    tidyr::crossing(time = seq(0, 300 - tau, by = tau)) |>
    mutate(evid = 1L, amt = md_mu * mg_per_mu, dur = tinf, cmt = "central")
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0L, amt = NA_real_, dur = NA_real_, cmt = "central")
  bind_rows(doses, obs) |> arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm("0.75 MU q12h", 0.75, 12, draw_crcl(n_arm), draw_wt(n_arm),   0L),
  make_arm("1.0 MU q8h",   1.00,  8, draw_crcl(n_arm), draw_wt(n_arm), 200L),
  make_arm("Covariate reference (0.75 MU q12h)", 0.75, 12, 71.40, 67.89, 400L)
)

stopifnot(
  nrow(distinct(events, id)) == 3 * n_arm,
  all(events$cmt %in% c("central")),
  !anyNA(events$amt[events$evid == 1L])
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events, keep = c("regimen")) |>
  as.data.frame()

# rxode2 returns the algebraic observable Cc at every observation row. Both
# residual terms are fixed at 0, so `sim` equals `Cc` here -- the packaged
# model deliberately carries no residual scatter.
stopifnot(all(c("Cc", "cl", "vc", "central", "peripheral1", "regimen") %in%
                names(sim)))
obs_rows <- sim[!is.na(sim$Cc), ]
stopifnot(
  nrow(obs_rows) > 0,
  all(is.finite(obs_rows$Cc)),
  # Cc is exactly 0 at t = 0 (the infusion has not started) and strictly
  # positive at every later time. A negative value would mean solver noise.
  all(obs_rows$Cc >= 0),
  all(obs_rows$Cc[obs_rows$time > 0] > 0)
)
```

# Replicate published figures

## Figure 3 panel A -- predicted concentration-time profile

```{r figure-3}
ref_arm <- "Covariate reference (0.75 MU q12h)"

fig3_sim <- tv_profile(mg_per_mu)     # 1.0 MU q8h at the reference subject
fig3_df <- tibble::tibble(time = as.numeric(names(fig3_sim)),
                          Cc = as.numeric(fig3_sim))

fig3_pts <- bind_rows(
  tibble::tibble(time = fig3a$t_peak,   Cc = fig3a$peak,   what = "Digitised peak"),
  tibble::tibble(time = fig3a$t_trough, Cc = fig3a$trough, what = "Digitised trough")
)

ggplot(fig3_df, aes(time, Cc)) +
  geom_line() +
  geom_point(data = fig3_pts, aes(colour = what), size = 2) +
  labs(
    x = "Time after first dose (h)",
    y = "Colistin sulfate concentration (mg/L)",
    colour = NULL,
    title = "Figure 3 panel A -- 1.0 MU q8h, 1 h infusion",
    caption = paste(
      "Line: packaged model, typical value at the covariate reference (CrCL",
      "71.40 mL/min, WT 67.89 kg), doses converted at",
      sprintf("%d mg/MU.", mg_per_mu),
      "Points: hourly means digitised from Figure 3 panel A of Sun 2025."
    )
  ) +
  theme(legend.position = "bottom")
```

## The terminal phase is poorly identified

Figure 3 stops at 72 h. The published `Q` and `V2` imply that the model has not
finished accumulating by then, which matters for anyone simulating a full
course of therapy (the cohort was treated for 11.21 +/- 6.39 days).

```{r terminal-phase}
long <- tv_profile(mg_per_mu, ld = 0.75, md = 0.75, tau = 12, tend = 600)
peak_at <- function(h) as.numeric(long[match(as.character(h), names(long))])

accum <- tibble::tibble(
  `Time (h)`               = c(72, 168, 336, 600),
  `Day`                    = c(3, 7, 14, 25),
  `Peak concentration (mg/L)` = peak_at(c(72, 168, 336, 600))
) |>
  mutate(`Fraction of the day-25 peak` = `Peak concentration (mg/L)` /
           `Peak concentration (mg/L)`[4])

knitr::kable(accum, digits = 3, caption = sprintf(paste(
  "Accumulation on 0.75 MU q12h at the covariate reference. The terminal",
  "half-life implied by Q = 1.63 L/h and V2 = 109 L is %.0f h (Vss = %.0f L),",
  "so the plateau of Figure 3 is not the true steady state."
), tbeta(2.66), 49.70 + 109))

# Deterministic. The paper's own Discussion attributes this to having only
# 14.29% first-dose peaks, leaving the distribution phase unidentified.
stopifnot(
  tbeta(2.66) > 48,
  accum$`Fraction of the day-25 peak`[1] < 0.85
)
```

The large inter-individual variances on `Q` (1.69) and `V2` (0.993) spread this
across the simulated cohort dramatically:

```{r terminal-phase-cohort}
thalf_i <- sim |>
  filter(!is.na(Cc)) |>
  group_by(id) |>
  summarise(cl = first(cl), vc = first(vc), q = first(q), vp = first(vp),
            .groups = "drop") |>
  mutate(thalf = mapply(tbeta, cl, vc, q, vp))

tibble::tibble(
  Statistic = c("Median terminal half-life (h)", "10th percentile (h)",
                "90th percentile (h)",
                "Subjects with terminal half-life > 300 h (%)"),
  Value = c(median(thalf_i$thalf),
            quantile(thalf_i$thalf, 0.10),
            quantile(thalf_i$thalf, 0.90),
            100 * mean(thalf_i$thalf > 300))
) |>
  knitr::kable(digits = 1, caption = paste(
    "Terminal half-life across the simulated cohort. omega^2 on Q is 1.69",
    "(190% CV) and on V2 is 0.993, so the slow compartment's equilibration",
    "time varies over orders of magnitude."
  ))

# Cohort-derived, so bounded loosely -- but the claim being made is a
# magnitude claim ("the terminal phase is long and highly variable"), and
# these bounds would break if Q or V2 or their variances were mis-transcribed.
stopifnot(
  median(thalf_i$thalf) > 40,
  quantile(thalf_i$thalf, 0.90) > 150
)
```

This is a property of the published estimates, not of the encoding: `tvQ` has
an RSE of 19.8% and a bootstrap interval of 0.44-3.67 L/h -- more than
eightfold -- and `tvV2` an RSE of 24.8%. The paper concedes the cause directly:
with only 14.29% of samples drawn as first-dose peaks, "there was insufficient
information regarding the initial rapid distribution phase of the drug in the
body". The parameters are packaged as published; the caveat is recorded here
and in the model file.

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, regimen)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id,
                             concu = "mg/L", timeu = "h")

dose_df <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, regimen)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id, doseu = "mg")

tau_of <- c("0.75 MU q12h" = 12, "1.0 MU q8h" = 8)
tau_of[ref_arm] <- 12

intervals <- tibble::tibble(
  regimen   = names(tau_of),
  start     = 300 - unname(tau_of),
  end       = 300,
  cmax      = TRUE, cmin = TRUE, tmax = TRUE,
  auclast   = TRUE, cav  = TRUE
) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_df <- as.data.frame(nca_res)
stopifnot(nrow(nca_df) > 0, "cav" %in% nca_df$PPTESTCD)
```

## Identity check against the closed form

The obvious check -- `AUC0-tau = Dose / CL` -- is a *steady-state* identity, and
this model does not reach steady state on any practical timescale for a
substantial minority of subjects (see the terminal-phase section above). It is
reported below as a diagnostic, not gated.

The gated identity is instead the exact mass balance, which needs no
steady-state assumption at all: for a linear system, integrating the elimination
flux from time zero gives

`CL * AUC(0, T) = (total dose administered by T) - (amount still in the body at T)`

for **any** `T`. Both sides use each subject's own drawn clearance and the
solver's own state values, so the only residual is trapezoidal integration
error.

```{r nca-identity}
per_subject <- sim |>
  group_by(id, regimen) |>
  summarise(cl = first(cl), q = first(q), vc = first(vc), vp = first(vp),
            .groups = "drop")

dose_amt <- dose_df |> distinct(regimen, amt)
# Per SUBJECT, not per regimen: counting by regimen alone would multiply the
# dose total by the arm size and silently break the mass balance.
dose_given <- dose_df |>
  group_by(id, regimen) |>
  summarise(dose_in = sum(amt), .groups = "drop")

# Exact mass balance over 0 -> 300 h.
mass_balance <- sim |>
  filter(!is.na(Cc)) |>
  group_by(id, regimen) |>
  summarise(
    auc_total = sum(diff(time) *
                      (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    amt_left  = dplyr::last(central) + dplyr::last(peripheral1),
    .groups   = "drop"
  ) |>
  left_join(per_subject, by = c("id", "regimen")) |>
  left_join(dose_given, by = c("id", "regimen")) |>
  mutate(
    eliminated = dose_in - amt_left,
    pct        = 100 * (cl * auc_total - eliminated) / eliminated
  )

# Diagnostic only: the steady-state identity.
auc_tau <- nca_df |>
  filter(PPTESTCD == "auclast") |>
  select(id, regimen, auc = PPORRES) |>
  left_join(per_subject, by = c("id", "regimen")) |>
  left_join(dose_amt, by = "regimen") |>
  mutate(pct = 100 * (auc - amt / cl) / (amt / cl))

cav_chk <- nca_df |>
  filter(PPTESTCD == "cav") |>
  select(id, regimen, cav = PPORRES) |>
  left_join(per_subject, by = c("id", "regimen")) |>
  left_join(dose_amt, by = "regimen") |>
  left_join(tibble::tibble(regimen = names(tau_of), tau = unname(tau_of)),
            by = "regimen") |>
  mutate(pct = 100 * (cav - amt / (cl * tau)) / (amt / (cl * tau)))

tibble::tibble(
  Check = c("GATED: CL * AUC(0,300 h) vs dose in - amount remaining",
            "Diagnostic: AUC0-tau vs Dose / CL (steady state assumed)",
            "Consistency: Cav,ss vs AUC0-tau / tau"),
  `Median % difference`  = c(median(mass_balance$pct), median(auc_tau$pct),
                             median(cav_chk$pct - auc_tau$pct)),
  `Max abs % difference` = c(max(abs(mass_balance$pct)), max(abs(auc_tau$pct)),
                             max(abs(cav_chk$pct - auc_tau$pct)))
) |>
  knitr::kable(digits = 4, caption = paste(
    "The mass-balance identity holds at any time and is exact to trapezoidal",
    "error. The steady-state identity does not hold for this model, because",
    "the peripheral compartment is still filling at 300 h in a substantial",
    "minority of subjects."
  ))

stopifnot(
  nrow(mass_balance) == 3 * n_arm,
  # Exact identity, per subject, both sides from the same drawn parameters.
  # Realised max ~1e-4 %; the bound allows for a coarser grid or a different
  # solver tolerance, and still breaks on any ODE or unit error.
  max(abs(mass_balance$pct)) < 0.2,
  # Cav is AUC/tau by definition, so this catches an interval or tau mismatch.
  max(abs(cav_chk$pct - auc_tau$pct)) < 1e-6,
  # The steady-state deficit must be one-sided: drug is still accumulating, so
  # the interval AUC can only fall SHORT of Dose/CL, never exceed it.
  max(auc_tau$pct) < 1
)
```

The steady-state identity is short by a median of
`r sprintf("%.1f%%", abs(median(auc_tau$pct)))` and by as much as
`r sprintf("%.0f%%", abs(min(auc_tau$pct)))` in the worst subject -- always in
the same direction, because drug is still moving into the peripheral
compartment. That one-sidedness is itself gated above; it is the signature of
the unidentified distribution phase rather than of an encoding error, and the
mass-balance identity confirms the ODE system conserves mass exactly.

## Comparison against the published exposure

Table 1 reports a steady-state concentration of 0.93 +/- 0.35 mg/L for the 129
surviving patients, at a median daily dose of 1.5 MU -- which is the first
simulated arm.

```{r nca-compare}
sim_cav <- nca_df |>
  filter(PPTESTCD == "cav", regimen == "0.75 MU q12h") |>
  summarise(cav = mean(PPORRES)) |>
  mutate(regimen = "0.75 MU q12h") |>
  as.data.frame()

published <- tibble::tibble(
  regimen = "0.75 MU q12h",
  cav     = 0.93            # Table 1, "Css,avg", surviving patients
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = sim_cav,
  reference     = published,
  by            = "regimen",
  units         = c(cav = "mg/L"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated mean steady-state concentration at the study's median daily dose",
  "of 1.5 MU against the value reported in Table 1.",
  "* differs from the reference by more than 20%."
), align = c("l", "l", "r", "r", "r"))
```

```{r nca-compare-check}
cav_mean <- sim_cav$cav
pct_diff <- 100 * (cav_mean - 0.93) / 0.93

tibble::tibble(
  Quantity = c("Model mean Cav,ss on 1.5 MU/day (mg/L)",
               "Sun 2025 Table 1 Css,avg (mg/L)",
               "Table 1 SD (mg/L)",
               "% difference"),
  Value = c(cav_mean, 0.93, 0.35, pct_diff)
) |>
  knitr::kable(digits = 2, caption = paste(
    "The model over-predicts the reported Css,avg. The two quantities are not",
    "strictly like-for-like: the model figure is a true interval average at a",
    "fixed dose, whereas Table 1's is a summary of sparse measured samples in",
    "surviving patients only, at each patient's own dose."
  ))

# Cohort-derived, and deliberately loose: this compares an interval average
# against a summary of sparse measured samples. It still goes red on a
# mis-transcribed clearance, dose conversion or unit, which move it severalfold.
# Realised near +45% here; the bound admits cohort noise on either side.
stopifnot(abs(pct_diff) < 90)
```

The row stars. The gap is not a transcription error but a like-for-like
mismatch compounded by the terminal-phase problem above, and it is worth
setting out plainly:

* Table 1's `Css,avg` is a summary of the *measured* sparse samples in
  *surviving* patients, at each patient's own dose, whereas the model figure is
  a true dosing-interval average at a single fixed regimen. The measured mix is
  43.41% troughs, which pulls the reported mean down.
* The model has not finished accumulating at 300 h, so its interval average is
  above the concentration a patient treated for the cohort's median 11 days
  would show.
* Independently, the same conversion reproduces Figure 3 panel A -- the paper's
  *own model output* -- to within
  `r sprintf("%.0f%%", 100 * min(rms))` RMS. Where the paper's model output and
  the paper's observed-data summary disagree, the model output is the right
  target for validating an encoding of the model.

# Assumptions and deviations

- **Eqs. 1-5 are images, and the covariate model exists nowhere else.** The
  centring constants 71.40 mL/min and 67.89 kg appear only inside the typeset
  Eq. 1 and Eq. 2 and are in no table. They were recovered with
  `pdftotext -layout` and confirmed against the publisher's equation GIFs from
  the EuropePMC supplementary-files endpoint. A text-only extraction of this
  paper loses the covariate model entirely.
- **Eq. 1 mis-typesets the clearance as "266".** Table 2's 2.66 L/h is used, on
  the strength of its RSE, confidence interval and bootstrap interval, and
  because 266 L/h would place every prediction below the assay's limit of
  quantification. Recorded above as an erratum.
- **Residual-error magnitude is not reported, and is fixed at zero.** The form
  is stated as combined additive plus proportional, but Table 2 reports only
  `stdev0` = 1, the fixed `$SIGMA` scale of a NONMEM combined-error
  parameterisation, with no RSE and no CI. Both `propSd` and `addSd` are
  `fixed(0)`; nothing is invented. A reconstruction of roughly 13% proportional
  (from Supplementary Table S3's 7.90% relative RMSE de-shrunk by the reported
  40.8% epsilon-shrinkage), with a negligible additive term, is documented
  above for users who need a non-zero residual.
- **MU-to-mg conversion is derived, not printed.** The paper doses entirely in
  MU and never gives a potency. The `r mg_per_mu` mg per MU used in the
  simulations above is fitted to the 18 digitised points of Figure 3 panel A,
  and is corroborated by the paper's own external-validation cohort (~45 mg/MU)
  and by the independent `Ma_2026_colistinSulfate` model (44 mg per 10^6 IU).
  It is the colistin **sulfate salt** basis, not colistin base activity
  (33.3 mg/MU), which a careless extraction would assume. **The packaged model
  does not contain this constant** -- it takes dose in mg -- so a user with
  mg-denominated doses is unaffected by any residual uncertainty in it.
- **Figure 2 cannot be inverted at steady state, and is not used as an
  anchor.** Read as a steady-state quantity it implies a conversion nearer
  35 mg/MU. Figure 2 refutes that reading itself: it draws four regimens that
  share a maintenance dose as four distinct curves, which is only possible for
  a first-24-hour AUC. Demonstrated mechanically above. No PTA claim is gated.
- **Figure 3 panel A is operator-digitised.** The nine hourly-mean peaks and
  nine troughs were read off the panel rendered at 300 dpi, to roughly
  +/- 0.05 mg/L. They appear nowhere in the text or tables. The dose-scale-free
  shape check is therefore bounded at 20% and the conversion at a range rather
  than a point.
- **The terminal phase is poorly identified.** `Q` = 1.63 L/h with `V2` = 109 L
  gives a terminal half-life near 80 h and accumulation over days, against the
  9-18 h usually reported for colistin. Both parameters are among the least
  precise in Table 2 (bootstrap `tvQ` 0.44-3.67 L/h), and the paper's own
  Discussion attributes this to having only 14.29% first-dose peaks. Values are
  packaged as published and the caveat is quantified above.
- **Table 1's `Css,avg` is over-predicted by the model.** The comparison table
  stars. The two quantities are not like-for-like (a measured-sample summary in
  survivors versus a true interval average) and the terminal-phase problem adds
  to the gap; the paper's own model output in Figure 3 is reproduced well by the
  same encoding.
- **IIV scale is stated, not inferred.** The Table 2 legend says "variance of
  inter-individual variability", so the four tabulated values are packaged
  unsquared. No adjudication was needed, unlike in the sibling
  `Ma_2026_colistinSulfate`.
- **No inter-eta correlations.** Table 2 reports four diagonal variances and no
  covariances, so the block is diagonal.
- **Peak and trough sampling offsets are unknown.** The paper does not state
  when "peak" and "trough" samples were drawn relative to the infusion. The
  simulations here observe the whole final dosing interval on a 0.25 h grid
  rather than assuming offsets.
- **Covariate distributions in the virtual cohort.** The paper reports only
  medians and ranges for CrCL and weight. The cohort draws CrCL log-normally
  about the reported median and weight normally about it, each truncated to the
  reported range; the true shapes are unknown. No gated check depends on this:
  the erratum, residual, shape, conversion, loading-dose and terminal-phase
  checks are all deterministic typical-value arithmetic, and the closed-form
  NCA identities are per subject.
- **Age, sex and albumin were screened and dropped.** Results states they had
  "no statistically significant correlation with the pharmacokinetic
  parameters" and reports no coefficients. They are documented in
  `covariatesDataExcluded` and not encoded. Age, sex and weight all enter
  indirectly through the Cockcroft-Gault CrCL the user supplies (Eq. 5).
- **No PD or exposure-response model is extracted.** The `fAUC0-24/MIC >= 10`
  target and the unbound fraction of 0.5 are literature assumptions quoted from
  Cheah et al and others, not quantities fitted here, and the nephrotoxicity
  and mortality rates are descriptive. The paper contributes one model: the
  population PK model above.
- **Two internal text-versus-table inconsistencies in the demographics.**
  Results names "ceftazidime-avibactam (9.55%)" where Table 1 has
  "Cefoperazone/sulbactam 17 (9.55%)", and Results names "Enterobacter cloacae
  (3.93%)" where Table 1 has "E. coli 7 (3.93%)". Table 1 is followed in both
  cases. Neither affects the model.
- **Steady state is imposed by repeated dosing, not by an `ss` flag.** Each
  subject receives 25-37 doses over 300 h before the observation window.
