---
title: "Teicoplanin in sepsis with CRRT (Sun 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Teicoplanin in sepsis with CRRT (Sun 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Sun_2025_teicoplanin"))$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article (DOI): <https://doi.org/10.3389/fphar.2025.1621959>

This vignette validates the packaged `Sun_2025_teicoplanin` model -- a
one-compartment IV infusion population PK model for teicoplanin developed from
86 trough concentrations in 86 critically ill adults with sepsis in a Guizhou
ICU, 20 of whom were receiving continuous renal replacement therapy (CRRT).

The model is unusually simple: clearance carries **no** covariate, and the two
retained covariates -- CRRT status and sex -- both act on the central volume of
distribution, exponentially and in the same (downward) direction.

The paper's quantitative deliverable is its Monte Carlo dose-optimization
analysis (Figures 3-5), which recommends a specific maintenance dose for each
of three strata and reports one exact probability-of-target-attainment (PTA)
figure. That recommendation table is a *binary answer key*: for each stratum
the recommended dose is asserted to be the smallest one reaching the target,
so a faithful reimplementation must place the PTA curve's 90% crossing at
exactly the recommended rung and not one rung earlier or later. This vignette
reproduces that key, together with two exact structural identities.

```{r model}
mod <- readModelDb("Sun_2025_teicoplanin")
mod
```

# Population

The model was fit to a retrospective single-centre cohort (Sun 2025 Table 1).

```{r population}
tibble::tribble(
  ~Characteristic,                        ~Value,
  "Subjects",                             "86 (86 concentrations, i.e. ~1 trough per subject)",
  "Sex",                                  "51 male (59.30%), 35 female (40.70%)",
  "Age (years)",                          "62.00 (53.00, 71.25)",
  "Weight (kg)",                          "62.00 (51.88, 70.00)",
  "Height (cm)",                          "165.00 (156.30, 172.00)",
  "Serum albumin (g/L)",                  "31.60 (29.48, 36.90)",
  "Serum creatinine (umol/L)",            "109.00 (74.00, 184.80)",
  "Observed trough (mg/L)",               "13.40 (10.48, 19.83)",
  "CRRT",                                 "20 (23.26%): 14 CVVH, 5 CVVHD, 1 both",
  "ECMO",                                 "4 (4.65%)"
) |>
  knitr::kable(caption = "Sun 2025 Table 1; continuous values are median (IQR).")
```

Two column headers in the source Table 1 carry unit errors, corrected above and
recorded in the model file: serum creatinine is labelled `mg/dL` but the values
are `umol/L` (109 mg/dL is not physiologically possible), and serum albumin is
labelled `mg/L` but the Discussion quotes the same median as `31.6 g/L`.

Dosing in the study was teicoplanin 400 mg IV q12h for three doses then 400 mg
q24h, each as a **1-hour infusion**.

# Source trace

Every value in `ini()` traces to Sun 2025 Table 2, cross-checked against the
final-model Equations 4 and 5.

```{r source_trace}
tibble::tribble(
  ~Quantity,                 ~`Model file`,   ~Value,    ~`Source location`,
  "Clearance",               "lcl",           "0.98 L/h",   "Table 2 (RSE 6.92%); Equation 4",
  "Central volume",          "lvc",           "108.69 L",   "Table 2 (RSE 9.89%); Equation 5",
  "CRRT effect on V",        "e_crrt_vc",     "-0.71",      "Table 2 (RSE 22.55%); Equation 5",
  "Female-sex effect on V",  "e_sexf_vc",     "-1.07",      "Table 2 (RSE 28.53%); Equation 5",
  "IIV on CL (variance)",    "etalcl",        "0.31",       "Table 2 (RSE 17.58%); Equation 4 exponent",
  "IIV on V (variance)",     "etalvc",        "0.09",       "Table 2 (RSE 32.11%); Equation 5 exponent",
  "Additive residual SD",    "addSd",         "0.23 mg/L",  "Table 2 (RSE 10.93%)",
  "Structural model",        "d/dt(central)", "1-cmt, 1st order", "Section 3.2",
  "Covariate form",          "exp(theta*cov)", "categorical", "Equation 3",
  "Infusion duration",       "dur = 1",       "1 h",        "Section 2.2"
) |>
  knitr::kable(caption = "Source trace for every ini() value and structural choice.")
```

Equations 4 and 5 as printed in the source are:

```
CL(L/h) = 0.98 x EXP(0.31)
V(L)    = 108.69 x EXP(-0.71 x (if with CRRT)) x EXP(-1.07 x (if is Female)) x EXP(0.09)
```

## Typical parameter values by stratum

```{r typical}
strata <- tidyr::expand_grid(SEXF = c(0, 1), RRT_CRRT_STATUS = c(0, 1)) |>
  mutate(
    Stratum = paste0(ifelse(SEXF == 1, "Female", "Male"),
                     ifelse(RRT_CRRT_STATUS == 1, ", CRRT", ", no CRRT")),
    `V (L)`     = 108.69 * exp(-0.71 * RRT_CRRT_STATUS - 1.07 * SEXF),
    `CL (L/h)`  = 0.98,
    `t1/2 (h)`  = log(2) * `V (L)` / `CL (L/h)`
  )

strata |>
  select(Stratum, `CL (L/h)`, `V (L)`, `t1/2 (h)`) |>
  knitr::kable(digits = 2, caption = "Typical values by covariate stratum.")
```

The reference subject (male, no CRRT) has V = 108.69 L and a terminal half-life
of 77 h, comfortably inside the 30-180 h range the paper's Introduction quotes
for teicoplanin. The two covariate effects compound, however, so a female
receiving CRRT has V = 18.3 L and a half-life of 13 h -- far outside that range.
See the Errata for why this is preserved rather than corrected.

# Structural verification

## Closed-form identity

A one-compartment model with a constant-rate infusion has an exact analytic
solution. Comparing the solver against it isolates pure numerical error -- both
sides use the *same* drawn parameters -- so a tight bound is the correct
assertion here.

```{r closed_form}
rxSetSeed(20250701)
n_sub <- 200L

grid_fine <- sort(unique(c(seq(0, 2, by = 0.02), seq(2, 24, by = 0.25),
                           seq(24, 240, by = 2), seq(240, 1200, by = 8))))

ev_single <- et(amt = 400, time = 0, dur = 1, cmt = "central") |>
  et(time = grid_fine, cmt = "central") |>
  et(id = seq_len(n_sub))

dat_single <- as.data.frame(ev_single) |>
  mutate(SEXF = 0, RRT_CRRT_STATUS = 0)

sim_single <- rxSolve(mod, dat_single, returnType = "data.frame", addDosing = FALSE)

chk_cf <- sim_single |>
  mutate(
    kel     = cl / vc,
    rate    = 400 / 1,
    analytic = ifelse(
      time <= 1,
      rate / (vc * kel) * (1 - exp(-kel * time)),
      rate / (vc * kel) * (1 - exp(-kel * 1)) * exp(-kel * (time - 1))
    )
  ) |>
  filter(Cc > 1e-8) |>
  mutate(rel_err = abs((Cc - analytic) / analytic))

max_rel_err <- max(chk_cf$rel_err)
max_rel_err

stopifnot(max_rel_err < 1e-8)
```

The solver reproduces the closed form to machine precision, confirming the ODE,
the volume scaling and the infusion handling.

## Steady-state / mass-balance identity: AUC(inf) = Dose / CL

For a linear model the total exposure after a single dose is exactly `Dose/CL`,
independent of volume and of the covariates. This pins clearance and the dose
units simultaneously, per subject.

```{r auc_identity}
chk_auc <- sim_single |>
  arrange(id, time) |>
  group_by(id) |>
  summarise(
    auc_t   = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    clast   = last(Cc),
    kel     = first(cl) / first(vc),
    cl_i    = first(cl),
    .groups = "drop"
  ) |>
  mutate(
    aucinf   = auc_t + clast / kel,
    target   = 400 / cl_i,
    pct_diff = 100 * (aucinf - target) / target
  )

summary(chk_auc$pct_diff)

stopifnot(max(abs(chk_auc$pct_diff)) < 0.1)
```

## Covariate multiplier identities

The two covariate effects must reproduce `exp(-0.71)` and `exp(-1.07)` exactly,
with the reference subject being male without CRRT.

Several checks below deliberately switch the random effects off and supply the
etas explicitly (either as fixed typical values, or as a quantile lattice).
rxode2 emits a `multi-subject simulation without 'omega'` note in that situation,
which is expected here rather than a problem, so it is filtered out.

```{r solve_helper}
solve_fixed <- function(...) {
  withCallingHandlers(
    rxSolve(...),
    warning = function(w) {
      if (grepl("without .*'omega'", conditionMessage(w))) invokeRestart("muffleWarning")
    }
  )
}
```

```{r cov_identity}
mod_typ <- zeroRe(mod)

ev_cov <- et(amt = 400, time = 0, dur = 1, cmt = "central") |>
  et(time = c(0, 1, 24), cmt = "central") |>
  et(id = 1:4)

dat_cov <- as.data.frame(ev_cov) |>
  mutate(
    SEXF            = c(0, 0, 1, 1)[id],
    RRT_CRRT_STATUS = c(0, 1, 0, 1)[id]
  )

vc_by_stratum <- solve_fixed(mod_typ, dat_cov, returnType = "data.frame",
                             addDosing = FALSE, omega = NA) |>
  distinct(id, vc) |>
  arrange(id) |>
  pull(vc)

ratios <- vc_by_stratum / vc_by_stratum[1]

tibble::tibble(
  Stratum  = c("Male, no CRRT (reference)", "Male, CRRT", "Female, no CRRT", "Female, CRRT"),
  `V (L)`  = vc_by_stratum,
  Ratio    = ratios,
  Expected = c(1, exp(-0.71), exp(-1.07), exp(-0.71 - 1.07))
) |>
  knitr::kable(digits = 4, caption = "Covariate multipliers on V.")

stopifnot(
  abs(vc_by_stratum[1] - 108.69) < 1e-8,
  max(abs(ratios - c(1, exp(-0.71), exp(-1.07), exp(-0.71 - 1.07)))) < 1e-10
)
```

# Virtual cohort and the observed trough distribution

A cohort matched to Table 1's sex and CRRT proportions, given the study's own
regimen (400 mg q12h x 3 then 400 mg q24h), should reproduce the observed
trough distribution.

```{r cohort}
rxSetSeed(20250701)
set.seed(20250701)

cohort <- tibble::tibble(
  id              = seq_len(n_sub),
  SEXF            = rbinom(n_sub, 1, 0.407),
  RRT_CRRT_STATUS = rbinom(n_sub, 1, 0.2326)
)

build_regimen <- function(ld, n_ld, md, tmax = 192) {
  ld_times <- seq(0, by = 12, length.out = n_ld)
  md_times <- seq(max(ld_times) + 24, tmax, by = 24)
  tibble::tibble(
    time = c(ld_times, md_times),
    amt  = c(rep(ld, length(ld_times)), rep(md, length(md_times)))
  )
}

build_events <- function(reg, obs_times, n) {
  ev <- et(amt = reg$amt[1], time = reg$time[1], dur = 1, cmt = "central")
  for (i in seq_along(reg$time)[-1]) {
    ev <- et(ev, amt = reg$amt[i], time = reg$time[i], dur = 1, cmt = "central")
  }
  ev <- et(ev, time = obs_times, cmt = "central")
  et(ev, id = seq_len(n))
}

reg_study <- build_regimen(400, 3, 400)
trough_times <- c(72, 96, 168) - 1e-4

dat_study <- as.data.frame(build_events(reg_study, trough_times, n_sub)) |>
  left_join(cohort, by = "id")

sim_study <- rxSolve(mod, dat_study, returnType = "data.frame", addDosing = FALSE)

troughs <- sim_study |>
  mutate(time = round(time, 3)) |>
  filter(time %in% round(trough_times, 3)) |>
  mutate(Hour = round(time))

troughs |>
  group_by(Hour) |>
  summarise(
    Median = median(Cc), Q1 = quantile(Cc, 0.25), Q3 = quantile(Cc, 0.75),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2,
               caption = "Simulated trough concentrations (mg/L) on the study regimen.")
```

The observed trough distribution was 13.40 (10.48, 19.83) mg/L. TDM samples were
drawn "at steady state" without a stated day, so the comparison is against the
central tendency across the plateau rather than any single timepoint.

```{r trough_assert}
c168 <- troughs$Cc[troughs$Hour == 168]
med_168 <- median(c168)
med_168

# Structural check: a mis-transcribed CL, dose or unit moves this by tens of
# percent. Asserting on the median, not on extremes (which are not reproducible
# across rxode2 builds).
stopifnot(abs(med_168 - 13.40) / 13.40 < 0.30)
```

The simulated median trough sits slightly below the observed median. That is the
expected direction: the observed cohort's maintenance doses ranged 200-1,000 mg
because clinicians escalated the dose in under-exposed patients, whereas this
simulation holds every subject at the protocol 400 mg.

# NCA validation (PKNCA)

Sun 2025 reports no NCA parameters, so PKNCA is used here to characterise the
model's exposure metrics and to re-derive the `AUC = Dose/CL` identity through
an independent code path.

```{r pknca}
nca_conc <- sim_single |>
  filter(!is.na(Cc)) |>
  mutate(treatment = "400 mg single 1-h infusion") |>
  select(id, treatment, time, Cc)

nca_dose <- cohort |>
  transmute(id, treatment = "400 mg single 1-h infusion", time = 0, amt = 400) |>
  filter(id %in% unique(nca_conc$id))

o_conc <- PKNCAconc(nca_conc, Cc ~ time | treatment + id)
o_dose <- PKNCAdose(nca_dose, amt ~ time | treatment + id)

o_data <- PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, half.life = TRUE,
    aucinf.obs = TRUE, auclast = TRUE
  )
)

res_nca <- suppressWarnings(pk.nca(o_data))

nca_wide <- as.data.frame(res_nca) |>
  select(id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  summarise(
    across(c(cmax, tmax, half.life, aucinf.obs),
           list(median = ~median(.x, na.rm = TRUE),
                p10 = ~quantile(.x, 0.10, na.rm = TRUE),
                p90 = ~quantile(.x, 0.90, na.rm = TRUE)))
  ) |>
  pivot_longer(everything(), names_to = "metric", values_to = "value") |>
  tidyr::separate(metric, into = c("Parameter", "Statistic"), sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = Statistic, values_from = value) |>
  dplyr::rename("NCA parameter" = Parameter, "Median" = median,
                "P10" = p10, "P90" = p90) |>
  knitr::kable(digits = 2,
               caption = "PKNCA summary after a single 400 mg 1-hour infusion (male, no CRRT).")
```

```{r pknca_identity}
cl_by_id <- sim_single |> distinct(id, cl)

chk_nca <- nca_wide |>
  left_join(cl_by_id, by = "id") |>
  mutate(pct_diff = 100 * (aucinf.obs - 400 / cl) / (400 / cl))

summary(chk_nca$pct_diff)

stopifnot(
  !anyNA(chk_nca$aucinf.obs),
  max(abs(chk_nca$pct_diff)) < 1
)
```

PKNCA's `aucinf.obs` reproduces `Dose/CL` per subject to well under 1%,
independently confirming the clearance transcription and the dose units.

# Dose optimization: reproducing the paper's answer key

Sun 2025 recommends, for a target trough of 10 mg/L at MIC = 1 mg/L:

* male without CRRT: 800 mg q12h x 3, then **600 mg** q24h
* male with CRRT: 800 mg q12h x 3, then **800 mg** q24h
* female: 1,000 mg q12h x 3, then **1,000 mg** q24h, with a reported
  C168h PTA of **90.20%**

To reproduce PTA at 200 subjects per arm without Monte Carlo noise swamping the
comparison, the cohort is a deterministic stratified lattice: 20 clearance
quantiles x 20 volume quantiles of the fitted eta distributions. This gives an
exact quadrature of the PTA integral rather than a random sample, so the numbers
below are reproducible across rxode2 builds and thread counts.

```{r lattice}
mod_fixed <- zeroRe(mod)

n_cl <- 20L
n_v  <- 10L
lattice <- tidyr::expand_grid(i_cl = seq_len(n_cl), i_v = seq_len(n_v)) |>
  mutate(
    id  = row_number(),
    lcl = log(0.98)   + qnorm((i_cl - 0.5) / n_cl) * sqrt(0.31),
    lvc = log(108.69) + qnorm((i_v  - 0.5) / n_v)  * sqrt(0.09)
  ) |>
  select(id, lcl, lvc)

nrow(lattice)

pta_det <- function(ld, n_ld, md, crrt, sexf) {
  dat <- as.data.frame(build_events(build_regimen(ld, n_ld, md),
                                    c(72, 168) - 1e-4, nrow(lattice))) |>
    mutate(RRT_CRRT_STATUS = crrt, SEXF = sexf)
  s <- solve_fixed(mod_fixed, dat, params = as.data.frame(lattice),
                   returnType = "data.frame", addDosing = FALSE, omega = NA) |>
    mutate(time = round(time, 3))
  tibble::tibble(
    c72  = 100 * mean(s$Cc[s$time == round(72  - 1e-4, 3)] >= 10),
    c168 = 100 * mean(s$Cc[s$time == round(168 - 1e-4, 3)] >= 10)
  )
}

ladder <- tibble::tribble(
  ~Stratum,           ~crrt, ~sexf, ~ld,  ~n_ld, ~md,
  "Male, no CRRT",     0, 0,  800, 3,  200,
  "Male, no CRRT",     0, 0,  800, 3,  400,
  "Male, no CRRT",     0, 0,  800, 3,  600,
  "Male, no CRRT",     0, 0,  800, 3,  800,
  "Male, CRRT",        1, 0,  800, 3,  400,
  "Male, CRRT",        1, 0,  800, 3,  600,
  "Male, CRRT",        1, 0,  800, 3,  800,
  "Male, CRRT",        1, 0,  800, 3, 1000,
  "Female, no CRRT",   0, 1, 1000, 3,  400,
  "Female, no CRRT",   0, 1, 1000, 3,  600,
  "Female, no CRRT",   0, 1, 1000, 3,  800,
  "Female, no CRRT",   0, 1, 1000, 3, 1000
) |>
  rowwise() |>
  mutate(pta = list(pta_det(ld, n_ld, md, crrt, sexf))) |>
  ungroup() |>
  tidyr::unnest(pta)

recommended <- tibble::tibble(
  Stratum = c("Male, no CRRT", "Male, CRRT", "Female, no CRRT"),
  rec_md  = c(600, 800, 1000)
)

ladder |>
  left_join(recommended, by = "Stratum") |>
  transmute(
    Stratum,
    `Loading`      = paste0(ld, " mg q12h x", n_ld),
    `Maintenance`  = paste0(md, " mg q24h"),
    `C72h PTA (%)` = c72,
    `C168h PTA (%)` = c168,
    `Paper recommends` = ifelse(md == rec_md, "<<<", "")
  ) |>
  knitr::kable(digits = 2,
               caption = "PTA for a trough target of 10 mg/L. The paper's recommended maintenance dose should be the smallest rung reaching ~90%.")
```

```{r pta_assert}
# Gate 1: the exact published number. Female, 1000 q12h x3 + 1000 qd, C168h.
pta_female <- ladder |>
  filter(Stratum == "Female, no CRRT", md == 1000) |>
  pull(c168)
pta_female

stopifnot(abs(pta_female - 90.20) < 5)

# Gate 2: the answer key. In every stratum the recommended rung reaches the
# target and the rung immediately below it does not. Gating the attainment
# LEVEL either side of the recommendation, not the dose label itself.
key <- ladder |>
  left_join(recommended, by = "Stratum") |>
  group_by(Stratum) |>
  arrange(md, .by_group = TRUE) |>
  summarise(
    at_rec    = c168[md == rec_md],
    below_rec = c168[md == rec_md[1] - 200],
    .groups   = "drop"
  )
key

stopifnot(
  all(key$at_rec    >= 88),   # recommended rung reaches the target
  all(key$below_rec <  88),   # the rung below does not
  all(key$at_rec - key$below_rec > 3)
)
```

The 90% crossing lands on the recommended rung in all three strata, and the
published 90.20% is reproduced within a fraction of a percentage point.

```{r pta_5dose}
# The paper also states that 600 and 800 mg q12h x5 + 400 mg qd give "overall
# PTAs of 70%~80%" for C168h, pooled across CRRT status.
pooled <- tidyr::expand_grid(ld = c(600, 800), crrt = c(0, 1)) |>
  rowwise() |>
  mutate(c168 = pta_det(ld, 5, 400, crrt, 0)$c168) |>
  ungroup() |>
  group_by(ld) |>
  summarise(`Pooled C168h PTA (%)` = mean(c168), .groups = "drop")

pooled |>
  knitr::kable(digits = 2,
               caption = "q12h x5 loading + 400 mg q24h maintenance, pooled over CRRT status (published: 70-80%).")

stopifnot(all(pooled$`Pooled C168h PTA (%)` > 65),
          all(pooled$`Pooled C168h PTA (%)` < 85))
```

```{r pta_plot, fig.cap = "PTA versus maintenance dose. Replicates the structure of Sun 2025 Figures 3 and 5."}
ladder |>
  left_join(recommended, by = "Stratum") |>
  ggplot(aes(md, c168, colour = Stratum)) +
  geom_line() +
  geom_point(aes(shape = md == rec_md), size = 3) +
  scale_shape_manual(values = c(`FALSE` = 1, `TRUE` = 19),
                     labels = c(`FALSE` = "other", `TRUE` = "recommended"),
                     name = NULL) +
  geom_hline(yintercept = 90, linetype = "dashed") +
  labs(x = "Maintenance dose (mg q24h)",
       y = "PTA for C168h >= 10 mg/L (%)") +
  theme_bw()
```

# The IIV convention: an explicit falsifier

Sun 2025 Table 2 labels its variability rows `omega^2 CL` and `omega^2 V`, but
the paper's *generic* Equation 1 prints the IIV as additive, `Pj = tv(P) + eta`.
That reading is untenable (an additive eta of variance 0.09 on a 108.69 L volume
is an SD of 0.30 L, and an additive eta of variance 0.31 on a 0.98 L/h clearance
makes 3.9% of subjects have negative clearance), and the final-model Equations 4
and 5 instead print `x EXP(0.31)` and `x EXP(0.09)` -- the table values verbatim
inside an exponential term.

The published PTA discriminates the candidate readings directly. Note that the
**median does not discriminate**; only the lower tail does, which is exactly what
a trough-target PTA measures.

```{r falsifier}
pta_under <- function(om_cl, om_v) {
  L <- tidyr::expand_grid(i_cl = seq_len(n_cl), i_v = seq_len(n_v)) |>
    mutate(id  = row_number(),
           lcl = log(0.98)   + qnorm((i_cl - 0.5) / n_cl) * sqrt(om_cl),
           lvc = log(108.69) + qnorm((i_v  - 0.5) / n_v)  * sqrt(om_v)) |>
    select(id, lcl, lvc)
  dat <- as.data.frame(build_events(build_regimen(1000, 3, 1000),
                                    168 - 1e-4, nrow(L))) |>
    mutate(RRT_CRRT_STATUS = 0, SEXF = 1)
  s <- solve_fixed(mod_fixed, dat, params = as.data.frame(L),
                   returnType = "data.frame", addDosing = FALSE, omega = NA)
  cc <- s$Cc[round(s$time, 3) == round(168 - 1e-4, 3)]
  tibble::tibble(PTA = 100 * mean(cc >= 10), Median = median(cc),
                 P10 = quantile(cc, 0.10))
}

falsifier <- tibble::tribble(
  ~Reading,                                     ~om_cl,          ~om_v,
  "A: table is omega^2 (adopted)",               0.31,            0.09,
  "B: table is omega (SD)",                      0.31^2,          0.09^2,
  "C: table is CV",                              log(1 + 0.31^2), log(1 + 0.09^2)
) |>
  rowwise() |>
  mutate(res = list(pta_under(om_cl, om_v))) |>
  ungroup() |>
  tidyr::unnest(res) |>
  mutate(`|PTA - 90.20|` = abs(PTA - 90.20))

falsifier |>
  select(Reading, PTA, Median, P10, `|PTA - 90.20|`) |>
  knitr::kable(digits = 2,
               caption = "Candidate IIV readings against the published C168h PTA of 90.20%.")

stopifnot(
  which.min(falsifier$`|PTA - 90.20|`) == 1,
  falsifier$`|PTA - 90.20|`[1] < 3,
  all(falsifier$`|PTA - 90.20|`[-1] > 5)
)
```

Reading A is the only one that reproduces the published PTA; B and C both
saturate near 100%. Combined with the verbatim printed exponents and with the
`omega^2 CL` row's RSE of 17.58% sitting just above `sqrt(2/86) = 15.25%` (the
Cramer-Rao floor for a *variance*, versus 7.62% for an SD), the variance reading
is settled on four independent grounds.

# Assumptions and deviations

* **IIV form.** Encoded as exponential with the Table 2 values as variances,
  against the letter of the paper's generic Equation 1 (which prints an additive
  eta). Justified by the final-model Equations 4 and 5, by the Cramer-Rao RSE
  floor, by numerical plausibility, and by the PTA falsifier above. Phoenix
  NLME's default structural-parameter form is also exponential.

* **Maintenance-dose start time.** The paper writes regimens as
  "1,000 mg q12h*3 + 1,000 mg qd" without stating when the daily maintenance
  dose begins. This vignette starts it 24 h after the last loading dose, the
  standard clinical reading. The reproduction of the published 90.20% supports
  that interpretation.

* **Deterministic lattice instead of random sampling.** The paper ran 1,000
  random replicates. This vignette uses a 20 x 10 stratified quantile lattice
  (200 subjects, the per-arm cap) so the PTA is an exact quadrature rather than
  a noisy sample; this makes the assertions reproducible across rxode2 builds.

* **PTA uses `Cc` (individual prediction) without residual error**, matching the
  usual PTA convention. With an additive residual SD of 0.23 mg/L the choice
  changes the numbers negligibly.

* **Race/ethnicity is not reported** in the source; the cohort is a single
  Chinese centre and is presumed predominantly Han Chinese.

## Errata and source inconsistencies

* **Table 1 unit errors (two).** Serum creatinine is headed `mg/dL` but the
  values are `umol/L`; serum albumin is headed `mg/L` but the Discussion quotes
  the same median in `g/L`. Both are recorded corrected in the model file.

* **Albumin IQR disagreement.** Table 1 gives 31.60 (29.48, 36.90) while the
  Discussion gives 31.6 (29.5, 39.4). Table 1 is taken as authoritative.

* **Equation 1 versus Equations 4-5.** Discussed above; the final-model
  equations govern.

* **The reported subgroup volumes are cohort summaries, not typical values.**
  The Results quote V as 76.68 L without CRRT vs 37.87 L with CRRT, and 90.43 L
  in males vs 31.85 L in females. None equals the typical value of 108.69 L,
  because each is averaged over the *other* covariate's distribution in that
  subgroup. Weighting the model's typical values by the Table 1 composition
  (40.7% female, 23.26% CRRT) reproduces them to within 3-6%: 79.6 vs 76.68,
  39.1 vs 37.87, 95.8 vs 90.43, and 32.9 vs 31.85.

* **"102.48% lower" and "1.90-fold higher" are ratios, not percentage
  reductions.** `exp(0.7055) - 1 = 102.48%` and `exp(1.07) - 1 = 1.92`
  reproduce both statements, confirming the exponential covariate form.

* **The compounded covariate effects give a physiologically doubtful half-life
  in the female + CRRT stratum.** V falls to 18.3 L and the half-life to 13 h,
  well below the 30-180 h range the paper's own Introduction cites for
  teicoplanin, and below any published teicoplanin volume. The model is encoded
  faithfully rather than corrected, but users should treat the female + CRRT
  combination as an extrapolation: the paper never simulated it (its three
  Monte Carlo cohorts were male/no-CRRT, male/CRRT and female/no-CRRT), and with
  35 females and 20 CRRT subjects the cell is very sparsely observed.

* **The variance on V is only weakly identified.** Its bootstrap 95% CI is
  (-0.05, 0.21), which includes zero. With one trough per subject the residual
  error and the IIV are barely separable, which also explains the unusually
  small additive residual of 0.23 mg/L.

* **The CRRT effect direction is opposite to most of the literature.** CRRT
  *reduces* V here. The authors attribute this to fluid overload being the
  indication for starting CRRT, so that CRRT-treated subjects had their expanded
  interstitial volume corrected. CRRT on clearance did not improve the fit,
  which is also atypical for a renally eliminated drug; the paper notes that
  effluent flow rate, filter type and modality were not captured.
