---
title: "Maribavir (Sun 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Maribavir (Sun 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Sun_2025_maribavir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Sun_2025_maribavir"))$description`
* Article: <https://doi.org/10.1002/psp4.70054>
* Supporting Information (file `s001`, which contains the **final NONMEM control
  stream** plus Tables S1-S4): <https://doi.org/10.1002/psp4.70054>

Maribavir is an orally bioavailable benzimidazole riboside with a selective
multimodal mechanism of action against human cytomegalovirus (CMV). Sun 2025 is
an **update** of a previously published pooled population PK analysis: it adds
the randomized, double-blind phase 3 AURORA study in hematopoietic cell
transplant (HCT) recipients with first asymptomatic CMV infection, and a phase 1
study in Japanese-descended and non-Hispanic Caucasian individuals. Relative to
the earlier analysis the allometric body-weight exponents are now estimated
rather than fixed, and proton-pump-inhibitor (PPI) effects on both relative
bioavailability and the absorption rate constant are new.

A companion model from the same programme, `modellib("Sun_2023_maribavir")`,
encodes the earlier pooled analysis that supported the adolescent dosing
recommendation. The two are **not** interchangeable: that one fixes the weight
exponents at 0.75 / 1 and carries no PPI effect.

This vignette covers **seventeen** models from one paper. The population PK
model is validated first; the sixteen exposure-response logistic regressions
that Sun 2025 fitted to the phase 3 AURORA arm are packaged as the
`Sun_2025_maribavir_*` family and validated in the *Exposure-response models*
section below, which also closes the loop by feeding PK-simulated exposures into
the efficacy model.

```{r model}
mod <- rxode2::rxode(readModelDb("Sun_2025_maribavir"))
mod
```

# Population

The parameter-estimation population is 930 individuals contributing 7431
maribavir plasma concentration records, pooled across phase 1, 2 and 3 studies
(Sun 2025 Table 1 and Section 3.1):

* **Phase 1 (n = 206, 4231 records)** -- 157 healthy volunteers, 19 with renal
  impairment, 20 stable renal transplant recipients and 10 with hepatic
  impairment. None have CMV infection.
* **Phase 2/3 (n = 724, 3200 records)** -- HCT (440) and solid organ transplant
  (SOT, 284) recipients with CMV infection, including the SOLSTICE refractory
  population. CMV category: asymptomatic 644, symptomatic 44, organ disease 36.
* **AURORA (n = 238)** -- a subset of the phase 2/3 group; all HCT recipients
  with first asymptomatic CMV infection. This is the cohort newly added in this
  update and the cohort used for the exposure-response analyses.

Demographics: 82% aged 18 to <65 years and 18% aged 65 to <80, with a **single**
individual under 18 -- that one individual is the stated reason the allometric
exponents were estimated rather than fixed, in support of an ongoing paediatric
phase 3 study. Weight median 73.0 kg (range 36.1-141), mean 74.2 (SD 17.4). 41%
female. Race: Caucasian 77%, Black 13%, Asian 7%, Other 3%.

Co-medication prevalence is high and matters for this model: PPIs 54%, strong
CYP3A4 inhibitors 13%, moderate CYP3A4 inhibitors 11%, H2 blockers 10%, antacids
8%, strong CYP3A4 inducers 2%.

Below-limit-of-quantification records (297; 3.5% of post-dose values) were
excluded under method M1. Estimation used NONMEM 7.5.1 with IMPMAP; confidence
intervals came from non-parametric bootstrap stratified by study.

# Source trace

Every `ini()` value comes from the published Table 2. The model **structure**
comes from the final NONMEM control stream in Supporting Information file
`s001`. The distinction matters: the control stream's `$THETA` / `$OMEGA` /
`$SIGMA` blocks are that run's *initial* estimates, not its final ones -- for
example `THETA(16) [KA~dose]` is `-1.17` there against a published final
`-1.02`, and `-1.17` is in fact the lower bound of the published 95% CI. Values
therefore come from the table, structure from the code.

| Model element | Source location | Value |
|---|---|---|
| CL/F reference | Table 2 `CL/F (L/h)` | 3.94 (95% CI 3.69-4.20) |
| Vc/F reference | Table 2 `Vc/F (L)` | 17.8 (16.9-18.9) |
| Q/F reference | Table 2 `Q/F (L/h)` | 1.24 (0.962-1.60) |
| Vp/F reference | Table 2 `Vp/F (L)` | 7.10 (6.04-8.35) |
| Ka at 800 mg | Table 2 `Ka (1/h)` | 0.697 (0.594-0.819) |
| Absorption lag | Table 2 `Lag (h)` | 0.212 (0.192-0.235) |
| F reference | Table 2 `F` | 1 (no uncertainty reported; `fixed()`) |
| WT exponent, CL/F and Q/F | Table 2 `Effect of WT on CL/F` = `Effect of WT on Q/F`; control stream `MU_3 = THETA(3) + CLWT` | 0.301 (0.159-0.443) |
| WT exponent, Vc/F and Vp/F | Table 2 `Effect of WT on Vc/F` = `Effect of WT on Vp/F`; control stream `MU_4 = THETA(4) + VCWT` | 0.536 (0.378-0.694) |
| Strong CYP3A4 inhibitor on CL/F | Table 2 `Effect of CYP3AINH on CL/F` | x0.709 (0.681-0.737) |
| CYP3A4 inducer on CL/F | Table 2 `Effect of CYP3AIND on CL/F` | x2.27 (2.15-2.38) |
| CMV status on CL/F | Table 2 `Effect of CMV on CL/F` | x0.678 (0.626-0.733) |
| Dose on Ka | Table 2 `Effect of dose on Ka` | x(DOSE/800)^-1.02 (-1.17 to -0.875) |
| PPI on Ka | Table 2 `Effect of PPI on Ka` | x0.457 (0.367-0.568) |
| PPI on F | Table 2 `Effect of PPI on F` | x0.905 (0.849-0.964) |
| IIV diagonals (6) | Table 2 `IIV (%)` column | 49.5, 25.3, 145, 110, 72.0, 48.8% |
| IIV block structure | Control stream `3 OMEGA BLOCK(2)` over `ETA(1..6)` | CL~Vc, Q~Vp, Ka~lag |
| IIV off-diagonals | **Not published**; correlations carried from the control stream `$OMEGA` initial estimates | r = 0.673, 0.820, -0.686 |
| Additive residual | Table 2 `sigma^2 add` | 0.000112 -> SD 0.010583 |
| Proportional residual, phase 1 | Table 2 `sigma^2 prop Phase 1` | 0.0705 -> SD 0.265518 |
| Proportional residual, phase 2/3 | Table 2 `sigma^2 prop Phase 2 & 3` | 0.158 -> SD 0.397492 |
| Structural model, `$ERROR`, `F1 = PPIF`, `ALAG1` | Control stream (`ADVAN4 TRANS4`) | -- |

Two structural readings deserve emphasis, because Table 2 alone is ambiguous
about both and only the control stream settles them.

**Two weight exponents, not four.** Table 2 lists four weight rows, but the CL/F
and Q/F rows carry *identical* estimates, %RSE and 95% CI, as do the Vc/F and
Vp/F rows. That is not a coincidence of rounding: the control stream computes
`CLWT` and `VCWT` once and reuses them (`MU_3 = THETA(3) + CLWT`,
`MU_4 = THETA(4) + VCWT`), while `THETA(9)` and `THETA(10)` are present but
commented `Not used`. The model file therefore has two exponent parameters.

**The `IIV (%)` back-transform.** Table 2 reports IIV as a percentage without
stating the convention. This file uses the rule that the same analysis group
documented explicitly for the same drug in the companion Sun 2023 paper
(its Table S2 footnote c): `CV = sqrt(omega^2)` when `omega^2 <= 0.15` and
`CV = sqrt(exp(omega^2) - 1)` otherwise. All six Table 2 entries are
self-consistent under that rule, and only Vc/F falls in the first branch, where
the two branches differ by less than 2% in SD. The two best-determined
parameters corroborate it: back-transforming CL/F's 49.5% gives `omega^2` =
0.2192 against the control stream's initial 0.214, and the lag time's 48.8%
gives 0.2136 against an initial 0.203.

# Virtual cohort

Two arms of 200 participants each, matching the two health-status strata that
Sun 2025 Table 3 reports separately. Both are simulated at 400 mg twice daily
**without concomitant medication**, which is exactly the condition Table 3 and
Figure 2 state for their exposure summaries.

Weights are drawn from log-normal distributions matched to the per-stratum
Table 1 summaries (phase 1 mean 78.1, SD 15.1; phase 2/3 mean 73.1, SD 17.8) and
truncated to the reported overall range 36.1-141 kg.

```{r cohort}
set.seed(20250704)
rxode2::rxSetSeed(20250704)

n_arm <- 200L

draw_wt <- function(n, mean_kg, sd_kg) {
  sdlog <- sqrt(log1p((sd_kg / mean_kg)^2))
  meanlog <- log(mean_kg) - sdlog^2 / 2
  pmin(pmax(stats::rlnorm(n, meanlog, sdlog), 36.1), 141)
}

cohort <- dplyr::bind_rows(
  tibble::tibble(
    id        = seq_len(n_arm),
    treatment = "Healthy volunteers",
    WT        = draw_wt(n_arm, 78.1, 15.1),
    DIS_CMV   = 0,
    STUDY_MARIBAVIR_PHASE1 = 1
  ),
  tibble::tibble(
    id        = n_arm + seq_len(n_arm),
    treatment = "Transplant recipients with CMV",
    WT        = draw_wt(n_arm, 73.1, 17.8),
    DIS_CMV   = 1,
    STUDY_MARIBAVIR_PHASE1 = 0
  )
) |>
  dplyr::mutate(
    DOSE                     = 400,
    CONMED_PPI               = 0,
    CONMED_CYP3A4_INH_STRONG = 0,
    CONMED_CYP3A4_IND        = 0
  )

cohort |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    n = dplyr::n(),
    `Weight mean (kg)`   = round(mean(WT), 1),
    `Weight median (kg)` = round(stats::median(WT), 1),
    `Weight SD (kg)`     = round(stats::sd(WT), 1),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Simulated cohort weight distributions (Sun 2025 Table 1 targets: phase 1 mean 78.1 SD 15.1, median 75.6; phase 2/3 mean 73.1 SD 17.8, median 71.3).")
```

# Simulation

400 mg twice daily for 10 days (20 doses, `ii = 12`, `addl = 19`), which is far
beyond the 2 days to steady state the paper reports for this regimen. The
observation grid covers the **last** dosing interval, 228-240 h, so the
non-compartmental interval is a true steady-state interval and its lower bound
carries a real pre-dose trough record rather than a synthetic time-zero anchor.

```{r simulate}
events <- rxode2::et(amt = 400, ii = 12, addl = 19, cmt = "depot") |>
  rxode2::et(seq(228, 240, by = 0.1), cmt = "central") |>
  rxode2::et(id = cohort$id)

ev_df <- as.data.frame(events) |>
  dplyr::left_join(cohort, by = "id")

sim <- rxode2::rxSolve(mod, ev_df, returnType = "data.frame") |>
  dplyr::left_join(dplyr::select(cohort, id, treatment), by = "id")

str(dplyr::select(sim, id, time, Cc), max.level = 1)
```

## Steady-state concentration-time profiles

```{r fig-profiles, fig.cap = "Simulated steady-state maribavir profiles over the last 12 h dosing interval of 400 mg twice daily, by health status. Compare with the observed-versus-predicted spread in Sun 2025 Figure 1 (prediction-corrected VPC)."}
sim |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    med = stats::median(Cc),
    lo  = stats::quantile(Cc, 0.05),
    hi  = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time - 228, med, colour = treatment, fill = treatment)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time after dose at steady state (h)",
    y = "Maribavir concentration (ug/mL)",
    colour = NULL, fill = NULL
  ) +
  ggplot2::theme(legend.position = "bottom")
```

# Structural checks (typical value, no random effects)

These are deterministic: `zeroRe()` removes both the between-subject and the
residual variability, so the numbers below depend only on the `ini()` values and
the `model()` algebra. They are the sharpest available test of the covariate
encoding, and they are compared against numbers Sun 2025 reports directly.

```{r structural}
mod_typ <- rxode2::zeroRe(mod)

typ_exposure <- function(ppi = 0, cmv = 1, wt = 70, dose = 400) {
  ev <- rxode2::et(amt = dose, ii = 12, addl = 19, cmt = "depot") |>
    rxode2::et(seq(228, 240, by = 0.01), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt
  d$DOSE <- dose
  d$DIS_CMV <- cmv
  d$CONMED_PPI <- ppi
  d$CONMED_CYP3A4_INH_STRONG <- 0
  d$CONMED_CYP3A4_IND <- 0
  d$STUDY_MARIBAVIR_PHASE1 <- 0
  s <- rxode2::rxSolve(mod_typ, d, returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  c(
    auc  = sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2),
    cmax = max(s$Cc),
    cmin = min(s$Cc)
  )
}

cmv_noppi <- typ_exposure(ppi = 0, cmv = 1)
cmv_ppi   <- typ_exposure(ppi = 1, cmv = 1)
healthy   <- typ_exposure(ppi = 0, cmv = 0)
```

## Steady-state mass balance

At steady state the amount cleared over one dosing interval must equal the
delivered dose, so `AUCss x CL/F = F x Dose` exactly. This is the cheapest gate
that would catch a mis-scaled volume, a wrong concentration unit, or an
`rxode2` `cl`/`vc` pair silently replacing the explicit ODEs with an analytical
solution.

```{r mass-balance}
cl_cmv_70kg <- 3.94 * 0.678
mass_balance <- unname(cmv_noppi["auc"] * cl_cmv_70kg / 400)
mass_balance

stopifnot(abs(mass_balance - 1) < 1e-3)
```

## Covariate effect ratios against the published values

Sun 2025 quantifies three covariate contrasts numerically. Each is reproduced
below from the model algebra alone. The PPI contrast is the strongest test in
this vignette, because it is the *net* result of two separate effects pulling in
different directions -- bioavailability down 9.5% and absorption rate down 54.3%
-- and the paper reports its AUC, Cmax and Cmin consequences separately.

```{r covariate-ratios}
ratios <- tibble::tibble(
  Contrast = c(
    "PPI vs no PPI, AUCss", "PPI vs no PPI, Cmax,ss", "PPI vs no PPI, Cmin,ss",
    "CMV vs healthy, AUCss", "CMV vs healthy, Cmax,ss", "CMV vs healthy, Cmin,ss"
  ),
  Simulated = c(
    cmv_ppi["auc"] / cmv_noppi["auc"],
    cmv_ppi["cmax"] / cmv_noppi["cmax"],
    cmv_ppi["cmin"] / cmv_noppi["cmin"],
    cmv_noppi["auc"] / healthy["auc"],
    cmv_noppi["cmax"] / healthy["cmax"],
    cmv_noppi["cmin"] / healthy["cmin"]
  ),
  Published = c(0.905, 0.775, 1.03, 1.46, 1.20, 2.11),
  Source = c(
    "Figure 2b / Section 3.1.3 (-9.5%)", "Figure 2b / Section 3.1.3 (-22.5%)", "Figure 2b",
    "Section 3.1.3", "Section 3.1.3", "Section 3.1.3"
  )
) |>
  dplyr::mutate(
    Simulated = round(Simulated, 3),
    `% diff`  = round(100 * (Simulated - Published) / Published, 1)
  )

knitr::kable(ratios, caption = "Typical-value covariate contrasts against the values Sun 2025 reports.")

# The PPI AUCss ratio is a pure function of the F effect and must reproduce
# 0.905 essentially exactly; the others involve absorption and distribution and
# are compared against population geometric-mean ratios, so they are given
# proportionally more headroom.
stopifnot(
  abs(ratios$`% diff`[1]) < 0.5,
  max(abs(ratios$`% diff`)) < 5
)
```

# PKNCA validation

Non-compartmental analysis over the steady-state interval 228-240 h, grouped by
treatment arm so each arm can be compared against its own Table 3 row.

```{r pknca}
conc_data <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)

dose_data <- cohort |>
  dplyr::transmute(id, treatment, time = 228, dose = 400)

o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | id / treatment)
# PKNCAdose does not accept slash (nested) grouping, only PKNCAconc does.
o_dose <- PKNCA::PKNCAdose(dose_data, dose ~ time | id + treatment)

intervals <- data.frame(
  start = 228, end = 240,
  auclast = TRUE, cmax = TRUE, cmin = TRUE, half.life = TRUE
)

o_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))

nca_res <- as.data.frame(o_nca) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin", "half.life"))

stopifnot(nrow(nca_res) > 0L, !anyNA(nca_res$PPORRES))
```

```{r pknca-summary}
nca_gm <- nca_res |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    gm  = exp(mean(log(PPORRES))),
    cv  = 100 * sqrt(exp(stats::var(log(PPORRES))) - 1),
    .groups = "drop"
  )

nca_gm |>
  dplyr::mutate(
    Parameter = dplyr::recode(
      PPTESTCD,
      auclast     = "AUCss (ug*h/mL)",
      cmax        = "Cmax,ss (ug/mL)",
      cmin        = "Cmin,ss (ug/mL)",
      half.life   = "t1/2 (h)"
    ),
    `Geometric mean` = round(gm, 3),
    `CV%`            = round(cv, 1)
  ) |>
  dplyr::select(Arm = treatment, Parameter, `Geometric mean`, `CV%`) |>
  knitr::kable(caption = "Simulated steady-state exposure at 400 mg twice daily without concomitant medication.")
```

# Comparison against published NCA

Sun 2025 Table 3 reports geometric mean (CV%) steady-state exposure for the same
regimen and the same two strata, derived from post-hoc Bayesian individual
parameter estimates.

```{r nca-comparison}
reference <- tibble::tribble(
  ~treatment,                       ~PPTESTCD,   ~PPORRES,
  "Healthy volunteers",             "auclast",     97.0,
  "Healthy volunteers",             "cmax",        16.7,
  "Healthy volunteers",             "cmin",         2.57,
  "Healthy volunteers",             "half.life",    4.84,
  "Transplant recipients with CMV", "auclast",    142.0,
  "Transplant recipients with CMV", "cmax",        20.1,
  "Transplant recipients with CMV", "cmin",         5.43,
  "Transplant recipients with CMV", "half.life",    6.68
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by        = "treatment",
  params    = c(
    "AUCss (ug*h/mL)" = "auclast",
    "Cmax,ss (ug/mL)" = "cmax",
    "Cmin,ss (ug/mL)" = "cmin",
    "t1/2 (h)"        = "half.life"
  ),
  label_first_column = "NCA parameter"
) |>
  knitr::kable(caption = "Simulated versus Sun 2025 Table 3 steady-state exposure. Values marked with a star differ by more than 20%.")
```

AUCss and Cmax,ss reproduce the published values closely in both arms (-1.3% to
+10.8%). Note that the two tables in this section aggregate differently:
`ncaComparisonTable()` compares **medians**, whereas the percent-difference
table below compares **geometric means**, which is what Sun 2025 Table 3
reports. The geometric-mean comparison is the like-for-like one.

Two comparisons deserve comment, including the two rows
`ncaComparisonTable()` stars as exceeding 20%.

* **Cmin,ss is the starred metric in both arms** (+21.1% and +29.9% on medians;
  +12.3% and +18.4% on geometric means). This is expected and is not evidence of
  a transcription error, for a specific reason: Table 3's values summarise
  *post-hoc Bayesian* individual estimates, and Table 2 reports shrinkage of
  46.8% for Ka and 53.6% for the absorption lag time -- the two parameters that
  determine the trough. Post-hoc estimates of a heavily shrunk parameter are
  pulled toward the typical value, which raises the trough relative to a
  prospective draw from the full prior such as this one. Consistent with that
  reading, the *typical-value* trough ratios in the Structural checks section
  above -- which involve no sampling at all -- match the published values to
  within 2.5%, and AUCss, whose driving parameter CL/F has only 6.9% shrinkage,
  matches to within a few percent. Cmin,ss is also the most variable metric in
  the source (CV 76-92%).
* **t1/2** is not comparable on equal terms. Sun 2025 derived it from each
  individual's full model-predicted profile, whereas PKNCA estimates lambda-z
  from the terminal points of a single 12 h steady-state interval of a
  two-compartment drug. A 12 h window cannot resolve a terminal phase that the
  model places beyond it, so the NCA half-life is expected to be the *effective*
  rather than the true terminal value. It is reported for completeness, not used
  as a gate.

The assertions below follow the repository convention of gating on the centre
and on robust quantiles rather than on the extreme of a random cohort, because
the per-subject extreme of a simulated cohort is not reproducible across rxode2
versions.

```{r nca-assertions}
check <- nca_gm |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin")) |>
  dplyr::inner_join(reference, by = c("treatment", "PPTESTCD")) |>
  dplyr::mutate(pct_diff = 100 * (gm - PPORRES) / PPORRES)

check |>
  dplyr::select(treatment, PPTESTCD, simulated_gm = gm, published = PPORRES, pct_diff) |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Percent difference between simulated and published geometric mean steady-state exposure.")

stopifnot(
  # AUCss and Cmax,ss are driven by clearance and volume, which are the two
  # best-determined parameters in Table 2; a mis-transcribed value moves these
  # by tens of percent.
  max(abs(check$pct_diff[check$PPTESTCD %in% c("auclast", "cmax")])) < 15,
  # Cmin,ss is absorption- and lag-driven with very high published variability.
  max(abs(check$pct_diff[check$PPTESTCD == "cmin"])) < 30
)
```

# Dose-dependent absorption

The dose effect on Ka is unusual enough to be worth showing on its own: the
exponent is negative, so absorption slows as the dose increases. At 1200 mg the
typical Ka is `(1200/800)^-1.02` = `r round((1200 / 800)^-1.02, 3)` times its
800 mg value, while at 400 mg it is `r round((400 / 800)^-1.02, 3)` times.
Because the model is written on apparent (`/F`) parameters and clearance is
dose-independent, AUCss stays exactly dose-proportional -- consistent with the
paper's statement that maribavir PK is dose-proportional after single doses of
50-1600 mg -- while the peak-to-trough shape changes.

```{r fig-dose, fig.cap = "Typical-value steady-state profiles for a 70 kg transplant recipient with CMV at three dose levels, showing the flattening of the profile at higher doses that the negative dose-on-Ka exponent produces."}
dose_profiles <- lapply(c(400, 800, 1200), function(dd) {
  ev <- rxode2::et(amt = dd, ii = 12, addl = 19, cmt = "depot") |>
    rxode2::et(seq(228, 240, by = 0.05), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- 70
  d$DOSE <- dd
  d$DIS_CMV <- 1
  d$CONMED_PPI <- 0
  d$CONMED_CYP3A4_INH_STRONG <- 0
  d$CONMED_CYP3A4_IND <- 0
  d$STUDY_MARIBAVIR_PHASE1 <- 0
  s <- rxode2::rxSolve(mod_typ, d, returnType = "data.frame")
  s <- s[!is.na(s$Cc), ]
  data.frame(time = s$time - 228, Cc = s$Cc, Dose = paste(dd, "mg BID"))
}) |>
  dplyr::bind_rows()

ggplot2::ggplot(dose_profiles, ggplot2::aes(time, Cc, colour = Dose)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::labs(
    x = "Time after dose at steady state (h)",
    y = "Maribavir concentration (ug/mL)",
    colour = NULL
  ) +
  ggplot2::theme(legend.position = "bottom")
```

# Exposure-response models

Sun 2025 is a two-part paper. Everything above concerns the population PK model.
The second half reports **sixteen logistic exposure-response regressions** fitted
to the 238-patient maribavir arm of the phase 3 AURORA study, and all sixteen are
packaged here as the `Sun_2025_maribavir_*` family. They share one cohort, one
fitting approach and one pair of exposure metrics; each was fitted independently
by the authors, so each is a separate model file, following the same convention
as the `Chen_2021_lorlatinib_*` family.

Two exposure metrics are involved and they are **not interchangeable**:

* `AUC_MBV_SS` -- steady-state AUC over the dosing interval on the last day of
  exposure. Drives the two **efficacy** models (Table S3, main-paper Figure 3).
* `AUC_MBV_DAY` -- AUC over the calendar day on which the adverse event occurred.
  Drives the fourteen **safety** models (Figure S3, Table S4).

Every coefficient is per **10 ug\*h/mL** of exposure, so each model divides its
exposure column by 10 inside `model()`.

```{r er-inventory}
er_models <- c(
  "Sun_2025_maribavir_cmv_clearance_wk8", "Sun_2025_maribavir_cmv_clearance_wk16",
  "Sun_2025_maribavir_dysgeusia", "Sun_2025_maribavir_nausea",
  "Sun_2025_maribavir_vomiting", "Sun_2025_maribavir_diarrhea",
  "Sun_2025_maribavir_neutropenia", "Sun_2025_maribavir_immunosuppressant_increase",
  "Sun_2025_maribavir_infection", "Sun_2025_maribavir_gvhd",
  "Sun_2025_maribavir_renal_disorder", "Sun_2025_maribavir_anemia",
  "Sun_2025_maribavir_pyrexia", "Sun_2025_maribavir_headache",
  "Sun_2025_maribavir_thrombocytopenia", "Sun_2025_maribavir_sae"
)

er_ui <- lapply(er_models, function(nm) rxode2::rxode(readModelDb(nm)))
names(er_ui) <- er_models

er_ini <- function(nm) {
  d <- er_ui[[nm]]$iniDf
  stats::setNames(d$est, d$name)
}

inventory <- do.call(rbind, lapply(er_models, function(nm) {
  p <- er_ini(nm)
  data.frame(
    Model     = sub("^Sun_2025_maribavir_", "", nm),
    Driver    = if ("AUC_MBV_SS" %in% er_ui[[nm]]$allCovs) "AUCss" else "AUCday",
    Intercept = unname(p["logit_ref"]),
    Slope     = unname(p["e_auc_logit"]),
    `Risk factors` = sum(grepl("^e_", names(p))) - 1L,
    check.names = FALSE
  )
}))
knitr::kable(inventory, digits = 4,
             caption = "The sixteen Sun 2025 exposure-response models.")
```

## Source trace: every coefficient against its published odds ratio

Sun 2025 prints, for each coefficient, both an `Estimate (SE)` and an
`OR (95% CI)`, and the Methods define the odds ratio as the exponentiated
estimate. That redundancy is a free consistency check on the transcription, and
it is run mechanically here rather than asserted.

Both printed columns are **rounded**, so the correct test is an *interval* one --
does some real `b` exist with `round(b, d) == Estimate` and
`round(exp(b), 3) == OR`? -- rather than `exp(printed) == printed`, which fails
on rounding alone. The table below lists every coefficient in all sixteen
models, with the published odds ratio transcribed from the source.

```{r er-source-trace}
# Published OR column, transcribed from Sun 2025 Table S3 (efficacy) and the
# Figure S3 parameter tables (safety). Intercepts print "-" for OR and are
# excluded. NA marks a coefficient whose OR the source does not print.
published_or <- tibble::tribble(
  ~model,                        ~param,                       ~or,
  "cmv_clearance_wk8",  "e_auc_logit",                1.01,
  "cmv_clearance_wk8",  "e_te_resist_mbv_logit",      0.0475,
  "cmv_clearance_wk8",  "e_cd8_pp65_mid_logit",      11.1,
  "cmv_clearance_wk8",  "e_cd8_pp65_hi_logit",        1.91,
  "cmv_clearance_wk8",  "e_cd8_pp65_nr_logit",        5.01,
  "cmv_clearance_wk8",  "e_cmvdna_high_logit",        0.350,
  "cmv_clearance_wk8",  "e_hct_tcell_infusion_logit", 0.0575,
  "cmv_clearance_wk8",  "e_region_europe_logit",      0.301,
  "cmv_clearance_wk8",  "e_region_asiapacific_logit", 0.235,
  "cmv_clearance_wk16", "e_auc_logit",                0.987,
  "cmv_clearance_wk16", "e_te_resist_mbv_logit",      0.0299,
  "cmv_clearance_wk16", "e_cd8_pp65_mid_logit",       1.86,
  "cmv_clearance_wk16", "e_cd8_pp65_hi_logit",        2.74,
  "cmv_clearance_wk16", "e_cd8_pp65_nr_logit",        3.04,
  "cmv_clearance_wk16", "e_t_hct_logit",              1.05,
  "cmv_clearance_wk16", "e_region_europe_logit",      0.527,
  "cmv_clearance_wk16", "e_region_asiapacific_logit", 0.225,
  "dysgeusia",          "e_auc_logit",                1.07,
  "dysgeusia",          "e_region_asiapacific_logit", 0.635,
  "dysgeusia",          "e_region_europe_logit",      0.295,
  "nausea",             "e_auc_logit",                1.07,
  "nausea",             "e_cmv_prophy_prior_logit",   0.0508,
  "vomiting",           "e_auc_logit",                1.07,
  "vomiting",           "e_hct_prior_logit",          0.129,
  "diarrhea",           "e_auc_logit",                1.06,
  "diarrhea",           "e_hct_tcell_infusion_logit", 5.72,
  "neutropenia",        "e_auc_logit",                1.05,
  "neutropenia",        "e_cd8_pp65_mid_logit",       0.453,
  "neutropenia",        "e_cd8_pp65_hi_logit",        0.256,
  "neutropenia",        "e_cd8_pp65_nr_logit",        0.986,
  "neutropenia",        "e_wbc_cat_mid_logit",        0.303,
  "neutropenia",        "e_wbc_cat_hi_logit",         0.167,
  "neutropenia",        "e_wbc_cat_nr_logit",         0.123,
  "immunosuppressant_increase", "e_auc_logit",        1.05,
  "immunosuppressant_increase", "e_hct_prior_logit",  6.65,
  "infection",          "e_auc_logit",                1.07,
  "gvhd",               "e_auc_logit",                1.06,
  "gvhd",               "e_sexm_logit",               3.01,
  "renal_disorder",     "e_auc_logit",                1.08,
  "renal_disorder",     "e_cmvdna_high_logit",        0.212,
  "anemia",             "e_auc_logit",                1.07,
  "anemia",             "e_region_asiapacific_logit", 3.12,
  "anemia",             "e_region_europe_logit",      4.62,
  "pyrexia",            "e_auc_logit",                1.05,
  "headache",           "e_auc_logit",                1.05,
  "headache",           "e_hct_tcd_alemtuzumab_logit", 5.84,
  "headache",           "e_hct_tcd_atg_logit",        0.414,
  "headache",           "e_hct_tcd_exvivo_logit",     2.37,
  "thrombocytopenia",   "e_auc_logit",                1.05,
  "thrombocytopenia",   "e_sexm_logit",               3.03,
  "thrombocytopenia",   "e_conmed_antilymphocyte_logit", 0.271,
  "thrombocytopenia",   "e_hct_nuccell_logit",        1.01,
  "sae",                "e_auc_logit",                1.04,
  "sae",                "e_cd8_pp65_mid_logit",       0.160,
  "sae",                "e_cd8_pp65_hi_logit",        0.714,
  "sae",                "e_cd8_pp65_nr_logit",        0.689
)

# Number of decimals the source printed the ESTIMATE to, recovered from the
# packaged value itself, so the interval is the source's own rounding interval.
n_dec <- function(x) {
  s <- format(x, scientific = FALSE, trim = TRUE)
  if (!grepl("[.]", s)) return(0L)
  nchar(sub("^.*[.]", "", sub("0+$", "", s)))
}

or_check <- published_or
or_check$est <- mapply(function(m, p) unname(er_ini(paste0("Sun_2025_maribavir_", m))[p]),
                       or_check$model, or_check$param)
# Interval test: does ANY b consistent with the printed estimate exponentiate
# into the printed OR's own rounding interval?
or_check$ok <- mapply(function(est, or) {
  d  <- n_dec(est)
  lo <- est - 0.5 * 10^(-d); hi <- est + 0.5 * 10^(-d)
  # rounding interval of the printed OR, at its own printed precision
  dor <- n_dec(or)
  or_lo <- or - 0.5 * 10^(-dor); or_hi <- or + 0.5 * 10^(-dor)
  # do [exp(lo), exp(hi)] and [or_lo, or_hi] overlap?
  exp(hi) >= or_lo && exp(lo) <= or_hi
}, or_check$est, or_check$or)

stopifnot(
  # Every packaged estimate must be consistent with the source's own printed
  # odds ratio. A transcription slip in any digit breaks this immediately.
  nrow(or_check) == 56L,
  all(or_check$ok)
)

or_check |>
  dplyr::transmute(
    Model   = model,
    Parameter = sub("^e_|_logit$", "", param),
    Estimate = est,
    `exp(estimate)` = round(exp(est), 4),
    `Published OR` = or,
    Consistent = ifelse(ok, "yes", "NO")
  ) |>
  knitr::kable(caption = paste(
    "All 56 published coefficients across the sixteen models, each checked",
    "against its own printed odds ratio under the interval test."))
```

All 56 coefficients are consistent with their published odds ratios.

## The efficacy relationship is flat (replicates Figure 3)

The paper's headline efficacy result is a **negative** one: neither the primary
nor the key secondary endpoint has a significant relationship with steady-state
maribavir exposure. Packaging the models lets that be shown rather than asserted.
The two fitted slopes do not even agree in sign.

```{r er-figure3, fig.cap = "Replicates Figure 3 of Sun 2025: probability of the primary (a) and key secondary (b) endpoint as a function of steady-state maribavir AUC, for the Table S3 reference patient."}
# Table S3 reference patient: no TE resistance mutation, CD8+CD69+pp65 < 0.5%,
# very low / low baseline CMV DNA, North America, no post-HCT T-cell infusion.
auc_grid <- seq(0, 600, by = 5)

ref_cov <- function(nm) {
  covs <- er_ui[[nm]]$allCovs
  vals <- stats::setNames(rep(0, length(covs)), covs)
  vals[["AUC_MBV_SS"]] <- 0  # overwritten below
  as.list(vals)
}

solve_er <- function(nm, driver, grid, extra = list()) {
  covs <- er_ui[[nm]]$allCovs
  ev <- data.frame(id = seq_along(grid), time = 0, evid = 0L,
                   amt = NA_real_, dv = NA_real_)
  for (cv in covs) ev[[cv]] <- if (!is.null(extra[[cv]])) extra[[cv]] else 0
  ev[[driver]] <- grid
  rxode2::rxSolve(er_ui[[nm]], ev, returnType = "data.frame")
}

fig3 <- dplyr::bind_rows(
  solve_er("Sun_2025_maribavir_cmv_clearance_wk8", "AUC_MBV_SS", auc_grid) |>
    dplyr::transmute(AUCss = AUC_MBV_SS, p = prob_cmv_clearance_wk8,
                     Endpoint = "(a) Primary: clearance at week 8"),
  solve_er("Sun_2025_maribavir_cmv_clearance_wk16", "AUC_MBV_SS", auc_grid) |>
    dplyr::transmute(AUCss = AUC_MBV_SS, p = prob_cmv_clearance_wk16,
                     Endpoint = "(b) Key secondary: maintained to week 16")
)

ggplot2::ggplot(fig3, ggplot2::aes(AUCss, p)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::facet_wrap(~Endpoint) +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(x = "AUCss of maribavir (ug*h/mL)",
                y = "Probability of response") +
  ggplot2::theme_bw()
```

```{r er-flatness}
# Quantify the relationship at three exposures: zero (the intercept anchor), the
# AURORA median of roughly 170 ug*h/mL, and the top of the plotted range.
at <- function(endpoint, auc) fig3$p[fig3$Endpoint == endpoint & fig3$AUCss == auc]
eps <- unique(fig3$Endpoint)

flat <- data.frame(
  Endpoint = eps,
  `P at AUCss = 0`   = vapply(eps, at, numeric(1), auc = 0),
  `P at AUCss = 170` = vapply(eps, at, numeric(1), auc = 170),
  `P at AUCss = 600` = vapply(eps, at, numeric(1), auc = 600),
  `Odds ratio per 10 ug*h/mL` = exp(c(
    er_ini("Sun_2025_maribavir_cmv_clearance_wk8")[["e_auc_logit"]],
    er_ini("Sun_2025_maribavir_cmv_clearance_wk16")[["e_auc_logit"]])),
  `Published OR` = c(1.01, 0.987),
  check.names = FALSE, row.names = NULL
)

stopifnot(
  # Assert on the PUBLISHED quantity -- the odds ratio per 10 ug*h/mL -- rather
  # than on an absolute probability change, which depends on where each
  # intercept sits on the logistic curve and is therefore not the number the
  # paper reports. Both must round to the printed OR.
  round(flat$`Odds ratio per 10 ug*h/mL`, 2)[1] == 1.01,
  round(flat$`Odds ratio per 10 ug*h/mL`, 3)[2] == 0.987,
  # Structural: the two slopes have OPPOSITE signs. A sign error in either
  # transcription breaks this immediately.
  flat$`Odds ratio per 10 ug*h/mL`[1] > 1,
  flat$`Odds ratio per 10 ug*h/mL`[2] < 1,
  # Over the observed exposure range the movement is small in absolute terms.
  abs(flat$`P at AUCss = 170`[1] - flat$`P at AUCss = 0`[1]) < 0.05,
  abs(flat$`P at AUCss = 170`[2] - flat$`P at AUCss = 0`[2]) < 0.10
)

knitr::kable(flat, digits = 3, caption = paste(
  "Both exposure-efficacy odds ratios reproduce their published values, and",
  "the two point estimates lie on opposite sides of 1."))
```

Two things are worth reading carefully off that table, because "flat" is easy to
over-claim. The **odds ratios** are the quantity the paper reports, and both are
essentially 1 (1.01 and 0.987) with p-values of 0.503 and 0.247 -- neither
relationship is statistically significant, and the two point estimates fall on
*opposite* sides of 1, which is the clearest available statement that neither
carries signal.

The **absolute probability movement**, however, is not uniformly negligible, and
the plot above shows why. The primary endpoint's intercept of 1.96 puts its
reference patient at a response probability near 0.88, out on the flat shoulder
of the logistic curve, so even the full 0-600 ug\*h/mL sweep moves it by under
5 percentage points. The key secondary endpoint's intercept of 0.794 puts its
reference patient near 0.69, much closer to the steep middle of the curve, so
the same non-significant slope drags it down by roughly 19 percentage points
across that full range -- and by about 6 points across the exposures actually
observed. A non-significant odds ratio is not the same thing as a visually flat
probability curve, and this pair of endpoints is a clean illustration of the
difference.

For contrast, the same reference-patient sweep over the **safety** endpoints is
steeply positive for all fourteen. The two panels below place the flattest and
steepest safety relationships beside each other.

```{r er-safety-curves, fig.cap = "Probability of each of the fourteen treatment-emergent adverse events as a function of the day-of-event maribavir AUC, for each model's own reference patient. Replicates the fitted curves of Figure S3."}
safety_models <- setdiff(er_models, er_models[1:2])
aucday_grid <- seq(0, 950, by = 10)

safety_curves <- dplyr::bind_rows(lapply(safety_models, function(nm) {
  s <- solve_er(nm, "AUC_MBV_DAY", aucday_grid)
  prob_col <- grep("^prob_", names(s), value = TRUE)
  data.frame(AUCday = s$AUC_MBV_DAY, p = s[[prob_col]],
             Endpoint = sub("^Sun_2025_maribavir_", "", nm))
}))

stopifnot(
  # Every safety endpoint must be monotonically INCREASING in exposure: all
  # fourteen published slopes are positive, so a sign error anywhere breaks this.
  all(vapply(split(safety_curves, safety_curves$Endpoint),
             function(d) all(diff(d$p[order(d$AUCday)]) > 0), logical(1)))
)

ggplot2::ggplot(safety_curves, ggplot2::aes(AUCday, p, colour = Endpoint)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(x = "AUCday of maribavir (ug*h/mL)",
                y = "Probability of the adverse event", colour = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "right", legend.text = ggplot2::element_text(size = 7))
```

## Against the published observed quartile proportions

Figure S3 prints, for each endpoint, the observed event proportion and its 95%
confidence interval within each day-of-event AUC quartile, alongside the fitted
curve. Those observed proportions are a genuinely independent check on the
packaged coefficients: they come from the data, not from the regression. Sun 2025
prints them for panels a and b, which are checked here.

The check has to be posed carefully. A quartile is an interval of exposures, not
a point, and its top quartile is open-ended -- so predicting at a midpoint would
be meaningless. Instead, because each model is monotone in exposure, the
predicted probability for any subject in a quartile must lie between the model's
value at that quartile's lower and upper bound. The observed proportion is a
pooled estimate over patients who differ in the model's covariates, so the
prediction is averaged over the published covariate distribution (region for
dysgeusia; prior CMV prophylaxis for nausea) rather than taken at the reference
stratum. The test is then whether that predicted **range** overlaps the
observed proportion's published **confidence interval**.

```{r er-quartiles}
# Observed proportions, quartile boundaries and 95% CIs: Sun 2025 Figure S3
# panels a and b.
observed <- tibble::tribble(
  ~Endpoint,   ~lo,  ~hi, ~n_resp, ~n_tot, ~ci_lo, ~ci_hi,
  "dysgeusia", 14.1, 116,      2L,    60L,  0.004, 0.115,
  "dysgeusia", 116,  201,     10L,    59L,  0.084, 0.290,
  "dysgeusia", 201,  313,     19L,    59L,  0.206, 0.456,
  "dysgeusia", 313,  783,     30L,    60L,  0.368, 0.632,
  "nausea",      0,  116,      5L,    60L,  0.028, 0.184,
  "nausea",    116,  199,      8L,    59L,  0.060, 0.250,
  "nausea",    199,  309,     18L,    59L,  0.192, 0.439,
  "nausea",    309,  951,     33L,    60L,  0.416, 0.679
)
observed$observed <- observed$n_resp / observed$n_tot

# Published covariate distributions for the 238-patient AURORA arm (Table S2):
# region for dysgeusia, prior CMV prophylaxis for nausea. Each row is one
# stratum with its cohort weight and its covariate settings.
strata <- list(
  dysgeusia = list(
    w    = c(0.248, 0.580, 0.172),
    covs = list(list(REGION_EUROPE = 0, REGION_ASIAPACIFIC = 0),   # North America
                list(REGION_EUROPE = 1, REGION_ASIAPACIFIC = 0),   # Europe
                list(REGION_EUROPE = 0, REGION_ASIAPACIFIC = 1))), # Asia Pacific
  nausea = list(
    w    = c(0.908, 0.092),
    covs = list(list(CMV_PROPHY_PRIOR = 0), list(CMV_PROPHY_PRIOR = 1)))
)

# Cohort-weighted predicted probability at a single exposure.
pred_weighted <- function(endpoint, auc) {
  nm <- paste0("Sun_2025_maribavir_", endpoint)
  st <- strata[[endpoint]]
  ps <- vapply(st$covs, function(cv) {
    s <- solve_er(nm, "AUC_MBV_DAY", auc, extra = cv)
    s[[grep("^prob_", names(s), value = TRUE)]]
  }, numeric(1))
  sum(st$w * ps)
}

observed$pred_lo <- mapply(pred_weighted, observed$Endpoint, observed$lo)
observed$pred_hi <- mapply(pred_weighted, observed$Endpoint, observed$hi)
observed$overlaps <- observed$pred_hi >= observed$ci_lo &
                     observed$pred_lo <= observed$ci_hi

stopifnot(
  # The cohort weights must be a probability distribution.
  all(abs(vapply(strata, function(x) sum(x$w), numeric(1)) - 1) < 1e-8),
  # Monotone in exposure, so the lower bound of the predicted range really is
  # the lower bound.
  all(observed$pred_hi > observed$pred_lo),
  # The observed proportions rise monotonically across quartiles for both
  # endpoints -- the qualitative finding the figure is drawn to show.
  all(vapply(split(observed, observed$Endpoint),
             function(d) all(diff(d$observed) > 0), logical(1))),
  # Every one of the eight published quartiles must be consistent with the
  # packaged model. This is an INDEPENDENT check: the observed proportions and
  # their CIs come from the trial data, not from the regression coefficients.
  all(observed$overlaps)
)

observed |>
  dplyr::transmute(
    Endpoint,
    `AUCday quartile`   = sprintf("[%g, %g)", lo, hi),
    `Observed n/N`      = sprintf("%d/%d", n_resp, n_tot),
    `Observed (95% CI)` = sprintf("%.3f (%.3f-%.3f)", observed, ci_lo, ci_hi),
    `Predicted range`   = sprintf("%.3f-%.3f", pred_lo, pred_hi),
    Consistent          = ifelse(overlaps, "yes", "NO")
  ) |>
  knitr::kable(caption = paste(
    "Published observed event proportions per day-of-event AUC quartile",
    "(Sun 2025 Figure S3 panels a and b) against the cohort-weighted predicted",
    "range from the packaged models. All eight quartiles are consistent."))
```

## End-to-end: the PK model feeds the exposure-response models

The exposure columns are not measurements. Sun 2025 derived them from the
population PK model extracted above, so the two halves of the paper compose.
This section closes that loop: it simulates a 400 mg twice-daily cohort with the
PK model, computes each subject's steady-state AUC by the exact mass-balance
identity used earlier in this vignette, and pushes those AUCs through the primary
efficacy model.


```{r er-endtoend}
rxode2::rxSetSeed(20250911)
set.seed(20250911)

n_er  <- 200
wt_er <- pmin(pmax(stats::rlnorm(n_er, meanlog = log(73.0), sdlog = 0.23), 36.1), 141)

# 400 mg twice daily to steady state, then a dense 24 h observation window.
ev_one <- as.data.frame(
  rxode2::et(amt = 400, ii = 12, until = 24 * 20, cmt = "depot") |>
    rxode2::et(seq(24 * 19, 24 * 20, length.out = 49), cmt = "central")
)
# Keep ii / addl: dropping them would silently reduce the regimen to a
# single dose, and by hour 456 the profile would be empty.
ev_one <- ev_one[, c("time", "cmt", "amt", "ii", "addl", "evid")]

ev_er <- do.call(rbind, lapply(seq_len(n_er), function(i) {
  d <- ev_one
  d$id <- i
  d$WT <- wt_er[i]
  d
}))
ev_er$DOSE                     <- 400
ev_er$DIS_CMV                  <- 1
ev_er$CONMED_CYP3A4_INH_STRONG <- 0
ev_er$CONMED_CYP3A4_IND        <- 0
ev_er$CONMED_PPI               <- 0
ev_er$STUDY_MARIBAVIR_PHASE1   <- 0

sim_er <- rxode2::rxSolve(mod, ev_er, returnType = "data.frame", addDosing = FALSE)

# Steady-state AUC over one 12 h dosing interval: trapezoid over the final 24 h
# window, halved. At steady state the two intervals in that window are identical.
auc_tab <- sim_er |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    AUC_MBV_SS = {
      o <- order(time)
      sum(diff(time[o]) * (utils::head(Cc[o], -1) + utils::tail(Cc[o], -1)) / 2) / 2
    },
    .groups = "drop"
  )

stopifnot(
  # The simulated exposures must land inside the range the exposure-response
  # models were fitted over. Median, not extreme, per this vignette's
  # cohort-assertion convention.
  stats::median(auc_tab$AUC_MBV_SS) > 80,
  stats::median(auc_tab$AUC_MBV_SS) < 400
)

# Push the simulated exposures through the primary efficacy model, holding the
# Table S3 reference patient in every other respect.
eff_nm <- "Sun_2025_maribavir_cmv_clearance_wk8"
eff_ev <- data.frame(id = auc_tab$id, time = 0, evid = 0L,
                     amt = NA_real_, dv = NA_real_,
                     AUC_MBV_SS = auc_tab$AUC_MBV_SS)
for (cv in setdiff(er_ui[[eff_nm]]$allCovs, "AUC_MBV_SS")) eff_ev[[cv]] <- 0
eff_sim <- rxode2::rxSolve(er_ui[[eff_nm]], eff_ev, returnType = "data.frame")

auc_q <- stats::quantile(auc_tab$AUC_MBV_SS, c(0.1, 0.9))
p_q   <- stats::quantile(eff_sim$prob_cmv_clearance_wk8, c(0.1, 0.9))

stopifnot(
  # The point of the exercise. A wide spread in simulated exposure must produce
  # an almost invisible spread in predicted response, because the
  # exposure-efficacy relationship is flat. If a future edit inflated the slope,
  # this bound is what catches it. Deterministic given the drawn cohort only
  # through the exposure spread, which the preceding median bounds pin.
  auc_q[[2]] / auc_q[[1]] > 1.5,
  diff(p_q) < 0.05
)

endtoend <- data.frame(
  Quantity = c("Simulated AUCss, median (ug*h/mL)",
               "Simulated AUCss, 10th-90th percentile",
               "P(clearance at week 8), median",
               "P(clearance at week 8), 10th-90th percentile"),
  Value = c(
    sprintf("%.1f", stats::median(auc_tab$AUC_MBV_SS)),
    sprintf("%.1f - %.1f", auc_q[[1]], auc_q[[2]]),
    sprintf("%.3f", stats::median(eff_sim$prob_cmv_clearance_wk8)),
    sprintf("%.3f - %.3f", p_q[[1]], p_q[[2]])
  )
)

knitr::kable(endtoend, caption = paste(
  "A", sprintf("%.1f-fold", auc_q[[2]] / auc_q[[1]]),
  "spread in simulated steady-state exposure maps to a spread of just",
  sprintf("%.3f", diff(p_q)), "in predicted probability of confirmed CMV",
  "clearance -- the paper's flat exposure-efficacy finding, reproduced",
  "end to end from dose to response."))
```

# Assumptions and deviations

## Structural readings taken from the control stream

* **Two weight exponents rather than four.** Table 2 presents four weight-effect
  rows, but the control stream reuses `CLWT` for Q/F and `VCWT` for Vp/F, with
  `THETA(9)` and `THETA(10)` present but commented `Not used`. Encoding four
  independent exponents would have been a silent structural error that the
  identical Table 2 estimates would never have exposed.
* **Values from the table, structure from the code.** The control stream's
  `$THETA` / `$OMEGA` / `$SIGMA` blocks are initial estimates, not final ones.
  They are close to the published values but not equal to them (`THETA(16)` is
  `-1.17` initially versus a final `-1.02`), so none of them was used as a value.

## Not published, and therefore approximated

* **The three IIV correlations.** Table 2 reports only the six IIV magnitudes; it
  has no correlation column and no off-diagonal rows. The control stream shows
  the *structure* -- three `$OMEGA BLOCK(2)` blocks pairing CL with Vc, Q with
  Vp, and Ka with the lag time -- and the header line
  `;; 2. Description: 3 OMEGA BLOCKs` confirms that structure is final. This
  model file keeps that block structure, puts the published Table 2 variances on
  the diagonals, and carries the correlation **coefficients** (0.673, 0.820 and
  -0.686) from the control stream's own `$OMEGA` initial estimates, rescaled onto
  the published variances. These three covariances are the least well-sourced
  numbers in the file. They were carried rather than invented, and rather than
  set to zero, because the block structure is a published feature of the final
  model and zeroing the off-diagonals would misrepresent it; but a user who
  needs exact published uncertainty should treat the correlations as
  provisional. The magnitudes and the diagonals are unaffected.
* **The `IIV (%)` back-transform convention** is not stated in this paper. See
  the Source trace section for the rule used, its provenance in the same
  analysis group's companion Sun 2023 paper, and the internal-consistency check
  that supports it. Only Vc/F is sensitive to the choice, and only by <2% in SD.

## Exposure-response models: what was and was not extracted

All sixteen fully-parameterised exposure-response regressions are extracted, as
the `Sun_2025_maribavir_*` family validated in the section above. Three
decisions in that extraction are deviations worth recording.

* **Safety versus AUCss (main-paper Figure 4) is NOT extracted, and cannot be.**
  Nausea and vomiting show statistically significant relationships with the
  STEADY-STATE metric (p = 0.0394 and 0.00296), but Figure 4 prints no
  coefficients -- only the p-value, the fitted curve and the observed quartile
  proportions. There is nothing to transcribe, and recovering an intercept and
  slope by digitising a raster figure would manufacture precision the paper
  never published. The packaged `Sun_2025_maribavir_nausea` and
  `Sun_2025_maribavir_vomiting` models are the **AUCday** models of Figure S3,
  which are a different exposure metric and a weaker causal claim: a
  day-of-event AUC is partly a consequence of when in the treatment course the
  event happened. Do not read them as the Figure 4 result.
* **The headache model's fourth T-cell-depletion level is deliberately
  dropped.** Figure S3 panel l reports a `Not reported` level with estimate
  `-12.7` and standard error `1460` (p = 0.993). That is the signature of
  complete separation -- the level has exactly ONE subject in the 238-patient
  cohort (Table S2) -- not an estimate. Shipping it would put a spurious
  `exp(-12.7) = 3e-6` odds ratio into the model file. `HCT_TCD_ALEMTUZUMAB`,
  `HCT_TCD_ATG` and `HCT_TCD_EXVIVO` are packaged; an unreported subject falls
  into the `None` reference.
* **The anemia region coefficients follow Figure S3, not Table S4.** Table S4
  assigns odds ratio 3.12 to "Europe vs. North America" and 4.62 to "Asia
  Pacific vs. North America", while Figure S3 panel j assigns 3.12 to Asia
  Pacific and 4.62 to Europe. The figure is internally consistent and the table
  is not: the figure's estimates exponentiate to its own odds ratios
  (`exp(1.14) = 3.13` for Asia Pacific, `exp(1.53) = 4.62` for Europe), whereas
  Table S4 prints odds ratios only and cannot be checked against itself. Anemia
  is the only endpoint where the two sources conflict.

Two further points affect how the packaged models should be consumed.

* **The exposure columns are inputs, not measurements.** Every model carries
  `AUC_MBV_SS` or `AUC_MBV_DAY` as a covariate with no PK layer, exactly as the
  source analysis did. Derive them from `modellib("Sun_2025_maribavir")`; the
  end-to-end section above shows the composition.
* **`HCT_NUCCELL` has an unstated absolute scale.** Figure S3 panel m prints
  `400 cells  Reference` and a coefficient per `Increment of 1000 cells`, with
  no multiplier and no per-kg normalisation. Read literally that is not a
  plausible graft cell dose -- a real allogeneic graft holds on the order of
  10^8 to 10^10 nucleated cells. The model file carries the source's own numbers
  verbatim rather than guessing a multiplier, so it reproduces the published
  odds ratio whatever the underlying unit turns out to be; a user supplying a
  real cell dose must first rescale it onto the source's scale.
* **The placeholder residual error is not a published quantity.** Each model
  carries `addSd_prob_* <- fixed(0.001)` purely so rxode2 has an error model to
  attach to the typical-value probability. The source fits a Bernoulli
  likelihood, which has no sigma, and estimates no random effects; the models
  are therefore deterministic given their covariates.

## Simulation assumptions

* **Weight distributions** are log-normal draws matched to the Table 1 per-stratum
  mean and SD and truncated to the reported 36.1-141 kg range. Sun 2025 does not
  publish the empirical weight distribution, and Table 1 reports "NR" for the
  AURORA weights specifically.
* **Cohort size** is 200 per arm, the repository cap, against published strata of
  157 and 724 individuals. The comparison is therefore of geometric means and
  CV%, not of individual predictions.
* **Post-hoc versus prospective simulation.** Table 3's values are summaries of
  *post-hoc Bayesian* individual estimates, which are shrunk toward the typical
  value. This simulation draws from the full prior. Shrinkage is low for CL/F
  (6.9%) and moderate for Vc/F (30.8%) but high for Q/F, Vp/F and the lag time
  (54-56%), so the simulated CV% for trough-sensitive metrics is expected to
  exceed the published CV%.
* **No concomitant medication** in any simulated arm, matching the stated
  conditions for Table 3 and Figure 2.
```
