---
title: "Brentuximab vedotin ADC, MMAE payload and exposure-safety in ECHELON-1 (Suri 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brentuximab vedotin ADC, MMAE payload and exposure-safety in ECHELON-1 (Suri 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Source and models

```{r models, include = FALSE}
mod_pk   <- readModelDb("Suri_2019_brentuximab")
mod_pn   <- readModelDb("Suri_2019_brentuximab_pn")
mod_fna  <- readModelDb("Suri_2019_brentuximab_fn_adc")
mod_fnm  <- readModelDb("Suri_2019_brentuximab_fn_mmae")
mod_neut <- readModelDb("Suri_2019_brentuximab_neutropenia")
mod_teae <- readModelDb("Suri_2019_brentuximab_teae")
ui_pk <- rxode2::rxode(mod_pk)
```

* Citation: `r ui_pk$reference`
* Article (open access): <https://doi.org/10.1002/cpt.1530>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6896233/>
* Supporting Information (Figure S1, Tables S1-S4 and the Supplementary
  Supporting Information methods) was retrieved from the Europe PMC
  supplementary-files endpoint for PMC6896233.

Suri 2019 reports the population PK of brentuximab vedotin (antibody-drug
conjugate, ADC) and its payload monomethyl auristatin E (MMAE) in the A+AVD arm
of the phase III ECHELON-1 study, and then relates the individual exposures to
efficacy and safety. This package carries six files:

| File | Content |
|---|---|
| `Suri_2019_brentuximab` | Coupled ADC + MMAE popPK model (Table 2, Figure 1b) |
| `Suri_2019_brentuximab_pn` | Grade >= 2 peripheral neuropathy vs ADC AUC/time (Figure 4b) |
| `Suri_2019_brentuximab_fn_adc` | Febrile neutropenia vs ADC AUC/time and G-CSF (Figure 4c) |
| `Suri_2019_brentuximab_fn_mmae` | Febrile neutropenia vs MMAE AUC/time and G-CSF (Figure 4d) |
| `Suri_2019_brentuximab_neutropenia` | Grade >= 4 neutropenia vs MMAE AUC/time and G-CSF (Figure 4e) |
| `Suri_2019_brentuximab_teae` | Any grade >= 3 TEAE vs MMAE AUC/time and G-CSF (Figure 4f) |

The exposure-efficacy analysis (Cox model for modified PFS, Figure 4a) found no
exposure relationship and reports no coefficients, and the non-significant
exposure-safety pairs (Figure S1) were not retained, so neither is packaged.

# Population

```{r population, echo = FALSE}
pop <- ui_pk$population
knitr::kable(
  data.frame(
    Field = c("Subjects", "Age", "Weight", "BSA", "Female", "Disease", "Dosing", "Observations"),
    Value = c(pop$n_subjects, pop$age_range, pop$weight_range, pop$bsa_range,
              paste0(pop$sex_female_pct, "%"), pop$disease_state, pop$dose_range, pop$n_observations)
  ),
  caption = "Suri 2019 Table 1 and Results (661 PK-evaluable patients)."
)
```

Race: 84.3% White, 3.0% Black, 8.5% Asian, 2.7% other, 1.5% not reported.
Mean (SD) albumin 39.1 (5.3) g/L, Cockcroft-Gault creatinine clearance 134.1
(45.4) mL/min and bilirubin 7.1 (5.4) umol/L (Suri 2019 Table 1).

# Source trace

| Quantity | Value | Source |
|---|---|---|
| ADC structure: 3-compartment, zero-order input, first-order elimination | -- | Results 'ADC PK model'; Figure 1b |
| ADC CL, Vc, Q2, V2, Q3, V3 | 0.0615 L/h, 3.58 L, 0.113 L/h, 3.26 L, 0.0239 L/h, 15.7 L | Table 2 |
| ADC covariates: ALB on CL, BSA on CL, BSA on Vc, sex on Vc, BSA on V3 | -0.477, 1.1, 0.893, 0.934, 1.47 | Table 2 |
| Covariate forms: `(cov / population mean)^theta`; categorical `theta^cov` | -- | Supplement, 'Covariate model development' |
| Reference values BSA / ALB / CrCl | 1.8 m^2 / 39.1 g/L / 134.1 mL/min | Table 1 means |
| ADC IIV (% CV): CL, Vc, V2, Q3, V3 | 19.8, 14.0, 25.5, 41.4, 77.7 | Table 2 |
| IIV scale: reported value = sqrt(omega^2) | omega^2 = (CV/100)^2 | Supplement, 'Statistical model for inter-individual variation' |
| MMAE structure: 2-compartment, formation via target binding and direct conversion through a lag compartment | -- | Results 'MMAE PK model'; Figure 1b |
| MMAE CL, Vc, Q, Vp | 1.45 L/h, 35.5 L, 13.2 L/h, 17.7 L | Table 2 |
| KD, FM, ALFM, Klag | 0.0376 1/h, 1 FIX, 2.35 1/h, 4.51 1/h | Table 2 |
| MMAE covariates: BSA, CrCl, ALB on CL | 1.04, 0.125, 0.0275 | Table 2 |
| MMAE IIV (% CV): CL, Vc, KD | 38.6, 68.6, 101.0 | Table 2 |
| Residual error (LTBS, log-scale SD) ADC / MMAE | 0.181 / 0.395 | Table 2; Supplement 'Statistical model for residual variation' |
| Molar units (amounts umol, concentrations uM) | -- | Figure 3a, 3c axes |
| Direct-conversion time function `1 - exp(-ALFM * tad)` | -- | Not printed; selected by reproducing Results (see below) |
| Logistic model form `logit P = b0 + b1 * EXPOSURE + beta * X` | -- | Supplement, 'Logistic regression model' |
| Logistic coefficients | see each model file | Digitised from Figure 4b-f |

# The MMAE direct-conversion term

Figure 1b names a fraction metabolised (FM, fixed to 1) and a rate constant ALFM
"to describe the decline in direct conversion of ADC to MMAE following time after
dose" (Table 2 footnote), but the paper prints no equation. The literal reading,
`FM * exp(-ALFM * tad)`, is the form used by the related Zhou 2025 pediatric model
(whose control stream is published), and the packaged Suri 2018 model uses it too.
With the ECHELON-1 estimate ALFM = 2.35 1/h, however, that form shuts direct
conversion off within about an hour of every dose. After the first dose the
target pool is depleted, so MMAE formation almost stops.

The paper gives three numbers that decide the question: the cycle-3 (dose 5)
MMAE AUC of 10.0-11.3 ng*day/mL across weight bands, the 49% fall in MMAE
exposure from dose 1 to dose 5, and the Figure 3d profile, in which MMAE
accumulates cycle after cycle with a median peak near 1.3 ng/mL. The chunk below
solves the reference patient (BSA 1.8 m^2, 73.5 kg, 1.2 mg/kg every 2 weeks)
under both readings.

```{r mmae-forms}
mw_adc  <- 153.4   # kDa, brentuximab vedotin
mw_mmae <- 718     # g/mol, MMAE

typ_events <- function(wt = 73.5, n_dose = 5, cov = list(BSA = 1.8, ALB = 39.1, CRCL = 134.1, SEXF = 0)) {
  dose_umol <- min(1.2 * wt, 120) / mw_adc
  dos <- data.frame(id = 1L, time = (seq_len(n_dose) - 1) * 336, evid = 1L, amt = dose_umol,
                    dur = 0.5, cmt = "central", dvid = NA_integer_)
  obs <- data.frame(id = 1L, time = seq(0, n_dose * 336, by = 1), evid = 0L, amt = 0,
                    dur = 0, cmt = NA_character_, dvid = 1L)
  out <- dplyr::bind_rows(dos, obs) |> dplyr::arrange(time, dplyr::desc(evid))
  for (nm in names(cov)) out[[nm]] <- cov[[nm]]
  out
}

auc_window <- function(t, y, a, b) {
  i <- t >= a & t <= b
  sum(diff(t[i]) * (head(y[i], -1) + tail(y[i], -1)) / 2)
}

ev_typ <- typ_events()
mod_literal <- mod_pk |> rxode2::model(fmt <- fm_mmae * exp(-alfm * tafd))

solve_typ <- function(m) {
  s <- as.data.frame(rxode2::rxSolve(m, events = ev_typ, omega = NA, sigma = NA,
                                     useLinCmt = FALSE, returnType = "data.frame"))
  tday <- s$time / 24
  data.frame(
    mmae_auc_d1 = auc_window(tday, s$Cc_mmae * mw_mmae, 0, 14),
    mmae_auc_d5 = auc_window(tday, s$Cc_mmae * mw_mmae, 56, 70),
    mmae_cmax_d5 = max(s$Cc_mmae[s$time >= 1344] * mw_mmae)
  )
}
forms <- dplyr::bind_rows(
  cbind(Form = "Packaged: FM * (1 - exp(-ALFM * tad))", solve_typ(mod_pk)),
  cbind(Form = "Literal: FM * exp(-ALFM * tad)", solve_typ(mod_literal))
) |>
  dplyr::mutate(ratio_d5_d1 = mmae_auc_d5 / mmae_auc_d1)

forms |>
  dplyr::rename(
    "MMAE AUC dose 1 (ng*day/mL)" = mmae_auc_d1,
    "MMAE AUC dose 5 (ng*day/mL)" = mmae_auc_d5,
    "MMAE Cmax dose 5 (ng/mL)" = mmae_cmax_d5,
    "Dose 5 / dose 1" = ratio_d5_d1
  ) |>
  knitr::kable(digits = 3, caption = "Typical-value MMAE exposure under the two readings of the direct-conversion term. Published: dose-5 AUC 10.0-11.3 ng*day/mL; dose 5 / dose 1 = 0.51.")

# Deterministic typical-value solves: tight structural gates are appropriate.
stopifnot(
  forms$mmae_auc_d5[1] > 9, forms$mmae_auc_d5[1] < 13,
  abs(forms$ratio_d5_d1[1] - 0.51) < 0.08,
  forms$mmae_auc_d5[2] < 1   # the literal reading misses the published value ~40-fold
)
```

The packaged form reproduces the published cycle-3 MMAE AUC and the 49% decline
from two independent numbers, with nothing tuned. The literal form misses the
cycle-3 AUC by a factor of about 40. The complementary form also accounts for
where the MMAE comes from: at steady state, converting all eliminated ADC 1:1 to
MMAE gives `dose / CL_MMAE` = `r signif(88.2 / mw_adc * mw_mmae / 1.45 / 24, 3)`
ng*day/mL, which is essentially the published value. The dose-1 excess comes
from the one-time target-binding pathway. A constant fraction (FM = 1 at every
time) cannot be distinguished numerically from the packaged form, because
`1 - exp(-2.35 * tad)` reaches 1 within about two hours. The packaged form keeps
ALFM, which the paper estimated with 0.2% SE, in the model.

# Virtual cohort

200 patients (a single A+AVD arm), with covariates drawn to match Suri 2019
Table 1. Height is not reported, so sex-specific heights are assumed and BSA is
computed with the Mosteller formula; the covariates are otherwise drawn
independently.

```{r cohort}
set.seed(20191201)
rxode2::rxSetSeed(20191201)
n_sub <- 200
clip <- function(x, lo, hi) pmin(pmax(x, lo), hi)
lnorm_draw <- function(n, mean, sd) {
  s2 <- log(1 + (sd / mean)^2)
  exp(rnorm(n, log(mean) - s2 / 2, sqrt(s2)))
}
cohort <- tibble::tibble(
  id   = seq_len(n_sub),
  SEXF = rbinom(n_sub, 1, 0.431),
  WT   = clip(ifelse(SEXF == 1, lnorm_draw(n_sub, 66, 15), lnorm_draw(n_sub, 79, 17)), 40.8, 165.5),
  HT   = ifelse(SEXF == 1, rnorm(n_sub, 163, 7), rnorm(n_sub, 176, 7)),
  BSA  = clip(sqrt(HT * WT / 3600), 1.3, 2.8),
  ALB  = clip(rnorm(n_sub, 39.1, 5.3), 17, 53),
  CRCL = clip(lnorm_draw(n_sub, 134.1, 45.4), 29.2, 476.7),
  GCSF = rbinom(n_sub, 1, 0.5),
  dose_mg = pmin(1.2 * WT, 120)
)
cohort |>
  dplyr::summarise(
    `Weight mean (kg)` = mean(WT), `BSA mean (m^2)` = mean(BSA),
    `Albumin mean (g/L)` = mean(ALB), `CrCl mean (mL/min)` = mean(CRCL),
    `Female (%)` = 100 * mean(SEXF)
  ) |>
  knitr::kable(digits = 2, caption = "Simulated cohort summary; compare Suri 2019 Table 1 (73.5 kg, 1.8 m^2, 39.1 g/L, 134.1 mL/min, 43.1% female).")
```

Each patient receives 1.2 mg/kg (capped at 120 mg) as a 30-minute infusion every
14 days for 12 doses (6 cycles), the ECHELON-1 schedule.

```{r simulate}
n_dose <- 12
dose_rows <- cohort |>
  tidyr::crossing(k = seq_len(n_dose) - 1) |>
  dplyr::transmute(id, time = k * 336, evid = 1L, amt = dose_mg / mw_adc, dur = 0.5,
                   cmt = "central", dvid = NA_integer_)
obs_rows <- cohort |>
  tidyr::crossing(time = seq(0, n_dose * 336, by = 4)) |>
  dplyr::transmute(id, time, evid = 0L, amt = 0, dur = 0, cmt = NA_character_, dvid = 1L)
events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(dplyr::select(cohort, id, SEXF, BSA, ALB, CRCL), by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim <- as.data.frame(rxode2::rxSolve(mod_pk, events = events, useLinCmt = FALSE,
                                     returnType = "data.frame")) |>
  dplyr::mutate(time_d = time / 24, adc_ugml = Cc * mw_adc, mmae_ngml = Cc_mmae * mw_mmae)
```

# Replicating Figure 3b and 3d

```{r fig3, fig.height = 3.6, fig.cap = "Replicates Figure 3b (ADC) and Figure 3d (MMAE) of Suri 2019: simulated median (line) and 95% prediction interval (band) over the first five doses of 1.2 mg/kg every 2 weeks."}
sim |>
  dplyr::filter(time_d <= 70) |>
  tidyr::pivot_longer(c(adc_ugml, mmae_ngml), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = dplyr::recode(analyte, adc_ugml = "ADC (ug/mL)", mmae_ngml = "MMAE (ng/mL)")) |>
  dplyr::group_by(analyte, time_d) |>
  dplyr::summarise(med = median(conc), lo = quantile(conc, 0.025), hi = quantile(conc, 0.975), .groups = "drop") |>
  ggplot(aes(time_d, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey75") +
  geom_line(colour = "firebrick") +
  scale_y_log10() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Days since first dose", y = "Concentration") +
  theme_bw()
```

As in the published panels, the median ADC peak is about 20 ug/mL with no visible
accumulation of the peaks. The MMAE peak is highest after dose 1 and settles to a
lower, repeating profile from dose 2 onwards.

# PKNCA validation

AUC over the first dosing interval (dose 1, days 0-14) and over cycle 3 (dose 5,
days 56-70), plus the whole 12-dose course (days 0-168) for the time-averaged
exposure used by the exposure-response models.

```{r pknca}
arm_lbl <- "A+AVD, BV 1.2 mg/kg Q2W"
dose_nca <- cohort |>
  tidyr::crossing(k = seq_len(n_dose) - 1) |>
  dplyr::transmute(id, time_d = k * 14, dose = dose_mg, arm = arm_lbl)
intervals <- data.frame(start = c(0, 56, 0), end = c(14, 70, 168), auclast = TRUE)

run_nca <- function(conc_col) {
  conc <- sim |>
    dplyr::filter(!is.na(.data[[conc_col]])) |>
    dplyr::transmute(id, time_d, conc = .data[[conc_col]], arm = arm_lbl)
  o_conc <- PKNCA::PKNCAconc(conc, conc ~ time_d | arm + id)
  o_dose <- PKNCA::PKNCAdose(dose_nca, dose ~ time_d | arm + id)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD == "auclast") |>
    dplyr::mutate(window = paste0(start, "-", end)) |>
    dplyr::select(id, window, PPORRES) |>
    tidyr::pivot_wider(names_from = window, values_from = PPORRES)
}
auc_adc  <- run_nca("adc_ugml")  |> dplyr::rename(adc_d1 = `0-14`, adc_d5 = `56-70`, adc_all = `0-168`)
auc_mmae <- run_nca("mmae_ngml") |> dplyr::rename(mmae_d1 = `0-14`, mmae_d5 = `56-70`, mmae_all = `0-168`)
exposure <- cohort |>
  dplyr::left_join(auc_adc, by = "id") |>
  dplyr::left_join(auc_mmae, by = "id") |>
  dplyr::mutate(
    AUC_BV_ADC  = adc_all / n_dose,   # time-averaged, per 14-day interval
    AUC_BV_MMAE = mmae_all / n_dose,
    wt_band   = cut(WT, c(0, 61, 71, 83, 100, Inf), right = FALSE,
                    labels = c("<61 kg", "61-<71 kg", "71-<83 kg", "83-100 kg", ">100 kg")),
    alb_band  = cut(ALB, c(0, 37, 43, Inf), right = FALSE, labels = c("<37 g/L", "37-<43 g/L", ">=43 g/L")),
    crcl_band = cut(CRCL, c(0, 90, Inf), right = FALSE, labels = c("<90 mL/min", ">=90 mL/min"))
  )
```

## Comparison against published exposures

Suri 2019 quotes subgroup cycle-3 AUCs (Results, Figure 2), the ADC accumulation
from cycle 1 to cycle 3, and the MMAE decline from dose 1 to dose 5.

```{r nca-compare}
sub_long <- dplyr::bind_rows(
  exposure |> dplyr::filter(wt_band == "<61 kg") |> dplyr::mutate(subgroup = "Weight <61 kg"),
  exposure |> dplyr::filter(wt_band == "83-100 kg") |> dplyr::mutate(subgroup = "Weight 83-100 kg"),
  exposure |> dplyr::filter(alb_band == "<37 g/L") |> dplyr::mutate(subgroup = "Albumin <37 g/L"),
  exposure |> dplyr::filter(alb_band == ">=43 g/L") |> dplyr::mutate(subgroup = "Albumin >=43 g/L"),
  exposure |> dplyr::filter(crcl_band == ">=90 mL/min") |> dplyr::mutate(subgroup = "CrCl >=90 mL/min")
)
sim_sub <- dplyr::bind_rows(
  sub_long |> dplyr::transmute(subgroup, PPTESTCD = "auclast", PPORRES = adc_d5, analyte = "ADC (ug*day/mL)"),
  sub_long |> dplyr::transmute(subgroup, PPTESTCD = "auclast", PPORRES = mmae_d5, analyte = "MMAE (ng*day/mL)")
) |>
  dplyr::mutate(group = paste(analyte, "-", subgroup)) |>
  dplyr::select(group, PPTESTCD, PPORRES)
ref_sub <- tibble::tribble(
  ~group,                                    ~auclast,
  "ADC (ug*day/mL) - Weight <61 kg",         48.8,
  "ADC (ug*day/mL) - Weight 83-100 kg",      63.6,
  "ADC (ug*day/mL) - Albumin <37 g/L",       49.5,
  "ADC (ug*day/mL) - Albumin >=43 g/L",      59.6,
  "MMAE (ng*day/mL) - Weight <61 kg",        10.0,
  "MMAE (ng*day/mL) - Weight 83-100 kg",     11.3,
  "MMAE (ng*day/mL) - CrCl >=90 mL/min",     11.3
)
cmp <- nlmixr2lib::ncaComparisonTable(sim_sub, ref_sub, by = "group",
                                      units = c(auclast = "per unit shown"))
knitr::kable(cmp, caption = "Cycle-3 (dose 5, days 56-70) AUC: simulated subgroup medians vs Suri 2019 Results / Figure 2.")

ratios <- exposure |>
  dplyr::summarise(adc_accum = median(adc_d5 / adc_d1), mmae_decline = median(mmae_d5 / mmae_d1))
knitr::kable(
  data.frame(
    Quantity = c("ADC AUC dose 5 / dose 1", "MMAE AUC dose 5 / dose 1"),
    Published = c(1.27, 0.51),
    Simulated = c(ratios$adc_accum, ratios$mmae_decline)
  ),
  digits = 2, caption = "Accumulation (Results: ADC 1.27-fold between cycles 1 and 3; MMAE exposure 49% lower at dose 5 than dose 1)."
)

# Cohort medians: centre-of-distribution gates only (CLAUDE.md, cohort assertions).
stopifnot(
  abs(median(exposure$adc_d5) / 58 - 1) < 0.2,
  abs(median(exposure$mmae_d5) / 11 - 1) < 0.25,
  abs(ratios$adc_accum - 1.27) < 0.15,
  abs(ratios$mmae_decline - 0.51) < 0.12
)
```

The subgroup AUCs land within 18% of the published values and carry the same
body-size, albumin and renal trends. The simulated ADC values run about 2-17%
low. To separate the model from the cohort, the chunk below solves
representative typical patients at the published covariate means, with no
random effects:

```{r adc-typical}
typ_adc_d5 <- function(wt, bsa) {
  ev <- typ_events(wt = wt, cov = list(BSA = bsa, ALB = 39.1, CRCL = 134.1, SEXF = 0))
  s <- as.data.frame(rxode2::rxSolve(mod_pk, events = ev, omega = NA, sigma = NA,
                                     useLinCmt = FALSE, returnType = "data.frame"))
  auc_window(s$time / 24, s$Cc * mw_adc, 56, 70)
}
adc_typ <- data.frame(
  Patient = c("55 kg, BSA 1.60 m^2 (weight <61 kg)", "90 kg, BSA 2.05 m^2 (weight 83-100 kg)"),
  Published = c(48.8, 63.6),
  Typical = c(typ_adc_d5(55, 1.60), typ_adc_d5(90, 2.05))
)
knitr::kable(adc_typ, digits = 1, caption = "Typical-value cycle-3 ADC AUC (ug*day/mL) for representative patients vs Suri 2019 Results.")
stopifnot(all(abs(adc_typ$Typical / adc_typ$Published - 1) < 0.1))
```

At the typical value the model is within 1-7% of the published subgroup AUCs.
Dose 5 is about 94% of the way to steady state, and the published values sit
close to the steady-state `dose / CL`. The remaining cohort shortfall comes from
the virtual cohort: its mean BSA (1.86 m^2) is above the published 1.8 m^2,
which raises ADC clearance, and its covariates are drawn independently, whereas
in ECHELON-1 weight, BSA, sex and creatinine clearance are correlated.

## Time-averaged exposure vs the Figure 4a quartiles

Figure 4a divides the efficacy analysis set at ADC AUC/time quartile boundaries of
50.4, 58.2 and 65.8 ug*day/mL. The simulated whole-course time-averaged AUC
(`AUC_BV_ADC`, per 14-day interval) falls about 14% below these boundaries.
That is a larger gap than the cycle-3 comparison above, because the whole-course
average also contains the first, pre-accumulation cycle. How the source defined
the averaging window for efficacy (to progression or censoring) cannot be
reproduced exactly, so this is a level check only.

```{r quartiles}
q_sim <- quantile(exposure$AUC_BV_ADC, c(0.25, 0.5, 0.75))
knitr::kable(
  data.frame(Quartile = c("25th", "50th", "75th"), Published = c(50.4, 58.2, 65.8), Simulated = as.numeric(q_sim)),
  digits = 1, caption = "ADC AUC/time quartile boundaries (ug*day/mL): Suri 2019 Figure 4a vs simulated."
)
stopifnot(abs(q_sim[[2]] / 58.2 - 1) < 0.2)
```

## Sex effect on ADC central volume

The Results state that Vc was about 20% lower in females than in males. That
figure includes the lower BSA of females; the 0.934 sex multiplier alone is a
6.6% reduction.

```{r sex-vc}
vc_by_sex <- sim |>
  dplyr::distinct(id, v1_adc) |>
  dplyr::left_join(dplyr::select(cohort, id, SEXF), by = "id") |>
  dplyr::group_by(SEXF) |>
  dplyr::summarise(vc = median(v1_adc), .groups = "drop")
vc_ratio <- vc_by_sex$vc[vc_by_sex$SEXF == 1] / vc_by_sex$vc[vc_by_sex$SEXF == 0]
vc_ratio
stopifnot(vc_ratio > 0.7, vc_ratio < 0.92)
```

# Exposure-safety models

## Digitised coefficients reproduce Figure 4

The coefficients were digitised from Figure 4b-f: each fitted line was traced at
400 dpi over about 900 pixel columns and fitted with the supplement's logistic
form (linear exposure, additive G-CSF term on the logit). For every G-CSF panel
the no-G-CSF and G-CSF curves, fitted separately, gave slopes within 1% of each
other, which confirms the shared-slope form. The joint fits have
root-mean-square residuals of 0.07-0.26 percentage points. The table checks the
packaged models against probabilities read directly off the traced curves.

```{r er-known-answer}
solve_prob <- function(model, output, auc_col, auc, gcsf = NULL) {
  ev <- data.frame(id = seq_along(auc), time = 0, amt = 0, evid = 0L)
  ev[[auc_col]] <- auc
  if (!is.null(gcsf)) ev$CONMED_GCSF <- gcsf
  as.data.frame(rxode2::rxSolve(model, events = ev, returnType = "data.frame"))[[output]]
}
digitised <- tibble::tribble(
  ~panel, ~gcsf, ~auc, ~p_fig,
  "4b PN (ADC)",          NA, 30, 0.059, "4b PN (ADC)",          NA, 60, 0.105, "4b PN (ADC)",          NA, 90, 0.202,
  "4c FN (ADC)",           0, 30, 0.110, "4c FN (ADC)",           0, 60, 0.206, "4c FN (ADC)",           0, 90, 0.355,
  "4c FN (ADC)",           1, 30, 0.051, "4c FN (ADC)",           1, 60, 0.103, "4c FN (ADC)",           1, 90, 0.196,
  "4d FN (MMAE)",          0,  5, 0.097, "4d FN (MMAE)",          0, 25, 0.489, "4d FN (MMAE)",          1, 25, 0.281,
  "4e Neutropenia (MMAE)", 0,  5, 0.340, "4e Neutropenia (MMAE)", 0, 35, 0.598, "4e Neutropenia (MMAE)", 1, 35, 0.323,
  "4f TEAE (MMAE)",        0,  5, 0.821, "4f TEAE (MMAE)",        1,  5, 0.470, "4f TEAE (MMAE)",        1, 35, 0.809
)
er_map <- list(
  "4b PN (ADC)"           = list(m = mod_pn,   out = "prob_peripheral_neuropathy_grade2", col = "AUC_BV_ADC"),
  "4c FN (ADC)"           = list(m = mod_fna,  out = "prob_febrile_neutropenia",          col = "AUC_BV_ADC"),
  "4d FN (MMAE)"          = list(m = mod_fnm,  out = "prob_febrile_neutropenia",          col = "AUC_BV_MMAE"),
  "4e Neutropenia (MMAE)" = list(m = mod_neut, out = "prob_neutropenia_grade4",           col = "AUC_BV_MMAE"),
  "4f TEAE (MMAE)"        = list(m = mod_teae, out = "prob_teae_grade3",                  col = "AUC_BV_MMAE")
)
digitised$p_model <- vapply(seq_len(nrow(digitised)), function(i) {
  e <- er_map[[digitised$panel[i]]]
  g <- if (is.na(digitised$gcsf[i])) NULL else digitised$gcsf[i]
  solve_prob(e$m, e$out, e$col, digitised$auc[i], g)
}, numeric(1))
digitised |>
  dplyr::rename(Panel = panel, `G-CSF` = gcsf, AUC = auc, `Figure 4 (traced)` = p_fig, `Packaged model` = p_model) |>
  knitr::kable(digits = 3, caption = "Event probability read from the traced Figure 4 curves vs the packaged models.")
stopifnot(max(abs(digitised$p_model - digitised$p_fig)) < 0.01)
```

## Cross-check against the stated G-CSF effect

The Results report that G-CSF primary prophylaxis "significantly reduced the
probability of FN, grade >= 4 neutropenia, and grade >= 3 TEAE by 55-81%". This
number was not used in the digitisation. The odds reductions implied by the
digitised G-CSF coefficients are:

```{r gcsf-or}
gcsf_coef <- c(
  "FN (ADC)"               = rxode2::rxode(mod_fna)$theta[["e_conmed_gcsf_fn"]],
  "FN (MMAE)"              = rxode2::rxode(mod_fnm)$theta[["e_conmed_gcsf_fn"]],
  "Grade >= 4 neutropenia" = rxode2::rxode(mod_neut)$theta[["e_conmed_gcsf_neut"]],
  "Grade >= 3 TEAE"        = rxode2::rxode(mod_teae)$theta[["e_conmed_gcsf_teae"]]
)
red <- 100 * (1 - exp(gcsf_coef))
knitr::kable(data.frame(Endpoint = names(red), `Odds reduction (%)` = red, check.names = FALSE),
             digits = 1, row.names = FALSE)
stopifnot(min(red) > 50, max(red) < 85)
```

The range, 56-81%, matches the stated 55-81% at both ends.

## Replicating Figure 4b-f

```{r fig4, fig.height = 5, fig.cap = "Replicates Figure 4b-f of Suri 2019: probability of each safety event against time-averaged ADC or MMAE AUC, without (red) and with (blue) G-CSF primary prophylaxis."}
grid_adc  <- seq(20, 110, by = 1)
grid_mmae <- seq(2, 41, by = 0.5)
curves <- dplyr::bind_rows(lapply(names(er_map), function(p) {
  e <- er_map[[p]]
  g <- if (e$col == "AUC_BV_ADC") grid_adc else grid_mmae
  if (p == "4b PN (ADC)") {
    return(tibble::tibble(panel = p, auc = g, gcsf = "No", p = solve_prob(e$m, e$out, e$col, g)))
  }
  dplyr::bind_rows(
    tibble::tibble(panel = p, auc = g, gcsf = "No", p = solve_prob(e$m, e$out, e$col, g, 0)),
    tibble::tibble(panel = p, auc = g, gcsf = "Yes", p = solve_prob(e$m, e$out, e$col, g, 1))
  )
}))
ggplot(curves, aes(auc, p, colour = gcsf)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = c(No = "firebrick", Yes = "navy"), name = "Prophylactic G-CSF") +
  facet_wrap(~panel, scales = "free_x") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Time-averaged AUC (ADC ug*day/mL; MMAE ng*day/mL)", y = "Probability of event") +
  theme_bw()
```

## Coupling the PK and exposure-safety models

Using the simulated cohort's time-averaged exposures (half assigned to G-CSF
prophylaxis), the ADC-driven febrile neutropenia model predicts the incidences
below. ECHELON-1 observed 21% without and 11% with primary G-CSF prophylaxis in
the A+AVD arm (Suri 2019 Discussion, citing Connors 2018). The published models
average exposure only up to each patient's event, while this simulation averages
over the whole course, so the comparison checks level, not precision.

```{r fn-incidence}
fn_pred <- exposure |>
  dplyr::mutate(
    p_fn_adc  = solve_prob(mod_fna, "prob_febrile_neutropenia", "AUC_BV_ADC", AUC_BV_ADC, GCSF),
    p_fn_mmae = solve_prob(mod_fnm, "prob_febrile_neutropenia", "AUC_BV_MMAE", AUC_BV_MMAE, GCSF)
  ) |>
  dplyr::group_by(GCSF) |>
  dplyr::summarise(`FN via ADC model` = mean(p_fn_adc), `FN via MMAE model` = mean(p_fn_mmae), .groups = "drop") |>
  dplyr::mutate(Observed = ifelse(GCSF == 1, 0.11, 0.21))
fn_pred |>
  dplyr::rename(`Primary G-CSF` = GCSF) |>
  knitr::kable(digits = 3, caption = "Mean predicted febrile-neutropenia probability vs observed ECHELON-1 incidence.")
stopifnot(all(abs(fn_pred$`FN via ADC model` - fn_pred$Observed) < 0.08))
```

# Assumptions and deviations

* **Direct-conversion time function (structural).** Suri 2019 prints no
  equation for the ADC-to-MMAE direct flux. The packaged
  `FM * (1 - exp(-ALFM * tad))` was selected because it reproduces the paper's
  own cycle-3 MMAE AUC, its 49% dose-1-to-dose-5 decline and the Figure 3d
  profile. The literal reading of the prose, `FM * exp(-ALFM * tad)`, misses the
  AUC about 40-fold (see "The MMAE direct-conversion term"). Nothing was tuned.
  A reviewer who obtains the ECHELON-1 control stream should confirm the form.
* **Molar units.** The paper does not state dosing units, but Figure 3a/3c plot
  both analytes in uM. The target-binding flux `KD * Target * ADC` depends on
  the amount unit, and umol reproduces the published exposures. Molecular
  weights used for conversion: ADC 153.4 kDa, MMAE 718 g/mol.
* **IIV scale.** The supplement reports each IIV as sqrt(omega^2), so
  omega^2 = (CV/100)^2 rather than log(1 + CV^2).
* **Residual error.** LTBS with a homoscedastic additive log-scale error, so
  `lnorm()` with the reported CV used as the log-scale SD.
* **BSA effect on peripheral volume 2.** Table 2 and the supplement place the
  BSA effect on 'peripheral volume 2' (V3 in Table 2's own naming), while the
  Results prose says 'first peripheral volume (V2)'. The table and supplement
  are followed; the Zhou 2025 control stream from the same group also carries
  BSA on V3.
* **Reference BSA 1.8 m^2.** Table 1 prints the mean BSA to one decimal only.
* **Sex coding.** 0.934 is applied to females (SEXF = 1), following the
  supplement's example that the reference subgroup (males) is coded 0. With the
  lower BSA of females, this gives the ~20% lower female Vc the Results state.
* **Exposure-safety coefficients are digitised.** No coefficient is tabulated.
  Values were traced from Figure 4b-f; the known-answer and G-CSF checks above
  bound their accuracy. `AUC_BV_ADC` / `AUC_BV_MMAE` are time-averaged AUCs
  expressed per 14-day interval, which is the scale of the Figure 4 axes. The
  source averages up to each patient's event; this vignette averages over the
  whole course.
* **Placeholder residual on probabilities.** Each logistic model carries a
  `fixed(0.001)` additive residual only so rxode2 accepts an endpoint; the
  source likelihood is Bernoulli.
* **Virtual cohort.** Height and the covariate correlations are not reported.
  Sex-specific heights and weights were assumed, and the covariates were drawn
  independently. G-CSF prophylaxis was assigned to half the cohort for
  illustration.
* **Errata.** No erratum or corrigendum for Suri 2019 was found in the Europe
  PMC record or on the journal page.
