---
title: "Svensson_2016_bedaquiline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Svensson_2016_bedaquiline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
set.seed(20160001)
```

# Model and source

* Citation: Svensson E. M., Dosne A.-G., Karlsson M. O. (2016). Population Pharmacokinetics of Bedaquiline and Metabolite M2 in Patients With Drug-Resistant Tuberculosis: The Effect of Time-Varying Weight and Albumin. *CPT Pharmacometrics Syst Pharmacol* 5(12):682-691. doi:10.1002/psp4.12147.
* DDMORE Foundation Model Repository: [DDMODEL00000219](https://repository.ddmore.eu/model/DDMODEL00000219).
* Article: <https://doi.org/10.1002/psp4.12147>

The packaged model is a three-compartment population PK model for the antimycobacterial bedaquiline (BDQ) coupled to a one-compartment model for its N-desmethyl metabolite M2. Oral absorption goes through two transit compartments. Time-varying body weight (allometric) and serum-albumin (power and free-fraction) effects act on disposition, with additional Black-race and age effects on clearance.

# Population

Svensson 2016 pooled n = 335 adult patients with multidrug-resistant tuberculosis (MDR-TB) from two phase II trials: C208 (placebo-controlled, two-stage) and C209 (open-label). Subjects received oral bedaquiline 400 mg once daily for 2 weeks (loading), followed by 200 mg three times weekly for 22 weeks, on top of an individualized second-line MDR-TB background regimen. Body weight and albumin were measured longitudinally throughout the 24-week treatment phase and 96-week follow-up.

The population was a mix of Asian and Black subjects (study-stage dependent) with median age 32 years (range 18-68), median weight 55-57 kg, and median albumin 3.6-3.8 g/dL increasing toward 4.0-4.1 g/dL at end of treatment (Tables 1 and 2 of Svensson 2016). HIV co-infection was permitted if CD4+ counts exceeded 250 cells/microL.

The same information is available programmatically:

```{r population}
ui <- rxode2::rxode(readModelDb("Svensson_2016_bedaquiline"))
str(ui$meta$population, max.level = 1)
```

# Source trace

Per-parameter origins are recorded as inline comments next to each `ini()` entry in `inst/modeldb/ddmore/Svensson_2016_bedaquiline.R`. The table below summarises the audit trail.

| Parameter / equation | Value | Source |
|---|---|---|
| `lmat` (mean absorption time, h) | log(0.66 × 6) ≈ log(3.97) | THETA(9); Svensson 2016 Table 3 "MAT, fraction of 6 hours" |
| `logitfr` (fraction of MAT in delay) | logit(0.466) | THETA(10); Table 3 "FR" |
| `lcl` (CL/F) | log(2.616) | THETA(11); Table 3 "CL/F = 2.62 L/h" |
| `lvc` (Vc/F) | log(198.34) | THETA(12); Table 3 "V/F = 198 L/70 kg" |
| `lq` (Q1/F) | log(3.658) | THETA(13); Table 3 "Q1/F = 3.66 L/h" |
| `lvp` (Vp1/F) | log(8549.06) | THETA(14); Table 3 "VP1/F = 8550 L/70 kg" |
| `lq2` (Q2/F) | log(7.335) | THETA(15); Table 3 "Q2/F = 7.34 L/h" |
| `lvp2` (Vp2/F) | log(2690.91) | THETA(16); Table 3 "VP2/F = 2690 L/70 kg" |
| `lcl_m2` (CLM2/(F×fm)) | log(10.0496) | THETA(17); Table 3 "CLM2/(F×fm) = 10.0 L/h" |
| `lvc_m2` (Vm2/(F×fm)) | log(2203.71) | THETA(18); Table 3 "VM2/(F×fm) = 2200 L/70 kg" |
| `e_wt_cl` | 0.181 | THETA(19); Table 3 "Allometric scaling clearances = 0.18" |
| `e_wt_vc` | 1.0 (FIX) | THETA(20); Table 3 footnote, allometric volumes fixed to 1 |
| `e_alb_vc` | 1.0 (FIX) | THETA(21); Table 3 "Time varying effect of protein binding on disposition - 1 Fix" |
| `e_alb_cl` | 1.640 | THETA(22); Table 3 "Individual time varying effect of albumin = 1.64" |
| `e_race_black_cl` | 0.839 | THETA(23); Table 3 "Effect of black race on CL/CM2 = 0.84" |
| `e_age_cl` | 0.00881 | THETA(24); Table 3 "Age effect on CL/CLM2 = 0.0088" |
| `lfdepot` | log(1) FIX | derived; the .mod's 1.8002 is a unit conversion (mg→nmol/mL) absorbed into output units |
| Block IIV CL / CLM2 (variances 0.153, 0.135, 0.212) | | .mod $OMEGA BLOCK(2); Table 3 BSV CL = 41 % CV, BSV CLM2 = 49 % CV, ρ = 75 % |
| Diagonal IIV V, Q1, VM2, F, MAT (var 0.172, 0.181, 0.150, 0.0803, 1.16) | | .mod $OMEGA blocks 13-15, 10, 8 |
| `propSd` (BDQ residual) | sqrt(0.0518) ≈ 0.228 | .mod $SIGMA block 3; Table 3 "Proportional residual error BDQ = 23.1 % CV" |
| `propSd_m2` (M2 residual) | sqrt(0.0367) ≈ 0.191 | .mod $SIGMA block 4 diagonal; Table 3 "Proportional residual error M2 = 19.3 % CV" |
| Two-transit-cmt absorption ODE chain | n/a | .mod $DES, depot/transit1/transit2/central |
| 3-cmt BDQ disposition ODEs | n/a | .mod $DES BDQC, BDQPERI1, BDQPERI2 |
| 1-cmt M2 disposition ODE | n/a | .mod $DES M2 |
| Albumin-FU power on `vc_obs` (output only) | n/a | .mod $ERROR `VE = VB*ALLVE*COVFUIP` |

**On the choice of source.** The DDMORE bundle ships an `Output_real_*.lst` listing labelled as the run on the original real dataset. Inspection shows that listing is an `ESTIMATION STEP OMITTED: YES` (i.e. evaluation-only) run: it does not minimise; it reports `R MATRIX ALGORITHMICALLY SINGULAR` and `COVARIANCE STEP ABORTED`; and its `INITIAL ESTIMATE OF THETA` block disagrees with the bundled `.mod`'s `$THETA` for THETA(3), THETA(12), THETA(14), THETA(16), THETA(18), and THETA(24) by factors of 100 to 1000 — implying the listing was generated by a different (internally rescaled) `.mod`. The bundled `.mod`'s `$THETA` block, by contrast, matches Svensson 2016 Table 3 and Table 2 exactly. The packaged model therefore uses the `.mod` `$THETA` (and equivalently the publication's reported point values) as the authoritative source for parameter values; all values were independently confirmed against the published Tables.

# Virtual cohort

Original observed data are not publicly available. The figures below use a virtual population whose covariate distributions approximate the Svensson 2016 demographics. Body weight and albumin are time-varying in the source paper but supplied here as constants per subject (one observation per id) to keep the demonstration self-contained — see Assumptions and deviations for context.

```{r cohort}
n_subjects <- 200L

cohort <- tibble::tibble(
  id         = seq_len(n_subjects),
  WT         = pmax(35, pmin(120, rnorm(n_subjects, mean = 57,  sd = 12))),
  ALB        = pmax(2.0, pmin(5.5, rnorm(n_subjects, mean = 3.7, sd = 0.5))),
  AGE        = pmax(18, pmin(70, rnorm(n_subjects, mean = 32,  sd = 10))),
  RACE_BLACK = rbinom(n_subjects, size = 1, prob = 0.38)
)
summary(cohort)
```

The clinical regimen is 400 mg once-daily for 2 weeks (loading), then 200 mg three-times-weekly for 22 weeks. To keep the simulation tractable the demonstration runs a 14-day single-cohort window covering the initial loading phase plus one washout day (336 h). Sampling times are dense within the first 24 h and sparser thereafter.

```{r events}
loading_doses <- expand.grid(
  id     = cohort$id,
  TIME   = seq(0, 13 * 24, by = 24)
) |>
  dplyr::mutate(
    EVID = 1L,
    AMT  = 400,
    CMT  = "depot"
  )

sample_grid <- c(
  seq(0, 24, by = 0.5),       # day 1 dense
  seq(25, 14 * 24, by = 6)    # remainder sparser
)

samples <- expand.grid(
  id   = cohort$id,
  TIME = sample_grid
) |>
  dplyr::mutate(
    EVID = 0L,
    AMT  = 0,
    CMT  = "Cc"
  )

events <- dplyr::bind_rows(loading_doses, samples) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::arrange(id, TIME, dplyr::desc(EVID)) |>
  dplyr::mutate(DV = NA_real_)

stopifnot(!anyDuplicated(unique(events[, c("id", "TIME", "EVID")])))
nrow(events)
```

# Simulation

```{r simulate-typical}
mod <- rxode2::rxode(readModelDb("Svensson_2016_bedaquiline"))
mod_typical <- rxode2::zeroRe(mod)

sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events,
  returnType = "data.frame",
  keep = c("WT", "ALB", "AGE", "RACE_BLACK")
)
head(sim_typical[, c("id", "time", "Cc", "Cc_m2", "WT", "ALB", "RACE_BLACK")])
```

```{r simulate-vpc}
sim_vpc <- rxode2::rxSolve(
  mod,
  events = events,
  returnType = "data.frame",
  keep = c("WT", "ALB", "AGE", "RACE_BLACK")
)
```

# Replicate published figures

## Bedaquiline and M2 trajectory during loading

```{r figure-bdq-vpc, fig.height = 5}
sim_summary <- sim_vpc |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    BDQ_Q05 = quantile(Cc,    0.05, na.rm = TRUE),
    BDQ_Q50 = quantile(Cc,    0.50, na.rm = TRUE),
    BDQ_Q95 = quantile(Cc,    0.95, na.rm = TRUE),
    M2_Q05  = quantile(Cc_m2, 0.05, na.rm = TRUE),
    M2_Q50  = quantile(Cc_m2, 0.50, na.rm = TRUE),
    M2_Q95  = quantile(Cc_m2, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(-time, names_to = c("analyte", "stat"), names_sep = "_") |>
  tidyr::pivot_wider(names_from = stat, values_from = value)

ggplot(sim_summary, aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y", labeller = ggplot2::as_labeller(c(
    BDQ = "Bedaquiline (Cc)",
    M2  = "Metabolite M2 (Cc_m2)"
  ))) +
  scale_y_log10() +
  labs(
    x = "Time (days after start of treatment)",
    y = "Plasma concentration (mg/L)",
    title = "Loading-phase VPC: bedaquiline and M2",
    caption = paste("Replicates the lower panel of Figure 3 of Svensson 2016 (time after dose 0-24 h",
                    "extended to the loading window 0-14 days) on the simulated cohort.")
  )
```

## Black-race effect on bedaquiline clearance

```{r figure-black-effect, fig.height = 4.5}
typical_70kg <- tibble::tibble(
  id         = 1:2,
  WT         = 70,
  ALB        = 4.04,
  AGE        = 32,
  RACE_BLACK = c(0L, 1L)
)

ev_compare <- expand.grid(
  id   = typical_70kg$id,
  TIME = c(0, sample_grid)
) |>
  dplyr::mutate(EVID = 0L, AMT = 0, CMT = "Cc") |>
  dplyr::bind_rows(
    tibble::tibble(id = typical_70kg$id, TIME = 0, EVID = 1L, AMT = 400, CMT = "depot")
  ) |>
  dplyr::left_join(typical_70kg, by = "id") |>
  dplyr::arrange(id, TIME, dplyr::desc(EVID)) |>
  dplyr::mutate(DV = NA_real_)

sim_compare <- rxode2::rxSolve(
  mod_typical,
  events = ev_compare,
  returnType = "data.frame",
  keep = c("RACE_BLACK")
)

sim_compare |>
  dplyr::mutate(group = ifelse(RACE_BLACK == 1, "Black race", "non-Black reference")) |>
  ggplot(aes(time, Cc, colour = group)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(
    x = "Time after single 400 mg dose (h)",
    y = "Bedaquiline Cc (mg/L)",
    colour = NULL,
    title  = "Effect of Black race on bedaquiline disposition",
    caption = paste("Black patients have ~84 % higher CL (e_race_black_cl = 0.84,",
                    "Svensson 2016 Table 3) and therefore lower exposure for the same dose.")
  )
```

# PKNCA validation

Single-dose NCA on bedaquiline within the first 24 h after the day-1 dose. The treatment grouping variable here is `RACE_BLACK` so per-group Cmax / AUC0-24 can be inspected; in a real validation a clinically meaningful stratification (dose group, regimen) would be used.

```{r pknca}
sim_day1 <- sim_vpc |>
  dplyr::filter(time >= 0, time <= 24) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(treatment = ifelse(RACE_BLACK == 1, "Black", "non-Black")) |>
  dplyr::select(id, time, Cc, treatment)

dose_df_day1 <- events |>
  dplyr::filter(EVID == 1, TIME == 0) |>
  dplyr::mutate(treatment = ifelse(RACE_BLACK == 1, "Black", "non-Black")) |>
  dplyr::select(id, time = TIME, amt = AMT, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_day1,      Cc  ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df_day1,  amt ~ time | treatment + id)

intervals <- data.frame(
  start    = 0,
  end      = 24,
  cmax     = TRUE,
  tmax     = TRUE,
  auclast  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
nca_summary <- summary(nca_res)
nca_summary
```

The numeric values above are the simulated cohort's NCA parameters per `RACE_BLACK` stratum. The publication does not report a paired NCA table — it uses VPCs and a covariate-effect plot (Figure 4) for goodness-of-fit. Instead, the Black-vs-non-Black AUC0-24 ratio is the load-bearing comparison: an effect coefficient of 0.84 on CL implies an AUC ratio (Black / non-Black) of `1 / (1 + 0.84) ≈ 0.54`, i.e. Black subjects should have ~46 % lower AUC0-24 at typical 70 kg / 4.04 g/dL / 32 y. The simulated cohort is mixed across covariates, so the observed ratio also reflects the covariate distribution; the deterministic-typical comparison in the prior chunk demonstrates the underlying covariate effect cleanly.

# Assumptions and deviations

The packaged model differs from the published Svensson 2016 NONMEM control stream in the following ways. None of these affect the typical-value PK predictions for users who supply ALB and WT as time-varying covariates equivalent to those the publication's semi-physiological models would predict.

- **Albumin and weight time courses are not embedded in the model.** The publication fits a self-limiting logistic albumin model (A0 → Ass with T1/2_return) and a linear weight model (WT0 → WT120 over 120 weeks) simultaneously with the PK and uses individually predicted time-varying ALB and WT inside the disposition equations. The packaged model expects the user to supply ALB and WT as time-varying columns in the dataset (or as constants if a fixed-baseline simulation suffices). The published albumin model parameters (A0 = 3.65 g/dL, Ass = 4.04 g/dL, T1/2_return = 20.4 weeks per Table 2) and weight model parameters (WT0 = 56.6 kg, WT120 = 62.6 kg, linear) are reported in Table 2 for users who wish to derive equivalent trajectories externally.
- **Box-Cox transformations on baseline-albumin / steady-state-albumin / 120-week-weight IIVs are dropped** (the relevant IIVs are dropped along with the albumin and weight sub-models above).
- **Between-occasion variability (BOV) on F and MAT is dropped.** The published model splits study C208 into two occasions (one per period of rich PK sampling) with BOV variances 0.0382 (F) and 1.16 (MAT). For library-scale simulation the packaged model retains only the BSV component of MAT (variance 1.16, identical magnitude to the BOV) and drops the BOV-on-F entirely. Users running multi-occasion designs should add OCC handling externally.
- **BSV on residual error (BSVRUV1, BSVRUV2 in the .mod) is dropped.** The published model includes etas on the residual SD itself (var 0.0539 / 0.0523, correlation 56 %); the packaged model uses fixed proportional SDs equal to the population CV (sqrt(SIGMA) = 22.8 % for BDQ, 19.1 % for M2). This loses inter-individual variability in residual variability but preserves the population-mean residual magnitude.
- **Cross-output residual correlation (BDQ vs M2 residuals, ρ = 43 %) is dropped.** nlmixr2 has no idiomatic encoding for cross-output residual correlation in this model class; the packaged residuals are independent.
- **The unit-conversion factor 1.8002 in the .mod's `F1 = 1.8002 * EXP(BOVF + BSVF)` is dropped.** The .mod expresses concentrations in nmol/mL = µmol/L; the packaged model uses mg/L (= µg/mL). To recover the published nmol/mL output, multiply Cc and Cc_m2 by `1000 / MW`, with MW (bedaquiline) = 555.5 g/mol → factor 1.8002, and MW (M2) ≈ 541.5 g/mol → factor 1.847.
- **The .mod's `V_PK` vs `V_obs` distinction is preserved.** The albumin-driven free-fraction adjustment (THETA(21) = 1 FIX) acts only on the observation equations (`Cc <- central / vc_obs`, `Cc_m2 <- central_m2 / vc_m2_obs`), not on the ODE clearance terms (`d/dt(central) <- ... -cl*central/vc -...`). This matches the source `.mod`'s `V` (in $PK / $DES) vs `VE` (in $ERROR) bookkeeping. It is a deliberate model-level inconsistency in the publication, retained here for fidelity.
- **Final estimates are taken from the bundled `.mod` `$THETA` rather than the bundled `Output_real_*.lst`.** The `.lst` is an evaluation-only run with a different (internally rescaled) `.mod` driving its FINAL ESTIMATE block, so its values disagree with the publication for several thetas. The `.mod` `$THETA` matches Svensson 2016 Tables 2 and 3 exactly and is the authoritative source.
- **Population race breakdown.** Table 1 reports per-stage race counts (e.g., C209 was 38.2 % Asian, 38.2 % Black, ~24 % Other). The pooled percentages above are approximate; consult the source Table for cohort-specific numbers.
