---
title: "Cefotaxime (Swartling 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefotaxime (Swartling 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Swartling_2022_cefotaxime"))$reference`
* Description: `r rxode2::rxode(readModelDb("Swartling_2022_cefotaxime"))$description`
* Article: <https://doi.org/10.1007/s00228-021-03218-6> (open access, PMC8748331)
* Online Resource 2 (final NONMEM control stream) and Online Resource 1
  (MAP-estimation precision, Table S1) are published with the article.

```{r model}
mod <- rxode2::rxode(readModelDb("Swartling_2022_cefotaxime"))
mod
```

# Population

Swartling 2022 is a sub-study of ACCIS (Antibiotic Concentrations in Critical
Ill ICU Patients in Sweden, ACTRN12616000167460), a prospective observational
multi-centre PK study run at seven ICUs - one of them a burn unit - in five
Swedish hospitals between December 2015 and July 2017. Adults treated with
intravenous cefotaxime for a proven or suspected infection were included within
24 h of starting treatment; patients on renal replacement therapy were excluded
from this analysis.

Fifty-one patients (18 in the burn unit, 33 in general ICUs) contributed **263
total serum cefotaxime concentrations**, a median of 6 per patient, drawn at
mid-interval and just before the next dose on up to three consecutive days from
the first day of treatment. Doses were 1000-3000 mg as a 5-min infusion, 2-6
times daily, most often 1000 mg t.i.d. and then 2000 mg t.i.d. Table 1 reports a
median age of 64 years (range 23-90), 35% female, body weight at ICU admission
92 kg (IQR 81-102, range 55-124), SAPS3 55 (range 25-89) and Cockcroft-Gault
estimated creatinine clearance (eCLcr) 94 mL/min over days 1-3 (IQR 65-138,
range 5-258).

```{r population}
str(readModelDb("Swartling_2022_cefotaxime")()$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The table
below collects them. "OR2" is Online Resource 2, the final NONMEM control stream.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment IV model, linear elimination | `central` + `peripheral1` | Results 'Pharmacokinetic modelling'; OR2 `ADVAN3 TRANS4`, `S1 = V1` |
| `lcl` (CL) | 11.1 L/h (RSE 8.2%) | Table 2 |
| `lvc` (Vc) | 5.13 L (RSE 28%) | Table 2 |
| `lvp` (Vp) | 18.2 L (RSE 12%) | Table 2 |
| `lq` (Q) | 14.5 L/h (RSE 19%) | Table 2 |
| `e_wt_cl_q` | 0.75, fixed | Table 2 footnotes a, d; OR2 `(BWAI/92)**0.75` |
| `e_wt_vc_vp` | 1, fixed | Table 2 footnotes b, c; OR2 `(BWAI/92)` |
| Weight reference 92 kg | cohort median | Table 1 (weight at ICU admission); Table 2 footnotes |
| `e_crcl_cl` | 6.67 per 1000 mL/min (RSE 17%) | Table 2 'theta cov (eCLcr <= 120)'; footnote a; OR2 `THETA(7)*((COGBWA-94)/1000)` |
| eCLcr centring 94 mL/min | cohort median | Table 1; Table 2 footnote a; OR2 |
| `crcl_cap` | 120 mL/min, fixed | Results ('fixed slope of zero at values above 120 mL/min ... equals truncating eCLcr'); Table 2 footnote a; OR2 `IF(COGBWA.GT.120)` branch |
| `scale_etalvc` | 1.31 (RSE 39%) | Table 2 'f CL,Vc'; OR2 `V1 = TVV1*EXP(ETA(1)*THETA(6))` |
| `etalcl` | 0.2401 = 0.49^2 | Table 2 'IIV CL (% CV)' 49 (shrinkage 0.8%); scale resolved below |
| Shared eta on CL and Vc | - | Results; OR2 `CL = TVCL*EXP(ETA(1))`, single `$OMEGA` |
| `propSd` | 0.333 | Table 2 'Proportional residual error (% CV)' 33.3; OR2 `W = IPRED*THETA(5)`, `$SIGMA 1 FIX` |
| Typical CL at eCLcr 0: 4.2 L/h; 0.74 L/h per 10 mL/min | - | Results, Figure 3 |

## Resolving the %CV convention for the IIV

Table 2 reports IIV as '% CV' - 49% on CL and 64% on Vc - and footnote b states
that the Vc value 'was derived as CV for CL times the estimated f CL,Vc'. The
model multiplies the eta itself by the factor, so that multiplication is exact
only if the '% CV' is the standard deviation of eta (`sqrt(omega^2)`). Under the
exact log-normal form `CV = sqrt(exp(omega^2) - 1)` the same model would print a
different Vc CV:

```{r cv-convention}
cvCl <- 0.49
fVc <- 1.31
cvTable <- data.frame(
  convention = c(
    "CV% = 100 * sqrt(omega^2)",
    "CV% = 100 * sqrt(exp(omega^2) - 1)"
  ),
  omega2_cl = c(cvCl^2, log(1 + cvCl^2))
)
cvTable$vc_cv_pct <- c(
  100 * fVc * sqrt(cvTable$omega2_cl[1]),
  100 * sqrt(exp(fVc^2 * cvTable$omega2_cl[2]) - 1)
)
cvTable |>
  dplyr::rename(
    "Convention" = convention,
    "omega^2 on CL" = omega2_cl,
    "Implied Vc CV (%)" = vc_cv_pct
  ) |>
  knitr::kable(digits = 3)
# Only the SD reading reproduces the printed 64%.
stopifnot(
  round(cvTable$vc_cv_pct[1]) == 64,
  round(cvTable$vc_cv_pct[2]) != 64
)
```

Only the first convention gives the printed 64%, and the proportional residual
row in the same table is also a standard deviation (the control stream estimates
it as `THETA(5)` with `$SIGMA 1 FIX`). The model therefore uses
`omega^2 = 0.49^2 = 0.2401`.

# Replicate Figure 3: clearance versus estimated creatinine clearance

Figure 3 draws the typical clearance of a 92-kg patient against eCLcr: a line
from 4.2 L/h at eCLcr 0 rising 0.74 L/h per 10 mL/min to a plateau at 120
mL/min. The model's typical clearance is solved on a grid of eCLcr values with
the random effects zeroed.

```{r fig3}
modTyp <- rxode2::zeroRe(mod)
crclGrid <- c(0, seq(5, 250, by = 5))
evFig3 <- data.frame(
  id = seq_along(crclGrid),
  time = 0,
  evid = 0,
  amt = 0,
  cmt = "central",
  WT = 92,
  CRCL = crclGrid
)
fig3 <- rxode2::rxSolve(modTyp, evFig3, returnType = "data.frame")
fig3$CRCL <- crclGrid

ggplot(fig3, aes(CRCL, cl)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = 120, linetype = "dashed", colour = "grey50") +
  coord_cartesian(ylim = c(0, 45)) +
  labs(
    x = "eCLcr (mL/min)",
    y = "Typical cefotaxime CL (L/h)",
    caption = "Replicates the model line of Figure 3 of Swartling 2022 (body weight 92 kg)."
  ) +
  theme_bw()

clAt <- function(x) fig3$cl[fig3$CRCL == x]
fig3Check <- data.frame(
  quantity = c(
    "CL at eCLcr 94 mL/min (L/h)",
    "CL at eCLcr 0 (y-intercept, L/h)",
    "Change in CL per 10 mL/min below 120 (L/h)",
    "CL at eCLcr 120 and above (plateau, L/h)"
  ),
  paper = c(11.1, 4.2, 0.74, NA),
  model = c(
    stats::approx(fig3$CRCL, fig3$cl, 94)$y,
    clAt(0),
    (clAt(100) - clAt(50)) / 5,
    clAt(200)
  )
)
fig3Check |>
  dplyr::rename(
    "Quantity" = quantity,
    "Swartling 2022" = paper,
    "Model" = model
  ) |>
  knitr::kable(digits = 3)

stopifnot(
  abs(fig3Check$model[1] - 11.1) < 1e-6,
  # 11.1 * (1 - 6.67 * 94 / 1000) = 4.14; the paper's 4.2 is from unrounded
  # estimates, so a 0.1 L/h tolerance is the rounding band, not slack.
  abs(fig3Check$model[2] - 4.2) < 0.1,
  abs(fig3Check$model[3] - 0.74) < 0.005,
  # Flat above the 120 mL/min cap.
  abs(clAt(120) - clAt(250)) < 1e-9
)
```

The Figure 3 plateau reads about 13 L/h, matching the model's
`r round(clAt(200), 2)` L/h.

# Virtual cohort

Covariates are drawn to match Table 1. Body weight at ICU admission is normal
(median 92 kg, SD 15.6 kg from the IQR 81-102) and eCLcr log-normal (median 94
mL/min, log-SD 0.56 from the IQR 65-138). Values outside the observed ranges
(55-124 kg; 5-258 mL/min) are redrawn, not clamped, so the cohort keeps its
shape inside the band. eCLcr is held constant within a patient; the model
accepts a time-varying `CRCL` column, as the paper's data had.

```{r cohort}
N_ARM <- 200L
set.seed(20220101)
drawBand <- function(n, draw, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- draw(n)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}
makeCohort <- function(n) {
  data.frame(
    WT = drawBand(n, function(k) stats::rnorm(k, 92, 15.6), 55, 124),
    CRCL = drawBand(n, function(k) exp(stats::rnorm(k, log(94), 0.56)), 5, 258)
  )
}
cohort <- makeCohort(N_ARM)
summary(cohort)
```

# Replicate Figure 2: concentrations over three days of treatment

Figure 2 is a prediction-corrected VPC of mid-interval and trough samples
against time after dose. Two regimens are simulated for the cohort - the two
most common, 1000 mg and 2000 mg every 8 h as 5-min infusions - with samples
at mid-interval (4 h) and trough (8 h) after every dose over three days,
including residual error.

```{r sim-vpc}
DOSE_TIMES <- seq(0, 64, by = 8)
makeEvents <- function(amt, idOffset) {
  ids <- seq_len(N_ARM) + idOffset
  doses <- expand.grid(time = DOSE_TIMES, id = ids)
  doses$evid <- 1L
  doses$amt <- amt
  doses$rate <- amt / (5 / 60)
  obs <- expand.grid(time = c(DOSE_TIMES + 4, DOSE_TIMES + 8), id = ids)
  obs$evid <- 0L
  obs$amt <- 0
  obs$rate <- 0
  ev <- rbind(doses, obs)
  ev$cmt <- "central"
  ev$WT <- cohort$WT[ev$id - idOffset]
  ev$CRCL <- cohort$CRCL[ev$id - idOffset]
  ev$regimen <- paste(amt, "mg q8h")
  ev[order(ev$id, ev$time, -ev$evid), ]
}
events <- rbind(makeEvents(1000, 0L), makeEvents(2000, N_ARM))

rxode2::rxSetSeed(20220102)
simVpc <- rxode2::rxSolve(mod, events, returnType = "data.frame", keep = "regimen")
simVpc <- simVpc |>
  dplyr::mutate(
    tad = ifelse(time %% 8 == 0, 8, time %% 8),
    day = floor((time - 1e-9) / 24) + 1
  ) |>
  dplyr::filter(time <= 72)

vpcSummary <- simVpc |>
  dplyr::group_by(regimen, tad) |>
  dplyr::summarise(
    p10 = stats::quantile(sim, 0.10),
    p50 = stats::median(sim),
    p90 = stats::quantile(sim, 0.90),
    .groups = "drop"
  )

# Observed percentiles read by the maintainers from Figure 2 of Swartling 2022
# (prediction-corrected, all regimens pooled); approximate raster readings.
figure2Obs <- data.frame(
  tad = c(4, 8, 4, 8, 4, 8),
  conc = c(7, 1.7, 2.4, 0.55, 17, 8),
  statistic = rep(c("median", "10th percentile", "90th percentile"), each = 2)
)

ggplot(vpcSummary, aes(tad)) +
  geom_ribbon(aes(ymin = p10, ymax = p90, fill = regimen), alpha = 0.25) +
  geom_line(aes(y = p50, colour = regimen), linewidth = 1) +
  geom_point(
    data = figure2Obs, aes(tad, conc, shape = statistic),
    size = 3, inherit.aes = FALSE
  ) +
  scale_y_log10() +
  scale_x_continuous(breaks = c(4, 8)) +
  labs(
    x = "Time after dose (h)",
    y = "Cefotaxime serum concentration (mg/L)",
    fill = "Simulated 10th-90th percentile",
    colour = "Simulated median",
    shape = "Figure 2 observed",
    caption = "Replicates Figure 2 of Swartling 2022 (points: observed percentiles read from the figure)."
  ) +
  theme_bw()

knitr::kable(
  vpcSummary |>
    dplyr::rename(
      "Regimen" = regimen,
      "Time after dose (h)" = tad,
      "10th percentile (mg/L)" = p10,
      "Median (mg/L)" = p50,
      "90th percentile (mg/L)" = p90
    ),
  digits = 2
)
```

The Figure 2 observed median (about 7 mg/L at mid-interval and 1.7 mg/L at
trough) lies between the simulated 1000 mg and 2000 mg medians, as expected for
a cohort treated mostly with those two regimens.

## Trough concentrations by treatment day

Table 1 reports observed troughs (samples drawn within 1 h before the next
dose), pooled over regimens: median 1.8 mg/L overall, 2.6 on day 1, 2.2 on day
2 and 1.4 on day 3.

```{r troughs}
troughs <- simVpc |>
  dplyr::filter(tad == 8, day <= 3) |>
  dplyr::group_by(regimen, day) |>
  dplyr::summarise(
    median = stats::median(sim),
    q25 = stats::quantile(sim, 0.25),
    q75 = stats::quantile(sim, 0.75),
    .groups = "drop"
  )
table1 <- data.frame(
  day = 1:3,
  obs = c("2.6 (0.9-7.5)", "2.2 (0.8-4.4)", "1.4 (0.7-2.8)")
)
troughs |>
  dplyr::left_join(table1, by = "day") |>
  dplyr::rename(
    "Regimen" = regimen,
    "Treatment day" = day,
    "Simulated median (mg/L)" = median,
    "Simulated Q1 (mg/L)" = q25,
    "Simulated Q3 (mg/L)" = q75,
    "Table 1 observed median (IQR), all regimens" = obs
  ) |>
  knitr::kable(digits = 2)

pooledMedian <- stats::median(simVpc$sim[simVpc$tad == 8])
# The pooled simulated trough median for the two dominant regimens sits
# close to the observed 1.8 mg/L. A mis-transcribed CL, volume or unit moves
# it several-fold; the 2-fold band admits the unknown regimen mix and the
# per-cohort draw.
stopifnot(pooledMedian > 1.8 / 2, pooledMedian < 1.8 * 2)
```

The pooled simulated trough median is `r round(pooledMedian, 2)` mg/L against the
observed 1.8 mg/L. The observed decline from day 1 to day 3 is not reproduced:
the model holds eCLcr constant in each virtual patient, whereas Table 1 shows
eCLcr rising from a median 85 mL/min on day 1 to 105 on day 3, and the paper
reports that the treatment day itself was not retained as a covariate.

# PKNCA validation at steady state

A typical 92-kg patient with eCLcr 94 mL/min is dosed every 8 h for three days
at 1000 mg and 2000 mg (5-min infusions). NCA is run on the final dosing
interval (64-72 h), where the model is at steady state.

```{r nca}
ncaTimes <- sort(unique(c(64, 64 + c(5 / 60, 0.25, 0.5, 1, 1.5, 2, 3, 4, 5, 6, 7, 8))))
makeTypical <- function(amt, id) {
  doses <- data.frame(id = id, time = DOSE_TIMES, evid = 1L, amt = amt, rate = amt / (5 / 60))
  obs <- data.frame(id = id, time = c(0, ncaTimes), evid = 0L, amt = 0, rate = 0)
  ev <- rbind(doses, obs)
  ev$cmt <- "central"
  ev$WT <- 92
  ev$CRCL <- 94
  ev$treatment <- paste(amt, "mg q8h")
  ev[order(ev$time, -ev$evid), ]
}
evTyp <- rbind(makeTypical(1000, 1L), makeTypical(2000, 2L))
simTyp <- rxode2::rxSolve(
  modTyp, evTyp,
  returnType = "data.frame", keep = "treatment",
  rtol = 1e-10, atol = 1e-12
)

concNca <- simTyp |>
  dplyr::filter(!is.na(Cc), time >= 64 | time == 0) |>
  dplyr::mutate(Cc = pmax(Cc, 0)) |>
  dplyr::select(id, time, Cc, treatment)
doseNca <- evTyp |>
  dplyr::filter(evid == 1, time == 64) |>
  dplyr::select(id, time, amt, treatment)

concObj <- PKNCA::PKNCAconc(concNca, Cc ~ time | treatment + id)
doseObj <- PKNCA::PKNCAdose(doseNca, amt ~ time | treatment + id)
intervals <- data.frame(
  start = 64, end = 72,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)
ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))
ncaTab <- as.data.frame(ncaRes$result) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
ncaTab |>
  dplyr::rename(
    "Regimen" = treatment,
    "Cmax (mg/L)" = cmax,
    "Tmax (h)" = tmax,
    "Cmin (mg/L)" = cmin,
    "AUC0-8 (mg*h/L)" = auclast
  ) |>
  knitr::kable(digits = 2)

# Mass balance at steady state: CL * AUCtau = dose. Checks the typical CL and
# the dose / volume units through the full ODE path. The trapezoid on the NCA
# grid is the dominant error (the peak is only 5 min wide), measured about 1%.
auc <- ncaTab$auclast[match(c("1000 mg q8h", "2000 mg q8h"), ncaTab$treatment)]
massBalance <- 11.1 * auc / c(1000, 2000)
massBalance
stopifnot(all(abs(massBalance - 1) < 0.05))
```

The paper reports no NCA parameters, so there is no published NCA table to
compare against; the model-predicted Cmax and trough above are consistent with
the Figure 2 mid-interval and trough percentiles.

# Assumptions and deviations

* **IIV scale.** Table 2 prints IIV as '% CV'. The model uses `omega^2 = 0.49^2`
  because only the standard-deviation reading reproduces the printed Vc CV
  (49 x 1.31 = 64); see "Resolving the %CV convention" above.
* **eCLcr cap as a fixed parameter.** The paper's piecewise-linear covariate
  with a slope fixed to zero above 120 mL/min is encoded as
  `min(CRCL, crcl_cap)` with `crcl_cap <- fixed(120)`, which the paper states is
  equivalent. The breakpoint was chosen by stepwise evaluation of candidate
  breakpoints, not estimated, so it is fixed.
* **Allometric exponents** 0.75 (CL, Q) and 1 (Vc, Vp) are hard-coded in the
  control stream and are encoded as `fixed()` parameters.
* **Covariate definitions.** `WT` is body weight AT ICU ADMISSION (not the
  pre-admission weight), and `CRCL` is the raw Cockcroft-Gault estimate in
  mL/min computed with that weight - not BSA-normalized. Supplying a
  BSA-normalized eGFR would rescale the covariate effect.
* **Virtual cohort.** Weight and eCLcr distributions are reconstructed from the
  Table 1 medians and IQRs (normal and log-normal respectively), independent of
  each other; the paper does not report their correlation. eCLcr is held
  constant within each virtual patient.
* **Regimens.** Only the two most common regimens (1000 mg and 2000 mg every
  8 h) are simulated; the paper does not tabulate the regimen mix. The extra
  dose some patients received in the first dosing interval is not simulated.
* **Figure 2 comparison.** Figure 2 is prediction-corrected and pools all
  regimens; its percentiles were read by the maintainers from the raster image
  and are approximate. They are shown for orientation only and are not used in
  any assertion.
* **Population restriction.** The model was developed without patients on renal
  replacement therapy and is not intended for them. Only total cefotaxime was
  measured; desacetylcefotaxime is not modelled.
* **No erratum** was found for this article (checked 2026-09-30).
