---
title: "Etirinotecan pegol and its four metabolites (Sy 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Etirinotecan pegol and its four metabolites (Sy 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Sy_2018_etirinotecanPegol"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1007/s00280-018-3562-3>
* Description: `r ui$description`

Etirinotecan pegol (EP, NKTR-102) is a four-arm polyethylene-glycol conjugate of
irinotecan designed to release irinotecan slowly, converting irinotecan's
short-lived, high-peak SN-38 exposure into a low-peak, sustained one. Sy 2018
fit one integrated system to five analytes measured in the same subjects: EP
itself, the released irinotecan, and irinotecan's metabolites SN-38, SN-38
glucuronide (SN-38G) and APC. Each analyte has its own two-compartment
disposition; the metabolic cascade links them.

The paper builds the model in two stages. A 3-analyte model (EP, irinotecan,
SN-38) is fit first; its individual parameter estimates are then held fixed
while the SN-38G and APC data are added, giving the 5-analyte model. The two
between-subject correlations `Corr(CL, V1)` and `Corr(k2e, V2*)` could not be
fixed in Monolix and were re-estimated at stage two. Per the library's
"replicate the author's structure" policy, a base-plus-final development
sequence is packaged as **one** model file holding the final (5-analyte) model,
with the stage-one values taken from the same Table 4.

## Compartment scaling: what the volumes mean

Only EP's central volume is a real volume. For every metabolite, the conversion
fraction and the volume of distribution are not separately identifiable, so
Sy 2018 estimate aggregate ratios:

| Analyte | Aggregate volume | Definition (Sy 2018 Table 4) |
|---|---|---|
| Etirinotecan pegol | `vc` | `V1`, a true volume |
| Irinotecan | `vc_irinotecan` | `V2* = V2 / F12` |
| SN-38 | `vc_sn38` | `V3** = V3 / (F12 F23)` |
| SN-38G | `vc_sn38g` | `V4* = V4 / (F12 (F23 + F25) F34)` |
| APC | `vc_apc` | `V5* = V5 / F25`, set equal to `V3*` because `V5 = V3` was assumed |

`F12` is the fraction of eliminated EP appearing as irinotecan, `F23` and `F25`
the fractions of eliminated irinotecan appearing as SN-38 and APC, and `F34` the
fraction of eliminated SN-38 appearing as SN-38G. None of them is identifiable
on its own.

The packaged `model()` block carries **amounts** in its states, obtained by
multiplying each of the paper's concentration ODEs (Eqs. 2-4, 8-9) by that
analyte's own aggregate central volume. In amount form every metabolic step
becomes a plain first-order flux out of its precursor, and `state / volume`
returns the measured plasma concentration. Only EP's state is on the true molar
scale.

## Concentration and dose units

All five analytes were converted to molar concentrations before modelling
(Sy 2018 Methods, "Bioanalytical assays"). EP concentrations were first scaled
to their irinotecan content using the study-specific irinotecan loading factors
of 9.4% (06-IN-IR001) and 9.5% (07-PIR-02), then converted with irinotecan's
molecular weight of 586.678. Model concentrations are therefore **nmol/L**, and
EP concentrations are nmol/L of irinotecan equivalents.

The dose must be supplied in the same currency: irinotecan-equivalent nmol.

```{r units}
mw <- c(ep = 586.678, irinotecan = 586.678, sn38 = 382.404, sn38g = 568.53, apc = 618.687)
loading_factor <- 0.094 # 06-IN-IR001 irinotecan loading factor, Sy 2018 Methods

ep_dose_nmol <- function(mg_per_m2, bsa) {
  mg_per_m2 * bsa * loading_factor * 1e6 / mw[["ep"]]
}
ep_dose_nmol(145, 1.86)
```

Sy 2018 Figures 4 and 5 report EP exposures in **total-EP mass** units, so
reproducing them needs the inverse conversion (divide out the loading factor):

```{r units-back}
ep_nM_to_mg_per_L <- function(nM) nM * mw[["ep"]] * 1e-6 / loading_factor
sn38_nM_to_ng_per_mL <- function(nM) nM * mw[["sn38"]] / 1000
```

# Population

Sy 2018 pooled 83 patients with advanced solid tumours from two phase 1 studies:
67 of 76 enrolled in 06-IN-IR001 (an open-label dose-escalation study using
three weekly doses every 4 weeks, once every 2 weeks, or once every 3 weeks) and
16 of 18 in 07-PIR-02 (etirinotecan pegol 100 or 125 mg/m^2 every 3 weeks with
cetuximab). EP was given as a 90-minute intravenous infusion. Rich sampling
followed the first and third doses in 06-IN-IR001 and the first dose in
07-PIR-02, with sparse pre-dose sampling thereafter.

Baseline characteristics (Sy 2018 Table 2): median age 60 years (25-81), median
weight 72.3 kg (43.9-153.6), median body surface area 1.86 m^2 (1.36-2.74),
45.8% female, 94% White. Median eGFR was 79.1 mL/min/1.73 m^2 (34.9-216.6), with
31% normal, 48% mildly impaired and 19% moderately impaired renal function.
UGT1A1\*28 genotype was available for everyone: 38.6% none, 48.2% one copy,
10.8% two copies and 2.4% indeterminate, the last pooled into the wild-type
category. The analysis set held 1414, 1777, 1769, 1731 and 1167 quantifiable
concentrations for EP, irinotecan, SN-38, SN-38G and APC.

The same information is available programmatically via `ui$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Sy_2018_etirinotecanPegol.R` carries an in-file
comment naming its source location. They are collected here for review. The
display equations were read from the typeset PDF; the preprocessed markdown
drops all of them.

| Equation / parameter | Value | Source location |
|---|---|---|
| EP central ODE | `d[EP]/dt = Rin/V1 - (CL/V1 + k1p)[EP] + kp1[EPp]` | Eq. 2 |
| Irinotecan central ODE | `d[Iri]/dt = (CL/V2*)[EP] - (k2e + k2p)[Iri] + kp2[Irip]` | Eq. 3 |
| SN-38 central ODE | `d[SN38]/dt = (1/V3*) k2e V2* [Iri] - (k3e + k3p)[SN38] + kp3[SN38p]` | Eq. 4 |
| SN-38G central ODE | `d[SN38G]/dt = (1/V4*) k3e V3* [SN38] - (k4e + k4p)[SN38G] + kp4[SN38Gp]` | Eq. 8 (printed `k4e`; see Deviations) |
| APC central ODE | `d[APC]/dt = (1/V5*) k2e V2* [Iri] - (k5e + k5p)[APC] + kp5[APCp]` | Eq. 9, with the SN-38 : APC split of Results, "5-Analyte pharmacokinetic model" |
| EP clearance covariate model | `CL = theta_CL (AGE/60)^t_AGE (BSA/1.86)^t_BSA (eGFR/84.1)^t_eGFR exp(eta)` | Eq. 5 |
| EP volume covariate model | `V1 = theta_V1 (BSA/1.86)^t_BSA exp(eta)` | Eq. 6 |
| SN-38 UGT1A1 covariate model | `k3e = theta_k3e exp(t_UGT1A1 I[TA(7)/TA(7)]) exp(eta)` | Eq. 7 |
| eGFR definition | `175 SCr^-1.154 Age^-0.203 0.742^I(F) 1.212^I(Black)` | Eq. 1 |
| `lcl` 0.237 L/h, `lvc` 5.05 L, `lk12` 6.78e-3, `lk21` 5.8e-4 | | Table 4, EP block |
| `e_age_cl` -0.271, `e_bsa_cl` 1.32, `e_crcl_cl` 0.2, `e_bsa_vc` 1.1 | | Table 4, EP block |
| `lvc_irinotecan` 1.8 L, `lkel_irinotecan` 27.6, `lk12_irinotecan` 18.8, `lk21_irinotecan` 3.2e-3 | | Table 4, Irinotecan block |
| `lkel_sn38` 0.0602, `lk12_sn38` 0.23, `lk21_sn38` 8.75e-3 | | Table 4, SN38 block |
| `lvc_sn38` 365 L | **Not from Table 4.** Back-solved from the reference SN-38 Cmax of 4.0 ng/mL via `AUC(SN-38) = Dose/(V3* k3e)`; see Deviations | Figure 5, "Reference 60 yrs" row |
| `e_snp_ugt1a1_rs8175347_hom_kel_sn38` -0.67 | | Table 4, SN38 block |
| `lvc_sn38g` 11.6 L, `lkel_sn38g` 1.41, `lk12_sn38g` 0.548, `lk21_sn38g` 0.104 | | Table 4, SN38G block |
| `lfm_sn38` 0.631, `lkel_apc` 0.0235, `lk12_apc` 0.0236, `lk21_apc` 1.39e-3 | | Table 4, APC block |
| All IIV variances | | Table 4, right-hand column |
| `Corr(CL, V1)` 0.713, `Corr(k2e, V2*)` -0.755 | | Table 4, 5-analyte "Correlations" block |
| All residual-error terms | | Table 4, "Multiplicative error" / "Additive error" rows |
| Reference EP AUC 6.87 mg.h/mL, Cmax 50.4 mg/L over six cycles | | Figure 4, "Reference 60 yrs" row |
| Reference SN-38 AUC 3.51 mg.h/L, Cmax 4.0 ng/mL over six cycles | | Figure 5, "Reference 60 yrs" row |
| Homozygous SN-38 AUC 5.80 mg.h/L, Cmax 5.4 ng/mL | | Figure 5, "TA(7)/TA(7)" row |

# Structural verification

Two mechanical checks come first, because both failure modes are silent.

`rxode2` recognises a `cl` / `vc` pair and can solve a system analytically from
those two variables, discarding the explicit `d/dt()` bodies and dropping
peripheral states from the output. Assert that all ten declared states survive
the solve.

Separately, a model with five declared endpoints assigns each observable its own
compartment slot. Assert that every endpoint slot sits **above** all ten ODE
states, which proves no state was renumbered by the endpoint declarations.

```{r structure}
stopifnot(length(ui$state) == 10L)
ui$predDf[, c("var", "dvid", "cmt")]

stopifnot(
  # every endpoint slot is above every ODE state: nothing was inserted among them
  all(ui$predDf$cmt > length(ui$state)),
  setequal(ui$predDf$var, c("Cc", "Cc_irinotecan", "Cc_sn38", "Cc_sn38g", "Cc_apc"))
)
```

With five endpoints declared, observation records must carry `cmt =` an
**endpoint name**, not an ODE state name; `rxSolve()` then returns every model
variable as a column regardless of which endpoint was named.

```{r typical-solve}
mod_typ <- rxode2::zeroRe(ui)

solve_typ <- function(mg_per_m2, bsa, age, crcl, ugt_hom, n_cycles = 6, by = 0.25,
                      tmax = 504 * n_cycles) {
  ev <- rxode2::et(
    amt = ep_dose_nmol(mg_per_m2, bsa), dur = 1.5,
    ii = 504, addl = n_cycles - 1, cmt = "central"
  )
  ev <- rxode2::et(ev, seq(0, tmax, by = by), cmt = "Cc")
  ev <- as.data.frame(ev)
  ev$AGE <- age
  ev$BSA <- bsa
  ev$CRCL <- crcl
  ev$SNP_UGT1A1_RS8175347_HOM <- ugt_hom
  rxode2::rxSolve(
    mod_typ, ev,
    returnType = "data.frame", useLinCmt = FALSE,
    atol = 1e-12, rtol = 1e-10, maxsteps = 1e6
  )
}

ref <- solve_typ(145, bsa = 1.86, age = 60, crcl = 84.1, ugt_hom = 0)
stopifnot(all(ui$state %in% names(ref)))
```

The ten states are all present, so the explicit ODE system was integrated.

A third structural check pins EP's own disposition to its closed form. EP's
central compartment is a standard two-compartment system in `kel`, `k12` and
`k21`, so its macro rate constants follow analytically; if `rxode2` had matched
a one-compartment kernel the terminal slope would be `kel` instead of `beta`.

```{r closed-form}
kel <- 0.237 / 5.05
k12 <- 6.78e-3
k21 <- 5.8e-4
s <- kel + k12 + k21
alpha <- (s + sqrt(s^2 - 4 * kel * k21)) / 2
beta <- (s - sqrt(s^2 - 4 * kel * k21)) / 2

# a single dose, followed long enough for the terminal phase to dominate
single <- solve_typ(145, bsa = 1.86, age = 60, crcl = 84.1, ugt_hom = 0,
                    n_cycles = 1, by = 4, tmax = 24 * 400)
late <- single[single$time >= 24 * 200 & single$time <= 24 * 400, ]
slope_obs <- -stats::coef(stats::lm(log(late$Cc) ~ late$time))[[2]]

c(alpha_t_half_h = log(2) / alpha, beta_t_half_d = log(2) / beta / 24)
stopifnot(
  # the solved terminal slope is beta, not kel: a one-compartment collapse
  # would give kel = 0.0469 /h, eighty times larger
  abs(slope_obs / beta - 1) < 0.05,
  abs(slope_obs / kel - 1) > 0.5
)
```

EP's terminal half-life of about 57 days is the feature the PEG conjugate was
built for, and it is what makes every downstream metabolite formation-rate
limited.

# Mass balance across the metabolic cascade

For a linear cascade, integrating each ODE to infinity gives a chain of exact
recovery identities. Each analyte's total elimination flux must equal the flux
arriving from its precursor, and EP's must equal the dose. These identities are
independent of the peripheral compartments, which net to zero over all time.

```{r mass-balance}
# a single dose integrated far past the terminal phase, so that the recovery
# identities are not truncated by the 57-day terminal half-life
tail_ev <- rxode2::et(amt = ep_dose_nmol(145, 1.86), dur = 1.5, cmt = "central")
tail_ev <- rxode2::et(
  tail_ev,
  # fine early so the trapezoidal AUC of the fast-turnover metabolites is
  # accurate, coarse later where every profile decays at EP's terminal rate
  sort(unique(c(seq(0, 12, by = 0.02), seq(12, 168, by = 0.5),
                seq(168, 24 * 2000, by = 6)))),
  cmt = "Cc"
)
tail_ev <- as.data.frame(tail_ev)
tail_ev$AGE <- 60
tail_ev$BSA <- 1.86
tail_ev$CRCL <- 84.1
tail_ev$SNP_UGT1A1_RS8175347_HOM <- 0
inf <- rxode2::rxSolve(mod_typ, tail_ev, returnType = "data.frame", useLinCmt = FALSE,
                       atol = 1e-12, rtol = 1e-10, maxsteps = 1e7)

auc_of <- function(v) PKNCA::pk.calc.auc.last(conc = inf[[v]], time = inf$time)
dose1 <- ep_dose_nmol(145, 1.86)
fm <- 0.631
# named once rather than inlined three times: V3* is the one parameter in this
# model that is not the printed Table 4 value (see "Assumptions and
# deviations"), and it enters both the SN-38 and the APC identity because the
# paper assumes V5 = V3
v3 <- 365

balance <- tibble::tibble(
  Identity = c(
    "CL * AUC(EP) / Dose",
    "k2e * V2* * AUC(irinotecan) / Dose",
    "k3e * V3* * AUC(SN-38) / Dose",
    "k4e * V4* * AUC(SN-38G) / (k3e * V3* * AUC(SN-38))",
    "k5e * V5* * AUC(APC) / ((1 - fm)/fm * k2e * V2* * AUC(irinotecan))"
  ),
  Value = c(
    0.237 * auc_of("Cc") / dose1,
    27.6 * 1.8 * auc_of("Cc_irinotecan") / dose1,
    0.0602 * v3 * auc_of("Cc_sn38") / dose1,
    1.41 * 11.6 * auc_of("Cc_sn38g") / (0.0602 * v3 * auc_of("Cc_sn38")),
    0.0235 * v3 * auc_of("Cc_apc") /
      ((1 - fm) / fm * 27.6 * 1.8 * auc_of("Cc_irinotecan"))
  )
)
knitr::kable(balance, digits = 4)

stopifnot(all(abs(balance$Value - 1) < 0.01))
```

All five identities close to within 1%. The last two are the ones that matter
for transcription: the SN-38G identity fails by a factor of 23 if Eq. 8's
printed `k4e` is taken literally instead of being read as `k3e`, and the APC
identity fails by a factor of `fm / (1 - fm)` if the SN-38 : APC flux ratio is
dropped from Eq. 9.

That flux ratio is also the only place `F(irinotecan->SN38)` enters the model, so
a two-solve perturbation confirms it is load-bearing.

```{r fm-perturbation}
perturb <- rxode2::rxSolve(
  mod_typ, tail_ev, returnType = "data.frame", useLinCmt = FALSE,
  params = c(lfm_sn38 = log(0.4)),
  atol = 1e-12, rtol = 1e-10, maxsteps = 1e7
)
rel_change <- c(
  EP = max(abs(perturb$Cc - inf$Cc)) / max(inf$Cc),
  SN38 = max(abs(perturb$Cc_sn38 - inf$Cc_sn38)) / max(inf$Cc_sn38),
  APC = max(abs(perturb$Cc_apc - inf$Cc_apc)) / max(inf$Cc_apc)
)
round(rel_change, 6)
stopifnot(
  # fm moves APC and nothing else
  rel_change[["APC"]] > 0.5,
  rel_change[["EP"]] < 1e-8,
  rel_change[["SN38"]] < 1e-8
)
```

# Replicating Figure 4: covariate effects on etirinotecan pegol exposure

Sy 2018 simulated 500 virtual patients receiving six cycles of 145 mg/m^2 every
3 weeks and reported cumulative EP AUC and Cmax for a reference subject (60
years, median covariates, eGFR 84.1 mL/min/1.73 m^2) and for five covariate
scenarios. Because every parameter is log-normal and the exposures are products
of log-normals, the cohort median equals the typical-value prediction, so the
published medians can be reproduced deterministically.

```{r fig4}
scenarios <- tibble::tribble(
  ~scenario,                   ~age, ~crcl,
  "Reference 60 yrs",            60,  84.1,
  "45 yrs",                      45,  84.1,
  "75 yrs",                      75,  84.1,
  "Normal renal function",       60, 105.0,
  "Mild renal impairment",       60,  75.0,
  "Moderate renal impairment",   60,  45.0
)

fig4_sim <- dplyr::bind_rows(lapply(seq_len(nrow(scenarios)), function(i) {
  s <- solve_typ(
    145, bsa = 1.86, age = scenarios$age[i], crcl = scenarios$crcl[i], ugt_hom = 0
  )
  tibble::tibble(
    scenario = scenarios$scenario[i],
    cmax = ep_nM_to_mg_per_L(max(s$Cc)),
    auclast = ep_nM_to_mg_per_L(PKNCA::pk.calc.auc.last(conc = s$Cc, time = s$time))
  )
}))

fig4_pub <- tibble::tibble(
  scenario = scenarios$scenario,
  # Figure 4 prints EP AUC in mg.h/mL; 6.87 mg.h/mL = 6870 mg.h/L
  auclast = c(6.87, 6.41, 7.22, 6.53, 6.92, 7.56) * 1000,
  cmax = c(50.4, 50.2, 50.5, 50.2, 50.4, 50.6)
)

fig4_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = fig4_sim,
  reference = fig4_pub,
  by = "scenario",
  units = c(cmax = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(fig4_cmp)
```

Every row agrees with Figure 4. The Cmax column is essentially flat across
scenarios in both the publication and the replication, because neither age nor
renal function acts on `V1`; the AUC column reproduces the publication's
ordering (AUC rises with age and falls with renal function, the latter because
`theta_CL,eGFR` is positive so worse renal function means lower EP clearance).

The AUC replication runs slightly below the publication because a
six-cycle cumulative AUC truncated at 3024 h omits the tail of a compound whose
terminal half-life is 57 days; the published value is nearer the
extrapolated-to-infinity figure.

```{r fig4-gate}
chk4 <- fig4_sim |>
  dplyr::left_join(fig4_pub, by = "scenario", suffix = c("_sim", "_pub")) |>
  dplyr::mutate(
    pct_cmax = 100 * (cmax_sim - cmax_pub) / cmax_pub,
    pct_auc = 100 * (auclast_sim - auclast_pub) / auclast_pub
  )
stopifnot(
  # structural: a mis-transcribed CL, volume, dose or unit conversion moves
  # these by tens of percent
  max(abs(chk4$pct_cmax)) < 5,
  max(abs(chk4$pct_auc)) < 10,
  # the publication's own ordering must be preserved
  identical(order(chk4$auclast_sim), order(chk4$auclast_pub))
)
```

# Replicating Figure 5: UGT1A1\*28 and SN-38 exposure

Sy 2018 report that a UGT1A1\*28 homozygote has median SN-38 AUC and Cmax about
1.7-fold and 1.3-fold those of a wild-type or heterozygous patient. The
underlying mechanism is Eq. 7: `exp(-0.67) = 0.512`, so a homozygote retains
about half the SN-38 elimination capacity.

These are **ratios**, so they are independent of the SN-38 aggregate volume and
test the covariate model directly.

```{r fig5-ratio}
sn38_hom <- solve_typ(145, bsa = 1.86, age = 60, crcl = 84.1, ugt_hom = 1)

ratio <- c(
  AUC = PKNCA::pk.calc.auc.last(conc = sn38_hom$Cc_sn38, time = sn38_hom$time) /
    PKNCA::pk.calc.auc.last(conc = ref$Cc_sn38, time = ref$time),
  Cmax = max(sn38_hom$Cc_sn38) / max(ref$Cc_sn38)
)
published_ratio <- c(AUC = 5.80 / 3.51, Cmax = 5.4 / 4.0)
knitr::kable(
  tibble::tibble(
    Metric = names(ratio),
    Replicated = round(unname(ratio), 3),
    `Figure 5` = round(unname(published_ratio), 3)
  )
)
stopifnot(
  abs(ratio[["AUC"]] / published_ratio[["AUC"]] - 1) < 0.10,
  abs(ratio[["Cmax"]] / published_ratio[["Cmax"]] - 1) < 0.15,
  # the AUC effect must exceed the Cmax effect, as Figure 5 shows
  ratio[["AUC"]] > ratio[["Cmax"]]
)
```

The covariate model reproduces both of Figure 5's ratios, and reproduces the
qualitative point the paper makes with them: because SN-38 kinetics after EP are
formation-rate limited, halving SN-38 elimination raises AUC substantially more
than it raises Cmax.

## SN-38 absolute scale

The same comparison on **absolute** values. Note that one cell of this table is
circular: `lvc_sn38` was back-solved from the wild-type Cmax of 4.0 ng/mL (see
"Assumptions and deviations"), so that entry is pinned by construction and
cannot fail. The other three cells are genuine out-of-sample checks -- the
wild-type AUC tests the integration window and the two homozygous cells test the
UGT1A1 covariate arm, and none of the three entered the back-solve.

```{r fig5-absolute}
fig5_sim <- tibble::tibble(
  genotype = c("Wild-type / heterozygous", "TA(7)/TA(7) homozygous"),
  cmax = sn38_nM_to_ng_per_mL(c(max(ref$Cc_sn38), max(sn38_hom$Cc_sn38))),
  auclast = c(
    PKNCA::pk.calc.auc.last(conc = ref$Cc_sn38, time = ref$time),
    PKNCA::pk.calc.auc.last(conc = sn38_hom$Cc_sn38, time = sn38_hom$time)
  ) * mw[["sn38"]] * 1e-6
)
fig5_pub <- tibble::tibble(
  genotype = fig5_sim$genotype,
  cmax = c(4.0, 5.4),
  auclast = c(3.51, 5.80)
)
knitr::kable(
  nlmixr2lib::ncaComparisonTable(
    simulated = fig5_sim, reference = fig5_pub, by = "genotype",
    units = c(cmax = "ng/mL", auclast = "mg*h/L"), tolerance_pct = 20
  )
)

fig5_pct_diff <- 100 * (as.matrix(fig5_sim[, c("cmax", "auclast")]) /
  as.matrix(fig5_pub[, c("cmax", "auclast")]) - 1)
dimnames(fig5_pct_diff) <- list(fig5_sim$genotype, c("Cmax", "AUC"))
round(fig5_pct_diff, 1)

# The three non-circular cells: wild-type AUC, and both homozygous cells. These
# are typical-value (zeroRe) solves compared against the paper's own printed
# summary numbers, so the residual difference is digitisation and six-cycle
# truncation of a long terminal phase -- not per-subject sampling. A 15% bound
# is therefore the right shape here.
stopifnot(abs(fig5_pct_diff["Wild-type / heterozygous", "AUC"]) < 15)
stopifnot(all(abs(fig5_pct_diff["TA(7)/TA(7) homozygous", ]) < 15))
```

With `lvc_sn38` set from the Figure 5 Cmax anchor, all four cells agree with the
paper. Because SN-38's ODE is exactly

```
d[SN38]/dt = k2e V2* [irinotecan] / V3* - (k3e + k3p)[SN38] + kp3[SN38p]
```

and integrating it to infinity gives `AUC(SN-38) = Dose / (V3* k3e)`, `V3*` is
the only parameter that can put a pure scale error on SN-38 without disturbing
any other analyte -- which is what made the back-solve well posed. The printed
Table 4 value of 80 L puts this table 4.6-fold high on every cell; see
"Assumptions and deviations" for why the printed value was not used.

# Virtual cohort

A 200-subject cohort receiving one 145 mg/m^2 cycle, with covariates drawn to
match Sy 2018 Table 2. Weight is not a model covariate; BSA is.

```{r cohort}
rxode2::rxSetSeed(20180321)
set.seed(20180321)
n_sub <- 200L

clamp <- function(x, lo, hi) pmin(pmax(x, lo), hi)
cohort <- tibble::tibble(
  id = seq_len(n_sub),
  AGE = clamp(round(stats::rnorm(n_sub, 59, 11)), 25, 81),
  BSA = clamp(stats::rnorm(n_sub, 1.88, 0.24), 1.36, 2.74),
  CRCL = clamp(exp(stats::rnorm(n_sub, log(79.1), 0.32)), 34.9, 216.6),
  SNP_UGT1A1_RS8175347_HOM = stats::rbinom(n_sub, 1, 0.108)
)
summary(cohort[, c("AGE", "BSA", "CRCL")])
mean(cohort$SNP_UGT1A1_RS8175347_HOM)

grid <- sort(unique(c(
  seq(0, 6, by = 0.25), seq(8, 48, by = 4), seq(60, 504, by = 12)
)))
ev_cohort <- lapply(seq_len(n_sub), function(i) {
  e <- rxode2::et(amt = ep_dose_nmol(145, cohort$BSA[i]), dur = 1.5, cmt = "central")
  e <- rxode2::et(e, grid, cmt = "Cc")
  e <- as.data.frame(e)
  e$id <- cohort$id[i]
  e$AGE <- cohort$AGE[i]
  e$BSA <- cohort$BSA[i]
  e$CRCL <- cohort$CRCL[i]
  e$SNP_UGT1A1_RS8175347_HOM <- cohort$SNP_UGT1A1_RS8175347_HOM[i]
  e
})
ev_cohort <- dplyr::bind_rows(ev_cohort)

sim <- rxode2::rxSolve(
  ui, ev_cohort, returnType = "data.frame", useLinCmt = FALSE,
  atol = 1e-10, rtol = 1e-8, maxsteps = 5e5
)
stopifnot(dplyr::n_distinct(sim$id) == n_sub)
```

## A consequence of log-normal variability on a fraction

Sy 2018 state that all interindividual variability was log-normal, and Table 4
gives `F(irinotecan->SN38)` an omega variance of 0.0528. A log-normal centred on
0.631 with that variance puts roughly 2% of subjects above 1, where the
`(1 - fm) / fm` APC flux ratio turns negative. This is a property of the
published variability model, not of the transcription; it is reported rather
than repaired, and those subjects are excluded from the APC panels below.

```{r fm-admissible}
fm_by_id <- sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(fm_sn38 = dplyr::first(fm_sn38), .groups = "drop")
n_inadmissible <- sum(fm_by_id$fm_sn38 >= 1)
c(n_subjects = n_sub, n_fm_at_or_above_1 = n_inadmissible,
  pct = round(100 * n_inadmissible / n_sub, 1))

apc_ok <- fm_by_id$id[fm_by_id$fm_sn38 < 1]
stopifnot(
  # the published omega is admissible for the large majority of subjects
  n_inadmissible / n_sub < 0.06,
  all(sim$Cc_sn38[sim$time > 0] > 0),
  all(sim$Cc_sn38g[sim$time > 0] > 0),
  all(sim$Cc_apc[sim$time > 0 & sim$id %in% apc_ok] > 0)
)
```

## Concentration-time profiles

```{r profiles, fig.height = 6}
analyte_labels <- c(
  Cc = "Etirinotecan pegol", Cc_irinotecan = "Irinotecan", Cc_sn38 = "SN-38",
  Cc_sn38g = "SN-38G", Cc_apc = "APC"
)
prof <- sim |>
  dplyr::filter(id %in% apc_ok) |>
  dplyr::select(id, time, dplyr::all_of(names(analyte_labels))) |>
  tidyr::pivot_longer(-c(id, time), names_to = "analyte", values_to = "conc") |>
  dplyr::mutate(analyte = factor(analyte_labels[analyte], levels = analyte_labels))

# every analyte starts at exactly zero, so the time-zero record PKNCA needs is
# already present and is kept; it is dropped only for the log-scale plot
stopifnot(all(prof$conc[prof$time == 0] == 0))

prof_q <- prof |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(analyte, time) |>
  dplyr::summarise(
    lo = stats::quantile(conc, 0.05), md = stats::median(conc),
    hi = stats::quantile(conc, 0.95), .groups = "drop"
  )

ggplot2::ggplot(prof_q, ggplot2::aes(time / 24)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.25) +
  ggplot2::geom_line(ggplot2::aes(y = md)) +
  ggplot2::facet_wrap(~analyte, scales = "free_y", ncol = 2) +
  ggplot2::scale_x_continuous("Time (days)") +
  ggplot2::scale_y_log10("Concentration (nmol/L)") +
  ggplot2::labs(
    title = "Etirinotecan pegol 145 mg/m^2, single 90-min infusion",
    subtitle = "Median with 5th-95th percentile band, 200 virtual patients"
  ) +
  ggplot2::theme_bw()
```

Every metabolite tracks EP's 57-day terminal phase rather than its own
disposition half-life. That formation-rate limitation is the paper's central
pharmacological claim about the conjugate, and it emerges from the parameters
without being imposed.

# PKNCA validation

Non-compartmental analysis of the cohort's first cycle, one grouping level per
analyte. The concentration frame is filtered only on `!is.na()`, so the time-zero
record is retained and no AUC-range warning is raised.

```{r pknca}
conc_data <- prof |>
  dplyr::filter(!is.na(conc)) |>
  dplyr::arrange(id, analyte, time)

dose_data <- cohort |>
  dplyr::filter(id %in% apc_ok) |>
  dplyr::transmute(id, time = 0, amt = ep_dose_nmol(145, BSA)) |>
  tidyr::expand_grid(analyte = factor(analyte_labels, levels = analyte_labels))

o_conc <- PKNCA::PKNCAconc(as.data.frame(conc_data), conc ~ time | id / analyte)
# PKNCAdose rejects slash grouping, so the same two levels are given with `+`
o_dose <- PKNCA::PKNCAdose(as.data.frame(dose_data), amt ~ time | id + analyte)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(
    start = 0, end = 504,
    cmax = TRUE, tmax = TRUE, auclast = TRUE
  )
)
nca_res <- suppressWarnings(PKNCA::pk.nca(o_data))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(analyte, PPTESTCD, PPORRES) |>
  dplyr::group_by(analyte, PPTESTCD) |>
  dplyr::summarise(value = stats::median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)

nca_wide |>
  dplyr::rename(
    "Analyte" = analyte, "Cmax (nmol/L)" = cmax, "Tmax (h)" = tmax,
    "AUClast (nmol*h/L)" = auclast
  ) |>
  knitr::kable(digits = c(0, 2, 2, 1))
```

```{r pknca-gate}
stopifnot(
  setequal(as.character(nca_wide$analyte), unname(analyte_labels)),
  # EP peaks at the end of the 90-min infusion; every metabolite peaks later,
  # which is the signature of formation-rate-limited kinetics
  nca_wide$tmax[nca_wide$analyte == "Etirinotecan pegol"] <= 1.5,
  all(nca_wide$tmax[nca_wide$analyte != "Etirinotecan pegol"] > 1.5)
)
```

## Comparison against the published exposures

Sy 2018 report no NCA table, but Figures 4 and 5 give the model-predicted
reference exposures over six cycles, which are the appropriate published
comparators. They were compared above by scenario (Figure 4, all rows within
tolerance) and by genotype (Figure 5, both ratios reproduced and all four
absolute cells within tolerance, one of which is pinned by the `lvc_sn38`
back-solve described in "Assumptions and deviations"). The combined view:

```{r combined-comparison}
combined_sim <- tibble::tibble(
  analyte = c("Etirinotecan pegol", "SN-38"),
  cmax = c(ep_nM_to_mg_per_L(max(ref$Cc)), sn38_nM_to_ng_per_mL(max(ref$Cc_sn38))),
  auclast = c(
    ep_nM_to_mg_per_L(PKNCA::pk.calc.auc.last(conc = ref$Cc, time = ref$time)),
    PKNCA::pk.calc.auc.last(conc = ref$Cc_sn38, time = ref$time) * mw[["sn38"]] * 1e-6
  )
)
combined_pub <- tibble::tibble(
  analyte = combined_sim$analyte, cmax = c(50.4, 4.0), auclast = c(6870, 3.51)
)
knitr::kable(
  nlmixr2lib::ncaComparisonTable(
    simulated = combined_sim, reference = combined_pub, by = "analyte",
    units = c(cmax = "mg/L or ng/mL", auclast = "mg*h/L"), tolerance_pct = 20
  )
)
```

# Assumptions and deviations

## SN-38 aggregate volume: Table 4 contradicts Figures 3 and 5 by 4.6-fold

This is the one material conflict in the source, and it governs the absolute
SN-38 and APC concentrations the packaged model predicts.

**What the model uses.** `lvc_sn38 = log(365)`, **not** the 80 L printed in
Table 4 for `V3** = V3 / (F12 F23)`. This is a deliberate, documented deviation
from the published parameter table -- the only one in this model -- and it is the
one number in the file that does not come from the paper's own tables. The
value was back-solved from the reference SN-38 Cmax of 4.0 ng/mL printed in
Figure 5, inverting `AUC(SN-38) = Dose / (V3* k3e)` (which follows exactly from
Eq. 4). The Cmax anchor was preferred over the cumulative-AUC anchor because it
needs no integration window; the AUC anchor independently gives 369 L, agreeing
to 1.3%.

**Why the printed value was not used.**

1. It is the only entry in Table 4 whose standard error exceeds its estimate
   (80 +/- 190, a relative standard error of 238%). Every other volume is
   reported at 7-9% (`V2*` 1.8 +/- 0.13, `V4*` 11.6 +/- 0.95), and no other
   parameter in the table exceeds 32%.
2. `AUC(SN-38) = Dose / (V3* k3e)` follows exactly from Eq. 4, so `V3*` is the
   only parameter that can scale SN-38 without changing anything else. With
   `V3* = 80` the model over-predicts the paper's own Figure 5 reference values
   by 4.55-fold on Cmax and 4.61-fold on AUC. The agreement of those two factors
   to within 1.3% is what rules out a structural transcription error: a wrong
   equation would move Cmax and AUC by different amounts. It is one number.
3. The Figure 3 goodness-of-fit panels place both observed and
   individual-predicted SN-38 between about 0.4 and 25 nmol/L across the whole
   dose range. With `V3* = 80` the model predicts 47.6 nmol/L for a typical
   patient at 145 mg/m^2, above the top of the paper's own cloud before any dose
   escalation or IIV is considered. With `V3* = 365` it predicts about
   10.4 nmol/L, inside that cloud.
4. The Table 4 additive residual error for SN-38 is 0.383 nmol/L, which is 4% of
   the Figure 5 reference Cmax of 4.0 ng/mL (10.5 nmol/L) but under 1% of what
   `V3* = 80` predicts. An additive term exists to describe the
   near-quantitation-limit end of the range, and 4% is the plausible magnitude.

**What the substitution does and does not touch.** Only SN-38 and APC move.
SN-38G's level depends on the SN-38 *amount* rather than its concentration, so
it is independent of `V3*` and already reproduced the Figure 3 range under
either value. APC moves with SN-38 because the model sets `vc_apc <- vc_sn38`
(the paper assumes `V5 = V3`), so the override propagates to APC automatically --
which is the intended behaviour, since Figure 3's APC panel has the same scale
problem under the printed value. EP and irinotecan are untouched.

**Why this was decided the way it was.** Substituting 365 L replaces a published
parameter-table value with one back-solved from a figure in the same paper, so
it is not a call this workflow makes on its own: the standing policy sanctions
digitising figures for parameters the tables *omit*, whereas here the table does
print a value. It was therefore escalated as an unresolvable value conflict
between two places in one source and ruled on explicitly (maintainer decision,
2026-09-21) in favour of the figure-anchored value, on the grounds that the
printed 80 L cannot be reconciled with three other independent statements the
same paper makes. Both readings were implemented and gated before the ruling;
the printed value is recoverable in one line for anyone who wants the literal
transcription:

```r
mod <- nlmixr2lib::modellib("Sy_2018_etirinotecanPegol")
mod <- rxode2::ini(mod, lvc_sn38 = log(80))
```

**What this buys.** With the substitution, the SN-38 absolute comparison above
becomes a real gate rather than a flagged deviation, and it passes on the three
cells that did not enter the back-solve. Everything that never depended on the
SN-38 concentration scale is unchanged and still passes: the mass-balance
identities, the EP terminal-slope check, the Figure 4 EP replication across all
six covariate scenarios, and both Figure 5 UGT1A1 ratios (which are ratios, so
they were always invariant to `V3*`).

## Equation transcription

* **Eq. 8 prints `k4e` where `k3e` is meant.** SN-38G's formation term is
  printed as `(1/V4*) k4e V3 [SN38]`, which would form a metabolite at its own
  elimination rate constant. The surrounding text ("Glucuronidation of SN38 to
  SN38G is governed by the rate constant `F34 k3e`") and Figure 2b (`k34 = k3e`)
  both give `k3e`, and the mass-balance identity above closes only with `k3e`.
* **Eqs. 4 and 9 omit the SN-38 : APC split that the text supplies.** As printed,
  the SN-38 and APC formation terms are identical, which would leave
  `F(irinotecan->SN38)` absent from the model despite being estimated in Table 4
  with a standard error and an IIV. Results, "5-Analyte pharmacokinetic model"
  states that "the fraction for the formation of SN38 and subsequently one minus
  that fraction to describe the formation of APC accounted for the difference in
  the concentrations of the two metabolites", and that "the volume of the central
  compartment for APC was assumed to be the same as that for SN38". With a shared
  volume, the concentration difference between the two is exactly the flux ratio
  `(1 - fm) / fm`, which is how the packaged model applies it. The perturbation
  check above confirms `fm_sn38` then moves APC and nothing else.
* **`V3*` is defined twice.** Table 4 defines it as `V3 / (F12 F23)` while the
  5-analyte text redefines it as `V3 / (F12 (F23 + F25))`. The two differ by
  `F23 / (F23 + F25)`, which is `fm_sn38` itself, so the two readings are the
  same model once the flux ratio is placed as above; only the bookkeeping name
  for the number changes. The packaged model uses the Table 4 definition, which
  is where the numeric value comes from.

## Reference values and covariate encoding

* **eGFR reference is 84.1, not 79.1.** Eq. 5 normalises eGFR by 84.1
  mL/min/1.73 m^2, and the simulation reference population in Results says
  "renal function (84.1 mL/min)". The Table 4 caption instead says the typical
  patient has eGFR 79.1, which is the Table 2 cohort median. The printed
  equation is used, per the standing text-versus-equation rule; the difference
  is 1.2% on EP clearance.
* **The eGFR equation is MDRD, not CKD-EPI or Cockcroft-Gault.** Table 1 labels
  the covariate "eGFR CG" and its Comments column says "Calculated from serum
  creatinine using Corrected CKD-EPI", but Eq. 1 as printed is the
  IDMS-traceable four-variable MDRD equation. The covariate is registered as
  `CRCL`, whose canonical definition covers MDRD- and CKD-EPI-estimated GFR
  alike; the assay actually used is recorded in
  `covariateData$CRCL$description`.
* **EP central volume is 5.05 L, not 5.5 L.** The Abstract says 5.5 L; Table 4
  reports 5.05 +/- 0.14. The parameter table is used.
* **UGT1A1 is a single homozygous indicator.** Eq. 7 carries
  `I[UGT1A1 = TA(7)/TA(7)]` only, so wild-type and heterozygous subjects share
  the reference `k3e`. The two subjects with indeterminate genotype were pooled
  into the wild-type category by the authors.
* **Interindividual variances are used as printed.** Table 4's right-hand column
  is headed "variance" and its parenthesised CV% equals `sqrt(variance)` for
  every row, so the tabulated numbers are omega variances on the log scale and
  need no `log(CV^2 + 1)` conversion.
* **Off-diagonals come from the 5-analyte block.** `Corr(CL, V1)` and
  `Corr(k2e, V2*)` are reported twice, as 0.759 / -0.751 in the 3-analyte block
  and as 0.713 / -0.755 in the 5-analyte "Correlations" block. The latter are
  used, because they are the final model's values and were the only two
  parameters re-estimated at stage two.
* **No parameter is wrapped in `fixed()`.** The stage-one parameters were held
  fixed while fitting stage two, but every value in Table 4 carries a standard
  error from its own estimation step, so none was fixed in the sense the
  convention records (a value held at a constant the authors never estimated).
* **CYP3A4 co-medication is not in the model.** A CYP3A4/5 inducer effect on
  `k25` reached significance but was discarded by the authors because its
  direction was opposite to the expected mechanism (Results, "5-Analyte
  pharmacokinetic model"). It is recorded in `population$co_medication` rather
  than in `covariatesDataExcluded`, because no canonical column exists for a
  generic weak/moderate CYP3A4 inducer class.

## Simulation assumptions

* The virtual cohort's covariate distributions were chosen to reproduce the
  medians and ranges of Table 2; Sy 2018 publish no covariate correlation
  structure, so age, BSA, eGFR and genotype are drawn independently.
* The dose conversion uses the 06-IN-IR001 irinotecan loading factor of 9.4%.
  07-PIR-02 used 9.5%; the 1% difference is below every tolerance used here.
* Figures 4 and 5 report medians over 500 simulated patients. Because the model
  is linear and every random effect is log-normal, the median of the simulated
  exposures equals the typical-value prediction, so the replications above are
  run deterministically with `rxode2::zeroRe()`.
