---
title: "Linagliptin (Tadayasu 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linagliptin (Tadayasu 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Tadayasu_2014_linagliptin"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Tadayasu_2014_linagliptin"))$description`
* Article: <https://doi.org/10.18433/j3s304>

# Population

Tadayasu 2014 analysed a single Phase II trial in Japanese patients with type
2 diabetes mellitus (T2DM). 72 patients were randomised to placebo, 0.5 mg,
2.5 mg, or 10 mg linagliptin orally once daily for 28 days; the final
population PK/PD model uses only the 2.5 mg and 10 mg arms (18 patients each,
n = 36 total), because the 0.5 mg arm violated the quasi-equilibrium
assumption (Tadayasu 2014 Methods, Patient population).

Demographics in the modelled cohort (Tadayasu 2014 Methods):

* Age 40 - 69 years (mean 60.2 in 2.5 mg arm, 59.1 in 10 mg arm).
* Body mass index 18.4 - 34.4 kg/m^2 (mean 26.0 in 2.5 mg arm, 23.8 in 10 mg
  arm).
* 25 % female (only 9 of the 36 patients).
* 100 % Japanese.
* Baseline fasting plasma glucose 154.7 +/- 25.1 mg/dL (2.5 mg arm) and
  158.4 +/- 28.6 mg/dL (10 mg arm); baseline HbA1c 7.1 +/- 0.5 % and
  7.2 +/- 0.9 % respectively.
* Individual antidiabetic treatment discontinued 14 days prior to the first
  drug administration.

The model is fit to 498 + 520 = 1018 plasma concentrations and 520 + 538 =
1058 DPP-4 inhibition observations after exclusion of 9 records with protocol
violations (Tadayasu 2014 Table 2).

The same metadata are available programmatically via
`readModelDb("Tadayasu_2014_linagliptin")$population`.

# Source trace

Every `ini()` parameter in
`inst/modeldb/specificDrugs/Tadayasu_2014_linagliptin.R` carries an inline
source-trace comment. The table below collects them in one place for review;
numeric values are reproduced verbatim from Tadayasu 2014 Table 3.

| Equation / parameter             | Value         | Source location                                                |
|----------------------------------|---------------|----------------------------------------------------------------|
| 2-cmt TMDD-QE structural ODE     | n/a           | Tadayasu 2014 Figure 1 (model schematic)                       |
| QE binding (central + peripheral)| n/a           | Tadayasu 2014 Methods, Pharmacokinetic model                   |
| Occupancy PD: DPP-4 inh = Emax * Cb/BMAX | n/a   | Tadayasu 2014 Methods, Pharmacokinetic/Pharmacodynamic model   |
| `lfdepot` (F1 fixed = 1)         | 1 (fixed)     | Table 3 row F1                                                 |
| `lka` (KA)                       | 1.63 1/h      | Table 3 row KA                                                 |
| `lcl` (CL/F1)                    | 121 L/h       | Table 3 row CL/F1                                              |
| `lvc` (V2/F1)                    | 633 L         | Table 3 row V2/F1                                              |
| `lq` (Q3/F1)                     | 73.0 L/h      | Table 3 row Q3/F1                                              |
| `lvp` (V3/F1)                    | 683 L         | Table 3 row V3/F1                                              |
| `lbmax` (BMAX, central)          | 6.07 nmol/L   | Table 3 row BMAX                                               |
| `lkd` (KD)                       | 0.108 nmol/L  | Table 3 row KD                                                 |
| `lamax2` (AMAX2/F1, peripheral)  | 534 nmol      | Table 3 row AMAX2/F1                                           |
| `emax` (Emax fraction)           | 0.925         | Table 3 row EMAX (92.5 %)                                      |
| IIV F1 (CV 46.7 %)               | omega^2 = 0.19727 | Table 3 row 'IIV in F1'                                    |
| IIV KA (CV 73.6 %)               | omega^2 = 0.43288 | Table 3 row 'IIV in KA'                                    |
| IIV CL (CV 68.8 %)               | omega^2 = 0.38754 | Table 3 row 'IIV in CL'                                    |
| IIV BMAX (CV 14.2 %)             | omega^2 = 0.01996 | Table 3 row 'IIV in BMAX'                                  |
| OMEGA matrix                     | diagonal      | Tadayasu 2014 Results, Pharmacokinetic model (BMAX-CL block tested but not retained) |
| PK proportional residual (CV 27.0 %) | propSd = 0.270 | Table 3 row 'Proportional residual variability PK'         |
| PD residual Y = Y_hat + (100 - Y_hat) * eps | propSd_Dpp4Act = 0.201 | Table 3 row 'Proportional residual variability PD' |

# Virtual cohort

Original observed concentrations were not released. The simulations below use
small virtual cohorts whose dose levels match the published 5 mg, 2.5 mg, and
10 mg arms (the 5 mg arm is the simulation that supported the dose-selection
conclusion of the paper).

```{r cohort}
set.seed(20260520)

n_per_arm <- 60L

make_arm <- function(arm_label, dose_mg, id_offset) {
  dose_times <- seq(0, by = 24, length.out = 28)              # 28-day QD regimen
  ss_obs <- 27 * 24 + c(0, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 10, 12, 16, 20, 24)
  obs_times <- sort(unique(c(
    seq(0, 27 * 24, by = 6),                                  # coarse pre-SS sampling
    ss_obs                                                    # dense day-28 SS profile
  )))
  id_seq <- id_offset + seq_len(n_per_arm)
  dose_rows <- expand.grid(id = id_seq, time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, cmt = "depot")
  obs_rows  <- expand.grid(id = id_seq, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "Cc")
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::mutate(cohort = arm_label, DOSE = dose_mg) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm("2.5 mg QD", dose_mg = 2.5, id_offset =   0L),
  make_arm("5 mg QD",   dose_mg = 5,   id_offset = 200L),
  make_arm("10 mg QD",  dose_mg = 10,  id_offset = 400L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Tadayasu_2014_linagliptin")
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("cohort", "DOSE")
) |>
  as.data.frame() |>
  dplyr::as_tibble()
```

For deterministic typical-value replication of the per-arm steady-state
profile (no between-subject variability), zero out the random effects:

```{r simulate-typical}
mod_typical <- mod |> rxode2::zeroRe()
typical_events <- events |> dplyr::filter(id %in% c(1L, 201L, 401L))
sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = typical_events,
  keep   = c("cohort", "DOSE")
) |>
  as.data.frame() |>
  dplyr::as_tibble()
```

# Replicate published figures

## Figure 4 -- predicted steady-state DPP-4 inhibition

Figure 4 of Tadayasu 2014 shows the predicted steady-state DPP-4 inhibition
time profile for 2.5 mg, 5 mg, and 10 mg linagliptin once daily, with the
80 % inhibition target marked in red. The simulation here replicates the
median trajectory plus 90 % prediction interval over a single day-28 dosing
interval.

```{r figure-4}
fig4 <- sim |>
  dplyr::filter(time >= 27 * 24, time <= 28 * 24) |>
  dplyr::mutate(time_h = time - 27 * 24) |>
  dplyr::group_by(cohort, time_h) |>
  dplyr::summarise(
    Q05 = quantile(Dpp4Inh, 0.05, na.rm = TRUE),
    Q50 = quantile(Dpp4Inh, 0.50, na.rm = TRUE),
    Q95 = quantile(Dpp4Inh, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(fig4, aes(time_h, Q50, colour = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 80, colour = "red", linetype = "dashed") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(
    x = "Time after last dose (h)",
    y = "DPP-4 inhibition (%)",
    title = "Figure 4 -- Predicted steady-state DPP-4 inhibition",
    caption = "Replicates Figure 4 of Tadayasu 2014. Solid line = median, ribbon = 90 % PI; red dashed line = 80 % target."
  )
```

## Figure 3A -- linagliptin steady-state concentration VPC

Figure 3A of the paper shows a VPC of linagliptin plasma concentrations on
day 28; this chunk replicates the 2.5 mg and 10 mg arms (the two arms used in
fitting).

```{r figure-3a}
fig3a <- sim |>
  dplyr::filter(time >= 27 * 24, time <= 28 * 24, cohort %in% c("2.5 mg QD", "10 mg QD")) |>
  dplyr::mutate(time_h = time - 27 * 24) |>
  dplyr::group_by(cohort, time_h) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(fig3a, aes(time_h, Q50, colour = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(
    x = "Time after last dose (h)",
    y = "Linagliptin Cc (ng/mL, log scale)",
    title = "Figure 3A -- Linagliptin Cc VPC at steady state",
    caption = "Replicates Figure 3A of Tadayasu 2014 (2.5 mg and 10 mg arms only). Solid line = median, ribbon = 90 % PI."
  )
```

# Reproducing published E24,ss values

The simulation paper reports median predicted DPP-4 inhibition 24 h after the
last steady-state dose (E24,ss): 79.2 % (2.5 mg), 84.2 % (5 mg), and 87.7 %
(10 mg). The typical-value chunk below reproduces these directly.

```{r e24ss}
e24 <- sim_typical |>
  dplyr::filter(round(time, 1) == 28 * 24) |>
  dplyr::select(cohort, DOSE, Cc_ng_mL = Cc, Dpp4Inh) |>
  dplyr::mutate(
    paper_E24ss = dplyr::case_when(
      DOSE == 2.5 ~ 79.2,
      DOSE == 5   ~ 84.2,
      DOSE == 10  ~ 87.7
    ),
    abs_diff_pct = round(Dpp4Inh - paper_E24ss, 2)
  )
knitr::kable(
  e24,
  digits   = 2,
  caption  = "Typical-value E24,ss vs Tadayasu 2014 Simulation Results."
)
```

# PKNCA validation

PKNCA-derived steady-state Cmax, Tmax, AUCtau, and half-life from the
day-28 simulated dosing interval. The treatment grouping `| cohort + id` keeps
per-dose-group rollups available for comparison.

```{r pknca}
# Day-28 (steady-state) interval: only the 24 h after the last dose are kept
sim_nca <- sim |>
  dplyr::filter(time >= 27 * 24, time <= 28 * 24, !is.na(Cc)) |>
  dplyr::mutate(time = time - 27 * 24) |>
  dplyr::select(id, time, Cc, cohort)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | cohort + id)

# Dose rows: only the steady-state dose at the start of the chosen interval
dose_df <- events |>
  dplyr::filter(evid == 1L, time == 27 * 24) |>
  dplyr::mutate(time = 0) |>
  dplyr::select(id, time, amt, cohort)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id)

intervals <- data.frame(
  start          = 0,
  end            = 24,
  cmax           = TRUE,
  tmax           = TRUE,
  auclast        = TRUE,
  half.life      = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
suppressWarnings({
  nca_res <- PKNCA::pk.nca(nca_data)
})

nca_summary <- summary(nca_res)
knitr::kable(
  nca_summary,
  caption = "Simulated steady-state NCA parameters by dose group (day 28 dosing interval)."
)
```

# PD-side validation -- DPP-4 activity NCA

PKNCA is set up here on the DPP-4 activity output (= 100 - inhibition) as a
sanity check that the per-arm activity profile remains low and is dose-ordered.

```{r pknca-dpp4}
sim_pd <- sim |>
  dplyr::filter(time >= 27 * 24, time <= 28 * 24, !is.na(Dpp4Act)) |>
  dplyr::mutate(time = time - 27 * 24) |>
  dplyr::select(id, time, Dpp4Act, cohort)

conc_pd <- PKNCA::PKNCAconc(sim_pd, Dpp4Act ~ time | cohort + id)
nca_pd  <- PKNCA::PKNCAdata(conc_pd, dose_obj, intervals = data.frame(
  start = 0, end = 24, auclast = TRUE
))
suppressWarnings({
  nca_pd_res <- PKNCA::pk.nca(nca_pd)
})
knitr::kable(
  summary(nca_pd_res),
  caption = "Simulated AUC of DPP-4 activity over the day-28 dosing interval (lower AUC = more inhibition)."
)
```

# Assumptions and deviations

- **Custom PD residual encoded as proportional error on DPP-4 activity.** The
  paper's residual model on DPP-4 inhibition is `Y = Y_hat + (100 - Y_hat) *
  eps` with `eps ~ N(0, sigma^2)`. For a symmetric Gaussian residual this is
  algebraically identical to a proportional residual on DPP-4 activity
  (= 100 - inhibition), since `100 - Y = (100 - Y_hat) * (1 - eps)`. The
  model file therefore exposes `Dpp4Act` as the residual-bearing output with
  `Dpp4Act ~ prop(propSd_Dpp4Act)`, and emits `Dpp4Inh = 100 - Dpp4Act` as a
  derived quantity. Users with DPP-4 inhibition data should convert to
  activity (`activity = 100 - inhibition`) before fitting.
- **Concentration unit conversion (nmol/L -> ng/mL).** Tadayasu 2014 reports
  linagliptin concentrations, BMAX, KD, and AMAX2 in nmol/L (or nmol).
  Doses are clinical (2.5, 5, 10 mg). The model uses nmol/L internally so
  the published parameters can be used verbatim, but converts the `Cc`
  output to ng/mL via the linagliptin molecular weight 472.54 g/mol so the
  observation unit is dimensionally consistent with the mg dose. The
  conversion is `Cc_ng_mL = Cc_nmolL * MW / 1000`. The same pattern is used
  in the companion non-Japanese `Retlich_2015_linagliptin` model.
- **0.5 mg arm excluded.** The paper drops the 0.5 mg arm because the
  quasi-equilibrium assumption fails when free linagliptin concentrations
  drop below BMAX (Tadayasu 2014 Discussion, Pharmacokinetics). The
  packaged model inherits this scope. Simulating sub-2.5 mg doses with this
  model is outside the validated range; the predictions will be biased
  because the TMDD-QE assumption no longer holds.
- **OMEGA matrix is diagonal.** The paper tested a BMAX-CL OMEGA block with
  R = 0.837 but did not retain it in the final model because it
  overestimated the variability (Tadayasu 2014 Results, Pharmacokinetic
  model). The packaged model follows the final-model choice.
- **No subject-level covariates.** The paper does not report covariate
  effects on any structural parameter (the Japanese cohort was demographically
  narrow: 100 % Japanese, 75 % male, age 40 - 69, controlled T2DM). The
  packaged `covariateData` is therefore empty.
- **Simulation scope.** The packaged model is suitable for simulating
  Japanese T2DM patients at doses 2.5 - 10 mg linagliptin PO once daily.
  Extrapolation to other ethnicities, doses < 2.5 mg, paediatric patients,
  or patients with renal or hepatic impairment is not supported by the
  source data; see `Retlich_2015_linagliptin` for the broader Caucasian /
  multi-region popPK/PD model with covariate effects.
