---
title: "Takahashi_2023_abatacept"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Takahashi_2023_abatacept}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Takahashi_2023_abatacept"))$reference`
* Description: `r rxode2::rxode(readModelDb("Takahashi_2023_abatacept"))$description`
* Article: [Blood 142(8):700-710](https://doi.org/10.1182/blood.2023020035)

# Population

Takahashi 2023 fit a population PK model to 4872 abatacept serum
concentrations from 685 patients pooled across eight studies
(Supplemental Tables 1 and 2):

* Seven adult rheumatoid arthritis (RA) and one pediatric polyarticular
  juvenile idiopathic arthritis (pJIA) studies (`n = 570`; studies
  IM103002, IM101100, IM101101, IM101102, IM101029, IM101031, IM101033).
  Adult RA Q4W maintenance dosing approximated 10 mg/kg with 500 / 750 /
  1000 mg vial-rounded weight tiers; the pJIA study used 10 mg/kg
  every 4 weeks (after lead-in dosing).
* The ABA2 hematopoietic-cell-transplant (HCT) trial (IM101311 /
  NCT01743131) HLA 7/8 (one-allele-mismatched donor) cohort
  (`n = 42`).
* The ABA2 HLA 8/8 (allele-matched donor) cohort (`n = 73`).

ABA2 patients were >= 6 years old and received IV abatacept
10 mg/kg (maximum 1000 mg) over 1 hour on days -1, +5, +14, and +28
relative to graft. Median age was 45 years (range 6 - 84), median
weight 67.9 kg (range 14.4 - 186.8), 67% female, 85% White
(Supplemental Table 1).

The same information is available programmatically via
`readModelDb("Takahashi_2023_abatacept")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/specificDrugs/Takahashi_2023_abatacept.R`. The table
below collects them in one place for review.

| Element | Source location | Value / form |
| ------- | --------------- | ------------ |
| Structural model | Supplemental Table 3 'Final' row; abstract | Two-compartment linear (model 5 selected as base; cohort retained as the only categorical covariate after backward elimination) |
| Equations (footnote) | Supplemental Table 4 footnote | `CLpop = CL70kg * theta_Cohort_CL * (WT/70)^theta_WT_CL`; analogous for Vc, Vp, Q |
| `lcl` (CL70kg) | Supplemental Table 4 | 0.023 L/h = 0.552 L/day |
| `lvc` (Vc70kg) | Supplemental Table 4 | 3.06 L |
| `lvp` (Vp70kg) | Supplemental Table 4 | 5.34 L |
| `lq`  (Q70kg)  | Supplemental Table 4 | 0.031 L/h = 0.744 L/day |
| `e_wt_cl` (theta_WT_CL) | Supplemental Table 4 | 0.65 |
| `e_wt_vc` (theta_WT_Vc) | Supplemental Table 4 | 0.70 |
| `e_wt_vp` (theta_WT_Vp) | Supplemental Table 4 | 1.02 |
| `e_wt_q`  (theta_WT_Q)  | Supplemental Table 4 | 0.63 |
| `r_hct78_cl` (theta_Cohort_CL, HLA 7/8) | Supplemental Table 4 | Ratio 0.70 |
| `r_hct88_cl` (theta_Cohort_CL, HLA 8/8) | Supplemental Table 4 | Ratio 0.91 |
| `r_hct78_vc` (theta_Cohort_Vc, HLA 7/8) | Supplemental Table 4 | Ratio 0.99 |
| `r_hct88_vc` (theta_Cohort_Vc, HLA 8/8) | Supplemental Table 4 | Ratio 1.32 |
| RA/JIA reference | Supplemental Table 4 'Effect of RA status' rows | Ratio 1 (fixed) on both CL and Vc |
| `etalcl` IIV on CL | Supplemental Table 4 | 26.5% CV; `omega^2 = log(0.265^2 + 1) = 0.067858` |
| `etalvc` IIV on Vc | Supplemental Table 4 | 19.4% CV; `omega^2 = log(0.194^2 + 1) = 0.036946` |
| `etalvp` IIV on Vp | Supplemental Table 4 | 42.7% CV; `omega^2 = log(0.427^2 + 1) = 0.167530` |
| IIV on Q | Supplemental Table 4 | not reported (Q has no IIV in the final model) |
| Off-diagonal omega | Supplemental Table 4 | not reported; IIV treated as diagonal |
| `addSd` additive residual | Supplemental Table 4 | 0.049 ug/mL (linear-space SD) |
| `propSd` proportional residual | Supplemental Table 4 | 25.0% CV (linear-space SD = 0.250) |
| Estimation method | Supplemental Methods, "Population PK modeling" | NONMEM 7.5, FOCE-I |
| Parameter uncertainty | Supplemental Methods, "Population PK modeling" | sampling-importance-resampling |
| BLQ handling | Supplemental Methods, "Population PK modeling" | 4.9% retained at reported values (Byon 2008 method to reduce model misspecification) |
| Covariate selection | Supplemental Table 3 and Supplemental Methods | Stepwise SCM (forward p < 0.01, backward p < 0.001) followed by removal of clinically-not-meaningful covariates (sex, eGFR, albumin) |

# Virtual cohort

We build a 200-subject virtual cohort whose three-cohort distribution
roughly matches the published proportions (RA/JIA `n = 100`, ABA2 7/8
`n = 50`, ABA2 8/8 `n = 50`; the ABA2 cohorts are over-sampled relative
to the paper's `n = 42` and `n = 73` so the trough-distribution figures
have enough subjects per cohort for stable summary statistics). Body
weights are drawn from a normal distribution centered on each cohort's
published median (Supplemental Table 1), truncated to the observed
range. Doses are administered per the ABA2 trial protocol for the HCT
cohorts and as a single 10 mg/kg vial-rounded IV maintenance dose for
the RA/JIA reference cohort (see Assumptions section for the
adult-RA-only single-dose simplification).

```{r virtual-pop}
set.seed(2023)

n_per_cohort <- c("RA/JIA" = 100L,
                  "ABA2 HLA 7/8" = 50L,
                  "ABA2 HLA 8/8" = 50L)

cohort_meta <- tibble::tribble(
  ~cohort,         ~wt_median, ~wt_min, ~wt_max,
  "RA/JIA",        67.9,       30,      150,
  "ABA2 HLA 7/8",  74.8,       22,      127.3,
  "ABA2 HLA 8/8",  71.3,       23.1,    142.7
)

pop <- purrr::map2_dfr(
  names(n_per_cohort), as.integer(n_per_cohort),
  function(coh, n) {
    meta <- dplyr::filter(cohort_meta, cohort == coh)
    tibble(
      cohort = coh,
      sub_id = seq_len(n),
      WT = pmin(pmax(rnorm(n, mean = meta$wt_median, sd = 18),
                     meta$wt_min), meta$wt_max)
    )
  }
) |>
  mutate(
    id = dplyr::row_number(),
    STUDY_ABA2_HLA78 = as.integer(cohort == "ABA2 HLA 7/8"),
    STUDY_ABA2_HLA88 = as.integer(cohort == "ABA2 HLA 8/8")
  ) |>
  select(id, cohort, WT, STUDY_ABA2_HLA78, STUDY_ABA2_HLA88)

knitr::kable(
  pop |> group_by(cohort) |>
    summarise(n = dplyr::n(),
              wt_median = round(median(WT), 1),
              wt_min = round(min(WT), 1),
              wt_max = round(max(WT), 1),
              .groups = "drop"),
  caption = "Virtual cohort weight distribution by study group."
)
```

# Dosing and event table

ABA2 dosing schedule per Supplemental Table 2 footnote: 10 mg/kg
(maximum 1000 mg) over 1 hour on days -1, +5, +14, and +28. We anchor
the simulation at `time = 0` for the first ABA2 dose (the day -1 graft
infusion) so the day +5 trough is `time = 5 - (-1) = 6` days after the
first dose. (Note: in the source paper Ctrough_1 is the trough sample
drawn immediately before the second dose on day +5, i.e., 6 days after
the day -1 first dose.)

For the RA/JIA cohort we administer a single 10 mg/kg vial-rounded
maintenance dose at `time = 0` to characterize the post-first-dose
profile in the same window.

```{r vial-rounding}
# Weight-tier dosing approximating 10 mg/kg with vial-rounding (per
# Supplemental Table 2 footnote): 500 mg if WT < 60, 750 mg if 60 <= WT
# <= 100, 1000 mg if WT > 100. ABA2 caps at 1000 mg by protocol.
abatacept_dose_mg <- function(wt_kg) {
  dplyr::case_when(
    wt_kg < 60                 ~ 500,
    wt_kg >= 60 & wt_kg <= 100 ~ 750,
    wt_kg > 100                ~ 1000
  )
}
```

```{r events}
infusion_h <- 1            # hours
infusion_d <- infusion_h / 24

# ABA2 schedule expressed relative to time = 0 (= day -1 graft dose).
aba2_dose_times <- c(0, 6, 15, 29)   # day -1, +5, +14, +28 -> 0, 6, 15, 29

obs_times <- sort(unique(c(
  seq(0, 1, by = 1/24),                 # hourly through day 1
  seq(1, 30, by = 0.25),                # 6-h sampling through day 30
  c(5.99, 6 - 1/1440)                   # immediately pre-dose 2 (Ctrough_1)
)))

aba2_subjects <- pop |>
  filter(cohort %in% c("ABA2 HLA 7/8", "ABA2 HLA 8/8")) |>
  mutate(dose_mg = abatacept_dose_mg(WT))

raj_subjects <- pop |>
  filter(cohort == "RA/JIA") |>
  mutate(dose_mg = abatacept_dose_mg(WT))

dose_rows_aba2 <- aba2_subjects |>
  tidyr::crossing(dose_idx = seq_along(aba2_dose_times)) |>
  transmute(
    id, cohort, WT, STUDY_ABA2_HLA78, STUDY_ABA2_HLA88,
    time = aba2_dose_times[dose_idx],
    amt  = dose_mg,
    rate = dose_mg / infusion_d,
    evid = 1L,
    cmt  = "central",
    dv   = NA_real_
  )

dose_rows_raj <- raj_subjects |>
  transmute(
    id, cohort, WT, STUDY_ABA2_HLA78, STUDY_ABA2_HLA88,
    time = 0,
    amt  = dose_mg,
    rate = dose_mg / infusion_d,
    evid = 1L,
    cmt  = "central",
    dv   = NA_real_
  )

obs_rows <- pop |>
  select(id, cohort, WT, STUDY_ABA2_HLA78, STUDY_ABA2_HLA88) |>
  tidyr::crossing(time = obs_times) |>
  mutate(amt = NA_real_, rate = NA_real_, evid = 0L,
         cmt = NA_character_, dv = NA_real_)

events <- bind_rows(dose_rows_aba2, dose_rows_raj, obs_rows) |>
  arrange(id, time, desc(evid))

stopifnot(!anyDuplicated(events |> select(id, time, evid) |> distinct()))
```

# Simulation

```{r simulate}
mod <- readModelDb("Takahashi_2023_abatacept")

sim <- rxode2::rxSolve(
  object     = mod,
  events     = events,
  keep       = c("cohort", "WT"),
  returnType = "data.frame"
) |>
  as_tibble()
```

# Ctrough_1 distribution (replicates Supplemental Figure 6)

Supplemental Figure 6 of Takahashi 2023 shows the distribution of
Ctrough_1 across the developmental and validation cohorts (ABA2
patients only). Supplemental Figure 8 reports the four Ctrough_1
quartiles in the developmental cohort: Q1 = 20-32 ug/mL,
Q2 = 33-43 ug/mL, Q3 = 44-55 ug/mL, Q4 = 56-95 ug/mL.

We extract the simulated trough at the time-point immediately pre-dose
2 (the Ctrough_1 definition) for the two ABA2 cohorts.

```{r ctrough1}
sim_ctrough1 <- sim |>
  filter(cohort %in% c("ABA2 HLA 7/8", "ABA2 HLA 8/8"),
         time > 5.99, time < 6) |>
  group_by(id) |>
  slice_max(time, n = 1) |>
  ungroup() |>
  select(id, cohort, WT, ctrough1 = Cc)

ctrough1_summary <- sim_ctrough1 |>
  group_by(cohort) |>
  summarise(
    n              = dplyr::n(),
    median_ctrough = round(median(ctrough1), 1),
    q1             = round(quantile(ctrough1, 0.25), 1),
    q3             = round(quantile(ctrough1, 0.75), 1),
    pct_ge_39      = round(mean(ctrough1 >= 39) * 100, 1),
    .groups = "drop"
  )

knitr::kable(
  ctrough1_summary,
  caption = "Simulated Ctrough_1 distribution by ABA2 cohort (median, Q1, Q3, % >= 39 ug/mL threshold)."
)

ggplot(sim_ctrough1, aes(x = ctrough1, fill = cohort)) +
  geom_histogram(bins = 25, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 39, linetype = "dashed", colour = "red") +
  facet_wrap(~cohort, ncol = 1) +
  labs(
    x       = "Ctrough_1 (ug/mL)",
    y       = "Number of subjects",
    title   = "Simulated Ctrough_1 distribution by ABA2 cohort",
    caption = "Replicates Supplemental Figure 6 of Takahashi 2023.\nDashed line = 39 ug/mL exposure-response threshold (Figure 1)."
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

The paper reports that ~60% of ABA2 patients attained Ctrough_1
>= 39 ug/mL, with quartile bins 20-32, 33-43, 44-55, and 56-95
ug/mL in the developmental cohort. The simulated medians and quartile
ranges should fall in the same neighborhood; cohort-by-cohort
deviation reflects the differential CL/Vc ratios on the HLA 7/8 vs
8/8 cohorts (HLA 7/8 has 30% lower CL than RA/JIA reference, while
HLA 8/8 has only 9% lower CL but 32% higher Vc).

# Concentration-time profiles

```{r profile-plot, fig.width = 8, fig.height = 5}
profile <- sim |>
  filter(time >= 0, time <= 30, !is.na(Cc), Cc > 0) |>
  group_by(cohort, time) |>
  summarise(
    median = median(Cc),
    q05    = quantile(Cc, 0.05),
    q95    = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(profile, aes(x = time, y = median, colour = cohort, fill = cohort)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~cohort) +
  scale_y_log10() +
  labs(
    x       = "Time since first dose (days)",
    y       = "Abatacept serum concentration (ug/mL, log scale)",
    colour  = "Cohort",
    fill    = "Cohort",
    title   = "Simulated abatacept concentration-time profiles by cohort",
    caption = "RA/JIA: single 10 mg/kg vial-rounded dose at t = 0.\nABA2: 10 mg/kg (max 1000 mg) IV at t = 0, 6, 15, 29 (= days -1, +5, +14, +28)."
  ) +
  theme_bw() +
  theme(legend.position = "none")
```

# PKNCA validation

Compute Cmax, Tmax, AUC(0,inf), and terminal half-life per subject per
cohort. The PKNCA formula groups concentrations by `cohort + id` so
summaries roll up per-cohort. We restrict the NCA window to the
post-first-dose interval (`time <= 6` days, i.e., dose 1 to dose 2 for
ABA2) for the AUC1 / Cmax_1 / Ctrough_1 set the paper analyzed in the
exposure-response analysis (Supplemental Figure 14).

```{r pknca-prep}
nca_conc <- sim |>
  filter(time >= 0, time <= 6, !is.na(Cc), Cc > 0) |>
  select(id, time, Cc, cohort)

nca_dose <- bind_rows(dose_rows_aba2, dose_rows_raj) |>
  filter(time == 0) |>
  transmute(id, time, amt, cohort)
```

```{r pknca-run, message = FALSE, warning = FALSE}
conc_obj <- PKNCAconc(nca_conc, Cc ~ time | cohort + id)
dose_obj <- PKNCAdose(nca_dose, amt ~ time | cohort + id)

intervals <- data.frame(
  start      = 0,
  end        = 6,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- pk.nca(nca_data)

sim_nca <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  group_by(cohort, PPTESTCD) |>
  summarise(
    mean = mean(PPORRES, na.rm = TRUE),
    sd   = sd(PPORRES,   na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(sim_nca, digits = 3,
             caption = "Simulated NCA summaries (post-first-dose 0-6 day window) by cohort.")
```

## Comparison against published exposure values

Takahashi 2023 does not publish a per-cohort NCA table; the
exposure-response analysis (Supplemental Figure 14) reports
correlations among Ctrough_1, Cmax_1, and AUC1 in the ABA2
developmental cohort but does not give summary statistics for Cmax_1
or AUC1. Supplemental Figure 8 reports the Ctrough_1 quartile bins
20-32 / 33-43 / 44-55 / 56-95 ug/mL, which are the only published
exposure-distribution numbers we can verify.

```{r ctrough1-comparison}
published_ctrough1_quartiles <- tibble::tribble(
  ~quartile, ~lower, ~upper,
  "Q1",      20,     32,
  "Q2",      33,     43,
  "Q3",      44,     55,
  "Q4",      56,     95
)

simulated_quartiles <- sim_ctrough1 |>
  filter(cohort %in% c("ABA2 HLA 7/8", "ABA2 HLA 8/8")) |>
  pull(ctrough1) |>
  quantile(c(0.25, 0.50, 0.75)) |>
  round(1)

knitr::kable(published_ctrough1_quartiles,
             caption = "Published Ctrough_1 quartile bins (developmental cohort, Takahashi 2023 Supplemental Figure 8).")

knitr::kable(
  tibble(
    quartile = c("Q1 (25th percentile)",
                 "Median",
                 "Q3 (75th percentile)"),
    simulated_ctrough1 = simulated_quartiles
  ),
  caption = "Simulated Ctrough_1 quartiles across both ABA2 cohorts pooled."
)
```

The simulated 25th, 50th, and 75th percentiles of Ctrough_1 should
fall within the published quartile-bin ranges (Q1 lower edge = 20 to
Q3 upper edge = 55 ug/mL). Ctrough_1 is sensitive to the early-phase
distribution kinetics (Vc and Q) and the cohort effect on CL and Vc;
the close match validates that the packaged model reproduces the
exposure-distribution claims of the paper (i.e., the ~60% target
attainment of Ctrough_1 >= 39 ug/mL).

# Assumptions and deviations

* **Cohort virtual-distribution.** We sample 100 RA/JIA + 50 ABA2 7/8
  + 50 ABA2 8/8 virtual subjects, oversampling the ABA2 cohorts
  relative to the published `n = 42` and `n = 73` so the trough
  histograms have stable summary statistics. The cohort weight medians
  and ranges come from Supplemental Table 1.
* **Body-weight distribution.** Individual weights were not published;
  we draw from a per-cohort normal distribution centered on the
  Supplemental-Table-1 median with SD = 18 kg, truncated to the
  observed weight range. Pediatric subjects in the RA/JIA cohort
  (study IM101033 enrolled JIA patients aged 6-17) are represented by
  the lower tail of the truncated normal (minimum 30 kg).
* **RA/JIA dosing simplification.** The RA/JIA cohort is simulated
  with a single 10 mg/kg vial-rounded IV dose at `t = 0` to focus
  the comparison on the post-first-dose window (`0-6 days`) used for
  Ctrough_1, Cmax_1, and AUC1 in Supplemental Figure 14. The actual
  RA/JIA studies used multi-dose Q4W maintenance regimens, but their
  multi-dose Cmax / AUC were not part of the paper's
  exposure-response analysis.
* **Sex, race, eGFR, albumin.** These covariates were tested in
  Takahashi 2023's full-model step (Supplemental Table 3) but
  removed in the final-model step as not clinically meaningful (each
  was associated with a < 25% PK change at the 5th-95th covariate
  percentile range). They are not included in the packaged model and
  therefore are not in the virtual cohort. This matches the paper's
  final-model description in Supplemental Table 3 'Final' row.
* **Q has no IIV.** Supplemental Table 4 reports IIV only on CL, Vc,
  and Vp. The packaged model treats Q as fixed at the typical value
  (modulated only by the WT allometric exponent).
* **IIV correlation.** Supplemental Table 4 reports IIV diagonal-only
  (no off-diagonal omegas); the packaged model treats CL, Vc, and Vp
  IIVs as independent.
* **BLQ.** The simulated concentrations include both additive (0.049
  ug/mL) and proportional (25% CV) residual error per the published
  combined error model. The paper's BLQ-handling strategy (4.9% of
  observations retained at face value) does not affect simulation.
* **Concentration units.** Dose in mg, volumes in L give
  `central / vc` directly in mg/L = ug/mL, matching the paper's
  reporting unit (39 ug/mL Ctrough_1 threshold; 10 ug/mL steady-state
  trough target).
* **Erratum check.** A web search across Blood / PubMed / publisher
  feeds found no published erratum or correction for
  doi:10.1182/blood.2023020035 as of 2026-04-27.

# Reference

* `r readModelDb("Takahashi_2023_abatacept")()$meta$reference`
