---
title: "Vancomycin (Takada 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Takada 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Takada_2025_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Takada_2025_vancomycin"))$description`
* Article: <https://doi.org/10.1186/s40780-025-00423-8>
* Supplement (Additional Files 1-10, Springer ESM):
  <https://static-content.springer.com/esm/art%3A10.1186%2Fs40780-025-00423-8/MediaObjects/40780_2025_423_MOESM1_ESM.docx>
  (files `MOESM1` through `MOESM10`)

Note on supplement numbering: the paper's "Additional File *N*" labels do not
line up with the `MOESM` file indices on the publisher's server. The two
load-bearing files here are the renal-function equations (the paper's Additional
File 2: Table 2, served as `MOESM4`) and the clearance nomogram (Additional
File 6: Table 5, served as `MOESM7`). Citations below give both the paper's
label and the file index so either route reaches the right table.

Takada 2025 develops a two-compartment intravenous population PK model for
vancomycin in Japanese inpatients aged 75 years and older with a body mass
index below 25 kg/m^2. The analysis is motivated by a specific clinical
failure mode: creatinine-based renal-function equations systematically
underestimate clearance in patients of advanced age with low muscle mass, so a
model that carries only creatinine clearance mis-predicts vancomycin clearance
in exactly the patients most at risk. The paper's contribution is the addition
of **serum albumin** as a second covariate on clearance -- albumin correlates
with muscle mass in this age group -- which lowered the objective function by
5.11 points over the creatinine-clearance-only model (Table 2, model 4 versus
model 3) and gave the lowest mean absolute and mean squared prediction error of
any tested model in the subgroup with serum creatinine below 0.6 mg/dL.

# Population

The model was fit to 417 vancomycin concentrations (65 peaks, 352 troughs) from
159 patients treated at Yokohama General Hospital between August 2016 and
September 2024 (Takada 2025 Table 1). Median age was 84 years (range 75-99),
with 49.7% aged 85 or older; 42.1% were female. The cohort was small and frail:
median body weight 47 kg (range 26-70), median body mass index 18.6 kg/m^2
(range 11.0-24.8, with 48.4% below 18.5), and median serum albumin 2.3 g/dL
(range 1.2-4.2, with 90.6% below 3.0 g/dL). Median Cockcroft-Gault creatinine
clearance was 51.5 mL/min (range 9.7-121.2) and median serum creatinine
0.64 mg/dL (range 0.22-3.00), with 42.1% below 0.60 mg/dL. Daily maintenance
doses were chosen by the treating physician, median 1500 mg/day (range
250-3000). Patients on dialysis, patients treated for fewer than three days,
and patients with no vancomycin sample were excluded.

A separate 133-patient multicentre cohort (eight hospitals, September 2020 to
December 2023) was used only for external validation of predictive performance
and contributed nothing to the parameter estimates reproduced here.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Takada_2025_vancomycin")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Takada_2025_vancomycin.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at reference covariates) | 1.96 L/h | Table 3, theta1 (SE 0.06, CV 3.11%, 95% CI 1.84-2.08) |
| `e_crcl_cl` (exponent on CLcr) | 0.63 | Table 3, theta2 (SE 0.06, CV 9.72%, 95% CI 0.51-0.76) |
| `e_alb_cl` (exponent on Alb) | 0.22 | Table 3, theta3 (SE 0.09, CV 41.75%, 95% CI 0.03-0.40) |
| `lq` (intercompartmental clearance) | 4.86 L/h | Table 3, theta4 (SE 0.91, CV 18.63%, 95% CI 3.08-6.64); Abstract agrees. See Errata for the conflicting 3.24 in the Results narrative |
| `lvc` (central volume) | 31.78 L | Table 3, theta5 (SE 3.86, CV 12.16%, 95% CI 24.19-39.38) |
| `lvp` (peripheral volume) | 53.64 L | Table 3, theta6 (SE 4.22, CV 7.86%, 95% CI 45.36-61.93) |
| `etalcl` (IIV variance on CL) | 0.11 | Table 3 footnote: "eta ... normally distributed with mean 0 and variance omega^2, etaCL = 0.11" |
| `propSd`, `addSd` | 0 (fixed) | Not reported. Additional File 1: Table 1 (file MOESM3) lists the three candidate residual models evaluated but no estimate is published anywhere. See Errata |
| CL covariate equation | `1.96 * (CLcr/3.09)^0.63 * (Alb/2.3)^0.22 * exp(etaCL)` | Table 3 header row and Results, "Development of a population pharmacokinetic model" |
| Reference CLcr = 3.09 L/h | 51.5 mL/min | Table 1 modeling-cohort median estimated creatinine clearance |
| Reference Alb = 2.3 g/dL | 23 g/L | Table 1 modeling-cohort median serum albumin |
| CLcr estimating equation | `([140 - age] * BW) / (SCr * 72)`, female `* 0.85` | Additional File 2: Table 2 (file MOESM4), Equations 1-2 |
| Two-compartment structure | n/a | Methods ("one- or two-compartment models of the first-order elimination were fitted"); Results, "A two-compartment model was optimal for VCM" |
| Published CL nomogram | 70-cell grid | Additional File 6: Table 5 (file MOESM7) |
| Published dose nomogram | 70-cell grid | Table 4 |
| Published safety nomogram | 70-cell risk grid | Additional File 10: Table 8 (file MOESM10); legend: low risk below 10%, moderate 10 to below 25%, high 25% or above |
| MIC distribution used for target attainment | 0.25: 0.1%, 0.5: 11.6%, 1.0: 79.3%, 2.0: 9.0% (ug/mL) | Methods, Monte Carlo section, citing the 2019 Japanese MRSA guidelines |

```{r model}
mod <- readModelDb("Takada_2025_vancomycin")
ui <- rxode2::rxode(mod)
ui$iniDf |>
  dplyr::filter(!is.na(ntheta)) |>
  dplyr::select(name, est, fix, label) |>
  dplyr::rename(
    "Parameter" = name, "Estimate" = est, "Fixed" = fix, "Label" = label
  ) |>
  knitr::kable(caption = "Packaged `ini()` values (log-scale entries shown as stored).")
```

# Check 1 -- the published clearance nomogram (Additional File 6: Table 5, file MOESM7)

Takada 2025 prints a 70-cell grid of typical clearance over creatinine
clearance 1.2-5.1 L/h (20-85 mL/min) and serum albumin 1.5-3.5 g/dL. That grid
is a direct, closed-form test of the covariate equation: if the packaged
parameters and the unit conversions in `model()` are right, the equation must
reproduce every published cell.

There is one wrinkle, which is a property of the paper and not of this
implementation. The paper's own nomogram substitutes the *constant* 0.11 into
the `exp(etaCL)` term, treating the between-subject variance as a fixed
multiplier of `exp(0.11) = 1.1163`. That is confirmed by the Discussion, which
quotes CL = 2.73 and 3.29 L/h at CLcr 5.1 L/h for albumin 1.5 and 3.5 g/dL --
values that are only recovered with the extra factor. The packaged model treats
`etaCL` correctly, as a mean-zero random effect with variance 0.11, so its
typical-value clearance is exactly `exp(0.11)` = 11.6% below every printed cell.
Both readings are shown below.

```{r nomogram-closed-form}
crcl_grid <- c(1.2, 1.5, 1.8, 2.1, 2.4, 2.7, 3.0, 3.3, 3.6, 3.9, 4.2, 4.5, 4.8, 5.1)
alb_grid  <- c(1.5, 2.0, 2.5, 3.0, 3.5)

# Takada 2025 Additional File 6: Table 5 (file MOESM7), transcribed as printed.
published_cl <- matrix(
  c(1.09, 1.25, 1.41, 1.55, 1.69, 1.82, 1.95, 2.07, 2.19, 2.30, 2.41, 2.52, 2.62, 2.73,
    1.16, 1.34, 1.50, 1.65, 1.80, 1.94, 2.07, 2.20, 2.33, 2.45, 2.57, 2.68, 2.80, 2.91,
    1.22, 1.40, 1.58, 1.74, 1.89, 2.04, 2.18, 2.32, 2.45, 2.57, 2.70, 2.82, 2.94, 3.05,
    1.27, 1.46, 1.64, 1.81, 1.97, 2.12, 2.27, 2.41, 2.55, 2.68, 2.81, 2.93, 3.06, 3.18,
    1.31, 1.51, 1.70, 1.87, 2.04, 2.20, 2.35, 2.49, 2.64, 2.77, 2.91, 3.04, 3.16, 3.29),
  nrow = length(alb_grid), byrow = TRUE,
  dimnames = list(paste0("Alb ", alb_grid), paste0("CLcr ", crcl_grid))
)

cl_equation <- function(crcl, alb, eta = 0) {
  1.96 * (crcl / 3.09)^0.63 * (alb / 2.3)^0.22 * exp(eta)
}

recomputed_paper <- outer(alb_grid, crcl_grid, function(a, c) cl_equation(c, a, eta = 0.11))
recomputed_model <- outer(alb_grid, crcl_grid, function(a, c) cl_equation(c, a, eta = 0))

max_dev_paper <- max(abs(round(recomputed_paper, 2) - published_cl))
ratio <- published_cl / recomputed_model
rel_dev <- abs(ratio / exp(0.11) - 1)

stopifnot(
  # Every published cell is reproduced to within one unit in the last printed
  # decimal place when the paper's exp(0.11) convention is applied. (The
  # epsilon absorbs binary floating-point representation of 0.01, not any
  # real deviation.)
  max_dev_paper <= 0.01 + 1e-9,
  # ... and the offset from the packaged (eta = 0) typical value is exactly
  # exp(0.11) in all 70 cells, to within the granularity of the published
  # two-decimal rounding. A signature, not scatter: real parameter error would
  # vary systematically across the grid rather than sitting on a constant.
  all(rel_dev < 0.015),
  length(published_cl) == 70L
)

cat(sprintf(
  "All %d published cells reproduced; max absolute deviation %.3f L/h.\n",
  length(published_cl), max_dev_paper
))
cat(sprintf(
  "published / packaged-typical ratio: %.4f to %.4f (exp(0.11) = %.4f);\n  max relative deviation from exp(0.11): %.2f%%.\n",
  min(ratio), max(ratio), exp(0.11), 100 * max(rel_dev)
))
```

The Discussion's two quoted values are recovered exactly:

```{r nomogram-discussion}
tibble::tibble(
  `Quantity` = c("CL at CLcr 5.1 L/h, Alb 1.5 g/dL", "CL at CLcr 5.1 L/h, Alb 3.5 g/dL"),
  `Takada 2025 Discussion (L/h)` = c(2.73, 3.29),
  `Recomputed with exp(0.11) (L/h)` = round(c(cl_equation(5.1, 1.5, 0.11), cl_equation(5.1, 3.5, 0.11)), 2),
  `Packaged typical value (L/h)` = round(c(cl_equation(5.1, 1.5), cl_equation(5.1, 3.5)), 2)
) |>
  knitr::kable(caption = "Takada 2025 Discussion cross-check of the clearance equation.")
```

# Virtual cohort

Original observed data are not publicly available. The simulations below use a
virtual population whose demographics approximate the published modeling-cohort
characteristics (Takada 2025 Table 1). Age, sex, body weight, serum creatinine,
and serum albumin are sampled from truncated distributions matched to the
published medians and ranges; creatinine clearance is then **computed** with the
paper's own Cockcroft-Gault equation (Additional File 2: Table 2, file MOESM4) rather than
sampled directly, so the covariate pipeline is exercised end to end.

```{r cohort}
set.seed(20250307)

n_subj <- 159L  # matches the Takada 2025 modeling cohort

# Sample from a log-normal truncated to the published range, calibrated so the
# median equals the published median.
rtrunc_lnorm <- function(n, med, lo, hi, cv) {
  out <- numeric(0)
  while (length(out) < n) {
    draw <- stats::rlnorm(2 * n, meanlog = log(med), sdlog = sqrt(log(cv^2 + 1)))
    out <- c(out, draw[draw >= lo & draw <= hi])
  }
  out[seq_len(n)]
}

draw_subjects <- function(n) {
  tibble::tibble(
    AGE     = round(rtrunc_lnorm(n, med = 84,   lo = 75,   hi = 99,   cv = 0.07)),
    SEXF    = stats::rbinom(n, 1L, 0.421),
    WT      = round(rtrunc_lnorm(n, med = 47,   lo = 26,   hi = 70,   cv = 0.20), 1),
    CREAT   = round(rtrunc_lnorm(n, med = 0.64, lo = 0.22, hi = 3.00, cv = 0.55), 2),
    ALB_gdL = round(rtrunc_lnorm(n, med = 2.3,  lo = 1.2,  hi = 4.2,  cv = 0.22), 1)
  ) |>
    dplyr::mutate(
      # Takada 2025 Additional File 2: Table 2 (file MOESM4), Equations 1-2.
      CRCL = ((140 - AGE) * WT) / (CREAT * 72) * ifelse(SEXF == 1L, 0.85, 1),
      # Canonical ALB is SI g/L; the paper reports g/dL.
      ALB  = ALB_gdL * 10
    )
}

# Demographics are sampled independently, so a few draws combine into a
# creatinine clearance outside the observed range. Reject those rather than
# extrapolate the covariate model past the data it was fit to (Takada 2025
# Table 1 CLcr range 9.7-121.2 mL/min; the Limitations note that even the
# published extremes are sparsely populated).
cohort <- draw_subjects(8L * n_subj) |>
  dplyr::filter(CRCL >= 9.7, CRCL <= 121.2) |>
  dplyr::slice_head(n = n_subj) |>
  dplyr::mutate(id = seq_len(n_subj))
stopifnot(nrow(cohort) == n_subj)

# The cohort is only credible if the DERIVED creatinine clearance lands on the
# published distribution, so check it rather than assume it.
cohort_check <- tibble::tribble(
  ~Characteristic,                  ~`Takada 2025 Table 1`,     ~Simulated,
  "Age (years), median [range]",    "84 [75-99]",               sprintf("%.0f [%.0f-%.0f]", median(cohort$AGE), min(cohort$AGE), max(cohort$AGE)),
  "Female (%)",                     "42.1",                     sprintf("%.1f", 100 * mean(cohort$SEXF)),
  "Body weight (kg), median [range]", "47 [26-70]",             sprintf("%.0f [%.0f-%.0f]", median(cohort$WT), min(cohort$WT), max(cohort$WT)),
  "Serum creatinine (mg/dL), median [range]", "0.64 [0.22-3.00]", sprintf("%.2f [%.2f-%.2f]", median(cohort$CREAT), min(cohort$CREAT), max(cohort$CREAT)),
  "Serum albumin (g/dL), median [range]", "2.3 [1.2-4.2]",      sprintf("%.1f [%.1f-%.1f]", median(cohort$ALB_gdL), min(cohort$ALB_gdL), max(cohort$ALB_gdL)),
  "CLcr (mL/min), median [range]",  "51.5 [9.7-121.2]",         sprintf("%.1f [%.1f-%.1f]", median(cohort$CRCL), min(cohort$CRCL), max(cohort$CRCL))
)

knitr::kable(
  cohort_check,
  caption = "Virtual cohort versus the Takada 2025 modeling-cohort demographics. CLcr is derived from the sampled demographics via the paper's Cockcroft-Gault equation, not sampled."
)

# The derived CLcr median must land near the published 51.5 mL/min for the
# cohort to be a fair stand-in; a gate that cannot go red is not a gate.
stopifnot(abs(median(cohort$CRCL) - 51.5) < 8)
```

# Simulation

Vancomycin is given as an intermittent intravenous infusion. Takada 2025 does
not state the infusion duration, so a one-hour infusion is assumed; steady-state
exposure (`AUCss`), which is what the paper's targets and nomograms are built
on, does not depend on that choice. Three regimens spanning the published dose
range are simulated at steady state (`ss = 1`, `ii = 12`): 500, 750, and
1000 mg every 12 hours, i.e. 1000, 1500 (the cohort median), and 2000 mg/day.

```{r events}
tau <- 12  # dosing interval (h)
obs_times <- seq(0, 24, by = 0.25)

make_arm <- function(cohort, dose_mg, label, id_offset) {
  covs <- cohort |>
    dplyr::mutate(id = id + id_offset, regimen = label) |>
    dplyr::select(id, regimen, CRCL, ALB)

  doses <- covs |>
    dplyr::mutate(
      time = 0, amt = dose_mg, rate = dose_mg / 1,  # 1-hour infusion
      evid = 1L, cmt = "central", ii = tau, ss = 1L, addl = 1L
    )
  obs <- covs |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(
      amt = NA_real_, rate = NA_real_, evid = 0L,
      cmt = "central",  # the ODE state, never the observable `Cc`
      ii = 0, ss = 0L, addl = 0L
    )
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(cohort,  500, "500 mg q12h",  id_offset =   0L),
  make_arm(cohort,  750, "750 mg q12h",  id_offset = 200L),
  make_arm(cohort, 1000, "1000 mg q12h", id_offset = 400L)
)
# Check the event table directly: wrapping this in `unique()` first would strip
# the very duplicates being tested for, making the assertion unfalsifiable.
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))

sim <- rxode2::rxSolve(mod, events = events, keep = c("regimen", "CRCL", "ALB")) |>
  as.data.frame()
stopifnot(nrow(sim) > 0L, !all(is.na(sim$Cc)))
```

# Figure 1 -- steady-state concentration-time profile

Takada 2025 Figure 1 is a visual predictive check of vancomycin concentration
against time after dosing, with the 5th, 50th, and 95th percentiles of the
prediction overlaid on the observations. The observed concentrations are not
available, so the panel below shows the simulated percentile bands alone. The
bands reflect between-subject variability in clearance (variance 0.11) and in
the sampled covariates only; because Takada 2025 publishes no residual-error
estimate, the packaged model carries none, so these bands are narrower than the
paper's.

```{r figure-1}
# rxSolve returns observation rows only, and carries no `evid` column.
sim |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = stats::quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~regimen) +
  labs(
    x = "Time after dose (h)", y = "Vancomycin concentration (ug/mL)",
    title = "Steady-state vancomycin profiles",
    caption = "Structure of Figure 1 of Takada 2025 (5th / 50th / 95th percentiles); observed data not available."
  ) +
  theme_bw()
```

# PKNCA validation

Non-compartmental analysis is run over the steady-state dosing interval
(0 to `r tau` h), which under `ss = 1` is a genuine steady-state interval.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(time <= tau) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

# Time-zero records are produced by the observation grid; assert rather than assume.
stopifnot(all(sim_nca |> dplyr::group_by(id) |> dplyr::summarise(z = any(time == 0)) |> dplyr::pull(z)))

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | regimen + id,
  concu = "ug/mL", timeu = "h"
)

# One dose row per subject anchors the interval so `cl.last` is Dose_tau / AUC_tau.
dose_df <- events |>
  dplyr::filter(evid == 1, time == 0) |>
  dplyr::select(id, time, amt, regimen)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | regimen + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, cmin = TRUE, tmax = TRUE,
  auclast = TRUE, cav = TRUE, cl.last = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result)
stopifnot(nrow(nca_tbl) > 0L)
```

## Mass-balance gate: AUCss must equal daily dose divided by clearance

At steady state the exposure identity `AUC(0-24) = daily dose / CL` holds
exactly for a linear model, independent of compartment count, infusion
duration, and dosing interval. It is therefore the sharpest available check
that the packaged clearance -- covariate terms, unit conversions, and all --
is wired correctly.

```{r mass-balance}
auc_tau <- nca_tbl |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(regimen, id, auc_tau = PPORRES)

daily_dose <- c("500 mg q12h" = 1000, "750 mg q12h" = 1500, "1000 mg q12h" = 2000)

# Each subject's realised clearance, recovered from the simulation output. This
# is the right comparator: it carries that subject's covariates AND their draw
# of etaCL, so the identity below tests the whole chain, not just the typical
# value.
cl_realised <- sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl_sim = dplyr::first(cl), .groups = "drop")
stopifnot(nrow(cl_realised) == 3L * n_subj, !anyNA(cl_realised$cl_sim))

mb <- auc_tau |>
  dplyr::mutate(
    auc_ss24 = auc_tau * (24 / tau),
    dose_day = unname(daily_dose[as.character(regimen)])
  ) |>
  dplyr::left_join(cl_realised, by = "id") |>
  dplyr::mutate(
    cl_nca = dose_day / auc_ss24,
    pct_diff_realised = 100 * (cl_nca - cl_sim) / cl_sim
  )
stopifnot(nrow(mb) == 3L * n_subj, !anyNA(mb$pct_diff_realised))

mb |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    `Subjects` = dplyr::n(),
    `Median AUCss 0-24 (ug*h/mL)` = round(stats::median(auc_ss24), 1),
    `Median NCA CL (L/h)` = round(stats::median(cl_nca), 3),
    `Median model CL (L/h)` = round(stats::median(cl_sim), 3),
    `Max |% diff|` = round(max(abs(pct_diff_realised)), 3),
    .groups = "drop"
  ) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(caption = "Mass-balance gate: NCA-derived clearance (daily dose / AUCss) against each subject's model clearance.")

# NCA on a 0.25-h grid recovers each subject's clearance to better than 0.1%.
# The bound is set to the accuracy actually achieved, not a loose one, so a
# future change to the model or the observation grid trips it.
stopifnot(max(abs(mb$pct_diff_realised)) < 0.1)
```

## Comparison against the published clearance nomogram

The paper reports no observed NCA table, but Additional File 6: Table 5 (file MOESM7)
publishes typical clearance across a covariate grid, which the simulation can
be made to reproduce directly. A deterministic arm (`omega = NA`, so
`etaCL = 0`) is simulated at each grid cell, and `cl.last`
(`Dose_tau / AUC_tau`) is compared against the published value.

```{r deterministic-grid}
grid_cells <- tidyr::crossing(alb_gdL = alb_grid, crcl_Lh = crcl_grid) |>
  dplyr::mutate(
    id        = dplyr::row_number(),
    cell      = sprintf("Alb %.1f | CLcr %.0f", alb_gdL, crcl_Lh / 0.06),
    CRCL      = crcl_Lh / 0.06,
    ALB       = alb_gdL * 10,
    # Index the published matrix BY NAME-free position lookup, so the join can
    # never silently transpose if `crossing()` changes its ordering.
    published = published_cl[cbind(match(alb_gdL, alb_grid), match(crcl_Lh, crcl_grid))]
  )
stopifnot(nrow(grid_cells) == 70L, !anyNA(grid_cells$published))

grid_dose <- 1000
grid_events <- dplyr::bind_rows(
  grid_cells |>
    dplyr::select(id, cell, CRCL, ALB) |>
    dplyr::mutate(time = 0, amt = grid_dose, rate = grid_dose / 1,
                  evid = 1L, cmt = "central", ii = tau, ss = 1L, addl = 1L),
  grid_cells |>
    dplyr::select(id, cell, CRCL, ALB) |>
    tidyr::crossing(time = seq(0, tau, by = 0.25)) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L,
                  cmt = "central", ii = 0, ss = 0L, addl = 0L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

grid_sim <- rxode2::rxSolve(mod, events = grid_events, omega = NA,
                            keep = c("cell", "CRCL", "ALB")) |>
  as.data.frame()

grid_conc <- grid_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, cell)
grid_dose_df <- grid_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, cell)

grid_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(grid_conc), Cc ~ time | cell + id,
                   concu = "ug/mL", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(grid_dose_df), amt ~ time | cell + id, doseu = "mg"),
  intervals = data.frame(start = 0, end = tau, auclast = TRUE, cl.last = TRUE)
))

grid_cl <- as.data.frame(grid_nca$result) |>
  dplyr::filter(PPTESTCD == "cl.last") |>
  dplyr::select(cell, cl_nca = PPORRES)

grid_cmp <- grid_cells |>
  dplyr::select(cell, alb_gdL, crcl_Lh, published) |>
  dplyr::left_join(grid_cl, by = "cell") |>
  dplyr::mutate(ratio = cl_nca / published)

stopifnot(
  nrow(grid_cmp) == 70L,
  !anyNA(grid_cmp$cl_nca),
  # Simulated typical-value CL sits exactly exp(-0.11) below every published
  # cell -- the deterministic offset explained above, with no extra scatter
  # beyond the published two-decimal rounding.
  all(abs(grid_cmp$ratio / exp(-0.11) - 1) < 0.015)
)
cat(sprintf(
  "All %d grid cells simulated. Simulated / published CL ratio: %.4f to %.4f (exp(-0.11) = %.4f);\n  max relative deviation %.2f%%.\n",
  nrow(grid_cmp), min(grid_cmp$ratio), max(grid_cmp$ratio), exp(-0.11),
  100 * max(abs(grid_cmp$ratio / exp(-0.11) - 1))
))
```

A readable subset is rendered side by side below. Every row differs from the
published value by the same 10.4%, which is `exp(-0.11)` expressed as a
percentage difference -- the paper's nomogram convention, not a discrepancy in
the packaged parameters.

```{r nca_compare}
subset_cells <- grid_cmp |>
  dplyr::filter(alb_gdL %in% c(1.5, 2.5, 3.5), crcl_Lh %in% c(1.2, 2.4, 3.6, 5.1))

simulated_nca <- subset_cells |>
  dplyr::transmute(cell, PPTESTCD = "cl.last", PPORRES = cl_nca)
reference_nca <- subset_cells |>
  dplyr::transmute(cell, cl.last = published)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = as.data.frame(simulated_nca),
  reference     = as.data.frame(reference_nca),
  by            = "cell",
  units         = c(cl.last = "L/h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated steady-state clearance versus Takada 2025 Additional File 6: Table 5 (file MOESM7). Cell labels are 'Alb (g/dL) | CLcr (mL/min)'. * differs from reference by >20%.",
  align = c("l", "l", "r", "r", "r")
)

# No row should be flagged: the offset is a uniform ~10.4%, well inside the
# 20% tolerance. A non-NULL footnote here would mean something else moved.
stopifnot(is.null(attr(cmp, "footnote")))
```

## Reproducing the published dose nomogram (Table 4)

Takada 2025 Table 4 gives, for each covariate cell, the smallest daily
maintenance dose (in 250 mg steps) reaching a probability of target attainment
of at least 85% for `AUCss/MIC >= 400`. The full Monte Carlo behind that table
depends on unpublished assumptions (whether `etaCL` was resampled, the exact
covariate distributions), but its deterministic floor is recoverable: at
MIC = 1 ug/mL the typical patient needs `AUCss >= 400`, hence
`daily dose >= 400 * CL`. The published dose must therefore be at least that
floor, and -- since target attainment across the MIC distribution needs only
modest headroom -- should sit within one 250 mg step of it.

```{r dose-nomogram}
published_dose <- matrix(
  c( 500, 750, 750, 750,  750, 1000, 1000, 1000, 1000, 1250, 1250, 1250, 1500, 1500,
     750, 750, 750, 750, 1000, 1000, 1000, 1000, 1250, 1250, 1500, 1500, 1500, 1500,
     750, 750, 750, 750, 1000, 1000, 1000, 1250, 1250, 1500, 1500, 1500, 1500, 1500,
     750, 750, 750, 1000, 1000, 1000, 1250, 1250, 1500, 1500, 1500, 1500, 1500, 1500,
     750, 750, 750, 1000, 1000, 1000, 1250, 1500, 1500, 1500, 1500, 1500, 1500, 1500),
  nrow = length(alb_grid), byrow = TRUE,
  dimnames = dimnames(published_cl)
)

# The floor uses `recomputed_paper` -- the clearance equation evaluated with the
# paper's own exp(0.11) convention -- rather than the printed two-decimal grid.
# Both stay inside the paper's arithmetic, but the printed grid's rounding can
# push 400 * CL across a 250 mg boundary and manufacture a spurious extra step
# (it does so in exactly one of the 70 cells).
floor_dose <- ceiling(400 * recomputed_paper / 250) * 250
gap <- published_dose - floor_dose

stopifnot(
  length(published_dose) == 70L,
  all(gap >= 0),    # never below the AUCss/MIC = 400 floor
  all(gap <= 250)   # never more than one 250 mg step above it
)

tibble::tibble(
  `Check` = c(
    "Cells evaluated",
    "Published dose at or above the AUCss/MIC = 400 floor",
    "Published dose exactly at the floor",
    "Published dose one 250 mg step above the floor",
    "Published dose more than one step above the floor"
  ),
  `Result` = c(
    length(published_dose),
    sum(gap >= 0), sum(gap == 0), sum(gap == 250), sum(gap > 250)
  )
) |>
  knitr::kable(caption = "Takada 2025 Table 4 against the deterministic AUCss/MIC >= 400 dose floor implied by the paper's own clearance nomogram.")
```

## Steady-state exposure against the paper's therapeutic window

Takada 2025 targets `AUCss` between 400 and 600 ug*h/mL: at least 400 for
efficacy (`AUCss/MIC >= 400` at the modal MIC of 1 ug/mL, which covers 79.3% of
the surveillance isolates cited from the 2019 Japanese MRSA guidelines) and
below 600 to limit acute kidney injury risk.

Of the three simulated regimens, only 1000 mg/day places the cohort median
inside that window. At 1500 mg/day -- the dose actually prescribed most often in
the source cohort -- the simulated median exposure is well above 600 ug*h/mL and
most subjects exceed the safety threshold.

That is not a discrepancy with the paper; it is the paper's central argument
restated. Clearance in this population is low (typical 1.96 L/h versus
2.45-4.73 L/h in previous Japanese analyses), so the doses the paper actually
recommends are *below* observed practice: at a common creatinine clearance of
40 mL/min, Table 4 gives 750 mg/day at an albumin of 1.5 g/dL -- less than the
guideline's 1000 mg -- and 1000 mg/day at 2.0-3.5 g/dL. The cohort's observed
median of 1500 mg/day is one to two 250 mg steps above the recommendation for a
typical patient, which is precisely why the paper concludes that early
therapeutic drug monitoring and a one-step dose reduction should be considered
for patients at risk of kidney injury.

The check below makes that concrete: applying the paper's own Table 4 dose to
each covariate cell keeps typical-value exposure at or below roughly
640 ug*h/mL everywhere on the grid, whereas the cohort's observed median dose
does not.

```{r exposure-window}
mb |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    `Median AUCss (ug*h/mL)` = round(stats::median(auc_ss24), 1),
    `5th pctile` = round(stats::quantile(auc_ss24, 0.05), 1),
    `95th pctile` = round(stats::quantile(auc_ss24, 0.95), 1),
    `% with AUCss >= 400` = round(100 * mean(auc_ss24 >= 400), 1),
    `% with AUCss > 600` = round(100 * mean(auc_ss24 > 600), 1),
    .groups = "drop"
  ) |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(caption = "Simulated steady-state exposure against the Takada 2025 efficacy (AUCss/MIC >= 400 at MIC 1 ug/mL) and safety (AUCss > 600 ug*h/mL) thresholds.")
```

## The paper's own safety nomogram (Additional File 10: Table 8, file MOESM10)

Additional File 10 classifies each of the 70 nomogram cells by the probability
that the Table 4 dose produces an `AUCss` above 600 ug*h/mL: low risk (below
10%), moderate (10 to below 25%), or high (25% or above). Three cells are
flagged high-risk, all at the lowest creatinine clearances; the remaining 67 are
low-risk.

Two things are checked here. First, that the paper's Table 4 doses do keep
typical-value exposure bounded -- the deterministic side, which reproduces. Second,
that the *probabilities* do not follow from the reported between-subject
variance -- which they do not, and that failure is itself informative.

```{r safety-nomogram}
# Typical-value AUCss under the paper's own recommended dose, using the paper's
# own exp(0.11) clearance convention so the comparison stays inside its
# arithmetic.
auc_typ_grid <- published_dose / recomputed_paper

# Additional File 10 (Table 8) as printed: TRUE where the cell carries the
# high-risk symbol (probability of AUCss > 600 at or above 25%).
high_risk <- matrix(FALSE, nrow = length(alb_grid), ncol = length(crcl_grid),
                    dimnames = dimnames(published_cl))
high_risk[1, 2] <- TRUE  # Alb 1.5, CLcr 1.5 L/h
high_risk[2, 1] <- TRUE  # Alb 2.0, CLcr 1.2 L/h
high_risk[3, 1] <- TRUE  # Alb 2.5, CLcr 1.2 L/h

# Deterministic side: the recommended doses bound typical exposure, and the
# high-risk cells are the most-exposed corner of the grid.
rank_of_high <- rank(-as.vector(auc_typ_grid))[as.vector(high_risk)]
stopifnot(
  sum(high_risk) == 3L,
  all(auc_typ_grid <= 650),          # no cell's typical exposure runs away
  all(rank_of_high <= 5L)            # all three sit in the top 5 of 70 by AUCss
)

# Probability side: propagate the packaged IIV (log-normal CL, variance 0.11)
# analytically. AUCss = dose / CL and log(CL) is normal, so
# log(AUCss) ~ N(log(auc_typ), 0.11).
p_over_600 <- 1 - stats::pnorm(log(600 / auc_typ_grid) / sqrt(0.11))

cat(sprintf(
  "Typical-value AUCss across the 70 Table 4 cells: %.0f to %.0f ug*h/mL.\n",
  min(auc_typ_grid), max(auc_typ_grid)))
cat(sprintf(
  "High-risk cells rank %s of 70 by typical AUCss (1 = highest).\n",
  paste(sort(rank_of_high), collapse = ", ")))
cat(sprintf(
  "P(AUCss > 600) implied by the reported variance 0.11: %.0f%% to %.0f%%\n  across all 70 cells; Additional File 10 classifies %d of 70 as below 10%%.\n",
  100 * min(p_over_600), 100 * max(p_over_600), sum(!high_risk)))

# The implied risk exceeds 10% in EVERY cell, so the published classification
# cannot be recovered from the reported variance. Asserted so that the
# contradiction is a tested claim rather than a remark in prose.
stopifnot(all(p_over_600 > 0.10))
```

The deterministic bound holds: every Table 4 dose keeps typical exposure inside
the low-to-mid 600s at worst, and the three cells the paper flags as high-risk
are the three most-exposed cells on the grid. The probabilities, however, are
not reproducible -- a log-normal clearance random effect of variance 0.11 puts
*every* cell above the 10% risk band, while the paper places 67 of 70 below it.
That is the same inconsistency documented in the Errata below, seen from the
other side: the paper's Monte Carlo appears to have inflated the typical value
by `exp(0.11)` while propagating little or no clearance random effect. Its
Methods describe generating covariates from a normal distribution and computing
clearance from the final model, and never mention sampling `etaCL`.

# Assumptions and deviations

## Errata and internal inconsistencies in the source

- **Intercompartmental clearance: 4.86 versus 3.24 L/h.** Table 3 reports
  `Q = 4.86 L/h` with a standard error of 0.91, a coefficient of variation of
  18.63%, and a 95% confidence interval of 3.08-6.64; the Abstract repeats
  4.86. The Results narrative ("Development of a population pharmacokinetic
  model for VCM targeting older patients") instead states
  "Clearance between the central and peripheral compartments (Q) (L/h) = 3.24".
  The packaged model uses **4.86**: it is the value in the parameter table with
  its full uncertainty summary, it is corroborated by the Abstract, and 3.24 is
  not consistent with the tabulated confidence interval's centre. The narrative
  value appears to be a transcription error.

- **`etaCL = 0.11` is the between-subject variance, but the paper's nomograms
  use it as a fixed multiplier.** The Table 3 footnote defines eta as
  "normally distributed with mean 0 and variance omega^2, etaCL = 0.11", so
  0.11 is the variance of the clearance random effect (34.1% coefficient of
  variation on the exponential scale). The packaged model encodes it that way
  (`etalcl ~ 0.11`). The paper's own derived outputs -- the Discussion's
  2.73 / 3.29 L/h, Additional File 6: Table 5 (file MOESM7), and by extension Table 4 and
  Additional Files 7-10 -- instead substitute the constant 0.11 into
  `exp(etaCL)`, inflating typical clearance by `exp(0.11)` = 11.6%. That is
  arithmetically incompatible with a mean-zero random effect, and it is also
  incompatible with the paper reporting shrinkage, individual predictions, and
  conditional weighted residuals, all of which require an actual random effect.
  The parameter count corroborates the reading: the Table 2 base model has
  seven parameters, which is exactly four structural thetas plus one omega plus
  two residual-error parameters. Check 1 above quantifies the offset in all 70
  nomogram cells; it is exactly `exp(0.11)` everywhere, with no scatter.

- **The Table 8 risk classification does not follow from the reported
  variance.** Additional File 10 (Table 8) classifies 67 of the 70 nomogram
  cells as carrying below a 10% probability of `AUCss` exceeding 600 ug*h/mL.
  Propagating the reported between-subject variance of 0.11 through
  `AUCss = dose / CL` puts every one of those cells above 10% (16% to 58%
  across the grid), so the published probabilities cannot be recovered from the
  published variability. This is the `exp(0.11)` problem seen from the other
  side, and it is consistent with the Methods, which describe generating
  covariates from a normal distribution and computing clearance from the final
  model but never mention sampling `etaCL`. The deterministic content of
  Table 8 does reproduce: the three high-risk cells are the three cells with the
  highest typical-value exposure. Quantified in the safety-nomogram check above.

- **The "41.5%" figure is not interpretable as printed.** The Results state that
  "an AUCss of > 600 ug*h/mL was observed in 41.5% of some CLcr and Alb levels
  (Additional File 10: Table 8) (Table 4)". Table 8 contains no percentages --
  it is a 70-cell grid of three risk symbols -- and 41.5% is not the fraction of
  flagged cells either (3 of 70, or 4.3%). The figure is most consistent with a
  probability reached in a single worst-case cell, but the paper does not print
  the underlying numbers, so nothing in this vignette depends on it.

- **Median creatinine clearance: 3.06 versus 3.09 L/h.** The Results narrative
  quotes a modeling-cohort median CLcr of 3.06 L/h [51.0 mL/min] while Table 1
  reports 51.5 mL/min, and the model equation normalises to 3.09 L/h
  (= 51.5 mL/min). The packaged model uses **3.09 L/h**, the value that appears
  in the model equation itself and matches Table 1.

## Parameters the paper does not publish

- **Residual error.** Takada 2025 evaluated additive, multiplicative, and
  combined additive-plus-multiplicative residual models (Methods; Additional
  File 1: Table 1, file MOESM3, which prints the three Phoenix code forms) but reports
  neither the selected model nor any residual-error estimate anywhere in the
  article or its ten supplementary files. Table 3 lists only the six structural
  thetas. Both `propSd` and `addSd` are therefore encoded as `fixed(0)` rather
  than invented. The seven-parameter base model of Table 2 implies the
  two-parameter combined form was used; the Phoenix combined model
  `Cobs = C + eps * sqrt[1 + C^2 * (Cmultstdev/sigma)^2]` is algebraically
  identical to nlmixr2's `add(sigma) + prop(Cmultstdev)`, so the packaged error
  structure matches the source's form even though the magnitudes are unknown.
  Consequence: simulations from this model carry between-subject variability
  but no residual error, so the percentile bands in the Figure 1 panel are
  narrower than the paper's visual predictive check.

- **Between-subject variability on Q, Vc, and Vp.** None is reported; the
  Results state that Vc "was a fixed value". No random effect is included on
  those parameters.

## Simulation assumptions

- **Infusion duration** is not stated by the source; one hour is assumed. The
  steady-state exposure checks are unaffected -- `AUCss` depends only on dose
  and clearance -- but the peak concentrations in the Figure 1 panel do depend
  on it.
- **Virtual-cohort covariate distributions.** Age, sex, body weight, serum
  creatinine, and serum albumin are drawn from log-normal distributions
  truncated to the Table 1 ranges and centred on the Table 1 medians;
  correlations among them (which the paper does not report, beyond a Pearson
  correlation of -0.12 between albumin and creatinine clearance) are ignored.
  Creatinine clearance is derived from the sampled demographics using the
  paper's Cockcroft-Gault equation rather than sampled, and the resulting
  distribution is checked against Table 1 in the cohort section.
- **Simulated regimens** (500 / 750 / 1000 mg q12h) are representative points
  inside the published dose range; they are not a regimen the paper reports
  results for individually.
- **The Table 4 reproduction is a bound, not an exact replication.** The
  paper's probability-of-target-attainment simulation resamples covariates and
  MIC from published distributions using assumptions it does not fully specify.
  What is reproduced here is the deterministic floor that any such simulation
  must respect, plus the finding that the published doses never exceed it by
  more than one 250 mg step.
