---
title: "Roxadustat (Takada 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Roxadustat (Takada 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Takada_2022_roxadustat"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Takada_2022_roxadustat"))$description`
* Article: <https://doi.org/10.1111/bcp.15023> (open access; PMC9292185)

# Population

Takada et al. pooled 367 Japanese patients (256 male, 111 female) with
anaemia of dialysis-dependent chronic kidney disease from four Astellas
studies of oral roxadustat given three times weekly (Table 1): the phase II
haemodialysis study 1517-CL-0304 (n = 92, after an ESA washout) and three
phase III studies, 1517-CL-0308 (haemodialysis, ESA-untreated, n = 74),
1517-CL-0307 (haemodialysis, conversion from ESA, n = 145) and 1517-CL-0302
(peritoneal dialysis, n = 56). Mean (SD) age was 64.4 (10.9) years, weight
59.5 (11.8) kg and BMI 22.8 (3.5) kg/m^2 (Table 3). Initial doses of 50, 70 or
100 mg were titrated on haemoglobin to 20-250 mg. Sparse samples (1285
concentrations) were taken at study visits at any time relative to dose and
dialysis.

Phosphate binders were common: 82.0% of patients took at least one during the
study (Table 3): calcium carbonate 47.7%, lanthanum carbonate 37.9%,
sevelamer hydrochloride or bixalomer 21.5%, ferric citrate 16.1% and
sucroferric oxyhydroxide 4.6%. The phase III protocols required roxadustat to
be taken at least 1 hour before or after a phosphate binder; the phase II study
did not.

The same information is available programmatically via
`readModelDb("Takada_2022_roxadustat")()$population`.

# Source trace

Every `ini()` value in `inst/modeldb/specificDrugs/Takada_2022_roxadustat.R`
carries an in-file comment pointing at its source. All final estimates come
from Table 4 of the article.

| Component | Value | Source |
|-----------|-------|--------|
| Two-compartment disposition, first-order absorption with lag time | -- | Results 3.2; Conclusion |
| Log-normal IIV, `P_i = theta * exp(eta_i)` | -- | Methods 2.4 equation |
| Categorical covariate model `P_i = theta1 * theta2^(0 or 1)` | -- | Methods 2.5 equation |
| CL/F | 0.923 L/h | Table 4 |
| Vc/F | 14.6 L | Table 4 |
| ka | 0.63 1/h | Table 4 |
| Q/F | 0.134 L/h | Table 4 |
| Vp/F | 2.89 L | Table 4 |
| ALAG1 | 0.287 h | Table 4 |
| Age >= 65 y on CL/F | 0.792 | Table 4; Results 3.4 |
| Sevelamer/bixalomer (SBUSE) on F1 | 0.744 | Table 4 |
| Calcium carbonate (CUSE) on F1 | 0.931 | Table 4 |
| Lanthanum carbonate (LUSE) on F1 | 0.969 | Table 4 |
| Ferric citrate (FUSE) on F1 | 0.744 | Table 4 |
| Sucroferric oxyhydroxide (SCUSE) on F1 | 0.837 | Table 4 |
| No time separation of PBs in phase II on F1 | 0.935 | Table 4 and its footnote |
| IIV CL/F | 41.7% -> omega^2 = 0.1739 | Table 4 (see "IIV scale" below) |
| IIV Vc/F | 22.0% -> omega^2 = 0.0484 | Table 4 |
| IIV ka | 184.1% -> omega^2 = 3.389 | Table 4 |
| Combined residual error `W = sqrt((IPRED * theta_prop)^2 + theta_add^2)` | -- | Methods 2.4 equation |
| Proportional residual SD | 0.439 | Table 4 |
| Additive residual SD | 1.88 ng/mL | Table 4 |
| Dose mg, V in L -> `Cc = 1000 * central / vc` in ng/mL | -- | Methods 2.2 (ng/mL assay) |

## IIV scale

Table 4 reports IIV as "%CV" without saying how it was computed from
`omega^2`. The printed 95% confidence intervals settle it: they are
`omega^2 +/- 1.96 * SE` (with SE = RSE x `omega^2`) back-transformed by a
square root, so the reported percentage is `100 * sqrt(omega^2)`. The two
readings are nearly indistinguishable for the moderate CL/F and Vc/F rows, but
the wide ka row separates them: only the `sqrt(omega^2)` reading reproduces its
154.2-209.8% interval.

```{r iiv-scale}
iiv <- data.frame(
  parameter = c("CL/F", "Vc/F", "ka"),
  cv = c(0.417, 0.220, 1.841),
  rse = c(0.096, 0.503, 0.152),
  lo_pub = c(0.376, 0.026, 1.542),
  hi_pub = c(0.455, 0.310, 2.098)
)
iiv <- iiv |>
  dplyr::mutate(
    # Reading A: %CV = sqrt(omega^2)
    om2_a = cv^2,
    lo_a = sqrt(om2_a * (1 - 1.96 * rse)),
    hi_a = sqrt(om2_a * (1 + 1.96 * rse)),
    # Reading B: %CV = sqrt(exp(omega^2) - 1)
    om2_b = log(1 + cv^2),
    lo_b = sqrt(exp(om2_b * (1 - 1.96 * rse)) - 1),
    hi_b = sqrt(exp(om2_b * (1 + 1.96 * rse)) - 1)
  )
iiv |>
  dplyr::transmute(
    Parameter = parameter,
    "Published 95% CI" = sprintf("%.1f-%.1f%%", 100 * lo_pub, 100 * hi_pub),
    "sqrt(omega^2) reading" = sprintf("%.1f-%.1f%%", 100 * lo_a, 100 * hi_a),
    "log(1 + CV^2) reading" = sprintf("%.1f-%.1f%%", 100 * lo_b, 100 * hi_b)
  ) |>
  knitr::kable()

# The sqrt(omega^2) reading reproduces every published bound to within the
# table's rounding; the log(1 + CV^2) reading misses the wide ka row by >15
# percentage points.
stopifnot(
  all(abs(iiv$lo_a - iiv$lo_pub) < 0.002),
  all(abs(iiv$hi_a - iiv$hi_pub) < 0.002),
  abs(iiv$hi_b[3] - iiv$hi_pub[3]) > 0.15
)
```

# Covariate effects on exposure (Figure 4)

The paper summarises each covariate as the percentage change in roxadustat
AUCinf, where `AUCinf = Dose / (CL/F)` with CL/F adjusted for the covariates
(Methods 2.7). Because each phosphate binder acts on F1, the AUCinf change is
`F1 factor - 1`; age acts on CL/F, so AUCinf changes by `1/0.792 - 1 = +26.3%`
(Results 3.4). The check below solves the packaged model for a typical patient
(no IIV) after a single 100 mg dose, computes AUCinf with PKNCA for each
covariate setting, and compares the ratio with Figure 4.

```{r fig4-events}
mod <- rxode2::rxode2(readModelDb("Takada_2022_roxadustat"))
mod_typ <- rxode2::zeroRe(mod)

no_cov <- data.frame(
  AGE = 60,
  CONMED_SEVELAMER_BIXALOMER = 0,
  CONMED_CALCIUM_CARBONATE = 0,
  CONMED_LANTHANUM_CARBONATE = 0,
  CONMED_FERRIC_CITRATE = 0,
  CONMED_SUCROFERRIC_OXYHYDROXIDE = 0,
  STUDY_PHASE2 = 0
)

scenarios <- list(
  "Reference" = no_cov,
  "Age >= 65 y" = dplyr::mutate(no_cov, AGE = 70),
  "Sevelamer/bixalomer" = dplyr::mutate(no_cov, CONMED_SEVELAMER_BIXALOMER = 1),
  "Sucroferric oxyhydroxide" = dplyr::mutate(no_cov, CONMED_SUCROFERRIC_OXYHYDROXIDE = 1),
  "Ferric citrate" = dplyr::mutate(no_cov, CONMED_FERRIC_CITRATE = 1),
  "Calcium carbonate" = dplyr::mutate(no_cov, CONMED_CALCIUM_CARBONATE = 1),
  "Lanthanum carbonate" = dplyr::mutate(no_cov, CONMED_LANTHANUM_CARBONATE = 1),
  # The phase II factor applies only on top of a coadministered binder;
  # compare calcium carbonate without vs with time separation.
  "Calcium carbonate, no time separation" =
    dplyr::mutate(no_cov, CONMED_CALCIUM_CARBONATE = 1, STUDY_PHASE2 = 1),
  # STUDY_PHASE2 alone (no binder) must not change exposure.
  "Phase II, no binder" = dplyr::mutate(no_cov, STUDY_PHASE2 = 1)
)

obs_times <- c(0, seq(0.25, 12, by = 0.25), seq(13, 48, by = 1), seq(54, 480, by = 6))
fig4_events <- dplyr::bind_rows(lapply(seq_along(scenarios), function(i) {
  covs <- scenarios[[i]]
  dplyr::bind_rows(
    data.frame(id = i, time = 0, evid = 1L, amt = 100, cmt = "depot"),
    data.frame(id = i, time = obs_times, evid = 0L, amt = 0, cmt = "central")
  ) |>
    dplyr::mutate(scenario = names(scenarios)[i]) |>
    cbind(covs)
}))
```

```{r fig4-solve}
fig4_sim <- rxode2::rxSolve(
  mod_typ,
  events = fig4_events,
  keep = "scenario",
  addDosing = FALSE
) |>
  as.data.frame()

fig4_conc <- fig4_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, scenario)
fig4_dose <- fig4_events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, scenario)

fig4_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(fig4_conc, Cc ~ time | scenario + id),
  PKNCA::PKNCAdose(fig4_dose, amt ~ time | scenario + id),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, cmax = TRUE)
))
fig4_res <- as.data.frame(fig4_nca) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(scenario, aucinf = PPORRES)

auc_ref <- fig4_res$aucinf[fig4_res$scenario == "Reference"]
auc_ca <- fig4_res$aucinf[fig4_res$scenario == "Calcium carbonate"]

fig4_pub <- data.frame(
  scenario = c(
    "Age >= 65 y", "Sevelamer/bixalomer", "Sucroferric oxyhydroxide",
    "Ferric citrate", "Calcium carbonate", "Lanthanum carbonate",
    "Calcium carbonate, no time separation", "Phase II, no binder"
  ),
  # Figure 4 prints -20.8% for age, which is the change in CL/F (0.792 - 1);
  # the corresponding AUCinf change is the +26.3% stated in Results 3.4.
  published_pct = c(26.3, -25.6, -16.3, -25.6, -6.9, -3.1, -6.5, 0),
  published_source = c(
    "Results 3.4 (+26.3%)", "Figure 4", "Figure 4", "Figure 4", "Figure 4",
    "Figure 4", "Figure 4 ('No time of separation of PBs in Phase II')",
    "Table 4 footnote (factor acts on binder intake only)"
  )
)

fig4_cmp <- fig4_pub |>
  dplyr::left_join(fig4_res, by = "scenario") |>
  dplyr::mutate(
    comparator = ifelse(
      scenario == "Calcium carbonate, no time separation", auc_ca, auc_ref
    ),
    simulated_pct = 100 * (aucinf / comparator - 1)
  )

fig4_cmp |>
  dplyr::transmute(
    Covariate = scenario,
    "Simulated AUCinf change (%)" = round(simulated_pct, 1),
    "Published (%)" = published_pct,
    Source = published_source
  ) |>
  knitr::kable()

# Deterministic typical-value solves: the AUCinf ratio is exact up to the
# PKNCA extrapolation of the terminal phase, so a 0.3 percentage-point
# tolerance covers the published one-decimal rounding.
stopifnot(all(abs(fig4_cmp$simulated_pct - fig4_cmp$published_pct) < 0.3))
```

The closed form `AUCinf = F1 * Dose / (CL/F)` gives the same reference value:

```{r closed-form}
auc_closed <- 1 * 100 / 0.923 * 1000 # mg / (L/h) -> mg*h/L; x 1000 -> ng*h/mL
c(PKNCA = auc_ref, closed_form = auc_closed)
stopifnot(abs(auc_ref / auc_closed - 1) < 0.01)
```

# Virtual cohort

The observed data are not public. The cohort below approximates the phase III
population: three arms at the phase III starting doses of 50, 70 and 100 mg
three times weekly (Monday / Wednesday / Friday, i.e. doses at 0, 48 and 96 h
of each week) for four weeks, 100 patients per arm. Age is drawn from a normal
distribution with the pooled mean and SD (Table 3) truncated to 30-90 years.
Each phosphate binder is assigned per patient, held constant over the four
weeks, at its phase III pooled frequency (Table 3), with time separation
(`STUDY_PHASE2 = 0`). No dose titration is simulated.

```{r cohort}
set.seed(20220702)
rxode2::rxSetSeed(20220702)
n_per_arm <- 100

dose_times <- as.vector(outer(c(0, 48, 96), 168 * (0:3), "+"))
obs_times <- sort(unique(c(seq(0, 168 * 4, by = 2), 504 + c(0.5, 1, 1.5, 2.5, 3, 4, 5, 6, 8, 10))))

make_arm <- function(dose_mg, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  covs <- data.frame(
    id = ids,
    AGE = pmin(pmax(stats::rnorm(n_per_arm, 64.4, 10.9), 30), 90),
    CONMED_SEVELAMER_BIXALOMER = stats::rbinom(n_per_arm, 1, 0.164),
    CONMED_CALCIUM_CARBONATE = stats::rbinom(n_per_arm, 1, 0.458),
    CONMED_LANTHANUM_CARBONATE = stats::rbinom(n_per_arm, 1, 0.367),
    CONMED_FERRIC_CITRATE = stats::rbinom(n_per_arm, 1, 0.207),
    CONMED_SUCROFERRIC_OXYHYDROXIDE = stats::rbinom(n_per_arm, 1, 0.062),
    STUDY_PHASE2 = 0
  )
  doses <- tidyr::crossing(id = ids, time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = dose_mg, cmt = "depot")
  obs <- tidyr::crossing(id = ids, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::left_join(covs, by = "id") |>
    dplyr::mutate(treatment = paste(dose_mg, "mg TIW")) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_arm(50, 0L),
  make_arm(70, 1000L),
  make_arm(100, 2000L)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = "treatment",
  addDosing = FALSE
) |>
  as.data.frame()
```

# Concentration versus time after dose (Figure 3)

Figure 3 of the paper is a prediction-corrected VPC of plasma roxadustat versus
time after the last dose, with most samples at 48-72 hours because sampling
followed the dialysis schedule. The plot below shows the simulated individual
concentrations (including residual error) against time after the most recent
dose during week 4, with the 5th, 50th and 95th percentiles. It is not
prediction-corrected, so it is a visual analogue rather than a reproduction.

```{r fig3}
week4 <- sim |>
  dplyr::filter(time >= 504, time <= 672, !is.na(sim)) |>
  dplyr::mutate(
    last_dose = 504 + c(0, 48, 96)[findInterval(time - 504, c(0, 48, 96))],
    tad = time - last_dose,
    tad_bin = cut(tad, breaks = c(-0.01, 2, 6, 12, 24, 36, 48, 60, 72))
  )

vpc_stats <- week4 |>
  dplyr::group_by(treatment, tad_bin) |>
  dplyr::summarise(
    tad = median(tad),
    p05 = quantile(sim, 0.05),
    p50 = median(sim),
    p95 = quantile(sim, 0.95),
    .groups = "drop"
  )

ggplot(week4, aes(tad, pmax(sim, 1))) +
  geom_point(alpha = 0.05, size = 0.6, colour = "steelblue") +
  geom_line(data = vpc_stats, aes(tad, p50), colour = "red", linewidth = 1) +
  geom_line(data = vpc_stats, aes(tad, p05), colour = "red", linetype = "dashed") +
  geom_line(data = vpc_stats, aes(tad, p95), colour = "red", linetype = "dashed") +
  scale_y_log10() +
  facet_wrap(~treatment) +
  labs(
    x = "Time after last dose (h)",
    y = "Roxadustat plasma concentration (ng/mL)",
    caption = "Simulated week-4 concentrations; red lines: 5th, 50th, 95th percentiles. Visual analogue of Figure 3 of Takada 2022."
  ) +
  theme_bw()
```

# Steady-state NCA (PKNCA)

The paper publishes no NCA table, so the PKNCA block below characterises the
week-4 Monday-to-Wednesday dosing interval (504-552 h) by arm. The structural
checks are that exposure is dose proportional (the model is linear) and that
the median AUC over the 48-hour interval matches the closed form
`F1 * Dose / (CL/F)` for the cohort's own individual parameters, because with
a 48-hour interval nearly the whole dose is eliminated before the next one.

```{r pknca}
conc_ss <- sim |>
  dplyr::filter(time >= 504, time <= 552, !is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
dose_ss <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_ss, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_ss, amt ~ time | treatment + id),
  intervals = data.frame(
    start = 504, end = 552,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, cmin = TRUE
  )
))

nca_tbl <- as.data.frame(nca_ss) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = signif(median(PPORRES, na.rm = TRUE), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::arrange(as.numeric(sub(" mg TIW", "", treatment)))

nca_tbl |>
  dplyr::rename(
    Arm = treatment,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-48 (ng*h/mL)" = auclast,
    "Cmin (ng/mL)" = cmin
  ) |>
  knitr::kable(caption = "Median week-4 NCA over the 48-hour interval (504-552 h).")

auc_by_id <- as.data.frame(nca_ss) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, treatment, auclast = PPORRES)
indiv <- sim |>
  dplyr::filter(time == 504) |>
  dplyr::distinct(id, cl, fdepot)
auc_chk <- auc_by_id |>
  dplyr::left_join(indiv, by = "id") |>
  dplyr::mutate(
    dose = as.numeric(sub(" mg TIW", "", treatment)),
    auc_closed = fdepot * dose / cl * 1000,
    pct_diff = 100 * (auclast / auc_closed - 1),
    auc_per_mg = auclast / dose
  )

auc_dn <- auc_chk |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(median_auc_per_mg = median(auc_per_mg), .groups = "drop")
auc_dn

stopifnot(
  # Centre of the interval-AUC vs closed-form distribution. A wrong unit,
  # dose or clearance would move this by tens of percent.
  # No tail-quantile gate: with a 184% CV on ka, roughly 8% of patients
  # absorb so slowly that much of a dose spills past the 48-hour interval.
  abs(median(auc_chk$pct_diff)) < 10,
  # Dose-normalised exposure is similar across arms (independent cohorts, so
  # only sampling noise separates them).
  max(auc_dn$median_auc_per_mg) / min(auc_dn$median_auc_per_mg) < 1.5
)
```

# Assumptions and deviations

* **IIV scale.** Table 4 labels IIV as %CV. The printed confidence intervals
  are reproduced only when the percentage is `100 * sqrt(omega^2)`, so the
  model uses `omega^2 = (CV/100)^2` (0.1739, 0.0484, 3.389 for CL/F, Vc/F and
  ka). The very large ka variance (184% CV) is as published; the authors
  attribute it to the few absorption-phase samples.
* **Residual error.** Methods 2.4 defines `W = sqrt((IPRED * theta_prop)^2 +
  theta_add^2)` with `EPS` fixed to variance 1, so `propSd = 0.439` and
  `addSd = 1.88` ng/mL are standard deviations, combined as in nlmixr2's
  default `prop() + add()`.
* **Phase II "no time separation" factor.** Table 4 lists a separate 0.935
  factor on F1 for phosphate binders taken without time separation in the
  phase II study; the Table 4 footnote calls it "the additional effect of the
  phosphate binders due to the uncontrolled intake". The paper does not print
  the NONMEM code, so the model applies the factor once, when `STUDY_PHASE2 = 1`
  and at least one of the five binders is coadministered. This matches the
  single -6.5% AUCinf change in Figure 4. Whether the original code applied it
  once per binder (for patients taking several) is not stated.
* **Multiple binders.** The categorical model `theta1 * theta2^(0 or 1)` is
  applied once per binder, so the F1 factors multiply when a patient takes more
  than one binder. The paper does not show a combination, but this follows
  from the Methods 2.5 equation.
* **Age.** Only the categorical age >= 65 years effect is retained (a patient
  aged exactly 65 is in the >= 65 group). The model takes continuous `AGE` and
  applies the threshold internally.
* **Figure 4 age row.** Figure 4 is labelled "Change of roxadustat AUCinf" but
  prints -20.8% for age >= 65 years, which is the change in CL/F. Results 3.4
  states that AUCinf increases by 26.3%, which the model reproduces.
* **Time-varying concomitant medications.** The source built the binder
  indicators from daily administration records; the virtual cohort holds each
  patient's binders constant for simplicity. The indicators may vary over
  time in a user's data set.
* **Table 3 versus text.** The Results text gives 131 patients on lanthanum
  carbonate; Table 3 gives 139. The cohort uses the Table 3 phase III
  frequencies.
* **Screened, not retained.** Sex, body weight, BMI, serum albumin, AST, ALT,
  ALP, total protein, dialysis modality (haemodialysis versus peritoneal
  dialysis), clopidogrel, dose, oral iron and multivalent cation-containing
  drugs were tested and not retained (Results 3.2-3.3). Covariates are
  recorded in the model's `covariatesDataExcluded` metadata.
* **Errata.** No erratum or correction was found for this article (Europe PMC
  and the journal landing page, checked 2026-10-02).
