---
title: "Nemolizumab in prurigo nodularis (Takechi 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nemolizumab in prurigo nodularis (Takechi 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  error      = FALSE
)
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20250929)
```

# Model and source

* Citation: Takechi T, Shimizu J, Kabashima K, Ieiri I (2025). *Quantitative
  evaluation of nemolizumab pharmacokinetics and efficacy in prurigo nodularis:
  a population pharmacokinetics and model-based meta-analysis approach.*
  Dermatol Ther (Heidelb) 15(12):3615-3632.
* Article: <https://doi.org/10.1007/s13555-025-01554-4>
* Supplement (Figs. S1-S10, Table S1) retrieved from the EuropePMC
  supplementary-file endpoint for PMC12619860. It contains figures and the
  study-design table only -- **no NONMEM control stream and no additional
  parameter table**.

Nemolizumab is a humanized monoclonal antibody against interleukin-31 receptor
A (IL-31RA). IL-31 signalling on sensory neurons and immune cells is a
principal driver of itch, and nemolizumab is approved for prurigo nodularis
(PN) and atopic dermatitis (AD). Takechi 2025 is a three-part pharmacometric
analysis -- and the meta-analysis part fits two endpoints -- so the paper yields
four models, all validated here:

| Part | nlmixr2lib model | What it describes |
| --- | --- | --- |
| PopPK | `Takechi_2025_nemolizumab` | Serum nemolizumab concentration, one compartment, first-order SC absorption |
| PopPD | `Takechi_2025_nemolizumab_ppnrs` | Weekly average Peak Pruritus NRS (PP-NRS, 0-10) under placebo and nemolizumab |
| MBMA | `Takechi_2025_nemolizumab_mbma_iga` | Study-arm IGA success rate, nemolizumab vs dupilumab, 6 RCTs |
| MBMA | `Takechi_2025_nemolizumab_mbma_ppnrs` | Study-arm PP-NRS success rate over the same 6 RCTs, with an arm-severity covariate |

The four are **independent models fitted to different data**, exactly as the
authors built them, so they are four model files sharing this one vignette.
The PopPD model has no PK layer: the authors looked for an exposure-response
relationship and did not find one, because the 30 mg and 60 mg arms of the
pivotal study responded alike.

The two meta-analysis models share a structure but not a covariate model: the
PP-NRS endpoint retains the arm's moderate-IGA proportion on the nemolizumab
maximum effect, and the IGA endpoint does not (Table 4 prints a dash in that
row of the IGA column). Part 4 sets out how that covariate's functional form was
recovered, since the paper does not print it.

```{r load-models}
pk         <- rxode2::rxode(readModelDb("Takechi_2025_nemolizumab"))
pd         <- rxode2::rxode(readModelDb("Takechi_2025_nemolizumab_ppnrs"))
mbma       <- rxode2::rxode(readModelDb("Takechi_2025_nemolizumab_mbma_iga"))
mbma_ppnrs <- rxode2::rxode(readModelDb("Takechi_2025_nemolizumab_mbma_ppnrs"))
```

# Population

**PopPK** (Table 1, "All" column; Supplementary Table S1). 680 patients
contributing 4389 nemolizumab concentrations across seven studies -- six in AD
(including two paediatric studies in children aged 6-12 years) and one in PN.
Median age 37 years (range 6-84), median weight 62.8 kg (range 16-151), median
serum albumin 4.4 g/dL (range 2.0-5.2), 386 male / 294 female. Race, summed
across the per-study counts of Table 1, is 125 White / 18 Black or African
American / 535 Asian / 2 Other, which totals exactly the 680 stated subjects;
all non-Asian subjects come from the single multinational phase 2a study
CIM003JG. Placebo subjects were excluded from the PopPK dataset (Table S1
footnote a). Doses ranged from 0.003 mg/kg single doses in phase 1 up to 60 mg
Q4W.

The 153 PN patients (905 observations) were held out as an **external
validation** set rather than used for estimation, which is the basis of the
paper's conclusion that AD and PN share the same PK.

**PopPD** (Table 1, "Patients with PN" column, footnote-b rows). All 229
patients of the phase II/III study M525101-11 (jRCT2011200017, Yokozeki et al.
2024), contributing 3833 weekly-average PP-NRS observations over 112 days.
Median age 51 years (range 13-84), median weight 61.2 kg (range 32.8-109.6),
107 male / 122 female, all Japanese. Baseline weekly average PP-NRS median 8.6
(range 6.4-10); baseline IGA 3 in 112 patients and 4 in 117. Arms were placebo
(76), nemolizumab 30 mg Q4W after a 60 mg loading dose (77) and nemolizumab
60 mg Q4W (76).

**MBMA** (Table 3). 1170 participants in 13 arms of six randomized controlled
trials: Yokozeki et al. (jRCT2011200017), Staender et al. (SPR.115828),
OLYMPIA 1, OLYMPIA 2 for nemolizumab, and PRIME and PRIME2 for dupilumab.
Arm-mean age 46.7-59.7 years, arm-mean weight 60.4-87.1 kg, baseline PP-NRS
8.3-8.6, proportion with IGA 4 at baseline 28.0-54.5%.

# Source trace

Every value in the four `ini()` blocks, with the source location. Values are
transcribed from the "Original data / Estimate" columns; the bootstrap medians
and 95% CIs in the adjacent columns were used only as a transcription check.

```{r source-trace, echo = FALSE}
trace <- tibble::tribble(
  ~Model,  ~Parameter,             ~Value,          ~Source,
  "PopPK", "lcl",                  "log(0.340)",    "Table 2, CL/F (L/day); also the displayed CL/F equation",
  "PopPK", "lvc",                  "log(8.44)",     "Table 2, V/F (L); also the displayed V/F equation",
  "PopPK", "lka",                  "log(0.548)",    "Table 2, ka (1/day)",
  "PopPK", "e_wt_cl",              "fixed(0.75)",   "Displayed CL/F equation exponent (no RSE reported)",
  "PopPK", "e_wt_vc",              "fixed(1)",      "Displayed V/F equation, printed without an exponent",
  "PopPK", "e_alb_cl",             "-1.52",         "Table 2, Covariate effect of ALB",
  "PopPK", "etalcl + etalvc",      "0.135 / 0.0896 / 0.148", "Table 2, IIV CL/F, Covariance, IIV V/F",
  "PopPK", "etalka",               "0.404",         "Table 2, IIV ka",
  "PopPK", "expSd",                "0.164866",      "Table 2, log normal error (CV%) = 16.6; sqrt(log(1 + 0.166^2))",
  "PopPD", "lpmax",                "0.618",         "Table 2, Pmax (log scale; see Assumptions)",
  "PopPD", "lkp",                  "log(2.74e-3)",  "Table 2, Kp (x 10^-3/day)",
  "PopPD", "lemax",                "1.41",          "Table 2, Edrug (log scale; see Assumptions)",
  "PopPD", "lkdrug",               "log(28.7e-3)",  "Table 2, Kd (x 10^-3/day)",
  "PopPD", "etalpmax / etalkp / etalemax / etalkdrug", "1.55 / 10.4 / 1.19 / 2.10", "Table 2, IIV Pmax / Kp / Edrug / Kd",
  "PopPD", "addSd",                "sqrt(0.582)",   "Table 2, Additive error (variance; see Assumptions)",
  "MBMA IGA",  "e0",                 "-4.92",         "Table 4, IGA column, row A",
  "MBMA IGA",  "pmax_placebo",       "2.99",          "Table 4, IGA column, row Pmax",
  "MBMA IGA",  "lkp",                "log(0.123)",    "Table 4, IGA column, row Kp (/week)",
  "MBMA IGA",  "edrug_nemolizumab",  "1.71",          "Table 4, IGA column, row Edrug,nemolizumab",
  "MBMA IGA",  "lkdrug_nemolizumab", "log(0.416)",    "Table 4, IGA column, row Kd,nemolizumab (/week)",
  "MBMA IGA",  "edrug_dupilumab",    "1.49",          "Table 4, IGA column, row Edrug,dupilumab",
  "MBMA IGA",  "lkdrug_dupilumab",   "log(0.134)",    "Table 4, IGA column, row Kd,dupilumab (/week)",
  "MBMA IGA",  "eta_study_eff",      "0.226",         "Table 4, IGA column, row ISV on EFF (variance)",
  "MBMA IGA",  "addSd",              "sqrt(0.619)",   "Table 4, IGA column, row Additive error (variance)",
  "MBMA PP-NRS", "e0",                         "-4.38",        "Table 4, PP-NRS column, row A",
  "MBMA PP-NRS", "pmax_placebo",               "2.64",         "Table 4, PP-NRS column, row Pmax",
  "MBMA PP-NRS", "lkp",                        "log(0.277)",   "Table 4, PP-NRS column, row Kp (/week)",
  "MBMA PP-NRS", "edrug_nemolizumab",          "1.63",         "Table 4, PP-NRS column, row Edrug,nemolizumab",
  "MBMA PP-NRS", "e_igamod_edrug_nemolizumab", "-1.84",        "Table 4, PP-NRS column, row PPM effect on Edrug,nemolizumab (coefficient printed; FORM reconstructed from Fig. 4 -- see Part 4)",
  "MBMA PP-NRS", "edrug_dupilumab",            "1.93",         "Table 4, PP-NRS column, row Edrug,dupilumab",
  "MBMA PP-NRS", "lkdrug_nemolizumab",         "log(1.93)",    "Table 4, PP-NRS column, row Kd,nemolizumab (/week)",
  "MBMA PP-NRS", "lkdrug_dupilumab",           "log(0.119)",   "Table 4, PP-NRS column, row Kd,dupilumab (/week)",
  "MBMA PP-NRS", "eta_study_eff",              "0.105",        "Table 4, PP-NRS column, row ISV on EFF (variance)",
  "MBMA PP-NRS", "addSd",                      "sqrt(0.738)",  "Table 4, PP-NRS column, row Additive error (variance)",
  "MBMA PP-NRS", "IGA_MOD_PCT centring 59.0",  "59.0",         "Table 3, median over the 13 arms of (100 - 'IGA4 at baseline (%)')"
)
knitr::kable(trace, caption = "Source trace for all four model files.")
```

Structural equations: the PopPK model structure is Results "PopPK Analysis"
paragraph 1 plus the two displayed fixed-effect equations; the PopPD equations
are the two displayed equations in Methods "Pharmacodynamic Model"; the MBMA
equations are the displayed logit / EFF / f0 / fdrug group and the displayed
residual equation in Methods "Model-Based Meta-analysis", shared by both
endpoints. The one structural element with **no** printed source is the
functional form attaching `PPM` to `Edrug,nemolizumab` in the PP-NRS
meta-analysis; Part 4 derives it from the paper's own Fig. 4 and shows that the
two competing readings are falsified.

## Reading the omega scale from the covariance row

Table 2 prints the IIV rows as bare numbers with no CV% label, so whether they
are variances or standard deviations has to be settled from the table itself.
The `Covariance for CL/F and V/F` row settles it: only one reading gives a
legal correlation.

```{r omega-scale}
corr_if_variance <- 0.0896 / sqrt(0.135 * 0.148)
corr_if_sd       <- 0.0896 / sqrt(0.135^2 * 0.148^2)
c(`as variances` = corr_if_variance, `as SDs` = corr_if_sd)

# A correlation must lie in [-1, 1]. The SD reading is impossible.
stopifnot(
  abs(corr_if_variance) < 1,
  abs(corr_if_sd) > 1
)
```

The IIV rows are therefore **variances**, and by the same table's convention so
are the `ISV on EFF` and `Additive error` rows of Table 4.

# Part 1 -- Population pharmacokinetics

## Typical-value structure

With the etas zeroed, the model must reproduce the paper's two displayed
equations exactly for any weight and albumin.

```{r pk-typical}
pk0 <- rxode2::zeroRe(pk)

pk_typ <- function(wt, alb_gL, dose = 60, tmax = 112, dt = 0.25) {
  ev <- rxode2::et(amt = dose, cmt = "depot") |>
    rxode2::et(seq(0, tmax, by = dt)) |>
    as.data.frame()
  ev$id  <- 1L
  ev$WT  <- wt
  ev$ALB <- alb_gL
  as.data.frame(rxode2::rxSolve(pk0, ev))
}

s70 <- pk_typ(70, 45)   # 70 kg, albumin 4.5 g/dL = 45 g/L -> both covariates at reference
s40 <- pk_typ(40, 30)   # 40 kg, albumin 3.0 g/dL = 30 g/L

# Paper's equations:
#   CL/F = 0.340 * (WT/70)^0.75 * (ALB_gdL/4.5)^-1.52
#   V/F  = 8.44  * (WT/70)
cl_paper <- function(wt, alb_gL) 0.340 * (wt / 70)^0.75 * ((alb_gL * 0.1) / 4.5)^-1.52
vc_paper <- function(wt)         8.44  * (wt / 70)

chk <- tibble::tibble(
  quantity = c("CL/F at 70 kg / 45 g/L", "V/F at 70 kg", "CL/F at 40 kg / 30 g/L", "V/F at 40 kg"),
  model    = c(s70$cl[1], s70$vc[1], s40$cl[1], s40$vc[1]),
  paper    = c(cl_paper(70, 45), vc_paper(70), cl_paper(40, 30), vc_paper(40))
) |>
  dplyr::mutate(rel_err = abs(model - paper) / paper)
knitr::kable(chk, digits = 6,
             caption = "Model parameters vs the paper's displayed equations.")

# Deterministic identity: exact to machine precision, so assert tightly.
stopifnot(all(chk$rel_err < 1e-10))
```

The reference values also reproduce Table 2 directly: CL/F 0.340 L/day and
V/F 8.44 L at 70 kg and 4.5 g/dL, giving an elimination half-life of

```{r pk-halflife}
thalf <- log(2) * s70$vc[1] / s70$cl[1]
round(thalf, 1)
stopifnot(abs(thalf - 17.2) < 0.2)   # deterministic; 17.2 days
```

which is in the usual range for an IgG cleared by non-specific catabolism.

## Closed-form check and mass balance

The ODE solve is checked against the analytic one-compartment first-order
absorption solution using the *same* parameters, so the only difference is
numerical -- a tight bound is correct here.

```{r pk-closed-form}
ka <- s70$ka[1]; cl <- s70$cl[1]; vc <- s70$vc[1]; kel <- cl / vc
bateman <- 1000 * 60 * ka / (vc * (ka - kel)) * (exp(-kel * s70$time) - exp(-ka * s70$time))

rel <- abs(s70$Cc - bateman) / pmax(bateman, 1e-8)
max(rel[s70$time > 0 & bateman > 1])
stopifnot(max(rel[s70$time > 0 & bateman > 1]) < 1e-4)
```

Mass balance: everything dosed is either still in the depot, in the central
compartment, or has been eliminated.

```{r pk-mass-balance}
# Eliminated amount = integral of CL * C dt, by the trapezoid rule.
# Cc is ng/mL, so divide by 1000 to get mg/L before multiplying by CL in L/day.
rate <- s70$cl * s70$Cc / 1000
elim <- c(0, cumsum(0.5 * (head(rate, -1) + tail(rate, -1)) * diff(s70$time)))
total <- s70$depot + s70$central + elim
max(abs(total - 60))
# Deterministic: the only error is the trapezoid discretisation of the
# elimination integral on the 0.25-day grid. Realised 0.0072 mg.
stopifnot(max(abs(total - 60)) < 0.02)
```

## Concentration-time profile and assay range

The paper's assay is validated over 100-6400 ng/mL (Methods, "Pharmacokinetic
and Pharmacodynamic Assessments"), which is an independent check that the model
is on the right concentration scale.

```{r pk-profile, fig.cap = "Typical nemolizumab profile after a single 60 mg SC dose, 70 kg subject at reference albumin."}
cmax_typ <- max(s70$Cc)
tmax_typ <- s70$time[which.max(s70$Cc)]
c(Cmax_ng_mL = round(cmax_typ), Tmax_day = tmax_typ)

# Cmax must land inside the validated assay range, not orders of magnitude out.
stopifnot(cmax_typ > 100, cmax_typ < 6400)

ggplot(s70, aes(time, Cc)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(100, 6400), linetype = "dashed", colour = "grey50") +
  labs(x = "Time (day)", y = "Serum nemolizumab (ng/mL)",
       subtitle = "Dashed lines: validated assay range 100-6400 ng/mL")
```

## Virtual cohort

A 150-subject cohort at the multiple-dose regimen used in the PN study: 30 mg
Q4W after a 60 mg loading dose, over the 16-week initial treatment phase.
Weight and albumin are drawn to match the PN cohort of Table 1.

```{r pk-cohort}
n_pk <- 150L
set.seed(11)
cov_pk <- tibble::tibble(
  id  = seq_len(n_pk),
  WT  = pmin(109.6, pmax(32.8, rlnorm(n_pk, log(61.2), 0.24))),
  ALB = pmin(51, pmax(27, rnorm(n_pk, 43, 3)))     # g/L; PN median 4.3 g/dL
)

ev_pk <- rxode2::et(amt = 60, cmt = "depot") |>                       # loading dose
  rxode2::et(amt = 30, cmt = "depot", time = 28, ii = 28, addl = 2) |> # 30 mg Q4W
  rxode2::et(seq(0, 112, by = 1), cmt = "central") |>                  # observations on the ODE state
  as.data.frame()

d_pk <- merge(cov_pk, ev_pk)
d_pk <- d_pk[order(d_pk$id, d_pk$time, -d_pk$evid), ]

sim_pk <- as.data.frame(rxode2::rxSolve(pk, d_pk))
stopifnot(nrow(sim_pk) > 0, !anyNA(sim_pk$Cc))
```

```{r pk-vpc, fig.cap = "Simulated nemolizumab concentrations, 30 mg Q4W with a 60 mg loading dose (150 subjects). Ribbon: 5th-95th percentile; line: median."}
pctl <- sim_pk |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(lo = quantile(sim, 0.05), md = median(sim), hi = quantile(sim, 0.95),
                   .groups = "drop")

ggplot(pctl, aes(time)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80") +
  geom_line(aes(y = md), linewidth = 0.8) +
  labs(x = "Time (day)", y = "Serum nemolizumab (ng/mL)")
```

## PKNCA validation

Non-compartmental analysis of the typical-value (`zeroRe`) single-dose profile.
Because the etas are zeroed, AUC(0-inf) must equal Dose/(CL/F) analytically,
which makes this a real gate rather than a self-consistency check.

```{r pknca}
nca_in <- pk_typ(70, 45, dose = 60, tmax = 240, dt = 0.25) |>
  dplyr::transmute(id = 1L, time, Cc, treatment = "60 mg SC single dose") |>
  dplyr::filter(!is.na(Cc))

# A time-zero record must be present or PKNCA warns per subject.
stopifnot(any(nca_in$time == 0))

o_conc <- PKNCA::PKNCAconc(nca_in, Cc ~ time | id / treatment)
o_dose <- PKNCA::PKNCAdose(
  data.frame(id = 1L, time = 0, dose = 60),
  dose ~ time | id
)
o_data <- PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf,
                         cmax = TRUE, tmax = TRUE, half.life = TRUE,
                         aucinf.obs = TRUE)
)
res <- suppressWarnings(as.data.frame(PKNCA::pk.nca(o_data)))
nca <- setNames(res$PPORRES, res$PPTESTCD)

auc_analytic <- 60 / cl_paper(70, 45) * 1000   # mg / (L/day) -> ng*day/mL

nca_tab <- tibble::tibble(
  `NCA parameter` = c("Cmax (ng/mL)", "Tmax (day)", "t-half (day)", "AUC0-inf (ng*day/mL)"),
  `PKNCA`         = c(nca[["cmax"]], nca[["tmax"]], nca[["half.life"]], nca[["aucinf.obs"]]),
  `Analytic`      = c(cmax_typ, tmax_typ, log(2) * vc / cl, auc_analytic)
) |>
  dplyr::mutate(`% difference` = 100 * (PKNCA - Analytic) / Analytic)
knitr::kable(nca_tab, digits = 2,
             caption = "PKNCA on the typical-value profile vs the analytic one-compartment result.")

# Deterministic (no IIV, no residual): a tight bound is correct here.
stopifnot(max(abs(nca_tab$`% difference`)) < 1)
```

The paper reports **no** NCA table of its own -- its PK validation is entirely
GOF plots and VPCs (Fig. 1, Figs. S1-S5) -- so there is no published Cmax /
AUC to compare against, and this section deliberately validates against the
analytic identity instead.

# Part 2 -- Population PD: weekly average PP-NRS

## Reading Pmax and Edrug off Table 2

Table 2 prints `Pmax = 0.618` and `Edrug = 1.41` with no units. Taking them at
face value as PP-NRS point reductions is wrong: they are the **log-scale
thetas** of the paper's stated exponential (log-normal) IIV
parameterisation. This subsection reproduces the adjudication, because it is
the single most consequential reading in the extraction.

The paper's Fig. 2 plots the observed 5th, 50th and 95th percentiles of the
weekly average PP-NRS for each arm. Those observed percentiles, read from the
figure, are the target:

```{r ppnrs-targets}
days <- c(0, 28, 56, 84, 112)
obs_fig2 <- list(
  placebo     = rbind(p05 = c(7.0, 6.1, 4.4, 3.0, 2.7),
                      p50 = c(8.3, 8.0, 7.8, 8.0, 7.7),
                      p95 = c(10,  10,  10,  10,  10)),
  nemolizumab = rbind(p05 = c(7.0, 1.6, 0.3, 0.0, 0.1),
                      p50 = c(8.5, 6.0, 4.6, 3.6, 3.1),
                      p95 = c(10,  8.9, 8.5, 8.9, 8.2))
)
```

Simulate the source's own design under both readings. 200 subjects per arm.

```{r ppnrs-adjudicate}
n_pd <- 200L

pd_cohort <- function(dose, base_median) {
  set.seed(if (dose > 0) 21 else 22)
  b <- pmin(10, pmax(6.4, rnorm(n_pd, base_median, 0.9)))
  d <- expand.grid(time = days, id = seq_len(n_pd))
  d$SCORE_PPNRS         <- b[d$id]
  d$DOSE_NEMOLIZUMAB_MG <- dose
  d$evid <- 0L
  d$amt  <- 0
  d[order(d$id, d$time), ]
}

run_pd <- function(model, dose, base_median) {
  s <- as.data.frame(rxode2::rxSolve(model, pd_cohort(dose, base_median), nSub = n_pd))
  sapply(split(s$sim, s$time), quantile, c(0.05, 0.50, 0.95))
}

# The shipped model = log-scale reading.
q_log <- list(placebo     = run_pd(pd, 0,  8.4),
              nemolizumab = run_pd(pd, 30, 8.6))

# The linear reading, for comparison only: overwrite the two thetas with
# log(0.618) and log(1.41) so the typical maxima become 0.618 and 1.41 points.
pd_linear <- rxode2::ini(pd, lpmax = log(0.618), lemax = log(1.41))
q_lin <- list(placebo     = run_pd(pd_linear, 0,  8.4),
              nemolizumab = run_pd(pd_linear, 30, 8.6))
```

```{r ppnrs-adjudicate-table, echo = FALSE}
row_of <- function(q, arm, pct) q[[arm]][pct, ]
adj <- tibble::tibble(
  Arm        = rep(c("placebo", "nemolizumab"), each = 5),
  Day        = rep(days, 2),
  `Observed (Fig. 2)` = c(obs_fig2$placebo["p50", ], obs_fig2$nemolizumab["p50", ]),
  `Log-scale reading` = c(row_of(q_log, "placebo", 2), row_of(q_log, "nemolizumab", 2)),
  `Linear reading`    = c(row_of(q_lin, "placebo", 2), row_of(q_lin, "nemolizumab", 2))
)
knitr::kable(adj, digits = 2,
             caption = "Median weekly average PP-NRS: observed (Fig. 2) vs the two readings of Table 2.")
```

The linear reading misses the active-arm median at day 112 by more than 3
points -- more than the entire observed placebo-arm change -- and it cannot be
rescued by the IIV, because with `Edrug = 1.41` as a point reduction the
*median* subject's drug effect is capped at 1.41 points however large the etas
are. The log-scale reading tracks both arms.

```{r ppnrs-adjudicate-gate}
err_log <- abs(row_of(q_log, "nemolizumab", 2) - obs_fig2$nemolizumab["p50", ])
err_lin <- abs(row_of(q_lin, "nemolizumab", 2) - obs_fig2$nemolizumab["p50", ])
c(log_scale_max_err = max(err_log), linear_max_err = max(err_lin))

# Cohort-derived, so bounds are deliberately loose (see the note below). The
# separation between the two readings is ~3 points at day 112, far outside any
# thread-count noise, which is what makes this gate meaningful.
stopifnot(
  max(err_log) < 1.5,   # log-scale reading tracks the observed median
  max(err_lin) > 2.5    # linear reading does not
)
```

These are cohort statistics, so the bounds admit the sampling noise rather than
pinning one run: `rxSetSeed()` fixes rxode2's stream per solver thread but not
across thread counts, so a CI runner draws a different cohort. Realised
`max(err_log)` was 0.8 and `max(err_lin)` 3.7 while authoring; the two readings
are separated by about 3 points, so bounds of 1.5 and 2.5 sit well outside the
noise while still going red on a mis-transcribed theta.

## Full VPC against Figure 2

```{r ppnrs-vpc-table, echo = FALSE}
vpc_tab <- do.call(rbind, lapply(c("placebo", "nemolizumab"), function(arm) {
  tibble::tibble(
    Arm = arm, Day = days,
    `Sim 5th`  = q_log[[arm]][1, ], `Obs 5th`  = obs_fig2[[arm]]["p05", ],
    `Sim 50th` = q_log[[arm]][2, ], `Obs 50th` = obs_fig2[[arm]]["p50", ],
    `Sim 95th` = q_log[[arm]][3, ], `Obs 95th` = obs_fig2[[arm]]["p95", ]
  )
}))
knitr::kable(vpc_tab, digits = 1,
             caption = "Simulated vs observed PP-NRS percentiles (observed values read from Takechi 2025 Fig. 2).")
```

The 50th and 95th percentiles track the observed data in both arms. The 5th
percentile runs far below it -- in the active arm it goes negative, because the
model has no floor at 0 and the additive residual is applied to an already
small predicted score.

**This is the paper's own model behaving as the paper says it does**, not an
extraction error. Results, "PopPD Analysis": *"The observed 50th and 95th
percentiles were almost within the 95% prediction interval (PI) of the model
simulation. The simulated 5th percentile slightly underestimated the severity
of pruritus during early treatment, suggesting a modest overprediction of early
response in some patients."* Fig. 2's lower prediction band sits clipped
against the zero axis for the same reason. It is recorded here as a deviation
and excluded from the gate.

```{r ppnrs-vpc-gate}
gate <- do.call(rbind, lapply(c("placebo", "nemolizumab"), function(arm) {
  data.frame(arm = arm, pct = c("50th", "95th"),
             err = c(max(abs(q_log[[arm]][2, ] - obs_fig2[[arm]]["p50", ])),
                     max(abs(q_log[[arm]][3, ] - obs_fig2[[arm]]["p95", ]))))
}))
knitr::kable(gate, digits = 2, caption = "Maximum absolute deviation from the observed percentile, by arm.")
stopifnot(nrow(gate) == 4L, max(gate$err) < 1.5)
```

```{r ppnrs-plot, fig.cap = "Simulated median PP-NRS (lines) against the observed medians read from Takechi 2025 Fig. 2 (points)."}
plot_df <- dplyr::bind_rows(
  tibble::tibble(arm = "placebo",     day = days, sim = q_log$placebo[2, ],
                 obs = obs_fig2$placebo["p50", ]),
  tibble::tibble(arm = "nemolizumab", day = days, sim = q_log$nemolizumab[2, ],
                 obs = obs_fig2$nemolizumab["p50", ])
)
ggplot(plot_df, aes(day, colour = arm)) +
  geom_line(aes(y = sim), linewidth = 0.8) +
  geom_point(aes(y = obs), size = 2.5) +
  ylim(0, 10) +
  labs(x = "Time (day)", y = "Weekly average PP-NRS", colour = NULL,
       subtitle = "Lines: simulated median. Points: observed median (Fig. 2).")
```

## Onset rates

The paper's rapid-onset claim is a statement about `Kd` relative to `Kp`.

```{r ppnrs-onset}
onset <- tibble::tibble(
  effect          = c("placebo", "nemolizumab"),
  `rate (1/day)`  = c(2.74e-3, 28.7e-3),
  `half-life (d)` = log(2) / c(2.74e-3, 28.7e-3),
  `% of maximum expressed by day 112` = 100 * (1 - exp(-c(2.74e-3, 28.7e-3) * 112))
)
knitr::kable(onset, digits = 1, caption = "Onset of the placebo and nemolizumab effects.")

stopifnot(
  abs(28.7e-3 / 2.74e-3 - 10.5) < 0.1,                       # deterministic
  onset$`% of maximum expressed by day 112`[2] > 95,
  onset$`% of maximum expressed by day 112`[1] < 30
)
```

Only about a quarter of the placebo maximum is expressed within the 112-day
observation window, against essentially all of the nemolizumab maximum. That
asymmetry -- not the ratio of the maxima alone -- is what produces the roughly
eight-fold separation between the two arms in Fig. 2.

# Part 3 -- Model-based meta-analysis of IGA success rates

This model is a **study-arm-level** model. Its output is the proportion of an
arm achieving at least a 2-point decrease in IGA, not an individual patient's
outcome, and its only random effect is between-study.

## Reproducing Figure 4

Fig. 4B plots the typical (population-mean) time course of the IGA success rate
for placebo, dupilumab and nemolizumab, and Fig. 4D gives simulated medians at
weeks 4 and 16. With the between-study eta zeroed, the model is deterministic,
so these are exact regression tests.

```{r mbma-fig4}
mbma0 <- rxode2::zeroRe(mbma)

arms <- tibble::tribble(
  ~arm,          ~DOSE_NEMOLIZUMAB_MG, ~DOSE_DUPILUMAB_MG,
  "placebo",     0,                    0,
  "dupilumab",   0,                    300,
  "nemolizumab", 30,                   0
)
weeks <- c(4, 16, 24)

d_mbma <- arms |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::crossing(time = weeks) |>
  dplyr::mutate(evid = 0L, amt = 0) |>
  dplyr::arrange(id, time)

sim_mbma <- as.data.frame(rxode2::rxSolve(mbma0, as.data.frame(d_mbma)))
sim_mbma$arm <- arms$arm[sim_mbma$id]

fig4 <- sim_mbma |>
  dplyr::transmute(Arm = arm, Week = time, `Model (%)` = 100 * pResp) |>
  dplyr::left_join(
    tibble::tribble(
      ~Arm,          ~Week, ~`Read from Fig. 4 (%)`,
      "placebo",     4,     2.5,
      "placebo",     16,    8.8,
      "placebo",     24,    11.0,
      "dupilumab",   4,     4.0,
      "dupilumab",   16,    26.5,
      "dupilumab",   24,    34.0,
      "nemolizumab", 4,     8.5,
      "nemolizumab", 16,    34.0,
      "nemolizumab", 24,    41.0
    ), by = c("Arm", "Week")
  ) |>
  dplyr::mutate(`Difference (points)` = `Model (%)` - `Read from Fig. 4 (%)`)
knitr::kable(fig4, digits = 1,
             caption = "IGA success rate: model vs values read from Takechi 2025 Figs. 4B and 4D.")

# Deterministic (zeroRe), so the only error is figure read-off. A
# mis-transcribed theta moves these by many percentage points.
stopifnot(nrow(fig4) == 9L, max(abs(fig4$`Difference (points)`)) < 1.5)
```

Nine independent points across three arms and three time points are reproduced
to within figure read-off precision, which validates the intercept, the placebo
maximum and rate, and both drugs' maxima and rates simultaneously.

```{r mbma-curves, fig.cap = "Typical IGA success-rate time courses. Replicates Figure 4B of Takechi 2025."}
d_curve <- arms |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::crossing(time = seq(0, 24, by = 0.5)) |>
  dplyr::mutate(evid = 0L, amt = 0) |>
  dplyr::arrange(id, time)
curve <- as.data.frame(rxode2::rxSolve(mbma0, as.data.frame(d_curve)))
curve$arm <- arms$arm[curve$id]

ggplot(curve, aes(time, 100 * pResp, colour = arm)) +
  geom_line(linewidth = 0.8) +
  ylim(0, 80) +
  labs(x = "Time (week)", y = "IGA success rate (%)", colour = NULL)
```

Nemolizumab is above dupilumab throughout, and the gap is widest early: at week
4 the model gives

```{r mbma-early}
early <- fig4 |> dplyr::filter(Week == 4) |> dplyr::select(Arm, `Model (%)`)
knitr::kable(early, digits = 1, caption = "Week-4 IGA success rate.")

nemo4 <- early$`Model (%)`[early$Arm == "nemolizumab"]
dupi4 <- early$`Model (%)`[early$Arm == "dupilumab"]
stopifnot(nemo4 > 2 * dupi4)   # deterministic
```

which is the quantitative content of the paper's "particularly notable rapid
onset of effects during the early treatment phase". It is driven by the onset
rate constants, not by the maxima: nemolizumab's `Kd` is 0.416/week against
dupilumab's 0.134/week, while its maximum effect (1.71) is only slightly larger
than dupilumab's (1.49).

## Arm-size weighting of the residual

The paper's residual is scaled by the binomial standard error of the arm:
`success rate = Pr + eps * sqrt(Pr * (1 - Pr) / N)`. The model encodes the
prediction-dependent factor `sqrt(Pr * (1 - Pr))` exactly but ships the `N = 1`
residual, because `N` is a property of the arm being simulated rather than of
the model -- the same convention as `Boucher_2018_naproxen_mbma`. Divide by
`sqrt(N_arm)` to reproduce a real arm.

```{r mbma-residual}
set.seed(7)
n_rep  <- 400L
N_arm  <- 190L   # OLYMPIA 1 nemolizumab arm (Table 3)

d_rep <- data.frame(id = seq_len(n_rep), time = 24,
                    DOSE_NEMOLIZUMAB_MG = 30, DOSE_DUPILUMAB_MG = 0,
                    evid = 0L, amt = 0)
rep_sim <- as.data.frame(rxode2::rxSolve(mbma0, d_rep, nSub = n_rep))

sd_shipped  <- sd(rep_sim$sim)
sd_expected <- sqrt(0.619) * sqrt(mean(rep_sim$pResp) * (1 - mean(rep_sim$pResp)))
sd_for_arm  <- sd_shipped / sqrt(N_arm)

tibble::tibble(
  quantity = c("simulated residual SD (N = 1)",
               "expected addSd * sqrt(Pr(1-Pr))",
               paste0("rescaled for an arm of N = ", N_arm)),
  value    = c(sd_shipped, sd_expected, sd_for_arm)
) |>
  knitr::kable(digits = 4, caption = "Residual SD of the shipped model and its per-arm rescaling.")

# Cohort-derived SD from 400 draws: allow generous sampling noise.
stopifnot(abs(sd_shipped - sd_expected) / sd_expected < 0.20)
# The rescaled per-arm SD must be a plausible few percentage points.
stopifnot(sd_for_arm > 0.005, sd_for_arm < 0.05)
```

# Part 4 -- Model-based meta-analysis of PP-NRS success rates

The same six trials, the same logistic structure, the second endpoint: the
proportion of an arm achieving at least a **4-point improvement in the weekly
average PP-NRS**. Two things differ from Part 3. The onset rate constants are
far more separated (nemolizumab 1.93/week against dupilumab 0.119/week, a
factor of sixteen, versus a factor of three for IGA), and this endpoint retains
a covariate.

## The covariate and its functional form

Table 4's PP-NRS column carries a row absent from the IGA column:
`PPM effect on Edrug,nemolizumab = -1.84`, where `PPM` is the paper's
abbreviation for the *proportion of patients classified as moderate by IGA* in
a study arm. It is registered here as `IGA_MOD_PCT`, an arm-level percentage,
and is the complement of Table 3's `IGA4 at baseline (%)` column.

```{r ppnrs-ppm-arms}
arm_ppm <- tibble::tibble(
  iga4_pct = c(46.1, 54.5, 52.6, 39.0, 53.0, 35.4, 43.7, 47.3, 41.0,
               29.3, 28.0, 39.5, 37.2)
) |>
  dplyr::mutate(IGA_MOD_PCT = 100 - iga4_pct)

ppm_median <- median(arm_ppm$IGA_MOD_PCT)
cat("13 arms; IGA_MOD_PCT range",
    sprintf("%.1f-%.1f%%", min(arm_ppm$IGA_MOD_PCT), max(arm_ppm$IGA_MOD_PCT)),
    "median", sprintf("%.1f%%", ppm_median), "\n")

# The centring constant hardcoded in model() must be the Table 3 median.
stopifnot(nrow(arm_ppm) == 13L, ppm_median == 59.0)
```

**The paper never prints the covariate equation.** Only the coefficient is
given. The form encoded in the model file is a fractional deviation centred on
that median arm,

$$E_{drug,nemo} = 1.63 \times \left(1 - 1.84 \left(\frac{\mathtt{IGA\_MOD\_PCT}}{100} - 0.590\right)\right)$$

and it was chosen because it is the only one of three candidate readings that
reproduces the paper's own Fig. 4. Fig. 4's caption states the simulation used
*"the recommended dose of each drug in Japan"*, i.e. the Yokozeki 30 mg Q4W
arm, whose `IGA_MOD_PCT` is `100 - 54.5 = 45.5`. That fixes the covariate value
and makes the three readings testable against three plotted numbers.

```{r ppnrs-form-selection}
# Closed-form logistic prediction, independent of the shipped model, so this
# genuinely discriminates between the candidate forms.
pr_ppnrs <- function(week, edrug_nemo) {
  f0 <- -4.38 + 2.64 * (1 - exp(-0.277 * week))
  100 * plogis(f0 + edrug_nemo * (1 - exp(-1.93 * week)))
}
ppm_ref  <- 0.455          # Yokozeki nemolizumab 30 mg Q4W arm
ppm_cent <- 0.590          # Table 3 median

candidates <- tibble::tibble(
  Form = c("fractional deviation (shipped)", "normalised power",
           "additive logit shift", "no covariate"),
  Edrug = c(1.63 * (1 - 1.84 * (ppm_ref - ppm_cent)),
            1.63 * (ppm_ref / ppm_cent)^(-1.84),
            1.63 - 1.84 * (ppm_ref - ppm_cent),
            1.63)
) |>
  dplyr::mutate(
    `Week 4 (%)`  = pr_ppnrs(4,  Edrug),
    `Week 16 (%)` = pr_ppnrs(16, Edrug),
    `Week 24 (%)` = pr_ppnrs(24, Edrug)
  )

knitr::kable(candidates, digits = 2,
             caption = paste("Candidate covariate forms for the nemolizumab arm of Fig. 4.",
                             "Fig. 4C reads ~36% at week 4 and ~56% at week 16;",
                             "the Fig. 4A plateau is ~57%."))

shipped <- candidates[candidates$Form == "fractional deviation (shipped)", ]
# Deterministic arithmetic on transcribed thetas -- exact, not cohort-derived.
stopifnot(
  abs(shipped$`Week 4 (%)`  - 36) < 2,
  abs(shipped$`Week 16 (%)` - 56) < 2,
  abs(shipped$`Week 24 (%)` - 57) < 2
)
# ... and every alternative must MISS at week 24, or the test proves nothing.
others <- candidates[candidates$Form != "fractional deviation (shipped)", ]
stopifnot(all(abs(others$`Week 24 (%)` - 57) > 3))
```

The power function -- the form the paper's Methods states for the *PopPD*
analysis -- overshoots by 14 percentage points, the additive shift undershoots
by 4, and dropping the covariate undershoots by 10. Only the fractional
deviation lands. This is a **reconstruction**, and it is recorded as one in the
model file and in the Assumptions below.

## Reproducing Figure 4A and 4C

With the covariate form settled, the placebo and dupilumab arms are an
independent check: neither carries the covariate, so if the rest of the column
were mis-transcribed they would miss too.

```{r ppnrs-fig4}
mbma_p0 <- rxode2::zeroRe(mbma_ppnrs)

arms_p <- tibble::tribble(
  ~arm,          ~DOSE_NEMOLIZUMAB_MG, ~DOSE_DUPILUMAB_MG, ~IGA_MOD_PCT,
  "placebo",     0,                    0,                  45.5,
  "dupilumab",   0,                    300,                45.5,
  "nemolizumab", 30,                   0,                  45.5
)

d_p <- arms_p |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::crossing(time = c(4, 16, 24)) |>
  dplyr::mutate(evid = 0L, amt = 0) |>
  dplyr::arrange(id, time)

sim_p <- as.data.frame(rxode2::rxSolve(mbma_p0, as.data.frame(d_p)))
sim_p$arm <- arms_p$arm[sim_p$id]

fig4a <- sim_p |>
  dplyr::transmute(Arm = arm, Week = time, `Model (%)` = 100 * pResp) |>
  dplyr::left_join(
    tibble::tribble(
      ~Arm,          ~Week, ~`Read from Fig. 4 (%)`,
      "placebo",     4,     7,
      "placebo",     16,    14,
      "placebo",     24,    15,
      "dupilumab",   4,     13,
      "dupilumab",   16,    46,
      "dupilumab",   24,    52,
      "nemolizumab", 4,     36,
      "nemolizumab", 16,    56,
      "nemolizumab", 24,    57
    ), by = c("Arm", "Week")
  ) |>
  dplyr::mutate(`Difference (points)` = `Model (%)` - `Read from Fig. 4 (%)`)

knitr::kable(fig4a, digits = 1,
             caption = "PP-NRS success rate: model vs values read from Takechi 2025 Figs. 4A and 4C.")

# Deterministic (zeroRe), so the only error is figure read-off. The week-4 and
# week-16 targets are the open and closed circles of the Fig. 4C forest plot;
# the week-24 targets are the right-hand end of the Fig. 4A curves. These were
# read off a 150 dpi render of the publisher's page and are good to roughly a
# point, so the gate is set at 2 -- still far tighter than the many-point shift
# a mis-transcribed theta produces.
stopifnot(nrow(fig4a) == 9L, max(abs(fig4a$`Difference (points)`)) < 2)
```

```{r ppnrs-curves, fig.cap = "Typical PP-NRS success-rate time courses for the Yokozeki 30 mg Q4W arm severity. Replicates Figure 4A of Takechi 2025."}
d_pc <- arms_p |>
  dplyr::mutate(id = dplyr::row_number()) |>
  tidyr::crossing(time = seq(0, 24, by = 0.5)) |>
  dplyr::mutate(evid = 0L, amt = 0) |>
  dplyr::arrange(id, time)
curve_p <- as.data.frame(rxode2::rxSolve(mbma_p0, as.data.frame(d_pc)))
curve_p$arm <- arms_p$arm[curve_p$id]

ggplot(curve_p, aes(time, 100 * pResp, colour = arm)) +
  geom_line(linewidth = 0.8) +
  ylim(0, 80) +
  labs(x = "Time (week)", y = "PP-NRS success rate (%)", colour = NULL)
```

The nemolizumab curve is essentially at its plateau by week 4 while dupilumab
is still climbing at week 24 -- the clearest quantitative statement in the paper
of the rapid-itch-relief claim, and a much starker separation than the IGA
endpoint of Part 3 shows.

## Covariate direction across the observed arm range

A lower `IGA_MOD_PCT` means a more severe arm, and the negative coefficient
makes such an arm respond more. The effect is not small over the range the
meta-analysis actually spans.

```{r ppnrs-covariate-range}
d_cov <- tibble::tibble(IGA_MOD_PCT = sort(arm_ppm$IGA_MOD_PCT)) |>
  dplyr::mutate(id = dplyr::row_number(),
                DOSE_NEMOLIZUMAB_MG = 30, DOSE_DUPILUMAB_MG = 0,
                time = 24, evid = 0L, amt = 0)

cov_sim <- as.data.frame(rxode2::rxSolve(mbma_p0, as.data.frame(d_cov)))
cov_sim$IGA_MOD_PCT <- d_cov$IGA_MOD_PCT[cov_sim$id]

cov_tab <- cov_sim |>
  dplyr::transmute(`Arm moderate-IGA (%)` = IGA_MOD_PCT,
                   `Week-24 success (%)`  = 100 * pResp)
knitr::kable(cov_tab, digits = 1,
             caption = "Week-24 nemolizumab PP-NRS success rate across the 13 arms' observed severity range.")

# Monotone decreasing in IGA_MOD_PCT: a negative coefficient, exactly.
stopifnot(all(diff(cov_tab$`Week-24 success (%)`) < 0))
# The spread across the observed range is material, not cosmetic.
stopifnot(diff(range(cov_tab$`Week-24 success (%)`)) > 10)
```

## Arm-size weighting of the residual

Identical convention to Part 3: the model ships the `N = 1` residual and the
`1 / sqrt(N_arm)` factor is applied downstream.

```{r ppnrs-residual}
set.seed(11)
n_rep_p <- 400L
N_arm_p <- 77L   # Yokozeki nemolizumab 30 mg Q4W arm (Table 3)

d_rep_p <- data.frame(id = seq_len(n_rep_p), time = 24,
                      DOSE_NEMOLIZUMAB_MG = 30, DOSE_DUPILUMAB_MG = 0,
                      IGA_MOD_PCT = 45.5, evid = 0L, amt = 0)
rep_p <- as.data.frame(rxode2::rxSolve(mbma_p0, d_rep_p, nSub = n_rep_p))

sd_shipped_p  <- sd(rep_p$sim)
sd_expected_p <- sqrt(0.738) * sqrt(mean(rep_p$pResp) * (1 - mean(rep_p$pResp)))
sd_for_arm_p  <- sd_shipped_p / sqrt(N_arm_p)

tibble::tibble(
  quantity = c("simulated residual SD (N = 1)",
               "expected addSd * sqrt(Pr(1-Pr))",
               paste0("rescaled for an arm of N = ", N_arm_p)),
  value    = c(sd_shipped_p, sd_expected_p, sd_for_arm_p)
) |>
  knitr::kable(digits = 4, caption = "Residual SD of the shipped PP-NRS model and its per-arm rescaling.")

# Cohort-derived SD from 400 draws: allow generous sampling noise.
stopifnot(abs(sd_shipped_p - sd_expected_p) / sd_expected_p < 0.20)
stopifnot(sd_for_arm_p > 0.005, sd_for_arm_p < 0.10)
```

# Assumptions and deviations

**1. `Pmax` and `Edrug` of the PopPD model are read as log-scale thetas.**
Table 2 prints them without units while labelling `Kp` and `Kd`
`(x 10^-3/day)`. The evidence is set out in full in the "Reading Pmax and
Edrug" section above and in the model file: a unitless column, an unrescuable
3.7-point miss of the Fig. 2 active-arm median under the linear reading, a
matching 50th and 95th percentile under the log-scale reading, and the paper's
own statement that its simulated 5th percentile ran too low. The paper never
states the scale explicitly, so this is an inference, but it is a
figure-falsified one rather than a preference.

**2. The PopPD equation that combines the two response components with the
baseline is not printed in the paper.** Methods "Pharmacodynamic Model" gives
`Placebo response(t)` and `Drug response(t)` but no line assembling them into a
PP-NRS prediction. The additive assembly used here --
`PP-NRS = baseline - placebo response - drug response` -- follows from the
paper's own definition of `Pmax` and `Edrug` as *"the maximum effects, defined
as the reduction in PP-NRS scores"* and is what reproduces Fig. 2. This is a
genuine reporting gap in the source, not a converter loss: the PDF, the PMC
full-text XML and the supplement all contain only the two component equations.

**3. Variance-vs-SD reading of the variability rows.** Table 2's IIV rows are
variances, proved above from the CL/F-V/F covariance row. The same convention
is applied to Table 2's `Additive error` (0.582 -> SD 0.763) and Table 4's
`ISV on EFF` and `Additive error`, on the grounds that every row in those
tables is on the raw NONMEM scale except the one the authors explicitly
relabelled `log normal error (CV%)`. Reading Table 2's additive error instead
as an SD of 0.582 changes the residual by 24% and no structural conclusion; a
user who prefers that reading can set `addSd = 0.582`.

**4. Serum albumin units.** Table 1 heads the albumin row "Serum albumin
(g/L)" but tabulates 4.3-4.5 with a range of 2.0-5.2. Those are **g/dL**: 45
g/L is a normal serum albumin and 4.5 g/L is not survivable, and the model
equation's own reference of 4.5 confirms it. The Table 1 header is a
publication unit-label error. The covariate register stores `ALB` in SI g/L, so
the model applies the register's documented inline conversion
`alb_gdL <- ALB * 0.1` before the power term.

**5. Allometric exponents are `fixed()`.** Neither 0.75 nor the implicit 1 on
V/F appears as an estimated row in Table 2 with an %RSE or a bootstrap CI; both
are printed only inside the displayed equations, so both are encoded as
structural assumptions.

**6. Observed percentiles in Part 2 and the Fig. 4 values in Part 3 are read
from figures.** The paper tabulates neither, so the comparison targets were
read off Fig. 2 and Figs. 4B/4D at the resolution of the publisher's
figure files. The gates are set to tolerate read-off error (1.5 points /
1.5 percentage points) while still failing on a mis-transcribed parameter,
which moves these quantities by many points. No parameter value in any model
file is figure-derived -- every `ini()` entry comes from Table 2 or Table 4.

**7. The MBMA residual ships at `N = 1`.** See "Arm-size weighting" above.

**8. No PK-PD link.** The PopPD and MBMA models take no concentration input.
This is the authors' structure, not a simplification: no exposure-response
relationship was identifiable, so the drug effects are constants selected by a
dose column used only as an on/off switch.

**9. No 0-10 floor on the PP-NRS prediction.** The source describes no
clamping, and clamping would change the residual distribution near zero.
Simulated values below 0 occur for strong responders and are visible in the
paper's own Fig. 2, whose lower prediction band sits against the axis.

**10. Simulation scope.** Both `Takechi_2025_nemolizumab_mbma_iga` and
`Takechi_2025_nemolizumab_mbma_ppnrs` predict study-arm success rates and must
not be used to simulate individual patients; their only random effect is
between-study. `Takechi_2025_nemolizumab_ppnrs` does describe individual
subjects.

**11. The PP-NRS covariate's functional form is reconstructed, not quoted.**
Table 4 prints the coefficient `-1.84` but the paper contains no equation
attaching `PPM` to `Edrug,nemolizumab` anywhere -- not in Methods, not in the
supplement, and there is no NONMEM control stream on disk. The fractional
deviation `1.63 * (1 + (-1.84) * (IGA_MOD_PCT/100 - 0.590))` used here is the
reading that reproduces three independent plotted values of Fig. 4, where the
two obvious alternatives miss by 14 and 4 percentage points respectively; Part 4
runs that discrimination as a test. The centring constant `0.590` is likewise
not printed -- it is the median of Table 3's 13 arms, which the vignette
recomputes and asserts. A reader who has access to the authors' control stream
should check this term before relying on the model outside the 45.5-72.0%
severity range the meta-analysis spans.

**12. `IGA_MOD_PCT` scale.** The covariate register stores this column as a
**percent** (0-100), matching its arm-level siblings `TUMTP_SQUAM_PCT`,
`PS_ECOG_0_PCT` and `RACE_ASIAN_PCT`; the model converts to a fraction inline.
The paper's `PPM` is a fraction, so a user transcribing directly from Table 4's
notation must multiply by 100 before supplying the column.

# Session info

```{r session-info}
sessionInfo()
```
