---
title: "Ozoralizumab (Takeuchi 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ozoralizumab (Takeuchi 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Takeuchi_2023_ozoralizumab"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Takeuchi_2023_ozoralizumab"))$description`
* Article: [J Clin Pharmacol. 2024;64(4):418-427](https://doi.org/10.1002/jcph.2380)
  (open access)

# Population

Takeuchi 2023 pooled two clinical trials of subcutaneous ozoralizumab in
Japanese patients with rheumatoid arthritis (RA): the OHZORA trial (Phase
II/III, with concomitant methotrexate; jRCT2080223971; n = 363) and the
NATSUZORA trial (Phase III, without methotrexate; jRCT2080223973; n = 131).
After exclusion of placebo-treated patients, protocol violations, BLQ records,
and outliers (CWRES absolute value > 6), the analysis dataset comprised 3412
plasma concentrations from 494 patients dosed at 30 or 80 mg every 4 weeks for
up to 52 weeks.

Baseline demographics (Table 2): mean age 56 (SD 12, range 21-84) years,
76% female, mean body weight 59 (SD 13, range 35-112) kg, mean baseline
albumin 3.9 (SD 0.3) g/dL, mean baseline eGFR 88 (SD 20, range 35-174)
mL/min/1.73 m^2 (Japanese formula), mean CDAI 32 (SD 11), mean hs-CRP
1.6 (SD 2.0) mg/dL. 363 of 494 patients (74%) received concomitant
methotrexate; 185 of 494 (37%) developed antidrug antibodies. The
final-model equation centres body weight at 56.65 kg and eGFR at
85.95 mL/min/1.73 m^2 — the population medians, which sit below the
right-skewed means.

The same information is available programmatically via
`readModelDb("Takeuchi_2023_ozoralizumab")$population`.

# Source trace

Every structural parameter, covariate effect, IIV element, and residual-error
term is taken from Takeuchi 2023 Table 3 (p. 421) and the final-model
equations on p. 422.

| Equation / parameter           | Value                  | Source location                       |
|--------------------------------|------------------------|---------------------------------------|
| `lka` (Ka)                     | `log(0.0343)` 1/h      | Table 3, Ka row                       |
| `lcl` (CL/F, L/h)              | `log(9.20 / 1000)` L/h | Table 3, CL/F row (paper reports mL/h)|
| `lvc` (Vd/F)                   | `log(4.91)` L          | Table 3, Vd/F row                     |
| `e_wt_cl`                      | `0.847`                | Table 3, "WT on CL/F"                 |
| `e_egfr_cl`                    | `0.191`                | Table 3, "eGFR on CL/F"               |
| `e_nomtx_cl`                   | `0.126`                | Table 3, "MTX on CL/F"                |
| `e_male_cl`                    | `0.117`                | Table 3, "Sex on CL/F"                |
| `e_ada_cl`                     | `0.0967`               | Table 3, "ADA on CL/F"                |
| `e_wt_vc`                      | `0.469`                | Table 3, "WT on Vd/F"                 |
| `e_male_vc`                    | `0.134`                | Table 3, "Sex on Vd/F"                |
| `var(etalcl)`                  | `0.0423`               | Table 3, omega^2_CL/F                 |
| `cov(etalcl, etalvc)`          | `0.00806`              | Table 3, omega_CL/F-Vd/F              |
| `var(etalvc)`                  | `0.0230`               | Table 3, omega^2_Vd/F                 |
| `var(etalka)`                  | 0 (fixed; not modelled)| Table 3, omega^2_Ka fixed             |
| `propSd`                       | `sqrt(0.0347)`         | Table 3, sigma^2 = 0.0347             |
| Structure (1-cmt + first-order SC absorption + linear elimination) | n/a | Methods p. 420 / Equations p. 422 |
| WT reference (population median)   | 56.65 kg               | Final-model equation, p. 422          |
| eGFR reference (population median) | 85.95 mL/min/1.73 m^2  | Final-model equation, p. 422          |

## Parameterization notes

- **Unit conversion on CL/F.** Takeuchi 2023 reports CL/F in mL/h. The model
  file converts to L/h (`lcl <- log(9.20 / 1000)`) so that with Vd/F in L, the
  ODE term `cl/vc` has units 1/h and `Cc = central / vc` yields mg/L = ug/mL.
- **Multiplicative `(1 + theta * cov)` covariate form.** Takeuchi parameterises
  binary covariates as `P = theta1 * (1 + theta3 * cov)` (Methods p. 420),
  rather than the more common `exp(theta * cov)`. The model file uses the
  paper's form unchanged.
- **Sex / MTX reference-category flips.** Takeuchi encodes Male = 1 with
  female as the TVCL/TVVd reference, and MTX = 1 with on-MTX as the TVCL
  reference. To preserve the published TVCL = 9.20 mL/h and TVVd = 4.91 L
  while storing the data under the canonical `SEXF` (1 = female) and
  `CONMED_MTX` (1 = on MTX) columns, the model applies the effects as
  `(1 + e_male_* * (1 - SEXF))` and `(1 + e_nomtx_cl * (1 - CONMED_MTX))`.
  The published coefficient values are unchanged.
- **eGFR formula.** Takeuchi computes eGFR with the Japanese formula
  `194 * Scr^-1.094 * age^-0.287 * (0.739 if female)` (Methods p. 420),
  not MDRD or CKD-EPI. The canonical `CRCL` column carries the precomputed
  eGFR value in mL/min/1.73 m^2.
- **Ka has no IIV.** The paper fixed `omega^2_Ka` to 0 because of high
  shrinkage (Methods p. 421). The model file omits `etalka` accordingly;
  Ka is the typical value for every individual.
- **Proportional residual error.** Takeuchi reports `sigma^2 = 0.0347`
  (NONMEM proportional `Y = IPRED * (1 + EPS)`, `EPS ~ N(0, sigma^2)`).
  In nlmixr2's `prop(propSd)` notation `propSd = sqrt(sigma^2) = 0.186`.

# Virtual cohort

The simulations below use a virtual cohort whose covariate distributions
approximate the Takeuchi 2023 Table 2 demographics. Subject-level observed
data are not publicly released.

```{r cohort}
set.seed(20260428)
n_subj <- 150

cohort <- tibble::tibble(
  id         = seq_len(n_subj),
  WT         = pmin(pmax(rlnorm(n_subj, log(56.65), 0.20), 35, 112)),
  SEXF       = rbinom(n_subj, 1, 0.76),
  CRCL       = pmin(pmax(rnorm(n_subj, mean = 88, sd = 20), 35, 174)),
  CONMED_MTX = rbinom(n_subj, 1, 0.74),
  ADA_POS    = rbinom(n_subj, 1, 0.37)
)
```

```{r events}
tau     <- 4 * 7 * 24       # q4w dosing interval, in hours (672 h)
n_doses <- 6                # 6 monthly doses ~= 5 half-lives, last cycle at SS
dose_h  <- seq(0, tau * (n_doses - 1), by = tau)

ss_start <- tau * (n_doses - 1)
ss_end   <- ss_start + tau

# Observation grid: dense early (first 2 cycles) for absorption-phase
# resolution and dense over the steady-state cycle for NCA. Coarser bridge
# in between to keep the event table small enough for a 5-minute render.
obs_h <- sort(unique(c(
  seq(0, 2 * tau, by = 24),                # daily through first two cycles
  dose_h + 1, dose_h + 6, dose_h + 24,     # near-dose absorption resolution
  seq(2 * tau, ss_start, by = 7 * 24),     # weekly bridge
  seq(ss_start, ss_end, by = 12)           # 12-hour grid for NCA cycle
)))

ev_dose <- cohort |>
  tidyr::crossing(time = dose_h) |>
  dplyr::mutate(amt = 30, cmt = "depot", evid = 1L)
ev_obs <- cohort |>
  tidyr::crossing(time = obs_h) |>
  dplyr::mutate(amt = 0, cmt = NA_character_, evid = 0L)

events <- dplyr::bind_rows(ev_dose, ev_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, amt, cmt, evid,
                WT, SEXF, CRCL, CONMED_MTX, ADA_POS)
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("Takeuchi_2023_ozoralizumab"))
keep_cols <- c("WT", "SEXF", "CRCL", "CONMED_MTX", "ADA_POS")

sim <- as.data.frame(rxode2::rxSolve(mod, events = events, keep = keep_cols))
```

# Replicate published figures

## Concentration-time profile over the dosing horizon

Trough-only sampling in the original trials precludes a direct dense
concentration-time replication of the source figures. The block below
shows the simulated 5/50/95th percentile profile over the 6-cycle dosing
horizon at the labelled 30 mg q4w regimen, summarised on a per-day grid.

```{r profile, fig.width = 7.5, fig.height = 4}
profile <- sim |>
  dplyr::filter(!is.na(Cc), time > 0) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(profile, aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20, fill = "steelblue") +
  geom_line(linewidth = 0.7, colour = "steelblue") +
  scale_y_log10() +
  labs(
    x = "Time since first dose (days)",
    y = "Ozoralizumab Cc (ug/mL)",
    title = "Simulated 5-50-95 percentile profile: 30 mg SC q4w",
    caption = paste0("Virtual RA cohort (N = ", n_subj, "); ", n_doses,
                     " monthly doses; final cycle at steady state.")
  ) +
  theme_minimal()
```

## Steady-state cycle

```{r ss-cycle, fig.width = 7, fig.height = 4}
ss_summary <- sim |>
  dplyr::filter(time >= ss_start, time <= ss_end, !is.na(Cc)) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(ss_summary, aes((time - ss_start) / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20, fill = "steelblue") +
  geom_line(linewidth = 0.7, colour = "steelblue") +
  labs(
    x = "Time within dosing interval (days)",
    y = "Ozoralizumab Cc (ug/mL)",
    title = "Steady-state q4w cycle (cycle 6 of 6)",
    caption = "Median and 90% prediction interval over the final 28-day interval."
  ) +
  theme_minimal()
```

# PKNCA validation

NCA on the steady-state q4w cycle (cycle 6). Time within the cycle is
re-zeroed so PKNCA computes per-interval Cmax, Cmin, AUC, and Cavg.

```{r pknca}
nca_conc <- sim |>
  dplyr::filter(time >= ss_start, time <= ss_end, !is.na(Cc)) |>
  dplyr::mutate(time_nom = time - ss_start, treatment = "30mg_Q4W") |>
  dplyr::select(id, time = time_nom, Cc, treatment)

nca_dose <- cohort |>
  dplyr::mutate(time = 0, amt = 30, treatment = "30mg_Q4W") |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start   = 0,
  end     = tau,
  cmax    = TRUE,
  cmin    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_summary <- summary(nca_res)
nca_summary
```

## Typical-patient steady-state exposures vs published Tables S1 / S2

Takeuchi 2023 Supplemental Tables S1 and S2 report median (95% prediction
interval) Cmax, AUCtau, and Ctrough at steady state for the typical patient
and for several covariate-perturbed sub-cohorts (1000 simulations each).
The block below repeats the typical-patient simulation by fixing IIV at
zero (`rxode2::zeroRe()`) and reading off the maximum, minimum, and
trapezoidal AUC over the final dosing interval. This is a deterministic
typical-patient check; it does **not** reproduce the 95% prediction
interval — that requires the stochastic block above.

```{r typical}
mod_typical <- mod |> rxode2::zeroRe()

ev_typ <- function(cov_row, dose_amt = 30) {
  ev_dose <- cov_row |>
    tidyr::crossing(time = dose_h) |>
    dplyr::mutate(amt = dose_amt, cmt = "depot", evid = 1L)
  obs_t <- sort(unique(c(seq(ss_start, ss_end, by = 6), dose_h)))
  ev_obs <- cov_row |>
    tidyr::crossing(time = obs_t) |>
    dplyr::mutate(amt = 0, cmt = NA_character_, evid = 0L)
  dplyr::bind_rows(ev_dose, ev_obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::select(id, time, amt, cmt, evid,
                  WT, SEXF, CRCL, CONMED_MTX, ADA_POS)
}

ss_metrics <- function(cov_row, label) {
  ev <- ev_typ(cov_row)
  out <- as.data.frame(rxode2::rxSolve(mod_typical, events = ev))
  ss <- out |>
    dplyr::filter(time >= ss_start, time <= ss_end, !is.na(Cc)) |>
    dplyr::arrange(time)
  tibble::tibble(
    scenario     = label,
    Cmax_sim     = max(ss$Cc),
    Ctrough_sim  = ss$Cc[which.max(ss$time)],
    AUCtau_sim   = sum(diff(ss$time) *
                       (head(ss$Cc, -1) + tail(ss$Cc, -1)) / 2)
  )
}

# Typical-patient covariate template:
# WT 56.7 kg, eGFR 86, female (SEXF=1), MTX yes (CONMED_MTX=1), ADA negative (=0)
typical_cov <- tibble::tibble(id = 1L, WT = 56.7, SEXF = 1L, CRCL = 86,
                              CONMED_MTX = 1L, ADA_POS = 0L)

scenarios <- dplyr::bind_rows(
  ss_metrics(typical_cov,                            "Typical patient"),
  ss_metrics(typical_cov |> dplyr::mutate(WT = 42.5),  "WT 42.5 kg"),
  ss_metrics(typical_cov |> dplyr::mutate(WT = 83.2),  "WT 83.2 kg"),
  ss_metrics(typical_cov |> dplyr::mutate(SEXF = 0L),  "Male"),
  ss_metrics(typical_cov |> dplyr::mutate(CRCL = 58),  "eGFR 58 mL/min/1.73 m^2"),
  ss_metrics(typical_cov |> dplyr::mutate(CRCL = 121), "eGFR 121 mL/min/1.73 m^2"),
  ss_metrics(typical_cov |> dplyr::mutate(CONMED_MTX = 0L), "Without MTX"),
  ss_metrics(typical_cov |> dplyr::mutate(ADA_POS = 1L),    "ADA-positive")
)

# Published medians from Takeuchi 2023 Supplemental Tables S1 / S2.
published <- tibble::tribble(
  ~scenario,                   ~Cmax_pub, ~AUCtau_pub, ~Ctrough_pub,
  "Typical patient",                7.28,        3230,         2.51,
  "WT 42.5 kg",                     8.93,        4110,         3.45,
  "WT 83.2 kg",                     5.59,        2320,         1.60,
  "Male",                           6.55,        2930,         2.31,
  "eGFR 58 mL/min/1.73 m^2",        7.60,        3430,         2.80,
  "eGFR 121 mL/min/1.73 m^2",       7.02,        3010,         2.24,
  "Without MTX",                    6.84,        2880,         2.06,
  "ADA-positive",                   6.95,        2980,         2.23
)

comparison <- dplyr::left_join(scenarios, published, by = "scenario") |>
  dplyr::mutate(
    Cmax_pct_diff    = 100 * (Cmax_sim    - Cmax_pub)    / Cmax_pub,
    AUCtau_pct_diff  = 100 * (AUCtau_sim  - AUCtau_pub)  / AUCtau_pub,
    Ctrough_pct_diff = 100 * (Ctrough_sim - Ctrough_pub) / Ctrough_pub
  ) |>
  dplyr::select(scenario,
                Cmax_pub, Cmax_sim, Cmax_pct_diff,
                AUCtau_pub, AUCtau_sim, AUCtau_pct_diff,
                Ctrough_pub, Ctrough_sim, Ctrough_pct_diff)

knitr::kable(
  comparison, digits = c(0, 2, 2, 1, 0, 0, 1, 2, 2, 1),
  caption = paste("Typical-patient (IIV zeroed) steady-state exposures vs.",
                  "Takeuchi 2023 Supplemental Tables S1 / S2 medians (1000",
                  "stochastic simulations). All scenarios at 30 mg SC q4w.")
)
```

The published values are medians from 1000 stochastic simulations with IIV
and RUV; the simulated values here are deterministic (typical-patient,
IIV zeroed). For a log-normal individual parameter the median equals the
typical value, so for Cmax / AUCtau / Ctrough at fixed covariates the two
quantities should agree to within numerical and SS-approach precision.

## Replicate Figure 3 forest plot

Takeuchi 2023 Figure 3 plots the ratio of CL/F, Vd/F, Cmax, and AUCtau in a
covariate-perturbed patient relative to the typical patient. The forest
points are deterministic (covariate effects on typical values), without
the stochastic 90% prediction interval that the source plot overlays.

```{r forest, fig.width = 7, fig.height = 4.5}
forest <- comparison |>
  dplyr::filter(scenario != "Typical patient") |>
  dplyr::mutate(
    Cmax_ratio    = Cmax_sim    / comparison$Cmax_sim[1],
    AUCtau_ratio  = AUCtau_sim  / comparison$AUCtau_sim[1],
    Ctrough_ratio = Ctrough_sim / comparison$Ctrough_sim[1]
  ) |>
  dplyr::select(scenario, Cmax_ratio, AUCtau_ratio, Ctrough_ratio) |>
  tidyr::pivot_longer(-scenario, names_to = "metric", values_to = "ratio") |>
  dplyr::mutate(
    metric = factor(metric,
                    levels = c("Cmax_ratio", "AUCtau_ratio", "Ctrough_ratio"),
                    labels = c("Cmax", "AUCtau", "Ctrough")),
    scenario = factor(scenario, levels = rev(unique(scenario)))
  )

ggplot(forest, aes(ratio, scenario)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_point(size = 2.4) +
  facet_wrap(~ metric, nrow = 1) +
  labs(
    x = "Ratio vs typical patient",
    y = NULL,
    title = "Steady-state exposure ratios at 30 mg q4w",
    caption = "Replicates Figure 3 of Takeuchi 2023 (point estimates only)."
  ) +
  theme_minimal()
```

# Errata

No errata, corrigenda, or correction notices were identified at extraction
time (Wiley Online Library landing page; PubMed correction-notice search;
Google Scholar). The full-model equations on p. 422 are internally
consistent with Table 3 (p. 421) and with Supplemental Tables S1-S3
under the typical-patient simulation above.

# Assumptions and deviations

- **Sex / MTX reference-category flips.** Takeuchi encodes Male = 1 with
  female-reference TVCL/TVVd, and MTX = 1 with on-MTX-reference TVCL.
  Stored under canonical `SEXF` (1 = female) and `CONMED_MTX` (1 = on MTX),
  the effects are applied as `(1 + e_male_* * (1 - SEXF))` and
  `(1 + e_nomtx_cl * (1 - CONMED_MTX))` so the published TVCL = 9.20 mL/h
  and TVVd = 4.91 L are preserved. The published coefficient signs and
  magnitudes are unchanged.
- **eGFR formula.** The covariate-column register's canonical `CRCL`
  reference values are documented for MDRD (US) and CKD-EPI conventions.
  Takeuchi uses the Japanese eGFR formula
  `194 * Scr^-1.094 * age^-0.287 * (0.739 if female)` with reference value
  85.95 mL/min/1.73 m^2. Users applying this model to a non-Japanese
  cohort should supply CRCL computed by the same Japanese formula or
  document the substitution.
- **Ka has no IIV.** `omega^2_Ka` was fixed to 0 in the published model
  because of high shrinkage. Ka is therefore the typical value for every
  simulated subject. This is faithful to the paper but means individual-
  level absorption variability is not captured.
- **Virtual-cohort covariate distributions.** The virtual cohort is
  constructed to match the published marginal Table 2 demographics, not
  joint distributions: WT lognormal centred at the population median
  (56.65 kg) and truncated to [35, 112] kg; SEXF Bernoulli(0.76); CRCL
  N(88, 20) truncated to [35, 174]; CONMED_MTX Bernoulli(0.74); ADA_POS
  Bernoulli(0.37). Covariate correlations (e.g., albumin / disease
  activity / age co-variation) are not modelled because Takeuchi did not
  report them.
- **No subject-level observed data.** OHZORA / NATSUZORA subject-level
  data are held by Taisho Pharmaceutical Co., Ltd. and not publicly
  released. The simulated comparison against Tables S1 / S2 substitutes
  for a true VPC.
- **Typical-patient comparison.** The published Tables S1 / S2 report
  medians from 1000 stochastic simulations with IIV and RUV; the
  comparison block uses a deterministic typical-patient simulation
  (`zeroRe()`). For the typical-patient log-normal parameterisation the
  median equals the typical value, so the comparison is appropriate;
  the 95% prediction interval the paper reports is not reproduced here.
