---
title: "Rivipansel (Tammara 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rivipansel (Tammara 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

The integrated population PK model published by Tammara and Harnisch
(2017) was used to inform the rivipansel phase III dose-selection
exercise in pediatric patients with sickle cell disease (SCD). The
model pools 109 subjects from three rivipansel phase I studies and the
phase II study (NCT01119833) in adolescents and adults hospitalized
for vaso-occlusive crisis (VOC). It is a three-compartment IV linear
disposition model with creatinine clearance and body weight as the
only structural covariates.

* Citation: Tammara BK, Harnisch LO. Dose Selection Based on Modeling
  and Simulation for Rivipansel in Pediatric Patients Aged 6 to 11
  Years With Sickle Cell Disease. CPT Pharmacometrics Syst Pharmacol.
  2017;6(12):845-854.
* Article: <https://doi.org/10.1002/psp4.12263>

# Population

The integrated dataset pooled 109 subjects aged 12-51 years across
four studies: rivipansel study 101 (single IV dose 2-40 mg/kg in
healthy adults, n = 40), study 102 (multiple IV doses 2-20 mg/kg q8h
plus a 40 mg/kg loading dose followed by 20 mg/kg q8h in healthy
adults, n = 32), study 103 / NCT00911495 (two IV doses 20 mg/kg
loading + 10 mg/kg in adults with SCD not in VOC, n = 15), and the
phase II study NCT01119833 (Telen 2015) in patients with SCD aged
12-60 years hospitalized for VOC, n = 76 (56 adults and 20 children
aged 12-17 years). Rivipansel is renally cleared with about 60%
protein binding and a low volume of distribution; phase I half-life
is approximately 7-8 hours.

The same population metadata is available programmatically:

```{r meta, eval = FALSE}
readModelDb("Tammara_2017_rivipansel")$population
```

# Source trace

Every `ini()` value carries a trailing comment in
`inst/modeldb/specificDrugs/Tammara_2017_rivipansel.R` pointing to its
source location. The table below collects them for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Three-compartment IV linear disposition with central + peripheral1 + peripheral2 | -- | Methods, "Study design and data sources" |
| Typical CL = 1.25 * (CRCL/150)^0.468 * (1 + 0.234 * STUD) | -- | Table 1 footnote b |
| Typical V1 = 6.24 * (WT/70)^0.569 | -- | Table 1 footnote c |
| Typical V2 = 4.02 * (WT/70)^0.569 | -- | Table 1 footnote c |
| Typical V3 = 0.656 * (WT/70)^0.569 | -- | Table 1 footnote c |
| `lcl` (CL, L/h) | 1.25 | Table 1, CL row |
| `lvc` (V1, L) | 6.24 | Table 1, V1 row |
| `lq` (Q_rapid, L/h) | 2.62 | Table 1, Q_rapid row |
| `lvp` (V2, L) | 4.02 | Table 1, V2 row |
| `lq2` (Q_slow, L/h) | 0.0316 | Table 1, Q_slow row |
| `lvp2` (V3, L) | 0.656 | Table 1, V3 row |
| `e_study_riv201_cl` (additive SCD shift, fraction) | 0.234 | Table 1, "Study effect on CL" row |
| `e_crcl_cl` (CRCL exponent on CL) | 0.468 | Table 1, "Exponent for CRCL on CL" row |
| `e_wt_vc_vp_vp2` (shared WT exponent) | 0.569 | Table 1, "Exponent for WT on V1, V2, V3" row |
| IIV %CV on CL / V1 / Q_rapid / V2 / Q_slow / V3 | 18.8 / 24.1 / 19.8 / 15.5 / 15.5 / 15.3 | Table 1, interindividual variability rows |
| Additive residual SD (phase I / study 201) | 0.29 / 0.165 ug/mL | Table 1, residual variability rows |
| Proportional residual %CV (phase I / study 201) | 9.52 / 23.2 | Table 1, residual variability rows |

# Virtual cohort

The original observed data are not publicly available. The cohort
below approximates the phase II SCD population, with body weight and
creatinine clearance distributions drawn from typical SCD-cohort
ranges and `STUDY_RIV201` set to 1 for every subject (the simulation
target is the SCD population for which the dose-selection exercise
applies).

```{r cohort}
set.seed(12263)  # DOI suffix
n_subj <- 100

# Body weight (kg). Phase II included adolescents and adults; weight
# distribution sampled to span a typical adolescent-adult SCD range
# (Tammara 2017 Figure 2; observed phase II weights ranged roughly
# 40-110 kg with median near 70 kg).
WT <- pmin(pmax(rnorm(n_subj, mean = 70, sd = 15), 45), 110)

# Creatinine clearance (mL/min, NOT BSA-normalized; raw Cockcroft-Gault
# scale per Tammara 2017). SCD hyperfiltration shifts the typical CRCL
# in adults upward of the 120 mL/min anchor; the simulated distribution
# centres on 130 mL/min and spans 60-200 mL/min, matching the observed
# range in Tammara 2017 Figure 4 (adult panel).
CRCL <- pmin(pmax(rnorm(n_subj, mean = 130, sd = 30), 60), 200)

# Phase II SCD study indicator: 1 for every subject in the target
# population (the simulation context is the SCD cohort).
STUDY_RIV201 <- rep(1L, n_subj)

cohort <- tibble(
  id           = seq_len(n_subj),
  WT           = WT,
  CRCL         = CRCL,
  STUDY_RIV201 = STUDY_RIV201
)
```

# Simulation

Rivipansel is administered as a 20-minute IV infusion. The
dose-selection regimen recommended by Tammara 2017 for phase III is a
40 mg/kg loading dose followed by 20 mg/kg every 12 hours; we simulate
that regimen out to seven days so steady state is reached.

```{r events}
infusion_dur <- 20 / 60  # 20 minutes expressed in hours
tau          <- 12       # dosing interval (hours)
maint_doses  <- 13       # 13 maintenance doses after loading -> 14 doses, 7 days

# Per-subject dosing: loading dose at time 0, then maintenance doses
# at 12-hour intervals. amt is in mg (weight-based mg/kg dose times
# subject weight); rate is amt / 20-min infusion duration.
make_doses <- function(row) {
  load_amt  <- 40 * row$WT
  maint_amt <- 20 * row$WT
  dose_times <- c(0, tau * seq_len(maint_doses))
  amt        <- c(load_amt, rep(maint_amt, maint_doses))
  data.frame(
    id           = row$id,
    time         = dose_times,
    amt          = amt,
    rate         = amt / infusion_dur,
    evid         = 1L,
    cmt          = "central",
    WT           = row$WT,
    CRCL         = row$CRCL,
    STUDY_RIV201 = row$STUDY_RIV201
  )
}

doses <- cohort |>
  split(seq_len(nrow(cohort))) |>
  lapply(make_doses) |>
  bind_rows()

# Observation grid: every 0.5 h for the first 24 h to characterise the
# loading-dose phase, then every 2 h out to 168 h (7 days). The final
# dosing interval (156-168 h) is the steady-state window used for NCA.
obs_times <- sort(unique(c(
  seq(0, 24, by = 0.5),
  seq(24, 168, by = 2)
)))
obs <- expand_grid(
  id = cohort$id,
  time = obs_times
) |>
  mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central") |>
  left_join(cohort, by = "id")

events <- bind_rows(doses, obs) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
mod <- readModelDb("Tammara_2017_rivipansel")
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("WT", "CRCL", "STUDY_RIV201")) |>
  as.data.frame() |>
  filter(!is.na(Cc))
```

# Replicate published figures

```{r figure-5-like, fig.width = 7, fig.height = 4.5}
# Approximate Tammara 2017 Figure 5 (adult panel column on the 40/20
# mg/kg regimen): predicted concentration-time trajectory in adult SCD
# subjects on the recommended regimen. Median and 5th/95th prediction
# intervals over 168 h.
sim |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Rivipansel concentration (ug/mL)",
    title = "Adult SCD prediction interval on 40/20 mg/kg q12h",
    caption = paste(
      "Replicates the adult panel of Tammara 2017 Figure 5 (rightmost",
      "row; 40/20 mg/kg regimen). Dashed line: 10 ug/mL preclinical",
      "efficacy threshold."
    )
  )
```

```{r figure-6-like, fig.width = 6.5, fig.height = 4}
# Approximate Tammara 2017 Figure 6 (high-dose panel for adults): the
# Cavg,ss distribution in adults on the 20 mg/kg q12h maintenance
# regimen. Computed analytically from each subject's individual CL
# rather than from the simulated concentration trace (per the paper:
# "Cavg,ss simulations did not contain residual error as they reflect
# a model-based summary measure across the whole dosing interval").
mod_typical <- mod  # individual CL from simulation already includes IIV
cavg_ss <- sim |>
  group_by(id) |>
  slice(1) |>
  ungroup() |>
  mutate(
    maint_amt = 20 * WT,
    cl_i      = exp(log(1.25)) * (CRCL / 150)^0.468 *
                  (1 + 0.234 * STUDY_RIV201),
    Cavg_ss   = (maint_amt / tau) / cl_i
  )

ggplot(cavg_ss, aes(x = "20 mg/kg q12h", y = Cavg_ss)) +
  geom_boxplot(fill = "darkseagreen", width = 0.4, outlier.alpha = 0.4) +
  geom_hline(yintercept = 10, linetype = "dashed", colour = "red") +
  labs(
    x = NULL,
    y = "Cavg,ss (ug/mL)",
    title = "Adult Cavg,ss on the recommended 40/20 mg/kg regimen",
    caption = paste(
      "Replicates the adult-panel portion of Tammara 2017 Figure 6",
      "(right panel, 40/20 mg/kg). Dashed line: 10 ug/mL preclinical",
      "efficacy threshold."
    )
  )
```

# PKNCA validation

Validation against the published Cavg,ss summary (Figure 6) and the
phase I terminal half-life (Methods narrative: mean phase I half-life
7-8 hours). The steady-state NCA window is the final dosing interval
(156-168 h) at the 20 mg/kg q12h maintenance dose; results are
grouped by the single treatment label.

```{r pknca}
sim_nca <- sim |>
  mutate(treatment = "40/20 mg/kg q12h") |>
  select(id, time, Cc, treatment)

dose_df <- events |>
  filter(evid == 1) |>
  mutate(treatment = "40/20 mg/kg q12h") |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | treatment + id,
  concu = "ug/mL", timeu = "hour"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id,
  doseu = "mg"
)

start_ss <- max(dose_df$time)
end_ss   <- start_ss + tau

intervals <- data.frame(
  start    = c(0,        start_ss),
  end      = c(tau,      end_ss),
  cmax     = c(TRUE,     TRUE),
  tmax     = c(TRUE,     TRUE),
  cmin     = c(FALSE,    TRUE),
  auclast  = c(TRUE,     TRUE),
  cav      = c(FALSE,    TRUE),
  ctrough  = c(FALSE,    TRUE)
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

nca_tbl <- as.data.frame(nca_res$result)
nca_summary <- nca_tbl |>
  group_by(PPTESTCD, start, end) |>
  summarise(
    median = median(PPORRES, na.rm = TRUE),
    q05    = quantile(PPORRES, 0.05, na.rm = TRUE),
    q95    = quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(start, PPTESTCD)

knitr::kable(
  nca_summary,
  digits = 3,
  caption = paste(
    "Simulated NCA parameters for the 40/20 mg/kg q12h regimen.",
    "Start = 0 covers the loading-dose interval; start =", start_ss,
    "covers the steady-state interval (last maintenance dose +",
    tau, "h)."
  )
)
```

## Terminal half-life from a single-dose simulation

The paper reports a mean phase I terminal half-life of 7-8 hours. To
verify the model reproduces that value, simulate a single 40 mg/kg
dose in a typical 70-kg phase I subject (`STUDY_RIV201 = 0`,
`CRCL = 100`) and compute the terminal half-life via PKNCA.

```{r halflife}
typ <- tibble(
  id           = 1L,
  WT           = 70,
  CRCL         = 100,
  STUDY_RIV201 = 0L
)

typ_events <- bind_rows(
  data.frame(
    id   = 1L, time = 0, amt = 40 * 70,
    rate = (40 * 70) / infusion_dur, evid = 1L, cmt = "central",
    WT = 70, CRCL = 100, STUDY_RIV201 = 0L
  ),
  data.frame(
    id   = 1L, time = seq(0, 72, by = 0.25),
    amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central",
    WT = 70, CRCL = 100, STUDY_RIV201 = 0L
  )
) |>
  arrange(time, desc(evid))

mod_zero <- rxode2::zeroRe(mod)
typ_sim <- rxode2::rxSolve(mod_zero, events = typ_events,
                           keep = c("WT", "CRCL", "STUDY_RIV201")) |>
  as.data.frame() |>
  filter(!is.na(Cc), time > 0) |>
  mutate(id = 1L, treatment = "single 40 mg/kg")

conc_obj_typ <- PKNCA::PKNCAconc(
  typ_sim |> select(id, time, Cc, treatment),
  Cc ~ time | treatment + id,
  concu = "ug/mL", timeu = "hour"
)
dose_obj_typ <- PKNCA::PKNCAdose(
  data.frame(id = 1L, time = 0, amt = 40 * 70,
             treatment = "single 40 mg/kg"),
  amt ~ time | treatment + id,
  doseu = "mg"
)
intervals_typ <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)
nca_typ <- suppressWarnings(PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj_typ, dose_obj_typ, intervals = intervals_typ)
))
typ_tbl <- as.data.frame(nca_typ$result) |>
  select(PPTESTCD, PPORRES)
knitr::kable(
  typ_tbl,
  digits = 3,
  caption = "Typical-value NCA after a single 40 mg/kg IV infusion in a phase I 70-kg subject (CRCL = 100 mL/min)."
)
```

The published phase I mean half-life of 7-8 hours is the comparator
against the `half.life` row above.

# Assumptions and deviations

- The phase II SCD residual error magnitudes (`addSd_study201`,
  `propSd_study201`) and the phase I magnitudes (`addSd_phase1`,
  `propSd_phase1`) are estimated separately in Tammara 2017 Table 1.
  The model file selects between them per observation via the
  `STUDY_RIV201` indicator. Simulations targeting the SCD population
  use `STUDY_RIV201 = 1`, which selects `propSd = 0.232`,
  `addSd = 0.165 ug/mL`.
- Tammara 2017 expresses CRCL in raw mL/min (Cockcroft-Gault for
  adults, Schwartz with BSA adjustment for children to give an
  absolute mL/min value), not normalised to 1.73 m^2. The canonical
  `CRCL` covariate accepts both conventions; this model file's
  `covariateData[[CRCL]]$units` is `"mL/min"` to flag the
  non-BSA-normalised use.
- Body weight and CRCL distributions for the virtual cohort are
  approximated from the figures in the paper (Figure 2 for weight,
  Figure 4 for creatinine clearance in adults). The exact baseline
  demographics table for the integrated dataset is not reproduced in
  the publication.
- The Cavg,ss distribution figure uses the closed-form
  Cavg,ss = (dose / tau) / CL_i per subject, matching the paper's
  approach ("Cavg,ss simulations did not contain residual error as
  they reflect a model-based summary measure across the whole
  dosing interval").
- The model does not encode the pediatric (6-11 year) demographic
  extrapolation from the CDC growth charts and the Hoste creatinine
  relationship that the paper develops for its simulation exercise.
  Users extrapolating to younger pediatric populations must supply
  the appropriate WT, CRCL covariate values themselves; the
  structural model and the SCD hyperfiltration effect are assumed by
  the paper to apply unchanged in younger children.
- No `eta` correlation structure was reported in Table 1; the six
  IIV terms are encoded as independent log-normal random effects.
