---
title: "Cabozantinib (Tan 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cabozantinib (Tan 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta}
ui <- rxode2::rxode(readModelDb("Tan_2024_cabozantinib"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1007/s40262-024-01379-y>
* Supplement (NONMEM control streams, Tables S1-S3):
  <https://doi.org/10.1007/s40262-024-01379-y> (Supplementary Information,
  `40262_2024_1379_MOESM1_ESM.docx`)

Tan et al. (2024) took the cabozantinib population PK model from the FDA
registration file, evaluated it against a real-world therapeutic-drug-monitoring
(TDM) cohort of 27 patients with metastatic renal cell carcinoma (mRCC) treated
at Leiden University Medical Center, re-estimated the two parameters that did
not transfer (apparent clearance and the proportional residual error), and then
used the resulting model to design dosing regimens that cut cabozantinib drug
expenses without changing exposure. Cabozantinib is flat-priced -- the 20 mg,
40 mg and 60 mg tablets cost the same -- so a regimen that reaches the same
exposure from fewer tablets is directly a cost saving.

# Population

The model was evaluated and partially re-estimated on 27 mRCC patients
contributing 75 TDM observations, treated between August 2018 and December 2021
(Tan 2024 Table 1). Mean age was 65 years (range 39-85), mean weight 78 kg
(range 49-105), and 70.3% were male. Mean baseline creatinine clearance was
70 mL/min/1.73m^2 (CKD-EPI) and mean baseline ALT 46 U/L. Patients were mostly
pre-treated (pazopanib 42.9%, nivolumab with or without ipilimumab 34.3%,
sunitinib 11.4%); 8.5% were treatment-naive. IMDC prognosis was intermediate in
63.0%. The median administered dose was 40 mg once daily (range 20-60) over a
median of 75 treatment days (range 11-552). Median observed concentration was
603 ng/mL (range 135-1471) and median trough 632 ng/mL (range 308-1134).

Two points about this population matter when reading the model. First, the
27-patient cohort is far too small to have supported a model of this complexity
on its own: every parameter other than CL/F and the residual error is inherited
from the registration analysis, which pooled 63 healthy participants from a
phase I study with 325 mRCC patients from a phase III study. Second, ethnicity
was not recorded in the TDM dataset, so Tan 2024 assumed all patients were
Caucasian (section 2.4.2) and the NONMEM control stream hardcodes `RACE = 1`.
The Asian-race effect on CL/F is therefore carried in the model as published but
was never exercised by this cohort.

The full metadata is available programmatically:

```{r population}
str(ui$population)
```

# Model structure

Absorption is the distinctive feature: the dose splits between two parallel
first-order depots, each with its own lag time (Tan 2024 Fig. 1). A fraction
`F1 = 0.675` enters a fast depot (`ka1 = 0.568 /h`, lag 0.459 h) and the
remaining 32.5% enters a slow depot (`ka2 = 0.102 /h`, lag 16.8 h). Disposition
is two-compartment with linear elimination. The fast absorption rate constant
falls with dose as `(DOSE / 60 mg)^-0.5`.

Because the two depots are dosed as a pair, an event table for this model must
place a dose record on **both** `depot` and `depot2` with the same amount; the
`f()` multipliers in the model split it into the `F1` and `1 - F1` fractions,
reproducing NONMEM's `F1` / `F2 = 1 - F1` bioavailability pair.

```{r structure}
cat(paste(deparse(ui$model), collapse = "\n"))
```

# Source trace

Every `ini()` entry in `inst/modeldb/specificDrugs/Tan_2024_cabozantinib.R`
carries an in-file comment naming its origin. The table collects them for
review. "Table 2 final" is the *Estimates of the final POPPK model* column of
Tan 2024 Table 2; "CS" is the second NONMEM control stream in the Supplementary
Material (*NONMEM CONTROL STREAM SIMULATION WITH THE FINAL CABOZANTINIB PK
MODEL*). Only the two rows marked **estimated** were fitted by Tan 2024; the
rest are fixed.

| Parameter | Value | Fixed? | Source location |
|-----------|-------|--------|-----------------|
| `lcl` | CL/F = 3.11 L/h | **estimated** (RSE 5%) | Table 2 final; bootstrap median 3.11, 95% CI 2.73-3.58 (section 3.3.2) |
| `lvc` | Vc/F = 81.5 L | fixed | Table 2 final; CS `$THETA (81.5) FIX ; V2` |
| `lvp` | Vp/F = 213 L | fixed | Table 2 final; CS `$THETA (213) FIX ; V3` |
| `lq` | Q/F = 14.2 L/h | fixed | Table 2 final; CS `$THETA (14.2) FIX ; Q` |
| `lka` | ka1 = 0.568 /h | fixed | Table 2 final; CS `$THETA (0.568) FIX ; KA1` |
| `lka2` | ka2 = 0.102 /h | fixed | Table 2 final; CS `$THETA (0.102) FIX ; KA2` |
| `ltlag` | ALAG1 = 0.459 h | fixed | Table 2 final; CS `$THETA (0.459) FIX ; ALAG1` |
| `ltlag2` | ALAG2 = 16.8 h | fixed | Table 2 final; CS `$THETA (16.8) FIX ; ALAG2` |
| `lffo` | F1 = 0.675 | fixed | Table 2 final; CS `$THETA (0.675) FIX ; F1` |
| `lfdepot` | F = 1 (anchor) | fixed | Neither CS carries an F term; apparent parameters absorb F |
| `e_dose_ka` | -0.5 | fixed | Table 2 "Dose exponent on Ka1"; CS `$PK KA1 = TVKA1*EXP(ETA(3))*EXP(ALPHA*LOG(DOS/60))` |
| `e_sexf_cl` | -0.21 (multiplier 0.79) | fixed | Table 2 "Female on CL/F" = 0.79; section 2.4.1 "21% lower CL/F"; CS `$THETA (0.21) FIX ; Gender` |
| `e_race_asian_cl` | -0.27 (multiplier 0.73) | fixed | Table 2 "Asian on CL/F" = 0.73; section 2.4.1 "27% lower CL/F" |
| `e_fed_highfat_f` | +0.5 | fixed (assumption) | Section 2.6 "the bioavailability of the drug will increase by 50%" |
| `etalcl`, `etalvc` block | 0.213 / 0.44 / 1.06 | fixed | Table 2 Omega2_CL, Omega2_CL:Vc, Omega2_Vc; CS `$OMEGA BLOCK(2) FIX` |
| `etalka` | 0.437 | fixed | Table 2 Omega2_Ka; CS `$OMEGA 0.437 FIX ; KA1` |
| `etalffo` | 0.05 | fixed | Table 2 final Omega2_F1 (registration value 0.385); CS `$OMEGA 0.05 FIX ; F1`; section 3.2 |
| `propSd` | 0.335 | **estimated** (RSE 10%) | Table 2 final "Residual error"; CS `$THETA (0.335) FIX ; Prop.RE (sd)` |
| `addSd` | 0.001 ng/mL | fixed | CS `$THETA (0.001) FIX ; Add.RE (sd)`; not listed in Table 2 |

The control stream's `$ERROR` block is
`W = SQRT(THETA(10)**2 + (THETA(11)*IPRED)**2)` with `$SIGMA 1 FIX`, i.e. a
combined additive-plus-proportional error whose variances add. That is
nlmixr2's default `combined2` parameterisation for `prop() + add()`, so the two
thetas map across directly as standard deviations.

# Regimens

Tan 2024 Fig. 2 defines five simulated regimens, compared in two pairs against
two references. All doses are oral tablets.

```{r regimens}
regimens <- tibble::tribble(
  ~regimen,                ~dose_mg, ~pattern,  ~fed, ~reference,
  "40 mg QD fasted",             40, "qd",         0, "40 mg QD fasted",
  "60 mg 2d+1skip fasted",       60, "skip",       0, "40 mg QD fasted",
  "40 mg 2d+1skip fed",          40, "skip",       1, "40 mg QD fasted",
  "20 mg QD fasted",             20, "qd",         0, "20 mg QD fasted",
  "40 mg q72h fed",              40, "q72h",       1, "20 mg QD fasted"
)
knitr::kable(regimens, caption = "Simulated regimens (Tan 2024 Fig. 2).")
```

Every regimen delivers the same amount of drug per 72 h as its reference once
the assumed 50% food effect is applied: 3 x 40 mg = 120 mg for the fasted
reference; 2 x 60 mg = 120 mg for the skip-day regimen; 2 x 40 mg x 1.5 =
120 mg-equivalent for the fed skip-day regimen; and 40 mg x 1.5 = 60 mg
equivalent against 3 x 20 mg = 60 mg for the second pair. That equality is the
whole basis of the paper's cost argument, and it is what the closed-form gate
below tests.

```{r dose-times}
# Dosing schedules over the 936 h simulated horizon. The 2-days-on / 1-day-skip
# pattern has period 72 h with doses at t and t + 24 within each cycle.
horizon <- 936
dose_times <- function(pattern) {
  switch(pattern,
    qd   = seq(0, horizon, by = 24),
    skip = sort(c(seq(0, horizon, by = 72), seq(24, horizon, by = 72))),
    q72h = seq(0, horizon, by = 72)
  )
}

# The paper's assessment window: a 72 h interval at steady state.
ss_start <- 864
ss_end   <- 936
stopifnot(ss_end - ss_start == 72, ss_start %% 72 == 0, ss_start %% 24 == 0)
```

At 864 h the model is at steady state for practical purposes: the terminal
half-life implied by the fixed disposition parameters is

```{r thalf}
cl <- 3.11; vc <- 81.5; vp <- 213; q <- 14.2
k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
s <- k10 + k12 + k21
beta <- 0.5 * (s - sqrt(s^2 - 4 * k10 * k21))
t_half_terminal <- log(2) / beta
t_half_terminal
```

so 864 h is `r round(864 / t_half_terminal, 1)` terminal half-lives of
accumulation. (With the registration-file CL/F of 2.23 L/h the same calculation
gives 99.3 h, reproducing the "extremely long terminal half-life, around 99 h"
quoted throughout the paper.)

```{r thalf-check}
k10_reg <- 2.23 / vc
s_reg <- k10_reg + k12 + k21
beta_reg <- 0.5 * (s_reg - sqrt(s_reg^2 - 4 * k10_reg * k21))
# Paper abstract / section 1 / section 4: "around 99 h", "~ 99 h".
stopifnot(abs(log(2) / beta_reg - 99) < 1.5)
```

# Event-table construction

```{r make-events}
# A dose record goes on BOTH depots with the same amount; the model's f()
# multipliers split it into F1 and 1 - F1. Observations sit on `central`, the
# ODE state -- never on `Cc`, which is an algebraic observable.
make_arm <- function(dose_mg, pattern, fed, n_sub, id_offset = 0L,
                     obs_times = seq(ss_start, ss_end, by = 0.25)) {
  dt <- dose_times(pattern)
  ev <- rxode2::et(amt = dose_mg, time = dt, cmt = "depot") |>
    rxode2::et(amt = dose_mg, time = dt, cmt = "depot2") |>
    rxode2::et(obs_times, cmt = "central")
  one <- as.data.frame(ev)
  one$DOSE        <- dose_mg
  one$FED_HIGHFAT <- fed
  one$SEXF        <- 0    # male reference: the control stream simulates GEND = 1
  one$RACE_ASIAN  <- 0    # section 2.4.2: all subjects assumed non-Asian
  out <- one[rep(seq_len(nrow(one)), n_sub), , drop = FALSE]
  out$id <- rep(seq_len(n_sub) + id_offset, each = nrow(one))
  out
}
```

# Closed-form validation of steady-state exposure

For a linear model at steady state the AUC over one full dosing cycle is fixed
by mass balance alone: every milligram that is absorbed must be cleared, so

$$\mathrm{AUC}_{\tau} = \frac{F \cdot D_{\tau}}{\mathrm{CL}/F}$$

independent of `ka1`, `ka2`, the lag times, the depot split, and the peripheral
compartment. This is the sharpest available check on the disposition
parameters and the unit scaling, and because both sides use the same parameter
values it is a pure numerical-accuracy comparison -- a tight bound is the
correct assertion here.

```{r closed-form}
closed_form <- function(dose_mg, pattern, fed) {
  dt <- dose_times(pattern)
  n_in_window <- sum(dt >= ss_start & dt < ss_end)
  dose_per_cycle <- n_in_window * dose_mg * (1 + 0.5 * fed)
  dose_per_cycle * 1000 / 3.11   # mg -> ng, / (L/h) -> ng*h/mL
}

trap_auc <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

typ <- lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  ev <- make_arm(r$dose_mg, r$pattern, r$fed, n_sub = 1L)
  s <- rxode2::rxSolve(
    rxode2::zeroRe(ui), ev,
    returnType = "data.frame", addDosing = FALSE, maxsteps = 500000L
  )
  s <- s[s$time >= ss_start & s$time <= ss_end, ]
  s <- s[order(s$time), ]
  data.frame(
    regimen   = r$regimen,
    auc_sim   = trap_auc(s$time, s$Cc),
    auc_exact = closed_form(r$dose_mg, r$pattern, r$fed)
  )
}) |> bind_rows() |>
  mutate(pct_diff = 100 * (auc_sim - auc_exact) / auc_exact)

knitr::kable(
  typ |> rename("Regimen" = regimen,
                "Simulated AUC (ng*h/mL)" = auc_sim,
                "Dose / (CL/F) (ng*h/mL)" = auc_exact,
                "% diff" = pct_diff),
  digits = c(0, 0, 0, 3),
  caption = "Typical-value AUC over the 864-936 h interval against the closed form Dose/(CL/F)."
)
```

```{r closed-form-gate}
# Deterministic: same parameters on both sides, so the only difference is
# trapezoidal error on a 0.25 h grid plus the residual approach to steady
# state (864 h is ~11.8 terminal half-lives). Realised max |diff| 0.09%.
# A mis-transcribed CL/F, volume, dose or the ng/mL scaling moves this by
# tens of percent, so 0.5% is a gate that can still go red.
stopifnot(max(abs(typ$pct_diff)) < 0.5)
```

All five regimens land on the closed form, and -- as the design intends -- the
three 120 mg-equivalent regimens share one AUC and the two 60 mg-equivalent
regimens share another.

# Virtual cohort simulation

```{r cohort}
n_per_arm <- 200   # per-arm cap; the paper used 1000 NONMEM subproblems

sim_list <- lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  # Re-seed before EVERY arm, with the seed from the paper's own $SIMULATION
  # record. Because each arm draws the same number of subjects, this makes the
  # five arms share one set of individual etas -- common random numbers. The
  # comparison the paper makes is a within-subject one (does this patient get
  # the same exposure on the cheaper regimen?), so pairing the arms is both the
  # scientifically correct contrast and a large variance reduction: it removes
  # the between-arm sampling noise that would otherwise dominate the AUC ratios
  # below. The `ffo`-tail and Table 3 comparisons are then computed over 200
  # distinct subjects rather than 1000.
  rxode2::rxSetSeed(19950705)
  ev <- make_arm(r$dose_mg, r$pattern, r$fed, n_sub = n_per_arm,
                 id_offset = (i - 1L) * n_per_arm)
  s <- rxode2::rxSolve(
    ui, ev, returnType = "data.frame", addDosing = FALSE,
    maxsteps = 500000L, keep = c("DOSE", "FED_HIGHFAT")
  )
  s$regimen <- r$regimen
  s
})
sim <- bind_rows(sim_list)
nrow(sim)
```

```{r cohort-sanity}
# Every arm solved, every subject present, no NA concentrations.
stopifnot(
  !anyNA(sim$Cc),
  dplyr::n_distinct(sim$regimen) == nrow(regimens),
  all(table(unique(sim[c("regimen", "id")])$regimen) == n_per_arm)
)

# Common random numbers actually took: the k-th subject of every arm must carry
# the same individual parameters. Asserted rather than assumed, because the
# paired AUC-ratio gate further down is only meaningful if it held.
#
# `ka` is the exception -- it is dose-dependent by construction,
# ka = ka_ref * (DOSE/60)^-0.5 -- so it is compared after dividing the dose
# term back out. That makes this assertion do double duty: it also confirms the
# dose-power covariate is applied with the published exponent, since any other
# exponent leaves a residual spread across the 20/40/60 mg arms.
etas <- sim |>
  distinct(regimen, id, cl, vc, ka, ffo, DOSE) |>
  mutate(ka_ref = ka * (DOSE / 60)^0.5) |>
  arrange(regimen, id) |>
  group_by(regimen) |>
  mutate(k = row_number()) |>
  ungroup()
eta_spread <- etas |>
  group_by(k) |>
  summarise(across(c(cl, vc, ka_ref, ffo), ~ diff(range(.x))), .groups = "drop")
stopifnot(max(as.matrix(eta_spread[, -1])) < 1e-10)
```

## The absorption fraction can exceed 1

The control stream gives `F1` an exponential random effect,
`F1 = 0.675 * exp(eta)`, so `F1` crosses 1 whenever
`eta > log(1 / 0.675) = 0.393` -- and then the slow-depot fraction `1 - F1`
goes negative. This is the numerical failure Tan 2024 section 3.2 reports:
at the registration-file variance of 0.385 the runs terminated, and the authors
shrank the variance to 0.05. Shrinking it does not remove the tail, only
narrows it.

```{r ffo-tail}
sd_eta <- sqrt(0.05)
frac_over_1 <- function(omega2) {
  pnorm(log(1 / 0.675), mean = 0, sd = sqrt(omega2), lower.tail = FALSE)
}
tibble::tibble(
  `Omega^2 on F1`          = c(0.385, 0.05),
  `Source`                 = c("FDA registration file", "Tan 2024 final model"),
  `P(F1 > 1)`              = round(c(frac_over_1(0.385), frac_over_1(0.05)), 4)
) |> knitr::kable(caption = "Probability that the fast-depot fraction exceeds 1.")

# All five arms share one set of subjects (common random numbers), so the tail
# is a property of the 200 distinct subjects, not of 1000 arm-subject rows.
observed_over_1 <- sim |>
  filter(regimen == regimens$regimen[1]) |>
  distinct(id, ffo) |>
  summarise(pct = 100 * mean(ffo > 1)) |>
  pull(pct)
observed_over_1
```

```{r ffo-tail-gate}
# The analytic tail probability at omega^2 = 0.05 is 3.9%; at n = 200 the
# binomial standard error is 1.4 points. The bound is wide enough for the
# sampling noise of any cohort the model can produce, and still red if the eta
# were dropped (0%) or left at the registration value 0.385 (26%).
stopifnot(observed_over_1 > 0.5, observed_over_1 < 12)
```

Subjects in that tail get a negative amount deposited in the slow depot. Total
bioavailable dose is unaffected -- `F1 + (1 - F1) = 1` regardless -- so
steady-state AUC is untouched, but the concentration-time *shape* for those
subjects is not physiological. This is a property of the published model, faithfully
reproduced here rather than patched; see Errata.

## Replicating Figure 4

```{r figure4, fig.height = 7}
plot_pairs <- tibble::tribble(
  ~panel, ~arms,
  "a: 40 mg QD fasted vs 60 mg 2d+1skip fasted",
  c("40 mg QD fasted", "60 mg 2d+1skip fasted"),
  "b: 40 mg QD fasted vs 40 mg 2d+1skip fed",
  c("40 mg QD fasted", "40 mg 2d+1skip fed"),
  "c: 20 mg QD fasted vs 40 mg q72h fed",
  c("20 mg QD fasted", "40 mg q72h fed")
)

pi_bands <- sim |>
  group_by(regimen, time) |>
  summarise(
    med = median(Cc),
    lo  = quantile(Cc, 0.05),
    hi  = quantile(Cc, 0.95),
    .groups = "drop"
  )

fig4 <- lapply(seq_len(nrow(plot_pairs)), function(i) {
  pi_bands |>
    filter(regimen %in% plot_pairs$arms[[i]]) |>
    mutate(panel = plot_pairs$panel[i])
}) |> bind_rows()

ggplot(fig4, aes(time - ss_start, med, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(
    x = "Time within the 72 h steady-state interval (h)",
    y = "Cabozantinib concentration (ng/mL)",
    colour = NULL, fill = NULL
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Replicates Figure 4 of Tan 2024 (median with 90% prediction interval). Panel a
shows the higher peak and deeper trough of the skip-day regimen against the
flatter 40 mg QD profile; panels b and c add the assumed food effect.

# Exposure metrics against Tan 2024 Table 3

```{r metrics}
metrics <- sim |>
  arrange(regimen, id, time) |>
  group_by(regimen, id) |>
  summarise(
    cmin = min(Cc),
    cmax = max(Cc),
    auc  = trap_auc(time, Cc),
    .groups = "drop"
  ) |>
  mutate(cavg = auc / 72)

summary_tbl <- metrics |>
  group_by(regimen) |>
  summarise(
    cmin = mean(cmin), cmax = mean(cmax),
    auc  = mean(auc),  cavg = mean(cavg),
    .groups = "drop"
  )
```

```{r published}
# Tan 2024 Table 3, "final POPPK model" (mean of 1000 simulations).
published <- tibble::tribble(
  ~regimen,                 ~cmin, ~cmax,   ~auc, ~cavg,
  "40 mg QD fasted",          529,   727,  34270,   476,
  "60 mg 2d+1skip fasted",    469,   765,  34279,   476,
  "40 mg 2d+1skip fed",       504,   819,  36697,   510,
  "20 mg QD fasted",          269,   387,  17534,   244,
  "40 mg q72h fed",           221,   587,  18088,   251
)

cmp <- summary_tbl |>
  rename(cmin_sim = cmin, cmax_sim = cmax, auc_sim = auc, cavg_sim = cavg) |>
  left_join(published, by = "regimen") |>
  transmute(
    Regimen        = regimen,
    `Cmin pub`     = cmin,     `Cmin sim` = round(cmin_sim),
    `Cmin % diff`  = round(100 * (cmin_sim - cmin) / cmin, 1),
    `Cmax pub`     = cmax,     `Cmax sim` = round(cmax_sim),
    `Cmax % diff`  = round(100 * (cmax_sim - cmax) / cmax, 1),
    `AUC pub`      = auc,      `AUC sim`  = round(auc_sim),
    `AUC % diff`   = round(100 * (auc_sim - auc) / auc, 1)
  )
knitr::kable(
  cmp,
  caption = paste(
    "Simulated cohort (n =", n_per_arm,
    "per arm) against Tan 2024 Table 3. Cmin in ng/mL, Cmax in ng/mL,",
    "AUCss,72h in ng*h/mL."
  )
)
```

Trough concentrations reproduce the published values across all five arms.
Peak concentrations run high, which is expected: the observation grid here is
0.25 h, and a coarser grid cannot resolve the true peak, so `max(Cc)` is
grid-dependent in a way `Cmin` and `AUC` are not.

The **AUC and Cavg columns of Table 3 do not reproduce**, and the reason is not
in this model. See Errata below for the arithmetic; briefly, Table 3's own
Cmin and Cavg values are mutually impossible for the two once-daily reference
arms.

```{r metrics-gate}
# Cmin is the metric Table 3 reports that is both grid-robust and internally
# consistent, so it is the one worth gating on. This comparison has a published
# constant on one side and a drawn cohort on the other, so cohort noise is
# irreducible: with omega^2_CL = 0.213 the standard error on a cohort mean is
# about 3.4% at n = 200, and under common random numbers that draw shifts all
# five arms together. Gate on the centre first -- a mis-transcribed clearance,
# dose, volume or the ng/mL scaling moves every arm by tens of percent -- and
# keep only a loose envelope on the worst arm, which is the quantity that is
# not reproducible across rxode2 versions. Realised: median -3.8%, max 6.8%.
cmin_diff <- cmp$`Cmin % diff`
stopifnot(
  abs(median(cmin_diff)) < 12,
  max(abs(cmin_diff)) < 18
)
```

## Regimen equivalence: the paper's actual claim

The cost argument does not depend on the absolute exposure numbers, only on the
*ratio* between a proposed regimen and its reference. Tan 2024 applies the
narrow-therapeutic-index bioequivalence window of 90-111% (section 2.6).

```{r equivalence}
ref_of <- setNames(regimens$reference, regimens$regimen)
ratios <- summary_tbl |>
  mutate(reference = ref_of[regimen]) |>
  left_join(summary_tbl |> select(reference = regimen, auc_ref = auc),
            by = "reference") |>
  filter(regimen != reference) |>
  transmute(
    Regimen              = regimen,
    Reference            = reference,
    `AUCss,72h ratio (%)` = round(100 * auc / auc_ref, 1),
    `Cavg ratio (%)`      = round(100 * cavg / (auc_ref / 72), 1)
  )
knitr::kable(ratios, caption = "Exposure of each proposed regimen relative to its reference.")
```

```{r equivalence-gate}
# Two gates, from weakest to strongest.
#
# (1) The paper's own narrow-therapeutic-index bioequivalence window (90-111%,
#     section 2.6). This is the published claim, asserted verbatim.
stopifnot(all(ratios$`AUCss,72h ratio (%)` > 90),
          all(ratios$`AUCss,72h ratio (%)` < 111))

# (2) Much sharper, and available only because the arms are paired by common
#     random numbers: mass balance forces each ratio to exactly 100% at steady
#     state when the delivered dose per 72 h is equal, and pairing removes the
#     between-arm sampling noise that would otherwise leave several percent of
#     scatter. What survives is trapezoidal round-off and the residual approach
#     to steady state -- realised |ratio - 100| <= 0.03%. This gate does real
#     work: it goes red if a dose amount, a dosing pattern or the 50% food
#     multiplier is wrong, since any of those breaks the 120 mg-equivalent
#     (or 60 mg-equivalent) balance the paper's cost argument rests on.
stopifnot(max(abs(ratios$`AUCss,72h ratio (%)` - 100)) < 1)
```

Every proposed regimen is bioequivalent to its reference on AUC, reproducing the
paper's central conclusion: 60 mg for 2 days then a skip day matches 40 mg daily
(saving one third of the tablets), and 40 mg every 72 h with a high-fat meal
matches 20 mg daily (saving two thirds).

# PKNCA validation

The metrics above are computed by hand from a dense grid so that they match the
paper's definitions exactly (a minimum, a maximum and a trapezoidal integral
over a fixed 864-936 h window). PKNCA provides the independent check, computing
steady-state NCA over the same interval on the 40 mg QD reference arm.

```{r pknca-data}
nca_arm <- "40 mg QD fasted"

sim_nca <- sim |>
  filter(regimen == nca_arm) |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, regimen)

# PKNCA integrates from the interval start, so the interval start must carry a
# concentration row. The grid begins exactly at ss_start, so the row is present;
# assert it rather than assume it.
stopifnot(all(tapply(sim_nca$time, sim_nca$id, min) == ss_start))

dose_nca <- expand.grid(
  id   = unique(sim_nca$id),
  time = dose_times("qd")
) |>
  mutate(amt = 40, regimen = nca_arm) |>
  filter(time >= ss_start - 24, time < ss_end)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | regimen + id,
                             doseu = "mg")

intervals <- data.frame(
  start = ss_start, end = ss_end,
  cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-table}
nca_tbl <- as.data.frame(nca_res$result) |>
  group_by(PPTESTCD) |>
  summarise(mean = mean(PPORRES), .groups = "drop")
knitr::kable(nca_tbl |> rename("NCA parameter" = PPTESTCD, "Mean" = mean),
             digits = 1,
             caption = "PKNCA steady-state NCA over 864-936 h, 40 mg QD fasted arm.")
```

```{r pknca-gate}
getp <- function(code) {
  v <- nca_tbl$mean[nca_tbl$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique PKNCA row for '", code, "'")
  v
}
manual <- summary_tbl |> filter(regimen == nca_arm)

# PKNCA and the hand-rolled metrics read the same grid, so they must agree to
# within trapezoidal round-off; this gate catches a window or unit mismatch
# between the two paths, not model error.
stopifnot(
  abs(getp("cmax")    - manual$cmax) / manual$cmax < 0.01,
  abs(getp("cmin")    - manual$cmin) / manual$cmin < 0.01,
  abs(getp("auclast") - manual$auc)  / manual$auc  < 0.01,
  abs(getp("cav")     - manual$cavg) / manual$cavg < 0.01
)

# And the same closed-form anchor as before, now via PKNCA, on a cohort rather
# than the typical subject. The cohort mean of Dose/CL_i exceeds the
# typical-value Dose/CL by exp(omega^2/2) because CL is log-normal.
auc_expected <- 120 * 1000 / 3.11 * exp(0.213 / 2)
100 * (getp("auclast") - auc_expected) / auc_expected
stopifnot(abs(100 * (getp("auclast") - auc_expected) / auc_expected) < 12)
```

The PKNCA `auclast` for this arm is
`r round(getp("auclast"))` ng*h/mL against a closed-form cohort expectation of
`r round(auc_expected)` ng*h/mL -- i.e. the model's steady-state exposure is
governed by `Dose / (CL/F)` as it must be, and Table 3's 34270 ng*h/mL for the
same arm is roughly 20% below that.

# Assumptions and deviations

## Errata: Tan 2024 Table 3 AUC and Cavg columns are not reproducible

Tan 2024 Table 3 (and Supplementary Table S3, the same simulation run with the
registration-file CL/F) reports four exposure metrics per regimen. `Cmin,ss`
reproduces here; `AUCss,72h` and the `Cavg,ss` derived from it do not, running
roughly 13-19% below what the published model produces. Three independent lines
of evidence place the error in the paper, not in this implementation.

**1. The published table is internally impossible for the reference arms.**
Table 3 footnote 3 defines `Cavg,ss` as the mean of `AUCss,72h / 72`, so
`Cavg` and `Cmin` are means of per-subject quantities over the same subjects,
and `Cmin_i <= Cavg_i` must hold pointwise for a once-daily regimen. Yet:

```{r errata-impossible}
published |>
  filter(regimen %in% c("40 mg QD fasted", "20 mg QD fasted")) |>
  transmute(Regimen = regimen,
            `Cmin,ss (ng/mL)` = cmin,
            `Cavg,ss (ng/mL)` = cavg,
            `Cmin > Cavg?`    = cmin > cavg) |>
  knitr::kable(caption = "Tan 2024 Table 3: the reported trough exceeds the reported average concentration in both once-daily reference arms.")
```

A once-daily regimen's trough cannot sit above its own interval average. The
three skip-day / q72h arms have deep enough troughs that the inequality does not
trip there, but the two arms it does trip are the *references* against which
every relative change in the table is computed.

**2. The reported AUC contradicts mass balance.** At steady state
`AUCss,72h = D_72h / (CL/F)` exactly. For the 40 mg QD arm that is
120 mg / 3.11 L/h = `r round(120 * 1000 / 3.11)` ng*h/mL for the typical
subject, and `r round(auc_expected)` ng*h/mL as a cohort mean. Table 3 reports
34270. The published `Cmin` (529) and `Cmax` (727) for the same arm do bracket
the mass-balance `Cavg` of `r round(auc_expected / 72)` ng/mL, so the paper's
concentration metrics and its AUC metrics disagree with each other.

**3. The AUC error is arm-dependent, not a single scale factor.** Table 3
reports the fed 2-days-plus-skip regimen at +7.08% AUC versus the 40 mg QD
reference, but both deliver 120 mg-equivalent per 72 h under the paper's own
50% food-effect assumption, so at steady state the ratio must be 1. The
`Cavg` column repeats the same +6.76%.

The likely mechanism is the AUC bookkeeping compartment: the control stream
integrates `DADT(5) = A(3)/V2`, which accumulates in mg*h/L (= ug*h/mL) while
`IPRED` is scaled separately to ng/mL by `IPRED = A(3)*1000/V2`. A unit
reconciliation applied to the AUC compartment output would explain a systematic
offset, though not the arm-dependence.

**None of this undermines the paper's conclusions.** Its claim is that the
proposed regimens are bioequivalent to their references, and that claim is a
*ratio* -- which this implementation reproduces more cleanly than the published
table does (see the equivalence table above: all three ratios land within 0.03%
of 100%, against Table 3's own +7.08% for a regimen that delivers exactly the
reference dose per 72 h). The gate in this
vignette therefore asserts on `Cmin` and on the equivalence ratios, and reports
the AUC discrepancy rather than tuning anything to hide it.

## Other assumptions and deviations

* **Table 3's two fed arms are not consistent with a single food multiplier.**
  Back-solving the multiplier that would reproduce each fed arm's published
  exposure gives roughly 1.61 for the 40 mg 2-days-plus-skip arm and roughly
  1.54 for the 40 mg q72h arm -- neither equal to the stated 1.5, nor to each
  other. Both back-solves run through the `AUCss,72h` column, which the errata
  above shows is unreliable, so this is most likely another symptom of the same
  problem rather than an independent one. This vignette encodes the multiplier
  the paper states in words -- "the bioavailability of the drug will increase by
  50%" (section 2.6) -- rather than back-solving a different value per arm.
* **The food effect is a simulation assumption, not a fitted parameter.**
  It appears in neither control stream. `FED_HIGHFAT = 0` recovers the fitted
  model exactly. Tan 2024 section 4 notes the underlying food-effect study used
  the *capsule* formulation while the TDM cohort took tablets, and that the two
  are similar but not bioequivalent.
* **`F1 > 1` in about 4% of subjects** is reproduced rather than patched, since
  the exponential eta on `F1` is what the control stream specifies and what
  produced the numerical problem Tan 2024 documents. Total bioavailable dose is
  unaffected; the affected subjects' absorption *shape* is not physiological.
* **The Asian-race effect on CL/F is carried but never exercised.** Section
  2.4.2 assumed all patients Caucasian because ethnicity was not recorded; the
  control stream hardcodes `RACE = 1`. All simulations here use
  `RACE_ASIAN = 0`.
* **Sex is fixed to male in the simulations.** Both control streams simulate
  `GEND = 1`. Note the orientation trap documented in the model file: the Tan
  2024 dataset codes `SEX == 0` as *female* (that is the branch applying the
  0.79 multiplier), the inverse of the canonical `SEXF`.
* **Cmax is grid-dependent and is excluded from the gates.** `max(Cc)` over a
  discrete grid is an upward-biased estimate of the true peak, and the bias
  depends on how finely the grid resolves it. The 0.25 h grid used here is
  probably finer than the paper's (which is not stated), which is the likely
  reason simulated `Cmax` tends to run above Table 3; the realised deviations
  span roughly -5% to +15% across arms, mixing that bias with cohort noise.
  `Cmin` and `AUC` are grid-robust at this resolution and carry the gates
  instead.
* **The `AUC` bookkeeping compartment of the control stream is not
  reproduced.** rxode2 returns the concentration directly, so the vignette
  integrates externally.
* **Cohort size is 200 per arm**, against the paper's 1000 NONMEM subproblems.
  With `omega^2_CL = 0.213` that gives a standard error of about 3.4% on the
  cohort mean, which is the dominant source of scatter in the comparison
  against Table 3.
* **The five arms are paired by common random numbers.** Re-seeding before each
  arm gives every arm the same 200 individual parameter draws, so a regimen
  comparison is a within-subject contrast rather than a difference between two
  independently drawn cohorts. Whether Tan 2024 did the same is not stated (the
  control stream requests `SUBPROBLEMS=1000` from a single `$SIMULATION`
  record). Pairing does not change any single arm's expected exposure -- the
  Table 3 comparison is unaffected -- but it removes the between-arm noise from
  the AUC ratios, which is what lets the equivalence gate above assert 100%
  rather than merely the paper's 90-111% window.
