---
title: "Luvometinib (Tan 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Luvometinib (Tan 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Tan_2023_luvometinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Tan_2023_luvometinib"))$description`
* Article: <https://doi.org/10.3389/fphar.2023.1101991> (open access)

The paper names the drug by its development code, FCN-159. Its
international nonproprietary name is luvometinib (approved in China in 2025),
which is the name used for the model file.

# Population

The model was fit to 1030 plasma concentrations from 45 Chinese adults in two
phase I studies, both dosing luvometinib tablets once daily in the fasting
state (Tan 2023 Section 2.2 and Supplementary Tables S1-S2):

* **FCN-159-001** (NCT03932253): 33 patients with advanced melanoma with
  aberrant (phase Ia) or mutant (phase Ib) NRAS, doses 0.2-15 mg. Phase Ia
  opened with a single-dose PK run-in sampled to 72 h before continuous dosing.
* **FCN-159-002** (NCT04954001): 12 adults with neurofibromatosis type 1
  (NF1), doses 4-12 mg.

Pooled demographics (Table S2): median age 49 years (range 20-71), median
weight 63 kg (43-106), median body surface area (BSA) 1.66 m^2 (1.33-2.16),
median total protein 71.75 g/L (59-82), 26 men and 19 women. The model was
then used, with no pediatric data, to project exposure in children with NF1
aged 2-17 years.

The same information is available programmatically:

```{r population}
str(readModelDb("Tan_2023_luvometinib")()$population, give.attr = FALSE)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
table collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | log(0.5) 1/h | Table 2, KA |
| `ltlag` | log(0.211) h | Table 2, ALAG1 |
| `lcl` | log(13.2) L/h | Table 2, CL/F |
| `lvc` | log(48.7) L | Table 2, VC/F |
| `lq` | log(35.1) L/h | Table 2, Q/F |
| `lvp` | log(314) L | Table 2, VP/F |
| `e_bsa_cl`, `e_bsa_q`, `e_bsa_vp`, `e_bsa_vc` | 1.11, 2.20, 3.84, 2.41 | Table 2, '- BSA' rows; Table 1 theta8-theta11 |
| `e_tpro_vc` | -3.1 | Table 2, 'VC/F - TP'; Table 1 theta12 |
| `e_sexf_vp` | 0.751 | Table 2, 'VP/F - SEX'; Table 1 theta13 |
| `e_tumtp_nf1_fdepot` | 1.29 | Table 2, F1 and its footnote; Table 1 |
| `etalcl`, `etalvc`, `etalq`, `etalvp` | 0.03546, 0.39315, 0.06541, 0.06787 | Table 2 IIV 19%, 69.4%, 26%, 26.5% as log(1 + CV^2) |
| `propSd` | 0.258 | Table 2, proportional residual 25.8% |
| Size scaling `(BSA/1.66)^theta` on CL/F, Vc/F, Q/F, Vp/F | n/a | Table 1; Methods 2.4.2 (median BSA 1.66 m^2) |
| Vc/F `(TP/68.95)^theta12` | n/a | Table 1 |
| Vp/F `(1 + theta13 * SEX)`, SEX = 1 female | n/a | Table 1 and footnote |
| F1 = 1 (melanoma), theta7 (NF1) | n/a | Table 1; Table 2 footnote |
| Two compartments, first-order absorption with lag, linear elimination | n/a | Sections 2.4.1 and 3.2 (ADVAN4 TRANS4) |
| Proportional residual error | n/a | Section 3.2 |

# Covariate effects (Figure 2)

Figure 2 of the paper shows the steady-state AUC and Cmax of each covariate
level relative to a reference subject (BSA 1.66 m^2, male, total protein
68.95 g/L). The points are medians over simulated subjects; since every
covariate scales a log-normally distributed parameter, the median ratio is the
ratio of typical-value predictions, so the figure is replicated here with the
random effects set to zero. Each scenario receives 8 mg once daily for 28
days, and the ratios are taken over the last dosing interval. The ratios do
not depend on dose or on disease type.

```{r fig2-sim}
mod <- readModelDb("Tan_2023_luvometinib")
mod_typical <- rxode2::zeroRe(mod)

fig2_ref <- tibble::tribble(
  ~covariate,      ~level,      ~BSA, ~TPRO, ~SEXF, ~auc_pub, ~cmax_pub,
  "Reference",     "reference", 1.66, 68.95, 0,     1.00,     1.00,
  "BSA",           "2.01 m^2",  2.01, 68.95, 0,     0.80,     0.71,
  "BSA",           "1.81 m^2",  1.81, 68.95, 0,     0.90,     0.85,
  "BSA",           "1.55 m^2",  1.55, 68.95, 0,     1.08,     1.13,
  "BSA",           "1.42 m^2",  1.42, 68.95, 0,     1.19,     1.31,
  "Total protein", "79.68 g/L", 1.66, 79.68, 0,     1.00,     1.33,
  "Total protein", "74.83 g/L", 1.66, 74.83, 0,     1.00,     1.17,
  "Total protein", "67.8 g/L",  1.66, 67.80, 0,     1.00,     0.97,
  "Total protein", "62.7 g/L",  1.66, 62.70, 0,     1.00,     0.82,
  "Sex",           "female",    1.66, 68.95, 1,     1.00,     0.97
) |>
  mutate(id = row_number())

tau <- 24
n_dose <- 28
t_last <- (n_dose - 1) * tau
fig2_events <- fig2_ref |>
  select(id, BSA, TPRO, SEXF) |>
  mutate(TUMTP_NF1 = 0) |>
  tidyr::crossing(
    dplyr::bind_rows(
      tibble(time = seq(0, t_last, by = tau), evid = 1, amt = 8, cmt = "depot"),
      tibble(time = t_last + seq(0, tau, by = 0.05), evid = 0, amt = 0, cmt = "central")
    )
  ) |>
  arrange(id, time, desc(evid))

fig2_sim <- rxode2::rxSolve(
  mod_typical, events = fig2_events,
  rtol = 1e-10, atol = 1e-12, returnType = "data.frame"
)

fig2_exposure <- fig2_sim |>
  filter(time >= t_last) |>
  group_by(id) |>
  summarise(
    cmax = max(Cc),
    auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )
fig2 <- fig2_ref |>
  left_join(fig2_exposure, by = "id") |>
  mutate(
    auc_model = auc / auc[covariate == "Reference"],
    cmax_model = cmax / cmax[covariate == "Reference"]
  )

fig2 |>
  select(covariate, level, auc_pub, auc_model, cmax_pub, cmax_model) |>
  dplyr::rename(
    "Covariate" = covariate, "Level" = level,
    "AUC ratio, Figure 2" = auc_pub, "AUC ratio, model" = auc_model,
    "Cmax ratio, Figure 2" = cmax_pub, "Cmax ratio, model" = cmax_model
  ) |>
  knitr::kable(digits = 2, caption = "Steady-state exposure relative to the reference subject: Figure 2 of Tan 2023 vs the packaged model.")
```

```{r fig2-plot}
fig2 |>
  filter(covariate != "Reference") |>
  select(covariate, level, auc_pub, auc_model, cmax_pub, cmax_model) |>
  pivot_longer(c(auc_pub, auc_model, cmax_pub, cmax_model),
               names_to = c("metric", "source"), names_sep = "_") |>
  mutate(
    metric = ifelse(metric == "auc", "AUC,ss", "Cmax,ss"),
    source = ifelse(source == "pub", "Figure 2", "Model"),
    level = factor(level, levels = rev(unique(fig2$level)))
  ) |>
  ggplot(aes(value, level, colour = source, shape = source)) +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_point(size = 2.5) +
  facet_grid(covariate ~ metric, scales = "free_y", space = "free_y") +
  scale_x_log10() +
  labs(x = "Fold change relative to reference", y = NULL, colour = NULL, shape = NULL,
       caption = "Replicates the point estimates of Figure 2 of Tan 2023.")
```

The AUC ratios follow directly from CL/F and match Figure 2 to within 0.01
for every level, and the BSA rows match for Cmax to within 0.03. The
total-protein and sex rows for Cmax do not match; see Assumptions and
deviations.

```{r fig2-gate}
fig2_chk <- fig2 |> filter(covariate != "Reference")
stopifnot(
  # AUC: a closed-form consequence of the CL/F covariate model.
  all(abs(fig2_chk$auc_model - fig2_chk$auc_pub) < 0.02),
  # Cmax for BSA; a deterministic typical-value solve, so the bound only
  # needs to absorb the two-decimal rounding of the figure labels.
  all(abs(fig2_chk$cmax_model - fig2_chk$cmax_pub)[fig2_chk$covariate == "BSA"] < 0.04)
)
```

# Single-dose profiles and NCA

The melanoma study opened with a single dose followed by sampling to 72 h
(Table S1). The paper reports no NCA table of its own; it states that time to
peak is within 3 h (Introduction; 'ranging from 1 to 3 h', Discussion) and
that the terminal half-life is 31.8-57.4 h (Discussion). The simulation below
follows single 4, 8 and 12 mg doses in a virtual melanoma cohort of 100
subjects per dose group, sampled to 14 days so the terminal phase is resolved.

```{r sd-cohort}
rxode2::rxSetSeed(20230123)
set.seed(20230123)

# Covariates drawn to approximate Table S2 (FCN-159-001 column for melanoma,
# FCN-159-002 column for NF1). BSA is log-normal around the study median and
# total protein normal around it, each truncated to the reported range.
draw_covariates <- function(n, bsa_med, bsa_rng, tp_med, tp_rng, p_female, nf1) {
  tibble(
    BSA = pmin(pmax(exp(rnorm(n, log(bsa_med), 0.1)), bsa_rng[1]), bsa_rng[2]),
    TPRO = pmin(pmax(rnorm(n, tp_med, 5), tp_rng[1]), tp_rng[2]),
    SEXF = rbinom(n, 1, p_female),
    TUMTP_NF1 = nf1
  )
}

n_sd <- 100
sd_doses <- c(4, 8, 12)
sd_obs <- c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 2, 2.5, 3, 4, 6, 8, 12, 24, 48, 72,
            96, 120, 168, 240, 336)
sd_subjects <- bind_rows(lapply(seq_along(sd_doses), function(i) {
  draw_covariates(n_sd, 1.74, c(1.36, 2.16), 72.5, c(59, 82), 17 / 33, 0) |>
    mutate(id = (i - 1) * n_sd + seq_len(n_sd), dose = sd_doses[i])
}))
sd_events <- sd_subjects |>
  tidyr::crossing(tibble(time = c(0, sd_obs), evid = c(1, rep(0, length(sd_obs))))) |>
  mutate(
    amt = ifelse(evid == 1, dose, 0),
    cmt = ifelse(evid == 1, "depot", "central"),
    treatment = paste0(dose, " mg")
  ) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(sd_events[, c("id", "time", "evid")])))

sd_sim <- rxode2::rxSolve(mod, events = sd_events, keep = c("treatment", "dose"),
                          returnType = "data.frame")
```

```{r sd-plot}
sd_sim |>
  filter(time > 0) |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(x = "Time after single dose (h)", y = "Luvometinib Cc (ng/mL)",
       caption = "Median and 90% prediction interval, melanoma patients.")
```

```{r sd-nca}
sd_conc <- sd_sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
sd_conc <- bind_rows(
  sd_conc,
  sd_conc |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

sd_dose <- sd_events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

sd_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sd_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(sd_dose, amt ~ time | treatment + id),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))

sd_summary <- as.data.frame(sd_nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median)

sd_summary |>
  select(treatment, cmax, tmax, aucinf.obs, half.life) |>
  dplyr::rename(
    "Dose" = treatment, "Cmax (ng/mL)" = cmax, "Tmax (h)" = tmax,
    "AUC0-inf (ng*h/mL)" = aucinf.obs, "t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 1, caption = "Median single-dose NCA of the simulated melanoma cohort.")
```

The median simulated Tmax is between 1 and 3 h as reported, and the median
terminal half-life (about 35-39 h) lies inside the reported 31.8-57.4 h range.
That range is not tied to a stated dose or population. In the model the
half-life depends mainly on Vp/F, which is 75% larger in women, so women have
the longer half-lives.

```{r sd-gate}
sd_hl <- as.data.frame(sd_nca) |> filter(PPTESTCD == "half.life")
sd_tmax <- as.data.frame(sd_nca) |> filter(PPTESTCD == "tmax")
stopifnot(
  # Tmax 'within 3 h' / 'from 1 to 3 h'; the typical-value Tmax is 1.7 h.
  median(sd_tmax$PPORRES) >= 1, median(sd_tmax$PPORRES) <= 3,
  # Half-life: the PKNCA cohort median is about 35-39 h per dose group. A
  # mis-transcribed peripheral volume or intercompartmental clearance moves it
  # by a factor of several.
  median(sd_hl$PPORRES, na.rm = TRUE) > 20,
  median(sd_hl$PPORRES, na.rm = TRUE) < 60
)
```

# Pediatric BSA-band dosing (Figure 5)

The paper projected steady-state AUC0-24 in children aged 2-17 years by
drawing BSA uniformly between 0.55 and 2.15 m^2 and assigning integer doses
by BSA band, with between-subject variability and the adult bioavailability
(Section 2.6). Figure 5 shows the resulting bands (2 mg for 0.55-0.7 m^2 up to
8 mg for 1.875-2.125 m^2, about 4 mg/m^2) alongside adults with NF1 at 8 mg.
The box-plot medians below were digitised by the maintainers from the
figure's raster image (axis calibrated on its 0-2000 gridlines, about 3.3
ng*h/mL per pixel).

Children are simulated as NF1 patients; total protein is held at the 68.95
g/L reference and sex is drawn 50:50, neither of which affects AUC. The
adult arm uses the pooled BSA distribution (the figure labels it 1.33-2.16
m^2) with NF1 bioavailability.

```{r fig5-cohort}
bands <- tibble::tribble(
  ~band,              ~lo,   ~hi,   ~dose, ~auc_fig5,
  "0.55-0.7 m^2",     0.55,  0.70,  2,     626,
  "0.71-0.85 m^2",    0.71,  0.85,  3,     714,
  "0.86-1.125 m^2",   0.86,  1.125, 4,     709,
  "1.126-1.375 m^2",  1.126, 1.375, 5,     668,
  "1.376-1.625 m^2",  1.376, 1.625, 6,     638,
  "1.626-1.875 m^2",  1.626, 1.875, 7,     615,
  "1.875-2.125 m^2",  1.875, 2.125, 8,     591
)
adult_fig5 <- 768

n_band <- 120
n_adult <- 200
ped_subjects <- bind_rows(lapply(seq_len(nrow(bands)), function(i) {
  tibble(
    id = (i - 1) * n_band + seq_len(n_band),
    BSA = runif(n_band, bands$lo[i], bands$hi[i]),
    TPRO = 68.95,
    SEXF = rbinom(n_band, 1, 0.5),
    TUMTP_NF1 = 1,
    dose = bands$dose[i],
    treatment = bands$band[i]
  )
}))
adult_subjects <- draw_covariates(n_adult, 1.66, c(1.33, 2.16), 71.75, c(59, 82), 19 / 45, 1) |>
  mutate(id = nrow(ped_subjects) + seq_len(n_adult), dose = 8, treatment = "Adults 8 mg")
fig5_subjects <- bind_rows(ped_subjects, adult_subjects)

ss_obs <- t_last + c(0, 0.25, 0.5, 0.75, 1, 1.25, 1.5, 2, 2.5, 3, 4, 6, 8, 12, 16, 20, 24)
fig5_events <- fig5_subjects |>
  tidyr::crossing(bind_rows(
    tibble(time = seq(0, t_last, by = tau), evid = 1),
    tibble(time = ss_obs, evid = 0)
  )) |>
  mutate(
    amt = ifelse(evid == 1, dose, 0),
    cmt = ifelse(evid == 1, "depot", "central")
  ) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(fig5_events[, c("id", "time", "evid")])))

fig5_sim <- rxode2::rxSolve(mod, events = fig5_events, keep = c("treatment", "BSA"),
                            returnType = "data.frame")
```

```{r fig5-nca}
fig5_conc <- fig5_sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
fig5_dose <- fig5_events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

fig5_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(fig5_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(fig5_dose, amt ~ time | treatment + id),
  intervals = data.frame(start = t_last, end = t_last + tau, auclast = TRUE, cmax = TRUE)
))

fig5_published <- bind_rows(
  bands |> select(treatment = band, auclast = auc_fig5),
  tibble(treatment = "Adults 8 mg", auclast = adult_fig5)
)
fig5_cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = fig5_nca,
  reference = fig5_published,
  by = "treatment",
  params = "auclast",
  units = c(auclast = "ng*h/mL"),
  tolerance_pct = 10
)
knitr::kable(
  fig5_cmp,
  caption = "Median steady-state AUC0-24: simulation vs Figure 5 of Tan 2023 (digitised). * differs by more than 10%."
)
```

```{r fig5-plot}
fig5_auc <- as.data.frame(fig5_nca) |>
  filter(PPTESTCD == "auclast") |>
  mutate(treatment = factor(treatment, levels = c(bands$band, "Adults 8 mg")))
ggplot(fig5_auc, aes(treatment, PPORRES, fill = treatment == "Adults 8 mg")) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(data = fig5_published |>
               mutate(treatment = factor(treatment, levels = levels(fig5_auc$treatment))),
             aes(treatment, auclast), inherit.aes = FALSE, colour = "red", shape = 4, size = 3) +
  geom_hline(yintercept = adult_fig5, linetype = 2, colour = "red") +
  scale_fill_manual(values = c("white", "grey70"), guide = "none") +
  coord_cartesian(ylim = c(0, 2000)) +
  labs(x = NULL, y = "AUC0-24,ss (ng*h/mL)",
       caption = "Replicates Figure 5 of Tan 2023; red crosses are the digitised published medians.") +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

The simulated pediatric medians lie within about 10% of the published ones
and the adult median within a few percent. The model's pediatric medians fall more slowly with
BSA than the published ones (higher in the largest bands, lower in the
smallest); see Assumptions and deviations.

```{r fig5-gate}
fig5_med <- fig5_auc |>
  group_by(treatment) |>
  summarise(median_auc = median(PPORRES), .groups = "drop") |>
  mutate(treatment = as.character(treatment)) |>
  inner_join(fig5_published, by = "treatment") |>
  mutate(pct_diff = 100 * (median_auc / auclast - 1))
stopifnot(
  nrow(fig5_med) == nrow(bands) + 1,
  # Typical-value AUCs differ from the digitised medians by -8% to +8%; the
  # simulated medians add about 2-3% sampling noise at 120 subjects per band
  # (one render gave -8.9% to +7.5%, median +1.0%). A mis-transcribed CL/F
  # or F1 shifts every band together and fails the median check.
  abs(median(fig5_med$pct_diff)) < 5,
  all(abs(fig5_med$pct_diff) < 15)
)
```

# Assumptions and deviations

- **Drug name.** The paper uses the development code FCN-159; the file uses
  the INN luvometinib.
- **IIV scale.** Table 2 gives IIV as percentages without a formula. They are
  read as the coefficient of variation of a log-normal parameter and
  converted with omega^2 = log(1 + CV^2). Reading them as 100 x omega instead
  changes only Vc/F materially (variance 0.482 instead of 0.393). No
  covariance between etas was reported, so the etas are uncorrelated.
- **Residual error.** Methods 2.4.1 says combined additive and proportional
  error was used for model selection, but Results 3.2 and Table 2 report only
  a proportional term (25.8%), which is what the model uses.
- **Total-protein reference.** The Vc/F effect is centred on 68.95 g/L, the
  constant in the Table 1 equation and the forest-plot reference. Table S2
  gives a pooled median of 71.75 g/L, so the centring value is not the
  tabulated median. The equation value is kept.
- **Sex percentages.** Table S2 prints the sex and diagnosis percentages
  against a denominator of 48 instead of 45 (26 men shown as 54.17%). The
  counts are used: 19 of 45 subjects (42.2%) were women.
- **NF1 bioavailability.** NF1 and melanoma patients came from separate
  studies, so the F1 effect is equally a study effect. The authors attribute
  it to bioavailability because both Cmax and AUC were higher in NF1.
- **Figure 2, total protein and sex on Cmax (not reproduced).** The model
  gives Cmax,ss ratios of 0.93, 0.99, 1.07 and 1.12 for total protein of
  62.7, 67.8, 74.83 and 79.68 g/L, where Figure 2 shows 0.82, 0.97, 1.17 and
  1.33. Results 3.3 likewise states a 30% higher Cmax,ss at 76.89 g/L, where
  the model gives about 9%. The published effect is about 2.6 times larger on
  the log scale than the printed exponent of -3.1 produces through Vc/F. The
  same model reproduces the BSA rows of the same figure, which also act
  through Vc/F (exponent 2.41), so the mismatch is specific to the
  total-protein term. No other reading of the Table 1 equation reproduces the
  figure, and parameters were not tuned to match it. Figure 2 also shows
  women with about 3% lower Cmax,ss than men, while the model gives about 2%
  higher, a difference of about 5 points from a 75% larger peripheral volume.
  Neither row is gated above.
- **Figure 5 slope.** The published pediatric medians fall from about 710 to
  590 ng*h/mL between the 3 mg and 8 mg bands, which corresponds to a BSA
  exponent on CL/F near 1.24. The model, with the Table 2 exponent of 1.11,
  falls from about 680 to 640. The paper does not describe any change to the
  model for the pediatric simulation, so the Table 2 values are kept.
- **Virtual cohorts.** Covariate distributions are drawn to approximate Table
  S2: BSA log-normal (log-scale SD 0.1) around the study median and total
  protein normal (SD 5 g/L) around it, each truncated to the reported range.
  Children have BSA uniform within each band as in the paper, with total
  protein at the reference value; children are assumed to have the adult NF1
  bioavailability, as in the paper.
- **Figure 4 (VPC)** pools several dose levels and two sampling schedules per
  panel without listing the per-subject doses, so it is not replicated.
- **Errata.** No correction notice is linked to the article in Europe PMC
  as of 2026-10-11.
