---
title: "Levofloxacin (Tao 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Levofloxacin (Tao 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Tao_2026_levofloxacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Tao_2026_levofloxacin"))$description`
* Article: <https://doi.org/10.1128/aac.01853-25>
* Supplement (Tables S1-S3, Figures S1-S4): `AAC01853-25-s0001.docx`, available
  from the Europe PMC open-access record for
  [PMC13231918](https://europepmc.org/article/MED/10.1128/aac.01853-25)

Tao and colleagues developed the first population PK model for intravenous
levofloxacin in children with severe refractory *Mycoplasma pneumoniae*
pneumonia (SRMPP), then used a Cox proportional hazards model to derive a
steady-state exposure target and Monte Carlo simulation to recommend a dose.
This vignette packages the PK model and checks it against the paper's own
answer keys: the reported weight-normalized clearance, the observed
steady-state AUC distribution (Table S2), and the 12-scenario probability of
target attainment tables (Table 3 and Table S3).

# Population

The model was fit to 293 scavenged plasma samples from 191 Chinese pediatric
patients (100 male, 91 female) aged 0.16 to 16 years, treated at a single
center between April 2023 and April 2024 (Table 1). Mean age was 7.07 years
(SD 2.77) and mean body weight 24.05 kg (SD 9.00). All patients had normal
renal function: mean serum creatinine 31.77 umol/L (SD 8.03) and mean eGFR by
the modified Schwartz equation 149.61 mL/min/1.73 m^2 (SD 34.48). Levofloxacin
was given as a 0.5 to 2 h syringe-pump infusion at 8 to 10 mg/kg per dose, q12h
below 5 years and q24h from 5 years, capped at 750 mg/day; the mean daily dose
was 11.32 mg/kg (SD 3.06). All patients received concomitant glucocorticoids.

The design is sparse and opportunistic -- roughly 1.5 samples per child -- which
is why the authors could estimate interindividual variability on clearance only
and fixed it to zero on the central volume, the peripheral volume and
intercompartmental clearance because of high shrinkage. Only 7 patients were
under 2 years of age.

A separate exposure-response cohort of 161 patients (Table S2, after excluding
4 who discontinued for adverse effects) had a median steady-state AUC(0-24 h) of
35.59 mg*h/L (IQR 33.11 to 41.54) on a median daily dose of 10.00 mg/kg. That
distribution is the main external anchor used below.

```{r population}
pop <- rxode2::rxode(readModelDb("Tao_2026_levofloxacin"))$population
str(pop, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Tao_2026_levofloxacin.R`. The table below
collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (theta CL) | 6.85 L/h | Table 2, fixed effects, `theta_CL` (RSE 4.33%); bootstrap median 6.88, 95% CI 6.34-7.48 |
| `lvc` (theta V1) | 25.04 L | Table 2, `theta_V1` (RSE 8.78%); bootstrap median 25.15, 95% CI 20.98-29.94 |
| `lq` (theta Q) | 1.20 L/h | Table 2, `theta_Q` (RSE 29.54%); bootstrap median 1.19, 95% CI 0.64-2.72 |
| `lvp` (theta V2) | 5.28 L | Table 2, `theta_V2` (RSE 18.84%); bootstrap median 5.32, 95% CI 3.51-8.38 |
| `e_wt_cl` | 1.11 | Table 2, `theta_BW_CL` (RSE 12.88%); bootstrap median 1.09, 95% CI 0.76-1.29 |
| `e_creat_cl` | -0.20 | Table 2, `theta_SCr_CL` (RSE 24.86%); bootstrap median -0.20, 95% CI -0.31 to -0.10 |
| `e_wt_vc` | 1.83 | Table 2, `theta_BW_V1` (RSE 14.66%); bootstrap median 1.77, 95% CI 1.24-2.18 |
| `etalcl` (variance) | 0.013 | Table 2, interindividual variability, row `CL (omega 2)` (RSE 21.08%); bootstrap median 0.012, 95% CI 0.0073-0.018 |
| `propSd` | 0.350 | Table 2, residual variability, `Proportional error (%)` = 35.0 (RSE 5.90%); bootstrap median 35.0, 95% CI 31.0-39.0 |
| `cl <- ... * (WT/24.05)^e_wt_cl * (CREAT/31.77)^e_creat_cl` | n/a | Table 2 footnote equation: `CL = theta_CL * (BW/24.05)^theta_BW_CL * (SCr/31.77)^theta_SCr_CL * exp(eta_CL)` |
| `vc <- ... * (WT/24.05)^e_wt_vc` | n/a | Table 2 footnote equation: `V1 = theta_V1 * (BW/24.05)^theta_BW_V1` |
| Two-compartment, first-order elimination | n/a | Results, "PopPK analysis of levofloxacin": "A two-compartment model with first-order elimination best described the data" |
| Exponential IIV, proportional residual error | n/a | Methods, "Population pharmacokinetic modeling"; Results: "Residual variability was best captured by a proportional model" |
| IIV on V1, V2 and Q fixed to zero | n/a | Results: "Interindividual variability for the central (V1) and peripheral (V2) volumes of distribution and intercompartmental clearance was fixed to zero due to high shrinkage" |
| Reference values 24.05 kg and 31.77 umol/L | n/a | Table 1 cohort **means** (not the medians 22.50 kg / 30.70 umol/L); the Table 2 footnote equations normalize to them explicitly |

Two numbers external to Table 2 are used as validation targets below and are
*not* model inputs:

| Target | Value | Source location |
|--------|-------|-----------------|
| Weight-normalized typical CL | 0.28 L/kg/h | Discussion, first paragraph after the model summary |
| Observed steady-state AUC(0-24 h), n = 161 | median 35.59 mg*h/L (IQR 33.11-41.54) | Table S2, on a median daily dose of 10.00 mg/kg |
| Efficacy target AUC(0-24 h) | 30.74 mg*h/L | Results, exposure-response; HR 2.03 (95% CI 1.12-3.68), P = 0.019 |
| Probability of target attainment, 12 scenarios | Table 3 (8/9/10 mg/kg) and Table S3 (9/10/11 mg/kg) | |

# Model structure

```{r model}
mod <- readModelDb("Tao_2026_levofloxacin")
mod
```

Dosing is intravenous into `central`; there is no depot. The infusion duration
is set to 1 h throughout this vignette, the midpoint of the 0.5 to 2 h range in
Methods. For a linear model the steady-state AUC over a dosing interval does
not depend on the infusion duration at all, so nothing below is sensitive to
that choice; only the peak concentration is.

# Typical-value profile and the steady-state AUC identity

The single most informative deterministic check on this model is that the
steady-state AUC over one dosing interval equals the daily dose divided by
clearance. That identity holds for any linear disposition model at steady
state, so it simultaneously exercises the ODE system, the infusion handling,
the NCA window and the covariate model -- and it is what the paper's entire
exposure-response and dose-finding argument rests on, because
`AUC(ss,0-24h) = daily dose / CL` is how the authors' Monte Carlo target
attainment is generated.

```{r typical-events}
tau_h        <- 24        # dosing interval for the q24h reference schedule
n_days       <- 7         # doses given before the NCA interval; terminal t1/2 is ~4 h
infusion_dur <- 1         # h, midpoint of the 0.5-2 h range in Methods
ss_start     <- (n_days - 1) * tau_h
ss_end       <- ss_start + tau_h

# One subject: the reference child at the Table 1 cohort means.
ref_wt    <- 24.05
ref_creat <- 31.77

make_events <- function(subjects, mgkg_per_dose, tau, id_offset = 0L,
                        label = "reference") {
  subjects <- subjects |>
    dplyr::mutate(
      id  = id_offset + dplyr::row_number(),
      amt = pmin(mgkg_per_dose * .data$WT, 750 * tau / 24)  # 750 mg/day cap, Methods
    )
  dose_times <- seq(0, ss_start + tau_h - tau, by = tau)
  doses <- subjects |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(evid = 1L, dur = infusion_dur, cmt = "central")
  obs <- subjects |>
    dplyr::select(-"amt") |>
    tidyr::crossing(time = seq(ss_start, ss_end, by = 0.25)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(regimen = label) |>
    dplyr::arrange(.data$id, .data$time, dplyr::desc(.data$evid))
}

ref_subject <- tibble::tibble(WT = ref_wt, CREAT = ref_creat)
ev_typ <- make_events(ref_subject, mgkg_per_dose = 10, tau = tau_h,
                      label = "10 mg/kg q24h")
```

```{r typical-solve}
sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_typ,
  keep = c("WT", "CREAT", "regimen"), returnType = "data.frame"
)
# rxSolve drops `id` entirely for a single subject (see the vignette failure
# patterns reference); restore it so the PKNCA grouping formula works.
if (is.null(sim_typ$id)) sim_typ$id <- 1L

cl_typ <- unique(round(sim_typ$cl, 10))
vc_typ <- unique(round(sim_typ$vc, 10))
stopifnot(length(cl_typ) == 1L, length(vc_typ) == 1L)

tibble::tibble(
  Quantity = c("CL (L/h)", "CL/WT (L/kg/h)", "V1 (L)", "V1/WT (L/kg)",
               "Q (L/h)", "V2 (L)", "Terminal half-life (h)"),
  Model = c(
    cl_typ, cl_typ / ref_wt, vc_typ, vc_typ / ref_wt,
    unique(sim_typ$q), unique(sim_typ$vp),
    {
      k10 <- cl_typ / vc_typ
      k12 <- unique(sim_typ$q) / vc_typ
      k21 <- unique(sim_typ$q) / unique(sim_typ$vp)
      s <- k10 + k12 + k21
      log(2) / ((s - sqrt(s^2 - 4 * k10 * k21)) / 2)
    }
  )
) |>
  knitr::kable(digits = 3,
               caption = "Typical-value parameters at the Table 1 cohort means (24.05 kg, 31.77 umol/L).")
```

The paper reports the weight-normalized typical clearance in the Discussion as
0.28 L/kg/h. The model gives `r sprintf("%.4f", cl_typ / ref_wt)` L/kg/h, which
rounds to the published value.

```{r typical-check-cl}
# Deterministic: no cohort, no RNG. A mis-transcribed theta_CL or reference
# weight moves this immediately.
stopifnot(abs(cl_typ / ref_wt - 0.28) < 0.005)
```

```{r typical-nca}
nca_conc_typ <- sim_typ |>
  dplyr::filter(!is.na(.data$Cc)) |>
  dplyr::select("id", "time", "Cc", "regimen")

nca_dose_typ <- ev_typ |>
  dplyr::filter(.data$evid == 1L) |>
  dplyr::select("id", "time", "amt", "regimen")

ss_intervals <- data.frame(
  start = ss_start, end = ss_end,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  auclast = TRUE, cav = TRUE
)

res_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_conc_typ, Cc ~ time | regimen + id,
                   concu = "mg/L", timeu = "h"),
  PKNCA::PKNCAdose(nca_dose_typ, amt ~ time | regimen + id, doseu = "mg"),
  intervals = ss_intervals
))

typ_tbl <- as.data.frame(res_typ$result) |>
  dplyr::select("PPTESTCD", "PPORRES") |>
  tidyr::pivot_wider(names_from = "PPTESTCD", values_from = "PPORRES")

daily_dose <- 10 * ref_wt
auc_identity <- daily_dose / cl_typ

tibble::tibble(
  Quantity = c("AUC(0-24 h) at steady state by PKNCA (mg*h/L)",
               "Daily dose / CL (mg*h/L)",
               "Relative difference"),
  Value = c(typ_tbl$auclast, auc_identity,
            typ_tbl$auclast / auc_identity - 1)
) |>
  knitr::kable(digits = c(0, 6),
               caption = "Closed-form gate: steady-state AUC over one interval equals daily dose / CL.")
```

```{r typical-check-auc}
# Both sides use the SAME drawn parameters, so the only difference is PKNCA's
# interpolation error on a 0.25 h grid. That is pure numerical error, not
# per-subject physical variation, so a tight bound is correct here and must not
# be loosened. PKNCA's default "lin up/log down" method realises -3.3e-4 for
# this profile; a pure linear trapezoid over the same grid agrees with
# daily dose / CL to 1e-12, which is how the residual is known to be
# interpolation rather than a structural error. 1e-3 leaves 3x headroom and
# still goes red on anything mechanistic, which moves this by percent or more.
cat(sprintf("relative difference: %+.3e\n", typ_tbl$auclast / auc_identity - 1))
stopifnot(abs(typ_tbl$auclast / auc_identity - 1) < 1e-3)

# Cmin at steady state is well above zero, i.e. the profile has not decayed
# into solver noise over the interval used for NCA.
stopifnot(typ_tbl$cmin > 0, all(sim_typ$Cc >= 0))
```

```{r figure-2-typical}
sim_typ |>
  dplyr::mutate(tad = .data$time - ss_start) |>
  ggplot(aes(.data$tad, .data$Cc)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Levofloxacin concentration (mg/L)",
    title = "Typical-value steady-state profile, 10 mg/kg q24h, 24.05 kg child",
    caption = "Semi-log panel corresponding to Figure 2B of Tao 2026."
  ) +
  theme_bw()
```

# Virtual cohort and the observed steady-state AUC distribution

The exposure-response cohort of 161 patients had a median steady-state
AUC(0-24 h) of 35.59 mg*h/L (IQR 33.11 to 41.54) on a median daily dose of
10.00 mg/kg (Table S2). Those AUCs were produced by maximum-a-posteriori
Bayesian forecasting from this same model, so reproducing the *centre* of that
distribution is a direct check that the clearance and its covariate model were
transcribed correctly.

The cohort below draws body weight and serum creatinine as independent
lognormals matched to the Table 1 medians and IQRs, truncated to the 9 to 45 kg
and 19 to 45 umol/L windows spanned by the paper's own Monte Carlo scenarios
(Table 3) so nothing is simulated outside the observed covariate range. The
independence assumption is discussed under Assumptions and deviations.

```{r cohort}
# set.seed() seeds R's RNG for the covariate draws. It does NOT seed rxode2's
# simulation RNG, whose streams are partitioned per solver thread -- so the eta
# draws below differ between a 2-core CI runner and a 16-thread workstation.
# Every assertion in this vignette is written to hold for any cohort the model
# can produce.
set.seed(20260914)
rxode2::rxSetSeed(20260914)

n_cohort <- 200  # per arm; validation cohorts are capped at 200

# Lognormal parameters matched to the Table 1 medians and IQRs:
#   meanlog = log(median); sdlog = (log(Q3) - log(Q1)) / (2 * qnorm(0.75))
ln_par <- function(med, q1, q3) {
  c(meanlog = log(med), sdlog = (log(q3) - log(q1)) / (2 * stats::qnorm(0.75)))
}
wt_par    <- ln_par(22.50, 18.00, 28.00)   # Table 1 body weight
creat_par <- ln_par(30.70, 26.80, 37.00)   # Table 1 serum creatinine

rtrunc_lnorm <- function(n, par, lo, hi) {
  x <- stats::rlnorm(n, par[["meanlog"]], par[["sdlog"]])
  while (any(bad <- x < lo | x > hi)) {
    x[bad] <- stats::rlnorm(sum(bad), par[["meanlog"]], par[["sdlog"]])
  }
  x
}

cohort <- tibble::tibble(
  WT    = rtrunc_lnorm(n_cohort, wt_par,    9, 45),
  CREAT = rtrunc_lnorm(n_cohort, creat_par, 19, 45)
)

tibble::tibble(
  Covariate = c("Body weight (kg)", "Serum creatinine (umol/L)"),
  `Simulated median (IQR)` = c(
    sprintf("%.2f (%.2f, %.2f)", median(cohort$WT),
            quantile(cohort$WT, 0.25), quantile(cohort$WT, 0.75)),
    sprintf("%.2f (%.2f, %.2f)", median(cohort$CREAT),
            quantile(cohort$CREAT, 0.25), quantile(cohort$CREAT, 0.75))
  ),
  `Table 1 median (IQR)` = c("22.50 (18.00, 28.00)", "30.70 (26.80, 37.00)")
) |>
  knitr::kable(caption = "Virtual cohort covariates against Tao 2026 Table 1.")
```

```{r cohort-simulate}
ev_cohort <- make_events(cohort, mgkg_per_dose = 10, tau = tau_h,
                         label = "10 mg/kg q24h")
stopifnot(!anyDuplicated(unique(ev_cohort[, c("id", "time", "evid")])))

sim_cohort <- rxode2::rxSolve(
  mod, events = ev_cohort,
  keep = c("WT", "CREAT", "regimen"), returnType = "data.frame"
)
stopifnot(!is.null(sim_cohort$id), all(sim_cohort$Cc >= 0))
```

```{r figure-2-vpc}
sim_cohort |>
  dplyr::mutate(tad = .data$time - ss_start) |>
  dplyr::group_by(.data$tad) |>
  dplyr::summarise(
    Q05 = quantile(.data$Cc, 0.05), Q50 = median(.data$Cc),
    Q95 = quantile(.data$Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(.data$tad, .data$Q50)) +
  geom_ribbon(aes(ymin = .data$Q05, ymax = .data$Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Levofloxacin concentration (mg/L)",
    title = "Simulated 5th / 50th / 95th percentiles, 10 mg/kg q24h",
    caption = paste(
      "Corresponds to the simulated percentiles of Figure 2B of Tao 2026.",
      "Between-subject variability enters through clearance only (a single eta)",
      "plus the covariate spread, so the band is narrower than an observed VPC,",
      "which additionally carries the 35% proportional residual error."
    )
  ) +
  theme_bw()
```

```{r cohort-nca}
nca_conc <- sim_cohort |>
  dplyr::filter(!is.na(.data$Cc)) |>
  dplyr::select("id", "time", "Cc", "regimen")

nca_dose <- ev_cohort |>
  dplyr::filter(.data$evid == 1L) |>
  dplyr::select("id", "time", "amt", "regimen")

res_cohort <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_conc, Cc ~ time | regimen + id,
                   concu = "mg/L", timeu = "h"),
  PKNCA::PKNCAdose(nca_dose, amt ~ time | regimen + id, doseu = "mg"),
  intervals = ss_intervals
))

auc_subject <- as.data.frame(res_cohort$result) |>
  dplyr::filter(.data$PPTESTCD == "auclast") |>
  dplyr::select(id = "id", auc = "PPORRES")
stopifnot(nrow(auc_subject) == n_cohort, !anyNA(auc_subject$auc))

# Per-subject cross-check of the daily-dose/CL identity across the whole
# cohort, not just the typical subject: same drawn parameters on both sides,
# so this stays a tight numerical bound. It is looser than the typical-subject
# bound above only because PKNCA's lin-up/log-down interpolation error grows
# with the decay rate, and the lightest children in the cohort eliminate
# roughly 2.5x faster than the 24.05 kg reference child. Realised maximum is
# 1.4e-3, so 1e-2 leaves 7x headroom; a structural error is percent-level.
cl_subject <- sim_cohort |>
  dplyr::group_by(.data$id) |>
  dplyr::summarise(cl = dplyr::first(.data$cl), WT = dplyr::first(.data$WT),
                   .groups = "drop")
chk <- auc_subject |>
  dplyr::inner_join(cl_subject, by = "id") |>
  dplyr::mutate(auc_closed = 10 * .data$WT / .data$cl,
                rel = .data$auc / .data$auc_closed - 1)
cat(sprintf("max |relative difference| across the cohort: %.3e\n", max(abs(chk$rel))))
stopifnot(nrow(chk) == n_cohort, max(abs(chk$rel)) < 1e-2)
```

```{r cohort-compare}
sim_summary <- tibble::tibble(
  regimen     = "10 mg/kg q24h",
  auclast     = median(auc_subject$auc),
  cav         = median(auc_subject$auc) / tau_h
)

published <- tibble::tribble(
  ~regimen,          ~auclast, ~cav,
  "10 mg/kg q24h",   35.59,    35.59 / 24
)

nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = published,
  by        = "regimen",
  units     = c(auclast = "mg*h/L", cav = "mg/L"),
  tolerance_pct = 20
) |>
  knitr::kable(
    caption = paste(
      "Simulated median steady-state exposure against the Tao 2026 Table S2",
      "observed median for the 161-patient exposure-response cohort",
      "(median daily dose 10.00 mg/kg). * marks rows differing by >20%."
    ),
    align = c("l", "l", "r", "r", "r")
  )
```

```{r cohort-check}
auc_median <- median(auc_subject$auc)
# Centre, not extreme. The published 35.59 mg*h/L is a median over 161 real
# patients; the simulated median over 200 virtual ones has a sampling SE near
# 1% (IIV CV is only 11.4%), and the covariate draw contributes a little more.
# A 10% band is ample headroom for that while still going red on a
# mis-transcribed clearance, reference weight or dose, each of which moves the
# whole distribution by tens of percent.
stopifnot(abs(auc_median / 35.59 - 1) < 0.10)

# Robust spread check: the simulated interquartile range should sit inside a
# plausible multiple of the observed one. It is expected to be NARROWER, because
# the observed AUCs are MAP-Bayesian estimates that absorb residual error and
# real dose heterogeneity that the fixed-10-mg/kg simulation does not carry.
iqr_sim <- diff(unname(quantile(auc_subject$auc, c(0.25, 0.75))))
iqr_obs <- 41.54 - 33.11
stopifnot(iqr_sim < 1.5 * iqr_obs)
```

# Covariate model

Body weight enters both clearance and central volume with estimated exponents
well above the conventional allometric 0.75 and 1: 1.11 on clearance and 1.83
on central volume. The consequence worth seeing is that weight-normalized
central volume *rises* with body size rather than staying flat, so the model
should not be extrapolated outside the 9 to 45 kg window the paper simulated.

```{r covariate-grid}
cov_grid <- tidyr::crossing(
  WT    = seq(9, 45, by = 1),
  CREAT = c(19, 31.77, 45)
) |>
  dplyr::mutate(id = dplyr::row_number(), amt = 100, evid = 1L, time = 0,
                dur = infusion_dur, cmt = "central")

cov_sim <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = dplyr::bind_rows(
    cov_grid,
    cov_grid |> dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, time = 1)
  ) |> dplyr::arrange(.data$id, .data$time, dplyr::desc(.data$evid)),
  keep = c("WT", "CREAT"), returnType = "data.frame"
) |>
  dplyr::group_by(.data$id) |>
  dplyr::summarise(WT = dplyr::first(.data$WT), CREAT = dplyr::first(.data$CREAT),
                   cl = dplyr::first(.data$cl), vc = dplyr::first(.data$vc),
                   .groups = "drop")

cov_sim |>
  tidyr::pivot_longer(c("cl", "vc"), names_to = "param", values_to = "value") |>
  dplyr::mutate(
    per_kg = .data$value / .data$WT,
    param = factor(.data$param, c("cl", "vc"),
                   c("CL / WT (L/kg/h)", "V1 / WT (L/kg)")),
    CREAT = factor(.data$CREAT, c(19, 31.77, 45),
                   c("SCr 19 umol/L", "SCr 31.77 umol/L (mean)", "SCr 45 umol/L"))
  ) |>
  ggplot(aes(.data$WT, .data$per_kg, colour = .data$CREAT)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~param, scales = "free_y") +
  labs(x = "Body weight (kg)", y = "Weight-normalized parameter",
       colour = NULL,
       title = "Covariate model over the range simulated by Tao 2026",
       caption = paste(
         "Serum creatinine enters clearance only, so the V1 panel has three",
         "superimposed lines."
       )) +
  theme_bw() + theme(legend.position = "bottom")
```

```{r covariate-check}
ref_row <- cov_sim |> dplyr::filter(abs(.data$CREAT - 31.77) < 1e-8)
# Deterministic reproduction of the printed covariate exponents. Fitting
# log(parameter) on log(WT) must return the Table 2 exponents exactly, because
# the relation is an exact power law; an exponent typed into the wrong slot
# (CL vs V1) shows up here immediately.
slope_cl <- unname(coef(lm(log(ref_row$cl) ~ log(ref_row$WT)))[2])
slope_vc <- unname(coef(lm(log(ref_row$vc) ~ log(ref_row$WT)))[2])
stopifnot(abs(slope_cl - 1.11) < 1e-8, abs(slope_vc - 1.83) < 1e-8)

# Serum creatinine exponent, likewise exact.
scr_row <- cov_sim |> dplyr::filter(abs(.data$WT - 24) < 1e-8)
slope_scr <- unname(coef(lm(log(scr_row$cl) ~ log(scr_row$CREAT)))[2])
stopifnot(abs(slope_scr - (-0.20)) < 1e-8)
```

# Probability of target attainment (Table 3 and Table S3)

This is the paper's principal result and its richest answer key: 12 covariate
scenarios, each with a printed probability of reaching the
AUC(0-24 h) >= 30.74 mg*h/L efficacy target, under three regimens in Table 3 and
three more in Table S3.

Because the model is linear, steady-state exposure is exactly
`AUC = daily dose / CL_i`, and `CL_i` is lognormal with the typical value taken
from the model and log-scale variance 0.013. The attainment probability for a
scenario therefore has a closed form,
`P(AUC_i >= target) = P(CL_i <= daily dose / target) = Phi(log((daily dose / target) / CL_typ) / omega)`,
with no Monte Carlo noise at all. The typical clearances below are read out of
the packaged model rather than recomputed from the printed equation, so the
comparison against the paper's printed percentages is a genuine external check.

```{r pta-scenarios}
# Tao 2026 Table 3 / Table S3, columns "Age (years)", "Body weight (kg)" and
# "Serum creatinine (umol/L)". Scenarios 1-6 are dosed q12h in Table 3 and
# scenarios 7-12 q24h, per the two header blocks of that table.
scenarios <- tibble::tribble(
  ~scenario, ~age_band, ~WT, ~CREAT,
  1L,  "1-3",    9,  19,
  2L,  "1-3",   11,  21,
  3L,  "1-3",   13,  28,
  4L,  "3-5",   15,  24,
  5L,  "3-5",   17,  28,
  6L,  "3-5",   19,  30,
  7L,  "5-10",  21,  27,
  8L,  "5-10",  24,  31,
  9L,  "5-10",  27,  37,
  10L, "10-16", 33,  34,
  11L, "10-16", 39,  41,
  12L, "10-16", 45,  45
)

# Read the typical clearance for each scenario out of the model.
pta_ev <- scenarios |>
  dplyr::mutate(id = .data$scenario, amt = 100, evid = 1L, time = 0,
                dur = infusion_dur, cmt = "central")
pta_ev <- dplyr::bind_rows(
  pta_ev,
  pta_ev |> dplyr::mutate(amt = NA_real_, evid = 0L, dur = NA_real_, time = 1)
) |>
  dplyr::arrange(.data$id, .data$time, dplyr::desc(.data$evid))

cl_scen <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = pta_ev,
  keep = c("WT", "CREAT"), returnType = "data.frame"
) |>
  dplyr::group_by(id = .data$id) |>
  dplyr::summarise(cl_typ = dplyr::first(.data$cl), .groups = "drop")

scenarios <- scenarios |>
  dplyr::inner_join(cl_scen, by = c("scenario" = "id"))
stopifnot(nrow(scenarios) == 12L, !anyNA(scenarios$cl_typ))

auc_target <- 30.74                       # Results, exposure-response analysis
omega_cl   <- sqrt(0.013)                 # Table 2, row 'CL (omega 2)'

pta_closed <- function(daily_mgkg, WT, cl_typ) {
  cl_threshold <- daily_mgkg * WT / auc_target
  100 * stats::pnorm(log(cl_threshold / cl_typ) / omega_cl)
}
```

```{r pta-mc-crosscheck}
# Cross-check the closed form against a full ODE + PKNCA Monte Carlo run for
# one scenario, so the closed form is demonstrated rather than assumed.
rxode2::rxSetSeed(20260915)
mc_scen <- scenarios |> dplyr::filter(.data$scenario == 8L)
mc_cohort <- tibble::tibble(WT = mc_scen$WT, CREAT = mc_scen$CREAT)[rep(1, n_cohort), ]

mc_ev <- make_events(mc_cohort, mgkg_per_dose = 10, tau = tau_h,
                     label = "scenario 8, 10 mg/kg q24h")
mc_sim <- rxode2::rxSolve(mod, events = mc_ev, keep = c("WT", "CREAT", "regimen"),
                          returnType = "data.frame")

mc_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(
    mc_sim |> dplyr::filter(!is.na(.data$Cc)) |>
      dplyr::select("id", "time", "Cc", "regimen"),
    Cc ~ time | regimen + id, concu = "mg/L", timeu = "h"
  ),
  PKNCA::PKNCAdose(
    mc_ev |> dplyr::filter(.data$evid == 1L) |>
      dplyr::select("id", "time", "amt", "regimen"),
    amt ~ time | regimen + id, doseu = "mg"
  ),
  intervals = ss_intervals
))

mc_auc <- as.data.frame(mc_res$result) |>
  dplyr::filter(.data$PPTESTCD == "auclast") |> dplyr::pull("PPORRES")
stopifnot(length(mc_auc) == n_cohort, !anyNA(mc_auc))

mc_pta     <- 100 * mean(mc_auc >= auc_target)
closed_pta <- pta_closed(10, mc_scen$WT, mc_scen$cl_typ)

tibble::tibble(
  Source = c("ODE + PKNCA Monte Carlo (n = 200)", "Closed form", "Tao 2026 Table 3"),
  `PTA (%)` = c(mc_pta, closed_pta, 83.9)
) |>
  knitr::kable(digits = 1,
               caption = "Scenario 8 (24 kg, 31 umol/L) at 10 mg/kg q24h.")

# n = 200 around p = 0.87 has a binomial SE of ~2.4 percentage points, so the
# bound below is roughly 4 SE and cannot be met by a mis-specified model.
stopifnot(abs(mc_pta - closed_pta) < 10)
```

```{r pta-table}
regimens <- tibble::tribble(
  ~label,             ~daily_mgkg, ~source,
  "8 mg/kg q24h",     8,           "Table 3",
  "9 mg/kg q24h",     9,           "Table 3 / S3",
  "10 mg/kg q24h",    10,          "Table 3 / S3",
  "11 mg/kg q24h",    11,          "Table S3"
)

# Printed PTA for scenarios 7-12, Table 3 (8/9/10 mg/kg q24h) and Table S3
# (11 mg/kg q24h).
printed_q24 <- tibble::tribble(
  ~scenario, ~`8 mg/kg q24h`, ~`9 mg/kg q24h`, ~`10 mg/kg q24h`, ~`11 mg/kg q24h`,
  7L,  25.7, 50.2, 84.4, 96.3,
  8L,  14.6, 53.3, 83.9, 97.3,
  9L,  19.1, 62.9, 88.8, 98.0,
  10L, 23.2, 48.3, 79.6, 95.8,
  11L, 14.6, 54.6, 84.9, 97.7,
  12L, 19.3, 55.5, 86.4, 97.0
) |>
  tidyr::pivot_longer(-"scenario", names_to = "label", values_to = "printed")

model_q24 <- scenarios |>
  dplyr::filter(.data$scenario >= 7L) |>
  tidyr::crossing(regimens) |>
  dplyr::mutate(model = pta_closed(.data$daily_mgkg, .data$WT, .data$cl_typ)) |>
  dplyr::select("scenario", "age_band", "WT", "CREAT", "label", "daily_mgkg",
                "model")

pta_q24 <- model_q24 |>
  dplyr::inner_join(printed_q24, by = c("scenario", "label")) |>
  dplyr::mutate(diff = .data$model - .data$printed)
stopifnot(nrow(pta_q24) == 24L)

pta_q24 |>
  dplyr::arrange(.data$daily_mgkg) |>
  dplyr::mutate(
    # pivot_wider orders new columns by first appearance, so sort on the
    # numeric dose first -- an alphabetical label sort would put 10 and 11
    # before 8 and 9.
    label = factor(.data$label, levels = unique(.data$label)),
    cell  = sprintf("%.1f / %.1f", .data$model, .data$printed)
  ) |>
  dplyr::select("scenario", "age_band", "WT", "CREAT", "label", "cell") |>
  tidyr::pivot_wider(names_from = "label", values_from = "cell") |>
  dplyr::arrange(.data$scenario) |>
  dplyr::rename("Scenario" = "scenario", "Age (years)" = "age_band",
                "WT (kg)" = "WT", "SCr (umol/L)" = "CREAT") |>
  knitr::kable(
    caption = paste(
      "Probability of attaining AUC(0-24 h) >= 30.74 mg*h/L, ages 5-16",
      "(Tao 2026 scenarios 7-12). Each cell is model / published."
    )
  )
```

```{r pta-check}
# The 9, 10 and 11 mg/kg columns are the gate. They are deterministic on the
# model side (closed form, no RNG), so the only variation is the paper's own
# Monte Carlo noise at n = 1000 per scenario.
gate <- pta_q24 |> dplyr::filter(.data$daily_mgkg >= 9)
stopifnot(nrow(gate) == 18L)
stopifnot(mean(abs(gate$diff)) < 6, max(abs(gate$diff)) < 10)

# The 8 mg/kg column is excluded and reported as a known deviation below.
excluded <- pta_q24 |> dplyr::filter(.data$daily_mgkg == 8)
stopifnot(nrow(excluded) == 6L)

tibble::tibble(
  Column = c("9 mg/kg q24h", "10 mg/kg q24h", "11 mg/kg q24h", "8 mg/kg q24h"),
  `Mean absolute difference (percentage points)` = vapply(
    c(9, 10, 11, 8),
    function(d) mean(abs(pta_q24$diff[pta_q24$daily_mgkg == d])),
    numeric(1)
  ),
  `In gate` = c("yes", "yes", "yes", "no (see Assumptions and deviations)")
) |>
  knitr::kable(digits = 2,
               caption = "Agreement with the published PTA, by regimen column.")
```

## The 8 mg/kg column is internally inconsistent in the source

For a linear model, `AUC = daily dose / CL`, so scaling the mg/kg dose scales
every scenario's exposure by the same factor. The *ranking* of the 12 scenarios
by attainment probability must therefore be identical in every mg/kg column of
Table 3 -- whichever scenario has the highest PTA at 9 mg/kg must also have the
highest at 8, 10 and 11. The published 9, 10 and 11 mg/kg columns satisfy this;
the 8 mg/kg column does not, and is close to reversed.

```{r pta-ranking}
rank_tbl <- pta_q24 |>
  dplyr::group_by(.data$label) |>
  dplyr::summarise(
    `Rank correlation with the model` =
      stats::cor(.data$printed, .data$model, method = "spearman"),
    .groups = "drop"
  ) |>
  dplyr::arrange(.data$label)

knitr::kable(rank_tbl, digits = 3, caption = paste(
  "Spearman correlation between the published and model-predicted PTA across",
  "scenarios 7-12. A linear model forces the same scenario ordering in every",
  "column."
))
```

```{r pta-ranking-check}
# The 9/10/11 columns rank the six scenarios essentially as the model does.
hi <- rank_tbl |> dplyr::filter(.data$label != "8 mg/kg q24h")
stopifnot(nrow(hi) == 3L, all(hi$`Rank correlation with the model` > 0.7))

# The 8 mg/kg column does not, and reproducibly so: this is a property of the
# printed table, not of a random draw, so it is asserted rather than tolerated.
lo <- rank_tbl |> dplyr::filter(.data$label == "8 mg/kg q24h")
stopifnot(nrow(lo) == 1L, lo$`Rank correlation with the model` < 0)
```

## Ages 1 to 5 years, q12h

Table 3 reports 100% attainment for every 1 to 5 year scenario at 8, 9 and
10 mg/kg q12h, i.e. 16, 18 and 20 mg/kg/day. The model reproduces that.

```{r pta-q12}
pta_q12 <- scenarios |>
  dplyr::filter(.data$scenario <= 6L) |>
  tidyr::crossing(tibble::tibble(per_dose = c(8, 9, 10))) |>
  dplyr::mutate(
    label = sprintf("%d mg/kg q12h", .data$per_dose),
    model = pta_closed(2 * .data$per_dose, .data$WT, .data$cl_typ),
    printed = 100
  )
stopifnot(nrow(pta_q12) == 18L)

pta_q12 |>
  dplyr::arrange(.data$per_dose) |>
  dplyr::mutate(
    label = factor(.data$label, levels = unique(.data$label)),
    cell = sprintf("%.1f / %.0f", .data$model, .data$printed)
  ) |>
  dplyr::select("scenario", "age_band", "WT", "CREAT", "label", "cell") |>
  tidyr::pivot_wider(names_from = "label", values_from = "cell") |>
  dplyr::arrange(.data$scenario) |>
  dplyr::rename("Scenario" = "scenario", "Age (years)" = "age_band",
                "WT (kg)" = "WT", "SCr (umol/L)" = "CREAT") |>
  knitr::kable(caption = paste(
    "Probability of attaining AUC(0-24 h) >= 30.74 mg*h/L, ages 1-5",
    "(Tao 2026 scenarios 1-6, q12h). Each cell is model / published."
  ))

stopifnot(all(pta_q12$model > 99.9))
```

## The authors' recommended regimens and the q24h / q12h equivalence

The paper concludes that 11 mg/kg q24h or 5.5 mg/kg q12h give at least 90%
attainment across ages 1 to 16, and the Table S3 footnote states that
"9 mg/kg q24h and 4.5 mg/kg q12h, 10 mg/kg q24h and 5 mg/kg q12h, and
11 mg/kg q24h and 5.5 mg/kg q12h demonstrated equivalent PTA". For a linear
model that equivalence is exact, because both schedules deliver the same daily
dose and steady-state AUC over 24 h depends only on the daily dose.

```{r recommended}
recommended <- scenarios |>
  dplyr::mutate(
    `11 mg/kg q24h`  = pta_closed(11, .data$WT, .data$cl_typ),
    `5.5 mg/kg q12h` = pta_closed(2 * 5.5, .data$WT, .data$cl_typ)
  )

recommended |>
  dplyr::select("scenario", "age_band", "WT", "CREAT",
                "11 mg/kg q24h", "5.5 mg/kg q12h") |>
  dplyr::rename("Scenario" = "scenario", "Age (years)" = "age_band",
                "WT (kg)" = "WT", "SCr (umol/L)" = "CREAT") |>
  knitr::kable(digits = 1, caption = paste(
    "Model PTA for the two regimens Tao 2026 recommends. The paper's claim is",
    ">=90% attainment across all 12 scenarios, and that the two schedules are",
    "equivalent."
  ))
```

```{r recommended-check}
# Exact by construction for a linear model; assert it rather than assume it,
# because a dosing bug (for example an interval-dependent bioavailability or a
# mis-set infusion duration) would break it.
stopifnot(max(abs(recommended$`11 mg/kg q24h` -
                    recommended$`5.5 mg/kg q12h`)) < 1e-9)

# The paper's headline claim: at least 90% attainment in every scenario.
stopifnot(all(recommended$`11 mg/kg q24h` >= 90))

# And the complementary claim for the guideline regimen from 5 years. The
# paper's own wording is that "10 mg/kg q24 h approached relatively acceptable
# levels (80%-90%)" (Results, Monte Carlo simulations), so the band the paper
# states is the gate, not a stricter "all below 90" that the paper never
# claims. The model puts five of six scenarios below 90% and scenario 9 at
# 90.7% against a printed 88.8%.
inadequate <- pta_q24 |> dplyr::filter(.data$daily_mgkg == 10)
stopifnot(
  nrow(inadequate) == 6L,
  all(inadequate$model > 79), all(inadequate$model < 91),
  sum(inadequate$model < 90) >= 4L
)

# 11 mg/kg strictly improves on 10 mg/kg in every scenario -- a monotonicity
# the linear model forces, and a cheap guard against a dose-handling bug.
dose_step <- recommended |>
  dplyr::select("scenario", pta11 = "11 mg/kg q24h") |>
  dplyr::inner_join(inadequate |> dplyr::select("scenario", pta10 = "model"),
                    by = "scenario")
stopifnot(nrow(dose_step) == 6L, all(dose_step$pta11 > dose_step$pta10))
```

Safety was assessed against a threshold of 108 mg*h/L, the exposure of the
highest adult dose of 750 mg q24h, with no simulated regimen exceeding it.

```{r safety}
safety_threshold <- 108  # Methods, Monte Carlo simulations
risk <- scenarios |>
  dplyr::mutate(
    daily_mg = 11 * .data$WT,
    # P(AUC > threshold) = P(CL < daily dose / threshold)
    risk_pct =
      100 * stats::pnorm(log((.data$daily_mg / safety_threshold) / .data$cl_typ) /
                           omega_cl),
    # AUC at the typical clearance, as a multiple of the safety threshold: the
    # interpretable number, since the risk itself underflows to zero.
    auc_typ = .data$daily_mg / .data$cl_typ
  )

risk |>
  dplyr::mutate(
    `Typical AUC(0-24 h) (mg*h/L)` = sprintf("%.1f", .data$auc_typ),
    `Fraction of the 108 mg*h/L threshold` = sprintf("%.2f",
                                                     .data$auc_typ / safety_threshold),
    `Risk of exceeding it (%)` = sprintf("%.2g", .data$risk_pct)
  ) |>
  dplyr::select("scenario", "age_band", "WT", "Typical AUC(0-24 h) (mg*h/L)",
                "Fraction of the 108 mg*h/L threshold", "Risk of exceeding it (%)") |>
  dplyr::rename("Scenario" = "scenario", "Age (years)" = "age_band",
                "WT (kg)" = "WT") |>
  knitr::kable(
    caption = paste(
      "Toxicity risk at the recommended 11 mg/kg q24h regimen. The typical",
      "exposure sits at roughly a third of the safety threshold in every",
      "scenario, so the risk underflows; it is printed to two significant",
      "figures rather than rounded to zero."
    )
  )

# The paper's criterion was <10% risk of toxicity; the model puts it far below.
stopifnot(all(risk$risk_pct < 10))
# Guard against the risk column being trivially zero for the wrong reason (a
# broken threshold or a zero daily dose): typical exposure must be a real,
# positive fraction of the threshold, well under 1.
stopifnot(all(risk$auc_typ > 0), max(risk$auc_typ / safety_threshold) < 0.6)
```

# Assumptions and deviations

* **Infusion duration set to 1 h.** Methods give 0.5 to 2 h without a
  per-patient value. Steady-state AUC over a dosing interval is independent of
  infusion duration for a linear model, so no validation result above depends
  on this; only the peak concentration does.
* **Body weight and serum creatinine simulated independently.** Figure S2 shows
  a Pearson correlation matrix among the Table 1 covariates but prints only two
  coefficients in the text (body weight with age, r = 0.84, and with height,
  r = 0.88). No weight-creatinine coefficient is published, so the virtual
  cohort draws the two marginals independently from the Table 1 medians and
  IQRs. The paper's own Table 3 scenarios pair the two by within-age-band
  percentile and therefore do imply a positive association; a correlated draw
  would narrow the simulated AUC spread slightly and would not move the median,
  which is what the cohort gate asserts.
* **Covariate distributions are lognormal fits to Table 1.** Table 1 reports
  mean, SD, median and IQR but not the full distribution. Weight and creatinine
  are drawn as lognormals matched to the published medians and IQRs and
  truncated to the 9 to 45 kg and 19 to 45 umol/L windows spanned by the Table 3
  scenarios, so nothing is simulated outside the covariate range the authors
  themselves simulated.
* **The exposure-response layer is not encoded as a model.** The paper's
  efficacy analysis is a Cox proportional hazards model on a *binary* split of
  AUC(0-24 h) at 30.74 mg*h/L (HR 2.03, 95% CI 1.12-3.68), adjusted for
  pulmonary necrosis, plus a second Cox model for cough resolution at
  33.72 mg*h/L (HR 1.43, 95% CI 1.02-2.00). No baseline hazard function is
  reported, so no time-to-event model can be reconstructed from the published
  output; only the hazard ratio and the cutoff are available. The cutoff is
  carried here as the simulation target, which is how the authors use it, and
  is recorded in the model's `description`. No time-to-event parameters have
  been invented.
* **Interindividual variability on V1, V2 and Q is absent, not zero-valued.**
  The authors fixed these to zero for high shrinkage. They are omitted from
  `ini()` rather than written as `~ fixed(0)`, because a zero-variance diagonal
  makes OMEGA singular and breaks the Cholesky sampler `rxSolve()` uses. The
  practical consequence is that simulated between-subject spread in peak
  concentration is narrower than a real population's.
* **Reference values are the cohort means, not the medians.** The Table 2
  footnote equations normalize to 24.05 kg and 31.77 umol/L, which are the
  Table 1 *means*; the medians are 22.50 kg and 30.70 umol/L. Using the medians
  would shift every typical value.
* **Known deviation: the 8 mg/kg q24h column of Table 3 is not reproduced and
  is internally inconsistent.** The published values (25.7, 14.6, 19.1, 23.2,
  14.6, 19.3% for scenarios 7 to 12) rank the six scenarios in close to the
  reverse of the order the 9, 10 and 11 mg/kg columns rank them, with a
  Spearman correlation against the model of
  `r sprintf("%.2f", rank_tbl$"Rank correlation with the model"[rank_tbl$label == "8 mg/kg q24h"])`
  against `r sprintf("%.2f", min(hi$"Rank correlation with the model"))` or
  better for the other three. For a linear model the ordering is forced to be
  identical across columns, so the inconsistency is in the printed table rather
  than in this encoding: scaling the dose cannot reorder the scenarios. The
  column is excluded from the numeric gate and asserted to be
  anti-correlated instead, so a future change that accidentally "fixes" it
  would be caught. The 9, 10 and 11 mg/kg columns -- 18 published cells -- are
  reproduced to a mean absolute difference of
  `r sprintf("%.1f", mean(abs(gate$diff)))` percentage points, against the
  paper's Monte Carlo noise at n = 1000 per scenario.
* **Simulated exposure spread is narrower than the observed.** The Table S2
  AUCs were derived by MAP Bayesian forecasting from real, heterogeneous doses
  and absorb residual error; the simulation gives every subject exactly
  10 mg/kg/day and carries a single eta with a log-scale SD of 0.114. The
  centre matches; the spread is expected to be tighter, and the gate asserts
  the centre plus an upper bound on spread.
* **Do not extrapolate to renal impairment or to infants.** The cohort has no
  renally impaired children (mean eGFR 149.61 mL/min/1.73 m^2) and only 7
  patients under 2 years, so neither the creatinine power term nor the weight
  exponents are informed outside the observed range. The authors say so
  explicitly in their limitations.
