---
title: "Methotrexate (Taylor 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methotrexate (Taylor 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Taylor_2026_methotrexate"))$reference`
* Description: `r rxode2::rxode(readModelDb("Taylor_2026_methotrexate"))$description`
* Article: <https://doi.org/10.1007/s40262-026-01618-4> (open access, CC BY-NC 4.0)

Taylor 2026 asks a narrow methodological question: does putting cystatin C into
the glomerular filtration rate (GFR) estimate improve the prediction of
methotrexate clearance? The answer is the covariate model on `cl`. The CKD-EPI
creatinine-cystatin C 2021 equation beat serum creatinine, the creatinine-only
CKD-EPI equation and the cystatin C-only CKD-EPI equation, dropping the
objective function value by 117.1 units and explaining roughly 11.5% of the
total inter-individual variability in clearance.

# Population

Eighty adults with histologically confirmed lymphoma were enrolled
prospectively at a single center (Mayo Clinic, Rochester, MN) between January
2018 and December 2019, contributing 80 high-dose methotrexate administrations
and 427 serum methotrexate concentrations. The cohort is elderly (median age
68.6 years, 64% aged 65 or older), predominantly male (68%) and predominantly
Caucasian (93%), with a median weight of 80.5 kg and a median body surface area
of 1.97 m^2 (Taylor 2026 Table 1). Renal function at baseline was close to
normal: CKD-EPI creatinine-cystatin C eGFR 88 +/- 24 mL/min/1.73 m^2, with only
five patients (6%) carrying a history of chronic kidney disease. Patients with
any-stage AKI at baseline, patients on renal replacement therapy, and patients
whose infusion ran beyond 4 h were excluded.

Methotrexate was given as a 4 h intravenous infusion at a protocol dose of
1.5 g/m^2 (11% of patients), 3.5 g/m^2 (43%) or 8 g/m^2 (46%). Delivered doses
were materially lower than the protocol nominal - the median delivered dose in
the 8 g/m^2 stratum was 11.5 g against a nominal 8 x 1.97 = 15.8 g - consistent
with dose reduction in an elderly population.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Taylor_2026_methotrexate"))$population
str(pop[c("species", "n_subjects", "n_observations", "age_median", "regions")])
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Taylor_2026_methotrexate.R` carries an in-file
comment naming its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL)           | 11.8 L/h | Table 2, "CL (L/h)"; RSE 5.1%, bootstrap 11.8 (10.8-12.8) |
| `lvc` (V1)           | 42.2 L | Table 2, "V1 (L)"; RSE 6.4%, bootstrap 41.9 (37.8-46.5) |
| `lq` (Q)             | 0.35 L/h | Table 2, "Q (L/h)"; RSE 11.6%, bootstrap 0.35 (0.29-0.43) |
| `lvp` (V2)           | 6.67 L | Table 2, "V2 (L)"; RSE 10.4%, bootstrap 6.63 (5.56-7.88) |
| `e_crcl_cl`          | 0.80 | Table 2, "theta eGFR"; RSE 11.5%, bootstrap 0.79 (0.64-0.95) |
| `e_alb_cl`           | 0.69 | Table 2, "theta Albumin"; RSE 21.3%, bootstrap 0.67 (0.42-0.97) |
| `etalcl` (IIV on CL) | 0.02 (variance) | Table 2, "IIV CL"; RSE 24.3%, shrinkage 13.3% |
| `etalvp` (IIV on V2) | 0.03 (variance) | Table 2, "IIV V2"; RSE 23.9%, shrinkage 31% |
| `propSd`             | sqrt(0.20) | Table 2, "Residual error" (variance); RSE 11.6%, shrinkage 12.2% |
| eGFR reference 76 mL/min/1.73 m^2 | n/a | Eq. 1, page 470 (denominator); ESM Table S2 model 8 `(eGFR/76)^theta`; Sect. 3.2.2 "reference value set to the median GFR value" |
| Albumin reference 4 g/dL | n/a | Eq. 1, page 470 (denominator); ESM Table S2 model 3 `(Alb/4)^theta` |
| `cl <- ... * (CRCL / 76)^e_crcl_cl * (alb_gdL / 4)^e_alb_cl` | n/a | Eq. 1, page 470 |
| Two-compartment ODE system | n/a | Sect. 2.2.1 (NONMEM ADVAN3 TRANS4, CL / V1 / Q / V2) |
| Proportional residual error | n/a | Sect. 2.2.1 ("proportional error structure implemented via the log-transformed both sides approach") |

Table 2's footnote - "IIV and residual error are reported as the variance of the
selected parameter" - is what licenses using 0.02, 0.03 and 0.20 directly as
`omega^2` and `sigma^2` rather than as standard deviations. That reading is
independently confirmed by the paper's own prose, which is checked numerically
below.

## Published values transcribed from the paper

```{r published}
# Taylor 2026 Table 2 - final model parameter estimates.
tab2 <- tibble::tribble(
  ~parameter,       ~estimate, ~rse_pct, ~boot_median, ~boot_lo, ~boot_hi,
  "CL (L/h)",            11.8,      5.1,         11.8,     10.8,     12.8,
  "theta eGFR",          0.80,     11.5,         0.79,     0.64,     0.95,
  "theta Albumin",       0.69,     21.3,         0.67,     0.42,     0.97,
  "V1 (L)",              42.2,      6.4,         41.9,     37.8,     46.5,
  "Q (L/h)",             0.35,     11.6,         0.35,     0.29,     0.43,
  "V2 (L)",              6.67,     10.4,         6.63,     5.56,     7.88,
  "IIV CL (variance)",   0.02,     24.3,         0.02,     0.01,     0.03,
  "IIV V2 (variance)",   0.03,     23.9,         0.03,     0.02,     0.05,
  "Residual (variance)", 0.20,     11.6,         0.20,     0.16,     0.23
)

# Taylor 2026 Table 3 - baseline characteristics and delivered dose by
# protocol-dose stratum, median (IQR). "Albumin at baseline" is headed
# "(mg/L)" in Table 3, which is a units typo: Table 1 and Eq. 1 both put
# albumin in g/dL, and 3.9 mg/L is not a physiological albumin value.
tab3 <- tibble::tribble(
  ~arm,          ~n,  ~dose_med, ~dose_lo, ~dose_hi, ~crcl_med, ~crcl_lo, ~crcl_hi,
  "<=3.5 g/m2",  43L,       5.1,      3.4,      6.9,      82.5,     69.3,     96.5,
  "8 g/m2",      37L,      11.5,      9.1,    13.75,      80.2,     63.8,     95.4
) |>
  dplyr::mutate(
    alb_med = c(3.9, 3.8), alb_lo = c(3.3, 3.5), alb_hi = c(4.2, 4.1),  # g/dL
    cl_med  = c(11.9, 11.4)                                             # L/h, post hoc
  )

# Taylor 2026 Table 4 - model-estimated 4 h serum methotrexate concentration
# (umol/L), median (IQR). The row totals span both AKI strata.
tab4 <- tibble::tribble(
  ~arm,         ~conc_med, ~conc_lo, ~conc_hi,
  "<=3.5 g/m2",     129.2,     84.1,    178.8,
  "8 g/m2",         315.7,    240.2,    534.5
)

knitr::kable(tab2, caption = "Taylor 2026 Table 2, transcribed.")
```

## Internal consistency of the reported variance scale

Table 2 gives the IIV on clearance as 0.02 while the Results text quotes the IIV
on clearance in percent. If the percentages are `omega` (the standard deviation
on the log scale) then the two must reconcile exactly, and they do - which is a
free check that the model file has not misread a variance as an SD.

```{r variance-scale}
# Sect. 3.2.2: the eGFR term reduced IIV on CL "from 28.5 to 17%", and albumin
# then explained "2.9% of the inter-individual variability". So the final IIV
# on CL implied by the prose is 17 - 2.9 = 14.1%.
iiv_from_prose <- 17.0 - 2.9
iiv_from_tab2  <- 100 * sqrt(tab2$estimate[tab2$parameter == "IIV CL (variance)"])

c(prose_pct = iiv_from_prose, sqrt_of_table2_pct = round(iiv_from_tab2, 2))

stopifnot(abs(iiv_from_prose - iiv_from_tab2) < 0.1)
```

The prose figure and the square root of the tabulated value agree to within
0.1 percentage points, so Table 2's 0.02 is `omega^2` and the model file's
`etalcl ~ 0.02` is the correct encoding. The same footnote governs `etalvp` and
the residual error, hence `propSd <- sqrt(0.20)`.

# Units

Taylor 2026 reports methotrexate concentrations in umol/L throughout (Tables 3
and 4, the 160 umol/L AKI threshold, the LC-MS lower limit of detection of
10 ng/mL "~0.02 umol/L") and reports CL in L/h and V1 in L. The amount unit
that pairs with those is therefore **umol**, which is what the model file
declares. Doses reported in grams are converted with the methotrexate molecular
weight.

```{r units}
MW_MTX <- 454.44  # g/mol; not a Taylor 2026 value - see Assumptions below
g_to_umol <- function(g) g / MW_MTX * 1e6
```

# Virtual cohort

The individual patient data are not public ("not publicly available because of
institutional privacy policies"), so the cohorts below are virtual populations
whose delivered dose and covariate distributions reproduce the medians and
interquartile ranges of Taylor 2026 Table 3, stratum by stratum. Log-normal
distributions are fitted to each published median / IQR pair.

```{r cohort}
set.seed(20260830)

N_PER_ARM <- 150L  # <= 200 per arm

# Fit a log-normal to a published median and IQR.
lnorm_from_iqr <- function(med, lo, hi) {
  c(meanlog = log(med), sdlog = (log(hi) - log(lo)) / (2 * stats::qnorm(0.75)))
}

# Sampling grid: dense through the infusion and distribution phase, then out to
# 96 h, the last of the paper's protocol sampling times (4, 12, 24, 48, 72, 96 h).
t_obs <- c(seq(0, 12, by = 0.25), seq(12.5, 24, by = 0.5), seq(25, 96, by = 1))

make_cohort <- function(arm_label, n, id_offset, dose_p, crcl_p, alb_p) {
  subj <- tibble::tibble(
    id     = id_offset + seq_len(n),
    arm    = arm_label,
    dose_g = stats::rlnorm(n, dose_p[["meanlog"]], dose_p[["sdlog"]]),
    CRCL   = stats::rlnorm(n, crcl_p[["meanlog"]], crcl_p[["sdlog"]]),
    # ALB is supplied in canonical SI g/L; the paper's g/dL values x 10.
    ALB    = stats::rlnorm(n, alb_p[["meanlog"]], alb_p[["sdlog"]]) * 10
  )
  dplyr::bind_rows(
    subj |>
      dplyr::mutate(time = 0, amt = g_to_umol(dose_g), evid = 1L,
                    cmt = "central", dur = 4),
    tidyr::crossing(subj, time = t_obs) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", dur = NA_real_)
  )
}

events <- dplyr::bind_rows(
  make_cohort("<=3.5 g/m2", N_PER_ARM, 0L,
              lnorm_from_iqr(5.1, 3.4, 6.9),
              lnorm_from_iqr(82.5, 69.3, 96.5),
              lnorm_from_iqr(3.9, 3.3, 4.2)),
  make_cohort("8 g/m2", N_PER_ARM, 1000L,
              lnorm_from_iqr(11.5, 9.1, 13.75),
              lnorm_from_iqr(80.2, 63.8, 95.4),
              lnorm_from_iqr(3.8, 3.5, 4.1))
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# Disjoint id ranges across the two arms, and no duplicated event key. Note the
# absence of unique() here: anyDuplicated(unique(x)) is 0 by construction and
# would be a check that cannot fail.
stopifnot(
  anyDuplicated(events[, c("id", "time", "evid")]) == 0L,
  dplyr::n_distinct(events$id) == 2L * N_PER_ARM
)
```

The simulated cohort should reproduce the published medians it was built from;
if it does not, the sampling code is wrong before the model ever runs.

```{r cohort-check}
cohort_summary <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n             = dplyr::n(),
    dose_g_median = median(dose_g),
    crcl_median   = median(CRCL),
    alb_gdL_median = median(ALB) / 10,
    .groups       = "drop"
  ) |>
  dplyr::left_join(
    tab3 |> dplyr::select(arm, dose_med, crcl_med, alb_med), by = "arm"
  )

knitr::kable(cohort_summary, digits = 2,
             caption = "Simulated cohort vs Taylor 2026 Table 3 medians.")

stopifnot(
  with(cohort_summary, all(abs(dose_g_median / dose_med - 1) < 0.10)),
  with(cohort_summary, all(abs(crcl_median   / crcl_med - 1) < 0.10)),
  with(cohort_summary, all(abs(alb_gdL_median / alb_med  - 1) < 0.10))
)
```

# Simulation

```{r simulate}
mod <- readModelDb("Taylor_2026_methotrexate")
ui  <- rxode2::rxode(mod)

# `omega` is passed explicitly on every solve in this vignette. rxode2 keeps the
# omega in the solve options attached to the *compiled* model, so a later
# zeroRe() run would otherwise inherit this stochastic solve's omega and quietly
# re-sample etas (and vice versa). Both directions fail silently, so each solve
# below is also followed by a guard asserting that IIV is present / absent.
sim <- rxode2::rxSolve(mod, events = events, keep = c("arm", "dose_g"),
                       omega = ui$omega) |>
  as.data.frame()

if (is.null(sim$id)) sim$id <- 1L

stopifnot(
  all(sim$Cc >= 0, na.rm = TRUE),
  # IIV must actually have been applied: cl and vp carry etas, so they must vary.
  dplyr::n_distinct(round(sim$cl, 8)) > 1L,
  dplyr::n_distinct(round(sim$vp, 8)) > 1L
)
```

`Cc` is the individual prediction; the residual error is carried separately by
`sim`, so the figures and NCA below are free of measurement noise.

# Structural verification

Before comparing anything against the paper, confirm the ODE system in the model
file is the two-compartment infusion model the paper describes. The check
compares the solver against an independent closed-form solution for a zero-order
infusion into the central compartment of a two-compartment system, written out
from the hybrid rate constants. Both sides use the same parameter values, so the
only difference is numerical integration error and the bound can be tight.

```{r closed-form}
CL <- 11.8; V1 <- 42.2; Q <- 0.35; V2 <- 6.67   # Taylor 2026 Table 2
k10 <- CL / V1; k12 <- Q / V1; k21 <- Q / V2
sum_k <- k10 + k12 + k21
alpha <- (sum_k + sqrt(sum_k^2 - 4 * k10 * k21)) / 2
beta  <- (sum_k - sqrt(sum_k^2 - 4 * k10 * k21)) / 2

# Concentration during a zero-order infusion at rate R into the central
# compartment of a two-compartment model, evaluated at time t <= infusion end.
conc_infusion <- function(t, R) {
  (R / V1) * (
    (k21 - alpha) / (-alpha * (alpha - beta)) * (1 - exp(-alpha * t)) +
    (k21 - beta)  / ( beta  * (alpha - beta)) * (1 - exp(-beta  * t))
  )
}

# Typical-value solve at the covariate reference point (eGFR 76, albumin 4 g/dL)
# so cl == 11.8 exactly and the two sides are directly comparable.
mod_typ <- rxode2::zeroRe(mod)
ref_dose_g <- 11.5
ev_ref <- dplyr::bind_rows(
  tibble::tibble(id = 1L, time = 0, amt = g_to_umol(ref_dose_g), evid = 1L,
                 cmt = "central", dur = 4, CRCL = 76, ALB = 40),
  tibble::tibble(id = 1L, time = seq(0, 4, by = 0.05), amt = NA_real_, evid = 0L,
                 cmt = "central", dur = NA_real_, CRCL = 76, ALB = 40)
) |>
  dplyr::arrange(time, dplyr::desc(evid))

# omega = NA is the only sentinel that actually suppresses eta sampling here;
# zeroRe() alone is not enough once a stochastic solve has run in this session.
sim_ref <- rxode2::rxSolve(mod_typ, ev_ref, omega = NA) |> as.data.frame()

# Guard: with the etas suppressed, cl must be exactly the typical value at the
# covariate reference point. If eta sampling leaked back in, this trips before
# the closed-form comparison can be misread as a model error.
stopifnot(abs(unique(round(sim_ref$cl, 8)) - 11.8) < 1e-6)

sim_ref$closed_form <- conc_infusion(sim_ref$time, g_to_umol(ref_dose_g) / 4)

max_rel_err <- max(abs(sim_ref$Cc - sim_ref$closed_form) /
                     pmax(sim_ref$closed_form, 1e-12))
signif(max_rel_err, 3)

stopifnot(max_rel_err < 1e-6)
```

The solver reproduces the analytic infusion solution to better than one part in
a million, so the ODE transcription of the ADVAN3 mass balance is correct.

The derived disposition half-lives are a useful sanity anchor: methotrexate is
routinely described as biphasic with a terminal half-life of roughly 8-15 h in
adults with normal renal function.

```{r half-lives}
c(distribution_t_half_h = log(2) / alpha, terminal_t_half_h = log(2) / beta)
```

Next, confirm the covariate function is Eq. 1 exactly. `cl` is returned by
`rxSolve` as a model variable, so it can be checked directly against the
published equation evaluated on each subject's own covariates.

```{r covariate-function}
cov_check <- sim |>
  dplyr::group_by(id) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    # Eq. 1 with the eta removed: cl / exp(eta) must equal the covariate factors
    # times 11.8. Recover the eta-free ratio via the typical CL.
    eq1_factor   = (CRCL / 76)^0.80 * ((ALB * 0.1) / 4)^0.69,
    cl_over_eta  = cl / eq1_factor   # == 11.8 * exp(etalcl)
  )

# The spread of cl_over_eta must be exactly the IIV on CL and nothing else.
c(median_cl_over_eta = median(cov_check$cl_over_eta),
  sd_log_cl_over_eta = sd(log(cov_check$cl_over_eta)),
  target_omega       = sqrt(0.02))

stopifnot(
  # Median of a log-normal eta is 1, so the median must recover the typical CL.
  abs(median(cov_check$cl_over_eta) / 11.8 - 1) < 0.03,
  # And the residual spread must be the declared omega, not something else.
  abs(sd(log(cov_check$cl_over_eta)) / sqrt(0.02) - 1) < 0.15
)
```

Dividing each subject's simulated `cl` by Eq. 1's covariate factor leaves a
quantity whose median is the published 11.8 L/h and whose log-scale spread is
the published `sqrt(0.02)`. Nothing else is acting on clearance.

# Replicate published figures

```{r figure-1}
# Structurally replicates Figure 1 of Taylor 2026: the prediction-corrected
# visual predictive check of the final model. The observed data underlying the
# published figure are not available, so this shows the simulated median and
# 10th / 90th percentiles from the packaged model over the same time window.
# Base-R subset (not dplyr::filter) so this plotting-only drop of the t = 0 row
# is never confused with the PKNCA input filter, which must keep time = 0.
sim[sim$time > 0, ] |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q10 = quantile(Cc, 0.10), Q50 = median(Cc), Q90 = quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(x = "Time after start of infusion (h)",
       y = "Serum methotrexate (umol/L)",
       title = "Simulated concentration-time profile by protocol-dose stratum",
       caption = paste("Structurally replicates Figure 1 of Taylor 2026;",
                       "median with 10th-90th percentile band."))
```

Taylor 2026 Figure 2 plots the predicted probability of any-stage AKI against
the 4 h methotrexate concentration from a logistic regression. That regression's
intercept is not published (only the odds ratios), so the curve itself cannot be
reproduced; the exposure-response numbers that *are* published are used below
instead.

# PKNCA validation

For a 4 h zero-order infusion with no absorption phase, Cmax occurs exactly at
the end of the infusion. The paper's "model-estimated 4 h methotrexate
concentration" (Table 4) is therefore the same quantity as Cmax, which lets the
published table be used as an NCA reference.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 record per subject; for an infusion starting at t = 0 the
# pre-dose concentration is 0.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "umol/L", timeu = "h")

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "umol")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, auclast = TRUE,
  half.life = TRUE, clast.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result)
```

## Structural identity: AUC0-inf x CL = Dose

For any linear model, extrapolated exposure times clearance recovers the dose
exactly, per subject and independent of the infusion duration or the covariate
values. Both sides of this comparison use the same drawn parameters, so the
discrepancy is pure numerical error and the bound is tight.

```{r auc-identity}
auc_chk <- nca_tbl |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(id, arm, auc = PPORRES) |>
  dplyr::inner_join(
    sim |> dplyr::group_by(id) |> dplyr::summarise(cl = dplyr::first(cl),
                                                   .groups = "drop"),
    by = "id"
  ) |>
  dplyr::inner_join(dose_df |> dplyr::select(id, amt), by = "id") |>
  dplyr::mutate(pct_diff = 100 * (auc * cl / amt - 1))

c(n = nrow(auc_chk),
  max_abs_pct = max(abs(auc_chk$pct_diff)),
  median_pct  = median(auc_chk$pct_diff))

stopifnot(
  nrow(auc_chk) == 2L * N_PER_ARM,   # a lookup that matched nothing must not pass
  max(abs(auc_chk$pct_diff)) < 0.5
)
```

## Tmax lands exactly at the end of the infusion

```{r tmax-identity}
tmax_vals <- nca_tbl$PPORRES[nca_tbl$PPTESTCD == "tmax"]

stopifnot(
  length(tmax_vals) == 2L * N_PER_ARM,
  all(abs(tmax_vals - 4) < 1e-9)
)

```

## Terminal phase: quantifiability and lambda-z

The simulated profiles run to 96 h, the last of the paper's protocol sampling
times. By then a large share of subjects have fallen below the LC-MS lower limit
of detection quoted in Sect. 2.1 (10 ng/mL, i.e. about 0.02 umol/L). That is a
faithful reflection of the real study rather than a defect: sampling there ran
"every 6-24 h until the patient achieves discharge status", and discharge is
itself triggered by methotrexate falling to a low concentration, so the real
profiles also terminate rather than continue to an assay-limited tail.

```{r lloq-fraction}
clast <- nca_tbl$PPORRES[nca_tbl$PPTESTCD == "clast.obs"]

c(n_subjects           = length(clast),
  pct_clast_below_lloq = round(100 * mean(clast < 0.02), 1),
  median_clast_umol_L  = signif(median(clast), 3))
```

Because the packaged model is solved without residual error, a sub-LLOQ tail is
not solver noise and does not corrupt lambda-z. The question that matters for
the NCA above is how much leverage the extrapolated tail actually has on
`aucinf.obs`, and the answer is almost none.

```{r auc-extrapolation}
extrap <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("auclast", "aucinf.obs")) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(pct_extrap = 100 * (1 - auclast / aucinf.obs))

c(median_pct_extrapolated = signif(median(extrap$pct_extrap), 3),
  max_pct_extrapolated    = signif(max(extrap$pct_extrap), 3))

stopifnot(
  nrow(extrap) == 2L * N_PER_ARM,
  # Under 2% extrapolated for every subject, so the AUC identity checked above
  # is a statement about the observed profile, not about the lambda-z fit.
  max(extrap$pct_extrap) < 2
)
```

Finally, the NCA terminal half-life must recover the analytic terminal rate
constant of the two-compartment system, computed from each subject's *own*
`cl` and `vp` (both of which carry etas). Both sides use the same drawn
parameters, so the only sources of difference are numerical integration and
PKNCA's automatic choice of lambda-z window.

```{r half-life-identity}
beta_i <- sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), vc = dplyr::first(vc),
                   q  = dplyr::first(q),  vp = dplyr::first(vp),
                   .groups = "drop") |>
  dplyr::mutate(
    k10 = cl / vc, k12 = q / vc, k21 = q / vp,
    sk  = k10 + k12 + k21,
    beta           = (sk - sqrt(sk^2 - 4 * k10 * k21)) / 2,
    thalf_analytic = log(2) / beta
  )

hl_chk <- nca_tbl |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::select(id, thalf_nca = PPORRES) |>
  dplyr::inner_join(beta_i, by = "id") |>
  dplyr::mutate(pct_diff = 100 * (thalf_nca / thalf_analytic - 1))

c(n            = nrow(hl_chk),
  median_pct   = signif(median(hl_chk$pct_diff), 3),
  q95_abs_pct  = signif(quantile(abs(hl_chk$pct_diff), 0.95), 3))

stopifnot(
  nrow(hl_chk) == 2L * N_PER_ARM,   # a lookup that matched nothing must not pass
  # Centre: a mis-transcribed Q or V2 would move the whole distribution.
  abs(median(hl_chk$pct_diff)) < 2,
  # Envelope: robust to which subjects sample a fast lambda-z window.
  quantile(abs(hl_chk$pct_diff), 0.95) < 4
)
```

The NCA half-life sits a fraction of a percent below the analytic value across
the cohort. The small negative bias is expected and is a property of the NCA,
not of the model: PKNCA's lambda-z window reaches back into concentrations that
still carry a trace of the distribution phase, which steepens the fitted slope
slightly. The hybrid rate constants of the packaged ODE system are therefore
correct.

## Comparison against the published 4 h concentrations

Table 4 reports the **median of the per-patient post hoc (empirical Bayes)
estimates** of the 4 h concentration, computed on each patient's own delivered
dose and longitudinal eGFR. That statistic cannot be reconstructed exactly
without the individual data set. The reproducible comparison is a deterministic
one: solve the typical-value model at each stratum's published median dose and
median covariates, and read off Cmax. This has no Monte Carlo noise, so it is
stable across rxode2 versions, and it is a genuine gate on units and scale - a
umol/mg mix-up moves it by a factor of 454, a wrong V1 or a wrong dose by a
factor of two.

```{r reference-subjects}
ev_arms <- dplyr::bind_rows(lapply(seq_len(nrow(tab3)), function(i) {
  row <- tab3[i, ]
  dplyr::bind_rows(
    tibble::tibble(id = i, arm = row$arm, time = 0,
                   amt = g_to_umol(row$dose_med), evid = 1L, cmt = "central",
                   dur = 4, CRCL = row$crcl_med, ALB = row$alb_med * 10),
    tibble::tibble(id = i, arm = row$arm, time = t_obs, amt = NA_real_,
                   evid = 0L, cmt = "central", dur = NA_real_,
                   CRCL = row$crcl_med, ALB = row$alb_med * 10)
  )
})) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_arms <- rxode2::rxSolve(mod_typ, events = ev_arms, keep = "arm",
                            omega = NA) |>
  as.data.frame()

# One subject per arm, no IIV: cl must be a single deterministic value per arm.
cl_per_arm <- sim_arms |>
  dplyr::group_by(id) |>
  dplyr::summarise(n_cl = dplyr::n_distinct(round(cl, 8)), .groups = "drop")

stopifnot(
  dplyr::n_distinct(sim_arms$id) == nrow(tab3),
  all(cl_per_arm$n_cl == 1L)
)

arms_nca <- sim_arms |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Same defensive time-zero guarantee as the cohort NCA above.
arms_nca <- dplyr::bind_rows(
  arms_nca,
  arms_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

nca_arms <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(arms_nca,
                   Cc ~ time | arm + id, concu = "umol/L", timeu = "h"),
  PKNCA::PKNCAdose(ev_arms |> dplyr::filter(evid == 1L) |>
                     dplyr::select(id, time, amt, arm),
                   amt ~ time | arm + id, doseu = "umol"),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         aucinf.obs = TRUE, half.life = TRUE)
))

published <- tab4 |>
  dplyr::transmute(arm, cmax = conc_med, tmax = 4)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_arms,
  reference     = published,
  by            = "arm",
  units         = c(cmax = "umol/L", tmax = "h"),
  tolerance_pct = 25
)

knitr::kable(
  cmp,
  caption = paste("Typical-value simulation at each stratum's published median",
                  "dose and covariates vs Taylor 2026 Table 4.",
                  "* differs from reference by >25%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r reference-assertions}
cmax_cmp <- cmp[grepl("Cmax", cmp[[1]], fixed = TRUE), ]
pct <- as.numeric(gsub("[^0-9.+-]", "", cmax_cmp[["% diff"]]))

stopifnot(
  nrow(cmax_cmp) == 2L,        # guard: the lookup must actually have matched
  all(is.finite(pct)),
  all(abs(pct) < 25)
)
```

The typical-value Cmax runs about 10-20% above the published medians in both
strata, in the same direction and of similar magnitude. Candidate explanations,
none of which can be settled from the published material and none of which was
used to adjust any parameter:

* **The statistics are not the same quantity.** Table 4 is the median of 37-43
  individual post hoc predictions, each on its own dose and its own longitudinal
  eGFR trajectory; the deterministic run is a single prediction at the median
  dose and the median *baseline* covariates. Baseline eGFR (82.5 and
  80.2 mL/min/1.73 m^2) sits above the longitudinal median of 76 that centers
  Eq. 1, which is a real difference but only worth a few percent.
* **Sampling time.** Post-infusion the distribution half-life is
  `r sprintf("%.1f", log(2) / alpha)` h, so concentrations fall steeply just
  after 4 h. The offsets implied by the two strata are computed below and are
  both under an hour, which is within ordinary clinical sampling slack.
* **No body-size term.** The model carries no weight or BSA covariate on any
  parameter, yet dose scales with BSA. Patients receiving the largest doses are
  the largest patients, and a fixed 42.2 L central volume cannot track that.

```{r implied-offset}
implied_offset <- vapply(seq_len(nrow(tab4)), function(i) {
  prof <- sim_arms[sim_arms$id == i & sim_arms$time >= 4, ]
  prof$time[which.min(abs(prof$Cc - tab4$conc_med[i]))] - 4
}, numeric(1))

tibble::tibble(arm = tab4$arm,
               published_umol_L = tab4$conc_med,
               implied_delay_min = round(60 * implied_offset)) |>
  knitr::kable(caption = paste("Post-4 h delay that would reconcile the",
                               "deterministic simulation with Table 4."))
```

Both strata imply a delay under an hour, which makes the sampling-time
explanation internally coherent - but it remains a hypothesis, not a finding.

## Individual clearance against the published post hoc medians

```{r cl-comparison}
cl_cmp <- tab3 |>
  dplyr::mutate(
    cl_simulated = 11.8 * (crcl_med / 76)^0.80 * (alb_med / 4)^0.69,
    pct_diff     = 100 * (cl_simulated / cl_med - 1)
  ) |>
  dplyr::select(arm, cl_med, cl_simulated, pct_diff) |>
  dplyr::rename(
    "Stratum"                            = arm,
    "Published median post hoc CL (L/h)" = cl_med,
    "Eq. 1 at median covariates (L/h)"   = cl_simulated,
    "% diff"                             = pct_diff
  )

knitr::kable(cl_cmp, digits = 2,
             caption = "Taylor 2026 Table 3 post hoc clearance vs Eq. 1.")

stopifnot(nrow(cl_cmp) == 2L, all(abs(cl_cmp[["% diff"]]) < 8))
```

Evaluating Eq. 1 at each stratum's median baseline covariates lands within about
4% of the published median post hoc clearance in both strata. Clearance - the
parameter this paper is actually about - is transcribed correctly.

# Exposure-response context

Taylor 2026 identifies 160 umol/L as the optimal 4 h discriminatory threshold
for any-stage AKI in the `<=` 3.5 g/m^2 stratum, with sensitivity 0.77,
specificity 0.87, and an odds ratio of 22 (95% CI 3.7-89.4). Those four numbers
over-determine the 2x2 table, so reconstructing it and recovering the published
odds ratio confirms the exposure-response figures have been read correctly.

```{r roc-consistency}
n_aki  <- 13L; n_no_aki <- 30L  # <= 3.5 g/m2 stratum: 13 of 43 developed AKI
sens <- 0.77; spec <- 0.87

tp <- round(sens * n_aki); fn <- n_aki - tp
fp <- round((1 - spec) * n_no_aki); tn <- n_no_aki - fp
or_reconstructed <- (tp * tn) / (fp * fn)

c(above_threshold = tp + fp, of_n = n_aki + n_no_aki,
  odds_ratio_reconstructed = round(or_reconstructed, 1),
  odds_ratio_published = 22)

stopifnot(abs(or_reconstructed - 22) < 3)
```

The reconstruction recovers the published odds ratio, and implies that roughly
`r round(100 * (round(0.77 * 13) + round(0.13 * 30)) / 43)`% of that stratum
exceeded the threshold. For reference, in the simulated `<=` 3.5 g/m^2 cohort:

```{r threshold-fraction}
frac_over <- sim |>
  dplyr::filter(arm == "<=3.5 g/m2", abs(time - 4) < 1e-9) |>
  dplyr::summarise(pct_over_160 = 100 * mean(Cc > 160)) |>
  dplyr::pull(pct_over_160)

round(frac_over, 1)
```

The simulated fraction is higher than the reconstructed one, by the same margin
and for the same reasons as the Cmax comparison above. This is context, not a
validation of the logistic-regression sub-model, which is not part of the
packaged PK model.

# Assumptions and deviations

* **Molecular weight.** Doses are published in grams and concentrations in
  umol/L, so a molecular weight is needed to connect them. `MW_MTX = 454.44`
  g/mol is a physical constant, not a Taylor 2026 value; it matches the constant
  already used by `Taylor_2020_methotrexate.R` in this package. The model file
  itself is unit-agnostic - it declares `dosing = "umol"` and the conversion
  lives only in this vignette.
* **eGFR held constant per subject.** Taylor 2026 modelled the CKD-EPI
  creatinine-cystatin C eGFR as a **longitudinal** covariate, re-derived at each
  sampling time. The per-patient trajectories appear only as an unlabelled
  spaghetti plot (ESM Fig. S2E) with no tabulated values, so each simulated
  subject carries a single baseline eGFR for the whole profile. This is the main
  structural simplification in this vignette;
  `CRCL` is a time-varying covariate in the model and accepts a per-record
  column when trajectories are available.
* **Albumin reference and units.** Eq. 1 centers albumin on 4 g/dL. The
  canonical register unit for `ALB` is SI g/L, so `model()` applies
  `alb_gdL <- ALB * 0.1` and the cohort supplies g/L. Taylor 2026 Table 3 heads
  the albumin row "(mg/L)" while Table 1 and Eq. 1 use g/dL; the Table 3 header
  is a units typo - 3.9 mg/L is not a physiological albumin concentration.
* **eGFR units.** Table 1 heads the renal block "Estimated kidney function
  (mL/min)", pooling the raw Cockcroft-Gault estimate with the three
  BSA-normalized CKD-EPI equations. Sect. 2.2.2 is explicit that the CKD-EPI
  values are mL/min/1.73 m^2, and that is the unit recorded for `CRCL`.
* **Covariate distributions.** Delivered dose, eGFR and albumin are drawn from
  log-normal distributions fitted to the published median / IQR pairs of
  Table 3. The paper does not report the distributional shapes, and dose,
  weight and renal function are certainly correlated in the real cohort; the
  simulated covariates are drawn independently.
* **Residual error scale.** Taylor 2026 fitted on log-transformed
  concentrations, so the reported residual variance of 0.20 is additive on the
  log scale. That maps to a proportional error in nlmixr2's linear space, giving
  `propSd = sqrt(0.20) = 0.447`. Under a strictly log-normal reading the
  coefficient of variation would instead be `sqrt(exp(0.20) - 1) = 0.471`; the
  proportional form is the package convention for log-transformed-both-sides
  models and the difference is under 3 percentage points of CV.
* **Supplement.** The ESM (`40262_2026_1618_MOESM1_ESM.docx`) was retrieved and
  reviewed. It holds no parameter values beyond those in Table 2, but it does
  independently corroborate both centering constants: Table S2 records the
  tested covariate forms as `(eGFR/76)^theta` (model 8, the retained one) and
  `(Alb/4)^theta` (model 3), matching Eq. 1's printed denominators. Table S2
  also disambiguates a phrasing in the Results text: the "OFV = -25.5 vs 13.5"
  quoted for cystatin C versus creatinine are *absolute* objective function
  values, whose changes against the base model are -81.5 and -42.9. Figures
  S1-S4 are images (concentration-time spaghetti, longitudinal renal-function
  trajectories, goodness-of-fit, AKI histogram) and carry no extractable
  parameter.
* **Not externally validated.** The paper states the PK model and the logistic
  regression did not undergo external validation, and the exposure-response
  models "were not intended for individual risk prediction".
* **Exposure-response sub-model not packaged.** The logistic regression relating
  the 4 h concentration to AKI is not part of the PK model and is not
  implemented here; its intercept is not published.
