---
title: "Lumefantrine PK and 28-day recurrence (Tchaparian 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lumefantrine PK and 28-day recurrence (Tchaparian 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Tchaparian 2016 reports two models, extracted here as two files and validated in
this single vignette.

* `Tchaparian_2016_lumefantrine` -- the two-compartment population PK model
  (Table 2).
* `Tchaparian_2016_lumefantrine_recurrence` -- the adjusted Cox
  proportional-hazards model for 28-day recurrent parasitaemia (Table 3), plus
  the paper's comparative day-3-versus-day-7 exposure analysis.

* Citation: `r rxode2::rxode2(readModelDb("Tchaparian_2016_lumefantrine"))$reference`
* Article: [J Infect Dis. 2016;214(8):1243-1251](https://doi.org/10.1093/infdis/jiw338)
* Open-access record: [PMC5034953](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5034953/)
* Supplement: retrieved from the EuropePMC supplementary-files endpoint for
  PMC5034953 (`supp_jiw338_jiw338supp.docx`). It is load-bearing for this
  extraction -- see "The supplement settles the omega scale" below.

```{r load-models}
pk <- rxode2::rxode2(readModelDb("Tchaparian_2016_lumefantrine"))
pd <- rxode2::rxode2(readModelDb("Tchaparian_2016_lumefantrine_recurrence"))
```

## The supplement settles the omega scale

Table 2's variability columns are headed only `IIV, %` and `IOV (%)`, which is
exactly the ambiguity that ships as a wrong model. The supplement resolves it in
so many words:

> "Due to the sparsity of the data, IIV and IOV of Q/F and V2/F were each fixed
> to be 50% (**CV**)."

and the Results text corroborates it for clearance ("2.19 L/h, with 8% (**CV**)
interindividual variability"). Both columns are therefore log-normal **CV%**,
not variances, so every entry is converted as `omega^2 = log(CV^2 + 1)`. Reading
them as variances instead would have shrunk the coefficient of variation on
bioavailability from 34.9% to 59% on the log scale -- a different model.

The supplement also supplies three further facts used above: that correlation
among CL/F, V1/F, Q/F and V2/F was induced through variability on F rather than
through an estimated covariance block (which is why this model has no `ini()`
correlation block); that the residual was combined additive plus proportional;
and that the final model centred weight and age on a one-year-old "for ease of
interpretation", which is what anchors the Table 2 estimates at 8.43 kg and 12
months rather than at the cohort median of 9.0 kg.

# Population

`r rxode2::rxode2(readModelDb("Tchaparian_2016_lumefantrine"))$description`

The PK analysis set is 101 Ugandan children aged 6 months to 2 years, treated
for 207 malaria episodes and contributing 806 evaluable capillary whole-blood
concentrations; the outcomes analysis set is 100 children over 222 episodes.
Baseline characteristics are Table 1. Every child weighed under 14 kg and so
received the identical regimen: one artemether-lumefantrine tablet (20 mg
artemether plus 120 mg lumefantrine) twice daily for three days, six doses and
720 mg of lumefantrine in total. Morning doses were given in clinic with
fat-containing milk to standardise the well-known food effect on lumefantrine
absorption; evening doses were given at home.

Two features of the analysis matter for reading everything below.

**The matrix is capillary whole blood on filter paper, not venous plasma.** The
assay had an LLOQ of 132 ng/mL and an LOD of 52 ng/mL; 188 of 806 samples
(23.3%) fell below the LOD and were retained as left-censored rather than
discarded, which is the only reason the model can describe the 14-day tail at
all. The authors devote a Discussion paragraph to the point that venous plasma
runs higher, so the apparent volumes and exposures here are not directly
comparable with the plasma-based lumefantrine models elsewhere in this library
(`Hoglund_2015_lumefantrine`, `Kloprogge_2018_lumefantrine`,
`Ding_2026_lumefantrine`).

**The recurrence endpoint is dominated by reinfection, not recrudescence.** At
63 days, 95.9% of recurrences genotyped as new infection. Tororo district has an
entomological inoculation rate up to 562 infective bites per person-year, so the
hazard modelled below should be read as post-treatment prophylaxis rather than
as a cure rate.

# Source trace

Every `ini()` value and every non-obvious `model()` equation, with the place in
the source it came from.

```{r source-trace, echo = FALSE}
trace_pk <- tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "lcl", "log(2.19) L/h", "Table 2, CL/F point estimate (RSE 8%); Results repeats 'CL/F in a typical 12-month-old child is 2.19 L/h'",
  "lvc", "log(83.2) L", "Table 2, V1/F point estimate per 8.43 kg (RSE 7%)",
  "lq", "log(0.23) L/h", "Table 2, Q/F point estimate (RSE 35%)",
  "lvp", "log(441) L", "Table 2, V2/F point estimate per 8.43 kg (RSE 74%)",
  "lka (fixed)", "log(0.45) 1/h", "Table 2 footnote c; Methods: ka 'fixed to the previously reported value of 0.45 h-1'",
  "lfdepot (fixed)", "log(1)", "Table 2 footnote c, F = 1 -- the apparent-parameter anchor",
  "e_wt_cl_q (fixed)", "0.75", "Table 2 footnote a: 'Pk * [WT/8.43]^0.75'; Methods imposes it unconditionally",
  "e_wt_vc_vp (fixed)", "1", "Methods: volumes multiplied by [weight/reference weight]",
  "e_age_fdepot", "0.596", "Table 2 'Age effect' (RSE 38%), footnote b; Results equation F_i = [1 * Age_i/12]^0.596",
  "WT_REF", "8.43 kg", "Table 2 footnote a; Supplement: final model centred on a one-year-old",
  "AGE_REF_MO", "12 months", "Results covariate equation; Table 2 footnote b",
  "etalcl", "log(0.08^2+1) = 0.00638", "Table 2 CL/F IIV 8% CV (RSE 174)",
  "etalvc", "log(0.152^2+1) = 0.02284", "Table 2 V1/F IIV 15.2% CV (RSE 46)",
  "etalq, etalvp (fixed)", "log(0.5^2+1) = 0.22314", "Table 2 footnote c; Supplement: 'IIV and IOV of Q/F and V2/F were each fixed to be 50% (CV)'",
  "etalfdepot", "log(0.349^2+1) = 0.11494", "Table 2 F row, IIV 34.9% CV (RSE 23)",
  "etaiov_cl_*", "log(0.093^2+1) = 0.00861", "Table 2 CL/F IOV 9.3% (RSE 105)",
  "etaiov_vc_*", "log(0.086^2+1) = 0.00737", "Table 2 V1/F IOV 8.6% (RSE 133)",
  "etaiov_q_*, etaiov_vp_* (fixed)", "log(0.5^2+1) = 0.22314", "Table 2 footnote c; Supplement as above",
  "etaiov_fdepot_*", "log(0.594^2+1) = 0.30220", "Table 2 F row, IOV 59.4% (RSE 8)",
  "propSd", "0.378", "Table 2 Proportional 37.8% (RSE 5); Results names it a standard deviation",
  "addSd", "19.2 ng/mL", "Table 2 Additive 19.2 ng/mL (RSE 18)",
  "Cc = 1000 * central / vc", "unit conversion", "Dose in mg and vc in L give mg/L = ug/mL; Table 2 and Figures 1-2 report ng/mL"
)
knitr::kable(trace_pk, caption = "Source trace, population PK model (Tchaparian_2016_lumefantrine).")
```

```{r source-trace-pd, echo = FALSE}
trace_pd <- tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "conc_cut (fixed)", "200 ng/mL", "Table 3 row heading '<200 ng/mL'; Results ROC analysis, area under the curve 0.684",
  "e_conc_lumefantrine_168h_low_haz_nots", "log(2.97)", "Table 3, 'Without TMP-SMZ use' block: adjusted HR 2.97 (1.59-5.55), P = .0007",
  "e_conc_lumefantrine_168h_low_haz_ts", "log(0.50)", "Table 3, 'With TMP-SMZ use' block: adjusted HR 0.50 (0.22-1.13), P = .10",
  "e_conmed_tmpsmx_haz", "log(0.75)", "Table 3, 'TMP-SMZ use / Yes': adjusted HR 0.75 (0.43-1.29), P = .30",
  "age_cut1, age_cut2 (fixed)", "1.0, 1.5 years", "Table 3 age strata 6-12, 12-18 and >=18 months",
  "e_age_haz_12to18", "log(1.87)", "Table 3, age 12-18 mo: adjusted HR 1.87 (0.83-4.22), P = .13",
  "e_age_haz_ge18", "log(1.53)", "Table 3, age >=18 mo: adjusted HR 1.53 (0.64-3.63), P = .34",
  "e_residence_rural_haz", "log(1.21)", "Table 3, 'Residence / Rural': adjusted HR 1.21 (0.43-3.43), P = .71",
  "hgb_ref (fixed)", "10.0 g/dL", "Table 1 PK-outcomes column, haemoglobin median (centring value; see deviations)",
  "e_hgb_haz", "log(0.93)", "Table 3 haemoglobin, footnote a continuous: adjusted HR 0.93 (0.83-1.05), P = .25",
  "para_ref (fixed)", "17603 parasites/uL", "Table 1 PK-outcomes column, geometric mean (centring value; see deviations)",
  "e_para_haz", "log(1.04)", "Table 3 parasite density, footnote a continuous: adjusted HR 1.04 (0.95-1.14), P = .35",
  "waz_cut (fixed)", "-2", "Methods: 'a weight-for-age z score with a -2 cutoff', WHO standards; Table 3 footnote b",
  "e_waz_underweight_haz", "log(1.04)", "Table 3, 'Underweight / Yes': adjusted HR 1.04 (0.47-2.31), P = .93",
  "conc_day3_ref (fixed)", "2777 ng/mL", "Results: day 3 median 2777 ng/mL (IQR 1672-4760, n = 187)",
  "conc_day7_ref (fixed)", "216 ng/mL", "Results: day 7 median 216 ng/mL (IQR 136-345, n = 216)",
  "e_conc_lumefantrine_72h_haz", "log(0.51)", "Results: '49% reduced hazard' per ln unit on day 3, P = .002; Discussion prints 0.51",
  "e_conc_lumefantrine_168h_haz", "log(0.80)", "Results: '20% reduced hazard' per ln unit on day 7, P = .002; Discussion prints 0.80"
)
knitr::kable(trace_pd, caption = "Source trace, recurrence model (Tchaparian_2016_lumefantrine_recurrence). All coefficients are natural logs of the printed hazard ratios, which is the scale a Cox model estimates on.")
```

# Structural checks on the PK model

These run on the typical individual with the random effects zeroed, so they are
deterministic: no cohort is drawn and the tolerances are numerical rather than
statistical.

## The ODEs really do integrate the published two-compartment model

The model names its parameters `cl` / `vc` / `q` / `vp`, which in rxode2 can
trigger the solved-system path and silently discard an explicitly written
`d/dt()`. It does not here -- but that is worth proving rather than assuming, so
the solve is compared against the closed-form oral two-compartment
biexponential built independently from the same numbers.

```{r closed-form}
pk0 <- rxode2::zeroRe(pk)
stopifnot(identical(pk$state, c("depot", "central", "peripheral1")))
stopifnot(isFALSE(pk$predDf$linCmt[1]))

wt_med <- 9.1
age_med_mo <- 14.4
dose_mg <- 120

ev1 <- rxode2::et(amt = dose_mg, cmt = "depot", time = 0) |>
  rxode2::et(seq(0, 1000, by = 0.05), cmt = "central")
ev1 <- as.data.frame(ev1)
ev1$WT <- wt_med
ev1$AGE <- age_med_mo / 12
ev1$OCC <- 1

s1 <- rxode2::rxSolve(pk0, ev1, returnType = "data.frame", atol = 1e-12, rtol = 1e-10)
s1 <- s1[!is.na(s1$Cc) & s1$time > 0, ]

# Independent closed form, built from the Table 2 numbers directly.
cl_i <- 2.19 * (wt_med / 8.43)^0.75
vc_i <- 83.2 * (wt_med / 8.43)
q_i <- 0.23 * (wt_med / 8.43)^0.75
vp_i <- 441 * (wt_med / 8.43)
ka_i <- 0.45
f_i <- (age_med_mo / 12)^0.596
kel_i <- cl_i / vc_i
k12_i <- q_i / vc_i
k21_i <- q_i / vp_i
bsum <- kel_i + k12_i + k21_i
bprod <- kel_i * k21_i
alpha <- (bsum + sqrt(bsum^2 - 4 * bprod)) / 2
beta <- (bsum - sqrt(bsum^2 - 4 * bprod)) / 2
amp <- f_i * dose_mg * ka_i / vc_i
analytic <- 1000 * amp * (
  (k21_i - alpha) / ((ka_i - alpha) * (beta - alpha)) * exp(-alpha * s1$time) +
    (k21_i - beta) / ((ka_i - beta) * (alpha - beta)) * exp(-beta * s1$time) +
    (k21_i - ka_i) / ((alpha - ka_i) * (beta - ka_i)) * exp(-ka_i * s1$time)
)
rel_err <- max(abs(s1$Cc - analytic) / analytic)

# Pure numerical agreement between two exact forms of the same system, so this
# bound is a machine-precision bound and is deliberately tight. Realised
# 1.1e-11; anything above 1e-6 means the ODEs are not the system above.
stopifnot(rel_err < 1e-6)
sprintf("closed-form biexponential agreement: max relative error %.2e over %d points", rel_err, nrow(s1))
```

## Dose, clearance, bioavailability and units are mutually consistent

`CL/F * AUC(0-inf) = F * Dose` must hold exactly for a linear model with a
single input path. This is one gate that simultaneously exercises the
transcribed clearance, the allometric exponent, the age-on-F power function, the
720 mg course and the ng/mL unit conversion.

```{r auc-identity}
ev2 <- rxode2::et(amt = dose_mg, cmt = "depot", time = seq(0, 60, by = 12)) |>
  rxode2::et(seq(0, 20000, by = 0.5), cmt = "central")
ev2 <- as.data.frame(ev2)
ev2$WT <- wt_med
ev2$AGE <- age_med_mo / 12
ev2$OCC <- 1
s2 <- rxode2::rxSolve(pk0, ev2, returnType = "data.frame", atol = 1e-12, rtol = 1e-10)
s2 <- s2[!is.na(s2$Cc), ]
auc_numeric <- sum(diff(s2$time) * (head(s2$Cc, -1) + tail(s2$Cc, -1)) / 2) / 1000
auc_identity <- f_i * 6 * dose_mg / cl_i

stopifnot(abs(auc_numeric / auc_identity - 1) < 1e-3)
sprintf(
  "AUC(0-inf): numeric %.2f vs F*Dose/CL %.2f ug*h/mL (relative difference %.1e); published median 347.8",
  auc_numeric, auc_identity, abs(auc_numeric / auc_identity - 1)
)
```

## The age-on-bioavailability function reproduces the Discussion

The Discussion states the model "predicts a point estimate of relative
bioavailability of 0.7 for a 6-month-old child as compared to 1.5 in a
2-year-old child, with a value of 1 for the reference 1-year-old population".
Those three numbers are an independent check on the 0.596 exponent and on the
12-month reference age.

```{r f-age}
f_at <- function(months) (months / 12)^0.596
f_tab <- tibble::tibble(
  `Age` = c("6 months", "12 months (reference)", "24 months"),
  `Model F` = round(f_at(c(6, 12, 24)), 3),
  `Discussion` = c(0.7, 1.0, 1.5)
)
knitr::kable(f_tab, caption = "Relative bioavailability versus age (Tchaparian 2016 Discussion).")

# The paper rounds to one decimal, so an absolute tolerance on the rounding step
# is the right form here (see also the LOD check below).
stopifnot(all(abs(f_at(c(6, 12, 24)) - c(0.7, 1.0, 1.5)) < 0.05))
```

## Time to the limit of detection

Results states that "half of all children are predicted to reach the LOD
(52 ng/mL) in approximately 10 days". This is the only statement the paper makes
about the shape of the terminal decline, so it is the only available check on
`Q/F` and `V2/F` -- the two least precisely estimated parameters in Table 2
(RSE 35% and 74%).

```{r lod}
s_lod <- s2[s2$time > 100, ]
t_lod <- s_lod$time[which(s_lod$Cc < 52)[1]] / 24

stopifnot(abs(t_lod - 10) < 1.5)
sprintf("typical child crosses the 52 ng/mL LOD at %.2f days (paper: approximately 10 days)", t_lod)
```

# Virtual cohort

The cohort is calibrated to Tchaparian 2016 Table 1 rather than to an external
growth standard, because the model's two covariates are weight and age and it is
the paper's own marginal distributions that the published exposure summaries are
conditioned on. Age is drawn uniformly across the observed range and weight from
a weight-for-age line anchored on the cohort median, with log-normal scatter
tuned so that the simulated weight median and range match the published ones.
This cohort runs visibly lighter than the WHO median for age -- the paper reports
a median weight-for-age z score of -0.74 -- which is the intended behaviour.

Cohort size is 200, the per-arm cap for this repository; a single treatment arm
is simulated because every child in the study received the same regimen.

```{r cohort}
rxode2::rxSetSeed(20260922L)
set.seed(20260922L)
n_sub <- 200L

age_mo <- runif(n_sub, 6.6, 22.2)
wt_kg <- (6.95 + 0.149 * age_mo) * exp(rnorm(n_sub, 0, 0.115))
wt_kg <- pmin(pmax(wt_kg, 6.1), 13.0)

cohort <- tibble::tibble(id = seq_len(n_sub), WT = wt_kg, AGE = age_mo / 12, OCC = 1)

cohort_tab <- tibble::tibble(
  Characteristic = c("Weight, kg", "Age, months"),
  Simulated = c(
    sprintf("%.1f (%.1f-%.1f)", median(wt_kg), min(wt_kg), max(wt_kg)),
    sprintf("%.1f (%.1f-%.1f)", median(age_mo), min(age_mo), max(age_mo))
  ),
  `Table 1` = c("9.1 (6.1-13.0)", "14.4 (6.6-22.2)")
)
knitr::kable(cohort_tab, caption = "Simulated cohort versus Tchaparian 2016 Table 1, median (range).")

# Gate the cohort itself: if the covariate distributions drift, every exposure
# comparison below becomes uninterpretable. Bounds are on the median only --
# the extremes of a random cohort are not reproducible across rxode2 builds.
stopifnot(
  abs(median(wt_kg) / 9.1 - 1) < 0.10,
  abs(median(age_mo) / 14.4 - 1) < 0.10
)
```

# Simulation

Six doses of 120 mg at 0, 12, 24, 36, 48 and 60 h, matching the study's
twice-daily three-day course. The observation grid is log-spaced through the
dosing period so that the peak is resolved, then dense to 96 h and coarser out
to 1512 h (63 days, the paper's secondary follow-up horizon).

Note the landmark arithmetic: Methods define day 3 as "12 hours after the last
dose" and day 7 as "108 hours after the last dose". With the last of the six
doses at 60 h, the two landmarks are **72 h** and **168 h** after the first dose.

```{r simulate}
obs_times <- sort(unique(c(
  exp(seq(log(0.05), log(60), length.out = 60)),
  seq(0, 96, by = 1),
  seq(100, 1512, by = 4)
)))

ev_one <- rxode2::et(amt = dose_mg, cmt = "depot", time = seq(0, 60, by = 12)) |>
  rxode2::et(obs_times, cmt = "central")
ev_one <- as.data.frame(ev_one)

# Covariates must be attached to a materialized data frame: assigning them to
# an rxEt object silently drops them.
events <- do.call(rbind, lapply(seq_len(n_sub), function(i) {
  one <- ev_one
  one$id <- cohort$id[i]
  one$WT <- cohort$WT[i]
  one$AGE <- cohort$AGE[i]
  one$OCC <- cohort$OCC[i]
  one
}))

sim <- rxode2::rxSolve(pk, events, returnType = "data.frame")
obs <- sim |> dplyr::filter(!is.na(Cc))

per_id <- obs |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cmax = max(Cc),
    tmax = time[which.max(Cc)],
    c72 = Cc[which.min(abs(time - 72))],
    c168 = Cc[which.min(abs(time - 168))],
    cl_i = dplyr::first(cl),
    f_ind = dplyr::first(fdepot),
    .groups = "drop"
  ) |>
  dplyr::mutate(auc_model = f_ind * 6 * dose_mg / cl_i)
```

`Cc` is the individual prediction and carries no residual error, which is the
right quantity to compare against the paper's model-predicted summaries.

# Replicating Figure 2

Figure 2 of Tchaparian 2016 shows the predicted concentration distribution
against the observed data, with the median and nested prediction intervals
shaded. Below is the same construction from the packaged model, with the
assay LOD and LLOQ and the 200 ng/mL efficacy threshold marked.

```{r figure2, fig.cap = "Replicates Figure 2 of Tchaparian 2016: predicted capillary whole-blood lumefantrine concentration versus time, median with 50% and 90% prediction intervals. Horizontal lines mark the assay LOD (52 ng/mL), the LLOQ (132 ng/mL) and the 200 ng/mL day-7 efficacy threshold; vertical lines mark the day 3 and day 7 landmarks."}
band <- obs |>
  dplyr::filter(time > 0, time <= 504) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    lo90 = quantile(Cc, 0.05), lo50 = quantile(Cc, 0.25),
    med = median(Cc),
    hi50 = quantile(Cc, 0.75), hi90 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(band, aes(x = time / 24)) +
  geom_ribbon(aes(ymin = lo90, ymax = hi90), alpha = 0.18, fill = "steelblue") +
  geom_ribbon(aes(ymin = lo50, ymax = hi50), alpha = 0.35, fill = "steelblue") +
  geom_line(aes(y = med), linewidth = 0.9) +
  geom_hline(yintercept = c(52, 132, 200), linetype = c("dotted", "dotted", "dashed"),
             colour = c("grey40", "grey40", "firebrick")) +
  geom_vline(xintercept = c(3, 7), linetype = "dotdash", colour = "grey55") +
  scale_y_log10() +
  labs(
    x = "Days after the first dose",
    y = "Lumefantrine, capillary whole blood (ng/mL)"
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
sim_nca <- obs |>
  dplyr::mutate(treatment = "AL 120 mg BID x 3 d") |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time-zero record so PKNCA does not warn about an AUC interval
# starting before the first measurement. Pre-dose concentration is 0 for an
# extravascular model.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::mutate(treatment = "AL 120 mg BID x 3 d") |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, clast.obs = TRUE, lambda.z = TRUE, span.ratio = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
knitr::kable(
  as.data.frame(summary(nca_res)),
  caption = "PKNCA non-compartmental summary of the simulated cohort, median [geometric CV%] unless noted."
)
```

## Why AUC is compared on the model-based value, not on `aucinf.obs`

Table 2 footnote d states that the published AUC(0-inf) is the "median of values
calculated from individual estimates (modes) of CL and F" -- that is,
`F * Dose / CL`, not a trapezoidal NCA result. The comparison below therefore
uses the same construction, so that the two sides are matched by definition
rather than approximately.

That choice is worth a check rather than an assumption, because this model's
formal terminal phase is very slow: `Q/F` is 0.23 L/h against a `V2/F` of 441 L,
giving a beta half-life near 60 days. The 1512 h observation window is shorter
than one such half-life, so PKNCA's span ratio comes out below 1 and its
`lambda.z` is fitted to a phase the window does not fully cover.

```{r span-ratio}
nca_long <- as.data.frame(nca_res)
span_med <- median(nca_long$PPORRES[nca_long$PPTESTCD == "span.ratio"], na.rm = TRUE)
aucinf_med <- median(nca_long$PPORRES[nca_long$PPTESTCD == "aucinf.obs"], na.rm = TRUE) / 1000

sprintf(
  "median span ratio %.2f; PKNCA aucinf.obs median %.1f vs model-based F*Dose/CL median %.1f ug*h/mL (%+.2f%%)",
  span_med, aucinf_med, median(per_id$auc_model),
  100 * (aucinf_med / median(per_id$auc_model) - 1)
)

# Two separate claims, both measured across 1/2/4/16 solver threads.
#  (1) The window really is shorter than one terminal half-life. Realised
#      span ratio 0.65-0.78, so a bound of 0.95 holds while still going red if
#      the grid or the peripheral compartment changes materially.
#  (2) Despite (1), the extrapolation recovers the model-based AUC closely --
#      realised -0.09% to -0.11%, so the two constructions agree and neither is
#      biased here. A 5% bound keeps ~45x headroom yet still fails on a unit
#      or dose error, which would move it by orders of magnitude.
# Do NOT write `aucinf_med < median(per_id$auc_model)`: the two differ by ~0.1%
# and that ordering is a coin flip from one cohort draw to the next.
stopifnot(
  span_med < 0.95,
  abs(aucinf_med / median(per_id$auc_model) - 1) < 0.05
)
```

So the under-characterised terminal phase does not in fact distort the
trapezoidal result at this sampling density -- the two AUC constructions agree to
about 0.1%. `Cmax` is taken from PKNCA, where the NCA and model definitions
coincide exactly.

# Comparison against the published summaries

```{r nca-compare}
simulated_summary <- tibble::tibble(
  treatment = "AL 120 mg BID x 3 d",
  cmax = median(per_id$cmax),
  aucinf.obs = median(per_id$auc_model) * 1000,
  c72 = median(per_id$c72),
  c168 = median(per_id$c168)
)

published <- tibble::tibble(
  treatment = "AL 120 mg BID x 3 d",
  cmax = 3900,
  aucinf.obs = 347.8 * 1000,
  c72 = 2777,
  c168 = 216
)

cmp <- tibble::tibble(
  Parameter = c(
    "Cmax (ng/mL)", "AUC0-inf (ng*h/mL)",
    "Day 3 (72 h) concentration (ng/mL)", "Day 7 (168 h) concentration (ng/mL)"
  ),
  Simulated = signif(unlist(simulated_summary[, -1]), 4),
  Published = signif(unlist(published[, -1]), 4),
  `Difference, %` = round(100 * (unlist(simulated_summary[, -1]) / unlist(published[, -1]) - 1), 1),
  Source = c(
    "Discussion, 'approximately 3.9 ug/mL'",
    "Table 2 secondary parameter, median of F*Dose/CL",
    "Results, measured median (n = 187)",
    "Results, measured median (n = 216)"
  )
)
knitr::kable(cmp, caption = "Simulated cohort medians versus Tchaparian 2016. All comparisons are on the median; see the note below on why the spread is not compared.")
```

Bounds below are on the **median only**, and were set from renders at 1, 2, 4, 8
and 16 solver threads rather than from a single run. `rxSetSeed()` fixes
rxode2's stream per thread count, not across thread counts, so a bound tight
enough to sit inside that spread passes where it is written and fails in CI.

```{r nca-gate}
pct <- function(sim, ref) 100 * (sim / ref - 1)

# Realised across 1/2/4/8/16 threads:
#   AUC   +0.6 to -5.3 %     -> 20% bound, ~4x headroom
#   C168  -6.2 to +7.2 %     -> 25% bound
#   Cmax  -4.8 to -1.0 %     -> 20% bound
#   C72  +14.1 to +17.3 %    -> 25% bound; a systematic offset, see deviations
# A mis-transcribed dose, clearance, allometric exponent or unit conversion
# moves these by tens of percent, so every bound can still go red.
stopifnot(
  abs(pct(median(per_id$auc_model), 347.8)) < 20,
  abs(pct(median(per_id$c168), 216)) < 25,
  abs(pct(median(per_id$cmax), 3900)) < 20,
  abs(pct(median(per_id$c72), 2777)) < 25
)
```

The simulated **spread** is wider than the published interquartile ranges (for
example AUC 208-531 against 244-511 at 16 threads) and is deliberately not
gated. The published intervals are computed over individual *modes* -- empirical
Bayes estimates, which shrink toward the population typical value -- whereas the
simulation draws from the full variability model, which for bioavailability is
34.9% interindividual plus 59.4% interoccasion variability, about 72% CV on a
single occasion. Wider simulated spread is the expected consequence of
shrinkage, not a defect.

# The recurrence model

## Reproducing Table 3 cell by cell

The Cox model is algebraic and deterministic, so each adjusted hazard ratio in
Table 3 can be recovered exactly by perturbing one covariate away from the
reference cohort: a 6-to-under-12-month, urban-resident, not-underweight child
not receiving trimethoprim-sulfamethoxazole prophylaxis, whose day 7
concentration is at or above 200 ng/mL and whose haemoglobin and parasite
density sit at the centring values.

```{r table3}
ref_row <- tibble::tibble(
  CONC_LUMEFANTRINE_168H = 250, CONC_LUMEFANTRINE_72H = 2777,
  CONMED_TMPSMX = 0, AGE = 0.75, RESIDENCE_RURAL = 0,
  HGB = 10, PARA = 17603, WAZ = 0
)

hr_of <- function(...) {
  row <- ref_row
  changes <- list(...)
  for (nm in names(changes)) row[[nm]] <- changes[[nm]]
  rxode2::rxSolve(pd, row, rxode2::et(0), returnType = "data.frame")$hr[1]
}

table3 <- tibble::tribble(
  ~Covariate, ~Contrast, ~Model, ~`Table 3`,
  "(reference cohort)", "all covariates at reference", hr_of(), 1.00,
  "Day 7 concentration", "<200 ng/mL, without prophylaxis", hr_of(CONC_LUMEFANTRINE_168H = 150), 2.97,
  "TMP-SMZ prophylaxis", "receiving, day 7 >=200 ng/mL", hr_of(CONMED_TMPSMX = 1), 0.75,
  "Age group", "12-18 months", hr_of(AGE = 15 / 12), 1.87,
  "Age group", ">=18 months", hr_of(AGE = 20 / 12), 1.53,
  "Residence", "rural vs urban", hr_of(RESIDENCE_RURAL = 1), 1.21,
  "Haemoglobin", "+1 g/dL", hr_of(HGB = 11), 0.93,
  "Parasite density", "x e (one natural-log unit)", hr_of(PARA = 17603 * exp(1)), 1.04,
  "Underweight", "weight-for-age z below -2", hr_of(WAZ = -2.5), 1.04
) |>
  dplyr::mutate(Model = round(Model, 4))

knitr::kable(table3, caption = "Adjusted hazard ratios for 28-day recurrent parasitaemia: packaged model versus Tchaparian 2016 Table 3.")

# Deterministic algebra against printed constants, so this is exact to machine
# precision. Realised 1.1e-16.
stopifnot(max(abs(table3$Model / table3$`Table 3` - 1)) < 1e-8)
```

The within-prophylaxis low-exposure contrast is the paper's interaction and is
recovered as a ratio of two cells rather than from a single coefficient:

```{r interaction}
hr_ts_low <- hr_of(CONC_LUMEFANTRINE_168H = 150, CONMED_TMPSMX = 1)
hr_ts_ok <- hr_of(CONMED_TMPSMX = 1)
stopifnot(abs(hr_ts_low / hr_ts_ok - 0.50) < 1e-8)
sprintf(
  "within the prophylaxis stratum, <200 ng/mL vs >=200 ng/mL gives HR %.4f (Table 3: 0.50)",
  hr_ts_low / hr_ts_ok
)
```

## The four-way stratification of Figure 4

Figure 4 stratifies the cumulative recurrence risk by prophylaxis status and by
the 200 ng/mL threshold. The model's four relative hazards reproduce the
qualitative pattern the figure and Discussion describe: exposure matters only in
the absence of prophylaxis, and the ordering reverses when prophylaxis is
present.

```{r figure4, fig.cap = "Relative hazard of 28-day recurrent parasitaemia across the four strata of Tchaparian 2016 Figure 4, at the reference covariate pattern. The reference cell is no prophylaxis with an adequate day 7 concentration."}
strata <- tidyr::expand_grid(
  tmpsmx = c(0, 1),
  low = c(FALSE, TRUE)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    hr = hr_of(CONMED_TMPSMX = tmpsmx, CONC_LUMEFANTRINE_168H = if (low) 150 else 250)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    Prophylaxis = factor(ifelse(tmpsmx == 1, "TMP-SMZ", "no TMP-SMZ"),
      levels = c("no TMP-SMZ", "TMP-SMZ")
    ),
    Exposure = factor(ifelse(low, "day 7 <200 ng/mL", "day 7 >=200 ng/mL"),
      levels = c("day 7 >=200 ng/mL", "day 7 <200 ng/mL")
    )
  )

ggplot(strata, aes(x = Prophylaxis, y = hr, fill = Exposure)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_text(aes(label = sprintf("%.2f", hr)),
    position = position_dodge(width = 0.7), vjust = -0.4, size = 3.2
  ) +
  scale_fill_manual(values = c("day 7 >=200 ng/mL" = "grey70", "day 7 <200 ng/mL" = "firebrick")) +
  labs(x = NULL, y = "Relative hazard of 28-day recurrence") +
  expand_limits(y = 3.4) +
  theme_bw() +
  theme(legend.position = "top")

# The direction of the interaction is the paper's headline and is a property of
# the printed coefficients, not of a draw, so it is asserted directly.
stopifnot(
  strata$hr[strata$tmpsmx == 0 & strata$low] > strata$hr[strata$tmpsmx == 0 & !strata$low],
  strata$hr[strata$tmpsmx == 1 & strata$low] < strata$hr[strata$tmpsmx == 1 & !strata$low]
)
```

## Day 3 is the stronger predictor

The paper's comparative analysis reports that in children without prophylaxis, a
one-unit rise in natural-log concentration reduces the 28-day hazard by 49% on
day 3 but only 20% on day 7. The two outputs `hr_c3` and `hr_c7` carry those two
fits as exposure-only relative hazards, in which the unprinted adjustment terms
cancel.

```{r day3-vs-day7}
ratio_of <- function(col, ref_val) {
  row <- ref_row
  row[[col]] <- ref_val * exp(1)
  out <- rxode2::rxSolve(pd, row, rxode2::et(0), returnType = "data.frame")
  if (col == "CONC_LUMEFANTRINE_72H") out$hr_c3[1] else out$hr_c7[1]
}

slope_tab <- tibble::tibble(
  Landmark = c("Day 3 (72 h)", "Day 7 (168 h)"),
  `Model HR per ln unit` = round(c(
    ratio_of("CONC_LUMEFANTRINE_72H", 2777),
    ratio_of("CONC_LUMEFANTRINE_168H", 216)
  ), 4),
  Published = c(0.51, 0.80)
)
knitr::kable(slope_tab, caption = "Hazard ratio per one-unit increase in natural-log lumefantrine concentration, children not receiving prophylaxis (Tchaparian 2016 Results and Discussion).")

stopifnot(max(abs(slope_tab$`Model HR per ln unit` / slope_tab$Published - 1)) < 1e-8)
```

## Linking the two models

The recurrence model's exposure column is meant to be generated by the PK model,
which is what makes the pair usable together. Feeding the simulated cohort's own
day 7 concentrations through the hazard model gives the exposure distribution the
threshold acts on.

```{r link}
linked <- per_id |>
  dplyr::transmute(
    id,
    CONC_LUMEFANTRINE_168H = c168,
    CONC_LUMEFANTRINE_72H = c72,
    CONMED_TMPSMX = 0, AGE = 0.75, RESIDENCE_RURAL = 0,
    HGB = 10, PARA = 17603, WAZ = 0
  )

linked_hr <- rxode2::rxSolve(pd, linked, rxode2::et(0), returnType = "data.frame")
frac_low <- mean(linked$CONC_LUMEFANTRINE_168H < 200)

link_tab <- tibble::tibble(
  Quantity = c(
    "Children below the 200 ng/mL day 7 threshold, %",
    "Median relative hazard, no prophylaxis",
    "Median relative hazard if all were on prophylaxis"
  ),
  Value = c(
    round(100 * frac_low, 1),
    round(median(linked_hr$hr), 3),
    round(median(rxode2::rxSolve(pd, dplyr::mutate(linked, CONMED_TMPSMX = 1),
      rxode2::et(0), returnType = "data.frame")$hr), 3)
  )
)
knitr::kable(link_tab, caption = "The PK model feeding the recurrence model, reference covariate pattern otherwise.")

# The paper's measured day 7 median of 216 ng/mL sits just above the 200 ng/mL
# threshold, so roughly half the cohort should fall below it. A wide bound: the
# quantity is a proportion of a random cohort near a threshold, and the bound
# still fails if the exposure distribution is off by the factor that a
# mis-transcribed dose or clearance would produce.
stopifnot(frac_low > 0.2, frac_low < 0.7)
```

# Assumptions and deviations

**Day 3 concentrations run about 15% above the measured median, reproducibly.**
Across 1, 2, 4, 8 and 16 solver threads the simulated day 3 median was +14.1% to
+17.3% against the measured 2777 ng/mL, while every other comparison sat within
7%. This is recorded as a deviation rather than absorbed into a wider bound. The
likely mechanism is named by the paper itself: "the lack of sufficient sampling
times during the absorption phase, making it difficult to detect an absorption
delay and related kinetics", which is also why `ka` was fixed to 0.45 1/h from a
different study rather than estimated. Day 3 is 12 h after the last dose and
closest to the peak, so it is precisely where an unmodelled absorption lag would
bias predictions upward. Note also that the published 2777 ng/mL is a *measured*
median carrying assay variability and real home-dosing times, whereas the
simulated value is a noise-free individual prediction at exactly 72 h.

**The model's formal terminal half-life is far longer than lumefantrine's
accepted value.** `V2/F` of 441 L with `Q/F` of 0.23 L/h implies a beta half-life
near 60 days, against the roughly 4.5 days the paper's own Introduction quotes
for lumefantrine. This is a faithful transcription, not an error: `V2/F` carries
an RSE of 74% and the deep compartment holds very little mass, so the
*observable* decline is much faster and does match the paper -- the typical child
crosses the LOD at 9.6 days against the paper's "approximately 10 days". The
practical consequence is narrower than it first appears: PKNCA's span ratio over
a 1512 h window is 0.65 to 0.78, so `lambda.z` is fitted to a phase the window
does not fully cover, yet the extrapolated `aucinf.obs` still lands within about
0.1% of `F * Dose / CL`. Users sampling over a materially shorter window should
not assume that continues to hold.

**The base of the parasite-density logarithm is inferred.** Methods describe the
Table 3 covariate only as "parasite density (log transformed)" without naming a
base. The natural log is used here because the same paper names natural log
explicitly for the day 3 and day 7 concentration analyses one subsection later. A
log10 reading would rescale `e_para_haz` by `ln(10)`, making the effect 2.30
times steeper per log10 unit; the covariate is not significant either way
(HR 1.04, 95% CI 0.95-1.14, P = .35), so the choice does not affect any published
conclusion, but a user fitting this term to new data should confirm the base.

**Centring values for the two continuous Cox covariates are the cohort
summaries, not published reference values.** The paper prints per-unit hazard
ratios for haemoglobin and parasite density but no centring value. Haemoglobin is
centred on the Table 1 median (10.0 g/dL) and parasite density on the Table 1
geometric mean (17,603 parasites/uL), so that `hr` is 1 at a typical child. This
is presentational only: every hazard *ratio* between two covariate values is
unchanged by the choice.

**The prophylaxis main effect and the exposure effect are not printed on one
consistent parameterisation.** Table 3 gives the low-exposure effect within each
prophylaxis stratum (2.97 and 0.50) because the interaction is significant, but
lists the prophylaxis effect (0.75) in the main-effect block, where the
day-7-concentration row reads 1.21 rather than the 2.97 the interaction model
implies. The two blocks therefore come from different fits, and the prophylaxis
coefficient in the interaction model is not identifiable from the table. The
model uses 0.75 as the prophylaxis effect within the adequate-exposure stratum
and flags it here. The direction is corroborated twice independently: the
prophylaxis stratum had a higher adequate-response rate (65.9% versus 59.1%) and
the Discussion attributes the attenuated exposure-response to "the independent
protective effects of TMP-SMZ against malaria". The marginal 1.21 row is not
encoded, since it contradicts the retained interaction model.

**Coincidence warning on the value 1.21.** Table 3 prints 1.21 twice, once for
rural residence (95% CI 0.43-3.43) and once for the day-7-concentration main
effect (95% CI 0.71-2.07). `e_residence_rural_haz` is the residence row. The two
must not be conflated.

**No baseline hazard, by construction.** A Cox regression is semiparametric, so
`h0(t)` was never produced by the fit; the recurrence model returns relative
hazards only and offers no survivor function. Figure 4 and Supplemental Figure S3
carry the paper's empirical Kaplan-Meier description of the absolute time course.

**Interoccasion variability is encoded for five occasions.** Monolix reports one
interoccasion magnitude per parameter shared across occasions, so occasion 1
carries the estimated variance and occasions 2 to 5 repeat it through `fixed()`,
following `Stoschus_2025_phenobarbital` and `Ding_2026_vancomycin`. Five spans
the observed maximum episodes per child (Table 1). Simulations here use `OCC = 1`.
This encoding raises a non-mu-referenced eta warning from rxode2 at parse time,
which is inherent to indicator-multiplexed interoccasion variability and is
shared by both registered precedents; it affects estimation, not simulation.

**Left censoring is not reproduced.** The fit retained 188 below-LOD samples
(23.3%) as left-censored, which is what informs the 14-day tail. The simulations
here are uncensored, so simulated concentrations below 52 ng/mL appear as real
values where the study could only record a censoring event.

**Race and ethnicity are not reported** beyond the cohort being Ugandan children
resident in Tororo district, and are not covariates in either model.

**Cohort construction.** Weight and age were calibrated to the Table 1 marginals
rather than drawn from WHO growth curves, because the published exposure
summaries are conditioned on this cohort's own distributions and the cohort runs
below the WHO median for age (reported median weight-for-age z score -0.74).
