---
title: "TerHeine_2014_tamoxifen"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TerHeine_2014_tamoxifen}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("TerHeine_2014_tamoxifen"))$reference`
* Description: `r rxode2::rxode2(readModelDb("TerHeine_2014_tamoxifen"))$description`
* Article: <https://doi.org/10.1111/bcp.12388>
* DDMORE Foundation Model Repository entry: `DDMODEL00000212`

# Population

The Ter Heine 2014 publication reports a popPK study of tamoxifen and its
active metabolite endoxifen at steady state in adult breast-cancer patients
on standard 20 mg PO QD chronic dosing. The full publication PDF was not on
disk during model extraction; only the Table 2 of final estimates was
embedded in the DDMORE bundle (`Output_real_tamoxifen.pdf`), so the
demographic ranges captured in `population` are partial — the
`age_range` / `weight_range` fields carry a TODO marker. The DDMORE
bundle ships a simulated dataset of 100 hypothetical 20-mg-QD subjects
whose CYP2D6 and CYP3A4 / 5 individual-activity covariates span a wide
range (CYP2D6 ≈ 20-21000 ng/L; CYP3A4 ≈ 20-130 ng/L), reflecting the
clinically observed variability in CYP-probe-derived metabolic activity.

The same information is available programmatically via
`rxode2::rxode2(readModelDb("TerHeine_2014_tamoxifen"))$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/ddmore/TerHeine_2014_tamoxifen.R`. The
table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (k12, 1/h)     | 1.90  | PDF Table 2 (RSE 20.2%) |
| `llag` (tlag, h)     | 0.455 | PDF Table 2 (RSE 10.4%) |
| `lcl` (CL20/F, L/h)  | 9.34  | PDF Table 2 (RSE 6.20%) |
| `lvc` (V2/F, L)      | 753   | PDF Table 2 (RSE 9.00%) |
| `lq`  (Q/F, L/h)     | 61.8  | PDF Table 2 (RSE 65.4%) |
| `lcl_endx` (CL23/F, L/h) | 0.324 | PDF Table 2 (RSE 17.0%) |
| `e_cyp2d6_cl_endx`   | 0.262 | PDF Table 2 (RSE 14.0%) |
| `e_cyp3a4_cl_endx`   | 0.157 | PDF Table 2 (RSE 72.0%) |
| `etalcl` (ω CL20/F)  | %CV 37.8 → log-variance 0.1336 | PDF Table 2 (RSE 19.2%) |
| `etalvc` (ω V2/F)    | %CV 26.6 → log-variance 0.0683 | PDF Table 2 (RSE 53.9%) |
| `etalcl_endx` (ω CL23/F) | %CV 25.5 → log-variance 0.0630 | PDF Table 2 (RSE 19.3%) |
| ρ(CL20-V2/F) → block covariance | 0.613 → 0.0586 | PDF Table 2 (RSE 31.2%) |
| `propSd` (σ_TAM)     | 0.138 | PDF Table 2 (RSE 11.3%) |
| `propSd_endx` (σ_ENDX) | 0.189 | PDF Table 2 (RSE 10.1%) |
| Endoxifen Vc (fixed, L) | 400 | `.mdl` GROUP_VARIABLES; cite Ahmad et al. CPT 88(6):814-817 (2010) |
| Endoxifen CL (fixed, L/h) | 5.1 | `.mdl` GROUP_VARIABLES; cite Ahmad et al. CPT 88(6):814-817 (2010) |
| Tamoxifen MW (g/mol) | 371.51456 | `.mdl` GROUP_VARIABLES |
| Endoxifen MW (g/mol) | 373.48738 | `.mdl` GROUP_VARIABLES |
| CYP2D6 reference (ng/L) | 1560 | `.mdl` GROUP_VARIABLES (population median) |
| CYP3A4 reference (ng/L) | 44.7  | `.mdl` GROUP_VARIABLES (population median) |
| Equation: depot decay (`d/dt(depot) <- -ka * depot`) | n/a | `.mdl` MODEL_PREDICTION DEQ block |
| Equation: hepatic QSS (`c_hep <- (ratein + q*c2) / (q + cl_endx_form)`) | n/a | `.mdl` MODEL_PREDICTION DEQ block |
| Equation: parent central ODE | n/a | `.mdl` MODEL_PREDICTION DEQ block |
| Equation: metabolite central ODE | n/a | `.mdl` MODEL_PREDICTION DEQ block |
| Equation: ALAG mechanism on depot | n/a | `.mdl` MODEL_PREDICTION DEQ (manual `if (T >= ALAG1)` switch) — replaced here with the equivalent standard NONMEM `lag(depot)` mechanism |

# Virtual cohort

We exercise the model with a small set of CYP-phenotype scenarios, all on
the standard 20 mg PO QD chronic regimen at steady state. The scenarios
cover:

* **typical** — population-median CYP2D6 (1560 ng/L) and CYP3A4 (44.7 ng/L);
* **CYP2D6-low** — 156 ng/L (1/10 of the median, approximating poor-metabolizer activity);
* **CYP2D6-high** — 15600 ng/L (10 × the median, approximating ultrarapid-metabolizer activity);
* **CYP3A4-low** — 4.47 ng/L (1/10 of the median);
* **CYP3A4-high** — 447 ng/L (10 × the median).

```{r cohort}
set.seed(42)

scenarios <- tribble(
  ~scenario,        ~CYP2D6, ~CYP3A4,
  "typical",        1560,    44.7,
  "CYP2D6-low",      156,    44.7,
  "CYP2D6-high",   15600,    44.7,
  "CYP3A4-low",     1560,     4.47,
  "CYP3A4-high",    1560,    447
)

# Build event tables: 20 mg QD steady-state with II=24h. Observation grid
# spans one steady-state dosing interval at half-hour resolution.
# rxSolve reads CYP2D6 / CYP3A4 from columns in the events data frame.
build_events <- function(scenarios) {
  obs_t <- seq(0, 24, by = 0.25)
  out <- vector("list", nrow(scenarios))
  for (i in seq_len(nrow(scenarios))) {
    sc   <- scenarios[i, ]
    n_obs <- length(obs_t)
    dose_row <- data.frame(
      id = i, time = 0, evid = 1, amt = 20, cmt = 1L, ii = 24, ss = 1,
      scenario = sc$scenario, CYP2D6 = sc$CYP2D6, CYP3A4 = sc$CYP3A4
    )
    obs_tam <- data.frame(
      id = i, time = obs_t, evid = 0, amt = 0, cmt = 4L, ii = 0, ss = 0,
      scenario = sc$scenario, CYP2D6 = sc$CYP2D6, CYP3A4 = sc$CYP3A4
    )
    obs_endx <- data.frame(
      id = i, time = obs_t, evid = 0, amt = 0, cmt = 5L, ii = 0, ss = 0,
      scenario = sc$scenario, CYP2D6 = sc$CYP2D6, CYP3A4 = sc$CYP3A4
    )
    out[[i]] <- rbind(dose_row, obs_tam, obs_endx)
  }
  dplyr::bind_rows(out)
}

events <- build_events(scenarios)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("TerHeine_2014_tamoxifen"))

# Typical-value (no IIV): each scenario is one subject at the typical
# parameter values, so we zero out the random effects.
mod_typ <- rxode2::zeroRe(mod)

sim <- rxode2::rxSolve(
  mod_typ,
  events = events,
  keep   = c("scenario", "CYP2D6", "CYP3A4")
)
sim <- as.data.frame(sim)
```

# Replicate published behaviour

Ter Heine 2014 reports a complete Table 2 of point estimates but does not
publish a clean reference Cmax / Cmin / AUC table for steady state. The
deterministic typical-value profiles below are shown so that a reviewer
can confirm the steady-state magnitudes are biologically plausible
(tamoxifen ≈ 100-300 nmol/L; endoxifen ≈ 5-50 nmol/L on 20 mg QD chronic
dosing in adults — typical clinical-pharmacology ranges) and that the
covariate sensitivities go in the correct direction (CYP2D6-high
increases endoxifen formation; CYP2D6-low decreases it).

```{r profile-tam, fig.width=7, fig.height=4}
sim_tam <- sim |>
  dplyr::filter(CMT == 4) |>
  dplyr::select(id, time, scenario, Cc)

ggplot(sim_tam, aes(time, Cc, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after dose (h)", y = "Tamoxifen Cc (nmol/L)",
       colour = "scenario",
       title = "Tamoxifen steady-state profile, 20 mg PO QD",
       caption = "Typical-value simulation; CYP covariates affect endoxifen formation only, so the parent profile is invariant across scenarios.") +
  theme_bw()
```

```{r profile-endx, fig.width=7, fig.height=4}
sim_endx <- sim |>
  dplyr::filter(CMT == 5) |>
  dplyr::select(id, time, scenario, Cc_endx)

ggplot(sim_endx, aes(time, Cc_endx, colour = scenario)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after dose (h)", y = "Endoxifen Cc (nmol/L)",
       colour = "scenario",
       title = "Endoxifen steady-state profile, 20 mg PO QD",
       caption = "CYP2D6-high / CYP3A4-high increase endoxifen formation; the *-low scenarios reduce it.") +
  theme_bw()
```

# PKNCA validation

Steady-state NCA over the dosing interval (`tau = 24 h`), one row per
scenario.

```{r pknca-tam}
tau <- 24

sim_nca_tam <- sim |>
  dplyr::filter(CMT == 4, !is.na(Cc)) |>
  dplyr::select(id, time, Cc, scenario)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, scenario)

conc_obj_tam <- PKNCA::PKNCAconc(
  sim_nca_tam, Cc ~ time | scenario + id,
  concu = "nmol/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | scenario + id, doseu = "mg"
)

intervals_ss <- data.frame(
  start    = 0,
  end      = tau,
  cmax     = TRUE,
  tmax     = TRUE,
  cmin     = TRUE,
  auclast  = TRUE,
  cav      = TRUE,
  ctrough  = TRUE
)

nca_tam <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj_tam, dose_obj, intervals = intervals_ss
))
knitr::kable(
  as.data.frame(nca_tam$result) |>
    dplyr::select(scenario, id, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES),
  caption = "Steady-state NCA — tamoxifen (Cc) by CYP-phenotype scenario"
)
```

```{r pknca-endx}
sim_nca_endx <- sim |>
  dplyr::filter(CMT == 5, !is.na(Cc_endx)) |>
  dplyr::select(id, time, Cc_endx, scenario) |>
  dplyr::rename(Cc = Cc_endx)

conc_obj_endx <- PKNCA::PKNCAconc(
  sim_nca_endx, Cc ~ time | scenario + id,
  concu = "nmol/L", timeu = "h"
)

nca_endx <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj_endx, dose_obj, intervals = intervals_ss
))
knitr::kable(
  as.data.frame(nca_endx$result) |>
    dplyr::select(scenario, id, PPTESTCD, PPORRES) |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES),
  caption = "Steady-state NCA — endoxifen (Cc_endx) by CYP-phenotype scenario"
)
```

The covariate sensitivities are the headline finding of Ter Heine 2014:
endoxifen exposure rises with CYP2D6 activity (the CYP2D6-high scenario
yields a markedly higher endoxifen Cmax / AUC than CYP2D6-low) and rises
more modestly with CYP3A4/5 activity, while tamoxifen exposure is
invariant under either covariate (CYP2D6 and CYP3A4 enter only on the
formation clearance to endoxifen).

# Comparison against the DDMORE-shipped simulated dataset

The DDMORE bundle ships a simulated dataset
(`Simulated_tamoxifen.csv`, 100 hypothetical subjects, 20 mg QD steady
state) along with `Output_simulated_tamoxifen.lst` — a NONMEM re-fit of
the model to that dataset using the DDMORE convertor's translation. We
do **not** attempt a row-by-row IPRED comparison against that dataset
here, because:

1. The bundle's `IPRED` column carries **individual** predictions
   (Empirical Bayes etas applied), not population-typical predictions,
   so it cannot be matched against a typical-value simulation without
   recovering each subject's eta vector.
2. The DDMORE convertor's parameter estimates on the simulated dataset
   differ from the original publication's estimates (especially on
   `tlag` — 2.02 h in the bundle re-fit vs 0.455 h in the publication),
   because the convertor implements the absorption lag differently from
   standard NONMEM (see Errata below). The model file in this package
   uses the **publication** values.

The cleanest sanity check the bundle does support is dimensional /
order-of-magnitude: the simulated `DV` values for tamoxifen at steady
state span roughly 50-400 nmol/L and for endoxifen span roughly
3-15 nmol/L, which agrees with our typical-value steady-state simulations
above and with the clinical-pharmacology literature for chronic 20 mg QD
tamoxifen in adult breast-cancer patients.

# Assumptions and deviations

- **Source-publication PDF not on disk during extraction.** Only Table 2
  of the original Ter Heine 2014 publication was embedded in the DDMORE
  bundle (`Output_real_tamoxifen.pdf`). The full publication
  (Br J Clin Pharmacol 2014;78:572-586, doi:10.1111/bcp.12388) was not
  available locally, so the `population$age_range` and
  `population$weight_range` fields carry TODO markers. Demographic
  fields can be filled in via a follow-up edit once the full PDF is on
  disk.

- **Parameter values from the publication, not from the DDMORE re-fit.**
  The bundle's `Output_simulated_tamoxifen.lst` reports a NONMEM re-fit
  of the model to a simulated dataset using the DDMORE convertor's
  translation. The convertor implements the absorption lag inside the
  ODE system rather than via the standard NONMEM `ALAG` mechanism, and
  this implementation difference produces materially different
  estimates of `tlag` (2.02 h re-fit vs 0.455 h published) and smaller
  shifts in several other parameters. The
  `Model_Accommodations.txt` in the bundle calls out the deviation.
  Per the `extract-literature-model` skill, this model file uses the
  **publication** values; the DDMORE convertor's `.lst` re-fit values
  are not used.

- **Standard NONMEM `lag(depot)` substituted for the DDMORE `if`-switch.**
  The DDMORE-converted `.mdl` implements the absorption lag manually
  inside `$DES` via an `if (T >= ALAG1)` switch, leaving the gut
  compartment loaded at the dose value through `t = 0..ALAG1` and then
  starting first-order decay. The standard NONMEM `ALAG` mechanism (used
  here, `lag(depot) <- exp(llag)`) instead delays the dose's appearance
  in the depot until `t = lag`, so `depot = 0` for `t < lag`. Both forms
  yield the same systemic exposure once `t >= lag`; the published
  parameters were originally fitted with the standard mechanism, so
  using `lag(depot)` here keeps the implementation aligned with the
  parameter values.

- **Residual-error correlation dropped.** The original publication
  reports a residual-error correlation between the parent and metabolite
  observations (ρ = 62.2% per PDF Table 2). The DDMORE-shipped `.mdl`
  drops that correlation and parameterises the residual SDs as separate
  THETAs (per `Model_Accommodations.txt`); we follow the DDMORE
  implementation here. Re-introducing the correlated residual would
  require switching to a multivariate residual-error formulation that is
  outside the standard nlmixr2 multi-output `prop()` syntax.

- **Endoxifen disposition fixed from a separate publication.** The
  endoxifen elimination clearance (5.1 L/h) and central volume (400 L)
  are fixed from Ahmad et al. (2010) Clin Pharmacol Ther 88(6):814-817,
  not estimated by Ter Heine 2014. The fixed values are hard-coded
  inside `model()` rather than being declared as `ini()` parameters
  (with `fixed()`) because they are literature constants, not
  parameters of the Ter Heine 2014 fit.

- **Implicit mass-conserving parent-to-metabolite conversion.** The
  source `.mdl` tracks the metabolite compartment in tamoxifen-mass
  equivalents (i.e., `d/dt(central_endx) = cl_endx_form * c_hep -
  kel_endx * central_endx` with no molar-mass scaling on the formation
  flux), then converts the compartment mass to plasma concentration via
  the metabolite's own MW (`Mendx`). This is dimensionally consistent
  only if `Mtam ≈ Mendx`; here `Mtam = 371.515` and `Mendx = 373.487`
  g/mol differ by ≈ 0.5%, so the implicit assumption introduces
  negligible error and we replicate the source's scaling exactly.

- **Convention check warning on dimensional consistency between dosing
  unit and concentration unit.** `checkModelConventions()` flags
  `units$dosing = "mg"` and `units$concentration = "nmol/L"` as
  dimensionally incompatible. This is intentional: the model
  internally performs a molar-mass conversion via the `Mtam` and
  `Mendx` group variables, so the dose is in mass units (mg) but the
  output concentration is in molar units (nmol/L = nM) to match the
  publication's reporting convention.

- **CYP2D6 / CYP3A4 covariate columns added to the
  `inst/references/covariate-columns.md` register.** These were not in
  the canonical register before this extraction; both have been added
  with general scope and a TODO note to consolidate the various
  CYP-phenotype encodings (continuous probe-derived activity,
  copy-number-corrected score, categorical PM/IM/EM/UM phenotype,
  activity-score sum) and to register the rest of the canonical
  drug-metabolizing-CYP set prospectively.
