---
title: "Piperacillin (Thorsted 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Piperacillin (Thorsted 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Thorsted_2019_piperacillin")()$reference`
* Description: `r readModelDb("Thorsted_2019_piperacillin")()$description`
* Article: <https://doi.org/10.1093/jac/dkz270> (open access, PMC6916132)

# Population

Thorsted et al. 2019 studied 43 children (aged 1-18 years, median 12) with
cancer and chemotherapy-induced fever at the Department of Pediatric Oncology,
Aarhus University Hospital, Denmark (April 2016 - January 2018). Children could
contribute more than one febrile episode, and 89 episodes (1-4 per child) were
analysed. Piperacillin/tazobactam (8:1) was given as a 5-minute IV infusion
approximately every 8 h at 300 mg/kg/day piperacillin, capped at the adult
16000 mg/day. The 482 serum samples (19 below the 0.5 mg/L LLOQ, handled with
the M3 method) were assayed for **unbound** piperacillin by UPLC after
ultrafiltration, so the model describes free concentrations directly.
Episode-level body weight had median 39.4 kg (range 9.5-107) and the
Schwartz-estimated GFR a median of 172.4 mL/min/1.73 m^2, i.e. a
hyperfiltrating population; 63% of the children were male, neutropenia was
present in 87% of episodes and bacteraemia in 11% (Table 1).

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Two-compartment disposition, first-order elimination from central | -- | Results, 'Pharmacokinetic modelling' (dOFV = -18.7 vs one-compartment) |
| `lcl` | log(15.4 L/h) at 70 kg | Table 2 |
| `lvc` | log(16.0 L) at 70 kg | Table 2 |
| `lq` | log(0.237 L/h) at 70 kg | Table 2 |
| `lvp` | log(3.40 L) at 70 kg | Table 2 |
| `e_wt_cl`, `e_wt_q` | fixed 0.75 | Methods 'Pharmacokinetic modelling'; Table 2 footnote b |
| `e_wt_vc`, `e_wt_vp` | fixed 1.0 | Methods 'Pharmacokinetic modelling'; Table 2 footnote b |
| Allometric reference weight | 70 kg | Table 2 footnote b |
| `etaiov_cl_1` ... `etaiov_cl_4` | log(1 + 0.166^2) = 0.027183 | Table 2 'CV%CL' 16.6%, inter-fever-episode variability |
| No IIV on any parameter | -- | Results: IIV 'became insignificant' once IOV was added |
| `propSd` | 0.332 | Table 2 'CV%ERR' 33.2% |
| Occasion = febrile episode, 1-4 per child | -- | Methods 'Pharmacokinetic modelling'; Results 'Patient characteristics' |

# Model

```{r model}
mod <- readModelDb("Thorsted_2019_piperacillin")
mod_typical <- rxode2::zeroRe(mod)
```

## Closed-form half-lives

The paper reports distribution and terminal half-lives of 0.708 h and 10.1 h
for a 70-kg patient. These follow exactly from the Table 2 parameters through
the two-compartment eigenvalues, so this is a tight transcription gate.

```{r halflives}
cl <- 15.4
vc <- 16.0
q <- 0.237
vp <- 3.40
k10 <- cl / vc
k12 <- q / vc
k21 <- q / vp
s <- k10 + k12 + k21
alpha <- (s + sqrt(s^2 - 4 * k10 * k21)) / 2
beta <- (s - sqrt(s^2 - 4 * k10 * k21)) / 2
hl <- c(alpha = log(2) / alpha, beta = log(2) / beta)
knitr::kable(
  data.frame(
    "Half-life" = c("t1/2,alpha (h)", "t1/2,beta (h)"),
    Paper = c(0.708, 10.1),
    Model = signif(hl, 3),
    check.names = FALSE
  )
)
stopifnot(abs(hl[["alpha"]] - 0.708) < 0.005, abs(hl[["beta"]] - 10.1) < 0.05)
```

# Typical concentration-time courses (Figure 3)

Figure 3 shows typical free concentrations for 15, 40 and 65 kg children given
300 mg/kg/day (capped at 16000 mg/day) as (a) three 5-minute infusions, (b)
three 3-hour infusions, and (c) a continuous infusion.

```{r fig3}
wts <- c(15, 40, 65)
daily_dose <- function(wt) min(300 * wt, 16000)

make_regimen <- function(wt, regimen) {
  dd <- daily_dose(wt)
  ev <- switch(regimen,
    "SI 5 min q8h" = rxode2::et(amt = dd / 3, dur = 5 / 60, ii = 8, addl = 2, cmt = "central"),
    "EI 3 h q8h" = rxode2::et(amt = dd / 3, dur = 3, ii = 8, addl = 2, cmt = "central"),
    "CI" = rxode2::et(amt = dd, dur = 24, cmt = "central")
  )
  ev <- rxode2::et(ev, seq(0, 24, by = 0.05), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt
  d$OCC <- 1
  d$id <- 1
  d
}

grid <- expand.grid(wt = wts, regimen = c("SI 5 min q8h", "EI 3 h q8h", "CI"), stringsAsFactors = FALSE)
fig3 <- Map(function(wt, regimen) {
  s <- rxode2::rxSolve(mod_typical, make_regimen(wt, regimen), returnType = "data.frame")
  data.frame(time = s$time, Cc = s$Cc, wt = paste(wt, "kg"), regimen = regimen)
}, grid$wt, grid$regimen) |>
  dplyr::bind_rows() |>
  dplyr::mutate(regimen = factor(regimen, levels = c("SI 5 min q8h", "EI 3 h q8h", "CI")))

ggplot(fig3, aes(time, Cc, colour = wt)) +
  geom_line() +
  geom_hline(yintercept = c(0.5, 2, 4, 16), linetype = "dashed", colour = "grey50") +
  facet_wrap(~regimen) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 24, 8)) +
  labs(
    x = "Time (h)", y = "Unbound piperacillin (mg/L)", colour = "Body weight",
    caption = "Replicates Figure 3 of Thorsted 2019 (typical patient, 300 mg/kg/day)."
  )
```

Consistent with the paper, lighter children reach lower troughs because the
allometric exponent on CL (0.75) is smaller than on volume (1.0), which
shortens their half-life. After the 5-minute infusion the peaks sit in the few
hundred mg/L range (about 340-410 mg/L in the first interval; the paper
quotes roughly 500 mg/L, read by eye from its log-scale Figure 3).

```{r fig3-check}
peaks <- fig3 |>
  dplyr::filter(regimen == "SI 5 min q8h", time <= 8) |>
  dplyr::group_by(wt) |>
  dplyr::summarise(Cmax = max(Cc), C8 = Cc[which.min(abs(time - 8))], .groups = "drop")
knitr::kable(peaks, digits = 2)
# Peaks after the 5-min infusion stay at or below the paper's ~500 mg/L.
stopifnot(all(peaks$Cmax < 550), all(peaks$Cmax > 250))
```

# Dose required for target attainment (Figure 5)

The corrected Figure 5 (erratum doi:10.1093/jac/dkz429) prints, for typical
15, 40 and 65 kg children at MIC = 4 mg/L (the MIC90), the steady-state dose
required for each target together with the resulting AUCss and Cmax,ss, for
(a) the licensed regimen (30-minute infusion q6h) and (b) a 3-hour extended
infusion q8h.

```{r fig5-reference}
fig5_ref <- data.frame(
  regimen = rep(c("(a) 30 min q6h", "(b) EI 3 h q8h"), each = 6),
  target = rep(rep(c("50% fT>4xMIC", "100% fT>MIC"), each = 3), 2),
  wt = rep(c(15, 40, 65), 4),
  dose_mg = c(2353, 2803, 3459, 7078, 9254, 10299, 1203, 1984, 2652, 10095, 13800, 14625),
  auclast = c(518.5, 296.1, 237.9, 1560, 977.3, 708.4, 248.5, 196.6, 182.4, 2087, 1365, 1007),
  cmax = c(526.9, 251.4, 185.2, 1585, 829.7, 551.3, 81.1, 63.3, 57.3, 680.4, 439.2, 316.4)
)
fig5_ref$treatment <- paste(fig5_ref$regimen, fig5_ref$target, paste(fig5_ref$wt, "kg"), sep = " | ")
```

Because the model is linear, the steady-state profile scales with dose, so the
required dose is found by bisection on one steady-state profile per weight and
regimen. The threshold is 16 mg/L for 50% fT>4xMIC and 4 mg/L for 100% fT>MIC.

```{r target}
ss_profile <- function(wt, tau, tinf, amt = 100 * wt, n = 40, by = 0.005) {
  ev <- rxode2::et(amt = amt, dur = tinf, ii = tau, addl = n - 1, cmt = "central")
  ev <- rxode2::et(ev, seq((n - 1) * tau, n * tau, by = by), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt
  d$OCC <- 1
  d$id <- 1
  s <- rxode2::rxSolve(mod_typical, d, returnType = "data.frame")
  s <- s[s$time >= (n - 1) * tau, ]
  data.frame(time = s$time - (n - 1) * tau, Cc = s$Cc, c_per_mg = s$Cc / amt)
}

required_dose <- function(p, threshold, frac) {
  ft <- function(dose) mean(p$c_per_mg * dose > threshold)
  lo <- 1
  hi <- 1e6
  for (i in 1:60) {
    mid <- sqrt(lo * hi)
    if (ft(mid) >= frac) hi <- mid else lo <- mid
  }
  hi
}

regimens <- list(
  "(a) 30 min q6h" = c(tau = 6, tinf = 0.5),
  "(b) EI 3 h q8h" = c(tau = 8, tinf = 3)
)
thresholds <- list("50% fT>4xMIC" = c(thr = 16, frac = 0.5), "100% fT>MIC" = c(thr = 4, frac = 1))

fig5_sim <- fig5_ref[, c("regimen", "target", "wt", "treatment")]
fig5_sim$dose_mg <- NA_real_
for (i in seq_len(nrow(fig5_sim))) {
  r <- regimens[[fig5_sim$regimen[i]]]
  th <- thresholds[[fig5_sim$target[i]]]
  p <- ss_profile(fig5_sim$wt[i], r[["tau"]], r[["tinf"]])
  fig5_sim$dose_mg[i] <- required_dose(p, th[["thr"]], th[["frac"]])
}

dose_cmp <- merge(
  fig5_ref[, c("treatment", "dose_mg")], fig5_sim[, c("treatment", "dose_mg")],
  by = "treatment", suffixes = c("_paper", "_model")
)
dose_cmp$pct_diff <- 100 * (dose_cmp$dose_mg_model / dose_cmp$dose_mg_paper - 1)
dose_cmp |>
  dplyr::rename(
    "Regimen | target | weight" = treatment,
    "Paper dose (mg)" = dose_mg_paper,
    "Model dose (mg)" = dose_mg_model,
    "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 1)
```

For the extended-infusion regimen (panel b) every model dose is within 2% of
the published dose, and the two licensed-regimen doses at 65 kg are within 1%
and 4%. At 15 and 40 kg the licensed-regimen doses (panel a) come out 7-10%
higher than published. This is not a transcription error in the model: panel a is not
internally consistent with panel b under any single set of linear PK
parameters. The steady-state AUC over a dosing interval equals dose / CL
whatever the regimen, so the paper's own dose / AUCss ratio recovers the
clearance it simulated with:

```{r implied-cl}
fig5_ref |>
  dplyr::filter(target == "50% fT>4xMIC") |>
  dplyr::mutate(cl_implied = dose_mg / auclast, cl_model = 15.4 * (wt / 70)^0.75) |>
  dplyr::select(regimen, wt, cl_implied, cl_model) |>
  dplyr::rename(
    "Regimen" = regimen, "Body weight (kg)" = wt,
    "Paper dose / AUCss (L/h)" = cl_implied, "Model CL (L/h)" = cl_model
  ) |>
  knitr::kable(digits = 2)
```

Panel b implies the Table 2 clearance at every weight, and so does panel a at
65 kg, but at 15 and 40 kg panel a implies a clearance about 6% lower -- the
same size as the dose gap. The erratum already corrected dose and AUC values
in this figure once; the residual inconsistency is recorded here and not
tuned away. The Results text quotes panel a as 0.67-1.96 (50% fT>4xMIC) and
1.98-5.78 (100% fT>MIC) times 80 mg/kg.

```{r target-check}
b <- dose_cmp$pct_diff[grepl("^\\(b\\)", dose_cmp$treatment)]
a65 <- dose_cmp$pct_diff[grepl("^\\(a\\).*65 kg$", dose_cmp$treatment)]
a_all <- dose_cmp$pct_diff[grepl("^\\(a\\)", dose_cmp$treatment)]
# Deterministic typical-value calculations: tight where the paper is
# self-consistent, a documented envelope for the self-inconsistent panel a.
stopifnot(all(abs(b) < 3), all(abs(a65) < 5), all(abs(a_all) < 12))
```

# Steady-state exposure with PKNCA (Figure 5)

The typical steady-state interval at each model-derived dose is analysed with
PKNCA (one typical profile per regimen, target and weight) and compared with
the AUCss and Cmax,ss printed in Figure 5.

```{r pknca}
conc <- lapply(seq_len(nrow(fig5_sim)), function(i) {
  r <- regimens[[fig5_sim$regimen[i]]]
  p <- ss_profile(fig5_sim$wt[i], r[["tau"]], r[["tinf"]], amt = fig5_sim$dose_mg[i], by = 0.02)
  data.frame(id = i, treatment = fig5_sim$treatment[i], time = p$time, Cc = p$Cc)
}) |>
  dplyr::bind_rows() |>
  dplyr::filter(!is.na(Cc))
doses <- data.frame(id = seq_len(nrow(fig5_sim)), treatment = fig5_sim$treatment, time = 0, amt = fig5_sim$dose_mg)
tau_by_id <- ifelse(grepl("^\\(a\\)", fig5_sim$treatment), 6, 8)
intervals <- data.frame(
  id = seq_len(nrow(fig5_sim)), treatment = fig5_sim$treatment,
  start = 0, end = tau_by_id, auclast = TRUE, cmax = TRUE
)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(doses, amt ~ time | treatment + id)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

tbl <- nlmixr2lib::ncaComparisonTable(
  nca, fig5_ref[, c("treatment", "auclast", "cmax")],
  by = "treatment",
  units = c(auclast = "mg*h/L", cmax = "mg/L")
)
knitr::kable(tbl)
```

Starred rows (> 20% difference) would flag a problem; the AUC and Cmax rows
follow the dose comparison above (panel b and 65 kg within a few percent, the
self-inconsistent panel a rows at 15 and 40 kg somewhat further off).

```{r nca-assert}
res <- as.data.frame(nca$result)
chk <- merge(
  res[res$PPTESTCD %in% c("auclast", "cmax"), c("treatment", "PPTESTCD", "PPORRES")],
  tidyr::pivot_longer(fig5_ref[, c("treatment", "auclast", "cmax")], -treatment,
    names_to = "PPTESTCD", values_to = "ref"
  ),
  by = c("treatment", "PPTESTCD")
)
chk$pct <- 100 * (chk$PPORRES / chk$ref - 1)
stopifnot(
  all(abs(chk$pct[grepl("^\\(b\\)", chk$treatment)]) < 5),
  all(abs(chk$pct) < 20)
)
```

# Variability around the typical profile (Figure 5b)

A stochastic cohort of 200 episodes per weight (IOV on CL, one episode each,
proportional residual error excluded as in a prediction interval of the
typical profile) reproduces the 90% prediction interval of the 50% fT>4xMIC
extended-infusion panel of Figure 5b.

```{r fig5b}
rxode2::rxSetSeed(20190827)
n_per_arm <- 200
n_dose <- 10
tau <- 8
ei50 <- fig5_sim[fig5_sim$regimen == "(b) EI 3 h q8h" & fig5_sim$target == "50% fT>4xMIC", ]
cohort <- lapply(seq_len(nrow(ei50)), function(i) {
  ev <- rxode2::et(amt = ei50$dose_mg[i], dur = 3, ii = tau, addl = n_dose - 1, cmt = "central")
  ev <- rxode2::et(ev, seq((n_dose - 1) * tau, n_dose * tau, by = 0.1), cmt = "central")
  d <- as.data.frame(ev)
  ids <- seq_len(n_per_arm) + (i - 1) * n_per_arm
  d <- dplyr::bind_rows(lapply(ids, function(id) dplyr::mutate(d, id = id)))
  d$WT <- ei50$wt[i]
  d$OCC <- 1
  d$weight <- paste(ei50$wt[i], "kg")
  d
}) |> dplyr::bind_rows()

sim <- rxode2::rxSolve(mod, cohort, keep = "weight", returnType = "data.frame")
pi_df <- sim |>
  dplyr::filter(time >= (n_dose - 1) * tau) |>
  dplyr::mutate(tad = time - (n_dose - 1) * tau) |>
  dplyr::group_by(weight, tad) |>
  dplyr::summarise(med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95), .groups = "drop")

ggplot(pi_df, aes(tad, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.3) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 16, linetype = "dashed") +
  facet_wrap(~weight) +
  scale_y_log10() +
  labs(
    x = "Time after last dose (h)", y = "Unbound piperacillin (mg/L)",
    caption = "Replicates Figure 5b (top row) of Thorsted 2019: median and 90% PI at the dose required for 50% fT>4xMIC."
  )
```

# Assumptions and deviations

* **IOV encoding.** The paper's only random effect on a structural parameter is
  inter-occasion (between-febrile-episode) variability on CL, with no IIV in
  the final model. rxode2 cannot simulate the `eta ~ var | occ` syntax, so the
  model multiplexes four per-occasion etas through the canonical `OCC` column
  (children contributed 1-4 episodes), with occasions 2-4 fixed to the
  occasion-1 variance. Because there is no IIV, simulating each episode as a
  separate subject with `OCC = 1` (as done above) is statistically equivalent.
* **CV to variance.** Table 2 reports the IOV as 16.6% CV; it was converted to
  a log-scale variance with `log(1 + CV^2)`. The proportional residual CV of
  33.2% is used directly as `propSd`.
* **Allometric exponent on Q.** Table 2 footnote b states the 0.75 exponent for
  "CL"; the Methods state 0.75 applies to clearances in general, so it is also
  applied to Q.
* **Figure 5a self-inconsistency.** At 15 and 40 kg the licensed-regimen
  panel implies (through dose / AUCss) a clearance about 6% below the Table 2
  value that panel b and the 65-kg panel a reproduce, and the model-derived
  required doses are correspondingly 7-10% higher. The model
  was not adjusted; the gap is reported in the comparison tables above.
* **Target attainment definition.** fT>threshold is computed on a 0.005-h grid
  over one steady-state dosing interval of the typical (no IOV) profile; the
  paper does not state its grid, which may explain the remaining 1-2%
  differences in panel b.
* **Unbound concentrations.** `Cc` is the free (ultrafiltrate) serum
  concentration; no protein-binding correction should be applied when using
  the model for fT>MIC.
* **Tazobactam** was not measured and is not modelled.
* **Erratum** (doi:10.1093/jac/dkz429, J Antimicrob Chemother 2020;75:254-255)
  corrects the Results sentence on the licensed regimen (the 100% fT>MIC range
  was misprinted as 2.10-9.25 and should read 1.98-5.78 times 80 mg/kg) and
  some dose / AUCss values in Figure 5. No model parameter or equation is
  affected. The corrected 1.98-5.78 range is the one checked above. The
  Discussion's 182-248 mg*h/L AUC range was not amended by the erratum.
* Appendix S1 (supplementary data) describes the optimal sampling design only
  and carries no parameters of the final model.
