---
title: "Somatropin / rhGH translational human (Thorsted 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Somatropin / rhGH translational human (Thorsted 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("Thorsted_2016_somatropin_human")
ui  <- rxode2::rxode(mod)
cat(ui$reference, sep = "\n")
```

* Description: `r ui$description`
* Article: [Br J Pharmacol 173(11):1742-55](https://doi.org/10.1111/bph.13473)

This is the translational (allometrically-scaled rat-to-human)
companion model to `Thorsted_2016_somatropin_rat`. Parameter values
are obtained from the rat fit by scaling clearances and Vmax with
exponent 0.75, distribution volumes with exponent 0.9 (the
empirically-selected best-fit exponent for human i.v. data per
Thorsted 2016 Figure 4), and first-order absorption / IGF-1 turnover
rate constants with exponent -0.25, around a reference adult body
weight of 70 kg (Thorsted 2016 Table 3). The s.c. absorption model
is the corrected form (F2 reduced from 0.833 to 0.500 and a transit
compartment added to the ka1 path; Thorsted 2016 Figure 5 narrative).
The GH effect-delay (CPLAG) chain is intentionally dropped for the
human prediction.

# Population

```{r population}
str(ui$population)
```

The translational target is a 70-kg GH-deficient adult male as
described in the Laursen 1996 validation cohort cited by Thorsted
2016 (10 patients, withdrawn from rhGH treatment for 1 week to
washout IGF-1 prior to study, dosed at 5 ug/kg i.v. bolus and at 33.5
ug/kg s.c.). The packaged model is intended for forward simulation
rather than re-fitting to human data.

# Source trace

| Equation / parameter | Rat value (Table 2) | Scaling | Human value (Table 3) | Source |
|---|---|---|---|---|
| CL | 0.0285 L/h | 0.75 | 3.88 L/h | Table 3 |
| Vmax | 11.5 ug/h | 0.75 | 1565 ug/h | Table 3 |
| KM | 358 ug/L | - (not scaled, see Errata) | 358 ug/L | Table 3 (Methods text) |
| Vc | 0.0069 L | 0.9 | 2.51 L | Table 3 |
| Q | 0.0101 L/h | 0.75 | 1.37 L/h | Table 3 |
| Vp | 0.0081 L | 0.9 | 2.94 L | Table 3 |
| ka1 | 3.02 1/h | -0.25 | 0.587 1/h | Table 3 |
| ka2 | 1.22 1/h | -0.25 | 0.237 1/h | Table 3 |
| F1 | 0.0316 | - | 0.0316 | Table 3 (kept as in rat) |
| F2 | 0.833 | corrected to 0.500 | 0.500 | Results / Figure 5 narrative |
| kout | 0.0913 1/h | -0.25 | 0.0178 1/h | Table 3 |
| R0 | 29.4 ng/mL | fixed to human pop mean | 65 ng/mL | Table 3 / Methods (Laursen 1996) |
| Emax | 9.88 | - | 9.88 | Table 3 |
| EC50 | 16.3 ug/L | - | 16.3 ug/L | Table 3 |
| GH effect-delay chain | active in rat | dropped | not present | Methods (Prediction of human PKPD) |
| Proportional residual error (PK) | 0.233 | fixed at 0.25 | 0.25 | Methods |
| Additive residual SD (PK) | 0.279 ng/mL | fixed at 0.25 | 0.25 ng/mL | Methods |
| Additive residual SD (IGF-1) | 21.3 ng/mL | fixed at 10 | 10 ng/mL | Methods |
| IIV CL / Vp / ka2 / R0 | inherited from rat | inherited from rat | inherited from rat | Methods (variability as estimated in PKPD model) |

# Virtual cohort

```{r cohort}
set.seed(20260516)

ref_wt <- 70

make_sc_cohort <- function(n, dose_ug_per_kg, regimen_label, id_offset = 0L,
                           obs_grid = c(0, 0.5, 1, 2, 4, 6, 8, 12, 24, 36, 48)) {
  ids <- id_offset + seq_len(n)
  dose_amt <- dose_ug_per_kg * ref_wt
  doses_depot  <- tidyr::expand_grid(id = ids, time = 0, evid = 1,
                                     amt = dose_amt, cmt = "depot")
  doses_depot2 <- doses_depot |> dplyr::mutate(cmt = "depot2")
  obs <- tidyr::expand_grid(id = ids, time = obs_grid) |>
    dplyr::mutate(evid = 0, amt = 0, cmt = "Cc")
  events <- dplyr::bind_rows(doses_depot, doses_depot2, obs) |>
    dplyr::arrange(id, time, evid) |>
    dplyr::mutate(WT = ref_wt, cohort = regimen_label)
  events
}

make_iv_cohort <- function(n, dose_ug_per_kg, regimen_label, id_offset = 0L,
                           obs_grid = c(0, 0.08, 0.17, 0.33, 0.5, 1, 2, 4, 6,
                                        8, 12, 24)) {
  ids <- id_offset + seq_len(n)
  dose_amt <- dose_ug_per_kg * ref_wt
  doses <- tidyr::expand_grid(id = ids, time = 0, evid = 1,
                              amt = dose_amt, cmt = "central")
  obs <- tidyr::expand_grid(id = ids, time = obs_grid) |>
    dplyr::mutate(evid = 0, amt = 0, cmt = "Cc")
  events <- dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, evid) |>
    dplyr::mutate(WT = ref_wt, cohort = regimen_label)
  events
}

events <- dplyr::bind_rows(
  make_iv_cohort(20,   5, "IV 5 ug/kg",     id_offset =   0L),
  make_sc_cohort(20, 33.5, "SC 33.5 ug/kg", id_offset = 100L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
```

# Typical-value simulation

```{r typical-pk}
mod_typ <- rxode2::zeroRe(mod)
sim_typ <- rxode2::rxSolve(mod_typ, events = events,
                           keep = c("cohort", "WT")) |>
  as.data.frame()
```

```{r figure-4b-typical-pk}
sim_typ |>
  dplyr::filter(time > 0) |>
  dplyr::distinct(cohort, time, .keep_all = TRUE) |>
  ggplot(aes(time, Cc, colour = cohort)) +
  geom_line(size = 0.8) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "rhGH plasma concentration (ng/mL)",
       title = "Typical-value rhGH PK in 70 kg adult (scaled from rat)",
       caption = "Replicates the shape of Thorsted 2016 Figure 4B (i.v.) and Figure 5B (corrected s.c.).")
```

# IGF-1 induction after a single SC dose

```{r typical-pd}
events_pd <- make_sc_cohort(
  n = 10, dose_ug_per_kg = 33.5, regimen_label = "SC 33.5 ug/kg",
  id_offset = 500L,
  obs_grid = c(0, 1, 2, 4, 8, 12, 24, 36, 48, 72)
)
sim_pd <- rxode2::rxSolve(mod_typ, events = events_pd,
                          keep = c("cohort", "WT")) |>
  as.data.frame()
```

```{r figure-5c-igf1}
sim_pd |>
  dplyr::filter(time > 0) |>
  dplyr::distinct(time, .keep_all = TRUE) |>
  ggplot(aes(time, IGF1)) +
  geom_line(size = 0.8, colour = "tomato") +
  geom_hline(yintercept = 65, linetype = "dashed") +
  labs(x = "Time (h)", y = "IGF-1 (ng/mL)",
       title = "Typical-value IGF-1 after 33.5 ug/kg s.c. rhGH in 70 kg adult",
       caption = "Dashed line at R0 = 65 ng/mL. Replicates Thorsted 2016 Figure 5C (corrected absorption).")
```

# NCA validation of rhGH after i.v. dosing

```{r pknca-iv}
sim_nca <- sim_typ |>
  dplyr::filter(time > 0, grepl("^IV ", cohort)) |>
  dplyr::select(id, time, Cc, cohort)

dose_df <- events |>
  dplyr::filter(evid == 1, grepl("^IV ", cohort)) |>
  dplyr::select(id, time, amt, cohort)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | cohort + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | cohort + id)

intervals <- data.frame(
  start      = 0,
  end        = 24,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
knitr::kable(summary(nca_res),
             caption = "Simulated NCA for rhGH after i.v. bolus (typical-value 70 kg adult).")
```

# Assumptions and deviations

- **Translational, not fitted.** This model is a forward simulation
  construct built by allometrically scaling the rat fit. It is not a
  fit to human data. Residual error parameters are fixed at the values
  used by Thorsted 2016 for the human-simulation validation (Methods),
  not estimated.

- **KM = 358 ug/L (text) versus 403 ug/L (Table 3).** Thorsted 2016
  Methods (Prediction of human PKPD) state: "The constant KM was not
  scaled." Table 3 column "Predicted" reports KM = 403 ug/L for the
  human-scaled model, which is internally inconsistent with the text
  (no scaling exponent should leave the value unchanged at 358). The
  packaged model uses the text-consistent value 358; treat this as a
  minor typographic discrepancy in Table 3.

- **ka2 IIV = 9.3% CV (Table 2) versus 19.3% (Table 3).** Thorsted
  2016 Methods state that the human-prediction variability is "as
  estimated in the PKPD model" (i.e., inherited unchanged from the
  rat fit). Table 2 reports rat ka2 IIV at 9.3% CV; Table 3 reports
  human ka2 IIV at 19.3%, inconsistent with the text. The packaged
  model uses 9.3% (Table 2) for both rat and human.

- **GH effect-delay (CPLAG) chain dropped for human prediction.** The
  rat IGF-1 response is delayed by a three-compartment effect chain
  feeding the Emax stimulation; Thorsted 2016 explicitly drops this
  chain for the human prediction (Methods: "The effect delay for rats
  was not included for predictions, as patients had only been deprived
  of rhGH for 1 week in order to washout IGF-1"). The packaged human
  model therefore drives IGF-1 stimulation directly from plasma
  rhGH concentration.

- **Bodyweight gain dropped.** The rat model's linear bodyweight gain
  driven by IGF-1 above baseline is not included in the human model,
  consistent with Thorsted 2016 (which uses bodyweight gain only as a
  preclinical efficacy biomarker, not a human end-point).

- **Parallel-absorption dosing convention.** SC doses require two
  event records per administration in user-supplied datasets (one
  with `cmt = "depot"` and one with `cmt = "depot2"`, same `amt`).
  IV doses use a single event with `cmt = "central"`.
