---
title: "Turoctocog alfa (Tiede 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Turoctocog alfa (Tiede 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Tiede_2021_turoctocogAlfa"))$reference`
* Description: `r rxode2::rxode(readModelDb("Tiede_2021_turoctocogAlfa"))$description`
* Article: <https://doi.org/10.3324/haematol.2019.241554> (open access)

Tiede et al. used a population PK model of factor VIII activity (FVIII:C)
after turoctocog alfa (NovoEight) to predict each patient's FVIII:C time
course from diary-recorded prophylactic doses in the guardian 1, 2 and 3
trials. They then related the predicted FVIII:C to diary-reported
spontaneous bleeds. The population PK parameters are printed in Table 2
(citing Jimenez-Yuste 2015, doi:10.1111/jth.12816) and the structural
equations in the Results. This article packages that PK model. The
exposure-response part of the paper, a negative-binomial regression of
annualized bleeding rate (ABR) on five FVIII:C categories, is reproduced
below as a calculation on simulated profiles.

# Population

The guardian programme enrolled previously treated patients with severe
hemophilia A (FVIII <= 1%) without inhibitors: adults and adolescents
(>= 12 years; guardian 1) and children (0-11 years; guardian 3), who could
continue into the guardian 2 extension trial. The exposure-response analysis
population (Table 1) comprised 168 adults/adolescents (mean age 28.98 years,
SD 12.15; mean weight 73.5 kg, SD 18.13) and 63 children (mean age 6.08 years,
SD 2.91; mean weight 24.6 kg, SD 10.03). Rich single-dose PK profiles (up to
48 h after a dose) were collected in a PK subgroup of 22 adults/adolescents and
28 children. Post-dose FVIII:C from routine visits of all patients was added to
the PK data. FVIII:C was measured by one-stage clot assay at a central
laboratory. Prophylaxis was 20-50 IU/kg every second day or 20-60 IU/kg three
times weekly, depending on age.

The same information is available programmatically via
`readModelDb("Tiede_2021_turoctocogAlfa")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` | log(3.02 dL/h) | Table 2: CL (70 kg, 20 y) = 302 mL/h |
| `lvc` | log(34.6 dL) | Table 2: V (70 kg) = 3.46 L |
| `e_wt_cl` | 0.95 | Table 2: allometric exponent on CL |
| `e_wt_vc` | 0.86 | Table 2: allometric exponent on V |
| `e_age_cl` | -0.01 per year | Table 2: age effect on CL |
| `etalcl` | log(1 + 0.32^2) = 0.0975 | Table 2: CV of CL = 0.32; Methods: log-normal, uncorrelated |
| `etalvc` | log(1 + 0.22^2) = 0.0473 | Table 2: CV of V = 0.22 |
| `propSd`, `addSd` | fixed 0 | Methods: combined proportional + additive error; magnitudes not reported |
| `vc = V * (WT/70)^e_wt_vc` | n/a | Results equation for V(W), Wref = 70 kg |
| `cl = CL * (WT/70)^e_wt_cl * (1 + e_age_cl * (AGE - 20))` | n/a | Results equation for CL(W, A), Aref = 20 years |
| `d/dt(central) = -cl/vc * central` | n/a | Results equation Cp(t) = D/V * exp(-CL/V * t) |

# Structural check against the printed closed form

The paper prints the one-compartment bolus solution
`Cp(t) = D/V * exp(-CL/V * t)`. A typical-value solve of the packaged model
for a 70-kg, 20-year-old patient given 50 IU/kg must reproduce it.

```{r closed-form}
mod <- readModelDb("Tiede_2021_turoctocogAlfa")
mod_typical <- rxode2::zeroRe(mod)

ev_ref <- data.frame(
  id = 1L,
  time = c(0, 0, 0.5, 1, 2, 4, 8, 12, 24, 36, 48),
  evid = c(1L, rep(0L, 10)),
  amt = c(50 * 70, rep(0, 10)),
  cmt = "central",
  WT = 70,
  AGE = 20
)
sim_ref <- rxode2::rxSolve(
  mod_typical,
  events = ev_ref,
  rtol = 1e-10,
  atol = 1e-12
) |>
  as.data.frame()

closed_form <- 50 * 70 / 34.6 * exp(-3.02 / 34.6 * sim_ref$time)
# Same parameters on both sides: pure integration error (~1e-9 measured).
stopifnot(max(abs(sim_ref$Cc / closed_form - 1)) < 1e-6)

knitr::kable(
  data.frame(
    quantity = c("Peak FVIII:C after 50 IU/kg (IU/dL)", "Half-life (h)"),
    value = c(50 * 70 / 34.6, log(2) * 34.6 / 3.02)
  ),
  digits = 2,
  caption = "Typical-value (70 kg, 20 years) derived quantities."
)
```

The typical half-life is 7.9 h. The age term raises clearance by 1% per year
below 20 years and lowers it by 1% per year above. Because the exponent on
CL (0.95) exceeds that on V (0.86), V/CL scales with WT^-0.09, which on its
own lengthens the half-life at low body weight; in young children the age
term more than offsets this, giving a slightly shorter half-life than in the
70-kg, 20-year-old reference, while older, heavier adults have a longer one.

```{r covariate-effects}
typ <- expand.grid(WT = c(15, 25, 50, 70, 90), AGE = c(3, 8, 20, 40)) |>
  mutate(
    cl_dLh = 3.02 * (WT / 70)^0.95 * (1 + -0.01 * (AGE - 20)),
    vc_dL = 34.6 * (WT / 70)^0.86,
    thalf_h = log(2) * vc_dL / cl_dLh
  )
typ |>
  filter(
    (WT == 15 & AGE == 3) |
      (WT == 25 & AGE == 8) |
      (WT == 50 & AGE == 20) |
      (WT == 70 & AGE == 20) |
      (WT == 90 & AGE == 40)
  ) |>
  dplyr::rename(
    "Weight (kg)" = WT,
    "Age (years)" = AGE,
    "CL (dL/h)" = cl_dLh,
    "V (dL)" = vc_dL,
    "Half-life (h)" = thalf_h
  ) |>
  knitr::kable(digits = 2, caption = "Typical PK parameters by weight and age.")
```

# Virtual cohort

Original patient data are not public. The virtual cohort draws age and body
weight independently from normal distributions with the Table 1 means and
standard deviations of each age group, redrawing values outside plausible
bounds (adults/adolescents 12-70 years and 35-150 kg; children 0.5-11.9 years
and 8-60 kg). The group sizes are those of Table 1 (168 and 63).

```{r cohort}
set.seed(20210701)
rxode2::rxSetSeed(20210701)

draw_bounded <- function(n, mean, sd, lower, upper) {
  x <- rnorm(n, mean, sd)
  bad <- x < lower | x > upper
  while (any(bad)) {
    x[bad] <- rnorm(sum(bad), mean, sd)
    bad <- x < lower | x > upper
  }
  x
}

subjects <- bind_rows(
  tibble(
    id = 1:168,
    group = "Adults/adolescents",
    AGE = draw_bounded(168, 28.98, 12.15, 12, 70),
    WT = draw_bounded(168, 73.5, 18.13, 35, 150)
  ),
  tibble(
    id = 168 + 1:63,
    group = "Children",
    AGE = draw_bounded(63, 6.08, 2.91, 0.5, 11.9),
    WT = draw_bounded(63, 24.6, 10.03, 8, 60)
  )
)
stopifnot(!anyDuplicated(subjects$id))

subjects |>
  group_by(group) |>
  summarise(
    n = n(),
    age_mean = mean(AGE),
    age_sd = sd(AGE),
    wt_mean = mean(WT),
    wt_sd = sd(WT),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 1, caption = "Virtual cohort vs Tiede 2021 Table 1.")
```

# Single-dose PK and NCA

The PK sessions of the pivotal trials gave a single 50 IU/kg dose with
sampling up to 48 h. The same design is simulated here (extended to 72 h for
the terminal phase), and PKNCA computes the NCA parameters by age group.
Tiede 2021 does not report NCA values, so there is no published comparison
table; the check is internal consistency of the simulated profiles with the
individual parameters.

```{r single-dose}
obs_times <- c(0, 0.25, 0.5, 1, 2, 4, 6, 8, 10, 12, 24, 30, 36, 48, 60, 72)
ev_sd <- subjects |>
  rowwise() |>
  reframe(
    id = id,
    group = group,
    WT = WT,
    AGE = AGE,
    time = c(0, obs_times),
    evid = c(1L, rep(0L, length(obs_times))),
    amt = c(50 * WT, rep(0, length(obs_times))),
    cmt = "central"
  ) |>
  arrange(id, time, desc(evid))

sim_sd <- rxode2::rxSolve(
  mod,
  events = ev_sd,
  keep = "group",
  rtol = 1e-10,
  atol = 1e-12
) |>
  as.data.frame()

sim_sd |>
  group_by(group, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = median(Cc),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  filter(time > 0) |>
  ggplot(aes(time, Q50, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)",
    y = "FVIII:C (IU/dL = %)",
    colour = NULL,
    fill = NULL,
    caption = "Median and 90% interval after a single 50 IU/kg dose."
  )
```

```{r pknca}
stopifnot(all(sim_sd$Cc >= -1e-6 * max(sim_sd$Cc)))
conc_df <- sim_sd |>
  filter(!is.na(Cc)) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(id, time, Cc, group)
dose_df <- ev_sd |>
  filter(evid == 1) |>
  select(id, time, amt, group)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | group + id)
dose_obj <- PKNCA::PKNCAdose(
  dose_df,
  amt ~ time | group + id,
  route = "intravascular"
)
intervals <- data.frame(
  start = 0,
  end = Inf,
  cmax = TRUE,
  aucinf.obs = TRUE,
  half.life = TRUE,
  cl.obs = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca_res$result) |>
  select(group, id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Individual cl and half-life from the model output.
indiv <- sim_sd |>
  group_by(id) |>
  summarise(cl_model = first(cl), vc_model = first(vc), .groups = "drop")
chk <- nca_wide |>
  left_join(indiv, by = "id") |>
  mutate(
    pct_cl = 100 * (cl.obs / cl_model - 1),
    pct_thalf = 100 * (half.life / (log(2) * vc_model / cl_model) - 1)
  )
# NCA clearance and half-life agree with the individual model values;
# residual differences are the trapezoidal AUC error and the extrapolation.
stopifnot(
  abs(median(chk$pct_cl)) < 3,
  quantile(abs(chk$pct_cl), 0.9) < 6,
  abs(median(chk$pct_thalf)) < 3
)

nca_wide |>
  group_by(group) |>
  summarise(
    across(c(cmax, aucinf.obs, half.life, cl.obs), median),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Age group" = group,
    "Cmax (IU/dL)" = cmax,
    "AUCinf (IU*h/dL)" = aucinf.obs,
    "Half-life (h)" = half.life,
    "CL (dL/h)" = cl.obs
  ) |>
  knitr::kable(digits = 2, caption = "Median simulated NCA after 50 IU/kg.")
```

# Figure 1: predicted FVIII:C in the activity ranges

Figure 1 of Tiede 2021 shows the first week of prophylaxis for two patients
dosed three times weekly with a 2-2-3 day interval pattern, with the
predicted FVIII:C time divided into the ranges 0-1%, >1-5%, >5-15%, >15-50%
and >50%. The patients' weights and doses are not reported, so the figure
below shows the typical 70-kg, 20-year-old patient given an illustrative
40 IU/kg on days 0, 2 and 4.

```{r figure-1}
ev_fig1 <- data.frame(
  id = 1L,
  time = c(0, 48, 96, seq(0, 168, by = 0.25)),
  evid = c(rep(1L, 3), rep(0L, length(seq(0, 168, by = 0.25)))),
  amt = c(rep(40 * 70, 3), rep(0, length(seq(0, 168, by = 0.25)))),
  cmt = "central",
  WT = 70,
  AGE = 20
) |>
  arrange(time, desc(evid))
sim_fig1 <- rxode2::rxSolve(mod_typical, events = ev_fig1) |> as.data.frame()

ggplot(sim_fig1, aes(time / 24, Cc)) +
  geom_hline(yintercept = c(1, 5, 15, 50), linetype = "dotted", colour = "grey50") +
  geom_line() +
  scale_y_log10(breaks = c(0.1, 1, 5, 15, 50, 100)) +
  labs(
    x = "Time (days)",
    y = "Predicted FVIII:C (%)",
    caption = "Illustrative analogue of Figure 1 of Tiede 2021 (typical patient, 40 IU/kg three times weekly)."
  )
```

# Table 3: proportion of time in each FVIII:C range

Tiede 2021 Table 3 gives the proportion of prophylaxis time spent in each
predicted FVIII:C range. Those proportions were computed from each patient's
actual diary doses, which are not available. Two weeks of an illustrative
three-times-weekly regimen of 40 IU/kg (days 0, 2, 4 of each week, the middle
of the 20-60 IU/kg label range) are simulated for the virtual cohort, and the
second week is classified into the paper's five ranges. The comparison is a
plausibility check of the model's decay rate, not a reproduction: the dose
and interval of each patient drive these proportions.

```{r table-3}
grid <- seq(168, 336, by = 0.25)
dose_times <- c(0, 48, 96, 168, 216, 264)
ev_proph <- subjects |>
  rowwise() |>
  reframe(
    id = id,
    group = group,
    WT = WT,
    AGE = AGE,
    time = c(dose_times, grid),
    evid = c(rep(1L, length(dose_times)), rep(0L, length(grid))),
    amt = c(rep(40 * WT, length(dose_times)), rep(0, length(grid))),
    cmt = "central"
  ) |>
  arrange(id, time, desc(evid))

sim_proph <- rxode2::rxSolve(mod, events = ev_proph, keep = "group") |>
  as.data.frame() |>
  filter(time < 336)

fviii_breaks <- c(-Inf, 1, 5, 15, 50, Inf)
fviii_labels <- c("0-1%", ">1-5%", ">5-15%", ">15-50%", ">50%")
time_in_range <- sim_proph |>
  mutate(range = cut(Cc, fviii_breaks, labels = fviii_labels, right = TRUE)) |>
  count(group, id, range) |>
  group_by(group, id) |>
  mutate(frac = n / sum(n)) |>
  ungroup() |>
  complete(nesting(group, id), range, fill = list(n = 0L, frac = 0))

sim_tab3 <- time_in_range |>
  group_by(group, range) |>
  summarise(simulated = round(100 * mean(frac), 1), .groups = "drop")

published_tab3 <- tibble(
  range = factor(rep(fviii_labels, 2), levels = fviii_labels),
  group = rep(c("Adults/adolescents", "Children"), each = 5),
  pivotal = c(15.58, 27.27, 25.97, 28.57, 3.90, 21.74, 26.09, 21.74, 21.74, 8.70),
  extension = c(12.85, 26.10, 25.50, 29.32, 6.22, 16.60, 26.73, 20.73, 23.04, 12.90)
)

tab3 <- sim_tab3 |>
  left_join(published_tab3, by = c("group", "range"))

# Broad plausibility bound on the 0-1% band: a 10-fold clearance or volume
# unit error drives it towards 0% or 100%. Published: 12.9-21.7%.
below1 <- time_in_range |>
  filter(range == "0-1%") |>
  group_by(group) |>
  summarise(pct = 100 * mean(frac), .groups = "drop")
stopifnot(all(below1$pct > 3), all(below1$pct < 40))

tab3 |>
  dplyr::rename(
    "Age group" = group,
    "FVIII:C range" = range,
    "Simulated, 40 IU/kg 3x weekly (%)" = simulated,
    "Table 3, pivotal (%)" = pivotal,
    "Table 3, extension (%)" = extension
  ) |>
  knitr::kable(caption = "Proportion of time in each predicted FVIII:C range.")
```

# Exposure-response: annualized bleeding rate by FVIII:C range

Tiede 2021 Table 4 reports the negative-binomial ABR of spontaneous bleeds
(and spontaneous joint bleeds) in each predicted FVIII:C range, pooled over
trial phases and age groups. The ABR decreases monotonically with FVIII:C.
Combined with the time each simulated patient spends in each range, the
Table 4 rates give an expected ABR under a given regimen. Table 5 of the paper
gives the same rates split by trial phase or by age group (not crossed); the
subgroup rates are much lower in the extension phase and in children at the
same FVIII:C range, which the authors attribute to joint status and
treatment history rather than FVIII:C.

```{r table-4}
tab4 <- tibble(
  range = factor(fviii_labels, levels = fviii_labels),
  pye = c(116.6, 214.5, 197.1, 223.7, 63.6),
  bleeds = c(303, 396, 371, 154, 13),
  joint_bleeds = c(241, 335, 337, 133, 9),
  abr_spont = c(4.16, 2.65, 2.14, 0.76, 0.21),
  abr_joint = c(3.44, 2.28, 1.99, 0.67, 0.15)
) |>
  mutate(crude_rate = bleeds / pye)

tab4 |>
  dplyr::rename(
    "FVIII:C range" = range,
    "PYE" = pye,
    "Spontaneous bleeds" = bleeds,
    "Spontaneous joint bleeds" = joint_bleeds,
    "ABR, spontaneous (NB estimate)" = abr_spont,
    "ABR, joint (NB estimate)" = abr_joint,
    "Bleeds / PYE" = crude_rate
  ) |>
  knitr::kable(digits = 2, caption = "Tiede 2021 Table 4.")

expected_abr <- time_in_range |>
  left_join(tab4 |> select(range, abr_spont, abr_joint), by = "range") |>
  group_by(group, id) |>
  summarise(
    abr_spont = sum(frac * abr_spont),
    abr_joint = sum(frac * abr_joint),
    .groups = "drop"
  )

expected_abr |>
  group_by(group) |>
  summarise(
    median_abr_spont = median(abr_spont),
    median_abr_joint = median(abr_joint),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Age group" = group,
    "Expected spontaneous ABR (median)" = median_abr_spont,
    "Expected joint ABR (median)" = median_abr_joint
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Expected ABR under 40 IU/kg three times weekly, using the Table 4 range-specific rates."
  )
```

For reference, the observed overall spontaneous ABR in the trials was 1237
bleeds over 815 patient-years (1.52 per year; Table 3).

# Assumptions and deviations

- **Parameter source.** Table 2 of Tiede 2021 cites the population PK
  parameters to Jimenez-Yuste 2015 (J Thromb Haemost 13:370-379). That paper
  was not accessible to the maintainers (publisher access blocked); all values
  here are as printed in Tiede 2021 Table 2 and its Results equations. Tiede
  2021 describes the PK data pool as the pivotal-trial PK subgroups plus
  routine post-dose samples from all patients (n = 231), while the
  Jimenez-Yuste 2015 abstract describes 76 patients from six trials; the model
  is packaged as printed in Tiede 2021.
- **Typical values multiply the covariate equations.** The printed equations
  for V(W) and CL(W, A) show only the covariate factors; the Table 2 reference
  values (for 70 kg and 20 years) are the multipliers, as their footnotes
  state.
- **IIV scale.** Table 2 reports inter-individual variability as a CV (0.32 on
  CL, 0.22 on V) for log-normal etas; the variances are
  `omega^2 = log(1 + CV^2)`. Reading the CV as the eta SD instead gives
  variances of 0.1024 and 0.0484, about 5% and 2% larger.
- **Residual error.** The Methods state a combined proportional and additive
  residual error model but no magnitudes are reported, so both are fixed to 0.
  Simulated `Cc` is therefore the individual prediction, which is also what
  the paper's exposure-response analysis used.
- **No endogenous FVIII baseline.** The model follows the printed closed form
  with no baseline FVIII:C term; all patients had severe hemophilia A
  (FVIII <= 1%).
- **Age and weight.** The paper does not state whether age and weight were
  time-varying. The Table 1 values are at entry into the trial programme.
  Age and weight are drawn independently in the virtual cohort (their
  correlation within each age group is not reported).
- **Exposure-response sub-model not packaged.** The ABR estimates are
  negative-binomial regression estimates per FVIII:C category (Tables 4 and 5,
  Figure 2); the overdispersion and the parametric model mentioned in the
  Methods are not reported. They are reproduced as tables and as an
  expected-ABR calculation in this article rather than as a model output.
- **Illustrative regimens.** Figure 1 and Table 3 depend on each patient's
  diary doses and intervals, which are not reported. The article uses 40 IU/kg
  three times weekly for both age groups; the resulting time-in-range
  proportions are a plausibility check only.
- **Population count.** The abstract states n = 187 patients (815
  patient-years), while Results and Table 1 give 231 patients (168 + 63). The
  model metadata uses 231.
- No erratum or correction to Tiede 2021 was found (checked 2026-09-28).
