---
title: "Ethambutol in African children (Tikiso 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ethambutol in African children (Tikiso 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  error = FALSE
)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)

rxode2::rxSetSeed(20220425)
set.seed(20220425)
```

Tikiso et al. (2022) pooled intensive steady-state ethambutol pharmacokinetic
data from 188 African children with tuberculosis enrolled in three studies --
DATiC (South Africa and Malawi), the DNDi study of lopinavir super-boosted with
additional ritonavir in children with HIV and TB (South Africa), and SHINE
(South Africa and Zambia) -- to inform paediatric ethambutol dosing.

```{r load_model}
mod <- nlmixr2lib::readModelDb("Tikiso_2022_ethambutol")
```

Reference: Tikiso T, McIlleron H, Abdelwahab MT, Bekker A, Hesseling A,
Chabala C, Davies G, Zar HJ, Rabie H, Andrieux-Meyer I, Lee J, Wiesner L,
Cotton MF, Denti P (2022). Population pharmacokinetics of ethambutol in African
children: a pooled analysis. *J Antimicrob Chemother* 77:1949-1959.
[doi:10.1093/jac/dkac127](https://doi.org/10.1093/jac/dkac127)

## Population

| Characteristic | DATiC | DNDi | SHINE | Combined |
|---|---|---|---|---|
| Children (male) | 79 (33) | 84 (36) | 25 (16) | 188 (85) |
| PK samples | 368 | 471 | 173 | 1012 |
| Age (years) | 2.6 (0.3-11.6) | 1.6 (0.3-6.8) | 3.1 (0.3-12.6) | 1.9 (0.3-12.6) |
| Weight (kg) | 11.1 (4.2-26.7) | 8.8 (3.9-14.9) | 11.8 (6.4-34.5) | 9.6 (3.9-34.5) |
| Fat-free mass (kg) | 9.1 (3.5-22.6) | 6.8 (3.3-13.6) | 9.2 (4.6-29.2) | 7.7 (3.3-29.2) |
| HIV-positive | 19 | 84 | 0 | 103 |
| On LPV/r-based ART | 11 | 84 | 0 | 95 |

Values are median (range), reproducing Table 1 of the source. Children received
ethambutol dihydrochloride 15-25 mg/kg once daily per WHO guidelines (median
20.2 mg/kg), as whole, crushed, syringe-delivered or nasogastric-tube tablets.
**Doses in the model are mg of ethambutol base**: the authors multiplied the
dihydrochloride dose by 0.737 (MW 204.31 / 277.23), so a dihydrochloride dose
must be converted the same way before simulation. Sampling was at steady state
before an observed dose and up to 8-12 h after it.

## Source trace

Every `ini()` value carries an in-file comment naming its source location; this
table is the vignette-level summary.

| Quantity | Value | Source |
|---|---|---|
| Structure | 2-compartment, Savic transit absorption into a first-order depot | Results, Structural model; supplementary control stream `$DES` |
| CL/F, Vc/F, Q/F, Vp/F (FFM 7.7 kg, mature) | 15.9 L/h, 44.3 L, 11.5 L/h, 86.2 L | Table 2 + footnote b; control stream `TVFFM = 7.7` |
| ka, MTT, transit n | 1.43 1/h, 40.4 min, 4.82 | Table 2 |
| F | 1 fixed | Table 2 |
| Allometry | FFM, exponents 0.75 / 1 fixed | Methods; control stream `ALLMCL_FFM_CH`, `ALLMV_FFM_CH` |
| Paediatric FFM equation | Al-Sallami multiplier on Janmahasatian, sex-specific constants | Control stream `$PK` |
| Maturation | `PMA^g / (PMA^g + PMA50^g)`, PMA50 10.8 months, g 3.25; PMA = AGE + GA/52 years | Table 2; Methods; control stream `MATCL` |
| Study effect on absorption | ka x 0.764, MTT / 0.764 in DNDi and SHINE | Table 2 + footnote c |
| LPV/r effect on F | -32.0% | Table 2 |
| Age on F | +0.0853 per year below a 3.16-year hinge, flat above | Table 2; control stream `AGE_BIO` with `THETA(17)` FIX 0 |
| BSV CL | 11.6% | Table 2 |
| BOV F / ka / MTT | 33.1% / 63.7% / 48.5%; F BOV x 1.37 for the unobserved dose | Table 2; control stream `SCALE_BOVBIO` |
| Residual error | 17.7% proportional + 0.0168 mg/L additive (fixed) | Table 2 + footnote d |

The supplementary NONMEM control stream fixes the structure, the covariate
equations and the random-effect layout. Its `$THETA` / `$OMEGA` values are
initial estimates for the final run, so Table 2 supplies every value.

## Structural verification

These are exact identities of the published parameterisation, compared against
the model's own solve with the same parameters on both sides, so they are
asserted tightly.

```{r helpers}
# Daily dosing into the depot for eight days, dense observations over the last
# 24 h interval on the central compartment (rxode2 returns Cc at those rows).
ss_events <- function(amt_df, obs_step = 0.05) {
  doses <- amt_df |>
    dplyr::rowwise() |>
    dplyr::reframe(
      id = id, time = seq(0, 168, by = 24), amt = amt, evid = 1L,
      cmt = "depot"
    )
  obs <- amt_df |>
    dplyr::select(id) |>
    dplyr::cross_join(data.frame(time = seq(168, 192, by = obs_step))) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::left_join(dplyr::select(amt_df, -amt), by = "id")
}

trap <- function(time, conc) {
  sum(diff(time) * (utils::head(conc, -1) + utils::tail(conc, -1)) / 2)
}

typical <- function(cov, amt = 150) {
  ev <- ss_events(data.frame(id = 1L, amt = amt, cov))
  out <- rxode2::rxSolve(rxode2::zeroRe(mod), ev, returnType = "data.frame")
  out[out$time >= 168 & !duplicated(out$time), ]
}

ref_child <- data.frame(
  WT = 10, HT = 82, SEXF = 0, AGE = 4, GA = 39,
  CONMED_LOPINAVIR = 0, STUDY_DNDI_SHINE = 0, OCC = 2
)
```

**Steady-state mass balance.** Over one steady-state interval the eliminated
amount equals the absorbed dose, so `CL * AUCtau = F * Dose`. This pins the unit
chain (mg, L, mg/L) and proves that the explicit Savic transit density delivers
exactly one dose per administration with the ordinary depot bolus suppressed.

```{r gate_mass}
tv <- typical(ref_child)
mass <- tv$cl[1] * trap(tv$time, tv$Cc) / (tv$fdepot[1] * 150)
stopifnot(
  abs(mass - 1) < 1e-4,
  identical(rxode2::rxode2(mod)$state, c("depot", "central", "peripheral1"))
)
c(ffm_kg = tv$ffm[1], cl_L_h = tv$cl[1], cl_auc_over_f_dose = mass)
```

**Covariate identities quoted in the Abstract.** With the default gestational
age of 39 weeks, clearance is at 50% of its mature value at 10.8 months
postmenstrual age (about 2 months after birth) and at 99% by 3 years; and the
bioavailability at birth is `1 - 0.0853 * 3.16` = 73.1% of that in children at
or above the 3.16-year hinge. Lopinavir/ritonavir lowers F by exactly 32%, and
the DNDi / SHINE study effect slows absorption without changing exposure.

```{r gate_covariates}
at_age <- function(age, ...) {
  cov <- ref_child
  cov$AGE <- age
  extra <- list(...)
  for (nm in names(extra)) cov[[nm]] <- extra[[nm]]
  typical(cov)[1, ]
}
m_pma50 <- at_age(10.8 / 12 - 39 / 52)$mat_cl
m_2mo <- at_age(2 / 12)$mat_cl
m_3y <- at_age(3)$mat_cl
f_birth <- at_age(0)$fdepot
f_lpv <- at_age(4, CONMED_LOPINAVIR = 1)$fdepot

auc_datic <- trap(tv$time, tv$Cc)
tv_shine <- typical(transform(ref_child, STUDY_DNDI_SHINE = 1))
auc_shine <- trap(tv_shine$time, tv_shine$Cc)

stopifnot(
  abs(m_pma50 - 0.5) < 1e-6,
  abs(m_3y - 0.99) < 0.005,
  abs(f_birth - 0.731) < 0.001,
  abs(f_lpv - 0.68) < 1e-6,
  abs(auc_shine / auc_datic - 1) < 1e-3,
  max(tv_shine$Cc) < max(tv$Cc)
)
c(
  maturation_at_pma50 = m_pma50, maturation_2_months = m_2mo,
  maturation_3_years = m_3y, F_at_birth = f_birth, F_on_LPVr = f_lpv,
  Cmax_ratio_DNDi_SHINE_vs_DATiC = max(tv_shine$Cc) / max(tv$Cc)
)
```

## Virtual cohort

The individual data are not public, so the cohort is synthetic. Ages are drawn
per study from a log-normal centred on the Table 1 median and truncated to the
study's range. Weight and height follow approximate WHO median growth curves,
scaled down to the cohort's undernutrition: the factor 0.83 maps the WHO median
weight at the pooled median age (1.9 years, about 11.5 kg) onto the Table 1
pooled median of 9.6 kg; height is scaled by 0.95. Gestational age is 39 weeks
(the paper's default when unknown) and 55% of children are female.

```{r cohort_helpers}
who <- data.frame(
  age = c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13),
  wt = c(3.3, 6.0, 7.6, 8.7, 9.3, 10.6, 11.9, 14.1, 16.1, 18.1, 20.3, 22.6,
         25.2, 28.2, 31.6, 35.4, 40.0, 45.0),
  ht = c(49.5, 60.5, 66.5, 70.9, 74.6, 81.0, 86.8, 95.6, 102.8, 109.6, 115.9,
         121.8, 127.5, 132.9, 138.4, 143.9, 149.5, 156.0)
)
wt_factor <- 0.83
ht_factor <- 0.95

draw_age <- function(n, median, lo, hi) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rlnorm(4 * n, log(median), 0.8)
    out <- c(out, x[x >= lo & x <= hi])
  }
  out[seq_len(n)]
}

grow <- function(age) {
  n <- length(age)
  data.frame(
    AGE = age,
    WT = stats::approx(who$age, who$wt, age)$y * wt_factor *
      exp(stats::rnorm(n, 0, 0.10)),
    HT = stats::approx(who$age, who$ht, age)$y * ht_factor *
      exp(stats::rnorm(n, 0, 0.03))
  )
}

make_arm <- function(n, mix, lpv) {
  parts <- lapply(seq_len(nrow(mix)), function(i) {
    k <- round(n * mix$frac[i])
    cbind(
      grow(draw_age(k, mix$median[i], mix$lo[i], mix$hi[i])),
      STUDY_DNDI_SHINE = mix$study[i]
    )
  })
  arm <- do.call(rbind, parts)
  arm$SEXF <- stats::rbinom(nrow(arm), 1, 0.55)
  arm$GA <- 39
  arm$CONMED_LOPINAVIR <- lpv
  arm$OCC <- 2
  arm
}
```

The two arms mirror the study composition of Table 1: children not on
lopinavir/ritonavir come from DATiC (68 of 93) and SHINE (25), and children on
lopinavir/ritonavir from DNDi (84 of 95) and DATiC (11). Each child receives
the cohort-median 20.2 mg/kg of ethambutol dihydrochloride once daily, i.e.
`20.2 * 0.737` mg/kg of ethambutol base.

```{r cohort}
n_arm <- 200
mix_no_lpv <- data.frame(
  frac = c(68, 25) / 93, median = c(2.6, 3.1), lo = 0.3, hi = c(11.6, 12.6),
  study = c(0, 1)
)
mix_lpv <- data.frame(
  frac = c(84, 11) / 95, median = c(1.6, 2.6), lo = 0.3, hi = c(6.8, 11.6),
  study = c(1, 0)
)
cohort <- dplyr::bind_rows(
  dplyr::mutate(make_arm(n_arm, mix_no_lpv, 0), arm = "No LPV/r"),
  dplyr::mutate(make_arm(n_arm, mix_lpv, 1), arm = "LPV/r")
) |>
  dplyr::mutate(id = dplyr::row_number(), amt = 20.2 * 0.737 * WT)

cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    n = dplyr::n(),
    "Age (years)" = sprintf("%.1f (%.1f-%.1f)", median(AGE), min(AGE), max(AGE)),
    "Weight (kg)" = sprintf("%.1f (%.1f-%.1f)", median(WT), min(WT), max(WT)),
    .groups = "drop"
  ) |>
  dplyr::rename(Arm = arm) |>
  knitr::kable(caption = "Simulated cohort, median (range).")
```

## Simulation

```{r simulate}
ev <- ss_events(dplyr::select(cohort, -arm), obs_step = 0.25)
sim <- rxode2::rxSolve(mod, ev, returnType = "data.frame") |>
  dplyr::left_join(dplyr::select(cohort, id, arm), by = "id")
stopifnot(!anyNA(sim$Cc))

ss <- sim |>
  dplyr::filter(time >= 168) |>
  dplyr::mutate(tad = time - 168)
```

```{r fig_profiles, fig.cap = "Simulated steady-state ethambutol plasma concentrations (median and 5th-95th percentiles of individual predictions), at 20.2 mg/kg ethambutol dihydrochloride once daily, by lopinavir/ritonavir co-treatment. Compare with the Figure 1 VPC of Tikiso 2022 stratified by HIV treatment."}
ss |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, p50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Ethambutol (mg/L)",
    colour = NULL, fill = NULL
  ) +
  theme_bw()
```

## PKNCA validation

```{r pknca}
conc <- ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, arm, tad, Cc)
dose <- cohort |>
  dplyr::transmute(id, arm, tad = 0, amt)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ tad | arm + id)
dose_obj <- PKNCA::PKNCAdose(dose, amt ~ tad | arm + id)
intervals <- data.frame(
  start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE
)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)
```

The paper's Discussion reports the model-predicted steady-state ethambutol
Cmax for the study children at the median 20.2 mg/kg dose: 1.66 mg/L (IQR
1.21-2.15) at 2.74 h without lopinavir/ritonavir and 0.882 mg/L (0.669-1.28)
at 3.0 h with it.

```{r nca_compare}
published <- data.frame(
  arm = c("No LPV/r", "LPV/r"),
  cmax = c(1.66, 0.882),
  tmax = c(2.74, 3.0)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = dplyr::filter(nca_res, PPTESTCD %in% c("cmax", "tmax")),
  reference = published,
  by = "arm",
  units = c(cmax = "mg/L", tmax = "h"),
  tolerance_pct = 20
)
cmp |>
  dplyr::rename("Arm" = arm) |>
  knitr::kable(caption = "Simulated vs published median steady-state Cmax and Tmax.")

auc_summary <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    value = sprintf(
      "%.2f (%.2f-%.2f)", median(PPORRES), quantile(PPORRES, 0.25),
      quantile(PPORRES, 0.75)
    ),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value)
auc_summary |>
  dplyr::rename(
    "Arm" = arm, "Cmax (mg/L)" = cmax, "AUC0-24 (mg*h/L)" = auclast
  ) |>
  knitr::kable(caption = "Simulated steady-state exposure, median (IQR).")
```

Median Cmax reproduces the published values in both arms. Median Tmax is
about 0.7 h earlier than published in both arms (flagged rows). The shift is
the same in both arms, so it does not come from the lopinavir/ritonavir or the
study effects. The typical DATiC child's Tmax is about 1.8 h from the Table 2
ka (1.43 1/h) and MTT (40.4 min). Tmax rises with both quantities, so a median
of 2.74 h would need a typical absorption roughly twice as slow as Table 2
reports. The published Tmax values are post hoc predictions for the actual
children, whose individual absorption estimates (between-occasion variability
of 63.7% on ka and 48.5% on MTT, crushed tablets in the youngest) need not
centre on the typical values. The rate constants are kept exactly as
published.

```{r nca_gate}
med_cmax <- nca_res |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(med = median(PPORRES), .groups = "drop")
med_no <- med_cmax$med[med_cmax$arm == "No LPV/r"]
med_lpv <- med_cmax$med[med_cmax$arm == "LPV/r"]
stopifnot(
  # Centre of the distribution: a mis-transcribed clearance, volume, dose
  # conversion or absorption parameter moves the median by tens of percent.
  abs(med_no / 1.66 - 1) < 0.25,
  abs(med_lpv / 0.882 - 1) < 0.25,
  # The co-treatment contrast: published 0.882 / 1.66 = 0.53.
  abs((med_lpv / med_no) / (0.882 / 1.66) - 1) < 0.25
)
c(no_lpv = med_no, lpv = med_lpv, ratio = med_lpv / med_no)
```

## Weight-band dosing (Figures 2 and 3)

The paper simulated exposures by WHO weight band with the WHO-recommended doses
and with its proposed optimised doses (Table 3, mg of ethambutol
dihydrochloride), and compared Cmax with 2-6 mg/L and AUC0-24 with 16-29
mg*h/L. Here 25 children per weight band (200 per arm) are drawn uniformly
within each band, with age obtained by inverting the scaled WHO weight-for-age
curve; study is set to the DATiC reference.

```{r bands}
bands <- data.frame(
  band = c("3-3.9", "4-5.9", "6-7.9", "8-11.9", "12-14.9", "15-19.9",
           "20-24.9", "25-35.9"),
  lo = c(3, 4, 6, 8, 12, 15, 20, 25),
  hi = c(3.9, 5.9, 7.9, 11.9, 14.9, 19.9, 24.9, 35.9),
  who = c(75, 100, 150, 200, 300, 400, 400, 600),
  opt_no_lpv = c(200, 300, 400, 600, 800, 800, 800, 1200),
  opt_lpv = c(300, 500, 800, 800, 1200, 1600, 1600, 1600)
)
band_kids <- bands |>
  dplyr::rowwise() |>
  dplyr::reframe(band = band, WT = stats::runif(25, lo, hi)) |>
  dplyr::mutate(
    AGE = pmax(stats::approx(who$wt * wt_factor, who$age, WT, rule = 2)$y, 0.05),
    HT = stats::approx(who$age, who$ht, AGE)$y * ht_factor,
    SEXF = stats::rbinom(dplyr::n(), 1, 0.55),
    GA = 39, STUDY_DNDI_SHINE = 0, OCC = 2
  ) |>
  dplyr::left_join(bands, by = "band")

band_cohort <- dplyr::bind_rows(
  dplyr::mutate(band_kids, CONMED_LOPINAVIR = 0, arm = "No LPV/r",
                regimen = "WHO", salt_mg = who),
  dplyr::mutate(band_kids, CONMED_LOPINAVIR = 1, arm = "LPV/r",
                regimen = "WHO", salt_mg = who),
  dplyr::mutate(band_kids, CONMED_LOPINAVIR = 0, arm = "No LPV/r",
                regimen = "Optimised", salt_mg = opt_no_lpv),
  dplyr::mutate(band_kids, CONMED_LOPINAVIR = 1, arm = "LPV/r",
                regimen = "Optimised", salt_mg = opt_lpv)
) |>
  dplyr::mutate(id = dplyr::row_number(), amt = salt_mg * 0.737)

band_ev <- ss_events(
  dplyr::select(band_cohort, id, amt, WT, HT, AGE, SEXF, GA,
                STUDY_DNDI_SHINE, OCC, CONMED_LOPINAVIR),
  obs_step = 0.25
)
band_sim <- rxode2::rxSolve(mod, band_ev, returnType = "data.frame") |>
  dplyr::filter(time >= 168)
stopifnot(!anyNA(band_sim$Cc))

band_exp <- band_sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cmax = max(Cc), auc = trap(time, Cc), .groups = "drop"
  ) |>
  dplyr::left_join(dplyr::select(band_cohort, id, band, arm, regimen), by = "id") |>
  dplyr::mutate(
    band = factor(band, levels = bands$band),
    regimen = factor(regimen, levels = c("WHO", "Optimised"))
  )
```

```{r fig2, fig.cap = "Simulated steady-state ethambutol Cmax by weight band with WHO-recommended (left) and optimised (right) dosing; shaded band 2-6 mg/L. Replicates Figure 2 of Tikiso 2022."}
ggplot(band_exp, aes(band, cmax, fill = arm)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 2, ymax = 6,
           alpha = 0.2, fill = "orange") +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~regimen) +
  labs(x = "Weight band (kg)", y = "Cmax (mg/L)", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig3, fig.cap = "Simulated steady-state ethambutol AUC0-24 by weight band with WHO-recommended (left) and optimised (right) dosing; shaded band 16-29 mg*h/L. Replicates Figure 3 of Tikiso 2022."}
ggplot(band_exp, aes(band, auc, fill = arm)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 16, ymax = 29,
           alpha = 0.2, fill = "orange") +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~regimen) +
  labs(x = "Weight band (kg)", y = "AUC0-24 (mg*h/L)", fill = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The paper concludes that WHO doses leave most children below the adult Cmax
and AUC ranges, and that roughly doubled (no LPV/r) or tripled (LPV/r) doses
bring the typical child into them.

```{r band_gate}
band_med <- band_exp |>
  dplyr::group_by(regimen, arm) |>
  dplyr::summarise(
    cmax = median(cmax), auc = median(auc), .groups = "drop"
  )
knitr::kable(
  dplyr::rename(band_med, "Regimen" = regimen, "Arm" = arm,
                "Median Cmax (mg/L)" = cmax, "Median AUC0-24 (mg*h/L)" = auc),
  digits = 2,
  caption = "Median simulated steady-state exposure across all weight bands."
)
who_med <- dplyr::filter(band_med, regimen == "WHO")
opt_med <- dplyr::filter(band_med, regimen == "Optimised")
stopifnot(
  all(who_med$cmax < 2),
  all(who_med$auc < 16),
  all(opt_med$cmax > 2 & opt_med$cmax < 6),
  all(opt_med$auc > 16 & opt_med$auc < 29)
)
```

## Assumptions and deviations

- **Dose basis.** Every disposition parameter is apparent and relative to the
  dose expressed as ethambutol base (dihydrochloride x 0.737). The paper notes
  that comparing them with reports that dosed the salt requires multiplying by
  1.357.
- **Table 2 values, not control-stream initials.** The supplementary control
  stream's `$THETA` and `$OMEGA` are the starting values of the final run (for
  example CL 21.8 L/h on the salt-dose scale, i.e. about 16.1 L/h after the
  0.737 conversion). Table 2 is used for every value.
- **Variability scale.** Table 2 percentages are read as the log-scale omega
  (`omega^2 = (CV/100)^2`): the control-stream initials for BOV on ka (0.403)
  and MTT (0.233) reproduce the tabulated 63.7% and 48.5% as `sqrt(omega^2)`
  (63.5%, 48.3%), not as `sqrt(exp(omega^2) - 1)` (70.4%, 51.0%).
- **Inflated BOV on F for the unobserved dose** is encoded by fixing the
  occasion-1 variance to `1.37^2 * 0.331^2`, an exact re-expression of the
  control stream's scaling of the occasion-1 deviate. The simulations above use
  occasion 2 (the observed dose) throughout.
- **Between-occasion etas are not mu-referenced.** rxode2 has no NONMEM-style
  occasion level, so the two-occasion BOV is expanded into indicator-selected
  etas and rxode2 warns that these etas are not mu-referenced. This only
  affects estimation speed-ups, not simulation.
- **Explicit transit input.** The Savic transit density is written out with
  `podo()` / `tad()` exactly as in the control stream `$DES`, instead of
  rxode2's `transit()` built-in, which yields a zero input when combined with
  `f(depot) <- 0`. Each dose restarts the transit input; with MTT about 0.7 h
  against a 24 h interval no dose is lost, as the mass-balance check confirms.
- **FFM is computed inside the model** from `WT`, `HT`, `AGE` and `SEXF` with
  the control stream's paediatric equation, so these four columns are needed
  even though only FFM enters the allometry. The reference FFM is the 7.7 kg
  cohort median (`TVFFM`), which Table 2 footnote b describes as a 10 kg child.
- **Clearance maturation is not normalised**: Table 2 CL is the fully mature
  value, so a 4-year-old has 99.6% of it. PMA50 is coded in months (Table 2);
  postmenstrual age is `(AGE + GA/52) * 12` months.
- **Censoring-specific error inflation** (extra additive error for imputed
  BLQ records, near-infinite error for trailing BLQ records) belongs to the
  estimation dataset and is not part of the simulation model.
- **Virtual cohort.** Growth curves are approximate WHO medians scaled to the
  cohort's undernutrition (factors 0.83 for weight and 0.95 for height),
  ages are log-normal by study, gestational age is 39 weeks, and every child
  receives exactly 20.2 mg/kg rather than a weight-band dose. The published
  Cmax values are post hoc predictions for the actual children, so the
  comparison checks the centre of the distribution rather than its tails.
- **Weight-band simulations** use the DATiC reference for the study effect
  (the paper does not state which study its simulations assumed) and 25
  children per band instead of the paper's more than 50,000.
