---
title: "Timmermann_2019_brodalumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Timmermann_2019_brodalumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Timmermann_2019_brodalumab"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Timmermann_2019_brodalumab"))$description`
* Article: [Basic Clin Pharmacol Toxicol. 2019;125(1):16-25](https://doi.org/10.1111/bcpt.13202)
  (open access via [PMC6850607](https://pmc.ncbi.nlm.nih.gov/articles/PMC6850607/))

# Population

Timmermann 2019 pooled six clinical studies (2 phase I, 1 phase IIb with open-label
extension, 3 phase III) into a population PK analysis set of 622 adults with
moderate-to-severe plaque psoriasis. The dataset included 7725 quantifiable and
2508 below-LLOQ brodalumab concentrations following IV or SC single / multiple
doses of 70-700 mg administered over 1-2015 days. The approved regimen is 210 mg
SC at weeks 0, 1, 2 followed by 210 mg every 2 weeks ("Q2W+1").

Baseline demographics from Table 2: median age 46 years (range 18-75), 33% female
(417 of 622 were male), median body weight 87.8 kg (range 43-186), 93% Caucasian
(1% Asian, 1% Black, 5% other per the corresponding Table 2 breakdown in the
source), median BMI 29.6 kg/m^2 (range 16.7-66.1), median baseline PASI 17.7
(range 8.8-60.6). Healthy volunteers were deliberately excluded because IL-17RA
expression (and therefore target-mediated disposition) differs between diseased
and healthy populations. Race was not tested as a covariate because 93% of
patients were Caucasian; anti-drug-antibody status was not tested due to low and
transient positivity.

The same information is available programmatically via
`readModelDb("Timmermann_2019_brodalumab")$population`.

# Source trace

Every structural parameter, covariate effect, IIV element, and residual-error
term below is taken from Timmermann 2019 Table 3. The reference patient is a
90 kg adult with plaque psoriasis; all body-weight normalizations use
`WT / 90`.

| Equation / parameter                     | Value                                   | Source location |
|------------------------------------------|-----------------------------------------|-----------------|
| `lka`  (Ka)                              | `log(0.300)` 1/day                      | Table 3, Ka row |
| `lvc`  (V1)                              | `log(4.68)` L                           | Table 3, V1 row |
| `lcl`  (CL)                              | `log(0.155)` L/day                      | Table 3, CL row |
| `lvm`  (Vmax)                            | `log(6.07)` mg/day                      | Table 3, Vmax row |
| `lq`   (Q)                               | `log(0.328)` L/day                      | Table 3, Q row |
| `lvp`  (V2)                              | `log(2.41)` L                           | Table 3, V2 row |
| `km`   (Km, fixed)                       | `0.02` ug/mL                            | Table 3, Km row (fixed) |
| `lfdepot` (F, fixed)                     | `log(0.548)`                            | Table 3, F row (fixed at 54.8%) |
| `e_wt_cl`   (WT/90 exponent on CL)       | `0.767`                                 | Table 3: Power of weight on CL |
| `e_wt_vc`   (WT/90 exponent on V1)       | `0.938`                                 | Table 3: Power of weight on V1 |
| `e_wt_vm`   (WT/90 exponent on Vmax)     | `0.769`                                 | Table 3: Power of weight on Vmax |
| `var(etalka)`                            | `log(0.626^2 + 1) = 0.33065`            | Table 3: Ka IIV 62.6% CV |
| `var(etalcl)`                            | `log(0.575^2 + 1) = 0.28565`            | Table 3: CL IIV 57.5% CV |
| `var(etalvc)`                            | `log(0.255^2 + 1) = 0.06300`            | Table 3: V1 IIV 25.5% CV |
| `cov(etalcl, etalvc)`                    | `0.75 * sqrt(0.28565 * 0.06300) = 0.10061` | Table 3: CL-V1 correlation 0.75 |
| `var(etalvm)`                            | `log(0.247^2 + 1) = 0.05922`            | Table 3: Vmax IIV 24.7% CV |
| `var(etalq)`                             | `log(0.910^2 + 1) = 0.60328`            | Table 3: Q IIV 91% CV |
| `var(etalvp)`                            | `log(1.890^2 + 1) = 1.51997`            | Table 3: V2 IIV 189% CV |
| `CcpropSd`                               | `0.355` (fraction)                      | Table 3: Proportional residual error 35.5% CV |
| `CcaddSd`                                | `3.00` ug/mL                            | Table 3: Additive residual error (SD, per footnote) |
| Structure (2-cmt + first-order SC absorption + parallel linear + MM elimination) | n/a | Section 2.3 Methods and ODE description |

## Parameterization notes

- **Parallel linear + Michaelis-Menten clearance.** Timmermann 2019 parameterizes
  elimination from the central compartment as a sum of first-order linear
  clearance (`CL`) and saturable Michaelis-Menten elimination with parameters
  `Vmax` (mg/day) and `Km` (ug/mL). Explicit ODEs are used rather than
  `linCmt()` because of the non-linear term.
- **Separate bioavailability.** Because phase I included IV (700 mg bolus) and SC
  arms, the authors identified the SC bioavailability F separately and fixed it
  at 54.8% before the covariate analysis. In the nlmixr2 implementation,
  `f(depot) <- exp(lfdepot)` applies F only to doses entering the `depot`
  compartment; IV doses that the user places directly in `central` are not
  scaled.
- **Fixed Km.** Km was fixed at 0.02 ug/mL, a value carried over from the
  previous analysis (Endres 2014). The authors note that Km is effectively an
  artificial constant because the non-linear term saturates well below 1 ug/mL
  and a zero-order elimination would describe the data equivalently but allow
  negative concentrations; Km was kept for pharmacological plausibility.
- **CV% to log-normal variance.** Timmermann 2019 Table 3 reports between-subject
  variability as CV% on the linear-parameter scale, with multiplicative
  exponential (log-normal) random effects per Section 2.3. The nlmixr2 internal
  variance is therefore `omega^2 = log(CV^2 + 1)`.
- **CL-V1 correlation.** Table 3 reports a 0.75 correlation between the etas on
  CL and V1, encoded as a 2x2 block in `ini()` with the off-diagonal computed
  as the correlation times `sqrt(var_cl * var_vc)`.
- **Combined residual error.** Table 3 footnote states "Proportional and
  additive errors are given as %CV and standard deviation." `CcpropSd = 0.355`
  (fraction) and `CcaddSd = 3.00` ug/mL are used directly.

# Virtual cohort

The simulations below use a virtual cohort whose body-weight distribution
approximates Timmermann 2019 Table 2. No subject-level observed data were
released with the paper.

```{r cohort}
set.seed(20260424)
n_subj <- 400

cohort <- tibble::tibble(
  id = seq_len(n_subj),
  WT = pmin(pmax(rnorm(n_subj, mean = 87.8, sd = 20), 43, 186))
)
```

The cohort receives the approved labelled regimen: 210 mg SC at weeks 0, 1, 2,
followed by 210 mg Q2W. The dosing horizon is extended well past the time to
reach 90% of steady state (reported as 10 weeks for the reference 90 kg
patient) to ensure the final Q2W cycle used for NCA is at steady state.

```{r events}
tau <- 14L                          # Q2W interval (days) after the loading phase
loading_days <- c(0, 7, 14)         # 210 mg at weeks 0, 1, 2
maint_days   <- seq(28, 28 + tau * 40, by = tau)
dose_days    <- c(loading_days, maint_days)

obs_days <- sort(unique(c(
  seq(0, max(dose_days) + tau, by = 1),
  dose_days + 0.25,
  dose_days + 1,
  dose_days + 3,
  dose_days + 7
)))

build_events <- function(cohort, dose_amt) {
  ev_dose <- cohort |>
    tidyr::crossing(time = dose_days) |>
    dplyr::mutate(amt = dose_amt, cmt = "depot", evid = 1L)
  ev_obs <- cohort |>
    tidyr::crossing(time = obs_days) |>
    dplyr::mutate(amt = 0, cmt = NA_character_, evid = 0L)
  dplyr::bind_rows(ev_dose, ev_obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::select(id, time, amt, cmt, evid, WT)
}

events <- build_events(cohort, 210)
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("Timmermann_2019_brodalumab"))
sim <- as.data.frame(rxode2::rxSolve(mod, events = events, keep = "WT"))
```

# Replicate published figures

## Figure 4 - Typical-patient concentration-time profile (210 mg Q2W+1, 90 kg)

Timmermann 2019 Figure 4 shows the typical 90 kg patient's predicted
concentration-time profile over the first 24 weeks of 210 mg Q2W+1 dosing
together with a zoom on the first two weeks (loading phase). The block below
reproduces the typical-patient profile with all random effects zeroed.

```{r figure-4-typical, fig.width = 7.5, fig.height = 5}
mod_typical <- mod |> rxode2::zeroRe()

typ_events <- build_events(tibble::tibble(id = 1L, WT = 90), 210)
sim_typ <- as.data.frame(rxode2::rxSolve(mod_typical, events = typ_events, keep = "WT"))

ggplot(sim_typ |> dplyr::filter(time <= 168), aes(time, Cc)) +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  labs(
    x = "Time (days)",
    y = "Brodalumab Cc (ug/mL)",
    title = "Typical 90 kg patient: 210 mg Q2W+1 (first 24 weeks)",
    caption = "Reproduces the typical-patient trace underlying Timmermann 2019 Figure 4."
  ) +
  theme_minimal()
```

## Figure 2 - VPC percentiles over the first 24 weeks (virtual cohort)

The simulated 5th, 50th, and 95th percentiles across the virtual cohort
(N = 400, weight distribution per Table 2) for the first 168 days of 210 mg
Q2W+1 dosing.

```{r figure-2-vpc, fig.width = 7.5, fig.height = 4.5}
vpc <- sim |>
  dplyr::filter(!is.na(Cc), time > 0, time <= 168) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(vpc, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  labs(
    x = "Time (days)",
    y = "Brodalumab Cc (ug/mL)",
    title = "Simulated 5-50-95 percentiles: 210 mg Q2W+1 (first 24 weeks)",
    caption = "Virtual psoriasis cohort (N = 400). Replicates the 0-168 day panel of Timmermann 2019 Figure 2."
  ) +
  theme_minimal()
```

## Figure 3 - Fold-change in AUCss with body weight

Timmermann 2019 Figure 3 shows the geometric-mean fold-change in AUCss (relative
to the 90 kg reference) for subjects of 60, 80, 90, 100, and 120 kg on 210 mg
Q2W+1. The block below reproduces that relationship using 200 subjects at each
fixed weight and computing AUC over the final Q2W dosing interval.

```{r figure-3-auc-weight, fig.width = 7, fig.height = 4}
ss_start <- max(maint_days)
ss_end   <- ss_start + tau
wt_levels <- c(60, 80, 90, 100, 120)

make_wt_cohort <- function(wt, n_per, id_offset) {
  tibble::tibble(id = id_offset + seq_len(n_per), WT = wt)
}

auc_by_wt <- lapply(seq_along(wt_levels), function(i) {
  wt <- wt_levels[i]
  c_wt <- make_wt_cohort(wt, 200, id_offset = (i - 1L) * 1000L)
  ev_wt <- build_events(c_wt, 210)
  sim_wt <- as.data.frame(rxode2::rxSolve(mod, events = ev_wt, keep = "WT"))
  ss <- sim_wt |>
    dplyr::filter(time >= ss_start, time <= ss_end, !is.na(Cc)) |>
    dplyr::arrange(id, time)
  ss |>
    dplyr::group_by(id) |>
    dplyr::summarise(
      AUCss = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      WT = dplyr::first(WT), .groups = "drop"
    )
}) |> dplyr::bind_rows()

ref_auc <- auc_by_wt |>
  dplyr::filter(WT == 90) |>
  dplyr::summarise(geo_mean = exp(mean(log(AUCss)))) |>
  dplyr::pull(geo_mean)

fold_change <- auc_by_wt |>
  dplyr::group_by(WT) |>
  dplyr::summarise(
    geo_mean_ratio = exp(mean(log(AUCss))) / ref_auc,
    .groups = "drop"
  )

ggplot(fold_change, aes(WT, geo_mean_ratio)) +
  geom_hline(yintercept = c(0.8, 1, 1.25), linetype = c("dotted", "solid", "dotted"),
             colour = c("steelblue", "grey40", "steelblue")) +
  geom_point(size = 3) +
  scale_y_continuous(limits = c(0.4, 2.4)) +
  labs(
    x = "Body weight (kg)",
    y = "Geometric-mean AUCss ratio (reference = 90 kg)",
    title = "Fold-change in AUCss with body weight (210 mg Q2W+1)",
    caption = "Replicates Figure 3 of Timmermann 2019; expected ~>2-fold at 60 kg and ~0.5 at 120 kg."
  ) +
  theme_minimal()
```

# PKNCA validation

Non-compartmental analysis of the final (steady-state) Q2W dosing interval of
the 210 mg regimen. Cmax, Tmax, Cmin/Ctrough, AUC0-tau, and Cavg are computed
per simulated subject.

```{r pknca}
nca_conc <- sim |>
  dplyr::filter(time >= ss_start, time <= ss_end, !is.na(Cc)) |>
  dplyr::mutate(time_nom = time - ss_start, treatment = "210mg_Q2W") |>
  dplyr::select(id, time = time_nom, Cc, treatment)

nca_dose <- cohort |>
  dplyr::mutate(time = 0, amt = 210, treatment = "210mg_Q2W") |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id)

intervals <- data.frame(
  start   = 0,
  end     = tau,
  cmax    = TRUE,
  cmin    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
summary(nca_res)
```

## Comparison against Timmermann 2019 Table 4

Table 4 reports Cmax at week 1, Cmax at steady state, tmax at steady state, and
AUCss for a reference 90 kg patient dosed 210 mg Q2W+1. The block below
simulates 500 reference (90 kg) patients to extract the corresponding
distributional statistics and compares mean, median, and CV% against the
published values.

```{r table-4-comparison}
set.seed(20260424)
ref_cohort <- tibble::tibble(id = seq_len(500), WT = 90)
ref_events <- build_events(ref_cohort, 210)
sim_ref <- as.data.frame(rxode2::rxSolve(mod, events = ref_events, keep = "WT"))

week1_start <- 0                    # Cmax within the first dosing week (after loading dose 1)
week1_end   <- 7
week1_tbl <- sim_ref |>
  dplyr::filter(time >= week1_start, time <= week1_end, !is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::summarise(Cmax_week1 = max(Cc), .groups = "drop")

ss_tbl <- sim_ref |>
  dplyr::filter(time >= ss_start, time <= ss_end, !is.na(Cc)) |>
  dplyr::arrange(id, time) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    Cmax_ss = max(Cc),
    tmax_ss = time[which.max(Cc)] - ss_start,
    AUC_ss  = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  )

geo_cv <- function(x) {
  if (any(!is.finite(log(x)))) NA_real_
  else sqrt(exp(sd(log(x))^2) - 1) * 100
}

summary_sim <- tibble::tibble(
  Parameter = c("Cmax week 1 (ug/mL)", "Cmax SS (ug/mL)", "tmax SS (d)", "AUCss (ug*d/mL)"),
  Mean_sim   = c(mean(week1_tbl$Cmax_week1), mean(ss_tbl$Cmax_ss),
                 mean(ss_tbl$tmax_ss), mean(ss_tbl$AUC_ss)),
  Median_sim = c(median(week1_tbl$Cmax_week1), median(ss_tbl$Cmax_ss),
                 median(ss_tbl$tmax_ss), median(ss_tbl$AUC_ss)),
  CV_sim     = c(geo_cv(week1_tbl$Cmax_week1), geo_cv(ss_tbl$Cmax_ss),
                 NA_real_,                     geo_cv(ss_tbl$AUC_ss))
)

published <- tibble::tibble(
  Parameter   = c("Cmax week 1 (ug/mL)", "Cmax SS (ug/mL)", "tmax SS (d)", "AUCss (ug*d/mL)"),
  Mean_pub    = c(9.95, 20.2, NA_real_, 225),
  Median_pub  = c(9.57, 16.1, 4,        160),
  CV_pub      = c(50.7, 76.8, NA_real_, 92.8)
)

comparison <- published |>
  dplyr::left_join(summary_sim, by = "Parameter") |>
  dplyr::mutate(
    Mean_pct_diff   = 100 * (Mean_sim   - Mean_pub)   / Mean_pub,
    Median_pct_diff = 100 * (Median_sim - Median_pub) / Median_pub
  )

knitr::kable(comparison, digits = 2,
  caption = "Simulated vs. published (Timmermann 2019 Table 4) secondary PK parameters for a reference 90 kg patient on 210 mg Q2W+1. Differences > 20% are investigated in the text below, not tuned away.")
```

# Assumptions and deviations

- **Virtual-cohort weight distribution.** Body weight is drawn from `N(87.8, 20)`
  kg truncated to [43, 186] to approximate Timmermann 2019 Table 2 (PK dataset
  median 87.8 kg, range 43-186). No subject-level observed data were released
  so the exact shape is unknown; sensitivity of the VPC to the weight
  distribution is expected to be modest because weight enters the model only
  as allometric exponents on CL, V1, and Vmax.
- **Race, sex, age, PASI, ADA.** The final model does not include race, sex,
  age, baseline PASI, or ADA status as covariates. These are not required to
  simulate from the model and are therefore omitted from the virtual-cohort
  construction.
- **Km artificial.** Km is fixed at 0.02 ug/mL (well below the LLOQ of 0.05
  ug/mL). As Timmermann 2019 notes in the Discussion, the non-linear term is
  effectively zero-order over the dosing range and Km should not be given a
  strict pharmacological interpretation.
- **Residual error units.** Additive residual error 3.00 ug/mL is taken from
  Table 3; the footnote explicitly states additive error is reported as
  standard deviation. This is relatively large (~15% of steady-state Cmax) and
  likely absorbs a meaningful fraction of the BLQ-related noise, consistent
  with the M3 method being used for censored data.
- **IV route.** IV doses in the original phase I trial (700 mg bolus) were
  modelled as bolus to central. In the nlmixr2 implementation the user must
  set `cmt = "central"` on any IV dose row (and `cmt = "depot"` on SC rows).
  No lag time or infusion handling is required.
- **AUC computation.** The comparison table uses linear-trapezoidal integration
  over the final simulated Q2W dosing interval. Timmermann 2019 Table 4 uses
  "standard non-compartmental analysis" on 1000 simulated 90 kg patients; the
  shape of the curve is identical so the method difference is negligible. The
  tmax for a reference patient is model-derived from the same SS interval.
- **No unit conversion needed.** Dose is mg and concentration is ug/mL
  (equivalent to mg/L), so `central / vc` yields ug/mL directly; this matches
  Timmermann 2019.
