---
title: "Vancomycin MIPD model switch (Tong 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin MIPD model switch (Tong 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

## What this paper is, and what is extracted from it

Tong 2026 (doi:10.1093/jacamr/dlag016) is not a model-development paper. It is a
large quasi-experimental interrupted-time-series study of what happens to
clinical outcomes when a model-informed precision dosing (MIPD) service
*switches* the population PK model it doses vancomycin with. Across nineteen US
hospital systems and 90 295 patients, two switches were evaluated:

| BMI stratum | Pre-intervention model | Post-intervention model |
|---|---|---|
| BMI < 40 kg/m2 | modified Goti | capped Thomson |
| BMI >= 40 kg/m2 | Carreno | Hughes |

The interrupted-time-series analysis itself is a statistical model and is out of
scope for nlmixr2lib. What *is* in scope is the supplement: its Code section
carries the complete NONMEM control stream for all four models exactly as
implemented in the InsightRX Nova software, and Tables S1 and S2 tabulate their
parameters. Those four streams are the source for the four model files extracted
here.

Two points make this paper the primary source for these implementations rather
than a redundant restatement of the four upstream publications:

* **Three of the four are modified.** The "modified Goti" recomputes
  Cockcroft-Gault creatinine clearance internally and carries a residual-error
  model that differs from Goti 2018 as published; the "capped Thomson" adds the
  150 mL/min creatinine-clearance cap that gives it its name; and the Carreno
  model has been *translated from a non-parametric fit into a parametric one*,
  which required residual-variability assumptions original to this
  implementation.
* **All four are fixed priors.** Tong 2026 estimated no population parameters.
  Every value is fixed and used as a prior for MAP Bayesian estimation of
  individual parameters, so `fixed()` is the faithful encoding throughout.

```{r load-models}
model_names <- c(
  goti    = "Tong_2026_vancomycin_goti",
  thomson = "Tong_2026_vancomycin_thomson",
  carreno = "Tong_2026_vancomycin_carreno",
  hughes  = "Tong_2026_vancomycin_hughes"
)
mods <- lapply(model_names, function(nm) readModelDb(nm)())
uis  <- lapply(model_names, function(nm) rxode2::rxode(readModelDb(nm)))
names(mods) <- names(uis) <- names(model_names)

tibble::tibble(
  Model      = names(model_names),
  `Model file` = unname(model_names),
  Stratum    = c("BMI < 40 (pre)", "BMI < 40 (post)",
                 "BMI >= 40 (pre)", "BMI >= 40 (post)"),
  Compartments = vapply(uis, function(u) paste(u$state, collapse = ", "), character(1))
) |>
  knitr::kable(caption = "The four MIPD models extracted from the Tong 2026 supplement.")
```

## Population

The cohorts below are the populations these models were *applied to* in Tong
2026, not the populations they were estimated from. Each model file records both:
`population` describes the application cohort (Table 1), and
`population$notes` records the upstream development cohort from Table S1 / S2.

```{r population}
pop_tab <- tibble::tribble(
  ~Characteristic,             ~`BMI < 40`,            ~`BMI >= 40`,
  "Patients",                  "87 586",               "2 709",
  "Treatment courses",         "94 991",               "2 964",
  "Samples",                   "192 013",              "6 572",
  "Sex (male/female)",         "39 961 / 55 030",      "1 272 / 1 692",
  "Serum creatinine (mg/dL)",  "0.90 (0.05-25.3)",     "0.95 (0.10-10.9)",
  "Height (cm)",               "170.2 (72.0-256.5)",   "167.6 (93.3-200.7)",
  "Weight (kg)",               "79.0 (20.4-173.8)",    "133.0 (43.8-318.0)",
  "Age (years)",               "65.8 (18.0-90+)",      "60.0 (18.2-90+)"
)
knitr::kable(
  pop_tab,
  caption = paste(
    "Tong 2026 Table 1 (median (min-max) or count). Patients were adults (>= 18 y)",
    "on intravenous vancomycin with at least two doses and one concentration;",
    "haemodialysis patients were excluded."
  )
)
```

Development cohorts, from Tables S1 and S2, are much smaller than the
application cohorts, and strikingly so for the Carreno model:

```{r dev-pop}
tibble::tibble(
  Model = names(model_names),
  `Development patients` = c(1812L, 398L, 12L, 100L),
  `Development samples`  = c(2765L, 1557L, 71L, 276L),
  `Upstream publication` = c("Goti 2018", "Thomson 2009", "Carreno 2017", "Hughes 2024")
) |>
  knitr::kable(caption = "Development populations (Tong 2026 Tables S1 and S2).")
```

## Source trace

Every `ini()` value in the four model files comes from the supplement's Code
section (the NONMEM control streams) and is cross-checked against the printed
equations and %CV rows in Tables S1 and S2. The table below records the mapping;
the model files carry the same trace inline, per parameter.

```{r source-trace}
trace <- tibble::tribble(
  ~Model,     ~Quantity,                    ~`Control-stream line`,                                  ~`Table S1 / S2 cross-check`,
  "goti",     "CL 4.5 L/h",                 "$THETA(1) 4.5 FIX",                                     "CL = 4.5 * (CRCLi/120)^0.8",
  "goti",     "Vc 58.4 L",                  "$THETA(2) 58.4 FIX, TVV = THETA(2)*(WT/70)",            "V = 58.4 * WT/70",
  "goti",     "Vp 38.4 L (no WT scaling)",  "$THETA(3) 38.4 FIX, TVV2 = THETA(3)",                   "V2 = 38.4",
  "goti",     "Q 6.5 L/h",                  "$THETA(4) 6.5 FIX",                                     "Q = 6.5",
  "goti",     "CRCL exponent 0.8",          "$THETA(5) 0.8 FIX",                                     "exponent in (CRCLi/120)^0.8",
  "goti",     "Dialysis factors 0.7 / 0.5", "$THETA(6) 0.7 FIX, $THETA(7) 0.5 FIX",                  "not tabulated (structural)",
  "goti",     "CrCL cap 150 mL/min",        "IF(CRCLi.GT.150.0) CRCLi = 150.0",                      "If CRCL > 150, then CRCLi = 150",
  "goti",     "IIV 39.8 / 81.6 / 57.1 %CV", "$OMEGA BLOCK(3) FIX 0.1584, 0.6659, 0.326",             "IIV on CL/V/V2 = 39.8 / 81.6 / 57.1",
  "goti",     "Error 1.42 + 19.7 %",        "$ERROR ADD = 1.42, PROP = 0.197",                       "Additive 1.42, proportional 19.7",
  "thomson",  "CL 2.99 L/h at CrCL 66",     "$THETA(1) 2.99 FIX",                                    "CL = 2.99 * (1 + 0.0154*(CRCLi-66))",
  "thomson",  "Vc 0.675 L/kg",              "$THETA(2) 0.675 FIX, TVV = THETA(2)*WT",                "V = 0.675 * WT",
  "thomson",  "CRCL slope 0.0154",          "$THETA(3) 0.0154 FIX",                                  "slope in (1 + 0.0154*(CRCLi-66))",
  "thomson",  "Q 2.28 L/h",                 "$THETA(4) 2.28 FIX",                                    "Q = 2.28",
  "thomson",  "Vp 0.732 L/kg",              "$THETA(5) 0.732 FIX, TVV2 = THETA(5)*WT",               "V2 = 0.732 * WT",
  "thomson",  "CrCL cap 150 mL/min",        "IF(EGFR.GT.CRCLCAP) CRCLcapped = CRCLCAP",              "If CRCL > 150, then CRCLi = 150",
  "thomson",  "IIV 27 / 15 %CV + cov",      "$OMEGA BLOCK(2) FIX 0.0729 / 0.01 0.0225",              "IIV on CL/V = 27 / 15",
  "thomson",  "IIV Q 49, V2 130 %CV",       "$OMEGA 0.2401 FIX; $OMEGA 1.69 FIX",                    "IIV on Q/V2 = 49 / 130",
  "thomson",  "Error 1.6 + 15 %",           "$ERROR ADD = 1.6, PROP = 0.15",                         "Additive 1.6, proportional 15",
  "carreno",  "CL intercept 0.18 L/h",      "$THETA(5) 0.18 FIX (TVSCLINTER)",                       "CL = 0.18 + 0.036 * CRCL",
  "carreno",  "CL slope 0.036",             "$THETA(2) 0.036 FIX (TVSCLSLOPE)",                      "CL = 0.18 + 0.036 * CRCL",
  "carreno",  "Vc 25.76 L",                 "$THETA(1) 25.76 FIX",                                   "V = 25.76",
  "carreno",  "k12 2.29 /h, k21 1.44 /h",   "$THETA(3) 2.29 FIX, $THETA(4) 1.44 FIX",                "Q = 2.29*25.76; V2 = Q/1.44",
  "carreno",  "IIV 45.3/55.6/105.6/120.1/23.9", "$OMEGA BLOCK(5) 0.20559 ... 0.057068",              "IIV rows of Table S2",
  "carreno",  "Error 0.1 + 10 %",           "$ERROR ADD = 0.1, PROP = 0.1",                          "Additive 0.1, proportional 10",
  "hughes",   "CL 5.0627 L/h at CrCL 100",  "$THETA(1) 5.0627",                                      "CL = 5.06 * (CRCL/100)^0.8509",
  "hughes",   "Vc 64.339 L at FFM 70",      "$THETA(2) 64.339",                                      "V = 64.339 * FFM/70",
  "hughes",   "Q 6.2402 L/h",               "$THETA(3) 6.2402",                                      "Q = 6.2402",
  "hughes",   "Vp 59.019 L at FFM 70",      "$THETA(4) 59.019",                                      "V2 = 59.019 * FFM/70",
  "hughes",   "FFM exponent 1 (fixed)",     "$THETA(7) 1 FIX  ; allo V1/V2",                         "linear FFM/70 in both volumes",
  "hughes",   "CRCL exponent 0.8509",       "$THETA(8) 0.8509",                                      "exponent in (CRCL/100)^0.8509",
  "hughes",   "IIV 21.8 / 18.0 / 79.7 %CV", "$OMEGA BLOCK(3) 0.0473154 ... 0.634656",                "IIV on CL/V/V2 = 21.8 / 18.0 / 79.7",
  "hughes",   "Error 0.001 + 13.72 %",      "$THETA(6) 0.001, $THETA(5) 0.1372",                     "Additive '-', proportional 13.3 (see Errata)"
)
knitr::kable(
  trace,
  caption = paste(
    "Source trace. 'Control-stream line' refers to the Tong 2026 Supplementary",
    "data, Code section; Tables S1 (BMI < 40) and S2 (BMI >= 40) provide the",
    "independent cross-check."
  )
)
```

A structural detail worth recording, because it is easy to get wrong: both
control streams that compute Cockcroft-Gault internally write the sex factor as
`0.85**(1-SEX)`, i.e. **SEX = 1 denotes male**. The Hughes stream confirms this
independently, because its `IF(SEX.EQ.0)` branch selects the *female*
Janmahasatian fat-free-mass coefficients. The model files therefore store the
canonical `SEXF` (1 = female) and write the factor as `0.85^SEXF`.

## Check 1: the models reproduce the printed Table S1 / S2 equations

The strongest available source-trace check is deterministic: for a reference
patient, solve each model and confirm the individual parameters equal the
equations printed in Tables S1 and S2. No simulation noise, no seed dependence.

```{r typical-values}
solve_typical <- function(mod, cov) {
  ev <- rxode2::et(amt = 1000, cmt = "central", dur = 1) |>
    rxode2::et(c(0.5, 2), cmt = "central")
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, params = cov,
                                returnType = "data.frame"))
}

# Reference patient for the BMI < 40 models: 70 kg, 40 y, male, SCr 0.9 mg/dL.
cg_ref  <- (140 - 40) * 70 * 0.85^0 / (72 * 0.9)
s_goti  <- solve_typical(mods$goti,
  data.frame(AGE = 40, WT = 70, SEXF = 0, CREAT = 0.9, RRT_HEMODIAL_STATUS = 0))
s_thom  <- solve_typical(mods$thomson, data.frame(WT = 70, CRCL = 100))

# Reference patient for the BMI >= 40 models: 133 kg, 167.6 cm, 60 y, female.
bmi_ref  <- 133 / (167.6 / 100)^2
ffm_ref  <- 9270 * 133 / (8780 + 244 * bmi_ref)
crcl_ref <- (140 - 60) * ffm_ref * 0.85 / (72 * 0.95)
s_carr   <- solve_typical(mods$carreno, data.frame(CRCL = 100))
s_hugh   <- solve_typical(mods$hughes,
  data.frame(WT = 133, HT = 167.6, AGE = 60, SEXF = 1, CREAT = 0.95))

tv <- tibble::tribble(
  ~Model, ~Parameter, ~Simulated, ~`Table S1 / S2 equation`, ~Expected,
  "goti",    "CL (L/h)", s_goti$cl[1], "4.5 * (CRCLi/120)^0.8",   4.5 * (cg_ref / 120)^0.8,
  "goti",    "Vc (L)",   s_goti$vc[1], "58.4 * WT/70",            58.4 * 70 / 70,
  "goti",    "Vp (L)",   s_goti$vp[1], "38.4",                    38.4,
  "goti",    "Q (L/h)",  s_goti$q[1],  "6.5",                     6.5,
  "thomson", "CL (L/h)", s_thom$cl[1], "2.99 * (1 + 0.0154*(CRCL-66))", 2.99 * (1 + 0.0154 * (100 - 66)),
  "thomson", "Vc (L)",   s_thom$vc[1], "0.675 * WT",              0.675 * 70,
  "thomson", "Vp (L)",   s_thom$vp[1], "0.732 * WT",              0.732 * 70,
  "thomson", "Q (L/h)",  s_thom$q[1],  "2.28",                    2.28,
  "carreno", "CL (L/h)", s_carr$cl[1], "0.18 + 0.036 * CRCL",     0.18 + 0.036 * 100,
  "carreno", "Vc (L)",   s_carr$vc[1], "25.76",                   25.76,
  "carreno", "Q (L/h)",  s_carr$q[1],  "k12 * V = 2.29 * 25.76",  2.29 * 25.76,
  "carreno", "Vp (L)",   s_carr$vp[1], "Q / k21 = Q / 1.44",      2.29 * 25.76 / 1.44,
  "hughes",  "CL (L/h)", s_hugh$cl[1], "5.0627 * (CRCL/100)^0.8509", 5.0627 * (crcl_ref / 100)^0.8509,
  "hughes",  "Vc (L)",   s_hugh$vc[1], "64.339 * FFM/70",         64.339 * ffm_ref / 70,
  "hughes",  "Vp (L)",   s_hugh$vp[1], "59.019 * FFM/70",         59.019 * ffm_ref / 70,
  "hughes",  "Q (L/h)",  s_hugh$q[1],  "6.2402",                  6.2402
) |>
  mutate(`% diff` = 100 * (Simulated - Expected) / Expected)

tv |>
  mutate(across(c(Simulated, Expected), ~ round(.x, 4)),
         `% diff` = signif(`% diff`, 3)) |>
  knitr::kable(caption = paste(
    "Check 1. Individual parameters solved from each model file against the",
    "equation printed in Tong 2026 Table S1 / S2. The reference patients are",
    sprintf("70 kg / 40 y / male / SCr 0.9 (Cockcroft-Gault %.1f mL/min) and", cg_ref),
    sprintf("133 kg / 167.6 cm / 60 y / female (FFM %.1f kg, CrCL %.1f mL/min).",
            ffm_ref, crcl_ref)
  ))

# The printed equations are exact algebra, so this must match to solver precision.
stopifnot(max(abs(tv$`% diff`)) < 1e-6)
```

That every one of the sixteen rows matches to solver precision is the check that
the transcription of the four control streams is correct: it exercises the
Cockcroft-Gault derivation, the Janmahasatian fat-free-mass branch, the sex
polarity conversion, the linear-versus-power renal terms, the per-kg versus
per-70-kg volume conventions, and the Carreno k12 / k21 to Q / Vp derivation.

## Check 1b: the closed-form and ODE solvers agree

`rxSolve()` defaults to `useLinCmt = TRUE`, which rewrites a recognisably linear
two-compartment system into a closed form. When such a model is written straight
from stored micro-constants, that rewrite can silently drop `peripheral1` and
solve a *one-compartment* model instead -- with total AUC unchanged, so an
exposure check will not catch it. The readout that moves is the terminal
half-life.

All four models are therefore written so the peripheral transfer is expressed as
`q` and `vp`, and the equivalence is asserted mechanically rather than assumed.
This is a regression gate: an earlier draft of the Carreno file stored `k12` and
`k21` directly and collapsed to 5.22 h against a true 13.82 h.

```{r lincmt-equivalence}
beta_half_life <- function(kel, k12, k21) {
  b <- kel + k12 + k21
  log(2) / ((b - sqrt(b^2 - 4 * kel * k21)) / 2)
}

# Sampling runs to 200 h and the slope is taken over 100-200 h. The capped
# Thomson model has a small Q against a large peripheral volume, so its
# distribution phase is still resolving at 72 h; by 100 h all four models are in
# the true terminal phase and the fitted slope is the beta root to 4 figures.
lincmt_ev <- as.data.frame(
  rxode2::et(amt = 1500, cmt = "central", dur = 1.5) |>
    rxode2::et(seq(0, 200, by = 0.5), cmt = "central")
)
lincmt_ev$id <- 1L

lincmt_cov <- list(
  goti    = list(WT = 79, AGE = 65, SEXF = 1, CREAT = 0.9,
                 RRT_HEMODIAL_STATUS = 0),
  thomson = list(WT = 79, CRCL = 90),
  carreno = list(CRCL = 90),
  hughes  = list(WT = 133, HT = 167.6, AGE = 60, SEXF = 1, CREAT = 0.95)
)

lincmt_chk <- dplyr::bind_rows(lapply(names(lincmt_cov), function(nm) {
  ev <- lincmt_ev
  for (cn in names(lincmt_cov[[nm]])) ev[[cn]] <- lincmt_cov[[nm]][[cn]]
  hl <- vapply(c(TRUE, FALSE), function(lin) {
    s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mods[[nm]]), ev,
                                       useLinCmt = lin, returnType = "data.frame"))
    s <- s[s$time >= 100 & s$Cc > 0, ]
    -log(2) / stats::coef(stats::lm(log(Cc) ~ time, data = s))[["time"]]
  }, numeric(1))
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mods[[nm]]), ev,
                                     returnType = "data.frame"))
  tibble::tibble(
    Model            = nm,
    `t1/2 closed form (h)` = hl[[1]],
    `t1/2 ODE (h)`         = hl[[2]],
    `t1/2 analytic (h)`    = beta_half_life(s$kel[1], s$k12[1], s$k21[1]),
    `t1/2 if collapsed (h)` = log(2) / s$kel[1]
  )
}))

lincmt_chk |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 3))) |>
  knitr::kable(caption = paste(
    "Check 1b. Terminal half-life recovered from the closed-form solver, from",
    "the ODE solver, and from the analytic beta root of the two-compartment",
    "system. The last column is what a collapsed one-compartment rewrite would",
    "return; it is roughly half the correct value in every case, which is what",
    "makes this gate sensitive."
  ))

# All three must agree: the two solvers exactly, and both against the analytic
# beta root to log-linear regression precision.
stopifnot(
  max(abs(lincmt_chk$`t1/2 closed form (h)` /
          lincmt_chk$`t1/2 ODE (h)` - 1)) < 1e-6,
  max(abs(lincmt_chk$`t1/2 ODE (h)` /
          lincmt_chk$`t1/2 analytic (h)` - 1)) < 1e-3,
  # The gate would be vacuous if the collapsed value were close to the correct
  # one; assert that it is not, so a future regression cannot slip through.
  all(lincmt_chk$`t1/2 if collapsed (h)` <
        0.75 * lincmt_chk$`t1/2 analytic (h)`)
)
```

## Check 2: steady-state exposure equals Dose / CL

For any linear model, the AUC over one dosing interval at steady state equals
Dose / CL exactly. This is independent of the number of compartments and of the
distribution parameters, so it is a clean test that the clearance
parameterisation and the dose-to-concentration unit chain are right in all four
models. Dosing runs long enough for the slowest subject in the cohort to reach
steady state (120 doses at q12h, about 60 days; the slowest subjects here have
terminal half-lives near 70 h).

```{r cohorts}
rxode2::rxSetSeed(20260828)
set.seed(20260828)

n_per_arm <- 100  # well within the 200-per-arm cap
tau       <- 12
n_dose    <- 120
ss_start  <- (n_dose - 1) * tau
ss_end    <- n_dose * tau

# BMI < 40 cohort, marginals from Tong 2026 Table 1.
coh_lo <- tibble::tibble(
  id    = seq_len(n_per_arm),
  WT    = pmin(pmax(exp(rnorm(n_per_arm, log(79), 0.26)), 20.4), 173.8),
  AGE   = pmin(pmax(rnorm(n_per_arm, 65.8, 15), 18), 90),
  SEXF  = rbinom(n_per_arm, 1, 0.579),
  CREAT = pmin(pmax(exp(rnorm(n_per_arm, log(0.90), 0.45)), 0.05), 25.3),
  RRT_HEMODIAL_STATUS = 0
) |>
  mutate(
    # The MIPD pipeline computes capped Cockcroft-Gault once and feeds it to
    # whichever model needs CRCL as a data item (the Thomson variant).
    CRCL = pmin((140 - AGE) * WT * 0.85^SEXF / (72 * CREAT), 150),
    HT   = 170.2,
    dose = pmin(pmax(round(15 * WT / 250) * 250, 500), 2000)
  )

# BMI >= 40 cohort, marginals from Tong 2026 Table 1.
coh_hi <- tibble::tibble(
  id    = seq_len(n_per_arm),
  WT    = pmin(pmax(exp(rnorm(n_per_arm, log(133), 0.24)), 43.8), 318.0),
  HT    = pmin(pmax(rnorm(n_per_arm, 167.6, 11), 93.3), 200.7),
  AGE   = pmin(pmax(rnorm(n_per_arm, 60.0, 15), 18), 90),
  SEXF  = rbinom(n_per_arm, 1, 0.571),
  CREAT = pmin(pmax(exp(rnorm(n_per_arm, log(0.95), 0.45)), 0.10), 10.9),
  RRT_HEMODIAL_STATUS = 0
) |>
  mutate(
    CRCL = pmin((140 - AGE) * WT * 0.85^SEXF / (72 * CREAT), 150),
    dose = pmin(pmax(round(15 * WT / 250) * 250, 500), 2500)
  )

tibble::tibble(
  Cohort = c("BMI < 40", "BMI >= 40"),
  `WT median` = c(median(coh_lo$WT), median(coh_hi$WT)),
  `AGE median` = c(median(coh_lo$AGE), median(coh_hi$AGE)),
  `CREAT median` = c(median(coh_lo$CREAT), median(coh_hi$CREAT)),
  `CRCL median` = c(median(coh_lo$CRCL), median(coh_hi$CRCL)),
  `Dose median (mg)` = c(median(coh_lo$dose), median(coh_hi$dose))
) |>
  mutate(across(where(is.numeric), ~ round(.x, 1))) |>
  knitr::kable(caption = paste(
    "Virtual cohorts, 100 subjects per arm. Marginal distributions are matched",
    "to the Tong 2026 Table 1 medians and truncated to its observed ranges;",
    "the joint distribution is not recoverable from the paper (see Assumptions)."
  ))
```

```{r build-events}
build_events <- function(coh) {
  ev <- lapply(seq_len(nrow(coh)), function(i) {
    e <- as.data.frame(
      rxode2::et(amt = coh$dose[[i]], cmt = "central", dur = 1,
                 ii = tau, addl = n_dose - 1) |>
        rxode2::et(seq(ss_start, ss_end, by = 0.1), cmt = "central")
    )
    e$id <- coh$id[[i]]
    e
  })
  ev <- dplyr::bind_rows(ev)
  # Covariates are attached to the materialised data frame, never to the rxEt
  # object (rxode2 silently drops assignments made to an rxEt).
  ev <- dplyr::left_join(
    ev,
    dplyr::select(coh, id, WT, HT, AGE, SEXF, CREAT, RRT_HEMODIAL_STATUS, CRCL),
    by = "id"
  )
  dplyr::arrange(ev, id, time, dplyr::desc(evid))
}

ev_lo <- build_events(coh_lo)
ev_hi <- build_events(coh_hi)

solve_apriori <- function(mod, ev) {
  as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, returnType = "data.frame"))
}

sims <- list(
  goti    = solve_apriori(mods$goti,    ev_lo),
  thomson = solve_apriori(mods$thomson, ev_lo),
  carreno = solve_apriori(mods$carreno, ev_hi),
  hughes  = solve_apriori(mods$hughes,  ev_hi)
)
stopifnot(vapply(sims, function(s) all(!is.na(s$Cc)), logical(1)))
stopifnot(vapply(sims, function(s) all(s$Cc >= 0), logical(1)))
```

These are *a priori* predictions in the sense the paper uses: population-level
predictions from the patient's covariates with no concentration data and hence no
individual random effects, which is what drives the initial dosing decision.
`zeroRe()` supplies exactly that, and makes the result deterministic rather than
seed-dependent.

```{r pknca}
# ID offsets keep subjects disjoint across the four model arms so PKNCA never
# pools two different subjects under one id.
id_offset <- c(goti = 0L, thomson = 1000L, carreno = 2000L, hughes = 3000L)

conc_all <- dplyr::bind_rows(lapply(names(sims), function(nm) {
  sims[[nm]] |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(id = id + id_offset[[nm]], time = time, Cc = Cc,
                     treatment = nm)
}))

dose_all <- dplyr::bind_rows(lapply(names(sims), function(nm) {
  coh <- if (nm %in% c("goti", "thomson")) coh_lo else coh_hi
  # PKNCA needs a dose record inside the steady-state interval.
  tibble::tibble(id = coh$id + id_offset[[nm]], time = ss_start,
                 amt = coh$dose, treatment = nm)
}))

conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc_all), Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_all), amt ~ time | treatment + id,
                             doseu = "mg")

intervals_ss <- data.frame(
  start = ss_start, end = ss_end,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  cav = TRUE, auclast = TRUE
)
# PKNCA's `ctrough` is deliberately not requested: it is defined at a dosing
# time, and this interval ends one dosing interval after the last dose, so it
# would be NA for every subject. `cmin` -- the minimum over the interval -- is
# the trough at steady state, and is checked against the simulated profile
# directly below.

nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                         intervals = intervals_ss))
nca_tbl <- as.data.frame(nca_ss$result)
```

```{r ss-identity}
auc_ss <- nca_tbl |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::transmute(id, treatment, auc_tau = PPORRES)

cl_ss <- dplyr::bind_rows(lapply(names(sims), function(nm) {
  sims[[nm]] |>
    dplyr::group_by(id) |>
    dplyr::summarise(cl = dplyr::first(cl), .groups = "drop") |>
    dplyr::mutate(id = id + id_offset[[nm]], treatment = nm)
}))

dose_lookup <- dose_all |> dplyr::select(id, treatment, amt)

ss_chk <- auc_ss |>
  dplyr::inner_join(cl_ss, by = c("id", "treatment")) |>
  dplyr::inner_join(dose_lookup, by = c("id", "treatment")) |>
  dplyr::mutate(pred = amt / cl,
                pct_diff = 100 * (auc_tau - pred) / pred)

ss_chk |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    n = dplyr::n(),
    `median AUCtau (mg*h/L)` = round(median(auc_tau), 1),
    `median Dose/CL (mg*h/L)` = round(median(pred), 1),
    `max |% diff|` = signif(max(abs(pct_diff)), 3),
    .groups = "drop"
  ) |>
  knitr::kable(caption = paste(
    "Check 2. PKNCA AUC over the steady-state dosing interval against the exact",
    "linear-system identity Dose / CL, for every subject in every arm."
  ))

# Both sides use the same drawn parameters, so the only difference is numerical
# (solver tolerance plus trapezoidal error on a 0.1 h grid). A tight bound is
# correct here and should stay tight.
stopifnot(max(abs(ss_chk$pct_diff)) < 0.25)
```

## Check 3: the two BMI < 40 clearance models cross over near 50 mL/min

The paper's central pharmacological finding is that the modified Goti model
estimates AUC *higher* than the capped Thomson model, and that this is what
shifted target attainment when sites switched. The mechanism is entirely in the
two clearance submodels: Goti uses a power function of creatinine clearance
normalised at 120 mL/min, Thomson a linear function centred at 66 mL/min. Because
one is concave and the other linear, they cross.

```{r crossover, fig.cap = "A priori clearance under the two BMI < 40 models as a function of creatinine clearance. The models cross near 49 mL/min; above it Goti predicts lower clearance and therefore higher AUC, which is the direction Tong 2026 reports."}
sweep <- tibble::tibble(CRCL = seq(20, 150, by = 1)) |>
  mutate(
    `modified Goti`  = 4.5 * (CRCL / 120)^0.8,
    `capped Thomson` = 2.99 * (1 + 0.0154 * (CRCL - 66))
  )

crossover <- sweep$CRCL[which.min(abs(sweep$`modified Goti` - sweep$`capped Thomson`))]

sweep |>
  tidyr::pivot_longer(-CRCL, names_to = "Model", values_to = "CL") |>
  ggplot(aes(CRCL, CL, colour = Model)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = crossover, linetype = "dashed", colour = "grey40") +
  annotate("text", x = crossover + 3, y = max(sweep$`capped Thomson`) * 0.35,
           label = paste0("crossover ~", crossover, " mL/min"),
           hjust = 0, size = 3.2, colour = "grey30") +
  labs(x = "Creatinine clearance (mL/min)", y = "A priori CL (L/h)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")

# Structural consequences of the two functional forms.
stopifnot(
  crossover > 40, crossover < 60,
  # Above the crossover Thomson clears faster, so Goti gives the higher AUC.
  with(sweep, all((`capped Thomson` > `modified Goti`)[CRCL > crossover + 5])),
  # Below it the ordering reverses.
  with(sweep, all((`capped Thomson` < `modified Goti`)[CRCL < crossover - 5]))
)
```

The population median creatinine clearance implied by Table 1 (age 65.8 y, 79 kg,
58% female, SCr 0.90 mg/dL) is about 77 mL/min, comfortably above the crossover.
So most patients sit in the region where Goti predicts the higher AUC, and a
low-renal-function minority sits in the region where the ordering reverses.

## Check 4: a priori AUC agreement, replicating Figure S4

Tong 2026 reports a Bland-Altman comparison of a priori AUC between the two
BMI < 40 models: the modified Goti model estimated AUCs +17.5 mg*h/L higher than
the capped Thomson model on average, one patient in 13 (7.7%) differed by more
than 20%, and the discordance was "driven primarily by the Goti model estimating
higher AUCs" (5.9% of patients).

```{r bland-altman, fig.cap = "Replicates Figure S4 of Tong 2026. Bland-Altman comparison of a priori steady-state AUC24 between the modified Goti and capped Thomson models over the virtual BMI < 40 cohort. Dashed line is the mean difference; dotted lines are the +/- 20% relative-difference boundaries."}
auc24 <- auc_ss |>
  dplyr::filter(treatment %in% c("goti", "thomson")) |>
  dplyr::mutate(id = id - id_offset[treatment],
                auc24 = 2 * auc_tau) |>
  dplyr::select(id, treatment, auc24) |>
  tidyr::pivot_wider(names_from = treatment, values_from = auc24)

ba <- auc24 |>
  dplyr::mutate(
    mean_auc = (goti + thomson) / 2,
    diff_auc = goti - thomson,
    pct_diff = 100 * diff_auc / mean_auc
  )

ggplot(ba, aes(mean_auc, diff_auc)) +
  geom_hline(yintercept = 0, colour = "grey70") +
  geom_hline(yintercept = mean(ba$diff_auc), linetype = "dashed",
             colour = "steelblue") +
  geom_line(aes(y = 0.2 * mean_auc), linetype = "dotted", colour = "grey40") +
  geom_line(aes(y = -0.2 * mean_auc), linetype = "dotted", colour = "grey40") +
  geom_point(alpha = 0.6, size = 1.6) +
  labs(x = "Mean of the two a priori AUC24 estimates (mg*h/L)",
       y = "Goti - Thomson (mg*h/L)") +
  theme_bw()

ba_stats <- tibble::tribble(
  ~Metric,                                    ~Published, ~Simulated,
  "Mean difference, Goti - Thomson (mg*h/L)", "+17.5",    sprintf("%+.1f", mean(ba$diff_auc)),
  "Patients differing by > 20%",              "7.7%",     sprintf("%.1f%%", 100 * mean(abs(ba$pct_diff) > 20)),
  "Goti higher by > 20%",                     "5.9%",     sprintf("%.1f%%", 100 * mean(ba$pct_diff > 20)),
  "Thomson higher by > 20%",                  "1.8%",     sprintf("%.1f%%", 100 * mean(ba$pct_diff < -20))
)
knitr::kable(ba_stats, caption = paste(
  "Check 4. Published a priori AUC agreement statistics (Tong 2026 Results and",
  "Figure S4) against the virtual cohort. The 1.8% 'Thomson higher' figure is",
  "the published 7.7% total less the 5.9% Goti-higher share."
))

# Direction and asymmetry are structural and must reproduce. The MAGNITUDE is
# not reproducible from Table 1's marginal summaries -- see Assumptions -- so it
# is bounded loosely rather than asserted at +17.5.
stopifnot(
  mean(ba$diff_auc) > 0,                                     # Goti higher on average
  median(ba$diff_auc) > 0,
  mean(ba$diff_auc) < 150,                                   # same order of magnitude
  mean(abs(ba$pct_diff) > 20) > 0.02,                        # a real discordant minority
  mean(abs(ba$pct_diff) > 20) < 0.35,
  mean(ba$pct_diff > 20) > mean(ba$pct_diff < -20)            # asymmetry in the published direction
)
```

The direction of the bias and the asymmetry of the discordance both reproduce.
The magnitude is larger than the published +17.5 mg*h/L, for reasons set out under
Assumptions: the joint covariate distribution is not recoverable from Table 1,
and the real cohort's doses were individualised by MIPD rather than set to a flat
15 mg/kg q12h.

## Check 5: steady-state NCA across the four models

```{r nca-summary}
nca_wide <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "cav", "auclast", "tmax")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_wide |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))) |>
  dplyr::rename(
    "Model"              = treatment,
    "AUCtau (mg*h/L)"    = auclast,
    "Cav (mg/L)"         = cav,
    "Cmax (mg/L)"        = cmax,
    "Cmin (mg/L)"        = cmin,
    "Tmax (h)"           = tmax
  ) |>
  knitr::kable(caption = paste(
    "Check 5. Median steady-state NCA parameters by model (PKNCA, over the",
    "final q12h dosing interval). Columns are renamed by name, never",
    "positionally."
  ))

# Cav must equal AUCtau / tau by construction; a mismatch would mean the
# interval passed to PKNCA is not the interval we think it is.
cav_chk <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("cav", "auclast")) |>
  tidyr::pivot_wider(id_cols = c(id, treatment), names_from = PPTESTCD,
                     values_from = PPORRES) |>
  dplyr::mutate(pct = 100 * (cav - auclast / tau) / (auclast / tau))
stopifnot(max(abs(cav_chk$pct)) < 1e-6)

# Vancomycin Tmax at steady state is the end of the 1 h infusion.
stopifnot(all(abs(nca_wide$tmax - 1) < 1e-8))

# PKNCA's Cmin must equal the minimum of the simulated profile over the same
# interval, for every subject in every arm. Both sides come from the same solve,
# so this is exact arithmetic and the bound stays tight. It confirms the
# interval handed to PKNCA is the interval we intended.
cmin_pknca <- nca_tbl |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::transmute(id, treatment, cmin = PPORRES)

cmin_sim <- dplyr::bind_rows(lapply(names(sims), function(nm) {
  sims[[nm]] |>
    dplyr::filter(time >= ss_start, time <= ss_end) |>
    dplyr::group_by(id) |>
    dplyr::summarise(c_min = min(Cc), c_end = dplyr::last(Cc), .groups = "drop") |>
    dplyr::mutate(id = id + id_offset[[nm]], treatment = nm)
}))

cmin_chk <- dplyr::inner_join(cmin_pknca, cmin_sim, by = c("id", "treatment"))
stopifnot(
  nrow(cmin_chk) == 4L * n_per_arm,
  max(abs(cmin_chk$cmin - cmin_chk$c_min)) < 1e-9
)

# Where in the interval that minimum falls is itself informative. After 120 q12h
# doses the cohort is at steady state to within a fraction of a percent, so the
# concentration at the start of the interval and at its end agree closely; the
# residual difference is the last sliver of accumulation in the slowest
# subjects, which is why the minimum sits at the interval start rather than at
# its end.
ss_gap <- max(abs(cmin_chk$c_end - cmin_chk$c_min) / cmin_chk$c_min)
stopifnot(ss_gap < 0.01)
```

A terminal half-life cannot be estimated from a single 12 h steady-state
interval, so it is taken from a separate single-dose simulation with sampling
carried out to 96 h.

```{r half-life}
single_dose_hl <- function(mod, cov, dose) {
  ev <- rxode2::et(amt = dose, cmt = "central", dur = 1) |>
    rxode2::et(seq(0, 96, by = 0.25), cmt = "central")
  s <- as.data.frame(rxode2::rxSolve(rxode2::zeroRe(mod), ev, params = cov,
                                     returnType = "data.frame"))
  s <- s[!is.na(s$Cc), ]
  s$id <- 1L
  s
}

hl_specs <- list(
  goti    = list(mod = mods$goti,    dose = 1000,
                 cov = data.frame(AGE = 65, WT = 79, SEXF = 1, CREAT = 0.9,
                                  RRT_HEMODIAL_STATUS = 0)),
  thomson = list(mod = mods$thomson, dose = 1000,
                 cov = data.frame(WT = 79, CRCL = 77)),
  carreno = list(mod = mods$carreno, dose = 1500,
                 cov = data.frame(CRCL = 77)),
  hughes  = list(mod = mods$hughes,  dose = 1500,
                 cov = data.frame(WT = 133, HT = 167.6, AGE = 60, SEXF = 1,
                                  CREAT = 0.95))
)

hl_conc <- dplyr::bind_rows(lapply(names(hl_specs), function(nm) {
  sp <- hl_specs[[nm]]
  single_dose_hl(sp$mod, sp$cov, sp$dose) |>
    dplyr::transmute(id = 1L, time, Cc, treatment = nm)
}))
hl_dose <- tibble::tibble(
  id = 1L, time = 0,
  amt = vapply(hl_specs, function(sp) sp$dose, numeric(1)),
  treatment = names(hl_specs)
)

hl_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(as.data.frame(hl_conc), Cc ~ time | treatment + id,
                   concu = "mg/L", timeu = "h"),
  PKNCA::PKNCAdose(as.data.frame(hl_dose), amt ~ time | treatment + id,
                   doseu = "mg"),
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         half.life = TRUE, aucinf.obs = TRUE, cl.obs = TRUE)
))

hl_tbl <- as.data.frame(hl_res$result) |>
  dplyr::filter(PPTESTCD %in% c("half.life", "aucinf.obs", "cl.obs", "cmax")) |>
  tidyr::pivot_wider(id_cols = treatment, names_from = PPTESTCD,
                     values_from = PPORRES)

hl_tbl |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ round(.x, 2))) |>
  dplyr::rename("Model" = treatment, "t1/2 (h)" = half.life,
                "AUCinf (mg*h/L)" = aucinf.obs, "CL/F (L/h)" = cl.obs,
                "Cmax (mg/L)" = cmax) |>
  knitr::kable(caption = paste(
    "Single-dose NCA at each model's reference patient, sampled to 96 h.",
    "CL/F recovered by NCA is compared against the model's own CL below."
  ))

# NCA-recovered clearance must return the model's structural CL. This is the
# same-parameters comparison as Check 2, so it stays tight.
cl_model <- vapply(names(hl_specs), function(nm) {
  sp <- hl_specs[[nm]]
  s <- single_dose_hl(sp$mod, sp$cov, sp$dose)
  s$cl[1]
}, numeric(1))
cl_cmp <- hl_tbl |>
  dplyr::mutate(cl_model = cl_model[treatment],
                pct = 100 * (cl.obs - cl_model) / cl_model)
stopifnot(max(abs(cl_cmp$pct)) < 1.0)
```

## Check 6: the paper's model-stability argument is visible in the IIV

Tong 2026 argues in the Discussion that the Carreno model, "developed as a
non-parametric two-compartment model on only 12 patients", gives MAP Bayesian
estimation so much flexibility that "small variations in creatinine clearance or
drug concentrations can result in very different PK parameter estimates", and
that this instability was a reason to prefer the simpler Hughes model even though
outcomes did not change significantly. That claim is qualitative in the paper,
but it is a direct consequence of the two variance structures, so it is checkable.

```{r iiv-spread, fig.cap = "Between-subject spread of steady-state AUC24 in the BMI >= 40 cohort under the two models, with inter-individual variability switched on. The Carreno model's much wider distribution is the flexibility Tong 2026 identifies as a source of unstable MAP estimates."}
solve_iiv <- function(mod, ev) {
  as.data.frame(rxode2::rxSolve(mod, ev, returnType = "data.frame"))
}

iiv <- dplyr::bind_rows(lapply(c("carreno", "hughes"), function(nm) {
  solve_iiv(mods[[nm]], ev_hi) |>
    dplyr::filter(time >= ss_start) |>
    dplyr::group_by(id) |>
    dplyr::summarise(
      auc24 = 2 * sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      .groups = "drop"
    ) |>
    dplyr::mutate(Model = nm)
}))

ggplot(iiv, aes(Model, auc24)) +
  geom_boxplot(width = 0.45, outlier.alpha = 0.4) +
  scale_y_log10() +
  labs(x = NULL, y = "Steady-state AUC24 with IIV (mg*h/L, log scale)") +
  theme_bw()

spread <- iiv |>
  dplyr::group_by(Model) |>
  dplyr::summarise(
    `median AUC24` = round(median(auc24), 1),
    `geometric CV %` = round(100 * sqrt(exp(var(log(auc24))) - 1), 1),
    `90% interval width / median` = round(
      diff(quantile(auc24, c(0.05, 0.95))) / median(auc24), 2),
    .groups = "drop"
  )
knitr::kable(spread, caption = paste(
  "Check 6. Between-subject dispersion of steady-state AUC24 under the two",
  "BMI >= 40 models. Reflects each model's published IIV structure:",
  "Carreno carries 55.6% CV on the renal clearance slope and 23.9% on the",
  "intercept; Hughes carries 21.8% CV on clearance."
))

# The Carreno model must be the more dispersed of the two -- that is the
# structural consequence of its variance terms, and the paper's argument.
carreno_cv <- spread$`geometric CV %`[spread$Model == "carreno"]
hughes_cv  <- spread$`geometric CV %`[spread$Model == "hughes"]
stopifnot(length(carreno_cv) == 1L, length(hughes_cv) == 1L,
          carreno_cv > hughes_cv)
```

## Assumptions and deviations

**Encoded as fixed priors throughout.** Tong 2026 estimated no population
parameters; the four models are priors for MAP Bayesian estimation of individual
parameters. Every `ini()` entry is therefore wrapped in `fixed()`. This is
directly supported by the NONMEM `FIX` flags on the modified-Goti and
modified-Thomson streams. It is an inference, not a flag, for the other two: the
Carreno `$OMEGA BLOCK(5)` and all of the Hughes `$THETA` and `$OMEGA` records
except `THETA(7)` carry no `FIX` flag in the supplement listing, most likely
because those records were pasted from the upstream estimation control streams.
Treating them as estimated would misrepresent the study, in which they were held
constant.

**Variance convention is the control stream's, not `log(CV^2 + 1)`.** All four
streams store omega values that satisfy `sqrt(omega^2) = %CV` from Tables S1 and
S2, i.e. the `omega^2 = CV^2` shorthand rather than the exact log-normal
`omega^2 = log(CV^2 + 1)`. The control-stream values are used verbatim, because
they are the implemented model. This is the main numerical difference from the
sibling `Goti_2018_vancomycin` extraction, which was made from the Goti 2018
publication and applied the exact log-normal conversion (0.14705 versus 0.1584
for clearance).

**Differences from the existing `Goti_2018_vancomycin` model file.** The two are
deliberately not merged, because they are different models from different
sources. Beyond the variance convention above:

* Peripheral volume. The InsightRX stream sets `TVV2 = THETA(3)` with no weight
  term, and Table S1 prints "V2 = 38.4". `Goti_2018_vancomycin` scales `vp` by
  `(WT/70)` on the strength of the Goti 2018 Methods statement that volume
  parameters were normalised to 70 kg.
* Residual error. This implementation uses 1.42 mg/L additive and 19.7%
  proportional; Goti 2018 as published reports 3.4 mg/L and 22.7%.
* Creatinine clearance is computed inside the model here, and the elderly
  low-creatinine truncation described by Goti 2018 is absent -- consistent with
  Tong 2021 (reference 19), which found that age-adjusting serum creatinine
  *decreased* predictive performance and was therefore not adopted.

**Hughes proportional error: 0.1372 versus a tabulated 13.3%.** The control
stream gives `$THETA(5) 0.1372` while Table S2 reports 13.3%. This is not a
rounding difference. The control-stream value is used as authoritative, on the
grounds that it is the machine-readable implemented model and is more precise;
the discrepancy is recorded here rather than silently reconciled. The
corresponding additive term, `$THETA(6) 0.001` mg/L, is effectively nil and is
consistent with Table S2 printing "-" for that row.

**The Carreno peripheral transfer is routed through Q and Vp.** The Carreno
`$DES` block drives the ODEs directly from the micro-constants `K12` and `K21`.
The model file keeps `lk12` and `lk21` as the stored fixed effects, matching
`$THETA(3)` and `$THETA(4)`, but computes `q = k12 * vc` and `vp = q / k21` and
drives the ODEs from those. This is an exact algebraic re-routing, not a change
of model, and it is necessary: written the literal way, `rxSolve()`'s default
`useLinCmt = TRUE` rewrite drops `peripheral1` and silently solves a
one-compartment model (terminal half-life 5.22 h against a true 13.82 h, Cmax
52.8 against 27.1 mg/L). Because total AUC is unaffected by that collapse, an
exposure check would not reveal it, so Check 1b asserts the half-life against the
analytic beta root instead.

**The AUC accumulator compartment is omitted.** Three of the four streams declare
a third state, `DADT(3) = A(1)/V`, which integrates concentration to give AUC for
MIPD reporting. It is a quadrature accumulator, not a biological compartment: it
has no analyte, no specimen, and no influence on the PK. It is omitted from all
four model files, and AUC is obtained by NCA instead. Relatedly, the Hughes
stream uses the closed-form `ADVAN3 TRANS4` and so has no third state at all,
yet its `$ERROR` block still assigns `AUC = A(3)`; that is a harmless leftover in
the source listing.

**`CL_HEMO` is carried as `fixed(0)`.** The modified-Goti and modified-Thomson
streams compute `CLTOT = CL + CL_HEMO`, where `CL_HEMO` is a per-patient additive
haemodialysis clearance supplied by the MIPD software. Tong 2026 excluded
haemodialysis patients entirely, so it is identically zero throughout the study.
Encoding it as `fixed(0)` keeps the structural term visible and auditable against
the stream without inventing a covariate column for a quantity the paper never
varies.

**Covariate polarity conversion.** Both internally-computing streams use
`SEX = 1` for male. The canonical column is `SEXF` (1 = female), so
`SEXF = 1 - SEX` and the Cockcroft-Gault factor `0.85**(1-SEX)` becomes
`0.85^SEXF`. The Hughes stream's `IF(SEX.EQ.0)` selection of the female
Janmahasatian coefficients independently confirms the polarity, so this is not an
assumption.

**Creatinine clearance units.** The InsightRX pipeline supplies `CRCL` in L/h and
both data-item streams convert to mL/min inline (`* 16.66667`). These model files
take `CRCL` directly in mL/min, which is the scale on which the 66 mL/min
centring value, the 150 mL/min cap and the 0.036 slope are all expressed. `CRCL`
here is raw Cockcroft-Gault creatinine clearance and is *not* BSA-normalised,
which the canonical register permits with per-model documentation.

**The virtual cohorts are marginal, not joint.** Table 1 reports only a median
and a min-max per covariate, so weight, age, sex and serum creatinine are
simulated independently (log-normal for the positive continuous covariates,
truncated to the published ranges) and height is fixed at the cohort median for
the BMI < 40 arm, which does not use it. Real weight, age and creatinine are
correlated, and the resulting creatinine-clearance distribution is therefore
wider than the true one. Dosing is a flat 15 mg/kg q12h rounded to 250 mg,
whereas the real cohort's doses were individualised by MIPD. This is why Check 4
reproduces the *direction* and the *asymmetry* of the published a priori AUC
disagreement but overstates its magnitude, and why that check bounds the mean
difference loosely instead of asserting the published +17.5 mg*h/L. Checks 1, 2,
3 and 5 do not depend on the cohort construction: Checks 1 and 3 are
deterministic algebra, and Checks 2 and 5 compare two quantities computed from
the same drawn parameters, so their tolerances are tight by design.

**Race and ethnicity are not reported** in Tong 2026 for either cohort and are
absent from `population$race_ethnicity`.

**No published NCA table to compare against.** Tong 2026 reports AUC target
attainment and interrupted-time-series effect sizes, not NCA summary statistics,
so `ncaComparisonTable()` is not used. Check 2 supplies the equivalent
quantitative gate by comparing PKNCA's steady-state AUC against the exact
Dose / CL identity for every subject, and Check 1 pins every structural parameter
against the printed Table S1 / S2 equations.

## Reference

Tong DMH, Brooks JT, Keizer RJ, Hughes JH. Vancomycin target attainment improved
following population pharmacokinetic model switch: a large-scale
quasi-experimental study of precision dosing. *JAC Antimicrob Resist.* 2026.
doi:10.1093/jacamr/dlag016

Upstream model publications: Goti V et al. *Ther Drug Monit* 2018;40:212-221
(doi:10.1097/FTD.0000000000000490); Tong DMH et al. *Ther Drug Monit*
2021;43:139-140 (doi:10.1097/FTD.0000000000000819); Thomson AH et al.
*J Antimicrob Chemother* 2009;63:1050-1057 (doi:10.1093/jac/dkp085);
Carreno JJ et al. *Antimicrob Agents Chemother* 2017;61:e02478-16
(doi:10.1128/aac.02478-16); Hughes MSA et al. *Ther Drug Monit* 2024;46:575-583
(doi:10.1097/FTD.0000000000001214).
