---
title: "Prostate cancer signaling logic-based ODE network (Traynard 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prostate cancer signaling logic-based ODE network (Traynard 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Traynard_2017_prostateSignaling_qsp"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Traynard_2017_prostateSignaling_qsp"))$description`
* Article: <https://doi.org/10.1002/psp4.12225> (PMCID: PMC5572374, open access)
* Supplementary materials (model files, fitted parameters, training data, fitting
  script): <https://github.com/saezlab/CPT_QSPtutorial> (GPLv3)
* Training data source: Lescarbeau RM, Kaplan DL (2014) BMC Cancer 14:325,
  <https://doi.org/10.1186/1471-2407-14-325>

Traynard 2017 is a CPT:PSP **tutorial** on logic modeling, but its worked example
is a real model fitted to real published data: a 20-node logic-based ODE network
of prostate cancer signaling, trained with `CellNOptR` / `CNORode2017` against 44
perturbation conditions of the Lescarbeau 2014 LNCaP phosphoproteome dataset. The
45 fitted parameters (20 node lifetimes `tau`, 25 edge strengths `k`) are published
in the supplementary repository, so the model is reproducible end to end. That is
what this file packages.

The paper's *other* model -- a continuous-time Boolean (MaBoSS) Markov process over
the full, uncompressed network including a `Survival` phenotype node -- is a
stochastic process over Boolean states rather than an ODE system and is not
packaged here. See **Assumptions and deviations** below.

# Population

The "subject" in this model is a **perturbation condition**, not an organism. The
system is the **LNCaP human prostate cancer cell line**, an androgen-sensitive line
used to study the transition to castration-resistant growth.

Lescarbeau 2014 perturbed LNCaP cells with combinations of five ligands (EGF,
IGF-1, IL-6, TNF-alpha and the androgen DHT) and five kinase inhibitors (PI3K,
MEK, IKK, mTOR and p38), giving 44 conditions, and assayed eight phosphosites
(AKT, RPS6, GSK3, ERK1/2, p38, JNK, HSP27, Stat3) at 30 min, 4 h and 24 h with an
antibody-based panel. Traynard 2017 used the 30 min and 4 h timepoints only,
reasoning that phosphorylation signaling reaches a semi-steady state within a few
hours while a 24 h window would require modeling transcriptional rewiring. The
docetaxel arm of Lescarbeau 2014 was excluded because it produced little
phosphoproteome variation and its target (beta-tubulin) lies outside the network.

Measurements were normalized per phosphosite as the log2 fold change versus the
unperturbed basal state and then linearly rescaled onto `[0, 1]` with **0.5 at
basal**. Every model state therefore starts at 0.5, and an unstimulated ligand
input node is also held at 0.5 rather than 0.

The same information is available programmatically:

```{r population}
readModelDb("Traynard_2017_prostateSignaling_qsp")()$population[
  c("species", "n_subjects", "n_conditions", "n_measured_nodes", "disease_state")
]
```

# Model structure

Each network node `j` carries a normalized activity on `[0, 1]` and relaxes toward
the continuous homologue of its Boolean rule:

$$\frac{dx_j}{dt} = \tau_j \left( B_j\left(f(x_{i_1}, k_{i_1 \to j}), \dots\right) - x_j \right)\left(1 - \mathrm{inh}_j\right)$$

* `tau_j` is the node lifetime parameter. `tau_j = 0` freezes the node at its
  initial (basal) value -- Traynard 2017 states this explicitly for IGF1_R, TNFR
  and IKK.
* `f` is the edge transfer function. The supplement's fitting script selects
  `transfer_function = 4`, which `CNORode2017` dispatches to
  `FG_transfer_function()`:
  $$f(x, n, k) = 1 - \frac{(1-x)^n}{(1-x)^n + k^n}\left(1 + k^n\right)$$
  a monotone sigmoid with `f(0) = 0`, `f(1) = 1`, and `f` identically zero when
  `k = 0` (the paper's "when k = 0 the dynamic of AKT is independent from PI3K").
  The Hill exponent `n` was held at the CNORode default of 3 (`opt_n = FALSE`).
* `B_j` is the multilinear (HillCube) homologue of node `j`'s Boolean rule.
  All four multi-input nodes in the trained model (AR, PI3K, RAS, JNK) are OR
  gates, whose homologue is `1 - prod(1 - f_i)`.
* `inh_j` is the inhibitor covariate for that node; a fully inhibited node has a
  zero derivative and so is pinned at the basal value 0.5.

```{r loadmodel}
mod <- rxode2::rxode2(readModelDb("Traynard_2017_prostateSignaling_qsp"))
rxode2::rxState(mod)
```

# Source trace

Values that are not in the article PDF come from the supplementary GitHub
repository the article's Supplementary Materials section points to. Every file
named below is in that repository.

```{r sourcetrace, echo = FALSE}
st <- tibble::tribble(
  ~Item, ~Source,
  "ODE right-hand side `tau * (B - x) * (1 - inh)`", "Article, 'Training the logic-based ODE model with CellNOpt' (AKT instance printed in text); general form CNORode2017 `src/rhsODE.c`",
  "Transfer function f(x, n, k)", "`transferFun = 4` in supplement `CellNOptR_optimisation.R`; dispatched to `FG_transfer_function` in CNORode2017 `src/sim_logic_ode.c`, defined in `src/FG_transfer_function.c`",
  "Hill exponent n = 3, not estimated", "Supplement `CellNOptR_optimisation.R`: `createLBodeContPars(..., default_n = 3, opt_n = FALSE)`",
  "OR gate for multi-input nodes", "Article Figure 2b caption: 'All other nodes in the model with more than one input edge are modeled with a simple OR gate'",
  "Network topology (20 states, 25 edges)", "Supplement `PriorKnowledgeNetwork.sif` after CellNOptR compression; independently confirmed by supplement `trainedmodel.bnd` and Article Figure 3a",
  "20 node lifetimes tau", "Supplement `logicODEparameters_nodes.txt`; independently reproduced as `$u_<node>` in supplement `trainedmodel.cfg`",
  "25 edge strengths k", "Supplement `logicODEparameters_edges.txt`",
  "k(Rac -> JNK) = 6.7e-05, k(AKT -> AR) = 8.4e-04", "Printed in the article main text, section 'Simulation of a logical model in different conditions'",
  "tau(IGF1_R) = tau(TNFR) = tau(IKKa) = 0", "Supplement `logicODEparameters_nodes.txt`; stated in the article text after Table 1 and shown as white boxes in Figure 3a",
  "Initial condition 0.5 for every state", "Article, normalization step 3: 'all initial conditions are set to 0.5 (which is the basal state...)'; CNORode2017 `src/simulateODE.c` sets the whole state array to 0.5",
  "Unstimulated stimulus node = 0.5", "Supplement `CellNOptR_optimisation.R`: `cnolist$valueStimuli[cnolist$valueStimuli==0]=0.5`",
  "Inhibitor pins its node at the initial value", "CNORode2017 `src/rhsODE.c`: the whole derivative is multiplied by `(1 - inhibitor_array[j])`",
  "Training data (44 conditions x 2 timepoints x 8 phosphosites)", "Supplement `perturbationData_LNCaP_MIDAS.csv`, shipped here as `inst/extdata/Traynard_2017_perturbationData_LNCaP_MIDAS.csv`",
  "Validation targets r = 0.66 and MSE = 0.017", "Article Figure 3c annotation"
)
knitr::kable(st, caption = "Source of every structural element and parameter block.")
```

# Experimental design and simulation

The MIDAS training file is shipped with the package. The design matrix is
recovered from its `TR:` columns, recoding an absent stimulus to the normalized
basal value 0.5 exactly as the supplement's fitting script does.

```{r design}
midas_path <- system.file(
  "extdata", "Traynard_2017_perturbationData_LNCaP_MIDAS.csv",
  package = "nlmixr2lib"
)
stopifnot(nzchar(midas_path))

midas <- read.csv(midas_path, check.names = FALSE)
names(midas) <- gsub(":", "_", sub("^TR:", "TR_", sub("^DV:", "DV_", sub("^DA:", "DA_", names(midas)))))

design <-
  midas |>
  dplyr::filter(DA_ALL == 0) |>
  dplyr::transmute(
    id = dplyr::row_number(),
    STIM_EGF_NORM = ifelse(TR_EGF == 1, 1, 0.5),
    STIM_IGF1_NORM = ifelse(TR_IGF_1 == 1, 1, 0.5),
    STIM_IL6_NORM = ifelse(TR_IL6 == 1, 1, 0.5),
    STIM_TNFA_NORM = ifelse(TR_TNFa == 1, 1, 0.5),
    STIM_DHT_NORM = ifelse(TR_DHT == 1, 1, 0.5),
    INH_PI3K = TR_PI3Ki,
    INH_MEK = TR_MEKi,
    INH_MTOR = TR_mTORi,
    INH_P38 = TR_p38i,
    INH_IKK = TR_IKKai
  )

stopifnot(nrow(design) == 44L)
knitr::kable(
  head(design, 8),
  caption = "First eight of the 44 perturbation conditions (0.5 = ligand absent)."
)
```

The model is deterministic -- Traynard 2017 reports no between-subject variability
and no residual error -- so a single solve per condition is the whole cohort. 44
conditions is well inside the 200-per-arm vignette cap.

```{r simulate}
measured <- c("AKT", "RPS6", "GSK3a", "ERK1_2", "p38", "JNK", "HSP27", "Stat3")

ev_fit <- merge(design["id"], as.data.frame(rxode2::et(c(0, 30, 240))), by = NULL)
ev_fit <- dplyr::arrange(ev_fit, id, time)

sim_fit <- rxode2::rxSolve(mod, events = ev_fit, params = design, returnType = "data.frame")

obs_long <-
  midas |>
  dplyr::filter(DA_ALL > 0) |>
  dplyr::mutate(id = rep(seq_len(44), times = 2L)) |>
  dplyr::select(id, time = DA_ALL, dplyr::all_of(paste0("DV_", measured))) |>
  tidyr::pivot_longer(-c(id, time), names_to = "node", values_to = "observed") |>
  dplyr::mutate(node = sub("^DV_", "", node))

pred_long <-
  sim_fit |>
  dplyr::select(id, time, dplyr::all_of(measured)) |>
  tidyr::pivot_longer(-c(id, time), names_to = "node", values_to = "predicted")

fit <- dplyr::inner_join(obs_long, pred_long, by = c("id", "time", "node"))
stopifnot(nrow(fit) == 44L * 2L * 8L, !anyNA(fit$observed), !anyNA(fit$predicted))
```

# Replicating Figure 3c: goodness of fit against the published metrics

Figure 3c of Traynard 2017 plots simulated against measured normalized values over
all 704 training observations and annotates the panel with two numbers:
**Pearson correlation = 0.66** and **MSE = RSS/N = 0.017**. Those are printed
values, so they are an exact answer key for the whole transcription -- the 45
parameters, the transfer function, the OR gates, the inhibitor semantics, the
initial conditions and the stimulus recoding all have to be right simultaneously
for them to come out.

```{r fig3c-metrics}
pearson_r <- cor(fit$observed, fit$predicted)
mse <- mean((fit$observed - fit$predicted)^2)
cod <- 1 - sum((fit$observed - fit$predicted)^2) / sum((fit$observed - mean(fit$observed))^2)

knitr::kable(
  tibble::tibble(
    Metric = c("Pearson correlation r", "MSE = RSS/N", "Coefficient of determination"),
    Published = c("0.66", "0.017", "(shown graphically only, Figure 3d)"),
    Reproduced = c(sprintf("%.4f", pearson_r), sprintf("%.5f", mse), sprintf("%.4f", cod))
  ),
  caption = "Figure 3c annotations versus this package's solve of the same model."
)

# These are printed point values compared against a fully deterministic solve --
# no random cohort is involved, so the only difference that can arise is ODE
# solver tolerance. A tight bound is the correct assertion here.
stopifnot(
  abs(round(pearson_r, 2) - 0.66) < 1e-8,
  abs(round(mse, 3) - 0.017) < 1e-8
)
```

```{r fig3c-plot, fig.cap = "Replicates Figure 3c of Traynard 2017: model simulation versus normalised measured values over all 44 conditions, 2 timepoints and 8 phosphosites."}
ggplot2::ggplot(fit, ggplot2::aes(observed, predicted)) +
  ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  ggplot2::geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey60") +
  ggplot2::geom_point(size = 0.8, alpha = 0.6) +
  ggplot2::geom_smooth(method = "lm", formula = y ~ x, colour = "grey40", se = TRUE) +
  ggplot2::coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  ggplot2::labs(
    x = "normalised measured values", y = "model simulation",
    subtitle = sprintf("Pearson correlation = %.2f; MSE = RSS/N = %.3f", pearson_r, mse)
  ) +
  ggplot2::theme_bw()
```

# Replicating Figure 3b: per-condition fitting-error heatmap

Figure 3b shows the mean squared error of each measured phosphoprotein in each
experimental condition on a white-to-dark-orange scale capped at 0.25, above a
green/red grid of the experimental design. The same quantity is reproduced here.

```{r fig3b, fig.height = 5.2, fig.cap = "Replicates Figure 3b (lower panel) of Traynard 2017: per-condition, per-phosphoprotein fitting error, with the experimental design below."}
err <-
  fit |>
  dplyr::group_by(id, node) |>
  dplyr::summarise(mse = mean((observed - predicted)^2), .groups = "drop") |>
  dplyr::mutate(node = factor(node, levels = rev(c("HSP27", "ERK1_2", "Stat3", "p38", "AKT", "RPS6", "GSK3a", "JNK"))))

design_long <-
  design |>
  dplyr::mutate(
    EGF = STIM_EGF_NORM == 1, IGF_1 = STIM_IGF1_NORM == 1, IL6 = STIM_IL6_NORM == 1,
    TNFa = STIM_TNFA_NORM == 1, DHT = STIM_DHT_NORM == 1,
    `PI3K i` = INH_PI3K == 1, `IKK i` = INH_IKK == 1, `MEK i` = INH_MEK == 1,
    `mTOR i` = INH_MTOR == 1, `p38 i` = INH_P38 == 1
  ) |>
  dplyr::select(id, EGF, IGF_1, IL6, TNFa, DHT, `PI3K i`, `IKK i`, `MEK i`, `mTOR i`, `p38 i`) |>
  tidyr::pivot_longer(-id, names_to = "row", values_to = "on") |>
  dplyr::mutate(
    kind = ifelse(row %in% c("EGF", "IGF_1", "IL6", "TNFa", "DHT"), "Stimulus", "Inhibitor"),
    row = factor(row, levels = rev(c("EGF", "IGF_1", "IL6", "TNFa", "DHT", "PI3K i", "IKK i", "MEK i", "mTOR i", "p38 i")))
  )

p_err <-
  ggplot2::ggplot(err, ggplot2::aes(factor(id), node, fill = pmin(mse, 0.25))) +
  ggplot2::geom_tile(colour = "grey85", linewidth = 0.1) +
  ggplot2::scale_fill_gradientn(
    colours = c("#FFFFF0", "#FEE08B", "#F46D43", "#8C2D04"), limits = c(0, 0.25),
    name = "Mean\nSquared\nError (MSE)"
  ) +
  ggplot2::labs(x = NULL, y = "Measured phosphoproteins") +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_blank(), panel.grid = ggplot2::element_blank())

p_des <-
  ggplot2::ggplot(design_long, ggplot2::aes(factor(id), row)) +
  ggplot2::geom_tile(ggplot2::aes(fill = ifelse(on, kind, NA_character_)), colour = "grey85", linewidth = 0.1) +
  ggplot2::scale_fill_manual(
    values = c(Stimulus = "#7CB342", Inhibitor = "#E53935"), na.value = "white",
    name = NULL, na.translate = FALSE
  ) +
  ggplot2::labs(x = "Experimental condition", y = "Experimental condition") +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.x = ggplot2::element_blank(), panel.grid = ggplot2::element_blank())

print(p_err)
print(p_des)
```

As in the published panel, the fit is good for most conditions (the bulk of the
heatmap is near-white) and the worst errors concentrate in a handful of
inhibitor-containing conditions, notably for p38 and Stat3.

```{r fig3b-worst}
condition_label <-
  design |>
  dplyr::transmute(
    id = id,
    condition = trimws(paste0(
      ifelse(STIM_EGF_NORM == 1, "EGF ", ""), ifelse(STIM_IGF1_NORM == 1, "IGF1 ", ""),
      ifelse(STIM_IL6_NORM == 1, "IL6 ", ""), ifelse(STIM_TNFA_NORM == 1, "TNFa ", ""),
      ifelse(STIM_DHT_NORM == 1, "DHT ", ""), ifelse(INH_PI3K == 1, "+PI3Ki ", ""),
      ifelse(INH_MEK == 1, "+MEKi ", ""), ifelse(INH_MTOR == 1, "+mTORi ", ""),
      ifelse(INH_P38 == 1, "+p38i ", ""), ifelse(INH_IKK == 1, "+IKKi ", "")
    ))
  ) |>
  dplyr::mutate(condition = ifelse(nzchar(condition), condition, "control"))

err |>
  dplyr::slice_max(mse, n = 6) |>
  dplyr::left_join(condition_label, by = "id") |>
  dplyr::transmute(Condition = condition, Phosphoprotein = as.character(node), MSE = round(mse, 3)) |>
  knitr::kable(caption = "The six worst-fitting condition / phosphoprotein cells.")
```

# Replicating Figure 3b (upper panels): example time courses

Figure 3b also shows four example fits of simulated activity against the measured
points. The same comparison is drawn here for four representative conditions.

```{r fig3b-timecourse, fig.cap = "Model simulation (line) against the measured normalised activity (points) for four representative condition / phosphoprotein pairs, in the style of the Figure 3b upper panels."}
ev_dense <- merge(design["id"], as.data.frame(rxode2::et(seq(0, 240, by = 2))), by = NULL)
ev_dense <- dplyr::arrange(ev_dense, id, time)
sim_dense <- rxode2::rxSolve(mod, events = ev_dense, params = design, returnType = "data.frame")

examples <- tibble::tribble(
  ~id, ~node,
  which(design$STIM_EGF_NORM == 1 & rowSums(design[, c("INH_PI3K", "INH_MEK", "INH_MTOR", "INH_P38", "INH_IKK")]) == 0)[1], "ERK1_2",
  which(rowSums(design[, 2:6] == 1) == 0 & rowSums(design[, 7:11]) == 0)[1], "p38",
  which(design$STIM_DHT_NORM == 1 & rowSums(design[, c("INH_PI3K", "INH_MEK", "INH_MTOR", "INH_P38", "INH_IKK")]) == 0)[1], "AKT",
  which(design$INH_PI3K == 1)[1], "RPS6"
) |>
  dplyr::mutate(panel = sprintf("condition %d: %s", id, node))

ex_pred <-
  examples |>
  dplyr::left_join(
    sim_dense |> dplyr::select(id, time, dplyr::all_of(measured)) |>
      tidyr::pivot_longer(-c(id, time), names_to = "node", values_to = "predicted"),
    by = c("id", "node")
  )
ex_obs <- dplyr::inner_join(examples, fit, by = c("id", "node"))

ggplot2::ggplot(ex_pred, ggplot2::aes(time, predicted)) +
  ggplot2::geom_line(colour = "#1f4e99", linetype = "dashed") +
  ggplot2::geom_point(data = ex_obs, ggplot2::aes(time, observed), shape = 2, size = 2.4) +
  ggplot2::geom_point(data = ex_obs, ggplot2::aes(time, predicted), colour = "#1f4e99", size = 2.4) +
  ggplot2::facet_wrap(~panel) +
  ggplot2::coord_cartesian(ylim = c(0, 1)) +
  ggplot2::labs(x = "time (min)", y = "activity") +
  ggplot2::theme_bw()
```

# Validation checks

This model has no drug, no dose and no plasma concentration, so a PKNCA
non-compartmental analysis is not the right instrument. The checks below are the
mechanistic-model equivalents: exact structural identities, an independent
closed-form steady state, and the paper's own qualitative claims.

## 1. Nodes with `tau = 0` are frozen at basal, in every condition

`tau_IGF1_R`, `tau_TNFR` and `tau_IKKa` are all exactly zero in the fitted
parameter set, so the whole right-hand side of those three ODEs vanishes and the
states cannot leave their initial value.

```{r gate-tau0}
frozen <- c("IGF1_R", "TNFR", "IKKa")
dev_frozen <- max(abs(as.matrix(sim_dense[, frozen]) - 0.5))
dev_frozen
stopifnot(dev_frozen == 0)
```

## 2. The system does not respond to IGF-1 or TNF-alpha

Traynard 2017 draws a biological conclusion from those three null lifetimes:
*"IGF1-R, TNFR, and IKK are associated with null tau parameters, which means that
they become independent from their regulator activities. The system therefore does
not depend on the inputs IGF-1 and TNF alpha."* If the packaged model is right,
turning those two ligands on must change nothing, anywhere, at any time -- exactly,
not approximately.

```{r gate-inputs}
states <- rxode2::rxState(mod)
control <- dplyr::mutate(
  design[1, ],
  STIM_EGF_NORM = 0.5, STIM_IGF1_NORM = 0.5, STIM_IL6_NORM = 0.5,
  STIM_TNFA_NORM = 0.5, STIM_DHT_NORM = 0.5,
  INH_PI3K = 0, INH_MEK = 0, INH_MTOR = 0, INH_P38 = 0, INH_IKK = 0
)
tgrid <- as.data.frame(rxode2::et(seq(0, 240, by = 5)))
solve_one <- function(p) {
  p$id <- 1L
  rxode2::rxSolve(mod, events = cbind(id = 1L, tgrid), params = p, returnType = "data.frame")
}

base_traj <- as.matrix(solve_one(control)[, states])
dev_igf1 <- max(abs(as.matrix(solve_one(dplyr::mutate(control, STIM_IGF1_NORM = 1))[, states]) - base_traj))
dev_tnfa <- max(abs(as.matrix(solve_one(dplyr::mutate(control, STIM_TNFA_NORM = 1))[, states]) - base_traj))

c(IGF1 = dev_igf1, TNFa = dev_tnfa)
stopifnot(dev_igf1 == 0, dev_tnfa == 0)
```

## 3. A fully inhibited node is pinned at the basal value

```{r gate-inhibitor}
inh_map <- c(PI3K = "INH_PI3K", MEK = "INH_MEK", mTOR = "INH_MTOR", p38 = "INH_P38", IKKa = "INH_IKK")
stimulated <- dplyr::mutate(control, STIM_EGF_NORM = 1, STIM_DHT_NORM = 1, STIM_IL6_NORM = 1)

dev_inh <- vapply(names(inh_map), function(nm) {
  p <- stimulated
  p[[inh_map[[nm]]]] <- 1
  max(abs(solve_one(p)[[nm]] - 0.5))
}, numeric(1))

dev_inh
stopifnot(all(dev_inh == 0))
```

## 4. The ODE steady state matches an independent closed-form fixed point

The trained network is a directed acyclic graph apart from a single
`AR -> PI3K -> AKT -> AR` feedback loop. Its steady state can therefore be computed
without any ODE solver: substitute forward through the cascade, and resolve the
one loop by fixed-point iteration. The transfer function used below is written out
again from the `CNORode2017` C source rather than reused from the model object, so
this is an independent check of the packaged ODEs, not a restatement of them.

```{r gate-closedform}
fg <- function(x, k, n = 3) 1 - (1 - x)^n / ((1 - x)^n + k^n) * (1 + k^n)

closed_form_ss <- function(EGF, IGF1, IL6, TNFa, DHT) {
  EGFR <- fg(EGF, 0.550337329553758)
  IGF1_R <- 0.5 # tau = 0
  IL6R <- fg(IL6, 0.78134194522735)
  TNFR <- 0.5 # tau = 0
  # The one feedback loop AR -> PI3K -> AKT -> AR, resolved by fixed-point
  # iteration; the map is a contraction here so it converges in a few steps.
  AR <- AKT <- PI3K <- 0.5
  repeat {
    prev <- c(AR, PI3K, AKT)
    AR <- 1 - (1 - fg(AKT, 0.000835340126624366)) * (1 - fg(DHT, 0.590287658825809))
    PI3K <- 1 - (1 - fg(AR, 1)) * (1 - fg(EGFR, 1)) * (1 - fg(IGF1_R, 1))
    AKT <- fg(PI3K, 0.352193224711085)
    if (max(abs(c(AR, PI3K, AKT) - prev)) < 1e-15) break
  }
  mTOR <- fg(AKT, 1)
  Jak <- fg(IL6R, 1)
  RAS <- 1 - (1 - fg(Jak, 0.674347577514922)) * (1 - fg(EGFR, 0.25869659979145))
  Rac <- fg(RAS, 0.284328823821596)
  p38 <- fg(Rac, 1)
  c(
    EGFR = EGFR, IGF1_R = IGF1_R, IL6R = IL6R, TNFR = TNFR, AR = AR, PI3K = PI3K,
    AKT = AKT, mTOR = mTOR, RPS6 = fg(mTOR, 0.303674761072917),
    GSK3a = fg(AKT, 0.558225196843653), Jak = Jak, Stat3 = fg(Jak, 0.28320076784032),
    RAS = RAS, MEK = fg(RAS, 0.490290337695327),
    ERK1_2 = fg(fg(RAS, 0.490290337695327), 0.462172528684214), Rac = Rac, p38 = p38,
    HSP27 = fg(p38, 0.993983089528085),
    JNK = 1 - (1 - fg(TNFR, 0.361795499785572)) * (1 - fg(Rac, 0.0000673323536405412)),
    IKKa = 0.5
  )
}

scenarios <- tibble::tribble(
  ~label, ~EGF, ~IGF1, ~IL6, ~TNFa, ~DHT,
  "control", 0.5, 0.5, 0.5, 0.5, 0.5,
  "EGF", 1.0, 0.5, 0.5, 0.5, 0.5,
  "DHT", 0.5, 0.5, 0.5, 0.5, 1.0,
  "EGF + IL6 + DHT", 1.0, 0.5, 1.0, 0.5, 1.0
)

ss_long_grid <- as.data.frame(rxode2::et(c(0, 2e6)))
ss_dev <- vapply(seq_len(nrow(scenarios)), function(i) {
  p <- dplyr::mutate(
    control,
    STIM_EGF_NORM = scenarios$EGF[i], STIM_IGF1_NORM = scenarios$IGF1[i],
    STIM_IL6_NORM = scenarios$IL6[i], STIM_TNFA_NORM = scenarios$TNFa[i],
    STIM_DHT_NORM = scenarios$DHT[i]
  )
  p$id <- 1L
  sv <- rxode2::rxSolve(mod, events = cbind(id = 1L, ss_long_grid), params = p, returnType = "data.frame")
  ana <- closed_form_ss(scenarios$EGF[i], scenarios$IGF1[i], scenarios$IL6[i], scenarios$TNFa[i], scenarios$DHT[i])
  max(abs(unlist(sv[nrow(sv), states]) - ana[states]))
}, numeric(1))

knitr::kable(
  dplyr::mutate(scenarios["label"], `max abs deviation` = signif(ss_dev, 3)),
  caption = "ODE solution at t = 2e6 min versus the closed-form fixed point."
)

# Both sides use the same parameter values, so the only difference is solver
# tolerance -- a tight bound is correct here.
stopifnot(all(ss_dev < 1e-10))
```

## 5. The EGF arm dominates, and the IL-6 arm barely moves

Traynard 2017: *"IL6-R is associated with a very small value for tau(IL6-R),
suggesting that the dynamics of the system will predominantly depend on pathways
activated by EGF."*

```{r gate-egf-dominance}
arm_shift <- function(stim) {
  p <- control
  p[[stim]] <- 1
  max(abs(as.matrix(solve_one(p)[, states]) - base_traj))
}
shifts <- vapply(
  c(EGF = "STIM_EGF_NORM", IL6 = "STIM_IL6_NORM", DHT = "STIM_DHT_NORM"),
  arm_shift, numeric(1)
)
round(shifts, 4)

# Deterministic solves, so these are exact numbers, not a sampled cohort. The
# EGF arm moves the network more than twenty times as far as the IL-6 arm, and
# the IL-6 arm barely leaves basal at all.
stopifnot(
  shifts[["EGF"]] > 10 * shifts[["IL6"]],
  shifts[["DHT"]] > 10 * shifts[["IL6"]],
  shifts[["IL6"]] < 0.05
)
```

The EGF arm shifts the network by up to `r sprintf("%.2f", shifts[["EGF"]])`
activity units and the androgen arm by `r sprintf("%.2f", shifts[["DHT"]])`,
against `r sprintf("%.3f", shifts[["IL6"]])` for IL-6 -- the IL-6 receptor's
lifetime parameter is small enough (3.1e-04 per min) that its arm has hardly
begun to move by 240 min.

## 6. Androgen and PI3K inhibition at the node level

Table 1 of Traynard 2017 compares experimental survival proportions against
*MaBoSS* survival probabilities. Survival is not a state of the logic-based ODE
model (see Assumptions below), but the node-level behaviour that drives the
paper's conclusion -- DHT raises PI3K/AKT signaling, and PI3K inhibition abrogates
it -- is reproducible here.

```{r table1-nodes}
arm <- function(lab, dht, pi3ki) {
  p <- dplyr::mutate(control, STIM_DHT_NORM = dht, INH_PI3K = pi3ki)
  p$id <- 1L
  sv <- rxode2::rxSolve(mod, events = cbind(id = 1L, tgrid), params = p, returnType = "data.frame")
  tibble::tibble(
    Condition = lab,
    AR = sv$AR[nrow(sv)], PI3K = sv$PI3K[nrow(sv)], AKT = sv$AKT[nrow(sv)],
    mTOR = sv$mTOR[nrow(sv)], RPS6 = sv$RPS6[nrow(sv)]
  )
}
t1 <- dplyr::bind_rows(
  arm("control (DHT absent)", 0.5, 0),
  arm("DHT = 1", 1.0, 0),
  arm("PI3K inhibited", 0.5, 1),
  arm("DHT = 1 + PI3K inhibited", 1.0, 1)
)
knitr::kable(dplyr::mutate(t1, dplyr::across(-Condition, ~ round(.x, 3))),
  caption = "Node activity at 240 min in the Table 1 androgen / PI3K-inhibitor conditions."
)

# DHT must raise AR and the PI3K axis; PI3K inhibition must pin the axis at basal
# and block the DHT effect downstream of PI3K.
stopifnot(
  t1$AR[2] > t1$AR[1],
  t1$PI3K[2] > t1$PI3K[1],
  t1$AKT[2] > t1$AKT[1],
  t1$PI3K[3] == 0.5, t1$PI3K[4] == 0.5,
  t1$AKT[4] == t1$AKT[3]
)
```

DHT raises AR and propagates through PI3K, AKT, mTOR and RPS6; inhibiting PI3K
pins the axis at basal and the DHT signal cannot reach AKT at all, which is the
node-level counterpart of the survival result the paper reports (survival
0.87 with DHT, 0.33 with PI3K inhibition, 0.63 with both).

# Assumptions and deviations

* **The MaBoSS Boolean model is not packaged.** Traynard 2017 builds two models.
  The one packaged here is the logic-based ODE model trained with CellNOptR. The
  second is a continuous-time Markov process over Boolean states, simulated with
  MaBoSS, over the full uncompressed network including `Survival`, `Cell_cycle`,
  `MYC`, `Caspase8`, `Caspase9`, `p53`, `NFkB` and `beta_catenin`. That model is
  not an ODE system -- it is a stochastic jump process whose master equation runs
  over `2^n` Boolean states -- so it cannot be expressed in rxode2, and the eight
  extra nodes it carries were removed from the ODE model by CellNOptR's
  compression step and have no fitted parameters. The survival probabilities in
  Table 1, the node-inhibition scan in Figure 5 and the epistasis network in
  Figure 6 are all MaBoSS outputs and are therefore out of scope for this file.
  The MaBoSS model definition is in the supplement as `trainedmodel.bnd` /
  `trainedmodel.cfg` for anyone who wants it.
* **The trained model has 25 edges; the published prior knowledge network file has
  26 candidate edges among the surviving nodes.** `PriorKnowledgeNetwork.sif`
  contains a final line `TNFa -> PI3K` that is absent from the fitted parameter
  file `logicODEparameters_edges.txt`, absent from the trained MaBoSS model
  (`trainedmodel.bnd` gives PI3K the rule `AR | EGFR | IGF1_R`), and not drawn in
  Figure 3a. The `.sif` file has classic-Mac CR line endings and no terminating
  newline on that last line, which is the likely reason it was not read at
  training time. Two independent published artifacts agree on the 25-edge trained
  model, so that is what is packaged; the discrepancy is recorded here rather than
  silently resolved.
* **Parameters are on the natural scale, not log-transformed.** `tau` and `k` are
  bounded on `[0, 1]` by the fitting script and several estimates sit exactly on a
  bound (`tau = 0` for three nodes, `k = 1` for six edges), so a log
  parameterization is both undefined and structurally wrong here.
* **No IIV and no residual error.** Traynard 2017 reports a single deterministic
  fit obtained by minimizing a residual sum of squares plus a steady-state
  penalty. No parameter standard errors, confidence intervals, between-subject
  variances or residual-error terms are reported anywhere in the paper or its
  supplement; uncertainty is reported only as a bootstrap performance
  distribution in Figure 3d. Nothing was invented to fill those gaps.
* **Ligand and inhibitor concentrations are not recoverable from this paper.** The
  applied concentrations live in Lescarbeau 2014, the upstream data paper. The
  `STIM_*` covariates are therefore expressed on the model's own normalized
  activity scale (0.5 = absent, 1 = applied) and the `INH_*` covariates are binary
  flags, which is exactly the resolution the trained model uses.
* **Inhibition pins a node at basal rather than driving it to zero.** This is the
  `CNORode2017` convention (`src/rhsODE.c` multiplies the whole derivative by
  `1 - inhibitor`), and it is coherent with the normalization in which 0.5 is the
  unperturbed basal state: the inhibitor prevents the node from responding, it
  does not abolish baseline activity. The covariates accept values on `[0, 1]` so
  a partial inhibition can be simulated, although the Lescarbeau 2014 design used
  0/1 only.
* **The GSK3 antibody measures an inhibitory site.** Traynard 2017 inverted the
  sign of the edges to and from GSK3 to account for this before fitting; the
  `GSK3a` state and its fitted `k_AKT_GSK3a` therefore already carry that
  inversion, and no further sign handling is applied here.
* **The Hill exponent is a single shared parameter.** CNORode carries one `n` per
  edge, but the fitting script fixed every one of them at the default of 3
  (`opt_n = FALSE`), so a single `nhill` parameter is an exact representation of
  the published model.
* **Source is the supplementary repository, not the article PDF.** The article
  prints the structural ODE form and two of the 25 edge strengths
  (`k(Rac -> JNK)` and `k(AKT -> AR)`, both of which match the supplement
  exactly); all remaining parameter values come from the GitHub repository the
  article's Supplementary Materials section designates. The relevant files are
  archived alongside the article PDF in this project's literature directory.
* **Transfer-function and right-hand-side semantics were read from source.** The
  article does not write out `FG_transfer_function` or the inhibitor factor; those
  were read from the C sources of `CNORode2017`, the package the supplement's own
  fitting script installs and calls. The Figure 3c answer key (Pearson
  correlation 0.66 and MSE 0.017) confirms the reading.

# Session info

```{r sessioninfo}
sessionInfo()
```
