---
title: "Omecamtiv mecarbil (Trivedi 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Omecamtiv mecarbil (Trivedi 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Trivedi_2021_omecamtivMecarbil"))$reference`
* Description: `r rxode2::rxode(readModelDb("Trivedi_2021_omecamtivMecarbil"))$description`
* Article: <https://doi.org/10.1111/cts.13118>
* Supplement (Supplementary Materials PDF and the Simcyp output workbook
  deposited as Supplementary Table S1): <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8604240/>

# What this model is, and what it is not

Trivedi 2021 measured the effect of a single 50 mg oral dose of omecamtiv
mecarbil (OM), an in-vitro BCRP inhibitor, on a single 10 mg oral dose of
rosuvastatin in 14 healthy adults. They then built a Simcyp (V17.1) PBPK model
to extrapolate the interaction to 50 mg twice-daily dosing.

The PBPK analysis has two layers, and only one of them can be encoded outside
Simcyp.

* **Omecamtiv mecarbil (the perpetrator) is a two-compartment model.** The
  authors fitted an ordinary two-compartment model to data after a 35 mg
  intravenous dose with SAAM II (`k12`, `k21`, `k10`, `Vc`). They fixed that
  distribution in Simcyp's *minimal* PBPK model, with a single adjusting
  compartment (SAC) whose volume is `Vsac = Vc * k12 / k21`. Then they fitted
  a first-order absorption rate `ka` and an apparent oral clearance `CLpo` to
  oral data. Every input is printed in Table 3, and the as-run values are in
  the Simcyp workbook deposited as Supplementary Table S1. That layer is what
  this package ships.
* **Rosuvastatin (the victim) and the DDI are not reproducible.** Rosuvastatin
  runs on the Simcyp library compound file: a full PBPK model with the ADAM
  gut model, a permeability-limited liver, and OATP1B1/1B3, BCRP and MRP4
  transporter kinetics (Supplementary Table S3). The interaction acts through
  a BCRP `Ki` of 0.05 uM in the gut, liver and kidney. Both depend on the
  platform's physiology and its transporter-abundance scaling, and neither is
  published. The DDI ratios (Table 6 and Supplementary Table S2) are therefore
  not reproduced here.

# Population

The omecamtiv mecarbil compound model draws on six clinical studies in healthy
adults (Trivedi 2021 Table 1):

| Study | OM dose | Route | n | Women | Age (y) | Role |
|---|---|---|---|---|---|---|
| Study 5 | 35 mg single dose | IV | 7 | 0% | 21-45 | Distribution (SAAM II) |
| Study 1 | 50 mg single dose | oral | 14 | 43% | 22-54 | `ka`, `CLpo` fit |
| Study 2 | 25 mg b.i.d., 7 days | oral | 20 | 30% | 22-45 | `ka`, `CLpo` fit |
| Study 3 | 50 mg b.i.d., 7 days | oral | 13 | 8% | 20-35 | Verification |
| Study 4 | 25 mg b.i.d. | oral | 13 | 50% | 22-54 | Verification |
| OM-rosuvastatin DDI | 50 mg single dose | oral | 14 | 57% | 18-50 | DDI study |

The DDI-study participants (Supplementary Table S1) had a mean (SD) age of
34.1 (9.71) years and a mean weight of 70.4 (13.1) kg. They were 71.4% White,
21.4% Black and 7.1% multiple race. Table 3 converts the fitted volumes to per-kg
values using a mean body weight of 83.79 kg.

```{r population}
readModelDb("Trivedi_2021_omecamtivMecarbil")()$population[c(
  "n_subjects",
  "n_studies",
  "sex_female_pct",
  "dose_range"
)]
```

# Source trace

Every value is the as-run Simcyp compound-file input from the deposited S1
workbook (sheet `Input Sheet`, column `Inhibitor 1`, compound
`AMG 423_target formulation`). Each one agrees with the rounded value in
Table 3.

| Parameter / equation | Value | Source location |
|---|---|---|
| `lka` (ka) | 0.2234 1/h | S1 workbook `ka (1/h)`; Table 3 0.220 'Parameterized'; Results 0.22 (0.18-0.26) |
| `lcl` (CLpo, CL/F) | 10.81 L/h | S1 workbook `CL (po) (L/h)`; Table 3 10.8 'Parameterized'; Results 10.8 (10.3-11.3) |
| `lvss` (Vss) | 3.8575 L/kg | S1 workbook `Vss (L/kg)`; Table 3 3.86 L/kg (323.3 L / 83.79 kg) |
| `lvp` (Vsac) | 2.369 L/kg | S1 workbook `Volume [Vsac] (L/kg)`; Table 3 2.37 L/kg (198.5 L / 83.79 kg) |
| `lk12` (SAC kin) | 0.31955 1/h | S1 workbook `SAC kin (1/h)`; Table 3 / Results k12 = 0.319 (0.186-0.452) |
| `lk21` (SAC kout) | 0.2 1/h | S1 workbook `SAC kout (1/h)`; Table 3 / Results k21 = 0.200 (0.136-0.265) |
| `etalka`, `etalcl`, `etalvss` | log(1 + 0.3^2) = 0.0862 | S1 workbook `CV ka (%)`, `CV CL (po) (%)`, `CV Vss (%)` = 30 |
| `propSd` | 0 (fixed) | No residual-error model in a Simcyp simulation |
| fa = 1, no lag | n/a | S1 workbook `fa` = 1, `lag time (h)` = 0; Methods 'first-order kinetics' |
| `vc = WT * (Vss - Vsac)` | 1.4885 L/kg typical | Methods (Vss = Vc + Vsac); Table 3 Vc 1.49 L/kg |
| systemic-volume floor 0.035 L/kg | n/a | Smallest `Inhibitor 1 Vsys` among the 50 S1 workbook subjects (0.0347 L/kg) |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Methods 'OM model development' (two-compartment, k12 / k21) |
| `Cc = 1000 * central / vc` | n/a | mg / L to ng/mL (Tables 4-5 units) |

The Table 3 arithmetic closes on the paper's own numbers:

```{r table3-arithmetic}
vc_L <- 124.7
k12 <- 0.319
k21 <- 0.200
k10 <- 0.098
wt_mean <- 83.79
arith <- data.frame(
  Quantity = c(
    "Vsac = Vc * k12 / k21 (L)",
    "Vss = Vc + Vsac (L)",
    "Vc per kg (L/kg)",
    "Vsac per kg (L/kg)",
    "Vss per kg (L/kg)"
  ),
  Recomputed = c(
    vc_L * k12 / k21,
    vc_L * (1 + k12 / k21),
    vc_L / wt_mean,
    198.5 / wt_mean,
    323.3 / wt_mean
  ),
  Printed = c(198.5, 323.3, 1.49, 2.37, 3.86),
  `As run (S1 workbook)` = c(NA, NA, 3.8575 - 2.369, 2.369, 3.8575),
  check.names = FALSE
)
arith$`% diff vs printed` <- 100 * (arith$Recomputed / arith$Printed - 1)
knitr::kable(arith, digits = 4, caption = "Trivedi 2021 Table 3 and Results arithmetic.")
stopifnot(all(abs(arith$`% diff vs printed`) < 0.5))
```

# Model structure checks

```{r load-model}
mod <- readModelDb("Trivedi_2021_omecamtivMecarbil")
ui <- rxode2::rxode(mod)
# The model has both `cl` and `vc`; confirm rxode2 integrates the explicit
# ODEs rather than silently swapping in a linCmt() solution.
stopifnot(is.null(ui$linCmt))
mod_typ <- rxode2::zeroRe(mod)

# Typical-value terminal half-life from the micro-constants at the
# Simcyp reference body weight of 80.7 kg (S1 workbook).
wt_ref <- 80.7
kel <- 10.81 / ((3.8575 - 2.369) * wt_ref)
sum_k <- kel + 0.31955 + 0.2
beta <- (sum_k - sqrt(sum_k^2 - 4 * kel * 0.2)) / 2
cat(sprintf("Typical terminal half-life at %.1f kg: %.1f h\n", wt_ref, log(2) / beta))
```

For a linear model, the oral `AUC(0,inf)` must equal `Dose / CLpo` whatever the
body weight or the distribution. This is how Simcyp's in vivo `CLpo` input is
defined. The check solves the typical subject with tight tolerances.

```{r auc-identity}
ev_id <- rxode2::et(amt = 50, cmt = "depot") |>
  rxode2::et(c(seq(0, 24, by = 0.1), seq(25, 1500, by = 1))) |>
  as.data.frame() |>
  dplyr::mutate(id = 1L, WT = wt_ref)
sim_id <- as.data.frame(rxode2::rxSolve(
  mod_typ,
  ev_id,
  rtol = 1e-10,
  atol = 1e-12,
  returnType = "data.frame"
))
conc_id <- sim_id |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = 1L, Cc = pmax(Cc, 0))
nca_id <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_id, Cc ~ time | id),
  PKNCA::PKNCAdose(data.frame(id = 1L, time = 0, amt = 50), amt ~ time | id),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, cmax = TRUE)
))
auc_inf <- as.data.frame(nca_id) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::pull(PPORRES)
auc_expected <- 50 * 1000 / 10.81
cat(sprintf(
  "AUC(0,inf) = %.1f ng*h/mL; Dose / CLpo = %.1f ng*h/mL (%.3f%% apart)\n",
  auc_inf,
  auc_expected,
  100 * (auc_inf / auc_expected - 1)
))
# Trapezoidal error on a 0.1 h / 1 h grid is ~0.01%; 0.5% is 50x headroom.
stopifnot(abs(auc_inf / auc_expected - 1) < 0.005)
```

# Known-answer check against the deposited Simcyp run

The S1 workbook is the Simcyp output of the OM-rosuvastatin DDI simulation. It
holds 5 trials of 10 subjects, 18-65 years, with 50% women in the input sheet.
Omecamtiv mecarbil is the inhibitor, given as a single 50 mg oral dose. The
workbook reports each virtual subject's omecamtiv mecarbil Cmax, Tmax and
`AUC(0-48)` (sheet `AUC0(Inh 1)(CPlasma)`), and the population mean, 5th and
95th percentile profile (sheet `Conc Profiles CSys(CPlasma)`). The values below
were transcribed from those sheets.

```{r deposit-data}
deposit_median <- c(Cmax = 127.702, Tmax = 4.5251, AUC48 = 3619.68)
deposit_profile <- tibble::tribble(
  ~time, ~mean, ~p5, ~p95,
  0.0000, 0.000, 0.00000, 0.000,
  0.4800, 91.356, 15.89200, 373.430,
  0.9600, 136.710, 28.14900, 518.310,
  1.9200, 170.090, 43.78500, 565.660,
  2.8800, 174.910, 51.96600, 533.690,
  4.0801, 167.660, 59.04000, 421.700,
  6.0001, 148.590, 64.34000, 307.720,
  7.9201, 130.530, 64.46600, 240.230,
  12.0001, 102.550, 51.27300, 168.050,
  16.0801, 84.369, 40.95600, 135.160,
  23.7601, 62.650, 21.52800, 95.306,
  36.0001, 42.454, 3.86910, 71.183,
  48.0000, 30.400, 0.65996, 57.636
)
```

The typical subject at the Simcyp reference weight should sit at the centre of
the deposited distribution. It has no inter-individual variability, and the
reduction uses bioavailability 1:

```{r deposit-typical}
ev_48 <- rxode2::et(amt = 50, cmt = "depot") |>
  rxode2::et(seq(0, 48, by = 0.05)) |>
  as.data.frame() |>
  dplyr::mutate(id = 1L, WT = wt_ref)
sim_48 <- as.data.frame(rxode2::rxSolve(mod_typ, ev_48, returnType = "data.frame")) |>
  dplyr::mutate(id = 1L)
nca_48 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_48, Cc ~ time | id),
  PKNCA::PKNCAdose(data.frame(id = 1L, time = 0, amt = 50), amt ~ time | id),
  intervals = data.frame(start = 0, end = 48, cmax = TRUE, tmax = TRUE, auclast = TRUE)
))
typ_48 <- as.data.frame(nca_48) |>
  dplyr::select(PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
typ_cmp <- data.frame(
  Metric = c("Cmax (ng/mL)", "Tmax (h)", "AUC0-48 (ng*h/mL)"),
  `Typical reduction` = c(typ_48$cmax, typ_48$tmax, typ_48$auclast),
  `Deposited median (50 subjects)` = unname(deposit_median),
  check.names = FALSE
)
typ_cmp$`% diff` <- 100 * (typ_cmp$`Typical reduction` / typ_cmp$`Deposited median (50 subjects)` - 1)
knitr::kable(typ_cmp, digits = 2, caption = "Typical subject vs the deposited Simcyp virtual subjects.")
# Measured: Cmax +0.4%, Tmax +7.2%, AUC0-48 -2.0%. A dose, clearance or
# volume transcription error moves these by tens of percent.
stopifnot(all(abs(typ_cmp$`% diff`) < 10))
```

Next, a stochastic cohort of the same design (200 subjects, 50 mg single
dose). Body weight is drawn as normal with mean 80.7 kg (the Simcyp reference
weight) and SD 13 kg (the DDI study's SD), redrawn outside 50-120 kg. The
robust summaries are compared with the deposited subjects:

```{r deposit-cohort}
set.seed(8604240)
rxode2::rxSetSeed(8604240)

draw_wt <- function(n, mean = 80.7, sd = 13, lo = 50, hi = 120) {
  wt <- numeric(0)
  while (length(wt) < n) {
    x <- stats::rnorm(n, mean, sd)
    wt <- c(wt, x[x >= lo & x <= hi])
  }
  wt[seq_len(n)]
}

n_dep <- 200L
ev_dep <- rxode2::et(amt = 50, cmt = "depot") |>
  rxode2::et(c(0, seq(0.24, 48, by = 0.24))) |>
  rxode2::et(id = seq_len(n_dep)) |>
  as.data.frame() |>
  dplyr::left_join(data.frame(id = seq_len(n_dep), WT = draw_wt(n_dep)), by = "id")
sim_dep <- as.data.frame(rxode2::rxSolve(mod, ev_dep, returnType = "data.frame"))

nca_dep <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(dplyr::filter(sim_dep, !is.na(Cc)), Cc ~ time | id),
  PKNCA::PKNCAdose(data.frame(id = seq_len(n_dep), time = 0, amt = 50), amt ~ time | id),
  intervals = data.frame(start = 0, end = 48, cmax = TRUE, tmax = TRUE, auclast = TRUE)
))
coh <- as.data.frame(nca_dep) |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(median = stats::median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)
coh_cmp <- data.frame(
  Metric = c("Median Cmax (ng/mL)", "Median Tmax (h)", "Median AUC0-48 (ng*h/mL)"),
  `Simulated cohort` = c(coh$cmax, coh$tmax, coh$auclast),
  `Deposited Simcyp subjects` = unname(deposit_median),
  check.names = FALSE
)
coh_cmp$`% diff` <- 100 * (coh_cmp$`Simulated cohort` / coh_cmp$`Deposited Simcyp subjects` - 1)
knitr::kable(coh_cmp, digits = 2, caption = "Cohort medians vs the deposited Simcyp virtual subjects.")
# Medians are robust to which subjects land in the tails; measured
# differences are within about 10% across seeds (AUC0-48 runs 7-9% low).
stopifnot(all(abs(coh_cmp$`% diff`) < 20))
```

```{r deposit-profile-plot, fig.width = 7, fig.height = 4.5}
coh_prof <- sim_dep |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    mean = mean(Cc),
    p5 = stats::quantile(Cc, 0.05),
    p95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  )
ggplot() +
  geom_ribbon(data = coh_prof, aes(time, ymin = p5, ymax = p95), fill = "steelblue", alpha = 0.25) +
  geom_line(data = coh_prof, aes(time, mean), colour = "steelblue", linewidth = 1) +
  geom_point(data = deposit_profile, aes(time, mean), colour = "black") +
  geom_errorbar(data = deposit_profile, aes(time, ymin = p5, ymax = p95), colour = "black", width = 0.6) +
  labs(
    x = "Time after 50 mg oral dose (h)",
    y = "Omecamtiv mecarbil plasma concentration (ng/mL)",
    caption = paste(
      "Blue: this model, 200 virtual subjects (mean and 5th-95th percentile).",
      "Black: deposited Simcyp run, 50 subjects (S1 workbook).",
      sep = "\n"
    )
  ) +
  theme_bw()
```

The deposited mean profile peaks near 2.9 h, well before the median
individual Tmax of 4.5 h. This comes from subjects whose systemic volume is
very small. When a subject's sampled Vss falls close to the fixed Vsac, their
systemic compartment shrinks and they get a tall, early spike. The reduction
reproduces this because it applies the variability to Vss and not to Vc (see
[Assumptions and deviations](#assumptions-and-deviations)).

# Replicating Tables 4 and 5 (Simcyp-simulated omecamtiv mecarbil exposure)

Tables 4 and 5 compare the Simcyp-simulated exposures of the four oral studies
with the observed values. Each virtual study is simulated here with the design
of Table 1: 10 trials of the study's size (130-200 subjects), a single dose or
twice-daily dosing, and a final morning dose on the reported day. Table 4
footnotes define the intervals: `AUC(0-144)` after the single dose, and
`AUC(0-12)` on day 1 and day 8. Table 5 reports study 4 on day 15.

```{r studies-design}
make_study <- function(label, dose, n, n_doses, id_offset) {
  dose_times <- (seq_len(n_doses) - 1) * 12
  last <- max(dose_times)
  obs <- sort(unique(c(
    seq(0, 12, by = 0.1),
    seq(last, last + 12, by = 0.1),
    if (n_doses == 1) seq(12, 144, by = 0.5) else seq(0, last, by = 1)
  )))
  ids <- id_offset + seq_len(n)
  dose_rows <- expand.grid(id = ids, time = dose_times) |>
    dplyr::mutate(evid = 1L, amt = dose, cmt = "depot")
  obs_rows <- expand.grid(id = ids, time = obs) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::left_join(data.frame(id = ids, WT = draw_wt(n)), by = "id") |>
    dplyr::mutate(study = label, last_dose = last) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  make_study("Study 1: 50 mg single dose", 50, 140L, 1L, 0L),
  make_study("Study 2: 25 mg b.i.d., day 8", 25, 200L, 15L, 1000L),
  make_study("Study 3: 50 mg b.i.d., day 8", 50, 130L, 15L, 2000L),
  make_study("Study 4: 25 mg b.i.d., day 15", 25, 130L, 29L, 3000L)
)
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))

sim <- as.data.frame(rxode2::rxSolve(
  mod,
  events,
  keep = c("study", "last_dose"),
  returnType = "data.frame"
))
```

```{r studies-plot, fig.width = 7, fig.height = 5}
sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(tad_last = time - last_dose) |>
  dplyr::filter(tad_last >= 0, tad_last <= 12 | grepl("single", study)) |>
  dplyr::group_by(study, tad_last) |>
  dplyr::summarise(
    median = stats::median(Cc),
    p5 = stats::quantile(Cc, 0.05),
    p95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::filter(tad_last <= 48) |>
  ggplot(aes(tad_last, median)) +
  geom_ribbon(aes(ymin = p5, ymax = p95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~study, scales = "free") +
  labs(
    x = "Time after the (last) dose (h)",
    y = "Omecamtiv mecarbil (ng/mL)",
    caption = "Median and 5th-95th percentile. Compare with Figures S3 and S4 of Trivedi 2021."
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
conc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(Cc = pmax(Cc, 0)) |>
  dplyr::select(id, time, Cc, study, last_dose)
dose <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, study)

intervals <- dplyr::bind_rows(
  data.frame(
    study = "Study 1: 50 mg single dose",
    start = c(0, 0),
    end = c(144, Inf),
    cmax = c(TRUE, FALSE),
    tmax = c(TRUE, FALSE),
    auclast = c(TRUE, FALSE),
    aucinf.obs = c(FALSE, TRUE)
  ),
  data.frame(
    study = c("Study 2: 25 mg b.i.d., day 8", "Study 3: 50 mg b.i.d., day 8"),
    start = 0,
    end = 12,
    cmax = TRUE,
    tmax = TRUE,
    auclast = TRUE,
    aucinf.obs = FALSE
  ),
  data.frame(
    study = c(
      "Study 2: 25 mg b.i.d., day 8",
      "Study 3: 50 mg b.i.d., day 8",
      "Study 4: 25 mg b.i.d., day 15"
    ),
    start = c(168, 168, 336),
    end = c(180, 180, 348),
    cmax = TRUE,
    tmax = TRUE,
    auclast = TRUE,
    aucinf.obs = FALSE
  )
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc, Cc ~ time | study + id, concu = "ng/mL", timeu = "h"),
  PKNCA::PKNCAdose(dose, amt ~ time | study + id, doseu = "mg"),
  intervals = intervals
))
nca_df <- as.data.frame(nca) |>
  dplyr::mutate(
    arm = dplyr::case_when(
      start == 0 & grepl("b.i.d.", study) ~ paste(sub(", day.*", "", study), "day 1"),
      TRUE ~ study
    )
  )
```

Table 4 and Table 5 report the Simcyp-simulated Cmax and AUC as arithmetic
means and Tmax as a median. The simulated cohort is summarised the same way
before comparison. The day-8 and day-15 pre-dose concentrations are read
directly from the simulation at the time of the final dose.

```{r nca-compare}
sim_summary <- nca_df |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    PPORRES = if (PPTESTCD[1] == "tmax") stats::median(PPORRES) else mean(PPORRES),
    .groups = "drop"
  )
ctrough <- sim |>
  dplyr::filter(!is.na(Cc), grepl("b.i.d.", study), time == last_dose) |>
  dplyr::group_by(arm = study) |>
  dplyr::summarise(PPORRES = mean(Cc), .groups = "drop") |>
  dplyr::mutate(PPTESTCD = "ctrough")
sim_summary <- dplyr::bind_rows(sim_summary, ctrough)

# Trivedi 2021 Table 4 (studies 1-2) and Table 5 (studies 3-4), 'Simulated'
# rows. auclast is AUC(0-144) for study 1 and AUC(0-12) otherwise; study 4
# 'AUCtau' is AUC(0-12) on day 15.
published <- tibble::tribble(
  ~arm, ~cmax, ~tmax, ~auclast, ~aucinf.obs, ~ctrough,
  "Study 1: 50 mg single dose", 218, 4.6, 4956, 5227, NA,
  "Study 2: 25 mg b.i.d. day 1", 114, 4.6, 930, NA, NA,
  "Study 2: 25 mg b.i.d., day 8", 256, 3.1, 2512, NA, 162,
  "Study 3: 50 mg b.i.d. day 1", 216, 4.6, 1768, NA, NA,
  "Study 3: 50 mg b.i.d., day 8", 506, 3.1, 5026, NA, 326,
  "Study 4: 25 mg b.i.d., day 15", 266, 3.1, 2624, NA, NA
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = published,
  by = "arm",
  units = c(
    cmax = "ng/mL",
    tmax = "h",
    auclast = "ng*h/mL",
    aucinf.obs = "ng*h/mL",
    ctrough = "ng/mL"
  ),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "This model vs the Simcyp-simulated values of Trivedi 2021 Tables 4-5",
    "(Cmax, AUC and pre-dose concentration as means; Tmax as medians)."
  )
)
```

The AUC means depend almost entirely on `CLpo` and its variability. They run
4-12% below the Simcyp means after the single dose and at steady state, and
18-21% below for the day-1 `AUC(0-12)`. The direction is consistent with the
platform adding clearance variability through its population physiology: the
mean of `Dose / CL` rises with the spread of `CL`. This reduction carries only
the compound file's 30% CV. As a check, `Dose / CLpo` for 50 mg is 4625 ng*h/mL,
and Simcyp's mean study-1 `AUC(0,inf)` of 5227 ng*h/mL is 13% above it. A
log-normal 30% CV alone raises the mean by only 4.4%.

The Cmax means are driven by the right tail, the few subjects with a very
small systemic volume. That makes them sensitive to the platform's handling of
those draws and to which subjects are sampled. They run 5-29% low, are shown
for comparison, and are not gated.

```{r nca-gate}
pct <- function(arm, code) {
  s <- sim_summary$PPORRES[sim_summary$arm == arm & sim_summary$PPTESTCD == code]
  r <- published[[code]][published$arm == arm]
  100 * (s / r - 1)
}
auc_diff <- c(
  pct("Study 1: 50 mg single dose", "auclast"),
  pct("Study 1: 50 mg single dose", "aucinf.obs"),
  pct("Study 2: 25 mg b.i.d., day 8", "auclast"),
  pct("Study 3: 50 mg b.i.d., day 8", "auclast"),
  pct("Study 4: 25 mg b.i.d., day 15", "auclast")
)
print(round(auc_diff, 1))
# Structural: a mis-transcribed clearance or dose moves every mean AUC by
# tens of percent. Measured median about -9% (range -4% to -12%); the bounds
# leave room for the cohort draw, per the robust-quantile rule.
stopifnot(
  abs(stats::median(auc_diff)) < 15,
  all(abs(auc_diff) < 25)
)
```

The observed clinical values (Tables 4-5, 'Observed' rows) are shown for
context. The authors' acceptance criterion was simulated within two-fold of
observed. The largest misses in the paper itself are the study-1 Tmax
(simulated 4.6 h vs observed 9.0 h) and the study-1 Cmax (218 vs 112 ng/mL).

```{r observed}
observed <- tibble::tribble(
  ~Arm, ~`Cmax (ng/mL)`, ~`Tmax (h)`, ~`AUC (ng*h/mL)`,
  "Study 1: 50 mg single dose (AUC0-144)", "112 +/- 19.1", "9.0 (3.0-12.0)", "4430 +/- 1040",
  "Study 2: 25 mg b.i.d. day 1 (AUC0-12)", "77.4 +/- 21.6", "4.0 (0.5-8.0)", "687 +/- 210",
  "Study 2: 25 mg b.i.d. day 8 (AUC0-12)", "256 +/- 71.2", "2.0 (0.5-4.0)", "2570 +/- 739",
  "Study 3: 50 mg b.i.d. day 1 (AUC0-12)", "154 +/- 22", "3.0 (0.5-6.0)", "1330 +/- 217",
  "Study 3: 50 mg b.i.d. day 8 (AUC0-12)", "537 +/- 91.7", "3.0 (0.5-4.0)", "5490 +/- 1000",
  "Study 4: 25 mg b.i.d. day 15 (AUCtau)", "229 +/- 15", "2.0 (1.0-11.8)", "2520 +/- 15"
)
knitr::kable(observed, caption = "Observed omecamtiv mecarbil exposure, Trivedi 2021 Tables 4-5.")
```

# Assumptions and deviations

* **Only the omecamtiv mecarbil layer is encoded.** The rosuvastatin PBPK
  model and the BCRP-mediated interaction (Tables 2 and 6, Supplementary
  Tables S2-S4) need the Simcyp full-PBPK/ADAM rosuvastatin compound file and
  the platform's transporter-abundance scaling. Neither is published, so the
  rosuvastatin AUC and Cmax ratios cannot be reproduced. The compound file
  also lists CYP2C8 and CYP3A4 inhibition, CYP3A4 induction, and OATP1B1/1B3,
  BCRP and MRP4 inhibition constants for omecamtiv mecarbil. Those act only on
  a co-simulated substrate and are not part of this model.
* **Bioavailability 1 with the apparent oral clearance.** Simcyp's in vivo
  `CLpo` input fixes the oral `AUC(0,inf)` at `Dose / CLpo` (fa = 1, no gut
  metabolism). The platform still routes the absorbed dose through a liver
  compartment. That applies an internal hepatic first-pass fraction computed
  from its own hepatic blood flow, which is not printed. Here the absorbed
  dose enters the systemic compartment directly, and elimination uses
  `CLpo`. The typical subject still matches the deposited Simcyp medians
  within 7.5% (Cmax, Tmax and `AUC(0-48)`).
* **The SAAM II intravenous clearance is not carried.** The IV fit gave
  `k10 = 0.098 1/h`, i.e. `CL = k10 * Vc = 12.2 L/h`. That is above the oral
  `CLpo = 10.8 L/h`, so taken literally it implies a bioavailability of 1.13.
  The final Simcyp model uses only `CLpo` for elimination (Table 3 has no
  `k10` row), and this model follows it. It is therefore an oral model. It is
  not a validated description of the 35 mg intravenous data.
* **Variability.** The 30% CVs on `ka`, `CLpo` and `Vss` are Simcyp's
  compound-file inputs, encoded log-normally. They are not population
  estimates. Simcyp's population file adds more variability through its
  physiology (hepatic blood flow, liver volume), which is not reproduced. As
  in Simcyp, `Vsac` carries no variability, so all the `Vss` variability falls
  on the systemic volume. That gives the systemic volume a CV of about 70% here
  (log-normal Vss). The deposited run's 50 subjects show a CV of 59%
  (`Inhibitor 1 Vsys` cv 0.59). The deposited Vss is also less skewed
  (skewness 0.15) than a log-normal 30% CV would give. The platform's
  sampling distribution for Vss is not stated, so the log-normal form is an
  assumption.
* **Systemic-volume floor.** About 5% of `Vss` draws fall below `Vsac`. For
  those subjects the platform keeps a small positive systemic volume and
  shortens the SAC instead. The rule is not published. The model uses a floor
  of 0.035 L/kg, the smallest systemic volume among the 50 deposited
  subjects. The floor never acts at the typical value.
* **Liver lumped into the systemic compartment.** Simcyp's systemic volume
  excludes the liver, about 0.02 L/kg in the deposited run. The workbook's
  SAC clearances (`CLin` 37.86 L/h, `CLout` 38.24 L/h) are consistent with
  that. The paper's own Table 3 value `Vc = Vss - Vsac = 1.49 L/kg` is used
  instead. It differs by about 1.4%.
* **Virtual cohorts.** Body weight is drawn as normal with mean 80.7 kg (the
  Sim-Healthy Volunteers reference weight in the S1 workbook) and SD 13 kg
  (the DDI study), redrawn outside 50-120 kg. Age and sex have no effect in
  this reduction, so they are not simulated.
* **Study numbering in the supplement.** The Supplementary Materials section
  'Development of the OM compound file' calls the single 50 mg dose study 'Study
  4' and the 25 mg b.i.d. study 'Study 1'. The main text (Table 1, Table 4)
  uses 'Study 1' and 'Study 2' for the same studies. The main-text numbering
  is used here.
* **Study 4 observed AUC SD.** Table 5 prints the observed study 4 `AUCtau`
  as 2520 +/- 15 ng*h/mL and Cmax as 229 +/- 15 ng/mL. An SD of 15 on an AUC
  of 2520 is implausibly small next to every other row. It is reproduced as
  printed.
