---
title: "Proinsulin conversion to insulin in type 2 diabetes (Trujillo 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Proinsulin conversion to insulin in type 2 diabetes (Trujillo 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, include = FALSE}
ui <- rxode2::rxode(readModelDb("Trujillo_2025_proinsulin_qsp"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.3390/pharmaceutics17121522> (open access, PMC12736522)
* Supplement: `pharmaceutics-17-01522-s001.zip`, published with the article and
  recovered from the Europe PMC `supplementaryFiles` endpoint. It holds
  Table S1 (the 15 ODEs), Table S2 (the 42 flux and rule expressions) and
  Table S3 (the parameter values for all four virtual patients). **The main
  text prints Equations 1-6 only and contains no parameter values at all**, so
  the supplement is the model.

Trujillo 2025 is an in-silico hypothesis test, not a fitted population
analysis. The authors took an existing SimBiology "Diabetes QSP model" of
glucose homeostasis, added a proinsulin sub-model, built four virtual patients
(VPs), and simulated a hypothetical drug that converts circulating proinsulin
to insulin. The finding was negative: proinsulin conversion produces only a
~0.2% HbA1c reduction, and the programme was stopped. Nothing in the paper is
estimated, so the packaged model carries no IIV and no residual error.

The system spans gastric emptying of carbohydrate and non-carbohydrate
calories, intestinal glucose absorption, hepatic glucose output regulated by a
glucose x insulin / glucagon ratio (`HRS`), muscle uptake via
insulin-independent GLUT1 and insulin- and proinsulin-responsive GLUT4, brain
and other-tissue uptake, renal filtration / reabsorption / excretion,
glucose-stimulated insulin (GSIS) and proinsulin (GSPS) secretion amplified by
active GLP-1, two-compartment insulin disposition, one-compartment proinsulin
disposition, glucagon secretion with a T2DM dysregulation factor, and 30-day /
90-day glucose delay states that generate HbA1c.

# Population

```{r population, echo = FALSE}
pop <- ui$population
tibble::tibble(Field = names(pop), Value = vapply(pop, paste, character(1), collapse = " ")) |>
  knitr::kable()
```

The four virtual patients are model constructs, not fitted individuals. They
were calibrated against fasting insulin, proinsulin and the proinsulin/insulin
ratio reported in the literature and measured in baseline samples from 307
participants with T2DM in two completed Merck Phase 3 studies (Study 11 /
MK-3102-011, NCT01717313, n = 181; Study 24 / MK-3102-024, NCT01755156,
n = 127). The underlying Diabetes QSP model was separately validated against
the insulin glargine arm of Zinman 2012 (Figure S1).

# Source trace

Every packaged value comes from the supplement unless the table says otherwise.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Element, ~`Source location`,
  "All 15 ODEs (gastric, gut, glucose, tubular, insulin x2, GLP-1 x2, glucagon, delays x2, proinsulin, muscle x2)", "Supplementary Table S1",
  "All 42 flux and rule expressions (GSIS, GSPS, HRS, Ra_Liver, BoundG4, renal, incretin, glucagon, A1c, ...)", "Supplementary Table S2",
  "53 of the 55 tabulated parameter values, VPHealthy column (bfx and Km_MIns are omitted -- see below)", "Supplementary Table S3",
  "VPT2DM-1 / -2 / -3 values for the 10 parameters that vary", "Supplementary Table S3, columns 2-4",
  "ProinsulinConv = 1/K_CPro * PlasmaProinsulin (the drug term)", "Main text Equation (6)",
  "The conversion term is ADDED to represent the drug (so it is off at baseline)", "Main text Section 2.2",
  "K_CPro = 20 min chosen to give a ~7 pM/min conversion rate", "Main text Section 2.6 and Table S3",
  "OGTT design: 16 h fast, then 75 g oral glucose", "Main text Section 2.6",
  "Fasting and postprandial glucose / insulin, HbA1c per VP", "Main text Table 1",
  "Fasting insulin, proinsulin, proinsulin/insulin per VP", "Main text Table 2",
  "Vmax_G6P = 560 mg/min", "NOT REPORTED -- derived below from the paper's own numbers"
) |>
  knitr::kable()
```

## The one parameter the paper never prints: `Vmax_G6P`

The `Ra_Liver` rule in Table S2 is
`Ra_Liver = Vmax_G6P / (HRS * Km_Ra_Liver + 1)`, but `Vmax_G6P` has no row in
Table S3 and appears nowhere in the article. Without it the model cannot be
integrated. It was **derived from the paper's own reported values**, not
substituted from outside the paper, along three independent lines:

1. **The model's own reference point.** `HRS` is built so that it equals
   exactly 1 when plasma glucose, effective plasma insulin and glucagon are at
   their tabulated reference values (`NormalGlucose` 90 mg/dL,
   `NormalInsulin` 5 uU/mL, `NormalGlucagon` 75 pmol). At that point
   `Ra_Liver = Vmax_G6P / (1 * 3 + 1) = Vmax_G6P / 4` for VPHealthy.
   `Vmax_G6P = 560` makes basal hepatic glucose release exactly 140 mg/min, or
   2.0 mg/kg/min for a 70 kg adult.
2. **Inversion of the fasting steady state.** Requiring VPHealthy to reproduce
   its Table 1 fasting glucose of 93 mg/dL recovers the same value to the two
   significant figures that 93 is reported to. This is computed below rather
   than asserted.
3. **Cross-check on VPs not used in the derivation** -- see the next section.

Lines 2 and 3 are computed, not asserted: the scan and the inversion are run
against the model itself under "Confirming `Vmax_G6P`" below, once the
steady-state helper has been defined.

# Virtual patients

Ten parameters differ between the four VPs. The values below are Table S3
verbatim.

```{r vps}
vp_pars <- list(
  VPHealthy = c(lkm_ra_liver = log(3),        lkm_ins_g4 = log(37.6137),
                lkm_pro_g4   = log(5187.19),  lvmax_gsis = log(0.2),
                lkm_gsis     = log(155),      lvmax_gsps = log(0.2),
                lkm_gsps     = log(350),      ln_gsps    = log(4.5),
                afx          = 1),
  `VPT2DM-1` = c(lkm_ra_liver = log(1.5),     lkm_ins_g4 = log(75.2274),
                lkm_pro_g4   = log(10374.38), lvmax_gsis = log(0.32),
                lkm_gsis     = log(193.75),   lvmax_gsps = log(0.32),
                lkm_gsps     = log(437.5),    ln_gsps    = log(4),
                afx          = 0.8),
  `VPT2DM-2` = c(lkm_ra_liver = log(2.4),     lkm_ins_g4 = log(47.017125),
                lkm_pro_g4   = log(6483.9875), lvmax_gsis = log(0.175),
                lkm_gsis     = log(221.42857), lvmax_gsps = log(0.175),
                lkm_gsps     = log(500),      ln_gsps    = log(4.3),
                afx          = 0.9),
  `VPT2DM-3` = c(lkm_ra_liver = log(2.1),     lkm_ins_g4 = log(53.73386),
                lkm_pro_g4   = log(7410.271), lvmax_gsis = log(0.142857),
                lkm_gsis     = log(310),      lvmax_gsps = log(0.142857),
                lkm_gsps     = log(700),      ln_gsps    = log(4.2),
                afx          = 0.8)
)
```

## Six of the ten VP-varying rows are generated by exactly two factors

Ten rows of Table S3 vary across the VPs. Six of them are not independent
numbers: they are reproduced to six significant figures by an
insulin-resistance factor `IR` and the tabulated beta-cell factor `bfx`. The
remaining four (`n_GSPS`, `afx`, and `bfx` and `Km_MIns` themselves) are set
independently and are not regenerated here.

```{r two-factor}
IR  <- c(1, 2, 1.25, 1.428571)      # = 3 / Km_Ra_Liver
bfx <- c(1, 0.8, 0.7, 0.5)          # tabulated directly in Table S3

generated <- tibble::tibble(
  VP             = names(vp_pars),
  Km_Ra_Liver    = 3 / IR,
  Km_Ins_GLUT4   = 37.6137 * IR,
  Km_Pro_GLUT4   = 5187.19 * IR,
  Km_GSIS        = 155 / bfx,
  Km_GSPS        = 350 / bfx,
  Vmax_GSIS_GSPS = 0.2 * IR * bfx
)
knitr::kable(generated, digits = 6)

tabulated <- list(
  Km_Ra_Liver    = c(3, 1.5, 2.4, 2.1),
  Km_Ins_GLUT4   = c(37.6137, 75.2274, 47.017125, 53.73386),
  Km_Pro_GLUT4   = c(5187.19, 10374.38, 6483.9875, 7410.271),
  Km_GSIS        = c(155, 193.75, 221.42857, 310),
  Km_GSPS        = c(350, 437.5, 500, 700),
  Vmax_GSIS_GSPS = c(0.2, 0.32, 0.175, 0.142857)
)
for (nm in names(tabulated)) {
  stopifnot(max(abs(generated[[nm]] / tabulated[[nm]] - 1)) < 1e-5)
}
cat("All 24 values (6 parameters x 4 VPs) regenerate from IR and bfx to within 1e-5\n")
cat("relative error. n_GSPS and afx vary across VPs but are NOT generated this way.\n")
```

This settles a question the supplement leaves open. Two rows of Table S3 --
`bfx` and `Km_MIns` (37.6137 in all four VPs) -- are referenced by **no** flux
or rule in Table S2. They are not orphans and they are not an omission: they
are the *generators* of the table, not runtime inputs. Accordingly they are
documented in the model file but deliberately absent from `ini()`, where an
unreferenced parameter would be an error.

# Fasting steady state

The model has no dosing at rest, so the untreated fasting state is the
system's steady state. The vignette reaches it by integrating 400 simulated
days with no meals; this also serves as the starting state for every
subsequent simulation.

```{r fasting}
state_names <- ui$state
init_guess <- c(gastric_cho = 0, gastric_other = 0, gut_glucose = 0,
                glucose = 93 * 140, tubular_glucose = 136, insulin = 0.14,
                insulin_per = 0.87, glp1 = 5, glp1_inactive = 10,
                glucagon = 72, glucose_delay30 = 93, glucose_delay90 = 93,
                proinsulin = 0.028, insulin_musc = 4.7, proinsulin_musc = 0.9)
stopifnot(setequal(names(init_guess), state_names))

solve_to_fasting <- function(pars, treated = 0, days = 400) {
  ev <- as.data.frame(rxode2::et(seq(0, days * 1440, length.out = 200),
                                 cmt = "glucose"))
  ev$ON_TREATMENT <- treated
  out <- rxode2::rxSolve(ui, ev, params = pars, inits = init_guess,
                         atol = 1e-10, rtol = 1e-8)
  res <- out[nrow(out), ]
  # The two HbA1c delay states are pure low-pass filters of plasma glucose with
  # 30-day and 90-day time constants, so they are the slowest states in the
  # system by three orders of magnitude. 400 simulated days leaves the 90-day
  # state ~5e-6 mg/dL short of its limit; pin both to the converged plasma
  # glucose, which is their exact steady-state value, so that the state handed
  # to the OGTT simulations is a true fixed point of all 15 ODEs.
  res$glucose_delay30 <- res$glucosePlasma
  res$glucose_delay90 <- res$glucosePlasma
  res$hba1c <- 1.4 + res$glucosePlasma / 28
  res
}

fasting <- lapply(vp_pars, solve_to_fasting)
```

## Against Table 1 and Table 2

```{r fasting-table, echo = FALSE}
reported <- tibble::tibble(
  VP         = names(vp_pars),
  glu_rep    = c(93, 140, 131, 167),
  ins_rep    = c(28, 17, 34, 13),
  pro_rep    = c(5, 7, 25, 3),
  ratio_rep  = c(0.18, 0.41, 0.73, 0.23)
)
fast_tab <- reported |>
  mutate(
    glu_mod   = vapply(fasting, function(x) x$glucosePlasma, numeric(1)),
    ins_mod   = vapply(fasting, function(x) x$insulinPlasma, numeric(1)),
    pro_mod   = vapply(fasting, function(x) x$proinsulinPlasma, numeric(1)),
    ratio_mod = pro_mod / ins_mod
  ) |>
  transmute(
    VP,
    `Glucose model (mg/dL)` = round(glu_mod, 1), `Glucose reported` = glu_rep,
    `Insulin model (pM)`    = round(ins_mod, 1), `Insulin reported` = ins_rep,
    `Proinsulin model (pM)` = round(pro_mod, 2), `Proinsulin reported` = pro_rep,
    `P/I model` = round(ratio_mod, 2), `P/I reported` = ratio_rep
  )
knitr::kable(fast_tab)
```

**VPHealthy and VPT2DM-3 reproduce their published baselines; VPT2DM-1 and
VPT2DM-2 do not.** VPHealthy lands on all four reported quantities
(93.0 / 28.2 / 5.46 / 0.19 against 93 / 28 / 5 / 0.18). VPT2DM-3 reproduces
fasting glucose to 0.8% and fasting insulin to 8%, and overshoots fasting
proinsulin by about a quarter (3.8 pM against a reported 3 pM given to one
significant figure).

Note that VPHealthy's *glucose* agreement is not independent evidence --
`Vmax_G6P` was inverted from exactly that number. The load-bearing checks are
insulin, proinsulin and the P/I ratio for VPHealthy, and all three quantities
for VPT2DM-3, none of which entered the derivation:

```{r fasting-assert}
h <- fasting[["VPHealthy"]]
s <- fasting[["VPT2DM-3"]]
stopifnot(
  abs(h$glucosePlasma    /  93 - 1) < 0.01,   # Table 1
  abs(h$insulinPlasma    /  28 - 1) < 0.02,   # Table 1 / Table 2
  abs(h$proinsulinPlasma /   5 - 1) < 0.10,   # Table 2 (reported to 1 sig fig)
  abs(s$glucosePlasma    / 167 - 1) < 0.02,   # Table 1
  abs(s$insulinPlasma    /  13 - 1) < 0.10,   # Table 1 / Table 2
  abs(s$proinsulinPlasma /   3 - 1) < 0.30    # Table 2 (reported to 1 sig fig)
)
cat(sprintf("VPHealthy glucose %.2f / insulin %.2f / proinsulin %.2f\n",
            h$glucosePlasma, h$insulinPlasma, h$proinsulinPlasma))
cat(sprintf("VPT2DM-3  glucose %.2f / insulin %.2f / proinsulin %.2f\n",
            s$glucosePlasma, s$insulinPlasma, s$proinsulinPlasma))
```

The VPT2DM-1 and VPT2DM-2 discrepancies are characterised in the Errata section
below; they are a property of the published tables, not of this implementation.

## Confirming `Vmax_G6P`

The one value the paper never prints. Both the sensitivity scan and the
inversion promised in the Source trace section are computed here against the
model itself.

```{r vmax-g6p}
fasting_glucose_at <- function(vg6p) {
  p <- vp_pars[["VPHealthy"]]
  p["lvmax_g6p"] <- log(vg6p)
  solve_to_fasting(p)$glucosePlasma
}
g6p_scan <- tibble::tibble(
  Vmax_G6P = c(500, 560, 600),
  `Ra_Liver at HRS = 1 (mg/min)` = Vmax_G6P / 4,
  `VPHealthy fasting glucose (mg/dL)` = vapply(Vmax_G6P, fasting_glucose_at, numeric(1)),
  `Reported (Table 1)` = 93
)
knitr::kable(g6p_scan, digits = 2)

# Invert the fasting steady state: which Vmax_G6P reproduces the reported
# 93 mg/dL exactly?
inverted <- stats::uniroot(function(v) fasting_glucose_at(v) - 93,
                           interval = c(400, 800), tol = 1e-4)$root
cat(sprintf("Inversion on the reported 93 mg/dL gives Vmax_G6P = %.1f, i.e. %.0f to the\n",
            inverted, signif(inverted, 2)))
cat("two significant figures 93 is reported to -- the value the model file carries.\n")
stopifnot(abs(signif(inverted, 2) - 560) < 1e-8)
```

## Structural checks

Three identities that must hold exactly if the equations were transcribed
correctly.

```{r structural}
# 1. Glucagon mass balance. PlasmaGlucagonClearance = PlasmaGlucagon / 15, so
#    at steady state glucagon = 15 * PlasmaGlucagonSecretion. For VPHealthy
#    (afx = 1) the dysregulation term vanishes.
g <- h$glucosePlasma
ggn_secretion <- 3.25 + 20 * 1 / ((0.02^4) * (g^4) + 1)
stopifnot(abs(h$glucagonPlasma / (15 * ggn_secretion) - 1) < 1e-4)

# 2. HbA1c is a pure low-pass filter of glucose: at steady state both delay
#    states equal plasma glucose, so A1c = 1.4 + glucose / 28.
stopifnot(abs(h$hba1c - (1.4 + g / 28)) < 1e-6)

# 3. Insulin steady state = secretion / (CL/V), with the peripheral
#    compartment at equilibrium and therefore contributing zero net flux.
stopifnot(abs(h$insulin_per / 31.6 - h$insulin / 5.11) < 1e-6)

cat(sprintf("glucagon %.3f = 15 x %.4f | HbA1c %.4f = 1.4 + %.2f/28 | peripheral insulin at equilibrium\n",
            h$glucagonPlasma, ggn_secretion, h$hba1c, g))
```

# Oral glucose tolerance test

Section 2.6 specifies the OGTT completely: a 16 h fast, then a 75 g oral
glucose load. Carbohydrate enters the gastric compartment in kcal; at 4 kcal/g
a 75 g load is 300 kcal, and the `250 * GastricCarbohydrateEmptying` term of
Table S1 delivers 250 mg of glucose to the intestine per kcal. The two
conversions cancel exactly:

```{r ogtt-mass-balance}
stopifnot(identical(300L * 250L, 75000L))   # 300 kcal x 250 mg/kcal = 75 g
cat("300 kcal of carbohydrate delivers exactly 75000 mg = 75 g of glucose to the gut.\n")
```

```{r ogtt}
run_ogtt <- function(pars, start_state, treated = 0, k_cpro = NULL,
                     tmax = 300) {
  st <- unlist(start_state[, state_names])
  ev <- rxode2::et(amt = 300, cmt = "gastric_cho", time = 0) |>
    rxode2::et(seq(0, tmax, by = 2), cmt = "glucose")
  ev <- as.data.frame(ev)
  ev$ON_TREATMENT <- treated
  p <- pars
  if (!is.null(k_cpro)) p["lk_cpro"] <- log(k_cpro)
  rxode2::rxSolve(ui, ev, params = p, inits = st, atol = 1e-10, rtol = 1e-8) |>
    as.data.frame()
}

ogtt <- Map(function(p, f) run_ogtt(p, f), vp_pars, fasting)
```

## Replicates Figure 2 of Trujillo 2025

Figure 2 overlays simulated OGTT glucose (A), insulin (B) and proinsulin (C)
for the three diabetic VPs on the Yoshioka 1988 reference data.

```{r fig2, fig.height = 5.2}
ogtt_long <- bind_rows(lapply(names(ogtt), function(nm) {
  ogtt[[nm]] |>
    transmute(VP = nm, time,
              `Glucose (mg/dL)` = glucosePlasma,
              `Insulin (pM)`    = insulinPlasma,
              `Proinsulin (pM)` = proinsulinPlasma)
})) |>
  pivot_longer(-c(VP, time), names_to = "Analyte", values_to = "value")

ggplot(filter(ogtt_long, VP != "VPHealthy"),
       aes(time, value, colour = VP)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Analyte, scales = "free_y", ncol = 1) +
  labs(x = "Time after 75 g oral glucose (min)", y = NULL,
       title = "Replicates Figure 2 of Trujillo 2025: simulated OGTT in the T2DM VPs") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Against the Table 1 postprandial values

Table 1 reports a single "postprandial" glucose and insulin per VP without
stating the sampling time; both the simulated peak and the 120 min value are
shown.

```{r ogtt-table, echo = FALSE}
tibble::tibble(
  VP = names(ogtt),
  `Glucose peak (model)` = round(vapply(ogtt, function(x) max(x$glucosePlasma), numeric(1)), 1),
  `Glucose 120 min (model)` = round(vapply(ogtt, function(x) x$glucosePlasma[which.min(abs(x$time - 120))], numeric(1)), 1),
  `Glucose reported` = c(152, 209, 196, 240),
  `Insulin peak (model)` = round(vapply(ogtt, function(x) max(x$insulinPlasma), numeric(1)), 1),
  `Insulin 120 min (model)` = round(vapply(ogtt, function(x) x$insulinPlasma[which.min(abs(x$time - 120))], numeric(1)), 1),
  `Insulin reported` = c(180, 150, 270, 110)
) |>
  knitr::kable()
```

Simulated peak OGTT glucose runs above the reported postprandial values in
every VP, by 2 mg/dL (VPT2DM-1, 211 vs 209) to 35 mg/dL (VPT2DM-2, 231 vs 196).
The paper administers its OGTT after 52 weeks of three meals a day, whereas
this vignette starts from the pure fasting steady state (the meal schedule is
not reported -- see Errata), and Table 1 does not state the sampling time its
"postprandial" values refer to, so an offset of this size is expected. The
ordering across VPs is reproduced on glucose. Peak insulin is not ordered
correctly for VPT2DM-1 and VPT2DM-2, which is the same column-exchange signal
the Errata section characterises.

# Proinsulin conversion therapy

The hypothetical drug adds `ProinsulinConv = 1/K_CPro * PlasmaProinsulin`
(Equation 6) to the proinsulin balance and to the plasma insulin balance.
Section 2.2 states that this term was *added* to the model to represent the
therapy, so it is absent from the untreated base model; the packaged model
gates it on the `ON_TREATMENT` covariate.

## The gating is confirmed numerically

If the conversion term were active in the untreated base model, VPHealthy's
fasting proinsulin would sit near 1.3 pM rather than the 5 pM of Table 2:

```{r gating}
untrt <- fasting[["VPHealthy"]]
trt   <- solve_to_fasting(vp_pars[["VPHealthy"]], treated = 1)
untreated  <- untrt$proinsulinPlasma
treated_ss <- trt$proinsulinPlasma
cat(sprintf("Fasting proinsulin: untreated %.2f pM, treated %.2f pM (Table 2 reports 5 pM)\n",
            untreated, treated_ss))
stopifnot(abs(untreated / 5 - 1) < 0.10, treated_ss < 2)
```

The exact structural check is the steady-state proinsulin mass balance. Setting
`d(PlasmaProinsulin)/dt = 0` in Table S1 gives

`ProinsulinSecretion = PlasmaProinsulin * (CL_Proinsulin/V_Proinsulin + ON_TREATMENT/K_CPro)`,

with `ProinsulinSecretion = GSPS * Incretin` rebuilt from the simulated glucose
and active GLP-1. This must hold to solver tolerance in **both** arms.

```{r gating-balance}
V_PRO <- 5.11; CL_PRO <- 0.084; K_CPRO <- 20

proinsulin_balance <- function(row, on_treatment) {
  gsps     <- 0.2 * row$glucosePlasma^4.5 / (350^4.5 + row$glucosePlasma^4.5)
  incretin <- 5 * row$glp1 / (23 + row$glp1)
  secretion <- gsps * incretin                                  # nmol/min
  amount    <- row$proinsulinPlasma / 1000 * V_PRO              # nmol
  removal   <- amount * (CL_PRO / V_PRO + on_treatment / K_CPRO)
  c(secretion = secretion, removal = removal, relerr = removal / secretion - 1)
}
bal <- rbind(Untreated = proinsulin_balance(untrt, 0),
             Treated   = proinsulin_balance(trt, 1))
knitr::kable(bal, digits = 8)
stopifnot(max(abs(bal[, "relerr"])) < 1e-5)
```

The treated/untreated ratio is therefore the analytic removal ratio
`(CL/V) / (CL/V + 1/K_CPro)` **times** the change in secretion. Secretion is not
constant across the arms: converting proinsulin adds insulin, which lowers
fasting glucose slightly, which lowers `GSPS`. Because `GSPS` is far below its
Km the correction is exactly the glucose ratio raised to `n_GSPS`:

```{r gating-ratio}
open_loop  <- (CL_PRO / V_PRO) / (CL_PRO / V_PRO + 1 / K_CPRO)
feedback   <- (trt$glucosePlasma / untrt$glucosePlasma)^4.5
observed   <- treated_ss / untreated
stopifnot(abs(observed / (open_loop * feedback) - 1) < 1e-3)
cat(sprintf("observed %.5f = open-loop %.5f x glucose feedback %.5f (= %.5f)\n",
            observed, open_loop, feedback, open_loop * feedback))
cat(sprintf("Removal fraction %.1f%%; fasting glucose falls %.3f mg/dL on treatment.\n",
            100 * (1 - observed), untrt$glucosePlasma - trt$glucosePlasma))
```

## Replicates Figure 3 of Trujillo 2025

Figure 3 shows proinsulin (A), insulin (B) and glucose (C) during an OGTT with
and without conversion therapy in VPT2DM-2.

```{r fig3, fig.height = 5.2}
vp_fig3 <- "VPT2DM-2"
trt_cmp <- bind_rows(
  run_ogtt(vp_pars[[vp_fig3]], fasting[[vp_fig3]], treated = 0) |> mutate(Arm = "Untreated"),
  run_ogtt(vp_pars[[vp_fig3]], fasting[[vp_fig3]], treated = 1) |> mutate(Arm = "Conversion therapy")
) |>
  transmute(Arm, time,
            `Proinsulin (pM)` = proinsulinPlasma,
            `Insulin (pM)`    = insulinPlasma,
            `Glucose (mg/dL)` = glucosePlasma) |>
  pivot_longer(-c(Arm, time), names_to = "Analyte", values_to = "value")

ggplot(trt_cmp, aes(time, value, colour = Arm)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~Analyte, scales = "free_y", ncol = 1) +
  labs(x = "Time after 75 g oral glucose (min)", y = NULL,
       title = paste("Replicates Figure 3 of Trujillo 2025 (", vp_fig3, ")")) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## The two quantitative treatment claims

Section 2.6 chose the conversion rate as "~7 pM/min", and Section 3.2 reports
that "insulin levels increase by ~10 pM at the peak of secretion".

```{r trt-claims}
trt_summary <- bind_rows(lapply(names(vp_pars), function(nm) {
  o0 <- run_ogtt(vp_pars[[nm]], fasting[[nm]], treated = 0)
  o1 <- run_ogtt(vp_pars[[nm]], fasting[[nm]], treated = 1)
  tibble::tibble(
    VP = nm,
    `Peak conversion rate (pM/min)` = max(o0$proinsulinPlasma) / 20,
    `Peak insulin untreated (pM)`   = max(o0$insulinPlasma),
    `Peak insulin treated (pM)`     = max(o1$insulinPlasma),
    `Insulin rise (pM)`             = max(o1$insulinPlasma) - max(o0$insulinPlasma),
    `Mean glucose change (mg/dL)`   = mean(o1$glucosePlasma) - mean(o0$glucosePlasma)
  )
}))
knitr::kable(trt_summary, digits = 2)

stopifnot(
  # "conversion rates of up to ~7 pM/min were explored" -- the tabulated
  # K_CPro = 20 min produces peak rates that bracket 7 pM/min across the VPs.
  min(trt_summary$`Peak conversion rate (pM/min)`) < 7,
  max(trt_summary$`Peak conversion rate (pM/min)`) > 7,
  # "insulin levels increase by ~10 pM at the peak of secretion"
  all(trt_summary$`Insulin rise (pM)` > 2),
  all(trt_summary$`Insulin rise (pM)` < 20),
  # the paper's central negative finding: glucose barely moves
  all(abs(trt_summary$`Mean glucose change (mg/dL)`) < 5)
)
cat(sprintf("Peak conversion rates span %.1f-%.1f pM/min, bracketing the ~7 pM/min of Section 2.6;\n",
            min(trt_summary$`Peak conversion rate (pM/min)`),
            max(trt_summary$`Peak conversion rate (pM/min)`)))
cat(sprintf("insulin rises %.1f-%.1f pM at the OGTT peak against the ~10 pM of Section 3.2;\n",
            min(trt_summary$`Insulin rise (pM)`), max(trt_summary$`Insulin rise (pM)`)))
cat(sprintf("mean OGTT glucose falls by only %.2f-%.2f mg/dL across the VPs.\n",
            min(abs(trt_summary$`Mean glucose change (mg/dL)`)),
            max(abs(trt_summary$`Mean glucose change (mg/dL)`))))
```

The last column is the paper's central result. Because HbA1c in this model is
exactly `1.4 + mean_glucose / 28`, a mean-glucose change of 0.8-2.6 mg/dL maps
to an HbA1c change of 0.03-0.09%, the same direction and order of magnitude as
the ~0.2% the paper reports, and far below any clinically meaningful threshold.
The vignette's figure is smaller than the paper's because it is taken over a
5 h OGTT rather than over a year of three-meal-a-day feeding, which cannot be
reproduced (the meal schedule is unreported -- see Errata).

## Replicates Figure S2: varying the conversion rate

```{r figS2}
rates <- c(200, 100, 50, 20, 10)   # K_CPro (min); smaller = faster conversion
figS2 <- bind_rows(lapply(rates, function(k) {
  run_ogtt(vp_pars[["VPT2DM-2"]], fasting[["VPT2DM-2"]], treated = 1, k_cpro = k) |>
    mutate(`K_CPro (min)` = factor(k, levels = rates))
}))
ggplot(figS2, aes(time, proinsulinPlasma, colour = `K_CPro (min)`)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after 75 g oral glucose (min)", y = "Plasma proinsulin (pM)",
       title = "Replicates Figure S2: proinsulin during an OGTT vs conversion rate") +
  theme_bw()
```

Faster conversion (smaller `K_CPro`) drives proinsulin toward zero in the
fasting and late postprandial phases but leaves a transient peak during the
absorption phase -- exactly the limitation the Discussion describes ("the
postprandial rate of proinsulin (and insulin) secretion increases dramatically
during an OGTT such that a small transient amount of proinsulin remained").

## Replicates Figure 4B: raising proinsulin secretion

Section 3.2 raised `Vmax_GSPS` 2-fold and 3-fold to create VPs with higher
proinsulin/insulin ratios.

```{r fig4b}
ratio_scan <- bind_rows(lapply(names(vp_pars)[-1], function(nm) {
  bind_rows(lapply(c(1, 2, 3), function(m) {
    p <- vp_pars[[nm]]
    p["lvmax_gsps"] <- p["lvmax_gsps"] + log(m)
    f <- solve_to_fasting(p)
    tibble::tibble(VP = nm, Multiplier = m,
                   `P/I ratio` = f$proinsulinPlasma / f$insulinPlasma)
  }))
}))
knitr::kable(pivot_wider(ratio_scan, names_from = Multiplier,
                         values_from = `P/I ratio`,
                         names_prefix = "Vmax_GSPS x"), digits = 3)

# The fasting proinsulin/insulin ratio is exactly linear in Vmax_GSPS: fasting
# insulin does not depend on Vmax_GSPS at all, and proinsulin is proportional
# to it. This is a structural identity, so assert it tightly.
lin <- ratio_scan |>
  group_by(VP) |>
  summarise(r2 = `P/I ratio`[Multiplier == 2] / `P/I ratio`[Multiplier == 1],
            r3 = `P/I ratio`[Multiplier == 3] / `P/I ratio`[Multiplier == 1],
            .groups = "drop")
stopifnot(max(abs(lin$r2 - 2)) < 1e-3, max(abs(lin$r3 - 3)) < 1e-3)
knitr::kable(lin, digits = 5)
```

The ratio scales exactly 1:2:3 with `Vmax_GSPS`, reproducing the paper's
statement that "the proinsulin/insulin ratios increased 2-3-fold". The
absolute range differs from the reported 0.53-2.33 only because VPT2DM-2's
tabulated baseline ratio does not match its reported one -- see Errata.

# Assumptions and deviations

* **`Vmax_G6P` is not reported anywhere on disk.** It is derived from the
  paper's own numbers as documented above (`560 mg/min`), giving exactly
  140 mg/min basal hepatic glucose release at the model's own reference point
  and reproducing VPHealthy's reported fasting glucose. It is flagged inline
  in `ini()` and is the only value in the model file that is not a direct
  transcription. No value was taken from outside the paper.
* **The conversion term is gated on `ON_TREATMENT`.** Table S3 lists
  `K_CPro = 20` for all four VPs with no on/off column, but Section 2.2 says
  the term was "added" to represent the drug. The untreated reading is
  confirmed numerically above (fasting proinsulin 5.5 pM untreated vs 1.3 pM
  treated, against the 5 pM of Table 2). Note that Equation (5) as printed
  already carries a `ProinsulinConv` term, and the Discussion refers to "the
  maximum proinsulin conversion in the base model"; but Table S2 defines only
  one `ProinsulinConv` expression and Table S3 gives `K_CPro = 20` in all four
  VP columns with no untreated value, so the tabulated rate can only be the
  treated one. Reproducing Table 2 requires the term to be off at baseline.
* **The meal schedule is not reported.** `FoodCarbonhydrateIntake` and
  `FoodOtherIntake` are inputs in Table S1, but the paper never states the
  caloric content or composition of the "three meals a day". The 52-week
  meal-fed simulations behind Figure 3D (HbA1c over time) and the absolute
  HbA1c values of Table 1 therefore **cannot be reproduced**, and are not
  attempted here. Calibrating a meal so that VPHealthy reaches its reported
  HbA1c of 4.9 requires about 530 kcal/day, which is not a plausible diet, so
  the gap is real rather than a matter of picking a sensible default. All
  validation in this vignette is restricted to the fasting steady state and
  the fully specified 75 g OGTT.
* **`bfx` and `Km_MIns` are omitted from `ini()`.** They are tabulated in
  Table S3 but referenced by no flux in Table S2. As shown above they are the
  generators of the VP parameter table, not runtime inputs; an unreferenced
  `ini()` parameter is an error in rxode2.
* **No IIV and no residual error.** The paper reports none: it is a
  deterministic hand-built model with four hand-constructed virtual patients.
* **Default `ini()` is VPHealthy**, the only VP that reproduces every reported
  baseline quantity. The other three parameter sets are given above and in the
  model file's source-trace comments.
* Table S2 spells one rule `ProinsulinEqulibrium` while Table S1 references
  `ProinsulinEquilibrium`. Transcribed as the same quantity; noted in the
  model file.
* Two Hill exponents are hardcoded to 4 in Table S2 (`GSIS` and the glucagon
  suppression term) rather than being tabulated, as is the 15 min glucagon
  clearance time constant and the 30-day / 90-day delay constants. All are
  carried verbatim.

# Errata: VPT2DM-1 and VPT2DM-2 do not reproduce their published baselines

VPHealthy and VPT2DM-3 reproduce Tables 1 and 2 on every quantity. VPT2DM-1
and VPT2DM-2 do not, and the pattern is systematic: **their Table S3 columns
appear to be exchanged relative to Tables 1 and 2.**

```{r errata}
swap_check <- bind_rows(lapply(c("VPT2DM-1", "VPT2DM-2", "VPT2DM-3"), function(obs_vp) {
  swapped <- c(`VPT2DM-1` = "VPT2DM-2", `VPT2DM-2` = "VPT2DM-1",
               `VPT2DM-3` = "VPT2DM-3")[[obs_vp]]
  rep_row <- filter(reported, VP == obs_vp)
  err <- function(par_vp) {
    f <- fasting[[par_vp]]
    abs(log(f$glucosePlasma / rep_row$glu_rep)) +
      abs(log(f$insulinPlasma / rep_row$ins_rep)) +
      abs(log(f$proinsulinPlasma / rep_row$pro_rep))
  }
  tibble::tibble(`Reported as` = obs_vp,
                 `|log fold-error|, columns as printed` = err(obs_vp),
                 `|log fold-error|, columns exchanged`  = err(swapped))
}))
knitr::kable(swap_check, digits = 3)
cat(sprintf("Total: as printed %.3f, exchanged %.3f\n",
            sum(swap_check[[2]]), sum(swap_check[[3]])))
```

Three independent lines point the same way:

1. **Numerically**, exchanging the two columns cuts the total fasting
   log fold-error by roughly 2.5-fold, and it also fixes the OGTT peak insulin
   (VPT2DM-1 330 pM vs 150 reported as printed, 167 pM vs 150 exchanged).
2. **The paper's own prose.** Section 2.5 describes VPT2DM-1 and VPT2DM-2 as
   showing "increasing insulin resistance" in that order. The insulin
   resistance factor derived above is 2 for the printed VPT2DM-1 column and
   1.25 for the printed VPT2DM-2 column -- decreasing, not increasing. Under
   the exchange it reads 1.25 then 2.
3. **The hyper-proinsulinemia claim.** Section 3.1 says VPT2DM-2 "was
   engineered to have severe hyper-proinsulinemia" and Table 2 gives it the
   highest proinsulin of any VP (25 pM, ratio 0.73). Only the exchanged
   assignment gives VPT2DM-2 the highest simulated proinsulin.

**The packaged model ships Table S3 exactly as printed.** The tables are both
from the paper and both are primary; resolving the conflict by silently
relabelling the parameter columns would substitute this analysis for the
authors'. Users reproducing Figures 3 and 4 for a named VP should be aware of
the ambiguity and may wish to exchange the two parameter sets.

# Session info

```{r session, echo = FALSE}
sessionInfo()
```
