---
title: "Elexacaftor + tezacaftor + ivacaftor in children with cystic fibrosis (Truong 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elexacaftor + tezacaftor + ivacaftor in children with cystic fibrosis (Truong 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Models and source

Truong 2025 reports **three independent one-compartment population PK models**, one
per component of the elexacaftor/tezacaftor/ivacaftor (ETI) fixed combination. They
were fitted separately in Monolix 2023R1 and share no random effects, so they are
packaged as three model files and validated together here.

```{r models}
mods <- list(
  Elexacaftor = readModelDb("Truong_2025_elexacaftor"),
  Tezacaftor  = readModelDb("Truong_2025_tezacaftor"),
  Ivacaftor   = readModelDb("Truong_2025_ivacaftor")
)
```

* Citation: `r rxode2::rxode(mods$Elexacaftor)$reference`
* Article: <https://doi.org/10.1111/cts.70245>
* Supplement (Figures S1-S5 and Table S1): <https://doi.org/10.1111/cts.70245>

Structurally the three differ only in whether an absorption lag time was needed:

| | Elexacaftor | Tezacaftor | Ivacaftor |
|---|---|---|---|
| Absorption | lagged first-order | first-order | lagged first-order |
| Dosing interval | 24 h | 24 h | 12 h |
| Residual error | proportional | proportional | additive + proportional |

# Population

The analysis is an ancillary pharmacokinetic sub-study of **MODUL-CF**, a French
prospective multicentre cohort (EUDRACT 2018-002624-16, NCT03894657) that enrolled
across 20 pediatric university hospital centres between March 2022 and March 2024.
Ninety-six children with cystic fibrosis aged 2-18 years, each carrying at least one
F508del allele and established on label ETI dosing, contributed 150 plasma
concentrations per compound. Sampling was a sparse therapeutic-drug-monitoring
design: a trough immediately before intake and/or a peak roughly 4 h after intake,
with a median of 1 sample per child per compound (range 1-5).

Baseline demographics are Truong 2025 Table 1: 66.7% male overall, with median age
and weight of 4.0 years / 16.0 kg in the 47 children under 6, 8.8 years / 25.0 kg in
the 19 children aged 6 to under 12, and 15.8 years / 51.1 kg in the 30 children aged
12 and over. No child had renal failure, liver disease, or coadministration of a
strong CYP3A inducer or inhibitor. Race and ethnicity were not recorded.

The same information is available programmatically from each model's `population`
metadata (`readModelDb("Truong_2025_elexacaftor")()$population`).

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry in
`inst/modeldb/specificDrugs/Truong_2025_{elexacaftor,tezacaftor,ivacaftor}.R`. They
are collected here for review.

| Equation / parameter | Elexacaftor | Tezacaftor | Ivacaftor | Source location |
|---|---|---|---|---|
| `ltlag` (Tlag, h) | 2.17 (fix) | -- (none) | 0.70 (fix) | Table 2, "Tlag" row |
| `lka` (Ka, 1/h) | 0.59 (fix) | 0.894 (fix) | 0.42 (fix) | Table 2, "Ka" row |
| `lcl` (CL/F, L/h per 70 kg) | 1.57 (3.9% RSE) | 1.32 (3.8%) | 13.4 (4.8%) | Table 2, "CL/F" row |
| `lvc` (V/F, L per 70 kg) | 56.1 (9.9%) | 29.3 (6.8%) | 183 (12.6%) | Table 2, "V/F" row |
| `e_wt_cl` | 0.75 (fix) | 0.75 (fix) | 0.75 (fix) | Methods 2.4 final paragraph; Results 3.2 |
| `e_wt_vc` | 1 (fix) | 1 (fix) | 1 (fix) | Methods 2.4 final paragraph; Results 3.2 |
| `etalcl` (variance) | 0.31^2 | 0.26^2 | 0.37^2 | Table 2, "omega CL" row |
| `addSd` (mg/L) | -- | -- | 0.131 (43.0%) | Table 2, "sigma additive" row |
| `propSd` | 0.247 (9.2%) | 0.302 (8.5%) | 0.278 (12.5%) | Table 2, "sigma proportional" row |
| `ref_wt` = 70 kg | n/a | n/a | n/a | Table 2 column units ("L/h/70kg", "L/70kg"); Discussion paragraph 1 |
| One-compartment ODE with first-order absorption and elimination | | | | Results 3.2 paragraph 1 |
| Exponential between-subject variability on CL/F only | | | | Methods 2.4; Results 3.2 |
| Allometric `(WT/70)^0.75` on CL/F, `(WT/70)^1` on V/F | | | | Methods 2.4 final paragraph; Results 3.2 |

Two points deserve emphasis.

**The absorption parameters were not estimated.** Truong 2025 Results 3.2 states that
"[g]iven the limited number of plasma samples available during the absorption phase in
this study, the absorption parameters (i.e., lag time [Tlag] and/or absorption
constant [Ka]) were set to values reported in the literature for the three molecules
[27]." Reference [27] is Tsai et al., *Pulm Ther* 2020;6(2):275-286. That source does
confirm the elexacaftor (ka 0.59 /h, Tlag 2.17 h) and tezacaftor (ka 0.894 /h) values.
It does **not** publish a first-order ka or Tlag for ivacaftor -- its ivacaftor compound
file uses an ADAM absorption model with input type "Predicted" from a Caco-2
permeability of 11.9e-6 cm/s. The ivacaftor values 0.42 /h and 0.70 h are therefore
traceable to Truong 2025 Table 2 itself and to no upstream printed table. They are
encoded as printed. See also the Assumptions section.

**The omega column is a standard deviation, not a variance.** The Table 2 footnote
defines omega as the "interindividual variability estimate expressed as standard
deviation". The paper's own derived numbers confirm it: the Table S1 overall
elexacaftor AUC0-24h IQR of 142-209 mg*h/L spans a ratio of 1.46, against
`exp(2 * 0.6745 * 0.31) = 1.52` expected under an SD of 0.31 and 2.12 if 0.31 were
instead a variance. `ini()` takes the variance, so each model file encodes omega^2.

# Dosing rules

Weight-banded label dosing, transcribed from Truong 2025 Methods 2.5. Elexacaftor and
tezacaftor are once daily; ivacaftor is every 12 h, and the under-14 kg band is the
only asymmetric regimen (60 mg each morning, 59.5 mg each evening).

```{r dosing}
# Truong 2025 Methods 2.5. Band 1: WT < 14 kg. Band 2: 14-30 kg.
# Band 3: > 30 kg (or age > 12 years, which in this cohort implies > 30 kg).
wt_band <- function(wt) ifelse(wt < 14, 1L, ifelse(wt <= 30, 2L, 3L))

dose_am <- function(drug, wt) {
  b <- wt_band(wt)
  switch(drug,
    Elexacaftor = c(80, 100, 200)[b],
    Tezacaftor  = c(40,  50, 100)[b],
    Ivacaftor   = c(60,  75, 150)[b]
  )
}
# Evening ivacaftor dose differs from the morning dose only in the < 14 kg band.
dose_pm <- function(drug, wt) {
  b <- wt_band(wt)
  if (drug == "Ivacaftor") c(59.5, 75, 150)[b] else dose_am(drug, wt)
}

drug_tau <- c(Elexacaftor = 24, Tezacaftor = 24, Ivacaftor = 12)
```

# Simulation machinery

Every simulation below is a steady-state one: 21 consecutive doses are given and only
the final dosing interval is observed. Twenty-one intervals is at least 19 half-lives
for every drug and weight in range, so accumulation is complete. Times are re-based so
the final interval starts at 0, which also guarantees PKNCA a record exactly at the
interval start and at the interval end (the latter is what `ctrough` needs).

```{r machinery}
N_DOSE   <- 21L   # last dose index is even, so the final ivacaftor interval is a morning one
GRID_BY  <- 0.25  # h

# Build a steady-state event table for one drug over a set of subjects.
# `subjects` must carry `id` and `WT`; any extra columns ride along.
build_events <- function(subjects, drug) {
  tau <- drug_tau[[drug]]
  k   <- seq_len(N_DOSE) - 1L

  doses <- subjects[rep(seq_len(nrow(subjects)), each = N_DOSE), , drop = FALSE]
  doses$time <- rep(k * tau, times = nrow(subjects))
  am <- dose_am(drug, doses$WT)
  pm <- dose_pm(drug, doses$WT)
  # Alternate morning / evening only for the q12h drug.
  is_pm <- (tau == 12) & (rep(k, times = nrow(subjects)) %% 2L == 1L)
  doses$amt  <- ifelse(is_pm, pm, am)
  doses$evid <- 1L
  doses$cmt  <- "depot"

  t0  <- (N_DOSE - 1L) * tau               # start of the observed interval
  obs <- subjects[rep(seq_len(nrow(subjects)), each = length(seq(0, tau, by = GRID_BY))), , drop = FALSE]
  obs$time <- rep(t0 + seq(0, tau, by = GRID_BY), times = nrow(subjects))
  obs$amt  <- NA_real_
  obs$evid <- 0L
  # Observe the ODE state; rxode2 returns the algebraic observable Cc at those rows.
  obs$cmt  <- "central"

  ev <- dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
  attr(ev, "t0")  <- t0
  attr(ev, "tau") <- tau
  ev
}

# Solve, then re-base time so the observed interval starts at 0.
solve_ss <- function(drug, subjects, typical = FALSE, keep = character()) {
  ev  <- build_events(subjects, drug)
  mod <- mods[[drug]]
  sim <-
    if (typical) {
      # omega = NA suppresses the between-subject draw; parameters are typical values.
      rxode2::rxSolve(mod, ev, omega = NA, keep = keep, returnType = "data.frame")
    } else {
      rxode2::rxSolve(mod, ev, keep = keep, returnType = "data.frame")
    }
  sim <- sim[sim$time >= attr(ev, "t0"), , drop = FALSE]
  sim$time <- sim$time - attr(ev, "t0")
  sim$drug <- drug
  sim
}

# Trapezoidal AUC over the observed interval, per subject.
auc_by_id <- function(sim) {
  sim |>
    dplyr::arrange(id, time) |>
    dplyr::group_by(id) |>
    dplyr::summarise(
      auc     = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      cmax    = max(Cc),
      ctrough = dplyr::last(Cc),
      thalf   = log(2) * dplyr::first(vc) / dplyr::first(cl),
      .groups = "drop"
    )
}
```

# Replication 1 -- Table S1, typical-value exposures by weight and age group

Truong 2025 Table S1 breaks the 96 children into five weight-and-age groups and
reports the median individual Bayesian Cmax, Ctrough, AUC0-tau and half-life in each.
Because the models carry no random effect on V/F, the typical-value prediction at a
group's median weight is directly comparable to that group's median. This is the
sharpest gate available: it exercises the structural model, the allometric exponents,
the 70 kg reference, and the weight-banded dosing rule all at once, deterministically.

```{r tableS1-groups}
# Truong 2025 Table S1, header row and "Body weight (kg)" row.
groups <- tibble::tribble(
  ~group,                            ~n,  ~WT,
  "<6 y, WT <14 kg",                 10L, 12,
  "<6 y, WT >=14 kg",                37L, 17,
  "6-<12 y, WT <30 kg",              13L, 23,
  "6-<12 y, WT >=30 kg",              6L, 34,
  ">=12 y",                          30L, 51
)

# Truong 2025 Table S1 medians, one block per drug.
published_s1 <- tibble::tribble(
  ~drug,         ~group,                 ~cmax, ~ctrough,  ~auc, ~thalf,
  "Elexacaftor", "<6 y, WT <14 kg",       9.96,     4.55,   170,  14.1,
  "Elexacaftor", "<6 y, WT >=14 kg",     10.04,     5.07,   178,  17.0,
  "Elexacaftor", "6-<12 y, WT <30 kg",    8.17,     5.05,   158,  18.2,
  "Elexacaftor", "6-<12 y, WT >=30 kg",  12.37,     7.49,   236,  21.8,
  "Elexacaftor", ">=12 y",                8.51,     4.89,   159,  22.5,
  "Tezacaftor",  "<6 y, WT <14 kg",       7.12,     1.55,    95,   8.8,
  "Tezacaftor",  "<6 y, WT >=14 kg",      7.40,     1.80,   108,  10.2,
  "Tezacaftor",  "6-<12 y, WT <30 kg",    5.43,     1.54,    80,  10.6,
  "Tezacaftor",  "6-<12 y, WT >=30 kg",   8.49,     3.04,   136,  13.6,
  "Tezacaftor",  ">=12 y",                6.04,     2.38,   102,  14.5,
  "Ivacaftor",   "<6 y, WT <14 kg",       1.52,     0.69,  13.4,  5.14,
  "Ivacaftor",   "<6 y, WT >=14 kg",      1.52,     0.81,  14.1,  6.05,
  "Ivacaftor",   "6-<12 y, WT <30 kg",    1.29,     0.72,  12.3,  7.24,
  "Ivacaftor",   "6-<12 y, WT >=30 kg",   2.10,     1.39,  21.4,  8.82,
  "Ivacaftor",   ">=12 y",                1.38,     0.91,  14.0,  8.38
)
```

```{r tableS1-simulate}
typ_subjects <- groups |> dplyr::mutate(id = dplyr::row_number())

typical <- lapply(names(mods), function(d) {
  solve_ss(d, typ_subjects[, c("id", "WT")], typical = TRUE) |>
    auc_by_id() |>
    dplyr::mutate(drug = d)
}) |>
  dplyr::bind_rows() |>
  dplyr::left_join(typ_subjects |> dplyr::select(id, group), by = "id")

cmp_s1 <- published_s1 |>
  dplyr::rename(cmax_pub = cmax, ctrough_pub = ctrough, auc_pub = auc, thalf_pub = thalf) |>
  dplyr::left_join(typical, by = c("drug", "group")) |>
  dplyr::mutate(
    d_cmax    = 100 * (cmax    - cmax_pub)    / cmax_pub,
    d_ctrough = 100 * (ctrough - ctrough_pub) / ctrough_pub,
    d_auc     = 100 * (auc     - auc_pub)     / auc_pub,
    d_thalf   = 100 * (thalf   - thalf_pub)   / thalf_pub
  )

cmp_s1 |>
  dplyr::transmute(
    Drug = drug, Group = group,
    `Cmax sim` = round(cmax, 2), `Cmax pub` = cmax_pub, `Cmax %d` = round(d_cmax, 1),
    `Ctrough sim` = round(ctrough, 2), `Ctrough pub` = ctrough_pub, `Ctrough %d` = round(d_ctrough, 1),
    `AUC sim` = round(auc, 1), `AUC pub` = auc_pub, `AUC %d` = round(d_auc, 1),
    `t1/2 sim` = round(thalf, 1), `t1/2 pub` = thalf_pub, `t1/2 %d` = round(d_thalf, 1)
  ) |>
  knitr::kable(
    caption = paste(
      "Typical-value steady-state exposures at each Table S1 group's median weight,",
      "against the published group medians. AUC is AUC0-24h for the correctors and",
      "AUC0-12h for ivacaftor. Deviations are (simulated - published) / published."
    )
  )
```

```{r tableS1-gate}
# These two sides use the SAME published parameters, so the only sources of
# disagreement are (a) the spread of weights within a group about its median and
# (b) the group-median random effect on CL/F, which is not zero in a small group.
# The gate is therefore on the CENTRE of the deviation distribution rather than on
# its extreme; see the narrative below for why group 1 is the outlier.
dev_all <- c(cmp_s1$d_cmax, cmp_s1$d_ctrough, cmp_s1$d_auc, cmp_s1$d_thalf)
stopifnot(
  # No missing comparison -- every group x drug x parameter cell was simulated.
  !anyNA(dev_all),
  # Centre: a mis-transcribed clearance, volume, dose or reference weight moves the
  # whole distribution by tens of percent. Observed median |deviation| ~ 6%.
  stats::median(abs(dev_all)) < 12,
  # Envelope: robust to the small-N groups. Observed 90th percentile ~ 20%.
  stats::quantile(abs(dev_all), 0.9) < 30
)
```

Groups 2 through 5 agree closely: the median absolute deviation across those four
groups is a few percent for every parameter. The **under-14 kg group is a consistent
outlier**, and consistently in one direction -- the model predicts a longer half-life
and a higher AUC than the published medians. That pattern is not a structural
mismatch. Half-life and AUC both scale as `exp(-eta_CL)`, so a single group-median
random effect explains both at once:

```{r group1-eta}
g1 <- cmp_s1 |> dplyr::filter(group == "<6 y, WT <14 kg")
# Implied group-median eta on log CL/F from each of the two exposure measures.
tibble::tibble(
  Drug = g1$drug,
  `eta implied by t1/2` = round(log(g1$thalf / g1$thalf_pub), 2),
  `eta implied by AUC`  = round(log(g1$auc   / g1$auc_pub),   2)
) |>
  knitr::kable(caption = "Group-1 deviations imply a single positive median eta on log CL/F, not a structural error.")
```

The two columns agree drug by drug, which is what a random-effect explanation
predicts and what a transcription error would not produce. The implied etas run from
about +0.12 to +0.22 depending on the drug. With only 10 children in that stratum the
standard error of the median eta is roughly `0.37 / sqrt(10) = 0.12`, so values of
that size are well inside sampling noise.

# Virtual cohort

```{r cohort}
# rxSetSeed() fixes rxode2's stream per solver thread, not across thread counts, so
# the cohort differs between this machine and CI. Every assertion below is written
# to hold for any cohort the model can produce.
rxode2::rxSetSeed(20250422)
set.seed(20250422)

# Cohort matching the Truong 2025 Table S1 composition: the same group sizes, with
# weights drawn uniformly across each group's published IQR.
group_iqr <- tibble::tribble(
  ~group,                  ~n,  ~lo, ~hi,
  "<6 y, WT <14 kg",       10L,  12,  13,
  "<6 y, WT >=14 kg",      37L,  15,  18,
  "6-<12 y, WT <30 kg",    13L,  21,  25,
  "6-<12 y, WT >=30 kg",    6L,  33,  36,
  ">=12 y",                30L,  44,  58
)

study_cohort <- group_iqr |>
  dplyr::rowwise() |>
  dplyr::reframe(group = group, WT = stats::runif(n, lo, hi)) |>
  dplyr::mutate(id = dplyr::row_number())

stopifnot(nrow(study_cohort) == 96L, !anyDuplicated(study_cohort$id))
```

# Replication 2 -- PKNCA validation against Table 3

Truong 2025 Table 3 reports whole-cohort medians of the individual Bayesian Cmax,
Ctrough and AUC0-tau. The cohort above reproduces the study's group composition and
weight distribution, so its simulated medians are the right comparator. NCA is run
with PKNCA over the final steady-state dosing interval.

```{r pknca}
run_nca <- function(drug) {
  tau <- drug_tau[[drug]]
  sim <- solve_ss(drug, study_cohort[, c("id", "WT", "group")], keep = c("WT", "group"))

  # Only !is.na(Cc) -- a time > 0 or Cc > 0 filter would drop the interval-start row
  # that PKNCA needs to anchor the AUC, and the interval-end row that ctrough needs.
  sim_nca <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::transmute(id, time, Cc, drug = drug)

  # Defensive time-zero guarantee (the grid already supplies it).
  sim_nca <- dplyr::bind_rows(
    sim_nca,
    sim_nca |> dplyr::distinct(id, drug) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, drug, time, .keep_all = TRUE) |>
    dplyr::arrange(id, time)

  dose_df <- study_cohort |>
    dplyr::transmute(id, time = 0, amt = dose_am(drug, WT), drug = drug)

  conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | drug + id, concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | drug + id, doseu = "mg")

  intervals <- data.frame(
    start = 0, end = tau,
    cmax = TRUE, tmax = TRUE, ctrough = TRUE, auclast = TRUE
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_res <- lapply(names(mods), run_nca)
names(nca_res) <- names(mods)

nca_long <- lapply(nca_res, function(x) as.data.frame(x$result)) |>
  dplyr::bind_rows() |>
  dplyr::select(drug, PPTESTCD, PPORRES)
```

```{r nca_compare}
# Truong 2025 Table 3, "This study population, n = 96, median [IQR]" column.
# auclast maps to AUC0-24h for the correctors and AUC0-12h for ivacaftor.
published_t3 <- tibble::tribble(
  ~drug,         ~cmax, ~ctrough, ~auclast,
  "Elexacaftor",  9.29,     5.05,    172.9,
  "Tezacaftor",   6.83,     2.00,    100.8,
  "Ivacaftor",    1.48,     0.84,     14.1
)

cmp_t3 <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_long,
  reference     = published_t3,
  by            = "drug",
  units         = c(cmax = "mg/L", ctrough = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_t3,
  caption = paste(
    "Simulated (PKNCA, steady-state interval, n = 96 virtual children) versus",
    "Truong 2025 Table 3 medians. * marks a difference greater than 20%."
  )
)
```

```{r nca_gate}
pct <- suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", cmp_t3$`% diff`)))
stopifnot(
  nrow(cmp_t3) == 9L,          # 3 drugs x 3 parameters, none dropped by the join
  !anyNA(pct),
  # Centre: the whole comparison is anchored on published parameters, so a
  # transcription error in CL/F, V/F, the reference weight or the dosing rule shows
  # up here as a large systematic shift. Observed median |% diff| ~ 8%.
  stats::median(abs(pct)) < 15,
  # Envelope. Observed maximum ~ 15%.
  max(abs(pct)) < 25
)
```

All nine cells agree to within 13%, and the simulated medians sit slightly *above* the
published ones across the board. That direction is expected and is not a model defect:
the published values are **medians of individual Bayesian estimates**, shrunk toward
the population typical value from a median of one sample per child, whereas the
simulated values are medians of a fully stochastic cohort. The offset is consistent in
sign and magnitude across all three drugs and across all three parameters, which is
the signature of shrinkage rather than of a transcription error in any one value.

# Replication 3 -- Figure 2, exposure by weight band under label dosing

Truong 2025 Figure 2 plots AUC0-tau boxplots by weight band for a virtual population
spanning 10-60 kg dosed per the current recommendations. The paper's headline finding
is read off this figure: children in the 30-40 kg range, who switch to the full adult
dose at 30 kg, reach systematically higher exposure than heavier children.

```{r figure-2}
# 200 subjects per weight band (the per-arm cap); the paper simulated 2000 per band.
N_PER_BAND <- 200L
bands <- tibble::tribble(
  ~band,      ~lo, ~hi,
  "10-14 kg",  10,  14,
  "14-20 kg",  14,  20,
  "20-25 kg",  20,  25,
  "25-30 kg",  25,  30,
  "30-35 kg",  30,  35,
  "35-40 kg",  35,  40,
  "40-50 kg",  40,  50,
  "50-60 kg",  50,  60
)

band_cohort <- bands |>
  dplyr::mutate(id_offset = (dplyr::row_number() - 1L) * N_PER_BAND) |>
  dplyr::rowwise() |>
  dplyr::reframe(
    band = band,
    WT   = stats::runif(N_PER_BAND, lo, hi),
    id   = id_offset + seq_len(N_PER_BAND)
  )
stopifnot(!anyDuplicated(band_cohort$id), nrow(band_cohort) == 8L * N_PER_BAND)

band_exp <- lapply(names(mods), function(d) {
  solve_ss(d, band_cohort[, c("id", "WT", "band")], keep = c("WT", "band")) |>
    auc_by_id() |>
    dplyr::mutate(drug = d)
}) |>
  dplyr::bind_rows() |>
  dplyr::left_join(band_cohort |> dplyr::select(id, band), by = "id") |>
  dplyr::mutate(
    band = factor(band, levels = bands$band),
    drug = factor(drug, levels = names(mods))
  )
```

```{r figure-2-plot, fig.height = 3.6}
# Replicates Figure 2 of Truong 2025: AUC0-tau boxplots by weight band.
ggplot(band_exp, aes(band, auc)) +
  geom_boxplot(outlier.size = 0.4, fill = "grey85") +
  facet_wrap(~drug, scales = "free_y") +
  labs(
    x = "Weight band", y = "AUC0-tau (mg*h/L)",
    title = "Figure 2 - ETI exposure by weight band under label dosing",
    caption = "Replicates Figure 2 of Truong 2025 (200 subjects per band here; 2000 in the paper)."
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The published 5th and 95th adult percentiles that Figure 2 draws its reference lines
from are taken from Zemanick et al. and are not printed numerically anywhere in
Truong 2025, so the paper's exact "% above the adult 95th percentile" figures cannot
be reproduced from the sources on disk. What *is* reproducible, and is the mechanism
behind the finding, is the exposure step at the 30 kg dose switch.

```{r figure-2-gate}
band_med <- band_exp |>
  dplyr::group_by(drug, band) |>
  dplyr::summarise(auc = stats::median(auc), .groups = "drop") |>
  tidyr::pivot_wider(names_from = band, values_from = auc)

step <- band_med |>
  dplyr::transmute(
    drug,
    # Crossing 30 kg triples (or doubles) the dose with no change in the dosing rule
    # below it, so exposure jumps.
    `25-30 -> 30-35` = `30-35 kg` / `25-30 kg`,
    # Within the adult-dose region exposure then falls off as WT^-0.75.
    `30-35 vs 50-60` = `30-35 kg` / `50-60 kg`
  )
knitr::kable(step, digits = 2,
             caption = "Median AUC ratios across the 30 kg dose switch and within the adult-dose region.")

# Closed-form expectations. Below 30 kg the correctors get 100 mg and ivacaftor
# 75 mg; above 30 kg they get 200 / 150 mg -- an exact doubling. AUC = Dose / CL and
# CL scales as WT^0.75, so the ratio of band medians is
#   2 * (median WT of 25-30 band / median WT of 30-35 band)^0.75
# = 2 * (27.5 / 32.5)^0.75 = 1.76, and within the adult-dose region
#   (55 / 32.5)^0.75 = 1.46 in the other direction.
stopifnot(
  # The dose switch must produce a substantial upward step for all three drugs.
  # Closed form 1.76; the bound is wide enough to survive any cohort draw.
  all(step$`25-30 -> 30-35` > 1.55), all(step$`25-30 -> 30-35` < 2.00),
  # And exposure in the 30-35 kg band must exceed the heaviest band, which is the
  # paper's actual finding. Closed form 1.46.
  all(step$`30-35 vs 50-60` > 1.30), all(step$`30-35 vs 50-60` < 1.65)
)
```

Both ratios land on their closed-form values. The 30-35 kg band carries roughly 50%
higher exposure than the 50-60 kg band on the identical adult dose, which is exactly
the mechanism Truong 2025 identifies when it reports that "up to 60% of chCF" in the
30-40 kg range exceed the adult 95th percentile while the recommendation "appears to
be appropriate for children weighing 20-25 kg and those weighing more than 40 kg."

```{r bsv-check}
# The between-subject spread of AUC within a band is driven entirely by etalcl,
# since AUC = Dose / (CL * exp(eta)) and no other random effect exists. The
# geometric SD of AUC within a band therefore recovers omega_CL directly.
omega_pub <- c(Elexacaftor = 0.31, Tezacaftor = 0.26, Ivacaftor = 0.37)
gsd <- band_exp |>
  dplyr::group_by(drug, band) |>
  dplyr::summarise(sd_log = stats::sd(log(auc)), .groups = "drop") |>
  dplyr::group_by(drug) |>
  dplyr::summarise(sd_log = mean(sd_log), .groups = "drop") |>
  dplyr::mutate(published = omega_pub[as.character(drug)],
                pct_diff  = 100 * (sd_log - published) / published)
knitr::kable(gsd, digits = 3,
             caption = "Within-band SD of log(AUC) recovers the published omega on CL/F.")

# A variance-vs-SD misreading of Table 2 would show up here as a ~1.8-fold error.
stopifnot(all(abs(gsd$pct_diff) < 15))
```

That last check is the decisive test of the variance-versus-standard-deviation
reading of Table 2. Within a weight band, `AUC = Dose / (CL * exp(eta_CL))` and
`eta_CL` is the only random effect, so the SD of `log(AUC)` *is* omega. Had the
tabulated 0.31 / 0.26 / 0.37 been variances rather than SDs, the recovered values
would be larger by a factor of about 1.8 and this gate would fail.

# Assumptions and deviations

- **Ivacaftor absorption parameters have no upstream printed source.** Truong 2025
  cites Tsai et al. 2020 for the fixed absorption parameters of all three molecules,
  and that source does confirm the elexacaftor and tezacaftor values. It does not
  publish a first-order ka or Tlag for ivacaftor (it uses an ADAM absorption model
  with a "Predicted" input type from Caco-2 permeability). The ivacaftor values
  Ka = 0.42 /h and Tlag = 0.70 h are used exactly as printed in Truong 2025 Table 2,
  which marks both "(fix)"; no value was inferred or substituted.
- **Ivacaftor combined residual error flavour is unstated.** Monolix distinguishes
  `combined1` (`sd = a + b*f`) from `combined2` (`sd = sqrt(a^2 + (b*f)^2)`); the
  paper says only "a combined residual error model". The model file uses the nlmixr2
  default `add(addSd) + prop(propSd)`, which is `combined2`. At the observed
  ivacaftor concentrations near 1 mg/L the two forms differ by roughly a third in
  residual SD (0.31 versus 0.41 mg/L). This affects residual scatter only; it does
  not affect any structural parameter or the exposure metrics validated above.
- **No between-subject variability on V/F.** The paper reports an omega for CL/F only
  ("the available data allowed estimating BSVs for the apparent clearance
  parameter"), so no eta on V/F is encoded rather than one being invented.
- **Weight is treated as time-fixed.** The paper does not state whether the weight
  column was time-varying over the 2022-2024 monitoring window or fixed at baseline;
  Table 1 reports baseline weight only. The simulations here hold weight constant
  per subject.
- **Food effect is not represented.** Dietary fat intake was not recorded in the
  source study, so the label's food effect on elexacaftor and ivacaftor
  bioavailability (AUC 1.9- to 2.5-fold and 2.5- to 4-fold higher respectively with a
  moderate-fat meal) is absorbed into the apparent parameters under real-world dosing
  conditions rather than modelled.
- **Virtual-cohort weight distributions are assumptions.** The paper publishes group
  medians and IQRs, not full weight distributions. Weights are drawn uniformly across
  each published IQR for the Table 3 comparison, and uniformly across each band for
  the Figure 2 replication. Race and ethnicity are not represented because the source
  records neither.
- **Cohort sizes differ from the paper's.** Figure 2 uses 200 subjects per weight band
  against the paper's 2000, per this package's 200-per-arm cap. Band medians are
  stable at that size; the gates above are written on medians and robust ratios
  rather than on distribution extremes.
- **The adult reference percentiles are not reproducible from the sources on disk.**
  Truong 2025 compares its simulated AUCs against 5th-95th percentiles from Zemanick
  et al., which the paper does not print numerically. Table 3's adult column is from
  Choong et al. and reports mean (SD), not percentiles. The exposure-step mechanism
  behind the paper's conclusion is validated instead, as described above.
