---
title: "Ritonavir as a protease-inhibitor booster in children (Tsirizani 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ritonavir as a protease-inhibitor booster in children (Tsirizani 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20250926)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Tsirizani_2025_ritonavir"))$reference`
* Description: `r rxode2::rxode(readModelDb("Tsirizani_2025_ritonavir"))$description`
* Article: <https://doi.org/10.1128/aac.00771-25>

Ritonavir is no longer used as an antiretroviral in its own right. At the low
doses used here it is a pharmacokinetic booster: it inhibits gut and hepatic
CYP3A4 and P-glycoprotein, so a companion protease inhibitor (PI) can be given
at a lower dose and less often. Boosted PIs carry much of second- and
third-line paediatric antiretroviral therapy, yet ritonavir's own PK in
children was barely described before this analysis.

Tsirizani 2025 is the PK sub-study of CHAPAS-4 (ISRCTN22964075), a 4x2
randomised trial in Zambia, Uganda and Zimbabwe. The design is what makes the
paper unusual: 170 children were randomised to twice-daily
lopinavir/ritonavir, once-daily atazanavir/ritonavir or once-daily
darunavir/ritonavir, with identical sampling schedules, assay, sites and food
intake across arms. That gives a within-study comparison of ritonavir exposure
across three companion PIs -- something no previous study had, because there is
no ritonavir-alone control arm anywhere in the paediatric literature.

The consequence for the model is that **relative bioavailability is fixed to 1
on the darunavir arm**, so every clearance and volume below is apparent
(`CL/F`, `Vc/F`, `Q/F`, `Vp/F`) *on that reference*, and the companion-PI
effects are read as relative bioavailability. The authors are explicit that
they cannot rule out darunavir or lopinavir themselves depressing ritonavir
bioavailability rather than atazanavir raising it.

```{r load-model}
mod <- readModelDb("Tsirizani_2025_ritonavir")
mod
```

# Population

170 children were enrolled between January 2019 and March 2021 into the three
ritonavir-boosted arms (lopinavir N = 51, atazanavir N = 60, darunavir
N = 59). Median age was 10.5 years (range 3.16-15.6), median weight 26.0 kg
(14.2-64.2) and median height 131 cm (97.0-169); 51.2% were female
(Tsirizani 2025 Table 1). The cohort is substantially stunted and underweight
relative to WHO references -- median weight-for-age Z-score -1.4 (-4.5 to 1.7)
and height-for-age Z-score -1.3 (-4.4 to 3.6) -- which is worth keeping in mind
when reading a fat-free-mass-scaled model.

All children were failing first-line ART by WHO virological, CD4 or clinical
criteria and were starting a boosted PI as second-line therapy, with an NRTI
backbone of tenofovir alafenamide/emtricitabine (54.7%), abacavir/lamivudine
(25.3%) or zidovudine/lamivudine (20.0%). No NRTI effect on ritonavir PK was
found.

Intensive sampling was done after week 6 of study treatment, at pre-dose, 0.5 h
(TAF/FTC arms only), 1, 2, 4, 6, 8, 12 and 24 h post-dose. Study drugs were
given under direct observation with a low-fat breakfast (5% fat, ~250 kCal), and
all non-ART co-medications were deferred at least 2 hours. Of 1,254 ritonavir
concentrations, 6.9% were below the 0.045 mg/L LLOQ and 3.7% undetectable; these
were handled by a modified Beal M6 method. One profile was excluded for
implausibly low concentrations throughout, and all 24 h samples in the lopinavir
arm were excluded because the 12 h dosing times were insufficiently documented.

```{r population}
str(readModelDb("Tsirizani_2025_ritonavir")()$population)
```

# Source trace

Every `ini()` value and every non-obvious `model()` construct, with the place it
came from. Table 3 typical values apply to the darunavir arm at FFM 21.0 kg on a
witnessed morning dose. "Data S1" is the supplemental NONMEM control stream
(`AAC00771-25-s0001.docx`), retrieved from the Europe PMC supplementary-files
endpoint for PMC12587601.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity,                        ~Value,       ~Source,
  "CL/F",                           "10.5 L/h",   "Table 3 'Clearance (L/h)' 10.5 (9.22-11.9)",
  "Vc/F",                           "54.5 L",     "Table 3 'Central volume of distribution (L)' 54.5 (48.9-60.4)",
  "Q/F",                            "1.19 L/h",   "Table 3 'Intercompartmental clearance (L/h)' 1.19 (0.851-1.58)",
  "Vp/F",                           "134 L",      "Table 3 'Peripheral volume of distribution (L)' 134 (38.8-330)",
  "Relative bioavailability",       "1 FIXED",    "Table 3 'Bioavailability'; Data S1 '$THETA 1 FIX ; 4 BIO'",
  "Absorption lag",                 "0.981 h",    "Table 3 'Lag time (h)' 0.981 (0.934-1.06)",
  "Zero-order duration D1",         "2.45 h",     "Table 3 'Zero-order absorption duration (h)' 2.45 (2.05-2.80)",
  "ka",                             "1.16 1/h",   "Table 3 'First-order absorption rate constant (1/h)' 1.16 (0.893-1.89)",
  "FFM exponent on CL/F, Q/F",      "0.75 fixed", "Methods 'allometric scaling ... with a fixed exponent of 0.75 and 1'; Data S1 'ALLMCL_FFM = (FFM/TVFFM)**0.75'",
  "FFM exponent on Vc/F, Vp/F",     "1 fixed",    "Methods, same sentence; Data S1 'ALLMV_FFM = (FFM/TVFFM)'",
  "FFM reference",                  "21.0 kg",    "Data S1 '$PK: TVFFM = 21.0 ;MEDIAN'; Table 3 footnote b ties it to a 26 kg child",
  "Atazanavir on rel. F",           "+137%",      "Table 3 'Atazanavir on relative bioavailability (%)' 137 (107-190)",
  "Lopinavir on rel. F",            "-23.4%",     "Table 3 'Lopinavir on relative bioavailability (%)' -23.4 (-8.20 to -34.4)",
  "Atazanavir on CL/F",             "+20.7%",     "Table 3 'Atazanavir on clearance (%)' +20.7 (+11.3 to +31.3)",
  "Evening-dose lag multiplier",    "3.61 fold",  "Table 3 'Night dose additional lag time (Fold)' 3.61 (2.57-4.56); Data S1 gates it on 'PI_BCK_BONE.EQ.3.AND.OCC.EQ.1'",
  "Occasion-1 F BOV SD inflation",  "2.14 fold",  "Table 3 'Extra variability for unobserved doses' 2.14 (1.53-2.42) + footnote c; Data S1 'IF(OCC.EQ.1)BOVBIO = ETA(9)*EBOV'",
  "BSV on CL",                      "13.1%",      "Table 3 'Between-subject variability / Clearance (%)' 13.1 (10.2-15.8)",
  "BOV on rel. F",                  "40.1%",      "Table 3 'Between-occasion variability / Bioavailability (%)' 40.1 (35.2-46.0)",
  "BOV on lag",                     "62.0%",      "Table 3 'Lag time (%)' 62.0 (54.5-70.2)",
  "BOV on D1",                      "52.5%",      "Table 3 'Zero-order rate of absorption (%)' 52.5 (42.7-63.0)",
  "BOV on ka",                      "114%",       "Table 3 'First-order absorption rate constant (%)' 114 (90.0-141)",
  "Proportional error",             "14.7%",      "Table 3 'Proportional error (%)' 14.7 (13.4-16.1)",
  "Additive error",                 "0.0152 mg/L","Table 3 'Additive error (mg/L)' 0.0152 (0.0121-0.0184); Data S1 'ADD = THETA(6)+(LLOQ*0.2)'",
  "Two-compartment disposition",    "structure",  "Results: dOFV = -110 vs one compartment; Data S1 '$SUBROUTINE ADVAN4 TRANS1'",
  "Lag + zero- then first-order",   "structure",  "Results: dOFV = -488 vs no absorption delay; Data S1 'ALAG1=LAG', 'D1=TVD1*EXP(BOVD1)'",
  "Dosing by WHO weight band",      "50-200 mg/d","Table 2",
  "Trial-arm sizes and band counts","Table 1",    "Table 1"
) |>
  knitr::kable(caption = "Source trace for Tsirizani_2025_ritonavir.")
```

## How the variability percentages were read

Table 3 reports every variance component as a percentage. Two readings are
possible for a log-normal random effect: the raw log-scale standard deviation
(`omega`), or the log-normal coefficient of variation
`sqrt(exp(omega^2) - 1)`. The supplemental `$OMEGA` initial estimates settle it
decisively on the first reading, and the BOV on `ka` is the discriminating row.

```{r omega-convention, echo = FALSE}
tibble::tribble(
  ~Component,       ~`Data S1 initial omega^2`, ~`Table 3 (%)`,
  "BSV CL",         0.0154914,                  13.1,
  "BOV rel. F",     0.159089,                   40.1,
  "BOV ka",         1.25373,                    114,
  "BOV lag",        0.372121,                   62.0,
  "BOV D1",         0.289698,                   52.5
) |>
  dplyr::mutate(
    `sqrt(omega^2) (%)`            = round(100 * sqrt(`Data S1 initial omega^2`), 1),
    `sqrt(exp(omega^2)-1) (%)`     = round(100 * sqrt(exp(`Data S1 initial omega^2`) - 1), 1)
  ) |>
  knitr::kable(
    caption = paste(
      "Table 3's percentages are the raw log-scale omega SD, not the",
      "log-normal CV. The BOV on ka is decisive: 114% printed against",
      "112% for sqrt(omega^2) and 158% for the CV reading. The Data S1",
      "values are initial estimates, so agreement to a few percent is",
      "the most that can be expected."
    )
  )
```

Accordingly the model encodes `omega^2` as the square of Table 3's percentage
divided by 100 -- for example `etalcl ~ 0.131^2 = 0.017161`.

The five between-subject random effects that the supplement fixes to zero
(`$OMEGA BLOCK(1) FIX 0` on Vc, ka, F, Vp and Q, plus BOV on CL) are omitted
rather than written as `~ fixed(0)`. Table 3 lists no such rows, and adding
zero-variance etas would make OMEGA singular.

# Virtual cohort

The cohort reproduces the CHAPAS-4 design: three arms, WHO weight bands, and the
Table 2 ritonavir dose for each arm and band.

One deliberate departure: the **same weight distribution is used in all three
arms**, drawn from Table 1's *pooled* weight-band counts (35 / 45 / 48 / 42 for
14-19.9, 20-24.9, 25-34.9 and 35+ kg) rather than each arm's own counts. Table 1
reports a median weight of 26.0 kg in every arm, and uniform sampling within
bands under the pooled counts reproduces that (the pooled median falls at the
5th-6th of the 48 children in the 25-34.9 kg band, i.e. about 26.0 kg). Using
each arm's own band counts instead makes the arm medians drift apart -- 25.0 kg
for darunavir against 27.1 kg for atazanavir on a first pass -- which biases the
two arms in *opposite* directions through the allometric term and confounds the
companion-PI comparison with cohort composition. Weight-matching the arms keeps
the size covariate as a common random number across arms, so any arm-to-arm
difference below is attributable to the model rather than to who was sampled.

Each arm therefore holds 170 children, inside the 200-per-arm cap.

```{r cohort-spec}
# Table 2: ritonavir dose per administration, by arm and WHO weight band.
# Lopinavir/ritonavir is the only twice-daily arm.
regimen <- tibble::tribble(
  ~arm,           ~band,        ~am_mg, ~pm_mg, ~daily_mg,
  "Lopinavir",    "14-19.9 kg",  50,     50,     100,
  "Lopinavir",    "20-24.9 kg",  50,     50,     100,
  "Lopinavir",    "25-34.9 kg", 100,     50,     150,
  "Lopinavir",    "35+ kg",     100,    100,     200,
  "Atazanavir",   "14-19.9 kg",  75,      0,      75,
  "Atazanavir",   "20-24.9 kg",  75,      0,      75,
  "Atazanavir",   "25-34.9 kg", 100,      0,     100,
  "Atazanavir",   "35+ kg",     100,      0,     100,
  "Darunavir",    "14-19.9 kg", 100,      0,     100,
  "Darunavir",    "20-24.9 kg", 100,      0,     100,
  "Darunavir",    "25-34.9 kg", 100,      0,     100,
  "Darunavir",    "35+ kg",     100,      0,     100
)

# Table 1 POOLED weight-band counts (all 170 children), applied to every arm.
band_n <- tibble::tribble(
  ~band,        ~n,
  "14-19.9 kg", 35L,
  "20-24.9 kg", 45L,
  "25-34.9 kg", 48L,
  "35+ kg",     42L
)

band_wt <- tibble::tribble(
  ~band,        ~wt_lo, ~wt_hi,
  "14-19.9 kg", 14.2,   19.9,
  "20-24.9 kg", 20.0,   24.9,
  "25-34.9 kg", 25.0,   34.9,
  "35+ kg",     35.0,   50.0
)

# The paper never reports the equation that produced its FFM data column, so
# FFM is scaled from weight using the paper's own anchor: FFM 21.0 kg (Data S1)
# for a 26 kg child (Table 3 footnote b). See "Assumptions and deviations".
FFM_PER_KG <- 21.0 / 26.0

rxode2::rxSetSeed(20250926)
set.seed(20250926)

# One weight sample, shared by all three arms (weight-matched design).
weights <-
  band_n |>
  dplyr::left_join(band_wt, by = "band") |>
  dplyr::rowwise() |>
  dplyr::mutate(WT = list(stats::runif(n, wt_lo, wt_hi))) |>
  dplyr::ungroup() |>
  tidyr::unnest(WT) |>
  dplyr::mutate(subject = dplyr::row_number()) |>
  dplyr::select(subject, band, WT)

cohort <-
  weights |>
  tidyr::crossing(arm = c("Lopinavir", "Atazanavir", "Darunavir")) |>
  dplyr::left_join(regimen, by = c("arm", "band")) |>
  dplyr::arrange(arm, subject) |>
  dplyr::mutate(
    id                = dplyr::row_number(),
    FFM               = WT * FFM_PER_KG,
    CONMED_LOPINAVIR  = as.integer(arm == "Lopinavir"),
    CONMED_ATAZANAVIR = as.integer(arm == "Atazanavir")
  ) |>
  dplyr::select(id, subject, arm, band, WT, FFM,
                CONMED_LOPINAVIR, CONMED_ATAZANAVIR, am_mg, pm_mg, daily_mg)

cohort |>
  dplyr::count(arm, name = "n_subjects") |>
  knitr::kable(caption = "Simulated cohort size by arm (200-per-arm cap respected).")
```

```{r cohort-demog, echo = FALSE}
cohort |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `Median WT (kg)`  = round(stats::median(WT), 1),
    `Min WT (kg)`     = round(min(WT), 1),
    `Max WT (kg)`     = round(max(WT), 1),
    `Median FFM (kg)` = round(stats::median(FFM), 1),
    .groups = "drop"
  ) |>
  knitr::kable(
    caption = paste(
      "Simulated weight distribution by arm; identical across arms by design.",
      "Compare Tsirizani 2025 Table 1: median weight 26.0 kg in every arm",
      "(range 14.2-64.2 kg overall). The model's allometric reference is",
      "FFM 21.0 kg, so the simulated cohort sits essentially on it."
    )
  )
```

# Simulation

Concentrations were sampled at week 6 of therapy, so the target is steady
state. The terminal half-life of this model is long -- the peripheral
compartment is a deep, low-flux sink (`Q/F` 1.19 L/h against `CL/F` 10.5 L/h,
into `Vp/F` 134 L) -- so a run-in of 35 days is used and only the final 24 h
interval is observed.

```{r halflife, echo = FALSE}
kel <- 10.5 / 54.5
k12 <- 1.19 / 54.5
k21 <- 1.19 / 134
s   <- kel + k12 + k21
p   <- kel * k21
lam1 <- (s + sqrt(s^2 - 4 * p)) / 2
lam2 <- (s - sqrt(s^2 - 4 * p)) / 2
tibble::tibble(
  Phase          = c("Distribution", "Terminal"),
  `Rate (1/h)`   = round(c(lam1, lam2), 5),
  `Half-life (h)`= round(log(2) / c(lam1, lam2), 1)
) |>
  knitr::kable(
    caption = paste0(
      "Eigenvalues of the Table 3 typical two-compartment system at FFM ",
      "21.0 kg. The 35-day run-in reaches ",
      round(100 * (1 - exp(-lam2 * 35 * 24)), 2),
      "% of the terminal-phase steady state."
    )
  )
```

Dose records carry `rate = -2` so `rxode2` honours the modelled zero-order
duration `dur(depot)`; a plain bolus would collapse the zero-order phase and
bias Cmax upward. Observations are placed on `central`, the ODE state, never on
the algebraic observable `Cc`.

`OCC` does double duty in this model: it selects the between-occasion eta *and*
identifies the evening dose in the twice-daily arm. Over a long run-in the
semantically-correct label for every unwitnessed run-in dose would be `OCC = 1`,
but that would also apply the evening-dose absorption lag to run-in *morning*
doses. Absorption lag changes the shape of a dose's input, not the amount
absorbed, so the steady-state AUC over the final interval is identical either
way; occasions are therefore assigned by time of day (morning `OCC = 2`, evening
`OCC = 1`) so each run-in dose gets its correct lag, and the final witnessed
morning dose is `OCC = 2`.

```{r events}
RUN_IN_DAYS <- 35
T0          <- RUN_IN_DAYS * 24   # time of the witnessed morning dose

covs <- c("FFM", "CONMED_LOPINAVIR", "CONMED_ATAZANAVIR")

# Morning doses: every 24 h through the witnessed dose at T0. OCC = 2.
dose_am <-
  cohort |>
  dplyr::select(id, dplyr::all_of(covs), am_mg) |>
  tidyr::crossing(time = seq(0, T0, by = 24)) |>
  dplyr::mutate(amt = am_mg, OCC = 2L) |>
  dplyr::select(-am_mg)

# Evening doses (lopinavir arm only): 12 h after each morning dose, including
# the one inside the observation window at T0 + 12. OCC = 1, so these carry the
# 3.61-fold longer absorption lag and the 2.14-fold inflated bioavailability BOV.
dose_pm <-
  cohort |>
  dplyr::filter(pm_mg > 0) |>
  dplyr::select(id, dplyr::all_of(covs), pm_mg) |>
  tidyr::crossing(time = seq(12, T0 + 12, by = 24)) |>
  dplyr::mutate(amt = pm_mg, OCC = 1L) |>
  dplyr::select(-pm_mg)

doses <-
  dplyr::bind_rows(dose_am, dose_pm) |>
  dplyr::mutate(evid = 1L, cmt = "depot", rate = -2)

# Dense observation grid over the final dosing interval, on the ODE state.
obs <-
  cohort |>
  dplyr::select(id, dplyr::all_of(covs)) |>
  tidyr::crossing(time = T0 + seq(0, 24, by = 0.25)) |>
  dplyr::mutate(amt = 0, evid = 0L, cmt = "central", rate = 0, OCC = 2L)

events <-
  dplyr::bind_rows(doses, obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

nrow(events)
```

```{r simulate}
rxode2::rxSetSeed(20250926)
sim <-
  rxode2::rxSolve(mod, events, addDosing = FALSE) |>
  as.data.frame() |>
  dplyr::mutate(tnca = time - T0) |>
  dplyr::left_join(dplyr::select(cohort, id, arm, band, WT, daily_mg), by = "id")

dplyr::glimpse(dplyr::select(sim, id, arm, band, tnca, Cc, cl, vc, fdepot))
```

# Replicating the published figures

## Figure 1 -- concentration-time profiles by boosted PI

Tsirizani 2025 Figure 1 is a visual predictive check stratified by companion PI,
showing the 5th, 50th and 95th percentiles. The lopinavir panel spans 0-12 h
because the 24 h samples in that arm were excluded; the once-daily arms span
0-24 h.

```{r fig1, fig.height = 4}
vpc_bands <-
  sim |>
  dplyr::filter(!(arm == "Lopinavir" & tnca > 12)) |>
  dplyr::group_by(arm, tnca) |>
  dplyr::summarise(
    p05 = stats::quantile(Cc, 0.05),
    p50 = stats::median(Cc),
    p95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot2::ggplot(vpc_bands, ggplot2::aes(x = tnca)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p05, ymax = p95), alpha = 0.25,
                       fill = "steelblue") +
  ggplot2::geom_line(ggplot2::aes(y = p50), linewidth = 0.9) +
  ggplot2::geom_hline(yintercept = 0.045, linetype = "dotted") +
  ggplot2::facet_wrap(~arm) +
  ggplot2::labs(
    title = "Replicates Figure 1 of Tsirizani 2025",
    subtitle = "Median (line) and 5th-95th percentile band; dotted line is the 0.045 mg/L LLOQ",
    x = "Time after the witnessed morning dose (h)",
    y = "Ritonavir concentration (mg/L)"
  ) +
  ggplot2::theme_bw()
```

The ordering and separation of the arms is the paper's headline result:
atazanavir well above darunavir, lopinavir below it and flatter -- the lopinavir
profile is damped both by the 23.4% lower relative bioavailability and by the
split of the daily dose across two administrations.

## Figure 2 -- covariate effects on AUC(0-24 h)

Figure 2 is a forest plot of AUC(0-24 h) ratios against a reference child of
26 kg on 100 mg ritonavir with 800 mg darunavir, with every weight category also
held at 100 mg ritonavir. Its printed ratios are the sharpest validation target
in the paper, because they are ratios of typical values and so are free of
residual error and of between-subject variability.

Two of the three are also **free of the fat-free-mass assumption**: the
companion-PI ratios compare arms at the same weight, so the FFM term cancels
exactly. The weight ratios depend only on FFM being proportional to weight, not
on the value of the constant.

```{r fig2-setup}
# Typical-value (deterministic) solves: zero all random effects.
mod_typ <- rxode2::zeroRe(mod)

fig2_cases <- tibble::tribble(
  ~label,     ~facet,      ~WT,   ~CONMED_LOPINAVIR, ~CONMED_ATAZANAVIR, ~published,
  "REF",      "Reference", 26.0,  0L,                0L,                 1.00,
  "17 kg",    "Weight",    17.0,  0L,                0L,                 1.41,
  "22.5 kg",  "Weight",    22.5,  0L,                0L,                 1.18,
  "26 kg",    "Weight",    26.0,  0L,                0L,                 1.00,
  "30 kg",    "Weight",    30.0,  0L,                0L,                 0.931,
  "35 kg",    "Weight",    35.0,  0L,                0L,                 0.845,
  "DRV",      "Boosted PI", 26.0, 0L,                0L,                 1.00,
  "LPV",      "Boosted PI", 26.0, 1L,                0L,                 0.762,
  "ATV",      "Boosted PI", 26.0, 0L,                1L,                 1.99
) |>
  dplyr::mutate(id = dplyr::row_number(), FFM = WT * FFM_PER_KG)

# All Figure 2 cases hold the ritonavir dose at 100 mg/day. The lopinavir case
# is twice daily (50 mg AM + 50 mg PM) because that is how the arm is dosed;
# steady-state AUC over 24 h is unaffected by the split, which the check below
# confirms.
f2_am <-
  fig2_cases |>
  dplyr::select(id, FFM, CONMED_LOPINAVIR, CONMED_ATAZANAVIR) |>
  tidyr::crossing(time = seq(0, T0, by = 24)) |>
  dplyr::mutate(amt = ifelse(CONMED_LOPINAVIR == 1L, 50, 100), OCC = 2L)

f2_pm <-
  fig2_cases |>
  dplyr::filter(CONMED_LOPINAVIR == 1L) |>
  dplyr::select(id, FFM, CONMED_LOPINAVIR, CONMED_ATAZANAVIR) |>
  tidyr::crossing(time = seq(12, T0 + 12, by = 24)) |>
  dplyr::mutate(amt = 50, OCC = 1L)

f2_obs <-
  fig2_cases |>
  dplyr::select(id, FFM, CONMED_LOPINAVIR, CONMED_ATAZANAVIR) |>
  tidyr::crossing(time = T0 + seq(0, 24, by = 0.05)) |>
  dplyr::mutate(amt = 0, evid = 0L, cmt = "central", rate = 0, OCC = 2L)

f2_events <-
  dplyr::bind_rows(
    dplyr::mutate(dplyr::bind_rows(f2_am, f2_pm), evid = 1L, cmt = "depot", rate = -2),
    f2_obs
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

f2_sim <-
  rxode2::rxSolve(mod_typ, f2_events, addDosing = FALSE) |>
  as.data.frame() |>
  dplyr::mutate(tnca = time - T0)
```

`AUC(0-24 h)` over the final interval is obtained analytically with the
trapezoidal rule on the 0.05 h grid, which is dense enough that the numerical
error is far below the comparison tolerance.

```{r fig2-auc}
auc24 <- function(t, c) sum(diff(t) * (utils::head(c, -1) + utils::tail(c, -1)) / 2)

fig2_obs <-
  f2_sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(auc = auc24(tnca, Cc), .groups = "drop")

fig2 <-
  fig2_cases |>
  dplyr::left_join(fig2_obs, by = "id") |>
  dplyr::group_by(facet) |>
  dplyr::mutate(
    # Within each facet the reference row is the 26 kg darunavir case.
    ref_auc   = auc[abs(WT - 26) < 1e-9 & CONMED_LOPINAVIR == 0L & CONMED_ATAZANAVIR == 0L][1],
    simulated = auc / ref_auc
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(pct_diff = 100 * (simulated / published - 1))

fig2 |>
  dplyr::select(facet, label, published, simulated, pct_diff) |>
  dplyr::rename(
    "Covariate group" = facet,
    "Level"           = label,
    "Published ratio" = published,
    "Simulated ratio" = simulated,
    "% diff"          = pct_diff
  ) |>
  knitr::kable(
    caption = "Simulated AUC(0-24 h) ratios against the values printed beside Tsirizani 2025 Figure 2.",
    digits = 3
  )
```

```{r fig2-plot, fig.height = 4}
ggplot2::ggplot(fig2, ggplot2::aes(x = simulated, y = factor(label, levels = rev(label)))) +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed") +
  ggplot2::geom_point(ggplot2::aes(colour = "Simulated"), size = 2.5) +
  ggplot2::geom_point(ggplot2::aes(x = published, colour = "Published"), size = 2.5,
                      shape = 4, stroke = 1.2) +
  ggplot2::facet_grid(facet ~ ., scales = "free_y", space = "free_y") +
  ggplot2::scale_colour_manual(values = c(Simulated = "steelblue", Published = "firebrick")) +
  ggplot2::labs(
    title = "Replicates Figure 2 of Tsirizani 2025",
    subtitle = "Ritonavir AUC(0-24 h) relative to a 26 kg child on 100 mg ritonavir with darunavir",
    x = "Ratio to reference", y = NULL, colour = NULL
  ) +
  ggplot2::theme_bw()
```

```{r fig2-gate}
pi_rows <- dplyr::filter(fig2, facet == "Boosted PI")
wt_rows <- dplyr::filter(fig2, facet == "Weight")

stopifnot(
  # The companion-PI ratios are exact properties of the encoded covariate
  # model -- the FFM term cancels because both arms sit at 26 kg -- so they
  # are held to a tight bound. A sign error, a percent-vs-fraction slip, or a
  # swapped reference category would blow this immediately.
  max(abs(pi_rows$pct_diff)) < 3,
  # The weight ratios additionally assume FFM is proportional to weight. The
  # residual gap is the curvature of the real (unreported) FFM equation across
  # 17-35 kg, which is why this bound is looser than the one above.
  max(abs(wt_rows$pct_diff)) < 8
)
```

The companion-PI ratios land within `r sprintf("%.1f%%", max(abs(pi_rows$pct_diff)))`
of the published values. The weight ratios are within
`r sprintf("%.1f%%", max(abs(wt_rows$pct_diff)))`, and every simulated weight
ratio is *further from 1* than the published one -- exactly the signature of a
real FFM equation in which fat-free mass grows more slowly than total weight, so
that the true clearance trend across the weight bands is shallower than a
constant FFM/weight ratio predicts.

## Closed-form steady-state check

For a linear model at steady state, the AUC over one dosing interval is
`F_rel * daily dose / (CL/F)` exactly, whatever the absorption model does. That
identity is computed here from the **printed Table 3 numbers**, not from the
model object, so it is a genuine check on the encoded ODE rather than a
tautology.

```{r closed-form}
closed_form <- tibble::tribble(
  ~arm,         ~f_rel,           ~cl_mult,     ~daily_mg,
  "Darunavir",  1,                1,            100,
  "Lopinavir",  1 - 0.234,        1,            100,
  "Atazanavir", 1 + 1.37,         1 + 0.207,    100
) |>
  dplyr::mutate(
    # CL/F = 10.5 L/h at FFM 21.0 kg (Table 3 + Data S1 TVFFM), the reference
    # 26 kg child. Table 3 fixes the FFM exponent on clearance at 0.75.
    cl_ref     = 10.5 * cl_mult,
    auc_closed = f_rel * daily_mg / cl_ref
  )

ode_auc <-
  fig2 |>
  dplyr::filter(facet == "Boosted PI") |>
  dplyr::transmute(
    arm = c(DRV = "Darunavir", LPV = "Lopinavir", ATV = "Atazanavir")[label],
    auc_ode = auc
  )

cf <-
  closed_form |>
  dplyr::left_join(ode_auc, by = "arm") |>
  dplyr::mutate(pct_diff = 100 * (auc_ode / auc_closed - 1))

cf |>
  dplyr::select(arm, auc_closed, auc_ode, pct_diff) |>
  dplyr::rename(
    "Arm"                       = arm,
    "Closed form (mg*h/L)"      = auc_closed,
    "ODE solve (mg*h/L)"        = auc_ode,
    "% diff"                    = pct_diff
  ) |>
  knitr::kable(
    caption = paste(
      "Steady-state AUC(0-24 h) for a 26 kg child on 100 mg/day ritonavir:",
      "F_rel * dose / (CL/F) from Table 3 against the numerical ODE solve.",
      "Both sides of the comparison use the same drawn parameters, so the",
      "residual is pure numerical error (finite run-in plus trapezoidal",
      "integration) and a tight bound is the correct assertion."
    ),
    digits = 4
  )

stopifnot(max(abs(cf$pct_diff)) < 1)
```

The lopinavir row is the informative one: it is dosed 50 mg twice daily against
100 mg once daily for the other two arms, yet reproduces the same closed form,
confirming that the evening-dose lag and the split dosing redistribute exposure
within the interval without changing its integral.

# PKNCA validation

Non-compartmental analysis is run on the stochastic cohort with `PKNCA`, over
the single steady-state dosing interval. Two grids are used, because they answer
different questions:

* the **dense** 0.25 h grid, which estimates the true AUC of the simulated
  profile;
* the **paper's sparse grid** (0, 0.5, 1, 2, 4, 6, 8, 12 and 24 h post-dose),
  which is what Tsirizani 2025 actually had available when it computed the
  "model-based AUC(0-24 h)" of Table 4.

```{r pknca}
nca_conc_dense <-
  sim |>
  dplyr::select(id, arm, band, tnca, Cc) |>
  dplyr::filter(!is.na(Cc))

paper_times <- c(0, 0.5, 1, 2, 4, 6, 8, 12, 24)
nca_conc_sparse <- dplyr::filter(nca_conc_dense, tnca %in% paper_times)

nca_dose <-
  dplyr::bind_rows(
    dplyr::transmute(cohort, id, arm, band, tnca = 0,  dose = am_mg),
    dplyr::transmute(dplyr::filter(cohort, pm_mg > 0), id, arm, band,
                     tnca = 12, dose = pm_mg)
  )

intervals <- data.frame(
  start = 0, end = 24,
  auclast = TRUE, cmax = TRUE, tmax = TRUE, cav = TRUE, cmin = TRUE
)

run_nca <- function(conc) {
  o_conc <- PKNCA::PKNCAconc(
    as.data.frame(conc), Cc ~ tnca | arm + id,
    concu = "mg/L", timeu = "h"
  )
  o_dose <- PKNCA::PKNCAdose(
    as.data.frame(nca_dose), dose ~ tnca | arm + id,
    doseu = "mg"
  )
  PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals),
                verbose = FALSE)
}

res_dense  <- run_nca(nca_conc_dense)
res_sparse <- run_nca(nca_conc_sparse)

long_of <- function(res) {
  as.data.frame(res) |>
    dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "tmax", "cav", "cmin")) |>
    dplyr::left_join(dplyr::select(cohort, id, band), by = "id") |>
    dplyr::select(id, arm, band, PPTESTCD, PPORRES)
}

sim_dense  <- long_of(res_dense)
sim_sparse <- long_of(res_sparse)

sim_dense |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    median = stats::median(PPORRES),
    q25    = stats::quantile(PPORRES, 0.25),
    q75    = stats::quantile(PPORRES, 0.75),
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(names_from = PPTESTCD,
                     values_from = c(median, q25, q75)) |>
  knitr::kable(caption = "Dense-grid NCA summary by arm.", digits = 3)
```

```{r pknca-grid-effect}
grid_effect <-
  dplyr::inner_join(
    dplyr::filter(sim_dense,  PPTESTCD == "auclast") |>
      dplyr::rename(dense = PPORRES),
    dplyr::filter(sim_sparse, PPTESTCD == "auclast") |>
      dplyr::rename(sparse = PPORRES),
    by = c("id", "arm", "band", "PPTESTCD")
  ) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `Dense grid`  = stats::median(dense),
    `Paper grid`  = stats::median(sparse),
    `Sparse/dense` = stats::median(sparse) / stats::median(dense),
    .groups = "drop"
  )

grid_effect |>
  knitr::kable(
    caption = paste(
      "Median AUC(0-24 h) on the dense 0.25 h grid versus the paper's nine",
      "sampling times, by arm."
    ),
    digits = 3
  )
```

The sparse grid turns out to matter little, and not in one direction: the
sparse/dense ratio spans
`r sprintf("%.3f", min(grid_effect[["Sparse/dense"]]))`-`r sprintf("%.3f", max(grid_effect[["Sparse/dense"]]))`.
Two errors of opposite sign roughly cancel -- linear interpolation across the
12-24 h gap over-reads a convex declining curve, while the 0-4 h absorption
phase is under-sampled -- and the lopinavir arm, whose second dose lands at
12 h, is the only one with a visible (about 5% low) bias. The sparse-grid
results are carried into the next section anyway, because they are the closest
match to what Tsirizani 2025 actually had in hand, but nothing below turns on
the choice of grid.

# Comparison against the published NCA

Tsirizani 2025 Table 4 reports model-based AUC(0-24 h) and *observed* Cmax,
stratified by companion PI. The two columns are not equally comparable, and the
distinction matters for reading the table below:

* **AUC(0-24 h) is model-based** -- a model output, so simulating it is a
  model-versus-model comparison and disagreement points at the encoding.
* **Cmax is observed** -- the largest of nine measured samples in real children.
  The paper never claims its model reproduces those maxima except through the
  Figure 1 VPC. With between-occasion variability of 114% on `ka`, 62% on the
  lag and 52% on the zero-order duration, individual peaks are enormously
  variable in both height and timing, and a nine-point grid catches the true
  peak only sometimes. Cmax is therefore the weakest of the checks in this
  vignette, not the strongest.

The pooled per-arm rows are the primary comparison; band-level rows are shown
descriptively because each rests on only 11-18 children in the paper.

```{r nca-comparison}
published_arm <- tibble::tribble(
  ~arm,          ~auclast, ~cmax,
  "Lopinavir",    9.83,    0.66,
  "Atazanavir",  17.6,     2.38,
  "Darunavir",    8.31,    1.12
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = dplyr::filter(sim_sparse, PPTESTCD %in% c("auclast", "cmax")),
  reference     = published_arm,
  by            = "arm",
  units         = c(auclast = "mg*h/L", cmax = "mg/L"),
  tolerance_pct = 20
)

cmp |>
  dplyr::rename("Boosted PI" = arm) |>
  knitr::kable(
    caption = paste(
      "Simulated steady-state NCA on the paper's sampling grid versus",
      "Tsirizani 2025 Table 4 medians (AUC model-based, Cmax observed).",
      "* differs from the published median by more than 20%."
    ),
    digits = 3
  )
```

Before asserting anything, the Monte-Carlo error of these medians has to be
known. `rxode2::rxSetSeed()` fixes the random stream per solver thread but not
across thread counts, so a different machine -- or the same machine with a
different core count -- draws a different cohort. An assertion with less
headroom than the Monte-Carlo error would be a coin flip, not a check.

```{r mcse}
mcse <-
  sim_sparse |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    median = stats::median(PPORRES),
    # Bootstrap standard error of the median, expressed as a percentage of it.
    mcse_pct = 100 * stats::sd(replicate(
      500, stats::median(sample(PPORRES, length(PPORRES), replace = TRUE))
    )) / stats::median(PPORRES),
    .groups = "drop"
  )

mcse |>
  dplyr::mutate(PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  dplyr::rename(
    "Boosted PI"        = arm,
    "NCA parameter"     = PPTESTCD,
    "Simulated median"  = median,
    "MC-SE (% of median)" = mcse_pct
  ) |>
  knitr::kable(
    caption = paste(
      "Bootstrap Monte-Carlo standard error of each simulated median",
      "(170 children per arm). Cmax is roughly twice as noisy as AUC,",
      "because the 114% between-occasion variability on ka scatters both the",
      "height and the timing of individual peaks."
    ),
    digits = 2
  )
```

```{r nca-comparison-gate}
gate <-
  cmp |>
  dplyr::mutate(pct = suppressWarnings(as.numeric(gsub("[^0-9.-]", "", `% diff`))))

auc_gate <- dplyr::filter(gate, grepl("AUC", gate[[1]], fixed = TRUE))

# The structural gate on this model is NOT this comparison -- it is the
# deterministic pair above: the Figure 2 companion-PI ratios (held to 3%) and
# the closed-form steady-state identity (held to 1%). Those are exact,
# thread-count-independent, and strictly better at catching a mis-transcription:
# a percent-versus-fraction slip on the atazanavir clearance effect would move
# the Figure 2 atazanavir ratio by 17% and blow its 3% bound instantly, while
# barely moving a cohort median.
#
# This comparison is against published medians of 51-60 real children, filtered
# through an unreported FFM equation and an unknown within-band weight
# distribution, and its own Monte-Carlo error is 4-7% (above). It is therefore
# asserted on the CENTRE, with headroom of at least twice the Monte-Carlo error,
# so that a re-draw on a machine with a different core count cannot flip it.
stopifnot(
  nrow(auc_gate) == 3L,
  stats::median(abs(gate$pct)) < 20,
  all(abs(gate$pct) < 30)
)
```

```{r nca-comparison-bands, echo = FALSE}
published_band <- tibble::tribble(
  ~arm,          ~band,        ~auclast, ~cmax,
  "Lopinavir",   "14-19.9 kg",  8.57,    0.54,
  "Lopinavir",   "20-24.9 kg", 10.3,     0.52,
  "Lopinavir",   "25-34.9 kg", 11.3,     1.04,
  "Lopinavir",   "35+ kg",      9.83,    0.80,
  "Atazanavir",  "14-19.9 kg", 19.8,     2.68,
  "Atazanavir",  "20-24.9 kg", 14.7,     2.30,
  "Atazanavir",  "25-34.9 kg", 20.3,     2.76,
  "Atazanavir",  "35+ kg",     14.2,     1.76,
  "Darunavir",   "14-19.9 kg", 11.8,     1.63,
  "Darunavir",   "20-24.9 kg",  9.68,    1.27,
  "Darunavir",   "25-34.9 kg",  7.86,    1.14,
  "Darunavir",   "35+ kg",      6.95,    0.81
)

sim_band <-
  sim_sparse |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
  dplyr::group_by(arm, band, PPTESTCD) |>
  dplyr::summarise(simulated = stats::median(PPORRES), .groups = "drop")

published_band |>
  tidyr::pivot_longer(c(auclast, cmax), names_to = "PPTESTCD",
                      values_to = "published") |>
  dplyr::left_join(sim_band, by = c("arm", "band", "PPTESTCD")) |>
  dplyr::mutate(
    PPTESTCD = nlmixr2lib::ncaParamLabel(PPTESTCD),
    pct_diff = 100 * (simulated / published - 1)
  ) |>
  dplyr::rename(
    "Boosted PI"  = arm,
    "Weight band" = band,
    "Parameter"   = PPTESTCD,
    "Published"   = published,
    "Simulated"   = simulated,
    "% diff"      = pct_diff
  ) |>
  knitr::kable(
    caption = paste(
      "Descriptive band-level comparison against Tsirizani 2025 Table 4.",
      "Band medians in the paper rest on 11-18 children each, so scatter of",
      "this size in both directions is expected and no assertion is made on",
      "individual rows."
    ),
    digits = 3
  )
```

The pooled per-arm agreement is within
`r sprintf("%.0f%%", max(abs(auc_gate$pct)))` on the model-based AUC in all
three arms and within `r sprintf("%.0f%%", max(abs(gate$pct)))` on the observed
Cmax, and it reproduces the paper's central finding quantitatively: atazanavir
roughly doubles ritonavir exposure relative to darunavir, and lopinavir sits
slightly below it.

The band-level rows scatter in both directions without a systematic trend, which
is what a comparison of 11-18-child published medians against a simulated cohort
should look like. The one row that stands out is atazanavir at 25-34.9 kg
(published AUC 20.3 mg*h/L, simulated 15.2), and it is worth noting that the
published value is *not* monotone across the atazanavir bands -- 19.8, 14.7,
20.3, 14.2 -- while the model, whose only weight-related term is a monotone
allometric one, cannot reproduce a zig-zag of that shape at any parameter values.
The alternation is most likely small-sample noise in bands of 12-18 children,
compounded by the footnoted children who received 100 mg instead of 75 mg in the
two lightest bands.

```{r cmin-note, echo = FALSE}
sim_sparse |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    `Median Cmin (mg/L)` = round(stats::median(PPORRES), 3),
    `Fraction below LLOQ 0.045 mg/L` = round(mean(PPORRES < 0.045), 3),
    .groups = "drop"
  ) |>
  knitr::kable(
    caption = paste(
      "Simulated end-of-interval concentrations. Tsirizani 2025 reports that",
      "6.9% of the 1,254 measured concentrations were below the 0.045 mg/L",
      "LLOQ and 3.7% undetectable; troughs are where those records come from."
    )
  )
```

# Assumptions and deviations

## Values not stated in the paper

* **The fat-free-mass equation is not reported.** `FFM` entered the analysis as
  a pre-computed data column (`$INPUT ... FFM`), and the Methods cite only
  Holford & Anderson 2017 (reference 21) for the allometric-size *theory*, not
  a prediction equation. Neither the main text nor the supplement gives the
  formula, so no equation is imported here. Instead FFM is scaled from weight
  using the paper's own anchor: `TVFFM = 21.0` kg (Data S1) for the 26 kg child
  that Table 3 footnote b names as the reference, i.e. `FFM = 0.808 * WT`. This
  is an approximation -- fat-free mass does not grow in strict proportion to
  weight across 14-64 kg in children -- and the Figure 2 weight-ratio
  comparison above quantifies the resulting error at under 6%, with a
  consistent sign. A user with measured weight, height, age and sex should
  supply `FFM` directly rather than rely on this scaling.
* **The FFM reference value 21.0 kg is recoverable only from the supplement.**
  The main text gives the 26 kg weight anchor but never the corresponding
  fat-free mass. Without Data S1 the allometric term could not be reproduced at
  all.

## Reconciling the additive residual error

Data S1 builds the additive error as `ADD = THETA(6) + (LLOQ * 0.2)` with
`LLOQ = 0.045`, so the total additive standard deviation at an uncensored record
is `THETA(6) + 0.009`. Table 3's printed 0.0152 mg/L is that **total**, not
`THETA(6)`: the control-stream initial estimate is `THETA(6) = 0.00647`, and
`0.0152 - 0.009 = 0.0062` is a 4% move from it, in line with every other
parameter's initial-to-final drift (1-8%, except the night-lag term at 23%).
Reading 0.0152 as `THETA(6)` itself would require a 135% move, far outside that
pattern. The model therefore encodes `addSd = 0.0152` mg/L, which is both the
printed value and the additive SD that actually applies to a measurable
concentration.

The censoring machinery in Data S1 -- the further `CENS_THR * 0.5` inflation on
M6-censored records, and the `ADD = 1e10` blanking of trailing undetectable
values -- is data-handling for estimation, not part of the predictive model, and
is deliberately not encoded.

## Internal inconsistencies in the source

* **Table 1's lopinavir dose row is per administration, not per day.** Table 1
  reports "RTV total daily dose (mg)" for the lopinavir arm as 100 [50.0-100],
  but Table 2 gives that arm 100-200 mg of ritonavir per *day* (up to 2 + 2
  tablets of 200/50 mg lopinavir/ritonavir). The accompanying mg/kg row settles
  it: 2.65 mg/kg [2.02-4.00] against a 26 kg median implies about 69 mg, and the
  smallest weight band on a true 100 mg/day would already exceed 5 mg/kg. The
  once-daily arms have no such ambiguity, and their ranges reconcile exactly
  (atazanavir 100/64.2 = 1.56 and 100/14.5 = 6.90; darunavir 100/47.0 = 2.13 and
  100/14.5 = 6.90). Simulation therefore uses the Table 2 daily doses.
* **The atazanavir clearance effect is +20.7% in Table 3 and 20.0% in the
  Results text.** Table 3, the final-parameter table, is used; the control
  stream's initial estimate (1.20875) corroborates 20.7% over 20.0%.
* **The Discussion states lopinavir lowers AUC(0-24 h) by 34.8%.** The ratio
  printed beside Figure 2 is 0.762 (95% CI 0.698-0.847), i.e. -23.8%, which
  matches Table 3's -23.4% bioavailability effect and the closed-form
  calculation above. The 34.8% figure is not reproducible from any combination
  of the published parameters and is treated as an error in the Discussion;
  the paired atazanavir figure in the same sentence (+99.0%) does match
  Figure 2's 1.99 and is used as published.
* **The Discussion's summary paragraph says atazanavir "slightly reduced
  clearance".** Table 3, the Results text and a later Discussion paragraph all
  say clearance is *higher* with atazanavir. The increase is encoded.
* **Table 3's between-occasion row for the zero-order phase is labelled
  "Zero-order rate of absorption (%)", but the parameter is a duration.** The
  control stream carries that random effect on `D1` (`D1 = TVD1*EXP(BOVD1)`),
  and Table 3's own structural row two lines above is "Zero-order absorption
  duration (h)". The 52.5% is therefore variability in the *duration*, which is
  how it is encoded (`etaiov_d1_*`); the word "rate" in that one label is loose.

## Modelling choices

* **Zero-variance random effects are omitted, not encoded as `fixed(0)`.** Data
  S1 fixes between-subject variability on Vc, ka, F, Vp and Q, and
  between-occasion variability on CL, to zero. Writing them as `~ fixed(0)`
  would make OMEGA singular; Table 3 lists no such rows either.
* **Occasion assignment during the simulated run-in.** `OCC` both selects the
  BOV eta and identifies the evening dose. Run-in doses are labelled by time of
  day so each gets its correct absorption lag; the steady-state AUC of the final
  interval is unaffected, since absorption lag changes a dose's input shape and
  not the amount absorbed. The witnessed dose is `OCC = 2`, matching the Table 3
  reference condition.
* **Weight is sampled uniformly within each WHO band, and matched across
  arms.** Table 1 reports band counts and a per-arm median weight of 26.0 kg but
  not the within-band distribution, so uniform sampling is the least-assumption
  choice; it reproduces the reported 26.0 kg median when the *pooled* band
  counts are used (the simulated median is 26.3 kg). The same weight vector is
  given to all three arms so that arm-to-arm differences in the comparison
  tables are attributable to the model rather than to cohort composition -- an
  earlier draft using each arm's own band counts produced medians of 25.0 kg
  (darunavir) against 27.1 kg (atazanavir), which shifted the two arms in
  opposite directions through the allometric term and inflated every
  discrepancy by roughly 8 percentage points.
* **Weight sampling in the top band.** The `35+ kg` band is sampled from
  35-50 kg. The cohort's single heaviest child (64.2 kg, in the atazanavir arm)
  is not represented; per-arm maxima in Table 1 are 49.5 kg (lopinavir),
  64.2 kg (atazanavir) and 47.0 kg (darunavir).
* **Race and ethnicity are not modelled.** Tsirizani 2025 does not report them
  by category; all participants were enrolled in Zambia, Uganda and Zimbabwe.

## Scope limit stated by the authors

The authors are explicit that this model **should not be used to simulate
ritonavir doses above 100 mg per administration**. Ritonavir shows
dose-dependent saturable pharmacokinetics at the 200-500 mg doses once used
therapeutically, and this dataset -- confined to boosting doses, with dose
changes confounded by companion PI, formulation and weight -- could not
characterise that saturation. A saturable liver-elimination model was tested and
rejected (dOFV = -0.390), with the estimated Michaelis constant close to the
maximum observed concentration, which is itself evidence that the data carry no
information about saturation.
