---
title: "Cholecalciferol and vitamin D3 metabolites in CKD (Tuey 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cholecalciferol and vitamin D3 metabolites in CKD (Tuey 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Tuey_2024_cholecalciferol"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.3390/ijms252212279> (open access, PMC11595143)
* Trial registration: ClinicalTrials.gov NCT02360644

# Population

Tuey 2024 enrolled 29 adults with non-dialysis chronic kidney disease (stages
1-5) and vitamin D insufficiency or deficiency, defined as a total 25D3
concentration below 30 ng/mL. Baseline characteristics (Table 1): 59% female;
median age 61 years (range 29-73); median weight 92.0 kg (70.7-135.3); median
BMI 32.6 kg/m^2 (25.6-43.4); median eGFR 37 mL/min/1.73 m^2 (11-97); median
baseline 25D3 18 ng/mL (7-29). Race was reported as 66% White and 34% Black.

Each subject received a single 5000 I.U. oral dose of cholecalciferol, with
plasma sampled at baseline and 0.5, 1, 2, 4, 8, 12, 24, 48, 168 and 336 h
(Methods 4.1). 310 concentrations entered the model; 212 of the parent VitD3
observations (over 72%) were below the limit of quantification and were handled
with the Beal M3 method. The model was fitted in Phoenix NLME v8.3 using the
QRPEM engine.

The same information is available programmatically via
`readModelDb("Tuey_2024_cholecalciferol")()$population`.

# Model structure

The model (Tuey 2024 Figure 1) is a single simultaneous fit of the parent and
three sequential metabolites:

* **Cholecalciferol (VitD3)** -- two-compartment disposition
  (`central` / `peripheral1`) with first-order absorption from `depot` and a
  constant zero-order endogenous production rate `ksyn` entering `central`.
  All parameters are apparent (divided by the unknown oral bioavailability F).
* **25D3** -- one compartment, formed first-order from VitD3. The formed
  fraction `fm_25d3` is fixed to 1, i.e. VitD3 has no elimination pathway other
  than 25-hydroxylation.
* **1,25D3** and **24,25D3** -- one compartment each, formed first-order from
  25D3. `fm_125d3` is fixed to 0.017, and the remaining `1 - fm_125d3` of 25D3
  clearance forms 24,25D3.

Every species carries an estimated endogenous pre-dose baseline concentration
used as that compartment's initial condition. All concentrations are molar
(nmol/L) because Methods 4.5 converted every observation from ng/mL before
combining the four analytes into one data set.

```{r structure, echo = FALSE}
ui$iniDf |>
  dplyr::filter(!is.na(ntheta)) |>
  dplyr::transmute(
    Parameter = name,
    `Value on the ini() scale` = est,
    Fixed = ifelse(fix, "yes", ""),
    Label = label
  ) |>
  knitr::kable(digits = 4, caption = "Parameters as stored in the packaged model.")
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Tuey_2024_cholecalciferol.R`. The table
below collects them in one place for review. Every value comes from Table 2 of
Tuey 2024, read from the published PDF layout; note that the preprocessed
markdown rendering of that table is scrambled and should not be used.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lrbase` (C0) | 0.98 nmol/L | Table 2, RSE 41.7% |
| `lka` (ka) | fixed 0.054 1/h | Table 2; Discussion restates 0.054 |
| `ksyn` (kendog) | fixed 0.55 nmol/h | Table 2; Discussion "fixed endogenous rate of 0.55 nmol/h" |
| `lvc` (Vc/F) | 21.3 L | Table 2, RSE 22.2% |
| `lcl` (CL/F) | 1.4 L/h | Table 2, RSE 42.4% |
| `lvp` (Vp/F) | fixed 50 L | Table 2 |
| `lq` (Q/F) | fixed 0.44 L/h | Table 2; Discussion restates 0.44 L/h |
| `lrbase_25d3` (C0m1) | 43.5 nmol/L | Table 2, RSE 4.1% |
| `lvc_25d3` (Vm1) | 58.3 L | Table 2, RSE 14.8% |
| `lcl_25d3` (CLm1) | 0.02 L/h | Table 2, RSE 52.2% |
| `lrbase_125d3` (C0m2) | 0.20 nmol/L | Table 2, RSE 6.9% |
| `lvc_125d3` (Vm2) | 71.5 L | Table 2, RSE 206.8% |
| `lcl_125d3` (CLm2) | 0.08 L/h | Table 2, RSE 47.7% |
| `lrbase_2425d3` (C0m3) | 2.2 nmol/L | Table 2, RSE 9.4% |
| `lvc_2425d3` (Vm3) | 105.2 L | Table 2, RSE 140.5% |
| `lcl_2425d3` (CLm3) | 0.40 L/h | Table 2, RSE 53.4% |
| `fm_25d3` (fm1) | fixed 1 | Results 2.2 / Methods 4.6 |
| `fm_125d3` (fm2) | fixed 0.017 | Results 2.2 / Methods 4.6 (from the Sawyer 2022 PBPK model) |
| `propSd` (sigma1) | 0.125 | Table 2, 12.5% |
| `propSd_25d3` (sigma2) | 0.657 | Table 2, 65.7% |
| `propSd_125d3` (sigma3) | 0.172 | Table 2, 17.2% |
| `propSd_2425d3` (sigma4) | 0.166 | Table 2, 16.6% |
| IIV on C0, Vc/F, CL/F | not published | Results 2.2 declares the terms; no OMEGA magnitude is reported |
| `d/dt(central)` etc. | n/a | Figure 1 schematic + Results 2.2 narrative |

# Simulation setup

Tuey 2024 publishes no IIV magnitudes (see *Assumptions and deviations*), so the
packaged model is deterministic and every simulation below is a typical-value
prediction. Cohorts are therefore one subject per dose arm rather than a virtual
population -- with all etas fixed at zero, additional subjects would be exact
duplicates and would add no validation value.

```{r constants}
# Molecular weights used by Tuey 2024 Methods 4.5 to convert ng/mL to nmol/L.
MW <- c(vitd3 = 384.64, d25 = 400.64, d125 = 416.64, d2425 = 416.64)

# 1 I.U. of vitamin D3 is defined as 0.025 ug of cholecalciferol.
iu_to_nmol <- 0.025 * 1000 / MW[["vitd3"]]
stopifnot(all.equal(5000 * iu_to_nmol, 324.979, tolerance = 1e-4))

nmol_per_L_to_ng_per_mL <- function(x, mw) x * mw / 1000

mod <- readModelDb("Tuey_2024_cholecalciferol")

# The model declares four endpoints (Cc, Cc_25d3, Cc_125d3, Cc_2425d3), so
# rxode2 builds a dvid -> compartment map that places each endpoint in a
# pseudo-compartment AFTER the six real ODE states (dvid 1-4 -> slots 7-10).
# An observation row therefore has to identify the endpoint, and a bare
# `cmt = "<ODE state>"` fails with
#   'dvid'->'cmt' or 'cmt' on observation record or on a undefined compartment
# The fix is `dvid` -- NOT naming an algebraic observable in `cmt`, which would
# inject a compartment slot and renumber the states. Observation rows name the
# real `central` state and carry `dvid = 1L`; rxSolve then returns all four
# observables as columns regardless. `dvid` is set on the dose rows too so the
# column is not NA-typed. Verified numerically identical to the endpoint-named
# form on rxode2 5.1.7.
build_events <- function(dose_iu, id, ii = 24, until_h, obs_times) {
  dose_rows <- data.frame(
    id = id, time = seq(0, until_h, by = ii),
    amt = dose_iu * iu_to_nmol, evid = 1L, cmt = "depot", dvid = 1L
  )
  if (dose_iu == 0) dose_rows <- dose_rows[0, ]
  obs_rows <- data.frame(
    id = id, time = obs_times, amt = NA_real_, evid = 0L, cmt = "central",
    dvid = 1L
  )
  rbind(dose_rows, obs_rows) |> dplyr::arrange(time, dplyr::desc(evid))
}

solve_typical <- function(events, keep = character()) {
  out <- rxode2::rxSolve(
    mod, events, omega = NA,
    keep = keep, returnType = "data.frame"
  )
  # rxSolve omits the `id` column entirely when the event table holds a
  # single subject, so restore it for the downstream grouping / PKNCA code.
  if (is.null(out$id)) out$id <- unique(events$id)
  out
}
```

# Study design: single 5000 I.U. oral dose

This reproduces the sampling design actually studied (Methods 4.1) and
corresponds to the four VPC panels of Figure 3.

```{r study-sim}
study_times <- sort(unique(c(seq(0, 336, by = 0.5), c(0.5, 1, 2, 4, 8, 12, 24, 48, 168, 336))))
study_ev <- build_events(dose_iu = 5000, id = 1L, until_h = 0, obs_times = study_times)
study <- solve_typical(study_ev)

stopifnot(nrow(study) > 0, !anyNA(study$Cc), !anyNA(study$Cc_25d3))
```

```{r figure-3, echo = FALSE}
study |>
  transmute(
    time,
    `VitD3`    = nmol_per_L_to_ng_per_mL(Cc, MW[["vitd3"]]),
    `25D3`     = nmol_per_L_to_ng_per_mL(Cc_25d3, MW[["d25"]]),
    `1,25D3`   = nmol_per_L_to_ng_per_mL(Cc_125d3, MW[["d125"]]),
    `24,25D3`  = nmol_per_L_to_ng_per_mL(Cc_2425d3, MW[["d2425"]])
  ) |>
  pivot_longer(-time, names_to = "analyte", values_to = "conc") |>
  mutate(analyte = factor(analyte, levels = c("VitD3", "25D3", "1,25D3", "24,25D3"))) |>
  ggplot(aes(time, conc)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~analyte, scales = "free_y") +
  labs(
    x = "Time (h)", y = "Concentration (ng/mL)",
    title = "Typical-value profiles after a single 5000 I.U. oral dose",
    caption = "Corresponds to the four VPC panels of Figure 3 of Tuey 2024 (typical value, no IIV)."
  ) +
  theme_bw()
```

The parent rises and falls over the first ~48 h while 25D3 accumulates slowly
and the two dihydroxy metabolites track it, which is the qualitative pattern
Figure 3 shows.

# Structural check: baseline flux balance

The three metabolite baselines are estimated independently, but the two
dihydroxy baselines are over-determined: if the system is at endogenous steady
state before dosing, then for each dihydroxy metabolite

```
formation = elimination
fm * CLm1 * C0m1 = CLm_x * C0m_x
```

so `C0m2` and `C0m3` are predicted by `C0m1`, `CLm1`, `fm_125d3` and the
respective metabolite clearance. This is an exact structural identity that the
paper never states, and it is a strong falsifier for the direction of the
metabolic chain: it only holds if `fm_125d3` splits the **25D3 clearance**
(rather than, say, the VitD3 clearance) and if 24,25D3 receives the complement.

```{r flux-balance}
# ui$theta holds each estimate on its ini() scale, so the l-prefixed
# parameters need exponentiating and the bare ones (fm_*) do not.
theta <- ui$theta
natural <- function(nm) if (startsWith(nm, "l")) exp(theta[[nm]]) else theta[[nm]]

fm_125d3     <- natural("fm_125d3")
cl_25d3      <- natural("lcl_25d3")
rbase_25d3   <- natural("lrbase_25d3")
cl_125d3     <- natural("lcl_125d3")
rbase_125d3  <- natural("lrbase_125d3")
cl_2425d3    <- natural("lcl_2425d3")
rbase_2425d3 <- natural("lrbase_2425d3")

# Guard: the values pulled out of the model must be the published ones.
stopifnot(
  all.equal(fm_125d3, 0.017), all.equal(cl_25d3, 0.02),
  all.equal(rbase_25d3, 43.5), all.equal(cl_125d3, 0.08),
  all.equal(rbase_125d3, 0.20), all.equal(cl_2425d3, 0.40),
  all.equal(rbase_2425d3, 2.2)
)

flux <- tibble::tibble(
  Metabolite = c("1,25D3", "24,25D3"),
  `Predicted baseline (nmol/L)` = c(
    fm_125d3       * cl_25d3 * rbase_25d3 / cl_125d3,
    (1 - fm_125d3) * cl_25d3 * rbase_25d3 / cl_2425d3
  ),
  `Published baseline (nmol/L)` = c(rbase_125d3, rbase_2425d3)
) |>
  mutate(`Difference (%)` = 100 * (`Predicted baseline (nmol/L)` -
                                     `Published baseline (nmol/L)`) /
           `Published baseline (nmol/L)`)

knitr::kable(flux, digits = 3,
             caption = "Steady-state flux balance predicts the two dihydroxy baselines from the 25D3 baseline.")

# Both predictions must land within 10% of the published estimates.
stopifnot(nrow(flux) == 2L, all(abs(flux$`Difference (%)`) < 10))
```

Both baselines fall out of the flux balance to within 8%, confirming the chain
direction and the role of `fm_125d3` as reported.

# Replicating the published dose-ranging simulation (Figure 4)

Tuey 2024 simulated 25D3 for 600, 1000, 2000, 5000 and 10,000 I.U./day over six
months (Methods 4.9, Results 2.5, Figure 4) and reported two endpoint sets: the
maximum mean 25D3 concentration at the end of treatment, and the time to reach
the 30 ng/mL repletion target.

```{r figure-4-sim}
dose_levels <- c(600, 1000, 2000, 5000, 10000)
six_months_h <- 24 * 182
obs_grid <- seq(0, six_months_h + 24, by = 1)

sim4 <- bind_rows(lapply(seq_along(dose_levels), function(i) {
  ev <- build_events(dose_levels[i], id = i, until_h = six_months_h, obs_times = obs_grid)
  ev$treatment <- paste0(format(dose_levels[i], big.mark = ",", trim = TRUE), " I.U./day")
  solve_typical(ev, keep = "treatment")
}))

# Each arm is solved separately and carries its own id and label; confirm all
# five arms are present and none collapsed into another.
stopifnot(
  length(unique(sim4$id)) == length(dose_levels),
  length(unique(sim4$treatment)) == length(dose_levels),
  all(table(sim4$treatment) == length(obs_grid))
)

sim4 <- sim4 |>
  mutate(
    conc_25d3 = nmol_per_L_to_ng_per_mL(Cc_25d3, MW[["d25"]]),
    treatment = factor(treatment,
                       levels = paste0(format(dose_levels, big.mark = ",", trim = TRUE),
                                       " I.U./day"))
  )
```

```{r figure-4, echo = FALSE}
ggplot(sim4, aes(time / 24, conc_25d3)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 30, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 60, linetype = "dotted", colour = "grey40") +
  facet_wrap(~treatment, scales = "free_y") +
  labs(
    x = "Time (days)", y = "25D3 (ng/mL)",
    title = "25D3 over six months of daily cholecalciferol",
    caption = paste("Replicates Figure 4A-E of Tuey 2024. Dashed line = 30 ng/mL",
                    "repletion target; dotted line = 60 ng/mL upper target.")
  ) +
  theme_bw()
```

## Comparison against the published simulation endpoints

```{r figure-4-compare}
published <- tibble::tibble(
  treatment  = levels(sim4$treatment),
  max_pub    = c(38.1, 54.1, 95.3, 218.5, 424.03),   # Results 2.5, ng/mL
  t30_pub    = c(1488, 840, 360, 144, 96)            # Results 2.5, h
)

simulated <- sim4 |>
  group_by(treatment) |>
  summarise(
    max_sim = max(conc_25d3),
    t30_sim = {
      hit <- time[conc_25d3 >= 30]
      if (length(hit)) min(hit) else NA_real_
    },
    .groups = "drop"
  )

cmp <- left_join(simulated, published, by = "treatment") |>
  mutate(
    max_pct = 100 * (max_sim - max_pub) / max_pub,
    t30_pct = 100 * (t30_sim - t30_pub) / t30_pub
  )

cmp |>
  dplyr::rename(
    "Regimen"                    = treatment,
    "Max 25D3, simulated (ng/mL)" = max_sim,
    "Max 25D3, published (ng/mL)" = max_pub,
    "Max diff (%)"                = max_pct,
    "Time to 30 ng/mL, simulated (h)" = t30_sim,
    "Time to 30 ng/mL, published (h)" = t30_pub,
    "Time diff (%)"               = t30_pct
  ) |>
  dplyr::relocate("Regimen", "Max 25D3, simulated (ng/mL)",
                  "Max 25D3, published (ng/mL)", "Max diff (%)") |>
  knitr::kable(digits = 1,
               caption = "Simulated vs. published six-month simulation endpoints (Tuey 2024 Results 2.5).")

# The end-of-treatment plateau is the well-conditioned endpoint and must
# reproduce to within 5% at every dose level.
stopifnot(nrow(cmp) == 5L, all(abs(cmp$max_pct) < 5))
```

The end-of-treatment 25D3 concentration reproduces every published dose level to
within 2%. This is the endpoint worth asserting on: at steady state the 25D3
plateau is set by the daily input and `CLm1` alone, so it is insensitive to the
unpublished IIV.

The time-to-30 ng/mL endpoint runs 12-31% late. That comparison is
intrinsically ill-conditioned rather than a model discrepancy: 25D3 approaches
its plateau asymptotically, so near the 30 ng/mL crossing the curve is nearly
flat and a small vertical error maps to a large horizontal one. The check below
quantifies that -- evaluated at the *published* crossing times, the simulated
concentrations are all within 11% of 30 ng/mL, while the crossing times differ
by up to 31%.

```{r crossing-conditioning}
conditioning <- cmp |>
  rowwise() |>
  mutate(
    conc_at_pub_t30 = approx(
      sim4$time[sim4$treatment == treatment],
      sim4$conc_25d3[sim4$treatment == treatment],
      xout = t30_pub
    )$y,
    slope_ng_per_day = approx(
      sim4$time[sim4$treatment == treatment],
      sim4$conc_25d3[sim4$treatment == treatment],
      xout = t30_pub + 24
    )$y - conc_at_pub_t30
  ) |>
  ungroup() |>
  transmute(
    treatment,
    conc_at_pub_t30,
    vertical_pct = 100 * (conc_at_pub_t30 - 30) / 30,
    slope_ng_per_day,
    horizontal_pct = t30_pct
  )

conditioning |>
  dplyr::rename(
    "Regimen"                              = treatment,
    "Simulated 25D3 at published t30 (ng/mL)" = conc_at_pub_t30,
    "Vertical difference vs 30 ng/mL (%)"  = vertical_pct,
    "Local slope (ng/mL/day)"              = slope_ng_per_day,
    "Horizontal difference in t30 (%)"     = horizontal_pct
  ) |>
  knitr::kable(digits = 2,
               caption = paste("The crossing-time comparison amplifies a small vertical",
                               "error: agreement on concentration is much tighter than",
                               "agreement on the time at which a threshold is crossed."))

# Vertical agreement must be strictly better than horizontal agreement at
# every dose level -- that is what makes this an conditioning artefact rather
# than a structural error.
stopifnot(
  nrow(conditioning) == 5L,
  all(abs(conditioning$vertical_pct) < 15),
  all(abs(conditioning$vertical_pct) < abs(conditioning$horizontal_pct))
)
```

A residual few-percent vertical offset in the rise phase remains, and it has a
plausible mechanistic explanation. The published numbers are *means over 20
replicates simulated with IIV* on C0, Vc/F and CL/F (Methods 4.9), whereas this
is a typical-value curve. The rise phase is governed by VitD3 disposition, which
carries all three IIV terms, so a log-normal IIV on CL/F makes the mean rise
faster than the typical-value curve. The plateau, by contrast, is governed by
`CLm1`, which carries no IIV -- which is exactly why the plateau agrees to
within 2% while the approach to it does not. The IIV magnitudes are unpublished,
so the mean-with-IIV curve cannot be reproduced.

# PKNCA validation

NCA is run on the parent cholecalciferol profile after the single 5000 I.U. dose
that was actually studied. Because VitD3 is endogenous, the drug-attributable
concentration is obtained by subtracting a matched no-dose simulation rather
than by subtracting a constant baseline: the model's endogenous production rate
`ksyn` was fixed from an earlier model iteration and is not in balance with the
estimated baseline C0, so the undosed profile drifts and a constant-baseline
correction would leave a negative tail.

```{r nca-baseline}
nca_doses <- dose_levels
nca_times <- sort(unique(c(seq(0, 336, by = 0.25), c(0.5, 1, 2, 4, 8, 12, 24, 48, 168, 336))))

# Dosed and undosed arms, matched one-for-one on the same time grid.
dosed <- bind_rows(lapply(seq_along(nca_doses), function(i) {
  ev <- build_events(nca_doses[i], id = i, until_h = 0, obs_times = nca_times)
  ev$treatment <- paste0(format(nca_doses[i], big.mark = ",", trim = TRUE), " I.U.")
  solve_typical(ev, keep = "treatment")
}))

undosed <- build_events(0, id = 1L, until_h = 0, obs_times = nca_times) |>
  solve_typical() |>
  transmute(time, Cc_endog = Cc)

sim_nca <- dosed |>
  left_join(undosed, by = "time") |>
  mutate(Cc = Cc - Cc_endog) |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

# The endogenous correction must remove the baseline exactly at time zero and
# must never produce a negative concentration.
stopifnot(
  nrow(sim_nca) > 0,
  all(abs(sim_nca$Cc[sim_nca$time == 0]) < 1e-8),
  all(sim_nca$Cc >= -1e-8)
)
sim_nca$Cc <- pmax(sim_nca$Cc, 0)

# Built directly from the dose ladder (keyed by id) rather than by relying on
# the row order of a distinct() over the simulation output.
dose_df <- tibble::tibble(
  id        = seq_along(nca_doses),
  treatment = paste0(format(nca_doses, big.mark = ",", trim = TRUE), " I.U."),
  time      = 0,
  amt       = nca_doses * iu_to_nmol
)
stopifnot(setequal(dose_df$treatment, unique(sim_nca$treatment)))
```

```{r nca-run}
conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  mutate(treatment = factor(treatment, levels = paste0(
    format(nca_doses, big.mark = ",", trim = TRUE), " I.U."))) |>
  arrange(treatment) |>
  dplyr::rename(
    "Dose"                  = treatment,
    "Cmax (nmol/L)"         = cmax,
    "Tmax (h)"              = tmax,
    "AUClast (nmol*h/L)"    = auclast,
    "AUC0-inf (nmol*h/L)"   = aucinf.obs,
    "t-half (h)"            = half.life
  ) |>
  knitr::kable(digits = 3,
               caption = "PKNCA summary for baseline-corrected parent cholecalciferol.")
```

Tuey 2024 reports no NCA table, so there is nothing to compare against
directly. Two internal consistency checks are available instead.

```{r nca-checks}
# Attach the dose BY TREATMENT KEY, never by row position: pivot_wider()
# returns treatments in alphabetical order ("1,000 I.U." before "600 I.U."),
# so zipping the dose ladder on positionally would silently mispair every row.
nca_chk <- nca_wide |>
  left_join(dose_df |> select(treatment, dose_nmol = amt), by = "treatment") |>
  mutate(cl_implied = dose_nmol / aucinf.obs,
         cmax_per_dose = cmax / dose_nmol)
stopifnot(nrow(nca_chk) == length(nca_doses), !anyNA(nca_chk$dose_nmol))

# 1. Strict dose proportionality. The model is linear in dose, so Cmax and AUC
#    must scale exactly with dose and the implied CL/F must be constant. The
#    tolerance is set just above the numerical spread actually achieved
#    (~5e-7), so it is a real regression guard rather than a formality.
stopifnot(
  diff(range(nca_chk$cl_implied)) / mean(nca_chk$cl_implied) < 1e-5,
  diff(range(nca_chk$cmax_per_dose)) / mean(nca_chk$cmax_per_dose) < 1e-5
)

# 2. The implied CL/F must recover the published CL/F of 1.4 L/h.
cl_published <- 1.4
cl_recovered <- mean(nca_chk$cl_implied)

tibble::tibble(
  Quantity = c("CL/F implied by NCA (Dose / AUC0-inf)", "CL/F published (Table 2)"),
  `Value (L/h)` = c(cl_recovered, cl_published)
) |>
  knitr::kable(digits = 3, caption = "NCA recovers the published apparent clearance.")

# Tolerance set to the accuracy actually achieved (0.04%), not to a loose
# round number, so a future change to the dose conversion would be caught.
stopifnot(abs(cl_recovered - cl_published) / cl_published < 0.01)
```

Dose proportionality holds to within 1e-5 across a 17-fold dose range, as it
must for a model that is linear in dose, and `Dose / AUC0-inf` recovers the
published apparent clearance of 1.4 L/h to within 0.04% -- confirming that the
dose-unit conversion (I.U. to nmol) and the absorption / disposition
parameterisation are mutually consistent.

# Assumptions and deviations

* **IIV magnitudes are not published.** Results 2.2 states that IIV terms on
  C0, Vc/F and CL/F "were included", but Table 2's CV% column is the relative
  standard error of each estimate, not an OMEGA -- Results 2.2 reads those same
  numbers as precision ("estimated with adequate precision with the exception
  of Vm2 and Vm3", whose CV% are 206.8 and 140.5). No OMEGA is reported
  anywhere in the paper. The three etas are therefore carried at `fixed(0)`,
  preserving the random-effect structure without inventing variances. The
  packaged model is consequently deterministic, and a VPC in the style of
  Figure 3 cannot be reproduced.
* **The Results prose ng/mL values are internally inconsistent and are not
  used.** Results 2.2 gives each baseline in both units, but the pairs do not
  reconcile at the paper's own stated molecular weights: C0m1 is printed as
  "108.57 ng/mL (43.5 nmol/L)", yet 43.5 nmol/L is 17.4 ng/mL. The nmol/L
  column is authoritative on three independent grounds: it is the scale the
  model was fitted on (Methods 4.5); 43.5 nmol/L = 17.4 ng/mL matches the Table 1
  median baseline 25D3 of 18 ng/mL, whereas 108.57 ng/mL would put every subject
  six-fold above the study's own `<30 ng/mL` enrolment ceiling; and only the
  nmol/L reading reproduces the published six-month simulation endpoints. The
  same mismatch affects C0 (2.54 vs 0.377 ng/mL), C0m2 (0.48 vs 0.083 ng/mL) and
  C0m3 (printed as "528 ng/mL", which is neither the nmol/L equivalent nor a
  consistent pg/mL value).
* **Results-versus-Table conflict on the fixed parameters, resolved in favour of
  Table 2.** Results 2.2 states that ka, Vp/F and Q/F "were fixed to value from
  the literature or from previous iterations of the model" and quotes
  0.323 1/h, 2333 L and 0.185 L/h; Table 2 gives 0.054, 50 and 0.44. The
  Discussion resolves this explicitly, stating that the iteration-derived values
  "resulted in better model performance compared to using fixed values reported
  from the scarce literature" and restating ka = 0.054 1/h and Q/F = 0.44 L/h.
  Table 2 is used throughout.
* **`peripheral1` is initialised at distribution equilibrium.** Tuey 2024
  reports only the central baseline C0. The peripheral compartment is started at
  `C0 * Vp/F` (equal concentrations, the distribution-equilibrium condition),
  because a subject sitting at their lifelong endogenous baseline should not
  exhibit a distributional transient at t = 0. Starting it empty instead changes
  the six-month simulation endpoints by at most 0.21% (largest at the lowest
  dose, 600 I.U./day, and smaller at every higher dose), so this assumption is
  immaterial to the validation above.
* **The model is not at endogenous steady state at t = 0.** `ksyn` was fixed at
  0.55 nmol/h from an earlier model iteration while CL/F was subsequently
  estimated at 1.4 L/h, so the implied steady-state parent concentration
  (`ksyn / CL` = 0.393 nmol/L) does not equal the estimated baseline C0 of
  0.98 nmol/L, and an undosed profile drifts downward. This is a property of the
  published parameter set, faithfully reproduced rather than corrected; it is
  also why the NCA above subtracts a matched no-dose simulation instead of a
  constant baseline.
* **No covariates are implemented.** Tuey 2024 screened weight, BMI, age, sex,
  race, ethnicity, eGFR, six SNPs (CYP2R1, CYP27B1, CYP24A1, GC, and two VDR
  variants), PTH and FGF-23, and retained none (Results 2.3). The full screen is
  recorded in the model file's `covariatesDataExcluded` metadata so the
  provenance is preserved without declaring covariates the model never uses.
* **Dose units.** Doses are converted from I.U. to nmol using the definition
  1 I.U. = 0.025 ug cholecalciferol and the paper's molecular weight of
  384.64 g/mol, giving 325 nmol for the studied 5000 I.U. dose. Tuey 2024 does
  not print this conversion; it is the standard definition, and the resulting
  simulation reproduces the published endpoints.
* **Time-to-threshold is reported but not asserted on.** See the conditioning
  table above -- crossing-point comparisons on an asymptotically flat curve
  amplify small vertical errors, so the end-of-treatment plateau is used as the
  quantitative gate instead.
