---
title: "Edoxaban (Ueshima 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Edoxaban (Ueshima 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
mod <- readModelDb("Ueshima_2025_edoxaban")
ui  <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1186/s40780-025-00453-2> (open access, PMC12131839)

# Population

Ueshima 2025 is a retrospective real-world therapeutic-drug-monitoring study of
131 Japanese adults with atrial fibrillation treated with once-daily oral
edoxaban (Lixiana) at three Japanese centres between January 2017 and August
2019. Each patient contributed exactly **one** plasma sample, drawn at steady
state 9 to 47 h after the last dose, so the analysis dataset holds 131
observations from 131 patients and contains no absorption-phase information at
all. That single fact drives every structural choice in the model: NONMEM
ADVAN1 TRANS2 (one compartment, no absorption, dose straight into the central
compartment), an apparent volume of distribution fixed to an external value,
and an apparent clearance as the only estimated structural parameter.

Baseline characteristics (Ueshima 2025 Table 1, median with range): age 72.2
years (35.2-92.6), body weight 61.4 kg (37.3-97.3), sex 83 male / 48 female,
serum creatinine 0.89 mg/dL (0.56-2.00), Cockcroft-Gault creatinine clearance
61.8 mL/min (20.0-135.5), AST 25 IU/L (11-92), ALT 19 IU/L (5-77). Daily doses
were 15 mg (n = 3), 30 mg (n = 70) and 60 mg (n = 58). Sixteen patients took a
CYP3A4 and/or P-glycoprotein inhibitor (amiodarone 5, diltiazem 6, verapamil
5). Observed edoxaban concentrations were 19.5 ng/mL (1.7-152.0), with a lower
limit of quantification of 1 ng/mL.

The paper's headline results are **negative** pharmacogenomic findings:
`CYP3A5*3` (rs776746) and the `ABCB1` 1236C>T / 2677G>T,A / 3435C>T
polymorphisms, and the 2677/3435 haplotype, were all screened and none affected
edoxaban pharmacokinetics. Those screened-but-not-retained covariates are
recorded in the model file's `covariatesDataExcluded` list so the provenance of
the screen survives with the model.

```{r population}
str(ui$population)
```

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F at CRCL 61.8 mL/min) | 28.2 L/h | Table 2, `theta1` (95% CI 26.9-29.5; bootstrap median 28.1); Eq. 10 |
| `e_crcl_cl` (power exponent on CRCL/61.8) | 0.692 | Table 2, `theta3` (95% CI 0.582-0.801; bootstrap median 0.693); Eq. 10 |
| `lvc` (Vd/F at WT 70 kg) | 336.4 L, fixed | Table 2, `theta2` ("336.4 fixed"); Methods Eq. 3 = 193 L central volume x 1.743 Asian fold-change from reference [22] (Krekels 2016, ENGAGE AF-TIMI 48) |
| `e_wt_vc` (exponent on WT/70) | 1, fixed | Methods, PPK modeling: "with the exponent fixed to 1 due to a lack of individual data"; Eq. 3 |
| `etalcl` (IIV on CL/F) | 26.4 CV%, variance 0.0697 | Table 2, `omega1` (95% CI 18.8-34.0; bootstrap median 25.9). Footnote c: the parenthesised 27.4 is the eta SHRINKAGE, not an RSE |
| `expSd` (residual) | 58.7 CV%, fixed | Table 2, `sigma` ("58.7 fixed"); Methods Eq. 1 and text: 14.6% (healthy subjects) + 44.1% (AF-patient increment), both from reference [22] |
| `d/dt(central)` (1 cmt, no absorption) | n/a | Methods, PPK modeling: "a 1-compartment model without first-order absorption was employed ... (ADVAN1 TRANS2)" |
| `Cc = central / vc * 1000` | n/a | Unit conversion only: mg / L = ug/mL; the paper reports ng/mL throughout (Table 1, Figs. 2-3) |
| Reference CRCL 61.8 mL/min | n/a | Table 1 cohort median, and printed inside Eq. 10 and the Table 2 parameter header |
| Reference WT 70 kg | n/a | Eq. 3 (inherited from the ENGAGE AF-TIMI 48 parameterisation, not a cohort median) |

Note on the variability scale: Table 2 labels the two rows `omega1 (CV%)` and
`sigma (CV%)`, and footnote b states they are "the coefficient of variations of
the inter-individual variability for CL/F and residual variability". The
reported number is therefore the standard-deviation parameter itself expressed
as a percentage -- the usual NONMEM `%CV = 100 * sqrt(omega^2)` convention --
so `omega1 = 0.264` (variance 0.0697) and `sigma = 0.587` on the log scale.
The "two variability terms" gate below tests that reading term by term.

# Structural check against the paper's own printed clearances

The Discussion prints CL/F for three renal-function scenarios, computed by the
authors from Eq. 10. Those three numbers are an exact answer key for the
transcription of `lcl` and `e_crcl_cl`, and they are deterministic -- no
simulation, no cohort -- so the assertion can be exact.

```{r clearance-answer-key}
cl_paper <- function(crcl) 28.2 * (crcl / 61.8)^0.692   # Ueshima 2025 Eq. 10
v_paper  <- function(wt)   336.4 * (wt / 70)            # Ueshima 2025 Eq. 3

answer_key <- tibble::tibble(
  CRCL          = c(100, 60, 30),
  scenario      = c("Normal renal function", "Mild renal impairment",
                    "Severe renal impairment"),
  published_cl  = c(39.3, 27.6, 17.1),   # Ueshima 2025 Discussion, paragraph 3
  model_cl      = round(cl_paper(c(100, 60, 30)), 1)
)
knitr::kable(
  answer_key |>
    dplyr::rename(
      "CLcr (mL/min)"        = CRCL,
      "Scenario"             = scenario,
      "Published CL/F (L/h)" = published_cl,
      "Encoded CL/F (L/h)"   = model_cl
    ),
  caption = "Ueshima 2025 Discussion CL/F answer key vs. the encoded equation."
)

stopifnot(identical(answer_key$model_cl, answer_key$published_cl))
```

The paper additionally compares those to the ENGAGE AF-TIMI 48 values
(34.6 / 24.8 / 18.5 L/h); we do not assert against those, since they come from
a different model.

# Virtual cohort

Original observed data are not publicly available. Two virtual cohorts are used
below.

The first replicates the paper's own Monte Carlo design (Ueshima 2025,
Model-based simulation): patients at fixed creatinine clearances of 30, 60, 90
and 120 mL/min, dosed 30 mg once daily at CLcr 30 and 60 mg once daily
otherwise, per the Japanese package insert. The paper does not state the body
weight used for that simulation; 70 kg (the model's own reference weight, so
Vd/F = 336.4 L exactly) reproduces the Figure 3 medians and is used here -- see
Assumptions and deviations.

Steady state is imposed directly with `ss = 1`, `ii = 24` rather than by
simulating a burn-in, so the profile is the exact steady-state solution.

```{r cohort}
# set.seed() seeds R's RNG, NOT rxode2's simulation RNG, and rxode2's streams
# are partitioned per solver thread -- so this cohort differs between a
# 16-thread workstation and a 2-core CI runner. Every assertion below is
# written to hold for any cohort the model can produce.
set.seed(20250908)

n_per_arm <- 200L
tau       <- 24

arms <- tibble::tibble(
  arm  = factor(
    c("CLcr 30, 30 mg", "CLcr 60, 60 mg", "CLcr 90, 60 mg", "CLcr 120, 60 mg"),
    levels = c("CLcr 30, 30 mg", "CLcr 60, 60 mg", "CLcr 90, 60 mg", "CLcr 120, 60 mg")
  ),
  CRCL = c(30, 60, 90, 120),
  dose = c(30, 60, 60, 60)
)

make_arm <- function(i, id_offset) {
  ids <- id_offset + seq_len(n_per_arm)
  covs <- list(WT = 70, CRCL = arms$CRCL[i], arm = arms$arm[i])
  dosing <- tibble::tibble(
    id = ids, time = 0, amt = arms$dose[i], evid = 1L, cmt = "central",
    ii = tau, ss = 1L
  )
  obs <- tidyr::crossing(id = ids, time = seq(0, tau, by = 0.25)) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", ii = 0, ss = 0L)
  dplyr::bind_rows(dosing, obs) |>
    dplyr::mutate(WT = covs$WT, CRCL = covs$CRCL, arm = covs$arm) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(
  lapply(seq_len(nrow(arms)), function(i) make_arm(i, (i - 1L) * n_per_arm))
)
stopifnot(
  !anyDuplicated(unique(events[, c("id", "time", "evid")])),
  length(unique(events$id)) == n_per_arm * nrow(arms)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = as.data.frame(events),
                       keep = c("arm", "WT", "CRCL")) |>
  as.data.frame()

# `Cc` is the individual prediction (structural model + eta, no residual
# error); `sim` additionally carries the log-normal residual and is therefore
# the column that corresponds to a simulated OBSERVED concentration, which is
# what Ueshima 2025 Figure 3 plots.
stopifnot(all(c("Cc", "sim", "cl", "vc") %in% names(sim)), all(sim$Cc > 0))
```

## Gate: the packaged model really is the paper's ODE

The model writes an explicit `d/dt(central)` but names its parameters `cl` and
`vc`, which lets rxode2 recognise a one-compartment linear system and solve it
analytically. That substitution is only safe because the ODE's rate constant IS
`cl / vc`; the two checks below confirm the packaged object behaves as the
paper's equations say it should, rather than silently solving something else.

```{r gate-two-solve}
# 1. Typical-value profile must equal the analytical steady-state solution of
#    Eqs. 3 and 10. Both sides use the same parameters, so the only difference
#    is numerical -- a tight bound is correct here.
mod0 <- rxode2::zeroRe(mod)
# One representative subject per arm (the first ID of each arm's block).
typical_ids <- (seq_len(nrow(arms)) - 1L) * n_per_arm + 1L
sim0 <- rxode2::rxSolve(
  mod0,
  events = as.data.frame(dplyr::filter(events, id %in% typical_ids)),
  keep = c("arm", "WT", "CRCL")
) |>
  as.data.frame()

closed_form <- function(t, dose, crcl, wt = 70) {
  k <- cl_paper(crcl) / v_paper(wt)
  (dose / v_paper(wt)) * 1000 * exp(-k * t) / (1 - exp(-k * tau))
}

sim0 <- sim0 |>
  dplyr::mutate(
    dose     = arms$dose[match(CRCL, arms$CRCL)],
    analytic = closed_form(time, dose, CRCL),
    pct_diff = 100 * (Cc - analytic) / analytic
  )
cat(sprintf("max |%% diff| typical-value solve vs analytical solution: %.5f%%\n",
            max(abs(sim0$pct_diff))))
stopifnot(max(abs(sim0$pct_diff)) < 0.01)

# 2. Perturbing the estimated clearance must move the profile. A model whose
#    ODE had been discarded in favour of an unrelated solved form would not
#    respond correctly to this.
mod_slow <- rxode2::zeroRe(mod) |> rxode2::ini(lcl = log(28.2 / 2))
sim_slow <- rxode2::rxSolve(
  mod_slow,
  events = as.data.frame(dplyr::filter(events, id == n_per_arm + 1L))
) |>
  as.data.frame()
trough_ref  <- sim0$Cc[sim0$CRCL == 60 & abs(sim0$time - tau) < 1e-8]
trough_slow <- sim_slow$Cc[abs(sim_slow$time - tau) < 1e-8]
# Halving CL/F raises the 24 h trough of a mono-exponential; the analytical
# ratio for this arm is deterministic, so assert it exactly.
ratio_expected <- closed_form(tau, 60, 60) / {
  k2 <- (cl_paper(60) / 2) / v_paper(70)
  (60 / v_paper(70)) * 1000 * exp(-k2 * tau) / (1 - exp(-k2 * tau))
}
cat(sprintf("trough ratio (published CL / halved CL): observed %.4f, analytic %.4f\n",
            trough_ref / trough_slow, ratio_expected))
stopifnot(abs(trough_ref / trough_slow / ratio_expected - 1) < 1e-3)
```

# Replicate Figure 3 -- simulated steady-state trough concentrations

Ueshima 2025 Figure 3 plots box-and-whisker summaries of the predicted trough
concentration 24 h after the last dose at four creatinine clearances, over the
grey band and dashed line that mark the interquartile range (13.9-47.4 ng/mL)
and median (25.0 ng/mL) of observed trough concentrations in Asians reported by
Chao et al. (reference [10]).

```{r figure-3, fig.cap = "Replicates Figure 3 of Ueshima 2025."}
chao <- list(median = 25.0, q25 = 13.9, q75 = 47.4)   # Ueshima 2025 Results / Fig. 3 caption
lloq <- 1.0                                            # Ueshima 2025 Edoxaban assay

trough <- sim |>
  dplyr::filter(abs(time - tau) < 1e-8) |>
  dplyr::mutate(dose = arms$dose[match(CRCL, arms$CRCL)])

ggplot(trough, aes(x = arm, y = sim)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = chao$q25, ymax = chao$q75,
           fill = "grey70", alpha = 0.45) +
  geom_hline(yintercept = chao$median, linetype = "dashed") +
  geom_hline(yintercept = lloq, linetype = "dotdash") +
  geom_boxplot(width = 0.5, outlier.size = 0.6) +
  scale_y_log10() +
  labs(x = NULL, y = "Predicted trough concentration (ng/mL)",
       title = "Figure 3 - steady-state trough by creatinine clearance",
       caption = paste("Replicates Figure 3 of Ueshima 2025. Grey band and dashed line:",
                       "observed IQR and median in Asians (Chao et al.).",
                       "Dot-dash line: assay LLOQ 1 ng/mL.")) +
  theme_bw()
```

```{r figure-3-numbers}
fig3 <- trough |>
  dplyr::group_by(arm, CRCL, dose) |>
  dplyr::summarise(
    median_sim = median(sim),
    q25_sim    = quantile(sim, 0.25),
    q75_sim    = quantile(sim, 0.75),
    .groups    = "drop"
  ) |>
  dplyr::mutate(
    typical  = closed_form(tau, dose, CRCL),
    pct_diff = 100 * (median_sim - typical) / typical,
    iqr_over_median = (q75_sim - q25_sim) / median_sim
  )
knitr::kable(
  fig3 |>
    dplyr::mutate(dplyr::across(median_sim:iqr_over_median, \(x) round(x, 3))) |>
    dplyr::rename(
      "Arm"                        = arm,
      "CLcr (mL/min)"              = CRCL,
      "Dose (mg)"                  = dose,
      "Simulated median (ng/mL)"   = median_sim,
      "Simulated Q1 (ng/mL)"       = q25_sim,
      "Simulated Q3 (ng/mL)"       = q75_sim,
      "Typical value (ng/mL)"      = typical,
      "% diff (median vs typical)" = pct_diff,
      "IQR / median"               = iqr_over_median
    ),
  caption = "Simulated steady-state trough concentrations and the analytical typical value."
)

# The cohort median estimates the typical value; assert on the CENTRE, with
# headroom for which subjects a given rxode2 build / thread count happens to
# draw. 200 subjects with a total log-scale SD of ~0.86 give a median standard
# error of about 8%.
stopifnot(max(abs(fig3$pct_diff)) < 25)
```

## Gate: the two variability terms reach the simulation at the encoded scale

Reading `omega1 (CV%) = 26.4` and `sigma (CV%) = 58.7` as standard deviations
rather than as lognormal coefficients of variation (or, worse, as variances) is
the one transcription decision in Table 2 that a structural check cannot catch.
Three checks pin it down. Note that the trough is a strongly non-linear
function of clearance -- `d log C(24) / d log CL = -24k / (1 - exp(-24k))`,
about -2.3 for the 60 mL/min arm -- so the trough spread is roughly 2.4x the
clearance spread and cannot be compared to `omega1` directly. Test each term
where it enters instead.

```{r gate-variability}
# A. Every individual profile is the closed-form solution evaluated at that
#    individual's own cl and vc. This is deterministic per subject, so the
#    bound is numerical.
ind <- sim |>
  dplyr::filter(abs(time - tau) < 1e-8) |>
  dplyr::mutate(
    dose     = arms$dose[match(CRCL, arms$CRCL)],
    k_i      = cl / vc,
    analytic = (dose / vc) * 1000 * exp(-k_i * tau) / (1 - exp(-k_i * tau)),
    rel      = Cc / analytic - 1
  )
cat(sprintf("max |Cc / individual closed form - 1| over %d subjects: %.3g\n",
            nrow(ind), max(abs(ind$rel))))
stopifnot(max(abs(ind$rel)) < 1e-4)

# B. IIV enters on cl only, so sd(log(cl)) recovers omega1 = 0.264 directly.
sd_log_cl <- sd(log(ind$cl / (28.2 * (ind$CRCL / 61.8)^0.692)))
cat(sprintf("sd(log eta multiplier on CL/F): %.4f (encoded omega1 = 0.264)\n", sd_log_cl))
stopifnot(abs(sd_log_cl / 0.264 - 1) < 0.15)

# C. `sim` is `Cc` times the log-normal residual, so their log ratio recovers
#    sigma = 0.587 directly.
sd_log_eps <- sd(log(ind$sim / ind$Cc))
cat(sprintf("sd(log(sim / Cc)): %.4f (encoded sigma = 0.587)\n", sd_log_eps))
stopifnot(abs(sd_log_eps / 0.587 - 1) < 0.15)
```

Ueshima 2025 (Discussion, final paragraph) reports that the predicted troughs at
CLcr 90 and 120 mL/min "tended to be lower than the median of previously
observed concentrations", while the lower-clearance arms sat within the
observed range. Both halves of that claim are reproduced:

```{r figure-3-claim}
claim <- fig3 |>
  dplyr::mutate(
    within_chao_iqr = median_sim > chao$q25 & median_sim < chao$q75,
    below_chao_med  = median_sim < chao$median
  )
knitr::kable(
  claim |>
    dplyr::select(arm, median_sim, within_chao_iqr, below_chao_med) |>
    dplyr::mutate(median_sim = round(median_sim, 1)) |>
    dplyr::rename(
      "Arm"                       = arm,
      "Simulated median (ng/mL)"  = median_sim,
      "Within Chao IQR 13.9-47.4" = within_chao_iqr,
      "Below Chao median 25.0"    = below_chao_med
    ),
  caption = "Ueshima 2025 Discussion claim about Figure 3, reproduced."
)
stopifnot(
  claim$within_chao_iqr[claim$CRCL %in% c(30, 60)],
  claim$below_chao_med[claim$CRCL %in% c(90, 120)]
)
```

# PKNCA validation

Steady-state non-compartmental analysis over the 24 h dosing interval. The
model is dosed directly into the central compartment, so the time-zero record
is the steady-state peak, not a pre-dose zero -- no synthetic `Cc = 0` row is
added, and its presence is asserted rather than assumed.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Every subject must already carry a record at the interval start; PKNCA
# otherwise silently anchors AUC at the first available time.
stopifnot(
  all(tapply(sim_nca$time, sim_nca$id, \(x) any(abs(x) < 1e-8))),
  all(tapply(sim_nca$time, sim_nca$id, \(x) any(abs(x - tau) < 1e-8)))
)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start     = 0,
  end       = tau,
  cmax      = TRUE,
  tmax      = TRUE,
  cmin      = TRUE,
  auclast   = TRUE,
  cav       = TRUE,
  ctrough      = TRUE,
  half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against the paper's printed equations

Ueshima 2025 publishes no NCA table -- it is a sparse-sampling
therapeutic-drug-monitoring study with one concentration per patient. The
reference column below is therefore the **analytical steady-state solution of
the paper's own printed equations** (Eq. 3 and Eq. 10), which is the thing this
vignette exists to check: does the packaged encoding reproduce what the paper
printed? For a one-compartment model dosed into the central compartment,

* `AUC0-24,ss` = 1000 x Dose / (CL/F),
* `Cmax,ss` = 1000 x (Dose / (Vd/F)) / (1 - exp(-k x 24)) at `Tmax = 0`,
* `Ctrough` = `Cmin` = `Cmax,ss` x exp(-k x 24),
* `Cav` = `AUC0-24,ss` / 24, and
* `t1/2` = log(2) / k, with `k` = (CL/F) / (Vd/F).

The comparison uses the typical-value (zero random effects) simulation, so both
sides are deterministic and the agreement should be numerical-precision tight.

```{r pknca-typical}
nca_typ_conc <- sim0 |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)
nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(nca_typ_conc, Cc ~ time | arm + id),
  PKNCA::PKNCAdose(
    events |>
      dplyr::filter(evid == 1, id %in% nca_typ_conc$id) |>
      dplyr::select(id, time, amt, arm),
    amt ~ time | arm + id
  ),
  intervals = intervals
))

published_eq <- arms |>
  dplyr::mutate(
    k         = cl_paper(CRCL) / v_paper(70),
    auclast   = 1000 * dose / cl_paper(CRCL),
    cmax      = 1000 * (dose / v_paper(70)) / (1 - exp(-k * tau)),
    tmax      = 0,
    cmin      = cmax * exp(-k * tau),
    ctrough      = cmin,
    cav       = auclast / tau,
    half.life = log(2) / k
  ) |>
  dplyr::select(arm, cmax, tmax, cmin, ctrough, cav, auclast, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_typ,
  reference     = published_eq,
  by            = "arm",
  units         = c(cmax = "ng/mL", cmin = "ng/mL", ctrough = "ng/mL",
                    cav = "ng/mL", auclast = "ng*h/mL",
                    tmax = "h", half.life = "h"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste("Typical-value steady-state NCA vs. the analytical solution of",
                  "Ueshima 2025 Eqs. 3 and 10. * differs by >20%."),
  align   = c("l", "l", "r", "r", "r")
)

# `% diff` from ncaComparisonTable() is FORMATTED CHARACTER, so assert on
# numbers recomputed here rather than parsing the display column.
typ_wide <- as.data.frame(nca_typ) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
chk <- published_eq |>
  dplyr::left_join(typ_wide, by = "arm", suffix = c("_ref", "_sim"))

# Drive the comparison off an explicit parameter list and demand that both
# columns exist for every one of them. A silently-missing column would
# otherwise drop that parameter from `worst` and leave a gate that cannot go
# red.
compare_params <- c("cmax", "cmin", "ctrough", "cav", "auclast", "half.life")
pct_by_param <- vapply(compare_params, function(p) {
  ref <- chk[[paste0(p, "_ref")]]
  sm  <- chk[[paste0(p, "_sim")]]
  if (is.null(ref) || is.null(sm) || anyNA(ref) || anyNA(sm)) {
    stop("missing or NA reference/simulated column for '", p, "'")
  }
  max(abs(100 * (sm - ref) / ref))
}, numeric(1))
print(round(pct_by_param, 4))
worst <- max(pct_by_param)
cat(sprintf("worst |%% diff| across %s: %.4f%%\n",
            paste(compare_params, collapse = ", "), worst))
stopifnot(worst < 0.5, all(chk$tmax_sim == 0))
```

## Cohort-level NCA

The same NCA over the full 200-subject-per-arm cohort. Between-subject
variability is on clearance only, so exposure metrics vary while `Tmax` and
`Cmax` per unit volume do not.

```{r pknca-cohort}
nca_summary <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "ctrough", "auclast", "half.life")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    median = median(PPORRES),
    q05    = quantile(PPORRES, 0.05),
    q95    = quantile(PPORRES, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    Parameter = nlmixr2lib::ncaParamLabel(
      PPTESTCD,
      units = c(cmax = "ng/mL", ctrough = "ng/mL",
                auclast = "ng*h/mL", half.life = "h")
    )
  ) |>
  dplyr::select(arm, Parameter, median, q05, q95)

knitr::kable(
  nca_summary |>
    dplyr::mutate(dplyr::across(median:q95, \(x) signif(x, 4))) |>
    dplyr::rename("Arm" = arm, "Median" = median,
                  "5th percentile" = q05, "95th percentile" = q95),
  caption = "Cohort steady-state NCA, median and 90% interval across subjects."
)

# The cohort median AUC estimates 1000 * Dose / CL/F. Assert on the centre.
auc_med <- nca_summary |>
  dplyr::filter(grepl("^AUC", Parameter)) |>
  dplyr::left_join(published_eq |> dplyr::select(arm, auclast), by = "arm") |>
  dplyr::mutate(pct = 100 * (median - auclast) / auclast)
cat(sprintf("AUC0-24 cohort median vs analytical, %% diff by arm: %s\n",
            paste(round(auc_med$pct, 1), collapse = ", ")))
stopifnot(max(abs(auc_med$pct)) < 25)
```

# Replicate Figure 2 -- concentration vs. time after last dose

Ueshima 2025 Figure 2 is a prediction-corrected VPC of the observed
concentrations against time after the last dose, with the 5th, 50th and 95th
percentiles of 1,000 simulated datasets. The panel below simulates the study
cohort itself -- 131 patients at the observed dose split, with body weight and
creatinine clearance drawn to match the Table 1 medians and ranges -- and plots
the same three percentiles.

```{r figure-2, fig.cap = "Replicates Figure 2 of Ueshima 2025."}
n_study <- 131L
# Log-normal draws centred on the Table 1 medians, truncated to the Table 1
# ranges. The paper reports only median and range, so the dispersion is an
# assumption; see Assumptions and deviations.
rtrunc_lnorm <- function(n, med, lo, hi, sdlog) {
  x <- rep(NA_real_, n)
  todo <- seq_len(n)
  while (length(todo)) {
    cand <- med * exp(stats::rnorm(length(todo), 0, sdlog))
    ok <- cand >= lo & cand <= hi
    x[todo[ok]] <- cand[ok]
    todo <- todo[!ok]
  }
  x
}

study <- tibble::tibble(
  id   = seq_len(n_study),
  WT   = rtrunc_lnorm(n_study, 61.4, 37.3, 97.3, 0.17),
  CRCL = rtrunc_lnorm(n_study, 61.8, 20.0, 135.5, 0.38),
  amt  = rep(c(15, 30, 60), times = c(3L, 70L, 58L))   # Ueshima 2025 Table 1
)

study_ev <- dplyr::bind_rows(
  study |> dplyr::mutate(time = 0, evid = 1L, cmt = "central", ii = tau, ss = 1L),
  tidyr::crossing(id = study$id, time = seq(0, 48, by = 0.5)) |>
    dplyr::left_join(study |> dplyr::select(id, WT, CRCL), by = "id") |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", ii = 0, ss = 0L)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# Observations beyond 24 h describe the profile after the LAST dose, i.e. a
# missed / delayed dose -- which is exactly the situation Figure 2's 24-47 h
# samples represent -- so dosing is not continued past time 0.
study_sim <- rxode2::rxSolve(mod, events = as.data.frame(study_ev),
                             keep = c("WT", "CRCL")) |>
  as.data.frame()

study_pct <- study_sim |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(sim, 0.05), Q50 = quantile(sim, 0.50),
    Q95 = quantile(sim, 0.95), .groups = "drop"
  )

ggplot(study_pct, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = lloq, linetype = "dotdash") +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 48, by = 12)) +
  labs(x = "Time after last dose (h)", y = "Edoxaban concentration (ng/mL)",
       title = "Figure 2 - simulated 5th / 50th / 95th percentiles",
       caption = paste("Replicates Figure 2 of Ueshima 2025.",
                       "Dot-dash line: assay LLOQ 1 ng/mL.")) +
  theme_bw()
```

```{r figure-2-check}
# Ueshima 2025 Table 1: the 131 observed concentrations, sampled 9-47 h after
# the last dose, had a median of 19.5 ng/mL and a range of 1.7-152.0.
#
# This CANNOT be turned into a point comparison of medians. The observed median
# is taken over the sampling times that actually occurred, and the paper does
# not report their distribution; Figure 2 shows them clustered near 24 h with a
# tail out to 48 h and a group near 12 h. The grid below weights 9-47 h
# uniformly, which puts more mass on the late, low-concentration end than the
# real sampling did, so the simulated median sits below the observed one by
# construction. What the simulation can legitimately be asked is whether the
# published summary statistics fall inside the distribution it produces.
window <- study_sim |> dplyr::filter(time >= 9, time <= 47) |> dplyr::pull(sim)
window_q <- quantile(window, c(0.05, 0.25, 0.50, 0.75, 0.95))
print(round(window_q, 1))
cat(sprintf("simulated median over the 9-47 h window: %.1f ng/mL (observed 19.5; %.1f-%.1f is the simulated IQR)\n",
            window_q[["50%"]], window_q[["25%"]], window_q[["75%"]]))

# Containment checks, all robust to which cohort a given machine draws:
stopifnot(
  # The observed median lies inside the simulated central 90%.
  19.5 > window_q[["5%"]], 19.5 < window_q[["95%"]],
  # The simulated distribution spans the observed range rather than sitting
  # entirely inside or entirely outside it.
  min(window) < 1.7, max(window) > 152.0
)

# Trough-time comparison, which removes the sampling-time weighting problem:
# the median 24 h concentration should be the same order as the observed
# median, since 24 h is where the paper's sampling density peaks.
med24 <- median(study_sim$sim[abs(study_sim$time - tau) < 1e-8])
cat(sprintf("simulated median at 24 h: %.1f ng/mL\n", med24))
stopifnot(med24 > 19.5 / 3, med24 < 19.5 * 3)
```

# Assumptions and deviations

* **Body weight for the Figure 3 replication.** Ueshima 2025 does not state the
  body weight used in its Monte Carlo simulation. 70 kg is used here, which is
  the model's own reference weight and therefore gives Vd/F = 336.4 L exactly;
  it reproduces the Figure 3 medians. The cohort median weight is 61.4 kg,
  which would lower Vd/F to 295 L and raise the trough by about 20%.
* **Covariate distributions for the Figure 2 replication.** Table 1 reports only
  medians and ranges for body weight and creatinine clearance, so the log-normal
  dispersions (`sdlog` 0.17 and 0.38) are chosen to fill the published ranges
  and are an assumption, not a published quantity. Weight and creatinine
  clearance are drawn independently even though the Cockcroft-Gault equation
  makes them correlated in reality; this widens the simulated spread slightly.
* **No non-renal clearance arm.** The paper fitted CL/F as the sum of an
  apparent renal and an apparent non-renal component (Eq. 4), but the non-renal
  population mean converged to 0.01 L/h and was judged negligible, so the
  published final model (Eq. 10) is the single power term encoded here. The
  paper itself notes this is likely an underestimate -- an intravenous study
  puts renal clearance at 49.1% of total -- because no urine or metabolite
  samples were collected. A user applying this model outside the fitted
  creatinine-clearance range of 20.0-135.5 mL/min will therefore extrapolate a
  term that carries the whole of clearance.
* **Vd/F and the residual error are not this paper's estimates.** Both were
  fixed to values from Krekels 2016 (ENGAGE AF-TIMI 48, reference [22]):
  Vd/F = 336.4 L (193 L x 1.743 Asian fold-change) and sigma = 58.7 CV%
  (14.6% healthy-subject residual + 44.1% AF-patient increment). They are
  encoded with `fixed()` and are recorded as such in the model file. The
  weight exponent on Vd/F is likewise `fixed(1)`, not an allometric 0.75 or 1
  chosen by this analysis -- the paper fixed it "due to a lack of individual
  data".
* **No absorption, and therefore no `Tmax`.** The model dose enters `central`
  directly (NONMEM ADVAN1 TRANS2), so the simulated `Tmax` is 0 and `Cmax` is
  the immediate post-dose concentration. This is a property of the published
  model, not of the encoding: with only one sample per patient on the
  elimination phase, no absorption parameter was identifiable. Simulated peak
  concentrations should not be compared to observed edoxaban `Cmax`, which
  occurs 1-2 h post-dose in richly sampled studies.
* **Pharmacogenomic covariates are documented, not implemented.** `CYP3A5*3`
  and the three `ABCB1` polymorphisms, the 2677/3435 haplotype, sex, age, AST,
  ALT and concomitant CYP3A4 / P-glycoprotein inhibitor use were all screened
  and none was retained. They appear in the model file's
  `covariatesDataExcluded` metadata and are absent from `model()`. The paper
  flags the small number of exposed patients (2 with `CYP3A5*1/*1`, 16 on an
  inhibitor) as a limitation of the negative co-medication and genotype
  findings.
* **Figure 3 arms are simulated at steady state via `ss = 1`** rather than by a
  multi-day burn-in, which gives the exact steady-state solution rather than an
  approach to it.
