---
title: "B cell immune response QSP, mouse (Ugolkov 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{B cell immune response QSP, mouse (Ugolkov 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Ugolkov_2026_bcell_mouse_qsp"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.3389/fimmu.2026.1745710>
* Supplement (equations, RxODE code, Table S1): <https://www.frontiersin.org/articles/10.3389/fimmu.2026.1745710/full#supplementary-material>

`r ui$description`

This is an endogenous / mechanistic systems model: there is no drug, no dosing
event and no concentration-time profile, so PKNCA is not the right validation
target. The vignette instead follows the endogenous validation pattern --
steady-state hold, mass balance, dimensional analysis, perturbation recovery --
and then replicates the paper's published figures.

# Population

The model was built from aggregated data across 21 published mouse studies
identified by a systematic PubMed / Google Scholar search (Supplementary
Table S1); no individual-animal data were used, so there are no per-subject
demographics. Strains span C57BL/6 (with C57BL/6J, CBA/J, CBA/N) and BALB/c
(with BALB/cJRj), ages 5-24 weeks where reported, both sexes. Studies of
genetically modified or diseased animals were excluded.

Data entered the model in two roles. Steady-state counts of immature,
transitional type 1 (T1) and naive B cells in bone marrow, blood, spleen and
lymph nodes parameterised the homeostasis layer. Longitudinal
antibody-secreting cell (ASC) counts following immunization -- with virus-like
particles, LCMV, *Plasmodium chabaudi*, sheep red blood cells, keyhole limpet
hemocyanin and NP-ova/CFA, given intraperitoneally, intravenously or
intradermally -- calibrated the activation layer. Plasma IgG kinetics from five
further studies were held out for external validation.

Two tissue-scaling conventions are baked into the calibration targets and are
reproduced by the model: bone marrow ASC counts are two-femur counts multiplied
by 7.9 (a femur is 12.7% of total murine marrow), and lymph node ASC counts are
single-node counts multiplied by 22, the number of murine lymph nodes.

```{r population}
str(ui$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/endogenous/Ugolkov_2026_bcell_mouse_qsp.R`. The tables
below collect the equations and parameters in one place.

## Equations

All 20 ODEs come from the Supplementary Equations, which the model file cites
line by line.

| Model line | Source | Content |
|---|---|---|
| `kmat_1`, `k1`, `k3`, `kmat_2`, `k7`, `k5`, `kdeg_spl` | Supp Eq 1-7 | The seven parameters solved analytically from steady state |
| `d/dt(imm_bone)` | Supp Eq 8 | Immature B cell generation and maturation in bone marrow |
| `d/dt(t1_bone)` | Supp Eq 9 | T1 B cells in bone marrow |
| `d/dt(t1_blood)` | Supp Eq 10 | T1 B cells in blood |
| `d/dt(t1_spleen)` | Supp Eq 11 | T1 B cells in spleen |
| `d/dt(naive_blood)` | Supp Eq 12 | Naive B cells in blood |
| `d/dt(naive_spleen)` | Supp Eq 13 | Naive B cells in spleen |
| `d/dt(naive_ln)` | Supp Eq 14 | Naive B cells in lymph nodes |
| `d/dt(antigen0)` | Supp Eq 15 | Empirical antigen forcing function |
| `d/dt(antigen1..5)`, `d/dt(antigen)` | Supp Eq 16-21 | Six-compartment antigen transit chain |
| `d/dt(asc_spleen)` | Supp Eq 22 | ASC generation and exchange, spleen |
| `d/dt(asc_ln)` | Supp Eq 23 | ASC generation and exchange, lymph nodes |
| `d/dt(asc_bone)` | Supp Eq 24 | Saturable ASC influx into the bone marrow niche |
| `d/dt(asc_blood)` | Supp Eq 25 | ASC in blood |
| `d/dt(asc_peripheral)` | Supp Eq 26 | ASC in non-lymphoid peripheral tissues |
| `ascTotal` | Supp Eq 27 | ASC pool driving systemic IgG |
| `d/dt(igg_blood)` | Supp Eq 28 | Antigen-specific plasma IgG |

## Parameters

```{r param-table}
ui$iniDf |>
  dplyr::filter(!is.na(.data$ntheta)) |>
  dplyr::transmute(
    Parameter = .data$name,
    `Value in file` = signif(.data$est, 6),
    Estimated = ifelse(.data$fix, "fixed", "estimated"),
    Label = .data$label
  ) |>
  knitr::kable(caption = "ini() entries. Log-transformed entries (l-prefixed) are the five parameters Ugolkov 2026 estimated; every other entry is fixed or a steady-state anchor.")
```

The seven parameters derived analytically inside `model()` are not `ini()`
entries. Their values, and how they compare with Table 1, are shown in the
Errata section below.

## Units

Two amount scales coexist, exactly as in the deposited RxODE code, whose
comments read `#cell / 10^6` for the B cell anchors.

| Quantity | Units | Notes |
|---|---|---|
| time | day | |
| `imm_bone`, `t1_*`, `naive_*` | 10^6 cell | Table 1 prints these as e.g. `1.97 * 10^6 cell` |
| `antigen0..5`, `antigen` | dimensionless | Empirical forcing function, no units assigned |
| `asc_spleen`, `asc_ln`, `asc_bone`, `asc_blood`, `asc_peripheral` | cell | Absolute counts |
| `igg_blood` | pg | |
| `k1`-`k14`, `kmat_1..3`, `kdeg_spl`, `kdeg_ln`, `kdeg_igg` | 1/day | |
| `ksyn_imm` | 10^6 cell/day | Table 1 prints `20 * 10^6 cell/day` |
| `ksin_igg` | pg/cell/day | |
| `vmax` | cell/day | |
| `khalf` | cell | |
| `mtt` | day | |

The scale conversion happens exactly once, in Supp Eq 22-23. The paper writes
the activated precursor pool as `NaiveSpleen / 10^6` with `NaiveSpleen` in
absolute cells, the `10^6` being the physiological antigen-specific precursor
frequency of 1 in 10^6. Because `naive_spleen` is already carried in units of
10^6 cells, that division is already applied and `kmat_3 * antigen *
naive_spleen` correctly reads as an absolute number of activated precursors per
day -- which is how the deposited code writes it.

## Dimensional analysis

Checking the two structurally interesting lines.

* **Supp Eq 22.** `kmat_3 * antigen * naive_spleen` is
  `(1/day) * (1) * (10^6 cell)`. With `naive_spleen` in units of 10^6 cells,
  the numeric value equals `NaiveSpleen_absolute / 10^6`, i.e. the count of
  antigen-specific precursors, so the term is `cell/day` -- matching
  `d/dt(asc_spleen)` in `cell/day`. `k8 * asc_spleen` and `k9 * asc_blood` are
  `(1/day) * cell`, also `cell/day`. Consistent.
* **Supp Eq 24.** `vmax * asc_blood / (khalf + asc_blood)` is
  `(cell/day) * cell / cell = cell/day`; `k12 * asc_bone` is `(1/day) * cell`.
  Consistent. Note that `khalf` carries units of cells, so the ratio is
  dimensionless, and `vmax` alone sets the maximum influx rate.
* **Supp Eq 28.** `ksin_igg * ascTotal` is `(pg/cell/day) * cell = pg/day`;
  `kdeg_igg * igg_blood` is `(1/day) * pg`. Consistent.

# Steady-state and mass balance

Without perturbation the homeostasis layer should hold at the observed
steady-state counts indefinitely. The antigen response never depletes naive B
cells (the paper omits that term deliberately: at a precursor frequency of
1 in 10^6 the loss is negligible), so the seven B cell states should be flat
across the whole 350-day simulation.

```{r steady-state}
times <- seq(0, 350, by = 0.5)
sim <- rxode2::rxSolve(
  ui, rxode2::et(times),
  atol = 1e-10, rtol = 1e-8, maxsteps = 500000, useLinCmt = FALSE
) |> as.data.frame()

homeo <- tibble::tibble(
  state = c("imm_bone", "t1_bone", "t1_blood", "t1_spleen",
            "naive_blood", "naive_spleen", "naive_ln"),
  baseline = c(1.97, 0.76, 0.004, 4.40, 4.64, 24.30, 1.90)
) |>
  dplyr::rowwise() |>
  dplyr::mutate(max_rel_drift = max(abs(sim[[state]] / baseline - 1))) |>
  dplyr::ungroup()

homeo |>
  dplyr::rename(
    "State" = state,
    "Baseline (10^6 cell)" = baseline,
    "Max relative drift over 350 d" = max_rel_drift
  ) |>
  knitr::kable(digits = 5, caption = "Homeostasis hold.")
```

The bone-marrow and T1 chain (`imm_bone`, `t1_bone`, `t1_blood`, `t1_spleen`)
holds to solver precision: those four states are balanced exactly by Supp
Eq 1-4. The naive triad drifts by up to 2.3%, and that drift is not a
transcription error -- it is a direct and reproducible consequence of Supp
Eq 5, which is analysed in the Errata section.

```{r steady-state-assert}
# Deterministic model, no random effects: these are numerical-identity checks,
# so they are asserted tightly.
stopifnot(
  # Exactly balanced states: machine precision.
  max(homeo$max_rel_drift[1:4]) < 1e-6,
  # Naive triad: bounded by the Supp Eq 5 imbalance quantified in Errata.
  max(homeo$max_rel_drift[5:7]) < 0.05
)
```

## Mass-balance / flux check

At the reported steady state each homeostasis ODE should evaluate to zero.
Computing the fluxes symbolically from the file's own derived parameters:

```{r mass-balance}
p <- sim[1, ]
bl <- c(imm_bone = 1.97, t1_bone = 0.76, t1_blood = 0.004, t1_spleen = 4.40,
        naive_blood = 4.64, naive_spleen = 24.30, naive_ln = 1.90)

flux <- tibble::tibble(
  Equation = c("Supp Eq 8 (imm_bone)", "Supp Eq 9 (t1_bone)",
               "Supp Eq 10 (t1_blood)", "Supp Eq 11 (t1_spleen)",
               "Supp Eq 12 (naive_blood)", "Supp Eq 13 (naive_spleen)",
               "Supp Eq 14 (naive_ln)"),
  `d/dt at baseline` = c(
    20 - p$kmat_1 * bl[["imm_bone"]],
    p$kmat_1 * bl[["imm_bone"]] - p$k1 * bl[["t1_bone"]],
    p$k1 * bl[["t1_bone"]] + 1.01 * bl[["t1_spleen"]] - p$k3 * bl[["t1_blood"]],
    p$k3 * bl[["t1_blood"]] - 1.01 * bl[["t1_spleen"]] - p$kmat_2 * bl[["t1_spleen"]],
    10.08 * bl[["naive_spleen"]] - p$k5 * bl[["naive_blood"]] +
      p$k7 * bl[["naive_ln"]] - 4.9 * bl[["naive_blood"]],
    p$kmat_2 * bl[["t1_spleen"]] - p$kdeg_spl * bl[["naive_spleen"]] -
      10.08 * bl[["naive_spleen"]] + p$k5 * bl[["naive_blood"]],
    4.9 * bl[["naive_blood"]] - p$k7 * bl[["naive_ln"]] - 0.01 * bl[["naive_ln"]]
  )
)
flux |> knitr::kable(digits = 8, caption = "Net flux at the reported steady state (10^6 cell/day).")
```

Six of the seven balance to zero. Supp Eq 14 leaves a residual of about
+0.22 x 10^6 cell/day, i.e. the lymph node influx exceeds the modelled efflux
by roughly 1% -- again the Supp Eq 5 substitution, quantified in Errata. The
residual is exactly `kdeg_ln * (NaiveSpleen_ss - NaiveLN_ss)`:

```{r mass-balance-assert}
stopifnot(
  max(abs(flux$`d/dt at baseline`[1:6])) < 1e-8,
  abs(flux$`d/dt at baseline`[7] - 0.01 * (24.30 - 1.90)) < 1e-8
)
```

## Perturbation recovery

Displacing the naive spleen pool to half and to twice its baseline should
return it to baseline, confirming a single stable attractor rather than a
coincidental initial condition.

```{r perturbation}
perturb <- function(mult) {
  rxode2::rxSolve(
    ui, rxode2::et(seq(0, 20, by = 0.05)),
    inits = c(naive_spleen = 24.30 * mult),
    atol = 1e-10, rtol = 1e-8, maxsteps = 500000, useLinCmt = FALSE
  ) |>
    as.data.frame() |>
    dplyr::mutate(start = paste0(mult, "x baseline"))
}
pert <- dplyr::bind_rows(perturb(0.5), perturb(1), perturb(2))

ggplot(pert, aes(time, naive_spleen, colour = start)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 24.30, linetype = "dashed") +
  labs(x = "Time (day)", y = "Naive B cells in spleen (10^6 cell)",
       colour = "Initial condition",
       title = "Perturbation recovery of the naive spleen pool",
       caption = "Dashed line is the reported steady state of 24.30 x 10^6 cells.")
```

```{r perturbation-assert}
final <- pert |>
  dplyr::group_by(start) |>
  dplyr::slice_tail(n = 1) |>
  dplyr::ungroup()
# All three trajectories converge to the same attractor. Deterministic, so the
# tolerance only needs to admit the ~1% Supp Eq 5 offset from 24.30 exactly.
stopifnot(diff(range(final$naive_spleen)) < 1e-3,
          all(abs(final$naive_spleen / 24.30 - 1) < 0.05))
```

# Replicate published figures

## Figure 2 -- ASC dynamics across tissues

Ugolkov 2026 Figure 2 shows model-predicted ASC counts against the pooled
experimental data in lymph nodes (A), spleen (B), bone marrow (C) and blood
(D). The Results section reports the predicted Cmax and Tmax for each tissue,
which is what is checked here.

```{r figure-2, fig.height = 5.5}
asc_long <- sim |>
  dplyr::select(time, asc_ln, asc_spleen, asc_bone, asc_blood) |>
  tidyr::pivot_longer(-time, names_to = "state", values_to = "cells") |>
  dplyr::mutate(tissue = dplyr::recode(
    state,
    asc_ln = "A. Lymph nodes", asc_spleen = "B. Spleen",
    asc_bone = "C. Bone marrow", asc_blood = "D. Blood"))

ggplot(asc_long, aes(time, cells)) +
  geom_line(colour = "firebrick", linewidth = 0.7) +
  facet_wrap(~tissue, scales = "free_y") +
  labs(x = "Time (day)", y = "ASC (cells)",
       title = "Figure 2 -- ASC dynamics following antigen exposure",
       caption = "Replicates Figure 2 of Ugolkov 2026 (model prediction only; the published panels also overlay pooled experimental data).")
```

```{r figure-2-table}
published <- tibble::tribble(
  ~Quantity,                          ~Published,
  "ASC lymph nodes Cmax (cell)",      2.6e5,
  "ASC lymph nodes Tmax (day)",       24,
  "ASC spleen Cmax (cell)",           2.7e4,
  "ASC spleen Tmax (day)",            21,
  "ASC bone marrow at day 200 (cell)", 2.3e4,
  "ASC blood Cmax (cell)",            4.6e3,
  "ASC blood Tmax (day)",             24,
  "IgG Tmax (day)",                   30
)
simulated <- c(
  max(sim$asc_ln), sim$time[which.max(sim$asc_ln)],
  max(sim$asc_spleen), sim$time[which.max(sim$asc_spleen)],
  sim$asc_bone[which.min(abs(sim$time - 200))],
  max(sim$asc_blood), sim$time[which.max(sim$asc_blood)],
  sim$time[which.max(sim$igg_blood)]
)
cmp <- published |>
  dplyr::mutate(Simulated = signif(simulated, 4),
                `Difference (%)` = round(100 * (simulated - Published) / Published, 1))
cmp |> knitr::kable(caption = "Simulated vs. values reported in the Results section of Ugolkov 2026.")
```

```{r figure-2-assert}
# Deterministic solve against the paper's own reported numbers. Realised
# differences are 0.8-8.2%; 12% leaves headroom for solver settings while
# still catching a mis-transcribed rate constant or steady-state anchor,
# which move these quantities by tens of percent (see Errata, where the
# Table 1 k7 variant misses the ASC Cmax values by ~90%).
stopifnot(max(abs(cmp$`Difference (%)`)) < 12)
```

The paper's own Figure 2 discussion notes that the observed Tmax values carry
wide inter-study spread (spleen Tmax 17 days with SE 9.45; blood Tmax 14 days
against a predicted 24), so the model-vs-data Tmax offsets in the publication
are larger than the model-vs-reported-prediction offsets checked here.

## Figure 3 -- external validation on IgG

IgG was not used in calibration. Figure 3 shows normalised IgG against five
held-out studies; the paper reports a predicted peak at day 30.

```{r figure-3}
sim |>
  dplyr::mutate(igg_norm = igg_blood / max(igg_blood)) |>
  ggplot(aes(time, igg_norm)) +
  geom_line(colour = "steelblue", linewidth = 0.7) +
  labs(x = "Time (day)", y = "Normalised antigen-specific IgG",
       title = "Figure 3 -- external validation against plasma IgG",
       caption = "Replicates the model curve of Figure 3A of Ugolkov 2026, normalised to its own maximum as the paper does.")
```

## Figure 4 -- local sensitivity to kmat_3 and Vmax

The paper's central mechanistic claim is a tissue-specific partition: ASC in
secondary lymphoid organs are governed by the ASC maturation rate `kmat_3`,
while ASC in the bone marrow are governed by the survival-niche capacity
`Vmax`. Figure 4 varies each by +/-20%.

```{r figure-4, fig.height = 5.5}
solve_with <- function(par, value, label) {
  pars <- c(lkmat_3 = log(294.19), lvmax = log(490.60))
  pars[[par]] <- log(value)
  rxode2::rxSolve(
    ui, rxode2::et(seq(0, 350, by = 1)), params = pars,
    atol = 1e-10, rtol = 1e-8, maxsteps = 500000, useLinCmt = FALSE
  ) |>
    as.data.frame() |>
    dplyr::mutate(varied = label, level = paste0(round(100 * value /
      ifelse(par == "lkmat_3", 294.19, 490.60)), "%"))
}

sens <- dplyr::bind_rows(
  solve_with("lkmat_3", 294.19 * 0.8, "kmat_3 +/- 20%"),
  solve_with("lkmat_3", 294.19,       "kmat_3 +/- 20%"),
  solve_with("lkmat_3", 294.19 * 1.2, "kmat_3 +/- 20%"),
  solve_with("lvmax",   490.60 * 0.8, "Vmax +/- 20%"),
  solve_with("lvmax",   490.60,       "Vmax +/- 20%"),
  solve_with("lvmax",   490.60 * 1.2, "Vmax +/- 20%")
) |>
  dplyr::select(time, varied, level, asc_spleen, asc_ln, asc_bone) |>
  tidyr::pivot_longer(c(asc_spleen, asc_ln, asc_bone),
                      names_to = "state", values_to = "cells") |>
  dplyr::mutate(tissue = dplyr::recode(
    state, asc_spleen = "Spleen", asc_ln = "Lymph nodes", asc_bone = "Bone marrow"))

ggplot(sens, aes(time, cells, colour = level)) +
  geom_line(linewidth = 0.6) +
  facet_grid(tissue ~ varied, scales = "free_y") +
  labs(x = "Time (day)", y = "ASC (cells)", colour = "Parameter level",
       title = "Figure 4 -- local sensitivity",
       caption = "Replicates the family-of-curves panels (A-F) of Figure 4 of Ugolkov 2026.")
```

The tornado panels (Figure 4G-I) extend this to all five estimated parameters,
plotting the percentage change in the day-200 ASC level from a +/-20% change in
each. Reproducing that analysis:

```{r figure-4-tornado}
base_par <- c(lk9 = log(8.53), lk12 = log(0.02), lk13 = log(16.26),
              lkmat_3 = log(294.19), lvmax = log(490.60))
at_day200 <- function(pars) {
  rxode2::rxSolve(ui, rxode2::et(c(0, 200)), params = pars,
                  atol = 1e-10, rtol = 1e-8, maxsteps = 500000,
                  useLinCmt = FALSE) |>
    as.data.frame() |>
    dplyr::filter(time == 200)
}
ref200 <- at_day200(base_par)

tornado <- do.call(rbind, lapply(names(base_par), function(p) {
  do.call(rbind, lapply(c(0.8, 1.2), function(mult) {
    pp <- base_par
    pp[[p]] <- log(exp(base_par[[p]]) * mult)
    r <- at_day200(pp)
    data.frame(
      Parameter = sub("^l", "", p),
      Change = ifelse(mult < 1, "-20%", "+20%"),
      Spleen = 100 * (r$asc_spleen / ref200$asc_spleen - 1),
      `Lymph nodes` = 100 * (r$asc_ln / ref200$asc_ln - 1),
      `Bone marrow` = 100 * (r$asc_bone / ref200$asc_bone - 1),
      check.names = FALSE
    )
  }))
}))
knitr::kable(tornado, digits = 1,
             caption = "Change in day-200 ASC level from a +/-20% parameter change (%). Replicates the tornado panels G-I of Figure 4 of Ugolkov 2026.")
```

```{r figure-4-assert}
cell <- function(param, change, tissue) {
  v <- tornado[[tissue]][tornado$Parameter == param & tornado$Change == change]
  if (length(v) != 1L) stop("no unique tornado row for ", param, " ", change)
  v
}
# Deterministic solves: the structure below is exact and reproducible. These
# assertions encode the paper's central mechanistic claim -- a tissue-specific
# partition of control -- as sign and zero-effect structure rather than as the
# tornado's magnitudes, which are NOT reproduced (see Assumptions below).
stopifnot(
  # kmat_3 drives the secondary lymphoid organs, in the reported direction ...
  cell("kmat_3", "+20%", "Spleen") > 15,
  cell("kmat_3", "+20%", "Lymph nodes") > 15,
  cell("kmat_3", "-20%", "Spleen") < -15,
  # ... and has NO effect on bone marrow (Figure 4I shows no kmat_3 bar).
  abs(cell("kmat_3", "+20%", "Bone marrow")) < 1,
  # Vmax is the mirror image: it drives bone marrow ...
  cell("vmax", "+20%", "Bone marrow") > 15,
  # ... and is negligible in spleen and lymph nodes.
  abs(cell("vmax", "+20%", "Spleen")) < 2,
  abs(cell("vmax", "+20%", "Lymph nodes")) < 2,
  # k12 acts on bone marrow with the INVERTED sign the paper reports.
  cell("k12", "+20%", "Bone marrow") < -10,
  # k9 is blood-to-spleen, so it moves the spleen and not the lymph nodes
  # (Figure 4H shows no k9 bar).
  cell("k9", "+20%", "Spleen") > 15,
  abs(cell("k9", "+20%", "Lymph nodes")) < 1,
  # k13 diverts ASC to peripheral tissue, depleting both lymphoid organs.
  cell("k13", "+20%", "Spleen") < -10,
  cell("k13", "+20%", "Lymph nodes") < -10
)
```

The **direction** of every effect, and the set of parameters with **no** effect,
match Figure 4G-I exactly: `kmat_3` and `Vmax` partition control between the
secondary lymphoid organs and the bone marrow; `k12` acts on the bone marrow
with the inverted sign the paper highlights; `k9` moves the spleen but not the
lymph nodes; `k13` depletes both. The **magnitudes** do not match, and that is
recorded as a deviation below.

## Figure 5 -- inter-study heterogeneity through Vmax

Refitting each of the six calibration datasets while estimating only `Vmax`
gave a range of 26.66 to 836 cells/day around the pooled value. Figure 5 shows
that varying this one parameter spans the observed heterogeneity in bone
marrow ASC.

```{r figure-5}
vmax_grid <- c(26.66, 100, 300, 490.60, 836)
het <- dplyr::bind_rows(lapply(vmax_grid, function(v) {
  rxode2::rxSolve(
    ui, rxode2::et(seq(0, 350, by = 1)), params = c(lvmax = log(v)),
    atol = 1e-10, rtol = 1e-8, maxsteps = 500000, useLinCmt = FALSE
  ) |>
    as.data.frame() |>
    dplyr::mutate(Vmax = factor(v, levels = vmax_grid))
}))

ggplot(het, aes(time, asc_bone, colour = Vmax)) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time (day)", y = "ASC in bone marrow (cells)",
       colour = "Vmax (cell/day)",
       title = "Figure 5 -- inter-study heterogeneity captured by Vmax alone",
       caption = "Replicates Figure 5 of Ugolkov 2026. Range 26.66-836 cell/day is the span of the six individually refitted datasets.")
```

```{r figure-5-assert}
span <- het |>
  dplyr::filter(time == 200) |>
  dplyr::summarise(lo = min(asc_bone), hi = max(asc_bone))
# The paper attributes an order-of-magnitude spread in bone marrow ASC across
# studies (Slifka 1998 vs Schrock 2019) to this parameter alone.
stopifnot(span$hi / span$lo > 10)
```

The day-200 bone marrow ASC count spans a factor of
`r round(span$hi / span$lo)` across the refitted `Vmax` range, consistent with
the paper's observation that the maximal bone marrow ASC count reported by
Slifka 1998 was "nearly two orders of magnitude higher" than that of
Schrock 2019.

# Assumptions and deviations

## Errata -- two Table 1 values are not reproducible

**This is the one substantive deviation in this extraction.** Table 1 prints
`kdeg_ln = 0.05` and `k7 = 62.62`, but the deposited Supplementary RxODE model
code sets `kdeg_ln = 0.01` and derives `k7` from Supp Eq 5. The model file uses
the deposited code, because only the code reproduces the paper's own published
outputs. The evidence:

```{r errata}
# A compact standalone re-implementation lets k5, k7 and kdeg_spl be supplied
# directly, which the packaged model does not allow (it derives them inside
# model()). Structure is identical to the packaged model.
errata_mod <- rxode2::rxode({
  ktr <- 7 / 18
  d/dt(imm_bone)     <- 20 - 10.152284 * imm_bone
  d/dt(t1_bone)      <- 10.152284 * imm_bone - 26.315789 * t1_bone
  d/dt(t1_blood)     <- 26.315789 * t1_bone + 1.01 * t1_spleen - 6111 * t1_blood
  d/dt(t1_spleen)    <- 6111 * t1_blood - 1.01 * t1_spleen - 4.5454545 * t1_spleen
  d/dt(naive_blood)  <- 10.08 * naive_spleen - k5 * naive_blood +
    k7 * naive_ln - 4.9 * naive_blood
  d/dt(naive_spleen) <- 4.5454545 * t1_spleen - kdeg_spl * naive_spleen -
    10.08 * naive_spleen + k5 * naive_blood
  d/dt(naive_ln)     <- 4.9 * naive_blood - k7 * naive_ln - kdeg_ln * naive_ln
  tclamp <- max(t, 1e-3)
  d/dt(antigen0) <- (393.12 / tclamp^2 - 66.98 / tclamp) *
    exp(189.1 - 393.12 / tclamp - 66.98 * log(tclamp))
  d/dt(antigen1) <- ktr * antigen0 - ktr * antigen1
  d/dt(antigen2) <- ktr * antigen1 - ktr * antigen2
  d/dt(antigen3) <- ktr * antigen2 - ktr * antigen3
  d/dt(antigen4) <- ktr * antigen3 - ktr * antigen4
  d/dt(antigen5) <- ktr * antigen4 - ktr * antigen5
  d/dt(antigen)  <- ktr * antigen5 - ktr * antigen
  d/dt(asc_spleen) <- 294.19 * antigen * naive_spleen - 2.52 * asc_spleen + 8.53 * asc_blood
  d/dt(asc_ln)     <- 294.19 * antigen * (naive_ln * 22) - 2.72 * asc_ln + 143.6 * asc_blood
  d/dt(asc_bone)   <- (490.6 * asc_blood) / (0.1 + asc_blood) - 0.02 * asc_bone
  d/dt(asc_blood)  <- 2.72 * asc_ln + 2.52 * asc_spleen + 0.02 * asc_bone +
    0.02 * asc_peripheral -
    (8.53 + 143.6 + 16.26 + 490.6 / (0.1 + asc_blood)) * asc_blood
  d/dt(asc_peripheral) <- 16.26 * asc_blood - 0.02 * asc_peripheral
  d/dt(igg_blood) <- 22 * (asc_spleen + asc_bone + asc_blood + asc_ln) - 0.12 * igg_blood
  imm_bone(0) <- 1.97;    t1_bone(0) <- 0.76;   t1_blood(0) <- 0.004
  t1_spleen(0) <- 4.40;   naive_blood(0) <- 4.64
  naive_spleen(0) <- 24.30; naive_ln(0) <- 1.90
})

run_variant <- function(kdeg_ln, k7, k5, kdeg_spl, label) {
  rxode2::rxSolve(
    errata_mod, rxode2::et(seq(0, 350, by = 0.5)),
    params = c(kdeg_ln = kdeg_ln, k7 = k7, k5 = k5, kdeg_spl = kdeg_spl),
    atol = 1e-10, rtol = 1e-8, maxsteps = 500000
  ) |>
    as.data.frame() |>
    dplyr::summarise(
      Variant = label,
      `ASC LN Cmax` = signif(max(asc_ln), 3),
      `ASC spleen Cmax` = signif(max(asc_spleen), 3),
      `ASC blood Cmax` = signif(max(asc_blood), 3),
      `IgG Tmax (day)` = time[which.max(igg_blood)],
      `naive_ln at 350 d (x baseline)` = signif(naive_ln[dplyr::n()] / 1.90, 3)
    )
}

variants <- dplyr::bind_rows(
  run_variant(0.01, 11.8384, 52.7373, 0.8130453,
              "Deposited RxODE code (used in the model file)"),
  run_variant(0.05, 62.62, 73.53, 4.78,
              "Table 1 as printed"),
  run_variant(0.05, 11.9163, 52.7692, 0.8191358,
              "Strict steady-state solve of Supp Eq 14")
) |>
  dplyr::bind_rows(tibble::tibble(
    Variant = "Published (Results section)",
    `ASC LN Cmax` = 2.6e5, `ASC spleen Cmax` = 2.7e4,
    `ASC blood Cmax` = 4.6e3, `IgG Tmax (day)` = 30,
    `naive_ln at 350 d (x baseline)` = 1
  ))
variants |> knitr::kable(caption = "Which parameterisation reproduces the paper.")
```

```{r errata-assert}
dep <- variants[variants$Variant == "Deposited RxODE code (used in the model file)", ]
tab <- variants[variants$Variant == "Table 1 as printed", ]
stopifnot(
  # The deposited code lands within 12% of every published quantity ...
  abs(dep$`ASC LN Cmax` / 2.6e5 - 1) < 0.12,
  abs(dep$`ASC spleen Cmax` / 2.7e4 - 1) < 0.12,
  abs(dep$`ASC blood Cmax` / 4.6e3 - 1) < 0.12,
  # ... and Table 1 as printed misses them by an order of magnitude and
  # collapses the naive lymph node pool.
  tab$`ASC LN Cmax` / 2.6e5 < 0.2,
  tab$`naive_ln at 350 d (x baseline)` < 0.1
)
```

Reading the table:

* **Deposited RxODE code** reproduces every published Cmax and the IgG Tmax to
  within 8%, and holds the naive lymph node pool at baseline.
* **Table 1 as printed** misses the ASC Cmax values by roughly 90% and drives
  the naive lymph node pool to about 2% of baseline, because `k7 = 62.62`
  is more than five times the efflux the lymph node balance can sustain.
  `62.62` is recoverable only by substituting the *spleen* anchor into the
  influx term of Supp Eq 5, `(4.9 * 24.30 - 0.05 * 1.90) / 1.90 = 62.62`,
  which no printed equation says.
* **Strict steady-state solve of Supp Eq 14**, `k7 = (k6 * NaiveBlood_ss -
  kdeg_ln * NaiveLN_ss) / NaiveLN_ss = 11.92`, is the biologically correct
  derivation and gives an exactly flat homeostasis layer. It is *not* used,
  because the extraction policy is to reproduce the model the authors built and
  ran; the difference from the deposited code is under 2% on every output.

Note that Table 1's downstream values `k5 = 73.53` and `kdeg_spl = 4.78` are
internally consistent with `k7 = 62.62`: substituting 62.62 into Supp Eq 6 gives
73.53, and that in turn into Supp Eq 7 gives 4.78. The packaged model applies
those same two equations to its own `k7 = 11.84` and therefore derives
`k5 = 52.74` and `kdeg_spl = 0.813` instead. So the whole Table 1 divergence
propagates from just two printed entries -- `k7` and `kdeg_ln` -- while the four
derived values that Table 1 *does* reproduce are unaffected:

```{r derived-params}
derived <- tibble::tibble(
  Parameter = c("kmat_1", "k1", "k3", "kmat_2", "k7", "k5", "kdeg_spl"),
  `In model file` = signif(c(p$kmat_1, p$k1, p$k3, p$kmat_2,
                             p$k7, p$k5, p$kdeg_spl), 6),
  `Table 1` = c(10.15, 26.32, 6111, 4.55, 62.62, 73.53, 4.78)
) |>
  dplyr::mutate(Agrees = ifelse(
    abs(`In model file` / `Table 1` - 1) < 0.01, "yes", "NO -- see above"))
derived |>
  knitr::kable(caption = "The seven parameters derived inside model() from Supp Eq 1-7, against Table 1.")
```

```{r derived-params-assert}
# Deterministic algebra, so these are exact-value checks. They pin both halves:
# the four values that must keep matching Table 1, and the three that must keep
# taking the deposited-code values rather than silently drifting to Table 1's.
stopifnot(
  derived$Agrees == c("yes", "yes", "yes", "yes",
                      "NO -- see above", "NO -- see above", "NO -- see above"),
  abs(p$k7 - 11.8384) < 1e-3,
  abs(p$k5 - 52.7373) < 1e-3,
  abs(p$kdeg_spl - 0.8130453) < 1e-6
)
```

## Errata -- the Figure 4 tornado magnitudes are not reproducible

The direction and the zero-effect structure of Figure 4G-I are reproduced
exactly (see the assertions above), but the magnitudes are not. The paper's
tornado panels show a +20% change in `kmat_3` raising day-200 spleen and lymph
node ASC by roughly +215%, and a +20% change in `Vmax` raising bone marrow ASC
by roughly +245%; the Results text describes the `kmat_3` effect as having
"nearly doubled ASC levels in both the spleen and lymph nodes, whereas a 20%
decrease in this parameter led to a ~50% reduction", and attributes such
"ultrasensitive responses" to "the nonlinear structure of the model equations,
including the saturable transport kinetics".

Solving the published equations gives a strictly **linear** +/-20% response in
every case. This is structurally inevitable rather than a transcription
artefact:

* ASC generation in Supp Eq 22-23 is exactly proportional to `kmat_3`, and the
  spleen and lymph node pools exchange with blood through linear rate
  constants, so those pools are linear in `kmat_3`.
* The only nonlinearity is the saturable bone marrow influx of Supp Eq 24-25.
  With `khalf` fixed at 0.1 cell against circulating ASC counts in the
  thousands, `Vmax * asc_blood / (khalf + asc_blood)` is saturated to within
  0.01% of `Vmax` throughout. The paper says as much -- "this small value
  mathematically implies near-maximal influx even at low circulating ASC
  numbers" -- but the consequence is that the term behaves as a **constant**
  influx, so the bone marrow steady state is simply `Vmax / k12`. That makes
  bone marrow ASC exactly linear in `Vmax` (+20% -> +20%) and exactly
  proportional to `1 / k12` (+20% -> -16.7%), and makes it entirely
  independent of `kmat_3`, `k9` and `k13` -- which is precisely the pattern of
  empty bars in Figure 4I.

So the saturable term cited as the source of ultrasensitivity is, at the
published `khalf`, the one term that cannot produce it. The model file
reproduces the published equations and parameters; the tornado magnitudes are
recorded here as a known, unreproduced discrepancy rather than accommodated by
changing any value.

## Other assumptions and deviations

* **Supp Eq 5 is reproduced, not corrected.** Its degradation term uses the
  spleen steady-state anchor where a strict solve of Supp Eq 14 would use the
  lymph node anchor. Both the printed equation and the deposited code write it
  that way, so the model file writes it that way, with the consequence
  quantified above (a residual `kdeg_ln * (NaiveSpleen_ss - NaiveLN_ss)` in the
  lymph node balance, and up to 2.3% drift in the naive states over 350 days).
* **Singularity guard on the antigen forcing function.** Supp Eq 15 contains
  `1/t`, `1/t^2` and `log(t)` and is undefined at `t = 0`. The model file
  clamps `t` to a floor of `1e-3` day. The guard is numerically inert:
  `exp(189.1 - 393.12/t - 66.98*log(t))` underflows to exactly zero for
  `t <= 1e-3`, which is also the true limit of Supp Eq 15 as `t -> 0+`.
* **Antigen function constants are literals.** The values `393.12`, `66.98` and
  `189.1` are printed as bare numbers in Supp Eq 15 and in the deposited code;
  they are unnamed shape constants of an empirical curve fitted to the
  viral-load data of Castillo-Mendez 2007, so they are reproduced as literals
  rather than promoted to named parameters.
* **`mtt = 18` days and the initial conditions are not in Table 1.** They come
  from the Results narrative ("A mean transit time of 18 days was selected to
  match the observed average time to reach the maximum ASC counts") and from
  the deposited RxODE model code respectively.
* **`Vmax` is 490.60**, the Table 1 and deposited-code value. The inter-study
  heterogeneity section quotes 492.8 cells/day as "the typical calibrated
  value"; the 0.4% difference is not resolved by any on-disk source and does
  not affect any conclusion.
* **No IIV and no residual error.** Ugolkov 2026 minimised a -2 log-likelihood
  whose per-variable sigma was computed from the residuals themselves and
  tabulates no residual-error magnitude for any output. The published 95%
  intervals are parameter-uncertainty bands from the inverse Hessian, and the
  paper states explicitly that they "should not be interpreted as a prediction
  interval for individual experimental observations". No variance components
  were invented.
* **ASC elimination is absent by design**, not by omission: the paper reports
  ASC levels stable to 300 days post-immunization.
* **Naive B cells are not depleted by activation**, again by design, because
  the antigen-specific precursor frequency is 1 in 10^6.
* **The published figures overlay pooled experimental data** that are not
  redistributable here; the replications above show the model curves only, and
  are checked numerically against the Cmax / Tmax values the paper reports in
  its Results text.
