---
title: "Glycochenodeoxycholic acid 3-O-sulfate, an OATP1B3 / OAT3 biomarker (Ujihira 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Glycochenodeoxycholic acid 3-O-sulfate, an OATP1B3 / OAT3 biomarker (Ujihira 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ujihira_2025_glycochenodeoxycholicAcidSulfate"))$reference`
* Article: <https://doi.org/10.1002/cpt.70023>

Glycochenodeoxycholic acid 3-O-sulfate (GCDCA-S) is a sulfated bile-acid
conjugate proposed by the International Transporter Consortium as a Tier 2
endogenous biomarker of hepatic OATP1B3 and renal OAT3. Ujihira 2025 is the
first mechanistic population-PK model for it, and it quantifies two things that
were previously unknown: the biomarker's zero-order synthesis rate, and the
split between its hepatobiliary and renal elimination routes.

The distinctive feature of this extraction is that the model is a **single
coupled three-drug system**. The paper fit the GCDCA-S turnover model
simultaneously with the perpetrator PK of rifampicin and probenecid (Figure 2),
and the two inhibition constants plus the probenecid fold-change `X` are joint
estimates across that system. The model file therefore carries all five ODE
states in one place, so a user can dose a perpetrator and read the biomarker
response directly, without a two-step simulate-then-feed workflow.

```{r model}
mod <- rxode2::rxode(readModelDb("Ujihira_2025_glycochenodeoxycholicAcidSulfate"))
mod
```

The model declares four endpoints, so rxode2 places their slots *after* the five
ODE states. Observation records in every event table below therefore use
`cmt = "Cc"` and not `cmt = "central"`: with more than one declared endpoint,
`cmt = "central"` is rejected with `'dvid'->'cmt' or 'cmt' on observation
record`. Because `Cc` already has an endpoint slot from the model definition,
naming it on an observation row injects nothing and renumbers nothing, and
`rxSolve()` returns every model variable as an output column regardless of which
endpoint is named.

```{r endpoint-map}
mod$predDf[, c("var", "cond", "cmt", "dvid")]
mod$state
```

# Population and biological context

```{r population}
pop <- readModelDb("Ujihira_2025_glycochenodeoxycholicAcidSulfate")()$population
str(pop, max.level = 1)
```

The model was developed on 24 healthy White volunteers pooled from three
crossover studies, each with a control occasion and an inhibitor occasion
(Ujihira 2025 Table 1): Tatosian et al. 2021 (n = 6, 3 male / 3 female, single
600 mg oral rifampicin), Willemin et al. 2021 (n = 6, all female, probenecid
500 mg four times daily on days 1-7 after two day-0 loading doses), and an
unpublished study of the same design as Willemin (n = 12, all male). 430
GCDCA-S plasma samples and 175 GCDCA-S
urine samples were fit simultaneously with 153 perpetrator plasma samples in
Monolix 2024R2.

Observed GCDCA-S plasma baseline was approximately 80 nmol/L, with high
between-subject variability (CV 34-69% across the three studies) and
comparable within-subject diurnal variability (CV 34-43%) -- both larger than
for the Tier 1 biomarker coproporphyrin I. Six diurnal-fluctuation functions
(Table S2) and a sex effect on the synthesis rate (Table S3) were each tested
and each failed to improve the fit, so neither appears in the final model.

# Source trace

Every structural equation and every `ini()` value, with its location in the
source.

| Model element | Value | Source location |
| --- | --- | --- |
| Plasma turnover ODE, no inhibitor | `dC/dt = [ksyn - (C*CLh + C*CLR)] / Vc` | Eq. 1, main text p. 1535 |
| Plasma turnover ODE, with inhibitors | `dC/dt = [ksyn - (C*CLh/(1+Cu_RIF/Ki_u_OATP1B3)*(1/X) + C*CLR/(1+Cu_PROB/Ki_u_OAT3))] / Vc` | Eq. 2, main text p. 1535 |
| Urine ODE, no inhibitor | `dA/dt = C*CLR` | Eq. 3, main text p. 1535 |
| Urine ODE, with probenecid | `dA/dt = C*CLR/(1+Cu_PROB/Ki_u_OAT3)` | Eq. 4, main text p. 1535 |
| Rifampicin structure | 1-compartment, zero-order absorption after a lag, linear elimination | Methods "Structural models"; Figure 2 left panel |
| Probenecid structure | 1-compartment, first-order absorption, linear elimination | Methods "Structural models"; Figure 2 right panel |
| `lksyn` = log(1.0) umol/h | 1.0 (RSE 18%) | Table 2, GCDCA-S `k syn` |
| `lvc` = fixed(log(4.8)) L | 4.8 (FIXED) | Table 2, GCDCA-S `V c,GCDCA-S` |
| `lcl_renal` = log(0.31) L/h | 0.31 (RSE 8%) | Table 2, GCDCA-S `CL R,GCDCA-S` |
| `lcl_nonren` = log(15) L/h | 15 (RSE 13%) | Table 2, GCDCA-S `CL h,GCDCA-S` |
| `ltlag_rif` = log(0.56) h | 0.56 (RSE 20%) | Table 2, RIF `Tlag RIF` |
| `ld1_rif` = log(0.84) h | 0.84 (RSE 41%) | Table 2, RIF `Tk0 RIF` |
| `lvc_rif` = log(35) L | 35 (RSE 5%) | Table 2, RIF `V RIF` |
| `lcl_rif` = log(6.7) L/h | 6.7 (RSE 8%) | Table 2, RIF `CL RIF` |
| `lkiu_oatp1b3` = log(0.009) umol/L | 0.009 (RSE 29%) | Table 2, RIF `K i,unbound,OATP1B3` |
| `lka_prob` = log(1.1) 1/h | 1.1 (RSE 37%) | Table 2, PROB `ka PROB` |
| `lvc_prob` = log(17) L | 17 (RSE 10%) | Table 2, PROB `V PROB` |
| `lcl_prob` = log(0.95) L/h | 0.95 (RSE 7%) | Table 2, PROB `CL PROB` |
| `lkiu_oat3` = log(2.7) umol/L | 2.7 (RSE 12%) | Table 2, PROB `K i,unbound,OAT3` |
| `e_prob_clh` = -0.4118 | `X` = 1.7 (RSE 5%), encoded as 1/X - 1 | Table 2, PROB `X`; Eq. 2 |
| `etalksyn` = 0.3164 | IIV 61% (RSE 15%), log(1 + 0.61^2) | Table 2, GCDCA-S IIV |
| `etalcl_renal` = 0.0560 | IIV 24% (RSE 25%) | Table 2, GCDCA-S IIV |
| `etaltlag_rif` = 0.1920 | IIV 46% (RSE 32%) | Table 2, RIF IIV |
| `etald1_rif` = 0.4657 | IIV 77% (RSE 37%) | Table 2, RIF IIV |
| `etalcl_rif` = 0.0319 | IIV 18% (RSE 35%) | Table 2, RIF IIV |
| `etalvc_prob` = 0.0917 | IIV 31% (RSE 28%) | Table 2, PROB IIV |
| `etalcl_prob` = 0.0704 | IIV 27% (RSE 20%) | Table 2, PROB IIV |
| `etalkiu_oat3` = 0.0975 | IIV 32% (RSE 27%) | Table 2, PROB IIV |
| `propSd` = 0.45 | 45% (RSE 3%) | Table 2, `sigma prop - GCDCA-S plasma` |
| `propSd_Ugcdcas` = 0.29 | 29% (RSE 1%) | Table 2, `sigma prop - GCDCA-S urine` |
| `propSd_rif` = 0.20 | 20% (RSE 14%) | Table 2, `sigma prop - RIF` |
| `addSd_rif` = fixed(0.2) umol/L | 0.2 (FIXED) | Table 2, `sigma add - RIF` |
| `propSd_prob` = 0.23 | 23% (RSE 8%) | Table 2, `sigma prop - PROB` |
| `addSd_prob` = fixed(0.001) umol/L | 0.001 (FIXED) | Table 2, `sigma add - PROB` |
| `fu_rif` = fixed(0.11) | **not in Ujihira 2025** -- see Errata | Barnett 2018 Table 1 footnote c |
| `fu_prob` = fixed(0.062) | **not in Ujihira 2025** -- see Errata | Ahmad 2021 Table 1 footnote |

## Units of every ODE term (dimensional analysis)

The whole system runs in `umol` and `umol/L` (= `uM`), which is why perpetrator
doses must be supplied in `umol` rather than `mg`.

| Term | Units | Check |
| --- | --- | --- |
| `central`, `urine` | umol | state = amount |
| `Cc = central / vc` | umol / L = uM | matches the paper's `C_GCDCA-S` (uM) |
| `ksyn` | umol/h | matches the paper's `k syn` (umol/h) |
| `cl_nonren`, `cl_renal` | L/h | matches `CLh`, `CLR` (L/h) |
| `(cl_nonren_eff + cl_renal_eff) * Cc` | L/h * umol/L = umol/h | balances `d/dt(central)` |
| `d/dt(urine) = cl_renal_eff * Cc` | umol/h | matches the paper's `A_GCDCA-S` (umol) |
| `cu_rif / kiu_oatp1b3` | uM / uM = unitless | inhibition term is dimensionless |
| `e_prob_clh` | unitless | multiplicative factor on `cl_nonren` |
| `central_rif / vc_rif` | umol / L = uM | matches `Cu,RIF` scale after `fu_rif` |

The paper writes Eq. 1 in terms of concentration with the whole bracket divided
by `Vc`; the file uses the algebraically identical amount form
(`d/dt(central) = ksyn - CL*Cc` with `Cc = central/vc`), which is the standard
rxode2 idiom and makes the urine state directly comparable to the paper's
`A_GCDCA-S` in umol.

```{r constants}
# Molecular weights used only to convert the published mg doses into the umol
# the model consumes. Rifampicin 822.94 g/mol and probenecid 285.34 g/mol are
# the values already carried by the nlmixr2lib covariate / model registers
# (PubChem CID 135398735 and 4911 agree to 4 significant figures).
MW_RIF  <- 822.94
MW_PROB <- 285.34

dose_umol <- function(mg, mw) mg / mw * 1000

# Typical-value (deterministic) model for the structural checks.
mod0 <- rxode2::zeroRe(mod)

# Analytic baseline from Eq. 1 with dC/dt = 0: Css = ksyn / (CLh + CLR).
ksyn <- 1.0; clh <- 15; clr <- 0.31; vc <- 4.8
css_analytic <- ksyn / (clh + clr)
c(css_umol_per_L = css_analytic, css_nmol_per_L = css_analytic * 1000)
```

# Steady-state check

With no perpetrator on board the model must sit at the analytic baseline
`ksyn / (CLh + CLR)` indefinitely.

```{r steady-state}
ev_ss <- rxode2::et(seq(0, 168, by = 0.5), cmt = "Cc") |> as.data.frame()
ev_ss$CONMED_PROBENECID <- 0

ss <- rxode2::rxSolve(mod0, ev_ss, returnType = "data.frame")

max_drift <- max(abs(ss$Cc - css_analytic))
c(analytic = css_analytic, simulated = ss$Cc[1], max_abs_drift = max_drift)

stopifnot(max_drift < 1e-10)
```

The simulated baseline holds at `r signif(css_analytic, 6)` umol/L
(`r signif(css_analytic * 1000, 4)` nmol/L) over 7 days with no drift.

# Perturbation-recovery

The model-body initial condition `central(0) <- ksyn/(CLh + CLR) * vc` overrides
`rxSolve(inits = )`, so the state is displaced with events instead: a bolus dose
that doubles the central amount, and an `evid = 6` multiply event that halves
it. Both must return to the same attractor.

```{r perturbation}
c0 <- css_analytic * vc

perturb <- function(ev, label) {
  ev <- as.data.frame(ev)
  ev$CONMED_PROBENECID <- 0
  out <- rxode2::rxSolve(mod0, ev, returnType = "data.frame")
  out$arm <- label
  out
}

up <- perturb(
  rxode2::et(amt = c0, cmt = "central", time = 0) |>
    rxode2::et(seq(0, 6, by = 0.02), cmt = "Cc"),
  "2x baseline (bolus)"
)
dn <- perturb(
  rxode2::et(amt = 0.5, cmt = "central", time = 0, evid = 6) |>
    rxode2::et(seq(0, 6, by = 0.02), cmt = "Cc"),
  "0.5x baseline (multiply)"
)

recov <- dplyr::bind_rows(up, dn)

c(final_up = tail(up$Cc, 1), final_dn = tail(dn$Cc, 1), target = css_analytic)
stopifnot(abs(tail(up$Cc, 1) - css_analytic) < 1e-9,
          abs(tail(dn$Cc, 1) - css_analytic) < 1e-9)
```

```{r perturbation-plot, fig.cap = "Perturbation-recovery: displacing GCDCA-S to twice and half its baseline returns it to the same steady state."}
ggplot2::ggplot(recov, ggplot2::aes(time, Cc, colour = arm)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = css_analytic, linetype = "dashed") +
  ggplot2::labs(x = "Time (h)", y = "GCDCA-S plasma (umol/L)", colour = NULL) +
  ggplot2::theme_bw()
```

Recovery is fast because the uninhibited system's relaxation half-life is
`ln(2) * Vc / (CLh + CLR)` =
`r signif(log(2) * vc / (clh + clr), 3)` h. This is a load-bearing property of
the model: under strong OATP1B3 inhibition `CLh` collapses, the effective
half-life lengthens by more than an order of magnitude, and the biomarker
therefore accumulates slowly rather than jumping to a new quasi-steady state.

# Mass-balance / flux check

At the baseline, production must exactly balance elimination, and the renal
fraction of the total flux is the fraction excreted in urine, `fe`.

```{r mass-balance}
flux_in    <- ksyn
flux_hep   <- clh * css_analytic
flux_renal <- clr * css_analytic
fe         <- clr / (clh + clr)

c(production = flux_in,
  hepatobiliary = flux_hep,
  renal = flux_renal,
  net = flux_in - flux_hep - flux_renal,
  fe_percent = 100 * fe,
  biliary_percent = 100 * (1 - fe))

stopifnot(abs(flux_in - flux_hep - flux_renal) < 1e-12)
```

`fe` = `r signif(100 * fe, 3)`% reproduces the paper's central mechanistic
finding: biliary excretion is the primary route of elimination for GCDCA-S
(reported as ~95%, with `fe` < 5%; Results and Discussion, and Table S6 which
infers `fe` < 3.1-6.9% from four independent clinical DDI studies).

A second, sharper identity: with no inhibitor present, the operational renal
clearance `Ae(0-24) / AUC(0-24)` must recover the `CLR` parameter exactly.

```{r clr-identity}
ev_24 <- rxode2::et(seq(0, 24, by = 0.02), cmt = "Cc") |> as.data.frame()
ev_24$CONMED_PROBENECID <- 0
s24 <- rxode2::rxSolve(mod0, ev_24, returnType = "data.frame")

trapz <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
clr_operational <- (max(s24$Ugcdcas) - min(s24$Ugcdcas)) / trapz(s24$time, s24$Cc)

c(clr_parameter = clr, clr_operational = clr_operational)
stopifnot(abs(clr_operational - clr) < 1e-6)
```

# Rifampicin OATP1B3 interaction

The paper's headline predictive check is the GCDCA-S plasma AUC ratio
(`AUCR0-24h`) after a single oral rifampicin dose. Reported values: predicted
AUCR of **13** at 600 mg against an observed range of **10-16**, and predicted
**9.3** at 300 mg against an observed **4.3-5.9** (Results, "GCDCA-S model
verification"; Table S4 lists the 600 mg value as 13).

```{r rif-sim}
sim_rif <- function(mg, label) {
  ev <- if (is.na(mg)) {
    rxode2::et(seq(0, 24, by = 0.02), cmt = "Cc")
  } else {
    rxode2::et(amt = dose_umol(mg, MW_RIF), cmt = "central_rif",
               rate = -2, time = 0) |>
      rxode2::et(seq(0, 24, by = 0.02), cmt = "Cc")
  }
  ev <- as.data.frame(ev)
  ev$CONMED_PROBENECID <- 0
  out <- rxode2::rxSolve(mod0, ev, returnType = "data.frame")
  out$treatment <- label
  out$id <- 1L
  out
}

rif <- dplyr::bind_rows(
  sim_rif(NA,  "Control"),
  sim_rif(300, "RIF 300 mg"),
  sim_rif(600, "RIF 600 mg")
)
```

`rate = -2` on the dose record is required so rxode2 uses the modelled
zero-order duration `dur(central_rif) <- d1_rif` rather than treating the dose
as an instantaneous bolus.

```{r rif-nca}
conc_rif <- dplyr::filter(rif, !is.na(Cc))

o_conc <- PKNCA::PKNCAconc(conc_rif, Cc ~ time | id / treatment)
intervals <- data.frame(start = 0, end = 24,
                        auclast = TRUE, cmax = TRUE, tmax = TRUE)
nca_rif <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, intervals = intervals))

nca_tab <- as.data.frame(nca_rif) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

auc_ctl <- nca_tab$auclast[nca_tab$treatment == "Control"]

# Published reference values joined BY KEY, never by row position, so a change
# in PKNCA's output ordering cannot silently transpose them.
published <- data.frame(
  treatment      = c("RIF 300 mg", "RIF 600 mg"),
  published_AUCR = c(9.3, 13),
  observed_AUCR  = c("4.3-5.9", "10-16")
)

aucr <- nca_tab |>
  dplyr::filter(treatment != "Control") |>
  dplyr::mutate(AUCR = auclast / auc_ctl) |>
  dplyr::inner_join(published, by = "treatment") |>
  dplyr::select(
    "Treatment"          = treatment,
    "AUC0-24h (uM*h)"    = auclast,
    "Cmax (uM)"          = cmax,
    "Simulated AUCR"     = AUCR,
    "Published AUCR"     = published_AUCR,
    "Observed AUCR"      = observed_AUCR
  )

knitr::kable(aucr, digits = 2,
             caption = "GCDCA-S plasma AUC ratio after single-dose rifampicin, simulated vs. Ujihira 2025.")
```

`PKNCA` reports "No dose information provided" because GCDCA-S is endogenous and
has no dose record; only the interval-based `auclast` / `cmax` / `tmax` are
requested, none of which need a dose.

```{r rif-assert}
aucr_600 <- aucr[["Simulated AUCR"]][aucr$Treatment == "RIF 600 mg"]
aucr_300 <- aucr[["Simulated AUCR"]][aucr$Treatment == "RIF 300 mg"]

# 600 mg: must land inside the paper's observed range of 10-16 and within 15%
# of the paper's own predicted 13.
stopifnot(aucr_600 > 10, aucr_600 < 16, abs(aucr_600 / 13 - 1) < 0.15)
# 300 mg: must reproduce the paper's PREDICTED 9.3 within 15%. The paper is
# explicit that its own prediction overestimates the observed 4.3-5.9 here, so
# the target is the prediction, not the observation.
stopifnot(abs(aucr_300 / 9.3 - 1) < 0.15)
```

This is the strongest available end-to-end check on the extraction: the AUCR is
a product of the mg-to-umol conversion, the rifampicin PK parameters, the
unbound fraction `fu_rif`, the inhibition constant `Ki,u,OATP1B3`, and the
GCDCA-S clearance split. An error in any one of them would move it.

```{r rif-plot, fig.cap = "GCDCA-S plasma response to single-dose rifampicin (replicates the pattern of Ujihira 2025 Figure 3a and Figure 4b)."}
ggplot2::ggplot(rif, ggplot2::aes(time, Cc, colour = treatment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::labs(x = "Time (h)", y = "GCDCA-S plasma (umol/L)", colour = NULL) +
  ggplot2::theme_bw()
```

```{r rif-pk-plot, fig.cap = "Rifampicin perpetrator PK driving the interaction (replicates Ujihira 2025 Figure 4c)."}
ggplot2::ggplot(dplyr::filter(rif, treatment != "Control"),
                ggplot2::aes(time, Cc_rif, colour = treatment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::labs(x = "Time (h)", y = "Rifampicin plasma (umol/L)", colour = NULL) +
  ggplot2::theme_bw()
```

# Probenecid OAT3 interaction

For OAT3 the paper's metric is renal clearance, not plasma AUC: Table S4 reports
a `CLR` ratio (`CLR,+inhibitor / CLR,control`) of **0.1** at steady state under
the probenecid regimen. The regimen is 500 mg at 6 pm and 11 pm on day 0, then
500 mg at 7 am, 1 pm, 6 pm and 11 pm on days 1-7 (Table 1).

```{r prob-sim}
# t = 0 is the day-0 6 pm dose; the day-1 onward pattern repeats every 24 h.
dose_times <- c(0, 5, as.vector(outer(c(13, 19, 24, 29), 24 * (0:6), "+")))
dose_times <- sort(dose_times[dose_times <= 24 * 7.5])
grid <- seq(0, 24 * 7.5, by = 0.05)

sim_prob <- function(on) {
  ev <- if (on) {
    rxode2::et(amt = dose_umol(500, MW_PROB), cmt = "depot_prob",
               time = dose_times) |>
      rxode2::et(grid, cmt = "Cc")
  } else {
    rxode2::et(grid, cmt = "Cc")
  }
  ev <- as.data.frame(ev)
  ev$CONMED_PROBENECID <- as.integer(on)
  rxode2::rxSolve(mod0, ev, returnType = "data.frame")
}

prob_ctl <- sim_prob(FALSE)
prob_on  <- sim_prob(TRUE)

# Steady-state 24 h window (day 6 -> day 7).
window_metrics <- function(s) {
  k <- s$time >= 144 & s$time <= 168
  auc <- trapz(s$time[k], s$Cc[k])
  ae  <- max(s$Ugcdcas[k]) - min(s$Ugcdcas[k])
  c(AUC = auc, Ae = ae, CLR = ae / auc)
}

w_ctl <- window_metrics(prob_ctl)
w_on  <- window_metrics(prob_on)

prob_tab <- data.frame(
  Metric = c("AUC0-24h (uM*h)", "Ae0-24h (umol)", "CLR (L/h)"),
  Control = as.numeric(w_ctl),
  Probenecid = as.numeric(w_on),
  Ratio = as.numeric(w_on) / as.numeric(w_ctl)
)
knitr::kable(prob_tab, digits = 4,
             caption = "GCDCA-S at steady state under the Willemin probenecid regimen.")
```

```{r prob-assert}
clr_ratio <- unname(w_on["CLR"] / w_ctl["CLR"])
auc_ratio <- unname(w_on["AUC"] / w_ctl["AUC"])

c(clr_ratio = clr_ratio, published_clr_ratio = 0.1,
  auc_ratio = auc_ratio, X = 1.7)

# Table S4 reports the CLR ratio to one significant figure as 0.1; require the
# simulated ratio to be consistent with that rounding (stated as an explicit
# interval rather than round(), which is rounding-mode dependent at 0.125).
stopifnot(clr_ratio > 0.05, clr_ratio < 0.15)
# The plasma AUC ratio must recover X = 1.7 to within 5%: CLh dominates total
# clearance, so reducing it 1.7-fold raises plasma exposure 1.7-fold, while the
# OAT3 inhibition of the small renal arm contributes almost nothing to AUC.
stopifnot(abs(auc_ratio / 1.7 - 1) < 0.05)
```

The two ratios separate the two mechanisms exactly as the paper describes them
(Figure S3): OAT3 inhibition is read out almost entirely in `CLR` (ratio
`r signif(clr_ratio, 3)`), while the plasma AUC moves only by the
concentration-independent `X` factor (`r signif(auc_ratio, 3)` against
`X` = 1.7). This is why the paper recommends `CLR` rather than plasma AUC as
the GCDCA-S metric for OAT3-mediated DDI.

```{r prob-plot, fig.cap = "GCDCA-S plasma and renal clearance under multiple-dose probenecid (replicates the pattern of Ujihira 2025 Figure 3a-b)."}
prob_plot <- dplyr::bind_rows(
  dplyr::mutate(prob_ctl, arm = "Control"),
  dplyr::mutate(prob_on,  arm = "Probenecid 500 mg QID")
)
ggplot2::ggplot(prob_plot, ggplot2::aes(time / 24, Cc, colour = arm)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::labs(x = "Time (days)", y = "GCDCA-S plasma (umol/L)", colour = NULL) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

# Between-subject variability and the observed baseline

The paper reports an observed GCDCA-S plasma baseline of approximately
80 nmol/L (Results) and 85 nmol/L in Table S5. The typical value implied by the
parameters is lower, at `r signif(css_analytic * 1000, 4)` nmol/L, because
`ksyn` carries 61% log-normal IIV and the reported baselines are arithmetic
means, not medians.

```{r cohort}
set.seed(20250101)
ev_cohort <- rxode2::et(seq(0, 24, by = 1), cmt = "Cc") |> as.data.frame()
ev_cohort$CONMED_PROBENECID <- 0

cohort <- rxode2::rxSolve(mod, ev_cohort, nSub = 200, returnType = "data.frame")

baseline_by_subject <- cohort |>
  dplyr::filter(time == 0) |>
  dplyr::summarise(
    median_nM = median(Cc) * 1000,
    mean_nM   = mean(Cc) * 1000,
    cv_pct    = 100 * sd(Cc) / mean(Cc)
  )
baseline_by_subject

# Analytic log-normal mean of the baseline: median * exp(omega^2 / 2).
c(analytic_mean_nM = css_analytic * 1000 * exp(0.3163576 / 2),
  reported_mean_nM = 80)
```

The simulated cohort mean baseline lands close to the reported ~80 nmol/L, and
the simulated baseline CV is consistent with the observed 34-69% range across
the three development studies.

```{r cohort-plot, fig.cap = "Simulated baseline GCDCA-S distribution in 200 virtual subjects, no inhibitor."}
ggplot2::ggplot(dplyr::filter(cohort, time == 0),
                ggplot2::aes(Cc * 1000)) +
  ggplot2::geom_histogram(bins = 30, fill = "grey70", colour = "white") +
  ggplot2::geom_vline(xintercept = 80, linetype = "dashed") +
  ggplot2::labs(x = "Baseline GCDCA-S plasma (nmol/L)", y = "Subjects",
                caption = "Dashed line: reported observed baseline of ~80 nmol/L.") +
  ggplot2::theme_bw()
```

# Assumptions and deviations

**Unbound fractions are not reported by Ujihira 2025.** The model's inhibition
terms are driven by *unbound* perpetrator concentrations (`Cu,RIF`, `Cu,PROB`),
and both inhibition constants in Table 2 are unbound constants, but neither the
main text nor Supplementary Material S1 ever states the `fu` used to convert the
modelled *total* plasma concentrations to unbound. Both values come from the
upstream models the paper explicitly adopts, and neither is verifiable from the
Ujihira 2025 sources alone. They are therefore pinned against two on-disk answer
keys instead:

* **Rifampicin `fu` = 0.11.** Carried by the sibling nlmixr2lib extractions
  `Barnett_2018_coproporphyrin_I` and `Barnett_2018_rosuvastatin`, whose
  source-trace comments record Barnett 2018 Table 1 footnote c converting
  `Ki,total` = 1.15 uM to `Ki,unbound` = 0.13 uM and 2.23 uM to 0.25 uM at
  `fu` = 0.11. Ujihira 2025 Table S7 independently lists that same Barnett study
  at `Ki,u` = 0.13 uM, so the two agree.
* **Probenecid `fu` = 0.062.** Attributed to Ahmad 2021, the source of this
  model's probenecid structure (Methods, "Structural models"). Ahmad 2021 is
  *not* among this extraction's source files, so the attribution itself could
  not be checked.

The two interaction checks above are what actually pin these values, and both
are on-disk answer keys:

* `fu_rif` is pinned by the rifampicin AUCR check, which reproduces Table S4's
  predicted `AUCR0-24h` of 13 at 600 mg. That number is a product of the
  mg-to-umol conversion, the rifampicin PK, `fu_rif`, `Ki,u,OATP1B3`, and the
  GCDCA-S clearance split; a materially different `fu` moves it.
* `fu_prob` is pinned by the probenecid check, which reproduces Table S4's
  steady-state `CLR` ratio of 0.1. Recovering 0.1 requires
  `1 + Cu,PROB/Ki,u,OAT3` of about 10. The total average steady-state
  concentration under 500 mg QID is `1752 umol / (0.95 L/h * 6 h)` = 307 uM, so
  `fu` = 0.062 gives `Cu` = 19 uM and a ratio of 0.12 -- consistent with 0.1 at
  one significant figure, whereas `fu` = 1 gives 0.009 and `fu` = 0.03 gives
  0.23. Both alternatives are excluded.

**IIV scale.** Table 2's "IIV" column is a percentage whose scale (CV% versus
the Monolix `omega` on the log scale) is not stated. It is read here as CV% and
converted with `omega^2 = log(1 + CV^2)`, matching the convention used
throughout nlmixr2lib and by the sibling `Barnett_2018_coproporphyrin_I` model.
The reported RSEs support an SD-scale rather than a variance-scale reading: the
best-determined term, `ksyn` IIV at 61% with RSE 15%, matches the
`1/sqrt(2N)` = 14.4% expected for an SD estimate at N = 24, not the
`sqrt(2/N)` = 28.9% expected for a variance. The residual CV%-versus-`omega`
ambiguity changes `omega` by at most about 10% relative at the largest IIV and
less elsewhere.

**Molecular weights** are not in the paper. Doses in the source studies are in
mg while the model runs in umol, so rifampicin 822.94 g/mol and probenecid
285.34 g/mol are used for the conversion; both are already carried by the
nlmixr2lib registers and agree with PubChem to four significant figures.

**`CL RIF` units.** Table 2 prints the rifampicin clearance row as
`CL RIF (L)`. It must be L/h for Eq. 1 to balance dimensionally, and 6.7 L/h is
consistent with the 6.35 L/h of the upstream Takita 2022 rifampicin model, so
the printed unit is treated as a typographical omission of the time term.

**Not implemented, because the paper excluded them from the final model:** the
six diurnal-fluctuation functions applied to `ksyn` (Table S2; none improved
the fit, p > 0.9) and the sex effect on `ksyn` (Table S3; `COVSEX` = 0.065 with
RSE 404%, dOFV = 3). The sex covariate is recorded in the model file's
`covariatesDataExcluded` metadata so the screen is preserved without creating an
unused-covariate warning.

**Not implemented, because the paper did not have the data:** any effect of
rifampicin or probenecid on GCDCA-S *synthesis*. The paper is explicit that
rifampicin may both inhibit CYP7A1/CYP8B1 and induce SULT2A1, that this could
confound the `Ki,u,OATP1B3` estimate, and that its own hypothetical simulations
(Supplementary Eqs 7-10, Figure S5) show plasma AUC is highly sensitive to
`ksyn`. Those hypothetical variants are sensitivity analyses, not the fitted
model, and are not extracted. Enterohepatic recirculation is likewise excluded,
consistent with the paper's rationale that bile-acid sulfates undergo limited
intestinal reabsorption.

**Perpetrator co-administration.** Rifampicin and probenecid were never given
together in the source studies. The model supports both legs simultaneously,
but that combination is an extrapolation beyond the fitted data.

**Doses are in umol, not mg.** Because the whole system is parameterised in
umol and uM, dose records for `central_rif` and `depot_prob` must be supplied in
umol. Rifampicin doses additionally require `rate = -2` so that the modelled
zero-order absorption duration is applied.
