---
title: "Capecitabine (Urien 2005)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Capecitabine (Urien 2005)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Urien S, Rezai K, Lokiec F. Pharmacokinetic modelling of 5-FU
  production from capecitabine -- a population study in 40 adult patients with
  metastatic cancer. *J Pharmacokinet Pharmacodyn.* 2005;32(5-6):817-833.
* Article: <https://doi.org/10.1007/s10928-005-0018-2>

The model describes capecitabine and the three sequential metabolites in the
5-FU production pathway -- 5'-DFCR (5'-deoxy-5-fluorocytidine) -> 5'-DFUR
(5'-deoxy-5-fluorouridine) -> 5-FU (5-fluorouracil) -- following oral
capecitabine administration. The structure is a four-compartment chain
(Figure 2 of Urien 2005, Appendix A equations A.1-A.4). Capecitabine has a
real apparent volume V1/F (338 L); each metabolite has its apparent volume
fixed to 1 L (not identifiable from the data), so the metabolite-to-metabolite
fluxes are described directly by first-order rate constants K23, K34, K40
(paper notation).

# Population

The model was fit to 1426 plasma concentrations (373 capecitabine, 354 5'-DFCR,
363 5'-DFUR, 336 5-FU) from 40 adult patients (25 male, 15 female) with
metastatic cancer enrolled in two phase I studies at Centre Rene Huguenin,
Saint-Cloud, France. Capecitabine 1400-2300 mg/m^2/day (median 2000) was
administered orally every 12 hours within 30 minutes of breakfast or dinner;
75 PK courses (most patients had two PK days, days 1 and 15) contributed to
the dataset. Patients received capecitabine in combination with either
irinotecan (200-250 mg/m^2 IV over 90 min on day 1) or irofulven (0.4 mg/kg
IV over 30 min on day 1), depending on the specific phase I study.

Baseline demographics from Table I: age 30-73 years (median 54.5), body weight
41.5-95 kg (median 68), height 150-178 cm (median 169), body surface area
1.40-2.10 m^2 (median 1.80), serum albumin 30-46 g/L (median 37), serum
creatinine 58-113 umol/L (median 85), and total bilirubin 3-22 umol/L
(median 8.8). The same information is available programmatically via
`readModelDb("Urien_2005_capecitabine")$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Urien_2005_capecitabine.R`. The
table below collects them in one place.

| Equation / parameter             | Value                       | Source location                                            |
|----------------------------------|-----------------------------|------------------------------------------------------------|
| `lka`                            | log(2.07 1/h)               | Table II row Ka (mean 2.07)                                |
| `llag`                           | log(0.28 h)                 | Table II row TLAG (mean 0.28)                              |
| `lvc`                            | log(338 L)                  | Table II row V1 (mean 338)                                 |
| `lcl`                            | log(218 L/h)                | Table II row TV.CL10 (mean 218); page 826 covariate eqn   |
| `e_tbili_cl`                     | +0.32                       | Table II row "BILT effect on CL10" (mean +0.32); page 826 |
| `lcl_dfcr`                       | log(12.9 L/h)               | Table II row CL12 (mean 12.9)                              |
| `lvc_dfcr` / `lvc_dfur` / `lvc_5fu` | log(1 L) FIXED          | Page 820 ("metabolite distribution volumes are not identifiable and fixed to 1") |
| `lk_dfur_form`                   | log(10.7 1/h)               | Table II row K23 (mean 10.7)                               |
| `lk_5fu_form`                    | log(5.30 1/h)               | Table II row K34 (mean 5.30); see Errata for 5.30 vs 5.70 |
| `e_tbili_k_5fu_form`             | -0.36                       | Table II row "BILT effect on K34" (mean -0.36); page 826  |
| `lkel_5fu`                       | log(66 1/h)                 | Table II row K40 (mean 66)                                 |
| `etallag`                        | omega^2 = 0.793             | Table II row "ISV TLAG %" (110% -> log(1+1.10^2))         |
| `etalvc`                         | omega^2 = 1.047             | Table II row "ISV V1 %" (136% -> log(1+1.36^2))           |
| `etalcl`                         | omega^2 = 0.0319            | Table II row "ISV CL10 %" (18% -> log(1+0.18^2))          |
| `etalk_dfur_form`                | omega^2 = 0.223             | Table II row "ISV K23 %" (50% -> log(1+0.50^2))           |
| `etalk_5fu_form`                 | omega^2 = 0.0606            | Table II row "ISV K34 %" (25% -> log(1+0.25^2))           |
| `etalkel_5fu`                    | omega^2 = 0.109             | Table II row "ISV K40 %" (34% -> log(1+0.34^2))           |
| `addSd`                          | 3.83 umol/L                 | Table II row "Res. variability CAP umol"                  |
| `addSd_dfcr`                     | 3.72 umol/L                 | Table II row "Res. variability 5'-DFCR umol"              |
| `addSd_dfur`                     | 5.81 umol/L                 | Table II row "Res. variability 5'-DFUR umol"              |
| `addSd_5fu`                      | 0.64 umol/L                 | Table II row "Res. variability 5-FU umol"                 |
| `d/dt(depot)`                    | -ka * depot                 | rxode2 explicit form of the analytical Ka * D input (Appendix A) |
| `d/dt(central)`                  | ka*depot - (kel + kform_dfcr)*central | Appendix A eq A.1 (with X1 = central)            |
| `d/dt(central_dfcr)`             | kform_dfcr*central - kform_dfur*central_dfcr | Appendix A eq A.2                          |
| `d/dt(central_dfur)`             | kform_dfur*central_dfcr - kform_5fu*central_dfur | Appendix A eq A.3                       |
| `d/dt(central_5fu)`              | kform_5fu*central_dfur - kel_5fu*central_5fu | Appendix A eq A.4                          |

# Virtual cohort

Original observed data are not publicly available. The cohort below uses 100
virtual patients whose body weight, body surface area, and total bilirubin
distributions match the Table I demographics (pooled male/female; the paper
does not report covariate effects on weight or BSA, only on TBILI). Each
patient receives a single oral capecitabine dose of 4596 umol -- the
normalisation dose used in Figure 4 of Urien 2005, equivalent to ~1650 mg
(capecitabine MW 359.35 g/mol).

```{r cohort}
set.seed(20051209)
n_subj <- 100L

# Sample baseline covariates from log-normal-ish distributions whose median
# / range bracket Table I. TBILI is the only structurally relevant covariate
# in the model; WT and BSA are carried for documentation / NCA labelling
# only.
subj <- tibble::tibble(
  id    = seq_len(n_subj),
  WT    = round(rlnorm(n_subj, meanlog = log(68),  sdlog = 0.18), 1),
  BSA   = round(rlnorm(n_subj, meanlog = log(1.80), sdlog = 0.10), 2),
  TBILI = pmin(pmax(round(rlnorm(n_subj, meanlog = log(8.8), sdlog = 0.45), 1), 3), 22)
)

dose_amt <- 4596 # umol -- single oral dose, matches Figure 4 caption

obs_times <- seq(0.05, 12, by = 0.05)

# Build the event table with rxode2::et(): one dose row at t=0 into the
# `depot` compartment, then one observation row per element of obs_times
# with cmt = "Cc". The cmt = "Cc" placeholder triggers rxSolve to compute
# every output variable in the model (Cc, Cc_dfcr, Cc_dfur, Cc_5fu) at
# each sampled time -- the named output is irrelevant for downstream
# plotting because the simulation result carries all output columns.
make_one_subject <- function(row) {
  ev <- rxode2::et(amt = dose_amt, cmt = "depot", time = 0)
  ev <- rxode2::et(ev, time = obs_times, cmt = "Cc")
  ev <- as.data.frame(ev)
  ev$id    <- row$id
  ev$WT    <- row$WT
  ev$BSA   <- row$BSA
  ev$TBILI <- row$TBILI
  ev
}

events <- do.call(
  rbind,
  lapply(seq_len(nrow(subj)), function(i) make_one_subject(subj[i, ]))
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Urien_2005_capecitabine")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("WT", "BSA", "TBILI")
) |>
  as.data.frame()
```

For deterministic replication of typical-value profiles (Figure 4 of the
paper depicts log-scale predicted concentrations from the bootstrap mean
parameters), zero out the random effects and re-solve a single typical-
TBILI patient.

```{r simulate-typical}
mod_typ <- mod |> rxode2::zeroRe()
events_typ <- {
  ev <- rxode2::et(amt = dose_amt, cmt = "depot", time = 0)
  ev <- rxode2::et(ev, time = obs_times, cmt = "Cc")
  ev <- as.data.frame(ev)
  ev$id    <- 1L
  ev$TBILI <- 8.8
  ev
}
sim_typ <- rxode2::rxSolve(mod_typ, events = events_typ) |>
  as.data.frame()
```

# Replicate published figures

## Figure 4 -- log-scale predicted profiles

Replicates Figure 4 of Urien 2005. The paper's Figure 4 shows log-scale
observed and predicted plasma concentrations of capecitabine and the three
metabolites against time, normalised for a 4596 umol dose. The simulation
below shows the typical-value (zeroRe) prediction stacked against the
stochastic VPC median at the same dose.

```{r figure-4}
sim_long <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::select(id, time, Cc, Cc_dfcr, Cc_dfur, Cc_5fu) |>
  tidyr::pivot_longer(
    cols      = c(Cc, Cc_dfcr, Cc_dfur, Cc_5fu),
    names_to  = "analyte",
    values_to = "conc"
  ) |>
  dplyr::mutate(analyte = factor(
    analyte,
    levels = c("Cc", "Cc_dfcr", "Cc_dfur", "Cc_5fu"),
    labels = c("capecitabine", "5'-DFCR", "5'-DFUR", "5-FU")
  ))

vpc_summary <- sim_long |>
  dplyr::group_by(time, analyte) |>
  dplyr::summarise(
    Q05 = quantile(conc, 0.05, na.rm = TRUE),
    Q50 = quantile(conc, 0.50, na.rm = TRUE),
    Q95 = quantile(conc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

typ_long <- sim_typ |>
  dplyr::filter(time > 0) |>
  dplyr::select(time, Cc, Cc_dfcr, Cc_dfur, Cc_5fu) |>
  tidyr::pivot_longer(
    cols      = c(Cc, Cc_dfcr, Cc_dfur, Cc_5fu),
    names_to  = "analyte",
    values_to = "conc"
  ) |>
  dplyr::mutate(analyte = factor(
    analyte,
    levels = c("Cc", "Cc_dfcr", "Cc_dfur", "Cc_5fu"),
    labels = c("capecitabine", "5'-DFCR", "5'-DFUR", "5-FU")
  ))

ggplot(vpc_summary, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20, fill = "steelblue") +
  geom_line(colour = "steelblue") +
  geom_line(
    data    = typ_long,
    mapping = aes(time, conc),
    colour  = "black",
    linetype = "dashed"
  ) +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(
    x       = "Time after dose (h)",
    y       = "Plasma concentration (umol/L)",
    title   = "Figure 4 replication -- log-scale predicted profiles",
    caption = "Solid line / ribbon: stochastic VPC median (5th-95th percentile, 100 virtual patients). Dashed line: typical-value (zeroRe) prediction at TBILI = 8.8 umol/L."
  )
```

## Bilirubin sensitivity (paper page 826 covariate model)

The paper reports that varying bilirubin by +/-30% from the median 8.8 umol/L
produced -2% to +2% changes in 5'-DFUR AUC and +12% to -8% changes in
capecitabine, 5'-DFCR, and 5-FU exposures (page 830 Discussion).

```{r bilirubin-sensitivity}
bilirubin_grid <- c(6.16, 8.8, 11.44) # 8.8 +/- 30%

sens <- lapply(bilirubin_grid, function(b) {
  ev <- events_typ |> dplyr::mutate(TBILI = b)
  rxode2::rxSolve(mod_typ, events = ev) |>
    as.data.frame() |>
    dplyr::filter(time > 0) |>
    dplyr::mutate(TBILI_label = sprintf("TBILI = %.2f umol/L", b))
}) |>
  dplyr::bind_rows()

sens_long <- sens |>
  dplyr::select(time, TBILI_label, Cc, Cc_dfcr, Cc_dfur, Cc_5fu) |>
  tidyr::pivot_longer(
    cols      = c(Cc, Cc_dfcr, Cc_dfur, Cc_5fu),
    names_to  = "analyte",
    values_to = "conc"
  ) |>
  dplyr::mutate(analyte = factor(
    analyte,
    levels = c("Cc", "Cc_dfcr", "Cc_dfur", "Cc_5fu"),
    labels = c("capecitabine", "5'-DFCR", "5'-DFUR", "5-FU")
  ))

ggplot(sens_long, aes(time, conc, colour = TBILI_label)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  scale_y_log10() +
  labs(
    x      = "Time after dose (h)",
    y      = "Plasma concentration (umol/L)",
    colour = NULL,
    title  = "Bilirubin sensitivity at +/-30% of the median (8.8 umol/L)"
  )
```

# PKNCA validation

PKNCA is run separately for each of the four analytes. The treatment
grouping variable is `analyte` (one entry per analyte) so per-analyte
NCA summaries can be compared against the AUC ranges shown in
Figure 6 of Urien 2005.

```{r pknca-cap}
sim_nca_cap <- sim |>
  dplyr::filter(time > 0, !is.na(Cc)) |>
  dplyr::transmute(id, time, Cc, treatment = "capecitabine")

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::transmute(id, time, amt, treatment = "capecitabine")

conc_obj_cap <- PKNCA::PKNCAconc(
  sim_nca_cap, Cc ~ time | treatment + id,
  concu = "umol/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id,
  doseu = "umol"
)

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  aucinf.obs  = TRUE,
  half.life   = TRUE
)

nca_cap <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj_cap, dose_obj,
                                          intervals = intervals))
nca_cap_summary <- summary(nca_cap)
knitr::kable(nca_cap_summary,
             caption = "Capecitabine simulated NCA (single oral dose, 4596 umol).")
```

```{r pknca-dfcr}
sim_nca_dfcr <- sim |>
  dplyr::filter(time > 0, !is.na(Cc_dfcr)) |>
  dplyr::transmute(id, time, conc = Cc_dfcr, treatment = "5'-DFCR")

conc_obj_dfcr <- PKNCA::PKNCAconc(
  sim_nca_dfcr, conc ~ time | treatment + id,
  concu = "umol/L", timeu = "h"
)
dose_df_dfcr <- dose_df |> dplyr::mutate(treatment = "5'-DFCR")
dose_obj_dfcr <- PKNCA::PKNCAdose(
  dose_df_dfcr, amt ~ time | treatment + id,
  doseu = "umol"
)
nca_dfcr <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj_dfcr, dose_obj_dfcr, intervals = intervals
))
knitr::kable(summary(nca_dfcr),
             caption = "5'-DFCR simulated NCA.")
```

```{r pknca-dfur}
sim_nca_dfur <- sim |>
  dplyr::filter(time > 0, !is.na(Cc_dfur)) |>
  dplyr::transmute(id, time, conc = Cc_dfur, treatment = "5'-DFUR")

conc_obj_dfur <- PKNCA::PKNCAconc(
  sim_nca_dfur, conc ~ time | treatment + id,
  concu = "umol/L", timeu = "h"
)
dose_df_dfur <- dose_df |> dplyr::mutate(treatment = "5'-DFUR")
dose_obj_dfur <- PKNCA::PKNCAdose(
  dose_df_dfur, amt ~ time | treatment + id,
  doseu = "umol"
)
nca_dfur <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj_dfur, dose_obj_dfur, intervals = intervals
))
knitr::kable(summary(nca_dfur),
             caption = "5'-DFUR simulated NCA.")
```

```{r pknca-5fu}
sim_nca_5fu <- sim |>
  dplyr::filter(time > 0, !is.na(Cc_5fu)) |>
  dplyr::transmute(id, time, conc = Cc_5fu, treatment = "5-FU")

conc_obj_5fu <- PKNCA::PKNCAconc(
  sim_nca_5fu, conc ~ time | treatment + id,
  concu = "umol/L", timeu = "h"
)
dose_df_5fu <- dose_df |> dplyr::mutate(treatment = "5-FU")
dose_obj_5fu <- PKNCA::PKNCAdose(
  dose_df_5fu, amt ~ time | treatment + id,
  doseu = "umol"
)
nca_5fu <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj_5fu, dose_obj_5fu, intervals = intervals
))
knitr::kable(summary(nca_5fu),
             caption = "5-FU simulated NCA.")
```

## Comparison against published exposures

The paper reports per-individual AUCs in Figure 6 (capecitabine vs each
metabolite, with Pearson r values) but does not tabulate population mean /
median NCA values directly. The visible AUC ranges in Figure 6 are roughly:

| Analyte         | Approximate AUC range from Figure 6 (umol*h/L) |
|-----------------|-----------------------------------------------|
| capecitabine    | 15-45                                         |
| 5'-DFCR         | 10-65                                         |
| 5'-DFUR         | 25-90                                         |
| 5-FU            | 2-12                                          |

```{r exposure-comparison}
auc_summary <- function(nca_obj) {
  rows <- nca_obj$result |>
    dplyr::filter(PPTESTCD == "aucinf.obs", !is.na(PPORRES)) |>
    dplyr::summarise(
      median = median(PPORRES, na.rm = TRUE),
      P05    = quantile(PPORRES, 0.05, na.rm = TRUE),
      P95    = quantile(PPORRES, 0.95, na.rm = TRUE),
      n      = sum(!is.na(PPORRES))
    )
  rows
}

cap_auc  <- auc_summary(nca_cap)  |> dplyr::mutate(analyte = "capecitabine")
dfcr_auc <- auc_summary(nca_dfcr) |> dplyr::mutate(analyte = "5'-DFCR")
dfur_auc <- auc_summary(nca_dfur) |> dplyr::mutate(analyte = "5'-DFUR")
fu_auc   <- auc_summary(nca_5fu)  |> dplyr::mutate(analyte = "5-FU")

reported <- tibble::tibble(
  analyte         = c("capecitabine", "5'-DFCR", "5'-DFUR", "5-FU"),
  reported_low    = c(15, 10, 25, 2),
  reported_high   = c(45, 65, 90, 12)
)

dplyr::bind_rows(cap_auc, dfcr_auc, dfur_auc, fu_auc) |>
  dplyr::left_join(reported, by = "analyte") |>
  dplyr::transmute(
    Analyte         = analyte,
    `Sim median (umol*h/L)`     = round(median, 1),
    `Sim 5%-95% (umol*h/L)`     = sprintf("%.1f - %.1f", P05, P95),
    `Paper Fig 6 range (umol*h/L)` = sprintf("%d - %d", reported_low, reported_high)
  ) |>
  knitr::kable(caption = "Simulated single-dose AUC0-inf vs Figure 6 individual-AUC ranges from Urien 2005.")
```

# Assumptions and deviations

- **TBILI source-column rename.** The paper uses `BILT` (its NONMEM short
  label for total bilirubin) as the covariate column. The canonical
  `inst/references/covariate-columns.md` register names this `TBILI` and
  registers `BILT` as a source alias; downstream users should rename
  `BILT -> TBILI` (no value transformation, no reference-category flip)
  before passing the dataset to `rxSolve()`.
- **Inter-occasion variability on Ka not represented.** Urien 2005
  reports a 167% IOV on Ka (Table II "IOV Ka %") that arose after
  deletion of the inter-subject variability on Ka. IOV requires an
  occasion column on the dosing dataset and is not represented in this
  static model file. Simulations therefore underpredict the
  course-to-course spread in absorption rate; the absorption-related
  TLAG variability (ISV 110%) does propagate.
- **Residual-error covariances not propagated.** Urien 2005 reports
  bootstrap covariances between residual-error pairs
  (cov(eps_capecitabine, eps_5'-DFCR) = 3.1; cov(eps_5'-DFUR,
  eps_5-FU) = 2.6). nlmixr2's residual-error syntax in `model()` does
  not directly support cross-output residual covariances; the four
  additive residual SDs are independent here. The impact is on
  individual-level prediction intervals (the published model treats
  consecutive metabolite measurements as having correlated noise);
  population-level predictions are unaffected.
- **K34 typical value 5.30 vs 5.70.** The covariate equation on page
  826 reads "K34 (h^-1) = 5.70 * (BILT/8.8)^-0.36", whereas Table II
  (the bootstrap-validated final estimates table) reports the K34
  bootstrap mean as 5.30 1/h with median 4.30 (95% CI 3.0-7.60). The
  number in the page-826 formula is within the 95% CI but does not
  match the reported bootstrap mean; we follow the standing rule that
  the official summary table is authoritative and use 5.30. The
  population-typical K34 values from page 826 (5.70) and Table II
  (5.30) differ by ~7%, which is within the 95% CI half-width and does
  not materially affect simulation predictions.
- **FO method.** The paper switched from FOCE-INTER to FO because FOCE
  produced abnormal terminations on this complex four-compartment
  chain (Results page 823). The reported parameters are FO estimates;
  FOCE estimates would differ slightly but the FO point estimates are
  what the published model is. nlmixr2 simulation does not depend on
  the estimation method; this note is for users who would re-fit the
  model to new data.
- **Combination therapy not modelled.** Patients received capecitabine
  with either irinotecan or irofulven (Methods). The paper tested
  combined-drug indicators as covariates but they did not reach the
  retention criteria (>=11 OFV-unit improvement, RSE <50%). The
  packaged model has no combination-therapy covariate; this matches
  Urien 2005's final model exactly.
- **Pediatric / non-cancer use.** The model was developed in adults
  with metastatic cancer receiving second- or third-line chemotherapy
  (TBILI 3-22 umol/L). Extrapolation to populations outside this range
  -- pediatrics, hepatic impairment beyond TBILI 22 umol/L, healthy
  volunteers -- is not supported by the source data.
