---
title: "Quizartinib and AC886 (Vaddady 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quizartinib and AC886 (Vaddady 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
mod <- readModelDb("Vaddady_2024_quizartinib")
ui  <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1111/cts.70074> (open access, PMC11600623)
* Supporting Information: model Equations 8-20 and both final NONMEM control
  streams are distributed with the article and were used throughout this
  extraction.

`r ui$description`

# Population

The analysis pooled 932 subjects contributing 14,160 quizartinib and 13,399
AC886 plasma concentrations across 13 studies (nine Phase 1, two Phase 2, two
Phase 3), summarised in Vaddady 2024 Table 1. The pooled analysis set comprises
294 newly diagnosed FLT3-ITD-positive AML patients from the Phase 3
QuANTUM-First trial (AC220-A-U302), 365 relapsed/refractory (R/R)
FLT3-ITD-positive AML patients, and 273 non-AML subjects (healthy volunteers
and subjects with hepatic impairment). Baseline demographics (Vaddady 2024
Table 2): age 18-91 years (median 50.0), weight 36.8-153 kg (median 72.0),
46.9% female, 65.4% White / 18.1% Asian / 8.5% Black / 4.8% Other. Doses ranged
from 20 to 90 mg/day, given as tablet (86.4%) or oral solution (13.6%). CYP3A
inhibitor use was common: 19.7% strong, 18.9% moderate, 9.1% weak, 52.3% none.

The same information is available programmatically from the model's
`population` metadata:

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

Every `ini()` value carries an in-file comment naming its source location. The
table below collects them for review. All typical values and variances come
from **Vaddady 2024 Table 3**; the closed-form covariate relationships come from
**Supporting Information Equations 6-20**.

| Model element | Value | Source location |
|---|---|---|
| `lcl` quizartinib CL/F | 6.65 L/h | Table 3, Eq 8 |
| `lvc` quizartinib Vc/F | 371 L | Table 3, Eq 11 |
| `lq` quizartinib Q1/F | 40.7 L/h | Table 3, Eq 12 |
| `lvp` quizartinib Vp1/F | 312 L | Table 3, Eq 13 |
| `lq2` quizartinib Q2/F | 0.757 L/h | Table 3, Eq 14 |
| `lvp2` quizartinib Vp2/F | 91.9 L | Table 3, Eq 15 |
| `ltlag` absorption lag | 0.196 h | Table 3 |
| `lka` absorption rate | 1.10 1/h | Table 3, Eq 10 |
| `ld1` zero-order input duration | 0.710 h | Table 3 |
| `lfdepot` Frel at R/R AML reference | 1 (fixed) | Eq 9 (reference level) |
| `e_wt_cl`, `e_wt_vc` allometric exponents | 0.75, 1 (fixed) | Eq 6 |
| `e_cyp3a4_inh_strong_cl` | -0.301 | Table 3, Eq 8 |
| `e_race_black_cl` | -0.261 | Table 3, Eq 8 |
| `e_sexf_vc` | -0.169 | Table 3, Eq 11 |
| `e_age_vp` | 0.0152 1/year | Table 3, Eq 13 |
| `e_nonaml_ka` | -0.188 | Table 3, Eq 10 |
| `e_cyp3a4_inh_strong_fdepot` | 0.273 | Table 3, Eq 9 |
| `e_cyp3a4_inh_mod_fdepot` | 0.116 | Table 3, Eq 9 |
| `e_nonaml_fdepot` | 1.73 (a level, not a fractional change) | Table 3, Eq 9 |
| `e_trtph_*_fdepot` induction / consolidation / continuation | -0.419 / -0.192 / 0.418 | Table 3, Eq 9 |
| `boxcox_fdepot` | -1.28 | Table 3, Eq 2 |
| `lcl_ac886` AC886 CL/F | 4.61 L/h | Table 3, Eq 17 |
| `lvc_ac886` AC886 Vc/F | 8.93 L | Table 3, Eq 18 |
| `lq_ac886` AC886 Q/F | 3.76 L/h | Table 3, Eq 19 |
| `lvp_ac886` AC886 Vp/F | 68.5 L | Table 3, Eq 20 |
| `fmet_base` conversion fraction | 0.5 (fixed) | Methods, Eq 16 |
| `e_nonaml_cl_ac886` | 0.843 | Table 3, Eq 17 |
| `e_race_black_cl_ac886` | 0.488 | Table 3, Eq 17 |
| `e_cyp3a4_inh_strong_cl_ac886` | 0.298 | Table 3, Eq 17 |
| `e_cyp3a4_inh_strong_vc_ac886` | 2.79 | Table 3, Eq 18 |
| `e_trtph_*_fmet` induction / consolidation / continuation | 0.715 / 0.272 / -0.249 | Table 3, Eq 16 |
| IIV variances (`etalcl`, `etalq`, `etalvc`, `etaltlag`, `etalka`, `etald1`) | 0.695, 0.691, 0.186, 0.647, 0.423, 0.821 (as SD) | Table 3 |
| IIV Frel, AML / non-AML | 0.444 / 0.256 (as SD) | Table 3 |
| IIV CL AC886, AML / non-AML | 0.740 / 0.516 (as SD) | Table 3 |
| IIV Vc AC886 | 1.36 (as SD) | Table 3 |
| `propSd`, `propSd_ac886` | 0.440, 0.452 | Table 3 (AML patients) |
| Reference subject (75 kg, 47 years, male, R/R AML, no CYP3A, not Black) | n/a | Eq 6, Eq 13, Figure 4 caption |
| Molecular weights 560.68 / 576.67 g/mol | n/a | Methods |
| ODE structure | n/a | Figure 1a/1b; Supporting Information NONMEM streams |

## Three confirmations that Table 3 holds final estimates on the SD scale

1. **Final, not initial.** The final AC886 control stream re-states every
   quizartinib parameter as a `FIX` row at higher precision (`6.6545 FIX ; 1_CL`
   against Table 3's 6.65; `371.079 FIX` against 371; `40.7235 FIX` against
   40.7). The non-`FIX` `$THETA` rows in both streams are initial estimates and
   differ materially (quizartinib CL initial 6.85443; AC886 CL initial 4.81815
   against a final 4.61).
2. **Table 3's "(CV)" column is the log-scale SD, not a variance and not a true
   lognormal CV.** Each value squares to the corresponding `$OMEGA` / `$SIGMA`
   entry: `0.695^2 = 0.483` against `2_IIV_CL_AML = 0.481313`;
   `0.740^2 = 0.548` against AC886 `2_CLM = 0.542698`;
   `1.36^2 = 1.850` against `3_V5 = 1.85491`;
   `0.440^2 = 0.194` against `$SIGMA 1_RUV_AML = 0.194347`.
3. **The paper's own prose reads the column the same way.** The Abstract
   reports "approximately 70% coefficient of variation for systemic clearances"
   and the Discussion "a CV of 44% for Frel and 74% for CLAC886" -- i.e. the
   omega values 0.695, 0.444 and 0.740 multiplied by 100, with no lognormal
   back-transformation. (A true lognormal CV for omega = 0.695 would be
   `sqrt(exp(0.695^2) - 1) = 79%`.)

The model file therefore stores each variance as `omega^2` with the reported
omega visible in the expression.

# Model structure notes

Three implementation points are worth stating before the simulations.

**Sequential zero- then first-order absorption.** The dose enters `depot` at a
zero-order rate over `D1 = 0.710 h`, beginning after `Tlag = 0.196 h`, and then
transfers to `central` by first-order `ka`. rxode2 applies a modelled `dur()`
only to dose records carrying `rate = -2`; a plain bolus silently ignores it and
collapses Tmax onto the lag time. **Every dose record below sets `rate = -2`.**

**Two declared endpoints.** The model declares both `Cc` and `Cc_ac886` with
residual error, so rxode2 builds endpoint pseudo-compartments after the six ODE
states. Observation records therefore address the **endpoint name**
(`cmt = "Cc"`), not an ODE state; `cmt = "central"` fails with a
`'dvid'->'cmt' ... undefined compartment` error for this model class. Every
`rxSolve()` call also passes `useLinCmt = FALSE`, because the automatic
ODE-to-linCmt conversion corrupts the dvid mapping. `rxSolve()` returns both
observables as columns regardless of which endpoint the row names.

**The fMET reparameterization.** No intravenous data were available, so the
parent-to-metabolite conversion fraction is not identifiable and is fixed at
0.5. Following the authors' control stream, the model routes the entire parent
elimination flux into `central_ac886` (`K25 = CL/V2`, `K20 = 0`) and divides all
four AC886 disposition parameters by fMET. The fMET factors cancel in every rate
constant, so fMET acts purely as a multiplicative scale on predicted AC886
concentrations -- exactly as the authors intended, and the reason they could
state that "the model fit is not affected by the value of fMET".

# Virtual cohort

Original observed data are not publicly available. The deterministic scenarios
below use the paper's own reference subject (Vaddady 2024 Figure 4 caption:
75 kg, 47 years, male, R/R AML patient, no CYP3A coadministration, not Black)
with one covariate varied at a time -- the construction Figure 4 uses.

```{r covariates}
reference_subject <- tibble::tibble(
  WT = 75, AGE = 47, SEXF = 0, RACE_BLACK = 0, DIS_AML = 1,
  CONMED_CYP3A4_INH_STRONG = 0, CONMED_CYP3A4_INH_MOD = 0,
  TRTPH_INDUCTION = 0, TRTPH_CONSOLIDATION = 0, TRTPH_CONTINUATION = 0
)

vary <- function(...) {
  out <- reference_subject
  repl <- list(...)
  out[names(repl)] <- repl
  out
}

scenarios <- dplyr::bind_rows(
  vary()                                     |> dplyr::mutate(scenario = "Reference (R/R AML)"),
  vary(CONMED_CYP3A4_INH_STRONG = 1)         |> dplyr::mutate(scenario = "Strong CYP3A inhibitor"),
  vary(CONMED_CYP3A4_INH_MOD = 1)            |> dplyr::mutate(scenario = "Moderate CYP3A inhibitor"),
  vary(DIS_AML = 0)                          |> dplyr::mutate(scenario = "Non-AML subject"),
  vary(RACE_BLACK = 1)                       |> dplyr::mutate(scenario = "Black race"),
  vary(SEXF = 1)                             |> dplyr::mutate(scenario = "Female"),
  vary(TRTPH_INDUCTION = 1)                  |> dplyr::mutate(scenario = "Induction"),
  vary(TRTPH_CONSOLIDATION = 1)              |> dplyr::mutate(scenario = "Consolidation"),
  vary(TRTPH_CONTINUATION = 1)               |> dplyr::mutate(scenario = "Continuation")
) |>
  dplyr::mutate(id = dplyr::row_number(), .before = 1)

knitr::kable(
  scenarios |> dplyr::select(scenario, WT, AGE, SEXF, RACE_BLACK, DIS_AML,
                             CONMED_CYP3A4_INH_STRONG, CONMED_CYP3A4_INH_MOD,
                             TRTPH_INDUCTION, TRTPH_CONSOLIDATION, TRTPH_CONTINUATION),
  caption = "Covariate scenarios. One factor is varied at a time from the Vaddady 2024 Figure 4 reference subject."
)
```

Event tables are built as plain data frames (not `rxode2::et()`) so that the
`rate = -2` flag and the endpoint-named `cmt` column survive intact.

```{r event-builder}
DOSE_MG   <- 60    # a QuANTUM-First tablet strength (Vaddady 2024 Table 1)
TAU       <- 24    # once-daily dosing
N_DAYS_SS <- 42    # long enough for the slow deep compartment to equilibrate

build_events <- function(covariates, dose = DOSE_MG, n_days = N_DAYS_SS,
                         tau = TAU, grid = seq(0, TAU, by = 0.25)) {
  last_dose_time <- (n_days - 1) * tau
  doses <- covariates |>
    dplyr::mutate(time = 0, amt = dose, evid = 1L, rate = -2,
                  ii = tau, addl = as.integer(n_days - 1), cmt = "depot")
  obs <- covariates |>
    tidyr::crossing(time = last_dose_time + grid) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, rate = NA_real_,
                  ii = 0, addl = 0L, cmt = "Cc")
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events_ss <- build_events(scenarios)
stopifnot(!anyDuplicated(unique(events_ss[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate-ss}
sim_ss <- rxode2::rxSolve(
  mod, events = events_ss,
  omega = NA,            # typical-value (deterministic) replication of Figure 4
  useLinCmt = FALSE,     # the ODE-to-linCmt auto-conversion breaks the dvid map
  keep = c("scenario"),
  returnType = "data.frame"
)

# rxSolve silently drops subjects on some failures -- assert the count.
stopifnot(dplyr::n_distinct(sim_ss$id) == nrow(scenarios))
stopifnot(!anyNA(sim_ss$Cc), !anyNA(sim_ss$Cc_ac886))
```

## Steady-state concentration-time profiles

```{r figure-ss-profiles}
sim_ss |>
  dplyr::mutate(time_in_interval = time - min(time), .by = id) |>
  tidyr::pivot_longer(c(Cc, Cc_ac886), names_to = "moiety", values_to = "conc") |>
  dplyr::mutate(moiety = dplyr::recode(moiety,
                                       Cc = "Quizartinib", Cc_ac886 = "AC886")) |>
  ggplot(aes(time_in_interval, conc, colour = scenario)) +
  geom_line() +
  facet_wrap(~moiety, scales = "free_y") +
  labs(x = "Time within the dosing interval (h)", y = "Concentration (ng/mL)",
       colour = NULL,
       title = "Steady-state profiles, 60 mg once daily",
       caption = "Typical-value predictions for the Vaddady 2024 Figure 4 covariate scenarios.") +
  theme(legend.position = "bottom")
```

The quizartinib profile is strikingly flat across the dosing interval -- a
consequence of the very slow second peripheral compartment
(`Q2 = 0.757 L/h` into `Vp2 = 91.9 L`, a distribution half-life of roughly
84 h), which is what makes quizartinib a once-daily drug with a low
peak-to-trough ratio.

# PKNCA validation

Steady-state NCA over the final 24 h dosing interval, one PKNCA block per
moiety (the model has two outputs), grouped by scenario.

```{r pknca-setup}
last_dose_time <- (N_DAYS_SS - 1) * TAU

dose_df <- events_ss |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, scenario, amt) |>
  dplyr::mutate(time = last_dose_time)   # the dose that opens the SS interval

nca_intervals <- data.frame(
  start   = last_dose_time,
  end     = last_dose_time + TAU,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

run_nca <- function(conc_col) {
  conc_df <- sim_ss |>
    dplyr::select(id, scenario, time, Cc = dplyr::all_of(conc_col)) |>
    dplyr::filter(!is.na(Cc))
  conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc_df), Cc ~ time | scenario + id,
                               concu = "ng/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | scenario + id,
                               doseu = "mg")
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = nca_intervals))
}

nca_quiz  <- run_nca("Cc")
nca_ac886 <- run_nca("Cc_ac886")
```

```{r pknca-table}
tidy_nca <- function(res, moiety) {
  as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "tmax", "auclast", "cav")) |>
    dplyr::select(scenario, PPTESTCD, PPORRES) |>
    dplyr::mutate(moiety = moiety)
}

nca_long <- dplyr::bind_rows(
  tidy_nca(nca_quiz, "Quizartinib"),
  tidy_nca(nca_ac886, "AC886")
)

nca_long |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::select(moiety, scenario, auclast, cmax, cmin, tmax, cav) |>
  dplyr::arrange(dplyr::desc(moiety), scenario) |>
  dplyr::rename(
    "Moiety"                = moiety,
    "Scenario"              = scenario,
    "AUCss (ng*h/mL)"       = auclast,
    "Cmax,ss (ng/mL)"       = cmax,
    "Cmin,ss (ng/mL)"       = cmin,
    "Tmax,ss (h)"           = tmax,
    "Cavg,ss (ng/mL)"       = cav
  ) |>
  knitr::kable(digits = c(0, 0, 0, 1, 1, 2, 1),
               caption = "Steady-state NCA (PKNCA) over the final 24 h dosing interval, 60 mg once daily.")
```

## Structural identity check: AUCss = Frel x Dose / CL

At steady state, mass balance requires that the AUC over one dosing interval
equal `Frel * Dose / CL` exactly, for every subject. This is a per-subject
identity rather than a comparison of medians, so it is a sharp test of the
covariate algebra, the allometric scaling, and the units conversion all at once.

```{r identity-check}
identity_tbl <- as.data.frame(nca_quiz$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(scenario, auc_nca = PPORRES) |>
  dplyr::left_join(
    sim_ss |> dplyr::distinct(scenario, frel, cl),
    by = "scenario"
  ) |>
  # 1000 converts mg/L to ng/mL, matching the model's observation equation.
  dplyr::mutate(auc_theory = 1000 * frel * DOSE_MG / cl,
                pct_diff   = 100 * (auc_nca - auc_theory) / auc_theory)

identity_tbl |>
  dplyr::select(scenario, auc_nca, auc_theory, pct_diff) |>
  dplyr::rename("Scenario" = scenario,
                "AUCss by PKNCA (ng*h/mL)" = auc_nca,
                "Frel x Dose / CL (ng*h/mL)" = auc_theory,
                "% difference" = pct_diff) |>
  knitr::kable(digits = c(0, 0, 0, 3),
               caption = "Per-scenario steady-state mass-balance identity.")

# The residual gap is the finite approach to steady state after 42 daily doses,
# not a structural error. Assert it is under 0.5% everywhere.
stopifnot(nrow(identity_tbl) == nrow(scenarios))
stopifnot(all(abs(identity_tbl$pct_diff) < 0.5))
```

# Replicating the published covariate effects (Figure 4)

Vaddady 2024 Figure 4 is a forest plot of the relative difference in `AUCss`,
`Cmax,ss` and `Cmin,ss` from the reference subject, for quizartinib (panel a),
AC886 (panel b), and their sum (panel c). The paper's raster figure carries no
extractable numbers, but the Abstract, Results and Discussion state several of
the ratios explicitly. Those statements are the validation targets below.

```{r figure-4-ratios}
ratio_vs_reference <- function(df, moiety_label) {
  wide <- df |>
    tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
  ref <- wide |> dplyr::filter(scenario == "Reference (R/R AML)")
  wide |>
    dplyr::mutate(
      moiety = moiety_label,
      AUCss  = auclast / ref$auclast,
      Cmaxss = cmax    / ref$cmax,
      Cminss = cmin    / ref$cmin
    ) |>
    dplyr::select(moiety, scenario, AUCss, Cmaxss, Cminss)
}

# Panel (c) sums the two moieties on the ng/mL scale (Vaddady 2024 Methods:
# the molecular weights differ by under 3%, so the authors summed directly).
total_nca <- {
  conc_df <- sim_ss |>
    dplyr::mutate(Cc = Cc + Cc_ac886) |>
    dplyr::select(id, scenario, time, Cc)
  conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc_df), Cc ~ time | scenario + id,
                               concu = "ng/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | scenario + id,
                               doseu = "mg")
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = nca_intervals))
}

forest <- dplyr::bind_rows(
  ratio_vs_reference(tidy_nca(nca_quiz,  "x") |> dplyr::select(-moiety), "(a) Quizartinib"),
  ratio_vs_reference(tidy_nca(nca_ac886, "x") |> dplyr::select(-moiety), "(b) AC886"),
  ratio_vs_reference(tidy_nca(total_nca, "x") |> dplyr::select(-moiety), "(c) Quizartinib + AC886")
) |>
  dplyr::filter(scenario != "Reference (R/R AML)")

forest |>
  tidyr::pivot_longer(c(AUCss, Cmaxss, Cminss),
                      names_to = "metric", values_to = "ratio") |>
  ggplot(aes(ratio, scenario, colour = metric, shape = metric)) +
  geom_vline(xintercept = 1, colour = "grey40") +
  geom_vline(xintercept = c(0.8, 1.25), linetype = "dashed", colour = "grey60") +
  geom_point(position = position_dodge(width = 0.5), size = 2) +
  facet_wrap(~moiety) +
  scale_x_log10() +
  labs(x = "Ratio vs the reference subject", y = NULL, colour = NULL, shape = NULL,
       title = "Covariate effects on steady-state exposure",
       caption = paste("Replicates Figure 4 of Vaddady 2024. Dashed lines are the",
                       "80%-125% bioequivalence margins used in that figure.")) +
  theme(legend.position = "bottom")
```

## Comparison against the published effect sizes

Vaddady 2024 reports no absolute NCA table, so the comparison below is against
the fold-changes stated in the paper's own text rather than against
`ncaComparisonTable()` inputs (which expect absolute NCA parameters).

```{r published-comparison}
lookup <- function(moiety_label, scen, metric = "AUCss") {
  v <- forest[[metric]][forest$moiety == moiety_label & forest$scenario == scen]
  if (length(v) != 1L) {
    stop("no unique row for ", moiety_label, " / ", scen, " / ", metric)
  }
  v
}

published <- tibble::tribble(
  ~`Published claim`,                                                          ~Source,                 ~Published, ~Simulated,
  "Strong CYP3A inhibitor increases quizartinib AUCss 1.8-fold",               "Abstract; Results",     1.8,        lookup("(a) Quizartinib", "Strong CYP3A inhibitor"),
  "Quizartinib exposure ~1.7-fold higher in non-AML subjects",                 "Results",               1.7,        lookup("(a) Quizartinib", "Non-AML subject"),
  "Newly diagnosed AML, induction: 0.6-fold vs R/R reference",                 "Abstract; Results",     0.6,        lookup("(a) Quizartinib", "Induction"),
  "Newly diagnosed AML, consolidation: 0.8-fold vs R/R reference",             "Results",               0.8,        lookup("(a) Quizartinib", "Consolidation"),
  "Newly diagnosed AML, continuation: 1.4-fold vs R/R reference",              "Abstract; Results",     1.4,        lookup("(a) Quizartinib", "Continuation"),
  "Induction exposure ratio 0.58 (95% CI 0.54-0.62)",                          "Discussion",            0.58,       lookup("(a) Quizartinib", "Induction"),
  "Continuation exposure ratio 1.42 (95% CI 1.31-1.51)",                       "Discussion",            1.42,       lookup("(a) Quizartinib", "Continuation"),
  "AC886: Black race gives almost 0.7-fold exposure",                          "Results",               0.7,        lookup("(b) AC886", "Black race")
) |>
  dplyr::mutate(
    `% difference` = 100 * (Simulated - Published) / Published,
    Flag = ifelse(abs(`% difference`) > 20, "*", "")
  )

published |>
  knitr::kable(digits = c(0, 0, 2, 3, 1, 0),
               caption = paste("Simulated steady-state exposure ratios against the",
                               "fold-changes stated in Vaddady 2024.",
                               "* marks a difference above 20%."))

# No claim may drift by more than 20% -- this gate must be able to go red.
stopifnot(nrow(published) == 8L)
stopifnot(all(abs(published$`% difference`) <= 20))
```

Every stated effect size reproduces. The two Discussion values, quoted to three
significant figures, are matched to within 0.2%: the model returns
`r sprintf("%.3f", lookup("(a) Quizartinib", "Induction"))` against a published
0.58 for induction and
`r sprintf("%.3f", lookup("(a) Quizartinib", "Continuation"))` against a
published 1.42 for continuation. That precision is expected rather than
lucky -- the treatment phase acts only on `Frel`, so the ratio collapses
analytically to `1 + theta_phase` (`1 - 0.419 = 0.581` and
`1 + 0.418 = 1.418`), and recovering it through a full ODE solve plus PKNCA
integration confirms the absorption model, the dosing flags and the NCA
interval as well as the covariate algebra.

# Between-subject variability

The final section exercises the stochastic layer, including the two
disease-status-switched IIV strata and the Box-Cox transformed `Frel` eta for
AML patients. A single 60 mg dose is simulated in 200 subjects per arm.

```{r vpc-cohort}
set.seed(20241125)
N_PER_ARM <- 200

make_arm <- function(n, dis_aml, label, id_offset) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    # Weight and age sampled to approximate Vaddady 2024 Table 2 (median 72.0 kg,
    # median 50.0 years); the paper reports medians and ranges, not distributions.
    WT  = pmin(pmax(rnorm(n, mean = 72, sd = 16), 36.8), 153),
    AGE = pmin(pmax(rnorm(n, mean = 50, sd = 14), 18), 91),
    SEXF = rbinom(n, 1, 0.469),
    RACE_BLACK = rbinom(n, 1, 0.085),
    DIS_AML = dis_aml,
    CONMED_CYP3A4_INH_STRONG = 0, CONMED_CYP3A4_INH_MOD = 0,
    TRTPH_INDUCTION = 0, TRTPH_CONSOLIDATION = 0, TRTPH_CONTINUATION = 0,
    arm = label
  )
}

cohort <- dplyr::bind_rows(
  make_arm(N_PER_ARM, 1L, "AML patients",     id_offset = 0L),
  make_arm(N_PER_ARM, 0L, "Non-AML subjects", id_offset = N_PER_ARM)
)

events_sd <- build_events(cohort, n_days = 1, grid = c(seq(0, 24, by = 1),
                                                       seq(30, 336, by = 6)))
stopifnot(!anyDuplicated(unique(events_sd[, c("id", "time", "evid")])))

sim_sd <- rxode2::rxSolve(
  mod, events = events_sd, useLinCmt = FALSE,
  keep = c("arm"), returnType = "data.frame"
)
stopifnot(dplyr::n_distinct(sim_sd$id) == 2 * N_PER_ARM)
```

```{r vpc-plot}
sim_sd |>
  tidyr::pivot_longer(c(Cc, Cc_ac886), names_to = "moiety", values_to = "conc") |>
  dplyr::mutate(moiety = dplyr::recode(moiety,
                                       Cc = "Quizartinib", Cc_ac886 = "AC886")) |>
  dplyr::summarise(
    Q05 = quantile(conc, 0.05), Q50 = median(conc), Q95 = quantile(conc, 0.95),
    .by = c(time, arm, moiety)
  ) |>
  dplyr::filter(Q05 > 0) |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~moiety) +
  scale_y_log10() +
  labs(x = "Time after a single 60 mg dose (h)", y = "Concentration (ng/mL)",
       colour = NULL, fill = NULL,
       title = "Median and 5th-95th percentile prediction interval",
       caption = "200 subjects per arm. Non-AML subjects carry a 1.73-fold Frel and a narrower Frel IIV.") +
  theme(legend.position = "bottom")
```

The AML arm shows the wider spread the authors describe: the `Frel` IIV is
0.444 in AML patients against 0.256 in non-AML subjects, and the AC886 clearance
IIV is 0.740 against 0.516. The two arms' medians differ by roughly the 1.73-fold
`Frel` multiplier for quizartinib, while the AC886 medians nearly coincide
because the 1.73-fold higher formation input is offset by 1.843-fold higher
AC886 clearance in non-AML subjects.

```{r iiv-check}
# The reported IIV asymmetry must show up in the simulated individual
# parameters; assert it rather than eyeballing the ribbon.
iiv_summary <- sim_sd |>
  dplyr::distinct(id, arm, frel, cl_ac886) |>
  dplyr::summarise(sd_log_frel = sd(log(frel)),
                   sd_log_cl_ac886 = sd(log(cl_ac886)),
                   .by = arm)

knitr::kable(iiv_summary, digits = 3,
             caption = "Realised SD on the log scale of the simulated individual parameters.")

aml    <- iiv_summary[iiv_summary$arm == "AML patients", ]
nonaml <- iiv_summary[iiv_summary$arm == "Non-AML subjects", ]
stopifnot(nrow(aml) == 1L, nrow(nonaml) == 1L)
stopifnot(aml$sd_log_frel > nonaml$sd_log_frel)
stopifnot(aml$sd_log_cl_ac886 > nonaml$sd_log_cl_ac886)
# Non-AML Frel carries a plain exponential eta, so its realised log-scale SD
# should land near the reported omega of 0.256.
stopifnot(abs(nonaml$sd_log_frel - 0.256) < 0.05)
```

The AML `Frel` log-scale SD is deliberately **not** compared against 0.444:
that eta is Box-Cox transformed with `lambda = -1.28`, so the realised spread of
`Frel` is a nonlinear function of the underlying eta and does not equal the
reported omega. The non-AML arm, which carries a plain exponential eta, is the
one that can be checked directly against its reported omega, and it is.

# Assumptions and deviations

* **Residual error is carried for AML patients only.** Vaddady 2024 reports two
  residual SDs per moiety -- quizartinib 0.440 (AML) / 0.102 (non-AML) and
  AC886 0.452 (AML) / 0.312 (non-AML). nlmixr2 cannot express a
  covariate-switched residual error, so the model file carries the AML-patient
  value for each moiety: the population this analysis targets, and 659 of 932
  subjects. The non-AML values are recorded in the `ini()` comments and here.
  This follows the `Lahu_2010_roflumilast.R` precedent.
* **`DIS_AML` has an inverted reference category.** Vaddady 2024 Methods: "The
  AML patient group was set as the reference category versus which the effect of
  non-AML subjects was estimated." The source NONMEM column `AML3` is the
  *non-AML* indicator. The canonical `DIS_AML` column carries the AML indicator,
  so every non-AML effect is applied as `(1 - DIS_AML)`. This is a pure
  relabelling with no sign change to any published coefficient.
* **`Frel` for non-AML subjects is a level, not a fractional change.** Table 3
  reports 1.73 where every other categorical row reports a fractional change;
  the control stream confirms it (`F1AML2 = THETA(12)` directly, not
  `1 + THETA(12)`). The model applies it as
  `1 + (e_nonaml_fdepot - 1) * (1 - DIS_AML)` so the published 1.73 stays
  verbatim in `ini()`.
* **All three treatment phases carry their own indicator.** The `TRTPH_<phase>`
  family preamble normally folds a protocol's induction phase into the
  all-indicators-zero reference. Vaddady 2024 instead makes R/R AML patients on
  monotherapy the reference, "for whom no distinct treatment phases were
  reported", so induction, consolidation and continuation each need an
  indicator. The three columns were ratified for this extraction.
* **Amount units.** The source control streams carry amounts in micromoles and
  convert at the observation (`CP = A(2)*560.68/V2`). This model file carries mg
  (the natural dosing unit) and instead scales the AC886 formation flux by the
  molecular-weight ratio `576.67 / 560.68`, which is algebraically equivalent;
  observations are `1000 * amount / volume` to reach ng/mL.
* **Covariate distributions in the stochastic cohort are assumed.** Vaddady 2024
  Table 2 reports medians and ranges, not distributions. Weight and age are
  drawn from truncated normals matched to the published medians and ranges; sex
  and race are drawn from the published proportions.
* **A modelled `dur()` needs `rate = -2`.** Any event table built for this model
  must set `rate = -2` on dose records (or supply `dur =` directly), or rxode2
  silently ignores `D1` and collapses Tmax onto the lag time. This is a property
  of rxode2, not of the published model.
* **Non-mu-referenced etas.** The two disease-status-switched IIV strata cannot
  be mu-referenced, so nlmixr2 emits "some etas defaulted to non-mu referenced"
  when the model is loaded. This is inherent to the authors' structure and
  affects estimation efficiency only, not simulation.

## Errata and inconsistencies in the source

Two direction-of-effect slips in the paper's prose contradict its own Figure 4
and Table 3. Neither affects the extracted parameters; both are recorded so a
reader comparing this vignette against the paper is not misled.

1. **The Discussion inverts the treatment-phase ratio labels.** It reads: "The
   ratio of quizartinib steady-state exposure between R/R versus newly diagnosed
   AML patients was 0.58 ... during the induction phase, and it increased to
   1.42 ... during continuation." Those numbers are the *newly diagnosed versus
   R/R* ratios, not the reverse: the Abstract and Results both state the
   direction correctly ("dose-normalized AUCss values were 0.6-fold during
   induction ... and 1.4-fold during continuation compared to R/R AML
   patients"), and Table 3's coefficients give `1 - 0.419 = 0.581` and
   `1 + 0.418 = 1.418` for newly diagnosed patients relative to the R/R
   reference. The magnitudes are right; the two groups are named the wrong way
   round.
2. **"Approximately 40% lower" total exposure is a reciprocal slip.** The
   Discussion states that total quizartinib plus AC886 exposure in AML patients
   "was approximately 40% lower compared to the non-AML population (Figure 4c)".
   The model gives a non-AML to AML total-exposure ratio of
   `r sprintf("%.2f", lookup("(c) Quizartinib + AC886", "Non-AML subject"))`,
   i.e. non-AML exposure is about 40% *higher*, which corresponds to AML
   exposure being about 28% lower. A genuine 40% reduction would require a ratio
   of 1.67. The Figure 4c value the sentence cites is consistent with the
   model's 1.4; the prose converted "1.4-fold higher" into "40% lower" instead
   of "29% lower".
