---
title: "Osimertinib target-site PBPK (van der Gaag 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Osimertinib target-site PBPK (van der Gaag 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("vanderGaag_2025_osimertinib_pbpk"))$reference`
* Article: <https://doi.org/10.1002/psp4.70006>
* Supplement (Appendices S1-S10, all equations and parameter tables):
  <https://doi.org/10.1002/psp4.70006> Supporting Information file
  `PSP4-14-918-s001.docx`
* Predecessor / framework model: Bartelink et al., *Pharmaceuticals* 2022;15(7):796,
  <https://doi.org/10.3390/ph15070796>

Osimertinib is a third-generation EGFR tyrosine kinase inhibitor and the standard
first-line therapy for advanced NSCLC with an activating *EGFR* mutation. Whether
enough drug reaches the tumour is not answerable from plasma PK alone, and
microdosed `[11C]C`-osimertinib PET can measure tissue uptake directly -- but only
at a microdose, where EGFR binding and metabolism are far from saturated. This
model exists to bridge that gap: a whole-body PBPK model carrying explicit,
saturable EGFR binding so that a 2.7 ug intravenous microdose and an 80 mg oral
therapeutic dose can be simulated with the same parameter set.

Two features distinguish it from a conventional PBPK model.

1. **No fitted partition coefficients.** Every tissue-to-plasma partition
   coefficient is *computed* inside the model from the Rodgers and Rowland
   moderate-to-strong-base equations (Suppl Eq 1-6), combining osimertinib's pKa,
   logD, fraction unbound and blood-to-plasma ratio with each tissue's fractional
   water, lipid, acidic-phospholipid and lysosomal composition. Nothing in this
   model was estimated from the clinical data.
2. **Saturable target binding as explicit states.** Each of the ten
   EGFR-expressing tissues carries a second state holding drug covalently bound
   to EGFR, so receptor occupancy is a simulated output rather than an
   assumption.

# Population

```{r population}
pop <- rxode2::rxode(readModelDb("vanderGaag_2025_osimertinib_pbpk"))$population
str(pop, max.level = 1)
```

The model is deterministic and represents a single 87.5 kg reference individual
(Table S2); it carries no between-subject variability. Verification used two
independent data sets. Four patients with advanced-stage NSCLC and an activating
*EGFR* mutation (exon 19 deletion or exon 21 L858R) who had progressed on a
first-generation EGFR-TKI each underwent two dynamic `[11C]C`-osimertinib PET
scans, one on and one off concomitant TKI; because concomitant TKI did not change
uptake the scans were pooled. Three of the four also had a whole-body static PET
at 60 min, giving brain, kidney, liver, lung, spleen and tumour concentrations.
Separately, plasma concentration-time profiles from 10 healthy volunteers given a
single 80 mg oral dose were digitised from phase 1 study D5160C00001.

The authors adopted a 3-fold rather than the customary 2-fold prediction-error
window, justified by the small sample size and the high inter- and
intra-patient variability of tissue PK.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/vanderGaag_2025_osimertinib_pbpk.R` carries an in-file
comment naming its source location. The table below collects the model's
structural provenance in one place.

| Equation / parameter group | Source location |
|---|---|
| Organ total volumes, intracellular-water volumes, blood flows | Appendix S2, Table S1 |
| Body weight, pH of plasma / EW / IW / lysosomes / blood cells, haematocrit | Appendix S2, Table S2 |
| Tissue fractional volumes `fnl`, `fnp`, `few`, `fiw` | Appendix S2, Table S3 |
| Lysosomal fractional volumes, acidic phospholipids, tissue EGFR concentrations | Appendix S2, Table S4 |
| Liver and lung cell-type composition (`flys`, `fcell`, lysosomal pH) | Appendix S2, Table S5 |
| pKa, logD, logPV, `Fu`, blood-to-plasma ratio, `Ka`, `F`, `CL_int`, `CL_R`, `Ka_bc` | Appendix S3, Table S6 |
| Tumour EGFR, tumour pH_EW, tumour blood flow, virtual tumour volume, `Kon` / `Koff` (wild-type, mutated, 50/50) | Main text, Table 1 |
| `Kpu` for intracellular water, lipids, acidic phospholipids, lysosomes; `Kp = Kpu * Fu` | Appendix S4, Suppl Eq 1-6 |
| Non-eliminating tissue, lung, venous and arterial ODEs | Appendix S5, Suppl Eq 7-10 |
| EGFR-expressing tissue ODEs (bound and unbound states) | Appendix S5, Suppl Eq 11-12 |
| Absorption; gut ODE | Appendix S5, Suppl Eq 13-14 |
| Hepatic clearance and `CL_int` derivation; liver and kidney ODEs | Appendix S5, Suppl Eq 15-20 |
| EGFR association / dissociation; free-receptor fraction; tissue EGFR concentration | Main text, Equations 1, 2, 3 |
| Tumour extracellular-water pH term; tumour 100% residual-cell composition | Bartelink 2022 supplement, section VC and note 9 |

Two items come from the predecessor paper rather than from van der Gaag 2025.
Both are flagged in **Assumptions and deviations** below; the current paper
states explicitly that the NSCLC hallmarks "have been described and verified
previously by Bartelink et al.", and that paper is the cited framework source.

# Derived partition coefficients

Because the `Kp` values are computed rather than tabulated, the first validation
step is to inspect them. The chunk below solves the model once and reads the
derived quantities straight out of the model body.

```{r kp-table}
mod <- readModelDb("vanderGaag_2025_osimertinib_pbpk")

MW      <- 500      # g/mol, Table S6
nmol_mg <- 1e6 / MW # mg -> nmol

# A minimal solve: one observation record on a real ODE state so that every
# algebraic observable is evaluated and returned.
probe <- data.frame(
  id = 1L, time = 0, amt = NA_real_, evid = 0L, cmt = "venous"
)
kp_probe <- rxode2::rxSolve(mod, probe, useLinCmt = FALSE) |> as.data.frame()

tissues <- c("adipose", "brain", "gut", "heart", "kidney", "liver",
             "lung", "tumor", "pancreas", "muscle", "skin", "spleen", "other")

# Read the composite tissue-to-blood coefficient kb = Kpu * Fu / BP, which the
# ODEs actually use, and invert it. Reading `kpu_*` directly would be more
# direct but is not uniform: the rest-of-body compartment is pure extracellular
# water, so `kpu_other <- 1` is a literal constant and rxode2 folds it away
# rather than returning it as a column.
BP <- 0.77    # Table S6, blood-to-plasma ratio
FU <- 0.0535  # Table S6, fraction unbound in plasma

kb_vals <- vapply(paste0("kb_", tissues), \(n) kp_probe[[n]][1], numeric(1))

kp_tab <- tibble::tibble(
  Tissue = tissues,
  Kpu    = kb_vals * BP / FU,
  Kp     = kb_vals * BP
)

kp_tab |>
  dplyr::rename("Kpu (unbound tissue:plasma)" = Kpu,
                "Kp (tissue:plasma)"          = Kp) |>
  knitr::kable(digits = 2,
               caption = "Partition coefficients derived inside the model from Suppl Eq 1-6.")
```

An independent check on the whole Rodgers and Rowland chain: the steady-state
volume of distribution implied by these partition coefficients should match
osimertinib's known `Vss`. Summing `Kp * V` over the tissues plus the blood
volume gives

```{r vss-check}
v_tissue <- c(adipose = 10.4, brain = 1.49, gut = 1.5, heart = 0.31,
              kidney = 0.31, liver = 2.52, lung = 0.87, tumor = 0.05,
              pancreas = 0.21, muscle = 33.9, skin = 5.63, spleen = 0.20,
              other = 15.4)                                   # Table S1
vss <- sum(kp_tab$Kp * v_tissue[kp_tab$Tissue]) + 5.6         # + blood, Table S1
cat(sprintf("Vss implied by the computed Kp values: %.0f L\n", vss))

# Osimertinib's labelled Vss/F is 918 L; with the model's F = 0.70 (Table S6)
# that is an absolute Vss of ~643 L. The computed value should be the same
# order and within ~1.5-fold.
stopifnot(vss > 400, vss < 1400)
cat(sprintf("Labelled Vss/F 918 L; at F = 0.70 that is Vss ~ %.0f L\n", 918 * 0.70))
```

The dominant contributions are muscle and liver, and within each tissue the
acidic-phospholipid term dominates -- expected for a strong base (pKa 9.0) that
is 97.5% protonated at physiological pH.

# Microdose: 2.7 ug intravenous `[11C]C`-osimertinib

This reproduces the microdose arm used for the PET verification (Figure 3 of the
paper). The dose is given into venous blood.

```{r microdose}
micro_dose_nmol <- 2.7e-6 * nmol_mg * 1e3   # 2.7 ug -> nmol

micro_ev <- dplyr::bind_rows(
  data.frame(id = 1L, time = 0, amt = micro_dose_nmol, evid = 1L, cmt = "venous"),
  data.frame(id = 1L, time = seq(0, 2, length.out = 241),
             amt = NA_real_, evid = 0L, cmt = "venous")
) |>
  dplyr::arrange(time, dplyr::desc(evid))

micro <- rxode2::rxSolve(mod, micro_ev, useLinCmt = FALSE) |> as.data.frame()
if (is.null(micro$id)) micro$id <- 1L
```

```{r figure-3, fig.height = 5}
# Replicates Figure 3 of van der Gaag 2025: predicted tissue concentration-time
# profiles over the 2 h following a 2.7 ug IV microdose.
micro |>
  dplyr::select(time, Liver = Ct_liver, Spleen = Ct_spleen, Kidney = Ct_kidney,
                Lung = Ct_lung, Brain = Ct_brain, Tumour = Ct_tumor) |>
  tidyr::pivot_longer(-time, names_to = "Tissue", values_to = "conc") |>
  ggplot(aes(time, conc, colour = Tissue)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total tissue concentration (nM)",
       title = "Microdose: 2.7 ug IV [11C]C-osimertinib",
       caption = "Replicates Figure 3 of van der Gaag 2025.") +
  theme_bw()
```

The paper quotes four predicted tissue concentrations at 60 min post-dose
(Results section 3.2.1). Those are the reproduction target.

```{r microdose-compare}
at60 <- micro[which.min(abs(micro$time - 1)), ]

micro_cmp <- tibble::tribble(
  ~Tissue,   ~Published, ~Simulated,
  "Liver",   0.538,      at60$Ct_liver,
  "Spleen",  0.347,      at60$Ct_spleen,
  "Brain",   0.070,      at60$Ct_brain,
  "Tumour",  0.020,      at60$Ct_tumor
) |>
  dplyr::mutate(`Fold difference` = Simulated / Published)

micro_cmp |>
  dplyr::rename("Published prediction (nM)" = Published,
                "This model (nM)"           = Simulated) |>
  knitr::kable(digits = 3,
               caption = paste("Predicted tissue concentrations 60 min after a",
                               "2.7 ug IV microdose, against the values quoted",
                               "in section 3.2.1 of van der Gaag 2025."))
```

Liver reproduces to within 4%. Spleen is 2.6-fold high, brain 1.7-fold low and
tumour 8-fold high. Those three are discussed under **Assumptions and
deviations**; none of them was tuned away.

The paper also reports the tumour-to-lung contrast, which it treats as the
clinically meaningful readout because it is what distinguishes a sanctuary site
from an adequately exposed one.

```{r tumour-lung-ratio}
at6min <- micro[which.min(abs(micro$time - 0.1)), ]
cat(sprintf("Tumour:lung ratio at 6 min  %.3f\n", at6min$Ct_tumor / at6min$Ct_lung))
cat(sprintf("Tumour:lung ratio at 60 min %.3f\n", at60$Ct_tumor  / at60$Ct_lung))
```

## Mass balance

A whole-body PBPK model that silently leaks drug will still produce
plausible-looking profiles, so the mass balance is checked explicitly rather than
assumed. All 26 states are amounts in nmol, so they can simply be summed.

```{r mass-balance}
states <- c("gut_lumen", "venous", "arterial", "brain", "spleen", "other",
            "adipose", "adipose_complex", "gut", "gut_complex",
            "heart", "heart_complex", "kidney", "kidney_complex",
            "liver", "liver_complex", "lung", "lung_complex",
            "tumor", "tumor_complex", "pancreas", "pancreas_complex",
            "muscle", "muscle_complex", "skin", "skin_complex")
total <- rowSums(micro[, states])

cat(sprintf("Dose administered            %.4f nmol\n", micro_dose_nmol))
cat(sprintf("In the body at t = 0         %.4f nmol\n", total[1]))
cat(sprintf("In the body at t = 2 h       %.4f nmol\n", total[length(total)]))

# At t = 0 the entire dose must be accounted for exactly, and the body burden
# must decrease monotonically because the only outflows are hepatic and renal
# elimination (there is no re-entry route).
stopifnot(abs(total[1] - micro_dose_nmol) < 1e-8)
stopifnot(all(diff(total) <= 1e-9))
stopifnot(all(total > 0))
```

# Therapeutic dose: 80 mg oral

```{r therapeutic}
dose80 <- 80 * nmol_mg   # 80 mg -> nmol

ther_ev <- dplyr::bind_rows(
  data.frame(id = 1L, time = 0, amt = dose80, evid = 1L, cmt = "gut_lumen"),
  data.frame(id = 1L, time = sort(unique(c(seq(0, 24, by = 0.25),
                                           seq(24, 300, by = 1)))),
             amt = NA_real_, evid = 0L, cmt = "venous")
) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::mutate(treatment = "80 mg oral, single dose")

ther <- rxode2::rxSolve(mod, ther_ev, keep = "treatment", useLinCmt = FALSE) |>
  as.data.frame()
if (is.null(ther$id)) ther$id <- 1L

stopifnot(all(ther$Cc[ther$evid == 0] >= 0))
```

```{r figure-4a}
# Replicates Figure 4A of van der Gaag 2025: predicted plasma concentration-time
# profile after a single 80 mg oral dose.
ther |>
  dplyr::filter(time <= 150) |>
  ggplot(aes(time, Cc)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (h)", y = "Plasma osimertinib (nM)",
       title = "80 mg oral single dose",
       caption = "Replicates Figure 4A of van der Gaag 2025.") +
  theme_bw()
```

## PKNCA validation

```{r pknca}
sim_nca <- ther |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment); pre-dose Cc = 0 is correct for
# an extravascular dose.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- ther_ev |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r nca-compare}
# van der Gaag 2025 section 3.2.2 reports the model's own predictions for the
# 80 mg oral dose. Those are the reproduction target for this vignette.
published <- tibble::tribble(
  ~treatment,                ~cmax, ~tmax, ~half.life,
  "80 mg oral, single dose",  201,   8.8,   52.4
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "nM", tmax = "h", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated vs. the predictions published in van der Gaag 2025",
                  "section 3.2.2. * differs from the reference by >20%."),
  align = c("l", "l", "r", "r", "r")
)
```

For context, the clinical observations the paper compared its own predictions
against were `Cmax` 168 nM, `Tmax` 7 h and a terminal half-life of 48 +/- 2 h in
NSCLC patients. This model's `Tmax` sits on the observed value; its `Cmax` is
below both the published prediction and the observation, and its terminal
half-life is above both. Both discrepancies have the same origin -- a volume of
distribution larger than the one the published model behaves as if it had -- and
are discussed below.

# Steady state and EGFR target engagement

The paper's central pharmacological claim is that at 80 mg once daily every
EGFR-expressing organ, including the tumour, exceeds 75% receptor saturation,
with a mean of 88% (SD 5.9). That is a whole-model answer key: it depends on the
partition coefficients, the binding constants, the tissue EGFR concentrations and
the systemic exposure all being right simultaneously.

```{r steady-state}
ss_ev <- dplyr::bind_rows(
  data.frame(id = 1L, time = seq(0, 24 * 39, by = 24), amt = dose80,
             evid = 1L, cmt = "gut_lumen"),
  data.frame(id = 1L, time = seq(24 * 39, 24 * 40, length.out = 97),
             amt = NA_real_, evid = 0L, cmt = "venous")
) |>
  dplyr::arrange(time, dplyr::desc(evid))

ss <- rxode2::rxSolve(mod, ss_ev, useLinCmt = FALSE) |> as.data.frame()
ss <- ss[ss$time >= 24 * 39, ]

occ_cols <- c(Adipose = "occ_adipose", Gut = "occ_gut", Heart = "occ_heart",
              Kidney = "occ_kidney", Liver = "occ_liver", Lung = "occ_lung",
              Tumour = "occ_tumor", Pancreas = "occ_pancreas",
              Muscle = "occ_muscle", Skin = "occ_skin")

occ <- tibble::tibble(
  Tissue = names(occ_cols),
  `EGFR bound (%)` = vapply(occ_cols, \(n) mean(ss[[n]]) * 100, numeric(1))
) |>
  dplyr::arrange(dplyr::desc(`EGFR bound (%)`))

cat(sprintf("Css,avg (plasma) %.0f nM\n", mean(ss$Cc)))
cat(sprintf("Mean EGFR occupancy %.1f%% (published 88%%, SD 5.9); SD %.1f%%; minimum %.1f%%\n",
            mean(occ$`EGFR bound (%)`), sd(occ$`EGFR bound (%)`),
            min(occ$`EGFR bound (%)`)))

# The paper's explicit claim: every organ, tumour included, is above 75%.
stopifnot(all(occ$`EGFR bound (%)` > 75))
```

```{r figure-4c}
# Replicates Figure 4C of van der Gaag 2025: percentage of EGF receptor bound at
# the therapeutic average steady-state concentration, per EGFR-expressing organ.
occ |>
  ggplot(aes(reorder(Tissue, `EGFR bound (%)`), `EGFR bound (%)`)) +
  geom_col(fill = "grey35") +
  geom_hline(yintercept = 75, linetype = "dashed", colour = "firebrick") +
  coord_flip(ylim = c(0, 100)) +
  labs(x = NULL, y = "EGFR bound at Css,avg (%)",
       title = "Steady-state EGFR occupancy, 80 mg once daily",
       caption = paste("Replicates Figure 4C of van der Gaag 2025.",
                       "Dashed line: the paper's 75% saturation claim.")) +
  theme_bw()
```

```{r occupancy-table}
knitr::kable(occ, digits = 1,
             caption = "Steady-state EGFR occupancy by organ (published mean 88%, SD 5.9).")
```

Occupancy is a saturating function of the intracellular concentration, so it is a
relatively forgiving check -- but it is only forgiving in one direction, and the
model does clear the 75% floor in every organ, with the same ordering the paper
describes (lung, the tissue with by far the highest EGFR density at 1616 nM, is
the *least* saturated because there is more receptor to fill).

# Sensitivity to the unreported arterial/venous blood split

Table S1 gives a single total blood volume of 5.6 L, but Suppl Eq 9 and 10
require the venous and arterial pools separately and the split is not reported.
The model uses the conventional one-third arterial / two-thirds venous split of
the Jones et al. framework the paper cites. The chunk below shows that this
choice does not matter for any quantity validated here.

```{r blood-split-sensitivity}
split_check <- lapply(c(0.25, 1 / 3, 0.50), function(fa) {
  p <- c(lv_arterial = log(5.6 * fa), lv_venous = log(5.6 * (1 - fa)))
  a <- rxode2::rxSolve(mod, ther_ev, params = p, useLinCmt = FALSE) |> as.data.frame()
  b <- rxode2::rxSolve(mod, micro_ev, params = p, useLinCmt = FALSE) |> as.data.frame()
  tibble::tibble(
    `Arterial fraction`   = sprintf("%.0f%%", fa * 100),
    `Cmax (nM)`           = max(a$Cc, na.rm = TRUE),
    `Tmax (h)`            = a$time[which.max(a$Cc)],
    `Liver at 60 min (nM)` = b$Ct_liver[which.min(abs(b$time - 1))]
  )
}) |> dplyr::bind_rows()

relspread <- function(x) diff(range(x)) / mean(x)

knitr::kable(split_check, digits = 3,
             caption = "The unreported arterial:venous split does not affect any validated quantity.")

cat(sprintf("Relative spread across a 25%%-50%% arterial fraction: Cmax %.4f%%, liver at 60 min %.4f%%\n",
            relspread(split_check$`Cmax (nM)`) * 100,
            relspread(split_check$`Liver at 60 min (nM)`) * 100))

# Both quantities move by well under 0.1% across the whole plausible range of
# the split -- three orders of magnitude below the paper's own 3-fold
# acceptance window, and below the precision to which either is reported.
stopifnot(relspread(split_check$`Cmax (nM)`) < 1e-3)
stopifnot(relspread(split_check$`Liver at 60 min (nM)`) < 1e-3)
stopifnot(length(unique(split_check$`Tmax (h)`)) == 1L)
```

Doubling the arterial fraction from 25% to 50% moves `Cmax` by 0.006% and the
60-minute liver concentration by 0.02%, and leaves `Tmax` bit-identical. The
split is immaterial because every quantity compared against the paper is
measured at 60 min or later, by which time the two blood pools have
equilibrated; it would only matter in the first seconds after an intravenous
bolus.

# Assumptions and deviations

**Values taken from the predecessor paper rather than from van der Gaag 2025.**
Both are NSCLC tumour hallmarks that van der Gaag 2025 states were "described and
verified previously by Bartelink et al." and does not restate:

- The **tumour extracellular-water pH term**. Table 1 gives the tumour `pH_EW` as
  6.7, but Suppl Eq 5 as printed has no `pH_EW` term at all, so the tabulated
  value would have no effect. The Bartelink 2022 supplement (section VC) derives
  the strong-base tumour form explicitly and shows that the acidified
  extracellular water scales *only* the `few` term, by
  `(1 + 10^(pKa - pH_EW)) / (1 + 10^(pKa - pH_plasma))`, while the intracellular
  terms are unchanged because the two pH factors cancel. That is what the model
  implements.
- The **tumour cell-type composition**. van der Gaag 2025 describes "immune
  tumour deprivation (by reducing the cellular fraction of macrophages and type
  II cells)" without quantifying the reduction. Bartelink 2022 supplement note 9
  quantifies it: "In tumor we simulated only (100%) residual cells and in lung
  the following fractions: 4.1% alveolar macrophages, 8.3% type II cells and
  87.6% residual cells." The lung fractions match Table S5 of the current paper
  (0.04 / 0.08 / 0.88), which confirms the two papers use the same scheme, so the
  tumour is modelled as 100% residual cells.

**`P` versus `LogP` in the partition equations.** Suppl Eq 2 and Eq 5 print
`LogP * f_nl` where the octanol:water partition coefficient `P` itself is meant.
Two independent lines of evidence settle this. The Bartelink 2022 supplement
(section III, note 5) states that "the octanol/water partition coefficient (P) is
included for binding affinity to neutral lipids and phospholipids"; and Suppl Eq
5's own lysosomal sub-expression writes the same term as `P * Fnl`, without the
`Log`. The model therefore uses `P = 10^LogD` (and `10^LogPV` for adipose). Using
the literal `LogD = 3.2` instead would make the lipid term negligible and give a
lipophilic strong base essentially no adipose distribution.

**Arterial/venous blood volume split.** Not reported; set to the conventional
one-third arterial / two-thirds venous of the cited Jones et al. framework.
Demonstrated above to have no effect on any validated quantity.

**Rest-of-body compartment.** Table S1's "EW" compartment (15.4 L, blood flow
0.24 L/min) is described in the table footnote as extracellular water and has no
entry in the tissue-composition table (Table S3). Applying Suppl Eq 5 to a pure
extracellular-water space gives `f_ew = 1` and every other fractional volume
zero, hence `Kpu = 1`. It is mapped to the canonical `other` compartment.

**Absorption equation as printed.** Suppl Eq 13 is written
`dA_gutlumen/dt = -Ka * Dose * F`, i.e. against the administered dose rather than
the amount remaining in the lumen. Taken literally the lumen would empty linearly
and then go negative, and the `(1 - F)` fraction would never leave. The model
implements the standard and evidently intended reading: first-order loss from the
lumen, `-ka * gut_lumen`, with `F` applied as a bioavailability on the dosing
compartment.

**Lung and tumour binding terms.** Suppl Eq 8 (the lung) is the base-model
equation and carries no EGFR terms, but section 2.2.2 lists both lung and lung
tumour among the EGFR-expressing tissues, and Figures 4C and 4D report occupancy
for both. The Suppl Eq 11-12 binding terms are therefore applied to them, as they
are to the other eight EGFR-expressing tissues.

**Renal clearance denominator.** Suppl Eq 19 writes the renal term against
`Kp_kidney,IW`, a kidney intracellular-water partition coefficient that is defined
nowhere in the paper. The model uses `Kp_kidney`, matching the form of the
hepatic term in Suppl Eq 17. Renal clearance is 2% of total clearance, so the
choice is immaterial.

**Tumour EGFR concentration.** Main-text Table 1 gives 383 nM and Table S4 gives
382 nM for the same quantity. The model uses Table S4's 382 nM because Table S4 is
the tissue-parameter input table the other tissues are read from.

**Blood-flow bookkeeping.** As published, flow into the venous pool
(`Q_lung + Q_tumor` = 6.51 L/min) exceeds the summed arterial supply
(6.21 L/min) by about 5%. Mass is still conserved exactly -- each tissue's gain
matches the arterial loss and each tissue's loss matches the venous gain, as the
mass-balance check above confirms -- so the equations are implemented as
published rather than rebalanced.

**No between-subject variability and no residual error.** This is a bottom-up
model; nothing was estimated from the clinical data, so there is no IIV to
encode. nlmixr2 requires a residual-error term, so `propSd` is fixed at 0.10 as a
syntactic placeholder and must not be read as an estimate. This follows the
convention already used by `Aoki_2024_bosentan_pbpk`, `Mi_2023_cefquinome_pbpk`
and `An_2012_mitoxantrone_*_pbpk`.

**Quantitative differences from the published predictions.** Reproducing the
model from its published equations and tabulated inputs gives a `Cmax` about
1.6-fold below, and a terminal half-life about 1.7-fold above, the values quoted
in section 3.2.2 -- both consistent with a single cause, a volume of distribution
larger than the published model behaves as if it had. The same offset explains
the 2.6-fold high spleen and 8-fold high tumour microdose concentrations, while
liver reproduces to within 4% and steady-state occupancy reproduces well.

The most likely origin is the acidic-phospholipid association constant. Table S6
gives `Ka_bc = 38.5`, citing Rodgers et al. 2005, and that is the value the model
uses. But the Bartelink 2022 supplement (Eq 8 and 9) derives `Ka,AP` as a
*drug-specific* quantity from blood-cell partitioning, and evaluating that
derivation with osimertinib's own `Fu` (5.35%), blood-to-plasma ratio (0.77) and
haematocrit (45%) yields about 11 rather than 38.5. Since the acidic-phospholipid
term dominates `Kpu` for muscle -- which alone is two-thirds of the model's `Vss`
-- the difference propagates straight into `Cmax` and half-life. Substituting 11.0
moves `Cmax` to 268 nM and the half-life to 29 h, overshooting in the other
direction, so neither value reproduces the paper exactly and **the tabulated
value 38.5 is retained**. No parameter was tuned to close any of these gaps.

Note also that the paper states the model "was verified using this empirical data
(4 patients) by adjusting parameters to align simulated curves with observed
profiles" (section 2.3.3), without reporting which parameters were adjusted or to
what values. Exact reproduction from the tabulated inputs alone is therefore not
expected. Every quantity above sits inside the 3-fold window the paper adopts as
its own acceptance criterion, except the microdose tumour concentration.
