---
title: "Generic monoclonal antibody population PK (van den Berg 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generic monoclonal antibody population PK (van den Berg 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Article: [mAbs 2025;17(1):2512217](https://doi.org/10.1080/19420862.2025.2512217)
* Supplement: Supplementary Sections M1-M5, Figures S1-S7 and Table S1
  (`KMAB_A_2512217_SM8525.docx`), plus the full 160-model dataset
  (`KMAB_A_2512217_SM8524.xlsx`), both retrievable from the Europe PMC
  `supplementaryFiles` endpoint for PMC12128664.

This paper is a systematic meta-analysis of **160 published two-compartment
population PK models covering 69 marketed canonical IgG monoclonal antibodies**.
Rather than cataloguing those models only, the authors build an original
deliverable from them: a *generic* mAb model whose every parameter is the
across-model median, which they simulate in NONMEM 7.5.0 to show that "one
model" produces plausible IgG kinetics. That generic model, and the two
weight-covariate configurations the authors simulate alongside it, are what this
vignette validates.

Three model files come out of the paper:

```{r models}
mods <- list(
  generic  = rxode2::rxode(readModelDb("vandenBerg_2025_genericMab_mbma")),
  combI    = rxode2::rxode(readModelDb("vandenBerg_2025_genericMab_allometricClVc_mbma")),
  combII   = rxode2::rxode(readModelDb("vandenBerg_2025_genericMab_allometricAll_mbma"))
)
tibble::tibble(
  Model = names(mods),
  `nlmixr2lib name` = c(
    "vandenBerg_2025_genericMab_mbma",
    "vandenBerg_2025_genericMab_allometricClVc_mbma",
    "vandenBerg_2025_genericMab_allometricAll_mbma"
  ),
  Role = c(
    "Generic mAb model: depot + 2 compartments, IIV on CL/Vc/ka, no covariates (Figure 5c,d; Suppl. M5)",
    "Covariate combination I: body weight on CL and Vc only (Figure 4h; Suppl. Figure S7)",
    "Covariate combination II: body weight on CL, Vc, Vp and Q (Figure 4h; Suppl. Figure S7)"
  )
) |>
  knitr::kable(caption = "Models contributed by van den Berg 2025.")
```

* Citation: `r mods$generic$reference`

# Population

The "population" behind this model is not a trial cohort but a body of
literature. IMGT/mAb-DB was queried for marketed canonical IgG monoclonal
antibodies, PubMed was searched per mAb, and 303 candidate population PK models
were filtered down to **160 models of 69 mAbs** by excluding one-compartment
models, paediatric-only models, subcutaneous models that fixed bioavailability
to 100%, models without a linear clearance component, models fixing more than
two of CL/Vc/Vp/Q, PBPK and non-population models, non-i.v./s.c. routes,
FcRn-engineered mAbs, and models with insufficient reporting (Figure 1a,
Methods "PK models acquisition").

Those 160 models were fitted to a summed **143,094 patients** across the 157
models that reported a patient count (Supplementary Table S1 column `N`);
patients are double-counted wherever several models were fitted to overlapping
datasets, so that figure is an upper bound rather than a cohort size. The median
(IQR) model used 5250 (1788; 10223) samples, i.e. 8.6 (5.7; 13.3) samples per
patient (Results "Sample size matters").

By model (Table 1): IgG1 71.9%, IgG2 8.8%, IgG4 18.1%, IgG2/4 1.2%; chimeric
20.0%, humanised 35.0%, fully human 45.0%; i.v. only 67.5%, s.c. only 0.6%,
i.v./s.c. 31.9%. Nonlinear elimination was present in 47.5% of models and
time-dependent clearance in 19.4% -- neither is carried into the generic model,
which is deliberately a linear-clearance description of IgG homeostasis.

The same information is available programmatically:

```{r population}
str(mods$generic$population, max.level = 1)
```

# Source trace

Every `ini()` entry carries an in-file comment pointing at its source location.
The table below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL) | 0.22 L/day | Figure 5c caption; Suppl. M5 `$THETA 0.22 FIX`; Results median (IQR) 0.22 (0.17; 0.29) |
| `lvc` (Vc) | 3.42 L | Figure 5c caption; Suppl. M5 `$THETA 3.42 FIX`; Results median (IQR) 3.42 (2.96; 3.99) |
| `lvp` (Vp) | 2.68 L | Figure 5c caption; Suppl. M5 `$THETA 2.68 FIX`; Results median (IQR) 2.68 (2.09; 3.54) |
| `lq` (Q) | 0.54 L/day | Figure 5c caption; Suppl. M5 `$THETA 0.54 FIX`; Results median (IQR) 0.54 (0.36; 0.84) |
| `lka` (ka) | 0.25 /day | Figure 5c caption; Suppl. M5 `$THETA 0.25 FIX ; T_ka (/d)`; Results median (IQR) 0.25 (0.18; 0.29) |
| `lfdepot` (F) | 0.69 | Figure 5c caption; Suppl. M5 `$THETA 0.69 FIX` and `F1 = THETA(6)`; Results median (IQR) 69 (61; 77)% |
| `etalcl` | var 0.097 (31.9 CV%) | Figure 5c caption; Suppl. M5 `$OMEGA 0.097 FIX ; IIV_CL`; Figure 3a |
| `etalvc` | var 0.056 (24 CV%) | Figure 5c caption; Suppl. M5 `$OMEGA 0.056 FIX ; IIV_V1`; Figure 3a |
| `etalka` | var 0.26 (54.8 CV%) | Figure 5c caption; Suppl. M5 `$OMEGA 0.26 FIX ; IIV_ka` |
| `propSd` (generic) | 1e-5 | Suppl. M5 `$THETA 0.00001 FIX ; T_Prop_RE (sd)` with `$SIGMA 1 FIX`; Methods: "No RUV or covariates were introduced" |
| `e_wt_cl`, `e_wt_vc` (combination I) | 0.69, 0.67 | Suppl. Figure S7 caption: "I) CL_WT = 0.69, Vc_WT = 0.67" |
| `e_wt_cl`, `e_wt_vc`, `e_wt_vp`, `e_wt_q` (combination II) | 0.75, 0.85, 0.85, 0.77 | Suppl. Figure S7 caption: "II) CL_WT = 0.75, Vc_WT = 0.85, Vp_WT = 0.85, Q_WT = 0.77" |
| Reference weight 70 kg | 70 kg | Suppl. Figure S7 caption worked example `CL40kg = 0.22 x (40/70)^0.75`; Results median (IQR) reference weight 70 (69; 75) kg |
| `d/dt(depot)`, `d/dt(central)`, `d/dt(peripheral1)` | n/a | Suppl. M5 `$DES` (`DEPOT` / `CENTRAL` / `PERI`), with `K20 = CL/VC`, `K23 = Q/VC`, `K32 = Q/VP` |
| `f(depot) <- exp(lfdepot)` | n/a | Suppl. M5 `$PK`, `F1 = THETA(6)` |
| `Cc <- central / vc` | n/a | Suppl. M5 `$PK`, `S2 = VC` |
| Terminal half-life formula | n/a | Suppl. Section M4 (Toutain and Bousquet-Melou 2004) |

## The published medians reproduce from the deposited dataset

The paper deposits the full 160-model dataset (Supplementary Table S1). The
structural medians quoted in the Results and in the Figure 5 caption reproduce
from it exactly -- CL 0.2200, Vc 3.4230, Vp 2.6800, Q 0.5390 over all 160
models, and ka 0.2485, F 0.6945 over the 52 models built on both i.v. and s.c.
data -- along with every interquartile range. That check was run once at
extraction time against `KMAB_A_2512217_SM8524.xlsx`; it is recorded here rather
than re-run in the vignette because the spreadsheet is not redistributed with
this package.

# Part 1 -- the generic model

## Virtual cohort

The paper simulated 500 virtual individuals given a single 100 mg dose by either
route. This vignette uses 200 per arm, which is the cohort cap for validation
vignettes and is ample to characterise the median and the 5th/95th percentiles.

```{r cohort}
set.seed(20250521)

n_per_arm <- 200L
obs_times <- sort(unique(c(
  seq(0, 14, by = 0.5), seq(15, 60, by = 1), seq(62, 168, by = 2)
)))

make_arm <- function(n, arm, dose_cmt, id_offset) {
  subj <- tibble::tibble(id = id_offset + seq_len(n), treatment = arm)
  doses <- subj |>
    mutate(time = 0, amt = 100, evid = 1L, cmt = dose_cmt)
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_arm(n_per_arm, "100 mg IV", "central", 0L),
  make_arm(n_per_arm, "100 mg SC", "depot",   n_per_arm)
)

stopifnot(!anyDuplicated(events[events$evid == 0L, c("id", "time")]))
```

Observation rows point at the ODE state `central`; `Cc` is an algebraic
observable and is returned as a column regardless.

## Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mods$generic, events = events, keep = "treatment", useLinCmt = FALSE
) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim$id) == 2L * n_per_arm)
```

## Replicating Figure 5c and 5d

Figure 5c and 5d of van den Berg 2025 overlay the generic model's median and
5th/95th percentiles on the typical-value profiles of the individual published
i.v. (5c) and s.c. (5d) models. The published models are not redistributable, so
only the generic-model envelope is reproduced here.

```{r figure-5cd}
# Replicates Figure 5c (IV) and 5d (SC) of van den Berg 2025: median and
# 5th/95th percentile concentration-time profile of the generic model after a
# single 100 mg dose.
# rxSolve returns observation records only -- there is no evid column in its output.
sim |>
  group_by(treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time (days)", y = "Serum concentration (mg/L)",
    title = "Generic mAb model, single 100 mg dose",
    caption = "Replicates Figures 5c and 5d of van den Berg 2025 (generic-model envelope only)."
  ) +
  theme_bw()
```

## Terminal half-life -- the paper's headline number

The paper reports a terminal half-life of **21.3 days** for the generic model,
with 5th and 95th percentiles of **13.4** and **35.5** days (Results "Generic
model"). Half-life is computed here with the paper's own formula from
Supplementary Section M4,

`t1/2 = ln(2) / beta` where
`2*beta = (k12 + k21 + k10) - sqrt((k12 + k21 + k10)^2 - 4*k21*k10)`,

applied to each simulated individual's own CL, Vc, Vp and Q.

```{r terminal-halflife}
terminal_half_life <- function(cl, vc, vp, q) {
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  s <- k12 + k21 + k10
  log(2) / ((s - sqrt(s^2 - 4 * k21 * k10)) / 2)
}

thalf <- sim |>
  filter(treatment == "100 mg IV") |>
  distinct(id, cl, vc, vp, q) |>
  mutate(t_half = terminal_half_life(cl, vc, vp, q))

thalf_summary <- tibble::tibble(
  Quantity = c("Median", "5th percentile", "95th percentile"),
  Simulated = c(
    median(thalf$t_half),
    quantile(thalf$t_half, 0.05),
    quantile(thalf$t_half, 0.95)
  ),
  Published = c(21.3, 13.4, 35.5)
) |>
  mutate(`% diff` = (Simulated - Published) / Published * 100)

knitr::kable(
  thalf_summary,
  digits = 2,
  caption = "Terminal half-life (days) of the generic model, 200 virtual individuals, versus the 500-individual values reported in van den Berg 2025 Results 'Generic model'."
)

# Agreement within 10% is expected: the paper drew a different (larger) random
# sample, so the residue is Monte-Carlo noise, not a structural discrepancy.
stopifnot(all(abs(thalf_summary$`% diff`) < 10))
```

The typical-value (no-IIV) half-life is a pure function of the four fixed
disposition parameters and should land next to the simulated median:

```{r typical-halflife}
typ <- as.data.frame(rxode2::rxSolve(
  rxode2::zeroRe(mods$generic),
  events = filter(events, treatment == "100 mg IV", id == 1),
  useLinCmt = FALSE
))
cat(sprintf(
  "Typical-value terminal half-life: %.2f days\n",
  terminal_half_life(typ$cl[1], typ$vc[1], typ$vp[1], typ$q[1])
))
```

## Structural identities

Two identities must hold exactly for a linear two-compartment model given an
i.v. bolus, and both are checked per subject rather than on a median (a median
carries sampling noise; the per-subject identity does not):

* `Cmax = Dose / Vc`
* `AUC0-inf = Dose / CL`

and, comparing arms of the same individual model, `F = AUC_SC / AUC_IV = 0.69`.

```{r identities}
iv <- sim |> filter(treatment == "100 mg IV")

cmax_check <- iv |>
  group_by(id) |>
  summarise(cmax = max(Cc), predicted = 100 / first(vc), .groups = "drop") |>
  mutate(rel = cmax / predicted - 1)

stopifnot(max(abs(cmax_check$rel)) < 1e-6)
cat(sprintf(
  "Cmax = Dose/Vc holds for all %d IV subjects (max relative deviation %.2e)\n",
  nrow(cmax_check), max(abs(cmax_check$rel))
))
```

## PKNCA validation

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)

# Guarantee a time-zero record per subject; for the SC arm the pre-dose
# concentration is 0, and the IV arm already carries its t = 0 bolus row.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | treatment + id,
  concu = "mg/L", timeu = "day"
)

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id, doseu = "mg"
)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_wide <- as.data.frame(nca_res) |>
  select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  group_by(treatment) |>
  summarise(
    across(c(cmax, tmax, aucinf.obs, half.life), median),
    .groups = "drop"
  ) |>
  rename(
    "Treatment" = treatment, "Cmax (mg/L)" = cmax, "Tmax (day)" = tmax,
    "AUC0-inf (mg*day/L)" = aucinf.obs, "t1/2 (day)" = half.life
  ) |>
  knitr::kable(
    digits = 2,
    caption = "Median PKNCA parameters by arm, generic model, single 100 mg dose."
  )
```

`AUC0-inf = Dose / CL` per subject, and the ratio of the arms recovers the
bioavailability:

```{r auc-identity}
auc_iv <- nca_wide |>
  filter(treatment == "100 mg IV") |>
  left_join(distinct(iv, id, cl), by = "id") |>
  mutate(rel = aucinf.obs / (100 / cl) - 1)

cat(sprintf(
  "AUC0-inf = Dose/CL for all %d IV subjects (max relative deviation %.3f%%)\n",
  nrow(auc_iv), 100 * max(abs(auc_iv$rel))
))
stopifnot(max(abs(auc_iv$rel)) < 0.01)

f_recovered <-
  median(nca_wide$aucinf.obs[nca_wide$treatment == "100 mg SC"]) /
  median(nca_wide$aucinf.obs[nca_wide$treatment == "100 mg IV"])
cat(sprintf(
  "Recovered bioavailability F = %.3f (model value 0.69)\n", f_recovered
))
stopifnot(abs(f_recovered - 0.69) < 0.07)
```

The arms share the same `etalcl` draws only in distribution, not subject by
subject, so the recovered `F` carries a little Monte-Carlo noise; the tolerance
above reflects that rather than a structural doubt.

## Comparison against the published values

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,  ~half.life,
  "100 mg IV", 21.3
)

nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  units = c(half.life = "day"),
  tolerance_pct = 20
) |>
  knitr::kable(
    caption = "Simulated versus published NCA. * marks a >20% difference. The only NCA-style value the paper publishes for the generic model is the terminal half-life (Results 'Generic model')."
  )
```

# Part 2 -- the two weight-covariate configurations

Body weight was a covariate in 84.4% of the collected models, but *where* the
authors of those models put it varies. The two most common configurations are
weight on CL and Vc only (combination I, 20 models) and weight on all four
disposition parameters (combination II, 15 models). van den Berg 2025 simulate
both, parameterised with the median exponents inside each subset, and observe
that they push terminal half-life in **opposite directions** as weight falls:
24.4 days for combination I versus 19 days for combination II at 40 kg, "28%
longer with combination I" (Results "Covariate model"; Figure 4h; Figure S7).

These simulations are deterministic -- the authors ran them "without
inter-individual variability or residual variability".

```{r allometric-sim}
wt_grid <- seq(40, 100, by = 5)

allo_events <- tibble::tibble(id = seq_along(wt_grid), WT = wt_grid) |>
  tidyr::crossing(time = c(0, sort(unique(c(
    seq(0.25, 21, by = 0.25), seq(22, 120, by = 1), seq(124, 400, by = 4)
  ))))) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central") |>
  bind_rows(
    tibble::tibble(
      id = seq_along(wt_grid), WT = wt_grid, time = 0,
      amt = 100, evid = 1L, cmt = "central"
    )
  ) |>
  arrange(id, time, desc(evid))

solve_allo <- function(mod, label) {
  rxode2::rxSolve(mod, events = allo_events, keep = "WT", useLinCmt = FALSE) |>
    as.data.frame() |>
    mutate(combination = label)
}

allo <- bind_rows(
  solve_allo(mods$combI,  "I: weight on CL, Vc"),
  solve_allo(mods$combII, "II: weight on CL, Vc, Vp, Q")
)
```

## Replicating Figure 4h

```{r figure-4h}
# Replicates Figure 4h of van den Berg 2025: terminal half-life versus body
# weight under the two weight-covariate configurations.
allo_thalf <- allo |>
  distinct(combination, WT, cl, vc, vp, q) |>
  mutate(t_half = terminal_half_life(cl, vc, vp, q))

ggplot(allo_thalf, aes(WT, t_half, colour = combination, linetype = combination)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.4) +
  labs(
    x = "Body weight (kg)", y = "Terminal half-life (days)",
    colour = NULL, linetype = NULL,
    title = "Terminal half-life versus body weight",
    caption = "Replicates Figure 4h of van den Berg 2025."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r figure-4h-check}
at40 <- allo_thalf |> filter(WT == 40)

tibble::tibble(
  Combination = at40$combination,
  `Simulated t1/2 at 40 kg (day)` = at40$t_half,
  `Published t1/2 at 40 kg (day)` = c(24.4, 19.0)
) |>
  mutate(`% diff` = (`Simulated t1/2 at 40 kg (day)` /
                       `Published t1/2 at 40 kg (day)` - 1) * 100) |>
  knitr::kable(
    digits = 2,
    caption = "Terminal half-life at 40 kg versus van den Berg 2025 Results 'Covariate model'."
  )

pct_longer <- 100 * (at40$t_half[1] / at40$t_half[2] - 1)
cat(sprintf(
  "Combination I is %.1f%% longer than combination II at 40 kg (paper: 28%%)\n",
  pct_longer
))
stopifnot(abs(pct_longer - 28) < 3)

# The crossover direction is the paper's point: at 40 kg combination I gives the
# LONGER half-life, at 100 kg the SHORTER one.
at100 <- allo_thalf |> filter(WT == 100)
stopifnot(at40$t_half[1] > at40$t_half[2], at100$t_half[1] < at100$t_half[2])
```

## Exposure, by PKNCA

The paper adds that "after a single 100 mg dose, total drug exposure in a 40 kg
individual was only 10% higher with combination 2 compared to combination I",
and that the difference is eliminated after multiple doses. The integration
window behind that 10% is not stated; it is reproduced below over a three-cycle
(63-day) window, which is the horizon Figure S7C-E displays.

```{r allometric-pknca}
allo_nca_input <- allo |>
  filter(!is.na(Cc)) |>
  mutate(arm = paste(combination, WT, sep = " @ ")) |>
  select(arm, id = WT, time, Cc)

allo_conc <- PKNCA::PKNCAconc(
  allo_nca_input, Cc ~ time | arm + id, concu = "mg/L", timeu = "day"
)
allo_dose <- PKNCA::PKNCAdose(
  distinct(allo_nca_input, arm, id) |> mutate(time = 0, amt = 100),
  amt ~ time | arm + id, doseu = "mg"
)

allo_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  allo_conc, allo_dose,
  intervals = data.frame(
    start = 0, end = c(63, Inf),
    aucall = c(TRUE, FALSE), aucinf.obs = c(FALSE, TRUE),
    half.life = c(FALSE, TRUE)
  )
))

allo_summary <- as.data.frame(allo_res) |>
  filter(PPTESTCD %in% c("aucall", "aucinf.obs", "half.life")) |>
  tidyr::separate_wider_delim(arm, " @ ", names = c("combination", "wt_lab")) |>
  select(combination, WT = id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

exposure_ratio <- allo_summary |>
  filter(WT == 40) |>
  summarise(
    `AUC0-63d ratio (II / I)` =
      aucall[startsWith(combination, "II")] / aucall[startsWith(combination, "I:")],
    `AUC0-inf ratio (II / I)` =
      aucinf.obs[startsWith(combination, "II")] / aucinf.obs[startsWith(combination, "I:")]
  )

knitr::kable(
  exposure_ratio, digits = 3,
  caption = "Exposure ratio at 40 kg after a single 100 mg IV dose. The paper's ~10% figure is reproduced over the 63-day (three q3w cycle) window; extrapolated to infinity the two configurations differ by only ~3%, which is the paper's 'eliminated after multiple doses' observation, since AUC over a dosing interval at steady state equals Dose/CL and the two CL exponents differ by only 0.06."
)

stopifnot(
  exposure_ratio$`AUC0-63d ratio (II / I)` > 1.05,
  exposure_ratio$`AUC0-63d ratio (II / I)` < 1.20
)
```

```{r allometric-nca-table}
allo_summary |>
  filter(WT %in% c(40, 70, 100)) |>
  arrange(combination, WT) |>
  rename(
    "Combination" = combination, "Weight (kg)" = WT,
    "AUC0-63d (mg*day/L)" = aucall,
    "AUC0-inf (mg*day/L)" = aucinf.obs,
    "t1/2 (day)" = half.life
  ) |>
  knitr::kable(
    digits = 2,
    caption = "PKNCA parameters for the two weight-covariate configurations. At 70 kg the two are identical by construction, since both reduce to the across-model medians."
  )
```

At 70 kg both configurations must collapse onto the un-scaled generic
disposition parameters:

```{r seventy-kg-identity}
at70 <- allo_summary |> filter(WT == 70)
stopifnot(abs(diff(at70$aucinf.obs)) / mean(at70$aucinf.obs) < 1e-6)
cat("Both configurations agree exactly at the 70 kg reference weight.\n")
```

# Between-model versus between-subject variability

One of the paper's central observations is that the spread of parameter
estimates *across published models* is comparable to the between-subject
variability *within* a model -- which is what motivates using medians as a
priori estimates in the first place. That between-model spread is a property of
the literature, not of the generic model, so it is **not** encoded in the model
file; it is tabulated here for reference (Results "Structural PK estimates" and
"Stochastic model", Figures 1c,d and 3).

```{r variability-table}
tibble::tribble(
  ~Parameter, ~`Between-model CV% (Figure 1)`, ~`Between-subject CV%, base models (Figure 3b)`,
  "CL", 54.6, 43.0,
  "Vc", 25.0, 28.8,
  "Vp", 73.9, NA_real_,
  "Q",  108.0, NA_real_,
  "ka", 34.0, NA_real_,
  "F",  61.0, NA_real_
) |>
  knitr::kable(
    caption = "Variability in population estimates across the 160 models, next to the median between-subject variability reported in the base models. Only CL and Vc have both."
  )
```

# Assumptions and deviations

* **This is a meta-analysis, and the model it produces is a meta-analytic
  construct.** Every value is a median over published model estimates, not an
  estimate from data. All parameters are consequently `fixed()`, matching the
  paper's own NONMEM control stream, in which every `$THETA` and `$OMEGA` line
  carries `FIX`. Do not read the `ini()` values as fitted point estimates.
* **The random effects are between-SUBJECT, not between-study.** This is a
  meta-analysis *of models* rather than a regression meta-analysis of trial
  summaries, and the authors deliberately used the median published
  between-subject omegas as individual-level IIV in a 500-individual simulation.
  They therefore carry the standard `etal<param>` names rather than the
  `eta_study_*` MBMA convention. The separate between-*model* spread (table
  above) is not encoded.
* **Cohort size.** The paper used 500 virtual individuals; this vignette uses
  200 per arm, the validation-vignette cap. The residual disagreement in the
  half-life percentiles (a few percent) is Monte-Carlo noise from the smaller,
  differently-seeded sample.
* **Intravenous dosing is a bolus here.** The paper notes that in the collected
  models "i.v. doses are administered directly into the central compartment over
  the course of an hour", but its own generic-model control stream (Suppl. M5)
  carries no infusion rate, so a bolus is used. Over a 21-day half-life the
  choice is immaterial except at the first few observations, and it makes
  `Cmax = Dose/Vc` an exact identity to test against.
* **`ka` units.** The abstract prints the absorption rate constant as
  "0.25 L/d", which is dimensionally impossible. The Figure 5c caption and the
  Supplementary M5 `$THETA` comment both give `/d`, and that is what the model
  uses.
* **Residual error is a placeholder, not an estimate.** The paper introduced no
  RUV. The generic model carries the control stream's `propSd = 1e-5`
  verbatim; the two allometric configurations, which the paper ran "without
  inter-individual variability or residual variability", carry a structural
  `propSd = 0`. Re-specify a residual-error model before fitting data with any
  of the three.
* **Specimen is unstated.** The 160 source models pool serum and plasma assays
  (the paper reports only the detection method, ELISA or electrochemiluminescence).
  `compartmentData` records `serum` with `verified = FALSE` because the source
  does not state a matrix for the generic model.
* **The third common covariate configuration is not extracted.** The paper
  identifies "weight and sex on CL and Vc" (19 models) as the third most common
  combination but neither simulates it nor reports the median coefficients for
  that subset, so it cannot be parameterised from on-disk sources.
* **Screened but unused covariates.** Median effects for body weight, albumin,
  sex and ADA status across the 160 models are recorded in the generic model's
  `covariatesDataExcluded` metadata (allometric exponents 0.62 on CL and 0.56 on
  Vc when estimated; albumin power -0.90 on CL at a 40 g/L reference; 19% higher
  CL in males; 1.39-fold CL in ADA-positive patients). None of them is part of
  the generic model, and the paper does not report a joint covariate model that
  combines them, so none is encoded in `model()`.
* **Nonlinear and time-dependent elimination are out of scope by construction.**
  47.5% of the collected models included nonlinear kinetics and 19.4%
  time-dependent clearance, and the paper shows the linear CL component is 26%
  lower in mixed models (0.18 versus 0.25 L/day). The generic model describes
  linear IgG homeostasis only; a target-mediated component must be added
  explicitly for any mAb where it matters.
* **The 10% single-dose exposure claim depends on an unstated integration
  window.** It reproduces over a 63-day window (Figure S7's display horizon) but
  not to infinity, where the difference is ~3%. Both numbers are shown above
  rather than picking the flattering one.
