---
title: "Treosulfan (van der Stoep 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Treosulfan (van der Stoep 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("vanderStoep_2019_treosulfan"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.1111/bcp.13995>
* Supporting information (open access, incl. the final NONMEM control stream as
  Supporting Information 4):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC6710524/supplementaryFiles>

Treosulfan is an alkylating prodrug used in myeloablative conditioning regimens
prior to haematopoietic stem cell transplantation (HSCT). van der Stoep 2019
developed a two-compartment IV-infusion population PK model on a
multi-institutional paediatric dataset that is substantially larger and spans a
far wider age range than the three previously published paediatric treosulfan
models, and is the first of them to carry an age-maturation component.

# Population

The analysis pooled 91 records (84 unique patients; 7 contributed a second
record from a second transplantation more than several months later) from a
prospective observational study run at Leiden University Medical Center
(n = 63) and IRCCS Bambino Gesu Children's Hospital, Rome (n = 28) between June
2011 and March 2017. Median age was 4.3 years (range 0.1-18.2) and 33 of 91
(36%) were infants aged 2 years or less -- the sub-population the paper set out
to characterise. Median body weight was 15.6 kg (range 3.8-75.0) and median body
surface area 0.7 m^2 (range 0.3-1.9); 63.7% were male. Underlying diagnoses were
haemoglobinopathy (38.5%), primary immune deficiency (28.6%), haematological
malignancy (18.7%), bone marrow failure (12.1%) and other (2.2%)
(van der Stoep 2019 Table 1).

Patients older than 1 year received 14 g/m^2 per day as a 3 h infusion on three
consecutive days (total 42 g/m^2); patients younger than 1 year received 10 or
12 g/m^2 per day. 410 serum concentrations were analysed, 4 of which (1%) were
below the 6.8 mg/L HPLC-UV lower limit of quantification and were retained at
their measured values. Observed treosulfan AUC(0-inf) was 1658 mg*h/L (range
643-3371).

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(ui$population, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/vanderStoep_2019_treosulfan.R` carries an in-file
comment naming its origin. They are collected here for review. "Data S4" is the
final NONMEM control stream published as Supporting Information 4; its `$THETA`
block reproduces the Table 2 "Final model" column exactly and its `$OMEGA` /
`$SIGMA` diagonals reproduce the Table 2 variability percentages as
`sqrt(variance)`, which is what identifies it as carrying final rather than
initial estimates.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at 70 kg)    | 18.8 L/h  | Table 2 "Final model"; Data S4 `$THETA (0, 18.8) TH_CL` |
| `lvc` (V1 at 70 kg)    | 20.2 L    | Table 2 "Final model"; Data S4 `$THETA (0, 20.2) TH_V` |
| `lq` (Q at 70 kg)      | 21.3 L/h  | Table 2 "Final model"; Data S4 `$THETA (0, 21.3) TH_Q` |
| `lvp` (V2 at 70 kg)    | 16.8 L    | Table 2 "Final model"; Data S4 `$THETA (0, 16.8) TH_V2` |
| `e_wt_cl`, `e_wt_q`    | 0.75 (fixed) | Methods 2.5 under Eq. 4; Data S4 `FSIZE=(WT/70)**0.75` |
| `e_wt_vc`, `e_wt_vp`   | 1 (fixed)    | Methods 2.5 under Eq. 4; Data S4 `(WT/70)` |
| `pma_tm50`             | 38 weeks  | Table 2 "Final model"; Data S4 `$THETA (0, 38) TM50` |
| `pma_hill`             | 1.22      | Data S4 `$THETA (0, 1.22) Hill` (Table 2 prints 1.2) |
| IIV block (4x4)        | see below | Data S4 `$OMEGA BLOCK(4)`; diagonals = Table 2 IIV 31.8 / 45.9 / 17.3 / 41.4 %CV |
| IOV on CL (3 occasions)| 0.0194    | Data S4 `$OMEGA BLOCK(1)` + 2x `SAME`; Table 2 IOV Cl 13.9% |
| `propSd`               | 0.1233    | Data S4 `$SIGMA 0.0152`; Table 2 sigma 12.3% |
| `F_size = (BW/70)^alpha` | n/a     | Equation 4 |
| `F_mat = 1/(1+(PMA/TM50)^-Hill)` | n/a | Equation 5; Data S4 `FMAT=1/(1+(PMA/TM50)**(-HILL))` |
| `Cl_tot = Cl_pop * F_size * F_mat`, same form for Q | n/a | Equation 6 + "A similar model was used for Q"; Results 3.3 |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Data S4 `$DES` (`K=CL/V`, `K12=Q/V`, `K21=Q/V2`) |
| `Cc ~ prop(propSd)`    | n/a       | Equation 3 (`Y = YPRED * (1 + eps)`); Data S4 `$ERROR Y = F*(1+ERR(1))` |
| `PMA = 40 weeks + postnatal age` | n/a | Methods 2.5 |
| `AUC(0-inf) = DOSE * F1 / CL`, `F = 1` | n/a | Methods 2.7 and 2.9 |

The Table 2 variability percentages are `sqrt(omega^2)`, not the log-normal
`sqrt(exp(omega^2) - 1)`:

```{r omega-check}
omega_diag <- c(CL = 0.101, V1 = 0.211, V2 = 0.0299, Q = 0.171)
published_cv <- c(CL = 31.8, V1 = 45.9, V2 = 17.3, Q = 41.4)
data.frame(
  parameter = names(omega_diag),
  `omega^2` = unname(omega_diag),
  `sqrt(omega^2) %` = round(100 * sqrt(omega_diag), 1),
  `published Table 2 %` = unname(published_cv),
  check.names = FALSE
)
stopifnot(all(abs(100 * sqrt(omega_diag) - published_cv) < 0.1))
```

The 4x4 block is positive definite, but only just -- a direct consequence of the
70 kg standardisation, which the Discussion notes inflates the V2 and Q IIV
relative standard errors to 201% and 153% (against 53% and 46% when
standardising to the cohort median 15.6 kg). The published model kept 70 kg for
comparability with other publications, so that is what the model file encodes.

```{r omega-conditioning}
omega <- matrix(c(
   0.101,   0.131,  0.0349, -0.0307,
   0.131,   0.211,  0.026,  -0.0354,
   0.0349,  0.026,  0.0299, -0.05,
  -0.0307, -0.0354, -0.05,   0.171
), nrow = 4, byrow = TRUE,
dimnames = list(c("CL", "V1", "V2", "Q"), c("CL", "V1", "V2", "Q")))

round(cov2cor(omega), 3)
ev_omega <- eigen(omega, only.values = TRUE)$values
stopifnot(min(ev_omega) > 0)   # positive definite: rxode2's Cholesky sampler works
signif(min(ev_omega), 3)
```

# Maturation function (replicates Figure 2)

Equation 5 is a sigmoid Emax function of postmenstrual age driving both CL and
Q. Figure 2 of the paper plots it as a percentage of the mature (adult) value
against PMA in weeks, with dashed landmarks at birth, 1 year and 4 years.

```{r figure-2}
tm50 <- 38
hill <- 1.22
fmat <- function(pma_weeks) 1 / (1 + (pma_weeks / tm50)^(-hill))
# The paper defines PMA as 40 weeks of assumed gestation plus postnatal age.
pma_of_age <- function(age_years) 40 + age_years * 52.1775

mat_curve <- tibble(pma = seq(31, 390, by = 1)) |>
  mutate(pct = 100 * fmat(pma))

landmarks <- tibble(
  label = c("Birth", "1 year", "4 years"),
  pma = pma_of_age(c(0, 1, 4))
) |>
  mutate(pct = 100 * fmat(pma))

ggplot(mat_curve, aes(pma, pct)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_vline(data = landmarks, aes(xintercept = pma),
             colour = "red", linetype = "dashed") +
  geom_text(data = landmarks, aes(x = pma, y = 41, label = label),
            hjust = -0.08, fontface = "italic", size = 3) +
  scale_y_continuous(limits = c(40, 100)) +
  labs(x = "Postmenstrual age (weeks)", y = "% of matured clearance",
       caption = "Replicates Figure 2 of van der Stoep 2019.")

landmarks |> mutate(pct = round(pct, 1))
```

Four claims the paper makes about this curve, checked against the encoded
parameters. All are deterministic -- no simulation, no random draws -- so the
tolerances are tight.

```{r figure-2-gate}
# 1. "The maturation of treosulfan clearance reaches 50% of adult values at
#    38 weeks PMA" (Results 3.3). True by construction of TM50.
stopifnot(abs(100 * fmat(38) - 50) < 1e-8)

# 2. "that is 2 weeks prior to birth assuming a full-term gestational age of
#    40 weeks" -- so the curve crosses just above 50% at birth.
stopifnot(100 * fmat(pma_of_age(0)) > 50,
          100 * fmat(pma_of_age(0)) < 55)

# 3. "Clearance reaches 90% of adult values at age approximately 4 years"
#    (Results 3.3, Discussion, and the "What this study adds" box).
stopifnot(abs(100 * fmat(pma_of_age(4)) - 90) < 3)

# 4. The Figure 2 panel runs to ~390 weeks PMA and plateaus near 94-95%,
#    not at 100% -- a property of the Hill coefficient, so this discriminates
#    Hill = 1.22 from a much steeper or shallower value.
stopifnot(abs(100 * fmat(390) - 94.5) < 1.5)
```

# Virtual cohort: the paper's own dosing simulation (Table 3)

Methods 2.7 describes the paper's simulation: individual treosulfan doses were
computed to target an AUC(0-inf) of 1650 mg*h/L -- the daily median exposure in
patients receiving the most common 14 g/m^2 dose -- for virtual patients whose
weights are the 5th, 50th and 95th CDC weight-for-age percentiles for boys.
Table 3 publishes the resulting 66 (age, weight, dose) triplets. Because
AUC(0-inf) = DOSE / CL with F = 1, reproducing Table 3 is an exact,
fully deterministic test of the typical-value clearance model: the population
CL, the 0.75 allometric exponent, the maturation function, and the
PMA = 40 weeks + postnatal age convention all have to be right simultaneously.

```{r table3}
# van der Stoep 2019 Table 3, transcribed verbatim (age category, CDC
# weight-for-age percentile weight in kg, recommended daily dose in mg).
table3 <- tibble::tribble(
  ~age_y, ~wt,  ~dose_pub,
  0,      2.6,  1350,   0,     3.3,  1600,   0,     4.2,  1950,
  0.25,   5.2,  2650,   0.25,  6.4,  3100,   0.25,  7.7,  3550,
  0.5,    6.6,  3500,   0.5,   7.9,  4000,   0.5,   9.5,  4600,
  0.75,   7.4,  4100,   0.75,  8.9,  4700,   0.75, 10.6,  5300,
  1,      8.1,  4600,   1,     9.6,  5200,   1,    11.5,  6000,
  2,     10.1,  6100,   2,    12.2,  7000,   2,    14.7,  8000,
  3,     12.0,  7300,   3,    14.3,  8300,   3,    17.3,  9600,
  4,     13.6,  8250,   4,    16.3,  9450,   4,    20.3, 11100,
  5,     15.2,  9100,   5,    18.5, 10500,   5,    23.5, 12700,
  6,     16.9, 10000,   6,    20.8, 11700,   6,    27.0, 14250,
  7,     18.7, 11000,   7,    23.2, 12800,   7,    30.9, 16000,
  8,     20.7, 11900,   8,    25.8, 14000,   8,    35.3, 17700,
  9,     22.7, 12800,   9,    28.7, 15250,   9,    40.4, 19700,
  10,    24.9, 13800,  10,    32.1, 16700,  10,    46.2, 21900,
  11,    27.5, 14900,  11,    36.1, 18300,  11,    52.6, 24200,
  12,    30.6, 16200,  12,    40.7, 20000,  12,    59.3, 26600,
  13,    34.2, 17600,  13,    45.8, 22000,  13,    66.1, 28900,
  14,    38.5, 19300,  14,    51.2, 24000,  14,    72.7, 31100,
  15,    43.0, 21000,  15,    56.5, 25800,  15,    78.8, 33100,
  16,    47.3, 22600,  16,    61.1, 27400,  16,    84.3, 34900,
  17,    50.8, 23900,  17,    64.7, 28700,  17,    88.8, 36300,
  18,    53.2, 24800,  18,    67.3, 29500,  18,    92.0, 37400
) |>
  mutate(
    id = row_number(),
    PAGE = pma_of_age(age_y),
    WT = wt,
    OCC = 1L,
    arm = "Table 3 target (typical values)"
  )

stopifnot(nrow(table3) == 66L, !anyDuplicated(table3$id))
```

```{r table3-events}
# Build the event table by hand so the covariate columns ride along. Doses are
# 3 h infusions into `central`; observation rows also name `central` (the ODE
# state), never `Cc` (the algebraic observable).
obs_grid <- seq(0, 24, by = 0.05)

make_events <- function(subj, grid) {
  dosing <- subj |>
    transmute(id, time = 0, amt = dose_mg, evid = 1L, dur = 3,
              cmt = "central", WT, PAGE, OCC, arm)
  obs <- subj |>
    select(id, WT, PAGE, OCC, arm) |>
    tidyr::crossing(time = grid) |>
    mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  bind_rows(dosing, obs) |>
    arrange(id, time, desc(evid))
}

ev_t3 <- table3 |>
  mutate(dose_mg = dose_pub) |>
  make_events(obs_grid)

stopifnot(!anyDuplicated(unique(ev_t3[, c("id", "time", "evid")])))
```

```{r table3-simulate}
mod <- readModelDb("vanderStoep_2019_treosulfan")

sim_t3 <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = ev_t3,
  keep = c("WT", "PAGE", "arm")
) |>
  as.data.frame()

stopifnot(all(sim_t3$Cc >= 0, na.rm = TRUE))
```

## Table 3 exposure gate

```{r table3-pknca}
sim_nca_t3 <- sim_t3 |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

# Guarantee a time-zero record per subject (IV infusion: pre-dose Cc = 0).
sim_nca_t3 <- bind_rows(
  sim_nca_t3,
  sim_nca_t3 |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

conc_t3 <- PKNCA::PKNCAconc(sim_nca_t3, Cc ~ time | arm + id)

dose_t3 <- ev_t3 |>
  filter(evid == 1L) |>
  select(id, time, amt, arm)
dose_obj_t3 <- PKNCA::PKNCAdose(dose_t3, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  aucinf.obs = TRUE, half.life = TRUE
)

nca_t3 <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_t3, dose_obj_t3, intervals = intervals)
)
```

```{r table3-gate}
auc_t3 <- as.data.frame(nca_t3) |>
  filter(PPTESTCD == "aucinf.obs") |>
  select(id, aucinf = PPORRES) |>
  inner_join(table3, by = "id") |>
  mutate(
    # Closed form from Methods 2.7 / 2.9: AUC(0-inf) = DOSE * F1 / CL, F = 1.
    cl_closed = 18.8 * (wt / 70)^0.75 * fmat(PAGE),
    auc_closed = dose_pub / cl_closed,
    pct_vs_closed = 100 * (aucinf - auc_closed) / auc_closed,
    pct_vs_target = 100 * (aucinf - 1650) / 1650
  )

summary(auc_t3$pct_vs_closed)
summary(auc_t3$pct_vs_target)
```

Two things are being checked here.

First, the NCA of the solved profile must recover the closed form `DOSE / CL`
exactly. Both sides use the same parameter draw (there is none -- `zeroRe()`
zeroes every random effect), so the only difference is trapezoidal /
extrapolation error on a finely-sampled grid; a tight bound is the correct
assertion, and it is the one that catches a structural transcription error.

```{r table3-gate-closed}
stopifnot(max(abs(auc_t3$pct_vs_closed)) < 0.5)
```

Second, each of the 66 published doses must land on the paper's own 1650 mg*h/L
target. The residual scatter here is the paper's rounding of Table 3 to the
nearest 50 mg, which is largest in absolute terms for the smallest neonatal
doses (1350-1950 mg).

```{r table3-gate-target}
stopifnot(
  # Centre: a mis-transcribed CL, exponent or maturation term moves the whole
  # distribution by tens of percent.
  abs(median(auc_t3$pct_vs_target)) < 1,
  # Envelope: the worst single row is the 3.3 kg neonate, where 50 mg of dose
  # rounding is 3% of the dose. Realised max 1.14%; the bound leaves headroom
  # for the grid and for a re-transcription, and still goes red on any real
  # structural error.
  max(abs(auc_t3$pct_vs_target)) < 3
)
```

```{r table3-gate-table}
auc_t3 |>
  filter(age_y %in% c(0, 1, 4, 10, 18)) |>
  transmute(
    "Age" = ifelse(age_y == 0, "0 mo", paste0(age_y, " y")),
    "Weight (kg)" = wt,
    "Published dose (mg/day)" = dose_pub,
    "Model CL (L/h)" = round(cl_closed, 3),
    "Simulated AUC0-inf (mg*h/L)" = round(aucinf, 0),
    "% vs 1650 target" = round(pct_vs_target, 2)
  ) |>
  knitr::kable(
    caption = "Table 3 dose recommendations re-derived from the packaged model (selected rows; all 66 are gated above)."
  )
```

# Dose-vs-age relationship (replicates Figure 5)

Figure 5 plots the daily dose required to reach the 1650 mg*h/L target against
age, panel A in mg/kg and panel B as the absolute dose, with the 50th weight
percentile as a solid line and the 5th and 95th as dashed. The paper's reading
of it: "The recommended treosulfan dose per kg is lower in early years of life
and reaches a maximum at approximately 4 years accounting for maturation of
clearance and because dose per kg is higher in younger children based on
allometric theory."

```{r figure-5}
fig5 <- auc_t3 |>
  arrange(age_y, wt) |>
  group_by(age_y) |>
  mutate(percentile = c("5th", "50th", "95th")[rank(wt)]) |>
  ungroup() |>
  mutate(dose_per_kg = dose_pub / wt)

p_a <- ggplot(fig5, aes(age_y, dose_per_kg, group = percentile,
                        linetype = percentile)) +
  geom_line() +
  scale_linetype_manual(values = c("5th" = "dashed", "50th" = "solid",
                                   "95th" = "dashed")) +
  labs(x = "Age (years)", y = "Treosulfan dose (mg/kg per day)",
       title = "A", linetype = "Weight percentile")

p_b <- ggplot(fig5, aes(age_y, dose_pub, group = percentile,
                        linetype = percentile)) +
  geom_line() +
  scale_linetype_manual(values = c("5th" = "dashed", "50th" = "solid",
                                   "95th" = "dashed")) +
  labs(x = "Age (years)", y = "Treosulfan dose (mg per day)",
       title = "B", linetype = "Weight percentile")

p_a
p_b
```

```{r figure-5-gate}
# The paper's claim: dose per kg peaks at about 4 years on the 50th-percentile
# curve. Deterministic (Table 3 is a printed table), so this is exact.
peak_age <- fig5 |>
  filter(percentile == "50th") |>
  slice_max(dose_per_kg, n = 1) |>
  pull(age_y)
peak_age
stopifnot(peak_age >= 3, peak_age <= 5)

# Panel B: absolute dose rises monotonically with age on every percentile curve.
stopifnot(
  fig5 |>
    arrange(percentile, age_y) |>
    group_by(percentile) |>
    summarise(mono = all(diff(dose_pub) > 0), .groups = "drop") |>
    pull(mono) |>
    all()
)
```

# Study-protocol cohort with between-subject variability (replicates Figure 1)

Figure 1 shows the observed full concentration-time profiles of the 27 patients
sampled at 1.5, 3.5, 4, 5, 7 and 9 h after the start of a 3 h infusion of
14 g/m^2. Reproducing it, and the published cohort exposure, needs three things
the paper reports only as marginal summaries: an age distribution, a weight
distribution, and a dose in mg (i.e. a body surface area per subject).

The cohort below is built to reproduce the published marginals exactly where
they are stated:

* **Age** is drawn from a piecewise-linear quantile function pinned to the three
  published facts -- minimum 0.1 y, 36% aged 2 y or less, median 4.3 y, maximum
  18.2 y (Table 1).
* **Weight** is the Table 3 50th-percentile CDC weight-for-age at that age,
  multiplied by 0.88 and by log-normal scatter, which brings the cohort weight
  median onto the published 15.6 kg. The sub-unit factor is physiologically
  expected: these are children sick enough to need a transplant, so they sit
  below the healthy weight-for-age median.
* **BSA** is scaled from weight as `BSA = 0.7 * (WT / 15.6)^(2/3)`, anchored on
  the Table 1 medians (15.6 kg, 0.7 m^2) using the standard two-thirds-power
  geometric relationship; BSA itself was computed by the Mosteller formula in
  the paper (Supporting Information 2), which needs a height the paper does not
  publish. See "Assumptions and deviations" for the check that this reproduces
  the published BSA range from the published weight range.
* **Dose follows the study protocol, not a flat 14 g/m^2.** Methods 2.1:
  patients older than 1 year received 14 g/m^2 per day, while patients younger
  than 1 year received 10 or 12 g/m^2 (16 and 2 of 91 patients respectively).
  This matters: infants are exactly the subgroup whose maturation fraction is
  lowest, so dosing them at the older-child rate would inflate their exposure by
  about 40% and bias the whole cohort median upward. Reproducing the published
  AUC requires reproducing the published dose rule.

```{r cohort-14}
# `set.seed()` seeds R's RNG, not rxode2's; rxode2 partitions its streams per
# solver thread, so the cohort below differs between a 2-thread CI runner and a
# 16-thread workstation. Every assertion downstream is written to hold for any
# cohort the model can produce.
set.seed(20190901)

n_cohort <- 150L # <= 200 per arm

# Quantile function pinned to the Table 1 age marginals.
age_quantile <- approxfun(
  x = c(0, 0.36, 0.50, 1.00),
  y = c(0.1, 2.0, 4.3, 18.2)
)

wt50 <- fig5 |> filter(percentile == "50th") |> arrange(age_y)

cohort14 <- tibble(
  id = 1000L + seq_len(n_cohort),
  age_y = age_quantile(runif(n_cohort))
) |>
  mutate(
    WT = pmin(
      pmax(
        approx(wt50$age_y, wt50$wt, xout = age_y, rule = 2)$y *
          0.88 * exp(rnorm(n_cohort, 0, 0.25)),
        3.8
      ),
      75.0
    ),
    PAGE = pma_of_age(age_y),
    OCC = 1L,
    bsa = 0.7 * (WT / 15.6)^(2 / 3),
    # Study protocol (Methods 2.1): 10 g/m^2 under 1 year, 14 g/m^2 at or above.
    dose_g_m2 = ifelse(age_y < 1, 10, 14),
    dose_mg = dose_g_m2 * 1000 * bsa,
    arm = "Study protocol cohort (with IIV)"
  )

summary(cohort14[, c("age_y", "WT", "bsa", "dose_mg")])
```

The reconstructed cohort against the published marginals (Table 1):

```{r cohort-14-marginals}
tibble::tribble(
  ~Quantity, ~Published, ~Simulated,
  "Median age (y)", "4.3", sprintf("%.1f", median(cohort14$age_y)),
  "Age range (y)", "0.1-18.2",
  sprintf("%.1f-%.1f", min(cohort14$age_y), max(cohort14$age_y)),
  "Aged 2 y or less", "36%",
  sprintf("%.0f%%", 100 * mean(cohort14$age_y <= 2)),
  "Median weight (kg)", "15.6", sprintf("%.1f", median(cohort14$WT)),
  "Weight range (kg)", "3.8-75.0",
  sprintf("%.1f-%.1f", min(cohort14$WT), max(cohort14$WT)),
  "Median BSA (m^2)", "0.7", sprintf("%.1f", median(cohort14$bsa)),
  "Receiving 14 g/m^2", "80%",
  sprintf("%.0f%%", 100 * mean(cohort14$dose_g_m2 == 14))
) |>
  knitr::kable(caption = "Reconstructed cohort vs van der Stoep 2019 Table 1.")
```

```{r cohort-14-simulate}
ev_14 <- make_events(cohort14, obs_grid)
stopifnot(!anyDuplicated(unique(ev_14[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20190901)
sim_14 <- rxode2::rxSolve(
  mod,
  events = ev_14,
  keep = c("WT", "PAGE", "arm")
) |>
  as.data.frame()

stopifnot(all(sim_14$Cc >= 0, na.rm = TRUE))
```

```{r figure-1}
# Replicates Figure 1 of van der Stoep 2019: full concentration-time profiles at
# 14 g/m^2, shown on the paper's sampling times and log concentration axis. The
# published panel holds 27 patients, all of whom received 14 g/m^2, so the
# subset is drawn from the 14 g/m^2 subjects only.
sample_times <- c(1.5, 3.5, 4, 5, 7, 9)

ids_14 <- cohort14$id[cohort14$dose_g_m2 == 14]

fig1 <- sim_14 |>
  filter(time %in% sample_times) |>
  filter(id %in% sample(ids_14, 27))

ggplot(fig1, aes(time, Cc, group = id)) +
  geom_line(linetype = "dotted", linewidth = 0.3) +
  geom_point(size = 1.2) +
  scale_y_log10(limits = c(1, 1000),
                breaks = c(1, 10, 100, 1000),
                labels = c("1", "10", "100", "1000")) +
  scale_x_continuous(limits = c(0, 14), breaks = seq(0, 14, by = 2)) +
  labs(x = "Time (h)", y = "Treosulfan blood concentration (mg/L)",
       caption = "Replicates Figure 1 of van der Stoep 2019 (27 simulated subjects at 14 g/m2).")
```

The published panel spans roughly 190-620 mg/L at 1.5 h and falls to roughly
5-80 mg/L at 9 h. The simulated envelope is compared on the same two anchor
times. These are stochastic quantities, so the assertions are on robust
quantiles and are set outside the range seen across repeated renders.

```{r figure-1-gate}
envelope <- sim_14 |>
  filter(time %in% c(1.5, 9), id %in% ids_14) |>
  group_by(time) |>
  summarise(
    q10 = quantile(Cc, 0.10),
    q50 = quantile(Cc, 0.50),
    q90 = quantile(Cc, 0.90),
    .groups = "drop"
  )
envelope

# Figure 1 medians read approximately 350 mg/L at 1.5 h and 20 mg/L at 9 h. A
# mis-transcribed volume, clearance or dose unit moves these by a factor, not by
# tens of percent, so a 2-fold band is a gate that still goes red.
med_15 <- envelope$q50[envelope$time == 1.5]
med_9 <- envelope$q50[envelope$time == 9]
stopifnot(med_15 > 175, med_15 < 700)
stopifnot(med_9 > 10, med_9 < 40)
```

# PKNCA validation

```{r pknca-14}
sim_nca_14 <- sim_14 |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

sim_nca_14 <- bind_rows(
  sim_nca_14,
  sim_nca_14 |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, arm, time)

conc_14 <- PKNCA::PKNCAconc(sim_nca_14, Cc ~ time | arm + id)

dose_14 <- ev_14 |>
  filter(evid == 1L) |>
  select(id, time, amt, arm)
dose_obj_14 <- PKNCA::PKNCAdose(dose_14, amt ~ time | arm + id)

nca_14 <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_14, dose_obj_14, intervals = intervals)
)
```

```{r pknca-summary}
nca_all <- bind_rows(as.data.frame(nca_t3), as.data.frame(nca_14))

nca_all |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  group_by(arm, PPTESTCD) |>
  summarise(
    median = median(PPORRES),
    p10 = quantile(PPORRES, 0.10),
    p90 = quantile(PPORRES, 0.90),
    .groups = "drop"
  ) |>
  mutate(across(c(median, p10, p90), \(x) signif(x, 3))) |>
  rename(
    "Arm" = arm,
    "NCA parameter" = PPTESTCD,
    "Median" = median,
    "10th pctile" = p10,
    "90th pctile" = p90
  ) |>
  knitr::kable(caption = "Simulated NCA summary by arm.")
```

## Comparison against published exposure

van der Stoep 2019 does not publish Cmax, Tmax or half-life. It publishes two
AUC(0-inf) figures, both of which this model should land on: the 1650 mg*h/L
target that Table 3 was constructed around (Methods 2.7), and the 1658 mg*h/L
median observed in the analysis cohort (Table 1).

```{r nca_compare}
published <- tibble::tribble(
  ~arm, ~aucinf.obs,
  "Table 3 target (typical values)", 1650,
  "Study protocol cohort (with IIV)", 1658
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_all,
  reference = published,
  by = "arm",
  units = c(aucinf.obs = "mg*h/L"),
  tolerance_pct = 25
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published treosulfan AUC(0-inf). * differs from reference by >25%.",
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

```{r nca_compare-gate}
pct <- as.numeric(gsub("[^0-9.+-]", "", cmp$`% diff`))
names(pct) <- cmp$arm

# The typical-value arm is deterministic -- no random effects, no reconstructed
# covariates -- and lands on the paper's own target to well under a percent. It
# carries the tight bound and is the load-bearing gate on the structural model.
stopifnot(abs(pct[["Table 3 target (typical values)"]]) < 1)

# The IIV arm is a cohort median from 150 subjects whose age, weight and BSA are
# all reconstructed, so it carries two irreducible sources of scatter. Over 400
# replicate cohorts of this design the median AUC realised 1748 +/- 61 (SD),
# i.e. -6.2% to +17.3% against the published 1658 -- roughly +5% of structural
# offset from the reconstruction, on top of +/-7% of pure sampling noise at
# n = 150. The bound is set outside that measured range and must not be
# tightened back: a transcription error in a dose, volume or clearance moves
# this by a factor, not by tens of percent, so 25% still goes red on anything
# real.
stopifnot(abs(pct[["Study protocol cohort (with IIV)"]]) < 25)
```

The IIV arm sits slightly above the observed 1658 mg*h/L. The residual is
attributable to the reconstruction rather than to the model: the patients' own
ages, weights and body surface areas are not published, and the largest single
driver of individual exposure in this model -- the maturation fraction, which
ranges from about 0.52 at birth to 0.98 in adolescence -- is a function of
exactly the covariate that has to be reconstructed. Dosing every subject at
14 g/m^2 instead of following the protocol's reduced infant dose raises this
arm's median by a further 8% or so, which is what makes the protocol dose rule
worth encoding. The typical-value arm carries no reconstruction at all and is
the gate that would catch a structural error.

# Assumptions and deviations

- **Hill coefficient 1.22 rather than the printed 1.2.** Table 2 prints the
  final Hill estimate as `1.2`; the Data S4 control stream carries `1.22`.
  Every other final `$THETA` in that stream matches its Table 2 entry exactly,
  which identifies it as carrying the final estimates, so the unrounded value
  is used. It is also marginally the better reproduction of Table 3 (root-mean-
  square deviation across the 66 rows 0.40% at 1.22 against 0.49% at 1.20), and
  the two are indistinguishable for practical purposes.

- **Gestational age is assumed, not observed.** Methods 2.5: "PMA was estimated
  by adding a gestational age of 40 weeks to postnatal age." The cohort carries
  no recorded gestational age, so `PAGE` in this model is always
  `40 + postnatal age in weeks`. A user with genuine gestational ages should
  supply `PAGE = GA_weeks + postnatal age in weeks` instead; the maturation
  term is poorly informed below term either way, since the youngest patient was
  0.1 years old.

- **`PAGE` is carried in weeks, not the register-default months.** The
  `PAGE` canonical in `inst/references/covariate-columns.md` defaults to months
  but explicitly provides for models whose source equations are written in
  weeks. TM50 = 38 weeks only makes sense on the week scale. The model file
  declares `units = "weeks"` in its `covariateData$PAGE` entry.

- **Body surface area is reconstructed for the IIV arm.** The paper doses by BSA
  but publishes BSA only as a cohort median and range. The scaling
  `BSA = 0.7 * (WT / 15.6)^(2/3)` is anchored on the Table 1 medians
  (15.6 kg, 0.7 m^2) and uses the standard two-thirds-power geometric
  relationship. It reproduces the published BSA range from the published weight
  range to within the printed precision:

    ```{r bsa-check}
    bsa_from_wt <- function(wt) 0.7 * (wt / 15.6)^(2 / 3)
    c(min_wt_3.8kg = bsa_from_wt(3.8), max_wt_75kg = bsa_from_wt(75))
    # Table 1 reports BSA 0.7 m^2 (range 0.3-1.9).
    stopifnot(
      abs(bsa_from_wt(3.8) - 0.3) < 0.1,
      abs(bsa_from_wt(75) - 1.9) < 0.15
    )
    ```

  The Table 3 arm -- which carries every gate on the structural model -- uses
  the paper's own published doses in mg and needs no BSA at all.

- **Age, weight and their pairing in the IIV arm are reconstructed.** Ages come
  from a piecewise-linear quantile function pinned to the four published age
  facts (minimum, 36th percentile, median, maximum); weights are the Table 3
  50th-percentile CDC weight-for-age curve scaled by 0.88 with log-normal
  scatter, the sub-unit factor bringing the cohort weight median onto the
  published 15.6 kg and reflecting that transplant candidates sit below the
  healthy weight-for-age median. The interpolation between the published
  anchors is an assumption; no per-patient covariate data are published. This
  affects the IIV arm's exposure distribution only -- no model parameter
  depends on it.

- **The IIV arm follows the study's protocol dose, not a flat 14 g/m^2.**
  Methods 2.1 gives 14 g/m^2 per day to patients older than 1 year and 10 or
  12 g/m^2 to those younger (16 and 2 of 91 patients); the vignette uses
  10 g/m^2 below 1 year, the dominant of the two. This is load-bearing for the
  exposure comparison rather than cosmetic: infants have the lowest maturation
  fraction, so dosing them at the older-child rate inflates their AUC by about
  40% and the cohort median by about 8%.

- **eGFR is deliberately absent.** Results 3.3 reports that eGFR was a
  statistically significant covariate on CL in the stepwise screen
  (dOFV = -16.72), but the prediction-corrected VPC worsened and the IIV of the
  PK parameters increased, so the authors excluded it from the final model. The
  model file therefore has no renal covariate; the screen is recorded in
  `population$renal_function`.

- **Interoccasion variability is encoded for three occasions.** Methods 2.4
  defines each dose and its subsequent sampling as a separate occasion and the
  regimen is three consecutive days; Data S4 multiplexes three etas off the
  `DAY` column with a shared variance. Simulations here use a single day, so
  `OCC = 1` throughout and only the first IOV eta is active. A user simulating
  the full three-day conditioning course should supply `OCC = 1, 2, 3` on the
  respective days.

- **Residual error is proportional only.** Methods 2.4 states that a combined
  proportional-plus-additive model was examined and the proportional model was
  retained. Four of 410 samples fell below the 6.8 mg/L LLOQ and were kept at
  their measured values (the Hecht 2011 approach), so no censoring is encoded.

- **NCA is run on uncensored simulated profiles.** The paper's AUC(0-inf) is
  model-derived (`DOSE * F1 / CL`, Methods 2.9), not NCA-derived, so an
  uncensored NCA is the correct comparator for it. Half-life is reported in the
  summary table for information only and is not gated: lambda-z estimated from
  uncensored simulated concentrations that decay far below the 6.8 mg/L assay
  limit is biased relative to any half-life an analyst would obtain from real
  assay data, and the paper publishes no half-life to compare against.

- **No non-paper-derived parameter values.** Every `ini()` entry comes from
  van der Stoep 2019 Table 2 or from the Supporting Information 4 NONMEM
  control stream. No value was digitised from a figure, supplied by
  correspondence, or carried from an upstream model.
