---
title: "Cefazolin in pregnancy, empirical vs semiphysiological (van Hasselt 2014)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefazolin in pregnancy, empirical vs semiphysiological (van Hasselt 2014)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load-models, include = FALSE}
ui_emp <- rxode2::rxode(readModelDb("vanHasselt_2014_cefazolin_empirical"))
ui_semi <- rxode2::rxode(readModelDb("vanHasselt_2014_cefazolin_semiphysiological"))
ui_crcl <- rxode2::rxode(readModelDb("vanHasselt_2014_crcl_pregnancy"))
```

# Model and source

This paper contributes **three** model files, which is how the authors built it:
two competing final population PK models for cefazolin, and the separately
developed creatinine-clearance model that the second of them consumes.

| Model file | What it is |
|---|---|
| `vanHasselt_2014_cefazolin_empirical` | Final PK model with a linear gestational-age effect on clearance |
| `vanHasselt_2014_cefazolin_semiphysiological` | Final PK model with clearance driven by the gestational rise in creatinine clearance |
| `vanHasselt_2014_crcl_pregnancy` | The creatinine-clearance trajectory itself (Methods 2.4.1, Table 3, Figure 3) |

* Citation: `r ui_emp$reference`
* Article: <https://doi.org/10.1155/2014/897216>
* Corrigendum: <https://doi.org/10.1155/2015/124035>

The base model (Table 2, first estimates column) is not packaged separately: it
is a model-building intermediate that both final models supersede.

# The corrigendum is load-bearing

**The version of this paper archived in PubMed Central is the uncorrected one.**
The 2015 corrigendum states that the two covariate equations in the Table 2
footnotes "have been erroneously switched", and gives the corrected assignment:

| | Corrected (corrigendum) |
|---|---|
| Empirical model | `CL = theta_CL0 + theta_CLPreg * (1 + GA/40)` |
| Semiphysiological model | `CL = theta_CL0 + theta_CLPreg * (CrCL(t) / CrCL_0)` |

Extracting from the archived parent alone would therefore attach each covariate
equation to the **wrong** model. The corrected assignment was not taken on the
corrigendum's word; it is confirmed independently two ways:

1. **It is the only assignment consistent with the paper's own Methods**, and
   this is the decisive evidence. Each footnote equation is verbatim a numbered
   equation from the Methods section that *defines that model*. Equation (3) is
   the linear covariate relation normalised by `NCOV`, which Methods 2.3 says is
   "the maximum GA of 40" for linear GA models -- that section builds the
   *empirical* model. Equation (6), `CL_i = theta_CL0 + theta_CLpreg *
   (CrCL_i(t)/CrCL_i0)`, is introduced in Methods 2.4.2, which builds the
   *semiphysiological* model. The archived footnotes pair each equation with the
   other section's model.
2. **Only the corrected assignment reproduces the difference between the two
   approaches that the paper reports.** Results 3.3.3 states that the empirical
   and semiphysiological predictions diverge "especially in early pregnancy".
   Under the switched reading the two models are numerically *indistinguishable*
   in early pregnancy -- 0.2% apart at GW 15 -- and diverge only at term, which
   inverts the paper's own finding. Checked numerically below.

A caution on what does **not** discriminate, recorded so a future reader does
not mistake it for evidence: clearance at the cohort median GA of 33 weeks is
**not** a usable test. All four readings land within 8% of the base model's
0.49 L/min, and the switched empirical reading (0.469) is actually marginally
*closer* to it than the corrected one (0.515). The median-GA arithmetic is a
consistency check on the corrected forms, not a way to choose between them.

The corrigendum also supplies the model's reference frame, which the parent
never states: `CL0`, `VC`, `VP` and `Q` "were fit based on the free cefazolin
concentration", so `CLtotal = CLfree * fu`. Both PK models therefore treat
`central / vc` as the **unbound** concentration and reconstruct total cefazolin
algebraically as `Cc = Cunbound / fu`, following the shipped
`Fauchet_2015_lopinavir_unbound` precedent.

```{r corrigendum-check}
# Methods eq 4 trajectory, typical values from Table 3.
crcl_traj <- function(ega) 97.83 + 83.83 * ega / (13.3 + ega)

cl_base <- 0.49 # Table 2, base model clearance (L/min)

# Empirical column thetas (Table 2): CL0 = 0.119, CLPreg = 0.217.
# Semiphysiological column thetas (Table 2): CL0 = 0.142, CLPreg = 0.212.
# Corrected: empirical takes the linear GA form, semiphysiological the ratio.
cl_emp_corrected <- function(ega) 0.119 + 0.217 * (1 + ega / 40)
cl_semi_corrected <- function(ega) 0.142 + 0.212 * (crcl_traj(ega) / crcl_traj(0))
# Switched (as archived): the two covariate forms swapped.
cl_emp_switched <- function(ega) 0.119 + 0.217 * (crcl_traj(ega) / crcl_traj(0))
cl_semi_switched <- function(ega) 0.142 + 0.212 * (1 + ega / 40)

ga_med <- 33 # Table 1, median gestational age

tibble::tibble(
  Model = c("Empirical", "Semiphysiological"),
  `CL corrected` = c(cl_emp_corrected(ga_med), cl_semi_corrected(ga_med)),
  `CL as-printed (switched)` = c(cl_emp_switched(ga_med), cl_semi_switched(ga_med)),
  `Base model CL` = cl_base
) |>
  knitr::kable(digits = 3, caption = "Clearance (L/min) at the cohort median GA of 33 weeks. Note that this comparison does NOT discriminate between the two readings.")
```

The discriminating comparison is the *gap between the two models* in early
pregnancy, which Results 3.3.3 reports as the regime where they diverge most.

```{r corrigendum-gap}
ga_early <- 15 # gestational weeks; the earliest level plotted in Figure 4

gap <- function(emp, semi, ega) semi(ega) / emp(ega) - 1

tibble::tibble(
  `Gestational week` = c(15, 25, 35, 40),
  `Gap, corrected (%)` = 100 * gap(cl_emp_corrected, cl_semi_corrected, c(15, 25, 35, 40)),
  `Gap, switched (%)` = 100 * gap(cl_emp_switched, cl_semi_switched, c(15, 25, 35, 40))
) |>
  knitr::kable(digits = 1, caption = "Relative difference between the empirical and semiphysiological clearances under each reading.")
```

```{r corrigendum-assert}
# Closed-form arithmetic on published constants -- no RNG, no cohort -- so
# exact bounds are appropriate here.
stopifnot(
  # (a) Consistency: both CORRECTED forms sit within 6% of the base model's
  # 0.49 L/min. This does not discriminate (see the caution above); it only
  # confirms the corrected forms are self-consistent with the base model.
  abs(cl_emp_corrected(ga_med) / cl_base - 1) < 0.06,
  abs(cl_semi_corrected(ga_med) / cl_base - 1) < 0.06,
  # (b) Discrimination: in early pregnancy the CORRECTED reading separates the
  # two approaches by several percent, as Results 3.3.3 describes...
  abs(gap(cl_emp_corrected, cl_semi_corrected, ga_early)) > 0.05,
  # ...whereas the SWITCHED reading collapses them onto each other there,
  # which contradicts the paper. Realised values: 7.9% vs 0.25%.
  abs(gap(cl_emp_switched, cl_semi_switched, ga_early)) < 0.01
)
```

# Population

The pooled analysis dataset covers **94 pregnant women** contributing 187
cefazolin observations across three sources (Methods 2.1): a prospective study
of 41 women undergoing in utero surgical intervention (153 observations, 2 g
every 8 h for two days, median GA 25 weeks), 24 samples at term caesarean
delivery after a 1 g bolus (GA fixed at 40 weeks for that study), and 10 fetal
interventions in 7 women after a single 2 g bolus (mean GA 27 weeks).

Baseline demographics (Table 1): median gestational age 33 weeks (range 17-40),
median age 31 years (20-42), median body weight 72 kg (54-99), and median serum
creatinine 0.64 mg/dL (0.33-0.88). Creatinine clearance was computed throughout
by Cockcroft-Gault using body weight.

Missing data were substantial and were imputed by pooled medians (Methods 2.5):
body weight was missing for 46% of patients, age for 55%, and some serum
creatinine values for 58%. In the semiphysiological arm specifically, missing
CrCL values were imputed by the typical trajectory rather than a median -- which
is part of why the paper finds CrCL uninformative as a plain empirical covariate
but informative once modelled.

The same information is available programmatically via
`readModelDb("vanHasselt_2014_cefazolin_empirical")()$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. They
are collected here for review. Table 2 columns are named as in the paper;
"Empirical" and "Semiphysiological" refer to its two final-model columns.

| Equation / parameter | Value | Source location |
|---|---|---|
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Results 3.1, "a two-compartmental model best described the data" |
| `Cc <- Cunbound / fu` | n/a | Results 3.1 eq (7), `C_total = C_free / f_u` |
| Unbound reference frame | n/a | Corrigendum clarification, `CL_total = CL_free * f_u` |
| **Empirical model** | | |
| `preg_cl <- 1 + EGA/40` | n/a | Corrigendum Table 2 footnote a; Methods eq (3) with `NCOV` = 40 |
| `lcl_nonren` | 0.119 L/min | Table 2, Clearance `theta_CL0`, empirical (RSE 58%) |
| `lcl_renal` | 0.217 | Table 2, Gestation effect `theta_CLPreg`, empirical (RSE 16%) |
| `lvc` | 33.1 L | Table 2, Central volume `V_C`, empirical (RSE 17%) |
| `lvp` | 12.8 L | Table 2, Peripheral volume `V_P`, empirical (RSE 27%) |
| `lq` | 0.326 L/min | Table 2, Intercompartmental clearance `Q`, empirical (RSE 25%) |
| `lfu` | 0.286 | Table 2, Free fraction `F_U`, empirical (RSE 5%) |
| `etalcl`, `etalvc`, `etalvp`, `etalfu` | 19.9, 47.6, 34.2, 17.5 CV% | Table 2, Between subject variability, empirical |
| `propSd`, `addSd` | 0.0328, 0.842 (variances) | Table 2, Proportional / Additive total concentration, empirical |
| `propSd_Cunbound`, `addSd_Cunbound` | 0.0158, 0.229 (variances) | Table 2, Proportional / Additive free concentration, empirical |
| **Semiphysiological model** | | |
| `preg_cl <- crcl / crcl_ega0` | n/a | Corrigendum Table 2 footnote b; Methods eq (6) |
| `lcl_nonren` | 0.142 L/min | Table 2, Clearance `theta_CL0`, semiphysiological (RSE 44%) |
| `lcl_renal` | 0.212 | Table 2, Gestation effect `theta_CLPreg`, semiphysiological (RSE 38%) |
| `lvc` | 14.1 L | Table 2, Central volume `V_C`, semiphysiological (RSE 25%) |
| `lvp` | 17.1 L | Table 2, Peripheral volume `V_P`, semiphysiological (RSE 7%) |
| `lq` | 0.436 L/min | Table 2, Intercompartmental clearance `Q`, semiphysiological (RSE 10%) |
| `lfu` | 0.291 | Table 2, Free fraction `F_U`, semiphysiological (RSE 9%) |
| `etalcl`, `etalvc`, `etalvp`, `etalfu` | 10.4, 101.5, 68, 19.1 CV% | Table 2, Between subject variability, semiphysiological |
| `propSd`, `addSd` | 0.0328, 0.681 (variances) | Table 2, Proportional / Additive total concentration, semiphysiological |
| `propSd_Cunbound`, `addSd_Cunbound` | 0.0153, 0.217 (variances) | Table 2, Proportional / Additive free concentration, semiphysiological |
| **Creatinine-clearance model** | | |
| `crcl <- crcl_ega0 + crcl_matspan * EGA/(crcl_ega50 + EGA)` | n/a | Methods eq (4) |
| `lcrcl_ega0` | 97.83 mL/min, FIXED | Table 3, Baseline CrCL `CrCL0` (upstream RSE 3.91%) |
| `lcrcl_matspan` | 83.83 mL/min, FIXED | Table 3, Maximum CrCL `CrCLMAX` (upstream RSE 12.48%) |
| `lcrcl_ega50` | 13.3 weeks, FIXED | Table 3, Time of half-maximum CrCL `CrCL50` (upstream RSE 37.59%) |
| `etalcrcl_ega0`, `etalcrcl_matspan`, `etalcrcl_ega50` | 31.4, 35.1, 111.8 CV% | Table 3, Between subject variability |
| `propSd` (CrCL model) | 0.0291 (variance) | Table 3, Proportional error `sigma_CrCL` (RSE 43%); Methods eq (5) |

# The gestational creatinine-clearance trajectory

`vanHasselt_2014_crcl_pregnancy` packages Methods equation (4) on its own. Its
time axis **is** maternal gestational age in weeks, so it can be solved directly
over a pregnancy.

```{r crcl-figure-3}
# crcl_ega50 = 13.3 is not a multiple of 0.25, so it is added explicitly:
# the half-span identity below must be evaluated at the exact week, not at a
# neighbouring grid point.
ega_grid <- sort(unique(c(seq(0, 42, by = 0.25), 13.3)))
crcl_typ <- rxode2::rxSolve(
  rxode2::zeroRe(ui_crcl),
  rxode2::et(ega_grid),
  returnType = "data.frame"
)

ggplot(crcl_typ, aes(time, crcl)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = 97.83, linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 13.3, linetype = "dotted", colour = "grey40") +
  labs(
    x = "Gestational age (weeks)", y = "Creatinine clearance (mL/min)",
    title = "Typical gestational rise in creatinine clearance",
    caption = paste(
      "Replicates the typical-value curve underlying Figure 3 of van Hasselt 2014.",
      "Dashed = nonpregnant baseline; dotted = crcl_ega50."
    )
  )
```

These are exact structural identities of equation (4), not cohort statistics, so
they are asserted exactly.

```{r crcl-identities}
# Fail loudly rather than silently returning a neighbouring grid point: a
# near-miss lookup would turn the exact identities below into approximate
# ones without anything going red.
crcl_at <- function(w) {
  i <- which(abs(crcl_typ$time - w) < 1e-9)
  if (length(i) != 1L) stop("no exact grid point at EGA ", w)
  crcl_typ$crcl[i]
}

stopifnot(
  # At EGA 0 the curve collapses to the nonpregnant baseline. This is the
  # anchor the semiphysiological PK model divides by, so if it drifts, that
  # model's covariate term silently stops being 1 at conception.
  isTRUE(all.equal(crcl_at(0), 97.83, tolerance = 1e-8)),
  # At EGA = crcl_ega50 exactly half the span has been covered.
  isTRUE(all.equal(crcl_at(13.3), 97.83 + 83.83 / 2, tolerance = 1e-6)),
  # Monotone increasing over gestation.
  all(diff(crcl_typ$crcl) > 0)
)

tibble::tibble(
  `Gestational age (weeks)` = c(0, 13.3, 20, 27, 33, 40),
  `CrCL (mL/min)` = vapply(c(0, 13.3, 20, 27, 33, 40), crcl_at, numeric(1))
) |>
  dplyr::mutate(`Fraction of span covered` = (`CrCL (mL/min)` - 97.83) / 83.83) |>
  knitr::kable(digits = 3, caption = "Typical CrCL trajectory at selected gestational ages.")
```

Note what `crcl_matspan` is **not**: at term (40 weeks) the curve has covered
only `r round(100 * (crcl_at(40) - 97.83) / 83.83, 1)`% of its span, reaching
`r round(crcl_at(40), 1)` mL/min rather than the asymptote of
`r round(97.83 + 83.83, 1)` mL/min. This is exactly why the canonical is named
`matspan` and not `max`.

# Virtual cohort and typical-value profiles

The paper's simulations (Methods 2.6) use 2 g cefazolin infused over 30 minutes
every 6 or 8 hours, evaluated at different periods of pregnancy against an
MIC90 of 4 mg/L for coagulase-negative *Staphylococcus*. Figure 4 plots the
resulting **free** cefazolin profiles (per the corrigendum, which corrects a
caption that had claimed free *and* total).

```{r regimens}
MIC90 <- 4 # mg/L, Methods 2.6
tau <- c(q6h = 360, q8h = 480) # minutes
gw_levels <- c(15, 25, 35) # gestational weeks, as in Figure 4
n_doses <- 8L # enough to reach steady state
```

```{r solve-helper}
# Both PK models declare TWO endpoints (Cc and Cunbound). Observation rows must
# therefore name a declared endpoint, and rxSolve needs useLinCmt = FALSE --
# its default ODE->linCmt conversion corrupts the dvid->cmt mapping for this
# model shape.
solve_profile <- function(ui, ega, ii, typical = TRUE, n = 1L, id_offset = 0L) {
  last_dose <- ii * (n_doses - 1L)
  grid <- seq(last_dose, last_dose + ii, by = 1)
  ev <- rxode2::et(amt = 2000, dur = 30, ii = ii, addl = n_doses - 1L, cmt = "central") |>
    rxode2::et(grid, cmt = "Cc")
  d <- as.data.frame(ev)
  d <- d[rep(seq_len(nrow(d)), times = n), ]
  d$id <- id_offset + rep(seq_len(n), each = nrow(d) / n)
  d$EGA <- ega
  mod <- if (typical) rxode2::zeroRe(ui) else ui
  out <- rxode2::rxSolve(mod, d, returnType = "data.frame", useLinCmt = FALSE)
  if (is.null(out$id)) out$id <- 1L + id_offset
  out$tad <- out$time - last_dose
  out$GW <- ega
  out$regimen <- names(tau)[match(ii, tau)]
  # Return a FIXED column set. The semiphysiological model emits its embedded
  # creatinine-clearance intermediates as extra columns, so the two models' raw
  # not row-bindable without this.
  out[, c("id", "time", "tad", "GW", "regimen", "cl", "Cc", "Cunbound")]
}
```

```{r typical-profiles}
grid_arms <- tidyr::expand_grid(
  model = c("Empirical", "Semiphysiological"),
  GW = gw_levels,
  regimen = names(tau)
)

typ <- do.call(rbind, lapply(seq_len(nrow(grid_arms)), function(i) {
  a <- grid_arms[i, ]
  ui <- if (a$model == "Empirical") ui_emp else ui_semi
  out <- solve_profile(ui, ega = a$GW, ii = tau[[a$regimen]])
  out$model <- a$model
  out
}))

stopifnot(nrow(typ) > 0, !anyNA(typ$Cunbound))
```

```{r figure-4, fig.height = 5}
typ |>
  dplyr::mutate(GW_lab = paste0("GW ", GW)) |>
  ggplot(aes(tad / 60, Cunbound, colour = regimen)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = MIC90, linetype = "dashed") +
  facet_grid(model ~ GW_lab) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Free cefazolin (mg/L)",
    colour = "Regimen",
    title = "Steady-state free cefazolin, 2 g over 30 min",
    caption = paste(
      "Replicates Figure 4 of van Hasselt 2014 (free cefazolin only, per the corrigendum).",
      "Dashed line = MIC90 of 4 mg/L."
    )
  )
```

## The paper's dosing conclusion

Results 3.3.3 concludes that moving from q8h to q6h "yielded prolonged
therapeutic concentrations" against the 4 mg/L MIC90. The time above MIC is
computed over the **whole** dosing interval, so the two regimens are compared on
a common footing.

```{r t-above-mic}
t_above_mic <- typ |>
  dplyr::group_by(model, GW, regimen) |>
  dplyr::summarise(
    pct_T_over_MIC = 100 * mean(Cunbound > MIC90),
    cmin = min(Cunbound),
    cmax = max(Cunbound),
    .groups = "drop"
  )

t_above_mic |>
  dplyr::rename(
    "Model" = model, "Gestational week" = GW, "Regimen" = regimen,
    "%T > MIC90" = pct_T_over_MIC, "Cmin (mg/L)" = cmin, "Cmax (mg/L)" = cmax
  ) |>
  knitr::kable(digits = c(0, 0, 0, 1, 3, 1), caption = "Steady-state free cefazolin exposure vs the 4 mg/L MIC90.")
```

```{r t-above-mic-assert}
wide <- t_above_mic |>
  dplyr::select(model, GW, regimen, pct_T_over_MIC) |>
  tidyr::pivot_wider(names_from = regimen, values_from = pct_T_over_MIC)

stopifnot(
  # Guard against a lookup that silently tests nothing.
  nrow(wide) == length(gw_levels) * 2L,
  all(c("q6h", "q8h") %in% names(wide)),
  # The paper's claim: q6h prolongs time above MIC90 relative to q8h, in
  # every arm. Deterministic typical-value solves, so this is exact.
  all(wide$q6h > wide$q8h),
  # And the trough at q8h sits below the MIC90 everywhere, which is the
  # finding that motivates the recommendation.
  all(t_above_mic$cmin[t_above_mic$regimen == "q8h"] < MIC90)
)
```

# PKNCA validation

The source paper reports no NCA table, so there is nothing to compare against
externally. The check that *is* available is a strong one: at steady state the
non-compartmental clearance `dose / AUCtau` must reproduce the model's own
clearance. Both sides of this comparison use the same drawn parameters, so the
residual difference is pure trapezoidal error and a tight bound is correct.

Because the disposition parameters are referenced to the **unbound**
concentration, `dose / AUCtau` computed on `Cunbound` recovers `CL_free`
directly.

```{r pknca}
nca_in <- typ |>
  dplyr::mutate(
    arm = paste(model, paste0("GW", GW), regimen, sep = " | "),
    conc_id = as.integer(factor(arm))
  ) |>
  dplyr::filter(!is.na(Cunbound)) |>
  dplyr::select(conc_id, arm, tad, Cunbound)

conc_obj <- PKNCA::PKNCAconc(nca_in, Cunbound ~ tad | arm + conc_id)

dose_df <- nca_in |>
  dplyr::distinct(conc_id, arm) |>
  dplyr::mutate(tad = 0, amt = 2000)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ tad | arm + conc_id)

ivals <- nca_in |>
  dplyr::group_by(conc_id, arm) |>
  dplyr::summarise(start = 0, end = max(tad), .groups = "drop") |>
  dplyr::mutate(auclast = TRUE, cmax = TRUE, cmin = TRUE, tmax = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = as.data.frame(ivals)))
```

```{r pknca-compare}
nca_wide <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("auclast", "cmax", "cmin")) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

model_cl <- typ |>
  dplyr::mutate(arm = paste(model, paste0("GW", GW), regimen, sep = " | ")) |>
  dplyr::group_by(arm) |>
  dplyr::summarise(cl_model = dplyr::first(cl), .groups = "drop")

cmp <- nca_wide |>
  dplyr::left_join(model_cl, by = "arm") |>
  dplyr::mutate(
    cl_nca = 2000 / auclast,
    pct_diff = 100 * (cl_nca / cl_model - 1)
  )

cmp |>
  dplyr::select(arm, cl_model, cl_nca, pct_diff, cmax, cmin) |>
  dplyr::rename(
    "Arm" = arm,
    "CL from model (L/min)" = cl_model,
    "CL from NCA, dose/AUCtau (L/min)" = cl_nca,
    "Difference (%)" = pct_diff,
    "Free Cmax (mg/L)" = cmax,
    "Free Cmin (mg/L)" = cmin
  ) |>
  knitr::kable(digits = c(0, 4, 4, 3, 1, 3), caption = "Steady-state mass-balance check: NCA clearance vs the model's own clearance.")
```

```{r pknca-assert}
stopifnot(
  nrow(cmp) == nrow(grid_arms),
  !anyNA(cmp$pct_diff),
  # Deterministic identity CL = dose/AUCtau at steady state; both sides use
  # the same drawn parameters, so the only error is trapezoidal
  # discretisation on a 1-minute grid. Realised max was 0.0045%, so this
  # bound is ~20x the observed error and still goes red on any structural
  # regression (a dropped peripheral compartment, a wrong reference frame or
  # a units slip all move it by whole percent).
  max(abs(cmp$pct_diff)) < 0.1
)
```

# Between-subject variability

The semiphysiological model carries a much smaller `omega_CL` than the
empirical one (10.4 vs 19.9 CV%, Table 2). Read on its own that looks like a
model that predicts far less variability in clearance, but it is not: the
missing variability has moved into the creatinine-clearance trajectory's own
three random effects, which this model embeds and which propagate into `cl`
through the `CrCL(t)/CrCL(0)` ratio.

The stochastic cohort below shows the consequence -- the two models produce a
**similar** realised spread in individual clearance despite the near-twofold
difference in their tabulated `omega_CL`. That is the point worth taking from
this section: `omega_CL` alone is not comparable between the two
parameterisations.

```{r vpc, fig.height = 4}
n_sub <- 100L # per arm; the 200/arm cap is a ceiling, not a target

vpc <- do.call(rbind, lapply(c("Empirical", "Semiphysiological"), function(m) {
  ui <- if (m == "Empirical") ui_emp else ui_semi
  out <- solve_profile(ui, ega = 25, ii = tau[["q8h"]], typical = FALSE, n = n_sub)
  out$model <- m
  out
}))

vpc |>
  dplyr::group_by(model, tad) |>
  dplyr::summarise(
    Q05 = quantile(Cunbound, 0.05),
    Q50 = quantile(Cunbound, 0.50),
    Q95 = quantile(Cunbound, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad / 60, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = MIC90, linetype = "dashed") +
  facet_wrap(~model) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)", y = "Free cefazolin (mg/L)",
    title = "Simulated free cefazolin at GW 25, 2 g q8h (median and 5th-95th percentiles)",
    caption = "Comparable to the visual predictive checks of Figure 1. Dashed line = MIC90."
  )
```

```{r vpc-assert}
spread <- vpc |>
  dplyr::group_by(model, id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop") |>
  dplyr::group_by(model) |>
  dplyr::summarise(cv = 100 * sd(cl) / mean(cl), .groups = "drop")

knitr::kable(
  dplyr::rename(spread, "Model" = model, "Realised CV% of individual CL" = cv),
  digits = 1,
  caption = "Between-subject spread in clearance across the simulated cohort."
)

stopifnot(
  nrow(spread) == 2L,
  # Both cohorts show a plausible clearance spread. Bounds are wide because
  # these are cohort statistics, which rxode2 draws differently at different
  # solver-thread counts; realised values here were 16.9% and
  # 16.4%. A mis-entered omega would move these by a factor, not a few points.
  all(spread$cv > 3), all(spread$cv < 45)
)
```

# Assumptions and deviations

- **The archived parent is uncorrected.** All covariate-equation assignments
  follow the 2015 corrigendum, not the PMC-archived 2014 text. Every affected
  in-file comment points at the corrigendum.
- **Between-subject variability scale.** Tables 2 and 3 report IIV as "CV%" for
  an exponential IIV model (Methods eq 1). This is read as
  `omega^2 = (CV/100)^2`, the usual NONMEM convention where a reported CV% is
  `100 * sqrt(omega^2)`. The exact-lognormal alternative
  `omega^2 = log(1 + CV^2)` would give materially smaller variances at the
  largest terms (101.5 CV% and 111.8 CV%), and the paper does not state which
  conversion it used. The chosen reading is consistent across all three model
  files.
- **Residual-error scale.** Tables 2 and 3 head their residual blocks
  "Residual unexplained variability variance(s)", so every tabulated sigma is
  treated as a variance and the `ini()` entry is its square root.
- **Time unit.** Clearances are reported in L/min and volumes in L, so the PK
  models use minutes as the time axis, unconverted.
- **The CrCL model's structural parameters are inherited, not fitted here.**
  All three are `fixed()`, taken from the upstream meta-analysis the paper cites
  as reference 13. Their tabulated RSEs are that meta-analysis's precision. The
  upstream meta-analysis itself was not available when this model was built;
  only the three point estimates it produced, as reprinted in Table 3, are used.
- **`CrCLMAX` units.** Table 3 leaves the unit cell blank for this row. mL/min
  is forced by equation (4), where the term is added to a mL/min baseline.
- **Two-stage covariate propagation.** The paper fitted the CrCL model
  separately and carried individual empirical Bayes estimates into the PK model.
  The packaged semiphysiological model instead embeds the trajectory with its
  three random effects, which is the faithful way to *simulate* from the
  combined two-stage model.
- **The empirical model must not be extrapolated below the observed GA range.**
  Its clearance does not collapse to a non-pregnant value at EGA 0 -- the
  gestational arm still contributes `cl_renal * 1` there. The paper makes
  exactly this point in its Discussion; the semiphysiological model is the one
  built for extrapolation.
- **No published NCA table exists** for this paper, so the PKNCA section is a
  self-consistency check against the model's own clearance rather than a
  comparison to published values.
- **Cosmetic corrigendum items**, recorded for completeness: Figure 1's x-axis
  is in minutes, not hours; Figure 4 depicts free cefazolin only.
