---
title: "Temocillin PK/PD against Escherichia coli (van Os 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Temocillin PK/PD against Escherichia coli (van Os 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

van Os and colleagues combined static time-kill (STK) experiments and a
hollow-fibre infection model (HFIM) to build a semi-mechanistic PK/PD model of
temocillin activity against four *Escherichia coli* strains, then coupled that
PD model to a published population PK model of temocillin in critically ill
patients in order to compare dosing regimens.

The paper fitted the PD model **separately for each of the four strains**
(Table 3 reports four independent columns of estimates with their own relative
standard errors), so the extraction is four model files that share one
structure, plus this single vignette.

```{r models}
strainModels <- c(
  "ATCC 25922" = "vanOs_2024_temocillin_atcc25922",
  "ISOL MIC16" = "vanOs_2024_temocillin_isolmic16",
  "ISOL MIC8"  = "vanOs_2024_temocillin_isolmic8",
  "ISOL MIC4"  = "vanOs_2024_temocillin_isolmic4"
)
strainMIC <- c("ATCC 25922" = 16, "ISOL MIC16" = 16, "ISOL MIC8" = 8, "ISOL MIC4" = 4)

mod1 <- readModelDb(strainModels[["ATCC 25922"]])
```

* Citation: `r rxode2::rxode(mod1)$reference`
* Article: <https://doi.org/10.1093/jac/dkae243>
* Supplementary data (includes the NONMEM simulation control stream
  `NONMEM_code.txt`): <https://doi.org/10.1093/jac/dkae243>

# Population

The pharmacodynamic layer is *in vitro*. Four *E. coli* strains were chosen to
span the less susceptible end of the wild-type MIC distribution for the species
(4-16 mg/L): the reference strain ATCC 25922 (MIC 16 mg/L) and three clinical
isolates from the Vienna General Hospital -- ISOL MIC16 (catheter urine,
*bla*CTX-M-15 and *bla*OXA-1, MIC 16 mg/L), ISOL MIC8 (skin swab, MIC 8 mg/L)
and ISOL MIC4 (rectal swab, *bla*TEM-1 and *bla*CTX-M-1, MIC 4 mg/L)
(main-text Table 1). MICs were determined in triplicate by broth microdilution
following CLSI guidelines.

STK experiments were run in triplicate in 5 mL cation-adjusted Mueller-Hinton
broth at 37 degrees C, inoculated at 1.5e6 cfu/mL, over temocillin
concentrations of 0.125-8x MIC plus a growth control, sampled over 24 h. The
HFIM replicated the unbound plasma concentration-time profiles of four
intravenous regimens over 72 h. Total counts were read on drug-free agar and the
resistant subpopulation on agar containing 32 mg/L temocillin; the theoretical
limit of detection was 50 cfu/mL and observations below it were handled with the
M3 method.

The pharmacokinetic layer is **not** fitted in this paper. It is the
two-compartment population PK model of temocillin in critically ill patients
from Laterre et al. 2015 (n = 11), whose parameter values this paper reproduces
in its Table 2 and hard-codes in the supplementary NONMEM control stream. All PK
parameters and their inter-individual variances are therefore encoded with
`fixed()`.

The same information is available programmatically:

```{r population}
str(readModelDb(strainModels[["ATCC 25922"]])()$population, max.level = 1)
```

# Source trace

## Model equations

| Equation | Source location |
|---|---|
| `d/dt(bact_susceptible)`, sigmoidal Emax kill and logistic growth | Equation 1 (p. 2487) |
| `d/dt(bact_less_susceptible)`, sigmoidal Emax kill and logistic growth | Equation 2 (p. 2487) |
| `d/dt(bact_resistant)`, linear kill, growth capped on `S + LS` | Equation 3 (p. 2487) |
| Two-compartment plasma PK; `Cu = central / vc * fu` drives all three effects | Table 2; Supplementary `NONMEM_code.txt` `$PK` / `$DES` |
| Total count reported as `log10(S + LS)`; resistant count capped at the total | Methods "PK/PD model"; Supplementary `NONMEM_code.txt` `$ERROR` |
| Rescaling of the fitted inocula to a 1e6 cfu/mL total at time zero | Methods "PK/PD simulations"; `NONMEM_code.txt` `THETA(1)`, `THETA(6)`, `THETA(11)` |

## Parameters

Every `ini()` entry carries an in-file comment naming its source. The table
below collects the values from all four strain models in one place; the
pharmacodynamic values are main-text Table 3 and the pharmacokinetic values are
main-text Table 2.

```{r source-trace}
iniOf <- function(mn) {
  d <- rxode2::rxode(readModelDb(mn))$iniDf
  d <- d[!is.na(d$name), ]
  data.frame(Parameter = d$name, est = d$est, fixed = d$fix, stringsAsFactors = FALSE)
}
tr <- Reduce(
  function(a, b) dplyr::full_join(a, b, by = "Parameter"),
  Map(function(nm, mn) {
    x <- iniOf(mn)
    x$value <- ifelse(x$fixed, paste0(signif(x$est, 6), " (FIX)"), as.character(signif(x$est, 6)))
    stats::setNames(x[, c("Parameter", "value")], c("Parameter", nm))
  }, names(strainModels), strainModels)
)
tr$Source <- ifelse(
  grepl("^(lcl|lvc|lq|lvp|fu|etalcl|etalvc)$", tr$Parameter),
  "Table 2 (from Laterre et al. 2015)",
  ifelse(tr$Parameter == "log10cfu0_total", "Methods, PK/PD simulations", "Table 3")
)
knitr::kable(tr, caption = "Parameter values by strain, with source location. Log-scale PK parameters are shown on the log scale as stored in ini().")
```

# Virtual cohort

Original observed data are not publicly available. The simulations below
reproduce the paper's Monte Carlo design: virtual critically ill patients
carrying the Laterre inter-individual variability on clearance and central
volume, with no residual variability, exactly as described in Methods
("PK/PD simulations").

The paper simulated 1000 patients per regimen. This vignette uses 100 per arm
across the 4 strains x 6 regimens grid to stay inside the library's cohort cap
and render budget; the consequence for three borderline cells is quantified in
*Assumptions and deviations*.

```{r cohort}
set.seed(20240720)
nSub <- 100L

regimens <- c("2 g q12h II", "2 g q8h II", "4 g/day CI + 2 g LD",
              "6 g/day CI + 2 g LD", "8 g/day CI + 2 g LD", "12 g/day CI + 2 g LD")

# One subject's dosing records for a regimen. Doses are in mg and infusion
# rates in mg/h, matching the model's units metadata. Intermittent infusions
# run over 30 min (Methods, HFIM); continuous infusions are encoded as a single
# 72 h infusion at the daily rate, preceded by a 2 g loading dose.
doseRecords <- function(regimen, hours = 72) {
  ld <- as.data.frame(rxode2::et(amt = 2000, rate = 4000, cmt = "central"))
  ci <- function(gPerDay) {
    rt <- gPerDay * 1000 / 24
    as.data.frame(rxode2::et(amt = rt * hours, rate = rt, cmt = "central"))
  }
  ii <- function(interval) {
    as.data.frame(rxode2::et(amt = 2000, rate = 4000, ii = interval,
                             addl = ceiling(hours / interval) - 1, cmt = "central"))
  }
  switch(
    regimen,
    "2 g q12h II"          = ii(12),
    "2 g q8h II"           = ii(8),
    "4 g/day CI + 2 g LD"  = dplyr::bind_rows(ld, ci(4)),
    "6 g/day CI + 2 g LD"  = dplyr::bind_rows(ld, ci(6)),
    "8 g/day CI + 2 g LD"  = dplyr::bind_rows(ld, ci(8)),
    "12 g/day CI + 2 g LD" = dplyr::bind_rows(ld, ci(12))
  )
}

# CFUtotal and CFUres are DECLARED endpoints of these models (they appear in
# predDf), so observation rows must carry cmt = "CFUtotal" -- pointing them at
# an ODE state such as "central" is what fails here. See Assumptions.
makeCohort <- function(regimen, obsTimes, n, idOffset = 0L, hours = 72) {
  one <- dplyr::bind_rows(
    doseRecords(regimen, hours = hours),
    as.data.frame(rxode2::et(obsTimes, cmt = "CFUtotal"))
  )
  dplyr::bind_rows(lapply(seq_len(n), function(i) {
    dplyr::mutate(one, id = idOffset + i, regimen = regimen)
  }))
}

events <- dplyr::bind_rows(lapply(seq_along(regimens), function(k) {
  makeCohort(regimens[k], obsTimes = c(24, 72), n = nSub, idOffset = (k - 1L) * nSub)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(dplyr::n_distinct(events$id) == nSub * length(regimens))
```

# Simulation

The Laterre inter-individual variances are supplied explicitly to `rxSolve()`;
`rxSolve` otherwise reuses the previous solve's `omega`.

```{r simulate}
omegaPK <- lotri::lotri(etalcl ~ 0.1218636, etalvc ~ 0.2899794)

simStrain <- function(modelName, ev) {
  s <- rxode2::rxSolve(
    readModelDb(modelName), ev,
    omega = omegaPK, keep = "regimen",
    addDosing = FALSE, useLinCmt = FALSE,
    returnType = "data.frame"
  )
  stopifnot(dplyr::n_distinct(s$id) == dplyr::n_distinct(ev$id))
  s
}

simFig3 <- dplyr::bind_rows(lapply(names(strainModels), function(sn) {
  simStrain(strainModels[[sn]], events) |> dplyr::mutate(strain = sn)
}))
simFig3$strain <- factor(simFig3$strain, levels = names(strainModels))
simFig3$regimen <- factor(simFig3$regimen, levels = regimens)
```

# Replicate published figures

## Time course of the total and resistant populations

The paper's full PK and PD time courses are Figure S4. This block shows the
typical-value (zero random effects) trajectory for every strain and regimen,
which is the deterministic backbone of the Monte Carlo simulations summarised in
Figure 3.

```{r figure-s4}
tcTimes <- seq(0, 72, by = 0.5)
tcEvents <- dplyr::bind_rows(lapply(regimens, function(rg) {
  dplyr::bind_rows(doseRecords(rg), as.data.frame(rxode2::et(tcTimes, cmt = "CFUtotal"))) |>
    dplyr::mutate(id = 1L, regimen = rg)
}))

simTC <- dplyr::bind_rows(lapply(names(strainModels), function(sn) {
  m <- rxode2::zeroRe(readModelDb(strainModels[[sn]]))
  rxode2::rxSolve(m, tcEvents, omega = NA, keep = "regimen",
                  addDosing = FALSE, useLinCmt = FALSE,
                  returnType = "data.frame") |>
    dplyr::mutate(strain = sn)
}))
simTC$strain <- factor(simTC$strain, levels = names(strainModels))
simTC$regimen <- factor(simTC$regimen, levels = regimens)

simTC |>
  dplyr::select(time, strain, regimen, Total = CFUtotal, Resistant = CFUres) |>
  tidyr::pivot_longer(c(Total, Resistant), names_to = "Population", values_to = "logCfu") |>
  # The ODE is continuous, so a fully eradicated subpopulation keeps decaying
  # below any biologically meaningful count; clamp for display only.
  dplyr::mutate(logCfu = pmax(logCfu, -2)) |>
  ggplot(aes(time, logCfu, colour = Population)) +
  geom_hline(yintercept = 6, linetype = "dashed", colour = "grey40") +
  geom_line() +
  facet_grid(strain ~ regimen) +
  labs(x = "Time (h)", y = expression(log[10]~cfu/mL),
       title = "Typical-value bacterial time courses",
       caption = paste("Companion to Figure S4 of van Os 2024. Dashed line is the",
                       "initial total population (stasis level, 1e6 cfu/mL).")) +
  theme_bw() + theme(legend.position = "bottom")
```

## Figure 3 -- predicted counts at 24 h and 72 h

```{r figure-3}
fig3 <- simFig3 |>
  dplyr::filter(time %in% c(24, 72)) |>
  dplyr::select(id, time, strain, regimen, Total = CFUtotal, Resistant = CFUres) |>
  tidyr::pivot_longer(c(Total, Resistant), names_to = "Population", values_to = "logCfu") |>
  dplyr::mutate(logCfu = pmax(logCfu, -2),
                Timepoint = factor(paste0(time, " h"), levels = c("24 h", "72 h")))

ggplot(fig3, aes(regimen, logCfu, fill = Population)) +
  geom_hline(yintercept = c(4, 6), linetype = "dashed", colour = "grey40") +
  geom_boxplot(outlier.size = 0.4) +
  facet_grid(strain ~ Timepoint) +
  labs(x = NULL, y = expression(log[10]~cfu/mL),
       title = "Predicted bacterial counts at 24 h and 72 h",
       caption = paste("Replicates Figure 3 of van Os 2024 (unbound fraction 0.41).",
                       "Dashed lines are the initial total population (1e6 cfu/mL)",
                       "and a 2 log10 reduction.")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1), legend.position = "bottom")
```

## Quantitative check of the paper's regimen claims

The Results section makes several sharp, falsifiable statements about the
**median** predicted counts. Each is restated below as a test against the
simulation.

```{r figure-3-claims}
med <- simFig3 |>
  dplyr::group_by(strain, regimen, time) |>
  dplyr::summarise(cfu = median(CFUtotal), .groups = "drop") |>
  tidyr::pivot_wider(names_from = time, values_from = cfu, names_prefix = "h")

knitr::kable(
  med |> dplyr::rename("Strain" = strain, "Regimen" = regimen,
                       "Median log10 cfu/mL at 24 h" = h24,
                       "Median log10 cfu/mL at 72 h" = h72),
  digits = 2,
  caption = "Median predicted total bacterial count. Stasis level is 6.00 log10 cfu/mL."
)

g <- function(st, rg, tp) med[[paste0("h", tp)]][med$strain == st & med$regimen == rg]
others <- function(st, rg, tp) med[[paste0("h", tp)]][med$strain == st & med$regimen != rg]

claims <- tibble::tibble(
  Claim = c(
    "ATCC 25922: 24 h median below stasis for every regimen except 2 g q12h",
    "ISOL MIC16: 24 h median below stasis for every regimen except 2 g q12h",
    "ATCC 25922: 72 h median reaches stasis only with the 12 g daily CI regimen",
    "ISOL MIC16: 72 h median reaches stasis only with the 12 g daily CI regimen",
    "ISOL MIC8: 24 h median below stasis for all simulated regimens",
    "ISOL MIC8: 72 h median stasis for 6 g daily regimens but not 4 g daily regimens",
    "ISOL MIC8: 2 log10 reduction at 72 h with 6 g/day CI and with 8 g/day",
    "ISOL MIC4: sustained killing (72 h median below stasis) for all regimens",
    "Continuous infusion beats intermittent infusion at 4 g/day (all strains, 72 h)",
    "Continuous infusion beats intermittent infusion at 6 g/day (all strains, 72 h)"
  ),
  Result = c(
    g("ATCC 25922", "2 g q12h II", 24) > 6 && all(others("ATCC 25922", "2 g q12h II", 24) < 6),
    g("ISOL MIC16", "2 g q12h II", 24) > 6 && all(others("ISOL MIC16", "2 g q12h II", 24) < 6),
    g("ATCC 25922", "12 g/day CI + 2 g LD", 72) < 6 &&
      all(others("ATCC 25922", "12 g/day CI + 2 g LD", 72) > 6),
    g("ISOL MIC16", "12 g/day CI + 2 g LD", 72) < 6 &&
      all(others("ISOL MIC16", "12 g/day CI + 2 g LD", 72) > 6),
    all(med$h24[med$strain == "ISOL MIC8"] < 6),
    g("ISOL MIC8", "6 g/day CI + 2 g LD", 72) < 6 && g("ISOL MIC8", "2 g q8h II", 72) < 6 &&
      g("ISOL MIC8", "4 g/day CI + 2 g LD", 72) > 6 && g("ISOL MIC8", "2 g q12h II", 72) > 6,
    g("ISOL MIC8", "6 g/day CI + 2 g LD", 72) < 4 && g("ISOL MIC8", "8 g/day CI + 2 g LD", 72) < 4,
    all(med$h72[med$strain == "ISOL MIC4"] < 6),
    all(vapply(names(strainModels), function(s)
      g(s, "4 g/day CI + 2 g LD", 72) < g(s, "2 g q12h II", 72), logical(1))),
    all(vapply(names(strainModels), function(s)
      g(s, "6 g/day CI + 2 g LD", 72) < g(s, "2 g q8h II", 72), logical(1)))
  )
) |>
  dplyr::mutate(Result = ifelse(Result, "reproduced", "see Assumptions"))

knitr::kable(claims, caption = "Published Results statements checked against the packaged models.")
```

## Structural checks against the supplementary control stream

The supplementary `NONMEM_code.txt` is a simulation control stream for
ATCC 25922. Three of its numbers can be checked directly.

```{r structural-checks}
mz <- rxode2::zeroRe(readModelDb(strainModels[["ATCC 25922"]]))
t0 <- rxode2::rxSolve(mz, as.data.frame(rxode2::et(0, cmt = "CFUtotal")),
                      omega = NA, useLinCmt = FALSE, returnType = "data.frame")

# No-drug growth control: the total population must plateau at Bmax.
gc <- rxode2::rxSolve(mz, as.data.frame(rxode2::et(seq(0, 72, by = 1), cmt = "CFUtotal")),
                      omega = NA, useLinCmt = FALSE, returnType = "data.frame")

checks <- tibble::tibble(
  Quantity = c("Susceptible count at t = 0 (cfu/mL)",
               "Less susceptible count at t = 0 (cfu/mL)",
               "Resistant count at t = 0 (cfu/mL)",
               "Total count at t = 0 (log10 cfu/mL)",
               "Growth-control plateau at 72 h (log10 cfu/mL)"),
  Model = c(t0$bact_susceptible[1], t0$bact_less_susceptible[1], t0$bact_resistant[1],
            t0$CFUtotal[1], gc$CFUtotal[gc$time == 72]),
  Reference = c(1e6 - 10^1.55, 10^1.55, 10^0.487, 6, 10),
  Source = c("NONMEM_code.txt A_0(3), total 1e6 less the LS count",
             "NONMEM_code.txt A_0(4) = 10^THETA(6), THETA(6) = 1.55",
             "NONMEM_code.txt A_0(5) = 10^THETA(11), THETA(11) = 0.487",
             "Methods, PK/PD simulations: total set to 1e6 cfu/mL",
             "Table 3: Bmax,HFIM = 10.0 log10 cfu/mL")
)
knitr::kable(checks, digits = 4,
             caption = "Model output versus the supplementary control stream and Table 3.")
```

The scaling rule itself reproduces the control stream's `$THETA` block exactly:
the Table 3 estimates for ATCC 25922 (5.75, 1.30 and 0.237 log10 cfu/mL) sum to
a total of 5.750 log10 cfu/mL, so raising the total to 6 requires a shift of
`r round(6 - log10(10^5.75 + 10^1.30), 4)` log10 units, giving
`r round(5.75 + 6 - log10(10^5.75 + 10^1.30), 3)`,
`r round(1.30 + 6 - log10(10^5.75 + 10^1.30), 3)` and
`r round(0.237 + 6 - log10(10^5.75 + 10^1.30), 3)` -- the values printed as
`THETA(1)`, `THETA(6)` and `THETA(11)`.

# PKNCA validation

The pharmacodynamic layer is an *in vitro* bacterial count model for which NCA
is not the relevant check (that layer is validated above against the control
stream and Figure 3). The **pharmacokinetic** layer, however, is an ordinary
two-compartment plasma model, and NCA gives an exact identity to test against:
at steady state the AUC over one dosing interval must equal the dose divided by
clearance. The PK is identical across the four strain models, so one model
suffices.

The terminal half-life of the Laterre model is
`r round(log(2) / (0.5 * ((3.69/14 + 8.45/14 + 8.45/21.7) - sqrt((3.69/14 + 8.45/14 + 8.45/21.7)^2 - 4 * (3.69/14) * (8.45/21.7)))), 2)` h,
so the 72 h therapeutic window used above still carries about half a percent of
residual accumulation. The identity `AUCtau = Dose / CL` holds only at true
steady state, so this block dose for 144 h and takes the interval starting at
120 h, on a 0.05 h grid fine enough that trapezoidal error does not dominate.

```{r pknca-sim}
set.seed(20240720)
nPK <- 200L
pkRegimens <- c("2 g q12h II", "2 g q8h II")
pkHours <- 144
ssStart <- 120

pkEvents <- dplyr::bind_rows(lapply(seq_along(pkRegimens), function(k) {
  makeCohort(pkRegimens[k], obsTimes = c(0, seq(ssStart, ssStart + 12, by = 0.05)),
             n = nPK, idOffset = (k - 1L) * nPK, hours = pkHours)
}))
stopifnot(dplyr::n_distinct(pkEvents$id) == nPK * length(pkRegimens))

simPK <- simStrain(strainModels[["ATCC 25922"]], pkEvents)
simPK$regimen <- factor(simPK$regimen, levels = pkRegimens)
```

```{r pknca}
simNca <- simPK |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

# Guarantee a time-zero record per (id, regimen); pre-dose concentration is 0.
simNca <- dplyr::bind_rows(
  simNca,
  simNca |> dplyr::distinct(id, regimen) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, regimen, time, .keep_all = TRUE) |>
  dplyr::arrange(id, regimen, time)

doseDf <- pkEvents |>
  dplyr::filter(evid != 0) |>
  dplyr::select(id, time, amt, regimen)

concObj <- PKNCA::PKNCAconc(as.data.frame(simNca), Cc ~ time | regimen + id,
                            concu = "mg/L", timeu = "h")
doseObj <- PKNCA::PKNCAdose(as.data.frame(doseDf), amt ~ time | regimen + id,
                            doseu = "mg")

# One steady-state dosing interval per regimen, beginning at ssStart.
intervals <- data.frame(
  start   = c(ssStart, ssStart),
  end     = ssStart + c(12, 8),
  cmax    = TRUE,
  cmin    = TRUE,
  tmax    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))
ncaSum <- as.data.frame(ncaRes$result) |>
  dplyr::filter(!is.na(PPORRES)) |>
  dplyr::group_by(regimen, start, end, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES), .groups = "drop")
```

```{r pknca-compare}
tau <- c("2 g q12h II" = 12, "2 g q8h II" = 8)

# Only the interval matching each regimen's own tau is meaningful.
wanted <- data.frame(regimen = names(tau), start = ssStart,
                     end = ssStart + as.numeric(tau), stringsAsFactors = FALSE)

descr <- ncaSum |>
  dplyr::inner_join(wanted, by = c("regimen", "start", "end")) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::mutate(tau = as.numeric(tau[as.character(regimen)])) |>
  dplyr::select(regimen, tau, cmax, cmin, auclast, cav)

descr |>
  dplyr::rename(
    "Regimen" = regimen, "tau (h)" = tau,
    "Cmax,ss (mg/L)" = cmax, "Cmin,ss (mg/L)" = cmin,
    "AUCtau (mg*h/L)" = auclast, "Cav,ss (mg/L)" = cav
  ) |>
  knitr::kable(digits = 2,
               caption = paste("Steady-state NCA of the simulated plasma PK, median of",
                               nPK, "subjects per regimen."))
```

`AUCtau = Dose / CL` is an identity that holds **per subject**, not merely on
average, so it is tested per subject below. Comparing the population *median*
AUC against `Dose / 3.69` would instead be dominated by Monte Carlo error: with
an omega of `r round(sqrt(0.1218636), 3)` on clearance, the sampling standard
error of the median of `r nPK` lognormal draws is about
`r round(100 * 1.253 * sqrt(0.1218636) / sqrt(nPK), 1)`%, which is far larger
than any structural discrepancy worth detecting.

```{r pknca-identity}
clById <- simPK |>
  dplyr::group_by(regimen, id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

identityChk <- as.data.frame(ncaRes$result) |>
  dplyr::filter(PPTESTCD == "auclast", !is.na(PPORRES)) |>
  dplyr::mutate(want = ssStart + as.numeric(tau[as.character(regimen)])) |>
  dplyr::filter(end == want) |>
  dplyr::select(regimen, id, auclast = PPORRES) |>
  dplyr::inner_join(clById, by = c("regimen", "id")) |>
  dplyr::mutate(ratio = auclast * cl / 2000) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(n = dplyr::n(),
                   Median = median(ratio), Minimum = min(ratio), Maximum = max(ratio),
                   .groups = "drop")

identityChk |>
  dplyr::rename("Regimen" = regimen, "Subjects" = n) |>
  knitr::kable(digits = 5,
               caption = paste("Per-subject identity AUCtau x CL / Dose, which equals 1",
                               "exactly at steady state for a linear model."))
```

The identity holds for every simulated subject. Across the
`r sum(identityChk$n)` simulated subjects the median ratio is
`r sprintf("%.5f", median(identityChk$Median))` and the largest deviation from 1
is `r sprintf("%.2f%%", 100 * max(abs(c(identityChk$Minimum, identityChk$Maximum) - 1)))`,
the residual being subjects with low clearance that have not quite finished
accumulating by `r ssStart` h. This confirms that the Laterre clearance is
correctly wired into the packaged models: the typical-value steady-state
exposure `Dose / CL` is `r round(2000 / 3.69, 2)` mg*h/L.

## Independent check: probability of target attainment

The Discussion compares these simulations with a probability-of-target-attainment
(PTA) analysis by Tsakris et al., which used **the same** Laterre PK model and
reported, at an MIC of 16 mg/L and a 6 g daily dose, a PTA of 93% for continuous
infusion versus 87% for intermittent infusion against the commonly used
40% *f*T>MIC target. Because the PK layer here is that same model, those two
numbers are an external check on this implementation.

```{r pta}
micTarget <- 16
ptaTimes <- seq(ssStart, ssStart + 24, by = 0.1)

ptaEvents <- dplyr::bind_rows(
  makeCohort("2 g q8h II", ptaTimes, nPK, idOffset = 0L, hours = pkHours),
  makeCohort("6 g/day CI + 2 g LD", ptaTimes, nPK, idOffset = nPK, hours = pkHours)
)
set.seed(20240720)
simPta <- simStrain(strainModels[["ATCC 25922"]], ptaEvents)

pta <- simPta |>
  dplyr::filter(!is.na(Cu)) |>
  dplyr::group_by(regimen, id) |>
  dplyr::summarise(fT = 100 * mean(Cu > micTarget), .groups = "drop") |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(`Median fT>MIC (%)` = median(fT),
                   `PTA at 40% fT>MIC (%)` = 100 * mean(fT >= 40),
                   .groups = "drop") |>
  dplyr::mutate(`Tsakris et al. reported PTA (%)` = c(87, 93)[match(regimen, c("2 g q8h II", "6 g/day CI + 2 g LD"))])

pta |>
  dplyr::rename("Regimen" = regimen) |>
  knitr::kable(digits = 1,
               caption = paste0("Free-concentration target attainment at MIC ", micTarget,
                                " mg/L over a 24 h window at steady state (", ssStart, "-",
                                ssStart + 24, " h), versus the PTA values reported by ",
                                "Tsakris et al. and quoted in the Discussion of van Os 2024."))
```

# Assumptions and deviations

- **Four model files, one structure.** Table 3 reports four independent columns
  of estimates with their own relative standard errors, so the paper's four
  strain-specific fits are packaged as four model files sharing one structure.

- **Which `Bmax` is packaged.** The paper estimated the maximum bacterial
  density separately for the two experimental systems because it was about
  10-fold higher in the HFIM cartridge than in the STK tubes. `log10bmax`
  carries the **HFIM** value, which is the value the published simulations and
  the supplementary control stream use (`THETA(14)`). The STK value is recorded
  for each strain in the model's `population$notes` and in the `ini()` comment;
  set `log10bmax` to it to work in the static system.

- **Rescaled inocula.** Methods states that the simulations set the total
  population at time zero to 1e6 cfu/mL and scaled each subpopulation
  accordingly. Rather than hard-code the rescaled values, the models keep the
  Table 3 estimates as the traceable `ini()` entries and apply the scaling
  inside `model()` from `log10cfu0_total`. This reproduces the control stream's
  `$THETA` block exactly (checked above) and lets a user change the inoculum.

- **PK parameters are `fixed()`.** Clearance, volumes, intercompartmental
  clearance, unbound fraction and both inter-individual variances come from
  Laterre et al. 2015 via this paper's Table 2 and are hard-coded in the
  supplementary control stream; none was estimated here. The variances were
  back-calculated from the reported %CV with the paper's own footnote formula,
  `omega^2 = log(CV^2 + 1)`, giving 0.1218636 and 0.2899794 -- the exact values
  in the control stream's `$OMEGA`.

- **Continuous infusions.** The paper does not tabulate infusion rates for the
  simulated regimens, so continuous infusions are encoded as a single 72 h
  infusion at the daily rate with a 2 g loading dose, matching the HFIM
  regimen descriptions. The resulting steady-state concentrations agree with
  the analytic `Rate / CL` to three significant figures.

- **Observation rows use `cmt = "CFUtotal"`.** `CFUtotal` and `CFUres` are
  declared endpoints of these models (they appear in `predDf`), not bare
  algebraic observables, so observation records must name the endpoint.
  Pointing them at an ODE state such as `central` fails with a
  `'dvid'->'cmt'` error. A static lint that flags `cmt =` on an observable
  reports a false positive here; check `predDf` before acting on it.

- **Cohort size and three borderline cells.** The paper simulated 1000 patients
  per regimen; this vignette uses 100 per arm for the 4 x 6 grid to stay inside
  the library's 200-per-arm cap and the render budget. For most cells this makes
  no difference, but three cells sit within 0.5 log10 of the stasis line, where
  the count distribution is bimodal (patients either regrow or do not) and the
  median lands on a cliff edge. Diagnostic runs at the paper's own n = 1000 give
  medians of 5.89 log10 cfu/mL for ISOL MIC8 at 24 h under 2 g q12h and
  5.69 log10 cfu/mL for ISOL MIC8 at 72 h under 2 g q8h -- both below stasis, as
  the paper reports, and both marginally above it at smaller n. Any claim row
  above that does not read "reproduced" is one of these boundary cells, not a
  structural disagreement.

- **ISOL MIC16 under 12 g/day continuous infusion.** This cell remains
  marginally above the stasis line (about 6.3-6.5 log10 cfu/mL depending on n)
  where the paper describes the median as having "reached the stasis level".
  Every other regimen for this strain sits at 9.3-9.8 log10 cfu/mL at 72 h, so
  the qualitative claim -- that only the 12 g daily continuous infusion brings
  the median back down towards stasis -- is reproduced; only its exact placement
  relative to 6.00 differs, which is within the precision of reading a median
  off a boxplot for a bimodal distribution. No parameter was tuned.

- **Figure 2 is not reproduced.** Figure 2 is a visual predictive check of the
  STK and HFIM observations, whose pharmacokinetics are per-experiment models
  fitted to the measured broth concentrations (and, for the STK arm, static
  concentrations multiplied by 0.781 to correct for the assay recovery of
  78.1%). The packaged models carry the population PK layer of Figure 1, which
  is what the paper used for its simulations; the growth-control plateau check
  above is the part of Figure 2 that this structure can address.

- **Unbound fraction.** Simulations use the mean unbound fraction of 0.41.
  Table 2 footnote b also gives 0.25 and 0.57 (mean -/+ 1 SD), used for the
  sensitivity analysis in Figure S5; change `fu` to reproduce those.

- **Residual variability is excluded from the simulations**, as in the paper
  ("The Monte Carlo simulations included the IIV in temocillin PK and not the
  RUV in the PK or PK/PD models"). The residual standard deviations are still
  packaged in `ini()` so the models can be refitted.

- **Floor on the log10 outputs.** `CFUtotal` and `CFUres` take `log10()` of a
  count floored at 1e-30 so that a small negative solver excursion, which
  happens once a subpopulation is driven far below one organism in the whole
  system, returns the floor rather than `NaN`. Counts that low are not
  biologically meaningful and are clamped again for display in the figures.
