---
title: "vanRongen_2018_midazolam"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vanRongen_2018_midazolam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("vanRongen_2018_midazolam"))$reference`
* Description: `r rxode2::rxode2(readModelDb("vanRongen_2018_midazolam"))$description`
* Article: <https://doi.org/10.1007/s40262-017-0579-4>
* PubMed: [PMID 28785981](https://pubmed.ncbi.nlm.nih.gov/28785981/)
* DDMORE Foundation Model Repository entry:
  [DDMODEL00000250](https://repository.ddmore.eu/model/DDMODEL00000250)

This model was extracted from the DDMORE Foundation Model Repository
bundle for `DDMODEL00000250` (scraped to
`dpastoor/ddmore_scraping/250/`). The bundle contains:

- `Executable_FinalModelCode.mod` -- NONMEM control stream for the
  final published model (the canonical run; see `Command.txt`).
- `Executable_AccessWeightModelCode.mod` -- companion control stream
  for an "access weight" sensitivity scenario in which the adolescent
  CL covariate is reparameterised as
  `THETA(1) * (WTAL/70)^0.75 + THETA(8) * WTAC` (allometric on
  weight-for-age-and-length plus a linear access-weight term, with
  the allometric exponent fixed at 0.75). This vignette and the
  packaged model implement the **final** model only; the access-weight
  scenario is not separately extracted (see Errata).
- `Output_real_FinalModelCode.lst` -- NONMEM listing of a re-fit on
  the bundle's `Simulated_DatafileMidaObesity.csv`. The
  `MINIMIZATION SUCCESSFUL` block (line 353; `OBJV = 524.451`,
  line 407) is the source of the final parameter estimates the
  packaged model uses.
- `Output_simulated_FinalModelCode.lst` -- companion `$SIMULATION`
  listing on the same simulated dataset.
- `Simulated_DatafileMidaObesity.csv` -- the simulated dataset (9
  subjects: ID 1-4 = adults, ID 31-50 = adolescents; PO + IV
  midazolam dosing).
- `DDMODEL00000250.rdf` -- purpose / research-stage / modelling-
  question metadata.
- `250.json`, `Command.txt` -- scraper / run metadata.

The original publication is **not on disk** in this worktree;
publication-level demographics, NCA tables, and figures are taken
from the PubMed abstract (PMID 28785981) only. Numerical values used
by the packaged model are the `Output_real_FinalModelCode.lst` final
estimates (i.e., from the bundle's re-fit on the simulated dataset),
not directly the publication's typical values. See **Assumptions and
deviations** below for the bundle-vs-publication comparison.

# Population

van Rongen 2018 reports a combined population PK analysis of
midazolam (a probe substrate of CYP3A) in 19 obese adolescents
(median total body weight 102.7 kg; range 62-149.5 kg) and 20
morbidly obese adults (median 144 kg; range 112-186 kg). The study's
goal was to disentangle the contributions of age, obesity, and
inflammation to the suppression of CYP3A-mediated clearance.
Subjects received both oral and intravenous midazolam dosing
(intravenous administration is captured in the bundle's simulated
dataset as a near-instantaneous infusion: AMT = 5000 microgram,
RATE = 29997 microgram/min ~= 10 s).

Numeric age range, sex distribution, and per-region breakdown are
not in the abstract and are not reproduced in the bundle, so those
fields of the model's `population` metadata are intentionally
`NA_character_` / `NA_real_`. Consumers needing those details
should consult the publication directly (DOI in the model's
`reference`).

The same information is available programmatically via the model's
`population` metadata
(`readModelDb("vanRongen_2018_midazolam")$population` after the
model is loaded).

# Source trace

Per-parameter origin (also recorded as in-file comments next to each
`ini()` entry of `inst/modeldb/ddmore/vanRongen_2018_midazolam.R`):

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl`             | log(0.540)              | `Output_real_FinalModelCode.lst` FINAL THETA TH 1 (CL adults; .mod $THETA initial 0.442) |
| `lvc`             | log(57.3)               | FINAL THETA TH 3 (V2; .mod initial 55.2 L) |
| `lq`              | log(1.27)               | FINAL THETA TH 4 (Q; .mod initial 1.14 L/min) |
| `lvp`             | log(166)                | FINAL THETA TH 5 (V3 at WT = 141.8 kg; .mod initial 172 L) |
| `lka`             | log(0.111)              | FINAL THETA TH 6 (KA = KTR; .mod initial 0.115 / min) |
| `lfdepot`         | log(0.684)              | FINAL THETA TH 2 (F1; .mod initial 0.562) |
| `e_adolescent_cl` | log(0.793 / 0.540)      | derived from FINAL THETA TH 7 / TH 1 (adolescent CL at WT = 104.7 kg) |
| `e_wt_cl`         | 1.05                    | FINAL THETA TH 8 (TBW power on adolescent CL; .mod initial 1.2) |
| `e_wt_vp`         | 3.36                    | FINAL THETA TH 9 (TBW power on adult V3; .mod initial 3.28) |
| `etalcl`          | 4.33e-06                | FINAL OMEGA(1,1); near-boundary, ETA shrinkage 98.97 % (.lst lines 356-368) |
| `etalfdepot`      | 9.22e-02                | FINAL OMEGA(2,2) -- IIV on F1 (ETA(2)) |
| `etalq`           | 4.32e-01                | FINAL OMEGA(3,3) -- IIV on Q (ETA(3)) |
| `etalka`          | 8.61e-02                | FINAL OMEGA(4,4) -- IIV on KA (ETA(4)) |
| `etalvc`          | 1.86e-02                | FINAL OMEGA(5,5) -- IIV on V2 (ETA(5)) |
| `etalvp`          | 7.96e-02                | FINAL OMEGA(6,6) -- IIV on V3 (ETA(6)) |
| `propSd`          | sqrt(0.0988) ~= 0.3143    | FINAL SIGMA(1,1) -- Y = F * (1 + EPS(1)) |
| `d/dt(depot)`            | n/a | .mod $DES line 52: `DADT(1) = -K14*A(1)` (K14 = KA) |
| `d/dt(transit1..5)`      | n/a | .mod $DES lines 55-59: 5-link `KTR` chain (KTR = KA) |
| `d/dt(central)`          | n/a | .mod $DES line 53: `DADT(2) = KTR*A(8) - K23*A(2) + K32*A(3) - K20*A(2)` |
| `d/dt(peripheral1)`      | n/a | .mod $DES line 54: `DADT(3) = K23*A(2) - K32*A(3)` |
| `f(depot) = exp(lfdepot + etalfdepot)` | n/a | .mod $PK line 29: `F1 = THETA(2) * EXP(ETA(2))` |
| `Cc = central / vc`      | n/a | .mod $PK line 38 + $ERROR: `S2 = V2`, `IPRED = F` |
| `ADOLESCENT switch on CL` | n/a | .mod $PK lines 26-27 (`IF (ID.LE.30) TVCL=THETA(1); IF (ID.GT.30) TVCL=THETA(7)*(TBW/104.7)**THETA(8)`) |
| `ADOLESCENT switch on Vp` | n/a | .mod $PK lines 32-33 (`IF (ID.LE.30) TVV3=THETA(5)*(TBW/141.8)**THETA(9); IF (ID.GT.30) TVV3=THETA(5)`) |
| `Cc ~ prop(propSd)`      | n/a | .mod $ERROR line 63: `Y = F * (1 + ERR(1))` (proportional in linear space) |

# Virtual cohort

Original observed data are not publicly available. The figures below
use a virtual cohort whose total-body-weight distributions
approximate the published trial subgroups: 50 morbidly obese adults
(WT centred at 144 kg, range 112-186 kg) and 50 obese adolescents
(WT centred at 102.7 kg, range 62-149.5 kg), each receiving a
single 7500 microgram oral midazolam dose followed approximately
2.5 hours later by a 5000 microgram IV bolus. (The bundle's
`Simulated_DatafileMidaObesity.csv` uses an analogous PO + IV
sequence per subject.)

```{r cohort}
set.seed(20260506L)

n_per_group <- 50L
po_dose_ug  <- 7500    # microgram, oral
iv_dose_ug  <- 5000    # microgram, IV bolus
iv_rate_ug  <- 29997   # microgram/min (matches bundle simulated dataset)
po_time_min <- 0
iv_time_min <- 162.9   # bundle's adult ID 1 second-dose time

obs_times <- c(0,
               seq(5, 60, by = 5),
               seq(75, 240, by = 15),
               seq(270, 720, by = 30))

make_cohort <- function(group_label, adolescent_flag, wt_values, id_offset = 0L) {
  n <- length(wt_values)
  ids <- id_offset + seq_len(n)

  dose_po <- tibble::tibble(
    id    = ids,
    time  = po_time_min,
    amt   = po_dose_ug,
    rate  = 0,
    evid  = 1L,
    cmt   = "depot",
    WT    = wt_values,
    ADOLESCENT = adolescent_flag,
    treatment  = group_label
  )

  dose_iv <- tibble::tibble(
    id    = ids,
    time  = iv_time_min,
    amt   = iv_dose_ug,
    rate  = iv_rate_ug,
    evid  = 1L,
    cmt   = "central",
    WT    = wt_values,
    ADOLESCENT = adolescent_flag,
    treatment  = group_label
  )

  obs <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(
      amt   = 0,
      rate  = 0,
      evid  = 0L,
      cmt   = "central"
    ) |>
    dplyr::left_join(
      dplyr::select(dose_po, id, WT, ADOLESCENT, treatment),
      by = "id"
    )

  dplyr::bind_rows(dose_po, dose_iv, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# Adults: WT centred at 144 kg, truncated to the published range
wt_adults <- pmin(pmax(round(rnorm(n_per_group, mean = 144, sd = 18)), 112), 186)
# Adolescents: WT centred at 102.7 kg, truncated to the published range
wt_adols  <- pmin(pmax(round(rnorm(n_per_group, mean = 102.7, sd = 21)), 62), 150)

events <- dplyr::bind_rows(
  make_cohort("morbidly obese adult", 0L, wt_adults, id_offset =   0L),
  make_cohort("obese adolescent",     1L, wt_adols,  id_offset = 100L)
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- rxode2::rxode2(readModelDb("vanRongen_2018_midazolam"))

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "WT", "ADOLESCENT")
) |>
  as.data.frame()
```

For the typical-value trajectory used in the figures below, zero out
the random effects so the prediction is deterministic per subject:

```{r simulate-typical}
mod_typical <- mod |> rxode2::zeroRe()
sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events,
  keep   = c("treatment", "WT", "ADOLESCENT")
) |>
  as.data.frame()
```

# Replicate published figures

The publication's main quantitative result is that obese adolescents
have a significantly higher CYP3A clearance than morbidly obese
adults (van Rongen 2018, abstract: 0.71 vs 0.44 L/min). The typical
concentration-time profile of the packaged model after a 7.5 mg
oral dose plus a 5 mg IV bolus reproduces this ordering: at any
given time, the morbidly-obese-adult concentration is higher than
the matched obese-adolescent concentration (lower CL -> higher
exposure).

```{r figure-conc-vs-time, fig.width = 7, fig.height = 4.5}
sim_typical |>
  dplyr::filter(time > 0, time <= 720) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Cc_median = stats::median(Cc, na.rm = TRUE),
    Cc_p10    = stats::quantile(Cc, 0.10, na.rm = TRUE),
    Cc_p90    = stats::quantile(Cc, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Cc_median, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Cc_p10, ymax = Cc_p90),
              alpha = 0.20, colour = NA) +
  geom_line() +
  scale_y_log10() +
  labs(
    x = "Time after first oral dose (min)",
    y = "Cc (microgram/L)",
    colour = NULL, fill = NULL,
    title = paste0(
      "Typical-value midazolam Cc (zero IIV / EPS) -- ",
      "10th, 50th, 90th percentiles across the WT distribution"
    ),
    caption = paste0(
      "7500 microgram oral dose at t = 0; ",
      "5000 microgram IV bolus at t = 162.9 min ",
      "(matches the bundle's Simulated_DatafileMidaObesity.csv)."
    )
  ) +
  theme(legend.position = "bottom")
```

```{r figure-cl-vs-wt, fig.width = 7, fig.height = 4}
typical_cl <- sim_typical |>
  dplyr::distinct(id, treatment, ADOLESCENT, WT) |>
  dplyr::mutate(
    cl_typical = ifelse(
      ADOLESCENT == 1,
      0.793 * (WT / 104.7)^1.05,
      0.540
    )
  )

ggplot(typical_cl, aes(WT, cl_typical, colour = treatment)) +
  geom_point() +
  geom_hline(
    yintercept = c(0.44, 0.71),
    linetype   = "dashed"
  ) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    x = "Total body weight (kg)",
    y = "Typical CL (L/min)",
    colour = NULL,
    title = "Typical CL vs WT -- bundle .lst final estimates",
    caption = paste0(
      "Solid points: model-implied typical CL using lst final estimates ",
      "(0.540 / 0.793 L/min at WT = 141.8 / 104.7). ",
      "Dashed lines: van Rongen 2018 abstract typical values ",
      "(0.44 adult, 0.71 adolescent -- match .mod $THETA initial values)."
    )
  ) +
  theme(legend.position = "bottom")
```

# PKNCA validation

PKNCA Cmax / AUCinf / half-life by treatment group, computed from
the stochastic simulation. The single oral dose at `t = 0` is the
NCA reference dose; the later IV bolus is excluded from the NCA
window via `end = iv_time_min - 1` so the oral exposure is not
contaminated.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time >= 0, time < iv_time_min) |>
  dplyr::transmute(id, time, Cc, treatment, WT, ADOLESCENT)

dose_df <- events |>
  dplyr::filter(evid == 1, time == po_time_min) |>
  dplyr::transmute(id, time, amt, treatment, WT, ADOLESCENT)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "microgram/L",
                             timeu = "min")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             route = "extravascular",
                             doseu = "microgram")

intervals <- data.frame(
  start      = 0,
  end        = iv_time_min - 1,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)

nca_summary <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    median = stats::median(PPORRES, na.rm = TRUE),
    p05    = stats::quantile(PPORRES, 0.05, na.rm = TRUE),
    p95    = stats::quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

knitr::kable(
  nca_summary,
  caption = "Simulated NCA parameters by treatment group (single 7500 microgram oral midazolam; PKNCA over the pre-IV window)."
)
```

## Comparison against published clearance

The publication's headline result is the difference in typical CL
between the two subgroups (van Rongen 2018 abstract): obese
adolescents 0.71 L/min vs morbidly obese adults 0.44 L/min. The
bundle's `Output_real_FinalModelCode.lst` final estimates are
0.793 L/min (adolescents) and 0.540 L/min (adults) -- i.e., 12 % and
22 % higher than the published typical values. The simulation-based
NCA-implied CL (oral CL/F = Dose x F / AUCinf) should match the
bundle's typical values to within Monte-Carlo and NCA-extrapolation
noise, **not** the publication's typical values, because the
simulation is driven by the bundle's final estimates.

```{r compare-cl}
po_dose_ug_const <- po_dose_ug
F_typical        <- 0.684

cl_compare <- nca_summary |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::mutate(
    cl_oral_implied_L_per_min =
      po_dose_ug_const * F_typical / median   # microgram / (microgram*min/L) = L/min
  ) |>
  dplyr::transmute(
    treatment,
    auc_inf_med_ug_min_per_L = round(median, 0),
    cl_oral_implied_L_per_min = round(cl_oral_implied_L_per_min, 3)
  ) |>
  dplyr::mutate(
    bundle_typical_CL_L_per_min = ifelse(
      grepl("adolescent", treatment, ignore.case = TRUE),
      0.793, 0.540
    ),
    publication_typical_CL_L_per_min = ifelse(
      grepl("adolescent", treatment, ignore.case = TRUE),
      0.71, 0.44
    )
  )

knitr::kable(
  cl_compare,
  caption = "Simulated AUCinf, NCA-implied oral CL, the bundle .lst typical CL, and the publication's typical CL by treatment group."
)
```

# Assumptions and deviations

- **Final estimates are from the bundle's re-fit on a simulated
  dataset, not directly from the publication.** The
  `Output_real_FinalModelCode.lst` listing was produced by re-running
  the .mod against the shipped `Simulated_DatafileMidaObesity.csv`
  (only 9 subjects, 138 records). The final estimates therefore
  differ slightly from the published typical values: CL adults
  0.540 vs 0.44 L/min (+22 %), CL adolescents 0.793 vs 0.71 L/min
  (+12 %), F1 0.684 vs 0.562 (+22 %). The .mod's `$THETA` initial
  values match the publication exactly, but per the
  `extract-literature-model` skill the packaged model uses the
  `.lst` final estimates as the canonical numbers; the publication
  values are documented here for reference.
- **`OMEGA(1,1)` (IIV on CL) sits at the lower boundary in the
  re-fit.** The .lst reports 4.33e-06 with a 98.97 % ETA shrinkage
  and the warning `PARAMETER ESTIMATE IS NEAR ITS BOUNDARY. THIS
  MUST BE ADDRESSED BEFORE THE COVARIANCE STEP CAN BE IMPLEMENTED`
  (lines 356-358). The eta on CL is retained in the packaged model
  for structural fidelity, but it contributes negligibly to
  simulation variability -- practical IIV on CL in this re-fit is
  effectively zero. A re-fit on the original (un-simulated)
  39-subject dataset would likely return a non-zero IIV on CL.
- **The "AccessWeightModelCode" companion scenario is not
  separately extracted.** The bundle ships
  `Executable_AccessWeightModelCode.mod`, an alternative
  parameterisation in which the adolescent CL covariate is split
  into an allometric `(WTAL/70)^0.75` term plus a linear
  access-weight `WTAC` term (allometric exponent fixed at 0.75).
  This sensitivity model is not the canonical run per
  `Command.txt` (`Executable model = Executable_FinalModelCode.mod`)
  and is omitted from `inst/modeldb/ddmore/`. If the access-weight
  parameterisation is needed for downstream analysis, extract it
  as a separate `vanRongen_2018a_midazolam.R` per the year-letter
  collision rule in `naming-conventions.md`.
- **Source dataset uses ID-coded sub-population, not a labelled
  ADOLESCENT column.** The .mod uses `IF (ID.LE.30) ... IF (ID.GT.30)`
  branches to select between the adult and adolescent CL / V3
  forms. The packaged model exposes this as a canonical
  `ADOLESCENT` (0/1) covariate; users must materialize the column
  themselves from age category. No implicit derivation from the ID
  range is performed.
- **Reference weights 104.7 kg (adolescents) and 141.8 kg
  (adults) come from the .mod source.** The publication abstract
  reports cohort medians of 102.7 kg (adolescents) and 144 kg
  (adults); the .mod's reference values differ by ~2 % and likely
  reflect a slightly different summary statistic (mean vs median or
  rounding) used at fitting time. The packaged model uses the
  .mod's values to remain mass-balance consistent with the
  published parameter estimates.
- **Original publication PDF is not on disk in this worktree.**
  The model's `description`, `reference`, and `population` fields
  are populated from the PubMed abstract (PMID 28785981) plus the
  DDMORE bundle metadata. A side-by-side comparison against the
  full publication's parameter table or a published NCA summary is
  not part of this vignette's scope; the validation here is the F.2
  bundle-self-consistency check plus a comparison against the
  abstract's two typical-CL numbers.
- **Population demographic detail is intentionally `NA`** for the
  fields the abstract does not report (numeric age range, sex
  distribution, regions). Consumers needing those details should
  consult the publication directly.
- **Single proportional residual error.** The .mod
  `$ERROR Y = F * (1 + ERR(1))` form is proportional in linear
  space; no additive component is fit. The packaged model uses
  `Cc ~ prop(propSd)` with `propSd = sqrt(SIGMA(1,1))` (linear-space
  SD), matching the source.
- **Bundle simulated dataset is a smoke-test cohort.** The 9
  subjects in `Simulated_DatafileMidaObesity.csv` (4 adults,
  5 adolescents) are a regression-test artifact, not a recreation
  of the 39-subject van Rongen 2018 trial. The virtual cohort built
  in this vignette mirrors the published WT distribution rather
  than the bundle's smoke-test set.
