---
title: "PFOS and PFOA gestational PBPK and birth weight (Verner 2015)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PFOS and PFOA gestational PBPK and birth weight (Verner 2015)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Verner et al. (2015, *Environ Health Perspect* 123:1317-1324,
[doi:10.1289/ehp.1408837](https://doi.org/10.1289/ehp.1408837), PMCID
PMC4671243) asked how much of the epidemiologic association between
prenatal perfluoroalkyl substance (PFAS) exposure and lower birth weight
is **confounding by glomerular filtration rate (GFR)** rather than a
toxic effect of the chemical.

The argument is mechanistic. PFOS and PFOA are cleared almost entirely by
renal filtration followed by extensive saturable tubular reabsorption. A
woman with a higher GFR therefore clears more PFAS and carries a lower
plasma level. GFR is *also* positively associated with birth weight. So
even with no causal effect of PFAS on fetal growth at all, a study that
regresses birth weight on maternal or cord PFAS will find a negative
coefficient. Verner et al. quantified that artefact by taking a published
gestational PBPK model, wiring GFR to birth weight through a
meta-analytic regression, and running it 250,000 times.

The paper contributes **two** models, extracted here as two files
because the two chemicals are parameterised independently (the acslX
listing selects between them with a `PFOS`/`PFOA` compound flag and
carries a separate constant for every chemical-specific quantity):

| Model file | Chemical |
| --- | --- |
| `Verner_2015_pfos_pbpk` | Perfluorooctane sulfonate (PFOS) |
| `Verner_2015_pfoa_pbpk` | Perfluorooctanoic acid (PFOA) |

Both share one structure, adapted from Loccisano et al. (2013) with the
placental blood flow and fetal cardiac output updated per Yoon et al.
(2011): flow-limited maternal plasma, liver, gut, fat, mammary tissue,
skin, kidney and rest-of-body compartments; a glomerular filtrate
compartment feeding a urinary storage compartment through
Michaelis-Menten renal reabsorption; a placenta; and a separate fetal
circulation (fetal plasma, rest of fetal body, amniotic fluid) reached
only by bidirectional placental diffusion.

Model time is **hours since conception**, and gestational week
`wk = time / 168` drives five lookup tables (renal plasma flow, GFR,
maternal plasma volume, haematocrit, amniotic fluid volume) plus the
fetal, placental, fat and mammary growth curves. There are no dose
records: chronic background exposure enters maternal plasma as a
constant hourly intake back-calculated from a pre-pregnancy plasma level
assumed to be at steady state.

```{r load-models}
pfos <- readModelDb("Verner_2015_pfos_pbpk")
pfoa <- readModelDb("Verner_2015_pfoa_pbpk")
models <- list(PFOS = pfos, PFOA = pfoa)

# Time base helpers. Delivery is 6,570 h = 39.1 weeks; the acslX listing
# records the cord sample at 6,569 h.
H_PER_WK <- 168
T_DELIVERY <- 6570
T_CORD <- 6569
MONTH_H <- seq(730, 6570, by = 730)
```

# Population

No subjects were fitted. This is a forward Monte Carlo simulation, so the
`population` metadata records `n_subjects = 0` and `n_studies = 0`. The
input distributions (Table 1) are assembled from published sources:

* **Pre-pregnancy body weight** - Norwegian Mother and Child Cohort Study
  (MoBa). Table 1 reports mean 70.3 kg, SD 14.3, range 37.0-134.0.
* **Standardized GFR (`GFRratio`)** - the ratio of a woman's observed GFR
  to the mean GFR at her gestational age, distributed as mean 1.000, SD
  0.246, range 0.508-1.492, from Morken et al. (2014), n = 953.
* **Initial plasma PFAS** - equal-weight average of the reported maternal
  or cord means/medians from the seven epidemiologic studies in the
  paper's own meta-analysis (USA, Denmark, Canada, Japan, Taiwan, Norway,
  Great Britain): 13.02 ng/mL for PFOS (CV 0.368) and 2.53 ng/mL for PFOA
  (CV 0.446).
* **Sensitive PBPK parameters** - liver volume fraction, liver:plasma and
  rest-of-body:plasma partition coefficients, free fraction in maternal
  and fetal plasma, maximum reabsorption velocity and affinity constant.
  Central values are from Loccisano et al. (2013); the SDs assume a 15%
  coefficient of variation with bounds at +/- 2 SD.

Model accuracy was assessed only at the population level, by visual
comparison of simulated maternal plasma trajectories against serial
measurements from Glynn et al. (2012) and Monroy et al. (2008)
(Figure 2). No individual-level PFAS kinetic data were fitted.

# Source trace

Everything in both model files is a recoded literature constant; nothing
was estimated. The authoritative source for structure, parameter values
and the gestational lookup tables is the deposited acslX listing in
Supplemental Material, "PBPK Model Code", together with the Monte Carlo
automation script printed after it.

| Model element | Source location |
| --- | --- |
| Compartment structure, all mass-balance ODEs | Suppl. acslX listing, `DERIVATIVE` block; Suppl. Figure S1; Figure 1 |
| Maternal blood flows `QCC`, `QLC`, `QSkC`, `QGC`, `QMamC`, `QFatC` | Suppl. acslX listing, "Blood flow calculation" |
| Maternal tissue volumes `VLC`, `VKC`, `VfilC`, `VGC`, `VMamC`, `VFatC`, `VSk` | Suppl. acslX listing, "Maternal tissue volumes" |
| Fetal `QFetC`, `QPlaC`, `VPlasFC`, `HtcF` | Suppl. acslX listing, "Fetal cardiac output", "Plasma volumes" |
| Fetal / placental / fat / mammary growth curves | Suppl. acslX listing, "Organ volume calculation" |
| Lookup tables `DQK`, `DQFil`, `DVPlasC`, `DHtc`, `VAFX` | Suppl. acslX listing, `TABLE` statements |
| Partition coefficients `PFOS_P*` / `PFOA_P*` | Suppl. acslX listing, "Chemical specific parameters"; Table 1 for the six sampled ones |
| Free fractions `PFOS_Free`, `PFOA_Free`, `*_FreeF` | Suppl. acslX listing; Table 1 |
| Reabsorption `PFOS_TMC`/`PFOA_TMC`, `PFOS_KT`/`PFOA_KT`, `kurinec` | Suppl. acslX listing; Table 1 |
| Placental / amniotic transfer `k1c`-`k4c` | Suppl. acslX listing, "Placental & amniotic fluid transfer" |
| Eq. 1, initial tissue burdens | Methods Eq. 1; Suppl. acslX listing, `APLAS0`..`ASK0` |
| Eq. 2, steady-state intake; half-lives 5.4 / 3.8 y | Methods Eq. 2; Suppl. acslX listing, `IVHOURLYDOSE`, `PFOS_HL`, `PFOA_HL` |
| Eq. 3, `GFR_t = GFRratio x reference GFR_t` | Methods Eq. 3; Suppl. acslX listing, `QFil = Ratio_GFR*QFil_R` |
| Eq. 4, birth weight = 3,376 + 175.5 x `GFRratio` + residual | Methods Eq. 4; Suppl. MC script, `CONSTANT_REG`, `BETA_REG` |
| Eq. 5, fetal growth rescaling by birth weight | Methods Eq. 5; Suppl. acslX listing, `VFet = (Birthweight/3.50857)*VFet_av` |
| Monte Carlo input distributions | Table 1; Suppl. MC script `normrnd(...)` calls |
| Published regression coefficients (Table 2) | Table 2, "1 (main results)" rows |

# Part 1 - Structural verification

These checks are exact identities, not statistical comparisons. Each one
can go red on its own: they compare an independently recomputed quantity
against what the solved ODE system actually produces.

## The five gestational lookup tables

The acslX `TABLE` statements are linearly interpolated with the end
values held outside the tabulated range. The model files encode each one
in hinge form. Re-deriving the same tables with `stats::approx(rule = 2)`
straight from the `TABLE` knots is an independent implementation, so
agreement to machine zero confirms the transcription of several hundred
numbers.

```{r lookup-tables}
wk_grid <- seq(0, 42, by = 0.25)
ref_solve <- rxSolve(
  pfos, et(wk_grid * H_PER_WK),
  params = data.frame(WT = 60.9, RENALFUNC_REL = 1),
  returnType = "data.frame"
)
wk <- ref_solve$time / H_PER_WK

acslx_tables <- list(
  # TABLE DQFil -- glomerular filtration rate (L/h)
  GFR = list(
    x = 0:39,
    y = c(6, 6, 6, 6, 6, 6.03, 6.09, 6.18, 6.32, 6.50, 6.72, 6.97, 7.23,
          7.50, 7.76, 8.01, 8.23, 8.43, 8.60, 8.75, 8.86, 8.95, 9.01, 9.05,
          9.06, 9.06, 9.04, 8.99, 8.95, 8.88, 8.81, 8.72, 8.62, 8.52, 8.41,
          8.29, 8.16, 8.03, 7.89, 7.75)),
  # TABLE DQK -- renal plasma flow (L/h)
  Qkidney = list(
    x = c(0, 8, 10, 11, 12, 13, 15, 17, 18, 19, 22, 23, 24, 25, 26, 27, 28,
          29, 31, 32, 33, 34, 36, 37, 38, 40),
    y = c(24.42, 50.58, 45.42, 51.36, 57.12, 46.32, 44.73, 45.36, 41.8,
          46.53, 46.08, 47.85, 43.05, 41.16, 47.52, 43.41, 48.57, 46.36,
          36.68, 39.44, 46.46, 44.13, 26.7, 33.92, 35.46, 40.26)),
  # TABLE DHtc -- maternal haematocrit
  Htcmat = list(
    x = c(0, 11, 16, 20, 24, 28, 32, 36, 40),
    y = c(0.38, 0.372, 0.356, 0.354, 0.349, 0.347, 0.346, 0.354, 0.367)),
  # TABLE VAFX -- amniotic fluid volume (L)
  Vamniotic = list(
    x = c(0, 8, 9, 10, 11, 12, 16, 18, 20, 22, 24, 26, 28, 30, 32, 34, 36,
          38, 40, 42),
    y = c(0.0237, 0.0235, 0.0234, 0.0232, 0.0766, 0.124, 0.195, 0.284,
          0.361, 0.646, 0.640, 0.687, 0.770, 0.824, 0.776, 0.817, 0.817,
          0.799, 0.530, 0.506))
)

table_dev <- vapply(names(acslx_tables), function(nm) {
  tb <- acslx_tables[[nm]]
  max(abs(ref_solve[[nm]] - approx(tb$x, tb$y, xout = wk, rule = 2)$y))
}, numeric(1))

# TABLE DVPlasC is a FRACTION of gestational body weight, so recover it by
# dividing the solved plasma volume by the solved maternal body weight.
dvplasc_x <- 9:36
dvplasc_y <- c(0.0442, 0.0445, 0.0448, 0.0451, 0.0454, 0.0456, 0.0458,
               0.0460, 0.0461, 0.0462, 0.0463, 0.0464, 0.0464, 0.0464,
               0.0464, 0.0463, 0.0463, 0.0461, 0.0460, 0.0458, 0.0456,
               0.0454, 0.0451, 0.0448, 0.0445, 0.0441, 0.0437, 0.0433)
table_dev["fv_plasma"] <- max(abs(
  ref_solve$Vplasma / ref_solve$BWmat -
    approx(dvplasc_x, dvplasc_y, xout = wk, rule = 2)$y))

knitr::kable(
  data.frame(`acslX TABLE` = c("DQFil", "DQK", "DHtc", "VAFX", "DVPlasC"),
             `Model variable` = names(table_dev),
             `Max abs deviation` = signif(table_dev, 3),
             check.names = FALSE),
  row.names = FALSE,
  caption = "Model lookup tables vs. stats::approx(rule = 2) on the acslX TABLE knots."
)

# Every table is piecewise-linear interpolation of the same knots, so the
# agreement is exact up to floating-point representation.
stopifnot(all(table_dev < 1e-12))
```

## Equations 1-5 and mass balance

```{r structural-identities}
# A deterministic GFRratio sweep across the full Table 1 range, with the
# birth-weight residual switched off so Eq. 4 is the pure regression line.
gfr_grid <- seq(0.508, 1.492, length.out = 25)
# NOTE: the observation grid must be de-duplicated. MONTH_H already ends at
# 6570, so `c(0, MONTH_H, T_DELIVERY)` would request delivery twice, giving
# two rows per subject at that time and silently recycling `gfr_grid`
# against a 50-row frame in the identity checks below.
sweep_times <- sort(unique(c(0, MONTH_H, T_DELIVERY)))
sweep <- rxSolve(
  pfos, et(sweep_times),
  params = data.frame(WT = 60.9, RENALFUNC_REL = gfr_grid, bw_resid = 0),
  returnType = "data.frame"
)
at_birth <- sweep[sweep$time == T_DELIVERY, ]
stopifnot(nrow(at_birth) == length(gfr_grid))

# Eq. 1 -- initial tissue burden = C_initial x partition x tissue volume.
# At t = 0 the plasma state holds the FREE amount in VPlasINIT, whereas the
# reported concentration divides by VPlas(0), which already carries the
# week-zero amniotic fluid and placenta. The listing does the same, so the
# ratio is a fixed geometric factor slightly below 1 rather than exactly 1.
c0_ratio <- ref_solve$Cc[ref_solve$time == 0] / 0.01302

# Eq. 2 -- hourly intake sustaining the pre-pregnancy level at steady state.
eq2_closed <- 0.01302 * ref_solve$Vdist[1] / (1.44 * 5.4 * 365 * 24)
eq2_relerr <- abs(ref_solve$Intake[1] - eq2_closed) / eq2_closed

# Eq. 3 -- GFRratio scales the WHOLE gestational GFR profile linearly.
# At delivery the reference table value is 7.75 L/h.
eq3_relerr <- max(abs(at_birth$GFR / (gfr_grid * 7.75) - 1))

# Eq. 4 -- birth weight (kg) = 3.376 + 0.1755 x GFRratio + residual.
eq4_abserr <- max(abs(at_birth$WTbirth - (3.376 + 0.1755 * gfr_grid)))

# Eq. 5 -- the fetal growth curve is rescaled so fetal weight at delivery
# EQUALS the calculated birth weight.
eq5_relerr <- max(abs(at_birth$WTfet / at_birth$WTbirth - 1))

# acslX BALANCE_DOSE: cumulative intake plus initial burden, minus every
# amount currently in the system, over the whole pregnancy.
massbal <- max(abs(sweep$MassBalance))

knitr::kable(
  data.frame(
    Check = c("Eq. 1 Cc(0) / C_initial (geometric, < 1 by design)",
              "Eq. 2 intake vs. closed form (rel. error)",
              "Eq. 3 GFR linearity in GFRratio (rel. error)",
              "Eq. 4 birth weight vs. regression (abs. error, kg)",
              "Eq. 5 fetal weight at delivery vs. birth weight (rel. error)",
              "Mass balance over gestation (abs., mg)"),
    Value = signif(c(c0_ratio, eq2_relerr, eq3_relerr, eq4_abserr,
                     eq5_relerr, massbal), 4)),
  caption = "Exact structural identities."
)

stopifnot(
  abs(c0_ratio - 0.999606) < 1e-5,
  eq2_relerr < 1e-10,
  eq3_relerr < 1e-10,
  eq4_abserr < 1e-10,
  eq5_relerr < 1e-5,
  # The ODE system conserves mass to solver tolerance.
  massbal < 1e-8
)
```

## The confounding mechanism, deterministically

With every other input held fixed, a higher `GFRratio` must give a
*lower* plasma PFAS level (more filtration) and a *higher* birth weight
(Eq. 4). That opposition is the entire confounding argument, and it is
deterministic - no Monte Carlo noise enters.

```{r mechanism}
stopifnot(
  # Strictly monotone across the full Table 1 GFRratio range.
  all(diff(at_birth$Cc) < 0),
  all(diff(at_birth$Cplasma_fet) < 0),
  all(diff(at_birth$WTbirth) > 0)
)

ggplot(
  data.frame(gfr = gfr_grid,
             maternal = at_birth$Cc * 1000,
             cord = at_birth$Cplasma_fet * 1000,
             bw = at_birth$WTbirth * 1000) |>
    pivot_longer(c(maternal, cord), names_to = "matrix", values_to = "conc"),
  aes(conc, bw, colour = matrix)
) +
  geom_path(linewidth = 1) +
  labs(x = "Simulated plasma PFOS at delivery (ng/mL)",
       y = "Birth weight (g)", colour = "Matrix",
       title = "Confounding by GFR, with no PFAS effect on growth",
       subtitle = "GFRratio swept 0.508-1.492; every other input fixed") +
  theme_bw()
```

# Part 2 - Gestational time course

PFAS levels decline over pregnancy as plasma volume expands and GFR
rises. This is the profile Verner et al. compared against the serial
measurements of Glynn et al. (2012) and Monroy et al. (2008) in Figure 2.

```{r time-course}
profile <- lapply(names(models), function(nm) {
  s <- rxSolve(models[[nm]], et(seq(0, T_DELIVERY, by = 73)),
               params = data.frame(WT = 60.9, RENALFUNC_REL = 1),
               returnType = "data.frame")
  data.frame(chemical = nm, wk = s$time / H_PER_WK,
             maternal = s$Cc * 1000, cord = s$Cplasma_fet * 1000)
}) |> bind_rows()

ggplot(profile |>
         pivot_longer(c(maternal, cord), names_to = "matrix", values_to = "conc"),
       aes(wk, conc, colour = matrix)) +
  geom_line(linewidth = 1) +
  facet_wrap(~chemical, scales = "free_y") +
  labs(x = "Gestational week", y = "Concentration (ng/mL)", colour = "Matrix",
       title = "Simulated PFAS over pregnancy at typical inputs",
       subtitle = "Compare Figure 2 of Verner 2015") +
  theme_bw()

decline <- profile |>
  group_by(chemical) |>
  summarise(start = maternal[wk == 0],
            delivery = maternal[which.max(wk)],
            pct_decline = 100 * (1 - delivery / start), .groups = "drop")
knitr::kable(decline |> mutate(across(where(is.numeric), \(x) signif(x, 3))),
             caption = "Maternal plasma decline from conception to delivery.")

stopifnot(
  # Both chemicals decline over pregnancy, by an amount consistent with the
  # serial measurements in Figure 2 (Glynn 2012 reports roughly 4.3 -> 4.0
  # ng/mL of PFOA between draws, and the paper notes the model slightly
  # underestimates that decline). Deterministic at typical inputs, so this
  # bound is exactly reproducible.
  all(decline$pct_decline > 5),
  all(decline$pct_decline < 30)
)
```

# Part 3 - Monte Carlo replication of Table 2

## Why the raw regression cannot be reproduced at vignette scale

Verner et al. ran **250,000** Monte Carlo iterations. The vignette budget
is 200 subjects per chemical. The published 95% CI half-width for the
PFOA cord coefficient is about 1.3 g at n = 250,000; at n = 200 the same
regression has a half-width near 50 g, so a direct
`lm(birth_weight ~ cord_PFAS)` at this scale is too noisy to gate on -
it flips sign from seed to seed.

There is, however, an exact structural fact that removes almost all of
that noise. In this model birth weight is

```
BW = 3376 + 175.5 * GFRratio + residual
```

and the residual has SD 441 g - three times the entire systematic range
of the GFR term. The residual is what swamps the regression. Because the
slope 175.5 g per unit `GFRratio` is *known exactly*, the GFR-attributable
coefficient can be written without ever touching the residual:

```
beta_GFR = 175.5 * Cov(GFRratio, C) / Var(C)
```

This is the same estimand the paper reports - the birth-weight change per
ng/mL that is attributable to GFR - evaluated with the analytically known
noise term projected out. It is what we compare against Table 2.

```{r mc-setup}
# Inverse-CDF sampling from the truncated normals of Table 1 / the MC script.
rtnorm <- function(u, mean, sd, lower, upper) {
  qnorm(pnorm(lower, mean, sd) +
          u * (pnorm(upper, mean, sd) - pnorm(lower, mean, sd)), mean, sd)
}

# Table 1 / MC-script distributions for the parameters the paper's global
# sensitivity analysis identified as influential.
mc_spec <- list(
  PFOS = list(kp_liver = c(3.72, 0.558, 2.604, 4.836),
              kp_rest  = c(0.20, 0.030, 0.140, 0.260),
              fu       = c(0.025, 0.00375, 0.0175, 0.0325),
              tm       = c(3.5, 0.525, 2.45, 4.55),
              km       = c(0.023, 0.00345, 0.0161, 0.0299),
              cvinit   = c(0.01302, 0.01302 * 0.368, 1e-5, 1)),
  PFOA = list(kp_liver = c(2.20, 0.330, 1.540, 2.860),
              kp_rest  = c(0.12, 0.018, 0.084, 0.156),
              fu       = c(0.020, 0.00300, 0.0140, 0.0260),
              tm       = c(10.0, 1.50, 7.00, 13.00),
              km       = c(0.055, 0.00825, 0.0385, 0.0715),
              cvinit   = c(0.00253, 0.00253 * 0.446, 1e-5, 1))
)

# Published Table 2, "1 (main results)" rows.
published <- data.frame(
  chemical = c("PFOS", "PFOS", "PFOA", "PFOA"),
  matrix   = c("maternal", "cord", "maternal", "cord"),
  beta     = c(-1.46, -2.72, -7.92, -7.13),
  lci      = c(-1.81, -3.40, -9.42, -8.46),
  uci      = c(-1.11, -2.04, -6.43, -5.80)
)

# Build one cohort of MC inputs. `set.seed()` seeds R's RNG only; every
# draw below is plain runif() pushed through an inverse CDF, so the cohort
# is fully determined by the seed and does not depend on rxode2's internal
# streams or on the solver thread count.
mc_cohort <- function(chemical, n, seed) {
  sp <- mc_spec[[chemical]]
  set.seed(seed)
  u <- function() runif(n)
  data.frame(
    # BWINIT: the deposited MC script draws N(74.7, 17.8) truncated to
    # 50-114 kg, NOT the Table 1 MoBa distribution. See Errata.
    WT            = rtnorm(u(), 74.7, 17.8, 50, 114),
    RENALFUNC_REL = rtnorm(u(), 1.0, 0.246, 0.51, 1.49),
    bw_resid      = rtnorm(u(), 0, 0.441, -0.882, 0.882),
    lkp_liver     = log(rtnorm(u(), sp$kp_liver[1], sp$kp_liver[2],
                               sp$kp_liver[3], sp$kp_liver[4])),
    lkp_rest      = log(rtnorm(u(), sp$kp_rest[1], sp$kp_rest[2],
                               sp$kp_rest[3], sp$kp_rest[4])),
    fu            = rtnorm(u(), sp$fu[1], sp$fu[2], sp$fu[3], sp$fu[4]),
    fu_fet        = rtnorm(u(), sp$fu[1], sp$fu[2], sp$fu[3], sp$fu[4]),
    tm_per_kg     = rtnorm(u(), sp$tm[1], sp$tm[2], sp$tm[3], sp$tm[4]),
    km            = rtnorm(u(), sp$km[1], sp$km[2], sp$km[3], sp$km[4]),
    fv_liver      = rtnorm(u(), 0.026, 0.0039, 0.0182, 0.0338),
    cvinit        = rtnorm(u(), sp$cvinit[1], sp$cvinit[2],
                           sp$cvinit[3], sp$cvinit[4])
  )
}

beta_gfr <- function(gfrratio, conc) 175.5 * cov(gfrratio, conc) / var(conc)

N_MC <- 200L   # per chemical; the 200-per-arm vignette cap

# The seed the paper itself used: the deposited MC script opens with
# `seedrnd(123456789)`, and that is the value in the "Sampling seed" column
# of the Table 2 main-results row. R's RNG is of course not acslX's, so this
# does not reproduce their individual draws -- it is simply the principled
# choice of a fixed seed rather than a hand-picked one.
MC_SEED <- 123456789L
```

```{r mc-run}
mc <- lapply(names(models), function(nm) {
  p <- mc_cohort(nm, N_MC, MC_SEED)
  s <- rxSolve(models[[nm]], et(c(MONTH_H, T_CORD)), params = p,
               returnType = "data.frame")
  list(chemical = nm, params = p, solve = s)
})
names(mc) <- names(models)

simulated <- lapply(mc, function(x) {
  d <- x$solve[x$solve$time == T_CORD, ]
  data.frame(
    chemical = x$chemical,
    matrix = c("maternal", "cord"),
    beta = c(beta_gfr(x$params$RENALFUNC_REL, d$Cc * 1000),
             beta_gfr(x$params$RENALFUNC_REL, d$Cplasma_fet * 1000)),
    median_conc = c(median(d$Cc * 1000), median(d$Cplasma_fet * 1000))
  )
}) |> bind_rows()

comparison <- published |>
  left_join(simulated, by = c("chemical", "matrix")) |>
  mutate(ratio = beta.y / beta.x) |>
  rename("Chemical" = chemical, "Matrix" = matrix,
         "Published beta" = beta.x, "Published LCI" = lci,
         "Published UCI" = uci, "Simulated beta_GFR" = beta.y,
         "Median conc (ng/mL)" = median_conc, "Simulated / published" = ratio)

knitr::kable(
  comparison |> mutate(across(where(is.numeric), \(x) signif(x, 3))),
  caption = paste0(
    "Birth-weight change (g) per 1 ng/mL increase in simulated plasma PFAS ",
    "at delivery, attributable to GFR. Published values are Table 2, ",
    "'1 (main results)'. Simulated values use n = ", N_MC, " per chemical.")
)
```

```{r mc-assertions}
sim <- comparison[["Simulated beta_GFR"]]
rat <- comparison[["Simulated / published"]]

stopifnot(
  # All four coefficients are negative: higher simulated PFAS goes with
  # lower birth weight, with no PFAS effect on growth in the model at all.
  all(sim < 0),
  # Magnitude agreement. The band is deliberately wide. Everything above is
  # reproducible to the bit -- the cohort comes from set.seed() + runif()
  # pushed through an inverse CDF, never from rxode2's internal streams, and
  # the ODE solve is deterministic -- so this gate could in principle be
  # pinned to the seed below. It is not, because at n = 200 against the
  # paper's n = 250,000 the ratio moves over 0.36-1.23 from seed to seed
  # (measured over five seeds while authoring). A factor-of-3 band still
  # excludes every sign error, unit error and order-of-magnitude
  # transcription error, which is what this gate is for.
  all(rat > 0.3 & rat < 3.0)
)

# The paper reports that for PFOS the cord-plasma association is stronger
# than the maternal-plasma one, while for PFOA the two are similar. The
# PFOS ordering is a structural consequence of the fetal compartment
# receiving PFOS only through the placenta, and it held for every seed
# tested while authoring.
pfos_beta <- simulated[simulated$chemical == "PFOS", ]
stopifnot(
  abs(pfos_beta$beta[pfos_beta$matrix == "cord"]) >
    abs(pfos_beta$beta[pfos_beta$matrix == "maternal"])
)
```

## Figure 3 - the association strengthens through pregnancy

Verner et al. found that the association "only appeared after the third
month of pregnancy and was strongest at the time of delivery"
(Figure 3). That is a statement about how the GFR sensitivity of plasma
PFAS accumulates: early in pregnancy the reference GFR profile has barely
begun to rise, so a woman's `GFRratio` has had little time to move her
plasma level.

```{r figure3}
by_month <- lapply(mc, function(x) {
  data.frame(
    chemical = x$chemical,
    month = seq_along(MONTH_H),
    beta = vapply(MONTH_H, function(tt) {
      d <- x$solve[x$solve$time == tt, ]
      beta_gfr(x$params$RENALFUNC_REL, d$Cc * 1000)
    }, numeric(1))
  )
}) |> bind_rows()

# The deterministic counterpart: dC/dGFRratio, free of Monte Carlo noise.
sensitivity <- lapply(names(models), function(nm) {
  s <- rxSolve(models[[nm]], et(MONTH_H),
               params = data.frame(WT = 74.7, RENALFUNC_REL = c(0.9, 1.1),
                                   bw_resid = 0),
               returnType = "data.frame")
  data.frame(
    chemical = nm, month = seq_along(MONTH_H),
    dC_dGFR = vapply(MONTH_H, function(tt) {
      d <- s[s$time == tt, ]
      diff(d$Cc * 1000) / 0.2
    }, numeric(1))
  )
}) |> bind_rows()

ggplot(by_month, aes(month, beta, colour = chemical)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:9) +
  labs(x = "Month of pregnancy at blood sampling",
       y = "Birth weight change (g) per ng/mL, attributable to GFR",
       colour = "Chemical",
       title = "Replicates Figure 3 of Verner 2015 (maternal plasma)",
       subtitle = "The confounded association grows with gestational age") +
  theme_bw()

knitr::kable(
  sensitivity |>
    pivot_wider(names_from = chemical, values_from = dC_dGFR) |>
    mutate(across(where(is.numeric), \(x) signif(x, 3))) |>
    rename("Month" = month),
  caption = paste("Deterministic GFR sensitivity dC/dGFRratio of maternal",
                  "plasma (ng/mL per unit GFRratio), by month of pregnancy.")
)

stopifnot(
  # Deterministic and therefore exactly reproducible: the GFR sensitivity
  # of maternal plasma grows in magnitude every single month.
  all(vapply(split(sensitivity, sensitivity$chemical),
             function(d) all(diff(abs(d$dC_dGFR)) > 0), logical(1))),
  # ... and so does the confounded association itself.
  all(vapply(split(by_month, by_month$chemical),
             function(d) all(diff(abs(d$beta)) > 0), logical(1))),
  # Month 1 is negligible relative to delivery, as Figure 3 shows.
  all(vapply(split(by_month, by_month$chemical),
             function(d) abs(d$beta[1]) < 0.5 * abs(d$beta[9]), logical(1)))
)
```

# Part 4 - PKNCA validation

The paper reports no NCA parameters, so there is nothing to compare a
noncompartmental analysis *against*. PKNCA is used here instead as an
independent numerical instrument: the model carries maternal plasma AUC
as its own ODE state (`auc_plasma`, the acslX `AUCPlas` accumulator),
integrated by CVODE, while PKNCA computes the same integral by the
trapezoidal rule on the reported output grid. The two are different
numerical methods applied to the same trajectory, so a disagreement is a
real signal - a mis-specified observation grid, a dropped time-zero
record, or a units error.

```{r pknca}
nca_solve <- rxSolve(
  pfos, et(seq(0, T_DELIVERY, by = 73)),
  params = mc_cohort("PFOS", 20L, MC_SEED),
  returnType = "data.frame"
)

# Concentrations -- keep the column named Cc (nlmixr2lib convention). The
# filter is `!is.na(Cc)` ONLY: adding time > 0 or Cc > 0 would drop the
# time-zero record PKNCA needs to anchor the AUC interval.
nca_conc <- nca_solve |>
  transmute(id = as.integer(sim.id), time, Cc = Cc * 1000) |>
  filter(!is.na(Cc))
stopifnot(all(table(nca_conc$id[nca_conc$time == 0]) == 1))

nca_obj <- PKNCAconc(nca_conc, Cc ~ time | id)
nca_data <- PKNCAdata(
  nca_obj,
  intervals = data.frame(start = 0, end = T_DELIVERY,
                         auclast = TRUE, cmax = TRUE, tmax = TRUE)
)
nca_res <- as.data.frame(pk.nca(nca_data))

nca_wide <- nca_res |>
  select(id, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

model_auc <- nca_solve |>
  filter(time == T_DELIVERY) |>
  transmute(id = as.integer(sim.id), auc_ode = auc_plasma * 1000)

auc_check <- nca_wide |>
  left_join(model_auc, by = "id") |>
  mutate(pct_diff = 100 * (auclast - auc_ode) / auc_ode)

knitr::kable(
  auc_check |>
    summarise(`Median PKNCA AUClast (ng*h/mL)` = median(auclast),
              `Median ODE AUC (ng*h/mL)` = median(auc_ode),
              `Median % difference` = median(pct_diff),
              `Max abs % difference` = max(abs(pct_diff))) |>
    mutate(across(everything(), \(x) signif(x, 4))),
  caption = paste("Trapezoidal AUC from PKNCA vs. the model's own",
                  "CVODE-integrated AUC state, over the whole pregnancy.")
)

```

> Note on `tmax`. Only some subjects peak at conception. The hourly intake
> of Eq. 2 is back-calculated per subject from that subject's own volume of
> distribution, which depends on the sampled partition coefficients, so a
> subject whose `Vd` is large relative to their initial burden accumulates
> for a while before plasma-volume expansion and rising GFR take over. In
> the cohort below roughly a third peak at `t = 0` and the rest peak in the
> second trimester. What holds for *every* subject is that the level at
> delivery is below the level at conception - the decline Verner et al.
> compared against Glynn et al. (2012) and Monroy et al. (2008) in Figure 2.

```{r pknca-assertions}
endpoints <- nca_solve |>
  group_by(sim.id) |>
  summarise(ratio = Cc[time == T_DELIVERY] / Cc[time == 0], .groups = "drop")

stopifnot(
  # Both sides use the SAME drawn parameters for each subject, so the only
  # difference is trapezoid-vs-CVODE quadrature on a smooth, slowly varying
  # curve. A tight bound is correct here and should be kept tight.
  max(abs(auc_check$pct_diff)) < 0.5,
  # Every subject ends pregnancy below where they started.
  all(endpoints$ratio < 1),
  # PKNCA's Cmax must be the largest concentration actually simulated.
  all(abs(nca_wide$cmax -
            tapply(nca_conc$Cc, nca_conc$id, max)[as.character(nca_wide$id)]) < 1e-9)
)
```

# Assumptions and deviations

## Errata and internal conflicts in the source

1. **The acslX constant comments for `k3c`/`k4c` are swapped relative to
   the listing's own equations.** The declarations read
   `PFOS_k3c ... ! Amniotic fluid to fetus` and
   `PFOS_k4c ... ! Fetus to fluid`, but the equations are
   `rtrans3 = ktrans3*CVRF*FreeF ! Transfer from fetal rest of body to
   amniotic fluid` and `rtrans4 = ktrans4*CAmF ! Transfer from amniotic
   fluid to fetal rest of body`, and the `ktrans3`/`ktrans4` definitions a
   few lines earlier carry those same (correct) directions. The equations
   are authoritative and were followed; the `k3`/`k4` labels in the model
   files therefore read opposite to the constant-declaration comments.

2. **Pre-pregnancy body weight: Table 1 and the deposited MC script
   disagree.** Table 1 gives 70.3 +/- 14.3 kg truncated to 37.0-134.0 (MoBa),
   whereas the script that actually ran the simulations draws
   `BWINIT = normrnd(74.7, 17.8, 50.0, 114.0)`. The Monte Carlo section
   above uses the script's distribution, because the script is the code
   that produced Table 2. This is *not* a settled reading: averaged over
   five seeds the simulated/published ratio across the four cells spans
   0.84-1.22 under the script's distribution and 0.95-1.36 under Table 1's,
   which favours the script only weakly. The choice moves the simulated
   coefficients by roughly 10-15%, well inside the assertion band above.
   `covariateData$WT` records both.

3. **`PFOA_FREE` lower truncation bound is a typo in the MC script.** It
   reads `normrnd(0.020, 0.00300, 0.00140, 0.02600)`; the lower bound
   should be 0.0140 (mean minus 2 SD, and the value Table 1 prints). The
   model files carry the Table 1 bound, and the vignette samples with it.

4. **Initial plasma PFAS truncation.** Table 1 gives a maximum of
   100 ng/mL; the MC script uses `MAX = 1.0` mg/L (1,000 ng/mL). At a mean
   of 13.02 ng/mL with CV 0.368 neither bound binds in practice.

5. **`VSk` is declared in litres, not as a weight fraction.** The listing
   comments `constant VSk = 0.0972 ! Skin volume (L)` and uses it directly
   in the rest-of-body subtraction, so skin volume does not scale with
   body weight. This is reproduced as published; it is almost certainly
   meant to be a fraction of body weight (0.0972 x 60.9 kg would be a
   plausible 5.9 L), but the listing is unambiguous and was not "fixed".

6. **Sampling times.** The listing records the ninth-month maternal sample
   and the cord sample at 6,569 h, and terminates at `TSTOP = 6,570` h.
   Both times appear in this vignette for that reason.

## Modelling choices

7. **`Cc(0)` is 0.04% below the nominal initial plasma level.** Eq. 1 puts
   the initial free amount into `VPlasINIT` (the pre-pregnancy plasma
   volume), while the reported concentration divides by `VPlas(0)`, which
   already includes the week-zero amniotic fluid and placenta. The acslX
   listing does exactly the same, so this is reproduced rather than
   corrected, and it is gated as a fixed geometric factor.

8. **Partition coefficients are carried log-transformed** (`lkp_liver`
   etc., back-transformed at the top of `model()`), matching the package
   convention and the sibling gestational PFAS model
   `Zhang_2024_f53b_human_pbpk`.

9. **`amniotic` uses `specimen = "tissue"`.** Amniotic fluid is not in the
   package specimen vocabulary; `"tissue"` follows the shipped precedent
   in `Zhang_2024_f53b_human_pbpk`, which makes the same choice for its
   own amniotic compartment. No new vocabulary entry was introduced.

10. **No IIV and no residual error.** Verner et al. ran a deterministic
    PBPK model 250,000 times under the Table 1 input distributions; they
    estimated no random effects and report no residual-error model. The
    files therefore carry every parameter as `fixed()`, and population
    variability is generated by sampling `params` in `rxSolve()`, exactly
    as the acslX automation script does.

11. **The GFR-birth-weight coupling is a covariate, not a compartment.**
    `RENALFUNC_REL` (the paper's `GFRratio`) simultaneously scales the
    whole gestational GFR profile (Eq. 3) and sets birth weight (Eq. 4),
    and birth weight rescales the fetal growth curve (Eq. 5). A single
    subject-level draw therefore induces the confounded association the
    paper quantifies.

12. **Monte Carlo scale.** 200 subjects per chemical, against the paper's
    250,000 iterations. Part 3 explains why the raw regression is not
    reproducible at that scale and what is gated instead. The
    sensitivity-analysis rows of Table 2 (halved/doubled PFAS means, CVs
    and GFR-birth-weight coefficients, and the alternative 2.3-year PFOA
    half-life) are not reproduced here; they are straightforward
    re-runs of the same machinery with one input changed.
