---
title: "Morphine and M6G brain PBPK (Verscheijden 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Morphine and M6G brain PBPK (Verscheijden 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Verscheijden_2021_morphine_pbpk")()$reference`
* Description: `r readModelDb("Verscheijden_2021_morphine_pbpk")()$description`
* Article: <https://doi.org/10.1371/journal.pcbi.1008786> (open access)
* Deposited code: S1 File of the article, the complete `deSolve`
  implementation of the paediatric model (2.6-16.4 y version).
* Structural brain framework and adult physiology:
  <https://doi.org/10.1371/journal.pcbi.1007117> (Verscheijden et al. 2019,
  S1 Table, adult column).

Verscheijden et al. built a whole-body physiologically based pharmacokinetic
(PBPK) model for morphine and its active metabolite morphine-6-glucuronide
(M6G), with a four-compartment brain (brain blood, brain mass, cranial CSF,
spinal CSF). Passive blood-brain and blood-CSF barrier permeability came from a
rat perfusion study, and active P-glycoprotein (Pgp) efflux of morphine was
scaled from MDCKII-Pgp transwell experiments by in vitro-in vivo extrapolation
(IVIVE), with postnatal Pgp maturation. Organ volumes, flows and tissue
composition depend on age, sex, weight and height, so one model file covers
neonates, children and adults. A receptor-occupancy module converts unbound
brain-mass concentrations of morphine and M6G into a relative response.

This vignette validates the pharmacokinetic part of the model: the IVIVE
clearance, the mass balance, the volume of distribution, the published
with/without-Pgp brain exposure ratio, and the concentration-time profiles of
the paper's verification scenarios. The receptor-occupancy output is shown
only as its relative-response curve.

# Model structure

```
 venous blood --> lung --> arterial blood --> 10 perfused tissues --> venous blood
      |  CL (blood)                |
      v                            +--> brain blood <--PSb, Pgp--> brain mass
  0.1 x CL x MW ratio                     ^   PSc        PSe  |  bulk flow
  forms M6G (venous)                      +--> cranial CSF <--+
                                                    ^ |
                                         spinal CSF -+ +- sink flows to blood
```

* 17 amount states for morphine (`venous`, `arterial`, `lung`, `adipose`,
  `bone`, `heart`, `kidney`, `muscle`, `skin`, `spleen`, `other`, `gut`,
  `liver`, `brain_vascular`, `brain`, `brain_csf_sas_cranial`,
  `brain_csf_sas_spinal`) and the same 17 with an `_m6g` suffix for M6G.
* IV doses go into `venous`. Clearance acts on the venous blood
  concentration; a molecular-weight-corrected fraction of the cleared morphine
  appears as M6G in venous blood.
* Outputs: `Cc` (plasma morphine), `Cc_m6g`, `Cbrain_u` (unbound brain mass,
  the paper's surrogate for microdialysis ECF), `Ccsf` (cranial CSF),
  `Ccsf_spinal` and the M6G equivalents, plus `effect_rel` (relative response).

# Population

The paper fits no parameter to individual data. Drug inputs come from the
literature and the authors' own MDCKII-Pgp experiments (Table 1), and 11
published studies plus 19 neonatal CSF biobank samples were used only for
verification (Table 2):

* Adults 19-69 y: neurological / neurosurgical patients after 0.38 mg/kg IV
  (Meineke 2002; plasma and ventricular CSF), and traumatic brain injury
  patients after 10 mg IV (Bouw 2001, Ederoth 2003; plasma and microdialysis
  ECF).
* Children 1-18 y with acute leukaemia after 0.25 mg/kg IV (Hain 1999; plasma
  and lumbar CSF), and three children 3.5-9.5 y on a 0.03 mg/kg/h infusion
  (Ketharanathan 2019; plasma and ECF).
* Neonates after 0.1 mg/kg IV at 1.1 and 29 days postnatal age (Pokela 1993).

The paper simulated 500 virtual individuals per scenario, matched to each
study's age range, dose and sex ratio. The `population` metadata of the model
file records these facts.

```{r population}
str(readModelDb("Verscheijden_2021_morphine_pbpk")()$population)
```

# Source trace

| Model element | Value | Source |
|---|---|---|
| Morphine CL, adults | 1.962 L/h/kg | Table 1, CLiv (adult) |
| Morphine CL, children and neonates | 60 x 1.62 x (WT/70)^(1.47 - 0.59 WT^4.62/(4.01^4.62 + WT^4.62)) L/h | Table 1, CLiv (2-18y), (neonates); S1 File line 706 |
| M6G CL, adults / children / neonates | 0.131 / 0.114 / 0.017 L/h/kg | Table 1, CLiv M6G |
| Fraction of CL forming M6G | 0.1 (2 y to adult), 0.044 (neonates) | Table 1 |
| MW morphine / M6G | 285.343 / 461.467 g/mol | Table 1 |
| LogP, pKa, EP morphine | 0.89, 8.21 (base), 1.34 | Table 1 |
| LogP, pKa, EP M6G | -2.9, 2.87 (acid) / 9.12 (base), 0.15 | Table 1 |
| Olive oil:water logP morphine / M6G | -0.35765 / -4.5835 | S1 File lines 572, 591 |
| M6G Kp scalar | 0.5 | Methods; S1 File line 599 |
| fu plasma / brain mass / CSF, morphine | 0.64 / 0.5 / 1 | Table 1 |
| fu plasma / brain mass / CSF, M6G | 0.83 / 0.99 / 1 | Table 1 |
| fu blood morphine | 0.64 / 1.14 | S1 File line 475 |
| PSb morphine / M6G | 0.2112 / 0.0072 L/h per kg brain | Table 1 |
| PSc | 0.5 x PSb | Table 1 |
| PSe | 300 L/h | Table 1 |
| Pgp IVIVE inputs | ER 1.30, Papp 2.12e-6 cm/s, SA 0.33 cm^2, 81.4 ug protein, 4.21 / 0.19 pmol/mg, 0.244 mg/g, 1400 g | Table 1; Methods Eqs 2-3 |
| Pgp maturation | 41% at term, 100% at 6 months | Methods |
| CSF production adult / child / neonate | 0.021 / 0.024 / 0.010 L/h | 2019 S1 Table; Methods |
| Bulk flow, spinal sink and outflow fractions | 0.25, 0.38, 0.9 | 2019 S1 Table; S1 File lines 403-411 |
| Adult haematocrit male / female | 0.43 / 0.38 | 2019 S1 Table |
| Paediatric organ volumes, flows, haematocrit, tissue composition | equations | S1 File lines 266-569 |
| Adult organ volumes, flows, tissue composition | equations | 2019 S1 Table, adult column |
| Rodgers and Rowland Kp (base; zwitterion for M6G) | equations | S1 File lines 571-610 |
| Morphine and M6G ODEs | equations | S1 File lines 91-128 |
| Receptor occupancy and response | KM 11.8 / 42.5 nM, BR50 59.26 / 17, Hill 4.217 / 2.2 | Methods Eqs 4-6; S1 File lines 964-983 |
| IIV (SD of log-normal draws) | CL 0.4, M6G CL 0.18, CSF production 0.1, bulk 0.08, sink 0.3, outflow 1, haematocrit 0.065 / 0.071 | S1 File lines 400-492, 707-711 |

# Virtual cohorts

Paediatric height and weight are generated with the sex-specific
age-to-height and height-to-weight equations of the deposited code (S1 File
lines 14-38), including their log-normal variability. The deposited code has
no adult generator, so adult height and weight are drawn from normal and
log-normal distributions with typical European values (see Assumptions).
Each arm has 100 subjects (50% female).

```{r cohort-functions}
paed_height <- function(age, female) {
  male_h <- 0.0000176179 * age^7 - 0.00119874 * age^6 + 0.0323848 * age^5 -
    0.444112 * age^4 + 3.2946 * age^3 - 13.2191 * age^2 + 33.75 * age + 52.62152
  fem_h <- -0.00000151027 * age^8 + 0.000121261 * age^7 - 0.0040023 * age^6 +
    0.070179 * age^5 - 0.708233 * age^4 + 4.1872 * age^3 - 14.3393 * age^2 +
    33.84778 * age + 51.535477
  ifelse(female == 1, fem_h * exp(rnorm(length(age), 0, 0.052)), male_h * exp(rnorm(length(age), 0, 0.057)))
}

paed_weight <- function(age, height, female) {
  male_w <- 7.826 * (1 - exp(age * -1.2)) + exp(height * 0.0209 + 0.023 * age)
  fem_w <- 5.454 * (1 - exp(age * -1.57)) + exp(height * 0.0224 + 0.019 * age)
  ifelse(female == 1, fem_w * exp(rnorm(length(age), 0, 0.10)), male_w * exp(rnorm(length(age), 0, 0.15)))
}

make_cohort <- function(n, age_lo, age_hi, arm, id0 = 0) {
  female <- rep(c(0, 1), length.out = n)
  age <- runif(n, age_lo, age_hi)
  if (age_lo >= 18) {
    ht <- ifelse(female == 1, rnorm(n, 165, 7), rnorm(n, 178, 7))
    wt <- ifelse(female == 1, 66, 80) * exp(rnorm(n, 0, 0.15))
  } else {
    ht <- paed_height(age, female)
    wt <- paed_weight(age, ht, female)
  }
  data.frame(id = id0 + seq_len(n), arm = arm, AGE = age, WT = wt, HT = ht, SEXF = female)
}

# One dosing record per subject plus an observation grid on the venous state.
# dur = 0 gives an IV bolus; dur > 0 a constant-rate infusion.
make_events <- function(cohort, dose_mg, dur, obs_times) {
  dose <- cohort |>
    mutate(time = 0, amt = dose_mg, rate = if (dur > 0) dose_mg / dur else 0, evid = 1L, cmt = "venous")
  obs <- tidyr::crossing(cohort, time = obs_times) |>
    mutate(amt = 0, rate = 0, evid = 0L, cmt = "venous")
  bind_rows(dose, obs) |> arrange(id, time, desc(evid))
}
```

```{r cohorts}
rxode2::rxSetSeed(20210304)
set.seed(20210304)
n_arm <- 100
obs_24 <- sort(unique(c(seq(0, 0.5, by = 0.0125), seq(0.5, 2, by = 0.1), seq(2.25, 24, by = 0.25))))
# "Single IV dose": a 5-min infusion into venous blood (see Assumptions)
iv_dur <- 5 / 60

coh_adult <- make_cohort(n_arm, 19, 69, "Adults 0.38 mg/kg (Fig 1A)")
coh_paed <- make_cohort(n_arm, 1.4, 15.9, "Children 0.25 mg/kg (Fig 1B)", id0 = 1000)
coh_adult_ecf <- make_cohort(n_arm, 32, 52, "Adults 10 mg (Fig 2A)", id0 = 2000)
coh_paed_ecf <- make_cohort(n_arm, 3.5, 9.5, "Children 0.03 mg/kg/h (Fig 2A)", id0 = 3000)

ev <- bind_rows(
  make_events(coh_adult, 0.38 * coh_adult$WT, iv_dur, obs_24),
  make_events(coh_paed, 0.25 * coh_paed$WT, iv_dur, obs_24),
  make_events(coh_adult_ecf, 10, iv_dur, obs_24),
  # continuous infusion over the whole 24-h window
  make_events(coh_paed_ecf, 0.03 * coh_paed_ecf$WT * 24, 24, obs_24)
)

bind_rows(coh_adult, coh_paed, coh_adult_ecf, coh_paed_ecf) |>
  group_by(arm) |>
  summarise(
    n = n(), `female (%)` = 100 * mean(SEXF),
    `age (y)` = sprintf("%.1f (%.1f-%.1f)", median(AGE), min(AGE), max(AGE)),
    `weight (kg)` = sprintf("%.1f (%.1f-%.1f)", median(WT), min(WT), max(WT)),
    `height (cm)` = sprintf("%.0f (%.0f-%.0f)", median(HT), min(HT), max(HT)),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Virtual cohorts: median (range).")
```

# Simulation

```{r simulate}
mod <- readModelDb("Verscheijden_2021_morphine_pbpk")
keep_cols <- c("Cc", "Cc_m6g", "Cbrain_u", "Ccsf", "Ccsf_spinal", "Cbrain_u_m6g", "Ccsf_m6g", "Ccsf_spinal_m6g")
sim <- rxode2::rxSolve(mod, ev, keep = "arm", returnType = "data.frame") |>
  select(id, time, arm, all_of(keep_cols), cl, bp)

summ <- sim |>
  pivot_longer(all_of(keep_cols), names_to = "output", values_to = "conc") |>
  group_by(arm, output, time) |>
  summarise(
    median = median(conc), lo = quantile(conc, 0.05), hi = quantile(conc, 0.95),
    .groups = "drop"
  )

plot_band <- function(d, title) {
  ggplot(d, aes(time, median)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), fill = "grey80") +
    geom_line() +
    facet_wrap(~panel, scales = "free_y") +
    scale_y_log10() +
    labs(x = "Time (h)", y = "Concentration (mg/L)", title = title,
         caption = "Line: median; band: 90% interval of the virtual cohort.")
}

panel_labels <- c(
  Cc = "Plasma morphine", Cc_m6g = "Plasma M6G",
  Ccsf = "Cranial CSF morphine", Ccsf_m6g = "Cranial CSF M6G",
  Ccsf_spinal = "Spinal CSF morphine", Ccsf_spinal_m6g = "Spinal CSF M6G",
  Cbrain_u = "Unbound brain (ECF) morphine"
)
```

# Replicate published figures

## Figure 1A -- adults, plasma and ventricular CSF

```{r fig1a, fig.height = 6}
summ |>
  filter(arm == "Adults 0.38 mg/kg (Fig 1A)", output %in% c("Cc", "Cc_m6g", "Ccsf", "Ccsf_m6g"), time > 0) |>
  mutate(panel = factor(panel_labels[output], levels = panel_labels[c("Cc", "Cc_m6g", "Ccsf", "Ccsf_m6g")])) |>
  plot_band("Adults, 0.38 mg/kg IV")
```

Replicates Figure 1A of Verscheijden 2021. The ventricular-drain samples are
compared with the cranial CSF compartment.

## Figure 1B -- children, plasma and lumbar CSF

```{r fig1b, fig.height = 6}
summ |>
  filter(arm == "Children 0.25 mg/kg (Fig 1B)", output %in% c("Cc", "Cc_m6g", "Ccsf_spinal", "Ccsf_spinal_m6g"), time > 0) |>
  mutate(panel = factor(panel_labels[output], levels = panel_labels[c("Cc", "Cc_m6g", "Ccsf_spinal", "Ccsf_spinal_m6g")])) |>
  plot_band("Children 1.4-15.9 y, 0.25 mg/kg IV")
```

Replicates Figure 1B of Verscheijden 2021. Lumbar-puncture samples are
compared with the spinal CSF compartment.

## Figure 2A -- plasma and brain ECF

```{r fig2a, fig.height = 6}
summ |>
  filter(grepl("Fig 2A", arm), output %in% c("Cc", "Cbrain_u"), time > 0) |>
  mutate(panel = paste(panel_labels[output], "|", sub(" \\(Fig 2A\\)", "", arm))) |>
  plot_band("Plasma and unbound brain (ECF) morphine")
```

Replicates Figure 2A of Verscheijden 2021. As in the paper, the unbound
brain-mass concentration stands in for the microdialysis ECF measurement. The
paediatric infusion reaches a brain plateau within the first day, while the
plasma profile is flat from the first hours; the paper notes that one measured
9.5-year-old needed more than 20 h to reach ECF equilibrium.

## Figure 2B -- neonatal plasma

```{r fig2b}
neo <- data.frame(
  id = 1:2, arm = c("1.1 days postnatal", "29 days postnatal"),
  AGE = c(1.1, 29) / 365.25, SEXF = 0
)
neo$HT <- 52.62152 + 33.75 * neo$AGE # deposited male height polynomial near birth
neo$WT <- 7.826 * (1 - exp(neo$AGE * -1.2)) + exp(neo$HT * 0.0209 + 0.023 * neo$AGE)
ev_neo <- make_events(neo, 0.1 * neo$WT, iv_dur, obs_24)
# Typical-value solves, one subject at a time
solve_each <- function(m, events) {
  bind_rows(lapply(split(events, events$id), function(e) {
    rxode2::rxSolve(m, e, keep = "arm", returnType = "data.frame") |> mutate(id = e$id[1])
  }))
}
sim_neo <- solve_each(rxode2::zeroRe(mod), ev_neo)

ggplot(filter(sim_neo, time > 0), aes(time, Cc, colour = arm)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Plasma morphine (mg/L)", colour = NULL,
       title = "Typical neonates, 0.1 mg/kg IV")
```

Replicates Figure 2B of Verscheijden 2021 (typical-value profiles). Neonatal
clearance uses the bodyweight-dependent-exponent equation, and M6G formation
falls to 4.4% of clearance in neonates.

## Figure 2D -- Pgp contribution to brain exposure

The paper reports that adding Pgp efflux lowers adult brain ECF exposure
1.3-fold (AUC without Pgp / AUC with Pgp = 1.3) after 0.38 mg/kg. Pgp is
removed by setting the net in vitro efflux ratio to 1, which makes the IVIVE
efflux clearance zero.

```{r fig2d}
typ_adult <- data.frame(id = 1, arm = "typical adult", AGE = 40, WT = 70, HT = 176, SEXF = 0)
obs_48 <- seq(0, 48, by = 0.05)
ev_typ <- make_events(typ_adult, 0.38 * 70, iv_dur, obs_48)
mod_typ <- rxode2::zeroRe(mod)
mod_nopgp <- mod_typ |> rxode2::ini(er_pgp = 1)

s_pgp <- rxode2::rxSolve(mod_typ, ev_typ, returnType = "data.frame")
s_nopgp <- rxode2::rxSolve(mod_nopgp, ev_typ, returnType = "data.frame")

trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
auc_ratio <- trap(s_nopgp$time, s_nopgp$Cbrain_u) / trap(s_pgp$time, s_pgp$Cbrain_u)

bind_rows(
  mutate(s_pgp, scenario = "Pgp included"),
  mutate(s_nopgp, scenario = "Pgp not included")
) |>
  filter(time > 0, time <= 24) |>
  ggplot(aes(time, Cbrain_u, colour = scenario)) +
  geom_line() +
  scale_colour_manual(values = c("Pgp included" = "red", "Pgp not included" = "black")) +
  labs(x = "Time (h)", y = "Unbound brain (ECF) morphine (mg/L)", colour = NULL,
       title = sprintf("Typical adult, 0.38 mg/kg IV: AUC ratio = %.3f (paper: 1.3)", auc_ratio))

stopifnot(abs(auc_ratio - 1.3) < 0.05)
```

Replicates Figure 2D of Verscheijden 2021. Both curves use the same typical
parameters, so the ratio is deterministic.

# Clearance placement

The deposited S1 File removes the cleared morphine from the arterial
compartment while computing it from the venous concentration (S1 File line 97,
`-(CL_iv * Cblv)` in `dCbla`). The 2019 framework code removes it from venous
blood (2019 S1 File line 81). The two forms clear the same mass, but the
arterial form lowers the concentration reaching the tissues and the brain.
The model file uses the venous form. Below, the deposited form is rebuilt by
swapping the four blood ODEs, and both are compared with values digitised by
the maintainers from the paper's own figures:

* Figure 2D, typical adult after 0.38 mg/kg with Pgp: unbound brain (ECF)
  morphine of 11.4, 11.8, 11.1, 8.2, 3.4, 1.4 and 0.59 ng/mL at 1, 2, 3, 5, 10,
  15 and 20 h.
* Figure 2A, 9.5-year-old on 0.03 mg/kg/h: median plateaus of about 26 ng/mL
  (plasma) and 12.2 ng/mL (ECF), an ECF:plasma ratio of 0.47. The ratio is used
  because the child's individual weight and clearance are not reported.

```{r clearance-placement}
mod_deposited <- rxode2::rxode2(mod_typ) |>
  rxode2::model(
    d/dt(venous) <- q_adipose * c_ad / kp_adipose * bp + q_bone * c_bo / kp_bone * bp +
      q_heart * c_he / kp_heart * bp + q_kidney * c_ki / kp_kidney * bp +
      q_muscle * c_mu / kp_muscle * bp + q_skin * c_sk / kp_skin * bp +
      q_liver * c_li / kp_liver * bp + q_brain * c_bb + q_rest * c_ot * bp - q_lung * c_ven,
    d/dt(arterial) <- q_lung * c_lung / kp_lung * bp -
      (q_rest + q_brain + q_adipose + q_bone + q_heart + q_kidney + q_muscle + q_skin + q_spleen + q_gut + q_ha) * c_art -
      cl * c_ven,
    d/dt(venous_m6g) <- cl * c_ven * f_form +
      q_adipose * m_ad / kp_adipose_m6g * bp_m6g + q_bone * m_bo / kp_bone_m6g * bp_m6g +
      q_heart * m_he / kp_heart_m6g * bp_m6g + q_kidney * m_ki / kp_kidney_m6g * bp_m6g +
      q_muscle * m_mu / kp_muscle_m6g * bp_m6g + q_skin * m_sk / kp_skin_m6g * bp_m6g +
      q_liver * m_li / kp_liver_m6g * bp_m6g + q_brain * m_bb + q_rest * m_ot / kp_other_m6g * bp_m6g -
      q_lung * m_ven,
    d/dt(arterial_m6g) <- q_lung * m_lung / kp_lung_m6g * bp_m6g -
      (q_rest + q_brain + q_adipose + q_bone + q_heart + q_kidney + q_muscle + q_skin + q_spleen + q_gut + q_ha) * m_art -
      cl_m6g * m_ven
  )

fig2d <- data.frame(
  time = c(1, 2, 3, 5, 10, 15, 20),
  digitised = c(11.4, 11.8, 11.1, 8.16, 3.44, 1.42, 0.589)
)
ecf_at <- function(m) {
  s <- rxode2::rxSolve(m, ev_typ, returnType = "data.frame")
  approx(s$time, s$Cbrain_u * 1000, fig2d$time)$y
}
fig2d$model <- ecf_at(mod_typ)
fig2d$deposited <- ecf_at(mod_deposited)
knitr::kable(fig2d, digits = 2, caption = "Typical adult ECF morphine (ng/mL) vs digitised Figure 2D.")

# Typical 9.5-year-old boy, height and weight from the deposited equations
age_c <- 9.5
ht_c <- 0.0000176179 * age_c^7 - 0.00119874 * age_c^6 + 0.0323848 * age_c^5 - 0.444112 * age_c^4 +
  3.2946 * age_c^3 - 13.2191 * age_c^2 + 33.75 * age_c + 52.62152
wt_c <- 7.826 * (1 - exp(age_c * -1.2)) + exp(ht_c * 0.0209 + 0.023 * age_c)
child <- data.frame(id = 1, arm = "9.5 y", AGE = age_c, WT = wt_c, HT = ht_c, SEXF = 0)
ev_child <- make_events(child, 0.03 * wt_c * 100, 100, c(0, 80))
ss_ratio <- function(m) {
  s <- rxode2::rxSolve(m, ev_child, returnType = "data.frame")
  s$Cbrain_u[s$time == 80] / s$Cc[s$time == 80]
}
ratios <- c(figure = 12.2 / 26, model = ss_ratio(mod_typ), deposited = ss_ratio(mod_deposited))
ratios

# The venous form tracks Figure 2D within 25% up to 20 h and the Figure 2A
# ratio within 10%; the deposited form is 40% or more below Figure 2D and
# about 20% below the Figure 2A ratio.
stopifnot(
  all(abs(fig2d$model / fig2d$digitised - 1) < 0.25),
  all(fig2d$deposited / fig2d$digitised < 0.65),
  abs(ratios[["model"]] / ratios[["figure"]] - 1) < 0.1,
  ratios[["deposited"]] / ratios[["figure"]] < 0.85
)
```

Both forms give the same AUC ratio with and without Pgp (1.30), so the
published ratio cannot tell them apart; the profiles can. The deposited form
also drives arterial blood negative in subjects whose clearance comes near
their cardiac output. Morphine blood clearance in adults is 1.962 L/h/kg, and
its IIV SD is 0.4, so this happens to a few percent of a virtual adult cohort.

# Deterministic checks

## Pgp in vitro-in vivo extrapolation

Methods Eqs 2-3: CLefflux,vitro = 2 x (ER - 1) x Papp,AB x SA / Procell, then
CLefflux,vivo = CLefflux,vitro x abundance(ex vivo) / abundance(in vitro) x
BMvPGB x BW. The deposited code hard-codes the result as 0.14 L/h.

```{r pgp-ivive}
pgp <- s_pgp$q_pgp[1]
pgp_hand <- 2 * (1.30 - 1) * 2.12e-6 * 0.33 / 81.4 * 1000 * (4.21 / 0.19) * 0.244 * 1400 * 3.6
c(model = pgp, by_hand = pgp_hand, deposited_code = 0.14)
stopifnot(abs(pgp - pgp_hand) < 1e-10, abs(pgp / 0.14 - 1) < 0.01)

# Postnatal maturation: 41% of adult activity at birth, full at 6 months
neo_pgp <- solve_each(mod_typ, ev_neo) |>
  group_by(id) |>
  slice(1) |>
  pull(f_pgp)
rbind(model = neo_pgp, by_hand = 0.41 + (1 - 0.41) * neo$AGE / 0.5)
stopifnot(all(abs(neo_pgp - (0.41 + (1 - 0.41) * neo$AGE / 0.5)) < 1e-10))
```

## Mass balance

Clearance removes morphine at `cl x C_venous`, and a fraction
`f_form = fm x MW(M6G)/MW(morphine)` of that mass appears as M6G. Over 48 h,
the dose must equal the cleared mass plus what is left in the body, and the M6G
formed must equal the M6G cleared plus the M6G left. The check uses a 1-h
infusion so the venous concentration is smooth enough for trapezoidal
integration (a bolus puts a spike into venous blood that decays within
minutes).

```{r mass-balance}
morph_states <- c("venous", "arterial", "lung", "adipose", "bone", "heart", "kidney", "muscle",
                  "skin", "spleen", "other", "gut", "liver", "brain_vascular", "brain",
                  "brain_csf_sas_cranial", "brain_csf_sas_spinal")
m6g_states <- paste0(morph_states, "_m6g")
dose_typ <- 0.38 * 70
s_mb <- rxode2::rxSolve(mod_typ, make_events(typ_adult, dose_typ, 1, obs_48), returnType = "data.frame")
last <- s_mb[nrow(s_mb), ]

cleared <- trap(s_mb$time, s_mb$cl * s_mb$c_ven)
left <- sum(unlist(last[morph_states]))
m6g_formed <- cleared * last$f_form
m6g_cleared <- trap(s_mb$time, s_mb$cl_m6g * s_mb$m_ven)
m6g_left <- sum(unlist(last[m6g_states]))

mb <- data.frame(
  quantity = c("Morphine dose (mg)", "Morphine cleared + remaining (mg)",
               "M6G formed (mg)", "M6G cleared + remaining (mg)"),
  value = c(dose_typ, cleared + left, m6g_formed, m6g_cleared + m6g_left)
)
knitr::kable(mb, digits = 4)
stopifnot(
  abs((cleared + left) / dose_typ - 1) < 0.005,
  abs((m6g_cleared + m6g_left) / m6g_formed - 1) < 0.005
)
```

## Volume of distribution

The paper states adult Vss values of 330 L (morphine) and 30 L (M6G) without
naming the reference individual or the reference matrix. Here Vss is computed
from the volumes and Kp values of the typical 70 kg adult: tissue volumes
times their tissue:plasma Kp, plus the blood volume times the blood:plasma
ratio, plus the brain compartments at their unbound-concentration equilibrium.
The plasma-referenced value is then divided by the blood:plasma ratio to give
the blood-referenced value.

```{r vss}
r <- s_pgp[1, ]
tissues <- c("lung", "adipose", "bone", "heart", "kidney", "muscle", "skin", "spleen", "gut", "liver")
v_t <- unlist(r[paste0("v_", tissues)])
v_blood <- r$v_venous + r$v_arterial
# Brain blood holds blood concentration; brain mass and CSF equilibrate with
# the unbound blood concentration.
brain_vol <- function(bp, fu_bb, fu_bm, fu_c) {
  r$v_bb * bp + r$v_bm * fu_bb * bp / fu_bm + (r$v_ccsf + r$v_scsf) * fu_bb * bp / fu_c
}
vss_plasma <- sum(v_t * unlist(r[paste0("kp_", tissues)])) + r$v_other +
  v_blood * r$bp + brain_vol(r$bp, r$fu_bb, 0.5, 1)
vss_plasma_m6g <- sum(v_t * unlist(r[paste0("kp_", tissues, "_m6g")])) + r$v_other * r$kp_other_m6g +
  v_blood * r$bp_m6g + brain_vol(r$bp_m6g, r$fu_bb_m6g, 0.99, 1)

vss <- data.frame(
  analyte = c("Morphine", "M6G"),
  `Vss plasma-referenced (L)` = c(vss_plasma, vss_plasma_m6g),
  `Vss blood-referenced (L)` = c(vss_plasma / r$bp, vss_plasma_m6g / r$bp_m6g),
  `Paper (L)` = c(330, 30),
  check.names = FALSE
)
knitr::kable(vss, digits = 1)
stopifnot(abs(vss$`Vss blood-referenced (L)`[1] / 330 - 1) < 0.15)
```

The blood-referenced morphine Vss is within about 4% of the published 330 L.
For M6G the two references bracket the published 30 L (22 L plasma, 35 L
blood). The difference is consistent with the paper's unstated reference
individual and matrix, so the M6G value is reported without a pass/fail check.

# PKNCA validation

Plasma morphine and M6G NCA by arm. Elimination acts on venous blood, so the
plasma clearance from NCA should equal the model's blood clearance times the
blood:plasma ratio. Both sides use the same drawn parameters, so the check is
tight.

```{r pknca}
# Single-dose arms only (the continuous-infusion arm has no terminal phase)
sim_bolus <- sim[!grepl("0.03 mg/kg/h", sim$arm), ]
dose_df <- ev[ev$evid == 1 & !grepl("0.03 mg/kg/h", ev$arm), c("id", "arm", "time", "amt")]

run_nca <- function(conc_col, analyte) {
  conc_df <- sim_bolus |>
    transmute(id, arm, time, conc = .data[[conc_col]]) |>
    filter(!is.na(conc))
  o_conc <- PKNCA::PKNCAconc(conc_df, conc ~ time | arm + id)
  o_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)
  intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
                          half.life = TRUE, cl.obs = TRUE)
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
  as.data.frame(res) |> mutate(analyte = analyte)
}

nca <- bind_rows(run_nca("Cc", "Morphine"), run_nca("Cc_m6g", "M6G"))

nca |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  group_by(analyte, arm, PPTESTCD) |>
  summarise(value = signif(median(PPORRES, na.rm = TRUE), 3), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = value) |>
  rename(
    Analyte = analyte, Arm = arm, `Cmax (mg/L)` = cmax, `Tmax (h)` = tmax,
    `AUC0-inf (h*mg/L)` = aucinf.obs, `t1/2 (h)` = half.life
  ) |>
  knitr::kable(caption = "Median plasma NCA parameters of the virtual cohorts.")

cl_model <- sim_bolus |>
  distinct(id, cl, bp) |>
  transmute(id, cl_plasma = cl * bp)
cl_chk <- nca |>
  filter(analyte == "Morphine", PPTESTCD == "cl.obs") |>
  select(id, cl_nca = PPORRES) |>
  inner_join(cl_model, by = "id") |>
  mutate(ratio = cl_nca / cl_plasma)
summary(cl_chk$ratio)
stopifnot(abs(median(cl_chk$ratio) - 1) < 0.03, quantile(abs(cl_chk$ratio - 1), 0.9) < 0.06)
```

The paper reports no NCA parameters, so there is no published table to compare
against. The plasma clearance from NCA matches the model's blood clearance times
the blood:plasma ratio. This confirms that the dose, clearance and output units
fit together.

# Receptor-occupancy output

The model converts unbound brain-mass concentrations into competitive mu-opioid
receptor occupancy (Methods Eqs 4-5) and then into a sigmoid relative response
(Eq 6), with separate BR50 and Hill values for morphine and M6G. The curve
below is the relative response for each analyte alone, as a function of its
receptor occupancy, which is the relationship shown in S5 Fig.

```{r occupancy-curve}
occ <- seq(0, 100, by = 1)
data.frame(
  occupancy = rep(occ, 2),
  response = c(occ^4.217 / (occ^4.217 + 59.26^4.217), occ^2.2 / (occ^2.2 + 17^2.2)),
  analyte = rep(c("Morphine (BR50 59.26, Hill 4.217)", "M6G (BR50 17, Hill 2.2)"), each = length(occ))
) |>
  ggplot(aes(occupancy, response, colour = analyte)) +
  geom_line() +
  labs(x = "Receptors bound (%)", y = "Relative response (fraction)", colour = NULL)
```

The time course of the morphine and M6G contributions after a 28 mg dose in a
typical adult replicates the split shown in Figure 3B: the parent compound
carries most of the response in the first hours.

```{r fig3b}
ev_28 <- make_events(typ_adult, 28, iv_dur, obs_24)
s28 <- rxode2::rxSolve(mod_typ, ev_28, returnType = "data.frame")
s28 |>
  select(time, Morphine = rr, M6G = rr_m6g) |>
  pivot_longer(-time, names_to = "analyte", values_to = "response") |>
  ggplot(aes(time, response, colour = analyte)) +
  geom_line() +
  labs(x = "Time (h)", y = "Relative response (fraction)", colour = NULL)

stopifnot(max(s28$rr) > max(s28$rr_m6g))
```

# Assumptions and deviations

* **Adult physiology.** The deposited code is the paediatric (2.6-16.4 y)
  version. Adult organ volumes, flows, haematocrit and tissue composition come
  from the adult column of the 2019 framework S1 Table, which the paper names
  as the source for its adult version.
* **Age boundaries.** Adult equations apply from 18 y. Neonatal clearance, M6G
  formation fraction and the 10 mL/h CSF production apply below 0.25 y; the
  paper does not state its neonatal cut-off.
* **Pgp maturation shape.** The paper gives two anchors (41% of adult at term,
  100% at 6 months). Linear interpolation between them is assumed.
* **Adult virtual demographics.** The deposited code has no adult height or
  weight generator. Adults here have height N(178, 7) cm (men) or N(165, 7) cm
  (women) and log-normal weight with medians of 80 kg and 66 kg and a 15% CV.
  The paper's sex ratios were not reported in the article, so every arm is 50%
  female.
* **Dose administration.** "Single IV dose" is simulated as a 5-min infusion
  into venous blood. This keeps the early peaks seen at the start of the
  Figure 2A and 2B profiles, and avoids the artificial venous spike a true
  bolus puts into the 2-3 L venous compartment at time zero. The deposited
  example code uses a 1-h infusion (`tend = 1`) for its own scenario. The paediatric ECF scenario (0.03 mg/kg/h) is simulated as a
  24-h infusion for the cohort and a 100-h infusion for the typical-child
  check; the infusion length in the source study is not given in the article.
* **Clearance placement.** The 2021 S1 File subtracts the cleared morphine
  (CL x venous concentration) from the arterial compartment. The model here
  removes it from venous blood, as in the 2019 framework code. The section
  "Clearance placement" above shows that the deposited form falls below the
  paper's own Figures 2A and 2D, while the venous form reproduces them. The
  same placement is used for M6G.
* **CSF sampling site.** Ventricular-drain samples (adults) are compared with
  the cranial CSF compartment, and lumbar-puncture samples (children) with the
  spinal CSF compartment.
* **Cohort size.** The paper simulated 500 individuals per scenario; this
  vignette uses 100 per arm.
* **Observed data.** The observed verification points of Figures 1-2 were not
  digitised. The simulated typical-value curves of Figure 2D and the median
  plateaus of the Figure 2A 9.5-year-old panels were digitised by the
  maintainers from the published figure, to check the clearance placement.
* **IIV.** The model's random effects reproduce the variability of the
  deposited virtual-population code (log-normal SDs as variances). The female
  haematocrit SD (0.071) is reached by rescaling the shared haematocrit eta.
* **Residual error.** None was reported. `propSd` is fixed to 0 so the model
  can be used for estimation-style workflows.
* **Pharmacodynamics.** The model returns the relative response
  (`effect_rel`, the sum of the morphine and M6G terms). The paper multiplies it
  by the maximum effect observed in each clinical study; that study-specific
  scaling is not part of the model.
