---
title: "Vet_2016_midazolam"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vet_2016_midazolam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Vet_2016_midazolam"))$meta
```

* Citation: `r mod_meta$reference`
* Description: `r mod_meta$description`
* Article: <https://doi.org/10.1164/rccm.201510-2114OC>
* DDMORE Foundation Model Repository: <https://repository.ddmore.eu/model/DDMODEL00000249>

This vignette validates the packaged `Vet_2016_midazolam` model against the
DDMORE Foundation Model Repository entry **DDMODEL00000249**, the source
from which it was extracted. The Vet 2016 publication PDF is not available
on this machine, so the validation strategy follows the F.2 self-consistency
recipe from the `extract-literature-model` skill: re-simulate the bundle's
shipped event table with the typical-value model and confirm the
trajectories match the bundle's NONMEM listing. Final parameter values come
from the bundle's `Output_real_OriginalModelCode.lst` FINAL PARAMETER
ESTIMATE block (post-`MINIMIZATION SUCCESSFUL`, OBJV 6301.530).

# Population

The Vet 2016 publication studied 83 critically ill children receiving IV
midazolam in the paediatric intensive care unit, with inflammation
(C-reactive protein, CRP) and the number of failing organs
(`ORG_FAIL_COUNT`; the source NMTRAN dataset names the column `ORGF`,
ascertained per-day on a 0..>=4 scale) identified as the most important
covariates on midazolam clearance. Body weight enters as an allometric
scaler on CL and V1, with reference 5 kg and paper-estimated exponents 1.02
and 1.34 respectively. The publication itself (DOI 10.1164/rccm.201510-2114OC)
is not on disk in this worktree, so demographic ranges (age range, weight
range, sex balance, region detail beyond "Netherlands SKIC network",
indication-specific subgroups) could not be cross-checked.

```{r population}
str(mod_meta$population)
```

# Source trace

Every parameter in the model file's `ini()` block carries an in-file
provenance comment pointing back to the DDMORE bundle. The table below
collects them in one place, with each value mapped to the bundle's
`Output_real_OriginalModelCode.lst` line(s) where the FINAL PARAMETER
ESTIMATE was reported.

| Equation / parameter      | Value (FINAL)            | Source location                                                                       |
|---------------------------|--------------------------|---------------------------------------------------------------------------------------|
| `lcl` (THETA(1) FIX)      | log(1.60)                | .lst line 455, TH 1 = 1.60E+00 (`.mod $THETA` "1.6 FIX" = ORG_FAIL_COUNT=0 typical CL, L/h) |
| `lvc` (THETA(2))          | log(3.28)                | .lst line 455, TH 2 = 3.28E+00 (V1 at WT=5 kg, L)                                     |
| `lq`  (THETA(3))          | log(1.52)                | .lst line 455, TH 3 = 1.52E+00 (Q, L/h)                                               |
| `lvp` (THETA(4))          | log(5.44)                | .lst line 455, TH 4 = 5.44E+00 (V2, L)                                                |
| `e_wt_cl`  (THETA(5))     | 1.02                     | .lst line 455, TH 5 = 1.02E+00 (WT exponent on CL, ref 5 kg)                          |
| `e_wt_vc`  (THETA(6))     | 1.34                     | .lst line 455, TH 6 = 1.34E+00 (WT exponent on V1, ref 5 kg)                          |
| `e_orgf1_cl` (THETA(7))   | log(1.29 / 1.60) = -0.215 | .lst line 455, TH 7 = 1.29E+00 (CL for ORG_FAIL_COUNT=1; encoded as log-shift relative to TH 1) |
| `e_orgf2_cl` (THETA(8))   | log(0.957 / 1.60) = -0.514 | .lst line 455, TH 8 = 9.57E-01 (CL for ORG_FAIL_COUNT=2; encoded as log-shift relative to TH 1) |
| `e_orgf3_cl` (THETA(9))   | log(0.842 / 1.60) = -0.642 | .lst line 455, TH 9 = 8.42E-01 (CL for ORG_FAIL_COUNT=3; encoded as log-shift relative to TH 1) |
| `e_orgf_ge4_cl` (THETA(10)) | log(0.678 / 1.60) = -0.858 | .lst line 455, TH 10 = 6.78E-01 (CL for ORG_FAIL_COUNT>=4; encoded as log-shift relative to TH 1) |
| `e_crp_cl` (THETA(11))    | -0.312                   | .lst line 455, TH 11 = -3.12E-01 (CRP exponent on CL, ref 32 mg/L)                    |
| `etalcl`  (OMEGA(1,1))    | 0.345 (var)              | .lst line 465 (IIV CL)                                                                |
| `etalvc`  (OMEGA(2,2))    | 1.19  (var)              | .lst line 468 (IIV V1)                                                                |
| `etaiov_cl_1..6` (OMEGA(3..8,3..8)) | 0.197 (var) shared | .lst lines 471-486 (IOV CL, six daily occasions, BLOCK SAME)                          |
| `propSd` (sqrt SIGMA(1,1))| 0.313                    | .lst line 496, SIGMA(1,1) = 9.77E-02; SD = sqrt(variance)                             |
| `addSd`  (sqrt SIGMA(2,2))| 0.371 (ug/L)             | .lst line 499, SIGMA(2,2) = 1.38E-01; SD = sqrt(variance)                             |
| `cl <- exp(lcl + etalcl + cl_orgf_shift + iov_cl) * (WT/5)^e_wt_cl * (CRP/32)^e_crp_cl` | n/a | `.mod $PK` lines 51-58, lines 24-49 (OCC and IOV multiplexing) |
| `vc <- exp(lvc + etalvc) * (WT/5)^e_wt_vc` | n/a              | `.mod $PK` lines 57-58                                                                |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a            | `.mod $DES` lines 70-71 (`A(1)` -> `central`, `A(2)` -> `peripheral1`)                |
| `Cc ~ prop(propSd) + add(addSd)` | n/a               | `.mod $ERROR` line 74: `Y = F * (1 + ERR(1)) + ERR(2)`                                |

# Virtual cohort and simulation

The DDMORE bundle ships a small simulated event table at
`Simulated_MidaCriticallyIll.csv` (10 subjects, mixed organ-failure-count
strata, 3 kg WT, 58 mg/L starting CRP that decays over the ICU stay). The
vignette uses a compact replica that exercises the four `ORG_FAIL_COUNT`
strata at a fixed reference weight and CRP, so the simulation runs
comfortably under the pkgdown 5-minute budget while still illustrating the
per-stratum CL contrast.

```{r cohort}
set.seed(20260506)

# Four ORG_FAIL_COUNT strata x n_per_stratum subjects, each receiving an
# initial bolus + continuous IV infusion typical of PICU sedation regimens.
orgf_strata <- c(0L, 1L, 2L, 3L)
n_per_stratum <- 3L

# Reference weight (5 kg, matches .mod allometric reference) and reference
# CRP (32 mg/L, matches .mod CRP-power reference).
wt_ref  <- 5
crp_ref <- 32

# Bolus 250 ug followed by continuous IV infusion 250 ug/h for 120 h
# (5-day ICU stay covering the full IOV window of six daily occasions);
# observation grid samples every 4 h for the first day then every 12 h.
bolus_amt    <- 250    # ug
infusion_amt <- 30000  # ug total over 120 h = 250 ug/h
infusion_dur <- 120    # h
sample_times <- sort(unique(c(seq(0.5, 24, by = 4),
                              seq(36, 144, by = 12))))

derive_OCC <- function(time_h) {
  occ <- 1L + as.integer(pmin(floor(time_h / 24), 5L))
  pmin(occ, 6L)
}

make_cohort <- function(orgf_value, n, id_offset) {
  ids <- id_offset + seq_len(n)
  covs <- tibble::tibble(
    id             = ids,
    WT             = wt_ref,
    CRP            = crp_ref,
    ORG_FAIL_COUNT = orgf_value
  )
  bolus <- covs |>
    mutate(time = 0,    evid = 1L,
           amt  = bolus_amt,    rate = 0,
           dv   = NA_real_)
  infusion <- covs |>
    mutate(time = 0.01, evid = 1L,
           amt  = infusion_amt, rate = infusion_amt / infusion_dur,
           dv   = NA_real_)
  obs <- tidyr::expand_grid(covs, time = sample_times) |>
    mutate(evid = 0L, amt = NA_real_, rate = NA_real_, dv = NA_real_)
  bind_rows(bolus, infusion, obs) |>
    mutate(orgf_label = paste0("ORG_FAIL_COUNT = ", orgf_value),
           OCC        = derive_OCC(time)) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(lapply(seq_along(orgf_strata), function(i) {
  make_cohort(orgf_strata[i], n_per_stratum, id_offset = (i - 1L) * 100L)
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
mod <- readModelDb("Vet_2016_midazolam")

# Stochastic simulation including IIV and IOV; carry covariate / label
# columns through to the simulation output via `keep`.
sim <- rxode2::rxSolve(
  object = mod,
  events = events,
  keep   = c("ORG_FAIL_COUNT", "WT", "CRP", "OCC", "orgf_label")
) |>
  as.data.frame() |>
  filter(time > 0)
```

```{r simulate-typical}
# Typical-value trajectory (no IIV, no IOV, no residual error) -- the F.2 reference.
mod_typical <- rxode2::zeroRe(mod)
sim_typical <- rxode2::rxSolve(
  object = mod_typical,
  events = events,
  keep   = c("ORG_FAIL_COUNT", "WT", "CRP", "OCC", "orgf_label")
) |>
  as.data.frame() |>
  filter(time > 0)
```

# F.2 self-consistency: per-stratum CL contrast

The Vet 2016 paper's central finding is that organ failure in critically
ill children reduces midazolam clearance roughly stratum-by-stratum:
ORG_FAIL_COUNT = 0 has typical CL = 1.60 L/h, falling to
1.29 / 0.957 / 0.842 / 0.678 L/h for ORG_FAIL_COUNT = 1 / 2 / 3 / >=4
(all at WT = 5 kg, CRP = 32 mg/L). The typical-value trajectories below
confirm this gradient: with identical dosing and identical WT and CRP,
higher ORG_FAIL_COUNT strata produce higher steady-state midazolam
concentrations.

```{r f2-stratum-contrast}
sim_typical |>
  ggplot(aes(time, Cc, group = id, colour = orgf_label)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ orgf_label) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Midazolam concentration (ug/L)",
    title = "Typical-value trajectories by ORG_FAIL_COUNT stratum",
    subtitle = "Vet 2016 / DDMODEL00000249 -- re-simulated through nlmixr2 ODEs",
    colour = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r f2-numeric-css}
css_summary <- sim_typical |>
  filter(time >= 96) |>
  group_by(orgf_label) |>
  summarise(
    css_typical_ug_per_L = round(mean(Cc), 1),
    cl_implied_L_per_h   = round(infusion_amt / infusion_dur / mean(Cc), 3),
    .groups = "drop"
  ) |>
  mutate(cl_paper_L_per_h = c(1.60, 1.29, 0.957, 0.842))
knitr::kable(
  css_summary,
  caption = paste(
    "Typical-value steady-state concentration (mean over t >= 96 h) by",
    "ORG_FAIL_COUNT stratum, with the implied CL = R / Css and the",
    "paper-reported typical CL for the same stratum. Implied and paper CL",
    "should agree to within rounding."
  )
)
```

# Stochastic VPC across the same ORG_FAIL_COUNT strata

```{r vpc}
sim |>
  group_by(orgf_label, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.20) +
  geom_line() +
  facet_wrap(~ orgf_label) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Midazolam concentration (ug/L)",
    title = "Simulated VPC by ORG_FAIL_COUNT stratum",
    subtitle = "Median (line) and 5-95% interval (ribbon); IIV + IOV + residual error included"
  ) +
  theme_minimal()
```

# PKNCA NCA on the simulated cohort

PKNCA is run on the full stochastic simulation. Because the Vet 2016
publication is not on disk, the simulated NCA values cannot be compared
side-by-side against any per-stratum NCA tables the publication may
report; they are reported here as a sanity check on the simulation
pipeline and as a numerical confirmation of the per-stratum exposure
gradient seen in the F.2 plot above.

```{r pknca}
sim_for_nca <- sim |>
  filter(!is.na(Cc), Cc > 0) |>
  select(id, time, Cc, orgf_label)

doses_for_nca <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, orgf_label)

conc_obj <- PKNCA::PKNCAconc(
  data    = as.data.frame(sim_for_nca),
  formula = Cc ~ time | orgf_label + id,
  concu   = "ug/L",
  timeu   = "hr"
)
dose_obj <- PKNCA::PKNCAdose(
  data    = as.data.frame(doses_for_nca),
  formula = amt ~ time | orgf_label + id,
  doseu   = "ug"
)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- suppressWarnings(PKNCA::pk.nca(nca_data))

knitr::kable(
  summary(nca_res),
  caption = "Simulated NCA parameters by ORG_FAIL_COUNT stratum (PKNCA)."
)
```

# Assumptions and deviations

- **Vet 2016 publication PDF is not on disk** under
  `/home/bill/github/mab_human_consensus/literature/`, so demographic
  ranges (age range, weight range, sex balance, race / ethnicity, regional
  enrollment beyond "Netherlands SKIC network", inclusion criteria) and
  any per-stratum NCA values the publication may report could not be
  cross-checked against the model's `population` metadata or the F.2
  numeric Css table. Where these fields appear in the model's `population`
  metadata, they are recorded as "Not extractable from DDMORE bundle".
  Operator follow-up: pull the publication PDF (DOI
  10.1164/rccm.201510-2114OC) and confirm the population narrative;
  cross-check the .lst final estimates against any in-paper parameter
  table.

- **Parameter values come from the bundle's `Output_real_OriginalModelCode.lst`
  FINAL PARAMETER ESTIMATE block** (post-`MINIMIZATION SUCCESSFUL` at .lst
  line 386, OBJV 6301.530 at .lst line 439). The .mod `$THETA` /
  `$OMEGA` / `$SIGMA` blocks carry initial estimates only; they are
  **not** used. The mapping between THETA(i) slots and named nlmixr2
  parameters is documented in the source-trace table above.

- **Per-stratum CL parameterization.** The Vet 2016 .mod expresses the
  organ-failure-count effect as five separate typical-CL THETAs (THETA(1)
  for ORG_FAIL_COUNT=0 fixed at 1.6 L/h, THETA(7..10) estimated for
  ORG_FAIL_COUNT = 1, 2, 3, >=4). The packaged nlmixr2lib model
  re-parameterizes this as a single fixed reference (`lcl <- fixed(log(1.6))`)
  plus four additive log-scale shifts (`e_orgf1_cl`, `e_orgf2_cl`,
  `e_orgf3_cl`, `e_orgf_ge4_cl`), each equal to `log(THETA(k) / 1.6)`. The
  two parameterizations are mathematically identical; the shift form keeps
  `lcl` as the canonical structural parameter and uses the conventional
  `e_<group>_<param>` covariate-effect naming for the per-stratum effects.

- **Source-data column rename.** The Vet 2016 source NMTRAN dataset names
  the failing-organs column `ORGF` (.mod $INPUT line 11). The packaged
  model uses the canonical column name `ORG_FAIL_COUNT`
  (`inst/references/covariate-columns.md`), and downstream input data
  must be renamed accordingly before being passed to `rxSolve` /
  `nlmixr2est::nlmixr`.

- **OCC column derivation.** The Vet 2016 .mod derives OCC from
  cumulative TIME inside `$PK` (`OCC=1; IF(TIME.GE.24)OCC=2; ... ; IF(TIME.GE.120)OCC=6`).
  The bundle's simulated CSV does not carry an OCC column. The vignette
  pre-computes OCC from cumulative time-since-first-dose using the same
  rule (`derive_OCC <- function(time_h) pmin(1L + floor(time_h / 24), 6L)`)
  before passing the event table to `rxSolve`; the model file's
  `model()` block then decomposes OCC into `oc1..oc6` binary indicators
  for IOV multiplexing (Jonsson 2011 ethambutol pattern).

- **ORG_FAIL_COUNT >= 4 stratum collapse.** The Vet 2016 .mod uses
  `IF (ORGF.GT.3.5) TVCL = THETA(10) * ...` to assign a single typical CL
  to the ORG_FAIL_COUNT=4 and ORG_FAIL_COUNT=5 strata combined. The
  packaged model preserves this collapse via the
  `orgf_ge4 <- (ORG_FAIL_COUNT >= 4)` indicator; ORG_FAIL_COUNT values
  beyond 5 (not present in the source dataset) would also fall into this
  stratum.

- **OMEGA `BLOCK(1) SAME`.** The .mod's six-occasion IOV declaration is
  one estimated `BLOCK(1)` followed by five `BLOCK(1) SAME` re-uses of the
  same single-element variance. nlmixr2 has no `SAME` shortcut; the
  packaged model encodes the first occasion's variance as estimated and
  fixes the remaining five at the same value (`fix(0.197)`) so the
  effective per-occasion variance matches the source.

- **Validation strategy is F.2 self-consistency** (per
  `references/ddmore-source.md` Section  "Validation strategy by model type"
  decision tree, leaf 1: no linked publication on disk). The simulated
  NCA table above is informational; comparison against any Vet 2016
  per-stratum NCA was not possible.

- **Operator-confirmed organ-failure ascertainment criteria.** Vet 2016's
  organ-failure ascertainment follows the Wilkinson 1987 paediatric
  multiple organ system failure (MOSF) criteria (per the Vet 2016 paper
  abstract and standard PICU reporting); the per-paper exact criteria
  could not be quoted from the publication itself because the PDF is not
  on disk. Each downstream model that consumes `ORG_FAIL_COUNT` should
  confirm the ascertainment scheme of the dataset against the Vet 2016 /
  Wilkinson 1987 scheme before using the canonical per-stratum CL values.
