---
title: "Parathyroid hormone rhPTH(1-84) (Visscher 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parathyroid hormone rhPTH(1-84) (Visscher 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Visscher_2025_parathyroidHormone"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Article: <https://doi.org/10.1111/bcp.16342>\n")
cat("* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11992657/>\n")
cat("* Supplement (Tables S1-S4, Figures S1-S2): `BCP-91-1233-s003.docx`, retrieved through the EuropePMC full-text repository\n")
```

`r ui$description`

# Population

A single 43-year-old male patient with iatrogenic (postsurgical) chronic
hypoparathyroidism following total thyroidectomy for medullary thyroid carcinoma.
Conventional therapy with alfacalcidol, calcium carbonate and chlortalidone kept
his serum calcium adequately controlled but did not restore the
parathyroid-hormone-dependent renal reabsorption of calcium, and he developed
severe hypercalciuria with recurrent nephrolithiasis. He was therefore switched
to rhPTH(1-84) (Natpar) and studied intensively over 13 days across five dosing
frequencies: 100 ug once daily, 50 ug twice daily, 35 ug three times daily,
25 ug four times daily, and finally off-label continuous subcutaneous infusion
by insulin pump at 100, 75, 50 and 25 ug/day (Supporting Information Table S1).
Note that the first four regimens all deliver approximately the same total daily
dose (100, 100, 105 and 100 ug/day respectively) and differ only in frequency --
the study is a frequency comparison at constant daily dose, which the validation
below exploits as a closed-form check.

Serum calcium, phosphate and creatinine were sampled up to nine times daily
(Table S2) and urine was collected over 10 intervals per day (Table S3). Body
weight is not reported, so the model carries no allometric scaling. Endogenous
PTH before the first rhPTH(1-84) dose was 0.58 pmol/L. The Abstract describes
the patient as 42 years old and Methods section 2.1 as 43; the Methods value is
used here.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(ui$population)
```

# Model structure

**PK.** One compartment with subcutaneous absorption through an Erlang-5 transit
chain -- the dose site (`depot`) plus four transit compartments, all five
transfers governed by the single reported rate constant `ka` = 8.39 /h. Methods
section 2.3: *"to match the delay for the absorption of PTH, five transit
compartments were included in the absorption compartment based on the fitting of
the model to the data."* Plasma concentration carries a constant additive
endogenous baseline (`bl_pth` = 0.58 pmol/L, the Edsim++ `C01.Cb` object):

$$C_c = \frac{A_{central}}{V_c} + C_b$$

**Organ-function-scaled elimination.** Visscher 2025 Equation (3) splits total
clearance into a renal arm carrying the fraction excreted unchanged and a
hepatic remainder, each scaled by its own dimensionless relative-organ-function
covariate (1 = normal):

$$CL_i = CL \cdot \left( f_e \cdot \mathrm{RENALFUNC\_REL} + (1 - f_e) \cdot \mathrm{HEPFUNC\_REL} \right)$$

with `fe` = 0.3 fixed from Hruska et al's canine estimate that renal clearance
accounts for on average 30% of PTH total clearance. Methods 2.3 states that
*"LF is the relative liver function (equal to 1 in this model)"*; no numeric
values of RF are reported and no reference GFR is given from which the ratio
could be recovered, so both covariates are held at 1 throughout this vignette
and the decomposition is exercised as a sensitivity check instead.

**PD.** Two independent, analyte-specific effect arms hang off `Cc`. Each is a
two-state concentration-driven delay chain -- the paper's transfer compartment
(`T01`/`T02`) feeding a virtual central compartment (`C03`/`C02`), both
transfers driven by the single reported equilibration rate constant -- followed
by a sigmoid Emax function. Supporting Information Figure S1 caption: *"The
delay on the effect parameters was modeled by using a transfer compartment (T02
and T01 object) connected to a virtual central compartment (C02 and C03
object). The virtual compartments do not participate in the mass balance of the
model."* Figure 4B of the paper plots the effect against the **virtual**
compartment C03, which identifies the second state as the effect driver.

The readout is the renal clearance of phosphate (or calcium) expressed as a
percentage of creatinine clearance -- a fractional excretion. Equation (2),
expanded as Supplementary Equation (5), is

$$CL\,(\%) = \frac{CL_X}{CL_{creatinine}} = \frac{C^X_{urine} \cdot C^{creatinine}_{plasma}}{C^X_{plasma} \cdot C^{creatinine}_{urine}}$$

**Emax is an asymptote, not an increment.** The readout is encoded as

$$CL_X(\%) = E_0 + (E_{max} - E_0) \cdot \frac{C_e^{\,\gamma}}{EC_{50}^{\,\gamma} + C_e^{\,\gamma}}$$

The calcium arm forces this reading: Table S4 gives `E0` = 2.57% and `Emax` =
2.05%, so `Emax` lies *below* `E0`. Only the asymptote form produces the
*decrease* the Conclusion describes (*"a permanent increase in phosphate
clearance was observed which corresponds to a permanent reduction in the calcium
clearance"*); an increment reading `E0 + Emax * f` would drive calcium clearance
up to 4.62% instead. Under the asymptote reading phosphate runs
14.18% -> 50.62% and calcium 2.57% -> 2.05%, both correct in direction.

**No variability.** Table S4 reports no OMEGA and no SIGMA: the model was fitted
deterministically to one patient's intensive data in Edsim++. It is therefore
encoded with no `eta` terms and no residual-error model, and no `zeroRe()` or
`omega = NA` is needed (or valid) when simulating.

# Source trace

Every `ini()` value is also carried as an in-file comment next to its entry in
`inst/modeldb/specificDrugs/Visscher_2025_parathyroidHormone.R`. All 16 come
from a single table.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lfdepot` (F) | 0.53 (fixed) | Table S4, `I01.F`; Methods 2.3 "bioavailability was set at a literature value of 0.53" |
| `bl_pth` (Cb) | 0.58 pmol/L (fixed) | Table S4, `C01.Cb`; Methods 2.3 "an endogenous PTH concentration of 0.58 pmol/L ... used as the baseline level" |
| `fe` (NFE) | 0.3 (fixed) | Table S4, `ORG.nfe`; Methods 2.3 "for our base model an NFE of 0.3 was used" (from Hruska et al, canine) |
| `lka` | 8.39 /h (SE 2.81) | Table S4, `I02.k` |
| `lcl` | 39.93 L/h (SE 2.79) | Table S4, `O01.CL` |
| `lvc` | 120.71 L (SE 11.71) | Table S4, `C01.V` |
| `lke0_phosphate` | 0.45 /h (SE 0.08) | Table S4, `T01.k` |
| `e0_phosphate` | 14.18% (SE 4.72) | Table S4, `E01.E0` |
| `emax_phosphate` | 50.62% (SE 9.65) | Table S4, `E01.Emax` |
| `lec50_phosphate` | 6.35 pmol/L (SE 1.05) | Table S4, `E01.EC50`; main text "an EC50 of 6.3 pmol/L" |
| `hill_phosphate` | 3.82 (SE 2.18) | Table S4, `E01.j` |
| `lke0_calcium` | 4.16 /h (SE 63.19) | Table S4, `T02.k` (dagger footnote) |
| `e0_calcium` | 2.57% (SE 4.51) | Table S4, `E02.E0` (dagger footnote) |
| `emax_calcium` | 2.05% (SE 3.85) | Table S4, `E02.Emax` (dagger footnote) |
| `lec50_calcium` | 3.56 pmol/L (SE 34.88) | Table S4, `E02.EC50` (dagger footnote) |
| `hill_calcium` | 3.42 (SE 100.71) | Table S4, `E02.j` (dagger footnote) |
| Erlang-5 absorption chain | n/a | Methods 2.3, "five transit compartments were included in the absorption compartment" |
| `CLi = CL * (fe * RF + (1 - fe) * LF)` | n/a | Equation (3), page 1235 |
| Two-state effect delay per analyte | n/a | Figure S1 caption; Results "adding a biocompartment (virtual compartment) ... adding one transit compartment" |
| Fractional-excretion readout | n/a | Equation (2), page 1235; Supplementary Equation (5) |
| Emax-as-asymptote parameterisation | n/a | Derived: Table S4 calcium `Emax` < `E0`, reconciled with the Conclusion |
| Dosing schedule reproduced below | n/a | Supporting Information Table S1 |
| CL/F, Vss/F reference values | 75 (70-81) L/h; 228 (206-250) L | Table 1, "PKPD model 25-100 ug" column (value with standard error) |

The dagger footnote to Table S4 reads *"These parameters were hard to fit since
they were very sensitive for the initial value upon fitting."* Every calcium-arm
standard error exceeds its estimate, in three cases by more than an order of
magnitude, and the calcium goodness-of-fit R-squared is 0.06 against 0.66 for
phosphate (Figure 3). The calcium arm is reproduced faithfully but must not be
read as equal-confidence with the phosphate arm.

# Simulation setup

The model is encoded natively in the source's molar units (pmol, pmol/L), so no
molecular weight enters the model file. Table S1 states the clinical doses in
micrograms, so the vignette converts them here.

```{r units}
# NON-PAPER PROVENANCE: the molecular weight of rhPTH(1-84) is not stated in
# Visscher 2025. 9425 g/mol is the standard value for the 84-residue peptide.
# It is used ONLY in this vignette, never in the model file. The paper's own
# arithmetic corroborates it: see the "EC50 and the target dose" section, where
# this MW puts the fitted EC50 at 98.3 ug/day, inside the paper's stated
# "between 75 and 100 ug per day".
MW_PTH <- 9425                                        # g/mol
ug_to_pmol <- function(ug) ug * 1e-6 / MW_PTH * 1e12   # ug -> pmol
pmol_per_day_to_ug <- function(pmol) pmol * MW_PTH / 1e12 * 1e6

BL_PTH <- 0.58   # pmol/L, endogenous baseline (Table S4, C01.Cb)
CL_POP <- 39.93  # L/h    (Table S4, O01.CL)
V_POP  <- 120.71 # L      (Table S4, C01.V)
F_SC   <- 0.53   # -      (Table S4, I01.F)

# Trapezoidal time-average, used by the closed-form gates below. A plain
# arithmetic mean over a solver grid is a left-Riemann sum and is only accurate
# to O(h); the gates here are asserted to 0.5% so the integration must be
# better than that.
time_avg <- function(t, y) {
  stopifnot(length(t) == length(y), length(t) > 1L, !anyNA(t), !anyNA(y))
  sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2) / (max(t) - min(t))
}

# Solve helper. The model declares no eta terms, so `omega = NA` must NOT be
# passed (rxode2 would evaluate dim(NA)[1] and fail). Both organ-function
# covariates default to 1 = normal.
solve_pth <- function(events, keep = character()) {
  out <- rxode2::rxSolve(ui, events = events, keep = keep) |> as.data.frame()
  if (is.null(out$id)) out$id <- 1L   # rxSolve omits `id` for a single subject
  out
}
```

# The 13-day study regimen (Table S1)

```{r regimen}
# Day d at clock hour h maps to model time (d - 1) * 24 + h.
bolus <- tibble::tribble(
  ~day, ~ug, ~clock,
  1,  100, 8,
  2,  100, 8,
  3,   50, c(8, 20),
  4,   50, c(8, 20),
  5,   35, c(8, 15, 22),
  6,   35, c(8, 15, 22),
  7,   25, c(8, 12.5, 17, 22),
  8,   25, c(8, 12.5, 17, 22)
) |>
  tidyr::unnest(clock) |>
  dplyr::mutate(time = (day - 1) * 24 + clock,
                amt = ug_to_pmol(ug), dur = NA_real_)

infusion <- tibble::tribble(
  ~day_start, ~ug_per_day, ~dur,
  9,  100, 48,   # days 9-10
  11,  75, 24,
  12,  50, 24,
  13,  25, 24
) |>
  dplyr::mutate(time = (day_start - 1) * 24,
                amt = ug_to_pmol(ug_per_day * dur / 24))

# Total delivered over the 13 days, as a bookkeeping check against Table S1.
total_ug <- sum(bolus$ug) + sum(infusion$ug_per_day * infusion$dur / 24)
total_ug

dose_rows <- dplyr::bind_rows(
  bolus |> dplyr::select(time, amt, dur),
  infusion |> dplyr::select(time, amt, dur)
) |>
  dplyr::mutate(evid = 1L, cmt = "depot")

# Observation rows are written on the ODE state `central`, never on the
# algebraic observable `Cc` -- naming an observable in `cmt` would auto-inject a
# compartment slot and renumber every state.
obs_rows <- tibble::tibble(
  time = seq(0, 14 * 24, by = 0.05),
  amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central"
)

events_study <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::mutate(id = 1L, RENALFUNC_REL = 1, HEPFUNC_REL = 1)

sim_study <- solve_pth(events_study)

regimen_of <- function(day) {
  dplyr::case_when(
    day %in% 1:2   ~ "QD 100 ug",
    day %in% 3:4   ~ "BID 50 ug",
    day %in% 5:6   ~ "TID 35 ug",
    day %in% 7:8   ~ "QID 25 ug",
    day %in% 9:10  ~ "Continuous 100 ug/day",
    day == 11      ~ "Continuous 75 ug/day",
    day == 12      ~ "Continuous 50 ug/day",
    day == 13      ~ "Continuous 25 ug/day",
    TRUE           ~ NA_character_
  )
}

sim_study <- sim_study |>
  dplyr::mutate(day = floor(time / 24) + 1,
                regimen = regimen_of(day),
                regimen = factor(regimen, levels = c(
                  "QD 100 ug", "BID 50 ug", "TID 35 ug", "QID 25 ug",
                  "Continuous 100 ug/day", "Continuous 75 ug/day",
                  "Continuous 50 ug/day", "Continuous 25 ug/day")))
```

## Figure 2 -- PTH concentration, calcium clearance and phosphate clearance

```{r figure-2, fig.height = 8}
# Replicates Figure 2 of Visscher 2025: (A) simulated PTH concentration in the
# central compartment, (B) simulated calcium clearance, (C) simulated phosphate
# clearance, all against time, with vertical dashed lines at the dosing-regimen
# changes of Table S1.
regimen_breaks <- c(2, 4, 6, 8, 10, 11, 12) * 24 / 24  # in days

panel_data <- sim_study |>
  dplyr::filter(time <= 13 * 24) |>
  dplyr::transmute(
    day_time = time / 24,
    `A: PTH concentration (pmol/L)` = Cc,
    `B: Calcium clearance (% of creatinine clearance)` = clrel_calcium,
    `C: Phosphate clearance (% of creatinine clearance)` = clrel_phosphate
  ) |>
  tidyr::pivot_longer(-day_time, names_to = "panel", values_to = "value")

ggplot(panel_data, aes(day_time, value)) +
  geom_vline(xintercept = regimen_breaks, linetype = "dashed",
             colour = "grey55", linewidth = 0.3) +
  geom_line(linewidth = 0.4) +
  facet_wrap(~panel, ncol = 1, scales = "free_y") +
  labs(x = "Time (days)", y = NULL,
       caption = paste("Replicates Figure 2 of Visscher 2025.",
                       "Dashed lines mark the dosing-regimen changes of Table S1."))
```

The simulation reproduces the paper's central qualitative finding: under
intermittent dosing the PTH concentration swings widely and never settles, and
the phosphate-clearance elevation therefore decays back toward baseline between
doses. From day 9 onward, with continuous infusion, PTH is flat and the
phosphate-clearance increase (and the mirror-image calcium-clearance decrease)
becomes permanent.

## Quantitative gate -- same daily dose, different frequency

Because the PK is linear and the first four regimens deliver essentially the
same total daily dose, the time-averaged concentration over a day at
steady state must satisfy

$$\overline{C_c} = C_b + \frac{D_{daily} \cdot F}{CL \cdot 24}$$

independent of how the daily dose is split. What frequency changes is the
peak-to-trough swing -- and, because the sigmoid is steep (Hill 3.82), the
*minimum* phosphate clearance reached between doses. That minimum is exactly
what the paper's "permanent versus transient effect" claim is about.

The identity holds only at steady state. Table S1 gives each of the first five
regimens two days, so the second day qualifies; the final three continuous steps
(75, 50 and 25 ug/day) last one day each and taper down from the preceding
higher level, so for those the converged quantity is the end-of-day
concentration rather than the daily average, and that is what is tested below.

```{r same-daily-dose}
daily_ug <- c("QD 100 ug" = 100, "BID 50 ug" = 100, "TID 35 ug" = 105,
              "QID 25 ug" = 100, "Continuous 100 ug/day" = 100,
              "Continuous 75 ug/day" = 75, "Continuous 50 ug/day" = 50,
              "Continuous 25 ug/day" = 25)

# Use the SECOND day of each two-day intermittent block (and the single day of
# each tapering continuous step) so the profile is as close to steady state as
# the study design allows.
gate_days <- c("QD 100 ug" = 2, "BID 50 ug" = 4, "TID 35 ug" = 6,
               "QID 25 ug" = 8, "Continuous 100 ug/day" = 10,
               "Continuous 75 ug/day" = 11, "Continuous 50 ug/day" = 12,
               "Continuous 25 ug/day" = 13)

stopifnot(setequal(names(daily_ug), names(gate_days)))

# Days 1-10 give each regimen two days, so the second day is at (or, given the
# 2.1 h half-life, indistinguishable from) steady state and its 24-hour average
# is the invariant to test. Days 11-13 are single 24-hour infusion steps that
# TAPER down from the preceding higher level, so their 24-hour average is a
# transient, not a steady state -- for those the quantity that has converged by
# the end of the day is the concentration itself, so the end-of-day value is
# tested against that step's closed-form Css instead.
at_steady_state <- c("QD 100 ug" = TRUE, "BID 50 ug" = TRUE, "TID 35 ug" = TRUE,
                     "QID 25 ug" = TRUE, "Continuous 100 ug/day" = TRUE,
                     "Continuous 75 ug/day" = FALSE,
                     "Continuous 50 ug/day" = FALSE,
                     "Continuous 25 ug/day" = FALSE)
stopifnot(setequal(names(at_steady_state), names(gate_days)))

swing_tab <- lapply(names(gate_days), function(rg) {
  # Hoist the lookups BEFORE tibble(): inside tibble() a just-created column
  # masks a same-named object from the enclosing scope, so a column called
  # `daily_ug` would shadow the lookup vector for any later expression.
  day_i <- gate_days[[rg]]
  ug_i <- daily_ug[[rg]]
  ss_i <- at_steady_state[[rg]]
  w <- sim_study |> dplyr::filter(time >= (day_i - 1) * 24, time <= day_i * 24)
  stopifnot(nrow(w) > 100L)   # a gate with no rows cannot go red
  closed_form <- BL_PTH + ug_to_pmol(ug_i) * F_SC / (CL_POP * 24)
  tibble::tibble(
    regimen = rg,
    daily_ug = ug_i,
    steady_state = ss_i,
    tested_quantity = if (ss_i) "24 h average" else "end-of-day value",
    simulated = if (ss_i) time_avg(w$time, w$Cc) else w$Cc[which.max(w$time)],
    closed_form = closed_form,
    swing_Cc = max(w$Cc) - min(w$Cc),
    min_phosphate = min(w$clrel_phosphate),
    max_calcium = max(w$clrel_calcium)
  )
}) |> dplyr::bind_rows() |>
  dplyr::mutate(pct_diff = 100 * (simulated - closed_form) / closed_form)

# Both forms of the identity must hold, to the accuracy actually achieved.
stopifnot(
  nrow(swing_tab) == 8L,
  sum(swing_tab$steady_state) == 5L,
  max(abs(swing_tab$pct_diff)) < 0.1
)
# The swing must fall monotonically as the same daily dose is split more finely,
# and must be essentially zero under continuous infusion.
same_dose <- swing_tab |>
  dplyr::filter(regimen %in% c("QD 100 ug", "BID 50 ug", "QID 25 ug",
                               "Continuous 100 ug/day"))
stopifnot(nrow(same_dose) == 4L,
          all(diff(same_dose$swing_Cc) < 0),
          dplyr::last(same_dose$swing_Cc) < 0.01)

swing_tab |>
  dplyr::select(-steady_state) |>
  dplyr::mutate(dplyr::across(c(simulated, closed_form, swing_Cc,
                                min_phosphate, max_calcium), \(x) round(x, 3)),
                pct_diff = round(pct_diff, 4)) |>
  dplyr::rename(
    "Regimen" = regimen,
    "Daily dose (ug)" = daily_ug,
    "Quantity tested" = tested_quantity,
    "Simulated (pmol/L)" = simulated,
    "Closed form (pmol/L)" = closed_form,
    "Difference (%)" = pct_diff,
    "Peak-to-trough swing (pmol/L)" = swing_Cc,
    "Minimum phosphate CL (%)" = min_phosphate,
    "Maximum calcium CL (%)" = max_calcium
  ) |>
  knitr::kable(caption = paste(
    "Same-daily-dose identity and the collapse of the peak-to-trough swing.",
    "The first four regimens deliver 100-105 ug/day and share one",
    "daily-average concentration; only the swing changes. Drug-free reference",
    "values are 14.18% for phosphate clearance and 2.57% for calcium",
    "clearance."
  ))
```

Reading the last two columns against the drug-free values (`e0_phosphate` =
14.18%, `e0_calcium` = 2.57%) is the quantitative form of the paper's
conclusion. Under 100 ug once daily the phosphate clearance falls all the way
back to 14.19% between doses -- the effect is fully transient. Splitting the
same 100 ug across more administrations raises the trough progressively, and
continuous infusion of the identical daily dose holds phosphate clearance at
approximately 32.9% with essentially no fluctuation, while calcium clearance is
held down at 2.11% instead of returning to 2.57%. That is what the paper means
by *"a permanent increase in phosphate clearance ... which corresponds to a
permanent reduction in the calcium clearance"*.

## Figure 4 -- hysteresis and the virtual compartment

```{r figure-4, fig.height = 3.6}
# Replicates Figure 4 of Visscher 2025: phosphate clearance plotted against the
# PTH concentration in (A) the central compartment C01, which shows hysteresis,
# and (B) the virtual compartment C03, which collapses the loop onto the
# underlying sigmoid. Uses the single 100 ug dose given on day 2 (model time 32 h)
# through to the start of the day-3 regimen, so the trace is one clean
# dose-to-dose cycle.
DOSE2_T <- 32
loop <- sim_study |> dplyr::filter(time >= DOSE2_T, time < DOSE2_T + 24)
stopifnot(nrow(loop) > 100L)

hyst <- dplyr::bind_rows(
  loop |> dplyr::transmute(conc = Cc,
                           panel = "A: central compartment (C01)",
                           clrel_phosphate, tad = time - DOSE2_T),
  loop |> dplyr::transmute(conc = effect_phosphate,
                           panel = "B: virtual compartment (C03)",
                           clrel_phosphate, tad = time - DOSE2_T)
)

ggplot(hyst, aes(conc, clrel_phosphate)) +
  geom_path(aes(colour = tad), linewidth = 0.5) +
  facet_wrap(~panel, scales = "free_x") +
  scale_colour_viridis_c(name = "Hours\nafter dose") +
  labs(x = "PTH concentration (pmol/L)",
       y = "Phosphate clearance (% of creatinine CL)",
       caption = "Replicates Figure 4 of Visscher 2025.")
```

```{r hysteresis-gate}
# Quantify what the figure shows. A hysteresis loop means the rising and the
# falling limb of the trace assign DIFFERENT effects to the same abscissa value.
# Split the cycle at the abscissa's maximum, interpolate both limbs onto a
# common grid, and take the largest disagreement. A single-valued curve (no
# hysteresis) gives ~0; a loop gives its width in percentage points.
branch_gap <- function(x, y, n = 200) {
  i <- which.max(x)
  up <- seq_len(i)
  dn <- i:length(x)
  if (length(up) < 5L || length(dn) < 5L) {
    stop("branch_gap: a limb has too few points to interpolate")
  }
  lo <- max(min(x[up]), min(x[dn]))
  hi <- min(max(x[up]), max(x[dn]))
  if (!(hi > lo)) stop("branch_gap: limbs do not overlap in x")
  g <- seq(lo, hi, length.out = n)
  ou <- order(x[up])
  od <- order(x[dn])
  yu <- stats::approx(x[up][ou], y[up][ou], g, ties = "ordered")$y
  yd <- stats::approx(x[dn][od], y[dn][od], g, ties = "ordered")$y
  max(abs(yu - yd), na.rm = TRUE)
}

gap_central <- branch_gap(loop$Cc, loop$clrel_phosphate)
gap_virtual <- branch_gap(loop$effect_phosphate, loop$clrel_phosphate)

c(loop_width_vs_central_pct = round(gap_central, 3),
  loop_width_vs_virtual_pct = round(gap_virtual, 5),
  ratio = round(gap_central / gap_virtual, 0))

# The virtual compartment must ELIMINATE the hysteresis, not merely reduce it.
# Thresholds are set to the accuracy actually achieved (35.4 and 0.013), not to
# a loose bound, so a structural regression in the delay chain shows up red.
stopifnot(
  is.finite(gap_central), is.finite(gap_virtual),
  gap_central > 30,          # a wide loop against plasma
  gap_virtual < 0.05,        # single-valued against the virtual compartment
  gap_central / gap_virtual > 500
)
```

Against the central compartment the loop is `r round(gap_central, 1)` percentage
points wide; against the virtual compartment it collapses to
`r signif(gap_virtual, 2)` percentage points -- a factor of
`r format(round(gap_central / gap_virtual, -1), big.mark = ",")` -- i.e. to a
single-valued curve. This reproduces the paper's Results statement that
*"Hysteresis was observed if the plasma concentrations were plotted against the
effect (Figure 4A). This was partly solved by adding a biocompartment (virtual
compartment). Hysteresis could be eliminated by adding one transit compartment
(Figure 4B)."* The delay is substantial: the plasma concentration peaks 1.1 h
after the dose but the virtual-compartment concentration driving the effect
peaks at 5.5 h.

# EC50 and the target dose

The paper reports a fitted phosphate EC50 of 6.35 pmol/L and states in the
Discussion that *"the determined EC50 corresponds to a continuous infusion with
a dose between 75 and 100 ug per day."* Because the effect-delay chain
equilibrates to the plasma concentration, at steady state the effect-site
concentration equals `Cc`, so the dose that puts the effect exactly at EC50 is
available in closed form -- and the sigmoid must then sit exactly halfway
between `e0` and `emax`. This is a joint check of the EC50 value, the asymptote
parameterisation, F, CL and the molecular weight used above.

```{r ec50-gate}
ec50_phosphate <- 6.35   # pmol/L, Table S4 E01.EC50

dose_for_css <- function(css_pmol_L) {
  pmol_per_day_to_ug((css_pmol_L - BL_PTH) * CL_POP / F_SC * 24)
}
dose_at_ec50 <- dose_for_css(ec50_phosphate)
dose_at_ec50

# The paper's stated window.
stopifnot(dose_at_ec50 > 75, dose_at_ec50 < 100)

# Simulate that continuous dose to steady state and confirm both the
# concentration and the exact-midpoint effect.
ndays <- 20
ev_ec50 <- dplyr::bind_rows(
  tibble::tibble(time = 0, amt = ug_to_pmol(dose_at_ec50 * ndays),
                 dur = ndays * 24, evid = 1L, cmt = "depot"),
  tibble::tibble(time = seq(0, ndays * 24, by = 1), amt = NA_real_,
                 dur = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  dplyr::mutate(id = 1L, RENALFUNC_REL = 1, HEPFUNC_REL = 1)

ss <- solve_pth(ev_ec50) |>
  dplyr::filter(time > 15 * 24, time < ndays * 24)
stopifnot(nrow(ss) > 50L)

css_sim <- mean(ss$Cc)
phos_sim <- mean(ss$clrel_phosphate)
phos_midpoint <- (14.18 + 50.62) / 2

c(dose_ug_per_day = round(dose_at_ec50, 1),
  Css_simulated = round(css_sim, 4), EC50_published = ec50_phosphate,
  phosphate_CL_simulated = round(phos_sim, 4),
  phosphate_CL_midpoint = phos_midpoint)

stopifnot(abs(css_sim - ec50_phosphate) < 0.01,
          abs(phos_sim - phos_midpoint) < 0.01)
```

A continuous infusion of `r round(dose_at_ec50, 1)` ug/day drives the steady-state
PTH concentration to the fitted EC50, and the phosphate clearance lands on
`r round(phos_sim, 2)`% -- exactly the midpoint of `e0` = 14.18% and `emax` =
50.62%. The dose sits inside the paper's stated 75-100 ug/day window, which
independently corroborates both the asymptote parameterisation and the
molecular weight used for the microgram conversion.

The patient was ultimately maintained on 75 ug/day. The paper reports that this
*"results in a PTH concentration of 7 pmol/L"*; the model predicts
`r round(dose_for_css(7), 1)` ug/day would be needed for 7 pmol/L, or
equivalently that 75 ug/day gives about
`r round(ug_to_pmol(75) / 24 * F_SC / CL_POP + BL_PTH, 2)` pmol/L. The 7 pmol/L
figure is an observed clinical measurement, not a model prediction, and the
paper itself notes (Discussion, Table 1) that its fitted CL and V sit on the
low side of the range Clarke et al reported and that a lower true bioavailability
would reconcile them. No parameter was adjusted to close this gap.

# Equation (3) sensitivity -- the organ-function decomposition

Neither `RENALFUNC_REL` nor `HEPFUNC_REL` has a reported value in the source, so
they are held at 1 above. The check below confirms that the packaged model
consumes them exactly as Equation (3) specifies, which is the reason the
decomposition was kept in the model rather than folded into a single clearance.

```{r eq3-gate}
eq3_probe <- function(rf, lf) {
  ev <- dplyr::bind_rows(
    tibble::tibble(time = 0, amt = ug_to_pmol(100), dur = NA_real_,
                   evid = 1L, cmt = "depot"),
    tibble::tibble(time = seq(0, 48, by = 0.05), amt = NA_real_,
                   dur = NA_real_, evid = 0L, cmt = "central")
  ) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    dplyr::mutate(id = 1L, RENALFUNC_REL = rf, HEPFUNC_REL = lf)
  s <- solve_pth(ev)
  # Baseline-corrected AUC over a window many half-lives long; for linear PK
  # AUC = Dose * F / CLi, so CLi is recoverable from the profile.
  auc <- sum(diff(s$time) *
               (utils::head(s$Cc - BL_PTH, -1) + utils::tail(s$Cc - BL_PTH, -1)) / 2)
  tibble::tibble(RENALFUNC_REL = rf, HEPFUNC_REL = lf,
                 cli_recovered = ug_to_pmol(100) * F_SC / auc,
                 cli_equation3 = CL_POP * (0.3 * rf + 0.7 * lf))
}

eq3 <- dplyr::bind_rows(
  eq3_probe(1.0, 1.0), eq3_probe(0.5, 1.0), eq3_probe(0.2, 1.0),
  eq3_probe(1.0, 0.5), eq3_probe(1.5, 1.0)
) |>
  dplyr::mutate(pct_diff = 100 * (cli_recovered - cli_equation3) / cli_equation3)

stopifnot(nrow(eq3) == 5L, max(abs(eq3$pct_diff)) < 0.5)

eq3 |>
  dplyr::mutate(dplyr::across(c(cli_recovered, cli_equation3), \(x) round(x, 3)),
                pct_diff = round(pct_diff, 3)) |>
  dplyr::rename(
    "RENALFUNC_REL" = RENALFUNC_REL, "HEPFUNC_REL" = HEPFUNC_REL,
    "CLi recovered from AUC (L/h)" = cli_recovered,
    "CLi from Equation (3) (L/h)" = cli_equation3,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(caption = paste(
    "Individual clearance recovered from the simulated profile matches",
    "Equation (3) across renal and hepatic function multipliers. With fe = 0.3,",
    "abolishing renal function entirely can remove at most 30% of total",
    "clearance."
  ))
```

# Baseline hold

With no dose the model must sit exactly at the reported endogenous baseline
indefinitely -- the plasma concentration at `bl_pth`, and both readouts at their
drug-free values. This is the standard steady-state check for a model with an
additive endogenous baseline and it also confirms that initialising the two
delay chains at `bl_pth` (rather than at zero) removes any spurious start-up
transient.

```{r baseline-hold}
ev_bl <- tibble::tibble(time = seq(0, 14 * 24, by = 1), amt = NA_real_,
                        dur = NA_real_, evid = 0L, cmt = "central") |>
  dplyr::mutate(id = 1L, RENALFUNC_REL = 1, HEPFUNC_REL = 1)
bl <- solve_pth(ev_bl)
stopifnot(nrow(bl) > 100L)

c(Cc_range = diff(range(bl$Cc)),
  phosphate_range = diff(range(bl$clrel_phosphate)),
  calcium_range = diff(range(bl$clrel_calcium)),
  Cc_value = unique(round(bl$Cc, 10)),
  phosphate_value = round(bl$clrel_phosphate[1], 4),
  calcium_value = round(bl$clrel_calcium[1], 4))

stopifnot(
  diff(range(bl$Cc)) < 1e-10,
  abs(bl$Cc[1] - BL_PTH) < 1e-10,
  diff(range(bl$clrel_phosphate)) < 1e-8,
  diff(range(bl$clrel_calcium)) < 1e-8
)
```

The drug-free readouts settle at 14.184% for phosphate and 2.569% for calcium
rather than exactly at `e0` = 14.18% and 2.57%, because the endogenous baseline
of 0.58 pmol/L occupies a small but non-zero position on each sigmoid. The
offsets (+0.004 and -0.001 percentage points) are far below the standard errors
of `e0` itself (4.72 and 4.51).

# PKNCA validation

The paper's Table 1 reports CL/F and Vss/F derived from its PKPD model and
compares them with Clarke et al's single-dose values. Running non-compartmental
analysis on simulated single-dose profiles recovers those two quantities
independently of the closed-form arithmetic above, and confirms dose linearity
across the four single-dose levels used in the study.

Because the model carries a non-decaying endogenous baseline, the
concentrations passed to PKNCA are baseline-corrected (`Cc - bl_pth`); without
that correction AUC to infinity is unbounded and no terminal slope exists.

```{r pknca}
doses_ug <- c(25, 35, 50, 100)

# Dense sampling through the absorption and distribution phase; a grid that does
# not resolve tmax understates AUC by several percent.
nca_grid <- sort(unique(c(seq(0, 4, by = 0.02), seq(4, 24, by = 0.25))))

make_arm <- function(i) {
  d <- doses_ug[[i]]
  dplyr::bind_rows(
    tibble::tibble(time = 0, amt = ug_to_pmol(d), dur = NA_real_,
                   evid = 1L, cmt = "depot"),
    tibble::tibble(time = nca_grid, amt = NA_real_, dur = NA_real_,
                   evid = 0L, cmt = "central")
  ) |>
    dplyr::mutate(id = i, treatment = paste0(d, " ug SC single dose"),
                  RENALFUNC_REL = 1, HEPFUNC_REL = 1)
}

events_nca <- dplyr::bind_rows(lapply(seq_along(doses_ug), make_arm)) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

# Disjoint ids across arms: duplicate ids would silently merge into one subject.
stopifnot(!anyDuplicated(unique(events_nca[, c("id", "time", "evid")])),
          dplyr::n_distinct(events_nca$id) == length(doses_ug))

sim_nca_raw <- solve_pth(events_nca, keep = "treatment")

# Baseline-corrected concentrations. Only `!is.na(Cc)` in the filter: adding
# `time > 0` or `Cc > 0` would drop the time-zero anchor PKNCA needs for AUC.
sim_nca <- sim_nca_raw |>
  dplyr::mutate(Cc = Cc - BL_PTH) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time-zero row per (id, treatment); pre-dose Cc = 0 is correct for
# a baseline-corrected extravascular profile.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

# Solver noise must not push the tail negative -- PKNCA would take log() of it.
stopifnot(all(sim_nca$Cc >= 0), nrow(sim_nca) > 0L)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events_nca |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE, vz.obs = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-table}
nca_res$result |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life",
                                "cl.obs", "vz.obs")) |>
  dplyr::mutate(parameter = nlmixr2lib::ncaParamLabel(
    PPTESTCD,
    units = c(cmax = "pmol/L", tmax = "h", aucinf.obs = "pmol*h/L",
              half.life = "h", cl.obs = "L/h", vz.obs = "L")
  )) |>
  dplyr::select(treatment, parameter, PPORRES) |>
  tidyr::pivot_wider(names_from = treatment, values_from = PPORRES) |>
  dplyr::mutate(dplyr::across(-parameter, \(x) signif(x, 4))) |>
  dplyr::rename("NCA parameter" = parameter) |>
  knitr::kable(caption = paste(
    "Simulated single-dose NCA across the four dose levels of Table S1,",
    "on baseline-corrected concentrations. CL/F, Vz/F and half-life are",
    "dose-independent, as required by the linear one-compartment structure;",
    "Cmax and AUC scale in proportion to dose."
  ))
```

## Comparison against published values

```{r nca-compare}
# Table 1 of Visscher 2025, "PKPD model 25-100 ug" column. The parentheses in
# that table are value +/- standard error: propagating the Table S4 SEs through
# the division by F reproduces them exactly (2.79 / 0.53 = 5.3 -> 70-81, and
# 11.71 / 0.53 = 22.1 -> 206-250), which is what identifies them as SEs rather
# than a range or a 95% confidence interval.
# Vss/F is compared against vz.obs because for a one-compartment model the
# steady-state and terminal-phase volumes are the same quantity, Vc.
published <- tibble::tibble(
  treatment = paste0(doses_ug, " ug SC single dose"),
  cl.obs = 75,
  vz.obs = 228
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = c("cl.obs", "vz.obs"),
  units = c(cl.obs = "L/h", vz.obs = "L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated NCA versus Visscher 2025 Table 1 (CL/F 75 [70-81] L/h,",
    "Vss/F 228 [206-250] L). * marks a difference above 20%.",
    "For reference, Clarke et al report CL/F 82 (61-96) L/h at 50 ug and",
    "90 (88-139) L/h at 100 ug, and Vss/F 267 (221-585) L and 443 (253-722) L."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
# Both quantities must land within a few percent of the published values, at
# every dose. Tightened to the accuracy actually achieved so a future
# regression is caught.
recovered <- nca_res$result |>
  dplyr::filter(PPTESTCD %in% c("cl.obs", "vz.obs")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(recovered) == length(doses_ug))
stopifnot(
  # CL/F = 39.93 / 0.53 = 75.34 L/h; Table 1 rounds to 75.
  all(abs(recovered$cl.obs - CL_POP / F_SC) < 0.05),
  all(abs(recovered$cl.obs - 75) / 75 < 0.01),
  # Vz/F = 120.71 / 0.53 = 227.75 L; Table 1 rounds to 228.
  all(abs(recovered$vz.obs - V_POP / F_SC) < 1),
  all(abs(recovered$vz.obs - 228) / 228 < 0.01)
)

half_lives <- nca_res$result |>
  dplyr::filter(PPTESTCD == "half.life") |> dplyr::pull(PPORRES)
stopifnot(length(half_lives) == length(doses_ug),
          all(abs(half_lives - log(2) * V_POP / CL_POP) < 0.02))

c(CL_F_simulated = round(unique(round(recovered$cl.obs, 3)), 3),
  CL_F_published = 75,
  Vz_F_simulated = round(unique(round(recovered$vz.obs, 2)), 2),
  Vss_F_published = 228,
  half_life_simulated = round(unique(round(half_lives, 3)), 3),
  half_life_analytic = round(log(2) * V_POP / CL_POP, 3))
```

CL/F and Vz/F are recovered to within 0.5% and 0.1% of Table 1 respectively, and
the terminal half-life matches the analytic `log(2) * V / CL` = 2.10 h. The
recovered values are identical at all four dose levels, confirming the linearity
of the structure. The 2.1 h half-life is consistent with the paper's citation of
Sikjaer et al, who observed that PTH plasma concentrations *"returned to normal
after approximately 16 h"* -- about 7.6 half-lives.

# Assumptions and deviations

- **Erlang-5 absorption: five compartments, five transfers.** Methods 2.3 says
  *"five transit compartments were included in the absorption compartment"*
  without an equation or a diagram of the chain, and only one absorption rate
  constant (`I02.k`) is reported. This is implemented as the standard
  transit-compartment form -- the dose enters the first of five compartments and
  five first-order transfers at `ka` carry it to `central`, giving a
  gamma(5, 1/ka) absorption-time distribution. The alternative literal reading,
  a dose site *plus* five further transit compartments (six transfers), is not
  excluded by the text; it would move the simulated tmax from 1.08 h to 1.24 h,
  further from the 15 min tmax that the paper cites from Sikjaer et al,
  which is the weak evidence favouring the reading used here. Every other
  validation gate in this vignette is insensitive to the choice, because they
  all depend on total exposure rather than on the shape of the absorption phase.
- **`RENALFUNC_REL` and `HEPFUNC_REL` are held at 1.** The paper fixes LF at 1
  explicitly. For RF it states that Equation (3) *"includes the variations in
  the GFR by correcting for the relative renal function"* and shows a rising GFR
  over the study in Figure S2, but it reports no numeric RF values and no
  reference GFR denominator, and Table S3 omits the 24-hour urine volume that
  the paper's own GFR equation requires -- so the per-interval ratio cannot be
  reconstructed from the supplement either. No reference GFR was invented; the
  covariates are exercised as a sensitivity analysis instead.
- **Emax is treated as an asymptote, not an increment.** This is a derived
  reading, not a stated one. It is forced by the calcium arm (`Emax` = 2.05% is
  below `E0` = 2.57%) together with the Conclusion's statement that continuous
  dosing produces *"a permanent reduction in the calcium clearance"*; the
  increment reading would make calcium clearance rise. See the model-structure
  section.
- **Both effect-delay chains are initialised at the endogenous baseline.** The
  paper does not state the initial conditions of the virtual compartments.
  Starting them at `bl_pth` rather than at zero is what makes the drug-free
  model sit at rest, as the baseline-hold check demonstrates; starting at zero
  would introduce a spurious multi-hour transient at the start of every
  simulation.
- **The two transfers within each effect-delay chain share one rate constant.**
  Table S4 reports a single `T01.k` / `T02.k` per analyte, and Figure S1 shows
  one transfer object per analyte, so no second rate is available.
- **NON-PAPER PROVENANCE: molecular weight 9425 g/mol.** Visscher 2025 states
  its doses in micrograms and its concentrations and EC50 in pmol/L but never
  gives a molecular weight. The model file is therefore encoded entirely in
  molar units and the conversion lives only in this vignette. The value is
  corroborated by the paper's own arithmetic: it puts the fitted EC50 at
  98.3 ug/day of continuous infusion, and the Discussion states that the EC50
  *"corresponds to a continuous infusion with a dose between 75 and 100 ug per
  day"*.
- **PKNCA is run on baseline-corrected concentrations.** The endogenous baseline
  does not decay, so uncorrected profiles have no terminal phase and an
  unbounded AUC to infinity. Correction by `bl_pth` is what makes
  `cl.obs` = Dose/AUCinf equal CL/F.
- **`Vss/F` is compared against `vz.obs`.** Table 1 reports a steady-state
  volume; for a one-compartment model the steady-state and terminal-phase
  volumes are both `Vc`, so the comparison is exact rather than approximate.
- **The calcium arm is poorly identified and is shipped as reported.** All five
  calcium parameters carry the Table S4 dagger footnote, every standard error
  exceeds its estimate (by more than 15-fold for `T02.k`, `E02.EC50` and
  `E02.j`), and the goodness-of-fit R-squared is 0.06 versus 0.66 for phosphate.
  The paper attributes this to concomitant alfacalcidol, calcium carbonate and
  chlortalidone perturbing calcium homeostasis, and selects phosphate as the
  effect parameter for that reason. The calcium arm is reproduced without
  modification, but conclusions should not be drawn from it.
- **The observed 7 pmol/L at 75 ug/day is not reproduced, and was not tuned
  to.** The model predicts about 4.98 pmol/L at that dose. The paper reports
  7 pmol/L as a clinical measurement in the patient and separately notes that
  its fitted CL and V lie below Clarke et al's range, with a lower true
  bioavailability as the likely reconciliation. No parameter was adjusted.
- **No inter-individual variability or residual error is available.** The source
  fitted a single patient deterministically in Edsim++ and Table S4 contains no
  OMEGA or SIGMA block, so none was invented. The model is a typical-value
  simulation model; `zeroRe()` and `omega = NA` are neither needed nor valid for
  it.
- **Concomitant medication is recorded in `population$co_medication` but is not
  a model covariate.** Alfacalcidol, calcium carbonate and chlortalidone are the
  paper's own explanation for the calcium arm's poor fit, but no drug-interaction
  parameters are estimated for them.
