---
title: "Cefepime-enmetazobactam (Vollmer 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cefepime-enmetazobactam (Vollmer 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)

rxode2::rxSetSeed(20231014)  # IDWeek presentation date; common random numbers
```

# The model

`Vollmer_2023_cefepime_enmetazobactam` is the joint population PK model for the
fixed-ratio (4:1) cefepime-enmetazobactam combination EXBLIFEP, pooled across
three Phase 1 studies, one Phase 2 study and the Phase 3 ALLIUM trial. Each drug
gets its own two-compartment intravenous disposition; the two are tied together
by a four-way OMEGA block whose cross-drug correlations exceed 0.93.

```{r load-model}
mod <- nlmixr2lib::modellib("Vollmer_2023_cefepime_enmetazobactam")
typ <- rxode2::zeroRe(mod)   # typical-value (all etas zero) version
mod
```

# Population

The analysis set is 663 adults contributing 5,070 cefepime plasma observations
(588 subjects) and 6,342 enmetazobactam plasma observations (649 subjects)
across five studies: AT-101, AT-102 and AT-103 (Phase 1; n = 83, 30, 19),
AT-201 (Phase 2; n = 43) and AT-301 / ALLIUM (Phase 3; n = 488). Twenty percent
of the analysis set (132/663) are healthy volunteers and 80% are patients with a
complicated urinary tract infection or acute pyelonephritis.

```{r population}
knitr::kable(
  tibble::tribble(
    ~Characteristic,                ~`Mean (range) or n (%)`,
    "Age (years)",                  "51.62 (17-94)",
    "Body weight (kg)",             "76.52 (45-135)",
    "Body mass index (kg/m^2)",     "26.54 (16.9-45.9)",
    "Body surface area (m^2)",      "1.89 (1.37-2.63)",
    "eGFR (mL/min/1.73 m^2)",       "81.03 (4.4-166)",
    "De-indexed eGFR (mL/min)",     "87.98 (5.21-195.84)",
    "Female",                       "310 (47%)",
    "White",                        "641 (97%)",
    "Acute pyelonephritis",         "252 (38%)",
    "Complicated UTI (all subtypes)", "279 (42%)",
    "Healthy volunteer",            "132 (20%)"
  ),
  caption = "Baseline demographics of the population PK analysis set (FDA NDA 216165 Integrated Review Table 100)."
)
```

Race could not be tested as a covariate at all: 97% of the analysis set was
White, and the US label states there was insufficient information to evaluate a
race effect. Body-surface area, body-mass index, creatinine clearance, albumin
and bilirubin were formally screened and not retained; they are documented in
the model's `covariatesDataExcluded` rather than declared as covariates.

# Source trace

The IDWeek abstract is the paper of record, but it publishes only the typical
values for a 70 kg cUTI patient and their %CV. The complete parameter set comes
from the two regulatory reproductions of the same sponsor study report,
AAI101-PK-21-01, which agree with each other value for value.

```{r source-trace}
knitr::kable(
  tibble::tribble(
    ~Quantity, ~Source,
    "Structural model (2 compartments per drug, linear CL, IV infusion)", "EMA EPAR EMA/63929/2024 p.46; FDA Integrated Review Figure 13",
    "Cl, V1, Q, V2 for both drugs", "FDA Table 101 'Fixed effects'; EMA Tables 5 and 6",
    "Q fixed, not estimated", "EMA EPAR p.46; FDA Table 101 RSE column reads 'fixed'",
    "Reference individual: 70 kg, eGFR 100 mL/min, age 50 y, female, healthy", "EMA EPAR p.46 and Table 6 note",
    "Covariate coefficients (9)", "FDA Table 101 'Covariate coefficients'; EMA Tables 5 and 6",
    "IIV standard deviations (9, stratified by infection)", "FDA Table 101 'Standard deviations'",
    "Random-effect correlations (7)", "FDA Table 101 'Correlations'; EMA Tables 5 and 6 for the within-drug ones",
    "Residual error (3)", "FDA Table 101 'Observational error'; EMA Tables 5 and 6",
    "Typical values for a 70 kg cUTI patient; terminal half-lives 2.2 / 2.0 h", "Vollmer 2023 IDWeek abstract 2529, embedded table",
    "Baseline demographics", "FDA Integrated Review Table 100",
    "Simulated steady-state exposures by renal function", "FDA Integrated Review Table 104",
    "Observed steady-state NCA at the approved regimen", "US EXBLIFEP label (NDA 216165) Table 5"
  ),
  caption = "Where every model element comes from."
)
```

# Check 1 -- the reference individual reproduces the published estimates

Every covariate term is normalised so that it equals 1 at the reference
individual. Solving at 70 kg, age 50, de-indexed eGFR 100 mL/min, female and
healthy must therefore return the `ini()` estimates exactly.

```{r reference-individual}
base_cov <- function(...) {
  out <- data.frame(WT = 70, AGE = 50, CRCL = 100, SEXF = 1, DIS_CUTI = 0)
  rep <- list(...)
  for (nm in names(rep)) out[[nm]] <- rep[[nm]]
  out
}

# A single time-zero observation is enough to read the derived parameters back.
params_at <- function(cov) {
  s <- rxode2::rxSolve(
    typ,
    cbind(cov, id = 1L) |> mutate(time = 0, amt = 0, evid = 0L, cmt = "Cc"),
    returnType = "data.frame"
  )
  c(cl = s$cl[1], vc = s$vc[1], q = s$q[1], vp = s$vp[1],
    cl_enm = s$cl_enm[1], vc_enm = s$vc_enm[1], q_enm = s$q_enm[1], vp_enm = s$vp_enm[1])
}

ref <- params_at(base_cov())
published <- c(cl = 5.95, vc = 11.2, q = 7.22, vp = 5.7,
               cl_enm = 7.68, vc_enm = 13.2, q_enm = 7.16, vp_enm = 5.28)

knitr::kable(
  data.frame(
    Parameter = c("Cefepime CL (L/h)", "Cefepime Vc (L)", "Cefepime Q (L/h)", "Cefepime Vp (L)",
                  "Enmetazobactam CL (L/h)", "Enmetazobactam Vc (L)", "Enmetazobactam Q (L/h)",
                  "Enmetazobactam Vp (L)"),
    Published = published,
    Model = ref,
    row.names = NULL
  ),
  caption = "Reference individual (70 kg, 50 y, de-indexed eGFR 100 mL/min, female, healthy) versus FDA Table 101."
)

# Same drawn parameters on both sides, so this is pure floating-point agreement
# and a tight bound is the correct one.
stopifnot(max(abs(ref / published - 1)) < 1e-10)
```

# Check 2 -- the cUTI effect reproduces the separately published infected volume

This is the check that pins the categorical covariate form. The EMA report and
the IDWeek abstract both state that the enmetazobactam central volume in
infected subjects is 15.3 L, against a reference estimate of 13.2 L. The
exponential form gives `13.2 * exp(0.145) = 15.26`; the proportional form
`13.2 * (1 + 0.145) = 15.11` does not round to 15.3.

```{r cuti-effect}
inf_par <- params_at(base_cov(DIS_CUTI = 1))

knitr::kable(
  data.frame(
    Form = c("Model as encoded, exp(beta)", "Alternative, (1 + beta)"),
    `Enmetazobactam Vc (L)` = c(inf_par[["vc_enm"]], 13.2 * (1 + 0.145)),
    Published = c(15.3, 15.3),
    check.names = FALSE
  ),
  caption = "Enmetazobactam central volume in an infected 70 kg patient, published as 15.3 L."
)

stopifnot(
  abs(round(inf_par[["vc_enm"]], 1) - 15.3) < 1e-8,  # rounds to the published 15.3
  abs(round(13.2 * 1.145, 1) - 15.3) > 0.1           # the alternative form does not
)
```

Cefepime carries no infection effect, so its central volume is unchanged at
11.2 L -- which is also the value the abstract prints for a "70 kg cUTI
patient", and that agreement is what independently pins the weight reference to
70 kg.

```{r cuti-effect-fep}
stopifnot(abs(inf_par[["vc"]] - 11.2) < 1e-10)
```

# Check 3 -- the OMEGA block and its positive-definiteness repair

FDA Table 101 prints all six correlations of the infected 4x4 block to three
decimals. Built from those printed values verbatim, the block is very slightly
**indefinite**, so `chol()` fails and `rxSolve()` cannot sample from it. That is
a rounding artifact rather than a contradiction, and the model file repairs it
with the nearest positive-definite correlation matrix. The repair has to be
shown to be publication-consistent: every adjusted correlation must still round
to the printed three-decimal value.

```{r omega-repair}
nm  <- c("clF", "vcF", "clE", "vcE")
sdv <- c(clF = 0.297, vcF = 0.42, clE = 0.356, vcE = 0.474)

R <- diag(4); dimnames(R) <- list(nm, nm)
R["vcF", "clF"] <- R["clF", "vcF"] <- 0.261
R["clE", "clF"] <- R["clF", "clE"] <- 0.932
R["vcE", "clF"] <- R["clF", "vcE"] <- 0.261
R["clE", "vcF"] <- R["vcF", "clE"] <- 0.339
R["vcE", "vcF"] <- R["vcF", "vcE"] <- 0.955
R["vcE", "clE"] <- R["clE", "vcE"] <- 0.438

lambda_printed <- min(eigen(R, only.values = TRUE)$values)
Rp <- as.matrix(Matrix::nearPD(R, corr = TRUE, keepDiag = TRUE, posd.tol = 1e-6)$mat)
delta <- max(abs((Rp - R)[upper.tri(R)]))

knitr::kable(
  data.frame(
    Correlation = c("Cefepime Vc-CL", "Cefepime CL-enmetazobactam CL",
                    "Cefepime CL-enmetazobactam Vc", "Cefepime Vc-enmetazobactam CL",
                    "Cefepime Vc-enmetazobactam Vc", "Enmetazobactam Vc-CL"),
    Printed  = R[upper.tri(R)][c(1, 2, 4, 3, 5, 6)],
    Repaired = round(Rp[upper.tri(Rp)][c(1, 2, 4, 3, 5, 6)], 6),
    `Rounds back` = round(Rp[upper.tri(Rp)][c(1, 2, 4, 3, 5, 6)], 3) ==
                    R[upper.tri(R)][c(1, 2, 4, 3, 5, 6)],
    check.names = FALSE
  ),
  caption = "Nearest-positive-definite repair of the infected correlation block."
)

# The repair is only legitimate if it stays inside the rounding interval of the
# printed three-decimal values, i.e. moves nothing by 0.0005 or more.
stopifnot(
  lambda_printed < 0,                              # the printed matrix really is indefinite
  delta < 0.0005,                                  # repair stays inside the rounding interval
  all(round(Rp, 3) == round(R, 3)),                # every entry rounds back to what was printed
  min(eigen(Rp, only.values = TRUE)$values) > 0    # and the result is usable
)
```

The model's assembled 9x9 OMEGA must then be positive definite, and reading its
correlations back must recover the published cross-drug values.

```{r omega-model}
# modellib() hands back the model FUNCTION; build the rxUi to read its OMEGA.
om   <- rxode2::rxode2(mod)$omega
corr <- cov2cor(om)

knitr::kable(
  data.frame(
    `Random-effect pair` = c("Cefepime CL - enmetazobactam CL (infected)",
                             "Cefepime Vc - enmetazobactam Vc (infected)",
                             "Enmetazobactam Vc - CL (healthy)"),
    Published = c(0.932, 0.955, 0.983),
    Model = round(c(corr["etalcl_inf", "etalcl_enm_inf"],
                    corr["etalvc_inf", "etalvc_enm_inf"],
                    corr["etalcl_enm_hlth", "etalvc_enm_hlth"]), 4),
    check.names = FALSE
  ),
  caption = "Cross-drug and within-drug IIV correlations versus FDA Table 101."
)

stopifnot(
  min(eigen(om, only.values = TRUE)$values) > 0,
  abs(corr["etalcl_inf", "etalcl_enm_inf"]     - 0.932) < 0.0005,
  abs(corr["etalvc_inf", "etalvc_enm_inf"]     - 0.955) < 0.0005,
  abs(corr["etalcl_enm_hlth", "etalvc_enm_hlth"] - 0.983) < 0.0005
)
```

The printed %CV column is a second, independent view of the same variances, and
all nine estimated omegas reproduce it under the log-normal identity
`CV = sqrt(exp(omega^2) - 1)`. That over-determination is what removes any doubt
about whether the "Standard deviations" column holds SDs or variances.

```{r omega-cv}
omega_sd <- c(0.161, 0.297, 0.42, 0.195, 0.15, 0.356, 0.0549, 0.474, 0.0593)
cv_pub   <- c(16, 30, 44, 20, 15, 37, 5, 50, 6)
cv_calc  <- 100 * sqrt(exp(omega_sd^2) - 1)

knitr::kable(
  data.frame(
    `Random effect` = c("Cefepime CL, healthy", "Cefepime CL, infected",
                        "Cefepime Vc, infected", "Cefepime Vp, both",
                        "Enmetazobactam CL, healthy", "Enmetazobactam CL, infected",
                        "Enmetazobactam Vc, healthy", "Enmetazobactam Vc, infected",
                        "Enmetazobactam Vp, both"),
    omega = omega_sd,
    `%CV printed` = cv_pub,
    `%CV from omega` = round(cv_calc, 1),
    check.names = FALSE
  ),
  caption = "Every printed omega reproduces its printed %CV as a log-normal SD."
)

stopifnot(all(round(cv_calc) == cv_pub))
```

# Check 4 -- terminal half-life from the solved ODE system

Both the abstract and the EMA report quote terminal plasma half-lives calculated
from the final parameters: 2.2 h for cefepime and 2.0 h for enmetazobactam in
infected subjects. Regressing the log-linear tail of a solved single-dose
profile exercises the whole four-compartment system rather than just the `ini()`
arithmetic.

```{r half-life}
solve_regimen <- function(model, cov, fep, enm, ii, dur, t_start, t_end, by = 0.02) {
  dose_times <- if (ii >= t_end) 0 else seq(0, t_end - ii, by = ii)
  dose <- bind_rows(
    tidyr::expand_grid(cov, time = dose_times) |>
      mutate(amt = fep, cmt = "central",     evid = 1L, dur = dur),
    tidyr::expand_grid(cov, time = dose_times) |>
      mutate(amt = enm, cmt = "central_enm", evid = 1L, dur = dur)
  )
  obs <- bind_rows(
    tidyr::expand_grid(cov, time = seq(t_start, t_end, by = by)) |>
      mutate(amt = NA_real_, cmt = "Cc",     evid = 0L, dur = NA_real_),
    tidyr::expand_grid(cov, time = seq(t_start, t_end, by = by)) |>
      mutate(amt = NA_real_, cmt = "Cc_enm", evid = 0L, dur = NA_real_)
  )
  ev <- bind_rows(dose, obs) |> arrange(id, time, desc(evid))
  out <- rxode2::rxSolve(model, ev, returnType = "data.frame")
  # rxSolve() drops `id` when the event table holds exactly one subject.
  if (!"id" %in% names(out)) {
    stopifnot(dplyr::n_distinct(cov$id) == 1L)
    out$id <- cov$id[1]
  }
  out |> distinct(id, time, .keep_all = TRUE)
}

half_life <- function(t, y) log(2) / -stats::coef(stats::lm(log(y) ~ t))[[2]]

sd_prof <- solve_regimen(typ, cbind(base_cov(DIS_CUTI = 1), id = 1L),
                         fep = 2000, enm = 500, ii = 1e6, dur = 2,
                         t_start = 0, t_end = 24)
tail_w  <- sd_prof$time >= 14 & sd_prof$time <= 24

hl <- c(cefepime       = half_life(sd_prof$time[tail_w], sd_prof$Cc[tail_w]),
        enmetazobactam = half_life(sd_prof$time[tail_w], sd_prof$Cc_enm[tail_w]))

knitr::kable(
  data.frame(Drug = c("Cefepime", "Enmetazobactam"),
             `Published t1/2 (h)` = c(2.2, 2.0),
             `Model t1/2 (h)` = round(hl, 3),
             check.names = FALSE, row.names = NULL),
  caption = "Terminal half-life regressed from the solved profile of an infected 70 kg reference patient."
)

stopifnot(abs(hl[["cefepime"]] - 2.2) < 0.05, abs(hl[["enmetazobactam"]] - 2.0) < 0.05)
```

```{r single-dose-plot, fig.cap = "Typical-value single-dose profiles, infected 70 kg reference patient, 2 g cefepime / 0.5 g enmetazobactam over 2 hours."}
sd_prof |>
  select(time, Cefepime = Cc, Enmetazobactam = Cc_enm) |>
  tidyr::pivot_longer(-time, names_to = "Analyte", values_to = "conc") |>
  ggplot(aes(time, conc, colour = Analyte)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total plasma concentration (mg/L)") +
  theme_bw()
```

# Check 5 -- the renal-function ladder of FDA Table 104

This is the sharpest available test of the covariate model. For each renal
function group, FDA Integrated Review Table 104 reports the simulated
steady-state Cmax and AUC0-24 of **both** drugs at the dose recommended for that
group, in 8,000 virtual subjects whose covariates were all fixed except
de-indexed eGFR (sponsor page 138: body weight 75.68 kg, age 54.65 years,
infected, sex 1:1).

The eGFR distribution *within* each band is not published, so it is not assumed.
Instead a **single** number per group is taken from the table -- the cefepime
AUC0-24 -- and inverted through the cefepime clearance model to recover the
de-indexed eGFR of the patient who would produce it. The other three quantities
per group are then genuine *predictions*: cefepime Cmax, enmetazobactam Cmax and
enmetazobactam AUC0-24, eighteen predictions in total.

Because every dosing interval here divides 24 h exactly, `AUC0-24` at steady
state is `AUC_tau * 24/tau` with no phase dependence, so the inversion is clean.

```{r ladder-table}
t104 <- tibble::tribble(
  ~band,      ~lo, ~hi, ~fep, ~enm, ~ii, ~dur, ~fepCmax, ~fepAUC,  ~enmCmax, ~enmAUC,
  "15-<30",    15,  30, 1000,  250,  12,    2,   82.050, 1205.561,   17.210, 257.076,
  "30-<60",    30,  60, 1000,  250,   8,    2,   65.639,  930.164,   13.342, 188.652,
  "60-<90",    60,  90, 2000,  500,   8,    2,  104.614, 1271.797,   20.830, 250.690,
  "90-<130",   90, 130, 2000,  500,   8,    2,   89.936,  973.021,   17.614, 186.945,
  ">=130",    130, Inf, 2000,  500,   8,    4,   51.716,  732.636,    9.836, 137.723,
  ">=150",    150, Inf, 2000,  500,   8,    4,   44.926,  614.844,    8.432, 114.002
)

# Invert the cefepime renal power term. Cefepime CL carries no weight or age
# effect, so the inversion needs nothing but the printed AUC.
t104 <- t104 |>
  mutate(cl_obs = (24 / ii) * fep / fepAUC,
         egfr   = 100 * (cl_obs / 5.95)^(1 / 0.834))

knitr::kable(
  t104 |>
    mutate(inside = egfr >= lo & egfr <= hi) |>
    select(`Renal group (mL/min)` = band, `Implied de-indexed eGFR` = egfr,
           `Inside the band` = inside) |>
    mutate(`Implied de-indexed eGFR` = round(`Implied de-indexed eGFR`, 1)),
  caption = "De-indexed eGFR recovered by inverting the cefepime AUC0-24 of FDA Table 104."
)

# Six for six: a mis-specified reference or exponent would push these out of
# their own bands. This is also what identifies the table as TOTAL rather than
# free cefepime -- rescaling by the unbound fraction 0.80 puts four of the six
# outside their bands.
stopifnot(all(t104$egfr >= t104$lo & t104$egfr <= t104$hi))
free_egfr <- 100 * ((t104$cl_obs * 0.80) / 5.95)^(1 / 0.834)
stopifnot(sum(free_egfr < t104$lo | free_egfr > t104$hi) >= 4)
```

```{r ladder-predict}
# Dose for 14 days so even the slowest renal group is unambiguously at steady
# state, then read the final dosing interval. Female and male are solved
# separately (sex acts on enmetazobactam Vp) and averaged 1:1 to match the
# sponsor's simulation.
ladder_pred <- lapply(seq_len(nrow(t104)), function(i) {
  r <- t104[i, ]
  per_sex <- lapply(c(1, 0), function(sf) {
    cov <- data.frame(id = 1L, WT = 75.68, AGE = 54.65, CRCL = r$egfr,
                      SEXF = sf, DIS_CUTI = 1)
    s <- solve_regimen(typ, cov, r$fep, r$enm, r$ii, r$dur,
                       t_start = 14 * 24 - r$ii, t_end = 14 * 24, by = 0.01)
    trapz <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)
    c(fepCmax = max(s$Cc), enmCmax = max(s$Cc_enm),
      fepAUC = trapz(s$time, s$Cc) * 24 / r$ii,
      enmAUC = trapz(s$time, s$Cc_enm) * 24 / r$ii)
  })
  as.data.frame(t(colMeans(do.call(rbind, per_sex))))
}) |> bind_rows()

ladder <- t104 |>
  select(band, ref_fepCmax = fepCmax, ref_fepAUC = fepAUC,
         ref_enmCmax = enmCmax, ref_enmAUC = enmAUC) |>
  bind_cols(ladder_pred) |>
  mutate(across(c(fepCmax, enmCmax, enmAUC, fepAUC), ~ .x)) |>
  transmute(
    `Renal group (mL/min)` = band,
    `Cefepime AUC0-24 (closure)` = round(100 * (fepAUC / ref_fepAUC - 1), 2),
    `Cefepime Cmax` = round(100 * (fepCmax / ref_fepCmax - 1), 1),
    `Enmetazobactam Cmax` = round(100 * (enmCmax / ref_enmCmax - 1), 1),
    `Enmetazobactam AUC0-24` = round(100 * (enmAUC / ref_enmAUC - 1), 1)
  )

knitr::kable(
  ladder,
  caption = "Per cent difference from FDA Table 104. The cefepime AUC0-24 column is the inverted quantity and closes on itself; the other three columns are predictions."
)

pred_err <- as.matrix(ladder[, 3:5])
stopifnot(
  max(abs(as.matrix(ladder[, 2]))) < 0.5,   # the inversion closes
  max(abs(pred_err)) < 10,                  # every prediction within 10%
  stats::median(abs(pred_err)) < 3          # and typically far better
)
```

# Check 6 -- a virtual cohort against the approved-regimen label NCA

The US label reports observed steady-state (Day 7) noncompartmental parameters
in cUTI patients with eGFR at least 60 mL/min receiving the approved 2 g
cefepime / 0.5 g enmetazobactam every 8 hours as a 2-hour infusion. A virtual
cohort matching the Phase 3 AT-301 demographics, restricted to de-indexed eGFR
of at least 60 mL/min, is simulated with full inter-individual variability and
run through PKNCA.

```{r cohort}
n_sub <- 200   # one arm; the skill caps cohorts at 200 participants per arm

# AT-301 demographics (FDA Table 100): weight mean 76.11 kg (45-135), age mean
# 55.04 y (17-94), de-indexed eGFR mean 85.37 mL/min (21.36-193.45), 54% female.
# Weight and eGFR are drawn log-normally so they stay positive; the spreads are
# chosen to reproduce the published means and to sit inside the published
# ranges, which the paper does not tabulate beyond mean and range.
draw_trunc <- function(n, mean_target, cv, lo, hi) {
  s  <- sqrt(log(cv^2 + 1))
  mu <- log(mean_target) - s^2 / 2
  x  <- stats::rlnorm(n * 20, mu, s)
  x  <- x[x >= lo & x <= hi]
  stopifnot(length(x) >= n)   # fail loudly rather than silently recycling
  x[seq_len(n)]
}

cohort <- data.frame(
  id       = seq_len(n_sub),
  WT       = draw_trunc(n_sub, 76.11, 0.22, 45, 135),
  AGE      = round(draw_trunc(n_sub, 55.04, 0.30, 18, 94)),
  CRCL     = draw_trunc(n_sub, 95, 0.30, 60, 193.45),   # restricted to eGFR >= 60 per the label
  SEXF     = stats::rbinom(n_sub, 1, 0.54),
  DIS_CUTI = 1
)

tau     <- 8
t_last  <- 7 * 24 - tau     # time of the final dose in a 7-day course
obs_grid <- sort(unique(c(seq(t_last, t_last + tau, by = 0.1), t_last)))

ev <- bind_rows(
  tidyr::expand_grid(cohort, time = seq(0, t_last, by = tau)) |>
    mutate(amt = 2000, cmt = "central",     evid = 1L, dur = 2),
  tidyr::expand_grid(cohort, time = seq(0, t_last, by = tau)) |>
    mutate(amt = 500,  cmt = "central_enm", evid = 1L, dur = 2),
  tidyr::expand_grid(cohort, time = obs_grid) |>
    mutate(amt = NA_real_, cmt = "Cc",     evid = 0L, dur = NA_real_),
  tidyr::expand_grid(cohort, time = obs_grid) |>
    mutate(amt = NA_real_, cmt = "Cc_enm", evid = 0L, dur = NA_real_)
) |> arrange(id, time, desc(evid))

sim <- rxode2::rxSolve(mod, ev, returnType = "data.frame") |>
  distinct(id, time, .keep_all = TRUE)

# Re-base the final dosing interval to time zero so PKNCA sees a clean
# steady-state interval that starts at a real observation.
nca_in <- sim |>
  filter(time >= t_last) |>
  mutate(time = time - t_last) |>
  select(id, time, Cc, Cc_enm) |>
  tidyr::pivot_longer(c(Cc, Cc_enm), names_to = "endpoint", values_to = "conc") |>
  mutate(analyte = ifelse(endpoint == "Cc", "Cefepime", "Enmetazobactam")) |>
  filter(!is.na(conc))

stopifnot(all(c(0) %in% nca_in$time))   # a real time-zero record exists

dose_in <- cohort |>
  select(id) |>
  tidyr::expand_grid(analyte = c("Cefepime", "Enmetazobactam")) |>
  mutate(time = 0, dose = ifelse(analyte == "Cefepime", 2000, 500))

conc_obj <- PKNCA::PKNCAconc(as.data.frame(nca_in), conc ~ time | analyte + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_in), dose ~ time | analyte + id,
                             doseu = "mg")

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, half.life = TRUE, cl.last = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r cohort-compare}
sim_nca <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "auclast", "half.life", "cl.last")) |>
  select(analyte, PPTESTCD, PPORRES)

# US label Table 5: mean (SD) at steady state, Day 7, cUTI, eGFR >= 60 mL/min.
label_nca <- data.frame(
  analyte   = c("Cefepime", "Enmetazobactam"),
  cmax      = c(99.8, 19.8),
  auclast   = c(379.5, 75.3),
  half.life = c(2.7, 2.6),
  cl.last   = c(5.8, 7.6)
)

tbl <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_nca,
  reference = label_nca,
  by = "analyte",
  units = c(cmax = "mg/L", auclast = "mg*h/L", half.life = "h", cl.last = "L/h")
)
knitr::kable(tbl, caption = "Simulated steady-state NCA (cohort median) versus the US EXBLIFEP label Table 5.")
attr(tbl, "footnote")
```

The exposure parameters land close to the label. Two rows deserve comment and
neither indicates a transcription problem.

* **Half-life.** The label's 2.7 h (cefepime) and 2.6 h (enmetazobactam) are
  noncompartmental terminal half-lives estimated from sparse late samples within
  an 8-hour interval, which systematically exceed the model's 2.2 h and 2.0 h
  true terminal half-lives (Check 4). The bias is a property of the estimator,
  not of the parameters.
* **Clearance.** `cl.last` here is `dose / AUClast` over one interval. Its
  agreement with the label's 5.8 and 7.6 L/h is the more meaningful check, and
  it also brackets the model's reference estimates of 5.95 and 7.68 L/h at
  de-indexed eGFR 100 mL/min.

```{r cohort-assert}
# sim_nca holds one row per subject, so collapse to the cohort MEDIAN first --
# the same statistic ncaComparisonTable() reports above.
wide <- sim_nca |>
  group_by(analyte, PPTESTCD) |>
  summarise(PPORRES = stats::median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

ratio <- function(a, p) {
  wide[[p]][wide$analyte == a] / label_nca[[p]][label_nca$analyte == a]
}

# Assertions are on the CENTRE of the cohort, not on any extreme, because the
# label cohort's covariate distribution is only published as means and ranges.
stopifnot(
  abs(ratio("Cefepime", "cmax") - 1) < 0.25,
  abs(ratio("Enmetazobactam", "cmax") - 1) < 0.25,
  abs(ratio("Cefepime", "auclast") - 1) < 0.30,
  abs(ratio("Enmetazobactam", "auclast") - 1) < 0.30,
  abs(ratio("Cefepime", "cl.last") - 1) < 0.30,
  abs(ratio("Enmetazobactam", "cl.last") - 1) < 0.30
)
```

```{r cohort-plot, fig.cap = "Simulated steady-state concentration-time profiles over the final 8-hour dosing interval (median and 5th-95th percentiles, 200 virtual cUTI patients)."}
sim |>
  filter(time >= t_last) |>
  mutate(time = time - t_last) |>
  select(time, Cefepime = Cc, Enmetazobactam = Cc_enm) |>
  tidyr::pivot_longer(-time, names_to = "Analyte", values_to = "conc") |>
  group_by(Analyte, time) |>
  summarise(med = stats::median(conc),
            lo  = stats::quantile(conc, 0.05),
            hi  = stats::quantile(conc, 0.95), .groups = "drop") |>
  ggplot(aes(time, med, colour = Analyte, fill = Analyte)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after the Day-7 dose (h)", y = "Total plasma concentration (mg/L)") +
  theme_bw()
```

# Check 7 -- variability is higher in patients than in healthy volunteers

The abstract's qualitative claim is that "variability was higher in cUTI
patients although differences in mean ENM or FEP exposures between healthy
subjects and cUTI patients were negligible". Both halves are testable, and the
comparison uses common random numbers so that the two arms differ only through
the infection covariate.

```{r infection-strata}
arm <- function(infected) {
  rxode2::rxSetSeed(20231014)          # reseed INSIDE the loop: common random numbers
  cov <- data.frame(
    id = seq_len(n_sub), WT = 76.11, AGE = 55, CRCL = 100,
    SEXF = 1, DIS_CUTI = infected
  )
  ev <- bind_rows(
    tidyr::expand_grid(cov, time = 0) |>
      mutate(amt = 2000, cmt = "central",     evid = 1L, dur = 2),
    tidyr::expand_grid(cov, time = 0) |>
      mutate(amt = 500,  cmt = "central_enm", evid = 1L, dur = 2),
    tidyr::expand_grid(cov, time = seq(0, 24, by = 0.25)) |>
      mutate(amt = NA_real_, cmt = "Cc",     evid = 0L, dur = NA_real_),
    tidyr::expand_grid(cov, time = seq(0, 24, by = 0.25)) |>
      mutate(amt = NA_real_, cmt = "Cc_enm", evid = 0L, dur = NA_real_)
  ) |> arrange(id, time, desc(evid))
  rxode2::rxSolve(mod, ev, returnType = "data.frame") |>
    distinct(id, time, .keep_all = TRUE) |>
    group_by(id) |>
    summarise(fep = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
              enm = sum(diff(time) * (head(Cc_enm, -1) + tail(Cc_enm, -1)) / 2),
              .groups = "drop")
}

hlth <- arm(0)
infd <- arm(1)

spread <- function(x) stats::IQR(log(x))   # log-scale IQR, robust to the tails

strata_tbl <- data.frame(
  Analyte = c("Cefepime", "Enmetazobactam"),
  `Median AUC ratio, infected / healthy` =
    round(c(stats::median(infd$fep) / stats::median(hlth$fep),
            stats::median(infd$enm) / stats::median(hlth$enm)), 3),
  `Log-scale IQR ratio, infected / healthy` =
    round(c(spread(infd$fep) / spread(hlth$fep),
            spread(infd$enm) / spread(hlth$enm)), 2),
  check.names = FALSE
)
knitr::kable(
  strata_tbl,
  caption = "Single-dose AUC0-24 in 200 virtual subjects per arm, identical except for infection status."
)

stopifnot(
  # "differences in mean exposures ... were negligible" for cefepime, whose
  # volume carries no infection term at all.
  abs(strata_tbl[1, 2] - 1) < 0.10,
  # "variability was higher in cUTI patients" -- for both drugs, decisively.
  strata_tbl[1, 3] > 1.4,
  strata_tbl[2, 3] > 1.4
)
```

# Assumptions and deviations

* **The paper of record publishes only part of the model.** The IDWeek abstract
  gives the typical values for a 70 kg cUTI patient, their %CV and the terminal
  half-lives; it names the covariates but prints no coefficients, no
  correlations and no residual error. The complete parameter set encoded here is
  transcribed from the FDA NDA 216165 Integrated Review (Table 101) and the EMA
  EPAR EMA/63929/2024 (Tables 5 and 6), which are two independent reproductions
  of the same sponsor study report, AAI101-PK-21-01. They agree with each other
  on every value, and with the abstract on every value the abstract prints.

* **EMA prose contradicts EMA tables on the weight exponents, and the tables
  win.** The EMA assessment report's "Updated model" paragraph on page 48 says
  the previous body-weight coefficients on the central volumes were "0.618 for
  cefepime and 0.802 for enmetazobactam". Its own Tables 5 and 6, on the two
  preceding pages, give 0.802 for cefepime and 0.618 for enmetazobactam -- as
  does FDA Table 101. Two independent tables outvote one sentence, so the model
  uses 0.802 (cefepime) and 0.618 (enmetazobactam).

* **The covariate reference values are not in the paper of record.** The 70 kg /
  100 mL/min / 50 years reference individual is stated only in the EMA report
  (page 46). The 70 kg element is independently confirmed by the abstract, whose
  "70 kg cUTI patient" cefepime Vc of 11.2 L equals the untransformed estimate;
  the 100 mL/min element is independently confirmed by Check 5, where inverting
  FDA Table 104 on a reference of 100 mL/min lands all six implied eGFRs inside
  their own renal bands.

* **The infected 4x4 correlation block needed a positive-definiteness repair.**
  Built from the printed three-decimal correlations verbatim the block is
  indefinite (smallest eigenvalue -1.7e-04) and `rxSolve()` cannot sample from
  it. The model uses the nearest positive-definite correlation matrix, which
  moves no correlation by more than 6.7e-05 -- inside the rounding interval of
  the printed values, so every repaired correlation rounds back to what was
  published. Check 3 asserts this.

* **Two correlations are printed with the same value.** FDA Table 101 lists both
  "CEF V1, FEP Cl in infected" and "ENM V1, FEP Cl in infected" as 0.261 with an
  RSE of 19. The first is independently confirmed by EMA Table 5 ("V1, Cl in
  cUTI = 0.261"); the second is a cross-drug term that appears only in the FDA
  table. Both are encoded as printed. If the repetition is a transcription slip
  in the FDA table rather than a genuine coincidence, the affected element is
  the cefepime-CL / enmetazobactam-Vc covariance.

* **Cefepime has no healthy-stratum central-volume variability.** FDA Table 101
  prints "-" for both the estimate and the RSE of `omega_FEP,V1_healthy`, so
  healthy volunteers take the typical cefepime central volume exactly. This is
  encoded by omitting the healthy arm of that multiplexed eta, not by setting a
  variance to zero.

* **The infection covariate is a single pooled indicator.** The source analysis
  carries one coefficient for the whole infected cohort rather than separate
  cUTI and acute-pyelonephritis levels, so `DIS_CUTI` is used alone (set it to 1
  for cUTI *and* acute pyelonephritis patients, 0 only for healthy volunteers)
  and `DIS_AP` is deliberately not paired with it.

* **The enmetazobactam additive residual error was converted from ng/mL.** FDA
  Table 101 reports `a_ENM = 40.9 ng/mL` because the analysis dataset held
  concentrations in ng/mL. This model works in mg/L, so the value is encoded as
  0.0409 mg/L. Cefepime has no additive component at all.

* **The sex effect is carried by the male level.** The published reference
  individual is female, so the enmetazobactam peripheral volume of 5.28 L is the
  female value and males carry `exp(0.198)`. The canonical `SEXF` column codes
  female as 1, so the model term is written on `(1 - SEXF)` and the parameter is
  named `e_sexmale_vp_enm`.

* **A later "updated model" exists and is not what is encoded.** The EMA report
  (page 48) describes a re-fit in which body weight scales the central *and*
  peripheral volumes with a single shared coefficient (0.407 and 0.553 for the
  two drugs), improving the objective function by 16 points. That variant is not
  the model the abstract, the FDA review or the EMA parameter tables report, and
  its full parameter set is not published, so this file encodes the final model
  of record. Note that the same sentence carrying those two numbers is the one
  that swaps the drugs, so their drug assignment should not be trusted either.

* **Concentrations are total, not free.** Multiply by 0.80 for cefepime (20%
  protein bound) and by 1.00 for enmetazobactam (binding negligible) to obtain
  the free concentrations that the `%fT > MIC` and `%fT > CT` targets are defined
  on. Check 5 confirms that FDA Table 104 is likewise on a total-drug basis.

* **The cohort covariate distributions in Check 6 are assumed.** FDA Table 100
  publishes only means and ranges, not distributional shapes or correlations, so
  weight, age and de-indexed eGFR are drawn independently from truncated
  log-normals calibrated to the published means. The assertions are therefore on
  the cohort centre, not on any extreme.

* **Not extracted from this source.** The EMA report also contains a separate
  plasma-and-epithelial-lining-fluid population PK model built from the 19
  healthy volunteers of Phase 1 study AT-103 (its Table 7), and a hemodialysis
  extension of the plasma model (FDA Figure 22). Neither is the model this
  abstract describes; both are separate analyses with their own primary sources.
