---
title: "Efavirenz (Vucicevic 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efavirenz (Vucicevic 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- rxode2::rxode2(readModelDb("Vucicevic_2025_efavirenz"))
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.7759/cureus.88533>

Vucicevic 2025 is a population pharmacokinetic-pharmacogenetic (PopPK-PGx)
re-analysis of a Serbian cohort of adults with HIV-1 who were previously
reported in a single-timepoint concentration study (Olagunju 2014, the
paper's reference 17). The analysis quantifies the effect of two *CYP2B6*
polymorphisms on apparent oral efavirenz clearance and then uses the fitted
model to simulate the probability of achieving the 1,000-4,000 ng/mL
therapeutic window under different pharmacogenetic backgrounds and different
non-adherence patterns.

# Population

```{r population}
pop <- mod$population
```

The analysis dataset comprised **`r pop$n_subjects` adults** with confirmed
HIV-1 infection, all of white Caucasian ethnicity, receiving 600 mg efavirenz
once daily as part of efavirenz-based antiretroviral therapy for at least
three months (Methods, "PopPK-PGx model development"). Median body weight was
76 kg (IQR 66-85) and median age 40 years (IQR 32-48); 67.4% were smokers
(Table 1). Patients on concomitant antituberculosis medication or other drugs
known to interact with efavirenz metabolism, patients under 18 years, and
pregnant women were excluded.

The design is **extremely sparse: exactly one steady-state plasma sample per
patient**, drawn between 2.83 and 13.42 h after the last dose (Results
paragraph 1; Appendix 1 Figure 4). That single-sample design is the reason the
absorption rate constant and the apparent volume of distribution could not be
estimated and were fixed to literature values.

Genotype frequencies in the cohort (Table 1):

```{r genotype-table, echo = FALSE}
knitr::kable(
  data.frame(
    Polymorphism = c(
      "CYP2B6 c.516G>T (rs3745274)",
      "CYP2B6 c.485-18C>T (rs4803419)",
      "NR1I3 c.540C>T (rs2307424)",
      "NR1I3 c.152-1089T>C (rs3003596)"
    ),
    `Flagged n (%)` = c("30 (33.71)", "12 (13.48)", "53 (59.60)", "58 (65.17)"),
    `Retained in final model` = c("Yes", "Yes", "No", "No"),
    check.names = FALSE
  ),
  caption = "Genotype frequencies in the Vucicevic 2025 cohort (Table 1)."
)
```

Neither *NR1I3* polymorphism showed a detectable effect on efavirenz CL/F
(Discussion paragraph 4); both are recorded in the model file's
`covariatesDataExcluded` metadata together with the demographic covariates
that were screened but not retained.

# Source trace

Every value in the model file, and where it comes from.

```{r source-trace, echo = FALSE}
knitr::kable(
  data.frame(
    Item = c(
      "Structural model",
      "ka",
      "V/F",
      "CL/F (CLp)",
      "CL/F covariate equation",
      "theta_RSA (CYP2B6 516G>T carrier)",
      "theta_RSB (CYP2B6 c.485-18C>T TT)",
      "IIV on CL/F",
      "Residual error",
      "RSA definition",
      "RSB definition",
      "Dosing regimen simulated",
      "Therapeutic window",
      "Reference Ctrough by genotype"
    ),
    Value = c(
      "One compartment, first-order absorption and elimination (ADVAN2 TRANS2)",
      "0.3 1/h (fixed)",
      "237 L (fixed)",
      "13.9 L/h (RSE 4%)",
      "CL/F = CLp * (1 + theta_RSA)^RSA * (1 + theta_RSB)^RSB",
      "-0.364 (RSE 14%)",
      "-0.268 (RSE 16%)",
      "CV 13.1% (RSE 35%), exponential / log-normal",
      "Proportional, SD 0.25 (RSE 11%)",
      "Binary carrier indicator for CYP2B6 516G>T",
      "Binary indicator, CYP2B6 c.485-18C>T TT genotype",
      "600 mg once daily",
      "1,000-4,000 ng/mL",
      "1,179.4 / 1,757.15 / 2,140.05 / 3,072.75 ng/mL"
    ),
    `Source location` = c(
      "Methods, 'PopPK-PGx model development' paragraph 3",
      "Methods paragraph 3 and Discussion paragraph 2 (cited to reference 20, Csajka 2003)",
      "Methods paragraph 3 and Discussion paragraph 2 ('237 L/70 kg', references 9, 11, 18, 19)",
      "Table 2, row CLp (bootstrap 13.9; 95% CI 12.73-15.21)",
      "Table 2, header row of the 'Fixed effects' block",
      "Table 2, row theta_RSA (bootstrap -0.368; 95% CI -0.457 to -0.255)",
      "Table 2, row theta_RSB (no bootstrap value reported)",
      "Table 2, row CV CL (%); Results paragraph 2 (bootstrap 13.4%; 95% CI 3.24-20.9%)",
      "Table 2, row Proportional error; Results paragraph 2 (bootstrap 0.244; 95% CI 0.182-0.299)",
      "Table 2 footnote",
      "Table 2 footnote (called 'carrier'), resolved to TT by Results paragraph 2 and the Table 3 row labels",
      "Methods, 'Model-based dose simulations'",
      "Methods, 'Model-based dose simulations'; Results, 'Model-derived dosing implications'",
      "Table 3, column 'Median C trough (95% CI)'"
    ),
    check.names = FALSE
  ),
  caption = "Source trace for the Vucicevic 2025 efavirenz model."
)
```

The `ini()` block reproduces `CL/F = CLp * (1 + theta_RSA)^RSA * (1 + theta_RSB)^RSB`
literally. Because `RSA` and `RSB` are 0/1 indicators, each factor collapses to
1 when its indicator is 0. The canonical covariate columns are 0/1/2 allele
counts, from which `model()` recovers the paper's indicators as
`RSA = (SNP_CYP2B6_RS3745274_T_COUNT >= 1)` (dominant / carrier model) and
`RSB = (SNP_CYP2B6_RS4803419_T_COUNT == 2)` (recessive / TT-only model).

# Virtual cohort

Table 3 of the paper simulates four genotype combinations at the standard
600 mg once-daily regimen. We reproduce the same four arms, each with 200
virtual subjects carrying inter-individual variability on CL/F.

```{r cohort}
tau        <- 24     # dosing interval (h)
n_doses    <- 21     # dose at t = 0, 24, ..., 480 -- >= 19 half-lives for the slowest arm
dose_mg    <- 600
n_per_arm  <- 200
t_last     <- tau * (n_doses - 1)

genotypes <- tibble::tribble(
  ~treatment,                     ~rs3745274_T, ~rs4803419_T,
  "516GG, c.485-18CC",            0L,           0L,
  "516GG, c.485-18TT",            0L,           2L,
  "516GT, c.485-18CC",            1L,           0L,
  "516GT, c.485-18TT",            1L,           2L
)
arm_levels <- genotypes$treatment

subjects <-
  genotypes |>
  tidyr::crossing(rep = seq_len(n_per_arm)) |>
  dplyr::mutate(
    id = dplyr::row_number(),
    SNP_CYP2B6_RS3745274_T_COUNT = rs3745274_T,
    SNP_CYP2B6_RS4803419_T_COUNT = rs4803419_T
  ) |>
  dplyr::select(
    id, treatment,
    SNP_CYP2B6_RS3745274_T_COUNT, SNP_CYP2B6_RS4803419_T_COUNT
  )

obs_times <- seq(t_last, t_last + tau, by = 0.25)

# Build the event table as a plain data.frame: covariate columns assigned onto
# an rxEt object are silently dropped by rxode2.
build_events <- function(subj, dose_times, obs_times) {
  doses <- subj |>
    tidyr::crossing(time = dose_times) |>
    dplyr::mutate(amt = dose_mg, evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

events <- build_events(subjects, seq(0, t_last, by = tau), obs_times)
nrow(events)
```

Observation records point at the `central` ODE state, never at the algebraic
observable `Cc`; rxode2 returns `Cc` as a column at those rows automatically.

# Simulation

```{r simulate}
set.seed(20250722)
sim <-
  rxode2::rxSolve(
    mod$simulationModel, events,
    omega = mod$omega,           # pass omega explicitly; rxSolve reuses the previous solve's omega
    keep = c(
      "treatment",
      "SNP_CYP2B6_RS3745274_T_COUNT", "SNP_CYP2B6_RS4803419_T_COUNT"
    )
  ) |>
  as.data.frame() |>
  dplyr::mutate(
    treatment = factor(treatment, levels = arm_levels),
    Cc_ngml   = Cc * 1000        # model concentration is mg/L = ug/mL; the paper reports ng/mL
  )

# Guards: no subjects silently dropped, and IIV really was applied.
stopifnot(dplyr::n_distinct(sim$id) == nrow(subjects))
stopifnot(dplyr::n_distinct(round(sim$cl, 8)) > length(arm_levels))
stopifnot(all(sim$Cc >= 0))
```

## Replication of Figure 2 -- steady-state profiles by genotype

```{r figure2, fig.width = 7, fig.height = 4.5, fig.cap = "Replicates Figure 2 of Vucicevic 2025: simulated steady-state efavirenz concentration-time profiles for a 600 mg once-daily regimen under four CYP2B6 genotype combinations. Lines are medians, ribbons the 5th-95th percentiles across 200 virtual subjects per arm. Dashed lines mark the 1,000-4,000 ng/mL therapeutic window."}
prof <-
  sim |>
  dplyr::mutate(tad = time - t_last) |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    med = stats::median(Cc_ngml),
    lo  = stats::quantile(Cc_ngml, 0.05),
    hi  = stats::quantile(Cc_ngml, 0.95),
    .groups = "drop"
  )

ggplot2::ggplot(prof, ggplot2::aes(tad, med, colour = treatment, fill = treatment)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = c(1000, 4000), linetype = "dashed") +
  ggplot2::labs(
    x = "Time after dose at steady state (h)",
    y = "Efavirenz concentration (ng/mL)",
    colour = "CYP2B6 genotype", fill = "CYP2B6 genotype"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

# PKNCA validation

Steady-state non-compartmental analysis over the final dosing interval
(recipe 3: `AUC0-tau`, `Cmax,ss`, `Cmin,ss`, `Cav,ss`).

```{r pknca}
sim_nca <-
  sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# `events` already carries `treatment` (it is built from `subjects`), so no
# join is needed -- joining it again would produce treatment.x / treatment.y.
dose_df <-
  events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | treatment + id,
  concu = "ug/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + id,
  doseu = "mg"
)

start_ss <- max(dose_df$time)
intervals <- data.frame(
  start   = start_ss,
  end     = start_ss + tau,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
summary(res)
```

## Structural identity check: AUC0-tau = Dose / (CL/F)

At steady state the area under one dosing interval must equal `Dose / CL`
exactly. Checking it on the typical-value (no random effects) solve is a
one-sigma test of the whole parameterisation.

```{r identity}
mod_typ <- rxode2::zeroRe(mod)
typ_events <- build_events(subjects[!duplicated(subjects$treatment), ],
                           seq(0, t_last, by = tau), obs_times)

typ <-
  rxode2::rxSolve(mod_typ, typ_events, omega = NA,
                  keep = "treatment") |>
  as.data.frame() |>
  dplyr::mutate(treatment = factor(treatment, levels = arm_levels))

# zeroRe guard: CL must be a single value per arm.
stopifnot(dplyr::n_distinct(round(typ$cl, 8)) == length(arm_levels))

typ_nca <-
  PKNCA::pk.nca(PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(
      typ |> dplyr::filter(!is.na(Cc)) |> dplyr::select(id, time, Cc, treatment),
      Cc ~ time | treatment + id, concu = "ug/mL", timeu = "h"
    ),
    PKNCA::PKNCAdose(
      typ_events |>
        dplyr::filter(evid == 1) |>
        dplyr::select(id, time, amt, treatment),
      amt ~ time | treatment + id, doseu = "mg"
    ),
    intervals = data.frame(start = start_ss, end = start_ss + tau, auclast = TRUE)
  ))

identity_tbl <-
  as.data.frame(typ_nca$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(treatment, auc_nca = PPORRES) |>
  dplyr::left_join(
    typ |>
      dplyr::group_by(treatment) |>
      dplyr::summarise(cl = dplyr::first(cl), .groups = "drop"),
    by = "treatment"
  ) |>
  dplyr::mutate(
    auc_closed_form = dose_mg / cl,
    pct_diff        = 100 * (auc_nca - auc_closed_form) / auc_closed_form
  )

stopifnot(max(abs(identity_tbl$pct_diff)) < 1)

identity_tbl |>
  dplyr::rename(
    "CYP2B6 genotype"       = treatment,
    "CL/F (L/h)"            = cl,
    "AUC0-tau, NCA"         = auc_nca,
    "Dose / (CL/F)"         = auc_closed_form,
    "% difference"          = pct_diff
  ) |>
  knitr::kable(
    digits = c(0, 3, 3, 3, 3),
    caption = "Steady-state AUC0-tau (ug*h/mL) from PKNCA against the closed-form Dose / (CL/F) identity, typical individual per genotype."
  )
```

The agreement is within `r sprintf("%.3f", max(abs(identity_tbl$pct_diff)))`%,
confirming that the covariate equation, the fixed volume, and the dosing units
compose correctly.

# Comparison against the published simulation

The paper reports no observed-data NCA parameters (a single sample per patient
makes NCA impossible). The published quantities available for comparison are
the simulated median trough concentrations in Table 3. At steady state on a
once-daily regimen, `Cmin` over a dosing interval is the trough, so the
comparison is made on the PKNCA `cmin` parameter.

```{r ctrough-comparison}
# Table 3, "Median C trough (95% CI)", converted from ng/mL to ug/mL.
reference_ctrough <- data.frame(
  treatment = arm_levels,
  cmin      = c(1179.40, 1757.15, 2140.05, 3072.75) / 1000
)

ctrough_tbl <- nlmixr2lib::ncaComparisonTable(
  simulated = as.data.frame(res$result) |> dplyr::filter(PPTESTCD == "cmin"),
  reference = reference_ctrough,
  by        = "treatment",
  params    = "cmin",
  units     = c(cmin = "ug/mL")
)
knitr::kable(
  ctrough_tbl,
  digits = 3,
  caption = "Simulated versus published (Vucicevic 2025 Table 3) median steady-state trough efavirenz concentrations by CYP2B6 genotype."
)
fn <- attr(ctrough_tbl, "footnote")
if (!is.null(fn)) cat(fn)

# ncaComparisonTable formats "% diff" as a display string ("-13.7%", with a
# trailing "*" when flagged); strip both to recover the numeric values.
ctrough_pct <- as.numeric(gsub("[%*]", "", ctrough_tbl[["% diff"]]))
stopifnot(!anyNA(ctrough_pct), max(abs(ctrough_pct)) < 20)
```

All four arms agree with the published medians inside the 20% flagging
tolerance; the simulated troughs run
`r sprintf("%.1f", min(abs(ctrough_pct)))`%
to
`r sprintf("%.1f", max(abs(ctrough_pct)))`%
below the paper, with the largest gap in the wild-type arm. See
*Assumptions and deviations* for why the residual gap is a property of the
paper's own simulation reporting rather than of the transcribed parameters.

## Steady-state peaks against Figure 2

Table 3 tabulates only troughs, but Figure 2 draws the whole steady-state
profile, so the peaks are a second, independent check on the same parameters.
This matters because the two candidate explanations for the trough gap make
opposite predictions here: a mis-transcribed CL/F or V/F displaces peak and
trough together, whereas a trough read at the wrong time within the interval
displaces only the trough.

Figure 2 is a four-panel faceted plot on a linear axis with dashed reference
lines drawn at exactly 1,000 and 4,000 ng/mL. Those two lines calibrate the
axis, so the black typical-profile trace can be digitized to about +/- 25 ng/mL
(roughly half the drawn line width). The values below were read by pixel
calibration against the two reference lines; they come from the figure, not
from the paper's text or tables.

```{r peaks}
# Digitized from Figure 2 of Vucicevic 2025 (pixel calibration against the
# 1,000 and 4,000 ng/mL dashed reference lines; ~ +/- 25 ng/mL).
figure2_digitized <- data.frame(
  treatment  = arm_levels,
  fig_cmax   = c(2404, 3082, 3438, 4496),
  fig_ctrough = c(1083, 1726, 2083, 3115)
)

peaks_tbl <-
  typ |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cmax_ngml = max(Cc) * 1000,
    cmin_ngml = min(Cc) * 1000,
    .groups   = "drop"
  ) |>
  dplyr::left_join(figure2_digitized, by = "treatment") |>
  dplyr::mutate(
    cmax_pct = 100 * (cmax_ngml - fig_cmax) / fig_cmax,
    cmin_pct = 100 * (cmin_ngml - fig_ctrough) / fig_ctrough
  ) |>
  dplyr::arrange(treatment)

# The peaks are the discriminating measurement: they must agree tightly.
stopifnot(max(abs(peaks_tbl$cmax_pct)) < 2)
# The troughs are expected to sit low; assert the direction and the magnitude.
stopifnot(all(peaks_tbl$cmin_pct < 0), max(abs(peaks_tbl$cmin_pct)) < 10)

peaks_tbl |>
  dplyr::rename(
    "CYP2B6 genotype"        = treatment,
    "Cmax,ss simulated"      = cmax_ngml,
    "Cmax,ss Figure 2"       = fig_cmax,
    "Cmax % diff"            = cmax_pct,
    "Ctrough,ss simulated"   = cmin_ngml,
    "Ctrough,ss Figure 2"    = fig_ctrough,
    "Ctrough % diff"         = cmin_pct
  ) |>
  knitr::kable(
    digits  = c(0, 0, 0, 2, 0, 0, 2),
    caption = "Typical-individual steady-state peak and trough efavirenz concentrations (ng/mL) against the profiles drawn in Figure 2 of Vucicevic 2025."
  )
```

The peaks reproduce the figure to within
`r sprintf("%.1f", max(abs(peaks_tbl$cmax_pct)))`% in every arm, while the
troughs sit `r sprintf("%.1f", min(abs(peaks_tbl$cmin_pct)))`-`r sprintf("%.1f", max(abs(peaks_tbl$cmin_pct)))`%
low in every arm and the Table 3 medians sit higher still. Peak agreement at
the 1% level rules out a mis-transcribed CL/F or V/F -- either would have moved
the peaks by the same proportion as the troughs -- so the residual disagreement
is localised to how the trough was read, not to the transcribed parameters.

## Probability of falling outside the therapeutic window

Table 3 also reports the percentage of simulated patients whose trough falls
below 1,000 ng/mL or above 4,000 ng/mL.

```{r window}
window_tbl <-
  sim |>
  dplyr::filter(time == max(time)) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    median_ctrough = stats::median(Cc_ngml),
    below_1000     = 100 * mean(Cc_ngml < 1000),
    above_4000     = 100 * mean(Cc_ngml > 4000),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    data.frame(
      treatment       = arm_levels,
      paper_below     = c(34.80, 0.02, 0.00, 0.00),
      paper_above     = c(0.00, 0.00, 0.00, 5.30)
    ),
    by = "treatment"
  )

window_tbl |>
  dplyr::rename(
    "CYP2B6 genotype"           = treatment,
    "Median Ctrough (ng/mL)"    = median_ctrough,
    "Simulated < 1,000 (%)"     = below_1000,
    "Simulated > 4,000 (%)"     = above_4000,
    "Published < 1,000 (%)"     = paper_below,
    "Published > 4,000 (%)"     = paper_above
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Simulated versus published (Table 3) proportions outside the 1,000-4,000 ng/mL therapeutic window."
  )
```

The ordering is reproduced -- the wild-type arm is the only one at material
risk of sub-therapeutic exposure and the double-variant arm the only one at
risk of supra-therapeutic exposure -- but the magnitudes differ. Table 3 of the
paper is internally inconsistent on this point: a log-normal distribution with
the published median (1,179.4 ng/mL) and published 95% interval
(793.94-1,720 ng/mL) implies about 20% of the wild-type arm below 1,000 ng/mL,
whereas the same table reports 34.8%. Neither number can be reproduced from the
other, so this row is reported for transparency rather than treated as a
validation target.

# Replication of Figure 3 -- missed doses

```{r missed-doses}
miss_times <- seq(t_last, t_last + 4 * tau, by = 0.5)
miss_events <- build_events(subjects[!duplicated(subjects$treatment), ],
                            seq(0, t_last, by = tau), miss_times)

miss <-
  rxode2::rxSolve(mod_typ, miss_events, omega = NA, keep = "treatment") |>
  as.data.frame() |>
  dplyr::mutate(
    treatment = factor(treatment, levels = arm_levels),
    Cc_ngml   = Cc * 1000,
    tad       = time - t_last
  )

# Counting convention: with n consecutive doses missed, the relevant trough is
# the concentration at the moment the n-th missed dose was due, i.e. at
# t_last + n * tau. n = 1 is therefore the ordinary steady-state trough.
missed_long <-
  miss |>
  dplyr::filter(tad %in% (tau * (1:4))) |>
  dplyr::mutate(missed = as.integer(tad / tau)) |>
  dplyr::select(treatment, missed, Cc_ngml)

# Smallest number of consecutive missed doses that puts the typical individual
# below the 1,000 ng/mL lower therapeutic limit, joined by key to the counts
# stated in Results, "Model-derived dosing implications".
published_missed <- data.frame(
  treatment    = arm_levels,
  paper_missed = c(1L, 2L, 2L, 3L)
)

missed_tbl <-
  missed_long |>
  tidyr::pivot_wider(names_from = missed, values_from = Cc_ngml,
                     names_prefix = "miss") |>
  dplyr::left_join(
    missed_long |>
      dplyr::filter(Cc_ngml < 1000) |>
      dplyr::group_by(treatment) |>
      dplyr::summarise(sim_missed = min(missed), .groups = "drop"),
    by = "treatment"
  ) |>
  dplyr::left_join(published_missed, by = "treatment") |>
  dplyr::arrange(treatment)

missed_tbl |>
  dplyr::rename(
    "CYP2B6 genotype"        = treatment,
    "1 missed dose"          = miss1,
    "2 missed doses"         = miss2,
    "3 missed doses"         = miss3,
    "4 missed doses"         = miss4,
    "Simulated n below 1,000" = sim_missed,
    "Published n"            = paper_missed
  ) |>
  knitr::kable(
    digits = 0,
    caption = "Typical-individual efavirenz concentration (ng/mL) at the time the n-th consecutive missed 600 mg dose was due, at steady state. The final two columns give the smallest n that drops the typical individual below the 1,000 ng/mL lower therapeutic limit, against the count stated by Vucicevic 2025."
  )
```

```{r figure3, fig.width = 7, fig.height = 4.5, fig.cap = "Replicates Figure 3 of Vucicevic 2025: typical-individual efavirenz concentration after the last taken 600 mg dose at steady state, with all subsequent doses missed. The dashed line is the 1,000 ng/mL lower therapeutic limit; vertical lines mark successive missed-dose times."}
ggplot2::ggplot(miss, ggplot2::aes(tad, Cc_ngml, colour = treatment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::geom_hline(yintercept = 1000, linetype = "dashed") +
  ggplot2::geom_vline(xintercept = tau * (1:4), linetype = "dotted",
                      colour = "grey50") +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time after the last taken dose (h)",
    y = "Efavirenz concentration (ng/mL, log scale)",
    colour = "CYP2B6 genotype"
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

The paper's Results, "Model-derived dosing implications", states that
concentrations fall out of the therapeutic range after **three** consecutive
missed doses for carriers of both polymorphisms, **two** for carriers of
either one, and **one** for carriers of neither. The typical-individual
simulation reproduces those counts: three for the double-variant arm and two
for each single-variant arm, matching exactly. The wild-type arm is the one
boundary case -- its typical individual sits at 1,017 ng/mL at the first
missed dose, 1.7% above the 1,000 ng/mL limit, so the strict threshold test
returns two rather than the stated one. That arm is precisely the one the
paper describes as already marginal: Table 3 puts 34.8% of it below
1,000 ng/mL at the ordinary steady-state trough, so the published "one missed
dose" is a statement about a population straddling the limit rather than about
a typical individual clearing it.

The counting convention matters here and the paper does not state it. The
counts above assess the trough at the moment the *n*-th missed dose was due
(`t = n * tau` after the last dose taken), so `n = 1` is the ordinary
steady-state trough. Reading `n` as *additional* intervals beyond the normal
trough instead (`t = (n + 1) * tau`) shifts every arm by one and reproduces
none of the published counts, which is what identifies the first convention as
the paper's.

# Assumptions and deviations

* **V/F is encoded as a constant 237 L, not as a weight-scaled term.** The
  paper writes the fixed value as "237 L/70 kg" (Methods paragraph 3) and as
  "237 L for a body weight (BW) of 70 kg" (Discussion paragraph 2), which could
  be read either as a plain constant whose literature source happened to be a
  ~70 kg subject, or as a linear weight-scaling term `237 * (WT / 70)`. Three
  things settle it in favour of the constant: (a) the covariate screen tested
  body weight **on CL/F only** (Methods paragraph 3: "These were tested as
  covariates for CL"); (b) the Table 2 final-model equation contains no weight
  term, and body weight was not retained anywhere in the final model; and (c)
  of the four references cited for the fixed V/F (references 9, 11, 18, 19),
  the one already available in this package -- Sanchez 2011, reference 11,
  `modellib("Sanchez_2011_efavirenz")`, fitted in a Caucasian HIV-infected
  cohort like this one -- reports V/F as a plain constant (247 L) with no
  covariate effects on V/F at all. The paper's own Figure 2 then settles it
  numerically: steady-state peak concentration is far more sensitive to V/F
  than trough concentration is, and the constant 237 L reproduces the four
  digitized peaks to within 1%, whereas the weight-scaled reading
  `237 * (76 / 70)` = 257 L predicts a wild-type peak of about 2,362 ng/mL
  against the 2,404 ng/mL drawn -- roughly a five-fold-worse fit, and low in
  every arm rather than scattered about zero.
* **`RSB` is the c.485-18C>T TT stratum, not any-T carriage.** The Table 2
  footnote calls `RSB` a "carrier" indicator, but Results paragraph 2 reports
  the effect "in patients with TT genotype (n=12)" with the same n as the
  Table 1 "carriers" row, and Table 3 contrasts the simulated strata as
  "c.485-18TT" versus "c.485-18CC". A ~45% any-T carrier frequency would be
  expected in a European-ancestry cohort against the 13.48% observed, which
  matches the expected TT-homozygote frequency. The model therefore pools CT
  heterozygotes with CC wild-type in the reference group.
* **`RSA` is a carrier (dominant) indicator.** No TT homozygotes for
  c.516G>T appear in the cohort (Results paragraph 2 reports GT n=30 versus GG
  n=60), so carrier and heterozygote coincide in the fitted data. Applying the
  model to a 516TT subject assigns the same -36.4% shift as to a heterozygote;
  that is an extrapolation beyond the fitted cohort, not a published claim.
* **Simulated troughs run 2-14% below the published Table 3 medians.** The
  paper's simulated medians cannot be reproduced by any single value of V/F:
  fitting V/F to the four published medians requires roughly 290 L, and the
  residual pattern (largest gap in the fastest-clearing arm, smallest in the
  slowest) is the signature of the trough being read at a slightly earlier time
  than 24 h post dose rather than of a mis-transcribed parameter. The paper's
  own Figure 2 discriminates between those two explanations, and it does so
  decisively: digitized against the figure's own 1,000 and 4,000 ng/mL
  reference lines, the simulated steady-state *peaks* match all four drawn
  peaks to within 1% (2,413 vs 2,404; 3,062 vs 3,082; 3,430 vs 3,438; 4,458 vs
  4,496 ng/mL), while the troughs are low by 3.5-6.1% in every arm. A
  mis-transcribed CL/F or V/F would move peak and trough by the same
  proportion, so the disagreement localises to how the trough was read, not to
  the parameters. Note also that the figure's own troughs sit below the Table 3
  medians the figure is supposed to depict, so the two published readouts of
  the same simulation do not agree with each other. Consistent
  with this, the *spread* of the published Table 3 confidence intervals is
  reproduced closely: the elasticity of trough concentration with respect to
  CL/F is -1.63 in the wild-type arm and -1.26 in the double-variant arm,
  which multiplied by the published 13.1% CV on CL/F gives log-scale SDs of
  0.21 and 0.17 against the 0.197 and 0.163 implied by the published intervals.
  That agreement confirms both the 13.1% IIV scale (it is a CV, not a variance)
  and that the published intervals reflect inter-individual variability only,
  with no residual error added. No parameter was tuned to close the median gap.
* **The missed-dose counting convention is inferred, not stated.** The paper
  reports that 1 / 2 / 2 / 3 consecutive missed doses take the four arms below
  1,000 ng/mL but never defines when the trough is assessed. Assessing it at
  the moment the *n*-th missed dose was due (`t = n * tau` after the last dose
  taken) reproduces three of the four published counts exactly and leaves the
  wild-type arm 1.7% above the limit at `n = 1`; assessing it one interval
  later (`t = (n + 1) * tau`) reproduces none of them. The first convention is
  adopted on that basis. Only the labelling of the columns depends on this
  choice -- the predicted concentrations themselves are reported raw in the
  table above, so a reader preferring the other convention can shift the
  header row.
* **`n = 89` versus `n = 90` genotyped.** Results paragraph 2 reports 30 GT
  and 60 GG subjects for c.516G>T, which sums to 90 against the stated cohort
  size of 89. All Table 1 percentages are consistent with a denominator of 89.
  The one-subject discrepancy does not affect any transcribed parameter.
* **Sex distribution.** Table 1 reports "Sex, male; n (%) 21 (23.60)", i.e.
  76.4% female. That is an unusual composition for a European HIV-1 cohort and
  may be a row-label transposition in the source, but sex was not retained in
  the final model, so nothing in the model file depends on it. The value is
  recorded in `population$sex_female_pct` as printed.
* **Non-paper provenance: the Figure 2 comparators are digitized.** The eight
  `fig_cmax` / `fig_ctrough` values in the *Steady-state peaks against Figure 2*
  section were read off the published figure by pixel calibration against its
  own 1,000 and 4,000 ng/mL dashed reference lines, not transcribed from text
  or a table; they carry roughly +/- 25 ng/mL of reading error. They are used
  only as validation comparators. **No `ini()` parameter in the model file
  derives from a figure** -- every parameter comes from Table 2 or the Methods
  text, as recorded in the source-trace table above.
* **Units.** The model works in mg and L, so `Cc` is mg/L = ug/mL; the paper
  reports concentrations in ng/mL. Conversions in this vignette are explicit
  (`Cc * 1000`).
* **Bootstrap value for theta_RSB.** Table 2 leaves the bootstrap column blank
  for `theta_RSB`; only the final-model point estimate and RSE are available.
* **No observed-data NCA comparison is possible.** With one sample per patient
  the source has no published Cmax / AUC / half-life to compare against; the
  PKNCA section characterises the simulated profiles and anchors them to the
  `AUC0-tau = Dose / (CL/F)` identity instead.

# Errata

No erratum or correction notice for Vucicevic 2025 (Cureus 17(7):e88533,
published 22 July 2025) was located at the time of extraction.

Two internal inconsistencies in the source are noted for the record; neither
affects any transcribed parameter:

* The Abstract states the analysis used **NONMEM v7.3**, while Methods
  paragraph 2 states **NONMEM version 7.4**. The model file records the
  Methods value.
* Results paragraph 2 quotes the CL/F relative standard error as **4.4%**,
  while Table 2 prints **4**. Both are recorded in the source trace above.
