---
title: "Brexanolone (Wald 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Brexanolone (Wald 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wald_2022_brexanolone"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wald_2022_brexanolone"))$description`
* Article: <https://doi.org/10.1007/s40262-022-01155-w> (open access)
* Electronic Supplementary Material (ESM 1, model development, Table S1 final
  parameters and VPCs): available from the article page.

Brexanolone is a synthetic, chemically identical form of the endogenous
neuroactive steroid allopregnanolone, given as a 60-h weight-based
continuous intravenous infusion for postpartum depression (PPD). The model
predicts allopregnanolone concentrations in plasma and in breast milk; the
milk concentration is the plasma concentration multiplied by an estimated
milk:plasma partition coefficient.

# Population

The population PK analysis pooled 156 women: 144 patients with PPD from the
placebo-controlled studies 202A (N = 10), 202B (N = 79) and 202C (N = 51) and
the open-label study 201 (N = 4), plus 12 healthy lactating volunteers
(<= 6 months postpartum) from the open-label lactation study 547-CLP-108
(Table 1). Median age was 27 years (range 18-42), median body weight 82.9 kg
(range 44.9-150; mean 85, SD 22), and mean BMI 31.3 kg/m^2. Race: 50.6% White,
32.1% Black or African American, 0.6% Asian, 3.2% other; 13.5% Hispanic. All
subjects received a single 60-h infusion titrated to 60 ug/kg/h (BRX60) or 90
ug/kg/h (BRX90). Plasma was sampled in all studies; breast milk only in
547-CLP-108. After exclusion of implausible samples and BLQ data, 1256
observations from 155 subjects were analysed in NONMEM 7.3 (ESM).

The same information is available programmatically via
`readModelDb("Wald_2022_brexanolone")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: two-compartment, linear elimination, IV infusion | -- | Main text 2.3.3 and 3.4; ESM 'Final PopPK Model' |
| `lcl` (CL at 82.9 kg) | 89.8 L/h | ESM Table S1 |
| `lvc` (V1 at 82.9 kg) | 117 L | ESM Table S1 |
| `lq` (Q at 82.9 kg) | 37.9 L/h | ESM Table S1 |
| `lvp` (V2 at 82.9 kg) | 470 L | ESM Table S1 |
| `lcmpr` (Kp, milk:plasma) | 1.36 | ESM Table S1; main text 3.4 |
| `e_wt_cl` (CL, Q) | 0.75 (fixed) | ESM 'PopPK Model Development'; Table S1 footnote |
| `e_wt_vc` (V1, V2) | 1 (fixed) | ESM 'PopPK Model Development'; Table S1 footnote |
| Reference weight | 82.9 kg | ESM Table S1 footnote; Table 1 median |
| `etalcl` variance | 0.0435 (21.1% CV) | ESM Table S1; ESM 'Final PopPK Model' |
| `etalcl`/`etalvc` covariance | 0.106 | ESM Table S1 |
| `etalvc` variance | 1.15 (147% CV) | ESM Table S1; ESM 'Final PopPK Model' |
| `propSd` (plasma) | 0.272 | ESM Table S1 'Proportional residual variability' |
| `propSd_Cmilk` (milk) | 0.272 | Same single estimate; see Assumptions |
| `Cmilk <- cmpr * Cc` | -- | Main text 2.3.3 and 3.4 (partition coefficient) |
| BRX90 schedule 30/60/90/60/30 ug/kg/h | 0-4/4-24/24-52/52-56/56-60 h | Main text 2.2 |

# Virtual cohorts

```{r cohort}
# rxSetSeed() fixes rxode2's draws on one machine but not across rxode2
# builds or thread counts, so every assertion below is on a median or a
# robust quantile, never on a cohort extreme.
set.seed(20220723)
rxode2::rxSetSeed(20220723)

mod <- readModelDb("Wald_2022_brexanolone")

# Infusion schedules (ug/kg/h). BRX90 is main text section 2.2; BRX60 follows
# the step pattern of Figure 4a (see Assumptions).
schedules <- list(
  BRX90 = data.frame(start = c(0, 4, 24, 52, 56), end = c(4, 24, 52, 56, 60),
                     rate = c(30, 60, 90, 60, 30)),
  BRX60 = data.frame(start = c(0, 4, 56), end = c(4, 56, 60),
                     rate = c(30, 60, 30))
)

# Log-normal weights truncated to an observed range.
draw_weights <- function(n, median_wt, sdlog, lo, hi) {
  wt <- exp(rnorm(n * 3, log(median_wt), sdlog))
  wt <- wt[wt >= lo & wt <= hi]
  wt[seq_len(n)]
}

# Dose records are weight-based zero-order infusions into `central`;
# observation rows sit on the `central` state with dvid = 1 so rxode2 returns
# both Cc and Cmilk at every sampled time.
make_events <- function(wt, regimen, obs_times, id_offset = 0L) {
  sch <- schedules[[regimen]]
  ids <- id_offset + seq_along(wt)
  doses <- data.frame(
    id = rep(ids, each = nrow(sch)),
    time = rep(sch$start, times = length(wt)),
    evid = 1L,
    amt = as.vector(outer(sch$rate * (sch$end - sch$start), wt)),
    rate = as.vector(outer(sch$rate, wt)),
    cmt = "central",
    dvid = NA_integer_,
    WT = rep(wt, each = nrow(sch))
  )
  obs <- data.frame(
    id = rep(ids, each = length(obs_times)),
    time = rep(obs_times, times = length(wt)),
    evid = 0L, amt = 0, rate = 0, cmt = "central", dvid = 1L,
    WT = rep(wt, each = length(obs_times))
  )
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(regimen = regimen) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

n_arm <- 200L
# Pooled analysis population (Table 1 'All'): median 82.9 kg, mean 85 kg,
# range 44.9-150 kg; sdlog 0.22 reproduces mean / median = 85 / 82.9.
wt_pool_90 <- draw_weights(n_arm, 82.9, 0.22, 44.9, 150)
wt_pool_60 <- draw_weights(n_arm, 82.9, 0.22, 44.9, 150)
grid <- sort(unique(c(seq(0, 144, by = 0.5), 24, 48, 52)))

events_fig4 <- dplyr::bind_rows(
  make_events(wt_pool_60, "BRX60", grid, id_offset = 0L),
  make_events(wt_pool_90, "BRX90", grid, id_offset = n_arm)
)
stopifnot(!anyDuplicated(events_fig4[events_fig4$evid == 0, c("id", "time")]))
```

# Simulation

Figure 4 of the paper shows smooth individual-prediction bands (1000
simulated subjects per regimen, weights resampled from the analysis
population). The width of its 95% interval at the plateau matches the CL
variability alone, so the replication below simulates between-subject
variability without residual error: `Cc` and `Cmilk` in the rxode2 output are
individual predictions.

```{r sim-fig4}
sim_fig4 <- rxode2::rxSolve(
  mod, events = events_fig4, keep = c("regimen", "WT"),
  returnType = "data.frame", useLinCmt = FALSE
)

band_fig4 <- sim_fig4 |>
  dplyr::select(id, time, regimen, Plasma = Cc, Milk = Cmilk) |>
  tidyr::pivot_longer(c(Plasma, Milk), names_to = "matrix", values_to = "conc") |>
  dplyr::group_by(regimen, matrix, time) |>
  dplyr::summarise(
    lo = quantile(conc, 0.025), med = median(conc), hi = quantile(conc, 0.975),
    .groups = "drop"
  ) |>
  dplyr::mutate(matrix = factor(matrix, levels = c("Plasma", "Milk")))
```

# Replicate published figures

```{r fig4, fig.cap = "Replicates Figure 4a-d of Wald 2022: median and 95% prediction interval of allopregnanolone in plasma and breast milk for the BRX60 and BRX90 regimens (200 simulated subjects per regimen)."}
ggplot(band_fig4, aes(time, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi, fill = regimen), alpha = 0.3) +
  geom_line(aes(colour = regimen)) +
  facet_grid(matrix ~ regimen, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 144, 24)) +
  labs(x = "Time after start of infusion (h)",
       y = "Allopregnanolone (ng/mL)", colour = NULL, fill = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

The paper's Figure 4b plateau median at the end of the 90 ug/kg/h step is about
78 ng/mL (95% interval about 50-117 ng/mL), and Figure 4d shows the milk median
at about 107 ng/mL. Figure 4a shows the BRX60 plateau median at about 54
ng/mL. These values were read from the published figure, so the check allows a
10% band on the medians.

```{r fig4-check}
plateau <- band_fig4 |>
  dplyr::filter((regimen == "BRX90" & time == 52) | (regimen == "BRX60" & time == 56)) |>
  dplyr::select(regimen, matrix, lo, med, hi)
plateau |>
  dplyr::rename("Regimen" = regimen, "Matrix" = matrix, "2.5th pct" = lo,
                "Median" = med, "97.5th pct" = hi) |>
  knitr::kable(digits = 1, caption = "Simulated plateau concentrations (ng/mL) at 52 h (BRX90) and 56 h (BRX60).")

med_of <- function(reg, mat) plateau$med[plateau$regimen == reg & plateau$matrix == mat]
stopifnot(
  abs(med_of("BRX90", "Plasma") / 78 - 1) < 0.10,
  abs(med_of("BRX90", "Milk") / 107 - 1) < 0.10,
  abs(med_of("BRX60", "Plasma") / 54 - 1) < 0.10
)
```

The paper states that the model predicts milk allopregnanolone below 10 ng/mL
36 h after the end of the infusion (96 h) in 95% of subjects (section 3.5).

```{r milk-96}
milk96_p95 <- sim_fig4 |>
  dplyr::filter(time == 96) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(p95_milk_ng_mL = quantile(Cmilk, 0.95), .groups = "drop")
milk96_p95 |>
  dplyr::rename("Regimen" = regimen, "95th pct milk conc (ng/mL)" = p95_milk_ng_mL) |>
  knitr::kable(digits = 2,
               caption = "95th percentile of simulated milk concentration 36 h after the end of infusion.")
# A 95th percentile of 200 subjects carries Monte Carlo noise of roughly 10%,
# so the check allows that margin above the paper's 10 ng/mL statement.
stopifnot(milk96_p95$p95_milk_ng_mL[milk96_p95$regimen == "BRX90"] < 12)
```

# Relative infant dose (Table 4)

Table 4 of the paper derives the relative infant dose (RID) for the 24-48 h
window of the BRX90 regimen (the day at the 90 ug/kg/h rate, a maternal dose of
2160 ug/kg/day): average milk concentration = milk AUC24-48 / 24 h, infant
daily dose = average milk concentration x 150 mL/kg/day, and RID = infant dose
/ maternal dose.

```{r rid}
conc_rid <- sim_fig4 |>
  dplyr::filter(regimen == "BRX90", !is.na(Cc)) |>
  dplyr::select(id, time, regimen, Plasma = Cc, Milk = Cmilk) |>
  tidyr::pivot_longer(c(Plasma, Milk), names_to = "matrix", values_to = "Cc")
dose_rid <- events_fig4 |>
  dplyr::filter(evid == 1, regimen == "BRX90", time == 0) |>
  dplyr::select(id, time, regimen, amt) |>
  tidyr::crossing(matrix = c("Plasma", "Milk"))

nca_rid <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_rid, Cc ~ time | regimen + matrix + id),
  PKNCA::PKNCAdose(dose_rid, amt ~ time | regimen + matrix + id, route = "intravascular"),
  intervals = data.frame(start = 24, end = 48, auclast = TRUE)
))

rid <- as.data.frame(nca_rid) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, matrix, PPORRES) |>
  tidyr::pivot_wider(names_from = matrix, values_from = PPORRES) |>
  dplyr::mutate(
    milk_avg = Milk / 24,
    infant_ug_kg_day = milk_avg * 150 / 1000,
    rid_pct = 100 * infant_ug_kg_day / 2160
  )

rid_summary <- data.frame(
  statistic = c("Minimum", "Median", "Maximum"),
  plasma_auc = c(min(rid$Plasma), median(rid$Plasma), max(rid$Plasma)),
  milk_auc = c(min(rid$Milk), median(rid$Milk), max(rid$Milk)),
  milk_avg = c(min(rid$milk_avg), median(rid$milk_avg), max(rid$milk_avg)),
  rid_pct = c(min(rid$rid_pct), median(rid$rid_pct), max(rid$rid_pct)),
  pub_plasma_auc = c(462, 1760, 3410),
  pub_milk_auc = c(628, 2394, 4638),
  pub_rid_pct = c(0.1818, 0.6926, 1.342)
)
rid_summary |>
  dplyr::rename(
    "Statistic" = statistic,
    "Plasma AUC24-48 (ng*h/mL)" = plasma_auc,
    "Milk AUC24-48 (ng*h/mL)" = milk_auc,
    "Average milk conc (ng/mL)" = milk_avg,
    "RID (%)" = rid_pct,
    "Published plasma AUC24-48" = pub_plasma_auc,
    "Published milk AUC24-48" = pub_milk_auc,
    "Published RID (%)" = pub_rid_pct
  ) |>
  knitr::kable(digits = c(0, 0, 0, 1, 3, 0, 0, 3),
               caption = "Relative infant dose, BRX90, 24-48 h (simulated vs Table 4 of Wald 2022).")

# The minimum and maximum depend on which subjects land in the tails, so
# only the medians are checked. Milk/plasma AUC is Kp = 1.36 by construction.
stopifnot(
  abs(median(rid$Plasma) / 1760 - 1) < 0.15,
  abs(median(rid$rid_pct) / 0.6926 - 1) < 0.15,
  abs(median(rid$Milk / rid$Plasma) - 1.36) < 1e-6
)
```

# PKNCA validation against Table 3

Table 3 is a non-compartmental analysis of the observed plasma data of the 11
healthy lactating volunteers in 547-CLP-108 (BRX90). The replication samples a
virtual 547-CLP-108 cohort (median weight 79.6 kg, range 60.4-128 kg; Table 1)
at the study's nominal plasma times (pre-infusion, 12, 24, 36, 48, 56, 60, 61,
62, 64 and 72 h; main text 2.2) and adds the proportional residual error,
because Table 3 summarises observed concentrations: the observed Cmax is a
maximum over noisy samples.

```{r nca-sim}
wt_108 <- draw_weights(n_arm, 79.6, 0.22, 60.4, 128)
t_108 <- c(0, 12, 24, 36, 48, 56, 60, 61, 62, 64, 72)
events_108 <- make_events(wt_108, "BRX90", t_108, id_offset = 1000L) |>
  dplyr::mutate(treatment = "BRX90 (547-CLP-108)")

sim_108 <- rxode2::rxSolve(
  mod, events = events_108, keep = c("treatment", "WT"),
  returnType = "data.frame", useLinCmt = FALSE
)

# `sim` is the plasma observation with residual error; floor at 0 so that a
# rare negative draw cannot enter the log-linear terminal-phase fit.
conc_108 <- sim_108 |>
  dplyr::mutate(Cc = pmax(sim, 0)) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, treatment, Cc)
# Pre-infusion record: allopregnanolone below the assay LLOQ (main text 3.2).
conc_108 <- conc_108 |>
  dplyr::mutate(Cc = dplyr::if_else(time == 0, 0, Cc))

# One dose record per subject carrying the total 60-h amount, so PKNCA's
# clearance is total dose / AUC0-inf as in Table 3. The dose is intravenous
# (F = 1), so the "CL/F" label PKNCA attaches below is the systemic CL.
dose_108 <- events_108 |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(amt = sum(amt) / 1000, time = 0, .groups = "drop")  # ug -> mg

intervals_108 <- data.frame(
  start = c(0, 0, 24),
  end = c(Inf, 60, 56),
  cmax = c(TRUE, FALSE, FALSE),
  tmax = c(TRUE, FALSE, FALSE),
  aucinf.obs = c(TRUE, FALSE, FALSE),
  half.life = c(TRUE, FALSE, FALSE),
  cl.obs = c(TRUE, FALSE, FALSE),
  auclast = c(FALSE, TRUE, TRUE),
  cav = c(FALSE, FALSE, TRUE)
)

nca_108 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_108, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_108, amt ~ time | treatment + id,
                   route = "intravascular", duration = 60),
  intervals = intervals_108
))
```

```{r nca-compare}
nca_long <- as.data.frame(nca_108) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life",
                                "cl.obs", "auclast", "cav")) |>
  dplyr::mutate(
    interval = dplyr::case_when(
      start == 0 & end == 60 ~ "0-60 h",
      start == 24 & end == 56 ~ "24-56 h",
      TRUE ~ "0-inf"
    ),
    # PKNCA clearance is mg / (ng*h/mL) = 1000 L/h; convert to L/h.
    PPORRES = dplyr::if_else(PPTESTCD == "cl.obs", PPORRES * 1000, PPORRES)
  )

# Table 3 reports geometric means (tmax: the paper's central value), so the
# simulated values are summarised the same way before comparison.
sim_summary <- nca_long |>
  dplyr::filter(!is.na(PPORRES), PPORRES > 0) |>
  dplyr::group_by(treatment, interval, PPTESTCD) |>
  dplyr::summarise(
    PPORRES = if (dplyr::first(PPTESTCD) == "tmax") median(PPORRES) else exp(mean(log(PPORRES))),
    .groups = "drop"
  )

published_108 <- data.frame(
  treatment = "BRX90 (547-CLP-108)",
  interval = c("0-inf", "0-inf", "0-inf", "0-inf", "0-inf",
               "0-60 h", "24-56 h", "24-56 h"),
  PPTESTCD = c("cmax", "tmax", "aucinf.obs", "half.life", "cl.obs",
               "auclast", "auclast", "cav"),
  PPORRES = c(89.7, 47.8, 3736.5, 11.3, 87.5, 3358.4, 2257.9, 70.6)
)

cmp_108 <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary,
  reference = published_108,
  by = c("treatment", "interval"),
  units = c(cmax = "ng/mL", tmax = "h", aucinf.obs = "ng*h/mL",
            half.life = "h", cl.obs = "L/h", auclast = "ng*h/mL", cav = "ng/mL"),
  tolerance_pct = 20
)
cmp_108 |>
  dplyr::rename("Treatment" = treatment, "Interval" = interval) |>
  knitr::kable(caption = "Plasma NCA: simulated 547-CLP-108 cohort vs Table 3 of Wald 2022 (geometric means; tmax median).")
if (!is.null(attr(cmp_108, "footnote"))) cat(attr(cmp_108, "footnote"), "\n")

pct_diff <- function(code, iv = "0-inf") {
  s <- sim_summary$PPORRES[sim_summary$PPTESTCD == code & sim_summary$interval == iv]
  r <- published_108$PPORRES[published_108$PPTESTCD == code & published_108$interval == iv]
  # A missing match would make the comparison below vacuously TRUE.
  stopifnot(length(s) == 1, length(r) == 1)
  100 * (s / r - 1)
}
# Geometric means of a 200-subject cohort; a mis-transcribed CL, volume or
# dose unit moves these by tens of percent.
stopifnot(
  abs(pct_diff("cmax")) < 15,
  abs(pct_diff("cav", "24-56 h")) < 15,
  abs(pct_diff("auclast", "0-60 h")) < 15,
  abs(pct_diff("auclast", "24-56 h")) < 15,
  abs(pct_diff("half.life")) < 20,
  abs(pct_diff("aucinf.obs")) < 15,
  abs(pct_diff("cl.obs")) < 15
)
```

# Assumptions and deviations

* **Residual error scale.** ESM Table S1 prints the proportional residual
  variability as 0.272 without saying whether it is a standard deviation or a
  variance. The model uses it as an SD (27.2%). A simulation of the
  547-CLP-108 design supports this reading. With SD 0.272 the geometric mean
  Cmax is about 88 ng/mL against the published 89.7, and the Cmax/Cavg ratio
  is about 1.30 against 1.27 (89.7/70.6). Reading 0.272 as a variance (SD
  0.52) gives a Cmax of about 103 ng/mL and a ratio of about 1.55. The PKNCA
  comparison above repeats this check.
* **Milk residual error.** Plasma and milk were fitted simultaneously, but
  Table S1 lists a single proportional residual variability. The same value is
  used for both matrices (`propSd` and `propSd_Cmilk`).
* **Partition coefficient.** Kp (1.36) was estimated from the 547-CLP-108
  plasma and milk data and then held fixed when the plasma model was updated
  with all five studies (Table S1 footnote). In the model it is an ordinary
  `ini()` value with no between-subject variability, as reported. Table S1
  prints its "Relative SE (%)" as 0.235; with the 95% CI 0.858-2.16 this is the
  SE of log(Kp), not a percentage.
* **IIV.** Only CL and V1 carry between-subject variability (Table S1), as an
  estimated covariance block (correlation 0.47).
* **BRX60 schedule.** The main text gives only the BRX90 schedule in full. For
  BRX60 the model uses 30 ug/kg/h for 0-4 h, 60 ug/kg/h for 4-56 h and
  30 ug/kg/h for 56-60 h, which matches the steps visible in Figure 4a.
* **Table 4 units.** The column "Daily dose in milk, ug/kg/day" holds values
  in ng/kg/day (median 14,960 = 99.73 ng/mL x 150 mL/kg/day). Only with
  ng/kg/day does dividing by the maternal 2160 ug/kg/day give the tabulated
  RID of 0.69%.
* **Weights.** Virtual cohorts are log-normal, truncated to the observed
  ranges in Table 1: the pooled population for Figure 4 and Table 4, and the
  547-CLP-108 arm for Table 3.
* **Endogenous allopregnanolone.** The model describes exogenous brexanolone
  only; there is no endogenous baseline. The paper notes that late milk
  samples are under-predicted because of undetected endogenous
  allopregnanolone (section 3.5).
* **Errata.** No erratum to the article was found (checked 2026-10-08).
