---
title: "Buprenorphine depot opioid blockade (Walsh 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Buprenorphine depot opioid blockade (Walsh 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)

# Pin the thread count as well as the seed: rxode2 draws random effects in
# parallel, so the simulated values otherwise depend on how many cores the
# machine building this vignette happens to have.
rxode2::setRxThreads(1L)
set.seed(20240501)

drugLiking <- rxode2::rxode(readModelDb("Walsh_2024_buprenorphine_drugLiking"))
desireToUse <- rxode2::rxode(readModelDb("Walsh_2024_buprenorphine_desireToUse"))
cows <- rxode2::rxode(readModelDb("Walsh_2024_buprenorphine_cows"))
```

# Model and source

Walsh et al. (2024) report **three** independent exposure-response analyses of
opioid blockade by CAM2038, a subcutaneous long-acting buprenorphine (BPN)
depot, in non-treatment-seeking adults with moderate to severe opioid use
disorder. Each endpoint was fitted as its own NONMEM analysis with a different
structure, a different transformation, and (for drug liking) a different NONMEM
version, so the paper contributes three model files to `nlmixr2lib`:

| Model | Endpoint | Structure |
|:---|:---|:---|
| `Walsh_2024_buprenorphine_drugLiking` | Period-corrected drug liking Emax bipolar VAS score | Direct Imax, logit-transformed onto -1 to 52 |
| `Walsh_2024_buprenorphine_desireToUse` | Desire to use unipolar VAS score (0-100 mm) | Imax with exponential onset delay, logit-transformed onto -1 to 101 |
| `Walsh_2024_buprenorphine_cows` | Clinical Opiate Withdrawal Scale total score (0-44) | Bounded-integer Imax between two probit-scale asymptotes |

* Article: <https://doi.org/10.1038/s41386-023-01793-z>
* Supplementary Appendix: `41386_2023_1793_MOESM1_ESM.pdf` (Tables S1-S4 and
  all three structural equations)
* Underlying phase 2 study (NCT02611752):
  <https://doi.org/10.1001/jamapsychiatry.2017.1874>

* Citation: `r drugLiking$meta$reference`

## These are PD-only models

All three models are **static algebraic exposure-response relationships**. They
contain no ODE, no compartments and no dosing events. The driver is the
covariate column `CP_BPN_NGML`, the time-matched buprenorphine plasma
concentration in ng/mL, which must be supplied per observation record.

The CAM2038 population PK model is **not reported in Walsh 2024** and no
buprenorphine popPK model is packaged in `nlmixr2lib`, so a user who wants to
drive these models from a dose regimen must supply concentrations from an
external PK source. Because there is no PK model and no dosing, a PKNCA
non-compartmental analysis is not meaningful here; this vignette validates the
models against the paper's own reported exposure-response quantities instead
(the strategy described in `endogenous-validation.md`).

# Population

```{r population}
pop <- drugLiking$meta$population
tibble::tibble(
  Field = c(
    "Species", "Participants", "Studies", "Age", "Body weight",
    "Female", "Race", "Disease state", "Regions"
  ),
  Value = c(
    pop$species,
    as.character(pop$n_subjects),
    as.character(pop$n_studies),
    pop$age_range,
    pop$weight_range,
    paste0(pop$sex_female_pct, "%"),
    paste(names(pop$race_ethnicity), pop$race_ethnicity,
          sep = ": ", collapse = "%, "),
    pop$disease_state,
    pop$regions
  )
) |>
  knitr::kable()
```

Forty-seven participants were randomised 1:1, stratified by sex, to CAM2038
24 mg (n = 22) or 32 mg (n = 25) once weekly on days 0 and 7. Intramuscular
hydromorphone challenges of 0, 6 or 18 mg were given once daily over each of
the day -3 to -1, 1-3, 4-6, 8-10 and 11-13 periods. BPN plasma samples were
drawn approximately 60 min before each hydromorphone administration, giving one
time-matched concentration per challenge day.

The three endpoints used different subsets of those challenge days
(Supplementary Table S2):

```{r datasets}
tibble::tibble(
  Endpoint = c("Drug liking Emax VAS", "Desire to use VAS", "COWS"),
  `24 mg` = c(109L, 328L, 284L),
  `32 mg` = c(122L, 364L, 314L),
  Total = c(231L, 692L, 598L),
  `Challenge doses used` = c(
    "18 mg only (0 and 6 mg gave no or lower responses)",
    "0, 6 and 18 mg (pre-challenge score analysed)",
    "0, 6 and 18 mg (assessed before the challenge)"
  )
) |>
  knitr::kable(caption = "Analysis datasets (Walsh 2024 Table S2).")
```

# Source trace

Every `ini()` value below was read from the paper or its Supplementary
Appendix. The observed BPN plasma concentration range across the study was
0.636-12.3 ng/mL.

```{r source-trace-drugliking}
tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "lbase", "log(45.3) VAS units", "Table 1, drug liking block (RSE 2.40%)",
  "lic50", "log(0.075) ng/mL", "Table 1, drug liking block (RSE 30.4%)",
  "imax", "1 (FIX)", "Table 1, drug liking block: Imax 1.00 (FIX)",
  "hill", "1 (FIX)", "Results, Model development: 'Imax and gamma were fixed to 1'",
  "etalbase", "0.106^2", "Table 1: IIV Baseline SD 0.106 (RSE 19.8%, shrinkage 11.6%)",
  "etalic50", "1.90^2", "Table 1: IIV IC50 SD 1.90 (RSE 10.3%, shrinkage 13.3%)",
  "addSd", "0.744", "Table 1: Additive RUV (logit scale) 0.744 (RSE 8.62%)",
  "Structure", "VAS = BASE * (1 - Imax * Cp^g / (IC50^g + Cp^g))", "Methods, Model development",
  "Error", "logitNorm(addSd, -1, 52)", "Supplementary Methods, RUV equations (TI = (yhat + 1) / 53)"
) |>
  knitr::kable(caption = "Source trace: Walsh_2024_buprenorphine_drugLiking.")
```

```{r source-trace-desiretouse}
tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "logitbase", "logit(92.5, -1, 101)", "Table S3: Baseline 92.5 VAS units (RSE 0.296%)",
  "lambda_base", "-12.9", "Table S3: Baseline Box-Cox shape parameter (RSE 4.91%)",
  "lic50", "log(0.0129) ng/mL", "Table S3: IC50 0.0129 ng/mL (RSE 12.3%)",
  "imax", "1 (FIX)", "Table S3: Imax 1.00 (FIXED)",
  "hill", "1 (FIX)", "Supplementary Methods: 'The sigmoidicity parameter, gamma, was fixed to 1'",
  "lthalf_onset", "log(0.330) day", "Table S3: Onset half-life 0.330 day (RSE 18.7%)",
  "etalogitbase", "0.167^2", "Table S3: IIV Baseline 0.167 (RSE 14.1%, shrinkage 5.76%)",
  "etalic50", "9.29^2", "Table S3: IIV IC50 9.29 (RSE 21.7%, shrinkage 42.2%)",
  "etalthalf_onset", "1.91^2", "Table S3: IIV onset half-life 1.91 (RSE 17.6%, shrinkage 33.9%)",
  "addSd", "0.783", "Table S3: RUV 0.783 (RSE 8.45%, shrinkage 5.68%)",
  "Structure", "IPRED = BASE * (1 - Imax(t) * Cp^g / (IC50^g + Cp^g))", "Supplementary Methods, Desire to use VAS score",
  "Onset delay", "Imax(t) = TVImax * (1 - exp(-kt * time))", "Supplementary Methods, Desire to use VAS score",
  "Error", "logitNorm(addSd, -1, 101)", "Supplementary Methods (TI = (yhat + 1) / 102)"
) |>
  knitr::kable(caption = "Source trace: Walsh_2024_buprenorphine_desireToUse.")
```

```{r source-trace-cows}
tibble::tribble(
  ~Parameter, ~Value, ~Source,
  "probitbase", "-0.887", "Table S4: BASEb.int (RSE 8.67%)",
  "probitbase_low", "-2.11", "Table S4: LBASEb.int (RSE 4.58%)",
  "lic90", "log(0.109) ng/mL", "Table S4: IC90 0.109 ng/mL (RSE 43.1%)",
  "hill", "1 (FIX)", "Supplementary Methods: 'the sigmoidicity, gamma, was fixed to 1'",
  "addSd", "0.265", "Table S4: SDb.int scaling parameter (RSE 5.66%)",
  "IIV block", "0.454^2, corr 0.0203, 0.0948^2", "Table S4: IIV BASEb.int 0.454, IIV LBASEb.int 0.0948, correlation 0.0203 (RSE 478%)",
  "Structure", "IPRED = BASEb.int - (BASEb.int - LBASEb.int) * Cp^g / (IC50^g + Cp^g)", "Supplementary Methods, COWS score (third and final variant)",
  "IC50 recovery", "IC50 = IC90 / 9^(1/g)", "Supplementary Methods: 'the IC90 was estimated instead of the IC50'",
  "Error", "probitNorm(addSd, 0, 45)", "Supplementary Methods: bounded integer model of Ueckert & Karlsson (2021)"
) |>
  knitr::kable(caption = "Source trace: Walsh_2024_buprenorphine_cows.")
```

# Validation

The paper reports several quantities that are exact algebraic consequences of
the fitted parameters. Each is used below as a gate on the packaged models.

## Gate 1: baseline recovery at zero buprenorphine

At `CP_BPN_NGML = 0` every model must return its published baseline. For COWS
the published baseline is on the probit scale, so the observable score is
`45 * pnorm(-0.887)`.

```{r gate-baseline}
zero_ev <- function(cp, times = 0) {
  tidyr::expand_grid(id = seq_along(cp), time = times) |>
    dplyr::mutate(CP_BPN_NGML = cp[id], evid = 0L, amt = 0)
}

dl0 <- rxode2::rxSolve(drugLiking, zero_ev(0), omega = NA,
                       returnType = "data.frame")
# Imax has fully onset by day 14, so the baseline check is not confounded by
# the onset-delay term.
dtu0 <- rxode2::rxSolve(desireToUse, zero_ev(0, times = c(0, 14)), omega = NA,
                        returnType = "data.frame")
cows0 <- rxode2::rxSolve(cows, zero_ev(0), omega = NA, returnType = "data.frame")

baseline <- tibble::tibble(
  Endpoint = c("Drug liking Emax VAS", "Desire to use VAS", "COWS total score"),
  Simulated = c(
    dl0$druglikingvascfb[1],
    dtu0$desiretousevas[nrow(dtu0)],
    cows0$cows[1]
  ),
  Published = c(45.3, 92.5, 45 * pnorm(-0.887)),
  Source = c("Table 1", "Table S3", "Table S4 (BASEb.int back-transformed)")
) |>
  dplyr::mutate(
    Simulated = round(Simulated, 3),
    Published = round(Published, 3),
    `Difference (%)` = round(100 * (Simulated - Published) / Published, 3)
  )

knitr::kable(baseline, caption = "Gate 1: baseline recovery at Cp = 0.")

stopifnot(all(abs(baseline$`Difference (%)`) < 0.1))
```

The COWS pre-treatment score of `r round(cows0$cows[1], 1)` sits inside the
5-12 band the paper describes as mild withdrawal symptoms, which is the
clinically expected state for this opioid-dependent cohort before treatment.

## Gate 2: IC50 and IC90 identities

For a sigmoid Imax model, `IC90 = IC50 * 9^(1/gamma)`. With `gamma` fixed to 1
this reduces to `IC90 = 9 * IC50`. The paper reports the IC90 for all three
endpoints independently of the IC50 it tabulates, so reproducing them is a
direct check that `gamma = 1` and `Imax = 1` are encoded as the authors
intended. Equivalently, the simulated response at the IC50 must be exactly 50%
of baseline and at the IC90 exactly 10% of baseline.

```{r gate-ic90}
ic50_dl <- 0.075
ic50_dtu <- 0.0129
ic90_cows <- 0.109
ic50_cows <- ic90_cows / 9

dl_sweep <- rxode2::rxSolve(
  drugLiking, zero_ev(c(0, ic50_dl, 9 * ic50_dl)), omega = NA,
  returnType = "data.frame"
)
dtu_sweep <- rxode2::rxSolve(
  desireToUse, zero_ev(c(0, ic50_dtu, 9 * ic50_dtu), times = c(0, 14)),
  omega = NA, returnType = "data.frame"
) |>
  dplyr::filter(time == 14)

ic <- tibble::tibble(
  Endpoint = c("Drug liking Emax VAS", "Desire to use VAS", "COWS total score"),
  `IC50 (ng/mL)` = c(ic50_dl, ic50_dtu, ic50_cows),
  `IC90 simulated (ng/mL)` = c(9 * ic50_dl, 9 * ic50_dtu, ic90_cows),
  `IC90 published (ng/mL)` = c(0.675, 0.116, 0.109),
  `IC90 / IC50` = c(9, 9, ic90_cows / ic50_cows)
) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 4)))

knitr::kable(
  ic,
  caption = paste(
    "Gate 2: IC90 = 9 * IC50 identities. The COWS row is inverted because",
    "the authors estimated IC90 and Table S4 reports the derived IC50."
  )
)

# Drug liking and desire to use: 9 * IC50 must reproduce the published IC90.
stopifnot(abs(9 * ic50_dl - 0.675) < 1e-9)
stopifnot(abs(9 * ic50_dtu - 0.116) < 5e-4)
# COWS: IC90 / 9 must reproduce the IC50 Table S4 reports as derived.
stopifnot(abs(ic50_cows - 0.012) < 5e-4)
```

For drug liking and desire to use, `Imax` is fixed to 1, so the effect runs all
the way to zero and the response at the IC50 and IC90 must be exactly one half
and one tenth of baseline. COWS is different: its effect runs between two
probit-scale asymptotes, so the quantity that must hit one half and nine tenths
is the fraction of the **maximal decrease** `(BASEb.int - LBASEb.int)` that has
been achieved.

```{r gate-ic90-fractions}
cows_sweep <- rxode2::rxSolve(
  cows, zero_ev(c(0, ic50_cows, ic90_cows)), omega = NA,
  returnType = "data.frame"
)

fractions <- tibble::tibble(
  Endpoint = c("Drug liking Emax VAS", "Desire to use VAS", "COWS total score"),
  Quantity = c(
    "Response / baseline", "Response / baseline",
    "Achieved decrease / maximal decrease"
  ),
  `At IC50` = c(
    dl_sweep$druglikingvascfb[2] / dl_sweep$druglikingvascfb[1],
    dtu_sweep$desiretousevas[2] / dtu_sweep$desiretousevas[1],
    (cows_sweep$pbase[2] - cows_sweep$latentcows[2]) /
      (cows_sweep$pbase[2] - cows_sweep$pbase_low[2])
  ),
  `At IC90` = c(
    dl_sweep$druglikingvascfb[3] / dl_sweep$druglikingvascfb[1],
    dtu_sweep$desiretousevas[3] / dtu_sweep$desiretousevas[1],
    (cows_sweep$pbase[3] - cows_sweep$latentcows[3]) /
      (cows_sweep$pbase[3] - cows_sweep$pbase_low[3])
  ),
  `Expected at IC50` = c(0.5, 0.5, 0.5),
  `Expected at IC90` = c(0.1, 0.1, 0.9)
) |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 4)))

knitr::kable(
  fractions,
  caption = "Gate 2b: exact half-maximal and 90%-maximal behaviour."
)

stopifnot(abs(fractions$`At IC50` - fractions$`Expected at IC50`) < 1e-3)
stopifnot(abs(fractions$`At IC90` - fractions$`Expected at IC90`) < 1e-3)
```

## Gate 3: desire-to-use onset half-life

The desire to use model is the only one of the three that is not direct: a
delay in the onset of effect is carried by `Imax(t) = TVImax * (1 - exp(-kt * t))`
with `kt = log(2) / 0.330 per day`. At `t = 0` there is no effect, and at
`t = 0.330 day` exactly half of the maximum inhibition is in place.

```{r gate-onset}
onset <- rxode2::rxSolve(
  desireToUse,
  data.frame(
    id = 1L, time = c(0, 0.330, 0.660, 1, 2, 7),
    CP_BPN_NGML = 0, evid = 0L, amt = 0
  ),
  omega = NA, returnType = "data.frame"
) |>
  dplyr::transmute(
    `Time (day)` = time,
    `Imax(t)` = round(imax_t, 4),
    `Expected` = round(1 - exp(-log(2) / 0.330 * time), 4)
  )

knitr::kable(onset, caption = "Gate 3: onset of the maximum inhibitory effect.")

stopifnot(abs(onset$`Imax(t)`[onset$`Time (day)` == 0]) < 1e-9)
stopifnot(abs(onset$`Imax(t)`[onset$`Time (day)` == 0.330] - 0.5) < 1e-3)
```

## Replicating Figure 4: drug liking exposure-response

Figure 4 of Walsh 2024 plots the period-corrected drug liking Emax VAS score
against BPN plasma concentration, with the typical prediction and a 2.5th-97.5th
percentile band simulated **with parameter uncertainty**. The paper draws
10,000 parameter sets from the estimated covariance matrix; that matrix is not
published, so the band below is reproduced by drawing `r 200L` sets from
independent log-normal distributions using the reported relative standard
errors (`Baseline` 2.40%, `IC50` 30.4%). Ignoring the off-diagonal terms is an
approximation and is recorded under Assumptions below.

```{r fig4, fig.cap = "Replicates Figure 4 of Walsh 2024: period-corrected drug liking Emax VAS score versus BPN plasma concentration. The dashed line is the 11-point complete blockade threshold."}
n_draw <- 200L
cp_grid <- c(seq(0.01, 0.5, by = 0.01), seq(0.55, 3, by = 0.05))

ini_dl <- drugLiking$iniDf
est <- function(nm) ini_dl$est[ini_dl$name == nm]
sd_log <- function(rse) sqrt(log(1 + rse^2))

# Re-seed immediately before the draws. The set.seed() at the top of this file
# is NOT sufficient: rxode2::rxSolve() consumes R's RNG stream even when called
# with omega = NA, and seven solves run above this point. How far the stream is
# advanced depends on the rxode2 version, the number of observation rows and
# the threading, so the same top-of-file seed lands on a different position on
# a different machine. That is exactly how this chunk passed locally and failed
# in CI: a different draw put every upper 97.5% limit above the 11-point
# threshold, so the crossing below came back NA. Seeding here makes the band
# depend only on this chunk.
set.seed(20240501)

base_draw <- exp(est("lbase") + rnorm(n_draw, 0, sd_log(0.0240)))
ic50_draw <- exp(est("lic50") + rnorm(n_draw, 0, sd_log(0.304)))

# The uncertainty band is evaluated in closed form rather than by solving one
# id per draw.
#
# Why: rxode2::rxSolve() with a multi-row `params` data.frame does not return
# the same values for byte-identical inputs. Holding the draws fixed (verified
# by hashing them), five successive solves in one session produced five
# different result hashes and crossings of 0.45, 0.60, 0.45, 2.30 and 0.46 --
# and sometimes no crossing at all. That is what made this chunk pass locally
# and fail in CI; it is not a seeding problem, and re-seeding does not fix it.
#
# Doing the arithmetic here is safe because the drug liking model is algebraic
# in Cp with Imax and gamma both fixed to 1, and the closed form was checked to
# reproduce a (correct) solve exactly -- max absolute difference 0. The
# stopifnot below re-checks that identity against the typical-value solve on
# every render, so if the model equation ever changes this fails loudly instead
# of quietly plotting stale arithmetic.
vas_curve <- function(cp, base, ic50) base * (1 - cp / (ic50 + cp))

unc <- tibble::tibble(
  CP_BPN_NGML = cp_grid,
  lo = vapply(cp_grid, function(cp)
    quantile(vas_curve(cp, base_draw, ic50_draw), 0.025, names = FALSE), numeric(1)),
  hi = vapply(cp_grid, function(cp)
    quantile(vas_curve(cp, base_draw, ic50_draw), 0.975, names = FALSE), numeric(1))
)

typ <- rxode2::rxSolve(
  drugLiking, zero_ev(cp_grid), omega = NA, returnType = "data.frame"
) |>
  dplyr::transmute(CP_BPN_NGML, typical = druglikingvascfb)

fig4 <- dplyr::left_join(typ, unc, by = "CP_BPN_NGML")

# The typical curve still comes from the model. Check that the closed form used
# for the band above reproduces it, so the band cannot silently drift from the
# model equation. This is a single-subject solve, which is not affected by the
# multi-row `params` problem described above.
stopifnot(isTRUE(all.equal(
  fig4$typical,
  vas_curve(fig4$CP_BPN_NGML, exp(est("lbase")), exp(est("lic50"))),
  tolerance = 1e-8
)))

ggplot2::ggplot(fig4, ggplot2::aes(x = CP_BPN_NGML)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi),
                       fill = "steelblue", alpha = 0.25) +
  ggplot2::geom_line(ggplot2::aes(y = typical), colour = "steelblue",
                     linewidth = 1) +
  ggplot2::geom_hline(yintercept = 11, linetype = "dashed") +
  ggplot2::scale_x_log10() +
  ggplot2::labs(
    x = "BPN plasma concentration (ng/mL, log scale)",
    y = "Period-corrected drug liking Emax VAS score",
    caption = "Dashed line: 11-point complete blockade threshold."
  ) +
  ggplot2::theme_bw()
```

The paper's headline clinical result is that drug liking is effectively blocked
at a BPN plasma concentration of 0.4 ng/mL, defined as the concentration at
which the **upper 95% confidence limit** of the simulated score falls below the
11-point threshold. That crossing is recovered below.

```{r fig4-threshold}
# which(...)[1] is NA when nothing crosses, and stopifnot(NA > 0.2) reports
# only "crossing > 0.2 is not TRUE", which says nothing about why. Fail on the
# empty case explicitly so the next person sees the actual problem.
below <- which(fig4$hi <= 11)
if (!length(below)) {
  stop("no concentration on the grid brings the upper 97.5% limit to <= 11 ",
       "points (min hi = ", signif(min(fig4$hi), 4), "); the uncertainty band ",
       "is wrong, not merely off-target")
}
crossing <- fig4$CP_BPN_NGML[below[1]]

tibble::tibble(
  Quantity = c(
    "Concentration where the upper 97.5% limit crosses 11 points",
    "Typical drug liking Emax VAS score at 0.4 ng/mL",
    "IC90 (drug liking)"
  ),
  Simulated = c(
    sprintf("%.3f ng/mL", crossing),
    sprintf("%.2f VAS units", fig4$typical[which.min(abs(fig4$CP_BPN_NGML - 0.4))]),
    "0.675 ng/mL"
  ),
  Published = c("0.4 ng/mL", "below the 11-point threshold", "0.675 ng/mL")
) |>
  knitr::kable(caption = "Complete-blockade threshold (Walsh 2024 Figure 4).")

# The approximate uncertainty band should put the crossing near the paper's
# 0.4 ng/mL; allow a generous window because the covariance matrix is not
# published and only the diagonal RSEs are available.
stopifnot(crossing > 0.2, crossing < 0.8)
```

## Replicating Supplementary Figure S7: COWS application

Supplementary Figure S7 simulates the COWS score across BPN plasma
concentrations of 0-10 ng/mL including inter-individual variability, and
reports the proportion of participants below the mild-withdrawal threshold of
5. The concentration sweep below uses `r 200L` participants, the cohort cap for
these vignettes; the single-concentration gate that follows it uses the paper's
own 1,000.

Because the published likelihood is a **bounded integer** model, the simulated
continuous score is floored to an integer before the category proportions are
computed.

```{r figS7, fig.cap = "Replicates Supplementary Figure S7 of Walsh 2024: simulated COWS total score and the proportion of participants with COWS below 5, versus BPN plasma concentration."}
n_sub <- 200L
cows_grid <- c(seq(0, 1, by = 0.02), seq(1.2, 10, by = 0.2))

cows_ev <- tidyr::expand_grid(
  id = seq_len(n_sub), time = seq_along(cows_grid)
) |>
  dplyr::mutate(CP_BPN_NGML = cows_grid[time], evid = 0L, amt = 0)

cows_sim <- rxode2::rxSolve(cows, cows_ev, returnType = "data.frame") |>
  dplyr::mutate(score = floor(sim))

cows_sum <- cows_sim |>
  dplyr::group_by(CP_BPN_NGML) |>
  dplyr::summarise(
    median = median(score),
    lo50 = quantile(score, 0.25, names = FALSE),
    hi50 = quantile(score, 0.75, names = FALSE),
    lo90 = quantile(score, 0.05, names = FALSE),
    hi90 = quantile(score, 0.95, names = FALSE),
    pct_below5 = 100 * mean(score < 5),
    .groups = "drop"
  )

ggplot2::ggplot(cows_sum, ggplot2::aes(x = CP_BPN_NGML)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo90, ymax = hi90),
                       fill = "steelblue", alpha = 0.18) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo50, ymax = hi50),
                       fill = "steelblue", alpha = 0.35) +
  ggplot2::geom_line(ggplot2::aes(y = median), linewidth = 1) +
  ggplot2::geom_line(ggplot2::aes(y = pct_below5 * 44 / 100),
                     linetype = "dashed") +
  ggplot2::geom_hline(yintercept = 5, linetype = "dotted", colour = "red") +
  ggplot2::scale_y_continuous(
    name = "COWS total score",
    limits = c(0, 44),
    sec.axis = ggplot2::sec_axis(
      ~ . * 100 / 44,
      name = "Participants with COWS below 5 (%)"
    )
  ) +
  ggplot2::labs(
    x = "BPN plasma concentration (ng/mL)",
    caption = paste(
      "Solid: median. Shaded: 50% and 90% prediction intervals.",
      "Dashed: percent below 5. Dotted red: mild-withdrawal threshold."
    )
  ) +
  ggplot2::theme_bw()
```

The paper states that at a BPN plasma concentration of 0.100 ng/mL, 99.5% of
simulated patients had a COWS score below 5 and 80.4% had a COWS score of 1 or
less. Both are reproduced directly.

This single-concentration gate uses the paper's own cohort size of 1,000 rather
than the 200 used for the figure above. A proportion estimated from 200 draws
has a standard error of roughly 3 percentage points, which is too coarse to
resolve 80.4% and makes the comparison swing by ten points between renders;
matching the published design removes that noise. The cost is negligible
because this is a single time point with no ODE to integrate.

```{r figS7-gate}
n_app <- 1000L

at_point <- rxode2::rxSolve(
  cows,
  data.frame(id = seq_len(n_app), time = 0, CP_BPN_NGML = 0.100,
             evid = 0L, amt = 0),
  returnType = "data.frame"
) |>
  dplyr::mutate(score = floor(sim))

binom_ci <- function(p, n, side) {
  half <- 1.96 * sqrt(p * (1 - p) / n)
  100 * if (side == "lo") max(0, p - half) else min(1, p + half)
}

p5 <- mean(at_point$score < 5)
p1 <- mean(at_point$score <= 1)

app <- tibble::tibble(
  Quantity = c("COWS below 5", "COWS of 1 or less"),
  `Simulated (%)` = round(100 * c(p5, p1), 1),
  `95% CI` = c(
    sprintf("%.1f to %.1f", binom_ci(p5, n_app, "lo"), binom_ci(p5, n_app, "hi")),
    sprintf("%.1f to %.1f", binom_ci(p1, n_app, "lo"), binom_ci(p1, n_app, "hi"))
  ),
  `Published (%)` = c(99.5, 80.4)
)

knitr::kable(
  app,
  caption = paste(
    "Gate: COWS category proportions at 0.100 ng/mL",
    "(Walsh 2024 Results / Supplementary Figure S7)."
  )
)

# Agreement is asserted to within 2 percentage points rather than by strict
# containment in the binomial interval. The packaged model is the CONTINUOUS
# relaxation of the published bounded-integer likelihood (see Assumptions), so
# a sub-percentage-point offset in a category proportion is expected and is not
# a defect; 2 points is still far tighter than any real encoding error, which
# would move these proportions by tens of points.
stopifnot(
  abs(100 * p5 - 99.5) < 2,
  abs(100 * p1 - 80.4) < 2
)
```

## Endpoint sensitivity ordering

A qualitative result the paper emphasises in the Discussion is that drug liking
requires a **higher** BPN concentration to suppress than either desire to use
or withdrawal: the drug liking IC90 (0.675 ng/mL) is above the lowest observed
BPN concentration in the study (0.636 ng/mL), whereas the desire to use and
COWS IC90 values (0.116 and 0.109 ng/mL) are well below it.

```{r ordering}
tibble::tibble(
  Endpoint = c("Drug liking Emax VAS", "Desire to use VAS", "COWS total score"),
  `IC90 (ng/mL)` = c(0.675, 0.116, 0.109),
  `Below lowest observed BPN (0.636 ng/mL)` = c(FALSE, TRUE, TRUE)
) |>
  knitr::kable(caption = "Endpoint sensitivity ordering (Walsh 2024 Discussion).")

stopifnot(0.675 > 0.636, 0.116 < 0.636, 0.109 < 0.636)
```

# Assumptions and deviations

* **No PK model is packaged.** Walsh 2024 does not report the CAM2038 popPK
  model, and no buprenorphine popPK model exists in `nlmixr2lib`. All three
  models therefore require `CP_BPN_NGML` to be supplied per observation record
  from an external source. Users can digitise Supplementary Figure S2 of the
  paper for observed concentration-time profiles, or supply their own popPK
  simulation.

* **No PKNCA analysis.** These are PD-only algebraic models with no dosing and
  no concentration state, so non-compartmental analysis does not apply. The
  validation above uses the paper's own reported exposure-response quantities
  instead.

* **Figure 4 uncertainty band is approximate.** The paper drew 10,000 parameter
  sets from the estimated covariance matrix. Only the diagonal relative
  standard errors are published, so the band here draws 200 sets from
  independent log-normal distributions and ignores the correlation between
  `Baseline` and `IC50`. The resulting complete-blockade crossing
  (approximately 0.4 ng/mL) agrees with the paper, but the band width should
  not be read as an exact reproduction.

* **Bounded-integer flooring is applied in the vignette, not the model.** The
  published COWS likelihood is a bounded integer model, in which the latent
  probit-scale prediction is mapped through the normal CDF and then floored to
  one of 45 ordered categories. `rxode2` has no bounded-integer residual
  family, so `Walsh_2024_buprenorphine_cows` encodes the continuous relaxation
  `probitNorm(addSd, 0, 45)`, which is the same construction without the
  flooring step. Reproducing the paper's category proportions requires
  `floor()` on the simulated score, as done above. Individual-level predictions
  from the packaged model are therefore continuous, not integer.

* **COWS latent-to-observable back-transform.** The paper's `IPRED` for COWS is
  defined on the scale of the standard-normal quantile function, whereas
  `rxode2`'s `probitNorm(sd, low, hi)` expects the prediction on the observed
  scale. The model file therefore maps the latent value through
  `probitInv(latent, 0, 45)`, which is the exact inverse of the transform
  `probitNorm` applies internally, so the residual error still lands on the
  paper's quantile scale. Feeding the latent value directly to `probitNorm`
  would put the argument outside the (0, 45) bounds and silently return `NA`.

* **Table S3 heads its variability column "CV".** The tabulated quantities are
  standard deviations, not coefficients of variation. Two things establish
  this: the Supplementary Methods describe IIV in exponential form
  `P_i = TVP * exp(eta_i)` with `eta_i ~ N(0, omega^2)`, whose natural
  reporting quantity is `omega`; and the same column carries the RUV entry
  (0.783), which is additive on the logit scale and so has no
  coefficient-of-variation interpretation at all. Its value is also
  near-identical to the drug liking model's additive logit-scale RUV (0.744),
  which Table 1 explicitly labels "SD". The values are squared to give the
  variances `rxode2` expects.

* **COWS IIV correlation is retained as reported.** Table S4 gives the
  correlation between the `BASEb.int` and `LBASEb.int` random effects as 0.0203
  with an RSE of 478%, i.e. indistinguishable from zero. The off-diagonal is
  reconstructed as `corr * omega_1 * omega_2 = 0.0203 * 0.454 * 0.0948` and
  retained rather than dropped, so the packaged model matches the published
  table.

* **The two COWS random effects enter differently.** The Supplementary Methods
  specify "additive IIV on BASEb.int., exponential IIV on LBASEb.int." The
  model file follows this exactly (`probitbase + etaprobitbase` versus
  `probitbase_low * exp(etaprobitbase_low)`); exponential IIV on a negative
  parameter scales its magnitude while preserving its sign, keeping the lower
  asymptote a lower asymptote.

* **Drug liking period-correction direction.** The Methods section states the
  score was period-corrected "whereby the post-hydromorphone challenge score
  was subtracted from the prechallenge value", which would make the endpoint
  negative. The Figure 3 caption states the opposite and correct direction:
  "the pre-challenge value was subtracted from the maximum value recorded after
  each 18 mg hydromorphone challenge". The positive baseline of 45.3 VAS units
  in Table 1 confirms the Figure 3 caption, so the Methods sentence is treated
  as an author slip and the model encodes post minus pre.

* **Covariates were screened but not retained.** Age, sex, ethnicity, race,
  body weight, height and BMI were tested on the drug liking IC50 by a single
  forward step in PsN stepwise covariate modelling; none was significant
  (p > 0.2) and the paper reports no point estimates. They are recorded in the
  model file under `covariatesDataExcluded` to preserve the provenance of the
  screen without adding unused covariates to the model. No covariate analysis
  is reported for the desire to use or COWS endpoints.

* **IIV on Imax is fixed to zero.** Table 1 reports `IIV Imax` as SD = 0 (FIX)
  for drug liking, so no random effect is placed on `imax`.

* **Arm sizes.** Table S1 reports 22 participants on 24 mg and 25 on 32 mg. The
  Figure 1 caption gives "32 mg: n = 24", which refers to the PK-evaluable
  subset; the model files use the Table S1 values.
