---
title: "Missing dosing history in population PK (Wang 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing dosing history in population PK (Wang 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Wang_2019_oral1cmt_base")()$meta$reference`
* Article: <https://doi.org/10.1002/psp4.12374>
* Open access: PMC6363138

Wang and Liu (2019) is a **methodology paper**, not a drug report. There is no
real molecule and there are no real patients. The authors defined the
population PK parameters of a *hypothetical* drug, simulated cohorts of 100
virtual subjects, then **deleted dosing records** from the simulated data sets
and compared how six analysis methods recovered the parameters they had
started from:

| Method | What it does |
|---|---|
| IDM | *Ideal method* -- fits the complete data set. The reference. |
| OM  | *Omit method* -- comments out the missing dose and the observations that follow it. |
| PDM | *Prescribed dose method* -- fills the gap with the prescribed dose (8 mg). |
| MDM | *Missing dose method* (Soy 2004) -- estimates an initial amount `A0` in the central compartment. |
| MDM2 | **New in this paper** -- MDM plus a patient-ID split after the gap, plus a two-population mixture on `A0`. |
| CIM | **New in this paper** -- *compartment initialization*: back-calculates `A0 = Conc x V` from the first observation after the gap, with no extra parameter. |

Because there is no drug, these models are filed under
`inst/modeldb/pharmacokinetics/` rather than `specificDrugs/`, following the
precedent of `Beal_2001_iv1cmt_bql` and the `Schoning_2026_oral1cmt_*` family,
and the file stem carries the structural descriptor `oral1cmt` in the slot
where a drug name would normally sit.

## What is, and is not, extractable here

The **data-generating model** is fully specified: Table 1 gives every value,
and the authors deposited the actual NONMEM `$SIMULATION` control stream in
Code Examples S1. That model is packaged here, in two files matching the
authors' own split ("base models and covariate models were separately
simulated"):

```{r models}
nlmixr2lib::modeldb |>
  dplyr::filter(name %in% c("Wang_2019_oral1cmt_base", "Wang_2019_oral1cmt_covariate")) |>
  dplyr::select(name, parameters, dosing) |>
  dplyr::rename("Model" = name, "Parameters" = parameters,
                "Dosing compartments" = dosing) |>
  knitr::kable()
```

The **fitted arms** (OM, PDM, MDM, MDM2, CIM, IDM) are *not* packaged, because
the paper publishes no point estimates for them at all -- only relative
estimation error (Figures 2, 3) and standardized RMSE (Tables 3, 4) summaries
over 100 replicates. The `$THETA` values in their deposited control streams
are deliberately offset *initial* estimates (CL 20 against a truth of 15,
V 120 against 100, `$OMEGA` 0.09 against 0.04), not results. An `ini()` block
for those arms would have nothing to put in it.

```{r load}
base <- readModelDb("Wang_2019_oral1cmt_base")()
cov  <- readModelDb("Wang_2019_oral1cmt_covariate")()
```

# Population

The virtual cohort (Table S1, and "PK data simulation"):

* 100 subjects per data set -- 50 men, 50 women.
* Weight normally distributed: men 70 (10) kg, women 60 (10) kg.
* Three doses at 0, 12 and 24 h (Table 2), oral or i.v. bolus.
* Each administered dose drawn from **Uniform(5, 10) mg**, although the
  *prescribed* dose is 8 mg every 12 h. This gap between prescription and
  reality is the whole point: it is what makes PDM biased.
* 12 samples per subject at 11.83, 12.25, 12.75, 13, 17, 23.83, 24.25, 24.75,
  25, 27, 31 and 35 h, giving 1,200 observations per data set.

# Source trace

Every value in both `ini()` blocks, with the location it came from. The paper
and the deposited control stream agree on every one.

| Parameter | Value | Table 1 | Code Examples S1 (`$PROB RUN# 011_1`) |
|---|---|---|---|
| `lka` | `log(2)` | `ka` = 2 (1/h) | `$THETA 2 ;[KA, 1/h]` |
| `lcl` | `log(15)` | `CL` = 15 L/h (set A) | `$THETA 15 ;[CL, L/h]` |
| `lvc` | `log(100)` | `V` = 100 L | `$THETA 100 ;[V, L]` |
| `e_wt_cl` | 0.75 | `beta1`, weight influence on CL | `$THETA 0.75 ;[CLWT]`, `CLWT=(WT/70)**THETA(4)` |
| `e_wt_vc` | 1 | `beta2`, weight influence on V | `$THETA 1.0 ;[VWT]`, `VWT=(WT/70)**THETA(6)` |
| `e_sexf_cl` | -0.3 | `alpha`, "Male: -0.3, female: 0" | `$THETA -0.3 ;[CLSEX]`, `TVCL=...*(1 + CLSEX*SEX)` |
| `etalcl` | 0.04 | variance of BSV on CL (20% CV) | `$OMEGA 0.04 omega(1,1)`, `CL=TVCL*EXP(ETA(1))` |
| `etalvc` | 0.04 | variance of BSV on V (20% CV) | `$OMEGA 0.04 omega(2,2)`, `V=TVV*EXP(ETA(2))` |
| `propSd` | 0.1 | `sigma1^2` = 0.01 (10% CV) | `$SIGMA 0.01 ;[P] sigma(1,1)` |
| `addSd` | 1 | `sigma2^2` = 1 (ng/mL) | `$SIGMA 1 ;[A] sigma(2,2)` |
| `F` = 1 | (no term) | "F ... 1 fixed" | no `F1` statement in `$PK` (ADVAN2 default) |
| ng/mL scaling | `* 1000` | concentrations reported in ng/mL | `S2=V/1000` |

# Structural verification against closed forms

This model defines variables named `cl` and `vc`, which can make rxode2 match
a system against its analytic kernel and silently discard the explicit
`d/dt()` bodies. Here the ODEs *are* ADVAN2 TRANS2, so an analytic match would
be equivalent rather than wrong -- but that must be demonstrated, not assumed.
First confirm both states survive the solve:

```{r states}
ui <- rxode2::rxode2(base)
stopifnot(identical(ui$state, c("depot", "central")))
```

Now pin the solved profile to the closed form for each route. Both sides use
the same parameter values, so the only difference is solver quadrature error
and a tight bound is the correct gate.

```{r closedform}
D   <- 7 # mg
cl0 <- 15; vc0 <- 100; ka0 <- 2; kel0 <- cl0 / vc0
typ <- rxode2::zeroRe(base)
grid <- sort(unique(c(seq(0, 36, by = 0.25), exp(seq(log(1e-3), log(1), length.out = 30)))))

# Oral: dose into the depot
s_oral <- rxode2::rxSolve(
  typ, rxode2::et(amt = D, cmt = "depot") |> rxode2::et(grid),
  returnType = "data.frame"
)
cf_oral <- 1000 * D * ka0 / (vc0 * (ka0 - kel0)) *
  (exp(-kel0 * s_oral$time) - exp(-ka0 * s_oral$time))

# IV bolus: dose straight into central
s_iv <- rxode2::rxSolve(
  typ, rxode2::et(amt = D, cmt = "central") |> rxode2::et(grid),
  returnType = "data.frame"
)
cf_iv <- 1000 * D / vc0 * exp(-kel0 * s_iv$time)

relerr <- function(a, b) max(abs(a - b) / pmax(b, 1e-9))
c(oral = relerr(s_oral$Cc, cf_oral), iv = relerr(s_iv$Cc, cf_iv))
stopifnot(
  relerr(s_oral$Cc, cf_oral) < 1e-5,
  relerr(s_iv$Cc, cf_iv) < 1e-5,
  all(s_iv$depot == 0) # an i.v. bolus must never populate the depot
)
```

## The covariate model reproduces its reference subject exactly

The authors' clearance line is `TVCL = THETA(1) * CLWT * (1 + CLSEX*SEX)` with
`SEX = 1` for **men**, so `THETA(1) = 15 L/h` is the clearance of a 70 kg
**woman**, and men clear 30% more slowly. The canonical covariate column is
`SEXF` (1 = female), so the effect is applied as
`(1 + e_sexf_cl * (1 - SEXF))` -- the construction ratified for
male-indicator sources in `Bajaj_2017_nivolumab.R` and
`Wada_2023_sparsentan.R`. That must land on exactly 15 and 10.5 L/h:

```{r covref}
typc <- rxode2::zeroRe(cov)
iv_ev <- function(sexf, wt) {
  e <- as.data.frame(rxode2::et(amt = D, cmt = "central") |> rxode2::et(grid))
  e$SEXF <- sexf
  e$WT <- wt
  e
}
s_w70 <- rxode2::rxSolve(typc, iv_ev(1, 70), returnType = "data.frame")
s_m70 <- rxode2::rxSolve(typc, iv_ev(0, 70), returnType = "data.frame")
s_w50 <- rxode2::rxSolve(typc, iv_ev(1, 50), returnType = "data.frame")

stopifnot(
  # 70 kg woman: both covariate terms are exactly 1, so CL = THETA(1) = 15
  relerr(s_w70$Cc, 1000 * D / vc0 * exp(-(15 / 100) * s_w70$time)) < 1e-5,
  # 70 kg man: CL = 15 * (1 - 0.3) = 10.5
  relerr(s_m70$Cc, 1000 * D / vc0 * exp(-(10.5 / 100) * s_m70$time)) < 1e-5,
  # mutation control: the weight term must actually be live
  max(abs(s_w50$Cc - s_w70$Cc) / s_w70$Cc) > 0.01
)
c(CL_woman70 = unique(s_w70$cl), CL_man70 = unique(s_m70$cl))
```

# Replicating Figure 1: the four missing-record cases

Figure 1 illustrates the i.v. dosing situation. The underlying subject is the
same in all four panels -- what changes is only which dosing records the
*analyst* can see. Solid segments are PK events with an available dosing
record; dashed segments follow a missing one.

```{r fig1, fig.height = 5.5}
doses <- c(6, 9, 8) # mg at 0, 12, 24 h
prof <- rxode2::rxSolve(
  typ,
  rxode2::et(amt = doses, time = c(0, 12, 24), cmt = "central") |>
    rxode2::et(seq(0, 36, by = 0.05)),
  returnType = "data.frame"
)

cases <- list(
  "(a) Complete records"          = integer(0),
  "(b) 1st dose missing (case 1)" = 1L,
  "(c) 2nd dose missing (case 2)" = 2L,
  "(d) 3rd dose missing (case 3)" = 3L
)
fig1 <- dplyr::bind_rows(lapply(names(cases), function(nm) {
  miss <- cases[[nm]]
  # a PK event is "unobserved" from its own dose until the next dose
  win <- if (length(miss)) c(c(0, 12, 24)[miss], c(12, 24, Inf)[miss]) else c(Inf, Inf)
  dplyr::mutate(prof, panel = nm,
                observed = !(time >= win[1] & time < win[2]))
}))
dose_pts <- dplyr::bind_rows(lapply(names(cases), function(nm) {
  data.frame(panel = nm, time = c(0, 12, 24),
             recorded = !(seq_len(3) %in% cases[[nm]]))
}))

ggplot(fig1, aes(time, Cc)) +
  geom_line(aes(linetype = observed, group = interaction(panel, observed))) +
  geom_vline(data = dose_pts, aes(xintercept = time, colour = recorded),
             linewidth = 0.4) +
  facet_wrap(~panel, ncol = 2) +
  scale_linetype_manual(values = c("TRUE" = "solid", "FALSE" = "22"),
                        name = "dosing record available") +
  scale_colour_manual(values = c("TRUE" = "grey50", "FALSE" = "firebrick"),
                      name = "dose recorded") +
  labs(x = "Time (h)", y = "Concentration (ng/mL)",
       caption = "Replicates Figure 1 of Wang and Liu (2019); i.v. dosing, doses 6/9/8 mg.") +
  theme_bw() + theme(legend.position = "bottom")
```

# Virtual cohort and simulated trial

Reproducing the paper's own design: 100 subjects, 50/50 sex split, weights per
Table S1, three doses drawn from Uniform(5, 10) mg, and the 12 sampling times
of Table 2.

```{r cohort}
rxode2::rxSetSeed(20190117)
set.seed(20190117)
n <- 100L
demog <- data.frame(
  id   = seq_len(n),
  SEXF = rep(c(0, 1), each = n / 2)
) |>
  dplyr::mutate(WT = round(stats::rnorm(n, ifelse(SEXF == 1, 60, 70), 10), 3))

samp <- c(11.83, 12.25, 12.75, 13, 17, 23.83, 24.25, 24.75, 25, 27, 31, 35)
amt_tbl <- data.frame(
  id   = rep(demog$id, each = 3),
  time = rep(c(0, 12, 24), times = n),
  amt  = round(stats::runif(3 * n, 5, 10))
)

ev <- dplyr::bind_rows(
  dplyr::transmute(amt_tbl, id, time, amt, evid = 1L, cmt = "depot"),
  data.frame(id = rep(demog$id, each = length(samp)),
             time = rep(samp, times = n),
             amt = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::left_join(demog, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim <- rxode2::rxSolve(cov, ev, returnType = "data.frame")
nrow(dplyr::filter(sim, !is.na(Cc)))
```

```{r cohortplot}
ggplot(dplyr::filter(sim, time > 0), aes(time, sim, group = id)) +
  geom_line(alpha = 0.12) +
  stat_summary(aes(group = 1), fun = median, geom = "line",
               colour = "firebrick", linewidth = 1) +
  labs(x = "Time (h)", y = "Observed concentration (ng/mL)",
       title = "Simulated oral cohort (n = 100), covariate model",
       subtitle = "Grey: individual observations with residual error. Red: median.") +
  theme_bw()
```

The cohort must match the demographics the paper specified:

```{r cohortcheck}
demog |>
  dplyr::group_by(Sex = ifelse(SEXF == 1, "Female", "Male")) |>
  dplyr::summarise(n = dplyr::n(), mean_WT = round(mean(WT), 1),
                   sd_WT = round(stats::sd(WT), 1), .groups = "drop") |>
  knitr::kable()

stopifnot(
  sum(demog$SEXF == 1) == 50L, sum(demog$SEXF == 0) == 50L,
  # Table S1 means, allowing for sampling error at n = 50 (SE = 10/sqrt(50) = 1.4)
  abs(mean(demog$WT[demog$SEXF == 0]) - 70) < 4,
  abs(mean(demog$WT[demog$SEXF == 1]) - 60) < 4,
  all(amt_tbl$amt >= 5 & amt_tbl$amt <= 10)
)
```

# PKNCA validation

The paper reports **no NCA table** -- its results are estimation-error
summaries, not exposure metrics -- so there is nothing to compare against
side by side. PKNCA is instead used for the structural identity that must
hold for any linear one-compartment model: after a single dose, individual
clearance times individual `AUCinf` recovers the dose.

```{r pknca}
iv_single <- dplyr::left_join(
  dplyr::bind_rows(
    data.frame(id = demog$id, time = 0, amt = 8, evid = 1L, cmt = "central"),
    data.frame(id = rep(demog$id, each = 40), amt = NA_real_, evid = 0L,
               cmt = "central",
               time = rep(sort(unique(c(exp(seq(log(1e-3), log(72), length.out = 39)), 0))),
                          times = n))
  ),
  demog, by = "id"
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

nca_sim <- rxode2::rxSolve(cov, iv_single, returnType = "data.frame")

conc_df <- nca_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, time, Cc, treatment = "IV 8 mg")
dose_df <- data.frame(id = demog$id, time = 0, dose = 8, treatment = "IV 8 mg")

o_conc <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                           concu = "ng/mL", timeu = "h")
o_dose <- PKNCA::PKNCAdose(dose_df, dose ~ time | treatment + id,
                           doseu = "mg")
res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  o_conc, o_dose,
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE, half.life = TRUE)
))

nca <- as.data.frame(res) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(id = as.integer(as.character(id)))

indiv <- nca_sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), vc = dplyr::first(vc), .groups = "drop")

chk <- dplyr::left_join(nca, indiv, by = "id") |>
  dplyr::mutate(
    # Cc is ng/mL = ug/L, so cl (L/h) * AUC (ug*h/L) is ug; the dose is 8 mg.
    dose_recovery = cl * aucinf.obs / (8 * 1000),
    hl_ratio      = half.life / (log(2) * vc / cl)
  )
```

```{r pkncatable}
chk |>
  dplyr::summarise(
    `Dose recovery (cl x AUCinf / dose)` = median(dose_recovery),
    `Half-life vs ln(2)*vc/cl`           = median(hl_ratio)
  ) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "Identity",
                      values_to = "Median ratio") |>
  dplyr::mutate(`Median ratio` = round(`Median ratio`, 5)) |>
  knitr::kable()
```

Both identities are asserted on the centre and on a robust quantile, never on
the cohort extreme -- the extreme of a random cohort is not reproducible
across rxode2 builds, whereas these medians are.

```{r pkncagate}
stopifnot(
  !anyNA(chk$dose_recovery), !anyNA(chk$hl_ratio),
  abs(median(chk$dose_recovery) - 1) < 0.005,
  stats::quantile(abs(chk$dose_recovery - 1), 0.9) < 0.02,
  abs(median(chk$hl_ratio) - 1) < 0.005,
  stats::quantile(abs(chk$hl_ratio - 1), 0.9) < 0.02
)
```

# Why the missing-dose methods are needed

Two of the paper's claims are arithmetic, and can be checked directly against
this model rather than restated as prose.

**PDM is biased because the prescription is not the dose.** The prescribed
dose is 8 mg; the administered doses are Uniform(5, 10). Filling a gap with
8 mg therefore misstates the true amount by a predictable amount:

```{r pdm}
pdm_err <- (8 - amt_tbl$amt) / amt_tbl$amt
c(`median absolute dose error` = median(abs(pdm_err)),
  `max absolute dose error`    = max(abs(pdm_err)))
stopifnot(median(abs(pdm_err)) > 0.05)
```

**CIM's identity is exact.** The compartment initialization method sets
`A0(t) = Conc(t) x V`, so re-initializing the central compartment from an
observed concentration must reproduce that concentration exactly. Solving a
fresh profile whose initial central amount is back-calculated from a sampled
concentration recovers the sampled value:

```{r cim}
# take a 70 kg woman, sample her true concentration at 11.83 h, then
# re-initialize a fresh solve from A0 = Conc * V and confirm it lands there.
probe <- rxode2::rxSolve(typc, iv_ev(1, 70), returnType = "data.frame")
conc_t <- approx(probe$time, probe$Cc, xout = 11.83)$y
V_i    <- unique(probe$vc)
A0     <- conc_t * V_i / 1000 # ng/mL -> mg/L, times L, gives mg

reinit <- as.data.frame(
  rxode2::et(amt = A0, cmt = "central") |> rxode2::et(c(0, 1e-6))
)
reinit$SEXF <- 1
reinit$WT <- 70
back <- rxode2::rxSolve(typc, reinit, returnType = "data.frame")

c(sampled = conc_t, reinitialized = back$Cc[1])
stopifnot(abs(back$Cc[1] - conc_t) / conc_t < 1e-6)
```

# Assumptions and deviations

* **Parameter set A is encoded.** Table 1 lists five sets A-E differing *only*
  in clearance (15, 20, 25, 30, 35 L/h). Set A is used because it is the set
  the authors' own deposited `$SIMULATION` control stream actually runs
  (`$THETA 15 ;[CL, L/h]`). Change `lcl` to `log(20)` ... `log(35)` for B-E.

* **The weight reference is 70 kg, not the dataset median.** The paper's
  printed equation normalizes a continuous covariate by "the median value of
  the continuous covariate in the dataset", which for a balanced cohort drawn
  at 70 (10) kg for men and 60 (10) kg for women is about 65 kg. The authors'
  deposited control stream instead hardcodes `CLWT=(WT/70)**THETA(4)` and
  `VWT=(WT/70)**THETA(6)`. The executed code governs, so 70 kg is used here.

* **Sex reference category is inverted relative to the source.** The source
  column `SEX` is a male indicator (`SEX=1 IF MALE, 0 IF FEMALE`), so the
  canonical `SEXF = 1 - SEX`. The published coefficient -0.3 is preserved
  verbatim and applied to `(1 - SEXF)`; consequently `lcl = log(15)` is the
  clearance of a 70 kg **woman**, and men are 30% lower. Verified numerically
  above.

* **Table 1's two omega row labels are transposed.** The row labelled
  `omega2 - V` is described as "Variance of BSV on CL" and the row labelled
  `omega2 - CL` as "Variance of BSV on V". Both values are 0.04, so the
  transposition has no numerical consequence; the control stream
  (`CL=TVCL*EXP(ETA(1))`, `$OMEGA 0.04 omega(1,1)`) settles the mapping.

* **The deposited data example is not from parameter set A.** Data Examples S1
  gives one subject (WT 64.395 kg, male, doses 6/9/8 mg) with concentrations.
  Its troughs decay far faster than set A allows: the observed 23.83 h trough
  is 2.51 ng/mL against 34.4 predicted for set A, and a least-squares fit at
  fixed V = 100 and ka = 2 returns CL ~ 46 L/h -- above every published set,
  and reachable from set E (35 L/h) with a ~1.4 SD individual eta but from set
  A only with a ~5.6 SD one. The deposited *control stream* and the deposited
  *data file* therefore come from different runs. This does not affect the
  encoded values, which are doubly sourced from Table 1 and `$THETA`, but the
  deposited data cannot be used as a validation target and is not used as one.

* **Fitted-arm models are deliberately absent**, as explained above: the paper
  publishes relative errors, not estimates.

* **Specimen is unverified.** The paper says only that "blood samples were
  collected during visits and later analyzed for drug concentrations
  (ng/mL)"; it never states plasma, serum or whole blood, so
  `compartmentData$central$verified` is `FALSE`.

* **The 8 mg nominal dose and the Uniform(5, 10) actual dose are both real
  features of the design**, not a transcription conflict -- the divergence
  between them is the mechanism the paper is studying.
