---
title: "Mycophenolic acid (Wang 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mycophenolic acid (Wang 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
options(cli.progress_show_after = Inf)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wang_2022_mycophenolic_acid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wang_2022_mycophenolic_acid"))$description`
* Article: <https://doi.org/10.3389/fphar.2022.859351> (open access)

Wang 2022 compared mycophenolic acid (MPA) pharmacokinetics in adult renal
transplant recipients sampled in the first post-transplant week with
recipients sampled 5.5-10 years after transplantation. A single
two-compartment model with a lagged first-order absorption was fitted to both
groups in Phoenix NLME, and the post-transplant stage was the only retained
covariate: it lowers CL/F and V/F in the stable state.

# Population

Ninety-nine adult Chinese renal transplant recipients at the First Affiliated
Hospital of Zhengzhou University (August 2019 - June 2021), all on triple
therapy with oral mycophenolate mofetil (MMF) dispersible tablets twice daily,
tacrolimus and corticosteroids. Table 1 (mean +/- SD, early / stable):

| Characteristic | Early stage (n = 51) | Stable state (n = 48) |
|----------------|----------------------|-----------------------|
| Male, n (%) | 43 (84.3%) | 38 (79.2%) |
| Age (years) | 33.39 +/- 8.10 | 42.29 +/- 9.25 |
| Weight (kg) | 64.02 +/- 10.45 | 65.31 +/- 10.54 |
| Post-transplant time (days) | 4.88 +/- 1.01 | 2499.94 +/- 467.26 |
| MMF dose (g/day) | 1.69 +/- 0.44 | 1.09 +/- 0.29 |
| Creatinine clearance (mL/min) | 62.55 +/- 20.75 | 74.51 +/- 19.29 |
| Serum albumin (g/L) | 41.94 +/- 4.95 | 46.07 +/- 2.24 |

Early-stage patients also received pantoprazole. 1079 MPA plasma
concentrations (561 early, 518 stable) were analysed, one full 12-h
steady-state profile per patient (pre-dose plus mainly 0.5, 1, 1.5, 2, 3, 4, 6,
8 and 12 h). The same information is available programmatically via
`rxode2::rxode(readModelDb("Wang_2022_mycophenolic_acid"))$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Two-compartment, first-order absorption with lag, linear elimination, no EHC | -- | Results 'Population Pharmacokinetics Models'; Discussion |
| `lka` (ka) | 1.36 1/h | Table 3 `tvka` |
| `lvc` (V/F, early stage) | 78.07 L | Table 3 `tvV/F` |
| `lvp` (V2/F) | 554.52 L | Table 3 `tvV2/F` |
| `lcl` (CL/F, early stage) | 23.36 L/h | Table 3 `tvCL/F` |
| `lq` (Q/F) | 29.53 L/h | Table 3 `tvQ/F` |
| `ltlag` (Tlag) | 0.23 h | Table 3 `tvTlag` |
| `e_posttx_stable_vc` | -2.54 | Table 3 `dVdStage` |
| `e_posttx_stable_cl` | -0.82 | Table 3 `dCLdStage` |
| Stage coding (0 = early, 1 = stable) and exponential form | -- | Methods 'Population Pharmacokinetics Modeling'; Results CL/F 10.25 L/h in the stable state |
| IIV variances (omega^2) | V 1.03, CL 0.20, ka 0.34, V2 1.72, Q 0.98, Tlag 0.74 | Table 3 |
| `propSd` | 0.37 | Table 3 `stdev0`; Methods proportional form `Cobs = Cpred x (1 + eps)` |
| Dose units (mg MMF, no molecular-weight conversion) | -- | Methods; Table 2 NCA CL/F = MMF dose / AUC |

# Typical-value checks

All random effects are switched off here, so the comparisons are exact.

```{r model}
mod <- rxode2::rxode(readModelDb("Wang_2022_mycophenolic_acid"))
stopifnot(identical(mod$state, c("depot", "central", "peripheral1")))

tau <- 12
# Day-4 dosing interval: doses at 0, 12, ..., 72 h, sampling 72-84 h.
n_dose <- 7
t_day4 <- (n_dose - 1) * tau
samp <- c(0, 0.5, 1, 1.5, 2, 3, 4, 6, 8, 12)

make_events <- function(ids, dose, obs_times, n_doses = n_dose) {
  one_id <- function(i, amt) {
    dplyr::bind_rows(
      data.frame(
        id = i, time = seq(0, by = tau, length.out = n_doses),
        amt = amt, evid = 1L, cmt = "depot"
      ),
      data.frame(id = i, time = obs_times, amt = 0, evid = 0L, cmt = "central")
    )
  }
  dose <- rep_len(dose, length(ids))
  dplyr::bind_rows(Map(one_id, ids, dose)) |>
    # Observation before dose at a shared time, so the 72-h sample is a trough.
    dplyr::arrange(id, time, dplyr::desc(evid == 0L))
}

typ <- function(stage, dose, obs_times, n_doses = n_dose) {
  ev <- make_events(1L, dose, obs_times, n_doses)
  ev$POSTTX_STABLE <- stage
  rxode2::rxSolve(mod, ev,
    omega = NA, sigma = NA, returnType = "data.frame",
    rtol = 1e-10, atol = 1e-12, maxsteps = 1e6
  )
}
```

## Stage effect on CL/F and V/F

```{r stage-effect}
ini_df <- mod$iniDf
th <- setNames(ini_df$est, ini_df$name)
stage_tab <- data.frame(
  Quantity = c("CL/F, early (L/h)", "CL/F, stable (L/h)", "V/F, early (L)", "V/F, stable (L)"),
  Model = c(
    exp(th[["lcl"]]), exp(th[["lcl"]] + th[["e_posttx_stable_cl"]]),
    exp(th[["lvc"]]), exp(th[["lvc"]] + th[["e_posttx_stable_vc"]])
  ),
  Published = c(23.36, 10.25, 78.07, 16.24)
) |>
  dplyr::mutate(`Difference (%)` = 100 * (Model / Published - 1))
knitr::kable(stage_tab,
  digits = 2,
  caption = "Typical CL/F and V/F by stage against the values quoted in the Abstract and Results."
)
cl_stable <- stage_tab$Model[2]
stopifnot(
  # Exponential form: 23.36 * exp(-0.82) = 10.29 vs the quoted 10.25 (rounding).
  abs(cl_stable / 10.25 - 1) < 0.01,
  abs(stage_tab$Model[1] / 23.36 - 1) < 1e-8,
  abs(stage_tab$Model[3] / 78.07 - 1) < 1e-8
)
```

The stable-state CL/F reproduces the quoted 10.25 L/h to rounding, which
confirms the exponential categorical form `CL/F = 23.36 x exp(-0.82 x Stage)`.
The stable-state V/F does **not** reproduce the quoted 16.24 L: Table 3's
`dVdStage = -2.54` gives 6.16 L. See "Assumptions and deviations" below; the
model encodes Table 3.

## Steady-state AUC equals dose / CL

After a long dosing train, the AUC over one interval equals dose / (CL/F) for a
linear model, independent of the absorption and distribution parameters.

```{r ss-auc}
trap <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
ss_check <- dplyr::bind_rows(lapply(0:1, function(stage) {
  t0 <- 59 * tau
  s <- typ(stage, 1000, t0 + seq(0, tau, by = 0.005), n_doses = 61)
  data.frame(
    stage = stage, auc_tau = trap(s$time, s$Cc),
    dose_over_cl = 1000 / exp(th[["lcl"]] + stage * th[["e_posttx_stable_cl"]])
  )
})) |>
  dplyr::mutate(rel_diff = auc_tau / dose_over_cl - 1)
knitr::kable(ss_check, digits = 5)
# Trapezoid error on a 0.005-h grid is ~1e-5; a wrong CL, stage effect or
# dose unit moves this by tens of percent.
stopifnot(all(abs(ss_check$rel_diff) < 1e-3))
```

## Typical day-4 profiles at the mean dose

```{r typical-profiles}
mean_dose <- c(early = 1690 / 2, stable = 1090 / 2) # Table 1 mean MMF g/day, bid
prof <- dplyr::bind_rows(
  typ(0, mean_dose[["early"]], t_day4 + seq(0, 12, by = 0.05)) |>
    dplyr::mutate(stage = "Early (845 mg bid)"),
  typ(1, mean_dose[["stable"]], t_day4 + seq(0, 12, by = 0.05)) |>
    dplyr::mutate(stage = "Stable (545 mg bid)")
)
ggplot(prof, aes(time - t_day4, Cc, colour = stage)) +
  geom_line() +
  labs(
    x = "Time after the day-4 morning dose (h)", y = "MPA (mg/L)", colour = NULL,
    caption = "Typical-value profiles; compare with Figure 1 of Wang 2022."
  ) +
  theme_bw()
```

# Virtual cohort and NCA against Table 2

Each stage gets 200 virtual patients. Per-subject daily MMF doses are drawn
from a normal distribution with the Table 1 mean and SD for that stage,
truncated to the Methods dose range (1.0-3.0 g/day early, 0.5-1.5 g/day
stable) and given as two equal doses. The paper's 10-point sampling schedule
is simulated with residual error, as Table 2 is an NCA of observed
concentrations: Cmax and Cmin are read directly from the sampled values and
AUC0-12 uses the linear trapezoidal rule.

```{r cohort}
set.seed(20220509)
n_per_stage <- 200
draw_dose <- function(n, mean_day, sd_day, lo, hi) {
  d <- numeric(0)
  while (length(d) < n) {
    x <- stats::rnorm(n, mean_day, sd_day)
    d <- c(d, x[x >= lo & x <= hi])
  }
  d[seq_len(n)] / 2
}
cohort <- data.frame(
  id = seq_len(2 * n_per_stage),
  POSTTX_STABLE = rep(0:1, each = n_per_stage),
  dose = c(
    draw_dose(n_per_stage, 1690, 440, 1000, 3000),
    draw_dose(n_per_stage, 1090, 290, 500, 1500)
  )
)
ev_cohort <- make_events(cohort$id, cohort$dose, t_day4 + samp) |>
  dplyr::left_join(dplyr::select(cohort, id, POSTTX_STABLE), by = "id")

rxode2::rxSetSeed(20220509)
sim <- rxode2::rxSolve(mod, ev_cohort, returnType = "data.frame", maxsteps = 1e6) |>
  dplyr::select(-dplyr::any_of(c("POSTTX_STABLE", "dose"))) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::mutate(treatment = ifelse(POSTTX_STABLE == 1, "Stable state", "Early stage"))
stopifnot(!anyNA(sim$sim), nrow(sim) == nrow(cohort) * length(samp))
```

```{r vpc, fig.height = 4}
sim |>
  dplyr::mutate(tad = time - t_day4) |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    p05 = quantile(sim, 0.05), p50 = median(sim), p95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~treatment) +
  labs(
    x = "Time after dose (h)", y = "MPA (mg/L)",
    caption = "Simulated 5th-50th-95th percentiles with residual error; compare with Figure 4 of Wang 2022."
  ) +
  theme_bw()
```

```{r nca}
# Interval-relative time: the dose of the sampled interval is at time 0.
conc_df <- sim |>
  dplyr::transmute(id, treatment, time = time - t_day4, conc = sim) |>
  dplyr::filter(!is.na(conc))
# A normal proportional error makes Cc * (1 + eps) negative when eps < -1
# (P = 0.0035 for sd 0.37). Such samples would read as below the assay limit,
# so they enter the NCA as zero.
stopifnot(mean(conc_df$conc < 0) < 0.02)
conc_df$conc <- pmax(conc_df$conc, 0)
dose_df <- cohort |>
  dplyr::mutate(
    treatment = ifelse(POSTTX_STABLE == 1, "Stable state", "Early stage"),
    time = 0
  ) |>
  dplyr::select(id, treatment, time, dose)

intervals <- data.frame(
  start = 0, end = 12, cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE
)
run_nca <- function(trt) {
  o_conc <- PKNCA::PKNCAconc(dplyr::filter(conc_df, treatment == trt), conc ~ time | treatment + id)
  o_dose <- PKNCA::PKNCAdose(dplyr::filter(dose_df, treatment == trt), dose ~ time | treatment + id)
  o_data <- PKNCA::PKNCAdata(o_conc, o_dose,
    intervals = intervals,
    options = list(auc.method = "linear")
  )
  as.data.frame(PKNCA::pk.nca(o_data))
}
nca_res <- dplyr::bind_rows(lapply(unique(conc_df$treatment), run_nca))

# Table 2 reports arithmetic means, so aggregate to means before comparing.
sim_mean <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "tmax", "auclast")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES), .groups = "drop")

published <- data.frame(
  treatment = c("Early stage", "Stable state"),
  auclast = c(32.61, 42.30),
  cmax = c(9.83, 16.87),
  cmin = c(0.96, 1.29),
  # Table 2 Tmax is on the dataset clock (74.10 and 72.82 h); the sampled
  # interval starts at 72 h.
  tmax = c(74.10, 72.82) - 72
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_mean,
  reference = published,
  by = "treatment",
  units = c(cmax = "mg/L", cmin = "mg/L", tmax = "h", auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp,
  caption = paste(
    "Simulated (mean of 200 virtual patients per stage) vs Wang 2022 Table 2",
    "(mean of observed NCA). * differs from the reference by more than 20%."
  )
)
attr(cmp, "footnote")
```

```{r nca-gate}
get_sim <- function(trt, p) {
  v <- sim_mean$PPORRES[sim_mean$treatment == trt & sim_mean$PPTESTCD == p]
  if (length(v) != 1L) stop("no unique simulated value for ", trt, " / ", p)
  v
}
auc_pct <- c(
  100 * (get_sim("Early stage", "auclast") / 32.61 - 1),
  100 * (get_sim("Stable state", "auclast") / 42.30 - 1)
)
auc_ratio <- get_sim("Stable state", "auclast") / get_sim("Early stage", "auclast")
stopifnot(
  # AUC0-12 is dose / CL-dominated: a wrong clearance, stage effect or dose
  # unit moves it by tens of percent. The mean of 200 subjects with
  # omega^2(CL) = 0.20 has a Monte Carlo SE of about 3%.
  all(abs(auc_pct) < 15),
  # Table 2: 42.30 / 32.61 = 1.30 despite a 35% lower stable-state dose.
  abs(auc_ratio / (42.30 / 32.61) - 1) < 0.2
)
```

AUC0-12 is reproduced within a few percent in both stages, as is the
stable-state Cmax. The early-stage Cmax runs about 15% below the observed mean
and the early-stage Tmax is earlier than observed (2.1 h after the dose): ka
and Tlag are shared by both stages, so the model cannot reproduce the slower
early-stage absorption that the observed Tmax suggests, and the stable-state
Tmax is correspondingly a little later than observed. Cmin is about 15-20%
below the observed means in both stages. None of these move the AUC, which is the quantity the paper's dosing
recommendations rest on.

# Monte Carlo dosing simulation (Figure 5)

Wang 2022 simulated six fixed regimens in each stage (0.25-1.5 g bid, 1000
virtual patients each) and reported the percentage of patients with a day-4
AUC0-12 in the 30-60 mg h/L target window. Here each arm has 200 virtual
patients; AUC0-12 is integrated from the individual predictions on a 0.1-h grid.

```{r pta}
regimens <- c(250, 500, 750, 1000, 1250, 1500)
n_pta <- 200
pta_design <- tidyr::expand_grid(POSTTX_STABLE = 0:1, dose = regimens) |>
  dplyr::mutate(arm = dplyr::row_number())
pta_cohort <- pta_design[rep(seq_len(nrow(pta_design)), each = n_pta), ] |>
  dplyr::mutate(id = dplyr::row_number())
ev_pta <- make_events(pta_cohort$id, pta_cohort$dose, t_day4 + seq(0, 12, by = 0.1)) |>
  dplyr::left_join(dplyr::select(pta_cohort, id, POSTTX_STABLE), by = "id")

rxode2::rxSetSeed(859351)
sim_pta <- rxode2::rxSolve(mod, ev_pta, returnType = "data.frame", maxsteps = 1e6)
stopifnot(!anyNA(sim_pta$Cc))
auc_pta <- sim_pta |>
  dplyr::group_by(id) |>
  dplyr::summarise(auc = trap(time, Cc), .groups = "drop") |>
  dplyr::left_join(pta_cohort, by = "id") |>
  dplyr::mutate(stage = ifelse(POSTTX_STABLE == 1, "Stable state", "Early stage"))

pta <- auc_pta |>
  dplyr::group_by(stage, dose) |>
  dplyr::summarise(simulated = 100 * mean(auc >= 30 & auc <= 60), .groups = "drop") |>
  dplyr::arrange(stage, dose) |>
  dplyr::mutate(
    # Results 'Model-Based Simulations'.
    published = c(0.1, 9.0, 41.2, 59.4, 62.0, 51.2, 7.3, 60.1, 53.3, 31.7, 15.0, 7.5),
    difference = simulated - published
  )
knitr::kable(
  dplyr::rename(pta,
    "Stage" = stage, "MMF dose (mg bid)" = dose,
    "Simulated in target (%)" = simulated, "Published in target (%)" = published,
    "Difference (points)" = difference
  ),
  digits = 1,
  caption = "Probability of a day-4 AUC0-12 of 30-60 mg h/L."
)
```

```{r pta-figure}
auc_pta |>
  ggplot(aes(factor(dose / 1000), auc)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(yintercept = c(30, 60), linetype = "dashed", colour = "red") +
  facet_wrap(~stage) +
  labs(
    x = "MMF dose (g bid)", y = "Day-4 AUC0-12 (mg h/L)",
    caption = "Replicates Figure 5 of Wang 2022."
  ) +
  theme_bw()
```

```{r pta-gate}
best <- pta |>
  dplyr::group_by(stage) |>
  dplyr::slice_max(simulated, n = 1, with_ties = FALSE)
pta_cell <- function(stg, d) {
  v <- pta$simulated[pta$stage == stg & pta$dose == d]
  if (length(v) != 1L) stop("no unique PTA cell for ", stg, " at ", d, " mg")
  v
}
mad_stage <- tapply(abs(pta$difference), pta$stage, mean)
stopifnot(
  nrow(pta) == 12L,
  # Binomial SE of one 200-subject cell is <= 3.5 points. Mean absolute
  # differences were 2.4 / 5.3 points (early / stable) in a 1000-subject run
  # by the maintainers and 1.7 / 6.1 in this cohort; a mis-transcribed CL,
  # stage effect or dose unit shifts whole columns by tens of points.
  all(mad_stage < 10),
  # The paper's recommendation: 1.0-1.5 g bid early, 0.50-0.75 g bid stable.
  best$dose[best$stage == "Early stage"] %in% c(1000, 1250, 1500),
  best$dose[best$stage == "Stable state"] %in% c(500, 750),
  # Crossing of the two stages: tens of points apart at both ends.
  pta_cell("Stable state", 500) - pta_cell("Early stage", 500) > 25,
  pta_cell("Early stage", 1250) - pta_cell("Stable state", 1250) > 20
)
mad_stage
```

The simulated target-attainment pattern reproduces the paper's: the early
stage needs 1.0-1.5 g bid and the stable state 0.50-0.75 g bid. The early
stage matches cell by cell. In the stable state the 0.5 g bid arm attains the
target somewhat less often than published and the 1.25-1.5 g bid arms
somewhat more often, i.e. the simulated stable-state AUC distribution is a
little wider than the paper's; the same pattern appeared in a 1000-patient
run by the maintainers (52% at 0.5 g bid, 22% at 1.25 g bid), so it is not
only Monte Carlo noise. The paper does not describe its simulation settings
beyond the regimens and the cohort size.

# Assumptions and deviations

* **Stable-state V/F.** The Abstract and Results quote a
  stable-state V/F of 16.24 L, but Table 3's `dVdStage = -2.54` with the
  exponential form that reproduces the quoted stable CL/F (10.25 L/h) gives
  78.07 x exp(-2.54) = 6.16 L. The coefficient is internally consistent with
  its own uncertainty (CV 15.32%, 95% CI -3.29 to -1.77) and bootstrap median
  (-2.36), and no other parameterisation reproduces 16.24 L
  (log(16.24 / 78.07) = -1.57 lies outside the CI). The model therefore
  encodes Table 3. A simulation by the maintainers at the stable-state mean
  dose gave a mean steady-state Cmax of about 15 mg/L with -2.54 versus about
  11 mg/L with V/F = 16.24 L, against an observed Table 2 Cmax of 16.87 mg/L,
  which also favours Table 3. AUC is unaffected by the choice.
* **Doses are mg of MMF.** The model has no MMF-to-MPA molecular-weight
  conversion: Table 2's NCA CL/F equals MMF dose / AUC, and the simulated
  AUC0-12 reproduces Table 2 with the dose entered as mg MMF. All apparent
  clearances and volumes are relative to the MMF dose.
* **Post-transplant stage covariate.** `POSTTX_STABLE` is the paper's own
  binary `Stage` column (0 = early, 1 = stable). The cohort contains only
  patients sampled on days 4-8 or after 5.5-10 years, so the model says
  nothing about intermediate post-transplant times and no day-count threshold
  is implied.
* **No IIV covariances.** Table 3 reports only diagonal variances; the matrix
  is diagonal.
* **Virtual cohort.** Per-subject doses are drawn from a truncated normal
  distribution built from Table 1's mean and SD; the paper does not give the
  dose distribution. No other covariate enters the model.
* **Day-4 sampling.** Table 2's Tmax values (74.10 and 72.82 h) indicate that
  the sampled interval starts at 72 h, matching the Methods statement that
  sampling was at least 3 days after the same dose and the Monte Carlo
  simulations' "AUCss,12h on day 4". The NCA and PTA simulations use seven
  doses at 0-72 h and sample 72-84 h.
* **Monte Carlo details.** The paper does not say whether its simulated AUCs
  included residual error; the PTA here uses individual predictions without
  residual error, which reproduces the published percentages closely.
* **Supplement.** Supplementary Tables S1-S2 and Figures S1-S2 (the
  enterohepatic-circulation model comparison, covariate screening and
  limited-sampling regression equations) are not needed for the final model,
  whose parameters are all in Table 3. No erratum was found in a literature
  check on 2026-10-02.
