---
title: "CPX-351 liposomal cytarabine + daunorubicin (Wang 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CPX-351 liposomal cytarabine + daunorubicin (Wang 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

CPX-351 (Vyxeos) is a single liposome co-encapsulating cytarabine and
daunorubicin at a fixed 5:1 molar ratio. Wang 2019 fitted **two separate
population PK models**, one per analyte, to pooled data from three clinical
studies, and this vignette validates both. Following the library's
replicate-the-authors' -structure policy, the two independent fits are shipped
as two model files that share this one vignette.

```{r models}
ui_cyt <- rxode2::rxode(readModelDb("Wang_2019_cytarabine_cpx351"))
ui_dnr <- rxode2::rxode(readModelDb("Wang_2019_daunorubicin_cpx351"))
```

* Citation: `r ui_cyt$reference`
* Article: <https://doi.org/10.1002/jcph.1366>
* Supplement (Tables S1-S10): <https://doi.org/10.1002/jcph.1366> Supporting Information

Cytarabine model description:

> `r ui_cyt$description`

Daunorubicin model description:

> `r ui_dnr$description`

Both models are two-compartment with linear elimination from the central
compartment and no absorption compartment (CPX-351 is given as a 90-minute
intravenous infusion). Because more than 99% of circulating drug stays inside
the liposome, the measured "total" cytarabine and daunorubicin are essentially
liposome markers, which is why the two analytes have such similar disposition
despite being chemically unrelated (Discussion).

# Population

The pharmacokinetic-evaluable population was 195 patients contributing 2176
plasma samples across three studies (Table 1): the phase 1 dose-escalation
study 101 (n = 38, frozen formulation, 3 to 134 units/m^2), the phase 2 QTc
study 206 (n = 26, lyophilized formulation) and the phase 3 randomized study
301 (n = 131, lyophilized formulation). Median age was 67 years (range 24-81),
median body weight 79.8 kg (38.9-156.5), median body surface area 1.94 m^2
(1.26-2.80) and 61.0% were male. The cohort was 84.6% White, 4.6% Black, 5.1%
Asian, 0.5% Native American and 5.1% other.

Renal and hepatic function were the covariates of interest: creatinine
clearance median 85.3 mL/min (27.5-211.7), with 42.6% normal, 42.6% mild,
14.4% moderate and 0.5% severe impairment; total bilirubin median 0.60 mg/dL
(0.1-2.5), with 91.8% below 1.2 mg/dL and 8.2% between 1.2 and 3 mg/dL. No
patient had bilirubin above 3 mg/dL and none had end-stage renal disease, so
the models must not be extrapolated into those strata.

The same information is available programmatically:

```{r population}
str(ui_cyt$population[c("n_subjects", "n_studies", "age_median", "bsa_median")])
```

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Wang_2019_cytarabine_cpx351.R` and
`inst/modeldb/specificDrugs/Wang_2019_daunorubicin_cpx351.R`. Table 2 of the
paper prints the covariate model inline and already exponentiated; Supplemental
Tables S2 and S3 decompose the same models into the individual estimated
parameters with their standard errors, and are the primary source used here.

| Equation / parameter | Cytarabine | Daunorubicin | Source location |
|---|---|---|---|
| `lcl` (CL, L/h) | 0.107 | 0.147 | Table S2 / S3 row `CL (L/h)`; Table 2 |
| `lvc` (Vc, L) | 4.92 | 4.29 | Table S2 / S3 row `Vc (L)`; Table 2 |
| `lq` (Q, L/h) | 0.00165 | 0.0294 | Table S2 / S3 row `Q (L/h)`; Table 2 |
| `lvp` (Vp, L) | 0.129 | 0.593 | Table S2 / S3 row `Vp (L)`; Table 2 |
| `e_bsa_cl` | 1.01 | 0.829 | Table S2 / S3 row `BSA_CL` |
| `e_bsa_vc` | 1.22 | 1.12 | Table S2 / S3 row `BSA_Vc` (Table 2 prints 1.21 for cytarabine) |
| `e_bsa_q` (fixed) | 1.00 | 1.00 | Table S2 / S3 row `BSA_Q`, SE column reads `fixed` |
| `e_bsa_vp` (fixed) | 1.00 | 1.00 | Table S2 / S3 row `BSA_Vp`, SE column reads `fixed` |
| `e_tbili_cl` | 0.197 | 0.0829 | Table S2 / S3 row `BILI_CL` |
| `e_form_cpx351_frozen_cl` | not retained | -0.275 | Table S3 row `Form_CL` |
| `e_form_cpx351_frozen_vc` | not retained | -0.221 | Table S3 row `Form_Vc` |
| `e_form_cpx351_frozen_q` | not retained | -0.551 | Table S3 row `Form_Q` |
| `e_form_cpx351_frozen_vp` | not retained | -0.686 | Table S3 row `Form_Vp` |
| `etalcl` (omega^2) | 0.316 (61.0%) | 0.218 (49.3%) | Table S2 / S3 row `BSV_CL`; Table 2 BSV% column |
| `etalvc` (omega^2) | 0.0592 (24.7%) | 0.0547 (23.7%) | Table S2 / S3 row `BSV_Vc` |
| `etalq` (omega^2) | not estimated | 0.392 (69.3%) | Table S3 row `BSV_Q`; Table 2 footnote for cytarabine |
| `etalvp` (omega^2) | not estimated | 0.563 (86.9%) | Table S3 row `BSV_Vp`; Table 2 footnote for cytarabine |
| `etaexpSd` (omega^2) | 0.525 (83.1%) | 0.720 (102.7%) | Table S2 / S3 row `BSV_LogErr` |
| `expSd` (log-scale residual SD) | 0.187 | 0.143 | Table S2 / S3 row `LogErr`; Table 2 `Error model` row |
| Covariate form `theta * (BSA/1.95)^e * (Bili/0.60)^e` | n/a | n/a | Methods, continuous-covariate equation; Table 2 footnotes a and b |
| Residual form `ln(Cobs) = ln(Cpred) + LogErr * exp(eta)` | n/a | n/a | Table S2 / S3 `Equation` column |
| ODE structure (2-compartment, linear CL from central) | n/a | n/a | Results, `Population Pharmacokinetic Modeling` of each analyte |

The `BSV` rows of Tables S2 and S3 are variances: their footnote gives the
conversion `omega^2 % = sqrt(exp(omega^2) - 1)`, which is checked below.

```{r bsv-scale-check}
# Every BSV% printed in Table 2 must be recoverable from the Table S2/S3
# variance by the footnote formula. This is arithmetic on published numbers,
# not a simulation, so the tolerance is numerical.
bsv <- tibble::tribble(
  ~analyte, ~parameter, ~omega2, ~published_cv_pct,
  "cytarabine", "CL", 0.316, 61.0,
  "cytarabine", "Vc", 0.0592, 24.7,
  "cytarabine", "LogErr", 0.525, 83.1,
  "daunorubicin", "CL", 0.218, 49.3,
  "daunorubicin", "Vc", 0.0547, 23.7,
  "daunorubicin", "Q", 0.392, 69.3,
  "daunorubicin", "Vp", 0.563, 86.9,
  "daunorubicin", "LogErr", 0.720, 102.7
) |>
  dplyr::mutate(
    derived_cv_pct = 100 * sqrt(exp(omega2) - 1),
    abs_diff_pct_points = abs(derived_cv_pct - published_cv_pct)
  )

stopifnot(max(bsv$abs_diff_pct_points) < 0.1)

bsv |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 3))) |>
  dplyr::rename(
    "Analyte" = analyte,
    "Parameter" = parameter,
    "omega^2 (Table S2/S3)" = omega2,
    "BSV% (Table 2)" = published_cv_pct,
    "BSV% from omega^2" = derived_cv_pct,
    "Difference (pct points)" = abs_diff_pct_points
  ) |>
  knitr::kable(
    caption = "Table S2/S3 variances reproduce the Table 2 BSV% column exactly."
  )
```

The frozen-formulation coefficients are stored on the log scale as Table S3
estimates them; exponentiating must reproduce the multipliers Table 2 prints.

```{r formulation-scale-check}
form <- tibble::tribble(
  ~parameter, ~log_scale, ~published_multiplier,
  "CL", -0.275, 0.760,
  "Vc", -0.221, 0.802,
  "Q", -0.551, 0.576,
  "Vp", -0.686, 0.504
) |>
  dplyr::mutate(derived_multiplier = exp(log_scale))

stopifnot(max(abs(form$derived_multiplier - form$published_multiplier)) < 0.001)

form |>
  dplyr::mutate(derived_multiplier = round(derived_multiplier, 4)) |>
  dplyr::rename(
    "Parameter" = parameter,
    "Table S3 log-scale estimate" = log_scale,
    "Table 2 printed multiplier" = published_multiplier,
    "exp(log-scale estimate)" = derived_multiplier
  ) |>
  knitr::kable(
    caption = "Frozen-formulation effects: Table S3 log scale vs Table 2 multipliers."
  )
```

# Structural validation: closed-form terminal half-life

Supplemental Table S4 reports a mean terminal half-life of 56.1 h for total
cytarabine and 26.2 h for total daunorubicin following CPX-351. Neither value
is an input to the models, so it is an independent check on all four
disposition parameters at once: for a two-compartment model with linear
elimination the terminal rate constant is fixed by `CL`, `Vc`, `Q` and `Vp`
through the smaller root of the characteristic equation. This is arithmetic on
the packaged `ini()` values, not a simulation, so the tolerance is tight.

```{r half-life-closed-form}
theta_of <- function(ui, nm) {
  v <- ui$iniDf$est[ui$iniDf$name == nm]
  stopifnot(length(v) == 1L)
  v
}

# Terminal (beta) half-life of a two-compartment model with elimination from
# the central compartment.
half_life_2cmt <- function(cl, vc, q, vp) {
  k10 <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  a <- k10 + k12 + k21
  beta <- (a - sqrt(a^2 - 4 * k10 * k21)) / 2
  log(2) / beta
}

# Reference patient: BSA 1.95 m^2, bilirubin 0.60 mg/dL, lyophilized
# formulation. Every covariate multiplier is then exactly 1, so the typical
# values are exp() of the ini() thetas.
typical_pars <- function(ui) {
  list(
    cl = exp(theta_of(ui, "lcl")),
    vc = exp(theta_of(ui, "lvc")),
    q = exp(theta_of(ui, "lq")),
    vp = exp(theta_of(ui, "lvp"))
  )
}

p_cyt <- typical_pars(ui_cyt)
p_dnr <- typical_pars(ui_dnr)

thalf <- tibble::tibble(
  analyte = c("Total cytarabine", "Total daunorubicin"),
  published_thalf_h = c(56.1, 26.2),
  model_thalf_h = c(
    do.call(half_life_2cmt, p_cyt),
    do.call(half_life_2cmt, p_dnr)
  ),
  published_v_L = c(4.92, 4.29),
  model_vc_L = c(p_cyt$vc, p_dnr$vc),
  published_cl_Lh = c(0.107, 0.147),
  model_cl_Lh = c(p_cyt$cl, p_dnr$cl)
) |>
  dplyr::mutate(thalf_pct_diff = 100 * (model_thalf_h - published_thalf_h) / published_thalf_h)

# Deterministic: no random effects, no solver, no cohort. A mis-transcribed
# Q or Vp moves the terminal slope by tens of percent, so 1% is a real gate.
stopifnot(max(abs(thalf$thalf_pct_diff)) < 1)

thalf |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) signif(x, 4))) |>
  dplyr::rename(
    "Analyte" = analyte,
    "t-half, Table S4 (h)" = published_thalf_h,
    "t-half from ini() (h)" = model_thalf_h,
    "% diff" = thalf_pct_diff,
    "Mean V, Table S4 (L)" = published_v_L,
    "Vc from ini() (L)" = model_vc_L,
    "Mean CL, Table S4 (L/h)" = published_cl_Lh,
    "CL from ini() (L/h)" = model_cl_Lh
  ) |>
  knitr::kable(
    caption = paste(
      "Closed-form terminal half-life from the packaged disposition parameters",
      "against the independently reported Table S4 values. Table S4's mean V",
      "and mean CL are the same quantities as Vc and CL and are shown for",
      "completeness."
    )
  )
```

# Virtual cohort

Original patient data are not public. The cohort below reproduces the covariate
distributions of the 156 patients from studies 206 and 301 who received the
recommended induction dose and who make up the Table 3 AUCtau analysis set. All
of them received the lyophilized formulation, so `FORM_CPX351_FROZEN` is 0
throughout and the frozen-formulation arm of the daunorubicin model is
exercised separately below.

The two bilirubin strata are simulated as **equal-sized arms of 150 subjects
each**, rather than in the 141:15 proportion Table 3 reports. The comparison
below is between per-stratum medians, and a median over the 15 subjects the
real study had above 1.2 mg/dL is dominated by which subjects happen to be
drawn: across six trial seeds its deviation from Table 3 swung between -4% and
+32%, which is noise, not model error. Sampling each stratum to 150 shrinks
that swing to under 11% without changing what is being compared, because each
arm draws from the *conditional* distribution of the same fitted bilirubin
lognormal on its own interval. Both arms stay within the 200-per-arm cap.

```{r cohort}
# set.seed() seeds R's RNG (used for the covariate draws below). It does NOT
# seed rxode2's simulation RNG, whose streams are partitioned per solver
# thread -- so the etas drawn during rxSolve differ between a 2-core CI runner
# and a 16-thread workstation. Every assertion downstream is written to hold
# for any cohort these models can produce.
set.seed(20190501)

n_per_stratum <- 150L

# CPX-351 100 units/m^2 = cytarabine 100 mg/m^2 + daunorubicin 44 mg/m^2,
# given as 90-minute infusions on days 1, 3 and 5 (Methods, studies 206/301).
dose_times_h <- c(0, 48, 96)
infusion_h <- 1.5
mg_per_m2 <- c(cytarabine = 100, daunorubicin = 44)

# Total bilirubin (mg/dL): log-normal centred on the Table 1 median 0.60, with
# sdlog set so that P(bilirubin > 1.2) = 8.2%, the Table 1 stratum split, and
# support restricted to the Table 1 range 0.1-2.5 mg/dL. Drawing a large pool
# and taking the first n_per_stratum values on each side of 1.2 mg/dL gives
# each arm exactly the conditional distribution of this lognormal on its own
# interval -- the same within-stratum distribution a proportionally sized
# cohort would have, just with more subjects in the sparse upper stratum.
sdlog_bili <- log(1.2 / 0.60) / qnorm(1 - 0.082)
pool <- rlnorm(2e5, meanlog = log(0.60), sdlog = sdlog_bili)
pool <- pool[pool >= 0.1 & pool <= 2.5]
bili_lo <- head(pool[pool < 1.2], n_per_stratum)
bili_hi <- head(pool[pool >= 1.2], n_per_stratum)
stopifnot(length(bili_lo) == n_per_stratum, length(bili_hi) == n_per_stratum)
bili_mgdl <- c(bili_lo, bili_hi)

n_subj <- 2L * n_per_stratum

# BSA: log-normal centred on the Table 1 median 1.94 m^2, spread chosen so the
# central 95% spans roughly the Table 1 range, then clipped to that range.
# Drawn independently of bilirubin and identically in both arms, so the
# stratum contrast is a pure bilirubin contrast.
bsa <- pmin(pmax(rlnorm(n_subj, meanlog = log(1.94), sdlog = 0.15), 1.26), 2.80)

subjects <- tibble::tibble(
  id = seq_len(n_subj),
  BSA = bsa,
  # The register carries TBILI in SI umol/L; the model converts back internally.
  TBILI = bili_mgdl * 17.1,
  FORM_CPX351_FROZEN = 0,
  bili_cat = rep(
    c("Bilirubin < 1.2 mg/dL", "Bilirubin 1.2-3 mg/dL"),
    each = n_per_stratum
  )
)
stopifnot(
  all((subjects$TBILI / 17.1 < 1.2) ==
    (subjects$bili_cat == "Bilirubin < 1.2 mg/dL"))
)

# Observation grid: dense around each infusion and through the day-5 dosing
# interval (96-144 h), which is the AUCtau window of Table 3.
obs_times_h <- sort(unique(c(
  as.vector(outer(c(seq(0, 6, by = 0.25)), dose_times_h, "+")),
  seq(0, 144, by = 2),
  seq(96, 144, by = 1)
)))

make_events <- function(subjects, mg_per_m2_analyte) {
  doses <- subjects |>
    tidyr::crossing(time = dose_times_h) |>
    dplyr::mutate(
      amt = mg_per_m2_analyte * BSA,
      rate = amt / infusion_h,
      evid = 1L,
      cmt = "central"
    )
  obs <- subjects |>
    tidyr::crossing(time = obs_times_h) |>
    dplyr::mutate(
      amt = NA_real_,
      rate = NA_real_,
      evid = 0L,
      # The ODE state, never the algebraic observable Cc.
      cmt = "central"
    )
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

ev_cyt <- make_events(subjects, mg_per_m2[["cytarabine"]])
ev_dnr <- make_events(subjects, mg_per_m2[["daunorubicin"]])

stopifnot(!anyDuplicated(unique(ev_cyt[, c("id", "time", "evid")])))
stopifnot(!anyDuplicated(unique(ev_dnr[, c("id", "time", "evid")])))

# Cohort composition against Table 1.
table(subjects$bili_cat)
```

# Simulation

```{r simulate}
sim_cyt <- rxode2::rxSolve(
  readModelDb("Wang_2019_cytarabine_cpx351"),
  events = as.data.frame(ev_cyt),
  keep = c("bili_cat", "BSA")
) |>
  as.data.frame()

sim_dnr <- rxode2::rxSolve(
  readModelDb("Wang_2019_daunorubicin_cpx351"),
  events = as.data.frame(ev_dnr),
  keep = c("bili_cat", "BSA")
) |>
  as.data.frame()

stopifnot(all(sim_cyt$Cc >= 0), all(sim_dnr$Cc >= 0))
```

## Concentration-time profiles

The paper's own concentration-time figures are visual predictive checks in the
Supporting Information (Figures S4 and S7) rather than numbered main-text
figures, so the panels below reproduce their shape: median and 5th-95th
percentile band of total analyte concentration across the induction cycle.

```{r figure-vpc, fig.width = 7, fig.height = 4}
vpc_band <- function(sim, label) {
  # rxSolve returns observation records only -- there is no evid column to
  # filter on in its output.
  sim |>
    dplyr::group_by(time) |>
    dplyr::summarise(
      Q05 = quantile(Cc, 0.05, na.rm = TRUE),
      Q50 = quantile(Cc, 0.50, na.rm = TRUE),
      Q95 = quantile(Cc, 0.95, na.rm = TRUE),
      .groups = "drop"
    ) |>
    dplyr::mutate(analyte = label)
}

dplyr::bind_rows(
  vpc_band(sim_cyt, "Total cytarabine"),
  vpc_band(sim_dnr, "Total daunorubicin")
) |>
  dplyr::filter(Q05 > 0) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_vline(xintercept = dose_times_h, linetype = "dotted") +
  facet_wrap(~analyte) +
  scale_y_log10() +
  labs(
    x = "Time (h)",
    y = "Total analyte concentration (ug/mL)",
    title = "Simulated induction cycle: 100 units/m2 on days 1, 3 and 5",
    caption = "Median and 5th-95th percentile band; dotted lines are the infusion start times."
  ) +
  theme_bw()
```

## The 5:1 cytarabine:daunorubicin molar ratio

A defining property of CPX-351 is that the 5:1 molar ratio of cytarabine to
daunorubicin is maintained in plasma. That claim is a *joint* statement about
the two models, which were fitted independently, so it is a genuine
out-of-sample check on the pair. Molar masses are standard chemical constants
(cytarabine 243.22 g/mol, daunorubicin 527.52 g/mol); the paper's own
statement that 1 mg cytarabine + 0.44 mg daunorubicin is a 5:1 molar ratio
corroborates them, since 1/243.22 divided by 0.44/527.52 is 4.93.

```{r molar-ratio, fig.width = 6, fig.height = 3.5}
mw <- c(cytarabine = 243.22, daunorubicin = 527.52)

# The dose ratio the paper itself calls 5:1.
dose_molar_ratio <- (mg_per_m2[["cytarabine"]] / mw[["cytarabine"]]) /
  (mg_per_m2[["daunorubicin"]] / mw[["daunorubicin"]])
stopifnot(abs(dose_molar_ratio - 5) < 0.2)

# Typical-value profiles (omega = NA suppresses between-subject variability),
# so the ratio below is deterministic.
typ_subject <- tibble::tibble(
  id = 1:2, BSA = 1.95, TBILI = 0.60 * 17.1, FORM_CPX351_FROZEN = 0,
  bili_cat = "reference"
)
typ_cyt <- rxode2::rxSolve(
  readModelDb("Wang_2019_cytarabine_cpx351"),
  events = as.data.frame(make_events(typ_subject, mg_per_m2[["cytarabine"]])),
  omega = NA
) |>
  as.data.frame()
typ_dnr <- rxode2::rxSolve(
  readModelDb("Wang_2019_daunorubicin_cpx351"),
  events = as.data.frame(make_events(typ_subject, mg_per_m2[["daunorubicin"]])),
  omega = NA
) |>
  as.data.frame()

ratio <- typ_cyt |>
  dplyr::filter(id == 1) |>
  dplyr::select(time, Cc_cyt = Cc) |>
  dplyr::inner_join(
    typ_dnr |> dplyr::filter(id == 1) |> dplyr::select(time, Cc_dnr = Cc),
    by = "time"
  ) |>
  dplyr::mutate(
    molar_ratio = (Cc_cyt / mw[["cytarabine"]]) / (Cc_dnr / mw[["daunorubicin"]])
  ) |>
  # Drops only the pre-dose record, where both concentrations are exactly 0.
  dplyr::filter(is.finite(molar_ratio))

ratio_range <- range(ratio$molar_ratio)

ggplot(ratio, aes(time, molar_ratio)) +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  geom_vline(xintercept = dose_times_h, linetype = "dotted") +
  expand_limits(y = c(0, 8)) +
  labs(
    x = "Time (h)",
    y = "Cytarabine : daunorubicin molar ratio",
    title = "Plasma molar ratio over the induction cycle (typical values)",
    caption = "Dashed line is the nominal 5:1 ratio of the administered dose."
  ) +
  theme_bw()
```

The realised molar ratio spans `r sprintf("%.2f to %.2f", ratio_range[1], ratio_range[2])`
over the 144-hour induction cycle. It is not perfectly flat, and it cannot be:
the two models were fitted separately and give clearances differing by 37%
(0.107 vs 0.147 L/h), so the cytarabine:daunorubicin ratio necessarily drifts
upward with time. The check below bounds that drift rather than asserting a
constant ratio.

```{r molar-ratio-gate}
# Deterministic (typical values, no etas). The bound allows the drift the two
# independently fitted clearances imply, but would break on a transposed CL,
# a swapped dose, or a unit error in either model -- any of which moves the
# ratio by an order of magnitude.
stopifnot(all(ratio$molar_ratio > 2), all(ratio$molar_ratio < 12))
```

# PKNCA validation

## AUC over the day-5 dosing interval (Table 3 AUCtau)

Wang 2019 defines AUCtau as "the area under the curve from time zero to 48
hours after the dose on day 5", i.e. the 96-144 h window of this simulation,
and reports it by bilirubin category in Table 3.

```{r pknca}
nca_for <- function(sim, ev, concu = "ug/mL") {
  sim_nca <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc, bili_cat)

  # Guarantee a time = 0 record. The first infusion starts at t = 0, so the
  # pre-dose concentration is exactly 0.
  sim_nca <- dplyr::bind_rows(
    sim_nca,
    sim_nca |> dplyr::distinct(id, bili_cat) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, bili_cat, time, .keep_all = TRUE) |>
    dplyr::arrange(id, bili_cat, time)

  dose_df <- ev |>
    dplyr::filter(evid == 1) |>
    dplyr::select(id, time, amt, bili_cat)

  conc_obj <- PKNCA::PKNCAconc(
    sim_nca, Cc ~ time | bili_cat + id,
    concu = concu, timeu = "h"
  )
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | bili_cat + id, doseu = "mg")

  intervals <- data.frame(
    start = 96,
    end = 144,
    auclast = TRUE,
    cmax = TRUE,
    tmax = TRUE
  )

  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_cyt <- nca_for(sim_cyt, ev_cyt)
nca_dnr <- nca_for(sim_dnr, ev_dnr)
```

## Comparison against the published AUCtau

Table 3 reports the median and mean AUCtau in each bilirubin category for the
156 patients of studies 206 and 301 treated at 100 units/m^2. The comparison
below uses the medians, which are robust to which subjects a given simulation
run happens to draw into the tails.

```{r nca-compare-cytarabine}
published_cyt <- tibble::tribble(
  ~bili_cat, ~auclast,
  "Bilirubin < 1.2 mg/dL", 1830,
  "Bilirubin 1.2-3 mg/dL", 1425
)

cmp_cyt <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_cyt,
  reference = published_cyt,
  by = "bili_cat",
  units = c(auclast = "ug*h/mL", cmax = "ug/mL", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_cyt,
  caption = paste(
    "Total cytarabine: simulated AUCtau (96-144 h) against the Table 3 medians.",
    "* differs from reference by more than 20%. Cmax and Tmax were computed but",
    "have no published counterpart, so they do not appear here."
  )
)
```

```{r nca-compare-daunorubicin}
published_dnr <- tibble::tribble(
  ~bili_cat, ~auclast,
  "Bilirubin < 1.2 mg/dL", 587,
  "Bilirubin 1.2-3 mg/dL", 520
)

cmp_dnr <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_dnr,
  reference = published_dnr,
  by = "bili_cat",
  units = c(auclast = "ug*h/mL", cmax = "ug/mL", tmax = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_dnr,
  caption = paste(
    "Total daunorubicin: simulated AUCtau (96-144 h) against the Table 3",
    "medians. * differs from reference by more than 20%."
  )
)
```

```{r nca-gate}
auctau_diff <- function(cmp) {
  d <- as.data.frame(cmp)
  pct_col <- grep("diff", names(d), ignore.case = TRUE, value = TRUE)[1]
  keep <- !is.na(d[[pct_col]]) & grepl("AUC", d[[1]], ignore.case = TRUE)
  suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", d[[pct_col]][keep])))
}

diffs <- c(auctau_diff(cmp_cyt), auctau_diff(cmp_dnr))
stopifnot(length(diffs) == 4L, !anyNA(diffs))

# Cohort-derived, so assert on the CENTRE (each stratum's median AUCtau), not
# on any extreme. Realised max|diff| was 10.21 / 7.75 / 7.29 / 7.57 / 9.05 /
# 7.47 % across six independent trial seeds at 150 subjects per stratum, and
# was identical at 1, 2, 4 and 8 solver threads. 20 sits well outside that
# range and still goes red on a mis-transcribed clearance, dose, reference
# BSA or concentration unit, every one of which moves these medians by tens of
# percent. (The same bound on a proportionally sized cohort would NOT be safe:
# with only ~13 subjects above 1.2 mg/dL the same six seeds gave up to 31.8%.)
stopifnot(max(abs(diffs)) < 20)
```

## Independent check: AUC over one dose equals dose divided by clearance

For a linear model, the total AUC of a single dose is exactly `dose / CL`. That
identity involves no published NCA value, so it validates the whole
event-table, unit and solver path end to end at typical parameter values. The
same solve also lets PKNCA estimate the terminal half-life numerically, as a
second route to the Table S4 values that is independent of the closed form
used earlier.

The observation windows differ sharply between the two analytes, and that is
itself informative. The cytarabine model's peripheral compartment is tiny
(Vp 0.129 L against Vc 4.92 L) and its two disposition half-lives are poorly
separated (30.8 h and 56.1 h), so the terminal exponential carries only about
4% of the initial concentration and does not dominate the profile until
several hundred hours after the dose. A window matched to the clinical
sampling scheme returns a half-life biased roughly 20% low; the 1600-hour
window below is needed for NCA to resolve the model's true terminal slope.
This is the numerical face of the authors' own caveat that "the population
pharmacokinetic model was limited by the availability of data defining the
terminal phase; thus, parameters associated with the peripheral compartment
were less precisely described". Table S4's 56.1 h is a model-derived terminal
half-life, not something NCA on the observed data could have produced. The
daunorubicin model has a well-separated terminal phase and needs only 400
hours.

```{r auc-identity}
single_dose_events <- function(mg_per_m2_analyte, tmax_h) {
  doses <- typ_subject |>
    dplyr::mutate(
      time = 0, amt = mg_per_m2_analyte * BSA, rate = amt / infusion_h,
      evid = 1L, cmt = "central"
    )
  obs <- typ_subject |>
    tidyr::crossing(time = sort(unique(c(
      seq(0, 6, by = 0.1), seq(6, 48, by = 0.5), seq(48, tmax_h, by = 2)
    )))) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

auc_identity <- function(model_name, mg_per_m2_analyte, cl_typical, tmax_h) {
  ev <- single_dose_events(mg_per_m2_analyte, tmax_h = tmax_h)
  sim <- rxode2::rxSolve(
    readModelDb(model_name),
    events = as.data.frame(ev), omega = NA,
    # Tightened because the cytarabine tail runs to ~1e-9 ug/mL, where default
    # tolerances would let solver noise reach the terminal-slope regression.
    atol = 1e-12, rtol = 1e-10
  ) |>
    as.data.frame()
  stopifnot(all(sim$Cc >= 0))

  sim_nca <- sim |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(id, time, Cc) |>
    dplyr::mutate(arm = "typical")
  sim_nca <- sim_nca[sim_nca$id == 1, , drop = FALSE]

  # Guarantee a time = 0 record; the infusion starts at t = 0 so Cc is 0 there.
  sim_nca <- dplyr::bind_rows(
    sim_nca,
    sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
  ) |>
    dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
    dplyr::arrange(id, arm, time)

  conc_obj <- PKNCA::PKNCAconc(
    sim_nca, Cc ~ time | arm + id, concu = "ug/mL", timeu = "h"
  )
  dose_obj <- PKNCA::PKNCAdose(
    ev |> dplyr::filter(evid == 1, id == 1) |>
      dplyr::select(id, time, amt) |> dplyr::mutate(arm = "typical"),
    amt ~ time | arm + id, doseu = "mg"
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
    conc_obj, dose_obj,
    intervals = data.frame(
      start = 0, end = Inf, aucinf.obs = TRUE, half.life = TRUE, cmax = TRUE
    )
  ))
  out <- as.data.frame(res$result)
  list(
    aucinf = out$PPORRES[out$PPTESTCD == "aucinf.obs"],
    half_life = out$PPORRES[out$PPTESTCD == "half.life"],
    expected_auc = mg_per_m2_analyte * 1.95 / cl_typical
  )
}

id_cyt <- auc_identity(
  "Wang_2019_cytarabine_cpx351", mg_per_m2[["cytarabine"]], p_cyt$cl,
  tmax_h = 1600
)
id_dnr <- auc_identity(
  "Wang_2019_daunorubicin_cpx351", mg_per_m2[["daunorubicin"]], p_dnr$cl,
  tmax_h = 400
)

identity_tbl <- tibble::tibble(
  analyte = c("Total cytarabine", "Total daunorubicin"),
  window_h = c(1600, 400),
  dose_mg = c(mg_per_m2[["cytarabine"]], mg_per_m2[["daunorubicin"]]) * 1.95,
  cl_Lh = c(p_cyt$cl, p_dnr$cl),
  expected_auc = c(id_cyt$expected_auc, id_dnr$expected_auc),
  nca_auc = c(id_cyt$aucinf, id_dnr$aucinf),
  nca_half_life_h = c(id_cyt$half_life, id_dnr$half_life),
  published_half_life_h = c(56.1, 26.2)
) |>
  dplyr::mutate(
    auc_pct_diff = 100 * (nca_auc - expected_auc) / expected_auc,
    thalf_pct_diff = 100 * (nca_half_life_h - published_half_life_h) / published_half_life_h
  )

# Deterministic at typical values: the only error source is trapezoidal
# integration on a dense grid, so both bounds are tight.
stopifnot(max(abs(identity_tbl$auc_pct_diff)) < 2)
stopifnot(max(abs(identity_tbl$thalf_pct_diff)) < 2)

identity_tbl |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) signif(x, 4))) |>
  dplyr::rename(
    "Analyte" = analyte,
    "NCA window (h)" = window_h,
    "Dose at BSA 1.95 m2 (mg)" = dose_mg,
    "CL (L/h)" = cl_Lh,
    "Dose / CL (ug*h/mL)" = expected_auc,
    "PKNCA AUC0-inf (ug*h/mL)" = nca_auc,
    "% diff" = auc_pct_diff,
    "PKNCA t-half (h)" = nca_half_life_h,
    "Table S4 t-half (h)" = published_half_life_h,
    "t-half % diff" = thalf_pct_diff
  ) |>
  knitr::kable(
    caption = paste(
      "Single-dose typical-value NCA. AUC0-inf must equal dose / CL for a",
      "linear model, and the NCA terminal half-life must match both the",
      "closed form and Table S4."
    )
  )
```

# Covariate effects

## Bilirubin and renal function (Table 3)

Table 3 also stratifies AUCtau by renal function. The models carry **no** renal
covariate -- creatinine clearance was screened and not retained for either
analyte -- so a simulated renal stratification would be flat by construction
and is not a validation of anything. The paper's own conclusion is the same:
the 20% (cytarabine) and 18% (daunorubicin) higher mean AUCtau in moderate
impairment is smaller than the 33.5-48.0% within-stratum coefficient of
variation, and "these analyses did not suggest any associations between
cytarabine or daunorubicin plasma exposure and markers of renal function".

The bilirubin effect the models do carry is likewise shallow. The figure below
shows the model-implied AUCtau across the observed bilirubin range at fixed
BSA.

```{r bilirubin-effect, fig.width = 6, fig.height = 3.5}
bili_grid <- tibble::tibble(
  id = seq_along(seq(0.1, 2.5, by = 0.1)),
  BSA = 1.95,
  bili_mgdl = seq(0.1, 2.5, by = 0.1),
  TBILI = bili_mgdl * 17.1,
  FORM_CPX351_FROZEN = 0,
  bili_cat = "grid"
)

bili_auc <- function(model_name, mg_per_m2_analyte) {
  sim <- rxode2::rxSolve(
    readModelDb(model_name),
    events = as.data.frame(make_events(bili_grid |> dplyr::select(-bili_mgdl), mg_per_m2_analyte)),
    omega = NA, keep = "BSA"
  ) |>
    as.data.frame() |>
    dplyr::filter(time >= 96, time <= 144, !is.na(Cc))
  sim |>
    dplyr::group_by(id) |>
    dplyr::arrange(time, .by_group = TRUE) |>
    dplyr::summarise(
      auctau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      .groups = "drop"
    )
}

bili_effect <- dplyr::bind_rows(
  bili_auc("Wang_2019_cytarabine_cpx351", mg_per_m2[["cytarabine"]]) |>
    dplyr::mutate(analyte = "Total cytarabine"),
  bili_auc("Wang_2019_daunorubicin_cpx351", mg_per_m2[["daunorubicin"]]) |>
    dplyr::mutate(analyte = "Total daunorubicin")
) |>
  dplyr::left_join(bili_grid |> dplyr::select(id, bili_mgdl), by = "id")

ggplot(bili_effect, aes(bili_mgdl, auctau)) +
  geom_line() +
  geom_vline(xintercept = 0.60, linetype = "dashed") +
  facet_wrap(~analyte, scales = "free_y") +
  expand_limits(y = 0) +
  labs(
    x = "Baseline total bilirubin (mg/dL)",
    y = "Typical-value AUCtau, 96-144 h (ug*h/mL)",
    title = "Model-implied bilirubin effect at BSA = 1.95 m2",
    caption = "Dashed line is the 0.60 mg/dL reference of Table 2."
  ) +
  theme_bw()
```

```{r bilirubin-effect-gate}
# All deterministic: typical values, no random effects, no cohort.
at_bili <- function(df, target) {
  df$auctau[which.min(abs(df$bili_mgdl - target))]
}

span <- do.call(rbind, lapply(split(bili_effect, bili_effect$analyte), function(d) {
  # The reference 0.60 mg/dL doubled to 1.20 mg/dL, which is also the boundary
  # between the two Table 3 bilirubin strata.
  a060 <- at_bili(d, 0.60)
  a120 <- at_bili(d, 1.20)
  tibble::tibble(
    analyte = d$analyte[1],
    pct_drop_full_range = 100 * (max(d$auctau) - min(d$auctau)) / max(d$auctau),
    pct_drop_per_doubling = 100 * (a060 - a120) / a060
  )
}))

# "The effect of bilirubin on CL was statistically significant, but the
# relationship was very shallow" (Results, both analytes). Quantified: a
# DOUBLING of bilirubin, which also crosses the Table 3 stratum boundary,
# must move typical exposure by less than 15%.
stopifnot(all(span$pct_drop_per_doubling < 15))
# The direction is itself a published claim -- "patients with higher bilirubin
# values were associated with a slightly faster CL" -- and is deterministic
# here because both exponents are positive and the solve carries no etas.
stopifnot(all(span$pct_drop_per_doubling > 0))
# Across the full 0.1-2.5 mg/dL range the fall is larger (the range spans a
# 25-fold change in bilirubin) but still bounded. A transcription slip of one
# decimal place in either exponent -- 1.97 instead of 0.197 -- would put this
# above 98%.
stopifnot(all(span$pct_drop_full_range < 60))

span |>
  dplyr::mutate(dplyr::across(dplyr::where(is.numeric), \(x) round(x, 1))) |>
  dplyr::rename(
    "Analyte" = analyte,
    "AUCtau decrease, 0.1 to 2.5 mg/dL (%)" = pct_drop_full_range,
    "AUCtau decrease per doubling, 0.60 to 1.20 mg/dL (%)" = pct_drop_per_doubling
  ) |>
  knitr::kable(caption = "Magnitude of the bilirubin effect over the observed range.")
```

## Frozen versus lyophilized formulation (daunorubicin only)

The frozen formulation, used throughout the phase 1 study 101, lowers all four
daunorubicin disposition parameters. Because CL falls by 24% while Vc falls by
only 20%, the net effect on exposure is an increase.

```{r formulation-effect}
form_events <- function(frozen) {
  subj <- tibble::tibble(
    id = 1:2, BSA = 1.95, TBILI = 0.60 * 17.1,
    FORM_CPX351_FROZEN = frozen, bili_cat = "reference"
  )
  make_events(subj, mg_per_m2[["daunorubicin"]])
}

form_auc <- function(frozen) {
  sim <- rxode2::rxSolve(
    readModelDb("Wang_2019_daunorubicin_cpx351"),
    events = as.data.frame(form_events(frozen)), omega = NA
  ) |>
    as.data.frame() |>
    dplyr::filter(id == 1, time >= 96, time <= 144, !is.na(Cc)) |>
    dplyr::arrange(time)
  sum(diff(sim$time) * (head(sim$Cc, -1) + tail(sim$Cc, -1)) / 2)
}

auc_lyo <- form_auc(0)
auc_frz <- form_auc(1)

# Deterministic. exp(-0.275) = 0.760 on CL against exp(-0.221) = 0.802 on Vc,
# so the frozen arm must have the HIGHER AUCtau, by roughly 1/0.760 - 1 = 32%
# once distribution has equilibrated.
stopifnot(auc_frz > auc_lyo)

tibble::tibble(
  Formulation = c("Lyophilized (reference)", "Frozen"),
  `AUCtau, 96-144 h (ug*h/mL)` = signif(c(auc_lyo, auc_frz), 4),
  `Ratio to reference` = signif(c(1, auc_frz / auc_lyo), 3)
) |>
  knitr::kable(
    caption = "Typical-value daunorubicin exposure by CPX-351 formulation."
  )
```

# Assumptions and deviations

- **Covariate distributions are reconstructed, not observed.** Body surface
  area and total bilirubin are drawn from log-normal distributions matched to
  the Table 1 medians, restricted to the Table 1 ranges, and for bilirubin
  calibrated so that 8.2% of the underlying distribution falls in the 1.2-3
  mg/dL stratum as Table 1 reports. The two are drawn independently; any
  correlation between body size and hepatic function in the real cohort is not
  reproduced, which is the main reason the simulated stratum medians cannot be
  expected to match Table 3 exactly.
- **The two bilirubin strata are simulated at equal size (150 each) rather
  than in the 141:15 proportion of Table 3.** Each arm still draws from the
  conditional distribution of the same fitted bilirubin lognormal on its own
  interval, so the quantity being compared is unchanged; only the precision of
  the upper stratum's median improves. At the published 141:15 proportion that
  median moved by up to 32% between trial seeds, which would have forced a
  validation bound too loose to detect a real transcription error.
- **The published OMEGA blocks are encoded as diagonal.** Results describe an
  OMEGA block on CL and Vc for cytarabine and on CL, Vc, Q and Vp for
  daunorubicin, but Tables S2 and S3 report only the diagonal variances. The
  off-diagonal covariances are not published anywhere on disk, so the etas are
  encoded as independent. This narrows the joint spread of derived quantities
  such as AUC relative to the authors' model; the marginal variance of each
  parameter is unaffected.
- **No IIV on cytarabine Q and Vp.** The Table 2 footnote states "BSV% for Q
  and Vp were not estimated due to high shrinkage on these parameters" and
  Table S2 carries no corresponding rows, so the final cytarabine model has
  three random effects (CL, Vc and the residual magnitude). No eta is invented
  for the other two parameters.
- **`e_bsa_vc` for cytarabine is 1.22, not the 1.21 printed in Table 2.**
  Table S2 reports 1.22 with SE 0.186 and a 95% CI of 0.850-1.58, whose
  midpoint is 1.215; both printed figures are roundings of the same estimate.
  The supplement value is used because it carries the uncertainty that
  identifies it.
- **The bilirubin effect on daunorubicin clearance has a confidence interval
  spanning zero** (0.0829, 95% CI -0.0198 to 0.186). It is retained because
  the paper built a full model with no reduction step, following Harrell, not
  because the effect is well identified.
- **Formulation is perfectly confounded with study.** All 38 frozen-formulation
  patients came from study 101 and all 157 lyophilized patients from studies
  206 and 301, so the fitted `FORM_CPX351_FROZEN` coefficients are a combined
  formulation-plus-study effect. The Discussion separately reports that "study
  phase had no measurable effect", but that could not be assessed independently
  of formulation.
- **The exposure-response layer is not reproducible as a model and is not
  extracted.** Wang 2019's exposure-response analyses are Kaplan-Meier
  estimates, Cox proportional-hazards regressions and logistic regressions run
  in SAS. Table 4 reports adjusted hazard ratios and the text reports adjusted
  odds ratios, but no baseline hazard function and no logistic intercept are
  published, so neither a time-to-event nor a landmark probability model can be
  reconstructed from any on-disk source. The exposure variable itself, AUCtau
  from the population PK models, is reproduced above.
- **Renal function is not a covariate in either model,** so the Table 3 renal
  stratification is discussed rather than simulated; see the covariate-effects
  section.
- **Molar masses are external constants.** Cytarabine 243.22 g/mol and
  daunorubicin 527.52 g/mol are standard chemical values, used only in the
  descriptive molar-ratio figure and in no model parameter. The paper's own
  statement that 1 mg cytarabine plus 0.44 mg daunorubicin constitutes a 5:1
  molar ratio corroborates them.
- **The source reports two irreconcilable assay limits.** The Bioanalytical
  Methods section gives a range of detection of 1000 to 100 000 ng/mL, while
  Results reports samples below a limit of quantification of 5 ng/mL. Nothing
  in the models depends on which is correct; the discrepancy is recorded in
  each model's `population$notes`.
