---
title: "Polymyxin B (Wang 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polymyxin B (Wang 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wang_2020_polymyxinB"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wang_2020_polymyxinB"))$description`
* Article: [Front Pharmacol. 2020;11:829](https://doi.org/10.3389/fphar.2020.00829)
  (open access, PMC7289991)

# Population

Wang 2020 is a single-centre prospective study run at the First Affiliated
Hospital of Zhengzhou University (China) between April 2018 and November 2019.
Forty-six adults with documented multidrug-resistant Gram-negative bacterial
infections, receiving intravenous polymyxin B sulfate for at least 72 hours,
contributed 331 plasma concentrations. Patients on renal replacement therapy
were excluded. On day 4 of therapy a pre-dose sample (C0h) and five to seven
post-dose samples (mainly 0.5, 1, 1.5, 2, 4, 6 and 8 h) were drawn within a
single dosing interval. Polymyxin B1 and B2 were measured by LC-MS/MS and
summed, so the modelled concentration is total polymyxin B.

The cohort (Table 1) was 39 men and 7 women (15.2% female), median age 46
(range 18-94) years, median weight 70 (45-98) kg, and median Cockcroft-Gault
creatinine clearance 89.3 (15.6-315.2) mL/min. *Klebsiella pneumoniae* (20)
and *Acinetobacter baumannii* (19) were the main pathogens. Daily doses were
100, 150 or 200 mg (median 1.91 mg/kg/day), given q12h mostly as 1-hour
infusions (37 patients; 0.5 h in 2, 2 h in 7).

The same information is available programmatically via the model's
`population` metadata (`rxode2::rxode(readModelDb("Wang_2020_polymyxinB"))$population`).

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Two-compartment disposition, IV infusion | -- | Results, "Population PK Model": OFV 1062.13 (one-compartment) vs 694.63 (two-compartment); two-compartment with proportional error chosen as base model |
| Exponential IIV, `P_i = theta * exp(eta_i)` | Equ. 1 | Methods, "Population Pharmacokinetics Analysis" |
| `lvc` = log(6.218) | V 6.218 L | Table 2, row "tvV" (SE 0.83; bootstrap median 5.960, 95% CI 4.169-8.090); Equ. 4 |
| `lvp` = log(11.922) | V2 11.922 L | Table 2, row "tvV2" (SE 1.74; bootstrap median 12.073); Equ. 5 |
| `lcl` = log(1.786) | CL 1.786 L/h | Table 2, row "tvCl" (SE 0.12; bootstrap median 1.771); Equ. 6 |
| `lq` = log(13.518) | Q 13.518 L/h | Table 2, row "tvQ" (SE 3.35; bootstrap median 14.427); Equ. 7 |
| `e_crcl_cl` = 0.362 | CrCL exponent on CL | Table 2, row "dCldCrCL" (SE 0.09; bootstrap 95% CI 0.196-0.513); exponent of Equ. 6 |
| CrCL normalising constant 105.9 mL/min | -- | Equ. 6, `Cl = 1.786 x (CrCl/105.9)^0.362 x exp(eta_Cl)`; Results text: "105.9 ml/min was the median of CrCL" |
| `etalvc`, `etalcl`, `etalvp` variances | 0.318, 0.208, 0.690 | Table 2, rows "w2 V", "w2 Cl", "w2 V2" (footnote: variance of IIV) |
| V-Cl, V-V2, Cl-V2 correlations | 0.713, 0.667, 0.571 | Table 2, rows "CorrV-Cl", "CorrV-V2", "CorrCl-V2"; Results text (dOFV = 30.60 for the block) |
| `etalq` variance | 1.508 | Table 2, row "w2 Q" |
| `propSd` | 0.110 | Table 2, row "Residual variability (s) stdev0"; proportional model `Cobs = Cpred x (1 + eps)` (Methods) |

The block covariances in `ini()` are `corr * sqrt(var_i * var_j)` from the
printed variances and correlations; the resulting matrix is positive definite.

# Virtual cohort and dosing regimens

Table 3 and Figure 3 of the paper simulate three regimens -- 100 mg loading
then 50 mg q12h, 150 mg loading then 75 mg q12h, and 150 mg loading then
100 mg q12h -- infused at 50 mg/h, at three fixed creatinine clearances
(31.3, 105.9 and 315.2 mL/min). The loading dose is taken as the first dose
at time 0 and the maintenance doses start at 12 h. The paper simulated 1000
patients per scenario; 200 per scenario are used here.

```{r cohort-setup}
mod <- readModelDb("Wang_2020_polymyxinB")
n_per_arm <- 200L # cap: never more than 200 participants per arm

regimens <- tibble::tribble(
  ~regimen,        ~ld, ~md,
  "50 mg q12h",    100,  50,
  "75 mg q12h",    150,  75,
  "100 mg q12h",   150, 100
)
crcl_levels <- c(31.3, 105.9, 315.2)

arms <- tidyr::expand_grid(regimens, CRCL = crcl_levels) |>
  dplyr::mutate(
    arm = dplyr::row_number(),
    treatment = paste0(regimen, ", CrCL ", CRCL)
  )

subjects <- arms |>
  tidyr::expand_grid(rep = seq_len(n_per_arm)) |>
  dplyr::mutate(id = (arm - 1L) * n_per_arm + rep)

dose_times <- seq(0, 96, by = 12)
doses <- subjects |>
  tidyr::expand_grid(time = dose_times) |>
  dplyr::mutate(
    amt = ifelse(time == 0, ld, md),
    rate = 50, # Methods: 'The infusion rate was set as 50 mg/h'
    evid = 1L,
    cmt = "central"
  )
obs <- subjects |>
  tidyr::expand_grid(time = sort(unique(c(seq(0, 96, by = 0.25))))) |>
  dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L, cmt = "central")

events <- dplyr::bind_rows(doses, obs) |>
  dplyr::select(id, time, amt, rate, evid, cmt, CRCL) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

rxode2::rxSetSeed(20200605)
sim <- rxode2::rxSolve(mod, events, returnType = "data.frame") |>
  dplyr::left_join(
    dplyr::select(subjects, id, arm, regimen, treatment),
    by = "id"
  )
arm_key <- dplyr::distinct(arms, arm, regimen, CRCL, treatment)
sim <- sim |>
  dplyr::select(-dplyr::any_of("CRCL")) |>
  dplyr::left_join(dplyr::select(arm_key, arm, CRCL), by = "arm")
```

# Replicate Figure 3: median profiles by CrCL

Replicates Figure 3 of Wang 2020 (median simulated concentration-time profiles
for the three regimens in panels by CrCL).

```{r figure-3, fig.width = 8, fig.height = 4}
med_prof <- sim |>
  dplyr::group_by(regimen, CRCL, time) |>
  dplyr::summarise(Cc = stats::median(Cc), .groups = "drop") |>
  dplyr::mutate(
    regimen = factor(regimen, levels = regimens$regimen),
    panel = factor(paste0("CrCL ", CRCL, " mL/min"),
                   levels = paste0("CrCL ", crcl_levels, " mL/min"))
  )

ggplot2::ggplot(med_prof, ggplot2::aes(time, Cc, colour = regimen)) +
  ggplot2::geom_line() +
  ggplot2::facet_wrap(~panel) +
  ggplot2::labs(
    x = "Time (h)", y = "Median polymyxin B concentration (mg/L)",
    colour = "Maintenance dose",
    caption = "Replicates Figure 3 of Wang 2020."
  ) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

# PKNCA validation: day-4 AUC24 and Css,avg (Table 3)

Table 3 reports the day-4 AUC over 24 hours and the average steady-state
concentration. Day 4 is read as 72-96 h after the first (loading) dose. NCA is
run per subject on the individual predictions `Cc` (no residual error), grouped
by the regimen x CrCL scenario.

```{r pknca}
nca_conc <- sim |>
  dplyr::filter(!is.na(Cc), time >= 72, time <= 96) |>
  dplyr::select(id, time, Cc, treatment)
nca_dose <- doses |>
  dplyr::filter(time >= 72, time < 96) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | treatment + id,
                             doseu = "mg")
intervals <- data.frame(start = 72, end = 96, auclast = TRUE, cav = TRUE,
                        cmax = TRUE, cmin = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_long <- as.data.frame(nca_res) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES)
```

The reference values are the Table 3 medians (P50); `ncaComparisonTable()`
also aggregates the simulated subjects by the median, so the two sides are
the same statistic.

```{r table-3-comparison}
table3 <- tibble::tribble(
  ~regimen,      ~CRCL, ~auc_p5, ~auc_p50, ~auc_p95, ~cav_p5, ~cav_p50, ~cav_p95,
  "50 mg q12h",   31.3,  36.93,   87.20,  192.19,   1.54,   3.63,   8.01,
  "50 mg q12h",  105.9,  23.85,   53.33,  118.77,   0.99,   2.22,   4.95,
  "50 mg q12h",  315.2,  15.54,   37.92,   84.49,   0.65,   1.58,   3.52,
  "75 mg q12h",   31.3,  56.60,  128.58,  295.12,   2.36,   5.36,  12.30,
  "75 mg q12h",  105.9,  35.15,   82.04,  180.81,   1.46,   3.42,   7.53,
  "75 mg q12h",  315.2,  22.46,   53.60,  121.32,   0.94,   2.23,   5.06,
  "100 mg q12h",  31.3,  75.39,  167.90,  364.58,   3.14,   7.00,  15.19,
  "100 mg q12h", 105.9,  49.23,  108.57,  247.01,   2.05,   4.52,  10.29,
  "100 mg q12h", 315.2,  30.29,   71.91,  167.40,   1.26,   3.00,   6.98
) |>
  dplyr::mutate(treatment = paste0(regimen, ", CrCL ", CRCL))

reference <- table3 |>
  dplyr::transmute(treatment, auclast = auc_p50, cav = cav_p50)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_long,
  reference = reference,
  by = "treatment",
  params = c("auclast", "cav"),
  units = c(auclast = "mg*h/L", cav = "mg/L"),
  tolerance_pct = 20
)
cmp |>
  dplyr::rename("Scenario" = treatment) |>
  knitr::kable(caption = paste(
    "Simulated day-4 (72-96 h) AUC and average concentration versus the",
    "Wang 2020 Table 3 medians. AUC0-24 on day 4 is reported by PKNCA as",
    "AUClast over the 72-96 h interval."
  ))
if (!is.null(attr(cmp, "footnote"))) cat(attr(cmp, "footnote"), "\n")
```

The simulated medians reproduce the Table 3 medians to within a few percent
in every scenario, including the covariate direction (lower CrCL, higher
exposure) and its magnitude across the 10-fold CrCL range.

The 5th and 95th percentiles of Table 3 are compared as well. These are
tail quantiles of a 200-subject cohort, so they are shown for context and
are not asserted on.

```{r table-3-percentiles}
sim_q <- nca_long |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    sim_p5 = stats::quantile(PPORRES, 0.05),
    sim_p50 = stats::median(PPORRES),
    sim_p95 = stats::quantile(PPORRES, 0.95),
    .groups = "drop"
  ) |>
  dplyr::inner_join(table3, by = "treatment") |>
  dplyr::mutate(pct_diff_p50 = 100 * (sim_p50 - auc_p50) / auc_p50) |>
  dplyr::arrange(match(treatment, table3$treatment))

sim_q |>
  dplyr::select(treatment, auc_p5, sim_p5, auc_p50, sim_p50, auc_p95,
                sim_p95, pct_diff_p50) |>
  dplyr::rename(
    "Scenario" = treatment,
    "P5 paper" = auc_p5, "P5 sim" = sim_p5,
    "P50 paper" = auc_p50, "P50 sim" = sim_p50,
    "P95 paper" = auc_p95, "P95 sim" = sim_p95,
    "P50 % diff" = pct_diff_p50
  ) |>
  knitr::kable(digits = 1, caption = paste(
    "Day-4 AUC24 (mg*h/L) percentiles: Wang 2020 Table 3 versus simulation."
  ))

stopifnot(
  # Structural: a mis-transcribed clearance, exponent or normalising constant
  # moves every scenario by tens of percent.
  abs(stats::median(sim_q$pct_diff_p50)) < 5,
  # Envelope across the nine scenarios, robust to the sampling noise of a
  # 200-subject median (about 4% for the ~46% CV on CL).
  stats::quantile(abs(sim_q$pct_diff_p50), 0.9) < 12
)
```

## Typical-value closed-form check

For a linear two-compartment model at true steady state, the AUC over one
24-hour window equals the dose given in that window divided by CL,
independently of the distribution parameters. With random effects zeroed and
dosing continued for 20 days (the typical terminal half-life is well under
one day), the integrated AUC must reproduce `2 x MD / CL` at every CrCL. This
exercises the covariate term and the ODE wiring without any random draw, so a
tight bound is appropriate.

```{r closed-form}
mod_tv <- rxode2::zeroRe(mod)
tv_subj <- tibble::tibble(id = seq_along(crcl_levels), CRCL = crcl_levels)
tv_events <- dplyr::bind_rows(
  tv_subj |>
    tidyr::expand_grid(time = seq(0, 480 + 12, by = 12)) |>
    dplyr::mutate(amt = 50, rate = 50, evid = 1L, cmt = "central"),
  tv_subj |>
    tidyr::expand_grid(time = seq(480, 504, by = 0.05)) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L,
                  cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

tv_sim <- rxode2::rxSolve(mod_tv, tv_events, returnType = "data.frame")
tv_check <- tv_sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    CRCL = dplyr::first(CRCL),
    cl = dplyr::first(cl),
    auc = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    cl_expected = 1.786 * (CRCL / 105.9)^0.362,
    auc_expected = 2 * 50 / cl_expected,
    pct_diff = 100 * (auc - auc_expected) / auc_expected
  )

tv_check |>
  dplyr::select(CRCL, cl, cl_expected, auc, auc_expected, pct_diff) |>
  dplyr::rename(
    "CrCL (mL/min)" = CRCL, "CL model (L/h)" = cl,
    "CL Equ. 6 (L/h)" = cl_expected, "AUC24 integrated" = auc,
    "2 x 50 mg / CL" = auc_expected, "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 4, caption = "Typical-value steady-state AUC24 at 50 mg q12h.")

stopifnot(
  max(abs(tv_check$cl - tv_check$cl_expected)) < 1e-8,
  max(abs(tv_check$pct_diff)) < 0.1
)
```

# Observed-cohort AUC0-12

The paper reports an observed (non-compartmental) AUC0-12h of
43.64 +/- 27.68 mg*h/L (range 8.50-122.84) across the 46 patients. A rough
virtual version of that cohort is simulated below: maintenance doses of 50,
75 or 100 mg q12h in the Table 1 proportions (25 : 15 : 8 patients), 1-hour
infusions, and a creatinine clearance drawn log-normally around the Table 1
median of 89.3 mL/min and truncated to the observed 15.6-315.2 mL/min range.
Dosing runs for 4 days without a loading dose (the cohort loading practice is
not reported per patient) and the day-4 interval 72-84 h is analysed. This is
descriptive only -- the paper does not report the covariate distribution in
enough detail for a sharp comparison.

```{r observed-cohort}
n_cohort <- 200L
rxode2::rxSetSeed(46)
cohort <- tibble::tibble(
  id = seq_len(n_cohort),
  md = sample(c(50, 75, 100), n_cohort, replace = TRUE,
              prob = c(25, 15, 8)),
  CRCL = pmin(pmax(exp(stats::rnorm(n_cohort, log(89.3), 0.75)), 15.6), 315.2)
)
coh_events <- dplyr::bind_rows(
  cohort |>
    tidyr::expand_grid(time = seq(0, 84, by = 12)) |>
    dplyr::mutate(amt = md, rate = md, evid = 1L, cmt = "central"),
  cohort |>
    tidyr::expand_grid(time = seq(72, 84, by = 0.1)) |>
    dplyr::mutate(amt = NA_real_, rate = NA_real_, evid = 0L,
                  cmt = "central")
) |>
  dplyr::select(id, time, amt, rate, evid, cmt, CRCL) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()
coh_sim <- rxode2::rxSolve(mod, coh_events, returnType = "data.frame")
coh_auc <- coh_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    auc = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  )
tibble::tibble(
  Source = c("Wang 2020 observed (n = 46)", "Simulated virtual cohort"),
  `Mean AUC0-12 (mg*h/L)` = c(43.64, mean(coh_auc$auc)),
  `SD` = c(27.68, stats::sd(coh_auc$auc)),
  `Min` = c(8.50, min(coh_auc$auc)),
  `Max` = c(122.84, max(coh_auc$auc))
) |>
  knitr::kable(digits = 2, caption = "Day-4 AUC0-12: observed versus a virtual cohort.")
```

# Assumptions and deviations

* **CrCL normalising constant.** Equ. 6 normalises CrCL by 105.9 mL/min and
  the Results text calls 105.9 "the median of CrCL", whereas Table 1 reports
  the cohort CrCL as median 89.3 (range 15.6-315.2) mL/min. The Methods also
  use 105.9 as the "medium" value of the Table 3 simulations. The printed
  equation is followed (105.9). The typical clearance at the Table 1 median
  of 89.3 mL/min is then 1.68 L/h rather than 1.786 L/h. The source of the
  discrepancy (e.g. median over samples rather than patients, or over a
  subset) is not stated.
* **Covariate units.** CrCL is the raw Cockcroft-Gault estimate in mL/min
  (not normalised to 1.73 m^2) and is stored in the canonical `CRCL` column
  with mL/min units, as in other raw Cockcroft-Gault models in the library.
* **Dose units.** The Methods quote the Chinese label as "50 to 100 million
  units (1 million units equal to 1 mg) twice daily"; the conventional
  polymyxin B conversion is 10,000 units per mg, so the stated unit numbers
  appear to be a translation slip. All doses in the paper's tables and
  simulations are in mg, and the model is dosed in mg.
* **Table 1 dose counts.** The daily-dose counts (25 + 15 + 8 = 48) exceed the
  46 patients and "25 (50.0%)" does not match 25/46; the proportions are used
  only for the descriptive virtual cohort.
* **Day-4 window for Table 3.** "AUC24h ... on day four" is read as 72-96 h
  after the loading dose, with maintenance doses starting at 12 h. The paper
  does not state the dosing clock explicitly.
* **Parameter classification.** The Table 2 footnote calls dCldCrCL a
  "fixed parameter coefficient"; in Phoenix NLME terminology this denotes a
  fixed effect (theta), and the parameter has an SE and bootstrap CI, so it is
  encoded as estimated.
* **Limited sampling strategies.** The Bayesian and multiple-linear-regression
  limited sampling equations for AUC0-12h (Tables 4-5) are TDM estimation
  tools derived from the observed data, not part of the population model, and
  are not encoded.
* **No errata** were found for this article (literature check, September 2026).
  The supplement (Supplementary Figure 1, CrCL versus individual CL) contains
  no parameter values.
