---
title: "Ciprofol (Wang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ciprofol (Wang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wang_2026_ciprofol"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wang_2026_ciprofol"))$description`
* Article: <https://doi.org/10.1186/s12871-026-03647-9> (BMC Anesthesiology 26(1), open access; PMCID PMC12930602)
* Trial registration: Chinese Clinical Trial Registry [ChiCTR2200058405](https://www.chictr.org.cn/)

No supplementary material accompanies this article (EuropePMC reports
`hasSuppl = "N"` for an open-access record, so the field is informative), and no
erratum or correction notice was found.

# Population

Wang 2026 enrolled 27 Chinese children aged 1 to 9 years with American Society
of Anesthesiologists physical status I or II, scheduled for elective urologic
surgery of anticipated duration greater than 2 h, at the Second Affiliated
Hospital and Yuying Children's Hospital of Wenzhou Medical University between
January and August 2023. Enrolment was balanced across three age strata
(toddlers 1-2 years, preschoolers 3-5 years, school-age children 6-9 years,
n = 9 each). One child withdrew when the operation was shortened and one was
excluded for a blocked arterial catheter, leaving 25 children (14 male,
11 female) in strata of 9 / 9 / 7. Inclusion required a body mass index between
the 25th and 75th percentile for age and sex, so the cohort is deliberately
non-obese.

Each child received a single 0.6 mg/kg intravenous bolus of ciprofol over 30 s
at induction, after midazolam 0.1-0.2 mg/kg premedication and together with
fentanyl 2.0 ug/kg. Thirteen arterial samples were planned per patient (a
pre-dose blank plus 2, 4, 6, 8, 10, 20, 30, 45, 60, 90, 120 and 180 min after
injection); 8 of 325 samples were not collected, mostly the 180 min
late-elimination sample, leaving 317 samples for the analysis. Ciprofol was
quantified by UPLC-APCI-MS/MS over 5-20000 ng/mL with an LLOQ of 5 ng/mL, and
no concentration fell below the limit of quantitation. Baseline demographics
and laboratory values are in Table 1 of the source: mean age 4.2 years
(range 1.0-9.0), mean weight 18.5 kg (range 9.8-37), mean BMI 16.7 kg/m^2,
mean BUN 5.2 mmol/L (range 2.9-7.6).

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Wang_2026_ciprofol")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Wang_2026_ciprofol.R`. The table below
collects them in one place for review. The paper reports every disposition
parameter per kilogram of body weight and in millilitres, so each `ini()` value
is the printed number divided by 1000 (mL to L); body weight is reapplied in
`model()` with a linear exponent.

| Equation / parameter | Paper value | `ini()` value | Source location |
|---|---|---|---|
| `lcl` (CL) | 31.2 mL/min/kg (RSE 5.0%) | `log(0.0312)` | Table 3, row `CL, ml/min/kg` |
| `lvc` (V1) | 506 mL/kg (RSE 8.1%) | `log(0.506)` | Table 3, row `V1, ml/kg` |
| `lq` (Q2) | 28.2 mL/min/kg (RSE 19.9%) | `log(0.0282)` | Table 3, row `Q2, ml/min/kg` |
| `lvp` (V2) | 231 mL/kg (RSE 14.7%) | `log(0.231)` | Table 3, row `V2, ml/kg` |
| `lq2` (Q3) | 19.8 mL/min/kg (RSE 10.8%) | `log(0.0198)` | Table 3, row `Q3, ml/min/kg` |
| `lvp2` (V3) | 1360 mL/kg (RSE 12.4%) | `log(1.36)` | Table 3, row `V3, ml/kg` |
| `e_bun_vc` | -0.821 (RSE 28.6%) | `-0.821` | Table 3, row `Covariate thetaBUN on V1` |
| BUN normalising median | not printed; recovered as 5 mmol/L | `(BUN / 5)` | Results, "Population covariant analysis" (V1 = 0.770 and 0.384 L/kg at BUN 3 and 7); Methods Eq. 5 |
| `etalcl` | eta(CL) 26.3% (RSE 9.5%, shrinkage 0.3%) | `0.263^2 = 0.069169` | Table 3, `Inter-individual variability` block |
| `etalvc` | eta(V1) 35.1% (RSE 11.1%, shrinkage 5.8%) | `0.351^2 = 0.123201` | Table 3, `Inter-individual variability` block |
| cov(`etalcl`, `etalvc`) | Corr_CL & V1 = -0.821 | `-0.821 * 0.263 * 0.351 = -0.075789` | Table 3, row `Corr_CL & V1`; Results, "Population PK model" |
| `etalq2` | eta(Q3) 38.3% (RSE 15.5%, shrinkage 7.8%) | `0.383^2 = 0.146689` | Table 3, `Inter-individual variability` block |
| `propSd` | 13.5% (RSE 9.6%) | `0.135` | Table 3, row `epsilon_prop` |
| `addSd` | 4.83 ug/L (RSE 28.8%) | `0.00483` (mg/L) | Table 3, row `epsilon_add` |
| Three-compartment disposition, IV | n/a | `d/dt(central)`, `d/dt(peripheral1)`, `d/dt(peripheral2)` | Results, "Population PK model" (dOFV = 65.8 vs two-compartment, p < 0.001) |
| Linear body-weight scaling of all disposition parameters | n/a | `* WT` on `cl`, `vc`, `q`, `vp`, `q2`, `vp2` | Table 3 per-kilogram units; Methods, "Model selection and covariate analysis" (allometric scaling and maturation tested, not retained) |
| Log-normal IIV | n/a | `exp(lX + etalX)` | Methods, "Structural model" (Eq. 1) |
| Combined residual error | n/a | `Cc ~ add(addSd) + prop(propSd)` | Methods, Eq. 4 |

# Virtual cohort

Original observed data are not publicly available. The cohort below reproduces
the three enrolment strata of Table 1, drawing weight, BUN and age per stratum
from normal distributions centred on the published stratum mean with a standard
deviation of (max - min) / 4 and truncated to the published range. 100 subjects
per stratum (the skill cap is 200 per arm) is ample for the checks below.

Each subject receives the study regimen: 0.6 mg/kg as a 30 s infusion into
`central`, with observations on the ODE state `central` at the paper's own
sampling times. Using the published sparse grid rather than a dense one matters
for the NCA comparison: the trapezoidal bias and the terminal-slope window then
match those of the paper's own non-compartmental analysis.

```{r cohort}
# set.seed() seeds R's RNG (the covariate draws); rxode2's simulation RNG is
# partitioned per solver thread, so the realised cohort differs between a
# 2-thread CI runner and a 16-thread workstation. Every assertion below is
# written to hold for any cohort this model can produce.
set.seed(20260912)
rxode2::rxSetSeed(20260912)

SAMPLE_TIMES <- c(2, 4, 6, 8, 10, 20, 30, 45, 60, 90, 120, 180)
LLOQ <- 0.005  # 5 ng/mL = 5 ug/L = 0.005 mg/L (Methods, assay validation)

rtnorm <- function(n, mean, lo, hi) pmin(pmax(rnorm(n, mean, (hi - lo) / 4), lo), hi)

make_cohort <- function(n, label, wt, bun, age, id_offset = 0L) {
  subj <- tibble(
    id      = id_offset + seq_len(n),
    stratum = label,
    WT      = rtnorm(n, wt[1],  wt[2],  wt[3]),
    BUN     = rtnorm(n, bun[1], bun[2], bun[3]),
    AGE     = rtnorm(n, age[1], age[2], age[3])
  )
  bind_rows(
    # 0.6 mg/kg given over 30 s (0.5 min) into the central compartment
    subj |> mutate(time = 0, amt = 0.6 * WT, evid = 1L, dur = 0.5, cmt = "central"),
    # Observations on the ODE state `central` -- never on the observable `Cc`.
    # time = 0 is the paper's pre-dose blank sample and anchors PKNCA's AUC.
    tidyr::crossing(subj, time = c(0, SAMPLE_TIMES)) |>
      mutate(amt = NA_real_, evid = 0L, dur = NA_real_, cmt = "central")
  ) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(
  make_cohort(100, "Toddlers (1-2 y)",     c(12.8,  9.8, 16.5), c(5.7, 4.0, 6.8), c(1.8, 1, 2),   0L),
  make_cohort(100, "Preschoolers (3-5 y)", c(18.3, 15.0, 23.5), c(5.1, 2.9, 7.6), c(3.9, 3, 5), 100L),
  make_cohort(100, "School-age (6-9 y)",   c(25.9, 19.5, 37.0), c(4.7, 3.8, 6.2), c(7.7, 6, 9), 200L)
)

# Disjoint ids across the three cohorts (duplicate ids silently merge in rxSolve)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(nrow(distinct(events, id)) == 300L)
```

# Simulation

```{r simulate}
mod <- readModelDb("Wang_2026_ciprofol")

sim <- rxode2::rxSolve(mod, events = events, keep = c("stratum", "WT", "BUN")) |>
  as.data.frame()

# Cc is the individual prediction (mg/L); `sim` carries the residual error and
# is the analogue of an observed concentration. The paper reports ug/L.
sim <- sim |>
  mutate(
    Cc_ugL  = Cc * 1000,
    obs_ugL = pmax(sim, LLOQ / 2) * 1000  # see "Assumptions and deviations"
  )
```

# Replicate published figures

```{r figure-2, fig.cap = "Replicates Figure 2 of Wang 2026: ciprofol concentration-time profiles by age stratum."}
sim |>
  filter(!is.na(Cc), time > 0) |>
  group_by(stratum, time) |>
  summarise(
    Q025 = quantile(obs_ugL, 0.025),
    Q50  = quantile(obs_ugL, 0.500),
    Q975 = quantile(obs_ugL, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50, colour = stratum, fill = stratum)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time after ciprofol injection (min)",
    y = "Ciprofol plasma concentration (ug/L)",
    colour = NULL, fill = NULL,
    title = "Figure 2 -- concentration-time profiles by age stratum",
    caption = "Median and 2.5th-97.5th percentiles of 100 simulated subjects per stratum."
  ) +
  theme(legend.position = "bottom")
```

Wang 2026 reports that "the plasma concentration-time curve of ciprofol was
similar across the three groups" and that Table 2 showed no significant
difference between strata (all p > 0.15). The simulated profiles overlap for
the same reason the published ones do: after per-kilogram dosing, weight and
age cancel out of the model entirely, and the only covariate that moves a
prediction is BUN.

# Typical-value replication of Table 4

Table 4 of Wang 2026 reports the paper's own simulation of a typical child
given 0.6 mg/kg over 30 s at three BUN levels. This is a deterministic,
typical-value prediction, so it is reproduced with `zeroRe()` and gated
tightly. It is the sharpest available check on the covariate model, because
the central volume at BUN 3 and 7 mmol/L is quoted in the Results text and
pins both the exponent and the (unprinted) normalising median.

```{r table-4}
mod_typical <- rxode2::zeroRe(mod)
WT_TYP <- 18.5  # cohort mean weight, Table 1; predictions are weight-invariant

typical_profile <- function(bun) {
  ev <- rxode2::et(amt = 0.6 * WT_TYP, dur = 0.5, cmt = "central") |>
    rxode2::et(seq(0, 180, by = 0.05), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- WT_TYP
  d$BUN <- bun
  rxode2::rxSolve(mod_typical, d, returnType = "data.frame") |>
    filter(!is.na(Cc)) |>
    mutate(BUN = bun, Cc_ugL = Cc * 1000)
}

profiles <- bind_rows(lapply(c(3, 5, 7), typical_profile))

at_time <- function(df, t) df$Cc_ugL[which.min(abs(df$time - t))]

table4 <- profiles |>
  group_by(BUN) |>
  summarise(
    C2min   = at_time(pick(everything()), 2),
    C180min = at_time(pick(everything()), 180),
    # AUC(0-inf) = trapezoid to 180 min + Clast / lambda_z from the terminal slope
    AUC = {
      d <- pick(everything())
      ug <- d$Cc_ugL
      auc180 <- sum(diff(d$time) * (head(ug, -1) + tail(ug, -1)) / 2)
      keep <- d$time >= 120
      lz <- -stats::coef(stats::lm(log(ug[keep]) ~ d$time[keep]))[[2]]
      auc180 + tail(ug, 1) / lz
    },
    V1_L_per_kg = 0.506 * (first(BUN) / 5)^(-0.821),
    .groups = "drop"
  )

published4 <- tibble::tribble(
  ~BUN, ~C2min_pub, ~C180min_pub, ~AUC_pub, ~V1_pub,
     3,      640.7,         19.6,    19231,   0.770,
     5,      880.3,         17.9,    19231,      NA,
     7,     1056.1,         17.1,    19231,   0.384
)

cmp4 <- table4 |>
  left_join(published4, by = "BUN") |>
  mutate(
    `C2min % diff`   = 100 * (C2min - C2min_pub) / C2min_pub,
    `C180min % diff` = 100 * (C180min - C180min_pub) / C180min_pub,
    `AUC % diff`     = 100 * (AUC - AUC_pub) / AUC_pub,
    `V1 % diff`      = 100 * (V1_L_per_kg - V1_pub) / V1_pub
  )

cmp4 |>
  transmute(
    `BUN (mmol/L)`            = BUN,
    `V1 simulated (L/kg)`     = round(V1_L_per_kg, 3),
    `V1 published (L/kg)`     = V1_pub,
    `C2min simulated (ug/L)`  = round(C2min, 1),
    `C2min published (ug/L)`  = C2min_pub,
    `C180min simulated (ug/L)` = round(C180min, 1),
    `C180min published (ug/L)` = C180min_pub,
    `AUC simulated (ug*min/L)` = round(AUC, 0),
    `AUC published (ug*min/L)` = AUC_pub
  ) |>
  knitr::kable(
    caption = "Typical-value replication of Wang 2026 Table 4 (0.6 mg/kg over 30 s) and of the two central volumes quoted in the Results text."
  )
```

```{r table-4-gate}
# Closed form: with a covariate-free clearance, AUC(0-inf) is exactly Dose / CL
# and is invariant to BUN -- which is why Table 4 prints the same AUC on all
# three rows. This is an analytic identity, so it is gated tightly.
auc_closed_form <- 0.6 / 0.0312 * 1000  # ug*min/L
stopifnot(abs(auc_closed_form - 19231) / 19231 < 0.001)

# Deterministic (zeroRe, fixed covariates) -- no cohort randomness, so tight
# bounds are correct here and will catch a mis-transcribed parameter.
stopifnot(
  # Recovers the two published central volumes exactly, confirming BOTH the
  # -0.821 exponent and the normalising median of 5 mmol/L.
  max(abs(cmp4$`V1 % diff`), na.rm = TRUE) < 0.5,
  # Numerical AUC vs the paper's; residual gap is trapezoid + extrapolation.
  max(abs(cmp4$`AUC % diff`)) < 2,
  max(abs(cmp4$`C180min % diff`)) < 2,
  # C2min sits between our t = 2.0 min and t = 2.5 min values: the paper does
  # not state whether "2 min" runs from the start or the end of the 30 s
  # injection, and the curve falls ~7% per 0.1 min there. Realised 2.4-4.8%.
  max(abs(cmp4$`C2min % diff`)) < 6
)
```

# PKNCA validation

The paper's Table 2 reports non-compartmental parameters computed from the
observed concentrations of each child. The simulation is therefore taken
through the same route: concentrations carrying residual error, on the
published sampling grid, with the pre-dose blank as the time-zero anchor.

```{r pknca}
# Concentrations in the paper's units (ug/L). Filter on !is.na() ONLY -- a
# `time > 0` or `Cc > 0` filter would drop the time-zero anchor.
nca_conc <- sim |>
  filter(!is.na(Cc)) |>
  mutate(Cc = ifelse(time == 0, 0, obs_ugL)) |>   # pre-dose blank is exactly 0
  select(id, time, Cc, stratum)

# Duplicate every subject into a pooled "All" group so the per-stratum and the
# pooled comparison land in a single table (paper Table 2 reports both).
nca_conc <- bind_rows(
  nca_conc,
  nca_conc |> mutate(id = id + 1000L, stratum = "All (n = 25)")
)

nca_dose <- events |>
  filter(evid == 1) |>
  transmute(id, time, amt = amt * 1000, stratum)  # mg -> ug
nca_dose <- bind_rows(
  nca_dose,
  nca_dose |> mutate(id = id + 1000L, stratum = "All (n = 25)")
)

# Time-zero guarantee (already present above; kept as a defensive no-op).
nca_conc <- bind_rows(
  nca_conc,
  nca_conc |> distinct(id, stratum) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, stratum, time, .keep_all = TRUE) |>
  arrange(id, stratum, time)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | stratum + id,
                             concu = "ng/mL", timeu = "min")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | stratum + id, doseu = "ug")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, aucinf.obs = TRUE,
  half.life = TRUE, cl.obs = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)

# cl.obs comes back as dose / AUC in L/min; the paper reports mL/min/kg.
wt_by_id <- distinct(bind_rows(
  events |> filter(evid == 1) |> select(id, WT),
  events |> filter(evid == 1) |> transmute(id = id + 1000L, WT)
))

nca_tbl <- as.data.frame(nca_res$result) |>
  left_join(wt_by_id, by = "id") |>
  mutate(PPORRES = ifelse(PPTESTCD == "cl.obs", PPORRES / WT * 1000, PPORRES)) |>
  select(-WT)

stopifnot(nrow(nca_tbl) > 0)
```

## Comparison against published NCA

```{r nca-compare}
published <- tibble::tribble(
  ~stratum,                ~cmax,  ~tmax, ~auclast, ~aucinf.obs, ~half.life, ~cl.obs,
  "Toddlers (1-2 y)",      976.3,      2,    15248,       16401,       54.3,    39.1,
  "Preschoolers (3-5 y)", 1048.2,      2,    17029,       19424,       64.4,    31.8,
  "School-age (6-9 y)",    881.2,      2,    15966,       17493,       57.6,    35.4,
  "All (n = 25)",          975.5,      2,    16090,       17795,       58.8,    35.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_tbl,
  reference = published,
  by        = "stratum",
  units     = c(cmax = "ug/L", tmax = "min", auclast = "ug*min/L",
                aucinf.obs = "ug*min/L", half.life = "min", cl.obs = "mL/min/kg"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published non-compartmental parameters (Wang 2026 Table 2). * differs from the reference by more than 20%.",
  digits  = 1
)
```

```{r nca-gate}
# Recompute the pooled comparison from the PKNCA result frame rather than
# parsing the formatted display table. ncaComparisonTable() aggregates by
# median, so the gate does too and the two agree by construction.
GATED <- c("cmax", "tmax", "auclast", "aucinf.obs", "half.life", "cl.obs")

pooled <- nca_tbl |>
  filter(stratum == "All (n = 25)", PPTESTCD %in% GATED) |>
  group_by(PPTESTCD) |>
  summarise(Simulated = median(PPORRES, na.rm = TRUE),
            n_subj = sum(!is.na(PPORRES)), .groups = "drop") |>
  left_join(
    published |>
      filter(stratum == "All (n = 25)") |>
      select(-stratum) |>
      tidyr::pivot_longer(everything(), names_to = "PPTESTCD", values_to = "Reference"),
    by = "PPTESTCD"
  ) |>
  mutate(`% diff` = 100 * (Simulated - Reference) / Reference)

# A lookup that matches no rows returns numeric(0), and all(logical(0)) is
# TRUE -- so confirm every gated parameter actually produced results before
# asserting on them.
stopifnot(
  nrow(pooled) == length(GATED),
  setequal(pooled$PPTESTCD, GATED),
  !anyNA(pooled$Simulated), !anyNA(pooled$Reference),
  all(pooled$n_subj > 250)
)

knitr::kable(pooled, digits = 2,
             caption = "Pooled (n = 300 simulated) values used by the gate below.")

get_pct <- function(code) {
  v <- pooled$`% diff`[pooled$PPTESTCD == code]
  if (length(v) != 1L || is.na(v)) stop("no unique pooled row for ", code)
  v
}

# Gate on the pooled n = 25 column: it is the best-determined reference in
# Table 2. The per-stratum columns rest on n = 7-9 subjects, so their means
# carry roughly +/- 30% sampling noise and gating them would be a coin flip --
# they are shown above for information, not asserted on.
#
# Realised on the authoring run (300 subjects, pooled medians): aucinf.obs
# -0.03%, auclast +0.75%, half.life -1.24%, cl.obs -4.99%, cmax -7.93%. The
# per-stratum medians scatter over roughly -15% to +10%, but that spread is
# dominated by the n = 7-9 published references rather than by the simulation,
# and the pooled median over 300 subjects is far steadier. A mis-transcribed
# clearance, volume, dose or unit moves these by tens of percent, so a 20%
# bound retains plenty of headroom and still goes red on a real error.
stopifnot(
  abs(get_pct("cmax"))       < 20,
  abs(get_pct("auclast"))    < 20,
  abs(get_pct("aucinf.obs")) < 20,
  abs(get_pct("cl.obs"))     < 20,
  abs(get_pct("half.life"))  < 20
)

# The paper states Cmax was reached 2 min after injection (the first sample).
tmax_all <- nca_tbl |> filter(PPTESTCD == "tmax", stratum == "All (n = 25)")
stopifnot(mean(tmax_all$PPORRES == 2) > 0.5)

# Residual-error draws can push a late concentration below zero; LLOQ
# censoring must leave the NCA free of non-finite exposure values. Cmax and
# AUC0-last do not depend on the terminal fit, so every subject must yield a
# finite positive value -- that is the check that catches a censoring bug.
obs_exposure <- nca_tbl |>
  filter(stratum != "All (n = 25)", PPTESTCD %in% c("cmax", "auclast"))
stopifnot(
  nrow(obs_exposure) == 600L,
  !anyNA(obs_exposure$PPORRES),
  all(obs_exposure$PPORRES > 0)
)

# AUC0-inf additionally needs a lambda_z fit, which PKNCA can legitimately
# fail to obtain for the occasional subject whose 120-180 min points are
# non-monotone after residual error. Bound the rate rather than forbidding it
# (realised 1 of 300 subjects, 0.3%, on the authoring run).
aucinf_fail <- nca_tbl |>
  filter(stratum != "All (n = 25)", PPTESTCD == "aucinf.obs")
stopifnot(
  nrow(aucinf_fail) == 300L,
  mean(is.na(aucinf_fail$PPORRES)) < 0.05,
  all(aucinf_fail$PPORRES > 0, na.rm = TRUE)
)
```

The simulated pooled exposures reproduce Table 2 closely: AUC0-inf agrees to
0.03%, AUC0-last to 0.8%, terminal half-life to 1.2%, Tmax exactly, apparent
clearance to 5.0% and Cmax to 7.9%. No row in the table -- pooled or
per-stratum -- exceeds the 20% flag. Per-stratum rows scatter more widely
(-15% to +10%), which is what n = 7-9 subjects per published stratum predicts;
the model itself makes no distinction between the strata, since weight and age
cancel out after per-kilogram dosing.

Two of the remaining gaps have identifiable causes rather than being residual
noise. Cmax reads about 8% low because the first sample is at 2 min while the
true model maximum occurs at the end of the 30 s infusion, and the published
mean is an average over 25 children whereas the table compares medians. The
non-compartmental clearance (33.7 mL/min/kg here, 35.5 in the paper) is higher
than the model's structural CL of 31.2 mL/min/kg because a 180 min sampling
window truncates the slow third compartment, so the log-linear extrapolation
understates AUC0-inf and the derived clearance is biased upward. The paper
carries the same discrepancy between its own Table 2 and Table 3.

# Assumptions and deviations

* **BUN normalising median recovered, not printed.** Wang 2026 Eq. 5 divides a
  continuous covariate by its median, but the paper never prints the median
  BUN; Table 1 reports the cohort *mean* as 5.2 mmol/L. The value is recovered
  exactly from the two central volumes the Results text does print:
  `506 * (3/ref)^(-0.821) = 770` gives `ref = 5.003`, and
  `506 * (7/ref)^(-0.821) = 384` gives `ref = 5.002`. Table 4 independently
  simulates at BUN 3, 5 and 7 mmol/L with 5 as the central level. The model
  uses 5 mmol/L, and the Table 4 replication above confirms both published
  volumes to better than 0.5%.
* **Two parameters share the value -0.821.** Table 3 prints -0.821 for both the
  BUN-on-V1 exponent and the CL-V1 IIV correlation. This is a genuine
  coincidence rather than a duplicated cell: the exponent is independently
  confirmed by the printed volume ratio (0.770 / 0.384 = 2.005 = (3/7)^-0.821),
  and the correlation is independently stated in the Results text ("a decrease
  in OFV by 15.039, with a correlation coefficient of -0.821 for the IIV of CL
  and V1").
* **IIV percentages read as log-scale standard deviations.** Table 3 reports
  IIV under a "%" heading and uses the same "%" convention for the
  residual-error rows, where a percentage can only be a standard deviation. The
  variances encoded are therefore `omega^2 = (percent/100)^2`. Reading the
  percentages instead as the approximate CV of a log-normal
  (`omega = sqrt(log(1 + CV^2))`) changes `omega` by under 2% relative at these
  magnitudes, so the choice is not load-bearing for any result here.
* **Additive residual error carries concentration units, not percent.** Table 3
  prints `epsilon_add` as 4.83 under a "%" column header carried down from the
  proportional row above it. An additive residual must have concentration
  units; 4.83 ug/L sits essentially at the assay LLOQ of 5 ng/mL, the magnitude
  an additive term takes when set by assay noise at the bottom of the
  calibration range. It is encoded as 0.00483 mg/L.
* **Body weight enters structurally, not as a fitted covariate effect.** Every
  Table 3 disposition parameter is per kilogram, so weight multiplies all six
  with a linear exponent and no reference weight. Concentrations are therefore
  invariant to weight given per-kilogram dosing. The paper explicitly tested
  standard allometric scaling and established age-dependent maturation
  functions on clearance and did not retain either.
* **Screened-but-unretained covariates are documented, not modelled.** Age,
  sex, BMI, ALT, AST, total bilirubin, albumin, creatinine, haemoglobin and
  total protein were screened and rejected; they are recorded in the model
  file's `covariatesDataExcluded` list so their provenance survives without
  appearing in `model()`.
* **Covariate distributions are assumed.** Table 1 reports mean, minimum and
  maximum per stratum but no distributional shape or correlation structure.
  Weight, BUN and age are drawn independently from truncated normals centred on
  the stratum mean with SD = (max - min) / 4. Only BUN affects a prediction.
* **LLOQ censoring.** The paper reports that no observed concentration fell
  below the 5 ng/mL LLOQ. Simulated residual-error draws at the 180 min tail do
  occasionally go below zero (about 0.5% of observations on the authoring run),
  which would make `aucinf.obs` non-finite. Simulated concentrations are
  therefore censored at LLOQ/2 = 2.5 ug/L before the NCA, and the gate asserts
  the resulting exposures are finite and positive.
* **The 2 min sampling convention is ambiguous.** The paper does not state
  whether "2 min after injection" runs from the start or the end of the 30 s
  bolus. Measuring from the start gives C2min 2.4-4.8% above Table 4;
  measuring from the end gives 2.3-4.5% below it. The published values sit
  between the two. The vignette measures from the start of the infusion, which
  is the rxode2 convention, and the gate is set at 6% to accommodate the
  ambiguity.
* **Terminal half-life agrees at the median but not in spread.** PKNCA selects
  the lambda_z window automatically; the paper regressed "the log-linear phase"
  without specifying a window. The pooled median half-life matches the
  published 58.8 min to 1.2%, but the simulated distribution is right-skewed
  (SD about 36 min against the published 16.8 min, mean about 65 min) because
  automatic window selection occasionally lands on a short, poorly-conditioned
  terminal segment. The comparison table and its gate both aggregate by median,
  which is why this shows as agreement; a mean-based comparison would read
  about 11% high. For one simulated subject in 300 the lambda_z fit failed
  outright, which propagates to `aucinf.obs`, `half.life` and `cl.obs`; the
  gate bounds that failure rate at 5% rather than forbidding it.
* **No supplement and no erratum.** EuropePMC reports `hasSuppl = "N"` on an
  open-access record, and no correction notice was found. Every value above
  comes from the main article's text, Table 1, Table 2, Table 3 or Table 4.
