---
title: "Methotrexate popPK models in pediatric ALL (Wang 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methotrexate popPK models in pediatric ALL (Wang 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Wang 2023 is an **external-evaluation** paper: its authors developed no model of
their own. Instead they re-implemented six previously published pediatric
acute-lymphoid-leukaemia (ALL) methotrexate population-PK models in NONMEM and
tested all six against one independent cohort. Table 2 of that paper tabulates
the full structural, covariate, between-subject-variability and residual-error
specification of each model -- complete enough to recompile them -- so the six
models, not the evaluation, are what is packaged here.

**Five of the six are packaged in this release.** The sixth (Zhang 2010) is
deferred; see [Assumptions and deviations](#assumptions-and-deviations).

* Article: <https://doi.org/10.3390/pharmaceutics15020569>

```{r model-list, echo = FALSE}
mtx_models <- c(
  "Aumente_2006_methotrexate",
  "Gao_2021_methotrexate",
  "Hui_2019_methotrexate",
  "MedellinGaribay_2020_methotrexate",
  "Jonsson_2011_methotrexate"
)

data.frame(
  Model = mtx_models,
  Structure = c("2-compartment", "3-compartment", "2-compartment",
                "2-compartment", "2-compartment"),
  Covariates = c(
    "Weight; age (10-year stratum)",
    "Weight (fixed allometry); serum creatinine",
    "BSA; eGFR; age; occasion (IOV)",
    "BSA; weight",
    "Weight only"
  ),
  check.names = FALSE
) |>
  knitr::kable(caption = "The five packaged models from Wang 2023 Table 2.")
```

Each model carries its own citation and description:

```{r model-refs, echo = FALSE, results = "asis"}
for (nm in mtx_models) {
  ui <- rxode2::rxode(readModelDb(nm))
  cat("* **", nm, "** -- ", ui$reference, "\n\n", sep = "")
}
```

# Population

The six development cohorts and the independent evaluation cohort are
tabulated side by side in Wang 2023 Table 1. All are children with ALL
receiving high-dose intravenous methotrexate; the mean age across studies was
5.0-7.5 years. Three models were built on Chinese children (Gao, Hui, Zhang),
the others on Spanish (Aumente), American/Nordic (Jonsson) and Mexican
(Medellin-Garibay) children. Cohort sizes range from 36 (Hui) to 311 (Gao)
patients.

The **evaluation cohort** -- the dataset all six models were tested against,
and the cohort reproduced in this vignette -- was collected retrospectively at
the Third Xiangya Hospital, Central South University, 2019-2022
(ChiCTR2000035264): 51 children contributing 354 methotrexate concentrations,
median age 5.5 years (range 1.0-13.0), median weight 19.0 kg (9.5-62.0),
median BSA 0.77 m^2 (0.41-1.60), 40 male / 11 female, median serum creatinine
0.31 mg/dL (0.16-0.71) and median eGFR 149.9 mL/min/1.73 m^2 -- i.e. uniformly
normal renal function. Prescribed doses ranged from 1.0 to 5.0 g/m^2.

Population metadata for any packaged model is available programmatically:

```{r population-metadata}
str(readModelDb("Gao_2021_methotrexate")()$population, max.level = 1)
```

# Source trace

Every packaged value traces to Wang 2023 Table 2 unless the row says otherwise.
The one systematic exception is Gao 2021, whose primary is open access
(PMC8313761) and which is therefore encoded from the primary -- Wang's Table 2
mis-transcribes that row twice (see Errata).

```{r source-trace, echo = FALSE}
data.frame(
  Model = c(rep("Aumente 2006", 4), rep("Gao 2021", 4), rep("Hui 2019", 4),
            rep("Medellin-Garibay 2020", 3), rep("Jonsson 2011", 3)),
  Quantity = c(
    "K12, K21 (micro-constants)", "CL (two age strata)", "V1 (two age strata)",
    "IIV on K12/K21/CL/V1; prop + add RUV",
    "CL, V1, V2, Q1, V3, Q2", "Allometric exponents 0.75 / 1.0 (FIXED)",
    "Serum creatinine on CL (-0.97%/umol/L)", "IIV on CL and Q1; log-additive RUV",
    "CL (BSA and eGFR power terms)", "V1 (BSA power term), Q (age power term), V2",
    "IIV on CL and V2; IOV on CL", "Proportional RUV",
    "CL = 6.5 x BSA^0.62; V1 = 0.36 x Weight", "Q = 0.41; V2 = 3.2",
    "IIV on CL/V1/V2; proportional RUV",
    "CL, V1, Q, V2 (all linear in weight)", "IIV on all four parameters",
    "Proportional RUV = 30% (ASSUMED)"
  ),
  Source = c(
    "Wang 2023 Table 2, Aumente row", "Wang 2023 Table 2, Aumente row",
    "Wang 2023 Table 2, Aumente row", "Wang 2023 Table 2, IIV and RUV columns",
    "Gao 2021 Table 2 (primary)", "Gao 2021 Table 2 footnote",
    "Gao 2021 Table 2 + Abstract", "Gao 2021 Table 2 'CV for IIV' column",
    "Wang 2023 Table 2, Hui row", "Wang 2023 Table 2, Hui row",
    "Wang 2023 Table 2, 'IIV (%) (IOV (%))' column", "Wang 2023 Table 2, RUV column",
    "Wang 2023 Table 2 + primary abstract", "Wang 2023 Table 2 + primary abstract",
    "Wang 2023 Table 2, IIV and RUV columns",
    "Wang 2023 Table 2, Jonsson row", "Wang 2023 Table 2, IIV column",
    "Wang 2023 Methods 2.3 (Table 2 RUV = 'NR')"
  ),
  check.names = FALSE
) |>
  knitr::kable(caption = "Source trace for the five packaged models.")
```

# Virtual cohort

The cohort reproduces Wang 2023 Table 1's evaluation dataset. Weight, height
and serum creatinine are drawn log-normally with the published medians and
tuned spreads, then truncated to the published ranges; sex is drawn at the
published 40/11 ratio.

Two covariates are **derived rather than drawn**, using the formulae the paper
itself specifies -- which makes the published Table 1 medians a check on the
cohort rather than an input to it:

* **BSA** by the Mosteller formula, `sqrt(height_cm * weight_kg / 3600)`.
* **eGFR** by the Bedside Schwartz formula (Wang 2023 Equation 1),
  `0.413 * height_cm / Scr_mgdL`.

```{r cohort}
rxode2::rxSetSeed(20230208)
set.seed(20230208)

n_sub <- 51L # matches Wang 2023's evaluation cohort exactly

rtrunc_lnorm <- function(n, med, lo, hi, cv) {
  out <- numeric(0)
  while (length(out) < n) {
    draw <- med * exp(stats::rnorm(n * 4L, 0, sqrt(log(1 + cv^2))))
    out <- c(out, draw[draw >= lo & draw <= hi])
  }
  out[seq_len(n)]
}

cohort <- data.frame(
  id  = seq_len(n_sub),
  WT  = rtrunc_lnorm(n_sub, 19.0, 9.5, 62.0, 0.45),
  HT  = rtrunc_lnorm(n_sub, 113, 73, 168, 0.20),
  AGE = rtrunc_lnorm(n_sub, 5.5, 1.0, 13.0, 0.55),
  SCR_MGDL = rtrunc_lnorm(n_sub, 0.31, 0.16, 0.71, 0.35)
) |>
  mutate(
    SEXF  = as.integer(seq_len(n_sub) > 40L), # Wang Table 1: 40 male / 11 female
    BSA   = sqrt(HT * WT / 3600),             # Mosteller
    CRCL  = 0.413 * HT / SCR_MGDL,            # Bedside Schwartz (Wang Eq. 1)
    CREAT = SCR_MGDL * 88.4,                  # mg/dL -> umol/L, for Gao 2021
    OCC   = 1L                                # single HD-MTX course
  )

summary_tbl <- cohort |>
  summarise(
    `Weight (kg)`     = median(WT),
    `Height (cm)`     = median(HT),
    `Age (years)`     = median(AGE),
    `BSA (m^2)`       = median(BSA),
    `Scr (mg/dL)`     = median(SCR_MGDL),
    `eGFR (mL/min/1.73m^2)` = median(CRCL)
  )
knitr::kable(summary_tbl, digits = 2,
             caption = "Simulated cohort medians (compare Wang 2023 Table 1).")
```

The two derived covariates reproduce the paper's own reported medians, which
validates both the cohort construction and the covariate definitions. Note both
gates are on the cohort **median**, not on any individual subject, so they do
not depend on which subjects land in the tails.

```{r cohort-gates}
# Wang 2023 Table 1 reports evaluation-cohort median BSA = 0.77 m^2;
# Results 3.2 reports median eGFR = 149.9 mL/min/1.73 m^2. Both are
# consequences of the drawn weight/height/Scr, not inputs to them.
med_bsa  <- median(cohort$BSA)
med_egfr <- median(cohort$CRCL)

stopifnot(
  abs(med_bsa - 0.77) < 0.05,
  abs(med_egfr - 149.9) / 149.9 < 0.10
)
c(BSA = med_bsa, eGFR = med_egfr)
```

# Simulation

Wang 2023 Methods 2.2 describes the administration schedule precisely: 10% of
the total dose infused over 0.5 h, the remaining 90% over the following 23.5 h.
Both the low-risk (3.0 g/m^2) and the intermediate/high-risk (5.0 g/m^2) dose
levels are simulated, which supplies the treatment grouping the NCA needs.

Doses are converted from g/m^2 to umol using the methotrexate molecular weight
of 454.44 g/mol, because every model in this set is parameterised in umol and
umol/L.

```{r events}
MW_MTX <- 454.44 # g/mol

make_events <- function(cohort, dose_g_m2) {
  dose_umol <- dose_g_m2 * cohort$BSA * 1e6 / MW_MTX

  dosing <- bind_rows(
    data.frame(id = cohort$id, time = 0.0, amt = dose_umol * 0.10,
               dur = 0.5, evid = 1L, cmt = "central"),
    data.frame(id = cohort$id, time = 0.5, amt = dose_umol * 0.90,
               dur = 23.5, evid = 1L, cmt = "central")
  )

  # Observations on the ODE STATE 'central'; rxode2 returns the algebraic
  # observable Cc as a column at those rows.
  obs_times <- sort(unique(c(seq(0, 48, by = 0.5), seq(49, 336, by = 1))))
  obs <- expand.grid(id = cohort$id, time = obs_times) |>
    mutate(amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central")

  bind_rows(dosing, obs) |>
    left_join(cohort, by = "id") |>
    arrange(id, time, desc(evid))
}

sim_one <- function(model_name, dose_g_m2) {
  mod <- readModelDb(model_name)
  ev <- make_events(cohort, dose_g_m2)
  rxode2::rxSolve(mod, events = ev, returnType = "data.frame",
                  addDosing = FALSE) |>
    mutate(model = model_name,
           dose_g_m2 = dose_g_m2,
           treatment = paste0(dose_g_m2, " g/m^2"))
}

sim <- bind_rows(lapply(
  c(3.0, 5.0),
  function(d) bind_rows(lapply(mtx_models, sim_one, dose_g_m2 = d))
))

nrow(sim)
```

## Concentration-time profiles

This reproduces the shape of Wang 2023 Figure 1 (observed methotrexate
concentration versus time after dosing) and shows, per model, the spread the
paper's Figure 2 goodness-of-fit panels summarise.

```{r fig-profiles, fig.cap = "Simulated methotrexate concentration-time profiles by model and dose level. Replicates the layout of Figure 1 of Wang 2023; the horizontal lines are the therapeutic thresholds quoted in that paper's Introduction."}
sim_q <- sim |>
  filter(time > 0) |>
  group_by(model, treatment, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_q, aes(time, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = c(0.1, 16, 65), linetype = "dashed",
             colour = "grey40", linewidth = 0.3) +
  scale_y_log10() +
  scale_x_continuous(limits = c(0, 168)) +
  facet_wrap(~model, ncol = 2) +
  labs(x = "Time after start of infusion (h)",
       y = "Methotrexate concentration (umol/L)",
       colour = "Dose", fill = "Dose") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Steady-state concentration at end of infusion

Wang 2023 Introduction defines Cp,ss as the concentration at the end of the
24 h infusion and quotes the efficacy thresholds: above 65 umol/L for
intermediate/high-risk children (the 5.0 g/m^2 arm), above 33 umol/L for
low-risk children (the 3.0 g/m^2 arm), and subtherapeutic below 16 umol/L.

The point of the paper is that these six models **disagree** with one another,
so this table is presented as a cross-model comparison rather than as a gate on
any single published number.

```{r cpss}
cpss <- sim |>
  filter(abs(time - 24) < 1e-6) |>
  group_by(model, treatment) |>
  summarise(
    `Median Cp,ss` = median(Cc),
    `P10`          = quantile(Cc, 0.10),
    `P90`          = quantile(Cc, 0.90),
    .groups = "drop"
  )

cpss |>
  knitr::kable(digits = 1,
               caption = "Simulated Cp,ss (umol/L) at the end of the 24 h infusion.")
```

```{r cpss-gate}
# Structural gate: across every model and both dose levels the median Cp,ss
# must land inside the clinically plausible decade for HD-MTX that Wang 2023
# spans in its Introduction (subtherapeutic < 16 umol/L; > 100 umol/L at 24 h
# is supratherapeutic). A mis-transcribed clearance, volume, dose or unit
# conversion moves a whole median out of this band immediately. Deliberately
# asserted on the MEDIAN, never on the cohort extremes, which are not
# reproducible across rxode2 builds.
stopifnot(all(cpss$`Median Cp,ss` > 5), all(cpss$`Median Cp,ss` < 150))

# The 5 g/m^2 arm must exceed the 3 g/m^2 arm for every model (the models are
# linear, so this is exact up to the drawn cohort).
wide <- tidyr::pivot_wider(cpss, id_cols = model, names_from = treatment,
                           values_from = `Median Cp,ss`)
stopifnot(all(wide[["5 g/m^2"]] > wide[["3 g/m^2"]]))
```

# Structural verification

## Typical parameter values against the paper's own text

Wang 2023 Results 3.1 quotes typical values it computed from the same Table 2
equations packaged here. Solving each model at **its own** development-cohort
median (Wang 2023 Table 1) with between-subject variability zeroed must
reproduce those numbers. These are exact, deterministic identities -- no random
draw is involved -- so they are asserted tightly.

```{r typical-gates}
typical_solve <- function(model_name, covs) {
  ui <- rxode2::zeroRe(rxode2::rxode(readModelDb(model_name)))
  ev <- data.frame(id = 1L, time = c(0, 1), amt = c(1000, NA),
                   dur = c(0.5, NA), evid = c(1L, 0L), cmt = "central")
  ev <- cbind(ev, as.data.frame(covs))
  s <- rxode2::rxSolve(ui, events = ev, returnType = "data.frame",
                       addDosing = FALSE)
  s[1, ]
}

typ <- list(
  Aumente_2006_methotrexate         = list(WT = 24.2, AGE = 5.0),
  Gao_2021_methotrexate             = list(WT = 19.0, CREAT = 26.0),
  Hui_2019_methotrexate             = list(BSA = 0.735, CRCL = 149.9,
                                           AGE = 5.29, OCC = 1L),
  MedellinGaribay_2020_methotrexate = list(BSA = 0.79, WT = 21.2),
  Jonsson_2011_methotrexate         = list(WT = 19.0)
)

typ_res <- do.call(rbind, lapply(names(typ), function(nm) {
  s <- typical_solve(nm, typ[[nm]])
  data.frame(model = nm, cl = s$cl, vc = s$vc)
}))

knitr::kable(typ_res, digits = 3,
             caption = "Typical CL and Vc at each model's own cohort median.")
```

```{r typical-assert}
getv <- function(m, col) typ_res[[col]][typ_res$model == m]

stopifnot(
  # "Typical estimates for MTX clearance in the included studies ranged from
  #  3.52 [Jonsson] to 7.73 [Hui] L/h" -- Wang 2023 Results 3.1.
  abs(getv("Jonsson_2011_methotrexate", "cl") - 3.52) < 0.01,
  abs(getv("Hui_2019_methotrexate", "cl") - 7.73) < 0.02,
  # "the typical central volume of distribution ... ranging from 7.5
  #  [Medellin-Garibay] to 24.1 [Jonsson] L" -- same paragraph.
  abs(getv("Jonsson_2011_methotrexate", "vc") - 24.1) < 0.05,
  abs(getv("MedellinGaribay_2020_methotrexate", "vc") - 7.6) < 0.1,
  # Gao 2021 Table 2 typical CL for a 19 kg child at the 26 umol/L reference.
  abs(getv("Gao_2021_methotrexate", "cl") - 6.9) < 0.01
)
```

## Terminal half-life against the analytic eigenvalue

The terminal half-life PKNCA estimates from the simulated curve must agree with
the one implied by the model's own micro-constants. Both sides use the same
parameter values, so the only difference is numerical -- a tight bound is
correct here, and would catch a compartment wired to the wrong state.

```{r analytic-halflife}
analytic_thalf <- function(model_name, covs) {
  s <- typical_solve(model_name, covs)
  k10 <- s$cl / s$vc
  k12 <- s$q / s$vc
  k21 <- s$q / s$vp
  if (!is.null(s$q2) && !is.na(s$q2)) {
    k13 <- s$q2 / s$vc
    k31 <- s$q2 / s$vp2
    A <- matrix(c(-(k10 + k12 + k13), k21, k31,
                  k12, -k21, 0,
                  k13, 0, -k31), nrow = 3, byrow = TRUE)
  } else {
    A <- matrix(c(-(k10 + k12), k21,
                  k12, -k21), nrow = 2, byrow = TRUE)
  }
  lambda <- Re(eigen(A)$values)
  log(2) / abs(max(lambda)) # least-negative eigenvalue = terminal slope
}

thalf_analytic <- vapply(names(typ),
                         function(nm) analytic_thalf(nm, typ[[nm]]),
                         numeric(1))
round(thalf_analytic, 2)
```

# PKNCA validation

Wang 2023 reports no non-compartmental analysis of its own -- it reports
prediction-error metrics instead -- so there is no published NCA table to
compare against. The NCA is therefore run as an **internal-identity** check:
PKNCA's terminal half-life on a typical-value (variability-zeroed) solve must
match the analytic eigenvalue computed above.

Running the NCA on the typical-value solve rather than the full-variability
cohort is deliberate: over a cohort with between-subject variability, the
half-life estimate is both `NA`-poisoned and tmax-selected, which makes it a
poor gate.

```{r pknca}
typical_profile <- function(model_name, covs, dose_g_m2 = 3.0) {
  ui <- rxode2::zeroRe(rxode2::rxode(readModelDb(model_name)))
  bsa <- if (!is.null(covs$BSA)) covs$BSA else sqrt(113 * covs$WT / 3600)
  dose_umol <- dose_g_m2 * bsa * 1e6 / MW_MTX

  dosing <- data.frame(
    id = 1L, time = c(0, 0.5),
    amt = c(dose_umol * 0.10, dose_umol * 0.90),
    dur = c(0.5, 23.5), evid = 1L, cmt = "central"
  )
  obs <- data.frame(
    id = 1L,
    time = sort(unique(c(seq(0, 48, by = 0.25), seq(49, 336, by = 1)))),
    amt = NA_real_, dur = NA_real_, evid = 0L, cmt = "central"
  )
  ev <- bind_rows(dosing, obs)
  for (nm in names(covs)) ev[[nm]] <- covs[[nm]]
  ev <- arrange(ev, time, desc(evid))

  s <- rxode2::rxSolve(ui, events = ev, returnType = "data.frame",
                       addDosing = FALSE) |>
    mutate(model = model_name, dose_umol = dose_umol)
  s
}

typ_prof <- bind_rows(lapply(names(typ),
                             function(nm) typical_profile(nm, typ[[nm]])))

# PKNCA input: filter ONLY on missing concentrations so the time-zero row
# survives (dropping it triggers the "AUC range starting before the first
# measurement" warning).
nca_conc <- typ_prof |>
  filter(!is.na(Cc)) |>
  mutate(id = model) |>
  select(id, time, Cc, model)

nca_dose <- typ_prof |>
  group_by(model) |>
  summarise(id = first(model), time = 0, amt = first(dose_umol),
            .groups = "drop") |>
  select(id, time, amt, model)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | model + id,
                             concu = "umol/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(nca_dose, amt ~ time | model + id,
                             doseu = "umol")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  aucinf.obs = TRUE, half.life = TRUE, clast.obs = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-table}
nca_wide <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "half.life", "auclast", "aucinf.obs")) |>
  select(model, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(`Analytic t1/2 (h)` = thalf_analytic[model]) |>
  # Rename BY NAME, never positionally via col.names.
  dplyr::rename(
    "Model"              = model,
    "Cmax (umol/L)"      = cmax,
    "Tmax (h)"           = tmax,
    "t1/2 (h), PKNCA"    = half.life,
    "AUClast (umol*h/L)" = auclast,
    "AUCinf (umol*h/L)"  = aucinf.obs
  )

knitr::kable(nca_wide, digits = 2,
             caption = paste("PKNCA parameters on the typical-value solve at",
                             "3 g/m^2, with the analytic terminal half-life",
                             "for comparison."))
```

```{r pknca-assert}
rel_diff <- abs(nca_wide[["t1/2 (h), PKNCA"]] - nca_wide[["Analytic t1/2 (h)"]]) /
  nca_wide[["Analytic t1/2 (h)"]]
names(rel_diff) <- nca_wide$Model
round(100 * rel_diff, 3)

# Same parameters on both sides -- the difference is pure numerical error in
# the log-linear terminal regression, so this is tight by construction.
stopifnot(all(rel_diff < 0.02))

# Tmax must sit at the end of ONE of the two infusion segments -- 0.5 h (end of
# the 10% loading infusion) or 24 h (end of the 90% maintenance infusion) --
# and never in between. Concentration rises throughout the loading infusion,
# then approaches the much lower maintenance input rate monotonically, so the
# maximum over the whole regimen is attained at one of the two endpoints.
# Which endpoint wins depends on the central volume; see the note below.
stopifnot(all(nca_wide[["Tmax (h)"]] %in% c(0.5, 24)))

# Equivalently and more strongly: Cmax equals the larger of the two
# end-of-segment concentrations, exactly.
endpoint_conc <- typ_prof |>
  filter(time %in% c(0.5, 24)) |>
  group_by(model) |>
  summarise(peak = max(Cc), .groups = "drop")
stopifnot(all(abs(
  endpoint_conc$peak[match(nca_wide$Model, endpoint_conc$model)] -
    nca_wide[["Cmax (umol/L)"]]
) < 1e-6))

# AUCinf must be at least AUClast, and the extrapolated tail must be
# negligible given sampling out to 336 h. Note the comparison is >=, not >:
# for Aumente, Hui and Medellin-Garibay the profile has decayed so far by
# 336 h (Clast of 8e-10, 2e-6 and 3e-16 umol/L respectively) that the
# extrapolated tail vanishes into floating-point precision and AUCinf equals
# AUClast exactly. That is the window doing its job, not a defect.
extrap_frac <- (nca_wide[["AUCinf (umol*h/L)"]] - nca_wide[["AUClast (umol*h/L)"]]) /
  nca_wide[["AUCinf (umol*h/L)"]]
names(extrap_frac) <- nca_wide$Model
round(100 * extrap_frac, 3)

stopifnot(
  all(nca_wide[["AUCinf (umol*h/L)"]] >= nca_wide[["AUClast (umol*h/L)"]]),
  all(extrap_frac >= 0),
  # Deterministic typical-value solves, so this is tight by construction.
  all(extrap_frac < 0.02)
)
```

Note the spread of terminal half-lives across the set -- roughly 6 h
(Medellin-Garibay) to 115 h (Gao). Gao's is by far the longest because it is
the only three-compartment model of the group, and its first peripheral
compartment (41 L, reached through an intercompartmental clearance of only
0.255 L/h) drains very slowly. That deep compartment carries little drug, which
is why the extrapolated AUC fraction stays under 1% despite the long terminal
slope.

## Cmax is not always Cp,ss under this regimen

Wang's regimen infuses 10% of the dose over 0.5 h and the remaining 90% over
23.5 h, so the loading segment runs at roughly five times the maintenance
infusion rate. For models with a small central volume the loading segment
therefore produces a transient peak that **exceeds** the end-of-infusion
concentration, and Cmax is attained at 0.5 h rather than at 24 h.

In this set that happens for Medellin-Garibay 2020, whose central volume
(7.6 L at the cohort median) is the smallest of the six models Wang evaluated:
its Cmax is about 56 umol/L at 0.5 h against a Cp,ss of about 35 umol/L at
24 h. Aumente 2006, with the next-smallest volume, is nearly tied between the
two endpoints. The remaining models peak at 24 h.

This matters when reading the paper's therapeutic thresholds: Wang defines
Cp,ss as the 24 h value specifically, so for these models Cp,ss is *not* the
maximum concentration the patient experiences.

# Published external-evaluation performance

For reference, Wang 2023 Table 3 reports how each model performed against the
independent 51-patient cohort. These are the paper's own results, reproduced
here as published; they are **not** recomputed by this vignette, because the
evaluation dataset is not public.

```{r published-performance, echo = FALSE}
data.frame(
  Model = c("Aumente 2006", "Gao 2021", "Hui 2019",
            "Medellin-Garibay 2020", "Zhang 2010 (not packaged)",
            "Jonsson 2011"),
  `IPRED median PE (%)` = c(6.52, -25.20, -10.43, -0.75, 1.16, 5.25),
  `IPRED MPE (%)`       = c(12.06, 1.33, 8.78, 22.71, 16.00, 64.44),
  `IPRED RMSE (%)`      = c(76.09, 63.96, 82.96, 75.55, 63.39, 152.25),
  `PRED median PE (%)`  = c(-31.70, -26.76, -33.23, -7.56, -29.92, 442.04),
  `PRED MPE (%)`        = c(-25.51, 3.22, -8.24, 6.39, 15.62, 780.87),
  `PRED RMSE (%)`       = c(81.15, 72.39, 62.88, 68.33, 145.92, 1182.24),
  check.names = FALSE
) |>
  knitr::kable(caption = paste(
    "Wang 2023 Table 3. Every model exceeded the 30% RMSE acceptability",
    "threshold; only Medellin-Garibay met the +/-20% median population bias",
    "criterion."))
```

Wang's conclusion is that all six models are usable for individual (a
posteriori) prediction but too imprecise for a priori model-informed precision
dosing without refinement, and that Jonsson 2011 -- the only weight-only model,
built on a 5-8 g/m^2 cohort that does not overlap the 1-5 g/m^2 evaluation
data -- fails badly on population predictions.

# Assumptions and deviations {#assumptions-and-deviations}

## Provenance: these are transcriptions from a review

**Four of the five models are transcribed from Wang 2023 Table 2, not from
their own primary publications**, because those primaries are not open access.
Each model file records this in its `reference` metadata and each is queued for
re-extraction from its primary. The exception is Gao 2021, which is open access
and is encoded from the primary.

## Wang 2023 Table 2 mis-transcribes the Gao 2021 row (twice)

Gao 2021 being open access made a direct comparison possible, and it found two
errors in Wang's rendering:

1. Wang prints the CL between-subject CV as **17.9%**; Gao's own Table 2 says
   **17.5%**.
2. Wang places the second eta on **V2** in its equation column and calls it
   **V1** in its IIV column -- disagreeing with itself. Gao places it on
   **Q1**, and says so twice (Table 2 populates the `CV for IIV` cell on the Q1
   row, and the Results state that adding serum creatinine "reduced the
   variability of Q1 by 29.2%").

`Gao_2021_methotrexate.R` follows the Gao primary. This finding is the direct
evidence that Wang's Table 2 is not a reliable transcription source, and is why
the other four models carry re-extraction notes.

## Gao 2021: the serum-creatinine term is unbounded

Gao's creatinine effect is linear, not a power term:
`CL = CL_typical * (1 + (Scr - 26) * (-0.0097))`. It reaches zero at
Scr = 129.1 umol/L and goes **negative** above that -- inside Gao's own
reported creatinine range (0.1-1.5 mg/dL = 8.8-132.6 umol/L). The published
equation is reproduced as printed rather than clamped. Simulations must keep
`CREAT` below about 129 umol/L; the cohort here has a maximum well below it.

Gao's own Table 2 footnote writes the equation as
`[CL = CL_typical x ((SCr-26) x 0.0097)]`, which is garbled -- it drops both
the `1 +` and the minus sign, and would make clearance zero at the reference
value. Wang's rendering is the coherent one and is used.

## Hui 2019: an unresolvable parenthesisation

Wang 2023 Table 2 prints Hui's renal term as
`(eGFR x 1.73/192 x BSA)^0.256`. Read with ordinary left-to-right precedence --
and a 400 dpi render confirms an inline solidus rather than a fraction bar --
this is `(eGFR * 1.73 * BSA / 192)^0.256`, which is what is encoded. That
reading also behaves like a properly normalised covariate: at the evaluation
cohort's median eGFR (149.9) and Hui's median BSA (0.735) the bracket evaluates
to 0.993, i.e. centred at approximately 1.

The alternative grouping `(eGFR x 1.73)/(192 x BSA)` cannot be excluded without
the primary and would raise typical clearance by about 17%. Hui 2019 is queued
for re-extraction.

## Jonsson 2011: the residual error is assumed, not published

Wang 2023 Table 2 records Jonsson's residual variability as **NR** (not
reported), and Wang supplied one for the evaluation (Methods 2.3: "a
proportional error of 30% was assumed for the residual variability if the
models did not report this information"). The packaged model carries
`propSd <- fixed(0.30)` so it reproduces the model Wang actually evaluated, but
**this is not a Jonsson 2011 estimate** and must not be read as one. It is
wrapped in `fixed()` precisely because nothing in either source estimated it.

## Aumente 2006: the 10-year boundary is undefined

Wang 2023 Table 2 gives one pair of CL/V1 equations for "age > 10 years" and
another for "age < 10 years", leaving age exactly 10.0 undefined (both
inequalities are strict). This implementation assigns age exactly 10 to the
older stratum. The choice changes the prediction only for a subject recorded at
exactly 10.0 years.

## Hui 2019: the occasion count is an implementation choice

Wang reports a single shared between-occasion variance (14.9% CV) for
clearance but does not state how many occasions the primary modelled. Four
occasion slots are provided, all carrying that same published variance with
only the first estimable (the NONMEM `OMEGA BLOCK(1) SAME` idiom). The slot
count invents no parameter value; it only bounds how many distinct courses a
user can encode. Set `OCC = 1` for a single-course simulation.

Wang's Results also mis-reads this row once, describing the 14.9% as an
inter-individual effect; Table 2's own column header
(`IIV (%) (IOV (%))`) makes 14.9% the inter-occasion value and 14.3% the
inter-individual one, and Table 2 is followed here.

## Zhang 2010 is deferred, not dropped

The sixth model Wang evaluated, Zhang 2010
(Int J Clin Pharmacol Ther 48:11-21, doi:10.5414/cpp48011), is **not packaged
in this release**. Its clearance equation as printed by Wang is

```
CL (L/h) = (5.04 * (1 - 0.278 * Gender) * BSA^0.777 + (OH/100)^0.514) * exp(eta_CL)
```

and two things in it cannot be resolved from any available source:

* **`OH` -- "pre-chemotherapy alkalinization volume" -- has no stated units.**
  The term is *additive* to the BSA term, in L/h, so the scale matters
  enormously: at a typical pre-hydration volume of ~1500 mL the term
  contributes about 4 L/h, comparable to the entire BSA term and roughly
  doubling typical clearance. Zhang's own abstract presents 5.04 L/h *as* the
  typical clearance, which is only consistent with the `OH` term being near
  zero. The two readings differ about two-fold and cannot be told apart from
  what is on disk.
* **`Gender`'s reference category is unstated.** `(1 - 0.278 * Gender)` means
  `Gender = 1` lowers clearance by 27.8%, but neither Wang nor Zhang's
  abstract says which sex that is. A flip changes typical clearance by ~38%.

The Zhang 2010 primary is not open access. Rather than ship a model with an
assumed covariate scale that can swing clearance two-fold, the paper has been
queued for acquisition and will be extracted from its primary, where both the
units and the coding will be stated. Its row is retained in the published
performance table above for completeness.

## Not carried over

Hui 2019's primary also reports a separate osteosarcoma model. Wang 2023 did
not evaluate it and it is not extracted here.
