---
title: "Aripiprazole once monthly (Wang 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aripiprazole once monthly (Wang 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Wang 2022 contributes two models to `nlmixr2lib`:

* `modellib("Wang_2022_aripiprazole")` -- the population PK model for oral
  aripiprazole and for aripiprazole once monthly (AOM), the intramuscular
  extended-release injectable suspension.
* `modellib("Wang_2022_aripiprazole_relapse")` -- the exposure-response
  time-to-event model relating the model-predicted aripiprazole minimum
  concentration to the time to exacerbation of psychotic symptoms / impending
  relapse.

The two are independent fits sharing one data programme: the popPK model
produces the individual `CMIN_ARI` values that the exposure-response model
consumes, but the survival regression was run in SAS on those predictions
rather than estimated jointly. They are therefore packaged as two model files
and validated in this one vignette.

* Citation: `r readModelDb("Wang_2022_aripiprazole")()$reference`
* Article: <https://doi.org/10.1002/cpdd.1022>

# Population

The population PK model was built from 6153 aripiprazole plasma concentrations
in 663 subjects pooled across five studies: four phase 1 studies with serial PK
sampling (31-98-206, 31-98-207, CN138020 and 31-05-244) and one phase 3 study
with sparse PK sampling (31-07-246). Fifty-two subjects were healthy volunteers
who received oral aripiprazole only; the remaining 611 had schizophrenia or
schizoaffective disorder. Body mass index ranged from 15 to 61 kg/m^2. CYP2D6
phenotype was extensive in 621 subjects and poor in 42; 13 extensive
metabolizers took a concomitant CYP2D6 inhibitor and 25 a concomitant CYP3A4
inhibitor. All AOM doses were given in the gluteus maximus except in study
CN138020 (13 subjects, non-dominant arm or midlateral thigh).

A sixth study, phase 3 study 31-07-247, was reserved for external validation.
The final model met the pre-specified acceptance criteria for the 400/300 mg AOM
arm (median %PPE -6.8%, median |%PPE| 29.2%) but **not** for the 50/25 mg arm
(median |%PPE| 41.3%, 75th percentile 74.6%), so the model must not be
extrapolated below 300 mg AOM. Every simulation in this vignette stays at or
above 300 mg AOM, or on the 10-30 mg/day oral range the paper used to define its
therapeutic window.

The exposure-response analysis set is a different, smaller population: 615
subjects with schizophrenia from the randomised stages of studies 31-07-246 and
31-07-247 (120 on oral aripiprazole, 495 on 400/300 mg AOM, plus the placebo
arms), contributing 85 impending-relapse events and 530 censored subjects. The
121 subjects on 50/25 mg AOM were excluded for the reason above.

The same information is available programmatically via each model's
`population` metadata (`readModelDb("Wang_2022_aripiprazole")()$population`).

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them.

## Population PK model (`Wang_2022_aripiprazole`)

| Equation / parameter | Value | Source location |
|---|---|---|
| `lr1` (R1, zero-order oral input rate) | 9.33 mg/h (fixed) | Table 1, row "R1: infusion rate of oral dose into depot compartment (mg/h)" |
| `lka_oral` | 0.540 /h (fixed) | Table 1, row "Ka: oral first-order absorption rate (1/h)" |
| `lka_im` | 0.000904 /h (%RSE 5.3) | Table 1, row "IM Ka: AOM first-order absorption rate (1/h)" |
| `e_bmi_ka_im` | -0.975 (%RSE 11.5) | Table 1, row "IM Ka: power for (BMI/28)" |
| `e_sexf_ka_im` | 0.346 (%RSE 28.9) | Table 1, row "IM Ka: proportional shift for men" |
| `lfdepot_im` (F2) | 1.48 (%RSE 4.9) | Table 1, row "F2: relative bioavailability for AOM" |
| `lcl_em` | 3.71 L/h (%RSE 4.0) | Table 1, row "CL/F: clearance for EM (L/h)" |
| `lcl_pm` | 1.88 L/h (%RSE 6.9) | Table 1, row "CL/F: clearance for PM (L/h)" |
| `e_cyp2d6_inh_cl` | -0.511 (fixed) | Table 1, row "CL/F: proportional change in CL/F for CYP2D6 inhibitor" |
| `e_cyp3a4_inh_cl` | -0.237 (fixed) | Table 1, row "CL/F: proportional change in CL/F for CYP3A4 inhibitor" |
| `lvc` | 93.4 L (%RSE 8.8) | Table 1, row "Vc/F: central volume (L)" |
| `lq` | 0.591 L/h (fixed) | Table 1, row "Q1/F: intercompartmental CL/F (L/h)" |
| `lvp` | 118 L (fixed) | Table 1, row "Vp1/F: peripheral volume (L)" |
| `lq2` | 28.8 L/h (fixed) | Table 1, row "Q2/F: second intercompartmental CL/F (L/h)" |
| `lvp2` | 134 L (fixed) | Table 1, row "Vp2/F: second peripheral volume (L)" |
| `etalcl` | 38.34 %CV (%RSE 6.9) | Table 1, IIV column, CL/F row |
| `etalvc` | 124.50 %CV (%RSE 15.2) | Table 1, IIV column, Vc/F row |
| `etalka_oral` | 65.88 %CV (fixed) | Table 1, IIV column, oral Ka row |
| `etalka_im` | 55.59 %CV (%RSE 8.2) | Table 1, IIV column, IM Ka row |
| `propSdPhase1` | 24.23 %CV (%RSE 8.4) | Table 1, row "Phase 1 RV (%CV)" |
| `propSdPhase3` | 28.11 %CV (%RSE 4.7) | Table 1, row "Phase 3 RV (%CV)" |
| AOM Ka covariate equation | `0.000904 * (BMI/28)^-0.975 * (1 + 0.346 * Male)` | Table 1 footnote, "Related equations" |
| CL/F covariate equation | `(3.71 * EM + 1.88 * PM) * (1 - 0.511 * CYP2D6) * (1 - 0.237 * CYP3A4)` | Table 1 footnote, "Related equations" |
| Three-compartment linear disposition | n/a | Results, "Base Model Development" and "Final PopPK Model" |
| Sigmoid (zero-order then first-order) oral absorption; first-order AOM absorption | n/a | Results, "Base Model Development"; Table 1 footnote |
| Which parameters were estimated vs carried over fixed | n/a | Results, "Base Model Development" and "Final PopPK Model" |

## Exposure-response model (`Wang_2022_aripiprazole_relapse`)

| Equation / parameter | Value | Source location |
|---|---|---|
| `lhaz_base` | -6.256 (= -1 x 6.256; SE 0.1474, 95% CI 5.97-6.55) | Table 2, row "Intercept" |
| `e_cmin_haz` | -1.484 (= -1 x 1.484; SE 0.2177, 95% CI 1.06-1.91) | Table 2, row "Predicted aripiprazole (Cmin >= 95 ng/mL) proximate to the event" |
| Hazard ratio `exp(-e_cmin_haz)` = 4.41 (95% CI 2.89-6.75) | 4.41 | Table 2, row "Calculated hazard ratio of expected survival time" |
| 95 ng/mL threshold | 95 ng/mL | Results, "Exposure-Response Analysis of Time to Relapse" |
| Exponential (constant-hazard) functional form | n/a | Results, "Exposure-Response Analysis of Time to Relapse" |
| Time unit = days | n/a | Figures 5 and 6 x-axis label, "Time to Relapse (days)" |

The only non-transcription step in either file is the survival-model
parameterisation. Wang 2022 Table 2 reports an accelerated-failure-time
exponential regression on log survival time (SAS `PROC LIFEREG`),
`log(T) = 6.256 + 1.484 * I(Cmin >= 95)`. For the exponential distribution the
constant hazard is the reciprocal of the expected survival time, so the
log-hazard parameters are the negatives of the printed ones. The sign is
confirmed by the paper's own derived quantity: `exp(1.484) = 4.41` must be a
hazard *reduction* for subjects at or above the threshold, which is exactly what
`exp(-e_cmin_haz)` gives below.

# Virtual cohort

Original observed data are not publicly available. The cohorts below use
covariate distributions that approximate the published trial demographics: 60%
men (Wang 2022 Methods, "Population PK Model Simulations", which specifies the
same split for the paper's own 10 000-subject virtual population) and a CYP2D6
poor-metabolizer prevalence of 42/663 = 6.3% matching the model development set.

BMI is drawn log-normally with a median of 28 kg/m^2 -- the reference value the
paper's own AOM Ka equation is centred on -- and a geometric CV of 22%. The
paper states only that the virtual BMI values were "log-normally distributed on
the basis of sex to be similar to the phase 3 population" and reports the
observed range (15-61 kg/m^2); the per-sex distribution parameters are in the
supplement, which is not on disk. See Assumptions and deviations.

```{r cohort}
# `set.seed()` seeds R's RNG only. rxode2's own simulation RNG is partitioned
# per solver thread, so the eta draws below differ between a 2-core CI runner
# and a 16-thread workstation. Every assertion in this vignette is therefore
# either deterministic (typical-value solve, closed-form identity) or written
# on a robust central statistic, never on a cohort extreme.
set.seed(20220211)

nPerArm <- 200L

makeSubjects <- function(n, id_offset = 0L, pmFraction = 42 / 663) {
  tibble::tibble(
    id        = id_offset + seq_len(n),
    SEXF      = as.numeric(seq_len(n) > round(0.60 * n)),   # 60% men
    BMI       = round(stats::rlnorm(n, meanlog = log(28), sdlog = 0.22), 2),
    CYP2D6_PM = as.numeric(seq_len(n) <= round(pmFraction * n)),
    CONMED_CYP2D6_INH = 0,
    CONMED_CYP3A4_INH = 0,
    STUDY_ARI_PHASE3  = 1
  )
}

# One arm = subjects x (dose records + observation grid).
# Oral doses use rate = -1 so rxode2 applies the modelled zero-order input
# rate R1 = 9.33 mg/h; AOM doses are ordinary bolus records into `depot2`.
makeArm <- function(subj, cohort, amt, cmt, ii, addl, obsTimes) {
  # Resolve the dose rate from the ARGUMENT before mutate() shadows `cmt` with
  # the newly created (length-n) column of the same name.
  doseRate <- if (identical(cmt, "depot")) -1 else 0
  doses <- subj |>
    dplyr::mutate(
      time = 0, evid = 1L, amt = amt, cmt = cmt,
      rate = doseRate,
      ii = ii, addl = addl
    )
  obs <- subj |>
    tidyr::expand_grid(time = obsTimes) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central", rate = 0, ii = 0, addl = 0L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(cohort = cohort) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

tauAom  <- 672   # 28 days, in hours
tauOral <- 24

# AOM: 31 doses (dose 31 at 20160 h); observe the first interval densely for
# the concentration-time profile and the last interval for steady-state NCA.
obsAom  <- sort(unique(c(seq(0, tauAom, by = 6),
                         seq(30 * tauAom, 31 * tauAom, by = 12))))
# Oral: 91 doses (dose 91 at 2160 h) -- about 12 terminal half-lives.
obsOral <- sort(unique(c(0, seq(90 * tauOral, 91 * tauOral, by = 0.5))))

events <- dplyr::bind_rows(
  makeArm(makeSubjects(nPerArm,   0L), "AOM 400 mg q28d", 400, "depot2", tauAom,  30L, obsAom),
  makeArm(makeSubjects(nPerArm, 200L), "Oral 10 mg QD",    10, "depot",  tauOral, 90L, obsOral),
  makeArm(makeSubjects(nPerArm, 400L), "Oral 30 mg QD",    30, "depot",  tauOral, 90L, obsOral)
)

stopifnot(
  !anyDuplicated(events[events$evid == 0L, c("id", "time")]),
  nPerArm <= 200L
)
```

# Simulation

```{r simulate}
mod <- readModelDb("Wang_2022_aripiprazole")

sim <- rxode2::rxSolve(mod, events = events, keep = c("cohort", "SEXF", "CYP2D6_PM", "BMI"))
```

A typical-value (`zeroRe`) solve supplies the deterministic checks. Because the
paper reports the AOM absorption half-life, the therapeutic-window bounds and
the CYP2D6-poor-metabolizer dose recommendation as typical-subject quantities,
these are the sharpest gates available.

```{r simulate-typical}
modTypical <- rxode2::zeroRe(mod)

typicalScenario <- function(label, amt, cmt, ii, addl, obsTimes,
                            SEXF = 0, BMI = 28, CYP2D6_PM = 0,
                            CONMED_CYP2D6_INH = 0, CONMED_CYP3A4_INH = 0) {
  subj <- tibble::tibble(
    id = 1L, SEXF = SEXF, BMI = BMI, CYP2D6_PM = CYP2D6_PM,
    CONMED_CYP2D6_INH = CONMED_CYP2D6_INH,
    CONMED_CYP3A4_INH = CONMED_CYP3A4_INH, STUDY_ARI_PHASE3 = 1
  )
  ev <- makeArm(subj, label, amt, cmt, ii, addl, obsTimes)
  rxode2::rxSolve(modTypical, events = ev, omega = NA, keep = "cohort") |>
    as.data.frame()
}

typ <- dplyr::bind_rows(
  typicalScenario("Oral 10 mg QD (EM)",   10, "depot",  tauOral, 90L, obsOral),
  typicalScenario("Oral 30 mg QD (EM)",   30, "depot",  tauOral, 90L, obsOral),
  typicalScenario("AOM 400 mg q28d (EM, men)",  400, "depot2", tauAom, 30L, obsAom),
  typicalScenario("AOM 400 mg q28d (EM, women)", 400, "depot2", tauAom, 30L, obsAom, SEXF = 1),
  typicalScenario("AOM 300 mg q28d (PM, men)",  300, "depot2", tauAom, 30L, obsAom, CYP2D6_PM = 1),
  typicalScenario("AOM 400 mg q28d (PM, men)",  400, "depot2", tauAom, 30L, obsAom, CYP2D6_PM = 1)
)

ssWindow <- function(df, cohort, start, end) {
  d <- df[df$cohort == cohort & df$time >= start & df$time <= end, ]
  list(
    cmin = min(d$Cc), cmax = max(d$Cc),
    auc  = sum(diff(d$time) * (utils::head(d$Cc, -1) + utils::tail(d$Cc, -1)) / 2) / 1000
  )
}
```

# Structural checks against the paper's own reported quantities

## AOM absorption half-life (Results, "Final PopPK Model")

> "For a typical subject with a BMI of 28 [kg/m^2], the estimated AOM absorption
> half-life was approximately 32 days for women and 24 days for men."

```{r check-ka-halflife}
kaIm <- function(BMI, SEXF) 0.000904 * (BMI / 28)^-0.975 * (1 + 0.346 * (1 - SEXF))
absHalfLifeDays <- c(
  women = log(2) / kaIm(28, 1) / 24,
  men   = log(2) / kaIm(28, 0) / 24
)
round(absHalfLifeDays, 2)

# Deterministic: no simulation, no random draw. Tight bound is correct here.
stopifnot(
  abs(absHalfLifeDays[["women"]] - 32) < 0.5,
  abs(absHalfLifeDays[["men"]]   - 24) < 0.5
)
```

## Terminal elimination half-life (Results, "Model Simulations")

> "The average terminal elimination half-life was approximately 7.5 days for men
> and women."

The terminal half-life is a property of the disposition parameters, so it is
read from the slowest eigenvalue of the three-compartment rate matrix rather
than from an NCA. An NCA of an AOM profile would return the *absorption*
half-life instead, because AOM elimination is absorption-rate limited
(absorption half-life about 28 days versus 7 days for elimination) -- the
flip-flop the paper describes in its Discussion.

```{r check-terminal-halflife}
cl <- 3.71; vc <- 93.4; q <- 0.591; vp <- 118; q2 <- 28.8; vp2 <- 134
rateMatrix <- matrix(
  c(-(cl + q + q2) / vc,  q / vp,  q2 / vp2,
    q / vc,              -q / vp,  0,
    q2 / vc,              0,      -q2 / vp2),
  nrow = 3, byrow = TRUE
)
lambdaZ <- max(Re(eigen(rateMatrix)$values))     # least negative eigenvalue
terminalHalfLifeDays <- log(2) / -lambdaZ / 24
round(terminalHalfLifeDays, 2)

stopifnot(terminalHalfLifeDays > 6.5, terminalHalfLifeDays < 8.0)
```

## Therapeutic window (Methods, "Definition of Therapeutic Window")

> "The median of the simulated 10-mg oral minimum predicted drug concentration
> at steady state (Cmin,ss) values (94.0 ng/mL) was used to establish the minimum
> of the therapeutic window ... the 75th percentile of the simulated 30-mg oral
> maximum predicted drug concentration at steady state (Cmax,ss) values
> (534 ng/mL) was selected as a conservative upper bound."

```{r check-therapeutic-window}
windowLower <- 94.0
windowUpper <- 534

oralSsStart <- 90 * tauOral
oral10 <- ssWindow(typ, "Oral 10 mg QD (EM)", oralSsStart, oralSsStart + tauOral)
oral30 <- ssWindow(typ, "Oral 30 mg QD (EM)", oralSsStart, oralSsStart + tauOral)

tibble::tibble(
  quantity  = c("Oral 10 mg QD typical Cmin,ss", "Oral 30 mg QD typical Cmax,ss"),
  simulated = c(oral10$cmin, oral30$cmax),
  published = c(windowLower, windowUpper),
  statistic = c("median over 10 000 subjects", "75th percentile over 10 000 subjects")
) |>
  dplyr::mutate(`% diff` = 100 * (simulated - published) / published) |>
  knitr::kable(digits = 1, caption = "Therapeutic-window bounds (Wang 2022 Methods).")

# The typical-value Cmin,ss is the natural estimate of the paper's cohort
# MEDIAN (clearance is log-normal, so the median subject is the typical
# subject). Deterministic, so a 10% bound is appropriate.
stopifnot(abs(oral10$cmin - windowLower) / windowLower < 0.10)
# The 30 mg bound is a 75th PERCENTILE of a cohort with 125 %CV IIV on Vc, so
# the typical-value Cmax,ss must sit below it -- but not absurdly so.
stopifnot(oral30$cmax < windowUpper, oral30$cmax > 0.6 * windowUpper)
```

## CYP2D6 poor metabolizers (Results, "Model Simulations", Figure 4 panel 2)

> "for subjects with a known poor CYP2D6 metabolizer status, median steady-state
> concentrations after dosing of 300-mg AOM (as recommended) remained centered
> within the therapeutic window, while concentrations following dosing of 400-mg
> AOM approached the upper limit of the therapeutic window."

This is the single tightest cross-check available: reproducing "400 mg AOM in a
poor metabolizer approaches 534 ng/mL" requires `lcl_pm`, `lfdepot_im`,
`lka_im`, `lvc` and the whole three-compartment disposition to be simultaneously
correct.

```{r check-pm-dose}
aomSsStart <- 30 * tauAom
pm300 <- ssWindow(typ, "AOM 300 mg q28d (PM, men)", aomSsStart, aomSsStart + tauAom)
pm400 <- ssWindow(typ, "AOM 400 mg q28d (PM, men)", aomSsStart, aomSsStart + tauAom)

tibble::tibble(
  scenario = c("300 mg AOM, CYP2D6 PM", "400 mg AOM, CYP2D6 PM"),
  `Cmin,ss (ng/mL)` = c(pm300$cmin, pm400$cmin),
  `Cmax,ss (ng/mL)` = c(pm300$cmax, pm400$cmax)
) |>
  knitr::kable(digits = 1,
               caption = "Typical-value steady state in CYP2D6 poor metabolizers; therapeutic window 94.0-534 ng/mL.")

stopifnot(
  # 300 mg stays inside the window, as the label recommendation requires.
  pm300$cmin > windowLower, pm300$cmax < windowUpper,
  # 400 mg "approaches the upper limit": within 5% of 534 and not above it.
  abs(pm400$cmax - windowUpper) / windowUpper < 0.05
)
```

## Steady-state AUC identity

For a linear model at steady state, `AUC0-tau = Dose * F / CL` exactly. Both
sides of this comparison use the same drawn parameters, so the difference is
pure numerical-integration error and a tight bound is correct.

```{r check-auc-identity}
aomMen <- ssWindow(typ, "AOM 400 mg q28d (EM, men)", aomSsStart, aomSsStart + tauAom)
aucClosedForm <- 400 * 1.48 / 3.71     # mg*h/L
c(solved = aomMen$auc, closed_form = aucClosedForm)

stopifnot(abs(aomMen$auc - aucClosedForm) / aucClosedForm < 0.01)
```

# Replicate published figures

```{r figure-1, fig.width = 7, fig.height = 4}
# Replicates Figure 1 of Wang 2022 (left column): mean aripiprazole
# concentration-time profile following the FIRST AOM injection.
sim |>
  dplyr::filter(cohort == "AOM 400 mg q28d", time <= tauAom) |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  labs(x = "Days since first injection", y = "Aripiprazole (ng/mL)",
       title = "Figure 1 (left column) - first 400 mg AOM injection",
       caption = "Median and 5th-95th percentiles, 200 virtual subjects. Replicates Figure 1 of Wang 2022.")
```

```{r figure-4-panel-2, fig.width = 7, fig.height = 4}
# Replicates Figure 4 panel 2 of Wang 2022: steady-state concentration versus
# days since the previous active dose, for CYP2D6 poor and extensive
# metabolizers, against the therapeutic window.
typ |>
  dplyr::filter(grepl("^AOM", cohort), time >= aomSsStart) |>
  dplyr::mutate(daysSinceDose = (time - aomSsStart) / 24) |>
  ggplot(aes(daysSinceDose, Cc, colour = cohort)) +
  geom_line() +
  geom_hline(yintercept = c(windowLower, windowUpper), linetype = "dashed") +
  labs(x = "Days since previous active dose", y = "Aripiprazole (ng/mL)",
       colour = NULL,
       title = "Figure 4 panel 2 - steady state by CYP2D6 phenotype and dose",
       caption = "Typical-value profiles; dashed lines are the 94.0-534 ng/mL therapeutic window. Replicates Figure 4 panel 2 of Wang 2022.")
```

# PKNCA validation

```{r pknca}
simNca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, cohort)

# Defensive time-zero record (pre-dose extravascular concentration is 0).
simNca <- dplyr::bind_rows(
  simNca,
  simNca |> dplyr::distinct(id, cohort) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, cohort, time, .keep_all = TRUE) |>
  dplyr::arrange(id, cohort, time)

concObj <- PKNCA::PKNCAconc(simNca, Cc ~ time | cohort + id,
                            concu = "ng/mL", timeu = "h")

doseDf <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, cohort)

doseObj <- PKNCA::PKNCAdose(doseDf, amt ~ time | cohort + id, doseu = "mg")

# One steady-state interval per cohort: the last complete dosing interval.
intervals <- data.frame(
  cohort  = c("AOM 400 mg q28d", "Oral 10 mg QD", "Oral 30 mg QD"),
  start   = c(aomSsStart, oralSsStart, oralSsStart),
  end     = c(aomSsStart + tauAom, oralSsStart + tauOral, oralSsStart + tauOral),
  cmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE,
  stringsAsFactors = FALSE
)

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))

ncaWide <- as.data.frame(ncaRes) |>
  dplyr::select(cohort, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

## Comparison against published NCA

Wang 2022 summarises each of these three regimens with a *different* statistic,
so each row below is summarised the same way the paper summarised it -- means
for the 400 mg AOM steady-state exposures (Results, "Model Simulations"), the
median for the 10 mg oral Cmin,ss and the 75th percentile for the 30 mg oral
Cmax,ss (Methods, "Definition of Therapeutic Window"). The statistic used is
named in the table caption and repeated per row below.

```{r nca_compare}
summariseCohort <- function(df, cohort, fn) {
  d <- df[df$cohort == cohort, ]
  tibble::tibble(
    cohort  = cohort,
    cmax    = fn(d$cmax),
    cmin    = fn(d$cmin),
    auclast = fn(d$auclast)
  )
}

simulatedSummary <- dplyr::bind_rows(
  summariseCohort(ncaWide, "AOM 400 mg q28d", mean),
  summariseCohort(ncaWide, "Oral 10 mg QD",   median),
  summariseCohort(ncaWide, "Oral 30 mg QD",   function(x) stats::quantile(x, 0.75))
)

# Published reference values. AUC0-tau,ss 178.5 mg*h/L = 178 500 ng*h/mL.
# The oral rows report only the one bound the paper published for that
# regimen, so the other cells are NA and are dropped from the comparison.
published <- tibble::tribble(
  ~cohort,            ~cmax, ~cmin, ~auclast,
  "AOM 400 mg q28d",  321.5, 195.9, 178500,
  "Oral 10 mg QD",    NA,     94.0, NA,
  "Oral 30 mg QD",    534,    NA,   NA
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulatedSummary |> dplyr::filter(!is.na(cmax)),
  reference = published,
  by        = "cohort",
  units     = c(cmax = "ng/mL", cmin = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated versus published NCA. Each cohort is summarised with the",
    "statistic Wang 2022 used for it: mean (AOM 400 mg), median (oral 10 mg),",
    "75th percentile (oral 30 mg). * marks a difference above 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_assertions}
aomRow <- simulatedSummary[simulatedSummary$cohort == "AOM 400 mg q28d", ]

# Cohort-level checks. These are means over a 200-subject draw with 125 %CV
# IIV on Vc, so they are sized as envelopes on the CENTRE, not on extremes.
stopifnot(
  abs(aomRow$auclast - 178500) / 178500 < 0.30,
  abs(aomRow$cmax    -  321.5) /  321.5 < 0.30,
  abs(aomRow$cmin    -  195.9) /  195.9 < 0.30
)
```

# Exposure-response: time to impending relapse

The exposure-response model is deterministic (no IIV, no residual error), so
every check below is an exact reproduction rather than a statistical one.

```{r er-simulate}
modEr <- readModelDb("Wang_2022_aripiprazole_relapse")

erEvents <- dplyr::bind_rows(
  tibble::tibble(id = 1L, CMIN_ARI =  40, cohort = "Cmin < 95 ng/mL"),
  tibble::tibble(id = 2L, CMIN_ARI = 200, cohort = "Cmin >= 95 ng/mL")
) |>
  tidyr::expand_grid(time = seq(0, 365, by = 1)) |>
  dplyr::mutate(evid = 0L, amt = 0, cmt = "cumhaz")

simEr <- rxode2::rxSolve(modEr, events = erEvents, keep = c("cohort", "CMIN_ARI")) |>
  as.data.frame()
```

```{r figure-6, fig.width = 7, fig.height = 4}
# Replicates Figure 6 of Wang 2022: model-predicted probability of survival
# (no relapse) versus time, for predicted Cmin below and at/above 95 ng/mL.
ggplot(simEr, aes(time, sur, colour = cohort)) +
  geom_line() +
  ylim(0, 1) +
  labs(x = "Time to relapse (days)", y = "Probability of survival (no relapse)",
       colour = NULL,
       title = "Figure 6 - relapse-free survival by predicted Cmin group",
       caption = "Replicates the model-predicted median (blue dashed) curves of Figure 6 of Wang 2022.")
```

```{r er-checks}
hazards <- simEr |>
  dplyr::group_by(cohort) |>
  dplyr::summarise(hazard = unique(round(hazard, 12)),
                   sur350 = sur[time == 350], .groups = "drop")
hazards

hazardRatio <- hazards$hazard[hazards$cohort == "Cmin < 95 ng/mL"] /
  hazards$hazard[hazards$cohort == "Cmin >= 95 ng/mL"]

# Table 2 hazard ratio (95% CI 2.89-6.75). Deterministic identity.
stopifnot(abs(hazardRatio - 4.41) < 0.01)

# Figure 6 read-off at about 350 days: the model-predicted (blue dashed)
# survival is about 0.49 in the [0, 95] panel and about 0.86 in the
# (95, 580] panel.
sur350 <- stats::setNames(hazards$sur350, hazards$cohort)
round(sur350, 3)
stopifnot(
  abs(sur350[["Cmin < 95 ng/mL"]]  - 0.49) < 0.05,
  abs(sur350[["Cmin >= 95 ng/mL"]] - 0.86) < 0.05
)

# The exponential model's own closed form, exp(-t / exp(intercept)).
stopifnot(
  abs(sur350[["Cmin < 95 ng/mL"]]  - exp(-350 / exp(6.256)))         < 1e-6,
  abs(sur350[["Cmin >= 95 ng/mL"]] - exp(-350 / exp(6.256 + 1.484))) < 1e-6
)
```

# Assumptions and deviations

- **Supplement not on disk.** Wang 2022 Tables S1-S4, Figures S1-S3 and
  Appendices S1-S2 were not available for this extraction. Nothing needed for
  either model is in them: Table 1 gives every final popPK estimate and both
  covariate equations, and Table 2 gives both survival parameters. Table S4
  holds the *base* model (before the phase 3 data were added), which the final
  model supersedes; Tables S2/S3 hold per-study covariate summaries, which is
  why the `population` metadata reports the demographic ranges only where the
  main text states them.

- **Virtual-cohort BMI distribution.** Log-normal, median 28 kg/m^2, geometric
  CV 22%. The paper's own virtual population used sex-specific log-normal BMI
  distributions "similar to the phase 3 population" but does not print the
  parameters (they would be in Table S2). The median is anchored on the 28
  kg/m^2 reference value of the paper's AOM Ka equation; the spread was chosen
  so the central 95% spans roughly 18-43 kg/m^2. The paper's observed range,
  15-61 kg/m^2, is wider because it includes cohort extremes. This affects only
  the cohort figures and the cohort NCA row; every structural check in this
  vignette is run at the reference BMI of 28.

- **Sex coding is inverted relative to the paper.** Wang 2022 writes the AOM Ka
  covariate on a `Male` indicator; the canonical column is `SEXF` (1 = female),
  so the model applies `(1 + e_sexf_ka_im * (1 - SEXF))`. This is an exact
  re-expression, verified above by reproducing the paper's own AOM absorption
  half-lives (31.95 versus "about 32" days for women; 23.74 versus "about 24"
  days for men).

- **Survival model reparameterised from AFT to hazard.** See the Source trace
  section. The exponential distribution makes the two parameterisations
  equivalent; the signs are confirmed by reproducing the paper's own 4.41
  hazard ratio and its Figure 6 survival curves.

- **IIV variances back-transformed from %CV.** Wang 2022 Table 1 reports IIV as
  percent coefficients of variation. The model file stores
  `omega^2 = log(CV^2 + 1)`, so the Vc IIV of 124.50 %CV becomes a variance of
  0.936 (omega 0.968) rather than the naive 1.245^2. This is the package
  convention for log-normally distributed random effects and is documented in
  the model file.

- **No IIV correlations.** Wang 2022 Results, "Covariate Analysis", states that
  the off-diagonal covariance elements could not be estimated. The r = 0.901 in
  the Table 1 footnote is a correlation between the *estimates* of `lcl_em` and
  `lfdepot_im` (estimation precision), not an IIV covariance, and is therefore
  deliberately not encoded as an omega block.

- **The CYP2D6 mixture model is not carried over.** For subjects whose CYP2D6
  status was unmeasured, Wang 2022 imputed the phenotype with a NONMEM mixture
  model assuming 90% extensive metabolizers. That is an estimation device for
  missing covariate data, not part of the final structural model, so
  `CYP2D6_PM` is carried here as an ordinary observed covariate.

- **Do not extrapolate below 300 mg AOM.** The model development data set
  contained no AOM dose below 300 mg, and the paper's own external validation
  showed the model does not adequately describe the 50/25 mg arm (median |%PPE|
  41.3%, 75th percentile 74.6%). Wang 2022 excluded that arm from its own
  exposure-response analysis rather than refit.

- **Terminal half-life is checked analytically, not by NCA.** AOM elimination is
  absorption-rate limited (flip-flop), so an NCA of an AOM profile returns the
  absorption half-life of roughly 28 days, not the 7.5-day terminal elimination
  half-life. The check above therefore reads the terminal half-life from the
  disposition rate matrix. The 7.00 days it returns is the typical-subject
  value; the paper's 7.5 days is an average over individual empirical-Bayes
  estimates that includes the 42 poor metabolizers, whose lower clearance gives
  them a longer terminal half-life.

- **Injection site.** Every simulation here represents gluteal administration.
  Injection site was screened as a covariate on AOM Ka and was not retained, but
  only 13 subjects (study CN138020) received arm or thigh injections, so the
  model carries no site term and should not be used to predict a site
  difference. The paper's Discussion proposes bridging to alternative sites via
  the 95 ng/mL Cmin threshold rather than via a PK covariate.

# Errata

No erratum, corrigendum or author correction to Wang 2022 (doi:10.1002/cpdd.1022)
was located.
