---
title: "Entinostat + nivolumab + ipilimumab HER2-negative breast cancer QSP (Wang 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Entinostat + nivolumab + ipilimumab HER2-negative breast cancer QSP (Wang 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r loadmodel}
# Build once and reuse: this is a 120-state model, and every extra build costs
# several seconds.
mod <- rxode2::rxode2(readModelDb("Wang_2020_entinostat_nivolumab_ipilimumab_qsp"))
```

# Model and source

* Citation: `r mod$reference`
* Article: <https://doi.org/10.3389/fbioe.2020.00141>

This is a deterministic quantitative systems pharmacology (QSP) model of
HER2-negative breast cancer under immune checkpoint blockade with nivolumab
(anti-PD-1) and ipilimumab (anti-CTLA-4), combined with the class I histone
deacetylase inhibitor entinostat. It is a member of the Popel-laboratory
immuno-oncology QSP family (see also `Anbari_2023_atezolizumab_cibisatamab_qsp`,
`Ippolito_2024_pacmilimab_qsp` and `Wang_2024_evorpacept_qsp`).

The authors built it in SimBiology and deposited the model with the article:
an SBML export (`Immune_Oncology_Model.xml`) plus a workbook of the full
model listing -- compartments (Supplementary Table S1), parameters (S2),
reactions and rate laws (S3), algebraic rules (S4), species and initial amounts
(S5), the single event (S6) and the MDSC-module parameters (S7). The packaged
model is a translation of the SBML deposit, with every value checked against
the printed tables (the eight places where they differ are listed under
*Assumptions and deviations*).

The paper's own contributions over the earlier platform are:

* an **entinostat PK/PD module** (Figure 2): a fraction `F` of each oral dose is
  absorbed through a zero-order buccal route over `D0`, the rest through a
  lagged first-order gastrointestinal route; distribution to the peripheral,
  tumour and lymph-node compartments; linear plus Michaelis-Menten clearance;
  inhibition of cancer-cell proliferation and of CCL2, nitric-oxide and
  arginase-I production;
* an **MDSC module**: CCL2-driven MDSC recruitment, arginase-I and nitric-oxide
  inhibition of effector T cell (Teff) killing, and arginase-I driven Treg
  expansion;
* **anti-CTLA-4 ADCC**: ipilimumab-bound CTLA-4 on Tregs drives Treg depletion.

There are no etas and no residual-error model. The authors did not fit
inter-individual variability: they generated 1500 virtual patients by Latin
hypercube sampling of the parameter ranges printed in brackets in Table S2, of
which 1196 reached a preselected pre-treatment tumour size. The model is
packaged as a typical-individual mechanism model.

## How the translation works

SimBiology converts units implicitly; rxode2 does not. Every parameter value
was therefore converted to one consistent unit system -- amounts in counts
(the export itself defines 1 cell = 1 molecule = 1/Avogadro mol), volumes in
litres, areas in dm^2, time in days -- and every one of the 154 rate laws, 21
rules and 18 initial assignments was checked by a dimensional-analysis pass
that requires each rate to reduce to amount/time. The ODE states hold
*amounts* (`q_<compartment>_<species>`); the `x_<compartment>_<species>`
variables are the SimBiology species values (concentrations for species that
SimBiology stores as concentrations). Integrating amounts is what SimBiology
does internally and handles the time-varying tumour volume without any
dilution term.

The four drug families (nivolumab, ipilimumab, durvalumab and entinostat,
including the two entinostat depots) are held in **nmol**, so dose records are
in nmol. All other states (cells, surface molecules, cytokines, antigens) are
held in counts.

Model outputs: `tumour_diameter` (cm, from the total tumour volume assuming a
sphere), `Cc_nivolumab`, `Cc_ipilimumab`, `Cc_entinostat` (central
concentrations, nmol/L) and the tumour-infiltrating cell densities
`teff_density`, `treg_density`, `mdsc_density` (cells per mL of tumour).

### Dosing

* Nivolumab, ipilimumab (and durvalumab): IV doses, in nmol, into
  `q_V_C_nivo`, `q_V_C_ipi` (`q_V_C_durv`).
* Entinostat: give the **same** nmol amount twice at each dosing time -- once
  into `q_V_C_ENT_Buccal` with `rate = -2` (the model supplies the zero-order
  duration `durP`) and once into `q_V_C_Dose2`. The model splits the dose with
  `f(q_V_C_ENT_Buccal) <- F_ENT_buccal` and `f(q_V_C_Dose2) <- 1 - F_ENT_buccal`,
  and applies the gastrointestinal lag `alag(q_V_C_Dose2) <- lagP`.

### The pre-treatment phase

As in the paper, a simulation starts from a single cancer cell (`q_V_T_C1(0) =
1`) with no T cells anywhere, and the tumour is grown until it reaches the
chosen pre-treatment diameter; therapy then starts. Because the pre-treatment
phase contains no doses, it is enough to run the untreated model once to find
the time at which the tumour reaches that diameter, and then to schedule the
first dose at that time.

# Population

The virtual population is described in the Methods ("In silico Virtual
Clinical Trial") and Results of Wang 2020. It mirrors the clinical trial
NCT02453620 (nivolumab plus entinostat, with or without ipilimumab, in 26
patients with HER2-negative breast cancer). Baseline parameter values were
estimated from triple-negative breast cancer (TNBC) data and the sampling
ranges from TNBC plus oestrogen-receptor-positive/HER2-negative data. 1196 of
1500 sampled parameter sets reached a pre-treatment tumour diameter drawn from
1.1-4.5 cm (Table S2, `initial_tumour_diameter`), and each regimen was then
simulated for 400 days. The packaged model carries the reference (baseline)
value of every parameter.

# Source trace

| Model element | Source |
|---|---|
| 9 compartments and capacities (`vol_*`) | Supplementary Table S1; SBML `listOfCompartments` |
| 196 `ini()` values: 8 compartment capacities, 187 model parameters and `F_ENT_buccal` | Supplementary Table S2 (value, unit and description in each line's comment); MDSC module also Table S7 |
| Reaction rates `v1`-`v154` | Supplementary Table S3 rows 1-154 (numbered identically; rate laws from the SBML `kineticLaw` elements, which carry the compartment-volume factors SimBiology adds to concentration-per-time rates) |
| Tumour volume `vol_V_T` and the other repeated-assignment rules | Supplementary Table S4 |
| Species, locations and initial amounts; synapse and CTLA-4 initial assignments | Supplementary Table S5 and Table S4 (`initialAssignment` rows) |
| Cancer-cell elimination below 0.9 cells (`c1_alive`) | Supplementary Table S6 |
| Entinostat absorption scheme (buccal zero-order over `durP`, lagged gastrointestinal first-order) | Figure 2A and Methods "Pharmacokinetics and Pharmacodynamics (PK/PD) of Entinostat"; Table S2 `lagP`, `durP`, `k_dose2`, `k_a1_ENT`, `k_a2_ENT` |
| `F_ENT_buccal` = 0.324 | Not printed; back-solved by the maintainers from the Results text (Cmax 15.4 ng/mL after 2 mg/m^2, body surface area 1.7 m^2) -- see below |
| Regimens (nivolumab 3 mg/kg q2w; entinostat 5 mg weekly; ipilimumab 1 mg/kg q6w x 4) | Results, "Efficacy of Anti-PD-1 Monotherapy..." and "Model-Predicted Anti-tumoral Effect in Triple Combination Therapy" |
| Entinostat PK targets (Cmax 15.4/30.8/46.3 ng/mL, AUC 105.5/211.0/316.8 ng h/mL, tmax 0.5 h) | Results, "Prediction of Entinostat Concentration in Tumor"; Figure 2B |

# Simulation set-up

```{r setup-sim}
# Dosing conversions for the vignette (not model parameters): body weight for
# the mg/kg antibody doses and molecular weights for mg -> nmol.
wt_kg <- 70
mw <- c(nivolumab = 146000, ipilimumab = 148000, entinostat = 376.41) # g/mol
nivo_nmol <- 3 * wt_kg / mw[["nivolumab"]] * 1e6
ipi_nmol <- 1 * wt_kg / mw[["ipilimumab"]] * 1e6
ent_nmol <- 5 / mw[["entinostat"]] * 1e6

# Default tolerances reproduce an rtol = 1e-8 solve of the 4000-day untreated
# run to about 1e-4 (relative) in every reported quantity, at about one-tenth
# of the cost. method = "lsoda": the default liblsoda solver took ~55 s on the
# 4 mg/m^2 entinostat profile (identical result, 0.1 s with lsoda).
SOLVE <- function(ev, params = NULL) {
  rxode2::rxSolve(mod, ev, params = params, method = "lsoda", maxsteps = 1e6, returnType = "data.frame")
}

# entinostat dose records: the same amount into both depots
ent_doses <- function(ev, time, amt, ii = 0, addl = 0) {
  ev |>
    rxode2::et(time = time, amt = amt, cmt = "q_V_C_ENT_Buccal", rate = -2, ii = ii, addl = addl) |>
    rxode2::et(time = time, amt = amt, cmt = "q_V_C_Dose2", ii = ii, addl = addl)
}
```

# Replicating Figure 2B: entinostat pharmacokinetics

Single oral doses of 2, 4 and 6 mg/m^2 with a body surface area of 1.7 m^2
(as in the Results). Entinostat kinetics do not depend on the tumour, so the
dose is given at time zero.

```{r ent-pk}
ent_grid <- sort(unique(c(seq(0, 2, by = 0.02), seq(2, 400, by = 0.5)))) / 24 # days
ent_levels <- c(2, 4, 6) # mg/m^2
ent_pk <- bind_rows(lapply(ent_levels, function(lv) {
  amt <- lv * 1.7 / mw[["entinostat"]] * 1e6
  ev <- ent_doses(rxode2::et(ent_grid), time = 0, amt = amt)
  s <- SOLVE(ev)
  data.frame(
    treatment = paste(lv, "mg/m^2"),
    time_h = s$time * 24,
    conc = s$Cc_entinostat * mw[["entinostat"]] / 1000 # nmol/L -> ng/mL
  )
}))
ent_pk$id <- as.integer(factor(ent_pk$treatment, levels = paste(ent_levels, "mg/m^2")))
```

```{r ent-pk-plot, fig.cap = "Replicates Figure 2B of Wang 2020: simulated entinostat plasma concentration after single oral doses of 2, 4 and 6 mg/m^2 (log time axis)."}
ggplot(filter(ent_pk, time_h >= 0.1), aes(time_h, conc, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  scale_x_log10() +
  labs(x = "Time (h)", y = "Entinostat plasma concentration (ng/mL)", colour = NULL) +
  theme_bw()
```

## Non-compartmental analysis

```{r ent-nca}
conc_df <- ent_pk |> select(id, treatment, time = time_h, conc)
dose_df <- conc_df |>
  distinct(id, treatment) |>
  mutate(time = 0, dose = c(2, 4, 6)[id] * 1.7)
o_conc <- PKNCA::PKNCAconc(conc_df, conc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_df, dose ~ time | treatment + id)
intervals <- data.frame(start = 0, end = 400, cmax = TRUE, tmax = TRUE, auclast = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))

reference <- data.frame(
  treatment = paste(ent_levels, "mg/m^2"),
  cmax = c(15.4, 30.8, 46.3),
  tmax = c(0.5, 0.5, 0.5),
  auclast = c(105.5, 211.0, 316.8)
)
tbl <- nlmixr2lib::ncaComparisonTable(
  nca, reference,
  by = "treatment",
  units = c(cmax = "ng/mL", tmax = "h", auclast = "ng*h/mL")
)
knitr::kable(tbl, caption = "Entinostat NCA: Results text of Wang 2020 (Reference) vs the packaged model (Simulated), AUC over 0-400 h (the absorption tail of the gastrointestinal route is complete by then).")
```

```{r ent-nca-check}
nca_res <- as.data.frame(nca$result)
cmax_sim <- nca_res |> filter(PPTESTCD == "cmax") |> arrange(treatment) |> pull(PPORRES)
auc_sim <- nca_res |> filter(PPTESTCD == "auclast") |> arrange(treatment) |> pull(PPORRES)
stopifnot(
  # Cmax is the quantity F_ENT_buccal was back-solved to: it must reproduce
  # 15.4 / 30.8 / 46.3 ng/mL (dose-proportional) to within 1 percent.
  all(abs(cmax_sim / c(15.4, 30.8, 46.3) - 1) < 0.01),
  # Entinostat is linear at these doses (C << Kc_ENT): AUC scales with dose.
  all(abs(auc_sim / auc_sim[1] / c(1, 2, 3) - 1) < 0.01)
)
```

`F_ENT_buccal` is not printed anywhere in the article or supplement (the
authors' dosing scripts are available only on request). Because the plasma
Cmax is set almost entirely by the buccal fraction, it was back-solved from the
2 mg/m^2 Cmax of 15.4 ng/mL; the 4 and 6 mg/m^2 values then follow by dose
proportionality.

**The published AUC is not reproduced.** With the printed parameters the
simulated AUC is about 40 percent of the published 105.5 / 211.0 / 316.8
ng h/mL (starred rows), and the gastrointestinal "hump" of Figure 2B around
10 h is about three times lower than drawn. The simulated tmax is 0.38 h
against the 0.5 h in the text (also starred); the peak falls between the
0.1-h points plotted in Figure 2B, and the reading below moves it to 0.56 h. No value of `F_ENT_buccal` can fix
this: over the dosing interval essentially the whole dose is absorbed, so the
AUC is dose / total clearance whatever the split, and the printed clearance
terms give a low-concentration clearance of about 79 L/h (`k_cln_ENT / Kc_ENT
x V_C` = 78 L/h from the Michaelis-Menten term, plus 0.9 L/h linear), against
about 32 L/h implied by the published AUC.

The published numbers are reproduced *jointly* if the Michaelis-Menten
capacity `k_cln_ENT` is 0.4 times its printed value (4.9e-5 instead of
1.22e-4 mol/L/h, below the printed range of 6e-5 to 1.22e-4) with the buccal
fraction re-solved to 0.184. That reading is shown below as a sensitivity
scenario only. It is **not** used in the packaged model: the printed value is
the authors' deposited parameter, the SBML deposit carries the same value, and
fitting a printed parameter to a reported simulation is not done in this
library.

```{r ent-pk-scenario}
p_kcln <- mod$iniDf$est[mod$iniDf$name == "k_cln_ENT"]
scen <- bind_rows(lapply(ent_levels, function(lv) {
  amt <- lv * 1.7 / mw[["entinostat"]] * 1e6
  s <- SOLVE(
    ent_doses(rxode2::et(ent_grid), time = 0, amt = amt),
    params = c(k_cln_ENT = 0.4 * p_kcln, F_ENT_buccal = 0.184)
  )
  th <- s$time * 24
  cc <- s$Cc_entinostat * mw[["entinostat"]] / 1000
  data.frame(
    treatment = paste(lv, "mg/m^2"),
    cmax = max(cc),
    tmax = th[which.max(cc)],
    auc_0_400 = sum(diff(th) * (head(cc, -1) + tail(cc, -1)) / 2)
  )
}))
scen |>
  dplyr::rename(
    "Dose" = treatment, "Cmax (ng/mL)" = cmax, "tmax (h)" = tmax,
    "AUC0-400 (ng h/mL)" = auc_0_400
  ) |>
  knitr::kable(digits = 2, caption = "Sensitivity scenario (not the packaged model): k_cln_ENT x 0.4 and F_ENT_buccal = 0.184. Published: Cmax 15.4 / 30.8 / 46.3 ng/mL, AUC 105.5 / 211.0 / 316.8 ng h/mL, tmax 0.5 h.")
stopifnot(all(abs(scen$auc_0_400 / c(105.5, 211.0, 316.8) - 1) < 0.02))
```

# Reference virtual patient

The reference parameter set is grown from a single cancer cell to a
pre-treatment diameter of 2.5 cm (inside the 1.1-4.5 cm range of Table S2;
the paper does not give a reference value), then treated for 400 days with
each of the paper's regimens.

```{r growth}
growth <- SOLVE(rxode2::et(seq(0, 4000, by = 5)))
t_star <- approx(growth$tumour_diameter, growth$time, 2.5)$y
t_star
```

```{r growth-plot, fig.cap = "Untreated growth of the reference virtual patient from one cancer cell. The dashed line marks the 2.5 cm pre-treatment diameter."}
ggplot(growth, aes(time / 365.25, tumour_diameter)) +
  geom_line() +
  geom_hline(yintercept = 2.5, linetype = "dashed") +
  scale_y_log10() +
  labs(x = "Time from the first cancer cell (years)", y = "Tumour diameter (cm)") +
  theme_bw()
```

```{r arms}
obs_t <- t_star + seq(0, 400, by = 2)
make_arm <- function(nivo = FALSE, ent = FALSE, ipi = FALSE) {
  ev <- rxode2::et(obs_t)
  if (nivo) ev <- rxode2::et(ev, time = t_star, amt = nivo_nmol, cmt = "q_V_C_nivo", ii = 14, addl = 28)
  if (ent) ev <- ent_doses(ev, time = t_star, amt = ent_nmol, ii = 7, addl = 57)
  if (ipi) ev <- rxode2::et(ev, time = t_star, amt = ipi_nmol, cmt = "q_V_C_ipi", ii = 42, addl = 3)
  ev
}
arm_def <- list(
  "Untreated" = make_arm(),
  "Nivolumab" = make_arm(nivo = TRUE),
  "Nivolumab + entinostat" = make_arm(nivo = TRUE, ent = TRUE),
  "Nivolumab + entinostat + ipilimumab" = make_arm(nivo = TRUE, ent = TRUE, ipi = TRUE)
)
arms <- bind_rows(lapply(names(arm_def), function(a) {
  s <- SOLVE(arm_def[[a]])
  s$arm <- a
  s
}))
arms <- arms |>
  mutate(
    day = time - t_star,
    arm = factor(arm, levels = names(arm_def)),
    pct_change = 100 * (tumour_diameter / 2.5 - 1),
    teff_treg = teff_density / treg_density
  )
```

```{r arms-plot, fig.cap = "Percentage change in tumour diameter of the reference virtual patient under each regimen (compare the spider plots of Figure 3). Dashed lines: RECIST +20 percent (progression) and -30 percent (partial response)."}
ggplot(arms, aes(day, pct_change, colour = arm)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(20, -30), linetype = "dashed") +
  labs(x = "Days from the start of therapy", y = "Change in tumour diameter (%)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

```{r arms-table}
arm_sum <- arms |>
  group_by(arm) |>
  summarise(
    diam_end = tumour_diameter[which.max(day)],
    pct_end = pct_change[which.max(day)],
    teff_treg_d56 = teff_treg[day == 56],
    teff_d56 = teff_density[day == 56],
    treg_d56 = treg_density[day == 56],
    .groups = "drop"
  )
arm_sum |>
  dplyr::rename(
    "Regimen" = arm, "Diameter at day 400 (cm)" = diam_end,
    "Change at day 400 (%)" = pct_end, "Teff/Treg at day 56" = teff_treg_d56,
    "Teff (cells/mL) at day 56" = teff_d56, "Treg (cells/mL) at day 56" = treg_d56
  ) |>
  knitr::kable(digits = 1, caption = "Reference virtual patient: tumour size and tumour-infiltrating T cells by regimen.")
```

The reference parameter set is a progressive-disease patient under every
regimen, which is the most common outcome in the paper's virtual cohort
(74.8 percent progressive disease under nivolumab; 68.9 percent under
nivolumab + entinostat). The mechanistic directions reported in the Results
are reproduced:

* nivolumab raises the tumour-infiltrating Teff density and the Teff/Treg
  ratio (it relieves PD-1 inhibition of Teffs);
* entinostat reduces the MDSC suppression term `H_MDSC_C1` (through its
  inhibition of nitric-oxide and arginase-I production) and slows growth
  slightly;
* adding ipilimumab lowers the Treg density through ADCC and so raises the
  Teff/Treg ratio further, with little extra change in tumour size (the paper:
  "While the number of responders remains the same, the mean post-treatment
  tumor volume is lower than that in the double combination therapy").

```{r arms-check}
a <- setNames(split(arm_sum, arm_sum$arm), levels(arm_sum$arm))
h_mdsc <- arms |>
  filter(day == 200) |>
  select(arm, H_MDSC_C1)
stopifnot(
  # untreated growth: the tumour grows under no therapy
  a[["Untreated"]]$pct_end > 20,
  # PD-1 blockade raises intratumoral Teff and the Teff/Treg ratio (Figure 6)
  a[["Nivolumab"]]$teff_treg_d56 > 1.2 * a[["Untreated"]]$teff_treg_d56,
  # entinostat lowers MDSC-mediated suppression of Teff killing
  h_mdsc$H_MDSC_C1[h_mdsc$arm == "Nivolumab + entinostat"] <
    h_mdsc$H_MDSC_C1[h_mdsc$arm == "Nivolumab"] - 0.05,
  # ipilimumab ADCC lowers tumour Tregs and raises Teff/Treg (Figure 6A, 6C)
  a[["Nivolumab + entinostat + ipilimumab"]]$treg_d56 < 0.97 * a[["Nivolumab + entinostat"]]$treg_d56,
  a[["Nivolumab + entinostat + ipilimumab"]]$teff_treg_d56 > a[["Nivolumab + entinostat"]]$teff_treg_d56
)
```

## Tumour mutational burden (Figures 4, 8 and 9)

The paper identifies tumour mutational burden -- in the model, the number of
tumour-specific T cell clones `n_T1_clones` -- as the strongest predictor of
response (ROC AUC 0.872, Figure 10), with response rising with TMB in Figure 8.
The reference patient is re-run under nivolumab + entinostat at the two ends of
the Table S2 range for `n_T1_clones` (4 and 1840).

```{r tmb}
tmb_case <- function(n_clones) {
  p <- c(n_T1_clones = n_clones)
  g <- SOLVE(rxode2::et(seq(0, 4000, by = 5)), params = p)
  ts <- approx(g$tumour_diameter, g$time, 2.5)$y
  ev <- ent_doses(
    rxode2::et(ts + seq(0, 400, by = 2)) |>
      rxode2::et(time = ts, amt = nivo_nmol, cmt = "q_V_C_nivo", ii = 14, addl = 28),
    time = ts, amt = ent_nmol, ii = 7, addl = 57
  )
  s <- SOLVE(ev, params = p)
  data.frame(
    tmb = paste("n_T1_clones =", n_clones),
    day = s$time - ts,
    pct_change = 100 * (s$tumour_diameter / 2.5 - 1),
    cancer_cells = s$q_V_T_C1
  )
}
tmb <- bind_rows(tmb_case(4), tmb_case(1840))
tmb_end <- tmb |>
  group_by(tmb) |>
  summarise(pct_end = pct_change[which.max(day)], c1_end = cancer_cells[which.max(day)], .groups = "drop")
tmb_end |>
  dplyr::rename("TMB" = tmb, "Change in diameter at day 400 (%)" = pct_end, "Cancer cells at day 400" = c1_end) |>
  knitr::kable(digits = 1, caption = "Nivolumab + entinostat at low and high tumour mutational burden.")
stopifnot(
  tmb_end$pct_end[tmb_end$tmb == "n_T1_clones = 4"] > 20, # progressive disease
  tmb_end$pct_end[tmb_end$tmb == "n_T1_clones = 1840"] < -30, # RECIST response
  tmb_end$c1_end[tmb_end$tmb == "n_T1_clones = 1840"] < 1 # cancer cells eliminated
)
```

```{r tmb-plot, fig.cap = "Nivolumab + entinostat in the reference patient at low and high tumour mutational burden. At high TMB the cancer cells are eliminated; the residual diameter is the minimum tumour-compartment volume plus the remaining T cells and dead-cell debris."}
ggplot(tmb, aes(day, pct_change, colour = tmb)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = c(20, -30), linetype = "dashed") +
  labs(x = "Days from the start of therapy", y = "Change in tumour diameter (%)", colour = NULL) +
  theme_bw()
```

## Nivolumab exposure

```{r nivo-pk}
nivo <- arms |> filter(arm == "Nivolumab", day <= 140)
trough_ss <- nivo$Cc_nivolumab[nivo$day == 140] * mw[["nivolumab"]] / 1e6 # nmol/L -> ug/mL
trough_ss
stopifnot(trough_ss > 20, trough_ss < 120)
```

The simulated nivolumab trough after 10 doses of 3 mg/kg every 2 weeks is
about `r round(trough_ss)` ug/mL. The paper does not report nivolumab
exposure, so this is a plausibility check only (tens of ug/mL for a
therapeutic IgG4 regimen); the antibody PK parameters are those of Table S2,
taken by the authors from Bajaj et al.

## Mass conservation in the synapse

Receptor and ligand totals in each immunological synapse are conserved by the
binding reactions (the antibodies are not consumed: SimBiology's binding
reactions list only the surface species). This is an exact identity of the
ODE system, so it is checked tightly.

```{r conservation}
tri <- arms |> filter(arm == "Nivolumab + entinostat + ipilimumab")
pd1_tot <- with(tri, q_syn_T_C1_PD1 + q_syn_T_C1_PD1_PDL1 + q_syn_T_C1_PD1_PDL2 +
  q_syn_T_C1_PD1_nivo + 2 * q_syn_T_C1_PD1_nivo_PD1)
ctla4_tot <- with(tri, q_syn_T_APC_CTLA4 + q_syn_T_APC_CD80_CTLA4 + 2 * q_syn_T_APC_CTLA4_CD80_CTLA4 +
  2 * q_syn_T_APC_CD80_CTLA4_CD80_CTLA4 + q_syn_T_APC_CD80_CTLA4_CD80 + q_syn_T_APC_CD86_CTLA4 +
  q_syn_T_APC_CD86_CTLA4_CD86 + q_syn_T_APC_CTLA4_ipi + 2 * q_syn_T_APC_CTLA4_ipi_CTLA4)
pd1_0 <- 60000 / 151.31041193043737 * 37.8 # T_PD1_total / A_Tcell * synapse area
ctla4_0 <- 400 # T_CTLA4_syn / A_syn * synapse area
c(pd1 = max(abs(pd1_tot / pd1_0 - 1)), ctla4 = max(abs(ctla4_tot / ctla4_0 - 1)))
stopifnot(max(abs(pd1_tot / pd1_0 - 1)) < 1e-4, max(abs(ctla4_tot / ctla4_0 - 1)) < 1e-4)
```

# Assumptions and deviations

* **Values that differ between Table S2 and the SBML deposit.** Table S2 is
  used, with the SBML value noted in the parameter's comment:
  `k_Treg` = 1/day (SBML: 2/day; Table S2 range 0.1-1, which excludes 2),
  `IC50_ENT_ArgI` = 5e-7 M (SBML: 3.74e-7 M; Table S7 also prints 5e-7 M),
  `k_a2_ENT` = 2.47/h (SBML: 2.5/h). `k_cln_ENT` uses the full-precision
  SBML value 1.22207654715555e-4 M/h, which Table S2 prints rounded as
  1.22e-4.
* **Durvalumab distribution and clearance.** Table S2 prints `q_P_durv`,
  `q_T_durv`, `q_LN_durv` and `k_cl_durv` in 1/second, which is dimensionally
  inconsistent with the rate laws (they multiply a concentration and must give
  an amount per time). The SBML deposit's values (the nivolumab-like volumetric
  flows and a clearance of 0.4 L/day) are used. Durvalumab is not dosed in the
  paper, so this affects no published result.
* **Entinostat buccal fraction.** `F_ENT_buccal` = 0.324 is not printed; it was
  back-solved by the maintainers from the published Cmax. The published AUC
  and the 10-h gastrointestinal peak of Figure 2B are not reproduced with the
  printed clearance (see the entinostat section); a 0.4-fold Michaelis-Menten
  capacity would reproduce them, but that is not the printed value and is not
  used.
* **Cancer-cell elimination event (Table S6).** SimBiology resets the cancer
  cell count to zero when it falls below 0.9 cells. rxode2 has no state-reset
  events, so the model switches proliferation off below 0.9 cells
  (`c1_alive`); the remaining fraction of a cell then decays through the death
  terms instead of vanishing instantly. Tumour regrowth after elimination is
  prevented in both cases.
* **Units.** Arginase-I is carried in the export's placeholder unit mU, which
  SimBiology defines as 1 mol/L; the conversion is kept exactly as exported, so
  `IC50_ArgI_CTL`, `EC50_ArgI_Treg` and `k_sec_ArgI` keep their printed
  relationship.
* **Dosing assumptions of this vignette.** Body weight 70 kg for the mg/kg
  antibody doses (the paper sizes V_C and V_P "based on patient weight" but
  does not print the weight); molecular weights 146 kDa (nivolumab), 148 kDa
  (ipilimumab) and 376.41 g/mol (entinostat); antibodies given as IV boluses
  (the infusion duration is not stated).
* **Reference patient.** The pre-treatment diameter of 2.5 cm is a choice
  within the Table S2 range; the paper's own response rates come from 1196
  sampled parameter sets whose sampling distributions are not stated beyond
  their ranges, so they are not re-simulated here.
* **State names.** The states keep the SimBiology `compartment.species`
  naming (`q_V_T_C1`, `q_syn_T_C1_PD1`, ...) so each one can be traced to its
  Table S5 row; they are not mapped onto the library's PK compartment names.
