---
title: "Teicoplanin (Wang 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Teicoplanin (Wang 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod      <- readModelDb("Wang_2023_teicoplanin")
mod_meta <- rxode2::rxode(mod)$meta
```

* Citation: `r mod_meta$reference`
* Article (DOI): <https://doi.org/10.2147/DDDT.S413662>

This vignette validates the packaged `Wang_2023_teicoplanin` model -- a
two-compartment intravenous-infusion population PK model for teicoplanin
built from 304 plasma concentrations in 108 critically ill adults in the
ICU of Guangdong Provincial People's Hospital. The final model carries a
single covariate: the CKD-EPI estimated glomerular filtration rate (eGFR)
as a **linear** term on clearance, centred at 50 mL/min/1.73 m^2.

The paper's quantitative deliverable is a Monte Carlo dosing-optimization
analysis. Supplementary Tables S2-1 and S2-2 report, for eight candidate
regimens crossed with four renal-function strata, the median (IQR) trough
concentration at 72 h and at 168 h **and** the corresponding probability of
target attainment -- 64 published trough medians and 64 published PTA values.
Those are pure forward predictions of the packaged model, which makes them an
unusually strong validation target, so this vignette reproduces all of them
and compares against the published numbers.

# Population

```{r population}
pop <- rxode2::rxode(mod)$population
```

The Wang 2023 cohort comprised 151 ICU patients contributing 347 plasma
teicoplanin concentrations. The population PK model was built on the
`r pop$n_subjects` patients who each contributed more than two samples
(`r pop$n_concentrations` concentrations); the remaining 43 single-sample
patients formed an external validation group. Median age was 56.5 years
(range 21-91) and median weight 61 kg (range 40-115); 38 of 108 subjects
(35.2%) were female.

The defining feature of the cohort is severe renal impairment: median
CKD-EPI eGFR 29.2 mL/min/1.73 m^2 (range 4.89-170), median Cockcroft-Gault
creatinine clearance 34.8 mL/min, median serum creatinine 161 umol/L, and
43 of 108 subjects (39.8%) receiving continuous renal replacement therapy.
Median baseline APACHE II score was 25.5 (range 10-37). Pneumonia was the
commonest infection site (73 of the recorded sites), followed by
intra-abdominal infection (13) and bacteraemia (10).

Teicoplanin was given by 1-hour intravenous infusion. Concentrations were
measured by HPLC-MS/MS quantifying the teicoplanin A2-2 component, the most
active of the five components of the teicoplanin mixture, as the surrogate
for teicoplanin (LLOQ 1.0 mg/L, linear to 100.0 mg/L).

# Source trace

Every value in the packaged model, and every equation in `model()`, traced
to its location in the source.

```{r source-trace, echo = FALSE}
trace <- tibble::tribble(
  ~Quantity,                    ~Value,      ~`Source location`,
  "CL (typical, eGFR = 50)",    "0.838 L/h", "Table 2, Final Model Estimate (RSE 8.4%); restated in Results and Abstract",
  "Vc",                         "14.4 L",    "Table 2 (RSE 7.2%); Table 2 note 'Vc (L) = 14.4'",
  "Q",                          "3.08 L/h",  "Table 2 (RSE 14.3%); Table 2 note 'Q (L/h) = 3.08'",
  "Vp",                         "51.6 L",    "Table 2 (RSE 10.7%); Table 2 note 'Vp (L) = 51.6 x e^0.387'",
  "theta eGFR-CL",              "0.00823",   "Table 2 (RSE 28.3%); Abstract 'eGFR adjustment factor of 0.00823'",
  "CL covariate equation",      "CL = 0.838 * (1 + (eGFR - 50) * 0.0082)", "Table 2 notes; generic form in Supplementary Table S1 notes",
  "omega CL",                   "37.7% -> var 0.142",  "Table 2, BSV block (RSE 20%)",
  "omega Vc",                   "26.6% -> var 0.0708", "Table 2, BSV block (RSE 43.4%)",
  "omega Q",                    "0 FIX (no eta)",      "Table 2, BSV block; Results 'inter-individual variabilities ... 0%'",
  "omega Vp",                   "62.2% -> var 0.387",  "Table 2, BSV block (RSE 18%); confirmed by 'e^0.387' in the Table 2 note",
  "Proportional residual error","31.7%",     "Table 2, RUV block (RSE 10.6%)",
  "Two-compartment structure",  "central + peripheral1", "Results 'the data showed a better fit to two-compartment model'",
  "IV infusion, 1 h",           "-",         "Methods, 'Dosing Regimen': 'The infusion time was 1 h.'",
  "Exponential IIV, proportional RUV", "-",  "Results 'Exponential and proportional error models were selected'"
)
knitr::kable(trace)
```

Two transcription points are worth recording explicitly.

**The minus sign in `(eGFR - 50)` is not text-extractable.** The Table 2
notes are typeset in a symbol font whose minus glyph is dropped by
`pdftotext` and by the markdown preprocessor, so the equation extracts as
`0:838 * (1 + (eGFR 50) * 0:0082)` with a bare gap. Rendering page 6 of the
PDF to an image resolves it as a subtraction, and the generic covariate form
printed in the Supplementary Table S1 notes,
`P_ij = P_tv,j * [1 + theta_j * (COV - COV_ave)] * e^eta_i`, corroborates it.
The reproduction of 64 published simulation outputs below is a further,
independent confirmation: the alternative readings (`+ 50`, or no centring at
all) move the trough predictions by tens of percent.

**`Vp (L) = 51.6 x e^0.387` in the Table 2 note is not a multiplier.**
Taken literally it would give 76.0 L, contradicting both the Table 2 estimate
and the Results text, which give the typical Vp as 51.6 L. The exponent is the
between-subject variance: 0.622^2 = 0.3869, and 62.2% is exactly the Table 2
`omega Vp`. The same identity fixes the omega convention for the whole table
-- the published percentages are omega (the standard deviation on the log
scale) times 100, not `sqrt(exp(omega^2) - 1)`.

# Deterministic checks

With between-subject and residual variability switched off, the model must
satisfy exact internal identities. These are the tightest gates available
because they contain no Monte Carlo noise.

```{r deterministic}
theta_cl <- 0.838
theta_e  <- 0.00823

# The covariate equation as printed in the Table 2 note.
cl_at <- function(egfr) theta_cl * (1 + (egfr - 50) * theta_e)

grid <- sort(unique(c(seq(0, 12, by = 0.1), seq(12, 48, by = 0.5),
                      seq(48, 336, by = 2))))
ev_single <- rbind(
  data.frame(id = 1L, time = 0, amt = 400, evid = 1L, dur = 1,
             cmt = "central", CRCL = 50),
  data.frame(id = 1L, time = grid, amt = NA_real_, evid = 0L, dur = NA_real_,
             cmt = "central", CRCL = 50)
)
ev_single <- ev_single[order(ev_single$time, -ev_single$evid), ]

sim_single <- rxode2::rxSolve(mod, ev_single, omega = NA,
                              returnType = "data.frame")
sim_single$id <- 1L
```

```{r pknca-single}
conc_obj <- PKNCA::PKNCAconc(dplyr::filter(sim_single, !is.na(Cc)),
                             Cc ~ time | id)
dose_obj <- PKNCA::PKNCAdose(data.frame(id = 1L, time = 0, dose = 400),
                             dose ~ time | id)
nca_single <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                         auclast = TRUE, aucinf.obs = TRUE,
                         half.life = TRUE, cl.obs = TRUE)
))
nca_val <- function(code) {
  v <- as.data.frame(nca_single$result)
  out <- v$PPORRES[v$PPTESTCD == code]
  if (length(out) != 1L) stop("no unique NCA row for '", code, "'")
  out
}

# Analytic terminal half-life of the two-compartment system.
k10 <- theta_cl / 14.4
k12 <- 3.08 / 14.4
k21 <- 3.08 / 51.6
beta_rate <- (k10 + k12 + k21 -
              sqrt((k10 + k12 + k21)^2 - 4 * k10 * k21)) / 2

det_tab <- tibble::tibble(
  Identity = c("AUC(0-inf) after a single 400 mg dose = Dose / CL",
               "PKNCA CL(obs) = model CL",
               "PKNCA terminal half-life = analytic beta half-life"),
  Expected = c(400 / theta_cl, theta_cl, log(2) / beta_rate),
  Observed = c(nca_val("aucinf.obs"), nca_val("cl.obs"), nca_val("half.life"))
) |>
  dplyr::mutate(`% diff` = 100 * (Observed - Expected) / Expected)
knitr::kable(det_tab, digits = c(0, 4, 4, 5))

# These are numerical-integration identities with no cohort sampling, so an
# exact bound is correct here (pattern: tighten deterministic gates). The
# realised deviations are ~0.006% (AUC, CL) and ~0.3% (half-life, which is a
# log-linear regression over a finite grid rather than a closed form).
stopifnot(
  abs(det_tab$`% diff`[1]) < 0.05,
  abs(det_tab$`% diff`[2]) < 0.05,
  abs(det_tab$`% diff`[3]) < 1
)
```

The AUC identity is the single most informative gate on the transcription:
it fails if the clearance value, the dose units, the concentration units or
the compartment scaling is wrong by any amount.

```{r cl-covariate-check}
cov_tab <- tibble::tibble(
  `eGFR (mL/min/1.73 m^2)` = c(4.89, 29.2, 30, 50, 60, 90, 130, 170),
  `CL (L/h)`               = cl_at(c(4.89, 29.2, 30, 50, 60, 90, 130, 170)),
  Note = c("cohort minimum", "cohort median", "stratum 1 boundary",
           "centring value: CL = theta_CL exactly", "stratum 2 boundary",
           "stratum 3 boundary", "stratum 4 boundary", "cohort maximum")
)
knitr::kable(cov_tab, digits = 3)

# At the centring value the covariate term is exactly 1.
stopifnot(abs(cl_at(50) - theta_cl) < 1e-12)
# The linear form stays positive across and well beyond the observed range.
stopifnot(cl_at(0) > 0)
```

# Recovering the published simulation design

The paper does not state what eGFR value was used inside each renal-function
stratum, nor whether the reported trough concentrations include residual
error. Both choices materially change the predictions, so rather than guess
we identify them from the 64 published trough medians.

```{r sim-helpers}
n_sub <- 200  # participants per arm

make_events <- function(load_dose, n_load, maint_dose, egfr, n = n_sub,
                        tmax = 180, obs_times = c(72, 168)) {
  load_times  <- seq(0, by = 12, length.out = n_load)
  # Maintenance starts one full 24 h interval after the last loading dose.
  # This is the only schedule that places a trough exactly at both 72 h and
  # 168 h, which is how the paper defines its two endpoints, for BOTH the
  # q12h x 3 and q12h x 5 loading arms.
  maint_times <- seq(max(load_times) + 24, tmax, by = 24)
  do.call(rbind, lapply(seq_len(n), function(i) {
    dosing <- data.frame(
      id = i, time = c(load_times, maint_times),
      amt = c(rep(load_dose, n_load), rep(maint_dose, length(maint_times))),
      evid = 1L, dur = 1, cmt = "central", CRCL = egfr
    )
    obs <- data.frame(
      id = i, time = obs_times, amt = NA_real_, evid = 0L, dur = NA_real_,
      cmt = "central", CRCL = egfr
    )
    x <- rbind(dosing, obs)
    x[order(x$time, -x$evid), ]
  }))
}

regimens <- tibble::tribble(
  ~regimen,                    ~load, ~n_load, ~maint, ~target, ~infection,
  "400 mg q12hx3, 400 mg qd",  400,   3L,      400,    10,      "Gram-positive",
  "400 mg q12hx5, 400 mg qd",  400,   5L,      400,    10,      "Gram-positive",
  "600 mg q12hx3, 600 mg qd",  600,   3L,      600,    10,      "Gram-positive",
  "600 mg q12hx5, 600 mg qd",  600,   5L,      600,    10,      "Gram-positive",
  "800 mg q12hx3, 400 mg qd",  800,   3L,      400,    15,      "Deep-seated",
  "800 mg q12hx3, 600 mg qd",  800,   3L,      600,    15,      "Deep-seated",
  "800 mg q12hx3, 800 mg qd",  800,   3L,      800,    15,      "Deep-seated",
  "800 mg q12hx5, 800 mg qd",  800,   5L,      800,    15,      "Deep-seated"
)

strata_levels <- c("eGFR <= 30", "eGFR 30-60", "eGFR 60-90", "eGFR 90-130")

# Run every regimen x stratum cell at one eGFR value and return the
# simulated 72 h and 168 h troughs. `Cc` is the individual prediction;
# `sim` additionally carries the proportional residual error.
sweep_egfr <- function(egfr_by_stratum) {
  rxode2::rxSetSeed(42)
  out <- list()
  for (k in seq_along(strata_levels)) {
    for (j in seq_len(nrow(regimens))) {
      r <- regimens[j, ]
      s <- rxode2::rxSolve(
        mod,
        make_events(r$load, r$n_load, r$maint, egfr_by_stratum[k]),
        returnType = "data.frame"
      )
      out[[length(out) + 1L]] <- tibble::tibble(
        stratum = strata_levels[k], regimen = r$regimen,
        infection = r$infection, target = r$target,
        time = rep(c(72, 168), each = n_sub),
        ipred = c(s$Cc[s$time == 72], s$Cc[s$time == 168]),
        obs   = c(s$sim[s$time == 72], s$sim[s$time == 168])
      )
    }
  }
  dplyr::bind_rows(out)
}
```

```{r published-tables, echo = FALSE}
# Wang 2023 Supplementary Tables S2-1 and S2-2, transcribed verbatim.
published <- tibble::tribble(
  ~stratum,      ~regimen,                   ~time, ~med,  ~q25,  ~q75,  ~pta,
  "eGFR <= 30",  "400 mg q12hx3, 400 mg qd",    72,  10.5,  7.1,  14.4,  52.8,
  "eGFR <= 30",  "400 mg q12hx5, 400 mg qd",    72,  13.7,  9.4,  18.9,  71.2,
  "eGFR <= 30",  "600 mg q12hx3, 600 mg qd",    72,  15.7, 10.6,  21.9,  77.9,
  "eGFR <= 30",  "600 mg q12hx5, 600 mg qd",    72,  20.7, 14.3,  29.2,  87.0,
  "eGFR <= 30",  "800 mg q12hx3, 400 mg qd",    72,  16.9, 11.7,  24.7,  59.2,
  "eGFR <= 30",  "800 mg q12hx3, 600 mg qd",    72,  18.7, 13.0,  26.1,  66.8,
  "eGFR <= 30",  "800 mg q12hx3, 800 mg qd",    72,  21.2, 14.9,  29.3,  74.3,
  "eGFR <= 30",  "800 mg q12hx5, 800 mg qd",    72,  27.1, 18.7,  37.3,  85.4,
  "eGFR 30-60",  "400 mg q12hx3, 400 mg qd",    72,   8.9,  6.1,  12.0,  39.4,
  "eGFR 30-60",  "400 mg q12hx5, 400 mg qd",    72,  11.0,  7.5,  15.7,  56.9,
  "eGFR 30-60",  "600 mg q12hx3, 600 mg qd",    72,  13.4,  8.8,  18.4,  68.8,
  "eGFR 30-60",  "600 mg q12hx5, 600 mg qd",    72,  16.6, 11.0,  23.3,  80.3,
  "eGFR 30-60",  "800 mg q12hx3, 400 mg qd",    72,  13.6,  9.3,  19.7,  43.5,
  "eGFR 30-60",  "800 mg q12hx3, 600 mg qd",    72,  15.4, 10.6,  22.7,  52.2,
  "eGFR 30-60",  "800 mg q12hx3, 800 mg qd",    72,  17.0, 11.8,  23.7,  59.2,
  "eGFR 30-60",  "800 mg q12hx5, 800 mg qd",    72,  22.8, 15.3,  33.6,  76.2,
  "eGFR 60-90",  "400 mg q12hx3, 400 mg qd",    72,   7.1,  4.7,  10.1,  26.0,
  "eGFR 60-90",  "400 mg q12hx5, 400 mg qd",    72,   9.2,  6.2,  13.5,  43.6,
  "eGFR 60-90",  "600 mg q12hx3, 600 mg qd",    72,  10.9,  7.2,  15.3,  55.7,
  "eGFR 60-90",  "600 mg q12hx5, 600 mg qd",    72,  13.5,  9.2,  20.4,  70.2,
  "eGFR 60-90",  "800 mg q12hx3, 400 mg qd",    72,  11.4,  7.5,  16.6,  31.7,
  "eGFR 60-90",  "800 mg q12hx3, 600 mg qd",    72,  13.4,  8.3,  18.9,  40.2,
  "eGFR 60-90",  "800 mg q12hx3, 800 mg qd",    72,  14.0,  9.6,  19.6,  44.2,
  "eGFR 60-90",  "800 mg q12hx5, 800 mg qd",    72,  18.8, 12.3,  27.1,  64.9,
  "eGFR 90-130", "400 mg q12hx3, 400 mg qd",    72,   5.5,  3.6,   8.2,  13.8,
  "eGFR 90-130", "400 mg q12hx5, 400 mg qd",    72,   7.5,  5.1,  11.1,  31.1,
  "eGFR 90-130", "600 mg q12hx3, 600 mg qd",    72,   8.7,  5.6,  12.7,  40.4,
  "eGFR 90-130", "600 mg q12hx5, 600 mg qd",    72,  11.2,  7.2,  16.1,  57.7,
  "eGFR 90-130", "800 mg q12hx3, 400 mg qd",    72,   8.9,  6.0,  13.4,  18.7,
  "eGFR 90-130", "800 mg q12hx3, 600 mg qd",    72,  10.3,  6.7,  14.9,  24.6,
  "eGFR 90-130", "800 mg q12hx3, 800 mg qd",    72,  10.9,  7.4,  16.5,  30.4,
  "eGFR 90-130", "800 mg q12hx5, 800 mg qd",    72,  14.0,  9.4,  21.8,  47.8,
  "eGFR <= 30",  "400 mg q12hx3, 400 mg qd",   168,  13.9, 10.0,  20.1,  74.9,
  "eGFR <= 30",  "400 mg q12hx5, 400 mg qd",   168,  15.0, 10.0,  21.0,  75.2,
  "eGFR <= 30",  "600 mg q12hx3, 600 mg qd",   168,  20.6, 14.1,  28.6,  90.0,
  "eGFR <= 30",  "600 mg q12hx5, 600 mg qd",   168,  22.8, 15.1,  31.8,  91.9,
  "eGFR <= 30",  "800 mg q12hx3, 400 mg qd",   168,  16.5, 11.5,  23.2,  57.9,
  "eGFR <= 30",  "800 mg q12hx3, 600 mg qd",   168,  21.5, 15.5,  29.7,  76.3,
  "eGFR <= 30",  "800 mg q12hx3, 800 mg qd",   168,  28.3, 19.8,  38.3,  86.8,
  "eGFR <= 30",  "800 mg q12hx5, 800 mg qd",   168,  30.4, 21.0,  41.6,  88.0,
  "eGFR 30-60",  "400 mg q12hx3, 400 mg qd",   168,  10.8,  7.5,  15.4,  55.9,
  "eGFR 30-60",  "400 mg q12hx5, 400 mg qd",   168,  11.5,  7.7,  17.0,  60.7,
  "eGFR 30-60",  "600 mg q12hx3, 600 mg qd",   168,  16.2, 11.2,  23.9,  80.5,
  "eGFR 30-60",  "600 mg q12hx5, 600 mg qd",   168,  17.2, 11.9,  24.9,  83.2,
  "eGFR 30-60",  "800 mg q12hx3, 400 mg qd",   168,  12.2,  8.0,  17.9,  36.2,
  "eGFR 30-60",  "800 mg q12hx3, 600 mg qd",   168,  17.3, 11.7,  24.5,  60.4,
  "eGFR 30-60",  "800 mg q12hx3, 800 mg qd",   168,  21.6, 14.5,  29.4,  73.5,
  "eGFR 30-60",  "800 mg q12hx5, 800 mg qd",   168,  23.6, 15.5,  34.4,  76.6,
  "eGFR 60-90",  "400 mg q12hx3, 400 mg qd",   168,   8.3,  5.5,  12.0,  37.7,
  "eGFR 60-90",  "400 mg q12hx5, 400 mg qd",   168,   9.3,  6.0,  13.6,  45.4,
  "eGFR 60-90",  "600 mg q12hx3, 600 mg qd",   168,  13.0,  8.5,  18.5,  66.4,
  "eGFR 60-90",  "600 mg q12hx5, 600 mg qd",   168,  13.9,  9.1,  20.3,  69.6,
  "eGFR 60-90",  "800 mg q12hx3, 400 mg qd",   168,   9.8,  6.5,  14.8,  23.2,
  "eGFR 60-90",  "800 mg q12hx3, 600 mg qd",   168,  13.5,  9.0,  20.4,  43.6,
  "eGFR 60-90",  "800 mg q12hx3, 800 mg qd",   168,  16.4, 11.2,  23.8,  56.3,
  "eGFR 60-90",  "800 mg q12hx5, 800 mg qd",   168,  18.6, 12.3,  26.4,  65.8,
  "eGFR 90-130", "400 mg q12hx3, 400 mg qd",   168,   6.5,  4.0,   9.7,  23.4,
  "eGFR 90-130", "400 mg q12hx5, 400 mg qd",   168,   6.9,  4.5,  10.3,  27.2,
  "eGFR 90-130", "600 mg q12hx3, 600 mg qd",   168,   9.9,  6.4,  14.7,  48.4,
  "eGFR 90-130", "600 mg q12hx5, 600 mg qd",   168,  10.5,  6.9,  15.1,  53.4,
  "eGFR 90-130", "800 mg q12hx3, 400 mg qd",   168,   7.2,  4.5,  10.7,   9.9,
  "eGFR 90-130", "800 mg q12hx3, 600 mg qd",   168,  10.0,  6.4,  14.9,  24.5,
  "eGFR 90-130", "800 mg q12hx3, 800 mg qd",   168,  12.9,  8.9,  19.1,  40.9,
  "eGFR 90-130", "800 mg q12hx5, 800 mg qd",   168,  13.8,  8.8,  20.4,  44.8
) |>
  dplyr::mutate(stratum = factor(stratum, levels = strata_levels))
stopifnot(nrow(published) == 64L)
```

## Which eGFR was used inside each stratum?

The two natural candidates are the midpoint of each band and its upper
bound. They are easy to tell apart: across the eGFR range the model's trough
predictions change by roughly a factor of two, so a 20-40 mL/min/1.73 m^2
difference in the assumed eGFR is worth 10-20% on every cell.

```{r design-egfr}
summarise_medians <- function(sw, value_col) {
  sw |>
    dplyr::group_by(stratum, regimen, time) |>
    dplyr::summarise(sim = stats::median(.data[[value_col]]), .groups = "drop") |>
    dplyr::inner_join(published, by = c("stratum", "regimen", "time")) |>
    dplyr::mutate(pct = 100 * (sim - med) / med)
}

sweep_mid  <- sweep_egfr(c(17.4, 45, 75, 110))
sweep_edge <- sweep_egfr(c(30, 60, 90, 130))

design_tab <- dplyr::bind_rows(
  summarise_medians(sweep_mid,  "ipred") |>
    dplyr::mutate(scheme = "stratum midpoint (17.4 / 45 / 75 / 110)"),
  summarise_medians(sweep_edge, "ipred") |>
    dplyr::mutate(scheme = "stratum upper bound (30 / 60 / 90 / 130)")
) |>
  dplyr::group_by(scheme) |>
  dplyr::summarise(
    `n cells`             = dplyr::n(),
    `median |% diff|`     = stats::median(abs(pct)),
    `90th pctile |% diff|`= stats::quantile(abs(pct), 0.9),
    `median % diff`       = stats::median(pct),
    .groups = "drop"
  )
knitr::kable(design_tab, digits = 1)
```

Simulating each stratum at its **upper bound** reproduces all 64 published
trough medians with a median absolute deviation of a few percent, whereas the
midpoint assumption leaves a systematic bias of roughly +16% on every cell.
The upper bound is also the sensible design choice for a paper whose purpose
is to recommend a regimen for a whole band: within a band the highest eGFR
gives the highest clearance and hence the lowest trough, so it is the worst
case the regimen must cover.

```{r design-egfr-gate}
edge_row <- design_tab[grepl("upper", design_tab$scheme), ]
mid_row  <- design_tab[grepl("midpoint", design_tab$scheme), ]
# The discrimination is large (a few percent against roughly sixteen), so a
# factor-of-two separation is a bound with real headroom rather than one read
# off a single run.
stopifnot(
  edge_row$`median |% diff|` < 10,
  edge_row$`median |% diff|` < 0.5 * mid_row$`median |% diff|`
)
```

All remaining comparisons use the stratum upper bound.

## Do the published values carry residual error?

The individual prediction and the simulated observation have the same median
-- the residual error is proportional and symmetric about zero -- so the
trough medians cannot distinguish them. The interquartile **width** can: the
proportional residual error is 31.7%, which widens the IQR appreciably.

```{r design-ruv}
iqr_ratio <- function(sw, value_col) {
  sw |>
    dplyr::group_by(stratum, regimen, time) |>
    dplyr::summarise(
      ratio = stats::quantile(.data[[value_col]], 0.75) /
              stats::quantile(.data[[value_col]], 0.25),
      .groups = "drop"
    )
}
pub_ratio <- published |> dplyr::mutate(ratio = q75 / q25)

ruv_tab <- tibble::tibble(
  Quantity = c("Published (Tables S2-1, S2-2)",
               "Simulated, individual prediction only",
               "Simulated, with proportional residual error"),
  `Median IQR width ratio (q75/q25)` = c(
    stats::median(pub_ratio$ratio),
    stats::median(iqr_ratio(sweep_edge, "ipred")$ratio),
    stats::median(iqr_ratio(sweep_edge, "obs")$ratio)
  )
)
knitr::kable(ruv_tab, digits = 2)

# The published spread matches the residual-error-inclusive simulation and is
# clearly wider than the individual predictions alone. Published ratio 2.146;
# realised simulated ratios across 2 / 4 / 16 solver threads were
# 2.162 / 2.125 / 2.13 with residual error and 1.820 / 1.803 / 1.81 without,
# so the two candidates are separated by ~0.33 and neither bound is read off
# a single run.
pub_r <- ruv_tab$`Median IQR width ratio (q75/q25)`[1]
ip_r  <- ruv_tab$`Median IQR width ratio (q75/q25)`[2]
ob_r  <- ruv_tab$`Median IQR width ratio (q75/q25)`[3]
stopifnot(abs(ob_r - pub_r) < 0.25, ip_r < pub_r - 0.10)
```

The published interquartile widths match the simulated **observations**, so
the paper's Monte Carlo reports concentrations that include the proportional
residual error. The PTA comparisons below therefore use the same quantity.

# Replicating Supplementary Tables S2-1 and S2-2

```{r replicate-troughs}
replication <- sweep_edge |>
  dplyr::group_by(stratum, regimen, infection, target, time) |>
  dplyr::summarise(
    sim_med = stats::median(obs),
    sim_pta = 100 * mean(obs >= dplyr::first(target)),
    .groups = "drop"
  ) |>
  dplyr::inner_join(published, by = c("stratum", "regimen", "time")) |>
  dplyr::mutate(
    med_pct = 100 * (sim_med - med) / med,
    pta_pp  = sim_pta - pta
  )
stopifnot(nrow(replication) == 64L)
```

```{r replicate-summary, echo = FALSE}
summary_tab <- tibble::tibble(
  Comparison = c("Median trough concentration (% difference)",
                 "Probability of target attainment (percentage points)"),
  `n cells`  = c(64L, 64L),
  Median     = c(stats::median(abs(replication$med_pct)),
                 stats::median(abs(replication$pta_pp))),
  `90th pctile` = c(stats::quantile(abs(replication$med_pct), 0.9),
                    stats::quantile(abs(replication$pta_pp), 0.9)),
  Maximum    = c(max(abs(replication$med_pct)), max(abs(replication$pta_pp)))
)
knitr::kable(summary_tab, digits = 1)
```

```{r replicate-gate}
# Bounds are set well outside what a single cohort draw gives. rxSetSeed()
# fixes the rxode2 stream per solver-thread count, so a CI runner draws a
# different cohort than a development box; these statistics are aggregates
# over 64 cells, which damps but does not remove that. Measured across
# 2 / 4 / 16 solver threads when authored:
#   trough |% diff|  median 3.98 / 4.11 / 4.40, 90th pctile 8.99 / 9.41 / 9.20
#   PTA |pp diff|    median 2.05 / 1.70 / 2.00, 90th pctile 4.92 / 5.68 / 6.00
# The bounds below sit at roughly three times the observed medians and 2.5
# times the observed 90th percentiles. A mis-transcribed clearance, dose,
# volume or covariate sign moves the troughs by tens of percent and the PTAs
# by tens of points, so they still go red on any real error -- the ruled-out
# stratum-midpoint design alone lands at a median of ~16%.
stopifnot(
  stats::median(abs(replication$med_pct))         < 12,
  stats::quantile(abs(replication$med_pct), 0.9)  < 25,
  stats::median(abs(replication$pta_pp))          < 6,
  stats::quantile(abs(replication$pta_pp), 0.9)   < 14
)
```

The full 72 h comparison, one row per regimen and stratum:

```{r trough-table-72, echo = FALSE}
replication |>
  dplyr::filter(time == 72) |>
  dplyr::select(stratum, regimen, med, sim_med, med_pct, pta, sim_pta, pta_pp) |>
  dplyr::arrange(stratum, regimen) |>
  dplyr::rename(
    "Stratum" = stratum, "Regimen" = regimen,
    "Published Cmin (mg/L)" = med, "Simulated Cmin (mg/L)" = sim_med,
    "% diff" = med_pct,
    "Published PTA (%)" = pta, "Simulated PTA (%)" = sim_pta,
    "pp diff" = pta_pp
  ) |>
  knitr::kable(digits = 1,
               caption = "Replicates Wang 2023 Supplementary Tables S2-1 and S2-2, Cmin 72 h rows.")
```

And at 168 h:

```{r trough-table-168, echo = FALSE}
replication |>
  dplyr::filter(time == 168) |>
  dplyr::select(stratum, regimen, med, sim_med, med_pct, pta, sim_pta, pta_pp) |>
  dplyr::arrange(stratum, regimen) |>
  dplyr::rename(
    "Stratum" = stratum, "Regimen" = regimen,
    "Published Css (mg/L)" = med, "Simulated Css (mg/L)" = sim_med,
    "% diff" = med_pct,
    "Published PTA (%)" = pta, "Simulated PTA (%)" = sim_pta,
    "pp diff" = pta_pp
  ) |>
  knitr::kable(digits = 1,
               caption = "Replicates Wang 2023 Supplementary Tables S2-1 and S2-2, Css 168 h rows.")
```

# Replicating Figures 3 and 4

```{r figure-3-4, echo = FALSE, fig.height = 6}
replication |>
  dplyr::mutate(
    endpoint = ifelse(time == 72, "Cmin 72 h", "Css 168 h"),
    panel = paste0(infection, " (target ", target, " mg/L)")
  ) |>
  tidyr::pivot_longer(c(pta, sim_pta), names_to = "source", values_to = "PTA") |>
  dplyr::mutate(source = ifelse(source == "pta", "Published", "Simulated")) |>
  ggplot(aes(x = stratum, y = PTA, colour = regimen, linetype = source,
             group = interaction(regimen, source))) +
  geom_line() +
  geom_point(size = 1) +
  geom_hline(yintercept = 50, linetype = "dotted") +
  facet_grid(endpoint ~ panel) +
  labs(x = NULL, y = "Probability of target attainment (%)",
       colour = "Regimen", linetype = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "bottom", legend.box = "vertical") +
  guides(colour = guide_legend(ncol = 2))
```

Replicates Figures 3 (Gram-positive panel, dotted 50% line, 10 mg/L target)
and 4 (deep-seated panel, 15 mg/L target) of Wang 2023. Solid and dashed
lines for the same regimen overlay closely across all four strata.

# Replicating Table 3: the recommended regimens

Table 3 of the paper names one regimen per infection type per stratum. The
check here is that the packaged model reproduces the published PTA for each
of those eight recommended cells.

```{r table-3}
recommended <- tibble::tribble(
  ~stratum,      ~infection,      ~regimen,
  "eGFR <= 30",  "Gram-positive", "400 mg q12hx3, 400 mg qd",
  "eGFR 30-60",  "Gram-positive", "400 mg q12hx5, 400 mg qd",
  "eGFR 60-90",  "Gram-positive", "600 mg q12hx3, 600 mg qd",
  "eGFR 90-130", "Gram-positive", "600 mg q12hx5, 600 mg qd",
  "eGFR <= 30",  "Deep-seated",   "800 mg q12hx3, 400 mg qd",
  "eGFR 30-60",  "Deep-seated",   "800 mg q12hx3, 600 mg qd",
  "eGFR 60-90",  "Deep-seated",   "800 mg q12hx3, 800 mg qd",
  "eGFR 90-130", "Deep-seated",   "800 mg q12hx5, 800 mg qd"
) |>
  dplyr::mutate(stratum = factor(stratum, levels = strata_levels))

rec_tab <- recommended |>
  dplyr::inner_join(replication, by = c("stratum", "infection", "regimen")) |>
  dplyr::select(stratum, infection, regimen, time, pta, sim_pta, pta_pp) |>
  dplyr::arrange(infection, stratum, time)
stopifnot(nrow(rec_tab) == 16L)

rec_tab |>
  dplyr::rename(
    "Stratum" = stratum, "Infection type" = infection, "Regimen" = regimen,
    "Endpoint (h)" = time, "Published PTA (%)" = pta,
    "Simulated PTA (%)" = sim_pta, "pp diff" = pta_pp
  ) |>
  knitr::kable(digits = 1, caption = "Replicates Wang 2023 Table 3.")

stopifnot(stats::median(abs(rec_tab$pta_pp)) < 8)
```

Four of the eight recommended cells fall below the paper's own stated
"optimal if PTA > 50%" criterion at one or both endpoints -- for example
deep-seated infection at eGFR 60-90 (published 44.2% at 72 h) and at
eGFR 90-130 (published 47.8% at 72 h and 44.8% at 168 h). The paper presents
these as the best available option rather than as regimens meeting the
threshold; the Discussion notes that still-higher regimens reported elsewhere
(15 mg/kg or 1000 mg) were judged not clinically applicable.

# NCA of the recommended standard regimen

The paper reports observed plasma Cmax and Cmin for the model-building
cohort (Table 1). Those are compared here against a PKNCA analysis of a
simulated cohort receiving the paper's own group-1 standard regimen
(400 mg q12h x 3, then 400 mg q24h) at the cohort median eGFR.

```{r nca-cohort}
obs_grid <- sort(unique(c(seq(0, 180, by = 1), 145, 169)))
ev_cohort <- do.call(rbind, lapply(seq_len(n_sub), function(i) {
  dosing <- data.frame(
    id = i, time = c(0, 12, 24, seq(48, 180, by = 24)),
    amt = 400, evid = 1L, dur = 1, cmt = "central", CRCL = 29.2
  )
  obs <- data.frame(id = i, time = obs_grid, amt = NA_real_, evid = 0L,
                    dur = NA_real_, cmt = "central", CRCL = 29.2)
  x <- rbind(dosing, obs)
  x[order(x$time, -x$evid), ]
}))
rxode2::rxSetSeed(42)
sim_cohort <- rxode2::rxSolve(mod, ev_cohort, returnType = "data.frame")
if (is.null(sim_cohort$id)) sim_cohort$id <- 1L
```

```{r nca-cohort-run}
# One steady-state-approaching dosing interval, 144-168 h. The filter is
# !is.na(Cc) only, so the interval-start record is retained.
nca_input <- sim_cohort |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(conc = sim)

conc_cohort <- PKNCA::PKNCAconc(nca_input, conc ~ time | id)
dose_cohort <- PKNCA::PKNCAdose(
  data.frame(id = seq_len(n_sub), time = 144, dose = 400), dose ~ time | id
)
nca_cohort <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_cohort, dose_cohort,
  intervals = data.frame(start = 144, end = 168,
                         cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE)
))
nca_res <- as.data.frame(nca_cohort$result)
stopifnot(nrow(nca_res) > 0L)
```

```{r nca-comparison}
reference <- data.frame(
  cmax = 39.9,   # Wang 2023 Table 1, observed plasma Cmax, model-establishment group
  cmin = 10.5    # Wang 2023 Table 1, observed plasma Cmin, model-establishment group
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  params    = c("cmax", "cmin"),
  units     = c(cmax = "mg/L", cmin = "mg/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, digits = 1)
attr(cmp, "footnote")
```

This comparison is deliberately **not** gated, and the Cmax row is expected
to be flagged. The two sides are not measuring the same thing: the published
Cmax and Cmin are per-patient extremes of two or three sparse clinical
samples pooled across all four dosing groups -- including the 91 patients on
irregular doses at irregular intervals -- whereas the simulated values come
from a dense grid under a single fixed 400 mg regimen at one eGFR. The Cmin
comparison is nevertheless close, which is the more informative of the two
because the paper's targets, its Monte Carlo endpoints and its dosing
recommendations are all trough-based.

```{r profile-figure, echo = FALSE}
sim_cohort |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = stats::quantile(sim, 0.05), p50 = stats::median(sim),
    p95 = stats::quantile(sim, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(x = time)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line(aes(y = p50)) +
  geom_hline(yintercept = c(10, 15), linetype = "dashed") +
  geom_vline(xintercept = c(72, 168), linetype = "dotted") +
  labs(x = "Time after first dose (h)", y = "Teicoplanin (mg/L)") +
  theme_bw()
```

Simulated concentration-time profile (median and 5th-95th percentiles,
n = `r n_sub`) for the standard regimen at the cohort median eGFR of 29.2
mL/min/1.73 m^2. Dashed lines mark the 10 and 15 mg/L trough targets; dotted
lines mark the paper's two endpoints at 72 h and 168 h. The long terminal
half-life (`r round(log(2) / beta_rate, 1)` h) is why the 168 h trough is
still appreciably above the 72 h trough: the profile has not reached true
steady state within the simulated week.

# Assumptions and deviations

* **Within-stratum eGFR is an inference, not a published value.** The paper
  reports simulations "stratified by eGFR" but never states the eGFR used
  inside each band. The stratum upper bound (30 / 60 / 90 / 130
  mL/min/1.73 m^2) is identified above by reproducing the 64 published
  trough medians, and is what this vignette uses. The stratum midpoint is
  ruled out by a systematic +16% bias.
* **Residual error is included in the reproduced trough distributions.** Also
  an inference, identified from the published interquartile widths. The
  medians are insensitive to this choice; the PTAs are not.
* **The maintenance schedule starts 24 h after the last loading dose.** The
  paper writes "a loading dose of 400 mg every 12 hours for three doses,
  followed a maintenance dose of 400 mg every 24 hours" without giving the
  gap. Starting maintenance one full interval after the last loading dose is
  the only reading that places a trough exactly at both published endpoints
  (72 h and 168 h) for the q12h x 3 and the q12h x 5 arms simultaneously.
* **`omega Q` is encoded as no random effect on Q.** Table 2 reports it as
  "0 FIX". Writing `etalq ~ fixed(0)` would make OMEGA singular and break
  rxode2's Cholesky sampler; omitting the eta is mathematically identical.
* **The reported `omega` percentages are treated as the log-scale standard
  deviation times 100**, so the packaged variances are the squared
  percentages. This is fixed by the paper's own printed `e^0.387` for Vp
  against its 62.2% table entry, not assumed.
* **Number of simulated subjects.** The paper does not state its Monte Carlo
  replicate count. This vignette uses `r n_sub` per arm, the package cap.
* **No body-weight or dose-per-kg scaling.** Body weight was screened and not
  retained (Supplementary Table S1), so all doses are absolute mg, as in the
  paper's own simulations.

# Errata and internal inconsistencies in the source

* **Delta OFV for eGFR on CL is reported inconsistently.** Results states
  "eGFR was a covariate significantly associated with inter-individual
  variability of systemic clearance (CL) (dOFV -23.189)". The Discussion
  gives "eGFR (-20.374)" and Supplementary Table S1 gives -20.374 against a
  base-model OFV of 1564.335 and a final-model OFV of 1543.961, which is
  self-consistent. The value -23.189 appears nowhere else; the closest
  Table S1 entry is -23.839, for serum creatinine on CL. The packaged model
  and its metadata use -20.374. No model parameter depends on this.
* **The residual error is reported as two transposed values.** Table 2 gives
  the proportional error as 31.7% (RSE 10.6%), but the Results text states
  "the intra-individual variability was 37.1%" -- the same digits in a
  different order. Table 2 is the correct one: its own bootstrap 95% CI for
  this parameter, 21.4-36.4%, contains 31.7% and excludes 37.1%. The packaged
  model uses `propSd = 0.317`. This matters for the reproduction above,
  because the residual error sets the interquartile widths that identify
  whether the published simulations carry residual error.
* **`Vp (L) = 51.6 x e^0.387` in the Table 2 note** reads as a multiplier but
  is the between-subject variance; see the Source trace section above. The
  typical Vp is 51.6 L per Table 2 and per the Results text.
* **The minus sign in the CL covariate equation is lost by text extraction**
  from the PDF; it was confirmed from the rendered page image and from the
  supplement's generic covariate form. See the Source trace section.
* **The Table 1 infection-site counts exceed the group sizes** (127 sites for
  108 model-establishment patients, 47 for 43 validation patients),
  consistent with patients carrying more than one infection site. The counts
  are recorded verbatim in the model's `population` metadata.
* **No erratum or corrigendum was found.** Crossref reports no `update-to`
  and no `relation` entries for doi:10.2147/DDDT.S413662 as of the extraction
  date.
