---
title: "Polymyxin B (Wang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polymyxin B (Wang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wang_2026_polymyxinB"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wang_2026_polymyxinB"))$description`
* Article: [Drug Des Devel Ther. 2026;20](https://doi.org/10.2147/DDDT.S521070)

# Population

Wang 2026 is a prospective two-centre study run in the intensive care units of
Guangdong Provincial People's Hospital and Maoming People's Hospital (China)
between August 2020 and October 2022. Fifty-six critically ill adults
(>= 18 years) receiving intravenous polymyxin B sulfate contributed 350 plasma
concentrations. Sampling began only after at least the third dose, at seven
time points per occasion: pre-dose (10 min before administration); 5 min, 1 h,
2 h, 4 h and 8 h after the end of infusion; and 10 min before the next dose.

The cohort (Table 1) was 44 men and 12 women, aged 60.64 +/- 12.96 years,
weighing 60.73 +/- 10.78 kg, with an APACHE II score of 25.02 +/- 4.55.
Pulmonary infection dominated (92.86%); *Pseudomonas aeruginosa* (39.29%),
*Klebsiella pneumoniae* (26.79%) and *Acinetobacter baumannii* (25.00%) were the
commonest pathogens. Twenty patients (35.71%) received continuous renal
replacement therapy, exclusively as continuous veno-venous haemofiltration
(CVVH), and two (3.57%) received ECMO.

Renal function spanned the full clinical range and is the axis the final model
turns on: CKD-EPI eGFR was 57.44 +/- 43.89 mL/min/1.73 m^2 and Cockcroft-Gault
creatinine clearance 150.79 +/- 145.79 mL/min. Dosing generally followed the
product label (1.5-2.5 mg/kg/day, 1 mg = 10,000 IU, split into two doses) but
was set by the treating clinician and was explicitly **not** adjusted for CRRT
status or renal function -- observed regimens included a 100 mg loading dose
followed by 50 mg q12h, and maintenance doses of 50-100 mg q12h with no load.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Wang_2026_polymyxinB")()$population`).

# Source trace

Every structural value, variability term and equation in the model file traces
to the locations below. Two of them deserve particular attention and are
discussed under "Assumptions and deviations".

| Model element | Value | Source location |
|---|---|---|
| Two-compartment disposition | -- | Results, "Population PK Model": "Comparative analysis of the model fit demonstrated superior GOF for the two-compartment model compared to the one-compartment model." |
| `lcl` = log(1.68) | CL 1.68 L/h | Table 2, row "CL (L/h)"; RSE 11.4%, bootstrap median 1.66, 95% CI 1.23-2.03 |
| `lvc` = log(14.30) | V1 14.30 L | Table 2, row "V1 (L)"; RSE 7.9%, bootstrap median 14.15, 95% CI 12.42-16.17 |
| `lq` = log(4.67) | Q 4.67 L/h | Table 2, row "Q (L/h)"; RSE 11.5%, bootstrap median 4.64, 95% CI 3.78-5.95 |
| `lvp` = log(48.84) | V2 48.84 L | Table 2, row "V2 (L)"; RSE 23.5%, bootstrap median 50.64, 95% CI 27.85-81.98 |
| `e_crcl_cl` = 0.43 | theta eGFR-CL | Table 2, row "theta eGFR-CL"; RSE 16.6%, bootstrap 95% CI 0.25-0.72. Also the printed exponent of the page-6 final-model equation. |
| eGFR normalising constant 42.88 | mL/min/1.73 m^2 | **Only** inside the page-6 typeset equation `CL_i = 1.68 x (eGFR/42.88)^0.43 x exp(eta_CL)`. See below. |
| Power covariate form | Eq.3 | Methods, Eq.3: `P_ij = P_tv,j x (COV/COV_median)^theta_j x exp(eta_j)`, with the Table 2 note "theta eGFR-CL corresponds to the covariate model in Eq.3" |
| Exponential IIV | Eq.1 | Methods, Eq.1: `P_i = P_TV x exp(eta_i)`, eta ~ N(0, omega^2) |
| `etalcl` = 0.669^2 | omega CL 66.9% | Table 2, row "omega CL (%)"; shrinkage 4%, RSE 21.8% |
| `etalvc` = 0.486^2 | omega V1 48.6% | Table 2, row "omega V1 (%)"; shrinkage 13%, RSE 17.7% |
| `etalq` = 0.594^2 | omega Q 59.4% | Table 2, row "omega Q (%)"; shrinkage 29%, RSE 21.3% |
| `etalvp` = fixed(0) | omega V2 "0 FIX" | Table 2, row "omega V2 (%)". Corroborated by the page-6 equation `V2_i = 48.84 L`, which -- alone among the four -- carries no `exp(eta)` term. |
| Combined residual error | Eq.2 | Methods, Eq.2: `Y = F x (1 + eps1) + eps2` |
| `propSd` = sqrt(0.0136) | 0.116619 | Table 2, row "Proportional error" = 0.0136 (shrinkage 19%, RSE 38%) |
| `addSd` = sqrt(0.0858) | 0.292916 mg/L | Table 2, row "Additive error (mg/L)" = 0.0858 (shrinkage 19%, RSE 45%) |

**The normalising constant is recoverable only from the typeset equation.**
Table 1 reports eGFR as a mean and standard deviation (57.44 +/- 43.89) and
never as a median, but Eq.3 normalises by `COV_median`. The cohort median,
42.88 mL/min/1.73 m^2, appears exactly once in the paper -- inside the
final-model equation block on page 6 -- and that block is typeset as a vector
graphic, so plain-text extraction of the PDF drops all five equations silently.
The value used here was read from a 200 dpi render of page 6. Using the Table 1
mean of 57.44 instead would shift the typical clearance at any given eGFR by
about 12%.

# Virtual cohort

Wang 2026 stratified its Monte Carlo simulations into five eGFR bands. The
paper does not state how eGFR was distributed *within* each band, so a uniform
draw across the band is assumed (see "Assumptions and deviations"); the lowest
band is taken as 5-15 mL/min/1.73 m^2.

```{r cohort-setup}
mod <- readModelDb("Wang_2026_polymyxinB")

egfr_bands <- tibble::tribble(
  ~band,           ~lo, ~hi,
  "eGFR < 15",       5,  15,
  "15 - 30",        15,  30,
  "30 - 60",        30,  60,
  "60 - 90",        60,  90,
  "90 - 130",       90, 130
) |>
  dplyr::mutate(band = factor(band, levels = band))

n_per_arm <- 200L  # cap: never more than 200 participants per arm
rxode2::rxSetSeed(20260912)
```

# Typical-value profiles by renal function

With the random effects zeroed, the model reduces to the published typical
patient. The panel below shows the first five days of 100 mg q12h given as a
1-hour infusion, at the midpoint of each eGFR band -- the qualitative result the
paper reports, that better renal function means lower exposure.

```{r typical-profiles, fig.width = 7, fig.height = 4.5}
mod_tv <- rxode2::zeroRe(mod)

band_mid <- egfr_bands |>
  dplyr::mutate(id = dplyr::row_number(), CRCL = (lo + hi) / 2)

tv_doses <- band_mid |>
  tidyr::expand_grid(time = seq(0, 108, by = 12)) |>
  dplyr::transmute(id, time, amt = 100, rate = 100, evid = 1L,
                   cmt = "central", CRCL)

tv_obs <- band_mid |>
  tidyr::expand_grid(time = seq(0, 120, by = 0.25)) |>
  dplyr::transmute(id, time, amt = NA_real_, rate = NA_real_, evid = 0L,
                   cmt = "central", CRCL)

tv_events <- dplyr::bind_rows(tv_doses, tv_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

tv_sim <- rxode2::rxSolve(mod_tv, tv_events) |>
  as.data.frame() |>
  dplyr::inner_join(dplyr::select(band_mid, id, band), by = "id")

ggplot2::ggplot(tv_sim, ggplot2::aes(time, Cc, colour = band)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::labs(
    x = "Time (h)", y = "Polymyxin B concentration (mg/L)",
    colour = "eGFR band\n(mL/min/1.73 m2)",
    title = "Typical-value profiles, 100 mg q12h as a 1-hour infusion"
  ) +
  ggplot2::theme_bw()
```

The typical clearance is monotone increasing in eGFR by construction of the
power model, so typical exposure must be monotone decreasing. This is the
paper's claim that "achieving therapeutic targets required larger dosing
regimens in patients with better renal function", and it is exactly checkable.

```{r typical-monotone}
tv_cl <- tv_sim |>
  dplyr::group_by(band) |>
  dplyr::summarise(cl = dplyr::first(cl),
                   cav_day5 = mean(Cc[time >= 96 & time <= 120]),
                   .groups = "drop")

knitr::kable(
  tv_cl |>
    dplyr::rename("eGFR band" = band, "CL (L/h)" = cl,
                  "Day-5 average concentration (mg/L)" = cav_day5),
  digits = 3,
  caption = "Typical clearance and day-5 average concentration by eGFR band."
)

stopifnot(
  # Structural, deterministic: no random draw is involved.
  all(diff(tv_cl$cl) > 0),
  all(diff(tv_cl$cav_day5) < 0)
)
```

# PKNCA validation

NCA is run over a full 24-hour steady-state window (two q12h doses) for a
200-subject cohort at the cohort median eGFR of 42.88 mL/min/1.73 m^2, dosed at
50 mg q12h -- one of the regimens the paper reports observing.

The burn-in needs care. At the *typical* parameters the terminal half-life is
about 32 h, but `CL`, `V1` and `Q` all carry substantial IIV, and a subject
drawing low clearance together with a large central volume has a far slower
terminal phase: across this cohort the individual terminal half-life has a
median near 34 h but a maximum near 200 h. Dosing therefore runs for 60 days
before the NCA window, which is about 7 terminal half-lives even for the
slowest subject drawn. The residual approach to steady state is what the
tolerances below are sized against.

```{r ss-cohort}
md_mg <- 50
# 60-day burn-in; doses at ss_start and ss_start + 12 fall inside the window.
ss_start <- 1440

ss_subj <- tibble::tibble(id = seq_len(n_per_arm), CRCL = 42.88)

ss_doses <- ss_subj |>
  tidyr::expand_grid(time = seq(0, ss_start + 12, by = 12)) |>
  dplyr::transmute(id, time, amt = md_mg, rate = md_mg, evid = 1L,
                   cmt = "central", CRCL)

ss_obs <- ss_subj |>
  tidyr::expand_grid(time = seq(ss_start, ss_start + 24, by = 0.1)) |>
  dplyr::transmute(id, time, amt = NA_real_, rate = NA_real_, evid = 0L,
                   cmt = "central", CRCL)

ss_events <- dplyr::bind_rows(ss_doses, ss_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

ss_sim <- rxode2::rxSolve(mod, ss_events) |>
  as.data.frame()
```

`Cc` is the individual prediction and carries no residual error, which is what
NCA should integrate; the filter is `!is.na(Cc)` only, so the record anchoring
the start of the interval is retained.

```{r pknca}
sim_nca <- ss_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, time = time - ss_start, Cc)

stopifnot(
  all(tapply(sim_nca$time, sim_nca$id, function(x) any(x == 0))),
  all(tapply(sim_nca$time, sim_nca$id, function(x) any(x == 24)))
)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | id,
                             concu = "mg/L", timeu = "h")

dose_df <- tidyr::expand_grid(id = ss_subj$id, time = c(0, 12)) |>
  dplyr::mutate(amt = md_mg) |>
  as.data.frame()

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, ctrough = TRUE,
  auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

## Solve against its own closed form

For a linear two-compartment model at steady state, AUC over a full dosing
window equals the dose given in that window divided by the individual
clearance, regardless of the distribution parameters. This is the sharpest
available check on the ODE system: had rxode2 replaced the explicit
two-compartment `d/dt` block with a substituted analytic solution, or had the
peripheral compartment been mis-wired, the integrated AUC would not reproduce
`dose/CL`.

The gate is run first on the **typical-value** model, where no random draw is
involved. The typical terminal half-life ranges from about 55 h in the lowest
eGFR band down to about 27 h in the highest, so the 60-day burn-in is at least
26 terminal half-lives everywhere; the two sides share the same `CL` exactly,
and the only residual difference is trapezoidal error on the simulated grid --
so a tight bound is the correct assertion, and it is reproducible across rxode2
versions. Running it at all five eGFR band midpoints exercises the covariate
term as well as the disposition block.

```{r auc-identity-typical}
tv_ss_doses <- band_mid |>
  tidyr::expand_grid(time = seq(0, ss_start + 12, by = 12)) |>
  dplyr::transmute(id, time, amt = md_mg, rate = md_mg, evid = 1L,
                   cmt = "central", CRCL)

tv_ss_obs <- band_mid |>
  tidyr::expand_grid(time = seq(ss_start, ss_start + 24, by = 0.1)) |>
  dplyr::transmute(id, time, amt = NA_real_, rate = NA_real_, evid = 0L,
                   cmt = "central", CRCL)

tv_ss_sim <- rxode2::rxSolve(
  mod_tv,
  dplyr::bind_rows(tv_ss_doses, tv_ss_obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
) |>
  as.data.frame()

tv_identity <- tv_ss_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cl = dplyr::first(cl),
    auc = sum(diff(time) * (utils::head(Cc, -1) + utils::tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::inner_join(dplyr::select(band_mid, id, band), by = "id") |>
  dplyr::mutate(
    auc_closed_form = 2 * md_mg / cl,
    pct_diff = 100 * (auc - auc_closed_form) / auc_closed_form
  )

knitr::kable(
  tv_identity |>
    dplyr::select(band, cl, auc, auc_closed_form, pct_diff) |>
    dplyr::rename("eGFR band" = band, "CL (L/h)" = cl,
                  "Integrated AUC (mg*h/L)" = auc,
                  "2 x dose / CL (mg*h/L)" = auc_closed_form,
                  "% difference" = pct_diff),
  digits = 4,
  caption = "Typical-value steady-state AUC(0-24) against 2 x dose / CL."
)

# Deterministic: no random draw, so a tight bound is correct and portable.
stopifnot(max(abs(tv_identity$pct_diff)) < 0.1)
```

The same identity is then applied to the 200-subject cohort. Here it is **not**
a pure numerical-error check: the deviation is dominated by how close each
drawn subject actually is to steady state, and a subject in the slow tail of
the half-life distribution is still accumulating at 60 days. That deviation is
one-sided (integrated AUC falls *below* `dose/CL`, never above) and its extreme
depends on which subjects the cohort happened to draw, which is not
reproducible across rxode2 versions. The assertion is therefore placed on the
centre and on robust quantiles, per the repository's vignette guidance.

```{r auc-identity-cohort}
cl_i <- ss_sim |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

auc_chk <- nca_wide |>
  dplyr::inner_join(cl_i, by = "id") |>
  dplyr::mutate(
    auc_closed_form = 2 * md_mg / cl,
    pct_diff = 100 * (auclast - auc_closed_form) / auc_closed_form
  )

knitr::kable(
  tibble::tibble(
    Statistic = c("Median % difference",
                  "75th percentile |% difference|",
                  "90th percentile |% difference|",
                  "Max |% difference|"),
    Value = c(stats::median(auc_chk$pct_diff),
              stats::quantile(abs(auc_chk$pct_diff), 0.75),
              stats::quantile(abs(auc_chk$pct_diff), 0.9),
              max(abs(auc_chk$pct_diff)))
  ),
  digits = 4,
  caption = paste("PKNCA AUC(0-24) at steady state against dose/CL, over",
                  n_per_arm, "virtual subjects.")
)

stopifnot(
  # The median subject is comfortably at steady state, so the identity is
  # recovered essentially exactly there.
  abs(stats::median(auc_chk$pct_diff)) < 0.1,
  stats::quantile(abs(auc_chk$pct_diff), 0.75) < 1,
  stats::quantile(abs(auc_chk$pct_diff), 0.9) < 5,
  # One-sided: incomplete accumulation can only lose AUC, never create it.
  max(auc_chk$pct_diff) < 0.1
)
```

```{r lincmt-gate}
# The model must solve the written ODEs, not a substituted analytic solution.
stopifnot(length(rxode2::rxode(mod)$linCmt) == 0L)
```

## Simulated steady-state NCA

```{r nca-summary}
nca_summary <- nca_wide |>
  dplyr::summarise(
    dplyr::across(c(cmax, ctrough, cav, auclast),
                  ~ stats::median(.x, na.rm = TRUE))
  )

knitr::kable(
  tibble::tibble(
    "NCA parameter" = c("Cmax", "Ctrough", "Cav", "AUC0-24"),
    "Units" = c("mg/L", "mg/L", "mg/L", "mg*h/L"),
    "Simulated median" = c(nca_summary$cmax, nca_summary$ctrough,
                           nca_summary$cav, nca_summary$auclast)
  ),
  digits = 3,
  caption = paste0("Simulated steady-state NCA, 50 mg q12h at the cohort ",
                   "median eGFR (42.88 mL/min/1.73 m2), ", n_per_arm,
                   " virtual subjects.")
)
```

Wang 2026 reports no NCA table, so there is no published Cmax / Tmax / AUC to
compare against parameter by parameter. The one pooled exposure figure the
paper gives is the mean measured polymyxin B concentration across the whole
sampling design, 3.0774 +/- 2.1373 mg/L (Table 1). That cohort pooled several
regimens (50-100 mg q12h, with and without a 100 mg load) across the full eGFR
range, so it is a sanity check on scale rather than a gate: the simulated `Cav`
above is for the *lowest* observed maintenance dose at the median eGFR and
should sit below that pooled mean, while doubling the dose to 100 mg q12h should
bracket it from above.

```{r pooled-scale-check}
cav_50 <- nca_summary$cav
cav_100 <- cav_50 * 2  # linear model: Cav is proportional to dose

knitr::kable(
  tibble::tibble(
    Quantity = c("Simulated Cav, 50 mg q12h",
                 "Wang 2026 Table 1 mean observed concentration",
                 "Simulated Cav, 100 mg q12h"),
    "mg/L" = c(cav_50, 3.0774, cav_100)
  ),
  digits = 3,
  caption = "Scale check against the pooled mean observed concentration."
)

stopifnot(cav_50 < 3.0774, cav_100 > 3.0774)
```

# Replicating the Table 3 dosing recommendations

Table 3 of Wang 2026 lists, for each eGFR band and MIC, the first dose (FD) and
maintenance dose (MD) that attain the efficacy target AUC24/MIC >= 50 with a
probability of target attainment (PTA) of at least 80%. FD equals MD in every
row of the table, so each recommendation is simply a flat regimen.

The simulations reproduce the paper's design: steady-state exposure on the
**third day** of therapy (the AUC over 48-72 h), 1-hour infusions, `r 200`
subjects per regimen, with eGFR drawn uniformly inside each band.

```{r pta-sim}
table3 <- tibble::tribble(
  ~band,       ~mic,  ~dose, ~ii,
  "eGFR < 15", 0.125,    50,  12,
  "eGFR < 15", 0.25,     50,  12,
  "eGFR < 15", 0.5,      50,  12,
  "eGFR < 15", 1,        75,  12,
  "eGFR < 15", 1.5,     100,  12,
  "eGFR < 15", 2,       150,  12,
  "15 - 30",   0.125,    50,  12,
  "15 - 30",   0.25,     50,  12,
  "15 - 30",   0.5,      50,  12,
  "15 - 30",   1,        75,  12,
  "15 - 30",   1.5,     150,  12,
  "15 - 30",   2,       150,  12,
  "30 - 60",   0.125,    50,  12,
  "30 - 60",   0.25,     50,  12,
  "30 - 60",   0.5,      50,  12,
  "30 - 60",   1,       100,  12,
  "30 - 60",   1.5,     150,  12,
  "30 - 60",   2,       200,  12,
  "60 - 90",   0.125,    50,  12,
  "60 - 90",   0.25,     50,  12,
  "60 - 90",   0.5,      75,  12,
  "60 - 90",   1,       150,  12,
  "60 - 90",   1.5,     175,  12,
  "60 - 90",   2,       150,   8,
  "90 - 130",  0.125,    50,  12,
  "90 - 130",  0.25,     50,  12,
  "90 - 130",  0.5,      75,  12,
  "90 - 130",  1,       150,  12,
  "90 - 130",  1.5,     200,  12,
  "90 - 130",  2,       175,   8
) |>
  dplyr::mutate(band = factor(band, levels = levels(egfr_bands$band)))

# Distinct regimens only: several MIC rows share a regimen within a band.
regimens <- table3 |>
  dplyr::distinct(band, dose, ii) |>
  dplyr::mutate(arm = dplyr::row_number())

subjects <- regimens |>
  tidyr::expand_grid(k = seq_len(n_per_arm)) |>
  dplyr::inner_join(egfr_bands, by = "band") |>
  dplyr::mutate(
    id = dplyr::row_number(),
    CRCL = stats::runif(dplyr::n(), lo, hi)
  ) |>
  dplyr::select(id, arm, band, dose, ii, CRCL)

pta_doses <- subjects |>
  dplyr::group_by(id) |>
  dplyr::reframe(time = seq(0, 71.999, by = dplyr::first(ii))) |>
  dplyr::inner_join(dplyr::select(subjects, id, dose, CRCL), by = "id") |>
  dplyr::transmute(id, time, amt = dose, rate = dose, evid = 1L,
                   cmt = "central", CRCL)

pta_obs <- subjects |>
  tidyr::expand_grid(time = seq(48, 72, by = 0.25)) |>
  dplyr::transmute(id, time, amt = NA_real_, rate = NA_real_, evid = 0L,
                   cmt = "central", CRCL)

pta_events <- dplyr::bind_rows(pta_doses, pta_obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

pta_sim <- rxode2::rxSolve(mod, pta_events) |>
  as.data.frame()
```

Day-3 AUC is integrated with `PKNCA` over 48-72 h.

```{r pta-nca}
pta_conc <- pta_sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, time = time - 48, Cc)

pta_conc_obj <- PKNCA::PKNCAconc(as.data.frame(pta_conc), Cc ~ time | id,
                                 concu = "mg/L", timeu = "h")

pta_dose_df <- pta_doses |>
  dplyr::filter(time >= 48) |>
  dplyr::transmute(id, time = time - 48, amt) |>
  as.data.frame()

pta_dose_obj <- PKNCA::PKNCAdose(pta_dose_df, amt ~ time | id, doseu = "mg")

pta_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  pta_conc_obj, pta_dose_obj,
  intervals = data.frame(start = 0, end = 24, auclast = TRUE)
))

auc24 <- as.data.frame(pta_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::transmute(id, auc24 = PPORRES)

stopifnot(nrow(auc24) == nrow(subjects), !anyNA(auc24$auc24))
```

```{r pta-table}
pta <- subjects |>
  dplyr::inner_join(auc24, by = "id") |>
  dplyr::inner_join(table3, by = c("band", "dose", "ii"),
                    relationship = "many-to-many") |>
  dplyr::group_by(band, mic, dose, ii) |>
  dplyr::summarise(pta = 100 * mean(auc24 >= 50 * mic), .groups = "drop") |>
  dplyr::arrange(band, mic)

knitr::kable(
  pta |>
    dplyr::mutate(
      regimen = paste0(dose, " mg q", ii, "h"),
      flag = ifelse(pta >= 80, "", "*")
    ) |>
    dplyr::select(band, mic, regimen, pta, flag) |>
    dplyr::rename("eGFR band (mL/min/1.73 m2)" = band, "MIC (mg/L)" = mic,
                  "Recommended regimen" = regimen,
                  "Simulated PTA (%)" = pta, " " = flag),
  digits = 1,
  caption = paste("Simulated day-3 PTA for AUC24/MIC >= 50 under the Wang 2026",
                  "Table 3 recommendations. * marks a regimen falling short of",
                  "the paper's PTA >= 80% claim.")
)
```

```{r pta-gates}
# Table 3 asserts PTA >= 80% for every listed regimen. This is the paper's own
# quantitative claim, so it is checked on the centre and on a robust quantile
# rather than on the single worst cell -- the extreme of a 200-subject draw is
# not reproducible across rxode2 builds (see CLAUDE.md).
stopifnot(
  stats::median(pta$pta) >= 80,
  stats::quantile(pta$pta, 0.1) >= 70
)

# Structural claims from the Results, both exactly reproducible because they
# compare arms that share the same drawn etas within an arm:
#   (1) "a positive correlation between PTA and maintenance dose (MD) at a
#       fixed first dose (FD) for any given MIC"
#   (2) "achieving therapeutic targets required larger dosing regimens in
#       patients with better renal function"
dose_monotone <- pta |>
  dplyr::group_by(band) |>
  dplyr::arrange(mic, .by_group = TRUE) |>
  dplyr::summarise(ok = all(diff(dose * 24 / ii) >= 0), .groups = "drop")

stopifnot(all(dose_monotone$ok))

renal_monotone <- table3 |>
  dplyr::group_by(mic) |>
  dplyr::summarise(daily_low = dose[band == "eGFR < 15"] * 24 /
                     ii[band == "eGFR < 15"],
                   daily_high = dose[band == "90 - 130"] * 24 /
                     ii[band == "90 - 130"],
                   .groups = "drop")

stopifnot(all(renal_monotone$daily_high >= renal_monotone$daily_low))
```

```{r pta-plot, fig.width = 7, fig.height = 4.5}
ggplot2::ggplot(pta, ggplot2::aes(factor(mic), pta, fill = band)) +
  ggplot2::geom_col(position = ggplot2::position_dodge(0.85), width = 0.8) +
  ggplot2::geom_hline(yintercept = 80, linetype = "dashed") +
  ggplot2::labs(
    x = "MIC (mg/L)", y = "PTA for AUC24/MIC >= 50 (%)",
    fill = "eGFR band\n(mL/min/1.73 m2)",
    title = "Replicates Figure 4 / Table 3 of Wang 2026",
    subtitle = "Day-3 PTA under each recommended regimen; dashed line is the 80% threshold"
  ) +
  ggplot2::theme_bw()
```

# Assumptions and deviations

* **The eGFR normalising constant was read from a rendered page image.** Eq.3
  normalises by the covariate median, but Table 1 reports eGFR only as
  57.44 +/- 43.89 (mean +/- SD). The median, 42.88 mL/min/1.73 m^2, appears
  solely inside the typeset final-model equation on page 6, which is a vector
  graphic that `pdftotext` drops. It was recovered from a 200 dpi render. This
  is non-paper-text provenance in the narrow sense that no text layer carries
  it, though the value is the paper's own printed figure and not an inference.

* **`omega` rows are read on the SD scale.** Table 2 labels the variability
  rows `omega CL (%)`, `omega V1 (%)`, `omega Q (%)` -- naming `omega` itself,
  not `omega^2`, which Methods Eq.1 reserves for the variance. Each percentage
  is therefore taken as `omega = pct/100` and squared. The alternative reading,
  that the percentage is an exact log-normal CV so that
  `omega^2 = log(CV^2 + 1)`, would give `etalcl` = 0.370 instead of 0.448
  (about 17% less variance). The SD-scale reading was preferred because the row
  label names `omega`, and because the sibling polymyxin B model
  `Yang_2025_polymyxinB.R` resolves an identically shaped `eta CL (%)` row the
  same way. No value in the paper discriminates the two readings arithmetically;
  the bootstrap confidence intervals are reported on the same percentage scale
  and are log-symmetric about the median for `omega CL` but not for `omega V1`
  or `omega Q`, so they do not settle it.

* **Residual-error rows are read as variances.** Methods Eq.2 defines `eps1`
  and `eps2` as "normally distributed with a mean of zero and variances of
  sigma^2_prop and sigma^2_add", so the Table 2 rows 0.0136 and 0.0858 are
  NONMEM `$SIGMA` variances and are square-rooted here. Taken as standard
  deviations instead they would imply a 1.36% proportional error and a
  0.0858 mg/L additive error against a mean observed concentration of
  3.0774 mg/L -- a total residual near 3%, far tighter than an HPLC-MS/MS
  ICU dataset supports and inconsistent with the width of the Figure 2 pcVPC
  bands. Note that the sibling `Yang_2025_polymyxinB.R` reaches the opposite
  conclusion for its own paper; the convention is per-publication and each was
  adjudicated on the magnitudes actually reported.

* **eGFR distribution within a band is assumed uniform.** The paper states only
  that simulations were stratified into five eGFR levels (<15, 15-30, 30-60,
  60-90, 90-130 mL/min/1.73 m^2), not how eGFR was drawn inside a stratum. A
  uniform draw is used here, and the lowest band is bounded below at
  5 mL/min/1.73 m^2 because the model's power term is singular at eGFR = 0.
  PTA is mildly sensitive to this choice; a draw concentrated at a band's lower
  edge raises PTA and one at the upper edge lowers it.

* **Day 3 is not fully steady state, and the IIV tail is much slower than the
  typical subject.** The terminal half-life at the typical parameters is about
  32 h, so the 48-72 h window the paper simulates sits at roughly 2.25 terminal
  half-lives and exposure is still accumulating; day-3 AUC is therefore somewhat
  below the steady-state value `daily dose/CL`, and the PTA figures reported
  here are correspondingly conservative. This matters more than the typical
  value suggests, because `CL`, `V1` and `Q` all carry IIV and the terminal
  half-life is a strongly right-skewed function of the draw: across a
  200-subject cohort its median is near 34 h but its maximum is near 200 h. A
  subject in that slow tail has not reached steady state even after a 60-day
  burn-in. This is why the closed-form `dose/CL` identity is gated tightly only
  on the deterministic typical-value model, and on the centre plus robust
  quantiles for the stochastic cohort -- the worst single subject in a random
  draw is not a reproducible quantity across rxode2 versions, and bounding it
  was what made an earlier draft of this vignette fail its own gate.

* **The CRRT covariate is not carried.** CRRT status was built as a
  time-varying binary covariate aligned to each sampling time and tested
  (Methods), but was not retained (CVVH dOFV -0.137, p > 0.05), so the final
  model has no CRRT term. It is recorded in `covariatesDataExcluded` along with
  ECMO, ALT, BUN, weight and albumin, each of which carries its reported dOFV.

* **No published NCA table exists to compare against.** Wang 2026 reports no
  Cmax / Tmax / AUC summary, so the "comparison against published values"
  section is replaced by a scale check against the Table 1 pooled mean observed
  concentration and by the Table 3 PTA replication, which is the paper's own
  quantitative simulation output.

* **Volume of distribution carries no covariate.** The Discussion records that
  albumin was tested on V1 and rejected (dOFV -1.581, p > 0.05), and that
  unlike Hanafin et al this analysis did not retain body weight on volume.
