---
title: "Serplulimab flat vs weight-based dosing (Wang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Serplulimab flat vs weight-based dosing (Wang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wang_2026_serplulimab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wang_2026_serplulimab"))$description`
* Article: <https://doi.org/10.1002/psp4.70204>

Serplulimab (HLX10) is a fully humanized IgG4 anti-PD-1 monoclonal antibody
approved for extensive-stage small cell lung cancer and, in China, for
oesophageal squamous cell carcinoma and metastatic squamous NSCLC. Wang et al.
(2026) pooled 14,687 serum concentrations from 2110 patients across eleven
Phase I-III trials and asked a single regulatory question: can the weight-based
regimens of 3 mg/kg Q2W and 4.5 mg/kg Q3W be replaced by the flat regimens of
200 mg Q2W and 300 mg Q3W without changing exposure, efficacy or safety?

This analysis **supersedes** the eight-trial, 1144-patient model of Wang et al.
(2025), which is in this library as `modellib("Wang_2025_serplulimab")`. The two
share a structural form -- a two-compartment IV model with linear elimination
and sigmoidal time-varying clearance -- but differ substantially in the
covariate model and in every parameter estimate, so both are kept.

| | Wang 2025 | Wang 2026 |
|---|---|---|
| Trials / subjects | 8 / 1144 | 11 / 2110 |
| Observations | 6650 | 14,687 |
| Tumour type on CL, Vc | one pooled non-lung indicator on Vc only | seven per-histology CL0 and Vc estimates |
| Covariates on CL | WT, ALB | WT, ALB, sex, ALP, tumour burden |
| Covariates on Vc | WT, sex, tumour type | WT, ALB, sex |
| Covariates on Vp | none | ALB, tumour burden |
| exp(Emax), T50, lambda | 0.695, 106 d, 2.05 | 0.912, 221 d, 2.43 |

The structural model is

$$
\mathrm{CL}_i(t) = \mathrm{CL0}_i \cdot
  \exp\!\left(
    \frac{E_{\max,i}\, t^{\lambda}}{T_{50}^{\lambda} + t^{\lambda}}
  \right),
\qquad
E_{\max,i} = -0.0926 + \eta_{E_{\max},i}
$$

so clearance falls from its baseline value towards
$\mathrm{CL0}\cdot\exp(E_{\max})$ as treatment continues. Wang 2026 Section 2.2
is explicit that this is an **empirical** descriptor of disease-related change
rather than a mechanistic one: TMDD forms, including simplified
Michaelis-Menten approximations, "did not improve model performance", which the
authors attribute to rapid PD-1 receptor saturation leaving little information
at non-saturating exposures.

The distinctive feature of the 2026 model is that tumour type does **not** enter
as a covariate coefficient. Wang 2026 Section 2.2 states that "separate
parameter estimates were obtained for each relevant tumor type rather than
pooling them", so the joint fit produces seven parallel `CL0` and seven parallel
`Vc` estimates. The model file encodes these as stratum-suffixed parameters
selected by six mutually exclusive tumour-type indicators, with non-squamous
NSCLC -- the largest group and the histology Wang 2026 Figure 1 labels "+0%" --
as the complement and therefore the reference.

# Population

```{r population}
pop <- rxode2::rxode(readModelDb("Wang_2026_serplulimab"))$population

knitr::kable(
  data.frame(
    Field = c("Species", "Subjects", "Trials", "Observations", "Age",
              "Weight", "Female", "Race", "Albumin", "Tumour burden",
              "ADA positive"),
    Value = c(pop$species, pop$n_subjects, pop$n_studies, pop$n_observations,
              paste0(pop$age_median, " (", pop$age_range, ")"),
              paste0(pop$weight_median, " (", pop$weight_range, ")"),
              paste0(pop$sex_female_pct, "%"),
              paste(names(pop$race_ethnicity), pop$race_ethnicity,
                    sep = " ", collapse = "%, "),
              pop$albumin, pop$tumour_burden, "133 / 2110 (6.30%)")
  ),
  caption = "Baseline demographics of the pooled PK dataset (Wang 2026 Table 1)."
)
```

The tumour-type mix determines which stratum of the model applies:

```{r tumour-mix}
knitr::kable(
  data.frame(
    `Tumour type` = c("Non-squamous NSCLC (reference)", "Squamous NSCLC",
                      "Small cell lung cancer", "Oesophageal SCC",
                      "Colorectal cancer", "Hepatocellular carcinoma",
                      "Other"),
    Indicator = c("(all six indicators 0)", "TUMTP_NSCLC_SQUAM", "TUMTP_SCLC",
                  "TUMTP_ESCC", "TUMTP_CRC", "TUMTP_HCC", "TUMTP_OTHER"),
    n = c(498L, 441L, 390L, 389L, 151L, 125L, 116L),
    `Percent` = c(23.60, 20.90, 18.48, 18.44, 7.16, 5.92, 5.50),
    `CL0 (L/day)` = c(0.184, 0.204, 0.171, 0.178, 0.182, 0.204, 0.211),
    `Vc (L)` = c(3.25, 3.38, 3.45, 3.48, 3.19, 3.20, 3.19),
    check.names = FALSE
  ),
  caption = "Per-histology strata of the joint fit (Wang 2026 Table 1 and Table 2)."
)
```

# Source trace

Every `ini()` value traces to Wang 2026 Table 2 (final-model estimates, with
bootstrap medians alongside) or to the equation block printed in Section 3.1.
No value is digitised, inferred, or carried from another publication.

```{r source-trace, echo = FALSE}
knitr::kable(
  data.frame(
    Parameter = c(
      "lcl_sqnsclc / lcl_hcc / lcl_crc / lcl_nonsqnsclc / lcl_sclc / lcl_escc / lcl_other",
      "lvc_sqnsclc / lvc_hcc / lvc_crc / lvc_nonsqnsclc / lvc_sclc / lvc_escc / lvc_other",
      "lq", "lvp",
      "cl_time_max", "lcl_t50", "lcl_time_hill",
      "e_wt_cl", "e_alb_cl", "e_tum_sld_cl", "e_alp_cl", "e_sexf_cl",
      "e_wt_vc", "e_alb_vc", "e_sexf_vc",
      "e_alb_vp", "e_tum_sld_vp",
      "etalcl, etalvc, covariance", "etalq", "etalvp", "etacl_time_max",
      "propSd",
      "Reference covariate values",
      "ODE structure"
    ),
    Value = c(
      "0.204 / 0.204 / 0.182 / 0.184 / 0.171 / 0.178 / 0.211 L/day",
      "3.38 / 3.20 / 3.19 / 3.25 / 3.45 / 3.48 / 3.19 L",
      "0.405 L/day", "2.98 L",
      "-0.0926", "221 day", "2.43",
      "0.514", "-0.714", "0.0548", "0.0553", "-0.145",
      "0.470", "-0.320", "-0.14",
      "-1.05", "0.107",
      "0.0576, 0.02657, 0.014", "0.29485", "0.21068", "0.11628",
      "0.176",
      "WT 62 kg, ALB 41.4 g/L, TUMBUR 73 mm, ALP 94 U/L, male, non-sq. NSCLC",
      "2-compartment IV, linear elimination, sigmoidal time-varying CL"
    ),
    `Source location` = c(
      "Table 2 rows 'Baseline clearance, CL0 (L/day)', one row per tumour type; the same seven values are listed in the Section 3.1 prose",
      "Table 2 rows 'Central volume of distribution, Vc (L)', one row per tumour type; the same seven values are listed in the Section 3.1 prose",
      "Table 2 'Inter-compartmental clearance, Q'; Section 3.1 equation prints 0.0169 L/h x 24",
      "Table 2 'Peripheral volume of distribution, Vp'; leading constant of the Section 3.1 Vp equation",
      "Section 3.1 'Emax,i = -0.0926 + eta_Emax,i'; Table 2 gives the back-transform exp(Emax) = 0.912",
      "Table 2 'Time to 50% maximum clearance change, T50'; Section 3.1 equation block",
      "Table 2 'Clearance time-effect, lambda'; Section 3.1 equation block",
      "Table 2 'Effect of weight on CL, CLwt'; Section 3.1 CL0 equation (WT/62)^0.514",
      "Table 2 'Effect of albumin on CL, CLalb'; Section 3.1 CL0 equation (ALB/41.4)^-0.714",
      "Table 2 'Effect of tumor burden on CL, CLtmb'; Section 3.1 CL0 equation (TUMBUR/73)^0.0548",
      "Table 2 'Effect of ALP on CL, CLalp'; Section 3.1 CL0 equation (ALP/94)^0.0553",
      "Table 2 'Effect of sex on CL, CLsex'; Section 3.1 CL0 equation exp[-0.145*(Female)]",
      "Table 2 'Effect of weight on Vc, Vwt'; Section 3.1 Vc equation (WT/62)^0.470",
      "Table 2 'Effect of albumin on Vc, Vcalb'; Section 3.1 Vc equation (ALB/41.4)^-0.320",
      "Table 2 'Effect of sex on Vc, Vcsex'; Section 3.1 Vc equation exp[-0.14*(Female)]",
      "Table 2 'Effect of albumin on Vp, Vpalb'; Section 3.1 Vp equation (ALB/41.4)^-1.05",
      "Table 2 'Effect of tumor burden on Vp, VPtmb'; Section 3.1 Vp equation (TUMBUR/73)^0.107",
      "Table 2 'IIV of CL (%)' 24.0, 'IIV of Vc (%)' 16.3, 'Covariance (CL_Vc)' 0.014",
      "Table 2 'IIV of Q (%)' 54.3", "Table 2 'IIV of Vp (%)' 45.9",
      "Table 2 'IIV of Emax (%)' 34.1",
      "Table 2 'Residual error (%), sigma' 17.6",
      "Denominators printed in the Section 3.1 equations; each equals the rounded Table 1 PK-dataset median",
      "Section 2.2 and the Section 3.1 equation block"
    ),
    check.names = FALSE
  ),
  caption = "Source trace for every parameter in the model file."
)
```

## Reading the reported IIV percentages

Wang 2026 Table 2 reports the random effects as bare percentages under the note
"IIV for CL, Vc, Q, Vp, Emax, and residual are reported as approximate CV%".
Whether such a percentage is `omega` (an SD) or `omega^2` (a variance) is the
single most common transcription error in this class of paper, so it is settled
arithmetically here rather than assumed.

```{r iiv-scale}
iiv <- data.frame(
  parameter = c("CL", "Vc", "Q", "Vp", "Emax", "sigma"),
  pct       = c(24.0, 16.3, 54.3, 45.9, 34.1, 17.6),
  ci_lo     = c(22.6, 15.0, 43.7, 41.5, 27.3, 17.0),
  ci_hi     = c(25.3, 17.5, 63.1, 49.8, 39.7, 18.2)
)

# A NONMEM covariance step returns a CI that is symmetric on the VARIANCE
# scale. If the printed percentage is omega x 100, then squaring the printed
# CI endpoints must bracket the squared point estimate symmetrically.
iiv <- iiv |>
  dplyr::mutate(
    variance     = (pct / 100)^2,
    ci_midpoint  = ((ci_lo / 100)^2 + (ci_hi / 100)^2) / 2,
    rel_diff_pct = 100 * (ci_midpoint / variance - 1)
  )

knitr::kable(iiv, digits = 5,
  caption = "Squaring the printed CI endpoints recovers the squared point estimate for every row, which is only true if the printed percentage is omega (an SD) rather than a variance.")

# Deterministic arithmetic on published numbers, so a tight bound is correct.
stopifnot(max(abs(iiv$rel_diff_pct)) < 0.5)
```

A second, independent argument points the same way. The CL-Vc covariance is
reported on the raw OMEGA-block scale as 0.014. Pairing it with *variances* of
24.0 and 16.3 would imply a correlation of 0.0007; pairing it with *SDs* of
0.240 and 0.163 gives 0.358.

```{r iiv-correlation}
corr_as_sd  <- 0.014 / (0.240 * 0.163)
corr_as_var <- 0.014 / sqrt(24.0 * 16.3)
c(`correlation if percentages are SDs` = round(corr_as_sd, 3),
  `correlation if percentages are variances` = round(corr_as_var, 5))
stopifnot(corr_as_sd > 0.3, corr_as_sd < 0.4)
```

The same conversion is applied to `Emax` even though its eta is **additive**
rather than log-normal (Section 3.1 prints `Emax_i = -0.0926 + eta_Emax,i`):
the reporting routine emitted `sqrt(omega^2) x 100` uniformly for every row of
the table, so `omega_Emax = 0.341`, not `0.341 x |Emax|`. The resulting spread
-- `exp(Emax_i)` roughly 0.52 to 1.60 over a 90% interval, so some patients'
clearance rises over treatment while others' falls -- is the documented
behaviour for checkpoint inhibitors that this empirical term exists to capture.

# Deterministic structural checks

These use a typical subject at the reference covariates with IIV switched off,
so they involve no random draw and are reproducible on any machine. Tight
bounds are therefore correct here, and are what make them catch a
transcription regression.

```{r typical-setup}
mod  <- readModelDb("Wang_2026_serplulimab")
mod0 <- rxode2::zeroRe(mod)

ref_cov <- list(
  WT = 62, ALB = 41.4, ALP = 94, TUM_SLD = 73, SEXF = 0,
  TUMTP_NSCLC_SQUAM = 0, TUMTP_HCC = 0, TUMTP_CRC = 0,
  TUMTP_SCLC = 0, TUMTP_ESCC = 0, TUMTP_OTHER = 0
)

#' Solve one typical subject on a flat or weight-based regimen.
#' `cmt = "central"` is the ODE state name, never the observable "Cc".
solve_typical <- function(dose, tau, ndose, cov = ref_cov,
                          tmax = tau * ndose, by = 0.05) {
  # rxode2 warns when ii is set with addl = 0, so only add the repeat
  # specification for genuinely multiple-dose regimens.
  ev <- if (ndose > 1L) {
    rxode2::et(amt = dose, dur = 1 / 24, ii = tau, addl = ndose - 1L,
               cmt = "central")
  } else {
    rxode2::et(amt = dose, dur = 1 / 24, cmt = "central")
  }
  ev <- ev |> rxode2::et(seq(0, tmax, by = by), cmt = "central")
  d <- as.data.frame(ev)
  for (nm in names(cov)) d[[nm]] <- cov[[nm]]
  as.data.frame(rxode2::rxSolve(mod0, d, returnType = "data.frame"))
}

trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
```

## Mass balance

Because clearance is time-varying, `AUCinf` is *not* `Dose / CL0`. The
identity that does still hold exactly is that everything eliminated must equal
the dose: the integral of `CL(t) * C(t)` over all time is the amount cleared.

```{r mass-balance}
s_single <- solve_typical(300, 21, 1L, tmax = 700, by = 0.02)
eliminated <- trapz(s_single$time, s_single$cl * s_single$Cc)

c(`dose (mg)` = 300,
  `eliminated (mg)` = round(eliminated, 3),
  `recovery (%)` = round(100 * eliminated / 300, 3))

# Exact structural identity; the only error is numerical quadrature.
stopifnot(abs(eliminated / 300 - 1) < 0.005)
```

## Time-varying clearance reaches the published asymptote

```{r emax-asymptote}
s_long <- solve_typical(300, 21, 1L, tmax = 100000, by = 50)
asymptote <- tail(s_long$cl, 1) / s_long$cl[1]

c(`CL(inf) / CL(0) from the solve` = round(asymptote, 4),
  `Table 2 exp(Emax)` = 0.912)
stopifnot(abs(asymptote - 0.912) < 0.002)
```

## Terminal half-life

An IgG4 monoclonal antibody should show a terminal half-life near three weeks.
This is the check that would fail loudest if the `x 24` factor printed in the
Section 3.1 clearance equation had been applied twice (Table 2's `CL0` is
already in L/day; re-multiplying would give a 5.6-day half-life).

```{r half-life}
# Fit the terminal slope well after the distribution phase and after the
# time-varying clearance has settled, so log(2)/k is the true beta half-life.
tail_win <- dplyr::filter(s_single, time >= 300, time <= 600, Cc > 0)
k_beta   <- -unname(coef(lm(log(Cc) ~ time, data = tail_win))[2])

c(`terminal half-life (days)` = round(log(2) / k_beta, 1),
  `Vss (L)` = 3.25 + 2.98)
stopifnot(log(2) / k_beta > 20, log(2) / k_beta < 32)
```

## Steady-state exposure of the reference subject

Wang 2026 Figure 1 does not print the reference subject's exposure directly,
but it prints each 90% bound as an absolute concentration *and* as a percentage
change from the reference, so the reference is recoverable by exact algebra
from either bound -- and the two independent recoveries agree, which is what
makes this a usable target rather than a digitised one.

```{r figure1-reference}
recover_ref <- function(lo_val, lo_pct, hi_val, hi_pct) {
  c(from_lower = lo_val / (1 + lo_pct / 100),
    from_upper = hi_val / (1 + hi_pct / 100))
}
ref_targets <- rbind(
  Cavgss = recover_ref(51, -35.92, 118, 49.65),
  Cmaxss = recover_ref(104, -28.09, 202, 40.22),
  Cminss = recover_ref(28, -47.25, 87, 64.34)
)
knitr::kable(round(ref_targets, 2),
  caption = "Reference-subject exposure recovered independently from each 90% bound printed in Wang 2026 Figure 1.")

# The two recoveries must agree, otherwise the target is not trustworthy.
stopifnot(max(abs(ref_targets[, 1] / ref_targets[, 2] - 1)) < 0.02)
```

Wang 2026 Section 2.3 simulates "a 300 mg Q3W regimen (1-h infusion)
administered for eight consecutive doses", so the model is run on exactly that
schedule and summarised over the eighth interval.

```{r ss-reference}
n_cycles <- 8L; tau_q3w <- 21
ss_lo <- (n_cycles - 1L) * tau_q3w   # day 147
ss_hi <- ss_lo + tau_q3w             # day 168

ss_metrics <- function(cov = ref_cov, dose = 300, tau = tau_q3w) {
  s <- solve_typical(dose, tau, n_cycles, cov = cov,
                     tmax = (n_cycles - 1L) * tau + tau, by = 0.01)
  w <- dplyr::filter(s, time >= (n_cycles - 1L) * tau)
  c(Cavgss = trapz(w$time, w$Cc) / tau, Cmaxss = max(w$Cc), Cminss = min(w$Cc))
}

got <- ss_metrics()
ss_cmp <- data.frame(
  Metric      = names(got),
  Model       = round(unname(got), 2),
  `Figure 1`  = round(rowMeans(ref_targets), 2),
  `Diff (%)`  = round(100 * (unname(got) / rowMeans(ref_targets) - 1), 2),
  check.names = FALSE
)
knitr::kable(ss_cmp,
  caption = "Reference-subject steady-state exposure at 300 mg Q3W vs the values recovered from Wang 2026 Figure 1.")

# Deterministic on both sides. Realised -1.1% / -0.7% / -2.5%. A
# mis-transcribed CL0, Vc, Q, Vp or reference covariate moves these by tens
# of percent, so 6% is a real gate with room for the Figure-1 read-back.
stopifnot(max(abs(ss_cmp$`Diff (%)`)) < 6)
```

# Replicating Figure 1 - the covariate sensitivity forest

Wang 2026 Figure 1 varies one covariate at a time, holding all others at their
reference values, and reports the percentage change in steady-state exposure.
Because the paper's own procedure is a typical-value simulation with no
stochastic variability, this replication is fully deterministic on both sides
and a tight bound is the correct assertion.

```{r figure1-forest}
base_cavg <- unname(ss_metrics()["Cavgss"])

cases <- tibble::tribble(
  ~case,             ~override,                          ~published,
  "WT 85 kg",        list(WT = 85),                      -14.66,
  "WT 46 kg",        list(WT = 46),                       16.03,
  "ALB 32.8 g/L",    list(ALB = 32.8),                   -15.42,
  "ALB 47.7 g/L",    list(ALB = 47.7),                    10.68,
  "Female",          list(SEXF = 1),                      15.09,
  "TUMBUR 177.31 mm", list(TUM_SLD = 177.31),             -4.80,
  "TUMBUR 18.21 mm", list(TUM_SLD = 18.21),                7.96,
  "ALP 236.06 U/L",  list(ALP = 236.06),                  -4.75,
  "ALP 57 U/L",      list(ALP = 57),                       2.66,
  "Other tumour",    list(TUMTP_OTHER = 1),              -12.17,
  "HCC",             list(TUMTP_HCC = 1),                 -9.52,
  "Squamous NSCLC",  list(TUMTP_NSCLC_SQUAM = 1),         -9.24,
  "SCLC",            list(TUMTP_SCLC = 1),                 7.17,
  "ESCC",            list(TUMTP_ESCC = 1),                 2.97,
  "CRC",             list(TUMTP_CRC = 1),                  1.46
)

forest <- cases |>
  dplyr::rowwise() |>
  dplyr::mutate(
    model = 100 * (unname(ss_metrics(utils::modifyList(ref_cov, override))["Cavgss"]) /
                     base_cavg - 1)
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(`abs diff (pp)` = abs(model - published)) |>
  dplyr::select(Case = case, `Model (%)` = model,
                `Figure 1A (%)` = published, `abs diff (pp)`)

knitr::kable(forest, digits = 2,
  caption = "Cavg,ss covariate sensitivity: model vs Wang 2026 Figure 1A.")

# Guard against a vacuous pass: every published row must have been evaluated.
stopifnot(nrow(forest) == nrow(cases), !anyNA(forest$`Model (%)`))
# Deterministic both sides; realised max 0.37 pp. 1.5 pp still goes red on a
# sign error or a materially wrong exponent, which is what this checks.
stopifnot(max(forest$`abs diff (pp)`) < 1.5)
```

Every row reproduces to well under half a percentage point, including the
tumour-type rows -- which is a joint check on all seven `CL0` strata, all seven
`Vc` strata, and the indicator-selection logic in `model()`.

```{r figure1-plot, fig.width = 7, fig.height = 5}
ggplot2::ggplot(forest,
    ggplot2::aes(x = `Figure 1A (%)`, y = `Model (%)`, label = Case)) +
  ggplot2::geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                       colour = "grey50") +
  ggplot2::geom_point(size = 2.5, colour = "steelblue") +
  ggplot2::labs(
    title = "Replicates Figure 1A of Wang 2026",
    subtitle = "Percentage change in steady-state Cavg for a one-covariate-at-a-time change",
    x = "Published (Wang 2026 Figure 1A)", y = "nlmixr2lib model"
  ) +
  ggplot2::theme_bw()
```

# Virtual cohort

Wang 2026 publishes marginal medians and ranges (Table 1) but no individual
covariate table and no covariate correlations, so each covariate is drawn
independently from a distribution centred on the published median and clipped
to the published range. Tumour type is drawn from the published histology
proportions.

```{r virtual-cohort}
set.seed(20260913)
n_sub <- 200L

tumour_levels <- c("Non-sq. NSCLC", "Sq. NSCLC", "SCLC", "ESCC",
                   "CRC", "HCC", "Other")
tumour_probs  <- c(498, 441, 390, 389, 151, 125, 116) / 2110

pop_cohort <- data.frame(
  ID      = seq_len(n_sub),
  # Table 1: WT 62.0 kg (32.9-131)
  WT      = pmin(pmax(rlnorm(n_sub, log(62.0), 0.20), 32.9), 131),
  # Table 1: ALB 41.4 g/L (23.8-67.9). SD is not published; 5.0 g/L reproduces
  # the 10th/90th percentiles (32.8 / 47.7 g/L) that Figure 1 uses as its
  # covariate-sensitivity bounds.
  ALB     = pmin(pmax(rnorm(n_sub, 41.4, 5.0), 23.8), 67.9),
  # Table 1: ALP 94.0 U/L (10.7-910), strongly right-skewed
  ALP     = pmin(pmax(rlnorm(n_sub, log(94.0), 0.55), 10.7), 910),
  # Table 1: TUMBUR 72.9 mm (0-350)
  TUM_SLD = pmin(pmax(rlnorm(n_sub, log(72.9), 0.60), 5), 350),
  # Table 1: 20.62% female
  SEXF    = rbinom(n_sub, 1, 0.2062),
  tumour  = sample(tumour_levels, n_sub, replace = TRUE, prob = tumour_probs)
) |>
  dplyr::mutate(
    TUMTP_NSCLC_SQUAM = as.integer(tumour == "Sq. NSCLC"),
    TUMTP_SCLC        = as.integer(tumour == "SCLC"),
    TUMTP_ESCC        = as.integer(tumour == "ESCC"),
    TUMTP_CRC         = as.integer(tumour == "CRC"),
    TUMTP_HCC         = as.integer(tumour == "HCC"),
    TUMTP_OTHER       = as.integer(tumour == "Other")
  )

# Exactly one histology per subject: the six indicators must never sum above 1.
stopifnot(with(pop_cohort,
  max(TUMTP_NSCLC_SQUAM + TUMTP_SCLC + TUMTP_ESCC +
      TUMTP_CRC + TUMTP_HCC + TUMTP_OTHER) <= 1L))

knitr::kable(
  data.frame(
    Covariate  = c("WT (kg)", "ALB (g/L)", "ALP (U/L)", "TUMBUR (mm)",
                   "Female (%)", "Non-sq. NSCLC (%)"),
    Simulated  = c(round(median(pop_cohort$WT), 1),
                   round(median(pop_cohort$ALB), 1),
                   round(median(pop_cohort$ALP), 1),
                   round(median(pop_cohort$TUM_SLD), 1),
                   round(100 * mean(pop_cohort$SEXF), 1),
                   round(100 * mean(pop_cohort$tumour == "Non-sq. NSCLC"), 1)),
    `Wang 2026 Table 1` = c(62.0, 41.4, 94.0, 72.9, 20.62, 23.60),
    check.names = FALSE
  ),
  caption = "Simulated cohort medians vs Wang 2026 Table 1."
)
```

# Dosing schedule and simulation

The four regimens compared in Wang 2026 Section 3.3 are simulated on the same
cohort, with the same individual random effects, so the flat-versus-weight-based
contrast isolates the dose change.

```{r events}
regimens <- tibble::tribble(
  ~regimen,          ~tau, ~ndose, ~flat_mg, ~mg_per_kg,
  "3 mg/kg Q2W",       14,    12L,       NA,          3,
  "200 mg Q2W",        14,    12L,      200,         NA,
  "4.5 mg/kg Q3W",     21,     8L,       NA,        4.5,
  "300 mg Q3W",        21,     8L,      300,         NA
)

make_events <- function(r) {
  dose_times <- seq(0, by = r$tau, length.out = r$ndose)
  ss_start   <- max(dose_times)
  obs_times  <- sort(unique(c(
    seq(0, 3, by = 0.25),                       # cycle-1 distribution phase
    seq(3, r$tau, by = 0.5),                    # rest of cycle 1
    seq(r$tau, ss_start, by = 7),               # weekly across the middle
    seq(ss_start, ss_start + r$tau, by = 0.5)   # last (steady-state) interval
  )))

  d_dose <- pop_cohort |>
    tidyr::crossing(TIME = dose_times) |>
    dplyr::mutate(
      AMT  = if (is.na(r$flat_mg)) r$mg_per_kg * WT else r$flat_mg,
      EVID = 1L, CMT = "central", DUR = 1 / 24
    )
  d_obs <- pop_cohort |>
    tidyr::crossing(TIME = obs_times) |>
    dplyr::mutate(AMT = NA_real_, EVID = 0L, CMT = "central", DUR = NA_real_)

  ev <- dplyr::bind_rows(d_dose, d_obs) |>
    dplyr::mutate(regimen = r$regimen,
                  ss_start = ss_start, ss_end = ss_start + r$tau) |>
    dplyr::arrange(ID, TIME, dplyr::desc(EVID))

  stopifnot(
    nrow(ev) > 0,
    all(c("WT", "ALB", "ALP", "TUM_SLD", "SEXF") %in% names(ev)),
    # A time-zero observation must exist for PKNCA (see pknca-recipes).
    any(ev$TIME == 0 & ev$EVID == 0L),
    # And one at each end of the final dosing interval.
    any(ev$TIME == ss_start & ev$EVID == 0L),
    any(ev$TIME == ss_start + r$tau & ev$EVID == 0L)
  )
  ev
}
```

```{r simulate}
sim <- lapply(seq_len(nrow(regimens)), function(i) {
  ev <- make_events(regimens[i, ])
  rxode2::rxSetSeed(20260913)   # same etas drawn for every regimen
  out <- rxode2::rxSolve(
    mod, as.data.frame(ev), returnType = "data.frame",
    keep = c("regimen", "WT", "ALB", "ALP", "TUM_SLD", "SEXF", "tumour")
  )
  out$ss_start <- unique(ev$ss_start)
  out$ss_end   <- unique(ev$ss_end)
  out
}) |>
  dplyr::bind_rows()

stopifnot(
  nrow(sim) > 0, !anyNA(sim$Cc), all(sim$Cc >= 0),
  dplyr::n_distinct(sim$regimen) == 4L
)
```

# Replicating Figure 3E - concentration-time profiles

```{r figure3e, fig.width = 7.5, fig.height = 5}
prof <- sim |>
  dplyr::filter(time <= 168) |>
  dplyr::group_by(regimen, time) |>
  dplyr::summarise(
    med = median(Cc),
    lo  = quantile(Cc, 0.05),
    hi  = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(week = time / 7)

ggplot2::ggplot(prof, ggplot2::aes(week, med,
                                   colour = regimen, fill = regimen)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi),
                       alpha = 0.15, colour = NA) +
  ggplot2::geom_line(linewidth = 0.6) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  ggplot2::labs(
    title = "Replicates Figure 3E of Wang 2026",
    subtitle = "Median and 90% prediction interval, n = 200 per regimen",
    x = "Time (weeks)", y = "Serplulimab concentration (ug/mL)"
  ) +
  ggplot2::theme_bw()
```

# Flat versus weight-based dosing

Wang 2026 Section 3.3 makes a sharp, checkable claim: because the two dose
changes are of the same magnitude and the simulations reuse one set of
individual PK parameters, "the relative changes in exposure ... were identical
at 6.59% in both comparisons".

That identity is exact for a linear model. For any subject,

$$
\frac{200}{3 \cdot \mathrm{WT}} = \frac{300}{4.5 \cdot \mathrm{WT}}
= \frac{66.67}{\mathrm{WT}}
$$

so the flat-to-weight-based exposure ratio is the same number for **every**
exposure metric and for **both** comparisons. This is the strongest available
structural gate on the model: any dose non-linearity, any failure of dose
proportionality, or any accidental covariate dependence on dose would break it.

```{r flat-vs-wt}
exposure_by_subject <- sim |>
  dplyr::filter(time >= ss_start) |>
  # rxSolve returns the subject index as lowercase `id`.
  dplyr::rename(ID = id) |>
  dplyr::group_by(regimen, ID, WT) |>
  dplyr::summarise(
    Cavgss = trapz(time, Cc) / (dplyr::first(ss_end) - dplyr::first(ss_start)),
    Cmaxss = max(Cc),
    Cminss = min(Cc),
    .groups = "drop"
  )

ratios <- exposure_by_subject |>
  tidyr::pivot_longer(c(Cavgss, Cmaxss, Cminss),
                      names_to = "metric", values_to = "value") |>
  tidyr::pivot_wider(names_from = regimen, values_from = value) |>
  dplyr::mutate(
    q2w_ratio = `200 mg Q2W` / `3 mg/kg Q2W`,
    q3w_ratio = `300 mg Q3W` / `4.5 mg/kg Q3W`,
    expected  = (200 / 3) / WT
  )

stopifnot(nrow(ratios) == 3L * n_sub)

# Exact identities of a linear model solved on one set of individual
# parameters. Only solver tolerance separates the two sides.
c(`max |Q2W ratio - 66.67/WT|` = max(abs(ratios$q2w_ratio - ratios$expected)),
  `max |Q3W ratio - 66.67/WT|` = max(abs(ratios$q3w_ratio - ratios$expected)),
  `max |Q2W ratio - Q3W ratio|` = max(abs(ratios$q2w_ratio - ratios$q3w_ratio)))

# Realised ~2e-15 (machine epsilon on a ratio of order 1). 1e-6 keeps nine
# orders of headroom for a different solver tolerance while still going red
# on any real loss of dose proportionality.
stopifnot(
  max(abs(ratios$q2w_ratio - ratios$expected))  < 1e-6,
  max(abs(ratios$q3w_ratio - ratios$expected))  < 1e-6,
  max(abs(ratios$q2w_ratio - ratios$q3w_ratio)) < 1e-6
)
```

The published +6.59% therefore pins one number and one number only: the
geometric-mean body weight of the analysis cohort, at
`200 / (3 x 1.0659) = 62.55 kg`. This is an unusually clean readback -- the
paper's headline PK-comparability result is, algebraically, a statement about
its own cohort's weight distribution.

```{r flat-vs-wt-gm}
gm <- function(x) exp(mean(log(x)))
implied_wt <- (200 / 3) / gm(ratios$q2w_ratio)

data.frame(
  Quantity = c("Geometric-mean exposure increase, flat vs weight-based (%)",
               "Implied geometric-mean body weight (kg)",
               "Wang 2026 Section 3.3 reported increase (%)",
               "Body weight implied by the published 6.59% (kg)",
               "Wang 2026 Table 1 median body weight (kg)"),
  Value = c(round(100 * (gm(ratios$q2w_ratio) - 1), 2),
            round(implied_wt, 2),
            6.59,
            round(200 / (3 * 1.0659), 2),
            62.0)
) |>
  knitr::kable(caption = "The published 6.59% is a statement about cohort body weight.")

# Cohort-derived, so assert on the geometric mean and give it room for the
# draw. The published 6.59% corresponds to 62.55 kg and Table 1's median is
# 62.0 kg; a simulated cohort's geometric-mean weight wanders a few percent.
stopifnot(implied_wt > 55, implied_wt < 72)
```

Wang 2026 Section 3.3 further reports that flat dosing raises exposure most in
the lightest weight quartile (+33.9%) and lowers it in the heaviest (-16.95%),
with Q2 and Q3 in between. That gradient is a direct consequence of the same
identity and is reproduced here.

```{r weight-quartiles}
wt_breaks <- quantile(pop_cohort$WT, c(0, 0.25, 0.5, 0.75, 1))
quartiles <- ratios |>
  dplyr::filter(metric == "Cavgss") |>
  dplyr::mutate(Q = cut(WT, breaks = wt_breaks, include.lowest = TRUE,
                        labels = c("Q1", "Q2", "Q3", "Q4"))) |>
  dplyr::group_by(Q) |>
  dplyr::summarise(
    n = dplyr::n(),
    `WT range (kg)` = sprintf("%.1f-%.1f", min(WT), max(WT)),
    `Cavg,ss change (%)` = round(100 * (gm(q2w_ratio) - 1), 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(`Wang 2026 (%)` = c(33.9, 12.93, 0.47, -16.95))

knitr::kable(quartiles,
  caption = "Change in Cavg,ss from weight-based to flat dosing, by body-weight quartile (Wang 2026 Section 3.3 / Figure S4).")

# Trend, not step-by-step monotonicity, and a magnitude bound rather than an
# exact match: the quartile boundaries depend on the simulated weight
# distribution, which is drawn independently of the paper's real one.
stopifnot(
  quartiles$`Cavg,ss change (%)`[1] > quartiles$`Cavg,ss change (%)`[4],
  quartiles$`Cavg,ss change (%)`[1] > 15,
  quartiles$`Cavg,ss change (%)`[4] < 0
)
```

# PKNCA validation

Non-compartmental analysis over the final dosing interval of each regimen,
using `PKNCA` rather than an inline trapezoidal rule.

```{r pknca}
conc_df <- sim |>
  dplyr::filter(time >= ss_start, time <= ss_end) |>
  dplyr::transmute(ID = id, regimen, time, Cc) |>
  dplyr::filter(!is.na(Cc))

# A record must sit exactly at each interval boundary, or PKNCA warns about an
# AUC range starting before the first measurement.
boundaries <- sim |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(lo = dplyr::first(ss_start), hi = dplyr::first(ss_end),
                   .groups = "drop")

boundary_check <- conc_df |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(tmin = min(time), tmax = max(time), .groups = "drop") |>
  dplyr::inner_join(boundaries, by = "regimen")

stopifnot(
  nrow(boundary_check) == 4L,
  all(boundary_check$tmin == boundary_check$lo),
  all(boundary_check$tmax == boundary_check$hi)
)

dose_df <- sim |>
  dplyr::group_by(regimen, ID = id) |>
  dplyr::summarise(time = dplyr::first(ss_start), .groups = "drop") |>
  dplyr::left_join(
    exposure_by_subject |> dplyr::distinct(regimen, ID, WT),
    by = c("regimen", "ID")
  ) |>
  dplyr::left_join(
    regimens |> dplyr::select(regimen, flat_mg, mg_per_kg), by = "regimen"
  ) |>
  dplyr::mutate(AMT = ifelse(is.na(flat_mg), mg_per_kg * WT, flat_mg)) |>
  dplyr::select(regimen, ID, time, AMT) |>
  as.data.frame()

conc_obj <- PKNCA::PKNCAconc(as.data.frame(conc_df), Cc ~ time | regimen + ID,
                             concu = "ug/mL", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, AMT ~ time | regimen + ID, doseu = "mg")

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = boundaries |>
    dplyr::transmute(regimen, start = lo, end = hi,
                     cmax = TRUE, tmax = TRUE, auclast = TRUE) |>
    as.data.frame()
))

nca_df <- as.data.frame(nca$result)
stopifnot(nrow(nca_df) > 0, !all(is.na(nca_df$PPORRES)))

nca_summary <- nca_df |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::left_join(boundaries, by = "regimen") |>
  dplyr::mutate(
    PPORRES = ifelse(PPTESTCD == "auclast", PPORRES / (hi - lo), PPORRES),
    PPTESTCD = ifelse(PPTESTCD == "auclast", "cavg", PPTESTCD)
  ) |>
  dplyr::group_by(regimen, PPTESTCD) |>
  dplyr::summarise(gm = gm(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = gm)

knitr::kable(nca_summary, digits = 1,
  caption = "Geometric-mean steady-state exposure by regimen (PKNCA over the final dosing interval).")
```

## Two published values of the same quantity disagree with each other

Wang 2026 prints two absolute steady-state concentrations in Section 3.3:
geometric-mean `Cmax,ss` of 138.45 ug/mL for 200 mg Q2W and 166.48 ug/mL for
300 mg Q3W. Figure 1B reports the **same** quantity for the **same** 300 mg
Q3W regimen from the **same** model, as a typical-subject value plus a
5th-95th percentile band for the population. Those two published statements do
not agree, and the discrepancy is worth resolving before either is used as a
validation target.

```{r figure1-vs-section33}
# Figure 1B population band for 300 mg Q3W. For a log-normally distributed
# exposure the geometric mean is the geometric midpoint of any symmetric
# percentile pair, so sqrt(p5 * p95) recovers it.
fig1b <- c(p5 = 104, p95 = 202)
fig1_gm  <- sqrt(fig1b[["p5"]] * fig1b[["p95"]])
fig1_typ <- mean(ref_targets["Cmaxss", ])   # recovered earlier from both bounds

data.frame(
  Source = c("Figure 1B, typical subject (recovered from both bounds)",
             "Figure 1B, population geometric mean sqrt(p5 x p95)",
             "Section 3.3, population geometric mean",
             "This model, typical subject",
             "This model, cohort geometric mean"),
  `Cmax,ss (ug/mL)` = round(c(
    fig1_typ, fig1_gm, 166.48,
    unname(ss_metrics()["Cmaxss"]),
    nca_summary$cmax[nca_summary$regimen == "300 mg Q3W"]
  ), 1),
  check.names = FALSE
) |>
  knitr::kable(caption = "Steady-state Cmax at 300 mg Q3W: four published or simulated statements of one quantity.")

# Figure 1B is internally coherent: its typical subject and its population
# geometric mean agree to a couple of percent, as they must for a
# log-normal population centred on the typical subject.
stopifnot(abs(fig1_gm / fig1_typ - 1) < 0.05)
```

Figure 1B is self-consistent -- its typical value and its population geometric
mean agree to within 1% -- and this model reproduces it. The Section 3.3 value
of 166.48 ug/mL sits about 15% above the population geometric mean implied by
the paper's own figure. The comparison below is therefore made against
Section 3.3 (the value a reader would quote), with that offset expected and
documented rather than tuned away.

```{r nca-comparison}
flat_regimens <- c("200 mg Q2W", "300 mg Q3W")

simulated_nca <- nca_summary |>
  dplyr::filter(regimen %in% flat_regimens) |>
  dplyr::select(regimen, cmax) |>
  as.data.frame()

reference_nca <- data.frame(
  regimen = flat_regimens,
  cmax    = c(138.45, 166.48)
)

# Guard against a vacuous comparison: both regimens must be present on both
# sides, with no missing value.
stopifnot(
  setequal(simulated_nca$regimen, flat_regimens),
  !anyNA(simulated_nca$cmax)
)

cmp_tbl <- nlmixr2lib::ncaComparisonTable(
  simulated_nca, reference_nca,
  by = "regimen",
  units = c(cmax = "ug/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_tbl,
  caption = "Simulated vs published geometric-mean steady-state Cmax (Wang 2026 Section 3.3).")
attr(cmp_tbl, "footnote")
```

The ratio between the two published values is the sharper check, because it is
independent of the cohort and of the absolute exposure level.

```{r nca-ratio}
pub_ratio <- 166.48 / 138.45
sim_ratio <- nca_summary$cmax[nca_summary$regimen == "300 mg Q3W"] /
  nca_summary$cmax[nca_summary$regimen == "200 mg Q2W"]

c(`published Cmax,ss ratio (300 Q3W / 200 Q2W)` = round(pub_ratio, 4),
  `simulated ratio` = round(sim_ratio, 4),
  `difference (%)` = round(100 * (sim_ratio / pub_ratio - 1), 2))

# Cohort-derived but a ratio of two arms sharing one set of etas, so most of
# the draw noise cancels. 8% leaves room for the covariate draw while still
# going red on a mis-transcribed dosing interval or volume.
stopifnot(abs(sim_ratio / pub_ratio - 1) < 0.08)
```

# Exposure-response analysis (documented, not encoded)

Wang 2026 devotes Sections 2.5, 3.4 and Figures 4-5 to exposure-response:
indication-specific Cox proportional-hazards models for overall survival in
squamous NSCLC, SCLC and ESCC, and logistic-regression models for objective
response rate in MSI-H patients and for five safety endpoints.

**None of these sub-models can be encoded, because no coefficient is published
anywhere in the sources on disk.** This was checked exhaustively rather than
assumed:

* The main text reports only outcomes ("a 0.1% difference in OS", "predicted
  ORR differing by < 1%", "maximum difference in AE probability ... 2.8%"), never
  a hazard ratio, intercept or slope.
* The Supporting Information (`psp470204-sup-0001-DataS1.docx`) contains one
  table -- Table S1, the trial summary -- and five figures. It contains no
  parameter table of any kind.
* The E-R figure panels themselves were opened and read at native resolution
  (Figure 4, panels A-E; Figure 5, panels A-E). Figure 4D plots the ORR-vs-Cmin1
  logistic curve and Figure 4E its two predicted values (38.1% and 38.2%), but
  no equation, intercept or slope is annotated in any panel.

By contrast, the predecessor paper *does* publish its Cox coefficients (Wang
2025 Table S4), and those are recorded in the
`covariatesDataExcluded` notes of `Wang_2025_serplulimab.R`.

The PK layer is complete and is what this model file encodes. Nothing about the
E-R analysis has been inferred or reconstructed.

# Assumptions and deviations

* **Covariate distributions.** Wang 2026 Table 1 publishes marginal medians and
  ranges only. The virtual cohort draws each covariate independently, so the
  real correlations between weight, albumin, ALP, tumour burden and sex are not
  reproduced. Every cohort-derived assertion in this vignette is therefore
  written on a centre or a robust quantile, never on an extreme.
* **Albumin and ALP dispersion.** No SD is published for either. Albumin uses
  5.0 g/L, which reproduces the 32.8 / 47.7 g/L 10th and 90th percentiles that
  Figure 1 uses; ALP uses a log-normal with sdlog 0.55 to match its strong right
  skew (median 94, maximum 910 U/L).
* **The `x 24` factor in the Section 3.1 equations.** The printed `CL0` and `Q`
  equations both end in `x 24`, converting an L/h theta to L/day. Table 2
  reports `CL0` and `Q` already in L/day, so the model uses Table 2's values
  directly and does **not** re-apply the factor. This is confirmed three ways:
  `0.0169 x 24 = 0.4056` reproduces Table 2's `Q = 0.405`; the resulting
  terminal half-life is 26 days, the textbook IgG4 value, where a further
  factor of 24 would give 5.6 days; and the reference-subject exposures
  recovered from Figure 1 are matched to within 2.5%.
* **`cl_time_max` is not log-transformed.** `checkModelConventions()` emits a
  warning suggesting `lcl_time_max`. The suggestion cannot be followed:
  `cl_time_max` is the Emax of a log-scale sigmoid and is negative (-0.0926), so
  it has no logarithm. `cl_time_max` is the canonical role name registered in
  `parameter-names.md` for exactly this structure, and the sibling model
  `Wang_2025_serplulimab.R` carries the identical warning. This is a known gap
  in the checker, not a naming deviation.
* **Tumour type as strata, not as a covariate coefficient.** Following
  Section 2.2, the seven histologies are encoded as seven parallel `CL0` and
  `Vc` estimates rather than as a reference plus six offsets. The six
  indicators are mutually exclusive and non-squamous NSCLC is their complement,
  so a subject with all six at 0 receives the reference stratum. A user who
  sets two indicators to 1 will get a nonsensical blend; the cohort chunk above
  asserts against that.
* **The two published `Cmax,ss` statements disagree with each other.** For
  300 mg Q3W, Section 3.3 reports a population geometric mean of 166.48 ug/mL,
  while Figure 1B reports a typical subject at 144.4 ug/mL with a 5th-95th
  percentile band of 104-202 ug/mL, whose geometric midpoint is 144.9 ug/mL.
  Figure 1B is internally coherent and this model reproduces it to within 1%;
  Section 3.3 sits about 15% higher. The simulated-versus-published comparison
  above is made against Section 3.3 and therefore carries that offset. It is
  **not** tuned away: the model agrees with the paper's figure, and the
  discrepancy is between two of the paper's own numbers. The ratio of the two
  Section 3.3 values is checked separately because it is insensitive to
  whatever caused the level shift, and it matches to 0.8%.
* **`Cmin1` from the simulation, not from PKNCA.** Where a trough at an exact
  time is needed it is read from the solve. PKNCA's `cmin` over an interval
  starting at zero returns the time-zero anchor rather than the trough.
* **No residual error on the validation quantities.** `Cc` is the individual
  prediction. Wang 2026's exposure metrics are likewise computed from
  model-predicted profiles, so this is the like-for-like comparison.

# Errata

* No erratum, corrigendum or author correction was located for this article. It
  was published online 2026 following acceptance on 12 January 2026; the PubMed
  Central record (PMC13140920) and the Wiley landing page for
  `doi:10.1002/psp4.70204` show no correction notice.
* **Trial-identifier discrepancy against the predecessor paper.** Wang 2026
  Table S1 and Wang 2025 assign different NCT numbers to three of the shared
  trials: HLX10-001 is NCT03468751 here (NCT03952403 in Wang 2025, which is this
  paper's HLX10-002-NSCLC301); HLX10-008-HCC201 is NCT03973112 here
  (NCT05246164 in Wang 2025); HLX10-010-MSI201 is NCT03941574 here
  (NCT04747236 in Wang 2025). The identifiers recorded in this model's
  `population$notes` are those of Wang 2026 Table S1, which is the authoritative
  source for this model's dataset. The discrepancy is noted but not resolved.
* **The 10 mg/kg Q2W reference arm does not scale.** Wang 2026 Section 3.3 gives
  a geometric-mean `Cmax,ss` of 383.6 ug/mL for 10 mg/kg Q2W against 138.45
  ug/mL for 200 mg Q2W. For a linear model simulated on one set of individual
  parameters, that ratio should be the dose ratio, `10 x 62.55 / 200 = 3.13`,
  but the printed values give 2.77. The most likely explanation is that the
  10 mg/kg arm -- studied only in the two Phase I trials -- was summarised over
  a different subject subset. This arm is therefore not used as a validation
  target here.
