---
title: "Caspofungin (Wang 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Caspofungin (Wang 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wang_2020_caspofungin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wang_2020_caspofungin"))$description`
* Article: <https://doi.org/10.1128/AAC.00687-20>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7577146/>

Wang 2020 studied caspofungin given as antifungal prophylaxis to adult lung
transplant recipients in the first days after surgery. Twelve patients needed
veno-venous ECMO; they were sampled on ECMO (the "ECMO group") and again after
weaning, as their own controls ("control group A"). Seven patients who never
needed ECMO formed "control group B". ECMO did not affect caspofungin PK. The
final two-compartment model keeps three covariates: operative time on CL and Vc,
male sex on Vc, and the SOFA score on Q.

# Population

Nineteen patients from one centre in Beijing, China (October 2017 - March 2018)
gave 271 plasma samples over 31 dosing-interval profiles (Wang 2020 Results and
Table 1). The ECMO group had a median age of 65 years, weight 64 kg, 25% women
and a median operative time of 4.0 h (IQR 3.5-5.0). Control group B had a median
age of 59 years, weight 65 kg, 28.6% women and a median operative time of 5.5 h
(5.0-5.8). The median SOFA score on the sampling day was 7 (6-10) in the ECMO
group, 6 (5-9) after weaning, and 7 (6-8) in control group B (Table 2).
Patients on renal replacement therapy were excluded. Every patient received
50 mg caspofungin intravenously every 24 h, with no loading dose. The first dose
was given immediately after surgery.

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Two-compartment model, IV infusion | -- | Results, 'Establishment of a pharmacokinetic model' |
| `lcl` (CL at T_SURG = 5 h) | 0.21 L/h | Table 4 |
| `lvc` (Vc, female, T_SURG = 5 h) | 2.21 L | Table 4 |
| `lvp` (Vp) | 2.87 L | Table 4 |
| `lq` (Q at SOFA = 7) | 0.84 L/h | Table 4 |
| `e_t_surg_cl` | 1.30 | Table 4, 'Theta OPT on CL' |
| `e_t_surg_vc` | 0.93 | Table 4, 'Theta OPT on VC' |
| `e_sex_vc` (additive, men) | 0.62 L | Table 4, 'Theta SEX on VC' |
| `e_sofa_q` | 1.98 | Table 4, 'Theta SOFA on Q' |
| `etalcl`, `etalvc`, `etalvp` | 0.04, 0.01, 0.23 | Table 4, interindividual variability |
| `addSd` | 0.73 mg/L | Table 4, residual error |
| CL = CL_TV (OPT/5)^theta exp(eta) | -- | Results, first final-model equation |
| Vc = (Vc_TV + SEX theta) (OPT/5)^theta exp(eta) | -- | Results, second final-model equation |
| Q = Q_TV (SOFA/7)^theta | -- | Not printed. Power form from Methods, 'Establishment of a covariate model'. Centring 7 = Table 2 median |
| Dose 50 mg q24h, sampling 0-24 h | -- | Methods, 'Drug regimen and the collection of pharmacokinetic data' |

# Virtual cohort

Each arm has 100 subjects. Operative time and SOFA are drawn log-normally, using
each group's median and an SD taken from its IQR. SOFA is rounded to an integer,
and non-positive draws are redrawn. The simulated ECMO patients are reused as
control group A, as in the study. Their SOFA score drops by one point after
weaning (the Table 2 medians fall from 7 to 6). Their weaning occasion is the
third dose, 48-72 h, which is the interval Table 3 reports for group A.

```{r cohort}
set.seed(2020)
rxode2::rxSetSeed(2020)
n_arm <- 100

draw_lnorm <- function(n, med, q25, q75, integer = FALSE, min_val = 0) {
  sdl <- log(q75 / q25) / 1.349
  out <- numeric(0)
  while (length(out) < n) {
    x <- exp(rnorm(n, log(med), sdl))
    if (integer) x <- round(x)
    out <- c(out, x[x > min_val])
  }
  out[seq_len(n)]
}

ecmo <- data.frame(
  id = seq_len(n_arm),
  group = "ECMO / control A",
  T_SURG = draw_lnorm(n_arm, 4.0, 3.5, 5.0),
  SOFA = draw_lnorm(n_arm, 7, 6, 10, integer = TRUE),
  SEXF = rbinom(n_arm, 1, 0.25)
)
ecmo$SOFA_A <- pmax(ecmo$SOFA - 1, 1)

ctlb <- data.frame(
  id = n_arm + seq_len(n_arm),
  group = "Control B",
  T_SURG = draw_lnorm(n_arm, 5.5, 5.0, 5.8),
  SOFA = draw_lnorm(n_arm, 7, 6, 8, integer = TRUE),
  SEXF = rbinom(n_arm, 1, 0.286)
)
ctlb$SOFA_A <- ctlb$SOFA

subj <- dplyr::bind_rows(ecmo, ctlb)
subj |>
  dplyr::group_by(group) |>
  dplyr::summarise(
    median_T_SURG = median(T_SURG), median_SOFA = median(SOFA),
    pct_female = 100 * mean(SEXF), .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "Virtual cohort covariates.")
```

The infusion duration is not stated in the paper. A 0.5 h infusion is used
(see Assumptions and deviations).

```{r events}
inf_dur <- 0.5
dose_times <- c(0, 24, 48)
obs_times <- sort(unique(c(seq(0, 72, by = 0.25), 24 + c(0.5, 1, 2, 4, 8, 12, 24))))

make_events <- function(s) {
  dose <- data.frame(
    id = s$id, time = dose_times, evid = 1, amt = 50, rate = 50 / inf_dur,
    cmt = "central"
  )
  obs <- data.frame(id = s$id, time = obs_times, evid = 0, amt = 0, rate = 0, cmt = "central")
  ev <- rbind(dose, obs)
  ev$T_SURG <- s$T_SURG
  ev$SEXF <- s$SEXF
  # SOFA is time-varying for the ECMO patients: the weaning occasion (the
  # 48-72 h interval) uses the post-weaning score.
  ev$SOFA <- ifelse(ev$time >= 48, s$SOFA_A, s$SOFA)
  ev$group <- s$group
  ev[order(ev$time, -ev$evid), ]
}
events <- do.call(rbind, lapply(split(subj, subj$id), make_events))
```

# Simulation

```{r simulate}
mod <- readModelDb("Wang_2020_caspofungin")
sim <- rxode2::rxSolve(mod, events, keep = c("group"), returnType = "data.frame")
```

## Replicate Figure 2 (median concentration-time curves)

Figure 2 of Wang 2020 shows median observed concentrations over the 24-48 h
interval (the 2nd dose) for the ECMO group and control group B, and over
48-72 h (the weaning occasion) for control group A. Below are the simulated
medians and 5th-95th percentiles over the same intervals. The points are the
observed Table 3 median Cmax and Cmin.

```{r fig2}
seg <- function(d, lo, lab) {
  d |>
    dplyr::filter(time >= lo, time <= lo + 24) |>
    dplyr::mutate(tad = time - lo, arm = lab)
}
fig_dat <- dplyr::bind_rows(
  seg(dplyr::filter(sim, group == "ECMO / control A"), 24, "ECMO group"),
  seg(dplyr::filter(sim, group == "ECMO / control A"), 48, "Control group A"),
  seg(dplyr::filter(sim, group == "Control B"), 24, "Control group B")
) |>
  dplyr::group_by(arm, tad) |>
  dplyr::summarise(
    med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
    .groups = "drop"
  )

obs_pts <- data.frame(
  arm = rep(c("ECMO group", "Control group A", "Control group B"), 2),
  tad = rep(c(0.5, 24), each = 3),
  conc = c(16.7, 18.3, 17.9, 3.5, 3.2, 3.5)
)

ggplot(fig_dat, aes(tad, med, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line() +
  geom_point(data = obs_pts, aes(tad, conc), size = 2.5, shape = 21, colour = "black") +
  labs(
    x = "Time after start of infusion (h)", y = "Caspofungin (mg/L)",
    colour = NULL, fill = NULL,
    caption = "Replicates Figure 2 of Wang 2020. Points: observed median Cmax / Cmin (Table 3)."
  ) +
  theme_bw()
```

# PKNCA validation

```{r pknca}
nca_intervals <- dplyr::bind_rows(
  data.frame(group = "ECMO / control A", arm = "ECMO group", start = 24, end = 48),
  data.frame(group = "ECMO / control A", arm = "Control group A", start = 48, end = 72),
  data.frame(group = "Control B", arm = "Control group B", start = 24, end = 48)
)

run_nca <- function(k) {
  iv <- nca_intervals[k, ]
  conc <- sim |>
    dplyr::filter(group == iv$group, !is.na(Cc)) |>
    dplyr::mutate(arm = iv$arm) |>
    dplyr::select(id, time, Cc, arm)
  dose <- events |>
    dplyr::filter(group == iv$group, evid == 1) |>
    dplyr::mutate(arm = iv$arm) |>
    dplyr::select(id, time, amt, arm)
  conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | arm + id, concu = "mg/L", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | arm + id, doseu = "mg")
  intervals <- data.frame(
    start = iv$start, end = iv$end,
    cmax = TRUE, cmin = TRUE, auclast = TRUE
  )
  as.data.frame(PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))$result)
}
nca_res <- dplyr::bind_rows(lapply(seq_len(nrow(nca_intervals)), run_nca))

nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) |>
  dplyr::group_by(arm, PPTESTCD) |>
  dplyr::summarise(
    median = median(PPORRES), q25 = quantile(PPORRES, 0.25),
    q75 = quantile(PPORRES, 0.75), .groups = "drop"
  ) |>
  dplyr::mutate(parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  dplyr::select(arm, parameter, median, q25, q75) |>
  dplyr::rename(
    "Group" = arm, "NCA parameter" = parameter, "Median" = median,
    "25th percentile" = q25, "75th percentile" = q75
  ) |>
  knitr::kable(digits = 2, caption = "Simulated NCA of the sampled interval (Cmax, Cmin in mg/L; AUC in mg.h/L).")
```

## Comparison against the published NCA (Table 3)

```{r nca-compare}
published <- data.frame(
  arm = c("ECMO group", "Control group A", "Control group B"),
  cmax = c(16.7, 18.3, 17.9),
  cmin = c(3.5, 3.2, 3.5),
  auclast = c(163.06, 147.11, 156.2)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "arm",
  params = c("cmax", "cmin", "auclast"),
  units = c(cmax = "mg/L", cmin = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs published medians (Wang 2020 Table 3). * differs by >20%.")
attr(cmp, "footnote")
```

The model reproduces control group B, whose median operative time (5.5 h) is
close to the 5 h centring value: Cmax and AUC are within about 10%. It
overpredicts the two groups with short operations (median 4.0 h). Their Cmax is
about 35% high, and their AUC is about 45% high on ECMO and 80% high after
weaning. Control group A also shows some accumulation in the model, because its
interval is the third dose. The observed group A AUC is, if anything, lower
than the ECMO group's. The final-model estimates in Table 4 cause this, not the
simulation. Two things show it:

* Table 3 reports individual clearances with medians of 0.27-0.31 L/h in every
  group. The Table 4 typical CL is 0.21 L/h at 5 h of operative time, and only
  0.16 L/h at the ECMO group's median of 4.0 h. Likewise, Table 3 reports
  individual Vc medians of 2.9-3.2 L, but the model gives a typical 2.30 L for
  a man at 4.0 h. With IIV variances of only 0.04 on CL and 0.01 on Vc, the
  model cannot give the Table 3 values.
* The Table 4 point estimates for CL (0.21), Vc (2.21) and Q (0.84) all lie
  outside their own bootstrap 95% CIs (0.23-0.29, 2.45-2.89 and 0.86-1.28). The
  Results say that they lie inside.

The comparison below uses typical values: the zero-IIV solve at each group's
median covariates, for a man. It sets the published estimates beside the
bootstrap medians from the same table. The bootstrap medians reduce the AUC
overprediction, which confirms that the Table 4 estimates cause the gap. The
model file keeps the published final estimates. The bootstrap medians are only
a diagnostic here and are never used in the model.

```{r typical-values}
typ_subj <- data.frame(
  arm = c("ECMO group", "Control group A", "Control group B"),
  T_SURG = c(4.0, 4.0, 5.5), SOFA = c(7, 6, 7), SEXF = 0,
  start = c(24, 48, 24)
)

typ_auc <- function(m) {
  out <- lapply(seq_len(nrow(typ_subj)), function(k) {
    s <- typ_subj[k, ]
    ev <- rxode2::et(amt = 50, rate = 50 / inf_dur, ii = 24, addl = 2, cmt = "central") |>
      rxode2::et(seq(0, 72, by = 0.05))
    ev <- as.data.frame(ev)
    ev$T_SURG <- s$T_SURG
    ev$SOFA <- s$SOFA
    ev$SEXF <- s$SEXF
    r <- rxode2::rxSolve(rxode2::zeroRe(m), ev, returnType = "data.frame")
    r <- r[r$time >= s$start & r$time <= s$start + 24, ]
    data.frame(
      arm = s$arm,
      auc = sum(diff(r$time) * (head(r$Cc, -1) + tail(r$Cc, -1)) / 2),
      cmin = r$Cc[nrow(r)]
    )
  })
  do.call(rbind, out)
}

boot_mod <- mod |>
  rxode2::ini(lcl = log(0.26), lvc = log(2.61), lvp = log(2.98), lq = log(1.00),
              e_t_surg_cl = 1.31, e_sex_vc = 0.64, e_t_surg_vc = 0.87, e_sofa_q = 2.04)

tv <- typ_auc(mod) |>
  dplyr::rename(auc_final = auc, cmin_final = cmin) |>
  dplyr::left_join(
    typ_auc(boot_mod) |> dplyr::rename(auc_boot = auc, cmin_boot = cmin),
    by = "arm"
  ) |>
  dplyr::left_join(published[, c("arm", "auclast")], by = "arm") |>
  dplyr::mutate(
    ratio_final = auc_final / auclast,
    ratio_boot = auc_boot / auclast
  )

tv |>
  dplyr::select(arm, auclast, auc_final, ratio_final, auc_boot, ratio_boot) |>
  dplyr::rename(
    "Group" = arm, "Observed median AUC (Table 3)" = auclast,
    "Typical AUC, Table 4 estimates" = auc_final, "Ratio (estimates)" = ratio_final,
    "Typical AUC, bootstrap medians" = auc_boot, "Ratio (bootstrap)" = ratio_boot
  ) |>
  knitr::kable(digits = 2, caption = "Typical-value AUC over the sampled interval (mg.h/L), male, group-median covariates.")

# Deterministic (no random draws), so tight bounds are correct. They pin the
# documented overprediction of the published estimates, and the bootstrap
# medians' smaller gap.
ratio <- setNames(tv$ratio_final, tv$arm)
stopifnot(
  abs(ratio[["ECMO group"]] - 1.48) < 0.03,
  abs(ratio[["Control group A"]] - 1.90) < 0.03,
  abs(ratio[["Control group B"]] - 1.14) < 0.03,
  all(tv$ratio_boot < tv$ratio_final)
)
```

## Structural checks

At steady state, the AUC over one dosing interval equals Dose / CL. The check
uses the typical-value model and runs the PKNCA AUC over the 20th interval.

```{r mass-balance}
ss_ev <- rxode2::et(amt = 50, rate = 50 / inf_dur, ii = 24, addl = 19, cmt = "central") |>
  rxode2::et(seq(0, 480, by = 0.05)) |>
  as.data.frame()
ss_ev$T_SURG <- 5
ss_ev$SOFA <- 7
ss_ev$SEXF <- 0
ss_ev$id <- 1L
ss <- rxode2::rxSolve(rxode2::zeroRe(mod), ss_ev, returnType = "data.frame")
ss$id <- 1L

conc_ss <- PKNCA::PKNCAconc(ss[, c("id", "time", "Cc")], Cc ~ time | id)
dose_ss <- PKNCA::PKNCAdose(ss_ev[ss_ev$evid == 1, c("id", "time", "amt")], amt ~ time | id)
ss_nca <- as.data.frame(PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_ss, dose_ss,
  intervals = data.frame(start = 456, end = 480, auclast = TRUE)
))$result)
auc_tau <- ss_nca$PPORRES[ss_nca$PPTESTCD == "auclast"]
c(auc_tau = auc_tau, dose_over_cl = 50 / 0.21)
stopifnot(abs(auc_tau * 0.21 / 50 - 1) < 0.005)
```

The covariate equations are checked against hand calculation.

```{r covariate-check}
chk <- ss[1, c("cl", "vc", "vp", "q")]
stopifnot(
  abs(chk$cl - 0.21) < 1e-8, abs(chk$vc - (2.21 + 0.62)) < 1e-8,
  abs(chk$vp - 2.87) < 1e-8, abs(chk$q - 0.84) < 1e-8
)
one <- function(T_SURG, SOFA, SEXF) {
  e <- data.frame(id = 1L, time = c(0, 1), evid = c(1, 0), amt = c(50, 0),
                  cmt = "central", T_SURG = T_SURG, SOFA = SOFA, SEXF = SEXF)
  r <- rxode2::rxSolve(rxode2::zeroRe(mod), e, returnType = "data.frame")
  r[1, c("cl", "vc", "q")]
}
m4 <- one(4, 10, 0)
stopifnot(
  abs(m4$cl - 0.21 * (4 / 5)^1.30) < 1e-8,
  abs(m4$vc - (2.21 + 0.62) * (4 / 5)^0.93) < 1e-8,
  abs(m4$q - 0.84 * (10 / 7)^1.98) < 1e-8
)
f4 <- one(4, 10, 1)
stopifnot(abs(f4$vc - 2.21 * (4 / 5)^0.93) < 1e-8)
```

# Assumptions and deviations

* **The Q equation is not printed.** The Results print only the CL and Vc
  equations, but Table 4 has an estimate for "Theta SOFA on Q". The Q term uses
  the paper's general continuous-covariate form, (Cov / Cov median)^theta
  (Methods), centred at a SOFA of 7. That is the median of both the ECMO group
  and control group B (Table 2). The weaning occasions had a median of 6, so
  the median over all 31 occasions may be 6 or 7. A centring of 6 would raise
  typical Q by (7/6)^1.98, or 36%, at any given SOFA score.
* **Sex coding.** In the printed Vc equation, sex is an additive shift,
  (Vc_TV + SEX x 0.62) L. The Methods describe categorical covariates as
  exp(theta) multipliers, but where the text and a printed equation disagree,
  the equation is followed. The abstract and Discussion say that male sex
  increases the distribution volume, so SEX = 1 is taken to mean male. It enters
  as (1 - SEXF). The 2.21 L typical value is therefore the female value.
* **Residual error.** The Methods describe a combined proportional-plus-additive
  error. The Results text calls it "proportional", and Table 4 lists a single
  "Additive (mg/liter)" term of 0.73. With the units in mg/L, this term is
  encoded as an additive SD. No proportional value is reported, so none is
  invented.
* **IIV scale.** The Table 4 interindividual-variability values (0.04, 0.01,
  0.23) are read as omega^2 (variances of eta). These are the values NONMEM
  estimates, and none of them is labelled as a CV%.
* **Infusion duration.** The paper does not state the infusion duration. In
  Figure 2 the median concentration is higher at 0.5 h than at 1 h. The
  infusion therefore ended by 0.5 h, and a 0.5 h infusion is used. The
  licensed ~1 h infusion would give a lower 0.5 h concentration than the one
  observed.
* **The published estimates do not reproduce the published exposures.** See
  the NCA comparison. The final-model estimates are extracted as published and
  have not been adjusted. The bootstrap medians are shown only as a diagnostic.
* **Cohort.** The covariate distributions come from group medians and IQRs. The
  weaning-occasion SOFA is the on-ECMO value minus one point. Figure 4 (the VPC)
  cannot be replicated without the observed data.
