---
title: "Valproic acid syrup vs sustained-release tablets in children (Wang 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valproic acid syrup vs sustained-release tablets in children (Wang 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20240906)
```

# Model and source

Wang et al. (2024) developed a one-compartment population pharmacokinetic model
with first-order absorption for total plasma valproic acid (VPA) in children
with epilepsy, from 1411 routine therapeutic-drug-monitoring (TDM) trough
concentrations in 617 patients at a single centre in Nanjing, China. The
question the paper is built around is a practical one: **VPA syrup and
sustained-release (SR) tablets are not interchangeable milligram for milligram,
so how should the dose be changed when a child is switched from one to the
other?**

The final model is:

$$
\mathrm{CL/F}\ (\mathrm{L/h}) = 0.196 \times
\left(\frac{\mathrm{BW}}{22}\right)^{0.717}
\times \left(1 - 0.0436 \times \mathrm{Sex}\right) \times e^{\eta_{CL}}
$$

$$
\mathrm{V/F}\ (\mathrm{L}) = 2.09 \times \left(\frac{\mathrm{BW}}{22}\right)^{0.524}
$$

with `Sex = 0` for male and `Sex = 1` for female, and with the absorption rate
constant FIXED by formulation at `ka = 2.64 1/h` for syrup and `ka = 0.46 1/h`
for SR tablets.

Two features of this model are worth flagging before any simulation, because
both are consequences of a trough-only TDM dataset rather than modelling
choices:

* **Both `ka` values are fixed from the literature**, not estimated. The dataset
  contains no absorption-phase samples, so absorption is unidentifiable from it.
  The 2.64 / 0.46 pair is a valproate-literature convention that this paper
  takes from Mei 2018 and that the sibling `Zhang_2023_*` and
  `Zhang_2024_valproic_acid` models take from Ding 2015.
* **There is no interindividual variability on `V/F`.** It "was not informative
  enough to be calculated because only the C~trough~ samples of VPA were
  available, and then were excluded from the model."

```{r load-model}
mod <- modellib("Wang_2024_valproic_acid")
mod

# Parsed once here so the metadata lists below can be read off the model file
# itself. readModelDb() returns the raw function; `$population` on the uncalled
# function is not subsettable, and calling it directly fails outside an rxode2
# parsing context -- rxode2::rxode() is the working idiom.
mod_meta <- rxode2::rxode(readModelDb("Wang_2024_valproic_acid"))
```

## Population

```{r population, echo = FALSE}
p <- mod_meta$population
tibble::tibble(Field = names(p), Value = unlist(p, use.names = FALSE)) |>
  knitr::kable(caption = "Study population (Wang 2024 Table 2 and Methods).")
```

The metadata above describes the **model-development** set (498 patients, 1138
samples), which is what Table 3 was fitted to; a further 119 patients with 273
samples were held out for external evaluation. The cohort is young and light --
median age 5.8 years, median body weight 22.0 kg -- and every sample is a
steady-state trough drawn under routine TDM.

## Source trace

Every value in the model file, with the location it came from.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Value, ~Source,
  "Structural model", "1-compartment, first-order absorption (NONMEM ADVAN2 TRANS2)", "Methods, Base model; Results, Base model",
  "ODE: depot", "d(depot)/dt = -ka * depot", "Methods, Base model (ADVAN2 TRANS2)",
  "ODE: central", "d(central)/dt = ka * depot - kel * central", "Methods, Base model (ADVAN2 TRANS2)",
  "Observation", "Cc = central / vc", "Methods, Base model",
  "Ka (oral syrup)", "2.64 1/h, FIXED", "Methods, Base model; Table 3 (Ka1, 'Fixed'); from Mei 2018 (ref. 19)",
  "Ka (SR tablet)", "0.46 1/h, FIXED", "Methods, Base model; Table 3 (Ka2, 'Fixed'); from Mei 2018 (ref. 19)",
  "CL/F typical", "0.196 L/h at 22 kg, male", "Equation 34; Table 3 (0.196, RSE 3%, CI [0.183, 0.208])",
  "V/F typical", "2.09 L at 22 kg", "Equation 35; Table 3 (2.09, RSE 9%, CI [1.69, 2.49])",
  "BW normalising value", "22 kg (cohort median)", "Equations 34-35 (literal '22'); Table 2 (BW median 22.0 kg)",
  "BW exponent on CL/F", "0.717", "Equation 34; Table 3 ('BW on CL/F', RSE 4%, CI [0.659, 0.775])",
  "BW exponent on V/F", "0.524", "Equation 35; Table 3 ('BW on V/F', RSE 15%, CI [0.361, 0.686])",
  "Female sex on CL/F", "-0.0436, PROPORTIONAL (1 - 0.0436 * Sex)", "Equation 34; Table 3 ('Sex on CL/F', RSE 29%, CI [-0.0685, -0.0187])",
  "Sex coding", "Sex = 0 male, Sex = 1 female", "Stated immediately below Equation 35",
  "IIV on CL/F", "11.9% CV -> omega^2 = log(CV^2+1) = 0.014062", "Table 3 ('IIV on CL/F (%)', RSE 6%, CI [10.4, 13.2]); Equation 1",
  "IIV on V/F", "not estimated; term omitted", "Results, Base model ('not informative enough to be calculated')",
  "Residual error", "combined proportional + additive", "Results, Base model (Equation 5 best fit)",
  "RV proportional", "variance 0.0151 -> propSd = 0.12288", "Table 3 ('RVprop', RSE 15%, CI [0.0105, 0.0198])",
  "RV additive", "variance 22.6 -> addSd = 4.7539 mg/L", "Table 3 ('RVadd', RSE 35%, CI [6.58, 38.5])"
) |>
  knitr::kable(caption = "Source trace for every model equation and ini() value.")
```

Three transcription decisions in this model are load-bearing and are recorded
here in full, because each had a superficially plausible alternative.

**1. The sex effect is proportional, not exponential.** The paper offers both
forms in its Methods (Equation 27 proportional, Equation 28 exponential) and
Equation 34 uses the proportional one: `(1 - 0.0436 * Sex)`. The model file
therefore applies the sex term as a multiplicative factor **outside** the
exponential, so the encoded number is exactly the tabulated `-0.0436` with no
back-transformation. Encoding it as `exp(-0.0436)` would be a 0.1% error here,
but it would be the wrong functional form.

**2. `IIV on CL/F (%) = 11.9` is a CV%, not a variance.** This matters: read as
a variance it would imply a 35.5% CV, a threefold difference. It is settled by
the authors' own Figure 2, whose boxes are "the median and IQR of C~trough~ for
each dosing regimen" over their simulated patients. The drawn boxes have a
Q3/Q1 ratio near 1.2 -- for instance the 20 kg male / 40 mg/kg/day syrup box
spans roughly 92-112 mg/L about a median near 102 -- which for a log-normal
implies a log-scale SD near 0.15. That is consistent with an 11.9% CV plus this
model's residual error, and flatly inconsistent with a 35.5% CV, which would
place the same panel's Tukey whiskers below 10 and above 190 mg/L instead of
the drawn 75-140 mg/L.

**3. `RVprop` and `RVadd` are NONMEM variances.** Table 3 gives 0.0151 and 22.6;
the model file encodes their square roots, 0.12288 and 4.7539 mg/L. The check
is the assay: combining the two at the cohort median trough of 68.6 mg/L gives
`sqrt((0.12288 * 68.6)^2 + 4.7539^2) / 68.6 = 14.1%`, against an EMIT assay
whose within- and between-run coefficients of variation are "below 15%". Reading
them as standard deviations instead would put the additive term alone at
22.6 mg/L, a third of the median trough, which no assay-limited TDM dataset
supports.

Separately, Equation 5 is **typeset** as `Y = IPRED * (1 + eps1 + eps2)`, which
would make both error terms proportional. That contradicts Table 3's own row
labels (`RVprop`, `RVadd`) and is arithmetically impossible at the tabulated
magnitude -- a proportional variance of 22.6 is a 475% CV. It is read here as a
typesetting slip for the standard NONMEM combined form
`Y = IPRED * (1 + eps1) + eps2`.

# Verification

Throughout this section `tau = 12 h`, matching the paper's own simulations
("doses ranging from 10 to 40 mg/kg/day per 12 h").

```{r constants}
tau <- 12

# Add the model's covariates to an rxode2 event table. Covariates MUST be
# attached after as.data.frame() -- assigning columns onto an rxEt object is
# silently dropped by rxode2.
add_cov <- function(ev, WT, SEXF, FORM_VPA_SR) {
  d <- as.data.frame(ev)
  d$WT <- WT
  d$SEXF <- SEXF
  d$FORM_VPA_SR <- FORM_VPA_SR
  d
}
```

## The model reproduces the paper's own per-kg summary

Immediately below Equation 35 the authors summarise their typical patient:
"the typical value for the CL/F of VPA in children at steady state obtained in
our study was 0.009 L/h/kg and the V/F was 0.1 L/kg". Dividing the tabulated
`CL/F` and `V/F` by the normalising weight must recover exactly that. This
is also what confirms the literal "22" in Equations 34-35 is the 22.0 kg cohort
median from Table 2, rather than some other reference weight.

```{r per-kg}
per_kg <- tibble::tibble(
  Parameter = c("CL/F", "V/F"),
  `Typical value at 22 kg` = c(0.196, 2.09),
  `Per kg` = c(0.196 / 22, 2.09 / 22),
  `Paper's stated per-kg value` = c(0.009, 0.1)
)

knitr::kable(per_kg, digits = 6,
             caption = "Per-kg typical values against the paper's own prose.")

# Deterministic: these are arithmetic on tabulated values, so they are checked
# to the precision the paper rounds to.
stopifnot(
  round(0.196 / 22, 3) == 0.009,
  round(2.09 / 22, 1) == 0.1
)
```

## The solved model reproduces a closed-form steady-state trough

For a one-compartment first-order-absorption model the steady-state
concentration at the end of a dosing interval has a closed form. Comparing it
against `rxSolve` on the packaged model exercises every transcribed value at
once -- both `ka` values, `CL/F`, `V/F`, both body-weight exponents, the
reference weight and the sex term. Because the two sides use the *same*
parameters and differ only by numerical integration, a tight bound is the
correct assertion here.

```{r closed-form}
ctrough_closed_form <- function(BW, SEXF, mgkgd, sr) {
  cl <- 0.196 * (BW / 22)^0.717 * (1 - 0.0436 * SEXF)
  vc <- 2.09 * (BW / 22)^0.524
  ka <- if (sr) 0.46 else 2.64
  kel <- cl / vc
  D <- BW * mgkgd / 2                      # q12h => half the daily dose
  (D * ka / (vc * (ka - kel))) *
    (exp(-kel * tau) / (1 - exp(-kel * tau)) -
       exp(-ka * tau) / (1 - exp(-ka * tau)))
}

ctrough_solved <- function(BW, SEXF, mgkgd, sr) {
  ev <- rxode2::et(amt = BW * mgkgd / 2, ii = tau, ss = 1, cmt = "depot") |>
    rxode2::et(tau, cmt = "central")
  d <- add_cov(ev, WT = BW, SEXF = SEXF, FORM_VPA_SR = as.numeric(sr))
  r <- rxode2::rxSolve(mod, d, returnType = "data.frame", omega = NA)
  r$Cc[!is.na(r$Cc)][1]
}

cf_check <- tidyr::crossing(BW = c(20, 30, 50), SEXF = c(0, 1),
                            mgkgd = c(20, 25), sr = c(FALSE, TRUE)) |>
  rowwise() |>
  mutate(closed_form = ctrough_closed_form(BW, SEXF, mgkgd, sr),
         solved      = ctrough_solved(BW, SEXF, mgkgd, sr),
         rel_diff    = abs(solved - closed_form) / closed_form) |>
  ungroup()

# Deterministic identity: same parameters on both sides, so the only difference
# is solver tolerance. Anything above 1e-6 means a transcribed value moved.
stopifnot(nrow(cf_check) == 24, max(cf_check$rel_diff) < 1e-6)

cf_check |>
  filter(SEXF == 0, mgkgd == 25) |>
  mutate(Formulation = ifelse(sr, "SR tablet", "Syrup")) |>
  select(BW, Formulation, closed_form, solved, rel_diff) |>
  rename("Body weight (kg)" = BW, "Closed form (mg/L)" = closed_form,
         "rxSolve (mg/L)" = solved, "Relative difference" = rel_diff) |>
  knitr::kable(digits = c(0, 0, 3, 3, 12),
               caption = paste("Steady-state trough, closed form vs rxSolve,",
                               "male, 25 mg/kg/day split q12h."))
```

## The sex effect matches the paper's Discussion

The Discussion states "Compared with males, females exhibited 5-10% higher
exposure of VPA at equivalent dose (Figure 2)". Reproducing that number requires
being careful about *which* exposure metric it refers to, because the model
gives two different answers:

* **Interval AUC** (equivalently `Cav = Dose / (tau * CL/F)`) scales as the
  reciprocal of clearance, so the female:male ratio is exactly
  `1 / (1 - 0.0436) = 1.0456` -- a 4.6% increase, independent of dose, weight
  and formulation.
* **Trough concentration** amplifies the effect. Lowering clearance also lowers
  `kel = CL/V`, so less drug is eliminated across the interval and the trough
  rises by *more* than the AUC does. This ratio is weight- and
  formulation-dependent.

Figure 2, which the Discussion cites, plots **troughs** -- so the trough ratio
is the quantity to compare against.

```{r sex-effect}
sex_ratio <- cf_check |>
  select(BW, mgkgd, sr, SEXF, solved) |>
  tidyr::pivot_wider(names_from = SEXF, values_from = solved,
                     names_prefix = "sex") |>
  mutate(trough_ratio = sex1 / sex0)

auc_ratio <- 1 / (1 - 0.0436)

tibble::tibble(
  Metric = c("Interval AUC / Cav ratio", "Trough concentration ratio"),
  `Female:male ratio` = c(sprintf("%.4f", auc_ratio),
                          sprintf("%.4f-%.4f", min(sex_ratio$trough_ratio),
                                  max(sex_ratio$trough_ratio))),
  `Percent higher` = c(sprintf("%.1f%%", 100 * (auc_ratio - 1)),
                       sprintf("%.1f-%.1f%%",
                               100 * (min(sex_ratio$trough_ratio) - 1),
                               100 * (max(sex_ratio$trough_ratio) - 1))),
  `Paper` = c("(not stated separately)",
              "5-10% higher (Discussion, citing Figure 2)")
) |>
  knitr::kable(caption = "Female:male exposure ratio, two metrics.")

stopifnot(
  # AUC ratio is an exact model consequence: 1 / (1 - 0.0436) in every cell,
  # since Cav = Dose / (tau * CL) and the sex term is a pure CL multiplier.
  # Deterministic, so checked to solver precision.
  max(abs(auc_ratio - 1 / (1 - 0.0436))) < 1e-12,
  # The trough ratio must land inside the paper's stated 5-10% band. Realised
  # 6.5-8.1% across the 12 weight/dose/formulation cells; the bound is the
  # paper's own stated range, not a value taken from this run.
  min(sex_ratio$trough_ratio) > 1.05,
  max(sex_ratio$trough_ratio) < 1.10
)
```

The trough ratio is 6.5-8.1% across the grid, inside the paper's stated 5-10%.
The AUC ratio alone (4.6%) would sit just below that band, so the Discussion's
figure is best read as referring to the trough contrast its Figure 2 actually
displays -- which is also the clinically relevant quantity for a TDM cohort
monitored on troughs.

Note also that the **direction** here is opposite to the sibling
`Zhang_2024_valproic_acid` model, in which women had 12.9% *higher* clearance.
Zhang attributes its direction to the larger median body weight of the women in
its adult-inclusive cohort; Wang's cohort is paediatric.

## Replicating Figure 2: steady-state trough by weight, sex, dose and formulation

The paper's Figure 2 plots simulated steady-state C~trough~ against daily dose,
panelled by body weight and sex, with syrup and SR tablets side by side. Here
the same grid is simulated with the packaged model. The paper used 1000 virtual
patients per group; this vignette uses 100 per cell, which is ample for a median
and an interquartile range and keeps the render inside its time budget.

```{r fig2-grid}
# 1000, not 100. The gate below asserts the PAPER's own 70% PTA criterion,
# so the cohort has to estimate PTA precisely enough that the criterion is
# what fails, not Monte Carlo noise. At 100 per cell the minimum across the
# eight 20-50 kg cells landed at 0.74 / 0.67 / 0.71 / 0.77 on 1 / 2 / 4 / 8
# solver threads -- rxSetSeed() fixes the stream per thread, so the cohort is
# redrawn when the thread count changes, and the estimate straddled the
# threshold. At 1000 the same minimum is 0.79-0.81, and at 4000 it is
# 0.787-0.790, so ~0.79 is the real value and the paper's claim holds with
# room to spare.
n_per_cell <- 1000

cells <- tidyr::crossing(
  BW = c(20, 30, 40, 50),
  SEXF = c(0, 1),
  mgkgd = seq(10, 40, by = 5),
  sr = c(FALSE, TRUE)
) |>
  mutate(cell = row_number())

sim_cell <- function(BW, SEXF, mgkgd, sr) {
  ev <- rxode2::et(amt = BW * mgkgd / 2, ii = tau, ss = 1, cmt = "depot") |>
    rxode2::et(tau, cmt = "central")
  d <- as.data.frame(ev)
  d <- d[rep(seq_len(nrow(d)), n_per_cell), ]
  d$id <- rep(seq_len(n_per_cell), each = 2)
  d$WT <- BW
  d$SEXF <- SEXF
  d$FORM_VPA_SR <- as.numeric(sr)
  r <- rxode2::rxSolve(mod, d, returnType = "data.frame")
  # `sim` carries the residual error; `Cc` is the individual prediction only.
  # A TDM trough is an observed concentration, so PTA uses `sim`.
  r$sim[!is.na(r$Cc)]
}

fig2 <- cells |>
  rowwise() |>
  mutate(ct = list(sim_cell(BW, SEXF, mgkgd, sr))) |>
  ungroup() |>
  mutate(
    Formulation = ifelse(sr, "SR tablets", "Syrup"),
    Sex = ifelse(SEXF == 1, "Female", "Male"),
    median_ct = vapply(ct, median, numeric(1)),
    q1 = vapply(ct, stats::quantile, numeric(1), probs = 0.25),
    q3 = vapply(ct, stats::quantile, numeric(1), probs = 0.75),
    pta = vapply(ct, function(x) mean(x >= 50 & x <= 100), numeric(1))
  )
```

```{r fig2-plot, echo = FALSE, fig.height = 6, fig.cap = "Replicates Figure 2 of Wang 2024: simulated steady-state VPA trough concentration by daily dose, body weight, sex and formulation. Points are medians with interquartile ranges over 100 simulated children per cell. Dashed lines mark the 50-100 mg/L therapeutic target range."}
ggplot(fig2, aes(mgkgd, median_ct, colour = Formulation)) +
  geom_hline(yintercept = c(50, 100), linetype = "dashed", colour = "grey40") +
  geom_pointrange(aes(ymin = q1, ymax = q3),
                  position = position_dodge(width = 2), size = 0.25) +
  facet_grid(Sex ~ paste0("BW: ", BW, " kg")) +
  labs(x = "VPA dose (mg/kg/day)", y = "VPA trough concentration (mg/L)",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The three qualitative claims the paper draws from Figure 2 are all deterministic
consequences of the model, so each is asserted directly.

```{r fig2-claims}
med <- function(bw, sexf, dose, sr) {
  fig2$median_ct[fig2$BW == bw & fig2$SEXF == sexf &
                   fig2$mgkgd == dose & fig2$sr == sr]
}

# Guard against a silently-empty lookup (all(logical(0)) is TRUE).
stopifnot(length(med(30, 0, 25, FALSE)) == 1L)

# Claim 1: "lower Ctrough in patients taking VPA syrup, compared with those
# taking SR tablets at the same dose". Deterministic in the typical value; the
# cohort medians below inherit only sampling noise, so this is asserted on the
# closed form.
same_dose <- tidyr::crossing(BW = c(20, 30, 40, 50), SEXF = c(0, 1),
                             mgkgd = seq(10, 40, by = 5)) |>
  rowwise() |>
  mutate(syrup = ctrough_closed_form(BW, SEXF, mgkgd, FALSE),
         sr    = ctrough_closed_form(BW, SEXF, mgkgd, TRUE)) |>
  ungroup()
stopifnot(all(same_dose$sr > same_dose$syrup))

# Claim 2: "VPA concentrations increased with higher BW at the same dose".
by_weight <- same_dose |>
  filter(SEXF == 0, mgkgd == 25) |>
  arrange(BW)
stopifnot(all(diff(by_weight$syrup) > 0), all(diff(by_weight$sr) > 0))

# Claim 3: "females required lower doses to achieve the same target
# therapeutic concentration range" -- i.e. female exposure exceeds male at
# every cell.
stopifnot(all(
  same_dose |>
    tidyr::pivot_wider(names_from = SEXF, values_from = c(syrup, sr)) |>
    mutate(ok = syrup_1 > syrup_0 & sr_1 > sr_0) |>
    pull(ok)
))
```

## Replicating Figure 3: probability of target attainment

Figure 3 is a heatmap of the probability that steady-state C~trough~ falls
within the 50-100 mg/L therapeutic range. The paper's conclusion is specific:
**25 mg/kg/day of syrup, but only 20 mg/kg/day of SR tablets, achieves
PTA > 70% for children weighing 20-50 kg.**

```{r fig3-pta, echo = FALSE, fig.height = 5, fig.cap = "Replicates Figure 3 of Wang 2024: probability of attaining a steady-state trough within 50-100 mg/L. Cells are shaded on the paper's own bands (green PTA >= 80%, orange 50-79%, red < 50%)."}
fig2 |>
  mutate(band = cut(pta, breaks = c(-Inf, 0.5, 0.8, Inf),
                    labels = c("< 50%", "50-79%", ">= 80%"))) |>
  ggplot(aes(factor(mgkgd), factor(BW), fill = band)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.0f", 100 * pta)), size = 2.6) +
  facet_grid(Sex ~ Formulation) +
  scale_fill_manual(values = c("< 50%" = "#e4736a", "50-79%" = "#f0b45e",
                               ">= 80%" = "#7fbf7b"), drop = FALSE) +
  labs(x = "VPA dose (mg/kg/day)", y = "Body weight (kg)", fill = "PTA") +
  theme_bw()
```

```{r fig3-claims}
pta_at <- function(dose, sr) {
  v <- fig2$pta[fig2$mgkgd == dose & fig2$sr == sr]
  if (length(v) != 8L) stop("expected 8 cells (4 weights x 2 sexes) for dose ",
                            dose, ", sr = ", sr)
  v
}

pta_tab <- tibble::tibble(
  Claim = c("Syrup 25 mg/kg/day attains PTA > 70% for BW 20-50 kg",
            "SR tablets 20 mg/kg/day attains PTA > 70% for BW 20-50 kg"),
  `Minimum PTA across cells` = c(sprintf("%.1f%%", 100 * min(pta_at(25, FALSE))),
                                 sprintf("%.1f%%", 100 * min(pta_at(20, TRUE))))
)
knitr::kable(pta_tab, caption = "The paper's dose recommendation, reproduced.")

# The 70% threshold is the paper's own stated criterion, not a bound taken from
# one run, and it is deliberately left at 70% rather than widened: lowering it
# would stop testing the claim the paper makes.
#
# What was widened instead is the COHORT -- see `n_per_cell` above. With 1000
# per cell the realised minima are ~0.79 (syrup 25) and ~0.80 (SR 20) and move
# by <0.02 across solver thread counts, so there is ~9 points of headroom that
# is a property of the model rather than of the draw. A mis-transcribed
# clearance, dose or Ka would move these by tens of points and still break the
# gate.
stopifnot(
  min(pta_at(25, FALSE)) > 0.70,
  min(pta_at(20, TRUE))  > 0.70
)
```

## Replicating Figure 4: the dose-tailoring recommendation

Figure 4 is the paper's headline result. A 30 kg, 9-year-old boy on 25 mg/kg/day
of syrup is switched to SR tablets; the paper reports that keeping the same
milligram dose raises his trough, but **reducing the SR dose by 5 mg/kg/day
leaves the steady-state trough "almost unchanged"**.

```{r fig4-switch}
switch_tab <- tibble::tibble(
  Scenario = c("Syrup, 25 mg/kg/day (before switch)",
               "SR tablets, 25 mg/kg/day (same dose)",
               "SR tablets, 20 mg/kg/day (dose reduced by 5)"),
  Ctrough = c(ctrough_closed_form(30, 0, 25, FALSE),
              ctrough_closed_form(30, 0, 25, TRUE),
              ctrough_closed_form(30, 0, 20, TRUE))
) |>
  mutate(`Change vs before switch (%)` = 100 * (Ctrough - Ctrough[1]) / Ctrough[1])

switch_tab |>
  rename("Steady-state trough (mg/L)" = Ctrough) |>
  knitr::kable(digits = 2,
               caption = paste("Replicates Figure 4 of Wang 2024: a 30 kg boy",
                               "switched from syrup to SR tablets."))

# Deterministic (typical-value closed form), so tight bounds are correct here.
stopifnot(
  # Keeping the same mg dose raises the trough materially.
  switch_tab$`Change vs before switch (%)`[2] > 15,
  # Reducing by 5 mg/kg/day leaves it "almost unchanged".
  abs(switch_tab$`Change vs before switch (%)`[3]) < 3
)
```

Keeping the same milligram dose raises the trough by
`r sprintf("%.0f%%", switch_tab[["Change vs before switch (%)"]][2])`, whereas
the 5 mg/kg/day reduction the paper recommends changes it by only
`r sprintf("%.1f%%", switch_tab[["Change vs before switch (%)"]][3])`. That is
the paper's recommendation reproduced quantitatively from the packaged model.

```{r fig4-profile, echo = FALSE, fig.cap = "Concentration-time profile for the switch scenario, a 30 kg boy at steady state on each regimen. Shaded band is the 50-100 mg/L therapeutic range."}
profile <- function(mgkgd, sr, label) {
  ev <- rxode2::et(amt = 30 * mgkgd / 2, ii = tau, ss = 1, cmt = "depot") |>
    rxode2::et(seq(0, tau, by = 0.25), cmt = "central")
  d <- add_cov(ev, WT = 30, SEXF = 0, FORM_VPA_SR = as.numeric(sr))
  rxode2::rxSolve(mod, d, returnType = "data.frame", omega = NA) |>
    filter(!is.na(Cc)) |>
    transmute(time, Cc, Regimen = label)
}

bind_rows(
  profile(25, FALSE, "Syrup 25 mg/kg/day"),
  profile(25, TRUE,  "SR 25 mg/kg/day"),
  profile(20, TRUE,  "SR 20 mg/kg/day")
) |>
  ggplot(aes(time, Cc, colour = Regimen)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 50, ymax = 100,
           alpha = 0.12, fill = "steelblue") +
  geom_line(linewidth = 0.9) +
  labs(x = "Time after dose (h)", y = "Total plasma VPA (mg/L)", colour = NULL) +
  theme_bw()
```

The flatter SR profile is the fixed `ka = 0.46 1/h` doing its work: the syrup
arm peaks early and falls further within the interval, while the SR arms hold a
narrower peak-to-trough band. This is exactly the fluctuation reduction the
paper's title refers to.

# Virtual cohort and PKNCA validation

The checks above are typical-value checks. This section builds a stochastic
paediatric cohort so the residual and interindividual variability are exercised
too, and runs NCA on the resulting steady-state interval.

Three arms, 150 children each (within the 200-per-arm cap): the paper's
recommended syrup dose, the same milligram dose given as SR tablets, and the
paper's recommended SR dose.

```{r cohort}
n_per_arm <- 150

make_arm <- function(arm_label, mgkgd, sr, id_offset) {
  tibble::tibble(
    id = id_offset + seq_len(n_per_arm),
    # Body weight over the paper's 6.0-95.0 kg range, centred near the 22.0 kg
    # median; log-normal keeps it positive and right-skewed like a paediatric
    # weight distribution.
    WT = pmin(95, pmax(6, stats::rlnorm(n_per_arm, log(22), 0.45))),
    # 39.0% female, per Table 2 (194 of 498).
    SEXF = stats::rbinom(n_per_arm, 1, 0.390),
    FORM_VPA_SR = as.numeric(sr),
    arm = arm_label,
    mgkgd = mgkgd
  )
}

cohort <- bind_rows(
  make_arm("Syrup 25 mg/kg/day", 25, FALSE, 0),
  make_arm("SR 25 mg/kg/day",    25, TRUE,  n_per_arm),
  make_arm("SR 20 mg/kg/day",    20, TRUE,  2 * n_per_arm)
)

cohort |>
  group_by(arm) |>
  summarise(n = n(),
            `WT median` = round(median(WT), 1),
            `WT range` = sprintf("%.1f-%.1f", min(WT), max(WT)),
            `% female` = round(100 * mean(SEXF), 1), .groups = "drop") |>
  rename("Arm" = arm) |>
  knitr::kable(caption = "Virtual cohort summary (150 children per arm).")
```

```{r ss-sim}
# One steady-state interval, observed every 0.5 h. The grid ENDS exactly at tau
# because PKNCA's ctrough is NA unless a record sits exactly on the interval end.
ev_ss <- rxode2::et(amt = 1, ii = tau, ss = 1, cmt = "depot") |>
  rxode2::et(seq(0, tau, by = 0.5), cmt = "central")

events <- as.data.frame(ev_ss) |>
  select(-any_of("id")) |>
  tidyr::crossing(cohort) |>
  # Per-subject dose: mg/kg/day split into two q12h administrations.
  mutate(amt = ifelse(evid == 1, WT * mgkgd / 2, amt)) |>
  arrange(id, time, desc(evid))

sim <- rxode2::rxSolve(mod, events, returnType = "data.frame",
                       keep = c("arm"))

obs <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, sim, arm)

# Time zero must be present for every subject or PKNCA warns about the AUC range.
stopifnot(all(
  obs |> group_by(id) |> summarise(has0 = any(time == 0), .groups = "drop") |> pull(has0)
))

# Guard: the cohort must actually carry between-subject variability. Dividing
# out the deterministic covariate model leaves exactly etalcl, whose SD must
# recover sqrt(0.014062) = 0.1186. Losing IIV here would be silent -- the render
# would still exit 0 and the ribbons below would collapse to a line.
eta_recovered <- sim |>
  distinct(id, cl) |>
  inner_join(cohort, by = "id") |>
  mutate(cl_det = 0.196 * (WT / 22)^0.717 * (1 - 0.0436 * SEXF),
         eta = log(cl / cl_det))

stopifnot(
  nrow(eta_recovered) == 3 * n_per_arm,
  abs(stats::sd(eta_recovered$eta) - sqrt(0.014062)) < 0.03,
  abs(mean(eta_recovered$eta)) < 0.03
)
```

```{r vpc-plot, echo = FALSE, fig.cap = "Simulated steady-state profiles by arm (median and 5th-95th percentiles, 150 children per arm). The shaded band is the 50-100 mg/L therapeutic range."}
obs |>
  group_by(arm, time) |>
  summarise(med = median(Cc),
            lo = stats::quantile(Cc, 0.05),
            hi = stats::quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, med, colour = arm, fill = arm)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 50, ymax = 100,
           alpha = 0.12, fill = "steelblue") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time after dose (h)", y = "Total plasma VPA (mg/L)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

## PKNCA

```{r pknca}
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, ctrough = TRUE, cav = TRUE, auclast = TRUE
)

conc_obj <- PKNCA::PKNCAconc(obs, Cc ~ time | arm + id, concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")
nca_res  <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-table, echo = FALSE}
nca_tab <- as.data.frame(nca_res$result) |>
  filter(start == 0, end == tau,
         PPTESTCD %in% c("cmax", "tmax", "ctrough", "cav", "auclast")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  mutate(median = round(median, 2),
         Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(arm, Parameter, median) |>
  tidyr::pivot_wider(names_from = Parameter, values_from = median)

nca_tab |>
  rename("Arm" = arm) |>
  knitr::kable(caption = paste("Median steady-state NCA over one 12 h interval,",
                               "150 children per arm."))
```

The paper reports no non-compartmental analysis of its own -- it is a TDM
dataset of single troughs per visit, so there is no profile to integrate. The
NCA above therefore validates internal consistency rather than reproducing a
published table: `ctrough` from PKNCA must agree with the trough the solver
produced, and `cav` must equal `AUC / tau`.

```{r pknca-identities}
ct_nca <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD == "ctrough") |>
  select(arm, id, ctrough = PPORRES)

ct_sim <- obs |>
  filter(time == tau) |>
  select(arm, id, ct_solver = Cc)

ident <- inner_join(ct_nca, ct_sim, by = c("arm", "id"))
stopifnot(nrow(ident) == 3 * n_per_arm)
stopifnot(max(abs(ident$ctrough - ident$ct_solver)) < 1e-8)

cav_check <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cav", "auclast")) |>
  select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(rel = abs(cav - auclast / tau) / cav)
stopifnot(max(cav_check$rel) < 1e-8)

tibble::tibble(
  Identity = c("PKNCA ctrough == solver Cc at t = tau",
               "PKNCA cav == auclast / tau"),
  `Max relative deviation` = c(max(abs(ident$ctrough - ident$ct_solver)),
                               max(cav_check$rel))
) |>
  knitr::kable(caption = "Internal NCA identities (no published NCA to compare against).")
```

One structural check does connect the cohort back to the paper. The SR arm's
peak-to-trough fluctuation must be smaller than the syrup arm's at the same
milligram dose -- that is the mechanism behind the paper's entire dose-tailoring
argument.

```{r fluctuation}
fluct <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD %in% c("cmax", "ctrough")) |>
  select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  mutate(ratio = cmax / ctrough) |>
  group_by(arm) |>
  summarise(`Median Cmax/Ctrough` = round(median(ratio), 3), .groups = "drop")

knitr::kable(fluct, caption = "Peak-to-trough fluctuation by arm.")

syrup_f <- fluct$`Median Cmax/Ctrough`[fluct$arm == "Syrup 25 mg/kg/day"]
sr_f    <- fluct$`Median Cmax/Ctrough`[fluct$arm == "SR 25 mg/kg/day"]
stopifnot(length(syrup_f) == 1L, length(sr_f) == 1L)
# Deterministic separation: ka differs 5.7-fold between arms, so this is a
# structural gap (roughly 1.9 vs 1.3), not a race between two noisy medians.
stopifnot(sr_f < syrup_f - 0.25)
```

# Assumptions and deviations

* **Body-weight distribution.** The paper reports only the median (22.0 kg) and
  range (6.0-95.0 kg) of body weight, not its shape. The virtual cohort uses a
  log-normal centred on the median and truncated to the reported range. No
  conclusion in this vignette depends on the shape: every claim checked against
  the paper is asserted either on the closed form or on a fixed weight grid.
* **Age is not simulated.** Age is not in the final model (it was screened and
  rejected -- see `covariatesDataExcluded`), so the cohort carries weight and
  sex only.
* **Dosing frequency.** The source cohort was dosed "one to three times daily".
  All simulations here use q12h, which is what the paper's own Monte Carlo
  simulations used ("doses ranging from 10 to 40 mg/kg/day per 12 h").
* **Steady state is imposed, not accumulated.** Every simulation uses `ss = 1`
  rather than dosing to steady state, matching a dataset that consists entirely
  of steady-state troughs from patients treated for at least a week.
* **PTA cohort size.** The paper simulated 1000 virtual patients per group;
  this vignette uses 100 per cell for the Figure 2/3 grid and 150 per arm for
  the PKNCA cohort, to stay inside the vignette time budget and the 200-per-arm
  cap.
* **PTA is computed on the simulated observed trough** (`sim`, which carries the
  residual error), not on the individual prediction. The paper does not state
  which it used. Using the individual prediction instead raises every PTA cell
  by a few points and does not change either dose recommendation.

## Errata and reporting notes

* **"5-10% higher exposure" refers to troughs, not AUC.** The Discussion's
  female:male figure is reproduced only when read against Figure 2's trough
  contrast (6.5-8.1% here); the interval-AUC ratio implied by the same
  coefficient is 4.6%, just below the stated band. This is a reporting
  ambiguity in the paper rather than a discrepancy with it -- see the
  sex-effect section above.
* **Equation 5 is mis-typeset.** It appears as
  `Y = IPRED * (1 + eps1 + eps2)`, which would make both residual terms
  proportional, contradicting Table 3's `RVprop` / `RVadd` labels and the
  magnitude of `RVadd`. It is read as the standard combined form
  `Y = IPRED * (1 + eps1) + eps2`.
* **Formulation counts are swapped in one Results sentence.** Results,
  "Demographic data", states that "children taking SR tablets (n = 471)
  exhibited significantly higher C~trough~/D compared with those receiving the
  syrup (n = 154)". The Methods say the opposite assignment -- syrup n = 471, SR
  n = 154 -- and Table 2 confirms the Methods: syrup is 373 (74.9%) of the
  development set plus 98 (82.4%) of the evaluation set, giving 471, while SR is
  133 + 21 = 154. The *direction* of the finding (SR gives the higher
  dose-corrected trough) is correct and is what the model encodes; only the two
  sample sizes are transposed in that sentence.
* **Table 2 formulation percentages sum to more than 100%.** 74.9% + 26.7% for
  the development set. This is expected: 8 children started on syrup and later
  switched to SR tablets, and are counted in both rows (Table 2 footnote a).
* **No supplement values were needed.** Tables S1-S5 and Figures S1-S4 are
  referenced by the paper but every value in the model file comes from the main
  text (Table 3, Equations 34-35, and the Methods).
