---
title: "Magnesium bone implant PBPK (Ward 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Magnesium bone implant PBPK (Ward 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Ward_2026_magnesium_pbpk"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Ward_2026_magnesium_pbpk"))$description`
* Article: <https://doi.org/10.1021/acsomega.5c06910> (ACS Omega 2026;11(4):5144-5153, PMC12878740, open access)
* Supporting Information A-C: `ao5c06910_si_001.pdf`, distributed with the article.

This is not a drug-disposition model. It is a deterministic, physiologically
based description of whole-body magnesium (Mg(II)) homeostasis in an average
healthy adult, perturbed by the slow corrosion of a biodegradable magnesium
bone implant. There is no dose, no absorption phase and no residual error, so
the validation below uses the endogenous / mechanistic patterns (steady-state
hold, closed-form steady states, characteristic time scales, perturbation
response) rather than non-compartmental analysis.

# Population

The model is not fitted to an individual-level data set. Its parameters
describe an "average" healthy 70 kg adult carrying about 1000.7 mmol of total
body Mg, assembled from the physiology literature in Supporting Information B
(Tables S2-S3): serum 3 L, red blood cells 2 L, bone 12.3 L and bulk soft
tissue (muscle, soft tissue and liver, lumped) 52.7 L. The exchange clearances
between blood and bone and between blood and tissue were then tuned so the
model reproduces the stable-isotope Mg tracer kinetics of Sojka 1997 and
Sabatier 2003 (Supporting Information B, Figure S2). The authors describe the
resulting values as "representative rather than definitive".

Reduced renal function is represented by lowering the urinary excretion
clearance `cl_renal` (the paper's `gamma`), with a matched reduction in
`rate_diet` so that pre-implant homeostasis `Cse = rate_diet / cl_renal` is
preserved; chronic renal failure is taken as roughly one third of the healthy
value (Section 2.2.3).

No long-term human validation data exist: the paper notes that no human study
has monitored serum Mg over the 2-12 month window in which a Mg implant
corrodes.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Ward_2026_magnesium_pbpk")()$population` --
`readModelDb()` returns the model *function*, so the trailing `()` evaluates it
to the model object).

```{r model}
mod <- rxode2::rxode2(readModelDb("Ward_2026_magnesium_pbpk"))

# Every published scenario is a change to one or more of the fixed ini()
# values, so scenarios are built by re-pinning them rather than by dosing.
solve_scenario <- function(times,
                           n_implant      = 1,
                           f_diet         = 1,
                           rate_diet      = 6,
                           cl_renal       = 10.9,
                           rate_implant   = 0.05,
                           v_implant_zone = 0.00527) {
  # Bind to distinct local names first: rxode2::ini() is non-standard-evaluating,
  # so `ini(n_implant = n_implant)` would resolve the right-hand side against the
  # model's own parameter rather than this function's argument.
  .n    <- n_implant
  .rho  <- f_diet
  .diet <- rate_diet
  .lgam <- log(cl_renal)
  .sig  <- rate_implant
  .vi   <- v_implant_zone
  m <- suppressMessages(
    mod |>
      rxode2::ini(n_implant      = .n) |>
      rxode2::ini(f_diet         = .rho) |>
      rxode2::ini(rate_diet      = .diet) |>
      rxode2::ini(lcl_renal      = .lgam) |>
      rxode2::ini(rate_implant   = .sig) |>
      rxode2::ini(v_implant_zone = .vi)
  )
  as.data.frame(rxode2::rxSolve(m, rxode2::et(times), atol = 1e-12, rtol = 1e-10))
}

# Log-spaced grid for the paper's log-x figures; fine linear grid for the
# time-constant checks.
t_log <- c(0, 10^seq(-3, 4, length.out = 500))
```

# Source trace

Every `ini()` value carries an in-file comment pointing at its source location
in `inst/modeldb/endogenous/Ward_2026_magnesium_pbpk.R`. They are collected
here for review. All parameter values come from **Table 1** (page 5145); the
"source" column of that table records whether each value was read from the
literature, derived from the Supporting Information B homeostasis ratios (H),
derived from the Table S2/S3 data (R), or is an estimate (I, V).

| File parameter | Paper symbol | Value | Units | Source location |
|----------------|--------------|-------|-------|-----------------|
| `v_serum` | `V_s` | 3 | L | Table 1 |
| `v_rbc` | `V_r` | 2 | L | Table 1 |
| `v_bone` | `V_N` | 12.3 | L | Table 1 |
| `v_tissue` | `V_Ttot` | 52.7 | L | Table 1 |
| `v_implant_zone` | `V_I` | 0.00527 | L | Table 1; chosen so `V_I / V_Ttot = 1e-4` (SI B, S15) |
| `lcl_renal` | `gamma` | log(10.9) | log(L/day) | Table 1; `= rate_diet / Cse` (SI B, S11) |
| `q_serum_bone` | `mu_1` | 6.05 | L/day | Table 1 |
| `q_bone_serum` | `mu_-1` | 0.0775 | L/day | Table 1; fixed by `mu_-1 / mu_1 = Cse / CNe` (SI B, S12) |
| `q_serum_tissue` | `k_1` | 138 | L/day | Table 1 |
| `q_tissue_serum` | `k_-1` | 34.7 | L/day | Table 1; fixed by `k_-1 / k_1 = Cse / CTe` (SI B, S12) |
| `rate_diet` | `phi_D` | 6 | mmol/day | Table 1 |
| `rate_implant` | `sigma` | 0.05 | mmol/day | Table 1; one 3.2 x 32 mm screw |
| `f_exposed_bone` | `phi` | 0.3 | - | Table 1 |
| `kp_unexposed_serum` | `xi_1` | 0.538 | - | Table 1 |
| `kp_rbc_serum` | `xi_2` | 0.452 | - | Table 1 |
| `kp_unexposed_tissue` | `xi_3` | 3.00 | - | Table 1 |
| `n_implant` | `n` | 1 | - | eq 9, Section 2.2.2 (scenario lever) |
| `f_diet` | `rho` | 1 | - | eq 10, Section 2.2.3 (scenario lever) |

| Model structure | Source location |
|-----------------|-----------------|
| `d/dt(serum)` | eq 1 (page 5146), with the `rho` dietary-control factor of eq 10 (page 5149) |
| `d/dt(bone)` | eq 2 (page 5146) |
| `d/dt(other)` | eq 3 (page 5146) |
| `d/dt(implant_zone)` | eq 4 (page 5146) |
| Initial conditions `Cse`, `CNe`, `CTe` | eq 5 (page 5146) |
| `n` implant scaling of `V_I`, `sigma`, `xi` | eq 9 (page 5147) |
| Total serum `Cc = serum * (1 + xi_1)` | Table 1, `C_hom = Cse + Csu = 0.85 mmol/L` |

## Compartment mapping

| Paper compartment | File compartment | Note |
|-------------------|------------------|------|
| Blood serum, `C_s` | `serum` | exposed (exchangeable) Mg only |
| Bone, `C_N` | `bone` | |
| Tissue, `C_T` | `other` | muscle + soft tissue + liver, lumped |
| Implant region, `C_I` | `implant_zone` | localized peri-implant tissue |

The paper's bulk tissue store `T` maps onto the nlmixr2lib canonical
lumped-remainder compartment `other` rather than a compartment literally named
`tissue`: the role is defined by the lumping, and the paper never resolves
muscle from liver from generic soft tissue. `implant_zone` is a new canonical
registered with this extraction (see `inst/references/compartment-names.md`).

## Units and dimensional analysis

Unlike most nlmixr2lib models, the states here are **concentrations**
(mmol/L), not amounts: the published system is written directly in
concentration, with the compartment volume and its exposed/unexposed
capacitance factor appearing as the divisor on the left-hand side. Every
"exchange rate constant" in Table 1 has units of L/day, i.e. it is a
volumetric clearance, not a first-order rate constant.

| ODE | Numerator units | Divisor units | Result |
|-----|-----------------|---------------|--------|
| `serum` | mmol/day + (L/day)(mmol/L) = mmol/day | `v_serum*(1+xi_1) + v_rbc*xi_2` = L | mmol/L/day |
| `bone` | (L/day)(mmol/L) = mmol/day | `phi*v_bone` = L | mmol/L/day |
| `other` | (L/day)(mmol/L) = mmol/day | `v_tissue*(1+xi_3)` = L | mmol/L/day |
| `implant_zone` | mmol/day + (L/day)(mmol/L) = mmol/day | `v_iz*(1+xi_3)` = L | mmol/L/day |

`xi = V_I / V_Ttot` and the four `xi_j` partition ratios are dimensionless, as
are `n_implant` and `f_diet`. The observation `Cc = serum * (1 + xi_1)` is
total (exposed + unexposed) serum Mg in mmol/L -- the clinically measured
quantity that the 0.65-1.05 mmol/L reference range and the 1.05 mmol/L
hypermagnesemia threshold refer to.

```{r reference-values}
# Reference quantities used throughout, all from Table 1.
phiD  <- 6       # dietary intake rate, mmol/day
gam0  <- 10.9    # healthy urinary excretion clearance, L/day
k1    <- 138     # serum -> tissue exchange clearance, L/day
kn1   <- 34.7    # tissue -> serum
mu1   <- 6.05    # serum -> bone
mun1  <- 0.0775  # bone -> serum
VT    <- 52.7; VN <- 12.3; Vs <- 3; Vr <- 2
xi1   <- 0.538; xi2 <- 0.452; xi3 <- 3; phi_b <- 0.3
xi0   <- 0.00527 / VT      # V_I / V_Ttot for a single implant

Chom  <- 0.85              # Table 1, homeostatic total serum Mg (mmol/L)
Chyp  <- 1.05              # Table 1, hypermagnesemia threshold
Csev  <- 2.9               # Table 1, severe / "dangerous" concentration
Chyp_norm <- Chyp / Chom   # Table 2 footnote: C*_hyp = C_hyp / C_hom
Csev_norm <- Csev / Chom
c(Chyp_norm = Chyp_norm, Csev_norm = Csev_norm)  # paper: ~1.24 and ~3.41
```

# Steady-state hold (no implant)

With no implant the system must sit at the homeostasis of eq 5 forever. The
initial conditions are computed inside `model()` from `rate_diet / cl_renal`
(eq 5) rather than from Table 1's rounded printed `Cse = 0.553`, so this check
is exact rather than merely close: `gamma = 10.9` is itself the rounded form of
`phiD / Cse`, and seeding from the printed `Cse` would leave the model 0.4% off
its own steady state.

```{r steady-state}
ss <- solve_scenario(times = t_log, n_implant = 0)

drift <- max(abs(c(ss$Cs_norm, ss$CN_norm, ss$CT_norm, ss$CI_norm) - 1))
drift

# Exact to machine precision, over 10^4 days.
stopifnot(drift < 1e-10)
```

# Closed-form steady states (eq 8)

Equation 8 gives the exact long-term normalised steady states of eqs 1-4:

```
(Cs*, CN*, CT*, CI*) = (1 + sigma/phiD,  1 + sigma/phiD,  1 + sigma/phiD,
                        1 + sigma/phiD + sigma*gamma/(xi*k1*phiD))
```

so serum, bone and tissue all rise by the same proportion `sigma/phiD`, and
only the implant zone -- which absorbs the whole release term into a tiny
volume -- runs higher. This is the model's sharpest structural check: it pins
`rate_implant`, `rate_diet`, `lcl_renal`, `q_serum_tissue` and the `n` scaling
of eq 9 simultaneously.

```{r eq8}
eq8 <- function(n) {
  sig <- n * 0.05
  xi  <- n * xi0
  c(Cs = 1 + sig / phiD,
    CI = 1 + sig / phiD + sig * gam0 / (xi * k1 * phiD))
}

n_grid <- c(1, 2, 5, 10, 20, 30, 40)
eq8_chk <- lapply(n_grid, function(n) {
  s <- solve_scenario(times = t_log, n_implant = n)
  p <- eq8(n)
  tibble(
    n          = n,
    Cs_sim     = tail(s$Cs_norm, 1),  Cs_eq8 = unname(p["Cs"]),
    CN_sim     = tail(s$CN_norm, 1),
    CT_sim     = tail(s$CT_norm, 1),
    CI_sim     = tail(s$CI_norm, 1),  CI_eq8 = unname(p["CI"])
  )
}) |> bind_rows()

eq8_chk |>
  mutate(across(-n, ~ signif(.x, 6))) |>
  dplyr::rename(
    "n implants" = n,
    "Cs* (sim)" = Cs_sim, "Cs* (eq 8)" = Cs_eq8,
    "CN* (sim)" = CN_sim, "CT* (sim)" = CT_sim,
    "CI* (sim)" = CI_sim, "CI* (eq 8)" = CI_eq8
  ) |>
  knitr::kable(caption = "Simulated long-term steady states vs the closed form of eq 8.")

# Simulation and closed form agree to solver tolerance for every n, and the
# paper's statement that Cs*, CN* and CT* are indistinguishable holds exactly.
stopifnot(
  max(abs(eq8_chk$Cs_sim - eq8_chk$Cs_eq8)) < 1e-6,
  max(abs(eq8_chk$CI_sim - eq8_chk$CI_eq8)) < 1e-4,
  max(abs(eq8_chk$Cs_sim - eq8_chk$CN_sim)) < 1e-6,
  max(abs(eq8_chk$Cs_sim - eq8_chk$CT_sim)) < 1e-6
)
```

For a single screw this gives `Cs* = 1.00833` -- the paper's "the rest of the
compartments eventually peaks at around 0.8% above homeostatic levels" -- and
`CI* = 7.59`, matching Figure 2.

# Figure 2 -- a single Mg screw implant

```{r figure-2}
# Replicates Figure 2 of Ward 2026: normalised concentrations over time for a
# single 3.2 x 32 mm screw releasing sigma = 0.05 mmol/day.
TMg_formula <- (1 / gam0) *
  ((1 + xi1) * Vs + xi2 * Vr + (mu1 / mun1) * phi_b * VN +
     (1 + xi3) * (k1 / kn1) * VT)

f2 <- solve_scenario(times = t_log, n_implant = 1) |>
  select(time, Cs_norm, CN_norm, CT_norm, CI_norm) |>
  pivot_longer(-time, names_to = "state", values_to = "value") |>
  mutate(state = recode(state,
                        Cs_norm = "Cs* (serum)", CN_norm = "CN* (bone)",
                        CT_norm = "CT* (tissue)", CI_norm = "CI* (implant zone)")) |>
  filter(time > 0)

ggplot(f2, aes(time, value, colour = state)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = Chyp_norm, linetype = "dashed") +
  geom_hline(yintercept = Csev_norm, linetype = "dashed", colour = "red") +
  geom_vline(xintercept = TMg_formula, linetype = "dotted") +
  scale_x_log10(limits = c(1e-3, 1e4)) +
  labs(x = "Time (days)", y = "Normalised concentration (Cj*)", colour = NULL,
       title = "Figure 2 - single Mg screw implant",
       caption = paste("Replicates Figure 2 of Ward 2026. Dashed lines: C*hyp and C*sev.",
                       "Dotted vertical: t = T_Mg."))
```

The implant zone reaches its elevated plateau of about 7.6 within roughly three
weeks and stays there, above both the hypermagnesemia (1.24) and the
"dangerous" (3.41) normalised thresholds, while serum, bone and tissue overlap
each other and rise by less than 1%. That is the paper's central conclusion for
a single small implant: a large local effect and a negligible systemic one.

# Figure 3 -- multiple or larger implants

Equation 9 scales `V_I`, `sigma` and `xi` together with the number of implants
`n`, on the assumption that the affected tissue regions do not overlap.

```{r figure-3}
# Replicates Figure 3 of Ward 2026 (A: serum; B: implant zone).
f3 <- lapply(n_grid, function(n) {
  solve_scenario(times = t_log, n_implant = n) |>
    transmute(time, n = factor(n, levels = n_grid), Cs_norm, CI_norm)
}) |> bind_rows() |> filter(time > 0)

ggplot(filter(f3, time >= 1e-2), aes(time, Cs_norm, colour = n)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = Chyp_norm, linetype = "dashed") +
  geom_vline(xintercept = TMg_formula, linetype = "dotted") +
  scale_x_log10(limits = c(1e-2, 1e4)) +
  labs(x = "Time (days)", y = "Cs* (normalised serum)", colour = "n implants",
       title = "Figure 3A - serum Mg for n implants",
       caption = "Replicates Figure 3A of Ward 2026. Dashed: C*hyp ~= 1.24.")

ggplot(f3, aes(time, CI_norm, colour = n)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = Chyp_norm, linetype = "dashed") +
  geom_hline(yintercept = Csev_norm, linetype = "dashed", colour = "red") +
  scale_x_log10(limits = c(1e-3, 1e4)) +
  labs(x = "Time (days)", y = "CI* (normalised implant zone)", colour = "n implants",
       title = "Figure 3B - implant-zone Mg for n implants",
       caption = "Replicates Figure 3B of Ward 2026.")
```

Figure 3B reproduces the paper's observation that the implant-zone curves are
essentially independent of `n` for the first 2-3 days: eq 9 scales the release
rate and the zone volume by the same factor, so the local concentration is
initially unaffected by how many implants there are.

# Characteristic time scales (Table 2)

Table 2 gives closed-form approximations for the two time constants, defined as
the time at which the change has completed a fraction `1 - exp(-1) = 0.632`:

```
T_1  = V_T (1 + xi_3) / k_-1                                     ~= 6 days
T_Mg = (1/gamma) [ (1 + xi_1) V_s + xi_2 V_r
                   + (mu_1/mu_-1) phi V_N + (1 + xi_3)(k_1/k_-1) V_Ttot ]  ~= 104 days
```

```{r time-scales}
t_fine <- seq(0, 1500, length.out = 150001)
s1     <- solve_scenario(times = t_fine, n_implant = 1)

# Time at which a monotone rise has completed 1 - exp(-1) of its total change.
t_632 <- function(x, tt) {
  target <- 1 + (tail(x, 1) - 1) * (1 - exp(-1))
  tt[which(x >= target)[1]]
}

T1_formula <- VT * (1 + xi3) / kn1
tab2 <- tibble(
  Quantity  = c("T_1 (implant zone)", "T_Mg (systemic)"),
  Simulated = c(t_632(s1$CI_norm, t_fine), t_632(s1$Cs_norm, t_fine)),
  Formula   = c(T1_formula, TMg_formula),
  Paper     = c(6, 104)
)
tab2 |>
  mutate(across(where(is.numeric), ~ signif(.x, 4))) |>
  dplyr::rename("Table 2 formula" = Formula, "Paper (rounded)" = Paper) |>
  knitr::kable(caption = "Simulated time constants vs the Table 2 closed forms.")

# Both closed forms are asymptotic approximations in xi << 1, so a few percent
# of disagreement with the full four-ODE solve is expected and is the quantity
# actually being validated.
stopifnot(
  abs(tab2$Simulated[1] - tab2$Formula[1]) / tab2$Formula[1] < 0.05,
  abs(tab2$Simulated[2] - tab2$Formula[2]) / tab2$Formula[2] < 0.05,
  abs(tab2$Formula[1] - 6)   < 0.5,
  abs(tab2$Formula[2] - 104) < 2
)
```

The paper attributes "around 74%" of the systemic buildup time to tissue and
"about 25%" to bone; the four additive terms of `T_Mg` reproduce that split
directly.

```{r tmg-split}
TMg_terms <- c(
  serum  = (1 + xi1) * Vs,
  rbc    = xi2 * Vr,
  bone   = (mu1 / mun1) * phi_b * VN,
  tissue = (1 + xi3) * (k1 / kn1) * VT
)
round(100 * TMg_terms / sum(TMg_terms), 1)  # paper: tissue ~74%, bone ~25%

stopifnot(
  abs(100 * TMg_terms[["tissue"]] / sum(TMg_terms) - 74) < 1.5,
  abs(100 * TMg_terms[["bone"]]   / sum(TMg_terms) - 25) < 1.5
)
```

# Hypermagnesemia threshold -- a binary answer key

Table 2 gives the critical release rate above which hypermagnesemia is
eventually reached, and Section 2.2.2 converts it to a number of screws:
`sigma_hyp ~= 1.41 mmol/day`, so `sigma_hyp / sigma_0 ~= 28-29` implants. That
makes `n = 28` versus `n = 29` a clean binary key which pins `rate_implant`,
`rate_diet`, `lcl_renal` and the `C_hyp / C_hom` normalisation all at once.

```{r hyper-threshold}
sigma_hyp <- function(n) phiD * (Chyp_norm - 1) / (1 - n * xi0)
sigma_hyp(28)   # paper: ~1.41 mmol/day
sigma_hyp(28) / 0.05  # paper: minimum n is "28-29"

thresh <- tibble(n = c(27, 28, 29, 30)) |>
  rowwise() |>
  mutate(Cs_star = tail(solve_scenario(times = t_log, n_implant = n)$Cs_norm, 1)) |>
  ungroup() |>
  mutate(hypermagnesemic = Cs_star > Chyp_norm)

thresh |>
  mutate(Cs_star = signif(Cs_star, 6)) |>
  dplyr::rename("n implants" = n, "Cs* (steady state)" = Cs_star,
                "Exceeds C*hyp" = hypermagnesemic) |>
  knitr::kable(caption = paste0("Hypermagnesemia threshold. C*hyp = 1.05/0.85 = ",
                                signif(Chyp_norm, 6), "."))

# The model crosses the threshold between 28 and 29 implants, exactly as
# reported. This is a strict binary assertion, not a tolerance band.
stopifnot(
  identical(thresh$hypermagnesemic, c(FALSE, FALSE, TRUE, TRUE)),
  abs(sigma_hyp(28) - 1.41) < 0.02
)
```

The onset time follows from Table 2's `T_hyp = T_Mg log(sigma / (sigma - sigma_hyp))`.

```{r t-hyp}
n_h    <- 40
sig_h  <- n_h * 0.05
Thyp_formula <- TMg_formula * log(sig_h / (sig_h - sigma_hyp(n_h)))

s40  <- solve_scenario(times = seq(0, 600, length.out = 60001), n_implant = n_h)
Thyp_sim <- seq(0, 600, length.out = 60001)[which(s40$Cs_norm >= Chyp_norm)[1]]

c(simulated = Thyp_sim, formula = Thyp_formula)  # both a few months, per Figure 6

# T_hyp is a T_Mg-based asymptotic approximation, so agreement to a few percent
# is the expected result.
stopifnot(abs(Thyp_sim - Thyp_formula) / Thyp_formula < 0.06)
```

# Figure 4 -- the scaled-up Sato plate

Sato et al. inserted a 30 x 10 x 1 mm Mg alloy plate in a rat. Scaled up to
human proportions (about 200 times in volume) this becomes a plate of roughly
424 x 141 x 1 mm releasing `sigma = 10 mmol/day`, with an affected tissue
volume of `V_I = 1.2 L` (`xi ~= 0.023`).

```{r figure-4}
# Replicates Figure 4 of Ward 2026: the scaled-up Sato plate.
f4 <- solve_scenario(times = t_log, n_implant = 1,
                     rate_implant = 10, v_implant_zone = 1.2)

xi_plate <- 1.2 / VT
sato_eq8 <- c(Cs = 1 + 10 / phiD,
              CI = 1 + 10 / phiD + 10 * gam0 / (xi_plate * k1 * phiD))
c(Cs_sim = tail(f4$Cs_norm, 1), Cs_eq8 = unname(sato_eq8["Cs"]),
  CI_sim = tail(f4$CI_norm, 1), CI_eq8 = unname(sato_eq8["CI"]))

stopifnot(
  abs(tail(f4$Cs_norm, 1) - sato_eq8[["Cs"]]) < 1e-6,
  abs(tail(f4$CI_norm, 1) - sato_eq8[["CI"]]) < 1e-4
)

f4 |>
  select(time, Cs_norm, CN_norm, CT_norm, CI_norm) |>
  pivot_longer(-time, names_to = "state", values_to = "value") |>
  mutate(state = recode(state,
                        Cs_norm = "Cs* (serum)", CN_norm = "CN* (bone)",
                        CT_norm = "CT* (tissue)", CI_norm = "CI* (implant zone)")) |>
  filter(time > 0) |>
  ggplot(aes(time, value, colour = state)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = Chyp_norm, linetype = "dashed") +
  geom_hline(yintercept = Csev_norm, linetype = "dashed", colour = "red") +
  geom_vline(xintercept = TMg_formula, linetype = "dotted") +
  scale_x_log10(limits = c(1e-3, 1e4)) +
  labs(x = "Time (days)", y = "Normalised concentration (Cj*)", colour = NULL,
       title = "Figure 4 - scaled-up Sato plate (sigma = 10 mmol/day)",
       caption = "Replicates Figure 4 of Ward 2026.")
```

The implant zone settles at `CI* = 8.45`, the roughly eight-fold local increase
the paper reports, and serum rises to `Cs* = 2.67` -- well past the
hypermagnesemia threshold but below the "dangerous" level of 3.41.

# Figure 7 -- reduced kidney function and dietary control

For a patient with severely compromised renal function the paper sets
`gamma = gamma_0 / 3` with a matched reduction of `phi_D` so that pre-implant
homeostasis is unchanged, then modulates external Mg intake by the factor
`rho` (eq 10). The steady states of eqs 2-4 with eq 10 are given by eq 11, and
eq 12 gives the intake reduction that restores homeostasis exactly:

```
rho = 1 - sigma / phi_D
```

```{r figure-7}
gam_ckd  <- gam0 / 3   # severely compromised kidney function
phiD_ckd <- phiD / 3   # matched, so Cse = phiD/gamma is unchanged
n_ckd    <- 20
sig_ckd  <- n_ckd * 0.05
rho_star <- 1 - sig_ckd / phiD_ckd     # eq 12
rho_star

f7 <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(r) {
  solve_scenario(times = t_log, n_implant = n_ckd, f_diet = r,
                 rate_diet = phiD_ckd, cl_renal = gam_ckd) |>
    transmute(time, rho = factor(r), Cs_norm, CI_norm)
}) |> bind_rows() |> filter(time > 0)

ggplot(filter(f7, time >= 1e-2), aes(time, Cs_norm, colour = rho)) +
  geom_line(linewidth = 0.6) +
  geom_hline(yintercept = c(1, Chyp_norm), linetype = c("solid", "dashed"),
             colour = c("grey60", "black")) +
  geom_vline(xintercept = TMg_formula, linetype = "dotted") +
  scale_x_log10(limits = c(1e-2, 1e4)) +
  labs(x = "Time (days)", y = "Cs* (normalised serum)", colour = "rho",
       title = "Figure 7A - dietary control, n = 20, gamma = gamma_0/3",
       caption = "Replicates Figure 7A of Ward 2026. Grey line: homeostasis.")
```

The eq 11 steady states, and the eq 12 exact-homeostasis result, both reproduce:

```{r eq11-12}
eq11 <- function(rho, n, phiD_l, gam_l) {
  sig <- n * 0.05
  xi  <- n * xi0
  c(Cs = rho + sig / phiD_l,
    CI = rho + sig / phiD_l + sig * gam_l / (xi * k1 * phiD_l))
}

eq11_chk <- lapply(c(0, 0.25, 0.5, 0.75, 1), function(r) {
  s <- solve_scenario(times = t_log, n_implant = n_ckd, f_diet = r,
                      rate_diet = phiD_ckd, cl_renal = gam_ckd)
  p <- eq11(r, n_ckd, phiD_ckd, gam_ckd)
  tibble(rho = r,
         Cs_sim = tail(s$Cs_norm, 1), Cs_eq11 = unname(p["Cs"]),
         CI_sim = tail(s$CI_norm, 1), CI_eq11 = unname(p["CI"]))
}) |> bind_rows()

eq11_chk |>
  mutate(across(-rho, ~ signif(.x, 6))) |>
  dplyr::rename("rho" = rho, "Cs* (sim)" = Cs_sim, "Cs* (eq 11)" = Cs_eq11,
                "CI* (sim)" = CI_sim, "CI* (eq 11)" = CI_eq11) |>
  knitr::kable(caption = "Dietary control: simulated steady states vs eq 11.")

# eq 12: halving intake restores homeostasis exactly for this scenario.
cs_at_rho_star <- tail(
  solve_scenario(times = t_log, n_implant = n_ckd, f_diet = rho_star,
                 rate_diet = phiD_ckd, cl_renal = gam_ckd)$Cs_norm, 1)
c(rho_star = rho_star, Cs_star = cs_at_rho_star)

stopifnot(
  max(abs(eq11_chk$Cs_sim - eq11_chk$Cs_eq11)) < 1e-6,
  max(abs(eq11_chk$CI_sim - eq11_chk$CI_eq11)) < 1e-4,
  abs(rho_star - 0.5) < 1e-12,
  abs(cs_at_rho_star - 1) < 1e-5
)
```

## The early-phase dip

Cutting dietary intake at `t = 0` while the implant contribution is still
building produces a fast drop in serum Mg over the first day or so, before the
slow implant-driven rise takes over. Section 2.3.1 quantifies it with a
minimum-concentration formula and states that "the maximum drop is predicted to
be about 8% of homeostatic levels when Mg intake is zero", using the healthy
Table 1 parameters.

The printed formula is mangled in the typeset PDF (see Errata below); the
reading consistent with the paper's own quoted 8% is

```
Cs*_min ~= 1 + (rho - 1) * Gamma,   Gamma = gamma / k_1
```

which is the dimensionless group `Gamma` the paper defines in Section 2.2.3.

```{r early-dip}
Gamma0 <- gam0 / k1
Gamma0                       # paper prints Gamma_0 ~= 0.0787

# Healthy parameters, intake cut to zero: measure the fast-phase minimum before
# the slow systemic rise sets in (the paper's "around 1-10 days" window).
dip <- solve_scenario(times = seq(0, 1, length.out = 2001),
                      n_implant = 20, f_diet = 0)
dip_pct     <- 100 * (1 - tail(dip$Cs_norm, 1))
formula_pct <- 100 * Gamma0

c(simulated_drop_pct = dip_pct, formula_drop_pct = formula_pct)  # paper: "about 8%"

stopifnot(abs(dip_pct - formula_pct) < 0.5, abs(formula_pct - 8) < 1)
```

The simulated fast-phase drop is `r round(dip_pct, 1)`% against the formula's
`r round(formula_pct, 1)`%, matching the paper's "about 8%" and confirming the
intended reading of the printed equation.

# Assumptions, deviations and errata

* **No IIV and no residual error.** The publication is a deterministic
  mechanistic simulation of an "average" individual and reports neither a
  variance model nor a residual-error model. None is invented; the model file
  carries typical values only, all wrapped in `fixed()`.

* **Initial conditions are computed, not transcribed.** `serum(0)` etc. are set
  from eq 5 (`rate_diet / cl_renal` and the exchange ratios) rather than from
  Table 1's printed homeostatic concentrations. Table 1's `gamma = 10.9` is the
  rounded form of `phi_D / Cse = 6 / 0.5525 = 10.86`, so `phi_D / gamma =
  0.5505` differs from the printed `Cse = 0.553` by 0.4%. Seeding from the
  printed value would leave the model 0.4% off its own steady state and destroy
  the exact-homeostasis check above. The same rounding chain explains `CNe`
  42.97 vs the printed 43.1, `CTe` 2.189 vs 2.20, and `Gamma_0` 0.0790 vs the
  printed 0.0787.

* **`V_T` in eq 3 is `V_Ttot`, not `(1 - xi) V_Ttot`.** The `(1 - xi)` factor
  appears only on the right-hand side flux. This is settled by Supporting
  Information B (S15), "V_I is chosen so that V_I/V_T = 0.0001 giving V_I =
  5.27 ml", and independently by Table 2's `T_1 = V_T (1 + xi_3) / k_-1`, which
  evaluates to 6.07 days only with the full 52.7 L.

* **Erratum -- the Supporting Information contradicts Table 1 on one ratio.**
  Supporting Information B (p. S12) states `mu_-1 / mu_1 ~= Cse / CNe ~=
  0.0385`. Table 1's own values give `0.0775 / 6.05 = 0.01281`, and
  `Cse / CNe = 0.553 / 43.1 = 0.01282`. The prose figure is a factor of three
  too large (`0.0385 = 3 x 0.0128`). **Table 1 is self-consistent and is
  used here**: back-solving the prose ratio would force `mu_-1 = 0.233` and
  `CNe = 14.3`, contradicting Table 1's printed `CNe = 43.1` and breaking the
  eq 8 steady states that the simulations above reproduce exactly.

* **Erratum -- Section 2.3.1's minimum-concentration formula is mangled.** It
  is typeset as `C*_smin ~= 1 + (rho - 1)/gamma k_1`, which is not
  reconstructible as printed. The reading `1 + (rho - 1) * gamma / k_1` gives
  7.9% for `rho = 0`, matching both the paper's own quoted "about 8%" and the
  simulated fast-phase drop of 7.8%; no other grouping of the same symbols does
  (`1/(gamma k_1)` gives 0.07%). That reading is used in the check above and is
  the dimensionless `Gamma = gamma / k_1` defined in Section 2.2.3.

* **Erratum -- Figure 6 axis label.** The x-axis of Figure 6 is labelled
  `sigma (mmol/L)`; `sigma` is a release *rate* in mmol/day everywhere else in
  the paper, including Table 1 and the Figure 6 caption's own text.

* **`V_I` is a modelling choice, not a measurement.** The peri-implant zone has
  no anatomic boundary. The authors pick `V_I / V_Ttot = 1e-4` (5.27 mL, about
  20 times the volume of the screw itself) and note that `C_I` scales as
  `1 / V_I` while serum, bone and tissue predictions are insensitive to it. The
  paper explicitly says its implant-region predictions "underestimate the
  recorded Mg excretion rate and the rate of Mg increase in localized tissue
  concentration; though quantitative agreement for the latter can be tuned by
  decreasing the volume `V_I`."

* **Scenario levers are `ini()` parameters.** `n_implant` (eq 9) and `f_diet`
  (`rho`, eq 10) are fixed at 1 in the packaged model, i.e. one screw and
  normal dietary intake. They are varied above by re-pinning them, which is how
  every published figure is reproduced. They are not estimated quantities.

* **No non-compartmental analysis.** There is no dose and no
  absorption-distribution-elimination profile, so PKNCA is not an applicable
  validation and is deliberately omitted in favour of the steady-state,
  closed-form, time-constant and perturbation checks above.
