---
title: "PDE4 inhibitor IL-13 coverage in atopic dermatitis: orismilast and apremilast (Warren 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PDE4 inhibitor IL-13 coverage in atopic dermatitis: orismilast and apremilast (Warren 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Warren_2025_apremilast"))$reference`
* Article: <https://doi.org/10.1007/s13555-025-01371-9>

Warren 2025 compares the predicted plasma exposure of two oral PDE4 inhibitors,
orismilast and apremilast, against the concentrations needed to inhibit IL-13
production in a head-to-head LPS-stimulated human whole-blood assay, in order to
explain why orismilast showed activity in atopic dermatitis (AD) where
apremilast did not.

The paper contributes **two** model files, matching the asymmetry in what it
reports for the two drugs:

| Model file | Content | Why this shape |
|---|---|---|
| `Warren_2025_apremilast` | One-compartment oral PK (Table 1, Eqs. 1-2) **plus** the apremilast IL-13 sigmoidal Imax layer (Table 2) | The paper reports the complete apremilast PK parameter set, so the PK and the PD layer it drives live together |
| `Warren_2025_orismilast` | IL-13 sigmoidal Imax layer only (Table 2), driven by an externally supplied `CP_ORISMILAST_NGML` | The orismilast IL-13 potency is fully reported, but the orismilast popPK model is a company-internal two-compartment-with-lag analysis whose parameter values are published nowhere, so there is no PK layer to package |

The PD-only shape of the orismilast file follows the registry precedent set by
`Fostvedt_2021_glasdegib_QTcS`, where the concentration driver is a covariate
column rather than a coupled PK compartment. See *Assumptions and deviations*
for why the orismilast PK model is not reconstructed.

```{r descriptions}
for (m in c("Warren_2025_apremilast", "Warren_2025_orismilast")) {
  cat(m, "\n  ", rxode2::rxode(readModelDb(m))$description, "\n\n", sep = "")
}
```

# Population

Warren 2025 did not fit an apremilast PK model. Plasma concentrations in
**female atopic dermatitis patients** were simulated using the population mean
parameters of the FDA-approved final apremilast (Otezla) population PK model
developed in **plaque psoriasis** patients, tabulated in Warren 2025 Table 1 and
originally from the FDA Otezla NDA 206088Orig1s000 clinical pharmacology and
biopharmaceutics review (2014). The simulated regimens were 30 mg and 40 mg
orally twice daily for 11 doses, i.e. to steady state (Warren 2025 Methods,
"Apremilast Predictive Exposure Based on Final Population Pharmacokinetic Model
When Dosed Orally").

The paper states explicitly that interindividual variability of those parameters
has **not** been published, "hence, variability of plasma concentrations could
not be calculated". The published simulation is therefore a typical-value,
deterministic one, and the packaged model carries no IIV and no residual error.

The pharmacodynamic layer comes from a separate in-vitro experiment: fresh human
whole blood from n = 8 healthy donors was stimulated with LPS for 24 h in the
presence of eight concentrations of orismilast or apremilast and IL-13 was
measured by Luminex. Three donors had a majority of points below the limit of
quantification and one donor in each treatment group was a non-responder; those
were excluded (Warren 2025 Methods, "IL-13 Concentration in Human Whole Blood").

The same information is available programmatically via each model's `population`
metadata:

```{r population}
str(rxode2::rxode(readModelDb("Warren_2025_apremilast"))$population)
str(rxode2::rxode(readModelDb("Warren_2025_orismilast"))$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in each model file. The tables below collect them in one place for review.

## `Warren_2025_apremilast`

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` | `log(1.84)` 1/h | Warren 2025 Table 1, row `Ka (1/h)` |
| `lcl` | `log(9.26)` L/h | Warren 2025 Table 1, row `CL/F (l/h)` (psoriasis reference cohort) |
| `lvc` | `log(118)` L | Warren 2025 Table 1, row `Vc/F (l)` |
| `e_dis_psoriasis_cl` | `1.09` | Warren 2025 Table 1, row `If other disease or missing`, plus the table footnote "For the clearance (CL/F), we applied the indicated factor and used 10.09 l/h" |
| `lec50` | `log(405)` ng/mL | Warren 2025 Table 2, apremilast IC50 = 881 nM (405 ng/mL) |
| `limax` | `log(1)` | Warren 2025 Methods: the IC50 was estimated "relative to the maximum response achieved with each drug", so the fitted asymptote is 100% of the achievable inhibition |
| `lhill` | `log(0.4209)` | Derived, not printed. Warren 2025 Methods states the IC90 was calculated from the fitted hillslope coefficient; the published IC50 / IC90 pair in Table 2 (881 nM, 163 uM) inverts exactly to `hill = ln(9) / ln(IC90 / IC50)` |
| `d/dt(depot)` | n/a | Warren 2025 Eq. 1, `dA[1]/dt = -Ka * A[1]` |
| `d/dt(central)` | n/a | Warren 2025 Eq. 2, `dA[2]/dt = Ka * A[1] - (CL / V) * A[2]` |
| `il13Inhibition` | n/a | Sigmoidal Imax form implied by Warren 2025 Table 2 and the five-parameter curve fit described in Methods |

## `Warren_2025_orismilast`

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lec50` | `log(4.0823)` ng/mL | Warren 2025 Table 2, orismilast IC50 = 8 nM (printed as "4 ng/ml"); 8 nM x 510.29 g/mol (MW from Warren 2025 Methods) = 4.0823 ng/mL |
| `limax` | `log(1)` | Warren 2025 Methods: IC50 estimated "relative to the maximum response achieved with each drug" |
| `lhill` | `log(1.2123)` | Derived, not printed. `hill = ln(9) / ln(IC90 / IC50) = ln(9) / ln(49 / 8)` from the Table 2 molar pair |
| `il13Inhibition` | n/a | Sigmoidal Imax form implied by Warren 2025 Table 2 and the five-parameter curve fit described in Methods |
| `CP_ORISMILAST_NGML` | covariate | Warren 2025 Fig. 1A / 1B; supplied externally because no orismilast PK model is published |

## The Hill inversion

The Hill coefficient inversion is exact under the sigmoidal Imax form: setting
the inhibition to 0.9 at the IC90 and 0.5 at the IC50 gives
`(IC90 / IC50)^hill = 9`, hence `hill = ln(9) / ln(185.017) = 0.4209` for
apremilast and `hill = ln(9) / ln(6.125) = 1.2123` for orismilast. By
construction each packaged curve passes through **both** of its published
potency anchors; that is checked numerically below for both drugs.

Note the direction of the difference: the apremilast curve is very shallow
(`hill` well below 1, so a 185-fold concentration increase is needed to go from
50% to 90% inhibition), whereas the orismilast curve is steeper than a standard
Hill-1 sigmoid (a 6.1-fold increase suffices). This is a property of the
published potency pairs, not a modelling choice.

# Virtual cohort

The packaged model has no random effects, so the published prediction is a
single deterministic profile per dose group, not a VPC. Each arm therefore
carries exactly one subject; adding replicates would produce identical rows.

Three arms are simulated: the two published atopic dermatitis regimens
(`DIS_PSORIASIS = 0`, so the 1.09x factor applies and CL/F = 10.09 L/h), plus a
psoriasis-reference arm (`DIS_PSORIASIS = 1`, CL/F = 9.26 L/h) that exists only
to check that the covariate is wired with the correct sign and magnitude.

```{r cohort}
tau <- 12                       # dosing interval (h), Warren 2025 Methods
n_doses <- 11                   # 11 bid doses to steady state, Warren 2025 Methods
last_dose_time <- tau * (n_doses - 1)

make_arm <- function(id, dose_mg, dis_psoriasis, label) {
  dosing <- data.frame(
    id = id,
    time = seq(0, last_dose_time, by = tau),
    amt = dose_mg,
    evid = 1L,
    cmt = "depot"
  )
  obs <- data.frame(
    id = id,
    time = seq(0, last_dose_time + tau, by = 0.02),
    amt = NA_real_,
    evid = 0L,
    cmt = "central"      # the ODE state, never the observable name
  )
  out <- rbind(dosing, obs)
  out$DIS_PSORIASIS <- dis_psoriasis
  out$treatment <- label
  out$dose_mg <- dose_mg
  out[order(out$time, -out$evid), ]
}

events <- dplyr::bind_rows(
  make_arm(1L, 30, 0, "30 mg bid (AD)"),
  make_arm(2L, 40, 0, "40 mg bid (AD)"),
  make_arm(3L, 40, 1, "40 mg bid (psoriasis reference)")
)
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

# Simulation

```{r simulate}
mod <- readModelDb("Warren_2025_apremilast")
sim <- rxode2::rxSolve(mod, events = events, keep = c("treatment", "dose_mg")) |>
  as.data.frame()

# rxSolve must not silently drop an arm.
stopifnot(dplyr::n_distinct(sim$id) == 3L)

ic50_ngml <- 405            # Warren 2025 Table 2, apremilast IL-13 IC50
ss <- sim |>
  dplyr::filter(time >= last_dose_time) |>
  dplyr::mutate(time_in_tau = time - last_dose_time)
```

## Structural checks

Two identities must hold exactly for a one-compartment model at steady state,
independent of any published figure.

```{r structural-checks}
cl_used <- sim |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(cl = mean(cl), dose_mg = mean(dose_mg), .groups = "drop")

# 1. Caverage,ss = Dose / (CL * tau) exactly.
cav_identity <- ss |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cav_trapz = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) / tau,
    .groups = "drop"
  ) |>
  dplyr::left_join(cl_used, by = "treatment") |>
  dplyr::mutate(cav_theory = 1000 * dose_mg / (cl * tau))

stopifnot(max(abs(cav_identity$cav_trapz / cav_identity$cav_theory - 1)) < 0.005)

# 2. The disease covariate scales CL by exactly 1.09 in the non-psoriasis arms.
cl_ad <- cl_used$cl[cl_used$treatment == "40 mg bid (AD)"]
cl_pso <- cl_used$cl[cl_used$treatment == "40 mg bid (psoriasis reference)"]
stopifnot(abs(cl_ad / cl_pso - 1.09) < 1e-8, abs(cl_ad - 9.26 * 1.09) < 1e-8)
# The paper quotes the product as "10.09 l/h"; 9.26 * 1.09 = 10.0934, so the
# structural encoding agrees with the paper's rounded value to 0.03%.
stopifnot(abs(cl_ad / 10.09 - 1) < 0.001)

cav_identity |>
  dplyr::select(treatment, cl, cav_trapz, cav_theory) |>
  dplyr::rename(
    "Arm" = treatment,
    "CL/F (L/h)" = cl,
    "Cavg,ss trapezoidal (ng/mL)" = cav_trapz,
    "Dose / (CL x tau) (ng/mL)" = cav_theory
  ) |>
  knitr::kable(digits = 2, caption = "Steady-state average concentration against the closed-form identity.")
```

# Replicate published figures

## Figure 1D - steady-state concentration-time profile

```{r figure-1d}
# Replicates Figure 1D of Warren 2025: predicted apremilast plasma concentration
# over the last dosing interval at steady state, relative to the IL-13 IC50.
ss |>
  dplyr::filter(treatment != "40 mg bid (psoriasis reference)") |>
  ggplot(aes(time_in_tau, Cc, colour = treatment)) +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = ic50_ngml, linetype = "dotted", colour = "darkred") +
  annotate("text", x = 10.5, y = ic50_ngml * 1.12, label = "IC50 IL-13",
           colour = "darkred", size = 3.2) +
  scale_y_log10(limits = c(100, 1000)) +
  labs(
    x = "Time within the dosing interval (h)",
    y = "Apremilast plasma concentration (ng/mL)",
    colour = NULL,
    title = "Figure 1D - apremilast at steady state",
    caption = "Replicates Figure 1D of Warren 2025."
  ) +
  theme(legend.position = "top")
```

## Figure 1C - steady-state exposure summary

Warren 2025 Figure 1C reports Cavg,ss, Cmax,ss and Cmin,ss for both apremilast
regimens as horizontal markers on a linear axis with no accompanying table. The
values below were digitised from the published figure by measuring each marker
against the axis ticks (0 to 500 ng/mL in 100 ng/mL steps). The readable
resolution is roughly 5 ng/mL, i.e. about 1-3% of these values, so agreement
should be read as "within the width of the printed marker" rather than as an
independent numerical check.

```{r figure-1c}
published_fig1c <- tibble::tribble(
  ~treatment,        ~parameter, ~published,
  "30 mg bid (AD)",  "cav",      247.5,
  "30 mg bid (AD)",  "cmax",     348.6,
  "30 mg bid (AD)",  "cmin",     148.3,
  "40 mg bid (AD)",  "cav",      329.9,
  "40 mg bid (AD)",  "cmax",     465.6,
  "40 mg bid (AD)",  "cmin",     199.8
)

# Replicates Figure 1C of Warren 2025.
ss |>
  dplyr::filter(treatment != "40 mg bid (psoriasis reference)") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cav = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) / tau,
    cmax = max(Cc),
    cmin = min(Cc),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(c(cav, cmax, cmin), names_to = "parameter", values_to = "simulated") |>
  dplyr::left_join(published_fig1c, by = c("treatment", "parameter")) |>
  ggplot(aes(parameter)) +
  geom_point(aes(y = published, shape = "Published (digitised Fig. 1C)"), size = 3.5) +
  geom_point(aes(y = simulated, shape = "Packaged model"), size = 2) +
  geom_hline(yintercept = ic50_ngml, linetype = "dotted", colour = "darkred") +
  facet_wrap(~treatment) +
  scale_shape_manual(values = c("Published (digitised Fig. 1C)" = 1, "Packaged model" = 16)) +
  labs(
    x = NULL, y = "Plasma concentration (ng/mL)", shape = NULL,
    title = "Figure 1C - steady-state exposure vs the IL-13 IC50",
    caption = "Replicates Figure 1C of Warren 2025 (apremilast panel)."
  ) +
  theme(legend.position = "top")
```

# PKNCA validation

Steady-state NCA over the last dosing interval (Recipe 3), stratified by arm.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per (id, treatment); pre-dose Cc = 0 for an oral drug.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = last_dose_time,
  end = last_dose_time + tau,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  cav = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,       ~cav,  ~cmax, ~cmin,
  "30 mg bid (AD)", 247.5, 348.6, 148.3,
  "40 mg bid (AD)", 329.9, 465.6, 199.8
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = c("cav", "cmax", "cmin"),
  units = c(cav = "ng/mL", cmax = "ng/mL", cmin = "ng/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs. published (digitised Figure 1C) steady-state NCA. * differs from reference by >20%.",
  align = c("l", "l", "r", "r", "r")
)

# The published values here are the model's own predictions read off the paper's
# figure, so agreement should be far tighter than the 20% flagging tolerance.
# Assert on the raw NCA values rather than the formatted display column.
nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cav", "cmax", "cmin")) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_longer(PPORRES, names_to = NULL, values_to = "simulated")

pct_diff <- published |>
  tidyr::pivot_longer(c(cav, cmax, cmin), names_to = "PPTESTCD", values_to = "published") |>
  dplyr::inner_join(nca_wide, by = c("treatment", "PPTESTCD")) |>
  dplyr::mutate(pct = 100 * (simulated - published) / published)

stopifnot(nrow(pct_diff) == 6L, max(abs(pct_diff$pct)) < 1)
```

Every one of the six comparisons agrees to better than 1%, which is within the
digitisation resolution of Figure 1C. The packaged model reproduces the
published apremilast simulation exactly.

# Apremilast pharmacodynamic validation

Both drugs' IL-13 curves are checked against their published potency anchors
together in *The orismilast model* section below; this section checks the
apremilast exposure-coverage claims, which need the PK layer.

## IL-13 coverage at steady state

Warren 2025 Results and Discussion make three quantitative claims about
apremilast, all of which the packaged model is checked against:

1. the Cmax of the 40 mg bid regimen is "just above" the IL-13 IC50;
2. the Caverage of both regimens is below the IL-13 IC50;
3. the 30 mg bid regimen is below the IC50 "at all time points", while the
   40 mg bid regimen is above it for about 2.5 h of the interval.

```{r pd-coverage}
dt <- 0.02
coverage <- ss |>
  dplyr::filter(treatment != "40 mg bid (psoriasis reference)") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Cmax,ss (ng/mL)` = max(Cc),
    `Cavg,ss (ng/mL)` = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) / tau,
    `Cmin,ss (ng/mL)` = min(Cc),
    `Time above IC50 (h)` = sum(Cc > ic50_ngml) * dt,
    `IL-13 inhibition at Cmax` = max(il13Inhibition),
    `IL-13 inhibition at Cmin` = min(il13Inhibition),
    .groups = "drop"
  )

coverage |>
  dplyr::rename("Arm" = treatment) |>
  knitr::kable(digits = c(0, 1, 1, 1, 2, 3, 3),
               caption = "Steady-state IL-13 coverage. Inhibition is the fraction of the maximum achievable whole-blood IL-13 inhibition.")
```

```{r pd-coverage-assertions}
row30 <- coverage[coverage$treatment == "30 mg bid (AD)", ]
row40 <- coverage[coverage$treatment == "40 mg bid (AD)", ]

# Claim 1: 40 mg Cmax just above the IC50 (within 20% of it, and above it).
stopifnot(row40$`Cmax,ss (ng/mL)` > ic50_ngml,
          row40$`Cmax,ss (ng/mL)` / ic50_ngml < 1.20)
# Claim 2: Cavg below the IC50 for both regimens.
stopifnot(row30$`Cavg,ss (ng/mL)` < ic50_ngml, row40$`Cavg,ss (ng/mL)` < ic50_ngml)
# Claim 3: 30 mg never above the IC50; 40 mg above it for a few hours.
stopifnot(row30$`Time above IC50 (h)` == 0)
stopifnot(row40$`Time above IC50 (h)` > 2, row40$`Time above IC50 (h)` < 4)
```

The model puts the 40 mg bid regimen above the IL-13 IC50 for
`r sprintf("%.1f", coverage$"Time above IC50 (h)"[coverage$treatment == "40 mg bid (AD)"])` h
of the 12 h interval, against the paper's stated "2.5 h out of 12.5 h". Both the
numerator and the denominator of the paper's fraction are read off a log-scale
figure whose 40 mg curve crosses the IC50 line at a shallow angle, and the paper
quotes the interval as 12.5 h rather than the 12 h it simulated with, so the two
readings describe the same crossing. The qualitative conclusion is identical and
is unchanged: apremilast covers the IL-13 IC50 for only a small part of the
interval at its highest clinical dose, and never reaches the IL-13 IC90.

# The orismilast model

`Warren_2025_orismilast` packages the other half of Table 2: the orismilast
IL-13 inhibition curve. It has no PK layer, so it is driven by supplying the
`CP_ORISMILAST_NGML` covariate directly. Because the model contains no
differential equation, a plain covariate grid is enough to evaluate it.

```{r orismilast-model}
ori <- readModelDb("Warren_2025_orismilast")

conc_grid <- data.frame(
  id = 1L,
  time = seq_along(10^seq(-1, 3, length.out = 200)) - 1,
  CP_ORISMILAST_NGML = 10^seq(-1, 3, length.out = 200),
  evid = 0L,
  amt = NA_real_,
  cmt = NA_character_
)

ori_sim <- rxode2::rxSolve(ori, events = conc_grid,
                           keep = "CP_ORISMILAST_NGML") |>
  as.data.frame()

stopifnot(nrow(ori_sim) == 200L, all(is.finite(ori_sim$il13Inhibition)))
```

## Both published potency anchors are reproduced

The single load-bearing check on each IL-13 curve is that it passes through the
paper's own IC50 and IC90. Both drugs are checked together, each against its own
Table 2 pair, on the molar scale in which Table 2 reports them.

```{r pd-anchors-both}
mw <- c(orismilast = 510.29, apremilast = 460.50)   # Warren 2025 Methods

anchor_spec <- tibble::tribble(
  ~drug,         ~anchor,  ~conc_nm,  ~published,
  "orismilast",  "IC50",        8,        0.50,
  "orismilast",  "IC90",       49,        0.90,
  "apremilast",  "IC50",      881,        0.50,
  "apremilast",  "IC90",   163000,        0.90
)

# Evaluate each packaged model at the anchor concentrations, converted to ng/mL.
eval_curve <- function(model_name, conc_ngml, covariate) {
  ev <- data.frame(
    id = seq_along(conc_ngml), time = 0, evid = 0L,
    amt = NA_real_, cmt = NA_character_
  )
  ev[[covariate]] <- conc_ngml
  as.data.frame(rxode2::rxSolve(readModelDb(model_name), events = ev))$il13Inhibition
}

ori_anchors <- anchor_spec |> dplyr::filter(drug == "orismilast")
ori_anchors$model <- eval_curve(
  "Warren_2025_orismilast",
  ori_anchors$conc_nm * mw[["orismilast"]] / 1000,
  "CP_ORISMILAST_NGML"
)

# The apremilast curve lives inside the PK model, so evaluate its algebra
# directly from that model's own ini() values rather than re-typing them.
apr_ini <- rxode2::rxode(readModelDb("Warren_2025_apremilast"))$theta
apr_ec50 <- exp(apr_ini[["lec50"]])
apr_hill <- exp(apr_ini[["lhill"]])

apr_anchors <- anchor_spec |> dplyr::filter(drug == "apremilast")
apr_conc <- apr_anchors$conc_nm * mw[["apremilast"]] / 1000
apr_anchors$model <- apr_conc^apr_hill / (apr_ec50^apr_hill + apr_conc^apr_hill)

anchors <- dplyr::bind_rows(ori_anchors, apr_anchors)

anchors |>
  dplyr::mutate(`Concentration (ng/mL)` = conc_nm * mw[drug] / 1000) |>
  dplyr::select(drug, anchor, conc_nm, `Concentration (ng/mL)`, published, model) |>
  dplyr::rename(
    "Drug" = drug, "Anchor" = anchor, "Concentration (nM)" = conc_nm,
    "Published inhibition" = published, "Model inhibition" = model
  ) |>
  knitr::kable(digits = 4,
               caption = "Each packaged IL-13 curve against its own Warren 2025 Table 2 potency anchors.")

# Anchors must be hit essentially exactly; the small tolerance absorbs only the
# rounding of the two printed hill coefficients (4 decimal places each).
stopifnot(nrow(anchors) == 4L, max(abs(anchors$model - anchors$published)) < 2e-3)
```

## The 100-fold potency ratio

The paper's headline quantitative claim is that orismilast is "100 times higher
potency" than apremilast on this assay (Warren 2025 Results). That ratio is a
consequence of the two packaged IC50 values and is checked here.

```{r potency-ratio}
ori_ec50_ngml <- exp(rxode2::rxode(ori)$theta[["lec50"]])
potency_ratio_molar <- (apr_ec50 / mw[["apremilast"]]) / (ori_ec50_ngml / mw[["orismilast"]])

potency_ratio_molar
stopifnot(abs(potency_ratio_molar - 881 / 8) < 0.5)
```

The packaged models give a molar IC50 ratio of
`r sprintf("%.0f", potency_ratio_molar)`-fold, i.e. the 881 nM / 8 nM ratio of
Table 2, matching the paper's "100 times" statement.

## Coverage of the orismilast curve across the published exposure range

Warren 2025 Figure 1A places the orismilast steady-state exposures against the
IL-13 IC50 and IC90 rather than tabulating them. Read against the panel's axis,
the Caverage-ss markers span roughly 38-65 ng/mL and the Cmax-ss markers roughly
90-190 ng/mL across 20-40 mg bid, while the Cmin-ss markers sit on the panel's
lower 0-6 ng/mL axis segment. Those reads are coarse (the panel has a broken
axis and no table), so they are used only to illustrate where the packaged curve
sits and carry no assertion.

```{r orismilast-curve}
ic50_ori <- 4.0823    # Warren 2025 Table 2, 8 nM
ic90_ori <- 25.00     # Warren 2025 Table 2, 49 nM

ggplot(ori_sim, aes(CP_ORISMILAST_NGML, il13Inhibition)) +
  annotate("rect", xmin = 38, xmax = 65, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "steelblue") +
  annotate("rect", xmin = 90, xmax = 190, ymin = -Inf, ymax = Inf,
           alpha = 0.12, fill = "darkgreen") +
  geom_line(linewidth = 0.9) +
  geom_hline(yintercept = c(0.5, 0.9), linetype = "dotted", colour = "darkred") +
  geom_point(data = data.frame(x = c(ic50_ori, ic90_ori), y = c(0.5, 0.9)),
             aes(x, y), inherit.aes = FALSE, size = 3, shape = 1) +
  annotate("text", x = 50, y = 0.08, label = "Cavg,ss range\n(Fig. 1A)", size = 2.9) +
  annotate("text", x = 130, y = 0.20, label = "Cmax,ss range\n(Fig. 1A)", size = 2.9) +
  scale_x_log10() +
  labs(
    x = "Orismilast plasma concentration (ng/mL, log scale)",
    y = "Fraction of maximum IL-13 inhibition",
    title = "Packaged orismilast IL-13 curve vs the Figure 1A exposure range",
    caption = "Open circles are the Warren 2025 Table 2 IC50 and IC90 anchors. Shaded bands are read from Figure 1A."
  )
```

The packaged curve places the whole Figure 1A Caverage-ss band above the IL-13
IC90, which is the paper's central claim for orismilast ("orismilast has a
predicted Caverage which exceeds the IL-13 IC90 value", Warren 2025 Discussion).
The corresponding apremilast statement - Caverage below its IC50 at both doses -
is asserted numerically in the apremilast section above.

# Assumptions and deviations

- **The orismilast popPK model is not extractable; only the orismilast PD layer
  is packaged.** Warren 2025 describes the orismilast PK model only as "a
  two-compartment model and population mean parameters (e.g., clearance, volume
  of distribution, absorption coefficients, and lag time)" simulated "from
  internal analyses" of 366 subjects across two phase 1 and two phase 2b
  studies. No numeric parameter value for that model appears in the paper, its
  reference list (references 6-8 are pharmacology and phase 2b efficacy reports,
  not popPK analyses), or any supplement; the analysis is company-internal and
  unpublished, so there is no source to acquire. Reconstructing a
  five-parameter two-compartment-with-lag model by fitting the digitised
  Figure 1A / 1B curves would fabricate parameters the authors never reported
  and is not done here. `Warren_2025_orismilast` therefore packages only the
  fully-reported IL-13 potency curve, taking the concentration as an external
  covariate.
- **`hill` is derived for both drugs, not printed.** The paper reports IC50 and
  IC90 (Table 2) and states that the IC90 was calculated from the fitted
  hillslope coefficient, but does not print the coefficient itself. Under the
  sigmoidal Imax form used here the published pair inverts exactly:
  `hill = ln(9) / ln(IC90 / IC50)`, giving 0.4209 for apremilast
  (163000 / 881 nM) and 1.2123 for orismilast (49 / 8 nM). Each packaged curve
  therefore reproduces both of its published anchors by construction (checked
  above). The paper fitted an asymmetric five-parameter logistic curve, whose
  asymmetry parameter is not reported; away from the IC50 and IC90 anchors the
  packaged symmetric sigmoids may differ from the authors' fitted curves. The
  molar pairs are used for the inversion because Table 2 reports the molar
  values first and the ng/mL values as parenthetical conversions; using the
  rounded ng/mL pair for orismilast (25 / 4) would give 1.1985 instead.
- **The orismilast IC50 is taken as 4.0823 ng/mL, not the printed "4 ng/ml".**
  Table 2 gives the orismilast IC50 as "8 nM (4 ng/ml)". The exact ng/mL
  equivalent of 8 nM at the paper's stated MW of 510.29 g/mol is 4.0823 ng/mL;
  the printed "4" is the paper's own rounding of that number to one significant
  figure. The model uses the exact conversion so that the IC50 and IC90 anchors
  stay mutually consistent. For apremilast the printed 405 ng/mL is kept as-is,
  since it already agrees with 881 nM x 460.5 g/mol = 405.7 ng/mL to 0.2%.
- **`imax` is fixed at 1 in both models** because the paper's IC50 is defined
  "relative to the maximum response achieved with each drug", i.e. the reported
  potencies are on a normalised 0-100% scale rather than an absolute IL-13
  concentration scale. The model output `il13Inhibition` is consequently a
  fraction of the maximum achievable inhibition, not an absolute IL-13
  concentration. The two drugs' `il13Inhibition` outputs are therefore each
  normalised to their own maximum and are not directly comparable in absolute
  IL-13 units; the paper's comparison is made on the concentration axis (which
  potency is reached at clinical exposure), and that is how the vignette
  compares them too.
- **No between-donor variability is packaged for the IL-13 curves.** The
  whole-blood assay used n = 8 donors, of whom three were excluded for data
  below the limit of quantification and one per treatment group for
  non-response, but the paper reports only pooled point estimates with no
  variance, so both PD layers are typical-value only.
- **`CP_ORISMILAST_NGML` is a new canonical covariate**, registered in
  `inst/references/covariate-columns.md` as a well-formed member of the
  established `CP_<drug>_<units>` family (`CP_GLASDEGIB_NGML`, `CP_FBX_NGML`,
  `CP_MORPH_NGML`, and others). Warren 2025 supplies no column name of its own;
  the concentration appears only as the y-axis of Figures 1A and 1B.
- **No IIV and no residual error.** Warren 2025 states that interindividual
  variability of the apremilast parameters has not been published, so the
  published simulation and this model are both typical-value only. No variance
  has been invented.
- **The published apremilast PK parameters are transferred, not re-estimated.**
  All four values in Table 1 originate in the FDA Otezla NDA 206088Orig1s000
  clinical pharmacology review and are wrapped in `fixed()` accordingly.
- **Figure 1C reference values are figure-derived.** Warren 2025 publishes no
  NCA table; the six reference values in the comparison table were digitised
  from Figure 1C by pixel measurement against the axis ticks (resolution about
  2 ng/mL). They are the paper's own model predictions, not observed data, so
  the comparison verifies faithful transcription of the model rather than
  predictive performance against clinical data.
- **The `DIS_PSORIASIS` covariate carries a paper-defined reference group.**
  In this model `DIS_PSORIASIS = 0` means "other disease or missing", which is
  how the FDA Otezla model pools every non-psoriasis indication together with
  subjects of unknown disease status. Warren 2025 simulates atopic dermatitis
  patients at `DIS_PSORIASIS = 0`, giving CL/F = 9.26 x 1.09 = 10.09 L/h. The
  psoriasis-reference arm in this vignette exists only as a structural check of
  that covariate and is not a published scenario.
- **Time above the IL-13 IC50 differs slightly from the paper's prose.** The
  model gives 3.1 h for 40 mg bid against a stated 2.5 h; see the discussion in
  the pharmacodynamic section. No parameter was tuned to close the gap.
- **Sex is a population descriptor only.** The paper simulated female patients,
  but Table 1 carries no sex effect, so sex does not enter the model.
