---
title: "Methadone (Watt 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Methadone (Watt 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Watt_2026_methadone"))$reference`
* Description: `r rxode2::rxode(readModelDb("Watt_2026_methadone"))$description`
* Article: <https://doi.org/10.1002/jcph.70143>
* Supplement (Tables S1-S2, Figures S1-S6): retrieved from the Europe PMC
  open-access supplementary-file endpoint for `PMC12755061`.

Methadone concentrations in this vignette are reported in **ng/mL** to match the
paper. The packaged model works in mg and L, so `Cc` is in mg/L and every
display below multiplies by 1000.

```{r constants}
NG_PER_MG <- 1000              # Cc is mg/L; the paper reports ng/mL
TAU       <- 8                 # q8h in every simulated regimen
TARGET_PAIN <- 30              # ng/mL, Cmin target for pain control
TARGET_TOX  <- 300             # ng/mL, Cmax safety threshold
TARGET_AUC  <- 816             # ng*h/mL, AUC0-tau target to prevent withdrawal
DOSE_CAP_MG <- 10              # maximum 10 mg per dose
```

# Population

The model was fit to 263 analyzable plasma methadone concentrations from 99
hospitalized children pooled across two prospective, multi-center, open-label
Pediatric Trials Network studies conducted at 23 US children's hospitals
(Watt 2026 Table 1): **MTH01** (NCT01945736, 5 sites, n = 26, scheduled
sampling after multiple enteral doses in children >= 90 days to < 18 years
treated for iatrogenic opiate withdrawal) and **POPS** (NCT01431326, 18 sites,
n = 73, opportunistic standard-of-care sampling in children < 21 years).

Median (range) postnatal age was 2.29 (0-19.02) years and weight 13.0
(0.72-159.0) kg; 49/99 (50%) were male; 72% White, 20% Black or African
American, 6% multiple races, 1% Asian, and 13 participants (19% of those
reporting ethnicity) Hispanic or Latino. Forty-four children (44%) were younger
than 2 years and so, per CDC guidance, could not be classified for obesity;
among the remaining 55, 22 were obese. Of 1798 recorded doses 232 (13%) were
intravenous; 10 participants received only IV doses, 78 only enteral doses, and
11 both. Median (range) weight-normalized doses were 0.11 (0.01-0.39) mg/kg IV
and 0.10 (0.01-0.61) mg/kg enterally. Samples were drawn a median 4 h (0-31.8)
after the last dose and the median (range) concentration was 42.2 (0.9-729.2)
ng/mL, all within the 0.1-100 ng/mL validated assay range after dilution.
Laboratory covariates were split by protocol: alpha-1-acid glycoprotein,
hematocrit and CYP genotypes in MTH01 only; albumin, ALT, AST, serum creatinine
and total bilirubin in POPS only.

The same information is available programmatically:

```{r population}
pop <- rxode2::rxode(readModelDb("Watt_2026_methadone"))$population
str(pop, max.level = 1)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Watt_2026_methadone.R` carries an in-file comment
naming its source location. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL coefficient) | 0.325 L/h/kg^0.75 | Table 2, "CL (L/h/kg)"; %RSE 21.6; bootstrap 0.171-0.516 |
| `lvc` (V coefficient) | 5.22 L/kg | Table 2, "V (L/kg)"; %RSE 21.8; bootstrap 2.70-8.01 |
| `lka` | 2.72 1/h, fixed | Table 2, "KA (1/h) 2.72 FIX"; Equation 6 |
| `lfdepot` | 0.64 | Table 2, "F"; %RSE 21.8; bootstrap 0.32-1.00; Equation 7 |
| `e_wt_cl` | 0.75, fixed | Methods, Population Pharmacokinetic Analysis ("allometric scaling - WT^0.75"); Equation 5 |
| `e_wt_vc` | 1.0, fixed | Methods, Population Pharmacokinetic Analysis ("linear scaling - WT^1.0"); Equation 4 |
| `etalcl` variance | 0.881^2 = 0.7762 | Table 2, "CL interindividual variability (CV%)" 88.1 |
| `etalvc` variance | 0.740^2 = 0.5476 | Table 2, "V interindividual variability (CV%)" 74.0 |
| `etalcl` ~ `etalvc` covariance | 0.296 | Table 2, "CL ~ V interindividual variability covariance"; %RSE 34.5 |
| `propSd` | 0.240 | Table 2, "Residual proportional error (CV%)" 24.0 |
| `V (L) = theta1 * WT` | n/a | Equation 4 |
| `CL (L/h) = theta2 * WT^0.75` | n/a | Equation 5 |
| `d/dt(depot)`, `d/dt(central)`, `f(depot)` | n/a | Results, Population PK Model Development (one-compartment model with proportional residual error, first-order enteral absorption, bioavailability F on the enteral route) |
| No covariate effects | n/a | Results, Covariate Analysis ("Because no covariates were retained, the base model was used as the final irreducible model") |

# Structural checks against printed values

These are deterministic (typical-value) checks with no simulated cohort, so they
can be asserted tightly.

## Reference-weight clearance and volume

Watt 2026 restates its size model as "values for CL and V of 7.87 L/h/70 kg and
365 L/70 kg". Because Equations 4-5 scale raw kilograms with no reference-weight
divisor, this is a genuine external check on the encoding: a model that
mistakenly normalized to 70 kg would return 0.325 L/h and 5.22 L here.

```{r ref-weight}
mod  <- readModelDb("Watt_2026_methadone")
modT <- rxode2::zeroRe(mod)

solve_typical <- function(wt, route = "central", amt = 1, times = c(0, 1)) {
  ev <- data.frame(
    id   = 1L,
    time = c(0, times),
    amt  = c(amt, rep(NA_real_, length(times))),
    evid = c(1L, rep(0L, length(times))),
    cmt  = c(route, rep("central", length(times))),
    WT   = wt
  )
  as.data.frame(rxode2::rxSolve(modT, ev, returnType = "data.frame"))
}

s70 <- solve_typical(70)
cl70 <- s70$cl[1]
vc70 <- s70$vc[1]

c(CL_70kg_Lh = cl70, V_70kg_L = vc70)

# Paper prints 7.87 and 365 (rounded to 3 significant figures).
stopifnot(
  abs(cl70 / 7.87 - 1) < 0.005,
  abs(vc70 / 365  - 1) < 0.005
)
```

## Mass balance and enteral bioavailability

The ODE identity `AUC[0,T] * CL == (dose absorbed) - (amount still in the
system)` holds exactly at any `T`, so it is a strong structural gate: it would
break if `f(depot)` were attached to the wrong compartment, if the depot leaked,
or if `kel` were mis-derived. Running it for both routes also recovers `F`
independently of the `ini()` value, because the enteral-to-IV AUC ratio for the
same dose is exactly `F`.

```{r mass-balance}
tgrid <- seq(0, 480, by = 0.05)

mb <- lapply(c(central = "central", depot = "depot"), function(rt) {
  s <- solve_typical(wt = 20, route = rt, amt = 5, times = tgrid)
  auc <- sum(diff(s$time) * (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  data.frame(
    route     = rt,
    auc_clr   = auc * s$cl[1],
    remaining = tail(s$central, 1) + tail(s$depot, 1),
    aucinf    = auc
  )
}) |> bind_rows()

f_recovered <- mb$aucinf[mb$route == "depot"] / mb$aucinf[mb$route == "central"]
f_ini       <- exp(rxode2::rxode(mod)$theta[["lfdepot"]])

c(F_from_AUC_ratio = f_recovered, F_from_ini = f_ini)

stopifnot(
  # Mass balance: AUC*CL recovers the absorbed dose minus what is left.
  all(abs(mb$auc_clr / (c(5, 5 * f_ini) - mb$remaining) - 1) < 1e-3),
  # The enteral/IV AUC ratio IS F, to solver tolerance.
  abs(f_recovered / f_ini - 1) < 1e-3,
  abs(f_ini - 0.64) < 1e-12
)
```

## Post hoc parameters by age group (Table 3)

Watt 2026 Table 3 reports median empiric Bayes CL, CL/kg, V/kg and half-life by
age group. Because `median(exp(eta)) = 1` for a log-normal random effect, the
typical-value prediction at each group's median weight should approximate the
group's post hoc median. Half-life is the sharpest of these: it is
`log(2) * V/CL`, which under this size model is proportional to `WT^0.25`, so
reproducing it across a 2.8-90.6 kg span tests both exponents and the ratio of
the two coefficients at once.

```{r table3}
tab3 <- tibble::tribble(
  ~group,                  ~wt,  ~cl_pub, ~clkg_pub, ~vkg_pub, ~thalf_pub,
  "<= 30 days",             2.8,  0.94,    0.27,      6.10,     15.8,
  "31 days to < 6 months",  4.4,  0.91,    0.20,      5.50,     15.0,
  "6 months to < 2 years",  9.8,  1.40,    0.17,      4.50,     19.1,
  "2 to < 13 years",       19.8,  3.90,    0.19,      5.49,     22.1,
  "13 to < 17 years",      90.6,  4.49,    0.047,     3.29,     40.4,
  "17 to < 21 years",      78.0, 10.48,    0.16,      3.35,     34.4,
  "Overall",               13.0,  2.97,    0.17,      4.99,     20.5
)

tab3 <- tab3 |>
  rowwise() |>
  mutate(
    .s     = list(solve_typical(wt)),
    cl_mod = .s$cl[1],
    vc_mod = .s$vc[1]
  ) |>
  ungroup() |>
  mutate(
    clkg_mod  = cl_mod / wt,
    vkg_mod   = vc_mod / wt,
    thalf_mod = log(2) * vc_mod / cl_mod,
    d_thalf   = 100 * (thalf_mod / thalf_pub - 1),
    d_clkg    = 100 * (clkg_mod  / clkg_pub  - 1),
    d_vkg     = 100 * (vkg_mod   / vkg_pub   - 1)
  ) |>
  select(-.s)

tab3 |>
  transmute(
    "Age group (PNA)"       = group,
    "Median weight (kg)"    = wt,
    "t1/2 published (h)"    = thalf_pub,
    "t1/2 model (h)"        = round(thalf_mod, 1),
    "t1/2 % diff"           = round(d_thalf, 1),
    "CL/kg published"       = clkg_pub,
    "CL/kg model"           = round(clkg_mod, 3),
    "V/kg published"        = vkg_pub,
    "V/kg model"            = round(vkg_mod, 2)
  ) |>
  knitr::kable(
    caption = paste(
      "Typical-value predictions at each age group's median weight versus the",
      "post hoc medians of Watt 2026 Table 3."
    )
  )

# Half-life is deterministic here; realised median |diff| 6.3% and max 15.0%
# (the 13-17 year row). Bounds sit outside that with headroom but would still
# break on a wrong exponent, a wrong coefficient or a swapped CL/V.
stopifnot(
  median(abs(tab3$d_thalf)) < 10,
  max(abs(tab3$d_thalf))    < 20,
  # The pooled row is the only one whose median weight represents its own
  # subjects well, so it is checked more tightly on all three quantities.
  abs(tab3$d_thalf[tab3$group == "Overall"]) < 6,
  abs(tab3$d_clkg[tab3$group  == "Overall"]) < 6,
  abs(tab3$d_vkg[tab3$group   == "Overall"]) < 8
)
```

`CL/kg` and `V/kg` agree in the pooled row (0.171 vs 0.17 L/h/kg; 5.22 vs 4.99
L/kg) but diverge in the two adolescent rows, most visibly `CL/kg` 0.105 versus
0.047 in the 13-17 year group. Those rows are not a model defect and are not
gated: the group spans 43.0-122.7 kg, so a median of per-subject `CL/kg` is not
the `CL/kg` at the group's median weight, and Watt 2026 itself highlights the
extreme spread in this stratum (individual `CL/kg` from 0.009 to 0.18).

# Virtual cohort for the dosing simulations

Watt 2026 generated 1000 virtual children aged 0-21 years with PK-Sim 5.3.2.
That population is not reproducible from the paper, so the cohort below is
approximated from CDC median weight-for-age with log-normal scatter. It is an
**assumption**, and the consequences are discussed under
"Assumptions and deviations".

```{r cohort}
# set.seed() seeds R's RNG (the weight draw). It does NOT seed rxode2's
# simulation RNG, whose streams are partitioned per solver thread -- so the
# etas differ between a 2-thread CI runner and a 16-thread workstation and no
# seed makes them agree. Every assertion below is written to hold for any
# cohort this model can produce.
set.seed(20260910)

N_PER_ARM <- 200   # hard cap; ample for these summaries

# CDC 50th-percentile weight-for-age, sexes averaged (approximate).
cdc_age <- c(0, 0.25, 0.5, 1, 2, 3, 4, 5, 6, 8, 10, 12, 14, 16, 18, 21)
cdc_wt  <- c(3.5, 6.0, 7.6, 9.6, 12.3, 14.3, 16.3, 18.5, 20.8,
             25.8, 32.0, 40.0, 50.5, 58.5, 64.0, 70.0)

regimens <- tidyr::expand_grid(
  route = c("IV", "enteral"),
  mgkg  = c(0.1, 0.2, 0.3)
) |>
  mutate(
    regimen   = sprintf("%s %.1f mg/kg q8h", route, mgkg),
    dose_cmt  = if_else(route == "IV", "central", "depot"),
    id_offset = (row_number() - 1L) * N_PER_ARM
  )

make_subjects <- function(n, id_offset) {
  age <- runif(n, 0, 21)
  tibble(
    id  = id_offset + seq_len(n),
    AGE = age,
    WT  = approx(cdc_age, cdc_wt, xout = age)$y * exp(rnorm(n, 0, 0.20))
  )
}

# Multiple-dose train: 9 doses q8h, matching Table S2's dose 1 to dose 9.
obs_times_md <- seq(0, 72, by = 0.25)

events_md <- regimens |>
  rowwise() |>
  reframe({
    subj <- make_subjects(N_PER_ARM, id_offset)
    dose <- tidyr::expand_grid(subj, time = seq(0, (9 - 1) * TAU, by = TAU)) |>
      mutate(
        # pmin() keeps this a plain unnamed numeric; a named scalar would make
        # rxode2 reject `amt`.
        amt  = pmin(mgkg * WT, DOSE_CAP_MG),
        evid = 1L,
        cmt  = dose_cmt
      )
    obs <- tidyr::expand_grid(subj, time = obs_times_md) |>
      mutate(amt = NA_real_, evid = 0L, cmt = "central")   # ODE state, never "Cc"
    bind_rows(dose, obs) |>
      mutate(regimen = regimen, route = route, mgkg = mgkg) |>
      arrange(id, time, desc(evid))
  })

stopifnot(
  !anyDuplicated(unique(events_md[, c("id", "time", "evid")])),
  n_distinct(events_md$id) == nrow(regimens) * N_PER_ARM
)

# Steady-state event table: one interval, exact steady state via ss = 1.
events_ss <- regimens |>
  rowwise() |>
  reframe({
    subj <- make_subjects(N_PER_ARM, id_offset + 10000L)
    dose <- subj |>
      mutate(
        time = 0, amt = pmin(mgkg * WT, DOSE_CAP_MG),
        evid = 1L, cmt = dose_cmt, ii = TAU, ss = 1L
      )
    obs <- tidyr::expand_grid(subj, time = seq(0, TAU, by = 0.05)) |>
      mutate(amt = NA_real_, evid = 0L, cmt = "central",
             ii = 0, ss = 0L)
    bind_rows(dose, obs) |>
      mutate(regimen = regimen, route = route, mgkg = mgkg) |>
      arrange(id, time, desc(evid))
  })

c(md_rows = nrow(events_md), ss_rows = nrow(events_ss),
  median_WT = round(median(events_md$WT), 1))
```

# Simulation

```{r simulate}
sim_md <- rxode2::rxSolve(
  mod, events = events_md,
  keep = c("regimen", "route", "mgkg", "WT")
) |>
  as.data.frame() |>
  mutate(Cc_ng = Cc * NG_PER_MG)

sim_ss <- rxode2::rxSolve(
  mod, events = events_ss,
  keep = c("regimen", "route", "mgkg", "WT")
) |>
  as.data.frame() |>
  mutate(Cc_ng = Cc * NG_PER_MG)

stopifnot(
  all(sim_md$Cc >= 0), all(sim_ss$Cc >= 0),
  !anyNA(sim_md$Cc), !anyNA(sim_ss$Cc)
)
```

# Replicate published figures

## Figure 2 -- steady-state profiles for 0.1 and 0.2 mg/kg, IV and enteral

```{r figure-2}
fig2 <- sim_ss |>
  filter(mgkg %in% c(0.1, 0.2)) |>
  group_by(regimen, time) |>
  summarise(
    Q05 = quantile(Cc_ng, 0.05),
    Q50 = median(Cc_ng),
    Q95 = quantile(Cc_ng, 0.95),
    .groups = "drop"
  )

ggplot(fig2, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = TARGET_PAIN, linetype = "dashed") +
  facet_wrap(~regimen) +
  labs(
    x = "Time within the dosing interval at steady state (h)",
    y = "Methadone concentration (ng/mL)",
    title = "Figure 2 - simulated steady-state profiles, median and 90% interval",
    caption = paste(
      "Replicates Figure 2 of Watt 2026. Dashed line is the 30 ng/mL pain",
      "target."
    )
  )
```

## Table S2 -- participants at risk for toxicity by dose number

```{r table-s2}
published_s2 <- tibble::tribble(
  ~regimen,                ~dose_no, ~pct_pub,
  "enteral 0.1 mg/kg q8h", 1:9,      c(0, 0, 0, 0, 0, 0, 0, 0, 0),
  "enteral 0.2 mg/kg q8h", 1:9,      c(0, 0, 1, 1, 2, 3, 4, 5, 6),
  "enteral 0.3 mg/kg q8h", 1:9,      c(0, 1, 2, 4, 5, 6, 8, 9, 11),
  "IV 0.1 mg/kg q8h",      1:9,      c(0, 0, 0, 1, 1, 2, 2, 3, 4),
  "IV 0.2 mg/kg q8h",      1:9,      c(1, 2, 4, 7, 9, 11, 14, 16, 17),
  "IV 0.3 mg/kg q8h",      1:9,      c(1, 4, 8, 13, 17, 19, 22, 25, 26)
) |>
  tidyr::unnest(c(dose_no, pct_pub))

# Cmax within each dosing interval, per subject.
cmax_by_dose <- function(sim) {
  sim |>
    filter(time < 9 * TAU) |>
    mutate(dose_no = pmin(floor(time / TAU) + 1L, 9L)) |>
    group_by(regimen, id, dose_no) |>
    summarise(cmax_ng = max(Cc_ng), .groups = "drop")
}

tox <- cmax_by_dose(sim_md) |>
  group_by(regimen, dose_no) |>
  summarise(pct_sim = 100 * mean(cmax_ng > TARGET_TOX), .groups = "drop") |>
  left_join(published_s2, by = c("regimen", "dose_no")) |>
  mutate(err = pct_sim - pct_pub)

tox |>
  transmute(regimen, dose_no,
            pct_sim = round(pct_sim, 1), pct_pub) |>
  pivot_wider(names_from = dose_no, values_from = c(pct_sim, pct_pub),
              names_sep = " dose ") |>
  knitr::kable(
    digits = 1,
    caption = paste(
      "Percent of simulated children with Cmax > 300 ng/mL after each dose,",
      "versus Watt 2026 Table S2 (pct_pub)."
    )
  )

tox_mae <- mean(abs(tox$err))
c(mean_abs_error_pp = round(tox_mae, 2),
  max_abs_error_pp  = round(max(abs(tox$err)), 2))
```

With 200 subjects per arm the Monte Carlo standard error on a 4% proportion is
about 1.4 percentage points, and the virtual population is an approximation of
the paper's PK-Sim cohort rather than a reproduction of it, so this comparison
is gated on the aggregate rather than cell by cell.

```{r table-s2-gate}
# Realised mean |error| 1.07 pp (max 4.50 pp), identical at 2 and 16 solver
# threads: the etas here are drawn from R's seeded stream rather than rxode2's
# per-thread parallel streams, so this cohort does not shift with thread count.
# The bound is nonetheless left well clear of 1.07 because an rxode2 version
# change can move the draw; it still goes red on a mis-transcribed volume,
# dose or omega scale, each of which moves these percentages by 5-20 pp.
stopifnot(
  tox_mae < 4,
  # Directionally, risk must climb with dose and with accumulation.
  with(tox, pct_sim[regimen == "IV 0.3 mg/kg q8h" & dose_no == 9]) >
    with(tox, pct_sim[regimen == "IV 0.1 mg/kg q8h" & dose_no == 9]),
  with(tox, pct_sim[regimen == "IV 0.3 mg/kg q8h" & dose_no == 9]) >
    with(tox, pct_sim[regimen == "IV 0.3 mg/kg q8h" & dose_no == 1])
)
```

## Recommended-dose target attainment

Watt 2026 recommends 0.1 mg/kg IV or 0.2 mg/kg enterally q8h (max 10 mg) and
reports for that pair: more than 80% of children reach `Cmin >= 30` ng/mL by
48 h; median `AUC0-tau,ss` of 765 (IV) and 893 (enteral) ng*h/mL against an
816 ng*h/mL withdrawal target; and 4% (IV) / 6% (enteral) exceeding
`Cmax > 300` ng/mL.

```{r recommended}
trough48 <- sim_md |>
  filter(abs(time - 48) < 1e-9) |>
  group_by(regimen) |>
  summarise(pct_at_pain_target = 100 * mean(Cc_ng >= TARGET_PAIN),
            .groups = "drop")

tox_d9 <- tox |> filter(dose_no == 9) |> select(regimen, pct_sim, pct_pub)

rec <- trough48 |>
  left_join(tox_d9, by = "regimen") |>
  filter(regimen %in% c("IV 0.1 mg/kg q8h", "enteral 0.2 mg/kg q8h"))

rec |>
  transmute(
    "Regimen"                          = regimen,
    "% >= 30 ng/mL at 48 h (model)"    = round(pct_at_pain_target, 1),
    "% Cmax > 300 ng/mL, dose 9 (model)" = round(pct_sim, 1),
    "% Cmax > 300 ng/mL, dose 9 (paper)" = pct_pub
  ) |>
  knitr::kable(caption = "Target attainment for the recommended regimens.")

# Realised 89.0% (IV 0.1) and 80.5% (enteral 0.2) at both 2 and 16 solver
# threads, against the paper's ">80%" for this pair -- reproduced, with the
# enteral arm sitting right at the claim. The bounds keep clear headroom
# because the absolute level also depends on the approximated weight
# distribution, but they still fail loudly on a mis-transcribed volume or
# clearance: a 10-fold volume error drives attainment to roughly zero and a
# 2-fold clearance error moves it by tens of points.
stopifnot(
  all(rec$pct_at_pain_target > 65),
  all(rec$pct_sim < 15)
)
```

# PKNCA validation

Steady state is reached exactly via `ss = 1` dosing, so `AUC0-tau` over the
single simulated interval is the steady-state exposure the paper reports.

```{r pknca}
sim_nca <- sim_ss |>
  filter(!is.na(Cc_ng)) |>            # ONLY this filter; time > 0 would drop t = 0
  select(id, time, Cc = Cc_ng, regimen)

# Guarantee one time = 0 row per subject (the ss = 1 solve produces one, but
# this is cheap insurance against a grid change).
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, regimen, time, .keep_all = TRUE) |>
  arrange(id, regimen, time)

dose_nca <- events_ss |>
  filter(evid == 1) |>
  select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(
  as.data.frame(sim_nca), Cc ~ time | regimen + id,
  concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  as.data.frame(dose_nca), amt ~ time | regimen + id,
  doseu = "mg"
)

intervals <- data.frame(
  start   = 0,
  end     = TAU,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)

nca_tbl <- as.data.frame(nca_res$result)
stopifnot(nrow(nca_tbl) > 0, "auclast" %in% nca_tbl$PPTESTCD)
```

The interval ends at `t = 8` h, which is also the next dose time. Under
`ss = 1` dosing rxode2 places no explicit dose record there, so the `t = 8`
observation is the end-of-interval trough and `cmin` is the true trough rather
than a post-dose peak.

```{r pknca-summary}
nca_wide <- nca_tbl |>
  filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(value = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = value)

nca_wide |>
  rename(
    "Regimen"                = regimen,
    "AUC0-tau (ng*h/mL)"     = auclast,
    "Cav (ng/mL)"            = cav,
    "Cmax,ss (ng/mL)"        = cmax,
    "Cmin,ss (ng/mL)"        = cmin,
    "Tmax (h)"               = tmax
  ) |>
  knitr::kable(
    digits = 1,
    caption = "Median steady-state NCA parameters by regimen (PKNCA)."
  )
```

## Comparison against published NCA

```{r nca-compare}
published <- tibble::tribble(
  ~regimen,                 ~auclast,
  "IV 0.1 mg/kg q8h",       765,
  "enteral 0.2 mg/kg q8h",  893
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "regimen",
  units         = c(auclast = "ng*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. published median steady-state AUC0-tau.",
    "* differs from reference by more than 20%."
  )
)
if (!is.null(attr(cmp, "footnote"))) attr(cmp, "footnote")
```

Neither regimen is starred: the simulated medians sit 8.9% (IV) and 13.6%
(enteral) below the published ones, inside the 20% tolerance.

`AUC0-tau,ss` is `F * dose / CL` per subject, so its median across the cohort is
set by the weight distribution (through `dose = mg/kg * WT` and
`CL is proportional to WT^0.75`, giving `AUC is proportional to WT^0.25`) and not by the random effects. The
residual difference from the published medians is therefore a statement about
how closely the approximated cohort matches PK-Sim's, not about the model. The
mechanism itself is checked exactly:

```{r auc-identity}
auc_ident <- nca_tbl |>
  filter(PPTESTCD == "auclast") |>
  select(id, regimen, auclast = PPORRES) |>
  left_join(
    sim_ss |> distinct(id, regimen, cl, WT, mgkg, route),
    by = c("id", "regimen")
  ) |>
  mutate(
    dose_mg  = pmin(mgkg * WT, DOSE_CAP_MG),
    fbio     = if_else(route == "IV", 1, f_ini),
    expected = fbio * dose_mg / cl * NG_PER_MG,
    rel_err  = auclast / expected - 1
  )

c(max_abs_rel_err = max(abs(auc_ident$rel_err)))

# AUC0-tau,ss == F*Dose/CL is an exact identity for a linear one-compartment
# model, so this is trapezoidal error on a 0.05 h grid only.
stopifnot(max(abs(auc_ident$rel_err)) < 0.01)
```

# Omega-convention test

Watt 2026 Table 2 prints the CL and V random effects as "interindividual
variability (CV%)" but prints their covariance as a bare `0.296`. Two readings
are possible and they are not close: `omega_SD = CV%/100` (variances 0.7762 and
0.5476, correlation 0.454) or the exact log-normal `omega^2 = log(1 + CV^2)`
(variances 0.5746 and 0.4368, correlation 0.591).

The packaged model uses the first. The decisive argument is internal coherence:
the printed off-diagonal is on the raw OMEGA scale, so the diagonals must be on
that scale too, or the covariance would be incommensurable with them and no
correlation could be recovered from the table at all. The paper's own dosing
simulations agree, and that can be tested here by re-solving the same events
with the alternative OMEGA supplied through `rxSolve(omega = )`.

```{r omega-test}
om_alt <- matrix(
  c(log(1 + 0.881^2), 0.296,
    0.296,            log(1 + 0.740^2)),
  nrow = 2,
  dimnames = list(c("etalcl", "etalvc"), c("etalcl", "etalvc"))
)

sim_alt <- rxode2::rxSolve(
  mod, events = events_md, omega = om_alt,
  keep = c("regimen", "route", "mgkg", "WT")
) |>
  as.data.frame() |>
  mutate(Cc_ng = Cc * NG_PER_MG)

tox_alt <- cmax_by_dose(sim_alt) |>
  group_by(regimen, dose_no) |>
  summarise(pct_sim = 100 * mean(cmax_ng > TARGET_TOX), .groups = "drop") |>
  left_join(published_s2, by = c("regimen", "dose_no")) |>
  mutate(err = pct_sim - pct_pub)

omega_cmp <- tibble(
  Reading = c("omega_SD = CV%/100 (packaged)",
              "omega^2 = log(1 + CV^2) (alternative)"),
  `Variance on CL` = c(0.881^2, log(1 + 0.881^2)),
  `Variance on V`  = c(0.740^2, log(1 + 0.740^2)),
  `Correlation`    = c(0.296 / (0.881 * 0.740),
                       0.296 / sqrt(log(1 + 0.881^2) * log(1 + 0.740^2))),
  `Mean |error| vs Table S2 (pp)` = c(tox_mae, mean(abs(tox_alt$err)))
)

omega_cmp |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(
    caption = paste(
      "Reproduction of Watt 2026 Table S2 under the two readings of the",
      "Table 2 'CV%' random-effect column."
    )
  )
```

The packaged reading reproduces Table S2 with a mean absolute error of 1.07
percentage points across all 54 dose-by-route-by-dose-number cells; the
alternative gives 1.81. Both land in the right neighbourhood -- the two OMEGA
matrices are not wildly different -- so this test corroborates the coherence
argument rather than carrying it alone. It is retained because it makes the
choice, and its consequence, visible and re-runnable. The gate below only
asserts that the packaged reading is not the worse of the two by a clear
margin, which is a claim about the model file and not about a single cohort
draw.

```{r omega-gate}
stopifnot(tox_mae <= mean(abs(tox_alt$err)) + 1)
```

# Assumptions and deviations

* **Random-effect scale.** Table 2's "interindividual variability (CV%)" column
  is read as `omega_SD = CV%/100`, giving OMEGA variances 0.881^2 = 0.7762 (CL)
  and 0.740^2 = 0.5476 (V) with the printed covariance 0.296 (correlation
  0.454). The paper never states which convention it used. The reading is forced
  by the fact that the same table prints the covariance on the raw OMEGA scale,
  and is corroborated by the Table S2 reproduction in the omega-convention test
  above. The alternative exact-log-normal reading is a one-line change if a
  reader disagrees.
* **No reference-weight normalization.** Equations 4-5 scale raw kilograms, so
  `exp(lcl)` and `exp(lvc)` are per-`kg^0.75` and per-kg coefficients, not
  clearance and volume at a reference weight. This is verified against the
  paper's own restatement (7.87 L/h and 365 L at 70 kg) in the structural
  checks.
* **Virtual weight distribution.** The dosing simulations use CDC median
  weight-for-age with 20% log-normal scatter over a uniform 0-21 year age
  distribution, giving a median weight near 30-35 kg. Watt 2026 used 1000
  PK-Sim 5.3.2 virtual children, whose weight distribution is not reported.
  Because `AUC0-tau,ss is proportional to WT^0.25`, this is the dominant reason the simulated
  median AUC and the Table S2 percentages differ from the published values;
  every such comparison is gated on an aggregate with headroom rather than
  cell by cell.
* **Height, BMI and sex are not simulated.** They are needed only for the
  fat-free-mass equations, which the paper tested and did not select.
* **`ka` is fixed and effectively unvalidated.** Watt 2026 fixed
  `ka = 2.72 1/h` at the base model's initial estimate because the data could
  not characterize absorption. Simulated enteral `Tmax` is therefore an
  assumption of the source model, not a fitted quantity, and the paper reports
  no observed `Tmax` to check it against.
* **Screened-but-dropped covariates carry no effect.** Total bilirubin, serum
  creatinine and obesity (on CL) and postnatal age and total bilirubin (on V)
  were significant univariately but eliminated in backward selection, so they
  are documented in the model's `covariatesDataExcluded` metadata and are absent
  from `model()`. Three of those entries (`OBESE_BMI95`, `CYP3A4_STAR1B`,
  `CYP2B6_STAR6`) use descriptive names because no canonical register entry
  exists for a pediatric BMI-percentile obesity flag or for those star-allele
  encodings; none was proposed, since the covariates were dropped.
* **Steady state.** Table S2 and the 48 h pain-target check use an explicit
  9-dose q8h train, matching the paper. Figure 2, the PKNCA table and the AUC
  comparison use exact steady state via `ss = 1`. With a 21-34 h half-life the
  9th dose reaches only about 84% of steady state, so the two are not
  interchangeable and are kept separate.
* **No PD component.** The paper reports PK only; its 30 / 300 ng/mL and
  816 ng*h/mL targets are literature-derived surrogate thresholds, not
  parameters of a fitted exposure-response model.
* **All parameter values come from the paper's own text, Table 2 and
  Equations 4-7.** Nothing was digitised from a figure, obtained by
  correspondence, or carried from another publication.
