---
title: "Vancomycin (Wei 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Wei 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 5, dpi = 100
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Wei_2022_vancomycin"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Wei_2022_vancomycin"))$description`
* Article (open access): <https://doi.org/10.3389/fphar.2022.1005791>

Wei 2022 is a retrospective population PK analysis of routine therapeutic drug
monitoring (TDM) data for intravenous vancomycin at Beijing Tiantan Hospital,
fitted in Phoenix NLME 8.3 (FOCE-ELS). The data were dominated by steady-state
troughs, so the volume of distribution could not be estimated and was fixed at
60.2 L, the value of an earlier one-compartment model in Chinese neurosurgical
adults (Jing et al., J Clin Pharm Ther 2020;45:88-96). Clearance depends on
CKD-EPI estimated glomerular filtration rate, body weight and concomitant
mannitol.

# Population

The model was built on 895 serum concentrations from 560 Chinese adults, 497 of
them postoperative neurosurgical patients (Table 1). Age was 52.41 +/- 15.11
years (18-89), body weight 69.74 +/- 13.05 kg (37.5-130), and 370 patients were
male and 190 female. CKD-EPI eGFR was 112.74 +/- 30.91 mL/min (3.52-244.48);
68.2% of patients had augmented renal clearance during monitoring and 2.72% had
severe renal insufficiency, most of them on renal replacement therapy.
Concomitant medications were meropenem (71.32%), mannitol (60.32%) and
diuretics (15.95%). Vancomycin was given as 0.5 or 1 g every 6 to 24 hours;
the infusion time was not recorded and a 1 h infusion was assumed throughout.

```{r population}
str(readModelDb("Wei_2022_vancomycin")()$population, max.level = 1)
```

# Source trace

Every value below is recorded as an in-file comment next to the corresponding
`ini()` entry in `inst/modeldb/specificDrugs/Wei_2022_vancomycin.R`.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | log(7.98) | Table 3, final model: CL = 7.98 L/h (RSE 1.90%); Eq. (5) |
| `lvc` | fixed(log(60.2)) | Table 3: V = 60.2 L with no RSE or CI; Eq. (6); Methods 2.2.1 (fixed from Jing 2020) |
| `e_crcl_cl` | 0.80 | Table 3 'eGFR on CL' = 0.80 (RSE 4.30%); exponent in Eq. (5) |
| `e_wt_cl` | 0.30 | Table 3 'BW on CL' = 0.30 (RSE 20.19%); exponent in Eq. (5) |
| `e_conmed_mannitol_cl` | 0.13 | Table 3 'Mannitol on CL' = 0.13 (RSE 17.85%); Eq. (5) `e^A`, A = 0.13 with mannitol |
| `etalcl` | 0.044983 = log(1 + 0.2145^2) | Table 3 'IIV CL (CV%)' = 21.45; exponential IIV, Eq. (1) |
| `propSd` | 0.25 | Table 3 sigma1 (multiplicative, CV) = 0.25 (RSE 6.45%) |
| `addSd` | 1.51 mg/L | Table 3 sigma2 (additive) = 1.51 mg/L (RSE 25.83%) |
| Reference eGFR 115.2 mL/min | n/a | Denominator in Eq. (5); Methods 2.2.2: continuous covariates centered at the median |
| Reference weight 70 kg | n/a | Denominator in Eq. (5) |
| eGFR equation | n/a | Eq. (4), CKD-EPI with Scr in mg/dL (see Assumptions) |
| One compartment, first-order elimination | n/a | Methods 2.2.1 |
| Combined residual error | n/a | Eq. (2): `Cobs = Cpred + eps * sqrt(1 + (Cpred * sigma1 / sigma2)^2)`, i.e. SD = `sqrt(sigma2^2 + (sigma1 * Cpred)^2)`, which is nlmixr2's `combined2` |
| 1 h IV infusion | n/a | Methods 2.1 and 2.2.4 |

# Deterministic checks against the paper's own arithmetic

```{r arithmetic}
mod <- readModelDb("Wei_2022_vancomycin")
ui <- rxode2::rxode(mod)
th <- ui$theta
cl_typ <- exp(th[["lcl"]])

checks <- tibble::tibble(
  Quantity = c(
    "Typical CL per kg at 70 kg (L/h/kg)",
    "CL ratio with vs without mannitol"
  ),
  Computed = c(cl_typ / 70, exp(th[["e_conmed_mannitol_cl"]])),
  Published = c(0.11, 1.14),
  `Source (Wei 2022)` = c(
    "Discussion: 'the typical value of vancomycin CL (0.11 L/h/kg)'",
    "Discussion: 'Co-medicated with mannitol resulted in a 14% increase in vancomycin CL'"
  )
)
stopifnot(
  abs(cl_typ / 70 - 0.11) < 0.005,
  abs(exp(th[["e_conmed_mannitol_cl"]]) - 1.14) < 0.005
)
knitr::kable(checks, digits = 3)
```

# Table 4: the published dose-selection simulations

Table 4 of Wei 2022 lists, for six typical patients (eGFR 90, 45 or 15 mL/min;
body weight 85 or 65 kg) with and without mannitol, the q12h doses (1 h
infusion) that give a steady-state AUC24 of about 400 and of about 600 h*mg/L,
together with the simulated AUC24 and its 5th-95th percentile interval from
1,000 Monte Carlo replicates.

```{r table4}
table4 <- tibble::tribble(
  ~egfr, ~wt, ~mannitol, ~target, ~dose, ~auc, ~lo, ~hi,
  90, 85, 1, 400, 1600, 401.47, 293.80, 574.70,
  90, 65, 1, 400, 1450, 400.46, 278.89, 563.07,
  45, 85, 1, 400, 900, 402.28, 283.58, 564.73,
  45, 65, 1, 400, 825, 398.85, 275.21, 560.15,
  15, 85, 1, 400, 375, 399.33, 279.04, 541.50,
  15, 65, 1, 400, 350, 399.18, 186.48, 546.08,
  90, 85, 1, 600, 2350, 600.50, 414.31, 844.34,
  90, 65, 1, 600, 2150, 596.63, 425.29, 864.34,
  45, 85, 1, 600, 1350, 600.04, 412.90, 858.12,
  45, 65, 1, 600, 1250, 607.36, 425.11, 848.61,
  15, 85, 1, 600, 560, 596.34, 416.70, 808.63,
  15, 65, 1, 600, 525, 602.02, 422.52, 809.92,
  90, 85, 0, 400, 1400, 402.19, 275.46, 577.10,
  90, 65, 0, 400, 1300, 405.62, 290.79, 567.37,
  45, 85, 0, 400, 800, 401.47, 286.27, 587.27,
  45, 65, 0, 400, 750, 406.38, 283.88, 582.65,
  15, 85, 0, 400, 330, 394.54, 277.72, 528.21,
  15, 65, 0, 400, 310, 398.11, 281.88, 528.18,
  90, 85, 0, 600, 2050, 593.86, 409.72, 834.99,
  90, 65, 0, 600, 1900, 598.44, 424.16, 877.44,
  45, 85, 0, 600, 1200, 607.06, 429.45, 854.37,
  45, 65, 0, 600, 1100, 600.75, 421.69, 847.62,
  15, 85, 0, 600, 500, 593.28, 421.08, 792.91,
  15, 65, 0, 600, 465, 597.17, 422.82, 792.28
) |>
  mutate(
    treatment = sprintf(
      "eGFR %d, %d kg, mannitol %s, %d mg q12h",
      egfr, wt, ifelse(mannitol == 1, "yes", "no"), dose
    )
  )
stopifnot(nrow(table4) == 24, !anyDuplicated(table4$treatment))
```

## Typical-value AUC24

At steady state the AUC over one 24 h day is the daily dose divided by
clearance, so the typical-patient AUC24 follows from the encoded parameters in
closed form. For a log-normal clearance the cohort median of AUC24 equals this
typical value, so it can be compared directly with the Table 4 point values
(which carry the paper's own Monte Carlo noise of roughly 1%).

```{r table4-typical}
typ <- table4 |>
  mutate(
    cl = cl_typ * (egfr / 115.2)^th[["e_crcl_cl"]] * (wt / 70)^th[["e_wt_cl"]] *
      exp(th[["e_conmed_mannitol_cl"]] * mannitol),
    auc_typ = 2 * dose / cl,
    pct_diff = 100 * (auc_typ - auc) / auc
  )
# Deterministic on the model side: all 24 cells reproduce within the
# paper's Monte Carlo precision.
stopifnot(all(abs(typ$pct_diff) < 3))
typ |>
  select(egfr, wt, mannitol, dose, auc, auc_typ, pct_diff) |>
  rename(
    "eGFR (mL/min)" = egfr, "Weight (kg)" = wt, "Mannitol" = mannitol,
    "Dose q12h (mg)" = dose, "Table 4 AUC24" = auc,
    "Typical-value AUC24" = auc_typ, "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 1, caption = "Typical-value AUC24 (h*mg/L) against Table 4 of Wei 2022.")
```

## Scale of the clearance IIV

Table 3 reports the clearance IIV only as 'CV%' (21.45). For a log-normal AUC24
the ratio of the 95th to the 5th percentile is `exp(2 * 1.645 * omega)`, so each
Table 4 interval implies an SD of log-clearance. The encoded
`omega = sqrt(log(1 + 0.2145^2))` is set against the alternative readings of
the reported number.

```{r iiv-scale}
implied_sd <- log(table4$hi / table4$lo) / (2 * qnorm(0.95))
omega_enc <- sqrt(rxode2::rxode(mod)$omega[1, 1])
iiv_tab <- tibble::tibble(
  Reading = c(
    "Median SD implied by the 24 Table 4 intervals",
    "Encoded: CV% -> omega^2 = log(1 + CV^2)",
    "Alternative: 21.45% read as omega^2 x 100"
  ),
  `SD of log CL` = c(median(implied_sd), omega_enc, sqrt(0.2145))
)
stopifnot(abs(median(implied_sd) - omega_enc) < 0.02)
knitr::kable(iiv_tab, digits = 3)
```

The intervals imply an SD of about 0.21, which agrees with the CV reading and
rules out a variance reading (which would give intervals more than twice as
wide).

## Stochastic replication with PKNCA

Each Table 4 cell is simulated as a separate arm of 200 virtual patients with
the clearance IIV, given q12h 1 h infusions for 21 days. At eGFR 15 mL/min the
half-life is about 27 h, so the run is long enough for every arm to reach steady
state. The steady-state AUC24 is computed with PKNCA over the 480-504 h dosing
day and compared with the published median.

```{r sim-table4}
n_per_arm <- 200
obs_times <- c(0, seq(480, 504, by = 0.5))
dose_times <- seq(0, 492, by = 12)

arm_events <- function(i) {
  arm <- table4[i, ]
  ids <- (i - 1) * n_per_arm + seq_len(n_per_arm)
  doses <- expand.grid(time = dose_times, id = ids)
  doses$evid <- 1L
  doses$amt <- arm$dose
  doses$rate <- arm$dose / 1
  obs <- expand.grid(time = obs_times, id = ids)
  obs$evid <- 0L
  obs$amt <- 0
  obs$rate <- 0
  bind_rows(doses, obs) |>
    mutate(
      cmt = "central",
      CRCL = arm$egfr,
      WT = arm$wt,
      CONMED_MANNITOL = arm$mannitol,
      treatment = arm$treatment
    )
}
events <- bind_rows(lapply(seq_len(nrow(table4)), arm_events)) |>
  arrange(id, time, desc(evid)) |>
  select(id, time, evid, amt, rate, cmt, CRCL, WT, CONMED_MANNITOL, treatment)

rxode2::rxSetSeed(20221005)
sim <- rxode2::rxSolve(
  rxode2::rxode(mod), select(events, -treatment),
  returnType = "data.frame", addCov = TRUE
) |>
  left_join(distinct(events, id, treatment), by = "id")
stopifnot(!anyNA(sim$Cc), all(sim$Cc >= 0))
```

```{r pknca}
conc <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(conc, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(start = 480, end = 504, auclast = TRUE, cmax = TRUE, cmin = TRUE)
nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

auc_sub <- as.data.frame(nca$result) |>
  filter(PPTESTCD == "auclast") |>
  select(id, treatment, auc_nca = PPORRES)
```

The trapezoidal AUC is first checked against the closed-form `2 * dose / CL`
for the same simulated subjects; both sides use identical parameter draws, so
the difference is pure numerical error.

```{r nca-closed-form}
cl_sub <- sim |>
  distinct(id, cl) |>
  left_join(distinct(events, id, treatment), by = "id") |>
  left_join(select(table4, treatment, dose), by = "treatment") |>
  mutate(auc_cf = 2 * dose / cl)
nca_cf <- inner_join(auc_sub, cl_sub, by = c("id", "treatment"))
stopifnot(
  nrow(nca_cf) == nrow(table4) * n_per_arm,
  all(abs(100 * (nca_cf$auc_nca - nca_cf$auc_cf) / nca_cf$auc_cf) < 1)
)
```

```{r nca-comparison}
reference <- table4 |>
  select(treatment, auclast = auc)
cmp <- ncaComparisonTable(
  nca, reference,
  by = "treatment", params = "auclast",
  units = c(auclast = "h*mg/L"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Median simulated steady-state AUC24 (480-504 h) against Table 4 of Wei 2022.")
```

```{r nca-assert}
arm_sum <- auc_sub |>
  group_by(treatment) |>
  summarise(
    sim_med = median(auc_nca),
    sim_lo = quantile(auc_nca, 0.05),
    sim_hi = quantile(auc_nca, 0.95),
    .groups = "drop"
  ) |>
  inner_join(table4, by = "treatment") |>
  mutate(pct_med = 100 * (sim_med - auc) / auc)
stopifnot(
  nrow(arm_sum) == 24,
  # Centre: a mis-transcribed clearance or exponent shifts every arm.
  abs(median(arm_sum$pct_med)) < 3,
  # Envelope over arms, robust to which 200-subject draw lands in the tails.
  quantile(abs(arm_sum$pct_med), 0.9) < 8
)
```

```{r fig-table4, fig.height = 6}
arm_sum |>
  mutate(
    patient = sprintf("eGFR %d, %d kg", egfr, wt),
    mannitol = ifelse(mannitol == 1, "Mannitol", "No mannitol"),
    target = paste("Target", target)
  ) |>
  ggplot(aes(x = patient)) +
  geom_pointrange(aes(y = auc, ymin = lo, ymax = hi, colour = "Wei 2022 Table 4"),
    position = position_nudge(x = -0.15)
  ) +
  geom_pointrange(aes(y = sim_med, ymin = sim_lo, ymax = sim_hi, colour = "nlmixr2lib simulation"),
    position = position_nudge(x = 0.15)
  ) +
  facet_grid(target ~ mannitol, scales = "free_y") +
  labs(
    x = NULL, y = "Steady-state AUC24 (h*mg/L), median and 5th-95th percentile",
    colour = NULL
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "bottom")
```

Replicates Table 4 of Wei 2022. The 90% intervals match the published ones in
width; the one visible outlier is the published lower bound of 186.48 for the
15 mL/min, 65 kg, mannitol, 350 mg row (see Assumptions and deviations).

# Steady-state troughs in a Table 1 cohort

The monitored concentrations were mostly steady-state troughs from a mix of
0.5 g and 1 g regimens given every 6 to 24 hours (mean 14.20 +/- 7.36 mg/L, 20.4%
within 15-20 mg/L); the paper does not tabulate how many patients received each
regimen. As an illustration, 200 virtual patients with covariates drawn from the
Table 1 summaries receive 1 g every 12, 8 or 6 hours (1 h infusions) for 10 days,
and the steady-state trough (residual error included) is summarised per regimen.
This section is descriptive, not a gate.

```{r cohort}
set.seed(2022)
n_cohort <- 200
cohort <- tibble::tibble(
  pid = seq_len(n_cohort),
  WT = pmin(pmax(rnorm(n_cohort, 69.74, 13.05), 37.5), 130),
  CRCL = pmin(pmax(rnorm(n_cohort, 112.74, 30.91), 3.52), 244.48),
  CONMED_MANNITOL = rbinom(n_cohort, 1, 0.6032)
)
regimens <- tibble::tibble(regimen = c("1 g q12h", "1 g q8h", "1 g q6h"), tau = c(12, 8, 6))
last_dose <- 216

ev_c <- bind_rows(lapply(seq_len(nrow(regimens)), function(r) {
  tau <- regimens$tau[r]
  arm <- mutate(cohort, id = (r - 1) * n_cohort + pid, regimen = regimens$regimen[r])
  bind_rows(
    tidyr::expand_grid(arm, time = seq(0, last_dose, by = tau)) |>
      mutate(evid = 1L, amt = 1000, rate = 1000),
    tidyr::expand_grid(arm, time = c(0, last_dose + seq(0, tau, by = 0.5))) |>
      mutate(evid = 0L, amt = 0, rate = 0)
  )
})) |>
  mutate(cmt = "central") |>
  arrange(id, time, desc(evid))

rxode2::rxSetSeed(20221006)
sim_c <- rxode2::rxSolve(
  rxode2::rxode(mod),
  select(ev_c, id, time, evid, amt, rate, cmt, CRCL, WT, CONMED_MANNITOL),
  returnType = "data.frame"
) |>
  left_join(distinct(ev_c, id, regimen), by = "id") |>
  left_join(regimens, by = "regimen")
trough <- sim_c |> filter(time == last_dose + tau)
stopifnot(nrow(trough) == nrow(regimens) * n_cohort, !anyNA(trough$sim))
trough |>
  group_by(regimen) |>
  summarise(
    `Median trough (mg/L)` = median(sim),
    `5th-95th percentile (mg/L)` = sprintf("%.1f-%.1f", quantile(sim, 0.05), quantile(sim, 0.95)),
    `% in 15-20 mg/L` = 100 * mean(sim >= 15 & sim <= 20),
    .groups = "drop"
  ) |>
  arrange(match(regimen, regimens$regimen)) |>
  rename(Regimen = regimen) |>
  knitr::kable(digits = 1)
```

```{r fig-cohort}
sim_c |>
  filter(time >= last_dose) |>
  group_by(regimen, time) |>
  summarise(
    p05 = quantile(sim, 0.05), p50 = median(sim), p95 = quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  mutate(regimen = factor(regimen, levels = regimens$regimen)) |>
  ggplot(aes(time - last_dose, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = c(15, 20), linetype = "dashed") +
  facet_wrap(~regimen, scales = "free_x") +
  labs(
    x = "Time after the last dose (h)", y = "Vancomycin (mg/L)",
    title = "Steady state in a Table 1 cohort: median and 90% prediction interval"
  ) +
  theme_bw()
```

With the typical clearance of 7.98 L/h and the fixed 60.2 L volume, the
half-life is about 5.2 h, so a 1 g q12h trough is low (around 4-5 mg/L) even
though the corresponding AUC24 is about 240 h*mg/L. The paper's own Table 4
reaches the same conclusion through AUC24: at eGFR 90 mL/min it needs
1.3-1.6 g q12h for an AUC24 of 400. The observed mean of 14.2 mg/L is matched by
the q6h arm here and exceeds the q8h arm. The Supplementary Appendix S2 VPC, which
pools all regimens and doses, shows a predicted median near 14 mg/L about
10 h after the dose.

# Assumptions and deviations

1. **IIV scale.** Table 3 reports the clearance IIV only as '21.45 CV%'. It is
   encoded as `omega^2 = log(1 + 0.2145^2) = 0.044983`; the 24 Table 4 AUC24
   intervals imply an SD of log-clearance of about 0.21, matching this reading
   (the variance-times-100 alternative would give SD 0.46).
2. **eGFR definition.** Eq. (4) prints the CKD-EPI creatinine equation with a
   144 multiplier for both sexes, labels the result mL/min and applies no
   1.73 m^2 term. Users should compute `CRCL` exactly as printed (Scr in mg/dL)
   for consistency with the fitted exponent and the 115.2 mL/min reference.
   The paper reports Table 1 eGFR as a mean (112.74); 115.2 is the median used
   for centering (Methods 2.2.2).
3. **Volume fixed from another study.** V = 60.2 L comes from Jing 2020 and was
   not estimated; it is encoded as `fixed()`. No IIV on V was reported, so none
   is included.
4. **Infusion duration.** The paper assumed 1 h infusions because actual
   infusion times were unavailable; the simulations here do the same.
5. **Mannitol flag.** The paper does not say whether mannitol was coded per
   sample or per patient. `CONMED_MANNITOL` can be supplied either way;
   the Table 4 simulations hold it constant within a patient.
6. **Probable typo in Table 4.** The lower 5% bound for the eGFR 15 mL/min,
   65 kg, mannitol, 350 mg q12h row is printed as 186.48, whereas every other
   400-target row lies between 275 and 294 and the implied SD for that row
   alone (0.33) is far from the other 23 (0.19-0.22). It is most likely
   286.48. The value is shown as printed and does not enter any assertion
   other than the median implied SD, which is insensitive to it.
7. **Trough predictions depend on the borrowed volume.** AUC24 depends only on
   clearance, which the trough-dominated data estimate well (RSE 1.9%). Trough
   and peak levels also depend on V, which was fixed at the Jing 2020 value
   rather than estimated, so they carry that study's uncertainty as well.
8. **Unretained covariates.** Age, height, BMI, serum creatinine, Cockcroft-Gault
   CLcr, renal replacement therapy, sex, meropenem and diuretics were screened
   (Table 2) but not retained; no final-model estimates exist for them.
