---
title: "Alfaxalone in Lewis and Sprague-Dawley rats (White 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alfaxalone in Lewis and Sprague-Dawley rats (White 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

White 2022 fitted two one-compartment population PK models for intravenous
alfaxalone in rats, on two nested datasets. Both are packaged:

* `White_2022_alfaxalone_rat_pop1` -- "population 1", 28 rats (16 Lewis plus
  12 Sprague-Dawley rats from White 2017). Sex and strain on clearance, strain
  on volume. The authors used this model to design the adjusted female
  Sprague-Dawley infusion (Figure 6).
* `White_2022_alfaxalone_rat_pop2` -- "population 2", the final 52-rat model
  (population 1 plus 24 new Sprague-Dawley rats). Log-centred body weight and
  sex on clearance, strain on volume.

* Citation: `r rxode2::rxode(readModelDb("White_2022_alfaxalone_rat_pop2"))$reference`
* Article: <https://doi.org/10.1002/prp2.1031>

```{r models}
mod1 <- readModelDb("White_2022_alfaxalone_rat_pop1")
mod2 <- readModelDb("White_2022_alfaxalone_rat_pop2")

# Typical-value solve. Every random effect in both models is held at zero
# (the variances are unpublished), so rxode2's "multi-subject simulation
# without 'omega'" warning is expected and muffled; any other warning passes.
solve_typical <- function(mod, events, ...) {
  withCallingHandlers(
    rxSolve(zeroRe(mod), events, returnType = "data.frame", ...),
    warning = function(w) {
      if (grepl("without 'omega'", conditionMessage(w), fixed = TRUE)) {
        invokeRestart("muffleWarning")
      }
    }
  )
}
```

# Population

Sixteen adult Lewis rats (8-12 weeks; 9 males 308 +/- 49 g, 7 females
222 +/- 9 g) and 24 adult Sprague-Dawley rats (9-12 weeks; 8 males
422 +/- 41 g, 16 females 304 +/- 15 g) were studied (Methods 2.2). Population 1
added the 12 Sprague-Dawley rats (6 male, 6 female) of White 2017 to the 16
Lewis rats; population 2 added the 24 new Sprague-Dawley rats to population 1
(Figure 1). All rats were cannulated under isoflurane, then received alfaxalone
(Alfaxan) as a 1.67 mg/kg/min loading infusion for 2.5 min followed by a
constant-rate infusion (CRI): Lewis rats 0.75 mg/kg/min for 60 min then
0.52 mg/kg/min; new Sprague-Dawley males 0.75 mg/kg/min throughout; new
Sprague-Dawley females, on the regimen designed with the population 1 model,
0.52 mg/kg/min for 60 min then 0.42 mg/kg/min (Methods 2.5). Plasma alfaxalone
was measured by LC-MS/MS (LLOQ 200 ng/mL).

The same information is available programmatically via
`readModelDb("White_2022_alfaxalone_rat_pop2")()$population`.

# Source trace

Both NLME fits used total (not per-kg) dose in Phoenix NLME 8.3, with
exponential random effects on every parameter (diagonal omega) and a "mixed
ratio" residual error (Methods 2.8.1). The paper reports the typical values and
covariate coefficients only.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure | one-compartment, IV infusion | Sections 3.4 and 3.9 |
| pop1 `lcl` | log(0.0252 L/min) | Section 3.4: CL_TV = 25.2 mL/min |
| pop1 `lvc` | log(0.57 L) | Section 3.4: Vd_TV = 0.57 L |
| pop1 `e_sexf_cl` | -0.841 | Eq 1 |
| pop1 `e_strain_sd_cl` | 0.478 | Eq 1 |
| pop1 `e_strain_sd_vc` | -0.0237 | Eq 2 |
| pop2 `lcl` | log(0.0352 L/min) | Section 3.9: CL_TV = 35.2 mL/min |
| pop2 `lvc` | log(0.51 L) | Section 3.9: Vd_TV = 0.51 L |
| pop2 `e_wt_cl` | 3.64 | Eq 3: (1 + LCBW * 3.64) |
| pop2 LCBW | log10(WT / 0.317 kg) | Methods 2.8.1 names it; base and centre recovered from Figure S7 source data (see below) |
| pop2 `e_sexf_cl` | -0.43 | Eq 3 |
| pop2 `e_strain_sd_vc` | -0.692 | Eq 4; Table 4 SD Vd = 0.26 L |
| `etalcl`, `etalvc` (both) | 0 (not reported) | Sections 3.4, 3.9: diagonal omega, values not printed |
| `addSd`, `propSd` (both) | 0 (not reported) | Methods 2.8.1: mixed ratio error, values not printed |
| Covariate coding | SEXF: male 0, female 1; STRAIN_SD: Lewis 0, SD 1 | Methods 2.8.1 |

## The log-centred body weight (LCBW)

The paper calls the clearance covariate "log of centralized body weight" but
prints neither the logarithm base nor the centring weight. The supplement's
Figure S7 (eta CL versus the weight covariate) is a vector graphic that
carries the plotted value of each rat. All 39 distinct LCBW values map to
whole-gram body weights when read as `log10(WT / 317 g)` (mean deviation from
an integer below 1e-6 g), and only under that reading; the resulting weights
span 208-488 g, matching the Methods 2.2 cohort. The same form reproduces the
Table 4 Lewis typical clearances at the Methods 2.2 group mean weights, as the
next section shows.

# Typical values against Tables 3 and 4

Tables 3 and 4 list the adjusted typical Vd and CL of each sex-strain group
(per animal). For population 2, footnote a states the weight term was applied
at each group's average weight; the Lewis means (308 and 222 g) are printed in
Methods 2.2, while the Sprague-Dawley group means in population 2 include the
unprinted White 2017 weights.

```{r typical-tables}
groups <- tibble::tribble(
  ~group,             ~SEXF, ~STRAIN_SD, ~WT,
  "Male Lewis",       0,     0,          0.308,
  "Female Lewis",     1,     0,          0.222,
  "Male SD",          0,     1,          NA,
  "Female SD",        1,     1,          NA
)
pub <- tibble::tribble(
  ~group,         ~vd1, ~cl1, ~vd2, ~cl2,
  "Male Lewis",   0.57, 25.2, 0.51, 33.5,
  "Female Lewis", 0.57, 10.9, 0.51, 10.1,
  "Male SD",      0.56, 40.6, 0.26, 48.6,
  "Female SD",    0.56, 17.5, 0.26, 20.6
)

typ_pop1 <- function(SEXF, STRAIN_SD) {
  c(
    cl = 25.2 * exp(-0.841 * SEXF + 0.478 * STRAIN_SD),
    vd = 0.57 * exp(-0.0237 * STRAIN_SD)
  )
}
typ_pop2 <- function(SEXF, STRAIN_SD, WT) {
  c(
    cl = 35.2 * (1 + 3.64 * log10(WT / 0.317)) * exp(-0.43 * SEXF),
    vd = 0.51 * exp(-0.692 * STRAIN_SD)
  )
}

# Solve the packaged models (not the formulas above) for the typical CL and
# Vd: one observation per group, all random effects zero.
ev_typ <- groups |>
  mutate(id = row_number(), time = 0, evid = 0, amt = 0, cmt = "central",
         WT = ifelse(is.na(WT), 0.317, WT))
s1 <- solve_typical(mod1, ev_typ, keep = "group")
s2 <- solve_typical(mod2, ev_typ, keep = "group")

tab <- pub |>
  mutate(
    sim_vd1 = s1$vc, sim_cl1 = 1000 * s1$cl,
    sim_vd2 = s2$vc
  )
tab |>
  select(group, vd1, sim_vd1, cl1, sim_cl1, vd2, sim_vd2) |>
  rename(
    "Group" = group,
    "Pop 1 Vd, Table 3 (L)" = vd1, "Pop 1 Vd, model (L)" = sim_vd1,
    "Pop 1 CL, Table 3 (mL/min)" = cl1, "Pop 1 CL, model (mL/min)" = sim_cl1,
    "Pop 2 Vd, Table 4 (L)" = vd2, "Pop 2 Vd, model (L)" = sim_vd2
  ) |>
  knitr::kable(digits = 3, caption = "Typical Vd and CL against Tables 3 and 4.")

# Population 2 clearance at the printed Lewis group mean weights.
lewis_cl2 <- 1000 * s2$cl[1:2]
knitr::kable(
  data.frame(
    "Group" = c("Male Lewis (308 g)", "Female Lewis (222 g)"),
    "Table 4 CL (mL/min)" = c(33.5, 10.1),
    "Model CL (mL/min)" = lewis_cl2,
    check.names = FALSE
  ),
  digits = 2,
  caption = "Population 2 typical CL at the Methods 2.2 Lewis mean weights."
)

stopifnot(
  # Rounding of the printed coefficients is the only source of difference.
  all(abs(tab$sim_vd1 / tab$vd1 - 1) < 0.01),
  all(abs(tab$sim_cl1 / tab$cl1 - 1) < 0.01),
  all(abs(tab$sim_vd2 / tab$vd2 - 1) < 0.02),
  all(abs(lewis_cl2 / c(33.5, 10.1) - 1) < 0.01),
  # The packaged models agree with the printed equations.
  abs(1000 * s1$cl[4] - typ_pop1(1, 1)[["cl"]]) < 1e-6,
  abs(1000 * s2$cl[2] - typ_pop2(1, 0, 0.222)[["cl"]]) < 1e-6
)
```

The Table 4 Sprague-Dawley clearances (48.6 and 20.6 mL/min) correspond to
group mean weights of `r round(317 * 10^((48.6 / 35.2 - 1) / 3.64))` g and
`r round(317 * 10^((20.6 / (35.2 * exp(-0.43)) - 1) / 3.64))` g, consistent
with pooling the new Sprague-Dawley rats (422 and 304 g) with the lighter
White 2017 animals.

# Replicate Figure 6 (population 1 design simulation)

Figure 6 shows the median and 5th-95th percentile simulated profiles that the
authors generated to design the female Sprague-Dawley regimen. The median
curves are typical-value profiles of the population 1 model: their rise and
post-infusion decay rates match the population 1 elimination rate constants
(0.073 /min male, 0.031 /min female) and not those of population 2. The figure
does not state the body weight it simulated; the plateaus (6.21 and
6.08 ug/mL) imply 336 g and 253 g at the population 1 Sprague-Dawley
clearances, and those weights are used below. Infusions stop at about 258 min
(male) and 238.5 min (female), read off the figure.

```{r cohort-helpers}
# One animal's infusion schedule: `starts` are the times each rate begins
# (the first is the 1.67 mg/kg/min loading infusion at time 0), `rates` the
# mg/kg/min rates, and `t_stop` the end of the last infusion.
make_rat <- function(id, WT, SEXF, STRAIN_SD, rates, starts, t_stop,
                     obs_times, group) {
  dur <- diff(c(starts, t_stop))
  doses <- data.frame(
    id = id, time = starts, evid = 1L, cmt = "central",
    rate = rates * WT, amt = rates * WT * dur
  )
  obs <- data.frame(
    id = id, time = obs_times, evid = 0L, cmt = "central",
    rate = 0, amt = 0
  )
  bind_rows(doses, obs) |>
    mutate(WT = WT, SEXF = SEXF, STRAIN_SD = STRAIN_SD, group = group) |>
    arrange(time, desc(evid))
}
```

```{r figure-6}
obs_t <- seq(0, 350, by = 1)
ev_f6 <- bind_rows(
  make_rat(1, 0.336, 0, 1, c(1.67, 0.75), c(0, 2.5), 258, obs_t, "Male SD"),
  make_rat(2, 0.253, 1, 1, c(1.67, 0.52, 0.42), c(0, 2.5, 60), 238.5, obs_t,
           "Female SD")
)
f6 <- solve_typical(mod1, ev_f6, keep = "group")

# Population 2 at the new Sprague-Dawley cohort mean weights (422 g, 304 g)
ev_f6b <- bind_rows(
  make_rat(1, 0.422, 0, 1, c(1.67, 0.75), c(0, 2.5), 258, obs_t, "Male SD"),
  make_rat(2, 0.304, 1, 1, c(1.67, 0.52, 0.42), c(0, 2.5, 60), 238.5, obs_t,
           "Female SD")
)
f6b <- solve_typical(mod2, ev_f6b, keep = "group")

bind_rows(
  f6 |> mutate(model = "Population 1 (design weights)"),
  f6b |> mutate(model = "Population 2 (cohort mean weights)")
) |>
  ggplot(aes(time, Cc, colour = model)) +
  geom_line() +
  facet_wrap(~group) +
  labs(x = "Time (min)", y = "Alfaxalone plasma concentration (ug/mL)",
       colour = NULL,
       caption = "Replicates the median curves of Figure 6 of White 2022.") +
  theme(legend.position = "bottom")

# Median curve digitised by the maintainers from Figure 6B (female SD).
fig6b <- tibble::tribble(
  ~time, ~Cc_pub,
  30,    5.18,
  40,    5.80,
  70,    6.49,
  100,   6.23,
  200,   6.08,
  250,   4.22,
  260,   3.07,
  280,   1.62,
  300,   0.85
)
fig6a <- tibble::tribble(
  ~time, ~Cc_pub,
  30,    5.70,
  40,    5.94,
  50,    6.11,
  100,   6.21,
  250,   6.21
)
cmp6 <- bind_rows(
  fig6a |> mutate(group = "Male SD"),
  fig6b |> mutate(group = "Female SD")
) |>
  left_join(f6 |> select(group, time, Cc), by = c("group", "time")) |>
  mutate(pct_diff = 100 * (Cc / Cc_pub - 1)) |>
  relocate(group)
cmp6 |>
  rename("Group" = group, "Time (min)" = time,
         "Figure 6 median (ug/mL)" = Cc_pub, "Model (ug/mL)" = Cc,
         "Difference (%)" = pct_diff) |>
  knitr::kable(digits = 2, caption = "Population 1 typical profile against the digitised Figure 6 medians.")

stopifnot(
  # Digitisation reads the curve to about 0.05 ug/mL; the decay points test
  # the population 1 volume, the plateaus the clearance.
  max(abs(cmp6$pct_diff[cmp6$time <= 260])) < 3,
  max(abs(cmp6$pct_diff)) < 6
)
```

The population 2 model, evaluated at the actual cohort mean weights, gives
the same plateaus (`r round(max(f6b$Cc[f6b$group == "Male SD" & f6b$time == 200]), 2)`
and `r round(max(f6b$Cc[f6b$group == "Female SD" & f6b$time == 200]), 2)`
ug/mL at 200 min) as population 1 did at the design weights, but rises and
decays faster because of its smaller Sprague-Dawley volume (see
"Population 2 Sprague-Dawley volume" below).

# Lewis-rat simulation and PKNCA

A virtual Lewis cohort (100 rats per sex, weights drawn from the Methods 2.2
means and SDs) receives the Lewis regimen: the loading infusion, 0.75 mg/kg/min
to 60 min, then 0.52 mg/kg/min to 240 min, followed by a washout. Both models
are simulated at typical values (the published random-effect variances are
not available).

```{r lewis-cohort}
set.seed(2022)
n_per <- 100
obs_lewis <- c(0, 1, 2.5, 5, 10, 20, 30, 45, 60, 90, 120, 180, 240,
               242, 245, 250, 255, 260, 270, 280, 300, 330, 360, 420)
wt_m <- pmin(pmax(rnorm(n_per, 0.308, 0.049), 0.22), 0.45)
wt_f <- pmin(pmax(rnorm(n_per, 0.222, 0.009), 0.20), 0.25)
ev_lewis <- bind_rows(
  lapply(seq_len(n_per), function(i) {
    make_rat(i, wt_m[i], 0, 0, c(1.67, 0.75, 0.52), c(0, 2.5, 60), 240,
             obs_lewis, "Male Lewis")
  }),
  lapply(seq_len(n_per), function(i) {
    make_rat(n_per + i, wt_f[i], 1, 0, c(1.67, 0.75, 0.52), c(0, 2.5, 60), 240,
             obs_lewis, "Female Lewis")
  })
)
stopifnot(!anyDuplicated(unique(ev_lewis[, c("id", "time", "evid")])))

sim_lewis <- bind_rows(
  solve_typical(mod1, ev_lewis, keep = c("group", "WT")) |>
    mutate(model = "pop1"),
  solve_typical(mod2, ev_lewis, keep = c("group", "WT")) |>
    mutate(model = "pop2")
) |>
  mutate(treatment = paste(model, group, sep = ": "))
```

```{r figure-3}
sim_lewis |>
  group_by(model, group, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(time, Q50, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~model) +
  labs(x = "Time (min)", y = "Alfaxalone plasma concentration (ug/mL)",
       colour = NULL, fill = NULL,
       caption = paste("Typical-value profiles across the weight distribution;",
                       "compare with the per-rat fits of Figure 3 of White 2022.")) +
  theme(legend.position = "bottom")
```

Table 2 reports the per-kg clearance and half-life of each Lewis rat from the
authors' separate two-stage deterministic analysis (individual fits to
per-kg doses). PKNCA computes the same quantities from the simulated
profiles; dose is supplied per kg so that `cl.obs` is per kg.

```{r pknca}
sim_nca <- sim_lewis |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(treatment, id, time)

dose_df <- bind_rows(
  ev_lewis |> mutate(model = "pop1"),
  ev_lewis |> mutate(model = "pop2")
) |>
  filter(evid == 1) |>
  mutate(
    treatment = paste(model, group, sep = ": "),
    amt = amt / WT,            # mg/kg
    duration = amt / (rate / WT)
  ) |>
  select(id, time, amt, duration, treatment)

conc_obj <- PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCAdose(dose_df, amt ~ time | treatment + id, duration = "duration")
intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, aucinf.obs = TRUE, cl.obs = TRUE, half.life = TRUE
)
nca_res <- pk.nca(PKNCAdata(conc_obj, dose_obj, intervals = intervals))

# cl.obs is in (mg/kg) / (ug/mL * min) = L/min/kg; report mL/min/kg.
nca_res$result <- nca_res$result |>
  mutate(PPORRES = ifelse(PPTESTCD == "cl.obs", 1000 * PPORRES, PPORRES))

published <- tibble::tribble(
  ~treatment,               ~cl.obs, ~half.life,
  "pop1: Male Lewis",       98.3,    13.5,
  "pop1: Female Lewis",     36.8,    64.7,
  "pop2: Male Lewis",       98.3,    13.5,
  "pop2: Female Lewis",     36.8,    64.7
)
cmp <- ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = c("cl.obs", "half.life"),
  units = c(cl.obs = "mL/min/kg", half.life = "min"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste("Simulated (median) vs Table 2 two-stage estimates for Lewis rats.",
                  "* differs from reference by >20%.")
)

# PKNCA must recover the model's own clearance (a check on the dose/duration
# bookkeeping, not on the paper): AUC to infinity of a linear model.
cl_chk <- nca_res$result |>
  filter(PPTESTCD == "cl.obs") |>
  left_join(sim_lewis |> distinct(id, treatment, WT, cl), by = c("id", "treatment")) |>
  mutate(rel = PPORRES / (1000 * cl / WT) - 1)
stopifnot(quantile(abs(cl_chk$rel), 0.9, na.rm = TRUE) < 0.02)
```

The `CL/F` rows are the intravenous clearance per kg. For male Lewis rats the
population 1 clearance and half-life lie within 20% of the Table 2 means;
population 2 gives a higher clearance (+12%) and, with its smaller Lewis
volume (0.51 L), a half-life about 25% shorter than Table 2. For female Lewis
rats the half-life rows, and the population 1 clearance, are starred: Table 2's female
half-life (64.7 min) and clearance (36.8 mL/min/kg) describe slower
elimination than either NLME model's typical female (44-49 mL/min/kg,
half-life about 36 min). Table 2 is a different analysis (per-rat fits on per-kg doses, with a
separately estimated volume for every rat; female Vdss 3.0 L/kg against the
NLME 0.57 L, i.e. about 2.6 L/kg at 222 g), and the NLME post-hoc female Lewis
clearances (40.8 mL/min/kg in Table 3, 40.4 in Table 4) also sit below the
typical values. The difference is not a transcription error in the packaged
models, whose typical values reproduce Tables 3 and 4.

# Population 2 Sprague-Dawley volume

Table 4 also lists post-hoc (empirical Bayes) mean Vd per kg. For the Lewis
groups these agree with the typical Vd divided by the group mean weight; for
the Sprague-Dawley groups they are 3-4 times larger than the printed typical
Vd of 0.26 L would give.

```{r posthoc-vd}
vd_tab <- tibble::tribble(
  ~group,         ~WT_g, ~vd_typ, ~vd_posthoc,
  "Male Lewis",   308,   0.51,    1.58,
  "Female Lewis", 222,   0.51,    3.03,
  "Male SD",      403,   0.26,    2.18,
  "Female SD",    297,   0.26,    3.84
) |>
  mutate(vd_typ_per_kg = vd_typ / (WT_g / 1000),
         ratio = vd_posthoc / vd_typ_per_kg)
vd_tab |>
  rename("Group" = group, "Mean weight (g)" = WT_g,
         "Typical Vd, Table 4 (L)" = vd_typ,
         "Typical Vd per kg (L/kg)" = vd_typ_per_kg,
         "Post-hoc mean Vd, Table 4 (L/kg)" = vd_posthoc,
         "Post-hoc / typical" = ratio) |>
  knitr::kable(digits = 2, caption = "Population 2 typical versus post-hoc volume (Sprague-Dawley weights back-solved from Table 4 CL).")
```

Post-hoc means of a log-normal random effect sit close to the typical value,
so a three- to four-fold excess suggests the Sprague-Dawley volume of the fit
is nearer 1 L than 0.26 L, i.e. that the printed strain coefficient may have
the wrong sign (0.51 * exp(+0.692) = 1.02 L). Equation 4 and the Table 4
typical value agree with each other, however, and the paper offers no
independent check of the population 2 volume (Figure 6 is a population 1
simulation). The packaged model therefore follows the printed equation; users
simulating Sprague-Dawley rats with population 2 should be aware that the
volume, and so the loading-phase and washout kinetics, may be understated.
Steady-state concentrations depend only on clearance and are unaffected.

# Assumptions and deviations

- **Random effects and residual error are not reported.** The paper states
  that random effects were estimated on all parameters (diagonal omega) and
  that a Phoenix mixed-ratio residual error was used, but prints no variance
  or error magnitude in the text, tables or supplement. `etalcl`, `etalvc`,
  `addSd` and `propSd` are held at zero; simulations are typical-value. The
  Figure S3/S6 box plots also show a random effect on the mixed-ratio
  parameter itself, which is not represented.
- **Residual error form.** Phoenix's mixed ratio model,
  `C + CEps * (1 + C * CMixRatio)`, is an additive SD plus a proportional SD
  summed linearly, encoded as `add() + prop() + combined1()`.
- **LCBW base and centre.** Recovered as `log10(WT / 0.317 kg)` from the
  Figure S7 source data and confirmed against the Table 4 Lewis clearances
  (see Source trace). The linear form `1 + 3.64 * LCBW` would give a negative
  clearance below about 168 g, well below the studied range (208-488 g).
- **Figure 6 weights.** The body weights behind the Figure 6 simulation are
  not printed; 336 g (male) and 253 g (female) are back-solved from the
  plateaus and the population 1 clearances. Infusion stop times are read off
  the figure.
- **Strain covariate.** `STRAIN_SD` takes the Lewis rat as its reference
  (Methods 2.8.1: Lewis = 0, SD = 1); the register entry records that the
  comparator strain is set per paper.
- **Population 2 Sprague-Dawley volume.** Encoded as printed (Eq 4,
  exp(-0.692)); see the section above for the evidence of a possible sign
  error.
- **Not packaged.** The two-stage deterministic one-compartment fits of the
  individual Lewis rats (Table 2) are individual analyses rather than a
  population model; they are used here only as a comparator. The blood
  pressure observations (Figures 2, 4, 5, 7) were not modelled by the authors.
- No erratum or correction notice for this article was found (checked
  2026-10-10).
