---
title: "Spatial CNS and brain-tumor PBPK of abemaciclib (Wickramasinghe 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial CNS and brain-tumor PBPK of abemaciclib (Wickramasinghe 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
ui <- rxode2::rxode(readModelDb("Wickramasinghe_2025_abemaciclib_cns_pbpk"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1002/psp4.70026>
* PubMed Central: <https://pmc.ncbi.nlm.nih.gov/articles/PMC12072228/>

`r ui$description`

The 9-CNS model resolves the human central nervous system into nine
well-stirred concentration compartments. Their canonical nlmixr2lib names and
the symbols the paper uses for them are:

| nlmixr2lib state | Paper symbol | Region |
|---|---|---|
| `brain_vascular` | `Cbb` | Brain blood (cerebral microvasculature) |
| `brain_csf_adjacent` | `Cbm1` | Brain parenchyma within ~2 mm of the CSF tract |
| `brain_deep` | `Cbm2` | Deep brain parenchyma (> 2 mm from the CSF tract) |
| `tumor_rim` | `CT1` | Infiltrative, non-enhancing tumor rim (pH 6.8) |
| `tumor_bulk` | `CT2` | Bulky, contrast-enhancing tumor (pH 6.5) |
| `tumor_core` | `CT3` | Tumor core (pH 6.2) |
| `brain_csf_ventricular` | `Cvcsf` | Ventricular CSF (pH 7.3) |
| `brain_csf_sas_cranial` | `Cccsf` | Cranial subarachnoid CSF (pH 7.3) |
| `brain_csf_sas_spinal` | `Cscsf` | Spinal subarachnoid CSF |

Every state holds a **total** drug concentration. Passive permeability moves
only unbound *and* unionized drug, so its driving concentration is
`lam<region> * fu<region> * C<region>`; transporter-mediated clearances act on
unbound drug whether ionized or not, so theirs is `fu<region> * C<region>`. The
model reports both: the vignette uses the total concentrations `Cbb` ... `Cscsf`
(the app's "Total" output tabs) and the model also returns `Cbbu` ... `Cccsfu`
(the "Unbound" tabs).

# Population

```{r population}
pop <- ui$population
```

The driving plasma profile is the population-mean abemaciclib
concentration-time profile determined in glioblastoma patients, attributed by
this tutorial to its companion development paper (Li 2025, *Clin Pharmacol
Ther*). Table S2 of the tutorial additionally carries paired observed
abemaciclib concentrations for **`r pop$n_subjects` patients** who underwent
tumor resection: non-enhancing tumor, contrast-enhancing tumor and CSF samples
taken 120-130 h after the start of twice-daily oral dosing. No demographics
(age, weight, sex, race) are reported in this paper, and the dose amount is not
stated. The system-specific parameters of Table 1 describe a reference adult
brain (1.2 L parenchyma, 0.15 L total CSF) carrying a 35 mL contrast-enhancing
tumor.

```{r population-print}
str(pop)
```

# Source trace

Every `ini()` entry carries an in-file comment pointing at its source location
in `inst/modeldb/specificDrugs/Wickramasinghe_2025_abemaciclib_cns_pbpk.R`. All
85 values are **fixed**; none is estimated in this paper.

| Model component | Source location |
|---|---|
| The nine differential equations | Data S1 (Supporting Information), one display equation per compartment |
| Compartment volumes `Vbb` ... `Vscsf` | Table 1 (descriptions and rounded values); Table S2 column `Sim1` (values used) |
| Blood and CSF flows `Qbrain`, `QCsink`, `QSsink`, `QSin*`, `QSout*`, `Qgly*` | Table 1; Table S2 `Sim1` |
| Paravascular / convective bulk flows `Qbulk*` | Table 1; Table S2 `Sim1` |
| Passive permeability clearances `PSB*`, `PST*`, `PSV`, `PSC`, `PSE*` | Table 2 (definitions), Equation 1 (`PSB = Papp,A-B * SA / lambda`); Table S2 `Sim1` |
| Transporter clearances `CLeff*`, `CLup*` | Table 2 (definitions and the RAF scaling footnote); Table S2 `Sim1` |
| Brain metabolism `CLmet1`, `CLmet2` | Table 2; Table S2 `Sim1` (both 0 for abemaciclib) |
| Unbound fractions `fubb` ... `fuccsf` | Table 2; Table S2 `Sim1` |
| Unionization fractions `lambb` ... `lamccsf` | Table 2 and its Henderson-Hasselbalch footnote; Table S2 `Sim1` |
| Interindividual variability (20% CV on every parameter) | Table S2 column `IIV1`; lognormal form identified from Table S3 (see Errata) |
| Driving plasma profile `CP_ABEMACICLIB_UM` | Table S2 columns `Time1` / `Plasma1` |
| Observed tumor and CSF concentrations | Table S2 columns `Ob*_T` / `Tumor*` |

The full per-parameter table, generated from the packaged model object:

```{r source-trace-table}
ui$iniDf |>
  dplyr::filter(!is.na(.data$ntheta)) |>
  dplyr::transmute(
    Parameter = .data$name,
    Value = signif(ifelse(grepl("^l", .data$name), exp(.data$est), .data$est), 5),
    Scale = ifelse(grepl("^l", .data$name), "log", "linear"),
    Fixed = .data$fix,
    Description = .data$label
  ) |>
  knitr::kable(caption = "All 85 fixed parameters of the 9-CNS model (Tables 1, 2 and S2).")
```

# The input function

The 9-CNS model is driven by the systemic plasma concentration supplied as the
time-varying covariate `CP_ABEMACICLIB_UM`. Plasma is a pure forcing function:
CNS uptake never depletes it, so the coupling is one-way. The 1010-point
population-mean profile of Table S2 is shipped with the package.

```{r input-function}
plasma <- read.csv(
  system.file("extdata", "Wickramasinghe_2025_abemaciclib_plasma_input.csv",
              package = "nlmixr2lib")
)
observed <- read.csv(
  system.file("extdata", "Wickramasinghe_2025_abemaciclib_observed_cns.csv",
              package = "nlmixr2lib")
)
stopifnot(nrow(plasma) == 1010L, nrow(observed) == 110L)

c(n_points = nrow(plasma),
  t_max = max(plasma$time),
  Cmax = max(plasma$CP_ABEMACICLIB_UM),
  t_at_Cmax = plasma$time[which.max(plasma$CP_ABEMACICLIB_UM)])
```

```{r input-function-plot}
ggplot(plasma, aes(time, CP_ABEMACICLIB_UM)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Plasma abemaciclib (umol/L)",
       title = "Input function: population-mean plasma profile",
       caption = "Table S2 columns Time1 / Plasma1 of Wickramasinghe 2025.")
```

The covariate must be interpolated **linearly**, as the app does. The output
grid of every simulation below is therefore the published 1010-point grid
itself: coarsening it to 0.7 h changes the predicted CNS concentrations by up
to 3%, so the profile is used at full resolution rather than thinned.

# Typical-value simulation: replicating Figure 2

Figure 2 of the paper is a screenshot of the app's "Concentration log plots
(Total)" tab: the mean total-concentration profile in each compartment, with
the observed tumor and CSF concentrations overlaid. It is a typical-value
prediction, so the random effects are zeroed.

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(readModelDb("Wickramasinghe_2025_abemaciclib_cns_pbpk"))

events_typical <- data.frame(
  id = 1L,
  time = plasma$time,
  amt = NA_real_,
  evid = 0L,
  cmt = "brain_vascular",
  CP_ABEMACICLIB_UM = plasma$CP_ABEMACICLIB_UM
)

sim_typical <- rxode2::rxSolve(
  mod_typical, events_typical,
  covsInterpolation = "linear", returnType = "data.frame"
)
if (is.null(sim_typical$id)) sim_typical$id <- 1L
nrow(sim_typical)
```

```{r figure-2, fig.height = 8}
# Replicates Figure 2 of Wickramasinghe 2025: mean total abemaciclib
# concentration-time profile in each of the nine CNS compartments, on a
# log10 axis, with the observed tumor and CSF concentrations overlaid.
panel_levels <- c(
  Cbb = "Cbb (brain blood)",
  Cbm1 = "Cbm1 (parenchyma adj. CSF)",
  Cbm2 = "Cbm2 (deep parenchyma)",
  CT1 = "CT1 (tumor rim)",
  CT2 = "CT2 (bulk tumor)",
  CT3 = "CT3 (tumor core)",
  Cvcsf = "Cvcsf (ventricular CSF)",
  Cccsf = "Cccsf (cranial SAS CSF)",
  Cscsf = "Cscsf (spinal SAS CSF)"
)

pred_long <- sim_typical |>
  dplyr::select(time, dplyr::all_of(names(panel_levels))) |>
  tidyr::pivot_longer(-time, names_to = "state", values_to = "conc") |>
  dplyr::mutate(panel = factor(panel_levels[state], levels = panel_levels))

# The paper overlays the tumor observations on Cbm2/CT1/CT2/CT3 and the CSF
# observations on Cvcsf/Cccsf/Cscsf (Figure 2 caption).
obs_long <- dplyr::bind_rows(
  observed |> dplyr::filter(matrix != "CSF") |>
    tidyr::crossing(state = c("Cbm2", "CT1", "CT2", "CT3")),
  observed |> dplyr::filter(matrix == "CSF") |>
    tidyr::crossing(state = c("Cvcsf", "Cccsf", "Cscsf"))
) |>
  dplyr::mutate(panel = factor(panel_levels[state], levels = panel_levels))

ggplot(pred_long, aes(time, conc)) +
  geom_line(colour = "blue") +
  geom_point(data = obs_long, aes(time, conc, colour = matrix, shape = matrix),
             alpha = 0.7, size = 1.3) +
  facet_wrap(~panel, ncol = 2, scales = "free_y") +
  scale_y_log10() +
  scale_colour_manual(values = c("CSF" = "darkgreen",
                                 "enhancing tumor" = "red",
                                 "non-enhancing tumor" = "black")) +
  labs(x = "Time (h)", y = "Total abemaciclib (umol/L)", colour = NULL, shape = NULL,
       title = "Figure 2 - simulated mean total concentration by CNS compartment",
       caption = "Replicates Figure 2 of Wickramasinghe 2025.") +
  theme(legend.position = "bottom")
```

## Quantitative check against Figure 2

Figure 2 is a log10-axis screenshot, so it can only be read to about a
half-decade. The check below pins each panel's value at the steady-state peak
(t = 124.61 h, the time of the plasma Cmax) against the value read off the
published figure.

```{r figure-2-check}
t_peak <- plasma$time[which.max(plasma$CP_ABEMACICLIB_UM)]
i_peak <- which.min(abs(sim_typical$time - t_peak))

fig2_read <- c(Cbb = 0.42, Cbm1 = 1.6, Cbm2 = 1.5, CT1 = 2.5, CT2 = 4,
               CT3 = 11, Cvcsf = 0.03, Cccsf = 0.03, Cscsf = NA)

fig2_cmp <- tibble::tibble(
  Compartment = names(panel_levels),
  Simulated = signif(unlist(sim_typical[i_peak, names(panel_levels)]), 4),
  `Read off Figure 2` = unname(fig2_read[names(panel_levels)])
) |>
  dplyr::mutate(`Ratio` = signif(Simulated / `Read off Figure 2`, 3))

knitr::kable(
  fig2_cmp,
  caption = paste0("Total concentration at the steady-state plasma peak (t = ",
                   round(t_peak, 2), " h). Cscsf is not shown in Figure 2.")
)

# A log10 screenshot can only be read to about a half-decade, but the model
# agrees with every readable panel far more tightly than that (worst case
# 0.033 in log10, i.e. 8%). Assert the accuracy actually achieved, so this is
# a regression guard rather than a formality.
ok <- !is.na(fig2_cmp$`Read off Figure 2`)
stopifnot(sum(ok) == 8L, all(abs(log10(fig2_cmp$Ratio[ok])) < 0.05))
```

The brain-blood compartment equilibrates with plasma almost instantly
(`Qbrain` = 39 L/h into `Vbb` = 0.063 L), so `Cbb` tracks the input function
exactly. That is the model behaving as published, not a degenerate case:

```{r cbb-identity}
cbb_ratio <- sim_typical$Cbb[i_peak] / max(plasma$CP_ABEMACICLIB_UM)
stopifnot(abs(cbb_ratio - 1) < 0.005)
signif(cbb_ratio, 5)
```

## The spatial gradient the model exists to predict

The whole point of resolving nine compartments rather than one is that the
predicted exposure is not uniform across the CNS. At the steady-state peak the
model predicts a monotone rise from CSF through parenchyma into the
progressively more disrupted tumor interior:

```{r spatial-gradient}
gradient <- sim_typical[i_peak, c("Cccsf", "Cbm2", "CT1", "CT2", "CT3")]
stopifnot(all(diff(unlist(gradient)) > 0))
tibble::tibble(
  Region = c("Cranial SAS CSF", "Deep parenchyma", "Tumor rim",
             "Bulk tumor", "Tumor core"),
  `Total (umol/L)` = signif(unlist(gradient), 3),
  `Fold vs CSF` = signif(unlist(gradient) / gradient$Cccsf, 3)
) |>
  knitr::kable(caption = "Predicted spatial gradient at the steady-state peak.")
```

# Comparison against the observed patient data

The observed concentrations in Table S2 are the paper's own validation data.
The model prediction is compared at each patient's actual sampling time, using
the compartment the paper pairs with each matrix: the non-enhancing tumor with
`tumor_rim` (`CT1`), the contrast-enhancing tumor with `tumor_bulk` (`CT2`) and
CSF with `brain_csf_sas_cranial` (`Cccsf`).

```{r observed-comparison}
state_for <- c("non-enhancing tumor" = "CT1",
               "enhancing tumor" = "CT2",
               "CSF" = "Cccsf")

pred_at <- function(state, times) {
  stats::approx(sim_typical$time, sim_typical[[state]], xout = times, rule = 2)$y
}

obs_cmp <- observed |>
  dplyr::mutate(
    state = unname(state_for[matrix]),
    predicted = mapply(pred_at, state, time),
    ratio = conc / predicted
  )
stopifnot(nrow(obs_cmp) == 110L, !anyNA(obs_cmp$predicted))

obs_cmp |>
  dplyr::group_by(Matrix = matrix) |>
  dplyr::summarise(
    N = dplyr::n(),
    `Paired compartment` = dplyr::first(state),
    `Observed median` = signif(stats::median(conc), 3),
    `Predicted (typical)` = signif(stats::median(predicted), 3),
    `Median obs/pred` = signif(stats::median(ratio), 3),
    `5th-95th pct of obs/pred` = paste(
      signif(stats::quantile(ratio, 0.05), 2), "-",
      signif(stats::quantile(ratio, 0.95), 2)
    ),
    .groups = "drop"
  ) |>
  knitr::kable(caption = paste(
    "Observed abemaciclib concentrations (Table S2, 39 glioblastoma patients)",
    "against the typical-value prediction at each patient's sampling time."
  ))
```

The typical-value prediction sits inside the observed spread for all three
matrices, which is the claim Figure 2 makes graphically. The observed
between-patient spread is far wider than the template's uniform 20% CV
produces (see Errata) - the paper does not fit that spread, and neither does
this vignette.

```{r observed-comparison-assert}
med_ratio <- obs_cmp |>
  dplyr::group_by(matrix) |>
  dplyr::summarise(m = stats::median(ratio), .groups = "drop")
# The typical-value prediction must land within 3-fold of the observed median
# in every matrix -- a real constraint given the predictions span 0.03 to 11
# umol/L across compartments.
stopifnot(nrow(med_ratio) == 3L, all(med_ratio$m > 1 / 3), all(med_ratio$m < 3))
signif(stats::setNames(med_ratio$m, med_ratio$matrix), 3)
```

# Virtual cohort with interindividual variability

Table S2 assigns the same 20% CV to every one of the 85 parameters and the app
uses it to generate individual profiles (its "individual concentration - time
profiles" tab). The cohort below is 100 virtual patients - well inside the
200-per-arm cap and matching the scale of the app's own default group size.

```{r cohort}
set.seed(20250518)
n_sub <- 100L

events_cohort <- data.frame(
  id = rep(seq_len(n_sub), each = nrow(plasma)),
  time = rep(plasma$time, times = n_sub),
  amt = NA_real_,
  evid = 0L,
  cmt = "brain_vascular",
  CP_ABEMACICLIB_UM = rep(plasma$CP_ABEMACICLIB_UM, times = n_sub)
)
stopifnot(!anyDuplicated(events_cohort[, c("id", "time", "evid")]))

sim_cohort <- rxode2::rxSolve(
  readModelDb("Wickramasinghe_2025_abemaciclib_cns_pbpk"),
  events_cohort,
  covsInterpolation = "linear", returnType = "data.frame"
)
dplyr::n_distinct(sim_cohort$id)
```

## The IIV reproduces the app's own virtual patients

Table S3 lists five virtual patients that the SpatialCNS-PBPK app generated
from `Sim1` + `IIV1`. Those draws identify the distributional form the paper
never states: across the 355 non-zero parameter draws, `log(simulated /
typical)` is normal (Shapiro-Wilk p = 0.15, skewness 0.12) with SD 0.194
against the `sqrt(log(1 + 0.20^2))` = 0.198 expected of a 20% CV lognormal,
while the untransformed ratios are clearly right-skewed (Shapiro-Wilk
p = 1.6e-8, skewness 0.94). The packaged model encodes that lognormal form;
the check below confirms the simulated cohort returns the 20% CV.

```{r iiv-check}
sim_par <- sim_cohort |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  dplyr::select(PSB2, CLeffbbb2, fubb, VT2, lamT1)

cv <- vapply(sim_par, function(x) stats::sd(x) / mean(x), numeric(1))
tibble::tibble(
  Parameter = names(cv),
  `Simulated CV` = signif(cv, 3),
  `Table S2 IIV1` = 0.20
) |>
  knitr::kable(caption = "Realised interindividual CV in the 100-patient cohort.")

# 100 draws give a CV estimate with a relative standard error near
# 1/sqrt(2*99) = 7%, so allow a 3-sigma band around 0.20.
stopifnot(all(cv > 0.20 * (1 - 3 * 0.071)), all(cv < 0.20 * (1 + 3 * 0.071)))
```

## Variability bands by compartment

```{r cohort-bands, fig.height = 8}
sim_cohort |>
  dplyr::select(id, time, dplyr::all_of(names(panel_levels))) |>
  tidyr::pivot_longer(-c(id, time), names_to = "state", values_to = "conc") |>
  dplyr::group_by(state, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(conc, 0.05),
    Q50 = stats::quantile(conc, 0.50),
    Q95 = stats::quantile(conc, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(panel = factor(panel_levels[state], levels = panel_levels)) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(colour = "steelblue4") +
  facet_wrap(~panel, ncol = 2, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total abemaciclib (umol/L)",
       title = "Median and 5th-95th percentile profiles, 100 virtual patients",
       caption = paste("Reproduces the app's mean-and-individual-profiles output",
                       "using the Table S2 IIV1 column."))
```

# PKNCA validation

The app computes Cmax, Tmax and AUC over a user-selected window; the paper's
own worked example is "any user-defined time interval (e.g., 72-96 h)". The
same window is used here, one NCA per compartment. Note that 72-96 h is a
24 h dosing day part-way through the accumulation phase, not a steady-state
interval: the plasma peaks are still climbing at that point (0.391 umol/L at
64 h against 0.419 umol/L at 124.6 h), and the CNS compartments are further
from equilibrium still.

```{r pknca}
nca_states <- names(panel_levels)

sim_nca <- sim_cohort |>
  dplyr::select(id, time, dplyr::all_of(nca_states)) |>
  tidyr::pivot_longer(-c(id, time), names_to = "compartment", values_to = "Cc") |>
  dplyr::filter(!is.na(Cc))

# Guarantee a time = 0 record per (id, compartment). Every CNS compartment
# starts empty, so Cc = 0 at t = 0 is the correct anchor.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, compartment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, compartment, time, .keep_all = TRUE) |>
  dplyr::arrange(compartment, id, time)

stopifnot(nrow(sim_nca) > 0, all(sim_nca$Cc >= 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | compartment + id)

intervals <- data.frame(
  start = 72, end = 96,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

# There are no dose events in this model -- the systemic PK enters as a
# covariate -- so the NCA runs without a PKNCAdose object.
nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, data.dose = NA, intervals = intervals)
)
```

```{r pknca-table}
nca_tab <- as.data.frame(nca_res) |>
  dplyr::group_by(compartment, PPTESTCD) |>
  dplyr::summarise(value = stats::median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = value) |>
  dplyr::mutate(
    compartment = factor(compartment, levels = nca_states),
    dplyr::across(dplyr::where(is.numeric), ~ signif(.x, 3))
  ) |>
  dplyr::arrange(compartment)

nca_tab |>
  dplyr::rename(
    "Compartment" = compartment,
    "Cmax (umol/L)" = cmax,
    "Cmin (umol/L)" = cmin,
    "Tmax (h)" = tmax,
    "AUC72-96 (umol*h/L)" = auclast
  ) |>
  knitr::kable(caption = paste(
    "Median NCA parameters over the app's default 72-96 h window,",
    "100 virtual patients."
  ))
```

**No published NCA table to compare against.** This tutorial demonstrates the
app's pharmacokinetic-parameter function but does not print numeric Cmax /
Tmax / AUC values for abemaciclib anywhere in the paper or its supplement, so
there is no reference row for `ncaComparisonTable()`. The table above is
therefore a functional check of the packaged model rather than a comparison
against published values; the quantitative validation of this extraction rests
on the Figure 2 replication and the observed-data comparison above.

The NCA does support two internal consistency checks:

```{r pknca-checks}
# 1. Peak-to-trough ordering must hold in every compartment.
stopifnot(all(nca_tab$cmax >= nca_tab$cmin))

# 2. Over a 24 h window, AUC72-96 must lie between 24 * Cmin and 24 * Cmax
#    for every compartment.
stopifnot(
  all(nca_tab$auclast >= 24 * nca_tab$cmin),
  all(nca_tab$auclast <= 24 * nca_tab$cmax)
)

# 3. Exposure, not just instantaneous concentration, must rise monotonically
#    along the spatial chain the model exists to resolve. This is a distinct
#    check from the peak-concentration gradient above: AUC integrates over the
#    whole window, so a compartment could peak high and still be under-exposed.
auc <- stats::setNames(nca_tab$auclast, as.character(nca_tab$compartment))
stopifnot(all(diff(auc[c("Cccsf", "Cbm2", "CT1", "CT2", "CT3")]) > 0))
signif(auc, 3)
```

## Why 72-96 h cannot show the damping of the deep compartments

The natural next check would be that the deep compartments damp the twice-daily
oscillation relative to brain blood. Over 72-96 h that check does not hold, and
the reason is not that the model fails to damp: the CNS compartments are still
*filling* over that window, so their peak-to-trough range is dominated by the
accumulation trend rather than by intra-interval fluctuation. Every deep
compartment is higher at 96 h than it was at 72 h:

```{r accumulation}
acc_states <- c("Cbm2", "CT1", "CT2", "CT3")
at <- function(tt) {
  vapply(acc_states, function(v) {
    stats::approx(sim_typical$time, sim_typical[[v]], xout = tt)$y
  }, numeric(1))
}
accum <- tibble::tibble(
  Compartment = acc_states,
  `C(72 h)` = signif(at(72), 3),
  `C(96 h)` = signif(at(96), 3),
  `Ratio` = signif(at(96) / at(72), 3)
)
knitr::kable(accum, caption = "The deep compartments are still accumulating across the app's example window.")
stopifnot(all(accum$Ratio > 1.05))
```

Dosing in the published plasma profile stops after the 124.61 h peak, so the
damping *is* visible once the driving oscillation is no longer competing with
accumulation. Over the 144-168 h washout, brain blood swings by a factor of
several while the tumor core barely moves:

```{r damping}
swing_over <- function(lo, hi) {
  k <- sim_typical$time >= lo & sim_typical$time <= hi
  vapply(names(panel_levels), function(v) {
    x <- sim_typical[[v]][k]
    (max(x) - min(x)) / min(x)
  }, numeric(1))
}
sw <- swing_over(144, 168)
tibble::tibble(
  Compartment = names(sw),
  `Peak-to-trough swing` = signif(sw, 3),
  `Relative to Cbb` = signif(sw / sw[["Cbb"]], 3)
) |>
  knitr::kable(caption = "Relative fluctuation over the 144-168 h washout window.")

# Brain blood is the most variable compartment of the nine, and the tumor core
# damps it by more than three-fold. Both are properties of the published
# parameter set, so both are regression guards.
stopifnot(
  sw[["Cbb"]] == max(sw),
  sw[["CT3"]] < sw[["Cbb"]] / 3.5
)
```

# Structural check: CSF fluid balance

Table 1 sets `Qsout = QCsink - Qsin1` explicitly "to maintain fluid balance".
That is a checkable identity on the packaged parameter values, and it is the
kind of constraint a transcription error would break.

```{r fluid-balance}
theta <- stats::setNames(ui$iniDf$est, ui$iniDf$name)
val <- function(nm) if (paste0("l", nm) %in% names(theta)) {
  exp(theta[[paste0("l", nm)]])
} else {
  theta[[nm]]
}

fb <- c(QCsink = val("QCsink"), QSin1 = val("QSin1"),
        QSout = val("QSout"), residual = val("QCsink") - val("QSin1") - val("QSout"))
signif(fb, 4)
stopifnot(abs(fb[["residual"]]) < 1e-9)

# CSF production is split 60/40 between the cranial and spinal routes
# (Table 1: Qsin1 = 60% and Qsin2 = 40% of the 0.021 L/h production rate).
csf_prod <- val("QSin1") + val("QSin2")
signif(c(`Qsin1 + Qsin2` = csf_prod,
         `Qsin1 fraction` = val("QSin1") / csf_prod), 4)
stopifnot(abs(csf_prod - 0.021) < 1e-9,
          abs(val("QSin1") / csf_prod - 0.6) < 1e-3)
```

The absorption split is the same 62% / 38% cranial-to-spinal division Table 1
states for the arachnoid-villi route, and the glymphatic terms are each half
their corresponding sink:

```{r fluid-balance-2}
signif(c(
  `QCsink / (QCsink + QSsink)` = val("QCsink") / (val("QCsink") + val("QSsink")),
  `Qglyccsf / QCsink` = val("Qglyccsf") / val("QCsink"),
  `Qglyscsf / QSsink` = val("Qglyscsf") / val("QSsink")
), 4)
stopifnot(
  abs(val("QCsink") / (val("QCsink") + val("QSsink")) - 0.619) < 0.005,
  abs(val("Qglyccsf") / val("QCsink") - 0.5) < 0.01,
  abs(val("Qglyscsf") / val("QSsink") - 0.5) < 0.01
)
```

# Assumptions and deviations

## Errata and source inconsistencies

- **The unionization fraction of the tumor core does not match its stated pH.**
  The `lam` values are internally consistent with a single monobasic pKa of
  7.94 for every compartment: `lambb` = 0.2238 at pH 7.4, `lambm1` / `lambm2` =
  0.154 at pH 7.2, `lamT1` = 0.0676 at pH 6.8, `lamT2` = 0.035 at pH 6.5 and
  `lamvcsf` / `lamccsf` = 0.1864 at pH 7.3 all reproduce to four significant
  figures. The exception is `lamT3` = 0.0114, which corresponds to pH 6.00
  rather than the tumor-core pH of 6.2 stated in the main text (pH 6.2 would
  give 0.0179). The **published value 0.0114 is used as-is** - it is the value
  that generates Figure 2 - and the discrepancy is recorded here rather than
  silently corrected.

- **`CLeffT1` is inconsistent between Table S2 and Table S3.** Table S2 gives
  `CLeffT1` = 0.0584 L/h, but the `CLeffT1sim` column of Table S3 is exactly
  zero for all five app-generated virtual patients, unlike every other non-zero
  parameter. The model uses **0.0584**, the Table S2 value, because that is the
  value which reproduces Figure 2.

- **`Vvcsf` is 0.025 L in Table S2 and 0.0251 L in Table 1.** Table 1 derives
  it as 16.7% of the 0.15 L total CSF volume (0.02505 L). The model uses the
  Table S2 value 0.025, which is what the published simulation ran on.

- **`fuscsf` is defined but never used.** Table 2 lists the unbound fraction in
  the cranial *and* spinal subarachnoid CSF as a single row, but Table S2
  supplies only `fuccsf` and the spinal-CSF differential equation in Data S1
  contains no binding term at all - only flow terms. No `fuscsf` parameter is
  therefore encoded, and no unbound spinal-CSF observable is reported.

## Interpretation choices

- **The 20% CV is the input-file template's default, not an estimated
  variance.** Table S2 assigns exactly 0.2 to all 85 parameters, which is the
  signature of an illustrative placeholder rather than a fitted covariance: a
  uniform independent 20% CV on volumes, on unbound fractions and on
  unionization fractions is not a physiological IIV model, since `lam` is a
  deterministic function of pKa and pH and cannot vary independently of it.
  The value is nonetheless encoded, because it is what the app simulates with
  and what produces its variability bands. **Users fitting or simulating real
  populations should replace it.** The observed between-patient spread in
  Table S2 is much wider than 20% CV, and the paper does not attempt to fit it.

- **The lognormal form of the IIV was identified, not assumed.** The paper
  never states the distribution. It was recovered from Table S3, the app's own
  five virtual patients, by the test reported in the "IIV reproduces the app's
  own virtual patients" section above. This is a derivation from an on-disk
  source, not a class-typical default.

- **Concentration units.** The paper labels its figure axes only
  "Concentrations" and never states a unit. The 9-CNS ODE system is linear and
  homogeneous in concentration, so any self-consistent unit works provided the
  supplied plasma profile and the interpreted outputs share it. `umol/L` is
  declared because it is the unit implied by the Table S2 values: the plasma
  peak of 0.4191 corresponds to 212 ng/mL at abemaciclib's molecular weight of
  506.6 g/mol, and the observed tumor (0.3-21) and CSF (0.002-0.2) values are
  in the range reported for abemaciclib brain-tumor tissue.

- **No residual-error model.** The paper reports no residual variability and
  the app is a deterministic simulator whose only stochastic element is the
  parameter-level IIV. No `addSd` / `propSd` is invented.

- **The dose is not encoded.** The model has no dose events at all: the
  systemic PK enters entirely through `CP_ABEMACICLIB_UM`. The paper does not
  state the abemaciclib dose that produced the Table S2 plasma profile, only
  that it is a population mean from glioblastoma patients; the median
  inter-peak interval of 11.4 h indicates twice-daily dosing.

- **Observed-data pairing.** Figure 2 overlays the non-enhancing tumor
  observations on `Cbm2` / `CT1` / `CT2` / `CT3` and the CSF observations on
  all three CSF panels, without asserting a single "correct" compartment for
  each matrix. The quantitative comparison in this vignette pairs the
  non-enhancing tumor with `CT1` (the infiltrative rim), the contrast-enhancing
  tumor with `CT2` (the bulk tumor) and CSF with `Cccsf`, following the paper's
  own anatomical definitions of those regions.

## Scope

- **The 4-CNS model is not extracted.** The paper also packages a
  4-compartment CNS model, used solely to validate the app's ODE solver against
  Simcyp v18 with ribociclib (Figure S3, Table S1). Its structure, differential
  equations and system-specific parameters are third-party, published in
  Gaohua et al., *Drug Metab Pharmacokinet* 2016;31:224-233, and appear nowhere
  in this paper or its supplement. Extracting it would require fabricating its
  ODEs.

- **Upstream development paper.** The 9-CNS model was originally developed and
  validated across six drugs in Li J, Wickramasinghe C, Jiang J, et al.,
  *Clin Pharmacol Ther* 2025;117(3):690-703 (reference 7). That paper is not on
  disk; it is the richer source for the other five drugs' parameter sets, but
  nothing in this extraction depends on it - the tutorial and its supplement
  are self-contained for abemaciclib.
