---
title: "Pamiparib (Wickramasinghe 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pamiparib (Wickramasinghe 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Wickramasinghe_2025_pamiparib"))$reference`
* Article: <https://doi.org/10.3390/pharmaceutics17040524>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12030534/>

The paper reports **two separately estimated fits** of the same simultaneously
modelled total-plus-unbound dataset, differing only in whether the apparent
volume and apparent clearance are referenced to the total or to the unbound
plasma concentration (Table 2, "Total Pamiparib" and "Unbound Pamiparib"
columns; OFV 9210 versus 9212). Following the library's replicate-the-author's-
structure policy, both are packaged:

| Model | Parameterisation | Typical V/F | Typical CL/F |
|---|---|---|---|
| `Wickramasinghe_2025_pamiparib` | referenced to **total** plasma concentration | 44 L | 2.59 L/h |
| `Wickramasinghe_2025_pamiparib_unbound` | referenced to **unbound** plasma concentration | 1060 L | 62.5 L/h |

Both files predict both endpoints: `Cc` (total plasma concentration) and `Cu`
(unbound plasma concentration), linked by the constant binding relationship
`Cu = Cc * fu`.

```{r models}
mod_total   <- rxode2::rxode2(readModelDb("Wickramasinghe_2025_pamiparib"))
mod_unbound <- rxode2::rxode2(readModelDb("Wickramasinghe_2025_pamiparib_unbound"))

# Pamiparib molecular weight, used throughout to convert the model's mass
# concentrations (mg/L) to the molar concentrations (nM) the paper reports.
# Discussion paragraph 2: "Pamiparib is a small-molecule drug (molecular
# weight: 298.3)".
MW_PAMIPARIB <- 298.3
mgL_to_nM <- function(x) x / MW_PAMIPARIB * 1e6
```

# Population

Wickramasinghe et al. 2025 developed the population PK model from a phase 0/II
study (ClinicalTrials.gov NCT04614909) that enrolled 41 patients with newly
diagnosed or recurrent glioblastoma at a single US centre. Patients received
oral pamiparib 60 mg twice daily for three days (nine doses) before surgical
tumour resection; plasma was sampled pre-dose and at 0.5, 1, 2, 4, 7 and 24 h
after the ninth dose, and total and unbound pamiparib were quantified by a
validated LC-MS/MS method.

Baseline demographics are from Table 1: median age 60 years (range 31-80),
median weight 80 kg (range 45-129), median height 173 cm (range 155-193),
median BSA 1.99 m^2 (range 1.41-2.53), 21 of 41 female, 38 of 41 white.
Eligibility required ECOG performance status <= 2, adequate marrow function,
serum transaminases <= 3x ULN, total bilirubin <= 1.5x ULN, and eGFR
>= 30 mL/min/1.73 m^2 by CKD-EPI. Median post-operative creatinine clearance
was 111 mL/min (range 59-169) versus a pre-dose median of 98 mL/min (range
39-154); the final covariate model uses the **post-operative** value. 34 of 41
patients received dexamethasone (median total dose 19 mg); no concomitant
medication was retained as a covariate.

The same information is available programmatically:

```{r population}
rxode2::rxode2(readModelDb("Wickramasinghe_2025_pamiparib"))$population
```

# Source trace

Per-parameter origin (also recorded as in-file comments next to each `ini()`
value). Table 2 reports the model coefficients (theta, beta) together with the
typical values TV_V and TV_CL those coefficients produce at the population
median covariates; Table 3 reports the same final-model coefficients alongside
the 500-replicate bootstrap and, critically, the **log-scale SDs of the random
effects** (`Ka_SD`, `V/F_SD`, `CL/F_SD`, `Fu_SD`), which is what `ini()` needs.

## `Wickramasinghe_2025_pamiparib` (total-drug parameterisation)

| Parameter | Value | Source location |
|---|---|---|
| `lka` | log(1.58) | Table 2, Total Pamiparib final model, theta1 (Ka) = 1.58 h^-1 (RSE 42%) |
| `lvc` | log(15) | Table 2, Total Pamiparib final model, theta2 (V/F) = 15 L (RSE 28%) |
| `lcl` | log(6.76) | Table 2, Total Pamiparib final model, theta3 (CL/F) = 6.76 L/h (RSE 38%) |
| `lfu` | log(0.041) | Table 2, Total Pamiparib final model, theta4 (Fu) = 0.041 (RSE 2.8%) |
| `e_crcl_vc` | 0.0094 | Table 2, beta1 (PCC on V/F) = 0.0094 per mL/min (RSE 25%) |
| `e_age_cl` | -0.016 | Table 2, beta2 (Age on CL/F) = -0.016 per year (RSE 37%) |
| `etalka` | 2.8900 | Table 3, Total Pamiparib, `Ka_SD` = 1.7 -> variance 1.7^2 |
| `etalvc` | 0.1521 | Table 3, Total Pamiparib, `V/F_SD` = 0.39 -> variance 0.39^2 |
| `etalcl` | 0.2209 | Table 3, Total Pamiparib, `CL/F_SD` = 0.47 -> variance 0.47^2 |
| `etalfu` | 0.0144 | Table 3, Total Pamiparib, `Fu_SD` = 0.12 -> variance 0.12^2 |
| `addSd`, `propSd`, `addSd_Cu`, `propSd_Cu` | 0 | Section 2.2.1 Equation (1) declares a combined error model g = a + b*f; **a and b are not reported anywhere in the paper** (see Errata) |

## `Wickramasinghe_2025_pamiparib_unbound` (unbound-drug parameterisation)

| Parameter | Value | Source location |
|---|---|---|
| `lka` | log(1.7) | Table 2, Unbound Pamiparib final model, theta1 (Ka) = 1.7 h^-1 (RSE 44%) |
| `lvc` | log(402) | Table 2, Unbound Pamiparib final model, theta2 (V/F) = 402 L (RSE 28%) |
| `lcl` | log(163) | Table 2, Unbound Pamiparib final model, theta3 (CL/F) = 163 L/h (RSE 39%) |
| `lfu` | log(0.042) | Table 2, Unbound Pamiparib final model, theta4 (Fu) = 0.042 (RSE 2.6%) |
| `e_crcl_vc` | 0.0087 | Table 2, beta1 (PCC on V/F) = 0.0087 per mL/min (RSE 26%) |
| `e_age_cl` | -0.016 | Table 2, beta2 (Age on CL/F) = -0.016 per year (RSE 38%) |
| `etalka` | 2.7556 | Table 3, Unbound Pamiparib, `Ka_SD` = 1.66 -> variance 1.66^2 |
| `etalvc` | 0.1521 | Table 3, Unbound Pamiparib, `V/F_SD` = 0.39 -> variance 0.39^2 |
| `etalcl` | 0.2401 | Table 3, Unbound Pamiparib, `CL/F_SD` = 0.49 -> variance 0.49^2 |
| `etalfu` | 0.0121 | Table 3, Unbound Pamiparib, `Fu_SD` = 0.11 -> variance 0.11^2 |

## Structure

| Equation | Source location |
|---|---|
| One compartment, first-order absorption without lag, first-order elimination | Section 2.2.1 and Section 3.1 |
| `Cu = Cc * fu` | Section 2.2.1: "The relationship between total (Cp) and unbound (Cu) drug concentrations was incorporated into the model using the equation Cu = Cp x Fu" |
| `V/F = theta2 * exp(beta1 * PCC)` (uncentered) | Table 2 footnote: "TV_V = theta2 * e^(beta1 * PCC), with the population median PCC being 111.5" |
| `CL/F = theta3 * exp(beta2 * Age)` (uncentered) | Table 2 footnote: "TV_CL = theta3 * e^(beta2 * Age), with the population median age being 60" |
| Log-normal (exponential) IIV on all PK parameters | Section 2.2.1 |

# Verification of the published typical values

The Table 2 footnotes state the functional form and the median covariate values
at which the typical values `TV_V/F` and `TV_CL/F` were evaluated. Solving each
packaged model with the random effects zeroed at those medians must reproduce
the published typical values.

```{r typical}
CRCL_MEDIAN <- 111.5  # Table 2 footnote (post-operative creatinine clearance, mL/min)
AGE_MEDIAN  <- 60     # Table 2 footnote (years)

typical_params <- function(mod) {
  ev <- rxode2::et(amt = 60, cmt = "depot")
  ev <- rxode2::et(ev, c(0, 1))
  ev <- as.data.frame(ev)
  ev$dvid <- ifelse(ev$evid == 0, 1L, NA_integer_)
  ev$CRCL <- CRCL_MEDIAN
  ev$AGE  <- AGE_MEDIAN
  s <- rxode2::rxSolve(rxode2::zeroRe(mod), ev, returnType = "data.frame",
                       useLinCmt = FALSE)
  c(ka = s$ka[1], vc = s$vc[1], cl = s$cl[1], fu = s$fu[1],
    thalf = log(2) * s$vc[1] / s$cl[1])
}

tv_total   <- typical_params(mod_total)
tv_unbound <- typical_params(mod_unbound)

tv_check <- tibble::tribble(
  ~Model,     ~Quantity,             ~Published, ~Reproduced,
  "total",   "Ka (1/h)",             1.58,       tv_total[["ka"]],
  "total",   "TV V/F (L)",           44,         tv_total[["vc"]],
  "total",   "TV CL/F (L/h)",        2.59,       tv_total[["cl"]],
  "total",   "Fu",                   0.041,      tv_total[["fu"]],
  "unbound", "Ka (1/h)",             1.7,        tv_unbound[["ka"]],
  "unbound", "TV Vu/F (L)",          1060,       tv_unbound[["vc"]],
  "unbound", "TV CLu/F (L/h)",       62.5,       tv_unbound[["cl"]],
  "unbound", "Fu",                   0.042,      tv_unbound[["fu"]]
) |>
  dplyr::mutate(`% diff` = round(100 * (Reproduced - Published) / Published, 2),
                Reproduced = round(Reproduced, 4))

knitr::kable(
  tv_check,
  caption = paste("Published typical values (Table 2, final covariate models)",
                  "versus the packaged models evaluated at the Table 2 footnote",
                  "median covariates.")
)
```

Every entry reproduces exactly except `TV V/F` of the total-drug model, which
lands at `r round(tv_total[["vc"]], 2)` L against a published 44 L. That gap is
a rounding artefact and not an encoding error: Table 2 reports `theta2` to two
significant figures (15 L), and 15 * exp(0.0094 * 111.5) = 42.8 L. Recovering
exactly 44 L would need `theta2` = 15.43. The published `theta2` is encoded as
printed; it is **not** tuned to hit the typical value.

```{r typical-gates}
# The unbound parameterisation and both clearance arms reproduce to <0.5%.
stopifnot(
  abs(tv_total[["cl"]]      / 2.59  - 1) < 0.005,
  abs(tv_total[["fu"]]      / 0.041 - 1) < 1e-6,
  abs(tv_unbound[["vc"]]    / 1060  - 1) < 0.005,
  abs(tv_unbound[["cl"]]    / 62.5  - 1) < 0.005,
  abs(tv_unbound[["fu"]]    / 0.042 - 1) < 1e-6,
  # Total-drug V/F carries the theta2 = 15 two-significant-figure rounding.
  abs(tv_total[["vc"]]      / 44    - 1) < 0.03
)
```

## Elimination half-life

Section 3.1 reports a population mean elimination half-life of about 12 h,
"consistent with the estimated T1/2 of 13 h in the first in-human study".

```{r halflife}
thalf <- c(total = tv_total[["thalf"]], unbound = tv_unbound[["thalf"]])
round(thalf, 2)

# Both parameterisations must land on the paper's stated ~12 h. These are
# closed-form log(2) * V / CL values at fixed covariates, so the bound is
# pinned to the accuracy actually achieved rather than left loose.
stopifnot(all(thalf > 11.4), all(thalf < 11.9))
```

## Consistency of the two parameterisations

The two fits are reparameterisations of the same joint model, so the unbound-
referenced volume and clearance should be close to their total-drug
counterparts divided by `fu`. They are not identical, because each was
estimated in a separate SAEM run.

```{r param-equivalence}
equiv <- tibble::tibble(
  Quantity = c("V/F (L)", "CL/F (L/h)"),
  `From total model` = c(tv_total[["vc"]], tv_total[["cl"]]),
  `From unbound model x fu` = c(tv_unbound[["vc"]] * tv_unbound[["fu"]],
                                tv_unbound[["cl"]] * tv_unbound[["fu"]])
) |>
  dplyr::mutate(
    `% diff` = round(100 * (`From unbound model x fu` / `From total model` - 1), 1),
    dplyr::across(where(is.numeric), ~ round(.x, 3))
  )

knitr::kable(equiv, caption = "Cross-parameterisation consistency at the median covariates.")

# Independent SAEM runs, so agreement is expected to be close but not exact.
stopifnot(all(abs(equiv[["% diff"]]) < 5))
```

# Virtual cohort

The observed data are not public (Data Availability Statement: available from
the corresponding author on reasonable request). Section 2.4 states that the
dose-optimisation simulations used "the estimated population parameters along
with their respective interindividual variability" in "a general patient
population", but does not state how age and creatinine clearance were sampled.
Here they are drawn from normal distributions centred on the Table 1 medians
with SDs chosen so that the Table 1 ranges span roughly +/- 2.2 SD of a
41-subject sample, then truncated to the observed ranges. That choice is
internally consistent: the covariate-explained variance it adds to the final
model's residual IIV reconstructs the base model's IIV almost exactly
(V/F: sqrt(0.39^2 + (0.0094 * 25)^2) = 0.46 -> 48% CV versus the base model's
53%; CL/F: sqrt(0.47^2 + (0.016 * 11.1)^2) = 0.50 -> 54% CV versus 53%).

```{r cohort}
set.seed(20250416)
N_PER_ARM <- 200L

cov_tbl <- tibble::tibble(
  subject = seq_len(N_PER_ARM),
  AGE  = pmin(pmax(stats::rnorm(N_PER_ARM, 60,    11.1), 31), 80),
  CRCL = pmin(pmax(stats::rnorm(N_PER_ARM, 111.5, 25.0), 59), 169)
)

regimens <- tibble::tibble(
  treatment = c("60 mg BID", "60 mg QD", "40 mg BID",
                "30 mg BID", "20 mg BID", "10 mg BID"),
  dose      = c(60, 60, 40, 30, 20, 10),
  ii        = c(12, 24, 12, 12, 12, 12)
) |>
  dplyr::mutate(id_offset = (dplyr::row_number() - 1L) * N_PER_ARM)

# Section 2.4 / Table 4: five days of treatment. Observations run on a 0.5 h
# grid across the whole 120 h window, refined to 0.05 h over the final dosing
# interval where the steady-state NCA is taken.
SIM_END <- 120

make_arm <- function(treatment, dose, ii, id_offset) {
  ids <- id_offset + cov_tbl$subject
  obs_t <- sort(unique(c(seq(0, SIM_END, by = 0.5),
                         seq(SIM_END - ii, SIM_END, by = 0.05))))
  dose_t <- seq(0, by = ii, length.out = floor(SIM_END / ii))
  doses <- tidyr::expand_grid(id = ids, time = dose_t) |>
    dplyr::mutate(evid = 1L, amt = dose, cmt = "depot", dvid = NA_integer_)
  # Both endpoints are algebraic transforms of the ODE states, so observation
  # rows carry dvid and no compartment. Naming an observable as the
  # compartment would auto-inject a slot for it and renumber the ODE states.
  obs <- tidyr::expand_grid(id = ids, time = obs_t) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, cmt = NA_character_, dvid = 1L)
  dplyr::bind_rows(doses, obs) |>
    dplyr::mutate(treatment = treatment, subject = id - id_offset) |>
    dplyr::left_join(cov_tbl, by = "subject") |>
    dplyr::select(-subject) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- do.call(
  dplyr::bind_rows,
  Map(make_arm, regimens$treatment, regimens$dose, regimens$ii, regimens$id_offset)
)

# IDs must be disjoint across arms or rxSolve silently merges subjects.
stopifnot(
  dplyr::n_distinct(events$id) == N_PER_ARM * nrow(regimens),
  nrow(dplyr::distinct(events, id, treatment)) == N_PER_ARM * nrow(regimens)
)
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(
  mod_total, events = events,
  keep = c("treatment", "AGE", "CRCL"),
  useLinCmt = FALSE
) |>
  as.data.frame() |>
  dplyr::mutate(Cu_nM = mgL_to_nM(Cu), Cc_nM = mgL_to_nM(Cc))

stopifnot(nrow(sim) > 0, all(is.finite(sim$Cu_nM)), all(sim$Cu_nM >= 0))
```

# Replicate Figure 4

Figure 4 of Wickramasinghe 2025 plots the 50th, 5th and 95th percentiles of the
model-predicted **unbound** pamiparib concentration-time profile over five days
for each dosing regimen, with a dashed line at the assumed target-engagement
threshold (a target engagement ratio of 5 against a PARP IC50 of about 1 nM,
i.e. 5 nM unbound).

```{r figure-4, fig.width = 7.2, fig.height = 5}
TARGET_NM <- 5  # Section 3.3: target engagement ratio of 5 x IC50 (~1 nM)

sim |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cu_nM, 0.05),
    Q50 = stats::quantile(Cu_nM, 0.50),
    Q95 = stats::quantile(Cu_nM, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(treatment = factor(treatment, levels = regimens$treatment)) |>
  ggplot(aes(x = time)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, fill = "steelblue") +
  geom_line(aes(y = Q50), colour = "firebrick") +
  geom_hline(yintercept = TARGET_NM, linetype = "dashed") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Unbound pamiparib (nM)",
    title = "Figure 4 - model-predicted unbound pamiparib by dosing regimen",
    caption = paste("Replicates Figure 4 of Wickramasinghe 2025.",
                    "Line = median, band = 5th-95th percentile,",
                    "dashed line = target engagement ratio of 5 (5 nM).")
  )
```

# PKNCA validation

Table 4 reports the model-predicted steady-state peak, trough and average
**unbound** plasma concentrations after five days of treatment. The NCA below
is taken over the final dosing interval of the simulation (108-120 h for the
twice-daily arms, 96-120 h for the once-daily arm), which is the same window
Table 4 describes.

```{r pknca}
# Only !is.na() -- a `time > 0` or `Cu > 0` filter would drop the interval-start
# record that PKNCA anchors AUC on.
sim_nca <- sim |>
  dplyr::filter(!is.na(Cu_nM)) |>
  dplyr::select(id, time, Cu_nM, treatment)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cu_nM ~ time | treatment + id,
  concu = "nM", timeu = "h"
)

dose_df <- events |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

# One interval per regimen: the last dosing interval of the 120 h simulation.
intervals <- regimens |>
  dplyr::transmute(
    treatment,
    start = SIM_END - ii,
    end   = SIM_END,
    cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE,
    auclast = TRUE, half.life = TRUE
  ) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_df <- as.data.frame(nca_res)
stopifnot(nrow(nca_df) > 0, all(regimens$treatment %in% nca_df$treatment))
```

The NCA-derived terminal half-life is a second, independent check on the ~12 h
the paper reports.

```{r halflife-nca}
hl <- nca_df |>
  dplyr::filter(PPTESTCD == "half.life") |>
  dplyr::summarise(median_h = stats::median(PPORRES, na.rm = TRUE)) |>
  dplyr::pull(median_h)

round(hl, 2)
stopifnot(hl > 11.5, hl < 12.3)
```

## Comparison against the published Table 4 medians

```{r nca_compare}
# Table 4: model-predicted steady-state unbound pamiparib concentrations (nM),
# 50th percentile column, following 5 days of treatment.
published <- tibble::tribble(
  ~treatment,  ~cmax, ~cmin, ~cav,
  "60 mg BID",   302,   175,  233,
  "60 mg QD",    193,    61,  115,
  "40 mg BID",   201,   117,  155,
  "30 mg BID",   151,    88,  116,
  "20 mg BID",   101,    58,   78,
  "10 mg BID",    50,    29,   39
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by        = "treatment",
  params    = c("cmax", "cmin", "cav"),
  units     = c(cmax = "nM", cmin = "nM", cav = "nM"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated versus published Table 4 median steady-state",
                  "unbound pamiparib concentrations.",
                  "* differs from the reference by >20%."),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca_compare_gate}
stopifnot(nrow(cmp) == 18L)  # 6 regimens x 3 parameters; guard against a silent join drop
worst <- max(abs(as.numeric(gsub("[^0-9.-]", "", cmp[["% diff"]]))), na.rm = TRUE)
worst

# All 18 median comparisons agree with Table 4 to within this bound, so no row
# is starred at the 20% tolerance. The residual bias is dominated by the
# theta2 = 15 rounding (which alone raises every simulated concentration by
# ~2.9%, see the typical-value section) plus Monte Carlo noise at 200 subjects
# per arm. ncaComparisonTable() attaches a "footnote" attribute only when some
# row exceeds the tolerance, so its absence is the machine-checkable form of
# "nothing was starred".
stopifnot(worst < 20, is.null(attr(cmp, "footnote")))
```

## Why the percentile bands are narrower than Table 4

The medians reproduce, but the 5th-95th spread of this simulation is materially
narrower than Table 4's, because the residual-error magnitudes the authors fitted
are not reported (see Errata) and therefore cannot be simulated.

```{r percentile-spread}
sim_pct <- nca_df |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "cav")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    P05 = stats::quantile(PPORRES, 0.05, na.rm = TRUE),
    P95 = stats::quantile(PPORRES, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(`Simulated P95/P05` = round(P95 / P05, 1)) |>
  dplyr::select(treatment, PPTESTCD, `Simulated P95/P05`)

published_spread <- tibble::tribble(
  ~treatment,  ~PPTESTCD, ~p05,  ~p95,
  "60 mg BID", "cmax",    105,   774,
  "60 mg BID", "cmin",     16,   581,
  "60 mg BID", "cav",      47,   673,
  "60 mg QD",  "cmax",     63,   485,
  "60 mg QD",  "cmin",    0.6,   253,
  "60 mg QD",  "cav",      14,   357,
  "40 mg BID", "cmax",     68,   515,
  "40 mg BID", "cmin",     11,   388,
  "40 mg BID", "cav",      31,   448,
  "30 mg BID", "cmax",     53,   387,
  "30 mg BID", "cmin",      8,   291,
  "30 mg BID", "cav",      24,   336,
  "20 mg BID", "cmax",     35,   258,
  "20 mg BID", "cmin",      5,   194,
  "20 mg BID", "cav",      16,   224,
  "10 mg BID", "cmax",     18,   129,
  "10 mg BID", "cmin",      3,    97,
  "10 mg BID", "cav",       8,   112
) |>
  dplyr::mutate(`Published P95/P05` = round(p95 / p05, 1)) |>
  dplyr::select(treatment, PPTESTCD, `Published P95/P05`)

spread_cmp <- dplyr::inner_join(published_spread, sim_pct,
                                by = c("treatment", "PPTESTCD")) |>
  dplyr::rename("NCA parameter" = PPTESTCD, "Regimen" = treatment) |>
  dplyr::mutate(Regimen = factor(Regimen, levels = regimens$treatment)) |>
  dplyr::arrange(Regimen, `NCA parameter`)

knitr::kable(
  spread_cmp,
  caption = paste("Ratio of the 95th to the 5th percentile.",
                  "The published spread is consistently wider because the",
                  "paper's simulation includes a residual-error component",
                  "whose magnitude it never reports.")
)

# The gate is directional: the simulation must be narrower than the paper's,
# never wider, on every row. If a future edit accidentally introduces residual
# variability this goes red.
stopifnot(all(spread_cmp[["Simulated P95/P05"]] <
              spread_cmp[["Published P95/P05"]]))
```

# Dose-reduction conclusion

Section 3.3 concludes that "doses as low as 20 mg BID may be adequate for
therapeutic effects, even in patients with rapid clearance, where the unbound
pamiparib Css,min was approximately 5 nM (i.e., 5th percentile)". Because the
5th percentile of the packaged simulation is not reproducible (previous
section), the conclusion is restated here on the quantity that is: the median
trough at 20 mg BID sits far above the 5 nM threshold, and every regimen from
10 to 60 mg BID keeps the median trough above it.

```{r dose-conclusion}
trough <- nca_df |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(`Median Css,min (nM)` = round(stats::median(PPORRES), 1),
                   .groups = "drop") |>
  dplyr::mutate(
    treatment = factor(treatment, levels = regimens$treatment),
    `Target engagement ratio` = round(`Median Css,min (nM)` / 1, 1)
  ) |>
  dplyr::arrange(treatment) |>
  dplyr::rename("Regimen" = treatment)

knitr::kable(
  trough,
  caption = paste("Median steady-state unbound trough and the implied target",
                  "engagement ratio against the PARP IC50 of ~1 nM.")
)

stopifnot(nrow(trough) == nrow(regimens),
          all(trough[["Target engagement ratio"]] > TARGET_NM))
```

# Assumptions and deviations

* **Residual error is not reported and is carried at zero.** Section 2.2.1
  states that a combined error model `g = a + b * f` (with `c` fixed at 1) was
  selected, but neither `a` nor `b` appears in Table 2, Table 3, or anywhere
  else in the paper. All four residual SDs (`addSd`, `propSd`, `addSd_Cu`,
  `propSd_Cu`) are therefore `fixed(0)`. Zero records "magnitude not
  published", **not** "the authors estimated zero residual variability".
  Simulations from these models reproduce the structural and between-subject
  spread only; the Table 4 5th and 95th percentiles are consequently not
  reproducible, as quantified above.
* **`theta2` rounding.** Table 2 reports `theta2` (V/F) for the total-drug model
  to two significant figures (15 L), which yields TV V/F = 42.8 L rather than
  the 44 L printed in the same table. The published `theta2` is encoded as
  printed and the 2.8% gap is carried, not tuned away. The unbound-drug model's
  `theta2` = 402 L has enough precision to reproduce its TV of 1060 L exactly.
* **Covariates enter uncentered.** Section 2.2.2 states that "continuous
  covariates ... were centered around their median values", but the Table 2
  footnote equations `TV_V = theta2 * e^(beta1 * PCC)` and
  `TV_CL = theta3 * e^(beta2 * Age)` are only self-consistent with the reported
  `theta` values in the **uncentered** form: centering would make TV_V = theta2
  = 15 L, contradicting the printed TV of 44 L. Per the text-versus-equation
  rule the printed equation is trusted, so `exp(lvc)` and `exp(lcl)` are
  covariate-free intercepts rather than typical values.
* **Post-operative, not pre-dose, creatinine clearance.** Table 1 reports both;
  the Table 2 footnote names the covariate "PCC, post-operation creatine
  clearance" with a median of 111.5 mL/min, so the post-operative value is what
  the `CRCL` column must be populated with. It is a raw creatinine clearance in
  mL/min and is **not** normalised to 1.73 m^2 BSA.
* **Virtual-cohort covariate distributions are assumed.** The paper does not say
  how (or whether) age and creatinine clearance were sampled for its "general
  patient population" simulation. Normal distributions centred on the Table 1
  medians and truncated to the Table 1 ranges are used here; the justification
  for the chosen SDs is given in the Virtual cohort section.
* **No IIV correlation structure is reported.** Neither Table 2 nor Table 3
  reports an off-diagonal omega, so the random effects are encoded as
  independent.
* **Molecular weight for unit conversion.** The models work in mg and L, so
  concentrations come out in mg/L; Table 4 is in nM. The conversion uses
  MW = 298.3 g/mol, taken from the Discussion ("molecular weight: 298.3").
  The paper does not print the conversion itself.
* **Observed data are not public.** The Data Availability Statement offers the
  raw data from the corresponding author on reasonable request, so all figures
  here are simulation-only; no observed-versus-predicted overlay is possible.

# Errata and source-table defects

* **Table 2 and Table 3 abbreviation lists mislabel two parameters.** Both read
  "Ka, elimination rate constant" (Ka is the *absorption* rate constant, as
  Section 2.2.1 and the Results text make clear) and "CL/F, apparent volume of
  distribution" (a copy-paste of the V/F gloss; CL/F is apparent oral
  clearance). Cosmetic, but they would mislead a reader transcribing from the
  abbreviation list alone.
* **Table 2, "IIV of Fu (%)" for the unbound base model reads 112.** Every
  neighbouring entry is 11-12 (total base 12, total final 12, unbound final 11),
  and the corresponding Table 3 `Fu_SD` is 0.11, so 112 is almost certainly a
  misprint for 11.2. The value is not used here: the packaged models take the
  **final** covariate models.
* **The "Table 2. Cont." header row is corrupted.** Its column-group labels read
  "Total Pamiparib | Unbound Pamiparib | Unbound Pamiparib | Unbound Pamiparib"
  rather than the Total/Unbound x Base/Final structure of the first part. The
  column identity used here was resolved by matching the theta values in the
  continuation rows to the first part of Table 2 and to Table 3, which agree.
* **The abstract rounds the covariate contributions differently from the
  Results.** The abstract says creatinine clearance and age explain "~22% and
  ~5%" of the IIV of V/F and CL/F; Section 3.1 gives "~23% and ~22%" for V/F
  (total, unbound) and "~6% and ~4%" for CL/F (total, unbound). These are
  descriptive statements only and do not enter the model.
* No erratum or corrigendum for this article was found on the MDPI article
  landing page or in PubMed as of the extraction date.
