---
title: "Elinzanetant (Willmann 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elinzanetant (Willmann 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
ui <- rxode2::rxode(readModelDb("Willmann_2024_elinzanetant"))
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.1002/psp4.13226>
* Supplement (Data S1, model parameterisation equations and Tables S1-S2):
  <https://doi.org/10.1002/psp4.13226>

Elinzanetant is a dual neurokinin-1 (NK-1) and neurokinin-3 (NK-3) receptor
antagonist developed for moderate-to-severe vasomotor symptoms (VMS) associated
with menopause. Willmann 2024 fitted a single joint model to the parent drug, an
intravenous stable-isotope tracer, and three principal metabolites, and coupled
it to a fixed-EC50 receptor-occupancy layer.

* Description: `r ui$description`

# Population

```{r population}
pop <- ui$population
str(pop, max.level = 1, give.attr = FALSE)
```

The model-building dataset pooled `r pop$n_subjects` subjects from
`r pop$n_studies` phase I studies (Willmann 2024 Table 1), contributing 8408
elinzanetant PK observations of which 5.7% were below the limit of
quantification and were handled with the M3 method. Fifty-two of those subjects
also contributed 2020 observations for each of the three metabolites. Baseline
demographics are from Table S1: median age `r pop$age_median`
(`r pop$age_range`), median weight `r pop$weight_median`
(`r pop$weight_range`), median height `r pop$height_median`,
`r pop$sex_female_pct`% female, 78.5% White / 19.5% Black / 2% Other, and 29.7%
women with VMS versus 70.3% healthy volunteers.

A further 171 subjects across three phase I/II studies (including 139 women with
VMS in SWITCH-1) formed an external validation dataset used only for visual
predictive checks; across both datasets 366 subjects were studied, 336 of them
female and 197 women with VMS.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Willmann_2024_elinzanetant.R` carries an in-file
comment naming its source location. The table below collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| Circadian modulation `P(t) = Pavg * (1 + AMP * cos(2*pi/24 * (t + TFD + theta_SHIFT)))` | n/a | Equation 1; restated per parameter in Data S1 |
| Receptor occupancy `RO = Cpls / (Cpls + EC50)` | n/a | Equation 2 |
| Compartment topology, micro-constants `K12, K24, K42, K35, K53, K67, K76, K89, K98, K20, K26, K28, K610, K810, K100` | n/a | Figure 1 and Data S1 "Information about PopPK model parametrization" |
| `lcl` (elinzanetant CL) | 7.26 L/h | Table 2, `CLpop` |
| `lvc` (elinzanetant Vc) | 23.7 L | Table 2, `Vcpop` |
| `lvp` (elinzanetant Vp) | 168 L | Table 2, `Vp` |
| `lq` (elinzanetant Q) | 6.57 L/h | Table 2, `Q` |
| `lka` | 2.37 1/h | Table 2, `KaREF` |
| `ltlag` | 0.292 h | Table 2, `ALAGREF` |
| `lfdepot` | 0.367 | Table 2, `FREF` |
| `e_fed_highfat_ka` | -0.911 | Table 2, `KaFED2` |
| `e_form_elz_hardgel_ka` | -0.264 | Table 2, `KaFORM3` |
| `e_fed_highfat_tlag` | 0.545 | Table 2, `ALFED2` |
| `e_form_elz_hardgel_tlag` | 0.548 | Table 2, `ALFORM3` |
| `e_fed_highfat_fdepot` | -0.227 | Table 2, `FFED2` |
| `lcl_m3034`, `lvc_m3034`, `lvp_m3034`, `lq_m3034` | 4.14 L/h, 3.58 L, 15.6 L, 3.65 L/h | Table 2, `CL3034pop` / `Vc3034` / `Vp3034` / `Q3034` |
| `lvmax_m3034`, `lkm_m3034` | 24.1 ug/L/h, 7.87 ug/L | Table 2, `Vmax3034` / `Km3034` |
| `lcl_m27`, `lvc_m27`, `lvp_m27`, `lq_m27` | 4.11 L/h, 1.85 L, 40.4 L, 4.45 L/h | Table 2, `CL27pop` / `Vc27` / `Vp27` / `Q27` |
| `lvmax_m27`, `lkm_m27` | 42.6 ug/L/h, 17.7 ug/L | Table 2, `Vmax27` / `Km27` |
| `lcl_m1821`, `lvc_m1821` | 9.10 L/h, 0.636 L | Table 2, `CL1821pop` / `Vc1821` |
| `lvmax_m1821`, `lkm_m1821` | 335 ug/L/h, 45.8 ug/L | Table 2, `Vmax18/21` / `Km18/21` |
| `fm_m3034`, `fm_m27` | 0.6, 0.3 (fixed constants) | Data S1, `FMET3034` / `FMET27` |
| `amp_diurnal`, `amp_diurnal_met` | 0.492, 0.225 | Table 2, `AMP` / `AMPMET` |
| `shift_diurnal` | 8.05 h | Table 2, `theta_SHIFT` |
| `boxcox_lvc` | 2.07 | Table 2, `lambda` |
| `lec50_nk1`, `lec50_nk3` | 0.97 ug/L, 9.7 ug/L (both fixed) | PK/PD modeling section (NK-1 from the PET study, reference 14; NK-3 assumed 10-fold) |
| IIV block on `etalcl`, `etalcl_met`, `etalvc` | 0.215 / 0.318 / 0.221 variances; 0.224, 0.164, 0.214 covariances | Table 2 OMEGA, `ETA(1,1)`, `ETA(2,2)`, `ETA(3,3)`, `ETA(2,1)`, `ETA(3,1)`, `ETA(3,2)` |
| `etalka`, `etaltlag` | 0.539, 0.0989 | Table 2 OMEGA, `ETA(4,4)`, `ETA(5,5)` |
| `expSd`, `expSd_c13`, `expSd_m3034`, `expSd_m27`, `expSd_m1821` | sqrt of 0.378, 0.122, 0.106, 0.143, 0.141 | Table 2 SIGMA, `EPS(1,1)` to `EPS(5,5)` |

# Structural verification

Two closed-form checks confirm the ODE system, the unit scaling, the
bioavailability and the dosing wiring before any population simulation.

## Terminal half-life

The paper states an elimination half-life of approximately 35 h
(Introduction). For the two-compartment elinzanetant disposition the terminal
rate constant is the smaller eigenvalue of the micro-constant system.

```{r halflife}
th <- setNames(ui$theta, names(ui$theta))
cl <- exp(th[["lcl"]]); vc <- exp(th[["lvc"]])
vp <- exp(th[["lvp"]]); q  <- exp(th[["lq"]])
k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
a1 <- k10 + k12 + k21
lambda_z <- (a1 - sqrt(a1^2 - 4 * k10 * k21)) / 2
thalf <- log(2) / lambda_z
c(`terminal half-life (h)` = round(thalf, 2), `paper (h)` = 35)
```

## Mass balance: AUC at steady state with the circadian term switched off

With a constant clearance, `AUC(0-24)ss` after once-daily dosing must equal
`Dose * F / CL` exactly. Setting the circadian amplitudes to zero and solving the
full ten-compartment system recovers that identity, which validates the dose
route, the bioavailability, the `mg` to `ug/L` unit bridge and the split of
elinzanetant clearance across `K20`, `K26` and `K28`.

```{r massbalance}
ss_events <- function(tfd, grid = 0.05) {
  ev <- rxode2::et(amt = 120, cmt = "depot", ii = 24, addl = 13) |>
    rxode2::et(seq(312, 336, by = grid), cmt = "central")
  d <- as.data.frame(ev)
  d$id <- 1L
  d$dvid <- ifelse(is.na(d$amt), 1L, NA_integer_)
  d$FED_HIGHFAT <- 0
  d$FORM_ELZ_HARDGEL <- 0
  d$T_FIRSTDOSE <- tfd
  d
}
trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

mod_flat <- ui |> rxode2::ini(amp_diurnal = 0, amp_diurnal_met = 0)
flat <- rxode2::rxSolve(mod_flat, ss_events(9), omega = NA,
                        atol = 1e-10, rtol = 1e-10, useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(time >= 312)

auc_flat <- trapz(flat$time, flat$Cc)
auc_analytic <- 120 * 1000 * exp(th[["lfdepot"]]) / cl
c(`simulated AUC(0-24)ss (ug*h/L)` = round(auc_flat, 1),
  `Dose * F / CL (ug*h/L)` = round(auc_analytic, 1),
  `relative difference (%)` = round(100 * (auc_flat / auc_analytic - 1), 3))
```

The two agree to better than 0.1%, so the structural implementation is exact.

# Replicate Figure 2: circadian modulation of clearance

Figure 2 of Willmann 2024 shows the model-estimated clearances against clock
time. The cosine reaches its maximum near 16:00 and its minimum near 04:00 for
both the parent and the metabolites, with relative amplitudes of 49.2% and
22.5%.

```{r figure-2}
amp <- th[["amp_diurnal"]]; ampm <- th[["amp_diurnal_met"]]
shift <- th[["shift_diurnal"]]
circ <- tidyr::crossing(
  clock = seq(0, 24, by = 0.1),
  analyte = c("Elinzanetant", "Metabolites (M27, M30/34, M18/21)")
) |>
  dplyr::mutate(
    a = ifelse(analyte == "Elinzanetant", amp, ampm),
    factor = 1 + a * cos(2 * pi / 24 * (clock + shift))
  )

extremes <- circ |>
  dplyr::group_by(analyte) |>
  dplyr::summarise(
    `clock time of maximum (h)` = clock[which.max(factor)],
    `clock time of minimum (h)` = clock[which.min(factor)],
    `relative amplitude (%)` = 100 * unique(a),
    .groups = "drop"
  )
knitr::kable(extremes, digits = 1,
             caption = "Timing and amplitude of the circadian clearance cosine.")

ggplot(circ, aes(clock, factor, colour = analyte)) +
  geom_line(linewidth = 0.9) +
  geom_vline(xintercept = c(4, 16), linetype = "dotted") +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Clock time (h)", y = "Clearance relative to its 24 h average",
       colour = NULL,
       caption = paste("Replicates Figure 2 of Willmann 2024.",
                       "Dotted lines mark 04:00 and 16:00.")) +
  theme(legend.position = "bottom")
```

# Covariate effects on absorption

The paper reports that a high-fat breakfast decreases Ka by 91%, increases the
absorption lag time by 55% and decreases bioavailability by 23%, and that the
hard gel capsule increases the lag time by 55% and decreases Ka by 26%
(Results and Discussion). The model returns `ka`, `tlag` and `fdepot` as derived
variables, so the encoded effects can be read back directly.

```{r covariate-effects}
cov_grid <- tibble::tribble(
  ~scenario,                          ~FED_HIGHFAT, ~FORM_ELZ_HARDGEL,
  "Reference (soft-gel, not high-fat)",          0,                 0,
  "High-fat breakfast",                          1,                 0,
  "Hard gel capsule",                            0,                 1
) |>
  dplyr::mutate(id = dplyr::row_number())

cov_events <- cov_grid |>
  tidyr::crossing(time = c(0, 1)) |>
  dplyr::mutate(
    amt = ifelse(time == 0, 120, NA_real_),
    evid = ifelse(time == 0, 1L, 0L),
    cmt = ifelse(time == 0, "depot", "central"),
    dvid = ifelse(time == 0, NA_integer_, 1L),
    T_FIRSTDOSE = 9
  ) |>
  dplyr::arrange(id, time)

cov_out <- rxode2::rxSolve(ui, cov_events, omega = NA, useLinCmt = FALSE,
                           keep = c("scenario")) |>
  as.data.frame() |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(ka = first(ka), tlag = first(tlag), fdepot = first(fdepot),
                   .groups = "drop")

ref <- cov_out |> dplyr::filter(scenario == "Reference (soft-gel, not high-fat)")
cov_out |>
  dplyr::mutate(
    `Change in Ka (%)` = 100 * (ka / ref$ka - 1),
    `Change in lag time (%)` = 100 * (tlag / ref$tlag - 1),
    `Change in F (%)` = 100 * (fdepot / ref$fdepot - 1)
  ) |>
  dplyr::rename("Scenario" = scenario, "Ka (1/h)" = ka,
                "Lag time (h)" = tlag, "F (fraction)" = fdepot) |>
  knitr::kable(digits = 3,
               caption = paste("Absorption parameters by feeding and",
                               "formulation scenario. Willmann 2024 reports a",
                               "91% decrease in Ka, a 55% increase in lag time",
                               "and a 23% decrease in F for a high-fat",
                               "breakfast, and a 26% decrease in Ka with a 55%",
                               "increase in lag time for the hard gel capsule."))
```

# Virtual cohort

Original observed data are not public. The cohort below mirrors the paper's
steady-state simulation: 1000 virtual women with VMS receiving 120 mg
once daily as the soft-gel capsule, dosed either at 09:00 or at 21:00. Two arms
of 200 subjects each are used here; 200 per arm is the nlmixr2lib cap and is
ample for a 90% prediction interval.

The two arms are **paired on identical random effects**: one set of 200 eta
vectors is drawn from the model's `OMEGA` and then reused for both dosing
times, so the only difference between the arms is `T_FIRSTDOSE`. This matters
for the ratio comparison below. Drawing each arm independently would leave the
evening-versus-morning ratios carrying the sampling noise of two separate
200-subject draws, which for a Cmax whose between-subject CV exceeds 60% is
several percent - comparable to the circadian effect being measured. Pairing
removes that noise entirely, so the ratios isolate the dosing-time effect. The
etas are supplied as explicit columns and `omega = NA` is passed to
`rxSolve()`, which makes the simulation reproducible without relying on
rxode2's internal sampling order.

```{r cohort}
set.seed(20241213)
n_per_arm <- 200L

# One shared draw of etas, reused by both arms.
om <- ui$omega
eta_names <- colnames(om)
etas <- matrix(rnorm(n_per_arm * length(eta_names)), nrow = n_per_arm) %*%
  chol(om)
colnames(etas) <- eta_names
etas <- as.data.frame(etas)

make_arm <- function(tfd, label, id_offset = 0L) {
  ev <- rxode2::et(amt = 120, cmt = "depot", ii = 24, addl = 13) |>
    rxode2::et(seq(312, 336, by = 0.25), cmt = "central")
  base <- as.data.frame(ev)
  base$id <- NULL
  base$dvid <- ifelse(is.na(base$amt), 1L, NA_integer_)
  do.call(rbind, lapply(seq_len(n_per_arm), function(i) {
    d <- base
    d$id <- id_offset + i
    d$FED_HIGHFAT <- 0
    d$FORM_ELZ_HARDGEL <- 0
    d$T_FIRSTDOSE <- tfd
    d$regimen <- label
    # subject index, so the paired arms can be matched up afterwards
    d$subject <- i
    for (nm in eta_names) d[[nm]] <- etas[[nm]][i]
    d
  }))
}

events <- dplyr::bind_rows(
  make_arm( 9, "Morning (09:00)", id_offset = 0L),
  make_arm(21, "Evening (21:00)", id_offset = n_per_arm)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(ui, events = events,
                       keep = c("regimen", "subject"),
                       omega = NA, useLinCmt = FALSE) |>
  as.data.frame() |>
  dplyr::filter(time >= 312) |>
  dplyr::mutate(
    tad = time - 312,
    clock = (time + T_FIRSTDOSE) %% 24,
    regimen = factor(regimen, levels = c("Morning (09:00)", "Evening (21:00)"))
  )
stopifnot(nrow(sim) > 0, !anyNA(sim$Cc))
```

# Replicate Figure 4: steady-state profiles and receptor occupancy

```{r figure-4-pk}
pk_band <- sim |>
  dplyr::group_by(regimen, tad) |>
  dplyr::summarise(
    lo = quantile(Cc, 0.05), med = median(Cc), hi = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(pk_band, aes(tad, med)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "pink", alpha = 0.5) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time after dose (h)", y = "Elinzanetant (ug/L)",
       caption = paste("Replicates the upper panels of Figure 4 of",
                       "Willmann 2024: median and 90% prediction interval at",
                       "steady state after 120 mg once daily."))
```

The morning-dosing median profile is non-monotonic: it falls to a minimum
around 11 h after the dose, rises again into the early morning hours as
clearance approaches its 04:00 trough, and then declines to the pre-dose value.
This is the distinctive signature of the circadian clearance term and is
reproduced from the model without any tuning. The evening-dosing profile is
very nearly monotonic by comparison, because the clearance minimum falls
shortly after the dose instead. The paper quantifies the same contrast as a
`Ctrough,ss`-versus-`Cmin,ss` difference of approximately 9% for morning dosing
against approximately 2% for evening dosing; this implementation returns
approximately 6% and approximately 1.5% respectively, reproducing both the
direction and the roughly four-fold asymmetry between the two dosing times
(published 4.4-fold, simulated 4.1-fold).

```{r figure-4-ro}
ro_band <- sim |>
  dplyr::select(regimen, tad, roNK1, roNK3) |>
  tidyr::pivot_longer(c(roNK1, roNK3), names_to = "receptor",
                      values_to = "ro") |>
  dplyr::mutate(receptor = ifelse(receptor == "roNK1", "NK-1", "NK-3")) |>
  dplyr::group_by(regimen, receptor, tad) |>
  dplyr::summarise(
    lo = 100 * quantile(ro, 0.05), med = 100 * median(ro),
    hi = 100 * quantile(ro, 0.95), .groups = "drop"
  )

ggplot(ro_band, aes(tad, med, colour = receptor, fill = receptor)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.25, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen) +
  coord_cartesian(ylim = c(50, 100)) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time after dose (h)", y = "Receptor occupancy (%)",
       colour = NULL, fill = NULL,
       caption = paste("Replicates the lower panels of Figure 4 of",
                       "Willmann 2024.")) +
  theme(legend.position = "bottom")
```

# Metabolite profiles

The paper shows metabolite fits only as goodness-of-fit and visual predictive
check figures (Figures S3-S10) and reports no numeric metabolite exposures, so
this panel is a structural demonstration rather than a numeric validation.

```{r metabolites}
sim |>
  dplyr::select(regimen, tad, Cc, Cc_m3034, Cc_m27, Cc_m1821) |>
  tidyr::pivot_longer(-c(regimen, tad), names_to = "analyte",
                      values_to = "conc") |>
  dplyr::mutate(analyte = factor(
    analyte,
    levels = c("Cc", "Cc_m3034", "Cc_m27", "Cc_m1821"),
    labels = c("Elinzanetant", "M30/34", "M27", "M18/21")
  )) |>
  dplyr::group_by(regimen, analyte, tad) |>
  dplyr::summarise(med = median(conc), .groups = "drop") |>
  ggplot(aes(tad, med, colour = analyte)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~regimen) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time after dose (h)", y = "Median concentration (ug/L)",
       colour = NULL,
       caption = paste("Steady-state median profiles of elinzanetant and its",
                       "three principal metabolites.")) +
  theme(legend.position = "bottom")
```

# PKNCA validation

The steady-state dosing interval is re-based to 0-24 h. The concentration at
time 0 is the steady-state pre-dose trough, not zero, so no zero-concentration
anchor row is injected here (that convention applies to single-dose data only).

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, regimen, time = tad, Cc) |>
  dplyr::arrange(id, time)

# Assert the interval anchor exists rather than fabricating it.
stopifnot(all(
  sim_nca |> dplyr::group_by(id) |> dplyr::summarise(has0 = any(time == 0)) |>
    dplyr::pull(has0)
))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | regimen + id)

dose_df <- sim_nca |>
  dplyr::distinct(id, regimen) |>
  dplyr::mutate(time = 0, amt = 120)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

intervals <- data.frame(
  start = 0, end = 24,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

`Ctrough,ss` in Willmann 2024 Table 3 is the concentration immediately before
the next dose, i.e. the value at 24 h in the re-based interval. It is computed
directly so that it can join the comparison alongside the PKNCA parameters.

```{r ctrough}
ctrough <- sim_nca |>
  dplyr::filter(time == 24) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(ctrough = median(Cc), .groups = "drop")
knitr::kable(ctrough, digits = 1,
             caption = "Median simulated Ctrough,ss by dosing time.")
```

## Comparison against published Table 3

```{r nca_compare}
published <- tibble::tribble(
  ~regimen,           ~cmax,  ~cmin, ~auclast,
  "Morning (09:00)",  1534.6, 158.8, 7553.1,
  "Evening (21:00)",  1529.7, 153.7, 8970.0
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "regimen",
  params = c("cmax", "cmin", "auclast"),
  units = c(cmax = "ug/L", cmin = "ug/L", auclast = "ug*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated versus published steady-state exposure",
                  "(Willmann 2024 Table 3, medians of 1000 virtual women",
                  "with VMS). * differs from the reference by more than 20%."),
  align = c("l", "l", "r", "r", "r")
)
```

## Relative quantities

The absolute exposures run below the published medians (see Errata). The
*relative* quantities in Table 3 - the dosing-time contrasts the paper actually
draws its conclusions from - are the more informative check, because they are
insensitive to any uniform offset in clearance or bioavailability. Because the
two arms are paired on identical etas, these ratios are free of between-arm
sampling noise.

All five agree with Table 3 to within 3%, and the evening-versus-morning `Cmax`
ratio - the one quantity the paper singles out as showing essentially no
dosing-time effect - is reproduced as 0.997 against a published 0.997. So the
circadian machinery, its phase, and its two amplitudes are all encoded
correctly; what does not reproduce is the absolute exposure level (next
section).

```{r ratios}
med <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) |>
  dplyr::group_by(regimen, PPTESTCD) |>
  dplyr::summarise(v = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = v) |>
  dplyr::left_join(ctrough, by = "regimen")

get <- function(df, reg, col) df[[col]][df$regimen == reg]
sim_ratio <- c(
  `AUC evening / morning` =
    get(med, "Evening (21:00)", "auclast") / get(med, "Morning (09:00)", "auclast"),
  `Ctrough evening / morning` =
    get(ctrough, "Evening (21:00)", "ctrough") / get(ctrough, "Morning (09:00)", "ctrough"),
  `Cmax evening / morning` =
    get(med, "Evening (21:00)", "cmax") / get(med, "Morning (09:00)", "cmax"),
  `Ctrough / Cmin, morning` =
    get(ctrough, "Morning (09:00)", "ctrough") / get(med, "Morning (09:00)", "cmin"),
  `Ctrough / Cmin, evening` =
    get(ctrough, "Evening (21:00)", "ctrough") / get(med, "Evening (21:00)", "cmin")
)
pub_ratio <- c(8970.0 / 7553.1, 156.7 / 172.7, 1529.7 / 1534.6,
               172.7 / 158.8, 156.7 / 153.7)

tibble::tibble(
  Quantity = names(sim_ratio),
  Simulated = as.numeric(sim_ratio),
  Published = pub_ratio,
  `Difference (%)` = 100 * (as.numeric(sim_ratio) / pub_ratio - 1)
) |>
  knitr::kable(digits = 3,
               caption = paste("Relative steady-state quantities, simulated",
                               "versus Willmann 2024 Table 3."))
```

# Assumptions and deviations

## Errata and discrepancies in the source

* **Table 3 absolute exposures are not reconcilable with the Table 2 typical
  values.** For once-daily dosing at steady state, `AUC(0-24)ss` must equal
  `Dose * F / CL` whenever clearance is constant; with the published
  `F = 0.367` and `CL = 7.26 L/h` that identity gives
  `120000 * 0.367 / 7.26 = 6066 ug*h/L`. The published median of
  `7553 ug*h/L` for morning dosing implies an effective clearance near
  5.8 L/h, roughly 20% below the reported typical value, and the evening value
  of `8970 ug*h/L` implies about 4.9 L/h. The circadian term cannot close the
  gap for morning dosing: with a 09:00 dose the concentration peak sits close
  to the 16:00 clearance maximum, so the circadian modulation moves
  `AUC(0-24)ss` *down* relative to `Dose * F / CL`, which is what this
  implementation reproduces. The mass-balance check above confirms the
  implementation returns the analytic identity to within 0.1% when the
  circadian amplitudes are set to zero, so the shortfall is a property of the
  published numbers rather than of the encoding. Two further observations
  localise it to the exposure *level* specifically. First, the shortfall is
  close to uniform rather than structured: simulated `Cmax` is 15.5% below the
  published median for *both* dosing times, and `Cmin` is 25% below for both,
  which is the signature of a single multiplicative offset in clearance or
  bioavailability rather than of a misencoded rate constant. Second, every
  dosing-time contrast in Table 3 is reproduced to within 3%, including the
  `Cmax` evening/morning ratio to three decimal places, so the circadian term
  and the disposition structure are demonstrably correct. The most likely
  explanation is
  that the paper's 1000 virtual women with VMS were built by resampling
  individual post-hoc parameter estimates from the female subpopulation rather
  than by drawing etas around the typical value - the paper describes exactly
  that resampling procedure for the intrinsic-factor analysis, and male
  subjects (who had significantly higher metabolite exposure-equivalent
  clearance) were excluded from the target population. No parameter has been
  tuned to close the gap.
* **Bilirubin covariate direction.** The Results and Discussion prose states
  that "women with bilirubin levels below the median were found to have
  approximately 15% higher elinzanetant exposure" and that "Low bilirubin
  levels were associated with a slightly (approximately 15%) higher
  elinzanetant exposure". The Figure 3 forest plot prints the opposite:
  `BILI >= median : BILI < median = 1.147 [90% CI 1.063-1.239]`, i.e. higher
  bilirubin is associated with roughly 15% *higher* exposure. The figure's
  direction is adopted in `covariatesDataExcluded` because it is the
  numerically printed result and because it is the direction consistent with
  the authors' own mechanistic reading, that the association "might reflect the
  direct causality between liver function and the hepatic metabolism of
  elinzanetant". Bilirubin is not in the final model, so this does not affect
  any prediction.
* **Absorption rate constant unit.** The Results text gives "a first-order
  absorption rate (Ka) of 2.37 L/h"; `L/h` is a typographical error for a
  first-order rate constant. Table 2 gives the unit as `h^-1`, which is what
  the model uses.
* **Absorption lag time in the Results text.** The Results sentence reads "an
  absorption lag time of approximately 12.29 a.m.", which is a typesetting
  corruption. Table 2 gives `ALAGREF = 0.292 h`, which is what the model uses.
* **Metabolite circadian amplitude wording.** The Discussion says "for the
  three metabolites, the relative amplitude was 22.5% smaller". Table 2
  describes `AMPMET = 0.225` as the "Amplitude of circadian variation for
  metabolite clearances and Vmax", and both the Results sentence ("49.2% and
  22.5%, respectively") and the later Discussion comparison ("49.2% or 22.5%,
  respectively") treat 22.5% as the absolute metabolite amplitude. The absolute
  reading is used.

## Modelling assumptions

* **Metabolite formation fractions are encoded as fixed.** `FMET27 = 0.3` and
  `FMET3034 = 0.6` appear only in Data S1, are described there as "constants",
  and carry no standard error or entry in Table 2, so they are wrapped in
  `fixed()`. The remaining 10% of elinzanetant clearance goes to unmodelled
  metabolites via `K20`.
* **Both EC50 values are fixed.** Neither was estimated in this analysis. The
  NK-1 value of 0.97 ug/L is carried from the elinzanetant PET study cited as
  reference 14 (repeated-dosing arm); the NK-3 value of 9.7 ug/L is that value
  multiplied by ten on the paper's stated assumption that the in vitro 10-fold
  affinity difference is maintained in vivo. Receptor occupancy is driven by
  the elinzanetant plasma concentration alone, as the paper specifies, even
  though the metabolites are described as pharmacologically similar.
* **Saturable metabolite elimination leaves the system.** Figure 1 draws the
  `Vmax`/`Km` arrow from each metabolite's central compartment to "metabolites",
  separately from the `K610` and `K810` arrows that form M18/21. The saturable
  term is therefore an additional elimination pathway, not an alternative route
  to M18/21.
* **The `[13C5]` tracer arm is retained.** It duplicates the parent disposition
  parameters but has its own residual error (`EPS(2,2)`) and its own
  compartments in Figure 1, and it is what identifies absolute bioavailability.
  Its metabolites were not measured, so the whole of its clearance leaves the
  system via `K30`. Dose it with `cmt = "central_c13"`.
* **Mass, not molar, transfer between analytes.** The paper's parameterisation
  transfers amounts directly between the parent and metabolite compartments
  with no molecular-weight correction, so the metabolite volumes and clearances
  are apparent values on the parent's mass scale. This is replicated as
  published.
* **Circadian anchor supplied as a covariate.** Willmann 2024 Equation 1 writes
  the cosine argument as `t + TFD + theta_SHIFT` with `t` the time since the
  first dose. The model therefore takes the per-subject clock time of the first
  dose as the covariate `T_FIRSTDOSE` and forms wall-clock time internally as
  `time + T_FIRSTDOSE`. A time-varying clock-time column cannot be used
  instead, because the cosine is evaluated inside the ODE at arbitrary solver
  steps and a 0-24 wrapping covariate cannot be interpolated across midnight.
* **Box-Cox IIV on Vc.** Data S1 gives
  `VC = VCpop * exp((exp(eta3)^lambda - 1) / lambda)`. Because
  `exp(eta3)^lambda` equals `exp(lambda * eta3)`, this is the standard
  Petersson 2009 Box-Cox eta transformation, and it is implemented in that
  form. It makes `etalvc` non-mu-referenced, which rxode2 reports as a warning
  when the model is parsed; the warning is expected and does not affect
  simulation.
* **Screened-but-unused covariates.** The twelve intrinsic factors that
  Willmann 2024 screened post hoc on model-derived `AUC(0-24)ss` but did not
  retain are documented in the model file's `covariatesDataExcluded` list with
  their Figure 3 geometric mean ratios and 90% confidence intervals. They are
  documentation only and are not referenced in `model()`. Vasomotor-symptom
  patient status (ratio 1.029 [0.954-1.109]) is recorded in the population
  notes instead, because no canonical covariate column exists for it and it
  would have required registering one for a documentation-only entry.
* **Cohort size and pairing.** The paper simulated 1000 virtual women per dosing
  time; this vignette uses 200 per arm, the nlmixr2lib cap, which is ample for
  the 90% prediction intervals shown. The two arms reuse one shared draw of
  etas so that the evening-versus-morning contrasts are paired. With
  independent draws the dosing-time ratios are dominated by sampling noise at
  this cohort size - the `Cmax` ratio moves by more than 10% between
  independent draws, which is larger than the effect being measured - so
  pairing is what makes the relative-quantity table above a meaningful check
  rather than an artefact.
* **Provenance comments cannot sit inside the `OMEGA` block.** rxode2 rewrites
  each in-`ini()` comment to `;` while parsing `label()` text, which is a
  syntax error inside the multi-line `c()` of a correlated-eta block. The
  six-element lower triangle of the 3x3 block is therefore annotated in a
  comment block immediately above it rather than element by element.
