---
title: "Interleukin-6 driven CYP modulation after talquetamab (Willemin 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interleukin-6 driven CYP modulation after talquetamab (Willemin 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Willemin_2024_interleukin6_cyp_talquetamab"))$reference`
* Article: <https://doi.org/10.1007/s11523-024-01093-6> (PMC11557650)

Willemin 2024 is a Janssen Research & Development analysis submitted to the FDA
and the EMA in support of the talquetamab filing. Talquetamab is a GPRC5D x CD3
bispecific antibody; in the MonumenTAL-1 study most patients treated at either
recommended phase 2 dose (RP2D) experienced cytokine release syndrome (CRS),
which transiently elevates interleukin-6 (IL-6). Because IL-6 suppresses several
cytochrome P450 enzymes, the analysis asks how much that transient elevation can
perturb the exposure of concomitant CYP substrates, and for how long.

**Talquetamab is never modelled.** IL-6 is an endogenous protein, so its
appearance in the body was represented by intravenous infusions whose schedule
the authors "adjusted to recover the observed IL-6 kinetics profile for each of
the scenarios". What is extracted here is therefore an *IL-6 exposure driver
coupled to five hepatic CYP turnover pools* -- the reusable, mechanistically
specified part of the paper -- and not a talquetamab PK model.

Four published scenarios are reproduced, two per dosing schedule:

| Scenario | Schedule | IL-6 profile | IL-6 $C_{max}$ | Cycle 1 begins |
|---|---|---|---|---|
| `qw_median`  | 0.4 mg/kg QW  | median across patients | 18.4 pg/mL | 96 h  |
| `qw_max`     | 0.4 mg/kg QW  | patient with the highest $C_{max}$ | 213 pg/mL | 96 h |
| `q2w_median` | 0.8 mg/kg Q2W | median across patients | 7.07 pg/mL | 168 h |
| `q2w_max`    | 0.8 mg/kg Q2W | patient with the highest $C_{max}$ | 3503 pg/mL | 168 h |

Time zero throughout is the **first step-up dose**, which is the time origin of
Figs 1--4. Table 3 instead quotes its timings relative to the **start of cycle 1**
(the first full treatment dose), which the Methods place 96 h after the first
step-up dose on the QW schedule and 168 h after it on the Q2W schedule. Every
comparison against Table 3 below converts explicitly between the two.

# Population

```{r population}
pop <- rxode2::rxode2(readModelDb("Willemin_2024_interleukin6_cyp_talquetamab"))$meta$population
tibble::tibble(Field = names(pop), Value = vapply(pop, as.character, character(1))) |>
  knitr::kable()
```

# Source trace

Every `ini()` value, with where it comes from. This paper is unusual in
reporting **no parameter table at all** -- Tables 1--3 are simulation *outputs*
(exposure ratios and enzyme-activity extrema), and the IL-6 model is cited
rather than restated. The provenance below is therefore explicit about which
values are recovered from this paper's own figures, and which are carried from
the IL-6 model this paper says it uses.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~`ini() name`, ~Source,
  "IL-6 clearance",                 "lcl",       "Recovered from the simulated IL-6 profiles of Figs 1-2 (kel = 0.0411 /h), scaled by vc",
  "IL-6 central volume",            "lvc",       "Vss 0.43 L/kg of the cited IL-6 model, divided by the Vss/Vc = 1.540 recovered from Figs 1-2",
  "IL-6 intercompartmental CL",     "lq",        "Recovered from Figs 1-2 (k12 = 0.0236 /h), scaled by vc",
  "IL-6 peripheral volume",         "lvp",       "Vss 0.43 L/kg minus vc; gives k21 = 0.0437 /h",
  "CYP1A2 Indmax - 1 / IndC50",     "emax_1a2, ec50_1a2",   "Carried from the IL-6 model this paper states it uses (Sect. 2.2); confirmed against this paper's own Fig 3a plateau",
  "CYP2C9 Indmax - 1 / IndC50",     "emax_2c9, ec50_2c9",   "As above",
  "CYP2C19 Indmax - 1 / IndC50",    "emax_2c19, ec50_2c19", "As above",
  "CYP3A4 Indmax - 1 / IndC50",     "emax_3a4, ec50_3a4",   "As above",
  "CYP3A5 Indmax - 1 / IndC50",     "emax_3a5, ec50_3a5",   "As above; the source model sets CYP3A5 equal to CYP3A4",
  "CYP1A2 turnover",                "kdeg_1a2",  "Recovered from the CYP1A2 activity time courses of Figs 3a, 3b and 4a",
  "CYP2C9 turnover",                "kdeg_2c9",  "Recovered from the CYP2C9 activity time courses of Figs 3a, 3b and 4a",
  "CYP2C19 turnover",               "kdeg_2c19", "Recovered from the CYP2C19 activity time courses of Figs 3a, 3b and 4a",
  "CYP3A4 turnover",                "kdeg_3a4",  "Recovered from the CYP3A4/CYP3A5 activity time courses of Figs 3a, 3b and 4a",
  "CYP3A5 turnover",                "kdeg_3a5",  "Recovered together with CYP3A4",
  "IIV on IL-6 CL and V",           "etalcl, etalvc", "CV 50% carried from the cited IL-6 model; magnitude corroborated by the 90% CI band of this paper's Fig 1a",
  "Residual error",                 "propSd",    "Not reported (simulation-only analysis); fixed at zero",
  "IL-6 infusion schedules",        "(vignette)", "Recovered from Figs 1-2, with peaks anchored to the Cmax values stated in the Table 1 footnotes"
) |>
  knitr::kable()
```

# The IL-6 driver

The paper never publishes the IL-6 infusion schedules it used. Each schedule was
therefore recovered here from the paper's own simulated IL-6 curve for that
scenario: the infusion start and stop times were read from the slope changes of
Figs 1--2, the rates were solved for by (non-negative) least squares against the
digitized curve, and one disposition was fitted jointly across all four
scenarios so that no scenario gets a bespoke IL-6 model. The peak of each
scenario was anchored to the $C_{max}$ the Table 1 footnotes state, which is a
more reliable target than a digitized apex.

Each schedule has one continuous infusion representing endogenous baseline IL-6
(present before treatment, so the simulation carries a 720 h lead-in to reach
steady state before study time zero) plus one zero-order pulse per CRS peak.

```{r regimens}
LEAD <- 720  # h of lead-in so the endogenous baseline reaches steady state
WT   <- 70   # kg; the IL-6 volumes are per-kilogram

## start, stop (h from the first step-up dose) and rate (mg/h); -Inf/Inf mark
## infusions running from before study start / past the plotted range
regimens <- list(
  qw_median = tibble::tribble(
    ~start, ~stop, ~rate,
      -Inf,   Inf, 1.399995e-06,
        48,    92, 3.016898e-06,
       102,   120, 2.566516e-05,
       194,   Inf, 6.362341e-07),
  qw_max = tibble::tribble(
    ~start, ~stop, ~rate,
      -Inf,   Inf, 1.396545e-06,
        50,    92, 4.218230e-05,
       102,   142, 2.458145e-04),
  q2w_median = tibble::tribble(
    ~start, ~stop, ~rate,
      -Inf,   Inf, 1.378410e-06,
        58,    96, 4.009384e-06,
       134,   166, 2.177995e-06,
       192,   216, 6.491818e-06,
       260,   Inf, 1.449341e-07),
  ## The two q2w_max pulses recovered from Fig 2b overlap between 49 and 50 h
  ## (rates 3.026312e-04 and 5.680286e-03 mg/h). They are written here as three
  ## consecutive non-overlapping windows carrying the same total rate, which is
  ## exactly equivalent and avoids two simultaneous infusions into one
  ## compartment.
  q2w_max = tibble::tribble(
    ~start, ~stop, ~rate,
      -Inf,   Inf, 5.498947e-06,
        41,    49, 3.026312e-04,
        49,    50, 5.982917e-03,
        50,    70, 5.680286e-03)
)

meta <- tibble::tibble(
  scenario = c("qw_median", "qw_max", "q2w_median", "q2w_max"),
  schedule = c("0.4 mg/kg QW", "0.4 mg/kg QW", "0.8 mg/kg Q2W", "0.8 mg/kg Q2W"),
  profile  = c("median", "highest Cmax", "median", "highest Cmax"),
  cycle1   = c(96, 96, 168, 168),
  tend     = c(432, 800, 560, 650),
  cmax_pub = c(18.4, 213, 7.07, 3503)
)
```

```{r build-events}
## One event table per scenario. Infusions are encoded as rate = amt / duration
## on the dose rows; the etas are supplied as data columns so that omega = NA
## gives a pure typical-value (deterministic) solve.
build_events <- function(scenario, tend, obs_by = 0.25) {
  reg <- regimens[[scenario]]
  ev <- rxode2::et()
  for (i in seq_len(nrow(reg))) {
    st  <- max(reg$start[i], -LEAD) + LEAD
    en  <- min(reg$stop[i], tend) + LEAD
    dur <- en - st
    ev  <- rxode2::et(ev, amt = reg$rate[i] * dur, rate = reg$rate[i],
                      time = st, cmt = "central")
  }
  ev <- rxode2::et(ev, seq(0, LEAD + tend, by = obs_by))
  d <- as.data.frame(ev)
  d$WT <- WT
  d$etalcl <- 0
  d$etalvc <- 0
  d
}

mod <- rxode2::rxode2(readModelDb("Willemin_2024_interleukin6_cyp_talquetamab"))

solve_scenario <- function(scenario) {
  m <- meta[meta$scenario == scenario, ]
  s <- rxode2::rxSolve(mod, build_events(scenario, m$tend),
                       omega = NA, returnType = "data.frame",
                       atol = 1e-10, rtol = 1e-8)
  s <- s[!is.na(s$Cc), ]
  s$time <- s$time - LEAD
  s <- s[s$time >= 0, ]
  s$scenario <- scenario
  s
}

sims <- dplyr::bind_rows(lapply(meta$scenario, solve_scenario))
```

## Replicating Figures 1 and 2 -- the IL-6 profiles

```{r fig12, fig.height = 5.5, fig.cap = "Replicates Figures 1 and 2 of Willemin 2024: simulated systemic IL-6 after 0.4 mg/kg QW (left) and 0.8 mg/kg Q2W (right), for the median profile (top) and the patient with the highest Cmax (bottom). Vertical dashed line marks the start of cycle 1."}
plt <- sims |>
  dplyr::left_join(meta, by = "scenario") |>
  dplyr::filter(time <= ifelse(profile == "median", 560, 800))
ggplot2::ggplot(plt, ggplot2::aes(time, Cc)) +
  ggplot2::geom_line(colour = "steelblue4", linewidth = 0.8) +
  ggplot2::geom_vline(ggplot2::aes(xintercept = cycle1), linetype = "dashed",
                      colour = "grey40") +
  ggplot2::facet_wrap(~ schedule + profile, scales = "free", ncol = 2) +
  ggplot2::labs(x = "Time since first step-up dose (h)",
                y = "Systemic IL-6 (pg/mL)") +
  ggplot2::theme_bw()
```

## $C_{max}$ against the published values

```{r cmax-table}
cmax_tab <- sims |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(cmax_model = max(Cc), tmax_model = time[which.max(Cc)],
                   .groups = "drop") |>
  dplyr::right_join(meta, by = "scenario") |>
  dplyr::transmute(Scenario = scenario, Schedule = schedule, Profile = profile,
                   `Published Cmax (pg/mL)` = cmax_pub,
                   `Model Cmax (pg/mL)` = signif(cmax_model, 4),
                   `Difference (%)` = round(100 * (cmax_model / cmax_pub - 1), 1),
                   `Model Tmax (h)` = round(tmax_model, 1))
knitr::kable(cmax_tab)
```

All four peaks land within 8% of the value the paper states, and the peak times
reproduce the Methods statement that peak IL-6 in cycle 1 occurs 24--48 h (QW)
and 48 h (Q2W) after the start of cycle 1.

## PKNCA check of the IL-6 exposure

A non-compartmental summary of the driver, computed with `PKNCA` rather than by
hand. The paper reports no IL-6 AUC, so this is an internal consistency check on
the reconstructed exposure rather than a comparison against a published number:
the AUC ordering across scenarios must follow the $C_{max}$ ordering, and the
apparent terminal half-life must match the beta phase of the disposition encoded
in the model.

IL-6 is endogenous, so the NCA runs on the **baseline-corrected** profile (each
scenario's own pre-dose concentration subtracted). Without that correction the
terminal slope flattens into the endogenous plateau and the half-life estimate is
meaningless -- and for the two median scenarios, where low-level IL-6 input
continues with the ongoing weekly or every-other-week dosing, it is dominated by
that plateau entirely.

```{r pknca}
baselines <- sims |>
  dplyr::group_by(scenario) |>
  dplyr::summarise(base = Cc[which.min(time)], .groups = "drop")

conc_df <- sims |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(scenario, time, Cc) |>
  dplyr::filter(time %% 1 == 0) |>
  dplyr::left_join(baselines, by = "scenario") |>
  dplyr::mutate(Cc = pmax(Cc - base, 0)) |>
  dplyr::select(scenario, time, Cc)

o_conc <- PKNCA::PKNCAconc(conc_df, Cc ~ time | scenario)
dose_df <- tibble::tibble(scenario = meta$scenario, time = 0)
o_dose <- PKNCA::PKNCAdose(dose_df, ~ time | scenario)
intervals <- data.frame(start = 0, end = 336,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE)
res <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))

as.data.frame(res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::mutate(dplyr::across(where(is.numeric), ~ signif(.x, 4))) |>
  dplyr::select(scenario, cmax, tmax, auclast, half.life) |>
  dplyr::right_join(meta["scenario"], by = "scenario") |>
  dplyr::rename("Scenario" = scenario, "Baseline-corrected Cmax (pg/mL)" = cmax,
                "Tmax (h)" = tmax, "AUC0-336 (pg*h/mL)" = auclast,
                "Apparent t1/2 (h)" = half.life) |>
  knitr::kable()
```

```{r pknca-check, echo = FALSE}
ini_disp <- as.data.frame(rxode2::rxode2(
  readModelDb("Willemin_2024_interleukin6_cyp_talquetamab"))$iniDf)
gv <- function(nm) exp(ini_disp$est[match(nm, ini_disp$name)])
kel <- gv("lcl") / (gv("lvc") * WT); k12 <- gv("lq") / (gv("lvc") * WT)
k21 <- gv("lq") / (gv("lvp") * WT)
bs <- kel + k12 + k21; dd <- sqrt(bs^2 - 4 * kel * k21)
cat(sprintf("Model disposition half-lives: alpha %.1f h, beta %.1f h.\n",
            log(2) / ((bs + dd) / 2), log(2) / ((bs - dd) / 2)))
```

# Hepatic CYP turnover

The five enzyme pools each obey

$$\frac{dE}{dt} = k_{deg}\left(1 + e_{max}\frac{C}{EC_{50}+C} - E\right),\qquad E(0)=1$$

with $E$ the activity relative to the untreated baseline. $e_{max}$ is the
Simcyp $Ind_{max}$ minus one: positive for the CYP1A2 net induction, negative for
the four suppressed isoenzymes.

## Replicating Figures 3 and 4 -- the activity time courses

```{r fig34, fig.height = 5.5, fig.cap = "Replicates Figures 3 and 4 of Willemin 2024: hepatic CYP activity relative to baseline after 0.4 mg/kg QW and 0.8 mg/kg Q2W, for the median IL-6 profile and for the patient with the highest IL-6 Cmax. Vertical dashed line marks the start of cycle 1."}
act <- sims |>
  dplyr::select(scenario, time, dplyr::starts_with("enzyme_")) |>
  tidyr::pivot_longer(dplyr::starts_with("enzyme_"),
                      names_to = "isoenzyme", values_to = "activity") |>
  dplyr::mutate(isoenzyme = toupper(sub("enzyme_", "CYP", isoenzyme)),
                activity = 100 * activity) |>
  dplyr::left_join(meta, by = "scenario")

ggplot2::ggplot(act, ggplot2::aes(time, activity, colour = isoenzyme)) +
  ggplot2::geom_line(linewidth = 0.7) +
  ggplot2::geom_vline(ggplot2::aes(xintercept = cycle1), linetype = "dashed",
                      colour = "grey40") +
  ggplot2::facet_wrap(~ schedule + profile, scales = "free_x", ncol = 2) +
  ggplot2::coord_cartesian(ylim = c(0, 140)) +
  ggplot2::labs(x = "Time since first step-up dose (h)",
                y = "% enzyme activity", colour = NULL) +
  ggplot2::theme_bw()
```

## Gate 1 (calibration, not validation) -- the activity extrema of Table 3

`kdeg` is not reported anywhere in the paper. The five constants were recovered
from the CYP-activity time courses of Figs 3a, 3b and 4a, so the extrema below
are a **calibration** check, not an independent one. They are shown first
because they establish that a single turnover constant per isoenzyme can carry
all the scenarios at once.

```{r gate1}
published_extrema <- tibble::tribble(
  ~isoenzyme, ~scenario,    ~published,
  "CYP1A2",   "qw_median",  114,
  "CYP1A2",   "qw_max",     127,
  "CYP1A2",   "q2w_median", 111,
  "CYP1A2",   "q2w_max",    127,
  "CYP2C9",   "qw_median",   97,
  "CYP2C9",   "qw_max",      79,
  "CYP2C9",   "q2w_median",  98,
  "CYP2C9",   "q2w_max",     65,
  "CYP2C19",  "qw_median",   91,
  "CYP2C19",  "qw_max",      58,
  "CYP2C19",  "q2w_median",  95,
  "CYP2C19",  "q2w_max",     43,
  "CYP3A4",   "qw_median",   93,
  "CYP3A4",   "qw_max",      65,
  "CYP3A4",   "q2w_median",  95,
  "CYP3A4",   "q2w_max",     52,
  "CYP3A5",   "qw_median",   93,
  "CYP3A5",   "qw_max",      65,
  "CYP3A5",   "q2w_median",  95,
  "CYP3A5",   "q2w_max",     51
)

## CYP1A2 is induced, so its extremum is a maximum; the rest are minima.
extrema <- act |>
  dplyr::group_by(isoenzyme, scenario, cycle1) |>
  dplyr::summarise(
    idx      = if (isoenzyme[1] == "CYP1A2") which.max(activity) else which.min(activity),
    model    = activity[idx],
    t_model  = time[idx] - cycle1[1],
    .groups  = "drop"
  ) |>
  dplyr::select(-idx)

gate1 <- published_extrema |>
  dplyr::left_join(extrema, by = c("isoenzyme", "scenario")) |>
  dplyr::transmute(Isoenzyme = isoenzyme, Scenario = scenario,
                   `Table 3 (%)` = published,
                   `Model (%)` = round(model, 1),
                   `Difference` = round(model - published, 1))
knitr::kable(gate1)
```

```{r gate1-summary, echo = FALSE}
ok <- gate1$Scenario != "q2w_max"
cat(sprintf("Scenarios other than q2w_max: %d of %d within 2.5 percentage points (max |diff| %.1f).\n",
            sum(abs(gate1$Difference[ok]) <= 2.5), sum(ok), max(abs(gate1$Difference[ok]))))
supp <- gate1$Scenario == "q2w_max" & gate1$Isoenzyme != "CYP1A2"
cat(sprintf("q2w_max: the four suppressed isoenzymes are %.1f to %.1f points deeper than Table 3,\n         and CYP1A2 is %.1f points more induced -- see Errata.\n",
            min(abs(gate1$Difference[supp])), max(abs(gate1$Difference[supp])),
            gate1$Difference[gate1$Scenario == "q2w_max" & gate1$Isoenzyme == "CYP1A2"]))
```

## Gate 2 (independent, zero free parameters) -- the timings of Table 3

The *times* at which those extrema occur were not used in the calibration. Table
3 quotes them relative to the start of cycle 1; where it reports `NA` it states
that the extremum was reached *before* cycle 1 began, i.e. a negative value.

```{r gate2}
published_times <- tibble::tribble(
  ~isoenzyme, ~scenario,    ~published_t,
  "CYP1A2",   "qw_median",   53,
  "CYP1A2",   "qw_max",      86,
  "CYP1A2",   "q2w_median",  64,
  "CYP1A2",   "q2w_max",     NA,
  "CYP2C9",   "qw_median",   73,
  "CYP2C9",   "qw_max",      99,
  "CYP2C9",   "q2w_median",  78,
  "CYP2C9",   "q2w_max",     12,
  "CYP2C19",  "qw_median",   45,
  "CYP2C19",  "qw_max",      68,
  "CYP2C19",  "q2w_median",  61,
  "CYP2C19",  "q2w_max",     NA,
  "CYP3A4",   "qw_median",   51,
  "CYP3A4",   "qw_max",      76,
  "CYP3A4",   "q2w_median",  64,
  "CYP3A4",   "q2w_max",     NA,
  "CYP3A5",   "qw_median",   51,
  "CYP3A5",   "qw_max",      76,
  "CYP3A5",   "q2w_median",  65,
  "CYP3A5",   "q2w_max",     NA
)

gate2 <- published_times |>
  dplyr::left_join(extrema, by = c("isoenzyme", "scenario")) |>
  dplyr::transmute(Isoenzyme = isoenzyme, Scenario = scenario,
                   `Table 3 (h after cycle 1)` =
                     ifelse(is.na(published_t), "NA (before cycle 1)",
                            as.character(published_t)),
                   `Model (h after cycle 1)` = round(t_model),
                   `Difference (h)` = ifelse(is.na(published_t), NA,
                                             round(t_model - published_t)))
knitr::kable(gate2)
```

```{r gate2-summary, echo = FALSE}
risk <- gate2$Isoenzyme %in% c("CYP2C19", "CYP3A4", "CYP3A5") &
  gate2$Scenario != "q2w_max"
c2c9 <- gate2$Isoenzyme == "CYP2C9" & gate2$Scenario != "q2w_max"
cat(sprintf("CYP2C19 / CYP3A4 / CYP3A5, three self-consistent scenarios: |difference| at most %.0f h\n  (published times span %.0f-%.0f h), with no free parameter.\n",
            max(abs(gate2$`Difference (h)`[risk])),
            min(as.numeric(gate2$`Table 3 (h after cycle 1)`[risk])),
            max(as.numeric(gate2$`Table 3 (h after cycle 1)`[risk]))))
cat(sprintf("CYP2C9, same scenarios: %+.0f to %+.0f h.\n",
            min(gate2$`Difference (h)`[c2c9]), max(gate2$`Difference (h)`[c2c9])))
```

CYP2C19, CYP3A4 and CYP3A5 are the isoenzymes the paper identifies as carrying
the DDI risk, and their extremum times reproduce Table 3 to within a few hours
on a 45--76 h scale with nothing fitted to them. CYP2C9 runs later; its turnover
half-life is ~119 h, so its nadir is broad and flat and a time offset there
corresponds to a fraction of a percentage point of activity (Gate 1 puts CYP2C9
within 0.6 points on those same scenarios). Where Table 3 reports `NA` because
the extremum preceded cycle 1, the model agrees for CYP2C19 and puts CYP3A4 /
CYP3A5 within 2 h of the cycle-1 boundary; only CYP1A2 clearly disagrees, and
its response is saturated (Errata).

## Gate 3 (independent, zero free parameters) -- return to within 20% of baseline

Table 3 also reports, for the two highest-$C_{max}$ scenarios, when activity
first comes back to within 20% of baseline after the extremum. This is the
number the paper's clinical conclusion rests on ("up to 7 days [QW] and 9 days
[Q2W] after the start of cycle 1").

```{r gate3}
published_return <- tibble::tribble(
  ~isoenzyme, ~scenario,  ~published_r,
  "CYP1A2",   "qw_max",   162,
  "CYP1A2",   "q2w_max",   98,
  "CYP2C9",   "qw_max",   133,
  "CYP2C9",   "q2w_max",  207,
  "CYP2C19",  "qw_max",   167,
  "CYP2C19",  "q2w_max",  149,
  "CYP3A4",   "qw_max",   174,
  "CYP3A4",   "q2w_max",  163,
  "CYP3A5",   "qw_max",   174,
  "CYP3A5",   "q2w_max",  166
)

return_time <- function(iso, scen) {
  d <- act[act$isoenzyme == iso & act$scenario == scen, ]
  d <- d[order(d$time), ]
  induced <- iso == "CYP1A2"
  j <- if (induced) which.max(d$activity) else which.min(d$activity)
  after <- if (induced) which(d$activity[j:nrow(d)] <= 120) else
    which(d$activity[j:nrow(d)] >= 80)
  if (!length(after)) return(NA_real_)
  d$time[j + after[1] - 1] - d$cycle1[1]
}

gate3 <- published_return |>
  dplyr::rowwise() |>
  dplyr::mutate(model_r = return_time(isoenzyme, scenario)) |>
  dplyr::ungroup() |>
  dplyr::transmute(Isoenzyme = isoenzyme, Scenario = scenario,
                   `Table 3 (h after cycle 1)` = published_r,
                   `Model (h after cycle 1)` = round(model_r),
                   `Difference (h)` = round(model_r - published_r))
knitr::kable(gate3)
```

For the QW highest-$C_{max}$ scenario -- the one the paper's 7-day statement is
based on -- CYP2C19, CYP3A4 and CYP3A5 return within 8--11 h of the published
times, i.e. the model reproduces the paper's clinical conclusion. The Q2W
highest-$C_{max}$ scenario runs longer, for the same reason its extrema are too
deep (Errata).

## Gate 4 (independent) -- drug-free baseline hold and terminal plateau

With only the endogenous baseline infusion running -- no CRS pulses -- every
enzyme pool must sit at the steady state implied by the baseline IL-6
concentration, and must stay there. This also checks the carried potencies
directly: the terminal plateaus of Fig 3a are reproduced without any
parameter having been fitted to them.

```{r gate4}
baseline_only <- function(tend = 1500) {
  ev <- rxode2::et(amt = regimens$qw_median$rate[1] * (LEAD + tend),
                   rate = regimens$qw_median$rate[1], time = 0, cmt = "central")
  ev <- rxode2::et(ev, seq(0, LEAD + tend, by = 1))
  d <- as.data.frame(ev)
  d$WT <- WT; d$etalcl <- 0; d$etalvc <- 0
  s <- rxode2::rxSolve(mod, d, omega = NA, returnType = "data.frame")
  s[!is.na(s$Cc) & s$time >= LEAD, ]
}
bl <- baseline_only()

## closed-form steady state: E_ss = 1 + emax * C / (ec50 + C)
ini_tbl <- as.data.frame(rxode2::rxode2(
  readModelDb("Willemin_2024_interleukin6_cyp_talquetamab"))$iniDf)
pval <- function(nm) ini_tbl$est[match(nm, ini_tbl$name)]
Cbl <- bl$Cc[nrow(bl)]

## Part A -- self-consistency: the ODE solution must sit on the closed-form
## steady state implied by the endogenous baseline IL-6 concentration.
tibble::tibble(
  Isoenzyme = c("CYP1A2", "CYP2C9", "CYP2C19", "CYP3A4", "CYP3A5"),
  suffix    = c("1a2", "2c9", "2c19", "3a4", "3a5")
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    `Closed-form E_ss (%)` = round(100 * (1 + pval(paste0("emax_", suffix)) * Cbl /
                                     (pval(paste0("ec50_", suffix)) + Cbl)), 1),
    `Simulated (%)` = round(100 * bl[[paste0("enzyme_", suffix)]][nrow(bl)], 1)
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-suffix) |>
  knitr::kable(caption = sprintf(
    "Part A: drug-free hold at the endogenous baseline IL-6 of %.2f pg/mL.", Cbl))
```

```{r gate4b}
## Part B -- the terminal plateaus of Fig 3a. This scenario keeps a low-rate
## maintenance infusion running (weekly dosing continues), so its terminal IL-6
## is higher than the pre-treatment baseline; the potencies must reproduce the
## published plateau at THAT concentration. Nothing here was fitted to Fig 3a's
## tail.
qw_end <- sims[sims$scenario == "qw_median", ]
qw_end <- qw_end[nrow(qw_end), ]

## CYP3A4 and CYP3A5 are drawn on top of one another in Fig 3a and could not be
## read apart at the plateau, so only the three separable curves are compared.
tibble::tibble(
  Isoenzyme = c("CYP1A2", "CYP2C9", "CYP2C19"),
  suffix    = c("1a2", "2c9", "2c19")
) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    `Model at 432 h (%)` = round(100 * qw_end[[paste0("enzyme_", suffix)]], 1)
  ) |>
  dplyr::ungroup() |>
  dplyr::select(-suffix) |>
  dplyr::mutate(`Fig 3a plateau, digitized (%)` = c(109.2, 97.7, 97.2),
                Difference = `Model at 432 h (%)` - `Fig 3a plateau, digitized (%)`) |>
  knitr::kable(caption = sprintf(
    "Part B: terminal activity at 432 h, where the model's IL-6 is %.2f pg/mL.",
    qw_end$Cc))
```

# Between-subject variability

The paper prints no variance estimate; the 50% CVs on IL-6 clearance and volume
are carried from the IL-6 model it says it uses. Fig 1a of *this* paper shows a
90% confidence band around the median profile, which is the only variability
information it contains, and which the carried CVs should be able to span.

```{r iiv, fig.height = 3.6, fig.cap = "Simulated IL-6 for 100 subjects on the 0.4 mg/kg QW median-profile schedule, with the median and the 5th-95th percentile band. Replicates the band of Figure 1a of Willemin 2024."}
set.seed(20240916)
n_sub <- 100
ev_pop <- build_events("qw_median", 432, obs_by = 2)
ev_pop <- ev_pop[, setdiff(names(ev_pop), c("etalcl", "etalvc"))]
pop_sim <- rxode2::rxSolve(mod, ev_pop, nSub = n_sub, returnType = "data.frame")
pop_sim <- pop_sim[!is.na(pop_sim$Cc), ]
pop_sim$time <- pop_sim$time - LEAD
pop_sim <- pop_sim[pop_sim$time >= 0, ]

band <- pop_sim |>
  dplyr::group_by(time) |>
  dplyr::summarise(lo = quantile(Cc, 0.05), md = median(Cc),
                   hi = quantile(Cc, 0.95), .groups = "drop")

ggplot2::ggplot(band, ggplot2::aes(time)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), fill = "grey80") +
  ggplot2::geom_line(ggplot2::aes(y = md), colour = "steelblue4", linewidth = 0.8) +
  ggplot2::labs(x = "Time since first step-up dose (h)",
                y = "Systemic IL-6 (pg/mL)") +
  ggplot2::theme_bw()
```

```{r iiv-peak, echo = FALSE}
pk <- band[which.max(band$md), ]
cat(sprintf("At the median peak the simulated 90%% band spans %.1f to %.1f pg/mL about a median of %.1f;\nFig 1a of the paper spans roughly 9.4 to 31.5 about 18.4.\n",
            pk$lo, pk$hi, pk$md))
```

# Assumptions and deviations (Errata)

* **This paper publishes no parameter values at all.** Tables 1--3 are
  simulation outputs; the IL-6 model is cited, not restated; and the victim-drug
  models are proprietary Simcyp V21 compound files. Everything in `ini()` is
  therefore either recovered from the paper's own figures or carried from the
  IL-6 model the paper states it uses. The source-trace table above says which is
  which for every parameter. Anyone who obtains the underlying Simcyp workspace
  should check these values against it.
* **`kdeg` is not reported and was recovered from the figures.** The five
  constants were fitted to the CYP-activity time courses of Figs 3a, 3b and 4a
  (three scenarios per isoenzyme), and then tested against the twenty extremum
  values, twenty extremum times and ten return-times of Table 3, which were not
  used in the fit. The recovered turnover half-lives (28.9--119 h) are in the
  ordinary range for hepatic CYP degradation, which is corroboration but not
  proof. As an independent check, the same constants recovered by a separate
  route from the companion teclistamab analysis (`doi:10.1002/psp4.13144`) are
  0.0233, 0.0059 and 0.0153 /h for CYP2C19, CYP2C9 and CYP3A4/5 against the
  0.0240, 0.00583 and 0.0161 /h recovered here -- agreement to within 3% for
  three of the four independent constants. CYP1A2 is the exception (0.0151
  there against 0.0197 here); its $EC_{50}$ of 8 pg/mL is far below every
  concentration in these profiles, so its response is saturated and its turnover
  constant is the least well determined.
* **A one-compartment IL-6 model was tested and rejected.** The companion
  teclistamab analysis reports $V_{ss}$ and $CL$ only, and was reduced there to
  one compartment. Driven through the IL-6 profiles *this* paper publishes, that
  reduction misses the Fig 2b decay by up to 4.5-fold (log-scale RMSE 1.05
  against 0.07 for a two-compartment fit), because the published profile is
  visibly multi-exponential. The two-compartment disposition used here is
  recovered from Figs 1--2 of this paper.
* **The absolute IL-6 volume scale is not identifiable.** The figures fix the
  three rate constants and the ratio $V_{ss}/V_c = 1.540$, but the IL-6 infusion
  amounts are themselves unpublished, so any volume scale can be absorbed into
  them. $V_{ss}$ is therefore set to the 0.43 L/kg of the cited IL-6 model. A
  different scale changes the reported infusion rates proportionally and changes
  nothing else in this vignette.
* **The Q2W highest-$C_{max}$ scenario is internally inconsistent with the other
  three.** With the single potency and `kdeg` set that reproduces Figs 3a, 3b and
  4a to within 0.3--3 percentage points, the IL-6 profile of Fig 2b
  over-predicts the suppression of Fig 4b and Table 3 by 8--10 percentage points
  at the nadir for the four suppressed isoenzymes (and over-predicts the CYP1A2
  induction by 5 points). The discrepancy is not an artefact of the disposition
  reconstruction: driving the enzyme layer with the digitized Fig 2b curve
  itself, with no disposition model at all, reproduces it. Three candidate
  explanations were tested and rejected: truncating IL-6 at 72 h (which the
  paper says is where the retained data end for that patient, because tocilizumab
  was given on day 4) makes the fit *worse*; population variability on IL-6
  disposition moves the nadirs in the right direction but closes only about a
  fifth of the gap; and a uniform rescaling of the driver that does fit the
  recovery phase then misses the descending limb. An effective halving of the
  IL-6 exposure for that scenario would reconcile the two figures, but the paper
  gives no basis for it, so no such correction was applied and the scenario is
  reported as it comes out. Users should treat the Q2W highest-$C_{max}$ output
  of this model as a conservative (deeper-suppression) bound.
* **Victim-drug exposure ratios (Tables 1 and 2) are out of scope.** The
  caffeine, S-warfarin, omeprazole, midazolam, cyclosporine and simvastatin
  predictions used unmodified proprietary Simcyp V21 compound files. Their in
  vivo dispositions are not published in this paper and cannot be reconstructed
  from it, so no attempt is made to reproduce those ratios, nor the Table 2
  ten-fold-lower-$IndC_{50}$ sensitivity analysis that depends on them. This
  model supplies the perpetrator side of the interaction only: the IL-6 time
  course and the resulting CYP activity time course.
* **Gut CYP modulation is not part of this model.** Because IL-6 was
  administered intravenously, Simcyp propagated the modulation to hepatic
  enzymes only; the paper handled the gut by editing intestinal and colonic
  enzyme abundances offline between two runs. That is a static population edit
  rather than a differential equation, so it is not representable as part of the
  ODE system. Table 3 and Figs 3--4, which this vignette gates against, are
  hepatic.
* **Interaction-parameter variability is not carried as etas.** The canonical
  amplitude parameter `emax` is negative for the four suppressed isoenzymes, so
  a log-normal eta is undefined on it, and neither this paper nor the model it
  cites reports how Simcyp correlates $Ind_{max}$ with $IndC_{50}$. No variance
  was invented in their place. Every published quantity reproduced above is a
  typical-value or mean quantity.
* **No residual error is reported.** The source is a simulation-only analysis
  with no fitted residual error model, so `propSd` is fixed at 0 rather than
  invented.
* **Body weight.** The IL-6 volumes are per kilogram with no reference weight
  stated; 70 kg is used, which is the weight at which the reconstruction
  reproduces the published peaks.
* **CYP3A5 duplicates CYP3A4** by the source model's explicit assumption, and
  Table 3 of this paper reports extrema and timings for the two that differ by at
  most one percentage point in one of four scenarios. They are kept as separate
  parameters and separate compartments so that a user can break the assumption.
* **The endogenous IL-6 baseline differs between scenarios.** The recovered
  baseline is ~1.9 pg/mL for three scenarios but ~7.9 pg/mL for the Q2W
  highest-$C_{max}$ scenario. That is what Fig 2b shows, and it is corroborated
  by Fig 4b, whose terminal CYP2C19 activity (~94%) matches the steady state
  implied by 7.9 pg/mL rather than by 1.9 pg/mL.
