---
title: "Posaconazole (Winchell 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posaconazole (Winchell 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- rxode2::rxode2(nlmixr2lib::readModelDb("Winchell_2024_posaconazole"))
```

* Citation: `r nlmixr2lib::readModelDb("Winchell_2024_posaconazole")()$reference`
* Article: <https://doi.org/10.1128/aac.01197-23>
* Supplement (Tables S1-S3, Fig. S1-S4): <https://doi.org/10.1128/aac.01197-23>
  (file `aac.01197-23-s0001.pdf`; also retrievable from the EuropePMC
  `supplementaryFiles` endpoint for PMC10994819)

Winchell 2024 is the pediatric companion to the adult posaconazole tablet
population PK analysis of the same group, which is also packaged here as
[`vanIersel_2018_posaconazole`](vanIersel_2018_posaconazole.html) and is cited as
reference 18 of the present paper ("A one-compartment model with first-order
absorption was fit to the PK data set ... based on the previous analyses of
posaconazole PK in adults").

# Population

The analysis data set comprised 1,236 plasma posaconazole observations from 114
immunocompromised pediatric participants aged 2 to 17 years with documented or
anticipated neutropenia, enrolled in the phase 1b open-label sequential
dose-escalation study P097 (MK-5592-097, ClinicalTrials.gov NCT02452034).
Baseline characteristics are Winchell 2024 Table 1: median age 8 years (range
2-17), median body weight 28.6 kg (range 10.2-102), median eGFR 146
mL/min/1.73 m^2 (range 12.2-314). About 59% were male, 83% White and 87% not
Hispanic. Participants received 3.5 (n = 35), 4.5 (n = 31), or 6 mg/kg (n = 48)
posaconazole -- IV BID on day 1, IV QD on days 2-10, then either oral PFS
(n = 60) or IV (n = 54) QD on days 10-28 -- with the absolute dose capped at
300 mg. Eighty samples were excluded (33 from two participants with incomplete
oral PFS dose intakes, 31 with no sampling time, 13 with a duplicate sampling
time), and concentrations below the 5.00 ng/mL LLOQ were excluded rather than
modelled.

The two age strata that the paper simulates separately are 2 to <7 years
(n = 48, median weight 16 kg, range 10.2-41.7) and 7 to 17 years (n = 66,
median weight 45.4 kg, range 18.2-102).

The same information is available programmatically:

```{r population}
str(nlmixr2lib::readModelDb("Winchell_2024_posaconazole")()$population)
```

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One compartment, first-order absorption, IV into `central` | structure | Results "Final model"; Methods "Population PK model development" ("with IV administration being assumed to be directed into the central compartment") |
| `lcl` (CL) | 4.71 L/h (RSE 3.86%) | Table 2, "CL (L/h)" |
| `lvc` (Vc) | 112 L (RSE 5.18%) | Table 2, "Vc (L)" |
| `lka` (ka) | 0.212 1/h (RSE 17.9%) | Table 2, "KA (h-1)" |
| `logitfdepot` (F1) | 0.826 (RSE 5.58%), on the logit scale | Table 2, "F1"; Results "Final model" ("estimated bioavailability using a logit function (in order to constrain its value between 0 and 1)") |
| `e_wt_cl` | 0.624 (RSE 9.86%), estimated | Table 2, "alpha for CL"; Results "Final model" |
| `e_wt_vc` | 0.971 (RSE 7.86%), estimated | Table 2, "alpha for Vc"; Results "Final model" |
| `etalcl` variance | 0.371^2 = 0.137641 | Table 2, "IIV (CL)" = 37.1 with footnote b ("a CV%, calculated by the square root of Omega, multiplied by 100") |
| `etalvc` variance | 0.277^2 = 0.076729 | Table 2, "IIV (Vc)" = 27.7 with footnote b |
| `etalcl`-`etalvc` covariance | not reported; set to 0 | Results "Final model" states a correlation between CL and Vc exists but no magnitude is given anywhere in the paper or supplement |
| `etalogitfdepot` variance | 2.02^2 = 4.0804 | Table 2, "IIV (F1)" = 2.02 with footnote c ("shown as a standard deviation in the logit domain ... 95% of patients having F1 comprising between 8% and 99%") |
| `expSd` | 0.331 (RSE 4.71%) | Table 2, "Residual error, SD"; Results "Final model" ("an additive error model in the logarithmic scale") |
| Allometric reference weight | 28.6 kg | Table 1, "All (N = 114)" median weight. **Not stated by the paper as the covariate-model reference** -- see Assumptions and deviations |
| No covariate retained (age, weight, eGFR, sex, ethnicity) | -- | Results "Final model" ("therefore, the base structural model was considered the final model") |
| No food effect on bioavailability | -- | Results "Effect of food"; Fig. 1 |

# Exact structural checks

These four checks are independent of any assumption about the simulated cohort,
so they test the packaged model itself rather than the virtual population. Each
is a hard `stopifnot()`.

## The logit-domain bioavailability SD reproduces the paper's own F1 interval

Table 2 footnote c states that the F1 variability of 2.02 "corresponds to 95% of
patients having F1 comprising between 8% and 99%". That sentence is what fixes
2.02 as a standard deviation (not a variance) in the logit domain: reading it as
a variance would give a 95% interval of 23% to 99%.

```{r f1-interval}
ini_tbl <- mod$iniDf
# Fixed effects carry ntheta; random-effect variances carry neta1 == neta2.
logitf1 <- ini_tbl$est[!is.na(ini_tbl$ntheta) & ini_tbl$name == "logitfdepot"]
om_f1 <- sqrt(ini_tbl$est[!is.na(ini_tbl$neta1) &
                            ini_tbl$neta1 == ini_tbl$neta2 &
                            ini_tbl$name == "etalogitfdepot"])
stopifnot(length(logitf1) == 1, length(om_f1) == 1)
c(logitfdepot = logitf1, sd_logit_F1 = om_f1)

f1_ci <- plogis(logitf1 + c(-1.96, 1.96) * om_f1)
# The alternative reading, in which the tabulated 2.02 were a variance rather
# than an SD, so the logit-domain SD would be sqrt(2.02) = 1.42.
f1_ci_var <- plogis(logitf1 + c(-1.96, 1.96) * sqrt(om_f1))

rbind(`2.02 as logit-domain SD` = 100 * f1_ci,
      `2.02 as logit-domain variance` = 100 * f1_ci_var) |>
  round(1)

# Paper footnote c states "between 8% and 99%". The lower bound is what
# discriminates the two readings: the SD reading gives 8%, the variance
# reading gives 23%. The upper bound is 99.6%, which the paper reports
# truncated to 99%.
stopifnot(
  abs(100 * f1_ci[1] - 8) < 0.5,
  100 * f1_ci[2] >= 99, 100 * f1_ci[2] < 100,
  100 * f1_ci_var[1] > 20            # the variance reading is refuted
)
```

## Allometric scaling, dose proportionality, and the Cavg identity

At steady state a one-compartment linear model must satisfy
`Cavg = Dose * F / (CL * tau)` exactly, and clearance must scale as
`(WT / 28.6)^0.624`. Both are checked against a typical-value (no-IIV)
simulation of the final QD dosing interval. `omega = NA` suppresses the random
effects without mutating the shared model object.

```{r structural-checks}
tau <- 24

# Average concentration over a dosing interval by the trapezoidal rule. A plain
# mean() of grid points is biased for the IV arms, where the bolus makes C(0+)
# the peak and C(tau-) the trough.
trap_mean <- function(time, conc) {
  sum(diff(time) * (head(conc, -1) + tail(conc, -1)) / 2) /
    (max(time) - min(time))
}

typ_events <- bind_rows(
  # IV dose straight into central; PFS dose into depot.
  data.frame(id = 1:4, time = 0, amt = c(100, 100, 300, 300),
             cmt = rep(c("central", "depot"), each = 2),
             evid = 1L, ii = tau, ss = 1L, WT = c(20, 60, 20, 60)),
  expand.grid(id = 1:4, time = seq(0, tau, by = 0.1)) |>
    mutate(amt = 0, cmt = "central", evid = 0L, ii = 0, ss = 0L,
           WT = c(20, 60, 20, 60)[id])
) |>
  arrange(id, time, -evid)

typ <- rxode2::rxSolve(mod, typ_events, omega = NA, returnType = "data.frame",
                       keep = "WT") |>
  filter(!is.na(Cc))

# Allometric identity on CL and Vc.
cl_wt <- typ |> group_by(WT) |> summarise(cl = first(cl), vc = first(vc), .groups = "drop")
stopifnot(
  isTRUE(all.equal(cl_wt$cl[cl_wt$WT == 60] / cl_wt$cl[cl_wt$WT == 20],
                   (60 / 20)^0.624, tolerance = 1e-8)),
  isTRUE(all.equal(cl_wt$vc[cl_wt$WT == 60] / cl_wt$vc[cl_wt$WT == 20],
                   (60 / 20)^0.971, tolerance = 1e-8))
)

# Cavg = Dose * F / (CL * tau).  F = 1 for the IV rows, F1 for the PFS rows.
f1_typ <- plogis(logitf1)
cavg_check <- typ |>
  group_by(id, WT) |>
  summarise(cavg_sim = trap_mean(time, Cc), cl = first(cl),
            fdepot = first(fdepot), .groups = "drop") |>
  mutate(route = c("IV", "IV", "PFS", "PFS"),
         amt = c(100, 100, 300, 300),
         f = ifelse(route == "IV", 1, fdepot),
         cavg_closed = amt * f / (cl * tau) * 1000,
         pct = 100 * (cavg_sim / cavg_closed - 1))
cavg_check |> select(route, WT, amt, cavg_sim, cavg_closed, pct)

stopifnot(all(abs(cavg_check$pct) < 0.01))
stopifnot(isTRUE(all.equal(unname(f1_typ), 0.826, tolerance = 1e-3)))
```

Dose proportionality of the IV arm (uncapped) is then exact by construction:

```{r dose-proportionality}
prop_events <- bind_rows(
  data.frame(id = 1:2, time = 0, amt = c(4.5, 6) * 25, cmt = "central",
             evid = 1L, ii = tau, ss = 1L, WT = 25),
  expand.grid(id = 1:2, time = seq(0, tau, by = 0.1)) |>
    mutate(amt = 0, cmt = "central", evid = 0L, ii = 0, ss = 0L, WT = 25)
) |>
  arrange(id, time, -evid)

prop <- rxode2::rxSolve(mod, prop_events, omega = NA, returnType = "data.frame") |>
  filter(!is.na(Cc)) |>
  group_by(id) |>
  summarise(cavg = trap_mean(time, Cc), .groups = "drop")

ratio <- prop$cavg[2] / prop$cavg[1]
c(observed = ratio, expected = 6 / 4.5)
stopifnot(isTRUE(all.equal(ratio, 6 / 4.5, tolerance = 1e-6)))
```

# Virtual cohort

The paper's simulations drew 1,000 subjects per age stratum by jointly sampling
age and weight from the pooled P097 and P032 pediatric populations. That pooled
weight distribution is not reported, so the cohort here is reconstructed from the
only weight information the paper publishes -- the per-stratum median and range
of Winchell 2024 Table 1 -- as a log-normal whose median matches the reported
median and whose 0.5th/99.5th percentiles span the reported range. Weights are
taken at **deterministic quantiles** rather than random draws so the comparison
below is reproducible and free of sampling artifacts.

The random effects are likewise assigned at deterministic quantiles, as a Latin
hypercube: each eta's marginal distribution is reproduced exactly (up to the
quantile grid), and the same subject-level draws are reused across every dose and
formulation arm. This is the common-random-numbers device -- it makes the
between-arm contrasts exact rather than Monte Carlo estimates, so the PFS/IV
comparison below isolates bioavailability instead of mixing in sampling noise.

```{r cohort}
n_per_group <- 200

make_weights <- function(median_wt, lo, hi, n) {
  # log-normal matched to the reported median and range (range ~ +/- 2.576 SD)
  sdlog <- (log(hi) - log(lo)) / (2 * 2.576)
  q <- qlnorm(seq(0.5, n - 0.5) / n, meanlog = log(median_wt), sdlog = sdlog)
  pmin(pmax(q, lo), hi)
}

# Latin-hypercube eta draws: exact marginals, fixed permutation per eta.
lhs_normal <- function(n, sd, seed) {
  set.seed(seed)
  sd * qnorm((sample.int(n) - 0.5) / n)
}

om <- c(etalcl = sqrt(0.137641), etalvc = sqrt(0.076729),
        etalogitfdepot = 2.02)

make_subjects <- function(agegrp, median_wt, lo, hi, seed0) {
  data.frame(
    agegrp = agegrp,
    WT = make_weights(median_wt, lo, hi, n_per_group),
    etalcl = lhs_normal(n_per_group, om[["etalcl"]], seed0 + 1L),
    etalvc = lhs_normal(n_per_group, om[["etalvc"]], seed0 + 2L),
    etalogitfdepot = lhs_normal(n_per_group, om[["etalogitfdepot"]], seed0 + 3L)
  )
}

cohort <- bind_rows(
  make_subjects("2 to <7 years", 16, 10.2, 41.7, 100L),
  make_subjects("7 to 17 years", 45.4, 18.2, 102, 200L)
)

cohort |>
  group_by(agegrp) |>
  summarise(n = n(), median_wt = median(WT), min_wt = min(WT), max_wt = max(WT),
            geomean_wt = exp(mean(log(WT))), .groups = "drop")
```

The cohort's realised relative bioavailability distribution matches the analytic
logit-normal, including the geometric mean that drives the PFS/IV contrast:

```{r cohort-f1}
f1_cohort <- plogis(logitf1 + cohort$etalogitfdepot)
gm_f1_cohort <- exp(mean(log(f1_cohort)))
c(cohort_GM_F1 = gm_f1_cohort,
  cohort_p2.5 = unname(quantile(f1_cohort, 0.025)),
  cohort_p97.5 = unname(quantile(f1_cohort, 0.975)))
```

# Simulation of the final steady-state dosing interval

Winchell 2024 computed Cavg and Cmin "after the last dose of IV and PFS
administration" on day 28 of QD dosing. Eighteen days of QD dosing is far beyond
steady state for a drug with a `r round(log(2) / (4.71 / 112), 1)` h typical
half-life, so the final interval is simulated directly with `ss = 1` and
`ii = 24`, which is exact rather than approximate. IV arms dose into `central`;
PFS arms dose into `depot`, where the logit-scale relative bioavailability
applies. The 300 mg absolute cap is imposed as in the paper.

```{r simulate}
arms <- expand.grid(dose_mgkg = c(4.5, 6, 7.5), form = c("IV", "PFS"),
                    agegrp = unique(cohort$agegrp),
                    stringsAsFactors = FALSE)

obs_grid <- seq(0, tau, by = 0.5)

eta_cols <- c("etalcl", "etalvc", "etalogitfdepot")

events <- lapply(seq_len(nrow(arms)), function(k) {
  a <- arms[k, ]
  # Same subjects (weights AND etas) in every arm -- common random numbers.
  s <- cohort[cohort$agegrp == a$agegrp, ]
  # Distinct id blocks per arm: rxode2 keys on `id` alone, so reusing ids
  # across arms would collapse them into single multi-regimen subjects.
  ids <- (k - 1L) * n_per_group + seq_len(nrow(s))
  amt <- pmin(a$dose_mgkg * s$WT, 300)
  rows <- rep(seq_len(nrow(s)), each = length(obs_grid))
  bind_rows(
    cbind(data.frame(id = ids, time = 0, amt = amt,
                     cmt = if (a$form == "IV") "central" else "depot",
                     evid = 1L, ii = tau, ss = 1L, WT = s$WT),
          s[, eta_cols]),
    cbind(data.frame(id = ids[rows],
                     time = rep(obs_grid, times = nrow(s)),
                     amt = 0, cmt = "central", evid = 0L, ii = 0, ss = 0L,
                     WT = s$WT[rows]),
          s[rows, eta_cols])
  ) |>
    mutate(agegrp = a$agegrp, form = a$form, dose_mgkg = a$dose_mgkg)
}) |>
  bind_rows() |>
  arrange(id, time, -evid)

# omega = NA: the etas are supplied per subject from the Latin hypercube above
# rather than resampled internally, so the whole vignette is deterministic.
sim <- rxode2::rxSolve(mod, events, omega = NA, returnType = "data.frame",
                       keep = c("WT", "agegrp", "form", "dose_mgkg")) |>
  filter(!is.na(Cc)) |>
  mutate(dose_label = paste0(dose_mgkg, " mg/kg"))

nrow(sim)
```

`Cc` is the individual predicted (IPRED) concentration. Exposure metrics are
taken on that scale because the paper's Cavg and Cmin are model-derived exposure
summaries, not residual-error-contaminated observations.

# PKNCA validation

```{r pknca}
conc_obj <- PKNCA::PKNCAconc(
  data = sim,
  formula = Cc ~ time | agegrp + dose_label + form + id,
  concu = "ng/mL", timeu = "h"
)

dose_df <- events |>
  filter(evid == 1L) |>
  mutate(dose_label = paste0(dose_mgkg, " mg/kg")) |>
  select(id, time, amt, agegrp, dose_label, form)

dose_obj <- PKNCA::PKNCAdose(
  data = dose_df,
  formula = amt ~ time | agegrp + dose_label + form + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = 0, end = tau,
  cav = TRUE, cmin = TRUE, cmax = TRUE, tmax = TRUE, auclast = TRUE
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_tbl <- as.data.frame(nca$result) |>
  # PKNCA emits dependency/interval rows; keep only the requested interval.
  filter(start == 0, end == tau, !is.na(PPORRES))

nca_tbl |> count(PPTESTCD)
```

Per-subject Cavg and Cmin are summarised as geometric means to match the
supplement, which reports "Predicted geometric mean Cavg and Cmin".

```{r summarise-nca}
sim_summary <- nca_tbl |>
  filter(PPTESTCD %in% c("cav", "cmin")) |>
  group_by(agegrp, dose_label, form, PPTESTCD) |>
  summarise(PPORRES = exp(mean(log(PPORRES))),
            geo_cv = 100 * sqrt(exp(var(log(PPORRES))) - 1), .groups = "drop")

sim_summary |>
  select(-geo_cv) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  arrange(agegrp, form, dose_label) |>
  rename("Age group" = agegrp, "Dose" = dose_label, "Formulation" = form,
         "Cavg (ng/mL)" = cav, "Cmin (ng/mL)" = cmin) |>
  knitr::kable(digits = 1,
               caption = "Simulated steady-state geometric mean exposure.")
```

# Comparison against the published simulation (Supplementary Table 2)

Winchell 2024 Supplementary Table 2 reports predicted geometric mean Cavg and
Cmin for every age group / dose / formulation combination. All 24 values are
transcribed below.

```{r reference-s2}
reference_s2 <- tribble(
  ~agegrp,          ~dose_label,  ~form,  ~cav,    ~cmin,
  "2 to <7 years",  "4.5 mg/kg",  "IV",   1044.94,  515.44,
  "2 to <7 years",  "4.5 mg/kg",  "PFS",   796.61,  504.96,
  "2 to <7 years",  "6 mg/kg",    "IV",   1364.60,  656.24,
  "2 to <7 years",  "6 mg/kg",    "PFS",  1049.68,  654.95,
  "2 to <7 years",  "7.5 mg/kg",  "IV",   1713.79,  830.09,
  "2 to <7 years",  "7.5 mg/kg",  "PFS",  1320.77,  827.78,
  "7 to 17 years",  "4.5 mg/kg",  "IV",   1361.00,  795.48,
  "7 to 17 years",  "4.5 mg/kg",  "PFS",  1041.55,  734.80,
  "7 to 17 years",  "6 mg/kg",    "IV",   1748.46, 1035.21,
  "7 to 17 years",  "6 mg/kg",    "PFS",  1331.39,  947.19,
  "7 to 17 years",  "7.5 mg/kg",  "IV",   2001.55, 1210.21,
  "7 to 17 years",  "7.5 mg/kg",  "PFS",  1546.59, 1114.93
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_summary |> select(agegrp, dose_label, form, PPTESTCD, PPORRES),
  reference = reference_s2,
  by = c("agegrp", "dose_label", "form"),
  params = c("cav", "cmin"),
  units = c(cav = "ng/mL", cmin = "ng/mL")
)
cmp |>
  rename("Age group" = agegrp, "Dose" = dose_label, "Formulation" = form) |>
  knitr::kable(digits = 1,
               caption = "Simulated vs. Winchell 2024 Supplementary Table 2.")
attr(cmp, "footnote")
```

The IV arms -- which are independent of the bioavailability parameter and of its
random effect -- reproduce the published geometric means to within
`r sprintf("%.0f%%", max(abs(as.numeric(gsub("[^0-9.-]", "", cmp[["% diff"]][cmp$form == "IV"])))))`,
with the 7 to 17 year stratum agreeing most closely. Two systematic offsets are
present and are explained (not tuned away) below.

## Why the 2 to <7 year stratum sits low: the paper's cohort is not its Table 1 cohort

Because `Cavg = Dose / (CL * tau)` for the IV arms and `CL` scales as
`WT^0.624`, the ratio of the two strata's geometric mean Cavg values depends only
on the ratio of their geometric mean weights -- not on the allometric reference
weight, nor on any other parameter. Inverting the published Supplementary Table 2
IV values therefore recovers the weight ratio of the *paper's* simulated cohort:

```{r cohort-diagnostic}
iv_ref <- reference_s2 |> filter(form == "IV") |> select(agegrp, dose_label, cav)
ratio_tbl <- iv_ref |>
  pivot_wider(names_from = agegrp, values_from = cav) |>
  mutate(cavg_ratio = `7 to 17 years` / `2 to <7 years`,
         implied_wt_ratio = cavg_ratio^(1 / (1 - 0.624)))
ratio_tbl |>
  rename("Dose" = dose_label, "Cavg ratio (older/younger)" = cavg_ratio,
         "Implied geometric mean weight ratio" = implied_wt_ratio) |>
  knitr::kable(digits = 3)

table1_median_ratio <- 45.4 / 16
table1_median_ratio
```

The published Cavg values imply a between-stratum weight ratio of roughly 1.5 to
2.0, whereas Winchell 2024 Table 1 medians give `r round(table1_median_ratio, 2)`.
(The 7.5 mg/kg row is additionally compressed by the 300 mg cap.) The pooled
P097 + P032 cohort actually simulated is therefore lighter in the older stratum
and/or heavier in the younger stratum than the P097-only Table 1 figures used to
build the cohort above. This is a property of the unreported virtual population,
not of the packaged model -- consistent with the exact structural checks passing
and with the older stratum, whose reconstructed weights are closest to the
implied ones, agreeing best.

## Why the PFS arms sit ~20% low: shrinkage on the bioavailability random effect

For the PFS arms, `Cavg = Dose * F1 / (CL * tau)`, so the PFS/IV ratio of
geometric means is exactly the population geometric mean of `F1`. With the
reported logit-domain SD of 2.02, that geometric mean is well below the typical
value of 0.826, because the logit-normal distribution has a long left tail:

Because the same subjects (and the same etas) appear in the IV and PFS arms, the
simulated PFS/IV ratio of geometric mean Cavg must equal the cohort geometric
mean of F1 *exactly*. That is asserted rather than merely observed:

```{r f1-shrinkage}
pfs_iv_sim <- sim_summary |>
  filter(PPTESTCD == "cav") |>
  select(agegrp, dose_label, form, PPORRES) |>
  pivot_wider(names_from = form, values_from = PPORRES) |>
  mutate(ratio = PFS / IV)

pfs_iv_paper <- reference_s2 |>
  select(agegrp, dose_label, form, cav) |>
  pivot_wider(names_from = form, values_from = cav) |>
  mutate(ratio = PFS / IV)

c(cohort_GM_F1 = gm_f1_cohort,
  simulated_PFS_IV_ratio = mean(pfs_iv_sim$ratio),
  paper_PFS_IV_ratio = mean(pfs_iv_paper$ratio),
  typical_F1 = unname(f1_typ))

# The uncapped arms must reproduce GM(F1) to numerical precision.
uncapped <- pfs_iv_sim |> filter(dose_label != "7.5 mg/kg" | agegrp == "2 to <7 years")
stopifnot(all(abs(uncapped$ratio / gm_f1_cohort - 1) < 0.01))
```

The paper's own simulation behaves as if the logit-domain SD were about 1.0
rather than 2.02. Table 2 reports 42% shrinkage on this random effect, and
resampling shrunken post-hoc etas (SD ~ 2.02 x 0.58 ~ 1.2) would reproduce
almost exactly the ratio the supplement shows. The packaged model encodes the
**reported** parameter (2.02, which footnote c's 8%-99% interval confirms
exactly), not the smaller value implied by the supplement's simulation output, so
the PFS arms here are expected to sit systematically below Supplementary Table 2
by roughly the ratio
`r sprintf("%.0f%%", 100 * (gm_f1_cohort / mean(pfs_iv_paper$ratio) - 1))`.
The IV arms are unaffected and are the appropriate validation target for the
disposition parameters.

# Comparison against Supplementary Table 3 (known weight bands)

Supplementary Table 3 tabulates the percentage of virtual pediatric patients
whose Cavg falls below 500, within 500 to 2,500, and at or above 2,500 ng/mL for
PFS 6 mg/kg QD, stratified by **weight band**. Because the bands are explicit,
this comparison does not depend on the unreported cohort weight distribution.

```{r supp-table3}
bands <- tribble(
  ~band,        ~lo,  ~hi,
  "30-40 kg",    30,   40,
  "40-50 kg",    40,   50,
  "50-70 kg",    50,   70,
  "70-90 kg",    70,   90,
  "90-110 kg",   90,  110
)

band_etas <- cohort[cohort$agegrp == "7 to 17 years", eta_cols]

band_events <- lapply(seq_len(nrow(bands)), function(k) {
  b <- bands[k, ]
  wt <- seq(b$lo, b$hi, length.out = n_per_group)
  ids <- 100000L + (k - 1L) * n_per_group + seq_along(wt)
  rows <- rep(seq_along(wt), each = length(obs_grid))
  bind_rows(
    cbind(data.frame(id = ids, time = 0, amt = pmin(6 * wt, 300), cmt = "depot",
                     evid = 1L, ii = tau, ss = 1L, WT = wt),
          band_etas),
    cbind(data.frame(id = ids[rows],
                     time = rep(obs_grid, times = length(wt)),
                     amt = 0, cmt = "central", evid = 0L, ii = 0, ss = 0L,
                     WT = wt[rows]),
          band_etas[rows, ])
  ) |>
    mutate(band = b$band)
}) |>
  bind_rows() |>
  arrange(id, time, -evid)

band_sim <- rxode2::rxSolve(mod, band_events, omega = NA, returnType = "data.frame",
                            keep = c("WT", "band")) |>
  filter(!is.na(Cc))

band_cat <- band_sim |>
  group_by(band, id) |>
  summarise(cavg = trap_mean(time, Cc), .groups = "drop") |>
  group_by(band) |>
  summarise(`<500` = 100 * mean(cavg < 500),
            `500 to <2500` = 100 * mean(cavg >= 500 & cavg < 2500),
            `>=2500` = 100 * mean(cavg >= 2500), .groups = "drop")

paper_s3 <- tribble(
  ~band,        ~`<500`, ~`500 to <2500`, ~`>=2500`,
  "30-40 kg",     0.30,   89.00,  10.70,
  "40-50 kg",     0.00,   87.80,  12.20,
  "50-70 kg",     0.10,   88.10,  11.80,
  "70-90 kg",     0.50,   95.80,   3.70,
  "90-110 kg",    1.40,   96.60,   2.00
)

bind_rows(
  band_cat |> mutate(Source = "Simulated"),
  paper_s3 |> mutate(Source = "Winchell 2024 Table S3")
) |>
  relocate(Source) |>
  arrange(band, Source) |>
  rename("Weight band" = band) |>
  knitr::kable(digits = 1,
               caption = "Percentage of virtual patients by Cavg category, PFS 6 mg/kg QD.")
```

The qualitative pattern the paper draws from this table is reproduced: essentially
nobody falls below 500 ng/mL in any band, and the fraction at or above
2,500 ng/mL declines with increasing weight once the 300 mg cap binds at 50 kg.
The simulated upper tail is smaller than the published one, in the same direction
and for the same reason as the Supplementary Table 2 PFS offset above.

# Replication of Figure 3

Figure 3 of Winchell 2024 shows the distribution of simulated Cavg per age group
and formulation at 4.5, 6, and 7.5 mg/kg, with reference lines at 500 and
2,500 ng/mL and the geometric mean marked.

```{r fig3, fig.height = 5.5}
cavg_subject <- nca_tbl |>
  filter(PPTESTCD == "cav") |>
  left_join(distinct(sim, id, agegrp, form, dose_label), by = c("id", "agegrp", "form", "dose_label"))

gm_lines <- cavg_subject |>
  group_by(agegrp, form, dose_label) |>
  summarise(gm = exp(mean(log(PPORRES))), .groups = "drop")

ggplot(cavg_subject, aes(x = PPORRES, fill = form)) +
  geom_density(alpha = 0.45, colour = NA) +
  geom_vline(data = gm_lines, aes(xintercept = gm, colour = form),
             linetype = "dashed", show.legend = FALSE) +
  geom_vline(xintercept = c(500, 2500), linewidth = 0.4) +
  facet_grid(dose_label ~ agegrp) +
  scale_x_log10() +
  labs(x = "Steady-state Cavg (ng/mL)", y = "Density", fill = "Formulation",
       title = "Replicates Figure 3 of Winchell 2024",
       subtitle = "Solid lines: 500 and 2,500 ng/mL thresholds; dashed lines: geometric means") +
  theme_bw()
```

```{r fig3-target}
cavg_subject |>
  group_by(agegrp, dose_label, form) |>
  summarise(`% >=500 ng/mL` = 100 * mean(PPORRES >= 500),
            `% 500 to <2500` = 100 * mean(PPORRES >= 500 & PPORRES < 2500),
            .groups = "drop") |>
  rename("Age group" = agegrp, "Dose" = dose_label, "Formulation" = form) |>
  knitr::kable(digits = 1,
               caption = "Attainment of the Winchell 2024 PK target (Cavg >= 500 ng/mL).")
```

The paper's headline conclusion -- that the 6 mg/kg per day regimen puts more
than 90% of pediatric patients in both age strata above the 500 ng/mL Cavg
target, by either route -- is reproduced.

# Concentration-time profile over the IV to PFS switch

The clinical regimen starts IV and optionally switches to oral PFS between days
10 and 18. A typical-value profile over the switch illustrates the structural
model; it is not a published figure.

```{r switch-profile}
switch_wt <- 28.6
switch_amt <- min(6 * switch_wt, 300)
switch_events <- bind_rows(
  # IV BID on day 1, IV QD days 2-10, PFS QD days 10-14.
  data.frame(time = c(0, 12), amt = switch_amt, cmt = "central", evid = 1L),
  data.frame(time = 24 * (1:9), amt = switch_amt, cmt = "central", evid = 1L),
  data.frame(time = 24 * (10:14), amt = switch_amt, cmt = "depot", evid = 1L),
  data.frame(time = seq(0, 24 * 15, by = 0.25), amt = 0, cmt = "central", evid = 0L)
) |>
  mutate(id = 1L, WT = switch_wt, ii = 0, ss = 0L) |>
  arrange(time, -evid)

switch_sim <- rxode2::rxSolve(mod, switch_events, omega = NA,
                              returnType = "data.frame") |>
  filter(!is.na(Cc))

ggplot(switch_sim, aes(time / 24, Cc)) +
  geom_line() +
  geom_vline(xintercept = 10, linetype = "dashed") +
  geom_hline(yintercept = 500, colour = "grey40") +
  labs(x = "Time (days)", y = "Posaconazole concentration (ng/mL)",
       title = "Typical-value profile, 6 mg/kg QD in a 28.6 kg patient",
       subtitle = "Dashed line: switch from IV solution to oral PFS on day 10") +
  theme_bw()
```

# Assumptions and deviations

* **Allometric reference weight (28.6 kg) is not stated by the paper.** Winchell
  2024 reports estimated allometric exponents on CL and Vc but never prints the
  body-weight normalisation constant used in the NONMEM covariate model, and
  neither does the supplement. 28.6 kg -- the overall median weight of the
  analysis population (Table 1, "All (N = 114)") -- was adopted. It is supported
  three ways: (i) back-solving the Supplementary Table 2 IV Cavg geometric means
  brackets the reference at roughly 29-35 kg and the weight-band fit to
  Supplementary Table 3 gives ~35 kg; (ii) the conventional 70 kg is refuted
  decisively, overpredicting the Supplementary Table 2 IV Cavg values by 54-78%;
  (iii) the same authors' adult posaconazole tablet model, which this analysis
  was based on, normalises body weight to its own simulation-cohort median
  (`vanIersel_2018_posaconazole`). A value in the low 30s would fit the
  supplement's tables marginally better, but is printed nowhere in the paper;
  choosing it would be fitting the model to a validation target rather than
  transcribing the source, so the published median was used instead. Users who
  need the best agreement with Supplementary Table 3 specifically can rescale
  `lcl` and `lvc` accordingly.
* **The CL-Vc random-effect correlation is reported to exist but not
  quantified.** Results "Final model" lists "correlation between CL and Vc" among
  the final model's features, but no correlation coefficient or covariance
  appears in Table 2, in the text, or in the supplement. The `ini()` block keeps
  the two-eta block structure with the off-diagonal set to 0 rather than
  inventing a covariance. Simulated marginal distributions of CL and Vc are
  unaffected; only their joint dependence is.
* **The bioavailability random effect is encoded at its reported magnitude
  (logit-domain SD 2.02), which is larger than the paper's own simulations
  behave as if they used.** Footnote c's 8%-99% interval confirms 2.02 as the
  reported SD exactly, so that is what the model file carries. The consequence
  is a systematic ~20% underprediction of the Supplementary Table 2 and Table 3
  PFS results, quantified above and attributable to the 42% shrinkage the paper
  reports on this random effect. This was not tuned away.
* **IIV magnitudes are read as log-scale SDs, per Table 2 footnote b**
  ("a CV%, calculated by the square root of Omega, multiplied by 100"), so
  omega^2 = 0.371^2 and 0.277^2 directly, rather than via the lognormal
  omega^2 = log(1 + CV^2) conversion. Footnote c's treatment of the F1 entry as a
  bare logit-domain SD is consistent with that reading of the column.
* **IV doses are modelled as boluses into the central compartment.** This follows
  Methods ("IV administration being assumed to be directed into the central
  compartment"); the paper does not report an infusion duration in the model, and
  Cavg and Cmin are insensitive to it. Cmax and Tmax after an IV dose would be,
  so no IV Cmax comparison is attempted here.
* **The virtual cohort is reconstructed from Table 1, not from the paper's
  simulation cohort**, which pooled P097 with the unreported P032 weight
  distribution. The consequences are quantified in the cohort-diagnostic section
  above.
* **Concentrations below the 5.00 ng/mL LLOQ were excluded** from the original
  analysis rather than modelled with an M3-type likelihood; the packaged model
  carries no LLOQ handling.
* **Simulated exposures use IPRED (`Cc`), without residual error.** The paper's
  Cavg and Cmin are model-derived exposure metrics, so residual variability is
  not applicable to them.
* **Absolute dose cap.** The 300 mg cap of the source protocol and simulations is
  applied in the event tables, not in the model file, since it is a dosing rule
  rather than a model parameter.
