---
title: "Abrocitinib (Wojciechowski 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Abrocitinib (Wojciechowski 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wojciechowski_2022_abrocitinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wojciechowski_2022_abrocitinib"))$description`
* Article: <https://doi.org/10.1007/s40262-021-01104-z> (open access, Clin Pharmacokinet 2022;61:709-723)
* Supplement: Online Resource 1-4 at the same DOI; Online Resource 2 prints the
  final NONMEM control stream from which every model equation below is taken.

A correction notice (Clin Pharmacokinet, published online 18 February 2022)
changed only the article's licence to open access; no value was revised.

# Population

The model was fitted to 6206 abrocitinib plasma concentrations from 995
participants in 11 Pfizer studies run between May 2013 and July 2019: seven
phase I studies in healthy volunteers (including dedicated food, formulation,
drug-interaction, hepatic-impairment and QT studies), one phase II study in
plaque psoriasis, and one phase IIb and two phase III studies in
moderate-to-severe atopic dermatitis (Online Resource 1). By participant type
there were 165 healthy volunteers, 45 patients with psoriasis, 769 with atopic
dermatitis (90 of them adolescents aged 12-17 years) and 8 each with mild and
moderate hepatic impairment (Child-Pugh A and B). Overall 39.2% were female;
median age was 34 years (range 12-84) and median body weight 76 kg (range
34.4-180); 66.6% were White, 11.0% Black, 17.9% Asian (4.4% Japanese) and
4.0% other race (Table 1). Daily doses ranged from 3 to 800 mg.

The same information is available programmatically:

```{r population}
str(readModelDb("Wojciechowski_2022_abrocitinib")()$population)
```

# Source trace

Every value is in Table 2 of the paper; every equation is in the Online
Resource 2 control stream. The per-parameter origin is also recorded next to
each `ini()` entry in
`inst/modeldb/specificDrugs/Wojciechowski_2022_abrocitinib.R`.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` | log(22) L/h | Table 2 'CL' |
| `lvc` | log(87.8) L | Table 2 'Vc' |
| `lq` | log(1.16) L/h | Table 2 'Q' |
| `lvp` | log(8.25) L | Table 2 'Vp' |
| `lr1` | log(75.3) mg/h | Table 2 'k0' (zero-order rate into central) |
| `lamt_fo` | log(121) mg | Table 2 'AK1' (amount absorbed first-order) |
| `lka` | log(4.01) 1/h | Table 2 'ka' |
| `ltlag` | log(0.183) h | Table 2 'Effect of tablet formulations on ALAG1'; base ALAG1 = 0 (Online Resource 2) |
| `logitfdepot` | logit(0.5977), held constant | Methods 2.4.2; Online Resource 2 `POPF = 0.5977` |
| `e_rif_f`, `e_cypinh_f` | -2.08, 1.31 | Table 2 rifampin / fluconazole-or-fluvoxamine on F |
| `e_ph2b_f`, `e_ph3_f` | -1.02, -0.766 | Table 2 phase IIb / phase III tablet on F |
| `e_multidose_f` | 0.241 | Table 2 'Effect of multiple dosing on F' |
| `e_race_f`, `e_dis_f`, `e_hepimp_f` | 0.815, 0.489, 1.3 | Table 2 Asian/other race, psoriasis and AD, hepatic impairment on F |
| `e_adol_f`, `e_dose800_f`, `e_sexf_f` | -0.589, -0.778, 0.353 | Table 2 adolescent age, 800 mg dose, female sex on F |
| `e_highfat_amt_fo` | -1, held constant | Table 2 'Effect of high-fat meal on Ak1'; Online Resource 3 run 133 |
| `e_susp_amt_fo`, `e_ph2b_amt_fo` | 1.17, -0.68 | Table 2 suspension / phase IIb tablets on Ak1 |
| `e_rif_cl`, `e_fluco_cl`, `e_fluvox_cl` | 0.264, -0.541, -0.234 | Table 2 DDI effects on CL |
| `e_dose_cl` | -0.169 | Table 2 'Effect of effective daily dose on CL' |
| `cl_exp_famp` | -0.186 | Table 2 'Maximum change in CL with respect to time (TAFO)' |
| `lcl_exp_kdes` | log(log(2) / 21.6) 1/h | Table 2 half-life 21.6 h |
| `e_wt_cl`, `e_wt_vc` | 0.453, 0.52 | Table 2 weight on CL and Q / Vc and Vp |
| `etalcl`, `etalvc` | 0.577^2, 0.414^2, cov 0.326 x 0.577 x 0.414 | Table 2 IIV (CV = sqrt(omega^2), footnote a) and correlation |
| `propSd`, `addSd` | 0.437, 0.509 ng/mL | Table 2 RUV PRO / RUV ADD (SDs; EPS variance 1) |
| `e_studymod_propSd`, `e_studyhi_propSd` | 0.495, 1.16 | Table 2 study effects on RUV PRO; studies from Online Resource 2 |
| `logit_fi` / `fi` | sum of logit shifts, `expit()` | Online Resource 2 `FT`, `FI` |
| `ffo = min(1, amt_fo / DOSE)` | n/a | Online Resource 2 `POPFK1` |
| `cl` | weight, DDI, time and effective-dose terms | Online Resource 2 `CL`, `COVTAFOCL`, `COVEFFDOSRCL` |
| `f(depot)`, `f(central)`, `rate(central)` | `fi * ffo`, `fi * (1 - ffo)`, `r1` | Online Resource 2 `F1`, `F2`, `R2` |
| `alag(depot)`, `alag(central)` | tablet lag | Online Resource 2 `ALAG1`, `ALAG2 = ALAG1` |
| `Cc ~ add(addSd) + prop(propSd_study)` | n/a | Online Resource 2 `$ERROR` (`W = SQRT(IPRE^2 * RUVPRO^2 + RUVADD^2)`) |

# Dosing the model

The paper's dataset gives every administration two records with the same
amount: one into the depot (first-order arm) and one into the central
compartment with `RATE = -1`, so that the model supplies the zero-order rate.
The same is required here: a `depot` record with `rate = 0` and a `central`
record with `rate = -1`. The model then routes `fi * ffo` of the dose to the
depot and `fi * (1 - ffo)` to central at 75.3 mg/h, where `ffo` is
`min(1, 121 mg / dose)` for tablets. A 100 mg tablet dose is therefore
absorbed entirely first-order, and the zero-order arm only appears above
121 mg (or always, after a high-fat meal).

Covariates are columns of the event table. `DOSE_ABROCITINIB_MG` must equal
`amt`; `DOSE_ABROCITINIB_MGD` is the regimen's total daily dose; `MULTI_DOSE`
is 0 up to the second dose of a regimen and 1 from then on. Time-dependent
clearance runs on the solver time `t`, so the first dose of a treatment
period must sit at `t = 0`.

```{r helpers}
reference_covariates <- function(...) {
  base <- list(
    WT = 70, SEXF = 0, RACE_ASIAN = 0, RACE_OTHER = 0, ADOLESCENT = 0,
    DIS_PSORIASIS = 0, DIS_ATOPIC_DERMATITIS = 0,
    HEPIMP_MILD = 0, HEPIMP_MOD = 0,
    CONMED_RIFAMPICIN = 0, CONMED_FLUCONAZOLE = 0, CONMED_FLUVOXAMINE = 0,
    FED_HIGHFAT = 0,
    FORM_ABROCITINIB_SUSP = 0, FORM_ABROCITINIB_TAB_PH2B = 0,
    FORM_ABROCITINIB_TAB_PH3 = 1,
    DOSE_ABROCITINIB_MG = 200, DOSE_ABROCITINIB_MGD = 200,
    STUDY_B7451005 = 0, STUDY_B7451006 = 0, STUDY_B7451012 = 0,
    STUDY_B7451043 = 0
  )
  as.data.frame(utils::modifyList(base, list(...)))
}

# One subject's event table: n_dose administrations every `tau` hours, each
# as a depot record and a rate = -1 central record, plus observations on the
# central state. Covariates in `cov` (one row) are copied to every row;
# MULTI_DOSE switches to 1 at the second dose.
regimen_events <- function(id, cov, n_dose, tau = 24, obs_times) {
  dose_times <- (seq_len(n_dose) - 1) * tau
  amt <- cov$DOSE_ABROCITINIB_MG
  doses <- dplyr::bind_rows(
    data.frame(time = dose_times, amt = amt, evid = 1L, cmt = "depot", rate = 0),
    data.frame(time = dose_times, amt = amt, evid = 1L, cmt = "central", rate = -1)
  )
  obs <- data.frame(time = obs_times, amt = 0, evid = 0L, cmt = "central", rate = 0)
  ev <- dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(time, dplyr::desc(evid)) |>
    dplyr::mutate(id = id)
  ev <- cbind(ev, cov[rep(1, nrow(ev)), , drop = FALSE])
  ev$MULTI_DOSE <- as.integer(n_dose > 1 & ev$time >= tau)
  ev
}
```

# Reproducing the published covariate ratios

Figure 3 of the paper gives, for each covariate scenario, the median over
1000 simulated trials of the geometric-mean ratio of steady-state Cmax and
24-h AUC after 200 mg once daily, relative to a healthy, White, adult, male,
70 kg, fasted participant taking the phase III tablet. The Results text adds
three typical-patient comparisons. All are reproduced below with the
typical-value model (random effects set to zero) after 30 days of dosing;
PKNCA computes Cmax and AUC over the last dosing interval.

```{r scenarios}
mod <- readModelDb("Wojciechowski_2022_abrocitinib")
mod_typical <- rxode2::zeroRe(mod)

scenarios <- list(
  "Reference" = reference_covariates(),
  "Female" = reference_covariates(SEXF = 1),
  "Adolescent" = reference_covariates(ADOLESCENT = 1),
  "Asian/other" = reference_covariates(RACE_ASIAN = 1),
  "Rifampin DDI" = reference_covariates(CONMED_RIFAMPICIN = 1),
  "Fluconazole DDI" = reference_covariates(CONMED_FLUCONAZOLE = 1),
  "Fluvoxamine DDI" = reference_covariates(CONMED_FLUVOXAMINE = 1),
  "AD/psoriasis patient" = reference_covariates(DIS_ATOPIC_DERMATITIS = 1),
  "Mild/moderate hepatic impairment" = reference_covariates(HEPIMP_MILD = 1),
  "High body weight (111 kg)" = reference_covariates(WT = 111),
  "Low body weight (52 kg)" = reference_covariates(WT = 52),
  "Phase II tablet 400 mg, fasted" = reference_covariates(
    FORM_ABROCITINIB_TAB_PH3 = 0, DOSE_ABROCITINIB_MG = 400,
    DOSE_ABROCITINIB_MGD = 400
  ),
  "Phase II tablet 400 mg, high-fat meal" = reference_covariates(
    FORM_ABROCITINIB_TAB_PH3 = 0, DOSE_ABROCITINIB_MG = 400,
    DOSE_ABROCITINIB_MGD = 400, FED_HIGHFAT = 1
  ),
  "AD, White, 80 kg" = reference_covariates(WT = 80, DIS_ATOPIC_DERMATITIS = 1),
  "AD, Asian, 66 kg" = reference_covariates(
    WT = 66, DIS_ATOPIC_DERMATITIS = 1, RACE_ASIAN = 1
  ),
  "AD, adolescent, 61 kg" = reference_covariates(
    WT = 61, DIS_ATOPIC_DERMATITIS = 1, ADOLESCENT = 1
  )
)

n_days <- 30
last_day <- seq((n_days - 1) * 24, n_days * 24, by = 0.05)
ev_typ <- dplyr::bind_rows(lapply(seq_along(scenarios), function(i) {
  regimen_events(i, scenarios[[i]], n_dose = n_days, obs_times = last_day) |>
    dplyr::mutate(scenario = names(scenarios)[i])
}))

sim_typ <- rxode2::rxSolve(
  mod_typical, ev_typ, keep = "scenario", returnType = "data.frame"
)

conc_typ <- sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, scenario, time, Cc)
dose_typ <- ev_typ |>
  dplyr::filter(evid == 1, cmt == "depot") |>
  dplyr::select(id, scenario, time, amt)

nca_typ <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_typ, Cc ~ time | scenario + id),
  PKNCA::PKNCAdose(dose_typ, amt ~ time | scenario + id),
  intervals = data.frame(
    start = (n_days - 1) * 24, end = n_days * 24, cmax = TRUE, auclast = TRUE
  )
))

typ <- as.data.frame(nca_typ) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
  dplyr::select(scenario, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
ref <- typ[typ$scenario == "Reference", ]
```

```{r ratio-table}
published <- tibble::tribble(
  ~scenario, ~versus, ~pub_cmax, ~pub_auc, ~source,
  "Female", "Reference", 0.996, 0.999, "Figure 3",
  "Adolescent", "Reference", 0.686, 0.666, "Figure 3",
  "Asian/other", "Reference", 1.430, 1.510, "Figure 3",
  "Rifampin DDI", "Reference", 0.186, 0.128, "Figure 3",
  "Fluconazole DDI", "Reference", 2.010, 3.840, "Figure 3",
  "Fluvoxamine DDI", "Reference", 1.730, 2.320, "Figure 3",
  "AD/psoriasis patient", "Reference", 1.270, 1.310, "Figure 3",
  "Mild/moderate hepatic impairment", "Reference", 1.650, 1.780, "Figure 3 and Discussion",
  "High body weight (111 kg)", "Reference", 0.789, 0.800, "Figure 3",
  "Low body weight (52 kg)", "Reference", 1.170, 1.150, "Figure 3",
  "Phase II tablet 400 mg, high-fat meal", "Phase II tablet 400 mg, fasted", 1.010, 0.991, "Figure 3",
  "AD, Asian, 66 kg", "AD, White, 80 kg", 1.43, 1.48, "Results 3.4",
  "AD, adolescent, 61 kg", "AD, White, 80 kg", 0.86, 0.81, "Discussion (14% and 19% lower)"
)

ratios <- published |>
  dplyr::left_join(typ, by = "scenario") |>
  dplyr::left_join(
    dplyr::rename(typ, versus = scenario, cmax_ref = cmax, auc_ref = auclast),
    by = "versus"
  ) |>
  dplyr::mutate(
    sim_cmax = cmax / cmax_ref,
    sim_auc = auclast / auc_ref,
    diff_cmax_pct = 100 * (sim_cmax / pub_cmax - 1),
    diff_auc_pct = 100 * (sim_auc / pub_auc - 1)
  )

ratios |>
  dplyr::transmute(
    "Scenario" = scenario,
    "Compared with" = versus,
    "Cmax ratio, model" = signif(sim_cmax, 3),
    "Cmax ratio, paper" = pub_cmax,
    "AUC ratio, model" = signif(sim_auc, 3),
    "AUC ratio, paper" = pub_auc,
    "Paper source" = source
  ) |>
  knitr::kable(caption = "Steady-state ratios after 200 mg once daily (400 mg for the food rows): typical-value model vs. the paper.")
```

```{r ratio-gate}
# Deterministic on the model side; the paper's values are medians of 1000
# simulated trials, so a few percent separates the two.
gated <- dplyr::filter(ratios, scenario != "Female")
stopifnot(
  all(abs(gated$diff_auc_pct) < 3),
  all(abs(gated$diff_cmax_pct) < 6)
)
```

Apart from the female row, discussed below, every AUC ratio is reproduced to
within `r round(max(abs(gated$diff_auc_pct)), 1)`% and every Cmax ratio to
within `r round(max(abs(gated$diff_cmax_pct)), 1)`%. That is the agreement
expected between a typical-value calculation and the median of the paper's
simulated trials, which include random effects on CL and Vc.

**Female sex.** Table 2 and the control stream estimate a +0.353 shift on
logit F for females (`IF (SEX.EQ.2) COVSEXF = SEXF2`), which the model
reproduces. At the reference this raises steady-state AUC by about
`r signif(100 * (ratios$sim_auc[ratios$scenario == "Female"] - 1), 2)`%,
yet the Female row of Figure 3 is centred at 1.0 (0.996 for Cmax, 0.999 for
AUC). The paper does not discuss sex in its Results, so the figure cannot be
reconciled with the estimate it prints. The model keeps the published
estimate, and the row is excluded from the gate above.

```{r ratio-figure, fig.height = 5, fig.width = 7}
ratios |>
  dplyr::filter(versus == "Reference") |>
  dplyr::select(scenario, sim_cmax, pub_cmax, sim_auc, pub_auc) |>
  tidyr::pivot_longer(-scenario, names_to = c("source", "metric"), names_sep = "_") |>
  dplyr::mutate(
    source = dplyr::recode(source, sim = "Model (typical value)", pub = "Paper (Figure 3 median)"),
    metric = dplyr::recode(metric, cmax = "Steady-state Cmax", auc = "24-h AUC")
  ) |>
  ggplot(aes(value, scenario, shape = source, colour = source)) +
  annotate("rect", xmin = 0.8, xmax = 1.25, ymin = -Inf, ymax = Inf, alpha = 0.15, fill = "steelblue") +
  geom_vline(xintercept = 1, linetype = 2) +
  geom_point(size = 2) +
  scale_x_log10() +
  facet_wrap(~metric) +
  labs(x = "Ratio to the reference participant (log scale)", y = NULL, shape = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Replicates Figure 3 of Wojciechowski 2022 (median ratios only).

# Stochastic simulation in atopic dermatitis

A virtual cohort of atopic-dermatitis patients receives the phase III tablet
once daily for 28 days at 100 mg or 200 mg (200 per arm). Covariates follow
the atopic-dermatitis column of Table 1: 45.5% female, 21.1% Asian and 2.6%
other race, 11.7% adolescents, body weight log-normal around the 73.9 kg median
(truncated to the observed 34.4-180 kg). Adolescents are given a lighter weight
distribution; the paper reports a typical adolescent of 61 kg.

```{r cohort}
rxode2::rxSetSeed(20220121)
set.seed(20220121)
n_per_arm <- 200

make_cohort <- function(n, daily_dose, id_offset) {
  adolescent <- rbinom(n, 1, 0.117)
  wt <- exp(rnorm(n, log(ifelse(adolescent == 1, 61, 75)), 0.24))
  wt <- pmin(pmax(wt, 34.4), 180)
  race <- sample(c("asian", "other", "white_black"), n, replace = TRUE,
                 prob = c(0.211, 0.026, 0.763))
  obs <- sort(unique(c(seq(0, 24, by = 0.5), seq(27 * 24, 28 * 24, by = 0.5))))
  dplyr::bind_rows(lapply(seq_len(n), function(i) {
    cov <- reference_covariates(
      WT = wt[i], SEXF = rbinom(1, 1, 0.455),
      RACE_ASIAN = as.integer(race[i] == "asian"),
      RACE_OTHER = as.integer(race[i] == "other"),
      ADOLESCENT = adolescent[i], DIS_ATOPIC_DERMATITIS = 1,
      DOSE_ABROCITINIB_MG = daily_dose, DOSE_ABROCITINIB_MGD = daily_dose
    )
    regimen_events(id_offset + i, cov, n_dose = 28, obs_times = obs)
  })) |>
    dplyr::mutate(treatment = paste(daily_dose, "mg QD"))
}

events <- dplyr::bind_rows(
  make_cohort(n_per_arm, 100, id_offset = 0L),
  make_cohort(n_per_arm, 200, id_offset = n_per_arm)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "treatment")])$id))

sim <- rxode2::rxSolve(mod, events, keep = "treatment", returnType = "data.frame")
stopifnot(!anyNA(sim$Cc))
```

```{r vpc, fig.height = 4, fig.width = 7}
sim |>
  dplyr::mutate(
    day = ifelse(time < 24 * 27, "Day 1", "Day 28 (steady state)"),
    tad = ifelse(time < 24 * 27, time, time - 27 * 24)
  ) |>
  dplyr::group_by(treatment, day, tad) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, p50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = 3) +
  scale_y_log10() +
  facet_wrap(~day) +
  labs(x = "Time after dose (h)", y = "Abrocitinib (ng/mL)", colour = NULL, fill = NULL) +
  theme_bw()
```

Median and 90% interval of simulated concentrations on day 1 and day 28; the
dotted line is the 1 ng/mL lower limit of quantification. The paper's own
prediction-corrected VPC (Figure 2) uses the analysis dataset, which is not
public, so it is not reproduced here.

# PKNCA validation

PKNCA is run on the steady-state (day 28) interval of the cohort above, by
dose group, and on a cohort of 200 White, male, adult, 80 kg
atopic-dermatitis patients at 200 mg once daily with random effects. The
latter matches the typical patient for which the paper reports a
steady-state Cmax of 1123 ng/mL and AUC of 5662 ng h/mL (Results 3.4).

```{r pknca}
ref_cov <- reference_covariates(WT = 80, DIS_ATOPIC_DERMATITIS = 1)
obs_ref <- seq(27 * 24, 28 * 24, by = 0.25)
events_ref <- dplyr::bind_rows(lapply(seq_len(n_per_arm), function(i) {
  regimen_events(2L * n_per_arm + i, ref_cov, n_dose = 28, obs_times = obs_ref)
})) |>
  dplyr::mutate(treatment = "200 mg QD, White male 80 kg AD")
sim_ref <- rxode2::rxSolve(mod, events_ref, keep = "treatment", returnType = "data.frame")

conc_all <- dplyr::bind_rows(sim, sim_ref) |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
dose_all <- dplyr::bind_rows(events, events_ref) |>
  dplyr::filter(evid == 1, cmt == "depot") |>
  dplyr::select(id, time, amt, treatment)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_all, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(dose_all, amt ~ time | treatment + id),
  intervals = data.frame(
    start = 27 * 24, end = 28 * 24,
    cmax = TRUE, tmax = TRUE, auclast = TRUE, cmin = TRUE
  )
))

as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "cmin")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = signif(median(PPORRES), 3), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Group" = treatment,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-24 (ng*h/mL)" = auclast,
    "Cmin (ng/mL)" = cmin
  ) |>
  knitr::kable(caption = "Median steady-state (day 28) NCA by group.")
```

## Comparison against published NCA

```{r nca-compare}
published_nca <- tibble::tribble(
  ~treatment, ~cmax, ~auclast,
  "200 mg QD, White male 80 kg AD", 1123, 5662
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_nca,
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = "Simulated (median) vs. published steady-state exposure for the typical White adult atopic-dermatitis patient (80 kg, 200 mg QD). * differs from the reference by >20%."
)

ref_nca <- as.data.frame(nca_res) |>
  dplyr::filter(treatment == "200 mg QD, White male 80 kg AD")
med_auc <- median(ref_nca$PPORRES[ref_nca$PPTESTCD == "auclast"])
med_cmax <- median(ref_nca$PPORRES[ref_nca$PPTESTCD == "cmax"])
typ_ad <- typ[typ$scenario == "AD, White, 80 kg", ]
stopifnot(
  # Deterministic: the typical-value AUC is F * dose / CL, which the paper
  # reports to four figures.
  abs(typ_ad$auclast / 5662 - 1) < 0.01,
  abs(typ_ad$cmax / 1123 - 1) < 0.05,
  # Cohort medians of 200 subjects: with 57.7% CV on CL the median AUC has a
  # sampling error of about 5%, so the bound allows three of those.
  abs(med_auc / 5662 - 1) < 0.20,
  abs(med_cmax / 1123 - 1) < 0.20
)
```

The typical-value AUC (`r round(typ_ad$auclast)` ng h/mL) matches the
paper's 5662 ng h/mL to within rounding. The typical-value Cmax
(`r round(typ_ad$cmax)` ng/mL) is about 2% below the published 1123 ng/mL,
consistent with the published value coming from a simulation that included
random effects on Vc.

# Assumptions and deviations

* **Two dose records per administration.** The model needs a `depot` record
  (`rate = 0`) and a `central` record (`rate = -1`) with the same `amt`, as in
  the source dataset (Online Resource 2 `CMT` / `RATE` definitions).
  rxode2 keeps the modelled rate and lets `f(central)` shorten the infusion,
  as NONMEM does. When the whole dose fits under the first-order cap (any
  tablet dose up to 121 mg) the central record carries no drug, and the model
  then drops its absorption lag; this changes nothing physically and avoids
  an rxode2 solver failure on lagged zero-amount infusions.
* **Time on treatment.** The source drives the time-dependent clearance with
  `TAFO`, the time after the first dose of a treatment period, which resets
  after a washout. The model uses the solver time `t`, so each treatment
  period must start at `t = 0`; simulate crossover periods as separate
  subjects. NONMEM evaluates `$PK` only at data records, so its clearance is
  piecewise constant between records, while rxode2 updates it continuously.
* **Repeated-dosing flag.** `MULTI_DOSE` is 0 for single doses and for the
  first dose of a regimen and 1 from the second dose on ("single or first
  dose vs. multiple dosing", Results 3.2). The source dataset carries the
  flag as a column; how its first-dose records were coded is not stated.
* **Covariate coding.** The five-level patient type (`PTST`) is split into
  `DIS_PSORIASIS`, `DIS_ATOPIC_DERMATITIS`, `HEPIMP_MILD` and `HEPIMP_MOD`
  (all 0 = healthy volunteer), which are mutually exclusive. The four-level
  formulation (`FORMS`) is split into three indicators; with all three at 0
  the dose is the phase II 100 mg tablet, which is the reference of every
  formulation effect in the control stream. The paper's simulation reference
  is the phase III tablet (`FORM_ABROCITINIB_TAB_PH3 = 1`). The fluconazole
  and fluvoxamine shift on F applies once if either drug is present.
  Food-not-controlled records (`FOOD = 2`) are grouped with fasted, as in
  the control stream.
* **800 mg dose.** The -0.778 shift on logit F applies when
  `DOSE_ABROCITINIB_MG` is exactly 800 mg, as in the control stream
  (`DOSE.EQ.800`); it is not extrapolated to larger doses.
* **Female sex.** The model keeps the published +0.353 logit-F shift, which
  does not reproduce the near-unity Female row of Figure 3 (see above).
* **Censoring.** The paper fitted below-quantification concentrations with
  the M3 method (LLOQ 1 ng/mL); that affects estimation only, and simulations
  here return uncensored concentrations.
* **Units.** Table 2 prints the maximum time-dependent change in CL as
  "-0.186 %"; the control stream uses it as a fraction
  (`1 + TAFOCLDELTA * (1 - exp(...))`), so CL falls by 18.6% at steady state.
* **Virtual cohort.** The atopic-dermatitis covariate distribution is an
  approximation of Table 1 (independent sampling, log-normal weight); it is
  used only for the illustrative percentile plot and the per-dose NCA.
