---
title: "Allopurinol (Wright 2016)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Allopurinol (Wright 2016)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r readModelDb("Wright_2016_allopurinol")()$reference`
* Article: <https://doi.org/10.1111/bcp.12799>

The model is a one-compartment population PK-PD model for allopurinol (via the
active metabolite oxypurinol) and plasma urate in adults with gout. Oxypurinol
disposition is a one-compartment, first-order-absorption / first-order-
elimination structure with Ka fixed at 1.09 1/h. Apparent oxypurinol
clearance (CL/F_oxy) is allometrically scaled on fat-free mass and
power-scaled on Cockcroft-Gault creatinine clearance standardised to 70 kg;
when a thiazide or loop diuretic is coadministered, CL/F_oxy is reduced by
26%. Apparent oxypurinol volume (V/F_oxy) is allometrically scaled on total
body weight (exponent fixed at 1.0) and shares its IIV with CL/F_oxy via a
fixed fractional scaler (Bonate 2006 fractional-effect parameterisation;
F_v_oxy = 0.0355).

Plasma urate is described by a direct-effect sigmoidal Emax inhibition of
urate production on top of a baseline urate U0. The baseline is power-scaled
on renal function (lower renal function -> mildly higher U0; exponent -0.119)
and multiplicatively higher with concomitant diuretic (+14%). A turnover
model for urate was tested by the authors but did not improve the fit and
was unstable.

# Population

The model was fit to the pooled five-study cohort summarised in Wright 2016
Table 2: 133 gout patients (87% male) from New Zealand. Median age 60 years
(range 27 to 83), median total body weight 94 kg (range 51 to 171), median
creatinine clearance 68 mL/min (range 12 to 125, raw Cockcroft-Gault).
Roughly 20% of subjects were of Maori or Pacific Islander descent
consistent with the high regional prevalence of severe gout. 33% (44 of
133) were on a thiazide or loop diuretic at study entry. The pooled
dataset included 1105 oxypurinol and 1162 urate plasma concentrations
across the five studies (Study 1 was the largest, n = 74, from Stamp 2011).
Allopurinol daily doses ranged 50-700 mg (median 300 mg at study entry).

The same information is available programmatically via the model's
`population` metadata:

```{r show-population}
pop <- readModelDb("Wright_2016_allopurinol")()$population
str(pop, max.level = 1, give.attr = FALSE)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in
`inst/modeldb/specificDrugs/Wright_2016_allopurinol.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` = log(1.09) (fixed) | Ka = 1.09 1/h | Wright 2016 Table 3 (Ka fixed both base and final columns) |
| `lcl` = log(1.32) | CL/F_oxy = 1.32 L/h at reference | Wright 2016 Table 3 thetaCL (RSE 3.9%) |
| `lvc` = log(41.6) | V/F_oxy = 41.6 L at reference TBW | Wright 2016 Table 3 thetaV (RSE 3.0%) |
| `e_ffm_cl` = 0.75 (fixed) | FFM allometric exponent on CL | Wright 2016 Methods (Anderson and Holford) |
| `e_wt_vc` = 1.0 (fixed) | TBW allometric exponent on V | Wright 2016 Methods (Anderson and Holford) |
| `e_crcl_cl` = 0.587 | thetaRFexp | Wright 2016 Table 3 (RSE 11.7%) |
| `e_conmed_diur_cl` = 0.740 | thetadiuretic | Wright 2016 Table 3 (RSE 6.4%) |
| `lemax` = log(0.409) | Emax = 0.409 | Wright 2016 Table 3 Emax (RSE 12%) |
| `lec50` = log(83.9) | C50 = 83.9 umol/L | Wright 2016 Table 3 C50 (RSE 17.4%) |
| `lhill` = log(1.30) | lambda (Hill) = 1.30 | Wright 2016 Table 3 lambda (RSE 11%) |
| `lrbase` = log(0.511) | U0 = 0.511 mmol/L | Wright 2016 Table 3 U0 (RSE 2.3%) |
| `e_crcl_rbase` = -0.119 | thetaE0_RFexp | Wright 2016 Table 3 (RSE 21.6%) |
| `e_conmed_diur_rbase` = 1.14 | thetaE0_diuretic | Wright 2016 Table 3 (RSE 1.8%) |
| `etalcl` ~ 0.0586 | omega_CL = 24.2% | Wright 2016 Table 3 omegaCL (RSE 15.5%) |
| `F_v_oxy` = 0.0355 (fixed) | V fractional shared-eta scaler | Wright 2016 Table 3 F_omegaV_oxy |
| `etalka` ~ fixed(0.347) | omega_Ka = 58.9% fixed | Wright 2016 Table 3 (fixed both columns) |
| `(etalrbase, etalemax, etalec50)` block | omega_U0 14.2%, omega_Emax 35.9%, omega_C50 60.7%; Cov(emax, U0) = 0.025; Cov(emax, C50) = 0.193; Cov(U0, C50) = 0.011 | Wright 2016 Table 3 omega and Covar rows |
| `propSd` = 0.199 | Cc proportional residual 19.9% CV | Wright 2016 Table 3 oxypurinol sigma_prop |
| `addSd` = fixed(0.001) | Cc additive residual (fixed, negligible) | Wright 2016 Table 3 footer |
| `addSd_Eurate` = 0.037 | Urate additive residual 0.037 mmol/L (Study 1 / 3 representative) | Wright 2016 Table 3 Study 1 sigma_add |
| `propSd_Eurate` = fixed(0.001) | Urate proportional residual (fixed, negligible) | Wright 2016 Table 3 footer |
| ODE: `d/dt(depot) = -ka * depot` | n/a | Wright 2016 PK Models section |
| ODE: `d/dt(central) = ka * depot - kel * central` | n/a | Wright 2016 PK Models section |
| Observable: `Cc = central / vc * 1000 / MW_allop` (MW_allop = 136.11 g/mol) | n/a | Implicit molar conversion: dose mg allopurinol -> umol/L oxypurinol |
| Observable: `Eurate = U0 * (1 - Emax * Cc^lambda / (C50^lambda + Cc^lambda))` | n/a | Wright 2016 Results equation set (page 282) |

# Virtual cohort

Original observed data are not publicly available. The figures below use
virtual populations whose covariate distributions approximate the published
trial demographics: a male-majority adult-gout cohort with median total
body weight 94 kg and median CRCL standardised to 70 kg ~= 4.08 L/h
(= 68 mL/min in the raw Cockcroft-Gault expression).

Cohort size is kept at 200 subjects per arm per the skill convention
(adequate for VPC; larger cohorts increase render time without
validation benefit).

```{r cohort-helpers}
set.seed(20260630)

# Janmahasatian fat-free mass (kg) from total body weight (kg), height
# (cm), and SEXF (1 = female, 0 = male). Wright 2016 cites this formula
# as the FFM derivation used in the model fit.
ffm_janmahasatian <- function(WT, HT_cm, SEXF) {
  HT_m  <- HT_cm / 100
  BMI   <- WT / HT_m^2
  WHS_max <- ifelse(SEXF == 1, 37.99, 42.92)
  WHS_50  <- ifelse(SEXF == 1, 35.98, 30.93)
  WHS_max * HT_m^2 * (WT / (WHS_50 * HT_m^2 + WT))
}

# Build one cohort of n synthetic subjects. id_offset shifts the integer
# id range so cohorts can be bind_rows()-ed without colliding.
make_cohort <- function(n, daily_dose_mg, diuretic = 0L, id_offset = 0L,
                        cohort_label = "") {
  tibble::tibble(
    id     = id_offset + seq_len(n),
    SEXF   = rbinom(n, 1, prob = 0.13),                       # ~13% female per Wright 2016 Table 2
    WT     = pmin(pmax(round(rnorm(n, mean = 94, sd = 18)), 51), 171),
    HT_cm  = pmin(pmax(round(rnorm(n, mean = 173, sd = 9)), 150), 195),
    CRCL_mLmin = pmin(pmax(round(rnorm(n, mean = 68, sd = 22)), 20), 125),
    CONMED_DIUR = as.integer(diuretic),
    cohort   = cohort_label,
    treatment = paste0(daily_dose_mg, " mg/day"),
    daily_dose_mg = daily_dose_mg
  ) |>
    dplyr::mutate(
      FFM  = ffm_janmahasatian(WT, HT_cm, SEXF),
      # Wright 2016's CRCL is Cockcroft-Gault expressed in L/h
      # standardised to 70 kg: CRCL_L_h_70kg = CRCL_mLmin * (70 / WT) / 16.6667
      # 16.6667 mL/min per L/h conversion.
      CRCL = CRCL_mLmin * (70 / WT) / 16.6667
    )
}

# Build a dosing + observation event table for one cohort. The dose is
# allopurinol mg given q24h for `dose_days` consecutive days, with urate
# and oxypurinol sampling at one-hour intervals over the final
# 24-hour interval (= steady-state-relevant window) and a baseline
# pre-dose sample at t = 0.
build_events <- function(subj, dose_days = 28, obs_grid_h = NULL) {
  if (is.null(obs_grid_h)) {
    # observation grid: every 4 hours over the LAST 24 h plus a baseline
    last_day_start <- (dose_days - 1) * 24
    obs_grid_h <- c(0, seq(last_day_start, last_day_start + 24, by = 4))
  }
  doses <- subj |>
    dplyr::select(id, daily_dose_mg, FFM, WT, CRCL, CONMED_DIUR,
                  cohort, treatment) |>
    tidyr::expand_grid(time = seq(0, by = 24, length.out = dose_days)) |>
    dplyr::mutate(amt = daily_dose_mg, evid = 1L, cmt = "depot",
                  dvid = NA_integer_)
  obs_cc <- subj |>
    dplyr::select(id, FFM, WT, CRCL, CONMED_DIUR, cohort, treatment) |>
    tidyr::expand_grid(time = obs_grid_h) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "Cc",
                  dvid = 1L)
  obs_urate <- subj |>
    dplyr::select(id, FFM, WT, CRCL, CONMED_DIUR, cohort, treatment) |>
    tidyr::expand_grid(time = obs_grid_h) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "Eurate",
                  dvid = 2L)
  dplyr::bind_rows(doses, obs_cc, obs_urate) |>
    dplyr::arrange(id, time, evid)
}
```

```{r build-cohorts}
# Six dose levels at fixed adult-typical covariates; reproduces the
# Wright 2016 Figure 3 stratification (100 / 200 / 300 / 400 / 600 mg
# daily for ~3 months). Wright 2016 uses 1000 subjects per dose for
# its Figure 3 simulation; we use 200/arm per skill convention.
N_PER_ARM <- 200L
doses <- c(100, 200, 300, 400, 600)

cohorts <- dplyr::bind_rows(lapply(seq_along(doses), function(i) {
  make_cohort(n = N_PER_ARM,
              daily_dose_mg = doses[i],
              diuretic = 0L,
              id_offset = (i - 1L) * N_PER_ARM,
              cohort_label = paste0(doses[i], " mg/day"))
}))

events_dose <- build_events(cohorts, dose_days = 90)

# Assertion: ids disjoint and (id, time, evid) unique per skill convention.
stopifnot(!anyDuplicated(unique(events_dose[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
mod <- readModelDb("Wright_2016_allopurinol")
sim <- rxode2::rxSolve(mod, events = events_dose,
                       keep = c("cohort", "treatment", "WT",
                                "CRCL", "CONMED_DIUR")) |>
  as.data.frame()
nrow(sim)
```

# Replicate Figure 3 -- percent urate reduction by daily dose

Wright 2016 Figure 3 compares the model-predicted distribution of percent
urate reduction (from baseline) under 100, 200, 300, 400, and 600 mg daily
allopurinol against external literature data. We reproduce the model side
of that figure by computing the percent reduction from baseline (= U0 at
the typical patient covariates) at steady state.

```{r figure-3}
# Steady-state percent reduction from baseline urate, per subject:
# pick the urate observation at the last simulated time point in each
# subject's series (= near-steady-state under chronic q24h dosing).
ss <- sim |>
  dplyr::filter(!is.na(Eurate)) |>
  dplyr::group_by(id, cohort, treatment) |>
  dplyr::filter(time == max(time)) |>
  dplyr::ungroup()

# Baseline U0 per subject = Eurate at t = 0 (pre-dose).
baseline <- sim |>
  dplyr::filter(time == 0, !is.na(Eurate)) |>
  dplyr::select(id, baseline = Eurate)

ss <- ss |>
  dplyr::inner_join(baseline, by = "id") |>
  dplyr::mutate(pct_reduction = 100 * (baseline - Eurate) / baseline,
                treatment = factor(treatment, levels = paste0(doses, " mg/day")))

ggplot(ss, aes(treatment, pct_reduction)) +
  geom_boxplot(fill = "grey80") +
  labs(x = "Allopurinol daily dose",
       y = "% reduction in plasma urate from baseline",
       title = "Replicates the model side of Figure 3 (Wright 2016)",
       caption = "Each box = 200 simulated subjects at the typical adult-gout covariate distribution.") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

# Median per dose (for the comparison table below).
median_table <- ss |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(median_pct = round(median(pct_reduction), 1),
                   q25_pct    = round(quantile(pct_reduction, 0.25), 1),
                   q75_pct    = round(quantile(pct_reduction, 0.75), 1),
                   .groups = "drop")

median_table |>
  dplyr::rename("Treatment"                = treatment,
                "Median % reduction"       = median_pct,
                "Q1 % reduction"           = q25_pct,
                "Q3 % reduction"           = q75_pct) |>
  knitr::kable(caption = "Simulated steady-state percent urate reduction by daily allopurinol dose, typical-cohort covariates.")
```

# Replicate dosing predictions -- Table 4 (no diuretic, CLcr 50-70)

Wright 2016 Table 4 lists the maintenance dose required to achieve plasma
urate <= 0.36 mmol/L in > 75% of subjects across a grid of body weights and
CRCL strata. Here we reproduce the > 75% success criterion for the
CLcr 50-70 mL/min stratum at 70 kg total body weight, no diuretic (Table 4
target: 350 mg/day).

```{r table-4-check}
# Build a 70 kg adult, no-diuretic cohort with CRCL drawn from a
# uniform U(50, 70) mL/min, then test three candidate doses around
# the Table 4 target of 350 mg/day.
make_table4_cohort <- function(n, daily_dose_mg, id_offset, cohort_label) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    SEXF = 0L,
    WT = 70,
    HT_cm = 175,
    CRCL_mLmin = runif(n, 50, 70),
    CONMED_DIUR = 0L,
    cohort = cohort_label,
    treatment = paste0(daily_dose_mg, " mg/day"),
    daily_dose_mg = daily_dose_mg
  ) |>
    dplyr::mutate(
      FFM = ffm_janmahasatian(WT, HT_cm, SEXF),
      CRCL = CRCL_mLmin * (70 / WT) / 16.6667
    )
}

t4_cohorts <- dplyr::bind_rows(
  make_table4_cohort(N_PER_ARM, 300L, id_offset = 0L,             cohort_label = "300 mg/day"),
  make_table4_cohort(N_PER_ARM, 350L, id_offset = N_PER_ARM,      cohort_label = "350 mg/day"),
  make_table4_cohort(N_PER_ARM, 400L, id_offset = 2L * N_PER_ARM, cohort_label = "400 mg/day")
)
ev_t4 <- build_events(t4_cohorts, dose_days = 90)
stopifnot(!anyDuplicated(unique(ev_t4[, c("id", "time", "evid")])))

sim_t4 <- rxode2::rxSolve(mod, events = ev_t4,
                          keep = c("cohort", "treatment", "WT", "CRCL")) |>
  as.data.frame()

# Steady-state urate at the last sampled time.
ss_t4 <- sim_t4 |>
  dplyr::filter(!is.na(Eurate)) |>
  dplyr::group_by(id, treatment) |>
  dplyr::filter(time == max(time)) |>
  dplyr::ungroup() |>
  dplyr::mutate(below_target = Eurate <= 0.36)

success_table <- ss_t4 |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(pct_below_target = round(100 * mean(below_target), 1),
                   median_urate = round(median(Eurate), 3),
                   .groups = "drop")

success_table |>
  dplyr::rename("Treatment" = treatment,
                "% subjects with urate <= 0.36 mmol/L" = pct_below_target,
                "Median steady-state urate (mmol/L)" = median_urate) |>
  knitr::kable(caption = "Replicate of Wright 2016 Table 4 row: 70 kg, no diuretic, CLcr 50-70 mL/min. Table 4 target dose = 350 mg/day.")
```

# PKNCA validation on oxypurinol

Use PKNCA to compute Cmax, Tmax, and AUC over a 24-hour steady-state
interval for the 300 mg/day dose group (the typical clinical regimen).
Wright 2016 does not publish formal NCA for oxypurinol (only descriptive
medians per study, Table 2), so this section is a sanity check that the
PK structure produces plausible numbers rather than a head-to-head
comparison.

```{r pknca}
sim_for_nca <- sim |>
  dplyr::filter(treatment == "300 mg/day", !is.na(Cc))

# Carve out the last 24 h of the simulation as a steady-state interval.
ss_window_start <- max(sim_for_nca$time) - 24

sim_nca <- sim_for_nca |>
  dplyr::filter(time >= ss_window_start) |>
  dplyr::mutate(time = time - ss_window_start) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time=0 (relative to interval start) Cc value per (id, treatment).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |>
    dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = NA_real_)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time) |>
  tidyr::fill(Cc, .direction = "downup") |>
  dplyr::filter(!is.na(Cc))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)

dose_df <- events_dose |>
  dplyr::filter(evid == 1, treatment == "300 mg/day") |>
  dplyr::group_by(id) |>
  dplyr::filter(time == max(time)) |>
  dplyr::ungroup() |>
  dplyr::mutate(time = 0) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE
)
nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res <- PKNCA::pk.nca(nca_data)

nca_summary <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(median = round(median(PPORRES, na.rm = TRUE), 2),
                   .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::rename("Treatment"   = treatment,
                "Median Cmax (umol/L)" = cmax,
                "Median Tmax (h)"      = tmax,
                "Median AUC0-24 (umol*h/L)" = auclast) |>
  knitr::kable(caption = "Simulated oxypurinol steady-state NCA (300 mg/day, no diuretic, 200 subjects). Wright 2016 reports a typical patient's average oxypurinol concentration of ~70 umol/L at 226 mg/day (Discussion); the 300 mg/day arm here should sit slightly above that value.")
```

# Assumptions and deviations

- **Year on the citation.** The task metadata named the paper "Wright 2015"
  because the publisher posted the accepted-article PDF online on 9 October
  2015. The final, citeable BJCP volume issue (81:2, 277-289) is dated
  February 2016 with the same DOI (10.1111/bcp.12799); we use the 2016
  year throughout the model file, vignette, and filename for the canonical
  citation, matching how the paper is indexed in PubMed.
- **Per-study urate residual error.** Wright 2016 reports five different
  urate additive residual SDs across the five pooled studies (0.021,
  0.022, 0.037, 0.037, 0.054 mmol/L). The model file retains a single
  representative value of 0.037 mmol/L (Studies 1 and 3; Study 1 is the
  largest single sub-cohort at n = 74 of 133 and the median across
  studies). A user wanting to reproduce a specific sub-study's VPC should
  override `addSd_Eurate` accordingly.
- **CONMED_DIUR class membership.** Wright 2016's CONMED_DIUR captures
  thiazide diuretics OR loop diuretics (per Wright 2016 Table 4 footer).
  This is NARROWER than the Stocker 2012 oxypurinol precedent
  (`Stocker_2012_oxypurinol.R`), which pools thiazide + loop +
  spironolactone into the same column. The two papers use the same
  canonical CONMED_DIUR column name; users simulating across both models
  must populate the indicator according to each paper's own definition.
- **IIV scale convention.** Wright 2016 reports omega values as "(CV%)"
  but the reported PD covariance block is only positive-definite when the
  reported %CV is interpreted as the log-scale SD directly (i.e.,
  omega^2 = (CV / 100)^2) rather than the back-transformed CV (which
  would give omega^2 = log(1 + CV^2)). The model file uses the
  log-scale-SD interpretation throughout for internal consistency.
- **V's "fractional effect" IIV.** Wright 2016 notes that BSV on V was not
  identifiable; only a single random effect was estimated for both CL
  and V, with V's individual deviation a fixed fraction (F_v_oxy =
  0.0355) of CL's. We encode this exactly: V_i = V_pop *
  exp(F_v_oxy * etalcl), shared etalcl, with F_v_oxy held fixed.
- **Dose unit and molar conversion.** The dose entered into the model is
  allopurinol mass in mg. The model's apparent CL/F_oxy and V/F_oxy
  absorb the implicit F_oxy (fraction absorbed * fraction metabolised to
  oxypurinol; 1:1 molar) so the central compartment carries
  allopurinol-equivalent mg. The observable Cc converts to umol/L
  oxypurinol using the allopurinol molecular weight 136.11 g/mol; this
  matches the paper's apparent-CL definition and reproduces the cohort
  median oxypurinol concentration of ~70 umol/L at 226 mg/day reported
  in the Wright 2016 Discussion.
- **CRCL standardisation.** The model expects CRCL in L/h standardised to
  70 kg (Wright 2016 reference: 100 mL/min/70 kg = 6 L/h/70 kg). The
  virtual cohort code in this vignette converts raw Cockcroft-Gault
  mL/min to the 70-kg-standardised L/h form via
  `CRCL = CRCL_mLmin * (70 / WT) / 16.6667`. Users supplying CRCL
  directly should standardise externally.
- **Genotype covariates.** Wright 2016 tested four urate-transporter SNPs
  (SLC2A9, ABCG2, NPT1, URAT1) and one of them (ABCG2 T-allele
  homozygote) approached significance on C50 (+40%) but did not meet the
  backward-elimination criterion. The model file does not include this
  effect because it was not retained; users who want to explore it can
  add a multiplier on `lec50` manually.
- **No external NCA validation.** Wright 2016 does not publish formal NCA
  parameters for oxypurinol (only cohort medians; Table 2). The PKNCA
  block above is a sanity check, not a head-to-head comparison.
