---
title: "Sitafloxacin (Wu 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sitafloxacin (Wu 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20250214)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wu_2025_sitafloxacin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wu_2025_sitafloxacin"))$description`
* Article: <https://doi.org/10.3389/fphar.2025.1476158>

Sitafloxacin is a fourth-generation fluoroquinolone marketed in Japan, Thailand
and China. Wu 2025 pooled 12 clinical trials to build a population PK model, then
used it to derive pharmacokinetic/pharmacodynamic (PK/PD) cut-offs -- the
exposure-based half of a clinical susceptibility breakpoint -- against five
target pathogens.

```{r load-model}
mod <- readModelDb("Wu_2025_sitafloxacin")
mod
```

# Population

The estimation dataset held 3,294 plasma concentrations from 342 subjects across
12 studies (Wu 2025 Section 2.1 and Section 4.1). Eleven studies were run in
Japan -- 147 patients with respiratory-system infection, 12 subjects with renal
insufficiency and 135 healthy subjects -- and one phase 1 study in China
contributed 48 healthy subjects. 246 subjects (71.9%) were male and 96 (28.1%)
female. Baseline demographics (Wu 2025 Table 1) were age 43.0 +/- 21.7 years,
body weight 58.2 +/- 10.2 kg, height 164.7 +/- 9.6 cm, BMI 21.4 +/- 3.0 kg/m^2
and creatinine clearance 115.1 +/- 53.5 mL/min (mean +/- SD). Dosing spanned
single doses of 3-200 mg and multiple doses of 50 or 100 mg q12h or 100 mg q8h
for 7-14 days, all oral. 1.02% of concentration records were missing or excluded.

The model was fitted in NONMEM 7.4 using FOCE with interaction and qualified by a
1,000-replicate bootstrap (945 successful, 94.5% convergence) and a
1,000-replicate visual predictive check.

The same information is available programmatically:

```{r population}
str(readModelDb("Wu_2025_sitafloxacin")()$population)
```

# Source trace

Every value below is also carried as an in-file comment beside its `ini()` entry
in `inst/modeldb/specificDrugs/Wu_2025_sitafloxacin.R`.

Wu 2025 prints the final model as a block of display equations in Section 2.2,
immediately above the Table 2 abbreviation note:

```
CL   = 14.7 * (CRCL/106.88)^0.46                     * exp(eta1)   (L/h)
V2   = 89.8 * (WT/58.55)^0.966 * (AGE/31)^0.286      * exp(eta2)   (L)
V3   = 33.8                                                        (L)
Q    = 5.81                                                        (L/h)
D1   = 0.281 * (1 + FOOD)^1.59                       * exp(eta3)   (h)
TLAG = 0.205                                                       (h)
Absorption: zero-order, fasting D1 = 0.281 h; fed D1 = 0.846 h
            first-order, Ka = 5.19 1/h
```

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 14.7 L/h | Table 2, row `CL` (RSE 1.8%) |
| `lvc` (V2/F) | 89.8 L | Table 2, row `V2` (RSE 2.8%) |
| `lq` (Q/F) | 5.81 L/h | Table 2, row `Q` (RSE 6.9%) |
| `lvp` (V3/F) | 33.8 L | Table 2, row `V3` (RSE 3.6%) |
| `lka` (ka) | 5.19 1/h | Table 2, row `Ka` (RSE 16.3%) |
| `ld1` (D1, fasting) | 0.281 h | Table 2, row `D1` (RSE 4.2%); Section 2.2 "in fasting state D1(h) = 0.281" |
| `ltlag` (TLAG) | 0.205 h | Table 2, row `TLAG` (RSE 1.1%) |
| `e_crcl_cl` | 0.460 | Table 2, row `CRCL on CL` (RSE 6.5%); Section 2.2 equation `(CRCL/106.88)^0.46` |
| `e_wt_vc` | 0.966 | Table 2, row `WT on V2` (RSE 13.5%); Section 2.2 equation `(WT/58.55)^0.966` |
| `e_age_vc` | 0.286 | Table 2, row `Age on V2` (RSE 17.9%); Section 2.2 equation `(AGE/31)^0.286` |
| `e_fed_d1` | 1.59 | Table 2, row `Food on D1` (RSE 3.8%); Section 2.2 equation `(1 + FOOD)^1.59` |
| `etalcl` | var 0.0515 | Table 2, `eta1` under "Interindividual variability (omega^2)"; footnote "eta1, interindividual variability of CL" |
| `etalvc` | var 0.0635 | Table 2, `eta2`; footnote "eta2, ... of V2" |
| `etald1` | var 3.46 | Table 2, `eta3`; footnote "eta3, ... of D1" |
| `propSd` | sqrt(0.0361) = 0.19 | Table 2, `epsilon1` under "Residual variability (sigma^2)"; footnote "epsilon1, proportional residual error" |
| `addSd` | sqrt(2.35e-5) = 0.0048477 mg/L | Table 2, `epsilon2`; footnote "epsilon2, additive residuals" |
| Reference CRCL 106.88 mL/min, WT 58.55 kg, AGE 31 years | n/a | printed inside the Section 2.2 display equations |
| Two-compartment disposition; sequential zero-order then first-order absorption | n/a | Section 2.2, "A two-compartment model with a linear elimination model best described sitafloxacin PKs ... The absorption of sitafloxacin involves zero-order and first-order kinetics" |
| `eta4` (IOV, var 0.0279) | **not encoded** | Table 2, `eta4`; footnote gives only "interindividual variability of IOV" and no host parameter -- see Errata |

## Confirming the food effect arithmetic

The food effect is printed in an unusual `(1 + FOOD)^theta` power form rather than
the customary `exp(theta * FOOD)`. Wu 2025 supplies its own worked value for the
fed state in the same display equation, which lets the form and the indicator
orientation be checked directly.

```{r food-arithmetic}
d1_fasting <- 0.281
d1_fed_computed <- 0.281 * (1 + 1)^1.59
c(
  paper_fasting   = 0.281,
  model_fasting   = d1_fasting,
  paper_fed       = 0.846,
  computed_fed    = round(d1_fed_computed, 4)
)
stopifnot(abs(d1_fed_computed - 0.846) < 0.001)
```

The computed fed-state duration reproduces the paper's printed 0.846 h, so
`FED = 1` is the fed state and the literal power form is correct.

# Virtual cohort

The original subject-level data are not public. The cohort below reproduces the
Wu 2025 Table 1 marginal distributions, restricted to creatinine clearance
>= 50 mL/min because Wu 2025 Section 4.2 states the PK/PD simulations covered
"patients with normal renal function and mild renal insufficiency
(CRCL >= 50 mL/min)".

Age, weight and creatinine clearance are drawn as independent log-normals matched
to the published means and SDs. Independence and log-normality are assumptions --
Wu 2025 reports only marginal mean +/- SD and no correlation structure.

```{r cohort}
n_subj <- 200L

# Moment-match a log-normal to a reported mean and SD.
lnorm_par <- function(mean_val, sd_val) {
  sdlog <- sqrt(log(1 + (sd_val / mean_val)^2))
  list(meanlog = log(mean_val) - sdlog^2 / 2, sdlog = sdlog)
}

rlnorm_mom <- function(n, mean_val, sd_val) {
  p <- lnorm_par(mean_val, sd_val)
  stats::rlnorm(n, p$meanlog, p$sdlog)
}

# CRCL is redrawn until every subject clears the 50 mL/min floor Wu 2025 used.
draw_crcl <- function(n) {
  out <- rlnorm_mom(n, 115.1, 53.5)
  while (any(out < 50)) {
    bad <- out < 50
    out[bad] <- rlnorm_mom(sum(bad), 115.1, 53.5)
  }
  out
}

cohort <- tibble::tibble(
  id   = seq_len(n_subj),
  WT   = rlnorm_mom(n_subj, 58.2, 10.2),
  AGE  = pmin(pmax(rlnorm_mom(n_subj, 43.0, 21.7), 18), 90),
  CRCL = draw_crcl(n_subj),
  FED  = 0
)

summary(dplyr::select(cohort, WT, AGE, CRCL))
```

```{r cohort-check}
# The cohort must reproduce the published marginals it was built from.
stopifnot(
  abs(mean(cohort$WT) - 58.2) < 2,
  abs(mean(cohort$AGE) - 43.0) < 4,
  all(cohort$CRCL >= 50),
  nrow(cohort) == n_subj
)
```

# Simulation

Wu 2025 Section 4.2 simulated three regimens for 7 days: 50 mg q12h, 100 mg q24h
and 100 mg q12h. All three arms use the **same 200 subjects** (common random
numbers), so between-regimen differences are driven only by the regimen.

Dose records carry `rate = -2` so that rxode2 uses the model's `dur(depot)`; a
plain bolus would collapse the zero-order absorption window.

```{r regimens}
regimens <- tibble::tribble(
  ~regimen,        ~dose, ~tau, ~daily_dose,
  "50 mg q12h",     50,    12,   100,
  "100 mg q24h",   100,    24,   100,
  "100 mg q12h",   100,    12,   200
)

n_days   <- 7
t_end    <- n_days * 24            # 168 h
ss_start <- (n_days - 1) * 24      # 144 h -- final 24-hour window

# Coarse grid up to the final day, dense grid across the final 24 h where the
# steady-state AUC is measured.
obs_times <- sort(unique(c(
  seq(0, ss_start, by = 0.5),
  seq(ss_start, t_end, by = 0.1)
)))

build_events <- function(reg_row, cohort) {
  dose_times <- seq(0, t_end - reg_row$tau, by = reg_row$tau)

  doses <- tidyr::crossing(cohort, time = dose_times) |>
    dplyr::mutate(amt = reg_row$dose, evid = 1L, cmt = "depot", rate = -2)

  obs <- tidyr::crossing(cohort, time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central", rate = 0)

  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(regimen = reg_row$regimen) |>
    as.data.frame()
}

sim_one <- function(reg_row, cohort) {
  ev <- build_events(reg_row, cohort)
  set.seed(20250214)   # common random numbers: identical etas across regimens
  out <- rxode2::rxSolve(
    mod, events = ev,
    keep = c("WT", "AGE", "CRCL", "FED", "regimen"),
    returnType = "data.frame"
  )
  if (is.null(out$id)) out$id <- 1L
  out
}

sim <- do.call(
  rbind,
  lapply(seq_len(nrow(regimens)), function(i) sim_one(regimens[i, ], cohort))
)

# rxSolve silently drops subjects on some failure modes -- assert, do not assume.
stopifnot(
  dplyr::n_distinct(sim$id) == n_subj,
  dplyr::n_distinct(sim$regimen) == nrow(regimens),
  all(!is.na(sim$Cc)),
  all(sim$Cc >= 0)
)
nrow(sim)
```

## Typical-value concentration-time profile

```{r typical-profile, fig.cap = "Typical-value steady-state sitafloxacin profiles for the three Wu 2025 regimens. IIV is suppressed with rxode2::zeroRe()."}
mod_typ <- rxode2::zeroRe(mod)

typ_cohort <- tibble::tibble(
  id = 1L, WT = 58.55, AGE = 31, CRCL = 106.88, FED = 0
)

typ <- do.call(rbind, lapply(seq_len(nrow(regimens)), function(i) {
  ev <- build_events(regimens[i, ], typ_cohort)
  out <- rxode2::rxSolve(mod_typ, events = ev,
                         keep = c("regimen"), returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out
}))

ggplot(typ, aes(time, Cc, colour = regimen)) +
  geom_line(linewidth = 0.6) +
  labs(x = "Time (h)", y = "Sitafloxacin concentration (mg/L)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

At the reference covariate vector (CRCL 106.88 mL/min, WT 58.55 kg, AGE 31 years,
fasting) every covariate multiplier collapses to 1, so the individual parameters
must equal the Table 2 typical values exactly.

```{r reference-collapse}
ref_par <- typ |>
  dplyr::filter(regimen == "100 mg q24h") |>
  dplyr::slice(1) |>
  dplyr::select(cl, vc, vp, q, ka, d1, tlag)

published <- c(cl = 14.7, vc = 89.8, vp = 33.8, q = 5.81,
               ka = 5.19, d1 = 0.281, tlag = 0.205)

cmp_ref <- tibble::tibble(
  Parameter = names(published),
  Published = as.numeric(published),
  Simulated = round(as.numeric(ref_par[names(published)]), 4)
)
knitr::kable(cmp_ref, caption = "Typical-value parameters at the reference covariate vector vs Wu 2025 Table 2.")

stopifnot(all(abs(cmp_ref$Simulated - cmp_ref$Published) < 1e-6))
```

## Absorption lag

The model applies `alag(depot) <- tlag` with TLAG = 0.205 h, so the concentration
must be exactly zero before 0.205 h and positive afterwards.

```{r lag-check}
first_dose <- typ |>
  dplyr::filter(regimen == "100 mg q24h", time <= 1)

max_conc_before_lag <- max(first_dose$Cc[first_dose$time < 0.205])
min_conc_after_lag  <- min(first_dose$Cc[first_dose$time > 0.5])

c(max_before_lag = max_conc_before_lag, min_after_lag = round(min_conc_after_lag, 4))
stopifnot(max_conc_before_lag == 0, min_conc_after_lag > 0)
```

## Falsifier: the peripheral compartment is actually solved

`rxSolve()` defaults to `useLinCmt = TRUE`, which rewrites a recognisably linear
system into closed form. On some two-compartment models written with
micro-constants that rewrite silently drops the peripheral compartment and
solves a one-compartment model instead. Total AUC is unaffected by that failure
(it is still exactly dose / CL), so the `AUC24h = daily dose / CL` identity above
cannot detect it -- the only readout that moves decisively is the terminal
half-life, which comes back equal to `log(2) / kel`.

The check below compares the simulated terminal slope against the analytic beta
root of `x^2 - (kel + k12 + k21) x + kel * k21 = 0` computed from the published
Table 2 typical values, and prints the one-compartment value alongside so a
regression is unambiguous.

```{r terminal-halflife}
hl_times <- sort(unique(c(seq(0, 4, by = 0.05), seq(4, 72, by = 0.25))))

ev_hl <- dplyr::bind_rows(
  data.frame(time = 0, amt = 100, evid = 1L, cmt = "depot", rate = -2),
  data.frame(time = hl_times, amt = NA_real_, evid = 0L, cmt = "central", rate = 0)
) |>
  dplyr::mutate(id = 1L, WT = 58.55, AGE = 31, CRCL = 106.88, FED = 0) |>
  dplyr::arrange(time, dplyr::desc(evid)) |>
  as.data.frame()

hl_sim <- rxode2::rxSolve(mod_typ, events = ev_hl, returnType = "data.frame")

# Terminal slope from the log-linear tail, well past the distribution phase.
tail_fit  <- stats::lm(log(Cc) ~ time, data = subset(hl_sim, time >= 36 & Cc > 0))
hl_simval <- log(2) / -unname(coef(tail_fit)[2])

# Analytic beta root from the published Table 2 typical values.
kel_p  <- 14.7 / 89.8
k12_p  <- 5.81 / 89.8
k21_p  <- 5.81 / 33.8
beta_p <- min(Re(polyroot(c(kel_p * k21_p, -(kel_p + k12_p + k21_p), 1))))

c(
  t_half_analytic_beta = round(log(2) / beta_p, 4),
  t_half_simulated     = round(hl_simval, 4),
  t_half_if_collapsed_to_1cmt = round(log(2) / kel_p, 4)
)

stopifnot(
  "peripheral1 retained in the solve" = "peripheral1" %in% names(hl_sim),
  abs(hl_simval / (log(2) / beta_p) - 1) < 0.01
)
```

## Food effect

The food effect sits entirely on the zero-order absorption duration. Because a
zero-order input conserves the delivered dose, food must reshape the absorption
phase (lower, later peak) while leaving the exposure over a dosing interval
unchanged.

```{r food-effect, fig.cap = "Replicates the Wu 2025 Section 2.2 food effect: fed dosing stretches the zero-order absorption window from 0.281 h to 0.846 h."}
food_cohort <- tibble::tibble(
  id = 1:2, WT = 58.55, AGE = 31, CRCL = 106.88, FED = c(0, 1)
)
ev_food <- build_events(regimens[regimens$regimen == "100 mg q24h", ], food_cohort)
food <- rxode2::rxSolve(mod_typ, events = ev_food,
                        keep = c("FED"), returnType = "data.frame")

food_sum <- food |>
  dplyr::filter(time >= ss_start) |>
  dplyr::group_by(FED) |>
  dplyr::summarise(
    d1     = unique(round(d1, 4)),
    cmax   = max(Cc),
    tmax   = time[which.max(Cc)] - ss_start,
    auc24  = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  dplyr::mutate(state = ifelse(FED == 1, "fed", "fasting"), .before = 1) |>
  dplyr::select(-FED)

knitr::kable(food_sum, digits = 4,
             caption = "Fed vs fasting at steady state (100 mg q24h, reference subject).")

stopifnot(
  abs(food_sum$d1[food_sum$state == "fasting"] - 0.281) < 1e-3,
  abs(food_sum$d1[food_sum$state == "fed"]     - 0.846) < 1e-3,
  # Dose is conserved: AUC over the interval is unaffected by food.
  abs(diff(food_sum$auc24)) / mean(food_sum$auc24) < 0.005,
  # Cmax is reduced and Tmax delayed in the fed state.
  food_sum$cmax[food_sum$state == "fed"] < food_sum$cmax[food_sum$state == "fasting"],
  food_sum$tmax[food_sum$state == "fed"] > food_sum$tmax[food_sum$state == "fasting"]
)
```

```{r food-plot, echo = FALSE, fig.cap = "Fed vs fasting absorption phase after the day-7 dose."}
food |>
  dplyr::filter(time >= ss_start, time <= ss_start + 6) |>
  dplyr::mutate(state = ifelse(FED == 1, "fed", "fasting")) |>
  ggplot(aes(time - ss_start, Cc, colour = state)) +
  geom_line(linewidth = 0.6) +
  labs(x = "Time after day-7 dose (h)", y = "Concentration (mg/L)", colour = NULL) +
  theme_bw() + theme(legend.position = "top")
```

# NCA validation with PKNCA

Wu 2025 reports no NCA table, so there is no published Cmax/Tmax/AUC to compare
against. PKNCA is still the right instrument for the quantity the paper's PK/PD
analysis is built on -- the steady-state 24-hour AUC -- so it is computed here
over the final 24 h and then used as the falsifier below.

```{r pknca}
conc_df <- sim |>
  dplyr::filter(time >= ss_start, !is.na(Cc)) |>
  dplyr::select(id, time, Cc, regimen)

dose_df <- do.call(rbind, lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  tidyr::crossing(
    tibble::tibble(id = cohort$id),
    time = seq(ss_start, t_end - r$tau, by = r$tau)
  ) |>
    dplyr::mutate(amt = r$dose, regimen = r$regimen)
})) |> as.data.frame()

conc_obj <- PKNCA::PKNCAconc(
  data = as.data.frame(conc_df),
  formula = Cc ~ time | regimen + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  data = dose_df,
  formula = amt ~ time | regimen + id,
  doseu = "mg"
)

intervals <- data.frame(
  start = ss_start, end = t_end,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result)

auc24 <- nca_tbl |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(regimen, id, auc24_nca = PPORRES)

nca_summary <- nca_tbl |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
  dplyr::group_by(regimen, PPTESTCD) |>
  dplyr::summarise(
    median = median(PPORRES),
    p05    = quantile(PPORRES, 0.05),
    p95    = quantile(PPORRES, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD), .before = 1) |>
  dplyr::select(-PPTESTCD) |>
  dplyr::rename("Regimen" = regimen, "Median" = median,
                "5th pctile" = p05, "95th pctile" = p95)

knitr::kable(nca_summary, digits = 3,
             caption = "Simulated steady-state NCA over the final 24 h (200 subjects per regimen). Wu 2025 publishes no NCA table for comparison; these values feed the PK/PD analysis below. AUC in mg*h/L, concentrations in mg/L, times in h.")
```

## Falsifier: AUC24h,ss must equal daily dose / CL, per subject

For a linear model at steady state, the exposure over 24 h is exactly the daily
dose divided by the individual apparent clearance -- independent of how the daily
dose is split across the day, of absorption, and of food. This is a **per-subject
identity**, which is a far stronger check than comparing cohort medians.

```{r auc-identity}
subj_cl <- sim |>
  dplyr::group_by(regimen, id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

ident <- auc24 |>
  dplyr::left_join(subj_cl, by = c("regimen", "id")) |>
  dplyr::left_join(dplyr::select(regimens, regimen, daily_dose), by = "regimen") |>
  dplyr::mutate(
    auc24_theory = daily_dose / cl,
    rel_err      = (auc24_nca - auc24_theory) / auc24_theory
  )

ident |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    n              = dplyr::n(),
    max_abs_relerr = max(abs(rel_err)),
    .groups = "drop"
  ) |>
  dplyr::rename("Regimen" = regimen, "N subjects" = n,
                "Max |relative error|" = max_abs_relerr) |>
  knitr::kable(digits = 5,
               caption = "Per-subject agreement between the PKNCA steady-state AUC and daily dose / CL.")

# The residual gap is trapezoidal-integration error on a 0.1 h grid, not model error.
stopifnot(nrow(ident) == n_subj * nrow(regimens), max(abs(ident$rel_err)) < 0.01)
```

## Falsifier: equal daily doses give identical exposure

Wu 2025 reports near-identical PTAs for 50 mg q12h and 100 mg q24h (for example
99.9% and 99.9% against *S. pneumoniae* at the animal target). That is a
consequence of both regimens delivering 100 mg/day: the 24-hour AUC must match
subject by subject, and the 200 mg/day regimen must give exactly twice as much.

```{r dose-proportionality}
wide <- ident |>
  dplyr::select(regimen, id, auc24_nca) |>
  tidyr::pivot_wider(names_from = regimen, values_from = auc24_nca)

ratios <- tibble::tibble(
  Comparison = c("100 mg q24h / 50 mg q12h (equal daily dose)",
                 "100 mg q12h / 50 mg q12h (double daily dose)"),
  `Median ratio` = c(
    median(wide$`100 mg q24h` / wide$`50 mg q12h`),
    median(wide$`100 mg q12h` / wide$`50 mg q12h`)
  ),
  `Max deviation` = c(
    max(abs(wide$`100 mg q24h` / wide$`50 mg q12h` - 1)),
    max(abs(wide$`100 mg q12h` / wide$`50 mg q12h` - 2))
  )
)
knitr::kable(ratios, digits = 5,
             caption = "Per-subject steady-state AUC ratios between regimens.")

stopifnot(
  max(abs(wide$`100 mg q24h` / wide$`50 mg q12h` - 1)) < 0.01,
  max(abs(wide$`100 mg q12h` / wide$`50 mg q12h` - 2)) < 0.02
)
```

## Falsifier: the covariate relationships are recovered

Regressing each log parameter on its log covariates should return the published
exponents. The estimates are *not* noise-free: the IIV terms (`omega = 0.23` on
CL, `0.25` on V2) act as residual error in these regressions, so each recovered
exponent carries a genuine sampling standard error. With 200 subjects that error
is about 0.04 for the CRCL and AGE exponents and about 0.11 for the WT exponent
-- body weight has the narrowest spread of the three covariates
(`sd(log(WT))` around 0.18 versus 0.48 for age), so its exponent is the least
precisely identified.

A fixed absolute tolerance would therefore be a seed lottery rather than a test.
The check below is calibrated to the regression's own uncertainty instead: each
published exponent must fall inside the fitted 95% confidence interval, and must
be within 3 standard errors of the estimate. A second assertion bounds the
standard errors themselves, so the test retains the power to fail -- without it,
a degenerate cohort with huge standard errors would pass the coverage check
vacuously.

```{r covariate-recovery}
par_by_subj <- sim |>
  dplyr::filter(regimen == "100 mg q24h") |>
  dplyr::group_by(id) |>
  dplyr::summarise(
    cl = dplyr::first(cl), vc = dplyr::first(vc),
    CRCL = dplyr::first(CRCL), WT = dplyr::first(WT), AGE = dplyr::first(AGE),
    .groups = "drop"
  )

# CL has one covariate, so regressing log CL on log CRCL recovers the exponent
# up to the eta noise. Vc has two, so both exponents come from one regression.
fit_cl <- stats::lm(log(cl) ~ log(CRCL), data = par_by_subj)
fit_vc <- stats::lm(log(vc) ~ log(WT) + log(AGE), data = par_by_subj)

# (fit, coefficient index) for each published relationship.
fits <- list(list(fit_cl, 2L), list(fit_vc, 2L), list(fit_vc, 3L))

recovery <- tibble::tibble(
  Relationship = c("CRCL on CL/F", "WT on V2/F", "AGE on V2/F"),
  Published    = c(0.460, 0.966, 0.286),
  Recovered    = vapply(fits, function(f) unname(coef(f[[1]])[f[[2]]]), numeric(1)),
  StdError     = vapply(fits, function(f) summary(f[[1]])$coefficients[f[[2]], 2], numeric(1)),
  CIlow        = vapply(fits, function(f) confint(f[[1]])[f[[2]], 1], numeric(1)),
  CIhigh       = vapply(fits, function(f) confint(f[[1]])[f[[2]], 2], numeric(1))
) |>
  dplyr::mutate(
    ZfromPublished = abs(Recovered - Published) / StdError,
    PublishedInCI  = Published >= CIlow & Published <= CIhigh
  )

recovery |>
  dplyr::rename(
    "Published" = Published, "Recovered" = Recovered, "SE" = StdError,
    "95% CI lower" = CIlow, "95% CI upper" = CIhigh,
    "|est - pub| / SE" = ZfromPublished, "Published in CI" = PublishedInCI
  ) |>
  knitr::kable(digits = 4,
               caption = "Covariate exponents recovered from the simulated cohort vs Wu 2025 Table 2, with the sampling uncertainty of each regression coefficient.")

stopifnot(
  # Every published exponent is covered by its own 95% confidence interval, and
  # sits within 3 SE of the point estimate.
  all(recovery$PublishedInCI),
  all(recovery$ZfromPublished < 3),
  # Power guard: the standard errors must stay small enough that a materially
  # wrong exponent would be rejected above rather than absorbed into a wide CI.
  all(recovery$StdError < 0.15)
)
```

# PK/PD cut-offs (Wu 2025 Tables 3 and 4)

This is the paper's headline result and the strongest available answer key: 12
PK/PD cut-off cells and 12 PTA values across three regimens and four
pathogen-specific targets.

The PK/PD index is the **unbound** `fAUC24h/MIC`. Wu 2025 never reports a
sitafloxacin unbound fraction. The value used here, `fu = 0.388`, is **not from
Wu 2025**: it is carried from the separately packaged
`Rodjun_2023_sitafloxacin` model, where Rodjun 2023 sources it to Tanigawara
2013. It is applied only in this section, never inside the model file.

```{r fu-provenance}
fu <- 0.388   # NOT from Wu 2025. Rodjun 2023 Methods, sourced there to Tanigawara 2013.
rodjun <- readModelDb("Rodjun_2023_sitafloxacin")()
grep("0.388", rodjun$description, value = TRUE) |> substr(1, 200)
```

Because `AUC24h,ss = daily dose / CL` holds exactly per subject (verified above),
the probability of target attainment can be evaluated on a large sample of
individual clearances without re-solving the ODE system. That buys the precision
Wu 2025's Monte Carlo had -- their PTAs are quoted to 0.1%, which 200 simulated
subjects cannot resolve -- at negligible cost.

```{r pta-cohort}
n_mc <- 10000L

set.seed(4620)
mc <- tibble::tibble(
  CRCL = draw_crcl(n_mc),
  eta  = stats::rnorm(n_mc, 0, sqrt(0.0515))
) |>
  dplyr::mutate(cl = 14.7 * (CRCL / 106.88)^0.460 * exp(eta))

# Sanity: the closed form must agree with the ODE cohort's clearance distribution.
c(ode_median_cl = median(par_by_subj$cl), mc_median_cl = median(mc$cl))
stopifnot(abs(median(mc$cl) / median(par_by_subj$cl) - 1) < 0.10)

pta <- function(daily_dose, mic, target) {
  mean(fu * daily_dose / mc$cl / mic >= target)
}

mic_grid <- c(0.008, 0.015, 0.03, 0.06, 0.125, 0.25, 0.5, 1, 2, 4)

targets <- tibble::tribble(
  ~scenario,                                        ~target,
  "S. pneumoniae, animal model (fAUC24h/MIC 11.56)",  11.56,
  "S. pneumoniae, clinical (fAUC24h/MIC >= 30)",      30.00,
  "S. aureus, animal model (fAUC24h/MIC 18.32)",      18.32,
  "Gram-negatives, animal model (fAUC24h/MIC 15.06)", 15.06
)

pta_grid <- tidyr::crossing(targets, regimens, mic = mic_grid) |>
  dplyr::mutate(pta = mapply(pta, daily_dose, mic, target) * 100)
```

```{r pta-figure, fig.height = 6, fig.cap = "Replicates Wu 2025 Figure 3: probability of target attainment versus MIC for the three dosing regimens, by PK/PD target. Dashed line marks the 95% PTA criterion."}
ggplot(pta_grid, aes(mic, pta, colour = regimen)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1) +
  geom_hline(yintercept = 95, linetype = "dashed") +
  scale_x_log10(breaks = mic_grid, labels = mic_grid) +
  facet_wrap(~scenario, ncol = 2) +
  labs(x = "MIC (mg/L)", y = "PTA (%)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Comparison against the published cut-offs

Wu 2025 Section 4.2 defines the cut-off as "the highest value of the
corresponding MIC range" at which the PTA exceeds 95%.

```{r cutoff-table}
simulated_cutoffs <- pta_grid |>
  dplyr::filter(pta >= 95) |>
  dplyr::group_by(scenario, regimen) |>
  dplyr::summarise(
    cutoff_sim = max(mic),
    pta_at_cutoff_sim = pta[which.max(mic)],
    .groups = "drop"
  )

# Wu 2025 Table 3, transcribed. Every cell of the published table is represented.
published_cutoffs <- tibble::tribble(
  ~scenario,                                          ~regimen,       ~cutoff_pub, ~pta_pub,
  "S. pneumoniae, animal model (fAUC24h/MIC 11.56)",  "50 mg q12h",    0.06,        99.9,
  "S. pneumoniae, animal model (fAUC24h/MIC 11.56)",  "100 mg q24h",   0.06,        99.9,
  "S. pneumoniae, animal model (fAUC24h/MIC 11.56)",  "100 mg q12h",   0.125,       99.7,
  "S. pneumoniae, clinical (fAUC24h/MIC >= 30)",      "50 mg q12h",    0.03,        99.3,
  "S. pneumoniae, clinical (fAUC24h/MIC >= 30)",      "100 mg q24h",   0.03,        99.1,
  "S. pneumoniae, clinical (fAUC24h/MIC >= 30)",      "100 mg q12h",   0.06,        99.1,
  "S. aureus, animal model (fAUC24h/MIC 18.32)",      "50 mg q12h",    0.06,        97.7,
  "S. aureus, animal model (fAUC24h/MIC 18.32)",      "100 mg q24h",   0.06,        97.9,
  "S. aureus, animal model (fAUC24h/MIC 18.32)",      "100 mg q12h",   0.125,       97.0,
  "Gram-negatives, animal model (fAUC24h/MIC 15.06)", "50 mg q12h",    0.06,        99.3,
  "Gram-negatives, animal model (fAUC24h/MIC 15.06)", "100 mg q24h",   0.06,        99.1,
  "Gram-negatives, animal model (fAUC24h/MIC 15.06)", "100 mg q12h",   0.125,       98.8
)

cutoff_cmp <- published_cutoffs |>
  dplyr::left_join(simulated_cutoffs, by = c("scenario", "regimen")) |>
  dplyr::mutate(
    dilutions_off = round(log2(cutoff_sim / cutoff_pub)),
    agree = ifelse(dilutions_off == 0, "yes", paste0(dilutions_off, " dilution"))
  )

# Guard against a silently empty join: every published cell must have matched.
stopifnot(nrow(cutoff_cmp) == 12L, !anyNA(cutoff_cmp$cutoff_sim))

cutoff_cmp |>
  dplyr::select(scenario, regimen, cutoff_pub, cutoff_sim,
                pta_pub, pta_at_cutoff_sim, agree) |>
  dplyr::rename(
    "Scenario" = scenario, "Regimen" = regimen,
    "Cut-off, Wu 2025 (mg/L)" = cutoff_pub,
    "Cut-off, simulated (mg/L)" = cutoff_sim,
    "PTA at cut-off, Wu 2025 (%)" = pta_pub,
    "PTA at cut-off, simulated (%)" = pta_at_cutoff_sim,
    "Agreement" = agree
  ) |>
  knitr::kable(digits = 1,
               caption = "Simulated vs published PK/PD cut-offs and PTAs (Wu 2025 Table 3).")

n_exact <- sum(cutoff_cmp$dilutions_off == 0)
cat(sprintf("%d of 12 cells reproduce the published cut-off exactly; %d of 12 within one doubling dilution.\n",
            n_exact, sum(abs(cutoff_cmp$dilutions_off) <= 1)))
```

The published cut-offs also have an internal structure worth testing directly.
MIC values are reported on a doubling-dilution series whose printed labels are
conventionally rounded (`0.008, 0.015, 0.03, 0.06, 0.125, 0.25, ...`), so
"one dilution higher" is **not** a numeric ratio of exactly 2 -- the step from
0.06 to 0.125 is a factor of 2.083. The relationships between regimens are
therefore tested as positions in the dilution series rather than as ratios.

```{r cutoff-assert}
# Position of a printed MIC in the doubling-dilution series.
dil_index <- function(mic) match(mic, mic_grid)

step_check <- cutoff_cmp |>
  dplyr::select(scenario, regimen, cutoff_pub) |>
  tidyr::pivot_wider(names_from = regimen, values_from = cutoff_pub) |>
  dplyr::mutate(
    step_double_daily_dose = dil_index(`100 mg q12h`) - dil_index(`50 mg q12h`),
    step_equal_daily_dose  = dil_index(`100 mg q24h`) - dil_index(`50 mg q12h`)
  )

step_check |>
  dplyr::select(scenario, step_double_daily_dose, step_equal_daily_dose) |>
  dplyr::rename(
    "Scenario" = scenario,
    "Dilutions: 100 q12h vs 50 q12h (2x daily dose)" = step_double_daily_dose,
    "Dilutions: 100 q24h vs 50 q12h (same daily dose)" = step_equal_daily_dose
  ) |>
  knitr::kable(caption = "Internal structure of the Wu 2025 Table 3 cut-offs, in doubling dilutions.")

stopifnot(
  # A doubling-dilution series is the resolution of the published answer key, so
  # agreement to within one dilution across every cell is the meaningful bar.
  all(abs(cutoff_cmp$dilutions_off) <= 1),
  nrow(step_check) == 4L,
  # Doubling the daily dose moves the published cut-off exactly one dilution up.
  all(step_check$step_double_daily_dose == 1L),
  # Equal daily doses split differently give exactly the same cut-off, which is
  # the same identity the per-subject AUC check established above.
  all(step_check$step_equal_daily_dose == 0L)
)
```

# Assumptions and deviations

## Errata and open items

* **`eta4` (inter-occasion variability) is not encoded.** Wu 2025 Table 2 reports
  a fourth random effect with `omega^2 = 0.0279` (RSE 8.7%, shrinkage 29.1%,
  bootstrap 95% CI 0.0184-0.0394). The abbreviation footnote describes it only as
  "eta4, interindividual variability of IOV" and never names the structural
  parameter it acts on; the Section 2.2 display equations account for `eta1`,
  `eta2` and `eta3` only. No supplement or control stream is available
  (the EuropePMC `supplementaryFiles` endpoint returns HTTP 500 for PMC11868765
  and the Frontiers file endpoints return 404), and the supplement's documented
  contents -- a study-design table and mean concentration-time figures -- would
  not place an eta. Assigning the term to CL, D1 or F would be inventing model
  structure, so it is omitted and its reported value preserved in the model file
  comments. Simulated between-subject spread in AUC is therefore marginally
  narrower than the paper's, which if anything makes the PTAs above slightly
  optimistic at the extremes.
* **The unbound fraction is not from Wu 2025.** `fu = 0.388` used in the PK/PD
  section is carried from `Rodjun_2023_sitafloxacin` (Rodjun 2023 Methods,
  sourced there to Tanigawara 2013). Wu 2025 computes `fAUC24h/MIC` throughout
  but never states the unbound fraction it used. The model file deliberately does
  not carry an `fu` term.
* **CFR is not reproduced.** Wu 2025 Table 3 also reports cumulative fraction of
  response, which requires the EUCAST MIC frequency distributions for each
  pathogen. Those distributions are external to the paper and are not on disk, so
  only the PTA and cut-off columns are validated here.
* **No NCA answer key.** Wu 2025 publishes no Cmax/Tmax/AUC table, so the PKNCA
  results above are reported rather than compared, and the validation weight
  rests on the per-subject `AUC24h = daily dose / CL` identity and the PK/PD
  cut-off table.
* **`eta3` on D1 is extreme as published** (`omega^2 = 3.46`, i.e. `omega = 1.86`
  on the log scale). It is encoded exactly as reported. Because a zero-order
  input conserves the delivered dose, even large `eta3` draws leave AUC -- and
  therefore every PK/PD conclusion -- untouched, perturbing only Cmax and Tmax.

## Assumptions made because the paper does not say

* **Covariate distributions.** Wu 2025 Table 1 reports only marginal mean +/- SD.
  Age, weight and creatinine clearance are drawn here as independent log-normals
  moment-matched to those marginals. Real cohorts correlate age, weight and renal
  function; ignoring that correlation is the largest single approximation in the
  PTA reproduction.
* **Renal-function floor.** Creatinine clearance is truncated at 50 mL/min per
  Wu 2025 Section 4.2. The paper does not state the shape of the CRCL
  distribution it simulated from.
* **Reference values.** The normalizing constants 106.88 mL/min, 58.55 kg and 31
  years are printed inside the Section 2.2 equations but never labelled. They are
  close to, yet distinct from, the Table 1 means, and are most likely medians.
  They are used exactly as printed.
* **Fed state in the PK/PD simulations.** Wu 2025 does not state whether the
  simulated subjects were fed or fasting. The cohort is simulated fasting
  (`FED = 0`). This has no effect on any conclusion: the food effect acts only on
  the zero-order absorption duration, which the food-effect section above shows
  leaves the 24-hour AUC unchanged.
* **Cut-off criterion.** Wu 2025 says the PTA must be "greater than 95%"; a
  `>= 95%` rule is used here. No cell in the table sits close enough to 95% for
  the distinction to matter.
```
