---
title: "Meropenem (Wu 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meropenem (Wu 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wu_2022_meropenem"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wu_2022_meropenem"))$description`
* Article: [Children (Basel) 2022;9(12):1998](https://doi.org/10.3390/children9121998)

# Population

Wu 2022 pooled 110 opportunistic (scavenged) meropenem plasma
concentrations from 78 Chinese neonates and young infants with bacterial
infection, treated at the Children's Hospital of Hebei Province and
Beijing Children's Hospital (Wu 2022 Section 2.1, Section 3.1). Inclusion
required a postmenstrual age (PMA) below 48 weeks. Table 1 reports a
median (range) postnatal age (PNA) of 12 days (1-113), gestational age
(GA) of 36.5 weeks (26.3-42.1), PMA of 38.2 weeks (27.4-46.1), birth
weight of 2500 g (480-4100) and current weight (CW) of 2680 g
(980-5310). Doses were 9.3-40.4 mg/kg (median 19.7 mg/kg) per
administration. Forty-three infants had GA <= 37 weeks and 35 had GA > 37
weeks (Section 2.6). Sex distribution is not reported. The same
information is available programmatically via
`rxode2::rxode(readModelDb("Wu_2022_meropenem"))$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment, first-order elimination | n/a | Section 3.2 paragraph 1 |
| `lvc` (theta1) | 1.63 L | Table 2, row 'theta1' under V |
| `lcl` (theta2) | 0.503 L/h | Table 2, row 'theta2' under CL |
| `e_wt_vc` | 1 (fixed) | Section 3.2 paragraph 2 ('allometric coefficients of 1 for V'); Table 2 V equation |
| `e_wt_cl` | 0.75 (fixed) | Section 3.2 paragraph 2; Table 2 CL equation |
| `e_pna_cl` (theta3) | 0.209 | Table 2, row 'theta3' (F_age) |
| `e_ga_cl` (theta4) | 2.14 | Table 2, row 'theta4' (F_age) |
| `etalvc` | 0.03017 = log(0.175^2 + 1) | Table 2, IIV V 17.5% |
| `etalcl` | 0.22957 = log(0.508^2 + 1) | Table 2, IIV CL 50.8% |
| `propSd` | 0.288 | Table 2, residual variability 28.8%; Section 3.2 (proportional) |
| V = theta1 * (CW/2680) | n/a | Table 2 |
| CL = theta2 * (CW/2680)^0.75 * F_age | n/a | Table 2 |
| F_age = (PNA/12)^theta3 * (GA/36.5)^theta4 | n/a | Table 2 |
| Reference values 2680 g, 36.5 weeks, 12 days | n/a | Table 2 footnote (cohort medians) |
| Free fraction 0.98, target 70% fT > MIC, MIC 8 mg/L | n/a | Section 2.6 |

# Virtual cohort

The individual covariates are not published, so the cohort is generated
from the Table 1 summaries, one cohort per GA subgroup used in the
paper's dosing simulations (GA <= 37 weeks and GA > 37 weeks). GA is
drawn from a normal distribution centred on the median (36.5 weeks, SD
4.65 weeks) truncated to the subgroup's share of the 26.3-42.1 week
range; PNA from a log-normal with the Table 1 median (12 days) and a
log-scale SD chosen to reproduce the Table 1 mean (17.6 days), truncated
to 1-113 days. Current weight is linked to PMA by a linear relation
built from the Table 1 means and SDs of CW and PMA with an assumed
correlation of 0.85, truncated to the observed 980-5310 g range.

```{r cohort}
set.seed(20221219)

n_per_group <- 200L

rtrunc_norm <- function(n, mean, sd, lower, upper) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- rnorm(4L * n, mean, sd)
    out <- c(out, x[x >= lower & x <= upper])
  }
  out[seq_len(n)]
}

# PNA: log-normal with median 12 days; mean = median * exp(s^2 / 2) = 17.6
pna_sdlog <- sqrt(2 * log(17.6 / 12))

make_group <- function(group, ga_lower, ga_upper) {
  ga <- rtrunc_norm(n_per_group, 36.5, 4.65, ga_lower, ga_upper)
  pna_d <- rtrunc_norm(n_per_group, log(12), pna_sdlog, log(1), log(113)) |> exp()
  pma <- ga + pna_d / 7
  r <- 0.85
  cw_g <- 2661 + r * 927 / 4.40 * (pma - 37.5) +
    rnorm(n_per_group, 0, 927 * sqrt(1 - r^2))
  cw_g <- pmin(pmax(cw_g, 980), 5310)
  tibble(
    subj = seq_len(n_per_group),
    group = group,
    GA = ga,
    PNA = pna_d / 30.4375, # canonical PNA is in months
    PNA_days = pna_d,
    WT = cw_g / 1000
  )
}

cohort <- bind_rows(
  make_group("GA <= 37 weeks", 26.3, 37),
  make_group("GA > 37 weeks", 37.0001, 42.1)
)

cohort |>
  group_by(group) |>
  summarise(
    N = n(),
    `GA median (weeks)` = median(GA),
    `PNA median (days)` = median(PNA_days),
    `CW median (kg)` = median(WT),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "Virtual cohort covariates by GA subgroup.")
```

# Typical-value check against the reported weight-normalised parameters

Section 3.2 reports median (range) individual weight-normalised V of
0.60 (0.51-0.69) L/kg and CL of 0.16 (0.04-0.51) L/h/kg. At the
reference subject (CW 2.68 kg, PNA 12 days, GA 36.5 weeks) the model
returns exactly theta1 and theta2.

```{r typical}
mod <- readModelDb("Wu_2022_meropenem")

ref_ev <- data.frame(
  id = 1L, time = c(0, 1), evid = c(1L, 0L), amt = c(100, 0),
  cmt = "central", WT = 2.68, GA = 36.5, PNA = 12 / 30.4375
)
ref <- rxode2::rxSolve(rxode2::zeroRe(mod), ref_ev) |> as.data.frame()
stopifnot(
  abs(ref$vc[1] / 1.63 - 1) < 1e-6,
  abs(ref$cl[1] / 0.503 - 1) < 1e-6
)
c(V_per_kg = ref$vc[1] / 2.68, CL_per_kg = ref$cl[1] / 2.68)
```

The typical-subject V/kg (0.61 L/kg) matches the reported median of
0.60 L/kg. The typical-subject CL/kg (0.19 L/h/kg) sits above the
reported median individual estimate of 0.16 L/h/kg; with 50.8% IIV on CL
and the skew of the PNA distribution towards the first two weeks of life
the median individual estimate is expected to lie below the reference
value.

# Simulation

Meropenem is linear in this model, so concentrations scale with dose.
The dosing grid of Figure 4 (20-60 mg/kg; twice or three times daily;
1-5 h infusions) is therefore simulated once per frequency and infusion
duration at 1 mg/kg and rescaled for each dose level. Six doses are
given, and the PTA is evaluated over the sixth dosing interval.

```{r simulate}
regimens <- expand_grid(tau = c(12, 8), tinf = 1:5) |>
  mutate(reg = row_number())

n_dose <- 6L
dt <- 0.05

build_events <- function(cohort, regimens) {
  design <- cross_join(cohort, regimens) |>
    mutate(id = row_number())
  doses <- design |>
    cross_join(tibble(dose_n = seq_len(n_dose))) |>
    mutate(
      time = (dose_n - 1) * tau,
      evid = 1L,
      amt = 1 * WT, # 1 mg/kg
      rate = amt / tinf,
      cmt = "central"
    )
  obs <- design |>
    cross_join(tibble(k = 0:round(max(regimens$tau) / dt))) |>
    filter(k * dt <= tau + 1e-9) |>
    mutate(
      time = (n_dose - 1) * tau + k * dt,
      evid = 0L, amt = 0, rate = 0, cmt = "central"
    )
  bind_rows(doses, obs) |>
    select(id, time, evid, amt, rate, cmt, WT, GA, PNA, group, tau, tinf, reg) |>
    arrange(id, time, desc(evid))
}

events <- build_events(cohort, regimens)

rxode2::rxSetSeed(20221219)
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("group", "tau", "tinf", "reg")
) |>
  as.data.frame()
stopifnot(!anyNA(sim$Cc))
```

# Replicate Figure 4: probability of target attainment

Target attainment is free concentration (98% of total, Section 2.6)
above an MIC of 8 mg/L for at least 70% of the dosing interval.

```{r pta}
fu <- 0.98
mic <- 8
dose_levels <- c(20, 30, 40, 50, 60)

# Fraction of the final interval each subject spends above the MIC, per
# dose level (the unit-dose profile scaled by the dose in mg/kg).
ft <- sim |>
  select(id, time, Cc, group, tau, tinf) |>
  cross_join(tibble(dose = dose_levels)) |>
  group_by(id, group, tau, tinf, dose) |>
  summarise(ft_mic = mean(fu * Cc * dose > mic), .groups = "drop")

pta <- ft |>
  group_by(group, tau, tinf, dose) |>
  summarise(PTA = 100 * mean(ft_mic >= 0.7), .groups = "drop") |>
  mutate(freq = ifelse(tau == 12, "BID (q12h)", "TID (q8h)"))
```

```{r pta-figure, fig.width = 8, fig.height = 6}
ggplot(pta, aes(tinf, PTA, colour = factor(dose))) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 70, linetype = "dashed") +
  facet_grid(freq ~ group) +
  labs(
    x = "Infusion duration (h)",
    y = "PTA, 70% fT > MIC of 8 mg/L (%)",
    colour = "Dose (mg/kg)",
    caption = "Replicates Figure 4 of Wu 2022 with a virtual cohort."
  )
```

Section 3.4 quotes the PTA of the regimens it recommends. The table
compares them with the virtual-cohort values.

```{r pta-compare}
published <- tribble(
  ~group,           ~tau, ~tinf, ~dose, ~PTA_published,
  "GA <= 37 weeks",   12,     3,    60,          72.49,
  "GA <= 37 weeks",   12,     4,    50,          73.54,
  "GA <= 37 weeks",   12,     5,    40,          74.09,
  "GA <= 37 weeks",    8,     1,    30,          71.71,
  "GA > 37 weeks",     8,     2,    50,          72.50,
  "GA > 37 weeks",     8,     3,    40,          75.08
)

pta_cmp <- published |>
  inner_join(pta, by = c("group", "tau", "tinf", "dose")) |>
  mutate(diff = PTA - PTA_published)
stopifnot(nrow(pta_cmp) == nrow(published))

pta_cmp |>
  mutate(Regimen = sprintf("%g mg/kg q%gh, %g-h infusion", dose, tau, tinf)) |>
  select(group, Regimen, PTA_published, PTA, diff) |>
  rename(
    "GA subgroup" = group,
    "PTA, Wu 2022 (%)" = PTA_published,
    "PTA, simulated (%)" = PTA,
    "Difference (points)" = diff
  ) |>
  knitr::kable(digits = 1, caption = "PTA of the Section 3.4 regimens: published versus virtual cohort.")
```

```{r pta-gate}
# The published PTAs come from the authors' own 78 covariate sets; the
# virtual cohort only approximates their joint distribution, and each
# simulated PTA carries a binomial SE of about 3 points at n = 200, so
# the gate is on the centre of the six differences, not their extreme.
# Across four seeds the median difference ranged 0.0 to 3.0 points and the
# median absolute difference 3.1 to 4.3 points. A mis-transcribed CL, V or
# exponent moves these PTAs by tens of points.
stopifnot(
  abs(median(pta_cmp$diff)) < 6,
  median(abs(pta_cmp$diff)) < 8
)
```

The virtual-cohort PTAs for the six regimens Section 3.4 recommends
fall within a few percentage points of the published values, about the
size of the binomial Monte Carlo error at 200 subjects per subgroup.

# PKNCA validation

Steady-state NCA over the sixth dosing interval for the two regimens the
paper recommends (30 mg/kg q8h as a 1-h infusion for GA <= 37 weeks,
40 mg/kg q8h as a 3-h infusion for GA > 37 weeks), solved at the actual
doses.

```{r pknca-sim}
rec <- tribble(
  ~group,           ~dose, ~tau, ~tinf, ~treatment,
  "GA <= 37 weeks",    30,    8,     1, "30 mg/kg q8h 1-h (GA <= 37)",
  "GA > 37 weeks",     40,    8,     3, "40 mg/kg q8h 3-h (GA > 37)"
)

design_rec <- cohort |>
  inner_join(rec, by = "group") |>
  mutate(id = row_number())

dose_rec <- design_rec |>
  cross_join(tibble(dose_n = seq_len(n_dose))) |>
  mutate(
    time = (dose_n - 1) * tau, evid = 1L, amt = dose * WT,
    rate = amt / tinf, cmt = "central"
  )
obs_rec <- design_rec |>
  cross_join(tibble(time = seq(0, n_dose * 8, by = 0.1))) |>
  mutate(evid = 0L, amt = 0, rate = 0, cmt = "central")
events_rec <- bind_rows(dose_rec, obs_rec) |>
  select(id, time, evid, amt, rate, cmt, WT, GA, PNA, treatment, tau) |>
  arrange(id, time, desc(evid))

sim_rec <- rxode2::rxSolve(mod, events = events_rec, keep = c("treatment", "tau")) |>
  as.data.frame()
stopifnot(!anyNA(sim_rec$Cc))
```

```{r pknca}
t_start <- (n_dose - 1) * 8
t_end <- n_dose * 8

conc_df <- sim_rec |>
  filter(!is.na(Cc), time >= t_start, time <= t_end) |>
  mutate(Cc = pmax(Cc, 0)) |>
  select(id, time, Cc, treatment)

dose_df <- events_rec |>
  filter(evid == 1, time == t_start) |>
  select(id, time, amt, rate, treatment) |>
  mutate(duration = amt / rate)

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg", duration = "duration")
intervals <- data.frame(
  start = t_start, end = t_end,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
knitr::kable(summary(nca_res),
             caption = "Simulated steady-state NCA (sixth dosing interval) for the two recommended regimens.")
```

At steady state, AUC over one dosing interval equals dose / CL for each
subject. The check below compares PKNCA's AUC with that identity using
each subject's simulated clearance.

```{r auc-identity}
cl_df <- sim_rec |>
  group_by(id) |>
  summarise(cl = first(cl), .groups = "drop")

auc_chk <- as.data.frame(nca_res$result) |>
  filter(PPTESTCD == "auclast") |>
  select(id, treatment, auclast = PPORRES) |>
  inner_join(cl_df, by = "id") |>
  inner_join(dose_df |> select(id, amt), by = "id") |>
  mutate(pct_diff = 100 * (auclast / (amt / cl) - 1))

auc_chk |>
  group_by(treatment) |>
  summarise(
    `Median % difference` = median(pct_diff),
    `90th percentile of abs % difference` = quantile(abs(pct_diff), 0.9),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2, caption = "PKNCA AUC0-tau versus dose / CL.")

# After six doses only subjects with the slowest CL are short of steady
# state, and the 0.1-h trapezoid grid adds a small bias. Realised median
# about -0.01% and 90th percentile below 0.2%; a dose, unit or CL error
# shifts every subject by far more than these bounds.
stopifnot(
  abs(median(auc_chk$pct_diff)) < 0.5,
  quantile(abs(auc_chk$pct_diff), 0.9) < 2
)
```

The paper reports no NCA parameters, so no published-versus-simulated
NCA table is shown.

# Assumptions and deviations

- **PNA units.** The Table 2 footnote describes PNA as 'postnatal age in
  weeks' but, in the same footnote, gives 12 days as the median PNA used
  for centring, and Table 1 reports PNA in days (median 12.0). The model
  therefore takes PNA in days inside the F_age term. The package's
  canonical `PNA` column is in months; `model()` converts with
  PNA * 30.4375.
- **Weight units.** The paper uses current weight in grams centred on
  2680 g. The packaged model takes `WT` in kg centred on 2.68 kg, which
  is numerically identical.
- **IIV conversion.** Table 2 reports IIV as a percentage for an
  exponential model; it is converted with omega^2 = log(CV^2 + 1).
  Residual variability (28.8%) is encoded as a proportional SD of 0.288.
- **Virtual cohort.** The paper's dosing simulations reuse its own 78
  covariate sets, which are not published. The virtual cohort here is
  generated from the Table 1 summaries with an assumed CW-PMA
  correlation of 0.85, and GA drawn from a normal distribution centred
  on the median (the published GA distribution is left-skewed, mean 35.0
  versus median 36.5 weeks). The simulated PTAs are therefore an
  approximation of Figure 4, not a replication of the authors' cohort.
- **PTA evaluation window.** Section 2.6 does not state which dosing
  interval the PTA was evaluated on. The sixth interval (steady state
  for nearly every subject) is used here, without residual error.
- **External validation not reproduced.** The 16-infant external
  validation dataset is not published.
