---
title: "Fedratinib (Wu 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fedratinib (Wu 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
mod <- readModelDb("Wu_2020_fedratinib")
ui <- rxode2::rxode(mod)
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1007/s00280-020-04131-y>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7515950/>
* Electronic supplementary material (ESM 1, Supplemental Materials SM1-SM12):
  distributed with the open-access article.

Wu 2020 built a physiologically based pharmacokinetic (PBPK) drug-drug
interaction model of the JAK2 inhibitor fedratinib in Simcyp V17R1 and used it
to support the fedratinib label (dose reduction to 200 mg QD with a strong
CYP3A4 inhibitor). The Simcyp model uses the platform's minimal-PBPK
distribution model with a single adjusting compartment, enzyme-kinetic hepatic
clearance scaled from recombinant CYP intrinsic clearances, and mechanism-based
inhibition and induction of CYP3A4 by fedratinib itself.

To seed the minimal-PBPK distribution inputs, the authors first fitted a
conventional **two-compartment model with first-order absorption** to mean
plasma profiles from four healthy-volunteer studies of a single 500 mg dose
(Supplemental Material SM3). That compartmental model is what this package
provides.

**Why the PBPK model itself is not included.** The Simcyp model's clearance,
auto-inhibition and auto-induction terms act on hepatic and intestinal CYP
enzyme pools whose abundances, microsomal protein per gram of liver, liver
weight, enzyme degradation rates and intestinal physiology come from the Simcyp
"Healthy Volunteers" and "Cancer" population library files, which the authors
used "without further modification" and which are not published in the paper
or its supplement. The compound-specific inputs (SM4) are printed in full, but
without the platform's system parameters they cannot be turned into a
standalone ODE system. A population PK model of fedratinib in patients with
myelofibrosis, polycythemia vera and essential thrombocythemia is available as
`Ogasawara_2019_fedratinib`.

# Population

```{r population, echo = FALSE, results = "asis"}
cat(ui$population$notes, "\n")
```

The mean profiles came from studies TDU12620 (single ascending doses in healthy
men), BDR12462 (500 mg tablet vs capsule bioequivalence), FED12258 and ALI13451
(food effect on a single 500 mg dose in healthy men); the study list is SM1
Table 1 and the pooling is described in SM3. Neither the number of subjects
nor the demographics of the pooled set are reported.

# Source trace

| Model element | Value | Source location |
|---|---|---|
| Structure: 2-compartment, first-order absorption, additive error | -- | SM3 text |
| `lka` (ka) | 0.221 1/h | SM3 compartmental PK parameter table |
| `lcl` (CL/F) | 27.2 L/h | SM3 compartmental PK parameter table |
| `lvc` (V/F) | 107 L | SM3 compartmental PK parameter table; restated in SM2 "Distribution Parameters" |
| `lq` (CL2/F) | 30.8 L/h | SM3 compartmental PK parameter table; restated in SM2 |
| `lvp` (V2/F) | 797 L | SM3 compartmental PK parameter table; restated in SM2 |
| `addSd` | fixed 0 | SM3 names an additive error model; no magnitude printed |
| IIV | none | SM3: Phoenix NLME "naive pooled" algorithm |
| `Cc = 1000 * central / vc` | ng/mL | unit conversion (mg / L to ng/mL, the units of the paper's exposure tables) |
| Derived minimal-PBPK inputs Vss, Vsac, Q | 4.1 L/kg, 3.5 L/kg, 11 L/h | SM2 Eqs (1)-(3) with F = 0.36 and 81 kg; SM2 Table 1 and SM4 |
| Observed single-dose exposures | see comparison table | SM6 Table (healthy subjects, 300 and 500 mg single dose); main-text Table 1 (INT12893, 300 mg alone) |

# Internal consistency with the published PBPK inputs

SM2 converts the compartmental estimates into the Simcyp minimal-PBPK inputs
with Eqs (1)-(3): `Vss = (V/F + V2/F) * F`, `Vsac = V2/F * F` and
`Q = CL2/F * F`, using `F = 0.36` and a body weight of 81 kg for volumes per
kg. Recomputing the three inputs from this model's `ini()` values reproduces
the printed values, which checks the transcription of `V/F`, `V2/F` and
`CL2/F` against a second, independent place in the supplement.

```{r pbpk-inputs}
p <- ui$theta
f_sm2 <- 0.36
wt_sm2 <- 81
recomputed <- c(
  Vss_L_per_kg = (exp(p[["lvc"]]) + exp(p[["lvp"]])) * f_sm2 / wt_sm2,
  Vsac_L_per_kg = exp(p[["lvp"]]) * f_sm2 / wt_sm2,
  Q_L_per_h = exp(p[["lq"]]) * f_sm2
)
printed <- c(Vss_L_per_kg = 4.1, Vsac_L_per_kg = 3.5, Q_L_per_h = 11)

tibble::tibble(
  Input = names(printed),
  Printed = printed,
  Recomputed = signif(recomputed, 3),
  `Difference (%)` = round(100 * (recomputed / printed - 1), 1)
) |>
  knitr::kable(caption = "Simcyp minimal-PBPK inputs recomputed from the SM3 estimates (SM2 Eqs 1-3).")

# The printed inputs are rounded to two significant figures ('~4.1', '~3.5',
# '~11'), so agreement to within that rounding is the expectation; a
# transcription slip in V/F, V2/F or CL2/F would miss by far more.
stopifnot(all(abs(recomputed / printed - 1) < 0.03))
```

# Simulation

The model has no between-subject variability, so a single deterministic solve
per dose gives the typical profile. Single oral doses of 300 mg and 500 mg are
simulated, matching the healthy-volunteer single-dose scenarios the paper
reports.

```{r simulate}
doses <- c(300, 500)
obs_times <- sort(unique(c(seq(0, 24, by = 0.1), seq(24.5, 96, by = 0.5), seq(98, 720, by = 2))))

ev <- dplyr::bind_rows(
  tibble::tibble(id = seq_along(doses), time = 0, amt = doses, evid = 1L, cmt = "depot"),
  tidyr::expand_grid(id = seq_along(doses), time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
) |>
  dplyr::mutate(treatment = paste0(doses[id], " mg")) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim <- rxode2::rxSolve(mod, events = ev, keep = "treatment") |>
  as.data.frame()

stopifnot(nrow(sim) == length(doses) * length(obs_times))
```

```{r profile-plot, fig.cap = "Typical fedratinib plasma concentration after a single oral dose (compare with the healthy-volunteer single-dose panels a and b of Wu 2020 Figure 2, which show the Simcyp predictions and the observed mean data)."}
ggplot(dplyr::filter(sim, time <= 168), aes(time, Cc, colour = treatment)) +
  geom_line() +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Fedratinib (ng/mL)", colour = "Single dose") +
  theme_bw()
```

# NCA with PKNCA

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time = 0 row per subject (pre-dose Cc = 0 for an oral dose).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- ev |>
  dplyr::filter(evid == 1L) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::select(treatment, cmax, tmax, aucinf.obs, half.life) |>
  dplyr::rename(
    "Single dose" = treatment,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUC0-inf (ng*h/mL)" = aucinf.obs,
    "t1/2 (h)" = half.life
  ) |>
  knitr::kable(digits = 1, caption = "PKNCA results for the typical profile.")
```

The model is linear with no variability, so `AUC0-inf = Dose / (CL/F)` must
hold exactly, up to the trapezoidal error of the sampling grid. This is a
check of the solve against its own closed form, so a tight bound is
appropriate.

```{r auc-identity}
auc_identity <- nca_wide |>
  dplyr::mutate(
    dose = as.numeric(sub(" mg", "", treatment)),
    expected = 1000 * dose / exp(p[["lcl"]]),
    pct_diff = 100 * (aucinf.obs / expected - 1)
  )
auc_identity |>
  dplyr::select(treatment, aucinf.obs, expected, pct_diff) |>
  knitr::kable(digits = 2, caption = "AUC0-inf from PKNCA vs Dose / (CL/F).")
stopifnot(all(abs(auc_identity$pct_diff) < 1))
```

# Comparison against the published observed exposures

Wu 2020 reports observed single-dose exposures in healthy subjects in SM6
(300 mg and 500 mg, arithmetic mean with geometric mean in brackets) and, for
300 mg, in main-text Table 1 (the fedratinib-alone arm of the ketoconazole
study INT12893). A typical-value profile is closest to a geometric mean, so
the geometric means are used as the reference.

```{r nca-compare}
published <- tibble::tibble(
  treatment = c("300 mg", "500 mg"),
  cmax = c(497, 683), # SM6 table, healthy subjects single dose, GM
  aucinf.obs = c(6530, 13200) # SM6 table, healthy subjects single dose, GM
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "treatment",
  params = c("cmax", "aucinf.obs"),
  units = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "Typical-value simulation vs observed geometric means in healthy subjects",
    "(Wu 2020 SM6). Rows marked with * differ by more than 20%."
  )
)
```

The typical profile over-predicts the observed geometric means: AUC0-inf by
about 69% at 300 mg and 39% at 500 mg, and Cmax by about 26% and 53%. The
AUC gap is a property of the published estimate, not of the implementation:
the AUC identity above shows the model returns exactly `Dose / (CL/F)` with
`CL/F = 27.2 L/h`, while the observed geometric means imply a `CL/F` of about
46 L/h (300 mg) and 38 L/h (500 mg). The supplement is consistent with this.
SM2 ("Metabolism Parameters") reports that the compartmental `CL/F` was only
the starting value for Simcyp's retrograde clearance calculation and was then
raised to 48 L/h so that the PBPK model reproduced the observed dose-normalized
AUC of 30 ng*h/(mL*mg), which corresponds to a `CL/F` of 33.3 L/h. The
compartmental model was fitted to mean profiles pooled across four studies
(including food-effect and formulation arms), and the paper does not say which
subjects underlie the SM6 comparison rows. The parameters are reproduced as
published and have not been adjusted.

```{r nca-compare-check}
# Pin the size and sign of the discrepancy described above. The model is
# deterministic, so these ratios do not vary between runs or machines.
ratio <- nca_wide |>
  dplyr::inner_join(published, by = "treatment", suffix = c("_sim", "_obs")) |>
  dplyr::mutate(
    auc_ratio = aucinf.obs_sim / aucinf.obs_obs,
    cmax_ratio = cmax_sim / cmax_obs
  )
stopifnot(
  all(ratio$auc_ratio > 1.3 & ratio$auc_ratio < 1.8),
  all(ratio$cmax_ratio > 1.2 & ratio$cmax_ratio < 1.6)
)
```

# Assumptions and deviations

* **PBPK layer not implemented.** Only the SM3 compartmental model is
  provided; the Simcyp minimal-PBPK DDI model depends on unpublished Simcyp
  population-library system parameters (see "Model and source").
* **Residual error.** SM3 states an additive error model but gives no
  magnitude; `addSd` is fixed at 0 so the model is a typical-value model.
  Simulations from it carry no residual noise.
* **No between-subject variability.** The fit used the naive-pooled algorithm
  on mean profiles, which estimates none.
* **Population.** The number, sex and body size of the subjects behind the
  mean profiles are not reported. Three of the four studies enrolled healthy
  men only.
* **Formulation and food.** Two of the pooled studies were food-effect studies
  and one compared tablet with capsule. SM3 does not say which arms were
  included in the mean profiles; the model carries no food or formulation
  effect.
* **Parameter table label.** In the supplement the compartmental parameter
  table is cited as "Table 3 in Supplemental Material SM 3" in SM2; the values
  are also restated in the SM2 text (V/F 107 L, V2/F 797 L, CL2/F 30.8 L/h),
  which agree with the table.
