---
title: "Fentanyl in preterm and term newborns (Wu 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fentanyl in preterm and term newborns (Wu 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Wu_2022_fentanyl"))$reference`
* Description: `r rxode2::rxode(readModelDb("Wu_2022_fentanyl"))$description`
* Article: [Clin Pharmacokinet 2022;61(3):401-412](https://doi.org/10.1007/s40262-021-01076-0)
  (open access, PMC8891207). The electronic supplementary material (ESM)
  carries the final NONMEM control stream.

Wu and colleagues pooled 673 fentanyl plasma concentrations from 164 newborns
across the whole neonatal range -- gestational age 24 to 42 weeks -- and fitted
a two-compartment model in NONMEM 7.4. Clearance depends on two separate
maturation processes: prenatal maturation, captured by birth weight, and
postnatal maturation, captured by postnatal age. Combining the two was better
than postmenstrual age alone (dOFV = -91). Central volume scales with current
body weight through a *bodyweight-dependent exponent* (BDE). The weight
exponent is itself a power function of weight, so volume changes fastest in
the smallest neonates. The authors use the model to show that a flat
1 ug/kg/h infusion gives very different exposure across birth weight and
postnatal age, and they propose a birth-weight and postnatal-age banded
infusion regimen (their Table 3).

# Population

The analysis data set (Wu 2022 Table 1) combines two studies.

* **Dataset 1 (Helsinki).** 66 mechanically ventilated newborns at the Hospital
  for Children and Adolescents, University of Helsinki (Saarenmaa 2000). The
  planned regimen was 10.5 ug/kg over 1 h, then a continuous infusion (median
  1.5 ug/kg/h for a median 58 h). Arterial samples were taken at 2, 12, 24, 48
  and 60 h. The assay was a radioimmunoassay with an LLOQ of 1 ug/L. Median
  gestational age 31.25 weeks (25.30-42.30), birth weight 1632 g (760-4245),
  postnatal age at the start of treatment 0.46 days. Current weight was not
  recorded and was set equal to birth weight.
* **Dataset 2 (DINO).** 98 preterm infants in four Dutch NICUs (the DINO
  study, NCT02421068). Physicians chose boluses of 0.5-3 ug/kg (median 2.1 ug/kg
  over 3 min) and/or infusions of 0.5-3 ug/kg/h. Samples were scavenged during
  routine care and assayed by LC-MS/MS with an LLOQ of 0.3 ug/L. Median
  gestational age 27.10 weeks (23.90-31.90), birth weight 906 g (390-1905),
  postnatal age 4.5 days (0-68).

Combined: 61.6% male; median gestational age 28.95 weeks; median birth weight
1055 g; median current weight 1165 g; median postnatal age 1.1 days. The two
medians, 1055 g and 1165 g, are the centring values of the clearance and
volume covariate models. The same information is in
`readModelDb("Wu_2022_fentanyl")` under `population`.

# Source trace

| Model element | Value | Source |
|---|---|---|
| Structure: two-compartment, IV input, first-order elimination | ADVAN3 TRANS4 | Results 3.1; ESM control stream `$SUBROUTINES` |
| `lcl` (TVCL) | 0.31 L/h | Table 2 |
| `e_wt_birth_cl` (theta BW) | 1.47 | Table 2 (Eq. 2 in the text prints 1.57 -- see below) |
| `e_pna_cl` (theta PNA) | 0.505 | Table 2 (Eq. 2 prints 0.502) |
| CL = TVCL (BW/1055)^thetaBW (PNA+0.01)^thetaPNA | days, g | Eq. 2; Table 2 header row; ESM `$PK` |
| `lvc` (TVV1) | 10.6 L | Table 2 |
| `e_wt_vc` (BDE intercept L1) | 1.56 | Table 2 |
| `bde_m_vc` (BDE slope M) | -0.417 | Table 2 |
| V1 = TVV1 (CW/1165)^BDE, BDE = L1 (CW/1165)^M | g | Eq. 3; Table 2; ESM `$PK` |
| `lq` (Q) | 0.573 L/h | Table 2 |
| `lvp` (V2) | 3.37 L | Table 2 |
| `etalcl` | 0.444^2 = 0.197 | Table 2 'On CL (%)' 44.4% (see omega-scale section) |
| `etalvc` | 0.456^2 = 0.208 | Table 2 'On V1 (%)' 45.6% |
| `addSd_helsinki`, `propSd_helsinki` | 0.246 ug/L, 0.23 | Table 2, 'on dataset 1' |
| `addSd_dino`, `propSd_dino` | 0.0297 ug/L, 0.361 | Table 2, 'on dataset 2' |
| Residual form W = sqrt(ADD^2 + (PRO*IPRED)^2), `$SIGMA 1 FIX` | combined, SD scale | ESM `$ERROR` |
| Dataset indicator `STUDY_DINO` = ASY (0 = dataset 1, 1 = dataset 2) | | ESM `$INPUT` |

The covariates are carried in canonical units. Birth weight (`WT_BIRTH`) and
current weight (`WT`) are in kg, so the model divides by 1.055 and 1.165.
Postnatal age (`PNA`) is in months, and the model converts it to days
(`PNA * 30.4375`) before adding the source's 0.01-day offset.

```{r load-model}
mod <- readModelDb("Wu_2022_fentanyl")
mod_typ <- rxode2::zeroRe(mod())
th <- as.list(mod()$theta)
```

## Printed Eq. 2 exponents vs Table 2

In the article body, Eq. 2 prints the clearance exponents as 1.57 (birth
weight) and 0.502 (postnatal age). Table 2 prints 1.47 and 0.505. The Results
and Abstract quote five fold-changes in clearance. All five follow from the
Table 2 values, and the birth-weight ratios rule out 1.57. The ESM `$THETA`
initial for the birth-weight exponent is also 1.47. The model uses Table 2.

```{r eq2-check}
fold <- function(e_bw, e_pna) {
  c(
    "BW 2000 vs 1000 g" = 2^e_bw,
    "BW 3000 vs 1000 g" = 3^e_bw,
    "PNA 7 vs 1 day" = ((7 + 0.01) / (1 + 0.01))^e_pna,
    "PNA 14 vs 1 day" = ((14 + 0.01) / (1 + 0.01))^e_pna,
    "PNA 21 vs 1 day" = ((21 + 0.01) / (1 + 0.01))^e_pna
  )
}
published <- c(2.77, 5.03, 2.66, 3.77, 4.63) # Results 3.1
eq2_tab <- data.frame(
  comparison = names(fold(1, 1)),
  published = published,
  table2 = unname(fold(th$e_wt_birth_cl, th$e_pna_cl)),
  eq2 = unname(fold(1.57, 0.502))
)
eq2_tab |>
  dplyr::rename(
    "Clearance ratio" = comparison,
    "Published (Results 3.1)" = published,
    "Table 2 exponents (1.47, 0.505)" = table2,
    "Printed Eq. 2 exponents (1.57, 0.502)" = eq2
  ) |>
  knitr::kable(digits = 2)
# Deterministic: the Table 2 exponents reproduce every published ratio to the
# printed rounding; the Eq. 2 birth-weight exponent misses by 7-12%.
stopifnot(
  all(abs(eq2_tab$table2 / eq2_tab$published - 1) < 0.005),
  abs(eq2_tab$eq2[1] / eq2_tab$published[1] - 1) > 0.05
)
```

## IIV scale

Table 2 gives the IIV as percentages (44.4% on CL, 45.6% on V1) without saying
whether they are `100*omega` or the exact log-normal CV
`100*sqrt(exp(omega^2) - 1)`. The ESM control stream helps, because its
`$OMEGA` initial values (0.2 and 0.207) are, like its `$THETA` initial values,
close to a previous run's finals. Read as `100*omega`, the initial values
land within 1% of the printed finals. The exact-CV reading puts them 5-6% away.
The same group (Voller 2019, midazolam) used the `100*omega` convention. The
model therefore encodes `omega^2 = (CV/100)^2`.

```{r omega-scale}
omega_init <- c(CL = 0.2, V1 = 0.207)
omega_tab <- data.frame(
  parameter = names(omega_init),
  printed = c(44.4, 45.6),
  sd_reading = 100 * sqrt(omega_init),
  cv_reading = 100 * sqrt(exp(omega_init) - 1),
  row.names = NULL
)
omega_tab |>
  dplyr::rename(
    "Parameter" = parameter,
    "Printed final (%)" = printed,
    "Initial as 100*omega (%)" = sd_reading,
    "Initial as exact CV (%)" = cv_reading
  ) |>
  knitr::kable(digits = 1)
om <- diag(mod()$omega)
stopifnot(
  abs(sqrt(om[["etalcl"]]) - 0.444) < 1e-6,
  abs(sqrt(om[["etalvc"]]) - 0.456) < 1e-6
)
```

The two readings differ by less than 0.02 in omega^2 at this magnitude
(0.197 vs 0.180 on CL), so the choice has little effect on the simulated
prediction intervals.

# Closed-form checks against the paper's own derived numbers

The Discussion gives three quantities that follow directly from the volume
model: the BDE is 1.66 at a current weight of 1000 g and 1.05 at 3000 g, and
the median weight-normalised volume of distribution is 12 L/kg. The typical
V1 + V2 at the median weight of 1.165 kg reproduces the 12 L/kg.

```{r closed-form}
bde <- function(wt_kg) th$e_wt_vc * (wt_kg / 1.165)^th$bde_m_vc
v1 <- function(wt_kg) exp(th$lvc) * (wt_kg / 1.165)^bde(wt_kg)
cf <- data.frame(
  quantity = c("BDE at 1000 g", "BDE at 3000 g", "(V1 + V2)/CW at 1165 g (L/kg)"),
  published = c(1.66, 1.05, 12),
  model = c(bde(1.0), bde(3.0), (v1(1.165) + exp(th$lvp)) / 1.165)
)
cf |>
  dplyr::rename("Quantity" = quantity, "Published (Discussion)" = published, "Model" = model) |>
  knitr::kable(digits = 3)
stopifnot(all(abs(cf$model / cf$published - 1) < 0.01))
```

# Virtual neonates

Postnatal age advances with the simulation clock and drives clearance, so the
helper below writes an observation record every 0.5 h, each carrying the
current postnatal age in months. That keeps rxode2 from stepping over the
change in the covariate. The authors projected current weight with the
published growth curves of Anchieta et al. (their reference 31). Wu 2022 does
not print those curves numerically, so current weight is
held at birth weight. This is exact for the Helsinki cohort (the authors made
the same assumption) and a close approximation during the first days of life.
It is weaker for a neonate who starts treatment at 21 days, when current weight
is above birth weight.

```{r helpers}
make_subject <- function(id, bw, pna0, doses, tend, dt = 0.5, group = "") {
  obs <- data.frame(
    id = id, time = seq(0, tend, by = dt), evid = 0L, amt = 0, rate = 0,
    cmt = "central"
  )
  dose <- data.frame(
    id = id, time = doses$time, evid = 1L, amt = doses$amt,
    rate = doses$rate, cmt = "central"
  )
  d <- dplyr::bind_rows(dose, obs) |> dplyr::arrange(time, dplyr::desc(evid))
  d$WT_BIRTH <- bw
  d$WT <- bw
  d$PNA <- (pna0 + d$time / 24) / 30.4375
  d$STUDY_DINO <- 1L
  d$group <- group
  d$bw <- bw
  d$pna0 <- pna0
  d
}
# Continuous infusion of `ugkgh` ug/kg/h for `dur` h
infusion <- function(bw, ugkgh, dur) {
  data.frame(time = 0, amt = ugkgh * bw * dur, rate = ugkgh * bw)
}
# Boluses of `ugkg` ug/kg given over 3 min
boluses <- function(bw, ugkg, times) {
  data.frame(time = times, amt = ugkg * bw, rate = ugkg * bw / 0.05)
}
```

# Figure 1A: clearance vs postnatal age

Replicates Figure 1A of Wu 2022, the population clearance for birth weights of
850, 1500, 2500 and 3000 g over the first 30 days of life.

```{r fig1}
cl_typ <- function(bw_kg, pna_d) {
  exp(th$lcl) * (bw_kg / 1.055)^th$e_wt_birth_cl * (pna_d + 0.01)^th$e_pna_cl
}
fig1 <- expand.grid(pna = seq(0, 30, by = 0.25), bw = c(0.85, 1.5, 2.5, 3.0)) |>
  dplyr::mutate(cl = cl_typ(bw, pna), BW = factor(paste(bw * 1000, "g"), levels = paste(c(850, 1500, 2500, 3000), "g")))
ggplot(fig1, aes(pna, cl, colour = BW)) +
  geom_line() +
  labs(x = "Postnatal age (days)", y = "Clearance (L/h)", colour = "Birth weight") +
  theme_bw()
```

# Figure 3: 1 ug/kg/h continuous infusion for 7 days

Replicates Figure 3 of Wu 2022. The Results quote three medians after 48 h of
infusion: 2.1 ug/L at birth weight 850 g and postnatal age 0.5 days at the
start, 1.29 ug/L at 3000 g and 0.5 days, and 0.68 ug/L at 850 g and 21 days.
The deterministic comparison below uses the typical-value (`zeroRe`)
prediction. The stochastic medians in the figure come from 100 virtual
neonates per arm.

```{r fig3-typical}
arms3 <- expand.grid(bw = c(0.85, 1.5, 2.5, 3.0), pna0 = c(0.5, 7, 21))
ev3 <- dplyr::bind_rows(lapply(seq_len(nrow(arms3)), function(i) {
  make_subject(
    i, arms3$bw[i], arms3$pna0[i], infusion(arms3$bw[i], 1, 168), 168,
    group = paste0("BW ", arms3$bw[i] * 1000, " g, PNA ", arms3$pna0[i], " d")
  )
}))
typ3 <- rxode2::rxSolve(mod_typ, ev3, keep = c("group", "bw", "pna0"), returnType = "data.frame")
c48 <- function(bw, pna0) {
  v <- typ3$Cc[typ3$bw == bw & typ3$pna0 == pna0 & typ3$time == 48]
  if (length(v) != 1L) stop("no unique 48 h row for BW ", bw, " PNA ", pna0)
  v
}
fig3_chk <- data.frame(
  case = c("BW 850 g, PNA 0.5 d", "BW 3000 g, PNA 0.5 d", "BW 850 g, PNA 21 d"),
  published = c(2.1, 1.29, 0.68),
  model = c(c48(0.85, 0.5), c48(3.0, 0.5), c48(0.85, 21))
) |>
  dplyr::mutate(pct_diff = 100 * (model / published - 1))
fig3_chk |>
  dplyr::rename(
    "Neonate (start of infusion)" = case, "Published median at 48 h (ug/L)" = published,
    "Model typical value at 48 h (ug/L)" = model, "% diff" = pct_diff
  ) |>
  knitr::kable(digits = 2)
# Deterministic typical-value solve. The two early-start cases sit within 8%.
# The 21-day case reads 13% high because current weight is held at birth
# weight: at 21 days the real neonate is heavier, so V1 is larger and the
# 48-h concentration is lower.
stopifnot(
  abs(fig3_chk$pct_diff[1]) < 5,
  abs(fig3_chk$pct_diff[2]) < 10,
  abs(fig3_chk$pct_diff[3]) < 20
)
```

```{r fig3-stochastic}
rxode2::rxSetSeed(20220301)
n_per_arm <- 100
ev3_iiv <- dplyr::bind_rows(lapply(seq_len(nrow(arms3)), function(i) {
  dplyr::bind_rows(lapply(seq_len(n_per_arm), function(j) {
    make_subject(
      (i - 1) * n_per_arm + j, arms3$bw[i], arms3$pna0[i],
      infusion(arms3$bw[i], 1, 168), 168, dt = 2,
      group = paste0("PNA ", arms3$pna0[i], " d at start")
    )
  }))
}))
sim3 <- rxode2::rxSolve(mod(), ev3_iiv, keep = c("group", "bw", "pna0"), returnType = "data.frame")
med3 <- sim3 |>
  dplyr::group_by(group, bw, time) |>
  dplyr::summarise(med = median(Cc), .groups = "drop") |>
  dplyr::mutate(BW = factor(paste(bw * 1000, "g"), levels = paste(c(850, 1500, 2500, 3000), "g")))
ggplot(med3, aes(time, med, colour = BW)) +
  geom_line() +
  facet_wrap(~group) +
  labs(
    x = "Time since start of infusion (h)", y = "Median fentanyl concentration (ug/L)",
    colour = "Birth weight",
    caption = "Replicates Figure 3 of Wu 2022 (current weight held at birth weight)."
  ) +
  theme_bw()
```

As in the paper, concentrations rise for 1-2 days before they peak. They fall
as birth weight rises and, within a birth weight, as postnatal age at the start
of treatment rises. The later decline the paper shows comes partly from
current weight increasing over the week, which this simulation leaves out.

# Figure 5: intermittent boluses with and without a loading dose

Replicates Figure 5 of Wu 2022, typical-value profiles at postnatal age 0.5
days. (A) 2 ug/kg every 4 h for 2 days. (B) A 5 ug/kg loading dose over 3 min,
then 2 ug/kg every 4 h. The Results state that the loading dose reaches
"around 0.75-1.2 ug/L" immediately at 850 g and "around 0.5 ug/L" in larger
neonates.

```{r fig5}
q4h <- seq(0, 44, by = 4)
arms5 <- expand.grid(bw = c(0.85, 1.5, 2.5, 3.0), regimen = c("A: 2 ug/kg q4h", "B: 5 ug/kg load + 2 ug/kg q4h"), stringsAsFactors = FALSE)
ev5 <- dplyr::bind_rows(lapply(seq_len(nrow(arms5)), function(i) {
  bw <- arms5$bw[i]
  d <- if (startsWith(arms5$regimen[i], "A")) {
    boluses(bw, 2, q4h)
  } else {
    dplyr::bind_rows(boluses(bw, 5, 0), boluses(bw, 2, q4h[-1]))
  }
  make_subject(i, bw, 0.5, d, 48, dt = 0.05, group = arms5$regimen[i])
}))
typ5 <- rxode2::rxSolve(mod_typ, ev5, keep = c("group", "bw"), returnType = "data.frame") |>
  dplyr::mutate(BW = factor(paste(bw * 1000, "g"), levels = paste(c(850, 1500, 2500, 3000), "g")))
ggplot(typ5, aes(time, Cc, colour = BW)) +
  geom_line() +
  facet_wrap(~group) +
  labs(x = "Time (h)", y = "Typical fentanyl concentration (ug/L)", colour = "Birth weight") +
  theme_bw()
load_peak <- typ5 |>
  dplyr::filter(startsWith(group, "B"), time <= 4) |>
  dplyr::group_by(BW) |>
  dplyr::summarise(peak_after_load = max(Cc), .groups = "drop")
knitr::kable(load_peak |> dplyr::rename("Birth weight" = BW, "Peak after loading dose (ug/L)" = peak_after_load), digits = 2)
stopifnot(nrow(load_peak) == 4L)
```

The model gives about 0.7 ug/L at 850 g and about 0.5 ug/L at 1500-3000 g. That
matches the paper's "around 0.5 ug/L" for larger neonates and falls just below
the 0.75 ug/L lower end quoted for 850 g. This is a typical-value prediction at
postnatal age 0.5 days. The paper's statement covers several postnatal ages
and is read from a figure of medians.

# PKNCA: the proposed regimen (Table 3) vs the published AUC envelope

Wu 2022 Table 3 proposes 0.5/0.7/0.9 ug/kg/h below 1500 g and 0.7/0.9/1.2
ug/kg/h at or above 1500 g, for postnatal ages of 1-2, 3-6 and >= 7 days at the
start of treatment. For this regimen the authors report (Results 3.2.1 and ESM
Fig. S5) daily AUCs of 9-16 ug*h/L on day 1, 12-26 on day 4 and 11-24 on day 7,
and concentrations of 0.6-1.2 ug/L after 24 h. The simulation below gives each
typical neonate a constant rate chosen from its postnatal age at the start. It
excludes birth weight >= 2500 g with postnatal age > 7 days, as the paper does.
PKNCA computes the three daily AUCs per neonate.

```{r table3-sim}
rate_t3 <- function(bw, pna0) {
  band <- 1L + (pna0 >= 3) + (pna0 >= 7)
  if (bw < 1.5) c(0.5, 0.7, 0.9)[band] else c(0.7, 0.9, 1.2)[band]
}
arms_t3 <- expand.grid(bw = c(0.85, 1.5, 2.5, 3.0), pna0 = c(1, 3, 7))
arms_t3$rate <- mapply(rate_t3, arms_t3$bw, arms_t3$pna0)
arms_t3$treatment <- paste0("BW ", arms_t3$bw * 1000, " g, PNA ", arms_t3$pna0, " d, ", arms_t3$rate, " ug/kg/h")
ev_t3 <- dplyr::bind_rows(lapply(seq_len(nrow(arms_t3)), function(i) {
  make_subject(i, arms_t3$bw[i], arms_t3$pna0[i], infusion(arms_t3$bw[i], arms_t3$rate[i], 168), 168,
    dt = 0.25, group = arms_t3$treatment[i]
  )
}))
typ_t3 <- rxode2::rxSolve(mod_typ, ev_t3, keep = c("group"), returnType = "data.frame") |>
  dplyr::rename(treatment = group)

conc_df <- typ_t3 |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)
dose_df <- ev_t3 |>
  dplyr::filter(evid == 1) |>
  dplyr::transmute(id, treatment = group, time, amt)
o_conc <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id)
o_dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)
intervals <- data.frame(
  start = c(0, 72, 144), end = c(24, 96, 168),
  auclast = TRUE
)
nca_t3 <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))
auc_t3 <- as.data.frame(nca_t3$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::mutate(day = c(`0` = "Day 1", `72` = "Day 4", `144` = "Day 7")[as.character(start)])
stopifnot(nrow(auc_t3) == 3L * nrow(arms_t3))

c24_t3 <- typ_t3 |> dplyr::filter(time == 24)
stopifnot(nrow(c24_t3) == nrow(arms_t3))

envelope <- data.frame(
  quantity = c("Day 1", "Day 4", "Day 7", "Concentration at 24 h (ug/L)"),
  published_low = c(9, 12, 11, 0.6),
  published_high = c(16, 26, 24, 1.2)
)
sim_range <- auc_t3 |>
  dplyr::group_by(quantity = day) |>
  dplyr::summarise(sim_low = min(PPORRES), sim_high = max(PPORRES), .groups = "drop") |>
  dplyr::bind_rows(data.frame(
    quantity = "Concentration at 24 h (ug/L)",
    sim_low = min(c24_t3$Cc), sim_high = max(c24_t3$Cc)
  ))
env_tab <- dplyr::left_join(envelope, sim_range, by = "quantity") |>
  dplyr::mutate(quantity = ifelse(startsWith(quantity, "Day"), paste(quantity, "AUC (ug*h/L)"), quantity))
env_tab |>
  dplyr::rename(
    "Quantity" = quantity, "Published low" = published_low, "Published high" = published_high,
    "Simulated low" = sim_low, "Simulated high" = sim_high
  ) |>
  knitr::kable(digits = 2)
```

```{r table3-per-arm}
auc_t3 |>
  dplyr::select(treatment, day, PPORRES) |>
  tidyr::pivot_wider(names_from = day, values_from = PPORRES) |>
  dplyr::left_join(c24_t3 |> dplyr::select(treatment, Cc), by = "treatment") |>
  dplyr::rename(
    "Typical neonate and Table 3 rate" = treatment,
    "Day 1 AUC (ug*h/L)" = `Day 1`, "Day 4 AUC (ug*h/L)" = `Day 4`,
    "Day 7 AUC (ug*h/L)" = `Day 7`, "C at 24 h (ug/L)" = Cc
  ) |>
  knitr::kable(digits = 2)
```

```{r table3-gate}
# Deterministic typical-value solve. Day 4, day 7 and the 24 h concentration
# fall inside the published envelopes. The day-1 maximum (16.8 ug*h/L at
# 1500 g, PNA 7 d, 1.2 ug/kg/h) is 5% above the published 16. The gate allows
# 10% on the envelope edges, which still fails on a mis-transcribed clearance,
# volume or dose unit, since any of those moves the whole range by tens of
# percent.
rng <- function(q) unlist(env_tab[env_tab$quantity == q, c("sim_low", "sim_high")])
stopifnot(
  all(rng("Concentration at 24 h (ug/L)") >= 0.6 & rng("Concentration at 24 h (ug/L)") <= 1.2),
  all(rng("Day 4 AUC (ug*h/L)") >= 12 & rng("Day 4 AUC (ug*h/L)") <= 26),
  all(rng("Day 7 AUC (ug*h/L)") >= 11 & rng("Day 7 AUC (ug*h/L)") <= 24),
  all(rng("Day 1 AUC (ug*h/L)") >= 9 * 0.9 & rng("Day 1 AUC (ug*h/L)") <= 16 * 1.1)
)
```

# PKNCA mass balance: AUC(0-inf) x CL = dose

With postnatal age held constant (so clearance is time-invariant), a single
2 ug/kg bolus must give `AUCinf * CL = dose` for every typical neonate. This
checks the ODE system and the dose, volume and clearance units.

```{r mass-balance}
arms_mb <- expand.grid(bw = c(0.5, 0.85, 1.5, 3.0), pna0 = c(1, 7))
ev_mb <- dplyr::bind_rows(lapply(seq_len(nrow(arms_mb)), function(i) {
  bw <- arms_mb$bw[i]
  obs_t <- sort(unique(c(seq(0, 1, by = 0.05), seq(1, 24, by = 0.5), seq(24, 1500, by = 4))))
  d <- dplyr::bind_rows(
    data.frame(id = i, time = 0, evid = 1L, amt = 2 * bw, rate = 2 * bw / 0.05, cmt = "central"),
    data.frame(id = i, time = obs_t, evid = 0L, amt = 0, rate = 0, cmt = "central")
  ) |> dplyr::arrange(time, dplyr::desc(evid))
  d$WT_BIRTH <- bw
  d$WT <- bw
  d$PNA <- arms_mb$pna0[i] / 30.4375 # held constant
  d$STUDY_DINO <- 1L
  d$treatment <- paste0("BW ", bw * 1000, " g, PNA ", arms_mb$pna0[i], " d")
  d
}))
typ_mb <- rxode2::rxSolve(mod_typ, ev_mb, keep = "treatment", returnType = "data.frame")
nca_mb <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_mb |> dplyr::filter(!is.na(Cc)) |> dplyr::select(id, treatment, time, Cc), Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(ev_mb |> dplyr::filter(evid == 1) |> dplyr::select(id, treatment, time, amt), amt ~ time | treatment + id),
  intervals = data.frame(start = 0, end = Inf, aucinf.obs = TRUE)
))
mb <- as.data.frame(nca_mb$result) |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::left_join(typ_mb |> dplyr::distinct(treatment, cl), by = "treatment") |>
  dplyr::left_join(ev_mb |> dplyr::filter(evid == 1) |> dplyr::distinct(treatment, amt), by = "treatment") |>
  dplyr::mutate(ratio = PPORRES * cl / amt)
mb |>
  dplyr::select(treatment, PPORRES, cl, amt, ratio) |>
  dplyr::rename(
    "Typical neonate" = treatment, "AUCinf (ug*h/L)" = PPORRES,
    "CL (L/h)" = cl, "Dose (ug)" = amt, "AUCinf x CL / dose" = ratio
  ) |>
  knitr::kable(digits = 3)
stopifnot(nrow(mb) == nrow(arms_mb), all(abs(mb$ratio - 1) < 0.01))
```

# Assumptions and deviations

* **Eq. 2 exponents.** The article's Eq. 2 prints 1.57 and 0.502. Table 2
  prints 1.47 and 0.505. Table 2 is used because it reproduces all five
  published clearance fold-changes, and the ESM initial value of the
  birth-weight exponent is also 1.47 (see "Printed Eq. 2 exponents vs
  Table 2").
* **IIV scale.** The printed IIV percentages are read as `100*omega` (see "IIV
  scale"). The exact-CV reading would give omega^2 of 0.180 and 0.189 instead
  of 0.197 and 0.208.
* **Proportional residual error.** Table 2 labels the proportional rows "(%)"
  but prints 0.23 and 0.361. With `$SIGMA 1 FIX` and
  `W = SQRT(ADD**2 + (PRO*IPRED)**2)` in the ESM `$ERROR` block, these are
  fractional SDs (23% and 36.1%).
* **Dataset indicator.** The residual magnitudes depend on the dataset (ESM
  column `ASY`), encoded as `STUDY_DINO` (1 = DINO dataset 2, 0 = Helsinki
  dataset 1). The simulations here use `STUDY_DINO = 1`. Only simulated
  observations with residual error depend on it; `Cc` does not.
* **Current body weight.** The authors projected current weight from birth
  weight with the Anchieta growth curves. Those curves are not given
  numerically, so these simulations hold current weight at birth weight. This
  mainly affects the 21-day arm of Figure 3.
* **Table 3 rate.** The simulations hold each neonate's Table 3 rate at the
  value for its postnatal age at the start of treatment. The paper does not
  say whether its Fig. S5 simulation stepped the rate up as postnatal age
  crossed a band boundary.
* **Postnatal-age units.** The source uses days. The canonical `PNA` column
  is in months, and the model converts it back (`PNA * 30.4375`).
