---
title: "Baicalein (Xiang 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Baicalein (Xiang 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Xiang_2018_baicalein"))$reference`
* Description: `r rxode2::rxode(readModelDb("Xiang_2018_baicalein"))$description`
* Article: <https://doi.org/10.3389/fphar.2018.00793>

# Population

The model was fit to a cellular in-vitro time-kill / cytokine-release
experiment in the murine RAW264.7 macrophage cell line (Center of Cellular
Resources, Chinese Academy of Sciences, Shanghai). Cells were inoculated at
1 x 10^6 cells/mL, pre-treated with baicalein at 0, 10, 20, or 40 uM for
0.5 h, then stimulated with LPS (1 ug/mL, Escherichia coli O55:B5) for an
additional 1, 2, 4, 8, 12, and 24 h. TNF-alpha and IL-6 in the cell-culture
supernatant were quantified by ELISA (MULTI SCIENCES Mouse kits;
intra/inter-assay CV 2.3-3.0% and 2.1-5.3%), iNOS expression was measured by
Western blot and normalised to the t = 0 control as
(iNOS/GAPDH) / (iNOS0/GAPDH0), and NO was quantitated by the Griess reaction
calibrated against sodium nitrite standards (Xiang 2018 Materials and
Methods). Reported sample sizes per condition were n = 3 for TNF-alpha,
IL-6, and NO, and n = 6 for iNOS (Xiang 2018 Figure 3 caption).

The same information is available programmatically via
`rxode2::rxode(readModelDb("Xiang_2018_baicalein"))$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/pharmacodynamics/Xiang_2018_baicalein.R`.
The table below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lalpha` (alpha)         | log(0.0832) | Xiang 2018 Table 1, row 1 (a) |
| `lkin_tnf` (kinTNF)      | log(3740)   | Xiang 2018 Table 1, row 2 (k_inTNFa) |
| `lkout_tnf` (koutTNF)    | log(0.0463) | Xiang 2018 Table 1, row 3 (k_outTNFa) |
| `lkin_il6` (kinIL6)      | log(0.353)  | Xiang 2018 Table 1, row 4 (k_inIL-6) |
| `lkout_il6` (koutIL6)    | log(0.143)  | Xiang 2018 Table 1, row 5 (k_outIL-6) |
| `ltau1` (tau1)           | log(1.38)   | Xiang 2018 Table 1, row 6 (tau_1) |
| `lkin_inos` (kiniNOS)    | log(0.00169)| Xiang 2018 Table 1, row 7 (k_iniNOS) |
| `ltau2` (tau2)           | log(1.41)   | Xiang 2018 Table 1, row 8 (tau_2) |
| `lkin_no` (kinNO)        | log(0.0605) | Xiang 2018 Table 1, row 9 (k_inNO) |
| `ldelta` (delta)         | log(1.35)   | Xiang 2018 Table 1, row 10 (delta) |
| `kout_inos`              | fixed(0)    | Xiang 2018 Results, Model Simulation paragraph 1 (koutiNOS fixed to 0) |
| `kout_no`                | fixed(0)    | Xiang 2018 Results, Model Simulation paragraph 1 (koutNO fixed to 0) |
| `lrbase_tnf`             | log(1000)   | Xiang 2018 Figure 4A, control curve at t = 0 (digitised) |
| `lrbase_il6`             | log(100)    | Xiang 2018 Figure 4B, control curve at t = 0 (digitised) |
| `lrbase_inos`            | log(1)      | Xiang 2018 Materials and Methods, Western Blotting (normalisation definition) |
| `lrbase_no`              | log(1)      | Xiang 2018 Figure 4D, control curve at t = 0 (digitised) |
| `propSd_tnf,_il6,_inos,_no` | fixed(0.10) | Placeholder; weighting strategy in Xiang 2018 Pharmacodynamic Modeling does not state the final-model residual-error form |
| Eq 1 (dTNF/dt)           | n/a         | Xiang 2018 Eq 1, page 4 |
| Eq 2 (f(Bai))            | n/a         | Xiang 2018 Eq 2, page 4 |
| Eq 3 (TNFa_0)            | n/a         | Xiang 2018 Eq 3, page 4 |
| Eq 4 (dIL-6/dt)          | n/a         | Xiang 2018 Eq 4, page 4 |
| Eq 5 (IL-6_0)            | n/a         | Xiang 2018 Eq 5, page 5 |
| Eq 6 (diNOS/dt)          | n/a         | Xiang 2018 Eq 6, page 5 |
| Eq 7 (dNO/dt)            | n/a         | Xiang 2018 Eq 7, page 5 |

## Dimensional analysis

Endogenous / mechanistic models in the package are required to carry a
worked dimensional analysis (per `references/endogenous-validation.md`).
The unit table below shows the units of every symbol on the right-hand
side of each ODE in `model()`; the left-hand side must equal
`[state units]/[time units]` (i.e., the state's reporting unit per hour).

| Term | Units of each factor | Product (= LHS) |
|------|----------------------|-----------------|
| `kin_tnf * (1 - f_bai)` | (pg/mL/h) * (unitless) | pg/mL/h = d/dt(TNF) |
| `kout_tnf * tnf`        | (1/h) * (pg/mL)        | pg/mL/h = d/dt(TNF) |
| `(tnf - tnf_delay1) / tau1` | (pg/mL) / (h) | pg/mL/h = d/dt(TNF_delay1) |
| `(tnf - tnf_delay2) / tau2` | (pg/mL) / (h) | pg/mL/h = d/dt(TNF_delay2) |
| `kin_il6 * tnf_delay1`  | (1/h) * (pg/mL)        | pg/mL/h = d/dt(IL-6) |
| `kout_il6 * il6`        | (1/h) * (pg/mL)        | pg/mL/h = d/dt(IL-6) |
| `kin_inos * tnf_delay2` | (1/h) * (pg/mL)        | pg/mL/h then normalized (note below) |
| `kout_inos * inos`      | (1/h) * (unitless)     | unitless/h |
| `kin_no * inos^delta`   | (uM/h) * (unitless)    | uM/h = d/dt(NO) |
| `kout_no * no`          | (1/h) * (uM)           | uM/h |

**Notes on apparent units.**

* `kinTNF` is reported with the unit `h^-1` in Xiang 2018 Table 1 row 2;
  Eq 1 requires the production-rate term to have the unit pg/mL/h, so
  `kinTNF` is interpreted as carrying units pg/mL/h. The Table 1 unit
  label is treated as a typesetting slip (the same column header was
  applied to all first-order rate constants). At the published value
  3.74 x 10^3, this interpretation is also numerically self-consistent
  with the observed t = 1 h peak (see the smoke check below).
* `kinNO` is reported with the unit `h^-1` in Xiang 2018 Table 1 row 9;
  Eq 7 with `iNOS^delta` unitless requires `kinNO` to carry units uM/h.
  Same Table-1 unit-label slip as `kinTNF`.
* `kiniNOS` (Table 1 row 7) carries units 1/h; Eq 6 multiplies the
  delayed TNF-alpha concentration in pg/mL, so the iNOS state's
  effective unit is iNOS-ratio per (pg/mL * h)) at unit TNF-alpha. iNOS
  is reported as the unitless ratio (iNOS/GAPDH) / (iNOS0/GAPDH0), so
  the "unit" of d/dt(iNOS) is iNOS-ratio per h; the dimensional
  consistency is therefore exact only if kiniNOS is read as having
  units (1/h) * (pg/mL)^-1 = 1/(h * pg/mL). The package retains the
  Table 1 value 0.00169 verbatim; downstream users who refit the model
  on a different TNF-alpha unit basis should rescale kiniNOS
  accordingly (see Assumptions and deviations below).

# Parameter table -- paper vs. file

```{r param-table, echo = FALSE}
tibble::tribble(
  ~Parameter,    ~`Paper (Table 1)`, ~`File (ini)`,        ~CV_pct_in_paper,
  "alpha",       0.0832,             exp(-2.4866),          "15%",
  "kinTNF",      3740,               exp(8.2267),           "6%",
  "koutTNF",     0.0463,             exp(-3.0719),          "32%",
  "kinIL6",      0.353,              exp(-1.0413),          "9%",
  "koutIL6",     0.143,              exp(-1.9445),          "23%",
  "tau1",        1.38,               exp(0.3221),           "11%",
  "kiniNOS",     0.00169,            exp(-6.3819),          "9%",
  "tau2",        1.41,               exp(0.3436),           "13%",
  "kinNO",       0.0605,             exp(-2.8042),          "37%",
  "delta",       1.35,               exp(0.3001),           "9%",
  "koutiNOS",    0,                  0,                     "fixed to 0",
  "koutNO",      0,                  0,                     "fixed to 0"
) |>
  knitr::kable(digits = 5,
               caption = "Xiang 2018 Table 1 vs the values stored in inst/modeldb/pharmacodynamics/Xiang_2018_baicalein.R. CV% in the source paper is the precision of the estimate (RSE), not between-subject variability.")
```

# Simulation -- replicate Figure 4

Xiang 2018 Figure 4 shows the model-predicted and measured time courses
of TNF-alpha (panel A), IL-6 (panel B), iNOS (panel C/E), and NO
(panel D) over 24.5 h in LPS-stimulated RAW264.7 macrophages at four
baicalein concentrations (0 = control, 10, 20, 40 uM). The simulation
below uses the packaged typical-value model with four virtual "wells"
(one per concentration), zero between-subject variability, and the
figure-derived baseline values.

```{r simulate}
mod <- rxode2::rxode(nlmixr2lib::readModelDb("Xiang_2018_baicalein"))

times <- seq(0, 24, by = 0.25)
concentrations <- c(0, 10, 20, 40)

make_well <- function(conc, id) {
  rxode2::et(amt = 0, time = 0) |>
    rxode2::et(times) |>
    as.data.frame() |>
    dplyr::mutate(id = id, CONC_BAI_UM = conc)
}

events <- dplyr::bind_rows(
  make_well(concentrations[1], id = 1L),
  make_well(concentrations[2], id = 2L),
  make_well(concentrations[3], id = 3L),
  make_well(concentrations[4], id = 4L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

sim <- rxode2::rxSolve(mod, events = events,
                       keep = "CONC_BAI_UM",
                       returnType = "data.frame") |>
  dplyr::as_tibble() |>
  dplyr::mutate(label = paste0("Baicalein ", CONC_BAI_UM, " uM"))
```

## Figure 4A -- TNF-alpha

Replicates Figure 4A of Xiang 2018 (TNF-alpha time course at 0, 10, 20,
40 uM baicalein in LPS-stimulated RAW264.7 macrophages).

```{r figure-4a}
ggplot(sim, aes(time, tnf, colour = label)) +
  geom_line() +
  labs(x = "Time post-LPS (h)",
       y = "TNF-alpha (pg/mL)",
       colour = "Baicalein conc.",
       title = "Figure 4A -- TNF-alpha trajectory",
       caption = "Replicates Figure 4A of Xiang 2018.")
```

## Figure 4B -- IL-6

```{r figure-4b}
ggplot(sim, aes(time, il6, colour = label)) +
  geom_line() +
  labs(x = "Time post-LPS (h)",
       y = "IL-6 (pg/mL)",
       colour = "Baicalein conc.",
       title = "Figure 4B -- IL-6 trajectory",
       caption = "Replicates Figure 4B of Xiang 2018.")
```

## Figure 4C/E -- iNOS

```{r figure-4ce}
ggplot(sim, aes(time, inos, colour = label)) +
  geom_line() +
  labs(x = "Time post-LPS (h)",
       y = "Relative iNOS expression ((iNOS/GAPDH) / (iNOS0/GAPDH0))",
       colour = "Baicalein conc.",
       title = "Figure 4C/E -- iNOS trajectory",
       caption = "Replicates Figures 4C and 4E of Xiang 2018.")
```

## Figure 4D -- NO

```{r figure-4d}
ggplot(sim, aes(time, no, colour = label)) +
  geom_line() +
  labs(x = "Time post-LPS (h)",
       y = "NO (uM)",
       colour = "Baicalein conc.",
       title = "Figure 4D -- NO trajectory",
       caption = "Replicates Figure 4D of Xiang 2018.")
```

# Smoke check -- t = 1 h TNF-alpha peak

Xiang 2018 Figure 4A shows that the TNF-alpha trace peaks at about 1 h
after LPS stimulation, then gradually declines through 12 h. The
packaged model's simple turnover form (constant LPS-driven production
plus first-order elimination) cannot reproduce the peak-and-decline; it
predicts a monotonic approach to the steady-state asymptote
kinTNF / koutTNF ~= 80800 pg/mL. The published model fit in Figure 4
appears to track the data more closely than the published equations
permit (see Assumptions and deviations).

The simulated t = 1 h TNF-alpha concentration is, however, in close
agreement with the observed peak (Xiang 2018 Figure 2B / 4A reads
~4500 pg/mL in the LPS-only control):

```{r smoke-tnf-peak}
sim |>
  dplyr::filter(abs(time - 1) < 1e-6) |>
  dplyr::select(label, time, tnf) |>
  dplyr::mutate(tnf = round(tnf, 0)) |>
  knitr::kable(caption = "Simulated TNF-alpha at t = 1 h post-LPS, by baicalein concentration. The LPS-only control value (~4600 pg/mL) is close to the Figure 4A observed peak (~4500 pg/mL).")
```

# Baseline-recovery smoke check

This model is not a quiescent steady-state system (the LPS-stimulated
kinTNF zero-order production is on from t = 0), so the classic
endogenous-model "set initial condition to baseline, expect no drift"
check is not applicable. A meaningful sanity test is to confirm that
when kinTNF is set to zero (no LPS), the four states stay near their
baselines for the duration of the experiment (linear-cascade
self-consistency).

```{r baseline-recovery}
# Build a parameter override: kin_tnf -> 0 (no LPS stimulation)
mod_no_lps <- rxode2::rxode(nlmixr2lib::readModelDb("Xiang_2018_baicalein"))
sim_no_lps <- rxode2::rxSolve(
  mod_no_lps,
  events = make_well(conc = 0, id = 1L),
  params = c(lkin_tnf = log(1e-12)),  # effectively zero
  returnType = "data.frame"
)
sim_no_lps |>
  dplyr::filter(time %in% c(0, 6, 12, 18, 24)) |>
  dplyr::select(time, tnf, il6, inos, no) |>
  dplyr::mutate(across(c(tnf, il6, inos, no), \(x) round(x, 3))) |>
  knitr::kable(caption = "Without LPS stimulation (kinTNF ~= 0), TNF-alpha decays slowly toward zero (driven only by koutTNF); IL-6 rises slowly because the model's IL-6 production is proportional to the residual TNF-alpha rather than to a TNF-alpha excursion from baseline. iNOS and NO drift up monotonically because koutiNOS and koutNO are fixed at zero. The package's baseline values are therefore the experimental t = 0 control measurements, not steady-state solutions of the unstimulated system.")
```

# Assumptions and deviations

The Xiang 2018 paper is a published cellular pharmacodynamic model with
a fully specified equation set (Eqs 1-7) and a complete typical-value
parameter table (Table 1), but several encoding decisions in the
package require attention.

* **Drug field corrected.** The task dispatcher's metadata listed
  the drug as "Frontiers in Pharmacology" (the journal name) -- a
  task-generator bug observed on several sibling tasks. The actual
  drug studied is baicalein (5,6,7-trihydroxy-2-phenyl-4H-1-benzopyran-
  4-one), the primary active flavonoid of Scutellaria baicalensis
  Georgi. Filename and vignette title corrected accordingly.

* **Species and model class.** The model was fit to an in-vitro
  experiment in the RAW264.7 mouse macrophage cell line, not to
  in-vivo human or animal data. `population$species` is set to
  "in vitro (RAW264.7 mouse macrophage cell line)" and the
  `description` field is prefixed with "In vitro (RAW264.7 mouse
  macrophage cell line)." per the SKILL Phase 1 step 3 species-
  recording requirement.

* **f(Bai) at C_Bai = 0.** Xiang 2018 Eq 2 writes the inhibition
  function as f(Bai) = alpha * ln(C_Bai). At C_Bai = 0 (the control
  wells with no baicalein) the bare ln is undefined. The package
  encodes f(Bai) = alpha * log(CONC_BAI_UM + 1), which (a) returns
  f = 0 in control wells without a divide-by-zero, and (b) differs
  from the bare ln by less than 5% at the three tested concentrations
  10, 20, 40 uM:

  - 10 uM: alpha*ln(10) = 0.192 vs alpha*ln(11) = 0.200 (delta 4.4%)
  - 20 uM: alpha*ln(20) = 0.249 vs alpha*ln(21) = 0.253 (delta 1.6%)
  - 40 uM: alpha*ln(40) = 0.307 vs alpha*ln(41) = 0.309 (delta 0.7%)

  The +1-shift convention is also the standard pharmacometric
  treatment of log-linear concentration-response with a controllable
  control arm; the paper's "f = 0 in control" condition appears in
  Eqs 3 and 5 ("TNF-alpha_0 = Control(TNF-alpha)").

* **Lag implementation -- single-compartment delay states (DDE
  approximation).** Eqs 4 and 6 use TNF-alpha(t - tau1) and
  TNF-alpha(t - tau2), which are true delay differential equation
  forms. rxode2 does not provide a native DDE solver, so the package
  encodes each delayed signal as a single first-order distribution
  compartment with mean transit time tau:

  ```
  d/dt(tnf_delay) = (tnf - tnf_delay) / tau
  ```

  The impulse response of this filter is a single-exponential ramp,
  not a Dirac shift -- a step in tnf at time t produces a smoothed
  response in tnf_delay over the next several tau, rather than an
  exact translation. For the LPS-stimulated TNF-alpha trace, which
  is itself a smooth approach to its asymptote rather than a step,
  the single-compartment delay is a reasonable approximation; the
  alternative is a multi-compartment transit chain (n delays each
  with mean transit time tau / n), which sharpens the response
  toward a step shift but adds n - 1 extra states. The user may
  build the alternative encoding by editing model() to chain n
  delay compartments.

* **Baseline values are figure-derived.** Xiang 2018 reports the
  initial conditions only as `Control(TNF-alpha)`, `Control(IL-6)`,
  and `(iNOS/GAPDH) / (iNOS0/GAPDH0) = 1` -- with the numeric values
  read off Figure 4 control curves at t = 0. The packaged defaults
  (1000 pg/mL for TNF-alpha, 100 pg/mL for IL-6, 1 for iNOS, 1 uM
  for NO) are operator digitisations of those figure points; users
  who have their own experimental t = 0 control measurements should
  override the corresponding `lrbase_*` parameters when refitting.
  iNOS is definitionally 1 at t = 0 per the Western-blot
  normalisation; the only figure-derived value with non-trivial
  uncertainty is bl_tnf.

* **Unit-label slips in Table 1.** As detailed in the Dimensional
  Analysis section, the Table 1 column "Estimate" for kinTNF and
  kinNO carries the unit label "h^-1" but the equations require
  pg/mL/h and uM/h respectively. The package preserves the Table 1
  numeric values (3.74 x 10^3 and 0.0605) and interprets the unit
  labels as typesetting slips -- a single column unit was applied
  to all first-order rate-constant rows including the two zero-order
  production rows. This interpretation is internally self-consistent
  with the observed t = 1 h peak (see smoke check above).

* **No IIV and no residual error are reported.** Xiang 2018
  Pharmacodynamic Modeling lists the weighting strategies considered
  (additive, power, multiplicative) but does not state the final
  choice. The reported CV% column in Table 1 is the precision of the
  estimate (RSE), not between-subject variability. The package
  carries placeholder 10% proportional residual on each of the four
  outputs (`propSd_tnf`, `propSd_il6`, `propSd_inos`, `propSd_no`)
  so the model is nlmixr2-fit-compatible; downstream users who refit
  the model on real data should replace these with whatever residual
  structure their data support. No `eta_*` IIVs are included -- the
  paper does not report between-subject (between-well or between-
  experiment) variability.

* **Non-canonical compartment names (lint warnings).**
  `checkModelConventions()` flags the six ODE-state names `tnf`,
  `tnf_delay1`, `tnf_delay2`, `il6`, `inos`, and `no` as non-canonical
  because they are not among the canonical PK/PD compartments (depot,
  central, peripheral1, ..., or registered metabolites). This is the
  same pattern used by other in-vitro / mechanism-model entries in
  the package whose compartments are biological-species states rather
  than drug-distribution compartments (Clewe 2018 TB MTP-GPDI:
  `fbugs`/`sbugs`/`nbugs`/`ar_off`/`ar_on`; Charbonneau 2021
  phenylalanine: `gut`/`phe`; Tetschke 2018 erythropoiesis: `thb`).
  These are kept as model-readable shorthand for the source species
  rather than renamed to numbered compartments, because the species
  identity is the meaningful entity in a cytokine-cascade mechanism.
  `tnf_delay1` and `tnf_delay2` are the single-compartment delay
  states implementing the DDE approximation described above.

* **Dosing/concentration unit mismatch (lint warning).**
  `checkModelConventions()` flags `units$dosing` (uM, baicalein) vs
  `units$concentration` numerator (pg, TNF-alpha/IL-6) as dimensionally
  incompatible. The mismatch is genuine -- baicalein concentration is
  reported in molar units while the cytokine readouts are in mass-
  concentration -- and reflects the inherent mixed-dimension nature
  of a small-molecule perturbant driving a protein-readout in-vitro
  PD experiment. No correction is appropriate; the warning is
  documented here per the SKILL convention for justified deviations.

* **Peak-and-decline gap.** The observed TNF-alpha trace in Figure 4A
  peaks at t = 1 h and then declines through 12 h before rising
  slowly to 24 h, while the model equation
  (constant kinTNF, first-order elimination koutTNF) predicts
  monotonic approach to the steady-state asymptote kinTNF / koutTNF
  ~= 80800 pg/mL. The model's t = 1 h prediction agrees with the
  observed peak, but later time points (especially t = 6-12 h)
  diverge upward. The paper's Figure 4 model curves nevertheless
  appear to track the data more closely than the published equations
  permit, suggesting either an unreported LPS-decay or self-feedback
  term in the model's actual simulation, or a parameter-set
  discrepancy between Table 1 and the figure's generating script.
  This is an acknowledged published-model limitation; the package
  encodes Eqs 1-7 verbatim with Table 1's parameter values and does
  NOT modify the equations to match Figure 4 visually.
