---
title: "Tofacitinib (Xie 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tofacitinib (Xie 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Xie_2019_tofacitinib"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Xie_2019_tofacitinib"))$description`
* Article: <https://doi.org/10.5414/CP203516>
* Open-access full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6704728/>

Tofacitinib is an oral Janus kinase inhibitor. Xie 2019 pooled the two phase 3
psoriatic-arthritis studies OPAL Broaden (NCT01877668) and OPAL Beyond
(NCT01882439) and fitted a one-compartment model with first-order absorption and
an absorption lag, parameterized in apparent oral clearance (CL/F) and apparent
volume of distribution (V/F). The analysis used a **full** covariate model --
every pre-specified covariate is retained with an estimated coefficient rather
than selected in or out by hypothesis testing -- so several of the coefficients
below have confidence intervals that include the null value. They are kept here
because that is how the authors reported the model.

```{r load}
mod <- rxode2::rxode2(readModelDb("Xie_2019_tofacitinib"))
```

# Population

Baseline demographics reproduce Table 2 of Xie 2019 (n = 650 tofacitinib-treated
patients with 3,252 plasma concentration measurements).

| Characteristic | Value |
|---|---|
| Age, years | mean 49.0 (SD 12.1), range 18-78 |
| Baseline body weight (BWT), kg | mean 84.8 (SD 19.0), range 38.1-159.7 |
| Baseline creatinine clearance (BCCL), mL/min | mean 123.4 (SD 37.7), range 49.1-348.5 |
| Baseline C-reactive protein (BCRP), mg/dL | mean 1.1 (SD 2.0), range 0.0-16.4 |
| Female | 360 (55.4%) |
| Race: White / Black / Asian / Other | 610 (93.9%) / 3 (0.5%) / 20 (3.1%) / 17 (2.6%) |
| Ethnicity: Hispanic / non-Hispanic | 69 (10.6%) / 581 (89.4%) |
| Dosing | tofacitinib 5 or 10 mg orally twice daily |

Sampling was sparse and trough-weighted: month 1 (pre-dose and 2 h post-dose),
months 4 and 6 (pre-dose and 0.5, 2 and 3 h post-dose), with pre-dose samples
drawn 12 +/- 2 h after the previous evening dose.

The **reference patient** used throughout the paper's covariate assessment
(Table 3 footnote b) is White, male, **Hispanic**, body weight 83.3 kg, age
50 years, BCRP 0.49 mg/dL, BCCL 120 mL/min. Note that the ethnicity reference is
the Hispanic group, which is the *smaller* of the two (10.6% of the cohort);
this is stated explicitly in Materials and methods ("For ethnicity, Hispanic
patients were used as the baseline typical patient for simulations").

```{r refpatient}
reference_patient <- list(
  AGE = 50, WT = 83.3, CRCL_BASE = 120, CRP = 0.49, SEXF = 0,
  RACE_BLACK = 0, RACE_ASIAN = 0, RACE_OTHER = 0, RACE_HISPANIC = 1
)
tau <- 12 # dosing interval, h
```

# Source trace

Every structural parameter, covariate effect and variance component, with the
location it was read from. All values are from the printed paper; no supplement
was required and none of the parameters were digitized from a figure.

| Model quantity | Source location | Value |
|---|---|---|
| One-compartment, first-order absorption + lag | Results, para. 2; Discussion para. 2 | structure |
| `lka` = log(Ka) | Table 3, `Ka, /h` | 13.8 (RSE 7.9%) |
| `lcl` = log(CL/F) | Table 3, `CL/F, L/h` | 20.4 (RSE 4.7%) |
| `lvc` = log(V/F) | Table 3, `V/F, L` | 110 (RSE 1.2%) |
| `ltlag` = log(lag time) | Table 3, `Lag time, h` | 0.3 (RSE 0.8%) |
| Continuous covariate form | Equation 1 | `theta_TV * (cov/cov_median)^theta_x` |
| Categorical covariate form | Equations 2 and 3 | `theta_TV * theta_x`; reference level gives `theta_TV` |
| `e_age_cl` | Table 3, covariate CL/F ~ Age | -0.20 |
| `e_wt_cl` | Table 3, covariate CL/F ~ BWT; footnote b | 0 (FIX) |
| `e_crcl_base_cl` | Table 3, covariate CL/F ~ BCCL | 0.32 |
| `e_crp_cl` | Table 3, covariate CL/F ~ BCRP | -0.02 |
| `e_race_black_cl` | Table 3, covariate CL/F ~ Black | 0.91 |
| `e_race_asian_cl` | Table 3, covariate CL/F ~ Asian | 0.95 |
| `e_race_other_cl` | Table 3, covariate CL/F ~ Other race | 0.96 |
| `e_race_hispanic_cl` | Table 3, covariate CL/F ~ Non-Hispanic | 1.12 |
| `e_sexf_cl` | Table 3, covariate CL/F ~ Female | 1.05 |
| `e_age_vc` | Table 3, covariate V/F ~ Age | -0.22 |
| `e_wt_vc` | Table 3, covariate V/F ~ BWT | 0.68 |
| `etalcl` | Table 3, `CL/F` IIV column | 31.7% |
| `etalka` | Table 3, `Ka` IIV column | 198% |
| `etaruv` | Table 3, IIV column on both proportional-error rows | 65.8% |
| `vc_eta_scale` | Table 3, `Scaling parameter`; Results para. 2 | 0.5 |
| `propSd_early` | Table 3, `Proportional error, TAD <= 5 hours, %` | 22.9 |
| `propSd_late` | Table 3, `Proportional error, TAD > 5 hours, %` | 52.6 |
| Reference patient covariate values | Table 3 footnote b; Materials and methods | 83.3 kg, 50 y, 0.49 mg/dL, 120 mL/min |
| Concentration units (ng/mL) | Figure 1 and Figure 2 axis labels | ng/mL |

Two structural points are worth spelling out because they are stated in prose
rather than in the table:

* **V/F carries no eta of its own.** Table 3 shows `-` in the IIV column for
  V/F. Instead, "a scaling parameter was used to describe the impact of the IIV
  of the V/F on the IIV of CL/F" (Results). The V/F individual deviation is
  therefore `vc_eta_scale * etalcl`, which is what produces the "covariance
  term" the Methods mention and forces the CL/F-V/F random-effect correlation to
  exactly 1.
* **The body-weight effect on CL/F is fixed to zero**, not absent: "initial runs
  estimated a biologically implausible, negative exponent" (Results). The
  zero-exponent term is retained in `model()` so the authors' full covariate
  structure is reproduced literally.

# Typical values for the reference patient

The first check is that the model returns the paper's Table 3 point estimates
for the reference patient, i.e. that every covariate term collapses to 1 there.

```{r typical}
tv <- rxode2::zeroRe(mod)

with_covariates <- function(ev, overrides = list()) {
  p <- utils::modifyList(reference_patient, overrides)
  out <- as.data.frame(ev)
  for (nm in names(p)) out[[nm]] <- p[[nm]]
  out
}

single <- rxode2::et(amt = 5, cmt = "depot") |> rxode2::et(1, cmt = "central")
param_at <- function(overrides, what) {
  rxode2::rxSolve(tv, with_covariates(single, overrides), returnType = "data.frame")[[what]][1]
}

typical <- data.frame(
  Parameter = c("CL/F (L/h)", "V/F (L)", "Ka (1/h)", "Lag time (h)"),
  Model = c(param_at(list(), "cl"), param_at(list(), "vc"),
            param_at(list(), "ka"), param_at(list(), "tlag")),
  Paper = c(20.4, 110, 13.8, 0.3)
)
knitr::kable(typical, digits = 4)

stopifnot(isTRUE(all.equal(typical$Model, typical$Paper, tolerance = 1e-8)))
```

# Reproducing the paper's covariate-impact statements

The Results section quantifies each covariate effect in prose. Because every
effect is a deterministic function of the reference-patient values, these are
exact arithmetic checks -- no simulation noise is involved, so a tight tolerance
is the correct one. Each row below restates one sentence of the paper.

The `Tolerance` column is set from the precision the paper printed each number
to, not from how close the model happens to land: a value quoted as "8.97%"
constrains the model far more tightly than one quoted as "~ 19%" or "32%".

```{r covariate-impacts}
base_cl <- param_at(list(), "cl")
base_vc <- param_at(list(), "vc")

impacts <- dplyr::bind_rows(
  # Quoted to two decimals in the Results, so this is the tightest row.
  data.frame(Quantity = "CL/F", Scenario = "Age 80 y vs 50 y", Quoted = "8.97%",
             Model = 100 * (param_at(list(AGE = 80), "cl") / base_cl - 1),
             Paper = -8.97, Tolerance = 0.02),
  # Quoted to one decimal.
  data.frame(Quantity = "V/F", Scenario = "Age 80 y vs 50 y", Quoted = "9.9%",
             Model = 100 * (param_at(list(AGE = 80), "vc") / base_vc - 1),
             Paper = -9.9, Tolerance = 0.12),
  data.frame(Quantity = "CL/F", Scenario = "BCCL 50 vs 120 mL/min", Quoted = "24.3%",
             Model = 100 * (param_at(list(CRCL_BASE = 50), "cl") / base_cl - 1),
             Paper = -24.3, Tolerance = 0.15),
  data.frame(Quantity = "CL/F", Scenario = "BCCL 60 vs 120 mL/min", Quoted = "19.8%",
             Model = 100 * (param_at(list(CRCL_BASE = 60), "cl") / base_cl - 1),
             Paper = -19.8, Tolerance = 0.15),
  data.frame(Quantity = "CL/F", Scenario = "BCRP 3.0 vs 0.49 mg/dL", Quoted = "3.6%",
             Model = 100 * (param_at(list(CRP = 3.0), "cl") / base_cl - 1),
             Paper = -3.6, Tolerance = 0.1),
  # Coefficient itself is rounded to two decimals in Table 3 (see the note below).
  data.frame(Quantity = "CL/F", Scenario = "Female vs male", Quoted = "5.4%",
             Model = 100 * (param_at(list(SEXF = 1), "cl") / base_cl - 1),
             Paper = 5.4, Tolerance = 0.45),
  data.frame(Quantity = "CL/F", Scenario = "Non-Hispanic vs Hispanic", Quoted = "12.3%",
             Model = 100 * (param_at(list(RACE_HISPANIC = 0), "cl") / base_cl - 1),
             Paper = 12.3, Tolerance = 0.35),
  # Quoted as "~ 19%" / "~ 20%" / "32%", i.e. to the nearest whole percent.
  data.frame(Quantity = "V/F", Scenario = "BWT 61 kg vs 83.3 kg", Quoted = "~19%",
             Model = 100 * (param_at(list(WT = 61), "vc") / base_vc - 1),
             Paper = -19, Tolerance = 0.5),
  data.frame(Quantity = "V/F", Scenario = "BWT 109 kg vs 83.3 kg", Quoted = "~20%",
             Model = 100 * (param_at(list(WT = 109), "vc") / base_vc - 1),
             Paper = 20, Tolerance = 0.5),
  data.frame(Quantity = "AUCss", Scenario = "BCCL 50 vs 120 mL/min", Quoted = "32%",
             Model = 100 * (base_cl / param_at(list(CRCL_BASE = 50), "cl") - 1),
             Paper = 32, Tolerance = 0.5)
) |>
  dplyr::mutate(Difference = Model - Paper, Pass = abs(Difference) < Tolerance)

impacts |>
  dplyr::rename("Paper wording" = Quoted, "Change from model (%)" = Model,
                "Change reported by paper (%)" = Paper,
                "Difference (pp)" = Difference, "Tolerance (pp)" = Tolerance) |>
  knitr::kable(digits = 2)
```

Every row agrees with the paper to within the precision the paper printed. The
two categorical rows -- female sex and non-Hispanic ethnicity -- carry the
loosest tolerances for a specific reason: the model gives exactly 5.00% and
12.00% because Table 3 prints the multiplicative factors rounded to two decimals
(1.05 and 1.12), while the Results prose quotes 5.4% and 12.3% computed from the
unrounded NONMEM estimates. Both prose values round back to the printed Table 3
factors (1.054 -> 1.05, 1.123 -> 1.12), so this is a reporting-precision gap in
the source, not a disagreement.

```{r covariate-impacts-assert}
stopifnot(all(impacts$Pass))
```

The three race factors are read back directly, since Table 3 reports them as
multiplicative factors on CL/F relative to the White reference.

```{r race-factors}
race_factors <- data.frame(
  Race = c("Black", "Asian", "Other"),
  Model = c(param_at(list(RACE_BLACK = 1), "cl"),
            param_at(list(RACE_ASIAN = 1), "cl"),
            param_at(list(RACE_OTHER = 1), "cl")) / base_cl,
  Paper = c(0.91, 0.95, 0.96)
)
knitr::kable(race_factors, digits = 4)

stopifnot(isTRUE(all.equal(race_factors$Model, race_factors$Paper, tolerance = 1e-8)))
# The body-weight exponent on CL/F is fixed to zero, so weight must not move CL/F.
stopifnot(isTRUE(all.equal(param_at(list(WT = 150), "cl"), base_cl)))
```

# Steady-state solve against the closed form

A one-compartment oral model with a lag has a closed-form steady-state solution,
so the ODE solve can be checked against it directly. Both sides use the same
parameter values and differ only by numerical integration error, so a tight
bound is appropriate here (this is the case the package conventions carve out
from the "assert on the centre, not the extremes" rule).

```{r closed-form}
n_load <- 28 # 29 doses over 14 days, well past steady state
t_ss <- n_load * tau

# Log-spaced grid: absorption is fast (Ka 13.8 /h) so the peak needs fine early
# resolution, while the trough needs coverage out to 12 h.
obs_times <- sort(unique(c(0, exp(seq(log(0.01), log(tau), length.out = 400)))))

ss_closed_form <- function(t, dose, ka, kel, v, tau, tlag) {
  # The lag shifts the periodic steady-state profile; it does not clip it.
  tt <- (t - tlag) %% tau
  amp <- dose * ka / (v * (ka - kel))
  1000 * amp * (exp(-kel * tt) / (1 - exp(-kel * tau)) -
                  exp(-ka * tt) / (1 - exp(-ka * tau)))
}

ss_profile <- function(dose) {
  ev <- rxode2::et(amt = dose, cmt = "depot", ii = tau, addl = n_load) |>
    rxode2::et(t_ss + obs_times, cmt = "central")
  rxode2::rxSolve(tv, with_covariates(ev), returnType = "data.frame", maxsteps = 100000L)
}

cf_check <- lapply(c(5, 10), function(dose) {
  d <- ss_profile(dose)
  cf <- ss_closed_form(d$time - t_ss, dose, d$ka[1], d$cl[1] / d$vc[1], d$vc[1], tau, d$tlag[1])
  data.frame(
    Dose = paste(dose, "mg BID"),
    `Cmax ODE` = max(d$Cc), `Cmax closed form` = max(cf),
    `Cmin ODE` = min(d$Cc), `Cmin closed form` = min(cf),
    `Max relative difference` = max(abs(d$Cc - cf) / cf),
    check.names = FALSE
  )
}) |> dplyr::bind_rows()

knitr::kable(cf_check, digits = c(0, 3, 3, 3, 3, 10))

stopifnot(all(cf_check$`Max relative difference` < 1e-4))
```

# Mass balance: CL/F x AUC(0-tau) = Dose at steady state

Because the model is parameterized in *apparent* clearance, bioavailability is
folded into CL/F and the steady-state identity is exact:
`CL/F * AUC(0-tau) = Dose`. This is the strongest single gate on the model --
it fails if the dose route, the depot, the concentration unit conversion, the
lag, or the elimination term is wrong.

```{r mass-balance-typical}
mb_typical <- lapply(c(5, 10), function(dose) {
  d <- ss_profile(dose)
  auc <- sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  data.frame(
    Dose = paste(dose, "mg BID"),
    `AUC(0-tau) simulated (ng*h/mL)` = auc,
    `Dose / (CL/F) (ng*h/mL)` = 1000 * dose / d$cl[1],
    `Relative difference` = abs(auc - 1000 * dose / d$cl[1]) / (1000 * dose / d$cl[1]),
    check.names = FALSE
  )
}) |> dplyr::bind_rows()

knitr::kable(mb_typical, digits = c(0, 3, 3, 10))

stopifnot(all(mb_typical$`Relative difference` < 1e-4))
```

To confirm the gate is not vacuous, the same computation is run against a
deliberately broken copy of the model in which the concentration scale factor is
wrong by 10x. A gate that cannot fail proves nothing.

```{r mass-balance-mutation}
broken <- rxode2::rxode2(
  paste(
    "cl <- 20.4; vc <- 110; ka <- 13.8; tlag <- 0.3; kel <- cl / vc",
    "d/dt(depot) <- -ka * depot",
    "d/dt(central) <- ka * depot - kel * central",
    "alag(depot) <- tlag",
    "Cc <- 10000 * central / vc", # <- deliberately 10x wrong
    sep = "\n"
  )
)
ev_b <- rxode2::et(amt = 5, cmt = "depot", ii = tau, addl = n_load) |>
  rxode2::et(t_ss + obs_times, cmt = "central")
db <- rxode2::rxSolve(broken, ev_b, returnType = "data.frame", maxsteps = 100000L)
auc_b <- sum(diff(db$time) * (head(db$Cc, -1) + tail(db$Cc, -1)) / 2)
mutation_rel_diff <- abs(auc_b - 1000 * 5 / 20.4) / (1000 * 5 / 20.4)
cat("Mutated model relative difference:", signif(mutation_rel_diff, 4), "\n")

stopifnot(mutation_rel_diff > 1) # the mutation must break the gate
```

# Virtual cohort

200 virtual subjects are drawn with covariate distributions matching Table 2,
and **each subject is simulated at both 5 and 10 mg twice daily**. Using one
cohort across both dose levels rather than two independent cohorts makes the
dose-proportionality check below an exact test of model linearity instead of a
comparison of two different random samples.

The paper reports only mean, SD and range for the continuous covariates, so the
distributional shape is an assumption -- truncated normals matched to the
reported mean and SD and clipped at the reported range. This is documented in
the Assumptions section.

```{r cohort}
rxode2::rxSetSeed(20190719)
set.seed(20190719)
n_sub <- 200

rtrunc_norm <- function(n, mean, sd, lower, upper) {
  x <- stats::rnorm(n, mean, sd)
  while (any(bad <- x < lower | x > upper)) {
    x[bad] <- stats::rnorm(sum(bad), mean, sd)
  }
  x
}

race <- sample(c("White", "Black", "Asian", "Other"), n_sub,
               replace = TRUE, prob = c(93.9, 0.5, 3.1, 2.6) / 100)

cohort <- data.frame(
  id = seq_len(n_sub),
  AGE = rtrunc_norm(n_sub, 49.0, 12.1, 18, 78),
  WT = rtrunc_norm(n_sub, 84.8, 19.0, 38.1, 159.7),
  CRCL_BASE = rtrunc_norm(n_sub, 123.4, 37.7, 49.1, 348.5),
  CRP = rtrunc_norm(n_sub, 1.1, 2.0, 0.0, 16.4),
  SEXF = stats::rbinom(n_sub, 1, 0.554),
  RACE_BLACK = as.integer(race == "Black"),
  RACE_ASIAN = as.integer(race == "Asian"),
  RACE_OTHER = as.integer(race == "Other"),
  RACE_HISPANIC = stats::rbinom(n_sub, 1, 0.106)
)

cohort |>
  dplyr::summarise(
    `Age (y)` = sprintf("%.1f (%.1f)", mean(AGE), sd(AGE)),
    `Weight (kg)` = sprintf("%.1f (%.1f)", mean(WT), sd(WT)),
    `BCCL (mL/min)` = sprintf("%.1f (%.1f)", mean(CRCL_BASE), sd(CRCL_BASE)),
    `BCRP (mg/dL)` = sprintf("%.2f (%.2f)", mean(CRP), sd(CRP)),
    `Female (%)` = sprintf("%.1f", 100 * mean(SEXF)),
    `Hispanic (%)` = sprintf("%.1f", 100 * mean(RACE_HISPANIC))
  ) |>
  tidyr::pivot_longer(dplyr::everything(), names_to = "Characteristic",
                      values_to = "Simulated cohort, mean (SD)") |>
  knitr::kable()
```

The absorption rate carries a very large inter-individual variability (198% CV),
so the slowest-absorbing subjects have a half-life of absorption on the order of
15 h rather than the typical 3 minutes. The 14-day loading period is sized from
that tail rather than from the typical half-life, so every subject -- not just
the typical one -- is at steady state when the NCA interval opens.

Each arm is solved in its own `rxSolve()` call, with the subject ids and the
rxode2 seed reset identically before each, so subject `i` receives the *same*
random-effect draw at both dose levels. The chunk asserts that the realized
individual CL/F, V/F and Ka are identical across the two arms; without that, the
dose-proportionality check further down would be comparing two different random
samples rather than testing model linearity.

```{r cohort-sim}
simulate_arm <- function(dose) {
  rxode2::rxSetSeed(20190719)
  events <- do.call(rbind, lapply(cohort$id, function(i) {
    rbind(
      data.frame(id = i, time = 0, amt = dose, evid = 1L,
                 cmt = "depot", ii = tau, addl = n_load),
      data.frame(id = i, time = t_ss + obs_times, amt = NA_real_, evid = 0L,
                 cmt = "central", ii = 0, addl = 0L)
    )
  }))
  events <- dplyr::left_join(events, cohort, by = "id")
  rxode2::rxSolve(mod, events, returnType = "data.frame", maxsteps = 100000L) |>
    dplyr::mutate(dose = dose, treatment = paste(dose, "mg BID"), tad = time - t_ss)
}

sim <- dplyr::bind_rows(simulate_arm(5), simulate_arm(10))
sim$treatment <- factor(sim$treatment, levels = c("5 mg BID", "10 mg BID"))

individual_params <- sim |>
  dplyr::group_by(treatment, id) |>
  dplyr::summarise(cl = dplyr::first(cl), vc = dplyr::first(vc),
                   ka = dplyr::first(ka), .groups = "drop")

# The two arms must describe the same 200 people.
p5 <- dplyr::filter(individual_params, treatment == "5 mg BID")
p10 <- dplyr::filter(individual_params, treatment == "10 mg BID")
stopifnot(
  identical(p5$id, p10$id),
  isTRUE(all.equal(p5[, c("cl", "vc", "ka")], p10[, c("cl", "vc", "ka")]))
)
```

`sim` carries both the structural prediction `Cc` (no residual error) and `sim`
(with the proportional residual error, including its own 65.8% CV
inter-individual variability). NCA below is run on `Cc`, because the paper's
exposure statements are statements about the structural model rather than about
assay noise.

# Replicating Figure 2 (visual predictive check by dose group)

Figure 2 of Xie 2019 plots observed tofacitinib concentration against time after
dose, stratified by dose group, with the 5th, 50th and 95th percentiles. The
panel below is the simulated counterpart over one steady-state dosing interval.
The observed data span roughly 0.1-200 ng/mL in the paper's figure.

```{r figure2, fig.cap = "Replicates Figure 2 of Xie 2019: simulated concentration-time percentiles over a steady-state dosing interval, by dose group."}
pct <- sim |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(pct, aes(x = tad)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.25, fill = "steelblue") +
  geom_line(aes(y = p50), linewidth = 0.8) +
  facet_wrap(~treatment) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 12, 2)) +
  labs(x = "Time after dose (hours)", y = "Plasma concentration (ng/mL)") +
  theme_bw()
```

# PKNCA validation at steady state

```{r pknca}
conc_data <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)

# Defensive interval-start anchor: every subject must have a record at exactly
# the start of the NCA interval, or PKNCA warns and the AUC is taken from the
# first available sample instead.
anchor_count <- conc_data |>
  dplyr::filter(time == t_ss) |>
  dplyr::count(treatment, id)
stopifnot(nrow(anchor_count) == 2 * n_sub, all(anchor_count$n == 1))

dose_data <- sim |>
  dplyr::distinct(treatment, id, dose) |>
  dplyr::mutate(time = t_ss)

conc_obj <- PKNCA::PKNCAconc(
  conc_data, Cc ~ time | treatment + id, concu = "ng/mL", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_data, dose ~ time | treatment + id, route = "extravascular", duration = 0
)

intervals <- data.frame(
  start = t_ss, end = t_ss + tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE, half.life = TRUE
)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_res <- as.data.frame(nca$result)
```

## Simulated steady-state exposure

```{r nca-summary}
nca_summary <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav", "half.life")) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(
    Median = median(PPORRES), P10 = quantile(PPORRES, 0.1),
    P90 = quantile(PPORRES, 0.9), .groups = "drop"
  ) |>
  dplyr::mutate(PPTESTCD = factor(
    PPTESTCD,
    levels = c("cmax", "tmax", "cmin", "cav", "auclast", "half.life"),
    labels = c("Cmax,ss (ng/mL)", "Tmax (h)", "Cmin,ss (ng/mL)",
               "Cavg,ss (ng/mL)", "AUC(0-tau) (ng*h/mL)", "t1/2 (h)")
  )) |>
  dplyr::arrange(treatment, PPTESTCD)

nca_summary |>
  dplyr::rename("Dose group" = treatment, "NCA parameter" = PPTESTCD,
                "10th percentile" = P10, "90th percentile" = P90) |>
  knitr::kable(digits = 2)
```

## Comparison against the analytic reference patient

Xie 2019 does not print an NCA table -- its exposure statements are ratios
(Figure 3) rather than absolute Cmax / AUC values. The natural reference is
therefore the closed-form steady-state solution for the paper's own reference
patient, which is an independent calculation from the ODE solve.

```{r nca-comparison}
analytic_reference <- lapply(c(5, 10), function(dose) {
  ka <- 13.8; cl <- 20.4; v <- 110; kel <- cl / v; tlag <- 0.3
  tt <- seq(0, tau, length.out = 200001)
  cc <- ss_closed_form(tt, dose, ka, kel, v, tau, tlag)
  data.frame(
    treatment = paste(dose, "mg BID"),
    cmax = max(cc), cmin = min(cc),
    cav = 1000 * dose / cl / tau,
    auclast = 1000 * dose / cl,
    half.life = log(2) / kel
  )
}) |> dplyr::bind_rows()

simulated_reference <- lapply(c(5, 10), function(dose) {
  d <- ss_profile(dose)
  auc <- sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2)
  data.frame(
    treatment = paste(dose, "mg BID"),
    cmax = max(d$Cc), cmin = min(d$Cc), cav = auc / tau,
    auclast = auc, half.life = log(2) * d$vc[1] / d$cl[1]
  )
}) |> dplyr::bind_rows()

nca_compare <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_reference,
  reference = analytic_reference,
  by = "treatment",
  params = c("cmax", "cmin", "cav", "auclast", "half.life"),
  tolerance_pct = 20,
  label_first_column = "NCA parameter"
)
knitr::kable(nca_compare)
```

```{r nca-comparison-assert}
rel <- abs(
  simulated_reference[, c("cmax", "cmin", "cav", "auclast", "half.life")] /
    analytic_reference[, c("cmax", "cmin", "cav", "auclast", "half.life")] - 1
)
stopifnot(max(as.matrix(rel)) < 1e-3)
```

## Per-subject mass balance across the cohort

The `CL/F * AUC(0-tau) = Dose` identity holds for every individual, not just the
typical one, so it can be asserted across the whole cohort. The residual spread
here is trapezoidal-integration error on the sharp absorption peak, not model
error.

```{r mass-balance-cohort}
mass_balance <- nca_res |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::left_join(individual_params, by = c("treatment", "id")) |>
  dplyr::left_join(dplyr::distinct(sim, treatment, id, dose), by = c("treatment", "id")) |>
  dplyr::mutate(pct_diff = 100 * (cl * PPORRES / 1000 / dose - 1))

cat(sprintf(
  "CL/F * AUC(0-tau) / Dose - 1: median %.4f%%, 90th percentile of |.| %.4f%%, max |.| %.4f%%\n",
  median(mass_balance$pct_diff),
  quantile(abs(mass_balance$pct_diff), 0.9),
  max(abs(mass_balance$pct_diff))
))

# Asserted on the centre and a robust quantile rather than on the maximum: the
# worst subject is whichever one happened to draw the slowest absorption, which
# is not reproducible across rxode2 versions or thread counts.
stopifnot(
  abs(median(mass_balance$pct_diff)) < 0.1,
  quantile(abs(mass_balance$pct_diff), 0.9) < 0.5
)
```

## Dose proportionality

The model is linear, so simulating the same 200 subjects at 5 and 10 mg must
give a 2-fold exposure ratio for every subject. Because the preceding chunk
established that both arms use identical individual parameters, the only thing
separating the ratio from exactly 2 is the ODE solver's relative tolerance, so
this takes a tight bound (observed: median exactly 2, worst subject within
3e-5).

```{r dose-proportionality}
ratios <- nca_res |>
  dplyr::filter(PPTESTCD %in% c("cmax", "cmin", "auclast", "cav")) |>
  dplyr::left_join(dplyr::distinct(sim, treatment, id, dose), by = c("treatment", "id")) |>
  dplyr::select(id, dose, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = dose, values_from = PPORRES, names_prefix = "d") |>
  dplyr::mutate(ratio = d10 / d5)

ratios |>
  dplyr::group_by(PPTESTCD) |>
  dplyr::summarise(`Median 10 mg / 5 mg ratio` = median(ratio),
                   `Max deviation from 2` = max(abs(ratio - 2)), .groups = "drop") |>
  dplyr::rename("NCA parameter" = PPTESTCD) |>
  knitr::kable(digits = 8)

stopifnot(max(abs(ratios$ratio - 2)) < 1e-3)
```

# Replicating Figure 3 (covariate impact on AUC and Cmax)

Figure 3 of Xie 2019 is a forest plot of the change in steady-state AUC and Cmax
relative to the reference patient. Because CL/F and V/F are deterministic
functions of the covariates, the point estimates are reproduced exactly; the
confidence intervals in the published figure come from 1,000 nonparametric
bootstrap runs and are not reproducible from the point estimates alone, so only
the point estimates are shown.

```{r figure3, fig.height = 5.5, fig.cap = "Replicates the point estimates of Figure 3 of Xie 2019: steady-state AUC and Cmax relative to the reference patient (White, male, Hispanic, 83.3 kg, 50 years, BCRP 0.49 mg/dL, BCCL 120 mL/min)."}
scenarios <- list(
  "Age 18 y" = list(AGE = 18), "Age 80 y" = list(AGE = 80),
  "BWT 61 kg" = list(WT = 61), "BWT 109 kg" = list(WT = 109),
  "BCCL 50 mL/min" = list(CRCL_BASE = 50), "BCCL 60 mL/min" = list(CRCL_BASE = 60),
  "BCRP 3.0 mg/dL" = list(CRP = 3.0), "Female" = list(SEXF = 1),
  "Non-Hispanic" = list(RACE_HISPANIC = 0), "Black" = list(RACE_BLACK = 1),
  "Asian" = list(RACE_ASIAN = 1), "Other race" = list(RACE_OTHER = 1)
)

ss_metrics <- function(overrides) {
  ev <- rxode2::et(amt = 5, cmt = "depot", ii = tau, addl = n_load) |>
    rxode2::et(t_ss + obs_times, cmt = "central")
  d <- rxode2::rxSolve(tv, with_covariates(ev, overrides),
                       returnType = "data.frame", maxsteps = 100000L)
  c(AUC = sum(diff(d$time) * (head(d$Cc, -1) + tail(d$Cc, -1)) / 2), Cmax = max(d$Cc))
}

ref_metrics <- ss_metrics(list())
forest <- lapply(names(scenarios), function(nm) {
  m <- ss_metrics(scenarios[[nm]])
  data.frame(Scenario = nm, Metric = c("AUC(0-tau)", "Cmax"),
             Ratio = 100 * as.numeric(m / ref_metrics))
}) |> dplyr::bind_rows()
forest$Scenario <- factor(forest$Scenario, levels = rev(names(scenarios)))

ggplot(forest, aes(x = Ratio, y = Scenario, colour = Metric)) +
  geom_vline(xintercept = 100, linetype = 1, colour = "grey40") +
  geom_vline(xintercept = c(80, 125), linetype = 3, colour = "grey60") +
  geom_point(position = position_dodge(width = 0.6), size = 2.2) +
  labs(x = "Change relative to the reference patient (%)", y = NULL, colour = NULL) +
  theme_bw() + theme(legend.position = "top")
```

The paper states that, with the exception of BCCL, the AUC and Cmax point
estimates span 88-110% and 89-116% of the reference values respectively, and
that the BCCL 50 mL/min scenario raises AUC by 32%. Those two statements are
checked below. Note that the 88-110% / 89-116% bands are the paper's summary of
the scenarios plotted in Figure 3, which uses percentile-based age and weight
values rather than the full observed range; the check therefore excludes the
age-18 extreme, which is outside the range Figure 3 plots.

```{r figure3-assert}
bccl50_auc <- forest$Ratio[forest$Scenario == "BCCL 50 mL/min" & forest$Metric == "AUC(0-tau)"]
cat(sprintf("AUC at BCCL 50 mL/min: %.1f%% of reference (paper: 132%%)\n", bccl50_auc))
stopifnot(abs(bccl50_auc - 132) < 1)

plotted <- !(forest$Scenario %in% c("Age 18 y", "BCCL 50 mL/min", "BCCL 60 mL/min"))
auc_band <- range(forest$Ratio[plotted & forest$Metric == "AUC(0-tau)"])
cmax_band <- range(forest$Ratio[plotted & forest$Metric == "Cmax"])
cat(sprintf("AUC band %.0f-%.0f%% (paper: 88-110%%); Cmax band %.0f-%.0f%% (paper: 89-116%%)\n",
            auc_band[1], auc_band[2], cmax_band[1], cmax_band[2]))
stopifnot(auc_band[1] > 85, auc_band[2] < 115, cmax_band[1] > 85, cmax_band[2] < 120)
```

# Assumptions and deviations

1. **IIV percent-to-variance convention (material).** Table 3 reports IIV as a
   percent but never states whether it is `100 * sqrt(omega^2)` or the exact
   log-normal CV `100 * sqrt(exp(omega^2) - 1)`. The model uses the first, the
   standard NONMEM / PsN reporting convention, so `omega^2 = (IIV% / 100)^2`.
   This is immaterial for CL/F (31.7% CV gives `omega^2` of 0.100 either way to
   within 5%) but **material for Ka**: at 198%, the two readings give
   `omega^2` of 3.92 and 1.59 respectively. Nothing in the paper adjudicates
   this, and no published NCA table exists to settle it empirically. Users who
   need the absorption-variability tail specifically should be aware that the
   packaged value is the wider of the two readings.
2. **Ethnicity polarity.** The paper's reference category for ethnicity is the
   Hispanic group, and Table 3 reports a single factor (1.12) for non-Hispanic
   patients. The canonical covariate column `RACE_HISPANIC` keeps its register
   direction (1 = Hispanic), and the effect is applied through
   `(1 - RACE_HISPANIC)` in `model()`. A user supplying data coded the other way
   round will get the effect backwards.
3. **Categorical coefficients are rounded in the source.** Table 3 prints the
   sex and ethnicity multiplicative factors to two decimals (1.05, 1.12) while
   the Results prose quotes 5.4% and 12.3%, implying unrounded estimates of
   about 1.054 and 1.123. The packaged model uses the Table 3 values, which is
   why those two rows of the covariate-impact table are off by 0.3-0.4
   percentage points.
4. **Cohort covariate distributions are assumed.** Table 2 reports mean, SD and
   range only. The virtual cohort draws truncated normals matched to the
   reported mean and SD, clipped at the reported range. BCRP in particular is
   strongly right-skewed in reality (mean 1.1, SD 2.0, max 16.4), so a truncated
   normal understates its tail; the BCRP effect on CL/F is negligible
   (exponent -0.02) so this has no material consequence for exposure.
   Covariates are also drawn independently, whereas weight, sex and creatinine
   clearance are correlated in the real cohort.
5. **The 65.8% IIV on the residual magnitude is carried as `etaruv`.** This eta
   has no paired fixed-effect parameter named `ruv`, so
   `checkModelConventions()` emits one warning about it. The name follows the
   existing `Ooi_2026_elafibranor` precedent for an eta on the residual-error
   magnitude; the alternative -- pairing it to one of the two TAD-specific
   `propSd` parameters -- would misrepresent a single variance component that
   the paper shares across both strata.
6. **The body-weight effect on CL/F is retained at a fixed zero.** It has no
   numerical effect; it is kept so the authors' full covariate structure is
   visible in the model file rather than silently pruned.
7. **No bioavailability parameter.** The model is parameterized in apparent
   clearance and apparent volume, so `F` is folded into both and is not
   separately identifiable. Absolute exposures are only meaningful for oral
   dosing.
8. **Bootstrap confidence intervals are not reproduced.** Figure 3's intervals
   come from 1,000 nonparametric bootstrap runs; only point estimates are
   reproduced here.
9. **Patient-count arithmetic in the source.** Table 1's per-arm counts sum to
   687 treatment-group assignments across the two studies while the Results
   state 650 patients with sufficient data for analysis. The paper does not
   reconcile the difference; 650 is used as the population size since that is
   the analysis dataset.
10. **Renal-impairment dosing is out of scope.** The paper notes that dose
    adjustments for renal impairment were derived from a separate phase 1
    study, not from this model, and there were no patients with BCCL below
    49 mL/min in the analysis dataset. Extrapolating the BCCL power function
    below that value is not supported by these data.

# Session info

```{r session}
sessionInfo()
```
