---
title: "Midazolam (Xie 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Midazolam (Xie 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-load}
mod <- rxode2::rxode(readModelDb("Xie_2025_midazolam"))
mod_typical <- rxode2::zeroRe(mod)
```

* Citation: `r mod$reference`
* Description: `r mod$description`
* Article: <https://doi.org/10.2147/DDDT.S495647>

Xie 2025 is the first prospective population pharmacokinetic study of
midazolam in mechanically ventilated Chinese ICU patients, and the first to
report an effect of the *NR1I2* (PXR) rs2461817 polymorphism on midazolam
clearance. Midazolam and its active metabolite 1-hydroxymidazolam (1-OH-MDZ)
were fitted simultaneously, each with a one-compartment model, with the
molar fraction of midazolam converted to 1-OH-MDZ fixed at 0.6.

# Population

Sixty-one adults (of 69 screened) admitted to the ICU of Fujian Medical
University Union Hospital between April 2020 and August 2022, all
mechanically ventilated for at least 24 h and sedated with a continuous
intravenous midazolam infusion titrated to a target Richmond
Agitation-Sedation Scale score. Baseline characteristics come from Xie 2025
Table 1: median weight 62.0 kg (38.0-87.6), median age 67 years (29-90),
68.9% male, median APACHE II 16 (2-24), median AST 37.0 IU/L (10.0-362.0),
median albumin 31.7 g/L, median creatinine clearance 70.77 mL/min. Patients
with hepatic coma or cirrhosis, with neurological conditions preventing
sedation assessment, or with haemodynamic instability requiring frequent dose
changes were excluded, which is why the cohort is described in the Discussion
as "mostly postoperative patients in relatively good health".

Two hundred and thirty-seven paired midazolam / 1-OH-MDZ arterial plasma
samples were drawn over the first 0-24 h of infusion (pre-dose and in the
windows 0-0.5, 1-3, 4-6 and 10-12 h) and assayed by LC-MS/MS with LLOQs of
0.5 ng/mL (midazolam) and 0.25 ng/mL (1-OH-MDZ). Observed concentrations
spanned 1.53-1622.36 ng/mL for midazolam and 0.28-166.19 ng/mL for 1-OH-MDZ.
Initial infusion rates were 2-6 mg/h.

The same information is available programmatically via the model's
`population` metadata:

```{r population}
str(mod$population, max.level = 1)
```

# Structural model

Xie 2025 Figure 2 draws the structure explicitly. Midazolam leaves its
central compartment at the total rate `CL_MDZ / V_MDZ`, split into a
non-metabolic arm carrying the fraction `1 - F_MET` and a metabolic arm
carrying `F_MET`, which feeds a separate one-compartment 1-OH-MDZ model
eliminated at `CL_1-OH-MDZ / V_1-OH-MDZ`.

The paper's Methods state that "the dose, infusion rate, and plasma
concentrations of MDZ (Mw = 325.77 g/mol) and 1-OH-MDZ (Mw = 341.77 g/mol)
were converted to molar equivalents", so `F_MET = 0.6` is a **molar**
fraction. This model file carries amounts in mg, so the metabolite formation
flux is scaled by the molecular-weight ratio
`0.6 * 341.77 / 325.77 = 0.62947` mg of 1-OH-MDZ formed per mg of midazolam
metabolised. Retaining that 4.9% correction is what reproduces the published
Figure 7 plateaus (see below); dropping it would shift them low by the same
factor.

```{r model-body}
mod
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Xie_2025_midazolam.R` carries an in-file comment
pointing at its origin. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: 1-cmt midazolam + 1-cmt 1-OH-MDZ, metabolic / non-metabolic split | n/a | Figure 2 (structural diagram); Methods "Structural Model and Choice of Statistical Model" |
| `lcl` (CL midazolam) | 22.6 L/h | Table 2, Final Model, "CL MDZ" (RSE 16%) |
| `lvc` (V midazolam) | 16.00 L | Table 2, Final Model, "V MDZ" (RSE 42%) |
| `lcl_1ohm` (apparent CL 1-OH-MDZ) | 67.10 L/h | Table 2, Final Model, "CL 1-OH-MDZ" (RSE 10%) |
| `lvc_1ohm` (apparent V 1-OH-MDZ) | 86.20 L | Table 2, Final Model, "V 1-OH-MDZ" (RSE 35%) |
| `fm_1ohm` (F_MET, molar) | 0.6, fixed | Table 2 ("F MET", RSE reported as "/"); Methods "Structural Model" ("fixed at 0.6 based on prior literature", ref 23 = Seng et al) |
| Molecular weights 325.77 / 341.77 g/mol | n/a | Methods "Structural Model" |
| `e_ast_cl` (AST on CL MDZ) | -0.397 | Table 2, Final Model, "ASTon CL MDZ" (RSE 47%) |
| `e_nr1i2_hom_cl` (rs2461817 on CL MDZ) | -0.405 | Table 2, Final Model, "rs2461817 on CL MDZ" (RSE 29%) |
| `e_wt_cl_1ohm` (BW on CL 1-OH-MDZ) | 1.58 | Table 2, Final Model, "BWon CL 1-OH-MDZ" (RSE 30%) |
| AST covariate form and reference (37 IU/L) | power | **Not printed by the paper** - back-solved from Results "Simulations" (see below); reference is the Table 1 AST median |
| BW covariate form and reference (62 kg) | power | **Not printed by the paper** - back-solved from Results "Simulations"; reference is the Table 1 weight median |
| rs2461817 covariate form | fractional `(1 + theta * GENE)` | **Not printed by the paper** - back-solved from Results "Simulations"; the exponential alternative is excluded numerically |
| `etalcl` | omega^2 = 0.302203 | Table 2, Final Model IIV "CL MDZ (CV%)" = 59.40; Table 2 footnote gives CV% = sqrt(exp(OMEGA) - 1) x 100 |
| `etalcl_1ohm` | omega^2 = 0.287379 | Table 2, Final Model IIV "CL 1-OH-MDZ (CV%)" = 57.70; same footnote |
| `propSd` | 0.400 | Table 2, Final Model RUV "Proportional (CV%)" = 40.00 |
| `addSd` | 3.43 ng/mL | Table 2, Final Model RUV "Additive (ng/mL)" = 3.43 |
| `addSd_1ohm` | 0.6 ng/mL | Table 2, Final Model RUV 1-OH-MDZ "Additive (ng/mL)" = 0.6 |

## Recovering the unprinted covariate equations

Xie 2025 reports the three covariate coefficients in Table 2 but never writes
the covariate equations, and no supplement carrying them is available. All
three functional forms and both reference values are nevertheless pinned
exactly, because the paper's Results "Simulations" section quotes the
resulting clearances to three significant figures. The table below is the
arithmetic; it is reproduced as a live check further down.

| Scenario | Paper's reported CL | Candidate form | Value |
|---|---|---|---|
| AST 22 IU/L | 27.8 L/h | `22.6 * (22/37)^-0.397` | 27.78 |
| AST 60 IU/L | 18.7 L/h | `22.6 * (60/37)^-0.397` | 18.65 |
| rs2461817 homozygous | 13.4 L/h | `22.6 * (1 - 0.405)` | 13.45 |
| rs2461817 homozygous (rejected) | 13.4 L/h | `22.6 * exp(-0.405)` | **15.07** |
| BW 54 kg | 53.9 L/h | `67.1 * (54/62)^1.58` | 53.94 |
| BW 65 kg | 72.3 L/h | `67.1 * (65/62)^1.58` | 72.30 |

The genotype effect is therefore a fractional shift, not the more common
exponential factor: the exponential form misses the paper's own number by
12%, while the fractional form matches it to the last printed digit.

# Virtual cohort

Original observed data are not publicly available. Two virtual cohorts are
used below: a small deterministic one that reproduces the paper's typical-value
simulation figures (Figures 5-7), and a 200-subject stochastic one for the
VPC-style figure (Figure 4).

```{r constants}
# Xie 2025 Methods "Structural Model": molecular weights used for the
# molar-equivalent conversion. Needed here only to plot Figure 4 on the
# paper's nmol/mL axis.
MW_MDZ  <- 325.77
MW_1OHM <- 341.77

# Xie 2025 Simulation section: "a loading dose of 3 mg and a maintenance dose
# of 4 mg/h ... based on the median weight (62 kg) of the study population",
# simulated "for 10 days".
LOADING_MG   <- 3
INFUSION_MGH <- 4
INFUSION_H   <- 240
```

```{r cohort-helpers}
# Build a plain data frame rather than an rxEt object: covariate columns
# assigned onto an rxEt are silently dropped by rxode2.
#
# The model declares two endpoints (Cc and Cc_1ohm), so rxode2 requires
# observation rows to be addressed by endpoint via cmt + dvid. That is the
# documented idiom for DECLARED multi-endpoint models; it is not the
# "cmt = <observable>" antipattern, which applies to models whose
# observables are plain algebraic intermediates.
make_events <- function(subjects, times, doses) {
  dose_rows <- subjects |>
    tidyr::crossing(doses) |>
    dplyr::mutate(evid = 1L, cmt = "central", dvid = NA_integer_,
                  analyte = NA_character_)

  obs_rows <- subjects |>
    tidyr::crossing(
      tibble::tibble(
        time    = rep(times, 2L),
        cmt     = rep(c("Cc", "Cc_1ohm"), each = length(times)),
        dvid    = rep(c(1L, 2L), each = length(times)),
        analyte = rep(c("midazolam", "1-OH-midazolam"), each = length(times))
      )
    ) |>
    dplyr::mutate(evid = 0L, amt = NA_real_, rate = NA_real_)

  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    as.data.frame()
}

# The paper's standard regimen: 3 mg IV loading dose, then 4 mg/h for 10 days.
paper_doses <- tibble::tibble(
  time = c(0, 0),
  amt  = c(LOADING_MG, INFUSION_MGH * INFUSION_H),
  rate = c(0, INFUSION_MGH)
)
```

```{r cohort-typical}
# One subject per published simulation arm. Figures 5 and 6 vary AST and
# genotype at the median weight; Figure 7 varies weight.
typical_arms <- tibble::tribble(
  ~id, ~figure,     ~arm,            ~AST, ~WT,  ~SNP_NR1I2_RS2461817_HOM,
  1L,  "Figure 5",  "AST 22 IU/L",     22,  62,  0,
  2L,  "Figure 5",  "AST 37 IU/L",     37,  62,  0,
  3L,  "Figure 5",  "AST 60 IU/L",     60,  62,  0,
  4L,  "Figure 6",  "GENE = 0",        37,  62,  0,
  5L,  "Figure 6",  "GENE = 1",        37,  62,  1,
  6L,  "Figure 7",  "BW 54 kg",        37,  54,  0,
  7L,  "Figure 7",  "BW 62 kg",        37,  62,  0,
  8L,  "Figure 7",  "BW 65 kg",        37,  65,  0
)

events_typical <- make_events(
  subjects = typical_arms,
  times    = seq(0, 250, by = 1),
  doses    = paper_doses
)
stopifnot(!anyDuplicated(unique(events_typical[, c("id", "time", "evid", "cmt")])))
```

```{r cohort-vpc}
set.seed(20250314)
n_vpc <- 200L

# Weight: truncated normal centred on the Table 1 median (62 kg) and clipped
# to the reported range (38.0-87.6 kg).
wt_vpc <- pmin(pmax(rnorm(n_vpc, mean = 62, sd = 10), 38), 87.6)

# AST: right-skewed, so a log-normal is used. The log-scale SD is chosen so
# that the 25th and 75th percentiles land on the paper's simulation strata
# (22 and 60 IU/L): log(60/37) / qnorm(0.75) = 0.717. Clipped to the Table 1
# range (10-362 IU/L).
ast_vpc <- pmin(pmax(rlnorm(n_vpc, meanlog = log(37), sdlog = 0.717), 10), 362)

# Genotype: Xie 2025 reports its rs2461817 genotype counts only in
# Supplementary Table 1, which is not available. See "Assumptions and
# deviations" - Hardy-Weinberg with a variant-allele frequency of 0.45.
gene_vpc <- rbinom(n_vpc, size = 1, prob = 0.45^2)

# Infusion rate: Results "Patients and Samples" - "The initial infusion rate
# for all patients ranged from 2 to 6 mg/h."
rate_vpc <- runif(n_vpc, 2, 6)

vpc_subjects <- tibble::tibble(
  id  = seq_len(n_vpc),
  AST = ast_vpc,
  WT  = wt_vpc,
  SNP_NR1I2_RS2461817_HOM = gene_vpc,
  infusion_rate = rate_vpc
)

# Continuous infusion only (no loading dose) over the paper's 0-24 h
# observation window, matching how the patients were actually dosed.
events_vpc <- vpc_subjects |>
  dplyr::mutate(time = 0, amt = infusion_rate * 24, rate = infusion_rate) |>
  dplyr::select(id, AST, WT, SNP_NR1I2_RS2461817_HOM, time, amt, rate) |>
  dplyr::mutate(evid = 1L, cmt = "central", dvid = NA_integer_,
                analyte = NA_character_) |>
  dplyr::bind_rows(
    vpc_subjects |>
      dplyr::select(id, AST, WT, SNP_NR1I2_RS2461817_HOM) |>
      tidyr::crossing(
        tibble::tibble(
          time    = rep(seq(0, 24, by = 0.5), 2L),
          cmt     = rep(c("Cc", "Cc_1ohm"), each = length(seq(0, 24, by = 0.5))),
          dvid    = rep(c(1L, 2L), each = length(seq(0, 24, by = 0.5))),
          analyte = rep(c("midazolam", "1-OH-midazolam"),
                        each = length(seq(0, 24, by = 0.5)))
        )
      ) |>
      dplyr::mutate(evid = 0L, amt = NA_real_, rate = NA_real_)
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

stopifnot(!anyDuplicated(unique(events_vpc[, c("id", "time", "evid", "cmt")])))
```

# Simulation

`useLinCmt = FALSE` is passed to every solve: rxode2's automatic ODE ->
`linCmt()` conversion corrupts the dvid -> cmt mapping for multi-output
models like this one.

```{r simulate-typical}
sim_typical <- rxode2::rxSolve(
  mod_typical, events = events_typical,
  keep = c("figure", "arm", "analyte"),
  useLinCmt = FALSE, returnType = "data.frame"
)

# rxSolve silently drops subjects on some event-table shapes; assert instead
# of assuming.
stopifnot(dplyr::n_distinct(sim_typical$id) == nrow(typical_arms))
```

```{r simulate-vpc}
sim_vpc <- rxode2::rxSolve(
  mod, events = events_vpc,
  keep = c("analyte", "infusion_rate"),
  useLinCmt = FALSE, returnType = "data.frame"
)

stopifnot(dplyr::n_distinct(sim_vpc$id) == n_vpc)
# Guard that IIV was actually applied - a check that cannot go red is worse
# than no check.
stopifnot(dplyr::n_distinct(round(sim_vpc$cl, 6)) > 1L)
```

# Replicate published figures

## Figure 5 - midazolam by AST stratum

```{r figure-5}
# Replicates Figure 5 of Xie 2025: typical midazolam profiles after a 3 mg
# loading dose plus 4 mg/h infusion for 10 days, by AST stratum.
sim_typical |>
  dplyr::filter(figure == "Figure 5", analyte == "midazolam") |>
  ggplot(aes(time, Cc, linetype = arm)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 300)) +
  labs(x = "Time (h)", y = "Midazolam concentration (ng/mL)",
       linetype = NULL,
       title = "Figure 5 - midazolam by AST level",
       caption = "Replicates Figure 5 of Xie 2025.")
```

## Figure 6 - midazolam by *NR1I2* rs2461817 genotype

```{r figure-6}
# Replicates Figure 6 of Xie 2025. GENE = 1 is the homozygous mutant;
# GENE = 0 pools wild-type homozygotes and heterozygotes.
sim_typical |>
  dplyr::filter(figure == "Figure 6", analyte == "midazolam") |>
  ggplot(aes(time, Cc, linetype = arm)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 400)) +
  labs(x = "Time (h)", y = "Midazolam concentration (ng/mL)",
       linetype = NULL,
       title = "Figure 6 - midazolam by NR1I2 rs2461817 genotype",
       caption = "Replicates Figure 6 of Xie 2025.")
```

## Figure 7 - 1-OH-midazolam by body weight

```{r figure-7}
# Replicates Figure 7 of Xie 2025: typical 1-OH-MDZ profiles under the same
# midazolam regimen, by body weight.
sim_typical |>
  dplyr::filter(figure == "Figure 7", analyte == "1-OH-midazolam") |>
  ggplot(aes(time, Cc_1ohm, linetype = arm)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 50)) +
  labs(x = "Time (h)", y = "1-hydroxymidazolam concentration (ng/mL)",
       linetype = NULL,
       title = "Figure 7 - 1-OH-midazolam by body weight",
       caption = "Replicates Figure 7 of Xie 2025.")
```

## Steady-state plateaus and clearances

Under a constant infusion the plateau is analytic, which makes this an exact
falsifier rather than an eyeball comparison. For midazolam,
`Css = rate / CL`; for the metabolite, all midazolam is eventually cleared, so
the metabolite formation rate at steady state is `F_MET * (MW ratio) * rate`
independently of midazolam clearance, giving
`Css_met = F_MET * (MW_1OHM / MW_MDZ) * rate / CL_1-OH-MDZ`. That independence
is why Xie 2025's Figure 7 caption does not need to state an AST level or a
genotype.

```{r plateau-table}
plateau <- sim_typical |>
  dplyr::filter(dplyr::near(time, INFUSION_H), analyte == "midazolam") |>
  dplyr::transmute(
    figure, arm,
    cl_sim      = cl,
    cl_1ohm_sim = cl_1ohm,
    css_sim     = Cc,
    css_1ohm_sim = Cc_1ohm,
    css_analytic = INFUSION_MGH / cl * 1000,
    css_1ohm_analytic = 0.6 * (MW_1OHM / MW_MDZ) * INFUSION_MGH / cl_1ohm * 1000
  )

# The simulated plateau must match the closed-form identity to solver
# precision, for both analytes.
stopifnot(
  max(abs(plateau$css_sim / plateau$css_analytic - 1)) < 1e-3,
  max(abs(plateau$css_1ohm_sim / plateau$css_1ohm_analytic - 1)) < 1e-3
)

# Clearances quoted in Xie 2025 Results "Simulations"; plateau concentrations
# read off Figures 5-7 (graphical, so approximate).
published_plateau <- tibble::tribble(
  ~arm,           ~cl_pub, ~css_pub, ~cl_1ohm_pub, ~css_1ohm_pub,
  "AST 22 IU/L",     27.8,      144,           NA,            NA,
  "AST 37 IU/L",     22.6,      177,           NA,            NA,
  "AST 60 IU/L",     18.7,      214,           NA,            NA,
  "GENE = 0",        22.6,      177,           NA,            NA,
  "GENE = 1",        13.4,      298,           NA,            NA,
  "BW 54 kg",          NA,       NA,         53.9,          46.5,
  "BW 62 kg",          NA,       NA,         67.1,          37.5,
  "BW 65 kg",          NA,       NA,         72.3,          35.0
)

plateau |>
  dplyr::left_join(published_plateau, by = "arm") |>
  dplyr::transmute(
    Figure = figure,
    Arm    = arm,
    `CL published (L/h)`      = dplyr::coalesce(cl_pub, cl_1ohm_pub),
    `CL simulated (L/h)`      = round(dplyr::if_else(is.na(cl_pub), cl_1ohm_sim, cl_sim), 2),
    `Css published (ng/mL)`   = dplyr::coalesce(css_pub, css_1ohm_pub),
    `Css simulated (ng/mL)`   = round(dplyr::if_else(is.na(css_pub), css_1ohm_sim, css_sim), 1)
  ) |>
  knitr::kable(
    caption = paste(
      "Steady-state clearance and plateau concentration by published",
      "simulation arm. Clearances are quoted verbatim in Xie 2025 Results",
      "'Simulations'; plateau concentrations are read off Figures 5-7 and",
      "are therefore graphical estimates. Figure 5 / 6 rows are midazolam;",
      "Figure 7 rows are 1-OH-midazolam."
    )
  )
```

Every simulated clearance reproduces the paper's quoted value to the printed
precision, and every plateau lands on the corresponding figure line.

## Figure 4 - VPC over the observation window

Xie 2025 plots Figure 4 on a nmol/mL axis (the units the model was fitted in),
so the simulated concentrations are converted back for comparison. The bands
are the 5th / 50th / 95th percentiles of the simulated concentrations
*including* residual variability (rxode2's `sim` column carries the residual
error; the `Cc` column does not).

```{r figure-4, fig.height = 6}
# Replicates Figure 4 of Xie 2025: VPC of midazolam (A) and 1-OH-MDZ (B)
# over the 0-24 h observation window.
vpc_bands <- sim_vpc |>
  dplyr::mutate(
    conc_molar = dplyr::if_else(analyte == "midazolam",
                                sim / MW_MDZ, sim / MW_1OHM)
  ) |>
  dplyr::filter(!is.na(conc_molar)) |>
  dplyr::group_by(analyte, time) |>
  dplyr::summarise(
    Q05 = quantile(conc_molar, 0.05),
    Q50 = quantile(conc_molar, 0.50),
    Q95 = quantile(conc_molar, 0.95),
    .groups = "drop"
  ) |>
  dplyr::mutate(analyte = factor(analyte,
                                 levels = c("midazolam", "1-OH-midazolam")))

ggplot(vpc_bands, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(colour = "firebrick") +
  facet_wrap(~analyte, ncol = 1, scales = "free_y") +
  labs(x = "Time after first dose (h)",
       y = "Concentration (nmol/mL)",
       title = "Figure 4 - simulated 5th / 50th / 95th percentiles",
       caption = "Replicates Figure 4 of Xie 2025 (axes in nmol/mL as published).")
```

The simulated midazolam band spans roughly 0-3 nmol/mL and the 1-OH-MDZ band
roughly 0-0.4 nmol/mL over 0-24 h, matching the ranges in Xie 2025 Figure 4A
and 4B.

# PKNCA validation

Xie 2025 reports no non-compartmental parameters, but it does report typical
clearances for seven covariate scenarios, and clearance is recoverable by NCA
as `Dose / AUC0-inf`. A single intravenous bolus with dense sampling is
simulated for each scenario so that PKNCA's `cl.obs` can be compared directly
against the published clearances. Doses are declared to PKNCA in ug so that
`cl.obs` (`ug / (h * ng/mL)`) is numerically L/h.

For the metabolite, the equivalent dose is the amount of 1-OH-MDZ actually
formed, `F_MET * (MW_1OHM / MW_MDZ) * dose`, so `cl.obs` recovers the
*apparent* metabolite clearance the paper reports.

```{r nca-simulate}
NCA_DOSE_MG <- 5
FM_MASS     <- 0.6 * MW_1OHM / MW_MDZ

nca_arms <- tibble::tribble(
  ~id,  ~analyte_group,     ~arm,            ~AST, ~WT,  ~SNP_NR1I2_RS2461817_HOM,
  1L,   "midazolam",        "AST 22 IU/L",     22,  62,  0,
  2L,   "midazolam",        "AST 37 IU/L",     37,  62,  0,
  3L,   "midazolam",        "AST 60 IU/L",     60,  62,  0,
  4L,   "midazolam",        "rs2461817 hom",   37,  62,  1,
  5L,   "1-OH-midazolam",   "BW 54 kg",        37,  54,  0,
  6L,   "1-OH-midazolam",   "BW 62 kg",        37,  62,  0,
  7L,   "1-OH-midazolam",   "BW 65 kg",        37,  65,  0
)

# Midazolam t1/2 is ~0.5 h and 1-OH-MDZ ~0.9 h, so 12 h is >= 13 terminal
# half-lives for both. Sampling is dense early to resolve the distribution
# phase, which is where a coarse grid would understate AUC.
nca_times <- c(0, seq(0.05, 2, by = 0.05), seq(2.25, 6, by = 0.25),
               seq(6.5, 12, by = 0.5))

events_nca <- make_events(
  subjects = nca_arms,
  times    = nca_times,
  doses    = tibble::tibble(time = 0, amt = NCA_DOSE_MG, rate = 0)
)

sim_nca_raw <- rxode2::rxSolve(
  mod_typical, events = events_nca,
  keep = c("analyte_group", "arm", "analyte"),
  useLinCmt = FALSE, returnType = "data.frame"
)

stopifnot(dplyr::n_distinct(sim_nca_raw$id) == nrow(nca_arms))
# Solver noise in the far tail would make PKNCA take log() of a negative
# value; assert the profiles stay non-negative.
stopifnot(min(sim_nca_raw$Cc) >= 0, min(sim_nca_raw$Cc_1ohm) >= 0)
```

## Midazolam

```{r nca-midazolam}
# Only !is.na(Cc) - never add time > 0 or Cc > 0, which drop the time-zero
# row PKNCA needs to anchor AUC0-*.
conc_mdz <- sim_nca_raw |>
  dplyr::filter(analyte == "midazolam",
                analyte_group == "midazolam",
                !is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

dose_mdz <- nca_arms |>
  dplyr::filter(analyte_group == "midazolam") |>
  dplyr::transmute(id, time = 0, amt = NCA_DOSE_MG * 1000, arm)

conc_obj_mdz <- PKNCA::PKNCAconc(conc_mdz, Cc ~ time | arm + id,
                                 concu = "ng/mL", timeu = "h")
dose_obj_mdz <- PKNCA::PKNCAdose(dose_mdz, amt ~ time | arm + id, doseu = "ug")

intervals_nca <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE,
  cl.obs     = TRUE
)

nca_mdz <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj_mdz, dose_obj_mdz, intervals = intervals_nca)
)
```

```{r nca-midazolam-compare}
# Xie 2025 Table 2 and Results "Simulations".
published_mdz <- tibble::tribble(
  ~arm,             ~cl.obs,
  "AST 22 IU/L",       27.8,
  "AST 37 IU/L",       22.6,
  "AST 60 IU/L",       18.7,
  "rs2461817 hom",     13.4
)

cmp_mdz <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_mdz,
  reference = published_mdz,
  by        = "arm",
  units     = c(cl.obs = "L/h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_mdz,
  caption = paste(
    "Midazolam clearance recovered by NCA (Dose / AUC0-inf) versus the",
    "typical clearances reported in Xie 2025. * differs from reference by",
    ">20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

## 1-OH-midazolam

```{r nca-metabolite}
conc_met <- sim_nca_raw |>
  dplyr::filter(analyte == "1-OH-midazolam",
                analyte_group == "1-OH-midazolam",
                !is.na(Cc_1ohm)) |>
  dplyr::select(id, time, Cc = Cc_1ohm, arm)

# The metabolite's "dose" is the amount of 1-OH-MDZ formed from the
# midazolam dose, in ug.
dose_met <- nca_arms |>
  dplyr::filter(analyte_group == "1-OH-midazolam") |>
  dplyr::transmute(id, time = 0, amt = NCA_DOSE_MG * 1000 * FM_MASS, arm)

conc_obj_met <- PKNCA::PKNCAconc(conc_met, Cc ~ time | arm + id,
                                 concu = "ng/mL", timeu = "h")
dose_obj_met <- PKNCA::PKNCAdose(dose_met, amt ~ time | arm + id, doseu = "ug")

nca_met <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj_met, dose_obj_met, intervals = intervals_nca)
)
```

```{r nca-metabolite-compare}
published_met <- tibble::tribble(
  ~arm,        ~cl.obs,
  "BW 54 kg",     53.9,
  "BW 62 kg",     67.1,
  "BW 65 kg",     72.3
)

cmp_met <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_met,
  reference = published_met,
  by        = "arm",
  units     = c(cl.obs = "L/h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp_met,
  caption = paste(
    "1-OH-midazolam apparent clearance recovered by NCA versus the typical",
    "apparent clearances reported in Xie 2025. * differs from reference by",
    ">20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-assert}
# Tighten the visual comparison into a hard assertion: NCA must recover the
# model's own clearance for every arm to well within 1%.
recovered <- dplyr::bind_rows(
  as.data.frame(nca_mdz$result),
  as.data.frame(nca_met$result)
) |>
  dplyr::filter(PPTESTCD == "cl.obs") |>
  dplyr::select(arm, cl_nca = PPORRES)

model_cl <- sim_nca_raw |>
  dplyr::group_by(arm, analyte_group) |>
  dplyr::summarise(
    cl_model = dplyr::first(dplyr::if_else(analyte_group == "midazolam",
                                           cl, cl_1ohm)),
    .groups = "drop"
  )

check_cl <- dplyr::inner_join(recovered, model_cl, by = "arm")
stopifnot(
  nrow(check_cl) == nrow(nca_arms),
  max(abs(check_cl$cl_nca / check_cl$cl_model - 1)) < 0.01
)
check_cl |>
  dplyr::transmute(
    Arm                    = arm,
    `Model CL (L/h)`       = round(cl_model, 3),
    `NCA CL (L/h)`         = round(cl_nca, 3),
    `Relative difference`  = sprintf("%.4f%%", 100 * (cl_nca / cl_model - 1))
  ) |>
  knitr::kable(caption = "NCA-recovered clearance versus the model's own clearance.")
```

Both analytes' clearances are recovered exactly; no starred rows appear in
either comparison table.

# Assumptions and deviations

- **Covariate equations are not printed by the paper.** Xie 2025 reports the
  three covariate coefficients in Table 2 but never writes the covariate model
  equations, and no supplement carrying them is on disk. The power form and
  the 37 IU/L / 62 kg reference values, and the fractional (rather than
  exponential) form of the genotype effect, were recovered by back-solving the
  paper's own reported typical clearances in Results "Simulations". All six
  recoverable checks agree to the printed precision, and the exponential
  alternative for the genotype effect is excluded numerically (15.07 vs the
  reported 13.4 L/h). This is a reconstruction, not a transcription; a reader
  who obtains the authors' control stream should confirm it.

- **Concentration units of the additive residual errors.** The model was
  fitted in molar units (Methods; Figure 4 is plotted in nmol/mL), but Table 2
  labels both additive residual SDs "ng/mL". The ng/mL reading is adopted here
  because the alternative is dimensionally impossible: the entire observed
  midazolam range tops out near 5 nmol/mL and 1-OH-MDZ near 0.49 nmol/mL, so
  additive SDs of 3.43 and 0.6 could not be nmol/mL. In ng/mL they are 6.9x
  and 2.4x the respective LLOQs, which is the expected magnitude.

- **Mass rather than molar internal units.** The authors fitted in molar
  equivalents; this file carries amounts in mg and reports concentrations in
  ng/mL, which is the unit used by Table 1, Figure 3 and Figures 5-7. The two
  parameterisations are exactly equivalent once the metabolite formation flux
  is scaled by `MW_1OHM / MW_MDZ = 1.0491`, which the model file does. The
  Figure 7 plateaus confirm the correction is required: without it the 62 kg
  plateau would be 35.8 rather than the published ~37.5 ng/mL.

- **AST is treated as time-fixed.** The paper collected laboratory values from
  the electronic medical record and does not state whether AST was updated
  during the infusion. It is carried here as a baseline, time-fixed covariate.

- **rs2461817 genotype frequency in the virtual cohort.** Xie 2025 reports its
  genotype counts only in Supplementary Table 1, which is not available. The
  VPC cohort assumes Hardy-Weinberg equilibrium (which the paper states holds
  for this locus) with a variant-allele frequency of 0.45, giving roughly 20%
  homozygous mutants. This affects only the width of the simulated VPC band,
  not any model parameter.

- **AST and weight distributions in the virtual cohort.** Weight is drawn from
  a normal distribution centred on the Table 1 median and clipped to the
  reported range; AST from a log-normal whose quartiles are set to the paper's
  own 25th / 75th percentile strata (22 and 60 IU/L) and clipped to the
  reported range. The paper reports medians and ranges only, so the
  distributional shapes are assumptions.

- **No IIV on either volume.** Table 2 reports IIV only for the two
  clearances, so neither volume carries an eta. This is the paper's structure,
  not a simplification.

- **1-OH-MDZ CL and V are apparent values.** Because F_MET could not be
  estimated, the metabolite disposition parameters are identifiable only as a
  ratio (Discussion). They are conditional on F_MET = 0.6 and should not be
  reinterpreted as physiological clearance and volume.

- **VPC sampling grid.** The paper sampled sparsely (pre-dose plus windows
  0-0.5, 1-3, 4-6 and 10-12 h). The VPC above uses a regular 0.5 h grid over
  0-24 h so the percentile bands are smooth; the underlying model and
  covariate distributions are unchanged.

- **Covariates screened but not retained** (age, sex, APACHE II, albumin, ALT,
  total bilirubin, CRP, creatinine clearance, concomitant propofol and
  methylprednisolone) are recorded in the model file's
  `covariatesDataExcluded` metadata rather than `covariateData`, since none
  appears in the final model.

- **Supplementary material not on disk.** Supplementary Table 1 (genotype
  counts and Hardy-Weinberg tests), Supplementary Table 2 (the covariate
  screening trace) and Supplementary Figure 1 (additional goodness-of-fit
  plots) were not available. None contains a final model parameter; every
  value in the model file comes from Table 2 or the Methods / Results text of
  the main article.
