---
title: "Tepotinib and MSC2571109A (Xiong 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tepotinib and MSC2571109A (Xiong 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Xiong_2022_tepotinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Xiong_2022_tepotinib"))$description`
* Article: <https://doi.org/10.1007/s00280-022-04423-5> (open access, CC BY 4.0)
* Supplement: Electronic Supplementary Material ESM 1-13 at the article landing page.

Tepotinib is a once-daily oral MET kinase inhibitor approved for non-small
cell lung cancer (NSCLC) with MET exon 14 skipping. Xiong et al. pooled 12
studies and fitted the parent drug first, then fitted its major circulating
metabolite MSC2571109A with the individual tepotinib parameters held fixed
(sequential approach). Both analytes live in one model file here because the
metabolite is formed from the parent's clearance and cannot be simulated
without it.

# Population

The tepotinib analysis set (Table 2) had 613 participants and 10,788
concentrations: 438 patients with cancer (NSCLC, including MET exon 14
skipping NSCLC from the pivotal VISION study, hepatocellular carcinoma,
colorectal, renal cell, gastroesophageal and other solid tumours) and 175
healthy participants, some with mild or moderate hepatic impairment. Median
age was 58 years (range 18-89), median weight 72.0 kg (35.5-136), 174 (28%)
were female, and race was 59% Caucasian, 5% Japanese, 24% other East Asian,
3% African origin, 4% Hispanic and 6% other or missing. Doses ranged from 30
to 1400 mg/day as capsules (CF1 non-micronised, CF2 micronised) or tablets
(TF1, TF1* finely micronised, TF2, TF3), given once daily, three times weekly
or as a single dose, fasted or fed. The MSC2571109A set had 464 participants
and 7197 concentrations.

The same information is available programmatically via
`readModelDb("Xiong_2022_tepotinib")()$population`.

# Source trace

Values come from Table 3 of the article; the functional form of every
covariate relationship comes from ESM 13, which writes the final tepotinib
covariate model out in full. The metabolite covariate forms follow the
Methods' Formula 1 (power model on continuous covariates, normalised to the
reference value) and Formula 2 (fractional difference, `1 + theta`, for a
categorical covariate), the same forms ESM 13 uses for the parent.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structure: 2-cmt parent, zero-order input of duration D1 into the depot then first-order ka; 2-cmt metabolite formed from `fm * CL/Vc * central` | n/a | Results; Figure 1 |
| `lcl` (CL/F) | 20.4 L/h | Table 3; ESM 13 |
| `lvc` (Vc/F) | 1020 L | Table 3 (ESM 13 prints 1030; see Assumptions) |
| `lka` (ka) | 0.278 1/h | Table 3 (ESM 13 prints 1.47; see the ka check below) |
| `lq` (Q/F) | 1.32 L/h | Table 3; ESM 13 |
| `lvp` (Vp/F) | 1180 L | Table 3; ESM 13 |
| `ld1` (D1) | 4.09 h | Table 3; ESM 13 |
| `lfdepot` (relative F) | 1 (held constant) | Table 3 |
| `e_fasted_d1`, `e_hepimp_d1` | -0.370, -0.332 | Table 3; ESM 13 |
| `e_dose_fdepot` | -0.0412 per 100 mg, `1 + e/100 * (DOSE - 500)` | Table 3; ESM 13 |
| `e_fasted_fdepot`, `e_fed_highfat_fdepot` | -0.209, 0.320 | Table 3; ESM 13 |
| `e_form_cf1_fdepot`, `e_form_tf3_fdepot` | -0.656, 0.154 | Table 3; ESM 13 |
| `e_wt_fdepot` | -0.475, `(WT / 72)^e` | Table 3; ESM 13 |
| `e_hepimp_fdepot` | -0.0729 | Table 3; ESM 13 |
| `e_fasted_ka`, `e_form_cf1_ka`, `e_form_tf1_ka`, `e_form_tf1fine_ka` | -0.561, -0.442, 0.305, 0.674 | Table 3; ESM 13 |
| `e_crcl_cl` | 0.199, `(eGFR / 97.28)^e` | Table 3; ESM 13 |
| `e_tumtp_hcc_cl`, `e_tumtp_crc_cl` | 0.130, -0.281 | Table 3; ESM 13 |
| `e_conmed_opioid_cl`, `e_study_0028_cl` | -0.167, -0.115 | Table 3; ESM 13 |
| `e_inr_q`, `e_alb_q` | 3.81 `(INR / 1.06)^e`, 4.14 `(ALB / 40)^e` | Table 3; ESM 13 |
| `e_age_vc`, `e_tumtp_nsclc_vc` | 0.219 `(AGE / 59)^e`, -0.232 | Table 3; ESM 13 |
| `e_dis_healthy_vp` | -0.810 (healthy volunteers) | Table 3; ESM 13 |
| `lcl_msc2571109a`, `lvc_msc2571109a`, `lq_msc2571109a`, `lvp_msc2571109a` | 40.2 L/h, 131 L, 106 L/h, 152 L | Table 3 |
| `lfm` | 1 (held constant) | Results, MSC2571109A population PK model |
| `e_crcl_cl_msc2571109a`, `e_wt_cl_msc2571109a` | 0.311, -0.696 | Table 3; Formula 1 |
| `e_tumtp_nsclc_cl_msc2571109a`, `e_tumtp_hcc_fm` | 0.498, -0.398 | Table 3; Formula 2 |
| `e_race_asian_northeast_q_msc2571109a` | 1.40 | Table 3; Formula 2 |
| `e_dis_healthy_vp_msc2571109a`, `e_hepimp_vc_msc2571109a` | 2.31, 0.520 | Table 3; Formula 2 |
| IIV CL/F (patients / healthy) | 0.335 / 0.128 | Table 3 |
| IIV ka, D1 | 0.653, 0.652 | Table 3 |
| IIV F (patients / CF1 doses / healthy) | 0.283 / 0.713 / 0.188 | Table 3 |
| IIV CLmet (patients / healthy), Vc,met, Qmet, Vp,met | 0.536 / 0.255, 0.859, 0.791, 0.248 | Table 3 |
| `propSd`, `propSd_msc2571109a` | 0.337, 0.298 | Table 3 |

# Units: free base versus labelled dose

Table 3 footnote a states that CL/F, Vc/F, Q/F and Vp/F (and the metabolite
CL, Vc, Q, Vp) "were multiplied by a factor of 0.9 to correct for the salt to
base molar weight ratio". The tabulated apparent parameters therefore apply
to the **free-base** amount: a 500 mg tablet of tepotinib hydrochloride
hydrate delivers 450 mg of tepotinib, so `amt = 450`. The dose effect on
bioavailability is written in ESM 13 on the **labelled** dose, `1 + (-0.0412 /
100) * (DOSE - 500 mg)`, so the `DOSE` covariate column carries the labelled
dose (500). Both conventions are reproduced below and are confirmed by the
Figure 3 replication, which lands on the published profile only with 450 mg
free base.

```{r helpers}
mod <- readModelDb("Xiong_2022_tepotinib")

# Covariates of the typical NSCLC patient of Figures 2 and 3 (59 years, 72 kg,
# serum albumin 40 g/L, eGFR 97.28, INR 1.06), taking the drug with food.
nsclc_covariates <- function(id, dose_label, tf3 = 1) {
  tibble(
    id = id,
    DOSE = dose_label,
    WT = 72, AGE = 59, CRCL = 97.28, ALB = 40, INR_BASE = 1.06,
    HEPIMP = 0, DIS_HEALTHY = 0,
    TUMTP_HCC = 0, TUMTP_CRC = 0, TUMTP_NSCLC = 1,
    CONMED_OPIOID = 0, STUDY_MS200095_0028 = 0,
    FED = 1, FED_HIGHFAT = 0,
    FORM_TEPOTINIB_CF1 = 0, FORM_TEPOTINIB_TF1 = 0,
    FORM_TEPOTINIB_TF1FINE = 0, FORM_TEPOTINIB_TF3 = tf3,
    RACE_ASIAN_NORTHEAST = 0
  )
}

# Once-daily oral regimen: dose records go to the depot with rate = -2 so the
# model's dur(depot) (the zero-order duration D1) applies. amt is the free-base
# amount (0.9 x the labelled dose). Observation rows sit on the ODE state
# 'central' with dvid = 1 (the model has two endpoints, so rxode2 needs the
# endpoint id); both Cc and Cc_msc2571109a come back on every row.
make_events <- function(n, dose_label, n_doses, obs_times, tf3 = 1,
                        id_offset = 0L, regimen = "") {
  ids <- id_offset + seq_len(n)
  doses <- tidyr::crossing(id = ids, time = 24 * (seq_len(n_doses) - 1)) |>
    mutate(evid = 1L, amt = 0.9 * dose_label, cmt = "depot", rate = -2,
           dvid = NA_integer_)
  obs <- tidyr::crossing(id = ids, time = obs_times) |>
    mutate(evid = 0L, amt = NA_real_, cmt = "central", rate = NA_real_,
           dvid = 1L)
  bind_rows(doses, obs) |>
    left_join(nsclc_covariates(ids, dose_label, tf3), by = "id") |>
    mutate(regimen = regimen) |>
    arrange(id, time, desc(evid)) |>
    # amt must precede the DOSE covariate column (rxode2 matches 'dose'
    # case-insensitively as an amount alias).
    relocate(id, time, evid, amt, cmt, rate)
}
```

# Replicate Figure 3: 500 mg QD TF3 with food

Figure 3 shows the median and the 5-95% prediction interval of tepotinib
concentrations for a typical NSCLC patient taking 500 mg TF3 once daily with
food, over the first 10 days. Between-subject variability is simulated with
the patient IIV terms; residual error is not added (the `Cc` column is the
individual prediction).

```{r figure-3}
rxode2::rxSetSeed(20220406)
ev_fig3 <- make_events(
  n = 200, dose_label = 500, n_doses = 10,
  obs_times = seq(0, 240, by = 1), regimen = "500 mg QD TF3 fed"
)
sim_fig3 <- rxode2::rxSolve(mod, events = ev_fig3, returnType = "data.frame")

fig3 <- sim_fig3 |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(fig3, aes(time / 24, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), fill = "palegreen", alpha = 0.8) +
  geom_line() +
  labs(
    x = "Time after first dose (day)", y = "Tepotinib plasma concentration (ng/mL)",
    title = "Figure 3: typical NSCLC patient, 500 mg QD TF3 with food",
    caption = "Replicates Figure 3 of Xiong 2022 (median and 5-95% prediction interval)."
  )
```

Values read off the published Figure 3 by the maintainers: median
peak about 460 ng/mL on day 1 at about 0.5 day, median peak about 1110 ng/mL
and median trough about 850 ng/mL over day 9-10, and a day-9 peak
prediction interval of roughly 590-2050 ng/mL.

```{r figure-3-check}
day1 <- fig3 |> filter(time <= 24)
day9 <- fig3 |> filter(time >= 216, time <= 240)
fig3_check <- tibble(
  quantity = c("Day-1 median peak", "Day-1 median tmax (h)", "Day-9 median peak",
               "Day-10 median trough", "Day-9 peak 5th percentile",
               "Day-9 peak 95th percentile"),
  figure_3 = c(460, 12, 1110, 850, 590, 2050),
  simulated = c(max(day1$Q50), day1$time[which.max(day1$Q50)], max(day9$Q50),
                tail(day9$Q50, 1), day9$Q05[which.max(day9$Q50)],
                day9$Q95[which.max(day9$Q50)])
) |>
  mutate(pct_diff = 100 * (simulated / figure_3 - 1))
knitr::kable(fig3_check, digits = 1,
             caption = "Simulated versus digitised Figure 3 values.")

# The medians are centre-of-distribution statistics of 200 subjects and move
# by a few percent between rxode2 builds; a 10% error in CL/F, F, the dose
# (e.g. 500 instead of 450 mg free base) or Vc moves them by more than 10%.
stopifnot(
  abs(fig3_check$pct_diff[c(1, 3, 4)]) < 12,
  # The tails of 200 subjects are noisier, so they get a wider envelope.
  abs(fig3_check$pct_diff[c(5, 6)]) < 30
)
```

## The absorption rate constant: Table 3 versus ESM 13

Table 3 gives ka = 0.278 1/h; the ESM 13 equation sheet prints ka = 1.47 1/h.
The two values give clearly different profile shapes, and Figure 3 decides
between them: with ka = 1.47 1/h the day-1 peak arrives at about 5.5 h and the
steady-state peak-to-trough ratio is about 1.6, whereas Figure 3 shows a broad
day-1 peak at about half a day and a peak-to-trough ratio of about 1.3. The
packaged model uses the Table 3 value.

```{r ka-check}
ev_typ <- make_events(1, dose_label = 500, n_doses = 10,
                      obs_times = seq(0, 240, by = 0.25))
typ_table3 <- rxode2::rxSolve(mod, events = ev_typ, omega = NA, sigma = NA,
                              returnType = "data.frame")
mod_esm13 <- mod |> rxode2::ini(lka = log(1.47))
typ_esm13 <- rxode2::rxSolve(mod_esm13, events = ev_typ, omega = NA, sigma = NA,
                             returnType = "data.frame")

profile_shape <- function(sim) {
  d1 <- sim |> filter(time <= 24)
  d9 <- sim |> filter(time >= 216, time <= 240)
  c(tmax_day1 = d1$time[which.max(d1$Cc)],
    peak_trough_day9 = max(d9$Cc) / tail(d9$Cc, 1))
}
ka_shape <- rbind(`Table 3 (ka = 0.278)` = profile_shape(typ_table3),
                  `ESM 13 (ka = 1.47)` = profile_shape(typ_esm13),
                  `Figure 3 (digitised)` = c(12, 1110 / 850))
knitr::kable(ka_shape, digits = 2,
             caption = "Typical-value profile shape under the two printed ka values.")
stopifnot(
  abs(ka_shape[1, "tmax_day1"] - 12) < 2,
  abs(ka_shape[1, "peak_trough_day9"] - 1110 / 850) < 0.1,
  abs(ka_shape[2, "peak_trough_day9"] - 1110 / 850) > 0.2
)

bind_rows(
  typ_table3 |> mutate(source = "Table 3: ka = 0.278 1/h"),
  typ_esm13 |> mutate(source = "ESM 13: ka = 1.47 1/h")
) |>
  ggplot(aes(time / 24, Cc, colour = source)) +
  geom_line() +
  labs(x = "Time after first dose (day)", y = "Tepotinib (ng/mL)", colour = NULL,
       title = "Typical NSCLC patient, 500 mg QD TF3 with food") +
  theme(legend.position = "bottom")
```

# Replicate Figure 2: steady-state AUC versus dose

Figure 2 plots the predicted tepotinib AUC at steady state against dose for
the typical NSCLC patient, with and without the dose effect on F. At steady
state AUC = F x amt / CL, so the curve follows in closed form from the model's
own parameters. The model is also solved at 100, 500 and 1400 mg to confirm the
closed form.

```{r figure-2}
f_dose <- function(dose) 1 - 0.0412 / 100 * (dose - 500)
fig2 <- tibble(dose = seq(100, 1400, by = 10)) |>
  mutate(
    with_dose_effect = f_dose(dose) * 0.9 * dose / 20.4 * 1000,
    dose_independent = 0.9 * dose / 20.4 * 1000
  )

ggplot(fig2, aes(dose)) +
  geom_line(aes(y = with_dose_effect, linetype = "With the estimated dose effect on F"), colour = "blue") +
  geom_line(aes(y = dose_independent, linetype = "Assuming a dose-independent F"), colour = "grey40") +
  scale_linetype_manual(values = c("dashed", "solid")) +
  labs(x = "Dose (mg)", y = "Predicted tepotinib AUCss (ng*h/mL)", linetype = NULL,
       title = "Figure 2: AUCss versus dose (typical NSCLC patient, TF2 reference)",
       caption = "Replicates Figure 2 of Xiong 2022.") +
  theme(legend.position = "bottom")
```

The figure's AUCss of about 22,000 ng*h/mL at 500 mg is reproduced with the
reference formulation (TF2, F = 1); applying the TF3 effect (+15.4%) would give
about 25,500 ng*h/mL. The ESM 2 bootstrap AUCss at 500 mg (22,274 ng*h/mL) and
at 250 mg (12,277 ng*h/mL) match the same TF2-reference values, so Figure 2 and
ESM 2's AUC rows evidently use F without the TF3 effect even though their
captions mention TF3.

```{r figure-2-check}
ev_auc <- bind_rows(lapply(seq_along(c(100, 250, 500, 1400)), function(i) {
  d <- c(100, 250, 500, 1400)[i]
  make_events(1, dose_label = d, n_doses = 90, obs_times = seq(2136, 2160, by = 0.25),
              tf3 = 0, id_offset = i - 1L, regimen = paste(d, "mg"))
}))
sim_auc <- rxode2::rxSolve(mod, events = ev_auc, omega = NA, sigma = NA,
                           keep = "regimen", returnType = "data.frame")
auc_check <- sim_auc |>
  group_by(regimen) |>
  summarise(auc_tau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2), .groups = "drop") |>
  mutate(dose = as.numeric(sub(" mg", "", regimen)),
         closed_form = f_dose(dose) * 0.9 * dose / 20.4 * 1000,
         pct_diff = 100 * (auc_tau / closed_form - 1)) |>
  arrange(dose)
knitr::kable(auc_check, digits = 1,
             caption = "Day-90 AUC over the dosing interval versus F x amt / CL.")
stopifnot(
  # 90 days approaches but does not quite reach steady state through the slow
  # peripheral compartment (k21 = Q/Vp = 0.0011 1/h); the residual shortfall
  # is about 1%.
  all(abs(auc_check$pct_diff) < 3),
  # Figure 2 read-offs: ~22,000 at 500 mg and ~39,000 at 1400 mg.
  abs(auc_check$closed_form[auc_check$dose == 500] / 22000 - 1) < 0.05,
  abs(auc_check$closed_form[auc_check$dose == 1400] / 39000 - 1) < 0.05,
  # Results: 'relative bioavailability for the supratherapeutic tepotinib dose
  # of 1000 mg is 0.79 compared with the Fpar for a dose of 500 mg'.
  abs(f_dose(1000) - 0.79) < 0.01
)
```

# Clearance versus eGFR (Results)

The Results give typical CL/F values of 16.2, 18.6, 19.7 and 20.9 L/h at eGFR
values of 30, 60, 80 and 110 mL/min/1.73 m^2 for a typical individual
receiving 500 mg.

```{r egfr-check}
egfr_check <- tibble(
  eGFR = c(30, 60, 80, 110),
  published = c(16.2, 18.6, 19.7, 20.9),
  model = 20.4 * (eGFR / 97.28)^0.199
)
knitr::kable(egfr_check, digits = 2, caption = "CL/F (L/h) versus eGFR.")
stopifnot(all(abs(egfr_check$model - egfr_check$published) < 0.1))
```

# PKNCA validation against ESM 2

ESM 2 tabulates bootstrap steady-state exposures for cancer patients at
500 mg and 250 mg once daily, with food, and an effective half-life of
32.1 h (the main text quotes about 32 h). The typical NSCLC patient taking TF3
with food is simulated for 90 days, and steady-state Cmax, Ctrough and
AUC0-24 come from PKNCA over the last dosing interval.

```{r pknca}
ev_ss <- bind_rows(
  make_events(1, dose_label = 500, n_doses = 90, obs_times = seq(2136, 2160, by = 0.25),
              id_offset = 0L, regimen = "500 mg QD"),
  make_events(1, dose_label = 250, n_doses = 90, obs_times = seq(2136, 2160, by = 0.25),
              id_offset = 1L, regimen = "250 mg QD")
)
stopifnot(!anyDuplicated(unique(ev_ss[, c("id", "time", "evid")])))
sim_ss <- rxode2::rxSolve(mod, events = ev_ss, omega = NA, sigma = NA,
                          keep = "regimen", returnType = "data.frame")

conc_ss <- sim_ss |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, regimen)
dose_ss <- ev_ss |>
  filter(evid == 1, time == 2136) |>
  select(id, time, amt, regimen)

conc_obj <- PKNCA::PKNCAconc(conc_ss, Cc ~ time | regimen + id)
dose_obj <- PKNCA::PKNCAdose(dose_ss, amt ~ time | regimen + id)
intervals <- data.frame(start = 2136, end = 2160, cmax = TRUE, cmin = TRUE,
                        auclast = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

published <- tibble::tribble(
  ~regimen,    ~cmax, ~cmin, ~auclast,
  "500 mg QD", 1236,  952,   22274,
  "250 mg QD", 683,   523,   12277
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published,
  by = "regimen",
  units = c(cmax = "ng/mL", cmin = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated typical NSCLC patient (TF3, fed) versus ESM 2 bootstrap means. * differs by >20%.")
```

No row differs by more than 20%. The simulated AUC0-24 is about 14% above
ESM 2 because ESM 2's AUC rows correspond to the TF2-reference F (see the
Figure 2 section), while its Cmax and Ctrough agree with TF3 to within 10%.
ESM 2 is not internally consistent on this point: its 500 mg AUCss implies an
average concentration of 928 ng/mL, below its own mean Ctrough of 952 ng/mL,
which no single concentration profile can produce.

```{r pknca-check}
sim_vals <- as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) |>
  select(regimen, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(published, by = "regimen", suffix = c("_sim", "_pub"))
stopifnot(
  all(abs(sim_vals$cmax_sim / sim_vals$cmax_pub - 1) < 0.1),
  all(abs(sim_vals$cmin_sim / sim_vals$cmin_pub - 1) < 0.15),
  all(abs(sim_vals$auclast_sim / sim_vals$auclast_pub - 1) < 0.2)
)
```

## Effective half-life

The effective half-life follows from an accumulation ratio R (steady state
over first dose) as t1/2,eff = tau x ln 2 / ln(R / (R - 1)); it does not depend
on F or the dose. ESM 2 reports 32.1 h (bootstrap 5th-95th percentile
31.4-32.9 h) and the text "~32 h", but neither says which accumulation ratio
was used. For a two-compartment drug the choice matters, so all three common
versions are shown.

```{r t-half-eff}
ev_acc <- make_events(1, dose_label = 500, n_doses = 90,
                      obs_times = c(seq(0, 24, by = 0.25), seq(2136, 2160, by = 0.25)))
sim_acc <- rxode2::rxSolve(mod, events = ev_acc, omega = NA, sigma = NA,
                           returnType = "data.frame")
auc_window <- function(d, a, b) {
  w <- d |> filter(time >= a, time <= b)
  sum(diff(w$time) * (head(w$Cc, -1) + tail(w$Cc, -1)) / 2)
}
t_eff <- function(R) 24 * log(2) / log(R / (R - 1))
day1 <- sim_acc |> filter(time <= 24)
day90 <- sim_acc |> filter(time >= 2136)
R_acc <- c(
  AUC0_24 = auc_window(sim_acc, 2136, 2160) / auc_window(sim_acc, 0, 24),
  Cmax = max(day90$Cc) / max(day1$Cc),
  Ctrough = tail(day90$Cc, 1) / tail(day1$Cc, 1)
)
t_half_eff <- tibble(ratio_basis = names(R_acc), R = R_acc, t_half_eff_h = t_eff(R_acc))
knitr::kable(t_half_eff, digits = 2,
             caption = "Typical NSCLC patient: effective half-life by accumulation-ratio basis (ESM 2: 32.1 h).")
# Known deviation: the AUC-based value is about 35.5 h, 11% above ESM 2's
# 32.1 h; the Cmax- and Ctrough-based values (about 30 h and 27.5 h) fall
# below it, so the published value lies inside the range of estimators.
# The estimator behind 32.1 h is not stated and ESM 2 averages over the
# covariates of the cancer patients, so the gap is recorded rather than
# chased. The envelope still breaks on a mis-transcribed CL/F, Vc/F or Q/F.
stopifnot(
  abs(t_half_eff$t_half_eff_h[t_half_eff$ratio_basis == "AUC0_24"] / 32.1 - 1) < 0.15,
  min(t_half_eff$t_half_eff_h) < 32.1, max(t_half_eff$t_half_eff_h) > 32.1
)
```

# MSC2571109A

The metabolite model has no published exposure summary to compare against
(ESM 11 is a forest plot of relative AUC), so its typical profile is shown for
the same regimen. Because the fraction metabolised is fixed to 1, the
metabolite amounts are expressed as parent-equivalent mass; the predicted
concentrations are unaffected by that choice (Discussion).

```{r metabolite}
typ_table3 |>
  select(time, Tepotinib = Cc, MSC2571109A = Cc_msc2571109a) |>
  tidyr::pivot_longer(-time, names_to = "analyte", values_to = "conc") |>
  ggplot(aes(time / 24, conc, colour = analyte)) +
  geom_line() +
  labs(x = "Time after first dose (day)", y = "Plasma concentration (ng/mL)",
       colour = NULL, title = "Typical NSCLC patient, 500 mg QD TF3 with food") +
  theme(legend.position = "bottom")

# The metabolite starts at zero, rises, and is present at steady state.
met_end <- tail(typ_table3$Cc_msc2571109a, 1)
stopifnot(typ_table3$Cc_msc2571109a[1] == 0, met_end > 0)
```

# Assumptions and deviations

- **ka.** Table 3 (0.278 1/h) and the ESM 13 equation sheet (1.47 1/h)
  disagree. The Table 3 value is used because only it reproduces the profile
  shape of Figure 3 (see the ka check above).
- **Vc/F.** Table 3 prints 1020 L and ESM 13 prints 1030 L. The 1% difference
  cannot be resolved from the figures; the Table 3 value is used because
  Table 3 is the parameter-estimate table and ESM 13 is demonstrably
  unreliable on ka.
- **Salt-to-base factor.** The tabulated apparent parameters apply to the
  free-base amount (Table 3 footnote a), so `amt` is 0.9 x the labelled dose,
  while the `DOSE` covariate is the labelled dose used by the ESM 13 dose
  effect on F.
- **IIV scale.** The Methods define IIV as log-normal "with standard deviation
  omega", and Table 3 reports each IIV next to "(CV)". The tabulated number is
  read as omega, so each variance is the value squared (for example CL/F
  0.335^2 = 0.1122). Reading it as a true CV, omega^2 = log(1 + CV^2), would
  give 0.1061 for CL/F and, for the widest terms, 0.355 instead of 0.426 (ka)
  and 0.505 instead of 0.738 (Vc,met). The Figure 3 prediction interval cannot
  separate the two readings.
- **Subgroup-specific IIV.** Table 3 reports separate IIV magnitudes for CL/F
  and F in healthy participants, for F on CF1 doses, and for metabolite CL in
  healthy participants. The control stream is not published, so the scoping
  is reconstructed as mutually exclusive etas: healthy participants draw the
  healthy-participant eta; patients draw the CF1 eta on CF1 dose records and
  the patient eta otherwise (CF1 was used only in patients, in study 001).
- **IIV on the fraction metabolised.** The Results state that IIV on FM was
  estimated, but Table 3 does not report its value. It is carried as
  `etalfm ~ fixed(0)`, so stochastic simulations understate metabolite
  between-subject variability.
- **No correlations between etas** are reported, so the omega matrix is
  diagonal.
- **Residual error.** The paper used an additive error on log-transformed
  concentrations; it is encoded as the equivalent proportional error in
  linear space, with the tabulated SDs 0.337 (tepotinib) and 0.298
  (MSC2571109A).
- **Metabolite covariate forms.** ESM 13 covers only the parent. The metabolite
  covariates follow the Methods' Formula 1 (power, normalised to the
  reference participant's eGFR 97.28 and weight 72 kg) and Formula 2
  (`1 + theta` for an indicator); "Patient/participant covariate on Vp,met"
  is applied to healthy participants, as ESM 13 does for the parent's Vp.
- **East Asian.** The metabolite "East Asian" covariate pools the paper's
  Japanese and Other East Asian categories and is mapped to
  `RACE_ASIAN_NORTHEAST`.
- **Serum albumin reference.** The Table 3 footnote's "4 g/L" is read as 40
  g/L, as in ESM 13 and the Figure 2 and 3 captions.
- **Figure 2 and ESM 2 formulation.** Both captions describe TF3 with food, but
  their AUC values match the reference formulation TF2 (F = 1); see the
  Figure 2 and PKNCA sections.
- **Effective half-life.** The model's AUC-accumulation effective half-life
  for the typical NSCLC patient is about 35.5 h, 11% above the 32.1 h of
  ESM 2; the paper does not state its estimator, and the Cmax- and
  Ctrough-based versions (about 30 h and 27.5 h) fall below it, so 32.1 h lies
  inside the range of the three estimators.
- **Figures 4 and 5** (AUCss by race and the covariate forest plot) are built
  from the individual covariates of the analysis data set, which are not
  public, and are not replicated.
- No erratum or correction notice is linked to this article in
  Europe PMC as of 2026-10-01.
