---
title: "Clopidogrel PBPK-PD in CAD with or without diabetes (Xu 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clopidogrel PBPK-PD in CAD with or without diabetes (Xu 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Xu et al. (2020) built a whole-body physiologically based
pharmacokinetic-pharmacodynamic (PBPK-PD) model for clopidogrel (CLOP), its
intermediate 2-oxo-clopidogrel (2-oxo-CLOP) and its active thiol metabolite
(CLOP-AM, the H4 isomer). The PD endpoint is inhibition of platelet
aggregation (IPA). They built the model for healthy adults and then scaled
it to coronary artery disease (CAD) patients with or without diabetes
mellitus (DM).

```{r model}
mod <- readModelDb("Xu_2020_clopidogrel_pbpk")
```

* Citation: `r rxode2::rxode(mod)$reference`
* Article: <https://doi.org/10.3389/fphar.2020.593982> (open access, PMC7845657)
* Description: `r rxode2::rxode(mod)$description`

Structure (paper Figure 1 and Equations 1-18):

* Each of the three analytes has the same perfusion-limited body (Eq 1):
  lung, heart, spleen, liver, kidney, brain, adipose, muscle, skin, rest of
  body (ROB), stomach wall, venous and arterial blood, and five gut-wall
  segments. Tissue `t` returns blood at `C_t / Kt/b`, where
  `Kt/b = Kt/p / Rbp`.
* Oral clopidogrel empties from the stomach lumen (Eq 2) through five
  gut-lumen segments (Eq 3). The duodenum, jejunum and ileum absorb it at
  `Ka,i` and the gut wall returns unbound drug to the lumen at the P-gp
  efflux constant `Kb,i` (Eqs 4-9). The cecum and colon do not absorb.
* The liver (Eqs 11-15) receives hepatic-artery flow plus the stomach,
  spleen and gut-wall outflows. CYP1A2, CYP2B6 and CYP2C19
  Michaelis-Menten oxidation of unbound clopidogrel forms 2-oxo-CLOP.
  CYP2B6, CYP2C9, CYP3A4 and CYP2C19 oxidation of 2-oxo-CLOP forms
  CLOP-AM. CES1 hydrolysis inactivates all three analytes. The microsomal
  kinetics are scaled to the whole liver by the microsomal protein content
  `PBSF` (Eq 12).
* PD (Eqs 16-17): unbound venous CLOP-AM irreversibly inactivates platelets:
  `dM/dt = kin - kout * M - kirre * C_ven,AM * fub,AM * M`, `M(0) = 1`,
  and `IPA = (1 - M) * 100`.
* Covariates: CYP2C19 phenotype sets the CYP2C19 `Vmax` of both oxidation
  steps (UM 1.58-fold, IM 0.5-fold, PM 0 relative to EM). CAD scales every
  blood flow by 0.90 (Eq 18) and `kirre` by 0.7. DM rescales the CYP and
  CES1 activities and replaces the gastrointestinal transit rates.

# Population

The physiology is a 70-kg reference adult (Table 1; the listed volumes sum to
70 L). The model is a bottom-up prediction model. Its structural parameters
come from in vitro kinetics (Kazui 2010 microsomal `Vmax`/`Km`, Caco-2
permeability) and the literature. The only parameter fitted to in vivo data
is `kirre`, which was fitted to one published IPA-time profile (Zhu 2008).
The comparison data are published healthy-volunteer studies (15 PK and 12
IPA reports) and studies in CAD (5 reports) and CAD + DM (5 reports)
patients from East Asia, Europe and North America. Doses range from single
doses of 75-900 mg to 300/75 mg and 600/150 mg loading/maintenance regimens.
The paper's visual predictive check estimated variances on
`Vmax,CYP2C19`, `CLint,CES1`, `Kt,i` and `kirre`, but it did not report
them, so the packaged model is typical-value only. The same information is
available as `rxode2::rxode(mod)$population`.

# Source trace

Every `ini()` value carries an in-file comment that points to its source.

| Quantity | Value(s) | Source |
|---|---|---|
| Organ volumes and blood flows (healthy) | Table 1, 18 rows | Table 1 'Health' columns |
| CAD blood-flow scaling | 0.90 x every flow | Eq 18 and text (Rerych 1978); Table 1 'CAD' columns |
| `fu`, `Rbp`, `fumic` (CLOP / 2-oxo / AM) | 0.02 / 0.0742 / 0.0791; 0.57 / 0.68 / 0.58; 0.015 / 0.180 / - | Table 2 |
| `Kt/p` for 11 tissues x 3 analytes | Table 2 | Table 2 (Schmitt 2008 method) |
| Gastric emptying `Kt,0` | 4.8 1/h | Methods 'In stomach' |
| Transit `Kt,1..5` | 4.2, 1.8, 2.4, 0.18, 0.06 1/h | Methods 'In gut lumen' |
| `fugut` | 0.02 | Methods 'In gut lumen' |
| `Ka,1..3` / `Kb,1..3` | 0.21, 0.26, 0.29 / 0.07, 0.12, 0.16 1/h | Methods, after Eqs 7-8 |
| Microsomal `Vmax`, `Km`, enzyme content | Table 3 | Table 3 (Kazui 2010) |
| `PBSF` | 55,120 mg | Methods 'In liver' |
| `CLint,CES1` CLOP / 2-oxo / AM | 276,650 / 2,200 / 529 L/h | Methods 'In liver' |
| CYP2C19 UM / IM / PM activity | 1.58 / 0.5 / 0 x EM | Methods 'In liver'; Table 3 footnotes b-c |
| DM CYP1A2/2B6/2C9/2C19/3A4 and CES1 | 1.23 / 0.55 / 1.26 / 0.54 / 0.62; 1.27 | Methods 'PBPK-PD Model in CAD Patients With DM' |
| DM transit, stomach..colon | 2.31, 2.30, 0.99, 1.32, 0.20, 0.04 1/h | same section |
| `kout`, `kin`, `M0` | 0.007804 1/h, `kout * M0`, 1 | Methods 'PD Kinetics' |
| `kirre` (healthy) and CAD factor | 47.576 mL/nmol/h; 0.7 | Methods 'PD Kinetics'; 'CAD Without DM' |
| Molecular weights | 321.82 / 337.82 / 355.83 g/mol | not printed; see Assumptions |
| Tissue, liver and gut-wall mass balances | | Eqs 1, 3, 9-11, 13, 15 |
| Lung, venous and arterial blood pools | | standard closure (not printed; see Assumptions) |

Two checks confirm the table values. First, the Table 1 flows close
exactly: the systemic organ flows plus the stomach and gut-wall flows sum to
the 5.27 L/min cardiac output. Second, the printed CES1 clearance of
clopidogrel is in L/h. With the Table 3 kinetics and `PBSF`, the linear-range
CYP intrinsic clearance is about 46,000 L/h, so CES1 is 86% of the total.
The paper states "85%".

# Simulation helpers

```{r helpers}
# One deterministic typical subject per treatment. Doses go into the stomach
# lumen; observation rows carry dvid = 1 (every output is computed on every
# row). Phenotype and disease are per-subject covariate columns.
make_subject <- function(id, treatment, dose_times, dose_amts, UM = 0, IM = 0,
                         PM = 0, IHD = 0, DIAB = 0, obs_times) {
  doses <- data.frame(
    id = id, time = dose_times, evid = 1L, amt = dose_amts,
    cmt = "stomach", dvid = NA_integer_
  )
  obs <- data.frame(
    id = id, time = obs_times, evid = 0L, amt = NA_real_,
    cmt = NA_character_, dvid = 1L
  )
  out <- rbind(doses, obs)
  out <- out[order(out$time, -out$evid), ]
  out$CYP2C19_UM <- UM
  out$CYP2C19_IM <- IM
  out$CYP2C19_PM <- PM
  out$DIS_IHD <- IHD
  out$DIS_DIAB <- DIAB
  out$treatment <- treatment
  out
}

phenotype_flags <- function(ph) {
  list(UM = as.integer(ph == "UM"), IM = as.integer(ph == "IM"), PM = as.integer(ph == "PM"))
}

solve_typical <- function(events, params = NULL) {
  out <- rxode2::rxSolve(mod, events = events, params = params, keep = "treatment",
                         returnType = "data.frame")
  as.data.frame(out)
}
```

# Healthy volunteers: single oral doses (Table 4, Figure 2)

```{r single-sim}
single_grid <- sort(unique(c(seq(0, 6, by = 0.02), seq(6, 24, by = 0.25))))
single_design <- tidyr::expand_grid(dose = c(75, 150, 300, 600), phenotype = c("UM", "EM", "IM", "PM"))
single_events <- bind_rows(lapply(seq_len(nrow(single_design)), function(i) {
  f <- phenotype_flags(single_design$phenotype[i])
  make_subject(
    id = i, treatment = paste0(single_design$dose[i], " mg ", single_design$phenotype[i]),
    dose_times = 0, dose_amts = single_design$dose[i],
    UM = f$UM, IM = f$IM, PM = f$PM, obs_times = single_grid
  )
}))
sim_single <- solve_typical(single_events)
```

```{r figure-2}
sim_single |>
  filter(grepl("^300 mg", treatment), time <= 12) |>
  select(time, treatment, Clopidogrel = Cc, `CLOP-AM` = Cc_h4) |>
  pivot_longer(-c(time, treatment), names_to = "analyte", values_to = "conc") |>
  ggplot(aes(time, conc, colour = treatment)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(
    x = "Time after dose (h)", y = "Plasma concentration (ng/mL)", colour = NULL,
    caption = "Replicates the model predictions of Figure 2 (300 mg single dose) of Xu 2020."
  )
```

## NCA with PKNCA

The paper reports one predicted `Cmax` and one predicted `AUC` per dose and
phenotype (Table 4, 'Pre' columns). It does not state the AUC window. Cmax does
not depend on the window, so it is the primary comparison. For AUC the
vignette reports `AUC0-24` and the two windows that come closest to the
paper's values; see the Assumptions section.

```{r single-nca}
nca_one <- function(sim, events, conc_col, analyte, start, end, partial_end) {
  conc <- sim |>
    filter(!is.na(.data[[conc_col]])) |>
    transmute(id, treatment, time, conc = .data[[conc_col]])
  dose <- events |>
    filter(evid == 1, time >= start, time < end) |>
    distinct(id, treatment, time, amt)
  conc_obj <- PKNCA::PKNCAconc(conc, conc ~ time | treatment + id,
                               concu = "ng/mL", timeu = "h")
  dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | treatment + id, doseu = "mg")
  intervals <- data.frame(
    start = c(start, start), end = c(end, partial_end),
    cmax = c(TRUE, FALSE), tmax = c(TRUE, FALSE), auclast = c(TRUE, FALSE),
    aucint.last = c(FALSE, TRUE)
  )
  res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
  as.data.frame(res$result) |>
    mutate(analyte = analyte) |>
    select(analyte, treatment, PPTESTCD, PPORRES)
}

nca_single <- bind_rows(
  nca_one(sim_single, single_events, "Cc", "Clopidogrel", 0, 24, 6),
  nca_one(sim_single, single_events, "Cc_h4", "CLOP-AM", 0, 24, 5)
)
```

```{r single-compare}
# Xu 2020 Table 4 'Pre' columns (model predictions), healthy volunteers.
pub_single <- tibble::tribble(
  ~analyte,      ~treatment,   ~aucint.last, ~cmax,
  "Clopidogrel", "75 mg EM",   2.15,   1.12,
  "Clopidogrel", "75 mg IM",   2.22,   1.15,
  "Clopidogrel", "75 mg PM",   2.29,   1.19,
  "Clopidogrel", "150 mg EM",  4.30,   2.23,
  "Clopidogrel", "150 mg IM",  4.44,   2.31,
  "Clopidogrel", "150 mg PM",  4.59,   2.38,
  "Clopidogrel", "300 mg EM",  8.61,   4.48,
  "Clopidogrel", "300 mg IM",  8.88,   4.62,
  "Clopidogrel", "300 mg PM",  9.18,   4.77,
  "CLOP-AM",     "75 mg EM",   13.85,  5.72,
  "CLOP-AM",     "75 mg IM",   10.26,  4.23,
  "CLOP-AM",     "75 mg PM",   6.86,   2.83,
  "CLOP-AM",     "150 mg EM",  27.56,  11.37,
  "CLOP-AM",     "150 mg IM",  20.41,  8.41,
  "CLOP-AM",     "150 mg PM",  13.66,  5.62,
  "CLOP-AM",     "300 mg UM",  71.88,  29.63,
  "CLOP-AM",     "300 mg EM",  54.56,  22.45,
  "CLOP-AM",     "300 mg IM",  40.42,  16.61,
  "CLOP-AM",     "300 mg PM",  27.07,  11.11,
  "CLOP-AM",     "600 mg UM",  140.92, 57.79,
  "CLOP-AM",     "600 mg EM",  106.96, 43.79,
  "CLOP-AM",     "600 mg IM",  79.28,  32.42,
  "CLOP-AM",     "600 mg PM",  53.18,  21.73
)

sim_single_cmp <- nca_single |>
  semi_join(pub_single, by = c("analyte", "treatment"))

cmp_single <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_single_cmp,
  reference = pub_single,
  by = c("analyte", "treatment"),
  params = c("cmax", "aucint.last"),
  units = c(cmax = "ng/mL", aucint.last = "ng*h/mL"),
  tolerance_pct = 20
)
cmp_single |>
  mutate(`NCA parameter` = sub("^aucint.last", "Partial AUC", `NCA parameter`)) |>
  rename("Analyte" = analyte, "Dose and phenotype" = treatment) |>
  knitr::kable(
    caption = paste(
      "Simulated versus Xu 2020 Table 4 predicted values (healthy volunteers).",
      "The partial AUC is AUC0-6 h for clopidogrel and AUC0-5 h for CLOP-AM",
      "(see Assumptions). * differs by more than 20%."
    ),
    digits = 2
  )
```

The model-predicted `AUC0-24` and `tmax` after 300 mg are:

```{r single-auc24}
nca_single |>
  filter(PPTESTCD %in% c("auclast", "tmax"), grepl("^300 mg", treatment)) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  rename("Analyte" = analyte, "Dose and phenotype" = treatment,
         "AUC0-24 (ng*h/mL)" = auclast, "tmax (h)" = tmax) |>
  knitr::kable(digits = 2, caption = "Simulated AUC0-24 and tmax after 300 mg.")
```

```{r single-gates}
cmax_check <- sim_single_cmp |>
  filter(PPTESTCD == "cmax") |>
  inner_join(pub_single, by = c("analyte", "treatment")) |>
  mutate(pct = 100 * (PPORRES - cmax) / cmax)
# Deterministic typical-value solve of the published equations: every
# predicted Cmax (23 dose x phenotype x analyte rows) is reproduced to within
# 3%. A wrong CYP2C19 factor, CES1 clearance, unit or stoichiometry moves
# whole groups by 10% or more.
stopifnot(nrow(cmax_check) == 23, all(abs(cmax_check$pct) < 3))

# Phenotype structure, independent of any scale: CLOP-AM Cmax relative to EM
# at 300 mg matches the paper's 1.320 (UM), 0.740 (IM) and 0.495 (PM).
cm <- setNames(cmax_check$PPORRES, paste(cmax_check$analyte, cmax_check$treatment))
stopifnot(
  abs(cm[["CLOP-AM 300 mg UM"]] / cm[["CLOP-AM 300 mg EM"]] - 29.63 / 22.45) < 0.01,
  abs(cm[["CLOP-AM 300 mg IM"]] / cm[["CLOP-AM 300 mg EM"]] - 16.61 / 22.45) < 0.01,
  abs(cm[["CLOP-AM 300 mg PM"]] / cm[["CLOP-AM 300 mg EM"]] - 11.11 / 22.45) < 0.01
)
```

# Healthy volunteers: loading and maintenance doses (Table 4)

Table 4 gives predicted values on day 7 (and day 5 in Simon 2011, with the same
prediction) after a 300 mg or 600 mg loading dose followed by 75 mg or 150 mg
once daily. The seventh dose is at 144 h.

```{r md-sim}
md_design <- tidyr::expand_grid(regimen = c("300/75", "600/150"), phenotype = c("UM", "EM", "IM", "PM"))
md_grid <- sort(unique(c(seq(0, 144, by = 1), seq(144, 168, by = 0.02))))
md_events <- bind_rows(lapply(seq_len(nrow(md_design)), function(i) {
  f <- phenotype_flags(md_design$phenotype[i])
  ld <- if (md_design$regimen[i] == "300/75") 300 else 600
  mdose <- if (md_design$regimen[i] == "300/75") 75 else 150
  make_subject(
    id = 100 + i, treatment = paste(md_design$regimen[i], "mg", md_design$phenotype[i]),
    dose_times = seq(0, 144, by = 24), dose_amts = c(ld, rep(mdose, 6)),
    UM = f$UM, IM = f$IM, PM = f$PM, obs_times = md_grid
  )
}))
sim_md <- solve_typical(md_events)
nca_md <- bind_rows(
  nca_one(sim_md, md_events, "Cc", "Clopidogrel", 144, 168, 150),
  nca_one(sim_md, md_events, "Cc_h4", "CLOP-AM", 144, 168, 149)
)

pub_md <- tibble::tribble(
  ~analyte,      ~treatment,        ~aucint.last, ~cmax,
  "Clopidogrel", "300/75 mg EM",    4.78,   2.33,
  "Clopidogrel", "300/75 mg IM",    4.93,   2.40,
  "Clopidogrel", "300/75 mg PM",    5.10,   2.49,
  "CLOP-AM",     "300/75 mg UM",    18.99,  7.71,
  "CLOP-AM",     "300/75 mg EM",    14.42,  5.85,
  "CLOP-AM",     "300/75 mg IM",    10.69,  4.32,
  "CLOP-AM",     "300/75 mg PM",    7.15,   2.89,
  "CLOP-AM",     "600/150 mg UM",   37.78,  15.32,
  "CLOP-AM",     "600/150 mg EM",   28.70,  11.62,
  "CLOP-AM",     "600/150 mg IM",   21.27,  8.60,
  "CLOP-AM",     "600/150 mg PM",   14.24,  5.74
)
sim_md_cmp <- nca_md |> semi_join(pub_md, by = c("analyte", "treatment"))
nlmixr2lib::ncaComparisonTable(
  simulated = sim_md_cmp, reference = pub_md, by = c("analyte", "treatment"),
  params = c("cmax", "aucint.last"),
  units = c(cmax = "ng/mL", aucint.last = "ng*h/mL"), tolerance_pct = 20
) |>
  mutate(`NCA parameter` = sub("^aucint.last", "Partial AUC", `NCA parameter`)) |>
  rename("Analyte" = analyte, "Regimen and phenotype" = treatment) |>
  knitr::kable(
    caption = paste(
      "Day-7 maintenance dose, simulated versus Xu 2020 Table 4 predicted values.",
      "Partial AUC as in the single-dose table. * differs by more than 20%."
    ),
    digits = 2
  )
```

The CLOP-AM rows agree to within 3%. The simulated day-7 CLOP-AM `Cmax` is
0.8% above the single-dose value, while the paper's is 2.3% above, so the
paper's model accumulates CLOP-AM slightly more.

The three clopidogrel rows (Kim 2014) do not agree: the paper's values are
about twice the simulated ones. The paper's own predictions are inconsistent
here. Its day-7 clopidogrel `Cmax` after 75 mg (2.33 ng/mL) is 2.08 times
its single 75 mg prediction (1.12 ng/mL). Yet the same maintenance dose in
CAD patients (Table 5, 75 mg on day 8: 1.14 ng/mL) shows essentially no
accumulation, and CAD changes only the blood flows, by 10%. The 2.33 value is
close to the paper's single 150 mg prediction (2.23 ng/mL). The clopidogrel
rows are therefore shown but not gated.

```{r md-gates}
md_check <- sim_md_cmp |>
  filter(PPTESTCD == "cmax") |>
  inner_join(pub_md, by = c("analyte", "treatment")) |>
  mutate(pct = 100 * (PPORRES - cmax) / cmax)
md_am <- filter(md_check, analyte == "CLOP-AM")
stopifnot(nrow(md_am) == 8, all(abs(md_am$pct) < 4))
# The paper's internal inconsistency for clopidogrel, from its own tables:
# healthy day-7 / single 75 mg is ~2.1, CAD day-8 / healthy single 75 mg ~1.0.
stopifnot(abs(2.33 / 1.12 - 2.08) < 0.01, abs(1.14 / 1.12 - 1) < 0.05)
```

# CAD patients with or without DM (Table 5, Results)

Table 5 gives the predicted values for CAD patients. The DM cohort (Results,
'Prediction and Validation of Pharmacokinetics and Pharmacodynamics in CAD
with DM Patients') received a single 600 mg dose. The paper did not state the
phenotype of this prediction; it is simulated here as an EM.

```{r cad-sim}
cad_rows <- tibble::tribble(
  ~treatment,              ~phenotype, ~ld,  ~md, ~ndose, ~IHD, ~DIAB,
  "CAD 75 mg MD d8 UM",    "UM",       75,   75,  8,      1,    0,
  "CAD 75 mg MD d8 EM",    "EM",       75,   75,  8,      1,    0,
  "CAD 75 mg MD d8 IM",    "IM",       75,   75,  8,      1,    0,
  "CAD 300 mg EM",         "EM",       300,  0,   1,      1,    0,
  "CAD 300 mg IM",         "IM",       300,  0,   1,      1,    0,
  "CAD 300 mg PM",         "PM",       300,  0,   1,      1,    0,
  "CAD 900 mg EM",         "EM",       900,  0,   1,      1,    0,
  "CAD 900 mg IM",         "IM",       900,  0,   1,      1,    0,
  "CAD 900 mg PM",         "PM",       900,  0,   1,      1,    0,
  "CAD + DM 600 mg EM",    "EM",       600,  0,   1,      1,    1
)
cad_events <- bind_rows(lapply(seq_len(nrow(cad_rows)), function(i) {
  r <- cad_rows[i, ]
  f <- phenotype_flags(r$phenotype)
  last <- 24 * (r$ndose - 1)
  make_subject(
    id = 200 + i, treatment = r$treatment,
    dose_times = seq(0, last, by = 24), dose_amts = c(r$ld, rep(r$md, r$ndose - 1)),
    UM = f$UM, IM = f$IM, PM = f$PM, IHD = r$IHD, DIAB = r$DIAB,
    obs_times = sort(unique(c(seq(0, last, by = 1), seq(last, last + 24, by = 0.02))))
  )
}))
sim_cad <- solve_typical(cad_events)
# Shift each subject so that its last dosing interval runs from 0 to 24 h.
shift_last <- function(d) {
  d |>
    group_by(id) |>
    mutate(time = time - max(time) + 24) |>
    ungroup() |>
    filter(time >= 0)
}
sim_cad_ss <- shift_last(sim_cad)
cad_events_ss <- shift_last(cad_events)
nca_cad <- bind_rows(
  nca_one(sim_cad_ss, cad_events_ss, "Cc", "Clopidogrel", 0, 24, 6),
  nca_one(sim_cad_ss, cad_events_ss, "Cc_h4", "CLOP-AM", 0, 24, 5)
)

pub_cad <- tibble::tribble(
  ~analyte,      ~treatment,             ~aucint.last, ~cmax,
  "Clopidogrel", "CAD 75 mg MD d8 UM",   2.25,   1.10,
  "Clopidogrel", "CAD 75 mg MD d8 EM",   2.33,   1.14,
  "Clopidogrel", "CAD 75 mg MD d8 IM",   2.41,   1.18,
  "CLOP-AM",     "CAD 75 mg MD d8 UM",   19.08,  7.80,
  "CLOP-AM",     "CAD 75 mg MD d8 EM",   14.49,  5.92,
  "CLOP-AM",     "CAD 75 mg MD d8 IM",   10.74,  4.38,
  "CLOP-AM",     "CAD 300 mg EM",        55.14,  22.80,
  "CLOP-AM",     "CAD 300 mg IM",        40.86,  16.88,
  "CLOP-AM",     "CAD 300 mg PM",        27.40,  11.29,
  "CLOP-AM",     "CAD 900 mg EM",        159.03, 65.10,
  "CLOP-AM",     "CAD 900 mg IM",        117.96, 48.23,
  "CLOP-AM",     "CAD 900 mg PM",        79.27,  32.41,
  "CLOP-AM",     "CAD + DM 600 mg EM",   52.46,  16.08
)
sim_cad_cmp <- nca_cad |> semi_join(pub_cad, by = c("analyte", "treatment"))
nlmixr2lib::ncaComparisonTable(
  simulated = sim_cad_cmp, reference = pub_cad, by = c("analyte", "treatment"),
  params = c("cmax", "aucint.last"),
  units = c(cmax = "ng/mL", aucint.last = "ng*h/mL"), tolerance_pct = 20
) |>
  mutate(`NCA parameter` = sub("^aucint.last", "Partial AUC", `NCA parameter`)) |>
  rename("Analyte" = analyte, "Population, regimen and phenotype" = treatment) |>
  knitr::kable(
    caption = paste(
      "CAD and CAD + DM, simulated versus Xu 2020 Table 5 and Results predicted",
      "values (last dosing interval). * differs by more than 20%."
    ),
    digits = 2
  )
```

```{r cad-gates}
cad_check <- sim_cad_cmp |>
  filter(PPTESTCD == "cmax") |>
  inner_join(pub_cad, by = c("analyte", "treatment")) |>
  mutate(pct = 100 * (PPORRES - cmax) / cmax)
stopifnot(nrow(cad_check) == 13, all(abs(cad_check$pct) < 5))
```

# Pharmacodynamics: IPA (Figures 3, 4G and 5E)

```{r pd-sim}
pd_design <- tidyr::expand_grid(
  population = c("Healthy", "CAD", "CAD + DM"), phenotype = c("UM", "EM", "IM", "PM")
) |>
  bind_rows(tibble::tribble(
    ~population, ~phenotype, ~md,
    "CAD",       "PM",       150,
    "CAD + DM",  "EM",       150,
    "CAD + DM",  "IM",       187.5,
    "CAD + DM",  "PM",       265.5
  )) |>
  mutate(md = ifelse(is.na(md), 75, md))
pd_events <- bind_rows(lapply(seq_len(nrow(pd_design)), function(i) {
  r <- pd_design[i, ]
  f <- phenotype_flags(r$phenotype)
  make_subject(
    id = 300 + i, treatment = paste0(r$population, ", ", r$phenotype, ", 300/", r$md, " mg"),
    dose_times = seq(0, 13 * 24, by = 24), dose_amts = c(300, rep(r$md, 13)),
    UM = f$UM, IM = f$IM, PM = f$PM,
    IHD = as.integer(r$population != "Healthy"), DIAB = as.integer(r$population == "CAD + DM"),
    obs_times = seq(0, 14 * 24, by = 1)
  ) |>
    mutate(population = r$population, phenotype = r$phenotype, md = r$md)
}))
sim_pd <- rxode2::rxSolve(mod, events = pd_events, keep = c("treatment", "population", "phenotype", "md"),
                          returnType = "data.frame") |>
  as.data.frame()
```

```{r figure-4g-5e}
sim_pd |>
  filter(md == 75) |>
  mutate(phenotype = factor(phenotype, levels = c("UM", "EM", "IM", "PM"))) |>
  ggplot(aes(time / 24, IPA, colour = phenotype)) +
  geom_line() +
  facet_wrap(~population) +
  labs(
    x = "Time (days)", y = "IPA (%)", colour = "CYP2C19",
    caption = paste(
      "300 mg loading dose then 75 mg once daily. Replicates the model",
      "predictions of Figures 3D, 4G and 5D-E of Xu 2020."
    )
  )
```

The paper states three quantitative PD results for the 300 mg loading dose
followed by 75 mg once daily, and the maintenance doses that restore the IPA
of CAD EM patients. It does not say at what time it read IPA off Figures 4G
and 5E. The comparison below uses the steady-state trough 24 h after the
14th dose (336 h). At the day-7 trough the ratio and difference claims agree
equally well (62.1, 22.1 and 12.7). The dose-adjusted arms are then still
2.4-4.4 IPA points below the reference, because a larger maintenance dose takes
longer to approach its steady state.

```{r pd-table}
ipa7 <- sim_pd |>
  filter(time == 336) |>
  select(population, phenotype, md, IPA)
ipa <- function(pop, ph, md = 75) ipa7$IPA[ipa7$population == pop & ipa7$phenotype == ph & ipa7$md == md]
pd_claims <- tibble::tribble(
  ~claim, ~paper, ~simulated,
  "CAD: IPA in PM / IPA in EM (x100)", 62, 100 * ipa("CAD", "PM") / ipa("CAD", "EM"),
  "CAD: IPA(UM) - IPA(PM), percentage points", 23.2, ipa("CAD", "UM") - ipa("CAD", "PM"),
  "CAD + DM: IPA(UM) - IPA(PM), percentage points", 13.7, ipa("CAD + DM", "UM") - ipa("CAD + DM", "PM")
)
knitr::kable(pd_claims, digits = 1, caption = "IPA claims from the Xu 2020 Results and Discussion.")

dose_up <- tibble::tribble(
  ~scenario,                          ~IPA,
  "CAD EM, 75 mg (reference)",        ipa("CAD", "EM"),
  "CAD PM, 150 mg",                   ipa("CAD", "PM", 150),
  "CAD + DM EM, 150 mg",              ipa("CAD + DM", "EM", 150),
  "CAD + DM IM, 187.5 mg",            ipa("CAD + DM", "IM", 187.5),
  "CAD + DM PM, 265.5 mg",            ipa("CAD + DM", "PM", 265.5)
)
knitr::kable(
  dose_up, digits = 1,
  caption = paste(
    "Maintenance doses that the paper reports restore the IPA of CAD EM",
    "patients on 75 mg (Figures 4G and 5E)."
  )
)
```

```{r pd-gates}
stopifnot(
  abs(pd_claims$simulated[1] - 62) < 4,
  abs(pd_claims$simulated[2] - 23.2) < 2.5,
  abs(pd_claims$simulated[3] - 13.7) < 2.5,
  # every adjusted dose lands within 2.5 IPA points of the CAD EM reference
  all(abs(dose_up$IPA[-1] - dose_up$IPA[1]) < 2.5)
)
```

# Sensitivity: which DM change lowers CLOP-AM exposure (Figure 7F)

The paper switched each DM change on one at a time in a CAD EM patient after a
300 mg dose. The transit change raised CLOP-AM exposure by 58%. The CES1,
CYP2C19 and CYP3A4 changes lowered it by 43%, 24% and 12%. Here each change
is isolated by setting the other DM multipliers back to their healthy values
through `params`.

```{r figure-7f}
th <- rxode2::rxode(mod)$theta
dm_neutral <- c(e_dm_cyp1a2 = 1, e_dm_cyp2b6 = 1, e_dm_cyp2c9 = 1,
                e_dm_cyp2c19 = 1, e_dm_cyp3a4 = 1, e_dm_ces1 = 1)
transit_healthy <- c(ktr_stomach_dm = th[["ktr_stomach"]], ktr_duodenum_dm = th[["ktr_duodenum"]],
                     ktr_jejunum_dm = th[["ktr_jejunum"]], ktr_ileum_dm = th[["ktr_ileum"]],
                     ktr_cecum_dm = th[["ktr_cecum"]], ktr_colon_dm = th[["ktr_colon"]])
scenarios <- list(
  "Transit rates" = dm_neutral,
  "CES1" = replace(c(dm_neutral, transit_healthy), "e_dm_ces1", th[["e_dm_ces1"]]),
  "CYP2C19" = replace(c(dm_neutral, transit_healthy), "e_dm_cyp2c19", th[["e_dm_cyp2c19"]]),
  "CYP3A4" = replace(c(dm_neutral, transit_healthy), "e_dm_cyp3a4", th[["e_dm_cyp3a4"]])
)
ev_cad <- make_subject(1, "CAD", 0, 300, IHD = 1, obs_times = seq(0, 24, by = 0.02))
auc_h4 <- function(s) {
  sum(diff(s$time) * (head(s$Cc_h4, -1) + tail(s$Cc_h4, -1)) / 2)
}
auc_ref <- auc_h4(solve_typical(ev_cad))
ev_dm <- mutate(ev_cad, DIS_DIAB = 1L)
fig7f <- tibble::tibble(
  change = names(scenarios),
  paper = c(58, -43, -24, -12),
  simulated = vapply(scenarios, function(p) {
    full <- th
    full[names(p)] <- p
    100 * (auc_h4(solve_typical(ev_dm, params = full)) / auc_ref - 1)
  }, numeric(1))
)
knitr::kable(fig7f, digits = 1,
             caption = "Percent change in CLOP-AM AUC0-24 from each isolated DM change (Figure 7F).")
stopifnot(all(abs(fig7f$simulated - fig7f$paper) < 4))
```

# Assumptions and deviations

* **Metabolites are formed 1:1 by mass.** Equations 13 and 15 add the
  metabolized clopidogrel (or 2-oxo-CLOP) amount directly to the metabolite
  with no molecular-weight ratio. The packaged model keeps that as-run form:
  its states are in mg, and molecular weights are used only to convert
  concentrations to the molar units of `Km` (uM) and `kirre` (mL/nmol/h).
  A molar-conserving variant that forms CLOP-AM mole-for-mole
  over-predicts every published CLOP-AM `Cmax` by a uniform 9% (the
  321.82/355.83 molecular-weight ratio). The as-run form reproduces them to
  within 1.5% after single doses and 3.5% in CAD or after repeated doses.
  The Kong 2020 vonoprazan code (doi:10.1038/s41401-019-0353-2) from the same group also
  runs in mass units.
* **Molecular weights are not printed in the paper.** Clopidogrel 321.82 and
  CLOP-AM 355.83 g/mol are taken from Jung 2024 (doi:10.1002/psp4.13053).
  2-oxo-clopidogrel (C16H16ClNO3S) is 321.82 + 16.00 = 337.82 g/mol.
* **Stomach-wall `Kt/p` is not printed.** Table 2 has no stomach row, but
  the liver equation needs `K_st/b`. The gut value is used for all three
  analytes. The stomach wall is a 0.15 L non-eliminating tissue, so this
  choice has a negligible effect on plasma concentrations.
* **The cecum and colon do not absorb.** The paper gives `Ka,i` and `Kb,i`
  (and radii) only for the duodenum, jejunum and ileum.
* **The liver sums the gut-wall flows over i = 1..5.** Eq 11 prints
  `sum(i = 0..5)` next to a separate stomach term. Summing i = 1..5 plus the
  stomach closes the Table 1 flows exactly on the 5.27 L/min cardiac output,
  so the i = 0 term would double-count the stomach.
* **Lung, venous and arterial blood pools** are the standard whole-body
  closure (lung between venous and arterial blood at cardiac output). The
  paper does not print them; the equations follow Kong 2020 from the same
  group.
* **`CLint,CES1,CLOP` is in L/h.** The paper prints "276,650 1/h". The 85%
  share of total intrinsic clearance that the paper cites holds only in L/h
  (86% with the Table 3 kinetics).
* **Enzyme content multiplies `Vmax`.** Eq 12 writes `Vmax` in
  pmol/min/pmol P450; the Table 3 enzyme content (pmol P450/mg protein)
  converts it to a per-mg-protein rate before `PBSF` scales it to the liver.
  The same 85% check confirms this.
* **CAD flows are 0.90 x the healthy flows** (Eq 18) rather than the
  rounded Table 1 'CAD' column. The two agree to the printed precision.
* **DM is only defined together with CAD.** The paper's DM cohort is CAD +
  DM, whose `kirre` equals the CAD value. `DIS_DIAB = 1` with
  `DIS_IHD = 0` is outside the paper's scenarios.
* **AUC window.** Table 4 does not state the AUC window of its predictions.
  AUC0-6 h reproduces the clopidogrel values and AUC0-5 h the CLOP-AM
  values, and the partial AUCs in the tables use those windows. The windows
  are inferred, so the gates use `Cmax` only; `AUC0-24` is shown
  separately.
* **No IIV or residual error.** The paper estimated variances for its
  visual predictive check but did not report them. The model is
  typical-value only, and the residual SDs are fixed at zero.
