---
title: "Xu_2011_sirukumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Xu_2011_sirukumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Xu_2011_sirukumab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Xu_2011_sirukumab"))$description`
* Article: [Br J Clin Pharmacol 72(2):270-281](https://doi.org/10.1111/j.1365-2125.2011.03964.x)

# Sirukumab population PK simulation

Simulate sirukumab concentration-time profiles using the final population
PK model from Xu et al. 2011. Sirukumab (CNTO 136) is a human anti-IL-6
monoclonal antibody developed for inflammatory and autoimmune diseases.
The Xu 2011 study was a double-blind, placebo-controlled, ascending
single-dose first-in-human (FIH) trial in healthy adult volunteers
(study C0524T01). A two-compartment model with zero-order IV input and
first-order elimination from the central compartment adequately
described the serum concentration-time data pooled across dose cohorts.

## Population

From Xu 2011 Table 1: 45 healthy adults were enrolled and 34 received
sirukumab across six cohorts (0.3, 1, 3, and 10 mg/kg in mixed-sex
cohorts of six subjects each except the 10 mg/kg cohort with n = 4, and
separate 6 mg/kg cohorts of six male and six female subjects to
evaluate sex). Baseline demographics (combined sirukumab arm): median
age 30 years (18-54), median weight 71.3 kg (49-99), 16% female,
71% White / 16% Black / 9% Asian / 4% other. The population PK dataset
comprised the 34 sirukumab-treated subjects. No subject developed
antibodies to sirukumab during the 20-week follow-up.

The same information is available programmatically as
`readModelDb("Xu_2011_sirukumab")` (the returned function's body holds
the `population` list literal).

## Source trace

| Element                         | Source location                                 | Value / form                                        |
| ------------------------------- | ----------------------------------------------- | --------------------------------------------------- |
| Structural model                | Xu 2011 Results, "Population PK analysis"       | 2-compartment, zero-order IV input, first-order CL  |
| CL (70 kg subject)              | Xu 2011 Table 4, theta1                         | 0.364 L/day                                         |
| V1 (70 kg subject)              | Xu 2011 Table 4, theta2                         | 3.28 L                                              |
| Q  (70 kg subject)              | Xu 2011 Table 4, theta3                         | 0.588 L/day                                         |
| V2 (70 kg subject)              | Xu 2011 Table 4, theta4                         | 4.97 L                                              |
| Allometric form                 | Xu 2011 Results, allometric scaling paragraph   | TV = theta * (WT/70)^exp; exponents fixed           |
| Allometric exponents CL, Q      | Xu 2011 Results                                 | 0.75 (fixed)                                        |
| Allometric exponents V1, V2     | Xu 2011 Results                                 | 1 (fixed)                                           |
| IIV on CL                       | Xu 2011 Table 4, IIV (%)                        | 24.3% CV (omega^2 = log(CV^2 + 1) = 0.057371)       |
| IIV on V1                       | Xu 2011 Table 4, IIV (%)                        | 19.3% CV (omega^2 = 0.036572)                       |
| IIV on Q                        | Xu 2011 Table 4, IIV (%)                        | 53.4% CV (omega^2 = 0.250880)                       |
| IIV on V2                       | Xu 2011 Table 4, IIV (%)                        | 28.3% CV (omega^2 = 0.077043)                       |
| Off-diagonal omega              | Xu 2011 Table 4                                 | None reported; IIV treated as diagonal              |
| Proportional residual error     | Xu 2011 Table 4, "Proportional error variability (%)" | 21.7 -> propSd = 0.217 (fraction)             |
| Additive residual error         | Xu 2011 Table 4, "Additive error (ug/L)"        | 0.0228 ug/L = 2.28e-5 ug/mL                         |
| Dose regimens                   | Xu 2011 Table 1 and Methods                     | 0.3, 1, 3, 6, 10 mg/kg IV over 10-15 min            |

## Virtual cohort

The per-subject body weights and demographics were not released with the
paper. We build a 34-subject virtual cohort whose cohort sizes match
Table 1 exactly and whose weights are sampled from a log-normal
distribution bounded to the published 49-99 kg range with median
71.3 kg.

```{r virtual-pop}
set.seed(2011)

cohorts <- tribble(
  ~treatment,          ~dose_mgkg, ~n,
  "0.3 mg/kg",            0.3,      6,
  "1 mg/kg",              1.0,      6,
  "3 mg/kg",              3.0,      6,
  "6 mg/kg (male)",       6.0,      6,
  "6 mg/kg (female)",     6.0,      6,
  "10 mg/kg",            10.0,      4
)

pop <- cohorts %>%
  group_by(treatment, dose_mgkg) %>%
  summarise(
    ID = list(seq_len(n[1])),
    .groups = "drop"
  ) %>%
  tidyr::unnest(ID) %>%
  mutate(
    ID = dplyr::row_number(),
    WT = pmin(pmax(rlnorm(dplyr::n(), meanlog = log(71.3), sdlog = 0.15), 49), 99),
    amt = dose_mgkg * WT
  )

pop
```

## Dosing and event table

Sirukumab was given as a single 10-15 min IV infusion. We use a 15 min
infusion (`dur = 15/60/24` day) with all doses delivered at time 0 into
the `central` compartment. Sampling times track the paper's PK-profile
design: dense sampling over the first 48 h, then out through 140 days
(20 weeks) of follow-up.

```{r events}
obs_times <- sort(unique(c(
  seq(0,   1,   by = 1/24),           # hourly on day 1
  seq(1,   7,   by = 0.5),            # twice daily through day 7
  seq(7,  28,   by = 1),              # daily through day 28
  seq(28, 140,  by = 7)               # weekly through week 20
)))

infusion_dur <- 15 / (60 * 24)          # 15 min in days

dose_rows <- pop %>%
  transmute(
    ID, treatment, WT,
    time = 0,
    amt,
    evid = 1L,
    cmt  = "central",
    dur  = infusion_dur,
    dv   = NA_real_
  )

obs_rows <- pop %>%
  select(ID, treatment, WT) %>%
  tidyr::crossing(time = obs_times) %>%
  mutate(
    amt  = NA_real_,
    evid = 0L,
    cmt  = NA_character_,
    dur  = NA_real_,
    dv   = NA_real_
  )

events <- bind_rows(dose_rows, obs_rows) %>%
  arrange(ID, time, desc(evid))
```

## Simulation

Simulate with between-subject variability so the spread across the
virtual cohort matches the paper's individual variability.

```{r simulate}
mod <- readModelDb("Xu_2011_sirukumab")

events_sim <- events %>% rename(id = ID)
sim <- rxSolve(object = mod, events = events_sim, returnType = "data.frame") %>%
  as_tibble() %>%
  left_join(pop %>% select(ID, treatment, dose_mgkg), by = c("id" = "ID"))
```

## Replicate Figure 3: concentration-time profiles by dose

Figure 3 of Xu 2011 shows mean (SD) sirukumab serum concentrations vs.
time by dose cohort on a log-linear scale following a single IV dose.

```{r figure-3, fig.width = 7.5, fig.height = 5}
fig3 <- sim %>%
  filter(time > 0, !is.na(Cc), Cc > 0) %>%
  group_by(treatment, time) %>%
  summarise(
    mean_Cc = mean(Cc, na.rm = TRUE),
    sd_Cc   = sd(Cc,   na.rm = TRUE),
    .groups = "drop"
  )

ggplot(fig3, aes(x = time, y = mean_Cc, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x       = "Time since dose (days)",
    y       = "Sirukumab concentration (ug/mL, log scale)",
    colour  = "Cohort",
    title   = "Figure 3 replication - simulated sirukumab exposure by dose cohort",
    caption = "Replicates Figure 3 of Xu et al. 2011 (mean across virtual cohort; cohort sizes match Table 1)."
  ) +
  theme_bw()
```

# PKNCA validation

Compute non-compartmental Cmax, Tmax, AUC(0,inf), and terminal half-life
per subject per cohort and compare to the published NCA values
(Xu 2011 Table 3). The PKNCA formula groups concentrations by
`treatment + id` so summaries are per-cohort.

```{r pknca-prep}
# Keep t = 0 (Cc = 0 before infusion starts) so PKNCA can integrate AUC from the
# dose time; filter only true NA values.
nca_conc <- sim %>%
  filter(time >= 0, !is.na(Cc)) %>%
  select(id, time, Cc, treatment)

nca_dose <- pop %>%
  transmute(id = ID, time = 0, amt, treatment)
```

```{r pknca-run, message = FALSE, warning = FALSE}
conc_obj <- PKNCAconc(nca_conc, Cc ~ time | treatment + id)
dose_obj <- PKNCAdose(nca_dose, amt ~ time | treatment + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- pk.nca(nca_data)

sim_nca <- as.data.frame(nca_res$result) %>%
  filter(PPTESTCD %in% c("cmax", "aucinf.obs", "half.life")) %>%
  group_by(treatment, PPTESTCD) %>%
  summarise(
    mean = mean(PPORRES, na.rm = TRUE),
    sd   = sd(PPORRES,   na.rm = TRUE),
    med  = median(PPORRES, na.rm = TRUE),
    .groups = "drop"
  )

sim_nca
```

## Comparison against Xu 2011 Table 3

Table 3 of Xu 2011 reports (mean +/- SD for Cmax and AUC(0,inf);
median [range] for terminal half-life):

```{r nca-comparison}
published <- tribble(
  ~treatment,          ~cmax_pub_mean, ~cmax_pub_sd, ~auc_pub_mean, ~auc_pub_sd, ~thalf_pub_median,
  "0.3 mg/kg",           7.9,            1.4,          82.1,          20.4,        20.9,
  "1 mg/kg",            19.2,            1.8,         167.3,          23.6,        19.9,
  "3 mg/kg",            60.1,           14.0,         540.2,         175.2,        18.5,
  "6 mg/kg (male)",    116.3,           11.6,        1225.0,         378.2,        29.6,
  "6 mg/kg (female)",  118.6,           19.2,        1262.0,         307.0,        25.0,
  "10 mg/kg",          248.8,           61.7,        2164.7,         658.5,        21.0
)

sim_cmax <- sim_nca %>% filter(PPTESTCD == "cmax") %>%
  transmute(treatment, cmax_sim_mean = mean, cmax_sim_sd = sd)
sim_auc  <- sim_nca %>% filter(PPTESTCD == "aucinf.obs") %>%
  transmute(treatment, auc_sim_mean = mean, auc_sim_sd = sd)
sim_thalf <- sim_nca %>% filter(PPTESTCD == "half.life") %>%
  transmute(treatment, thalf_sim_median = med)

compare <- published %>%
  left_join(sim_cmax,  by = "treatment") %>%
  left_join(sim_auc,   by = "treatment") %>%
  left_join(sim_thalf, by = "treatment") %>%
  mutate(
    cmax_pct_diff = 100 * (cmax_sim_mean - cmax_pub_mean) / cmax_pub_mean,
    auc_pct_diff  = 100 * (auc_sim_mean  - auc_pub_mean)  / auc_pub_mean
  )

knitr::kable(compare,
             digits  = 2,
             caption = "Simulated NCA vs. Xu 2011 Table 3: Cmax (ug/mL), AUC(0,inf) (ug*day/mL), terminal t1/2 (days).")
```

Dose-proportionality holds exactly by construction in the packaged
model (linear clearance, no absorption nonlinearity), so mean Cmax and
mean AUC scale linearly with the mg/kg dose across the six cohorts.
The paper noted that the observed median terminal half-life ranged
from 18.5 to 29.6 days across cohorts despite no mechanistic dose
dependence; this cohort-level variation is consistent with between-
subject variability at small n rather than a structural effect, and
the simulated median half-life falls within that reported range.

## Assumptions and deviations

* **Weight distribution.** Xu 2011 publishes only summary statistics
  (median 71.3 kg, range 49-99 kg). We draw weights from a log-normal
  distribution with median 71.3 kg and log-scale SD 0.15, clipped to the
  reported range.
* **Sex.** The 6 mg/kg cohorts are stratified by sex in the paper to
  evaluate sex effects, but sex is not a covariate in the final PK
  model. In simulation we treat "6 mg/kg (male)" and "6 mg/kg (female)"
  as cohort labels only; the structural parameters are identical.
* **Time-varying body weight.** The paper describes weight as a
  baseline covariate for allometric scaling. We hold `WT` constant
  over the 20-week follow-up.
* **Residual error magnitude.** Xu 2011 Table 4 lists "Additive error"
  with the units `ug/L` and value 0.0228. We preserve that value by
  converting to the model concentration units (`ug/mL`): 0.0228 ug/L
  = 2.28e-5 ug/mL. The additive term is therefore numerically
  negligible compared with the 21.7% proportional component across
  the observed concentration range, so the overall residual is
  dominated by the proportional term.
* **ADA.** No subjects developed antibodies to sirukumab in the study,
  so ADA is not a model covariate. Populations in which ADA incidence
  is non-trivial (e.g., patients on chronic dosing) are outside the
  validated scope of this model.

## Notes

* **Structural model:** 2-compartment with zero-order IV input into
  `central`, first-order elimination from `central`, and allometric
  weight scaling on CL and Q (exponent 0.75) and on V1 and V2
  (exponent 1) relative to a 70 kg reference.
* **IIV:** diagonal matrix on CL, V1, Q, and V2; no off-diagonal
  terms were reported in the source.
* **Terminal half-life** predicted from the structural parameters for
  a typical 70 kg subject is approximately 21 days (derived from
  CL, V1, Q, V2), consistent with the 18.5-29.6 day range of
  observed cohort-median half-lives reported in Table 3.

## Reference

* `r readModelDb("Xu_2011_sirukumab")()$meta$reference`
