---
title: "Enrofloxacin oral + intravenous popPK in largemouth bass (Xu 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enrofloxacin oral + intravenous popPK in largemouth bass (Xu 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Xu 2025 fitted the oral and the intravenous data **separately**, with different
structural models, so the paper contributes two model files:

* `Xu_2025_enrofloxacin_largemouthBass_oral` --- one-compartment, first-order
  absorption, 20 mg/kg oral gavage.
* `Xu_2025_enrofloxacin_largemouthBass_iv` --- two-compartment, 10 mg/kg
  intravenous bolus into the caudal vein.

```{r citation, echo = FALSE, results = "asis"}
cat("* Citation: ", rxode2::rxode2(readModelDb("Xu_2025_enrofloxacin_largemouthBass_oral"))$reference, "\n", sep = "")
```

* Article: [Animals 2025;15(10):1362](https://doi.org/10.3390/ani15101362)
  (open access, CC BY 4.0; PMC12108383)
* Supplement: none. The article has no supplementary material and the Data
  Availability Statement states that the individual data were not deposited and
  are "available upon request".

```{r descriptions, echo = FALSE, results = "asis"}
for (nm in c("Xu_2025_enrofloxacin_largemouthBass_oral",
             "Xu_2025_enrofloxacin_largemouthBass_iv")) {
  cat("**", nm, "** --- ", rxode2::rxode2(readModelDb(nm))$description, "\n\n", sep = "")
}
```

# Population

Both cohorts are healthy juvenile largemouth bass (*Micropterus salmoides*)
sourced from Huazhong Agricultural University and held at the Yangtze River
Fisheries Research Institute, Wuhan. The purchase lot of 60 fish weighed
243.11 +/- 54.86 g (Section 2.2); no per-fish weights are published. Fish
acclimated for 14 days on antibiotic-free feed in 480 L tanks at 26 L/min, with
water temperature held at 25.0 +/- 0.5 degC by air conditioning and dissolved
oxygen, total ammonia nitrogen, nitrite nitrogen and pH checked daily.

Each route used 24 fish randomised to four groups of six (Sections 2.3.1 and
2.3.2). Oral fish received a single 20 mg/kg gavage of a 20 mg/mL enrofloxacin
solution by plastic tube into the stomach; a fish that regurgitated was removed
and replaced. Intravenous fish received a single 10 mg/kg injection of the same
solution into the caudal vein; a fish that bled heavily or in which the needle
translocated was removed and replaced.

Sampling is **sparse** --- the design that motivates the whole paper. Fish have
too little blood for a full serial profile, so each fish contributes only four
samples, and the four groups interleave to cover 16 nominal times between 5 min
and 120 h:

| Group | Sampling times |
|-------|----------------|
| 1 | 5 min, 1 h, 8 h, 48 h |
| 2 | 10 min, 2 h, 12 h, 72 h |
| 3 | 15 min, 4 h, 16 h, 96 h |
| 4 | 0.5 h, 6 h, 24 h, 120 h |

Plasma enrofloxacin was measured by HPLC with fluorescence detection
(LOQ 0.01 ug/mL; recovery 83.3-103.1%; Table 1). Only the parent was assayed ---
the metabolite ciprofloxacin was not measured in this study.

The same information is available programmatically via each model's
`population` metadata, e.g.
`readModelDb("Xu_2025_enrofloxacin_largemouthBass_iv")()$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its source. They are
collected here for review. Both final models come from the
**"Without importing covariate of wt"** block of their table --- not the
"Importing covariate of wt" block and not the bootstrap block.

## Oral model (`..._oral`)

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (Ka) | 0.98 1/h | Table 4, "Without importing covariate of wt"; restated Sections 3.3, 4, 5 |
| `lvc` (V/F) | 6.82 L/kg | Table 4, same block; restated Sections 3.3, 4, 5 |
| `lcl` (CL/F) | 0.098 L/h/kg | Table 4, same block; restated Sections 3.3, 4, 5 |
| `etalka`, `etalvc`, `etalcl` | `fixed(0)` | Section 2.6 declares exponential IIV; magnitudes never published (see Errata) |
| `propSd` (stdev0) | 0.30 | Table 4, same block |
| `d/dt(depot)`, `d/dt(central)` | n/a | Section 2.6 ("one-compartment model with extravascular administration"); Section 4 confirms first-order absorption |
| Residual form `prop()` | n/a | Section 2.6: `OBSV = IPCN x (1 + eps)` |
| IIV form `exp(eta)` | n/a | Section 2.6: `P_i = tvP x exp(eta_Pi)` |

## Intravenous model (`..._iv`)

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lvc` (V1) | 0.57 L/kg | Table 6, "Without importing covariate of wt"; restated Sections 3.3, 5 |
| `lvp` (V2) | 1.28 L/kg | Table 6, same block; restated Sections 3.3, 5 |
| `lcl` (CL) | 0.012 L/h/kg | Table 6, same block; restated Sections 3.3, 4, 5 |
| `lq` (CL2) | 1.00 L/h/kg | Table 6, same block; restated Sections 3.3, 5 |
| `etalvc`, `etalvp`, `etalcl`, `etalq` | `fixed(0)` | Section 2.6 (IV "parameterizations were consistent with the above"); magnitudes never published |
| `propSd` (stdev0) | 0.14 | Table 6, same block |
| `k12 = q/vc`, `k21 = q/vp` | n/a | Section 2.6: `K12 = tvCL2/V1`, `K21 = tvCL2/V2` |
| `d/dt(central)`, `d/dt(peripheral1)` | n/a | Section 2.6 ("two-compartment model with intravenous administration"); Section 4 confirms no absorption |

## Secondary quantities used as validation targets

These are *not* parameters --- they are quantities Xu 2025 derived from the
parameters above, and they are what the gates in this vignette reproduce.

| Quantity | Published value | Source location |
|----------|-----------------|-----------------|
| Ke (oral) | 0.014 1/h | Section 3.3, 5; `Ke = tvCL/tvV` (Section 2.6) |
| T1/2Ke (oral) | 49.50 h | Section 3.3, 5 |
| AUC0-inf (oral, 20 mg/kg) | 204.08 ug.h/mL | Section 3.3, 5; `AUC = Dose/tvCL` (Section 2.6) |
| Kbeta (IV) | 0.0065 1/h | Section 3.3, 5; `Kbeta = tvCL/(tvV1 + tvV2)` (Section 2.6) |
| T1/2beta (IV) | 106.62 h | Section 3.3, 5 |
| AUC0-inf (IV, 10 mg/kg) | 833.33 ug.h/mL | Section 3.3, 5 |
| Vss (IV) | 1.85 L/kg | Section 4 ("the values of V ... 1.85 L/kg") |
| Bioavailability F | 12.24% | Abstract, Section 4 |
| AUC/MIC | 12755.0 / 816.32 / 408.16 | Section 4 |

# Assembling the two models

```{r models}
mod_oral <- readModelDb("Xu_2025_enrofloxacin_largemouthBass_oral")
mod_iv   <- readModelDb("Xu_2025_enrofloxacin_largemouthBass_iv")

dose_oral <- 20 # mg/kg, Section 2.3.1
dose_iv   <- 10 # mg/kg, Section 2.3.2
```

Both models are body-weight normalised throughout: the dose is mg/kg, volumes
are L/kg and clearances are L/h/kg, so `Cc = central/vc` is mg/L == ug/mL with
no weight column anywhere. That is why body weight is recorded in
`covariatesDataExcluded` rather than `covariateData` --- see Errata.

## The ODE system is live, not silently replaced

`rxode2` will auto-solve a model whose parameters happen to be named `cl`/`vc`
(and `q`/`vp`), discarding the explicit `d/dt()` block. If that happened here
the published `K12`/`K21` micro-constants would be bypassed. The cheap gate is
to collapse the peripheral arm and confirm the profile actually moves: it can
only move if the `d/dt(peripheral1)` equation is being integrated.

```{r ode-live-gate}
grid_iv <- seq(0, 480, by = 0.25)
ev_iv <- rxode2::et(amt = dose_iv, cmt = "central") |> rxode2::et(grid_iv)

sim_iv_typ <- rxode2::rxSolve(mod_iv, ev_iv, returnType = "data.frame")
sim_iv_noq <- rxode2::rxSolve(mod_iv, ev_iv, params = c(lq = log(1e-6)),
                              returnType = "data.frame")

c24_base <- sim_iv_typ$Cc[sim_iv_typ$time == 24]
c24_noq  <- sim_iv_noq$Cc[sim_iv_noq$time == 24]

# Killing intercompartmental clearance must change the 24 h concentration by a
# large margin. Observed here: 4.58 vs 10.58 ug/mL, a factor of 2.3.
stopifnot(c24_noq / c24_base > 1.5)

# The peripheral compartment must also be a genuine state in the solve output.
stopifnot(all(c("central", "peripheral1") %in% names(sim_iv_typ)))
```

# Deterministic typical-value profiles

Xu 2025 published no between-fish variances (Errata), so the etas are held at
zero and the typical-value solve *is* the population prediction. Every
published secondary parameter below is a typical-value quantity, so this is the
right object to validate against --- no cohort averaging is involved.

```{r typical-profiles}
grid_oral <- seq(0, 480, by = 0.25)
ev_oral <- rxode2::et(amt = dose_oral, cmt = "depot") |> rxode2::et(grid_oral)

sim_oral_typ <- rxode2::rxSolve(mod_oral, ev_oral, returnType = "data.frame")

typical <- dplyr::bind_rows(
  sim_oral_typ |> dplyr::transmute(arm = "Oral 20 mg/kg", time, Cc),
  sim_iv_typ   |> dplyr::transmute(arm = "IV 10 mg/kg",   time, Cc)
)
```

```{r figure-profiles}
# Not a replication of a published figure -- Xu 2025 publishes only diagnostic
# plots (Figures 1 and 2 are CWRES-vs-TAD and DV-vs-(I)PRED), and the
# individual concentration data behind them are not available. This is the
# typical-value profile implied by the two published parameter sets.
typical |>
  dplyr::filter(Cc > 1e-4) |>
  ggplot(aes(time, Cc, colour = arm)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Plasma enrofloxacin (ug/mL)",
       colour = NULL,
       title = "Typical-value profiles from the two published parameter sets",
       caption = "Xu 2025 Tables 4 and 6, covariate-free final models.")
```

# Closed-form validation gates

The published secondary parameters are algebraic functions of the estimates, so
they can be reproduced exactly rather than approximately. These gates would
break on any transcription error in the model files.

```{r closed-form-gates}
pars_oral <- sim_oral_typ[1, c("ka", "vc", "cl", "kel")]
pars_iv   <- sim_iv_typ[1, c("cl", "vc", "q", "vp", "kel", "k12", "k21")]

# `Rounding limited` marks the rows where the paper prints too few significant
# figures for the identity to close exactly: the two elimination rate constants
# are given to two significant figures (0.014, 0.0065), and both half-lives
# were then computed from those ROUNDED constants rather than from the
# unrounded ratio. See Errata item 3. The remaining rows are printed to enough
# digits that the identity is exact.
closed_form <- tibble::tribble(
  ~Quantity,                        ~Model,     ~Published, ~Reproduced,                                          ~`Rounding limited`,
  "Ke = CL/V (1/h)",                "oral",     0.014,      pars_oral$kel,                                        TRUE,
  "T1/2Ke = ln2/Ke (h)",            "oral",     49.50,      log(2) / pars_oral$kel,                               TRUE,
  "AUC0-inf = Dose/CL (ug.h/mL)",   "oral",     204.08,     dose_oral / pars_oral$cl,                             FALSE,
  "Vss = V1 + V2 (L/kg)",           "IV",       1.85,       pars_iv$vc + pars_iv$vp,                              FALSE,
  "Kbeta = CL/Vss (1/h)",           "IV",       0.0065,     pars_iv$cl / (pars_iv$vc + pars_iv$vp),               TRUE,
  "T1/2beta = ln2/Kbeta (h)",       "IV",       106.62,     log(2) / (pars_iv$cl / (pars_iv$vc + pars_iv$vp)),    TRUE,
  "K12 = CL2/V1 (1/h)",             "IV",       1.7544,     pars_iv$k12,                                          FALSE,
  "K21 = CL2/V2 (1/h)",             "IV",       0.7812,     pars_iv$k21,                                          FALSE,
  "AUC0-inf = Dose/CL (ug.h/mL)",   "IV",       833.33,     dose_iv / pars_iv$cl,                                 FALSE
) |>
  dplyr::mutate(`% diff` = 100 * (Reproduced - Published) / Published)

closed_form |>
  knitr::kable(digits = 4, caption = "Published secondary parameters vs. the packaged models.")
```

```{r closed-form-assert}
# Every row above is an algebraic identity in the fitted parameters, so the
# only slack is the paper's own rounding, and both tiers are deterministic --
# there is no cohort here, so these bounds do not vary between machines.
#
# Observed: the five exactly-printed rows agree to <= 0.007% (worst: K21,
# 0.0064%), and the four rounding-limited rows to <= 2.64% (worst: Ke, where
# the paper's 0.014 stands for 0.014370). The bounds below sit outside those
# ranges without being unfalsifiable: a single-digit transcription error in any
# theta moves the affected rows by tens of percent.
exact_rows <- !closed_form$`Rounding limited`
stopifnot(
  max(abs(closed_form$`% diff`[exact_rows])) < 0.05,
  max(abs(closed_form$`% diff`[!exact_rows])) < 5
)
```

## Bioavailability --- the cross-model gate

`F = 12.24%` is the one published number that both models must agree on: it is
computed from the oral AUC and the intravenous AUC together, so it cannot be
reproduced by either file alone and it catches a transcription error in either.

```{r bioavailability-gate}
auc_oral <- dose_oral / pars_oral$cl
auc_iv   <- dose_iv   / pars_iv$cl

F_pct <- 100 * (auc_oral / dose_oral) / (auc_iv / dose_iv)

# Equivalently F = CL_iv / (CL/F)_oral, since both AUCs are Dose/CL.
stopifnot(abs(F_pct - 100 * pars_iv$cl / pars_oral$cl) < 1e-8)

# Xu 2025 Abstract and Section 4 report 12.24%. Reproduced: 12.2449%.
stopifnot(abs(F_pct - 12.24) < 0.05)
sprintf("Bioavailability: reproduced %.4f%%, published 12.24%%", F_pct)
```

## The terminal slope actually produced by the two-compartment system

Xu 2025 defines `Kbeta = tvCL/(tvV1 + tvV2)`, which is the steady-state-volume
approximation to the terminal eigenvalue, not the eigenvalue itself. The two
coincide only when distribution is fast relative to elimination, so it is worth
checking that the approximation is benign for *these* estimates --- with a
slower peripheral compartment it would not be, and the published `T1/2beta`
would then not be reproducible from the ODE system at all.

```{r terminal-slope}
tail_iv <- sim_iv_typ |> dplyr::filter(time >= 300, Cc > 0)
beta_true <- -stats::coef(stats::lm(log(Cc) ~ time, data = tail_iv))[["time"]]
beta_paper <- pars_iv$cl / (pars_iv$vc + pars_iv$vp)

# True eigenvalue 0.006449 1/h vs the paper's Vss approximation 0.006486 1/h.
# They differ by 0.57% -- far inside the 6.1% RSE the paper reports for tvCL,
# so the approximation costs nothing here, but they are NOT identical: rounded
# to four decimals the eigenvalue is 0.0064 and the Vss form is 0.0065.
# Both give a terminal half-life within 1% of the published 106.62 h.
stopifnot(
  abs(beta_true - beta_paper) / beta_paper < 0.02,
  abs(log(2) / beta_true - 106.62) / 106.62 < 0.02,
  # The distribution phase must be fast relative to elimination for the Vss
  # form to be benign; k21 / beta is ~120 here. A slow-peripheral model would
  # break this, which is why the check exists.
  pars_iv$k21 / beta_true > 20
)
sprintf("terminal beta: eigenvalue %.6f vs CL/Vss %.6f (t1/2 %.2f vs %.2f h, published 106.62 h)",
        beta_true, beta_paper, log(2) / beta_true, log(2) / beta_paper)
```

# PKNCA validation

The closed-form gates above test the algebra. This section tests the *solved
profiles* independently, by running non-compartmental analysis over them with
PKNCA and comparing against the published AUC and half-life.

```{r pknca-conc}
# Dense grid to 2000 h so aucinf.obs extrapolates a small tail. Cc is the
# individual prediction (no residual error), which is what a typical-value
# NCA should consume.
grid_nca <- seq(0, 2000, by = 0.5)

nca_input <- dplyr::bind_rows(
  rxode2::rxSolve(mod_oral,
                  rxode2::et(amt = dose_oral, cmt = "depot") |> rxode2::et(grid_nca),
                  returnType = "data.frame") |>
    dplyr::transmute(id = 1L, arm = "Oral 20 mg/kg", time, Cc),
  rxode2::rxSolve(mod_iv,
                  rxode2::et(amt = dose_iv, cmt = "central") |> rxode2::et(grid_nca),
                  returnType = "data.frame") |>
    dplyr::transmute(id = 2L, arm = "IV 10 mg/kg", time, Cc)
) |>
  dplyr::filter(!is.na(Cc))

# Guarantee a time = 0 record per (id, arm). The oral solve already carries
# Cc = 0 at t = 0 and the IV solve carries the bolus C0, so this is defensive
# only -- but PKNCA warns per subject if it is ever missing.
nca_input <- dplyr::bind_rows(
  nca_input,
  nca_input |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

stopifnot(sum(nca_input$time == 0) == 2L)
```

```{r pknca-run}
# The grouping column is named `arm`, NOT `route`: PKNCA reserves `route` for
# its own dose-route column and a user column of that name collides with it.
conc_obj <- PKNCA::PKNCAconc(nca_input, Cc ~ time | arm + id)

dose_df <- tibble::tibble(
  id   = c(1L, 2L),
  arm  = c("Oral 20 mg/kg", "IV 10 mg/kg"),
  time = 0,
  amt  = c(dose_oral, dose_iv)
)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published values

```{r nca-compare}
# Xu 2025 publishes AUC0-inf and a terminal half-life per route (Sections 3.3
# and 5) but reports NO Cmax or Tmax for either route, so those two rows have
# no reference value and are omitted from the comparison.
published <- tibble::tribble(
  ~arm,             ~aucinf.obs, ~half.life,
  "Oral 20 mg/kg",  204.08,      49.50,
  "IV 10 mg/kg",    833.33,      106.62
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "arm",
  params        = c("aucinf.obs", "half.life"),
  units         = c(aucinf.obs = "ug*h/mL", half.life = "h"),
  tolerance_pct = 20
)

cmp |>
  knitr::kable(caption = "Simulated (PKNCA) vs. published NCA. * differs from reference by >20%.")
```

```{r nca-assert}
# Assert on the numbers, not on the formatted `% diff` column that
# ncaComparisonTable() returns as character.
nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

chk <- published |>
  dplyr::left_join(nca_wide, by = "arm", suffix = c("_ref", "_sim"))

# AUC0-inf is an identity (Dose/CL) that NCA must recover to within the
# trapezoidal error of the grid. Observed: < 0.02% on both arms.
stopifnot(max(abs(100 * (chk$aucinf.obs_sim - chk$aucinf.obs_ref) / chk$aucinf.obs_ref)) < 1)

# The half-lives are the loosest comparison in the vignette because the
# published values were computed from rounded rate constants (Errata).
# Observed: -2.5% (oral) and +0.8% (IV).
stopifnot(max(abs(100 * (chk$half.life_sim - chk$half.life_ref) / chk$half.life_ref)) < 5)
```

# PK/PD: AUC/MIC against fish pathogens

Section 4 converts the oral AUC0-inf into AUC/MIC ratios against three fish
pathogens, using MIC50 values quoted in the same paragraph, and compares them
against the conventional fluoroquinolone efficacy breakpoint of 125.

```{r auc-mic}
pathogens <- tibble::tribble(
  ~Pathogen,                  ~`MIC50 (ug/mL)`, ~`Published AUC/MIC`,
  "Flavobacterium columnare", 0.016,            12755.0,
  "Aeromonas hydrophila",     0.25,             816.32,
  "Aeromonas sobria",         0.5,              408.16
) |>
  dplyr::mutate(`Reproduced AUC/MIC` = auc_oral / `MIC50 (ug/mL)`)

pathogens |>
  # Per-column digits: the F. columnare MIC50 is 0.016 ug/mL and would render
  # as "0.02" under a scalar digits = 2.
  knitr::kable(digits = c(0, 3, 2, 2),
               caption = "Oral 20 mg/kg AUC0-inf / MIC50 vs. the values in Xu 2025 Section 4.")
```

```{r auc-mic-assert}
# Each ratio is AUC0-inf / MIC50 exactly; the only slack is the paper's
# rounding of 12755.06 to 12755.0. Observed: max |% diff| = 0.0005%.
stopifnot(
  max(abs(100 * (pathogens$`Reproduced AUC/MIC` - pathogens$`Published AUC/MIC`) /
            pathogens$`Published AUC/MIC`)) < 0.1,
  # All three exceed the AUC/MIC >= 125 breakpoint the paper cites.
  all(pathogens$`Reproduced AUC/MIC` >= 125)
)
```

This arithmetic also settles a labelling error in the paper --- see Errata.

# Study-design simulation

For orientation, the figure below overlays the paper's actual sparse-sampling
design on the typical-value profiles, with residual error applied at the
observed times. It is *not* a VPC: with no published between-fish variances
the only stochastic element is the residual error, so the spread below
understates the real between-fish spread that Figures 1 and 2 of the paper show.

```{r design-cohort}
# rxSetSeed() fixes rxode2's stream per solver thread, not across thread
# counts, so this cohort is not byte-identical across machines. Nothing below
# asserts on it.
rxode2::rxSetSeed(20250508)
set.seed(20250508)

sched <- list(
  c(5 / 60, 1, 8, 48),
  c(10 / 60, 2, 12, 72),
  c(15 / 60, 4, 16, 96),
  c(0.5, 6, 24, 120)
)

# 24 fish per arm, matching Sections 2.3.1 / 2.3.2 and well under the
# 200-per-arm cap.
make_arm <- function(cmt, amt, label, id_offset) {
  purrr_free <- lapply(seq_len(24), function(i) {
    times <- sched[[((i - 1L) %% 4L) + 1L]]
    dplyr::bind_rows(
      tibble::tibble(id = id_offset + i, time = 0, amt = amt, evid = 1,
                     cmt = cmt, arm = label),
      tibble::tibble(id = id_offset + i, time = times, amt = NA_real_, evid = 0,
                     cmt = if (cmt == "depot") "central" else cmt, arm = label)
    )
  })
  dplyr::bind_rows(purrr_free)
}

events <- dplyr::bind_rows(
  make_arm("depot",   dose_oral, "Oral 20 mg/kg", id_offset =  0L),
  make_arm("central", dose_iv,   "IV 10 mg/kg",   id_offset = 100L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r design-sim}
sim_design <- dplyr::bind_rows(
  rxode2::rxSolve(mod_oral,
                  events |> dplyr::filter(arm == "Oral 20 mg/kg"),
                  keep = "arm", returnType = "data.frame"),
  rxode2::rxSolve(mod_iv,
                  events |> dplyr::filter(arm == "IV 10 mg/kg"),
                  keep = "arm", returnType = "data.frame")
)

# `Cc` is the individual prediction; `sim` carries the residual error, which is
# what an observed sample would look like.
sim_design |>
  dplyr::filter(time > 0, sim > 0) |>
  ggplot(aes(time, sim)) +
  geom_point(aes(colour = arm), alpha = 0.5, size = 1.4) +
  geom_line(data = dplyr::filter(typical, Cc > 1e-3, time <= 120),
            aes(y = Cc, colour = arm), linewidth = 0.8) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Plasma enrofloxacin (ug/mL)",
       colour = NULL,
       title = "Sparse-sampling design (4 samples/fish, 24 fish/arm)",
       caption = "Points: simulated observations with residual error. Lines: typical-value prediction.")
```

```{r design-assert}
# Nothing here depends on which cohort was drawn: the residual error is
# proportional with SD 0.30 (oral) and 0.14 (IV), so simulated observations
# must stay positive and finite and must bracket the typical prediction.
stopifnot(
  all(is.finite(sim_design$sim)),
  nrow(sim_design |> dplyr::filter(time > 0)) == 2 * 24 * 4
)
```

# Assumptions and deviations

- **Between-fish variability is fixed at zero in both models.** Xu 2025
  Section 2.6 states that exponential IIV, `P_i = tvP x exp(eta_Pi)` with
  `eta ~ N(0, omega^2)`, was estimated on the PK parameters, and defines
  `CV(%) = 100 x sqrt(exp(omega^2) - 1)`. **The omega values are never
  printed anywhere in the paper.** That etas were estimated --- three in the
  oral model, four in the IV model --- is recoverable from the parameter
  counts in Tables 3 and 5 (the covariate-free oral model has 7 parameters =
  3 thetas + 1 stdev0 + 3 omegas; the covariate-free IV model has 9 = 4 + 1 +
  4, and each "-wt" scenario adds exactly one). Per the standing policy for
  unreported variability with structural values present, the etas are carried
  as `~ fixed(0)` rather than invented. **Consequence:** simulating from
  either model gives the typical-value profile, not a population. A user who
  wants a stochastic cohort must supply their own omega.
- **The "Coefficient of Variation (%)" columns of Tables 4 and 6 are relative
  standard errors, not inter-animal CVs, and are not the CV(%) defined in
  Section 2.6.** `0.50/6.82 = 7.3%` reproduces the printed 7.34 for `tvV`, and
  `0.12/1.00 = 12.0%` reproduces the printed 12.47 for `tvCL2`. Reading them
  as IIV would have produced two models with plausible-looking but entirely
  fabricated variances.
- **Body weight is in `covariatesDataExcluded`, not `covariateData`.** It was
  screened on every structural parameter of both models, selected on -2LL/AIC
  (V-wt orally, V2-wt intravenously), fitted, and then rejected in Section 3.3.
  Note also that weight scaling is *already* built into the units --- the dose
  is mg/kg and every volume and clearance is per kg --- so the rejected test
  was for a departure from exponent-1 proportionality, not for the presence of
  any weight effect. No per-fish weights are published, so the centring value
  for `dVdwt`/`dV2dwt` is unrecoverable and the covariate models could not be
  reproduced even if wanted.
- **Two files, not one.** The two routes were fitted separately, to different
  structural models (one-compartment oral, two-compartment IV), on disjoint
  fish. They are not two arms of one fit and are not encoded as one model with
  a route flag. The oral file's `V` and `CL` are apparent (per `F`); the IV
  file's are absolute.
- **The IV bolus is instantaneous.** Section 2.3.2 reports a caudal-vein
  injection from a 1 mL microinjector with no stated duration, so no infusion
  duration is modelled.
- **Only the parent drug is modelled.** The metabolite ciprofloxacin, which
  Xu's own earlier work in other species does measure, was not assayed here.
- Every parameter value in both files comes from the paper's own tables. No
  value was digitised from a figure, obtained by correspondence, or carried
  from an upstream model.

# Errata

Four discrepancies in the published article. None changes a parameter value;
all are recorded so a reader comparing this vignette against the PDF is not
surprised.

1. **Table 6's caption says "after oral administration"; it is the intravenous
   model.** The caption is a copy-paste of Table 4's. Its own Note, its
   parameter names (`tvV1`, `tvV2`, `tvCL2`), the Section 3.3 text that
   introduces it, and the Table 5 scenario list that precedes it all identify
   it as intravenous.

2. **Section 4 misorders the three pathogens against their AUC/MIC values.**
   The text reads "the AUC/MIC values were estimated to be 12755.0, 816.32,
   and 408.16 for *A. hydrophila*, *A. sobria*, and *F. columnare*". Dividing
   the published oral AUC0-inf of 204.08 by the MIC50 values quoted two
   sentences earlier gives `204.08/0.016 = 12755.06` for *F. columnare*,
   `204.08/0.25 = 816.32` for *A. hydrophila* and `204.08/0.5 = 408.16` for
   *A. sobria* --- so the pathogen list is in the reverse of the order the
   numbers require. The `auc-mic` chunk above reproduces all three ratios
   exactly, which is what identifies the pairing. The numbers are right; only
   the labels are swapped, and the paper's conclusion (all three far exceed
   the breakpoint of 125) is unaffected.

3. **The elimination rate constants are printed to two significant figures,
   and both half-lives were then computed from those rounded values.** The
   oral `Ke` is printed as 0.014 1/h where the identity `CL/V = 0.098/6.82`
   gives 0.014370 1/h (2.6% higher), and `T1/2Ke = ln2/0.014 = 49.51 h`
   matches the printed 49.50 h whereas the unrounded constant gives 48.24 h.
   Likewise the IV `Kbeta` is printed as 0.0065 1/h against an unrounded
   `0.012/1.85 = 0.006486 1/h`, and the printed `T1/2beta` of 106.62 h
   corresponds to `ln2/0.0065` rather than to the unrounded 106.86 h. These
   four rows are flagged `Rounding limited` in the closed-form table and gated
   at 5%; the five rows the paper prints at full precision are gated at 0.05%
   and agree to better than 0.007%.

4. **Table 2's Units column is scrambled** (initial naive-pooled estimates,
   not used by either model). It assigns "h" to `V` and `V1`, "/h" to `V2`
   and `CL`, and "L/h/kg" to `CL2`. The correct units are clear from Table 4
   and Table 6: volumes are L/kg and clearances are L/h/kg. No value in either
   model file is taken from Table 2.

Additionally, the paper's Section 2.6 states that both inter-individual **and
inter-occasion** variability were modelled, but every fish receives only a
single dose, so there is no occasion structure to estimate; no IOV term is
reported and none is encoded here.
