---
title: "Rat CNS PBPK for ten drugs (Yamamoto 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rat CNS PBPK for ten drugs (Yamamoto 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
```

```{r libraries, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The paper

Yamamoto and colleagues built a generic, physiologically based model of drug
distribution into the rat central nervous system (CNS) and then, without
fitting anything to the CNS data, used it to *predict* concentration-time
profiles for ten structurally diverse small molecules in brain extracellular
fluid (brain ECF), two cerebrospinal-fluid (CSF) sites and total brain tissue.

> Yamamoto Y, Valitalo PA, Huntjens DR, Proost JH, Vermeulen A, Krauwinkel W,
> Beukers MW, van den Berg DJ, Hartman R, Wong YC, Danhof M, van Hasselt JGC,
> de Lange ECM. Predicting Drug Concentration-Time Profiles in Multiple CNS
> Compartments Using a Comprehensive Physiologically-Based Pharmacokinetic
> Model. *CPT Pharmacometrics Syst Pharmacol.* 2017;6(11):765-777.
> [doi:10.1002/psp4.12250](https://doi.org/10.1002/psp4.12250)

The model is the first release of what the same Leiden group later called
LeiCNS-PK; the mouse successor is already in this package as the
`Saleh_2023_*_mouse_pbpk` family.

What makes the paper unusually reproducible is that **every CNS parameter is
either a rat physiological constant (Table 3) or an in-silico prediction from
the compound's molecular weight, logP and pKa (Table 4)**. Nothing in the CNS
half of the model was estimated. Only the plasma side -- an empirical one-,
two- or three-compartment model per drug -- was fitted, in NONMEM 7.3
(Table 2). The eleven NONMEM control streams are published verbatim as
Supplementary Material S3-S13, which is what this extraction is transcribed
from.

## Model structure

Nine CNS states sit downstream of the plasma model:

| State | Compartment | Rat volume (Table 3) |
|---|---|---|
| `brain_vascular` | brain microvasculature | 60 uL |
| `brain_ecf` | brain extracellular fluid | 290 uL |
| `brain_icf` | brain intracellular fluid | 1440 uL |
| `brain_lysosome` | lysosomes | 18 uL |
| `brain_csf_lv` | CSF, lateral ventricle | 50 uL |
| `brain_csf_tfv` | CSF, third + fourth ventricle | 50 uL |
| `brain_csf_cm` | CSF, cisterna magna | 17 uL |
| `brain_csf_sas` | CSF, subarachnoid space | 180 uL |

Plasma exchanges with `brain_vascular` at cerebral blood flow (1.2 mL/min).
`brain_vascular` is the donor compartment for **both** barriers: the
blood-brain barrier into `brain_ecf`, and the blood-CSF barrier into
`brain_csf_lv` and `brain_csf_tfv`. The CSF then flows in series
(LV -> TFV -> CM -> SAS) and returns to `brain_vascular`, so the CNS subsystem
is closed -- no drug is eliminated inside the brain.

Each barrier clearance is the sum of two routes (Eqs. 3-5):

* **paracellular**, `Qp = (Daq / width) * SA_para`, open only on the
  0.006 % (BBB) or 0.016 % (BCSFB) of surface area that tight junctions leave
  free;
* **transcellular**, `Qt = 0.5 * P0 * SA_trans`, on the remaining 99.8 %, with
  the factor 0.5 correcting for passage across two membranes.

Net active transport is folded into *asymmetry factors* (`AFin1-3`,
`AFout1-3`) that multiply the transcellular term only, and pH partitioning
into *pH-dependent factors* (`PHF1-7`) that scale each efflux clearance by the
ratio of uncharged fractions.

## The ten models

Each drug is a separate model file, because each carries its own plasma model
and its own set of drug-specific CNS parameters.

```{r list-models}
drugs <- c(
  "acetaminophen", "atenolol", "methotrexate", "morphine", "paliperidone",
  "phenytoin", "quinidine", "raclopride", "remoxipride", "risperidone"
)
model_names <- paste0("Yamamoto_2017_", drugs, "_rat_pbpk")
# modellib() returns the model function; rxode2() parses it into the rxUi
# object that rxSolve() and the $theta / $state accessors need.
models <- lapply(model_names, function(nm) suppressWarnings(rxode2(modellib(nm))))
names(models) <- drugs
data.frame(
  Drug = drugs,
  `Model file` = paste0(model_names, ".R"),
  `ODE states` = vapply(models, function(m) length(m$state), integer(1)),
  check.names = FALSE
) |>
  knitr::kable(row.names = FALSE)
```

`quinidine` is shown in full:

```{r show-quinidine}
models$quinidine
```

## Population

All ten datasets are male Wistar rats, 225-275 g, with femoral-artery
sampling and intracerebral / intraventricular microdialysis probes
(Supplementary Material S2). Table 1 of the paper gives the design per drug.

```{r population}
tab1 <- tibble::tribble(
  ~Drug,            ~n,  ~`Dose (mg/kg)`,              ~`Infusion (min)`, ~`CNS sites sampled`,
  "Acetaminophen",  16L, "15",                         "10",  "ECF, CSF-LV, CSF-CM",
  "Atenolol",        5L, "10",                          "1",  "ECF",
  "Methotrexate",   23L, "40, 80",                     "10",  "ECF, CSF-LV, CSF-CM",
  "Morphine",       83L, "4, 10, 40",                  "10",  "ECF",
  "Paliperidone",   21L, "0.5",                        "20",  "ECF, CSF-CM",
  "Phenytoin",      14L, "20, 30, 40",                 "10",  "ECF",
  "Quinidine",      41L, "10, 20",                     "10",  "ECF, CSF-LV, CSF-CM, total brain",
  "Raclopride",     19L, "0.56",                       "10",  "ECF, total brain",
  "Remoxipride",    94L, "0.7, 4, 5.2, 8, 14, 16",     "10 or 30", "ECF, CSF-LV, CSF-CM, total brain",
  "Risperidone",    16L, "2",                          "20",  "ECF, CSF-CM"
)
knitr::kable(tab1)
```

## Source trace

| Model element | Source |
|---|---|
| Nine CNS ODEs, plasma ODEs, total-brain output | Supplementary Material S3-S13, `$DES` and `$ERROR` blocks; Figure 1 |
| `lcl`, `lvc`, `lq`, `lvp`, `lq2`, `lvp2` | Table 2 (CL_PL, V_PL, Q_PL_PER1, V_PER1, Q_PL_PER2, V_PER2) |
| `etalcl`, `etalvc`, `etalq`, `etalvp`, `etalq2` | Table 2 interindividual variability rows, reported as CV; variance = CV^2 |
| `etaiov_1`, `etaiov_2` (morphine) | Table 2 interoccasional variability rows |
| `propSd`, `addSd` | Table 2 residual-error rows |
| `f_ehc`, `t_ehc` (acetaminophen) | Table 2 `Fraction` row (0.693); the 120 min switch is from control stream S3 `$PK` |
| `P0`, `Daq` | Table 4 (transmembrane permeability; aqueous diffusivity coefficient) |
| `AFin1-3`, `AFout1-3` | Table 4; closed forms in Supplementary Material S1 |
| `PHF1-7` | Table 5; Henderson-Hasselbalch Eqs. 10-17 |
| `BF` | Table 5; closed form in Supplementary Material S1 |
| `V_*`, `Q_*`, `SA_*`, `w_BBB` | Table 3 |
| `f_trans`, `f_para_BBB`, `f_para_BCSFB` | Table 3 footnotes b and c (99.8 %, 0.006 %, 0.016 %) |

## Validation 1 -- the combined clearances of Table 5

Table 5 publishes 16 derived clearances per drug. They are not model inputs
here: the model files carry only the *primitives* (`P0`, `Daq`, surface areas,
asymmetry factors, pH factors), so recomputing Table 5 from the encoded
primitives is a genuine end-to-end check of the parameter block and of
Eqs. 3-9.

```{r table5}
t5 <- tibble::tribble(
  ~drug, ~QBBB_in, ~QBBB_out, ~QtBBB, ~QpBBB,
  ~QBCSFB1_in, ~QBCSFB1_out, ~QtBCSFB1, ~QpBCSFB1,
  ~QBCSFB2_in, ~QBCSFB2_out, ~QtBCSFB2, ~QpBCSFB2,
  ~QBCM_in, ~QBCM_out, ~QLYSO_in, ~QLYSO_out,
  "acetaminophen", 0.16, 0.31, 0.014, 0.14, 0.019, 0.038, 6.8e-4, 0.018, 0.019, 0.040, 6.8e-4, 0.018, 0.33, 0.33, 0.16, 0.16,
  "atenolol", 0.12, 0.33, 0.0075, 0.11, 0.014, 0.034, 3.6e-4, 0.014, 0.014, 0.042, 3.6e-4, 0.014, 0.14, 0.068, 0.033, 0.00033,
  "methotrexate", 0.087, 4.8, 8.0e-5, 0.087, 0.011, 2.2, 3.8e-6, 0.011, 0.011, 4.9, 3.8e-6, 0.011, 0.0023, 0.0046, 0.0022, 0.21,
  "morphine", 0.14, 0.38, 0.033, 0.11, 0.015, 0.036, 0.0016, 0.014, 0.015, 0.044, 0.0016, 0.014, 0.61, 0.31, 0.15, 0.0015,
  "paliperidone", 0.33, 0.65, 0.24, 0.090, 0.023, 0.042, 0.011, 0.011, 0.023, 0.052, 0.011, 0.011, 4.4, 2.2, 1.1, 0.011,
  "phenytoin", 1.1, 4.4, 1.0, 0.11, 0.063, 0.38, 0.048, 0.014, 0.063, 0.38, 0.048, 0.014, 23, 23, 11, 11,
  "quinidine", 9.1, 5.1, 7.6, 0.10, 0.52, 0.21, 0.36, 0.013, 0.52, 0.21, 0.36, 0.013, 140, 70, 33, 0.34,
  "raclopride", 0.18, 0.18, 0.086, 0.099, 0.020, 0.013, 0.0041, 0.012, 0.017, 0.016, 0.0041, 0.012, 1.6, 0.80, 0.38, 0.0039,
  "remoxipride", 0.55, 0.69, 0.46, 0.096, 0.034, 0.040, 0.022, 0.012, 0.034, 0.047, 0.022, 0.012, 8.4, 4.4, 2.1, 0.022,
  "risperidone", 1.2, 1.2, 1.1, 0.091, 0.063, 0.063, 0.051, 0.012, 0.063, 0.073, 0.051, 0.012, 20, 10, 4.8, 0.049
)

derived <- function(m) {
  p <- as.list(m$theta)
  QpB <- (p$Daq / p$w_BBB) * p$SA_BBB * p$f_para_BBB
  QtB <- 0.5 * p$P0 * p$SA_BBB * p$f_trans
  Qp1 <- (p$Daq / p$w_BBB) * p$SA_BCSFB1 * p$f_para_BCSFB
  Qt1 <- 0.5 * p$P0 * p$SA_BCSFB1 * p$f_trans
  Qp2 <- (p$Daq / p$w_BBB) * p$SA_BCSFB2 * p$f_para_BCSFB
  Qt2 <- 0.5 * p$P0 * p$SA_BCSFB2 * p$f_trans
  c(
    QBBB_in = QpB + QtB * p$AFin1, QBBB_out = (QpB + QtB * p$AFout1) * p$PHF1,
    QtBBB = QtB, QpBBB = QpB,
    QBCSFB1_in = Qp1 + Qt1 * p$AFin2, QBCSFB1_out = (Qp1 + Qt1 * p$AFout2) * p$PHF2,
    QtBCSFB1 = Qt1, QpBCSFB1 = Qp1,
    QBCSFB2_in = Qp2 + Qt2 * p$AFin3, QBCSFB2_out = (Qp2 + Qt2 * p$AFout3) * p$PHF3,
    QtBCSFB2 = Qt2, QpBCSFB2 = Qp2,
    QBCM_in = p$P0 * p$SA_BCM * p$PHF4, QBCM_out = p$P0 * p$SA_BCM * p$PHF5,
    QLYSO_in = p$P0 * p$SA_LYSO * p$PHF6, QLYSO_out = p$P0 * p$SA_LYSO * p$PHF7
  )
}

chk5 <- lapply(drugs, function(d) {
  pr <- derived(models[[d]])
  pu <- unlist(t5[t5$drug == d, names(pr)])
  tibble::tibble(drug = d, param = names(pr), predicted = unname(pr),
                 published = unname(pu),
                 pct_diff = 100 * (unname(pr) - unname(pu)) / unname(pu))
}) |> bind_rows()

chk5 |>
  group_by(drug) |>
  summarise(
    `n compared` = n(),
    `n within 12%` = sum(abs(pct_diff) < 12),
    `median |% diff|` = round(median(abs(pct_diff)), 1),
    .groups = "drop"
  ) |>
  knitr::kable()
```

```{r table5-outliers}
chk5 |>
  filter(abs(pct_diff) >= 12) |>
  mutate(across(c(predicted, published), ~ signif(.x, 3)),
         pct_diff = round(pct_diff)) |>
  dplyr::rename("Drug" = drug, "Clearance" = param, "Predicted (mL/min)" = predicted,
                "Table 5 (mL/min)" = published, "% diff" = pct_diff) |>
  knitr::kable()
```

```{r table5-gate}
stopifnot(
  # 156 of the 160 published values must reproduce.
  sum(abs(chk5$pct_diff) < 12) >= 156,
  # and the centre of the whole comparison must be tight
  median(abs(chk5$pct_diff)) < 3
)
```

The four survivors are all cells that **Table 5's own neighbouring columns
contradict**, so they are typographical rather than structural:

* Quinidine `QBBB_out` is printed as 5.1, but Table 5 also prints
  `QpBBB` = 0.10, `QtBBB` = 7.6, `PHF1` = 0.80 and Table 4 prints
  `AFout1` = 1.0; its own footnote formula
  `(QpBBB + QtBBB * AFout1) * PHF1` then gives
  `(0.10 + 7.6) * 0.80 = 6.2`, which is what this implementation produces.
* Quinidine `QBCSFB1_out` and `QBCSFB2_out` are both printed as 0.21, against
  `(0.013 + 0.36) * 0.80 = 0.30` from the same row.
* Raclopride `QBCSFB1_in` is printed as 0.020, against
  `QpBCSFB1 + QtBCSFB1 = 0.012 + 0.0041 = 0.016` from the same row -- and the
  structurally identical `QBCSFB2_in` in the same row *is* printed as 0.017.

## Validation 2 -- steady-state Kp,uu against Table 4

This is the stronger gate. The asymmetry factors were *back-solved* by the
authors (Supplementary Material S1) so that the model reproduces the measured
unbound brain-to-plasma ratios at steady state. Those ratios (Table 4) are
therefore an answer key for the whole ODE system: the plasma-to-microvascular
exchange, both barriers, the surface-area fractions, the two-membrane factor,
and the closure of the CSF loop all have to be right for them to come back.

```{r ss-setup}
kpuu_pub <- tibble::tribble(
  ~drug, ~ecf, ~lv, ~cm,
  "acetaminophen", 0.51, 0.51, 0.51,
  "atenolol", 0.37, 0.37, 0.37,
  "methotrexate", 0.018, 0.0066, 0.0024,
  "morphine", 0.38, 0.38, 0.38,
  "paliperidone", 0.50, 0.50, 0.50,
  "phenytoin", 0.26, 0.26, 0.26,
  "quinidine", 1.5, 1.5, 1.5,
  "raclopride", 1.1, 1.1, 1.1,
  "remoxipride", 0.80, 0.80, 0.80,
  "risperidone", 0.97, 0.97, 0.97
)

steady_state <- function(m, pars = NULL) {
  ev <- et(amt = 1e12, rate = 1e6, cmt = "central")
  ev <- add.sampling(ev, c(3e4, 6e4))
  s <- suppressWarnings(rxSolve(zeroRe(m), ev, params = pars,
                                returnType = "data.frame",
                                atol = 1e-12, rtol = 1e-12))
  s[nrow(s), ]
}

kpuu <- lapply(drugs, function(d) {
  pars <- if (d == "morphine") c(DOSE_HIGH = 0, OCC = 1) else NULL
  s <- steady_state(models[[d]], pars)
  tibble::tibble(drug = d,
                 ecf = s$Cbrain_ecf / s$Cc,
                 lv = s$Cbrain_csf_lv / s$Cc,
                 cm = s$Cbrain_csf_cm / s$Cc)
}) |> bind_rows()

kpuu |>
  left_join(kpuu_pub, by = "drug", suffix = c("_pred", "_pub")) |>
  mutate(across(where(is.numeric), ~ signif(.x, 3))) |>
  dplyr::rename("Drug" = drug,
                "brainECF pred" = ecf_pred, "brainECF Table 4" = ecf_pub,
                "CSF-LV pred" = lv_pred, "CSF-LV Table 4" = lv_pub,
                "CSF-CM pred" = cm_pred, "CSF-CM Table 4" = cm_pub) |>
  knitr::kable()
```

```{r ss-gate}
ecf_err <- 100 * abs(kpuu$ecf - kpuu_pub$ecf) / kpuu_pub$ecf
stopifnot(
  # Every drug's brain-ECF Kp,uu reproduces. This is the structural gate:
  # any error in the BBB clearances, the surface-area fractions, the 0.5
  # two-membrane factor or the plasma / microvascular mass balance breaks it.
  max(ecf_err) < 6,
  median(ecf_err) < 3
)
round(ecf_err, 1)
```

All ten brain-ECF ratios return within a few percent -- the residual is
rounding in the two-significant-figure permeabilities and asymmetry factors
that Table 4 publishes. The CSF ratios are looser; see Errata.

## Concentration-time profiles

Typical-value profiles for every drug at its lowest Table 1 dose, in a 250 g
rat (dose in ng = mg/kg x 0.25 kg x 1e6).

```{r profiles, fig.height = 7}
dose_tab <- tibble::tribble(
  ~drug, ~mgkg, ~inf_min, ~tmax_min,
  "acetaminophen", 15, 10, 360,
  "atenolol", 10, 1, 240,
  "methotrexate", 40, 10, 360,
  "morphine", 4, 10, 360,
  "paliperidone", 0.5, 20, 480,
  "phenytoin", 20, 10, 480,
  "quinidine", 10, 10, 360,
  "raclopride", 0.56, 10, 240,
  "remoxipride", 0.7, 10, 240,
  "risperidone", 2, 20, 480
)
BW <- 0.25

profile_one <- function(d) {
  row <- dose_tab[dose_tab$drug == d, ]
  amt <- row$mgkg * BW * 1e6
  ev <- et(amt = amt, rate = amt / row$inf_min, cmt = "central")
  ev <- add.sampling(ev, seq(0, row$tmax_min, length.out = 300))
  pars <- if (d == "morphine") c(DOSE_HIGH = 0, OCC = 1) else NULL
  s <- suppressWarnings(rxSolve(zeroRe(models[[d]]), ev, params = pars,
                                returnType = "data.frame"))
  s |>
    select(time, Plasma = Cc, `Brain ECF` = Cbrain_ecf,
           `CSF-LV` = Cbrain_csf_lv, `CSF-CM` = Cbrain_csf_cm) |>
    pivot_longer(-time, names_to = "Compartment", values_to = "conc") |>
    mutate(drug = d)
}

prof <- bind_rows(lapply(drugs, profile_one))

ggplot(prof, aes(time, conc, colour = Compartment)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~drug, scales = "free", ncol = 3) +
  scale_y_log10() +
  labs(x = "Time (min)", y = "Unbound concentration (ng/mL)",
       title = "Replicates the layout of Supplementary Figure S1") +
  theme_bw(base_size = 9) +
  theme(legend.position = "top")
```

The ordering the paper emphasises is visible: for the efflux-transported
compounds (atenolol, methotrexate, phenytoin) brain ECF sits well below
plasma, while quinidine -- the one compound with net active influx -- sits
above it.

### Lysosomal trapping

The paper adds lysosomes specifically to capture trapping of basic compounds.
Because `PHF7` is the uncharged fraction at lysosomal pH 5.0 relative to
plasma pH 7.4, a strong base accumulates and a neutral compound does not.

```{r lysosome}
lyso <- lapply(c("quinidine", "phenytoin"), function(d) {
  row <- dose_tab[dose_tab$drug == d, ]
  amt <- row$mgkg * BW * 1e6
  ev <- et(amt = amt, rate = amt / row$inf_min, cmt = "central")
  ev <- add.sampling(ev, seq(0, row$tmax_min, length.out = 200))
  s <- suppressWarnings(rxSolve(zeroRe(models[[d]]), ev, returnType = "data.frame"))
  tibble::tibble(time = s$time, drug = d,
                 `Brain ICF` = s$Cbrain_icf, `Brain ECF` = s$Cbrain_ecf)
}) |> bind_rows() |>
  pivot_longer(c(`Brain ICF`, `Brain ECF`), names_to = "Compartment", values_to = "conc")

ggplot(lyso, aes(time, conc, colour = Compartment)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~drug, scales = "free_y") +
  labs(x = "Time (min)", y = "Unbound concentration (ng/mL)") +
  theme_bw()
```

```{r lysosome-gate}
# Quinidine (base, pKa 9.1) concentrates intracellularly relative to ECF;
# phenytoin (neutral, PHF4 = PHF5 = 1) equilibrates 1:1. Structural, not
# stochastic -- a tight bound is correct here.
ratio <- lyso |>
  pivot_wider(names_from = Compartment, values_from = conc) |>
  filter(time > 120) |>
  group_by(drug) |>
  summarise(r = median(`Brain ICF` / `Brain ECF`), .groups = "drop")
stopifnot(
  ratio$r[ratio$drug == "phenytoin"] > 0.98,
  ratio$r[ratio$drug == "phenytoin"] < 1.02,
  # PHF4 / PHF5 = 0.80 / 0.40 = 2 for quinidine
  abs(ratio$r[ratio$drug == "quinidine"] - 2) < 0.05
)
ratio
```

### Enterohepatic recirculation (acetaminophen)

Acetaminophen is the one drug with a thirteenth compartment: control stream S3
returns a fraction 0.693 of eliminated drug to plasma, switched on 120 min
after the dose.

```{r ehc}
amt <- 15 * BW * 1e6
ev <- et(amt = amt, rate = amt / 10, cmt = "central")
ev <- add.sampling(ev, seq(0, 600, length.out = 400))
ehc <- suppressWarnings(rxSolve(zeroRe(models$acetaminophen), ev,
                                returnType = "data.frame"))
ggplot(ehc, aes(time, Cc)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 120, linetype = 2, colour = "grey40") +
  scale_y_log10() +
  labs(x = "Time (min)", y = "Plasma concentration (ng/mL)",
       caption = "Dashed line: t_ehc = 120 min, when recirculation switches on") +
  theme_bw()
```

```{r ehc-gate}
# The recirculation term must actually bend the terminal slope. Compare the
# log-linear decay rate just before and just after the switch.
slope <- function(lo, hi) {
  d <- ehc[ehc$time > lo & ehc$time < hi, ]
  unname(coef(lm(log(Cc) ~ time, d))[2])
}
stopifnot(slope(130, 300) > slope(30, 110))
```

## PKNCA validation

Plasma non-compartmental analysis on the simulated typical-value profiles.
Because the CNS subsystem is closed -- every molecule that crosses either
barrier eventually returns to plasma through the subarachnoid space -- plasma
`AUC(0-inf)` must equal `Dose / CL` exactly. That is a real test of the
transcription: a sign error or a dropped return term anywhere in the nine CNS
ODEs would leak mass and break it.

Acetaminophen is the deliberate exception. Its enterohepatic compartment gives
the drug a second pass through plasma, so integrating the mass balance gives
`AUC = Dose / (CL * (1 - F_ehc))` once recirculation is active. Its AUC must
therefore come out *above* `Dose / CL`, and that excess is itself a check that
the recirculation term is wired up.

```{r pknca}
nca_one <- function(d) {
  row <- dose_tab[dose_tab$drug == d, ]
  amt <- row$mgkg * BW * 1e6
  ev <- et(amt = amt, rate = amt / row$inf_min, cmt = "central")
  # The tail runs to 3 x the plotting window, which is 10-20 terminal
  # half-lives for every drug here: far enough that the extrapolation to
  # infinity is negligible, but not so far that the solver's absolute
  # tolerance starts producing tiny negative concentrations (which make
  # PKNCA's log-trapezoidal rule return NaN).
  ev <- add.sampling(ev, sort(unique(c(
    0, seq(0, row$inf_min, length.out = 10),
    exp(seq(log(1), log(row$tmax_min * 3), length.out = 160))
  ))))
  pars <- if (d == "morphine") c(DOSE_HIGH = 0, OCC = 1) else NULL
  s <- suppressWarnings(rxSolve(zeroRe(models[[d]]), ev, params = pars,
                                returnType = "data.frame"))
  conc <- s |>
    filter(!is.na(Cc)) |>
    transmute(id = 1L, treatment = d, time = time, Cc = Cc)
  dosing <- data.frame(id = 1L, treatment = d, time = 0, dose = amt,
                       duration = row$inf_min)
  o_conc <- PKNCAconc(conc, Cc ~ time | treatment + id,
                      concu = "ng/mL", timeu = "min")
  o_dose <- PKNCAdose(dosing, dose ~ time | treatment + id,
                      duration = "duration", doseu = "ng")
  res <- pk.nca(PKNCAdata(
    o_conc, o_dose,
    intervals = data.frame(start = 0, end = Inf, auclast = TRUE,
                           aucinf.obs = TRUE, cmax = TRUE, tmax = TRUE,
                           half.life = TRUE)
  ))
  as.data.frame(res) |> mutate(treatment = d)
}

nca <- suppressWarnings(bind_rows(lapply(drugs, nca_one)))

nca_tab <- nca |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(
    tibble::tibble(
      treatment = drugs,
      cl = vapply(models, function(m) exp(unname(m$theta["lcl"])), numeric(1)),
      dose = dose_tab$mgkg[match(drugs, dose_tab$drug)] * BW * 1e6
    ),
    by = "treatment"
  ) |>
  mutate(
    `Dose/CL` = dose / cl,
    `% diff` = 100 * (aucinf.obs - `Dose/CL`) / `Dose/CL`
  )

nca_tab |>
  mutate(across(c(cmax, tmax, aucinf.obs, half.life, `Dose/CL`), ~ signif(.x, 4)),
         `% diff` = round(`% diff`, 2)) |>
  select(treatment, cmax, tmax, half.life, aucinf.obs, `Dose/CL`, `% diff`) |>
  dplyr::rename("Drug" = treatment, "Cmax (ng/mL)" = cmax, "Tmax (min)" = tmax,
                "t1/2 (min)" = half.life,
                "AUC0-inf, NCA (ng*min/mL)" = aucinf.obs,
                "Dose/CL (ng*min/mL)" = `Dose/CL`) |>
  knitr::kable()
```

```{r pknca-gate}
closed <- nca_tab[nca_tab$treatment != "acetaminophen", ]
ehc_row <- nca_tab[nca_tab$treatment == "acetaminophen", ]
stopifnot(
  # The nine drugs with a single elimination path: AUC must equal Dose / CL.
  # Trapezoidal AUC on a log-spaced grid, so allow a little numerical slack,
  # but no drug may leak out of the CNS subsystem.
  max(abs(closed$`% diff`)) < 2,
  median(abs(closed$`% diff`)) < 1,
  # Acetaminophen must sit above Dose / CL because 69.3 % of eliminated drug is
  # returned to plasma after 120 min, and below the 226 % excess that would
  # result if recirculation were active from time zero.
  ehc_row$`% diff` > 2,
  ehc_row$`% diff` < 226
)
```

The paper reports no non-compartmental summaries of its own, so there is no
published NCA table to compare against; the comparison above is the internal
mass-balance check.

## A stochastic cohort

Quinidine carries interindividual variability on clearance, intercompartmental
clearance and peripheral volume (Table 2). 100 animals, well inside the
200-per-arm cap.

```{r vpc}
rxSetSeed(20171011)
amt <- 10 * BW * 1e6
ev <- et(amt = amt, rate = amt / 10, cmt = "central")
ev <- add.sampling(ev, seq(0, 360, length.out = 120))
sim <- suppressWarnings(rxSolve(models$quinidine, ev, nSub = 100,
                                returnType = "data.frame"))

sim |>
  select(sim.id, time, Plasma = Cc, `Brain ECF` = Cbrain_ecf,
         `Total brain` = Cbrain_total) |>
  pivot_longer(-c(sim.id, time), names_to = "Compartment", values_to = "conc") |>
  group_by(Compartment, time) |>
  summarise(med = median(conc), lo = quantile(conc, 0.05),
            hi = quantile(conc, 0.95), .groups = "drop") |>
  ggplot(aes(time, med, colour = Compartment, fill = Compartment)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.7) +
  scale_y_log10() +
  labs(x = "Time (min)", y = "Concentration (ng/mL)",
       title = "Quinidine 10 mg/kg, 100 simulated rats (median, 5th-95th percentile)") +
  theme_bw()
```

```{r vpc-gate}
# Structural ordering that must hold for every animal at every time after the
# infusion: total brain > brain ECF for quinidine, because the binding factor
# (7.2) and the lysosomal / intracellular pools all add to the total.
late <- sim[sim$time > 60, ]
stopifnot(all(late$Cbrain_total > late$Cbrain_ecf))
```

## Assumptions and deviations

* **Ten files, one structure.** The authors published one generic CNS model
  and eleven control streams. Each drug becomes one model file, following the
  same shape as the `Saleh_2023_*_mouse_pbpk` family. Morphine's two control
  streams (S6 for 4 mg/kg, S7 for 10 and 40 mg/kg) are a single file with a
  `DOSE_HIGH` covariate, because Table 4 presents them as one drug with
  dose-stratified footnotes a and b.

* **Plasma parameters are taken from Table 2, not from the control streams.**
  The two agree exactly for nine of the ten drugs -- including which
  parameters carry interindividual variability, and with `$OMEGA` equal to the
  square of the tabulated coefficient of variation in every case. See Errata
  for atenolol.

* **The binding factor `BF` does not affect the ODE solution.** The control
  streams define `C4 = A(4)/V4*BF` and then use `C4/BF` at every occurrence in
  `$DES`, so `BF` cancels; it survives only in the total-brain output. This
  implementation uses the unbound concentration directly in the ODEs and
  applies `BF` only in `Cbrain_total`, which is algebraically identical and
  makes the cancellation explicit.

* **`Cbrain_total` is provided for nine drugs, not three.** The authors coded
  the total-brain output only in the three control streams that had total-brain
  data to compare against (quinidine, raclopride, remoxipride). Because `BF` is
  published in Table 5 for nine drugs, the same expression is emitted for all
  nine; methotrexate has `Kp` = NA and `BF` = NA and gets no total-brain output.
  Only the three the authors plotted were validated against data.

* **Rat body weight.** Table 1 gives doses in mg/kg while every volume and flow
  in Table 3 is absolute. This vignette uses 0.25 kg, the midpoint of the
  225-275 g range in Supplementary Material S2, to convert. The model files
  themselves take an absolute amount in ng and are weight-agnostic.

* **`Kp,uu` is relative to total plasma concentration.** The model drives
  barrier transport from the microvascular concentration, which equals the
  total plasma concentration at steady state -- unbound fraction in plasma
  never enters the ODEs. Table 4's `Kp,uu` values are reproduced on that basis
  (Validation 2), so the paper's `fu,p` column is descriptive only and is not
  encoded.

* **Equations 1 and 2 are not re-derived.** The molecular-weight and logP
  regressions for `Daq` and `P0` are reproduced verbatim only in references 16
  and 17, which are not on disk. The tabulated per-drug values of Table 4 are
  encoded directly instead. (They are self-consistent: evaluating
  `Daq = 10^(-4.113 - 0.4609*log10(MW)) * 60` and
  `P0 = 10^(0.939*logP - 6.21) * 60`, the forms the sibling LeiCNS mouse
  papers publish, returns every tabulated value for all ten drugs.)

## Errata and unresolved inconsistencies in the source

1. **Atenolol, Table 2 versus control stream S4.** Table 2 reports
   `V_PL` = 256 mL and no interindividual variability at all; S4 carries
   `THETA(10)` = 259 and `$OMEGA` = 0.02 (a 14.1 % coefficient of variation) on
   `V1`. Table 2 is used here, because it is the paper's reported final
   parameter table and it agrees exactly with the control stream for the other
   nine drugs, and because an `$OMEGA` of exactly 0.02 has the signature of a
   leftover initial estimate -- two sibling streams still carry explicit
   `; initial estimates` comments. Atenolol is therefore a typical-value-only
   plasma model here.

2. **Methotrexate `AFout2`.** Control stream S5 sets `AF22` = 46353.1. That is
   contradicted by three independent readings in the paper: Table 4 prints
   `AFout2` = 4.7 x 10^5, Table 5 prints `QBCSFB1_out` = 2.2 mL/min (which
   requires `AFout2` near 4.4 x 10^5), and the reported
   `Kp,uu,CSFLV` = 0.0066 requires roughly 3.4 x 10^5. The control-stream value
   would give `QBCSFB1_out` = 0.24 mL/min, an order of magnitude out. It reads
   as a misplaced decimal point, so Table 4's 4.7 x 10^5 is used. The two other
   methotrexate efflux factors, `AFout1` and `AFout3`, are taken from the
   control stream because they agree with Table 4.

3. **Table 5 cells contradicted by their own row.** Four of the 160 published
   combined clearances do not follow from Table 5's own components via Table
   5's own footnote formulas: quinidine `QBBB_out`, `QBCSFB1_out` and
   `QBCSFB2_out`, and raclopride `QBCSFB1_in`. Details under Validation 1. The
   primitives are encoded, so this implementation follows the formulas.

4. **CSF `Kp,uu` does not close as tightly as brain-ECF `Kp,uu`.** Running the
   published model to steady state returns Table 4's `Kp,uu,brainECF` within
   6 % for all ten drugs, but the CSF ratios within 10 % for only eight of ten:
   phenytoin (0.16 predicted versus 0.26 published), quinidine (1.72 versus
   1.5) and methotrexate CSF-LV (0.0048 versus 0.0066) deviate. The asymmetry
   factors are *inputs* to the simulation and were reproduced from Table 4, and
   the derived clearances reproduce Table 5 (Validation 1), so this is a
   property of the paper's own back-solution of `AFin2/AFout2` and
   `AFin3/AFout3` rather than of the transcription. The size of the deviation
   tracks how transcellular the blood-CSF barrier clearance is for that
   compound, which is what an error in a transcellular-only asymmetry factor
   would do. The closed forms in Supplementary Material S1 could not be
   evaluated independently: the supplement is a Word document whose equation
   rendering has lost its fraction structure, so numerator and denominator
   terms are run together. Anyone depending on absolute CSF predictions for
   these three compounds should treat them accordingly.

5. **The binding factor does not reproduce Table 4's `Kp`.** Solving the coded
   total-brain expression at steady state implies `Kp` near 7.0 for quinidine,
   5.4 for raclopride and 3.4 for remoxipride, against the 13, 11 and 5.5
   printed in Table 4. `BF` is nonetheless unambiguous -- Table 5 and the
   control streams give the same values (7.2, 8.5, 5.3) -- so it is encoded as
   published. The same Supplementary Material S1 rendering problem prevents
   checking the authors' `BF` derivation.

6. **An undocumented morphine covariate is left at its reference level.**
   Control streams S6 and S7 multiply plasma clearance by `THETA(13)` = 1.08
   when a data column `BLOCK` equals 1. `BLOCK` is defined nowhere in the
   paper, its supplement or Table 2, and Table 2 does not report the
   coefficient. The extraction uses the reference level (`BLOCK` = 0, factor
   1), which is what Table 2 describes.

7. **Author-line correction.** The article carries a publisher notice: it was
   published online 13 October 2017, an error in the author line was
   identified, and the corrected version was posted 27 October 2017. The
   correction does not touch any model value.
