---
title: "Benralizumab (Yan 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benralizumab (Yan 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Yan_2019_benralizumab"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Yan_2019_benralizumab"))$description`
* Article: <https://doi.org/10.1007/s40262-019-00738-4> (open access)

Benralizumab is an interleukin-5 receptor alpha-directed cytolytic monoclonal
antibody approved as add-on maintenance treatment for patients aged 12 years and
older with severe eosinophilic asthma. Yan 2019 is the pooled population PK
analysis of the phase I-III programme that supported that registration.

Three benralizumab population PK models are packaged in nlmixr2lib and they are
**not** interchangeable:

| Model | Scope | Distinguishing features |
|-------|-------|-------------------------|
| `modellib("Wang_2017_benralizumab")` | Early-phase healthy volunteers and patients with asthma | Includes a PD (eosinophil) layer; the ADA covariate uses a HIGH-TITER (>= 400) definition; Japanese-heritage effect on Vc |
| `modellib("Yan_2019_benralizumab")` (this page) | Nine pooled phase I-III asthma studies, 2317 patients | The registrational PK model; plain positive/negative ADA; MI-CP220 study bioavailability stratum |
| `modellib("Jin_2025_benralizumab")` | Twelve pooled studies, 2855 participants, adding Chinese and paediatric data | Updates this model; adds an Asian-race effect on CL, an AMES study bioavailability stratum, and a 200 mg dose effect on F |

Yan 2019 is the direct predecessor of Jin 2025, which calls it the "legacy
model". Where the two are compared below, the agreement is used as an
independent check on how an ambiguous Yan 2019 table row should be read.

# Population

The analysis pooled `r readModelDb("Yan_2019_benralizumab")()$population$n_subjects`
patients with asthma from
`r readModelDb("Yan_2019_benralizumab")()$population$n_studies` clinical studies
(Yan 2019 Table 1): five phase I/II studies (NCT00512486, NCT00659659,
NCT00768079, NCT00783289, and the phase IIb study NCT01238861, which the paper
also calls MI-CP220 or "CP220") and the four phase III studies SIROCCO
(NCT01928771), CALIMA (NCT01914757), ZONDA (NCT02075255) and BISE
(NCT02322775).

Baseline demographics come from Yan 2019 Tables 2 and 3: median age 50 years
(range 12-75), of whom only 57 patients (2.46%) were adolescents aged 12-17;
median body weight 77 kg (range 40-204.4); 63.96% female; 77.73% White, 9.8%
Asian, 4.96% African American and 7.51% Other. Anti-drug antibodies were
detected at some assessed visit in 362 patients (15.62%). No healthy volunteers
were enrolled. Doses spanned 0.0003-3 mg/kg intravenously and 2-200 mg
subcutaneously, but 80% of patients received the registrational 30 mg
subcutaneous dose.

The final updated model was fitted to 14,918 quantifiable serum (plasma for
NCT00659659) concentrations. ZONDA was first held out for external validation
(Yan 2019 Results 3.4, Figure 2) and then appended for the final re-estimation
(Results 3.5).

The same information is available programmatically:

```{r population}
str(readModelDb("Yan_2019_benralizumab")()$population, max.level = 1)
```

# Source trace

Every value below is from the **"Final updated model (Model 9)"** column of
Yan 2019 Table 5 unless stated otherwise. The base-model column of the same
table belongs to Model 5, an earlier eight-study fit with no covariates, and is
not used. The per-parameter origin is also recorded as an in-file comment beside
each `ini()` entry in `inst/modeldb/specificDrugs/Yan_2019_benralizumab.R`.

| Equation / parameter | Value in the model file | Source location |
|----------------------|-------------------------|-----------------|
| Two-compartment structure, first-order SC absorption, first-order elimination from central | `d/dt(depot)`, `d/dt(central)`, `d/dt(peripheral1)` | Methods 2.5.4 and Figure 1 |
| `lcl` | `log(0.291)` | Table 5, "CL, L/day" = 0.291 (90% CI 0.28-0.302) |
| `lvc` | `log(3.13)` | Table 5, "Vc, L" = 3.13 (90% CI 2.97-3.31) |
| `lq` | `log(0.738)` | Table 5, "Q, L/day" = 0.738 (90% CI 0.679-0.803) |
| `lvp` | `log(2.52)` | Table 5, "Vp, L" = 2.52 (90% CI 2.34-2.71) |
| `lka` | `log(log(2) / 3.54)` | Table 5, "ka, half-life; days" = 3.54 (90% CI 3.15-3.99); the paper parameterises the absorption HALF-LIFE, so `ka` is derived |
| `lfdepot` | `log(0.589)` | Table 5, "F" = 0.589 (90% CI 0.565-0.614) |
| `lfdepot_micp220` | `log(0.490)` | Table 5, "Change in F with study CP220, fraction" = 0.490 (90% CI 0.461-0.522) |
| `e_wt_cl` | `0.807` | Table 5, "Body weight on CL, power"; the Figure 4 caption prints the curve as `CL.(body weight/70)^0.807` |
| `e_wt_vc` | `0.803` | Table 5, "Body weight on Vc, power" (90% CI 0.627-0.979) |
| `e_wt_vp` | `0.528` | Table 5, "Body weight on Vp, power" (90% CI 0.351-0.706) |
| (no weight effect on Q) | -- | Table 5, "Body weight on Q, power" = 0 with no confidence interval, i.e. dropped by the Wald's Approximation Method step of Methods 2.5.7 |
| `e_ada_cl` | `log(2.24)` | Table 5, "ADAs on CL, fraction" = 2.24 (90% CI 2.18-2.3); the abstract and Discussion restate it as "+124%" |
| 70 kg reference weight | `(WT / 70)^...` | Methods 2.5.6 ("70 kg for body weight") and the Figure 4 caption |
| Power form for continuous covariates | `(WT / 70)^beta` | Equation 3 |
| Fractional-change form for categorical covariates | `exp(beta * indicator)` | Equation 4 |
| `etalcl` | `0.242^2` | Table 5, "etaCL, %CV" = 24.2 |
| `etalvc` | `0.244^2` | Table 5, "etaVc, %CV" = 24.4 |
| `etalq` | `fixed(0.008)` | Table 4, Model 8, "Fixed IIV(Q) = 0.008"; Table 5 prints the matching 8.94 %CV with no CI |
| `etalvp` | `0.447^2` | Table 5, "etaVp, %CV" = 44.7 |
| `etalka` | `0.831^2` | Table 5, "etaka (half-life), %CV" = 83.1 |
| `etalfdepot` | `0.171^2` | Table 5, "etaF, %CV" = 17.1 |
| `etalfdepot_micp220` | `0.350^2` | Table 5, "etaF (study CP220), %CV" = 35.0 |
| `expSd` | `0.367` | Table 5, "Proportional error (studies SIROCCO, CALIMA, ZONDA, BISE), %CV" = 36.7 |
| `expSd_Cc_early` | `0.250` | Table 5, "Proportional error, %CV" = 25.0; Results 3.2 assigns this row to the phase I-IIa studies |
| `expSd_Cc_micp220` | `0.545` | Table 5, "Proportional error (study CP220), %CV" = 54.5 |
| Log-normal residual error | `Cc ~ lnorm(...)` | Results 3.2, "the PK concentration data were log-transformed ... A log-normal residual error structure was used" |

## Two table-reading decisions worth recording

**Table 5 reports IIV as `100 * omega_SD`, not as an exact lognormal CV.**
Table 4 states that Model 8 holds `IIV(Q) = 0.008`, which in NONMEM is an omega
*variance*. Table 5 prints the corresponding `etaQ` as 8.94 %CV. Since
`100 * sqrt(0.008) = 8.944` while the exact lognormal CV
`100 * sqrt(exp(0.008) - 1) = 8.962` would have printed as 8.96, the tabulated
percentages are standard deviations. `nlmixr2`'s `~` takes a variance, so each
percentage is squared in the model file.

```{r iiv-convention}
c(
  as_sd = 100 * sqrt(0.008),
  as_exact_lognormal_cv = 100 * sqrt(exp(0.008) - 1),
  printed_in_table_5 = 8.94
)
```

**The MI-CP220 bioavailability row is an ABSOLUTE fraction, not a multiplier.**
Table 5 labels the row "Change in F with study CP220, fraction", which reads
like the `exp(theta)` multiplier of Equation 4 -- and for the row immediately
above it ("ADAs on CL, fraction" = 2.24) that is exactly what it is. For F it is
not, and three independent checks agree:

1. The base-model column of the same table gives 0.569 against a reference F of
   0.671, a ratio of 0.848. The successor analysis `Jin_2025_benralizumab`
   refits the same MI-CP220 data and reports `Fa1S220 / Fa1 = 0.457 / 0.539`,
   which is also 0.848. Reading the Yan 2019 row as a multiplier would make the
   relative bioavailability 0.569, not 0.848.
2. The final-model ratio `0.490 / 0.589 = 0.832` stays in the same place. Under
   the multiplier reading the relative bioavailability would have moved from
   0.569 to 0.490 between two fits of overlapping data.
3. The stratum carries its own separately tabulated IIV ("etaF (study CP220),
   %CV" = 35.0 versus 17.1 for the reference F). A separately estimated absolute
   fraction has its own omega; a shift applied to a shared parameter does not.

```{r micp220-adjudication}
tibble::tibble(
  analysis = c(
    "Yan 2019 base model (Model 5)",
    "Yan 2019 final model (Model 9)",
    "Jin 2025 final model"
  ),
  f_reference = c(0.671, 0.589, 0.539),
  f_micp220 = c(0.569, 0.490, 0.457)
) |>
  mutate(ratio = round(f_micp220 / f_reference, 3)) |>
  knitr::kable(
    caption = paste(
      "MI-CP220 bioavailability relative to the reference studies. The ratios",
      "agree across two independent analyses of overlapping data, which the",
      "relative-multiplier reading of the Yan 2019 row could not produce."
    )
  )
```

The three residual-error strata are read the same way, and corroborated the same
way: Yan 2019 prints the phase III stratum as 36.7 %CV and Jin 2025 tabulates
the same stratum as `Error_ADD3, log(ng/mL) = 0.367`.

# Model structure

```{r model}
mod <- readModelDb("Yan_2019_benralizumab")
ui <- rxode2::rxode2(mod)
ui$state
```

The model has three endpoints -- `Cc`, `Cc_early` and `Cc_micp220` -- that share
one prediction and differ only in which study stratum's residual error applies.
Observation rows therefore carry `cmt = "central"` (the ODE state) together with
`dvid = 1L` to select the `Cc` endpoint, and every `rxSolve()` call below passes
`useLinCmt = FALSE`, because rxode2's automatic ODE-to-`linCmt()` conversion
corrupts the `dvid`-to-`cmt` mapping for multi-endpoint models.

```{r endpoints}
stopifnot(
  identical(ui$state, c("depot", "central", "peripheral1")),
  # the explicit ODE system must be solved, not replaced by an analytic solution
  is.null(ui$linCmt)
)
```

# Structural checks against closed forms

These checks zero the random effects, so both sides of every comparison use the
same parameter values and the only difference is numerical integration error.
Tight bounds are therefore correct here; the cohort-level checks further down
use robust statistics instead.

```{r typical-solves}
mod_typ <- rxode2::zeroRe(ui)

# A fine early grid matters: with an absorption half-life of 3.54 days the peak
# has real curvature in the first fortnight, and a coarse grid understates AUC
# by enough to break the dose-recovery identity below.
t_grid <- sort(unique(c(seq(0, 30, by = 0.05), seq(30, 400, by = 0.5))))

make_typical <- function(label, route, wt, ada, micp220, id = 1L) {
  dose <- data.frame(
    id = id, treatment = label, time = 0, amt = 30, evid = 1L,
    cmt = if (route == "sc") "depot" else "central", dvid = NA_integer_
  )
  obs <- data.frame(
    id = id, treatment = label, time = t_grid, amt = NA_real_, evid = 0L,
    cmt = "central", dvid = 1L
  )
  out <- rbind(dose, obs)
  out$WT <- wt
  out$ADA_POS <- ada
  out$STUDY_MICP220 <- micp220
  out
}

arms <- list(
  sc_ref = make_typical("30 mg SC single dose", "sc", 70, 0, 0),
  iv_ref = make_typical("30 mg IV single dose", "iv", 70, 0, 0),
  sc_micp220 = make_typical("30 mg SC, study MI-CP220", "sc", 70, 0, 1),
  sc_ada = make_typical("30 mg SC, ADA positive", "sc", 70, 1, 0),
  sc_wt140 = make_typical("30 mg SC, 140 kg", "sc", 140, 0, 0)
)

# One subject per solve keeps the arms deterministic and independent.
solve_typical <- function(events, model = mod_typ) {
  out <- rxode2::rxSolve(
    model, events,
    omega = NA, useLinCmt = FALSE, returnType = "data.frame"
  )
  out$id <- events$id[1]
  out$treatment <- events$treatment[1]
  out
}

# Trapezoidal AUC to the last grid point plus a log-linear tail extrapolation.
auc_inf <- function(s) {
  n <- nrow(s)
  tail_rows <- (n - 60):n
  lambda_z <- -stats::coef(stats::lm(log(s$Cc[tail_rows]) ~ s$time[tail_rows]))[[2]]
  list(
    aucinf = sum(diff(s$time) * (utils::head(s$Cc, -1) + utils::tail(s$Cc, -1)) / 2) +
      utils::tail(s$Cc, 1) / lambda_z,
    lambda_z = lambda_z
  )
}

typ_solved <- lapply(arms, solve_typical)
typ <- lapply(typ_solved, function(s) {
  c(auc_inf(s), list(cl = s$cl[1], ka = s$ka[1], cmax = max(s$Cc)))
})
```

```{r structural-gates}
# Closed-form terminal half-life of a two-compartment model at the reference
# subject's parameter values (Table 5 estimates, 70 kg, ADA-negative).
terminal_half_life <- function(cl, vc, q, vp) {
  kel <- cl / vc
  k12 <- q / vc
  k21 <- q / vp
  a <- kel + k12 + k21
  log(2) / ((a - sqrt(a^2 - 4 * kel * k21)) / 2)
}
t_half_terminal <- terminal_half_life(0.291, 3.13, 0.738, 2.52)

checks <- tibble::tibble(
  check = c(
    "Dose recovery: cl * AUCinf / (F * Dose)",
    "Absolute bioavailability: AUC(SC) / AUC(IV)",
    "MI-CP220 stratum: AUC(MI-CP220) / AUC(reference)",
    "ADA effect: AUC(ADA+) / AUC(ADA-)",
    "Body weight on CL: AUC(140 kg) / AUC(70 kg)",
    "Absorption half-life recovered from ka",
    "Terminal half-life from the solved profile"
  ),
  simulated = c(
    typ$sc_ref$cl * typ$sc_ref$aucinf / (0.589 * 30),
    typ$sc_ref$aucinf / typ$iv_ref$aucinf,
    typ$sc_micp220$aucinf / typ$sc_ref$aucinf,
    typ$sc_ada$aucinf / typ$sc_ref$aucinf,
    typ$sc_wt140$aucinf / typ$sc_ref$aucinf,
    log(2) / typ$sc_ref$ka,
    log(2) / typ$iv_ref$lambda_z
  ),
  expected = c(
    1,
    0.589,
    0.490 / 0.589,
    1 / 2.24,
    2^-0.807,
    3.54,
    t_half_terminal
  ),
  source = c(
    "mass balance; F and CL from Table 5",
    "Table 5, 'F' = 0.589",
    "Table 5, 0.490 / 0.589",
    "Table 5, 'ADAs on CL, fraction' = 2.24",
    "Table 5, 'Body weight on CL, power' = 0.807",
    "Table 5, 'ka, half-life; days' = 3.54",
    "two-compartment closed form from Table 5"
  )
)

checks |>
  mutate(pct_diff = 100 * (simulated - expected) / expected) |>
  knitr::kable(
    digits = c(NA, 5, 5, NA, 4),
    caption = "Structural checks against closed forms."
  )

stopifnot(all(abs(checks$simulated / checks$expected - 1) < 0.002))
```

The dose-recovery check is worth reading closely. `cl * AUCinf == F * Dose`
holds only if the bioavailable amount reaching `central` is exactly
`fdepot * Dose`, so it simultaneously confirms that `f(depot)` is applied, that
the intravenous route bypasses it, and that the covariate terms multiply CL in
the intended direction.

The MI-CP220 check confirms that the encoding round-trips: a patient flagged
`STUDY_MICP220 = 1` gets exactly `0.490 / 0.589` of the reference exposure. It
does **not** by itself adjudicate the absolute-versus-multiplier reading of the
table row -- that evidence is the cross-analysis ratio agreement shown earlier.

## The gate is not vacuous

A check that cannot fail proves nothing. Perturbing the MI-CP220 bioavailability
to the value the multiplier reading would imply (`0.589 * 0.490 = 0.289`) must
break the ratio check.

```{r mutation-control}
mod_mutated <- ui |>
  rxode2::ini(lfdepot_micp220 = log(0.589 * 0.490)) |>
  rxode2::zeroRe()

as_shipped <- typ$sc_micp220$aucinf / typ$sc_ref$aucinf
if_multiplier <- auc_inf(solve_typical(arms$sc_micp220, mod_mutated))$aucinf /
  typ$sc_ref$aucinf

c(as_shipped = as_shipped, if_multiplier_reading = if_multiplier,
  target = 0.490 / 0.589)

# The shipped model matches the target; the mutated one does not.
stopifnot(
  abs(as_shipped - 0.490 / 0.589) < 0.002,
  abs(if_multiplier - 0.490 / 0.589) > 0.1
)
```

# PKNCA on the typical-value profiles

Yan 2019 publishes no non-compartmental analysis table. The only NCA-comparable
quantity it reports is the terminal elimination half-life, "approximately
15.5 days". The comparison below is run on the typical-value profiles so that it
is fully deterministic and reproducible on any machine, rather than on a random
cohort whose median half-life moves with the draw.

```{r pknca-typical}
typ_conc <- bind_rows(typ_solved[c("sc_ref", "iv_ref")]) |>
  select(id, time, Cc, treatment) |>
  mutate(id = as.integer(factor(treatment)))

# Only `!is.na(Cc)` -- filtering on time or on Cc > 0 would drop the time-zero
# row that PKNCA needs to anchor AUC from 0.
typ_conc <- typ_conc |>
  filter(!is.na(Cc))

typ_conc <- bind_rows(
  typ_conc,
  typ_conc |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(treatment, id, time)

typ_dose <- typ_conc |>
  distinct(id, treatment) |>
  mutate(time = 0, amt = 30)

typ_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(typ_conc, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(typ_dose, amt ~ time | treatment + id),
  intervals = data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
))

typ_nca_wide <- as.data.frame(typ_nca) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

typ_nca_wide |>
  rename(
    "Treatment" = treatment,
    "Cmax (mg/L)" = cmax,
    "Tmax (day)" = tmax,
    "AUC0-inf (mg*day/L)" = aucinf.obs,
    "t1/2 (day)" = half.life
  ) |>
  knitr::kable(digits = 3, caption = "Typical-value NCA (70 kg, ADA negative).")
```

```{r pknca-typical-checks}
auc_sc <- typ_nca_wide$aucinf.obs[typ_nca_wide$treatment == "30 mg SC single dose"]
auc_iv <- typ_nca_wide$aucinf.obs[typ_nca_wide$treatment == "30 mg IV single dose"]
hl_iv <- typ_nca_wide$half.life[typ_nca_wide$treatment == "30 mg IV single dose"]

c(
  f_from_pknca = auc_sc / auc_iv,
  f_published = 0.589,
  half_life_from_pknca = hl_iv,
  half_life_closed_form = t_half_terminal
)

stopifnot(
  # PKNCA must recover the published absolute bioavailability exactly.
  abs(auc_sc / auc_iv - 0.589) < 0.001,
  # ... and must agree with the closed-form terminal half-life. PKNCA's
  # automatic lambda-z window starts slightly inside the distribution phase,
  # which is why this is 1% rather than exact.
  abs(hl_iv / t_half_terminal - 1) < 0.01
)
```

## Comparison against the published value

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,               ~half.life,
  "30 mg IV single dose",   15.5
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = typ_nca,
  reference = published,
  by = "treatment",
  units = c(
    cmax = "mg/L", aucinf.obs = "mg*day/L",
    tmax = "day", half.life = "day"
  ),
  tolerance_pct = 20
)

cmp |>
  knitr::kable(
    caption = paste(
      "Simulated versus published NCA. Yan 2019 reports only the terminal",
      "half-life; * marks a >20% difference from the reference."
    )
  )

stopifnot(!any(grepl("\\*", as.character(unlist(cmp)))))
```

The typical-value terminal half-life is
`r round(t_half_terminal, 2)` days against the paper's "approximately
15.5 days", a difference of
`r sprintf("%.1f%%", 100 * (t_half_terminal - 15.5) / 15.5)`. The two are not
the same quantity: the closed form uses the Table 5 point estimates for a
reference 70 kg ADA-negative patient, whereas the paper's figure is a statistic
over individual patients, whose terminal half-life distribution is strongly
right-skewed. The cohort simulation below shows where 15.5 days sits in that
distribution.

# Virtual cohort

Original observed data are not publicly available. The cohort below matches the
pooled demographics of Yan 2019 Tables 2 and 3: body weight lognormal with
median 77 kg, truncated to the observed 40-204.4 kg range, and 15.62% anti-drug
antibody positive. No patient is assigned to MI-CP220, because the regimens
replicated here are the phase III ones.

```{r cohort}
# `set.seed()` seeds R's RNG for the covariate draws only. rxode2's simulation
# RNG is seeded separately with rxSetSeed() and its streams are partitioned per
# solver thread, so the drawn etas differ between a 2-core CI runner and a
# many-threaded workstation. Every cohort-level assertion below is written
# against a centre or a robust quantile, never against an extreme.
set.seed(20190311)

n_per_arm <- 200L

# meanlog = log(77) reproduces the median; sdlog = 0.2342 reproduces the
# mean/median ratio 79.14 / 77 reported in Table 2.
draw_weight <- function(n) {
  wt <- stats::rlnorm(n, meanlog = log(77), sdlog = 0.2342)
  pmin(pmax(wt, 40), 204.4)
}

make_cohort <- function(n, arm, id_offset = 0L) {
  tibble::tibble(
    id = id_offset + seq_len(n),
    arm = arm,
    WT = draw_weight(n),
    ADA_POS = stats::rbinom(n, 1, 0.1562),
    STUDY_MICP220 = 0
  )
}

cohort <- bind_rows(
  make_cohort(n_per_arm, "30 mg SC Q4W", id_offset = 0L),
  make_cohort(n_per_arm, "30 mg SC @4W+Q8W", id_offset = n_per_arm)
)

stopifnot(
  !anyDuplicated(cohort$id),
  all(cohort$WT >= 40), all(cohort$WT <= 204.4)
)

summary(cohort$WT)
mean(cohort$ADA_POS)
```

# Replicate Figure 3: phase III visual predictive check

Yan 2019 Figure 3 shows the final updated model's VPC for SIROCCO (labelled
study 17) and CALIMA (study 18) under the two registrational regimens: 30 mg
subcutaneously every 4 weeks, and 30 mg subcutaneously every 4 weeks for the
first three doses followed by every 8 weeks. The published panels run to roughly
56 weeks on a 1-5000 ng/mL log axis.

```{r simulate-phase3}
week <- 7
tau_q4w <- 4 * week
tau_q8w <- 8 * week

q4w_times <- seq(0, by = tau_q4w, length.out = 14L)
q8w_times <- c(
  0, tau_q4w, 2 * tau_q4w,
  seq(2 * tau_q4w + tau_q8w, by = tau_q8w, length.out = 5L)
)
dose_times <- list("30 mg SC Q4W" = q4w_times, "30 mg SC @4W+Q8W" = q8w_times)

last_dose_time <- c(
  "30 mg SC Q4W" = max(q4w_times),
  "30 mg SC @4W+Q8W" = max(q8w_times)
)
interval_length <- c("30 mg SC Q4W" = tau_q4w, "30 mg SC @4W+Q8W" = tau_q8w)

# Dense through the first absorption phase for the plot, and dense again over
# the final dosing interval of both arms so that the steady-state AUC below is
# integrated accurately. The absorption-rate IIV is large (omega 0.831 on the
# log scale), so a fast absorber peaks early and sharply.
obs_times <- sort(unique(c(
  seq(0, 28, by = 0.5),
  seq(28, 56 * week, by = 3.5),
  seq(min(last_dose_time), 56 * week, by = 0.25)
)))

doses <- cohort |>
  group_by(arm) |>
  group_modify(~ tidyr::expand_grid(.x, time = dose_times[[.y$arm]])) |>
  ungroup() |>
  mutate(amt = 30, evid = 1L, cmt = "depot", dvid = NA_integer_)

obs <- cohort |>
  tidyr::expand_grid(time = obs_times) |>
  mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)

events <- bind_rows(doses, obs) |> arrange(id, time, desc(evid))

rxode2::rxSetSeed(20190311)
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("arm", "WT", "ADA_POS"),
  useLinCmt = FALSE,
  returnType = "data.frame"
) |>
  mutate(Cc_ngml = Cc * 1000, week = time / 7)

stopifnot(nrow(sim) > 0, all(sim$Cc_ngml >= 0), all(is.finite(sim$Cc_ngml)))
```

```{r figure-3, fig.cap = "Replicates Figure 3 of Yan 2019: VPC of benralizumab serum concentration under the two phase III regimens."}
sim |>
  filter(time > 0) |>
  group_by(arm, week) |>
  summarise(
    Q05 = quantile(Cc_ngml, 0.05),
    Q50 = quantile(Cc_ngml, 0.50),
    Q95 = quantile(Cc_ngml, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(week, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time after first dose (weeks)",
    y = "Benralizumab concentration (ng/mL)",
    title = "Figure 3 - phase III VPC",
    caption = paste(
      "Replicates Figure 3 of Yan 2019. Band is the 5th-95th percentile of",
      "model-predicted exposure."
    )
  )
```

Yan 2019 publishes no digitised percentile values for Figure 3, so the
assertions below test structural features plus an exact steady-state mass
balance rather than band widths.

```{r figure-3-checks}
ss_window <- sim |> filter(week >= 40)

trough_by_arm <- ss_window |>
  group_by(arm, id) |>
  summarise(trough = min(Cc_ngml), .groups = "drop") |>
  group_by(arm) |>
  summarise(median_trough = median(trough), .groups = "drop")

trough_by_arm

# Per-patient steady-state mass balance over the final dosing interval:
# cl_i * AUCtau_i == fdepot_i * Dose exactly once steady state is reached.
# Both sides use the SAME drawn parameters, so the only discrepancy is
# trapezoidal integration error and a tight bound is the correct one.
ss_recovery <- sim |>
  mutate(
    window_start = last_dose_time[arm],
    window_end = last_dose_time[arm] + interval_length[arm]
  ) |>
  filter(time >= window_start, time <= window_end) |>
  group_by(arm, id, cl, fdepot) |>
  arrange(time, .by_group = TRUE) |>
  summarise(
    auc_tau = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
    .groups = "drop"
  ) |>
  mutate(recovery = cl * auc_tau / (fdepot * 30))

summary(ss_recovery$recovery)

stopifnot(
  # Exposure stays inside the decade range of the published panels. Robust
  # quantiles, not extremes: a single low-clearance patient may sit above the
  # plotted axis maximum without that being a model defect.
  quantile(sim$Cc_ngml[sim$week > 1], 0.95) < 5000,
  quantile(sim$Cc_ngml[sim$week > 1], 0.05) > 1,
  # The same dose given Q4W accumulates to a higher trough than given Q8W.
  trough_by_arm$median_trough[trough_by_arm$arm == "30 mg SC Q4W"] >
    trough_by_arm$median_trough[trough_by_arm$arm == "30 mg SC @4W+Q8W"],
  # Steady-state mass balance holds for every simulated patient.
  all(abs(ss_recovery$recovery - 1) < 0.01)
)
```

# Replicate Figure 4: body weight and ADA effects on clearance

Figure 4 plots individual benralizumab CL against body weight, with closed
circles for ADA-negative and open circles for ADA-positive patients, and a red
prediction curve the caption prints as `CL.(body weight/70)^0.807`.

```{r figure-4, fig.cap = "Replicates Figure 4 of Yan 2019: individual clearance versus body weight, split by ADA status."}
cl_by_subject <- sim |>
  distinct(id, arm, WT, ADA_POS, cl)

curve_df <- tibble::tibble(WT = seq(40, 205, length.out = 200)) |>
  mutate(cl = 0.291 * (WT / 70)^0.807)

ggplot(cl_by_subject, aes(WT, cl)) +
  geom_point(aes(shape = factor(ADA_POS)), alpha = 0.45) +
  geom_line(data = curve_df, colour = "red", linewidth = 1) +
  scale_shape_manual(
    values = c(`0` = 16, `1` = 1),
    labels = c(`0` = "ADA negative", `1` = "ADA positive"),
    name = NULL
  ) +
  labs(
    x = "Body weight (kg)",
    y = "Benralizumab CL (L/day)",
    title = "Figure 4 - body weight and ADA effects on CL",
    caption = "Replicates Figure 4 of Yan 2019. Red line is CL * (body weight / 70)^0.807."
  )
```

The scatter in the plot is inter-individual variability on clearance
(`etalcl`, 24.2 %CV), exactly as in the published figure, where the points are
individual post-hoc estimates. Two checks are therefore needed: an exact one on
the curve itself, and a robust one on the cohort.

```{r figure-4-curve-check}
# Exact check. With the random effects zeroed, the model's clearance must equal
# CL * (WT/70)^0.807 * 2.24^ADA_POS at every weight and both ADA states. This is
# deterministic, so it is checked to machine precision.
curve_grid <- tidyr::expand_grid(
  WT = c(40, 55, 70, 90, 120, 160, 204.4),
  ADA_POS = c(0, 1)
) |>
  mutate(id = row_number(), STUDY_MICP220 = 0)

curve_events <- bind_rows(
  curve_grid |> mutate(time = 0, amt = 30, evid = 1L, cmt = "depot", dvid = NA_integer_),
  curve_grid |>
    tidyr::expand_grid(time = c(1, 2)) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  arrange(id, time, desc(evid))

# The "multi-subject simulation without 'omega'" warning is expected and wanted
# here: the solve is deliberately deterministic across several subjects.
curve_check <- suppressWarnings(rxode2::rxSolve(
  mod_typ, curve_events,
  omega = NA, useLinCmt = FALSE,
  keep = c("WT", "ADA_POS"), returnType = "data.frame"
)) |>
  distinct(id, WT, ADA_POS, cl) |>
  mutate(published_curve = 0.291 * (WT / 70)^0.807 * 2.24^ADA_POS)

max(abs(curve_check$cl / curve_check$published_curve - 1))

stopifnot(
  nrow(curve_check) == nrow(curve_grid),
  all(abs(curve_check$cl / curve_check$published_curve - 1) < 1e-8)
)
```

```{r figure-4-cohort-check}
# Robust check on the cohort. Dividing out the weight effect leaves
# exp(etalcl) * 2.24^ADA_POS, whose median is 1 for ADA-negative patients and
# 2.24 for ADA-positive ones. Medians, not extremes -- individual points are
# spread by the 24.2 %CV clearance IIV.
normalised <- cl_by_subject |>
  mutate(cl_norm = cl / (0.291 * (WT / 70)^0.807)) |>
  group_by(`ADA status` = ifelse(ADA_POS == 1, "ADA positive", "ADA negative")) |>
  summarise(n = n(), median_normalised_cl = median(cl_norm), .groups = "drop")

normalised

med_neg <- normalised$median_normalised_cl[normalised$`ADA status` == "ADA negative"]
med_pos <- normalised$median_normalised_cl[normalised$`ADA status` == "ADA positive"]

stopifnot(
  all(normalised$n > 10),
  # median(exp(etalcl)) = 1 by construction; the bound covers sampling noise.
  abs(med_neg - 1) < 0.10,
  # ... and the ADA-positive group sits 2.24-fold above it.
  abs(med_pos / 2.24 - 1) < 0.20
)
```

Yan 2019 Figure 5 is the complementary negative result: benralizumab CL is flat
at 0.30-0.31 L/day across strata of estimated GFR and of baseline blood
eosinophil count. That figure needs no simulation to reproduce -- eGFR and
eosinophil count are absent from the final model, so CL is invariant to them by
construction. Both are recorded in `covariatesDataExcluded` with the paper's
reasoning, including why the eosinophil result is mechanistically load-bearing
for a cytolytic anti-eosinophil antibody.

# Cohort NCA and the terminal half-life

A separate single-dose cohort is used for the cohort-level NCA, because a
multiple-dose profile has no clean terminal phase. Both routes are simulated at
30 mg so the AUC ratio recovers the absolute bioavailability under full
between-subject variability.

```{r simulate-single-dose}
sd_cohort <- bind_rows(
  make_cohort(n_per_arm, "30 mg SC single dose", id_offset = 1000L),
  make_cohort(n_per_arm, "30 mg IV single dose", id_offset = 1000L + n_per_arm)
)

sd_obs_times <- sort(unique(c(seq(0, 28, by = 0.25), seq(28, 180, by = 2))))

sd_doses <- sd_cohort |>
  mutate(
    time = 0, amt = 30, evid = 1L, dvid = NA_integer_,
    cmt = ifelse(arm == "30 mg IV single dose", "central", "depot")
  )

sd_events <- bind_rows(
  sd_doses,
  sd_cohort |>
    tidyr::expand_grid(time = sd_obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dvid = 1L)
) |>
  arrange(id, time, desc(evid))

rxode2::rxSetSeed(20190312)
sim_sd <- rxode2::rxSolve(
  mod, sd_events,
  keep = c("arm", "WT", "ADA_POS"),
  useLinCmt = FALSE,
  returnType = "data.frame"
)

stopifnot(nrow(sim_sd) > 0, all(is.finite(sim_sd$Cc)))
```

```{r pknca-cohort}
sim_nca <- sim_sd |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment = arm)

# Guarantee a time = 0 record per patient. For the subcutaneous arm the
# pre-dose concentration is genuinely zero; for the intravenous arm rxSolve
# already returns the post-bolus value at time 0, which `distinct()` keeps.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, treatment) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, treatment, time, .keep_all = TRUE) |>
  arrange(id, treatment, time)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id),
  PKNCA::PKNCAdose(sd_doses |> select(id, time, amt, treatment = arm),
    amt ~ time | treatment + id
  ),
  intervals = data.frame(
    start = 0, end = Inf,
    cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
  )
))

nca_df <- as.data.frame(nca_res)

nca_df |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "half.life")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename(
    "Treatment" = treatment,
    "Cmax (mg/L)" = cmax,
    "Tmax (day)" = tmax,
    "AUC0-inf (mg*day/L)" = aucinf.obs,
    "t1/2 (day)" = half.life
  ) |>
  knitr::kable(digits = 3, caption = "Median simulated NCA parameters over the cohort.")
```

```{r pknca-cohort-checks}
# Per-patient mass balance from PKNCA's AUCinf on the subcutaneous arm. Same
# drawn parameters on both sides, so this stays tight.
sc_recovery <- nca_df |>
  filter(PPTESTCD == "aucinf.obs") |>
  inner_join(sim_sd |> distinct(id, arm, cl, fdepot), by = c("id", "treatment" = "arm")) |>
  filter(treatment == "30 mg SC single dose") |>
  mutate(recovery = cl * PPORRES / (fdepot * 30))

auc_median <- nca_df |>
  filter(PPTESTCD == "aucinf.obs") |>
  group_by(treatment) |>
  summarise(median_auc = median(PPORRES, na.rm = TRUE), .groups = "drop")

f_recovered <- auc_median$median_auc[auc_median$treatment == "30 mg SC single dose"] /
  auc_median$median_auc[auc_median$treatment == "30 mg IV single dose"]

c(
  median_sc_mass_balance = median(sc_recovery$recovery),
  f_recovered = f_recovered,
  f_published = 0.589
)

stopifnot(
  nrow(sc_recovery) == n_per_arm,
  abs(median(sc_recovery$recovery) - 1) < 0.005,
  quantile(abs(sc_recovery$recovery - 1), 0.95) < 0.02,
  # The two arms are different patients with independently drawn etas, so the
  # ratio of their median AUCs is a cohort statistic and carries sampling
  # noise; a loose relative bound is the honest one here.
  abs(f_recovered / 0.589 - 1) < 0.12
)
```

## Where the published 15.5 days sits

```{r half-life-distribution}
individual_half_life <- sim_sd |>
  distinct(id, arm, ADA_POS, cl, vc, q, vp) |>
  mutate(t_half = terminal_half_life(cl, vc, q, vp))

hl_summary <- individual_half_life |>
  group_by(`ADA status` = ifelse(ADA_POS == 1, "ADA positive", "ADA negative")) |>
  summarise(
    n = n(),
    `median t1/2 (day)` = median(t_half),
    `mean t1/2 (day)` = mean(t_half),
    .groups = "drop"
  )

hl_summary |>
  knitr::kable(digits = 2, caption = "Individual terminal half-life by ADA status.")

c(
  typical_value_ada_negative = t_half_terminal,
  published = 15.5
)
```

Three facts, in order of how much they matter:

1. Anti-drug antibodies raise CL 2.24-fold, which roughly halves the terminal
   half-life for the 15.62% of patients who develop them. The cohort split above
   shows the two groups separately.
2. Within the ADA-negative group the terminal half-life is right-skewed -- the
   mean sits above the median, and both sit above the 70 kg typical-value
   computation of `r round(t_half_terminal, 2)` days. The published 15.5 days
   falls inside that ADA-negative range.
3. The Discussion reports the same statistic stratified by age: "median
   estimates of approximately 14 days in adolescents, 15 days in adults
   (18-64 years of age), and 17 days in older adults". A single "approximately
   15.5 days" summarising that spread is consistent with what the model
   produces; it is not the typical-subject value.

```{r half-life-checks}
ada_neg_hl <- individual_half_life$t_half[individual_half_life$ADA_POS == 0]
ada_pos_hl <- individual_half_life$t_half[individual_half_life$ADA_POS == 1]

stopifnot(
  length(ada_neg_hl) > 50, length(ada_pos_hl) > 10,
  # ADA positivity shortens the terminal half-life (2.24-fold higher CL).
  median(ada_neg_hl) > median(ada_pos_hl),
  # The distribution is right-skewed, so the mean exceeds the median.
  mean(ada_neg_hl) > median(ada_neg_hl),
  # The published 15.5 days is within 25% of the ADA-negative cohort median.
  # Centre, not extreme, and loose enough to survive a different random draw.
  abs(median(ada_neg_hl) / 15.5 - 1) < 0.25,
  # The deterministic closed form is pinned exactly.
  abs(t_half_terminal - 14.62) < 0.05
)
```

# Assumptions and deviations

* **No published NCA table.** Yan 2019 reports model parameters, VPC figures and
  covariate-effect figures, but no Cmax / Tmax / AUC table. The comparison table
  above therefore has a single reference value, the terminal half-life, and the
  weight of the validation rests on the closed-form structural checks, the
  steady-state mass balance and the figure replications.
* **The MI-CP220 bioavailability row is read as an absolute fraction.** Table 5
  labels it "Change in F with study CP220, fraction", the same wording as the
  ADA row directly above it where the number *is* a multiplier. The reasoning
  for reading this one differently is set out in full under "Two table-reading
  decisions worth recording", and rests on the ratio agreement with the
  independent Jin 2025 refit plus the stratum's own separately tabulated IIV.
  This is the single interpretive decision in the extraction that a reviewer
  should check first.
* **IIV percentages are read as `100 * omega_SD`.** Pinned by the `IIV(Q)` value
  that Table 4 states as a variance and Table 5 prints as a percentage. Note
  that the successor paper Jin 2025 uses the *other* convention (the exact
  lognormal CV) for the same quantity, so the two tables cannot be compared
  percentage-to-percentage.
* **Residual error is encoded as `lnorm()`.** Yan 2019 Equation 2 writes a
  combined proportional-plus-additive error, but Results 3.2 states the data
  were log-transformed and a log-normal residual error used, and Table 5 reports
  only proportional components. Additive error on log-concentration is exactly
  `lnorm()`. The additive component of Equation 2 has no published estimate and
  is not carried.
* **No body weight effect on Q.** Table 5 lists "Body weight on Q, power" as 0
  with no confidence interval, i.e. the Wald's Approximation Method step dropped
  it. No zero-valued exponent is encoded; Q simply carries no covariate.
* **Intravenous doses are simulated as bolus injections.** The phase I/II
  studies administered benralizumab by intravenous infusion, but Yan 2019 does
  not report infusion durations and the model carries no `dur()` term. AUCinf is
  independent of infusion duration, so the bioavailability and half-life checks
  are unaffected; the simulated intravenous Cmax is an upper bound and is not
  compared against anything published.
* **Covariate distributions are assumed.** Body weight is drawn lognormal with
  the median and mean/median ratio of Yan 2019 Table 2 and truncated to the
  observed range; ADA status is drawn Bernoulli at the pooled 15.62% rate of
  Table 3 and held constant per patient, whereas the source dataset assessed it
  at each visit and therefore carried it as time-varying. The published race,
  sex and age distributions are recorded in the model's `population` metadata
  but are not simulated, because none of them entered the final model.
* **The VPC band is model-predicted exposure without residual error.** `Cc` from
  `rxSolve()` is the individual prediction; the published Figure 3 bands are
  prediction intervals that include the study-stratified residual error. The
  simulated band is therefore narrower than the published one, which is why the
  Figure 3 assertions test structural features rather than band widths.
* **Typographical error in the source.** Yan 2019 Tables 2 and 3 print
  "NCT00768079" twice in their column headers. The fourth column (n = 19) is
  NCT00783289, the fifth study named in Table 1 and in Results 3.2. This affects
  only the demographic tables, not any model parameter.
* **No supplement and no erratum.** EuropePMC records no supplementary files and
  no correction notice for PMC6584252, and the article text references neither.
  Every value used here comes from the main text and its tables.
