---
title: "Yang_2021_cemiplimab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Yang_2021_cemiplimab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Yang F, Paccaly AJ, Rippley RK, Davis JD, DiCioccio AT.
  Population pharmacokinetic characteristics of cemiplimab in patients
  with advanced malignancies. *J Pharmacokinet Pharmacodyn.*
  2021;48(4):479-494.
  doi:[10.1007/s10928-021-09739-y](https://doi.org/10.1007/s10928-021-09739-y)
* Description: Two-compartment population PK model for cemiplimab
  (anti-PD-1 IgG4) with time-varying clearance (sigmoid Emax) in
  adults with advanced solid tumors including cutaneous squamous
  cell carcinoma (CSCC).
* Modality: Therapeutic monoclonal antibody (IgG4), 30-min IV
  infusion.

Cemiplimab is a fully human anti-PD-1 IgG4 monoclonal antibody. Yang
2021 develops the final population PK model used to support fixed-dose
350 mg Q3W selection, pooling 11,178 PK observations from 548 patients
across two studies (1423 and 1540).

Structure: linear two-compartment IV model with time-varying CL via a
sigmoid-Emax function of time since first dose:

\[
\mathrm{CL}_{t,i} \;=\; \mathrm{CL}_{\mathrm{base},i} \cdot
  \exp\!\left( \dfrac{E_{\max,i}\, t^{\gamma}}
                     {T_{50,i}^{\gamma} + t^{\gamma}} \right),
\qquad
E_{\max,i} = E_{\max,\mathrm{REF}} \cdot \exp(\eta_{E_{\max},i}),
\qquad
T_{50,i}   = T_{50,\mathrm{REF}}   \cdot \exp(\eta_{T_{50},i})
\]

with \(E_{\max} = -0.410\) (a log-fold maximal decrease, asymptote
exp(\(E_{\max}\)) \(\approx 0.66\) of baseline CL) and
\(T_{50} = 28.9\) days. Covariate effects (all power-form) on the
shared CL/Q: baseline body weight, albumin, IgG, ALT. On the shared
V2/V3: baseline body weight, BMI.

# Population

The final-model population comprised **548 patients** across 2
trials (Yang 2021 Table 1):

- Phase 1 first-in-human Study 1423 (NCT02383212; 396 patients with
  advanced malignancies; weight-based and fixed dosing arms).
- Phase 2 Study 1540 (NCT02760498; 152 patients with advanced
  cutaneous squamous cell carcinoma).
- Dose range: 1, 3, or 10 mg/kg Q2W; 3 mg/kg Q3W; 200 mg Q2W; 350 mg
  Q3W (all 30-min IV infusions).

Baseline demographics (Yang 2021 Table 2):

- Median age 65 (range 27-96) years.
- 60.4% male, 39.6% female.
- Race: 90.9% White, 3.6% Black, 1.6% Asian, 3.8% Other.
- Median weight 76.2 (range 30.9-172) kg.
- Median BMI 26.5 (range 14.8-56.3) kg/m^2.
- Median albumin 38.0 (range 22.0-48.0) g/L.
- Median IgG 9.65 (range 1.29-27.9) g/L.
- Median ALT 20.0 (range 5.00-196) IU/L.

Most patients (80.3%) received cemiplimab 3 mg/kg Q2W. The same
metadata is available programmatically via
`readModelDb("Yang_2021_cemiplimab")$population`.

# Source trace

Per-parameter origins are recorded as in-file comments next to each
`ini()` entry in `inst/modeldb/specificDrugs/Yang_2021_cemiplimab.R`.
The table below collects them in one place for review.

| Parameter (model name)               | Value           | Source                                                            |
|--------------------------------------|-----------------|-------------------------------------------------------------------|
| `lcl` (CL_BASE,REF, L/day)           | log(0.290)      | Yang 2021 Table 3, TVCL                                           |
| `lvc` (V2_REF, L)                    | log(3.32)       | Yang 2021 Table 3, TVV2                                           |
| `lq`  (Q_REF, L/day)                 | log(0.638)      | Yang 2021 Table 3, TVQ                                            |
| `lvp` (V3_REF, L)                    | log(1.65)       | Yang 2021 Table 3, TVV3                                           |
| `emax` (EMAX_REF, log-fold)          | -0.410          | Yang 2021 Table 3, Emax                                           |
| `lt50` (T50_REF, days)               | log(28.9)       | Yang 2021 Table 3, T50                                            |
| `hill` (HILL, unitless)              | 2.79            | Yang 2021 Table 3, HILL                                           |
| `e_wt_clq` (power, WT on CL/Q)       | 0.477           | Yang 2021 Table 3, WGT_ON_CLQ                                     |
| `e_alb_clq` (power, ALB on CL/Q)     | -0.926          | Yang 2021 Table 3, ALB_ON_CLQ                                     |
| `e_igg_clq` (power, IGG on CL/Q)     | 0.184           | Yang 2021 Table 3, IGG_ON_CLQ                                     |
| `e_alt_clq` (power, ALT on CL/Q)     | -0.0795         | Yang 2021 Table 3, ALT_ON_CLQ                                     |
| `e_wt_vss` (power, WT on V2/V3)      | 0.970           | Yang 2021 Table 3, WGT_ON_VSS                                     |
| `e_bmi_vss` (power, BMI on V2/V3)    | -0.560          | Yang 2021 Table 3, BMI_ON_VSS                                     |
| IIV block `etalcl + etalvc`          | c(0.0870, 0.0422, 0.0432) | Yang 2021 Table 3, IIV_CLQ, IIV_CLQ:VSS, IIV_VSS        |
| `etaemax`                            | 0.228           | Yang 2021 Table 3, IIV_Emax                                       |
| `etalt50`                            | 0.610           | Yang 2021 Table 3, IIV_T50                                        |
| `propSd`                             | 0.188           | Yang 2021 Table 3, RUVCV                                          |
| `addSd` (mg/L)                       | 1.48            | Yang 2021 Table 3, RUVSD                                          |

Equations: structural two-compartment micro-constant form. The
final-model equations for CL_i, Q_i, V2_i, V3_i, T50_i, and Emax_i
appear in the "Final PopPK model" section of Yang 2021. Reference
covariate values (denominators in the power terms) are the population
medians from Yang 2021 Table 2: WT 76.2 kg, ALB 38 g/L, IGG 9.65 g/L,
ALT 20 IU/L, BMI 26.5 kg/m^2.

# Virtual cohort

Original observed data are not publicly available. The simulations
below use a virtual cohort whose demographics approximate the Yang
2021 population (Table 2). Continuous covariates are drawn from
log-normal distributions anchored to the reported median and a
plausible scale; the distributions are clipped to the reported ranges.

```{r cohort}
set.seed(2021)
n_subj <- 250

cohort <- tibble(
  ID  = seq_len(n_subj),
  WT  = pmin(pmax(rlnorm(n_subj, log(76.2), 0.27), 30.9), 172),
  BMI = pmin(pmax(rlnorm(n_subj, log(26.5), 0.22), 14.8), 56.3),
  ALB = pmin(pmax(rnorm(n_subj, 38.0, 5.0),         22.0), 48.0),
  IGG = pmin(pmax(rlnorm(n_subj, log(9.65), 0.30),  1.29), 27.9),
  ALT = pmin(pmax(rlnorm(n_subj, log(20.0), 0.55),  5.00), 196)
)
```

Two reference dosing regimens are compared: **3 mg/kg Q2W** (the
weight-based regimen used in 80.3% of the analysis population) and
**350 mg Q3W** (the fixed regimen ultimately recommended).

```{r events}
build_events <- function(pop, dose_mode = c("3mgkg_q2w", "350mg_q3w")) {
  dose_mode <- match.arg(dose_mode)
  if (dose_mode == "3mgkg_q2w") {
    interval_d <- 14
    n_doses    <- 12
    amt        <- pop$WT * 3
    label      <- "3 mg/kg Q2W"
  } else {
    interval_d <- 21
    n_doses    <- 8
    amt        <- rep(350, nrow(pop))
    label      <- "350 mg Q3W"
  }
  dose_times <- seq(0, by = interval_d, length.out = n_doses)
  obs_times  <- sort(unique(c(dose_times, seq(0, interval_d * n_doses, by = 1))))
  d_dose <- pop |>
    mutate(AMT = amt) |>
    tidyr::crossing(TIME = dose_times) |>
    mutate(EVID = 1, CMT = "central", DUR = 0.5 / 24, DV = NA_real_,
           treatment = label)
  d_obs <- pop |>
    tidyr::crossing(TIME = obs_times) |>
    mutate(AMT = NA_real_, EVID = 0, CMT = "central", DUR = NA_real_,
           DV = NA_real_, treatment = label)
  dplyr::bind_rows(d_dose, d_obs) |>
    dplyr::arrange(ID, TIME, dplyr::desc(EVID)) |>
    as.data.frame()
}

events_q2w <- build_events(cohort, "3mgkg_q2w")
events_q3w <- build_events(cohort, "350mg_q3w")
```

# Simulation

```{r simulate}
mod <- readModelDb("Yang_2021_cemiplimab")
sim_q2w <- rxSolve(mod, events = events_q2w, returnType = "data.frame")
sim_q3w <- rxSolve(mod, events = events_q3w, returnType = "data.frame")
sim <- dplyr::bind_rows(
  dplyr::mutate(sim_q2w, treatment = "3 mg/kg Q2W"),
  dplyr::mutate(sim_q3w, treatment = "350 mg Q3W")
)
```

# Concentration-time profiles

Yang 2021 Figure 5 shows visual predictive checks by dose group and
Figure 7 overlays observed-vs-simulated profiles at 350 mg Q3W. The
figure below reproduces the **median and 5-95% prediction interval**
from the packaged model for the two regimens that informed the
fixed-dose decision.

```{r pk-plot, fig.width = 7, fig.height = 5}
sim_summary <- sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(time, treatment) |>
  dplyr::summarise(
    median = stats::median(Cc, na.rm = TRUE),
    lo     = stats::quantile(Cc, 0.05, na.rm = TRUE),
    hi     = stats::quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(sim_summary, aes(time, median, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(
    x = "Time since first dose (days)",
    y = "Cemiplimab concentration (mg/L)",
    title = "Simulated cemiplimab PK: 3 mg/kg Q2W vs 350 mg Q3W",
    subtitle = paste0("Median and 90% prediction interval (N = ", n_subj,
                      " virtual patients per arm). Compare with Yang 2021 Figs. 5 and 7."),
    caption = "Model: Yang 2021 J Pharmacokinet Pharmacodyn 48(4):479-494"
  ) +
  theme_bw()
```

# Time-varying clearance

Yang 2021 reports that cemiplimab CL decreased on average by 35.9%
over time relative to baseline (from 0.326 to 0.208 L/day within 16
weeks of treatment). With \(E_{\max} = -0.410\), the asymptotic CL at
\(t \gg T_{50}\) is \(\mathrm{CL}_{\mathrm{base}} \cdot \exp(-0.410)
= 0.664\) of baseline (a 33.6% reduction). The typical-value CL(t)
profile below reproduces the time course at a typical reference subject
(median covariates, deterministic, etas = 0).

```{r time-dep-cl, fig.width = 6, fig.height = 4}
t_grid <- seq(0, 365, by = 5)
events_cl <- data.frame(
  ID   = 1,
  WT   = 76.2,
  BMI  = 26.5,
  ALB  = 38,
  IGG  = 9.65,
  ALT  = 20,
  TIME = c(0, t_grid),
  AMT  = c(350, rep(NA_real_, length(t_grid))),
  EVID = c(1, rep(0, length(t_grid))),
  CMT  = "central",
  DUR  = c(0.5 / 24, rep(NA_real_, length(t_grid))),
  DV   = NA_real_
)
mod_typ <- rxode2::zeroRe(mod)
sim_cl  <- rxSolve(mod_typ, events = events_cl, returnType = "data.frame")
sim_cl  <- sim_cl[sim_cl$time > 0, ]

ggplot(sim_cl, aes(time, cl / cl_base)) +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_hline(yintercept = 1,           linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = exp(-0.410), linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 16 * 7,      linetype = "dotted", colour = "darkred") +
  labs(
    x = "Time since first dose (days)",
    y = "CL(t) / CL(0)",
    title = "Time-varying clearance in Yang 2021",
    subtitle = paste0("Typical reference subject. Asymptote at exp(Emax) = ",
                      round(exp(-0.410), 3),
                      " (33.6% reduction). Red dotted line: 16-week mark.")
  ) +
  theme_bw()
```

# PKNCA validation

Yang 2021 Table 4 reports post-hoc steady-state AUC over 6 weeks and
trough concentrations by body-weight quartile at 3 mg/kg Q2W and 350
mg Q3W. The packaged model is used here to compute simulation-based
NCA over a 6-week steady-state window for the same dose regimens, and
the results are compared against the published values in the next
section.

```{r pknca}
nca_window <- function(sim, dose_interval_d, n_doses, label) {
  ss_start <- dose_interval_d * (n_doses - 3)            # last 3 doses span 6 weeks for both regimens
  ss_end   <- ss_start + 42                              # 6 weeks = 42 days
  sim_ss <- sim |>
    dplyr::filter(treatment == label,
                  !is.na(Cc),
                  time >= ss_start,
                  time <= ss_end) |>
    dplyr::mutate(time_rel = time - ss_start) |>
    dplyr::select(id, time_rel, Cc) |>
    dplyr::distinct(id, time_rel, .keep_all = TRUE)
  conc_obj <- PKNCA::PKNCAconc(sim_ss, Cc ~ time_rel | id)
  cohort_with_dose <- cohort |>
    dplyr::transmute(id = ID,
                     amt = if (label == "3 mg/kg Q2W") WT * 3 else 350)
  dose_df <- cohort_with_dose |>
    tidyr::crossing(time_rel = seq(0, by = dose_interval_d, length.out = ceiling(42 / dose_interval_d)))
  dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time_rel | id)
  intervals <- data.frame(start = 0, end = 42,
                          cmax = TRUE, cmin = TRUE,
                          tmax = TRUE, auclast = TRUE,
                          half.life = TRUE)
  nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
  PKNCA::pk.nca(nca_data)
}

nca_q2w <- nca_window(sim, 14, 12, "3 mg/kg Q2W")
nca_q3w <- nca_window(sim, 21,  8, "350 mg Q3W")

knitr::kable(summary(nca_q2w),
             caption = "Simulated NCA at steady state (6-week window), 3 mg/kg Q2W")
knitr::kable(summary(nca_q3w),
             caption = "Simulated NCA at steady state (6-week window), 350 mg Q3W")
```

# Comparison against published values

Yang 2021 Table 4 reports the following post-hoc estimates of
\(\mathrm{AUC}_{6\mathrm{wk,ss}}\) and \(C_{\mathrm{trough,ss}}\) by
body-weight quartile, pooled across all weight strata (overall N=548):

| Regimen      | Mean AUC_6wk,ss (day·mg/L) | Mean C_trough,ss (mg/L) |
|--------------|----------------------------:|-------------------------:|
| 3 mg/kg Q2W  | 3,800 (range 3,280-4,200)   | 67.5 (range 58.4-74.0)   |
| 350 mg Q3W   | 3,890 (range 3,190-4,510)   | 60.5 (range 49.0-70.3)   |

Computed roughly as the mean of the four quartile means in Yang 2021
Table 4 (Q1-Q4). The packaged model's PKNCA AUC_last and Cmin /
predose values over the 6-week window above are expected to be in the
same range (within ~20%); residual discrepancies reflect (a) the
difference between AUC_6wk,ss as the paper computes it (post-hoc
individual exposure averaged over the steady-state regimen) versus
the PKNCA AUC_last over a single 6-week interval starting at the
first dose of that interval, and (b) the virtual-cohort
covariate distribution not exactly reproducing the joint distribution
of WT, ALB, IGG, ALT, BMI in the Yang 2021 patients.

The two-regimen exposure equivalence (similar mean AUC_6wk,ss and
C_trough,ss between 3 mg/kg Q2W and 350 mg Q3W) is the central
finding informing the fixed-dose 350 mg Q3W selection. Both regimens
should produce overlapping prediction-interval bands in the
concentration-time plot above for a virtual cohort whose median
weight equals the population median (76.2 kg).

# Assumptions and deviations

- **`BLK_ON_T50` race effect on T50 is not implemented.** Yang 2021's
  final-model equation for \(T_{50,i}\) is written as
  \(T_{50,\mathrm{REF}} \cdot (\mathrm{BLK})^{\mathrm{BLK\_ON\_T50}}
  \cdot \exp(\eta)\) with `BLK_ON_T50 = 1.01` (RSE 29.2%). The
  formula is internally inconsistent: if `BLK` is a 0/1 indicator
  (Black = 1, non-Black = 0), then `0^1.01 = 0` would zero out
  T50 for the 96.4% of non-Black patients in the cohort. The paper
  does not show the NONMEM control stream and the supplementary DOCX
  was not available on disk, so the intended encoding could not be
  verified. The paper itself states all retained covariates had
  "limited impact (< 20%) on cemiplimab exposure" and were "not
  clinically meaningful" given the flat exposure-response
  relationship. The packaged model therefore uses
  \(T_{50,i} = T_{50,\mathrm{REF}} \cdot \exp(\eta_{T_{50}})\) for
  all races. Datasets with a `RACE_BLACK` column will be ignored by
  this model; this should be an immaterial omission per the paper's
  own assessment of clinical impact.
- **Time-varying covariates kept at baseline.** Yang 2021 mentions
  that body weight, albumin, and lactate dehydrogenase were assessed
  as time-varying covariates in addition to baseline values. The
  final-model equations expose only **baseline** values
  (`WGTBL`, `ALBBL`, `IGGBL`, `ALTBL`, `BMIBL`), so the packaged
  model treats these as time-fixed. Setting `WT`, `ALB`, etc. as
  time-varying in user data will treat the supplied values literally.
- **Reference covariate values.** Yang 2021's final-model equations
  are written generically as `(WGTBL_i / WGTBL_REF)^...`, etc. The
  paper does not state the reference values numerically in the
  equations, but the simulation section uses median values from
  Table 2 (WT 76.2 kg, ALB 38 g/L, IGG 9.65 g/L, ALT 20 IU/L, BMI
  26.5 kg/m^2). The packaged model uses these medians as the
  reference. (The paper's separate simulation section reports a
  slightly different mean weight of 75 kg used in fixed-dose
  simulations; the difference is immaterial because exposures are
  reported by quartile.)
- **Residual error parameterization.** Yang 2021 reports residual
  variability as `Y = F + F * ERR(1) + ERR(2)` with "log-transformation
  of the error model". The packaged model implements this as
  `Cc ~ add(addSd) + prop(propSd)`, the standard combined
  additive-plus-proportional residual error in nlmixr2's linear-space
  formulation. For VPC-style simulation the difference between the
  log-transformed and linear-space combined-error forms is
  immaterial; for likelihood-based estimation against new data, users
  who require a strict log-transform residual-error model should
  adapt the model accordingly.
- **IV infusion duration.** Yang 2021 administered cemiplimab as
  30-min IV infusions across all dose groups. The vignette
  simulations use `DUR = 0.5 / 24` day to match.
- **Virtual-cohort marginals.** Continuous covariates (WT, BMI, ALB,
  IGG, ALT) are simulated from independent log-normal / normal
  distributions anchored to the reported population medians and
  clipped to the published ranges. The joint covariate structure
  (e.g., correlation between WT and BMI) is not preserved.
