---
title: "Imatinib: fifteen published population PK models (Yang 2025 external evaluation)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imatinib: fifteen published population PK models (Yang 2025 external evaluation)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

```{r libraries, message = FALSE, warning = FALSE}
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(ggplot2)
```

## Provenance: this is a transcription from a secondary source

Read this section before using any model on this page.

Yang and colleagues (2025) did **not** develop a population PK model. They
performed a systematic review that identified fifteen previously published
population PK models of imatinib, and then evaluated all fifteen against a
real-world therapeutic-drug-monitoring (TDM) dataset of 39 adult and pediatric
patients treated at Nordic centres. Their conclusion is negative and is stated
in the title: the published models perform poorly, especially in children.

The standing policy for a secondary source is to skip it and extract the cited
primary papers instead. The carve-out that applies here is that Yang 2025 Table 1
**tabulates the parameters** of every one of the fifteen models to an unusually
complete standard: structural model, typical values, covariate coefficients with
their centring constants, between-subject variability, and residual-error
magnitudes, with table footnotes that resolve every categorical coefficient.
Fourteen of those fifteen models are therefore transcribed into this package;
the fifteenth, Jiang 2023, was already in the library, extracted directly from
its primary publication.

Consequences you must keep in mind:

* Every parameter value in these fourteen files comes from Yang 2025 Table 1,
  **not** from the paper named in each file's `reference` field. Each file says
  so in its `description` and repeats the citation chain in `reference`.
* Each file should be **re-extracted from its primary** when that paper is
  obtained. A review reprints values but routinely drops the host table's
  footnotes, its model-selection rationale, and any structure the summary row
  only gestures at.
* Yang 2025 also evaluated an **allometrically scaled variant** of most models,
  in which standard allometry replaced the published weight terms. Those variants
  are Yang 2025's own modification and are **not** encoded here. Each file is the
  model as published.
* Demographic detail beyond what Table 1 prints (weight ranges, sex splits, race)
  is unavailable from the secondary source and is recorded as such in each file's
  `population` metadata.

The one model whose transcription can be checked directly is Jiang 2023, because
the library already holds an independent extraction from its primary. Yang 2025
Table 1 gives CL/F = 9.72 x (RBC/3.7)^0.49 x theta_ABCG2 with theta = 0.879 (GT)
and 0.976 (TT), Vc/F = 229 L, ka = 1.22 /h, CV%(CL) 13.9% and a 29.6%
proportional residual. `Jiang_2023_imatinib.R`, extracted from Jiang 2023 itself,
carries exactly those ten numbers. That is a clean ten-of-ten fidelity check on
the secondary source's transcription, and it is why the carve-out was applied.

## The variability scale, and how it was settled

The single most dangerous ambiguity when transcribing a variance column out of a
review is its **scale**: a number like `0.305` may be a variance, a standard
deviation, or a coefficient of variation, and guessing wrong is a two-fold error
in omega. Yang 2025 does better than most secondary sources here, because it
labels each row with the scale it reports: `Var(eta_CL)` for Judson,
`omega_CL` for Yamakawa, `CV%(CL)` for the remaining rows, and `rho_CL-Vc` for
Gotta. Two rows nonetheless settle the question mechanically rather than by
trusting the label, because each reports a covariance alongside its variability
terms and a covariance constrains the implied correlation to lie in [-1, 1]:

* **Judson 2005** reports `Var(eta_CL) = 0.305`, `Var(eta_Vc) = 0.34`,
  `Cov = 0.237`. Read as variances, the correlation is
  0.237 / sqrt(0.305 x 0.34) = 0.736, which is admissible. Read as standard
  deviations, it is 0.237 / sqrt(0.093 x 0.116) = 2.29, which is impossible.
  The label `Var` is therefore correct.
* **Demetri 2009** reports `CV%(CL) = 34.6%`, `CV%(Vc) = 35.7%`, `Cov = 0.119`.
  Taking omega = CV gives a correlation of 0.119 / (0.346 x 0.357) = 0.963,
  admissible. Taking the exact log-normal conversion omega^2 = log(1 + CV^2)
  gives 0.119 / sqrt(0.1128 x 0.1198) = 1.024 > 1, impossible. So the CV%
  entries are converted as **omega = CV/100**, not by the exact log-normal
  formula.

The second rule matches the convention already used by the independently
extracted `Jiang_2023_imatinib.R`, whose primary prints the variance directly as
`exp(0.0192)` next to a tabulated IIV of 0.139 (0.139^2 = 0.0193). Every
`CV%` entry in these fourteen files is therefore converted as
`variance = (CV/100)^2`, and every `omega` entry (Yamakawa only) as
`variance = omega^2`.

## Source trace

Every parameter in every file traces to a single cell of Yang 2025 Table 1 (pages
876-877 of the article, spanning two printed pages), or to one of that table's
lettered footnotes. The per-file `ini()` blocks carry a trailing comment on each
line naming the cell; the table below records the structural claim per model.

```{r model-spec}
model_specs <- tibble::tribble(
  ~stem,                        ~label,                 ~primary_year, ~country,        ~n_subj, ~n_obs, ~diagnosis,
  "Judson_2005_imatinib",       "Judson 2005",          2005L,         "Europe + USA",   43L,     517L,   "GIST / STS",
  "Schmidli_2005_imatinib",     "Schmidli 2005",        2005L,         "Multi-country", 371L,    1930L,   "CML",
  "Widmer_2006_imatinib",       "Widmer 2006",          2006L,         "Switzerland",    59L,     321L,   "GIST / CML / ALL",
  "Petain_2008_imatinib",       "Petain 2008",          2008L,         "France",         67L,     505L,   "Solid tumours / GIST",
  "Demetri_2009_imatinib",      "Demetri 2009",         2009L,         "Europe + USA",   73L,      NA,    "GIST",
  "MenonAndersen_2009_imatinib","Menon-Andersen 2009",  2009L,         "USA",            41L,     842L,   "Ph+ leukemia / solid tumour",
  "Yamakawa_2011_imatinib",     "Yamakawa 2011",        2011L,         "Japan",          34L,     622L,   "CML",
  "Eechoute_2012_imatinib",     "Eechoute 2012",        2012L,         "Europe",         50L,    1743L,   "GIST",
  "Golabchifar_2014_imatinib",  "Golabchifar 2014",     2014L,         "Iran",           61L,     533L,   "CML",
  "Gotta_2014_imatinib",        "Gotta 2014",           2014L,         "Europe",       2478L,    4095L,   "CML",
  "DiPaolo_2014_imatinib",      "Di Paolo 2014",        2014L,         "Italy",          60L,     117L,   "CML",
  "Wang_2019_imatinib",         "Wang 2019",            2019L,         "China",         170L,     229L,   "CML",
  "Shriyan_2022_imatinib",      "Shriyan 2022",         2022L,         "India",          49L,     221L,   "CML",
  "Jiang_2023_imatinib",        "Jiang 2023",           2023L,         "China",         110L,      NA,    "GIST",
  "He_2023_imatinib",           "He 2023",              2023L,         "China",         230L,     424L,   "CML"
)

stopifnot(nrow(model_specs) == 15L)
# readModelDb() returns the model-defining function; rxode() evaluates it into
# the rxUi object that carries $state, $theta, $allCovs and can be solved.
models <- lapply(model_specs$stem, function(s) rxode2::rxode(nlmixr2lib::readModelDb(s)))
names(models) <- model_specs$stem
```

## Structural audit: an enumerating test against the paper's own prose

Yang 2025 Results section 3.1 makes a series of arithmetic claims about the
fifteen models it reviewed. Each is a count over the whole set, so each can be
checked by enumerating the transcribed model files rather than by inspecting a
few of them by hand. A claim that quantifies over every member of a set needs an
enumerating test; hand-picked examples cannot fail in the right way.

The structural features below are derived **from the model objects**, not
re-declared: the disposition compartment count from the presence of a
`peripheral1` state, zero-order absorption from a `d1` term, a transit chain from
`transit*` states, a lag time from a `tlag` term, and a metabolite sub-model from
`_ndmima`-suffixed states.

```{r structural-audit}
describe_structure <- function(mod) {
  states <- mod$state
  thetas <- names(mod$theta)
  parent_states <- states[!grepl("_ndmima$", states)]
  tibble::tibble(
    n_disposition_cmt = 1L + sum(grepl("^peripheral[0-9]*$", parent_states)),
    zero_order        = "ld1" %in% thetas,
    transit_chain     = any(grepl("^transit[0-9]+$", states)),
    lag_time          = "ltlag" %in% thetas,
    metabolite        = any(grepl("_ndmima$", states)),
    covariates        = paste(sort(mod$allCovs), collapse = ", ")
  )
}

structure_tbl <-
  dplyr::bind_cols(
    model_specs[, c("label", "country", "n_subj")],
    dplyr::bind_rows(lapply(models, describe_structure))
  ) |>
  dplyr::mutate(
    absorption = dplyr::case_when(
      transit_chain ~ "transit (5 compartments)",
      zero_order & lag_time ~ "zero-order + lag",
      zero_order ~ "zero-order",
      TRUE ~ "first-order"
    )
  )

structure_tbl |>
  dplyr::select(label, country, n_subj, n_disposition_cmt, absorption, metabolite, covariates) |>
  dplyr::rename(
    "Model" = label,
    "Country / region" = country,
    "N subjects" = n_subj,
    "Disposition compartments" = n_disposition_cmt,
    "Absorption" = absorption,
    "Metabolite sub-model" = metabolite,
    "Covariates" = covariates
  ) |>
  knitr::kable(caption = "Structural summary of all fifteen transcribed models, derived from the model files. Replicates Table 1 of Yang 2025.")
```

The counts Yang 2025 states in Results 3.1 are now assertions:

```{r structural-assertions}
# "A one-compartment model was used in almost all the included models (N = 14),
#  except one that used a two-compartment model."
stopifnot(sum(structure_tbl$n_disposition_cmt == 1L) == 14L)
stopifnot(sum(structure_tbl$n_disposition_cmt == 2L) == 1L)

# "The absorption phase was characterized by first-order (N = 7), zero-order
#  (N = 7), and a transit model (N = 1)."
stopifnot(sum(structure_tbl$absorption == "first-order") == 7L)
stopifnot(sum(grepl("^zero-order", structure_tbl$absorption)) == 7L)
stopifnot(sum(structure_tbl$transit_chain) == 1L)

# "One study found a lag time in imatinib absorption."
stopifnot(sum(structure_tbl$lag_time) == 1L)

# "Two studies also included the PK profile of the main metabolite N-desmethyl
#  imatinib in their model."
stopifnot(sum(structure_tbl$metabolite) == 2L)

# "Total body weight was the most frequently identified covariate (N = 5).
#  Hemoglobin (N = 2), white blood cell count (N = 2), plasma AGP (N = 2), and
#  albumin (N = 2) were other covariates found more than once."
covariate_counts <- table(unlist(lapply(models, function(m) m$allCovs)))
stopifnot(covariate_counts[["WT"]] == 5L)
stopifnot(covariate_counts[["HGB"]] == 2L)
stopifnot(covariate_counts[["WBC"]] == 2L)
stopifnot(covariate_counts[["AAG"]] == 2L)
stopifnot(covariate_counts[["ALB"]] == 2L)

cat("All structural counts from Yang 2025 Results 3.1 reproduce.\n")
```

## The reference subject and the typical-value simulation

The paper's own external dataset is not distributed, so the published
prediction-error metrics (Yang 2025 Table 3) cannot be reproduced here. What can
be reproduced, and is the paper's substantive finding, is the **dispersion
between the models themselves**: fifteen models of the same drug, given the same
patient and the same dose, disagree widely about the resulting exposure.

A single reference subject is defined and pushed through all fifteen models. It
is a 70 kg, 55-year-old man with laboratory values near the middle of the ranges
these cohorts report, at steady state on 400 mg once daily (the licensed adult
CML dose), on the day-29-or-later occasion that corresponds to TDM sampling.

```{r reference-subject}
ref_covs <- c(
  WT      = 70,     # kg
  AGE     = 55,     # years
  SEXF    = 0,      # male
  HGB     = 13,     # g/dL
  WBC     = 7,      # 10^9 cells/L
  ALB     = 38,     # g/L
  AAG     = 0.9,    # g/L (below the Di Paolo 0.94 g/L threshold)
  CRCL    = 90,     # mL/min/1.73 m^2 (above the He 2023 hinge at 85)
  RBC     = 3.7,    # 10^12 cells/L
  OCC     = 2,      # day >= 29 occasion, per Yang 2025 Table 1 footnote b
  TUM_VOL = 0,      # mm^3; no liver metastases
  # Time since the first dose, in hours. Set to 200 days so the Eechoute 2012
  # time-dependent absorption and bioavailability terms have effectively
  # decayed to their asymptotes.
  T_FIRSTDOSE = 4800,
  SNP_SLC22A1_RS683369    = 0,
  SNP_ABCG2_RS2231142_HET = 0,
  SNP_ABCG2_RS2231142_HOM = 0
)

dose_mg <- 400
tau <- 24            # h, once-daily dosing
n_days <- 30         # dosing days before the observed interval
```

Each model is dosed into whichever compartment its absorption model requires: a
`depot` for the first-order and transit models, and the `central` compartment
with `rate = -2` for the zero-order models, so that rxode2 uses the modelled
duration `dur(central) = d1` instead of collapsing the dose to a bolus.

```{r solve-typical}
build_events <- function(mod, obs_by) {
  # The zero-order-absorption models have no depot state: they are dosed
  # directly into `central` with rate = -2, which tells rxode2 to use the
  # modelled duration dur(central) = d1 rather than collapsing the dose to an
  # instantaneous bolus.
  if ("depot" %in% mod$state) {
    ev <- rxode2::et(amt = dose_mg, cmt = "depot",
                     ii = tau, until = tau * (n_days - 1))
  } else {
    ev <- rxode2::et(amt = dose_mg, cmt = "central", rate = -2,
                     ii = tau, until = tau * (n_days - 1))
  }
  # Observations across the final dosing interval, including its time-zero
  # record (PKNCA needs one; without it every subject warns that the AUC range
  # starts before the first measurement).
  obs_times <- seq(tau * (n_days - 1), tau * n_days, by = obs_by)

  # The two parent-metabolite models declare TWO endpoints (`Cc` and
  # `Cc_ndmima`), and rxode2 rejects an observation record that does not name
  # one of them. Single-endpoint models must NOT be given `cmt =` on their
  # observation rows: naming an algebraic observable there auto-injects a
  # compartment slot after the ODE states and renumbers everything. So the
  # endpoint is named only where the model declares more than one. Both
  # observables are returned as columns at every observation row either way.
  if (nrow(mod$predDf) > 1L) {
    rxode2::et(ev, obs_times, cmt = as.character(mod$predDf$var)[1L])
  } else {
    rxode2::et(ev, obs_times)
  }
}

solve_reference <- function(stem, mod) {
  ev <- build_events(mod, obs_by = 0.25)
  params <- ref_covs[names(ref_covs) %in% mod$allCovs]

  # omega = NA is the only reliable way to force a typical-value solve: rxode2
  # otherwise reuses the previous solve's omega for the same compiled model and
  # silently re-samples etas.
  out <- rxode2::rxSolve(mod, ev, params = params, omega = NA,
                         returnType = "data.frame")

  # rxSolve drops the id column entirely for a single-subject solve.
  out$id <- 1L
  out$stem <- stem
  out$time_rel <- out$time - tau * (n_days - 1)
  out[out$time_rel >= 0, , drop = FALSE]
}

sim_list <- Map(solve_reference, model_specs$stem, models)
sim_typ <- dplyr::bind_rows(sim_list) |>
  dplyr::left_join(model_specs[, c("stem", "label")], by = "stem")

# Guard against a silently stochastic "typical value" run: with omega = NA the
# individual clearance must be a single constant within each model.
stopifnot(
  sim_typ |>
    dplyr::group_by(stem) |>
    dplyr::summarise(n_cl = dplyr::n_distinct(round(cl, 8)), .groups = "drop") |>
    dplyr::pull(n_cl) |>
    max() == 1L
)
stopifnot(nrow(sim_typ) == 15L * length(seq(0, tau, by = 0.25)))
```

## Between-model dispersion at the licensed 400 mg dose

```{r figure-dispersion, fig.cap = "Steady-state imatinib concentration-time profiles over one 24 h dosing interval for the same reference subject on 400 mg once daily, predicted by each of the fifteen published models transcribed from Yang 2025 Table 1. The shaded band is the 1000-3000 ng/mL trough range recommended in the therapeutic-drug-monitoring literature that Yang 2025 cites."}
ggplot(sim_typ, aes(x = time_rel, y = Cc, group = label, colour = label)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1000, ymax = 3000,
           alpha = 0.12, fill = "grey30") +
  geom_line(linewidth = 0.6) +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(
    x = "Time after dose at steady state (h)",
    y = "Imatinib plasma concentration (ng/mL)",
    colour = "Model"
  ) +
  theme_bw() +
  theme(legend.text = element_text(size = 7))
```

```{r trough-spread}
trough_tbl <-
  sim_typ |>
  dplyr::filter(abs(time_rel - tau) < 1e-8) |>
  dplyr::transmute(label, cl = round(cl, 3), ctrough = round(Cc, 1)) |>
  dplyr::arrange(ctrough)

trough_tbl |>
  dplyr::rename(
    "Model" = label,
    "Clearance in the reference subject (L/h)" = cl,
    "Predicted steady-state trough (ng/mL)" = ctrough
  ) |>
  knitr::kable(caption = "Predicted 24 h steady-state trough concentration for one identical reference subject on imatinib 400 mg once daily, by model.")

cat(sprintf(
  "Predicted trough spans %.0f to %.0f ng/mL, a %.1f-fold range across models.\n",
  min(trough_tbl$ctrough), max(trough_tbl$ctrough),
  max(trough_tbl$ctrough) / min(trough_tbl$ctrough)
))
```

That fold-range is the quantitative form of Yang 2025's conclusion. The fifteen
models were fitted to different populations, different sampling designs and
different assays, and they do not agree on the exposure that a single
well-specified patient will achieve on the licensed dose. Yang 2025's external
evaluation found that none of them reached the pre-specified precision criterion
(median absolute prediction error at or below 30%) on real TDM data, even after
allometric scaling, and that the observations falling outside the 90% prediction
interval were disproportionately from children.

## Replicating the published typical-clearance range

Yang 2025 Results 3.1 states: "The typical apparent clearance was estimated to be
7.29-17.3 L/h." That range is over the fifteen models' typical values at their
own reference covariate values, so it is an exact check on the transcribed
`lcl` parameters, independent of everything else in the files.

```{r clearance-range}
typical_cl <- vapply(
  models,
  function(m) unname(exp(m$theta[["lcl"]])),
  numeric(1)
)

clearance_tbl <- tibble::tibble(
  label = model_specs$label,
  typical_cl = round(unname(typical_cl), 3)
) |>
  dplyr::arrange(typical_cl)

clearance_tbl |>
  dplyr::rename(
    "Model" = label,
    "Typical CL or CL/F at the model's own reference (L/h)" = typical_cl
  ) |>
  knitr::kable(caption = "Typical clearance parameter of each transcribed model. The minimum and maximum reproduce the 7.29-17.3 L/h range stated in Yang 2025 Results 3.1.")

stopifnot(isTRUE(all.equal(min(typical_cl), 7.29, tolerance = 1e-6)))
stopifnot(isTRUE(all.equal(max(typical_cl), 17.3, tolerance = 1e-6)))
cat("Typical clearance range reproduces Yang 2025 Results 3.1 exactly: 7.29 to 17.3 L/h.\n")
```

The minimum is Petain 2008 (the children-and-adults model) and the maximum is
Gotta 2014 (the 2478-patient routine-care cohort). Note that Schmidli 2005's
`lcl` of 13.8 L/h is its day-1 value; on the day-29-or-later occasion used for
TDM it falls to 10.62 L/h, which is why the trough table above and this table
order the models differently.

## PKNCA validation

Non-compartmental analysis is run on the simulated steady-state interval for
every model. The structural identity available at steady state is exact and
per-subject rather than a comparison of medians:

    AUC(0-tau) x CL = F x Dose

Because the typical-value solve fixes every random effect, any departure from
that identity is a coding error in the model file, in the dosing setup, or in
the NCA window, not sampling noise. This is a much stronger test than comparing
a simulated Cmax against a published one, and it is the only NCA check the
source paper's contents can support: Yang 2025 reports prediction-error metrics
against its own undistributed dataset, not NCA parameters.

```{r pknca}
conc_data <-
  sim_typ |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::transmute(id, label, time = time_rel, Cc)

dose_data <-
  model_specs |>
  dplyr::transmute(id = 1L, label, time = 0, dose_amount = dose_mg)

# The model label is the treatment-like grouping variable and is listed before
# `id`; PKNCAdose() rejects a nested (slash) grouping formula, so both objects
# use the additive form.
o_conc <- PKNCA::PKNCAconc(conc_data, Cc ~ time | label + id,
                           concu = "ng/mL", timeu = "h")
o_dose <- PKNCA::PKNCAdose(dose_data, dose_amount ~ time | label + id,
                           doseu = "mg")

nca_intervals <- data.frame(
  start = 0, end = tau,
  auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE
)

o_data <- PKNCA::PKNCAdata(o_conc, o_dose, intervals = nca_intervals)
nca_res <- PKNCA::pk.nca(o_data)

nca_wide <-
  as.data.frame(nca_res) |>
  dplyr::select(label, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

```{r pknca-identity}
# Model-specific bioavailability. Only Eechoute 2012 carries an explicit F term
# (its `fdepot` variable); for every other model F is folded into the apparent
# CL/F and Vc/F, so F = 1 and bind_rows has left the column NA.
if (!("fdepot" %in% names(sim_typ))) sim_typ$fdepot <- NA_real_

f_by_model <-
  sim_typ |>
  dplyr::group_by(label) |>
  dplyr::summarise(
    cl = unique(round(cl, 8)),
    fdepot = if (all(is.na(fdepot))) 1 else unique(round(fdepot, 8)),
    .groups = "drop"
  )

identity_tbl <-
  nca_wide |>
  dplyr::left_join(f_by_model, by = "label") |>
  dplyr::mutate(
    # auclast is in ng/mL * h and cl in L/h, so their product is in ug;
    # divide by 1000 to reach the mg in which the dose is expressed.
    amount_recovered_mg = auclast * cl / 1000,
    expected_mg = fdepot * dose_mg,
    pct_error = 100 * (amount_recovered_mg - expected_mg) / expected_mg
  )

identity_tbl |>
  dplyr::transmute(
    label,
    auclast = round(auclast, 0),
    cmax = round(cmax, 0),
    tmax = round(tmax, 2),
    cmin = round(cmin, 0),
    pct_error = round(pct_error, 3)
  ) |>
  dplyr::arrange(label) |>
  dplyr::rename(
    "Model" = label,
    "AUC0-tau (ng*h/mL)" = auclast,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "Cmin (ng/mL)" = cmin,
    "AUC x CL vs F x Dose, % error" = pct_error
  ) |>
  knitr::kable(caption = "Steady-state non-compartmental analysis of the reference subject on 400 mg once daily, with the mass-balance identity AUC0-tau x CL = F x Dose checked per model.")

# The identity is exact up to ODE solver tolerance and the trapezoidal
# approximation of AUC on a 0.25 h grid.
stopifnot(max(abs(identity_tbl$pct_error)) < 1)
cat(sprintf(
  "AUC(0-tau) x CL reproduces F x Dose for all %d models; worst absolute error %.3f%%.\n",
  nrow(identity_tbl), max(abs(identity_tbl$pct_error))
))
```

## The metabolite sub-models

Two of the fifteen models carry N-desmethyl imatinib (CGP74588) alongside the
parent. Both parameterise the metabolite as apparent with respect to the fraction
of imatinib metabolized to it (`CLm/fm`, `V1m/fm`, and for Menon-Andersen also
`Qm/fm` and `V2m/fm`), because `fm` is not identifiable from plasma data alone.
The `central_ndmima` state therefore holds an `fm`-scaled amount; the predicted
metabolite concentration is nonetheless the true one, because the same `fm`
divides both the clearance and the volume and cancels in their ratio.

```{r metabolite-figure, fig.cap = "Parent imatinib and N-desmethyl imatinib steady-state profiles for the reference subject on 400 mg once daily, from the two models that carry the metabolite. Both were developed on cohorts that included children."}
met_stems <- structure_tbl$label[structure_tbl$metabolite]

sim_typ |>
  dplyr::filter(label %in% met_stems) |>
  dplyr::select(label, time_rel, Imatinib = Cc, `N-desmethyl imatinib` = Cc_ndmima) |>
  tidyr::pivot_longer(c(Imatinib, `N-desmethyl imatinib`),
                      names_to = "Analyte", values_to = "conc") |>
  ggplot(aes(x = time_rel, y = conc, colour = Analyte)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~label) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time after dose at steady state (h)",
       y = "Plasma concentration (ng/mL)") +
  theme_bw()
```

```{r metabolite-ratio}
sim_typ |>
  dplyr::filter(label %in% met_stems, abs(time_rel - tau) < 1e-8) |>
  dplyr::transmute(
    label,
    parent = round(Cc, 1),
    metabolite = round(Cc_ndmima, 1),
    ratio_pct = round(100 * Cc_ndmima / Cc, 1)
  ) |>
  dplyr::rename(
    "Model" = label,
    "Imatinib trough (ng/mL)" = parent,
    "N-desmethyl imatinib trough (ng/mL)" = metabolite,
    "Metabolite-to-parent ratio (%)" = ratio_pct
  ) |>
  knitr::kable(caption = "Steady-state trough metabolite-to-parent ratio predicted by the two parent-metabolite models.")
```

## Between-subject variability: a stochastic cohort

The typical-value work above deliberately holds every random effect at zero. The
between-subject variability the models carry is large and differs markedly
between them, which is the other half of why they disagree. A 200-subject cohort
is simulated for three contrasting models: the smallest reported IIV on
clearance (Jiang 2023, 13.9%), the largest single-parameter IIV among the
one-compartment models (Golabchifar 2014, whose lag time carries 68.3%), and the
2478-patient routine-care model (Gotta 2014).

```{r stochastic-cohort}
n_sub <- 200
vpc_stems <- c("Jiang_2023_imatinib", "Gotta_2014_imatinib", "Golabchifar_2014_imatinib")

solve_cohort <- function(stem) {
  mod <- models[[stem]]
  ev <- build_events(mod, obs_by = 0.5)
  params <- ref_covs[names(ref_covs) %in% mod$allCovs]

  set.seed(20250908)
  out <- rxode2::rxSolve(mod, ev, params = params, nSub = n_sub,
                         returnType = "data.frame")
  # With covariates supplied as a plain named vector and subjects generated by
  # nSub, rxSolve names the subject column `sim.id` rather than `id`. Normalise
  # it so the assertion below actually measures the simulated cohort size
  # instead of silently reading a missing column.
  out$id <- out$sim.id
  out$stem <- stem
  out$time_rel <- out$time - tau * (n_days - 1)
  out[out$time_rel >= 0, , drop = FALSE]
}

sim_vpc <-
  dplyr::bind_rows(lapply(vpc_stems, solve_cohort)) |>
  dplyr::left_join(model_specs[, c("stem", "label")], by = "stem")

stopifnot(dplyr::n_distinct(sim_vpc$id) == n_sub)
```

```{r vpc-figure, fig.cap = "Simulated between-subject spread in steady-state imatinib concentrations for three contrasting models, 200 virtual subjects each sharing the reference subject's covariates. Solid line is the median, ribbon spans the 5th to 95th percentiles. Only the random effects differ between subjects."}
sim_vpc |>
  dplyr::group_by(label, time_rel) |>
  dplyr::summarise(
    p05 = quantile(Cc, 0.05),
    p50 = median(Cc),
    p95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(x = time_rel)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 1000, ymax = 3000,
           alpha = 0.12, fill = "grey30") +
  geom_ribbon(aes(ymin = p05, ymax = p95, fill = label), alpha = 0.25) +
  geom_line(aes(y = p50, colour = label), linewidth = 0.7) +
  facet_wrap(~label) +
  scale_x_continuous(breaks = seq(0, 24, by = 6)) +
  labs(x = "Time after dose at steady state (h)",
       y = "Imatinib plasma concentration (ng/mL)") +
  theme_bw() +
  theme(legend.position = "none")
```

```{r vpc-trough-summary}
sim_vpc |>
  dplyr::filter(abs(time_rel - tau) < 1e-8) |>
  dplyr::group_by(label) |>
  dplyr::summarise(
    p05 = round(quantile(Cc, 0.05), 0),
    p50 = round(median(Cc), 0),
    p95 = round(quantile(Cc, 0.95), 0),
    in_window_pct = round(100 * mean(Cc >= 1000 & Cc <= 3000), 1),
    .groups = "drop"
  ) |>
  dplyr::rename(
    "Model" = label,
    "5th percentile trough (ng/mL)" = p05,
    "Median trough (ng/mL)" = p50,
    "95th percentile trough (ng/mL)" = p95,
    "Within the 1000-3000 ng/mL window (%)" = in_window_pct
  ) |>
  knitr::kable(caption = "Simulated steady-state trough distribution on 400 mg once daily for three models, 200 virtual subjects each. The final column is the fraction of subjects whose predicted trough falls inside the therapeutic-drug-monitoring window Yang 2025 cites.")
```

The three models disagree not only on the typical trough but on how many patients
a fixed 400 mg dose would place inside the target window. That is the practical
case for TDM which Yang 2025 makes, and simultaneously the reason they conclude
that model-based dose individualisation for imatinib is not yet reliable,
particularly in children, for whom none of these cohorts provides real support.

## Assumptions and deviations

Recorded here because the source is a secondary one, so the gaps are systematic
rather than incidental.

1. **Every parameter comes from Yang 2025 Table 1, not the primary.** See the
   provenance section above. All fourteen new files should be re-extracted when
   their primaries are obtained.

2. **CV% is converted as `omega = CV/100`.** Settled mechanically by the Demetri
   2009 covariance, which is inadmissible under the exact log-normal conversion.
   Judson 2005's `Var(...)` entries are used as variances directly, settled by
   the same argument on its own covariance; Yamakawa 2011's `omega` entries are
   squared.

3. **Widmer 2006 carries no covariate.** Yang 2025 Table 1 footnote c records
   that the evaluation used "the demographic base model", so the alpha-1 acid
   glycoprotein relationship after which the primary is titled is not available
   from the secondary source. This is the largest single known gap between the
   transcription and its primary; it is documented in that file's
   `covariatesDataExcluded`.

4. **Demetri 2009 repeats its CL/F covariate exponents on Vc/F.** Yang 2025
   Table 1 prints `(ALB/38.3)^1.66 x (WBC/7)^-0.418` on both parameters. This
   was checked against the original PDF with `pdftotext -layout` rather than the
   trimmed markdown, so it is genuinely what the secondary source prints and not
   a table-flattening artifact. It remains unusual and is the highest-priority
   item to confirm against the primary.

5. **Eechoute 2012's transit-chain topology is an inference.** Table 1 gives the
   absorption model only as "T5" with rate constants Ktr and Ka and does not
   print the chain layout. A dosing depot followed by five transit compartments,
   with the final step into central at `ka`, is used. The alternative reading
   (dosing directly into `transit1`, with no separate depot) shortens the mean
   absorption time by one `1/ktr` step, about 4% of the total, so the choice is
   numerically near-immaterial.

6. **Eechoute 2012's shared `Vc-Ka` random effect.** Table 1 prints a single
   omega under the hyphenated label `CV%(Vc-Ka)`, in contrast to the separate
   per-parameter entries on the same row. It is implemented as one eta
   multiplying both `vc` and `ka`. If the primary instead reports two separate
   omegas of equal magnitude, the correction is confined to that one term.

7. **Gotta 2014's variability on the residual-error magnitude is omitted.**
   Table 1 reports `CV%(sigma_Prop) = 36.2%`, an inter-individual random effect
   on the size of the proportional residual error itself. nlmixr2's error-model
   syntax provides no supported way to attach an eta to a residual SD, so this
   component is not encoded. Typical-value predictions, the covariate model and
   the CL/V random effects are unaffected; only the simulated between-subject
   spread in residual scatter is narrower than the published model implies.

8. **He 2023's renal covariate has an inferred upper branch.** Table 1 footnote h
   is truncated: it gives `theta_eGFR = (eGFR/85)^0.25` when eGFR is below 85 but
   prints no "else" clause. The value above the hinge is taken to be 1, the only
   reading that keeps the function continuous at the hinge and the standard form
   for a renal-impairment-only covariate.

9. **Unit conversions applied at the call site.** Schmidli 2005 and He 2023 use
   hemoglobin in g/dL (the canonical `HGB` column permits either unit and each
   file declares which it uses). Di Paolo 2014's alpha-1 acid glycoprotein
   threshold is printed as 94 mg/dL and is tested as 0.94 g/L against the
   canonical `AAG` column. Eechoute 2012's liver-metastasis coefficient is per
   cm^3 and is divided by 1000 against the canonical `TUM_VOL` column, which is
   carried in mm^3. Additive residual errors printed in mg/L (Schmidli 249 ng/mL,
   Demetri 4 ng/mL, Yamakawa 400 ng/mL) are converted to the ng/mL in which `Cc`
   is reported.

10. **The occasion column is re-based.** Yang 2025 Table 1 footnote b codes the
    source occasion as 0 (day 1) and 1 (day 29 or later). The canonical `OCC`
    column is 1-based, so the three affected models (Schmidli 2005, Petain 2008,
    Demetri 2009) map those to `OCC = 1` and `OCC = 2` respectively and
    decompose to a binary indicator inside `model()`. The occasion drives a
    typical-value shift, not inter-occasion random variability; no eta is
    attached to it.

11. **The allometrically scaled variants are not encoded.** Yang 2025 evaluated
    each adult-only model twice, once as published and once with standard
    allometric scaling substituted for the published weight terms. Only the
    as-published form is in this package, because the scaled variant is the
    reviewers' modification rather than a published model.

12. **The reference subject is a construction of this vignette.** No single
    covariate vector is common to all fifteen cohorts, so the values in
    `ref_covs` were chosen to sit near the middle of the reported ranges and,
    where a model uses a threshold, on the reference side of it (AAG below Di
    Paolo's 0.94 g/L cut point, CRCL above He's 85 hinge, no liver metastases).
    A different reference subject would change the absolute troughs but not the
    finding of wide between-model dispersion.

13. **Yang 2025 Table 3 is not reproduced.** The prediction-error metrics
    (MDPE, MDAPE, F20, F30, RMSE) are computed against the authors' own
    39-patient TDM dataset, which is available only "from the corresponding
    author upon reasonable request" and is not distributed with the article.

## Registry additions made alongside this extraction

* `inst/references/compartment-names.md`: the metabolite suffix `ndmima`
  (N-desmethyl imatinib, CGP74588), following the `ndm<drug>` contraction
  established by `ndmsel`.
* `inst/references/covariate-columns.md`: `T_FIRSTDOSE` (time elapsed since the
  first dose of a treatment course, hours) and `SNP_SLC22A1_RS683369` (SLC22A1 /
  OCT1 c.480C>G L160F variant carrier indicator), both well-formed members of
  existing auto-approved canonical families.

## Session information

```{r session-info}
sessionInfo()
```
