---
title: "Polymyxin B (Yang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polymyxin B (Yang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Yang_2025_polymyxinB"))$reference`
* Description: `r rxode2::rxode(readModelDb("Yang_2025_polymyxinB"))$description`
* Article: [Front Pharmacol. 2025;16:1511088](https://doi.org/10.3389/fphar.2025.1511088)

# Population

Yang 2025 is a single-centre prospective study run in the intensive care unit of
the First Affiliated Hospital of Army Medical University (Chong Qing, China)
between August 2021 and July 2024. Ninety-five adults receiving intravenous
polymyxin B for a microbiologically confirmed carbapenem-resistant organism
(CRO) infection were split chronologically into a modelling set (80 patients,
184 concentrations) and an external validation set (15 patients, 30
concentrations). Patients receiving any form of renal replacement therapy during
polymyxin B treatment were excluded, so the cohort spans a wide but
non-dialysed range of renal function.

The modelling set (Table 1) had a median age of 60 years (IQR 47-74), a median
total body weight of 63 kg (IQR 55-74), 25% women, and an APACHE II score of
28 +/- 9.25. Carbapenem-resistant *Acinetobacter baumannii* was the commonest
pathogen (70 cases), followed by carbapenem-resistant Enterobacterales (56) and
carbapenem-resistant *Pseudomonas aeruginosa* (15); the lung was the infection
site in 81.25%. Cockcroft-Gault creatinine clearance had a median of
75.99 mL/min (IQR 38.46-130.58) and platelet count a median of
163.50 x 10^9/L (IQR 84.5-266.25) -- the two values that normalise the
covariate model. A loading dose was given to 97.5% of the modelling set and the
maintenance dose was 1.31 +/- 0.25 mg/kg per administration, which at the median
body weight is about 82.5 mg every 12 h.

Sampling was deliberately sparse and paired: after at least 48 h of therapy, one
sample immediately before an infusion (steady-state trough) and one immediately
after (steady-state peak). Concentrations were measured by validated
UPLC-MS/MS with polymyxin E2 as internal standard, with total polymyxin B
computed by summing the molar contributions of the PMB1 and PMB2 components.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Yang_2025_polymyxinB")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Yang_2025_polymyxinB.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL) | 2.03 L/h | Table 2, "CL (L h-1)"; RSE 5.30%, bootstrap 95% CI 1.81-2.24 |
| `lvc` (V) | 18 L | Table 2, "V(L)"; RSE 5.30%, bootstrap 95% CI 16.30-19.80 |
| `e_crcl_cl` | 0.26 | Table 2, "dCLdCrCL"; RSE 20.40%, bootstrap 95% CI 0.15-0.36 |
| `e_plt_cl` | -0.14 | Table 2, "dCLdPLT"; RSE 24.20%, bootstrap 95% CI -0.22 to -0.066 |
| CRCL normalising constant | 75.99 mL/min | Equation 6 prints the symbol `CrCLmedian`; the number is the modelling-set median of Table 1 |
| PLT normalising constant | 163.50 x 10^9/L | Equation 6 prints the symbol `PLTmedian`; the number is the modelling-set median of Table 1 |
| `etalcl` | variance 0.148225 (omega 0.385) | Table 2, "etaCL (%) 38.50"; RSE 10.70%, bootstrap 95% CI 29.70-46.60 |
| `etalvc` | `fixed(0)` | Equation 7 contains `exp(eta_V)` but Table 2 reports no variance for it |
| `propSd` | 0.30 | Table 2, "Proportional error"; RSE 8.20%, bootstrap 95% CI 0.25-0.35 |
| `addSd` | 0.21 mg/L | Table 2, "Additive error"; RSE 28.40%, bootstrap 95% CI 0.070-0.35 |
| `cl <- exp(lcl + etalcl) * (CRCL/75.99)^e_crcl_cl * (PLT/163.5)^e_plt_cl` | n/a | Equation 6 |
| `vc <- exp(lvc + etalvc)` | n/a | Equation 7 |
| `d/dt(central) <- -kel * central` | n/a | Results, "PopPK model analysis and validation": "A one-compartment model with first-order elimination best fit the population data" |
| `Cc ~ add(addSd) + prop(propSd)` | n/a | Results: "residual variability was described using both proportional and additive error models" |
| 1-hour infusion | n/a | Methods, "Monte Carlo simulation": "50 mg, 75 mg, 100 mg, and 125 mg every 12 h (q12h) with a 1-hour infusion time" |

## Two Table 2 rows that had to be resolved on scale, not read off

Table 2 reports its variability rows as bare numbers, and its abbreviation
footnote glosses `eta` as "variance of inter-individual variability". Read
literally that would make `etaCL = 0.385` a variance (omega = 0.62) and the two
residual rows NONMEM `$SIGMA` variances (proportional SD 0.548, additive SD
0.458). Both readings are refuted by numbers printed elsewhere in the same
paper, so this model encodes 38.50%, 0.30 and 0.21 on the **SD scale**.

**`etaCL` is an SD, not a variance.** Three independent checks agree.

1. The row is labelled `etaCL (%)`. A percentage is meaningless for a variance
   of a log-scale random effect.
2. Table 1 reports the model-based AUCss,24h of the modelling set as a median of
   74.07 with IQR 55.81-94.07. Because AUCss,24h was computed as the 24-hour
   dose divided by the individual clearance (Methods, "Data analysis"), that
   IQR is the IQR of 1/CL. Its 1.686-fold spread implies a total SD of log(CL)
   of about `log(94.07 / 55.81) / (2 * qnorm(0.75))` = 0.387. That total
   *already includes* the CrCL and PLT contributions, so it is an upper bound on
   omega -- and it excludes 0.62 outright.
3. The Table 3 band percentages can be reconstructed from Equation 6 only with
   omega near 0.385. The reconstruction is run as a gate below.

**The residual rows are SDs, not variances.** At the variance reading the
residual CV would be about 55%, and the external-validation metrics reported in
the Results (MAPE% 28.45, F30 73.33%, i.e. 73% of predictions within 30% of
observed) would be unreachable: for a 55% error SD only about
`r round(100 * (2 * stats::pnorm(0.30 / 0.61) - 1), 0)`% of predictions fall
within 30%. At the SD reading the combined residual CV over the observed
concentration range (Ctrough,ss 1.69 mg/L, Cpeak,ss 5.73 mg/L) is about 30-33%,
which puts F30 near 70% and MAPE near 25% -- consistent with what the paper
reports.

# Virtual cohort

Original observed data are not publicly available. The cohorts below are virtual
populations whose covariate distributions approximate the published Table 1
demographics of the modelling set.

`CRCL` and `PLT` are drawn as log-normals matched to the Table 1 median and
interquartile range, then truncated to a physiologically plausible window. The
dose is 82.53 mg every 12 h as a 1-hour infusion -- the median maintenance dose
of 1.31 mg/kg (Table 1) at the median body weight of 63 kg.

```{r cohort}
# set.seed() seeds R's RNG (used for the covariate draws below). It does NOT
# seed rxode2's eta sampler, whose streams are partitioned per solver thread,
# so the eta draw differs between this machine and a CI runner with a different
# thread count. Every assertion below is therefore written either on a
# deterministic (zeroRe) quantity or on a robust centre/quantile statistic.
set.seed(20250312)

n_sub <- 200L
tau <- 12      # dosing interval (h)
t_inf <- 1     # infusion duration (h)
dose_mg <- 1.31 * 63  # Table 1 median maintenance dose x median body weight

# Log-normal parameters matched to the Table 1 modelling-set median and IQR.
lnorm_from_median_iqr <- function(median_value, q25, q75) {
  list(
    meanlog = log(median_value),
    sdlog = log(q75 / q25) / (2 * stats::qnorm(0.75))
  )
}
crcl_par <- lnorm_from_median_iqr(75.99, 38.46, 130.58)
plt_par <- lnorm_from_median_iqr(163.50, 84.5, 266.25)

cohort <- tibble::tibble(
  id = seq_len(n_sub),
  CRCL = pmin(pmax(stats::rlnorm(n_sub, crcl_par$meanlog, crcl_par$sdlog), 5), 300),
  PLT = pmin(pmax(stats::rlnorm(n_sub, plt_par$meanlog, plt_par$sdlog), 10), 600)
)

# Event table built as a plain data frame so the dosing columns stay explicit.
# ss = 1 on the t = 0 dose puts the system directly at the steady state of the
# q12h regimen, which removes any dependence on the length of a burn-in; a
# second explicit dose at t = 12 completes a 24-hour steady-state window.
# PKNCA does not expand `addl`, so the second dose is written out as its own row.
obs_times <- seq(0, 24, by = 0.1)

dose_rows <- tidyr::expand_grid(
  cohort,
  tibble::tibble(time = c(0, tau), ss = c(1, 0), ii = c(tau, 0))
) |>
  dplyr::mutate(amt = dose_mg, evid = 1, cmt = "central", dur = t_inf)

obs_rows <- tidyr::expand_grid(cohort, tibble::tibble(time = obs_times)) |>
  dplyr::mutate(amt = NA_real_, evid = 0, cmt = "central", dur = NA_real_,
                ss = 0, ii = 0)

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, amt, evid, cmt, dur, ss, ii, CRCL, PLT) |>
  as.data.frame()

stopifnot(
  nrow(dose_rows) == 2L * n_sub,
  !anyNA(events$CRCL), !anyNA(events$PLT)
)
```

# Simulation

```{r simulate}
mod <- readModelDb("Yang_2025_polymyxinB")

# The solved object is named `sim_cohort`, not `sim`: rxSolve emits a column
# literally called `sim` (the residual-error-carrying simulated observation, as
# opposed to `Cc`, which is the individual prediction without residual error),
# and a data frame of the same name shadows it in confusing ways downstream.
# rxSolve returns observation records only, so there is no `evid` column to
# filter on.
sim_cohort <- rxode2::rxSolve(mod, events = events,
                              keep = c("CRCL", "PLT")) |>
  as.data.frame()

stopifnot(nrow(sim_cohort) == n_sub * length(obs_times))
```

A matching deterministic (typical-value) solve is used for every replication of
a published *median*. That substitution is exact here rather than approximate:
the model carries a single non-zero random effect, log-normal on `CL`, and both
the steady-state concentration profile and AUCss,24h are monotone in `CL`, so
the population median equals the typical-value profile.

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(mod)

events_typical <- events |>
  dplyr::filter(id == 1L) |>
  dplyr::mutate(CRCL = 75.99, PLT = 163.50)

sim_typical <- rxode2::rxSolve(mod_typical, events = events_typical) |>
  as.data.frame()
```

# Replicate published results

## Table 1 -- steady-state trough and peak concentrations

Table 1 reports the observed steady-state concentrations of the modelling set as
medians with IQR: Ctrough,ss 1.69 mg/L (0.78-3.29) and Cpeak,ss 5.73 mg/L
(3.96-7.64). Peak samples were drawn "immediately after" the infusion, i.e. at
the end of the 1-hour infusion, and trough samples "immediately before" the next
one.

```{r table1-conc}
peak_typ <- sim_typical$Cc[which.min(abs(sim_typical$time - t_inf))]
trough_typ <- sim_typical$Cc[which.min(abs(sim_typical$time - tau))]

table1_cmp <- tibble::tibble(
  Quantity = c("Cpeak,ss (end of 1-h infusion)", "Ctrough,ss (end of interval)"),
  Simulated = c(peak_typ, trough_typ),
  Published = c(5.73, 1.69),
  `Published IQR` = c("3.96-7.64", "0.78-3.29")
) |>
  dplyr::mutate(`Difference (%)` = 100 * (Simulated - Published) / Published)

knitr::kable(table1_cmp, digits = 2,
             caption = "Typical-value steady-state concentrations at the median dose (82.53 mg q12h) and median covariates, against the observed medians of Yang 2025 Table 1.")

# The typical-value profile IS the population median here (single log-normal
# eta, monotone response), so these gates are deterministic -- no cohort draw
# enters them.
stopifnot(
  peak_typ > 3.96, peak_typ < 7.64,
  trough_typ > 0.78, trough_typ < 3.29,
  abs(peak_typ - 5.73) / 5.73 < 0.05,
  abs(trough_typ - 1.69) / 1.69 < 0.02
)
```

## Table 3 -- AUCss,24h band probabilities across 40 simulated scenarios

Table 3 tabulates, for 40 combinations of maintenance dose (50, 75, 100,
125 mg q12h), creatinine clearance (30, 60, 90, 120, 150 mL/min) and platelet
count (85 and 266 x 10^9/L), the percentage of the simulated population whose
AUCss,24h falls below 50, inside 50-100, and above 100 mg h/L.

For a linear one-compartment model at steady state, AUCss,24h is exactly the
24-hour dose divided by the individual clearance -- which is also how Yang 2025
computed it (Methods, "Data analysis"). Individual clearance is log-normal about
the typical value of Equation 6, so the three band probabilities follow in
closed form from the typical clearance and omega. The typical clearance for each
scenario is read off a `zeroRe()` solve, so the check exercises the packaged
encoding of Equation 6 rather than a re-typed formula.

```{r table3-scenarios}
scenarios <- tidyr::expand_grid(
  dose = c(50, 75, 100, 125),
  CRCL = c(30, 60, 90, 120, 150),
  PLT = c(85, 266)
) |>
  dplyr::mutate(id = dplyr::row_number())

scen_events <- dplyr::bind_rows(
  scenarios |>
    dplyr::transmute(id, time = 0, amt = dose, evid = 1, cmt = "central",
                     dur = t_inf, ss = 1, ii = tau, CRCL, PLT),
  scenarios |>
    dplyr::transmute(id, time = tau, amt = NA_real_, evid = 0, cmt = "central",
                     dur = NA_real_, ss = 0, ii = 0, CRCL, PLT)
) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

scen_cl <- rxode2::rxSolve(mod_typical, events = scen_events) |>
  as.data.frame() |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl_typ = dplyr::first(cl), .groups = "drop")

omega_cl <- sqrt(0.148225)  # Table 2 etaCL 38.50% -> omega 0.385

band_prob <- function(auc_typ, lower, upper, omega) {
  stats::pnorm(log(upper / auc_typ) / omega) -
    stats::pnorm(log(lower / auc_typ) / omega)
}

table3_sim <- scenarios |>
  dplyr::inner_join(scen_cl, by = "id") |>
  dplyr::mutate(
    auc_typ = 2 * dose / cl_typ,
    below50 = 100 * band_prob(auc_typ, 0, 50, omega_cl),
    within = 100 * band_prob(auc_typ, 50, 100, omega_cl),
    above100 = 100 * band_prob(auc_typ, 100, Inf, omega_cl)
  )
```

```{r table3-published}
# Yang 2025 Table 3, transcribed. Columns: percentage of the simulated
# population with AUCss,24h < 50, within 50-100, and > 100 mg h/L.
table3_pub <- tibble::tribble(
  ~dose, ~CRCL, ~PLT, ~below50, ~within, ~above100,
  50,  30,  85, 35.2, 58.0,  6.8,   50,  30, 266, 24.3, 59.4, 16.3,
  50,  60,  85, 51.8, 45.0,  3.2,   50,  60, 266, 37.2, 55.8,  7.0,
  50,  90,  85, 64.4, 33.8,  1.8,   50,  90, 266, 50.2, 45.2,  4.6,
  50, 120,  85, 69.9, 29.0,  1.1,   50, 120, 266, 55.5, 41.2,  3.3,
  50, 150,  85, 76.4, 22.7,  0.9,   50, 150, 266, 63.8, 34.0,  2.2,
  75,  30,  85,  8.3, 58.9, 32.8,   75,  30, 266,  3.7, 46.2, 50.1,
  75,  60,  85, 17.4, 61.4, 21.2,   75,  60, 266, 10.5, 57.0, 32.5,
  75,  90,  85, 25.8, 60.0, 14.2,   75,  90, 266, 16.8, 60.7, 22.5,
  75, 120,  85, 30.6, 59.9,  9.5,   75, 120, 266, 20.5, 61.1, 18.4,
  75, 150,  85, 37.3, 55.3,  7.4,   75, 150, 266, 25.6, 60.8, 13.6,
  100,  30,  85,  2.2, 33.8, 64.0,  100,  30, 266,  0.6, 24.2, 75.2,
  100,  60,  85,  4.4, 48.0, 47.6,  100,  60, 266,  2.3, 35.5, 62.2,
  100,  90,  85,  8.4, 56.7, 34.9,  100,  90, 266,  4.7, 46.0, 49.3,
  100, 120,  85, 13.4, 56.7, 29.9,  100, 120, 266,  6.2, 50.0, 43.8,
  100, 150,  85, 15.6, 61.0, 23.4,  100, 150, 266,  8.4, 55.9, 35.7,
  125,  30,  85,  0.5, 17.0, 82.5,  125,  30, 266,  0.1, 10.4, 89.5,
  125,  60,  85,  1.4, 29.2, 69.4,  125,  60, 266,  0.4, 20.6, 79.0,
  125,  90,  85,  3.2, 38.2, 58.6,  125,  90, 266,  1.4, 28.8, 69.8,
  125, 120,  85,  4.8, 42.0, 53.2,  125, 120, 266,  2.0, 32.8, 65.2,
  125, 150,  85,  4.9, 50.6, 44.5,  125, 150, 266,  2.6, 39.2, 58.2
)
stopifnot(nrow(table3_pub) == 40L)

table3_cmp <- table3_pub |>
  tidyr::pivot_longer(c(below50, within, above100),
                      names_to = "band", values_to = "published") |>
  dplyr::inner_join(
    table3_sim |>
      tidyr::pivot_longer(c(below50, within, above100),
                          names_to = "band", values_to = "simulated"),
    by = c("dose", "CRCL", "PLT", "band")
  ) |>
  dplyr::mutate(diff_pp = simulated - published)

stopifnot(nrow(table3_cmp) == 120L)
```

```{r table3-gate}
table3_summary <- table3_cmp |>
  dplyr::group_by(band) |>
  dplyr::summarise(
    `Median |difference| (pp)` = stats::median(abs(diff_pp)),
    `90th pctile |difference| (pp)` = stats::quantile(abs(diff_pp), 0.9),
    `Max |difference| (pp)` = max(abs(diff_pp)),
    .groups = "drop"
  ) |>
  dplyr::rename("AUC band" = band)

knitr::kable(table3_summary, digits = 2,
             caption = "Reconstruction of Yang 2025 Table 3 from Equation 6 and omega = 0.385, across all 120 published cells. Differences are in percentage points.")

# Deterministic gate -- no random draw enters the reconstruction, so the
# residual difference is Yang 2025's own 1,000-draw Monte Carlo noise plus
# rounding of the two normalising medians.
stopifnot(
  stats::median(abs(table3_cmp$diff_pp)) < 2,
  stats::quantile(abs(table3_cmp$diff_pp), 0.9) < 3.5,
  max(abs(table3_cmp$diff_pp)) < 6,
  stats::cor(table3_cmp$simulated, table3_cmp$published) > 0.997
)
```

```{r figure-5}
# Replicates the substance of Figure 5 of Yang 2025: probability of target
# attainment against creatinine clearance, by maintenance dose and platelet
# count, at MIC 0.5 mg/L with the AUCss,24h/MIC >= 66.9 target.
pta_target <- 66.9 * 0.5

pta <- table3_sim |>
  dplyr::mutate(
    pta = 100 * (1 - stats::pnorm(log(pta_target / auc_typ) / omega_cl)),
    dose_label = factor(paste0(dose, " mg q12h"),
                        levels = paste0(c(50, 75, 100, 125), " mg q12h")),
    plt_label = paste0("PLT = ", PLT, " x 10^9/L")
  )

ggplot(pta, aes(CRCL, pta, colour = dose_label)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 90, linetype = "dashed") +
  facet_wrap(~plt_label) +
  coord_cartesian(ylim = c(0, 100)) +
  labs(x = "Creatinine clearance (mL/min)",
       y = "PTA (%) for AUCss,24h/MIC >= 66.9 at MIC 0.5 mg/L",
       colour = "Maintenance dose",
       caption = "Replicates Figure 5 / Results of Yang 2025.")

# Results: "At MIC <= 0.5 mg/L, PTA >= 90% was achieved in all groups except
# the 50 mg q12h maintenance dose group"; the Discussion narrows the exception
# to "the 50 mg maintenance dose group with high CrCL".
stopifnot(
  all(pta$pta[pta$dose > 50] >= 90),
  any(pta$pta[pta$dose == 50] < 90),
  all(pta$pta[pta$dose == 50 & pta$CRCL >= 90] < 90)
)
```

## Figure 2 -- steady-state concentration-time profile

```{r figure-2}
# Replicates the shape of Figure 2 (VPC) of Yang 2025 over one 24-hour
# steady-state window at the median maintenance dose.
sim_cohort |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05),
    Q50 = stats::quantile(Cc, 0.50),
    Q95 = stats::quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 24, by = 4)) +
  labs(x = "Time (h)", y = "Polymyxin B concentration (mg/L)",
       caption = paste("Steady-state profile, 82.53 mg q12h as a 1-h infusion;",
                       "median and 5th-95th percentile of 200 virtual subjects.",
                       "Compare Figure 2 of Yang 2025."))
```

# PKNCA validation

NCA is run over the full 24-hour steady-state window (two q12h doses), which is
the interval Yang 2025 reports AUCss,24h over.

```{r pknca}
sim_nca <- sim_cohort |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc)

# A record sits exactly on the interval end (t = 24) by construction of
# obs_times, which is what ctrough needs; and a t = 0 record anchors the interval
# start. Guard both defensively.
stopifnot(
  all(tapply(sim_nca$time, sim_nca$id, function(x) any(x == 0))),
  all(tapply(sim_nca$time, sim_nca$id, function(x) any(x == 24)))
)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | id,
                             concu = "mg/L", timeu = "h")

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | id, doseu = "mg")

intervals <- data.frame(
  start = 0,
  end = 24,
  cmax = TRUE,
  tmax = TRUE,
  cmin = TRUE,
  ctrough = TRUE,
  auclast = TRUE,
  cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

## Solve against its own closed form

For a linear one-compartment model at steady state, AUC over a full dosing
window equals the dose administered in that window divided by the individual
clearance. Both sides of this comparison use the *same* drawn `CL`, so the only
difference is trapezoidal error on the simulated grid and a tight bound is the
correct assertion.

```{r auc-identity}
cl_i <- sim_cohort |>
  dplyr::group_by(id) |>
  dplyr::summarise(cl = dplyr::first(cl), .groups = "drop")

auc_chk <- nca_wide |>
  dplyr::inner_join(cl_i, by = "id") |>
  dplyr::mutate(
    auc_closed_form = 2 * dose_mg / cl,
    pct_diff = 100 * (auclast - auc_closed_form) / auc_closed_form
  )

knitr::kable(
  tibble::tibble(
    Statistic = c("Median % difference", "90th percentile |% difference|",
                  "Max |% difference|"),
    Value = c(stats::median(auc_chk$pct_diff),
              stats::quantile(abs(auc_chk$pct_diff), 0.9),
              max(abs(auc_chk$pct_diff)))
  ),
  digits = 3,
  caption = "PKNCA AUC(0-24) at steady state against dose/CL, over 200 virtual subjects."
)

stopifnot(max(abs(auc_chk$pct_diff)) < 0.05)
```

## Comparison against published values

```{r nca_compare}
simulated_nca <- nca_wide |>
  dplyr::transmute(
    group = "82.53 mg q12h",
    cmax = cmax,
    ctrough = ctrough,
    auclast = auclast
  )

published <- tibble::tribble(
  ~group,          ~cmax, ~ctrough, ~auclast,
  "82.53 mg q12h",  5.73,  1.69,    74.07
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_nca,
  reference = published,
  by = "group",
  units = c(cmax = "mg/L", ctrough = "mg/L", auclast = "mg*h/L"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste("Simulated steady-state NCA against Yang 2025 Table 1",
                  "(observed Cpeak,ss and Ctrough,ss medians; model-based",
                  "AUCss,24h median). * differs from reference by >20%."),
  align = c("l", "l", "r", "r", "r")
)

# Centre-of-distribution gates against the published medians. The AUC target is
# the modelling set's model-based median, which was computed over the cohort's
# ACTUAL per-patient doses; the simulation fixes every subject at the median
# maintenance dose, so a moderate offset is expected and the bound is set on
# the centre rather than on any cohort extreme.
stopifnot(
  abs(stats::median(nca_wide$cmax) - 5.73) / 5.73 < 0.15,
  abs(stats::median(nca_wide$ctrough) - 1.69) / 1.69 < 0.15,
  abs(stats::median(nca_wide$auclast) - 74.07) / 74.07 < 0.25
)
```

# Assumptions and deviations

- **Normalising constants for the two covariates.** Equation 6 prints its two
  divisors only as the symbols `CrCLmedian` and `PLTmedian`. They are taken as
  the modelling-set medians of Table 1, 75.99 mL/min and 163.50 x 10^9/L. This
  is corroborated rather than assumed: the Table 3 reconstruction above
  reproduces all 120 published cells to a median of under 3 percentage points,
  which would not happen if either divisor were materially different, and the
  paper's own simulation grid uses the Table 1 modelling-set quartiles of PLT
  (85 and 266) as its two levels.
- **`etaCL` and the two residual rows read on the SD scale.** Table 2's
  abbreviation footnote calls `eta` a variance. That gloss is not followed, for
  the reasons set out in the source-trace section. Under the variance reading
  neither the Table 1 AUCss,24h IQR, nor the Table 3 band percentages, nor the
  external-validation MAPE/F30 metrics could be reproduced.
- **No IIV on volume of distribution.** Equation 7 is written
  `V = 18 L * exp(eta_V)`, but Table 2's inter-individual-variability block
  contains only the `etaCL` row, and no supplementary table on disk supplies a
  value (the article's deposited supplementary material is the base-model
  goodness-of-fit figure only). `etalvc` is therefore encoded as
  `fixed(0)` -- the random effect is declared, as the paper writes it, with a
  variance of zero rather than an invented magnitude. A user who wants to
  reinstate it can set a non-zero variance on `etalvc`.
- **Covariate distributions are virtual.** `CRCL` and `PLT` are drawn as
  log-normals matched to the Table 1 median and IQR of the modelling set, then
  truncated to 5-300 mL/min and 10-600 x 10^9/L. The paper does not publish the
  per-patient covariate values, nor the correlation between creatinine clearance
  and platelet count, which is treated here as zero.
- **Both covariates are treated as time-fixed.** The paper reports a single
  baseline value per patient in Table 1 and does not describe a time-varying
  covariate model, although platelet count and renal function do move over a
  13-day ICU course.
- **Dose.** Table 1 reports the maintenance dose as 1.31 +/- 0.25 mg/kg per
  administration and the median body weight as 63 kg; the simulated regimen uses
  the product, 82.53 mg q12h, for every subject. The paper's own Monte Carlo
  scenarios (50/75/100/125 mg q12h) are used for the Table 3 and Figure 5
  replications.
- **AUCss,24h definition.** Following the paper (Methods, "Data analysis"),
  model-based AUCss,24h is the 24-hour dose divided by the individual clearance.
  The PKNCA section computes it independently by trapezoidal integration of the
  simulated profile and confirms the two agree.
- **The PLT covariate is extrapolation-unsafe outside roughly
  85-266 x 10^9/L.** Yang 2025 excluded patients with elevated platelet counts
  from modelling (Limitations) and describes the platelet effect as a first
  report requiring confirmation.
- All parameter values come from the paper's own text and tables. No value was
  taken from a figure, from correspondence, or from an upstream model.
