---
title: "Osimertinib and AZ5104 (Yang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Osimertinib and AZ5104 (Yang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE,
                      fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(PKNCA)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
```

# Model

`Yang_2025_osimertinib` is the FLAURA2 update of the osimertinib
parent-metabolite population PK model. Yang 2025 pooled 52,338 plasma samples
from 2,196 patients with EGFR-mutated advanced NSCLC across six studies (AURA
and its extension, AURA2, AURA3, FLAURA, ADAURA and FLAURA2) and re-estimated
the structure first published by Brown 2017: first-order oral absorption into a
one-compartment parent disposition, feeding a one-compartment metabolite
(AZ5104) in series, with first-order elimination from both.

The analysis exists to answer one question -- does adding pemetrexed plus
platinum chemotherapy change osimertinib PK? -- and its answer is no. "Addition
of chemotherapy" was tested by forward inclusion on both parent and metabolite
clearance and was not significant on either (Table S2), so it does not appear in
the model at all. The covariates that were retained (weight, albumin, race) were
all judged to have no clinically meaningful effect on exposure.

```{r model}
mod <- nlmixr2lib::readModelDb("Yang_2025_osimertinib")
mod
```

# Population

Yang 2025 Table 1, overall pooled column (N = 2,196):

| Characteristic | Value |
|----|----|
| Age (years) | median 62.0 (range 25.0-91.0); mean 61.5, SD 10.7 |
| Weight (kg) | median 62.0 (range 29.0-128); mean 63.5, SD 14.1 |
| Body mass index (kg/m2) | median 23.4 (range 12.9-42.6) |
| Creatinine clearance (mL/min) | median 82.1 (range 20.5-196) |
| Serum albumin (g/L) | median 40.0 (range 17.0-53.3); mean 39.7, SD 4.92 |
| Sex | 1,416 female (64.5%), 780 male (35.5%) |
| Race | White 584 (26.6%), Asian excluding Chinese and Japanese 502 (22.9%), Chinese 501 (22.8%), Japanese 384 (17.5%), Other 225 (10.2%) |
| WHO performance status | 0: 896 (40.8%); 1: 1,300 (59.2%) |
| Line of therapy | first 38.5%, second 21.3%, third or later 25.4%, adjuvant 14.8% |

Dosing is osimertinib 80 mg once daily, with a permitted reduction to 40 mg once
daily for toxicity. In the FLAURA2 combination arm this was given with
pemetrexed 500 mg/m2 plus either cisplatin 75 mg/m2 or carboplatin AUC 5
mg/mL/min on Day 1 of 21-day cycles for four cycles, then pemetrexed
maintenance every three weeks.

The reference ("typical") patient used throughout Yang 2025, and used for every
typical-value check below, is a **White patient of 62 kg with serum albumin
40 g/L** -- the Table 1 medians and the denominators written into Eqs 2-5.

# Source trace

Every value in `ini()` and every equation in `model()`, with its location in the
source. The supplementary NONMEM control stream (`$PROBLEM FINAL MODEL`) carries
the same values at full precision as the rounded Table 2 entries, so it is the
quoted source where the extra digits matter.

| Model element | Value | Source |
|----|----|----|
| `lka` | 0.260946 1/h | Table 2 (0.26); Eq 1; `$THETA(3)` |
| `lcl` (CLptot/F) | 14.3452 L/h | Table 2 (14.35); Eq 2; `$THETA(1)` |
| `lvc` (Vp/F) | 1151.11 L | Table 2 (1,151); Eq 3; `$THETA(4)` |
| `lcl_az5104` (CLm/F) | 32.0484 L/h | Table 2 (32.05); Eq 4; `$THETA(2)` |
| `lvc_az5104` (Vm/F) | 151.014 L | Table 2 (151); Eq 5; `$THETA(6)` |
| `fm` | 0.25, fixed | Table 2 "0.25 (Fixed)"; Methods; `$THETA(5) 0.25 FIX` |
| `e_wt_cl` | 0.359774 | Table 2 (0.36); Eq 2; `$THETA(7)` |
| `e_alb_cl` | 0.674479 | Table 2 (0.67); Eq 2; `$THETA(8)` |
| `e_wt_vc` | 0.640227 | Table 2 (0.64); Eq 3; `$THETA(9)` |
| `e_alb_vc` | 1.56611 | Table 2 (1.57); Eq 3; `$THETA(10)` |
| `e_wt_cl_az5104` | 0.740376 | Table 2 (0.74); Eq 4; `$THETA(11)` |
| `e_alb_cl_az5104` | 0.722460 | Table 2 (0.72); Eq 4; `$THETA(12)` |
| `e_alb_vc_az5104` | **-0.652465** | Eq 5 exponent `-0.65`; `$THETA(13)`. Table 2 prints the magnitude only |
| `e_race_asian_oth_cl_az5104` | 0.161274 | Table 2 (0.16); Eq 4 factor 1.16; `$THETA(14)`, `ETHL = 1` |
| `e_race_japanese_cl_az5104` | 0.181966 | Table 2 (0.18); Eq 4 factor 1.18; `$THETA(16)`, `ETHL = 3` |
| `e_race_chinese_cl_az5104` | 0, fixed | `$THETA(15) 0 FIX`, `ETHL = 2`; Table S2 "not significant" |
| `e_race_other_cl_az5104` | 0, fixed | `$THETA(17) 0 FIX`, `ETHL = 99`; Table S2 "not significant" |
| IIV block CL/CLm | var 0.191276, cov 0.186815, var 0.237930 | `$OMEGA BLOCK(2)`; Table 2 (0.19 / 0.19 / 0.24) |
| `etalka`, `etalvc`, `etalvc_az5104` | 0.699828, 0.542208, 0.512184 | `$OMEGA`; Table 2 (0.70, 0.54, 0.51) |
| `propSd`, `addSd` | 0.219108, 29.5202 nM | Table 2 (0.22, 29.52); `$THETA(19)`, `$THETA(18)` |
| `propSd_az5104`, `addSd_az5104` | 0.231740, 0.358422 nM | Table 2 (0.23, 0.36); `$THETA(21)`, `$THETA(20)` |
| ODE rate constants | `K12 = KA`, `K20 = CL(1-FM)/V1`, `K23 = CL*FM/V1`, `K30 = CLM1/VM1` | `$PK` block of the control stream |
| Concentration scaling | `S2 = V1*499.61/1e6`, `S3 = VM1*485.59/1e6` | `$PK` block; gives nM |
| Reference WT / ALB | 62 kg, 40 g/L | Eqs 2-5 denominators; Table 1 medians; Figure 3 caption |

## Two things the parameter table alone would have got wrong

**The albumin exponent on AZ5104 volume is negative.** Table 2 lists "ALB on
Vm/F: 0.65" with no sign. Equation 5 writes `(ALB/40)^-0.65` and `$THETA(13)` is
`-0.652465`. Reading the table alone would invert this covariate. (Table 2's
bootstrap CI for this row is also printed reversed, "0.68 (1.28-0.42)".)

**Table 2's variability column is on the variance scale, not SD or CV.** The
paper separately reports %CV of 46, 52, 101, 85 and 82% for CLptot/F, CLm/F, Ka,
Vp/F and Vm/F. Back-transforming the tabulated values as variances via
`sqrt(exp(w) - 1)` reproduces all five; reading them as SDs does not:

```{r omega_scale}
w   <- c(CLptot = 0.191276, CLm = 0.237930, Ka = 0.699828, Vp = 0.542208, Vm = 0.512184)
pub <- c(46, 52, 101, 85, 82)
tibble::tibble(
  Parameter          = names(w),
  `Published %CV`    = pub,
  `If variance (%)`  = round(100 * sqrt(exp(w) - 1), 1),
  `If SD (%)`        = round(100 * sqrt(exp(w^2) - 1), 1)
)
```

```{r omega_scale_gate}
# The variance reading must reproduce every published %CV to within 1 point;
# the SD reading must not. Deterministic arithmetic, so this bound is exact.
stopifnot(
  all(abs(100 * sqrt(exp(w) - 1) - pub) < 1),
  any(abs(100 * sqrt(exp(w^2) - 1) - pub) > 10)
)
```

The `$OMEGA BLOCK(2)` off-diagonal is likewise a covariance, not the correlation
coefficient that Table 2's "Correlation" row label suggests; it implies a
CLptot/F--CLm/F correlation of `r round(0.186815 / sqrt(0.191276 * 0.237930), 2)`.
That is mechanistically sensible, because both parameters are apparent (`/F`)
and so share the same unknown bioavailability.

# Virtual cohort

200 patients, with weight and albumin drawn to match the Table 1 overall
marginal distributions (truncated to the reported ranges) and race sampled at
the Table 1 proportions. Covariates are drawn with R's own RNG, so the cohort is
identical across rxode2 versions.

```{r cohort}
set.seed(20250501)
n_sub <- 200

rtrunc_norm <- function(n, mean, sd, lo, hi) {
  x <- rnorm(n, mean, sd)
  while (any(bad <- x < lo | x > hi)) x[bad] <- rnorm(sum(bad), mean, sd)
  x
}

race_levels <- c("White", "Asian other", "Chinese", "Japanese", "Other")
race_probs  <- c(26.6, 22.9, 22.8, 17.5, 10.2) / 100

cohort <- tibble::tibble(
  id  = seq_len(n_sub),
  WT  = rtrunc_norm(n_sub, 63.5, 14.1, 29, 128),
  ALB = rtrunc_norm(n_sub, 39.7, 4.92, 17, 53.3),
  race_label = sample(race_levels, n_sub, replace = TRUE, prob = race_probs)
) |>
  dplyr::mutate(
    RACE_ASIAN_OTH = as.integer(race_label == "Asian other"),
    RACE_JAPANESE  = as.integer(race_label == "Japanese"),
    RACE_CHINESE   = as.integer(race_label == "Chinese"),
    RACE_OTHER     = as.integer(race_label == "Other")
  )

summary(cohort[, c("WT", "ALB")])
table(cohort$race_label)
```

```{r events}
dose_amt <- 80    # mg once daily
dose_int <- 24    # h
n_doses  <- 30    # 720 h; about 13 typical parent half-lives
ss_start <- (n_doses - 1) * dose_int

dose_rows <- tidyr::expand_grid(id = cohort$id, dose_idx = seq_len(n_doses)) |>
  dplyr::mutate(time = (dose_idx - 1) * dose_int, evid = 1L, amt = dose_amt,
                cmt = "depot", dvid = NA_integer_) |>
  dplyr::select(id, time, evid, amt, cmt, dvid)

# Dense over the first and the final (steady-state) interval, one trough per
# day in between. Both interval edges are present so PKNCA can anchor its
# window and resolve Ctrough without extrapolation.
obs_times <- sort(unique(c(
  c(0, 0.5, 1, 2, 3, 4, 6, 8, 10, 12, 16, 20, 24),
  seq(48, ss_start - dose_int, by = 24),
  ss_start + seq(0, 24, by = 0.5)
)))

# This model has exactly two endpoints (Cc, Cc_az5104), so observation records
# nominate their endpoint with dvid rather than cmt. rxode2 returns BOTH
# analyte columns at every observation row, so a single dvid = 1 grid is enough
# to read both. Naming an algebraic observable as though it were a compartment
# would instead inject a slot for it and renumber the ODE states.
obs_rows <- tidyr::expand_grid(id = cohort$id, time = obs_times) |>
  dplyr::mutate(evid = 0L, amt = 0, cmt = NA_character_, dvid = 1L)

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(cohort, by = "id") |>
  dplyr::arrange(id, time, dplyr::desc(evid))
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20250501)
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("WT", "ALB", "race_label"))

# Typical-value replication (no IIV) at the paper's reference patient.
mod_typical <- rxode2::zeroRe(mod)
ref_events <- events |>
  dplyr::filter(id == 1) |>
  dplyr::mutate(WT = 62, ALB = 40, RACE_ASIAN_OTH = 0, RACE_JAPANESE = 0,
                RACE_CHINESE = 0, RACE_OTHER = 0)
sim_ref <- rxode2::rxSolve(mod_typical, events = ref_events)
```

```{r profile_plot, fig.cap = "Simulated steady-state osimertinib and AZ5104 concentration-time profiles over the final 24 h dosing interval (200 virtual patients; median and 5th-95th percentile band). Compare with Yang 2025 Figure 2, which plots the observed concentrations across the six pooled studies."}
band <- sim |>
  dplyr::filter(time >= ss_start) |>
  dplyr::mutate(tad = time - ss_start) |>
  dplyr::select(tad, Osimertinib = Cc, AZ5104 = Cc_az5104) |>
  tidyr::pivot_longer(-tad, names_to = "Analyte", values_to = "conc") |>
  dplyr::summarise(
    med = median(conc), lo = quantile(conc, 0.05), hi = quantile(conc, 0.95),
    .by = c(tad, Analyte)
  )

ggplot2::ggplot(band, ggplot2::aes(tad, med)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = lo, ymax = hi), alpha = 0.25) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::facet_wrap(~Analyte, scales = "free_y") +
  ggplot2::labs(x = "Time after dose (h)", y = "Concentration (nM)") +
  ggplot2::theme_bw()
```

# Structural verification

## Closed-form mass balance

At steady state the AUC over one dosing interval must equal dose divided by
apparent clearance, with the dose converted to nmol using each analyte's own
molecular weight -- exactly the convention the control stream's `S2` and `S3`
encode. Both sides use the same parameters here, so the only difference is
numerical integration error and the bound is tight.

```{r mass_balance}
MW_parent <- 499.61
MW_az5104 <- 485.59

trapz <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

ref_ss <- sim_ref |> dplyr::filter(time >= ss_start) |> dplyr::arrange(time)
auc_parent_solved <- trapz(ref_ss$time, ref_ss$Cc)
auc_metab_solved  <- trapz(ref_ss$time, ref_ss$Cc_az5104)

auc_parent_closed <- (dose_amt * 1e-3 / MW_parent * 1e9) / 14.3452
auc_metab_closed  <- (0.25 * dose_amt * 1e-3 / MW_az5104 * 1e9) / 32.0484

tibble::tibble(
  Analyte       = c("Osimertinib", "AZ5104"),
  `Solved AUCtau (nM*h)` = round(c(auc_parent_solved, auc_metab_solved), 1),
  `Closed form (nM*h)`   = round(c(auc_parent_closed, auc_metab_closed), 1),
  `Difference (%)`       = round(100 * c(auc_parent_solved / auc_parent_closed,
                                         auc_metab_solved  / auc_metab_closed) - 100, 3)
)
```

```{r mass_balance_gate}
stopifnot(
  abs(auc_parent_solved / auc_parent_closed - 1) < 0.005,
  abs(auc_metab_solved  / auc_metab_closed  - 1) < 0.005
)
```

This gate is what proves the metabolite transfer convention. The control stream
moves amount from the central to the metabolite compartment 1:1, with **no molar
stoichiometric correction**, and only then forms the metabolite concentration
using the AZ5104 molecular weight. Inserting a molar correction would scale
AZ5104 exposure by `r round(485.59/499.61, 4)` and break this identity.

## Time to peak

Yang 2025 Results reports model-derived steady-state tmax medians of 6.7 h
(IQR 5.9-7.4) for osimertinib and 11.3 h (IQR 9.9-12.5) for AZ5104. These are
model-derived, so the typical-value profile should land on them. Note this is
*not* the same quantity as the Table S5 observed NCA tmax (4.50 h and 4.00 h),
which is bounded by a sparse sampling grid that stopped at 6 h post-dose.

```{r tmax}
tmax_parent <- ref_ss$time[which.max(ref_ss$Cc)] - ss_start
tmax_metab  <- ref_ss$time[which.max(ref_ss$Cc_az5104)] - ss_start

tibble::tibble(
  Analyte             = c("Osimertinib", "AZ5104"),
  `Typical tmax (h)`  = c(tmax_parent, tmax_metab),
  `Yang 2025 median`  = c(6.7, 11.3),
  `Yang 2025 IQR`     = c("5.9-7.4", "9.9-12.5")
)
```

```{r tmax_gate}
# Deterministic typical-value profile; must fall inside the published IQR.
stopifnot(
  tmax_parent >= 5.9, tmax_parent <= 7.4,
  tmax_metab  >= 9.9, tmax_metab  <= 12.5
)
```

## Half-lives

```{r halflife}
thalf_parent <- log(2) * 1151.11 / 14.3452
thalf_metab  <- log(2) * 151.014 / 32.0484
tibble::tibble(
  Analyte              = c("Osimertinib", "AZ5104"),
  `Typical t1/2 (h)`   = round(c(thalf_parent, thalf_metab), 1),
  `Yang 2025 median`   = c(47.7, 3.1),
  `Yang 2025 IQR`      = c("31.8-69.0", "2.2-4.5")
)
stopifnot(
  thalf_parent > 31.8, thalf_parent < 69.0,
  thalf_metab  >  2.2, thalf_metab  <  4.5
)
```

AZ5104's own elimination half-life is far shorter than the parent's, so the
metabolite is formation-rate limited and its observed decline tracks
osimertinib.

# Replicating Figure 3 (covariate forest plot)

Yang 2025 Results quotes two exact numbers off its own forest plot: "The ratio
of AZ5104 exposures for high body weight (89 kg) and for low body weight
(43 kg) vs. the reference population was 0.76 and 1.31, respectively" (the
published sentence has a typographical "AZ5102"). Because AZ5104 steady-state
AUC is inversely proportional to CLm/F, and weight enters CLm/F only through the
0.740376 power term, these are a closed-form two-point test of that exponent.

```{r forest}
wt_ratio <- function(wt) (wt / 62)^-0.740376
fp <- tibble::tibble(
  `Body weight (kg)`        = c(89, 43),
  Percentile                = c("95th", "5th"),
  `Model AZ5104 AUCss ratio` = round(wt_ratio(c(89, 43)), 3),
  `Yang 2025 Results`        = c(0.76, 1.31)
)
fp
```

```{r forest_gate}
# Closed-form against a printed number: exact to the paper's 2 decimal places.
stopifnot(all(abs(wt_ratio(c(89, 43)) - c(0.76, 1.31)) < 0.01))
```

The full simulated forest plot, over each retained covariate at the 5th and 95th
percentile (or against the White reference for race), reproduces the shape of
Figure 3 -- every effect sits inside or close to the 0.8-1.25 bioequivalence
band, which is the paper's basis for calling none of them clinically meaningful:

```{r forest_plot, fig.cap = "Covariate effects on steady-state exposure relative to the reference patient (White, 62 kg, albumin 40 g/L). Replicates Yang 2025 Figure 3. The shaded band is the 0.80-1.25 bioequivalence reference range."}
q05_wt <- 43; q95_wt <- 89; q05_alb <- 31; q95_alb <- 47

# AUCss ratios are inverse-clearance ratios; Cmaxss additionally moves with volume.
eff <- tibble::tribble(
  ~Covariate,                  ~Analyte,      ~ratio,
  "Weight 43 kg (5th)",        "Osimertinib", (q05_wt/62)^-0.359774,
  "Weight 89 kg (95th)",       "Osimertinib", (q95_wt/62)^-0.359774,
  "Albumin 31 g/L (5th)",      "Osimertinib", (q05_alb/40)^-0.674479,
  "Albumin 47 g/L (95th)",     "Osimertinib", (q95_alb/40)^-0.674479,
  "Weight 43 kg (5th)",        "AZ5104",      (q05_wt/62)^-0.740376,
  "Weight 89 kg (95th)",       "AZ5104",      (q95_wt/62)^-0.740376,
  "Albumin 31 g/L (5th)",      "AZ5104",      (q05_alb/40)^-0.722460,
  "Albumin 47 g/L (95th)",     "AZ5104",      (q95_alb/40)^-0.722460,
  "Asian other vs White",      "AZ5104",      1 / (1 + 0.161274),
  "Japanese vs White",         "AZ5104",      1 / (1 + 0.181966)
)

ggplot2::ggplot(eff, ggplot2::aes(ratio, Covariate)) +
  ggplot2::annotate("rect", xmin = 0.8, xmax = 1.25, ymin = -Inf, ymax = Inf,
                    alpha = 0.15, fill = "darkred") +
  ggplot2::geom_vline(xintercept = 1, linetype = "dashed") +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::facet_wrap(~Analyte, scales = "free_y") +
  ggplot2::labs(x = "AUCss ratio vs reference patient", y = NULL) +
  ggplot2::theme_bw()
```

```{r forest_direction_gate}
# Directional content of Figure 3: heavier patients and higher albumin both
# clear faster, so both reduce exposure; the two retained race effects raise
# metabolite clearance and so reduce AZ5104 exposure.
stopifnot(
  (q95_wt/62)^-0.359774  < 1, (q05_wt/62)^-0.359774  > 1,
  (q95_alb/40)^-0.674479 < 1, (q05_alb/40)^-0.674479 > 1,
  1 / (1 + 0.161274) < 1, 1 / (1 + 0.181966) < 1
)
```

# PKNCA validation

Steady-state NCA over the final 24 h dosing interval, for both analytes, in nM.

```{r pknca}
nca_frame <- function(conc_col) {
  sim |>
    dplyr::filter(time >= ss_start, !is.na(.data[[conc_col]])) |>
    dplyr::transmute(id, time = time - ss_start, conc = .data[[conc_col]],
                     treatment = "Osimertinib 80 mg QD")
}

conc_parent <- nca_frame("Cc")
conc_metab  <- nca_frame("Cc_az5104")

dose_df <- tibble::tibble(id = unique(conc_parent$id), time = 0,
                          amt = dose_amt, treatment = "Osimertinib 80 mg QD")

intervals <- data.frame(start = 0, end = 24,
                        cmax = TRUE, tmax = TRUE, cmin = TRUE,
                        auclast = TRUE, cav = TRUE)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

nca_parent <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_parent, conc ~ time | treatment + id,
                   concu = "nmol/L", timeu = "h"),
  dose_obj, intervals = intervals))

nca_metab <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_metab, conc ~ time | treatment + id,
                   concu = "nmol/L", timeu = "h"),
  dose_obj, intervals = intervals))
```

## Comparison against published NCA

Yang 2025 Table S5 reports observed steady-state NCA for the FLAURA2
osimertinib-plus-chemotherapy arm (N = 244), as geometric means.

```{r nca_compare}
simulated <- dplyr::bind_rows(
  as.data.frame(nca_parent$result) |> dplyr::mutate(analyte = "Osimertinib"),
  as.data.frame(nca_metab$result)  |> dplyr::mutate(analyte = "AZ5104")
)

reference <- tibble::tribble(
  ~analyte,      ~auclast, ~cmax,  ~cmin,
  "Osimertinib",    11840, 587.5,  392.5,
  "AZ5104",          1277,  59.07,  45.51
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = simulated,
  reference     = reference,
  by            = "analyte",
  params        = c("auclast", "cmax", "cmin"),
  units         = c(auclast = "nmol*h/L", cmax = "nmol/L", cmin = "nmol/L"),
  tolerance_pct = 20
)

knitr::kable(cmp, caption = paste(
  "Simulated steady-state NCA vs Yang 2025 Table S5 observed geometric means",
  "(FLAURA2 osimertinib plus chemotherapy arm, N = 244).",
  "* marks a difference greater than 20%."
))
```

```{r nca_gate}
# Structural gate on the centre of the distribution. A mis-transcribed
# clearance, dose or molecular weight moves these by tens of percent.
auc_p <- simulated |> dplyr::filter(analyte == "Osimertinib", PPTESTCD == "auclast")
auc_m <- simulated |> dplyr::filter(analyte == "AZ5104",      PPTESTCD == "auclast")
stopifnot(
  abs(median(auc_p$PPORRES) / 11840 - 1) < 0.25,
  abs(median(auc_m$PPORRES) /  1277 - 1) < 0.25
)
```

The metabolite-to-parent exposure ratio is an internally consistent check that
does not depend on either analyte's absolute calibration:

```{r mr_ratio}
mr_auc <- median(auc_m$PPORRES) / median(auc_p$PPORRES)
tibble::tibble(
  Metric               = "MRAUCss (AZ5104 / osimertinib)",
  Simulated            = round(mr_auc, 4),
  `Yang 2025 Table S5` = 0.1078
)
stopifnot(abs(mr_auc / 0.1078 - 1) < 0.30)
```

# Assumptions and deviations

* **The albumin effect on AZ5104 volume is negative.** Table 2 prints its
  magnitude (0.65) without a sign, and prints its bootstrap CI with the bounds
  reversed ("0.68 (1.28-0.42)"). Equation 5 and `$THETA(13) = -0.652465` both
  carry the negative sign and are taken as authoritative.

* **Table 2's between-subject-variability column is on the variance scale.**
  This is confirmed by the control stream's `$OMEGA` values and, independently,
  by back-transforming to the paper's own published %CV figures (chunk
  `omega_scale` above). The `$OMEGA BLOCK(2)` off-diagonal is a covariance, not
  the correlation coefficient its Table 2 row label implies.

* **Race coefficients are on a `(1 + theta)` fractional-deviation form**, not a
  power or exponential form. Table 2 lists 0.16 and 0.18; Equation 4 writes the
  corresponding multipliers as 1.16 and 1.18; the control stream's `CLM1ETHL`
  block computes `1 + THETA(n)`. Note this differs from the sibling
  `Brown_2017_osimertinib` model only in the estimated values, not the form.

* **Chinese and Other race indicators are retained with coefficients fixed at
  zero.** Backward elimination found both non-significant (Table S2) and the
  final control stream holds `$THETA(15)` and `$THETA(17)` at `0 FIX`. They are
  kept in the model file so the covariate screen stays auditable; they have no
  numerical effect.

* **The parent-to-metabolite transfer carries no molar stoichiometric
  correction.** Amount moves 1:1 and each analyte's concentration is then formed
  with its own molecular weight, per `S2` and `S3`. This is a deliberate
  departure from `Brown_2017_osimertinib`, which applies an explicit
  `mw_az5104 / mw_parent` factor in its ODE. Yang 2025's convention is what its
  own control stream fits, and it is what reproduces the Table S5 AZ5104
  exposure (to 0.6%, versus about 3% low with a molar correction).

* **Concentrations are in nM, not mg/L**, following the source throughout
  (assay ranges, additive residual errors, AUCss quartiles and Table S5 are all
  nM). `Brown_2017_osimertinib` uses mg/L and converts in its vignette; the two
  models are therefore not directly comparable without unit conversion.

* **Missing-covariate handling is not reproduced.** The control stream maps a
  missing weight or albumin (coded `-999`) onto a covariate factor of 1. This
  model expects observed values; Table 1 reports missingness of 0.4% for weight
  and 1.1% for albumin, so the effect on any simulation is negligible.

* **The cohort's covariate marginals are drawn independently.** Yang 2025 does
  not report the weight-albumin correlation or the joint distribution by race,
  so weight, albumin and race are sampled independently from the Table 1
  marginals. This widens the simulated exposure spread slightly relative to the
  real population.

* **Cmax is compared on `Cc` (without residual error).** Table S5's Cmax is an
  observed geometric mean from a sparse grid (pre-dose and 1, 2, 4 and 6 h at
  C3D1) that cannot resolve the model's 6.7 h median tmax, so the observed and
  model-derived peak quantities are not strictly the same measurement. The
  parent Cmax accordingly sits below the Table S5 value; AUC, which is far less
  sensitive to grid density, agrees closely.

## Not extracted: the exposure-response analyses

Yang 2025's second half is an exposure-response assessment of the FLAURA2
combination arm. It is deliberately **not** part of this model file, for two
reasons.

The efficacy analysis is a **Cox proportional-hazards model of
progression-free survival**. A Cox model is semi-parametric: it estimates
covariate coefficients against an unspecified baseline hazard that is never
estimated or published. Tables S3 and S4 report only minimum -2LL, likelihood
differences and p-values -- no hazard-ratio coefficients and no baseline hazard
-- so there is nothing that could be encoded as a simulable time-to-event model
without inventing a baseline hazard function.

The safety analysis is a set of logistic regressions of adverse-event
occurrence against exposure quartiles, reported graphically in Figure 5 with no
tabulated intercepts or slopes.

In any case both analyses are negative findings. No osimertinib or AZ5104
exposure metric (AUCss, Cmaxss, Cminss) was significant for PFS; the only
retained covariate was the number of pemetrexed cycles, which the authors
themselves caution is confounded by survivorship bias. No exposure-safety
relationship was found either, apart from an *inverse* association with AEs
leading to dose reduction that the paper interprets as non-causal. There is
therefore no drug-effect relationship to encode.
