---
title: "TQ-B3101 and TQ-B3101M (Yang 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TQ-B3101 and TQ-B3101M (Yang 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Yang_2021_TQB3101"))$reference`
* Description: `r rxode2::rxode(readModelDb("Yang_2021_TQB3101"))$description`
* Article: <https://doi.org/10.3389/fphar.2021.782518> (open access)

TQ-B3101 is an oral ALK/ROS1 kinase inhibitor that is rapidly converted by
amide hydrolysis to its active metabolite TQ-B3101M. Yang 2021 fitted the two
analytes sequentially in NONMEM 7.4 (FOCE-I): a one-compartment model with
first-order absorption for the parent was fitted first and fixed, then a
two-compartment model for the metabolite was added. All of the parent's
clearance forms the metabolite (the fraction metabolised, FM, was fixed to 1
for identifiability), and the metabolite's apparent clearance falls
exponentially with time since the first dose, by 41% at its asymptote (Table
3, footnote a). No covariate was retained.

The structures and molecular weights of both analytes are proprietary (Methods,
Analytical Methods). The packaged model therefore reports TQ-B3101M in
**TQ-B3101 mass equivalents**: `Cc_tqb3101m` is the concentration the metabolite
would have if each molecule weighed as much as a parent molecule. Multiplying by
the unpublished ratio MW(TQ-B3101M) / MW(TQ-B3101) gives the metabolite's own
mass concentration. The validation below keeps that unknown ratio visible
rather than guessing it.

# Population

The analysis pooled 40 patients from two Chinese studies (Table 1): a Phase 1
dose-escalation study in adults with advanced solid tumours (NCT03019276; single
doses of 100 or 200 mg, then 100, 200 or 300 mg once daily for 28 days) and a
Phase 2 single-arm study in adults and adolescents with relapsed or refractory
ALK-positive anaplastic large cell lymphoma (NCT04306887; 200, 250, 300 or 350
mg twice daily for 28 days). All doses were taken fasting. Table 2: 34 adults
(median age 51.5 years, range 28-73; median weight 59.0 kg, range 42.0-87.7) and
6 adolescents (median age 13.0 years, range 11-14; median weight 41.8 kg, range
32.9-68.0); 21 of 40 (52.5%) were male. The dataset held 375 quantifiable
TQ-B3101 and 658 quantifiable TQ-B3101M concentrations; the 340 and 42
below-quantitation samples were discarded.

The same information is available programmatically via
`readModelDb("Yang_2021_TQB3101")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Yang_2021_TQB3101.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | log(2850) L/h | Table 3 |
| `lvc` (V/F) | log(4200) L | Table 3 |
| `lka` (Ka) | log(51.9) 1/h | Table 3 |
| `fm` (FM) | fixed(1) | Methods, Pharmacokinetic Modeling; Figure 1 caption |
| `lcl_tqb3101m` (CLm0/Fm) | log(126) L/h | Table 3 |
| `lvc_tqb3101m` (Vcm/Fm) | log(2300) L | Table 3 |
| `lq_tqb3101m` (Qm/Fm) | log(113) L/h | Table 3 |
| `lvp_tqb3101m` (Vpm/Fm) | log(1480) L | Table 3 |
| `cl_exp_famp` (-TDPK) | -0.41 | Table 3 (TDPK = 0.41) |
| `lcl_exp_kdes` (KTDPK) | log(0.0363) 1/h | Table 3 |
| `etalcl`, `etalvc` | 0.0760, 0.1010 | Table 3 (28.1, 32.6 %CV); omega^2 = log(1 + CV^2) |
| `etalcl_tqb3101m`, `etalvc_tqb3101m`, `etalvp_tqb3101m` | 0.1100, 0.2484, 0.5300 | Table 3 (34.1, 53.1, 83.6 %CV) |
| `propSd`, `propSd_tqb3101m` | 0.711, 0.319 | Table 3 (sigma 71.1, 31.9 %CV) |
| `theta_i = theta_TV * exp(eta_i)` | n/a | Methods, Eq. 1 |
| `CLm(t) = CLm0 * [1 - TDPK * (1 - exp(-KTDPK * t))]` | n/a | Table 3, footnote a |
| Parent -> metabolite -> two-compartment ODEs | n/a | Figure 1 |

# Structural checks against the published half-lives

The Abstract and Results quote an elimination half-life of 1.0 h for TQ-B3101
and, for TQ-B3101M at steady state, distribution and elimination half-lives of
4.9 and 39.4 h. These follow in closed form from the typical values, using the
asymptotic metabolite clearance `CLm0 * (1 + cl_exp_famp)`.

```{r half-lives}
mod <- readModelDb("Yang_2021_TQB3101")
th <- rxode2::rxode2(mod)$theta

cl_p <- exp(th[["lcl"]])
vc_p <- exp(th[["lvc"]])
cl_m_inf <- exp(th[["lcl_tqb3101m"]]) * (1 + th[["cl_exp_famp"]])
vc_m <- exp(th[["lvc_tqb3101m"]])
q_m <- exp(th[["lq_tqb3101m"]])
vp_m <- exp(th[["lvp_tqb3101m"]])

k10 <- cl_m_inf / vc_m
k12 <- q_m / vc_m
k21 <- q_m / vp_m
sum_k <- k10 + k12 + k21
root <- sqrt(sum_k^2 - 4 * k10 * k21)

hl <- tibble::tibble(
  Quantity = c(
    "TQ-B3101 elimination half-life (h)",
    "TQ-B3101M distribution half-life at steady state (h)",
    "TQ-B3101M elimination half-life at steady state (h)",
    "TQ-B3101M apparent clearance at steady state (L/h)"
  ),
  Model = c(
    log(2) * vc_p / cl_p,
    log(2) / ((sum_k + root) / 2),
    log(2) / ((sum_k - root) / 2),
    cl_m_inf
  ),
  Published = c(1.0, 4.9, 39.4, 126 * (1 - 0.41))
) |>
  mutate(`Difference (%)` = 100 * (Model / Published - 1))

knitr::kable(hl, digits = 2, caption = "Closed-form half-lives versus Yang 2021 (Abstract; Results).")

# Deterministic identities of the typical values; the published half-lives are
# printed to one decimal, so each must agree to within that rounding (<= 3%).
stopifnot(all(abs(hl$`Difference (%)`) < 3))
```

# Time course of the metabolite clearance

```{r clearance-time}
tc <- tibble::tibble(time = seq(0, 240, by = 1)) |>
  mutate(cl_tqb3101m = exp(th[["lcl_tqb3101m"]]) *
    (1 + th[["cl_exp_famp"]] * (1 - exp(-exp(th[["lcl_exp_kdes"]]) * time))))

ggplot(tc, aes(time / 24, cl_tqb3101m)) +
  geom_line() +
  geom_hline(yintercept = cl_m_inf, linetype = "dashed") +
  labs(
    x = "Time since first dose (days)", y = "CLm/Fm (L/h)",
    title = "Typical TQ-B3101M apparent clearance",
    caption = "Table 3 footnote a of Yang 2021; dashed line = asymptote 74.3 L/h."
  )
```

The half-time of the decline is `log(2) / 0.0363` = 19.1 h, so the clearance is
within 1% of its asymptote after about 5.3 days. The authors attribute the
decline to auto-inhibition of CYP3A4 by the metabolite, as reported for
crizotinib (Discussion).

# Typical-value simulation and a steady-state identity

```{r typical}
mod_typ <- rxode2::zeroRe(mod)

ev_typ <- rxode2::et(amt = 300, cmt = "depot", ii = 12, addl = 55) |>
  rxode2::et(sort(unique(c(seq(0, 48, by = 0.1), seq(648, 672, by = 0.05)))), cmt = "central") |>
  as.data.frame() |>
  # Two residual-error endpoints: observation rows carry dvid = 1; a forward
  # solve still returns both Cc and Cc_tqb3101m on every row.
  mutate(dvid = ifelse(evid == 0, 1L, NA_integer_))
sim_typ <- rxode2::rxSolve(mod_typ, ev_typ, rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()

# At true steady state AUC0-tau of the metabolite equals fm * dose / CLm(inf)
# (the clearance has relaxed to within 1e-10 of its asymptote by day 28).
last <- sim_typ |> filter(time >= 660, time <= 672)
auc_num <- sum(diff(last$time) *
  (head(last$Cc_tqb3101m, -1) + tail(last$Cc_tqb3101m, -1)) / 2)
auc_closed <- 1000 * 300 / cl_m_inf
c(numeric = auc_num, closed_form = auc_closed)
# Solve against its own closed form: a pure numerical-error comparison.
# Trapezoidal error on the 0.05 h grid is ~1e-5 relative.
stopifnot(abs(auc_num / auc_closed - 1) < 1e-3)

sim_typ |>
  filter(time <= 48 | time >= 648) |>
  mutate(window = ifelse(time <= 48, "First two days", "Day 28 (steady state)")) |>
  select(time, window, `TQ-B3101` = Cc, `TQ-B3101M (TQ-B3101 mass eq.)` = Cc_tqb3101m) |>
  pivot_longer(-c(time, window), names_to = "analyte", values_to = "conc") |>
  filter(conc > 0) |>
  ggplot(aes(time, conc, colour = analyte)) +
  geom_line() +
  facet_wrap(~window, scales = "free_x") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Concentration (ng/mL)", colour = NULL,
    title = "Typical profiles, 300 mg TQ-B3101 twice daily"
  ) +
  theme(legend.position = "bottom")
```

The parent peaks within an hour and is essentially gone before the next dose
(half-life 1 h), while the metabolite peaks 1-3 h after the dose and
accumulates, consistent with the times to peak of 0.5-1 h and 1-3 h quoted in
the Discussion.

# Virtual adult cohorts

Original observed data are not available. Each regimen below is a separate arm
of 100 virtual patients who start at their first dose (the model's time-varying
clearance is indexed to time since the first dose). The model has no
covariates, so no demographic distribution is needed.

```{r cohort}
rxode2::rxSetSeed(20211)

make_arm <- function(n, regimen, amt, ii, n_doses, obs, id_offset) {
  ids <- id_offset + seq_len(n)
  dose_times <- (seq_len(n_doses) - 1) * ii
  dose <- tidyr::expand_grid(id = ids, time = dose_times) |>
    mutate(evid = 1L, amt = amt, cmt = "depot")
  obs_rows <- tidyr::expand_grid(id = ids, time = obs) |>
    mutate(evid = 0L, amt = 0, cmt = "central", dvid = 1L)
  bind_rows(dose, obs_rows) |>
    mutate(regimen = regimen) |>
    arrange(id, time, desc(evid))
}

obs_sd <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 24, 48, 72, 96, 144, 192, 240, 336)
obs_qd <- 648 + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24)
obs_bid <- 660 + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)

events <- bind_rows(
  make_arm(100, "100 mg single dose", 100, 24, 1, obs_sd, 0L),
  make_arm(100, "200 mg single dose", 200, 24, 1, obs_sd, 100L),
  make_arm(100, "100 mg QD, day 28", 100, 24, 28, obs_qd, 200L),
  make_arm(100, "200 mg QD, day 28", 200, 24, 28, obs_qd, 300L),
  make_arm(100, "300 mg BID, day 28", 300, 12, 56, obs_bid, 400L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

sim <- rxode2::rxSolve(mod, events = events, keep = "regimen") |>
  as.data.frame()
```

# PKNCA: accumulation of TQ-B3101M

The Methods report that in the Phase 1 study "the average AUC0-tau,ss of
TQ-B3101M after repeated dosing was 20-40% higher than the average AUC0-inf of
TQ-B3101M after a single dose" at 100 and 200 mg. With linear kinetics and a
constant clearance those two AUCs would be equal; the excess is produced by the
time-dependent clearance, so it is the most direct check of that part of the
model.

```{r pknca}
nca_input <- function(sim, analyte) {
  d <- sim |>
    filter(!is.na(.data[[analyte]])) |>
    transmute(id, time, regimen, Cc = pmax(.data[[analyte]], 0))
  # Time-zero rows for the single-dose arms (pre-dose concentration 0).
  bind_rows(
    d,
    d |> filter(grepl("single", regimen)) |> distinct(id, regimen) |>
      mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, regimen, time, .keep_all = TRUE) |>
    arrange(id, regimen, time)
}

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, regimen)

intervals <- data.frame(
  regimen = c("100 mg single dose", "200 mg single dose", "100 mg QD, day 28",
              "200 mg QD, day 28", "300 mg BID, day 28"),
  start = c(0, 0, 648, 648, 660),
  end = c(Inf, Inf, 672, 672, 672),
  cmax = TRUE,
  tmax = TRUE,
  aucinf.obs = c(TRUE, TRUE, FALSE, FALSE, FALSE),
  auclast = c(FALSE, FALSE, TRUE, TRUE, TRUE)
)

run_nca <- function(analyte) {
  conc <- PKNCA::PKNCAconc(nca_input(sim, analyte), Cc ~ time | regimen + id)
  dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc, dose, intervals = intervals))
}

nca_m <- run_nca("Cc_tqb3101m")
nca_p <- run_nca("Cc")

nca_m_df <- as.data.frame(nca_m$result)
auc_by_subject <- nca_m_df |>
  filter(PPTESTCD %in% c("aucinf.obs", "auclast")) |>
  select(regimen, id, PPORRES)
```

```{r pknca-accumulation}
acc <- auc_by_subject |>
  mutate(
    dose = ifelse(grepl("^100", regimen), "100 mg", "200 mg"),
    type = ifelse(grepl("single", regimen), "single", "repeated")
  ) |>
  filter(!grepl("BID", regimen)) |>
  group_by(dose, type) |>
  summarise(mean_auc = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = mean_auc) |>
  mutate(excess = 100 * (repeated / single - 1))

acc |>
  rename(
    "Dose" = dose,
    "Mean single-dose AUC0-inf (ng*h/mL, mass eq.)" = single,
    "Mean day-28 AUC0-24 (ng*h/mL, mass eq.)" = repeated,
    "Excess (%)" = excess
  ) |>
  knitr::kable(digits = 1, caption = "Repeated-dose versus single-dose TQ-B3101M exposure (published: 20-40% higher).")

# The two AUCs share the unknown molecular-weight ratio, so the ratio is free of
# it. Each per-dose excess compares two independent 100-subject means and
# carries ~7 percentage points of cohort noise, so only the pooled excess is
# gated here, with a bound a constant clearance (0%) or a mis-signed amplitude
# (negative) fails by several standard errors. The tight check is the
# deterministic typical-value excess below.
stopifnot(mean(acc$excess) > 0, mean(acc$excess) < 60)

# Typical-value excess: day-28 AUC0-24 = dose / CLm(inf) against the
# single-dose AUC0-inf of the typical subject (dense grid, numeric solve).
ev_sd <- rxode2::et(amt = 100, cmt = "depot") |>
  rxode2::et(c(seq(0, 24, by = 0.05), seq(24.5, 2000, by = 0.5)), cmt = "central") |>
  as.data.frame() |>
  mutate(dvid = ifelse(evid == 0, 1L, NA_integer_))
sd_typ <- rxode2::rxSolve(mod_typ, ev_sd, rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()
auc_sd_typ <- sum(diff(sd_typ$time) *
  (head(sd_typ$Cc_tqb3101m, -1) + tail(sd_typ$Cc_tqb3101m, -1)) / 2)
typ_excess <- 100 * ((1000 * 100 / cl_m_inf) / auc_sd_typ - 1)
typ_excess
# Deterministic: must fall in the 20-40% the Methods report.
stopifnot(typ_excess > 20, typ_excess < 40)
```

The simulated excess of repeated-dose over single-dose metabolite exposure is
`r paste(round(acc$excess), collapse = "% and ")`% at 100 and 200 mg, against
the 20-40% the paper reports from the Phase 1 non-compartmental analysis. The typical-value excess, free of cohort noise, is
`r round(typ_excess)`%, inside the published range; the per-dose cohort
values scatter around it by the sampling noise of two independent 100-subject
arms.

## Adult steady-state exposure

The Discussion quotes a TQ-B3101M AUC0-12h,ss of 4,251 ng*h/mL after 300 mg
TQ-B3101 twice daily. That number is in metabolite mass units, whereas the
packaged model is in parent mass equivalents, so the comparison below is
expected to differ by the unpublished factor MW(TQ-B3101M) / MW(TQ-B3101); the
paper does not say whether 4,251 is an observed or a simulated mean.

```{r adult-ss}
published_adult <- tibble::tibble(regimen = "300 mg BID, day 28", auclast = 4251)

adult_res <- nca_m
adult_res$result <- adult_res$result |> filter(regimen == "300 mg BID, day 28")

cmp_adult <- nlmixr2lib::ncaComparisonTable(
  simulated = adult_res,
  reference = published_adult,
  by = "regimen",
  params = "auclast",
  units = c(auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_adult, caption = "TQ-B3101M AUC0-12h,ss after 300 mg BID: simulated (TQ-B3101 mass equivalents) versus the Discussion value (metabolite mass units).")

adult_ratio <- 4251 / median(auc_by_subject$PPORRES[auc_by_subject$regimen == "300 mg BID, day 28"])
adult_ratio
```

The published-to-simulated ratio of about `r signif(adult_ratio, 2)` is the
factor that the missing molecular-weight ratio would have to supply, together
with any difference between the paper's summary statistic and the median used
here.

## TQ-B3101 (parent) NCA

```{r parent-nca}
as.data.frame(nca_p$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "auclast")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  knitr::kable(digits = 2, caption = "Median simulated TQ-B3101 NCA parameters by regimen (no published NCA values to compare).")
```

# Replicates Table 6 and Figure 4: paediatric BSA-tiered dosing

The paper's paediatric simulation is not part of the fitted model. It applied
body-weight allometry with fixed exponents of 0.75 on every clearance and 1 on
every volume, "scaled to a median weight of adult patients" (Methods, Clinical
Simulations). The median adult weight in Table 2 is 59.0 kg. The reference
weight is not otherwise stated; 59.0 kg is used here. The allometric lines are
added to a copy of the model inside this article only.

```{r ped-model}
mod_ped <- mod |>
  rxode2::model({
    cl <- exp(lcl + etalcl) * (WT / 59)^0.75
    vc <- exp(lvc + etalvc) * (WT / 59)
    cl_tqb3101m0 <- exp(lcl_tqb3101m + etalcl_tqb3101m) * (WT / 59)^0.75
    vc_tqb3101m <- exp(lvc_tqb3101m + etalvc_tqb3101m) * (WT / 59)
    q_tqb3101m <- exp(lq_tqb3101m) * (WT / 59)^0.75
    vp_tqb3101m <- exp(lvp_tqb3101m + etalvp_tqb3101m) * (WT / 59)
  })
```

The paper's 1,000 virtual children aged 6 to under 18 years came from the PK-Sim
Asian population database (Table 4: median weight 42.3 kg, range 15.8-102.3;
median BSA 1.3 m^2, range 0.7-2.2). That database is not reproduced here.
Instead, BSA was drawn from a normal distribution with mean 1.3 m^2 and SD 0.3
m^2 truncated to 0.74-2.2 m^2 (the Table 4 median, and a spread whose +/-3 SD
spans the Table 4 range), the first 100 children falling in each BSA tier of
Table 5 were kept, and body weight was obtained by inverting the weight-only
paediatric BSA formula of Costeff (1966), `BSA = (4 WT + 7) / (WT + 90)`. That formula maps Table 4's
median weight of 42.3 kg to 1.33 m^2 and its weight range to 0.66-2.16 m^2,
close to the tabulated BSA summary.

```{r ped-cohort}
tiers <- tibble::tibble(
  tier = c("0.74-0.89", "0.90-1.22", "1.23-1.38", "1.39-1.59", ">1.6"),
  bsa_lo = c(0.74, 0.90, 1.23, 1.39, 1.60),
  bsa_hi = c(0.89, 1.22, 1.38, 1.59, 2.20),
  dose = c(250, 350, 400, 450, 550)
)

obs_ped <- 13 * 24 + c(0, 0.5, 1, 2, 4, 6, 10, 12)  # Methods sampling grid, day 14

set.seed(20212)
bsa_pool <- rnorm(20000, mean = 1.3, sd = 0.3)
bsa_pool <- bsa_pool[bsa_pool >= 0.74 & bsa_pool <= 2.2]
ped_subjects <- tibble::tibble(
  BSA = bsa_pool,
  # Table 5 tiers are printed with gaps (0.89 | 0.90, 1.59 | 1.6); split them
  # at the midpoints.
  tier = as.character(cut(BSA, c(0.74, 0.895, 1.225, 1.385, 1.595, 2.2),
    labels = tiers$tier, include.lowest = TRUE
  ))
) |>
  group_by(tier) |>
  slice_head(n = 100) |>
  ungroup() |>
  left_join(select(tiers, tier, dose), by = "tier") |>
  mutate(
    id = seq_len(n()),
    WT = (90 * BSA - 7) / (4 - BSA)
  )
stopifnot(all(table(ped_subjects$tier) == 100L))

ped_events <- bind_rows(
  tidyr::expand_grid(ped_subjects, time = seq(0, 13 * 24, by = 12)) |>
    mutate(evid = 1L, amt = dose, cmt = "depot"),
  tidyr::expand_grid(ped_subjects, time = obs_ped) |>
    mutate(evid = 0L, amt = 0, cmt = "central", dvid = 1L)
) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(ped_events[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20213)
ped_sim <- rxode2::rxSolve(mod_ped, events = ped_events, keep = c("tier", "BSA")) |>
  as.data.frame()
```

```{r ped-nca}
ped_conc <- ped_sim |>
  filter(!is.na(Cc_tqb3101m)) |>
  transmute(id, tier, time, Cc = pmax(Cc_tqb3101m, 0))
ped_dose <- ped_events |>
  filter(evid == 1) |>
  select(id, tier, time, amt)
ped_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(ped_conc, Cc ~ time | tier + id),
  PKNCA::PKNCAdose(ped_dose, amt ~ time | tier + id),
  intervals = data.frame(start = 13 * 24, end = 13 * 24 + 12, cmax = TRUE, auclast = TRUE)
))

ped_ind <- as.data.frame(ped_nca$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(id, tier, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

table6 <- tibble::tibble(
  tier = tiers$tier,
  pub_cmax_gm = c(629.5, 637.1, 575.4, 581.7, 578.8),
  pub_auc_gm = c(6431.7, 6614.4, 6118.1, 6272.9, 6229.2)
)

geomean <- function(x) exp(mean(log(x)))
ped_summary <- ped_ind |>
  group_by(tier) |>
  summarise(sim_cmax_gm = geomean(cmax), sim_auc_gm = geomean(auclast), .groups = "drop") |>
  left_join(table6, by = "tier") |>
  mutate(
    tier = factor(tier, levels = tiers$tier),
    ratio_auc = pub_auc_gm / sim_auc_gm,
    ratio_cmax = pub_cmax_gm / sim_cmax_gm
  ) |>
  arrange(tier)
stopifnot(nrow(ped_summary) == 5L, !anyNA(ped_summary$pub_auc_gm))

ped_summary |>
  rename(
    "BSA tier (m^2)" = tier,
    "Sim. Cmax,ss GM (mass eq.)" = sim_cmax_gm,
    "Table 6 Cmax,ss GM" = pub_cmax_gm,
    "Sim. AUC0-12h,ss GM (mass eq.)" = sim_auc_gm,
    "Table 6 AUC0-12h,ss GM" = pub_auc_gm,
    "Table 6 / sim. (AUC)" = ratio_auc,
    "Table 6 / sim. (Cmax)" = ratio_cmax
  ) |>
  knitr::kable(digits = 2, caption = "Replicates Table 6 of Yang 2021: TQ-B3101M geometric-mean exposure by BSA tier (ng/mL, ng*h/mL). Simulated values are in TQ-B3101 mass equivalents; Table 6 is in metabolite mass units.")

# Structural check, independent of the unknown molecular-weight ratio: the
# BSA-tiered doses were chosen to give flat exposure across tiers, and with
# allometric scaling the published geometric means span only 8%. Without the
# weight scaling the tiers would span 2.2-fold (250 to 550 mg). The simulated
# tiers carry ~3.5% sampling error each, so a 25% band leaves ample headroom.
stopifnot(max(ped_summary$ratio_auc) / min(ped_summary$ratio_auc) < 1.25)
# Units check: the published-to-simulated factor is a molecular-weight ratio of
# a hydrolysis product to its parent, so it must be of order 1; a mg/ng or
# nmol slip would move it by orders of magnitude.
stopifnot(median(ped_summary$ratio_auc) > 0.5, median(ped_summary$ratio_auc) < 1.5)
```

```{r figure-4}
ped_ind |>
  mutate(tier = factor(tier, levels = tiers$tier)) |>
  pivot_longer(c(cmax, auclast), names_to = "metric", values_to = "value") |>
  mutate(metric = ifelse(metric == "cmax", "Cmax,ss (ng/mL)", "AUC0-12h,ss (ng*h/mL)")) |>
  ggplot(aes(tier, value)) +
  geom_boxplot(outlier.size = 0.5) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    x = "BSA tier (m^2)", y = "TQ-B3101M (TQ-B3101 mass equivalents)",
    title = "Paediatric TQ-B3101M exposure after BSA-tiered dosing",
    caption = "Replicates Figure 4 of Yang 2021 (simulated cohort described above)."
  )
```

Across the five tiers the ratio of the Table 6 geometric means to the simulated
ones is nearly constant (AUC ratios
`r paste(signif(ped_summary$ratio_auc, 3), collapse = ", ")`), so the
allometric bridge and the BSA-tiered doses reproduce the paper's flat exposure
profile. The common factor (median `r signif(median(ped_summary$ratio_auc), 3)`)
is below 1, as expected for a molecular-weight ratio of a hydrolysis product to
its parent. It is close to 275 / 300 = 0.917, the TQ-B3101-to-crizotinib dose
equivalence the Discussion quotes (the paper's Eq. 4 converts crizotinib doses
by the molecular-weight ratio of the two drugs), which is consistent with, but
does not establish, a metabolite of similar molecular weight to crizotinib.

The adult factor above (`r signif(adult_ratio, 3)`) is larger than the
paediatric one. The paediatric comparison is simulation against simulation
from the same model, whereas the paper does not say how the adult 4,251
ng*h/mL was obtained; if it is an observed mean from the few Phase 2 patients
on 300 mg BID, a difference of this size is within what a small sample and a
mean-versus-median comparison produce. Neither factor is folded into the model.

# Assumptions and deviations

- **Metabolite concentration units.** The molecular weights of TQ-B3101 and
  TQ-B3101M are proprietary. With the fraction metabolised fixed to 1, the
  packaged model yields TQ-B3101M in TQ-B3101 mass equivalents. The maintainers
  did not back-solve a molecular-weight ratio from Table 6 or from the adult
  AUC, because either would tune the model to a validation target; the implied
  factor is reported above instead.
- **Time-dependent clearance sign convention.** The paper's TDPK = 0.41 is a
  positive "maximum fraction reduction". The model encodes it as the signed
  fractional amplitude `cl_exp_famp = -0.41` in
  `CLm(t) = CLm0 * (1 + cl_exp_famp * (1 - exp(-KTDPK * t)))`, which is
  algebraically identical and follows the sign convention of `kel_exp_famp`
  (negative = decline). Time `t` is time since the first dose, so simulations
  must start at the first dose; a washout between a single-dose and a
  repeated-dose period, as in the Phase 1 design, is not modelled as a reset of
  the clearance.
- **IIV conversion.** Table 3 reports IIV as %CV; variances were computed as
  `omega^2 = log(1 + CV^2)`. No correlations are reported, so the etas are
  independent. Ka and Qm/Fm carry no IIV.
- **BQL handling.** The final model used the M1 method (below-quantitation
  samples discarded) for both analytes; the paper also evaluated M3 for the
  parent but reports only one set of estimates.
- **Paediatric simulation.** The allometric exponents (0.75 / 1), reference
  weight (59.0 kg, the adult median of Table 2) and BSA-tiered doses come from
  the paper's simulation, not from the fitted model, which retained no
  covariate. The virtual children were drawn uniformly within each BSA tier and
  their weights derived from the Costeff (1966) BSA formula because the PK-Sim
  population the paper used is not available; steady state was taken on day 14,
  after the clearance has settled, because the paper does not state the day.
- **Adult cohort.** The accumulation check uses separate virtual arms for the
  single-dose and repeated-dose regimens, each starting at its own first dose.
- No erratum or correction notice for Yang 2021 was found as of 2026-09-30.
