---
title: "COPD FEV1 combined aggregated + individual meta-analysis (Yang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{COPD FEV1 combined aggregated + individual meta-analysis (Yang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Yang 2026 builds a **combined aggregated-data + individual-patient-data (ADIPD)**
model of morning trough FEV1 in COPD, and in doing so builds two models that are
packaged here separately:

* `Yang_2026_copd_fev1_ipd` -- the individual-patient-data model, fit to 2,241
  patients in two fluticasone furoate / vilanterol trials.
* `Yang_2026_copd_fev1_adipd_mbma` -- the combined model, fit to those individual
  records jointly with 4,137 arm-mean FEV1 observations from 296 published
  trials covering 23 compounds.

* Citation: `r rxode2::rxode(readModelDb("Yang_2026_copd_fev1_adipd_mbma"))$reference`
* Article: <https://doi.org/10.1002/psp4.70059>
* Supporting Information (Tables S1-S5 and both NONMEM control streams):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12896389/supplementaryFiles>

```{r load-models}
ui_ipd   <- rxode2::rxode(readModelDb("Yang_2026_copd_fev1_ipd"))
ui_adipd <- rxode2::rxode(readModelDb("Yang_2026_copd_fev1_adipd_mbma"))
```

Both models are **purely algebraic** -- FEV1 is a closed-form function of time and
covariates, with no ODE states and no PK layer. Drug effects are driven by
per-arm total daily dose supplied as `CONMED_<drug>_DOSE` covariate columns.

```{r algebraic-check}
stopifnot(length(ui_ipd$state) == 0L, length(ui_adipd$state) == 0L)
```

# Population

The individual-patient data are the two 24-week, randomized, smoking-status
stratified, placebo-controlled fluticasone furoate / vilanterol trials
NCT01053988 (Kerwin 2013, n = 1025) and NCT01054885 (Martinez 2013, n = 1216),
pooled to 2,241 patients: mean age 62 years (SD 8.8, range 40-85), 69.7% male,
54.2% current smokers, GOLD spirometric stage 2 in 46.5% / stage 3 in 44.1% /
stage 4 in 8.66%, mean on-treatment FEV1 1.372 L (Supporting Information
Table S1).

The aggregated data are the 4,137 arm-mean trough FEV1 observations from 298
studies of 250,543 patients assembled for the predecessor meta-analysis
(Llanos-Paez 2023), covering 23 compounds in mono-, dual- and triple-therapy up
to 24 November 2020. The two individual-patient studies are among those 298 and
were **removed from the aggregated side** when fitting the combined model so
that no observation contributes twice (paper Section 2.1). Arm sizes span 18 to
5,724 patients, which is what makes the central-limit-theorem approximation of
Section 2.3.2 valid.

The same information is available programmatically:

```{r population, results = "asis"}
pop <- ui_adipd$population
cat(paste0("* **", names(pop), "**: ", vapply(pop, paste, character(1), collapse = "; "), collapse = "\n"))
```

# Source trace

Each `ini()` entry in the two model files carries an in-file comment naming its
source location. The table below collects the structural origins.

| Model component | Source location |
|-----------------|-----------------|
| Overall structure `FEV1 = B + PBO - DP + E + PostBD correction + err` | Paper Equation 10 (Section 2.4) |
| Individual-patient parameter values | Supporting Information Table S2 |
| Combined-model parameter values | Supporting Information Table S3 |
| Individual-patient model equations | Supporting Information, "NONMEM control stream for the IPD model" |
| Combined-model equations | Supporting Information, "NONMEM control stream for the combined ADIPD model" |
| Covariate relationships retained per model | Supporting Information Table S4 |
| Normal approximation to the arm-mean of a log-normal | Paper Section 2.3.2, Equations 1-8 |
| Linearization of the placebo mixture, the baseline step function and the severity covariate | Paper Section 2.3.1 |
| Post-SABD reversibility shift of 0.18 L | Paper Section 2.4, note following Equation 10 |
| Aggregated residual error scaled by `1/sqrt(N)` | Paper Equation 10, `err = eps_AD / sqrt(N_ij)` |

Individual-patient-model values are taken from **Table S2**, not from that
control stream's `$THETA` records, which carry that run's *initial* estimates
(1.18 vs the final 1.17 for baseline, and similar small differences elsewhere).
Combined-model values are taken from the control stream's `$THETA`, whose
converged values agree with Table S3 to every displayed digit while carrying
more precision (1.06908 vs the tabulated 1.07).

```{r param-counts}
n_par <- function(ui) sum(!is.na(ui$iniDf$ntheta))
n_eta <- function(ui) sum(!is.na(ui$iniDf$neta1))
data.frame(
  Model = c("Yang_2026_copd_fev1_ipd", "Yang_2026_copd_fev1_adipd_mbma"),
  `Fixed effects` = c(n_par(ui_ipd), n_par(ui_adipd)),
  `Random effects` = c(n_eta(ui_ipd), n_eta(ui_adipd)),
  check.names = FALSE
) |> knitr::kable()
```

# Simulation helpers

Neither model consumes rxode2 dose events, so an event table is simply a frame
of observation rows carrying the covariate columns. `evid = 0` throughout.

```{r helpers}
ipd_covs <- function(age = 62, sexf = 0, smoke = 1, gold = 3, vi = 0, ff = 0) {
  list(AGE = age, SEXF = sexf, SMOKE = smoke, DIS_COPD_GOLD = gold,
       CONMED_VILANTEROL_DOSE = vi, CONMED_FLUTICASONEFUROATE_DOSE = ff)
}

# Every CONMED_<drug>_DOSE column the combined model reads, defaulted to zero
# (an all-zero row is a placebo / background-therapy-only arm).
adipd_drugs <- c(
  "ACLIDINIUM", "ARFORMOTEROL", "AZD9668", "BATEFENTEROL", "BEA2180",
  "BECLOMETHASONE", "BUDESONIDE", "CILOMILAST", "FLUTICASONEFUROATE",
  "FLUTICASONEPROPIONATE", "FORMOTEROL", "GLYCOPYRRONIUM", "GSK233705",
  "INDACATEROL", "MOMETASONE", "OLODATEROL", "PH797804", "REVEFENACIN",
  "ROFLUMILAST", "SALMETEROL", "TIOTROPIUM", "UMECLIDINIUM", "VILANTEROL"
)

adipd_covs <- function(...) {
  base <- as.list(setNames(rep(0, length(adipd_drugs)),
                           paste0("CONMED_", adipd_drugs, "_DOSE")))
  base <- c(base, list(
    FORM_ACLIDINIUM_BID = 0, FORM_MOMETASONE_BID = 0, FORM_OLODATEROL_BID = 0,
    FORM_BATEFENTEROL_BID = 0, FORM_TIOTROPIUM_SMI = 0,
    FORM_TIOTROPIUM_OPENLABEL = 0,
    # Reference arm: aggregated record, reference covariate values, no
    # background therapy, no reconciliation term.
    DTYPE_AGGREGATED = 1, NARM = 200, MEAS_POSTBD = 0, FEV1_PBD_ANCHOR = 0,
    OCS_NONRESPONDER = 0, INCL_EXAC_REQUIRED = 0,
    AGE = 63.4, SEXF = 0.329, SMOKE = 0.463, DIS_COPD_GOLD = 3,
    DIS_COPD_GOLD_LOW = 2, DIS_COPD_GOLD_HIGH = 4,
    BGTHER_ICS_RUNIN_PCT = 0, BGTHER_LABA_RUNIN_PCT = 0,
    BGTHER_LAAC_RUNIN_PCT = 0, BGTHER_ICS_MAINT_PCT = 0,
    BGTHER_LABA_MAINT_PCT = 0, BGTHER_LAAC_MAINT_PCT = 0))
  mods <- list(...)
  for (n in names(mods)) base[[n]] <- mods[[n]]
  base
}

make_events <- function(covs_list, times) {
  do.call(rbind, lapply(seq_along(covs_list), function(i) {
    cbind(data.frame(id = i, time = times, evid = 0L, amt = NA_real_),
          as.data.frame(covs_list[[i]]))
  }))
}

# zeroRe() gives the typical-value prediction with every random effect at zero.
solve_typical <- function(ui, ev, ...) {
  rxode2::rxSolve(rxode2::zeroRe(ui), ev, returnType = "data.frame", ...)
}
```

# Check 1 -- the individual-patient model reproduces its closed form exactly

The individual-patient model is small enough to write out independently. Because
both sides use the same parameter values and no random draw is involved, this is
a pure numerical-identity check and a tight bound is the correct assertion.

```{r check-closed-form}
closed_form_ipd <- function(t, age, sexf, smoke, gold, vi, ff) {
  b <- 1.17 *
    (1 - 0.0144 * (age - 62)) *
    (1 + ifelse(gold <= 3, -0.47, -0.368) * (gold - 3)) *
    (1 - 0.0419 * (1 - smoke)) *
    (1 - 0.248 * sexf)
  dp <- 0.0241 * (1 - 1.89 * (gold - 3)) * t / 52
  ed50_vi <- exp(0.709); ed50_ff <- exp(2.43)
  emax_vi <- 0.117 / 25 * (ed50_vi + 25) *
    (1 - 0.0119 * (age - 62)) *
    (1 + ifelse(gold <= 3, -0.163, -0.533) * (gold - 3))
  emax_ff <- 0.0304 / 100 * (ed50_ff + 100)
  b - dp + (t > 0) * (emax_vi * vi / (vi + ed50_vi) + emax_ff * ff / (ff + ed50_ff))
}

grid_ipd <- expand.grid(
  age = c(45, 62, 78), sexf = c(0, 1), smoke = c(0, 1),
  gold = c(2, 3, 4), vi = c(0, 25), ff = c(0, 50, 100, 200)
)
ev_ipd <- make_events(
  lapply(seq_len(nrow(grid_ipd)), function(i) {
    with(grid_ipd[i, ], ipd_covs(age, sexf, smoke, gold, vi, ff))
  }),
  times = c(0, 4, 12, 24)
)
sim_ipd <- solve_typical(ui_ipd, ev_ipd)

expected_ipd <- closed_form_ipd(
  sim_ipd$time, sim_ipd$AGE, sim_ipd$SEXF, sim_ipd$SMOKE, sim_ipd$DIS_COPD_GOLD,
  sim_ipd$CONMED_VILANTEROL_DOSE, sim_ipd$CONMED_FLUTICASONEFUROATE_DOSE
)
max_abs_err <- max(abs(sim_ipd$FEV1 - expected_ipd))
cat(sprintf("%d covariate/time combinations; max |solve - closed form| = %.3e L\n",
            nrow(sim_ipd), max_abs_err))

# Deterministic identity: both sides are the same algebra on the same numbers,
# so the only difference possible is floating-point rounding.
stopifnot(nrow(sim_ipd) > 500, max_abs_err < 1e-10)
```

# Check 2 -- reference-dose efficacies are recovered exactly

The source parameterises each dose-response drug by its effect **at a reference
dose** rather than by Emax, via `Emax = EffRef / RefDose * (ED50 + RefDose)`.
Giving exactly the reference dose must therefore return exactly the published
reference efficacy. This is the check that catches a mis-transcribed ED50, a
mis-transcribed reference dose, or a units slip between the two -- all of which
would leave the model solving happily while predicting the wrong effect.

```{r check-refdose-ipd}
ev_ref <- make_events(list(
  ipd_covs(),                     # placebo
  ipd_covs(vi = 25),              # vilanterol at its 25 ug/day reference dose
  ipd_covs(ff = 100)              # fluticasone furoate at its 100 ug/day reference
), times = c(0, 24))
sim_ref <- solve_typical(ui_ipd, ev_ref)
wk24 <- sim_ref[sim_ref$time == 24, ]

ipd_ref_tab <- data.frame(
  Drug = c("Vilanterol 25 ug/day", "Fluticasone furoate 100 ug/day"),
  `Published reference efficacy (L)` = c(0.117, 0.0304),
  `Simulated effect vs placebo (L)` = round(wk24$FEV1[2:3] - wk24$FEV1[1], 6),
  check.names = FALSE
)
knitr::kable(ipd_ref_tab)

stopifnot(max(abs(ipd_ref_tab[[3]] - ipd_ref_tab[[2]])) < 1e-6)
```

The combined model applies one further multiplier before the effect reaches the
prediction: the predicted baseline scales the bronchodilator and
anti-inflammatory effects through `rel_bd` and `rel_ai`. At the reference
covariate values the arm-mean baseline is below the 1.2 L centring point, so
every drug's effect is attenuated by the same factor -- which makes the
identity below an exact test of all 13 dose-response drugs at once.

Six of the 13 also carry a class effect-onset time course, which is still a
fraction of a percent short of its plateau at week 24. That factor is read back
from the solve rather than assumed to be 1, so the identity below simultaneously
checks each drug's reference efficacy, its ED50, its reference dose and its
onset rate.

```{r check-refdose-adipd}
ref_doses <- list(
  ACLIDINIUM        = list(dose = 200, effref = 0.075184,  cls = "bd", tc = NA),
  BUDESONIDE        = list(dose = 320, effref = 0.0336889, cls = "ai", tc = "tc_ics"),
  FORMOTEROL        = list(dose = 18,  effref = 0.0699607, cls = "bd", tc = NA),
  INDACATEROL       = list(dose = 75,  effref = 0.12618,   cls = "bd", tc = "tc_laba"),
  GLYCOPYRRONIUM    = list(dose = 100, effref = 0.12743,   cls = "bd", tc = "tc_laac"),
  ROFLUMILAST       = list(dose = 500, effref = 0.0811084, cls = "ai", tc = "tc_pde4"),
  TIOTROPIUM        = list(dose = 18,  effref = 0.122236,  cls = "bd", tc = "tc_laac"),
  VILANTEROL        = list(dose = 25,  effref = 0.109036,  cls = "bd", tc = NA),
  REVEFENACIN       = list(dose = 175, effref = 0.148945,  cls = "bd", tc = NA),
  OLODATEROL        = list(dose = 5,   effref = 0.0889892, cls = "bd", tc = NA),
  BATEFENTEROL      = list(dose = 400, effref = 0.195297,  cls = "bd", tc = NA),
  FLUTICASONEFUROATE = list(dose = 100, effref = 0.0400295, cls = "ai", tc = NA)
)

ev_adipd_ref <- make_events(
  c(list(adipd_covs()),
    lapply(names(ref_doses), function(d) {
      a <- list(); a[[paste0("CONMED_", d, "_DOSE")]] <- ref_doses[[d]]$dose
      do.call(adipd_covs, a)
    })),
  times = c(0, 24)
)
sim_adipd_ref <- solve_typical(ui_adipd, ev_adipd_ref)
w24 <- sim_adipd_ref[sim_adipd_ref$time == 24, ]

# rel_bd / rel_ai are model outputs, so the expected value is read from the
# solve rather than hand-recomputed.
rel_bd <- w24$rel_bd[1]; rel_ai <- w24$rel_ai[1]
tc_at_24 <- function(nm) if (is.na(nm)) 1 else w24[[nm]][1]

adipd_ref_tab <- data.frame(
  Drug = names(ref_doses),
  `Reference dose (ug/day)` = vapply(ref_doses, function(x) x$dose, numeric(1)),
  `Published reference efficacy (L)` = vapply(ref_doses, function(x) x$effref, numeric(1)),
  `Onset fraction at week 24` = vapply(ref_doses, function(x) tc_at_24(x$tc), numeric(1)),
  `Expected effect (L)` = vapply(ref_doses, function(x)
    x$effref * tc_at_24(x$tc) * if (x$cls == "bd") rel_bd else rel_ai, numeric(1)),
  `Simulated effect vs placebo (L)` = w24$FEV1[-1] - w24$FEV1[1],
  check.names = FALSE, row.names = NULL
)
knitr::kable(adipd_ref_tab, digits = 6)

cat(sprintf("Baseline scaling at the reference arm: rel_bd = %.6f, rel_ai = %.6f\n",
            rel_bd, rel_ai))
stopifnot(
  nrow(adipd_ref_tab) == 12L,
  max(abs(adipd_ref_tab[["Simulated effect vs placebo (L)"]] -
            adipd_ref_tab[["Expected effect (L)"]])) < 1e-9
)
```

Tiotropium via the Respimat soft-mist inhaler is the one dose-response whose
ED50 is *derived* rather than estimated, so it is checked separately at its own
5 ug/day reference dose.

```{r check-tiotropium-smi}
ev_smi <- make_events(list(
  adipd_covs(),
  adipd_covs(CONMED_TIOTROPIUM_DOSE = 5, FORM_TIOTROPIUM_SMI = 1),
  adipd_covs(CONMED_TIOTROPIUM_DOSE = 18, FORM_TIOTROPIUM_OPENLABEL = 1)
), times = c(0, 24))
s_smi <- solve_typical(ui_adipd, ev_smi)
w_smi <- s_smi[s_smi$time == 24, ]
eff_smi <- w_smi$FEV1[2] - w_smi$FEV1[1]
eff_ol  <- w_smi$FEV1[3] - w_smi$FEV1[1]
tc_laac_24 <- w_smi$tc_laac[1]

cat(sprintf("Respimat 5 ug/day:    simulated %.6f L, expected %.6f L\n",
            eff_smi, 0.120457 * tc_laac_24 * rel_bd))
cat(sprintf("Open-label 18 ug/day: simulated %.6f L, expected %.6f L (0.918377 x blinded)\n",
            eff_ol, 0.918377 * 0.122236 * tc_laac_24 * rel_bd))
stopifnot(
  abs(eff_smi - 0.120457 * tc_laac_24 * rel_bd) < 1e-9,
  abs(eff_ol - 0.918377 * 0.122236 * tc_laac_24 * rel_bd) < 1e-9
)
```

# Check 3 -- covariate directions match Table S4

Supporting Information Table S4 tabulates which covariate relationships each
model retained and in which direction. The signs below are read off the packaged
models by perturbing one covariate at a time from the reference arm.

```{r check-covariate-signs}
delta_ipd <- function(mod_args, t = 24) {
  ev <- make_events(list(do.call(ipd_covs, list()),
                         do.call(ipd_covs, mod_args)), times = c(0, t))
  s <- solve_typical(ui_ipd, ev)
  s$FEV1[s$time == t][2] - s$FEV1[s$time == t][1]
}

sign_tab <- data.frame(
  Covariate = c("Age +10 years", "GOLD stage 3 -> 4", "Non-current smoker",
                "Female sex"),
  `Change in FEV1 (L)` = c(
    delta_ipd(list(age = 72)),
    delta_ipd(list(gold = 4)),
    delta_ipd(list(smoke = 0)),
    delta_ipd(list(sexf = 1))
  ),
  `Table S4 direction` = rep("negative on baseline", 4),
  check.names = FALSE
)
knitr::kable(sign_tab, digits = 4)

# All four are documented NEGATIVE relationships in Table S4, and all four are
# deterministic typical-value differences, so an exact sign test is valid here
# (contrast with a cohort-derived sign, which would not be).
stopifnot(all(sign_tab[[2]] < 0))
```

The combined model additionally retains a positive effect of the predicted
baseline on both drug-effect classes, and an exacerbation-history entry
criterion on baseline. Higher baseline must therefore give a *larger* drug
effect.

```{r check-baseline-on-effect}
eff_at_gold <- function(gold_low, gold_high) {
  ev <- make_events(list(
    adipd_covs(DIS_COPD_GOLD_LOW = gold_low, DIS_COPD_GOLD_HIGH = gold_high),
    adipd_covs(DIS_COPD_GOLD_LOW = gold_low, DIS_COPD_GOLD_HIGH = gold_high,
               CONMED_TIOTROPIUM_DOSE = 18)
  ), times = c(0, 24))
  s <- solve_typical(ui_adipd, ev)
  w <- s[s$time == 24, ]
  c(baseline = s$FEV1[s$time == 0][1], effect = w$FEV1[2] - w$FEV1[1])
}
milder  <- eff_at_gold(1, 3)   # mean stage 2
severer <- eff_at_gold(3, 4)   # mean stage 3.5

data.frame(
  Cohort = c("Milder (mean GOLD 2)", "More severe (mean GOLD 3.5)"),
  `Baseline FEV1 (L)` = c(milder[["baseline"]], severer[["baseline"]]),
  `Tiotropium 18 ug/day effect (L)` = c(milder[["effect"]], severer[["effect"]]),
  check.names = FALSE
) |> knitr::kable(digits = 4)

stopifnot(
  milder[["baseline"]] > severer[["baseline"]],   # higher severity, lower baseline
  milder[["effect"]]   > severer[["effect"]]      # higher baseline, larger effect
)
```

# Check 4 -- the aggregation-bias correction (paper Figure 2 and Section 3.2)

This is the paper's central methodological claim, and it is the reason the
combined model's typical baseline is 1.07 L where the aggregated-data-only model
reported 1.17 L (paper Section 4).

An individual baseline is log-normal about the study typical value. The
*arm mean* of `N` such baselines is **not** log-normal: by the central limit
theorem it is approximately normal, centred on the log-normal's **mean**
`theta * exp(omega^2 / 2)` rather than on its median `theta`, with standard
deviation shrunk by `sqrt(N)`. Fitting an aggregated arm mean as though it were
log-normal about `theta` therefore forces `theta` upward.

The check below compares the packaged model's encoded approximation against a
brute-force Monte Carlo of the actual arm mean.

```{r check-aggregation-bias}
set.seed(20260912)
cv_base <- ui_adipd$theta[["cv_base"]]
theta_b <- ui_adipd$theta[["base"]]

# What the model predicts for an aggregated arm mean at the reference arm.
model_arm_mean <- solve_typical(
  ui_adipd, make_events(list(adipd_covs(NARM = 200)), times = 0)
)$FEV1[1]

# Brute force: draw 200 individual log-normal baselines per arm, 20000 arms.
n_arm <- 200L; n_rep <- 20000L
arm_means <- colMeans(matrix(
  theta_b * exp(rnorm(n_arm * n_rep, 0, cv_base)), nrow = n_arm))

agg_tab <- data.frame(
  Quantity = c("Individual median (theta)",
               "Model's aggregated arm mean",
               "Monte-Carlo mean of the arm mean",
               "Model's arm-mean SD at N = 200",
               "Monte-Carlo SD of the arm mean"),
  Value = c(
    theta_b,
    model_arm_mean,
    mean(arm_means),
    theta_b * sqrt(exp(cv_base^2) - 1) * exp(cv_base^2 / 2) / sqrt(n_arm),
    sd(arm_means)
  )
)
knitr::kable(agg_tab, digits = 5)

rel_err_mean <- abs(mean(arm_means) - model_arm_mean) / model_arm_mean
rel_err_sd   <- abs(sd(arm_means) -
                      theta_b * sqrt(exp(cv_base^2) - 1) *
                      exp(cv_base^2 / 2) / sqrt(n_arm)) / sd(arm_means)
cat(sprintf("Relative error vs Monte Carlo: mean %.4f%%, SD %.3f%%\n",
            100 * rel_err_mean, 100 * rel_err_sd))

# 20000 replicates of a mean-of-200 gives a Monte-Carlo standard error of about
# 0.012% on the mean and about 0.5% on the SD, so these bounds sit several
# Monte-Carlo standard errors outside the noise while still failing loudly if
# the exp(omega^2/2) factor or the sqrt(N) shrinkage were dropped (either of
# which moves these by 2.8% and 100% respectively).
stopifnot(rel_err_mean < 0.005, rel_err_sd < 0.05)

# The uncorrected alternative -- treating the arm mean as log-normal about the
# MEDIAN with SD/sqrt(N) -- is biased low by exactly exp(omega^2/2).
bias <- (model_arm_mean - theta_b) / theta_b
cat(sprintf("Uncorrected log-normal arm mean would sit %.2f%% below the true mean.\n",
            100 * bias))
stopifnot(abs(bias - (exp(cv_base^2 / 2) - 1)) < 1e-10)
```

The 2.8% offset recovered here is the mechanism; the paper's own SSE found an
8-10% bias on the baseline typical value and its variance (Section 3.2,
Figure 3A) because there the estimator absorbs the discrepancy over a whole data
set rather than at a single arm.

# Check 5 -- the aggregated residual error shrinks as `1/sqrt(N)`

```{r check-residual-weight}
narm_grid <- c(18, 50, 200, 1000, 5724)   # the paper's reported arm-size range
ev_narm <- make_events(lapply(narm_grid, function(n) adipd_covs(NARM = n)),
                       times = 24)
s_narm <- solve_typical(ui_adipd, ev_narm)

resid_tab <- data.frame(
  `Arm size N` = narm_grid,
  `Residual SD (L)` = s_narm$sdFEV1,
  `addSd / sqrt(N)` = ui_adipd$theta[["addSd_FEV1"]] / sqrt(narm_grid),
  check.names = FALSE
)
knitr::kable(resid_tab, digits = 5)
stopifnot(max(abs(resid_tab[[2]] - resid_tab[[3]])) < 1e-12)
```

An individual-patient record instead carries the power error model, whose SD
rises with the prediction.

```{r check-residual-ipd}
ev_dt <- make_events(list(adipd_covs(DTYPE_AGGREGATED = 0, NARM = 1)),
                     times = c(0, 24))
s_dt <- solve_typical(ui_adipd, ev_dt)
expected_sd <- ui_adipd$theta[["powSd_FEV1"]] *
  s_dt$FEV1^ui_adipd$theta[["powExp_FEV1"]]
cat(sprintf("Individual-record residual SD at FEV1 = %.3f L: %.5f L\n",
            s_dt$FEV1[2], s_dt$sdFEV1[2]))
stopifnot(max(abs(s_dt$sdFEV1 - expected_sd)) < 1e-12)
```

# Check 6 -- the LABA / LAAC interaction is infra-additive

The three bronchodilator classes combine as
`(LABA^p + LAAC^p + MABA^p)^(1/p)` with `p = 1.35`. For `p > 1` this returns
*less* than the additive sum, so a dual bronchodilator delivers less than its two
monotherapies added together -- which is what the source estimated from the dual-
and triple-therapy arms.

```{r check-interaction}
p <- ui_adipd$theta[["int_labd"]]
ev_int <- make_events(list(
  adipd_covs(),
  adipd_covs(CONMED_VILANTEROL_DOSE = 25),                          # LABA alone
  adipd_covs(CONMED_UMECLIDINIUM_DOSE = 62.5),                      # LAAC alone
  adipd_covs(CONMED_VILANTEROL_DOSE = 25, CONMED_UMECLIDINIUM_DOSE = 62.5)
), times = c(0, 24))
s_int <- solve_typical(ui_adipd, ev_int)
w_int <- s_int[s_int$time == 24, ]
e <- w_int$FEV1 - w_int$FEV1[1]

int_tab <- data.frame(
  Arm = c("Placebo", "Vilanterol 25 (LABA)", "Umeclidinium 62.5 (LAAC)",
          "Dual LABA + LAAC"),
  `Effect vs placebo (L)` = e,
  check.names = FALSE
)
knitr::kable(int_tab, digits = 4)

additive <- e[2] + e[3]
cat(sprintf("Interaction exponent p = %.4f\n", p))
cat(sprintf("Dual effect %.4f L vs additive sum %.4f L (%.1f%% of additive)\n",
            e[4], additive, 100 * e[4] / additive))
stopifnot(
  p > 1,                       # the estimate is 1.35 (RSE 3.2%), well above 1
  e[4] > max(e[2], e[3]),      # dual still beats either monotherapy
  e[4] < additive              # but is infra-additive
)
```

# Check 7 -- effect-onset time courses reach their plateau

Corticosteroids and PDE4 inhibitors rise with a rate of
`exp(-0.951) = 0.386 / week`; the once-daily bronchodilator onsets are far
faster (9.4 and 11.5 / week), which is why only the anti-inflammatory arms show
a visible onset over a 24-week trial.

```{r onset-figure, fig.cap = "Effect onset over 24 weeks. The anti-inflammatory arms rise over several weeks; the bronchodilator arm is at plateau by the first observation."}
times <- c(0, seq(0.25, 24, by = 0.25))
ev_onset <- make_events(list(
  adipd_covs(),
  adipd_covs(CONMED_BUDESONIDE_DOSE = 320),
  adipd_covs(CONMED_ROFLUMILAST_DOSE = 500),
  adipd_covs(CONMED_TIOTROPIUM_DOSE = 18)
), times = times)
s_on <- solve_typical(ui_adipd, ev_onset)
s_on$Arm <- factor(s_on$id, labels = c("Placebo", "Budesonide 320 (ICS)",
                                       "Roflumilast 500 (PDE4)",
                                       "Tiotropium 18 (LAAC)"))
pbo <- s_on$FEV1[s_on$id == 1]
s_on$effect <- s_on$FEV1 - rep(pbo, times = 4)

ggplot(dplyr::filter(s_on, id != 1), aes(time, effect, colour = Arm)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (weeks)", y = "Effect vs placebo (L)",
       colour = NULL, title = "Drug-effect onset, Yang 2026 combined ADIPD model") +
  theme_bw() + theme(legend.position = "bottom")

# Half of the anti-inflammatory onset is reached at log(2)/0.386 = 1.80 weeks.
ics <- dplyr::filter(s_on, Arm == "Budesonide 320 (ICS)")
half_t <- approx(ics$effect / max(ics$effect), ics$time, xout = 0.5)$y
cat(sprintf("ICS onset half-time: %.3f weeks (expected %.3f)\n",
            half_t, log(2) / exp(ui_adipd$theta[["lonset_ai"]])))
stopifnot(abs(half_t - log(2) / exp(ui_adipd$theta[["lonset_ai"]])) < 0.05)

# Bronchodilator onset is essentially complete within the first week.
tio <- dplyr::filter(s_on, Arm == "Tiotropium 18 (LAAC)")
stopifnot(tio$effect[tio$time == 1] / max(tio$effect) > 0.99)
```

# Check 8 -- background therapy contributes in placebo arms

Background (non-randomized) therapy is present whether or not the arm received
a study drug, so it is deliberately *not* gated on active treatment. A placebo
arm in a trial where half the patients stay on inhaled corticosteroids therefore
sits above a placebo arm with no background therapy.

```{r check-background}
ev_bg <- make_events(list(
  adipd_covs(),
  adipd_covs(BGTHER_ICS_RUNIN_PCT = 50, BGTHER_ICS_MAINT_PCT = 50),
  adipd_covs(BGTHER_LAAC_RUNIN_PCT = 50, BGTHER_LAAC_MAINT_PCT = 50)
), times = c(0, 24))
s_bg <- solve_typical(ui_adipd, ev_bg)
w_bg <- s_bg[s_bg$time == 24, ]

data.frame(
  `Placebo arm` = c("No background therapy", "50% on background ICS",
                    "50% on background LAAC"),
  `FEV1 at week 24 (L)` = w_bg$FEV1,
  check.names = FALSE
) |> knitr::kable(digits = 4)

stopifnot(w_bg$FEV1[2] > w_bg$FEV1[1], w_bg$FEV1[3] > w_bg$FEV1[1])
```

# Check 9 -- post-bronchodilator records

A post-bronchodilator record is predicted from a baseline raised by the median
absolute reversibility of 0.18 L, and sees only `rel_postbd = 0.531` of the
long-acting bronchodilator effect.

```{r check-postbd}
ev_pbd <- make_events(list(
  adipd_covs(CONMED_TIOTROPIUM_DOSE = 18),
  adipd_covs(CONMED_TIOTROPIUM_DOSE = 18, MEAS_POSTBD = 1),
  adipd_covs(),
  adipd_covs(MEAS_POSTBD = 1)
), times = c(0, 24))
s_pbd <- solve_typical(ui_adipd, ev_pbd)
w_pbd <- s_pbd[s_pbd$time == 24, ]

eff_pre  <- w_pbd$FEV1[1] - w_pbd$FEV1[3]
eff_post <- w_pbd$FEV1[2] - w_pbd$FEV1[4]
base_shift <- s_pbd$FEV1[s_pbd$time == 0][4] - s_pbd$FEV1[s_pbd$time == 0][3]

cat(sprintf("Baseline shift for a post-SABD record: %.5f L (0.18 L x the lognormal-mean scaling %.5f)\n",
            base_shift, exp(ui_adipd$theta[["cv_base"]]^2 / 2)))
cat(sprintf("Tiotropium effect: pre-BD %.4f L, post-BD %.4f L, ratio %.4f (rel_postbd = %.4f)\n",
            eff_pre, eff_post, eff_post / eff_pre, ui_adipd$theta[["rel_postbd"]]))

# The baseline-on-effect multiplier rel_bd is built from the PRE-bronchodilator
# baseline for both record types (source control stream:
# 'RELBD = 1 + RELBDBSLslp * (B - 1.2)', with B the pre-SABD baseline), so the
# post/pre bronchodilator-effect ratio is an EXACT identity equal to
# rel_postbd, not merely a bound. The baseline shift is 0.18 L carried through
# the same log-normal-mean scaling as the baseline itself.
stopifnot(
  abs(base_shift - 0.18 * exp(ui_adipd$theta[["cv_base"]]^2 / 2)) < 1e-12,
  abs(eff_post / eff_pre - ui_adipd$theta[["rel_postbd"]]) < 1e-12
)
```

# Check 10 -- a stochastic per-study simulation (paper Figure 4)

Figure 4 of the paper simulates vilanterol 25 ug and vilanterol / fluticasone
furoate 25/200 ug over 24 weeks across a virtual population and plots the
week-24 FEV1 distribution faceted by sex and treatment, reporting that males
show higher FEV1 than females on average and that the two treatments perform
only slightly differently.

The combined model carries random effects at three levels and rxode2 draws one
level per solve, so this simulation treats **one rxode2 `id` as one study arm**
-- the level at which `eta_arm_base` and the `eta_study_*` effects are defined.
Cohort sizes are kept modest (60 arms per group) as this is an illustrative
reproduction, not a production VPC.

```{r figure4, fig.height = 5, fig.cap = "Week-24 FEV1 by sex and treatment, simulated from the combined ADIPD model. Reproduces the structure of Figure 4 of Yang 2026."}
rxode2::rxSetSeed(20260912)
n_arm_sim <- 60L

arms <- expand.grid(
  sex = c("Male", "Female"),
  trt = c("VI 25", "VI/FF 25/200"),
  rep = seq_len(n_arm_sim), stringsAsFactors = FALSE
)
ev_f4 <- make_events(lapply(seq_len(nrow(arms)), function(i) {
  adipd_covs(
    SEXF = if (arms$sex[i] == "Male") 0 else 1,
    DTYPE_AGGREGATED = 1, NARM = 200,
    CONMED_VILANTEROL_DOSE = 25,
    CONMED_FLUTICASONEFUROATE_DOSE = if (arms$trt[i] == "VI 25") 0 else 200
  )
}), times = c(0, 24))

sim_f4 <- rxode2::rxSolve(ui_adipd, ev_f4, returnType = "data.frame")
sim_f4 <- sim_f4[sim_f4$time == 24, ]
sim_f4$sex <- arms$sex[sim_f4$id]
sim_f4$trt <- arms$trt[sim_f4$id]

ggplot(sim_f4, aes(FEV1, colour = sex)) +
  geom_density(linewidth = 0.8) +
  facet_wrap(~ trt) +
  labs(x = "FEV1 at week 24 (L)", y = "Density", colour = NULL,
       title = "Simulated week-24 FEV1 by sex and treatment") +
  theme_bw() + theme(legend.position = "bottom")

f4_tab <- sim_f4 |>
  dplyr::group_by(trt, sex) |>
  dplyr::summarise(mean_FEV1 = mean(FEV1), sd_FEV1 = sd(FEV1), .groups = "drop") |>
  dplyr::rename("Treatment" = trt, "Sex" = sex,
                "Mean FEV1 (L)" = mean_FEV1, "SD across arms (L)" = sd_FEV1)
knitr::kable(f4_tab, digits = 4)

male_minus_female <- mean(sim_f4$FEV1[sim_f4$sex == "Male"]) -
  mean(sim_f4$FEV1[sim_f4$sex == "Female"])
cat(sprintf("Mean male - female difference: %.4f L\n", male_minus_female))

# The sex effect is a large, structural, typical-value difference (the
# coefficient implies a female/male baseline ratio of 0.75, i.e. roughly a
# 0.27 L gap at this baseline), so it dominates the cohort noise by a wide
# margin. The bound below is a MAGNITUDE band around that structural value,
# not a bound fitted to one draw: a mis-signed or dropped sex coefficient
# lands outside it, while resampling the cohort moves it by ~0.01 L.
stopifnot(male_minus_female > 0.15, male_minus_female < 0.40)

# Adding fluticasone furoate on top of vilanterol must help, but only slightly
# (the paper describes the two treatments as performing "slightly differently").
trt_gain <- mean(sim_f4$FEV1[sim_f4$trt == "VI/FF 25/200"]) -
  mean(sim_f4$FEV1[sim_f4$trt == "VI 25"])
cat(sprintf("Mean VI/FF - VI difference: %.4f L\n", trt_gain))
stopifnot(trt_gain > 0, trt_gain < 0.10)
```

# Assumptions and deviations

Everything below is a deliberate, documented choice; none changes a published
value.

**Data-interface re-encoding (both models).** The source data set identifies a
study arm's treatments through up to three positional slots, each a
`(drgNoN, avdostotN)` pair of an integer drug code and its average total daily
dose, with the regimen and device carried in parallel `dosfreqNoN`,
`dosfrmNSM` and `DRGOLN` slots. The packaged models instead read one named
`CONMED_<drug>_DOSE` column per compound plus named `FORM_*` selectors. This is
the same information keyed by drug name rather than by slot, it is
mathematically identical, and it matches the `CONMED_<INN>_DOSE` convention
already used by the repository's other multi-drug meta-analyses. Ten of the 23
compounds carry a constant effect with no dose-response; for those the model
keys on the dose column being non-zero, and the column value carries no
dose-proportionality -- the idiom already recorded for `CONMED_MTX_DOSE`.

**Post-bronchodilator baseline reconciliation (`FEV1_PBD_ANCHOR`).** For six of
the aggregated studies the source adds `DV * (1 - 0.884349)` to the prediction,
reconciling an absolute FEV1 that had been reconstructed from a
post-bronchodilator baseline. Using the dependent variable inside the prediction
cannot be expressed in rxode2 and is not meaningful for forward simulation, so
the value is exposed as an explicit input column that **must be zero for
simulation**. Setting it to zero reproduces the published model exactly for
every record outside those six studies, and the parameter itself
(`postbd_recon` = 0.884349, Table S3) is retained so the dependence stays
visible and auditable.

**Random-effect levels.** The source uses NONMEM `$LEVEL` with a study super-ID
plus an interoccasion-like arm random effect, needing **ten** separate N(0,1)
etas for the arm level because `$LEVEL` cannot nest an arm inside a study; only
one of the ten is ever active for a given record. The packaged model collapses
those to a single `eta_arm_base`, which is exact whenever one rxode2 `id` is one
study arm. rxode2 draws one level of random effects per solve, so a simulation
needing both a study and an arm level must solve once per level or supply the
etas explicitly; the naming (`eta_study_*` for study level, `eta_arm_base` for
arm level, unprefixed etas for subject level) records which is which.

**Convention-check warnings.** `checkModelConventions()` reports no errors on
either model and six `parameter_naming` warnings on the combined model, for etas
whose name does not resolve to a declared fixed-effect parameter. Five are the
missing-covariate imputation random effects (`eta_study_age`, `eta_study_smoke`,
`eta_study_bgics`, `eta_study_bglaba`, `eta_study_bglaac`), whose regressions
have hardcoded coefficients in the source control stream and therefore no
`ini()` parameter to pair with; the sixth is `eta_arm_base`, which names a
random-effect level (the study arm) that the checker's `eta_study_` /
`eta<param>` prefix rules do not yet recognise. The individual-patient model is
convention-clean.

**Dead parameters not carried.** Three Table S3 rows -- the imputation variances
for prior ICS, LABA and LAAC *medication history* -- feed only the
lowest-disease-severity adjustment `THETA(31)`, which was estimated to zero and
whose single consumer is commented out of the final control stream
(`LOWDSact = LOWDS`). Two further control-stream thetas fixed to zero
(`log.Placebo.T50`, `log.P.fast.plac.resp`) belong to the placebo mixture that
Section 2.3.1 removed, and `THETA(31)` itself likewise has no surviving
equation. None appears in any prediction, so none is encoded; the three
severity-covariate thetas that the source *does* keep as live-but-fixed-to-zero
structure (`e_gold_lo_base`, `e_gold_hi_base`, `e_gold_dps`) are retained so a
user can un-fix them.

**Missing-value sentinels.** The imputation branches are reached through the
source's in-band sentinels, preserved here: `-99` for a missing age, smoking
fraction or GOLD stage, and `9999` (class not used, treated as zero) or `7777`
(class used, fraction unreported) for the background-therapy columns. A complete
data set never reaches them.

**Sex coding.** The source codes sex as `SEX = 1` for male, centred at the
pooled male fraction 0.671. The canonical column is `SEXF` (1 = female), so the
identical algebra is written as `(0.329 - SEXF)` with the published coefficient
and sign unchanged. The resulting female/male baseline ratio of 0.747 agrees
with the 0.752 implied by the individual-patient model's own coefficient and
with the paper's repeated statement that female sex relates to a lower baseline.

**Erratum search.** No erratum or corrigendum for
doi:10.1002/psp4.70059 was found on the publisher's page or in the EuropePMC
record as of 2026-09-12.

**Upstream source.** The structural skeleton and the entire aggregated data set
are inherited from Llanos-Paez 2023 (J Pharmacokinet Pharmacodyn 50:297-314).
Every parameter value encoded here is re-estimated and reported in Yang 2026's
own Table S2 / Table S3, so no value is taken from the upstream paper; only the
narrative attribution of the structure is.

**A note on Table S3's fluticasone row.** Table S3 displays `THETA(38)` as
"Relative reference efficacy of fluticasone 200 ug q.d. compared to b.i.d = 0.04".
The control stream shows that theta used as
`TVDMX32QD = THETA(38)/REFDFF*(ED5032QD+REFDFF)` with `REFDFF = 100`, i.e. it is
the **fluticasone furoate** (drug 32) reference efficacy at 100 ug/day, not a
ratio involving fluticasone propionate (drug 9). The model follows the control
stream. This is also why the two molecules carry separate covariate columns.
