---
title: "Buagafuran (Yang 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Buagafuran (Yang 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Yang_2017_buagafuran_rat"))$reference`
* Description: `r rxode2::rxode(readModelDb("Yang_2017_buagafuran_rat"))$description`
* Article: <https://doi.org/10.3389/fphar.2017.00683> (open access, PMC5641330)

Buagafuran is an investigational anxiolytic agent (a synthetic analogue of a
constituent of agarwood) that had completed phase I trials in China at the time
of publication. Yang 2017 combines three pieces of work:

1. a GastroPlus whole-body PBPK model of buagafuran in the **rat**, fitted to
   plasma and brain concentrations after single 4 and 8 mg/kg oral doses;
2. a GastroPlus whole-body PBPK model in **human**, fitted to plasma
   concentrations after 30 / 60 / 120 mg single oral doses and 60 mg once daily
   for 7 days;
3. a **direct sigmoid Emax PD model** linking the rat brain concentration of
   buagafuran to the time spent on the open arms of an elevated plus-maze (EPM),
   the standard rodent read-out of unconditioned anxiety.

The two PBPK layers were then chained with the PD layer to predict a
potentially effective human dose of 30 mg t.i.d.

**Only the third component, the PD layer, is packaged in nlmixr2lib.** The
reasoning is in [Assumptions and deviations](#assumptions-and-deviations); in
short, the two PBPK layers are GastroPlus platform models whose tissue
partition coefficients, organ volumes and blood flows are software-database
outputs that the paper never prints, so they cannot be reproduced from the
publication. The PD layer is fully specified by the paper (with one parameter
recovered from the paper's own figure, documented below) and is self-contained
once the brain concentration is supplied.

```{r model}
mod <- readModelDb("Yang_2017_buagafuran_rat")
mod
```

# Population

The PD data come from male Wistar rats weighing 130-160 g, handled daily for a
week before the study and fasted for 12 h beforehand (Yang 2017, "PD in the
Rat"). The elevated plus-maze had two open arms and two closed arms, each
50 x 10 cm, extending from a 10 x 10 cm central platform raised 50 cm above the
floor; each rat explored freely for a **5 min** test period.

Two EPM arms were run. A dose-response arm (Group I) compared vehicle with 2, 4
and 8 mg/kg given orally, testing 10 min after dosing; it produced the inverted
U-shaped dose-response the paper emphasises, with the maximum effect at 4 mg/kg
and a slightly smaller effect at 8 mg/kg (Figure 4). A time-course arm (Group
II) gave a single 4 mg/kg oral dose and tested separate groups at 5, 10, 15, 30,
60, 90, 120 and 180 min (Figure 5). Each group was `n = 10`. **The PD model was
fitted to the time-course arm**, because 4 mg/kg is "the dose on the plateau of
the effectiveness dose-response curve" (Discussion).

The rats used for the plasma and brain PK (180-200 g) were a separate cohort
from the EPM rats.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
str(rxode2::rxode(readModelDb("Yang_2017_buagafuran_rat"))$population)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Yang_2017_buagafuran_rat.R`. The table
below collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `epm_openarm_time <- e0 + emax * conc^hill / (ec50^hill + conc^hill)` | n/a | Equation 4, Methods "PBPK/PD Modeling" |
| `conc <- CEFFECT` (direct link, no effect compartment) | n/a | Results "PD in the EPM Model": "No effect delays between the response and effects were found" |
| `le0` (E0) | 0.63 s | Methods, equation-4 paragraph: "E0 is the time spent by the control group in the open arms which is equal to 0.63" |
| `lec50` (EC50) | 10.6 ng/mL | Results "PD in the EPM Model": "EC50 was 10.6 ng/mL" |
| `lhill` (gamma) | 2.5 | Results "PD in the EPM Model": "Hill coefficient was 2.5"; restated in the Discussion |
| `lemax` (Emax) | 10.1 s | **Not printed anywhere in the paper.** Digitised from the fitted curve of Figure 6; see [Recovering Emax](#recovering-emax-from-figure-6) |
| `addSd` | 0 (fixed) | No residual-error model is reported for the PD fit |
| (no IIV) | n/a | No inter-individual variability is reported; the EPM data are group means of `n = 10` |

## Units

The model mixes two dimensions and neither is the usual plasma-concentration /
time pairing, so they are spelled out explicitly.

| Symbol | Quantity | Units |
|--------|----------|-------|
| `CEFFECT` / `conc` | buagafuran concentration in rat brain homogenate | ng/mL (stated in the equation-4 paragraph: "where C is in ng/mL") |
| `ec50` | half-maximal brain concentration | ng/mL |
| `hill` | Hill coefficient | unitless |
| `e0`, `emax`, `epm_openarm_time` | time spent on the open arms of a 5 min EPM trial | s (Figure 5 and Figure 6 y-axis labels) |
| `time` | time after the oral dose | min (Figure 5 x-axis) |

`conc^hill / (ec50^hill + conc^hill)` is a ratio of like powers and is therefore
unitless, so the right-hand side of equation 4 carries the units of `e0` and
`emax` (seconds) throughout. The model has no ODE state and no dose record:
`time` is carried only so that a user can line the response up with a brain
concentration-time profile, and it does not enter any equation.

# Figure 6 data

Figure 6 of Yang 2017 plots the eight observed EPM group means of the
time-course arm against the corresponding PBPK-predicted brain concentrations,
overlaid with the fitted sigmoid Emax curve. Both the markers and the curve were
digitised from a 300 dpi rendering of the published panel: the axes were
calibrated on the six y-axis ticks (0-15 s) and seven x-axis ticks (0-30 ng/mL)
located by pixel position, and the solid curve was traced column by column.

```{r figure6-data}
# Observed group means (open squares), Yang 2017 Figure 6, digitised.
observed <- tibble::tribble(
  ~CEFFECT, ~observed,
       5.63,      1.74,
       6.68,      4.36,
       9.73,      6.51,
       9.90,      1.74,
      17.59,      7.92,
      23.89,      7.24,
      26.62,     13.64,
      29.53,      9.89
)

# The published fitted curve (solid line), Yang 2017 Figure 6, digitised and
# thinned to ~1 ng/mL spacing. Columns where the trace crosses an observed
# marker were excluded before thinning.
published_curve <- tibble::tribble(
  ~CEFFECT, ~published,
      1.01,     0.649,
      2.01,     0.779,
      3.00,     1.012,
      4.00,     1.414,
      4.99,     1.920,
      5.99,     2.530,
      6.99,     3.205,
      8.01,     3.932,
      9.01,     4.619,
     10.01,     5.294,
     11.00,     5.878,
     12.00,     6.423,
     12.99,     6.916,
     13.99,     7.357,
     15.02,     7.734,
     16.01,     8.058,
     17.01,     8.343,
     18.00,     8.603,
     19.00,     8.811,
     19.99,     9.005,
     20.99,     9.161,
     21.99,     9.317,
     23.01,     9.433,
     23.98,     9.550,
     25.01,     9.654,
     26.00,     9.732,
     27.00,     9.810,
     27.99,     9.888,
     28.99,     9.939,
     29.98,     9.991
)
```

## Recovering Emax from Figure 6

The paper prints E0, EC50 and the Hill coefficient but never prints Emax. It is
recoverable because Figure 6's solid line is described in its own caption as
"the prediction for buagafuran" - it is the final model itself, not an
unadjusted side fit, so the digitised curve carries the model's own parameters.

Two fits to the digitised curve are shown below. The first holds E0, EC50 and
gamma at the printed values and estimates Emax alone. The second frees all four
parameters; because it recovers the three printed values to within about 2%, it
validates both the digitisation and the equation-4 form, and it puts a bound on
how much of the Emax estimate is digitisation error.

```{r recover-emax}
sigmoid <- function(conc, e0, emax, ec50, hill) {
  e0 + emax * conc^hill / (ec50^hill + conc^hill)
}

fit_emax_only <- nls(
  published ~ sigmoid(CEFFECT, 0.63, emax, 10.6, 2.5),
  data = published_curve, start = list(emax = 10)
)

fit_all_free <- nls(
  published ~ sigmoid(CEFFECT, e0, emax, ec50, hill),
  data = published_curve,
  start = list(e0 = 0.63, emax = 10, ec50 = 10.6, hill = 2.5)
)

recovery <- tibble::tibble(
  Parameter = c("E0 (s)", "Emax (s)", "EC50 (ng/mL)", "gamma"),
  Printed = c(0.63, NA_real_, 10.6, 2.5),
  `Emax-only fit` = c(0.63, unname(coef(fit_emax_only)["emax"]), 10.6, 2.5),
  `All-free fit` = unname(coef(fit_all_free)[c("e0", "emax", "ec50", "hill")]),
  `In model file` = c(0.63, 10.1, 10.6, 2.5)
)
knitr::kable(recovery, digits = 3,
             caption = "Recovery of the Figure 6 curve parameters by digitisation.")
```

The all-free fit reproduces the three printed values closely, so the three
digitisation-independent anchors agree with the digitisation. Both fits put Emax
near 10 s; the model file uses **10.1 s**, matching the paper's own
three-significant-figure style.

```{r recover-emax-checks}
emax_only <- unname(coef(fit_emax_only)["emax"])
all_free <- coef(fit_all_free)

stopifnot(
  # The all-free fit must recover the three PRINTED parameters. These bounds
  # are what makes the digitisation auditable: if the axis calibration were
  # wrong, or the traced line were not the published fit, these would fail.
  # Achieved: E0 0.625, EC50 10.58, gamma 2.546.
  abs(all_free[["e0"]] - 0.63) < 0.03,
  abs(all_free[["ec50"]] - 10.6) / 10.6 < 0.02,
  abs(all_free[["hill"]] - 2.5) / 2.5 < 0.05,
  # The two routes to Emax must agree, and both must agree with the value
  # committed to the model file. Achieved: 10.07 and 10.03 against 10.1.
  abs(emax_only - all_free[["emax"]]) < 0.2,
  abs(emax_only - 10.1) < 0.2,
  abs(all_free[["emax"]] - 10.1) < 0.2
)
```

# Simulation

The model has no ODE state, no dose record and no random effects: it is an
algebraic map from a brain concentration to an EPM response. A simulation is
therefore a set of observation records carrying the `CEFFECT` covariate.

Because the model declares no `eta`, do **not** pass `omega = NA` or apply
`rxode2::zeroRe()` - both are meaningless here and `omega = NA` errors on a
model with no IIV.

```{r simulate}
conc_grid <- data.frame(
  id = 1L,
  time = seq_along(seq(0, 35, by = 0.25)),
  CEFFECT = seq(0, 35, by = 0.25),
  evid = 0L,
  amt = 0
)

sim <- rxode2::rxSolve(mod, events = conc_grid) |>
  as.data.frame()

head(sim[, c("CEFFECT", "epm_openarm_time")])
```

# Replicate Figure 6

```{r figure-6, fig.width = 6, fig.height = 4.2}
# Replicates Figure 6 of Yang 2017: EPM response versus rat brain concentration.
ggplot() +
  geom_line(data = sim, aes(CEFFECT, epm_openarm_time), linewidth = 0.9) +
  geom_point(data = published_curve, aes(CEFFECT, published),
             shape = 3, size = 1.6, colour = "steelblue") +
  geom_point(data = observed, aes(CEFFECT, observed),
             shape = 0, size = 2.4) +
  coord_cartesian(xlim = c(0, 30), ylim = c(0, 15)) +
  labs(
    x = "Concentration of buagafuran in brain (ng/mL)",
    y = "Time spent on the open arms (s)",
    title = "Figure 6 - PD response versus brain concentration",
    caption = paste(
      "Replicates Figure 6 of Yang 2017. Solid line: nlmixr2lib model.",
      "Blue crosses: published fitted curve, digitised.",
      "Open squares: published observed group means, digitised."
    )
  )
```

# Validation

NCA is not an applicable validation for this model - there is no dose, no
compartment and no concentration-time profile to integrate - so the checks
below follow the pattern used for endogenous and mechanistic models:
definitional identities, boundary behaviour, agreement with the published fit,
and a dimensional check (the units table above).

Every quantity here is **deterministic**: the model carries no random effects,
so the same numbers come out on every machine and thread count, and the
assertions can be tight. Where a bound is loose it is because the *reference*
side is digitised, not because the model output varies.

## Definitional identities

Three properties follow from equation 4 by algebra alone and must hold to
solver precision. They are the cheapest way to catch a transcription error in
`e0`, `emax` or `ec50` - a wrong `ec50`, for instance, moves the half-maximal
point and breaks the second identity immediately.

```{r definitional}
anchor <- rxode2::rxSolve(
  mod,
  events = data.frame(
    id = 1L, time = 1:3, evid = 0L, amt = 0,
    CEFFECT = c(0, 10.6, 1e6)
  )
) |>
  as.data.frame()

e0_hat <- anchor$epm_openarm_time[1]
half_hat <- anchor$epm_openarm_time[2]
plateau_hat <- anchor$epm_openarm_time[3]

identities <- tibble::tibble(
  Property = c(
    "C = 0 gives E0",
    "C = EC50 gives E0 + Emax/2",
    "C -> infinity gives E0 + Emax"
  ),
  Expected = c(0.63, 0.63 + 10.1 / 2, 0.63 + 10.1),
  Model = c(e0_hat, half_hat, plateau_hat)
)
knitr::kable(identities, digits = 6,
             caption = "Definitional identities of the sigmoid Emax model.")

stopifnot(
  isTRUE(all.equal(e0_hat, 0.63, tolerance = 1e-8)),
  isTRUE(all.equal(half_hat, 0.63 + 10.1 / 2, tolerance = 1e-8)),
  isTRUE(all.equal(plateau_hat, 0.63 + 10.1, tolerance = 1e-6))
)
```

## Shape

The response must be strictly increasing in concentration, and with a Hill
coefficient above 1 the curve must be sigmoid, i.e. have an inflection at a
concentration below EC50. For the sigmoid Emax function the inflection sits at
`EC50 * ((gamma - 1) / (gamma + 1))^(1 / gamma)`, which is 6.99 ng/mL for
`EC50 = 10.6` and `gamma = 2.5`. The paper's own description - "a relatively
steeper sigmoid C-E curve was obtained with the Hill coefficient of 2.5" -
is what this check pins down.

```{r shape}
resp <- sim$epm_openarm_time
stopifnot(all(diff(resp) > 0))

inflection_theory <- 10.6 * ((2.5 - 1) / (2.5 + 1))^(1 / 2.5)
slope <- diff(resp) / diff(sim$CEFFECT)
inflection_model <- sim$CEFFECT[which.max(slope)]

c(theoretical = inflection_theory, from_model = inflection_model)

# The grid is 0.25 ng/mL, so the numerical maximum-slope point can only be
# located to within one grid step of the analytical value.
stopifnot(abs(inflection_model - inflection_theory) <= 0.25)
```

## Agreement with the published fitted curve

The strongest single check available: the packaged model is evaluated at the
concentrations of the digitised published curve and compared point by point.
The deviation here is dominated by digitisation error in the reference, not by
the model.

```{r agree-curve}
curve_cmp <- published_curve |>
  mutate(
    model = sigmoid(CEFFECT, 0.63, 10.1, 10.6, 2.5),
    diff_s = model - published
  )

summary_stats <- c(
  max_abs_diff_s = max(abs(curve_cmp$diff_s)),
  rms_diff_s = sqrt(mean(curve_cmp$diff_s^2))
)
summary_stats

# The digitised curve is read to roughly +/- 0.03 s (one pixel is about
# 0.026 s on the 300 dpi rendering), and the traced line has finite width, so
# a deviation of a few hundredths of a second is the floor. Achieved:
# max 0.09 s, RMS 0.04 s.
stopifnot(
  summary_stats[["max_abs_diff_s"]] < 0.20,
  summary_stats[["rms_diff_s"]] < 0.10
)
```

## Agreement with the published observed data

The observed EPM group means are what the curve was fitted to. The model is not
expected to pass through them - `n = 10` behavioural group means over a 5 min
trial are noisy, and Figure 6 shows the scatter plainly - but the model must sit
in the middle of them rather than systematically above or below.

```{r agree-observed}
obs_cmp <- observed |>
  mutate(
    model = sigmoid(CEFFECT, 0.63, 10.1, 10.6, 2.5),
    residual_s = observed - model
  )

knitr::kable(
  obs_cmp |>
    rename(
      "Brain conc (ng/mL)" = CEFFECT,
      "Observed (s)" = observed,
      "Model (s)" = model,
      "Residual (s)" = residual_s
    ),
  digits = 2,
  caption = "Observed EPM group means (Figure 6) against the packaged model."
)

# The fitted curve must be unbiased with respect to the data it was fitted to.
# The mean residual is a property of the published fit, not of the simulation,
# so this is a deterministic check on the transcription. Achieved: -0.08 s
# mean residual against an observed spread of 1.7 to 13.6 s.
stopifnot(abs(mean(obs_cmp$residual_s)) < 1)
```

## Concentration coverage

The Discussion notes that "the rat data provided appropriate coverage of the
EC50 region". That is a checkable claim about the data set: the observed brain
concentrations must bracket EC50 on both sides.

```{r coverage}
stopifnot(
  min(observed$CEFFECT) < 10.6,
  max(observed$CEFFECT) > 10.6,
  sum(observed$CEFFECT < 10.6) >= 2,
  sum(observed$CEFFECT > 10.6) >= 2
)
range(observed$CEFFECT)
```

# The dose-prediction claim

Yang 2017's headline result is that 30 mg t.i.d. is a potentially effective
human dose because it "appeared to reach EC50 for most of the day at
steady-state" (Figure 7A), while 60 mg b.i.d. gives a similar PD response
despite roughly doubling the brain concentration (Figure 7B).

The **PD half** of that claim is reproducible here - the flatness of the
response above EC50 is what the model encodes - but the **PK half** is not,
because the human brain concentration-time profile behind Figure 7 is a
GastroPlus PBPK output. What the packaged model does supply is the response at
any brain concentration a user cares to put in, and the table below shows where
the headroom runs out.

Note that the saturation does **not** begin at EC50. A sigmoid Emax curve is
symmetric in log concentration about EC50, so the doubling from EC50/2 to EC50
and the doubling from EC50 to 2 x EC50 add *exactly* the same response - the
assertions below check that identity rather than assuming a flattening that is
not there. The flattening sets in above 2 x EC50, by which point at most 15% of
the maximum response remains available however high the concentration goes.

```{r dose-claim}
tibble::tibble(
  `Brain conc (ng/mL)` = c(5.3, 10.6, 21.2, 30, 42.4),
  Note = c(
    "EC50 / 2",
    "EC50",
    "2 x EC50",
    "top of the observed range",
    "4 x EC50"
  )
) |>
  mutate(
    `Response (s)` = sigmoid(`Brain conc (ng/mL)`, 0.63, 10.1, 10.6, 2.5),
    `% of maximum` = 100 * (`Response (s)` - 0.63) / 10.1
  ) |>
  knitr::kable(digits = 2,
               caption = "Predicted EPM response across the concentration range.")

gain <- function(lo, hi) {
  sigmoid(hi, 0.63, 10.1, 10.6, 2.5) - sigmoid(lo, 0.63, 10.1, 10.6, 2.5)
}

# 1. Log-symmetry about EC50: the doubling below and the doubling above EC50
#    are identical. This is an algebraic property of the sigmoid Emax form and
#    holds to solver precision, so it is a genuine transcription check on
#    ec50 and hill.
stopifnot(isTRUE(all.equal(gain(5.3, 10.6), gain(10.6, 21.2), tolerance = 1e-8)))

# 2. Saturation above 2 x EC50: the next doubling adds well under half of what
#    the previous one did. Achieved ratio 0.343.
stopifnot(gain(21.2, 42.4) < 0.5 * gain(10.6, 21.2))

# 3. At 2 x EC50 the response is already past 80% of maximum, so no further
#    increase in exposure can add more than 20% of Emax. This is the
#    quantitative form of the paper's "PD response was not significantly
#    increased with elevated brain concentration" argument.
frac_at_2ec50 <- (sigmoid(21.2, 0.63, 10.1, 10.6, 2.5) - 0.63) / 10.1
stopifnot(frac_at_2ec50 > 0.80)
```

# Assumptions and deviations {#assumptions-and-deviations}

## The two PBPK layers are deliberately not packaged

Yang 2017 builds its rat and human PK with GastroPlus 8.5, using the ACAT
absorption model and the Poulin and Theil homogeneous method for tissue
partition coefficients. The paper prints the drug-specific inputs (Table 1:
MW, logP, solubility, particle radius, Caco-2 Papp, blood/plasma ratio, fu),
the optimised ASF coefficient C1 (0.6805 rat, 0.8005 human), the refined
hepatic clearances (36.8 mL/min/kg rat, 11.7 mL/min/kg human) and the optimised
brain partition coefficient (Kp = 6.7).

It does not print a single **volume** term anywhere in its eleven pages: no
central or steady-state volume, no organ volumes, no organ blood flows, and no
tissue partition coefficient other than brain ("Default values in GastroPlus
software were applied for Kp of other organs"). Those are software-database
outputs. Reconstructing the PBPK layers would therefore mean sourcing the
load-bearing parameters from a commercial software installation rather than
from the publication, which is not auditable against the paper and is outside
what nlmixr2lib packages. The clearance and brain Kp that *are* printed are
recorded in the model file's `covariateData$CEFFECT$notes` so the provenance of
the driving concentration is not lost.

A consequence worth stating plainly: this model cannot be dosed. It consumes a
brain concentration and returns a response. Users wanting the full
dose-to-response chain must supply their own rat or human brain PK.

## Emax is digitised, not printed

`Emax = 10.1 s` did not come from the paper's text or tables - it is not there.
It was recovered by digitising the fitted curve of Figure 6, as documented in
[Recovering Emax](#recovering-emax-from-figure-6) and asserted on above. The
supporting evidence that the digitisation is sound is that the same trace
independently reproduces the three parameters the paper *does* print (E0 0.625
vs 0.63, EC50 10.58 vs 10.6, gamma 2.546 vs 2.5). The printed values take
precedence and are what the model file carries; only Emax is figure-derived.
Its uncertainty is roughly +/- 0.05 s, i.e. under 1%.

The eight observed markers of Figure 6 used in the validation section are
likewise digitised. They are used only to check the fit, never to set a
parameter.

## No IIV and no residual error

The paper reports no OMEGA, no SIGMA, and no standard error or RSE on any of
the four PD parameters - the EPM data are group means of `n = 10` and GastroPlus
PDPlus was used for the fit. No random effects are therefore declared, and the
additive residual SD is `fixed(0)` rather than invented. A user refitting this
model to individual EPM data should free `addSd` and add an `eta` where the
data support one.

## Other assumptions

- `n_subjects = 90` in the `population` metadata assumes independent groups of
  10 for the vehicle control and each of the eight post-dose times of the
  time-course arm. The paper states `n = 10` per group but does not say whether
  animals were re-used; the single-trial design of the EPM implies they were
  not.
- The observed markers of Figure 6 are not labelled with their sampling times in
  the paper, so no time assignment is asserted here. They are used purely as
  (concentration, response) pairs, which is how Figure 6 presents them and how
  the model consumes them.
- The dose-response arm of the EPM (Figure 4), which shows the inverted U-shaped
  dose-response with a smaller effect at 8 mg/kg than at 4 mg/kg, is **not**
  captured by this model. A monotone sigmoid Emax cannot produce a downturn.
  That is the authors' own choice: they fitted the 4 mg/kg time course only, and
  the concentration range it spans (5.6-29.5 ng/mL) does not reach the
  concentrations the 8 mg/kg arm would have produced. Extrapolating this model
  above about 30 ng/mL is therefore extrapolating past both the data and the
  paper's own pharmacology.
- The units of `e0` and `emax` are seconds, read from the y-axis labels of
  Figures 5 and 6; the equation-4 paragraph gives the value 0.63 without a unit.
  The 5 min trial length makes seconds the only reading consistent with the
  figures.
