---
title: "Unecritinib and crizotinib (Yang 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unecritinib and crizotinib (Yang 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Yang_2021_unecritinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Yang_2021_unecritinib"))$description`
* Article: <https://doi.org/10.3389/fphar.2021.782518> (open access)

Unecritinib (development code TQ-B3101) is an oral ALK/ROS1/MET kinase inhibitor
that is rapidly converted by amide hydrolysis to its active metabolite
(TQ-B3101M). Yang 2021 fitted the two analytes sequentially in NONMEM 7.4
(FOCE-I): a one-compartment model with first-order absorption for the parent was
fitted first and fixed, then a two-compartment model for the metabolite was
added. All of the parent's clearance forms the metabolite (the fraction
metabolised, FM, was fixed to 1 for identifiability), and the metabolite's
apparent clearance falls exponentially with time since the first dose, by 41%
at its asymptote (Table 3, footnote a). No covariate was retained.

## Identity of the analytes

The paper refers to the two analytes only as TQ-B3101 and TQ-B3101M and states
that "the information of TQ-B3101 and TQ-B3101M including structure, molecular
formular and molecular weight cannot be provided due to a proprietary issue"
(Methods, Analytical Methods). The identities are now public:

* PubChem CID 71506874, **unecritinib**, lists TQ-B3101 as a synonym:
  C23H24Cl2FN5O2, 492.4 g/mol, the N-acetyl amide of crizotinib.
* Hydrolysis of that amide gives **crizotinib** (PubChem CID 11626560,
  C21H22Cl2FN5O, 450.3 g/mol). Later bioanalytical work names crizotinib as
  the active metabolite of unecritinib
  ([doi:10.1016/j.jpba.2024.116199](https://doi.org/10.1016/j.jpba.2024.116199)).
* The paper's own numbers agree. The Discussion equates 300 mg TQ-B3101 with
  275 mg crizotinib, and its Eq. 4 converts crizotinib doses to TQ-B3101 doses
  by the molecular-weight ratio of the two. 275 / 300 = 0.917, and
  450.3 / 492.4 = 0.914.

The data were fitted as molar concentrations: the LLOQ was 1 nmol/L for both
analytes and the goodness-of-fit plots are in molar units. With FM fixed to 1,
one mole of crizotinib therefore forms per mole of unecritinib cleared. The
packaged model applies the molecular-weight ratio to the formation flux, so the
metabolite states hold mg of crizotinib and `Cc_crizotinib` is in ng/mL of
crizotinib. The paper's paediatric Table 6 and adult exposures are in those
units, and the comparisons below are direct.

# Population

The analysis pooled 40 patients from two Chinese studies (Table 1): a Phase 1
dose-escalation study in adults with advanced solid tumours (NCT03019276; single
doses of 100 or 200 mg, then 100, 200 or 300 mg once daily for 28 days) and a
Phase 2 single-arm study in adults and adolescents with relapsed or refractory
ALK-positive anaplastic large cell lymphoma (NCT04306887; 200, 250, 300 or 350
mg twice daily for 28 days). All doses were taken fasting. Table 2: 34 adults
(median age 51.5 years, range 28-73; median weight 59.0 kg, range 42.0-87.7) and
6 adolescents (median age 13.0 years, range 11-14; median weight 41.8 kg, range
32.9-68.0); 21 of 40 (52.5%) were male. The dataset held 375 quantifiable
unecritinib and 658 quantifiable crizotinib concentrations; the 340 and 42
below-quantitation samples were discarded.

The same information is available programmatically via
`readModelDb("Yang_2021_unecritinib")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Yang_2021_unecritinib.R`.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | log(2850) L/h | Table 3 |
| `lvc` (V/F) | log(4200) L | Table 3 |
| `lka` (Ka) | log(51.9) 1/h | Table 3 |
| `fm` (FM) | fixed(1) | Methods, Pharmacokinetic Modeling; Figure 1 caption |
| `lcl_crizotinib` (CLm0/Fm) | log(126) L/h | Table 3 |
| `lvc_crizotinib` (Vcm/Fm) | log(2300) L | Table 3 |
| `lq_crizotinib` (Qm/Fm) | log(113) L/h | Table 3 |
| `lvp_crizotinib` (Vpm/Fm) | log(1480) L | Table 3 |
| `cl_exp_famp` (-TDPK) | -0.41 | Table 3 (TDPK = 0.41) |
| `lcl_exp_kdes` (KTDPK) | log(0.0363) 1/h | Table 3 |
| `etalcl`, `etalvc` | 0.0760, 0.1010 | Table 3 (28.1, 32.6 %CV); omega^2 = log(1 + CV^2) |
| `etalcl_crizotinib`, `etalvc_crizotinib`, `etalvp_crizotinib` | 0.1100, 0.2484, 0.5300 | Table 3 (34.1, 53.1, 83.6 %CV) |
| `propSd`, `propSd_crizotinib` | 0.711, 0.319 | Table 3 (sigma 71.1, 31.9 %CV) |
| `mw_unecritinib`, `mw_crizotinib` | 492.4, 450.3 g/mol | PubChem CIDs 71506874 and 11626560 (not in the paper) |
| `theta_i = theta_TV * exp(eta_i)` | n/a | Methods, Eq. 1 |
| `CLm(t) = CLm0 * [1 - TDPK * (1 - exp(-KTDPK * t))]` | n/a | Table 3, footnote a |
| Parent -> metabolite -> two-compartment ODEs | n/a | Figure 1 |

# Structural checks against the published half-lives

The Abstract and Results quote an elimination half-life of 1.0 h for unecritinib
and, for the metabolite at steady state, distribution and elimination
half-lives of 4.9 and 39.4 h. These follow in closed form from the typical
values, using the asymptotic metabolite clearance `CLm0 * (1 + cl_exp_famp)`.

```{r half-lives}
mod <- readModelDb("Yang_2021_unecritinib")
th <- rxode2::rxode2(mod)$theta

cl_p <- exp(th[["lcl"]])
vc_p <- exp(th[["lvc"]])
cl_m_inf <- exp(th[["lcl_crizotinib"]]) * (1 + th[["cl_exp_famp"]])
vc_m <- exp(th[["lvc_crizotinib"]])
q_m <- exp(th[["lq_crizotinib"]])
vp_m <- exp(th[["lvp_crizotinib"]])

k10 <- cl_m_inf / vc_m
k12 <- q_m / vc_m
k21 <- q_m / vp_m
sum_k <- k10 + k12 + k21
root <- sqrt(sum_k^2 - 4 * k10 * k21)

hl <- tibble::tibble(
  Quantity = c(
    "Unecritinib elimination half-life (h)",
    "Crizotinib distribution half-life at steady state (h)",
    "Crizotinib elimination half-life at steady state (h)",
    "Crizotinib apparent clearance at steady state (L/h)"
  ),
  Model = c(
    log(2) * vc_p / cl_p,
    log(2) / ((sum_k + root) / 2),
    log(2) / ((sum_k - root) / 2),
    cl_m_inf
  ),
  Published = c(1.0, 4.9, 39.4, 126 * (1 - 0.41))
) |>
  mutate(`Difference (%)` = 100 * (Model / Published - 1))

knitr::kable(hl, digits = 2, caption = "Closed-form half-lives versus Yang 2021 (Abstract; Results).")

# Deterministic identities of the typical values; the published half-lives are
# printed to one decimal, so each must agree to within that rounding (<= 3%).
stopifnot(all(abs(hl$`Difference (%)`) < 3))
```

# Time course of the metabolite clearance

```{r clearance-time}
tc <- tibble::tibble(time = seq(0, 240, by = 1)) |>
  mutate(cl_crizotinib = exp(th[["lcl_crizotinib"]]) *
    (1 + th[["cl_exp_famp"]] * (1 - exp(-exp(th[["lcl_exp_kdes"]]) * time))))

ggplot(tc, aes(time / 24, cl_crizotinib)) +
  geom_line() +
  geom_hline(yintercept = cl_m_inf, linetype = "dashed") +
  labs(
    x = "Time since first dose (days)", y = "CLm/Fm (L/h)",
    title = "Typical crizotinib apparent clearance after unecritinib dosing",
    caption = "Table 3 footnote a of Yang 2021; dashed line = asymptote 74.3 L/h."
  )
```

The half-time of the decline is `log(2) / 0.0363` = 19.1 h, so the clearance is
within 1% of its asymptote after about 5.3 days. The authors attribute the
decline to auto-inhibition of CYP3A4, as reported for crizotinib given as
crizotinib, whose mean CL/F falls from 100 L/h after a single dose to 60 L/h at
steady state (Discussion).

# Typical-value simulation and a steady-state identity

```{r typical}
mod_typ <- rxode2::zeroRe(mod)
mw_ratio <- 450.3 / 492.4

ev_typ <- rxode2::et(amt = 300, cmt = "depot", ii = 12, addl = 55) |>
  rxode2::et(sort(unique(c(seq(0, 48, by = 0.1), seq(648, 672, by = 0.05)))), cmt = "central") |>
  as.data.frame() |>
  # Two residual-error endpoints: observation rows carry dvid = 1; a forward
  # solve still returns both Cc and Cc_crizotinib on every row.
  mutate(dvid = ifelse(evid == 0, 1L, NA_integer_))
sim_typ <- rxode2::rxSolve(mod_typ, ev_typ, rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()

# At true steady state AUC0-tau of the metabolite equals
# fm * dose * MW ratio / CLm(inf) (the clearance has relaxed to within 1e-10 of
# its asymptote by day 28).
last <- sim_typ |> filter(time >= 660, time <= 672)
auc_num <- sum(diff(last$time) *
  (head(last$Cc_crizotinib, -1) + tail(last$Cc_crizotinib, -1)) / 2)
auc_closed <- 1000 * 300 * mw_ratio / cl_m_inf
c(numeric = auc_num, closed_form = auc_closed)
# Solve against its own closed form: a pure numerical-error comparison.
# Trapezoidal error on the 0.05 h grid is ~1e-5 relative.
stopifnot(abs(auc_num / auc_closed - 1) < 1e-3)

sim_typ |>
  filter(time <= 48 | time >= 648) |>
  mutate(window = ifelse(time <= 48, "First two days", "Day 28 (steady state)")) |>
  select(time, window, Unecritinib = Cc, Crizotinib = Cc_crizotinib) |>
  pivot_longer(-c(time, window), names_to = "analyte", values_to = "conc") |>
  filter(conc > 0) |>
  ggplot(aes(time, conc, colour = analyte)) +
  geom_line() +
  facet_wrap(~window, scales = "free_x") +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Concentration (ng/mL)", colour = NULL,
    title = "Typical profiles, 300 mg unecritinib twice daily"
  ) +
  theme(legend.position = "bottom")
```

The parent peaks within an hour and is essentially gone before the next dose
(half-life 1 h), while crizotinib peaks 1-3 h after the dose and accumulates,
consistent with the times to peak of 0.5-1 h and 1-3 h quoted in the
Discussion.

# Virtual adult cohorts

Original observed data are not available. Each regimen below is a separate arm
of 100 virtual patients who start at their first dose (the model's time-varying
clearance is indexed to time since the first dose). The model has no
covariates, so no demographic distribution is needed.

```{r cohort}
rxode2::rxSetSeed(20211)

make_arm <- function(n, regimen, amt, ii, n_doses, obs, id_offset) {
  ids <- id_offset + seq_len(n)
  dose_times <- (seq_len(n_doses) - 1) * ii
  dose <- tidyr::expand_grid(id = ids, time = dose_times) |>
    mutate(evid = 1L, amt = amt, cmt = "depot")
  obs_rows <- tidyr::expand_grid(id = ids, time = obs) |>
    mutate(evid = 0L, amt = 0, cmt = "central", dvid = 1L)
  bind_rows(dose, obs_rows) |>
    mutate(regimen = regimen) |>
    arrange(id, time, desc(evid))
}

obs_sd <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 24, 48, 72, 96, 144, 192, 240, 336)
obs_qd <- 648 + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 12, 16, 24)
obs_bid <- 660 + c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 10, 12)

events <- bind_rows(
  make_arm(100, "100 mg single dose", 100, 24, 1, obs_sd, 0L),
  make_arm(100, "200 mg single dose", 200, 24, 1, obs_sd, 100L),
  make_arm(100, "100 mg QD, day 28", 100, 24, 28, obs_qd, 200L),
  make_arm(100, "200 mg QD, day 28", 200, 24, 28, obs_qd, 300L),
  make_arm(100, "300 mg BID, day 28", 300, 12, 56, obs_bid, 400L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))

sim <- rxode2::rxSolve(mod, events = events, keep = "regimen") |>
  as.data.frame()
```

# PKNCA: accumulation of crizotinib

The Methods report that in the Phase 1 study "the average AUC0-tau,ss of
TQ-B3101M after repeated dosing was 20-40% higher than the average AUC0-inf of
TQ-B3101M after a single dose" at 100 and 200 mg. With linear kinetics and a
constant clearance those two AUCs would be equal; the excess is produced by the
time-dependent clearance, so it is the most direct check of that part of the
model.

```{r pknca}
nca_input <- function(sim, analyte) {
  d <- sim |>
    filter(!is.na(.data[[analyte]])) |>
    transmute(id, time, regimen, Cc = pmax(.data[[analyte]], 0))
  # Time-zero rows for the single-dose arms (pre-dose concentration 0).
  bind_rows(
    d,
    d |> filter(grepl("single", regimen)) |> distinct(id, regimen) |>
      mutate(time = 0, Cc = 0)
  ) |>
    distinct(id, regimen, time, .keep_all = TRUE) |>
    arrange(id, regimen, time)
}

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, regimen)

intervals <- data.frame(
  regimen = c(
    "100 mg single dose", "200 mg single dose", "100 mg QD, day 28",
    "200 mg QD, day 28", "300 mg BID, day 28"
  ),
  start = c(0, 0, 648, 648, 660),
  end = c(Inf, Inf, 672, 672, 672),
  cmax = TRUE,
  tmax = TRUE,
  aucinf.obs = c(TRUE, TRUE, FALSE, FALSE, FALSE),
  auclast = c(FALSE, FALSE, TRUE, TRUE, TRUE)
)

run_nca <- function(analyte) {
  conc <- PKNCA::PKNCAconc(nca_input(sim, analyte), Cc ~ time | regimen + id)
  dose <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc, dose, intervals = intervals))
}

nca_m <- run_nca("Cc_crizotinib")
nca_p <- run_nca("Cc")

auc_by_subject <- as.data.frame(nca_m$result) |>
  filter(PPTESTCD %in% c("aucinf.obs", "auclast")) |>
  select(regimen, id, PPORRES)
```

```{r pknca-accumulation}
acc <- auc_by_subject |>
  mutate(
    dose = ifelse(grepl("^100", regimen), "100 mg", "200 mg"),
    type = ifelse(grepl("single", regimen), "single", "repeated")
  ) |>
  filter(!grepl("BID", regimen)) |>
  group_by(dose, type) |>
  summarise(mean_auc = mean(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = type, values_from = mean_auc) |>
  mutate(excess = 100 * (repeated / single - 1))

acc |>
  rename(
    "Unecritinib dose" = dose,
    "Mean single-dose crizotinib AUC0-inf (ng*h/mL)" = single,
    "Mean day-28 crizotinib AUC0-24 (ng*h/mL)" = repeated,
    "Excess (%)" = excess
  ) |>
  knitr::kable(digits = 1, caption = "Repeated-dose versus single-dose crizotinib exposure (published: 20-40% higher).")

# Each per-dose excess compares two independent 100-subject means and carries
# ~7 percentage points of cohort noise, so only the pooled excess is gated
# here, with a bound a constant clearance (0%) or a mis-signed amplitude
# (negative) fails by several standard errors. The tight check is the
# deterministic typical-value excess below.
stopifnot(mean(acc$excess) > 0, mean(acc$excess) < 60)

# Typical-value excess: day-28 AUC0-24 = dose * MW ratio / CLm(inf) against
# the single-dose AUC0-inf of the typical subject (dense grid, numeric solve).
ev_sd <- rxode2::et(amt = 100, cmt = "depot") |>
  rxode2::et(c(seq(0, 24, by = 0.05), seq(24.5, 2000, by = 0.5)), cmt = "central") |>
  as.data.frame() |>
  mutate(dvid = ifelse(evid == 0, 1L, NA_integer_))
sd_typ <- rxode2::rxSolve(mod_typ, ev_sd, rtol = 1e-10, atol = 1e-12) |>
  as.data.frame()
auc_sd_typ <- sum(diff(sd_typ$time) *
  (head(sd_typ$Cc_crizotinib, -1) + tail(sd_typ$Cc_crizotinib, -1)) / 2)
typ_excess <- 100 * ((1000 * 100 * mw_ratio / cl_m_inf) / auc_sd_typ - 1)
typ_excess
# Deterministic: must fall in the 20-40% the Methods report.
stopifnot(typ_excess > 20, typ_excess < 40)
```

The simulated excess of repeated-dose over single-dose crizotinib exposure is
`r paste(round(acc$excess), collapse = "% and ")`% at 100 and 200 mg, against
the 20-40% the paper reports from the Phase 1 non-compartmental analysis. The
typical-value excess, free of cohort noise, is `r round(typ_excess)`%, inside
the published range; the per-dose cohort values scatter around it by the
sampling noise of two independent 100-subject arms.

## Adult steady-state exposure

The Discussion quotes a crizotinib (TQ-B3101M) AUC0-12h,ss of 4,251 ng*h/mL
after 300 mg unecritinib twice daily, against 3,880 ng*h/mL after 250 mg
crizotinib twice daily. The paper does not say whether 4,251 ng*h/mL is an
observed mean from the Phase 2 patients or a model prediction, nor on which day.

```{r adult-ss}
published_adult <- tibble::tibble(regimen = "300 mg BID, day 28", auclast = 4251)

adult_res <- nca_m
adult_res$result <- adult_res$result |> filter(regimen == "300 mg BID, day 28")

cmp_adult <- nlmixr2lib::ncaComparisonTable(
  simulated = adult_res,
  reference = published_adult,
  by = "regimen",
  params = "auclast",
  units = c(auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_adult, caption = "Crizotinib AUC0-12h,ss after 300 mg unecritinib BID: simulated median versus the Discussion value. * differs by >20%.")

adult_sim_auc <- auc_by_subject$PPORRES[auc_by_subject$regimen == "300 mg BID, day 28"]
c(median = median(adult_sim_auc), mean = mean(adult_sim_auc))
```

The simulated median is `r round(100 * (1 - median(adult_sim_auc) / 4251))`%
and the simulated mean `r round(100 * (1 - mean(adult_sim_auc) / 4251))`% below
the quoted value; the typical-value steady-state AUC, from the closed form
above, is `r round(auc_closed)` ng*h/mL. The paper gives neither the source nor
the sample size behind 4,251 ng*h/mL, so the difference is recorded here rather
than investigated further; it is not used to adjust the model.

## Unecritinib (parent) NCA

```{r parent-nca}
as.data.frame(nca_p$result) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "aucinf.obs", "auclast")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  knitr::kable(digits = 2, caption = "Median simulated unecritinib NCA parameters by regimen (the paper reports no parent NCA values to compare).")
```

# Replicates Table 6 and Figure 4: paediatric BSA-tiered dosing

The paper's paediatric simulation is not part of the fitted model. It applied
body-weight allometry with fixed exponents of 0.75 on every clearance and 1 on
every volume, "scaled to a median weight of adult patients" (Methods, Clinical
Simulations). The median adult weight in Table 2 is 59.0 kg. The reference
weight is not otherwise stated; 59.0 kg is used here. The allometric lines are
added to a copy of the model inside this article only.

```{r ped-model}
mod_ped <- mod |>
  rxode2::model({
    cl <- exp(lcl + etalcl) * (WT / 59)^0.75
    vc <- exp(lvc + etalvc) * (WT / 59)
    cl_crizotinib0 <- exp(lcl_crizotinib + etalcl_crizotinib) * (WT / 59)^0.75
    vc_crizotinib <- exp(lvc_crizotinib + etalvc_crizotinib) * (WT / 59)
    q_crizotinib <- exp(lq_crizotinib) * (WT / 59)^0.75
    vp_crizotinib <- exp(lvp_crizotinib + etalvp_crizotinib) * (WT / 59)
  })
```

The paper's 1,000 virtual children aged 6 to under 18 years came from the PK-Sim
Asian population database (Table 4: median weight 42.3 kg, range 15.8-102.3;
median BSA 1.3 m^2, range 0.7-2.2). That database is not reproduced here.
Instead, BSA was drawn from a normal distribution with mean 1.3 m^2 and SD 0.3
m^2 truncated to 0.74-2.2 m^2 (the Table 4 median, and a spread whose +/-3 SD
spans the Table 4 range), the first 100 children falling in each BSA tier of
Table 5 were kept, and body weight was obtained by inverting the weight-only
paediatric BSA formula of Costeff (1966), `BSA = (4 WT + 7) / (WT + 90)`. That
formula maps Table 4's median weight of 42.3 kg to 1.33 m^2 and its weight
range to 0.66-2.16 m^2, close to the tabulated BSA summary.

```{r ped-cohort}
tiers <- tibble::tibble(
  tier = c("0.74-0.89", "0.90-1.22", "1.23-1.38", "1.39-1.59", ">1.6"),
  dose = c(250, 350, 400, 450, 550)
)

obs_ped <- 13 * 24 + c(0, 0.5, 1, 2, 4, 6, 10, 12) # Methods sampling grid, day 14

set.seed(20212)
bsa_pool <- rnorm(20000, mean = 1.3, sd = 0.3)
bsa_pool <- bsa_pool[bsa_pool >= 0.74 & bsa_pool <= 2.2]
ped_subjects <- tibble::tibble(
  BSA = bsa_pool,
  # Table 5 tiers are printed with gaps (0.89 | 0.90, 1.59 | 1.6); split them
  # at the midpoints.
  tier = as.character(cut(BSA, c(0.74, 0.895, 1.225, 1.385, 1.595, 2.2),
    labels = tiers$tier, include.lowest = TRUE
  ))
) |>
  group_by(tier) |>
  slice_head(n = 100) |>
  ungroup() |>
  left_join(tiers, by = "tier") |>
  mutate(
    id = seq_len(n()),
    WT = (90 * BSA - 7) / (4 - BSA)
  )
stopifnot(all(table(ped_subjects$tier) == 100L))

ped_events <- bind_rows(
  tidyr::expand_grid(ped_subjects, time = seq(0, 13 * 24, by = 12)) |>
    mutate(evid = 1L, amt = dose, cmt = "depot"),
  tidyr::expand_grid(ped_subjects, time = obs_ped) |>
    mutate(evid = 0L, amt = 0, cmt = "central", dvid = 1L)
) |>
  arrange(id, time, desc(evid))
stopifnot(!anyDuplicated(unique(ped_events[, c("id", "time", "evid")])))

rxode2::rxSetSeed(20213)
ped_sim <- rxode2::rxSolve(mod_ped, events = ped_events, keep = c("tier", "BSA")) |>
  as.data.frame()
```

```{r ped-nca}
ped_conc <- ped_sim |>
  filter(!is.na(Cc_crizotinib)) |>
  transmute(id, tier, time, Cc = pmax(Cc_crizotinib, 0))
ped_dose <- ped_events |>
  filter(evid == 1) |>
  select(id, tier, time, amt)
ped_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(ped_conc, Cc ~ time | tier + id),
  PKNCA::PKNCAdose(ped_dose, amt ~ time | tier + id),
  intervals = data.frame(start = 13 * 24, end = 13 * 24 + 12, cmax = TRUE, auclast = TRUE)
))

# Table 6 medians (the comparison table pools the simulation by median).
table6_median <- tibble::tibble(
  tier = tiers$tier,
  cmax = c(631.1, 620.8, 586.0, 597.7, 589.6),
  auclast = c(6583.9, 6507.4, 6259.7, 6376.8, 6379.0)
)

cmp_ped <- nlmixr2lib::ncaComparisonTable(
  simulated = ped_nca,
  reference = table6_median,
  by = "tier",
  params = c("cmax", "auclast"),
  units = c(cmax = "ng/mL", auclast = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp_ped, caption = "Replicates Table 6 of Yang 2021: median crizotinib Cmax,ss and AUC0-12h,ss by BSA tier after BSA-tiered unecritinib dosing. * differs by >20%.")

ped_ind <- as.data.frame(ped_nca$result) |>
  filter(PPTESTCD %in% c("cmax", "auclast")) |>
  select(id, tier, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

ped_check <- ped_ind |>
  group_by(tier) |>
  summarise(sim_auc = median(auclast), sim_cmax = median(cmax), .groups = "drop") |>
  left_join(rename(table6_median, pub_auc = auclast, pub_cmax = cmax), by = "tier") |>
  mutate(
    pct_auc = 100 * (sim_auc / pub_auc - 1),
    pct_cmax = 100 * (sim_cmax / pub_cmax - 1)
  )
stopifnot(nrow(ped_check) == 5L, !anyNA(ped_check$pub_auc))
ped_check

# Structural: the BSA-tiered doses were chosen to give flat exposure, and the
# published tier medians span only 5%. Without the weight scaling the tiers
# would span 2.2-fold (250 to 550 mg). The simulated span is wider than the
# published one (realised 1.26 here, and 1.33 with a uniform-in-tier BSA draw)
# because weight is derived from BSA by the Costeff formula rather than drawn
# with height from PK-Sim, which shifts the within-tier weights; each tier
# median also carries ~4% sampling error. 1.5 sits outside that range and
# still fails a model without the allometric scaling.
stopifnot(max(ped_check$sim_auc) / min(ped_check$sim_auc) < 1.5)
# Level: a mis-transcribed clearance, a dropped molecular-weight ratio (-9%
# is inside the noise, but a mg/ng or molar slip is orders of magnitude) or a
# wrong reference weight moves every tier together.
stopifnot(abs(median(ped_check$pct_auc)) < 12, abs(median(ped_check$pct_cmax)) < 15)
```

```{r figure-4}
ped_ind |>
  mutate(tier = factor(tier, levels = tiers$tier)) |>
  pivot_longer(c(cmax, auclast), names_to = "metric", values_to = "value") |>
  mutate(metric = ifelse(metric == "cmax", "Cmax,ss (ng/mL)", "AUC0-12h,ss (ng*h/mL)")) |>
  ggplot(aes(tier, value)) +
  geom_boxplot(outlier.size = 0.5) +
  geom_hline(
    data = tibble::tibble(
      metric = c("Cmax,ss (ng/mL)", "AUC0-12h,ss (ng*h/mL)"),
      ref = c(621, 6530)
    ),
    aes(yintercept = ref), linetype = "dashed"
  ) +
  facet_wrap(~metric, scales = "free_y") +
  labs(
    x = "BSA tier (m^2)", y = "Crizotinib",
    title = "Paediatric crizotinib exposure after BSA-tiered unecritinib dosing",
    caption = paste(
      "Replicates Figure 4 of Yang 2021. Dashed lines: crizotinib target",
      "exposure after the paediatric crizotinib label dose (Table 6)."
    )
  )
```

The simulated tier medians agree with the Table 6 medians to within
`r round(max(abs(c(ped_check$pct_auc, ped_check$pct_cmax))))`%, with a median
difference across tiers of `r round(median(ped_check$pct_auc), 1)`% for AUC,
and sit near the crizotinib target exposure the tiered regimen was designed to
match (geometric-mean Cmax,ss 621 ng/mL, AUC0-12h,ss 6,530 ng*h/mL). The
allometric bridge, the tiered doses and the molecular-weight conversion
therefore reproduce the paper's dosing simulation. The tier-to-tier pattern is
less flat than in Table 6. The virtual children here get their weights from
BSA through the Costeff formula, not from the PK-Sim population the paper
used, and that moves the within-tier weights, most visibly in the wide 0.90-1.22
m^2 tier.

# Assumptions and deviations

- **Analyte identity and units.** The paper withholds the structures and
  molecular weights. The maintainers identified TQ-B3101 as unecritinib and
  TQ-B3101M as crizotinib from PubChem and later literature, and took the
  molecular weights (492.4 and 450.3 g/mol) from PubChem. The model applies
  their ratio to the formation flux because the source fitted molar data with
  a molar fraction metabolised of 1. That choice is supported by the paper's own
  300 mg ~ 275 mg dose equivalence and by the Table 6 replication above. It is
  not a fitted value.
- **Time-dependent clearance sign convention.** The paper's TDPK = 0.41 is a
  positive "maximum fraction reduction". The model encodes it as the signed
  fractional amplitude `cl_exp_famp = -0.41` in
  `CLm(t) = CLm0 * (1 + cl_exp_famp * (1 - exp(-KTDPK * t)))`, which is
  algebraically identical and follows the sign convention of `kel_exp_famp`
  (negative = decline). Time `t` is time since the first dose, so simulations
  must start at the first dose. A washout between a single-dose and a
  repeated-dose period, as in the Phase 1 design, is not modelled as a reset of
  the clearance.
- **IIV conversion.** Table 3 reports IIV as %CV; variances were computed as
  `omega^2 = log(1 + CV^2)`. No correlations are reported, so the etas are
  independent. Ka and Qm/Fm carry no IIV.
- **BQL handling.** The final model used the M1 method (below-quantitation
  samples discarded) for both analytes. The paper also evaluated M3 for the
  parent but reports only one set of estimates.
- **Paediatric simulation.** The allometric exponents (0.75 / 1), the
  reference weight (59.0 kg, the adult median of Table 2) and the BSA-tiered
  doses come from the paper's simulation, not from the fitted model, which
  retained no covariate. The PK-Sim population the paper used is not
  available. The virtual children were therefore drawn from a truncated normal
  BSA distribution matching Table 4, and their weights derived from the Costeff
  (1966) BSA formula. Steady state was taken on day 14, after the clearance has
  settled, because the paper does not state the day.
- **Adult cohort.** The accumulation check uses separate virtual arms for the
  single-dose and repeated-dose regimens, each starting at its own first dose.
- No erratum or correction notice for Yang 2021 was found in Crossref or
  Europe PMC as of 2026-09-30.
