---
title: "SGLT2 inhibitors: PK, urinary glucose excretion, FPG and HbA1c (Yao 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SGLT2 inhibitors: PK, urinary glucose excretion, FPG and HbA1c (Yao 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
rxode2::rxSetSeed(20230412)
```

# Model and source

Yao 2023 is a model-based meta-analysis (MBMA) of the three globally marketed
sodium-glucose co-transporter-2 (SGLT2) inhibitors available when the literature
search closed. It builds a chain that starts at dose and ends at a disease
endpoint:

```
dose -> PK (per drug) -> AUC(0-24 h) at steady state
     -> dUGEc (shared Emax, drug-specific EC50)
     -> FPG (linear in dUGEc)
     -> HbA1c (indirect-response turnover driven by FPG)
```

The paper fits this chain sequentially, because the PK, PD and endpoint data
come from different studies and different subjects. The extraction mirrors that
structure: three independent population PK models, one per drug, plus one
class-level PK/PD/endpoint model that the three feed through their
steady-state AUC.

```{r load-models}
pk_models <- c(
  dapagliflozin = "Yao_2023_dapagliflozin_mbma",
  canagliflozin = "Yao_2023_canagliflozin_mbma",
  empagliflozin = "Yao_2023_empagliflozin_mbma"
)

# readModelDb() returns the model FUNCTION; rxode2::rxode() resolves it to a
# ui object whose metadata fields can be read directly.
pk_ui <- lapply(pk_models, function(nm) rxode2::rxode(readModelDb(nm)))
ep_ui <- rxode2::rxode(readModelDb("Yao_2023_sglt2_endpoints_mbma"))
```

* Citation: `r ep_ui$reference`
* Article: <https://doi.org/10.1002/psp4.12934>
* Supplement (Appendix S1, Tables S1-S2, Figures S1-S8):
  <https://doi.org/10.1002/psp4.12934> (Supporting Information, file
  `psp4-12-487-s001.docx`)
* Models contributed by this paper:

```{r model-list, echo = FALSE}
knitr::kable(
  data.frame(
    Model = c(unname(pk_models), "Yao_2023_sglt2_endpoints_mbma"),
    Role = c(
      "Dapagliflozin population PK (2 transit compartments; food effect on Kt)",
      "Canagliflozin population PK (4 transit compartments)",
      "Empagliflozin population PK (2 transit compartments)",
      "Class-level AUC -> dUGEc -> FPG -> HbA1c endpoint model"
    )
  ),
  caption = "The four model files extracted from Yao 2023."
)
```

# Population

The PK layer pools 880 summary-level plasma concentrations from published
dapagliflozin, canagliflozin and empagliflozin studies: 201 dapagliflozin
subjects (177 healthy), 158 canagliflozin subjects (76 healthy) and 204
empagliflozin subjects (122 healthy). Mean age ranges from 38.1 to 46.8 years
and mean body weight from 73.3 to 76.4 kg across the three PK datasets
(Yao 2023 Table S2).

The endpoint layer pools 848 FPG and 1219 HbA1c study-arm means from 80 trials
in adults with type 2 diabetes: 8324 dapagliflozin, 7004 canagliflozin and 9489
empagliflozin patients (Yao 2023 Table S1). Those cohorts are older and heavier
than the PK cohorts (mean age 55.8-57.6 years, mean weight 79.0-84.9 kg) with a
mean baseline FPG near 160 mg/dL and a mean baseline HbA1c near 7.9%
(Table S2). Trials in patients with moderate or severe renal impairment or
hepatic insufficiency were excluded, as were trials using insulin, so the model
applies to patients with a glomerular filtration rate above
60 mL/min/1.73 m^2 -- a limitation the authors state explicitly.

Because this is a meta-analysis, **the unit of observation is a published
study-arm mean, not an individual measurement**. Every variance term in every
one of the four models is therefore a BETWEEN-STUDY variance, encoded as
`eta_study_*`. These models simulate study-arm mean trajectories; they are not
individual-subject models and must not be used as such.

```{r population, echo = FALSE}
pop <- ep_ui$population
knitr::kable(
  data.frame(
    Field = c("Species", "Patients", "Studies", "Disease state", "Dose range"),
    Value = c(
      pop$species,
      format(pop$n_subjects, big.mark = ","),
      pop$n_studies,
      pop$disease_state,
      pop$dose_range
    )
  ),
  caption = "Endpoint-model population metadata."
)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them.

```{r source-trace, echo = FALSE}
trace <- tibble::tribble(
  ~Quantity, ~Value, ~Source,
  "PK ODEs (2 transit + 2 compartment)", "structure", "Appendix S1 Eq. S1-S5",
  "PK ODEs, canagliflozin (4 transit + 2 compartment)", "structure", "Appendix S1 Eq. S6-S12",
  "Inter-study variability model", "P = theta * exp(eta)", "Appendix S1 Eq. S13",
  "Residual error model", "Y = F + F*eps1*W + eps2*W", "Appendix S1 Eq. S14",
  "Dapagliflozin CL/F, Vc/F, CLD/F, VT/F, Kt", "19.5, 82.0, 10.3, 122, 6.50", "Table 1",
  "Dapagliflozin food effect on Kt", "0.254", "Table 1 (Fed)",
  "Canagliflozin CL/F, Vc/F, CLD/F, VT/F, Kt", "12.0, 85.5, 9.77, 108, 6.38", "Table 1",
  "Empagliflozin CL/F, Vc/F, CLD/F, VT/F, Kt", "4.25, 30.6, 1.37, 28.3, 4.13", "Table 1",
  "dUGEc definition", "(UGE - UGE_base) / FPG_base", "Equation 1",
  "dUGEc Emax model", "Emax * AUC / (EC50 + AUC)", "Equation 2",
  "Emax (shared across the three drugs)", "0.606 g/(mg/dL)", "Table 2",
  "EC50 dapa / cana / empa", "56.6 / 2310 / 841 ng*h/mL", "Table 2",
  "FPG placebo model", "FPGbase + Pfmax*(1-exp(-Kfp*t)) + DISfp*t", "Equation 3",
  "FPG drug model", "FPG_placebo + SLOPEfd * dUGEc", "Equation 4 (see Errata)",
  "FPG baseline; Pfmax1-4", "160; 1.45, 1.90, -1.37, 4.30 mg/dL", "Table 2",
  "Kfp; DISfp; SLOPEfd", "0.340 /week; 3.13 mg/dL/100wk; -43.3", "Table 2",
  "HbA1c placebo model", "HbA1cbase - Phmax*(1-exp(-Khp*t)) + DIShp*t", "Equation 5 (see Errata)",
  "Kin definition", "Kout * HbA1cbase - Kin2", "Equation 6",
  "HbA1c drug ODE", "FPG/FPGbase*Kin + Kin2 - Kout*HbA1c", "Equation 7",
  "HbA1c total", "HbA1c_placebo + HbA1c_drug", "Equation 8",
  "HbA1c baseline; Phmax1-4", "7.92; -0.200, 0.0510, -0.230, -0.06 %", "Table 2",
  "Khp; DIShp; Kout; Kin2", "0.240 /week; 0.310 %/100wk; 0.200 /week; 0.500 %/week", "Table 2",
  "Residual variances (PK, PK/PD, FPG, HbA1c)", "0.457/0.462; 0.222/0.0646; 0.0330; 0.006", "Tables 1-2"
)
knitr::kable(trace, caption = "Source trace for every model equation and parameter.")
```

# Part 1 - Pharmacokinetics

## Typical-value profiles at the label doses

```{r pk-helpers}
# Typical-value solve: zeroRe() removes both the between-study eta terms and
# the residual error, leaving the population prediction.
solve_typical <- function(model_name, events, keep = character()) {
  mod <- rxode2::zeroRe(readModelDb(model_name))
  out <- rxode2::rxSolve(mod, events, keep = keep, returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out
}

# Observation rows are placed on the "central" ODE STATE, never on the "Cc"
# observable: referencing an observable as a compartment renumbers the
# compartment slots. rxode2 returns Cc as a column at these rows anyway.
pk_events <- function(dose_mg, n_days, obs_times, fed = 0) {
  dosing <- data.frame(
    id = 1L, time = seq(0, by = 24, length.out = n_days),
    amt = dose_mg, evid = 1L, cmt = "depot", FED = fed
  )
  obs <- data.frame(
    id = 1L, time = obs_times, amt = NA_real_, evid = 0L,
    cmt = "central", FED = fed
  )
  dplyr::arrange(dplyr::bind_rows(dosing, obs), time, dplyr::desc(evid))
}

label_dose <- c(dapagliflozin = 10, canagliflozin = 300, empagliflozin = 25)
```

```{r pk-profiles}
profile_times <- sort(unique(c(seq(0, 48, by = 0.1))))

pk_profiles <- dplyr::bind_rows(lapply(names(pk_models), function(drug) {
  ev <- pk_events(label_dose[[drug]], n_days = 2, obs_times = profile_times)
  # rxSolve() returns one row per OBSERVATION record; dosing records produce no
  # output row, so no evid filter is needed (and evid is not a returned column).
  solve_typical(pk_models[[drug]], ev) |>
    dplyr::mutate(drug = drug)
}))

ggplot(pk_profiles, aes(time, Cc, colour = drug)) +
  geom_line(linewidth = 0.8) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Plasma concentration (ng/mL)", colour = NULL,
    title = "Typical-value profiles at the label dose, first two daily doses",
    subtitle = "Dapagliflozin 10 mg, canagliflozin 300 mg, empagliflozin 25 mg"
  ) +
  theme_bw()
```

The three profiles reproduce the qualitative features Yao 2023 reports in
Figure S3 and the Discussion: a fast transit-limited absorption phase, a peak
inside the first two hours, and a terminal phase slow enough to give modest
accumulation on once-daily dosing.

## The dapagliflozin food effect

Food is the only covariate retained anywhere in the PK layer. Yao 2023 Table 1
reports the coefficient as a bare multiplier of 0.254 on the transit rate
constant, and the Discussion states that food "could significantly influence
the Kt parameter (p < 0.005) in dapagliflozin, which is consistent with a
longer time to maximum concentration of dapagliflozin under fed condition."

```{r pk-food}
food_times <- seq(0, 24, by = 0.05)
food <- dplyr::bind_rows(lapply(c(fasted = 0, fed = 1), function(f) {
  ev <- pk_events(10, n_days = 1, obs_times = food_times, fed = f)
  solve_typical("Yao_2023_dapagliflozin_mbma", ev, keep = "FED")
})) |>
  dplyr::mutate(state = ifelse(FED == 1, "fed", "fasted"))

food_summary <- food |>
  dplyr::group_by(state) |>
  dplyr::summarise(
    Tmax_h = time[which.max(Cc)],
    Cmax_ng_mL = max(Cc),
    .groups = "drop"
  )
knitr::kable(
  food_summary,
  digits = 2,
  caption = "Dapagliflozin 10 mg single dose: the fed state delays and blunts the peak."
)

# Deterministic (typical-value) quantities, so an exact bound is appropriate.
stopifnot(
  # Food must DELAY the peak, which is the direction the paper reports.
  food_summary$Tmax_h[food_summary$state == "fed"] >
    food_summary$Tmax_h[food_summary$state == "fasted"],
  # Food must LOWER the peak (slower input into an unchanged clearance).
  food_summary$Cmax_ng_mL[food_summary$state == "fed"] <
    food_summary$Cmax_ng_mL[food_summary$state == "fasted"]
)
```

## Steady-state AUC(0-24 h): an exact internal identity, and the published values

For a linear model at steady state the AUC over one dosing interval is exactly
`daily dose / CL`. The endpoint model consumes precisely that quantity, so the
identity is worth checking with a real NCA rather than assuming it.

Yao 2023 Results reports "the steady-state AUC in 24 h after long-term
treatment was 51.39 ng/mL*h for dapagliflozin, 83.6 ng/mL*h for canagliflozin,
and 235.3 ng/mL*h for empagliflozin." Those values are far too small to be
whole-dose exposures -- dapagliflozin 10 mg gives roughly 500-600 ng*h/mL -- but
they match `1000 / CL` for each drug to within a few tenths of a percent, i.e.
they are **dose-normalised**: AUC(0-24 h) at steady state per mg of daily dose.
The printed unit omits the `/mg`. Simulating a 1 mg once-daily regimen to
steady state and running PKNCA over the final dosing interval therefore tests
the transcription of every PK parameter that determines exposure.

```{r pk-nca}
tau <- 24
n_days_ss <- 21
start_ss <- (n_days_ss - 1) * tau
end_ss <- start_ss + tau

# Coarse sampling through the accumulation phase, dense sampling over the
# final interval so the trapezoidal AUC resolves the peak.
nca_times <- sort(unique(c(
  seq(0, start_ss, by = 4),
  seq(start_ss, end_ss, by = 0.05)
)))

nca_input <- dplyr::bind_rows(lapply(names(pk_models), function(drug) {
  ev <- pk_events(1, n_days = n_days_ss, obs_times = nca_times)
  solve_typical(pk_models[[drug]], ev) |>
    dplyr::mutate(treatment = drug, id = 1L)
}))

# The PKNCA input filter is !is.na(Cc) ONLY: filtering on time or on Cc > 0
# would drop the interval-start record and trigger the "AUC range starting
# before the first measurement" warning.
conc_df <- nca_input |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

dose_df <- dplyr::bind_rows(lapply(names(pk_models), function(drug) {
  data.frame(
    id = 1L, time = seq(0, by = tau, length.out = n_days_ss),
    amt = 1, treatment = drug
  )
}))

conc_obj <- PKNCA::PKNCAconc(conc_df, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start = start_ss, end = end_ss,
  cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pk-nca-compare}
published_auc <- tibble::tribble(
  ~treatment,      ~auclast,
  "dapagliflozin",   51.39,
  "canagliflozin",   83.60,
  "empagliflozin",  235.30
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = published_auc,
  by = "treatment",
  units = c(auclast = "ng*h/mL per mg/day"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "Simulated steady-state AUC(0-24 h) at 1 mg once daily versus the",
    "dose-normalised values reported in Yao 2023 Results.",
    "* differs from reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
```

```{r pk-nca-identity}
auc_sim <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(treatment, auclast = PPORRES)

check_auc <- auc_sim |>
  dplyr::left_join(published_auc, by = "treatment",
                   suffix = c("_sim", "_pub")) |>
  dplyr::mutate(
    cl_L_per_h = vapply(
      treatment,
      function(d) exp(pk_ui[[d]]$theta[["lcl"]]),
      numeric(1)
    ),
    auc_identity = 1000 / cl_L_per_h,
    pct_vs_identity = 100 * (auclast_sim / auc_identity - 1),
    pct_vs_published = 100 * (auclast_sim / auclast_pub - 1)
  )

knitr::kable(
  check_auc |>
    dplyr::rename(
      "Drug" = treatment,
      "AUC simulated" = auclast_sim,
      "AUC published" = auclast_pub,
      "CL/F (L/h)" = cl_L_per_h,
      "1000 / CL" = auc_identity,
      "% vs identity" = pct_vs_identity,
      "% vs published" = pct_vs_published
    ),
  digits = 3,
  caption = "The PKNCA AUC reproduces both the closed-form identity and the published dose-normalised value."
)

# Both sides are deterministic typical-value quantities (zeroRe(), fixed time
# grid), so exact bounds are appropriate here: no cohort is drawn, and the only
# error is trapezoidal discretisation of a smooth curve.
stopifnot(
  # NCA versus the closed form dose/CL: pure numerical integration error.
  max(abs(check_auc$pct_vs_identity)) < 0.5,
  # NCA versus the paper's own printed numbers: confirms CL was transcribed
  # correctly for all three drugs AND that the printed values are per mg/day.
  max(abs(check_auc$pct_vs_published)) < 1
)
```

Both bounds hold: the numerical AUC agrees with `1000 / CL` to well under a
percent, and matches all three published values to under 1%. That settles the
unit reading and validates the clearance transcription for every drug.

# Part 2 - Exposure to the translational biomarker dUGEc

Yao 2023 Equation 2 gives the biomarker as a shared-Emax, drug-specific-EC50
function of steady-state exposure. Because dUGEc has no dynamics, it can be
read straight off the endpoint model by solving it once per exposure level.

```{r ugec-helpers}
ep_events <- function(cov_rows, times) {
  cov_rows$.row <- seq_len(nrow(cov_rows))
  tidyr::crossing(cov_rows, time = times) |>
    dplyr::mutate(
      id = .row,
      amt = NA_real_,
      evid = 0L,
      # cmt names the ODE STATE (never an observable). The endpoint model has
      # three observation endpoints (UGEc, FPG, HbA1c), so rxode2 also needs a
      # dvid to resolve the dvid->cmt mapping; dvid = 1 selects UGEc, and every
      # observable is returned as its own column regardless.
      cmt = "hba1c_drug",
      dvid = 1L
    ) |>
    dplyr::select(-.row) |>
    dplyr::arrange(id, time) |>
    as.data.frame()
}

strata_cols <- function(stratum) {
  data.frame(
    TRT_T2DM_NAIVE = as.integer(stratum == "naive"),
    TRT_T2DM_NONNAIVE = as.integer(stratum == "non-naive"),
    TRT_T2DM_ADDON = as.integer(stratum == "add-on"),
    TRT_T2DM_MIXED = as.integer(stratum == "mixed")
  )
}

# AUC covariate row for one arm: only the treated drug's column is non-zero.
auc_cols <- function(drug, auc) {
  data.frame(
    AUC_DAPA = ifelse(drug == "dapagliflozin", auc, 0),
    AUC_CANA = ifelse(drug == "canagliflozin", auc, 0),
    AUC_EMPA = ifelse(drug == "empagliflozin", auc, 0)
  )
}

solve_endpoints <- function(cov_rows, times, keep) {
  mod <- rxode2::zeroRe(readModelDb("Yao_2023_sglt2_endpoints_mbma"))
  ev <- ep_events(cov_rows, times)
  out <- rxode2::rxSolve(mod, ev, keep = keep, returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out
}
```

```{r ugec-curve}
auc_grid <- 10^seq(1, 5, length.out = 60)

ugec_rows <- dplyr::bind_rows(lapply(names(pk_models), function(drug) {
  dplyr::bind_cols(
    data.frame(drug = drug, auc = auc_grid),
    auc_cols(drug, auc_grid),
    strata_cols("naive")[rep(1, length(auc_grid)), , drop = FALSE]
  )
}))

ugec_curve <- solve_endpoints(ugec_rows, times = 0, keep = c("drug", "auc"))

ggplot(ugec_curve, aes(auc, UGEc, colour = drug)) +
  geom_line(linewidth = 0.8) +
  scale_x_log10() +
  labs(
    x = "AUC(0-24 h) at steady state (ng*h/mL)",
    y = expression(Delta * "UGEc  (g/(mg/dL))"),
    colour = NULL,
    title = "Replicates Figure 2 of Yao 2023",
    subtitle = "Shared Emax of 0.606; EC50 56.6 / 2310 / 841 ng*h/mL"
  ) +
  theme_bw()
```

```{r ugec-checks}
ec50_pub <- c(dapagliflozin = 56.6, canagliflozin = 2310, empagliflozin = 841)
emax_pub <- 0.606

ugec_at <- function(drug, auc) {
  row <- dplyr::bind_cols(
    auc_cols(drug, auc),
    strata_cols("naive")[rep(1, length(auc)), , drop = FALSE]
  )
  solve_endpoints(row, times = 0, keep = character())$UGEc
}

# At AUC = EC50 the Emax expression must give exactly half of Emax, for every
# drug. This is a structural identity, so an exact bound is right.
half_max <- vapply(names(ec50_pub), function(d) ugec_at(d, ec50_pub[[d]]), numeric(1))
stopifnot(max(abs(half_max / (emax_pub / 2) - 1)) < 1e-6)

# The asymptote must approach Emax and never exceed it.
asymptote <- vapply(names(ec50_pub), function(d) ugec_at(d, 1e9), numeric(1))
stopifnot(
  max(abs(asymptote / emax_pub - 1)) < 1e-4,
  max(ugec_curve$UGEc) <= emax_pub
)

# A placebo arm (every AUC column 0) must give exactly zero biomarker response.
stopifnot(abs(ugec_at("none", 0)) < 1e-12)

knitr::kable(
  data.frame(
    Drug = names(ec50_pub),
    "EC50 (ng*h/mL)" = unname(ec50_pub),
    "dUGEc at EC50" = unname(half_max),
    "Emax/2" = emax_pub / 2,
    check.names = FALSE
  ),
  digits = 4,
  caption = "The Emax structure is recovered exactly at each drug's published EC50."
)
```

## Where the label doses sit on the exposure-response curve

```{r ugec-label-doses}
auc_label <- vapply(
  names(pk_models),
  function(d) 1000 * label_dose[[d]] / exp(pk_ui[[d]]$theta[["lcl"]]),
  numeric(1)
)

label_tab <- data.frame(
  Drug = names(auc_label),
  "Daily dose (mg)" = unname(label_dose[names(auc_label)]),
  "AUC(0-24 h),ss (ng*h/mL)" = unname(auc_label),
  "dUGEc (g/(mg/dL))" = vapply(names(auc_label),
                               function(d) ugec_at(d, auc_label[[d]]), numeric(1)),
  check.names = FALSE
)
label_tab$"Fraction of Emax" <- label_tab$"dUGEc (g/(mg/dL))" / emax_pub

knitr::kable(
  label_tab, digits = c(0, 0, 0, 4, 3),
  caption = "At their label doses the three drugs sit at a similar point on the shared Emax curve."
)

# The paper's central claim is that the three drugs reach a SIMILAR maximum
# response and differ only in potency. At label doses they should therefore
# land close together on the response axis even though their EC50 values span
# a 40-fold range. Deterministic, so a tight bound is appropriate.
stopifnot(
  diff(range(label_tab$"Fraction of Emax")) < 0.06,
  all(label_tab$"Fraction of Emax" > 0.5)
)
```

# Part 3 - FPG and HbA1c endpoints

## Placebo trajectories by treatment-history stratum

Yao 2023 estimates a separate maximal placebo effect on each endpoint in each
of four treatment-history strata. Figure 3 panels (a), (b), (e) and (f) show
the corresponding visual predictive checks.

```{r endpoint-placebo}
weeks <- seq(0, 104, by = 1)
strata <- c("naive", "non-naive", "add-on", "mixed")

placebo_rows <- dplyr::bind_rows(lapply(strata, function(s) {
  dplyr::bind_cols(
    data.frame(stratum = s),
    auc_cols("none", 0),
    strata_cols(s)
  )
}))

placebo <- solve_endpoints(placebo_rows, weeks, keep = "stratum")

placebo_long <- placebo |>
  dplyr::select(stratum, time, FPG, HbA1c) |>
  tidyr::pivot_longer(c(FPG, HbA1c), names_to = "endpoint", values_to = "value")

ggplot(placebo_long, aes(time, value, colour = stratum)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~endpoint, scales = "free_y") +
  labs(
    x = "Time after first dose (weeks)", y = NULL, colour = "Treatment history",
    title = "Placebo trajectories, replicating Figure 3 panels (a), (b), (e) and (f)",
    subtitle = "FPG in mg/dL; HbA1c in %"
  ) +
  theme_bw()
```

```{r endpoint-placebo-checks}
placebo_summary <- placebo |>
  dplyr::group_by(stratum) |>
  dplyr::summarise(
    fpg_0 = FPG[time == 0],
    fpg_min = min(FPG),
    fpg_104 = FPG[time == 104],
    hba1c_0 = HbA1c[time == 0],
    hba1c_min = min(HbA1c),
    hba1c_104 = HbA1c[time == 104],
    .groups = "drop"
  )
knitr::kable(placebo_summary, digits = 3,
             caption = "Placebo-arm endpoint summaries by treatment-history stratum.")

# Every arm must start exactly at the published baselines (Table 2).
stopifnot(
  max(abs(placebo_summary$fpg_0 - 160)) < 1e-8,
  max(abs(placebo_summary$hba1c_0 - 7.92)) < 1e-8
)

# The add-on stratum is the only one in which the placebo FPG effect is a
# DECREASE (Pfmax3 = -1.37 mg/dL); the paper attributes it to the confounding
# effect of the concomitant hypoglycaemic agents.
addon <- placebo_summary[placebo_summary$stratum == "add-on", ]
naive <- placebo_summary[placebo_summary$stratum == "naive", ]
stopifnot(addon$fpg_min < 160, addon$fpg_min < naive$fpg_min)

# HbA1c under placebo must DIP below baseline before disease progression
# carries it back above. The dip is the discriminating observation for the
# Equation 5 sign (see Assumptions and deviations): with the printed minus sign
# and Table 2's signed Phmax the naive arm would rise monotonically and never
# dip at all. Assert the existence and direction of the dip, plus a generous
# magnitude ceiling -- not a precise depth, which depends on how strongly the
# Equation 7 turnover state damps it.
stopifnot(
  naive$hba1c_min < 7.92,
  (7.92 - naive$hba1c_min) < 0.3,
  naive$hba1c_104 > 7.92,
  # The non-naive stratum is the one with a POSITIVE Phmax (+0.051%), so it is
  # the only stratum with no dip at all.
  placebo_summary$hba1c_min[placebo_summary$stratum == "non-naive"] >= 7.92 - 1e-8
)

# Disease progression on HbA1c: the paper states "The disease progression was
# estimated to be a 0.16% increase in HbA1c per year".
dishp_per_year <- ep_ui$theta[["dishp"]] / 100 * 52
stopifnot(abs(dishp_per_year - 0.16) < 0.01)
```

The HbA1c placebo shape is the load-bearing check here. Equation 5 as printed
subtracts `Phmax`, and Table 2 reports Phmax as a signed value, so a literal
reading double-negates and makes the naive placebo arm *rise* monotonically.
The model uses `+ Phmax`, which is what the paper's own Discussion says and
what Figure 3(e) shows, and the assertion above pins the resulting dip.

The typical-value dip is shallower than the placebo term alone would give
(0.058% at week 3 in the naive stratum, against 0.15% at week 12 for
`Phmax * (1 - exp(-Khp*t)) + DIShp*t` on its own). That is a consequence of
Equation 7: the HbA1c turnover state sees the *total* HbA1c in its elimination
term, so it partly compensates the placebo excursion. Figure 3(e) reads to a
dip of roughly 0.1-0.15%, between the two; it pools the naive and add-on strata
and plots the median of a stochastic simulation with substantial between-study
variability on the disease-progression slope, so it does not resolve the
difference. See Assumptions and deviations.

## Active-treatment trajectories

```{r endpoint-active}
active_rows <- dplyr::bind_rows(lapply(strata[c(1, 3)], function(s) {
  dplyr::bind_rows(
    dplyr::bind_cols(data.frame(stratum = s, arm = "placebo"),
                     auc_cols("none", 0), strata_cols(s)),
    dplyr::bind_rows(lapply(names(pk_models), function(d) {
      dplyr::bind_cols(
        data.frame(stratum = s, arm = d),
        auc_cols(d, auc_label[[d]]),
        strata_cols(s)
      )
    }))
  )
}))

active <- solve_endpoints(active_rows, weeks, keep = c("stratum", "arm"))

active_long <- active |>
  dplyr::select(stratum, arm, time, FPG, HbA1c) |>
  tidyr::pivot_longer(c(FPG, HbA1c), names_to = "endpoint", values_to = "value")

ggplot(active_long, aes(time, value, colour = arm)) +
  geom_line(linewidth = 0.8) +
  facet_grid(endpoint ~ stratum, scales = "free_y") +
  labs(
    x = "Time after first dose (weeks)", y = NULL, colour = NULL,
    title = "Replicates Figure 3 panels (c), (d), (g) and (h)",
    subtitle = "Label doses; naive and add-on strata"
  ) +
  theme_bw()
```

```{r endpoint-active-checks}
active_summary <- active |>
  dplyr::filter(stratum == "naive") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    fpg_ss = FPG[time == 52],
    hba1c_ss = HbA1c[time == 52],
    .groups = "drop"
  ) |>
  dplyr::mutate(
    d_fpg = fpg_ss - 160,
    d_hba1c = hba1c_ss - 7.92
  )

knitr::kable(
  active_summary |>
    dplyr::rename(
      "Arm" = arm, "FPG at week 52" = fpg_ss, "HbA1c at week 52" = hba1c_ss,
      "FPG change from baseline" = d_fpg, "HbA1c change from baseline" = d_hba1c
    ),
  digits = 3,
  caption = "Naive-stratum endpoint change from baseline at week 52, at each drug's label dose."
)

drug_rows <- active_summary[active_summary$arm != "placebo", ]

# Each drug must lower both endpoints relative to baseline, and the three must
# land close together (the paper's central claim: same Emax, different EC50, so
# similar efficacy at label doses). Deterministic typical-value solve.
stopifnot(
  all(drug_rows$d_fpg < -15),
  all(drug_rows$d_hba1c < -0.6),
  diff(range(drug_rows$d_fpg)) < 2,
  diff(range(drug_rows$d_hba1c)) < 0.1
)

# The HbA1c reductions must be clinically plausible: SGLT2-inhibitor phase III
# trials report roughly -0.6 to -1.0% total change from baseline at label dose.
stopifnot(all(drug_rows$d_hba1c > -1.2), all(drug_rows$d_hba1c < -0.5))
```

## The FPG onset rate, and why the drug term carries it

```{r endpoint-onset}
onset <- active |>
  dplyr::filter(stratum == "naive", arm == "dapagliflozin") |>
  dplyr::select(time, FPG)

fpg_span <- onset$FPG[onset$time == 0] - min(onset$FPG)
t90 <- min(onset$time[onset$FPG <= onset$FPG[onset$time == 0] - 0.9 * fpg_span])

kfp <- exp(ep_ui$theta[["lkfp"]])
kout <- exp(ep_ui$theta[["lkout"]])

knitr::kable(
  data.frame(
    Quantity = c(
      "Kfp (1/week)", "FPG response half-life (weeks)",
      "Kout (1/week)", "HbA1c turnover half-life (weeks)",
      "Simulated weeks to 90% of the FPG fall"
    ),
    Value = c(kfp, log(2) / kfp, kout, log(2) / kout, t90)
  ),
  digits = 3,
  caption = "Response time constants."
)

# Yao 2023 Discussion: Kfp of 0.34 /week is "a half-life of 2 weeks", and Kout
# of 0.20 /week is "a half-life of 3.5 weeks". Both are exact arithmetic.
stopifnot(
  abs(log(2) / kfp - 2) < 0.05,
  abs(log(2) / kout - 3.5) < 0.05
)

# The drug-arm FPG fall must be GRADUAL, not an instantaneous step at t = 0.
# Equation 4 as printed omits the (1 - exp(-Kfp*t)) factor and would make it a
# step; Figure 3(c) shows the simulated median leaving the 160 mg/dL baseline
# and reaching plateau at about 3/Kfp = 8.8 weeks. The model reproduces that.
stopifnot(
  abs(onset$FPG[onset$time == 0] - 160) < 1e-8,
  abs(t90 - log(10) / kfp) < 1.5
)
```

## Between-study variability

The four models carry between-study, not between-subject, variability. The
band below is therefore the spread of **study-arm means** a meta-analysis of
this size would see, which is what Figure 3's simulated percentiles represent.

```{r endpoint-itv}
n_studies <- 200

itv_rows <- dplyr::bind_rows(lapply(c("placebo", "dapagliflozin"), function(a) {
  dplyr::bind_cols(
    data.frame(arm = rep(a, n_studies)),
    auc_cols(if (a == "placebo") "none" else a,
             if (a == "placebo") 0 else auc_label[["dapagliflozin"]]),
    strata_cols("naive")[rep(1, n_studies), , drop = FALSE]
  )
}))

itv_ev <- ep_events(itv_rows, seq(0, 104, by = 4))
itv <- rxode2::rxSolve(
  readModelDb("Yao_2023_sglt2_endpoints_mbma"),
  itv_ev, keep = "arm", returnType = "data.frame"
)

itv_band <- itv |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    p05 = quantile(HbA1c, 0.05),
    p50 = median(HbA1c),
    p95 = quantile(HbA1c, 0.95),
    .groups = "drop"
  )

ggplot(itv_band, aes(time, p50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time after first dose (weeks)", y = "HbA1c (%)", colour = NULL, fill = NULL,
    title = "Between-study spread of study-arm mean HbA1c (naive stratum)",
    subtitle = paste0(n_studies, " simulated study arms; median and 5th-95th percentile")
  ) +
  theme_bw()
```

```{r endpoint-itv-checks}
# These are COHORT-derived quantities. rxSetSeed() fixes the draw only for a
# given solver thread count, so the bounds below are deliberately loose enough
# to survive a different cohort while still failing on a mis-transcribed
# parameter (which moves HbA1c by several tenths of a percent).
itv_end <- itv_band[itv_band$time == 104, ]
placebo_end <- itv_end$p50[itv_end$arm == "placebo"]
drug_end <- itv_end$p50[itv_end$arm == "dapagliflozin"]

stopifnot(
  # The drug arm must sit clearly below the placebo arm at the end of follow-up.
  placebo_end - drug_end > 0.4,
  # Both medians must stay inside a physiologically sensible HbA1c range.
  drug_end > 6, placebo_end < 9,
  # The between-study band must be non-degenerate but not implausibly wide.
  all(itv_end$p95 - itv_end$p05 > 0.05),
  all(itv_end$p95 - itv_end$p05 < 3)
)
```

# Assumptions and deviations

## Two printed equations reconciled against the paper's own outputs

Two of the eight printed equations do not reproduce the paper's own figures as
written. Both are transcribed here in the form that does, with the evidence
recorded next to the equation in
`inst/modeldb/specificDrugs/Yao_2023_sglt2_endpoints_mbma.R`.

1. **Equation 5 sign (HbA1c placebo response).** Printed as
   `HbA1c_placebo = HbA1c_baseline - Phmax * (1 - exp(-Khp*t)) + DIShp*t`,
   while Table 2 reports **signed** Phmax values (-0.200, +0.0510, -0.230,
   -0.06 %). Taking both literally double-negates and makes the naive placebo
   arm rise. The Results text states the opposite ("the estimated Phmax value
   in HbA1c for naive groups, add-on groups, and mixed groups were -0.20%,
   -0.23%, and -0.06% decrease from baseline while that of non-naive groups was
   0.051% increase from baseline"), and Figure 3(e) shows the simulated placebo
   median dipping about 0.15% below baseline before disease progression carries
   it back up. The companion FPG Equation 3 uses `+ Pfmax` with the same signed
   convention. **This model uses `+ Phmax`.**

2. **Equation 4 onset (FPG drug effect).** Printed as
   `FPG = FPG_placebo + SLOPEfd * dUGEc`, which makes the drug effect an
   instantaneous step at t = 0. The Discussion attributes the onset rate to the
   drug effect ("The FPG responses in drug effects with a Kfp of 0.34 weeks-1
   or a half-life of 2 weeks indicated a 2-week continuous treatment could show
   a significant decrease in FPG"), yet Kfp appears only in Equation 3. Figure 3
   panels (c) and (d) settle it: the simulated median starts at the 160 mg/dL
   baseline at t = 0 and reaches its plateau at about 9 weeks, which is
   3 / Kfp = 8.8 weeks. A step would put the median at the plateau already at
   t = 0. **This model multiplies the drug term by `(1 - exp(-Kfp * t))`** --
   the minimal change that reproduces the figure, and algebraically identical
   to folding `SLOPEfd * dUGEc` into the Pfmax slot of Equation 3.

## Reading of Equation 7

The unsubscripted `HbA1c` in the elimination term of Equation 7 is taken to be
the **total** of Equation 8 (placebo + drug). That is the literal reading of the
symbol and the only one under which the stated initial condition
`HbA1c_drug(0) = 0` is stationary when there is no drug and no placebo effect;
reading it as `HbA1c_drug` alone drives that state to the HbA1c baseline and
doubles the predicted HbA1c.

## The dapagliflozin food-effect equation is not printed

Table 1 reports the fed-state coefficient as a bare `0.254` with no covariate
equation anywhere in the paper or in Appendix S1. It is applied here as a
multiplier, `Kt_fed = Kt * 0.254`. That is the standard NONMEM encoding for a
covariate reported as a bare fraction, it is the only reading that produces the
"longer time to maximum concentration" the Discussion describes, and it moves
the simulated Tmax from roughly 1 h fasted to roughly 2 h fed, matching the
observed dapagliflozin high-fat-meal effect. The alternative reading
`Kt * (1 - 0.254)` changes Tmax by only about a third and would not be a
significant covariate at p < 0.005.

## Residual-error weighting

Appendix S1 Equation S14 writes the residual model as
`Y_obs = Y_pred + Y_pred * eps1 * W + eps2 * W` with `W` the square root of the
study sample size, so a **larger** study would carry a **larger** residual. The
usual NONMEM convention for a meta-analysis weight is the reverse (divide by
`sqrt(N)`), and the reported variances are only plausible under one direction
for the PK layer and the other for the endpoint layer: the PK proportional SD
of 0.676 is 68% at `W = 1`, implausible for study-arm means unless divided,
whereas the FPG additive SD of 0.18 mg/dL is implausibly small unless
multiplied. The direction is not resolvable from any on-disk source. The models
therefore carry the **reported variances back-transformed to SD at `W = 1`**,
verbatim, and every validation above uses `rxode2::zeroRe()` typical-value
solves so that no conclusion depends on the weighting. A downstream user
simulating a study of `N` subjects must decide the direction themselves.

## Other assumptions

* **Inter-study, not inter-individual.** Table 1 and Table 2 head the variance
  column "IIV (%)", but Appendix S1 section 1.2 defines it as inter-study
  variability. All etas are named `eta_study_*` and the models simulate
  study-arm means only.
* **IIV(%) to omega.** Exponential terms use `omega^2 = log((IIV/100)^2 + 1)`;
  the additive terms on Pfmax and Phmax use
  `omega^2 = ((IIV/100) * |value|)^2`, the same reading applied to the sibling
  T2DM meta-analysis `modellib("Li_2015_taspoglutide_mbma")`.
* **Apparent PK parameters.** The paper fits oral data only and reports no
  bioavailability term, so CL, Vc, CLD and VT are apparent (CL/F, Vc/F, ...).
* **Canagliflozin clearance.** The fitted CL/F of 12.0 L/h is roughly a third
  of the value implied by the labelled single-dose AUC, so `AUC_CANA` computed
  from this model is correspondingly larger than a label-derived AUC. The
  canagliflozin EC50 of 2310 ng*h/mL is scaled to the same internal exposure,
  so the pair is self-consistent; an externally derived canagliflozin AUC must
  **not** be fed to this endpoint model.
* **SLOPEfd units.** Table 2 prints "mg/dL^2". For `SLOPEfd * dUGEc` to have
  units of mg/dL with dUGEc in g/(mg/dL), the coefficient must be
  (mg/dL)^2 per g. The numeric value is used as printed.
* **Not extracted.** The external-validation exercise against ertugliflozin
  (Figure 4) reuses the endpoint model with an ertugliflozin PK/PD model whose
  parameters the paper does not tabulate, so no ertugliflozin model file is
  produced. The MPG-corrected variant of the biomarker (Figure S4) was rejected
  by the authors on objective-function value and is not extracted.

# Session information

```{r session-info, echo = FALSE}
sessionInfo()
```
