---
title: "Flurbiprofen enantiomers in plasma and CSF (Yao 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flurbiprofen enantiomers in plasma and CSF (Yao 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
```

# Model and source

Yao 2025 fitted **two separate population PK models**, one per flurbiprofen
enantiomer, to plasma and cerebrospinal-fluid (CSF) concentrations measured
after a single intravenous dose of the prodrug flurbiprofen axetil. The two
models share a structure but were built independently, have different retained
covariates, and are reported in two different tables, so they are packaged as
two model files linked to this one vignette.

```{r citation, echo = FALSE, results = "asis"}
mS <- readModelDb("Yao_2025_flurbiprofen_s")
mR <- readModelDb("Yao_2025_flurbiprofen_r")
cat("*", "Citation:", mS()$reference, "\n\n")
cat("* **S(+)-flurbiprofen** (`Yao_2025_flurbiprofen_s`):", mS()$description, "\n\n")
cat("* **R(-)-flurbiprofen** (`Yao_2025_flurbiprofen_r`):", mR()$description, "\n\n")
```

* Article: <https://doi.org/10.2147/DDDT.S542722>
* Trial registration: ClinicalTrials.gov NCT04128410

Both models use plasma as the central compartment and CSF as the peripheral
compartment, so the quantities Yao 2025 tabulates as `Vp` and `Q` are the CSF
distribution volume and the plasma-CSF intercompartmental clearance. The
abstract restates `Q` as a "CSF clearance" and `Vp` as a "CSF Vd"; there is no
separate elimination pathway out of the CSF compartment in either model.

# Population

67 of 70 enrolled adults (3 excluded for undetectable genotypes) undergoing
elective unilateral joint replacement under spinal anaesthesia at Peking
University People's Hospital, Beijing, between October 2019 and June 2020
(Yao 2025 Table 1). Median age 70 years (IQR 67-75, range 49-83), median
weight 70 kg (range 47-96), 57 of 67 female (85.1%), all Chinese. Every
patient received a single 100 mg intravenous injection of flurbiprofen axetil
(FEX, 5050E; Beijing Tide Pharmaceutical) at 2 mL/min.

Medical-ethics constraints allowed only **one CSF sample per participant**, so
patients were block-randomised into 10 groups of about 7 and each group was
sampled at a single nominal post-dose time (5, 10, ... 50 min), with the paired
venous plasma sample drawn simultaneously from the contralateral arm. The whole
dataset is therefore 67 plasma + 67 CSF concentrations, one pair per subject.
This is an extremely sparse design and the reported inter-individual variances
are correspondingly weakly identified.

Enantioselective LC-MS/MS on a CHIRALPAK-IG3 column quantified both isomers;
the plasma assay was linear over 0.1-10 ug/mL and the CSF assay over
1-100 ng/mL. Those two windows are used below as an independent plausibility
check on the model's predicted concentrations.

Programmatic access to the structured population metadata is via
`readModelDb("Yao_2025_flurbiprofen_s")()$population`.

```{r population-table, echo = FALSE}
readModelDb("Yao_2025_flurbiprofen_s")()$population[
  c("n_subjects", "age_median", "weight_median", "bmi_median", "bsa_median",
    "sex_female_pct")
] |>
  unlist() |>
  tibble::enframe(name = "Field", value = "Value") |>
  knitr::kable(caption = "Selected population metadata (Yao 2025 Table 1).")
```

# Source trace

Every `ini()` parameter carries an in-file source-trace comment next to its
value. The table below collects them in one place for review. `Vp` in the
source is the CSF compartment volume and is encoded `lvcsf`, following the
`Kumpulainen_2010_flurbiprofen` precedent for a named CSF compartment.

| Parameter | S(+) value | R(-) value | Source location |
|-----------|-----------|-----------|-----------------|
| `lvc` (Vc, plasma) | 25.6 L | 17.0 L | Table 2 / Table 3, Estimate column |
| `lcl` (CL, plasma) | 16.7 L/h | 11.8 L/h | Table 2 / Table 3; Discussion restates as 16.67 and 11.76 L/h |
| `lvcsf` (Vp, CSF) | 32.6 L | 79.1 L | Table 2 / Table 3; abstract restates as the "CSF Vd" |
| `lq` (Q, plasma-CSF) | 0.39 L/h | 0.45 L/h | Table 2 / Table 3; abstract restates as the "CSF CL" |
| `e_bsa_vc` (BSA on Vc) | not retained | 1.37 | Table 3 "BSA on Vc"; Methods Eq. I power form on median-normalised covariate |
| `e_snp_abcb1_rs1045642_ga_cl` | -1.52 | not retained | Table 2 "ABCB1 (rs1045642) on CL", GA row |
| `e_snp_abcb1_rs1045642_gg_cl` | 0.19 | not retained | Table 2 "ABCB1 (rs1045642) on CL", GG row |
| `e_snp_por_rs1057868_ga_cl` | not retained | -0.29 | Table 3 "POR (rs1057868) on CL", GA row |
| `e_snp_por_rs1057868_gg_cl` | not retained | -2.01 | Table 3 "POR (rs1057868) on CL", GG row |
| `etalvc` (IIV on Vc) | omega = 0.13 | omega = 0.03 | Table 2 / Table 3, "Interindividual variability" block |
| `etalcl` (IIV on CL) | omega = 0.16 | omega = 0.12 | Table 2 / Table 3 |
| `etalvcsf` (IIV on Vp) | omega = 0.25 | omega = 0.22 | Table 2 / Table 3 |
| `etalq` (IIV on Q) | omega = 0.15 | omega = 0.14 | Table 2 / Table 3 |
| `addSd` (plasma residual) | 0.003 mg/L | 0.003 mg/L | Table 2 / Table 3, "Plasma, additive error, sigma" |
| `propSd_Ccsf` (CSF residual) | 0.001 | 0.001 | Table 2 / Table 3, "CSF, multiplicative error, sigma" |
| Two-compartment ODE structure | n/a | n/a | Results: "Plasma and CSF were conceptualized as the central and peripheral compartments, respectively (Figure S1)" |
| Exponential IIV | n/a | n/a | Methods, Structural Model: "This study used the exponential model to describe the inter-individual variability" |
| Categorical covariate form | n/a | n/a | Methods, Population Covariate Analysis, Equation II (indicator variables inside an exponential, reference coded 0) |
| Continuous covariate form | n/a | n/a | Methods, Population Covariate Analysis, Equation I (power function after normalisation to the median) |

The tabulated `omega` values are SDs of eta on the log scale, per the footnote
of both tables ("omega, square root of interindividual variance for
parameters"); the model files carry `omega^2` as the internal variance. See
*Assumptions and deviations* for the `(%)` header inconsistency.

# Virtual cohort

Original observed data are not public (Data Sharing Statement: "The research
data are confidential"). The cohort below reproduces the Table 1 covariate
distributions: genotype frequencies at the exact observed counts, and BSA drawn
on **the paper's own BSA scale** (median 2.6 m^2, IQR 2.5-2.7, truncated to the
reported 2.0-3.2 range). That scale is not reproducible from the paper's height
and weight -- see *Assumptions and deviations* -- but the BSA exponent was
estimated against it, so the model must be driven on the same scale.

```{r cohort}
set.seed(20251009L)  # Yao 2025 publication date, 9 October 2025

n_sub <- 200L  # per enantiomer; 200/arm is the nlmixr2lib cohort cap

# Genotype frequencies exactly as counted in Yao 2025 Table 1.
abcb1_levels <- rep(c("AA", "GA", "GG"), times = c(12L, 40L, 15L))  # rs1045642
por_levels   <- rep(c("AA", "GA", "GG"), times = c(17L, 41L,  9L))  # rs1057868

# BSA on the paper's own scale: median 2.6, IQR 2.5-2.7 -> sd ~ 0.2/1.349.
draw_bsa <- function(n) {
  pmin(3.2, pmax(2.0, rnorm(n, mean = 2.6, sd = 0.2 / 1.349)))
}

cohort <- tibble(
  id     = seq_len(n_sub),
  BSA    = draw_bsa(n_sub),
  ABCB1  = sample(abcb1_levels, n_sub, replace = TRUE),
  POR    = sample(por_levels,   n_sub, replace = TRUE)
) |>
  mutate(
    SNP_ABCB1_RS1045642_GA = as.numeric(ABCB1 == "GA"),
    SNP_ABCB1_RS1045642_GG = as.numeric(ABCB1 == "GG"),
    SNP_POR_RS1057868_GA   = as.numeric(POR   == "GA"),
    SNP_POR_RS1057868_GG   = as.numeric(POR   == "GG")
  )

# Observation grid spans the study's own sampling window, 0-50 min.
obs_times_h <- seq(0, 50 / 60, length.out = 51L)

# Both model outputs (Cc, Ccsf) are ALGEBRAIC observables, not ODE states, so
# rxode2 requires dvid on the observation records and cmt = NA; it then returns
# Cc and Ccsf as columns on every observation row. Naming an observable as the
# compartment instead would inject an extra cmt slot and renumber the ODE
# states, so the observation records deliberately carry no compartment name.
make_events <- function(cov_df, times_h) {
  doses <- cov_df |>
    mutate(time = 0, amt = 100, evid = 1L,
           cmt = "central", dvid = NA_integer_)
  obs <- tidyr::expand_grid(id = cov_df$id, time = times_h) |>
    left_join(cov_df, by = "id") |>
    mutate(amt = NA_real_, evid = 0L,
           cmt = NA_character_, dvid = 1L)
  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    as.data.frame()
}

events <- make_events(cohort, obs_times_h)

# Real guard: (id, time, evid) must be unique. A dose and an observation both
# sit at t = 0 but carry different evid, so the triple is still unique.
stopifnot(anyDuplicated(events[, c("id", "time", "evid")]) == 0L)
stopifnot(sum(events$evid == 1L) == n_sub)
```

# Simulation

```{r simulate}
sim_S <- rxode2::rxSolve(
  mS, events = events,
  keep = c("BSA", "ABCB1", "POR")
) |>
  as.data.frame() |>
  mutate(enantiomer = "S(+)")

sim_R <- rxode2::rxSolve(
  mR, events = events,
  keep = c("BSA", "ABCB1", "POR")
) |>
  as.data.frame() |>
  mutate(enantiomer = "R(-)")

sim <- bind_rows(sim_S, sim_R)

# Typical-value (IIV-stripped) profiles for the closed-form gates below.
typ_cov <- cohort[1, ] |>
  mutate(BSA = 2.6,
         SNP_ABCB1_RS1045642_GA = 0, SNP_ABCB1_RS1045642_GG = 0,
         SNP_POR_RS1057868_GA = 0,   SNP_POR_RS1057868_GG = 0)
typ_events <- make_events(typ_cov, obs_times_h)

typ_S <- rxode2::rxSolve(rxode2::zeroRe(mS), events = typ_events) |>
  as.data.frame() |> mutate(enantiomer = "S(+)")
typ_R <- rxode2::rxSolve(rxode2::zeroRe(mR), events = typ_events) |>
  as.data.frame() |> mutate(enantiomer = "R(-)")
typ <- bind_rows(typ_S, typ_R)
```

# Replicate published figures

## Figure 2: visual predictive check, plasma and CSF, both enantiomers

Yao 2025 Figure 2 shows four VPC panels -- R(-) plasma, R(-) CSF, S(+) plasma,
S(+) CSF -- as observed concentrations against time after dose with the 5th,
50th and 95th percentiles of the model predictions. The published panels have no
digitisable axis labels reproduced in the open-access text, so the comparison
here is structural: the simulated percentile bands should sit inside the assay
calibration windows (0.1-10 ug/mL plasma, 1-100 ng/mL CSF) across the whole
5-50 min sampling window, which is the range the observations were drawn from.

```{r figure-2, fig.height = 6}
# rxSolve() returns observation records only, so there is no evid column to
# filter on here.
vpc_bands <- sim |>
  select(enantiomer, time, Cc, Ccsf) |>
  pivot_longer(c(Cc, Ccsf), names_to = "matrix", values_to = "conc") |>
  mutate(
    matrix = factor(matrix, levels = c("Cc", "Ccsf"),
                    labels = c("Plasma (mg/L)", "CSF (ng/mL)")),
    conc   = if_else(matrix == "CSF (ng/mL)", conc * 1000, conc),
    minutes = time * 60
  ) |>
  group_by(enantiomer, matrix, minutes) |>
  summarise(
    Q05 = quantile(conc, 0.05), Q50 = quantile(conc, 0.50),
    Q95 = quantile(conc, 0.95), .groups = "drop"
  )

ggplot(vpc_bands, aes(minutes)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(aes(y = Q50), linewidth = 0.9, colour = "steelblue4") +
  facet_grid(matrix ~ enantiomer, scales = "free_y") +
  labs(
    title = "Replicates Figure 2 of Yao 2025 (VPC, plasma and CSF)",
    subtitle = "Median with 5th-95th percentile band, 200 virtual subjects per enantiomer",
    x = "Time after dose (min)", y = NULL
  ) +
  theme_bw()
```

```{r figure-2-assay-window}
assay_check <- vpc_bands |>
  filter(minutes >= 5) |>
  group_by(enantiomer, matrix) |>
  summarise(lowest_p05 = min(Q05), highest_p95 = max(Q95), .groups = "drop") |>
  mutate(
    assay_window = if_else(matrix == "Plasma (mg/L)",
                           "0.1 - 10 mg/L", "1 - 100 ng/mL"),
    inside = if_else(
      matrix == "Plasma (mg/L)",
      lowest_p05 >= 0.1 & highest_p95 <= 10,
      lowest_p05 >= 1 & highest_p95 <= 100
    )
  )

assay_check |>
  rename(
    "Enantiomer" = enantiomer, "Matrix" = matrix,
    "Lowest 5th pct" = lowest_p05, "Highest 95th pct" = highest_p95,
    "Assay calibration range" = assay_window, "Entirely inside assay range" = inside
  ) |>
  knitr::kable(
    digits = 3,
    caption = "Simulated 5-50 min percentile envelope against the published assay calibration ranges."
  )
```

## Genotype and BSA covariate effects

Yao 2025 reports the covariate effects only as coefficients, not as figures.
The panels below show what those coefficients imply for the typical-value
plasma profile of each enantiomer.

```{r figure-covariates, fig.height = 4.5}
geno_grid <- function(model, snp_prefix, levels_lbl) {
  per_genotype <- lapply(levels_lbl, function(g) {
    cov <- typ_cov
    cov[[paste0(snp_prefix, "_GA")]] <- as.numeric(g == "GA")
    cov[[paste0(snp_prefix, "_GG")]] <- as.numeric(g == "GG")
    out <- rxode2::rxSolve(rxode2::zeroRe(model),
                           events = make_events(cov, obs_times_h)) |>
      as.data.frame()
    out$genotype <- g
    out
  })
  bind_rows(per_genotype)
}

geno_S <- geno_grid(mS, "SNP_ABCB1_RS1045642", c("AA", "GA", "GG")) |>
  mutate(enantiomer = "S(+), ABCB1 rs1045642")
geno_R <- geno_grid(mR, "SNP_POR_RS1057868", c("AA", "GA", "GG")) |>
  mutate(enantiomer = "R(-), POR rs1057868")

bind_rows(geno_S, geno_R) |>
  ggplot(aes(time * 60, Cc, colour = genotype)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~enantiomer) +
  labs(
    title = "Genotype effect on the typical-value plasma profile",
    subtitle = "Reference genotype is AA in both models (Yao 2025 Tables 2 and 3)",
    x = "Time after dose (min)", y = "Plasma concentration (mg/L)",
    colour = "Genotype"
  ) +
  theme_bw()
```

```{r figure-bsa, fig.height = 3.5}
bsa_curves <- lapply(c(2.0, 2.6, 3.2), function(b) {
  cov <- typ_cov |> mutate(BSA = b)
  out <- rxode2::rxSolve(rxode2::zeroRe(mR),
                         events = make_events(cov, obs_times_h)) |>
    as.data.frame()
  out$BSA_label <- sprintf("BSA = %.1f m^2", b)
  out
}) |>
  bind_rows()

ggplot(bsa_curves, aes(time * 60, Cc, colour = BSA_label)) +
  geom_line(linewidth = 0.9) +
  labs(
    title = "R(-)-flurbiprofen: BSA power effect on the central volume",
    subtitle = "Vc = 17.0 * (BSA / 2.6)^1.37 (Yao 2025 Table 3)",
    x = "Time after dose (min)", y = "Plasma concentration (mg/L)", colour = NULL
  ) +
  theme_bw()
```

# PKNCA validation

Yao 2025 reports **no** non-compartmental parameters -- no Cmax, Tmax, AUC or
half-life for either enantiomer -- so there is no published NCA table to compare
against. The NCA below is therefore validated against a **closed-form
reference** derived analytically from the published parameters, which is an
independent check on the ODE encoding, the dose record and PKNCA's settings
rather than a restatement of the model.

For a two-compartment IV bolus with micro-constants `k10 = CL/Vc`,
`k12 = Q/Vc`, `k21 = Q/Vcsf`, the plasma profile is
`C(t) = A*exp(-alpha*t) + B*exp(-beta*t)` where `alpha` and `beta` are the roots
of `lambda^2 - (k10+k12+k21)*lambda + k10*k21 = 0`, and

* `Cmax = D/Vc` at `t = 0`,
* `AUC(0,T) = A*(1-exp(-alpha*T))/alpha + B*(1-exp(-beta*T))/beta`.

```{r closed-form-reference}
closed_form <- function(dose, cl, vc, vcsf, q, tmax_h) {
  k10 <- cl / vc; k12 <- q / vc; k21 <- q / vcsf
  b <- k10 + k12 + k21
  disc <- sqrt(b^2 - 4 * k10 * k21)
  alpha <- (b + disc) / 2
  beta  <- (b - disc) / 2
  c0 <- dose / vc
  A <- c0 * (alpha - k21) / (alpha - beta)
  B <- c0 * (k21 - beta)  / (alpha - beta)
  list(
    cmax    = c0,
    tmax    = 0,
    auclast = A * (1 - exp(-alpha * tmax_h)) / alpha +
              B * (1 - exp(-beta  * tmax_h)) / beta,
    alpha   = alpha, beta = beta,
    thalf_alpha = log(2) / alpha, thalf_beta = log(2) / beta
  )
}

pars <- tibble::tribble(
  ~enantiomer, ~cl,  ~vc,  ~vcsf, ~q,
  "S(+)",      16.7, 25.6, 32.6,  0.39,
  "R(-)",      11.8, 17.0, 79.1,  0.45
)

ref_list <- Map(closed_form,
                dose = 100, cl = pars$cl, vc = pars$vc,
                vcsf = pars$vcsf, q = pars$q, tmax_h = 50 / 60)
names(ref_list) <- pars$enantiomer

reference_nca <- tibble(
  enantiomer = pars$enantiomer,
  cmax    = vapply(ref_list, function(x) x$cmax,    numeric(1)),
  tmax    = vapply(ref_list, function(x) x$tmax,    numeric(1)),
  auclast = vapply(ref_list, function(x) x$auclast, numeric(1))
)
```

```{r pknca}
# Typical-value profiles are the right input for the closed-form comparison:
# the reference is the typical-value analytic solution, and a log-normal eta on
# CL/Vc would push the cohort MEAN above the typical value.
nca_conc <- bind_rows(
  rxode2::rxSolve(rxode2::zeroRe(mS), events = typ_events) |>
    as.data.frame() |> mutate(enantiomer = "S(+)"),
  rxode2::rxSolve(rxode2::zeroRe(mR), events = typ_events) |>
    as.data.frame() |> mutate(enantiomer = "R(-)")
) |>
  mutate(id = 1L) |>
  # Filter on missingness ONLY. A `time > 0` or `Cc > 0` filter would drop the
  # time-zero record and trigger PKNCA's "AUC range starting before the first
  # measurement" warning for every subject.
  filter(!is.na(Cc)) |>
  select(id, enantiomer, time, Cc)

stopifnot(any(nca_conc$time == 0))  # time-zero record is mandatory for AUC

nca_dose <- nca_conc |>
  distinct(id, enantiomer) |>
  mutate(time = 0, dose = 100)

o_conc <- PKNCA::PKNCAconc(as.data.frame(nca_conc), Cc ~ time | id + enantiomer)
o_dose <- PKNCA::PKNCAdose(as.data.frame(nca_dose), dose ~ time | id + enantiomer)

intervals <- data.frame(
  start = 0, end = 50 / 60,
  cmax = TRUE, tmax = TRUE, auclast = TRUE
)

o_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals),
                       verbose = FALSE)
sim_nca <- as.data.frame(o_nca)
```

## Simulated versus closed-form NCA

```{r nca-comparison}
nca_tbl <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_nca,
  reference = reference_nca,
  by = "enantiomer",
  params = c("cmax", "tmax", "auclast"),
  units = c(cmax = "mg/L", tmax = "h", auclast = "mg*h/L")
)
knitr::kable(
  nca_tbl,
  digits = 4,
  caption = "PKNCA output over the study's 0-50 min window against the analytic two-compartment solution."
)
```

```{r nca-footnote, echo = FALSE, results = "asis"}
fn <- attr(nca_tbl, "footnote")
if (!is.null(fn)) cat("\n", fn, "\n")
```

Every parameter agrees with the analytic solution, which confirms that the ODE
system, the 100 mg dose record on `central`, the `dvid`-based observation records
and PKNCA's interval settings are all consistent.

## Clearance identity over the full profile

`AUC(0,inf) = Dose / CL` is an independent gate on the elimination term. It
requires integrating far past the 50-minute study window because of the model's
very long terminal phase (see below).

```{r clearance-identity}
long_events <- make_events(typ_cov, seq(0, 1200, length.out = 24001L))

cl_check <- lapply(seq_len(nrow(pars)), function(i) {
  mod <- if (pars$enantiomer[i] == "S(+)") mS else mR
  s <- rxode2::rxSolve(rxode2::zeroRe(mod), events = long_events) |>
    as.data.frame()
  auc_inf <- sum(diff(s$time) *
                   (head(s$Cc, -1) + tail(s$Cc, -1)) / 2)
  tibble(
    enantiomer = pars$enantiomer[i],
    `AUC(0,inf) integrated` = auc_inf,
    `Dose / CL` = 100 / pars$cl[i],
    `% diff` = 100 * (auc_inf - 100 / pars$cl[i]) / (100 / pars$cl[i])
  )
}) |>
  bind_rows()

knitr::kable(cl_check, digits = 4,
             caption = "Clearance identity AUC(0,inf) = Dose / CL.")
```

## The published parameters imply a very long terminal phase

The tabulated `Q` (0.39-0.45 L/h) is small relative to the CSF volume `Vp`
(32.6-79.1 L), so the CSF compartment behaves as a slowly-equilibrating deep
compartment and the model's terminal half-life is far longer than the
literature half-life of flurbiprofen (about 3-6 h). Because the data span only
5-50 min, this terminal phase is entirely extrapolation and is not identifiable
from the observations.

```{r halflife}
tibble(
  enantiomer = pars$enantiomer,
  `alpha t1/2 (h)` = vapply(ref_list, function(x) x$thalf_alpha, numeric(1)),
  `beta t1/2 (h)`  = vapply(ref_list, function(x) x$thalf_beta,  numeric(1)),
  `study window (h)` = 50 / 60
) |>
  knitr::kable(digits = 3,
               caption = "Analytic disposition half-lives from the published parameters.")
```

## Independent corroboration of the dose basis

Yao 2025 never states the amount entered in the modelling dataset, and labels
both `CL` and `Vc` "apparent" despite intravenous dosing. Zhang 2018 studied
the **same product (5050E, Beijing Tide) at the same centre with the same
one-sample-per-subject subarachnoid design**, and reported observed racemic
plasma concentrations of 3.48-14.56 mg/L. Summing this model's two enantiomer
predictions on a nominal 100 mg basis should land inside that observed range.

```{r dose-basis}
racemic <- typ |>
  filter(time >= 5 / 60) |>
  select(enantiomer, time, Cc) |>
  pivot_wider(names_from = enantiomer, values_from = Cc) |>
  mutate(racemic_total = `S(+)` + `R(-)`)

tibble(
  `Quantity` = c("Simulated racemic total, 5-50 min",
                 "Zhang 2018 observed racemic range"),
  `Low (mg/L)` = c(min(racemic$racemic_total), 3.48),
  `High (mg/L)` = c(max(racemic$racemic_total), 14.56)
) |>
  knitr::kable(digits = 2,
               caption = "Dose-basis cross-check against Zhang 2018 (same product, centre and design).")
```

The simulated racemic envelope sits inside the independently observed range,
supporting the nominal 100 mg flurbiprofen axetil dose basis recorded in both
model files.

# Assumptions and deviations

The paper is internally inconsistent in several places. Each item below records
what was done and why; no parameter was tuned to improve any comparison above.

1. **Dose basis is an assumption.** The paper never states the amount entered
   in the dataset. Flurbiprofen axetil 100 mg is a prodrug of racemic
   flurbiprofen, so the flurbiprofen-equivalent mass is about 70 mg
   (MW 244.3 / 348.4) and the per-enantiomer molar amount is about 35 mg -- yet
   both tables call `CL` and `Vc` "apparent" even though dosing was
   intravenous, which is the signature of a nominal-dose dataset. Both model
   files therefore use the **nominal 100 mg**, and every parameter is an
   apparent value on that basis. The Zhang 2018 cross-check above supports this
   reading; a 50 mg or 35 mg per-enantiomer basis would put the racemic total
   below the range Zhang 2018 observed. A user who prefers a different basis
   must rescale `Vc`, `Vcsf` and `CL` proportionally.

2. **Bolus, not a 5-minute infusion.** 100 mg of a 10 mg/mL formulation
   delivered at 2 mL/min is about a 5-minute injection, but neither table
   reports an infusion duration or rate parameter and the model is described
   only as two-compartment. The models encode an IV bolus into `central`. This
   matters only for the earliest sample (5 min).

3. **`Vc` for R(-) is 17.0 L, not 17.1 L.** Table 3 gives 17.0; the abstract
   twice says 17.1 for the same parameter. Table 3 is the designated
   final-model parameter table and is used.

4. **`CL` values are taken from the tables, not the Discussion.** Tables 2 and 3
   give 16.7 and 11.8 L/h; the Discussion restates the same estimates as 16.67
   and 11.76 L/h. The difference is under 0.4% and the tables are the
   designated source.

5. **Residual error follows the tables, not the R(-) prose.** Both tables list
   a plasma *additive* sigma and a CSF *multiplicative* sigma. The S(+) text
   calls this "a combined additive and proportional error model", which is
   consistent (the combination is across matrices). The R(-) text instead says
   the model "adopts a proportional residual error model", which contradicts
   its own Table 3 additive plasma row; the table is followed.

6. **Both residual-error magnitudes are implausibly small and are transcribed
   as published.** An additive plasma SD of 0.003 mg/L is 3 ng/mL, far below
   the 0.1 ug/mL plasma LLOQ, and a 0.1% proportional CSF error is tighter than
   any bioanalytical assay. The same pattern appears in Zhang 2018 (additive
   sigma 0.0023 mg/L) from the same group, so it reflects this group's Phoenix
   reporting convention rather than a transcription error. Consequence:
   simulated profiles are effectively noise-free apart from the IIV.

7. **`omega` is read as an SD, per the table footnote.** The
   "Interindividual variability" block is headed `(%)` but the footnote defines
   `omega` as the "square root of interindividual variance", so the values are
   SDs of eta on the log scale and the models carry `omega^2`. Reading them
   instead as CV fractions (`omega^2 = log(1 + CV^2)`) changes every variance
   by less than 3%, so the ambiguity is immaterial.

8. **The paper's BSA values are not reproducible from its own height and
   weight.** Table 1 reports median BSA 2.6 m^2 (range 2.0-3.2) for a cohort of
   median height 1.61 m and weight 70 kg, which give about 1.76 m^2 (DuBois) or
   1.79 m^2 (Mosteller); no standard formula yields 2.6, and the paper never
   states which formula it used. Because the exponent 1.37 was estimated
   against the paper's own BSA scale, the model normalises by the paper's
   median 2.6 m^2 and the virtual cohort samples BSA from the reported
   distribution rather than recomputing it. Driving the model with a correctly
   computed BSA would bias every predicted volume.

9. **Table 3's `Vp` bootstrap CI does not contain its own point estimate.**
   `Vp = 79.1 L` is reported with a bootstrap median of 55.3 and a 95% CI of
   33.5-58.3. The point estimate is used as published; the discrepancy is noted
   but not resolvable from the source.

10. **Two retained covariate effects have bootstrap CIs spanning or nearly
    spanning zero**: `BSA on Vc` (1.37; bootstrap median 0.59, CI -0.07 to
    1.96) and `ABCB1 GG on CL` (0.19; CI -0.16 to 0.93). They are retained
    because the final-model tables retain them.

11. **Genotype wild-type orientation is unresolved.** Both retained SNPs are
    reported as AA/GA/GG with AA as the reference category, but the paper never
    states which genotype is the wild type, and it writes the loci in cDNA
    nomenclature (`3435C>T`, `POR*28`) that cannot be mapped to A/G without
    knowing the assay's strand convention. For *ABCB1* rs1045642 the minus-strand
    mapping and the observed allele frequency both suggest AA is the *variant*
    homozygote; for *POR* rs1057868 the paper's own mechanism narrative
    ("compromised electron transfer from POR to CYP450") instead implies AA is
    the wild type, and the allele-frequency check is unusable because that locus
    deviates from Hardy-Weinberg equilibrium in this cohort. Because the model's
    predictions are identical either way, the covariates are named by the
    **reported genotype letters** (`SNP_ABCB1_RS1045642_GA` / `_GG`,
    `SNP_POR_RS1057868_GA` / `_GG`) so that no unverifiable wild-type claim is
    encoded. These are **not** poolable with the existing carrier indicator
    `SNP_ABCB1_RS1045642`, whose reference is the c.3435CC wild type.

12. **The model's terminal phase is extrapolation.** As tabulated above, the
    published `Q` and `Vp` imply a terminal half-life of about 59 h (S(+)) and
    127 h (R(-)), against a literature flurbiprofen half-life of 3-6 h. The
    data span only 5-50 min, so nothing in the study constrains the terminal
    phase. Use these models for the early distribution phase they were fitted
    to; do not use them for accumulation or steady-state predictions.

13. **The paper's genotyping panel and covariate list disagree.** The
    Pharmacogenetic methods section lists one SNP panel (CYP3A4\*1G, CYP3A5\*3,
    three *ABCB1* SNPs, *ABCG2*, POR\*28, two *PXR* SNPs, *CAR*) while the
    covariate-analysis section lists a different set (two *PXR*, two *POR*, two
    *ABCB1*, three *CYP2C9*, one *UGT1A9*); the abstract says 12 SNPs and the
    Results say 11; Table 1 lists `rs1057910` twice; and the R(-) covariate text
    mentions a `POR (rs2868177)` that appears nowhere else. Only the two SNPs
    that appear in the final-model tables are encoded. The screened-but-dropped
    covariates are recorded in each model's `covariatesDataExcluded`.

14. **Figure S1 and Table S1 were not available.** The supplement was not in
    the open-access package. Figure S1 is the structural-model schematic, which
    the Results text describes completely ("Plasma and CSF were conceptualized
    as the central and peripheral compartments"), and Table S1 lists genotyping
    primers, which carry no model parameters. No parameter value depends on the
    missing supplement.

15. **No published NCA to compare against.** The paper reports no Cmax, Tmax,
    AUC or half-life, so the PKNCA section is validated against the analytic
    two-compartment solution and, for the dose basis, against Zhang 2018's
    observed racemic concentrations. Nothing was tuned.

16. **Not modelled:** concomitant 1 mg IV midazolam and 15-20 mg subarachnoid
    ropivacaine, which every patient received; the paper does not model them
    either.
