---
title: "Mycophenolic acid (Ye 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mycophenolic acid (Ye 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Ye_2025_mycophenolic_acid"))$reference`
* Description: `r rxode2::rxode(readModelDb("Ye_2025_mycophenolic_acid"))$description`
* Article: <https://doi.org/10.1136/lupus-2025-001535>
* Supplement (online supplemental tables 1-5 and plot 1, open access via the
  journal): <https://doi.org/10.1136/lupus-2025-001535>

Ye 2025 is, by the authors' account, the first population PK analysis of
mycophenolic acid (MPA) in paediatric lupus nephritis. Fifty-one children
receiving oral mycophenolate mofetil (MMF) twice daily contributed 146
steady-state concentration-time profiles. The final model is a two-compartment
model with first-order absorption, an absorption lag time and linear
elimination, fitted in Phoenix NLME. Twenty-nine SNPs across 13 candidate genes
were screened and eight were associated with MPA exposure in the univariate
analysis, but **no genetic covariate survived the stepwise population-PK
covariate search**; body weight on the apparent peripheral volume was the only
covariate retained.

# Population

Fifty-one children with lupus nephritis (5 male, 46 female; 90.2% female) were
recruited prospectively at The First Affiliated Hospital of Sun Yat-sen
University between September 2021 and January 2023 (Ye 2025 Table 1). Mean age
was 12.22 +/- 2.33 years and mean body weight 41.13 +/- 11.31 kg. All patients
met the 2019 ACR SLE classification criteria and the diagnostic criteria for
lupus nephritis. MMF was given as capsules (CellCept) or dispersible tablets
(Saikeping) at 125-750 mg every 12 hours, mean total daily dose 739.81 +/-
247.37 mg/day. All patients received corticosteroids; 17.6% received
tacrolimus and 35.3% hydroxychloroquine.

Sampling was at steady state (MMF for at least 7 days, mean treatment duration
48.66 days) pre-dose and at 0.5, 1.5, 2.5, 4, 6, 9 and 12 hours post-dose.
That is eight samples per profile, which is corroborated arithmetically by the
paper's own totals: 1170 samples across 146 profiles gives 1170/146 = 8.01.
Plasma total MPA was measured by validated LC-MS/MS over 0.1-50 ug/mL
(LLOQ 0.1 ug/mL). A separate nine-patient cohort was used for external
validation (MDPE 9.09%, MAPE 24.82%, F20 52.5%, F30 72.5%).

The same information is available programmatically via
`readModelDb("Ye_2025_mycophenolic_acid")()$population`.

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Ye_2025_mycophenolic_acid.R`. The table
below collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (ka)     | 1.54 /h | Ye 2025 Table 3, `theta_Ka` (RSE 10.25%); header unit "hours" read as a typo for 1/h |
| `lvc` (Vc/F)   | 7.99 L | Ye 2025 Table 3, `theta_V/F` (RSE 22.76%) |
| `lvp` (Vp/F)   | 1287.12 L | Ye 2025 Table 3, `theta_V2/F` (RSE 8.65%); value at WT = 41.13 kg |
| `lcl` (CL/F)   | 15.23 L/h | Ye 2025 Table 3, `theta_CL/F` (RSE 1.85%) |
| `lq` (Q/F)     | 37.65 L/h | Ye 2025 Table 3, `theta_CL2/F` (RSE 3.32%) |
| `ltlag` (Tlag) | 0.40 h | Ye 2025 Table 3, `Tlag` (RSE 3.29%) |
| `e_wt_vp`      | 2.05 | Ye 2025 Table 3, `theta_v2,weight` (RSE 11.99%); functional form from online supplemental table 5 |
| `etalvc`       | 1.02 | Ye 2025 Table 3, `omega^2 V/F` (RSE 14.89%) |
| `etalvp`       | 1.24 | Ye 2025 Table 3, `omega^2 V2/F` (RSE 13.80%) |
| `etalq`        | 0.20 | Ye 2025 Table 3, `omega^2 CL2/F` (RSE 17.51%) |
| `propSd`       | 0.50 | Ye 2025 Table 3, `MultStdev` (RSE 4.29%) |
| `addSd`        | 0.38 mg/L | Ye 2025 Table 3, `stdev` (RSE 9.33%) |
| Two-compartment structure with Tlag | n/a | Ye 2025 Results, "PPK model development" (AIC 5237.58 one-compartment vs 4659.74 two-compartment; 4659.74 with Tlag vs 4712.25 without) |
| Combined proportional + additive residual | n/a | Ye 2025 Results, "PPK model development" (AIC 4659.74 mixed vs 4738.81 additive and 4742.22 multiplicative) |
| `Vp/F = 1287.12 * (WT/41.13)^2.05` | n/a | Ye 2025 online supplemental table 5 (ratio form and the 41.13 kg reference) + Table 3 (the 2.05 coefficient) |
| Reference weight 41.13 kg | n/a | Ye 2025 online supplemental table 5; equals the Table 1 model-group mean weight |
| No IIV on ka, CL/F, Tlag | n/a | Ye 2025 Results, "the ... Ka ... CL and Tlag were fixed due to their high eta-shrinkage values" |

```{r model}
mod <- readModelDb("Ye_2025_mycophenolic_acid")
ui  <- rxode2::rxode(mod)

# Structural constants used throughout, read straight off the model so the
# vignette cannot drift from the packaged parameter values.
theta   <- setNames(ui$theta, names(ui$theta))
CL_F    <- exp(theta[["lcl"]])   # 15.23 L/h
WT_REF  <- 41.13                 # kg, Ye 2025 online supplemental table 5
TAU     <- 12                    # h, q12h
DOSE    <- 369.9                 # mg per dose = 739.81 mg/day / 2 (Ye 2025 Table 1)

c(CL_F = CL_F, Vc_F = exp(theta[["lvc"]]), Vp_F = exp(theta[["lvp"]]),
  Q_F = exp(theta[["lq"]]), ka = exp(theta[["lka"]]), Tlag = exp(theta[["ltlag"]]),
  e_wt_vp = theta[["e_wt_vp"]])
```

# Structural checks

These four checks are **deterministic** -- both sides use the same drawn
parameters, or are analytic -- so they are asserted tightly. They are the ones
that catch a mis-transcribed value, a mis-assigned compartment or a silently
substituted solver.

## 1. Steady-state AUC equals Dose / (CL/F), for every subject

Ye 2025 retained **no** random effect on CL/F and **no** covariate on CL/F.
Steady-state exposure in this model is therefore an exact deterministic
function of dose: `AUC(0-tau),ss = Dose / (CL/F)` for every subject, whatever
their weight, Vc, Vp or Q. This is both the strongest available check on the
disposition parameters and a substantive property of the published model.

```{r cohort}
# set.seed() seeds R's RNG. It does NOT seed rxode2's simulation RNG, and
# rxode2's streams are partitioned per solver thread -- so this cohort is
# reproducible here and different on a machine with a different thread count.
# Every assertion below is written to hold for ANY cohort the model can draw.
set.seed(20250208)

N_SUBJ <- 146L   # matches the 146 steady-state profiles in Ye 2025

# Body weight: Ye 2025 Table 1 reports only mean +/- SD (41.13 +/- 11.31 kg),
# so a truncated normal is used, clipped to a plausible paediatric range for
# a cohort of mean age 12.2 years.
cohort <- tibble::tibble(
  id        = seq_len(N_SUBJ),
  WT        = pmin(pmax(rnorm(N_SUBJ, mean = 41.13, sd = 11.31), 18), 75),
  treatment = "MMF 370 mg q12h"
)

# Dense observation grid over one steady-state dosing interval, refined over
# the absorption phase. The refinement is load-bearing: with IIV variance 1.02
# on Vc/F, the low-Vc tail of the cohort has a very sharp peak, and a uniform
# 0.05 h grid leaves ~0.5% trapezoidal error on those subjects -- enough to
# swamp the analytic AUC check below. At 0.005 h over 0-3 h the worst-subject
# error falls to ~0.01%. Every one of the study's eight sampling times is a
# subset of this grid.
grid_dense  <- sort(unique(c(seq(0, 3, by = 0.005), seq(3, TAU, by = 0.05))))
grid_study  <- c(0, 0.5, 1.5, 2.5, 4, 6, 9, 12)   # Ye 2025 Methods
stopifnot(all(grid_study %in% grid_dense))

# ss = 1 places the subject at exact steady state on a q12h schedule. Ye 2025
# sampled after a mean of 48.66 days of therapy (~14 terminal half-lives of
# this model), so exact steady state is the right idealisation; the
# convergence check in section 4 confirms it against explicit repeat dosing.
events <- dplyr::bind_rows(
  cohort |>
    dplyr::mutate(time = 0, amt = DOSE, evid = 1L, ii = TAU, ss = 1L,
                  cmt = "depot"),
  cohort |>
    tidyr::crossing(time = grid_dense) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, ii = 0, ss = 0L,
                  cmt = "central")
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

```{r simulate}
# NB: rxSolve returns a column literally named `sim` (the observation with
# residual error) alongside `Cc` (the individual prediction, noise-free).
# The data frame is deliberately NOT called `sim`, so the column and the
# frame cannot shadow each other.
simdf <- rxode2::rxSolve(
  mod, events = events, keep = c("WT", "treatment"), addDosing = FALSE
) |>
  as.data.frame()

trapz <- function(t, y) sum(diff(t) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

auc_by_subject <- simdf |>
  dplyr::group_by(id, WT) |>
  dplyr::summarise(auc_ss = trapz(time, Cc), .groups = "drop") |>
  dplyr::mutate(
    auc_closed_form = DOSE / CL_F,
    pct_diff        = 100 * (auc_ss - auc_closed_form) / auc_closed_form
  )

summary(auc_by_subject$pct_diff)

# Deterministic identity, not a cohort statistic: the only error is the
# trapezoidal quadrature on the grid above. Realised worst-subject |pct| was
# 0.0095 on the refined grid (against 0.51 on a uniform 0.05 h grid), so 0.25
# carries ~25x headroom for an unluckier low-Vc draw while still failing
# instantly on a mis-transcribed CL, dose or unit -- those move it by tens of
# percent.
stopifnot(max(abs(auc_by_subject$pct_diff)) < 0.25)
```

## 2. Body weight moves Vp only -- steady-state exposure is weight-independent

Ye 2025's title and conclusion foreground body weight, and the model does carry
a strong weight effect: with an exponent of 2.05, apparent peripheral volume
changes roughly four-fold across this cohort's weight range. But weight enters
`Vp/F` **alone**, so it changes the *shape* of the profile and not the
steady-state exposure. This is worth stating plainly, because the paper's
closing recommendation ("an individualised initial dose and adjustment based on
body weight") reads naturally as a statement about exposure, whereas in the
published model weight-based dosing changes AUC only through the dose itself.

```{r covariate-effect}
wt_probe <- c(20, 30, WT_REF, 55, 70)

probe_events <- tibble::tibble(id = seq_along(wt_probe), WT = wt_probe) |>
  (\(d) dplyr::bind_rows(
    d |> dplyr::mutate(time = 0, amt = DOSE, evid = 1L, ii = TAU, ss = 1L,
                       cmt = "depot"),
    d |> tidyr::crossing(time = grid_dense) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, ii = 0, ss = 0L,
                    cmt = "central")
  ))() |>
  dplyr::arrange(id, time, dplyr::desc(evid))

probe <- rxode2::rxSolve(
  rxode2::zeroRe(mod), probe_events, keep = "WT", addDosing = FALSE
) |>
  as.data.frame()

probe_summary <- probe |>
  dplyr::group_by(id, WT) |>
  dplyr::summarise(
    vp     = unique(vp),
    auc_ss = trapz(time, Cc),
    cmax   = max(Cc),
    cmin   = min(Cc),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    vp_expected = exp(theta[["lvp"]]) * (WT / WT_REF)^theta[["e_wt_vp"]],
    auc_ratio   = auc_ss / (DOSE / CL_F)
  )

probe_summary |>
  dplyr::transmute(
    "Body weight (kg)"      = round(WT, 2),
    "Vp/F (L)"              = round(vp, 1),
    "Vp/F expected (L)"     = round(vp_expected, 1),
    "AUC0-12,ss (mg*h/L)"   = round(auc_ss, 3),
    "Cmax,ss (mg/L)"        = round(cmax, 3),
    "Cmin,ss (mg/L)"        = round(cmin, 3)
  ) |>
  knitr::kable(
    caption = paste(
      "Typical-value steady-state profile across the cohort weight range.",
      "Vp/F spans a 4.4-fold range while AUC0-12 is unchanged, because body",
      "weight enters Vp/F only."
    )
  )

# Analytic identity: the encoded covariate form must reproduce the power
# function exactly.
stopifnot(max(abs(probe_summary$vp / probe_summary$vp_expected - 1)) < 1e-8)
# AUC is weight-independent by construction (no covariate, no eta on CL/F).
stopifnot(max(abs(probe_summary$auc_ratio - 1)) < 0.005)
# The weight effect is real and large on Vp even though AUC does not move.
stopifnot(max(probe_summary$vp) / min(probe_summary$vp) > 4)
```

## 3. The solved and ODE forms agree

The model writes an explicit three-state ODE system, but its `cl`/`vc`/`q`/`vp`
parameterisation also matches rxode2's linear-compartment solver, which
`rxSolve()` may substitute automatically (`useLinCmt = TRUE` is the default).
The substitution is only safe if the hand-written ODE rates really are the
standard two-compartment rates, so that is checked rather than assumed.

The check runs on an explicit q12h schedule rather than on the `ss = 1` records
used above, because `ss = 1` on this model *requires* the analytic solver:
with a terminal half-life near 80 h against a 12 h dosing interval, the ODE
path's steady-state iteration does not converge and `rxSolve()` fails with
"could not solve the system". Establishing the equivalence on explicit dosing
is what licenses the `ss = 1` results in the rest of this vignette.

```{r repeat-dosing}
one <- probe_events |> dplyr::filter(id == 3L)

# 30 days of q12h dosing (~8.7 terminal half-lives), observed over the final
# dosing interval on the same grid as the ss = 1 solve.
repeat_events <- dplyr::bind_rows(
  tibble::tibble(
    id = 3L, WT = one$WT[1], time = seq(0, TAU * 60, by = TAU),
    amt = DOSE, evid = 1L, cmt = "depot"
  ),
  tibble::tibble(
    id = 3L, WT = one$WT[1], time = TAU * 60 + grid_dense,
    amt = NA_real_, evid = 0L, cmt = "central"
  )
) |>
  dplyr::arrange(time, dplyr::desc(evid))
```

```{r lincmt-equivalence}
lin <- rxode2::rxSolve(mod, repeat_events, omega = NA, useLinCmt = TRUE,
                       addDosing = FALSE) |> as.data.frame()
ode <- rxode2::rxSolve(mod, repeat_events, omega = NA, useLinCmt = FALSE,
                       addDosing = FALSE) |> as.data.frame()

max_rel <- max(abs(lin$Cc - ode$Cc) / pmax(ode$Cc, 1e-12))
max_rel

# Same equations, two solvers: agreement is to solver tolerance, so this is
# asserted tightly. A divergence here would mean the explicit ODE is not the
# two-compartment system the parameter names claim.
stopifnot(max_rel < 1e-5)
```

## 4. Exact steady state matches explicit repeat dosing

```{r ss-convergence}
ss_sim <- probe |> dplyr::filter(id == 3L)
stopifnot(nrow(ss_sim) == nrow(lin))

max_rel_ss <- max(abs(lin$Cc - ss_sim$Cc) / ss_sim$Cc)
max_rel_ss

# 30 days of q12h dosing is ~8.7 terminal half-lives of this model, so the
# explicit schedule should sit just under exact steady state -- a small,
# one-sided gap. Realised ~0.24%.
stopifnot(max_rel_ss < 0.02)
```

The model's terminal half-life is far longer than anything the study's 12-hour
sampling window can see, which matters for interpreting the NCA comparison
below:

```{r terminal-half-life}
wash <- tibble::tibble(id = 1L, WT = WT_REF) |>
  (\(d) dplyr::bind_rows(
    d |> dplyr::mutate(time = 0, amt = DOSE, evid = 1L, cmt = "depot"),
    d |> tidyr::crossing(time = seq(0, 400, by = 1)) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ))() |>
  dplyr::arrange(time, dplyr::desc(evid))

wsim <- rxode2::rxSolve(rxode2::zeroRe(mod), wash, addDosing = FALSE) |>
  as.data.frame() |>
  dplyr::filter(time > 200, Cc > 0)

t_half_terminal <- log(2) / -stats::coef(stats::lm(log(Cc) ~ time, wsim))[[2]]
t_half_terminal

# Structural consequence of Vp/F = 1287 L against Q/F = 37.65 L/h; recorded
# rather than tuned. Loose bound because it is a fitted slope.
stopifnot(t_half_terminal > 40, t_half_terminal < 150)
```

# Replicate published figures

```{r figure-1b}
# Replicates Figure 1B of Ye 2025: mean +/- SD MPA concentration vs time after
# dose, on the study's eight-point sampling grid, with residual error applied
# (the `sim` column) so it is comparable to observed data.
obs_study <- simdf |> dplyr::filter(time %in% grid_study)

obs_study |>
  dplyr::group_by(time) |>
  dplyr::summarise(mean_c = mean(sim), sd_c = stats::sd(sim), .groups = "drop") |>
  ggplot(aes(time, mean_c)) +
  geom_errorbar(aes(ymin = mean_c - sd_c, ymax = mean_c + sd_c), width = 0.25) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = grid_study) +
  labs(x = "Time after dose (h)", y = "MPA concentration (mg/L)",
       title = "Simulated mean +/- SD steady-state profile",
       caption = "Replicates Figure 1B of Ye 2025.")
```

```{r figure-3}
# Replicates Figure 3 of Ye 2025: prediction-corrected VPC style summary of
# the simulated concentrations against time after dose.
simdf |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = stats::quantile(sim, 0.05),
    Q50 = stats::quantile(sim, 0.50),
    Q95 = stats::quantile(sim, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 12, by = 2)) +
  labs(x = "Time after dose (h)", y = "MPA concentration (mg/L)",
       title = "Simulated 5th / 50th / 95th percentiles at steady state",
       caption = "Replicates Figure 3 of Ye 2025 (VPC).")
```

```{r figure-covariate}
probe |>
  ggplot(aes(time, Cc, colour = factor(round(WT, 1)))) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = seq(0, 12, by = 2)) +
  labs(x = "Time after dose (h)", y = "MPA concentration (mg/L)",
       colour = "Body weight (kg)",
       title = "Typical-value profiles across the cohort weight range",
       caption = paste(
         "Body weight enters Vp/F only, so heavier subjects have a lower peak",
         "and a higher trough at identical AUC0-12."))
```

# PKNCA validation

NCA is run on the study's own eight-point sampling grid, using the
residual-error observations, so that it is subject to the same sampling-density
limitations as Ye 2025 Table 2.

Ye 2025's assay has an LLOQ of 0.1 mg/L. The combined residual error (additive
SD 0.38 mg/L on troughs near 1.2 mg/L) puts a small fraction of simulated
observations below that -- some of them negative, which no assay can report and
which leaves PKNCA unable to define `clast` and therefore `auclast`. Values
below the LLOQ are substituted with LLOQ/2, the conventional handling.

```{r pknca}
LLOQ <- 0.1   # mg/L, Ye 2025 Methods (bioanalytical)

blq_fraction <- mean(obs_study$sim < LLOQ)
round(100 * blq_fraction, 2)   # percent of simulated observations below LLOQ

# Only `!is.na()` in this filter -- adding `time > 0` or `Cc > 0` would drop
# the time-zero row PKNCA needs to anchor AUC0-tau.
sim_nca <- obs_study |>
  dplyr::filter(!is.na(sim)) |>
  dplyr::transmute(id, time, Cc = pmax(sim, LLOQ / 2), treatment)

stopifnot(nrow(sim_nca) == N_SUBJ * length(grid_study))
stopifnot(all(sim_nca$Cc > 0))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")

dose_df <- cohort |> dplyr::mutate(time = 0, amt = DOSE) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start     = 0,
  end       = TAU,
  cmax      = TRUE,
  tmax      = TRUE,
  cmin      = TRUE,
  auclast   = TRUE,
  half.life = TRUE,
  cl.last   = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
```

```{r pknca-auc612}
# AUC6-12 is a second interval, reported separately by Ye 2025 (Table 2) as a
# surrogate for the enterohepatic-recirculation contribution.
nca_612 <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_obj, dose_obj,
  intervals = data.frame(start = 6, end = TAU, auclast = TRUE)
))

auc012 <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD == "auclast") |> dplyr::pull(PPORRES)
auc612 <- as.data.frame(nca_612$result) |>
  dplyr::filter(PPTESTCD == "auclast") |> dplyr::pull(PPORRES)

# Fail loudly rather than propagating NaN into the comparison below.
stopifnot(length(auc012) == N_SUBJ, length(auc612) == N_SUBJ,
          !anyNA(auc012), !anyNA(auc612))
```

## Comparison against published NCA

```{r nca-compare}
# Ye 2025 Table 2, "The pharmacokinetic parameters of MMF" (means).
published <- tibble::tribble(
  ~treatment,         ~cmax, ~tmax, ~cmin, ~auclast, ~half.life,
  "MMF 370 mg q12h",  15.08, 0.86,  0.80,  31.05,    5.68
)

# ncaComparisonTable() summarises a per-subject PKNCAresults by MEDIAN. Ye
# 2025 Table 2 reports means throughout, so the simulated side is
# pre-aggregated by mean and passed in that form (the documented override).
# On an eight-point grid this matters most for Tmax, which is quantised to
# the sampling times: its median snaps to a single grid value while its mean
# does not.
sim_means <- as.data.frame(nca_res$result) |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(PPORRES = mean(PPORRES, na.rm = TRUE), .groups = "drop")

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = sim_means,
  reference     = published,
  by            = "treatment",
  units         = c(cmax = "mg/L", cmin = "mg/L", tmax = "h",
                    auclast = "mg*h/L", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs. Ye 2025 Table 2 NCA, steady state, eight-point sampling",
    "grid. * differs from the reference by more than 20%."
  )
)
```

```{r nca-compare-auc612}
tibble::tibble(
  "NCA parameter" = c("AUC6-12 (mg*h/L)", "AUC6-12 / AUC0-12 (%)"),
  Reference       = c(7.02, 22.61),
  Simulated       = c(mean(auc612), 100 * mean(auc612) / mean(auc012))
) |>
  dplyr::mutate(
    "% diff" = round(100 * (Simulated - Reference) / Reference, 1),
    Reference = round(Reference, 2),
    Simulated = round(Simulated, 2)
  ) |>
  knitr::kable(
    caption = "Late-interval exposure vs. Ye 2025 Table 2 (medians of the simulated cohort)."
  )
```

### What matches, and what does not

```{r nca-assertions}
mean_of <- function(code, res = nca_res) {
  v <- as.data.frame(res$result) |>
    dplyr::filter(PPTESTCD == code) |> dplyr::pull(PPORRES)
  # Fail loudly rather than returning NaN from a zero-row filter, which would
  # make every check below pass vacuously.
  if (length(v) == 0L) stop("no NCA rows for '", code, "'")
  mean(v, na.rm = TRUE)
}

stopifnot(all(c("cmax", "tmax", "auclast", "cl.last") %in%
                as.data.frame(nca_res$result)$PPTESTCD))

checks <- tibble::tibble(
  Quantity  = c("AUC6-12 (mg*h/L)", "Tmax (h)", "AUC0-12 (mg*h/L)",
                "Cmax (mg/L)", "Cmin (mg/L)", "CL/F (L/h)"),
  Reference = c(7.02, 0.86, 31.05, 15.08, 0.80, 11.10),
  Simulated = c(mean(auc612), mean_of("tmax"), mean_of("auclast"),
                mean_of("cmax"), mean_of("cmin"), mean_of("cl.last")),
  Deviation = c(FALSE, FALSE, TRUE, TRUE, TRUE, TRUE)
) |>
  dplyr::mutate(`% diff` = 100 * (Simulated - Reference) / Reference)

checks |>
  dplyr::mutate(dplyr::across(c(Simulated, `% diff`), \(x) round(x, 2))) |>
  knitr::kable(caption = "Deviation = TRUE marks a reproducible disagreement, documented in Errata rather than tuned away.")

passing <- checks |> dplyr::filter(!Deviation)
stopifnot(nrow(passing) == 2L)

# Cohort-derived, so these are magnitude bounds set well outside the
# run-to-run spread rather than the value any single run produced.
# Realised +7.7% here and +14.7% under an earlier explicit-dosing variant,
# so the bound sits at roughly twice the worst value seen.
stopifnot(abs(passing$`% diff`[passing$Quantity == "AUC6-12 (mg*h/L)"]) < 35)
# Tmax is quantised to the eight sampling times, whose neighbours around the
# model's ~1.1 h peak are 0.5 and 1.5 h; one grid step is already -42% / +74%
# against the reported 0.86 h, so no tighter bound is meaningful on this
# design. Realised about +36%. It still goes red if absorption is
# mis-encoded: dropping Tlag or halving ka moves mean Tmax past 2 h.
# Realised about +26%.
stopifnot(abs(passing$`% diff`[passing$Quantity == "Tmax (h)"]) < 70)

# The decisive gate on this vignette's actual claim -- that the packaged
# model reproduces the PUBLISHED model. Steady-state exposure per gram of
# MMF is 1000 / (CL/F) by structural check 1, so this fails proportionally
# on any error in the clearance, the dose or the concentration units.
auc_per_gram_sim   <- 1000 * mean_of("auclast") / DOSE
auc_per_gram_model <- 1000 / CL_F
c(simulated = auc_per_gram_sim, model = auc_per_gram_model,
  `Ye 2025 Table 2` = 80.19)

# Band allows for the sparse eight-point trapezoid under-reading the peak
# and for the LLOQ substitution (realised about -6%); a 25% error in CL/F --
# the size of the Table 2 vs Table 3 conflict -- would break it.
stopifnot(abs(auc_per_gram_sim / auc_per_gram_model - 1) < 0.15)
```

The AUC0-12, Cmax and CL/F rows are starred, and all three are the **same
single discrepancy** seen from three directions. Because steady-state AUC in
this model is exactly `Dose / (CL/F)` (structural check 1), the AUC deviation
is fully determined by the clearance value and nothing else:

```{r cl-inconsistency}
# Ye 2025 reports three mutually inconsistent clearance values.
cl_from_model     <- CL_F                          # Table 3, theta_CL/F
cl_from_nca_mean  <- 11.10                         # Table 2, CL/F
cl_from_auc_dose  <- 1000 / 80.19                  # Table 2, AUC0-12/dose = 80.19 mg*h/L per g

c(`Table 3 theta_CL/F` = cl_from_model,
  `Table 2 CL/F`       = cl_from_nca_mean,
  `implied by Table 2 AUC0-12/dose` = cl_from_auc_dose)

# The simulated AUC deficit equals the clearance excess, to within the
# sampling-grid error -- i.e. the packaged model reproduces Table 3 exactly
# and it is Table 3 that disagrees with Table 2.
auc_ratio_predicted <- cl_from_auc_dose / cl_from_model
auc_ratio_observed  <- mean_of("auclast") / 31.05
c(predicted = auc_ratio_predicted, observed = auc_ratio_observed)

stopifnot(abs(auc_ratio_observed / auc_ratio_predicted - 1) < 0.35)
```

`theta_CL/F` = 15.23 L/h (Table 3) is 22% above the 12.47 L/h implied by the
paper's own dose-normalised AUC (Table 2, AUC0-12/dose = 80.19 mg*h/L per g)
and 37% above the mean NCA CL/F of 11.10 L/h (Table 2). Nothing in the encoding
can reconcile them: the packaged model reproduces Table 3, and Table 3
disagrees with Table 2. The residual Cmax shortfall on top of that is the
absorption-phase misfit Ye 2025 itself lists first among its limitations
("the observed discrepancies in figure 4 may stem from relatively sparse
sampling during the absorption phase"). Tmax (0.86 h reported) and the late
interval AUC6-12 are reproduced, so the discrepancy is confined to the peak
and the total area, not to the shape of the elimination phase.

# Assumptions and deviations

**Covariate functional form.** Ye 2025 online supplemental table 5 prints the
retained relationship as `V2,i = V2,pop * (body weight / 41.13)` with no
exponent, while Table 3 reports a coefficient `theta_v2,weight = 2.05` (RSE
11.99%, bootstrap 95% CI 0.39-2.60). It is encoded here as the power form
`Vp/F = 1287.12 * (WT/41.13)^2.05`. The supplement's equations omit the
estimated coefficient from *every* screened relationship, not just this one --
the genotype rows print `V_i = V_pop * e^(dsDNA = 1, 2 or 3)` with no theta
either -- so the omission is that table's notation, not a claim that the
exponent is 1. (This was confirmed against the raw OOXML of the supplement:
there is no dropped superscript.) A centred-linear reading
`Vp/F = 1287.12 * (1 + 2.05 * (WT - 41.13)/41.13)` was rejected because it
turns Vp/F negative below 21.1 kg, inside the plausible weight range of a
cohort with mean weight 41.13 +/- 11.31 kg. A power exponent is also the only
reading under which the printed ratio form is literally correct at its
special case (exponent 1) and under which a bootstrap CI spanning 1.0 is
natural.

**Which parameters carry a random effect.** Ye 2025 states that "the typical
value of absorption rate constant (Ka), typical value of clearance (CL) and
Tlag were fixed due to their high eta-shrinkage values", explaining that high
shrinkage "suggests that the random effect may be negligible". That reasoning
removes a random effect, not a THETA -- and all three thetas carry non-zero
RSEs and bootstrap medians that differ from the final estimates (1.69 vs 1.54,
16.27 vs 15.23, 0.42 vs 0.40), which a genuinely fixed THETA cannot. The etas
on ka, CL/F and Tlag are therefore encoded as dropped and the three thetas as
estimated, leaving exactly the three `omega^2` rows Table 3 prints.

**Table 3's shrinkage column is internally inconsistent.** The "Final model"
column prints a shrinkage beside `theta_CL/F` (18.30) and none beside
`theta_CL2/F`, while the "Bootstrap" column does the opposite (13.60 beside
CL2/F, none beside CL/F). The IIV rows and the table footnote both name
`omega^2 CL2/F`, and the text says CL's random effect was dropped, so the third
eta is placed on Q/F and the stray "(18.30)" is treated as a mis-set cell.

**Variability scale.** Table 3 labels the IIV rows `omega^2` and Phoenix NLME
reports the Omega matrix of an exponential IIV model as variances, so 1.02,
1.24 and 0.20 are used directly as eta variances. Reading them instead as
standard deviations would change the Vc variance by 2% and the Vp variance by
24%; it does not change any conclusion in this vignette, because the checks
that pass or fail here are driven by CL/F, which carries no eta.

**Absorption rate constant units.** Table 3 heads the row `theta_Ka (hours)`,
but ka is a first-order rate constant and the Discussion compares it with
literature values quoted "from 0.39 to 5.21 hours^-1". The header unit is read
as a typo for 1/h and the value 1.54 is used unchanged.

**Dose and molecular weight.** Doses are MMF mass in mg. Ye 2025 applies no
MMF-to-MPA molecular-weight conversion, so CL/F, Vc/F, Vp/F and Q/F are
apparent parameters absorbing both the molecular-weight ratio and oral
bioavailability; no `f(depot)` term is applied. The simulations use a single
dose level of 369.9 mg (half the reported mean total daily dose of 739.81
mg/day) rather than the reported 125-750 mg range, so that the comparison
against Ye 2025 Table 2 is not confounded by a dose distribution the paper
does not tabulate.

**Below-LLOQ handling.** Ye 2025 states an LLOQ of 0.1 mg/L but no BLQ rule.
The combined residual error puts a small percentage of simulated observations
below it, some of them negative; negative concentrations are unobservable and
leave PKNCA unable to define `clast`, hence `auclast`. Simulated values below
the LLOQ are replaced with LLOQ/2, the conventional substitution. This affects
Cmin materially (it is the statistic most exposed to the additive error near
the trough) and AUC0-12 only marginally.

**Body-weight distribution.** Ye 2025 Table 1 gives only mean +/- SD
(41.13 +/- 11.31 kg). The virtual cohort draws from a normal distribution with
those moments, truncated to 18-75 kg; neither the shape nor the truncation is
stated by the paper.

**Sampling schedule.** The Methods list pre-dose plus 0.5, 1.5, 2.5, 4, 6, 9
and 12 hours (eight samples), which the paper's own totals corroborate
(1170 samples / 146 profiles = 8.01). Figure 1's x-axis additionally carries a
1.0 h tick; the eight-point Methods schedule is used here.

**Steady state.** Simulations use `ss = 1` (exact steady state). Ye 2025
required a minimum of 7 days of therapy, which is only about two terminal
half-lives of this model, so the earliest-sampled patients would not have been
fully at steady state; the mean treatment duration of 48.66 days is about 14
terminal half-lives.

**Known deviations, not tuned.** Simulated AUC0-12, Cmax and CL/F differ from
Ye 2025 Table 2 by more than 20% and are excluded from the assertion gate but
kept visible in the comparison table above. The cause is a conflict internal to
the paper -- `theta_CL/F` = 15.23 L/h in Table 3 against 11.10 L/h in Table 2
and 12.47 L/h implied by Table 2's own dose-normalised AUC -- compounded by the
absorption-phase misfit the authors list as their first limitation. No
parameter was adjusted to close it.

**Supplemental table 5 reporting errors (not used by this model).** The
stepwise screening table reports an OFV that *increases* at every forward
inclusion step (base 4659.74, rising to 4666.72 when body weight on V2 is
added), which is the wrong direction for a likelihood-ratio inclusion, and it
holds the structural-parameter count at 6 for the covariate models as well as
the base model. Its "WBC" row also prints the albumin equation. These affect
the audit trail of the covariate search only; the final model parameters come
from Table 3 and are unaffected.

**Pharmacogenetics is not in the model.** Eight variants in *UGT1A9*, *UGT2B7*,
*ABCC2* and *CES1* were associated with dose-normalised MPA exposure in the
univariate analysis (online supplemental table 4), but none survived the
stepwise popPK covariate search, so no genetic covariate appears in the model
file. Ye 2025 says so explicitly: "the effect size may not be sufficient to
replace the covariate status of body weight in the population model."

**Enterohepatic recirculation is absent.** Ye 2025 could not fit an EHC
component (no samples around the second peak, few patients with a pronounced
EHC process), so the model has no EHC pathway and will not reproduce a
secondary peak.
