---
title: "Ceftazidime (Yin 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ceftazidime (Yin 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Yin_2026_ceftazidime"))$reference`
* Description: `r rxode2::rxode(readModelDb("Yin_2026_ceftazidime"))$description`
* Article: <https://doi.org/10.1128/aac.01810-25>

No supplementary material is deposited for this article, and no erratum or
corrigendum is recorded against the DOI (Crossref `update-to` / `updated-by`
both empty as of the extraction date). Every value below therefore comes from
the main text, Table 1 or Table 2.

# Population

Yin 2026 is a prospective, single-centre, open-label population PK study run at
Xiamen Maternity and Child Health Care Hospital between 2022 and 2024. Seventy-two
Chinese neonates with confirmed or suspected bacterial infection requiring
intravenous ceftazidime contributed 140 concentration measurements (median 2 per
patient, range 1-4). Postmenstrual age spanned 32.7-41.9 weeks (median 39.7),
gestational age 32.6-41.3 weeks (median 39.4), postnatal age 1-4 days (median 1),
and body weight 1.8-4.2 kg (median 3.2). Fourteen of the 72 (19.4%) were preterm
births, 29 (40.3%) were female, and serum creatinine ranged 23.7-88.0 umol/L
(median 56.7). Baseline characteristics are Table 1 of the source.

The distinguishing methodological feature is the sampling: capillary blood was
taken from the heel with a Capitainer quantitative dried blood spot (qDBS) device
at 0-0.5, 2-3 and 4-6 h after a dose, a metered 10 uL aliquot per card. **The
model was fitted directly to whole-blood concentrations**, so the packaged `CL`
and `V` are blood-referenced. This matters for any comparison with the many
plasma-based neonatal ceftazidime models, and is picked up again under
*Assumptions and deviations* below.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
pop <- rxode2::rxode(readModelDb("Yin_2026_ceftazidime"))$population
str(pop, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Yin_2026_ceftazidime.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (typical blood CL at 70 kg, full maturation) | `log(19.6)` L/h | Table 2, "CL (L/h/70 kg)" = 19.6 (RSE 4.2%; bootstrap 19.5, 95% CI 17.9-21.2) |
| `lvc` (typical blood V at 70 kg) | `log(70.6)` L | Table 2, "V (L/70 kg)" = 70.6 (RSE 7.5%; bootstrap 71.3, 95% CI 61.3-82.8) |
| `e_wt_cl` (allometric exponent on CL) | `fixed(0.75)` | Methods "PopPK modeling" / equation 1; fixed a priori, absent from Table 2 |
| `e_wt_vc` (allometric exponent on V) | `fixed(1.00)` | Methods "PopPK modeling" / equation 2; fixed a priori, absent from Table 2 |
| `tmat50` (PMA at 50% mature CL) | `fixed(47.7)` weeks | Methods "PopPK modeling" and Discussion; Rhodin et al. 2009 GFR model (source reference 25) |
| `hill_mat` (maturation Hill coefficient) | `fixed(3.40)` | Methods "PopPK modeling" and Discussion; Rhodin et al. 2009 (source reference 25) |
| `etalcl` (IIV on CL) | `0.0760` | Table 2, IIV CL = 28.1% CV; footnote c gives CV% = sqrt(exp(omega^2) - 1) x 100, so omega^2 = log(1 + 0.281^2) |
| `etalvc` (IIV on V) | `0.1653` | Table 2, IIV V = 42.4% CV; same inversion, omega^2 = log(1 + 0.424^2) |
| `propSd` (residual error) | `0.223` | Table 2, "Proportional error (%)" = 22.3 (RSE 45%; bootstrap 22.2, 95% CI 13.1-32.1) |
| `fmat <- PAGE^hill_mat / (PAGE^hill_mat + tmat50^hill_mat)` | n/a | Equation 1 (sigmoidal maturation term) |
| `cl <- exp(lcl + etalcl) * (WT / 70)^e_wt_cl * fmat` | n/a | Equation 1 |
| `vc <- exp(lvc + etalvc) * (WT / 70)^e_wt_vc` | n/a | Equation 2 |
| `d/dt(central) <- -kel * central` | n/a | Results "PopPK model building": "The concentration-time data were best described by a one-compartment model" |
| `Cc ~ prop(propSd)` | n/a | Table 2 footnote a (exponential error, reported and interpreted as proportional in the normal domain) |
| Reference weight 70 kg | n/a | Equations 1-2, "WT / 70 kg"; Table 2 parameter units |

The two blood-to-plasma conversion constants that appear in the paper's
simulation section (equations 3-5) are deliberately **not** in the model file;
see *Assumptions and deviations*.

# Typical-value reproduction

The paper states the typical blood CL and V standardised to 70 kg, and separately
quotes derived per-kilogram values at the cohort medians. The first check is that
the packaged model, with random effects zeroed, reproduces the standardised
values exactly and the derived blood values arithmetically. These are
deterministic quantities -- no cohort is drawn -- so the tolerances here are tight
on purpose.

```{r typical}
mod <- readModelDb("Yin_2026_ceftazidime")
ui  <- rxode2::rxode(mod)

# Median-neonate reference used by the paper's own comparison, from Table 1.
wt_med  <- 3.2   # kg
pma_med <- 39.7  # weeks (Table 1 median postmenstrual age)

fmat_med <- pma_med^3.4 / (pma_med^3.4 + 47.7^3.4)
cl_med   <- 19.6 * (wt_med / 70)^0.75 * fmat_med
v_med    <- 70.6 * (wt_med / 70)^1

typical <- tibble::tibble(
  Quantity = c(
    "CL standardised to 70 kg, fully mature (L/h)",
    "V standardised to 70 kg (L)",
    "Maturation fraction at PMA 39.7 weeks",
    "Blood CL at the median neonate (L/h)",
    "Blood CL at the median neonate (L/h/kg)",
    "Blood V at the median neonate (L)",
    "Blood V at the median neonate (L/kg)",
    "Terminal half-life at the median neonate (h)"
  ),
  Value = c(19.6, 70.6, fmat_med, cl_med, cl_med / wt_med,
            v_med, v_med / wt_med, log(2) * v_med / cl_med)
)

knitr::kable(typical, digits = 4,
             caption = "Typical-value quantities implied by Yin 2026 Table 2 and equations 1-2.")

# The packaged model must return exactly these values when the etas are zeroed.
ev_typ <- data.frame(
  id = 1L, time = c(0, 0.5), amt = c(25 * wt_med, NA), evid = c(1L, 0L),
  cmt = "central", dur = c(0.5, NA), WT = wt_med, PAGE = pma_med
)
sim_typ <- rxode2::rxSolve(rxode2::zeroRe(mod), events = ev_typ,
                           omega = NA, sigma = NA) |>
  as.data.frame()

stopifnot(
  # Deterministic: the model's own cl / vc against the hand-evaluated equations.
  abs(unique(sim_typ$cl) - cl_med) < 1e-9,
  abs(unique(sim_typ$vc) - v_med)  < 1e-9
)
```

# Virtual cohort

Original observed data are not publicly available, and the paper does not publish
the joint distribution of postmenstrual age and body weight (only the marginals of
Table 1 and the statement that the simulation cohort was drawn from the hospital's
neonatal database). The cohorts below therefore sample PMA uniformly over the
observed 32.7-41.9 week range and draw weight from a log-normal centred on a
linear PMA-weight trend anchored on the two Table 1 medians and the lower end of
the observed weight range. That is an explicit assumption; see *Assumptions and
deviations*.

Cohort sizes are 200 per arm.

```{r cohort}
# `set.seed()` seeds R's RNG (the covariate draw). It does NOT seed rxode2's
# simulation RNG, and rxode2's streams are partitioned per solver thread, so the
# eta draw differs between a 2-core CI runner and a many-thread workstation.
# Every assertion below is therefore written to hold for any cohort the model
# can produce.
set.seed(20260420)
rxode2::rxSetSeed(20260420)

n_arm <- 200L

# PMA-weight relation: 3.2 kg at the median PMA of 39.7 weeks, sloping to
# ~1.8 kg at the lowest observed PMA of 32.7 weeks (0.2 kg per week), with a
# 12% CV log-normal residual and truncation to the plausible neonatal range.
draw_subjects <- function(n, pma_lo, pma_hi, id_offset = 0L) {
  pma <- runif(n, pma_lo, pma_hi)
  wt  <- (3.2 + 0.2 * (pma - 39.7)) * exp(rnorm(n, 0, 0.12))
  tibble(
    id   = id_offset + seq_len(n),
    PAGE = pma,
    WT   = pmin(pmax(wt, 1.5), 4.5)
  )
}

# Build a dosing + observation event table for one regimen.
make_arm <- function(subj, mg_per_kg, tau, n_dose, grid = 0.05, dur = 0.5,
                     label) {
  dosing <- subj |>
    tidyr::crossing(tibble(.dose = seq_len(n_dose))) |>
    mutate(time = (.dose - 1) * tau, amt = mg_per_kg * WT, evid = 1L,
           cmt = "central", dur = dur) |>
    select(-.dose)
  obs <- subj |>
    tidyr::crossing(tibble(time = seq(0, n_dose * tau, by = grid))) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dur = NA_real_)
  bind_rows(dosing, obs) |>
    mutate(regimen = label) |>
    arrange(id, time, desc(evid))
}
```

The `cmt` on every observation row is `"central"` -- the ODE state -- never the
algebraic observable `Cc`; rxode2 returns `Cc` as an output column regardless.

## Cohort A -- single dose, structural checks

```{r cohort-a}
subj_a <- draw_subjects(n_arm, 32.7, 41.9, id_offset = 0L)
ev_a   <- make_arm(subj_a, mg_per_kg = 25, tau = 24, n_dose = 1L,
                   grid = 0.05, label = "25 mg/kg single dose")
stopifnot(!anyDuplicated(unique(ev_a[, c("id", "time", "evid")])))

sim_a <- rxode2::rxSolve(mod, events = as.data.frame(ev_a),
                         keep = c("WT", "PAGE", "regimen")) |>
  as.data.frame()
stopifnot(all(sim_a$Cc[!is.na(sim_a$Cc)] >= 0))
```

## Cohort B -- steady-state regimens from the paper's simulation grid

Five of the twelve regimens Yin 2026 simulated are enough to pin down the
paper's Figure 4 conclusions at the *Escherichia coli* breakpoint of 4 mg/L:
25 mg/kg every 6 h and every 8 h attain the target, 25 and 50 mg/kg every 12 h do
not, and 75 mg/kg every 12 h does. Each arm is dosed to steady state (the median
terminal half-life is ~3.3 h, so 24 h is several half-lives) and evaluated over
its final dosing interval.

```{r cohort-b}
regimens <- tibble::tribble(
  ~label,             ~mg_per_kg, ~tau, ~n_dose,
  "25 mg/kg q6h",     25,          6,    4L,
  "25 mg/kg q8h",     25,          8,    3L,
  "25 mg/kg q12h",    25,          12,   2L,
  "50 mg/kg q12h",    50,          12,   2L,
  "75 mg/kg q12h",    75,          12,   2L
)

ev_b <- do.call(bind_rows, lapply(seq_len(nrow(regimens)), function(i) {
  r <- regimens[i, ]
  subj <- draw_subjects(n_arm, 32.7, 41.9, id_offset = 1000L * i)
  make_arm(subj, r$mg_per_kg, r$tau, r$n_dose, grid = 0.05, label = r$label)
}))
stopifnot(!anyDuplicated(unique(ev_b[, c("id", "time", "evid")])))

sim_b <- rxode2::rxSolve(mod, events = as.data.frame(ev_b),
                         keep = c("WT", "PAGE", "regimen")) |>
  as.data.frame()
stopifnot(all(sim_b$Cc[!is.na(sim_b$Cc)] >= 0))
```

# Replicate published figures

```{r figure-1}
# Companion to Figure 1 of Yin 2026 (observed concentration vs time after dose,
# 0-6 h), and to the Figure 3 VPC: simulated whole-blood concentration after a
# single 25 mg/kg 30-minute infusion, over the paper's 0-6 h sampling window.
sim_a |>
  filter(time <= 6) |>
  group_by(time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  labs(x = "Time after dose (h)", y = "Whole-blood ceftazidime (mg/L)",
       title = "Simulated 5th / 50th / 95th percentiles, 25 mg/kg over 30 min",
       caption = paste("Companion to Figure 1 (observed scatter) and Figure 3",
                       "(VPC) of Yin 2026; the paper's qDBS samples were drawn",
                       "at 0-0.5, 2-3 and 4-6 h."))
```

```{r figure-maturation}
# The model's central claim (Results / Discussion): PMA drives clearance through
# the fixed Rhodin sigmoid, weight drives volume. Reproduces the shape of the
# maturation function the paper adopts (equation 1).
tibble(PAGE = seq(24, 60, by = 0.25)) |>
  mutate(fmat = PAGE^3.4 / (PAGE^3.4 + 47.7^3.4)) |>
  ggplot(aes(PAGE, fmat)) +
  geom_line() +
  geom_vline(xintercept = 47.7, linetype = "dashed") +
  annotate("rect", xmin = 32.7, xmax = 41.9, ymin = -Inf, ymax = Inf, alpha = 0.12) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Postmenstrual age (weeks)", y = "Fraction of mature clearance",
       title = "Renal maturation function (Yin 2026 equation 1)",
       caption = paste("TM50 = 47.7 weeks (dashed) and Hill = 3.4, both fixed to",
                       "Rhodin et al. 2009. Shaded band is the observed PMA range."))
```

```{r figure-ss}
# Steady-state whole-blood profiles over the final dosing interval, by regimen.
last_interval <- regimens |>
  mutate(t_start = (n_dose - 1) * tau, t_end = n_dose * tau) |>
  select(regimen = label, tau, t_start, t_end)

ss <- sim_b |>
  inner_join(last_interval, by = "regimen") |>
  filter(time >= t_start, time <= t_end) |>
  mutate(tad = time - t_start)

ss |>
  group_by(regimen, tad) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~regimen, scales = "free_x") +
  scale_y_log10() +
  labs(x = "Time since last dose (h)", y = "Whole-blood ceftazidime (mg/L)",
       title = "Steady-state profiles over the final dosing interval",
       caption = "Regimens from the simulation grid of Yin 2026 Methods / Figure 4.")
```

# PKNCA validation

The paper reports no NCA parameters, so the NCA here does two things instead:
it exercises the standard PKNCA pathway on the packaged model, and it provides
closed-form identity checks. For a one-compartment intravenous model, NCA on a
noise-free single-dose simulation must return `AUC(0-inf) = Dose / CL` and
`t1/2 = log(2) * V / CL` **for each individual subject**, using that subject's
own drawn `cl` and `vc`. Both sides of these comparisons use the same drawn
parameters, so the residual is pure numerical (trapezoid + lambda-z) error and
the tolerances are tight.

```{r pknca}
# Only `!is.na(Cc)` -- never a `time > 0` or `Cc > 0` filter, which would drop
# the time-zero anchor row that PKNCA needs for AUC(0-*).
sim_nca <- sim_a |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, regimen)

sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, regimen) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, regimen, time, .keep_all = TRUE) |>
  arrange(id, regimen, time)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | regimen + id)

dose_df <- ev_a |>
  filter(evid == 1) |>
  select(id, time, amt, regimen) |>
  as.data.frame()
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | regimen + id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))
nca_tbl <- as.data.frame(nca_res$result) |>
  select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)
```

```{r pknca-identity}
truth <- sim_a |>
  group_by(id) |>
  summarise(cl = first(cl), vc = first(vc), .groups = "drop") |>
  left_join(dose_df |> select(id, amt), by = "id") |>
  mutate(
    auc_theory      = amt / cl,
    halflife_theory = log(2) * vc / cl
  )

chk <- nca_tbl |>
  left_join(truth, by = "id") |>
  mutate(
    auc_pct  = 100 * (aucinf.obs - auc_theory) / auc_theory,
    hl_pct   = 100 * (half.life - halflife_theory) / halflife_theory,
    # PKNCA's cl.obs is dose / AUC, i.e. in dose-over-concentration units,
    # which for mg and mg/L is L/h -- directly comparable to the model's cl.
    clobs_pct = 100 * (cl.obs - cl) / cl
  )

identity_tbl <- tibble::tibble(
  Check = c("AUC(0-inf) vs Dose / CL",
            "Terminal half-life vs log(2) * V / CL",
            "PKNCA CL(obs) vs individual CL"),
  `Median % difference` = c(median(chk$auc_pct), median(chk$hl_pct),
                            median(chk$clobs_pct)),
  `Max abs % difference` = c(max(abs(chk$auc_pct)), max(abs(chk$hl_pct)),
                             max(abs(chk$clobs_pct)))
)

knitr::kable(identity_tbl, digits = 4,
             caption = paste("Closed-form identity checks. Both sides use the",
                             "same drawn parameters, so any difference is",
                             "numerical error in the NCA, not model error."))

# Same drawn parameters on both sides -> pure trapezoid / lambda-z error, so a
# tight all() bound is the correct assertion here (it is NOT a cohort-extreme
# assertion of the kind that varies between rxode2 builds).
stopifnot(
  max(abs(chk$auc_pct))   < 1.0,
  max(abs(chk$hl_pct))    < 1.0,
  max(abs(chk$clobs_pct)) < 1.0
)
```

## NCA summary by regimen

```{r pknca-ss}
ss_intervals <- last_interval |>
  transmute(start = t_start, end = t_end,
            cmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE) |>
  as.data.frame()

conc_ss <- sim_b |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, regimen) |>
  as.data.frame()

dose_ss <- ev_b |>
  filter(evid == 1) |>
  select(id, time, amt, regimen) |>
  as.data.frame()

nca_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc_ss, Cc ~ time | regimen + id),
  PKNCA::PKNCAdose(dose_ss, amt ~ time | regimen + id),
  intervals = bind_cols(ss_intervals, tibble(regimen = last_interval$regimen))
))

as.data.frame(nca_ss$result) |>
  filter(PPTESTCD %in% c("cmax", "cmin", "cav", "auclast")) |>
  group_by(regimen, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  dplyr::rename(
    "Regimen"                    = regimen,
    "AUC(tau) (mg*h/L)"          = auclast,
    "Cavg,ss (mg/L)"             = cav,
    "Cmax,ss (mg/L)"             = cmax,
    "Cmin,ss (mg/L)"             = cmin
  ) |>
  knitr::kable(digits = 2,
               caption = paste("Median steady-state whole-blood NCA over the",
                               "final dosing interval, by regimen."))
```

# Probability of target attainment

Yin 2026 evaluates 70% *f*T > MIC on **unbound plasma** concentrations, obtained
from the model's whole-blood predictions by equations 3-5:

$$C_{\text{plasma,unbound}} = f_u \times BP \times C_{\text{blood}}, \qquad
BP = C_{\text{plasma}} / C_{\text{blood}} = 0.72$$

The unbound fraction \(f_u\) is never given a numeric value anywhere in the
paper, so the target-attainment analysis cannot be reproduced exactly. It can be
*bounded*, which is enough to check the transcription. Write
\(\kappa = f_u \times BP\); then "unbound plasma above MIC" is identical to
"whole blood above MIC/\(\kappa\)", and because an unbound fraction cannot exceed
1, \(\kappa \le BP = 0.72\). Evaluating at \(\kappa = 0.72\) is therefore the
**most favourable admissible case** -- any real \(f_u < 1\) can only lower target
attainment.

```{r pta}
mic <- 4  # mg/L, the EUCAST E. coli breakpoint the paper highlights
bp  <- 0.72  # blood-to-plasma ratio as defined and used by Yin 2026 (reference 26)

# Fraction of the final dosing interval spent above a whole-blood threshold,
# per subject. The 0.05 h observation grid resolves the interval to <1%.
pta_at <- function(df, kappa, mic_value = mic) {
  df |>
    group_by(regimen, id) |>
    summarise(ft = mean(Cc > mic_value / kappa), .groups = "drop") |>
    group_by(regimen) |>
    summarise(pta = mean(ft >= 0.70), .groups = "drop")
}

# Yin 2026 Results "Model-based simulations" and Discussion, at MIC 4 mg/L:
# 25 mg/kg q6h and q8h attain PTA >= 90%; at a 12-h interval only the 75 and
# 100 mg/kg regimens do, so 25 and 50 mg/kg q12h must fall below 90%.
published_claim <- tibble::tribble(
  ~regimen,          ~attains,
  "25 mg/kg q6h",     TRUE,
  "25 mg/kg q8h",     TRUE,
  "25 mg/kg q12h",    FALSE,
  "50 mg/kg q12h",    FALSE,
  "75 mg/kg q12h",    TRUE
)

pta_max <- pta_at(ss, kappa = bp) |>
  inner_join(published_claim, by = "regimen") |>
  mutate(
    simulated = pta >= 0.90,
    agrees    = simulated == attains
  )

# Guard against a silently empty comparison (pattern 10): every published claim
# must have found a matching simulated arm.
stopifnot(nrow(pta_max) == nrow(published_claim))

pta_max |>
  transmute(
    Regimen = regimen,
    `PTA at kappa = 0.72` = sprintf("%.1f%%", 100 * pta),
    `Simulated: attains >= 90%?` = ifelse(simulated, "yes", "no"),
    `Yin 2026 Figure 4` = ifelse(attains, "yes", "no"),
    Agrees = ifelse(agrees, "yes", "no")
  ) |>
  knitr::kable(
    caption = paste("Target attainment for 70% fT > MIC at MIC = 4 mg/L,",
                    "evaluated at the largest admissible unbound scaling",
                    "(kappa = BP = 0.72, i.e. f_u = 1)."))
```

```{r pta-assertions}
get_pta <- function(reg) {
  v <- pta_max$pta[pta_max$regimen == reg]
  if (length(v) != 1L) stop("no unique PTA row for '", reg, "'")
  v
}

# These are cohort-derived, so only structural relations and wide-margin
# classifications are asserted (see known-vignette-failure-patterns pattern 12).
# Realised over four cohort scenarios (two seeds x two PMA-weight relations)
# and four solver-thread counts (1 / 2 / 4 / 8):
#   25 q6h  0.975-1.000   25 q8h  0.875-0.895   25 q12h 0.480-0.550
#   50 q12h 0.785-0.840   75 q12h 0.885-0.940
# 75 mg/kg q12h straddles the 90% line across those runs, which is precisely
# why its classification is reported but not asserted.
stopifnot(
  # Shortening the interval at a fixed dose must raise fT > MIC. Structural,
  # and the realised gaps (>= 0.09 and >= 0.32) are far outside sampling noise.
  get_pta("25 mg/kg q6h")  > get_pta("25 mg/kg q8h"),
  get_pta("25 mg/kg q8h")  > get_pta("25 mg/kg q12h"),
  # Raising the dose at a fixed interval must raise fT > MIC.
  get_pta("75 mg/kg q12h") > get_pta("50 mg/kg q12h"),
  get_pta("50 mg/kg q12h") > get_pta("25 mg/kg q12h"),
  # The paper's two clear-cut negatives at MIC 4 mg/L, with wide margins.
  # Both are cohort *means* over n_arm subjects, not cohort extremes, so they
  # are far steadier across rxode2 builds than a max()/min() would be; the
  # bounds sit 0.20 and 0.06 above the realised maxima quoted above.
  get_pta("25 mg/kg q12h") < 0.75,
  get_pta("50 mg/kg q12h") < 0.90,
  # The paper's clearest positive.
  get_pta("25 mg/kg q6h")  >= 0.90,
  # A mis-transcribed CL, V, dose or maturation term moves these PTAs by tens
  # of points; every arm must stay in a physically sensible band.
  all(pta_max$pta > 0.30), all(pta_max$pta <= 1.00)
)
```

At the most favourable admissible scaling the model reproduces the paper's
classification for `r sum(pta_max$agrees)` of the `r nrow(pta_max)` regimens.
The disagreements, where present, are the two regimens the paper places just
above the 90% line: see *Assumptions and deviations*.

```{r pta-figure}
kappa_grid <- seq(0.10, 1.00, by = 0.01)
pta_by_kappa <- do.call(bind_rows, lapply(kappa_grid, function(k) {
  pta_at(ss, kappa = k) |> mutate(kappa = k)
}))

pta_by_kappa |>
  ggplot(aes(kappa, pta, colour = regimen)) +
  annotate("rect", xmin = bp, xmax = Inf, ymin = -Inf, ymax = Inf, alpha = 0.10) +
  geom_line() +
  geom_hline(yintercept = 0.90, linetype = "dashed") +
  geom_vline(xintercept = bp, linetype = "dotted") +
  scale_y_continuous(labels = scales::percent) +
  labs(x = expression(kappa == f[u] %*% BP),
       y = "PTA for 70% fT > MIC at MIC = 4 mg/L",
       colour = "Regimen",
       title = "Target attainment vs. the unreported unbound scaling",
       caption = paste("Dashed line: the paper's 90% PTA criterion. Dotted line",
                       "and shading: kappa = BP = 0.72, above which f_u would",
                       "have to exceed 1 and which is therefore inadmissible."))
```

```{r pta-mic-curve}
# The shape of Figure 4, evaluated at the same maximal admissible kappa.
mic_grid <- c(0.25, 0.5, 1, 2, 4, 8, 16, 32)
pta_mic <- do.call(bind_rows, lapply(mic_grid, function(m) {
  pta_at(ss, kappa = bp, mic_value = m) |> mutate(MIC = m)
}))

pta_mic |>
  ggplot(aes(MIC, pta, colour = regimen)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 0.90, linetype = "dashed") +
  scale_x_log10(breaks = mic_grid) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "MIC (mg/L)", y = "PTA for 70% fT > MIC", colour = "Regimen",
       title = "Replicates the shape of Figure 4 of Yin 2026",
       caption = paste("Evaluated at kappa = BP = 0.72. The paper's own",
                       "conclusion that PTA falls away sharply above MIC 4-8",
                       "mg/L is reproduced."))

# The paper's Figure 4 shows PTA collapsing as MIC rises; assert the trend, not
# step-by-step monotonicity or an exact value.
trend <- pta_mic |>
  group_by(regimen) |>
  summarise(lo = pta[MIC == 0.25], hi = pta[MIC == 32], .groups = "drop")
# Realised at 1 / 4 / 8 solver threads: lo 0.99-1.00 for every arm, hi 0.000
# for four arms and up to 0.015 for 75 mg/kg q12h. The bounds below sit well
# outside that range while still going red if a dose, volume or clearance were
# mis-transcribed (which moves these by tens of points).
stopifnot(all(trend$lo > trend$hi), all(trend$lo >= 0.85), all(trend$hi < 0.50))
```

# Assumptions and deviations

* **The model is whole-blood referenced, not plasma.** Yin 2026 fitted the qDBS
  whole-blood concentrations directly (Discussion: "Whole-blood concentrations
  obtained from qDBS samples were used directly to develop the ceftazidime PopPK
  model"). `Cc` from this model is therefore a *whole-blood* ceftazidime
  concentration. Comparing it with a plasma-based neonatal ceftazidime model
  requires the paper's equations 3-5.

* **The unbound fraction `f_u` is never reported.** Equations 4 and 5 both
  contain it and no numeric value appears in the text, tables, or figure
  captions; there is no supplement. The blood-to-plasma and unbound conversions
  are consequently **not** encoded in the model file, and the target-attainment
  section above brackets \(\kappa = f_u \times BP\) by its physical ceiling
  \(\kappa \le BP = 0.72\) rather than substituting a literature value for
  `f_u`.

* **Known deviation: two of the paper's five PTA classifications sit on the
  boundary rather than reproducing cleanly.** Evaluated at \(\kappa = 0.72\) --
  the largest value an unbound fraction can produce, and so the most favourable
  admissible case -- the packaged model reproduces the paper's Figure 4
  conclusions at MIC 4 mg/L unambiguously for 25 mg/kg q6h (attains; realised
  97.5-100%) and for 25 and 50 mg/kg q12h (do not attain; 48-55% and 78.5-84%).
  The two remaining arms sit on the 90% line: 25 mg/kg q8h realised 87.5-89.5%
  and 75 mg/kg q12h realised 88.5-94.0% over four cohort scenarios (two seeds
  crossed with two PMA-weight relations) and four solver-thread counts, so
  75 mg/kg q12h agrees with the paper in some runs and not others. Reproducing
  the full published classification in every run would require
  \(\kappa \approx 0.76\)-\(0.81\), i.e. \(f_u \approx 1.06\)-\(1.12\), which is
  not physically possible. Three candidate explanations, none resolvable from
  the published material: `f_u` itself is unreported so \(\kappa\) can only be
  bounded, not set; the simulation cohort's joint PMA-weight distribution is not
  published and is approximated here (see the bullet below); and the paper does
  not state whether its target-attainment analysis was run at steady state or
  after a single dose -- the steady-state reading used here is the more
  favourable of the two. The discrepancy is a few percentage points and the
  *ordering* of all five regimens is reproduced exactly, so this is recorded as
  a deviation rather than a transcription error. The assertions in the PTA
  section deliberately gate on the structural orderings and the wide-margin
  classifications, never on the two borderline arms.

* **The paper's derived per-kilogram plasma values do not reconcile with its own
  Table 2.** The Discussion states "The plasma CL predicted by our model
  (0.49 L/h/kg)" and "current study: 2.32 L/kg", after "standardizing covariates
  to median values (PMA = 39.4 weeks, PNA = 1 day, and WT = 3.2 kg)" and applying
  the blood-to-plasma ratio 0.72. Evaluating equations 1 and 2 with Table 2's
  estimates gives blood CL = 0.211 L/h/kg and blood V = 1.009 L/kg at the cohort
  medians; converting with BP = 0.72 as described gives plasma CL = 0.293 L/h/kg
  and plasma V = 1.401 L/kg -- not 0.49 and 2.32. Both published figures instead
  require a common additional factor of about 2.30 on top of the blood values
  (2.32 / 1.009 = 2.30; 0.49 / 0.211 = 2.32), i.e. an extra ~1.66-fold beyond
  BP = 0.72. Because the same factor applies to clearance and to volume, it is a
  concentration-scale conversion rather than an error in either parameter, and
  it does not touch the packaged model, whose values are Table 2 verbatim. The
  check is reproduced below. Two related observations: the Discussion's
  "PMA = 39.4 weeks" is the Table 1 median *gestational* age; the Table 1 median
  postmenstrual age is 39.7 weeks (which is the value that makes the arithmetic
  above closest to internally consistent).

    ```{r derived-check}
    derived <- tibble::tibble(
      Quantity = c("Blood CL (L/h/kg)", "Blood V (L/kg)"),
      `From Table 2 + equations 1-2` = c(cl_med / wt_med, v_med / wt_med),
      `Converted with BP = 0.72`     = c(cl_med / wt_med / 0.72, v_med / wt_med / 0.72),
      `Stated in the Discussion`     = c(0.49, 2.32)
    ) |>
      mutate(`Implied extra factor` = `Stated in the Discussion` /
               `From Table 2 + equations 1-2`)

    knitr::kable(derived, digits = 3,
                 caption = paste("The Discussion's per-kilogram plasma values",
                                 "require a common ~2.3-fold factor on the",
                                 "blood-based Table 2 estimates."))

    # Deterministic arithmetic on fixed published constants -- no cohort is
    # drawn, so this bound is exact rather than noise-tolerant.
    stopifnot(abs(derived$`Implied extra factor`[2] - 2.30) < 0.02)
    ```

* **Residual error encoded as proportional, not log-normal.** The Methods say
  residual variability "was also modeled exponentially", but Table 2 reports the
  estimate on a row labelled "Proportional error (%)" and its footnote a states
  that the exponential (log-additive) form "approximates to a proportional error
  in the normal domain". The model therefore uses `Cc ~ prop(propSd)` with
  `propSd = 0.223`, matching the units and interpretation the paper itself
  assigns. At a 22.3% CV the difference from a strict `lnorm()` encoding is
  small (a log-normal with SD 0.223 on the log scale has a CV of
  `r sprintf("%.1f%%", 100 * sqrt(exp(0.223^2) - 1))`), but it is not zero, and a
  re-fit would not land on exactly the published number.

* **IIV variances back-transformed from the reported CV%.** Table 2 reports IIV
  as percent CV and its footnote c gives the conversion
  `CV% = sqrt(exp(omega^2) - 1) x 100`. The `ini()` block inverts it,
  `omega^2 = log(1 + CV^2)`, giving 0.0760 for CL and 0.1653 for V. No CL-V
  covariance is reported, so the etas are encoded as independent.

* **Covariate joint distribution assumed.** The paper publishes marginal PMA and
  weight distributions (Table 1) but not their joint distribution, and states only
  that the Monte Carlo cohort came from the hospital's neonatal database. The
  virtual cohort here draws PMA uniformly over the observed 32.7-41.9 week range
  and weight from a log-normal (12% CV) around a 0.2 kg/week linear trend anchored
  at the Table 1 medians. Neither the PMA distribution nor the correlation is
  the paper's, so absolute PTA values are illustrative; the \(\kappa\) inversion
  above is the part that is comparable to the source.

* **PMA is declared in weeks, not the register-default months.** The canonical
  `PAGE` covariate is registered in months, but its register entry explicitly
  allows a model to declare weeks when its source equations are written in weeks.
  Yin 2026 equation 1 and its fixed constants (TM50 = 47.7 weeks, Hill = 3.4) are
  only meaningful on the week scale, so `covariateData$PAGE$units` is `"weeks"`.
  Read that field before supplying the column.

* **Covariates screened but not retained** (gestational age, postnatal age, serum
  creatinine) are recorded in the model file's `covariatesDataExcluded` list, not
  in `covariateData`, because they are correctly absent from `model()`.

* **Assay detector reported inconsistently.** The Abstract and Methods opening
  say concentrations were quantified by UPLC-MS/MS, while Methods "Sample
  collection and determination" says UPLC with photodiode array detection
  (UPLC-PDA) with the assay range 0.25-50.0 ug/mL. This affects no model
  parameter.

* **No external validation and no reported NCA.** The paper performed a VPC and a
  bootstrap but no external validation, and reports no NCA parameters, so there
  is no published Cmax / AUC / half-life table to compare against. The NCA
  section above uses closed-form identities instead.
