---
title: "Emicizumab (Yoneyama 2017)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Emicizumab (Yoneyama 2017)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup, message = FALSE}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Model and source

* Citation: Yoneyama K, Schmitt C, Kotani N, Levy GG, Kasai R, Iida S,
  Shima M, Kawanishi T. A Pharmacometric Approach to Substitute for a
  Conventional Dose-Finding Study in Rare Diseases: Example of Phase III
  Dose Selection for Emicizumab in Hemophilia A.
  *Clin Pharmacokinet*. 2018 May;57(5):613-619.
* Article: [doi:10.1007/s40262-017-0616-3](https://doi.org/10.1007/s40262-017-0616-3)
* Trial registries:
  [NCT01290263](https://clinicaltrials.gov/study/NCT01290263) (Japanese
  single-ascending-dose phase I, healthy volunteers),
  [NCT01747980](https://clinicaltrials.gov/study/NCT01747980) (Caucasian
  single-ascending-dose phase I),
  [JapicCTI-121934](https://www.clinicaltrials.jp/) (Japanese
  multiple-ascending-dose phase I, severe hemophilia A patients).

Emicizumab (ACE910) is a recombinant, humanised, bispecific IgG4 monoclonal
antibody that simultaneously binds activated factor IX (FIXa) and factor X
(FX), thereby mimicking the cofactor function of activated factor VIII
(FVIII). It is approved for prophylaxis of bleeding in patients with
severe hemophilia A with or without FVIII inhibitors. Yoneyama 2017 is the
pharmacometric analysis used to substitute for a conventional dose-finding
study and select the three phase III maintenance regimens (1.5 mg/kg QW,
3 mg/kg Q2W, 6 mg/kg Q4W) following a 3 mg/kg QW loading dose for the
first 4 weeks.

This vignette packages the structural population PK model from Section 3.2
of Yoneyama 2017 (Table 2). The companion repeated time-to-event (RTTE)
bleeding-hazard model from Section 3.3 (Table 3) is **not** included in
nlmixr2lib at this time -- nlmixr2lib does not currently support TTE-style
hazard models. Users interested in coupling exposure to bleeding hazard
should consult Section 2.4 and Table 3 of the paper.

# Population

The popPK dataset pooled three studies (Table 1 of Yoneyama 2017):

* **24 Japanese healthy male adult volunteers** (single SC dose of 0.01,
  0.1, 0.3, or 1 mg/kg). An additional 6 Japanese volunteers received
  0.001 mg/kg but were excluded from the popPK dataset because all plasma
  concentrations were below the assay LLOQ (< 0.05 ug/mL).
* **18 Caucasian healthy male adult volunteers** (single SC dose of 0.1,
  0.3, or 1 mg/kg). One Caucasian volunteer developed anti-emicizumab
  neutralising antibodies (NAbs).
* **18 Japanese male adult/adolescent patients with severe hemophilia A**
  (FVIII activity < 1 IU/dL) across three cohorts: Cohort 1 (n = 6,
  1 mg/kg SC loading + 0.3 mg/kg QW SC), Cohort 2 (n = 6, 3 mg/kg SC
  loading + 1 mg/kg QW SC), Cohort 3 (n = 6, 3 mg/kg QW SC). 11 of 18
  patients carried FVIII inhibitors at baseline.

Pooled-cohort weights ranged 41-87 kg (median 60-70 kg by cohort); ages
ranged 12-58 years; all 60 subjects were male. The model standardises
typical values for a 70 kg, NAb-negative healthy volunteer (Table 2
footnote b); the canonical `DIS_HEALTHY` register convention shifts the
structural typicals to the patient state -- see "Source trace" below.

The same information is available programmatically via the model's
`population` metadata
(`readModelDb("Yoneyama_2017_emicizumab")$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each
`ini()` entry in `inst/modeldb/specificDrugs/Yoneyama_2017_emicizumab.R`.
The table below collects them in one place for review.

| Parameter / equation | Value | Source location |
|----------------------|-------|-----------------|
| `lka` (ka, derived from t_(1/2),abs = 1.56 days) | log(ln(2)/1.56) = -0.811 | Yoneyama 2017 Table 2 |
| `lcl` (patient-typical CL/F at WT 70 kg, ADA-) | log(0.222) + 0.232 = -1.273 | Yoneyama 2017 Table 2 (HV-typical 0.222 L/day + patient effect 0.232) |
| `lvc` (patient-typical Vd/F at WT 70 kg) | log(10.2) + 0.175 = 2.498 | Yoneyama 2017 Table 2 (HV-typical 10.2 L + patient effect 0.175) |
| `e_wt_cl` (allometric exponent on CL/F) | 0.75 (fixed) | Yoneyama 2017 Table 2 |
| `e_wt_vc` (allometric exponent on Vd/F) | 1 (fixed) | Yoneyama 2017 Table 2 |
| `e_dis_healthy_cl` (DIS_HEALTHY effect on log-CL/F) | -0.232 (= -1 x paper PATIENT effect) | Yoneyama 2017 Table 2 |
| `e_dis_healthy_vc` (DIS_HEALTHY effect on log-Vd/F) | -0.175 (= -1 x paper PATIENT effect) | Yoneyama 2017 Table 2 |
| `e_ada_pos_cl` (NAb effect on log-CL/F) | 2.01 | Yoneyama 2017 Table 2 |
| `tNab` (onset time of NAb effect, days) | 33.4 | Yoneyama 2017 Table 2 (NONMEM MTIME) |
| `etalcl + etalvc` block | Var(CL) = 0.0737; Cov(CL, Vc) = 0.0278; Var(Vc) = 0.0455 | Yoneyama 2017 Table 2 |
| `etalka` IIV | 0.502 (paper reports Var(t_(1/2),abs) = 0.502 on log-scale; equal magnitude on lka because ka = ln(2)/t_(1/2),abs) | Yoneyama 2017 Table 2 |
| Additive residual error | 0.0149 ug/mL (SD) | Yoneyama 2017 Table 2 |
| Proportional residual error | 0.128 (12.8 % CV) | Yoneyama 2017 Table 2 |
| `d/dt(depot)`, `d/dt(central)` | one-compartment SC PK with first-order absorption | Yoneyama 2017 Methods Section 2.3 |
| Eq. 1 (continuous covariate, power form) | `P = P_TV * (Cov/CovTV)^theta * exp(eta)` | Yoneyama 2017 Methods Section 2.3 |
| Eq. 2 (categorical covariate, exp form) | `P = P_TV * exp(Cov * theta) * exp(eta)` | Yoneyama 2017 Methods Section 2.3 |

# Covariate column naming

| Source column (paper) | Canonical column used here | Notes |
|-----------------------|----------------------------|-------|
| `BW` | `WT` | Body weight; reference 70 kg. |
| `PATIENT` (1 = severe hemophilia A patient, 0 = healthy volunteer) | `DIS_HEALTHY` (1 = healthy volunteer, 0 = patient) | Canonical orientation is the inverse of the paper's. The model file shifts the structural typicals so the model is mathematically identical to the paper's Eq. 2. |
| `NAb` | `ADA_POS` | Time-fixed once the subject has seroconverted; the 33.4-day onset time post first dose is encoded inside `model()` via `nab_active <- ADA_POS * (t > tNab)`. |

# Virtual cohort

Original observed data are not publicly available. The cohort below
reproduces the three phase III dosing regimens that Yoneyama 2017 used to
generate Figure 5 (typical-value plasma emicizumab concentration-time
profiles): 3 mg/kg QW loading for the first 4 weeks followed by 1.5 mg/kg
QW, 3 mg/kg Q2W, or 6 mg/kg Q4W maintenance. The reference subject is the
typical patient (DIS_HEALTHY = 0, ADA_POS = 0) at 70 kg, the
standardisation used in the paper for the typical-value simulations.

```{r cohort}
set.seed(20171206)  # online publication date 6 Dec 2017

mod <- readModelDb("Yoneyama_2017_emicizumab")

# Simulation horizon: 52 weeks total (4-week loading + 48-week maintenance)
sim_end_day  <- 52 * 7  # 364 days
loading_days <- seq(0, by = 7, length.out = 4)  # weeks 0, 1, 2, 3

build_regimen <- function(label, maintenance_dose_mg_per_kg,
                          maintenance_interval_days, n_sub = 200,
                          wt_mean = 70, wt_cv = 0.20, id_offset = 0L) {
  ids <- id_offset + seq_len(n_sub)
  wt  <- pmax(40, wt_mean * exp(rnorm(n_sub, 0, sqrt(log(1 + wt_cv^2)))))

  # Loading: 3 mg/kg QW x 4 weeks (weeks 0, 1, 2, 3)
  loading_doses <- tidyr::expand_grid(id = ids, time = loading_days) |>
    dplyr::left_join(tibble(id = ids, WT = wt), by = "id") |>
    dplyr::mutate(
      amt  = 3 * WT,                     # mg per kg x kg = mg
      evid = 1L,
      cmt  = "depot"
    )

  # Maintenance: starts at day 28 (= week 4) at the regimen-specific interval
  maint_times <- seq(from = 28, to = sim_end_day - 1,
                     by = maintenance_interval_days)
  maint_doses <- tidyr::expand_grid(id = ids, time = maint_times) |>
    dplyr::left_join(tibble(id = ids, WT = wt), by = "id") |>
    dplyr::mutate(
      amt  = maintenance_dose_mg_per_kg * WT,
      evid = 1L,
      cmt  = "depot"
    )

  # Observation grid: dense in the first 4 weeks, every 3.5 days thereafter.
  # Loading-window step relaxed from 1 to 2 days for vignette build budget;
  # post-loading step retained at 3.5 days for PKNCA Cmax fidelity within the
  # 7-day QW maintenance interval.
  obs_grid <- sort(unique(c(
    seq(0, 28, by = 2),
    seq(28, sim_end_day, by = 3.5)
  )))
  obs <- tidyr::expand_grid(id = ids, time = obs_grid) |>
    dplyr::left_join(tibble(id = ids, WT = wt), by = "id") |>
    dplyr::mutate(
      amt  = NA_real_,
      evid = 0L,
      cmt  = "Cc"
    )

  dplyr::bind_rows(loading_doses, maint_doses, obs) |>
    dplyr::arrange(id, time, evid) |>
    dplyr::mutate(
      DIS_HEALTHY = 0L,    # severe hemophilia A patient reference state
      ADA_POS     = 0L,    # NAb-negative subjects (the modal case)
      regimen     = label
    )
}

events <- dplyr::bind_rows(
  # n_sub downsampled from 200 per regimen for vignette build budget; VPC bands
  # at n=80 are visually indistinguishable from n=200 for this smooth SC profile.
  build_regimen("1.5 mg/kg QW",  1.5, 7,  n_sub = 80, id_offset =    0L),
  build_regimen("3 mg/kg Q2W",   3.0, 14, n_sub = 80, id_offset =   80L),
  build_regimen("6 mg/kg Q4W",   6.0, 28, n_sub = 80, id_offset =  160L)
)
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
# Stochastic VPC-style simulation across the three phase III regimens.
sim <- rxode2::rxSolve(
  mod, events = events,
  keep = c("WT", "DIS_HEALTHY", "ADA_POS", "regimen")
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))

# Typical-value (no IIV) replication of Figure 5 of Yoneyama 2017.
sim_typ <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = events,
  keep = c("WT", "DIS_HEALTHY", "ADA_POS", "regimen")
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc))
```

# Replicate published figures

## Figure 5 -- typical-value concentration-time profiles by phase III regimen

Yoneyama 2017 Figure 5 presents the simulated median (with 5th-95th
percentile shading) plasma emicizumab concentration-time profile at each
of the three selected phase III dosing regimens. The target efficacious
exposure of 45 ug/mL (dashed line) is reached at steady-state trough by
the QW and Q2W maintenance regimens and approached -- but slightly under
-- by the Q4W maintenance regimen, as Figure 5(c) of the paper shows.

```{r figure-5, fig.cap = "Replicates Figure 5 of Yoneyama 2017: 5-50-95 % VPC of plasma emicizumab concentration-time profiles at the three phase III dosing regimens (3 mg/kg QW loading x 4 weeks followed by 1.5 mg/kg QW, 3 mg/kg Q2W, or 6 mg/kg Q4W maintenance). Dashed line: target efficacious exposure of 45 ug/mL for >= 50 % zero-bleeding probability at steady state."}
sim |>
  dplyr::group_by(time, regimen) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::mutate(regimen = factor(regimen,
    levels = c("1.5 mg/kg QW", "3 mg/kg Q2W", "6 mg/kg Q4W"))) |>
  ggplot(aes(time / 7, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 45, linetype = "dashed", colour = "firebrick") +
  facet_wrap(~ regimen, ncol = 3) +
  labs(x = "Time (weeks)", y = "Plasma emicizumab Cc (ug/mL)",
       title = "Figure 5 -- VPC by phase III dosing regimen",
       caption = "Dashed line: 45 ug/mL target efficacious exposure (Yoneyama 2017 Section 3.4).")
```

## Steady-state trough comparison

Yoneyama 2017 Table S1 of the ESM (referenced from Section 3.5 of the
main text) lists predicted median steady-state trough concentrations
(C_trough,ss) for each phase III regimen. The simulated medians from the
VPC above are summarised at the last observation time prior to the next
maintenance dose for each regimen.

```{r ss-trough}
trough_times <- list(
  "1.5 mg/kg QW" = sim_end_day - 0.5,
  "3 mg/kg Q2W"  = sim_end_day - 0.5,
  "6 mg/kg Q4W"  = sim_end_day - 0.5
)
ss_trough <- sim |>
  dplyr::filter(time >= sim_end_day - 28) |>
  dplyr::group_by(regimen) |>
  dplyr::summarise(
    n             = dplyr::n_distinct(id),
    trough_median = stats::median(
      Cc[time == max(time[time < sim_end_day])], na.rm = TRUE),
    .groups = "drop"
  )
knitr::kable(ss_trough,
  caption = "Simulated median plasma emicizumab C_trough at week 52 by phase III regimen.")
```

# PKNCA validation

PKNCA is applied to the last steady-state cycle of each maintenance
regimen (last full dosing interval before week 52). Single-dose AUC,
Cmax, Tmax, and effective accumulated half-life are computed per subject
within each regimen for cross-comparison; the per-regimen summary table
is presented below.

```{r pknca}
# Per-regimen last-dose interval and last-dose time
last_dose_per_regimen <- list(
  "1.5 mg/kg QW" = list(t_dose = 357, interval = 7),
  "3 mg/kg Q2W"  = list(t_dose = 350, interval = 14),
  "6 mg/kg Q4W"  = list(t_dose = 336, interval = 28)
)
ss_window <- dplyr::bind_rows(lapply(names(last_dose_per_regimen), function(r) {
  ld <- last_dose_per_regimen[[r]]
  sim |>
    dplyr::filter(regimen == r, time >= ld$t_dose,
                  time <= ld$t_dose + ld$interval, Cc > 0) |>
    dplyr::mutate(
      t_in_interval = time - ld$t_dose,
      treatment     = r
    )
}))

sim_nca <- ss_window |>
  dplyr::transmute(id, time = t_in_interval, Cc, treatment)

# Per-subject doses (the last maintenance dose) - one row per subject per regimen
dose_nca <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id, regimen) |>
  dplyr::slice_max(time, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::transmute(id, time = 0, amt, treatment = regimen)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | treatment + id)

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  auclast    = TRUE,
  half.life  = TRUE
)

nca_data <- PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- PKNCA::pk.nca(nca_data)
nca_summary <- summary(nca_res)
knitr::kable(nca_summary,
  caption = "PKNCA over the last steady-state dosing interval, by phase III regimen.")
```

## Comparison against published values

| Quantity | Paper value (Yoneyama 2017) | Packaged-model typical |
|----------|------------------------------|------------------------|
| Elimination half-life (healthy volunteers) | 31.8 days | `r sprintf("%.1f", log(2) * 10.2 / 0.222)` days (= ln(2) x Vd/F / (CL/F) at HV typicals) |
| Elimination half-life (patients) | 30.1 days | `r sprintf("%.1f", log(2) * (10.2 * exp(0.175)) / (0.222 * exp(0.232)))` days (= ln(2) x Vd_patient / CL_patient) |
| Patient/HV CL/F ratio | 1.26-fold | `r sprintf("%.3f", exp(0.232))`-fold (= exp(0.232)) |
| Patient/HV Vd/F ratio | 1.19-fold | `r sprintf("%.3f", exp(0.175))`-fold (= exp(0.175)) |
| Absorption half-life | 1.56 days | 1.56 days (`exp(-lka) * ln(2)`) |
| Target efficacious exposure | 45 ug/mL | Reproduced visually in Figure 5 |
| NAb-positive CL/F multiplier (effect active) | exp(2.01) = 7.46 | exp(2.01) = `r sprintf("%.2f", exp(2.01))` |
| NAb onset time post first dose | 33.4 days | 33.4 days |

The simulated steady-state trough at 6 mg/kg Q4W is slightly below 45
ug/mL (matching the Section 3.5 narrative in the paper: "a maintenance
dose of 6 mg/kg Q4W was predicted to result in a median C_trough,ss
slightly lower than 45 ug/mL"). The QW and Q2W regimens reach the target
at steady state.

# Assumptions and deviations

* **Companion RTTE bleeding-hazard model not included.** Yoneyama 2017
  develops both a popPK model (Section 2.3, Table 2) and an inhibitory
  Emax repeated time-to-event (RTTE) bleeding-hazard model (Section 2.4,
  Table 3). nlmixr2lib does not currently encode TTE/RTTE models in the
  same package framework as continuous-PK/PD models; this vignette
  validates only the popPK layer (Figure 5 of the paper, plus the
  steady-state trough quantification of Section 3.5). Users seeking to
  reproduce the bleeding-frequency simulations of Figure 4 and Figure 6
  should consult Section 2.4 and Table 3 of the paper directly.

* **Canonical `DIS_HEALTHY` reference orientation.** Yoneyama 2017 Table 2
  reports typical values for a 70 kg, NAb-negative HEALTHY VOLUNTEER and
  encodes the cohort effect via a PATIENT flag (1 = patient). The
  canonical `DIS_HEALTHY` register orients the indicator as 1 = healthy
  volunteer with reference category 0 = patient. The packaged model
  shifts the structural typicals to the patient state
  (`lcl = log(0.222) + 0.232`, `lvc = log(10.2) + 0.175`) and negates the
  covariate coefficients (`e_dis_healthy_cl = -0.232`,
  `e_dis_healthy_vc = -0.175`) so the model is mathematically identical
  to Eq. 2 of the paper at every (`DIS_HEALTHY`, `ADA_POS`, `WT`) tuple.
  This is the same canonical-rotation pattern used in
  `Lu_2015_tacrolimus.R`.

* **NAb effect activation time.** The paper estimates the NAb-effect
  onset time as 33.4 days post the first SC dose of emicizumab and
  encodes it via the NONMEM MTIME/MPAST functionality (Section 2.3,
  Table 2 footnote e). The packaged `model()` reproduces this as
  `nab_active <- ADA_POS * (t > tNab)` with `tNab = 33.4` days, so the
  effect activates when simulation time exceeds 33.4 days for any subject
  with `ADA_POS = 1`. The NAb-effect estimate is informed by only one
  positive Caucasian healthy volunteer (Table 1 / Section 3.2); the wide
  95 % CI (1.26-2.74) in Table 2 reflects that single-subject
  identifiability. For most simulation use cases users will set
  `ADA_POS = 0` (the modal case in the dataset, 59 of 60 subjects).

* **Reference body weight 70 kg.** Yoneyama 2017 Table 2 footnote b
  states this explicitly ("Standardized for a 70 kg, NAb-negative healthy
  volunteer"). The Caucasian cohort median was 70 kg (Table 1); Japanese
  cohort medians were 60 kg for both healthy volunteers and Cohorts 1-3
  patients. The 70 kg reference therefore reflects the Caucasian-weighted
  pooled-cohort typical rather than the per-cohort median.

* **IIV on absorption rate `lka` vs the paper's IIV on `t_(1/2),abs`.**
  The paper parameterises absorption via the half-life `t_(1/2),abs` and
  reports `Var(t_(1/2),abs) = 0.502` on the log-normal scale. The
  packaged model uses the nlmixr2lib canonical absorption parameter
  `lka` and assigns the same variance magnitude (`etalka ~ 0.502`)
  because `ka = ln(2) / t_(1/2),abs` implies `eta_lka = -eta_l_t_half`
  and variance is sign-invariant. The marginal individual ka
  distributions are identical between the two parameterisations.

* **One-compartment SC model.** Yoneyama 2017 Section 2.3 states
  explicitly: "A one-compartment model with first-order absorption and
  elimination was employed for the structural model for SC dosing." No
  peripheral compartment is fit. The packaged model preserves this --
  the IV phase I single-ascending-dose data is not included in the
  Yoneyama 2017 analysis.

* **Combined additive-plus-proportional residual error.** The paper
  parameterises the additive term as a standard deviation (0.0149 ug/mL)
  and the proportional term as a coefficient of variation (12.8 %); the
  packaged model uses the matching nlmixr2 syntax
  `Cc ~ add(addSd) + prop(propSd)`.

* **Phase III regimens used for validation.** The Figure 5 simulations
  in this vignette use the phase III maintenance regimens identified in
  Section 3.5 of the paper: 3 mg/kg QW loading for the first 4 weeks
  followed by 1.5 mg/kg QW, 3 mg/kg Q2W, or 6 mg/kg Q4W. The actual
  phase III HAVEN-1 / HAVEN-3 / HAVEN-4 trials confirmed these regimens.
  The 45 ug/mL target efficacious exposure is from the integrated
  PopPK-RTTE simulation reported in Section 3.4 (Figure 4).

# Provenance summary

| Source on disk | Used for |
|----------------|----------|
| `Yoneyama_2018_A_Pharmacometric_Approach_to_Substitute__4df84e.pdf` | Main paper Methods, Results, Table 1, Table 2, Figure 5, Section 3.5 narrative on phase III regimen selection. |
| `Yoneyama_2018_A_Pharmacometric_Approach_to_Substitute__4df84e_trimmed.md` | Same content, trimmed-markdown form used during extraction for searchability. |

The Electronic Supplementary Material (ESM Methods, Table S1, Figures
S1-S6) is referenced in the paper but is not on disk for this
extraction; all parameter values used by the packaged model are present
in Table 2 of the main paper. No author correspondence was required.
