---
title: "Inotersen (Yu 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inotersen (Yu 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7.5,
  fig.height = 4.5
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Inotersen population PK/PD in hATTR polyneuropathy (Yu 2020)

Yu et al. (2020) developed the first population PK/PD model for an antisense
oligonucleotide therapeutic: inotersen, a 20-mer 2'-O-methoxyethyl gapmer that
reduces hepatic production of transthyretin (TTR). Plasma inotersen follows a
two-compartment model with first-order SC absorption and linear elimination
from the central compartment. Serum TTR follows an indirect-response model in
which inotersen inhibits the zero-order production of TTR:

$$\frac{dTTR}{dt} = k_{in}\left(1 - \frac{I_{max} C_p}{IC_{50} + C_p}\right) - k_{out}\,TTR$$

with $k_{in} = BL \cdot k_{out}$ so that TTR starts at its baseline. The PK
model was fit first and the individual (post hoc) PK parameters then drove the
PD fit; the packaged model carries both layers in one file.

* Citation: `r readModelDb("Yu_2020_inotersen")()$reference`
* Article: <https://doi.org/10.1089/nat.2019.0822>

# Population

The analysis pooled three Ionis studies (Yu 2020 Clinical studies section and
Table 1): a phase 1 single- and multiple-ascending-dose study in 51
inotersen-treated healthy volunteers (50-400 mg SC), the pivotal phase 2/3
NEURO-TTR study (300 mg SC on days 1, 3 and 5 then once weekly for 65 weeks)
and its open-label extension (300 mg SC weekly for up to 5 years). Of 202
subjects, 151 were patients with hereditary transthyretin amyloidosis with
polyneuropathy (hATTR-PN); 197 contributed 3,602 PK observations. Median age
was 57 years (range 25-81), median body weight 71.8 kg (37.0-140), median lean
body mass 51.6 kg (31.3-80.3), 69.8% were male, and 87.6% were Caucasian, 7.43%
African American and 3.47% Asian. Median baseline TTR was 22.1 mg/dL. PK
samples collected after the onset of antidrug antibodies were excluded.

The same information is available programmatically via
`readModelDb("Yu_2020_inotersen")()$population`.

# Source trace

Every `ini()` value carries an in-file comment naming its source row. They are
collected here for review.

| Parameter / equation | Value | Source |
|----------------------|-------|--------|
| `ka` | 0.261 1/h | Table 2 'Absorption rate constant, Ka' |
| `CL/F` (hATTR, LBM 51.6 kg) | 3.4 L/h | Table 2 'Clearance, CL' |
| `Vc/F` (hATTR, LBM 51.6 kg) | 20.7 L | Table 2 'Central volume, Vc' |
| `Q/F` | 0.266 L/h | Table 2 'Intercompartmental clearance, Q' |
| `Vp/F` | 230 L | Table 2 'Peripheral volume, Vp' |
| LBM exponent on CL/F and Q/F | 1 (fixed) | Table 2 'Lean body mass * CL/Q', power centred on median |
| LBM exponent on Vc/F and Vp/F | 1 (fixed) | Table 2 'Lean body mass * Vc/Vp', power centred on median |
| LBM reference | 51.6 kg | Table 1 lean body mass median |
| Healthy effect on CL/F | +0.111, `(1 + 0.111 * DIS_HEALTHY)` | Table 2 'Disease * CL', proportional |
| Healthy effect on Vc/F | -0.284, `(1 - 0.284 * DIS_HEALTHY)` | Table 2 'Disease * Vc', proportional |
| omega^2 CL, Vc, Vp | 0.071, 0.335, 0.677 | Table 2 |
| cov(Vc, CL), cov(CL, Vp) | 0.049, -0.145 | Table 2 (cov(Vc, Vp) not reported; set to 0) |
| PK residual | sigma^2 = 0.168 log-additive; `expSd = sqrt(0.168) = 0.4099` | Table 2 |
| `Imax` | 0.913 | Table 3 |
| `IC50` | 9.07 ng/mL | Table 3 |
| Baseline TTR (hATTR) | 20.4 mg/dL | Table 3 'Estimated baseline' |
| Healthy effect on baseline | +0.269, `(1 + 0.269 * DIS_HEALTHY)` | Table 3 and its footnote |
| `kout` | 0.00308 1/h | Table 3 |
| omega^2 IC50, kout; cov | 0.953, 0.288; -0.182 | Table 3 |
| omega^2 baseline | 0.0537 | Table 3 |
| PD residual | proportional 0.13, additive 1.35 mg/dL | Table 3 'Residual variability (log space), (SD)' |
| Indirect-response equation | see above | Population PD modeling section |

# Covariates

| Source | Canonical column | Notes |
|--------|------------------|-------|
| Lean body mass (kg) | `LBM` | Reference 51.6 kg. The paper does not state which LBM formula it used. |
| Disease status | `DIS_HEALTHY` | 1 = healthy volunteer, 0 = hATTR-PN patient (reference). |

# Closed-form checks on the typical PK

The paper derives four numbers from its typical PK estimates: alpha and beta
half-lives of 3.91 h and 26.9 days, a steady-state volume of 250.7 L, and a
trough accumulation ratio of 6.07 for weekly dosing. These use the same
parameters on both sides, so they are checked tightly.

```{r closed-form}
ini_df <- rxode2::rxode2(readModelDb("Yu_2020_inotersen"))$iniDf
th <- setNames(ini_df$est, ini_df$name)
cl <- exp(th[["lcl"]])
vc <- exp(th[["lvc"]])
q <- exp(th[["lq"]])
vp <- exp(th[["lvp"]])
kel <- cl / vc
k12 <- q / vc
k21 <- q / vp
s <- kel + k12 + k21
alpha <- (s + sqrt(s^2 - 4 * kel * k21)) / 2
beta <- (s - sqrt(s^2 - 4 * kel * k21)) / 2
checks <- data.frame(
  Quantity = c(
    "Alpha half-life (h)",
    "Beta half-life (days)",
    "Vss (L)",
    "Trough accumulation ratio, QW"
  ),
  Paper = c(3.91, 26.9, 250.7, 6.07),
  Model = c(
    log(2) / alpha,
    log(2) / beta / 24,
    vc + vp,
    1 / (1 - exp(-beta * 168))
  )
)
knitr::kable(checks, digits = 3)
stopifnot(all(abs(checks$Model / checks$Paper - 1) < 0.01))
```

# Virtual cohort and dosing regimens

Yu 2020 simulated the 151 hATTR-PN patients of the analysis dataset under four
regimens for 65 weeks (Simulations section). Subject-level covariates are not
published, so the virtual cohort below draws lean body mass from a truncated
log-normal around the pooled-cohort median and range (Table 1); all subjects
are patients (`DIS_HEALTHY = 0`). Each regimen uses the same 150 virtual
patients so that differences between arms come from the dosing alone.

Day 1 of the paper is time 0, so day `d` is `(d - 1) * 24` h; the last dose of
each regimen falls on day 449 (t = 10752 h), which is where Table 4 reports
the steady-state interval.

```{r cohort}
set.seed(20200301)
rxode2::rxSetSeed(20200301)
n_per_arm <- 150L

cohort <- tibble(
  subj = seq_len(n_per_arm),
  LBM = pmin(pmax(rlnorm(n_per_arm, log(51.6), 0.16), 31.3), 80.3),
  DIS_HEALTHY = 0L
)

t_last <- 448 * 24
regimens <- list(
  "300 mg QW" = list(amt = 300, times = seq(0, t_last, by = 168), tau = 168),
  "300 mg QOW" = list(amt = 300, times = seq(0, t_last, by = 336), tau = 336),
  "Loading + 300 mg QW" = list(
    amt = 300,
    times = c(0, 48, 96, seq(168, t_last, by = 168)),
    tau = 168
  ),
  "150 mg QW" = list(amt = 150, times = seq(0, t_last, by = 168), tau = 168)
)

make_arm <- function(arm_index, cohort, regimens) {
  reg <- regimens[[arm_index]]
  ids <- cohort |> mutate(id = subj + (arm_index - 1L) * nrow(cohort))
  dense <- c(seq(0, 12, by = 0.5), seq(14, reg$tau, by = 2))
  obs_t <- sort(unique(c(
    dense,
    seq(0, t_last + reg$tau, by = 24),
    t_last + dense
  )))
  doses <- ids |>
    crossing(time = reg$times) |>
    mutate(amt = reg$amt, evid = 1L, cmt = "depot", dvid = NA_integer_)
  obs <- ids |>
    crossing(time = obs_t) |>
    mutate(amt = 0, evid = 0L, cmt = NA_character_, dvid = 1L)
  bind_rows(doses, obs) |>
    mutate(regimen = names(regimens)[arm_index], tau = reg$tau) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(lapply(
  seq_along(regimens),
  make_arm,
  cohort = cohort,
  regimens = regimens
))
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
```

# Simulation

The model has two endpoints (`Cc` and `ttr`), so observation rows carry
`dvid = 1` with no compartment; rxode2 returns both outputs at every
observation time. Residual error is not added (`sigma = NA`); between-subject
variability is simulated.

```{r simulate}
mod <- readModelDb("Yu_2020_inotersen")
sim <- rxode2::rxSolve(
  mod,
  events = events,
  sigma = NA,
  useLinCmt = FALSE,
  keep = c("regimen", "tau"),
  returnType = "data.frame"
)
sim <- sim |> mutate(regimen = factor(regimen, levels = names(regimens)))
```

# Replicate published figures

## Figure 4: plasma trough concentrations

```{r figure-4}
trough <- sim |>
  filter(time %in% seq(168, t_last + 336, by = 168)) |>
  group_by(regimen, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05),
    Q50 = median(Cc),
    Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(trough, aes(time / 24 / 7, Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time (weeks)",
    y = "Inotersen plasma trough (ng/mL)",
    colour = NULL,
    fill = NULL,
    caption = "Replicates Figure 4 of Yu 2020 (median and 90% prediction interval)."
  ) +
  theme_bw()
```

## Figure 5: serum TTR

```{r figure-5}
ttr_prof <- sim |>
  filter(time %% 24 == 0) |>
  group_by(regimen, time) |>
  summarise(
    Q05 = quantile(ttr, 0.05),
    Q50 = median(ttr),
    Q95 = quantile(ttr, 0.95),
    .groups = "drop"
  )

ggplot(ttr_prof, aes(time / 24 / 7, Q50, colour = regimen, fill = regimen)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(
    x = "Time (weeks)",
    y = "Serum TTR (mg/dL)",
    colour = NULL,
    fill = NULL,
    caption = "Replicates Figure 5 of Yu 2020 (median and 90% prediction interval)."
  ) +
  theme_bw()
```

# PKNCA validation against Table 4

Table 4 reports geometric means of AUC over the dosing interval, Cmax and
Ctrough after the first dose (day 1) and after the last dose (day 449). NCA is
run on both intervals for each regimen, with Ctrough taken as the
concentration at the end of the interval (PKNCA `ctrough`), i.e. just before
the next dose. For the loading regimen the day-1
interval is not compared, because the paper's day-1 values for that arm mix a
48-hour trough with an AUC equal to the weekly arms' and the interval it used
is not stated.

```{r pknca}
# Each interval is re-anchored to its own dose so that it starts at time 0:
# PKNCA evaluates ctrough on time after dose, and the two periods are then
# separate PKNCA groups.
tau_by_reg <- vapply(regimens, `[[`, numeric(1), "tau")
periods <- c("Day 1" = 0, "Day 449" = t_last)

conc <- bind_rows(lapply(names(periods), function(p) {
  sim |>
    filter(!is.na(Cc)) |>
    mutate(time = time - periods[[p]], period = p) |>
    filter(time >= 0, time <= tau_by_reg[as.character(regimen)])
})) |>
  mutate(regimen = as.character(regimen)) |>
  select(id, time, Cc, regimen, period)

doses <- bind_rows(lapply(names(periods), function(p) {
  events |>
    filter(evid == 1L) |>
    mutate(time = time - periods[[p]], period = p) |>
    filter(time >= 0, time < tau)
})) |>
  select(id, time, amt, regimen, period)

intervals <- expand.grid(
  regimen = names(regimens),
  period = names(periods),
  stringsAsFactors = FALSE
) |>
  mutate(start = 0, end = unname(tau_by_reg[regimen])) |>
  filter(!(regimen == "Loading + 300 mg QW" & period == "Day 1")) |>
  mutate(auclast = TRUE, cmax = TRUE, ctrough = TRUE)

nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc, Cc ~ time | regimen + period + id),
  PKNCA::PKNCAdose(doses, amt ~ time | regimen + period + id),
  intervals = intervals
))

nca_res <- as.data.frame(nca$result) |>
  mutate(group = paste(regimen, period, sep = ", "))
```

```{r nca-compare}
# Table 4 geometric means, converted to ng*h/mL and ng/mL.
reference <- tribble(
  ~group, ~auclast, ~cmax, ~ctrough,
  "300 mg QW, Day 1", 84400, 6350, 5.62,
  "300 mg QOW, Day 1", 85500, 6330, 4.28,
  "150 mg QW, Day 1", 42300, 3160, 2.81,
  "300 mg QW, Day 449", 89900, 6390, 34.3,
  "300 mg QOW, Day 449", 90100, 6340, 14.8,
  "Loading + 300 mg QW, Day 449", 90000, 6340, 34.3,
  "150 mg QW, Day 449", 45000, 3180, 17.2
) |>
  as.data.frame()

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res |> select(group, PPTESTCD, PPORRES),
  reference = reference,
  by = "group",
  units = c(auclast = "ng*h/mL", cmax = "ng/mL", ctrough = "ng/mL")
)
knitr::kable(cmp, caption = "Simulated median vs Yu 2020 Table 4 geometric mean.")
if (!is.null(attr(cmp, "footnote"))) cat(attr(cmp, "footnote"))

pct <- as.numeric(gsub("[%*+]", "", cmp[["% diff"]]))
stopifnot(!anyNA(pct), length(pct) == 21L)
stopifnot(
  abs(median(pct)) < 10,
  quantile(abs(pct), 0.9) < 20
)
```

The simulated medians are compared with the paper's geometric means; for a
log-normally distributed exposure the two coincide. The paper also carried
parameter uncertainty (500 replicate datasets) that this simulation does not.

# TTR reduction against Table 5

Table 5 and the Results text report the mean percent change from baseline in
TTR at days 7, 14, 28, 85 and at steady state. Each simulated patient's change
is computed from their own baseline; the steady-state value is taken at day
449.

```{r table-5}
days <- c(7, 14, 28, 85, 449)
ttr_chg <- sim |>
  group_by(id) |>
  mutate(bl = ttr[time == 0][1]) |>
  ungroup() |>
  filter(time %in% ((days - 1) * 24)) |>
  mutate(day = time / 24 + 1, pct = 100 * (ttr / bl - 1)) |>
  group_by(regimen, day) |>
  summarise(simulated = mean(pct), .groups = "drop")

published <- tribble(
  ~regimen, ~day, ~paper,
  "300 mg QW", 7, -20.2,
  "300 mg QW", 14, -34.8,
  "300 mg QW", 28, -51.2,
  "300 mg QW", 85, -67.9,
  "300 mg QW", 449, -71.9,
  "Loading + 300 mg QW", 7, -32.5,
  "Loading + 300 mg QW", 14, -45.1,
  "Loading + 300 mg QW", 28, -57.5,
  "Loading + 300 mg QW", 85, -69.0,
  "Loading + 300 mg QW", 449, -71.9,
  "300 mg QOW", 7, -20.2,
  "300 mg QOW", 14, -24.7,
  "300 mg QOW", 28, -38.0,
  "300 mg QOW", 85, -53.8,
  "300 mg QOW", 449, -58.8,
  "150 mg QW", 7, -15.9,
  "150 mg QW", 14, -28.1,
  "150 mg QW", 28, -42.6,
  "150 mg QW", 85, -58.6,
  "150 mg QW", 449, -63.1
)

ttr_cmp <- published |>
  left_join(ttr_chg |> mutate(regimen = as.character(regimen)), by = c("regimen", "day")) |>
  mutate(difference = simulated - paper)
ttr_cmp |>
  dplyr::rename(
    Regimen = regimen,
    Day = day,
    "Paper (% change)" = paper,
    "Simulated (% change)" = simulated,
    "Difference (points)" = difference
  ) |>
  knitr::kable(digits = 1, caption = "Mean percent change from baseline TTR.")

stopifnot(
  abs(median(ttr_cmp$difference)) < 5,
  quantile(abs(ttr_cmp$difference), 0.9) < 10
)
```

The simulated reductions track the published ones within a few percentage
points at every time and regimen. The largest gaps (about 3-4 points more
suppression than published) are in the lower-exposure 300 mg QOW and 150 mg
QW arms at later times, where the steep part of the concentration-response
curve makes the mean sensitive to the tail of the wide IC50 distribution
(omega^2 = 0.953) and to the parameter uncertainty the paper included.

# Assumptions and deviations

- **Dose effect on clearance omitted.** Table 2 reports a 'Dose effect * CL'
  term (exponential, estimate 3.97) that raised apparent clearance at the
  50 mg phase 1 dose, described only as 'a simplified exponential model with
  dose administered being normalized to 300 mg dose'. The natural reading,
  `CL * exp(3.97 * (1 - DOSE / 300))`, would raise CL/F 7.3-fold at 150 mg and
  cannot be reconciled with the paper's own Table 4, where every exposure
  measure at 150 mg is almost exactly half of that at 300 mg. The functional
  form cannot be recovered from the publication, so the maintainers omitted
  the term; the paper itself states the nonlinearity is not clinically relevant
  at 150-300 mg. The model should not be used below 150 mg. The term is
  recorded in the model's `covariatesDataExcluded` metadata.
- **Orientation of the disease-status effects.** Only the PD effect's
  direction is stated (Table 3 footnote: the shift in baseline TTR of healthy
  volunteers relative to hATTR patients). The PK effects are read on the same
  orientation, with hATTR patients as the reference. This reproduces the
  Table 4 hATTR exposures above (the opposite orientation would lower the
  simulated AUC by about 10%) and the Discussion's statement that post hoc
  CL/F in hATTR patients (3.18 L/h) is close to the typical 3.40 L/h.
- **PD residual error.** Table 3 reports proportional (0.13) and additive
  (1.35) SDs 'in log space' for log-transformed TTR. An additive SD of 1.35 on
  the log scale is implausible, so the pair is read as the usual
  log-transform-both-sides form of a combined error,
  `W = sqrt(prop^2 + (add / IPRED)^2)`, and encoded as combined proportional +
  additive error on linear TTR in mg/dL.
- **Unreported covariance.** Table 2 reports covariances between Vc and CL
  and between CL and Vp but not between Vc and Vp; it is set to 0. The
  resulting 3 x 3 matrix is positive definite.
- **LBM formula.** The paper does not name the formula used for lean body
  mass. The virtual cohort draws LBM around the pooled-cohort median (51.6 kg,
  range 31.3-80.3 kg), which includes the 51 healthy volunteers.
- **Table 5 day-85 level for the loading regimen.** The TTR level printed for
  the loading regimen at day 85 (7.39 mg/dL) duplicates the 150 mg QW value
  and is inconsistent with the -69.0% change printed beside it; the
  comparison above uses the percent changes only.
- **Day convention.** Day `d` in the paper is taken as `(d - 1) * 24` h after
  the first dose, so the last dose falls on day 449 as in Table 4.
- **Parameter uncertainty and study design.** The paper's simulations drew
  from the parameter variance-covariance matrix and used the observed
  covariates of the 151 patients; this vignette uses fixed estimates and a
  synthetic cohort of 150 patients per arm.
- **Errata search.** No erratum or correction for Yu 2020 was found in
  Europe PMC as of September 2026.
