---
title: "Vancomycin (Yu 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Yu 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Yu_2025_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Yu_2025_vancomycin"))$description`
* Article (open access): <https://doi.org/10.1007/s40268-025-00523-8>
* No supplementary material accompanies the article; the Code Availability
  statement offers the NONMEM code on request from the corresponding author.

Yu 2025 is a single-center retrospective therapeutic-drug-monitoring study of IV
vancomycin in Chinese children. It has two quantitative products, and this
vignette exercises both:

1. **A one-compartment population PK model** (Table 3, with the parameter
   equations in Eqs. 12 and 13) - the model packaged here as
   `Yu_2025_vancomycin`.
2. **A Monte Carlo simulation of steady-state trough concentrations** across
   three renal-function strata and two guideline dosing regimens (Table 4,
   plotted in Fig. 3).

The first reproduces exactly. The second does **not** reproduce for the two
renal-impairment strata, and the section "Table 4 does not follow from Eq. 12
for the impaired strata" below localizes why.

# Population

The model was fit to 100 pediatric inpatients contributing 124 vancomycin
concentrations at Anhui Provincial Children's Hospital (Hefei, Anhui, China)
between September 2021 and November 2023 (Yu 2025 Sect. 2.1, Table 1). Median
age was 4 years (mean 5.17; range 0.0027-15), median actual body weight 24 kg
(mean 21.49; range 1.25-74), and median height 106.5 cm (range 37-178); 41% were
female. Median Schwartz eGFR was 173.42 mL/min/1.73 m^2 (mean 163.83; range
31.9-260.82), i.e. this cohort's renal function skews *supranormal*, which
matters below because the model's eGFR reference is that same 173.42.

Patients on renal replacement therapy or ECMO, with undetectable vancomycin
concentrations, or with incomplete clinical data were excluded. Dosing was
empirical and clinician-chosen; vancomycin was given as a 60-minute IV
intermittent infusion. Sampling was sparse and dominated by steady-state troughs
drawn 30 minutes before the fourth dose, giving only 124 concentrations from 100
patients - barely more than one sample per subject. Plasma vancomycin was
measured by enzyme amplification immunoassay (Siemens Viva-ProE) with a lower
limit of quantification of 2 mg/L; measurable samples below the LLOQ were kept.

That sampling design drives three structural decisions in the paper that a user
of this model needs to know about: a two-compartment model could not be
supported, interindividual variability on volume was dropped as unestimable, and
the residual error magnitude was fixed rather than estimated.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Yu_2025_vancomycin")()$population`).

# Source trace

Per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Yu_2025_vancomycin.R`. The table below
collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL at WT 70 kg, eGFR 173.42) | 8.22 L/h (RSE 9%) | Yu 2025 Table 3, `theta_CL` row |
| `lvc` (V at WT 70 kg) | 113 L (RSE 18%) | Yu 2025 Table 3, `theta_V` row |
| `e_wt_cl` | 0.75, fixed | Yu 2025 Eq. 8 `CL_i = CL_pop * (WT/70)^0.75` and Eq. 12; Sect. 2.4.1 "Clearance and V were scaled using fixed exponents of 0.75 and 1.0" |
| `e_wt_vc` | 1, fixed | Yu 2025 Eq. 9 `V_i = V_pop * (WT/70)` and Eq. 13; same Sect. 2.4.1 sentence |
| `e_crcl_cl` | 0.812 (RSE 15%) | Yu 2025 Table 3, `theta_GFR` row; applied as `(GFR/173.42)^theta_GFR` in Eq. 12 |
| Reference weight | 70 kg | Yu 2025 Eqs. 8, 9, 12, 13; confirmed in Sect. 4, which divides 8.22 L/h by 70 kg to obtain 0.117 L/h/kg |
| Reference eGFR | 173.42 mL/min/1.73 m^2 | Yu 2025 Eq. 12; equals the Table 1 cohort median, per Sect. 2.4.2 ("standardized by the median") |
| eGFR definition | `k * height (cm) / Scr (mg/dL)` | Yu 2025 Eq. 2 (Schwartz), with `k` in {0.33, 0.45, 0.55, 0.7} by age and sex; Eq. 1 gives the creatinine conversion 1 mg/dL = 88.4 umol/L |
| `etalcl` | omega^2 = 0.319225 | Yu 2025 Table 3: `omega_CL` 56.5% (RSE 10%); Eq. 3 gives the exponential model `P_i = theta_P * exp(eta_i)` |
| IIV on V | absent | Yu 2025 Sect. 3.2.1: "Because the estimated value of the inter-individual variability in V was too small, it was omitted in subsequent calculations" |
| `propSd` | 0.3, fixed | Yu 2025 Table 3: residual unexplained variability 30% (fixed); Sect. 3.2.1 selected the proportional model (Eq. 5); Sect. 4 describes the 0.2/0.3/0.4 sensitivity analysis behind the choice |
| One-compartment, first-order elimination, no depot | n/a | Yu 2025 Sect. 2.4.1: NONMEM ADVAN1 TRANS2; Sect. 2.2: 60-minute IV intermittent infusion |
| Renal strata and dosing regimens | n/a | Yu 2025 Sect. 2.6 and Table 4 |

Two source details deserve emphasis.

* **Table 3's typical values are for a 70 kg subject, not a typical study
  child.** The cohort's median weight is 24 kg and its maximum is 74 kg, so
  "CL = 8.22 L/h, V = 113 L" describes a hypothetical adult-sized patient at the
  allometric reference. Sect. 4 makes this explicit. Reading 113 L as a
  pediatric volume of distribution would be a 3-fold error at the cohort median
  weight, where the model gives `113 * 24/70 = 38.7 L`.
* **`omega_CL` = 56.5% is read as `100 * omega`, not as a log-normal %CV.** The
  paper reports the row as "Inter-individual variability `omega_CL` (%)" - the
  symbol is omega itself - and does not print the `%CV = sqrt(exp(omega^2) - 1)`
  definition that some papers use. Both readings are on the standard-deviation
  scale (see the Errata for why the reported number cannot be a *variance*), and
  they differ by only 7% in omega (0.565 versus 0.5265), so no conclusion in
  this vignette turns on the choice.

# The packaged model reproduces Table 3 exactly

```{r table3-typical}
mod <- readModelDb("Yu_2025_vancomycin")

typical_at <- function(WT, CRCL, etalcl = 0) {
  ev <- data.frame(
    id = 1L, WT = WT, CRCL = CRCL, etalcl = etalcl,
    time = 0, amt = NA_real_, evid = 0L, cmt = "central"
  )
  rxode2::rxSolve(mod, events = ev, omega = NA, returnType = "data.frame")
}

ref_subject <- typical_at(WT = 70, CRCL = 173.42)

table3 <- tibble::tibble(
  Parameter = c("Clearance (L/h)", "Volume of distribution (L)"),
  Published = c(8.22, 113),
  Model     = c(ref_subject$cl[1], ref_subject$vc[1])
)
stopifnot(max(abs(table3$Model - table3$Published)) < 1e-8)

knitr::kable(
  table3,
  caption = paste("Typical values at the Yu 2025 allometric reference subject",
                  "(WT 70 kg, eGFR 173.42 mL/min/1.73 m^2) versus Yu 2025 Table 3."),
  align = c("l", "r", "r")
)
```

The paper's own sanity check in Sect. 4 - "Based on a WT of 70 kg, the CL was
approximately 0.117 L/h/kg" - also reproduces, and confirms that the 70 kg
reference (rather than the 24 kg cohort median) is the right reading of Eq. 12.

```{r cl-per-kg}
cl_per_kg <- ref_subject$cl[1] / 70
stopifnot(abs(cl_per_kg - 0.117) < 0.0005)
sprintf("%.4f L/h/kg at 70 kg (Yu 2025 Sect. 4: 'approximately 0.117 L/h/kg')", cl_per_kg)
```

# Virtual cohort

Yu 2025 Sect. 2.6 fixes the simulated patient's weight at the cohort median of
24 kg and stratifies on Schwartz eGFR into normal (>= 90), mild insufficiency
(60-89) and moderate insufficiency (30-59) mL/min/1.73 m^2. Children with eGFR
<= 30 were excluded because they usually receive renal replacement therapy.

The paper does **not** report which eGFR value it used within each stratum - it
says only that 1000 virtual patients were simulated "using the initial dataset".
This vignette therefore samples eGFR uniformly across each stratum's stated band,
capping the normal stratum at the cohort maximum of 260.82. That choice is made
before looking at any result and is not tuned: the median of the normal
stratum's uniform draw (175.4) lands within 1.2% of the cohort median 173.42
that Eq. 12 uses as its reference, so it is the natural representative value.

```{r cohort}
set.seed(20250924)
n_per_arm <- 200L

strata <- tibble::tibble(
  renal   = c("Normal (eGFR >= 90)", "Mild (eGFR 60-89)", "Moderate (eGFR 30-59)"),
  egfr_lo = c(90, 60, 30),
  egfr_hi = c(260.82, 89, 59)
)

base_cohort <- strata |>
  dplyr::rowwise() |>
  dplyr::reframe(
    renal  = renal,
    CRCL   = runif(n_per_arm, egfr_lo, egfr_hi),
    # omega = 0.565 on the log scale; supplied as a column so the solve is
    # deterministic and the two regimens below share one cohort exactly.
    etalcl = rnorm(n_per_arm, 0, 0.565)
  ) |>
  dplyr::mutate(
    subject = dplyr::row_number(),
    WT      = 24  # Yu 2025 Sect. 2.6: "median patient WT (24 kg)"
  )

stopifnot(nrow(base_cohort) == 3L * n_per_arm)

base_cohort |>
  dplyr::group_by(renal) |>
  dplyr::summarise(
    n = dplyr::n(),
    `eGFR median` = round(median(CRCL), 2),
    `eGFR min` = round(min(CRCL), 1),
    `eGFR max` = round(max(CRCL), 1),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Simulated renal-function strata (Yu 2025 Sect. 2.6 bands).")
```

# Simulation of the two guideline regimens

Both regimens deliver the 60 mg/kg/day recommended by the Chinese 2020
vancomycin TDM guideline that Yu 2025 Sect. 2.6 cites: 20 mg/kg every 8 h and
15 mg/kg every 6 h, each as the 60-minute infusion of Sect. 2.2. Steady state is
imposed exactly with `ss = 1` on the time-zero dose rather than by simulating a
burn-in, so the time-zero record *is* the steady-state trough that Table 4
reports.

The same `base_cohort` (same weights, same eGFR draws, same `etalcl` draws) is
run through both regimens, which makes the linearity gate below an exact
per-subject identity rather than a distributional comparison.

```{r events}
regimens <- tibble::tibble(
  treatment = c("20 mg/kg q8h", "15 mg/kg q6h"),
  mg_per_kg = c(20, 15),
  tau       = c(8, 6),
  id_offset = c(0L, 10000L)
)

make_arm <- function(mg_per_kg, tau, id_offset, treatment) {
  subj <- base_cohort |>
    dplyr::mutate(
      id        = id_offset + subject,
      treatment = treatment,
      dose_mg   = mg_per_kg * WT
    )

  doses <- subj |>
    tidyr::expand_grid(dose_time = seq(0, 24 - tau, by = tau)) |>
    dplyr::transmute(
      id, subject, treatment, renal, WT, CRCL, etalcl,
      time = dose_time,
      amt  = dose_mg,
      rate = dose_mg / 1,  # 60-minute infusion (Yu 2025 Sect. 2.2)
      evid = 1L,
      cmt  = "central",
      ss   = ifelse(dose_time == 0, 1L, 0L),
      ii   = ifelse(dose_time == 0, tau, 0)
    )

  # Observation grid: 0.25 h throughout, refined to 0.05 h over each infusion
  # and its immediate post-infusion peak so that lin-up/log-down AUC reproduces
  # the analytic integral. Records are also placed immediately BEFORE each dose,
  # because an observation sharing a dose's timestamp is evaluated after that
  # dose and would otherwise miss the true trough.
  peak_times <- as.numeric(outer(seq(0, 24 - tau, by = tau),
                                 seq(0, 1.5, by = 0.05), "+"))
  grid <- sort(unique(c(seq(0, 24, by = 0.25),
                        seq(tau, 24, by = tau) - 1e-4,
                        peak_times[peak_times <= 24])))

  obs <- subj |>
    tidyr::expand_grid(time = grid) |>
    dplyr::transmute(
      id, subject, treatment, renal, WT, CRCL, etalcl, time,
      amt = NA_real_, rate = 0, evid = 0L, cmt = "central", ss = 0L, ii = 0
    )

  dplyr::bind_rows(doses, obs) |> dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(lapply(
  seq_len(nrow(regimens)),
  function(i) with(regimens[i, ], make_arm(mg_per_kg, tau, id_offset, treatment))
))

stopifnot(dplyr::n_distinct(events$id) == 6L * n_per_arm)
```

```{r simulate}
# etalcl is supplied as a column, so omega = NA gives a fully deterministic
# solve (no resampling of the random effects).
sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("subject", "treatment", "renal", "WT", "CRCL"),
  omega  = NA,
  returnType = "data.frame"
)

stopifnot(!anyNA(sim$Cc), all(sim$Cc >= 0))
```

# Replicate published figures and findings

## Steady-state concentration-time profiles (Fig. 3)

Yu 2025 Fig. 3 shows six panels - three renal strata by two regimens - of
simulated vancomycin concentration over time, with the median as a solid line
and the 5th-95th percentiles shaded. The panel below is the same figure over one
steady-state day.

```{r fig3}
sim |>
  dplyr::group_by(treatment, renal, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(colour = "steelblue") +
  geom_hline(yintercept = c(5, 15), linetype = "dashed", colour = "grey40") +
  facet_grid(treatment ~ renal) +
  labs(
    x = "Time within the steady-state day (h)",
    y = "Vancomycin concentration (mg/L)",
    title = "Replicates Figure 3 of Yu 2025 (WT 24 kg)",
    caption = paste("Line: median. Shading: 5th-95th percentiles. Dashed lines:",
                    "the 5-15 mg/L trough target of Yu 2025 Sect. 2.6.")
  )
```

The qualitative findings of Fig. 3 reproduce: at 24 kg the normal-renal-function
stratum sits inside the 5-15 mg/L trough target under both regimens, while both
impairment strata exceed the upper limit, and the q6h regimen gives the higher
trough of the two. The *magnitudes* in the impaired strata do not match Table 4;
that is dissected below.

## Clearance is a less-than-proportional function of eGFR

Yu 2025's Key Points name body weight and glomerular filtration rate as the two
important covariates on clearance. The eGFR exponent 0.812 being below 1 is what
makes the renal effect less than proportional.

```{r cl-vs-egfr}
egfr_grid <- seq(30, 261, by = 1)
tibble::tibble(
  CRCL = egfr_grid,
  `Yu 2025 (exponent 0.812)` = 8.22 * (24 / 70)^0.75 * (egfr_grid / 173.42)^0.812,
  `Proportional (exponent 1)` = 8.22 * (24 / 70)^0.75 * (egfr_grid / 173.42)
) |>
  tidyr::pivot_longer(-CRCL, names_to = "Scaling", values_to = "cl") |>
  ggplot(aes(CRCL, cl, linetype = Scaling)) +
  geom_line() +
  geom_vline(xintercept = 173.42, colour = "grey50") +
  labs(
    x = "Schwartz eGFR (mL/min/1.73 m^2)",
    y = "Clearance at WT 24 kg (L/h)",
    title = "Less-than-proportional eGFR effect on vancomycin clearance",
    caption = "Vertical line: the 173.42 mL/min/1.73 m^2 reference. Yu 2025 Table 3, Eq. 12."
  ) +
  theme(legend.position = "bottom")
```

# PKNCA validation

NCA is run over the steady-state 0-24 h window. The dose object declares
`duration = 1` so PKNCA treats the records as 60-minute infusions rather than
bolus doses.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment, renal)

# ss = 1 puts the steady-state trough on the time-zero record, so every subject
# already has a time-zero concentration; assert it rather than patching one in.
stopifnot(sum(sim_nca$time == 0) == 6L * n_per_arm)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment, renal)

conc_obj <- PKNCA::PKNCAconc(
  sim_nca, Cc ~ time | treatment + renal + id,
  concu = "mg/L", timeu = "h"
)
dose_obj <- PKNCA::PKNCAdose(
  dose_df, amt ~ time | treatment + renal + id,
  doseu = "mg", duration = 1
)

intervals <- data.frame(
  start = 0, end = 24,
  auclast = TRUE, cmax = TRUE, cmin = TRUE, cav = TRUE
)

nca_res <- suppressWarnings(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)

nca <- as.data.frame(nca_res$result) |>
  dplyr::filter(start == 0, end == 24) |>
  dplyr::select(id, treatment, renal, PPTESTCD, PPORRES)

stopifnot(dplyr::n_distinct(nca$id) == 6L * n_per_arm)
```

## Gate 1 - PKNCA AUC24 equals the analytic `Dose24 / CL`

For a linear one-compartment model at steady state, AUC over a full day is
exactly the daily dose divided by clearance, independent of the dosing interval
and of the infusion duration. Reproducing that from the ODE solution end to end
validates the structural model, the steady-state setup, and the NCA at once.

```{r gate-analytic}
auc24 <- nca |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, treatment, renal, AUC24 = PPORRES)

analytic <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::group_by(id) |>
  dplyr::summarise(dose_24h = sum(amt), .groups = "drop") |>
  dplyr::left_join(
    sim |> dplyr::group_by(id) |> dplyr::summarise(cl = dplyr::first(cl), .groups = "drop"),
    by = "id"
  ) |>
  dplyr::mutate(AUC24_analytic = dose_24h / cl)

gate1 <- auc24 |>
  dplyr::left_join(analytic, by = "id") |>
  dplyr::mutate(rel_err = AUC24 / AUC24_analytic - 1)

stopifnot(nrow(gate1) == 6L * n_per_arm, !anyNA(gate1$rel_err))
max_rel_err <- max(abs(gate1$rel_err))
max_rel_err
stopifnot(max_rel_err < 1e-3)
```

The worst per-subject discrepancy across all `r nrow(gate1)` subject-arms is
`r sprintf("%.4f%%", 100 * max_rel_err)`, which is the residual
lin-up/log-down interpolation error of the observation grid, not a model
discrepancy.

## Gate 2 - the two regimens deliver identical AUC24, subject by subject

20 mg/kg q8h and 15 mg/kg q6h are both 60 mg/kg/day. Because the model is linear
and both arms are the *same* cohort, every subject's AUC24 must be identical
between the two regimens - not merely similar in distribution. This is the
strongest available guard on the structural model and the steady-state setup,
because it would fail if either arm's dosing were mis-specified.

```{r gate-linearity}
gate2 <- auc24 |>
  dplyr::left_join(
    events |> dplyr::distinct(id, subject), by = "id"
  ) |>
  dplyr::select(subject, treatment, AUC24) |>
  tidyr::pivot_wider(names_from = treatment, values_from = AUC24) |>
  dplyr::mutate(rel_diff = `15 mg/kg q6h` / `20 mg/kg q8h` - 1)

stopifnot(nrow(gate2) == 3L * n_per_arm, !anyNA(gate2$rel_diff))
max_auc_diff <- max(abs(gate2$rel_diff))
max_auc_diff
stopifnot(max_auc_diff < 1e-3)
```

## Gate 3 - PKNCA `cmin` is the steady-state trough

At steady state every trough within the day is the same, so the `cmin` PKNCA
returns over 0-24 h must equal the concentration on the `ss = 1` time-zero
record. This gate ties the NCA summary used in the Table 4 comparison below back
to the raw solve.

```{r gate-cmin}
trough_ss <- sim |>
  dplyr::filter(time == 0) |>
  dplyr::select(id, subject, treatment, renal, trough_solve = Cc)

gate3 <- nca |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::select(id, cmin = PPORRES) |>
  dplyr::left_join(trough_ss, by = "id") |>
  dplyr::mutate(rel_err = cmin / trough_solve - 1)

stopifnot(nrow(gate3) == 6L * n_per_arm, !anyNA(gate3$rel_err))
max_cmin_err <- max(abs(gate3$rel_err))
max_cmin_err
stopifnot(max_cmin_err < 1e-8)
```

## Gate 4 - the simulated median trough is the typical-value trough

The trough is a monotone decreasing function of clearance, and `eta` has median
zero, so the median of the simulated cohort must converge to the typical-value
prediction at the stratum's median eGFR. This checks that the IIV was encoded on
the intended parameter and on the intended scale: an `omega` on the wrong scale
would not move the median, but an `eta` attached to the wrong parameter or a
mis-signed exponent would.

```{r gate-median}
typical_trough <- function(CRCL, mg_per_kg, tau) {
  ev <- rxode2::et(amt = mg_per_kg * 24, ii = tau, ss = 1, dur = 1, cmt = "central") |>
    rxode2::et(0, cmt = "central")
  out <- rxode2::rxSolve(
    mod, ev, params = c(WT = 24, CRCL = CRCL, etalcl = 0),
    omega = NA, returnType = "data.frame"
  )
  out$Cc[out$time == 0][1]
}

gate4 <- trough_ss |>
  dplyr::group_by(treatment, renal) |>
  dplyr::summarise(sim_median = median(trough_solve), .groups = "drop") |>
  dplyr::left_join(
    base_cohort |>
      dplyr::group_by(renal) |>
      dplyr::summarise(egfr_median = median(CRCL), .groups = "drop"),
    by = "renal"
  ) |>
  dplyr::left_join(regimens |> dplyr::select(treatment, mg_per_kg, tau), by = "treatment") |>
  dplyr::rowwise() |>
  dplyr::mutate(typical = typical_trough(egfr_median, mg_per_kg, tau)) |>
  dplyr::ungroup() |>
  dplyr::mutate(rel_diff = sim_median / typical - 1)

stopifnot(nrow(gate4) == 6L, !anyNA(gate4$rel_diff))
max_median_diff <- max(abs(gate4$rel_diff))
max_median_diff
# The tolerance is set by Monte Carlo noise, not by model error: the standard
# error of a log-normal median with omega = 0.565 and n = 200 is about 4.5%, so
# the worst of six arms is expected around 2 SE. The seed is fixed, so the
# realized value below is deterministic.
stopifnot(max_median_diff < 0.10)

gate4 |>
  dplyr::transmute(
    Regimen = treatment, `Renal stratum` = renal,
    `Stratum median eGFR` = round(egfr_median, 1),
    `Simulated median trough (mg/L)` = round(sim_median, 2),
    `Typical-value trough (mg/L)` = round(typical, 2),
    `% diff` = round(100 * rel_diff, 1)
  ) |>
  knitr::kable(
    caption = paste("Gate 4: the Monte Carlo median trough converges to the",
                    "typical-value prediction at each stratum's median eGFR.")
  )
```

# Comparison against the published Table 4 troughs

```{r table4-comparison}
reference_t4 <- tibble::tribble(
  ~treatment,      ~renal,                   ~cmin,
  "20 mg/kg q8h",  "Normal (eGFR >= 90)",    10.75,
  "20 mg/kg q8h",  "Mild (eGFR 60-89)",      19.35,
  "20 mg/kg q8h",  "Moderate (eGFR 30-59)",  29.87,
  "15 mg/kg q6h",  "Normal (eGFR >= 90)",    12.08,
  "15 mg/kg q6h",  "Mild (eGFR 60-89)",      22.16,
  "15 mg/kg q6h",  "Moderate (eGFR 30-59)",  32.24
)

simulated_t4 <- nca |>
  dplyr::filter(PPTESTCD == "cmin") |>
  dplyr::select(treatment, renal, PPTESTCD, PPORRES)

t4_table <- nlmixr2lib::ncaComparisonTable(
  simulated_t4, reference_t4,
  by     = c("treatment", "renal"),
  params = "cmin",
  units  = c(cmin = "mg/L")
)

knitr::kable(
  t4_table,
  caption = paste("Simulated versus Yu 2025 Table 4 median steady-state trough,",
                  "WT 24 kg. Simulated values are cohort medians.")
)
attr(t4_table, "footnote")
```

The normal-renal-function stratum reproduces within a few percent under both
regimens. Both impairment strata are over-predicted, by 30-51%.

## Table 4 does not follow from Eq. 12 for the impaired strata

The mismatch is *not* an artifact of the eGFR values chosen above. Inverting the
model - asking which eGFR would have to be assumed to reproduce each published
median - shows that the normal stratum's implied eGFR lands inside its stated
band while both impairment strata's implied values land *above* theirs.

```{r implied-egfr}
implied_egfr <- function(target, mg_per_kg, tau) {
  uniroot(function(g) typical_trough(g, mg_per_kg, tau) - target, c(10, 600))$root
}

implied <- reference_t4 |>
  dplyr::left_join(regimens |> dplyr::select(treatment, mg_per_kg, tau), by = "treatment") |>
  dplyr::left_join(strata, by = "renal") |>
  dplyr::rowwise() |>
  dplyr::mutate(egfr_implied = implied_egfr(cmin, mg_per_kg, tau)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    verdict = dplyr::case_when(
      egfr_implied > egfr_hi ~ "above band",
      egfr_implied < egfr_lo ~ "below band",
      TRUE                   ~ "inside band"
    )
  )

# The normal stratum reconciles; both impairment strata do not, and both fail in
# the same direction. Asserting the direction turns the finding into a guard: a
# future change to the covariate exponent or the reference eGFR would break it.
stopifnot(
  implied$verdict[implied$renal == "Normal (eGFR >= 90)"] == "inside band",
  implied$verdict[implied$renal != "Normal (eGFR >= 90)"] == "above band"
)

implied |>
  dplyr::transmute(
    Regimen = treatment, `Renal stratum` = renal,
    `Published median trough (mg/L)` = cmin,
    `Stated eGFR band` = sprintf("%g-%g", egfr_lo, egfr_hi),
    `Implied eGFR` = round(egfr_implied, 1),
    Verdict = verdict
  ) |>
  knitr::kable(
    caption = paste("eGFR that Eq. 12 would need in order to reproduce each",
                    "Yu 2025 Table 4 median trough at WT 24 kg.")
  )
```

Table 4 is internally consistent with the rest of the paper: its medians
reproduce the fold-increases quoted in Yu 2025 Sect. 4 (mild and moderate 0.8-
and 1.78-fold above normal at q8h; 0.83- and 1.67-fold at q6h) exactly to the
printed precision.

```{r table4-internal}
fold <- reference_t4 |>
  dplyr::group_by(treatment) |>
  dplyr::mutate(fold = cmin / cmin[renal == "Normal (eGFR >= 90)"] - 1) |>
  dplyr::ungroup() |>
  dplyr::filter(renal != "Normal (eGFR >= 90)")

published_fold <- c(0.80, 1.78, 0.83, 1.67)  # Yu 2025 Sect. 4, in the order below
stopifnot(max(abs(round(fold$fold, 2) - published_fold)) < 1e-8)

fold |>
  dplyr::transmute(
    Regimen = treatment, `Renal stratum` = renal,
    `Fold increase vs normal (from Table 4)` = round(fold, 3),
    `Quoted in Yu 2025 Sect. 4` = published_fold
  ) |>
  knitr::kable(caption = "Table 4 is internally consistent with the Sect. 4 narrative.")
```

So the discrepancy is between Table 4 and Eq. 12, not inside Table 4. Solving
for the eGFR exponent that *would* reproduce Table 4's renal ratios at the
stratum midpoints localizes it to that one parameter.

```{r implied-exponent}
typical_trough_exp <- function(CRCL, mg_per_kg, tau, e_crcl) {
  ev <- rxode2::et(amt = mg_per_kg * 24, ii = tau, ss = 1, dur = 1, cmt = "central") |>
    rxode2::et(0, cmt = "central")
  out <- rxode2::rxSolve(
    mod, ev,
    params = c(WT = 24, CRCL = CRCL, etalcl = 0, e_crcl_cl = e_crcl),
    omega = NA, returnType = "data.frame"
  )
  out$Cc[out$time == 0][1]
}

midpoints <- c(`Normal (eGFR >= 90)` = (90 + 260.82) / 2,
               `Mild (eGFR 60-89)` = 74.5,
               `Moderate (eGFR 30-59)` = 44.5)

published_ratio <- function(stratum, treatment) {
  num <- reference_t4$cmin[reference_t4$renal == stratum &
                             reference_t4$treatment == treatment]
  den <- reference_t4$cmin[reference_t4$renal == "Normal (eGFR >= 90)" &
                             reference_t4$treatment == treatment]
  if (length(num) != 1L || length(den) != 1L) {
    stop("no unique Table 4 row for '", stratum, "' / '", treatment, "'")
  }
  num / den
}

exponent_for <- function(stratum, treatment, mg_per_kg, tau) {
  ratio_target <- published_ratio(stratum, treatment)
  uniroot(function(e) {
    typical_trough_exp(midpoints[[stratum]], mg_per_kg, tau, e) /
      typical_trough_exp(midpoints[["Normal (eGFR >= 90)"]], mg_per_kg, tau, e) -
      ratio_target
  }, c(0.05, 1.5))$root
}

impaired <- c("Mild (eGFR 60-89)", "Moderate (eGFR 30-59)")

implied_exponents <- regimens |>
  dplyr::select(treatment, mg_per_kg, tau) |>
  tidyr::expand_grid(Stratum = impaired) |>
  dplyr::rowwise() |>
  dplyr::mutate(
    Exponent = exponent_for(Stratum, treatment, mg_per_kg, tau)
  ) |>
  dplyr::ungroup() |>
  dplyr::select(Regimen = treatment, Stratum, Exponent)

# Every implied exponent sits well below the Table 3 estimate of 0.812, and at
# or below the lower bootstrap bound of 0.55.
stopifnot(all(implied_exponents$Exponent < 0.7))

implied_exponents |>
  dplyr::mutate(
    `Implied eGFR exponent` = round(Exponent, 3),
    `Yu 2025 Table 3 estimate` = 0.812,
    `Table 3 bootstrap 2.5-97.5th` = "0.55-1.06",
    Exponent = NULL
  ) |>
  knitr::kable(
    caption = paste("eGFR exponent that would reproduce the Table 4 renal ratios",
                    "at the stratum midpoints, versus the Table 3 estimate.")
  )
```

Every implied exponent falls in the range 0.53-0.61, against the Table 3
estimate of 0.812 and its bootstrap interval of 0.55-1.06. In other words,
Table 4's renal-impairment effect is weaker than the model the paper reports. The
most likely proximate cause is that the Monte Carlo drew covariates "using the
initial dataset" (Sect. 2.6) rather than from the stated bands, so the effective
eGFR in each impaired stratum was higher than the band implies - the cohort has
only a handful of patients below 90 mL/min/1.73 m^2 (the Table 1 minimum is
31.9), and resampling a sparse tail will pull the stratum's realized eGFR
upward. The paper reports neither the per-stratum eGFR distribution nor the
simulation dataset, so this cannot be settled from the published record.

**Practical consequence.** The packaged model is faithful to Yu 2025 Table 3 and
Eqs. 12-13, which are the model. A user reproducing Table 4 or Fig. 3 for the
renal-impairment strata should expect troughs 30-51% higher than the paper
printed, and should treat the paper's dose-reduction suggestions in Sect. 4
(15 mg/kg q12h and 10 mg/kg q8h for moderate impairment) as conservative
relative to what Eq. 12 actually predicts.

# Assumptions and deviations

* **eGFR values within each renal stratum are assumed, not published.** Yu 2025
  Sect. 2.6 gives only the stratum bands and says 1000 virtual patients were
  simulated "using the initial dataset". This vignette samples eGFR uniformly
  across each stated band (normal capped at the Table 1 maximum of 260.82). The
  choice was fixed before any comparison was run.
* **`omega_CL` = 56.5% is read as `100 * omega`, giving omega^2 = 0.319225.**
  The paper labels the row with the symbol `omega_CL` and never defines a %CV
  transform. The alternative reading, `%CV = sqrt(exp(omega^2) - 1) * 100`,
  would give omega = 0.5265 instead of 0.565 - a 7% difference on the standard
  deviation that changes no conclusion here (it affects only the width of the
  simulated intervals, not any median). What *is* ruled out is the reported
  number being a variance: the asymptotic relative standard error of an
  estimated variance is at least `sqrt(2/N)` = 14.1% for N = 100 subjects,
  whereas Table 3 reports an RSE of 10%, which is only attainable on the
  standard-deviation scale.
* **Table 4's dispersion is labelled "IQR" but is far too wide for one.** The
  normal-stratum q8h entry is 10.75 (2.12-30.28), a 14-fold span; a log-normal
  with omega = 0.565 has an IQR spanning only 2.2-fold. Fig. 3's caption says
  its shading is the 5th-95th percentile, so the Table 4 parenthetical is
  probably a 5th-95th percentile range (or a range that also carries the
  residual error and the within-stratum eGFR spread). No assertion in this
  vignette depends on it, and only the medians are compared.
* **No residual error is applied in the simulations.** Yu 2025 Sect. 2.6 states
  that interindividual variability (`omega_CL`) was included in the Monte Carlo
  and does not mention the residual error. Since `propSd` was fixed rather than
  estimated, and since the trough medians compared here are unaffected by a
  zero-median multiplicative error, leaving it out is the cleaner comparison.
* **Steady state is imposed with `ss = 1` rather than by burn-in.** For the
  moderate-impairment stratum the terminal half-life at 24 kg reaches roughly 22
  h, so a burn-in would need several simulated days to converge; `ss = 1` gives
  the exact steady state directly.
* **The model is used only within the paper's stated range of validity.** Yu
  2025 Sect. 4 recommends against predicting concentrations above 20 mg/L,
  because the VPC's 95th percentile was poorly reproduced from sparse
  high-concentration data. The impaired-stratum profiles above sit above that
  threshold and are shown to replicate Fig. 3, not as endorsed predictions.
* **The paper reports no NCA parameters.** There is no published Cmax, Tmax,
  AUC, or half-life to compare against, so the PKNCA section validates against
  analytic identities (`AUC24 = Dose24 / CL`, dose linearity, `cmin` equal to
  the steady-state trough) and the published Table 4 trough medians instead.
* **No supplement, erratum, or corrigendum was located.** The article is open
  access with no supplementary material; a search of the journal landing page
  and PubMed found no correction notice as of the extraction date.
