---
title: "Zecchin_2016_tumorovarian"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zecchin_2016_tumorovarian}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Zecchin C, Gueorguieva I, Enas NH, Friberg LE. Models for change in tumour size, appearance of new lesions and survival probability in patients with advanced epithelial ovarian cancer. *Br J Clin Pharmacol*. 2016;82(3):717-727. doi:[10.1111/bcp.12994](https://doi.org/10.1111/bcp.12994). PubMed: [PMID 27136318](https://pubmed.ncbi.nlm.nih.gov/27136318/).
* DDMORE Foundation Model Repository entry: [DDMODEL00000217](https://repository.ddmore.eu/model/DDMODEL00000217). Source bundle in this extraction came from the GitHub mirror [`dpastoor/ddmore_scraping/217/`](https://github.com/dpastoor/ddmore_scraping/tree/master/217).

# Population

The Zecchin 2016 SLD model was estimated from 336 patients in a randomised Phase III trial comparing carboplatin monotherapy with carboplatin + gemcitabine combination chemotherapy in advanced (FIGO stage III/IV) epithelial ovarian cancer. The DDMORE bundle's `.lst` listing confirms the dataset size: `TOT. NO. OF INDIVIDUALS: 336` and `TOT. NO. OF OBS RECS: 1358`. Detailed baseline demographics (median age, weight, region) were not reproduced here because the linked publication PDF was not on disk for this extraction; the population block reports only what the PubMed abstract confirms.

The same information is available programmatically via the model's
`population` metadata at `readModelDb("Zecchin_2016_tumorovarian")$meta$population`
(after parsing through `rxode2::rxode()`).

# Source trace

Equations were extracted from `Executable_SLD.mod` (`$PK` and `$DES` blocks) and parameter values from the `FINAL PARAMETER ESTIMATE` block of `Output_real_SLD.lst`. The `.mod` `$THETA` / `$OMEGA` / `$SIGMA` lines hold initial values, not final estimates, and were not used for parameter values. The linked publication (Zecchin et al. 2016 BJCP, doi:10.1111/bcp.12994) was not on disk for this extraction, so per-table cross-checks against the publication were not performed; see the *Assumptions, deviations, and Errata* section below.

| Equation / parameter           | Value                          | Source location                                                                                |
|--------------------------------|--------------------------------|------------------------------------------------------------------------------------------------|
| `d/dt(tumorSize)`              | n/a                            | `Executable_SLD.mod` `$DES` (`DADT(1) = KG/1000 * A(1) - (KD0/1000 * E0 + KD1/100 * E1) * A(1)`) |
| Initial tumour size            | `ibase * 1000` mm              | `Executable_SLD.mod` `$PK` (`A_0(1) = IBASE*1000`)                                            |
| `lkg` (tumour growth rate)     | `log(0.611)`                   | `Output_real_SLD.lst` FINAL TH1                                                               |
| `lkd0` (carboplatin death)     | `log(0.0497)`                  | `Output_real_SLD.lst` FINAL TH2                                                               |
| `lkd1` (gemcitabine death)     | `log(0.0164)`                  | `Output_real_SLD.lst` FINAL TH3                                                               |
| `libase` (baseline SLD, m)     | `log(0.0713)`                  | `Output_real_SLD.lst` FINAL TH4                                                               |
| `addSd_tumorSize` (mm)         | `18.4`                         | `Output_real_SLD.lst` FINAL TH5                                                               |
| `etalkg`                       | `1.72`                         | `Output_real_SLD.lst` FINAL OMEGA(1,1)                                                        |
| `etalkd0_kd1` (shared)         | `1.09`                         | `Output_real_SLD.lst` FINAL OMEGA(2,2) — shared by KD0 and KD1 per `$PK` block                |
| `etalibase`                    | `0.515`                        | `Output_real_SLD.lst` FINAL OMEGA(3,3)                                                        |

# Virtual cohort and event table

The DDMORE bundle ships a `Simulated_SLD.csv` event dataset (336 subjects, ~16 records per subject, structured as cycle-start exposure events plus per-cycle SLD observations). Per the skill's DDMORE-source guidance, that file is "intentionally minimal" and not directly representative of the publication's clinical study; this vignette therefore builds a small virtual cohort that mimics the trial regimen at the structural level and lets the model generate the trajectories.

* 6 chemotherapy cycles, 21 days apart (`Q3W` carboplatin pacing — the standard advanced-ovarian-cancer regimen referenced by the DDMORE bundle).
* Two equally-allocated arms: carboplatin monotherapy (`AUC_GEM = 0`) and carboplatin + gemcitabine combination.
* Per-cycle carboplatin AUC drawn from a lognormal centred on the source-dataset median scale; per-cycle gemcitabine AUC drawn from a lognormal centred on its source-dataset scale (combination arm only). The numerical values absorb the source dataset's implicit unit conventions, which the Zecchin model then re-scales internally via the `/1000` (carboplatin and growth) and `/100` (gemcitabine) factors carried verbatim from the source `$DES` block.
* SLD observations every 21 days from baseline through day 252 (12 visits in total: 6 on-treatment + 6 follow-up).

```{r cohort}
set.seed(2025)

n_per_arm  <- 25L
cycle_days <- 21L
n_cycles   <- 6L

auc_carbo_typical <- 5     # source-dataset scale (median-ish AUC_CARBO)
auc_gem_typical   <- 10    # source-dataset scale (median-ish AUC_GEM, combo arm)

make_subject <- function(id, arm) {
  cycle_carbo <- rlnorm(n_cycles, log(auc_carbo_typical), 0.3)
  cycle_gem   <- if (arm == "combo") {
    rlnorm(n_cycles, log(auc_gem_typical), 0.3)
  } else {
    rep(0, n_cycles)
  }

  cycle_starts  <- seq(0L, by = cycle_days, length.out = n_cycles)
  followup_days <- seq(n_cycles * cycle_days, by = cycle_days, length.out = 7L)

  # Cycle-start records carry that cycle's per-cycle average AUCs; rxode2's
  # locf semantics on time-varying covariates then keep those AUC values
  # constant during ODE integration through the next cycle boundary.
  cycle_records <- tibble(
    id = id,
    time = cycle_starts,
    evid = 0L,
    AUC_CARBO = cycle_carbo,
    AUC_GEM   = cycle_gem
  )
  followup_records <- tibble(
    id = id,
    time = followup_days,
    evid = 0L,
    AUC_CARBO = 0,
    AUC_GEM   = 0
  )
  bind_rows(cycle_records, followup_records)
}

events <- bind_rows(
  lapply(seq_len(n_per_arm), function(i) make_subject(i, arm = "mono")),
  lapply(seq_len(n_per_arm), function(i) make_subject(i + n_per_arm, arm = "combo"))
) |>
  mutate(arm = if_else(id <= n_per_arm, "carboplatin alone", "carbo + gem"))

# Disjoint-id sanity check
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
head(events, 15)
```

# Typical-value simulation (replicates published mechanism)

A typical-value simulation (no IIV, no residual error) shows the deterministic SLD trajectory the model implies under each arm. This is the primary mechanistic-sanity check: does an "average" patient on combination chemotherapy have a faster SLD decline than a patient on carboplatin alone, and does SLD start rising again once treatment ends?

```{r simulate-typical}
mod <- readModelDb("Zecchin_2016_tumorovarian")
mod_typical <- mod |> rxode2::zeroRe()
sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events,
  keep   = c("arm")
) |>
  as.data.frame()

# Population-typical median trajectories per arm
typical_summary <- sim_typical |>
  group_by(arm, time) |>
  summarise(median_SLD = median(tumorSize), .groups = "drop")

ggplot(typical_summary, aes(time, median_SLD, colour = arm)) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = n_cycles * cycle_days, linetype = "dashed",
             colour = "grey60") +
  annotate("text", x = n_cycles * cycle_days + 3, y = max(typical_summary$median_SLD),
           label = "treatment ends", hjust = 0, size = 3.2, colour = "grey40") +
  labs(x = "Time (days)", y = "Typical SLD (mm)",
       title = "Mechanistic sanity: typical-value SLD trajectory by arm",
       caption = "Zecchin 2016 SLD model with no IIV / no residual error.") +
  theme_minimal()
```

The expected qualitative pattern holds:

* **Tumour shrinks during treatment.** Both arms show a downward trajectory while AUC_CARBO and AUC_GEM are non-zero (cycles 1-6, days 0-126). The combination arm declines faster than the carboplatin-only arm because both cytotoxic-death terms contribute additively (`kd0 * AUC_CARBO + kd1 * AUC_GEM`, `$DES` block).
* **Tumour regrows after treatment ends.** Once the AUC inputs drop to zero (after day 126), the death terms vanish and the growth term `kg * tumorSize` takes over, producing exponential regrowth. This matches the model's structural assumption of no resistance.
* **Initial value tracks `IBASE * 1000`.** At time 0 the trajectory begins at `exp(libase) * 1000` ≈ 71 mm, exactly the published baseline-SLD scale.

# Stochastic simulation (between-subject variability)

Including the model's IIV (typical of the source dataset's variability — `etalkg ~ 1.72`, `etalkd0_kd1 ~ 1.09`, `etalibase ~ 0.515`) shows the population spread of trajectories. The shared `etalkd0_kd1` introduces strong correlation between an individual's response to carboplatin and their response to gemcitabine, which is the source paper's mechanistic assumption (a single underlying "drug-sensitivity" scale per subject).

```{r simulate-iiv}
sim_iiv <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("arm")
) |>
  as.data.frame()

iiv_summary <- sim_iiv |>
  group_by(arm, time) |>
  summarise(
    Q05 = quantile(tumorSize, 0.05, na.rm = TRUE),
    Q50 = quantile(tumorSize, 0.50, na.rm = TRUE),
    Q95 = quantile(tumorSize, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(iiv_summary, aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 1) +
  geom_vline(xintercept = n_cycles * cycle_days, linetype = "dashed",
             colour = "grey60") +
  labs(x = "Time (days)", y = "SLD (mm)",
       title = "Stochastic SLD trajectory by arm (median + 5-95% interval)",
       caption = "50 virtual subjects (25 per arm). Random-effect variances from Output_real_SLD.lst FINAL OMEGA.") +
  theme_minimal()
```

The 5th-95th percentile bands are wide — consistent with the very large between-subject variability the source NONMEM run reports (CV on KG > 200% on the lognormal scale, with ETA-shrinkage of 55% on KG and 28% on KD; see *Assumptions and deviations* below for context).

# Mechanistic checks: dose-AUC-effect linearity

A second sanity check: at fixed tumour size, the death rate is linear in `AUC_CARBO` and `AUC_GEM`. Doubling the per-cycle AUC should approximately double the per-cycle relative SLD reduction (for moderate cycle counts where the exponential-decay approximation is valid).

```{r mechanism-linearity}
linear_check <- function(scale_factor) {
  events_scaled <- events |>
    mutate(AUC_CARBO = AUC_CARBO * scale_factor)
  rxode2::rxSolve(
    mod_typical,
    events = events_scaled,
    keep   = c("arm")
  ) |>
    as.data.frame() |>
    filter(time == n_cycles * cycle_days, arm == "carboplatin alone") |>
    summarise(SLD_at_end_of_treatment = median(tumorSize),
              scale_factor = scale_factor)
}

linear_table <- bind_rows(lapply(c(0, 0.5, 1, 2), linear_check))
knitr::kable(
  linear_table,
  caption = "Median typical SLD (carboplatin-alone arm) at the end of cycle 6 under different AUC_CARBO scaling factors. Larger drug exposure produces a smaller end-of-treatment SLD (zero-exposure recovers the pure-growth trajectory)."
)
```

# Validation strategy

The validation strategy for this DDMORE-source extraction is **mechanistic_sanity** (per the per-task validation flag `validation:mechanistic_sanity`). Per `references/ddmore-source.md` § *Validation strategy by model type*, the relevant checklist is `verification-checklist.md` § F.3 (count / hazard / typical-trajectory) plus the F.2 self-consistency substitute, since the linked Zecchin 2016 publication was not directly on disk for this extraction:

* **F.3 typical-value mechanistic check.** The deterministic SLD trajectory matches the model's mechanism (decline during treatment, regrowth after treatment ends, additive on the two drug effects, baseline = `IBASE * 1000`).
* **F.2 self-consistency.** The model parses (`buildModelDb()` + `checkModelConventions()` clean), `rxode2::rxSolve()` runs to completion across the 50-subject virtual cohort with no NaNs or negative SLD values across the 252-day window, and the simulated trajectories visually match the qualitative shape one would expect from the source `Executable_SLD.mod` `$DES` equation.

A side-by-side comparison against the bundle's `Output_simulated_SLD.lst` IPRED trajectory or against the published Zecchin 2016 figures is **not** included because the simulated-listing per-subject IPRED tables are not in the bundle and the publication PDF was not on disk for this extraction.

# Assumptions, deviations, and Errata

* **`checkModelConventions()` deviations.** `nlmixr2lib::checkModelConventions("Zecchin_2016_tumorovarian")` reports three warnings (no errors): (1) the `tumorSize` compartment is not on the canonical-compartment list; (2) the single-output observation variable should be `Cc`; (3) `units$concentration` does not contain `/` (the mass/volume divider). All three are intrinsic to a tumour-size-dynamics model and apply to every existing tumour-size model in the package (e.g., `tgi_no_sat_expo`, `oncology_xenograft_simeoni_2004`): SLD is a **length** (mm), not a drug concentration; the `Cc` canonical name is reserved for plasma drug concentrations and does not fit a tumour-size endpoint; the convention checker's mass/volume unit pattern is also drug-concentration-specific. Renaming `tumorSize` to `Cc` here would mislead readers about what the observation actually represents. The deviations are intentional.
* **`MINIMIZATION SUCCESSFUL` with caveats.** The source `.lst` reports `MINIMIZATION SUCCESSFUL` but adds `HOWEVER, PROBLEMS OCCURRED WITH THE MINIMIZATION. REGARD THE RESULTS OF THE ESTIMATION STEP CAREFULLY, AND ACCEPT THEM ONLY AFTER CHECKING THAT THE COVARIANCE STEP PRODUCES REASONABLE OUTPUT.` The reported correlation matrix shows THETA-THETA correlations near `±0.98-1.00`, indicating near-non-identifiability of the structural parameters. ETA-shrinkage on KG is `55.3%` and on the shared KD eta is `28.2%`. This is consistent with a small-N (336 patients), short-follow-up SLD model trying to estimate three structural parameters plus three random-effect variances simultaneously. The values are extracted as published; **the high correlations and shrinkage should be borne in mind when interpreting individual parameter estimates from this model.**
* **No external publication-table cross-check.** The publication PDF (Zecchin 2016, BJCP, doi:10.1111/bcp.12994) was not on disk for this extraction. Parameter values come exclusively from `Output_real_SLD.lst`; the `Model_Accomodations.text` file that DDMORE bundles usually ship was also not present in this directory. Citation metadata was confirmed via PubMed E-utilities (PMID 27136318).
* **No in-bundle `Model_Accomodations.text`.** The DDMORE bundle for `DDMODEL00000217` does not ship a `Model_Accomodations` reference file (one of the optional bundle artefacts described in `references/ddmore-source.md`). The publication mapping used here was reconstructed from the PubMed citation match for the task metadata.
* **M3 censored-likelihood handling omitted.** The source `$ERROR` block uses the M3 method to handle below-LLOQ SLD observations (`LLOQ = 5 mm`, `BQL` indicator column, `F_FLAG = 1` censoring branch with `Y = PHI((LLOQ - IPRED)/W)`). Forward simulation does not exercise the censoring likelihood, so the nlmixr2 translation drops the M3 branch and uses only the open-likelihood `Y = IPRED + ERR(1)*W` branch. Re-fitting this model in nlmixr2 with M3-style censoring would require explicit `cens()` handling on the dataset and is outside the scope of this packaged-model extraction.
* **Backward-interpolation of cycle exposures.** The source `$PK` block uses NONMEM `NEWIND` carry-forward logic (`OCB := CB`, `E0 := OCB`) to make the per-record `E0` value equal to the **previous** record's `CB` (and analogously for `E1` / `G`). The functional consequence is that `E0` is constant within each cycle, equal to the AUC value at that cycle's start. The vignette's virtual-cohort event table reproduces the same step-function pattern via cycle-start records that hold the new cycle's AUC value, which rxode2 carries forward via the default `locf` semantics for time-varying covariates.
* **Internal `/1000` and `/100` numerical scalings.** The source `$DES` block divides `KG` and `KD0` by 1000 and `KD1` by 100 inside the rate equation. These factors absorb the unit conventions of the source dataset's `CB` and `G` exposure columns (described in the `.mod` `$INPUT` comments as `ng/dL*day/n days in cycle` and `mol/10^6 cells*day/n days in cycle`, respectively). They are preserved verbatim in the rxode2 translation so the model reproduces the source NONMEM `FINAL` parameter values without re-scaling.
* **Virtual-cohort exposure values.** The vignette's per-cycle `AUC_CARBO` and `AUC_GEM` draws are centred on values that match the source-dataset distribution (median around 5 and 10 respectively in the `Simulated_SLD.csv` shipped with the bundle), not on standard clinical-trial AUC reporting units (e.g., carboplatin AUC `4-6 mg*min/mL` per the Calvert formula). The numerical values used are interpretable only in conjunction with the model's internal `/1000` and `/100` rescaling factors; downstream use of this model with a custom dataset must supply `AUC_CARBO` / `AUC_GEM` on the same source-dataset scale, not in raw clinical-trial AUC units.
* **No patient-level covariates.** The source final-model `$INPUT` and `$PK` blocks include no patient-level covariates beyond the time-varying drug-AUC inputs (no allometric weight scaling, no sex, no ECOG, no liver-metastasis indicator). The Zecchin 2016 paper title mentions liver-metastasis effects but those operate on the survival / new-lesion sub-models (DDMODEL00000218, the companion task), not on the SLD model captured here.
