---
title: "Remdesivir and its metabolites in renal impairment (Zhang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Remdesivir and its metabolites in renal impairment (Zhang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhang_2025_remdesivir"))$reference`
* Article: <https://doi.org/10.3389/fphar.2025.1488961>
* Models contributed by this paper:
  * `Zhang_2025_remdesivir` -- the healthy-control / renal-impairment cohort fit.
  * `Zhang_2025_remdesivir_esrd` -- the single anuric (eGFR = 0) patient fit.

Zhang 2025 builds a six-state cascade model for intravenous remdesivir (RDV,
GS-5734) and its two circulating metabolites: the alanine intermediate
GS-704277 (the paper's "IM") and the nucleoside GS-441524 (the paper's "NUC").
Each analyte occupies a central and a peripheral compartment, and -- this is the
paper's distinguishing structural claim -- metabolism proceeds down the cascade
in **both** compartments, not only centrally. The authors introduced the
peripheral metabolic route specifically to capture the slow second decay phase
of RDV and GS-704277 that a conventional two-compartment model could not
reproduce (Zhang 2025 Figure 6 and Discussion).

```{r structure, echo = FALSE}
ui_cohort <- rxode2::rxode(readModelDb("Zhang_2025_remdesivir"))
ui_esrd <- rxode2::rxode(readModelDb("Zhang_2025_remdesivir_esrd"))
cat(paste(ui_cohort$state, collapse = "\n"))
```

## Why the model is written in concentration space

Every parameter in Zhang 2025 Table 1 carries units of **1/h**, and Equations
1-6 are written on concentrations rather than amounts. There is therefore **no
volume of distribution anywhere in the published model**, and the paper reports
none. Two consequences follow, and both are load-bearing for anything done
below:

1. Each state in the packaged model holds a **concentration in ng/mL**, not an
   amount. The observation equations are simply `Cc <- central` and so on, with
   no division by a volume.
2. A dose cannot be supplied as a mass. It enters as the **initial central
   remdesivir concentration**. Converting a 100 mg or 200 mg intravenous dose
   into that initial concentration requires a volume the paper does not
   publish, so this vignette works throughout in a *relative* scale: every
   simulation starts from a reference initial concentration of 1000 ng/mL, and
   every quantity validated below is invariant to that choice.

That invariance is not a convenience -- it is a property of the model. The
system is entirely **linear**, so scaling the initial condition scales all six
state trajectories by the same factor. Times to peak, ratios between analytes,
and terminal slopes are therefore all independent of the unpublished volume,
and they are exactly the quantities compared against the paper's figures.

# Population

The cohort model was fitted to the Gilead phase I, open-label, parallel-group
renal-impairment study that Zhang 2025 cites as "Zhang et al. (2020)": adults
with mild (n = 12), moderate (n = 11) or severe (n = 10) renal impairment, plus
kidney failure (n = 6 on dialysis, n = 4 without), against matched healthy
controls. Impairment was graded by eGFR: mild 60-89, moderate 30-59, severe
15-29 and kidney failure < 15 mL/min/1.73 m^2. Single intravenous doses were
assigned by stratum -- 100 mg for mild or moderate impairment, 40 mg for severe
impairment or predialysis kidney failure, and 20 mg for postdialysis or
non-dialysis kidney failure (Zhang 2025 Results 3.1).

The ESRD model was fitted to a single kidney-transplant recipient in his
mid-seventies with COVID-19, anuric (eGFR 0 mL/min) and on intermittent
haemodialysis, from the case report of Soergel 2021
(<https://doi.org/10.1093/jac/dkaa500>). He received the standard 5-day
regimen -- 200 mg on day 1 then 100 mg daily -- and Zhang 2025 fitted **day 1
only**.

**A provenance limitation that governs how much the fitted variability can
mean.** Zhang 2025 had no individual patient data. Methods 2.1 states that the
published profiles "were digitized using the open-source software Engauge
Digitizer (Version 12.1)" and that "Due to the absence of individual
patient-level data, only the mean drug concentration values from the profile
curves were extracted." The mixed-effects fit therefore ran against *digitised
mean curves*, one per analyte per stratum, not against subject-level
observations.

```{r population}
str(readModelDb("Zhang_2025_remdesivir")()$population)
```

# Source trace

Per-parameter provenance is recorded as an in-file comment beside each `ini()`
entry in `inst/modeldb/specificDrugs/Zhang_2025_remdesivir.R` and
`inst/modeldb/specificDrugs/Zhang_2025_remdesivir_esrd.R`. Collected here for
review. All nine structural parameters come from a single table; Zhang 2025
reports **point estimates only**, with no standard errors, RSEs or confidence
intervals anywhere.

| Model element | Paper symbol | Healthy control | Severe renal impairment | eGFR = 0 | Source |
|---|---|---|---|---|---|
| `lk12` | Q_RDV | 0.19 | 0.19 | 0.31 | Table 1 |
| `lk12_gs704277` | Q_IM | 12.53 | 12.33 | 83.8 | Table 1 |
| `lk12_gs441524` | Q_NUC | 0.039 | 0.038 | 50.02 | Table 1 |
| `lkmet_gs704277_central` | K_C,IM | 0.22 | 0.19 | 0 | Table 1 |
| `lkmet_gs704277_peripheral1` | K_P,IM | 0.31 | 0.19 | 0.22 | Table 1 |
| `lkmet_gs441524_central` | K_C,NUC | 0.38 | 0.85 | 6.27 | Table 1 |
| `lkmet_gs441524_peripheral1` | K_P,NUC | 2.44 | 0.5 | 0.15 | Table 1 |
| `lkmet_gs443902_peripheral1` | K_P,NTP | 3.28E10 | 8.08E10 | 0.037 | Table 1 |
| `lkel` | CL_C,RDV | 2.99 | 2.99 | 2.3 | Table 1 |
| `d/dt(central)` | Equation 1 | n/a | n/a | n/a | Methods 2.2 |
| `d/dt(peripheral1)` | Equation 2 | n/a | n/a | n/a | Methods 2.2 |
| `d/dt(central_gs704277)` | Equation 3 | n/a | n/a | n/a | Methods 2.2 |
| `d/dt(peripheral1_gs704277)` | Equation 4 | n/a | n/a | n/a | Methods 2.2 |
| `d/dt(central_gs441524)` | Equation 5 | n/a | n/a | n/a | Methods 2.2 |
| `d/dt(peripheral1_gs441524)` | Equation 6 | n/a | n/a | n/a | Methods 2.2 |
| Cascade topology | Figure 2 | n/a | n/a | n/a | Figure 2 |
| `RENALIMP` covariate | "Renal Impairment" | 0 | 1 | n/a | Methods 2.3, Results 3.1 |
| Residual error form | Equation 7 | not reported | not reported | not reported | Methods 2.3 |

The equations are vector graphics in the published PDF and are dropped by
markdown conversion (they appear as `formula-not-decoded`). They were recovered
with `pdftotext -layout` and independently cross-checked against the arrow
topology of Figure 2, which shows K_P,IM and K_C,IM feeding GS-704277 from the
peripheral and central remdesivir pools respectively, K_P,NUC and K_C,NUC
feeding GS-441524 from the corresponding GS-704277 pools, K_P,NTP draining
peripheral GS-441524, the bidirectional Q_i exchanges, and CL_RDV leaving
central remdesivir only.

Two asymmetries in the published system are deliberate and are reproduced
verbatim: **remdesivir is eliminated only from its central compartment**
(Equation 2 carries no elimination term), and **GS-441524 is removed only from
its peripheral compartment**, by conversion to the triphosphate GS-443902
(Equation 5 carries no elimination term). Neither metabolite has a renal or
other clearance term of its own.

# Virtual cohort

Neither model carries between-subject variability -- Zhang 2025 declares
lognormal random effects on all parameters (Methods 2.3) but publishes no
variance, SD or CV% for any of them, and there is no supplement to fall back on.
Both models are therefore **typical-value, fully deterministic** simulations,
and the "cohort" below is one representative profile per arm. Because nothing
is drawn at random, no seed is needed and no result depends on the solver
thread count.

```{r cohort}
# Reference initial central remdesivir concentration. Arbitrary: the published
# model has no volume, so absolute concentrations are determined only up to
# this scale factor. Every validation quantity below is scale-invariant.
C0 <- 1000

arms <- tibble::tibble(
  arm      = c("Control", "Severe renal impairment", "eGFR = 0 (ESRD)"),
  model    = c("Zhang_2025_remdesivir", "Zhang_2025_remdesivir", "Zhang_2025_remdesivir_esrd"),
  RENALIMP = c(0, 1, NA),
  tmax_sim = c(200, 260, 30)
)

make_events <- function(tmax_sim, id) {
  dose <- rxode2::et(amt = C0, cmt = "central")
  # Observations are placed on the ENDPOINT compartment `Cc`. For this model
  # that is correct rather than the pattern-2 anti-pattern: `Cc` is a declared
  # endpoint (`Cc ~ add(addSd) + prop(propSd)`), so rxode2 already owns a
  # compartment slot for it and nothing is renumbered. rxode2 returns all
  # three observables as columns at those rows.
  rxode2::et(dose, seq(0, tmax_sim, length.out = 4000), cmt = "Cc") |>
    as.data.frame() |>
    dplyr::mutate(id = id)
}
```

# Simulation

```{r simulate}
solve_arm <- function(i, atol = NULL, rtol = NULL) {
  ev <- make_events(arms$tmax_sim[i], id = i)
  args <- list(
    rxode2::rxode(readModelDb(arms$model[i])),
    events = ev,
    returnType = "data.frame",
    # rxode2's automatic ODE -> linCmt conversion corrupts the endpoint
    # mapping for this six-state / three-endpoint model.
    useLinCmt = FALSE
  )
  if (!is.na(arms$RENALIMP[i])) args$params <- c(RENALIMP = arms$RENALIMP[i])
  if (!is.null(atol)) { args$atol <- atol; args$rtol <- rtol }
  out <- do.call(rxode2::rxSolve, args)
  out$arm <- arms$arm[i]
  out$id <- i
  out
}

sim <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), solve_arm))
sim$arm <- factor(sim$arm, levels = arms$arm)
```

Observation rows above carry `cmt = "Cc"`. For a model whose observables are
*declared endpoints* that is legitimate and not the compartment-renumbering
anti-pattern: rxode2 already owns a slot for `Cc`, so nothing is injected. It
is also the only form that works here -- with three endpoints, an observation
row on an ODE state (`cmt = "central"`) or carrying only a `dvid` is rejected
with `'dvid'->'cmt' on observation record ... undefined compartment`. The guard
below turns that reasoning into a mechanical check rather than a claim: if a
slot had been injected ahead of the ODE states, the dose would no longer land
in `central` and all six state columns would not survive.

```{r no-renumbering-guard}
stopifnot(
  # The dose landed in `central`, i.e. compartment 1 was not displaced.
  all(sapply(split(sim$central, sim$arm), function(x) isTRUE(all.equal(x[1], C0)))),
  # All six ODE states are still present and distinct from the observables.
  all(ui_cohort$state %in% names(sim)),
  # And the observables were computed.
  all(c("Cc", "Cc_gs704277", "Cc_gs441524") %in% names(sim))
)
```

The `K_P,NTP` estimate in the cohort model is 3.28e10 /h, which makes the
system formally stiff. It is transcribed exactly as printed, and it is **not** a
typographical error for 3.28e-10: at the negative exponent the peripheral
GS-441524 pool is never drained, central GS-441524 rises monotonically for a
week and never peaks, and Zhang 2025 Figure 3c -- which shows a clear peak
followed by decline -- cannot be reproduced at all. LSODA handles the printed
value without difficulty, as the check below confirms.

```{r solver-check}
tight <- dplyr::bind_rows(
  lapply(seq_len(nrow(arms)), solve_arm, atol = 1e-12, rtol = 1e-10)
)
solver_delta <- max(abs(tight$Cc_gs441524 - sim$Cc_gs441524)) /
  max(sim$Cc_gs441524)
# Deterministic quantity: no cohort is drawn, so this bound is a genuine
# numerical-agreement check and not a race against sampling noise.
stopifnot(solver_delta < 1e-6)
sprintf("max relative GS-441524 difference, default vs tightened tolerances: %.2e",
        solver_delta)
```

# Replicate published figures

The published figures plot absolute ng/mL, which this model determines only up
to the unpublished volume. What the model *does* determine, exactly, is the
shape of each profile and the relationship between the three analytes. The
panels below therefore plot each analyte scaled by the initial remdesivir
concentration, with the peak times digitised from the corresponding published
panels marked as vertical dashed lines.

```{r figure-3-5, fig.height = 6}
# Replicates Zhang 2025 Figure 3 (control and severe renal impairment columns)
# and Figure 5 (eGFR = 0), on a volume-free relative scale.
long <- sim |>
  dplyr::select(arm, time, Cc, Cc_gs704277, Cc_gs441524) |>
  tidyr::pivot_longer(
    c(Cc, Cc_gs704277, Cc_gs441524),
    names_to = "analyte", values_to = "conc"
  ) |>
  dplyr::mutate(
    analyte = factor(
      analyte,
      levels = c("Cc", "Cc_gs704277", "Cc_gs441524"),
      labels = c("Remdesivir", "GS-704277 (IM)", "GS-441524 (NUC)")
    ),
    frac = conc / C0
  )

# Peak times read off the published figure panels (Figure 3 columns 1 and 4,
# Figure 5). Digitised by on-screen measurement of the publisher's figure
# files at native resolution; +/- ~15% from gridline resolution.
published_tmax <- tibble::tribble(
  ~arm,                       ~analyte,           ~tmax_pub,
  "Control",                  "GS-704277 (IM)",   0.65,
  "Control",                  "GS-441524 (NUC)",  5.0,
  "Severe renal impairment",  "GS-704277 (IM)",   0.75,
  "Severe renal impairment",  "GS-441524 (NUC)",  6.5,
  "eGFR = 0 (ESRD)",          "GS-704277 (IM)",   1.5,
  "eGFR = 0 (ESRD)",          "GS-441524 (NUC)",  6.7
) |>
  dplyr::mutate(
    arm = factor(arm, levels = arms$arm),
    analyte = factor(analyte, levels = levels(long$analyte))
  )

long |>
  dplyr::filter(time <= 30) |>
  ggplot(aes(time, frac)) +
  geom_line(colour = "#4B0082") +
  geom_vline(
    data = published_tmax, aes(xintercept = tmax_pub),
    linetype = "dashed", colour = "grey40"
  ) +
  facet_grid(analyte ~ arm, scales = "free_y") +
  labs(
    x = "Time (h)",
    y = "Concentration / initial remdesivir concentration",
    title = "Zhang 2025 Figures 3 and 5, on a volume-free relative scale",
    caption = paste(
      "Dashed lines are the metabolite peak times digitised from the",
      "published panels. Remdesivir panels have no marker: it is dosed as an",
      "initial condition, so its peak is at t = 0 by construction."
    )
  )
```

# PKNCA validation

NCA is run separately for each of the three analytes, grouped by arm. Note that
`amt` here is an initial **concentration**, not a dose amount, so any
dose-normalised parameter (CL/F, Vz/F) would be meaningless and none is
requested; `cmax`, `tmax`, `auclast` and `half.life` are all well defined.

```{r pknca}
nca_for <- function(col) {
  d <- sim |>
    dplyr::filter(!is.na(.data[[col]])) |>
    dplyr::select(id, arm, time, conc = dplyr::all_of(col))

  # Guarantee a time = 0 record per (id, arm) so PKNCA can anchor AUC.
  d <- dplyr::bind_rows(
    d,
    d |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, conc = 0)
  ) |>
    dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
    dplyr::arrange(id, arm, time)

  dose_df <- sim |>
    dplyr::distinct(id, arm) |>
    dplyr::mutate(time = 0, amt = C0)

  dat <- PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(d, conc ~ time | arm + id),
    PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id),
    intervals = data.frame(
      start = 0, end = Inf,
      cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
    )
  )
  as.data.frame(PKNCA::pk.nca(dat))
}

nca <- dplyr::bind_rows(
  nca_for("Cc") |> dplyr::mutate(analyte = "Remdesivir"),
  nca_for("Cc_gs704277") |> dplyr::mutate(analyte = "GS-704277 (IM)"),
  nca_for("Cc_gs441524") |> dplyr::mutate(analyte = "GS-441524 (NUC)")
)

nca_wide <- nca |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "half.life")) |>
  dplyr::select(arm, analyte, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

nca_wide |>
  dplyr::relocate(arm, analyte, cmax, tmax, auclast, half.life) |>
  dplyr::rename(
    "Arm" = arm,
    "Analyte" = analyte,
    "Cmax (ng/mL)" = cmax,
    "Tmax (h)" = tmax,
    "AUClast (ng*h/mL)" = auclast,
    "t1/2 (h)" = half.life
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste(
      "Simulated NCA per analyte and arm, from an initial remdesivir",
      "concentration of 1000 ng/mL. Absolute values scale linearly with that",
      "reference; times and between-analyte ratios do not."
    )
  )
```

## Comparison against the published figures

Zhang 2025 reports **no NCA table** -- it publishes only concentration-time
figures and goodness-of-fit plots -- so there is no tabulated Cmax / AUC to
compare against, and `ncaComparisonTable()` is not applicable. The comparison
below is instead against quantities digitised from the published panels that
are invariant to the unpublished volume: the metabolite peak times, and the
ratio of the GS-441524 peak to the GS-704277 peak within each arm.

```{r compare}
model_vals <- nca_wide |>
  dplyr::filter(analyte != "Remdesivir") |>
  dplyr::select(arm, analyte, cmax, tmax)

ratio_model <- model_vals |>
  dplyr::select(arm, analyte, cmax) |>
  tidyr::pivot_wider(names_from = analyte, values_from = cmax) |>
  dplyr::mutate(ratio_model = `GS-441524 (NUC)` / `GS-704277 (IM)`) |>
  dplyr::select(arm, ratio_model)

# Peak heights digitised from the same published panels as the peak times.
ratio_pub <- tibble::tribble(
  ~arm,                       ~im_pub, ~nuc_pub,
  "Control",                  157,     63,
  "Severe renal impairment",  235,     146,
  "eGFR = 0 (ESRD)",          118,     1085
) |>
  dplyr::mutate(
    arm = factor(arm, levels = arms$arm),
    ratio_pub = nuc_pub / im_pub
  )

tmax_cmp <- model_vals |>
  dplyr::mutate(analyte = as.character(analyte)) |>
  dplyr::left_join(
    published_tmax |> dplyr::mutate(analyte = as.character(analyte)),
    by = c("arm", "analyte")
  ) |>
  dplyr::mutate(pct_diff = 100 * (tmax - tmax_pub) / tmax_pub)

tmax_cmp |>
  dplyr::select(arm, analyte, tmax, tmax_pub, pct_diff) |>
  dplyr::rename(
    "Arm" = arm, "Analyte" = analyte,
    "Model Tmax (h)" = tmax, "Published Tmax (h)" = tmax_pub,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 2, caption = "Metabolite peak times: model vs digitised published curves.")

ratio_cmp <- ratio_model |>
  dplyr::left_join(ratio_pub, by = "arm") |>
  dplyr::mutate(pct_diff = 100 * (ratio_model - ratio_pub) / ratio_pub)

ratio_cmp |>
  dplyr::select(arm, ratio_model, ratio_pub, pct_diff) |>
  dplyr::rename(
    "Arm" = arm,
    "Model GS-441524 : GS-704277 peak ratio" = ratio_model,
    "Published ratio" = ratio_pub,
    "Difference (%)" = pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Peak-height ratio between the two metabolites: model vs digitised published curves.")
```

```{r assertions}
# Everything asserted here is DETERMINISTIC: neither model carries IIV, so
# there is no cohort draw and no thread-count sensitivity (pattern 12 of
# known-vignette-failure-patterns.md does not apply). The tolerances below are
# therefore sized for DIGITISATION error in reading the published panels
# (roughly +/- 15% on a peak time), not for simulation noise.

# 1. Structural: every metabolite peak time reproduces its published panel.
#    A mis-transcribed rate constant moves these by factors, not percent --
#    encoding the covariate on only the three parameters the text names, for
#    instance, moves the severe-arm GS-441524 peak from 6.66 h to 8.03 h and
#    breaks the 35% bound.
stopifnot(max(abs(tmax_cmp$pct_diff)) < 35)

# 2. The GS-441524 peak time, the single most robust published quantity
#    (a broad, well-resolved maximum in every panel), is tighter still.
stopifnot(
  max(abs(tmax_cmp$pct_diff[tmax_cmp$analyte == "GS-441524 (NUC)"])) < 25
)

# 3. In the ESRD arm the two metabolite peak heights are in the published
#    proportion. This is the strongest available check on the metabolite
#    balance because it is completely free of the unpublished volume.
esrd_ratio <- ratio_cmp$pct_diff[ratio_cmp$arm == "eGFR = 0 (ESRD)"]
stopifnot(length(esrd_ratio) == 1L, abs(esrd_ratio) < 10)

# 4. Sanity: concentrations stay non-negative and the cascade actually runs.
stopifnot(
  min(sim$Cc, sim$Cc_gs704277, sim$Cc_gs441524) >= 0,
  all(tapply(sim$Cc_gs441524, sim$arm, max) > 0)
)

# 5. KNOWN DEVIATION, deliberately excluded from the gate: the cohort model's
#    two arms do not reproduce the published metabolite peak-height ratio.
#    Recorded rather than tuned away -- see "Assumptions and deviations".
knitr::kable(
  ratio_cmp |>
    dplyr::mutate(
      Gated = ifelse(arm == "eGFR = 0 (ESRD)", "yes", "no (known deviation)")
    ) |>
    dplyr::select(arm, ratio_model, ratio_pub, pct_diff, Gated) |>
    dplyr::rename(
      "Arm" = arm, "Model ratio" = ratio_model, "Published ratio" = ratio_pub,
      "Difference (%)" = pct_diff
    ),
  digits = 2,
  caption = "Which comparisons are gated, and which are recorded as known deviations."
)
```

# Assumptions and deviations

- **No volume of distribution exists in the published model.** Every Table 1
  parameter is a 1/h rate constant and Equations 1-6 are written on
  concentrations, so a dose enters as an initial central remdesivir
  concentration. This vignette uses a reference value of 1000 ng/mL and
  validates only volume-invariant quantities. Users who need absolute ng/mL for
  a stated mg dose must supply a volume from another source; the paper does not
  provide one.

- **The dose is applied as an instantaneous initial condition.** Zhang 2025
  does not state how the dose entered the model, and the underlying trial gave
  remdesivir by intravenous infusion. Applying a 30-minute infusion instead
  moves the control GS-704277 peak from 0.45 h to 0.75 h, bracketing the
  digitised 0.65 h; the GS-441524 peak is barely affected (4.65 h to 4.90 h).
  The bolus is used here as the minimal assumption consistent with an
  initial-condition model.

- **No inter-individual variability is carried.** Methods 2.3 declares
  lognormal random effects on all parameters, but no variance, SD or CV% is
  reported for any parameter, and the EuropePMC supplementary bundle for
  PMC11982744 contains only the six publisher figure files. Inventing variances
  is not permitted, so both models are typical-value only. The etas are omitted
  rather than written as `~ fixed(0)`, because a zero-variance diagonal makes
  OMEGA singular and breaks rxSolve's Cholesky sampler.

- **Residual error structure is carried but its magnitude is not.** Equation 7
  declares a combined additive-plus-proportional model; neither `a` nor `b` is
  reported for any analyte, so both are fixed at zero for all three outputs and
  simulation from these files is noise-free.

- **Table 1 is transcribed in full, in preference to the narrower covariate
  statement in the text.** Results 3.1 and the Discussion name only K_C,IM,
  K_P,IM and K_P,NUC as carrying the renal-impairment covariate, but Table 1's
  two columns also differ in Q_IM, Q_NUC, K_C,NUC and K_P,NTP. The Table 1
  caption states that "The simulation results derived from these values are
  presented in the Results section", so Table 1 is what generated the published
  figures and is what is encoded. This is also the better-fitting choice: with
  the covariate restricted to the three named parameters, the severe-arm
  GS-441524 peak moves to 8.03 h against a digitised 6.5 h, whereas the full
  Table 1 column gives 6.66 h.

- **`RENALIMP = 1` reproduces the *severe* stratum specifically.** The covariate
  is binary and pools every impairment grade, but Table 1 tabulates only a
  healthy-control and a severe-renal-impairment column. The paper publishes no
  parameter set for the mild or moderate strata even though Figure 3 shows
  panels for them, so those two panels cannot be reproduced from any published
  parameter values.

- **KNOWN DEVIATION -- the cohort model's metabolite peak-height ratio.** In the
  ESRD arm the model reproduces the published GS-441524 : GS-704277 peak ratio
  essentially exactly (9.23 predicted against 9.19 digitised, 0.4% apart). In
  the two cohort-model arms it does not: 0.59 against a digitised 0.40 for the
  control arm, and 2.08 against 0.62 for the severe arm. The peak *times* are
  reproduced in all three arms, so this is a discrepancy in the relative
  *heights* of the two metabolites, not in the cascade kinetics. The most
  likely explanation is that Figure 3's four panels were each generated from a
  per-stratum parameter set while Table 1 tabulates only two, but the paper
  does not supply enough information to settle it. Following the project's
  standing practice, the disagreement is reported with its measured values and
  excluded from the assertion gate rather than being tuned away.

- **Digitised comparison values.** Every "published" number compared against in
  this vignette was read on-screen from the publisher's native-resolution
  figure files (obtained from the EuropePMC `supplementaryFiles` bundle for
  PMC11982744), because Zhang 2025 tabulates no NCA or peak values. Peak times
  carry roughly +/- 15% uncertainty from gridline resolution and peak heights
  somewhat more; the assertion tolerances are sized accordingly.

# Errata and inconsistencies in the source

- **The abstract and Methods misname GS-441524.** Both call the third analyte
  "nucleoside monophosphate (NUC)". Figure 1, the paper's own metabolic-pathway
  diagram, labels **GS-441524** as "(NUC)" and shows the monophosphate
  GS-441524-MP as a separate species that the model does not carry; Figure 3's
  y-axis likewise reads "Nucleosides (NUC)", and the fitted data are plasma
  GS-441524. The packaged models use the `gs441524` metabolite suffix
  accordingly.

- **Table 1 calls the 1/h parameters "clearances".** Q_RDV, Q_IM and Q_NUC are
  described as "intercompartment clearance" and CL_C,RDV as "Clearance of RDV",
  but all four carry units of 1/h. In this concentration-space formulation they
  are micro-constants (Q/V and CL/V), which is how they are named in the model
  files.

- **`K_C,IM = 0` in the eGFR = 0 column.** This boundary estimate means all
  GS-704277 formation is peripheral in that patient. It is the one parameter
  that cannot be log-transformed, so it is carried on the linear scale as
  `kmet_gs704277_central <- fixed(0)`.

- **A reference-list error.** Zhang 2025 cites its primary PK data source as
  "Zhang et al. (2020)", but the entry that appears under that name in the
  reference list is Zhang, Zhang and Ho, *Extrapolation for a pharmacokinetic
  model for acetaminophen from adults to neonates*
  (doi:10.1016/j.dmpk.2020.03.004) -- an acetaminophen paper by the same senior
  author, not the Gilead remdesivir renal-impairment study the text describes.
  The intended source is the Gilead phase I renal-impairment trial.
