---
title: "Ribociclib intracranial CDK4/6 occupancy QSP (Zhang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ribociclib intracranial CDK4/6 occupancy QSP (Zhang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 7,
  fig.height = 4.5
)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Zhang_2026_ribociclib_qsp"))$reference`
* Article (open access, CC BY-NC-ND 4.0): <https://doi.org/10.1186/s12885-026-15561-x>
* PMC record: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC12888661/>

```{r load-model}
mod <- readModelDb("Zhang_2026_ribociclib_qsp")
ui  <- rxode2::rxode2(mod)
mod
```

Ribociclib (RIB) is a CDK4/6 inhibitor approved for hormone-receptor-positive,
HER2-negative advanced breast cancer at 600 mg once daily (OD). Zhang 2026 asks
whether that regimen delivers enough *intracranial* target engagement to treat
brain metastases while staying below a neutropenia-driven safety ceiling, and
answers it by chaining two layers: a whole-body PBPK model that predicts
ribociclib concentrations in plasma and cerebrospinal fluid (CSF), and a
target-engagement layer that converts the free CSF concentration into CDK4 and
CDK6 occupancy.

Ribociclib is roughly four-fold less potent against CDK6 than against CDK4
(Ki 39 nM versus 10 nM). That single fact drives the paper's entire conclusion:
at every regimen the authors simulate, CDK4 occupancy clears the 90% efficacy
bar well before CDK6 does, so the dose that satisfies *both* isoforms is the
binding constraint.

# What is extracted, and what is not

Zhang 2026 contains two modelling layers. The first is **not** extracted.

| Layer | Content | Extracted? | Why |
|---|---|---|---|
| 1 | Whole-body PBPK for ribociclib in plasma, brain and CSF, built in PK-Sim 9.1 | **No** | A platform port. The Methods state that "system-specific parameters were primarily derived from the built-in PK-Sim database" and that distribution used PK-Sim's built-in Rodgers and Rowland method. Table 1 publishes drug-specific inputs only. The paper reports **no volume term of any kind** -- no Vd, no Vss, no partition-coefficient table (only a scalar `Kp scale` of 3.0 multiplying PK-Sim's computed per-organ values), no organ volumes, no blood flows, no half-life, no terminal parameter. The only clearance printed is a renal `CL_R` of 1.22 L/h; everything else is a per-enzyme Vmax/Km that needs unpublished organ-level enzyme abundances to become a systemic clearance. No `.pksim5` project was deposited. The layer is not reproducible from the published sources, and filling the gaps from platform defaults would be fabrication. |
| 2 | CDK4/6 target engagement in CSF (Eq 1) | **Yes** | Eq 1 is printed in full and every constant it needs (`Ki` for both isoforms, `koff`, `f_up`, MW, `CDK4/6_0`, `Kp,uu`) is in main-text Table 1 or the Methods prose. |

Because layer 1 is not extracted, the ribociclib exposure is supplied as the
time-varying covariate `CP_RIB_NGML` (total plasma concentration, ng/mL). This
follows the `Zhang_2025_abemaciclib_qsp` and `Liang_2024_osimertinib_qsp`
precedents. `Richardson_2025_ribociclib` is an in-library plasma popPK model
that can serve as the driver, though it originates from a different
publication; everything below is driven instead from Zhang 2026's *own*
published predicted concentrations, so that no number in this vignette comes
from outside the paper.

Note that Zhang 2026 includes **no tumor compartment** and computes occupancy in
**CSF only** -- the Methods say "we used the free RIB concentration in CSF to
calculate CDK4/6 occupancy". This is a structural difference from the same
team's abemaciclib model, which carries both a plasma and a CSF occupancy limb.

# Population

```{r population}
pop <- ui$population
tibble::tibble(Field = names(pop), Value = vapply(pop, function(x) paste(as.character(x), collapse = "; "), character(1))) |>
  knitr::kable()
```

# Source trace

Every value in `ini()`, and the one equation in `model()`, traced to its
location in Zhang 2026.

```{r source-trace}
tibble::tribble(
  ~Quantity,                       ~`Model name`,  ~Value,      ~`Source location`,
  "Molecular weight",              "mw_rib",       "434.6 g/mol", "Table 1, physicochemical block, MW row (ref 18)",
  "Fraction unbound in plasma",    "fup",          "0.12",        "Table 1, physicochemical block, f_up row (ref 19)",
  "CSF-to-plasma unbound ratio",   "kpuu",         "1.03",        "Results, 'PBPK model validation for plasma PK': predicted Kp,uu 0.99 / 1.03 / 1.10 at 400 / 600 / 900 mg",
  "Ki (equilibrium Kd) for CDK4",  "kd4",          "10 nmol/L",   "Table 1, CDK4/6 occupancy block, 'K i (nM) 10/39', first of pair (ref 24)",
  "Ki (equilibrium Kd) for CDK6",  "kd6",          "39 nmol/L",   "Table 1, CDK4/6 occupancy block, 'K i (nM) 10/39', second of pair (ref 24)",
  "Dissociation rate constant",    "koff",         "3.78 /h",     "Table 1, koff row: 0.063 /min, x 60 min/h (ref 25)",
  "Starting CDK4/6 expression",    "cdk0",         "100 nmol/L",  "Methods prose beneath Eq 1: 'set to a value of 0.1 uM'",
  "Target-engagement ODE",         "d/dt(complex)","Eq 1",        "Methods, Eq 1 (recovered from the PDF; the trimmed markdown renders it as 'formula-not-decoded')"
) |>
  knitr::kable()
```

## Units table (dimensional analysis)

```{r units-table}
tibble::tribble(
  ~Step, ~Expression, ~Units,
  "Covariate in",           "CP_RIB_NGML",                    "ng/mL (total plasma)",
  "Free plasma",            "CP_RIB_NGML * fup",              "ng/mL",
  "Free CSF",               "CP_RIB_NGML * fup * kpuu",       "ng/mL",
  "Molar conversion",       "ccsf_ngml / mw_rib * 1000",      "(ng/mL) / (g/mol) * 1000 = nmol/L",
  "Association rate",       "koff / kd4",                     "(1/h) / (nmol/L) = 1/(nmol/L * h)",
  "Association flux",       "koff/kd4 * free_cdk4 * ccsf",    "1/(nmol/L * h) * nmol/L * nmol/L = nmol/L/h",
  "Dissociation flux",      "koff * complex",                 "1/h * nmol/L = nmol/L/h",
  "Occupancy",              "100 * complex / cdk0",           "percent"
) |>
  knitr::kable()
```

Both limbs of Eq 1 carry nmol/L/h, so the ODE is dimensionally consistent.

## Parameter table: paper vs. model file

```{r param-table}
iniDf <- ui$iniDf
iniDf |>
  dplyr::select(name, est, fix, label) |>
  knitr::kable()

# Every parameter must be fixed: this is a deterministic mechanism model.
stopifnot(all(iniDf$fix[!is.na(iniDf$ntheta)]))
# And there must be no random effects at all.
stopifnot(all(is.na(iniDf$neta1)))
```

# Building the exposure driver

Zhang 2026 Table 2 tabulates PBPK-**predicted** ribociclib exposures. Two sets
are used below, both from the predicted column so that the comparison is against
the paper's own model rather than against raw clinical observations.

```{r published-anchors}
# Predicted steady-state (Day 18) PLASMA exposures, Infante cohort.
plasma_anchors <- tibble::tribble(
  ~dose_mg, ~kpuu, ~cmax_ngml, ~cmin_ngml,
  400,      0.99,  1291,       524,
  600,      1.03,  2237,       797,
  900,      1.10,  3720,       1456
)

# Predicted free-CSF concentrations (Day 5), Tien and Johnson cohorts.
# Table 2 footnote: "Ct total free exposure in CSF".
csf_anchors <- tibble::tribble(
  ~label,               ~dose_mg, ~kpuu, ~csf_ngml,
  "Johnson, 400 mg OD, Ct",   400, 0.99,  48.1,
  "Johnson, 600 mg OD, Ct",   600, 1.03,  82.1,
  "Tien, 900 mg OD, C98h",    900, 1.10,  220.0,
  "Tien, 900 mg OD, C104h",   900, 1.10,  186.2,
  "Tien, 900 mg OD, C120h",   900, 1.10,  76.8
)

knitr::kable(plasma_anchors)
knitr::kable(csf_anchors)
```

The model consumes a *total plasma* concentration, so a published free-CSF
concentration is converted back with the paper's own relation
`C_CSF,free = Kp,uu * f_up * C_plasma,total`.

```{r driver-helpers}
FUP <- 0.12
MW  <- 434.6

plasma_from_csf <- function(csf_ngml, kpuu) csf_ngml / (FUP * kpuu)
csf_from_plasma <- function(cp_ngml, kpuu)  cp_ngml * FUP * kpuu
ngml_to_nm      <- function(x) x / MW * 1000

# Analytic (Langmuir) occupancy at equilibrium, for a free CSF concentration
# expressed in nmol/L. Derived below from Eq 1 by setting dN/dt = 0.
langmuir <- function(c_nm, kd) 100 * c_nm / (c_nm + kd)
```

An internal consistency check of that relation against the paper's two
independent prediction sets: the predicted steady-state plasma trough at 600 mg
OD, pushed through `Kp,uu * f_up`, should land near the separately predicted CSF
trough at the same dose.

```{r consistency-check}
derived_600 <- csf_from_plasma(797, 1.03)
c(derived_from_plasma = derived_600, published_csf = 82.1,
  pct_diff = 100 * (derived_600 - 82.1) / 82.1)
```

The two agree to about 20%, across different virtual populations (Infante's
breast-cancer / liposarcoma / colon-cancer cohort versus Johnson's glioblastoma
cohort), which supports using either as a driver.

```{r event-table}
# Observation-only event table: there are no dosing events in the extracted
# layer -- the drug enters entirely through the covariate. `cmt` points at a
# real ODE state, never at an algebraic observable.
obs_grid <- function(times) {
  rxode2::et(times) |>
    as.data.frame() |>
    dplyr::mutate(cmt = "complex_cdk4_csf")
}

# Deterministic model with no random effects, so one "subject" per arm suffices.
solve_const <- function(cp_ngml, kpuu, times = seq(0, 48, by = 0.25)) {
  ev <- obs_grid(times)
  ev$CP_RIB_NGML <- cp_ngml
  out <- rxode2::rxSolve(ui, ev, params = c(kpuu = kpuu), returnType = "data.frame")
  if (is.null(out$id)) out$id <- 1L
  out
}
```

# Validation

## 1. Baseline hold (zero exposure)

With no drug present, both complexes must stay at zero and both occupancies at
zero for all time.

```{r baseline-hold}
base <- solve_const(cp_ngml = 0, kpuu = 1.03)

stopifnot(
  max(abs(base$complex_cdk4_csf)) < 1e-10,
  max(abs(base$complex_cdk6_csf)) < 1e-10,
  max(abs(base$occCdk4Csf))       < 1e-8,
  max(abs(base$occCdk6Csf))       < 1e-8
)
c(max_complex_cdk4 = max(abs(base$complex_cdk4_csf)),
  max_complex_cdk6 = max(abs(base$complex_cdk6_csf)))
```

## 2. Mass balance: each CDK pool is conserved

`free_cdk4 + complex_cdk4_csf` must equal `cdk0` = 100 nmol/L at every time
point, and likewise for CDK6. This is exact by construction, so the tolerance is
solver noise rather than a modelling allowance.

```{r mass-balance}
mb <- solve_const(cp_ngml = plasma_from_csf(220.0, 1.10), kpuu = 1.10)

err4 <- max(abs(mb$totalCdk4Csf - 100))
err6 <- max(abs(mb$totalCdk6Csf - 100))
stopifnot(err4 < 1e-8, err6 < 1e-8)
c(max_massbalance_error_cdk4 = err4, max_massbalance_error_cdk6 = err6)
```

## 3. Analytic steady state of the binding layer

Setting `dN/dt = 0` in Eq 1 with a constant free concentration `C` gives

```
(koff/Kd) * (CDK0 - N) * C = koff * N
  =>  N / CDK0 = C / (C + Kd)
```

the Langmuir isotherm. `koff` cancels entirely, so the equilibrium occupancy
depends only on `C` and `Kd` -- which is why the `koff` units conflict noted in
the model file (Table 1 says min<sup>-1</sup>, the Eq 1 gloss says
h<sup>-1</sup>) cannot affect any steady-state result in this vignette.

```{r analytic-steady-state}
iso_check <- lapply(seq_len(nrow(csf_anchors)), function(i) {
  a  <- csf_anchors[i, ]
  s  <- solve_const(plasma_from_csf(a$csf_ngml, a$kpuu), a$kpuu)
  fin <- s[nrow(s), ]
  cnm <- ngml_to_nm(a$csf_ngml)
  tibble::tibble(
    label       = a$label,
    csf_ngml    = a$csf_ngml,
    csf_nM      = cnm,
    ode_cdk4    = fin$occCdk4Csf,
    exact_cdk4  = langmuir(cnm, 10),
    ode_cdk6    = fin$occCdk6Csf,
    exact_cdk6  = langmuir(cnm, 39)
  )
}) |> dplyr::bind_rows()

max_iso_err <- max(abs(c(iso_check$ode_cdk4 - iso_check$exact_cdk4,
                         iso_check$ode_cdk6 - iso_check$exact_cdk6)))
stopifnot(max_iso_err < 1e-6)

iso_check |>
  dplyr::mutate(
    csf_nM     = round(csf_nM, 2),
    ode_cdk4   = round(ode_cdk4, 4),
    exact_cdk4 = round(exact_cdk4, 4),
    ode_cdk6   = round(ode_cdk6, 4),
    exact_cdk6 = round(exact_cdk6, 4)
  ) |>
  knitr::kable()
c(max_abs_deviation_from_langmuir_pct = max_iso_err)
```

The ODE reproduces the closed form to better than 1e-6 percentage points. Both
sides use the same parameters, so this is pure numerical error and the tight
bound is the correct assertion.

## 4. Equilibration is fast relative to the dosing interval

The paper compares occupancy at peak and at trough. That reading is only valid
if the complex tracks the concentration essentially instantaneously. Starting
from zero complex at a 600 mg trough concentration, measure the time to reach
99% of the equilibrium value.

```{r equilibration-speed}
fast <- solve_const(797, 1.03, times = seq(0, 6, by = 0.01))
eq4  <- langmuir(ngml_to_nm(csf_from_plasma(797, 1.03)), 10)
eq6  <- langmuir(ngml_to_nm(csf_from_plasma(797, 1.03)), 39)

t99_4 <- min(fast$time[fast$occCdk4Csf >= 0.99 * eq4])
t99_6 <- min(fast$time[fast$occCdk6Csf >= 0.99 * eq6])

stopifnot(t99_4 < 1, t99_6 < 1)
c(t99_cdk4_h = t99_4, t99_cdk6_h = t99_6)
```

Both isoforms equilibrate in well under an hour against a 24 h (OD) or 12 h
(BID) dosing interval, so peak and trough occupancy are the Langmuir values at
peak and trough concentration. Every occupancy figure below rests on that.

## 5. Perturbation recovery (washout)

Drug removal must return both pools to zero occupancy.

```{r washout}
wash_ev <- obs_grid(seq(0, 24, by = 0.1))
wash_ev$CP_RIB_NGML <- ifelse(wash_ev$time <= 6, 797, 0)
wash <- rxode2::rxSolve(mod, wash_ev, params = c(kpuu = 1.03), returnType = "data.frame")

stopifnot(
  wash$occCdk4Csf[wash$time == 6] > 90,
  dplyr::last(wash$occCdk4Csf) < 1e-4,
  dplyr::last(wash$occCdk6Csf) < 1e-4
)

ggplot(
  wash |> tidyr::pivot_longer(c(occCdk4Csf, occCdk6Csf), names_to = "isoform", values_to = "occupancy"),
  aes(time, occupancy, colour = isoform)
) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = 6, linetype = "dashed") +
  labs(x = "Time (h)", y = "CSF CDK occupancy (%)",
       title = "Perturbation recovery: exposure withdrawn at 6 h") +
  theme_bw()
```

# Replicating the paper's occupancy results

## Occupancy at the paper's own predicted free-CSF concentrations

Each row takes a predicted CSF concentration straight from Table 2 and reports
the occupancy the extracted model returns.

```{r occupancy-at-anchors}
occ_anchor <- iso_check |>
  dplyr::transmute(
    `CSF anchor (Table 2, predicted)` = label,
    `C (ng/mL)`    = csf_ngml,
    `C (nmol/L)`   = round(csf_nM, 1),
    `CDK4 occ (%)` = round(ode_cdk4, 1),
    `CDK6 occ (%)` = round(ode_cdk6, 1)
  )
knitr::kable(occ_anchor)
```

## Steady-state peak and trough occupancy by regimen (Figure 3)

Driving instead from the predicted steady-state plasma Cmax and Cmin, which is
what Figure 3's regimen comparison rests on.

```{r regimen-table}
regimen <- plasma_anchors |>
  dplyr::mutate(
    csf_peak_ngml   = csf_from_plasma(cmax_ngml, kpuu),
    csf_trough_ngml = csf_from_plasma(cmin_ngml, kpuu),
    peak_cdk4  = langmuir(ngml_to_nm(csf_peak_ngml),   10),
    peak_cdk6  = langmuir(ngml_to_nm(csf_peak_ngml),   39),
    trough_cdk4 = langmuir(ngml_to_nm(csf_trough_ngml), 10),
    trough_cdk6 = langmuir(ngml_to_nm(csf_trough_ngml), 39)
  )

# Confirm the closed form against the ODE for every regimen, at both extremes.
ode_chk <- lapply(seq_len(nrow(regimen)), function(i) {
  r  <- regimen[i, ]
  pk <- solve_const(r$cmax_ngml, r$kpuu); tr <- solve_const(r$cmin_ngml, r$kpuu)
  c(dplyr::last(pk$occCdk4Csf) - r$peak_cdk4,  dplyr::last(pk$occCdk6Csf) - r$peak_cdk6,
    dplyr::last(tr$occCdk4Csf) - r$trough_cdk4, dplyr::last(tr$occCdk6Csf) - r$trough_cdk6)
})
stopifnot(max(abs(unlist(ode_chk))) < 1e-6)

regimen |>
  dplyr::transmute(
    `Regimen`          = paste0(dose_mg, " mg OD"),
    `Peak CDK4 (%)`    = round(peak_cdk4, 1),
    `Peak CDK6 (%)`    = round(peak_cdk6, 1),
    `Trough CDK4 (%)`  = round(trough_cdk4, 1),
    `Trough CDK6 (%)`  = round(trough_cdk6, 1)
  ) |>
  knitr::kable()
```

### Gate against the paper's stated claims

Zhang 2026 states two regimen results unambiguously in the Results section
"Simulations in BCBM for different dosing regimen":

* 600 mg OD -- "CDK4 occupancy exceeded 90% ... though CDK6 occupancy stayed
  below 90% but above 70%".
* 900 mg OD -- "Both CDK4 and CDK6 occupancies surpassed 90%".

Both are trough-referenced statements (Table 3 reports "Tough [trough] CDK4/6
occupancy in CSF"), so they are gated on the trough column.

```{r gate-claims}
g600 <- regimen[regimen$dose_mg == 600, ]
g900 <- regimen[regimen$dose_mg == 900, ]

# Guard that the lookups actually matched rows (a zero-row filter would make
# every all() below vacuously TRUE).
stopifnot(nrow(g600) == 1L, nrow(g900) == 1L)

stopifnot(
  # 600 mg OD: CDK4 clears 90%, CDK6 sits in the 70-90% band.
  g600$trough_cdk4 > 90,
  g600$trough_cdk6 > 70, g600$trough_cdk6 < 90,
  # 900 mg OD: both isoforms clear 90%.
  g900$trough_cdk4 > 90,
  g900$trough_cdk6 > 90,
  # The paper's mechanism: CDK4 engagement leads CDK6 at every dose.
  all(regimen$trough_cdk4 > regimen$trough_cdk6),
  all(regimen$peak_cdk4   > regimen$peak_cdk6)
)

tibble::tribble(
  ~Claim,                                        ~`Paper`,                 ~`Model`,
  "600 mg OD, trough CDK4",                      "> 90%",                  sprintf("%.1f%%", g600$trough_cdk4),
  "600 mg OD, trough CDK6",                      "70-90%",                 sprintf("%.1f%%", g600$trough_cdk6),
  "900 mg OD, trough CDK4",                      "> 90%",                  sprintf("%.1f%%", g900$trough_cdk4),
  "900 mg OD, trough CDK6",                      "> 90%",                  sprintf("%.1f%%", g900$trough_cdk6)
) |>
  knitr::kable()
```

Both published regimen claims are reproduced.

## Occupancy time course over a dosing interval

An illustrative 24 h steady-state profile at each dose, interpolated between the
paper's own published Cmax and Cmin anchors (log-linear decay from Cmax at an
assumed Tmax of 2 h down to Cmin at 24 h, and a log-linear rise before Tmax).
Tmax is the one quantity here that Zhang 2026 does not report -- see
"Assumptions and deviations".

```{r time-course}
TMAX <- 2

profile_for <- function(dose_mg) {
  r <- regimen[regimen$dose_mg == dose_mg, ]
  stopifnot(nrow(r) == 1L)
  tt <- seq(0, 24, by = 0.25)
  cp <- ifelse(
    tt <= TMAX,
    r$cmin_ngml * (r$cmax_ngml / r$cmin_ngml)^(tt / TMAX),
    r$cmax_ngml * (r$cmin_ngml / r$cmax_ngml)^((tt - TMAX) / (24 - TMAX))
  )
  ev <- obs_grid(tt)
  ev$CP_RIB_NGML <- cp
  out <- rxode2::rxSolve(mod, ev, params = c(kpuu = r$kpuu), returnType = "data.frame")
  out$dose <- paste0(dose_mg, " mg OD")
  out
}

tc <- dplyr::bind_rows(lapply(plasma_anchors$dose_mg, profile_for))

ggplot(
  tc |> tidyr::pivot_longer(c(occCdk4Csf, occCdk6Csf), names_to = "isoform", values_to = "occupancy"),
  aes(time, occupancy, colour = dose, linetype = isoform)
) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 90, linetype = "dotted") +
  coord_cartesian(ylim = c(50, 100)) +
  labs(x = "Time within the dosing interval (h)", y = "CSF CDK occupancy (%)",
       title = "Intracranial CDK4/6 occupancy over 24 h (replicates Figure 3 in form)",
       subtitle = "Dotted line: the paper's 90% efficacy threshold") +
  theme_bw()
```

The CDK4 traces sit above the 90% line across the whole interval at every dose,
while the CDK6 traces cross it only at 900 mg -- the paper's central finding.

# Assumptions and deviations

## The PBPK layer is not extracted

The whole-body PK-Sim 9.1 model that produces the ribociclib concentrations is
not reproducible from the published sources (see "What is extracted, and what is
not"). Every occupancy in this vignette is therefore **conditional on the
exposures the paper itself predicted**, which are reproduced verbatim from
Table 2 rather than re-simulated. A downstream user must supply their own
ribociclib exposure trajectory through `CP_RIB_NGML`.

## Deviation: CDK4 occupancy at 400 mg OD

Zhang 2026 states that "CDK4/6 occupancy failed to reach 90% at 400 mg OD
(Fig. 3A)". The extracted model agrees for CDK6 but not for CDK4.

```{r deviation-400}
g400 <- regimen[regimen$dose_mg == 400, ]
stopifnot(nrow(g400) == 1L)

johnson400 <- iso_check[iso_check$label == "Johnson, 400 mg OD, Ct", ]
stopifnot(nrow(johnson400) == 1L)

tibble::tribble(
  ~`Driver for 400 mg OD`,                          ~`CDK4 occ (%)`,               ~`CDK6 occ (%)`,
  "Predicted plasma Cmin (Infante), via Kp,uu*fup", round(g400$trough_cdk4, 1),    round(g400$trough_cdk6, 1),
  "Predicted CSF Ct (Johnson), used directly",      round(johnson400$ode_cdk4, 1), round(johnson400$ode_cdk6, 1)
) |>
  knitr::kable()

# CDK6 reproduces the paper's claim on both drivers; CDK4 exceeds 90% on both.
stopifnot(
  g400$trough_cdk6 < 90, johnson400$ode_cdk6 < 90,
  g400$trough_cdk4 > 90, johnson400$ode_cdk4 > 90
)
```

Both available drivers put CDK4 occupancy at 400 mg OD marginally above 90%
(92-94%), where the paper reports a failure. Three candidate causes, none of
which is resolvable from the on-disk sources:

1. The Figure 3 simulations use a BCBM virtual population aligned with Johnson's
   cohort, whereas the plasma anchors used here come from Infante's cohort and
   the CSF anchors are Day 5 rather than the Day 14 of Figure 3.
2. The paper's sentence is a compressed statement about the *pair* of isoforms
   ("CDK4/6 occupancy"), and Figure 3A may well show CDK4 near but not at 90%
   with CDK6 clearly below.
3. Figure 3 reports population means with 10th-90th percentiles (Supplementary
   Figure S1), whereas the calculation here is a typical-value calculation on a
   mean concentration; occupancy is concave in concentration, so the mean of the
   occupancy is below the occupancy of the mean.

The discrepancy is confined to one isoform at the lowest simulated dose and does
not affect either of the two regimen claims the paper states unambiguously, both
of which are gated above.

## Peak versus trough is ambiguous in the source

The paper defines its efficacy criterion twice, inconsistently. The Methods say
the target is "maximal target occupancy (i.e., peak occupancy levels) at or
above 90%"; Table 3 reports "Tough [trough] CDK4/6 occupancy in CSF". The
regimen table above reports both columns so a reader can apply either reading.
The gates use the trough column, which is the stricter of the two and the one
that matches Table 3.

```{r peak-vs-trough}
regimen |>
  dplyr::transmute(
    Regimen = paste0(dose_mg, " mg OD"),
    `CDK6 clears 90% on PEAK reading`   = peak_cdk6   > 90,
    `CDK6 clears 90% on TROUGH reading` = trough_cdk6 > 90
  ) |>
  knitr::kable()
```

## Twice-daily regimens are not reproducible from published numbers

Zhang 2026's headline recommendation is 300 mg BID, and Figure 3 also simulates
200 mg and 400 mg BID. **No plasma or CSF concentration is published for any
twice-daily regimen** -- Table 2 tabulates once-daily arms only. Reproducing the
BID arms would require running the un-extracted PBPK layer, so they are
deliberately omitted rather than reconstructed by extrapolating the once-daily
exposures (ribociclib exposure is visibly non-linear in these data: predicted
AUC rises 2.6-fold between 600 and 900 mg, driven by the CYP3A4 auto-inhibition
in the PBPK layer).

## Other assumptions

* **`Kp,uu` is a model output, not an input.** It is the one quantity in
  `ini()` that the un-extracted PBPK layer produced. The paper reports 0.99,
  1.03 and 1.10 at 400, 600 and 900 mg respectively (against observed 1.22, 1.29
  and 1.63); the model file defaults to the 600 mg value and each simulation
  above overrides it to the dose-matched value.
* **`koff` units conflict.** Table 1 gives 0.063 min<sup>-1</sup>; the prose
  gloss beneath Eq 1 calls `koff` an h<sup>-1</sup> quantity. The table is taken
  as authoritative (the same team's abemaciclib paper likewise tabulates `koff`
  in min<sup>-1</sup>). Section 4 above shows `koff` cancels out of every
  steady-state result, so the choice affects only the equilibration transient.
* **`Ki` is treated as an equilibrium dissociation constant** and stored under
  the canonical `kd` name, because the paper's own gloss states that
  "`koff`/`Ki` is equivalent to the association rate constant (`kon`)".
* **Tmax = 2 h** is assumed for the illustrative time-course figure only. It is
  not reported in Zhang 2026 and enters no gate.
* **No IIV and no residual error** are encoded. Zhang 2026 propagated
  variability through PK-Sim virtual-population physiology, which is part of the
  un-extracted layer.
* **The starting CDK4/6 expression of 0.1 uM is a default, not a measurement.**
  The paper states there are no reported human data, and lists the absence of
  experimentally determined CDK4/6 occupancy as a study limitation. Occupancy is
  a ratio to that pool, so it is insensitive to the choice except through
  target-mediated depletion, which is negligible here (free ribociclib in CSF is
  1-5x `cdk0` at the simulated doses).

## Errata and supplement

No erratum or corrigendum exists for this article as of the extraction date. The
single supplementary file (MOESM1) was retrieved and contains one figure
(S1, simulated intracranial CDK4/6 occupancy by regimen, presented as means with
10th-90th percentiles); it contains no parameter table and no equations. Eq 1
was recovered from the PDF because the trimmed markdown companion renders it as
a `formula-not-decoded` marker.
