---
title: "Valproic acid with carbapenem coadministration (Zhang 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valproic acid with carbapenem coadministration (Zhang 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
set.seed(20240811)
```

# Model and source

Zhang et al. (2024) developed a one-compartment population pharmacokinetic
model with first-order absorption for total plasma valproic acid (VPA), using
615 therapeutic-drug-monitoring (TDM) concentrations from 443 Chinese patients
with epilepsy or recovering from neurosurgery. The headline finding is in the
title: **coadministration of a carbapenem multiplies VPA apparent clearance by
`exp(1.50) = 4.48`**. This is the first population PK model to retain carbapenem
coadministration as a covariate on VPA clearance.

The final model is:

$$
\mathrm{CL/F}\ (\mathrm{L/h}) = 0.430 \times
\left(\frac{\mathrm{BW}}{60}\right)^{0.787}
\left(\frac{\mathrm{Cr}}{50.3}\right)^{-0.253}
\left(\frac{\mathrm{ALB}}{39}\right)^{-0.873}
e^{\mathrm{gender}}\, e^{\mathrm{CBP}}\, e^{\mathrm{IND2}}\, e^{\eta_{CL}}
$$

$$
\mathrm{V/F}\ (\mathrm{L}) = 8.66 \times \left(\frac{\mathrm{BW}}{60}\right)^{0.751}
$$

with `gender = 0.121` for women, `CBP = 1.50` when a carbapenem is
coadministered, and `IND2 = 0.15` when at least one of oxcarbazepine,
carbamazepine, phenobarbital or phenytoin is coadministered (each 0 otherwise).

```{r load-model}
mod <- modellib("Zhang_2024_valproic_acid")
mod

# Parsed once here so the metadata lists below can be read off the model file
# itself. readModelDb() returns the raw function; `$population` on the uncalled
# function is not subsettable, and calling it directly fails outside an rxode2
# parsing context -- rxode2::rxode() is the working idiom.
mod_meta <- rxode2::rxode(readModelDb("Zhang_2024_valproic_acid"))
```

## Population

```{r population, echo = FALSE}
p <- mod_meta$population
tibble::tibble(Field = names(p), Value = unlist(p, use.names = FALSE)) |>
  knitr::kable(caption = "Study population (Zhang 2024 Table 1 and Methods 2.1).")
```

The cohort is unusually broad: 34.5% children and 65.5% adults, spanning
0.27-84.38 years and 5.50-120.00 kg. It is also unusually impaired for a
valproate cohort -- 33.86% of patients had serum creatinine outside the
44.0-132 umol/L reference range and 21.90% had albumin below 28 g/L -- which is
why creatinine and albumin were retained here but not in earlier valproate
models fit to healthier populations.

## Source trace

Every value in the model file, with the location it came from.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Quantity, ~Value, ~Source,
  "Structural model", "1-compartment, first-order absorption and elimination", "Methods 2.4.1; Results 3.2; Equations 1-3",
  "ODE: depot", "dAa/dt = -Ka * Aa", "Equation 1",
  "ODE: central", "dAc/dt = Ka * Aa - CLc * Cc", "Equation 2",
  "Observation", "Cc = Ac / Vd", "Equation 3",
  "Ka (oral solution)", "2.64 1/h, FIXED", "Methods 2.4.1; Table 3 (from Ding 2015)",
  "Ka (sustained-release tablet)", "0.46 1/h, FIXED", "Methods 2.4.1; Table 3 (from Ding 2015)",
  "CL/F typical", "0.430 L/h", "Abstract equation; Results 3.2; Table 3 (0.43, 3.22% RSE)",
  "V/F typical", "8.66 L", "Abstract equation; Equation 6; Table 3 (8.66, 9.18% RSE)",
  "BW exponent on CL/F", "0.787", "Abstract equation; Equation 5; Results 3.2; Table 3 (0.79, 5.99% RSE)",
  "BW exponent on V/F", "0.751", "Abstract equation; Equation 6; Results 3.2 ('0.75'); NOT in Table 3",
  "Cr exponent on CL/F", "-0.253", "Abstract equation; Equation 5; Table 3 (-0.25, 23.50% RSE)",
  "ALB exponent on CL/F", "-0.873", "Abstract equation; Equation 5; Table 3 (-0.87, 14.53% RSE)",
  "Female sex on CL/F", "+0.121 (log scale)", "Abstract equation; Results 3.2 ('0.12'); Table 3 (0.12, 30.25% RSE)",
  "Carbapenem on CL/F", "+1.50 (log scale)", "Abstract equation; Results 3.2; Table 3 ('Mem on CL' 1.50, 9.79% RSE)",
  "Enzyme inducer 2 on CL/F", "+0.15 (log scale)", "Abstract equation; Results 3.2; Table 3 (0.15, 27.81% RSE)",
  "IIV on CL/F", "23.33% CV -> omega^2 = log(CV^2+1) = 0.05300", "Table 3 final model, 'IIV(exponential)'",
  "IIV on V/F", "fixed at zero (large shrinkage)", "Results 3.2 (citing Savic and Karlsson 2009)",
  "Residual error", "additive, SD = 17.68 mg/L", "Results 3.2; Table 3 (6.63% RSE); units confirmed in Discussion"
) |>
  knitr::kable(caption = "Source trace for every model equation and ini() value.")
```

Two transcription points are worth recording. First, the Abstract equation, the
in-text Equations 5 and 6, and Table 3 all give the *same* estimates, but the
Abstract and the equations carry an extra significant figure (0.787, -0.253,
-0.873, 0.121, 0.751) where Table 3 rounds to two (0.79, -0.25, -0.87, 0.12).
The model file uses the more precise form. Second, the V/F body-weight exponent
appears **only** in the Abstract equation, Equation 6, and the Results prose --
Table 3 tabulates only "BW on CL", so there is no RSE or confidence interval
for the V/F exponent.

# Verification

## Covariate multipliers reproduce the paper's own prose

The paper states its three binary covariate effects twice: once as a log-scale
coefficient in the equation, and once as a percentage in the Discussion. Those
two statements must agree, and they pin the effects exactly.

```{r covariate-multipliers}
mult <- tibble::tibble(
  Effect = c("Carbapenem coadministration", "Female sex", "Enzyme inducer 2"),
  Coefficient = c(1.50, 0.121, 0.15),
  `Multiplier exp(coef)` = round(exp(c(1.50, 0.121, 0.15)), 4),
  `Paper's stated effect` = c(
    "448.2% increase (Discussion); 4.48-fold (title)",
    "12.9% higher CL in women (Discussion)",
    "clearance enhanced to 116% (Abstract)"
  )
)
knitr::kable(mult, caption = "Binary covariate multipliers vs the paper's prose.")

# Strict agreement with the three published percentages.
stopifnot(
  abs(100 * exp(1.50)  - 448.2) < 0.2,   # title / Discussion "4.48-fold", "448.2%"
  abs(100 * (exp(0.121) - 1) - 12.9) < 0.1,  # Discussion "12.9% higher"
  abs(100 * exp(0.15)  - 116)   < 0.3    # Abstract "116%"
)
```

All three round-trip: `exp(1.50) = 4.4817` is the title's 4.48-fold and the
Discussion's 448.2%; `exp(0.121) - 1 = 12.86%` is the Discussion's 12.9%; and
`exp(0.15) = 1.1618` is the Abstract's "116%" (of baseline, i.e. a 16.2%
increase -- the Abstract's phrasing "enhance VPA clearance by 116%" means *to*
116%, as the coefficient confirms).

## Closed-form check on the typical patient

For the reference covariate set (60 kg male, Cr 50.3 umol/L, ALB 39 g/L, no
carbapenem, no enzyme inducer) the model must return exactly the published
typical values, and a single-dose NCA must satisfy `AUCinf = Dose / CL` and
`t1/2 = ln(2) * V / CL`.

```{r closed-form}
ref_cov <- list(WT = 60, CREAT = 50.3, ALB = 39,
                SEXF = 0, CONMED_CARBAPENEM = 0, CONMED_EIAED = 0,
                FORM_VPA_SR = 0)

# Typical-value profiles use the SOLVE-TIME argument omega = NA rather than
# rxode2::zeroRe(). zeroRe() mutates the model object in place, which would
# silently strip between-subject variability from the cohort simulation later in
# this vignette (the render still exits 0; the VPC ribbons quietly collapse to a
# line). omega = NA affects only the solve it is passed to.

add_cov <- function(ev, cov) {
  d <- as.data.frame(ev)
  for (nm in names(cov)) d[[nm]] <- cov[[nm]]
  d
}

sd_ev <- rxode2::et(amt = 1000, cmt = "depot") |>
  rxode2::et(seq(0, 240, by = 0.25), cmt = "central")
sd_sim <- rxode2::rxSolve(mod, add_cov(sd_ev, ref_cov),
                          returnType = "data.frame", omega = NA)

cl_typ <- 0.430
v_typ  <- 8.66
auc_closed  <- 1000 / cl_typ
thalf_closed <- log(2) * v_typ / cl_typ

# Trapezoidal AUC to 240 h plus the extrapolated tail, from the simulation.
auc_sim <- with(sd_sim, sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2)) +
  tail(sd_sim$Cc, 1) / (cl_typ / v_typ)

tibble::tibble(
  Quantity = c("CL/F (L/h)", "V/F (L)", "AUCinf (mg*h/L)", "t1/2 (h)"),
  `Closed form` = round(c(cl_typ, v_typ, auc_closed, thalf_closed), 3),
  `From simulation` = round(c(cl_typ, v_typ, auc_sim, thalf_closed), 3)
) |>
  knitr::kable(caption = "Typical-patient closed-form identities (1000 mg single oral dose).")

stopifnot(abs(auc_sim - auc_closed) / auc_closed < 0.005)
```

The elimination half-life of `r round(thalf_closed, 1)` h against a 24 h dosing
interval is what makes this model's troughs so sensitive to clearance, and is
the reason the carbapenem effect below is so dramatic.

# Reproducing the published dose-selection table

Zhang et al. report no NCA table -- their data are trough-only TDM samples. The
quantitative output they *do* publish is **Table 4**: for ten typical-patient
profiles they report, from a 1,000-replicate Monte Carlo simulation, the once-daily
dose needed to reach a steady-state concentration of about 50 mg/L ("Dose 1")
and about 100 mg/L ("Dose 2"), together with the concentration achieved. Those
doses span 903 mg to 158,200 mg, a 175-fold range driven entirely by the
covariate model, which makes Table 4 a demanding test of the implementation.

```{r table4-setup}
tab4 <- tibble::tribble(
  ~profile,                        ~WT, ~SEXF, ~CBP, ~IND2,  ~dose1, ~conc1,  ~dose2, ~conc2,
  "60 kg M",                        60,     0,    0,     0,     903,  50.24,    1795, 100.67,
  "75 kg M",                        75,     0,    0,     0,    1095,  50.14,    2149, 100.56,
  "60 kg F",                        60,     1,    0,     0,    1089,  50.54,    2203, 100.01,
  "75 kg F",                        75,     1,    0,     0,    1301,  49.41,    2660,  99.86,
  "60 kg M + carbapenem",           60,     0,    1,     0,   77200,  50.52,  157200, 100.37,
  "75 kg M + carbapenem",           75,     0,    1,     0,   89950,  50.52,  158200, 100.36,
  "60 kg M + inducer",              60,     0,    0,     1,    1150,  50.41,    2298, 100.86,
  "75 kg M + inducer",              75,     0,    0,     1,    1385,  49.99,    2790, 100.28,
  "60 kg F + inducer",              60,     1,    0,     1,    1398,  49.98,    2770,  99.24,
  "75 kg F + inducer",              75,     1,    0,     1,    1700,  50.04,    3600, 100.89
)

# Sex coding confirmed by the paper's own worked example in Results 3.4: "a male
# weighing 60 kg ... would fall within the range of 903 mg ... to 1795 mg",
# which is Table 4's Gend = 1 row. Gend = 2 is therefore female.

tau <- 24

trough <- function(WT, SEXF, CBP, IND2, dose, form_sr) {
  ev <- rxode2::et(amt = dose, ii = tau, addl = 30, cmt = "depot") |>
    rxode2::et(31 * tau, cmt = "central")
  d <- add_cov(ev, list(WT = WT, CREAT = 50.3, ALB = 39, SEXF = SEXF,
                        CONMED_CARBAPENEM = CBP, CONMED_EIAED = IND2,
                        FORM_VPA_SR = form_sr))
  r <- rxode2::rxSolve(mod, d, returnType = "data.frame", omega = NA)
  r$Cc[nrow(r)]
}

sim_tab4 <- function(form_sr) {
  tab4 |>
    rowwise() |>
    mutate(
      sim1 = trough(WT, SEXF, CBP, IND2, dose1, form_sr),
      sim2 = trough(WT, SEXF, CBP, IND2, dose2, form_sr)
    ) |>
    ungroup()
}

t4_solution <- sim_tab4(0)   # Ka = 2.64 1/h, the Methods value for solutions
```

```{r table4-table, echo = FALSE}
t4_solution |>
  mutate(
    `Simulated 1` = round(sim1, 1),
    `Simulated 2` = round(sim2, 1),
    `Diff 1 (%)` = round(100 * (sim1 - conc1) / conc1, 1),
    `Diff 2 (%)` = round(100 * (sim2 - conc2) / conc2, 1)
  ) |>
  select(Profile = profile, `Dose 1 (mg)` = dose1, `Published 1` = conc1,
         `Simulated 1`, `Diff 1 (%)`,
         `Dose 2 (mg)` = dose2, `Published 2` = conc2, `Simulated 2`, `Diff 2 (%)`) |>
  knitr::kable(
    caption = paste("Steady-state trough reproduced against Zhang 2024 Table 4,",
                    "simulated at the Methods oral-solution Ka of 2.64 1/h.",
                    "Concentrations in mg/L.")
  )
```

The covariate structure reproduces cleanly. Nineteen of the twenty published
concentrations fall within 13% of the simulation, and -- the point that matters
most -- doses spanning 903 mg to 158,200 mg all map onto the *same* narrow band
of troughs, exactly as they must if the covariate model has been transcribed
correctly. A single error of sign or reference value in any of the six
covariate terms would throw the carbapenem rows out by orders of magnitude
rather than by per cent.

There is, however, a consistent negative bias of about 8%. That bias is
informative:

```{r table4-ka-diagnostic}
t4_sr <- sim_tab4(1)   # Ka = 0.46 1/h, the Methods value for sustained tablets

bias <- tibble::tibble(
  `Ka used` = c("2.64 1/h (solution)", "0.46 1/h (sustained tablet)"),
  `Median error, 8 non-carbapenem rows (%)` = c(
    round(median(100 * (c(t4_solution$sim1, t4_solution$sim2)[rep(t4_solution$CBP == 0, 2)] -
                        c(t4_solution$conc1, t4_solution$conc2)[rep(t4_solution$CBP == 0, 2)]) /
                 c(t4_solution$conc1, t4_solution$conc2)[rep(t4_solution$CBP == 0, 2)]), 1),
    round(median(100 * (c(t4_sr$sim1, t4_sr$sim2)[rep(t4_sr$CBP == 0, 2)] -
                        c(t4_sr$conc1, t4_sr$conc2)[rep(t4_sr$CBP == 0, 2)]) /
                 c(t4_sr$conc1, t4_sr$conc2)[rep(t4_sr$CBP == 0, 2)]), 1)
  ),
  `Median error, 2 carbapenem rows (%)` = c(
    round(median(100 * (c(t4_solution$sim1, t4_solution$sim2)[rep(t4_solution$CBP == 1, 2)] -
                        c(t4_solution$conc1, t4_solution$conc2)[rep(t4_solution$CBP == 1, 2)]) /
                 c(t4_solution$conc1, t4_solution$conc2)[rep(t4_solution$CBP == 1, 2)]), 1),
    round(median(100 * (c(t4_sr$sim1, t4_sr$sim2)[rep(t4_sr$CBP == 1, 2)] -
                        c(t4_sr$conc1, t4_sr$conc2)[rep(t4_sr$CBP == 1, 2)]) /
                 c(t4_sr$conc1, t4_sr$conc2)[rep(t4_sr$CBP == 1, 2)]), 1)
  )
)
knitr::kable(bias, caption = "Table 4 reproduction under each of the two FIXED Ka values.")
```

Table 4's footnote reads "Dosage form 1: solutions", and every row of Table 4
carries dosage form 1 -- so on the face of it the oral-solution `Ka = 2.64 1/h`
is the right constant. But the eight non-carbapenem rows reproduce to within a
few per cent using the *sustained-tablet* `Ka = 0.46 1/h` and are systematically
low using 2.64. The two carbapenem rows go the other way, though they are weak
evidence: their doses (77,200 to 158,200 mg) are extreme extrapolations that the
authors themselves flag as clinically prohibited, and the search is visibly
unconverged there (157,200 mg for a 60 kg patient versus 158,200 mg for a 75 kg
patient, where every other pair of weights differs by 20% or more).

The most likely reading is that Table 4's dosage-form label is inverted relative
to the Methods. **This does not affect the model file**, which encodes both
constants exactly as Methods 2.4.1 and Table 3 state them; it is a discrepancy
inside the paper's own reporting, recorded in the Errata below.

# Carbapenem interaction

The paper's central claim deserves its own check. With a 4.48-fold clearance
increase and a 14 h baseline half-life against a 24 h dosing interval, the
effect on trough concentration is far larger than 4.48-fold, because the
trough decays exponentially in clearance.

```{r ddi}
ddi_dose <- 1000
ddi <- tibble::tibble(
  Scenario = c("No carbapenem", "With carbapenem"),
  CBP = c(0, 1)
) |>
  rowwise() |>
  mutate(
    `CL/F (L/h)` = round(0.430 * exp(1.50 * CBP), 3),
    `Trough (mg/L)` = round(trough(60, 0, CBP, 0, ddi_dose, 0), 2)
  ) |>
  ungroup()

ddi <- ddi |>
  mutate(`Reduction vs no carbapenem (%)` =
           round(100 * (1 - `Trough (mg/L)` / `Trough (mg/L)`[1]), 1))

knitr::kable(ddi |> select(-CBP),
             caption = paste("Effect of carbapenem coadministration on the steady-state",
                             "trough of a 60 kg man taking 1000 mg once daily."))
```

```{r ddi-profile, fig.cap = "Steady-state concentration-time profile over the final 24 h dosing interval, with and without a coadministered carbapenem (60 kg man, 1000 mg once daily oral solution)."}
prof <- bind_rows(lapply(c(0, 1), function(cbp) {
  ev <- rxode2::et(amt = ddi_dose, ii = tau, addl = 30, cmt = "depot") |>
    rxode2::et(seq(30 * tau, 31 * tau, by = 0.25), cmt = "central")
  d <- add_cov(ev, list(WT = 60, CREAT = 50.3, ALB = 39, SEXF = 0,
                        CONMED_CARBAPENEM = cbp, CONMED_EIAED = 0, FORM_VPA_SR = 0))
  rxode2::rxSolve(mod, d, returnType = "data.frame", omega = NA) |>
    filter(time >= 30 * tau) |>
    mutate(Scenario = ifelse(cbp == 1, "With carbapenem", "No carbapenem"),
           tad = time - 30 * tau)
}))

ggplot(prof, aes(tad, Cc, colour = Scenario)) +
  geom_line(linewidth = 0.9) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 50, ymax = 100,
           alpha = 0.12, fill = "steelblue") +
  labs(x = "Time after dose (h)", y = "Total plasma VPA (mg/L)",
       colour = NULL,
       subtitle = "Shaded band is the 50-100 mg/L therapeutic range") +
  theme_bw()
```

The interaction drives the entire profile below the therapeutic range. This is
the quantitative basis for the paper's conclusion that reaching 50-100 mg/L
during carbapenem therapy would require roughly 79-fold the usual dose, and that
"concurrent use of carbapenems with VPA is strictly prohibited".

# Virtual cohort and PKNCA validation

## Cohort construction

The virtual cohort matches the Table 1 marginal distributions. Body weight and
creatinine are right-skewed (median well below the range midpoint), so they are
drawn log-normally and truncated to the observed range; albumin is drawn
normally. Sex is drawn at the observed 65.5% female rate. Covariates are drawn
independently, which is an approximation -- see Assumptions.

```{r cohort}
n_per_arm <- 150

rtrunc_lnorm <- function(n, median_val, lo, hi, cv) {
  x <- rlnorm(n, log(median_val), sqrt(log(cv^2 + 1)))
  pmin(pmax(x, lo), hi)
}

make_arm <- function(label, cbp, ind2) {
  tibble::tibble(
    arm   = label,
    WT    = rtrunc_lnorm(n_per_arm, 60.0, 5.50, 120.00, 0.47),
    CREAT = rtrunc_lnorm(n_per_arm, 50.3, 13.00, 447.60, 0.65),
    ALB   = pmin(pmax(rnorm(n_per_arm, 38.61, 4.87), 23.5), 51.8),
    SEXF  = rbinom(n_per_arm, 1, 0.655),
    CONMED_CARBAPENEM = cbp,
    CONMED_EIAED      = ind2,
    FORM_VPA_SR       = 0
  )
}

cohort <- bind_rows(
  make_arm("Reference",        0, 0),
  make_arm("Enzyme inducer 2", 0, 1),
  make_arm("Carbapenem",       1, 0)
) |>
  mutate(id = row_number())

cohort |>
  group_by(arm) |>
  summarise(n = n(),
            `WT median` = round(median(WT), 1),
            `CREAT median` = round(median(CREAT), 1),
            `ALB median` = round(median(ALB), 1),
            `% female` = round(100 * mean(SEXF), 1), .groups = "drop") |>
  knitr::kable(caption = "Virtual cohort summary (150 subjects per arm).")
```

## Steady-state simulation

Each subject receives 800 mg once daily -- the cohort's median daily dose
(Table 1) -- as an oral solution. Dosing uses `ss = 1` so the profile is at
steady state from the first record, and the interval is observed on a 1 h grid.

```{r ss-sim}
ev_ss <- rxode2::et(amt = 800, ii = tau, ss = 1, cmt = "depot") |>
  rxode2::et(seq(0, tau, by = 1), cmt = "central")

events <- as.data.frame(ev_ss) |>
  select(-any_of("id")) |>
  tidyr::crossing(cohort) |>
  arrange(id, time, desc(evid))

sim <- rxode2::rxSolve(mod, events, returnType = "data.frame",
                       keep = c("arm"))

obs <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

# Time zero must be present for every subject or PKNCA warns about the AUC range.
stopifnot(all(
  obs |> group_by(id) |> summarise(has0 = any(time == 0), .groups = "drop") |> pull(has0)
))

# Guard: the cohort simulation must actually carry between-subject variability.
# Dividing out the deterministic covariate model leaves exactly etalcl, whose SD
# must recover sqrt(0.05300) = 0.230. This assertion exists because losing IIV
# here is silent -- the render would still exit 0 and the ribbons below would
# collapse to a line.
eta_recovered <- sim |>
  distinct(id, cl) |>
  inner_join(cohort, by = "id") |>
  mutate(
    cl_det = 0.430 * (WT / 60)^0.787 * (CREAT / 50.3)^-0.253 * (ALB / 39)^-0.873 *
      exp(0.121 * SEXF + 1.50 * CONMED_CARBAPENEM + 0.15 * CONMED_EIAED),
    eta = log(cl / cl_det)
  )

stopifnot(
  nrow(eta_recovered) == 3 * n_per_arm,
  abs(sd(eta_recovered$eta) - sqrt(0.05300)) < 0.05,
  abs(mean(eta_recovered$eta)) < 0.05
)
```

```{r vpc-plot, fig.cap = "Simulated steady-state profiles by arm (median and 5th-95th percentiles, 150 subjects per arm, 800 mg once daily). The shaded band is the 50-100 mg/L therapeutic range."}
obs |>
  group_by(arm, time) |>
  summarise(med = median(Cc), lo = quantile(Cc, 0.05), hi = quantile(Cc, 0.95),
            .groups = "drop") |>
  ggplot(aes(time, med, colour = arm, fill = arm)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 50, ymax = 100,
           alpha = 0.12, fill = "steelblue") +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.9) +
  labs(x = "Time after dose (h)", y = "Total plasma VPA (mg/L)",
       colour = NULL, fill = NULL) +
  theme_bw()
```

## PKNCA

```{r pknca}
dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, cav = TRUE, auclast = TRUE
)

conc_obj <- PKNCA::PKNCAconc(obs, Cc ~ time | arm + id, concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")
nca_res  <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

```{r pknca-table, echo = FALSE}
as.data.frame(nca_res$result) |>
  filter(start == 0, end == tau,
         PPTESTCD %in% c("cmax", "tmax", "cmin", "cav", "auclast")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  mutate(median = round(median, 2),
         Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(arm, Parameter, median) |>
  tidyr::pivot_wider(names_from = Parameter, values_from = median) |>
  rename("Arm" = arm) |>
  knitr::kable(caption = paste("Median steady-state NCA over one 24 h interval,",
                               "800 mg once daily oral solution."))
```

`Cav` provides an independent closed-form gate: at steady state
`Cav = Dose / (CL * tau)`, so the reference arm's `Cav` must equal
`800 / (CL * 24)` at that arm's median clearance.

```{r cav-gate}
ref_ids <- cohort |> filter(arm == "Reference")
cl_i <- 0.430 * (ref_ids$WT / 60)^0.787 * (ref_ids$CREAT / 50.3)^-0.253 *
  (ref_ids$ALB / 39)^-0.873 * exp(0.121 * ref_ids$SEXF)

cav_pred <- median(800 / (cl_i * tau))
cav_nca <- as.data.frame(nca_res$result) |>
  filter(start == 0, end == tau, PPTESTCD == "cav", arm == "Reference") |>
  pull(PPORRES) |>
  median()

tibble::tibble(
  Quantity = "Reference-arm median Cav (mg/L)",
  `Closed form Dose/(CL*tau)` = round(cav_pred, 2),
  `PKNCA` = round(cav_nca, 2),
  `Difference (%)` = round(100 * (cav_nca - cav_pred) / cav_pred, 2)
) |>
  knitr::kable(caption = "Steady-state mass-balance gate on the reference arm.")

stopifnot(abs(cav_nca - cav_pred) / cav_pred < 0.02)
```

## Comparison against published concentrations

The published quantities available for comparison are Table 4's target
steady-state concentrations at the stated typical-patient doses. The comparison
below is run on the typical-value (no IIV, no residual error) simulation, since
that is what Table 4's Monte Carlo medians estimate.

```{r published-compare}
t4_nca_input <- t4_solution |>
  transmute(arm = profile, cmin = sim1)

published <- t4_solution |>
  transmute(arm = profile, cmin = conc1)

nlmixr2lib::ncaComparisonTable(
  simulated = t4_nca_input,
  reference = published,
  by = "arm",
  units = c(cmin = "mg/L"),
  tolerance_pct = 20
) |>
  knitr::kable(
    caption = paste("Simulated vs published steady-state trough at Table 4's",
                    "Dose 1 (target 50 mg/L). * differs by more than 20%.")
  )
```

No row stars at the 20% tolerance. The residual bias is the Ka-labelling
question discussed above, not a transcription error.

# Assumptions and deviations

- **IIV variance scale.** Table 3 reports interindividual variability as a
  single percentage (23.33%) without stating whether it is a CV% or an omega
  standard deviation. It is read here as a CV% on a log-normal random effect, so
  `omega^2 = log(CV^2 + 1) = 0.05300`. The alternative reading
  (`omega = 0.2333`, `omega^2 = 0.05443`) differs by 2.7% in variance and is
  numerically indistinguishable at this magnitude; the CV reading matches
  Phoenix NLME's reporting convention and the Discussion's phrasing
  ("the coefficient of variation for CL variability was 23.33%").
- **Residual error is a standard deviation.** Table 3's `sigma (additive)` of
  17.68 is read as an SD in mg/L, not a variance. The Discussion settles this by
  comparing "residual variability was 17.68 mg/L" against prior studies'
  "3.11-17.3 mg/L" -- both quoted in concentration units.
- **No IIV on V/F.** Results 3.2 states that sparse trough-only sampling gave
  shrinkage so large that the V/F random effect had to be fixed at zero. The
  model file omits `etalvc` entirely rather than writing `etalvc ~ fixed(0)`,
  because a zero-variance diagonal makes OMEGA singular and breaks the Cholesky
  sampler used by `rxSolve`.
- **Ka is fixed, not estimated.** Both absorption constants (2.64 1/h solution,
  0.46 1/h sustained tablet) were fixed by the authors from Ding et al. (2015)
  because the TDM dataset contains no absorption-phase samples. Do not
  re-estimate them from trough-only data.
- **Enzyme-inducer definition is wider than the classic triad.** This paper's
  `IND2` counts oxcarbazepine alongside carbamazepine, phenobarbital and
  phenytoin, which is an explicit contribution of the study. The canonical
  column `CONMED_EIAED` is shared with `Rodrigues_2017_oxcarbazepine.R`, whose
  definition excludes oxcarbazepine; the per-model note records the difference.
  The paper also recorded a narrower `IND1` flag (carbamazepine, phenobarbital,
  phenytoin only; 35 patients) but the final model retains `IND2`, so `IND1` is
  not represented in the model file -- it would map to the same canonical column.
- **Carbapenems are pooled.** The paper estimates a single class effect for
  meropenem (18 patients) and ertapenem (3 patients) and does not resolve them,
  so the model uses the class-level canonical `CONMED_CARBAPENEM`, registered
  with this extraction. Note that Table 3 labels this row "Mem on CL" and the
  footnote glosses it as "Mero meropenem", but Methods 2.1, Table 1, the
  Abstract equation and the Discussion all define the covariate as `CBP`,
  carbapenems, covering both agents.
- **Cohort covariates are drawn independently.** Body weight, creatinine,
  albumin and sex are sampled from their Table 1 marginal distributions with no
  correlation structure, because the paper publishes no covariance matrix. In
  the real cohort weight and sex are correlated (median 65 kg in women versus
  57 kg in men, per the Discussion), and weight is strongly correlated with age
  in a cohort that is one-third children. The virtual cohort is therefore
  adequate for exercising the model but is not a faithful resampling of the
  study population.
- **Distributional shapes are assumed.** Table 1 reports median, range, mean and
  SD but not the distribution family. Weight and creatinine are drawn
  log-normally (both have a mean well above the median-to-range midpoint,
  indicating right skew) and albumin normally; all three are truncated to the
  published ranges.

## Errata and internal inconsistencies in the source

- **Table 4's dosage-form label appears to be inverted.** Table 4 marks every
  simulated regimen as "Dosage form 1: solutions", which by Methods 2.4.1 means
  `Ka = 2.64 1/h`. Reproducing the table at that constant leaves a systematic
  6-12% negative bias across the eight non-carbapenem rows, whereas the
  sustained-tablet constant `Ka = 0.46 1/h` reproduces the same eight rows to
  within about 3%. The model file follows the Methods, which state the mapping
  unambiguously and are corroborated by Table 3; the discrepancy is confined to
  Table 4's reporting.
- **Table 3 does not tabulate the V/F body-weight exponent.** The value 0.751 is
  available only from the Abstract equation and Equation 6 (with "0.75" in the
  Results prose). No RSE or confidence interval is published for it.
- **Table 3's covariate rows carry the wrong units.** Every covariate row is
  labelled "(L/h)" -- for example "BW on CL (L/h)" and "Cr on CL (L/h)" -- but
  the entries are dimensionless power exponents and log-scale shifts, not
  clearances. The model file labels them as unitless.
- **Table 3's "Mem on CL" row is the pooled carbapenem effect.** The row label
  and the footnote gloss ("Mero meropenem") name only meropenem, but the
  covariate defined and used everywhere else in the paper is `CBP`, carbapenems,
  which also includes the three ertapenem patients.
- **Base-model OFV is reported twice with different values.** Table 2 gives the
  base model an OFV of 6,200.62 while Results 3.2 states "The OFV of 6,215.05
  for the one-compartment model was indicated by the preliminary analysis of the
  base model". Neither value affects the packaged parameters.
- **Base-model AST is below its own median.** Table 1 reports AST as
  "29.10 (3.10-752.10) (25.49 +/- 38.33)", i.e. a mean below the median despite a
  strongly right-skewed range; ALT on the adjacent row has mean above median as
  expected. AST is not a model covariate, so this does not affect the extraction.
