---
title: "Nedosiran (Zhang 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nedosiran (Zhang 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhang_2024_nedosiran"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhang_2024_nedosiran"))$description`
* Article: <https://doi.org/10.1111/bcp.16194>

Nedosiran (RIVFLOZA) is a GalNAc-conjugated small interfering RNA that silences
hepatic *LDHA*, the gene encoding the lactate dehydrogenase isoform responsible
for the terminal glyoxylate-to-oxalate step in the liver. Zhang 2024 pooled five
clinical studies (PHYOX1, PHYOX2, PHYOX3, PHYOX5, PHYOX6) into a single
population PK/PD analysis and used it to confirm the approved weight-banded
dosing regimen.

The model is one jointly developed PK/PD analysis and is therefore packaged as a
single model file.

## Structure

The structure below is taken from Zhang 2024 Figure 3 (page 3183). **The number
of transit compartments in each absorption chain appears nowhere in the text or
in either parameter table** -- it is only readable from the Figure 3 schematic:

* **Slow pathway** (fraction `FR1` of the dose): input compartment `depot1`
  -- *k*<sub>a1</sub> --> `transit1` -- *k*<sub>a1</sub> --> `central`
  (two states in series).
* **Fast pathway** (fraction `1 - FR1`): input compartment `depot2`
  -- *k*<sub>a2</sub> --> `transit2` -- *k*<sub>a2</sub> --> `transit3`
  -- *k*<sub>a2</sub> --> `transit4` -- *k*<sub>a2</sub> --> `central`
  (four states in series).

So the paper's "dual *n*-transit absorption" is *n* = 2 for the slow pathway and
*n* = 4 for the fast pathway. Disposition is two-compartment with **parallel**
linear (`CL/F`) and Michaelis-Menten (`Vmax * C / (KM + C)`) elimination from
`central`; the paper attributes the linear arm to renal elimination and the
saturable arm to ASGPR-mediated hepatic uptake. A Sheiner effect compartment
with `ke0 = ln(2) / lambda` drives a sigmoid-Imax indirect-response model that
inhibits production of 24-h urinary oxalate.

# Population

The PK dataset comprised 1978 plasma nedosiran concentrations from 143
participants: 85 healthy adult volunteers without a primary hyperoxaluria (PH)
diagnosis, 46 patients with PH type 1 (PH1) and 12 with PH2. The PD dataset
comprised 588 24-h urinary oxalate observations from the 46 PH1 patients in
PHYOX1, PHYOX2 and PHYOX3, adjusted per 1.73 m^2 body surface area in
participants aged under 18 years. Placebo-arm PD data (n = 11) were excluded
after showing no apparent change over 28 weeks, and 13.1% of PK observations
fell below the 1.0 ng/mL LLOQ and were excluded.

Doses spanned 1.5, 3.0 and 6.0 mg/kg single subcutaneous doses (PHYOX1,
PHYOX6), a 170 mg single dose (PHYOX5), and repeat once-monthly flat doses of
170 mg (nedosiran sodium salt; 160 mg free-acid equivalent) or 136 mg for body
weight below 50 kg (PHYOX2, PHYOX3). All participants with eGFR below
30 mL/min/1.73 m^2 were non-PH volunteers from PHYOX5. Baseline demographics
are tabulated in Supporting Information Table S1, which is not available to this
extraction; the `population` metadata records that gap explicitly rather than
substituting values.

```{r population}
str(rxode2::rxode(readModelDb("Zhang_2024_nedosiran"))$population)
```

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Zhang_2024_nedosiran.R` carries an in-file comment
naming its source. They are collected here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F) | 5.69 L/h | Table 1 |
| `lvc` (Vc/F) | 147 L | Table 1 |
| `lq` (Q/F) | 2.45 L/h | Table 1 |
| `lvp` (Vp/F) | 5300 L | Table 1 |
| `lvmax` (Vmax) | 3.94 mg/h | Table 1 |
| `lkm` (KM) | 293 ng/mL | Table 1 |
| `logitffo` (FR1) | 69.80% | Table 1 |
| `lka1` | 0.22 1/h | Table 1 |
| `lka2` | 16.7 1/h | Table 1 |
| `e_wt_cl` | 0.75 FIX | Table 1 ("Exponent for WT on CL/F and Q/F") |
| `e_wt_vc` | 1 FIX | Table 1 ("Exponent for WT on Vc/F and Vp/F") |
| `e_crcl_cl` | 0.87 | Table 1 ("Exponent for eGFR on CL/F") |
| `e_crcl_vc` | 0.19 | Table 1 ("Exponent for eGFR on Vc/F") |
| `e_wt_ka` | 0.83 | Table 1 ("Exponent for WT on ka1 and ka2") |
| `e_dis_ph1_ka1` | 1.54 | Table 1 ("Covariate effect of PH1 on ka1") |
| `lrbase` (baseline Uox) | 1420 umol/24 h | Table 2 |
| `lkout` (Kout) | 0.338 1/week | Table 2 (converted to 1/h by /168) |
| `logitimax` (Imax) | 64.5% | Table 2 |
| `lic50` (IC50) | 1.04 ng/mL | Table 2 |
| `lhill` (gamma) | 2.56 | Table 2 |
| `lke0` | ln(2) / 21.9 weeks | Table 2 (lambda = 21.9 weeks); Figure 3 (`ke0 = ln(2)/lambda`) |
| `etalvc`, `etalka1`, `etalka2`, `etalvmax` | 27.1, 42.1, 27.6, 35.3 %CV | Table 1 |
| `etalogitffo` | 0.35 (additive) | Table 1 -- see "Recovering the variability scale" |
| `etalrbase`, `etalic50`, `etalogitimax` | 31.7, 86, 29.4 %CV | Table 2 |
| `propSd` | 27.70% | Table 1; Results 3.2 ("additive on the log-scale ... equates to a proportional RUV on the normal scale") |
| `addSd_uox` | 205 umol/24 h SD | Table 2 |
| Absorption chain lengths (2 slow, 4 fast) | n/a | **Figure 3 only** -- absent from the text and both tables |
| `Eff = Imax * Ceff^gamma / (Ceff^gamma + IC50^gamma)` | n/a | Figure 3 (inset equation) |
| `d/dt(UO24) = kin * (1 - Eff) - kout * UO24` | n/a | Figure 3 |
| `CL/F = 5.69 * (WT/70)^0.75 * (eGFR/90)^0.87` | n/a | Results 3.2 (equation; coefficients from Table 1) |
| `Vc/F = 147 * (WT/70)^1 * (eGFR/90)^0.19` | n/a | Results 3.2 |
| `ka1 = 0.22 * (WT/70)^0.83 * 1.54^PH1`, `ka2 = 16.7 * (WT/70)^0.83` | n/a | Results 3.2 |

# Recovering the variability scale

Zhang 2024 never states the scale on which two of its variability terms were
estimated, and both matter for simulation. This section documents how the
ambiguity was resolved -- from the paper's own numbers, not from assumption.

**The tables report standard deviations, not variances.** Table 2's additive
residual error row is labelled explicitly `(SD, umol/24 h) 205`. Read as a
variance it would imply a residual SD of 14 umol/24 h against a
1420 umol/24 h baseline, which is not credible. Every other variability entry in
Tables 1 and 2 is a `%CV`, which the table footnote defines as the
"approximate coefficient of variation" -- i.e. `100 * omega`. The one unlabelled
entry, `Between-subject variability for FR1 (additive) 0.35`, is therefore also
a standard deviation on FR1's estimation scale.

**That estimation scale must be bounded.** FR1 = 0.698 is a dose-split fraction.
An additive SD of 0.35 applied directly to it would put about 22% of subjects
outside [0, 1] and send the fast-pathway fraction `1 - FR1` negative. The same
argument applies to Imax = 0.645 with a 29.4 %CV: log-normal on the natural
scale puts `Imax > 1` for 6.8% of subjects, which drives `kin * (1 - Eff)`
negative and 24-h urinary oxalate below zero. Both are carried on the logit
scale in this implementation.

The Imax case is directly falsifiable against the paper's own simulation.
Figure 5C plots the median and 95% prediction interval of 24-h urinary oxalate
out to week 132; digitised from the published figure, the plateau is
approximately **245 / 485 / 990 umol/24 h**. At steady state the effect-compartment
concentration is far above IC50, so `Eff` is within 0.2% of `Imax` and the
plateau reduces to `rbase * (1 - Imax)`. That makes the two candidate encodings
separable with a few thousand draws and no ODE solve:

```{r variability-scale}
set.seed(20240701)
n_draw <- 20000
rbase <- 1420 * exp(rnorm(n_draw, 0, 0.317))     # Table 2: 31.7 %CV on baseline

encodings <- list(
  "logit-normal Imax (as packaged)" = plogis(qlogis(0.645) + rnorm(n_draw, 0, 0.294)),
  "log-normal Imax (counterfactual)" = exp(log(0.645) + rnorm(n_draw, 0, 0.294))
)

scale_check <- lapply(names(encodings), function(nm) {
  uox_ss <- rbase * (1 - encodings[[nm]])
  tibble(
    encoding = nm,
    `2.5th`  = quantile(uox_ss, 0.025),
    median   = quantile(uox_ss, 0.500),
    `97.5th` = quantile(uox_ss, 0.975),
    `% below zero` = 100 * mean(uox_ss < 0)
  )
}) |>
  bind_rows() |>
  bind_rows(tibble(encoding = "Zhang 2024 Figure 5C (digitised)",
                   `2.5th` = 245, median = 485, `97.5th` = 990,
                   `% below zero` = 0))

knitr::kable(
  scale_check, digits = 1,
  caption = paste("Steady-state 24-h urinary oxalate under the two candidate",
                  "Imax variability encodings, against the plateau digitised",
                  "from Zhang 2024 Figure 5C.")
)
```

The log-normal encoding predicts a negative 2.5th percentile -- an impossibility
that the published figure does not show -- while the logit encoding reproduces
all three published percentiles. The packaged model uses the logit encoding, and
the same reading (`0.35` as a standard deviation on a bounded estimation scale)
is applied to FR1.

# Virtual cohort

Original observed data are not publicly available. Zhang 2024 built its virtual
patients from WHO weight-for-age tables (age 10 and under, via a Z-score
approach) and from NHANES (over age 10); neither table is available to this
extraction, so the weight distributions below are plausible approximations of
the published age bands rather than reproductions of them. Cohort size is
200 per arm.

Five arms are simulated: the three age/weight bands of Figure 5 at normal renal
function, plus mild and moderate renal impairment for the reference age band
(Figure 6 and Table 3).

```{r cohort}
set.seed(20240715)

n_per_arm <- 200L
tau       <- 28 * 24        # once-monthly taken as every 28 days
n_dose    <- 34L            # last dose lands at week 132
t_last    <- (n_dose - 1L) * tau

# Weight bands: adults/adolescents at or above 50 kg, the same band below
# 50 kg, and children aged 6 to under 12 years. `band` selects the sampler so
# the arm table stays a plain data frame (no list columns).
arms <- tibble::tribble(
  ~arm,                                    ~band,     ~egfr_lo, ~egfr_hi,
  "Age >=12 y, >=50 kg, 170 mg (normal)",  "adult",         90,      120,
  "Age >=12 y, <50 kg, 136 mg (normal)",   "adult_low",     90,      120,
  "Age 6-<12 y, 3.5 mg/kg (normal)",       "child",         90,      120,
  "Age >=12 y, >=50 kg, 170 mg (mild RI)", "adult",         60,       89,
  "Age >=12 y, >=50 kg, 170 mg (mod RI)",  "adult",         30,       59
)

draw_weight <- function(band, n) {
  switch(band,
    adult     = 50 + rlnorm(n, log(26), 0.42),
    adult_low = pmin(49.9, 35 + rlnorm(n, log(7), 0.6)),
    child     = 18 + rlnorm(n, log(8), 0.45),
    stop("unknown weight band: ", band)
  )
}

assign_dose <- function(band, wt) {
  switch(band,
    adult     = rep(170, length(wt)),
    adult_low = rep(136, length(wt)),
    child     = 3.5 * wt,
    stop("unknown weight band: ", band)
  )
}

make_cohort <- function(arm, band, egfr_lo, egfr_hi, id_offset) {
  wt <- draw_weight(band, n_per_arm)
  subj <- tibble(
    id      = id_offset + seq_len(n_per_arm),
    arm     = arm,
    WT      = wt,
    CRCL    = runif(n_per_arm, egfr_lo, egfr_hi),
    DIS_PH1 = 1,
    amt     = assign_dose(band, wt)
  )

  # Two dose records per administration (one per parallel absorption pathway);
  # the model's f(depot1) / f(depot2) split the amount by FR1.
  dosing <- subj |>
    tidyr::crossing(time = seq(0, by = tau, length.out = n_dose),
                    cmt  = c("depot1", "depot2")) |>
    mutate(evid = 1L)

  # Observation grids: dense over the first dosing interval (Figures 5A / 6A),
  # weekly for urinary oxalate (Figures 5C / 6C and Table 3), and dense over the
  # final dosing interval for steady-state NCA.
  obs_times <- sort(unique(c(
    seq(0, 72, by = 1),
    seq(0, 132 * 168, by = 168),
    t_last + c(seq(0, 24, by = 1), seq(28, 96, by = 4), seq(120, tau, by = 24))
  )))
  obs <- subj |>
    select(-amt) |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "Cc")

  bind_rows(dosing, obs) |> arrange(id, time, desc(evid))
}

events <- do.call(
  bind_rows,
  lapply(seq_len(nrow(arms)), function(i) {
    do.call(make_cohort, c(as.list(arms[i, ]), id_offset = (i - 1L) * n_per_arm))
  })
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid", "cmt")])))
cat(sprintf("%d subjects across %d arms; %d event rows\n",
            dplyr::n_distinct(events$id), nrow(arms), nrow(events)))
```

# Simulation

`Cc` and `uox` are both declared endpoints (two `~` residual lines), so rxode2
assigns each an endpoint slot after the ten ODE states. Observation rows
therefore address the endpoint by name (`cmt = "Cc"`), not an ODE state; every
model variable is returned as a column regardless. `useLinCmt = FALSE` is
required because rxode2's automatic ODE-to-linCmt conversion corrupts the
endpoint mapping for models of this size.

```{r simulate}
mod <- readModelDb("Zhang_2024_nedosiran")

sim <- rxode2::rxSolve(
  mod, events = events,
  keep        = c("arm", "WT", "CRCL"),
  useLinCmt   = FALSE,
  returnType  = "data.frame"
)

stopifnot(dplyr::n_distinct(sim$id) == n_per_arm * nrow(arms))
```

The `stopifnot()` above is deliberate: `rxSolve()` can silently drop subjects,
so the subject count is asserted rather than assumed.

# Replicate published figures

## Figure 5A -- concentration-time profiles by age band

```{r figure-5a}
# Replicates Figure 5A of Zhang 2024: median simulated nedosiran concentration
# over the first 72 h, stratified by age/weight band (normal renal function).
sim |>
  filter(time <= 72, grepl("normal", arm, fixed = TRUE)) |>
  group_by(arm, time) |>
  summarise(Q50 = median(Cc), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = arm)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  scale_x_continuous(breaks = seq(0, 72, by = 12)) +
  labs(x = "Time since first dose (h)", y = "Median concentration (ng/mL)",
       colour = NULL,
       title = "Figure 5A - nedosiran concentration by age band",
       caption = "Replicates Figure 5A of Zhang 2024.") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 3))
```

```{r figure-5a-check}
sim |>
  filter(time <= 72, grepl("normal", arm, fixed = TRUE)) |>
  group_by(arm, time) |>
  summarise(Q50 = median(Cc), .groups = "drop") |>
  group_by(arm) |>
  summarise(
    `Median Cmax (ng/mL)` = max(Q50),
    `Tmax (h)`            = time[which.max(Q50)],
    `C at 72 h (ng/mL)`   = Q50[time == 72],
    .groups = "drop"
  ) |>
  rename(Arm = arm) |>
  knitr::kable(digits = 2,
               caption = paste("Median first-dose profile summaries. Zhang 2024",
                               "Figure 5A shows the reference band peaking near",
                               "500 ng/mL at 6-9 h and falling below 1 ng/mL by",
                               "72 h; PHYOX1 reported Tmax between 6 and 12 h."))
```

## Figure 5C -- 24-h urinary oxalate by age band

```{r figure-5c}
# Replicates Figure 5C of Zhang 2024: median and 95% prediction interval of
# 24-h urinary oxalate to week 132, stratified by age/weight band.
uox_weekly <- sim |>
  filter(time %in% seq(0, 132 * 168, by = 168)) |>
  mutate(week = time / 168)

uox_weekly |>
  filter(grepl("normal", arm, fixed = TRUE)) |>
  group_by(arm, week) |>
  summarise(Q025 = quantile(uox, 0.025), Q50 = median(uox),
            Q975 = quantile(uox, 0.975), .groups = "drop") |>
  ggplot(aes(week, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = c(460, 600), linetype = "dashed", colour = "grey30") +
  scale_x_continuous(breaks = seq(0, 132, by = 12)) +
  labs(x = "Time since first dose (weeks)", y = "24-h urinary oxalate (umol/24 h)",
       colour = NULL, fill = NULL,
       title = "Figure 5C - 24-h urinary oxalate by age band",
       caption = paste("Replicates Figure 5C of Zhang 2024. Dashed lines mark the",
                       "normal (<460) and near-normal (<600) limits.")) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 3), fill = guide_legend(nrow = 3))
```

```{r figure-5c-check}
uox_weekly |>
  filter(arm == arms$arm[1], week %in% c(0, 132)) |>
  group_by(week) |>
  summarise(`2.5th` = quantile(uox, 0.025), median = median(uox),
            `97.5th` = quantile(uox, 0.975), .groups = "drop") |>
  mutate(`Zhang 2024 Figure 5C (digitised)` =
           ifelse(week == 0, "710 / 1350 / 2490", "245 / 485 / 990")) |>
  rename(Week = week) |>
  knitr::kable(digits = 0,
               caption = paste("Reference band (age >=12 y, >=50 kg, 170 mg Q1M,",
                               "normal renal function) against the digitised",
                               "Figure 5C interval at baseline and at plateau."))
```

## Figure 6C and Table 3 -- renal-function stratification

```{r figure-6c}
# Replicates Figure 6C of Zhang 2024: 24-h urinary oxalate for the reference
# age band stratified by renal function.
uox_weekly |>
  filter(grepl(">=50 kg, 170 mg", arm, fixed = TRUE)) |>
  group_by(arm, week) |>
  summarise(Q025 = quantile(uox, 0.025), Q50 = median(uox),
            Q975 = quantile(uox, 0.975), .groups = "drop") |>
  ggplot(aes(week, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = c(460, 600), linetype = "dashed", colour = "grey30") +
  scale_x_continuous(breaks = seq(0, 132, by = 12)) +
  labs(x = "Time since first dose (weeks)", y = "24-h urinary oxalate (umol/24 h)",
       colour = NULL, fill = NULL,
       title = "Figure 6C - 24-h urinary oxalate by renal function",
       caption = "Replicates Figure 6C of Zhang 2024.") +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 3), fill = guide_legend(nrow = 3))
```

Table 3 of Zhang 2024 reports the percentage of PH1 patients reaching the normal
(<460 umol/24 h) and near-normal (<600 umol/24 h) ranges for the reference age
band under normal, mild and moderate renal function. Note that the Table 3
*caption* swaps the two impairment definitions relative to the Discussion; the
Discussion values (mild = eGFR 60-89, moderate = eGFR 30-59) are the standard
clinical bands, are consistent with the monotone ordering of the tabulated
percentages, and are what this vignette uses.

```{r table-3}
published_tbl3 <- tibble::tribble(
  ~renal,     ~week, ~pub_lt460, ~pub_lt600,
  "normal",      13,       31.3,       57.6,
  "normal",      26,       40.0,       67.6,
  "normal",      52,       42.4,       71.1,
  "normal",     128,       43.7,       72.1,
  "mild RI",     13,       34.1,       61.5,
  "mild RI",     26,       41.3,       69.9,
  "mild RI",     52,       43.5,       72.3,
  "mild RI",    128,       43.9,       72.9,
  "mod RI",      13,       36.0,       63.8,
  "mod RI",      26,       42.4,       71.0,
  "mod RI",      52,       43.8,       72.6,
  "mod RI",     128,       44.1,       73.1
)

uox_weekly |>
  filter(grepl(">=50 kg, 170 mg", arm, fixed = TRUE),
         week %in% c(13, 26, 52, 128)) |>
  mutate(renal = case_when(grepl("mild",   arm) ~ "mild RI",
                           grepl("mod",    arm) ~ "mod RI",
                           TRUE                 ~ "normal")) |>
  group_by(renal, week) |>
  summarise(sim_lt460 = 100 * mean(uox < 460),
            sim_lt600 = 100 * mean(uox < 600), .groups = "drop") |>
  left_join(published_tbl3, by = c("renal", "week")) |>
  arrange(match(renal, c("normal", "mild RI", "mod RI")), week) |>
  rename("Renal function" = renal, "Week" = week,
         "Simulated % <460" = sim_lt460, "Published % <460" = pub_lt460,
         "Simulated % <600" = sim_lt600, "Published % <600" = pub_lt600) |>
  knitr::kable(digits = 1,
               caption = paste("Percentage of simulated PH1 patients reaching the",
                               "normal and near-normal 24-h urinary oxalate ranges,",
                               "against Zhang 2024 Table 3."))
```

The normal-renal-function rows track the published percentages closely. The
renal ordering does **not** reproduce reliably, and that is expected rather than
a model defect: the published effect of renal function on this endpoint is only
about 1.4 percentage points across the three strata (42.4 -> 43.5 -> 43.8 at
week 52), because at steady state the effect-compartment concentration is far
above IC50 in every stratum, so `Eff` is essentially `Imax` regardless of
exposure. With 200 subjects per arm the binomial standard error on each
percentage is roughly 3.5 points, which is more than twice the effect being
resolved. The paper used 1000 subjects per arm. This vignette keeps 200 per arm
per the library's cohort cap, so these rows demonstrate the magnitude, not the
ordering.

# PKNCA validation

NCA is run over the final once-monthly dosing interval (steady state), with a
treatment grouping so results can be compared per arm.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc), time >= t_last) |>
  mutate(time_ss = time - t_last) |>
  select(id, arm, time_ss, Cc)

# Guarantee a time-zero record per subject so PKNCA can anchor the interval.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, arm) |> mutate(time_ss = 0, Cc = 0)
) |>
  distinct(id, arm, time_ss, .keep_all = TRUE) |>
  arrange(id, time_ss)

dose_nca <- events |>
  filter(evid == 1, time == t_last, cmt == "depot1") |>
  mutate(time_ss = 0) |>
  select(id, arm, time_ss, amt)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time_ss | arm + id)
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time_ss | arm + id)

intervals <- data.frame(start = 0, end = tau,
                        cmax = TRUE, tmax = TRUE, auclast = TRUE)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published exposure metrics

Zhang 2024 Figure 5B reports the model-predicted fold change in AUC<sub>0-tau,ss</sub>
for each age/weight band relative to the reference band, with 95% prediction
intervals. That is a per-arm ratio, so it is directly comparable to the
simulated NCA above.

```{r nca-compare}
auc_by_subject <- as.data.frame(nca_res) |>
  filter(PPTESTCD == "auclast") |>
  select(arm, id, auclast = PPORRES)

ref_arm    <- arms$arm[1]
ref_median <- median(auc_by_subject$auclast[auc_by_subject$arm == ref_arm])

published_fig5b <- tibble::tribble(
  ~arm,        ~pub_median, ~pub_lo, ~pub_hi,
  arms$arm[1],        1.00,    0.62,    1.52,
  arms$arm[2],        1.08,    0.69,    1.63,
  arms$arm[3],        0.94,    0.52,    1.47
)

auc_by_subject |>
  filter(arm %in% published_fig5b$arm) |>
  group_by(arm) |>
  summarise(sim_median = median(auclast) / ref_median,
            sim_lo     = quantile(auclast, 0.025) / ref_median,
            sim_hi     = quantile(auclast, 0.975) / ref_median,
            .groups = "drop") |>
  left_join(published_fig5b, by = "arm") |>
  mutate(across(where(is.numeric), \(x) round(x, 2)),
         `Simulated (95% PI)` = sprintf("%.2f (%.2f - %.2f)", sim_median, sim_lo, sim_hi),
         `Published (95% PI)` = sprintf("%.2f (%.2f - %.2f)", pub_median, pub_lo, pub_hi)) |>
  select(Arm = arm, `Simulated (95% PI)`, `Published (95% PI)`) |>
  knitr::kable(caption = paste("Fold change in AUC0-tau,ss relative to the",
                               "reference band, against Zhang 2024 Figure 5B."))
```

The two adult/adolescent bands agree closely (simulated 1.06 vs published 1.08
for the 136 mg band below 50 kg), which is the comparison that does not depend on
a weight table this extraction lacks. The paediatric band comes out low
(simulated ~0.72 vs published 0.94, a >20% difference) and the cause is the
virtual cohort, not the model: the paediatric dose is weight-based
(3.5 mg/kg), so the arm's exposure is driven directly by the weight
distribution, and Zhang 2024 built theirs from WHO weight-for-age tables that
are not available here. The Michaelis-Menten elimination arm amplifies the
sensitivity, because a smaller absolute dose saturates it less and therefore
yields less-than-proportional exposure -- the same nonlinearity the paper
invokes to justify the parallel elimination pathways. Shifting the paediatric
weight distribution upward closes the gap; no parameter has been tuned to
achieve that.

```{r nca-summary}
as.data.frame(nca_res) |>
  filter(PPTESTCD %in% c("cmax", "tmax", "auclast")) |>
  group_by(arm, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median) |>
  rename(Arm = arm, "Cmax,ss (ng/mL)" = cmax, "Tmax,ss (h)" = tmax,
         "AUC0-tau,ss (ng*h/mL)" = auclast) |>
  knitr::kable(digits = 1,
               caption = "Median steady-state NCA metrics by arm.")
```

# Assumptions and deviations

* **Table 1 supersedes the in-text equations.** Results section 3.2 prints a
  parameter set that differs from Table 1: CL/F 5.73 vs **5.69**, the eGFR-on-CL
  exponent 0.864 vs **0.87**, Vc/F 145 vs **147**, ka1 0.214 vs **0.22**, the
  WT-on-ka exponent 0.816 vs **0.83**, the PH1-on-ka1 factor 1.56 vs **1.54**,
  and ka2 16.6 vs **16.7**. These are not roundings of one another, so the two
  sets come from different model runs. This extraction uses Table 1 throughout:
  it is the formal parameter-estimates table (carrying %RSE and eta shrinkage),
  the text refers to it for "the parameter estimates for the POP-PK model", and
  it is the only source covering every parameter the model needs. Mixing the two
  sources would be worse than choosing one.
* **Absorption chain lengths come from Figure 3 alone.** Neither the text nor
  either table states how many transit compartments each pathway has; the 2-slow
  / 4-fast structure was read from the Figure 3 schematic.
* **Variability scale.** `0.35` for FR1 is read as a standard deviation on a
  bounded (logit) estimation scale, and the Imax `%CV` likewise, for the reasons
  and against the Figure 5C evidence set out in "Recovering the variability
  scale" above. The paper does not state either scale. The `%CV` entries are
  converted as `omega^2 = CV^2`, following the table footnote's definition of
  `%CV` as the "approximate coefficient of variation"; the alternative
  `omega^2 = log(1 + CV^2)` differs by under 2% here and does not change any
  conclusion.
* **Renal maturation function (RMF).** The published CL/F and Vc/F equations
  write the renal covariate as `eGFR * RMF`, citing Anderson & Holford 2011
  (reference 31), which is not available to this extraction. RMF is used only
  when deriving eGFR for the virtual paediatric cohort of the simulation section
  and equals 1 in adults, so no fitted parameter depends on its form. The
  `CRCL` column therefore carries the already-matured eGFR and no RMF form is
  reproduced.
* **Table 3 caption error.** The Table 3 caption gives "mild renal function
  (eGFR 30-59)" and "moderate renal function (eGFR 60-89)", which are swapped
  relative to the Discussion ("mild (eGFR = 60-89) and moderate (eGFR =
  30-59)") and relative to standard clinical definitions. The Discussion
  assignment is used here; it is also the one consistent with the monotone
  ordering of the tabulated percentages (lower eGFR, higher exposure, greater
  response).
* **Once-monthly dosing is taken as every 28 days.** The paper writes "Q1M"
  without defining it in days and reports outcomes on a weekly axis. Steady-state
  urinary oxalate is insensitive to this choice because the effect-compartment
  concentration sits far above IC50 either way.
* **Virtual cohort demographics are approximations.** Zhang 2024 generated
  virtual patients from WHO weight-for-age tables and NHANES; neither is
  available to this extraction. The weight and eGFR distributions here are
  plausible stand-ins for the published age and renal bands, so per-arm
  percentages will not match Table 3 exactly. This is the sole cause of the two
  flagged discrepancies above -- the paediatric AUC fold change and the renal
  ordering in Table 3 -- both of which are diagnosed in place. No parameter was
  adjusted to improve either.
* **No IIV on CL/F, Q/F, Vp/F or KM.** Table 1 reports between-subject
  variability only on Vc/F, ka1, ka2, Vmax and FR1; the model carries no other
  PK etas. Similarly, Table 2 reports PD variability only on baseline urinary
  oxalate, IC50 and Imax -- not on Kout, despite Methods section 2.4 describing
  a Kout eta in the *prior* model. The final model as tabulated is what is
  implemented.
* **Supporting Information is not on disk.** Tables S1-S3 and Figures S1-S4
  were not available. Table 1 and Table 2 carry the complete final parameter set
  and Figure 3 the complete structure, so nothing the model needs is missing;
  the gap affects baseline demographics, the covariate screening list, and the
  goodness-of-fit diagnostics only.
* **`cmt = "Cc"` on observation rows is correct here.** The model declares two
  endpoints, so rxode2 places endpoint slots after the ten ODE states and
  observation rows must address an endpoint rather than an ODE state.
  `nlmixr2libingest`'s `lint_vignette.R` flags this pattern as
  `[cmt-observable]`; that warning is a known false positive for declared
  multi-endpoint models, where `cmt = "central"` fails outright.
* **All simulated participants are PH1 patients** (`DIS_PH1 = 1`), matching the
  populations of Figures 5 and 6 and of Table 3.
