---
title: "Olanzapine (Zhang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Olanzapine (Zhang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhang_2025_olanzapine"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhang_2025_olanzapine"))$description`
* Article: <https://doi.org/10.3389/fpsyt.2025.1538996>

Zhang and colleagues fitted a one-compartment model with first-order oral
absorption to routine therapeutic-drug-monitoring (TDM) concentrations from 72
inpatients with major depressive disorder. The absorption rate constant was not
identifiable from sparse trough-dominated TDM data and was fixed to a published
value; apparent oral clearance and apparent volume were estimated. Of the 24
concomitant medications and the demographic and clinical-chemistry panel
screened, only body weight and paroxetine were retained: patients co-prescribed
paroxetine had 28.9% lower olanzapine CL/F.

This is the same group's companion analysis to `Zhang_2024_olanzapine`, which
studied olanzapine in schizophrenia and found concomitant aripiprazole rather
than paroxetine. The two share the structural model, the allometric form and
the linear-shift covariate parameterisation, but were fitted to different
cohorts and give materially different disposition estimates (V/F 197 L here
versus 854 L there).

# Population

Seventy-two inpatients with major depressive disorder treated at a single
centre in Xuzhou, Jiangsu, China contributed olanzapine TDM concentrations
collected between December 2020 and August 2023, with one to three
concentration samples per patient (Methods, Data Collection; Results, Patient
information). The cohort was 17 men and 55 women, mean age 48.14 years
(SD 20.94, range 16.00-87.90) and mean weight 61.83 kg (SD 11.82, range
40.00-92.00) (Table 1). Baseline laboratory values were broadly unremarkable:
mean albumin 39.51 g/L, creatinine 53.88 umol/L, hemoglobin 126.81 g/L.
Eighteen of the 72 patients were co-prescribed paroxetine hydrochloride tablets
(Table 2).

The paper reports no administered olanzapine doses, dose ranges, or sampling
times; the concentrations come from a routine TDM database rather than a
designed PK study. Every dosing regimen used below is therefore a scenario
constructed for validation, not a replication of the observed data.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Zhang_2025_olanzapine")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Zhang_2025_olanzapine.R`. The table below
collects them in one place.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (CL/F at 70 kg, no paroxetine) | 19.6 L/h | Table 3 (SE 7.1%; bootstrap median 19.4 [17.0, 21.9]); also printed in Equation (6) |
| `lvc` (V/F at 70 kg) | 197 L | Table 3 (SE 15.3%; bootstrap median 194 [154, 277]); also printed in Equation (7) |
| `lka` (ka, fixed) | 0.861 1/h | Table 3, "0.861 (fixed)"; Methods, Modeling: "absorption rate constants [Ka, fixed at 0.861/h (26)]". Reference 26 is Sun 2021, J Clin Pharmacol 61(11):1430-1441 |
| `e_wt_cl` (allometric exponent on CL/F, fixed) | 0.75 | Methods, Equation (3): "F is the allometric coefficient: 0.75 for the CL/F and 1 for the V/F", citing Anderson & Holford 2008; the exponent is printed again in Equation (6) |
| `e_wt_vc` (allometric exponent on V/F, fixed) | 1 | Methods, Equation (3); printed again in Equation (7) |
| `e_par_cl` (concomitant paroxetine on CL/F) | -0.289 | Table 3, theta_PAR (SE 30.5%; bootstrap median -0.283 [-0.420, -0.078]); assembled in Equation (6) as `(1 - 0.289 * PAR)` |
| `etalcl` (IIV on CL/F) | omega = 0.434, so variance 0.434^2 | Table 3, row `omega_CL/F` (SE 11.1%; bootstrap median 0.429 [0.336, 0.535]). Read as a standard deviation -- see Assumptions below |
| `propSd` (proportional residual error) | 0.153 | Table 3, row `sigma_1`, "residual variability, proportional error" (SE 16.6%) |
| `addSd` (additive residual error) | 1.005 ng/mL | Table 3, row `sigma_2`, "residual variability, additive error" (SE 45.0%) |
| IIV structure `Ai = TV(A) * exp(eta_i)` | n/a | Methods, Equation (1) |
| Residual structure `Bi = Ci + Ci*eps_1 + eps_2` | n/a | Methods, Equation (2) |
| Allometry `Di = Dstd * (Ei / Estd)^F`, `Estd = 70 kg` | n/a | Methods, Equation (3) |
| Categorical covariate form (linear shift) | n/a | Methods, Equation (5); realised in Equation (6) |
| `CL/F = 19.6 * (weight/70)^0.75 * (1 - 0.289 * PAR)` | n/a | Results, Modeling, Equation (6) |
| `V/F = 197 * (weight/70)` | n/a | Results, Modeling, Equation (7) |
| Clearance ratio 0.711:1 with vs without paroxetine | n/a | Results, Evaluation (Figure 1H); restated in the Discussion |
| Therapeutic window 20-80 ng/mL | n/a | Methods, Simulation (cites the paper's reference 28, Ding 2022) |
| Covariate screen (dOFV per drug) | n/a | Supplementary Table 1 |
| Initial-dose recommendations | n/a | Table 4 and Figure 4 |

```{r model}
mod <- readModelDb("Zhang_2025_olanzapine")
ui  <- rxode2::rxode(mod)
```

# Structural checks against the printed equations

The published model is fully specified by Equations (6) and (7), so the first
check is that the packaged model reproduces them exactly. `zeroRe()` removes
between-subject variability so the solved `cl` and `vc` are the typical values.

```{r structural-check}
# Published equations, transcribed independently of the model file.
cl_published <- function(wt, par) 19.6 * (wt / 70)^0.75 * (1 - 0.289 * par)
vc_published <- function(wt)      197  * (wt / 70)

weights <- c(40, 50, 60, 70, 80, 90, 100)

grid <- tidyr::expand_grid(WT = weights, CONMED_PAROXETINE = c(0, 1)) |>
  dplyr::mutate(id = dplyr::row_number(),
                arm = paste0(WT, " kg, ",
                             ifelse(CONMED_PAROXETINE == 1, "with", "no"),
                             " paroxetine"))

ev_struct <- grid |>
  dplyr::mutate(time = 0, amt = 10, evid = 1, cmt = "depot") |>
  dplyr::bind_rows(
    grid |> tidyr::crossing(time = 0:1) |>
      dplyr::mutate(amt = NA_real_, evid = 0, cmt = "central")
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_struct <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_struct,
  keep = c("WT", "CONMED_PAROXETINE", "arm")
) |>
  as.data.frame() |>
  dplyr::group_by(WT, CONMED_PAROXETINE, arm) |>
  dplyr::summarise(cl_model = mean(cl), vc_model = mean(vc), ka_model = mean(ka),
                   .groups = "drop") |>
  dplyr::mutate(cl_paper = cl_published(WT, CONMED_PAROXETINE),
                vc_paper = vc_published(WT))

stopifnot(
  nrow(sim_struct) == 14L,
  max(abs(sim_struct$cl_model - sim_struct$cl_paper)) < 1e-8,
  max(abs(sim_struct$vc_model - sim_struct$vc_paper)) < 1e-8,
  max(abs(sim_struct$ka_model - 0.861)) < 1e-8
)

sim_struct |>
  dplyr::transmute(
    Arm                 = arm,
    `CL/F model (L/h)`  = round(cl_model, 3),
    `CL/F Eq. 6 (L/h)`  = round(cl_paper, 3),
    `V/F model (L)`     = round(vc_model, 1),
    `V/F Eq. 7 (L)`     = round(vc_paper, 1)
  ) |>
  knitr::kable(caption = "Typical CL/F and V/F reproduce Equations (6) and (7) exactly.")
```

The Results and Discussion state the paroxetine effect as a clearance ratio:
"with the same weight, the clearance rates of olanzapine were 0.711:1 in
patients with major depressive disorder with or without paroxetine"
(Figure 1H). That ratio is `1 - 0.289` and is weight-independent, so it must
hold at every weight.

```{r clearance-ratio}
ratio <- sim_struct |>
  dplyr::select(WT, CONMED_PAROXETINE, cl_model) |>
  tidyr::pivot_wider(names_from = CONMED_PAROXETINE,
                     names_prefix = "par", values_from = cl_model) |>
  dplyr::mutate(ratio = par1 / par0)

# Deterministic (no IIV): an exact identity, so a tight bound is correct here.
stopifnot(nrow(ratio) == 7L, max(abs(ratio$ratio - 0.711)) < 1e-9)

ratio |>
  dplyr::transmute(`Weight (kg)` = WT,
                   `CL/F without paroxetine (L/h)` = round(par0, 2),
                   `CL/F with paroxetine (L/h)`    = round(par1, 2),
                   `Ratio (with:without)`          = round(ratio, 4)) |>
  knitr::kable(caption = "Replicates Figure 1H of Zhang 2025: a clearance ratio of 0.711:1 with vs without paroxetine, at every weight.")
```

# PKNCA validation against the published parameters

Zhang 2025 reports no NCA table, but the printed structural parameters imply
exact non-compartmental quantities for a single oral dose of a linear
one-compartment model:

* `AUC0-inf = Dose / (CL/F)` (mass balance; independent of `ka` and `V/F`),
* terminal `t1/2 = ln(2) * (V/F) / (CL/F)`, because `ka = 0.861 1/h` is an
  order of magnitude larger than `kel` (0.099 1/h without paroxetine at 70 kg,
  0.071 1/h with), so the terminal phase is elimination-rate-limited rather
  than flip-flop,
* `tmax = ln(ka / kel) / (ka - kel)`.

Running PKNCA over a typical-value (no-IIV) simulation therefore checks the
whole pipeline -- units, the mg-to-ng/mL conversion, the ODE system, the
covariate model -- against numbers derived from Table 3 alone.

```{r nca-sim}
# Single 10 mg oral dose; observe long enough for AUC0-inf to converge. The
# slowest arm (100 kg with paroxetine) has t1/2 = 10.7 h, so 240 h is >22
# half-lives.
obs_times <- sort(unique(c(seq(0, 24, by = 0.1), seq(24, 240, by = 0.5))))

ev_nca <- grid |>
  dplyr::mutate(time = 0, amt = 10, evid = 1, cmt = "depot") |>
  dplyr::bind_rows(
    grid |> tidyr::crossing(time = obs_times) |>
      dplyr::mutate(amt = NA_real_, evid = 0, cmt = "central")
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(ev_nca[, c("id", "time", "evid")])))

sim_nca_raw <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_nca,
  keep = c("WT", "CONMED_PAROXETINE", "arm")
) |>
  as.data.frame()

# Concentrations must stay non-negative for PKNCA's log-linear terminal fit.
stopifnot(all(sim_nca_raw$Cc >= 0, na.rm = TRUE))
```

```{r pknca}
sim_nca <- sim_nca_raw |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

# Guarantee a time = 0 row per (id, arm); pre-dose Cc = 0 for extravascular.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, arm) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, arm, time, .keep_all = TRUE) |>
  dplyr::arrange(id, arm, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "ng/mL", timeu = "h")

dose_df <- ev_nca |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start      = 0,
  end        = Inf,
  cmax       = TRUE,
  tmax       = TRUE,
  aucinf.obs = TRUE,
  half.life  = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against the values implied by Table 3

```{r nca-compare}
# Reference values derived from the PRINTED parameters only.
ka_pub <- 0.861
published <- grid |>
  dplyr::mutate(
    cl  = cl_published(WT, CONMED_PAROXETINE),
    vc  = vc_published(WT),
    kel = cl / vc,
    # 10 mg / (L/h) -> mg*h/L; * 1000 -> ng*h/mL
    aucinf.obs = 1000 * 10 / cl,
    half.life  = log(2) / kel,
    tmax       = log(ka_pub / kel) / (ka_pub - kel),
    cmax       = 1000 * (10 / vc) * (ka_pub / (ka_pub - kel)) *
      (exp(-kel * tmax) - exp(-ka_pub * tmax))
  ) |>
  dplyr::select(arm, cmax, tmax, aucinf.obs, half.life)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published,
  by            = "arm",
  units         = c(cmax = "ng/mL", tmax = "h",
                    aucinf.obs = "ng*h/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated NCA vs the closed-form values implied by Zhang 2025 Table 3",
    "(10 mg single oral dose). * marks a >20% difference."
  )
)
```

```{r nca-gate}
# These are deterministic comparisons -- a typical-value solve against a closed
# form built from the paper's printed numbers -- so the residual difference is
# pure numerical (grid resolution, log-linear terminal fit) and a tight bound
# is correct. Contrast with the cohort-derived attainment table below.
nca_wide <- as.data.frame(nca_res$result) |>
  dplyr::select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(published, by = "arm", suffix = c("_sim", "_pub"))

pct <- function(a, b) 100 * (a - b) / b
stopifnot(
  nrow(nca_wide) == 14L,
  max(abs(pct(nca_wide$aucinf.obs_sim, nca_wide$aucinf.obs_pub))) < 0.5,
  max(abs(pct(nca_wide$half.life_sim,  nca_wide$half.life_pub)))  < 0.5,
  max(abs(pct(nca_wide$cmax_sim,       nca_wide$cmax_pub)))       < 1.0,
  max(abs(pct(nca_wide$tmax_sim,       nca_wide$tmax_pub)))       < 2.0
)
```

`AUC0-inf` matches `Dose / (CL/F)` and the terminal half-life matches
`ln(2) * (V/F) / (CL/F)` to well under 1%, confirming that the packaged model,
its unit conversion and its covariate model all agree with Table 3.

# Replicate Figure 3: steady-state profiles at the recommended doses

Figure 3 of Zhang 2025 plots simulated olanzapine concentration-time profiles
for seven weight groups under four conditions -- once- or twice-daily dosing,
with or without paroxetine -- against the 20-80 ng/mL therapeutic band. The
panels below show the typical-value steady-state profile at the dose Table 4
recommends for each weight and condition. Once-daily dosing without paroxetine
is the one condition for which Table 4 recommends nothing (no dose from 0.1 to
1.0 mg/kg/day reached 55% attainment); it is shown at 0.6 mg/kg/day, the dose
the attainment sweep below finds best, and is excluded from the band gate.

```{r recommended-doses}
# Table 4 of Zhang 2025, transcribed. Weight bands are closed on the left.
# The once-daily / no-paroxetine row of Table 4 records no recommendation.
recommended_dose <- function(wt, par, ndose) {
  if (par == 0 && ndose == 1) NA_real_
  else if (par == 0 && ndose == 2) ifelse(wt < 56, 0.5, 0.4)
  else if (par == 1 && ndose == 1) ifelse(wt < 60, 0.5, 0.4)
  else ifelse(wt < 85, 0.3, 0.2)
}

conditions <- tidyr::expand_grid(
  CONMED_PAROXETINE = c(0, 1),
  ndose             = c(1, 2),
  WT                = weights
) |>
  dplyr::rowwise() |>
  dplyr::mutate(table4_dose = recommended_dose(WT, CONMED_PAROXETINE, ndose)) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    # No Table 4 recommendation for once-daily without paroxetine; use the
    # sweep-optimal 0.6 mg/kg/day so the panel is still informative.
    mg_per_kg_day = ifelse(is.na(table4_dose), 0.6, table4_dose),
    tau           = 24 / ndose,
    amt           = mg_per_kg_day * WT / ndose,
    condition     = paste0(ifelse(ndose == 1, "Once daily", "Twice daily"), ", ",
                           ifelse(CONMED_PAROXETINE == 1, "with", "no"),
                           " paroxetine"),
    wt_label      = paste0(WT, " kg")
  )

conditions |>
  dplyr::transmute(Condition = condition, `Weight (kg)` = WT,
                   `Table 4 dose (mg/kg/day)` = table4_dose,
                   `Dose simulated (mg/kg/day)` = mg_per_kg_day,
                   `Dose per administration (mg)` = round(amt, 1)) |>
  knitr::kable(caption = "Recommended initial doses transcribed from Table 4 of Zhang 2025. A blank Table 4 dose marks the once-daily, no-paroxetine condition, for which the paper recommends no dose.")
```

```{r figure-3-events}
# 10 days of run-in reaches steady state: the slowest arm (100 kg with
# paroxetine) has t1/2 = ln(2) / 0.0647 = 10.7 h, so 240 h is >22 half-lives.
run_in_h <- 240

make_arm <- function(row, id) {
  tau  <- row$tau
  dose <- tibble::tibble(
    id = id, time = seq(0, run_in_h - tau, by = tau),
    amt = row$amt, evid = 1, cmt = "depot"
  )
  last_dose <- max(dose$time)
  obs <- tibble::tibble(
    id = id, time = seq(last_dose, last_dose + tau, by = tau / 96),
    amt = NA_real_, evid = 0, cmt = "central"
  )
  dplyr::bind_rows(dose, obs) |>
    dplyr::mutate(WT = row$WT,
                  CONMED_PAROXETINE = row$CONMED_PAROXETINE,
                  condition = row$condition, wt_label = row$wt_label,
                  tau = tau)
}

ev_fig3 <- do.call(
  dplyr::bind_rows,
  lapply(seq_len(nrow(conditions)),
         function(i) make_arm(conditions[i, ], id = i))
) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

stopifnot(!anyDuplicated(unique(ev_fig3[, c("id", "time", "evid")])))
```

```{r figure-3}
sim_fig3 <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_fig3,
  keep = c("WT", "CONMED_PAROXETINE", "condition", "wt_label")
) |>
  as.data.frame() |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::group_by(id) |>
  dplyr::mutate(t_rel = time - min(time)) |>
  dplyr::ungroup() |>
  dplyr::mutate(wt_label = factor(wt_label, levels = paste0(weights, " kg")))

ggplot(sim_fig3, aes(t_rel, Cc, colour = wt_label)) +
  geom_hline(yintercept = c(20, 80), linetype = "dashed", colour = "red") +
  geom_line() +
  facet_wrap(~condition) +
  labs(x = "Time within the steady-state dosing interval (h)",
       y = "Olanzapine concentration (ng/mL)",
       colour = "Body weight",
       title = "Steady-state profiles at the Table 4 recommended doses",
       caption = "Replicates the structure of Figure 3 of Zhang 2025; red dashed lines are the 20-80 ng/mL therapeutic window.")
```

```{r figure-3-gate}
band <- sim_fig3 |>
  dplyr::group_by(condition, wt_label) |>
  dplyr::summarise(cmin = min(Cc), cmax = max(Cc),
                   cav  = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2) /
                     (max(time) - min(time)),
                   .groups = "drop")

# Deterministic (zeroRe) typical-value profiles -- no cohort noise anywhere in
# this chunk, so tight bounds are correct here. A mis-transcribed dose,
# clearance, volume or unit conversion moves these by tens of percent.
gated <- band |> dplyr::filter(condition != "Once daily, no paroxetine")
once_daily_band  <- band |> dplyr::filter(grepl("^Once daily",  condition))
twice_daily_band <- band |> dplyr::filter(grepl("^Twice daily", condition))
stopifnot(
  nrow(band) == 28L,
  nrow(gated) == 21L,
  # At the 21 doses Table 4 recommends, the typical patient's steady-state
  # TROUGH -- the quantity the paper's attainment criterion is applied to --
  # lies inside the therapeutic window at every weight.
  all(gated$cmin >= 20 & gated$cmin <= 80),
  # The interval-average and peak are a different story, and the contrast is
  # the structural reason Table 4 recommends twice-daily dosing. With a
  # terminal half-life of only 7-11 h, every once-daily peak clears the
  # 80 ng/mL ceiling by a wide margin, while no twice-daily peak does.
  all(once_daily_band$cmax  > 120),
  all(twice_daily_band$cmax < 100)
)

band |>
  dplyr::transmute(Condition = condition, Weight = wt_label,
                   `Ctrough (ng/mL)` = round(cmin, 1),
                   `Cav (ng/mL)`     = round(cav, 1),
                   `Cmax (ng/mL)`    = round(cmax, 1)) |>
  knitr::kable(caption = "Typical steady-state exposure at the simulated doses, against the 20-80 ng/mL window.")
```

The terminal half-life implied by Table 3 is short -- about 7 h without
paroxetine and 10 h with it -- so a 24 h dosing interval is more than two
half-lives wide. The consequence is visible in the table above: at any dose
large enough to hold the once-daily trough inside the window, the once-daily
peak overshoots 80 ng/mL roughly two-fold, whereas splitting the same daily
dose in two keeps the whole profile close to the band. That is the structural
reason Table 4 recommends no once-daily dose at all in the absence of
paroxetine, and why the paper concludes twice-daily administration is optimal.

# Target attainment (Table 4 and Figure 4)

Zhang 2025 simulated 1000 virtual patients per case against the 20-80 ng/mL
window over ten dose levels (0.1 to 1.0 mg/kg/day) and seven weight groups
(Methods, Simulation; results in Figure 4 and Table 4). The reconstruction
below applies the same criterion to the steady-state trough -- the quantity a
TDM service actually measures -- and asks two questions: does it reproduce the
attainment probabilities, and does it pick the same doses?

The model is linear in dose, so one simulated cohort per (condition, weight)
cell scales to all ten dose levels. That is verified explicitly rather than
assumed.

```{r attainment-cohort}
# rxSetSeed fixes rxode2's RNG per solver thread and NOT across thread counts,
# so this cohort differs between a 16-thread workstation and a 2-core CI
# runner. Every assertion below is written to hold for any such cohort.
rxode2::rxSetSeed(20250513)
n_per_arm <- 200L   # at the 200-per-arm cap

cells <- tidyr::expand_grid(
  CONMED_PAROXETINE = c(0, 1),
  ndose             = c(1, 2),
  WT                = weights
) |>
  dplyr::mutate(
    tau       = 24 / ndose,
    condition = paste0(ifelse(ndose == 1, "Once daily", "Twice daily"), ", ",
                       ifelse(CONMED_PAROXETINE == 1, "with", "no"),
                       " paroxetine"),
    cell      = dplyr::row_number()
  )

# Reference dose of 1.0 mg/kg/day; troughs at other doses scale linearly.
ref_mg_kg_day <- 1.0

make_cohort_arm <- function(row, id_offset) {
  tau       <- row$tau
  ids       <- id_offset + seq_len(n_per_arm)
  last_dose <- run_in_h - tau
  dose <- tidyr::expand_grid(id = ids, time = seq(0, last_dose, by = tau)) |>
    dplyr::mutate(amt = ref_mg_kg_day * row$WT / (24 / tau), evid = 1, cmt = "depot")
  # A single observation, exactly one dosing interval after the last dose.
  obs <- tibble::tibble(id = ids, time = last_dose + tau,
                        amt = NA_real_, evid = 0, cmt = "central")
  dplyr::bind_rows(dose, obs) |>
    dplyr::mutate(WT = row$WT, CONMED_PAROXETINE = row$CONMED_PAROXETINE,
                  condition = row$condition, cell = row$cell) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

# One rxSolve call per condition keeps each call's subject count moderate.
troughs <- lapply(unique(cells$condition), function(cond) {
  rows <- cells[cells$condition == cond, ]
  ev <- do.call(dplyr::bind_rows, lapply(seq_len(nrow(rows)), function(i) {
    make_cohort_arm(rows[i, ], id_offset = (i - 1L) * n_per_arm)
  }))
  stopifnot(!anyDuplicated(unique(ev[, c("id", "time", "evid")])))
  rxode2::rxSolve(mod, events = ev,
                  keep = c("WT", "CONMED_PAROXETINE", "condition", "cell")) |>
    as.data.frame() |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::select(condition, cell, WT, id, ctrough_ref = Cc)
}) |>
  dplyr::bind_rows()

stopifnot(nrow(troughs) == nrow(cells) * n_per_arm,
          all(troughs$ctrough_ref > 0))
```

```{r attainment-linearity}
# Verify the linear-in-dose scaling relied on below. The ODE system is linear
# in amount, so for any fixed set of individual parameters the steady-state
# trough is exactly proportional to dose. The check is run deterministically
# (zeroRe) across all 28 cells at two dose levels: comparing two stochastic
# solves instead would confound the property under test with the fact that
# rxode2 redraws etas per call.
lin_rows <- cells |> dplyr::mutate(id = dplyr::row_number())

make_lin_ev <- function(scale) {
  do.call(dplyr::bind_rows, lapply(seq_len(nrow(lin_rows)), function(i) {
    row <- lin_rows[i, ]
    tau <- row$tau
    last_dose <- run_in_h - tau
    dplyr::bind_rows(
      tibble::tibble(id = row$id, time = seq(0, last_dose, by = tau),
                     amt = scale * ref_mg_kg_day * row$WT / (24 / tau),
                     evid = 1, cmt = "depot"),
      tibble::tibble(id = row$id, time = last_dose + tau,
                     amt = NA_real_, evid = 0, cmt = "central")
    ) |>
      dplyr::mutate(WT = row$WT, CONMED_PAROXETINE = row$CONMED_PAROXETINE)
  })) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

solve_lin <- function(scale) {
  rxode2::rxSolve(rxode2::zeroRe(mod), events = make_lin_ev(scale),
                  keep = c("WT", "CONMED_PAROXETINE")) |>
    as.data.frame() |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::pull(Cc)
}

lin_ref <- solve_lin(1.0)
lin_04  <- solve_lin(0.4)

# Compared on a RELATIVE scale: the residual difference is the ODE solver's
# own tolerance (rtol 1e-6), not a departure from linearity, so an absolute
# bound would be unreachable at concentrations of 30-90 ng/mL. Measured
# maximum here is 2.6e-6, so 1e-4 keeps the check meaningful.
stopifnot(
  length(lin_ref) == 28L, length(lin_04) == 28L,
  all(lin_ref > 0),
  max(abs(lin_04 - 0.4 * lin_ref) / (0.4 * lin_ref)) < 1e-4
)
```

```{r attainment-sweep}
# Residual error is applied here rather than inside rxSolve so that the same
# eta draw is reused across dose levels, mirroring the common-random-numbers
# design of a dose sweep. Equation (2): Bi = Ci + Ci*eps_1 + eps_2.
set.seed(20250513)
eps1 <- rnorm(nrow(troughs), 0, 0.153)
eps2 <- rnorm(nrow(troughs), 0, 1.005)

# round() matters: seq(0.1, 1.0, by = 0.1)[3] is 0.30000000000000004, which
# would silently fail the equality join against Table 4's 0.3 further down.
dose_levels <- round(seq(0.1, 1.0, by = 0.1), 1)

sweep <- lapply(dose_levels, function(d) {
  cc <- troughs$ctrough_ref * d
  cc <- cc + cc * eps1 + eps2
  troughs |>
    dplyr::select(condition, WT) |>
    dplyr::mutate(mg_per_kg_day = d, in_window = cc >= 20 & cc <= 80)
}) |>
  dplyr::bind_rows() |>
  dplyr::group_by(condition, WT, mg_per_kg_day) |>
  dplyr::summarise(pta_pct = 100 * mean(in_window), .groups = "drop")

best <- sweep |>
  dplyr::group_by(condition, WT) |>
  dplyr::slice_max(pta_pct, n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::rename(best_dose = mg_per_kg_day, best_pta = pta_pct)
```

```{r attainment-plot}
sweep |>
  dplyr::mutate(wt_label = factor(paste0(WT, " kg"),
                                  levels = paste0(weights, " kg"))) |>
  ggplot(aes(mg_per_kg_day, pta_pct, colour = wt_label)) +
  geom_hline(yintercept = 55, linetype = "dotted") +
  geom_line() +
  geom_point(size = 0.8) +
  facet_wrap(~condition) +
  labs(x = "Olanzapine dose (mg/kg/day)",
       y = "Probability of a trough within 20-80 ng/mL (%)",
       colour = "Body weight",
       title = "Simulated target attainment across the ten dose levels",
       caption = "Replicates Figure 4 of Zhang 2025; the dotted line is the 55% threshold below which Table 4 makes no once-daily recommendation.")
```

## Is the Table 4 dose the one the sweep would choose?

The attainment-versus-dose curves are flat near their maxima -- adjacent dose
levels differ by well under a percentage point in most cells -- so which dose
is literally the arg-max is decided by Monte Carlo noise at 200 subjects per
cell, not by the model. The stable and meaningful quantity is the *attainment
lost* by using the Table 4 dose instead of the best dose in the sweep, and that
is what is tabulated and gated below.

```{r attainment-dose-selection}
selection <- best |>
  dplyr::left_join(
    conditions |> dplyr::select(condition, WT, table4_dose),
    by = c("condition", "WT")
  ) |>
  dplyr::left_join(
    sweep |> dplyr::rename(table4_dose = mg_per_kg_day, pta_at_table4 = pta_pct),
    by = c("condition", "WT", "table4_dose")
  ) |>
  dplyr::mutate(
    agrees    = !is.na(table4_dose) & abs(best_dose - table4_dose) < 1e-9,
    loss_pts  = best_pta - pta_at_table4
  )

stopifnot(nrow(selection) == 28L,
          sum(!is.na(selection$table4_dose)) == 21L,
          !anyNA(selection$loss_pts[!is.na(selection$table4_dose)]))

selection |>
  dplyr::mutate(wt_label = paste0(WT, " kg")) |>
  dplyr::transmute(
    Condition                    = condition,
    Weight                       = wt_label,
    `Table 4 dose (mg/kg/day)`   = table4_dose,
    `Attainment there (%)`       = round(pta_at_table4, 1),
    `Best simulated (mg/kg/day)` = best_dose,
    `Attainment at best (%)`     = round(best_pta, 1),
    `Lost by using Table 4 (pts)` = round(loss_pts, 1)
  ) |>
  knitr::kable(caption = "Attainment at the Table 4 dose against the best dose in the sweep, per condition and weight. A blank Table 4 column marks the once-daily, no-paroxetine condition, for which the paper recommends no dose.")
```

```{r attainment-gate}
# (a) Once daily without paroxetine: Table 4 states every dose from 0.1 to
#     1.0 mg/kg/day attains at most 55.0%. Bounded on BOTH sides, because only
#     the two-sided form discriminates the omega reading: the SD reading puts
#     this maximum near 51%, the variance reading near 35%. A cohort maximum
#     over 70 cells at 200 subjects each, so the bounds carry several standard
#     errors of headroom for Monte Carlo and thread-count noise.
qd_nopar <- sweep |> dplyr::filter(condition == "Once daily, no paroxetine")
stopifnot(nrow(qd_nopar) == 70L,
          max(qd_nopar$pta_pct) > 40,
          max(qd_nopar$pta_pct) < 65)

# (b) Attainment at the Table 4 dose, against the published range for that
#     condition. Asserted on the MEDIAN across the covered weights, not on the
#     extremes -- the tail weights are the ones that move between cohorts.
published_pta <- tibble::tribble(
  ~condition,                       ~table4_dose, ~published_range, ~lo,  ~hi,
  "Twice daily, no paroxetine",     0.5,          "68.5-68.9",      68.5, 68.9,
  "Twice daily, no paroxetine",     0.4,          "68.5-72.6",      68.5, 72.6,
  "Once daily, with paroxetine",    0.5,          "57.0-59.0",      57.0, 59.0,
  "Once daily, with paroxetine",    0.4,          "58.8-62.1",      58.8, 62.1,
  "Twice daily, with paroxetine",   0.3,          "74.4-76.7",      74.4, 76.7,
  "Twice daily, with paroxetine",   0.2,          "75.1-76.9",      75.1, 76.9
)

at_table4 <- conditions |>
  dplyr::filter(!is.na(table4_dose)) |>
  dplyr::select(condition, WT, table4_dose) |>
  dplyr::left_join(sweep, by = c("condition", "WT",
                                 "table4_dose" = "mg_per_kg_day")) |>
  # Fail loudly if the dose join misses: an unmatched row would otherwise
  # propagate an NA into the attainment comparison below.
  (\(x) { stopifnot(nrow(x) == 21L, !anyNA(x$pta_pct)); x })() |>
  dplyr::group_by(condition, table4_dose) |>
  dplyr::summarise(sim_median = median(pta_pct),
                   sim_lo = min(pta_pct), sim_hi = max(pta_pct),
                   n_weights = dplyr::n(), .groups = "drop") |>
  dplyr::left_join(published_pta, by = c("condition", "table4_dose")) |>
  dplyr::mutate(mid = (lo + hi) / 2,
                delta_pts = sim_median - mid)

stopifnot(
  nrow(at_table4) == 6L,
  !any(is.na(at_table4$published_range)),
  sum(at_table4$n_weights) == 21L,
  # Centre check: the reconstruction runs a few points below the paper, but a
  # variance-scale reading of omega would put it 20-30 points below.
  max(abs(at_table4$delta_pts)) < 10
)

at_table4 |>
  dplyr::transmute(
    Condition                        = condition,
    `Table 4 dose (mg/kg/day)`       = table4_dose,
    `Weights covered`                = n_weights,
    `Simulated attainment (%)`       = sprintf("%.1f-%.1f", sim_lo, sim_hi),
    `Simulated median (%)`           = round(sim_median, 1),
    `Zhang 2025 Table 4 (%)`         = published_range,
    `Median - published midpoint`    = round(delta_pts, 1)
  ) |>
  knitr::kable(caption = "Attainment at the Table 4 doses vs the probabilities Zhang 2025 reports.")

# (c) Dose optimality. Not asserted on the arg-max, which is Monte Carlo noise
#     here: the attainment curves are flat near their maxima, so the best dose
#     flips between adjacent levels from cohort to cohort. Asserted instead on
#     the attainment LOST by using the Table 4 dose, which is stable.
#
#     Scope of this check, stated honestly: it catches gross mis-transcription
#     -- a wrong clearance, a wrong dose conversion or a missing factor of 1000
#     moves the optimum by a factor and sends this loss to tens of points --
#     but it does NOT discriminate the SD-versus-variance reading of omega,
#     because the variance reading flattens the curves further and so also
#     scores a small loss (measured: 1.3 points, versus 0.7 for the SD
#     reading). Gate (b) above is what carries that distinction.
recommending <- selection |> dplyr::filter(!is.na(table4_dose))
stopifnot(nrow(recommending) == 21L, max(recommending$loss_pts) < 12)
```

The reconstruction reproduces all four of Table 4's qualitative conclusions:
once-daily dosing without paroxetine never reaches the 55% threshold; the
twice-daily no-paroxetine and once-daily with-paroxetine conditions land in the
high 50s to low 70s; and the twice-daily with-paroxetine condition is the best
of the four, in the mid 70s. Every Table 4 dose is also at or within a couple
of percentage points of the best dose available at that weight.

What the reconstruction cannot do is confirm Table 4's exact band boundaries
(56 kg, 60 kg, 85 kg). Those boundaries sit where two adjacent dose levels
cross, and at 200 subjects per cell the crossing point is not resolvable --
the two candidate doses differ by less than the Monte Carlo error. Recovering
them would need the paper's 1000 subjects per cell, which is above this
package's per-arm simulation cap.

# Assumptions and deviations

- **`omega_CL/F` = 0.434 is read as a standard deviation, not a variance.**
  Three independent lines of evidence support this. (1) Notation: Methods
  Equation (1) defines `eta` as having "variance omega^2", and the Table 3 row
  is labelled `omega_CL/F`, not `omega^2`; the same holds for `sigma_1` and
  `sigma_2` against Equation (2)'s "variance sigma^2". (2) The paper's own
  simulation output: reading 0.434 as a variance gives an inter-individual SD
  of 0.659 on the log scale, and target attainment then falls 20-30 percentage
  points below *every* cell of Table 4, whereas the SD reading reproduces them
  to within a few points. The attainment *level* is what separates the two
  readings; the selected doses do not, because the variance reading flattens
  the attainment-versus-dose curves and so keeps Table 4's doses near-optimal
  too. (3) The same group's companion model (`Zhang_2024_olanzapine`) reaches
  the same conclusion from the same table layout. The model therefore encodes
  `etalcl ~ 0.434^2`, i.e. 45.4% CV. `propSd = 0.153` and
  `addSd = 1.005 ng/mL` are read on the same convention.
- **The paper's target-attainment criterion is not stated precisely.** Methods,
  Simulation says only "the probability of achieving the target concentration"
  against 20-80 ng/mL, while the Figure 3 caption describes the plotted
  quantity as "the predicted olanzapine trough levels". This vignette therefore
  applies the criterion to the steady-state trough. Under that reading all four
  conditions reproduce, but the reconstruction sits systematically a few
  percentage points below the published values; the gap is small, consistent in
  sign across the recommended-dose cells, and does not change any conclusion.
- **Table 4's weight-band boundaries are not reproduced, and are not gated.**
  The 56 kg, 60 kg and 85 kg boundaries mark where two adjacent dose levels
  cross in attainment. At the 200-subjects-per-cell cap used here the two
  candidate doses at those weights differ by less than the Monte Carlo error,
  so the arg-max dose flips from cohort to cohort. The vignette gates the
  attainment *lost* by using the Table 4 dose (stable, and small everywhere)
  rather than the arg-max itself.
- **Residual error is applied outside `rxSolve`.** The dose sweep reuses one
  simulated cohort per (condition, weight) cell across all ten dose levels,
  which requires the residual draws to be held fixed alongside the eta draws.
  The proportional and additive terms are drawn once and combined per Equation
  (2). The linear-in-dose scaling this relies on is verified explicitly against
  a direct solve in the `attainment-linearity` chunk.
- **Doses and sampling times are constructed, not observed.** The source
  concentrations come from a routine TDM database and the paper reports no
  administered doses, dose ranges or sampling schedule, so no observed-data
  replication is possible. Every regimen simulated here is either a Table 4
  recommendation, a dose-sweep level from Methods Simulation, or a single 10 mg
  dose chosen to exercise the NCA identities.
- **Steady state is reached by simulation, not assumed.** A 240 h run-in is
  used throughout, which is more than 22 half-lives even for the slowest arm
  (100 kg with paroxetine, t1/2 = 10.7 h).
- **`ka` and both allometric exponents are `fixed()`.** `ka = 0.861 1/h` is
  quoted by Zhang 2025 from Sun 2021 (their reference 26) and printed as
  "0.861 (fixed)" in Table 3 with no SE and no bootstrap row. The exponents
  0.75 and 1 are asserted in Methods Equation (3) from Anderson & Holford
  (2008) and appear in no results table. None of the three is an estimate of
  this study.
- **No IIV on V/F or ka, and no correlation structure.** Table 3 reports a
  single `omega` row. This is encoded faithfully as a single-eta model rather
  than adding variability the paper did not report.
- **Body weight is time-fixed.** The paper reports one weight per patient from
  a retrospective medical record system.
- **`CONMED_PAROXETINE` is a new canonical covariate column**, registered in
  `inst/references/covariate-columns.md` in the same change as this model. It
  encodes the paper's `PAR` variable with the same 0/1 orientation.
- **Eighteen exposed patients.** The `-0.289` paroxetine coefficient rests on
  18 of 72 patients (Table 2). Its bootstrap 95% interval, [-0.420, -0.078],
  still reaches close to no effect, and any downstream use should carry that
  interval alongside the point estimate.
- **Screened-but-unretained covariates.** Age, sex, and the clinical-chemistry
  panel of Table 1 are recorded in the model's `covariatesDataExcluded`
  metadata. Globulin, total protein, mean corpuscular hemoglobin and mean
  corpuscular hemoglobin concentration were also screened but have no canonical
  covariate-column entry, so they are recorded in `population$notes` instead.
  Supplementary Table 1 gives the per-drug objective-function changes: only
  paroxetine (dOFV -8.295) and amlodipine (dOFV -5.935) were significant
  univariately, and amlodipine did not survive the second forward step
  (dOFV -2.134 on top of the paroxetine model).
