---
title: "Tucatinib (Zhang 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tucatinib (Zhang 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhang_2024_tucatinib"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhang_2024_tucatinib"))$description`
* Article: <https://doi.org/10.1007/s40262-024-01412-0>
* Correction: <https://doi.org/10.1007/s40262-024-01458-0>
* Supplement (Tables S1-S4, Fig. S1, and the final-model NONMEM control stream):
  <https://doi.org/10.1007/s40262-024-01412-0> (Electronic Supplementary Material,
  MOESM1 and MOESM2)

Tucatinib is an orally administered, highly selective HER2 tyrosine kinase
inhibitor approved at 300 mg twice daily (BID). Zhang 2024 pooled seven phase
I/II studies to build a population PK model and to ask whether any clinically
relevant covariate warrants a dose adjustment. The answer was no: tumor type
was the only covariate retained, and its effect was judged not clinically
meaningful.

# Population

The analysis dataset contained 3942 quantifiable tucatinib concentrations from
283 participants across seven studies (Table S1): 151 healthy participants
(53.4%) from four phase I studies, 63 patients with HER2+ metastatic breast
cancer (mBC, 22.3%) from ONT-380-004 and ONT-380-005, and 69 patients with
HER2+ metastatic colorectal cancer (mCRC, 24.4%) from the phase II
SGNTUC-017 (MOUNTAINEER) study.

Participants had a median age of 48 years (range 18-77), a median body weight
of 75.9 kg (range 40.7-146), and 39.9% were female (Tables S2 and S3). Race was
72.4% White, 13.1% Black, 8.8% Asian, with 4.9% missing. Hepatic function was
normal in 84.5% and NCI-mild in 14.8%; moderate and severe hepatic impairment,
and moderate and severe renal impairment, were not represented in sufficient
numbers to be evaluated.

Only fasted marketed-tablet data were used from the healthy-participant
studies. HER2CLIMB (ONT-380-206) was excluded because time since the preceding
dose was not recorded.

The same information is available programmatically via the model's `population`
metadata (`rxode2::rxode(readModelDb("Zhang_2024_tucatinib"))$population`).

# Model structure

A two-compartment model with linear elimination and first-order absorption
preceded by an absorption lag time (NONMEM `ADVAN4 TRANS4`, supplementary
control stream). Tumor type acts on two parameters:

* **CL** -- 51.9% lower in mBC and 70.5% lower in mCRC than in a healthy
  participant.
* **Relative bioavailability (Frel)** -- 63.7% lower in mCRC only. The mBC
  effect on Frel was fitted but removed, because it was poorly estimated
  (RSE 50.4%) and worsened the objective function.

Because both effects act in mCRC and partly offset, the *net* apparent
clearance CL/F is only 18.7% lower in mCRC, while in mBC the full 51.9%
reduction carries through. This is the mechanism behind the paper's headline
result: a 2.1-fold AUCss increase in mBC but only a 1.2-fold increase in mCRC,
and a *lower* Cmax in mCRC than in healthy participants.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Zhang_2024_tucatinib.R` carries an in-file comment
naming its source. They are collected here for review.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lcl` (CL/F, healthy) | 112 L/h | Table 1, "CL/F (L/h), healthy participants" (4.3% RSE) |
| `lvc` (Vc/F) | 125 L | Table 1, "Vc/F (L)" (11.3% RSE) |
| `lq` (Q/F) | 89.9 L/h | Table 1, "Q/F (L/h)" (5.4% RSE) |
| `lvp` (Vp/F) | 635 L | Table 1, "Vp/F (L)" (4.0% RSE) |
| `lka` (Ka) | 0.424 1/h | Table 1, "Ka (1/h)" (5.3% RSE) |
| `ltlag` (Tlag) | 0.392 h | Table 1, "Tlag (h)" (0.56% RSE) |
| `lfdepot` (Frel, healthy) | fixed at 1 | Supplementary NONMEM control stream, `$PK`: `TF1 = 1` |
| `e_tumtp_breast_cl` | -0.519 | Table 1, "Fractional change in CL/F in participants with HER2+ mBC" (6.3% RSE) |
| `e_tumtp_crc_cl` | -0.705 | Table 1, "Fractional change in CL/F in participants with HER2+ mCRC" (3.6% RSE) |
| `e_tumtp_crc_fdepot` | -0.637 | Table 1, "Fractional change in Frel in participants with HER2+ mCRC" (3.8% RSE) |
| `etalcl` variance | 0.184403 | Table 1, "BSV CV CL/F" = 45%; footnote `CV% = sqrt(exp(var) - 1) x 100%` |
| `etalvc` variance | 1.037623 | Table 1, "BSV CV Vc/F" = 135%; same footnote |
| `etalq` variance | 0.176237 | Table 1, "BSV CV Q/F" = 43.9%; same footnote |
| `etalvp` variance | 0.163889 | Table 1, "BSV CV Vp/F" = 42.2%; same footnote |
| `etalcl`/`etalvc` covariance | 0.205590 | Table 1, "Cov (CL/F, Vc/F)" = 0.470, read as a **correlation** -- see Assumptions and deviations |
| `addSd` | 1.07 ng/mL | Table 1, "Additive residual error SD (ng/mL)" (2.14% RSE) |
| `propSd` | 0.329 | Table 1, "Proportional residual error CV" (0.73% RSE) |
| Covariate model form `TV * (1 + theta * IND)` | n/a | Methods 2.2.2 "Categorical" equation; supplementary control stream `$PK` `CLTUM` / `F1TUM` blocks |
| ODE structure (depot -> central <-> peripheral1) | n/a | Methods 2.2; supplementary control stream `$SUBROUTINE ADVAN4 TRANS4` |
| `Cc = central / vc * 1000` | n/a | Supplementary control stream `$PK`: `S2 = V2/1000 ; dose mg conc ng/mL` |

# Closed-form gates on the covariate encoding

Tucatinib disposition is linear, so at steady state the AUC over one dosing
interval has an exact closed form that the paper itself uses (Methods 2.2.5):
`AUCss = Dose * Frel / CL`. That identity is an end-to-end check on `lcl`,
`lfdepot`, and the tumor-type covariate encoding all at once, and it is
independent of the ODE solver. The asserted values below are computed from the
`ini()` parameters, then checked against numbers the paper prints in its text.

```{r closed-form}
tuc <- rxode2::rxode(readModelDb("Zhang_2024_tucatinib"))
th  <- setNames(tuc$theta, names(tuc$theta))

cl_healthy <- exp(th[["lcl"]])
frel_ref   <- exp(th[["lfdepot"]])

cl_mbc  <- cl_healthy * (1 + th[["e_tumtp_breast_cl"]])
cl_mcrc <- cl_healthy * (1 + th[["e_tumtp_crc_cl"]])
frel_mcrc <- frel_ref * (1 + th[["e_tumtp_crc_fdepot"]])

dose_mg <- 300
# AUCss in ng*h/mL: (mg / (L/h)) = mg*h/L = ug*h/mL -> x1000 = ng*h/mL
auc_ss <- c(
  Healthy = dose_mg * frel_ref  / cl_healthy,
  mBC     = dose_mg * frel_ref  / cl_mbc,
  mCRC    = dose_mg * frel_mcrc / cl_mcrc
) * 1000

# Net apparent clearance CL/F (what an NCA of these profiles would report)
clf <- dose_mg / (auc_ss / 1000)

tibble::tibble(
  Arm            = names(auc_ss),
  `CL (L/h)`     = c(cl_healthy, cl_mbc, cl_mcrc),
  `Frel`         = c(frel_ref, frel_ref, frel_mcrc),
  `CL/F (L/h)`   = as.numeric(clf),
  `AUCss (ng*h/mL)` = as.numeric(auc_ss),
  `AUCss ratio vs healthy` = as.numeric(auc_ss / auc_ss[["Healthy"]])
) |>
  knitr::kable(digits = 3, caption = "Closed-form steady-state exposure implied by the packaged ini() values.")
```

These reproduce four separate statements in the paper's text:

```{r closed-form-gates}
ratio_mbc  <- unname(auc_ss[["mBC"]]  / auc_ss[["Healthy"]])
ratio_mcrc <- unname(auc_ss[["mCRC"]] / auc_ss[["Healthy"]])
clf_drop_mcrc <- 1 - unname(clf[["mCRC"]]) / cl_healthy

# Results 3.5 / Abstract: "2.1-fold" AUCss increase in mBC (90% CI 1.9, 2.3)
stopifnot(abs(ratio_mbc - 2.1) < 0.05, ratio_mbc > 1.9, ratio_mbc < 2.3)
# Results 3.5 / Discussion: "1.2-fold" (Discussion says 1.22) in mCRC (90% CI 1.0, 1.4)
stopifnot(abs(ratio_mcrc - 1.22) < 0.02, ratio_mcrc > 1.0, ratio_mcrc < 1.4)
# Results 3.4: mCRC has "an 18.7% lower CL/F"
stopifnot(abs(clf_drop_mcrc - 0.187) < 0.001)
# Results 3.4: mBC has "51.9% lower clearance"
stopifnot(abs((1 - cl_mbc / cl_healthy) - 0.519) < 1e-9)

c(`AUCss ratio mBC (paper 2.1)`   = ratio_mbc,
  `AUCss ratio mCRC (paper 1.22)` = ratio_mcrc,
  `CL/F reduction mCRC (paper 0.187)` = clf_drop_mcrc)
```

A second, independent gate: the apparent steady-state volume. `Vss/F` is
`(Vc/F + Vp/F) / Frel`, which the model makes 760 L in healthy participants and
mBC, but `760 / 0.363 = 2094` L in mCRC. Table 2 reports median Vss/F of 752 L
(healthy), 838 L (mBC) and **2130 L** (mCRC) -- the mCRC value confirms that
Frel divides the apparent volumes exactly as encoded.

```{r vss-gate}
vss_f <- c(Healthy = (exp(th[["lvc"]]) + exp(th[["lvp"]])) / frel_ref,
           mBC     = (exp(th[["lvc"]]) + exp(th[["lvp"]])) / frel_ref,
           mCRC    = (exp(th[["lvc"]]) + exp(th[["lvp"]])) / frel_mcrc)
# Table 2 (corrected) median Vss/F for mCRC = 2130 L
stopifnot(abs(vss_f[["mCRC"]] - 2130) / 2130 < 0.05)
round(vss_f, 1)
```

# Virtual cohort

Original participant-level data are not public. The cohorts below are 200
participants per arm, dosed 300 mg BID -- the approved regimen and the regimen
Table 2 tabulates. The only covariates the final model uses are the two
tumor-type indicators, so each arm is defined entirely by them; a healthy
participant is the reference (both indicators 0).

Seven days of BID dosing is more than five effective half-lives for every arm
(the paper reports effective half-lives of 11.9 h in mBC and 16.4 h in mCRC),
so the final interval is at steady state. A washout follows the last dose so
that the terminal half-life can be estimated; the washout window is scaled to
each arm's own half-life, because a single shared window would fit the fast
(healthy) arm's terminal slope to solver noise.

```{r cohort}
set.seed(20240808)

n_per_arm <- 200L
tau       <- 12      # h, BID
n_doses   <- 14L     # 7 days
t_lastdose <- tau * (n_doses - 1L)   # 156 h
t_ss_end   <- t_lastdose + tau       # 168 h

arms <- tibble::tribble(
  ~treatment,  ~TUMTP_BREAST, ~TUMTP_CRC, ~washout_h, ~id_offset,
  "Healthy",   0,             0,          48,         0L,
  "HER2+ mBC", 1,             0,          72,         1000L,
  "HER2+ mCRC", 0,            1,          120,        2000L
)

make_cohort <- function(treatment, TUMTP_BREAST, TUMTP_CRC, washout_h, id_offset,
                        n = n_per_arm) {
  subj <- tibble::tibble(
    id           = id_offset + seq_len(n),
    treatment    = treatment,
    TUMTP_BREAST = TUMTP_BREAST,
    TUMTP_CRC    = TUMTP_CRC
  )
  obs_times <- sort(unique(c(
    seq(0, tau, by = 0.25),                              # dense first interval
    seq(tau, t_lastdose, by = tau),                      # troughs in between
    seq(t_lastdose, t_ss_end, by = 0.25),                # dense steady-state interval
    seq(t_ss_end, t_ss_end + washout_h, by = 2)          # washout, arm-specific
  )))
  doses <- subj |>
    tidyr::crossing(time = seq(0, t_lastdose, by = tau)) |>
    dplyr::mutate(amt = dose_mg, evid = 1L, cmt = "depot")
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  dplyr::bind_rows(doses, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

events <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  do.call(make_cohort, as.list(arms[i, ]))
}))

# Disjoint IDs across arms are mandatory: rxSolve treats id as the subject key,
# and duplicated ids across cohorts silently merge into one subject.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(dplyr::n_distinct(events$id) == 3L * n_per_arm)
```

# Simulation

Between-subject variability on Vc/F is 135% CV. At that magnitude the default
LSODA solver silently returns `NA` for a substantial minority of subjects, so
`dop853` is used throughout.

```{r simulate}
mod <- readModelDb("Zhang_2024_tucatinib")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("treatment", "TUMTP_BREAST", "TUMTP_CRC"),
  method = "dop853"
) |>
  as.data.frame()

# Guard against the silent all-NA / partial-NA solver failure mode.
stopifnot(mean(is.na(sim$Cc)) < 0.001)
```

`Cc` is the individual prediction (IPRED). Table 2 of the paper reports
*model-predicted* exposures from post-hoc parameters, so IPRED -- not the
residual-error-perturbed `sim` column -- is the right comparator.

A deterministic typical-value profile is simulated alongside, matching the
paper's own approach in Methods 2.2.5 ("using typical model-estimated PK
parameters without including between-subject variability"). Setting the eta
columns to zero in the event table and passing `omega = NA` is preferred over
`zeroRe()`, which mutates the shared model object.

```{r simulate-typical}
events_typ <- events |>
  dplyr::filter(id %% 1000L == 1L) |>   # one representative subject per arm
  dplyr::mutate(etalcl = 0, etalvc = 0, etalq = 0, etalvp = 0)

sim_typ <- rxode2::rxSolve(
  mod,
  events = events_typ,
  keep   = c("treatment"),
  omega  = NA,
  method = "dop853"
) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim_typ$treatment) == 3L)
```

# Replicated figures

## Steady-state concentration-time profiles

```{r figure-profiles}
sim_typ |>
  dplyr::filter(time >= t_lastdose, time <= t_ss_end) |>
  dplyr::mutate(tad = time - t_lastdose) |>
  ggplot(aes(tad, Cc, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time after dose (h)", y = "Tucatinib concentration (ng/mL)",
       colour = NULL,
       title = "Typical-value steady-state profile, 300 mg BID",
       caption = "Compare with the steady-state exposures tabulated in Table 2 of Zhang 2024.") +
  theme_bw()
```

The ordering reproduces the paper's central finding: mBC sits well above
healthy across the whole interval, whereas mCRC has a *lower* peak than healthy
participants but a *higher* trough -- the signature of reduced bioavailability
combined with reduced clearance.

## Figure 2 -- prediction-corrected VPC (cohort spread)

```{r figure-2}
sim |>
  dplyr::filter(time >= t_lastdose, time <= t_ss_end) |>
  dplyr::mutate(tad = time - t_lastdose) |>
  dplyr::group_by(treatment, tad) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  ) |>
  ggplot(aes(tad, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 0.8) +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(x = "Time after dose (h)", y = "Tucatinib concentration (ng/mL)",
       title = "Simulated steady-state spread (median, 5th-95th percentile)",
       caption = "Analogous to the prediction-corrected VPC in Figure 2 of Zhang 2024.") +
  theme_bw()
```

## Figure 3 -- clearance stratified by tumor type

Figure 3 of the paper plots model-predicted CL/F in the two patient groups and
reports geometric means of 53.4 L/h (mBC) and 89.0 L/h (mCRC).

```{r figure-3}
clf_ind <- sim |>
  dplyr::group_by(id, treatment) |>
  dplyr::summarise(clf = mean(cl / fdepot), .groups = "drop")

gm <- clf_ind |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(`Geometric mean CL/F (L/h)` = exp(mean(log(clf))), .groups = "drop")

knitr::kable(gm, digits = 1,
             caption = "Replicates Figure 3 of Zhang 2024 (published geometric means: 53.4 L/h mBC, 89.0 L/h mCRC).")

# These are sampling estimates, not exact identities: the geometric mean of a
# log-normal CL converges to its typical value (53.87 and 91.02 L/h here), but
# with n = 200 the mean of the etas still has an SD of sqrt(0.1844/200) = 0.030,
# so a few percent of drift either way is expected. The 10% band accommodates
# that while still catching a mis-encoded covariate effect.
gm_v <- setNames(gm[[2]], gm[[1]])
stopifnot(abs(gm_v[["HER2+ mBC"]]  - 53.4) / 53.4 < 0.10)
stopifnot(abs(gm_v[["HER2+ mCRC"]] - 89.0) / 89.0 < 0.10)

clf_ind |>
  dplyr::filter(treatment != "Healthy") |>
  ggplot(aes(treatment, clf)) +
  geom_boxplot(outlier.alpha = 0.3, width = 0.5) +
  scale_y_log10() +
  labs(x = NULL, y = "Model-predicted CL/F (L/h)",
       title = "Figure 3 - tucatinib clearance stratified by tumor type",
       caption = "Replicates Figure 3 of Zhang 2024.") +
  theme_bw()
```

# PKNCA validation

NCA is run on the simulated individual predictions over three intervals per
arm: the first dosing interval (for accumulation ratios), the final dosing
interval (steady state), and an arm-specific post-dose washout window (terminal
half-life).

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time-zero record per subject; pre-dose Cc = 0 for an
# extravascular dose. Without it PKNCA warns once per subject that the AUC
# range starts before the first measurement.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

blank <- function(...) {
  base <- list(cmax = FALSE, tmax = FALSE, cmin = FALSE,
               auclast = FALSE, cav = FALSE, half.life = FALSE)
  base[names(list(...))] <- list(...)
  base
}

intervals <- dplyr::bind_rows(
  # first dosing interval - denominator of the accumulation ratios
  tibble::tibble(treatment = arms$treatment, start = 0, end = tau) |>
    dplyr::bind_cols(tibble::as_tibble(blank(cmax = TRUE, auclast = TRUE))),
  # final dosing interval - steady state
  tibble::tibble(treatment = arms$treatment, start = t_lastdose, end = t_ss_end) |>
    dplyr::bind_cols(tibble::as_tibble(blank(cmax = TRUE, tmax = TRUE, cmin = TRUE,
                                             auclast = TRUE, cav = TRUE))),
  # arm-specific washout - terminal half-life
  tibble::tibble(treatment = arms$treatment, start = t_ss_end,
                 end = t_ss_end + arms$washout_h) |>
    dplyr::bind_cols(tibble::as_tibble(blank(half.life = TRUE)))
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_df  <- as.data.frame(nca_res)
```

```{r pknca-assemble}
# Interval filtering matters: cmax and auclast are requested on two different
# intervals, so pooling without filtering would silently average the first-dose
# and steady-state values.
ss   <- nca_df |> dplyr::filter(start == t_lastdose)
fst  <- nca_df |> dplyr::filter(start == 0)
term <- nca_df |> dplyr::filter(start == t_ss_end, PPTESTCD == "half.life")

# Trough concentrations are read directly at the exact end-of-interval grid
# times rather than via PKNCA. `cmin` over the FIRST interval would return the
# pre-dose zero at t = 0, not the 12 h trough, so it is not usable as the
# accumulation-ratio denominator. Reading the concentration at a grid point is
# a lookup, not an NCA calculation, so no inline trapezoidal work is involved.
troughs <- sim |>
  dplyr::filter(time %in% c(tau, t_ss_end)) |>
  dplyr::mutate(which = ifelse(time == tau, "first", "ss")) |>
  dplyr::select(treatment, id, which, Cc) |>
  tidyr::pivot_wider(names_from = which, values_from = Cc) |>
  dplyr::mutate(PPTESTCD = "ctrough")

stopifnot(nrow(troughs) == 3L * n_per_arm, !anyNA(troughs$first), !anyNA(troughs$ss))

# Accumulation ratios: steady state over first dose, per subject.
rac <- dplyr::bind_rows(
  dplyr::inner_join(
    ss  |> dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
      dplyr::select(treatment, id, PPTESTCD, ss = PPORRES),
    fst |> dplyr::filter(PPTESTCD %in% c("auclast", "cmax")) |>
      dplyr::select(treatment, id, PPTESTCD, first = PPORRES),
    by = c("treatment", "id", "PPTESTCD")
  ),
  troughs |> dplyr::select(treatment, id, PPTESTCD, ss, first)
) |>
  dplyr::mutate(rac = ss / first)

rac_summary <- rac |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(Median = median(rac, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = Median) |>
  dplyr::rename("Rac, AUC" = auclast, "Rac, Cmax" = cmax, "Rac, Ctrough" = ctrough,
                "Arm" = treatment)

knitr::kable(rac_summary, digits = 2,
             caption = "Simulated median accumulation ratios at 300 mg BID. Published (Table 2 medians): Rac AUC 1.34 / 1.94 / 2.45; Rac Cmax 1.19 / 1.59 / 1.92; Rac Ctrough 1.57 / 2.38 / 2.96 for healthy / mBC / mCRC.")
```

## Comparison against the published NCA

Table 2 of Zhang 2024 (as amended by the 2024 correction) tabulates
model-predicted steady-state exposures at 300 mg BID. The medians are
transcribed below and compared against the simulated cohort;
`ncaComparisonTable()` pools the per-subject simulated values by median, which
matches the published statistic.

```{r nca_compare}
sim_long <- dplyr::bind_rows(
  ss   |> dplyr::filter(PPTESTCD %in% c("cmax", "auclast")) |>
    dplyr::select(treatment, id, PPTESTCD, PPORRES),
  term |> dplyr::filter(PPTESTCD == "half.life") |>
    dplyr::select(treatment, id, PPTESTCD, PPORRES),
  troughs |> dplyr::transmute(treatment, id, PPTESTCD, PPORRES = ss)
)

# Table 2 medians [min, max] as published; the correction of 2024-12-11 revised
# only the mCRC Ctrough,ss geometric mean (405 -> 197), leaving the medians as
# originally printed.
published <- tibble::tribble(
  ~treatment,   ~auclast, ~cmax, ~ctrough, ~half.life,
  "Healthy",     2730,     480,   88.4,     8.76,
  "HER2+ mBC",   5770,     759,   303,      16.2,
  "HER2+ mCRC",  3230,     402,   188,      21.3
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_long,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ng/mL", ctrough = "ng/mL",
                auclast = "ng*h/mL", half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(cmp,
             caption = "Simulated vs. published model-predicted steady-state NCA (Zhang 2024 Table 2 medians). * differs by >20%.",
             align = c("l", "l", "r", "r", "r"))
attr(cmp, "footnote")
```

```{r nca_compare_check}
# Regression guard, computed from the raw values rather than by parsing the
# formatted table (the friendly labels are unicode, e.g. "t<half>", so string
# matching on them is brittle). All four parameters across all three arms hold
# inside the same 20% band that ncaComparisonTable() stars on -- the observed
# worst case is -9.2% (mBC half-life), so 20% leaves roughly 2x headroom for
# solver/platform drift without letting a mis-encoded covariate effect through.
sim_med <- sim_long |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(sim = median(PPORRES, na.rm = TRUE), .groups = "drop")

ref_long <- published |>
  tidyr::pivot_longer(-treatment, names_to = "PPTESTCD", values_to = "ref")

chk <- dplyr::inner_join(sim_med, ref_long, by = c("treatment", "PPTESTCD")) |>
  dplyr::mutate(`% diff` = (sim - ref) / ref * 100)

stopifnot(nrow(chk) == 12L)   # 3 arms x 4 parameters, all matched
stopifnot(all(abs(chk$`% diff`) < 20))

chk |> dplyr::arrange(PPTESTCD, treatment) |> knitr::kable(digits = 1)
```

# Assumptions and deviations

* **`Cov (CL/F, Vc/F) = 0.470` in Table 1 is encoded as a correlation, not a
  covariance.** Table 1's BSV block reports variability on the normalised CV%
  scale, with the footnote `BSV CV% = sqrt(exp(variance) - 1) x 100%`. Applying
  that footnote gives variances of 0.184403 (CL/F, 45%) and 1.037623 (Vc/F,
  135%). Reading 0.470 as a covariance would imply a correlation of
  `0.470 / sqrt(0.184403 * 1.037623) = 1.074`, i.e. an Omega that is not
  positive semi-definite and that `rxSolve` cannot Cholesky-decompose. The
  largest covariance the reported CVs admit is 0.4429, even at the extremes of
  their rounding intervals, so the covariance reading is arithmetically
  excluded rather than merely unlikely. A correlation of 0.470 is feasible and
  is consistent with the rest of the Table 1 BSV block being reported on the
  normalised scale. The encoded covariance is therefore
  `0.470 * sqrt(0.184403 * 1.037623) = 0.205590`. The footnote's CV formula is
  independently corroborated by the supplementary control stream's `$OMEGA`
  initial estimates (0.15 -> 40.2% and 0.99 -> 130.0%, adjacent to the final
  45% and 135%).
* **A second, uncorrected inconsistency in Table 2.** The mCRC `Vss/F` column
  gives a median of 2130 L [1680, 4690] but a geometric mean of 829 L (20.9%
  CV). A geometric mean cannot sit that far below the minimum of its own
  range. The model reproduces the *median* (2094 L, i.e. 760 L divided by the
  mCRC Frel of 0.363) and not the geometric mean, which suggests the geometric
  mean was computed without the Frel division. The published correction of
  2024-12-11 addressed only the mCRC `Ctrough,ss` geometric mean and did not
  touch this cell. No model parameter depends on it.
* **Terminal half-life.** Simulated median terminal half-lives agree with the
  published medians to within 10% in all three arms (+7.1% healthy, -9.2% mBC,
  -0.5% mCRC), so `half.life` is held to the same 20% band as the exposure
  metrics. The residual spread is expected rather than structural: Table 2's
  half-lives are computed from post-hoc individual parameters, and the 135% CV
  on Vc/F makes the distribution of individual terminal slopes wide, so the
  median of a 200-subject sample need not land exactly on the published median
  of a 52- to 128-subject one. The washout window is scaled per arm so that
  each arm's terminal slope is fitted over a comparable number of its own
  half-lives. No parameter was tuned.
* **Absorption rate.** The paper notes a median observed Tmax of 1-2 h after a
  single 300 mg dose, but the popPK Ka of 0.424 1/h (absorption half-life
  1.6 h) yields a later simulated Tmax. The authors discuss this directly: the
  popPK Ka is a pooled compromise across highly variable absorption profiles,
  and agrees with the independent PBPK estimate (Tlag 0.5 h, Ka 0.6 1/h). The
  value is used as published.
* **Covariates screened but not retained.** Body weight, sex, race, age,
  albumin, creatinine clearance, ECOG performance status and NCI hepatic
  dysfunction category were all tested and dropped during backward elimination.
  They are recorded in the model file's `covariatesDataExcluded` metadata so
  the provenance of the covariate search is preserved without declaring
  covariates that `model()` never references.
* **Virtual cohort covariates.** Because the final model uses only the two
  tumor-type indicators, no demographic distributions had to be assumed; each
  arm is fully specified by its indicator pair. Arm sizes are 200 per arm
  rather than the published 128 / 52 / 68, which changes only the sampling
  precision of the simulated medians.
* **Regimen.** All arms are simulated at 300 mg BID, the approved dose and the
  regimen Table 2 tabulates. The 50 mg and 150 mg BID arms of SGNTUC-015 and
  the 350 mg BID minority in ONT-380-004/005 are not simulated; tucatinib
  disposition is linear in the model, so those arms scale proportionally.
* **New canonical covariate column.** `TUMTP_CRC` was added to
  `inst/references/covariate-columns.md` as part of this extraction. It is a
  well-formed member of the existing `TUMTP_<type>` family and is the sister
  indicator explicitly anticipated by the `TUMTP_BREAST` entry.
