---
title: "Nirmatrelvir (Zhang 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nirmatrelvir (Zhang 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-metadata, echo = FALSE}
ui <- rxode2::rxode(readModelDb("Zhang_2024_nirmatrelvir"))
```

* Citation: `r ui$reference`
* Description: `r ui$description`
* Article: <https://doi.org/10.2147/DDDT.S479561>
* PubMed Central: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11622681/>

Zhang 2024 is a single-centre retrospective population pharmacokinetic analysis
of nirmatrelvir in Chinese inpatients treated with the fixed
nirmatrelvir/ritonavir combination (Paxlovid) for mild-to-moderate COVID-19.
The distinguishing feature of the dataset is that it is *trough-dominated*: a
sparse sampling schedule was used and, as the authors put it, "the sampling
time points mainly represent valley concentrations". A trough-only dataset
carries almost no information about the absorption rate or the volume of
distribution, so the authors fixed `Ka` at 0.8 1/h and `V/F` at 39 L to
literature values and estimated only `CL/F`, its creatinine-clearance
exponent, and the interindividual variability on `CL/F`.

A different nirmatrelvir analysis is also packaged:
`modellib("Chan_2023_nirmatrelvir")`, a two-compartment model pooled across
eight phase I and phase II/III studies. This model is distinguished by its
population -- an elderly Chinese inpatient cohort with a median age of 76
years -- and by its trough-only sampling, which is why its structural model is
one-compartment with two of its three structural parameters fixed.

# Population

The model was built from 130 plasma samples contributed by 129 patients (65
male, 64 female) enrolled at Changxing People's Hospital, Zhejiang, between
December 2022 and June 2023. All patients received nirmatrelvir 300 mg with
ritonavir 100 mg twice daily for five days. Baseline demographics (Table 1 of
the paper) describe an elderly cohort: age 73.2 +/- 14.7 years (median 76.0,
range 18.0 to 97.0) and body weight 61.2 +/- 9.3 kg (median 61.1, range 37.5
to 96.0). Renal function spans the full clinical range, with creatinine
clearance 56.7 +/- 33.1 mL/min (median 52.9, range 4.8 to 289.2), which is why
creatinine clearance was the single covariate retained after stepwise forward
inclusion and backward elimination.

The advanced age of the cohort matters for interpretation. The authors note
that their typical `CL/F` of 3.41 L/h is lower than published values from
younger populations, and that the resulting exposure at 300 mg twice daily
(median AUCtau 71.07 ug*h/mL) is roughly twice that reported for young Chinese
and Western adult volunteers (32.01 and 33.35 ug*h/mL).

The same information is available programmatically:

```{r population}
str(ui$population)
```

# Source trace

Per-parameter provenance is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Zhang_2024_nirmatrelvir.R`. The table
collects them here for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model | 1-compartment, first-order absorption and elimination (ADVAN2 TRANS2) | Results, Model Development: "a one-compartment model with first-order absorption and first-order elimination (ADVAN2 TRANS2) was found to provide the best fit" |
| Final covariate equation | `CL/F = 3.41 * (WT/70)^0.75 * (CRCL/52.9)^theta * exp(eta)` | Results, Model Development (displayed inline in the text) |
| `lcl` | `log(3.41)` L/h | Table 2, "CL/F, L/h" = 3.41 (RSE 4.00%; bootstrap 3.36, 3.16-3.92) |
| `lvc` | `fixed(log(39))` L | Table 2, "V/F, L" = 39.00 (fixed); Results: "The V/F and Ka values were set at 39 L and 0.8 h-1" |
| `lka` | `fixed(log(0.8))` 1/h | Table 2, "Ka, h-1" = 0.80 (fixed); same Results sentence |
| `e_wt_cl` | `fixed(0.75)` | Methods, Model Development: "clearance terms multiplied by (WT/70)^0.75"; Discussion restates it |
| `e_wt_vc` | `fixed(1)` | Methods, Model Development: "volume terms multiplied by (WT/70)"; Discussion restates it |
| `e_crcl_cl` | 0.429 | Table 2, "Effect of CrCl on CL/F" = 0.429 (RSE 28.00%; bootstrap 0.40, 0.21-0.81) |
| CrCl reference 52.9 mL/min | 52.9 | Results equation `(CRCL/52.9)^theta`; equals the Table 1 cohort median creatinine clearance |
| Weight reference 70 kg | 70 | Methods, Model Development ("(WT/70)"); Results equation |
| `etalcl` | 0.241792 (= `log(0.523^2 + 1)`) | Table 2, "Interindividual variability, CV%" / "CL/F, L/h" = 52.30; text: "reduced the IIV from 58.9% to 52.3%" |
| `propSd` | `fixed(0.1)` | Table 2, "Residual variability, CV%" epsilon 1 = 10.00 (fixed) |
| `addSd` | `fixed(0.1)` ug/mL | Table 2, "Residual variability, CV%" epsilon 2 = 10.00 (fixed) |
| Validation targets | Cthrough / Cmax / AUCtau by renal function and by body weight | Tables 3 and 4 |

# Virtual cohort

The original observed concentrations are not public. The paper's own Monte
Carlo simulations (Tables 3 and 4, Figure 3) are the validation target, and
they are fully specified: 1,000 virtual subjects per group, oral nirmatrelvir
300 mg or 150 mg twice daily for 5 days, with body weight fixed at 65 kg while
creatinine clearance varies (Table 3) and creatinine clearance fixed at
100 mL/min while body weight varies (Table 4).

Nine distinct arms cover both tables; the 300 mg / 65 kg / CrCl 100 mL/min arm
appears in both and is simulated once. The cohort is reduced from the paper's
1,000 per group to 200 per group, the package cap for a validation vignette.

```{r cohort}
# set.seed() seeds R's RNG, NOT rxode2's simulation RNG, and rxode2 partitions
# its streams per solver thread -- so this cohort differs between a 2-core CI
# runner and a 16-thread workstation and no seed can make them agree. Every
# assertion below is therefore written either against the DETERMINISTIC
# typical-value prediction (which carries no draw at all) or with a tolerance
# wide enough for any cohort the model can produce.
set.seed(20241206)

n_per_arm <- 200L
tau       <- 12    # dosing interval, h
n_doses   <- 10L   # 5 days b.i.d.
ss_start  <- 108   # start of the 10th (day 5) dosing interval
ss_end    <- 120

arms <- tibble::tribble(
  ~arm,                       ~dose_mg, ~WT,  ~CRCL,
  "300 mg, 65 kg, CrCl 100",  300,      65,   100,
  "300 mg, 65 kg, CrCl 70",   300,      65,   70,
  "300 mg, 65 kg, CrCl 45",   300,      65,   45,
  "300 mg, 65 kg, CrCl 15",   300,      65,   15,
  "150 mg, 65 kg, CrCl 45",   150,      65,   45,
  "150 mg, 65 kg, CrCl 15",   150,      65,   15,
  "300 mg, 115 kg, CrCl 100", 300,      115,  100,
  "300 mg, 90 kg, CrCl 100",  300,      90,   100,
  "300 mg, 40 kg, CrCl 100",  300,      40,   100
)

# Dense over the first and the last (steady-state) dosing interval, where
# Cmax, Cthrough and AUCtau are read; coarse in between, where nothing is
# measured. A 0.25 h grid resolves Tmax (about 2.4 h) to well under 1%.
obs_times <- sort(unique(c(
  seq(0, tau, by = 0.25),
  seq(tau, ss_start, by = 2),
  seq(ss_start, ss_end, by = 0.25)
)))

# id_offset keeps subject IDs disjoint across arms. rxSolve uses id as the
# subject key, so colliding ids silently merge arms into single subjects that
# receive the summed dose.
make_arm <- function(arm_row, n, id_offset) {
  subj <- tibble(
    id      = id_offset + seq_len(n),
    arm     = arm_row$arm,
    dose_mg = arm_row$dose_mg,
    WT      = arm_row$WT,
    CRCL    = arm_row$CRCL
  )
  dosing <- subj |>
    tidyr::crossing(time = seq(0, by = tau, length.out = n_doses)) |>
    mutate(amt = dose_mg, evid = 1L, cmt = "depot")
  # cmt on observation rows is the ODE STATE, never the algebraic observable
  # "Cc" -- naming the observable auto-injects a compartment slot and
  # renumbers the ODE states.
  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central")
  bind_rows(dosing, obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, amt, evid, cmt, arm, dose_mg, WT, CRCL)
}

events <- do.call(
  bind_rows,
  lapply(seq_len(nrow(arms)), function(i) {
    make_arm(arms[i, ], n_per_arm, id_offset = (i - 1L) * n_per_arm)
  })
)

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(nrow(dplyr::distinct(events, id)) == n_per_arm * nrow(arms))
```

# Simulation

```{r simulate}
mod <- readModelDb("Zhang_2024_nirmatrelvir")

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep   = c("arm", "dose_mg", "WT", "CRCL")
) |>
  as.data.frame()

stopifnot(all(is.finite(sim$Cc)), all(sim$Cc >= 0))
```

The typical-value (population) prediction is obtained by zeroing the random
effects. Because the model's only random effect is a log-normal eta on `CL/F`,
and steady-state `AUCtau = Dose / CL`, the *median* of the simulated cohort and
the typical-value prediction coincide in expectation -- which makes the
typical-value profile the noise-free comparator for the paper's published
medians.

```{r simulate-typical}
mod_typical <- rxode2::zeroRe(ui)

events_typical <- do.call(
  bind_rows,
  lapply(seq_len(nrow(arms)), function(i) make_arm(arms[i, ], 1L, id_offset = i))
)

sim_typical <- rxode2::rxSolve(
  mod_typical,
  events = events_typical,
  keep   = c("arm", "dose_mg", "WT", "CRCL"),
  omega  = NA
) |>
  as.data.frame()
```

# Replicate published figures

```{r figure-3, fig.width = 7, fig.height = 4.5}
# Replicates Figure 3 of Zhang 2024: simulated nirmatrelvir concentration-time
# curves for 65 kg patients across renal function and dose. The paper plots
# the median (dark blue) with a 95% prediction interval (light blue shading).
fig3_arms <- arms$arm[arms$WT == 65]

sim |>
  filter(arm %in% fig3_arms) |>
  mutate(
    dose_label = paste0(dose_mg, " mg b.i.d."),
    crcl_label = factor(
      paste0("CrCl ", CRCL, " mL/min"),
      levels = paste0("CrCl ", c(100, 70, 45, 15), " mL/min")
    )
  ) |>
  group_by(dose_label, crcl_label, time) |>
  summarise(
    Q025 = quantile(Cc, 0.025),
    Q50  = quantile(Cc, 0.50),
    Q975 = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "navy") +
  facet_grid(dose_label ~ crcl_label) +
  labs(
    x = "Time (h)", y = "Nirmatrelvir concentration (ug/mL)",
    title = "Figure 3 - simulated profiles by renal function and dose",
    caption = "Replicates Figure 3 of Zhang 2024 (65 kg; median with 95% prediction interval)."
  )
```

```{r figure-weight, fig.width = 7, fig.height = 3}
# Companion to Table 4 of Zhang 2024: the effect of body weight at CrCl
# 100 mL/min on 300 mg b.i.d. exposure. The paper tabulates this rather than
# plotting it.
wt_arms <- arms$arm[arms$CRCL == 100]

sim |>
  filter(arm %in% wt_arms) |>
  mutate(wt_label = factor(paste0(WT, " kg"), levels = paste0(c(115, 90, 65, 40), " kg"))) |>
  group_by(wt_label, time) |>
  summarise(
    Q025 = quantile(Cc, 0.025),
    Q50  = quantile(Cc, 0.50),
    Q975 = quantile(Cc, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q025, ymax = Q975), fill = "steelblue", alpha = 0.3) +
  geom_line(colour = "navy") +
  facet_wrap(~wt_label, nrow = 1) +
  labs(
    x = "Time (h)", y = "Nirmatrelvir concentration (ug/mL)",
    title = "Table 4 - simulated profiles by body weight (CrCl 100 mL/min, 300 mg b.i.d.)",
    caption = "Companion to Table 4 of Zhang 2024 (median with 95% prediction interval)."
  )
```

# Internal identity check

Before comparing against the paper, confirm the packaged model is internally
consistent. For a linear one-compartment model at steady state the area under
the curve over one dosing interval is exactly `Dose / CL`, independent of `Ka`
and `V`. This check compares the trapezoidal AUC of the typical-value solve
over the day-5 interval against `Dose / CL` computed from the model equation.
Both sides use the same fixed parameters, so the only difference is
trapezoidal error -- a deterministic quantity that does not vary between
machines, and the bound is therefore tight.

```{r identity}
trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

identity_chk <- sim_typical |>
  filter(time >= ss_start, time <= ss_end) |>
  group_by(arm, dose_mg, WT, CRCL) |>
  summarise(auc_solved = trapz(time, Cc), .groups = "drop") |>
  mutate(
    # The model's own equation: CL/F = 3.41 * (WT/70)^0.75 * (CRCL/52.9)^0.429
    cl_closed  = 3.41 * (WT / 70)^0.75 * (CRCL / 52.9)^0.429,
    auc_closed = dose_mg / cl_closed,
    pct_diff   = 100 * (auc_solved - auc_closed) / auc_closed
  )

identity_chk |>
  select(arm, auc_closed, auc_solved, pct_diff) |>
  rename(
    "Arm"                      = arm,
    "Dose/CL (ug*h/mL)"        = auc_closed,
    "Solved AUCtau (ug*h/mL)"  = auc_solved,
    "% diff"                   = pct_diff
  ) |>
  knitr::kable(digits = 3, caption = "Steady-state AUCtau against the closed form Dose/CL.")

# Deterministic: trapezoidal error only, identical on every machine.
stopifnot(max(abs(identity_chk$pct_diff)) < 0.5)
```

# PKNCA validation

Non-compartmental parameters are computed with PKNCA over the day-5 dosing
interval (108 to 120 h), which is the window the paper's Tables 3 and 4
describe ("AUCtau, area under the concentration versus time curve through the
dosing interval (day 5); Cmax, maximum observed concentration (day 5)").

```{r pknca}
# Only `!is.na(Cc)` -- adding `time > 0` or `Cc > 0` would drop the time-zero
# record PKNCA uses to anchor the profile.
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, arm)

# Guarantee a time-zero record per subject; pre-dose Cc = 0 is correct for an
# extravascular first dose.
sim_nca <- bind_rows(
  sim_nca,
  sim_nca |> distinct(id, arm) |> mutate(time = 0, Cc = 0)
) |>
  distinct(id, arm, time, .keep_all = TRUE) |>
  arrange(id, time)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  filter(evid == 1L) |>
  select(id, time, amt, arm)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start   = ss_start,
  end     = ss_end,
  cmax    = TRUE,
  auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_wide <- as.data.frame(nca_res) |>
  select(arm, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Guard against a silently empty result (a gate that cannot go red is worse
# than no gate).
stopifnot(
  nrow(nca_wide) == n_per_arm * nrow(arms),
  all(is.finite(nca_wide$cmax)),
  all(is.finite(nca_wide$auclast))
)
```

Trough concentrations are read directly from the simulation at 12 h (the first
dosing interval) and 120 h (the tenth). PKNCA's `ctrough` anchors on a dose
record and is not defined for a bare `[108, 120]` window, and this profile
decreases monotonically to the end of the interval (`Ka` 0.8 1/h is well above
`kel`, which is at most 0.15 1/h across these arms), so the end-of-interval
value is the trough.

```{r troughs}
troughs <- sim |>
  filter(time %in% c(tau, ss_end)) |>
  mutate(dose_number = ifelse(time == tau, "1st (day 1)", "10th (day 5)")) |>
  group_by(arm, dose_number) |>
  summarise(ctrough = median(Cc), .groups = "drop")

troughs_typical <- sim_typical |>
  filter(time %in% c(tau, ss_end)) |>
  mutate(dose_number = ifelse(time == tau, "1st (day 1)", "10th (day 5)")) |>
  select(arm, dose_number, ctrough = Cc)
```

## Comparison against published NCA

Zhang 2024 Tables 3 and 4 report the median (95% CI) of each quantity over
1,000 simulated subjects. Both tables are transcribed below; the 300 mg /
65 kg / CrCl 100 mL/min arm is shared between them and appears once.

```{r published}
published <- tibble::tribble(
  ~arm,                       ~cmax,  ~auclast, ~ctrough_d1, ~ctrough_d5,
  "300 mg, 65 kg, CrCl 100",   8.31,   71.07,    2.36,        3.10,
  "300 mg, 65 kg, CrCl 70",    9.47,   85.21,    2.82,        3.99,
  "300 mg, 65 kg, CrCl 45",   10.92,  102.98,    3.39,        5.33,
  "300 mg, 65 kg, CrCl 15",   16.00,  164.30,    4.74,       10.10,
  "150 mg, 65 kg, CrCl 45",    5.46,   51.48,    1.70,        2.66,
  "150 mg, 65 kg, CrCl 15",    8.00,   82.11,    2.37,        5.05,
  "300 mg, 115 kg, CrCl 100",  5.14,   45.61,    1.58,        2.22,
  "300 mg, 90 kg, CrCl 100",   6.29,   54.82,    1.88,        2.57,
  "300 mg, 40 kg, CrCl 100",  12.30,  100.34,    3.27,        4.08
)
```

The simulated cohort is compared first. `ncaComparisonTable()` aggregates the
per-subject PKNCA output to the group median, which is the statistic the paper
reports.

```{r nca-compare}
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_wide |> select(arm, cmax, auclast),
  reference     = published |> select(arm, cmax, auclast),
  by            = "arm",
  units         = c(cmax = "ug/mL", auclast = "ug*h/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  digits  = 2,
  caption = "Simulated (200/arm) vs. published (1,000/arm) medians. * differs by >20%.",
  align   = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

The deterministic comparison follows. Because the only random effect is a
log-normal eta on `CL/F`, the model's population median equals its
typical-value prediction exactly, so the typical-value profile can be compared
to the published medians with no simulation noise at all.

```{r typical-compare}
typical_metrics <- sim_typical |>
  filter(time >= ss_start, time <= ss_end) |>
  group_by(arm) |>
  summarise(cmax = max(Cc), auclast = trapz(time, Cc), .groups = "drop") |>
  left_join(
    troughs_typical |>
      tidyr::pivot_wider(names_from = dose_number, values_from = ctrough) |>
      rename(ctrough_d1 = `1st (day 1)`, ctrough_d5 = `10th (day 5)`),
    by = "arm"
  )

typical_cmp <- typical_metrics |>
  tidyr::pivot_longer(-arm, names_to = "metric", values_to = "typical") |>
  left_join(
    published |> tidyr::pivot_longer(-arm, names_to = "metric", values_to = "published"),
    by = c("arm", "metric")
  ) |>
  mutate(
    metric = recode(
      metric,
      cmax        = "Cmax day 5 (ug/mL)",
      auclast     = "AUCtau day 5 (ug*h/mL)",
      ctrough_d1  = "Cthrough after 1st dose (ug/mL)",
      ctrough_d5  = "Cthrough after 10th dose (ug/mL)"
    ),
    pct_diff = 100 * (typical - published) / published
  )

typical_cmp |>
  arrange(metric, arm) |>
  rename(
    "Arm"               = arm,
    "Quantity"          = metric,
    "Model (typical)"   = typical,
    "Zhang 2024 median" = published,
    "% diff"            = pct_diff
  ) |>
  knitr::kable(
    digits  = 2,
    caption = "Deterministic typical-value prediction against the published medians.",
    align   = c("l", "l", "r", "r", "r")
  )
```

```{r typical-gate}
# Both sides of this comparison are fixed numbers: the model side carries no
# random draw (the etas are zeroed) and the paper side is transcribed, so the
# bound does NOT have to absorb cohort noise and is set tight. All 36 values
# agree to within 3.25% -- the residual is the paper's own Monte Carlo
# sampling error on 1,000 subjects (its AUCtau column runs about 3% high for
# the renal arms and about 1% low for the weight arms). A mis-transcribed
# clearance, exponent, reference value, dose or unit moves these quantities by
# tens of percent, so 5% fails instantly on any of them.
stopifnot(
  nrow(typical_cmp) == 4L * nrow(arms),
  !anyNA(typical_cmp$pct_diff),
  max(abs(typical_cmp$pct_diff)) < 5,
  abs(median(typical_cmp$pct_diff)) < 2
)

# The cohort medians are a draw and vary with the solver thread count; hold
# them to a wider, quantile-based bound rather than an extreme.
# (ncaComparisonTable() returns display-formatted columns, so recompute the
# percentage differences from the numeric PKNCA output.)
cohort_pct <- nca_wide |>
  group_by(arm) |>
  summarise(cmax = median(cmax), auclast = median(auclast), .groups = "drop") |>
  tidyr::pivot_longer(-arm, names_to = "metric", values_to = "simulated") |>
  left_join(
    published |>
      select(arm, cmax, auclast) |>
      tidyr::pivot_longer(-arm, names_to = "metric", values_to = "published"),
    by = c("arm", "metric")
  ) |>
  mutate(pct_diff = 100 * (simulated - published) / published)

stopifnot(
  nrow(cohort_pct) == 2L * nrow(arms),
  !anyNA(cohort_pct$pct_diff),
  abs(median(cohort_pct$pct_diff)) < 10,
  quantile(abs(cohort_pct$pct_diff), 0.9) < 20
)
```

## Renal impairment and the paper's dosing conclusion

The paper's clinical argument is that 150 mg twice daily in severe renal
impairment gives exposure comparable to 300 mg twice daily in normal renal
function -- the basis of its dose-reduction recommendation. That is a
statement about two of the simulated arms, and it can be checked directly.

```{r renal-conclusion}
ratio_tbl <- typical_metrics |>
  filter(arm %in% c("150 mg, 65 kg, CrCl 15", "300 mg, 65 kg, CrCl 100")) |>
  select(arm, cmax, auclast)

ratio_auc  <- ratio_tbl$auclast[ratio_tbl$arm == "150 mg, 65 kg, CrCl 15"] /
  ratio_tbl$auclast[ratio_tbl$arm == "300 mg, 65 kg, CrCl 100"]
ratio_cmax <- ratio_tbl$cmax[ratio_tbl$arm == "150 mg, 65 kg, CrCl 15"] /
  ratio_tbl$cmax[ratio_tbl$arm == "300 mg, 65 kg, CrCl 100"]

knitr::kable(
  data.frame(
    Quantity = c("AUCtau ratio", "Cmax ratio"),
    Ratio    = c(ratio_auc, ratio_cmax)
  ),
  digits  = 3,
  caption = paste(
    "150 mg b.i.d. at CrCl 15 mL/min relative to 300 mg b.i.d. at CrCl",
    "100 mL/min (65 kg, typical value). Zhang 2024 Discussion: 'drug exposure",
    "with 150/100 mg N/R in patients with CrCl of 15 mL/min is comparable to",
    "that in patients with normal renal function taking 300/100 mg N/R'."
  )
)

# Deterministic (typical values, no draw). Realised 1.13 for AUCtau and 0.94
# for Cmax; "comparable" is gated at within 35% either way, which the paper's
# own claim supports and which a halved or doubled dose would break.
stopifnot(ratio_auc > 0.75, ratio_auc < 1.35, ratio_cmax > 0.75, ratio_cmax < 1.35)
```

The Discussion also states that at 300 mg twice daily the moderate and severe
renal-impairment groups have a median Cmax above 10 ug/mL, and quotes AUCtau
increases of 44% and 131% relative to normal renal function.

```{r renal-increases}
auc_by_crcl <- typical_metrics |>
  filter(arm %in% arms$arm[arms$dose_mg == 300 & arms$WT == 65]) |>
  mutate(CRCL = arms$CRCL[match(arm, arms$arm)]) |>
  arrange(desc(CRCL))

auc_ref <- auc_by_crcl$auclast[auc_by_crcl$CRCL == 100]
auc_by_crcl <- auc_by_crcl |>
  mutate(pct_increase = 100 * (auclast / auc_ref - 1))

auc_by_crcl |>
  select(CRCL, cmax, auclast, pct_increase) |>
  rename(
    "CrCl (mL/min)"              = CRCL,
    "Cmax day 5 (ug/mL)"         = cmax,
    "AUCtau day 5 (ug*h/mL)"     = auclast,
    "% increase vs CrCl 100"     = pct_increase
  ) |>
  knitr::kable(digits = 1, caption = "Renal-function gradient at 300 mg b.i.d., 65 kg (typical value).")

# Zhang 2024 Discussion: 44% and 131% at CrCl 45 and 15. Deterministic on this
# side (typical values); realised 40.9% and 125.2%, so the tolerance below
# absorbs only the paper's Monte Carlo medians. A wrong CrCl exponent or
# reference value changes the gradient itself and breaks both.
inc45 <- auc_by_crcl$pct_increase[auc_by_crcl$CRCL == 45]
inc15 <- auc_by_crcl$pct_increase[auc_by_crcl$CRCL == 15]
stopifnot(
  abs(inc45 - 44) < 6,
  abs(inc15 - 131) < 9,
  # "median Cmax exceeded 10 ug/mL" for moderate and severe impairment.
  all(auc_by_crcl$cmax[auc_by_crcl$CRCL %in% c(45, 15)] > 10)
)
```

# Assumptions and deviations

- **The additive residual error is read as 0.1 ug/mL.** Table 2 lists both
  `epsilon 1` and `epsilon 2` as `10.00 (fixed)` under a single heading
  "Residual variability, CV%", and the Results text says only that the residual
  model was "additive plus constant coefficient of variation ... with values
  set at 10%". A CV cannot describe an additive term, so the table's shared
  header does not literally apply to `epsilon 2`. The reading used here is the
  one that makes the NONMEM combined error `Y = F + F*eps1 + eps2` consistent
  with a `$SIGMA 0.01 FIX 0.01 FIX` block: both standard deviations are 0.1,
  giving a 10% proportional term and an additive term of 0.1 ug/mL. That value
  is also exactly the assay's lower limit of quantification for nirmatrelvir
  (Methods, Laboratory Methods: "The linear ranges of nirmatrelvir and
  ritonavir were 0.1-10.0 ug/mL"), which is where an additive residual term is
  normally anchored. Both magnitudes were fixed by the authors, so neither is
  an estimate and the choice does not affect the typical-value or median
  predictions validated above; it affects only the width of a simulated
  residual-error band.
- **Interindividual variability was converted from CV% to a log-normal
  variance.** Table 2 reports IIV on `CL/F` as 52.30 under the heading
  "Interindividual variability, CV%", restated in the text as "reduced the IIV
  from 58.9% to 52.3%". The standard log-normal conversion
  `omega^2 = log(CV^2 + 1) = log(0.523^2 + 1) = 0.241792` is used. The
  alternative convention `omega^2 = CV^2 = 0.273529` would widen the
  prediction interval slightly; it does not move the median, and every
  assertion above is on a median or a typical value.
- **Creatinine clearance is entered in raw mL/min, not BSA-normalized.**
  Table 1 reports "Creatinine Clearance, mL/min" and the covariate reference
  52.9 mL/min is that table's median, so the model's `CRCL` column is the raw
  variant of the canonical column. The paper never states which equation
  estimated it -- no Cockcroft-Gault, MDRD or CKD-EPI attribution appears in
  the Methods, Results or table footnotes. Supplying a BSA-normalized value
  would rescale the renal term and, because the model also carries an
  allometric weight term, double-count body size.
- **Concentration units.** Doses are in mg and volumes in L, so the model's
  natural concentration unit is mg/L, which is numerically identical to the
  paper's ug/mL. No scaling is applied or needed.
- **`Ka` and `V/F` are fixed, not estimated.** The authors fixed them at
  0.8 1/h and 39 L from prior literature because the trough-dominated sparse
  sampling could not support estimating them (Results, Model Development;
  restated as the study's first limitation). Both values are printed in the
  paper's own Table 2, so nothing is carried in from an unavailable upstream
  source. The practical consequence is that this model should not be used to
  predict absorption-phase concentrations or Tmax outside the conditions the
  fixed values were borrowed for.
- **The cohort is 200 per arm, not the paper's 1,000.** 200 is the package cap
  for a validation vignette. The deterministic typical-value comparison, which
  is the primary gate, is unaffected by cohort size.
- **The specimen is plasma.** The Abstract says "serum concentrations" but the
  Methods describe plasma throughout ("100 uL plasma sample"; "the separated
  plasma was frozen at -80 C"; "PopPK of nirmatrelvir in plasma
  concentration-time data"). The Methods matrix is recorded in
  `compartmentData`.
- **Ritonavir is not modelled.** Nirmatrelvir was always given with ritonavir
  100 mg twice daily, and ritonavir concentrations were assayed, but the paper
  develops and reports a nirmatrelvir model only -- Table 2 has no ritonavir
  column and no separate ritonavir objective function is reported. The
  ritonavir boost is therefore folded into the apparent clearance rather than
  represented explicitly, and the model is valid only for the boosted
  combination.
- **The residual disagreement is systematic by table, and is the paper's, not
  the model's.** Every published value is reproduced, but the sign of the
  small residual tracks which table it came from: the four renal-function arms
  of Table 3 sit about 3.0 to 3.3% above the model's `Dose / CL`, while the
  three body-weight arms unique to Table 4 sit about 1.1 to 1.5% below it, and
  the arm shared by both tables agrees to 0.5%. For a linear model whose only
  random effect is a log-normal eta on clearance, the median steady-state
  AUCtau *is* `Dose / CL` by construction, so a per-table offset can only come
  from the paper's Monte Carlo run (1,000 subjects gives a median with roughly
  2% standard error at this IIV). It is recorded here rather than absorbed by
  widening a gate.
- **No covariates other than creatinine clearance and body weight.** Age,
  height, sex, white blood cell / neutrophil / lymphocyte counts, C-reactive
  protein, procalcitonin, alkaline phosphatase, glutamyl transpeptidase,
  alanine aminotransferase and aspartate aminotransferase were all screened
  and none was retained (Methods, Model Development; Results, Model
  Development). Body weight is in the model a priori rather than by selection.
