---
title: "Realgar-Indigo Naturalis Formula (Zhang 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Realgar-Indigo Naturalis Formula (Zhang 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
suppressPackageStartupMessages({
  library(nlmixr2lib)
  library(PKNCA)
  library(rxode2)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
})
```

# Model and source

* Citation: `r rxode2::rxode2(readModelDb("Zhang_2021_realgar_indigo"))$reference`
* Description: `r rxode2::rxode2(readModelDb("Zhang_2021_realgar_indigo"))$description`
* Article: <https://doi.org/10.2147/DDDT.S305244>

The oral tetra-arsenic tetra-sulfide (As4S4) formula studied here is the
Realgar-Indigo Naturalis Formula (RIF). Doses are expressed as mg of the
formula and the measured analyte is total plasma arsenic (ug/L), so the
clearance and volume are apparent values relative to the formula dose. A
literature check on 2026-09-28 (EuropePMC) found no erratum or correction
for this article.

# Population

Twelve Chinese children with acute promyelocytic leukemia (11 newly diagnosed,
1 relapsed) were studied at a single centre in Tianjin between July 2016 and
July 2019 (Zhang 2021 Table 1; ChiCTR-OIC-16010014). Median age was 8 years
(range 4-14) and median body weight 27 kg (range 16-63; mean 30.03, SD 14.06).
All patients received RIF at 60 mg/kg/day divided three times daily (median
540 mg per administration, range 320-1350 mg) together with ATRA, either as
maintenance (regimen A, n = 7) or consolidation (regimen B, n = 5) therapy.
Sex distribution was not reported. The 107 arsenic concentrations modelled
ranged from 0.1 to 75.0 ug/L.

# Source trace

| Element | Value | Source |
|---|---|---|
| Structure | one compartment, first-order absorption and elimination | Results, 'Population Pharmacokinetic Analysis' |
| `lka` | log(0.013) 1/h | Table 2, theta1 |
| `lcl` | log(1380) L/h at 27 kg | Table 2, theta2 |
| `lvc` | log(7080) L | Table 2, theta3 |
| `e_wt_cl` | 0.629 on (WT/27) | Table 2, theta4 and 'F WT-CL = (CW/27)^theta4' |
| `etalka` | 0.353^2 | Table 2, IIV Ka |
| `etalcl` | 0.167^2 | Table 2, IIV CL |
| `etalvc` | 0.787^2 | Table 2, IIV V |
| `addSd` | 3.619 ug/L | Table 2, residual variability (scale: see below) |
| Covariate screen | WT, age, creatinine, ALB, AST, ALT; only WT retained | Methods and Results |
| Cc = central / vc x 1000 | mg/L to ug/L | units of Table 1 dose and reported concentrations |

# Model

```{r model}
mod <- readModelDb("Zhang_2021_realgar_indigo")
ui <- rxode2::rxode2(mod)
ui
```

# Virtual cohort

Individual weights were not published. The cohort below draws weights from
a log-normal distribution matched to the Table 1 mean (30.03 kg) and SD
(14.06 kg), rejecting and redrawing any value outside the observed 16-63 kg
range. The same 200 body weights are used in every dose arm; the random
effects are drawn independently per arm (the dose-proportionality check below
holds them fixed across arms instead).

```{r cohort}
set.seed(8071704)
rxode2::rxSetSeed(8071704)
n_sub <- 200
cv_wt <- 14.06 / 30.03
sdlog_wt <- sqrt(log(1 + cv_wt^2))
meanlog_wt <- log(30.03) - sdlog_wt^2 / 2
draw_wt <- function(n) {
  out <- numeric(0)
  while (length(out) < n) {
    x <- stats::rlnorm(n, meanlog_wt, sdlog_wt)
    out <- c(out, x[x >= 16 & x <= 63])
  }
  out[seq_len(n)]
}
cohort <- data.frame(id = seq_len(n_sub), WT = draw_wt(n_sub))
summary(cohort$WT)
```

# Day-1 profile after the first administration

On day 1 samples were drawn over 28 h after a single administration, and the
Results report plasma arsenic of 0.17-5.6 ug/L 24 h after the dose. With
ka = 0.013 1/h (absorption half-life `r round(log(2) / 0.013)` h) against an
elimination rate constant of 1380/7080 = 0.195 1/h, the kinetics are
flip-flop: the terminal decline reflects absorption, not elimination.

```{r day1}
obs_times <- c(0, 1, 2, 4, 6, 7, 8, 9, 10, 16, 20, 24, 28)
ev_day1 <- bind_rows(
  cohort |> mutate(time = 0, amt = 20 * WT, evid = 1L, cmt = "depot"),
  tidyr::crossing(cohort, time = obs_times) |>
    mutate(amt = 0, evid = 0L, cmt = "central")
) |>
  arrange(id, time, desc(evid))
sim_day1 <- rxode2::rxSolve(mod, events = ev_day1, returnType = "data.frame")

typ_day1 <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = ev_day1 |> filter(id == 1) |> mutate(WT = 27, amt = ifelse(evid == 1, 540, 0)),
  returnType = "data.frame"
)
c24_typ <- typ_day1$Cc[typ_day1$time == 24]
c24_typ

sim_day1 |>
  group_by(time) |>
  summarise(
    p05 = quantile(Cc, 0.05), p50 = median(Cc), p95 = quantile(Cc, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, p50)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.3) +
  geom_line() +
  geom_line(data = typ_day1, aes(time, Cc), linetype = 2) +
  annotate("rect", xmin = 23.5, xmax = 24.5, ymin = 0.17, ymax = 5.6, alpha = 0.2, fill = "red") +
  labs(
    x = "Time after first administration (h)", y = "Plasma arsenic (ug/L)",
    title = "Day 1, 20 mg/kg single administration",
    caption = "Median and 5th-95th percentile of 200 simulated children; dashed = typical 27 kg child. Red band: reported 24 h range 0.17-5.6 ug/L."
  )

# The typical 27 kg child sits inside the reported 24 h range. Three
# administrations in the first 24 h would put it at about 12 ug/L, outside
# the range, which supports reading the day-1 samples as a single dose.
stopifnot(c24_typ > 0.17, c24_typ < 5.6)
```

# Steady state at 30, 45 and 60 mg/kg/day

The Monte Carlo simulation in the Results reports median steady-state trough
concentrations of 25.38, 38.17 and 50.66 ug/L for RIF 30, 45 and 60 mg/kg/day
given three times daily (10, 15 and 20 mg/kg per administration). Steady
state is imposed with `ss = 1`, so the absorption half-life of about two days
does not require a long loading period.

```{r steady-state}
arms <- data.frame(
  treatment = c("30 mg/kg/d", "45 mg/kg/d", "60 mg/kg/d"),
  mg_per_kg = c(10, 15, 20)
)
tau <- 8
ss_times <- seq(0, tau, by = 0.5)
ev_ss <- bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  cohort_i <- cohort |> mutate(id = id + (i - 1) * n_sub, treatment = arms$treatment[i])
  bind_rows(
    cohort_i |> mutate(
      time = 0, amt = arms$mg_per_kg[i] * WT, ii = tau, ss = 1L,
      evid = 1L, cmt = "depot"
    ),
    tidyr::crossing(cohort_i, time = ss_times) |>
      mutate(amt = 0, ii = 0, ss = 0L, evid = 0L, cmt = "central")
  )
})) |>
  arrange(id, time, desc(evid))

sim_ss <- rxode2::rxSolve(
  mod,
  events = ev_ss, returnType = "data.frame",
  rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12,
  keep = "treatment"
)
trough <- sim_ss |>
  filter(time == 0) |>
  group_by(treatment) |>
  summarise(median_cmin = median(Cc), .groups = "drop") |>
  mutate(published = c(25.38, 38.17, 50.66), pct_diff = 100 * (median_cmin / published - 1))
knitr::kable(
  trough |>
    rename(
      "Regimen" = treatment, "Simulated median Css,min (ug/L)" = median_cmin,
      "Published median (ug/L)" = published, "% diff" = pct_diff
    ),
  digits = 2,
  caption = "Median steady-state trough, simulated vs Zhang 2021 Results."
)

# The medians carry the cohort weight distribution, which was not published;
# a mis-transcribed CL/F, dose or unit would move them by far more than 10%.
stopifnot(all(abs(trough$pct_diff) < 10))
```

The published medians are exactly proportional to dose
(25.38 / 50.66 = `r round(25.38 / 50.66, 4)`, 38.17 / 50.66 =
`r round(38.17 / 50.66, 4)`), as the linear model requires when the same
virtual children receive each dose. The same holds for the model with
etas held fixed across arms:

```{r proportionality}
# Fixed-eta solve: one set of etas as data columns, typical-value model.
set.seed(20210427)
om <- ui$omega
# The omega matrix is diagonal, so independent normal draws suffice.
eta_draw <- sapply(colnames(om), function(nm) stats::rnorm(n_sub, 0, sqrt(om[nm, nm])))
ev_fix <- ev_ss |>
  mutate(base_id = (id - 1) %% n_sub + 1) |>
  left_join(data.frame(base_id = seq_len(n_sub), eta_draw), by = "base_id") |>
  select(-base_id)
sim_fix <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  events = ev_fix, returnType = "data.frame",
  rtol = 1e-10, atol = 1e-12, ssRtol = 1e-10, ssAtol = 1e-12,
  keep = "treatment"
)
med_fix <- sim_fix |>
  filter(time == 0) |>
  group_by(treatment) |>
  summarise(m = median(Cc), .groups = "drop")
# The supplied etas must actually reach the solve: with them, log(CL/F)
# varies beyond what body weight explains.
cl_resid <- sim_fix |>
  filter(time == 0) |>
  mutate(r = log(cl) - log(1380) - 0.629 * log(WT / 27)) |>
  pull(r)
stopifnot(sd(cl_resid) > 0.1)
ratio_30 <- med_fix$m[1] / med_fix$m[3]
ratio_45 <- med_fix$m[2] / med_fix$m[3]
c(ratio_30 = ratio_30, ratio_45 = ratio_45)
stopifnot(abs(ratio_30 - 0.5) < 1e-6, abs(ratio_45 - 0.75) < 1e-6)
```

# PKNCA over the steady-state dosing interval

```{r pknca}
conc_ss <- sim_ss |>
  filter(!is.na(Cc)) |>
  select(id, time, Cc, treatment)
dose_ss <- ev_ss |>
  filter(evid == 1) |>
  select(id, time, amt, treatment)
conc_obj <- PKNCA::PKNCAconc(conc_ss, Cc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose_ss, amt ~ time | treatment + id)
intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)
nca_res <- suppressMessages(
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
)
nca_long <- as.data.frame(nca_res)
knitr::kable(summary(nca_res), caption = "Steady-state NCA over one 8 h interval (Cc in ug/L, AUC in ug*h/L).")

sim_summary <- nca_long |>
  filter(PPTESTCD %in% c("cmin", "auclast")) |>
  group_by(treatment, PPTESTCD) |>
  summarise(PPORRES = median(PPORRES), .groups = "drop")
reference <- data.frame(
  treatment = arms$treatment,
  cmin = c(25.38, 38.17, 50.66)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  sim_summary, reference,
  by = "treatment", params = "cmin",
  units = c(cmin = "ug/L")
)
knitr::kable(cmp, caption = "Median steady-state Cmin: published Monte Carlo vs simulated.")
```

## AUC and individual clearance

The Results give the individual (post hoc) median and range of
weight-normalised CL/F as 45.26 (35.63-82.18) L/h/kg and state that the
steady-state AUC0-24 "ranged from 0.24 to 0.56 mg*h/L". A daily AUC at
60 mg/kg/day would be 60 / CL, i.e. 0.73-1.68 mg*h/L over that clearance
range, so the quoted range cannot be a 24 h AUC. It is reproduced exactly as
the AUC over one 8 h interval, 20 mg/kg / CL:

```{r auc-interval}
round(c(auc_low = 20 / 82.18, auc_high = 20 / 35.63), 3)
stopifnot(abs(20 / 82.18 - 0.24) < 0.005, abs(20 / 35.63 - 0.56) < 0.005)

auc60 <- nca_long |>
  filter(treatment == "60 mg/kg/d", PPTESTCD == "auclast") |>
  pull(PPORRES) / 1000
quantile(auc60, c(0.05, 0.5, 0.95))
# Median simulated interval AUC sits inside the published individual range.
stopifnot(median(auc60) > 0.24, median(auc60) < 0.56)

indiv <- sim_ss |>
  filter(treatment == "60 mg/kg/d", time == 0) |>
  transmute(cl_per_kg = cl / WT, v_per_kg = vc / WT)
data.frame(
  quantity = c("CL/F (L/h/kg)", "V/F (L/kg)"),
  typical_27kg = c(1380 / 27, 7080 / 27),
  simulated_median = c(median(indiv$cl_per_kg), median(indiv$v_per_kg)),
  published_median = c(45.26, 230.37),
  published_range = c("35.63-82.18", "85.96-495.68")
) |>
  knitr::kable(digits = 2, caption = "Weight-normalised apparent CL and V. Published values are post hoc estimates in 12 children.")
```

The simulated cohort medians land close to the published post hoc medians.
The published values are empirical-Bayes estimates for only 12 children, and
a median of 200 simulated children with a V/F IIV SD of 0.787 moves by several
percent between seeds, so this table is shown for context and not asserted;
the steady-state trough medians above are the quantitative check.

# Residual error scale

Table 2 prints 'Residual variability (%)' = 3.619 and the Results text says a
proportional model was chosen. The paper's own Figure 1B (observed vs
individual predicted concentrations) contradicts a 3.6% proportional error.
The points below were digitised by the maintainers from Figure 1B (single,
well-separated points only; the overlapping cluster below 5 ug/L is omitted).

```{r residual}
fig1b <- data.frame(
  dv = c(
    4.40, 10.92, 15.44, 8.21, 4.03, 11.53, 13.45, 14.29, 18.06, 22.50, 30.48,
    24.80, 28.11, 26.17, 36.46, 30.02, 32.64, 39.52, 39.17, 31.19, 40.69, 47.40,
    38.70, 43.19, 34.27, 40.96, 44.43, 53.84, 49.74, 55.36, 44.05, 50.48, 52.36,
    60.95, 56.88, 50.83, 52.96, 56.81, 61.43, 50.50, 62.91, 57.26, 64.80, 64.80,
    75.00, 71.67
  ),
  ipred = c(
    1.38, 4.68, 5.53, 6.52, 8.82, 10.38, 15.89, 17.87, 21.13, 21.60, 24.40,
    26.81, 26.83, 30.33, 30.50, 33.78, 34.30, 36.31, 39.01, 40.43, 41.39, 41.39,
    42.03, 42.78, 43.38, 43.70, 45.53, 46.10, 46.55, 46.57, 49.93, 52.20, 53.17,
    53.62, 54.49, 54.61, 55.77, 56.43, 56.45, 58.42, 61.43, 61.70, 65.39, 68.23,
    68.23, 68.65
  )
)
resid_bins <- fig1b |>
  mutate(ipred_bin = cut(ipred, c(0, 20, 40, 80))) |>
  group_by(ipred_bin) |>
  summarise(
    n = n(),
    abs_sd = sqrt(mean((dv - ipred)^2)),
    rel_sd = sqrt(mean((dv / ipred - 1)^2)),
    .groups = "drop"
  )
knitr::kable(resid_bins, digits = 3, caption = "Root-mean-square DV - IPRED residual by IPRED bin, digitised from Figure 1B.")

# Relative spread is roughly 10-fold larger at low concentrations while the
# absolute spread stays within a factor of 1.5: an additive, not a
# proportional, error. Individual-fit residuals are also shrunk toward zero,
# so their spread should not exceed sigma by much; 3-5 ug/L is consistent
# with an additive SD of 3.619 ug/L and not with a 3.6% proportional SD or
# with an additive variance of 3.619 (SD 1.90 ug/L).
stopifnot(
  resid_bins$rel_sd[1] / resid_bins$rel_sd[3] > 5,
  max(resid_bins$abs_sd) / min(resid_bins$abs_sd) < 2,
  min(resid_bins$abs_sd) > 3.619 * 0.75
)
```

# Assumptions and deviations

* **Residual error encoded as additive, 3.619 ug/L.** The paper's text and
  column header say proportional (%), but its Figure 1B shows residuals of
  roughly constant absolute size (above), which a 3.6% proportional error
  cannot produce. Users who prefer the printed description can replace
  `Cc ~ add(addSd)` with `Cc ~ prop(propSd)` and `propSd = 0.03619`.
* **IIV read as the SD of eta.** Table 2 lists the IIV rows as 0.353, 0.167
  and 0.787 under a '(%)' header. They are squared to give the eta variances.
  A variance reading is excluded because the CL row's printed RSE (39.3%) and
  the bootstrap intervals imply relative standard errors below the floor of
  sqrt(2/12) = 40.8% for a variance estimated from 12 subjects. The
  lognormal-CV reading cannot be excluded; it would give variances of 0.117,
  0.0275 and 0.486. The Results text says IIV was 'best described by an
  additive model', but Table 2 prints exponential equations
  (Ka = theta1 x EXP(eta1), etc.) and those equations are followed.
* **Dose is formula mass.** The model is dosed in mg of RIF formula, not mg
  of arsenic. This is what the reported CL/F and troughs imply: 540 mg every
  8 h / 1380 L/h = 49 ug/L, matching the reported trough medians.
* **AUC0-24 in the Results is the AUC over one 8 h interval** (see above).
* **Weight distribution** of the virtual cohort is an assumption (log-normal
  matched to the Table 1 mean and SD, restricted to 16-63 kg); individual
  weights were not published.
* Table 1 prints the ALT range as '9.0-4.4', an evident typo; ALT was not
  retained in the model.
