---
title: "Tebipenem (Zhang 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tebipenem (Zhang 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhang_2026_tebipenem"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhang_2026_tebipenem"))$description`
* Article: <https://doi.org/10.1111/cts.70453>
* Supporting Information (Appendix S1, two DOCX files inside a ZIP; holds the
  correction derivations, Tables S1-S2 and the authors' full `mrgsolve` /
  simulation R scripts): <https://doi.org/10.1111/cts.70453>
* Upstream population PK model, from which every parameter estimate is taken:
  Sato N et al., *Drug Metab Pharmacokinet* 2008;23(6):434-446,
  <https://doi.org/10.2133/dmpk.23.434>

Zhang 2026 did not estimate a population PK model. It adopted the Sato 2008
pediatric tebipenem model, re-expressed so that the dose supplied to the model
is tebipenem rather than the pro-drug tebipenem pivoxil, and used it to simulate
exposure in Bangladeshi children aged 24-59 months with shigellosis. Everything
in `ini()` is therefore `fixed()`.

# Population

The **parameter-estimation population** is Sato 2008's: 217 Japanese children
(112 boys, 105 girls) pooled from five clinical studies of oral tebipenem
pivoxil for otolaryngological infection or pneumonia, median (range) age 4.50
(0.67-15.23) years, weight 15.4 (7.06-49.5) kg, serum creatinine 0.30
(0.16-0.70) mg/dL and weight-normalized creatinine clearance 3.73 (1.46-6.97)
mL/min/kg (Sato 2008 Table 2 and Results). 395 plasma tebipenem concentrations
were fitted in NONMEM V with `ADVAN2` / `TRANS2` (one-compartment model with
first-order absorption), a relative (proportional) inter-subject error model and
an additive residual error model.

The **simulation target population** is different: 2249 Bangladeshi children
aged 24-59 months with suspected *Shigella* infection, screened between May and
September 2022 for trial NCT05121974 (Zhang 2026 Table 1). 59.5% were male;
girls had mean (SD) age 36.27 (9.66) months, height 91.68 (7.68) cm and weight
11.89 (2.64) kg, boys 35.58 (9.38) months, 92.66 (7.33) cm and 12.42 (2.58) kg.
This cohort weighs less than the WHO weight-for-age standard at every age
(Zhang 2026 Figure 2), which is why the authors constructed a bespoke virtual
population instead of a generic pediatric one.

Zhang 2026 compared the Sato 2008 model against the 15 Bangladeshi children in
the trial's pilot study and reported a median prediction error of 21.6% and a
median absolute prediction error of 40.9% (Zhang 2026 Methods 2.4, Figure S1);
the model was judged adequate and no re-estimation was attempted.

The same information is available programmatically via
`readModelDb("Zhang_2026_tebipenem")()$population`.

# Source trace

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment model, first-order absorption, absorption lag | structure | Sato 2008 Discussion ("we selected the one-compartment model with first-order absorption as a pharmacokinetic model"); NONMEM `ADVAN2` / `TRANS2`, Sato 2008 Methods |
| `lka` = log(5.85) | ka = 5.85 1/h | Sato 2008 Table 4; Table 5 theta1 = 5.85 +/- 1.02; Zhang 2026 Appendix S1 `mrgsolve` `$PARAM TVKA = 5.85` |
| `lcl` = log(0.363) | 0.363 L/h/kg | Sato 2008 Table 4 `CL/F (L/hr/kg) = 0.363 + 0.104 x Ccr`; Table 5 theta2 = 0.363 +/- 0.138; Zhang 2026 Eq. (6) |
| `e_crcl_cl` = 0.104 | 0.104 (L/h/kg)/(mL/min/kg) | Sato 2008 Table 4 (same equation); Table 5 theta3 = 0.104 +/- 0.0354; Zhang 2026 Eq. (6) |
| `lvc` = log(1.18) | 1.18 L/kg at age 1 y | Sato 2008 Table 4 `Vd/F (L/kg) = 1.18 x Age^-0.132`; Table 5 theta5 = 1.18 +/- 0.141; Zhang 2026 Eq. (7) |
| `e_age_vc` = -0.132 | exponent | Sato 2008 Table 4 (exponent printed in the equation); Table 5 theta6 = 0.132 +/- 0.0885; Zhang 2026 Eq. (7) |
| `ltlag` = log(0.239) | Tlag = 0.239 h | Sato 2008 Table 4; Table 5 theta8 = 0.239 +/- 0.0377; Zhang 2026 Appendix S1 `$PARAM TVALAG1 = 0.239` |
| `fu` = 0.58 | unbound fraction | Zhang 2026 Methods 2.5 (mean unbound plasma fraction 0.58, citing Rodvold 2022); Appendix S1 `freeTBPM <- IPRED * 0.58` |
| `etalcl` ~ 0.0407 | omega^2(CL/F) | Sato 2008 Table 5 (0.0407 +/- 0.0183); Table 4 omega(CL/F) = 20.2%; Appendix S1 `$OMEGA` element 1 |
| `etalvc` ~ 0.5558 | omega^2(Vd/F) | Zhang 2026 Appendix S1 `$OMEGA` element 2 = 0.5558; Sato 2008 Table 5 reports 0.558 +/- 0.0968, Table 4 omega(Vd/F) = 74.7% |
| `etalka` ~ 1.28 | omega^2(ka) | Sato 2008 Table 5 (1.28 +/- 0.981); Table 4 omega(ka) = 113%; Appendix S1 `$OMEGA` element 3 |
| `etaltlag` ~ 2.1 | omega^2(Tlag) | Sato 2008 Table 5 (2.10 +/- 1.28); Table 4 omega(Tlag) = 145%; Appendix S1 `$OMEGA` element 4 |
| `addSd` = 0.453 | sigma = 0.453 ug/mL | Sato 2008 Table 4 `sigma (ug/mL) 0.453`; Table 5 sigma^2 = 0.205 +/- 0.0584 (sqrt = 0.4528) |
| MW ratio 383.5 / 497.63 = 0.7707 on CL/F and V/F | correction | Zhang 2026 Supporting Information Appendix S1 (`CTS-2025-0490-s01.docx`), "Tebipenem molecular weight: 383.5 g/mol; Tebipenem-pivoxil molecular weight: 497.63 g/mol"; Zhang 2026 Eqs. (6) and (7) |
| Creatinine-clearance unit conversion (1000 / 60) | correction | Zhang 2026 Supporting Information Appendix S1, first paragraph; Zhang 2026 Eq. (6). Not needed here because `CRCL` is stored in Sato's native mL/min/kg |
| CrCl distribution N(3.72, 0.97) mL/min/kg | cohort input | Sato 2008 Table 2; Zhang 2026 Appendix S1 virtual-population script, "Ccr distribution from Sato et al. ... Mean (mL/min/kg) = 3.72, SD = 0.97" |
| 4 mg/kg TID, 9 doses q8h; 3 mg/kg QID, 12 doses q6h; dose rounded UP to a multiple of 5 mg | regimens | Zhang 2026 Methods 2.4 and 2.7; Appendix S1 `amt <- ceiling((4*WT)*2/10)/2*10` then `amt * 0.7707` |
| *Shigella* species 53% *flexneri* / 47% *sonnei* | PD input | Zhang 2026 Methods 2.5 (average of rural and urban Bangladeshi isolates in 2020, ref. 6) |
| MIC distributions | PD input | Zhang 2026 Methods 2.5: *S. flexneri* 0.016 / 0.032 / 0.063 ug/mL at 12.5% / 75.0% / 12.5%; *S. sonnei* 0.016 / 0.032 / 0.063 / 0.127 ug/mL at 1.35% / 66.2% / 21.6% / 10.8% (Fernandez Alvaro 2022, Vietnamese isolates) |
| Targets 40% fT > MIC and fAUC0-24/MIC/tau > 34.58 | PD targets | Zhang 2026 Methods 2.5 and 2.8 |
| Height-from-age and (Box-Cox) weight-from-height regressions | cohort input | Zhang 2026 Table S1 (Appendix S1) |

# Virtual cohort

The screening data set is not public, so the cohort below is reconstructed from
the published summary statistics: the sex-specific age / height / weight means
and SDs of Zhang 2026 Table 1 and the sex-specific linear regressions of Zhang
2026 Table S1. See "Assumptions and deviations" for exactly where this departs
from the authors' construction.

200 virtual participants are simulated (the per-arm cap for these vignettes);
Zhang 2026 simulated 500 trials of 66 participants, i.e. 33,000. The **same 200
participants and the same random effects** are used for both dosing regimens, as
in the paper.

```{r cohort}
set.seed(20260819)
n_sub <- 200L

# Zhang 2026 Table 1, screening data set (n = 2249), by sex.
tab1 <- tibble::tribble(
  ~sex,     ~age_mean, ~age_sd, ~ht_mean, ~ht_sd, ~wt_mean, ~wt_sd,
  "female", 36.27,     9.66,    91.68,    7.68,   11.89,    2.64,
  "male",   35.58,     9.38,    92.66,    7.33,   12.42,    2.58
)
# Zhang 2026 Table S1 (Appendix S1): sex-specific linear regressions.
tabS1 <- tibble::tribble(
  ~sex,     ~ht_int, ~ht_slope, ~ht_r2, ~wt_r2,
  "female", 68.8,    0.632,     0.629,  0.663,
  "male",   70.7,    0.616,     0.621,  0.649
)
demographics <- dplyr::left_join(tab1, tabS1, by = "sex")

age_lo <- 24  # months; trial inclusion criterion (Zhang 2026 Methods 2.1)
age_hi <- 59

# Age: beta distribution on [24, 59] months, moment-matched to the Table 1
# sex-specific mean and SD. Reproduces both moments exactly on a bounded
# support without needing the (unpublished) empirical age histogram.
draw_age <- function(n, mean_m, sd_m) {
  m <- (mean_m - age_lo) / (age_hi - age_lo)
  v <- (sd_m / (age_hi - age_lo))^2
  k <- m * (1 - m) / v - 1
  stopifnot(k > 0)
  age_lo + (age_hi - age_lo) * stats::rbeta(n, m * k, (1 - m) * k)
}

# Weight-normalized creatinine clearance: Sato 2008 Table 2 mean 3.72, SD 0.97
# mL/min/kg, truncated to the Sato 2008 observed range so no draw is
# non-physiological.
draw_crcl <- function(n) {
  x <- stats::rnorm(n, 3.72, 0.97)
  while (any(bad <- x < 1.46 | x > 6.97)) {
    x[bad] <- stats::rnorm(sum(bad), 3.72, 0.97)
  }
  x
}

mw_ratio <- 383.5 / 497.63  # tebipenem / tebipenem-pivoxil (Appendix S1)

cohort <-
  tibble::tibble(
    id  = seq_len(n_sub),
    # Zhang 2026 Methods 2.3: assigned male with probability 0.6.
    sex = ifelse(stats::runif(n_sub) < 0.6, "male", "female")
  ) |>
  dplyr::left_join(demographics, by = "sex") |>
  dplyr::mutate(
    age_months = draw_age(dplyr::n(), age_mean, age_sd),
    AGE        = age_months / 12,
    # Zhang 2026 Eqs. (1)-(2): height is linear in age with an additive
    # residual. The residual SD is not published, but it is fixed by the
    # published SD of height and the published R-squared:
    # SD_resid = SD_height * sqrt(1 - R^2).
    HT = ht_int + ht_slope * age_months +
      stats::rnorm(dplyr::n(), 0, ht_sd * sqrt(1 - ht_r2))
  ) |>
  dplyr::group_by(sex) |>
  dplyr::mutate(
    # Weight is generated conditionally on height with the published
    # height-weight correlation (rho = sqrt(R^2)) and a log-normal marginal
    # moment-matched to the Table 1 mean and SD.
    z_ht    = (HT - mean(HT)) / stats::sd(HT),
    sig_log = sqrt(log(1 + (wt_sd / wt_mean)^2)),
    mu_log  = log(wt_mean) - sig_log^2 / 2,
    rho     = sqrt(wt_r2),
    WT      = exp(mu_log + sig_log *
                    (rho * z_ht + sqrt(1 - rho^2) * stats::rnorm(dplyr::n())))
  ) |>
  dplyr::ungroup() |>
  dplyr::mutate(
    CRCL = draw_crcl(dplyr::n()),
    # Zhang 2026 Methods 2.4 / 2.7: mg/kg dose of tebipenem PIVOXIL rounded UP
    # to the nearest multiple of 5 mg, then converted to a tebipenem dose.
    amt_tid = ceiling(0.8 * WT) * 5 * mw_ratio,  # 4 mg/kg
    amt_qid = ceiling(0.6 * WT) * 5 * mw_ratio   # 3 mg/kg
  ) |>
  dplyr::select(id, sex, age_months, AGE, HT, WT, CRCL, amt_tid, amt_qid)

cohort |>
  dplyr::group_by(sex) |>
  dplyr::summarise(
    n = dplyr::n(),
    `Age (m)` = mean(age_months), `Age SD` = stats::sd(age_months),
    `Height (cm)` = mean(HT),     `Height SD` = stats::sd(HT),
    `Weight (kg)` = mean(WT),     `Weight SD` = stats::sd(WT),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2,
               caption = paste("Simulated cohort vs Zhang 2026 Table 1",
                               "(female 36.27 / 9.66 months, 91.68 / 7.68 cm,",
                               "11.89 / 2.64 kg; male 35.58 / 9.38, 92.66 / 7.33,",
                               "12.42 / 2.58)."))
```

```{r figure-1, fig.cap = "Weight versus age in the simulated cohort. Replicates the simulated (orange) points of Figure 1 of Zhang 2026; the observed screening data are not public. Dashed lines mark the Zhang 2026 Table 1 sex-specific mean weight."}
cohort |>
  ggplot(aes(age_months, WT)) +
  geom_point(alpha = 0.6, shape = 1) +
  geom_hline(
    data = tab1, aes(yintercept = wt_mean), linetype = "dashed", colour = "blue"
  ) +
  facet_wrap(~sex) +
  labs(x = "Age (months)", y = "Weight (kg)",
       caption = "Replicates the simulated arm of Figure 1 of Zhang 2026.")
```

# Simulation

Two regimens over the 3-day (72 h) treatment period, both driven by the same
cohort and the same random effects:

* 4 mg/kg tebipenem pivoxil three times daily (q8h), 9 doses - the protocol
  regimen of NCT05121974.
* 3 mg/kg tebipenem pivoxil four times daily (q6h), 12 doses - the alternative
  regimen of Zhang 2026 Methods 2.7, same total daily dose.

```{r simulate}
mod <- readModelDb("Zhang_2026_tebipenem")

# Zhang 2026 Appendix S1 simulates on a 0.1 h grid so fT > MIC can be counted
# by time point; the same grid is used here.
obs_times <- seq(0, 72, by = 0.1)

make_events <- function(coh, amt_col, ii, addl) {
  dosing <- coh |>
    dplyr::transmute(
      id, WT, AGE, CRCL,
      time = 0, amt = .data[[amt_col]], evid = 1L,
      cmt = "depot", ii = ii, addl = addl
    )
  obs <- coh |>
    dplyr::select(id, WT, AGE, CRCL) |>
    tidyr::crossing(time = obs_times) |>
    dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central",
                  ii = 0, addl = 0L)
  dplyr::bind_rows(dosing, obs) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

ev_tid <- make_events(cohort, "amt_tid", ii = 8, addl = 8L)
ev_qid <- make_events(cohort, "amt_qid", ii = 6, addl = 11L)

# Common random numbers: reseeding identically before each solve gives both
# regimens the same per-subject etas, as in Zhang 2026 ("the same 33,000
# virtual participants ... Shigella species and MIC value assignment to each
# virtual participant were unchanged").
solve_arm <- function(ev, label) {
  set.seed(101)
  rxode2::rxSetSeed(101)
  rxode2::rxSolve(mod, events = ev, returnType = "data.frame") |>
    dplyr::mutate(treatment = label)
}

sim_tid <- solve_arm(ev_tid, "4 mg/kg TID")
sim_qid <- solve_arm(ev_qid, "3 mg/kg QID")

# Gate 1: interindividual variability really was sampled (see the rxode2
# omega-reuse trap: a silently-dropped omega collapses every subject onto the
# typical value and still renders cleanly).
cl_tid <- sim_tid |> dplyr::distinct(id, cl) |> dplyr::arrange(id)
cl_qid <- sim_qid |> dplyr::distinct(id, cl) |> dplyr::arrange(id)
stopifnot(nrow(cl_tid) == n_sub, stats::sd(cl_tid$cl) > 0)

# Gate 2: the two arms really did share random effects.
stopifnot(isTRUE(all.equal(cl_tid$cl, cl_qid$cl)))

sim <- dplyr::bind_rows(sim_tid, sim_qid)
```

```{r figure-conc, fig.cap = "Simulated total plasma tebipenem over the 3-day treatment period. Line = median, ribbon = 5th-95th percentile. The horizontal line is the most frequent Shigella MIC (0.032 ug/mL) divided by the unbound fraction 0.58, i.e. the total concentration at which the free concentration equals that MIC."}
sim |>
  dplyr::filter(!is.na(Cc), Cc > 0) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = 0.032 / 0.58, linetype = "dashed", colour = "red") +
  facet_wrap(~treatment, ncol = 1) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total plasma tebipenem (ug/mL)")
```

# PK/PD target attainment

Each virtual participant is assigned a *Shigella* species (53% *flexneri*, 47%
*sonnei*) and, within species, an MIC drawn from the Vietnamese-isolate
distribution reported by Zhang 2026 Methods 2.5. The MIC assignment is shared
between regimens.

```{r pd-targets}
set.seed(555)
strain <- ifelse(stats::runif(n_sub) < 0.53, "flexneri", "sonnei")
mic <- vapply(
  strain,
  function(s) {
    if (identical(s, "flexneri")) {
      sample(c(0.016, 0.032, 0.063), 1L, prob = c(0.125, 0.750, 0.125))
    } else {
      sample(c(0.016, 0.032, 0.063, 0.127), 1L,
             prob = c(0.013514, 0.662162, 0.216216, 0.108108))
    }
  },
  numeric(1)
)
mic_tbl <- tibble::tibble(id = seq_len(n_sub), strain = strain, MIC = mic)

trapz <- function(x, y) sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)

pd <- sim |>
  dplyr::filter(!is.na(Ccfree)) |>
  dplyr::left_join(mic_tbl, by = "id") |>
  dplyr::mutate(tau = ifelse(treatment == "4 mg/kg TID", 8, 6)) |>
  dplyr::group_by(treatment, id, strain, MIC, tau) |>
  dplyr::summarise(
    # Zhang 2026 Methods 2.5: fraction of the 72 h period with free tebipenem
    # above the MIC, counted on the simulation grid.
    fT_over_MIC = mean(Ccfree >= MIC),
    # Zhang 2026 Methods 2.8: free AUC over the first 24 h.
    fAUC24      = trapz(time[time <= 24], Ccfree[time <= 24]),
    .groups     = "drop"
  ) |>
  dplyr::mutate(fAUC_MIC_tau = fAUC24 / MIC / tau)

pd_summary <-
  dplyr::bind_rows(
    pd |>
      dplyr::group_by(treatment) |>
      dplyr::summarise(target = "40% fT > MIC",
                       simulated = mean(fT_over_MIC > 0.4), .groups = "drop"),
    pd |>
      dplyr::group_by(treatment) |>
      dplyr::summarise(target = "fAUC0-24/MIC/tau > 34.58",
                       simulated = mean(fAUC_MIC_tau > 34.58), .groups = "drop")
  ) |>
  # Zhang 2026 Results 3.2, 3.4 and 3.5: median probability across 500 virtual
  # trials, with the 95% confidence interval.
  dplyr::left_join(
    tibble::tribble(
      ~treatment,      ~target,                     ~published, ~lo,    ~hi,
      "4 mg/kg TID",   "40% fT > MIC",              0.864,      0.773,  0.939,
      "3 mg/kg QID",   "40% fT > MIC",              0.924,      0.864,  0.985,
      "4 mg/kg TID",   "fAUC0-24/MIC/tau > 34.58",  0.545,      0.439,  0.652,
      "3 mg/kg QID",   "fAUC0-24/MIC/tau > 34.58",  0.773,      0.636,  0.864
    ),
    by = c("treatment", "target")
  ) |>
  dplyr::mutate(within_published_ci = simulated >= lo & simulated <= hi)

# A gate that cannot go red is worse than none: confirm every published row was
# actually matched before asserting on it.
stopifnot(nrow(pd_summary) == 4L, !anyNA(pd_summary$published))
stopifnot(all(pd_summary$within_published_ci))

pd_summary |>
  dplyr::transmute(
    `PK/PD target` = target,
    Regimen        = treatment,
    `Simulated (%)` = round(100 * simulated, 1),
    `Zhang 2026 median (%)` = round(100 * published, 1),
    `Zhang 2026 95% CI (%)` =
      sprintf("%.1f - %.1f", 100 * lo, 100 * hi),
    `Within published CI` = within_published_ci
  ) |>
  knitr::kable(
    caption = paste(
      "Proportion of virtual participants achieving each PK/PD target,",
      "against the median and 95% confidence interval reported across",
      "Zhang 2026's 500 virtual trials (Results 3.2, 3.4 and 3.5)."
    )
  )
```

```{r figure-3, fig.cap = "Per-participant fraction of the 72 h treatment period with free tebipenem above the assigned Shigella MIC. The dashed vertical line is the 40% target; the annotation gives the proportion above it, to be compared with the medians in Figure 3 of Zhang 2026 (86.4% for TID, 92.4% for QID)."}
pd |>
  ggplot(aes(fT_over_MIC)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = 0.4, linetype = "dashed", colour = "red") +
  facet_wrap(~treatment, ncol = 1) +
  labs(x = "fT > MIC (fraction of the 72 h treatment period)",
       y = "Virtual participants",
       caption = "Corresponds to Figure 3 of Zhang 2026.")
```

```{r figure-5, fig.cap = "Per-participant fAUC0-24/MIC/tau. The dashed vertical line is the logarithmic-killing target of 34.58 used in Zhang 2026 Methods 2.8; compare the proportion above it with the medians in Figure 5 of Zhang 2026 (54.5% for TID, 77.3% for QID)."}
pd |>
  ggplot(aes(fAUC_MIC_tau)) +
  geom_histogram(bins = 40) +
  geom_vline(xintercept = 34.58, linetype = "dashed", colour = "red") +
  facet_wrap(~treatment, ncol = 1) +
  scale_x_log10() +
  labs(x = "fAUC0-24 / MIC / tau", y = "Virtual participants",
       caption = "Corresponds to Figure 5 of Zhang 2026.")
```

# Cross-check against the upstream Sato 2008 cohort

Sato 2008 Table 6 reports empirical-Bayes secondary NCA parameters for its own
217-patient cohort at 4 mg/kg tebipenem pivoxil twice daily, across the whole
0.5-16 year age range: Cmax 3.48 +/- 1.67 ug/mL, Tmax 0.74 +/- 0.26 h, AUC0-24h
11.00 +/- 1.84 ug*h/mL and t1/2 1.05 +/- 0.68 h. Because this file's parameters
come from Sato 2008, a typical-value simulation at Sato's median covariates is
an independent check that the corrections were transcribed correctly.

```{r simulate-sato-typical}
# Sato 2008 Table 2 medians.
sato_age  <- 4.50   # years
sato_wt   <- 15.4   # kg
sato_crcl <- 3.73   # mL/min/kg
sato_amt  <- 4 * sato_wt * mw_ratio  # 4 mg/kg tebipenem pivoxil -> tebipenem

sato_times <- seq(0, 24, by = 0.02)
ev_sato <- dplyr::bind_rows(
  tibble::tibble(id = 1L, WT = sato_wt, AGE = sato_age, CRCL = sato_crcl,
                 time = 0, amt = sato_amt, evid = 1L, cmt = "depot",
                 ii = 12, addl = 1L),
  tibble::tibble(id = 1L, WT = sato_wt, AGE = sato_age, CRCL = sato_crcl,
                 time = sato_times, amt = NA_real_, evid = 0L,
                 cmt = "central", ii = 0, addl = 0L)
) |>
  dplyr::arrange(time, dplyr::desc(evid))

# omega = NA is required, not merely zeroRe(): rxode2 otherwise reuses the
# previous solve's omega and silently re-samples etas.
sim_sato <- rxode2::rxSolve(
  rxode2::zeroRe(mod), events = ev_sato, omega = NA, returnType = "data.frame"
)
if (is.null(sim_sato$id)) sim_sato$id <- 1L
sim_sato$treatment <- "Sato 4 mg/kg BID (typical)"

# Gate: the typical-value CL and V must equal the closed-form published
# equations, which also proves the eta draws really were suppressed.
cl_closed <- mw_ratio * (0.363 + 0.104 * sato_crcl) * sato_wt
vc_closed <- mw_ratio * 1.18 * sato_age^(-0.132) * sato_wt
stopifnot(
  isTRUE(all.equal(unique(sim_sato$cl), cl_closed, tolerance = 1e-8)),
  isTRUE(all.equal(unique(sim_sato$vc), vc_closed, tolerance = 1e-8)),
  isTRUE(all.equal(unique(sim_sato$tlag), 0.239, tolerance = 1e-8))
)
c(`CL (L/h)` = cl_closed, `V (L)` = vc_closed,
  `kel (1/h)` = cl_closed / vc_closed,
  `t1/2 (h)` = log(2) / (cl_closed / vc_closed))
```

# PKNCA validation

```{r pknca}
# IMPORTANT: filter on !is.na(Cc) only. Adding `time > 0` or `Cc > 0` drops the
# time-zero row PKNCA needs to anchor AUC0-*.
sim_nca <- dplyr::bind_rows(
  sim |> dplyr::select(id, time, Cc, treatment),
  sim_sato |> dplyr::select(id, time, Cc, treatment)
) |>
  dplyr::filter(!is.na(Cc))

# Guarantee a time = 0 row per (treatment, id); pre-dose Cc = 0 is correct for
# an extravascular model.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(treatment, id, time, .keep_all = TRUE) |>
  dplyr::arrange(treatment, id, time)

# One row per dose event: expand the ii / addl compression used in the event
# tables.
expand_doses <- function(coh, amt_col, ii, n_dose, label) {
  coh |>
    dplyr::select(id, amt = dplyr::all_of(amt_col)) |>
    tidyr::crossing(dose_number = seq_len(n_dose)) |>
    dplyr::mutate(time = (dose_number - 1) * ii, treatment = label) |>
    dplyr::select(id, time, amt, treatment)
}
dose_df <- dplyr::bind_rows(
  expand_doses(cohort, "amt_tid", 8, 9L, "4 mg/kg TID"),
  expand_doses(cohort, "amt_qid", 6, 12L, "3 mg/kg QID"),
  tibble::tibble(id = 1L, time = c(0, 12), amt = sato_amt,
                 treatment = "Sato 4 mg/kg BID (typical)")
)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ug/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

# Zhang 2026 Table S2 summarises Cmax and AUC0-72h; Sato 2008 Table 6
# summarises Cmax, Tmax, AUC0-24h and t1/2. The terminal-slope window for the
# Sato arm stops at 18 h, where the typical profile is still above the 0.03
# ug/mL lower limit of quantification of the Zhang 2026 assay.
intervals <- dplyr::bind_rows(
  tibble::tibble(
    treatment = c("4 mg/kg TID", "3 mg/kg QID"),
    start = 0, end = 72, cmax = TRUE, tmax = TRUE, auclast = TRUE,
    half.life = FALSE
  ),
  tibble::tibble(
    treatment = "Sato 4 mg/kg BID (typical)",
    start = 0, end = 24, cmax = TRUE, tmax = TRUE, auclast = TRUE,
    half.life = FALSE
  ),
  tibble::tibble(
    treatment = "Sato 4 mg/kg BID (typical)",
    start = 12, end = 18, cmax = FALSE, tmax = FALSE, auclast = FALSE,
    half.life = TRUE
  )
) |>
  as.data.frame()

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

## Structural check: AUC0-72 = total dose / CL

Under a one-compartment model with complete absorption over the observation
window, each participant's AUC0-72 must equal their total tebipenem dose divided
by their individual CL, minus only the small amount still in the body at 72 h.
This is a per-subject identity, so it is a far stronger check than comparing
medians.

```{r auc-identity}
auc72 <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast", start == 0, end == 72) |>
  dplyr::select(treatment, id, auclast = PPORRES)

identity_chk <- auc72 |>
  dplyr::left_join(
    dplyr::bind_rows(
      cohort |> dplyr::transmute(id, treatment = "4 mg/kg TID",
                                 total_dose = 9 * amt_tid),
      cohort |> dplyr::transmute(id, treatment = "3 mg/kg QID",
                                 total_dose = 12 * amt_qid)
    ),
    by = c("treatment", "id")
  ) |>
  dplyr::left_join(cl_tid, by = "id") |>
  dplyr::mutate(ratio = auclast / (total_dose / cl))

stopifnot(nrow(identity_chk) == 2L * n_sub, !anyNA(identity_chk$ratio))
# The exact upper bound is 1 (mass balance). Two numerical effects perturb it:
# the shortfall below 1 is the drug still un-eliminated at 72 h plus the
# absorption lag, both of which grow with the individual lag time (which
# reaches ~11 h in this cohort); the small excess above 1 in a handful of
# subjects is trapezoidal overestimation of the absorption peak for those whose
# sampled ka exceeds about 50 1/h, i.e. a peak narrower than the 0.1 h
# simulation grid. Neither exceeds 2%, whereas a mis-transcribed clearance,
# dose or molecular-weight ratio would move the ratio by tens of percent.
stopifnot(all(identity_chk$ratio <= 1.02),
          all(identity_chk$ratio >= 0.85),
          stats::median(identity_chk$ratio) > 0.99,
          stats::median(identity_chk$ratio) < 1.01)

identity_chk |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    `Median AUC0-72 / (Dose/CL)` = stats::median(ratio),
    `Minimum` = min(ratio), `Maximum` = max(ratio), .groups = "drop"
  ) |>
  dplyr::rename(Regimen = treatment) |>
  knitr::kable(digits = 4,
               caption = "Per-participant mass-balance identity for AUC0-72.")
```

Because AUC0-72 is `total dose / CL` and the dose is proportional to body weight
while CL is also proportional to body weight, weight very nearly cancels: the
only sources of AUC variability the published model admits are creatinine
clearance and the CL random effect.

```{r auc-cv}
cv <- function(x) stats::sd(x) / mean(x)
tibble::tibble(
  Quantity = c("CL", "total dose / CL (TID)", "AUC0-72 (TID, PKNCA)"),
  `CV (%)` = 100 * c(
    cv(cl_tid$cl),
    cv(identity_chk$total_dose[identity_chk$treatment == "4 mg/kg TID"] /
         identity_chk$cl[identity_chk$treatment == "4 mg/kg TID"]),
    cv(identity_chk$auclast[identity_chk$treatment == "4 mg/kg TID"])
  )
) |>
  knitr::kable(digits = 1,
               caption = paste(
                 "Coefficients of variation. Zhang 2026 Table S2 reports an",
                 "AUC0-72h mean of 52.98 and SD of 19.37 ug*h/mL, i.e. a CV of",
                 "36.6%, which the published model cannot produce - see",
                 "Assumptions and deviations."
               ))
```

## Comparison against published NCA

Each simulated value is pulled from one explicitly named interval. This matters
for `tmax`: over the whole 0-72 h window PKNCA reports the absolute time of the
global maximum, which - because tebipenem accumulates by about 0.2% per dose -
lands on the *last* dose. Sato 2008's Tmax is measured from a dose (Sato 2008
Equation 11, `Tmax = Tlag + ...`), so the comparable quantity is the
dose-relative `tmax` PKNCA returns for the 12-18 h interval.

```{r nca-select}
nca_all <- as.data.frame(nca_res)

nca_value <- function(trt, code, start_, end_, n_expected) {
  v <- nca_all |>
    dplyr::filter(treatment == trt, PPTESTCD == code,
                  start == start_, end == end_) |>
    dplyr::pull(PPORRES)
  if (length(v) != n_expected || anyNA(v)) {
    stop("expected ", n_expected, " non-missing values for ", trt, " / ", code)
  }
  stats::median(v)
}

sato_arm <- "Sato 4 mg/kg BID (typical)"

simulated_nca <- tibble::tribble(
  ~treatment,     ~PPTESTCD,   ~PPORRES,
  "4 mg/kg TID",  "cmax",      nca_value("4 mg/kg TID", "cmax", 0, 72, n_sub),
  "4 mg/kg TID",  "auclast",   nca_value("4 mg/kg TID", "auclast", 0, 72, n_sub),
  sato_arm,       "cmax",      nca_value(sato_arm, "cmax", 0, 24, 1L),
  sato_arm,       "tmax",      nca_value(sato_arm, "tmax", 12, 18, 1L),
  sato_arm,       "auclast",   nca_value(sato_arm, "auclast", 0, 24, 1L),
  sato_arm,       "half.life", nca_value(sato_arm, "half.life", 12, 18, 1L)
)
stopifnot(nrow(simulated_nca) == 6L, !anyNA(simulated_nca$PPORRES))
```

```{r nca_compare}
published <- tibble::tribble(
  ~treatment,    ~PPTESTCD,   ~PPORRES,
  # Zhang 2026 Table S2 (Appendix S1), 4 mg/kg TID, 33,000 virtual
  # participants. Table S2 reports mean 3.46 / SD 2.34 / median 2.80 for Cmax
  # and mean 52.98 / SD 19.37 / median 50.44 for AUC0-72h; the medians are used
  # because the simulated values are also medians.
  "4 mg/kg TID", "cmax",      2.80,
  "4 mg/kg TID", "auclast",   50.44,
  # Sato 2008 Table 6, 4 mg/kg BID, whole 0.5-16 year age range, mean of 217
  # patients: Cmax 3.48 +/- 1.67 ug/mL, Tmax 0.74 +/- 0.26 h, AUC0-24h 11.00
  # +/- 1.84 ug*h/mL, t1/2 1.05 +/- 0.68 h.
  sato_arm,      "cmax",      3.48,
  sato_arm,      "tmax",      0.74,
  sato_arm,      "auclast",   11.00,
  sato_arm,      "half.life", 1.05
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = simulated_nca,
  reference = published,
  by        = "treatment",
  units     = c(cmax = "ug/mL", tmax = "h", auclast = "ug*h/mL",
                half.life = "h"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = paste(
    "Simulated vs published NCA. The TID rows compare against Zhang 2026",
    "Table S2 (Cmax and AUC0-72h in the Bangladeshi virtual population);",
    "the Sato rows compare a typical-value simulation at Sato 2008's median",
    "covariates against Sato 2008 Table 6 (AUC0-24h). * differs from the",
    "reference by more than 20%."
  ),
  align = c("l", "l", "r", "r", "r")
)
attr(cmp, "footnote")
```

All four Sato 2008 secondary parameters fall inside the published mean +/- 1 SD,
which confirms that the molecular-weight and unit rescalings were transcribed
without changing the underlying Sato 2008 predictions.

```{r sato-within-sd}
sato_ref <- tibble::tribble(
  ~PPTESTCD,   ~mean, ~sd,
  "cmax",      3.48,  1.67,
  "tmax",      0.74,  0.26,
  "auclast",   11.00, 1.84,
  "half.life", 1.05,  0.68
)
sato_chk <- sato_ref |>
  dplyr::left_join(
    simulated_nca |> dplyr::filter(treatment == sato_arm) |>
      dplyr::select(PPTESTCD, simulated = PPORRES),
    by = "PPTESTCD"
  ) |>
  dplyr::mutate(within_1sd = abs(simulated - mean) <= sd)
stopifnot(nrow(sato_chk) == 4L, !anyNA(sato_chk$simulated),
          all(sato_chk$within_1sd))

sato_chk |>
  dplyr::transmute(
    `NCA parameter` = PPTESTCD,
    `Sato 2008 mean` = mean, `Sato 2008 SD` = sd,
    Simulated = simulated, `Within 1 SD` = within_1sd
  ) |>
  knitr::kable(digits = 3,
               caption = "Typical-value prediction against Sato 2008 Table 6.")
```

# Assumptions and deviations

* **Virtual-population construction.** Zhang 2026 generated height from age and
  Box-Cox-transformed weight from height with sex-specific linear regressions
  (Table S1), then inverted the Box-Cox transform. The Box-Cox exponents lambda
  and the regression residual standard deviations are not published (the
  deposited script computes them from the screening data set, which is not
  public). This vignette therefore keeps the published height-from-age
  regression and derives its residual SD algebraically from the published
  R-squared and the published SD of height
  (`SD_resid = SD_height * sqrt(1 - R^2)`), but replaces the Box-Cox
  weight-from-height step with a log-normal weight whose marginal mean and SD
  are moment-matched to Zhang 2026 Table 1 and whose correlation with height is
  the published `sqrt(R^2)`. Age is drawn from a beta distribution on
  [24, 59] months moment-matched to the Table 1 sex-specific mean and SD rather
  than bootstrapped from the (non-public) screening ages. The resulting cohort
  reproduces Table 1 to within a fraction of an SD on every marginal.
* **Cohort size.** 200 participants per regimen rather than Zhang 2026's 500
  trials of 66 (33,000 participants), per this package's per-arm cap. The
  consequence is that the target-attainment proportions here are single-sample
  estimates compared against the median and 95% CI of the paper's 500-trial
  distribution, not a reproduction of the trial-level distribution in Zhang 2026
  Figures 3 and 5.
* **Creatinine clearance truncation.** Zhang 2026's deposited script draws
  `Ccr ~ N(3.72, 0.97)` mL/min/kg with no bounds, which admits non-physiological
  and (very rarely) negative values. Draws here are truncated to Sato 2008's
  observed range 1.46-6.97 mL/min/kg.
* **IIV scale.** Sato 2008 states that "a relative error model was applied for
  inter-subject variation" and its Table 4 CV percentages are exactly
  `sqrt(omega^2)` (113%, 20.2%, 74.7%, 145% for ka, CL/F, Vd/F and Tlag), i.e.
  a proportional `P = theta * (1 + eta)` parameterization. Zhang 2026's
  deposited `mrgsolve` model applies the same variances as **log-normal**
  (`exp(eta)`) IIV, which inflates the effective CV (ka 113% becomes 161%). This
  file follows the deposited implementation, because that is what produced the
  published results and because it matches Zhang 2026 Table S2 (see the NCA
  comparison above).
* **Residual error scale.** The deposited `mrgsolve` model writes
  `$SIGMA 0.453`, which mrgsolve treats as a variance (SD 0.673). Sato 2008
  reports sigma = 0.453 ug/mL as a standard deviation (Table 4) and
  sigma^2 = 0.205 as the variance (Table 5). This file uses the Sato 2008
  standard deviation. Nothing published in Zhang 2026 depends on the choice:
  every analysis there is computed from `IPRED`, never from `DV`.
* **Absorption lag time.** The deposited `mrgsolve` model computes
  `double ALAG1` in `$MAIN`, which does **not** set a lag in mrgsolve (that
  requires `ALAG_GUT`), and instead applies the lag by shifting the reported
  time axis (`TIME_LAG = TIME + ALAG1`) and discarding `TIME_LAG > 72`. For a
  fixed-interval multiple-dose regimen the two are equivalent, and this file
  uses the proper `alag(depot) <- tlag`.
* **The two Zhang 2026 corrections are self-cancelling re-parameterizations.**
  Rescaling CL/F and V/F by the molecular-weight ratio 0.7707 while rescaling
  the dose by the same factor leaves `D/V` and `kel = CL/V` unchanged, and the
  1000/60 factor merely converts the creatinine-clearance input unit. The
  corrected model therefore predicts exactly the same concentration-time profile
  as Sato 2008; what changes is that the dose is now expressed as tebipenem.
  The cross-check against Sato 2008 Table 6 above confirms this directly. Sato
  2008's own equations are internally consistent as published: at the cohort
  median (Ccr 3.73 mL/min/kg) they give CL/F = 0.751 L/h/kg and, at 4 mg/kg
  twice daily, AUC0-24h = 2 * 4 / 0.751 = 10.65 ug*h/mL against the 11.00 +/-
  1.84 reported in Table 6.
* **Unresolved discrepancy in Zhang 2026 Table S2.** The simulated AUC0-72h mean
  (about 52) and median (about 50) reproduce Table S2's 52.98 and 50.44 to
  within roughly 2%, and every Cmax moment matches, but the simulated AUC0-72h
  SD is about 12 ug*h/mL against the 19.37 reported. Under this model AUC0-72 is
  identically `total dose / CL` (verified per participant above), and because
  both dose and CL are proportional to body weight the only admissible sources
  of AUC variability are creatinine clearance (CV 13.5%) and the CL random
  effect (CV 20.2%), which bound the AUC coefficient of variation at about 24%.
  Table S2's CV of 36.6% is therefore not attainable from the published model.
  No parameter was adjusted to close the gap; the discrepancy is reported as
  found.
* **fT > MIC denominator.** Zhang 2026 divides the count of grid points above
  the MIC by a fixed 720 (72 h at 0.1 h). This vignette uses the mean of the
  indicator over the 721 grid points from 0 to 72 h inclusive, a difference of
  one part in 721.
* **Not reproduced.** Zhang 2026 Figure 2 (comparison against the WHO
  weight-for-age standard) needs the WHO growth-standard tables, which are not
  part of the paper. Zhang 2026 Figure 4 and Table 2 (the non-inferiority trial
  simulation against ceftriaxone) are a binomial trial-outcome calculation
  layered on top of the PK/PD target attainment rather than a property of the PK
  model, so they are outside the scope of a model-validation vignette.
