---
title: "Imipenem popPK model review (Zhang 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imipenem popPK model review (Zhang 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: Zhang P, Zhao Y, Zhu J, Yang Y, Liang G, Wang X, Yu Z.
  Population pharmacokinetics of imipenem in different populations for
  individualized dosing: a systematic review.
  *Frontiers in Pharmacology* 2025;16:1738055.
* Article: <https://doi.org/10.3389/fphar.2025.1738055>
* Supplement (Table S1, study protocols): <https://www.frontiersin.org/articles/10.3389/fphar.2025.1738055/full#supplementary-material>

This vignette is unusual for the package: **its source is a secondary
one.** Zhang 2025 is a PRISMA-style systematic review that develops no
model of its own. It catalogues 18 previously published imipenem
population-PK models and tabulates, for each, the structural parameter
estimates (Table 2), the inter-individual and residual variability
(Table 2), the covariate screening and the final covariate equations
(Table 3), and the dosing and sampling protocol (Supplementary Table
S1).

The standing policy for a review is to skip it and queue its cited
primaries. That policy carries an explicit exception for a review that
**tabulates the parameters**, and this one does: Tables 2 and 3 together
are enough to write down every differential equation and initialise
every `ini()` value for most of the 18 constituent models. So the models
below are transcribed from the review, each carrying a
"re-verify against the primary" note in its `description` and
`reference` metadata.

Every model file in this family records `ALL PARAMETER VALUES ARE
SECONDARY` in its `population$notes`. Read the
"Assumptions and deviations" section at the foot of this page before
using any of them; it reports a **measured** audit of how faithfully the
review reproduces its primaries, and that audit found a real defect.

# Coverage of the review's 18 studies

Four of the review's 18 studies were already in `nlmixr2lib`, extracted
directly from their primary publications. Those were **not**
re-extracted; they are better provenance than anything this review can
supply, and they serve below as the validation set for the review's own
accuracy. Three studies could not be reconstructed from the review
alone. Nothing was silently dropped.

```{r coverage, echo = FALSE}
coverage <- tibble::tribble(
  ~Study, ~`Author (review's citation)`, ~Disposition, ~`Model file`,
   1L, "Ikawa et al. (2008)",             "Extracted",              "Ikawa_2008_imipenem",
   2L, "Lamoth et al. (2009)",            "Already in registry",    "Lamoth_2009_imipenem",
   3L, "Yoshizawa et al. (2013)",         "Extracted (2 files)",    "Yoshizawa_2013_imipenem_neonates / _children",
   4L, "Couffignal et al. (2014)",        "Already in registry",    "Couffignal_2014_imipenem",
   5L, "Li and Xie (2019)",               "Blocked: sieving coefficient unreported", NA_character_,
   6L, "Dong et al. (2019)",              "Extracted",              "Dong_2019_imipenem",
   7L, "Li et al. (2020)",                "Blocked: dialysate-compartment role + Qd units", NA_character_,
   8L, "Chen et al. (2020b)",             "Extracted",              "Chen_2020_imipenem",
   9L, "de Velde et al. (2020)",          "Extracted (NONMEM arm)", "deVelde_2020_imipenem",
  10L, "Nguyen et al. (2021)",            "Extracted",              "Nguyen_2021_imipenem",
  11L, "Jaruratanasirikul et al. (2021)", "Blocked: needs an ABW canonical", NA_character_,
  12L, "Por et al. (2021)",               "Extracted",              "Por_2021_imipenem",
  13L, "Dao et al. (2022)",               "Extracted",              "Dao_2022_imipenem",
  14L, "Dinh et al. (2022)",              "Extracted",              "Dinh_2022_imipenem",
  15L, "Bai et al. (2024)",               "Already in registry",    "Bai_2024_imipenem",
  16L, "Lafaurie et al. (2023)",          "Extracted (2 files)",    "Lafaurie_2023_imipenem_neutropenia / _recovery",
  17L, "Truong et al. (2025)",            "Extracted",              "Truong_2025_imipenem",
  18L, "Wang et al. (2024)",              "Already in registry",    "Wang_2024_imipenem"
)
knitr::kable(coverage)
```

**13 model files extracted from 11 studies, 4 already covered from their
primaries, 3 blocked.** The three blocked studies are:

* **Li and Xie (2019)** -- a CRRT cohort whose total clearance is
  `CL_CRRT + CL_body`, with `CL_CRRT = Sd x CRRTintensity x BW/1000`.
  The sieving coefficient `Sd` is never given a value by the review, and
  every patient in that study was on CRRT, so the extracorporeal arm is
  not optional. The review also cites this study inconsistently -- Table
  1 and Table 2 say "Li and Xie (2019)" while Supplementary Table S1
  says "Li et al., 2018" -- so the primary should be identified before
  re-extraction.
* **Li et al. (2020)** -- a three-compartment CRRT model whose third
  compartment is a dialysate space carrying its own volume (0.012 L),
  its own intercompartmental clearance and its own assayed effluent
  concentration. The package's canonical `dialysate` compartment is
  defined as a *cumulative collected amount*, not a volume-bearing
  distribution compartment, so the role needs ratifying. Separately, the
  dialysate-flow covariate `Qd` is centred on 500 with no unit stated;
  CRRT flows are prescribed in mL/h while the canonical `DFR` column is
  mL/min, and with an exponent of 1.944 a 60-fold unit error becomes a
  2800-fold clearance error.
* **Jaruratanasirikul et al. (2021)** -- needs an `ABW` (adjusted body
  weight) covariate canonical, which the register does not yet carry
  (`WT`, `IBW`, `BMI`, `BSA` and `FFM` are registered; `ABW` and `LBW`
  are not). Its central volume is `V1 = 13.8 - 0.348 x (ABW - 60)`.
  Note also that this study's Table 2 central volume (13.6 L) disagrees
  with the intercept printed in its Table 3 formula (13.8 L).

# Population

The review pools 18 studies from 10 countries published between 2008
and 2025, spanning neonates through the elderly. Sample sizes run from
10 patients (Ikawa 2008) to 247 (Chen 2020); only three studies enrolled
more than 100. The study populations are predominantly critically ill
adults, with specific cohorts for burns, intraabdominal infection,
haematological malignancy, neutropenia, AECOPD, CRRT, ECMO, neonates and
children.

Per-model demographics are in each file's `population` metadata; the
table below collects the cohorts for the 13 models extracted here from
the review's Table 1.

```{r population, echo = FALSE}
pop <- tibble::tribble(
  ~Model, ~Country, ~N, ~`Age`, ~`Weight (kg)`, ~`Patient group`,
  "Ikawa_2008_imipenem",                 "Japan",       10L, "43.7 +/- 14.9 y",  "56.7 +/- 10.5",  "Intraabdominal infections",
  "Yoshizawa_2013_imipenem_neonates",    "Japan",       60L, "10.5 +/- 8.3 d",   "2.93 +/- 0.7",   "Neonates",
  "Yoshizawa_2013_imipenem_children",    "Japan",       39L, "9.61 +/- 3.16 y",  "29.5 +/- 10.9",  "Children",
  "Dong_2019_imipenem",                  "China",       56L, "4.86 +/- 2.33 y",  "18.65 +/- 6.90", "Children, haematological malignancy",
  "Chen_2020_imipenem",                  "China",      247L, "67 (20-97) y",     "65 (37.5-110)",  "Critically ill, +/- ECMO",
  "deVelde_2020_imipenem",               "Netherlands", 26L, "51 (39-54) y",     "75 (66-85)",     "Critically ill",
  "Nguyen_2021_imipenem",                "Vietnam",     44L, "65 (60-72) y",     "50 (47-55)",     "AECOPD",
  "Por_2021_imipenem",                   "USA",         23L, "51-55 y",          "89.6-105.1",     "Burns, +/- CVVH",
  "Dao_2022_imipenem",                   "Switzerland", 82L, "GA 26.9 wk, PNA 21 d", "1.16 (0.5-4.1)", "Neonates (extremely preterm)",
  "Dinh_2022_imipenem",                  "Vietnam",     24L, "57.5 +/- 19.9 y",  "51.3 +/- 8.6",   "Critically ill",
  "Lafaurie_2023_imipenem_neutropenia",  "France",      16L, "37 (18.3-78.3) y", "65.5 (48-101)",  "Neutropenic adults (during)",
  "Lafaurie_2023_imipenem_recovery",     "France",      16L, "37 (18.3-78.3) y", "65.5 (48-101)",  "Neutropenic adults (recovered)",
  "Truong_2025_imipenem",                "Netherlands",151L, "63 (51-72) y",     "70 (61.2-82)",   "Critically ill and non-critically ill"
)
knitr::kable(pop)
```

Programmatic access, e.g.:

```{r population-metadata}
str(readModelDb("Dao_2022_imipenem")()$population[c("n_subjects", "ga_range", "weight_median")])
```

# Source trace

Every `ini()` value in every file carries an in-file comment naming its
origin. Because all 13 models share one secondary source, the trace
collapses to a single rule, stated here once and repeated per parameter
in the files:

| Quantity | Source location in Zhang 2025 |
|---|---|
| Structural typical values (CL, V1, V2, V3, Q, Q2, Q3, Ke, Kcp, Kpc) | **Table 2**, "Population typical value" column, the row for that study |
| Inter-individual variability | **Table 2**, "Inter-individual variability (IIV)" column |
| Residual variability (proportional, additive) | **Table 2**, "Residual variability (RV)" column |
| Covariate reference values and exponents | **Table 3**, "Formulation" column |
| Which covariates were screened but not retained | **Table 3**, "Covariates screened" vs "Covariates incorporated" |
| Covariate selection method and thresholds | **Table 3**, "Covariate analysis method" column |
| Demographics (N, age, weight, sex, country, design) | **Table 1** |
| Dose, sampling schedule, bioanalytical assay | **Supplementary Table S1** |
| Software and model-evaluation method | **Table 2**, last two columns |

The covariate equations transcribed from Table 3 are:

| Model | Covariate model (Zhang 2025 Table 3, verbatim) |
|---|---|
| `Ikawa_2008_imipenem` | none reported ("NR") |
| `Yoshizawa_2013_imipenem_*` | per-kg parameters; no other covariate reported ("NR") |
| `Dong_2019_imipenem` | `CL = 8.6 x (BW/18)^0.75 x (age/4.69)^0.265 x (CLcr/214)^0.509`; `V1 = 7.2 x (BW/18)`; `V2 = 6.51 x (BW/18)`; `Q = 0.996 x (BW/18)^0.75` |
| `Chen_2020_imipenem` | `CL = 8.88 x (CLcr/59.1)^0.295 x (BW/65.0)^0.306 x e^1.16 [ECMO] x e^eta` |
| `deVelde_2020_imipenem` | `Ke = 0.637 x ((eGFR CKD-EPI-abs)/119)^0.655 x e^eta` |
| `Nguyen_2021_imipenem` | `CL = 7.88 x (CLcr/75.54)^0.532 x e^eta` |
| `Por_2021_imipenem` | `Vc = 32.67 x (BW/99.5)^0.74 x (ALB/2.7)^-1.17 x e^eta`; `Vp = 41.23 x (ALB/2.7)^-3.68`; no CVVH: `CL = 15.31 x (CLcr/145.83)^0.46 x (BW/99.5)^0.33 x e^eta`; with CVVH: `CL = 13.78 x (BW/99.5)^0.75 x e^eta + 1.56` |
| `Dao_2022_imipenem` | `CL = 0.21 x (BW/1.16)^0.75 x (1 + 0.22 x (PNA-21)/21) x (1 + 1.31 x (GA-26.9)/26.9) x (46.6/SCr)^0.2`; `V = 0.73 x (BW/1.16)^0.75` |
| `Dinh_2022_imipenem` | `CL = 4.79 x e^(0.00642 x CLcr)` |
| `Lafaurie_2023_imipenem_*` | none reported ("NR") |
| `Truong_2025_imipenem` | `CL = 14.6 x (CLcr CG/87.6)^0.462` |

# Reference subjects

Each model is exercised below at its **own** reference covariates -- the
centring values printed inside its Table 3 formula -- so that every
covariate term evaluates to exactly 1 and the model's typical clearance
and volume must equal the values printed in Table 2. That is what makes
the checks in the next section genuine transcription gates: if a
reference constant, an exponent, a unit conversion or the covariate
algebra is wrong, the reproduced clearance will not match the published
one.

```{r refsubjects}
# Reference covariates, published typical CL (L/h) and published central
# volume Vc (L) for each model, all at the reference subject.
#
# The published values are typed in as literal constants read off Zhang
# 2025 Table 2 / Table 3 -- deliberately NOT read back out of the model
# object, which would make the comparison vacuous.
cfg <- list(
  Ikawa_2008_imipenem = list(
    cov = list(), cl_pub = 9.42, vc_pub = 4.66, dose = 500),
  Yoshizawa_2013_imipenem_neonates = list(
    # per-kg model: CL = (CLr + CLnr) * WT, V = 0.466 * WT, at the cohort mean 2.93 kg
    cov = list(WT = 2.93), cl_pub = (0.0783 + 0.138) * 2.93, vc_pub = 0.466 * 2.93, dose = 2.93 * 20),
  Yoshizawa_2013_imipenem_children = list(
    cov = list(WT = 29.5), cl_pub = (0.187 + 0.0711) * 29.5, vc_pub = 0.203 * 29.5, dose = 29.5 * 15),
  Dong_2019_imipenem = list(
    cov = list(WT = 18, AGE = 4.69, CRCL = 214), cl_pub = 8.6, vc_pub = 7.2, dose = 18 * 20),
  Chen_2020_imipenem = list(
    cov = list(CRCL = 59.1, WT = 65.0, ECMO_STATUS = 0), cl_pub = 8.88, vc_pub = 20.5, dose = 500),
  deVelde_2020_imipenem = list(
    # micro-constant model: CL is a derived quantity, kel * Vc = 0.637 * 29.6
    cov = list(CRCL = 119), cl_pub = 0.637 * 29.6, vc_pub = 29.6, dose = 500),
  Nguyen_2021_imipenem = list(
    cov = list(CRCL = 75.54), cl_pub = 7.88, vc_pub = 15.1, dose = 500),
  Por_2021_imipenem = list(
    # ALB supplied in canonical SI g/L; the model converts to the g/dL the
    # source was calibrated on, so 27 g/L == the reference 2.7 g/dL.
    cov = list(WT = 99.5, ALB = 27, CRCL = 145.83, RRT_CRRT_STATUS = 0),
    cl_pub = 15.31, vc_pub = 32.67, dose = 500),
  Dao_2022_imipenem = list(
    # PNA supplied in canonical months; 21 days = 21/30.4375 months.
    cov = list(WT = 1.16, PNA = 21 / 30.4375, GA = 26.9, CREAT = 46.6),
    cl_pub = 0.21, vc_pub = 0.73, dose = 1.16 * 20),
  Dinh_2022_imipenem = list(
    # UNCENTRED exponential covariate: the published 4.79 L/h is CL at CRCL = 0.
    cov = list(CRCL = 0), cl_pub = 4.79, vc_pub = 11.1, dose = 500),
  Lafaurie_2023_imipenem_neutropenia = list(
    cov = list(), cl_pub = 14.3, vc_pub = 20.7, dose = 1000),
  Lafaurie_2023_imipenem_recovery = list(
    cov = list(), cl_pub = 10.9, vc_pub = 14.5, dose = 1000),
  Truong_2025_imipenem = list(
    cov = list(CRCL = 87.6), cl_pub = 14.6, vc_pub = 28.7, dose = 1000)
)
length(cfg)
```

# Simulation

A single intravenous **bolus** into `central` is used for the
transcription gates. A bolus makes both checks closed-form and exact:
immediately after the dose the entire amount is in the central
compartment, so `Cmax = Dose / Vc`; and over an infinite interval a
linear system must clear the whole dose, so `AUC(0, inf) = Dose / CL`.
Neither identity involves any quantity read back out of the model.

All random effects are zeroed with `rxode2::zeroRe()`, so these are
deterministic typical-value solves. Nothing below depends on a drawn
cohort, which also means nothing below can pass on one machine and fail
on another because of solver-thread RNG partitioning.

```{r simulate}
# Observation grid: dense early so Cmax is captured at the bolus peak,
# long-tailed so the terminal phase is well characterised for AUCinf.
obs_times <- unique(c(
  0,
  seq(0.05, 2, by = 0.05),
  seq(2.25, 12, by = 0.25),
  seq(13, 24, by = 1),
  seq(26, 120, by = 2)
))

solve_one <- function(nm) {
  spec <- cfg[[nm]]
  mod  <- rxode2::zeroRe(readModelDb(nm))

  ev <- rxode2::et(amt = spec$dose, cmt = "central", time = 0) |>
    rxode2::et(obs_times)
  ev <- as.data.frame(ev)
  ev$id <- 1L
  # Covariates are added to the materialized data frame, never assigned
  # onto the rxEt object (rxode2 silently drops those assignments).
  for (cv in names(spec$cov)) ev[[cv]] <- spec$cov[[cv]]

  out <- rxode2::rxSolve(mod, events = ev)
  data.frame(model = nm, time = out$time, Cc = out$Cc)
}

sim <- dplyr::bind_rows(lapply(names(cfg), solve_one))
dplyr::glimpse(sim)
```

## Figure: typical-value profiles after a single IV bolus

Replicates, qualitatively, the clearance comparison of Zhang 2025 Figure
2: the wide between-study spread the review's abstract reports is
visible as a wide spread of terminal slopes.

```{r fig-profiles, echo = FALSE, warning = FALSE}
adult_models <- setdiff(names(cfg), c("Yoshizawa_2013_imipenem_neonates",
                                      "Yoshizawa_2013_imipenem_children",
                                      "Dong_2019_imipenem",
                                      "Dao_2022_imipenem"))
sim |>
  dplyr::filter(model %in% adult_models, time <= 12, Cc > 1e-3) |>
  ggplot2::ggplot(ggplot2::aes(time, Cc, colour = model)) +
  ggplot2::geom_line() +
  ggplot2::scale_y_log10() +
  ggplot2::labs(
    x = "Time after dose (h)", y = "Imipenem concentration (mg/L)",
    colour = NULL,
    title = "Adult models at their reference subjects, single IV bolus"
  ) +
  ggplot2::theme(legend.text = ggplot2::element_text(size = 7))
```

# PKNCA validation

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id = model, time, Cc) |>
  dplyr::mutate(treatment = "IV bolus")

dose_df <- tibble::tibble(
  id        = names(cfg),
  time      = 0,
  amt       = vapply(cfg, function(x) x$dose, numeric(1)),
  treatment = "IV bolus"
)

conc_obj <- PKNCA::PKNCAconc(as.data.frame(sim_nca), Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_df), amt ~ time | treatment + id,
                             doseu = "mg")

intervals <- data.frame(
  start       = 0,
  end         = Inf,
  cmax        = TRUE,
  tmax        = TRUE,
  aucinf.obs  = TRUE,
  aucpext.obs = TRUE,
  half.life   = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca <- as.data.frame(res)
head(nca)
```

## Gate 1 -- clearance identity `AUC(0, inf) x CL = Dose`

For a linear system the whole dose is eventually cleared, so
`Dose / AUC(0, inf)` recovers total clearance exactly. Comparing that
against the clearance **printed in Zhang 2025 Table 2** tests the
covariate reference constants, the covariate algebra, the unit
conversions (albumin g/L to g/dL, postnatal age months to days) and the
micro-constant-to-clearance mapping all at once.

```{r gate-cl}
aucinf <- nca |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::select(model = id, aucinf = PPORRES)
pext <- nca |>
  dplyr::filter(PPTESTCD == "aucpext.obs") |>
  dplyr::select(model = id, aucpext = PPORRES)

cl_check <- aucinf |>
  dplyr::left_join(pext, by = "model") |>
  dplyr::mutate(
    dose     = vapply(cfg[model], function(x) x$dose,   numeric(1)),
    cl_pub   = vapply(cfg[model], function(x) x$cl_pub, numeric(1)),
    cl_nca   = dose / aucinf,
    pct_diff = 100 * (cl_nca - cl_pub) / cl_pub
  ) |>
  dplyr::arrange(model)

knitr::kable(
  cl_check |>
    dplyr::transmute(
      Model            = model,
      `Published CL (L/h)` = round(cl_pub, 4),
      `NCA Dose/AUCinf (L/h)` = round(cl_nca, 4),
      `% difference`   = round(pct_diff, 3),
      `AUC extrapolated (%)` = round(aucpext, 3)
    ),
  caption = "Clearance recovered from the simulated profile against the value printed in Zhang 2025 Table 2."
)

# The identity is exact in exact arithmetic; the only error is numerical
# (ODE solver tolerance plus trapezoidal AUC on a finite grid plus the
# log-linear extrapolation of the tail). 1% is far outside that numerical
# error and far inside any real transcription defect: a mis-typed
# clearance digit, a wrong covariate reference or a dropped unit
# conversion moves this by tens of percent.
stopifnot(all(abs(cl_check$pct_diff) < 1))
# Guard against the tail being so poorly characterised that aucinf is
# mostly extrapolation, which would make the gate above meaningless.
stopifnot(all(cl_check$aucpext < 5))
```

## Gate 2 -- central-volume identity `Cmax = Dose / Vc`

Immediately after an IV bolus the entire dose sits in the central
compartment, so the peak concentration is `Dose / Vc` regardless of how
many peripheral compartments the model has.

```{r gate-vc}
vc_check <- nca |>
  dplyr::filter(PPTESTCD == "cmax") |>
  dplyr::select(model = id, cmax = PPORRES) |>
  dplyr::mutate(
    dose     = vapply(cfg[model], function(x) x$dose,   numeric(1)),
    vc_pub   = vapply(cfg[model], function(x) x$vc_pub, numeric(1)),
    vc_nca   = dose / cmax,
    pct_diff = 100 * (vc_nca - vc_pub) / vc_pub
  ) |>
  dplyr::arrange(model)

knitr::kable(
  vc_check |>
    dplyr::transmute(
      Model                 = model,
      `Published Vc (L)`    = round(vc_pub, 4),
      `NCA Dose/Cmax (L)`   = round(vc_nca, 4),
      `% difference`        = round(pct_diff, 3)
    ),
  caption = "Central volume recovered from the simulated peak against the value printed in Zhang 2025 Table 2."
)

# 2% rather than 1% here: the peak is read off a finite observation grid,
# so a model with very fast initial distribution (Ikawa 2008, whose
# combined outflow rate constant from central exceeds 10 /h) can lose a
# fraction of a percent between the dose and the first recorded sample.
# A mis-transcribed volume moves this by tens of percent.
stopifnot(all(abs(vc_check$pct_diff) < 2))
```

## Gate 3 -- the review's own cross-study ranges

Zhang 2025 states several numeric ranges in its abstract and Discussion
that its authors computed *across* the constituent studies. Reproducing
them from the packaged models checks the set as a whole, independently
of any single file. These bounds are quoted from the review, not chosen
from an observed run.

```{r gate-ranges}
# Typical clearance at each model's own reference subject, from Gate 1.
cl_typ <- setNames(cl_check$cl_nca, cl_check$model)

# Zhang 2025 Discussion: "Substantial interindividual variability in PK
# parameters was observed across studies utilizing a two-compartment
# model (CL: 4.79-15.31 L/h; V1: 7.2-32.67 L; V2: 2.9-41.23 L;
# Q: 0.996-24.3 L/h)."
#
# deVelde 2020 is a two-compartment model but is EXCLUDED from this
# range, because it is the one study in the review parameterised in
# micro-constants: Table 2 reports Ke, Kcp, Kpc and Vc and never a
# clearance, so the review's authors had no CL to include. Its implied
# clearance is Ke * Vc = 18.9 L/h, which is reported as a deviation row
# below rather than forced into the range.
two_cmt_here <- c("Chen_2020_imipenem",
                  "Dinh_2022_imipenem", "Dong_2019_imipenem",
                  "Por_2021_imipenem", "Truong_2025_imipenem",
                  "Yoshizawa_2013_imipenem_children")

claims <- tibble::tribble(
  ~Claim, ~`Review's value`, ~Achieved, ~Pass,
  "Lowest adult CL across the review is Dinh 2022 at 4.79 L/h",
  4.79, round(unname(cl_typ["Dinh_2022_imipenem"]), 3),
  isTRUE(all.equal(unname(cl_typ["Dinh_2022_imipenem"]), 4.79, tolerance = 1e-3)),

  "Highest CL among two-compartment models is Por 2021 at 15.31 L/h",
  15.31, round(unname(cl_typ["Por_2021_imipenem"]), 3),
  isTRUE(all.equal(unname(cl_typ["Por_2021_imipenem"]), 15.31, tolerance = 1e-3)),

  "Every two-compartment model's CL lies within 4.79-15.31 L/h",
  NA_real_, round(max(cl_typ[intersect(two_cmt_here, names(cl_typ))]), 3),
  all(cl_typ[intersect(two_cmt_here, names(cl_typ))] >= 4.79 - 1e-3 &
      cl_typ[intersect(two_cmt_here, names(cl_typ))] <= 15.31 + 1e-3),

  "Children's V1 is 0.203 L/kg (Yoshizawa 2013), the review's lower V1 bound",
  0.203, round(unname(cfg$Yoshizawa_2013_imipenem_children$vc_pub) / 29.5, 4),
  isTRUE(all.equal(unname(cfg$Yoshizawa_2013_imipenem_children$vc_pub) / 29.5, 0.203, tolerance = 1e-3)),

  "Burn patients' V1 is 32.67 L (Por 2021), the review's upper V1 bound",
  32.67, round(unname(vc_check$vc_nca[vc_check$model == "Por_2021_imipenem"]), 3),
  isTRUE(all.equal(unname(vc_check$vc_nca[vc_check$model == "Por_2021_imipenem"]), 32.67, tolerance = 2e-3))
)

knitr::kable(claims, caption = "Quantitative claims made by Zhang 2025, reproduced from the packaged models.")
stopifnot(all(claims$Pass))

# Deviation, reported rather than gated: the implied clearance of the one
# micro-constant-parameterised model exceeds every clearance the review
# tabulates.
cl_develde <- unname(cl_typ["deVelde_2020_imipenem"])
knitr::kable(
  tibble::tibble(
    Finding = paste(
      "deVelde 2020 implied CL = Ke x Vc, outside the review's",
      "two-compartment CL range (4.79-15.31 L/h) AND above the",
      "abstract's stated adult maximum of 16.2 L/h"
    ),
    `Implied CL (L/h)` = round(cl_develde, 2)
  ),
  caption = "Known deviation from the review's own summary ranges (not gated; see the text below)."
)
stopifnot(cl_develde > 15.31)  # the deviation is real and reproducible, not marginal
```

The deviation row is not a transcription defect in this package; it is
a property of the review. Zhang 2025's abstract says imipenem clearance
"varied from 4.79 to 16.2 L/h in adults" and its Discussion gives
4.79-15.31 L/h for two-compartment models. Both ranges are computed
from the clearances printed in Table 2 -- and **de Velde 2020 has no
clearance in Table 2**, because it is the only study in the review
parameterised in micro-constants (`Ke`, `Kcp`, `Kpc`, `Vc`). Its
implied clearance, `Ke x Vc = 0.637 x 29.6`, is `r round(cl_develde, 1)`
L/h, higher than every clearance the review tabulates. The review's
headline spread is therefore narrower than its own constituent models
support. This is recorded here rather than gated away, and is a further
reason to re-verify each model against its primary.

Two of the review's headline ranges cannot be closed from the models
extracted here, because their endpoints belong to studies that were
already in the registry or are blocked:

* The abstract's adult clearance range **4.79-16.2 L/h** takes its upper
  endpoint from Lamoth 2009, which is `Lamoth_2009_imipenem` (extracted
  from its primary and validated in its own vignette).
* The two-compartment `Q` range **0.996-24.3 L/h** takes its upper
  endpoint from Jaruratanasirikul 2021, which is blocked on the `ABW`
  canonical. The lower endpoint, 0.996 L/h, is `Dong_2019_imipenem` and
  is reproduced above.

## Gate 4 -- renal-function gradient

Seven of the 13 models carry a renal-function covariate on clearance,
all with a positive coefficient. Clearance must therefore increase
monotonically with creatinine clearance in every one of them. A sign
error, or a reference constant placed in the wrong position of a ratio,
breaks this.

```{r gate-renal}
renal_models <- c("Chen_2020_imipenem", "deVelde_2020_imipenem",
                  "Dinh_2022_imipenem", "Dong_2019_imipenem",
                  "Nguyen_2021_imipenem", "Por_2021_imipenem",
                  "Truong_2025_imipenem")
crcl_grid <- c(15, 30, 60, 90, 120, 150)

renal_cl <- function(nm) {
  spec <- cfg[[nm]]
  mod  <- rxode2::zeroRe(readModelDb(nm))
  vapply(crcl_grid, function(cr) {
    ev <- rxode2::et(amt = spec$dose, cmt = "central", time = 0) |>
      rxode2::et(obs_times)
    ev <- as.data.frame(ev)
    ev$id <- 1L
    cov <- spec$cov
    cov$CRCL <- cr
    for (cv in names(cov)) ev[[cv]] <- cov[[cv]]
    out <- rxode2::rxSolve(mod, events = ev)
    # Dose / AUC recovers CL without reading any model internal.
    spec$dose / PKNCA::pk.calc.auc(out$Cc, out$time, interval = c(0, Inf),
                                  auc.type = "AUCinf",
                                  lambda.z = PKNCA::pk.calc.half.life(
                                    out$Cc, out$time)$lambda.z,
                                  clast = PKNCA::pk.calc.clast.obs(out$Cc, out$time))
  }, numeric(1))
}

renal_tab <- vapply(renal_models, renal_cl, numeric(length(crcl_grid)))
rownames(renal_tab) <- paste0("CrCL ", crcl_grid, " mL/min")
knitr::kable(round(t(renal_tab), 3),
             caption = "Typical clearance (L/h) across creatinine clearance, all other covariates at their reference values.")

# Strictly increasing in every model. This is a deterministic solve of a
# monotone algebraic expression, not a noisy cohort statistic, so exact
# monotonicity is the right assertion here (contrast the cohort-derived
# quantities warned about in the failure-pattern catalogue).
stopifnot(all(apply(renal_tab, 2, function(x) all(diff(x) > 0))))
```

## Gate 5 -- steady-state exposure and the review's PK/PD index

Zhang 2025 states that imipenem is time-dependent and that the relevant
index is `%fT>MIC`. The table below reports steady-state exposure for
the adult models on a common regimen, 1000 mg as a 0.5 h infusion every
8 h, with `%T>MIC` computed on total concentration at an MIC of 2 mg/L
(the EUCAST *Pseudomonas aeruginosa* breakpoint). This is descriptive
rather than a gate against a published table -- the review reports no
per-model NCA values, so there is nothing to compare against -- but the
spread is the review's central finding, made quantitative.

```{r gate-ss}
adults_ss <- c("Chen_2020_imipenem", "deVelde_2020_imipenem",
               "Dinh_2022_imipenem", "Ikawa_2008_imipenem",
               "Lafaurie_2023_imipenem_neutropenia",
               "Lafaurie_2023_imipenem_recovery",
               "Nguyen_2021_imipenem", "Por_2021_imipenem",
               "Truong_2025_imipenem")
tau <- 8
mic <- 2

ss_one <- function(nm) {
  spec <- cfg[[nm]]
  mod  <- rxode2::zeroRe(readModelDb(nm))
  ev <- rxode2::et(amt = 1000, cmt = "central", dur = 0.5, time = 0,
                   ii = tau, addl = 14) |>
    rxode2::et(seq(0, 15 * tau, by = 0.02))
  ev <- as.data.frame(ev)
  ev$id <- 1L
  for (cv in names(spec$cov)) ev[[cv]] <- spec$cov[[cv]]
  out <- rxode2::rxSolve(mod, events = ev)
  # Last complete dosing interval = steady state.
  ss <- out[out$time >= 14 * tau & out$time <= 15 * tau, ]
  data.frame(
    Model        = nm,
    `Cmax (mg/L)` = round(max(ss$Cc), 2),
    `Cmin (mg/L)` = round(min(ss$Cc), 3),
    `T>MIC (%)`   = round(100 * mean(ss$Cc > mic), 1),
    check.names   = FALSE
  )
}

ss_tab <- dplyr::bind_rows(lapply(adults_ss, ss_one))
knitr::kable(ss_tab,
             caption = "Steady-state exposure on 1000 mg q8h (0.5 h infusion), typical subject, MIC 2 mg/L.")

# Sanity, not tuning: on 3 g/day every adult model must keep a typical
# subject above a 2 mg/L MIC for some but not all of the interval, and
# must produce a physically ordered profile. A model whose clearance or
# volume was transcribed wrong by an order of magnitude fails this.
stopifnot(all(ss_tab$`Cmax (mg/L)` > ss_tab$`Cmin (mg/L)`))
stopifnot(all(ss_tab$`Cmax (mg/L)` > mic))
stopifnot(all(ss_tab$`T>MIC (%)` >= 20, ss_tab$`T>MIC (%)` <= 100))
```

# Assumptions and deviations

## Transcription fidelity of the secondary source -- measured, not assumed

Four of the review's 18 studies were already in `nlmixr2lib`, extracted
from their **primary** publications: `Lamoth_2009_imipenem`,
`Couffignal_2014_imipenem`, `Bai_2024_imipenem` and
`Wang_2024_imipenem`. They form a validation set for the review's
accuracy, and the result decides how much weight the other 13 models can
carry.

**Structural parameters and covariate equations: 4 of 4 agree exactly.**
The review's Table 2 typical values and Table 3 covariate formulations
match the corresponding primary-sourced model files digit for digit --
including Lamoth's additive split clearance `CL = 10.7 + 4.79 x eGFR/100`
with `V = 33.5 x BW/70`, Couffignal's three-exponent
`V1 = 20.4 x (BW/77)^1.3 x (ALB/18)^-1.1`, Bai's
`CL = 11.357 x (CrCl/99.896)^0.473` and Wang's
`CL = 13.1 x (CLCR/71)^0.263`. The review transcribes structure and
covariate models reliably.

**Inter-individual variability: the column is NOT on a single scale, and
at least one entry is wrong.** The review prints IIV as a bare
percentage under the heading "Inter-individual variability (IIV)"
without ever stating the convention, and it reproduces each primary's
own reported quantity without harmonising. Checked against the four
primary-sourced files:

| Study | Review Table 2 | What the primary actually reports | Convention |
|---|---|---|---|
| Lamoth 2009 | `CL: 17%` | "17% +/- 6% proportional IIV" | apparent CV |
| Couffignal 2014 | `CL: 38%`, `V1: 31%` | Monolix `omega CL (%)` | omega **SD** as a percent |
| Bai 2024 | `CL: 35.748%` | Table 2 column headed "Iiv (CV%)" | apparent CV |
| Wang 2024 | `CL: 8.32%` | Table 2 row **`omega^2 CL = 0.0832`** | a **VARIANCE**, mis-rendered as a percent |

The Wang 2024 row is a genuine defect in the review: 0.0832 is a
variance, corresponding to an apparent CV of 29.5%, and the review
prints it as "8.32%" -- understating that model's between-subject
variability roughly three-and-a-half-fold. It was caught only because
that study's primary is independently in this package.

**What this package does about it.** Every model transcribed from this
review adopts one uniform, documented reading: the printed percentage is
an **apparent CV** of a log-normal random effect, so
`omega^2 = log(1 + CV^2)`. That reading is:

* exact for a source reporting NONMEM-style CV% (Lamoth, Bai above, and
  by inference most of the NONMEM studies in the review);
* modestly low for a source reporting an omega SD as a percent (the
  Monolix studies -- Couffignal, and hence probably also Dinh 2022,
  Lafaurie 2023 and Nguyen 2021). At CV 30% the two readings differ by
  4% in variance; the gap widens with CV, so the largest exposure is
  `Dinh_2022_imipenem`'s 110% CV on `Q`, where the readings differ by
  more than 50%;
* badly wrong for a row that is really a variance -- but the only known
  such row, Wang 2024, is already extracted from its primary and is
  therefore not among the 13 models here.

The alternative -- blocking all 13 models until every primary is
obtained -- was rejected because the per-study answer requires exactly
the primaries that the "re-verify against the primary" notes already
flag. **Treat every IIV magnitude in this family as provisional. The
structural parameters, covariate equations and residual errors are on
much firmer ground.**

## Per-model deviations and warnings

* **`Chen_2020_imipenem` -- residual error is implausibly small.** The
  review reports 6.2% proportional plus 0.003 mg/L additive for a sparse
  routine-TDM dataset of 580 samples in 247 critically ill patients. It
  is by a wide margin the smallest residual error among the 18 studies.
  Transcribed verbatim, but a user simulating residual-error-bearing
  observations from this model should treat the noise level as
  unverified.
* **`Chen_2020_imipenem` -- the ECMO coefficient is very large.**
  `e^1.16 = 3.19`, i.e. ECMO more than triples clearance, for a
  mechanism the review's own Discussion attributes mainly to an
  increased volume of distribution via circuit adsorption. No standard
  error is reported. Re-verify before using this model for ECMO dosing.
* **`Por_2021_imipenem` -- the albumin exponent on peripheral volume is
  extreme.** `-3.68` means a halving of albumin multiplies `Vp` by 12.8.
  For a 23-patient cohort this is almost certainly poorly identified.
  The review reports neither a standard error nor the observed albumin
  range, so the fitted domain is unknown.
* **`Por_2021_imipenem` -- albumin unit conversion.** The reference 2.7
  is g/dL (a US study; 2.7 g/L would be incompatible with life). The
  canonical `ALB` column is SI g/L, so `model()` converts inline with
  `alb_gdL <- ALB * 0.1`, per the register's instruction for models
  calibrated on g/dL.
* **`Dao_2022_imipenem` -- postnatal age unit reparameterisation.** The
  source centres PNA on 21 **days**; the canonical `PNA` column is
  months, so `model()` recovers days with `PNA * 30.4375` before forming
  the centred term, leaving the printed constants untouched. Same
  pattern as the registered Zhao 2018 and Bardhi 2026 entries.
* **`Dao_2022_imipenem` -- volume exponent 0.75.** The conventional
  allometric pairing is 0.75 on clearance and 1 on volume; this model
  uses 0.75 on both, as printed. Reproduced verbatim rather than
  "corrected", but a transcription slip in the secondary source is a
  live alternative.
* **`Dao_2022_imipenem` -- the two age covariates are centred LINEAR
  terms, not power terms.** The gestational-age factor reaches zero at
  6.4 weeks and goes negative below that; the postnatal-age factor
  reaches zero at -74.5 days. Both are outside any physical domain, but
  the model will not error if given nonsense -- it will return a
  nonsense clearance.
* **`Dinh_2022_imipenem` -- uncentred exponential covariate.** The
  published typical clearance of 4.79 L/h is the value at
  `CrCL = 0 mL/min`, not at any cohort-typical renal function. At a
  normal 100 mL/min the model gives 9.11 L/h. The review's abstract
  quotes 4.79 L/h as the lowest adult clearance it found, which is
  therefore an extrapolated intercept rather than any patient's
  clearance.
* **`Dinh_2022_imipenem` -- unit typo in the source.** Table 2 prints
  `V1: 11.1 h`. Encoded as 11.1 L; a volume cannot be in hours and the
  sibling entries in the same cell are in L and L/h.
* **`Yoshizawa_2013_imipenem_*` -- footnote erratum.** The Table 2
  footnote glosses the clearance symbols as "CLr, clearance of plasma
  drug; CLnr, clearance of urine drug", which is incoherent. Read as
  renal and non-renal clearance. Supporting evidence: the renal arm
  dominates in children (0.187 vs 0.0711 L/h/kg) while the non-renal arm
  dominates in neonates (0.138 vs 0.0783), which is the direction renal
  maturation predicts. Only the sum enters the plasma model, so this
  affects the interpretation of the two etas, not the predictions.
* **`Yoshizawa_2013_imipenem_*` -- linear per-kg scaling.** All
  parameters are reported per kilogram, i.e. weight exponent 1 on
  clearance as well as volume, unlike the allometric 0.75 used by the
  other paediatric models here.
* **`deVelde_2020_imipenem` -- only the NONMEM arm is encoded.** The
  study also fitted the data non-parametrically in Pmetrics. That arm's
  covariate model is printed in terms of per-subject posterior medians,
  which are not population parameters; and a non-parametric population
  is a discrete support-point distribution, not the log-normal random
  effect `ini()` expresses. It cannot be faithfully encoded even in
  principle. The review's Pmetrics "Gamma: 3.4" is an error-polynomial
  multiplier, not a residual SD.
* **`deVelde_2020_imipenem` -- absolute, not BSA-normalised, eGFR.** The
  covariate is CKD-EPI **multiplied by BSA**, in mL/min. Supplying a
  standard mL/min/1.73 m^2 eGFR understates the ratio for anyone larger
  than 1.73 m^2.
* **`Dong_2019_imipenem` -- allometric exponents left unwrapped.** The
  0.75 on clearance and 1 on volume may well have been fixed rather than
  estimated, but the review reports no standard errors and does not say,
  so they are not wrapped in `fixed()`. The same applies to
  `Dao_2022_imipenem`.
* **`Truong_2025_imipenem` -- dose not reported.** The review records
  "NR" for this study's dosage and says so explicitly in its Results.
  The steady-state table above therefore assumes a regimen for this
  model rather than reproducing the study's own.
* **`Truong_2025_imipenem` -- all-male cohort.** Table 1 records the sex
  split as 151/0. Unusual enough to warrant re-checking against the
  primary; it would make sex unidentifiable by construction rather than
  merely non-significant.
* **`Lafaurie_2023_imipenem_*` -- two files, not one.** The review
  reports the neutropenic and recovered phases with separate structural
  values, separate IIV **and separate residual error**. A per-phase
  residual error is the signature of two independent fits rather than
  one model carrying a phase covariate, so they are shipped as two
  files.

## What is not validated here

* **No published NCA comparison.** The review reports no Cmax, Tmax, AUC
  or half-life for any constituent study, so
  `nlmixr2lib::ncaComparisonTable()` has nothing to compare against.
  Gates 1 and 2 are closed-form identities instead; Gate 3 reproduces
  the review's own cross-study claims; Gate 5 is descriptive.
* **No observed data and no VPC.** No study's individual concentrations
  are available, and the review reproduces no goodness-of-fit or VPC
  figure in a digitisable form.
* **The gates test the packaged encoding, not the review's arithmetic.**
  Gates 1 and 2 confirm that each model's ODE system, covariate algebra
  and unit conversions deliver the clearance and volume printed in Table
  2 at the reference subject. They cannot confirm that Table 2 itself
  faithfully reproduces the primary -- only re-reading each primary can,
  which is what the "re-verify against the primary" notes ask for. The
  fidelity audit above is the closest available substitute, and it found
  the IIV-scale defect described there.
