---
title: "Valproic acid protein-binding non-linearity (Zhang 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Valproic acid protein-binding non-linearity (Zhang 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Zhang et al. (2023) is a two-part paper. The first part is a systematic review
and external evaluation of ten previously published paediatric valproic acid
(VPA) population PK models. The second part -- the part packaged here -- is an
original modelling study in which the authors fit **a base model plus five
alternative non-linearity strategies** to their own 202-child therapeutic drug
monitoring (TDM) cohort, in order to ask which functional form best captures
the concentration-dependent plasma protein binding that makes VPA clearance
dose-dependent.

Two of those six models are packaged in this release:

* `Zhang_2023_valproic_acid_base` -- the linear reference model.
* `Zhang_2023_valproic_acid_exponent` -- Model V, the simple exponent model, in
  which apparent clearance is a power function of the patient's own daily dose.

The remaining four (Model I one-binding-site, Model II Langmuir, Model III
dose-dependent maximum effect, Model IV linear non-saturable binding) are
**not yet packaged** -- see [Deferred models](#deferred-models) below for the
two specific blockers.

```{r citation, echo = FALSE, results = "asis"}
for (nm in c("Zhang_2023_valproic_acid_base",
             "Zhang_2023_valproic_acid_exponent")) {
  m <- readModelDb(nm)()
  cat("**`", nm, "`**\n\n", sep = "")
  cat("* Citation: ", m$reference, "\n", sep = "")
  cat("* Description: ", m$description, "\n\n", sep = "")
}
```

* Article: <https://doi.org/10.3389/fphar.2023.1228641>
* Supplement (Supplementary Tables S1-S3, Figures S1-S5):
  <https://www.frontiersin.org/articles/10.3389/fphar.2023.1228641/full#supplementary-material>

# Population

The evaluation cohort was 202 Chinese children with epilepsy (139 male / 63
female) treated with VPA at Wuhan Children's Hospital between January 2016 and
November 2018, contributing 255 total plasma VPA concentrations (Zhang 2023
Table 2). Age was 4.92 years (range 0.17-15.00), body weight 19.00 kg (range
4.00-70.00), daily dose 23.44 mg/kg/day (range 8.70-57.69, equivalent to
60-1250 mg/day), and the observed VPA serum concentration 50.40 mg/L (range
22.60-118.50). Serum albumin was 42.10 g/L (range 29.70-70.50). VPA was given
orally as a syrup (194 records) or a sustained-release tablet (61 records),
once, twice or three times daily.

Two features of this cohort shape every model below. First, **all samples were
troughs collected under steady-state conditions**, so absorption and
distribution are only weakly identified -- which is why the authors fixed the
absorption rate constants to literature values and why the estimated `V/F` is
large relative to the paediatric VPA literature. Second, patients with hepatic
or renal impairment, abnormal albumin, or poor adherence were excluded, so the
albumin range is narrow.

Concentrations were measured by gas chromatography (limit of detection 1 mg/L,
calibration range 12.5-150 mg/L, CV below 10%). 77 of 202 children received a
concomitant antiseizure medication -- most often levetiracetam (27),
oxcarbazepine (24) and topiramate (19) -- but no co-medication effect was
retained in any of the authors' own models, because so few patients were on the
classical enzyme inducers that dominate the older literature.

The same information is available programmatically via
`readModelDb("Zhang_2023_valproic_acid_base")()$population`.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| One-compartment, first-order absorption | n/a | Zhang 2023 Section 2.4 ("a one-compartment model with first-order absorption was used as base model") |
| `lka` (syrup) | 2.64 1/h, FIXED | Supplementary Table S3 footnote |
| `e_form_tablet_ka` | log(1.57 / 2.64), FIXED | Supplementary Table S3 footnote (conventional tablet Ka 1.57 1/h) |
| `e_form_vpa_sr_ka` | log(0.46 / 2.64), FIXED | Supplementary Table S3 footnote (SR tablet Ka 0.46 1/h) |
| **Base model** | | |
| `lcl` | 0.311 L/h (4.6% RSE) | Supplementary Table S3, "Base model" |
| `lvc` | 27.8 L (8.8% RSE) | Supplementary Table S3, "Base model" |
| `etalcl` | BSV 45.7% -> omega^2 0.18966 | Supplementary Table S3, "Base model" |
| `etalvc` | BSV 68.6% -> omega^2 0.38567 | Supplementary Table S3, "Base model" |
| `propSd` / `addSd` | 8.9% / 6.3 mg/L | Supplementary Table S3, "Base model" |
| **Model V (simple exponent)** | | |
| `CL/F = CLp/F * (DD/25)^k` | n/a | Equation 7 |
| `lcl` | 0.331 L/h (2.1% RSE) | Supplementary Table S3, "Model V" |
| `lvc` | 17.2 L (20.0% RSE) | Supplementary Table S3, "Model V" |
| `e_dose_vpa_mgkgd_cl` | 0.658 (7.7% RSE) | Supplementary Table S3, "Model V" (row `CL_DD`) |
| `etalcl` | BSV 24.3% -> omega^2 0.05737 | Supplementary Table S3, "Model V" |
| `etalvc` | BSV 45.1% -> omega^2 0.18515 | Supplementary Table S3, "Model V" |
| `propSd` / `addSd` | 14.1% / 4.2 mg/L | Supplementary Table S3, "Model V" |
| Reference daily dose 25 mg/kg/day | n/a | Equation 7 denominator; cohort mean 24.50 mg/kg/day (Table 2) |
| Observed concentration median 50.40 mg/L | reference value | Table 2 |
| Objective function values 1752.1 (base) / 1615.9 (Model V) | reference value | Supplementary Table S3 |
| Prediction metrics MDPE 1.50%, MAPE 17.68% (Model V) | reference value | Table 3 |

Between-subject variability is reported by Zhang 2023 as CV%, so the internal
variance is `omega^2 = log(CV^2 + 1)`.

# Virtual cohort

The original observations are not public. The cohort below approximates the
published demographics (Table 2) and is split into three daily-dose arms so the
dose-dependence of Model V can be exercised and tested. The reference arm is
25 mg/kg/day, the denominator of Equation 7 and the rounded cohort mean; the
flanking arms are exactly half and double that, which turns the paper's power
exponent into an exact arithmetic identity we can assert on.

```{r cohort}
set.seed(20231006)

n_per_arm <- 150L # cap is 200 per arm
tau <- 12         # twice-daily dosing interval (h)

# Uniform grid over one dosing interval. A uniform spacing matters: the
# structural checks below take a trapezoidal time-average of the profile, and
# an unevenly-spaced grid would weight the densely-sampled absorption phase
# more heavily than the terminal phase and bias that average per arm.
obs_times <- seq(0, tau, by = 0.25)

# The three dose arms are PAIRED: one set of subject covariates is drawn once
# and reused across all three arms (only the id offset and the daily dose
# change). Drawing weights independently per arm would leave sampling noise in
# the between-arm exposure ratios and make the exact dose-scaling identities
# below untestable.
# Covariates are placed on deterministic quantiles rather than drawn randomly.
# A random draw of n = 150 leaves several percent of sampling noise in the
# cohort median (one seed gave 21.6 kg against the published 19.0 kg), which
# propagates straight into the comparison against the published trough. Using
# quantiles pins the cohort median to the published value exactly and makes the
# whole vignette seed-independent.
subj_template <- tibble(
  # Weight: lognormal quantiles with median exactly the published 19.00 kg,
  # clamped to the published 4.00-70.00 kg range (Zhang 2023 Table 2).
  WT = pmin(pmax(round(stats::qlnorm(ppoints(n_per_arm), log(19), 0.45), 1), 4), 70),
  # 61 of 255 records (23.9%) were sustained-release tablet; the rest syrup.
  FORM_VPA_SR = as.integer(seq_len(n_per_arm) <= round(n_per_arm * 61 / 255)),
  FORM_TABLET = 0L
)

# The cohort must reproduce the published median weight exactly.
stopifnot(abs(median(subj_template$WT) - 19.00) < 0.05)

# Build one daily-dose arm. Steady state is imposed with ss = 1 on the first
# dose record: the apparent terminal half-life of these models is 35-60 h, so
# dosing up from zero would need days of run-in to reach the trough the paper
# actually observed.
make_arm <- function(dd_mgkgd, id_offset = 0L) {
  subj <- subj_template |>
    mutate(
      id = id_offset + seq_len(n()),
      DOSE_VPA_MGKGD = dd_mgkgd,
      arm = paste0(dd_mgkgd, " mg/kg/day"),
      amt_per_dose = DOSE_VPA_MGKGD * WT / (24 / tau)
    )

  doses <- subj |>
    mutate(time = 0, evid = 1L, amt = amt_per_dose,
           cmt = "depot", ii = tau, ss = 1L)

  obs <- subj |>
    tidyr::crossing(time = obs_times) |>
    mutate(evid = 0L, amt = NA_real_,
           # cmt is the ODE STATE name, never the observable "Cc"
           cmt = "central", ii = 0, ss = 0L)

  bind_rows(doses, obs) |>
    arrange(id, time, desc(evid)) |>
    select(id, time, evid, amt, cmt, ii, ss,
           WT, FORM_VPA_SR, FORM_TABLET, DOSE_VPA_MGKGD, arm)
}

events <- bind_rows(
  make_arm(12.5, id_offset = 0L),
  make_arm(25.0, id_offset = n_per_arm),
  make_arm(50.0, id_offset = 2L * n_per_arm)
)

# Disjoint IDs across arms are mandatory: duplicated ids are silently merged
# by rxSolve into one subject receiving the summed dose.
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
stopifnot(length(unique(events$id)) == 3L * n_per_arm)
# The arms are paired: the sorted weight vector must be identical in each arm.
wt_by_arm <- events |>
  distinct(arm, id, WT) |>
  group_by(arm) |>
  summarise(wt = list(sort(WT)), .groups = "drop")
stopifnot(length(unique(wt_by_arm$wt)) == 1L)

events |>
  group_by(arm) |>
  summarise(n = n_distinct(id),
            `Median weight (kg)` = median(WT),
            `Median dose per administration (mg)` = round(median(amt, na.rm = TRUE), 1),
            .groups = "drop") |>
  rename("Arm" = arm, "N" = n) |>
  knitr::kable(caption = "Virtual cohort by daily-dose arm (twice-daily dosing).")
```

# Simulation

```{r simulate}
mod_base <- readModelDb("Zhang_2023_valproic_acid_base")
mod_expo <- readModelDb("Zhang_2023_valproic_acid_exponent")

keep_cols <- c("arm", "WT", "DOSE_VPA_MGKGD", "FORM_VPA_SR")

sim_base <- rxode2::rxSolve(mod_base, events = events, keep = keep_cols) |>
  as.data.frame() |>
  mutate(model = "Base model")

sim_expo <- rxode2::rxSolve(mod_expo, events = events, keep = keep_cols) |>
  as.data.frame() |>
  mutate(model = "Model V (simple exponent)")

sim <- bind_rows(sim_base, sim_expo)

# rxSolve returns observation records only (no evid column), so every row
# below is an observation.
stopifnot(nrow(sim) > 0, !all(is.na(sim$Cc)))
```

Typical-value (zero between-subject variability) profiles are used for the
structural checks. Passing `omega = NA` to `rxSolve()` zeroes the random
effects for this call only, without mutating the shared model object.

```{r simulate-typical}
typ_base <- rxode2::rxSolve(mod_base, events = events, keep = keep_cols,
                            omega = NA) |>
  as.data.frame() |>
  mutate(model = "Base model")

typ_expo <- rxode2::rxSolve(mod_expo, events = events, keep = keep_cols,
                            omega = NA) |>
  as.data.frame() |>
  mutate(model = "Model V (simple exponent)")

typ <- bind_rows(typ_base, typ_expo)
```

# Structural checks

## The exponent is exactly the published 0.658

Equation 7 sets `CL/F = CLp/F * (DD/25)^k` with `k = 0.658`. Halving and
doubling the daily dose relative to the 25 mg/kg/day reference must therefore
scale typical clearance by exactly `0.5^0.658` and `2^0.658`. The base model,
having no dose covariate, must show a clearance ratio of exactly 1.

```{r exponent-identity}
k_pub <- 0.658

cl_typ <- typ |>
  group_by(model, arm, DOSE_VPA_MGKGD) |>
  summarise(cl = mean(cl), .groups = "drop") |>
  group_by(model) |>
  mutate(cl_ratio = cl / cl[DOSE_VPA_MGKGD == 25]) |>
  ungroup() |>
  mutate(expected = if_else(model == "Base model",
                            1,
                            (DOSE_VPA_MGKGD / 25)^k_pub))

cl_typ |>
  mutate(across(c(cl, cl_ratio, expected), \(x) round(x, 4))) |>
  rename("Model" = model, "Arm" = arm,
         "Daily dose (mg/kg/day)" = DOSE_VPA_MGKGD,
         "Typical CL/F (L/h)" = cl,
         "CL/F ratio vs 25 mg/kg/day" = cl_ratio,
         "Expected ratio" = expected) |>
  knitr::kable(caption = "Typical clearance scales with daily dose exactly as Equation 7 specifies.")

stopifnot(all(abs(cl_typ$cl_ratio - cl_typ$expected) < 1e-6))

# The reference-arm typical clearances must be the published point estimates.
ref_cl <- cl_typ |> filter(DOSE_VPA_MGKGD == 25)
stopifnot(
  abs(ref_cl$cl[ref_cl$model == "Base model"] - 0.311) < 1e-6,
  abs(ref_cl$cl[ref_cl$model == "Model V (simple exponent)"] - 0.331) < 1e-6
)
```

## Exposure rises sub-proportionally with dose under Model V

Because clearance itself increases with dose, Model V predicts that doubling
the daily dose raises the average steady-state concentration by a factor of
only `2 / 2^0.658 = 1.268`, not 2. The linear base model must double exactly.
This is the qualitative behaviour the paper is trying to capture, expressed as
an exact identity.

First check the solved profile against the analytic steady-state identity: for
any linear one-compartment model at steady state, the area under one dosing
interval equals `Dose / CL` exactly. Recovering that from the trapezoidal
integral of the simulated curve validates the ODE solve and the `ss = 1`
handling together, independently of any dose-scaling argument.

```{r ss-auc-identity}
trapz <- function(x, y) sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)

auc_typ <- typ |>
  arrange(model, id, time) |>
  group_by(model, arm, DOSE_VPA_MGKGD, id) |>
  summarise(auc_trap = trapz(time, Cc),
            cl = first(cl),
            .groups = "drop") |>
  # 1:1 join by id onto a per-subject summary table (not per-time-point rows)
  left_join(events |> filter(evid == 1) |> distinct(id, amt), by = "id") |>
  mutate(auc_exact = amt / cl,
         pct_err = 100 * (auc_trap - auc_exact) / auc_exact)

# Trapezoidal integration on a 0.25 h grid recovers Dose/CL to well under 1%.
stopifnot(max(abs(auc_typ$pct_err)) < 1)

auc_typ |>
  group_by(model, arm) |>
  summarise(`Median trapezoidal AUC0-tau (mg*h/L)` = round(median(auc_trap), 2),
            `Median Dose/CL (mg*h/L)` = round(median(auc_exact), 2),
            `Max abs error (%)` = round(max(abs(pct_err)), 3),
            .groups = "drop") |>
  rename("Model" = model, "Arm" = arm) |>
  knitr::kable(caption = paste("The solved steady-state profile reproduces the",
                               "analytic identity AUC0-tau = Dose/CL."))
```

Now the dose-scaling identity itself. Because the arms are paired on identical
subjects, the between-arm ratio of the average steady-state concentration
`Cav = AUC0-tau / tau` is exactly `(dose ratio) / (clearance ratio)`.

```{r dose-proportionality}
cav_typ <- auc_typ |>
  mutate(cav = auc_trap / tau) |>
  group_by(model, DOSE_VPA_MGKGD) |>
  summarise(cav = median(cav), .groups = "drop") |>
  group_by(model) |>
  mutate(cav_ratio = cav / cav[DOSE_VPA_MGKGD == 25]) |>
  ungroup() |>
  mutate(expected = if_else(
    model == "Base model",
    DOSE_VPA_MGKGD / 25,
    (DOSE_VPA_MGKGD / 25) / (DOSE_VPA_MGKGD / 25)^k_pub
  ))

cav_typ |>
  mutate(across(c(cav, cav_ratio, expected), \(x) round(x, 4))) |>
  rename("Model" = model, "Daily dose (mg/kg/day)" = DOSE_VPA_MGKGD,
         "Median typical Cav (mg/L)" = cav,
         "Cav ratio vs 25 mg/kg/day" = cav_ratio,
         "Expected ratio" = expected) |>
  knitr::kable(caption = "Base model exposure is dose-proportional; Model V is sub-proportional.")

stopifnot(all(abs(cav_typ$cav_ratio - cav_typ$expected) < 2e-3))
```

## Formulation switches the absorption rate constant

```{r formulation-check}
ka_check <- typ |>
  filter(model == "Base model") |>
  group_by(FORM_VPA_SR) |>
  summarise(ka = mean(ka), .groups = "drop")

stopifnot(
  abs(ka_check$ka[ka_check$FORM_VPA_SR == 0] - 2.64) < 1e-6,
  abs(ka_check$ka[ka_check$FORM_VPA_SR == 1] - 0.46) < 1e-6
)

ka_check |>
  mutate(formulation = if_else(FORM_VPA_SR == 1, "SR tablet", "Syrup (reference)")) |>
  select(formulation, ka) |>
  rename("Formulation" = formulation, "Ka (1/h)" = ka) |>
  knitr::kable(caption = "Formulation-specific fixed absorption rate constants.")
```

# Replicate published figures

## Supplementary Figure S5B -- clearance versus daily dose under Model V

Supplementary Figure S5B plots empirical Bayes estimates of VPA apparent
clearance against daily dose for the simple exponent model, with the
model-predicted typical clearance as a dashed line. The panel below reproduces
that structure from the packaged model: individual clearances from the
between-subject-variability simulation as points, and the typical-value curve
as the dashed line. As in the paper, the x-axis is the absolute daily dose in
mg/day, so the curve reflects the joint distribution of weight and mg/kg dose
in the cohort rather than being a pure function of the axis.

```{r figure-s5b}
ebe <- sim |>
  filter(model == "Model V (simple exponent)") |>
  group_by(id, arm, WT, DOSE_VPA_MGKGD) |>
  summarise(cl = first(cl), .groups = "drop") |>
  mutate(dd_mgd = DOSE_VPA_MGKGD * WT)

typ_curve <- tibble(dd_mgkgd = seq(8.7, 57.7, length.out = 200)) |>
  mutate(cl = 0.331 * (dd_mgkgd / 25)^k_pub,
         dd_mgd = dd_mgkgd * 19)

ggplot(ebe, aes(dd_mgd, cl)) +
  geom_point(shape = 1, colour = "steelblue", alpha = 0.6) +
  geom_line(data = typ_curve, linetype = "dashed", colour = "red", linewidth = 0.9) +
  labs(x = "Daily dose (mg/d)", y = "Apparent clearance (L/h)",
       title = "The simple exponent model",
       caption = paste("Replicates Supplementary Figure S5B of Zhang 2023.",
                       "Dashed line: typical CL/F at the cohort median 19 kg.")) +
  theme_bw()
```

## Steady-state concentration-time profiles

```{r figure-profiles}
sim |>
  group_by(model, arm, time) |>
  summarise(Q05 = quantile(Cc, 0.05, na.rm = TRUE),
            Q50 = quantile(Cc, 0.50, na.rm = TRUE),
            Q95 = quantile(Cc, 0.95, na.rm = TRUE),
            .groups = "drop") |>
  ggplot(aes(time, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~model) +
  labs(x = "Time within the dosing interval (h)",
       y = "Total plasma VPA (mg/L)",
       colour = "Daily dose", fill = "Daily dose",
       title = "Steady-state profiles over one 12 h interval",
       caption = paste("Median with 5th-95th percentile band, 150 subjects per arm.",
                       "Note the compressed spacing of the Model V arms:",
                       "clearance rises with dose.")) +
  theme_bw()
```

# PKNCA validation

The paper's own observations are steady-state troughs, so the NCA below is run
over one complete 12 h steady-state dosing interval (Recipe 3). Because
`ss = 1` puts the system at steady state from `time = 0`, the `time = 0` record
is the steady-state trough and is a genuine measurement -- the usual
`Cc = 0` pre-dose back-fill would be wrong here and is deliberately omitted.

```{r pknca}
sim_nca <- sim |>
  filter(!is.na(Cc)) |>
  select(model, id, time, Cc, arm)

# time = 0 is present by construction (it is the steady-state trough).
stopifnot(all(
  sim_nca |> group_by(model, id) |> summarise(has0 = any(time == 0), .groups = "drop") |> pull(has0)
))

dose_df <- events |>
  filter(evid == 1) |>
  select(id, time, amt, arm)

intervals <- data.frame(
  start = 0, end = tau,
  cmax = TRUE, tmax = TRUE, cmin = TRUE,
  auclast = TRUE, cav = TRUE
)

run_nca <- function(model_name) {
  conc_obj <- PKNCA::PKNCAconc(
    sim_nca |> filter(model == model_name) |> select(id, time, Cc, arm),
    Cc ~ time | arm + id, concu = "mg/L", timeu = "h"
  )
  dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id, doseu = "mg")
  PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
}

nca_base <- run_nca("Base model")
nca_expo <- run_nca("Model V (simple exponent)")
```

```{r pknca-table}
tidy_nca <- function(res, model_name) {
  as.data.frame(res$result) |>
    # PKNCA emits dependency rows; filter on the interval as well as the code
    filter(start == 0, end == tau,
           PPTESTCD %in% c("cmax", "tmax", "cmin", "auclast", "cav")) |>
    group_by(arm, PPTESTCD) |>
    summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
    mutate(model = model_name)
}

nca_tbl <- bind_rows(tidy_nca(nca_base, "Base model"),
                     tidy_nca(nca_expo, "Model V (simple exponent)"))

nca_tbl |>
  mutate(median = round(median, 2),
         Parameter = nlmixr2lib::ncaParamLabel(PPTESTCD)) |>
  select(model, arm, Parameter, median) |>
  tidyr::pivot_wider(names_from = Parameter, values_from = median) |>
  rename("Model" = model, "Arm" = arm) |>
  knitr::kable(caption = paste("Median steady-state NCA over one 12 h interval.",
                               "Cmin is the trough the paper's TDM samples measured."))
```

## Comparison against the published concentration

Zhang et al. report no NCA table -- their cohort is trough-only TDM data, and
their model metrics are prediction errors against individual observations that
are not public. The one published concentration that a simulation can be held
against is the observed steady-state trough itself: **median 50.40 mg/L**
(mean 54.34, range 22.60-118.50; Table 2), at a cohort median daily dose of
23.44 mg/kg/day. The 25 mg/kg/day arm is the closest match.

```{r nca-compare}
published <- tibble::tribble(
  ~arm,             ~cmin,
  "25 mg/kg/day",   50.40
)

cmp_base <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_base, reference = published, by = "arm",
  units = c(cmin = "mg/L"), tolerance_pct = 20
)
cmp_expo <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_expo, reference = published, by = "arm",
  units = c(cmin = "mg/L"), tolerance_pct = 20
)

bind_rows(
  cmp_base |> mutate(Model = "Base model"),
  cmp_expo |> mutate(Model = "Model V (simple exponent)")
) |>
  relocate(Model) |>
  knitr::kable(
    caption = paste("Simulated vs published steady-state trough (Zhang 2023 Table 2).",
                    "* differs from the reference by more than 20%."),
    align = c("l", "l", "l", "r", "r", "r")
  )
```

Both models reproduce the published median trough closely -- the base model
about 3% below it and Model V about 2% above -- and neither row is starred. For
Model V the agreement is a genuine cross-check rather than a coincidence: the
paper independently reports a median prediction error of **+1.50%** for this
model against its own observations (Table 3), and the simulation here lands at
+2.0% against the observed median. No parameter was tuned; the only change made
after first seeing these numbers was to place the cohort weights on
deterministic quantiles so the simulated median weight equals the published
19.00 kg exactly, instead of leaving several percent of sampling noise from a
150-subject random draw in the comparison.

Two caveats on what this check does and does not establish. It compares medians
across variable populations, so it validates the central tendency of `CL/F` and
the steady-state trough machinery, but says nothing about the variance
structure -- and the paper is explicit that variance was where every model it
evaluated performed badly (all models were rejected by the normalised
prediction distribution error global test). It also uses a single 25 mg/kg/day
twice-daily stratum in place of a cohort spanning 8.70-57.69 mg/kg/day and one
to three doses daily, so it cannot detect a misspecified dose-dependence; the
exact identities in the structural-checks section above are what pin that.

# Assumptions and deviations

* **Cohort composition is reconstructed, not published.** Weights are placed on
  lognormal quantiles with the median pinned to the published 19.00 kg and
  clamped to the published 4.00-70.00 kg range; the paper reports only summary
  statistics (Table 2), not the joint distribution of weight, age and dose.
  Deterministic quantiles are used in preference to a random draw so the cohort
  median matches the published value exactly and the vignette is
  seed-independent. The sustained-release fraction is set to 61/255 to match the
  record counts in Table 2. Age is not simulated at all, because no model in
  this paper uses it.
* **Twice-daily dosing was chosen for the simulation.** The paper states
  dosing was "one, two, or three times per day" without reporting the mix, so
  a single representative interval (`tau = 12 h`) is used throughout.
* **Steady state is imposed with `ss = 1`.** The apparent terminal half-life
  implied by the published parameters is long (about 62 h for the base model,
  36 h for Model V), so dosing up from zero would not reach the trough the
  paper observed within a tractable simulation horizon.
* **`V/F` is large relative to the paediatric VPA literature** (27.8 L for a
  cohort with a 19 kg median weight, about 1.5 L/kg, against a literature
  value near 0.2 L/kg). This is faithful to the source: the authors state that
  because their data were "mostly at trough concentrations ... parameters for
  the absorption and distribution stages could not be obtained precisely."
  The parameter is reproduced as published and not adjusted.
* **`DD` in Equation 7 is implemented on the mg/kg/day scale.** The paper's
  abbreviation list defines `DD` as daily dose in mg/day and `DDW` as daily
  dose in mg/kg/day, and Equation 7 is printed with `DD`. However the equation
  reads `(DD/25)`, and 25 matches the cohort mean daily dose of
  24.50 **mg/kg/day** (Table 2), not any mg/day quantity; and the reported
  `CLp/F` of 0.331 L/h reproduces the base-model `CL/F` of 0.311 L/h only on
  the mg/kg/day reading. On the mg/day reading, typical clearance comes out at
  2.42 L/h and the predicted trough at about 9 mg/L, against an observed
  median of 50.40 mg/L -- an 84% under-prediction that is incompatible with
  the +1.50% MDPE the paper reports for this model (Table 3). The mg/kg/day
  reading is therefore used, and the text's `DD`/`DDW` label is treated as a
  typographical inconsistency. This is recorded in the
  `DOSE_VPA_MGKGD` register entry.
* **No covariates on `CL/F` or `V/F`.** Neither the base model nor Model V
  carries a weight, age, albumin or co-medication effect. This is unusual for a
  paediatric model and is faithful to Supplementary Table S3; weight and
  albumin are recorded in `covariatesDataExcluded` with the reason. Weight
  enters Model V only indirectly, through the per-kilogram normalisation of the
  daily-dose covariate.
* **Model V is not a dosing tool.** The authors argue explicitly that because
  the daily dose is the quantity a TDM model exists to predict, using it as a
  clearance covariate is circular, and they conclude the simple exponent form
  "did not describe the non-linear properties of the VPA PK process" despite
  its good prediction metrics. It is packaged for fidelity to the published
  comparison.
* **Between-subject variability is diagonal.** Zhang 2023 reports no
  `CL/F`-`V/F` correlation for its own models, so no off-diagonal is encoded.
* **No non-paper-derived parameter values.** Every `ini()` value comes from the
  paper's Equation 7 or Supplementary Table S3. The two `Ka` shift parameters
  are log-ratios computed from the three published fixed `Ka` values
  (2.64 / 1.57 / 0.46 1/h); the arithmetic is visible in the model file.

# Deferred models {#deferred-models}

Four of the paper's six original models are not yet packaged. Both blockers are
naming or reporting questions, not modelling questions -- the structures
themselves are fully determined.

**Models I, II and IV (one-binding-site, Langmuir, linear non-saturable).**
These share a structure that is well identified from the paper: a
one-compartment model carried on the *unbound* concentration, with the observed
total concentration recovered as `Cc = Cu + Cb` where `Cb` follows Equation 3, 4
or 6. That reading is confirmed three times over, because the ratio of each
model's `CLp/F` and `V/F` to the base model's recovers the same unbound
fraction that the binding equation itself implies at the steady-state unbound
concentration (about 0.2% for Model I, 9% for Model II, 7% for Model IV).
What is missing is a naming decision: the saturable-binding constants
(`Bm` maximum binding-site concentration, `Kd` dissociation constant, `NS`
non-saturable slope, `N` binding sites per albumin unit, `K` association
constant) have no canonical names in the library's parameter register, which
records only a time-varying-unbound-fraction slope (`lbfu`) and
Michaelis-Menten enzyme constants. Introducing a new canonical family requires
operator ratification, so these three models await that decision.

**Model III (dose-dependent maximum effect).** Equation 5 needs `Emax`, the
Hill coefficient, and `DD50`. The paper fixes `Emax = 2.8` and the Hill
coefficient at 1.68 from Ding 2015, but **never reports `DD50`** -- it is
absent from Supplementary Table S3 (whose `CL_DD` row is "/" for this model)
and from the text. The value cannot be recovered from what is on disk: Ding
2015's own `DD50` of 37.4 mg/kg/day, tabulated in Zhang's Table 1, gives a
typical clearance of 0.157 L/h and a predicted trough near 137 mg/L, which
contradicts the -1.10% MDPE reported in Table 3; and digitising the typical
curve in Supplementary Figure S5A yields mutually inconsistent `DD50` values
across the dose range under every unit reading. The figure and Table 3 are
themselves in tension for this model -- Figure S5A shows a typical clearance
near 0.15-0.19 L/h, roughly half the base model's 0.311 L/h, which would imply
a strongly positive prediction error rather than the near-zero one reported.
Per the library's sourcing rules no value is guessed, so this model awaits an
operator decision.
