---
title: "Aztreonam + amoxicillin/clavulanate mutant-selection PK/PD (Zhang 2023)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Aztreonam + amoxicillin/clavulanate mutant-selection PK/PD (Zhang 2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Zhang 2023 combined in vitro susceptibility testing of NDM- and
serine-beta-lactamase co-producing *Escherichia coli* and *Klebsiella pneumoniae*
with population PK simulation, to ask whether clinical aztreonam and
amoxicillin/clavulanate regimens keep free drug concentrations above the mutant
prevention concentration (MPC) and out of the mutant selection window (MSW).

The PK layer is three independent two-compartment intravenous models, one per
drug, each transcribed from an upstream publication and used as a fixed
simulation prior. They are packaged separately because the authors used them
separately: nothing is jointly fitted and no parameter is shared.

```{r models}
mdlNames <- c(
  "Zhang_2023_aztreonam",
  "Zhang_2023_amoxicillin",
  "Zhang_2023_clavulanicAcid"
)
mAtm <- rxode2::rxode(readModelDb("Zhang_2023_aztreonam"))
mAmx <- rxode2::rxode(readModelDb("Zhang_2023_amoxicillin"))
mCla <- rxode2::rxode(readModelDb("Zhang_2023_clavulanicAcid"))
```

* Citation: `r mAtm$reference`
* Article: <https://doi.org/10.3390/pharmaceutics15010251>
* Supplement: <https://www.mdpi.com/article/10.3390/pharmaceutics15010251/s1>

```{r descriptions, echo = FALSE}
tibble::tibble(
  Model = mdlNames,
  Description = vapply(
    list(mAtm, mAmx, mCla), function(m) m$description, character(1)
  )
) |>
  knitr::kable()
```

# Population

There is no enrolled PK cohort. Zhang 2023 simulated 10,000 virtual subjects
(Section 2.4 and the Supplementary Materials script): sex in equal proportion,
height drawn from US anthropometric reference distributions, body weight derived
from height, and creatinine clearance drawn from a uniform distribution over
10 to 150 mL/min normalised to 1.73 m2. The lower bound of 10 mL/min excludes
patients on haemodialysis. The parameter estimates themselves come from the
upstream analyses cited in each model's `reference` field: an aztreonam analysis
in patients with normal and impaired renal function, and an
amoxicillin/clavulanic acid analysis in critically ill patients.

The same information is available programmatically:

```{r population}
readModelDb("Zhang_2023_aztreonam")()$population[c("species", "n_subjects", "renal_function")]
```

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()` entry
in `inst/modeldb/specificDrugs/Zhang_2023_*.R`. Collected here for review.
Section references are to the main text; "Suppl." is the Supplementary
Materials PDF.

| Equation / parameter | Value | Source location |
|---|---|---|
| Two-compartment IV ODE system | n/a | Suppl., "Detailed description of pharmacokinetic models" |
| Aztreonam `cl` at CRCL 100 | 4.93 L/h | Suppl. aztreonam equation and both RxODE code blocks (main text Section 2.4 prints 4.73; see Errata) |
| Aztreonam `e_crcl_cl` | 0.43 | Section 2.4; Suppl. aztreonam equation |
| Aztreonam `vc` at 70 kg | 7.43 L | Section 2.4; Suppl. aztreonam equation |
| Aztreonam `e_wt_vc` | 1.99 | Section 2.4; Suppl. aztreonam equation |
| Aztreonam `vp` | 6.44 L | Section 2.4; Suppl. aztreonam equation |
| Aztreonam `q` | 9.26 L/h | Section 2.4; Suppl. aztreonam equation |
| Aztreonam IIV on `cl`, `vc`, `vp` | 0.241, 0.509, 0.277 (log SD) | Section 2.4 (24.1%, 50.9%, 27.7% CV); Suppl. `rnorm(sd = )` arguments |
| Aztreonam `fu` | 0.44 | Section 2.4 ("plasma protein binding of aztreonam is 56%"); Suppl. `res$free <- res$value*(1-0.56)` |
| Aztreonam `relf` | 0.40 | Section 3.4 ("A 40% penetration into the ELF was assumed"); see Errata for the conflict with Section 2.4 |
| Amoxicillin `cl` at CRCL 102 | 10.3 L/h, linear in CRCL | Section 2.4; Suppl. amoxicillin equation |
| Amoxicillin `vc`, `vp`, `q` | 13.5 L, 14.1 L, 15.7 L/h | Section 2.4; Suppl. amoxicillin equation |
| Amoxicillin IIV on `cl`, `vc` | 0.399, 0.387 (log SD) | Section 2.4 (39.9%, 38.7% CV); Suppl. amoxicillin equation |
| Amoxicillin `fu` | 0.83 | Section 2.4 ("Amoxicillin plasma protein binding is 17%") |
| Clavulanic acid `cl` at CRCL 102 | 6.8 L/h, linear in CRCL | Section 2.4; Suppl. clavulanate equation |
| Clavulanic acid `vc`, `vp`, `q` | 7.6 L, 11.6 L, 10.4 L/h | Section 2.4; Suppl. clavulanate equation |
| Clavulanic acid IIV on `cl`, `vc` | 0.578, 0.347 (log SD) | Section 2.4 (57.8%, 34.7% CV); Suppl. clavulanate equation |
| Clavulanic acid `fu` | 0.75 | Section 2.4 ("Clavulanate is approximately 25% protein bound"), so 1 - 0.25 |
| Amoxicillin / clavulanate `relf` | 0.30 | Sections 2.4 and 3.4 |
| Dosing regimens by renal category | see below | Table 1; Suppl. `add.dosing` calls |
| Virtual population height and weight | see below | Suppl., "Detailed description of pharmacokinetic models" |
| Combination MIC per isolate | see below | Table 2, ATM/AMC/CA column |
| MPC per isolate | see below | Table 3, ATM/AMC/CA column |
| Plasma `fTMSW` and `fT>MPC` targets | see below | Table 4 |
| ELF `fTMSW` and `fT>MPC` targets | see below | Table 5 |
| PD index targets | 60% and 50% `fT>MIC` | Section 2.6 |

# Virtual cohort

Original observed data are not publicly available; the paper's own PK input was
a simulated cohort, which this vignette reconstructs from the Supplementary
Materials equations. Cohorts are capped at 200 subjects per arm.

```{r cohort-helpers}
NPER <- 200L
# Steady state is evaluated over the final 24 h of a 120 h dosing run. The
# slowest subjects (low CRCL, high Vc) have a terminal half-life near 6 h, so
# 96 h of dosing is more than 15 half-lives before the evaluation window.
T0 <- 96
T1 <- 120

# Suppl. virtual-population weight model. Heights are drawn from the sex-specific
# US reference distributions, then converted to weight; the /100 cm-to-m
# conversion appears only in the Suppl. RxODE script, not in its typeset equation.
simWeight <- function(n) {
  sexf <- round(stats::runif(n, 0, 1))
  htm <- round(stats::rnorm(n, 176.3, 0.17 * sqrt(4482)), 1)
  htf <- round(stats::rnorm(n, 162.2, 0.16 * sqrt(4857)), 1)
  wtm <- round(exp(3.28 + 1.92 * log(htm / 100)) * exp(stats::rnorm(n, 0, 0.14)), 1)
  wtf <- round(exp(3.49 + 1.45 * log(htf / 100)) * exp(stats::rnorm(n, 0, 0.17)), 1)
  ifelse(sexf == 1, wtf, wtm)
}

# Suppl. sampling grid (0.1 h for the first part of each interval, 0.25 h for the
# last 2 h), replicated across the final 24 h only.
obsGrid <- function(ii, t0 = T0) {
  t1 <- t0 + 24
  off <- c(seq(0, ii - 2, by = 0.1), seq(ii - 1.75, ii - 0.25, by = 0.25))
  sort(unique(c(as.vector(outer(off, seq(t0, t1 - ii, by = ii), "+")), t1)))
}

# One arm as a self-contained event table. `idOffset` keeps subject IDs disjoint
# across arms, which rxSolve requires: duplicate IDs are silently merged.
# `t0` opens the 24 h evaluation window; it is only moved off T0 by the arm-3
# diagnostic below.
makeArm <- function(n, crclLo, crclHi, ldAmt, mdAmt, ii, arm, idOffset, t0 = T0) {
  ev <- rxode2::et(amt = ldAmt, dur = 3, time = 0)
  ev <- rxode2::et(ev, amt = mdAmt, dur = 3, time = ii,
                   ii = ii, addl = ceiling((t0 + 24 - ii) / ii))
  ev <- rxode2::et(ev, obsGrid(ii, t0))
  base <- as.data.frame(ev)
  base$.row <- seq_len(nrow(base))
  subj <- data.frame(
    id = idOffset + seq_len(n),
    CRCL = round(stats::runif(n, crclLo, crclHi)),
    WT = simWeight(n),
    arm = arm,
    stringsAsFactors = FALSE
  )
  out <- merge(subj, base, by = NULL)
  out <- out[order(out$id, out$.row), setdiff(names(out), ".row")]
  rownames(out) <- NULL
  out
}

# rxSolve on an rxUi is quadratic in the number of subjects per call, so each arm
# is solved separately rather than binding all arms into one event table.
solveArm <- function(mod, ev) {
  as.data.frame(
    rxode2::rxSolve(mod, ev, keep = c("arm", "CRCL", "WT"), addDosing = FALSE)
  )
}
```

Table 1 dosing regimens, all as 3 h infusions with the same interval for both
drugs. The 2 g aztreonam and 1 g amoxicillin loading doses do not affect
steady-state metrics and are carried only for fidelity.

```{r regimens}
armLabels <- c("CrCL >50-150", "CrCL >30-50", "CrCL 10-30")
regimens <- tibble::tribble(
  ~arm,             ~crclLo, ~crclHi, ~atmLd, ~atmMd, ~atmIi, ~amxLd, ~amxMd, ~claLd, ~claMd, ~amcIi,
  armLabels[1],     51,      150,     2000,   1500,   6,      1000,   1000,   200,    200,    6,
  armLabels[2],     31,      50,      2000,    750,   6,      1000,   1000,   200,    200,    6,
  armLabels[3],     10,      30,      2000,    500,   8,      1000,    500,   200,    100,    12
)
knitr::kable(regimens, caption = "Regimens transcribed from Table 1 (doses in mg, intervals in h).")
```

# Simulation

```{r simulate}
# Seed immediately before this stochastic block. set.seed() governs the R-level
# covariate draws; rxSetSeed() governs rxode2's eta draws. Both are needed.
set.seed(20230111)
rxode2::rxSetSeed(20230111)

simAtm <- list()
simAmx <- list()
simCla <- list()
for (i in seq_len(nrow(regimens))) {
  r <- regimens[i, ]
  off <- 1000L * i
  simAtm[[r$arm]] <- solveArm(
    mAtm, makeArm(NPER, r$crclLo, r$crclHi, r$atmLd, r$atmMd, r$atmIi, r$arm, off)
  )
  simAmx[[r$arm]] <- solveArm(
    mAmx, makeArm(NPER, r$crclLo, r$crclHi, r$amxLd, r$amxMd, r$amcIi, r$arm, off + 100L)
  )
  simCla[[r$arm]] <- solveArm(
    mCla, makeArm(NPER, r$crclLo, r$crclHi, r$claLd, r$claMd, r$amcIi, r$arm, off + 200L)
  )
}
stopifnot(
  vapply(simAtm, function(d) length(unique(d$id)), integer(1)) == NPER,
  vapply(simAmx, function(d) length(unique(d$id)), integer(1)) == NPER,
  vapply(simCla, function(d) length(unique(d$id)), integer(1)) == NPER
)
```

## Pharmacodynamic metric definitions

`fT>MIC` and `fT>MPC` are the fractions of the 24 h window in which the free
drug concentration exceeds the MIC and the MPC. `fTMSW`, the time spent in the
mutant selection window, is their difference (Section 2.5 and Suppl.,
"Derivation of pharmacodynamic parameters"). Crossing times are found by linear
interpolation, as the Supplement specifies.

```{r pd-metrics}
fracAbove <- function(tt, cc, thr) {
  above <- cc > thr
  if (!any(above)) return(0)
  if (all(above)) return(1)
  n <- length(tt)
  dt <- diff(tt)
  a <- above[-n]
  b <- above[-1]
  full <- sum(dt[a & b])
  part <- 0
  idx <- which(xor(a, b))
  if (length(idx)) {
    f <- (thr - cc[idx]) / (cc[idx + 1L] - cc[idx])
    part <- sum(ifelse(a[idx], f, 1 - f) * dt[idx])
  }
  (full + part) / (tt[n] - tt[1])
}

# Per-subject fT above a threshold on the named concentration column.
ftPerSubject <- function(sim, concCol, thr, t0 = T0) {
  s <- sim[sim$time >= t0, ]
  vapply(split(s, s$id), function(d) {
    o <- order(d$time)
    fracAbove(d$time[o], d[[concCol]][o], thr)
  }, numeric(1))
}

# Mean fT>MPC and fTMSW over the cohort, in percent.
pdSummary <- function(sim, concCol, mic, mpc, t0 = T0) {
  ftMic <- ftPerSubject(sim, concCol, mic, t0)
  ftMpc <- ftPerSubject(sim, concCol, mpc, t0)
  c(ftmpc = 100 * mean(ftMpc), ftmsw = 100 * mean(ftMic - ftMpc))
}
```

## In vitro susceptibility of the seven isolates

Combination MIC values come from the ATM/AMC/CA column of Table 2 and MPC values
from the same column of Table 3. Isolates EC13 and EC48 share both an MIC of
4/4/2 and an MPC of 8/8/4, which is why the paper reports them as one row.

```{r isolates}
iso <- tibble::tribble(
  ~isolate,     ~micAtm, ~mpcAtm, ~micAmc, ~mpcAmc,
  "EC13, EC48",  4.00,    8,       4,       8,
  "LW13",        0.25,    4,       2,       4,
  "EC37",        0.25,    2,       1,       2,
  "LW8",         0.50,    1,       1,       1,
  "EC45",        0.25,    1,       1,       1
)
knitr::kable(iso, caption = "Combination MIC (Table 2) and MPC (Table 3) in mg/L.")
```

Two internal consistency checks on that transcription. LW8 and EC45 have an
amoxicillin MIC equal to their MPC, so their mutant selection window is empty
and `fTMSW` must be identically zero, which is what Tables 4 and 5 print. EC45
has a narrower aztreonam MIC-to-MPC gap than LW8 in the reverse direction, so
its aztreonam `fTMSW` must be slightly the larger of the two, which is also what
Table 4 prints (0.04% versus 0.03%).

```{r isolate-checks}
stopifnot(
  iso$micAmc[iso$isolate %in% c("LW8", "EC45")] ==
    iso$mpcAmc[iso$isolate %in% c("LW8", "EC45")],
  iso$micAtm[iso$isolate == "EC45"] < iso$micAtm[iso$isolate == "LW8"],
  iso$mpcAtm[iso$isolate == "EC45"] == iso$mpcAtm[iso$isolate == "LW8"]
)
```

# Replicate published figures

The Supplementary Materials plot free aztreonam concentration over time against
the 8 mg/L breakpoint for the high-renal-function arm. This reproduces it.

```{r figure-suppl}
simAtm[[armLabels[1]]] |>
  group_by(time) |>
  summarise(
    q025 = quantile(Ccfree, 0.025),
    q500 = quantile(Ccfree, 0.500),
    q975 = quantile(Ccfree, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, q500)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), fill = "grey70") +
  geom_line() +
  geom_hline(yintercept = 8, linetype = 2) +
  annotate("text", x = T0 + 12, y = 9.6, label = "Aztreonam breakpoint: 8 mg/L") +
  scale_y_log10(breaks = c(0.125, 0.5, 2, 8, 32, 128)) +
  coord_cartesian(ylim = c(0.1, 200)) +
  labs(
    x = "Time (h)", y = "Free aztreonam (mg/L)",
    title = "Aztreonam 2 g load then 1.5 g q6h, CrCL >50-150 mL/min",
    caption = "Reproduces the Supplementary Materials free-aztreonam figure (median and 95% interval)."
  ) +
  theme_bw()
```

Free steady-state profiles for all three drugs and all three renal arms.

```{r figure-all-drugs, fig.height = 6}
bind_rows(
  bind_rows(simAtm) |> mutate(drug = "Aztreonam"),
  bind_rows(simAmx) |> mutate(drug = "Amoxicillin"),
  bind_rows(simCla) |> mutate(drug = "Clavulanic acid")
) |>
  mutate(
    arm = factor(arm, levels = armLabels),
    drug = factor(drug, levels = c("Aztreonam", "Amoxicillin", "Clavulanic acid"))
  ) |>
  group_by(drug, arm, time) |>
  summarise(
    q025 = quantile(Ccfree, 0.025),
    q500 = quantile(Ccfree, 0.500),
    q975 = quantile(Ccfree, 0.975),
    .groups = "drop"
  ) |>
  ggplot(aes(time, q500)) +
  geom_ribbon(aes(ymin = q025, ymax = q975), fill = "grey75") +
  geom_line() +
  facet_grid(drug ~ arm) +
  scale_y_log10() +
  labs(
    x = "Time (h)", y = "Free concentration (mg/L)",
    title = "Steady-state free concentrations by drug and renal function category"
  ) +
  theme_bw()
```

# Validation against Tables 4 and 5

Tables 4 and 5 are a zero-parameter gate: every input (parameters, regimens,
covariate distributions, MIC, MPC) is fixed by the paper, so the 60 reported
cell means are predicted with no freedom to tune. The plasma metrics run on
`Ccfree` and the ELF metrics on `Celf`.

```{r gate-tables}
paperTab <- tibble::tribble(
  ~arm,         ~drug,         ~matrix,  ~isolate,     ~ftmpc, ~ftmsw,
  armLabels[1], "Aztreonam",   "Plasma", "EC13, EC48",  96.3,  3.20,
  armLabels[1], "Aztreonam",   "Plasma", "LW13",        99.6,  0.40,
  armLabels[1], "Aztreonam",   "Plasma", "EC37",        99.9,  0.12,
  armLabels[1], "Aztreonam",   "Plasma", "LW8",         99.9,  0.03,
  armLabels[1], "Aztreonam",   "Plasma", "EC45",        99.9,  0.04,
  armLabels[2], "Aztreonam",   "Plasma", "EC13, EC48",  95.8,  4.00,
  armLabels[2], "Aztreonam",   "Plasma", "LW13",        99.8,  0.20,
  armLabels[2], "Aztreonam",   "Plasma", "EC37",        99.9,  0.10,
  armLabels[2], "Aztreonam",   "Plasma", "LW8",         99.9,  0.04,
  armLabels[2], "Aztreonam",   "Plasma", "EC45",       100.0,  0.02,
  armLabels[3], "Aztreonam",   "Plasma", "EC13, EC48",  92.2,  7.20,
  armLabels[3], "Aztreonam",   "Plasma", "LW13",        99.3,  0.60,
  armLabels[3], "Aztreonam",   "Plasma", "EC37",        99.8,  0.10,
  armLabels[3], "Aztreonam",   "Plasma", "LW8",         99.9,  0.03,
  armLabels[3], "Aztreonam",   "Plasma", "EC45",        99.9,  0.05,
  armLabels[1], "Aztreonam",   "ELF",    "EC13, EC48",  94.2,  4.30,
  armLabels[1], "Aztreonam",   "ELF",    "LW13",        99.4,  0.60,
  armLabels[1], "Aztreonam",   "ELF",    "EC37",        99.8,  0.14,
  armLabels[1], "Aztreonam",   "ELF",    "LW8",         99.9,  0.03,
  armLabels[1], "Aztreonam",   "ELF",    "EC45",        99.9,  0.04,
  armLabels[2], "Aztreonam",   "ELF",    "EC13, EC48",  93.9,  5.80,
  armLabels[2], "Aztreonam",   "ELF",    "LW13",        99.7,  0.30,
  armLabels[2], "Aztreonam",   "ELF",    "EC37",        99.8,  0.10,
  armLabels[2], "Aztreonam",   "ELF",    "LW8",         99.9,  0.03,
  armLabels[2], "Aztreonam",   "ELF",    "EC45",        99.9,  0.04,
  armLabels[3], "Aztreonam",   "ELF",    "EC13, EC48",  89.7,  9.50,
  armLabels[3], "Aztreonam",   "ELF",    "LW13",        99.1,  0.80,
  armLabels[3], "Aztreonam",   "ELF",    "EC37",        99.8,  0.20,
  armLabels[3], "Aztreonam",   "ELF",    "LW8",         99.9,  0.04,
  armLabels[3], "Aztreonam",   "ELF",    "EC45",        99.9,  0.05,
  armLabels[1], "Amoxicillin", "Plasma", "EC13, EC48",  75.8, 14.40,
  armLabels[1], "Amoxicillin", "Plasma", "LW13",        90.0,  5.80,
  armLabels[1], "Amoxicillin", "Plasma", "EC37",        95.6,  2.60,
  armLabels[1], "Amoxicillin", "Plasma", "LW8",         98.2,  0.00,
  armLabels[1], "Amoxicillin", "Plasma", "EC45",        98.2,  0.00,
  armLabels[2], "Amoxicillin", "Plasma", "EC13, EC48",  96.9,  2.40,
  armLabels[2], "Amoxicillin", "Plasma", "LW13",        99.4,  0.50,
  armLabels[2], "Amoxicillin", "Plasma", "EC37",        99.7,  0.20,
  armLabels[2], "Amoxicillin", "Plasma", "LW8",         99.9,  0.00,
  armLabels[2], "Amoxicillin", "Plasma", "EC45",        99.9,  0.00,
  armLabels[3], "Amoxicillin", "Plasma", "EC13, EC48",  98.8,  1.00,
  armLabels[3], "Amoxicillin", "Plasma", "LW13",        99.8,  0.10,
  armLabels[3], "Amoxicillin", "Plasma", "EC37",        99.8,  0.10,
  armLabels[3], "Amoxicillin", "Plasma", "LW8",         99.9,  0.00,
  armLabels[3], "Amoxicillin", "Plasma", "EC45",        99.9,  0.00,
  armLabels[1], "Amoxicillin", "ELF",    "EC13, EC48",  28.2, 35.00,
  armLabels[1], "Amoxicillin", "ELF",    "LW13",        63.2, 22.00,
  armLabels[1], "Amoxicillin", "ELF",    "EC37",        84.5,  8.90,
  armLabels[1], "Amoxicillin", "ELF",    "LW8",         93.5,  0.00,
  armLabels[1], "Amoxicillin", "ELF",    "EC45",        93.5,  0.00,
  armLabels[2], "Amoxicillin", "ELF",    "EC13, EC48",  74.3, 19.30,
  armLabels[2], "Amoxicillin", "ELF",    "LW13",        93.6,  4.90,
  armLabels[2], "Amoxicillin", "ELF",    "EC37",        98.0,  1.40,
  armLabels[2], "Amoxicillin", "ELF",    "LW8",         99.4,  0.00,
  armLabels[2], "Amoxicillin", "ELF",    "EC45",        99.4,  0.00,
  armLabels[3], "Amoxicillin", "ELF",    "EC13, EC48",  82.5, 14.30,
  armLabels[3], "Amoxicillin", "ELF",    "LW13",        96.8,  2.70,
  armLabels[3], "Amoxicillin", "ELF",    "EC37",        99.1,  0.60,
  armLabels[3], "Amoxicillin", "ELF",    "LW8",         99.8,  0.00,
  armLabels[3], "Amoxicillin", "ELF",    "EC45",        99.8,  0.00
)

simFor <- function(drug, arm) if (drug == "Aztreonam") simAtm[[arm]] else simAmx[[arm]]
micFor <- function(drug, isolate) {
  r <- iso[iso$isolate == isolate, ]
  if (drug == "Aztreonam") c(r$micAtm, r$mpcAtm) else c(r$micAmc, r$mpcAmc)
}

gate <- paperTab
gate$simMpc <- NA_real_
gate$simMsw <- NA_real_
for (i in seq_len(nrow(gate))) {
  th <- micFor(gate$drug[i], gate$isolate[i])
  col <- if (gate$matrix[i] == "Plasma") "Ccfree" else "Celf"
  v <- pdSummary(simFor(gate$drug[i], gate$arm[i]), col, th[1], th[2])
  gate$simMpc[i] <- v[["ftmpc"]]
  gate$simMsw[i] <- v[["ftmsw"]]
}
gate <- gate |>
  mutate(dMpc = simMpc - ftmpc, dMsw = simMsw - ftmsw)
```

```{r gate-table-show, echo = FALSE}
gate |>
  mutate(across(c(simMpc, simMsw, dMpc, dMsw), \(x) round(x, 2))) |>
  select(
    Arm = arm, Drug = drug, Matrix = matrix, Isolate = isolate,
    `fT>MPC paper` = ftmpc, `fT>MPC sim` = simMpc, `fT>MPC diff` = dMpc,
    `fTMSW paper` = ftmsw, `fTMSW sim` = simMsw, `fTMSW diff` = dMsw
  ) |>
  knitr::kable(
    caption = paste(
      "Reproduction of Tables 4 (plasma) and 5 (ELF).",
      "Differences are simulated minus published, in percentage points."
    )
  )
```

Worst-case agreement per drug, matrix and renal arm:

```{r gate-summary, echo = FALSE}
gate |>
  mutate(arm = factor(arm, levels = armLabels)) |>
  group_by(Arm = arm, Drug = drug, Matrix = matrix) |>
  summarise(
    `Max |fT>MPC diff|` = round(max(abs(dMpc)), 2),
    `Max |fTMSW diff|` = round(max(abs(dMsw)), 2),
    `Mean fT>MPC diff` = round(mean(dMpc), 2),
    .groups = "drop"
  ) |>
  arrange(Arm, Drug, Matrix) |>
  knitr::kable()
```

The two higher renal-function arms reproduce. The CrCL 10-30 arm does not, for
either drug and in both matrices, under the regimen Table 1 prints. That arm is
therefore excluded from the gate and analysed separately below; nothing was
adjusted to close it.

```{r gate-assertions}
gateArms <- armLabels[1:2]
worst <- gate |>
  filter(arm %in% gateArms) |>
  group_by(drug, matrix) |>
  summarise(mpc = max(abs(dMpc)), msw = max(abs(dMsw)), .groups = "drop")
getWorst <- function(d, m, what) worst[[what]][worst$drug == d & worst$matrix == m]

# Bounds are set well outside the range observed while authoring, so that a
# different cohort draw still passes: rxode2 partitions its RNG per solver
# thread, so CI does not reproduce this machine's cohort and no seed can make it
# do so. Observed maxima at the time of writing, at 200 subjects per arm and
# again at 1500 to gauge Monte-Carlo noise, are quoted per line. Do not file
# these down to the observed values.
stopifnot(
  # Aztreonam reproduces both tables essentially exactly: the model, the
  # regimens, the weight and CRCL distributions and fu are all pinned.
  getWorst("Aztreonam", "Plasma", "mpc") < 4,    # observed 1.41 over the gated arms
  getWorst("Aztreonam", "Plasma", "msw") < 4,    # observed 1.42
  getWorst("Aztreonam", "ELF", "mpc") < 4,       # observed 1.60
  getWorst("Aztreonam", "ELF", "msw") < 4,       # observed 1.62
  # Amoxicillin carries a systematic offset of a few percentage points; see the
  # Errata note on the Table 4 amoxicillin block.
  getWorst("Amoxicillin", "Plasma", "mpc") < 8,  # observed 3.37
  getWorst("Amoxicillin", "Plasma", "msw") < 8,
  getWorst("Amoxicillin", "ELF", "mpc") < 14,    # observed 8.25
  getWorst("Amoxicillin", "ELF", "msw") < 14
)
```

## Known deviation: the CrCL 10-30 arm

Under Table 1 as printed, evaluated at steady state like the other two arms,
simulated attainment in the CrCL 10-30 arm falls well short of Tables 4 and 5:
aztreonam `fT>MPC` for EC13/EC48 comes out near 79% against a published 92.2%,
and amoxicillin ELF `fT>MPC` near 28% against a published 82.5%. The shortfall is
in the direction of too little drug, which cannot be a clearance error, because
the same models reproduce the two higher-CrCL arms.

The paper describes this arm inconsistently. Section 2.4 states that "the same
dosing interval was used for both aztreonam and amoxicillin/clavulanate", yet
Table 1 gives aztreonam q8h and amoxicillin/clavulanate q12h in this row alone;
Section 2.5 states that profiles were simulated at steady state; and the
Supplementary Materials script only ever spells out the CrCL >50-150 regimen.
The chunk below searches the readings those statements admit.

```{r arm3-alternatives}
set.seed(20231030)
rxode2::rxSetSeed(20231030)

# Two candidate readings, each evaluated over its own 24 h window.
altAtmFirst24 <- solveArm(mAtm, makeArm(NPER, 10, 30, 2000, 500, 8, "alt", 7000L, t0 = 0))
altAmx1gSs <- solveArm(mAmx, makeArm(NPER, 10, 30, 1000, 1000, 12, "alt", 7100L))

arm3 <- gate |> filter(arm == armLabels[3], isolate %in% c("EC13, EC48", "LW13"))
alt <- bind_rows(
  arm3 |>
    filter(drug == "Aztreonam") |>
    rowwise() |>
    mutate(reading = "500 mg q8h, first 24 h incl. 2 g load", altMpc = pdSummary(
      altAtmFirst24, if (matrix == "Plasma") "Ccfree" else "Celf",
      micFor("Aztreonam", isolate)[1], micFor("Aztreonam", isolate)[2], t0 = 0
    )[["ftmpc"]]) |>
    ungroup(),
  arm3 |>
    filter(drug == "Amoxicillin") |>
    rowwise() |>
    mutate(reading = "1000 mg q12h, steady state", altMpc = pdSummary(
      altAmx1gSs, if (matrix == "Plasma") "Ccfree" else "Celf",
      micFor("Amoxicillin", isolate)[1], micFor("Amoxicillin", isolate)[2]
    )[["ftmpc"]]) |>
    ungroup()
) |>
  mutate(dAlt = altMpc - ftmpc)
```

```{r arm3-alternatives-show, echo = FALSE}
alt |>
  mutate(across(c(simMpc, dMpc, altMpc, dAlt), \(x) round(x, 2))) |>
  select(
    Drug = drug, Matrix = matrix, Isolate = isolate, `fT>MPC paper` = ftmpc,
    `Table 1 as printed` = simMpc, `diff` = dMpc,
    `Alternative reading` = reading, `alt fT>MPC` = altMpc, `alt diff` = dAlt
  ) |>
  knitr::kable(
    caption = paste(
      "CrCL 10-30 arm: Table 1 as printed against the reading that reproduces",
      "the published column."
    )
  )
```

```{r arm3-assertions}
# Each alternative reproduces its own drug's published column, which localises
# the disagreement to the paper's simulation setup for this arm rather than to
# the PK models. The two readings are mutually inconsistent -- one is a
# first-dose window, the other a steady-state window -- so neither is adopted;
# the vignette keeps Table 1 as printed and records the deviation.
stopifnot(
  max(abs(alt$dAlt)) < 5,          # observed 1.14 at n=400 in the extraction work
  max(abs(alt$dMpc)) > 5           # the printed reading really is far off
)
```

Because the two reproducing readings disagree about the evaluation window, the
arm-3 column of Tables 4 and 5 cannot be attributed to a single self-consistent
setup, and no change was made to the models or to the transcribed regimens.

## Adjudicating the ELF encoding

Section 2.4 states ELF penetration rates of 55% for aztreonam and 30% for
amoxicillin/clavulanate; Section 3.4 states 40% for aztreonam. Neither section
says whether the rate multiplies the total or the unbound plasma concentration.
Table 5 settles both questions, because only one of the four readings reproduces
it. The discriminating isolate is EC13/EC48, the only one whose aztreonam mutant
selection window is wide enough for the choice to matter.

Because `Celf` is a constant multiple of `Cc`, each candidate reading is
evaluated by rescaling the threshold rather than re-solving.

```{r elf-adjudicate}
elfCandidates <- tibble::tribble(
  ~reading,                ~factor,
  "0.40 x total plasma",   0.40,
  "0.55 x total plasma",   0.55,
  "0.40 x free plasma",    0.40 * 0.44,
  "0.55 x free plasma",    0.55 * 0.44
)
target <- paperTab |>
  filter(drug == "Aztreonam", matrix == "ELF", isolate == "EC13, EC48",
         arm == armLabels[1])

elfCandidates <- elfCandidates |>
  rowwise() |>
  mutate(
    simMpc = pdSummary(simAtm[[armLabels[1]]], "Cc", 4 / factor, 8 / factor)[["ftmpc"]],
    simMsw = pdSummary(simAtm[[armLabels[1]]], "Cc", 4 / factor, 8 / factor)[["ftmsw"]]
  ) |>
  ungroup() |>
  mutate(errMpc = abs(simMpc - target$ftmpc), errMsw = abs(simMsw - target$ftmsw))
```

```{r elf-adjudicate-show, echo = FALSE}
elfCandidates |>
  mutate(across(c(simMpc, simMsw, errMpc, errMsw), \(x) round(x, 2))) |>
  select(
    Reading = reading,
    `fT>MPC sim` = simMpc, `fT>MPC err` = errMpc,
    `fTMSW sim` = simMsw, `fTMSW err` = errMsw
  ) |>
  knitr::kable(
    caption = paste0(
      "Candidate ELF readings against Table 5 for EC13/EC48, CrCL >50-150 ",
      "(published fT>MPC ", target$ftmpc, "%, fTMSW ", target$ftmsw, "%)."
    )
  )
```

```{r elf-assertions}
errOf <- function(r) elfCandidates$errMpc[elfCandidates$reading == r]

stopifnot(
  # The chosen reading reproduces the published cell.
  errOf("0.40 x total plasma") < 3,        # observed 0.08 at n=200, 1.04 at n=1500
  # Applying the rate to the FREE concentration is decisively wrong: it drops
  # fT>MPC by more than 30 points. This is the assertion that gives the gate its
  # discriminating power, so it is checked rather than merely displayed.
  errOf("0.40 x free plasma") > 15,        # observed 31.6 at n=1500
  # The Section 2.4 rate of 55% overshoots. The margin here is smaller than
  # above because fT>MPC saturates near 100%, so this is asserted only as a
  # directional overshoot rather than with a tight bound.
  elfCandidates$simMpc[elfCandidates$reading == "0.55 x total plasma"] -
    target$ftmpc > 1.5                     # observed +3.97 at n=1500
)
```

The conclusion is that `Celf` equals the **total** plasma concentration times
0.40 for aztreonam, and that the resulting ELF concentration is itself treated
as entirely unbound, consistent with Section 2.4's statement that no mucin
binding was assumed. The model files encode exactly that.

## Adjudicating the aztreonam clearance scalar

Section 2.4 prints `CL = 4.73*(CrCL/100)^0.43`. The Supplementary Materials print
4.93, in the typeset equation and in both RxODE code blocks. The published
Table 4 was produced by that code, so a paired comparison on the same cohort
decides which value the tables reflect. Reseeding before each solve gives both
arms common random numbers, so the contrast is almost free of Monte-Carlo noise.

```{r cl-adjudicate}
armCl <- regimens[1, ]
evCl <- local({
  set.seed(4931)
  makeArm(NPER, armCl$crclLo, armCl$crclHi, armCl$atmLd, armCl$atmMd,
          armCl$atmIi, armCl$arm, 9000L)
})

rxode2::rxSetSeed(4931)
simCl493 <- solveArm(mAtm, evCl)
rxode2::rxSetSeed(4931)
simCl473 <- solveArm(rxode2::ini(mAtm, lcl = log(4.73)), evCl)

paperPlasma1 <- paperTab |> filter(drug == "Aztreonam", matrix == "Plasma", arm == armLabels[1])
clCmp <- paperPlasma1 |>
  rowwise() |>
  mutate(
    sim493 = pdSummary(simCl493, "Ccfree", micFor("Aztreonam", isolate)[1],
                       micFor("Aztreonam", isolate)[2])[["ftmpc"]],
    sim473 = pdSummary(simCl473, "Ccfree", micFor("Aztreonam", isolate)[1],
                       micFor("Aztreonam", isolate)[2])[["ftmpc"]]
  ) |>
  ungroup() |>
  mutate(err493 = abs(sim493 - ftmpc), err473 = abs(sim473 - ftmpc))
```

```{r cl-adjudicate-show, echo = FALSE}
clCmp |>
  mutate(across(c(sim493, sim473, err493, err473), \(x) round(x, 2))) |>
  select(
    Isolate = isolate, `fT>MPC paper` = ftmpc,
    `sim CL 4.93` = sim493, `err 4.93` = err493,
    `sim CL 4.73` = sim473, `err 4.73` = err473
  ) |>
  knitr::kable(
    caption = "Aztreonam plasma fT>MPC under the two printed clearance scalars, common random numbers."
  )
```

```{r cl-assertions}
# Under common random numbers the two arms share every eta and covariate draw,
# so the contrast is paired and structural rather than a coin flip: the smaller
# clearance scalar must give every subject at least as much exposure, hence at
# least as much time above the MPC. That monotonicity is the assertion below
# that cannot be defeated by a different cohort.
stopifnot(all(clCmp$sim473 >= clCmp$sim493 - 1e-8))

# Both scalars overshoot the published column slightly, so the smaller one -
# which overshoots more - is the further away on every isolate, and 4.93
# reproduces Table 4 outright. Bounds carry margin over the values observed
# while authoring (max err 0.58 for 4.93, 1.36 for 4.73).
stopifnot(
  max(clCmp$err493) < 3,
  sum(clCmp$err493) < sum(clCmp$err473)
)
```

Both the printing count (three occurrences against one) and this gate support
4.93, which is what the model file encodes; the Section 2.4 value of 4.73 is
recorded as an erratum.

# PKNCA validation

The paper reports no noncompartmental parameters, so there is nothing to compare
against directly. PKNCA is instead used to derive steady-state exposure over one
dosing interval and to check it against the closed-form linear-model identity
`AUCtau = Dose / CL`. Both sides use the same per-subject drawn clearance, so
the only difference is numerical, and a tight bound is appropriate.

```{r pknca}
# All three drugs share a q6h interval in the CrCL >50-150 arm, so one PKNCA
# setup covers them. The final interval is 114 to 120 h, both endpoints present
# in the observation grid, so no time-zero back-extrapolation is requested.
tauStart <- 114
tauEnd <- 120

ncaConc <- bind_rows(
  simAtm[[armLabels[1]]] |> mutate(drug = "Aztreonam", mdAmt = armCl$atmMd),
  simAmx[[armLabels[1]]] |> mutate(drug = "Amoxicillin", mdAmt = armCl$amxMd),
  simCla[[armLabels[1]]] |> mutate(drug = "Clavulanic acid", mdAmt = armCl$claMd)
) |>
  filter(!is.na(Cc), time >= tauStart, time <= tauEnd)

concObj <- PKNCA::PKNCAconc(
  ncaConc |> select(id, time, Cc, drug),
  Cc ~ time | drug + id
)

doseObj <- PKNCA::PKNCAdose(
  ncaConc |>
    distinct(id, drug, mdAmt) |>
    mutate(time = tauStart, route = "intravascular", duration = 3) |>
    rename(amt = mdAmt),
  amt ~ time | drug + id
)

intervals <- data.frame(
  start = tauStart, end = tauEnd,
  auclast = TRUE, cmax = TRUE, tmax = TRUE, cmin = TRUE
)

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(concObj, doseObj, intervals = intervals))
```

```{r pknca-summary, echo = FALSE}
as.data.frame(ncaRes) |>
  filter(PPTESTCD %in% c("cmax", "cmin", "tmax", "auclast")) |>
  group_by(drug, PPTESTCD) |>
  summarise(median = median(PPORRES), .groups = "drop") |>
  pivot_wider(names_from = PPTESTCD, values_from = median) |>
  mutate(across(where(is.numeric), \(x) signif(x, 4))) |>
  rename(
    Drug = drug,
    `AUCtau (mg h/L)` = auclast,
    `Cmax (mg/L)` = cmax,
    `Cmin (mg/L)` = cmin,
    `Tmax (h)` = tmax
  ) |>
  knitr::kable(
    caption = "Median steady-state total-drug NCA over 114 to 120 h, CrCL >50-150 arm."
  )
```

```{r pknca-identity}
clById <- ncaConc |> distinct(drug, id, cl, mdAmt)
identity <- as.data.frame(ncaRes) |>
  filter(PPTESTCD == "auclast") |>
  select(drug, id, auctau = PPORRES) |>
  left_join(clById, by = c("drug", "id")) |>
  mutate(relErr = auctau / (mdAmt / cl) - 1)

cat(sprintf(
  "AUCtau vs Dose/CL: max |relative error| = %.2e, median %.2e, over %d pairs\n",
  max(abs(identity$relErr)), median(abs(identity$relErr)), nrow(identity)
))

# Both sides share the drawn per-subject clearance, so the only discrepancy is
# numerical. It is not integrator error but AUC discretisation: the paper's grid
# spaces the last 2 h of each interval at 0.25 h, and the trapezoidal rule
# overestimates the area under a convex decline there. The size of that residual
# therefore scales with how fast the fastest-declining subject falls, so the
# bound is set from the discretisation scale rather than filed to the observed
# maximum (9.7e-3 at the time of writing, median about 1e-3). It remains a real
# gate: any error in the dose, clearance or volume bookkeeping would move this
# by tens of percent, not by a few tenths of one percent.
stopifnot(
  nrow(identity) == 3L * NPER,
  max(abs(identity$relErr)) < 0.05
)
```

# Probability of target attainment

Section 2.6 sets the PD index target at 60% `fT>MIC` for aztreonam and 50%
`fT>MIC` for amoxicillin/clavulanate, and no target for clavulanate. Sections 3.3
and 3.4 make five quantitative claims about Figures 2 and 3, reproduced below.

```{r pta}
ptaAt <- function(sim, concCol, mic, target) {
  100 * mean(ftPerSubject(sim, concCol, mic) >= target)
}
micGrid <- c(0.25, 0.5, 1, 2, 4, 8, 16, 32)

ptaTab <- bind_rows(lapply(armLabels, function(a) {
  bind_rows(
    tibble(arm = a, drug = "Aztreonam", matrix = "Plasma", mic = micGrid,
           pta = vapply(micGrid, \(m) ptaAt(simAtm[[a]], "Ccfree", m, 0.60), numeric(1))),
    tibble(arm = a, drug = "Aztreonam", matrix = "ELF", mic = micGrid,
           pta = vapply(micGrid, \(m) ptaAt(simAtm[[a]], "Celf", m, 0.60), numeric(1))),
    tibble(arm = a, drug = "Amoxicillin", matrix = "Plasma", mic = micGrid,
           pta = vapply(micGrid, \(m) ptaAt(simAmx[[a]], "Ccfree", m, 0.50), numeric(1))),
    tibble(arm = a, drug = "Amoxicillin", matrix = "ELF", mic = micGrid,
           pta = vapply(micGrid, \(m) ptaAt(simAmx[[a]], "Celf", m, 0.50), numeric(1)))
  )
}))
```

```{r pta-figure, echo = FALSE, fig.height = 5.5}
ptaTab |>
  mutate(arm = factor(arm, levels = armLabels)) |>
  ggplot(aes(mic, pta, colour = arm, shape = arm)) +
  geom_hline(yintercept = 90, linetype = 2) +
  geom_line() +
  geom_point() +
  facet_grid(matrix ~ drug) +
  scale_x_log10(breaks = micGrid) +
  labs(
    x = "MIC (mg/L)", y = "Probability of target attainment (%)",
    colour = "Renal function", shape = "Renal function",
    title = "PTA for 60% fT>MIC (aztreonam) and 50% fT>MIC (amoxicillin)",
    caption = "Reproduces Figure 2 (plasma) and Figure 3 (ELF); dashed line is the 90% criterion."
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r pta-claims}
ptaOf <- function(d, m, a, mic) {
  ptaTab$pta[ptaTab$drug == d & ptaTab$matrix == m & ptaTab$arm == a & ptaTab$mic == mic]
}

# Claims are evaluated only over the two arms that reproduce Tables 4 and 5.
# The CrCL 10-30 arm is excluded for the reason given above, so claims that
# quantify over "every renal arm" are checked over the gated arms and marked.
claims <- tibble::tribble(
  ~claim, ~source, ~value, ~holds, ~gated,
  "Aztreonam plasma: >=90% PTA at MIC 8 mg/L",
  "Section 3.3, Figure 2",
  min(vapply(gateArms, \(a) ptaOf("Aztreonam", "Plasma", a, 8), numeric(1))),
  min(vapply(gateArms, \(a) ptaOf("Aztreonam", "Plasma", a, 8), numeric(1))) >= 90,
  TRUE,

  "Aztreonam ELF: >=90% PTA at MIC 8 mg/L",
  "Section 3.4, Figure 3",
  min(vapply(gateArms, \(a) ptaOf("Aztreonam", "ELF", a, 8), numeric(1))),
  min(vapply(gateArms, \(a) ptaOf("Aztreonam", "ELF", a, 8), numeric(1))) >= 90,
  TRUE,

  "Amoxicillin plasma: about 80% PTA at MIC 8 mg/L when CrCL >50 mL/min",
  "Section 3.3, Figure 2",
  ptaOf("Amoxicillin", "Plasma", armLabels[1], 8),
  abs(ptaOf("Amoxicillin", "Plasma", armLabels[1], 8) - 80) <= 15,
  TRUE,

  "Amoxicillin ELF: >=90% PTA at MIC 2 mg/L when CrCL >50 mL/min",
  "Section 3.4, Figure 3",
  ptaOf("Amoxicillin", "ELF", armLabels[1], 2),
  ptaOf("Amoxicillin", "ELF", armLabels[1], 2) >= 90,
  TRUE,

  "Amoxicillin ELF: >=90% PTA at MIC 4 mg/L when CrCL 30-50 mL/min",
  "Section 3.4, Figure 3",
  ptaOf("Amoxicillin", "ELF", armLabels[2], 4),
  ptaOf("Amoxicillin", "ELF", armLabels[2], 4) >= 90,
  TRUE,

  "Amoxicillin ELF: >=90% PTA at MIC 4 mg/L when CrCL 10-30 mL/min",
  "Section 3.4, Figure 3 (arm excluded from the gate)",
  ptaOf("Amoxicillin", "ELF", armLabels[3], 4),
  ptaOf("Amoxicillin", "ELF", armLabels[3], 4) >= 90,
  FALSE
)
```

```{r pta-claims-show, echo = FALSE}
claims |>
  mutate(value = round(value, 1)) |>
  select(
    Claim = claim, Source = source, `Simulated PTA (%)` = value,
    Reproduced = holds, Gated = gated
  ) |>
  knitr::kable(caption = "Quantitative PTA claims of Sections 3.3 and 3.4.")
```

```{r pta-assertions}
# The 90% criterion is a claim the paper itself states, so it is asserted
# directly rather than against a tolerance filed to an observed value. The
# amoxicillin plasma claim is "approximately 80%", so a 15-point window is used.
# The final row is displayed but not asserted: it depends on the CrCL 10-30 arm,
# which is a documented deviation. Widening a bound until that row passed would
# be tuning, so it is excluded instead.
stopifnot(all(claims$holds[claims$gated]))
```

# Assumptions and deviations

## Errata and internal inconsistencies in the source

* **Aztreonam clearance scalar, 4.73 against 4.93.** Section 2.4 prints
  `CL = 4.73*(CrCL/100)^0.43`. The Supplementary Materials print 4.93 three
  times: in the typeset aztreonam equation, in the sample `sim()` function, and
  in the `theta.aztreonam` block that actually drives the simulation loop. The
  model encodes **4.93**, on the printing count and on the paired gate above,
  which shows 4.93 reproducing Table 4 to within 0.1 percentage points at 1500
  subjects while 4.73 is systematically off. Section 2.4 is treated as a typo.
* **Aztreonam ELF penetration, 55% against 40%.** Section 2.4 gives ELF
  penetration rates of "55% and 30% ... for aztreonam and
  amoxicillin/clavulanate"; Section 3.4 gives "A 40% penetration ... for
  aztreonam". The model encodes **0.40**, because only that value reproduces
  Table 5 (see the adjudication above). The amoxicillin/clavulanate rate of 30%
  is consistent between the two sections and is used as printed.
* **The ELF penetration rate multiplies TOTAL, not unbound, plasma
  concentration.** Neither section states which. Table 5 is only reproducible
  under the total-plasma reading; the free-plasma reading is off by more than 30
  percentage points. The resulting ELF concentration is then treated as entirely
  unbound, which is consistent with Section 2.4's statement that no mucin
  binding was assumed in the ELF. A consequence worth noting is that the
  penetration ratio implicitly absorbs the plasma protein binding, so ELF free
  concentrations sit only slightly below plasma free concentrations for
  aztreonam (0.40 / 0.44 = 0.91).
* **Table 1 unit typo.** The CrCL >30-50 mL/min row prints the aztreonam
  maintenance dose as "750g"; 750 mg is intended, consistent with the surrounding
  regimens and the stated cap of 6 g daily.
* **The CrCL 10-30 arm of Tables 4 and 5 is not reproducible from the printed
  regimen.** Simulating Table 1 as printed (aztreonam 500 mg q8h,
  amoxicillin 500 mg q12h) at steady state undershoots the published column by
  up to 13 percentage points for aztreonam and up to 55 for amoxicillin in the
  ELF, while the same models reproduce the two higher-CrCL arms. The paper
  describes this arm inconsistently: Section 2.4 says both drugs shared a dosing
  interval, but Table 1 gives q8h and q12h in this row alone; Section 2.5 says
  the profiles are at steady state; and the Supplementary Materials script spells
  out only the CrCL >50-150 regimen. Two readings each reproduce their own drug's
  column to about 1 percentage point -- aztreonam 500 mg q8h evaluated over the
  **first** 24 h including the 2 g loading dose, and amoxicillin **1000** mg
  q12h at steady state -- but they disagree about the evaluation window, so
  neither is adopted. The vignette keeps Table 1 as printed, excludes this arm
  from the gate, and shows the diagnostic. No parameter or regimen was adjusted.
* **Table 4 amoxicillin standard error.** The CrCL >30-50 mL/min amoxicillin
  `fTMSW` cell for EC13/EC48 prints "2.4 +/- 7.7%", a standard error three times
  its own mean and two orders of magnitude larger than every other standard
  error in the table. At least one value in that block is misprinted. The
  amoxicillin gate bounds above are correspondingly looser than the aztreonam
  ones.
* **Amoxicillin reproduces with a systematic offset.** Simulated amoxicillin
  `fT>MPC` runs 1 to 5 percentage points above Table 4 across every isolate,
  while `fTMSW` matches to about 1 point. The aztreonam model, whose parameters
  and covariate model are specified in more detail, reproduces both tables
  essentially exactly, so the offset is most likely an unreported detail of the
  amoxicillin simulation setup rather than a transcription error. No parameter
  was adjusted to close it.
* **Section 3.3 self-inconsistency on the amoxicillin 8 mg/L target.** The same
  paragraph states that the regimens achieve at least 90% PTA for 50% `fT>MIC`
  at MIC up to 8 mg/L, and then that only about 80% was achieved at 8 mg/L for
  CrCL >50 mL/min. The simulation agrees with the second, more specific
  statement, which is the one gated above.

## Interpretation and encoding choices

* **Covariate exponents on amoxicillin and clavulanate clearance.** Both models
  scale clearance linearly with `CRCL/102`, with no power exponent. This is not
  a lost superscript: the aztreonam equations printed in the same two places do
  retain their `(CrCL/100)^0.43` and `(WT/70)^1.99` exponents under both
  `pdftotext` extraction modes, whereas the amoxicillin and clavulanate
  equations carry no exponent in either the main text or the supplement.
* **The reported "CV%" values are log-scale standard deviations.** Section 2.4
  labels them coefficients of variation; the Supplementary Materials script
  passes the identical numbers to `rnorm(sd = )` on the log scale. The `ini()`
  omega entries are therefore the squares of the printed values divided by 100
  (for example 24.1% becomes 0.241 squared, 0.058081). Under a log-normal
  distribution a log SD of 0.241 corresponds to an exact CV of 24.5%, so the
  paper used the small-variance approximation CV is approximately equal to the
  log SD.
* **No variability on `q`, and on `vp` for amoxicillin and clavulanate.** No CV
  is reported for these anywhere in the paper or supplement, and the simulation
  script draws no eta for them. They are carried as `fixed(0)` rather than
  invented.
* **No residual error.** Zhang 2023 simulated concentration-time profiles from
  between-subject variability only and never fitted concentration data, so no
  residual error is reported. `propSd` is carried as `fixed(0)` so that the
  endpoint is structurally present and a downstream re-fit has somewhere to put
  it.
* **Every parameter is `fixed()`.** Zhang 2023 estimated nothing; all values are
  literature priors re-used for simulation. Marking them fixed preserves that
  distinction.
* **Loading doses.** The 2 g aztreonam and 1 g amoxicillin/clavulanate loading
  doses are carried for fidelity but do not influence any steady-state metric.
  Table 1's "1/0.2 g followed by 500/100 mg q12h" for the CrCL 10-30 arm is read
  as a single loading dose followed by q12h maintenance; the reading does not
  affect the gates, which are evaluated at steady state.
* **Steady-state window.** Metrics are evaluated over 96 to 120 h. The paper
  states 24 h at steady state without naming a window; 96 h of dosing is well
  over 15 half-lives even for the slowest simulated subjects.
* **Clavulanate has no PD gate.** Section 2.6 states that dosing simulations for
  clavulanic acid were not undertaken because its PD target is not established,
  so `Zhang_2023_clavulanicAcid` is validated only through the exposure figure
  and the `AUCtau = Dose/CL` identity.
* **Non-paper-derived values.** None. Every parameter comes from the main text or
  the Supplementary Materials of Zhang 2023. The upstream publications were not
  consulted, so where Zhang 2023 and its own supplement disagree, the
  disagreement is resolved against the paper's own published tables rather than
  against the upstream source.
