---
title: "Infliximab (Zhao 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Infliximab (Zhao 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhao_2026_infliximab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhao_2026_infliximab"))$description`
* Article: <https://doi.org/10.1007/s40262-025-01565-6>
* Supplement (open access, two PDFs: Supplementary Tables 1-2 and Supplementary Figures 1-4): <https://doi.org/10.1007/s40262-025-01565-6>

```{r load-model}
mod <- rxode2::rxode(readModelDb("Zhao_2026_infliximab"))
```

# Population

Zhao 2026 is the first population-pharmacokinetic analysis of infliximab
restricted to children with inflammatory bowel disease who were 10 years old or
younger when infliximab was started. Therapeutic-drug-monitoring data were
collected retrospectively from 14 European and Canadian centres (patients
treated 2004-2016, data collected 2015-2019). The cohort comprised 104 children
with a median age of 8.2 years (range 1.2-10.0) and a median body weight of
25 kg (range 9.5-40.9) at the first infusion; 54 of 104 (52%) were female, 59
(57%) had Crohn's disease and 45 (43%) ulcerative colitis (Table 1). Sixteen
patients were under 6 years old, an age band in which infliximab is off-label.

Baseline inflammatory markers were high relative to published values for
children over 10 years: C-reactive protein 10.4 +/- 22.3 mg/L (median 3, range
0.02-302) and erythrocyte sedimentation rate 19.6 +/- 13.6 mm/h. Serum albumin
was 39.6 +/- 4.8 g/L (median 40.5, range 19.6-50). The median infusion was
6 mg/kg (range 3.5-15). Race and ethnicity are not reported.

The Abstract and Results 3.1 report 2150 measured concentrations, of which only
four were peak samples; Results 3.3 states that the model-development dataset
comprised 640 measurements, 95 of them (14.4%) below the limit of
quantification and handled with the M3 method. The near-absence of peak samples
is the source of the two fixed disposition parameters and, as shown below, of
this model's principal limitation.

The same information is available programmatically via the model's `population`
metadata:

```{r population}
pop <- readModelDb("Zhao_2026_infliximab")()$population
str(pop, max.level = 1)
```

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Zhao_2026_infliximab.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| Structural model: two compartments, intravenous | -- | Results 3.3, "A two-compartmental popPK model best described the IFX disposition in this population" |
| Covariate model form: power on median-normalised covariate, multiplied by allometric weight term | -- | Eq. 3 (continuous covariates); Methods 2.3.2, "values were normalized to the population median" |
| `lcl` (CL) | 0.779 L/day/65 kg | Table 2, final model; RSE 7%; bootstrap 0.775 (0.615-0.943) |
| `lvc` (V1) | 17.2 L/65 kg | Table 2, final model; RSE 16%; bootstrap 17.1 (10.831-23.629) |
| `lvp` (V2) | 1.21 L/65 kg, FIXED | Table 2, final model, "1.21 (FIX)"; value from Chung et al. and Clemente-Bautista et al. per Results 3.3 |
| `lq` (Q) | 0.0697 L/day/65 kg, FIXED | Table 2, final model, "0.0697 (FIX)"; same provenance |
| `e_wt_cl_q` | 0.75, FIXED | Eq. 3 and Eq. 4, `(body weight/65 kg)^0.75`; Results 3.3, "fixed at 0.75 for all CL terms ... a priori" |
| `e_wt_vc_vp` | 1, FIXED | Eq. 3 and Eq. 4 text, "if the parameter is volume of distribution, the exponential ... is 1"; Results 3.3 |
| `e_alb_cl` | -0.758 | Table 2, "ALB on CL"; RSE 32%; bootstrap -0.754 (-1.270 to -0.246) |
| Albumin centring value | 40.5 g/L | Results 3.3, "from the reference value of 40.5 g/L"; Table 1 total-cohort median |
| `e_crp_cl` | 0.0545 | Table 2, "CRP on CL"; RSE 36%; bootstrap 0.0541 (0.016-0.093) |
| CRP centring value | 3 mg/L | Results 3.3, "from the reference value of 3 mg/L"; Table 1 total-cohort median |
| `etalcl` | 0.156 (= log(1 + 0.411^2)) | Table 2, "CL (%CV)" = 41.1%; RSE 12%, shrinkage 15%; bootstrap 39.8% (26.88-52.35%) |
| `propSdSanquin` | 0.975 (= sqrt(0.951)) | Table 2, "Prop. RUV for Sanquin" = 0.951; RSE 27% |
| `propSdImmundiagnostik` | 0.884 (= sqrt(0.781)) | Table 2, "Prop. RUV for Immundiagnostik" = 0.781; RSE 9% |
| `propSdCaltag` | 0.785 (= sqrt(0.617)) | Table 2, "Prop. RUV for Caltag" = 0.617; RSE 9% |
| `propSdMatrixBiotek` | 0.769 (= sqrt(0.591)) | Table 2, "Prop. RUV for Matrix biotek" = 0.591; RSE 20% |
| `propSdBenHorin` | 0.943 (= sqrt(0.889)) | Table 2, "Prop. RUV for Shomron Ben-Horin" = 0.889; RSE 25% |
| `propSdPromonitor` | 0.956 (= sqrt(0.914)) | Table 2, "Prop. RUV for Promonitor" = 0.914; RSE 31% |
| Six-way assay stratification of the residual error | -- | Methods 2.3.1, error model evaluated "for each of the six different [IFX] assays"; Table 1 assay counts |
| Target trough concentration | 5 mg/L | Methods 2.3.5, per the ECCO/ESPGHAN guideline (van Rheenen et al.) |

The reported `Prop. RUV` values are NONMEM `$SIGMA` **variances**, so the SDs
that `prop()` requires are their square roots. The argument establishing that
scale is in [Assumptions and deviations](#assumptions-and-deviations); it is the
one non-obvious transcription decision in this extraction.

# Structural identities

These checks are deterministic -- they use `rxode2::zeroRe()` to suppress
between-subject variability, and every reference quantity is computed from the
values **printed in the paper** rather than from the model's own output, so a
mis-transcribed parameter, allometric exponent or unit makes them fail.

```{r typical-setup}
tmod <- rxode2::zeroRe(mod)

# Reference quantities from the PRINTED Table 2 values only. Nothing here reads
# the model object, so these are an independent check of the encoding.
printed <- list(cl = 0.779, vc = 17.2, vp = 1.21, q = 0.0697,
                e_cl = 0.75, e_v = 1, wt_ref = 65,
                alb_ref = 40.5, crp_ref = 3)

printed_pars <- function(wt) {
  list(cl = printed$cl * (wt / printed$wt_ref)^printed$e_cl,
       vc = printed$vc * (wt / printed$wt_ref)^printed$e_v,
       vp = printed$vp * (wt / printed$wt_ref)^printed$e_v,
       q  = printed$q  * (wt / printed$wt_ref)^printed$e_cl)
}

# Closed-form terminal half-life of a two-compartment IV model.
terminal_half_life <- function(p) {
  k10 <- p$cl / p$vc; k12 <- p$q / p$vc; k21 <- p$q / p$vp
  b <- k10 + k12 + k21
  log(2) / ((b - sqrt(b^2 - 4 * k10 * k21)) / 2)
}

solve_typical <- function(wt, amt, dose_times, obs_times, alb = printed$alb_ref,
                          crp = printed$crp_ref, assay = 2L) {
  ev <- rxode2::et(amt = amt, time = dose_times, cmt = "central") |>
    rxode2::et(obs_times, cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt; d$ALB <- alb; d$CRP <- crp; d$ASSAY_IFX <- assay
  out <- rxode2::rxSolve(tmod, d, returnType = "data.frame")
  out[!is.na(out$Cc), ]
}
```

## Single 5 mg/kg dose in the median 25 kg child

```{r identities}
wt <- 25
amt <- 5 * wt
p <- printed_pars(wt)
prof <- solve_typical(wt, amt, dose_times = 0, obs_times = seq(0, 400, by = 0.05))

# Trapezoidal AUC over the solved window, and the drug still in the body at the
# end of it.
auc_window <- sum(diff(prof$time) *
                    (head(prof$Cc, -1) + tail(prof$Cc, -1)) / 2)
remaining <- tail(prof$central, 1) + tail(prof$peripheral1, 1)

identities <- tibble::tribble(
  ~Identity, ~Reference, ~Model,
  "C(0) = dose / V1", amt / p$vc, prof$Cc[1],
  "AUC(0,inf) = dose / CL", amt / p$cl, auc_window,
  "AUC(0,T) * CL = dose in - amount remaining", amt - remaining, auc_window * p$cl,
  "Terminal half-life (days)", terminal_half_life(p), NA_real_
) |>
  mutate(`% diff` = 100 * (Model - Reference) / Reference)

knitr::kable(identities, digits = 4,
             caption = "Deterministic identities. Reference values use only the printed Table 2 parameters.")

# Mass balance and the dose/volume identity are exact algebra, so these are
# tight by construction; only solver tolerance separates the two sides.
stopifnot(
  abs(prof$Cc[1] - amt / p$vc) / (amt / p$vc) < 1e-6,
  abs(auc_window * p$cl - (amt - remaining)) / amt < 1e-4,
  # AUC(0,400) has not quite reached AUC(0,inf); 400 days is >27 terminal
  # half-lives, so the shortfall is far below 0.1%.
  abs(auc_window - amt / p$cl) / (amt / p$cl) < 1e-3
)

cat(sprintf("Terminal half-life at 25 kg: %.1f days\n", terminal_half_life(p)))
```

## Covariate sensitivity against the printed prose

Results 3.3 makes two quantitative claims about the covariate model. Both are
reproduced here from the model's own clearance output, so a wrong exponent or a
wrong centring constant breaks them.

```{r covariate-sensitivity}
cl_at <- function(alb, crp) {
  o <- solve_typical(25, 125, 0, c(0, 1), alb = alb, crp = crp)
  unique(o$cl)
}

# "CL decreased by approximately 17% for every 10 g/L increase in albumin from
# the reference value of 40.5 g/L". The printed 17% is reproduced by a 35 -> 45
# g/L step; a step starting at 40.5 itself gives 15.4%. Both readings confirm
# the exponent, which is what is load-bearing.
alb_35_45 <- 100 * (1 - cl_at(45, 3) / cl_at(35, 3))
alb_from_ref <- 100 * (1 - cl_at(50.5, 3) / cl_at(40.5, 3))

# "CL increased by approximately 1.6% for every 1 mg/L increase in CRP from the
# reference value of 3 mg/L". This pins BOTH the exponent and the centring
# value of 3 mg/L exactly.
crp_3_4 <- 100 * (cl_at(40.5, 4) / cl_at(40.5, 3) - 1)

claims <- tibble::tribble(
  ~Claim, ~Printed, ~Model,
  "CL change, albumin 35 -> 45 g/L", "-17%", sprintf("-%.1f%%", alb_35_45),
  "CL change, albumin 40.5 -> 50.5 g/L", "(not printed)", sprintf("-%.1f%%", alb_from_ref),
  "CL change, CRP 3 -> 4 mg/L", "+1.6%", sprintf("+%.2f%%", crp_3_4)
)
knitr::kable(claims, caption = "Printed covariate sensitivity statements (Results 3.3) vs the packaged model.")

stopifnot(
  abs(alb_35_45 - 17.3) < 0.5,
  abs(crp_3_4 - 1.58) < 0.05
)
```

# Non-compartmental analysis (PKNCA)

Zhao 2026 reports **no** NCA parameters -- the paper's only exposure metric is
`AUC` over weeks 6-14, used solely to compare assays and reported as p-values
rather than values (Table 5). The reference column below is therefore derived
from the paper's printed structural parameters, which is what is available to
check against, and is labelled as such.

```{r pknca}
nca_wt <- 25
nca_amt <- 5 * nca_wt
nca_p <- printed_pars(nca_wt)

# A sampling grid dense through the distribution phase so AUC is not
# understated, extended to a window many terminal half-lives long.
nca_times <- sort(unique(c(seq(0, 2, by = 0.1), seq(2.5, 14, by = 0.5),
                           seq(15, 60, by = 2), seq(64, 240, by = 8))))
nca_prof <- solve_typical(nca_wt, nca_amt, dose_times = 0, obs_times = nca_times)
stopifnot(all(nca_prof$Cc >= 0))

sim_nca <- nca_prof |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = 1L, treatment = "5 mg/kg IV, 25 kg") |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time-zero record so PKNCA's AUC interval starts at a measurement.
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = 0, Cc = nca_amt / nca_p$vc)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- data.frame(id = 1L, treatment = "5 mg/kg IV, 25 kg",
                      time = 0, amt = nca_amt)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "day")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id,
                             doseu = "mg")

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(
    conc_obj, dose_obj,
    intervals = data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                           aucinf.obs = TRUE, half.life = TRUE, cl.obs = TRUE)
  ),
  verbose = FALSE
)

reference <- tibble::tibble(
  treatment  = "5 mg/kg IV, 25 kg",
  cmax       = nca_amt / nca_p$vc,
  tmax       = 0,
  aucinf.obs = nca_amt / nca_p$cl,
  half.life  = terminal_half_life(nca_p),
  cl.obs     = nca_p$cl
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_res,
  reference = reference,
  by        = "treatment",
  units     = c(cmax = "mg/L", tmax = "day", aucinf.obs = "mg*day/L",
                half.life = "day", cl.obs = "L/day"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = paste(
  "PKNCA on the packaged model vs closed-form values from the paper's printed",
  "CL, V1, V2 and Q. The paper publishes no NCA parameters.",
  "* differs from reference by >20%."),
  align = c("l", "l", "r", "r", "r"))

# CL recovered by NCA must match Dose/AUC from the printed parameters. This is
# pure algebra on a deterministic profile, so it is tight.
nca_tbl <- as.data.frame(nca_res)
get_nca <- function(code) {
  v <- nca_tbl$PPORRES[nca_tbl$PPTESTCD == code]
  if (length(v) != 1L) stop("no unique NCA row for '", code, "'")
  v
}
stopifnot(
  abs(get_nca("cl.obs") - nca_p$cl) / nca_p$cl < 0.01,
  abs(get_nca("cmax") - nca_amt / nca_p$vc) / (nca_amt / nca_p$vc) < 0.01,
  abs(get_nca("half.life") - terminal_half_life(nca_p)) /
    terminal_half_life(nca_p) < 0.05
)
```

`cl.obs` from PKNCA reproduces the printed clearance for a 25 kg child, and
`half.life` reproduces the closed-form terminal half-life, confirming that the
allometric exponents and the two fixed disposition parameters are encoded as
printed.

# The paper's headline result: Table 6

Table 6 reports the percentage of a maintenance dosing interval spent below the
5 mg/L target for two typical patients -- a 5-year-old weighing 18 kg and an
8-year-old weighing 35 kg -- across three dose levels and three intervals, after
label induction at weeks 0, 2 and 6. Methods 2.3.5 states that each scenario was
simulated as 1000 virtual patients, so the reported percentage is a
population quantity: the fraction of patient-time below target across the
cohort, not a property of the median curve. Reproducing it therefore requires
the between-subject variability on clearance to be switched on.

```{r table6, fig.height = 5}
n_per_arm <- 200L  # per-arm cap; ample for this statistic

pct_below_target <- function(wt, dose_mg_kg, tau, n = n_per_arm, target = 5) {
  rxode2::rxSetSeed(20260910)  # common random numbers across scenarios
  amt <- dose_mg_kg * wt
  induction <- c(0, 14, 42)
  maintenance <- seq(42 + tau, by = tau, length.out = 30)
  t0 <- max(maintenance)
  ev <- rxode2::et(amt = amt, time = c(induction, maintenance), cmt = "central") |>
    rxode2::et(seq(t0, t0 + tau, by = 0.25), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- wt; d$ALB <- printed$alb_ref; d$CRP <- printed$crp_ref
  d$ASSAY_IFX <- 2L
  o <- rxode2::rxSolve(mod, d, nSub = n, returnType = "data.frame")
  o <- o[!is.na(o$Cc) & o$time >= t0 & o$time < t0 + tau, ]
  100 * mean(o$Cc < target)
}

table6 <- tibble::tribble(
  ~patient,              ~wt, ~dose, ~tau_wk, ~published,
  "5 years old (18 kg)",  18,   5.0,       4,         27,
  "5 years old (18 kg)",  18,   5.0,       6,         42,
  "5 years old (18 kg)",  18,   5.0,       8,         53,
  "5 years old (18 kg)",  18,   7.5,       4,          9,
  "5 years old (18 kg)",  18,   7.5,       6,         33,
  "5 years old (18 kg)",  18,   7.5,       8,         46,
  "5 years old (18 kg)",  18,  10.0,       4,          0,
  "5 years old (18 kg)",  18,  10.0,       6,         25,
  "5 years old (18 kg)",  18,  10.0,       8,         40,
  "8 years old (35 kg)",  35,   5.0,       4,         27,
  "8 years old (35 kg)",  35,   5.0,       6,         33,
  "8 years old (35 kg)",  35,   5.0,       8,         47,
  "8 years old (35 kg)",  35,   7.5,       4,          0,
  "8 years old (35 kg)",  35,   7.5,       6,         25,
  "8 years old (35 kg)",  35,   7.5,       8,         47,
  "8 years old (35 kg)",  35,  10.0,       4,          0,
  "8 years old (35 kg)",  35,  10.0,       6,         25,
  "8 years old (35 kg)",  35,  10.0,       8,         40
) |>
  rowwise() |>
  mutate(simulated = pct_below_target(wt, dose, tau_wk * 7)) |>
  ungroup() |>
  mutate(abs_diff = abs(simulated - published))

table6 |>
  transmute(Patient = patient,
            `Dose (mg/kg)` = dose,
            `Interval (weeks)` = tau_wk,
            `Published (%)` = published,
            `Simulated (%)` = round(simulated, 1),
            `Absolute difference (pp)` = round(abs_diff, 1)) |>
  knitr::kable(caption = paste(
    "Percentage of the maintenance dosing interval spent below 5 mg/L.",
    "Published values are Zhao 2026 Table 6."))

cat(sprintf("median |diff| = %.1f pp; 90th percentile = %.1f pp; max = %.1f pp\n",
            median(table6$abs_diff), quantile(table6$abs_diff, 0.9),
            max(table6$abs_diff)))
```

The centre of the comparison is what matters here: the simulated cohort is a
random draw, and `rxSetSeed()` fixes the draw only for a given solver-thread
count, so the per-cell values move between machines. The gate is therefore on
robust summaries across the 18 cells rather than on any single cell or on the
extremes.

```{r table6-gate}
# Measured at 1 / 2 / 4 / 16 solver threads: median |diff| 4.91 / 3.71 / 3.39 /
# 3.11 pp and 90th percentile 12.81 / 10.65 / 10.34 / 10.43 pp. The bounds sit
# outside that range with headroom -- do not tighten them back to one run. A
# mis-transcribed clearance, volume, dose or allometric exponent moves all 18
# cells by tens of percentage points and still breaks them.
stopifnot(
  median(table6$abs_diff) < 8,
  quantile(table6$abs_diff, 0.9) < 18
)

# The paper's central clinical conclusion, and the one that is robust to the
# cohort draw: the younger, lighter patient spends more of every interval below
# target than the older, heavier one at matched dose and interval. Asserted on
# the TREND across all nine matched pairs rather than pair-by-pair, because the
# q4w 10 mg/kg pair sits near zero where the ordering is a coin flip.
paired <- table6 |>
  select(wt, dose, tau_wk, simulated) |>
  pivot_wider(names_from = wt, values_from = simulated, names_prefix = "wt")
stopifnot(mean(paired$wt18 - paired$wt35) > 3)

# Monotone trends, asserted on the DETERMINISTIC typical-value curve so they
# are exact rather than a race between noisy statistics.
typical_pct <- function(wt, dose_mg_kg, tau, target = 5) {
  amt <- dose_mg_kg * wt
  maintenance <- seq(42 + tau, by = tau, length.out = 30)
  t0 <- max(maintenance)
  o <- solve_typical(wt, amt, c(0, 14, 42, maintenance),
                     seq(t0, t0 + tau, by = 0.05))
  o <- o[o$time >= t0 & o$time < t0 + tau, ]
  100 * mean(o$Cc < target)
}
trend <- expand.grid(wt = c(18, 35), dose = c(5, 7.5, 10), tau_wk = c(4, 6, 8)) |>
  rowwise() |>
  mutate(pct = typical_pct(wt, dose, tau_wk * 7)) |>
  ungroup()

for (w in c(18, 35)) {
  for (d in c(5, 7.5, 10)) {
    x <- trend$pct[trend$wt == w & trend$dose == d]
    stopifnot(all(diff(x) >= 0))          # longer interval never helps
  }
  for (tw in c(4, 6, 8)) {
    x <- trend$pct[trend$wt == w & trend$tau_wk == tw]
    stopifnot(all(diff(x) <= 0))          # higher dose never hurts
  }
}
```

Both of the paper's dosing conclusions follow. At the label-recommended
5 mg/kg every 8 weeks the typical young child is below target for roughly half
the interval, and only a 4-week interval brings the subtherapeutic fraction
close to zero -- Conclusions: "To maintain trough concentrations above 5 mg/L in
paediatric patients with IBD (<= 10 years old), the dosing interval during the
maintenance phase should be reduced to 4 weeks."

```{r fig34, fig.height = 5}
profile_for <- function(wt, dose_mg_kg, tau) {
  amt <- dose_mg_kg * wt
  maintenance <- seq(42 + tau, by = tau, length.out = 4)
  o <- solve_typical(wt, amt, c(0, 14, 42, maintenance),
                     seq(0, max(maintenance) + tau, by = 0.25))
  o$dose <- paste0(dose_mg_kg, " mg/kg")
  o$interval <- paste0("q", tau %/% 7, "w")
  o$patient <- if (wt == 18) "5 years old (18 kg)" else "8 years old (35 kg)"
  o
}
profiles <- do.call(rbind, lapply(c(18, 35), function(w)
  do.call(rbind, lapply(c(5, 7.5, 10), function(d)
    do.call(rbind, lapply(c(28, 42, 56), function(t) profile_for(w, d, t)))))))

ggplot(profiles, aes(time / 7, Cc, colour = interval)) +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  facet_grid(patient ~ dose) +
  scale_y_log10() +
  labs(x = "Time (weeks)", y = "Infliximab concentration (mg/L)",
       colour = "Maintenance\ninterval",
       title = "Replicates Figures 3 and 4 of Zhao 2026",
       subtitle = "Typical-value profiles; dashed line is the 5 mg/L target") +
  theme_bw()
```

# Label-recommended dosing across the cohort weight distribution

Figure 2 of Zhao 2026 simulates the label-recommended regimen over the study
population. Results 3.7 quantifies it: with the standard 8-week maintenance
interval, concentrations fell below 5 mg/L for 63% of the interval at 5 mg/kg,
50% at 7.5 mg/kg and 38% at 10 mg/kg, while median trough concentrations
exceeded 5 mg/L during induction (weeks 2-6).

These three numbers are the one place where the extraction does not cleanly
reproduce the paper, and the reason is a disagreement inside the source rather
than in the model: a cohort centred on 25 kg cannot be below target for *more*
of the interval (63%) than either of Table 6's typical patients at the same dose
and interval (53% at 18 kg, 47% at 35 kg). The table below therefore reports
both readings of "concentrations fell below 5 mg/L" -- the individual prediction
and the simulated observation -- and
[Assumptions and deviations](#assumptions-and-deviations) records what was
tested. Nothing was tuned to close the gap.

```{r fig2}
# Weight distribution matching Table 1 (median 25 kg, range 9.5-40.9, mean 25.3
# +/- 7.1). A truncated normal on the reported mean and SD reproduces both the
# median and the range.
rxode2::rxSetSeed(20260910)
set.seed(20260910)
cohort_wt <- pmin(pmax(rnorm(n_per_arm, 25.3, 7.1), 9.5), 40.9)

label_regimen <- function(dose_mg_kg) {
  rxode2::rxSetSeed(20260910)
  induction <- c(0, 14, 42)
  maintenance <- seq(42 + 56, by = 56, length.out = 6)
  subj <- lapply(seq_along(cohort_wt), function(i) {
    amt <- dose_mg_kg * cohort_wt[i]
    ev <- rxode2::et(amt = amt, time = c(induction, maintenance), cmt = "central") |>
      rxode2::et(seq(0, max(maintenance) + 56, by = 1), cmt = "central")
    d <- as.data.frame(ev)
    d$id <- i; d$WT <- cohort_wt[i]
    d$ALB <- printed$alb_ref; d$CRP <- printed$crp_ref; d$ASSAY_IFX <- 2L
    d
  })
  d <- do.call(rbind, subj)
  o <- rxode2::rxSolve(mod, d, returnType = "data.frame")
  o <- o[!is.na(o$Cc), ]
  o$dose <- paste0(dose_mg_kg, " mg/kg")
  o
}
fig2 <- do.call(rbind, lapply(c(5, 7.5, 10), label_regimen))

fig2_summary <- fig2 |>
  group_by(dose, time) |>
  summarise(median = median(Cc),
            p05 = quantile(Cc, 0.05),
            p95 = quantile(Cc, 0.95), .groups = "drop")

ggplot(fig2_summary, aes(time / 7, median)) +
  geom_ribbon(aes(ymin = p05, ymax = p95), alpha = 0.2) +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dashed") +
  facet_wrap(~dose) +
  scale_y_log10() +
  labs(x = "Time (weeks)", y = "Infliximab concentration (mg/L)",
       title = "Replicates Figure 2 of Zhao 2026",
       subtitle = paste("Label-recommended regimen: induction at weeks 0, 2, 6",
                        "then every 8 weeks; median with 5th-95th percentiles")) +
  theme_bw()
```

```{r fig2-checks}
# Median trough exceeds 5 mg/L during induction (weeks 2-6), i.e. at the week-2
# and week-6 pre-dose times.
induction_troughs <- fig2_summary |>
  filter(time %in% c(14, 42)) |>
  select(dose, time, median)
knitr::kable(induction_troughs, digits = 2,
             caption = "Median induction trough concentrations (weeks 2 and 6).")

# Fraction of the LAST 8-week maintenance interval below target, by dose, under
# both readings of "concentrations fell below 5 mg/L": the individual PREDICTION
# (Cc, the quantity Table 6 is reproduced with above) and the simulated
# OBSERVATION (sim, which carries the proportional residual error). Results 3.7
# does not say which it used, and with a residual SD near 0.9 the two differ
# substantially -- so both are shown rather than one being chosen to fit.
last_start <- 42 + 56 * 6
maint_below <- fig2 |>
  filter(time >= last_start, time < last_start + 56) |>
  group_by(dose) |>
  summarise(pct_pred = 100 * mean(Cc < 5),
            pct_obs = 100 * mean(sim < 5), .groups = "drop") |>
  mutate(published = c(63, 38, 50)[match(dose, c("5 mg/kg", "10 mg/kg", "7.5 mg/kg"))],
         diff_pred = pct_pred - published,
         diff_obs = pct_obs - published)

maint_below |>
  transmute(Dose = dose,
            `Published (%)` = published,
            `Simulated, prediction (%)` = round(pct_pred, 1),
            `Difference (pp)` = round(diff_pred, 1),
            `Simulated, observation (%)` = round(pct_obs, 1),
            `Difference (pp) ` = round(diff_obs, 1)) |>
  knitr::kable(caption = paste(
    "Percentage of the 8-week maintenance interval below 5 mg/L.",
    "Published values are Zhao 2026 Results 3.7."))

stopifnot(
  # Induction troughs above target at every dose (Results 3.7: "median IFX
  # trough concentrations exceeded 5 mg/L in the induction period"). A large
  # margin -- the lowest is around 24 mg/L -- so this is safe on any cohort.
  all(induction_troughs$median > 5),
  # Higher dose reduces subtherapeutic time. Published 63 -> 50 -> 38 is a
  # 25-point ordered effect, so the trend is safe to assert; asserted on the
  # endpoints rather than pairwise, and on both readings.
  maint_below$pct_pred[maint_below$dose == "10 mg/kg"] <
    maint_below$pct_pred[maint_below$dose == "5 mg/kg"],
  maint_below$pct_obs[maint_below$dose == "10 mg/kg"] <
    maint_below$pct_obs[maint_below$dose == "5 mg/kg"],
  # The observation reading brackets the published values. Measured differences
  # at 1/2/4/16 threads spanned -3 to +7 pp; the prediction reading is
  # reproducibly 5-10 pp low at every dose and is recorded as a deviation in
  # "Assumptions and deviations" rather than gated.
  median(abs(maint_below$diff_obs)) < 12
)
```

# Assay-stratified residual error

Stratifying the residual error by ELISA is the paper's central methodological
contribution. The `ASSAY_IFX` covariate selects one of six proportional
residual SDs per observation; because it enters **only** the error model, the
predicted exposure is identical across assays. That is the analytic form of the
paper's own conclusion (Table 5: all 15 pairwise comparisons of `AUC` over weeks
6-14 gave p > 0.05) and is a demonstration rather than a gate -- it holds by
construction of the model.

```{r assays}
assay_names <- c("Sanquin", "Immundiagnostik", "Caltag",
                 "Matrix Biotek", "Ben-Horin (in-house)", "Promonitor")

assay_sim <- do.call(rbind, lapply(seq_along(assay_names), function(k) {
  rxode2::rxSetSeed(20260910)
  ev <- rxode2::et(amt = 5 * 25, time = 0, cmt = "central") |>
    rxode2::et(seq(0, 56, by = 2), cmt = "central")
  d <- as.data.frame(ev)
  d$WT <- 25; d$ALB <- printed$alb_ref; d$CRP <- printed$crp_ref
  d$ASSAY_IFX <- k
  o <- rxode2::rxSolve(mod, d, nSub = n_per_arm, returnType = "data.frame")
  o <- o[!is.na(o$Cc), ]
  o$assay <- assay_names[k]
  o$code <- k
  o
}))

assay_tbl <- assay_sim |>
  group_by(code, assay) |>
  summarise(`Residual SD in model` = unique(round(propSdAssay, 3)),
            `Median individual prediction (mg/L)` = round(median(ipredSim), 3),
            `Observed-scale CV (%)` = round(100 * sd(sim / ipredSim), 1),
            .groups = "drop") |>
  mutate(`Table 2 variance` = c(0.951, 0.781, 0.617, 0.591, 0.889, 0.914)[code]) |>
  select(Code = code, Assay = assay, `Table 2 variance`,
         `Residual SD in model`, `Median individual prediction (mg/L)`,
         `Observed-scale CV (%)`)
knitr::kable(assay_tbl,
             caption = paste("Per-assay residual error. The individual",
                             "predictions are identical across assays because",
                             "the assay enters only the error model."))

stopifnot(
  # sqrt() of the printed Table 2 variances is what the model carries.
  all(abs(assay_tbl$`Residual SD in model` -
            round(sqrt(assay_tbl$`Table 2 variance`), 3)) < 1e-9),
  # Exposure is assay-independent: every assay yields the same median IPRED.
  diff(range(assay_tbl$`Median individual prediction (mg/L)`)) < 1e-6
)
```

```{r assay-plot, fig.height = 4}
ggplot(assay_sim, aes(factor(round(time)), sim / ipredSim)) +
  geom_boxplot(outlier.size = 0.3) +
  facet_wrap(~assay) +
  scale_y_log10() +
  geom_hline(yintercept = 1, linetype = "dashed") +
  labs(x = "Time (days)", y = "Observed / individual prediction",
       title = "Residual-error magnitude by infliximab ELISA",
       subtitle = "Proportional residual SD ranges from 0.769 (Matrix Biotek) to 0.975 (Sanquin)") +
  theme_bw()
```

# Comparison with older children and adults

Tables 3 and 4 place this model beside six published paediatric models
(> 10 to < 17 years) and six adult models. The comparison motivates the paper's
title claim, and it is reproduced here directly from those tables rather than
resimulated, since only this paper's model is packaged.

```{r comparison}
tibble::tribble(
  ~Population, ~`CL (L/day/65 kg)`, ~`V1 (L/65 kg)`, ~Source,
  "Children <= 10 years (this model)", "0.779", "17.2", "Table 2",
  "Children > 10 to < 17 years (range across 6 models)", "0.0151-0.353", "2.97-4.44", "Table 3",
  "Adults (range across 6 models)", "0.199-0.359", "3.41-5.03", "Table 4"
) |>
  knitr::kable(caption = "Typical parameter values by age group (Zhao 2026 Tables 2-4).")
```

Clearance in this cohort is roughly twice the highest value reported for older
children or adults, which the Discussion attributes to immature FcRn recycling
(citing Tian et al.) and to the higher inflammatory burden of these patients.
Central volume, however, is 3-5 times the published values, and the consequences
of that are discussed next.

# Assumptions and deviations {#assumptions-and-deviations}

**Residual error is on the variance scale.** Table 2's six `Prop. RUV` rows are
printed as bare decimals (0.591 to 0.951) with no percent sign and no unit,
while the inter-individual-variability row in the same table is explicitly
labelled `CL (%CV)` and carries a percent sign. That asymmetry already suggests
the residual rows are raw NONMEM `$SIGMA` values, which are variances. The
reported %RSEs settle it quantitatively. Splitting the 640 fitted observations
across assays in proportion to the Table 1 patient counts gives roughly 86, 192,
209, 37, 74 and 43 observations per stratum. The asymptotic %RSE of a variance
is `sqrt(2/n)`, and of an SD `sqrt(1/(2n))` -- a factor of two apart. Those
predictions are 15, 10, 10, 23, 16 and 22% on the variance scale versus 8, 5, 5,
12, 8 and 11% on the SD scale, against reported values of 27, 9, 9, 20, 25 and
31%. The variance reading tracks the reported precision across all six strata
(ratios 0.9 to 1.8, as expected for real estimates that exceed the asymptotic
floor); the SD reading is uniformly low by a factor of about 2.5. The model
therefore carries `sqrt()` of each printed value. Note that the two readings
differ by at most 27% in relative terms because the printed values sit near 1,
and that neither affects any quantity validated above: `zeroRe()` suppresses the
error entirely for the deterministic checks, and the Table 6 statistic is
computed from individual predictions.

**Central volume is large and the model should not be used for peaks.**
V1 = 17.2 L/65 kg gives `C(0) = 5 * 65 / 17.2 = 18.9 mg/L` after a 5 mg/kg dose,
independent of body weight because the volume exponent is 1. The observed peak
concentration after a 5 mg/kg infliximab infusion is around 100 mg/L, and the
comparator models in Tables 3 and 4 (V1 of 3-5 L/65 kg) reproduce that. The
discrepancy is a direct consequence of the dataset: Results 3.1 records that of
2150 concentrations "only four peak samples" were collected, and Results 3.3
states that V2 and Q could not be estimated at all for the same reason. V1 is
reported with a 16% RSE and a bootstrap interval of 10.8-23.6 L/65 kg, so this
is the paper's genuine estimate rather than a transcription error, but it is
identified almost entirely by trough kinetics. A consequence worth stating
explicitly: combining the high CL with the high V1 gives a terminal half-life of
about 14.5 days for a 25 kg child, **longer** than the 8-9 days implied by the
adult comparator models, which sits awkwardly with the Discussion's mechanistic
argument that immature FcRn recycling shortens infliximab half-life in young
children. **Use this model for trough concentrations and dosing-interval
questions, which is what it was built for and what is validated above; do not
use it to predict Cmax or early post-infusion concentrations.**

**Table 6's statistic is a population fraction, not a median-curve property.**
Methods 2.3.5 states that 1000 virtual patients were simulated per scenario but
does not define how "percentage of time interval below target" was computed. The
median-curve reading and the population patient-time reading were both tested:
the median curve under-predicts the q4w cells badly (11% versus a published 27%
for the 18 kg patient at 5 mg/kg), whereas the population patient-time fraction
reproduces all 18 cells with a median absolute difference of 3 percentage
points. The vignette uses the population reading and states the assumption here.

**Results 3.7 and Table 6 are not mutually consistent, and they favour
different readings of "below 5 mg/L".** Results 3.7 reports 63 / 50 / 38% of the
8-week maintenance interval below target at 5 / 7.5 / 10 mg/kg for a cohort
drawn from the study weight distribution, whose median weight is 25 kg. Table 6
reports 53% for an 18 kg patient and 47% for a 35 kg patient at 5 mg/kg q8w. A
cohort centred at 25 kg should land between those two typical-patient values,
i.e. between 47 and 53% -- not above both at 63%. The two published analyses
therefore disagree with each other, independently of this extraction.
Consistently with that, they favour different readings: the individual
*prediction* reproduces Table 6 (median absolute difference 3 pp over 18 cells)
but runs 5-10 pp below Results 3.7 at every dose, while the simulated
*observation*, which carries the residual error, brackets Results 3.7 (median
absolute difference about 3 pp) and would over-predict Table 6. Because the
residual SD is near 0.9, the two readings are far apart, and neither section
states which was used. The vignette uses the prediction reading throughout for
interpretability and consistency with the headline table, reports both readings
side by side against Results 3.7, and gates only the observation reading there.
The three prediction-reading differences are a reproducible deviation, not a
flickering one; they are recorded rather than tuned away.

Two candidate explanations for the Results 3.7 numbers were tested and rejected.
Using the observed albumin and CRP distributions (normal on 39.6 +/- 4.8 g/L,
truncated to 19.6-50; lognormal with median 3 and mean 10.4 mg/L, truncated to
0.02-302) instead of the cohort medians moves the prediction-reading fractions
by less than 1 pp, so the covariate spread is not the mechanism. Computing over
the first maintenance interval (weeks 6-14, the window the paper uses for its
`AUC` comparison) instead of at steady state moves them the wrong way, down by
6-7 pp, because the closely spaced induction doses leave that interval starting
higher than steady state.

**Two Table 6 cells for the 8-year-old appear to be carried over from the
5-year-old row.** Three cells drive most of the residual disagreement, all in
the 35 kg block: 5 mg/kg q4w (published 27%, simulated 13%), 7.5 mg/kg q8w
(published 47%, simulated 37%) and 10 mg/kg q6w and q8w (published 25 and 40%,
simulated 15 and 29%). The published 8-year-old values at 10 mg/kg (0, 25, 40)
are byte-identical to the 5-year-old values at 10 mg/kg, and the 8-year-old
7.5 mg/kg q8w value (47%) is identical to its own 5 mg/kg q8w value even though
the q6w value drops from 33 to 25% over the same dose increase. Duplication
across adjacent rows is the most economical explanation. The paper's prose
independently confirms the two cells that matter most -- Results 3.7: "reaching
53% in 5-year-old patients with IBD and 47% in 8-year-old patients with IBD at
an 8-week interval" -- and the model reproduces both (58 and 50%). The gate is
set on robust summaries across all 18 cells so that these cells are visible in
the table without being tuned away.

**The fixed V2 and Q carry an ambiguous reference weight.** Results 3.3 states
that V2 and Q were fixed to values from the published models of Chung et al. and
Clemente-Bautista et al. Table 3 prints those donor values as `V2 = 1.21 L/37.4
kg` and `Q = 0.0697 L/day/37.4 kg` (Chung) -- the **same numbers** that Table 2
and Table 3's own refined-model column carry against a **65 kg** reference. The
authors adopted the numerals without renormalising them from the donor's
reference weight, so the two tables disagree about what physical volume 1.21 L
denotes. This model encodes them as printed for Zhao's own model, at 65 kg,
because the paper's table for its own model is the authority. The alternative
reading makes V2 and Q 1.7 times larger at any body weight and lengthens the
terminal half-life by 13.2% (14.5 to 16.4 days at 25 kg). Table 6 was used to
try to arbitrate: the as-printed reading gives a median absolute difference of
3.11 percentage points over the 18 cells versus 3.45 for the alternative, which
favours the as-printed reading but is not decisive -- Q is an order of magnitude
below CL, so the peripheral compartment barely influences trough concentrations
and the statistic has little power to separate the two.

**CRP enters as a power term and is undefined at zero.** `(CRP/3)^0.0545` sends
clearance to zero as CRP approaches zero. The cohort minimum is 0.02 mg/L, at
which the multiplier is 0.762, so the form is well behaved over the observed
range, but a user simulating an exactly-zero CRP must floor the value at the
assay's lower reporting limit.

**The 65 kg allometric reference is far outside the data.** Body weights ranged
from 9.5 to 40.9 kg, so both `lcl` and `lvc` are extrapolations to a weight no
patient had. The authors chose 65 kg so their estimates would be comparable with
the adult and older-paediatric literature in Tables 3 and 4, and the Discussion
flags the consequence: "it could not be ignored that the covariate distribution
impacts scaling of CL and V1, ultimately impacting the estimation of the typical
values (Eq. 3)". All simulations here dose weights inside the observed range.

**Covariates screened but not retained are recorded, not encoded.** Age, height,
body surface area, sex, diagnosis type, concomitant immunomodulator use,
erythrocyte sedimentation rate and Physician's Global Assessment were all
screened (Methods 2.3.2) and rejected. The paper reports no point estimate,
objective-function change or confidence interval for any of them, so none can be
encoded; they are documented in the model file's `covariatesDataExcluded` list.
Notably ESR is retained on clearance by four of the six comparator models in
Table 3, which makes its rejection here a finding rather than an omission.

**Covariates are treated as static in these simulations.** Albumin and CRP are
laboratory values drawn alongside each therapeutic-drug-monitoring sample and
are time-varying in the source dataset; body weight changes as a child grows
over a multi-year treatment course. Every simulation here holds all three fixed,
at the cohort medians for albumin and CRP, matching the paper's own
typical-patient simulations.

**BLQ handling is not reproduced.** 95 of 640 measurements (14.4%) were below
the limit of quantification and were handled with the M3 method during
estimation. The packaged model emits continuous concentrations, so there is no
BLQ step in these simulations; the M3 method affects parameter estimation, not
forward simulation.

**Two assay labels in Supplementary Table 1 are not among the six modelled
strata.** Supplementary Table 1 maps 14 contributing centres onto vendor labels
and includes "apDIA" (Ghent), which contributed no patients to the PK analysis,
and "Sanquin (LC-MS/MS)" (Utrecht), whose observations fall in the Sanquin
stratum -- no separate LC-MS/MS residual magnitude is estimated. "Immune
diagnostic" (Dublin) and "Promonitor/Sanquin" (Tampere) are the Immundiagnostik
and Promonitor strata under variant spellings.

**Minor arithmetic and typesetting inconsistencies in the source.** The 14.4%
BLQ figure does not equal 95/640 = 14.8%. Table 1 prints two consecutive rows
both labelled "Female, n (%)" (54 (52%) and 50 (48%)); the second is evidently
male. The immunomodulator percentages (59.0% and 41.0%) are transposed relative
to their counts (46 and 58 of 104). The Table 1 Physician's Global Assessment
block cannot be reconciled across columns (total-cohort remission of 2 versus 10
in the under-6 column alone). None of these touch a parameter used by the model.

**A new canonical covariate column was registered.** `ASSAY_IFX` is an integer
1-6 code naming which commercial infliximab ELISA measured a sample, added to
`inst/references/covariate-columns.md` as an auto-approved member of the
`ASSAY_<method>` family. An integer stratum rather than five or six paired
binary indicators follows the guidance already recorded in the `ASSAY_CMIA`
entry for datasets mixing more than two assays, and the `STUDY_TACRO_FRANCKE`
precedent.
