---
title: "Tacrolimus with voriconazole co-therapy (Zhao 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tacrolimus with voriconazole co-therapy (Zhao 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- readModelDb("Zhao_2024_tacrolimus")
ui  <- rxode2::rxode(mod)
```

* Citation: `r ui$reference`
* Article: <https://doi.org/10.3389/fphar.2024.1439232>
* Description: `r ui$description`

Zhao 2024 develops a single population PK model, so this paper contributes one
model file.

# Population

The model was built on a retrospective, non-interventional single-centre
cohort of 19 adult first-time renal transplant recipients treated at The
Second Xiangya Hospital of Central South University, Changsha, China, between
January 2016 and March 2021 (ChiCTR2100048712). Every patient received oral
tacrolimus as part of a triple immunosuppressive regimen (with mycophenolate
mofetil and a glucocorticoid) and started voriconazole within 15 days of
transplantation.

Baseline characteristics (Zhao 2024 Table 1): 15/19 male (78.9%); median age
44 years (IQR 37.5-52.5); median weight 63 kg (IQR 51-72); median tacrolimus
dose 3.00 mg per administration (IQR 1.50-3.50); median observed tacrolimus
concentration 7.90 ng/mL (IQR 5.55-10.78); median time after operation 8 days
(IQR 4-11); median voriconazole concentration 0.00 ug/mL (IQR 0.00-0.50) with
an observed range of 0-3.38 ug/mL; median serum creatinine 237 umol/L (IQR
162.9-648.0). CYP3A5 genotyping was an inclusion requirement, but only
\*1/\*3 (n = 10) and \*3/\*3 (n = 9) carriers were enrolled -- no \*1/\*1 --
and neither CYP3A5 nor CYP2C19 genotype survived covariate selection.

167 whole-blood tacrolimus concentrations were collected (8-9 per patient),
the majority within 30 minutes before a dose, so the dataset is dominated by
troughs. That is why Ka had to be fixed and why an additive residual model was
selected.

The same information is available programmatically from
`rxode2::rxode(readModelDb("Zhao_2024_tacrolimus"))$population`.

# Source trace

Per-parameter origins are recorded as in-file comments beside each `ini()`
entry; they are collected here for review. Zhao 2024 prints **no display
equations at all** -- neither the structural model nor the covariate model --
so the functional forms below were reconstructed from the Table 3 and Table 4
footnotes ("exponent for `<covariate>`, as a covariate for `<parameter>`"),
from the covariate distributions in Table 1, and from the simulation settings
stated in section 3.4. This is documented in full under
*Assumptions and deviations*.

| Quantity | Value | Source location |
|----------|-------|-----------------|
| Structural model: one compartment, first-order absorption and elimination | n/a | Section 3.2.1, "a one-compartment model with first-order absorption and elimination ... was selected" |
| Residual model: additive | n/a | Section 3.2.1, "along with an additive residual model"; Table 2 model-comparison row `1a_Addictive` |
| `Ka` (fixed) | 8.39 1/h | Table 3, final model row `Ka`; fixed status from Table 3 base-model row `Ka (Fixed)` (Stderr 0.00, CI 8.39-8.39) and from the Discussion, "the Ka value was fixed during the model's development" |
| `V/F` | 2690 L | Table 3, final model row `V/F`; bootstrap mean 2655, 95% CI 1480-4060 (Table 4 `tvV`) |
| `CL/F` | 42.87 L/h | Table 3, final model row `CL/F`; bootstrap mean 42.00, 95% CI 30-50 (Table 4 `tvCl`) |
| Voriconazole coefficient on CL/F | -0.28 | Table 3 / Table 4 row `Theta VRC-CL` (Stderr 0.03; bootstrap 95% CI -0.99 to -0.21) |
| Voriconazole coefficient on V/F | -0.20 | Table 3 / Table 4 row `Theta VRC-V` (Stderr 0.04; bootstrap 95% CI -0.36 to -0.03) |
| Serum-creatinine exponent | -0.40 | Table 3 / Table 4 row `Theta CREA-V` (Stderr 0.11; bootstrap 95% CI -0.72 to -0.05) |
| Serum-creatinine normalising constant, 237 umol/L | n/a | Table 1 cohort median; section 3.4.1, "The CREA value, fixed at the median of 237 umol/L" |
| `omega^2` V | 0.02 | Table 4, row `omega 2 V` |
| `omega^2` CL | 0.16 | Table 4, row `omega 2 CL` |
| `sigma` (additive) | 3.50 ng/mL | Table 4, row `sigma`; bootstrap mean 3.41, 95% CI 3.06-3.74 |

# Structural verification against the closed form

With the random effects zeroed, the one-compartment first-order-absorption
model has an exact closed form, so a solve of the rxode2 model must reproduce
superposition of single-dose profiles to numerical precision. This check has
no random component, so a tight bound is the correct assertion here.

```{r closedform}
typ <- rxode2::zeroRe(mod)

kaRef   <- 8.39
clRef   <- 42.87
vcRef   <- 2690
doseMg  <- 3
dosingT <- seq(0, 60, by = 12)   # q12h for 3 days: 6 doses
obsT    <- seq(0, 84, by = 0.5)

evTyp <-
  dplyr::bind_rows(
    data.frame(id = 1L, time = dosingT, amt = doseMg, evid = 1L,
               cmt = "depot", dv = NA_real_),
    data.frame(id = 1L, time = obsT, amt = NA_real_, evid = 0L,
               cmt = "central", dv = NA_real_)
  ) |>
  dplyr::mutate(CONC_VORI_NGML = 0, CREAT = 237) |>
  dplyr::arrange(time, dplyr::desc(evid))

simTyp <- rxode2::rxSolve(typ, evTyp, returnType = "data.frame")

# Closed-form superposition, in ng/mL (dose mg / volume L = mg/L = 1000 ng/mL)
singleDose <- function(tau, dose, ka, cl, vc) {
  kel <- cl / vc
  out <- 1000 * dose * ka / (vc * (ka - kel)) * (exp(-kel * tau) - exp(-ka * tau))
  ifelse(tau < 0, 0, out)
}
closed <- vapply(
  simTyp$time,
  function(tt) sum(singleDose(tt - dosingT, doseMg, kaRef, clRef, vcRef)),
  numeric(1)
)

relDiff <- abs(simTyp$Cc - closed) / pmax(closed, 1e-8)
summary(relDiff)

stopifnot(max(relDiff) < 1e-3)
```

```{r closedform-plot, echo = FALSE}
ggplot2::ggplot(
  data.frame(time = simTyp$time, solved = simTyp$Cc, closed = closed),
  ggplot2::aes(time)
) +
  ggplot2::geom_line(ggplot2::aes(y = solved, colour = "rxode2 solve"), linewidth = 1) +
  ggplot2::geom_point(
    data = ~ dplyr::filter(.x, time %% 4 == 0),
    ggplot2::aes(y = closed, colour = "closed form"), size = 1.6
  ) +
  ggplot2::labs(
    x = "Time (h)", y = "Tacrolimus, whole blood (ng/mL)", colour = NULL,
    title = "3 mg q12h, typical patient at zero voriconazole and CREAT 237 umol/L"
  ) +
  ggplot2::theme_bw()
```

# Covariate effects: replicating the published dose-recommendation simulations

Zhao 2024 sections 3.4.1 and 3.4.2 report Monte Carlo simulations of the
**day-3 trough** (72 h after the first dose, on a q12h schedule, i.e. after six
doses) across a grid of voriconazole concentrations at fixed CREA = 237 umol/L
(Table 5) and across a grid of serum creatinines at fixed voriconazole = 0
(Table 6).

```{r day3-helper}
day3Trough <- function(cvrcUgMl, creat, doseMg) {
  ev <-
    dplyr::bind_rows(
      data.frame(time = dosingT, amt = doseMg, evid = 1L, cmt = "depot"),
      data.frame(time = 72,      amt = NA_real_, evid = 0L, cmt = "central")
    ) |>
    dplyr::mutate(id = 1L,
                  CONC_VORI_NGML = cvrcUgMl * 1000,
                  CREAT = creat) |>
    dplyr::arrange(time, dplyr::desc(evid))
  out <- rxode2::rxSolve(typ, ev, returnType = "data.frame")
  out$Cc[out$time == 72][1]
}
```

## Voriconazole concentration (Zhao 2024 Table 5)

```{r table5}
vGrid <- c(0, 0.5, 1, 1.5, 2, 2.5, 3, 3.5, 4, 4.5, 5, 5.5, 6, 6.5, 7)
# Published Table 5, tacrolimus 5.5 mg q12h column (mean day-3 trough, ng/mL)
table5Pub <- c(9.06, 9.63, 10.41, 11.13, 12.09, 13.31, 14.56, 16.37,
               18.18, 21.07, 24.82, 30.07, 38.14, 52.39, 80.36)

table5 <- data.frame(
  cvrc      = vGrid,
  published = table5Pub,
  model     = vapply(vGrid, day3Trough, numeric(1), creat = 237, doseMg = 5.5)
) |>
  dplyr::mutate(
    pubFold   = published / published[1],
    modelFold = model / model[1]
  )

knitr::kable(
  table5 |>
    dplyr::rename(
      "Voriconazole (ug/mL)"    = cvrc,
      "Published (ng/mL)"       = published,
      "This model (ng/mL)"      = model,
      "Published fold vs 0"     = pubFold,
      "This model fold vs 0"    = modelFold
    ),
  digits = 2,
  caption = "Replicates Zhao 2024 Table 5, 5.5 mg q12h column: day-3 trough vs voriconazole concentration at CREA = 237 umol/L."
)

# Direction and monotonicity are reproduced exactly; absolute levels are not.
stopifnot(
  all(diff(table5$model) > 0),
  table5$model[table5$cvrc == 0] > 0
)
```

The model reproduces the *direction* and the monotonicity of Table 5, but not
its absolute level: the published day-3 troughs run about 1.4 times higher than
the model at zero voriconazole, and the published curve is flatter than the
model over the observed 0-3.38 ug/mL range and steeper above it. This is a
property of the published table, not of the extraction -- see
*Assumptions and deviations*. No parameter has been tuned to close the gap.

```{r table5-plot, echo = FALSE}
table5 |>
  dplyr::select(cvrc, pubFold, modelFold) |>
  tidyr::pivot_longer(-cvrc, names_to = "source", values_to = "fold") |>
  dplyr::mutate(source = dplyr::recode(source,
                                       pubFold   = "Zhao 2024 Table 5",
                                       modelFold = "This model")) |>
  ggplot2::ggplot(ggplot2::aes(cvrc, fold, colour = source)) +
  ggplot2::geom_line(linewidth = 1) +
  ggplot2::geom_point() +
  ggplot2::geom_vline(xintercept = 3.38, linetype = "dashed") +
  ggplot2::annotate("text", x = 3.38, y = 8, hjust = -0.05, size = 3,
                    label = "observed C(VRC) ceiling") +
  ggplot2::labs(x = "Voriconazole concentration (ug/mL)",
                y = "Day-3 trough, fold change vs zero voriconazole",
                colour = NULL) +
  ggplot2::theme_bw()
```

## Serum creatinine (Zhao 2024 Table 6)

```{r table6}
cGrid <- c(40, 100, 160, 400, 600, 800, 1000, 1600, 1800, 2000)
# Published Table 6, tacrolimus 5.5 mg q12h row (mean day-3 trough, ng/mL)
table6Pub <- c(6.46, 8.58, 9.73, 12.42, 13.32, 14.16, 14.44, 15.46, 15.94, 15.60)

table6 <- data.frame(
  creat     = cGrid,
  published = table6Pub,
  model     = vapply(cGrid, function(cc) day3Trough(0, cc, 5.5), numeric(1))
) |>
  dplyr::mutate(
    pubFold   = published / published[1],
    modelFold = model / model[1]
  )

knitr::kable(
  table6 |>
    dplyr::rename(
      "Serum creatinine (umol/L)" = creat,
      "Published (ng/mL)"         = published,
      "This model (ng/mL)"        = model,
      "Published fold vs 40"      = pubFold,
      "This model fold vs 40"     = modelFold
    ),
  digits = 2,
  caption = "Replicates Zhao 2024 Table 6, 5.5 mg q12h row: day-3 trough vs serum creatinine at zero voriconazole."
)

foldRatio <- table6$modelFold[length(cGrid)] / table6$pubFold[length(cGrid)]
foldRatio

stopifnot(
  # Monotone increase in serum creatinine, as the paper reports.
  all(diff(table6$model) > 0),
  # The 40 -> 2000 umol/L fold-rise matches the published table within 10%.
  abs(foldRatio - 1) < 0.10,
  # Every intermediate fold-change is within 10% of the published fold-change.
  max(abs(table6$modelFold / table6$pubFold - 1)) < 0.10
)
```

Placing the -0.40 exponent on **CL/F** reproduces the published Table 6
fold-change profile closely: over the full 40-2000 umol/L grid the model
predicts a `r round(table6$modelFold[length(cGrid)], 2)`-fold rise in day-3
trough against the published `r round(table6$pubFold[length(cGrid)], 2)`-fold,
and every intermediate point agrees within 10%. Placing the same exponent on
V/F instead gives only a 1.80-fold rise at 72 h and inverts the direction
entirely by the time steady state is approached, so it cannot generate Table 6.
See *Assumptions and deviations* -- the paper is internally inconsistent about
which parameter this coefficient modifies.

# Simulated cohort

A 100-subject cohort at the cohort-median covariates, dosed 3 mg q12h, with the
published inter-individual and residual variability.

```{r cohort}
rxode2::rxSetSeed(20240910)
nSub <- 100L

evCohort <-
  dplyr::bind_rows(
    tidyr::expand_grid(id = seq_len(nSub), time = dosingT) |>
      dplyr::mutate(amt = doseMg, evid = 1L, cmt = "depot"),
    tidyr::expand_grid(id = seq_len(nSub), time = seq(0, 84, by = 2)) |>
      dplyr::mutate(amt = NA_real_, evid = 0L, cmt = "central")
  ) |>
  dplyr::mutate(CONC_VORI_NGML = 0, CREAT = 237) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

simCohort <- rxode2::rxSolve(mod, evCohort, returnType = "data.frame")

at72 <- dplyr::filter(simCohort, time == 72)

# ipredSim is the individual prediction (inter-individual variability only);
# sim additionally carries the additive residual error.
rbind(
  ipred = quantile(at72$ipredSim, c(0.05, 0.25, 0.5, 0.75, 0.95)),
  sim   = quantile(at72$sim,      c(0.05, 0.25, 0.5, 0.75, 0.95))
)
```

```{r cohort-plot, echo = FALSE}
simCohort |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    p05 = quantile(ipredSim, 0.05),
    p50 = median(ipredSim),
    p95 = quantile(ipredSim, 0.95),
    .groups = "drop"
  ) |>
  ggplot2::ggplot(ggplot2::aes(time)) +
  ggplot2::geom_ribbon(ggplot2::aes(ymin = p05, ymax = p95), alpha = 0.2) +
  ggplot2::geom_line(ggplot2::aes(y = p50), linewidth = 1) +
  ggplot2::labs(
    x = "Time (h)", y = "Tacrolimus, whole blood (ng/mL)",
    title = "3 mg q12h, 100 subjects: median and 5th-95th percentile of individual predictions"
  ) +
  ggplot2::theme_bw()
```

Two things are worth stating plainly here rather than glossed over.

**The additive residual drives simulated observations negative.** With
`sigma = 3.50` ng/mL and a typical day-3 trough of about 3.6 ng/mL at 3 mg
q12h, the lower tail of `sim` is below zero. That is an inherent consequence of
an additive error model fitted to a variable whose typical value is
single-digit ng/mL, not an extraction error -- the paper reports exactly this
sigma against a median observed concentration of 7.90 ng/mL, and its own
Discussion flags the residual as unexpectedly large. Users simulating
observations from this model should truncate at zero.

**The model's day-3 trough runs below the cohort's observed concentrations.**
At the cohort-median dose of 3 mg q12h the model predicts a day-3 trough near
3.6 ng/mL, against a median *observed* tacrolimus concentration of 7.90 ng/mL
in Table 1. Part of that gap is timing -- the observed samples have a median of
8 days post-operation, closer to steady state than day 3, and the model's
steady-state trough at the same dose is about 5.3 ng/mL -- and part of it is
the same systematic offset that separates this model from the paper's own
Table 5 and Table 6 (see *Errata*). Nothing has been adjusted to close it.

```{r cohort-assert}
kelRef <- clRef / vcRef
ssTrough <- 1000 * doseMg / vcRef * kaRef / (kaRef - kelRef) *
  (exp(-kelRef * 12) / (1 - exp(-kelRef * 12)) -
     exp(-kaRef * 12) / (1 - exp(-kaRef * 12)))

c(day3Typical       = day3Trough(0, 237, doseMg),
  steadyStateTypical = ssTrough,
  cohortMedianIpred  = median(at72$ipredSim),
  cohortObservedMedian = 7.90)

stopifnot(
  # Centre of the individual-prediction distribution, with headroom on both
  # sides; deliberately not an assertion on the tails, which are not stable
  # across rxode2 builds.
  median(at72$ipredSim) > 2,
  median(at72$ipredSim) < 6
)
```

# PKNCA validation

Zhao 2024 reports no non-compartmental analysis, so there is no published
Cmax / Tmax / AUC / half-life table to compare against. The NCA is therefore
used as an internal-identity check: for a single oral dose of a linear
one-compartment model with no residual error, PKNCA must recover
`AUC(0-inf) = Dose / (CL/F)` and `t(1/2) = ln(2) * (V/F) / (CL/F)` for each
voriconazole level, which independently confirms that the covariate terms are
wired onto the parameters the model file claims.

```{r pknca}
ncaArms <- data.frame(
  treatment = c("C(VRC) 0 ug/mL", "C(VRC) 1 ug/mL", "C(VRC) 2 ug/mL"),
  cvrc      = c(0, 1, 2)
)
ncaDose <- 5

ncaTimes <- sort(unique(c(seq(0, 4, by = 0.1), seq(4, 24, by = 1),
                          seq(24, 480, by = 6))))

evNca <-
  ncaArms |>
  dplyr::rowwise() |>
  dplyr::reframe(
    treatment = treatment,
    cvrc      = cvrc,
    dplyr::bind_rows(
      data.frame(time = 0,        amt = ncaDose,   evid = 1L, cmt = "depot"),
      data.frame(time = ncaTimes, amt = NA_real_,  evid = 0L, cmt = "central")
    )
  ) |>
  dplyr::mutate(
    id             = as.integer(factor(treatment)),
    CONC_VORI_NGML = cvrc * 1000,
    CREAT          = 237
  ) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

simNca <- rxode2::rxSolve(typ, evNca, returnType = "data.frame") |>
  dplyr::left_join(
    dplyr::distinct(evNca, id, treatment, cvrc),
    by = "id"
  )

concData <- simNca |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, treatment, time, Cc)

doseData <- concData |>
  dplyr::distinct(id, treatment) |>
  dplyr::mutate(time = 0, dose_mg = ncaDose)

ncaIntervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, auclast = TRUE,
  aucinf.obs = TRUE, half.life = TRUE
)

ncaRes <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(concData, Cc ~ time | treatment + id),
    PKNCA::PKNCAdose(doseData, dose_mg ~ time | treatment + id),
    intervals = ncaIntervals
  )
)

ncaWide <- as.data.frame(ncaRes) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

expected <- ncaArms |>
  dplyr::mutate(
    clExp = 42.87 * exp(-0.28 * cvrc),
    vcExp = 2690 * exp(-0.20 * cvrc),
    aucExp = 1000 * ncaDose / clExp,        # mg/L*h -> ng/mL*h
    thalfExp = log(2) * vcExp / clExp
  )

ncaCompare <- ncaWide |>
  dplyr::left_join(expected, by = "treatment") |>
  dplyr::mutate(
    aucPctDiff   = 100 * (aucinf.obs - aucExp) / aucExp,
    thalfPctDiff = 100 * (half.life - thalfExp) / thalfExp
  )

knitr::kable(
  ncaCompare |>
    dplyr::select(treatment, cmax, tmax, aucinf.obs, aucExp, aucPctDiff,
                  half.life, thalfExp, thalfPctDiff) |>
    dplyr::rename(
      "Arm"                       = treatment,
      "Cmax (ng/mL)"              = cmax,
      "Tmax (h)"                  = tmax,
      "AUC0-inf NCA (ng/mL*h)"    = aucinf.obs,
      "AUC0-inf Dose/CL"          = aucExp,
      "AUC % diff"                = aucPctDiff,
      "t1/2 NCA (h)"              = half.life,
      "t1/2 ln2*V/CL (h)"         = thalfExp,
      "t1/2 % diff"               = thalfPctDiff
    ),
  digits = 2,
  caption = "Single 5 mg oral dose: PKNCA against the model's own closed-form AUC and half-life at three voriconazole concentrations."
)

stopifnot(
  max(abs(ncaCompare$aucPctDiff)) < 1,
  max(abs(ncaCompare$thalfPctDiff)) < 1,
  # Exposure rises with voriconazole, as the paper reports.
  all(diff(ncaCompare$aucinf.obs) > 0)
)
```

Both identities hold to well under 1%, and AUC rises monotonically with
voriconazole concentration, confirming that `e_conc_vori_ngml_cl` and
`e_conc_vori_ngml_vc` act on CL/F and V/F respectively and in the direction the
paper reports.

# Assumptions and deviations

**The paper prints no equations.** Zhao 2024 contains no display equations of
any kind -- not for the structural model, not for the random-effect model, not
for the covariate model. The only description of the covariate functional forms
is the Table 3 and Table 4 footnote wording, "exponent for `<covariate>`, as a
covariate for `<parameter>`". Everything below follows from that wording plus
the covariate distributions in Table 1 and the simulation settings in section
3.4. None of it is tuned to a target.

1. **Voriconazole enters exponentially, not as a power.** The cohort median
   voriconazole concentration is exactly 0.00 ug/mL (Table 1) and many samples
   were drawn before voriconazole was started, so no `(C_VRC / median)^theta`
   form exists. The model uses `exp(theta * C_VRC)`, which equals 1 at
   `C_VRC = 0` -- the untreated reference -- and is the only Phoenix NLME
   continuous-covariate form compatible with a zero-valued covariate.

2. **Serum creatinine enters as a power normalised to 237 umol/L.** CREAT is
   strictly positive, the footnote calls the coefficient an exponent, and
   section 3.4.1 states the authors held CREA "fixed at the median of 237
   umol/L" for the voriconazole simulation, which identifies 237 as the
   normalising constant. An exponential form on the raw umol/L scale is
   arithmetically impossible here (`exp(-0.40 * 237)` underflows to zero).

3. **The serum-creatinine coefficient is placed on CL/F, and the paper
   contradicts itself about this.** The label `Theta CREA-V` and its footnote
   ("exponent for CREA, as a covariate for V") appear in both Table 3 and
   Table 4, and the Discussion says "higher CREA levels were associated with a
   further reduction in V/F". But the Abstract Results ("higher serum
   creatinine levels were associated with lower tacrolimus CL/F") and the
   section 5 Conclusion ("higher CREA also led to a reduction in tacrolimus
   CL/F") both say CL/F. CL/F is used here because the paper's own Table 6
   simulation adjudicates it, subject to one stated reading of the paper.

   The placement is confounded with the trough time, so both have to be
   scanned. Comparing each placement's predicted fold-change profile across the
   published 40-2000 umol/L grid against Table 6's:

   | Trough time (q12h doses) | CREA on CL/F: worst fold error | CREA on V/F: worst fold error |
   |---|---|---|
   | 36 h (3 doses, day 2)     | 29.1% | 4.1% |
   | 48 h (4 doses, day 3)     | 20.8% | 10.5% |
   | 60 h (5 doses, day 3)     | 12.4% | 19.5% |
   | **72 h (6 doses, day 3)** | **4.0%** | 26.8% |
   | 84 h (7 doses, day 4)     | 7.8% | 32.6% |

   The V/F placement reproduces Table 6 only at 36 h, which is day *2*. The
   CL/F placement reproduces it at 72 h -- the end-of-day-3 trough, after six
   complete q12h doses, which is the natural reading of "trough concentration
   on the third day after tacrolimus administration" (section 3.4.1) and of the
   Table 5 and Table 6 titles. At 72 h the CL/F placement predicts a 2.40-fold
   rise against the published 2.42-fold with every intermediate point inside
   10%, while the V/F placement gives only 1.80-fold; and because V/F does not
   set the steady-state trough, the V/F placement keeps degrading with time and
   eventually reverses the direction of the creatinine effect, contradicting
   the paper's own conclusion.

   This is the one substantive judgement call in the extraction, and it goes
   against a table label the paper repeats four times. A user who prefers that
   label can flip it by moving `e_creat_cl` from the `cl` line onto the `vc`
   line in `model()`; the coefficient value is unchanged either way, and the
   Table 6 assertion in this vignette will then fail loudly, which is the
   intended behaviour.

4. **`Ka` is an absorption, not an elimination, rate constant.** The Abstract
   and section 3.2.1 both call 8.39 1/h "the elimination rate constant (Ka)".
   Table 7 lists the same value under a column whose footnote reads "Ka,
   absorption rate constant", alongside Ka values of 0.419-13.1 1/h from six
   comparator tacrolimus models. An elimination rate constant of 8.39 1/h
   implies a 5-minute half-life, irreconcilable with the reported CL/F and V/F
   (which give kel = 0.0159 1/h, t1/2 = 43.5 h).

5. **`sigma = 3.50` is read as a standard deviation in ng/mL.** Table 4 gives
   a bare `sigma` with no unit. Phoenix NLME reports an additive residual as a
   standard deviation, and 3.50 ng/mL against a median observed concentration
   of 7.90 ng/mL is consistent with the Discussion's remark that the sigma
   value indicates substantial unexplained variability.

6. **`omega^2` values are used directly as variances.** Table 4 labels them
   `omega 2 V` and `omega 2 CL`, i.e. already on the variance scale, so no
   `log(CV^2 + 1)` conversion is applied. The Table 4 "CV%" entries for those
   two rows (6.67 and 6.32) are relative standard errors of the estimates, not
   distribution CVs -- the variances themselves imply CVs of 14% and 42%.

7. **Whole blood, not plasma.** Section 2.2 states tacrolimus was measured in
   whole-blood samples by chemiluminescence microparticle immunoassay. The
   Discussion's passing phrase "the population's typical V/F, derived from
   plasma concentration data" contradicts the Methods and is not followed.

## Errata and unreproducible published output

* **The published Monte Carlo tables cannot be reproduced from the published
  final-model parameters.** With any standard covariate parameterisation, the
  day-3 troughs in Table 5 and Table 6 sit systematically above what the
  reported `Ka`, `V/F`, `CL/F` and covariate coefficients generate (about
  1.4-fold for Table 5, about 1.6-fold for Table 6), and the two tables
  disagree with each other by about 19% at the one covariate combination they
  share (`C_VRC = 0`, CREA = 237 umol/L, 5.5 mg q12h: Table 5 gives 9.06 ng/mL,
  Table 6 interpolates to about 10.8 ng/mL). The Table 6 *fold-change* profile
  is reproduced within 10% (see above); the Table 5 fold-change profile is not
  -- the published curve is flatter than the model below the observed 3.38
  ug/mL voriconazole ceiling and much steeper above it. The extraction encodes
  the reported parameters; it does not attempt to reverse-engineer whatever
  additional structure or settings produced the simulation tables, and no
  parameter has been adjusted to reduce these gaps.
* **Table 1 lists the CYP2C19 genotype level `*1/*1` twice** (n = 6 and n = 1)
  where one of the two rows is presumably `*2/*2` or `*1/*17`. Neither
  genotype is in the final model, so this does not affect the extraction.
* **Table 3's base-model confidence intervals are inconsistent with their point
  estimates** (e.g. `V/F` 5291 L with a 2.5%-97.5% CI of 3.50-7.09, and `CL/F`
  32.14 L/h with a CI of 0.02-0.04); the columns appear to be on different
  scales. Only final-model values are extracted, and those are corroborated by
  the Table 4 bootstrap summary, which is internally consistent.
* **Section 3.2.1 reports base-model CV% values (13.31%, 22.91%, 18.05%) that
  do not appear in Table 3** (which lists 0.00, 17.18 and 17.15 for the same
  three parameters). Only final-model values are extracted.
* **The supplement was not retrieved.** Zhao 2024 references Supplementary
  Tables S1-S3 and Supplementary Figures S1-S3. All six are probability-of-
  target-attainment summaries and heat maps derived from the Table 5 and
  Table 6 simulations; none contains a model parameter, an equation, or a
  covariate definition, so no extracted value depends on them.
* **Section 3.4.2 states voriconazole was "fixed at to 0 umol/L"**; the unit is
  a typographical error for ug/mL (voriconazole is reported in ug/mL
  throughout Table 1 and Table 5).
