---
title: "Vancomycin (Zhao 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vancomycin (Zhao 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhao_2025_vancomycin"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhao_2025_vancomycin"))$description`
* Article: [Microbiol Spectr 2025;13(6):e02274-24](https://doi.org/10.1128/spectrum.02274-24)

```{r load-model}
mod <- readModelDb("Zhao_2025_vancomycin")
ui  <- rxode2::rxode(mod)
```

# Population

The model was built from 276 vancomycin concentrations in 126 neonates
admitted to the neonatal intensive care unit of Northwest Women's and
Children's Hospital (Xi'an, Shaanxi, China) between January 2019 and
December 2023. Inclusion required postnatal age at or below 28 days, at
least one measured vancomycin concentration, and at least three days of
vancomycin treatment. Neonates with congenital renal dysplasia or with
chronic or acute renal insufficiency were excluded, as were creatinine
values drawn within seven days of birth (which still reflect maternal
rather than neonatal renal function). Extremely preterm infants were
not included, which is what the title's "non-extremely preterm" refers
to.

Median gestational age at delivery was 32 weeks (IQR 29.38-37.85) and
median postmenstrual age at the start of vancomycin was 35.7 weeks
(IQR 32.9-39.98); 93/126 (73.8%) were preterm. Median birth weight was
1.45 kg and median weight at the start of vancomycin 1.98 kg
(IQR 1.35-2.98). 86/126 (68.3%) were male. Median serum creatinine was
32.48 umol/L (IQR 24.54-42.31), median daily fluid input 364 mL
(IQR 252.91-473.10) and median daily urine volume 200.5 mL. 42/126
(33.3%) received a concomitant diuretic. Vancomycin was given at
10-15 mg/kg per dose every 8 to 12 hours as a one-hour infusion, with
peaks and troughs drawn 0.5 h after and 0.5 h before an infusion, all
at steady state after at least four doses (Zhao 2025 Table 1 and
Materials and Methods).

The cohort was split by calendar time rather than at random: 112
patients (January 2019 to June 2023) built the model and 14 patients
(July 2023 to December 2023, 24 concentrations) were held out for
external validation, on which the paper reports a mean prediction error
of 2.74%, a mean absolute prediction error of 17.48%, and 75.00% /
83.33% of predictions within 20% / 30%.

The same information is available programmatically via
`rxode2::rxode(readModelDb("Zhao_2025_vancomycin"))$population`.

# The missing exponents

This paper needs one adjudication before it can be encoded, and it is
worth stating up front because it changes the model materially.

Zhao 2025 prints its final model as equations 6 and 7:

```
CL(L/h) = 0.14 x (WT/2.12) x (Scr/30.52) x (DFI/367.18) x e^A x exp(etaCL)
V       = 1.04 x (WT/2.12)
```

Every parenthesised covariate ratio is missing its superscript
exponent. This is **not** a text-extraction artifact: the EuropePMC
open-access bundle for PMC12131725 ships equations 1-7 as publisher
artwork (`spectrum.02274-24.m001.jpg` through `m007.jpg`), and the
rendered images for equations 6 and 7 carry no superscripts either. The
exponents were lost in production.

Table 3 supplies them -- rows `WTonCL` = 1.13, `Scr on CL` = -0.15,
`DFI on CL` = 0.14, and `WTonV` = 1.07 -- and the Methods state that
"an exponential model was used to evaluate the influence of covariates
on pharmacokinetic parameters", i.e. covariate ratios raised to an
estimated power. Two independent checks confirm that reading against
the paper's own numbers.

**Check 1: reading the equations literally is falsified by Table 4.**
If every exponent were really 1, then Table 4's recommended daily dose
(which the paper derives from clearance via equation 5,
`Dose = 400 x CL`) would rise ninefold from Scr = 10 to Scr = 90
umol/L, and sevenfold from DFI = 100 to DFI = 700 mL. Table 4 does the
opposite on creatinine and moves by 30% on fluid input.

**Check 2: the Table 3 exponents can be recovered from Table 4.**
Because Table 4 is a pure function of the clearance equation, taking
log-ratios across its rows and columns reads the exponents back out.

```{r missing-exponents}
# Table 4 of Zhao 2025, "Non-DA" (no concomitant diuretic) column,
# mg/kg/day, transcribed as printed.
tab4 <- tidyr::expand_grid(
  CREAT        = c(10, 30, 50, 70, 90),
  FLUID_IN_24H = c(100, 250, 400, 550, 700)
) |>
  dplyr::mutate(
    dose_DA    = c(26, 29, 31, 33, 34, 22, 25, 27, 28, 29, 21, 23, 25, 26, 27,
                   20, 22, 24, 25, 26, 19, 21, 23, 24, 25),
    dose_nonDA = c(32, 36, 38, 39, 41, 27, 30, 32, 34, 35, 25, 28, 30, 31, 32,
                   24, 27, 29, 30, 31, 23, 26, 28, 29, 30)
  )

# Exponent recovered from the creatinine axis, at each fixed fluid input.
scr_ref <- tab4 |>
  dplyr::group_by(FLUID_IN_24H) |>
  dplyr::summarise(
    e_creat = mean(log(dose_nonDA[CREAT != 10] / dose_nonDA[CREAT == 10]) /
                     log(CREAT[CREAT != 10] / 10)),
    .groups = "drop"
  )

# Exponent recovered from the fluid-input axis, at each fixed creatinine.
dfi_ref <- tab4 |>
  dplyr::group_by(CREAT) |>
  dplyr::summarise(
    e_dfi = mean(log(dose_nonDA[FLUID_IN_24H != 100] /
                       dose_nonDA[FLUID_IN_24H == 100]) /
                   log(FLUID_IN_24H[FLUID_IN_24H != 100] / 100)),
    .groups = "drop"
  )

# Diuretic effect recovered from the two dose columns.
diuretic_factor <- with(tab4, mean(dose_DA / dose_nonDA))

recovered <- tibble::tibble(
  Quantity = c("Serum-creatinine exponent on CL",
               "Daily-fluid-input exponent on CL",
               "Diuretic multiplier on CL"),
  `Zhao 2025 Table 3` = c(-0.15, 0.14, exp(-0.20)),
  `Recovered from Table 4` = c(mean(scr_ref$e_creat),
                               mean(dfi_ref$e_dfi),
                               diuretic_factor),
  `Literal reading of eq. 6` = c(1, 1, NA_real_)
)
knitr::kable(recovered, digits = 4,
             caption = paste("Covariate exponents recovered from the Zhao 2025 Table 4",
                             "dosing grid, versus the Table 3 estimates and versus",
                             "reading equation 6 literally."))

stopifnot(
  # The recovered exponents sit close to the Table 3 estimates. Table 4 is
  # printed to whole mg/kg/day, so a 1 mg rounding on a ~25 mg cell is ~4%
  # and the recovered exponents carry that much noise. Nothing here is
  # simulated -- both sides are fixed transcriptions -- so these bounds are
  # exactly reproducible and are set just outside the realised residuals
  # (0.0037, 0.0107, 0.0091).
  abs(mean(scr_ref$e_creat) - (-0.15)) < 0.02,
  abs(mean(dfi_ref$e_dfi) - 0.14) < 0.02,
  abs(diuretic_factor - exp(-0.20)) < 0.015,
  # And the literal reading is nowhere near: it has the creatinine effect
  # pointing the wrong way entirely.
  mean(scr_ref$e_creat) < 0
)
```

The packaged model therefore encodes equations 6 and 7 **with** the
Table 3 exponents:

```
CL = 0.14 * (WT/2.12)^1.13 * (CREAT/30.52)^-0.15 * (FLUID_IN_24H/367.18)^0.14 * exp(-0.20 * CONMED_DIURETIC) * exp(etaCL)
V  = 1.04 * (WT/2.12)^1.07
```

At the reference covariate values every ratio is 1, so `CL` reduces to
0.14 L/h -- exactly the "typical CL value of 0.14 L/hour" the Results
paragraph after equation 7 reports, which is a third consistency check
on the encoding.

# Source trace

Every numeric value in `ini()` carries an in-file comment pointing to
the Zhao 2025 source location. The table below collects them in one
place for review.

| Equation / parameter          | Value       | Source location                                                            |
|-------------------------------|-------------|----------------------------------------------------------------------------|
| Structural model              | 1 compartment, first-order elimination | Base model section; Table 3 reports only CL and V |
| Residual error model          | proportional | Results: "The examination of residual variability supported the adoption of a proportional error model" (equation 3) |
| IIV model                     | exponential | Equation 1, `P_i = P_TV * exp(eta_i)`                                      |
| `lcl` (CL)                    | 0.14 L/h    | Table 3, final model row "CL (L/hour)" (RSE 3.15%)                          |
| `lvc` (V)                     | 1.04 L      | Table 3, final model row "V (L)" (RSE 4.28%)                                |
| `e_wt_cl`                     | 1.13        | Table 3, final model row "WTonCL(L/hour)" (RSE 5.64%)                        |
| `e_creat_cl`                  | -0.15       | Table 3, final model row "Scr on CL (L/hour)" (RSE 31.34%)                   |
| `e_fluid_in_24h_cl`           | 0.14        | Table 3, final model row "DFI on CL (L/hour)" (RSE 27.03%)                   |
| `e_conmed_diuretic_cl`        | -0.20       | Table 3 row "DA on CL"; text after equation 6: "When diuretics were used concomitantly, A = -0.20" |
| `e_wt_vc`                     | 1.07        | Table 3, final model row "WTonV(L)" (RSE 7.61%)                              |
| `etalcl` (4.97% CV)           | 0.00246704  | Table 3, final model row "CL (%CV)"; also stated in prose in Results         |
| `propSd`                      | 0.180       | Table 3, final model row "Proportional (%CV)" = 18.0 (RSE 5.76%)             |
| WT reference 2.12 kg          |             | Equation 6 and equation 7 (cohort mean)                                      |
| Scr reference 30.52 umol/L    |             | Equation 6 (cohort mean)                                                     |
| DFI reference 367.18 mL/24h   |             | Equation 6 (cohort mean)                                                     |

# Virtual cohort

The virtual cohort reproduces the Table 1 marginal distributions of the
four model covariates. Weight, creatinine and fluid input are drawn
log-normally with the published median and interquartile range, and
truncated to the observed range; diuretic use is drawn at the published
33.3% prevalence and then split into two equal arms so that the
diuretic contrast can be read directly.

```{r cohort}
rxode2::rxSetSeed(20250514)   # Zhao 2025 publication date
set.seed(20250514)

n_per_arm <- 100L   # 200 subjects total, at the 200-per-arm vignette cap

# Draw a log-normal variate matching a published median and IQR.
rlnorm_iqr <- function(n, median, q25, q75, lower, upper) {
  sdlog <- (log(q75) - log(q25)) / (2 * stats::qnorm(0.75))
  x <- stats::rlnorm(n, meanlog = log(median), sdlog = sdlog)
  pmin(pmax(x, lower), upper)
}

subj <- tibble::tibble(
  id = seq_len(2L * n_per_arm),
  # Zhao 2025 Table 1: WT median 1.98 kg (IQR 1.35-2.98).
  WT = rlnorm_iqr(2L * n_per_arm, 1.98, 1.35, 2.98, lower = 0.8, upper = 5.0),
  # Table 1: Scr median 32.48 umol/L (IQR 24.54-42.31).
  CREAT = rlnorm_iqr(2L * n_per_arm, 32.48, 24.54, 42.31, lower = 8, upper = 95),
  # Table 1: DFI median 364 mL (IQR 252.91-473.10).
  FLUID_IN_24H = rlnorm_iqr(2L * n_per_arm, 364, 252.91, 473.10,
                            lower = 80, upper = 750),
  # Two equal arms so the diuretic contrast is directly readable; the
  # cohort prevalence in Table 1 is 33.3%.
  CONMED_DIURETIC = rep(c(0, 1), each = n_per_arm),
  arm = factor(rep(c("No diuretic", "Concomitant diuretic"), each = n_per_arm),
               levels = c("No diuretic", "Concomitant diuretic"))
) |>
  # Zhao 2025 Materials and Methods: 10-15 mg/kg per dose. Use the top of the
  # band, rounded to whole milligrams as it would be prescribed.
  dplyr::mutate(dose_mg = round(15 * WT))

knitr::kable(
  subj |>
    dplyr::group_by(arm) |>
    dplyr::summarise(
      n = dplyr::n(),
      `WT (kg), median [IQR]` = sprintf("%.2f [%.2f-%.2f]", median(WT),
                                        quantile(WT, .25), quantile(WT, .75)),
      `Scr (umol/L), median [IQR]` = sprintf("%.1f [%.1f-%.1f]", median(CREAT),
                                             quantile(CREAT, .25), quantile(CREAT, .75)),
      `DFI (mL/24h), median [IQR]` = sprintf("%.0f [%.0f-%.0f]",
                                             median(FLUID_IN_24H),
                                             quantile(FLUID_IN_24H, .25),
                                             quantile(FLUID_IN_24H, .75)),
      .groups = "drop"
    ),
  caption = "Virtual cohort covariate distributions (compare Zhao 2025 Table 1: WT 1.98 [1.35-2.98] kg, Scr 32.48 [24.54-42.31] umol/L, DFI 364 [253-473] mL)."
)
```

# Simulation

Vancomycin 15 mg/kg is given every 12 hours as a one-hour intravenous
infusion for eleven doses. Concentrations are sampled coarsely through
the accumulation phase and finely over the 96-120 h interval, which is
the steady-state 24-hour window used for NCA below. Observation rows
carry `cmt = "central"`, the model's ODE state.

```{r simulate}
dose_times <- seq(0, 120, by = 12)

dosing <- subj |>
  tidyr::crossing(time = dose_times) |>
  dplyr::mutate(
    evid = 1L,
    amt  = dose_mg,
    rate = dose_mg / 1,   # 1-hour IV infusion (Zhao 2025 Materials and Methods)
    cmt  = "central"
  )

obs_times <- sort(unique(c(seq(0, 96, by = 4), seq(96, 120, by = 0.25))))

obs <- subj |>
  tidyr::crossing(time = obs_times) |>
  dplyr::mutate(
    evid = 0L,
    amt  = NA_real_,
    rate = NA_real_,
    cmt  = "central"
  )

events <- dplyr::bind_rows(dosing, obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  as.data.frame()

sim <- rxode2::rxSolve(
  mod,
  events = events,
  keep = c("WT", "CREAT", "FLUID_IN_24H", "CONMED_DIURETIC", "arm", "dose_mg")
) |>
  as.data.frame()

stopifnot(
  # Fail loudly rather than silently validating nothing (failure pattern 10).
  nrow(sim) > 0,
  all(c("Cc", "cl", "vc") %in% names(sim)),
  !anyNA(sim$Cc),
  all(sim$Cc >= 0)
)
```

# Replicate published figures

Zhao 2025 Figure 1 plots observed individual clearance against serum
creatinine (panel A) and against daily fluid input (panel B), split by
diuretic use. The individual clearances themselves are not published,
but the typical-value surfaces the model implies over those two axes
are, and they are what the figure is asserting.

```{r figure-1, fig.width = 7, fig.height = 3.4}
cl_typ <- function(WT, CREAT, FLUID_IN_24H, CONMED_DIURETIC) {
  0.14 * (WT / 2.12)^1.13 * (CREAT / 30.52)^-0.15 *
    (FLUID_IN_24H / 367.18)^0.14 * exp(-0.20 * CONMED_DIURETIC)
}

fig1 <- dplyr::bind_rows(
  tidyr::expand_grid(CREAT = seq(10, 90, by = 1), CONMED_DIURETIC = c(0, 1)) |>
    dplyr::mutate(panel = "A: serum creatinine",
                  x = CREAT,
                  cl = cl_typ(2.12, CREAT, 367.18, CONMED_DIURETIC)),
  tidyr::expand_grid(FLUID_IN_24H = seq(100, 700, by = 5),
                     CONMED_DIURETIC = c(0, 1)) |>
    dplyr::mutate(panel = "B: daily fluid input",
                  x = FLUID_IN_24H,
                  cl = cl_typ(2.12, 30.52, FLUID_IN_24H, CONMED_DIURETIC))
) |>
  dplyr::mutate(Diuretic = ifelse(CONMED_DIURETIC == 1, "Yes", "No"))

ggplot(fig1, aes(x, cl, colour = Diuretic)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ panel, scales = "free_x") +
  labs(
    x = "Covariate value (umol/L in panel A, mL/24h in panel B)",
    y = "Typical CL (L/h) at WT = 2.12 kg",
    title = "Figure 1 - typical-value clearance vs creatinine and fluid input",
    caption = "Replicates the two relationships in Figure 1 of Zhao 2025 at the reference body weight. Clearance falls with creatinine, rises with fluid input, and is uniformly lower with a concomitant diuretic."
  )
```

```{r profile, fig.width = 7, fig.height = 3.6}
sim |>
  dplyr::filter(time <= 60) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
    Q95 = quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  facet_wrap(~ arm) +
  labs(
    x = "Time after first dose (h)", y = "Vancomycin Cc (mg/L)",
    title = "Simulated vancomycin concentration-time profile",
    caption = "15 mg/kg as a 1-hour IV infusion Q12H. Median with 5th-95th percentile band across 100 virtual neonates per arm."
  )
```

# Structural checks

Four checks on the packaged encoding, all deterministic or
closed-form.

```{r checks-structure}
# --- Check 1: the diuretic effect is exactly exp(-0.20) -------------------
# Same covariates, diuretic toggled: the ratio must be the published factor.
diuretic_ratio <- cl_typ(2.12, 30.52, 367.18, 1) / cl_typ(2.12, 30.52, 367.18, 0)

# --- Check 2: at the reference covariates, CL and V are the Table 3 values --
cl_at_ref <- cl_typ(2.12, 30.52, 367.18, 0)
vc_at_ref <- 1.04 * (2.12 / 2.12)^1.07

# --- Check 3: interindividual variability reads back off the simulation ----
# The model carries exactly ONE eta, so eta_i is recoverable from the solved
# individual clearance by dividing out the (deterministic) covariate model.
etas <- sim |>
  dplyr::distinct(id, cl, WT, CREAT, FLUID_IN_24H, CONMED_DIURETIC) |>
  dplyr::mutate(
    eta = log(cl / cl_typ(WT, CREAT, FLUID_IN_24H, CONMED_DIURETIC))
  )
omega_readback <- sd(etas$eta)
omega_published <- sqrt(log(1 + 0.0497^2))

# --- Check 4: the ODE is actually being solved ---------------------------
# rxode2 auto-solves a one-compartment model presented as a cl/vc pair,
# discarding the explicit d/dt. That is harmless here only because the ODE
# rate IS cl/vc; this check confirms the solved profile really does respond
# to the structural parameters rather than being a frozen artifact.
ev_one <- data.frame(
  id = 1L, time = c(0, seq(0.5, 24, by = 0.5)),
  evid = c(1L, rep(0L, 48)), amt = c(30, rep(NA_real_, 48)),
  rate = c(30, rep(NA_real_, 48)), cmt = "central",
  WT = 2.12, CREAT = 30.52, FLUID_IN_24H = 367.18, CONMED_DIURETIC = 0
)
mod_typ  <- mod |> rxode2::zeroRe()
base_run <- rxode2::rxSolve(mod_typ, ev_one) |> as.data.frame()
fast_run <- rxode2::rxSolve(mod_typ, ev_one,
                            params = c(lcl = log(0.28))) |> as.data.frame()
trough_ratio <- fast_run$Cc[nrow(fast_run)] / base_run$Cc[nrow(base_run)]

knitr::kable(
  tibble::tibble(
    Check = c("Diuretic multiplier on CL",
              "Typical CL at reference covariates (L/h)",
              "Typical V at reference weight (L)",
              "Interindividual SD of log CL",
              "24 h trough ratio when CL is doubled"),
    Model = c(diuretic_ratio, cl_at_ref, vc_at_ref, omega_readback, trough_ratio),
    `Zhao 2025` = c(exp(-0.20), 0.14, 1.04, omega_published, NA_real_)
  ),
  digits = 5,
  caption = "Structural checks against the Zhao 2025 published values."
)

stopifnot(
  # Deterministic identities -- these are exact arithmetic, so they are held
  # to floating-point tolerance.
  abs(diuretic_ratio - exp(-0.20)) < 1e-10,
  abs(cl_at_ref - 0.14) < 1e-10,
  abs(vc_at_ref - 1.04) < 1e-10,
  # Simulated quantity: the sd of 200 draws from a normal has a sampling CV of
  # 1/sqrt(2*199) = 5.0%, so this +/- 30-35% band is about six sampling SDs and
  # holds for any cohort the model can draw at any thread count. It still goes
  # red if the encoded omega were wrong by a factor of two (ratio 0.5 or 2.0).
  # Realised ratio 0.964 on this render.
  omega_readback > 0.70 * omega_published,
  omega_readback < 1.35 * omega_published,
  # Doubling clearance halves the half-life from 5.15 h to 2.57 h, so a 24-hour
  # trough drops by a further ~4.6 half-lives: the expected ratio is ~0.04, and
  # it was 0.042 here. If rxode2 had silently frozen the profile (it auto-solves
  # a cl/vc pair), this ratio would be exactly 1. Deterministic typical-value
  # solve, so the bound is set close to the expected value.
  trough_ratio < 0.15
)
```

# PKNCA validation

Zhao 2025 publishes no NCA table -- it reports model parameters, a
dosing grid, and external-validation prediction errors -- so
`nlmixr2lib::ncaComparisonTable()` has no published Cmax / Tmax / AUC /
half-life to sit beside. The NCA below instead serves as a closed-form
audit of the simulation: at steady state, the AUC over one 24-hour
window must equal the 24-hour dose divided by the individual clearance,
exactly. That identity is independent of the numerical solver and of
the trapezoidal rule, so any disagreement beyond integration error
means the encoded clearance and the solved profile have diverged.

```{r pknca}
ss_start <- 96
ss_end   <- 120

sim_nca <- sim |>
  dplyr::filter(!is.na(Cc), time >= ss_start, time <= ss_end) |>
  dplyr::select(id, time, Cc, arm)

dose_nca <- events |>
  dplyr::filter(evid == 1, time >= ss_start, time < ss_end) |>
  dplyr::select(id, time, amt, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id,
                             concu = "mg/L", timeu = "hr")
dose_obj <- PKNCA::PKNCAdose(dose_nca, amt ~ time | arm + id, doseu = "mg")

intervals <- data.frame(
  start = ss_start, end = ss_end,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE, cav = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj,
                                          intervals = intervals))

knitr::kable(
  summary(nca_res),
  caption = "Simulated steady-state NCA parameters by arm (15 mg/kg Q12H as a 1 h IV infusion; 24-hour window 96-120 h)."
)
```

```{r pknca-closed-form}
auc_nca <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::select(id, arm, auc_nca = PPORRES)

closed_form <- sim |>
  dplyr::distinct(id, arm, cl, dose_mg) |>
  dplyr::mutate(auc_closed = 2 * dose_mg / cl) |>   # two 12-hourly doses per 24 h
  dplyr::inner_join(auc_nca, by = c("id", "arm")) |>
  dplyr::mutate(pct_diff = 100 * (auc_nca - auc_closed) / auc_closed)

stopifnot(
  # Every subject was matched -- a silently empty join would make the
  # assertions below vacuously true (failure pattern 10).
  nrow(closed_form) == nrow(subj),
  # Both sides use the SAME drawn per-subject clearance, so the only
  # difference is trapezoidal integration error on a 0.25 h grid. This is
  # pure numerical error, not cohort variability, so a tight bound is correct
  # (the cohort-spread caveat about tightening to one run does not apply).
  # Realised maximum 0.015%; 0.3% is twenty times that and would still catch
  # any structural divergence between the encoded clearance and the solve.
  max(abs(closed_form$pct_diff)) < 0.3
)

knitr::kable(
  closed_form |>
    dplyr::group_by(arm) |>
    dplyr::summarise(
      `AUC0-24,ss NCA (mg*h/L), median` = median(auc_nca),
      `AUC0-24,ss = Dose/CL (mg*h/L), median` = median(auc_closed),
      `Max |% difference| across subjects` = max(abs(pct_diff)),
      .groups = "drop"
    ),
  digits = 3,
  caption = "Steady-state AUC from PKNCA versus the closed-form Dose/CL identity."
)
```

# Comparison against the published dosing table

Table 4 of Zhao 2025 is the paper's one published quantitative outcome
that the packaged model can be checked against end to end. It gives the
recommended initial vancomycin dose in mg/kg/day for a grid of serum
creatinine, daily fluid input and diuretic use, derived from the final
model through equation 5, `Dose = 400 x CL` (the daily dose achieving
AUC24/MIC = 400 at MIC = 1 mg/L).

Applying equation 5 to the packaged clearance at the reference body
weight reproduces the **structure** of Table 4 essentially exactly --
the ratio of published to model-predicted dose is constant across all
50 cells -- but sits systematically below it in **level**.

```{r table4-comparison}
tab4_cmp <- tab4 |>
  tidyr::pivot_longer(c(dose_DA, dose_nonDA),
                      names_to = "diuretic", values_to = "dose_published") |>
  dplyr::mutate(
    CONMED_DIURETIC = as.numeric(diuretic == "dose_DA"),
    # Equation 5 at the reference body weight, converted to mg/kg/day.
    dose_model = 400 * cl_typ(2.12, CREAT, FLUID_IN_24H, CONMED_DIURETIC) / 2.12,
    ratio = dose_published / dose_model
  )

ratio_cv <- 100 * sd(tab4_cmp$ratio) / mean(tab4_cmp$ratio)

knitr::kable(
  tibble::tibble(
    Quantity = c("Cells compared",
                 "Mean published / model dose ratio",
                 "CV of that ratio across cells (%)",
                 "Range of that ratio"),
    Value = c(as.character(nrow(tab4_cmp)),
              sprintf("%.3f", mean(tab4_cmp$ratio)),
              sprintf("%.2f", ratio_cv),
              sprintf("%.3f-%.3f", min(tab4_cmp$ratio), max(tab4_cmp$ratio)))
  ),
  caption = "Zhao 2025 Table 4 versus equation 5 applied to the packaged model."
)

stopifnot(
  nrow(tab4_cmp) == 50,
  # The covariate STRUCTURE of Table 4 is reproduced: the ratio is flat.
  # Table 4 is printed to whole mg/kg/day, so 1 mg of rounding on the
  # smallest (19 mg/kg/day) cell is already 5%, which is where the residual
  # 1.35% CV comes from. Nothing here is simulated, so the value is exactly
  # reproducible; 2% still fails if any exponent were mis-encoded, because a
  # wrong exponent tilts the ratio along that covariate's axis.
  ratio_cv < 2,
  # ... and the level offset is real and one-directional, not noise.
  mean(tab4_cmp$ratio) > 1.1
)
```

The ratio is `r sprintf("%.3f", mean(tab4_cmp$ratio))` with a CV of
only `r sprintf("%.2f", ratio_cv)`% across all 50 cells. A flat ratio
is a strong statement: it says every covariate exponent and the
diuretic multiplier in the packaged model match the ones that generated
Table 4, because any mis-encoded exponent would tilt the ratio along
that covariate's axis. The remaining constant factor of about 1.21 is a
**level** offset that the paper does not explain, and two mechanisms
are consistent with it:

* **The body weight used to build Table 4 is not stated.** Because the
  weight exponent on clearance is 1.13 rather than 1, dose per kilogram
  is not weight-invariant: it rises as `WT^0.13`, from 24.0 mg/kg/day
  at 1 kg to 29.5 mg/kg/day at 5 kg at the reference covariates. Table
  4 gives one number per cell over a simulated 1-5 kg range, so some
  representative weight was chosen and never reported. Evaluating at
  5 kg rather than at the 2.12 kg cohort mean shrinks the offset from
  about 21% to about 8%, but does not remove it.
* **Table 4 is a Monte Carlo target-attainment dose, not a
  typical-value dose.** The Methods state that each regimen underwent
  5,000 Monte Carlo simulations, so the tabulated dose is presumably
  the one attaining AUC24/MIC >= 400 in some fraction of the simulated
  population rather than in the typical patient. Any such fraction
  above the median pushes the dose above the equation-5 typical value.

Neither mechanism can be pinned down from what is printed, so this is
recorded as a documented deviation rather than absorbed into the model.
It does not affect the packaged parameters: the model file encodes
Table 3 and equations 6-7, and Table 4 is a downstream application of
them.

# Assumptions and deviations

- **The published equations 6 and 7 are incomplete, and the model file
  restores their exponents.** As set out in "The missing exponents"
  above, the covariate ratios are printed with no superscripts in both
  the article text and the publisher's own equation artwork. The
  packaged model uses the Table 3 exponents (1.13, -0.15, 0.14 on CL;
  1.07 on V), which are independently recovered from the Table 4 dosing
  grid in this vignette. This is the single most consequential decision
  in the extraction and any reviewer should check it first.
- **Reference values are cohort means, not medians.** The normalising
  constants in equations 6 and 7 (2.12 kg, 30.52 umol/L,
  367.18 mL/24h) are all slightly different from the Table 1 medians
  (1.98 kg, 32.48 umol/L, 364 mL). They are used exactly as printed.
- **Two Table 3 entries have printing errors that do not affect the
  model.** The relative standard errors for the Scr and DA effects are
  printed as negative numbers (-31.34 and -20.03); an RSE is a
  magnitude. The bootstrap 95% CI for the DFI effect is printed as
  "-0.087 to -0.21", which cannot bracket the positive point estimate
  of 0.13 and is a sign typo for 0.087 to 0.21. Only point estimates
  enter the model file, so neither error propagates.
- **The interindividual variability is unusually small and is encoded
  as published.** A 4.97% CV on clearance is far below what neonatal
  vancomycin models typically report. It is not a transcription error:
  the Results state it in prose, and the bootstrap corroborates it at
  4.80% (95% CI 2.83-6.77). The base model's 24.84% CV is the
  conventional-looking value, and almost all of it is absorbed by the
  four covariates. Users simulating unexplained between-subject spread
  should be aware that this model attributes very little of it to a
  random effect.
- **No interindividual variability on volume.** IIV on V was 0.90% CV
  with an RSE of 215.55% in the base model and was dropped from the
  final model, so V carries no eta. The consequence is that simulated
  peak concentrations are less variable than real neonatal data would
  be.
- **Variability scale.** Table 3 reports IIV and residual variability
  as "%CV" (defined in the Table 3 abbreviation footnote). The eta
  variance is taken as `log(1 + CV^2)`, the exact inverse of a
  log-normal CV; at 4.97% this differs from the naive `omega = CV`
  reading by 0.07%, so the choice is immaterial here.
- **Postmenstrual age is deliberately absent.** PMA was the second
  strongest univariate covariate on clearance (dOFV -113.1) and it
  entered the multivariate model, but it was the first term removed on
  backward elimination. Most neonatal vancomycin models retain a
  maturation term and this one does not; that is the paper's result,
  not an omission in the extraction. PMA and the other screened-but-not-
  retained covariates are recorded in the model file's
  `covariatesDataExcluded` list.
- **Daily fluid input required a new canonical covariate column.**
  `FLUID_IN_24H` (mL/24h) is registered in
  `inst/references/covariate-columns.md` as part of this extraction. It
  is deliberately distinct from `URINE_VOL_24H`, the output counterpart
  over the same window, which Zhao 2025 screened and did **not** retain
  -- collapsing the two would destroy the paper's finding.
- **Diuretic class composition.** `CONMED_DIURETIC` pools furosemide,
  spironolactone and hydrochlorothiazide, per the Materials and Methods
  list. The register requires the pooled class to be enumerated per
  model, and it is, in `covariateData[[CONMED_DIURETIC]]$notes`.
- **Table 4 level offset is documented, not fitted.** See the previous
  section. No parameter was adjusted to close it.
- **Virtual-cohort covariate distributions are reconstructed, not
  published.** Zhao 2025 reports medians and interquartile ranges but
  not the joint covariate distribution, so the cohort here draws each
  covariate independently from a log-normal matched to its published
  median and IQR. The paper reports a weight-DFI correlation of
  R = 0.7274 which this independent draw does not reproduce; that
  affects the spread of simulated exposures but not any of the
  structural checks above, all of which are evaluated per subject or in
  closed form.
- **Assay range.** Concentrations below 2 or above 50 mg/L were
  excluded from the original fit (Materials and Methods, exclusion
  criterion i). The simulation applies no such truncation.
