---
title: "Paracetamol and its metabolites in spinal muscular atrophy (Zhao 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Paracetamol and its metabolites in spinal muscular atrophy (Zhao 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r meta}
mod_meta <- nlmixr2est::nlmixr(readModelDb("Zhao_2025_paracetamol"))$meta
```

* Citation: `r mod_meta$reference`
* Article: <https://doi.org/10.1002/bcp.70028>
* Description: `r mod_meta$description`

# Population

Zhao et al. (2025) analysed a Danish single-centre trial (EudraCT
2018-002295-40) in which six adults with spinal muscular atrophy (SMA), six
children with SMA and 11 healthy controls received oral paracetamol
15 mg/kg every 6 h for three days, capped at a maximum single dose of 1 g
(Methods section 2.1). Blood was sampled hourly for 6-8 h on Days 1 and 3
after a pre-treatment baseline sample, giving 294 plasma samples per analyte
for each of paracetamol, paracetamol-glucuronide, paracetamol-sulphate and
the combined oxidative metabolites (cysteine conjugate + mercapturic acid).

The two groups differ sharply in size and in the two retained biochemical
covariates (Table 1 of the source): median body weight 30.5 kg (range 22-57)
in SMA versus 78.0 kg (51-103) in healthy controls; median plasma myoglobin
17 ng/mL (14-54) versus 34 ng/mL (17-74); median total bilirubin 4 umol/L
(2-25) versus 7 umol/L (3-15). Median age was 17 years (6-37) in the SMA
group and 25 years (20-36) in the controls; 10 of 23 participants (43.5%)
were female.

The same information is available programmatically via
`readModelDb("Zhao_2025_paracetamol")$population`.

# Source trace

Every `ini()` entry in
`inst/modeldb/specificDrugs/Zhao_2025_paracetamol.R` carries an in-file
comment naming its source location. The table below collects them for
review.

```{r source-trace, echo = FALSE}
source_trace <- tibble::tribble(
  ~Component,                    ~Value,             ~"Source location",
  "Structural model",            "1 cmt per compound, first-order absorption + lag, allometric WT", "Figure 1 (schematic) and Results section 3.2",
  "Covariate model form",        "Eq. 1 (categorical, proportional) and Eq. 2 (continuous, median-normalised power)", "Supplementary material 1, Methods for model building",
  "t_lag",                       "0.153 h (FIX)",    "Table 2, row 't lag (h) (FIX)' + table footnote",
  "k_a",                         "2.84 1/h",         "Table 2, final model column",
  "V_pcm/F",                     "63.5 L/70 kg",     "Table 2, final model column",
  "CL_p/F (leftover)",           "8.16 L/h/70 kg",   "Table 2, final model column",
  "CL_pg/F (glucuronide formation)",  "6.37 L/h/70 kg", "Table 2, final model column",
  "CL_ps/F (sulphate formation)",     "8.43 L/h/70 kg", "Table 2, final model column",
  "CL_pox/F (oxidative formation)",   "0.20 L/h/70 kg", "Table 2, final model column",
  "CL_gluc/F (glucuronide elimination)", "5.69 L/h/70 kg", "Table 2, final model column",
  "CL_sulf/F (sulphate elimination)",    "20.4 L/h/70 kg", "Table 2, final model column",
  "CL_ox/F (oxidative elimination)",     "3.72 L/h/70 kg", "Table 2, final model column",
  "Metabolite volumes",          "0.18 * V_pcm/F = 11.43 L/70 kg", "Methods section 2.2 and Table 2 rows V_glu/V_sulf/V_ox",
  "Allometric exponents",        "0.75 on CL, 1 on V (both FIX)", "Table 2 caption and Results section 3.2",
  "Disease effect on V_pcm/F",   "1.58",             "Table 2, 'Covariate effect on V_pcm/F -- Disease'",
  "Myoglobin on CL_p/F",         "-1.10 (per MYO/25)", "Table 2 + Results section 3.3 ('negative correlation')",
  "Bilirubin on CL_ps/F",        "+0.18 (per TBILI/6)", "Table 2 + Results section 3.3 ('positive correlation')",
  "Bilirubin on CL_ox/F",        "-0.177 (per TBILI/6)", "Table 2 + Results section 3.3 ('negative correlation')",
  "BSV (all six etas)",          "omega (log-SD) x 100, per Table 2 'BSV (%)'", "Table 2 + Discussion (CL_p/F CV 105% = sqrt(exp(0.862^2) - 1))",
  "Proportional residual error", "0.28 / 0.18 / 0.22 / 0.17", "Table 2, 'Residual variability -- Proportional'"
)
knitr::kable(source_trace, caption = "Source trace for the Zhao 2025 paracetamol model.")
```

## Two values whose sign the typeset table loses

Table 2 prints the myoglobin exponent as `1.10` and the bilirubin-on-CL_ox
exponent as `0.177`, but their bootstrap confidence intervals are printed in
*descending* order (`3.07 - 0.49` and `0.45 - 0.03`), which is only possible
for negative intervals, and Results section 3.3 states both correlations are
negative. The signs are confirmed numerically by supplementary Table S3:
the paracetamol half-life rises monotonically with myoglobin (2.20 h at
15 ng/mL to 3.69 h at 70 ng/mL, i.e. clearance falls), and the
oxidative-metabolite half-life rises monotonically with bilirubin (3.18 h at
3 umol/L to 4.19 h at 24 umol/L). Both exponents are therefore encoded as
negative.

## Concentration scale: molar, not paracetamol-mass

Zhao 2025 never states the amount scale of its metabolite compartments, and
the choice is load-bearing: paracetamol converts 1:1 on a *molar* basis to
each conjugate, so a parent-metabolite model with a single set of formation
and elimination clearances is only self-consistent when the compartments
carry moles. The paper's own simulation output settles it, because the
steady-state exposure ratio of a metabolite to the parent is
volume-independent -- it equals CL_formation / CL_elimination in whatever
amount unit the compartments carry, so it discriminates the amount scale
without depending on any volume. Reproducing the model on the molar scale
and converting each output to that species' own mass units matches
supplementary Table S2 to within 2.5%; a paracetamol-mass parameterisation
is 34-53% low. The check is run on the simulated 0-24 h window (the same
window Table S2 reports) in the "Structural identity" section below.

# Virtual cohort

The published simulations (Figures 2-4 and supplementary Tables S2 and S3)
all use SMA patients dosed 15 mg/kg every 6 h (capped at 1 g), varying one
covariate at a time and holding the others at the population median
(myoglobin 25 ng/mL, total bilirubin 6 umol/L -- the same medians Table 2
uses to normalise the two power models). Twelve typical-value scenarios
reproduce the three published simulation panels.

```{r scenarios}
mw <- c(pcm = 151.16, gluc = 327.29, sulf = 231.23, cysmer = 270.30)

scenarios <- dplyr::bind_rows(
  tibble::tibble(panel = "Body weight (Table S2)",
                 WT = c(70, 50, 35, 27), MYO = 25, TBILI = 6),
  tibble::tibble(panel = "Myoglobin (Table S3)",
                 WT = 70, MYO = c(15, 25, 40, 70), TBILI = 6),
  tibble::tibble(panel = "Bilirubin (Table S3)",
                 WT = 70, MYO = 25, TBILI = c(3, 6, 12, 24))
) |>
  dplyr::mutate(
    DIS_HEALTHY = 0,
    id        = dplyr::row_number(),
    dose_mg   = pmin(1000, 15 * WT),
    dose_umol = dose_mg / mw[["pcm"]] * 1000,
    scenario  = dplyr::case_when(
      panel == "Body weight (Table S2)" ~ paste0("WT ", WT, " kg"),
      panel == "Myoglobin (Table S3)"   ~ paste0("MYO ", MYO, " ng/mL"),
      TRUE                              ~ paste0("BILI ", TBILI, " umol/L")
    )
  )

knitr::kable(
  scenarios |> dplyr::select(panel, scenario, WT, MYO, TBILI, dose_mg, dose_umol),
  digits = 1,
  caption = "Twelve typical-value SMA scenarios reproducing the published simulations."
)
```

```{r events}
obs_times <- seq(0, 24, by = 0.1)

doses <- scenarios |>
  tidyr::expand_grid(time = c(0, 6, 12, 18)) |>
  dplyr::mutate(cmt = "depot", amt = dose_umol, evid = 1L) |>
  dplyr::select(id, time, cmt, amt, evid, WT, MYO, TBILI, DIS_HEALTHY)

obs <- scenarios |>
  dplyr::select(id, WT, MYO, TBILI, DIS_HEALTHY) |>
  tidyr::expand_grid(time = obs_times) |>
  dplyr::mutate(cmt = "Cc", amt = NA_real_, evid = 0L) |>
  dplyr::select(id, time, cmt, amt, evid, WT, MYO, TBILI, DIS_HEALTHY)

events <- dplyr::bind_rows(doses, obs) |>
  dplyr::arrange(id, time, dplyr::desc(evid))
```

# Simulation

Typical-value trajectories use `rxode2::zeroRe()` so that the comparison
against the published typical simulations is not blurred by the (very
large) between-subject variability on `k_a` and `CL_p/F`.

```{r simulate}
mod <- readModelDb("Zhao_2025_paracetamol")
mod_typ <- rxode2::zeroRe(mod)

sim <- rxode2::rxSolve(mod_typ, events = events, omega = NA, sigma = NA) |>
  as.data.frame() |>
  dplyr::left_join(dplyr::select(scenarios, id, panel, scenario), by = "id")

stopifnot(dplyr::n_distinct(sim$id) == nrow(scenarios))

# Convert each analyte from umol/L to its own mass concentration (ug/L),
# which is the scale the published figures and supplementary tables use.
sim <- sim |>
  dplyr::mutate(
    pcm_ugL    = Cc        * mw[["pcm"]],
    gluc_ugL   = Cc_gluc   * mw[["gluc"]],
    sulf_ugL   = Cc_sulf   * mw[["sulf"]],
    cysmer_ugL = Cc_cysmer * mw[["cysmer"]]
  )
```

# Replicate Figure 2 -- body-weight effect in SMA patients

```{r figure-2, fig.height = 6}
long_ugL <- function(d) {
  d |>
    tidyr::pivot_longer(
      c(pcm_ugL, gluc_ugL, sulf_ugL, cysmer_ugL),
      names_to = "analyte", values_to = "conc"
    ) |>
    dplyr::mutate(analyte = factor(
      analyte,
      levels = c("pcm_ugL", "gluc_ugL", "sulf_ugL", "cysmer_ugL"),
      labels = c("(A) Paracetamol", "(B) Glucuronide",
                 "(C) Sulphate", "(D) Oxidative metabolites")
    ))
}

sim |>
  dplyr::filter(panel == "Body weight (Table S2)") |>
  long_ugL() |>
  ggplot(aes(time, conc, colour = scenario)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (ug/L)", colour = "Scenario",
       title = "Body-weight effect in SMA patients (15 mg/kg q6h, capped at 1 g)",
       caption = "Replicates Figure 2 of Zhao 2025.") +
  theme_minimal()
```

# Replicate Figure 3 -- myoglobin effect in 70 kg SMA patients

```{r figure-3, fig.height = 6}
sim |>
  dplyr::filter(panel == "Myoglobin (Table S3)") |>
  long_ugL() |>
  ggplot(aes(time, conc, colour = scenario)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (ug/L)", colour = "Myoglobin",
       title = "Myoglobin effect in 70 kg SMA patients (1000 mg q6h)",
       caption = "Replicates Figure 3 of Zhao 2025.") +
  theme_minimal()
```

# Replicate Figure 4 -- bilirubin effect in 70 kg SMA patients

```{r figure-4, fig.height = 6}
sim |>
  dplyr::filter(panel == "Bilirubin (Table S3)") |>
  long_ugL() |>
  ggplot(aes(time, conc, colour = scenario)) +
  geom_line() +
  facet_wrap(~analyte, scales = "free_y") +
  labs(x = "Time (h)", y = "Concentration (ug/L)", colour = "Bilirubin",
       title = "Bilirubin effect in 70 kg SMA patients (1000 mg q6h)",
       caption = "Replicates Figure 4 of Zhao 2025.") +
  theme_minimal()
```

The absolute levels are the first check on the molar reading: at
myoglobin 25 ng/mL and bilirubin 6 umol/L the simulated 70 kg SMA patient
sits at roughly 3000-11000 ug/L paracetamol, 11000-18000 ug/L glucuronide,
2500-6000 ug/L sulphate and 400-700 ug/L oxidative metabolites, matching the
axis ranges of Figures 2-4 of the source.

# PKNCA validation

```{r pknca}
nca_one <- function(conc_col) {
  cdat <- sim |>
    dplyr::select(id, time, scenario, conc = dplyr::all_of(conc_col)) |>
    dplyr::filter(!is.na(conc))
  ddat <- scenarios |>
    dplyr::select(id, scenario, amt = dose_umol) |>
    tidyr::expand_grid(time = c(0, 6, 12, 18))
  obj <- PKNCA::PKNCAdata(
    PKNCA::PKNCAconc(cdat, conc ~ time | scenario + id),
    PKNCA::PKNCAdose(ddat, amt ~ time | scenario + id),
    intervals = data.frame(
      start = 0, end = 24,
      cmax = TRUE, tmax = TRUE, auclast = TRUE, half.life = TRUE
    )
  )
  as.data.frame(PKNCA::pk.nca(obj)$result) |>
    dplyr::mutate(analyte = conc_col)
}

nca <- dplyr::bind_rows(lapply(
  c("pcm_ugL", "gluc_ugL", "sulf_ugL", "cysmer_ugL"), nca_one
))

nca_wide <- nca |>
  dplyr::select(scenario, analyte, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

knitr::kable(
  nca_wide |>
    dplyr::rename("Scenario" = scenario, "Analyte" = analyte,
                  "Cmax (ug/L)" = cmax, "Tmax (h)" = tmax,
                  "AUC0-24 (ug*h/L)" = auclast, "t1/2 (h)" = half.life),
  digits = c(0, 0, 0, 2, 0, 2),
  caption = paste(
    "Simulated typical-value NCA over 0-24 h for all four analytes.",
    "Half-life is the NCA terminal half-life from the simulated profile,",
    "which for a metabolite is formation-rate limited and therefore tracks",
    "the parent -- it is not comparable to the V/CL half-lives the paper",
    "tabulates (see below)."
  )
)
```

## Comparison against the published AUC(0-24 h)

Supplementary Tables S2 and S3 report the median AUC(0-24 h) of 1000
simulated subjects for each scenario. The comparison below is against the
typical-value (zero-variability) trajectory, so a small systematic offset is
expected: with the very skewed between-subject distribution on `CL_p/F`
(omega = 0.862), the median of the individual total clearances exceeds the
typical value, so the published medians sit slightly below the typical-value
AUC.

```{r auc-reference}
ref_auc <- tibble::tribble(
  ~scenario,           ~pcm_ugL, ~gluc_ugL, ~sulf_ugL, ~cysmer_ugL,
  "WT 70 kg",            147330,    304768,     89652,       11152,
  "WT 50 kg",            143339,    298104,     88688,       11294,
  "WT 35 kg",            133833,    289267,     84298,       11141,
  "WT 27 kg",            123959,    263171,     76024,        9898,
  "MYO 15 ng/mL",        121164,    252827,     74728,        9197,
  "MYO 25 ng/mL",        146178,    301519,     89376,       11242,
  "MYO 40 ng/mL",        166944,    346387,    103558,       12789,
  "MYO 70 ng/mL",        178697,    366287,    109699,       13186,
  "BILI 3 umol/L",       152463,    315236,     81747,       10538,
  "BILI 6 umol/L",       146178,    301519,     89376,       11242,
  "BILI 12 umol/L",      141846,    298449,    100237,       12155,
  "BILI 24 umol/L",      131999,    271035,    104188,       11841
) |>
  tidyr::pivot_longer(-scenario, names_to = "analyte", values_to = "auclast")

auc_cmp <- nca_wide |>
  dplyr::select(scenario, analyte, simulated = auclast) |>
  dplyr::inner_join(ref_auc, by = c("scenario", "analyte")) |>
  dplyr::mutate(
    analyte = dplyr::recode(analyte,
      pcm_ugL = "Paracetamol", gluc_ugL = "Glucuronide",
      sulf_ugL = "Sulphate", cysmer_ugL = "Oxidative"),
    "% diff" = round(100 * (simulated - auclast) / auclast, 1)
  ) |>
  dplyr::rename("Scenario" = scenario, "Analyte" = analyte,
                "Published (ug*h/L)" = auclast,
                "Simulated (ug*h/L)" = simulated)

knitr::kable(auc_cmp, digits = 0,
             caption = "AUC(0-24 h): simulated typical value vs published median of 1000 subjects (supplementary Tables S2 and S3).")
```

```{r auc-summary}
summary(abs(auc_cmp[["% diff"]]))
```

## Comparison against the published half-lives

The half-lives in supplementary Tables S2 and S3 cannot be terminal
half-lives: the glucuronide (2.20 h) and sulphate (0.61 h) values are
*shorter* than the parent's (2.85 h), which is impossible for a metabolite
whose formation is rate-limiting. They are per-subject `ln(2) * V / CL`
values summarised over the simulated cohort. The check below computes the
same quantity analytically from the model parameters.

```{r halflife-check}
halflife_analytic <- function(WT, MYO, TBILI, DIS_HEALTHY = 0) {
  vc   <- 63.5 * 1.58 * (WT / 70) * (1 + (1 / 1.58 - 1) * DIS_HEALTHY)
  vmet <- 0.18 * vc
  allo <- (WT / 70)^0.75
  cl_left   <- 8.16 * allo * (MYO / 25)^(-1.10)
  cl_gluc   <- 6.37 * allo
  cl_sulf   <- 8.43 * allo * (TBILI / 6)^(0.18)
  cl_cysmer <- 0.20 * allo
  c(
    pcm    = log(2) * vc / (cl_left + cl_gluc + cl_sulf + cl_cysmer),
    gluc   = log(2) * vmet / (5.69 * allo),
    sulf   = log(2) * vmet / (20.4 * allo),
    cysmer = log(2) * vmet / (3.72 * allo * (TBILI / 6)^(-0.177))
  )
}

hl_sim <- scenarios |>
  dplyr::rowwise() |>
  dplyr::mutate(hl = list(halflife_analytic(WT, MYO, TBILI, DIS_HEALTHY))) |>
  dplyr::ungroup() |>
  tidyr::unnest_wider(hl) |>
  dplyr::select(scenario, pcm, gluc, sulf, cysmer) |>
  tidyr::pivot_longer(-scenario, names_to = "analyte", values_to = "simulated")

ref_hl <- tibble::tribble(
  ~scenario,        ~pcm, ~gluc, ~sulf, ~cysmer,
  "WT 70 kg",       2.85,  2.20,  0.61,    3.37,
  "WT 50 kg",       2.65,  2.04,  0.57,    3.12,
  "WT 35 kg",       2.40,  1.85,  0.51,    2.92,
  "WT 27 kg",       2.24,  1.74,  0.49,    2.61,
  "MYO 15 ng/mL",   2.20,  2.20,  0.61,    3.38,
  "MYO 25 ng/mL",   2.79,  2.17,  0.60,    3.31,
  "MYO 40 ng/mL",   3.14,  2.17,  0.60,    3.37,
  "MYO 70 ng/mL",   3.69,  2.22,  0.62,    3.32,
  "BILI 3 umol/L",  2.99,  2.29,  0.64,    3.18,
  "BILI 6 umol/L",  2.84,  2.20,  0.61,    3.37,
  "BILI 12 umol/L", 2.79,  2.21,  0.62,    3.79,
  "BILI 24 umol/L", 2.46,  2.13,  0.59,    4.19
) |>
  tidyr::pivot_longer(-scenario, names_to = "analyte", values_to = "published")

hl_cmp <- hl_sim |>
  dplyr::inner_join(ref_hl, by = c("scenario", "analyte")) |>
  dplyr::mutate(
    analyte = dplyr::recode(analyte,
      pcm = "Paracetamol", gluc = "Glucuronide",
      sulf = "Sulphate", cysmer = "Oxidative"),
    "% diff" = round(100 * (simulated - published) / published, 1)
  ) |>
  dplyr::rename("Scenario" = scenario, "Analyte" = analyte,
                "Published t1/2 (h)" = published,
                "Analytic ln(2)V/CL (h)" = simulated)

knitr::kable(hl_cmp, digits = 2,
             caption = "ln(2) * V / CL half-lives vs supplementary Tables S2 and S3.")
```

```{r halflife-assert}
# The three metabolite half-lives are pure parameter arithmetic with no
# between-subject variability on their elimination clearances, so they must
# reproduce the published values essentially exactly.
met_err <- hl_cmp |>
  dplyr::filter(Analyte != "Paracetamol") |>
  dplyr::pull("% diff") |>
  abs()
stopifnot(
  stats::median(met_err) < 2.5,
  max(met_err) < 8
)
c(median = stats::median(met_err), max = max(met_err))
```

## Allometric scaling is reproduced exactly

With a clearance exponent of 0.75 and a volume exponent of 1, `V / CL` --
and hence the half-life -- scales as `WT^0.25`. Supplementary Table S2's
body-weight panel reproduces that to two decimal places.

```{r allometry-check}
wt_panel <- scenarios |> dplyr::filter(panel == "Body weight (Table S2)")
tibble::tibble(
  "WT (kg)" = wt_panel$WT,
  "Published t1/2 ratio to 70 kg" =
    round(c(2.85, 2.65, 2.40, 2.24) / 2.85, 4),
  "(WT/70)^0.25" = round((wt_panel$WT / 70)^0.25, 4)
) |>
  knitr::kable(caption = "Half-life ratio across body weights vs the fixed allometric exponents.")
```

## Structural identity: metabolite exposure ratio equals CL_formation / CL_elimination

At steady state the amount of metabolite formed per dosing interval equals
the amount eliminated, so the metabolite:parent AUC ratio must equal
`CL_formation / CL_elimination` on the molar scale, independent of every
volume. This is asserted per scenario against the simulated molar AUCs.

```{r trapz-def}
trapz <- function(x, y) {
  sum(diff(x) * (utils::head(y, -1) + utils::tail(y, -1)) / 2)
}
```

```{r ratio-identity-run}
molar_auc <- sim |>
  dplyr::group_by(scenario, TBILI) |>
  dplyr::summarise(
    auc_pcm    = trapz(time, Cc),
    auc_gluc   = trapz(time, Cc_gluc),
    auc_sulf   = trapz(time, Cc_sulf),
    auc_cysmer = trapz(time, Cc_cysmer),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    ratio_gluc_obs   = auc_gluc / auc_pcm,
    ratio_gluc_exp   = 6.37 / 5.69,
    ratio_sulf_obs   = auc_sulf / auc_pcm,
    ratio_sulf_exp   = 8.43 * (TBILI / 6)^0.18 / 20.4,
    ratio_cysmer_obs = auc_cysmer / auc_pcm,
    ratio_cysmer_exp = 0.20 / (3.72 * (TBILI / 6)^(-0.177))
  )

# Over a finite 0-24 h window the identity holds only up to the metabolite
# still in the body at 24 h, so the observed ratio is slightly below the
# steady-state value; the deficit is largest for the slowest metabolite.
knitr::kable(
  molar_auc |>
    dplyr::transmute(
      Scenario = scenario,
      "Gluc obs/exp"   = ratio_gluc_obs   / ratio_gluc_exp,
      "Sulf obs/exp"   = ratio_sulf_obs   / ratio_sulf_exp,
      "Oxid obs/exp"   = ratio_cysmer_obs / ratio_cysmer_exp
    ),
  digits = 3,
  caption = "Simulated metabolite:parent molar AUC ratio divided by CL_formation / CL_elimination."
)

stopifnot(
  all(molar_auc$ratio_sulf_obs / molar_auc$ratio_sulf_exp > 0.9),
  all(molar_auc$ratio_sulf_obs / molar_auc$ratio_sulf_exp < 1.0),
  all(molar_auc$ratio_gluc_obs / molar_auc$ratio_gluc_exp < 1.0)
)
```

## Amount scale: molar versus paracetamol-mass, scored on the published window

The same simulated 0-24 h molar AUC ratios settle the amount-scale question
posed earlier. Reading the metabolite compartments as moles and converting
each output with that species' own molecular weight reproduces
supplementary Table S2's ratios; reading them as paracetamol-equivalent mass
does not.

```{r units-falsifier}
ref70 <- c(pcm = 147330, gluc = 304768, sulf = 89652, cysmer = 11152)
sim70 <- molar_auc |> dplyr::filter(scenario == "WT 70 kg")
mass_equiv <- c(
  gluc   = sim70$auc_gluc   / sim70$auc_pcm,
  sulf   = sim70$auc_sulf   / sim70$auc_pcm,
  cysmer = sim70$auc_cysmer / sim70$auc_pcm
)

units_cmp <- tibble::tibble(
  Metabolite = c("Glucuronide", "Sulphate", "Oxidative (cys + merc)"),
  "Published ratio (Table S2, 70 kg)" = unname(ref70[-1] / ref70[["pcm"]]),
  "Mass-equivalent model" = unname(mass_equiv),
  "Molar model x own MW" = unname(mass_equiv * mw[-1] / mw[["pcm"]])
) |>
  dplyr::mutate(
    "Mass % diff" = round(100 * (`Mass-equivalent model` -
      `Published ratio (Table S2, 70 kg)`) /
      `Published ratio (Table S2, 70 kg)`, 1),
    "Molar % diff" = round(100 * (`Molar model x own MW` -
      `Published ratio (Table S2, 70 kg)`) /
      `Published ratio (Table S2, 70 kg)`, 1)
  )

knitr::kable(
  units_cmp, digits = 3,
  caption = paste(
    "Metabolite:paracetamol AUC(0-24 h) ratio under the two candidate amount",
    "scales. Molecular weights are those tabulated in van Rongen 2016 Methods",
    "for the same analytes (paracetamol 151.16, glucuronide 327.29, sulphate",
    "231.23, cysteine conjugate 270.30 g/mol)."
  )
)

stopifnot(
  max(abs(units_cmp[["Molar % diff"]])) < 5,
  min(abs(units_cmp[["Mass % diff"]])) > 25
)
```

# Stochastic cohort

A 200-subject SMA cohort at the reference scenario (70 kg, myoglobin
25 ng/mL, bilirubin 6 umol/L) shows how far the published median of 1000
simulated subjects should sit below the typical-value AUC.

```{r stochastic}
set.seed(20250717)
n_sub <- 200L

ev_stoch <- dplyr::bind_rows(
  tidyr::expand_grid(id = seq_len(n_sub), time = c(0, 6, 12, 18)) |>
    dplyr::mutate(cmt = "depot", amt = 1000 / mw[["pcm"]] * 1000, evid = 1L),
  tidyr::expand_grid(id = seq_len(n_sub), time = seq(0, 24, by = 0.25)) |>
    dplyr::mutate(cmt = "Cc", amt = NA_real_, evid = 0L)
) |>
  dplyr::mutate(WT = 70, MYO = 25, TBILI = 6, DIS_HEALTHY = 0) |>
  dplyr::arrange(id, time, dplyr::desc(evid))

sim_stoch <- rxode2::rxSolve(mod, events = ev_stoch, sigma = NA) |>
  as.data.frame()

stopifnot(dplyr::n_distinct(sim_stoch$id) == n_sub)

auc_stoch <- sim_stoch |>
  dplyr::group_by(id) |>
  dplyr::summarise(auc_pcm = trapz(time, Cc) * mw[["pcm"]], .groups = "drop")

tibble::tibble(
  Quantity = c("Published median of 1000 subjects (Table S2, 70 kg)",
               "Typical value (zeroRe)",
               paste0("Median of ", n_sub, " simulated subjects")),
  "AUC0-24 paracetamol (ug*h/L)" = c(
    147330,
    auc_cmp[["Simulated (ug*h/L)"]][auc_cmp$Scenario == "WT 70 kg" &
                                      auc_cmp$Analyte == "Paracetamol"],
    stats::median(auc_stoch$auc_pcm)
  )
) |>
  knitr::kable(digits = 0,
               caption = "Typical value vs cohort median for the reference SMA scenario.")
```

```{r vpc, fig.height = 4}
sim_stoch |>
  dplyr::group_by(time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc * mw[["pcm"]], 0.05),
    Q50 = stats::quantile(Cc * mw[["pcm"]], 0.50),
    Q95 = stats::quantile(Cc * mw[["pcm"]], 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line() +
  labs(x = "Time (h)", y = "Paracetamol (ug/L)",
       title = "Simulated 70 kg SMA cohort, 1000 mg q6h",
       caption = paste0(n_sub, " subjects; between-subject variability only, residual error not added.")) +
  theme_minimal()
```

# Assumptions and deviations

- **Molar amount scale (determined, not stated).** Zhao 2025 does not state
  whether its compartments carry moles or paracetamol-equivalent mass. The
  model file uses `units$dosing = "umol"` / `units$concentration = "umol/L"`,
  determined from the paper's own supplementary Table S2 as shown in the
  "Concentration scale" section above (1-3% agreement on all three
  metabolite:parent AUC ratios, versus 43-111% error for a mass-equivalent
  reading). This matches the sibling paracetamol parent-metabolite model
  `vanRongen_2016_acetaminophen`, which states the molar convention
  explicitly. Users dosing in mg must convert with
  `dose_umol = dose_mg / 0.15116`.
- **Sign of two covariate exponents.** The myoglobin exponent on `CL_p/F`
  and the bilirubin exponent on `CL_ox/F` are printed without their minus
  signs in the typeset Table 2 (their bootstrap CIs print in descending
  order, which is only possible for negative intervals). Both are encoded as
  negative, consistent with the "negative correlation" statements in Results
  section 3.3 and with the monotone half-life trends in supplementary
  Table S3.
- **BSV is a log-scale SD, not a CV.** Table 2's "BSV (%)" block is the
  log-scale standard deviation times 100, so the nlmixr2 eta variances are
  `(BSV / 100)^2`. The determination is exact: the Discussion quotes the
  `CL_p/F` BSV as 105%, and `sqrt(exp(0.862^2) - 1) = 1.050` recovers that
  from the Table 2 value of 86.2%.
- **Residual error read as a standard deviation.** Table 2's "Proportional"
  residual block (0.28 / 0.18 / 0.22 / 0.17) is entered as proportional
  residual SDs, on the same standard-deviation scale as the BSV block in the
  same table. The supplementary paracetamol goodness-of-fit panel (observed
  vs individual-predicted, with the paired IWRES panel) shows relative
  residual scatter consistent with a ~0.28 SD rather than the ~0.53 SD that
  reading 0.28 as a NONMEM variance would imply.
- **No baseline compartment.** Pre-dose samples were below the LLOQ in most
  subjects and were fixed to their observed values during base-model
  building (Results section 3.2), so the published model carries no baseline
  term and neither does this implementation; all states start at zero.
- **Metabolite volumes are not identifiable.** They are fixed at 18% of the
  paracetamol volume from the literature (Methods section 2.2). The authors
  flag this as the probable cause of the implausibly large leftover-clearance
  fraction (47% in SMA, 23-27% in healthy controls, against a literature
  value near 4% for unchanged paracetamol; Discussion). The library model
  reproduces the published parameterisation without adjustment.
- **Simulated AUC compared against a cohort median.** Supplementary Tables
  S2 and S3 report the median AUC of 1000 simulated subjects, while the
  typical-value comparison above uses zero variability. Because `CL_p/F`
  carries omega = 0.862, the median individual total clearance exceeds the
  typical value and the published medians therefore run a few per cent below
  the typical-value AUC; the 200-subject cohort check quantifies the offset.
  The residual disagreement is not tuned away.
- **NCA half-life is not the paper's half-life.** The half-lives tabulated
  in supplementary Tables S2 and S3 are `ln(2) * V / CL` values, not terminal
  half-lives -- the glucuronide and sulphate entries are shorter than the
  parent's, which a formation-rate-limited metabolite cannot be. The
  comparison above therefore uses the analytic quantity; the PKNCA
  `half.life` column is reported separately and is not expected to match.
- **Disease indicator re-expressed as `DIS_HEALTHY`.** The paper's source
  column is the reverse-coded `disease` indicator (1 = SMA) with the healthy
  controls as the reference. The model file uses the existing canonical
  `DIS_HEALTHY` (reference category 0 = patient), shifting the structural
  typical to `lvc = log(63.5 * 1.58)` so that `1 / 1.58` at
  `DIS_HEALTHY = 1` restores the paper's printed healthy-control
  `V_pcm/F` of 63.5 L/70 kg. This follows `Cleary_2023_risdiplam`, which
  encodes the same SMA-vs-healthy contrast on the same canonical.
- **`MYO` is a new canonical covariate column,** registered in
  `inst/references/covariate-columns.md` alongside this extraction; it
  joins the plasma clinical-chemistry biomarkers (`ALT`, `AST`, `TBILI`,
  `ALP`, `LDH`, `CPK`).
- **Screened-but-unretained covariates.** Sex, age, BMI, fat-free mass, ALT,
  AST, ALP, LDH, creatine kinase, creatinine, potassium and sodium are
  recorded in `covariatesDataExcluded` because the paper screened them
  without retaining any of them; urea and glomerular filtration rate were
  also screened but have no unambiguous canonical column, so they are noted
  in `population$notes` instead.
- **No errata identified.** A search of the BJCP article landing page,
  PubMed and Europe PMC for corrections to `doi:10.1002/bcp.70028` returned
  none as of the extraction date (2026-08-17).
- **No NONMEM control stream on disk.** The Wiley supplement comprises the
  supplementary methods (DOCX), supplementary tables S1-S3 (DOCX) and
  figures S1-S9 (PDF/JPEG); no `.mod` / `.ctl` / `.lst` is deposited. Every
  parameter value comes from Table 2 of the main publication.
