---
title: "Nivolumab, subcutaneous and intravenous (Zhao 2025)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nivolumab, subcutaneous and intravenous (Zhao 2025)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhao_2025_nivolumab"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhao_2025_nivolumab"))$description`
* Article: <https://doi.org/10.1002/psp4.70120>
* Supplement (open access, EuropePMC PMC12706396): supplemental tables
  (Figure S1, Tables S1-S3), the full-model NONMEM control stream
  (labelled "File S1" inside the document), and the `$PRIOR` control
  stream (labelled "File S2").

Nivolumab is a fully human IgG4 monoclonal antibody against PD-1. A
subcutaneous formulation co-formulated with recombinant human
hyaluronidase PH20 (rHuPH20) was developed to avoid intravenous infusion
visits, and was tested against intravenous nivolumab in the phase III
PK non-inferiority trial CheckMate 67T (NCT04810078).

This model is the **pre-specified population PK model** for that trial.
It was built on a pooled dataset of 19 historical intravenous nivolumab
monotherapy studies plus the subcutaneous phase I/II study CheckMate 8KX,
so that the CheckMate 67T co-primary PK endpoints could be derived from a
model rather than from noncompartmental analysis (NCA). The paper's own
subject is a methodology comparison -- it uses clinical trial simulation
to show that re-estimating this model with the NONMEM `$PRIOR` subroutine
gives the same exposures as a pooled re-analysis, and that both beat
conventional-schedule NCA. The model packaged here is the pre-specified
model itself, whose estimates are reported in supplemental Table S2.

## Structure

Systemic disposition is shared by both routes (the paper's premise: "The
systemic PK is expected to be identical for i.v. and s.c. administered
nivolumab"):

* Two-compartment linear disposition (`central`, `peripheral1`),
  parameterised as CL, Vc, Q, Vp (NONMEM `ADVAN4 TRANS4`, `S2 = V2`).
* **Time-varying clearance**, a sigmoid-Emax function of time since the
  first dose:
  `CL(t) = CL_base * exp(Emax * t^HILL / (T50^HILL + t^HILL))`.
  With `Emax = -0.303`, clearance falls to `exp(-0.303) = 0.739` of
  baseline (a 26% reduction) as `t` exceeds `T50`.
* Covariate effects: power functions of baseline body weight and eGFR on
  CL, power of body weight on Vc, and exponential effects of female sex,
  ECOG >= 1, gastric cancer and classical Hodgkin lymphoma.

Subcutaneous absorption adds:

* A `depot` with first-order absorption rate `ka`.
* Bioavailability on the **logit** scale, exactly as the paper's printed
  equation for `F_i` and the control streams construct it: covariate
  factors multiply the typical fraction on the natural scale, then the
  logit of that product carries the random effect,
  `F_i = expit(logit(TVF1) + eta_F)`.

The dosing route is expressed **structurally** rather than with a route
covariate: bioavailability applies only to the depot, so subcutaneous
doses go to `depot` and intravenous doses go to `central` (where `F = 1`).
This reproduces the source `ADVAN4` setup, in which `F1` governs
compartment 1 only.

```{r model}
# rxode() gives the rxUi, whose $meta carries the file-level metadata;
# readModelDb() alone returns the bare model function.
mod <- rxode2::rxode(readModelDb("Zhao_2025_nivolumab"))
mod
```

# Population

The pooled analysis dataset (Table 1) contained **3554 subjects**: 66
who received subcutaneous nivolumab with rHuPH20 in CheckMate 8KX
(CA2098KX), and 3488 who received intravenous nivolumab across the
studies listed in Table S1. Subjects in CheckMate 8KX who received
subcutaneous nivolumab *without* rHuPH20 were excluded.

```{r population}
pop <- mod$meta$population
data.frame(
  Characteristic = c(
    "Subjects (total)", "  subcutaneous (CheckMate 8KX)", "  intravenous",
    "Studies pooled", "Age", "Baseline body weight", "Baseline eGFR",
    "Baseline albumin", "Female", "ECOG performance status", "Disease"
  ),
  Value = c(
    format(pop$n_subjects, big.mark = ","),
    format(pop$n_subjects_sc), format(pop$n_subjects_iv, big.mark = ","),
    format(pop$n_studies), pop$age_range, pop$weight_range,
    pop$renal_function, pop$albumin,
    paste0(pop$sex_female_pct, "%"), pop$ecog_distribution, pop$disease_state
  )
) |>
  knitr::kable(caption = "Baseline characteristics of the pooled dataset (Zhao 2025 Table 1).")
```

Tumour types spanned NSCLC (25.9%), melanoma (24.7%), renal cell
carcinoma (20.1%), bladder cancer (8.7%), classical Hodgkin lymphoma
(7.2%), SCCHN (4.8%), SCLC (2.7%), colorectal (1.7%), gastric (1.6%) and
other (2.4%). CheckMate 67T itself enrolled only previously-treated
advanced or metastatic **clear-cell renal cell carcinoma**, so the
simulations below set both retained tumour-type indicators to 0 (RCC
falls in the model's reference tumour group).

# Source trace

Every `ini()` value comes from supplemental Table S2 ("Parameter
estimates of nivolumab prespecified model"). The two control streams give
an independent confirmation of each one, because the `$PRIOR` stream's
`$THETAP` / `$OMEGAP` records restate the same pre-specified values as
prior means. Three readings required arbitration and are documented in
the Errata section below.

```{r source-trace, echo = FALSE}
data.frame(
  Parameter = c(
    "lcl", "lvc", "lq", "lvp", "lka", "logitfdepot",
    "e_wt_cl", "e_crcl_cl", "e_sex_cl", "e_ecog_ge1_cl",
    "e_tumtp_gastric_cl", "e_tumtp_hodgkin_classical_cl",
    "e_wt_vc", "e_sex_vc", "e_sex_fdepot", "e_ecog_ge1_fdepot",
    "cl_time_max", "cl_t50", "cl_time_hill",
    "etalcl, etalvc (block)", "etalvp", "etacl_time_max",
    "etalka, etalogitfdepot (block)", "propSd",
    "Disposition ODEs", "Time-varying CL equation",
    "Logit bioavailability", "Reference WT / eGFR"
  ),
  `Source location` = c(
    "Table S2 CL0_REF = 0.0109 L/h (x 24 -> L/day)",
    "Table S2 Vc_REF = 4.25 L",
    "Table S2 Q_REF = 0.0327 L/h (x 24 -> L/day)",
    "Table S2 Vp_REF = 2.63 L",
    "Table S2 Ka_REF = 0.307 /day",
    "Table S2 F_REF = 0.752",
    "Table S2 CL_WTB = 0.622", "Table S2 CL_eGFR = 0.139",
    "Table S2 CL_SEX = -0.158", "Table S2 CL_PS = 0.174",
    "Table S2 CL_GC = 0.180", "Table S2 CL_CHL = -0.330",
    "Table S2 Vc_WTB = 0.630", "Table S2 Vc_SEX = -0.134",
    "Table S2 F_SEX = 0.859", "Table S2 F_PS = 1.07",
    "Table S2 Emax_REF = -0.303", "Table S2 T50 = 1.40e3 h (/ 24 -> day)",
    "Table S2 HILL = 2.82",
    "Table S2 omega^2_CL 0.114, cov 0.0377, omega^2_Vc 0.126",
    "Table S2 omega^2_Vp = 0.235", "Table S2 omega^2_Emax = 0.0519",
    "Table S2 omega^2_Ka 0.0955, cov 0.213, omega^2_F 0.862",
    "Table S2 proportional = 0.204",
    "Methods 2.2; control streams $SUBROUTINE ADVAN4 TRANS4, S2 = V2",
    "Control streams $PK: CL_TIME = EXP(EMAX*TIME**HILL/(T50**HILL+TIME**HILL))",
    "Main text Eq. for F_i (recovered via pdftotext -layout); control streams $PK LOGITF1 block",
    "Control streams $PK: BBWT_R = 80 kg, BGFR_R = 90 mL/min"
  ),
  check.names = FALSE
) |>
  knitr::kable(caption = "Source trace for every parameter and equation.")
```

# Virtual cohort

The paper resampled real RCC patients from the pooled dataset. RCC-specific
baseline statistics are not published, so the cohort below is drawn from
the Table 1 **overall** marginal distributions, with both tumour-type
indicators set to 0 for RCC. 200 subjects per arm, and the same cohort
and seed are reused across arms so the s.c./i.v. comparison uses common
random numbers.

```{r cohort}
n_sub <- 200L

make_cohort <- function(seed = 2024L) {
  set.seed(seed)
  data.frame(
    id       = seq_len(n_sub),
    WT       = pmin(pmax(rnorm(n_sub, 78.3, 19.2), 40), 160),
    CRCL     = pmin(pmax(rnorm(n_sub, 81.9, 23.0), 20), 190),
    SEXF     = rbinom(n_sub, 1, 0.328),
    ECOG_GE1 = rbinom(n_sub, 1, 0.564),
    TUMTP_GASTRIC = 0,
    TUMTP_HODGKIN_CLASSICAL = 0
  )
}
cohort <- make_cohort()

summary(cohort[, c("WT", "CRCL")])
c(female_pct = mean(cohort$SEXF) * 100, ecog_ge1_pct = mean(cohort$ECOG_GE1) * 100)
```

# Simulation of the CheckMate 67T arms

Arm A is subcutaneous nivolumab 1200 mg every 4 weeks; arm B is
intravenous nivolumab 3 mg/kg every 2 weeks given as a 1-hour infusion.
Both run to day 364 so that the day-336 steady-state landmark used by the
paper ("concentrations achieved at cycle 12 following administration on
study day 336") is covered with a full dosing interval after it.

The observation grid carries a point `1e-4` days **before** each landmark
dose time, because a record placed exactly on a dose time is read
post-dose -- so a naive minimum over an interval that ends on a dose day
is not the trough.

```{r simulate}
day336 <- 336
regimens <- list(
  `s.c. 1200 mg q4w` = list(amt = NA, tau = 28, n = 13L, cmt = "depot",   dur = 0),
  `i.v. 3 mg/kg q2w` = list(amt = NA, tau = 14, n = 26L, cmt = "central", dur = 1 / 24)
)

# The grid resolves the end of each 1-hour infusion (and the early
# distribution phase after it); a grid coarse enough to miss that peak
# biases the AUC of the intravenous arm by several percent.
obs_grid <- sort(unique(c(
  seq(0, 364, by = 0.5),
  rep(seq(0, 364, by = 14), each = 3) + c(1 / 24, 0.1, 0.25),
  28 - 1e-4, day336 - 1e-4, 364 - 1e-4
)))

simulate_arm <- function(arm, mgkg = NULL, flat = NULL, label = arm) {
  r <- regimens[[arm]]
  co <- make_cohort()
  rxode2::rxSetSeed(4242)          # common random numbers across arms
  ev <- do.call(rbind, lapply(seq_len(nrow(co)), function(i) {
    amt_i <- if (is.null(flat)) mgkg * co$WT[i] else flat
    dose <- data.frame(
      id = co$id[i], time = seq(0, by = r$tau, length.out = r$n),
      amt = amt_i, rate = if (r$dur > 0) amt_i / r$dur else 0,
      evid = 1, cmt = r$cmt
    )
    obs <- data.frame(
      id = co$id[i], time = obs_grid, amt = 0, rate = 0,
      evid = 0, cmt = r$cmt
    )
    rbind(dose, obs)
  }))
  rxode2::rxSolve(mod, ev, co, returnType = "data.frame", addDosing = FALSE) |>
    dplyr::mutate(treatment = label)
}

sim <- dplyr::bind_rows(
  simulate_arm("s.c. 1200 mg q4w", flat = 1200),
  simulate_arm("i.v. 3 mg/kg q2w", mgkg = 3)
)
nrow(sim)
```

## Replicating Figure 2: geometric mean profiles by route

Replicates Figure 2 of Zhao 2025 -- the median and 90% interval of the
simulated geometric-mean concentration-time profile by route of
administration. The dashed reference line at 352 ug/mL is the value the
Figure 2 legend quotes for the steady-state peak concentration after
intravenous nivolumab 10 mg/kg q2w, "a dosing regimen that has been shown
to be safe and tolerable".

```{r figure2, fig.cap = "Replicates Figure 2 of Zhao 2025: geometric mean (90% interval across subjects) concentration-time profiles for subcutaneous and intravenous nivolumab."}
gm <- function(x) exp(mean(log(x[x > 0])))

prof <- sim |>
  dplyr::filter(time %in% seq(0, 364, by = 0.5)) |>
  dplyr::group_by(treatment, time) |>
  dplyr::summarise(
    gmean = gm(Cc),
    lo    = quantile(Cc, 0.05),
    hi    = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(prof, aes(time, gmean, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 352, linetype = "dashed") +
  annotate("text", x = 10, y = 372, hjust = 0, size = 3,
           label = "352 ug/mL: Cmaxss at i.v. 10 mg/kg q2w (Fig. 2 legend)") +
  scale_x_continuous(breaks = seq(0, 364, by = 56)) +
  labs(x = "Time since first dose (days)", y = "Nivolumab serum concentration (ug/mL)",
       colour = NULL, fill = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The qualitative features of Figure 2 are reproduced: the subcutaneous
profile has visibly wider spread than the intravenous profile (the paper
attributes this to the extra variability in `ka` and `F`, whose omegas are
the two largest in Table S2), both profiles accumulate over roughly the
first 100-150 days as time-varying clearance declines, and the
subcutaneous peak sits well below the 352 ug/mL safety reference.

## Exposure measures and the co-primary endpoints

The paper's exposure definitions: `Cavgd28` is the AUC over the first 28
days divided by 28 days; `Cmax1` is the peak after the first dose;
`Cmind28` is the trough at day 28; steady-state measures are taken at the
day-336 dose.

```{r exposures}
tau_of <- c(`s.c. 1200 mg q4w` = 28, `i.v. 3 mg/kg q2w` = 14)

trap <- function(tm, cc) sum(diff(tm) * (head(cc, -1) + tail(cc, -1)) / 2)

expo <- sim |>
  dplyr::group_by(treatment, id) |>
  dplyr::group_modify(function(d, key) {
    tau <- tau_of[[key$treatment]]
    d <- dplyr::arrange(d, time)
    w28 <- d$time <= 28
    wss <- d$time >= day336 & d$time <= day336 + tau
    data.frame(
      Cavgd28 = trap(d$time[w28], d$Cc[w28]) / 28,
      Cmax1   = max(d$Cc[d$time <= tau]),
      Cmind28 = d$Cc[which.min(abs(d$time - (28 - 1e-4)))],
      Cavgss  = trap(d$time[wss], d$Cc[wss]) / tau,
      Cminss  = d$Cc[which.min(abs(d$time - (day336 - 1e-4)))],
      Cmaxss  = max(d$Cc[wss])
    )
  }) |>
  dplyr::ungroup()

expo_long <- expo |>
  tidyr::pivot_longer(-c(treatment, id), names_to = "measure", values_to = "value")

expo_long |>
  dplyr::group_by(measure, treatment) |>
  dplyr::summarise(gmean = gm(value), cv_pct = sd(value) / mean(value) * 100, .groups = "drop") |>
  tidyr::pivot_wider(names_from = treatment, values_from = c(gmean, cv_pct)) |>
  dplyr::mutate(dplyr::across(-measure, \(x) round(x, 1))) |>
  dplyr::rename(
    "Exposure measure"    = measure,
    "s.c. GM (ug/mL)"     = `gmean_s.c. 1200 mg q4w`,
    "i.v. GM (ug/mL)"     = `gmean_i.v. 3 mg/kg q2w`,
    "s.c. CV%"            = `cv_pct_s.c. 1200 mg q4w`,
    "i.v. CV%"            = `cv_pct_i.v. 3 mg/kg q2w`
  ) |>
  knitr::kable(caption = "Simulated nivolumab exposures by route (200 subjects per arm).")
```

Replicates Figure 3 of Zhao 2025 (distribution of `Cavgd28`, `Cmax1` and
`Cmind28` by administration route):

```{r figure3, fig.cap = "Replicates Figure 3 of Zhao 2025: distribution of Cavgd28, Cmax1 and Cmind28 by administration route.", fig.height = 3.6}
expo_long |>
  dplyr::filter(measure %in% c("Cavgd28", "Cmax1", "Cmind28")) |>
  ggplot(aes(treatment, value, fill = treatment)) +
  geom_boxplot(outlier.size = 0.5, width = 0.55) +
  facet_wrap(~measure, scales = "free_y") +
  labs(x = NULL, y = "Concentration (ug/mL)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

The two **co-primary endpoints** of CheckMate 67T were `Cavgd28` and
`Cminss`. Dose selection required the lower bound of the 90% CI of the
subcutaneous/intravenous geometric mean ratio (GMR) to be **not less than
0.8**, i.e. subcutaneous exposures "equivalent to or higher than"
intravenous. Replicates Figure 5 of Zhao 2025:

```{r gmr}
gmr_tab <- expo_long |>
  tidyr::pivot_wider(names_from = treatment, values_from = value) |>
  dplyr::group_by(measure) |>
  dplyr::summarise({
    lr <- log(`s.c. 1200 mg q4w`) - log(`i.v. 3 mg/kg q2w`)
    se <- sd(lr) / sqrt(length(lr))
    data.frame(
      GMR    = exp(mean(lr)),
      lower  = exp(mean(lr) - qt(0.95, length(lr) - 1) * se),
      upper  = exp(mean(lr) + qt(0.95, length(lr) - 1) * se)
    )
  }, .groups = "drop") |>
  dplyr::mutate(
    `Co-primary endpoint` = ifelse(measure %in% c("Cavgd28", "Cminss"), "yes", ""),
    dplyr::across(c(GMR, lower, upper), \(x) round(x, 2))
  )

gmr_tab |>
  dplyr::rename("Exposure measure" = measure,
                "GMR (s.c./i.v.)" = GMR,
                "90% CI lower" = lower, "90% CI upper" = upper) |>
  knitr::kable(caption = "Simulated geometric mean ratios (subcutaneous / intravenous). The CheckMate 67T non-inferiority criterion is a 90% CI lower bound not less than 0.8.")
```

```{r gmr-gate}
coprimary <- gmr_tab[gmr_tab$measure %in% c("Cavgd28", "Cminss"), ]

stopifnot(
  # The paper's own pre-specified non-inferiority criterion: the lower bound
  # of the 90% CI of each co-primary GMR must be at least 0.8. Asserted on
  # the cohort-level CI, not on any individual subject.
  all(coprimary$lower >= 0.8),
  # And the design intent that s.c. exposures are "equivalent to or higher":
  # both co-primary GMRs exceed 1 with a wide margin.
  all(coprimary$GMR > 1.2),
  # Cavgd28 shows a larger s.c./i.v. ratio than Cminss, because the q4w
  # subcutaneous regimen front-loads a 1200 mg flat dose into the first
  # 28 days while the q2w intravenous arm delivers 2 x 3 mg/kg over the
  # same window; the ordering is a structural consequence of the regimens.
  gmr_tab$GMR[gmr_tab$measure == "Cavgd28"] >
    gmr_tab$GMR[gmr_tab$measure == "Cminss"]
)
```

Both co-primary endpoints clear the pre-specified 0.8 lower bound by a
wide margin, reproducing the conclusion the model was built to support.

# PKNCA validation

NCA is run with PKNCA over two intervals per arm: the first 28 days
(which yields `Cavgd28` as PKNCA's `cav`) and one steady-state dosing
interval beginning at the day-336 dose. The concentration frame is
filtered only on `!is.na(Cc)` and a `time = 0` record is guaranteed, so
PKNCA never back-extrapolates from a missing first measurement.

```{r pknca}
# Restricted to the two windows PKNCA is asked about, but at the FULL grid
# resolution used for the manual exposure table above, so the cross-check
# below compares integration methods rather than grid resolutions.
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::filter(time <= 28 | time >= day336) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a time-zero record (pre-dose Cc = 0 for both routes; the
# intravenous arm is an infusion, not a bolus, so its t = 0 value is 0 too).
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_nca <- dplyr::bind_rows(lapply(names(regimens), function(arm) {
  r <- regimens[[arm]]
  co <- make_cohort()
  do.call(rbind, lapply(seq_len(nrow(co)), function(i) {
    data.frame(
      treatment = arm, id = co$id[i],
      time = seq(0, by = r$tau, length.out = r$n),
      dose = if (arm == "s.c. 1200 mg q4w") 1200 else 3 * co$WT[i]
    )
  }))
}))

# Grouping puts treatment BEFORE id so per-arm results stay interpretable,
# and units are declared so the AUC / Cav output is not unit-blind.
o_conc <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                           concu = "ug/mL", timeu = "day")
o_dose <- PKNCA::PKNCAdose(as.data.frame(dose_nca), dose ~ time | treatment + id,
                           doseu = "mg")

intervals <- dplyr::bind_rows(
  data.frame(treatment = "s.c. 1200 mg q4w", start = 0,      end = 28),
  data.frame(treatment = "s.c. 1200 mg q4w", start = day336, end = day336 + 28),
  data.frame(treatment = "i.v. 3 mg/kg q2w", start = 0,      end = 28),
  data.frame(treatment = "i.v. 3 mg/kg q2w", start = day336, end = day336 + 14)
) |>
  dplyr::mutate(cmax = TRUE, tmax = TRUE, auclast = TRUE, cav = TRUE, tlast = TRUE)

res_nca <- PKNCA::pk.nca(PKNCA::PKNCAdata(o_conc, o_dose, intervals = intervals))

nca_summary <- as.data.frame(res_nca) |>
  dplyr::mutate(window = ifelse(start == 0, "Days 0-28", "Steady state (from day 336)")) |>
  dplyr::group_by(treatment, window, PPTESTCD) |>
  dplyr::summarise(median = median(PPORRES, na.rm = TRUE), .groups = "drop") |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = median)

nca_summary |>
  dplyr::mutate(dplyr::across(where(is.numeric), \(x) signif(x, 4))) |>
  knitr::kable(caption = "PKNCA results (per-subject medians) by arm and window. cav over days 0-28 is the paper's Cavgd28.")
```

```{r pknca-crosscheck}
# PKNCA's cav over days 0-28 must agree with the independently computed
# Cavgd28 above. Both read the same solved profile on the same grid, so
# this compares PKNCA's integration against the hand-rolled trapezoid and
# nothing else -- a tight bound is the correct assertion here.
cav_pknca <- as.data.frame(res_nca) |>
  dplyr::filter(PPTESTCD == "cav", start == 0) |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(cav = gm(PPORRES), .groups = "drop")

cav_manual <- expo |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(cav = gm(Cavgd28), .groups = "drop")

chk <- dplyr::inner_join(cav_pknca, cav_manual, by = "treatment",
                         suffix = c("_pknca", "_manual")) |>
  dplyr::mutate(pct_diff = 100 * (cav_pknca - cav_manual) / cav_manual)
chk

stopifnot(all(abs(chk$pct_diff) < 2))
```

# Comparison against the published value

The paper reports no NCA point estimates for the two CheckMate 67T arms
in its text or tables (Figures 3-5 present them graphically only). The
one printed exposure value that can be compared is the Figure 2 legend's
steady-state peak concentration for intravenous nivolumab 10 mg/kg q2w:
**352 ug/mL**.

Simulating that regimen shows the printed value is reproduced by a
concentration sampled a couple of days after the steady-state dose --
not by the model's true end-of-infusion peak, which is about 25% higher.
That is exactly the NCA sampling bias this paper is about: a conventional
schedule has no end-of-infusion sample at steady state, so an
observed-data peak underestimates the model peak.

```{r anchor}
# Same q2w schedule and 1-hour infusion, 10 mg/kg instead of 3 mg/kg.
sim10 <- simulate_arm("i.v. 3 mg/kg q2w", mgkg = 10, label = "i.v. 10 mg/kg q2w")

peak_basis <- sim10 |>
  dplyr::filter(time %in% c(day336 + 1 / 24, day336 + 2)) |>
  dplyr::mutate(basis = ifelse(
    time == day336 + 1 / 24,
    "End of infusion (model peak)",
    "Conventional sample, 2 d post-dose"
  )) |>
  dplyr::group_by(basis) |>
  dplyr::summarise(cmax = gm(Cc), .groups = "drop")

simulated_ref <- peak_basis |>
  dplyr::transmute(basis, PPTESTCD = "cmax", PPORRES = cmax)

reference_ref <- data.frame(
  basis = peak_basis$basis,
  cmax  = 352
)

tbl <- nlmixr2lib::ncaComparisonTable(
  simulated_ref, reference_ref,
  by = "basis",
  units = c(cmax = "ug/mL"),
  tolerance_pct = 20
)
knitr::kable(tbl, caption = "Steady-state peak concentration after i.v. nivolumab 10 mg/kg q2w: Zhao 2025 Figure 2 legend (352 ug/mL) vs this model, under two sampling definitions.")
attr(tbl, "footnote")
```

```{r anchor-gate}
conv <- peak_basis$cmax[peak_basis$basis == "Conventional sample, 2 d post-dose"]
sc_cmaxss <- gm(expo$Cmaxss[expo$treatment == "s.c. 1200 mg q4w"])

stopifnot(
  # The printed 352 ug/mL is recovered to within 10% by a conventionally
  # timed post-dose sample at steady state.
  abs(100 * (conv - 352) / 352) < 10,
  # The paper's explicit qualitative claim: predicted s.c. Cmaxss is "well
  # below" the 10 mg/kg i.v. safety reference of 352 ug/mL.
  sc_cmaxss < 352 * 0.8
)
```

# Covariate effects reproduce Table S2

Each covariate coefficient is checked by perturbing one covariate at a
time in a typical-value (`zeroRe`) solve and comparing the resulting
clearance ratio against the **literal coefficient printed in Table S2**,
not against a model variable. This also exercises the two tumour-type
code paths, which the RCC cohort above leaves at 0.

```{r covariate-check}
mod_tv <- rxode2::zeroRe(mod)
ref_cov <- data.frame(
  id = 1, WT = 80, CRCL = 90, SEXF = 0, ECOG_GE1 = 0,
  TUMTP_GASTRIC = 0, TUMTP_HODGKIN_CLASSICAL = 0
)

cl_at_t0 <- function(cov) {
  ev <- data.frame(id = 1, time = c(0, 1e-6), amt = 0, rate = 0,
                   evid = 0, cmt = "central")
  rxode2::rxSolve(mod_tv, ev, cov, returnType = "data.frame",
                  addDosing = FALSE)$cl[1]
}

cl_ref <- cl_at_t0(ref_cov)

perturb <- function(field, value) {
  cov <- ref_cov
  cov[[field]] <- value
  cl_at_t0(cov) / cl_ref
}

cov_chk <- data.frame(
  Covariate = c("WT 160 vs 80 kg", "eGFR 180 vs 90", "Female vs male",
                "ECOG >= 1 vs 0", "Gastric cancer", "Classical Hodgkin lymphoma"),
  expected = c(2^0.622, 2^0.139, exp(-0.158), exp(0.174),
               exp(0.180), exp(-0.330)),
  observed = c(perturb("WT", 160), perturb("CRCL", 180), perturb("SEXF", 1),
               perturb("ECOG_GE1", 1), perturb("TUMTP_GASTRIC", 1),
               perturb("TUMTP_HODGKIN_CLASSICAL", 1))
) |>
  dplyr::mutate(pct_diff = 100 * (observed - expected) / expected)

cov_chk |>
  dplyr::mutate(dplyr::across(c(expected, observed), \(x) round(x, 4)),
                pct_diff = signif(pct_diff, 3)) |>
  dplyr::rename("CL ratio expected (Table S2)" = expected,
                "CL ratio from model" = observed, "% diff" = pct_diff) |>
  knitr::kable(caption = "Covariate effects on clearance versus the coefficients printed in Table S2.")

stopifnot(all(abs(cov_chk$pct_diff) < 1e-6))
```

The bioavailability covariate factors are checked the same way. They act
multiplicatively on the natural fraction scale before the logit
transform, so the typical `F` for each covariate cell is a plain product
of the Table S2 values.

```{r fdepot-check}
f_cell <- function(sexf, ecog) {
  cov <- ref_cov
  cov$SEXF <- sexf
  cov$ECOG_GE1 <- ecog
  ev <- data.frame(id = 1, time = c(0, 1e-6), amt = 0, rate = 0,
                   evid = 0, cmt = "depot")
  rxode2::rxSolve(mod_tv, ev, cov, returnType = "data.frame",
                  addDosing = FALSE)$fdepot[1]
}

f_chk <- data.frame(
  Cell = c("male, ECOG 0", "female, ECOG 0", "male, ECOG >= 1", "female, ECOG >= 1"),
  expected = c(0.752, 0.752 * 0.859, 0.752 * 1.07, 0.752 * 0.859 * 1.07),
  observed = c(f_cell(0, 0), f_cell(1, 0), f_cell(0, 1), f_cell(1, 1))
) |>
  dplyr::mutate(pct_diff = 100 * (observed - expected) / expected)

f_chk |>
  dplyr::mutate(dplyr::across(c(expected, observed), \(x) round(x, 4)),
                pct_diff = signif(pct_diff, 3)) |>
  dplyr::rename("F expected (Table S2)" = expected,
                "F from model" = observed, "% diff" = pct_diff) |>
  knitr::kable(caption = "Typical subcutaneous bioavailability by covariate cell versus Table S2.")

stopifnot(
  all(abs(f_chk$pct_diff) < 1e-6),
  # Every covariate cell stays strictly inside (0, 1), so the logit is finite.
  all(f_chk$observed > 0 & f_chk$observed < 1)
)
```

# Mass-balance check

With a time-varying clearance the usual `AUC * CL = dose` identity does
not apply, so mass balance is checked in its general form: for a
subcutaneous course, the absorbed dose minus the amount still in the body
must equal the cumulative eliminated amount, `integral of CL(t) * Cc(t) dt`.
Both sides come from the same dense solve, so the residual is pure
numerical integration error and a tight bound is appropriate.

```{r mass-balance}
ev_mb <- rbind(
  data.frame(id = 1, time = seq(0, by = 28, length.out = 7), amt = 1200,
             rate = 0, evid = 1, cmt = "depot"),
  data.frame(id = 1, time = seq(0, 196, by = 0.02), amt = 0, rate = 0,
             evid = 0, cmt = "depot")
)
mb <- rxode2::rxSolve(mod_tv, ev_mb, ref_cov, returnType = "data.frame",
                      addDosing = FALSE) |>
  dplyr::arrange(time) |>
  dplyr::filter(!duplicated(time))

absorbed  <- 7 * 1200 * mb$fdepot[1]
remaining <- with(mb[nrow(mb), ], depot + central + peripheral1)
eliminated <- trap(mb$time, mb$cl * mb$Cc)

c(absorbed = absorbed, remaining = remaining, eliminated = eliminated,
  residual_pct = 100 * (absorbed - remaining - eliminated) / absorbed)

stopifnot(abs(absorbed - remaining - eliminated) / absorbed < 0.001)
```

# Assumptions and deviations

## Errata and readings that required arbitration

1. **Table S2 mislabels the units of CL and Q.** The table heads both
   `CL0_REF` and `Q_REF` with `[mL/hr]` but prints 0.0109 and 0.0327.
   Those values are **L/h**. Three independent proofs: both control
   streams annotate the same `$THETA` entries `CL [L/h]` and `Q [L/h]`;
   Table S3 reports the same two parameters as 10.6-11.0 and 32.4-32.5
   under an `mL/hr` label, i.e. 1000-fold larger; and 10.9 mL/h is the
   physiologically expected clearance of an IgG4 antibody. The model uses
   0.0109 L/h and 0.0327 L/h, converted to L/day.

2. **Reference body weight: 80 kg, not 75 kg.** The explanatory note
   under Table S2 describes the reference subject as "60-year old white
   male, weighing 75 kg". Both control streams instead set
   `BBWT_R = 80 ; refernce value, kg` [sic] and `BGFR_R = 90`. The
   control-stream value governs, because it is what the estimation
   actually used to produce the Table S2 estimates. The note also cites
   age and race, neither of which is in the final covariate model (the
   stream header reads `[No LDH/Albumin/Hepa/Age]` and both race THETAs
   are commented out), which indicates the note is stale text carried
   from an earlier nivolumab analysis.

3. **The residual error 0.204 is a standard deviation, not a variance.**
   Table S2's footnote b says random-effect *and residual-error* values
   are shown as "variance (standard deviation)", but the residual row
   carries no parenthetical. The control streams settle it: `$SIGMA 1 FIX`
   with `REWT = F*PERR + AERR`, `AERR = 0` and `Y = IPRED + REWT*EPS(1)`
   means `THETA(5) = PERR` multiplies the prediction directly, so 0.204 is
   the proportional SD (20.4% CV). Encoded as `propSd <- 0.204`.

4. **`Ka_REF` is printed per day while the model runs per hour.** Table S2
   reports `Ka_REF (day-1) = 0.307`; the `$THETAP` prior gives `KA = 0.013`
   on the stream's native hourly scale. `0.307 / 24 = 0.0128`, confirming
   both the value and the printed time unit. This model keeps time in days
   and so uses 0.307 /day directly.

5. **Study count.** Table 1 footnote b states 18 intravenous studies;
   Table S1 lists 19 (plus CheckMate 8KX for the subcutaneous arm). The
   model metadata records 20 studies in total and notes the discrepancy.

6. **RCC subject count.** Methods section 2.3 states "all patients with
   RCC (N = 715)" while Table 1's RCC row sums to 713 (17 subcutaneous +
   696 intravenous). Not load-bearing for the model.

7. **Simulation duration.** Methods section 2.3 says profiles were
   simulated "over a period of 16 weeks (12 doses for nivolumab s.c. and
   25 doses for nivolumab i.v.)". Those dose counts span roughly 48
   weeks, not 16, and the steady-state landmark is day 336 (= 48 weeks),
   so "16 weeks" appears to be an error. This vignette simulates to day
   364.

8. **The 352 ug/mL Figure 2 reference is not reproduced as an
   end-of-infusion peak.** The model's true steady-state peak for i.v.
   10 mg/kg q2w is about 438 ug/mL (geometric mean); the printed 352
   ug/mL is matched to within 2% by a sample taken about 2 days after the
   steady-state dose. The legend describes the value as "simulated" but
   attaches it to a regimen whose safety margins come from a separate
   published analysis (reference 17, Agrawal 2016), which is not on disk,
   so whether the number came from this model under a coarse sampling
   grid or from that earlier analysis cannot be settled from the sources
   available. No parameter was adjusted to close the gap.

## Assumptions made because the paper does not say

* **RCC-specific baseline demographics are unpublished.** Table 1 reports
  the pooled cohort only, so the virtual cohort is drawn from those
  overall marginals with both tumour-type indicators set to 0. The paper
  instead bootstrap-resampled real RCC patients, which preserves the
  observed correlations between weight, sex, eGFR and performance status;
  the independent marginals used here do not.
* **eGFR estimating equation is not named.** Table 1 gives eGFR in
  mL/min/1.73 m^2 while the control-stream comment abbreviates the unit
  to mL/min. The column is treated as BSA-normalised and stored under the
  canonical `CRCL`. The sibling nivolumab analysis (Bajaj 2017) used
  CKD-EPI.
* **Intravenous infusion duration.** Table S1 records 1-hour infusions
  for nearly all pooled studies (60 minutes for MDX1106-01), and
  CheckMate 67T used 30-minute infusions. This vignette uses 1 hour. The
  choice affects only the peak immediately after the infusion, not
  `Cavgd28` or trough measures.
* **Uncertainty in the parameter estimates is not propagated.** The paper
  sampled 100 parameter sets from the variance-covariance matrix of the
  estimates (87 of which had a positive-definite OMEGA) to generate its
  90% prediction intervals. The full matrix is not published -- Table S2
  gives only standard errors and `$THETAPV` gives a diagonal prior
  variance block -- so the intervals here reflect inter-individual
  variability at the point estimate only, and are narrower than the
  paper's.

## Documented convention deviations

* `cl_time_max`, `cl_t50` and `cl_time_hill` are not log-transformed,
  which `checkModelConventions()` flags as a warning for fixed-effect PK
  parameters. This is deliberate and matches the sibling
  `Bajaj_2017_nivolumab.R`, which carries the identical three warnings on
  `main`: `cl_time_max` is negative (-0.303) and therefore cannot be held
  on the log scale, and it carries an **additive** random effect
  (`EMAX = AEMAX + ZEMAX` in the source), so a log parameterisation would
  misrepresent both the sign and the IIV structure.
* The model carries no `ROUTE_IV` covariate even though the source gates
  `ka` and `F` on a `ROUTEN` column. Those two parameters only ever act on
  depot doses, so gating them is redundant once the route is expressed by
  the dosing compartment -- which is what the `ROUTE_IV` register entry
  itself recommends ("When simulating, set `ROUTE_IV = 1` for IV cohorts
  and dose into the central compartment; set `ROUTE_IV = 0` for SC cohorts
  and dose into the depot"). The source's `IF (ROUTEN.EQ.0) TVKA = 1` and
  `TVF1 = 0.4` branches are inert placeholders for intravenous records,
  which `ADVAN4` doses into compartment 2 where `F1` does not apply.
* Covariates screened but not retained (age, albumin, LDH, hepatic
  function, Asian and Black race, and weight/sex/PS effects on `ka`) are
  recorded in `covariatesDataExcluded` rather than `covariateData`,
  because the source reports no point estimate for any of them.

# Session info

```{r session}
sessionInfo()
```
