---
title: "Zheng_2016_sifalimumab"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Zheng_2016_sifalimumab}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

```{r setup}
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
library(PKNCA)
```

# Sifalimumab population PK simulation

Simulate sifalimumab serum concentration-time profiles using the final
population PK model from Zheng et al. 2016. Sifalimumab is a human
IgG1 kappa monoclonal antibody against interferon-alpha (IFN-alpha)
developed for systemic lupus erythematosus (SLE). The Zheng 2016
analysis used data from a phase IIb trial (MI-CP180) in which
298 SLE patients received fixed intravenous doses of 200, 600, or
1200 mg every 4 weeks for 52 weeks (14 doses total, with an
additional loading dose on day 15). A two-compartment model with
first-order elimination from the central compartment adequately
described the serum concentration-time data pooled across dose
cohorts. Body weight, baseline 21-gene type I interferon signature
score, dose, and baseline steroid use were identified as
statistically significant covariates on CL and V1 but accounted for
< 10% of PK variability.

Article: [Br J Clin Pharmacol 81(5):918-928](https://doi.org/10.1111/bcp.12864)

## Population

From Zheng 2016 Table 1: 298 SLE patients received sifalimumab
(102 at 200 mg, 102 at 600 mg, 94 at 1200 mg) out of 431 enrolled.
Baseline demographics (combined sifalimumab arm): median age
40 years (18-73), median weight 64.3 kg (39-131.3), 92% female,
59% White / 15% Asian / 7% African American / 4% Native American /
15% Other. 85% of patients were on baseline (concomitant)
corticosteroids. Median baseline 21-gene IFN signature score was
12.04 (range 0.32-38.59); median baseline SLEDAI-2K score was 10
(range 6-27). The population PK dataset comprised 3961 quantifiable
serum concentrations.

The same information is available programmatically as
`readModelDb("Zheng_2016_sifalimumab")` (the returned function's
body holds the `population` list literal).

## Source trace

| Element                                        | Source location                               | Value / form                                              |
| ---------------------------------------------- | --------------------------------------------- | --------------------------------------------------------- |
| Structural model                               | Zheng 2016 Results, "Base and final model"    | 2-compartment IV, first-order elimination                 |
| CL (typical patient)                           | Zheng 2016 Table 2, theta1                    | 0.184 L/day                                               |
| V1 (typical patient)                           | Zheng 2016 Table 2, theta2                    | 2.82 L                                                    |
| V2                                             | Zheng 2016 Table 2, theta3                    | 1.88 L                                                    |
| Q                                              | Zheng 2016 Table 2, theta4                    | 0.34 L/day                                                |
| WT exponent on CL                              | Zheng 2016 Table 2, theta5                    | 0.45 (reference 64.3 kg)                                  |
| BGENE21 exponent on CL                         | Zheng 2016 Table 2, theta6                    | 0.09 (reference 12.04)                                    |
| STEROID_BL fractional effect on CL             | Zheng 2016 Table 2, theta7                    | +0.11 (so CL x (1 + 0.11 * STEROID_BL))                   |
| WT exponent on V1                              | Zheng 2016 Table 2, theta8                    | 0.36 (reference 64.3 kg)                                  |
| DOSE exponent on V1                            | Zheng 2016 Table 2, theta9                    | 0.06 (reference 600 mg)                                   |
| STEROID_BL fractional effect on V1             | Zheng 2016 Table 2, theta10                   | -0.09 (so V1 x (1 - 0.09 * STEROID_BL))                   |
| BSV on CL (log-normal)                         | Zheng 2016 Table 2, BSV (CV %)                | 24% CV (omega^2 = log(CV^2 + 1) = 0.056016)               |
| BSV on V1 (log-normal)                         | Zheng 2016 Table 2, BSV (CV %)                | 16% CV (omega^2 = 0.025278)                               |
| IOV on V1 (per-occasion)                       | Zheng 2016 Table 2 / IOV paragraph            | 26% CV, seven infusions; **not implemented** (see below)  |
| BSV correlations                               | Zheng 2016 Results                            | None significant; IIV treated as diagonal                 |
| Proportional residual error                    | Zheng 2016 Table 2                            | 7.01% -> propSd = 0.0701 (fraction)                       |
| Additive residual error                        | Zheng 2016 Table 2                            | 1.14 ug/mL                                                |
| Covariate equation CL                          | Zheng 2016 Equations (Results)                | CL = theta1 * (WT/64.3)^0.45 * (BGENE21/12.04)^0.09 * (1 + 0.11 * STEROID_BL) |
| Covariate equation V1                          | Zheng 2016 Equations (Results)                | V1 = theta2 * (WT/64.3)^0.36 * (DOSE/600)^0.06 * (1 - 0.09 * STEROID_BL)       |
| Dose regimens                                  | Zheng 2016 Table 1 and Methods                | 200, 600, or 1200 mg IV q4w for 52 weeks; loading dose day 15 |

## Virtual cohort

The per-subject covariates were not released with the paper. We
build a 298-subject virtual cohort whose cohort sizes match Table 1
exactly (102/102/94 at 200/600/1200 mg) and whose body weights and
baseline 21-gene IFN signature scores are sampled from log-normal
distributions bounded to the published ranges (39-131.3 kg and
0.32-38.59 respectively). Baseline steroid use is drawn as a
Bernoulli variable with p = 0.85 to match the published 85% prevalence.

```{r virtual-pop}
set.seed(2016)

cohorts <- tribble(
  ~treatment,    ~dose_mg, ~n,
  "200 mg",         200,   102L,
  "600 mg",         600,   102L,
  "1200 mg",       1200,    94L
)

pop <- cohorts %>%
  rowwise() %>%
  do(tibble(treatment = .$treatment, dose_mg = .$dose_mg,
            subj = seq_len(.$n))) %>%
  ungroup() %>%
  mutate(
    ID         = dplyr::row_number(),
    WT         = pmin(pmax(rlnorm(dplyr::n(), meanlog = log(64.3), sdlog = 0.22),
                           39), 131.3),
    BGENE21    = pmin(pmax(rlnorm(dplyr::n(), meanlog = log(12.04), sdlog = 0.8),
                           0.32), 38.59),
    STEROID_BL = as.integer(runif(dplyr::n()) < 0.85),
    DOSE       = dose_mg
  ) %>%
  select(ID, treatment, dose_mg, WT, BGENE21, STEROID_BL, DOSE)

pop
```

## Dosing and event table

Sifalimumab was given as a 30-60 minute IV infusion every 4 weeks
for 52 weeks, with an additional loading dose on day 15. The full
schedule is 14 IV doses on days 1, 15, 29, 57, 85, 113, 141, 169,
197, 225, 253, 281, 309, and 337. We use a 30-minute infusion for
simulation. Observations are scheduled to give dense coverage of the
first dose, trough values at each dosing visit, a full profile after
the final (14th) dose, and follow-up to day 450 to capture terminal
elimination.

```{r events}
dose_days <- c(1, 15, 29, 57, 85, 113, 141, 169, 197, 225,
               253, 281, 309, 337) - 1  # model time starts at 0 on day 1
infusion_dur <- 30 / (60 * 24)  # 30 min in days

# Observation times: peri-dose window on dose days, dense after last dose
peri_dose <- sort(unique(unlist(lapply(dose_days, function(d) {
  d + c(-0.01, 0, 0.25, 1, 3, 7)  # pre-dose, end-of-infusion-ish, day 1, 3, 7 post
}))))
peri_dose <- peri_dose[peri_dose >= 0]

terminal <- seq(from = max(dose_days), to = max(dose_days) + 112, by = 3)

obs_times <- sort(unique(c(peri_dose, terminal)))

dose_rows <- pop %>%
  tidyr::crossing(time = dose_days) %>%
  transmute(
    ID, treatment, WT, BGENE21, STEROID_BL, DOSE,
    time,
    amt  = dose_mg,
    evid = 1L,
    cmt  = "central",
    dur  = infusion_dur,
    dv   = NA_real_
  )

obs_rows <- pop %>%
  select(ID, treatment, WT, BGENE21, STEROID_BL, DOSE) %>%
  tidyr::crossing(time = obs_times) %>%
  mutate(
    amt  = NA_real_,
    evid = 0L,
    cmt  = NA_character_,
    dur  = NA_real_,
    dv   = NA_real_
  )

events <- bind_rows(dose_rows, obs_rows) %>%
  arrange(ID, time, desc(evid))
```

## Simulation

Simulate with between-subject variability so the spread across the
virtual cohort matches the paper's reported variability.

```{r simulate}
mod <- readModelDb("Zheng_2016_sifalimumab")

events_sim <- events %>% rename(id = ID)
sim <- rxSolve(object = mod, events = events_sim, returnType = "data.frame") %>%
  as_tibble() %>%
  left_join(pop %>% select(ID, treatment, dose_mg),
            by = c("id" = "ID"))
```

## Replicate Figure 2: representative concentration-time profiles

Figure 2 of Zheng 2016 shows representative individual sifalimumab
serum concentration-time profiles for patients in each of the three
dose cohorts, overlaying observed and model-predicted concentrations
on a log-linear scale across the 52-week dosing interval. Below we
plot simulated median (and 5th-95th percentile) concentrations by
dose cohort.

```{r figure-2, fig.width = 7.5, fig.height = 5}
profile <- sim %>%
  filter(time > 0, !is.na(Cc), Cc > 0) %>%
  group_by(treatment, time) %>%
  summarise(
    Q05 = quantile(Cc, 0.05, na.rm = TRUE),
    Q50 = quantile(Cc, 0.50, na.rm = TRUE),
    Q95 = quantile(Cc, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

profile$treatment <- factor(profile$treatment,
                            levels = c("200 mg", "600 mg", "1200 mg"))

ggplot(profile, aes(x = time, y = Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ treatment) +
  scale_y_log10() +
  labs(
    x       = "Time since first dose (days)",
    y       = "Sifalimumab serum concentration (ug/mL, log scale)",
    title   = "Figure 2 replication - simulated concentration-time profiles by cohort",
    caption = "Replicates Figure 2 of Zheng et al. 2016 (median with 5th-95th percentile band)."
  ) +
  theme_bw()
```

# PKNCA validation

Compute non-compartmental Cmax, Cmin/Ctrough, AUC(tau), and
terminal half-life per subject per cohort at steady state (dosing
interval between the 13th and 14th doses, i.e., days 309-337) and
compare to the published exposure summary. The PKNCA formula groups
concentrations by `treatment + id` so summaries are per-cohort.

```{r pknca-prep}
# Steady-state interval: between the last two doses
ss_start <- dose_days[length(dose_days) - 1]  # day of 13th dose
ss_end   <- dose_days[length(dose_days)]      # day of 14th dose (last)

nca_conc <- sim %>%
  filter(time >= ss_start, time <= ss_end, !is.na(Cc)) %>%
  mutate(time_in_interval = time - ss_start) %>%
  select(id, time = time_in_interval, Cc, treatment)

nca_dose <- pop %>%
  transmute(id = ID, time = 0, amt = dose_mg, treatment)
```

```{r pknca-run, message = FALSE, warning = FALSE}
conc_obj <- PKNCAconc(nca_conc, Cc ~ time | treatment + id)
dose_obj <- PKNCAdose(nca_dose, amt ~ time | treatment + id)

intervals <- data.frame(
  start       = 0,
  end         = 28,
  cmax        = TRUE,
  cmin        = TRUE,
  tmax        = TRUE,
  auclast     = TRUE
)

nca_data <- PKNCAdata(conc_obj, dose_obj, intervals = intervals)
nca_res  <- pk.nca(nca_data)

sim_nca <- as.data.frame(nca_res$result) %>%
  filter(PPTESTCD %in% c("cmax", "cmin", "auclast")) %>%
  group_by(treatment, PPTESTCD) %>%
  summarise(
    mean = mean(PPORRES, na.rm = TRUE),
    sd   = sd(PPORRES,   na.rm = TRUE),
    med  = median(PPORRES, na.rm = TRUE),
    min  = min(PPORRES, na.rm = TRUE),
    max  = max(PPORRES, na.rm = TRUE),
    .groups = "drop"
  )

sim_nca$treatment <- factor(sim_nca$treatment,
                            levels = c("200 mg", "600 mg", "1200 mg"))
sim_nca <- sim_nca %>% arrange(treatment, PPTESTCD)

knitr::kable(sim_nca,
             digits  = 2,
             caption = "Steady-state NCA summary (between 13th and 14th doses) from the simulated 298-subject virtual cohort: Cmax and Cmin in ug/mL, AUC(tau=28 d) in ug*day/mL.")
```

## Comparison against Zheng 2016 exposure summary

Zheng 2016 does not publish a tabulated steady-state NCA by dose
cohort. Instead the paper reports (Results, "Steady-state exposures"
paragraph):

* **Peak serum concentrations (Cmax,ss) range 117-562 ug/mL** across
  the three cohorts combined.
* **Cmax,ss, Ctrough,ss and AUC(0,tau) increased in a dose-proportional
  manner** across 200-1200 mg.
* **Typical clearance** of 0.184 L/day implies a typical terminal
  half-life of roughly 24 days given the V1 = 2.82 L, V2 = 1.88 L,
  Q = 0.34 L/day structural parameters.

The simulated Cmax,ss range across the 298-subject virtual cohort
should span this 117-562 ug/mL envelope and scale approximately
linearly across the three dose cohorts.

```{r exposure-envelope}
cmax_range <- sim_nca %>%
  filter(PPTESTCD == "cmax") %>%
  transmute(treatment,
            cmax_mean = mean,
            cmax_med  = med,
            cmax_min  = min,
            cmax_max  = max) %>%
  arrange(treatment)

knitr::kable(cmax_range,
             digits  = 1,
             caption = "Simulated Cmax,ss by cohort (ug/mL). Overall min-max across cohorts should bracket the Zheng 2016 reported 117-562 ug/mL envelope.")
```

Dose-proportionality holds exactly by construction in the packaged
model (linear elimination, no non-linear saturation terms); the
small `(DOSE/600)^0.06` covariate on V1 introduces only a very
weak deviation (about +/-5% across the 200-1200 mg range), which
is consistent with the paper's description that the covariate
accounts for < 10% of PK variability.

# Assumptions and deviations

* **Weight distribution.** Zheng 2016 publishes only summary
  statistics (median 64.3 kg, range 39-131.3 kg). We draw weights
  from a log-normal distribution with median 64.3 kg and log-scale
  SD 0.22, clipped to the published range.
* **BGENE21 distribution.** Only median (12.04) and range (0.32-38.59)
  are published. We draw values from a log-normal distribution with
  median 12.04 and log-scale SD 0.8, clipped to the published range.
* **Steroid use correlation with weight.** The 85% steroid-use
  prevalence is sampled independently of weight and BGENE21; the
  paper does not describe joint distributions.
* **Time-varying weight.** Body weight is treated as a time-fixed
  baseline covariate in the simulation. The paper does not discuss
  intra-subject weight trajectories over the 52-week treatment
  period.
* **Inter-occasion variability (IOV) on V1.** The published final
  model included 26% CV IOV on V1 for the first seven infusions
  only. Occasion-specific random effects are outside the scope of a
  typical single-eta subject-level library model, so IOV is **not**
  implemented. The consequence is that individual peri-dose
  fluctuations in the simulation will be somewhat tighter than in
  the observed data; steady-state exposure summaries (Cmax,ss,
  AUC(tau), Ctrough,ss) across the full cohort are not materially
  affected.
* **Infusion duration.** Zheng 2016 specifies a 30-60 min IV
  infusion; we use 30 min. The choice is numerically immaterial
  given the 28-day dosing interval and ~24-day half-life.
* **Dose as covariate on V1.** The source uses a continuous
  `(Dose/600)^0.06` covariate on V1 driven by each subject's
  assigned fixed dose level. We preserve that parameterisation
  exactly; each subject's `DOSE` covariate equals the regimen
  dose (200, 600, or 1200 mg).

# Notes

* **Structural model:** 2-compartment with IV input into `central`,
  first-order elimination from `central`, no explicit depot or
  absorption compartment.
* **IIV:** diagonal matrix on CL and V1; no BSV on Q or V2.
* **Covariates:** WT, BGENE21, STEROID_BL on CL; WT, DOSE,
  STEROID_BL on V1. Collectively these accounted for < 10% of PK
  variability in the published final model.
* **Typical terminal half-life** predicted from the structural
  parameters for a typical 64.3 kg patient with BGENE21 = 12.04,
  DOSE = 600 mg, STEROID_BL = 0 is approximately 24 days
  (dominated by the central-compartment elimination rate
  CL/V1 = 0.184/2.82 = 0.0652 day^-1 plus the small peripheral
  redistribution term), consistent with the Results narrative.

## Reference

* `r readModelDb("Zheng_2016_sifalimumab")()$meta$reference`
