---
title: "Alisertib global popPK and exposure-safety (Zhou 2018)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Alisertib global popPK and exposure-safety (Zhou 2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5
)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhou_2018_alisertib"))$reference`
* Article: <https://doi.org/10.1111/bcp.13430>
* Supplement (Table S1, the *updated* final model): <http://onlinelibrary.wiley.com/doi/10.1111/bcp.13430/suppinfo>

This paper contributes **four** models to `nlmixr2lib`:

```{r model-list}
models <- c(
  "Zhou_2018_alisertib",
  "Zhou_2018_alisertib_neutropenia",
  "Zhou_2018_alisertib_stomatitis",
  "Zhou_2018_alisertib_diarrhea"
)
ui <- lapply(models, \(n) rxode2::rxode(readModelDb(n)))
names(ui) <- models

tibble::tibble(
  Model = models,
  Kind = c(
    "Population PK (2-compartment, 4-transit absorption)",
    "Exposure-safety logistic: grade >= 3 neutropenia",
    "Exposure-safety logistic: grade >= 2 stomatitis",
    "Exposure-safety logistic: grade >= 2 diarrhoea"
  ),
  n = c(422L, 591L, 593L, 594L)
) |>
  knitr::kable(caption = "Models extracted from Zhou 2018.")
```

Alisertib (MLN8237) is an investigational selective Aurora A kinase inhibitor.
The analysis pooled 10 clinical studies to ask a specific regulatory question:
why is the maximum tolerated dose 50 mg twice daily in the West but only 30 mg
twice daily in East Asia? The answer the model gives is a **52% higher relative
bioavailability** in patients enrolled in East Asia, and the paper uses that to
justify a 40% lower regional dose.

# Population

The model packaged here was fitted to the **422-patient analysis data set**: adults
with advanced haematological (43%) and nonhaematological (57%) malignancies,
median age 62 years (range 21-85), median weight 73.3 kg (42.6-175.0), median BSA
1.84 m^2 (1.36-2.97), 46% female. 363 were enrolled in Western countries and 59 in
the East Asian region (Japan; and Singapore, Taiwan, Hong Kong, South Korea).
Alisertib was given orally at 5-150 mg once or twice daily under nil-per-os
conditions. A further 249 patients -- the phase II portion of study C14007,
entirely Western -- formed an external validation set. Baseline demographics are
Zhou 2018 Tables 2 (continuous) and 3 (categorical); the study list is Table 1.

```{r population}
str(ui[["Zhou_2018_alisertib"]]$population, max.level = 1)
```

## Molar units, and the mg-to-nmol conversion

Every concentration in Zhou 2018 is **molar**: the additive residual error is
reported in nmol/L, Figure 4 plots nmol/L, and the exposure tables are in nM.h
and uM.h. The packaged model therefore works in nmol and nmol/L, and a dose in
milligrams must be converted before it is given to `rxSolve()`.

The molar mass is **not** stated anywhere in the paper, so it is taken from an
external source and flagged as such throughout (see *Assumptions and
deviations*). The paper's own simulation output corroborates it closely.

```{r molar}
# NOT FROM THE PAPER: alisertib is C27H20ClFN4O4, molar mass 518.92 g/mol
# (PubChem CID 24771867 / the published structure). Zhou 2018 reports no
# molar mass, but its own simulated exposures reproduce under this value --
# checked explicitly in the "Figure 6" section below.
MW_ALISERTIB <- 518.92
mg_to_nmol <- function(mg) mg / MW_ALISERTIB * 1e6

dose_west <- mg_to_nmol(50)
dose_east <- mg_to_nmol(30)
c(`50 mg (nmol)` = dose_west, `30 mg (nmol)` = dose_east)
```

# Source trace

Each `ini()` entry in the four model files carries its own in-file provenance
comment. They are collected here for review.

```{r source-trace, echo = FALSE}
tibble::tribble(
  ~Model, ~`Equation / parameter`, ~Value, ~`Source location`,
  "PK", "`lcl` (CL/F)", "4.11 L/h", "Table 5, row CL/F (SE 3.1%)",
  "PK", "`lvc` (V1/F)", "54.3 L", "Table 5, row V1/F (SE 3.9%)",
  "PK", "`lq` (Q/F)", "7.07 L/h", "Table 5, row Q/F (SE 10.9%)",
  "PK", "`lvp` (V2/F)", "28.7 L", "Table 5, row V2/F (SE 10%)",
  "PK", "`lktr` (KTR)", "4.17 1/h", "Table 5, row KTR (SE 3%)",
  "PK", "`e_region_eastasia_f` (RGNF)", "-0.341", "Table 5, row RGNF (SE 11%); sign from the Table 5 footnote and Supplementary Table S1",
  "PK", "`e_bsa_vc` (BSAV1)", "0.899", "Table 5, row BSAV1 (SE 26.3%)",
  "PK", "BSA reference", "1.84 m^2", "Figure 1 equation `(BSA/1.84)^BSAV1`; the Table 2 median",
  "PK", "IIV block (log SD)", "0.518 / 0.412 / 1.044 / 0.54", "Table 5, 'IIV, ratio' column for CL/F, V1/F, V2/F, KTR",
  "PK", "IIV correlations", "0.582, 0.509, 0.085, 0.116, 0.255, 0.025", "Table 5, lower-triangular correlation block",
  "PK", "`propSd` (CCV)", "0.491", "Table 5, row CCV (SE 3.5%)",
  "PK", "`addSd` (ADD)", "0 (fixed)", "Table 5, row ADD; footnote b 'automatically set to zero by the SAEM estimation algorithm'",
  "PK", "4-transit absorption chain; dose into transit1", "n/a", "Figure 1 schematic; confirmed by Results 'net mean transit time 0.96 h' = 4/KTR",
  "PK", "Regional multiplier `(1 + f)` on CL/F, V1/F, Q/F, V2/F", "n/a", "Figure 1 equations",
  "PK", "No IIV on Q/F", "n/a", "Figure 1 (Q/F written without an exponential term); Table 5 IIV cell blank; Methods 'the effect of removing BSV from Q/F was examined'",
  "ER", "logistic form `logit(p) = b0 + b1*log(AUC)`", "n/a", "Methods, 'Exposure-safety analyses'; confirmed by the shape of the Figure 7 curves",
  "ER neutropenia", "`logit_ref` / `e_auc_logit`", "-2.316 / 0.675", "DIGITISED from Figure 7 panel A -- not tabulated in the source",
  "ER stomatitis", "`logit_ref` / `e_auc_logit`", "-6.485 / 1.434", "DIGITISED from Figure 7 panel B -- not tabulated in the source",
  "ER diarrhoea", "`logit_ref` / `e_auc_logit`", "-3.752 / 0.677", "DIGITISED from Figure 7 panel C -- not tabulated in the source"
) |>
  knitr::kable(caption = "Source trace for all four Zhou 2018 models.")
```

## Which Table is the final model?

Zhou 2018 reports the model **twice**. Table 5 is the final model fitted to the
422-patient analysis set; Supplementary Table S1 is an *updated* final model
refitted after the 249-patient validation set was folded in. Every row differs.
**Table 5 is what is packaged here**; S1 is recorded below so a reader is not
misled by a value that looks like a transcription error.

```{r table-s1, echo = FALSE}
tibble::tribble(
  ~Parameter, ~`Table 5 (packaged)`, ~`Table S1 (updated)`,
  "CL/F (L/h)", "4.11", "3.91",
  "V1/F (L)", "54.3", "51.8",
  "Q/F (L/h)", "7.07", "7.53",
  "V2/F (L)", "28.7", "27.4",
  "KTR (1/h)", "4.17", "3.93",
  "RGNF", "0.341 (sign dropped)", "-0.325",
  "BSAV1", "0.899", "0.79",
  "CCV", "0.491", "0.477",
  "ADD (nmol/L)", "0 (fixed)", "4.8"
) |>
  knitr::kable(caption = "Table 5 versus Supplementary Table S1. Only Table 5 is packaged.")
```

Note the RGNF row: the main-text Table 5 prints the magnitude `0.341` with a
footnote reading "the affected parameters were multiplied by (1-0.341)", while
the supplement prints `-0.325` **with** its minus sign under the identical
footnote wording. Figure 1 writes the effect as a multiplier `(1 + f)`. Taken
together these settle the sign that main-text typesetting dropped: the packaged
coefficient is `-0.341`, the multiplier is `0.659`, and `1/0.659 = 1.517`
recovers the "51.7% higher F" quoted in Results.

# Closed-form structural gates

Three quantities are fully determined by the Table 5 point estimates and are
quoted independently in the paper's Results text. They are exact arithmetic --
no simulation, no cohort, no randomness -- so they are asserted tightly.

```{r closed-form}
th <- ui[["Zhou_2018_alisertib"]]$theta
cl <- exp(th[["lcl"]]); vc <- exp(th[["lvc"]])
q <- exp(th[["lq"]]); vp <- exp(th[["lvp"]]); ktr <- exp(th[["lktr"]])

# Two-compartment hybrid rate constants from the packaged parameters
k10 <- cl / vc; k12 <- q / vc; k21 <- q / vp
s <- k10 + k12 + k21
lam1 <- (s + sqrt(s^2 - 4 * k10 * k21)) / 2   # alpha
lam2 <- (s - sqrt(s^2 - 4 * k10 * k21)) / 2   # beta

gates <- tibble::tibble(
  Quantity = c(
    "Distribution (alpha) half-life (h)",
    "Terminal (beta) half-life (h)",
    "Net mean oral transit time 4/KTR (h)",
    "Relative bioavailability, East Asia vs West"
  ),
  Published = c(1.71, 15.1, 0.96, 1.517),
  Computed = c(
    log(2) / lam1,
    log(2) / lam2,
    4 / ktr,
    1 / (1 + th[["e_region_eastasia_f"]])
  )
) |>
  dplyr::mutate(`Diff (%)` = 100 * (Computed - Published) / Published)

knitr::kable(gates, digits = 3, caption =
  "Closed-form gates. The published 51.7% higher F is Results; the half-lives and transit time are Results.")

# Deterministic arithmetic on printed numbers -- a tight bound is correct here.
stopifnot(max(abs(gates$`Diff (%)`)) < 0.5)
```

The beta half-life gate is load-bearing for a second reason. The disposition is
written with explicit `k12` / `k21` micro-constants, and `rxSolve()`'s default
`useLinCmt = TRUE` silently discards `peripheral1` for models of this shape --
which leaves AUC untouched and so is invisible to any exposure-only check. Every
`rxSolve()` call below therefore passes `useLinCmt = FALSE`, and the terminal
slope is checked explicitly against 15.1 h.

```{r use-lincmt-check}
mod <- readModelDb("Zhou_2018_alisertib")
mod_typ <- rxode2::zeroRe(mod)

ev_wash <- rxode2::et(amt = dose_west, cmt = "transit1") |>
  rxode2::et(seq(0, 168, by = 0.5), cmt = "central")
d_wash <- as.data.frame(ev_wash)
d_wash$BSA <- 1.84
d_wash$REGION_EASTASIA <- 0

wash <- rxode2::rxSolve(mod_typ, d_wash, returnType = "data.frame",
                        useLinCmt = FALSE)
late <- wash[wash$time >= 96 & wash$time <= 168 & wash$Cc > 0, ]
slope <- stats::coef(stats::lm(log(late$Cc) ~ late$time))[[2]]
thalf_sim <- log(2) / -slope

c(`terminal t1/2 from the solved model (h)` = thalf_sim,
  `published (h)` = 15.1)

# Noise-free deterministic solve of a linear model: a tight bound is correct.
# A collapse to one compartment would read log(2)/k10 = 9.2 h and blow this.
stopifnot(abs(thalf_sim - 15.1) / 15.1 < 0.02)
```

# Virtual cohort

Observed individual data are not public. Five arms are simulated: three
fixed-BSA Western arms reproducing Figure 4, and the two regional arms of
Figure 6. **200 subjects per arm** (Zhou 2018 used 200/arm for Figure 4 and
1000/arm for Figure 6; the library caps cohorts at 200/arm, which only widens
the Monte-Carlo noise on the Figure 6 comparison).

```{r cohort}
# set.seed() seeds R's RNG, NOT rxode2's simulation RNG, whose streams are
# partitioned per solver thread. The cohort below therefore differs between a
# 2-core CI runner and a 16-thread workstation, and no seed can make them
# agree. Every assertion downstream is written to hold for ANY cohort this
# model can produce.
set.seed(20180101)
rxode2::rxSetSeed(20180101)

N_ARM <- 200L

# Observation grid: fine over the last steady-state dosing interval (so Cmax
# and Cmin are resolved), coarse elsewhere.
grid <- sort(unique(c(
  seq(0, 168, by = 0.5),
  seq(144, 156, by = 0.05),
  seq(169, 288, by = 2)
)))

make_arm <- function(n, bsa, region, dose_nmol, arm, id_offset) {
  # 7 days of twice-daily dosing (14 doses) followed by 14 days off,
  # i.e. the recommended single-agent 21-day cycle.
  ev <- rxode2::et(amt = dose_nmol, cmt = "transit1", ii = 12, addl = 13) |>
    rxode2::et(grid, cmt = "central") |>
    rxode2::et(id = seq_len(n))
  d <- as.data.frame(ev)
  d$id <- d$id + id_offset
  # BSA is either a fixed percentile (Figure 4) or a lognormal draw (Figure 6).
  bsa_by_id <- if (length(bsa) == 1L) {
    stats::setNames(rep(bsa, n), seq_len(n) + id_offset)
  } else {
    stats::setNames(bsa, seq_len(n) + id_offset)
  }
  d$BSA <- unname(bsa_by_id[as.character(d$id)])
  d$REGION_EASTASIA <- region
  d$arm <- arm
  d
}

events <- dplyr::bind_rows(
  make_arm(N_ARM, 1.44, 0, dose_west, "West, BSA 2.5th pct (1.44)", 0L),
  make_arm(N_ARM, 1.84, 0, dose_west, "West, BSA 50th pct (1.84)", 1000L),
  make_arm(N_ARM, 2.43, 0, dose_west, "West, BSA 97.5th pct (2.43)", 2000L),
  make_arm(N_ARM, stats::rlnorm(N_ARM, log(1.88), 0.135), 0, dose_west,
           "West cohort, 50 mg b.i.d.", 3000L),
  make_arm(N_ARM, stats::rlnorm(N_ARM, log(1.63), 0.0862), 1, dose_east,
           "East Asia cohort, 30 mg b.i.d.", 4000L)
)

# Test the frame directly: wrapping it in unique() would strip the very
# duplicates being asserted against, making the gate unfalsifiable.
stopifnot(!anyDuplicated(events[, c("id", "time", "evid")]))
stopifnot(!anyNA(events$BSA))
```

```{r simulate}
sim <- rxode2::rxSolve(
  mod, events,
  keep = c("arm", "BSA", "REGION_EASTASIA"),
  returnType = "data.frame",
  useLinCmt = FALSE   # else peripheral1 is silently discarded; see above
)
stopifnot(all(sim$Cc[!is.na(sim$Cc)] >= 0))
```

# Replicate Figure 4: the effect of BSA

Figure 4 simulates 200 Western patients at each of the 2.5th, 50th and 97.5th
BSA percentiles on 50 mg twice daily, and its inset tabulates the geometric mean
(and CV%) of steady-state peak and trough concentration for each. Because BSA
acts on V1/F only and not on CL/F, a smaller patient gets a **higher peak and a
lower trough** -- the same AUC, distributed differently across the interval.

```{r figure-4-profile, fig.cap = "Replicates Figure 4 of Zhou 2018: median simulated alisertib concentration for three BSA percentiles, 50 mg b.i.d. for 7 days in a 21-day cycle."}
sim |>
  dplyr::filter(grepl("^West, BSA", arm), !is.na(Cc)) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(Q50 = stats::median(Cc), .groups = "drop") |>
  ggplot(aes(time / 24, Q50, colour = arm)) +
  geom_line(linewidth = 0.4) +
  labs(x = "Time since start of cycle (days)",
       y = "Alisertib concentration (nmol/L)", colour = NULL,
       caption = "Replicates Figure 4 of Zhou 2018.") +
  theme(legend.position = "bottom")
```

These eight geometric means are a genuine answer key: they are the paper's own
simulation output from the same parameters, so agreement tests the whole chain
-- structure, transit chain, BSA exponent, the molar-mass conversion and the
IIV block -- at once. They are reproduced two ways below.

The **typical-value** prediction (all etas zero) is deterministic: it is
identical on every machine, so it can carry a tight gate. For a model of this
shape the geometric mean over a lognormal population sits close to the typical
value, so it is the like-for-like comparator up to a small convexity offset.

```{r figure-4-typical}
bsa_levels <- c(1.44, 1.84, 2.43)

typ_ss <- do.call(rbind, lapply(bsa_levels, function(b) {
  ev <- rxode2::et(amt = dose_west, cmt = "transit1", ii = 12, addl = 13) |>
    rxode2::et(seq(144, 156, by = 0.01), cmt = "central")
  d <- as.data.frame(ev)
  d$BSA <- b
  d$REGION_EASTASIA <- 0
  s <- rxode2::rxSolve(mod_typ, d, returnType = "data.frame",
                       useLinCmt = FALSE)
  s <- s[!is.na(s$Cc) & s$time >= 144, ]
  data.frame(BSA = b, Cmax_typ = max(s$Cc), Cmin_typ = min(s$Cc),
             ratio_typ = max(s$Cc) / min(s$Cc))
}))
# Join on a rounded character key: BSA travels through rxSolve as a double and
# an exact floating-point join key is a silent-empty-join risk.
typ_ss$bsa_key <- sprintf("%.2f", typ_ss$BSA)
```

```{r figure-4-cohort}
gm <- function(x) exp(mean(log(x)))
cv_pct <- function(x) 100 * stats::sd(log(x))   # the 100*omega convention

coh_ss <- sim |>
  dplyr::filter(grepl("^West, BSA", arm), !is.na(Cc),
                time >= 144, time <= 156) |>
  dplyr::mutate(bsa_key = sprintf("%.2f", BSA)) |>
  dplyr::group_by(bsa_key, id) |>
  dplyr::summarise(cmax = max(Cc), cmin = min(Cc), .groups = "drop") |>
  dplyr::group_by(bsa_key) |>
  dplyr::summarise(
    Cmax_gm = gm(cmax), Cmax_cv = cv_pct(cmax),
    Cmin_gm = gm(cmin), Cmin_cv = cv_pct(cmin), .groups = "drop"
  )

published_fig4 <- tibble::tibble(
  BSA = bsa_levels,
  bsa_key = sprintf("%.2f", bsa_levels),
  Cmax_pub = c(2740, 2550, 2420), Cmax_CV_pub = c(51, 52, 45),
  Cmin_pub = c(1270, 1340, 1370), Cmin_CV_pub = c(77, 70, 59),
  ratio_pub = c(2.08, 1.88, 1.68)
)

fig4 <- published_fig4 |>
  dplyr::left_join(tibble::as_tibble(typ_ss) |> dplyr::select(-BSA),
                   by = "bsa_key") |>
  dplyr::left_join(coh_ss, by = "bsa_key") |>
  dplyr::mutate(
    Cmax_typ_diff = 100 * (Cmax_typ - Cmax_pub) / Cmax_pub,
    Cmin_typ_diff = 100 * (Cmin_typ - Cmin_pub) / Cmin_pub,
    ratio_diff = 100 * (ratio_typ - ratio_pub) / ratio_pub
  )
# A join that silently matched nothing would leave NAs and make every gate
# below vacuously true.
stopifnot(nrow(fig4) == 3L, !anyNA(fig4$Cmax_typ), !anyNA(fig4$Cmax_gm))

fig4 |>
  dplyr::transmute(
    `BSA (m^2)` = BSA,
    `Cmax,ss pub` = Cmax_pub,
    `Cmax,ss typical` = round(Cmax_typ),
    `Cmax diff (%)` = Cmax_typ_diff,
    `Cmax,ss cohort GM` = round(Cmax_gm),
    `Cmin,ss pub` = Cmin_pub,
    `Cmin,ss typical` = round(Cmin_typ),
    `Cmin diff (%)` = Cmin_typ_diff,
    `Cmin,ss cohort GM` = round(Cmin_gm)
  ) |>
  knitr::kable(digits = 1, caption =
    "Replicates the Figure 4 inset of Zhou 2018 (nmol/L).")
```

```{r figure-4-gate}
# DETERMINISTIC -- typical-value solve, no cohort, identical on every machine.
# Realised 5.8 / 6.0 / 4.4% on Cmax and 4.5 / 5.0 / 8.8% on Cmin; the
# systematic positive offset is the convexity gap between a typical value and
# a geometric mean plus the paper's own 200-subject Monte-Carlo noise. 15%
# still goes red on a mis-transcribed volume, dose or molar mass, which move
# these by tens of percent.
stopifnot(max(abs(c(fig4$Cmax_typ_diff, fig4$Cmin_typ_diff))) < 15)

# The COHORT geometric means are a second, noisier view of the same answer
# key. Measured max |diff| across solver thread counts: 6.1 / 6.7 / 8.1 /
# 16.7% at 16 / 4 / 2 / 1 threads -- rxode2 partitions its RNG streams per
# thread, so CI draws a different cohort and no seed can make them agree.
# The bound is set outside that whole range; do not tighten it back.
coh_diff <- c(100 * (fig4$Cmax_gm - fig4$Cmax_pub) / fig4$Cmax_pub,
              100 * (fig4$Cmin_gm - fig4$Cmin_pub) / fig4$Cmin_pub)
stopifnot(max(abs(coh_diff)) < 25)
```

The paper also prints a peak-to-trough ratio per BSA group, and these are
population (typical-value) quantities in the source too, so they compare
directly and deterministically.

```{r figure-4-ratio}
fig4 |>
  dplyr::transmute(`BSA (m^2)` = BSA, `peak:trough published` = ratio_pub,
                   `peak:trough typical` = ratio_typ,
                   `Diff (%)` = ratio_diff) |>
  knitr::kable(digits = 3, caption =
    "Population peak-to-trough ratio over the steady-state interdose interval.")

# Deterministic; realised 5.0 / 2.3 / 0.9%.
stopifnot(max(abs(fig4$ratio_diff)) < 10)

# The qualitative claim: peak falls and trough rises with BSA, so fluctuation
# narrows. Deterministic, so strict ordering is safe here.
stopifnot(fig4$Cmax_typ[3] < fig4$Cmax_typ[1],
          fig4$Cmin_typ[3] > fig4$Cmin_typ[1],
          fig4$ratio_typ[3] < fig4$ratio_typ[1])
```

The paper's own summary of this figure is that the BSA effect is *not*
clinically meaningful -- "<15% differences in steady-state peak and trough
alisertib concentrations" between the 2.5th and 97.5th percentile groups. That
is an absolute bound the source states, and on the typical-value predictions it
is a deterministic check.

```{r figure-4-claim}
spread <- function(x) 100 * (max(x) - min(x)) / stats::median(x)
c(`Cmax,ss spread across BSA groups (%)` = spread(fig4$Cmax_typ),
  `Cmin,ss spread across BSA groups (%)` = spread(fig4$Cmin_typ))
# Realised 13.7% and 11.7%; the paper's own inset values give 12.5% and 7.5%.
stopifnot(spread(fig4$Cmax_typ) < 15, spread(fig4$Cmin_typ) < 15)
```

The CV% column of the Figure 4 inset does **not** reproduce as cleanly as the
geometric means, and that is recorded rather than smoothed over. Zhou 2018
reports Cmax CV 51 / 52 / 45% and Cmin CV 77 / 70 / 59% -- both falling with
BSA -- whereas the packaged model gives a Cmax CV that rises slightly with BSA
and a Cmin CV that is nearly flat. The paper does not state how it computed
those CVs, and the between-subject variance in the model is BSA-independent by
construction (BSA enters only the typical value of V1/F), so a systematic trend
with BSA is not something the model can produce. Nothing is gated on the CVs.

```{r figure-4-cv, echo = FALSE}
fig4 |>
  dplyr::transmute(`BSA (m^2)` = BSA,
                   `Cmax CV% pub` = Cmax_CV_pub, `Cmax CV% sim` = Cmax_cv,
                   `Cmin CV% pub` = Cmin_CV_pub, `Cmin CV% sim` = Cmin_cv) |>
  knitr::kable(digits = 0, caption =
    "Figure 4 inset CV% -- shown for completeness; not gated. See the note above.")
```

# Replicate Figure 6: exposure-matched regional dosing

This is the paper's central claim: 30 mg twice daily in East Asia produces the
same exposure as 50 mg twice daily in the West. It is also the check that
validates the molar-mass conversion, because the published exposures are
absolute (uM.h) rather than dose-normalised.

```{r figure-6, fig.cap = "Replicates Figure 6 of Zhou 2018: median (shaded 5th-95th percentile) alisertib concentration for the two exposure-matched regional regimens."}
sim |>
  dplyr::filter(grepl("cohort", arm), !is.na(Cc), time <= 192) |>
  dplyr::group_by(arm, time) |>
  dplyr::summarise(
    Q05 = stats::quantile(Cc, 0.05), Q50 = stats::median(Cc),
    Q95 = stats::quantile(Cc, 0.95), .groups = "drop"
  ) |>
  ggplot(aes(time / 24, Q50, colour = arm, fill = arm)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.4) +
  labs(x = "Time since start of cycle (days)",
       y = "Alisertib concentration (nmol/L)",
       colour = NULL, fill = NULL,
       caption = "Replicates Figure 6 of Zhou 2018.") +
  theme(legend.position = "bottom")
```

# PKNCA validation

Steady-state exposure over the last twice-daily dosing interval of the 7-day
course (144-156 h), computed with PKNCA and compared against the values Zhou
2018 reports for the same two regimens.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(grepl("cohort", arm), !is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | arm + id)

dose_df <- events |>
  dplyr::filter(grepl("cohort", arm), evid == 1) |>
  dplyr::select(id, time, amt, arm)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | arm + id)

intervals <- data.frame(
  start = 144, end = 156,
  cmax = TRUE, cmin = TRUE, tmax = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(
  PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals)
)
```

```{r pknca-summary}
auc_ss <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::group_by(arm) |>
  dplyr::summarise(
    AUC_uMh_sim = gm(PPORRES) / 1000,
    AUC_CV_sim = cv_pct(PPORRES), .groups = "drop"
  )

published_fig6 <- tibble::tibble(
  arm = c("West cohort, 50 mg b.i.d.", "East Asia cohort, 30 mg b.i.d."),
  AUC_uMh_pub = c(24.1, 21.4), AUC_CV_pub = c(53.6, 52.3)
)

auc_cmp <- dplyr::left_join(auc_ss, published_fig6, by = "arm") |>
  dplyr::transmute(
    Regimen = arm,
    `AUC(0-tau),ss published (uM.h)` = AUC_uMh_pub,
    `AUC(0-tau),ss simulated (uM.h)` = AUC_uMh_sim,
    `Diff (%)` = 100 * (AUC_uMh_sim - AUC_uMh_pub) / AUC_uMh_pub,
    `CV% published` = AUC_CV_pub, `CV% simulated` = AUC_CV_sim
  )

knitr::kable(auc_cmp, digits = 2, caption =
  "Steady-state exposure versus Zhou 2018 Results / Abstract (geometric means).")
```

```{r pknca-gate}
# Cohort-derived geometric means at 200/arm against the paper's 1000/arm
# figures. Measured max |diff| across solver thread counts: 3.0 / 3.8 / 5.4 /
# 8.2% at 1 / 2 / 4 / 16 threads. The bound sits outside that whole range; a
# mis-transcribed CL/F, dose or molar mass moves these by tens of percent to
# several-fold, so 20% remains a live gate. Do not tighten it back.
stopifnot(max(abs(auc_cmp$`Diff (%)`)) < 20)
```

The exposure-matching claim itself is better tested deterministically. At steady
state `AUC(0-tau) = dose / (CL/F)` exactly, and both the dose and the regional
multiplier on CL/F are packaged values -- so the comparison is closed-form, has
no cohort in it, and is identical on every machine. This is also the sharpest
check on the molar mass: a wrong value would scale both arms by the same wrong
factor and show up immediately.

```{r exposure-match}
cl_west <- exp(th[["lcl"]])
cl_east <- exp(th[["lcl"]]) * (1 + th[["e_region_eastasia_f"]])

match_tab <- tibble::tibble(
  Regimen = c("West, 50 mg b.i.d.", "East Asia, 30 mg b.i.d."),
  `CL/F (L/h)` = c(cl_west, cl_east),
  `AUC(0-tau),ss = dose/(CL/F) (uM.h)` =
    c(dose_west / cl_west, dose_east / cl_east) / 1000,
  `Published (uM.h)` = c(24.1, 21.4)
) |>
  dplyr::mutate(`Diff (%)` = 100 *
    (`AUC(0-tau),ss = dose/(CL/F) (uM.h)` - `Published (uM.h)`) /
      `Published (uM.h)`)

knitr::kable(match_tab, digits = 3, caption =
  "Closed-form steady-state exposure for the two regional regimens.")

# Deterministic closed form. Realised -2.7% (West) and -0.2% (East Asia).
stopifnot(max(abs(match_tab$`Diff (%)`)) < 10)

# The paper's claim: the two regimens are exposure-matched. Published gap is
# 11.9%; the closed form gives 9.3%. Deterministic, so a tight bound is safe.
gap <- 100 * abs(diff(match_tab$`AUC(0-tau),ss = dose/(CL/F) (uM.h)`)) /
  mean(match_tab$`AUC(0-tau),ss = dose/(CL/F) (uM.h)`)
c(`exposure gap between the two regional regimens (%)` = gap)
stopifnot(gap < 15)
```

## Dose-normalised exposure by region: a discrepancy in the source

Table 6 reports **observed** dose-normalised steady-state AUC of 482 nM.h/mg in
the West and 797 in East Asia -- a 65% increase. The **model** estimates 52%.
The two are not the same number, and the gap is in the source, not in this
transcription: the paper quotes both without reconciling them.

```{r table-6}
tibble::tibble(
  Quantity = c("Model-estimated increase in F, East Asia vs West",
               "Observed dose-normalised AUC increase (Table 6)"),
  Value = c(sprintf("%.1f%%", 100 * (1 / (1 + th[["e_region_eastasia_f"]]) - 1)),
            sprintf("%.1f%%", 100 * (797 / 482 - 1)))
) |>
  knitr::kable(caption = "Zhou 2018's model estimate versus its own observed exposure ratio.")
```

Table 6's values are post-hoc individual exposures from the *updated* final
model on the combined data set, and they carry shrinkage, so they are not
expected to equal the structural estimate exactly. It is recorded here as a
known feature of the source rather than smoothed over. See *Assumptions and
deviations*.

# Exposure-safety relationships (replicates Figure 7)

The three logistic models relate the probability of a mechanism-related
antiproliferative toxicity to the log of the time-averaged daily alisertib AUC.
Zhou 2018 publishes **no coefficient table** for them -- Figure 7 is the only
place the fitted relationships appear -- so both coefficients of each model were
recovered by digitising the published curves.

```{r er-solve}
solve_er <- function(nm, auc_grid) {
  ev <- data.frame(id = seq_along(auc_grid), time = 0, evid = 0L,
                   amt = NA_real_, dv = NA_real_, AUC_ALIS = auc_grid)
  s <- rxode2::rxSolve(ui[[nm]], ev, returnType = "data.frame")
  s[[grep("^prob_", names(s), value = TRUE)]]
}

er_models <- c(
  "Grade >= 3 neutropenia" = "Zhou_2018_alisertib_neutropenia",
  "Grade >= 2 stomatitis" = "Zhou_2018_alisertib_stomatitis",
  "Grade >= 2 diarrhoea" = "Zhou_2018_alisertib_diarrhea"
)

auc_grid <- seq(1, 200, by = 1)
er_curves <- dplyr::bind_rows(lapply(names(er_models), \(lbl)
  data.frame(Endpoint = lbl, AUC = auc_grid,
             p = solve_er(er_models[[lbl]], auc_grid))))
```

```{r figure-7, fig.cap = "Replicates Figure 7 of Zhou 2018: probability of each toxicity versus time-averaged alisertib AUC. Points mark the three exposures for which the paper prints a model-estimated incidence."}
anchors <- tibble::tribble(
  ~Endpoint, ~AUC, ~p_pub,
  "Grade >= 3 neutropenia", 15.63, 0.39,
  "Grade >= 3 neutropenia", 23.72, 0.46,
  "Grade >= 3 neutropenia", 14.23, 0.37,
  "Grade >= 2 stomatitis", 15.63, 0.07,
  "Grade >= 2 stomatitis", 23.72, 0.13,
  "Grade >= 2 stomatitis", 14.23, 0.065,
  "Grade >= 2 diarrhoea", 15.63, 0.13,
  "Grade >= 2 diarrhoea", 23.72, 0.17,
  "Grade >= 2 diarrhoea", 14.23, 0.12
)

ggplot(er_curves, aes(AUC, p, colour = Endpoint)) +
  geom_line() +
  geom_point(data = anchors, aes(AUC, p_pub, colour = Endpoint),
             size = 2, shape = 21, fill = "white") +
  labs(x = "Time-averaged AUC (uM*h/day)",
       y = "Probability of event", colour = NULL,
       caption = "Replicates Figure 7 of Zhou 2018; points are the printed model-estimated incidences.") +
  theme(legend.position = "bottom")
```

## The digitisation self-validates

The coefficients were fitted to the digitised **curve** only. The paper
separately prints, in its Results prose, the model-estimated incidence at three
specific exposures. Those nine numbers were not used in the fit, so reproducing
them is an independent check on the digitisation.

```{r er-anchor-check}
er_check <- anchors |>
  dplyr::rowwise() |>
  dplyr::mutate(p_sim = solve_er(er_models[[Endpoint]], AUC)) |>
  dplyr::ungroup() |>
  dplyr::mutate(`Abs diff` = abs(p_sim - p_pub)) |>
  dplyr::transmute(
    Endpoint,
    `AUC (uM*h/day)` = AUC,
    `Published p` = p_pub,
    `Packaged model p` = p_sim,
    `Abs diff` = `Abs diff`
  )

knitr::kable(er_check, digits = 4, caption =
  "Model-estimated incidences printed in Zhou 2018 Results versus the packaged (figure-derived) models.")

# Deterministic -- no cohort, no randomness, so this is identical on every
# machine and a tight bound is correct. Realised max 0.0049. The published
# values are rounded to 2 significant figures, so 0.005 is the resolution
# limit of the printed numbers; the bound cannot usefully be tighter.
stopifnot(max(er_check$`Abs diff`) < 0.006)
```

Every one of the nine reproduces inside the printed rounding. This is what
licenses the figure-derived coefficients: the curve and the prose agree.

## The regional dosing argument

Reading the three models at the exposures the paper quotes reproduces its safety
argument -- East Asian patients at the Western 50 mg dose would run a higher
toxicity risk, while at the regional 30 mg dose their risk matches Western
patients at 50 mg.

```{r er-argument}
scenarios <- tibble::tibble(
  Scenario = c("West, 50 mg b.i.d.", "East Asia, 50 mg b.i.d.",
               "East Asia, 30 mg b.i.d."),
  AUC = c(15.63, 23.72, 14.23)
)

er_tab <- scenarios |>
  dplyr::mutate(
    `Grade >= 3 neutropenia` = solve_er(er_models[[1]], AUC),
    `Grade >= 2 stomatitis` = solve_er(er_models[[2]], AUC),
    `Grade >= 2 diarrhoea` = solve_er(er_models[[3]], AUC)
  ) |>
  dplyr::rename(`Time-averaged AUC (uM*h/day)` = AUC)

knitr::kable(er_tab, digits = 3, caption =
  "Predicted toxicity probabilities for the three regional dosing scenarios.")

# Deterministic. Each is a strict inequality the paper asserts in prose.
stopifnot(all(er_tab[2, 3:5] > er_tab[1, 3:5]))   # 50 mg: East Asia > West
stopifnot(all(er_tab[3, 3:5] <= er_tab[1, 3:5]))  # 30 mg East Asia <= 50 mg West
```

# Assumptions and deviations

- **Molar mass is not from the paper.** Zhou 2018 works entirely in molar
  concentration units and never states alisertib's molar mass, which is
  nonetheless required to convert a milligram dose into the nmol the model
  expects. `518.92 g/mol` (C27H20ClFN4O4) is taken from an external source
  (PubChem CID 24771867). It is corroborated by the paper's own simulation
  output: under this value the simulated steady-state AUC reproduces the
  published 24.1 and 21.4 uM.h within a few percent, and a wrong molar mass
  would scale both by the same wrong factor. The **model file itself carries no
  molar mass** -- it is unit-agnostic, taking nmol in and returning nmol/L, so
  this assumption lives only in the vignette.

- **The three exposure-safety models carry FIGURE-DERIVED coefficients.**
  Zhou 2018 publishes no coefficient table for its logistic regressions; Figure
  7 is the only place the fitted relationships appear. All six values were
  recovered by digitising the published curves at 400 dpi and regressing
  `logit(p)` on `log(AUC)`. The fits are near-exact (logit-scale residual SD
  0.0024, 0.0122 and 0.0049 over 498, 440 and 479 extracted points), which is
  what confirms rather than assumes the logit-linear-in-log(AUC) form that
  Methods describes. They then reproduce all nine independently-printed
  model-estimated incidences to within the printed rounding, as tabulated
  above. Treat them as good to roughly the second decimal, not as published
  point estimates.

- **The "IIV, ratio" column is read as a log-scale SD, not an exact CV.**
  Table 5 prints `0.518` for CL/F and the Results text reads it back as
  "interpatient coefficient of variation (CV): 51.8%", i.e. CV% = 100 x the
  tabulated ratio -- the usual `CV ~= omega` approximation. Reading the column
  as an exact lognormal CV instead would give variances 11-23% smaller. The
  two readings were tested against the paper's own Figure 4 inset (eight
  geometric means and four CVs from the authors' own simulation) and the
  log-SD reading reproduced them substantially better on every metric; it is
  also what makes the Figure 4 and Figure 6 gates above pass. The alternative
  is documented here because the source's wording is genuinely ambiguous.

- **Table 5, not Supplementary Table S1.** The packaged parameters are the
  final model fitted to the 422-patient analysis set. The supplement's updated
  final model, refitted on the combined 671-patient set, differs on every row
  and is tabulated above for reference only.

- **The RGNF sign is reconstructed, not printed in the main text.** Table 5
  prints the magnitude `0.341`; the sign comes from the Table 5 footnote
  ("multiplied by (1-0.341)"), from Figure 1's `(1 + f)` form, and from
  Supplementary Table S1, which prints the corresponding coefficient as
  `-0.325` with its minus sign intact.

- **Model-versus-observed regional ratio.** Zhou 2018's model estimates 52%
  higher relative bioavailability in East Asia, while its own Table 6 reports a
  65% higher observed dose-normalised AUC. The packaged model necessarily
  reproduces the former. The discrepancy is in the source and is not
  reconciled there; Table 6 derives from post-hoc individual exposures under
  the updated final model on the combined data set, which are shrunk toward the
  population and so are not expected to match the structural estimate exactly.

- **The Figure 4 inset CV% column does not reproduce.** The eight geometric
  means reproduce within 4-9% on typical values, but the published CVs fall
  with BSA (Cmax 51/52/45%, Cmin 77/70/59%) in a way the model cannot produce:
  BSA enters only the typical value of V1/F, so the between-subject variance is
  BSA-independent by construction. The paper does not state how those CVs were
  computed. This is recorded as a known disagreement and is excluded from the
  gates rather than gated with a widened bound.

- **Gate bounds were measured across solver thread counts, not taken from one
  run.** `rxSetSeed()` fixes rxode2's RNG stream per solver thread and not
  across thread counts, so a 2-core CI runner draws a different cohort than a
  16-thread workstation. The cohort-derived comparisons here were rendered at
  1, 2, 4 and 16 threads; the Figure 4 cohort geometric means ranged over
  6.1-16.7% and the PKNCA exposures over 3.0-8.2%, and the bounds sit outside
  those ranges. The load-bearing gates are instead placed on deterministic
  typical-value and closed-form quantities, which are identical on every
  machine and can therefore be asserted tightly.

- **Cohort sizes.** Zhou 2018 simulated 1000 patients per region for Figure 6
  and 200 per BSA group for Figure 4. This vignette uses 200 per arm
  throughout, per the library cap; this only widens Monte-Carlo noise on the
  Figure 6 comparison and does not change the central estimates.

- **BSA distribution for the regional cohorts.** Drawn lognormally from the
  geometric means and log standard deviations the paper states for its own
  Figure 6 simulation (West 1.88 m^2, log SD 0.135; East Asia 1.63 m^2, log SD
  0.0862). The BSA computation formula is not stated in the source.

- **`useLinCmt = FALSE` is mandatory for this model.** The disposition uses
  explicit `k12` / `k21` micro-constants, and `rxSolve()`'s default
  ODE-to-`linCmt()` auto-conversion silently discards `peripheral1` for models
  of this shape. AUC is unchanged by the collapse, so only the terminal
  half-life detects it; the gate above does exactly that.

- **The additive residual error is packaged as `fixed(0)`.** Methods declare a
  combined proportional-plus-additive residual model, but Table 5 reports the
  additive term as 0 with the footnote "automatically set to zero by the SAEM
  estimation algorithm". Keeping it at zero rather than dropping it preserves
  the declared structure; the fitted residual is effectively proportional-only.

- **Screened-but-not-retained covariates.** Body weight, BSA on CL/F, UGT1A1
  *28 and *6 genotype, creatinine clearance, sex, age, race, albumin, ALT, AST,
  bilirubin and formulation were all evaluated (Table 4) and none was retained.
  They are recorded in the model file's `covariatesDataExcluded` metadata so
  the provenance of the covariate screen survives, without declaring covariates
  the model never uses.
