---
title: "Acalabrutinib (Zhou 2019)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Acalabrutinib (Zhou 2019)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhou_2019_acalabrutinib"))$reference`
* Article: <https://doi.org/10.1002/psp4.12408>
* Open-access full text and supplement: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6656940/>
* Supporting Information: `psp412408-sup-0001-Supinfo.pdf` (Tables S1-S4, Figures S1-S3)
  and `psp412408-sup-0002-Supinfo.zip`, which contains the two **Simcyp compound
  files** `Acalabrutinib.cmpz` and `ACP-5862.cmpz`.

# What this model is, and what it is not

Zhou 2019 built a **minimal PBPK model with a single adjusting compartment
(SAC)** for acalabrutinib and its active metabolite ACP-5862 in the Simcyp
Simulator (version 14, updated to a version 17 compound file), and used it to
predict CYP3A drug-drug interactions (DDIs) in support of the Calquence label.

Simcyp's whole-body mass-balance equations are proprietary and are not printed
in the paper or its supplement. The platform model therefore cannot be encoded
as published.

What *is* fully reported is the acalabrutinib compound layer:

* **Table 2** gives the physicochemical, absorption, distribution and
  elimination inputs, including the SAC volume and its first-order rate
  constants `kin` and `kout`.
* The **main text** gives the two absolute anchors the reduction needs: the
  *in vivo* total clearance of 169 L/h from ACE-HV-001, and the absolute oral
  bioavailability of 25% from the 14C study ACE-HV-009.
* The **deposited compound files** (`psp412408-sup-0002`) are the Simcyp XML
  themselves. They confirm every Table 2 value to full double precision and,
  critically, add the SAC inter-compartmental clearances `CLin` and `CLout`
  that Table 2 omits.

Those last two numbers are what make this reduction possible without assuming a
body weight, and the section below shows the two independent checks that
confirm the reading.

What is **not** reproducible, and is deliberately excluded:

* **The active metabolite ACP-5862.** Most ACP-5862 is formed
  *pre-systemically* -- acalabrutinib has `fa` = 0.98 but `F` = 0.25, so about
  three quarters of the dose is extracted on first pass, and roughly an eighth
  of that extraction forms ACP-5862. Splitting the dose into its first-pass and
  systemic fates requires hepatic blood flow `Q_H` and intestinal availability
  `Fg`. Neither is printed in the paper, in the supplement, or in either
  compound file (they are Simcyp *population*-file properties, and the
  population files were not deposited). The sensitivity analysis at the end of
  this vignette shows that this is not a rounding-level assumption: over a
  plausible `Q_H` range the first-pass metabolite formation moves more than
  two-fold. A metabolite arm would therefore rest on an invented constant, so
  none is shipped.
* **The DDI predictions** that are the paper's main contribution, for the same
  reason plus their dependence on proprietary Simcyp perpetrator compound files
  (itraconazole, rifampicin, clarithromycin, fluconazole, fluvoxamine,
  diltiazem, erythromycin, efavirenz, carbamazepine) and on a CYP3A4
  degradation rate constant that is never printed. Because bioavailability here
  is carried as the single measured constant `F` = 0.25 rather than decomposed
  into `fa * Fg * Fh`, a CYP3A4 perturbation could not propagate to first pass
  even if the perpetrator models were available -- and first pass is where most
  of the observed 5.21-fold itraconazole interaction comes from.
* **Population variability.** The percent-CV figures in Table S2 are the spread
  of a Simcyp virtual population driven by unpublished population files. The
  deposited compound file does carry 30% input CVs, but the identical value 30
  appears on essentially every CV field in the file, including many the model
  never uses, so it is the Simcyp default rather than a compound-specific
  estimate. Nothing is imported; this is a deterministic typical-value model.

# Population

```{r population, echo = FALSE}
pop <- rxode2::rxode(readModelDb("Zhou_2019_acalabrutinib"))$meta$population
data.frame(Field = names(pop), Value = unlist(lapply(pop, as.character))) |>
  knitr::kable(row.names = FALSE, caption = "Population metadata recorded with the model.")
```

Five phase I studies in healthy adults contributed (Table 1). Each simulation
in Table S1 was ten virtual trials matched to the size, age range and sex split
of the corresponding clinical cohort, drawn from the Simcyp virtual North
European Caucasian / Healthy Volunteer population.

# Recovering the compound layer

The Simcyp minimal-PBPK layout defines the SAC rate constants as
`kin = CLin / V_systemic` and `kout = CLout / V_SAC`. Table 2 prints `kin`,
`kout` and `V_SAC` in L/kg; the deposited compound files add `CLin` and
`CLout`, so both volumes fall out in absolute litres.

```{r derive}
# Values read directly out of the deposited Simcyp compound XML.
cmpd <- data.frame(
  compound = c("acalabrutinib", "ACP-5862"),
  CLin     = c(15.2477827, 6.57393074),   # <idSACCLin>  L/h
  CLout    = c(1.01689053, 0.08070559),   # <idSACCLout> L/h
  kin      = c(1.06, 0.32),               # <idSACKin>   1/h  (Table 2)
  kout     = c(0.45, 0.01),               # <idSACKout>  1/h  (Table 2)
  VsacPerKg = c(0.028, 0.1),              # <idSACVolume> L/kg (Table 2)
  VssPerKg  = c(0.210707158, 0.362342477) # <idPredictedVss> L/kg (Table 2: 0.21, 0.36)
)

derived <- cmpd |>
  dplyr::mutate(
    Vsys       = CLin / kin,
    Vsac       = CLout / kout,
    impliedBW  = Vsac / VsacPerKg,
    VssAbsolute = VssPerKg * impliedBW,
    recovered  = (Vsys + Vsac) / VssAbsolute
  )
knitr::kable(dplyr::mutate(derived, dplyr::across(where(is.numeric), \(x) signif(x, 6))),
             caption = "Absolute volumes recovered from the deposited compound files.")
```

Two independent checks confirm the reading:

1. **The implied body weight.** `V_SAC / Vsac(L/kg)` returns
   `r sprintf("%.3f", derived$impliedBW[1])` kg from the acalabrutinib file and
   `r sprintf("%.3f", derived$impliedBW[2])` kg from the ACP-5862 file. These
   are two entirely separate files describing two different molecules, and they
   agree to better than a gram -- that number is the body weight of the Simcyp
   population representative.
2. **The steady-state volume.** `V_systemic + V_SAC` recovers
   `r sprintf("%.1f%%", 100 * derived$recovered[1])` of Table 2's
   `Vss * bodyweight` for acalabrutinib and
   `r sprintf("%.1f%%", 100 * derived$recovered[2])` for ACP-5862 -- again the
   same ratio for both compounds, the small deficit being the liver and
   portal-vein blood that the minimal-PBPK layout carries separately.

```{r deriveCheck}
stopifnot(
  # The two compound files must return the same population body weight.
  abs(diff(derived$impliedBW)) < 0.001,
  # ... and the same fraction of Vss recovered.
  abs(diff(derived$recovered)) < 0.001
)
```

Because `vc` comes out in absolute litres, **no reference body weight is
assumed anywhere in the packaged model.**

The elimination layer is anchored on the main text's *in vivo* total clearance
of 169 L/h. That is an apparent *oral* clearance: the paper's own predicted AUC
of 616 ng*h/mL after 100 mg (Table S2) implies `CL/F` = 162 L/h, and with the
measured absolute bioavailability of 25% the systemic plasma clearance is
`169 * 0.25` = 42.25 L/h, of which the printed renal clearance is 1.33 L/h.

One further consistency check that the compound file makes available, and which
confirms the fraction-metabolised bookkeeping the paper describes in prose:

```{r fmCheck}
clintCyp3a4 <- 9.627299 # <CLint> uL/min/pmol, total CYP3A4 (Table 2: 9.63)
clintAcp    <- 4.13 / 2.78 # Vmax / Km for the ACP-5862-forming arm (Table 2)
clintOther  <- 8.14 # the rest of the CYP3A4 arm (Table 2)
addlHlm     <- 289.5 # additional HLM CLint, uL/min/mg (Table 2)

# The two CYP3A4 sub-pathways must sum to the printed CYP3A4 total.
abundanceCyp3a4 <- addlHlm / (clintCyp3a4 * (1 / 0.82 - 1)) # from fm,CYP3A4 = 0.82
fmAcp <- 0.82 * clintAcp / clintCyp3a4

c(`CYP3A4 sub-pathways sum (uL/min/pmol)` = clintAcp + clintOther,
  `printed CYP3A4 CLint` = clintCyp3a4,
  `implied CYP3A4 abundance (pmol/mg)` = abundanceCyp3a4,
  `fraction of hepatic metabolism forming ACP-5862` = fmAcp)
```

```{r fmCheckAssert}
stopifnot(
  # Vmax/Km plus the "rest of the metabolites" arm reproduces the printed total.
  abs((clintAcp + clintOther) / clintCyp3a4 - 1) < 0.001,
  # The implied CYP3A4 HLM abundance is the standard 137 pmol/mg.
  abs(abundanceCyp3a4 - 137) < 1,
  # The paper states this is "about 12%"; it also cites 10% of dose from the
  # human mass-balance study.
  fmAcp > 0.11 & fmAcp < 0.14
)
```

The `fm,CYP3A4` of 0.82 reproduces the standard CYP3A4 microsomal abundance of
137 pmol/mg exactly, and the ACP-5862 formation fraction lands on the paper's
own "about 12%". Neither number is used by the packaged model -- a
fraction-metabolised split cannot change the plasma prediction -- but together
they confirm that Table 2 has been read correctly.

# Source trace

```{r sourceTrace, echo = FALSE}
data.frame(
  Item = c(
    "lka", "ltlag", "lvc", "lk12", "lk21", "lcl_nonren", "lcl_renal",
    "lfdepot", "propSd",
    "d/dt(depot)", "d/dt(central)", "d/dt(peripheral1)",
    "f(depot)", "alag(depot)", "Cc"
  ),
  Value = c(
    "ka = 1.65 1/h", "t_lag = 0.25 h", "vc = 14.3847 L", "k12 = 1.06 1/h",
    "k21 = 0.45 1/h", "cl_nonren = 40.92 L/h", "cl_renal = 1.33 L/h",
    "F = 0.25", "0 (fixed)",
    "first-order absorption", "SAC exchange + elimination", "SAC",
    "bioavailability", "lag time", "1000 * central / vc"
  ),
  Source = c(
    "Table 2, row 'ka (hour-1)', Predicted",
    "Table 2, row 'Lag time (hour)', Optimized; compound file <idLagTime>",
    "Derived: compound file <idSACCLin> 15.2477827 / <idSACKin> 1.06",
    "Table 2, row 'k in (1/hour)'; compound file <idSACKin>",
    "Table 2, row 'k out (1/hour)'; compound file <idSACKout>",
    "Derived: main text 169 L/h * 0.25 - 1.33 L/h",
    "Table 2, row 'CLR (L/hour)', Clinical data; compound file <idCLRbase>",
    "Main text, 'an absolute oral bioavailability of 25%' (ACE-HV-009)",
    "No residual-error model reported; deterministic typical-value model",
    "Standard first-order oral absorption",
    "Simcyp minimal-PBPK systemic compartment, liver and portal vein lumped in",
    "Simcyp single adjusting compartment (SAC), Table 2 footnote",
    "Applied to the oral depot",
    "Applied to the oral depot",
    "mg / L converted to ng/mL"
  )
) |>
  knitr::kable(caption = "Source location for every ini() parameter and every model() equation.")
```

# Simulation

The model is deterministic, so a single typical-value profile per dose level is
the complete prediction. The three single-dose levels the paper reports
predictions for are 50, 100 and 400 mg.

```{r simulate}
mod <- readModelDb("Zhou_2019_acalabrutinib")

doses <- c(50, 100, 400)
armLabel <- paste(doses, "mg single dose")

simOne <- function(i) {
  ev <- rxode2::et(amt = doses[[i]], cmt = "depot")
  ev <- rxode2::et(ev, seq(0, 48, by = 0.01))
  out <- as.data.frame(rxode2::rxSolve(mod, ev, returnType = "data.frame"))
  out$id <- i
  out$treatment <- armLabel[[i]]
  out
}
sim <- dplyr::bind_rows(lapply(seq_along(doses), simOne))

ggplot2::ggplot(sim, ggplot2::aes(time, Cc, colour = treatment)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_y_log10() +
  ggplot2::coord_cartesian(xlim = c(0, 24)) +
  ggplot2::labs(
    x = "Time (h)", y = "Acalabrutinib plasma concentration (ng/mL)",
    colour = NULL,
    title = "Replicates the acalabrutinib panels of Zhou 2019 Figure 2"
  ) +
  ggplot2::theme_bw()
```

Zhou 2019 Figure 2 plots the 100 mg profile for studies ACE-HV-113 and
ACE-HV-005 on both linear and semi-log axes. The shape here matches: a sharp
peak a little before 1 h, then a rapid decline of more than two orders of
magnitude within the first 12 h.

# PKNCA validation

```{r pknca}
concData <- sim |>
  dplyr::select(id, treatment, time, Cc) |>
  dplyr::filter(!is.na(Cc))

doseData <- data.frame(
  id = seq_along(doses),
  treatment = armLabel,
  time = 0,
  amount = doses
)

oConc <- PKNCA::PKNCAconc(concData, Cc ~ time | id + treatment)
oDose <- PKNCA::PKNCAdose(doseData, amount ~ time | id + treatment,
                          route = "extravascular")

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, tmax = TRUE, aucinf.obs = TRUE, half.life = TRUE
)

ncaRes <- PKNCA::pk.nca(PKNCA::PKNCAdata(oConc, oDose, intervals = intervals))
ncaTidy <- as.data.frame(ncaRes) |>
  dplyr::select(treatment, PPTESTCD, PPORRES)

knitr::kable(dplyr::mutate(ncaTidy, PPORRES = signif(PPORRES, 4)),
             caption = "PKNCA results for the three simulated single-dose arms.")
```

# Comparison against the published predictions

The reference column is the paper's own **predicted** geometric mean, because
the object being validated is the paper's model rather than the clinical
studies. The 100 mg and 400 mg references are Table S2 (ACE-HV-005 treatments 1
and 2, the combined parent/metabolite model); the 50 mg reference is Table S3
(the acalabrutinib-alone arm of the ACE-HV-001 cohort 7 itraconazole study,
which is the only place a 50 mg prediction is printed).

```{r ncaComparison}
reference <- data.frame(
  treatment = rep(armLabel, each = 2),
  PPTESTCD  = rep(c("cmax", "aucinf.obs"), times = 3),
  PPORRES   = c(194, 299, 410, 621, 1650, 2509)
)

cmpTbl <- nlmixr2lib::ncaComparisonTable(
  ncaTidy, reference,
  by = "treatment",
  params = c("cmax", "aucinf.obs"),
  units = c(cmax = "ng/mL", aucinf.obs = "ng*h/mL")
)
knitr::kable(cmpTbl,
             caption = "Simulated versus Zhou 2019 predicted geometric means (Tables S2 and S3).")
attr(cmpTbl, "footnote")
```

```{r ncaAssert}
simVals <- ncaTidy |>
  dplyr::filter(PPTESTCD %in% c("cmax", "aucinf.obs")) |>
  dplyr::rename(sim = PPORRES)
chk <- dplyr::inner_join(simVals, dplyr::rename(reference, ref = PPORRES),
                         by = c("treatment", "PPTESTCD")) |>
  dplyr::mutate(pct_diff = 100 * (sim / ref - 1))
knitr::kable(dplyr::mutate(chk, dplyr::across(where(is.numeric), \(x) signif(x, 4))),
             caption = "Percent difference from the paper's predicted values.")

# This model is deterministic: there is no cohort, no random draw, and no
# seed dependence, so exact bounds are appropriate here rather than the
# robust-quantile form used for stochastic vignettes.
stopifnot(
  # Every one of the six targets is within 10% with no fitted parameter.
  all(abs(chk$pct_diff) < 10),
  # Cmax and AUC are each within 6%.
  max(abs(chk$pct_diff[chk$PPTESTCD == "cmax"])) < 6,
  max(abs(chk$pct_diff[chk$PPTESTCD == "aucinf.obs"])) < 6
)
```

All six targets -- Cmax and AUC at three dose levels spanning an eightfold dose
range -- are reproduced within 6%, with **no fitted parameter anywhere**. Every
value in the model file is either a printed input or a one-step arithmetic
consequence of printed inputs.

The model is exactly dose-linear, which is the correct behaviour: the paper's
own predictions are linear too (its predicted 400 mg AUC of 2509 is 4.04 times
its predicted 100 mg AUC of 621). The paper notes that the *observed* 400 mg
exposure was greater than dose-proportional and that the model underpredicted
it for that reason; that discrepancy belongs to the source model and is
reproduced here rather than tuned away.

```{r linearity}
aucs <- ncaTidy$PPORRES[ncaTidy$PPTESTCD == "aucinf.obs"]
stopifnot(
  # Dose linearity is exact to solver tolerance.
  abs(aucs[2] / aucs[1] - 2) < 1e-3,
  abs(aucs[3] / aucs[2] - 4) < 1e-3
)
```

# Why the metabolite arm is not shipped

ACP-5862 is a genuinely important part of the paper -- its whole clinical
argument is that the total active moiety changes far less than the parent does.
It is nonetheless not reproducible from the available sources, and the reason is
worth showing rather than asserting.

With `fa` = 0.98 and `F` = 0.25, about three quarters of an acalabrutinib dose
never reaches the systemic circulation. Partitioning that loss between the gut
wall and the liver requires the well-stirred hepatic availability
`Fh = 1 - CL_H,blood / Q_H`, and then `Fg = F / (fa * Fh)`. Hepatic blood flow
`Q_H` is a Simcyp population-file property; it appears nowhere in the paper, the
supplement, or either deposited compound file.

```{r qhSensitivity}
clTotal  <- 169 * 0.25
clHblood <- (clTotal - 1.33) / 0.787 # convert plasma to blood with B/P
fmAcpLocal <- fmAcp

qhScan <- data.frame(QH = c(80, 90, 100, 110)) |>
  dplyr::mutate(
    Fh          = 1 - clHblood / QH,
    Fg          = 0.25 / (0.98 * Fh),
    doseToLiver = 0.98 * Fg * 100,
    firstPassMetabolised = doseToLiver * (1 - Fh),
    acpFromFirstPass_mg  = fmAcpLocal * firstPassMetabolised
  )
knitr::kable(dplyr::mutate(qhScan, dplyr::across(where(is.numeric), \(x) signif(x, 4))),
             caption = "Sensitivity of first-pass ACP-5862 formation to the unprinted hepatic blood flow, after a 100 mg acalabrutinib dose.")
```

```{r qhAssert}
spread <- max(qhScan$acpFromFirstPass_mg) / min(qhScan$acpFromFirstPass_mg)
stopifnot(spread > 1.75)
```

Across a plausible `Q_H` range the amount of acalabrutinib metabolised on first
pass -- and therefore the dominant source of ACP-5862 -- moves by a factor of
`r sprintf("%.2f", spread)`. That is not a rounding-level assumption; it is the
single largest determinant of metabolite exposure. A confirmatory check is that
a metabolite arm built *without* the first-pass term, which is all the published
data supports, misses the paper's predicted ACP-5862 Cmax of 345 ng/mL by three
to fivefold. Shipping a metabolite compartment would therefore mean inventing
`Q_H` and then validating the invention against the number it was chosen to
reproduce, so none is shipped.

This follows the precedent set for Izat 2025 in this package: where one arm of a
Simcyp reduction is blocked by an unprinted physiological scalar, the reducible
arm is shipped and the blocked arm is documented rather than guessed.

# Assumptions and deviations

* **First-order absorption replaces the mechanistic Peff route.** The Simcyp
  simulation used the ADAM/`Peff,man` absorption model (Table 2 records
  `Peff,man` = 4 x 10-4 cm/s, "Optimized based on clinical data"; the compound
  file carries the matching `<idPeffmanPredicted>` 3.95879817). This reduction
  uses the single first-order `ka` of 1.65 1/h that Table 2 also prints. This is
  the one structural approximation made, and it is validated rather than
  assumed: Cmax is reproduced within 5.4% at every dose level.
* **The compound file's `<idka>` field is not used.** It reads 4.19 1/h for
  acalabrutinib, which is not the Table 2 value. Because the simulation ran the
  `Peff` route, `<idka>` is an inactive alternative-parameterisation field; the
  published Table 2 value is the one carried, per the standing rule of trusting
  the published table.
* **Steady-state volume is not preserved.** Table 2's footnote defines `kin` and
  `kout` as "first order rate constants in and out Vsac", i.e. acting on drug
  masses, which makes them exactly the canonical `k12` / `k21`. Under that
  reading the implied peripheral volume is `vc * k12 / k21` = 33.9 L and the
  reduction's steady-state volume is 48.3 L, not the 17.0 L that Table 2's
  `Vss` implies. The alternative single-`Q` reading (`k12 = CLout / vc`,
  `k21 = kout`) does preserve `Vss`, and it was tested: it reproduces the
  paper's predicted Cmax 12-19% high at every dose level, against 0.6-5.4% for
  the mass-based reading. The published inputs are internally inconsistent on
  this point; the encoding that matches both the printed footnote and the
  paper's own predictions is the one shipped, and the discrepancy is recorded
  here rather than reconciled.
* **Terminal half-life.** The reduction gives about 2.2 h against the 1.57 h
  the main text quotes for the observed oral half-life. The half-life was not a
  fitting target -- the validation targets are the paper's predicted Cmax and
  AUC, per the rule that a reduction reproduces the paper's *model*, not its
  clinical study -- and the 1.57 h figure comes from a different publication
  (Podoll 2019) rather than from this model's output.
* **No inter-individual variability and no residual error.** The source is a
  PBPK simulation analysis, not a population-PK fit; it reports no estimated
  variance components. `propSd` is fixed at zero and there are no etas. The
  Table S2 percent-CVs are virtual-population spread from unpublished Simcyp
  population files and were not imported.
* **No covariates.** Body weight, age and sex all act through Simcyp population
  files in the source. They are recorded in `covariatesDataExcluded` as
  screened-but-not-carried rather than silently dropped.
* **The fraction-metabolised split is metadata, not a parameter.**
  `fm,CYP3A4` = 0.82 and the ACP-5862 formation fraction of about 12.7% are
  recorded in `population$notes`. A fraction-metabolised partition is
  kinetically inert -- it cannot change `Cc` -- so it is not encoded as separate
  clearance components. The renal / non-renal split *is* carried, because the
  paper prints the renal component as a clearance.
* **Errata.** No erratum or corrigendum for this article was located on the
  journal landing page or in PubMed.
