---
title: "Ganitumab (Zhu 2013)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ganitumab (Zhu 2013)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Zhu_2013_ganitumab"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.1002/cpdd.48>

Ganitumab (AMG 479) is a fully human IgG1 monoclonal antibody antagonist of the
insulin-like growth factor-1 receptor (IGF1R). Zhu 2013 pooled three trials to
build a two-compartment population PK model and found that **pancreatic cancer
type**, not gemcitabine coadministration, was the dominant determinant of
ganitumab disposition: typical clearance was 1.7-fold higher and typical central
volume 1.3-fold higher in patients with metastatic pancreatic cancer than in
patients with other advanced solid cancers.

# Population

The estimation dataset comprised 99 patients contributing 1227 evaluable serum
ganitumab concentrations from three trials (Zhu 2013 Supplemental Table 1):

| Study | N with PK | Ganitumab dose | Cancer type | Regimen |
|---|---|---|---|---|
| Study 1 (NCT00630552) | 35 | 12 mg/kg | Metastatic pancreatic cancer | Ganitumab + gemcitabine |
| Study 2 (NCT00562380) | 53 | 1, 3, 10, 12 or 20 mg/kg | Advanced solid tumours | Ganitumab alone |
| Study 3 (NCT00974896) | 11 | 6 or 12 mg/kg | Advanced solid tumours | Ganitumab + gemcitabine |

All studies dosed intravenously once every 2 weeks. 62 of 99 patients (62.6%)
had non-pancreatic cancer and 46 of 99 (46.5%) received concomitant gemcitabine.
Two percent of samples were excluded as below the limit of quantification
(LLOQ 20.2 ng/mL) or in error. Fitting used NONMEM VI level 1.1 with
FOCE + INTERACTION.

A fourth trial (Study 4, NCT00626106; 104 patients with advanced breast cancer
receiving 12 mg/kg plus hormonal therapy, 802 concentrations) was held out and
used only for external qualification by visual predictive check; it is not part
of the estimation dataset.

**Zhu 2013 publishes no baseline-demographics table.** The Results state only
that "individual demographic data and disease characteristics for each study
were normally distributed and were similar among studies". Age, sex, race, and
the medians and ranges of body weight, albumin and creatinine are therefore
unavailable. This drives most of the Errata below, and it is why the virtual
cohorts here vary random effects rather than covariates.

The same information is available programmatically via
`readModelDb("Zhu_2013_ganitumab")()$population`.

# Source trace

Per-parameter origin is recorded as an in-file comment beside each `ini()` entry
in `inst/modeldb/specificDrugs/Zhu_2013_ganitumab.R`. Collected here for review:

| Equation / parameter | Value | Source location |
|---|---|---|
| Two-compartment linear structure, CL / Q / Vc / Vp | n/a | Methods, "Base Model Development"; confirmed in Results, "Base PK Model Generation" (dMOF = -1255.55 vs 1-compartment) |
| IIV form `theta_i = theta_typ * exp(eta_i)`, diagonal OMEGA | n/a | Methods equation (1) |
| Residual error, proportional only | n/a | Methods equation (2) allows add + prop; Results, "Base PK Model Generation" states the final structure used "a proportional error model"; Table 1 reports no additive term |
| Continuous covariate form `(Cov/Cov_ref)^theta_eff` | n/a | Methods equation (3) |
| `lcl` = log(0.679) | 0.679 L/day | Table 1, "Typical CL for non-pancreatic" (RSE 3.47%) |
| `lvc` = log(3.85) | 3.85 L | Table 1, "Typical Vc for non-pancreatic cancer" (RSE 2.91%) |
| `lq` = log(0.626) | 0.626 L/day | Table 1, "Q" (RSE 9.73%) |
| `lvp` = log(3.41) | 3.41 L | Table 1, "Vp" (RSE 8.39%) |
| `e_wt_cl` | 0.984 | Table 1, "WT on CL" (RSE 12.8%) |
| `e_alb_cl` | -0.859 | Table 1, "ALB on CL" (RSE 22.2%); sign from Results and Discussion prose, see Errata |
| `e_creat_cl` | -0.394 | Table 1, "CR on CL" (RSE 24.7%); sign from Results and Discussion prose, see Errata |
| `e_wt_vc` | 0.559 | Table 1, "WT on Vc" (RSE 21.6%) |
| `e_panc_cl` | log(1.154 / 0.679) | Table 1, typical CL pancreatic 1.154 (RSE 5.84%) vs non-pancreatic 0.679 L/day |
| `e_panc_vc` | log(5.13 / 3.85) | Table 1, typical Vc pancreatic 5.13 (RSE 4.21%) vs non-pancreatic 3.85 L |
| `etalcl` | 0.262^2 | Table 1, "IIV CL 26.2%" (RSE 18.8%) |
| `etalvc` | 0.217^2 | Table 1, "IIV Vc 21.7%" (RSE 17.7%) |
| `etalq` | 0.437^2 | Table 1, "IIV Q 43.7%" (RSE 38.7%) |
| `etalvp` | 0.602^2 | Table 1, "IIV Vp 60.2%" (RSE 25.7%) |
| `propSd` | 0.164 | Table 1, "Proportional error 16.4%" (RSE 9.26%) |
| WT reference 74.7 kg | derived | Recovered by inverting the four published steady-state AUC values in Results; see "Recovering the reference body weight" below |
| ALB reference 40 g/L, CREAT reference 1 mg/dL | assumed | Not reported anywhere in Zhu 2013 or its supplement; see Errata |

```{r ini-table, echo = FALSE}
ui$iniDf |>
  dplyr::filter(!is.na(.data$ntheta)) |>
  dplyr::transmute(
    Parameter = .data$name,
    Estimate  = signif(.data$est, 6),
    Label     = .data$label
  ) |>
  knitr::kable(caption = "Fixed-effect parameters as packaged.")
```

# Reference covariate values

The model is centred on a **non-pancreatic** patient at the cohort median
covariates, for whom every covariate term equals 1 and CL / Vc take their
published typical values.

```{r refs}
WT_REF    <- 74.7  # kg,    derived below
ALB_REF   <- 40    # g/L,   assumed (not published)
CREAT_REF <- 1     # mg/dL, assumed (not published)
```

## Recovering the reference body weight

Zhu 2013 equation (3) defines the reference as "the median of the covariate
among patients", but the paper tabulates no medians. The parenthetical
"(e.g., 70 kg for body weight)" in that same sentence illustrates *what a
reference value is*; it is not a statement of the value used.

The reference weight is nevertheless identifiable from the paper's own published
exposures. At the median covariates every covariate term is 1, so
`AUCss = Dose / CL_typical`, and with a mg/kg dose that inverts to
`WT_ref = AUCss * CL_typical / (mg/kg)`. Zhu 2013 Results reports four
steady-state AUC values, which must all give the same answer:

```{r wt-derivation}
auc_published <- tibble::tribble(
  ~arm,                      ~mgkg, ~auc_mg_day_mL, ~cl_typical,
  "Non-pancreatic 12 mg/kg",    12,           1.32,       0.679,
  "Non-pancreatic 20 mg/kg",    20,           2.20,       0.679,
  "Pancreatic 12 mg/kg",        12,           0.78,       1.154,
  "Pancreatic 20 mg/kg",        20,           1.29,       1.154
) |>
  mutate(wt_implied = auc_mg_day_mL * 1000 * cl_typical / mgkg)

auc_published |>
  dplyr::rename(
    "Arm"                        = arm,
    "Dose (mg/kg)"               = mgkg,
    "Published AUCss (mg*day/mL)" = auc_mg_day_mL,
    "Typical CL (L/day)"         = cl_typical,
    "Implied WT reference (kg)"  = wt_implied
  ) |>
  knitr::kable(digits = 2, caption = "Inverting the published steady-state AUC values for the reference weight.")
```

All four independently imply the same weight to three significant figures. The
competing candidate, 70 kg, reproduces none of the four published AUC values:

```{r wt-falsify}
candidate_check <- tidyr::expand_grid(
  wt_ref = c(70, 74.5, 74.7, 75.0),
  auc_published |> select(arm, mgkg, auc_mg_day_mL, cl_typical)
) |>
  mutate(auc_predicted = round(mgkg * wt_ref / cl_typical / 1000, 2)) |>
  group_by(wt_ref) |>
  summarise(
    n_matching = sum(auc_predicted == auc_mg_day_mL),
    predicted  = paste(sprintf("%.2f", auc_predicted), collapse = ", "),
    .groups    = "drop"
  )

candidate_check |>
  dplyr::rename(
    "Candidate WT reference (kg)"          = wt_ref,
    "Published AUCss reproduced (out of 4)" = n_matching,
    "Predicted AUCss (mg*day/mL)"          = predicted
  ) |>
  knitr::kable(caption = "Only 74.7 kg reproduces all four published steady-state AUC values.")

stopifnot(
  # Exactly one candidate reproduces all four; 70 kg reproduces none.
  candidate_check$n_matching[candidate_check$wt_ref == WT_REF] == 4L,
  candidate_check$n_matching[candidate_check$wt_ref == 70]     == 0L,
  sum(candidate_check$n_matching == 4L) == 1L,
  # And the packaged constant agrees with the mean of the four inversions.
  abs(mean(auc_published$wt_implied) - WT_REF) < 0.1
)
```

Albumin and creatinine admit no equivalent inversion: no published quantity
depends on them separately, so their reference values are assumed. See Errata.

# Virtual cohort

Original observed data are not publicly available, and Zhu 2013 publishes no
demographic distributions, so a covariate-varying virtual population would be
fabrication rather than replication. The cohorts below therefore hold covariates
at the reference values and vary only the model's own random effects, which are
published. Covariate *effects* are exercised separately, deterministically, in
the sensitivity section.

```{r cohort-helpers}
INFUSION_DUR <- 1 / 24  # days; see Errata -- Zhu 2013 does not state the duration
II           <- 14      # days between doses (Q2W, Supplemental Table 1 footnote a)

arms <- tibble::tribble(
  ~arm,                      ~TUMTP_PANC, ~mgkg,
  "Non-pancreatic 12 mg/kg",          0L,    12,
  "Non-pancreatic 20 mg/kg",          0L,    20,
  "Pancreatic 12 mg/kg",              1L,    12,
  "Pancreatic 20 mg/kg",              1L,    20
)

# Build one arm as a self-contained event table. `id_offset` keeps subject IDs
# disjoint across arms; duplicate IDs are silently merged by rxSolve.
make_arm <- function(arm_row, n, n_doses, obs_times, id_offset = 0L) {
  ids <- id_offset + seq_len(n)
  amt <- arm_row$mgkg * WT_REF
  doses <- tidyr::expand_grid(id = ids, time = II * (seq_len(n_doses) - 1L)) |>
    mutate(amt = amt, evid = 1L, cmt = "central", dur = INFUSION_DUR)
  obs <- tidyr::expand_grid(id = ids, time = obs_times) |>
    mutate(amt = NA_real_, evid = 0L, cmt = "central", dur = NA_real_)
  bind_rows(doses, obs) |>
    mutate(
      WT = WT_REF, ALB = ALB_REF, CREAT = CREAT_REF,
      TUMTP_PANC = arm_row$TUMTP_PANC,
      arm = arm_row$arm, mgkg = arm_row$mgkg
    ) |>
    arrange(id, time, desc(evid))
}
```

# Simulation

## Typical-value steady state

Twenty Q2W doses take every arm well past steady state (Zhu 2013 predicts steady
state in 7 weeks for non-pancreatic and 5 weeks for pancreatic cancer). The final
dosing interval is sampled finely so trapezoidal AUC resolves the post-infusion
peak.

```{r sim-typical}
n_doses  <- 20L
t_last   <- II * (n_doses - 1L)
grid_tau <- sort(unique(c(
  seq(0, 0.5, by = 0.005), seq(0.5, 2, by = 0.05), seq(2, II, by = 0.2), II
)))

events_ss <- bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  make_arm(arms[i, ], n = 1L, n_doses = n_doses,
           obs_times = t_last + grid_tau, id_offset = i - 1L)
}))
stopifnot(!anyDuplicated(unique(events_ss[, c("id", "time", "evid")])))

mod         <- readModelDb("Zhu_2013_ganitumab")
mod_typical <- rxode2::zeroRe(mod)
sim_ss <- rxode2::rxSolve(mod_typical, events = events_ss,
                          keep = c("arm", "mgkg", "TUMTP_PANC")) |>
  as.data.frame()
```

## Typical disposition parameters

```{r typical-params}
typ <- sim_ss |>
  group_by(arm) |>
  summarise(cl = first(cl), vc = first(vc), q = first(q), vp = first(vp),
            .groups = "drop") |>
  mutate(vss = vc + vp)

typ |>
  dplyr::rename(
    "Arm"           = arm,
    "CL (L/day)"    = cl,
    "Vc (L)"        = vc,
    "Q (L/day)"     = q,
    "Vp (L)"        = vp,
    "Vc + Vp (L)"   = vss
  ) |>
  knitr::kable(digits = 3, caption = "Typical disposition parameters at the reference covariates.")

np <- typ |> filter(arm == "Non-pancreatic 12 mg/kg")
pc <- typ |> filter(arm == "Pancreatic 12 mg/kg")

stopifnot(
  # Table 1 typical values, reproduced exactly by construction.
  abs(np$cl - 0.679) < 1e-9, abs(np$vc - 3.85) < 1e-9,
  abs(pc$cl - 1.154) < 1e-9, abs(pc$vc - 5.13) < 1e-9,
  abs(np$q  - 0.626) < 1e-9, abs(np$vp - 3.41) < 1e-9,
  # Discussion: "the typical ganitumab Vc + Vp in patients with non-pancreatic
  # cancer was 7.3 L", and "higher Vc + Vp was found in patients with
  # pancreatic cancer (8.5 L)". These are INDEPENDENT of the typical-value
  # round trip above -- they test that Vp is not double-counted per arm.
  round(np$vc + np$vp, 1) == 7.3,
  round(pc$vc + pc$vp, 1) == 8.5,
  # Abstract and Results: CL 1.7-fold and Vc 1.3-fold higher in pancreatic.
  round(pc$cl / np$cl, 1) == 1.7,
  round(pc$vc / np$vc, 1) == 1.3
)
```

## Terminal half-life

Zhu 2013 reports a "calculated median elimination half-life" of approximately
10 days in non-pancreatic and approximately 7 days in pancreatic cancer. The fit
below regresses `log(Cc)` well after a single dose, where the distribution phase
has decayed, and is cross-checked against the closed-form two-compartment
`beta` eigenvalue.

```{r half-life}
events_wash <- bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  make_arm(arms[i, ], n = 1L, n_doses = 1L,
           obs_times = seq(60, 140, by = 1), id_offset = i - 1L)
}))
sim_wash <- rxode2::rxSolve(mod_typical, events = events_wash, keep = "arm") |>
  as.data.frame()

thalf <- sim_wash |>
  group_by(arm) |>
  summarise(
    t_half_fitted = log(2) / -stats::coef(stats::lm(log(Cc) ~ time))[["time"]],
    cl = first(cl), vc = first(vc), q = first(q), vp = first(vp),
    .groups = "drop"
  ) |>
  mutate(
    k10 = cl / vc, k12 = q / vc, k21 = q / vp,
    s   = k10 + k12 + k21,
    beta = (s - sqrt(s^2 - 4 * k10 * k21)) / 2,
    t_half_closed_form = log(2) / beta,
    published = ifelse(grepl("^Pancreatic", arm), 7, 10)
  )

thalf |>
  select(arm, t_half_fitted, t_half_closed_form, published) |>
  dplyr::rename(
    "Arm"                          = arm,
    "Fitted terminal t1/2 (days)"  = t_half_fitted,
    "Closed-form t1/2 (days)"      = t_half_closed_form,
    "Zhu 2013 reported (days)"     = published
  ) |>
  knitr::kable(digits = 2, caption = "Terminal half-life vs the reported approximate values.")

stopifnot(
  # The washout fit must recover the analytic eigenvalue, not merely be close
  # to the paper's rounded figure.
  all(abs(thalf$t_half_fitted - thalf$t_half_closed_form) /
        thalf$t_half_closed_form < 0.01),
  # Zhu 2013 reports these only to the nearest day ("approximately"), so the
  # test is that each rounds to the published integer.
  all(round(thalf$t_half_closed_form) == thalf$published)
)
```

# PKNCA validation

NCA is run over the final dosing interval, which contains exactly one dose and
one full Q2W interval, giving AUCtau at steady state.

```{r pknca}
conc_ss <- sim_ss |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, arm)

dose_ss <- events_ss |>
  dplyr::filter(evid == 1L, time == t_last) |>
  dplyr::select(id, time, amt, arm)

conc_obj <- PKNCA::PKNCAconc(conc_ss, Cc ~ time | arm + id)
dose_obj <- PKNCA::PKNCAdose(dose_ss, amt ~ time | arm + id)

intervals <- data.frame(
  start = t_last, end = t_last + II,
  cmax = TRUE, tmax = TRUE, cmin = TRUE, auclast = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
```

## Comparison against published NCA

Zhu 2013 Results reports steady-state AUC in mg*day/mL; the model works in
ug/mL and days, so the published values are multiplied by 1000 to give
ug*day/mL. Note that `AUCtau,ss = Dose / CL` exactly, so this comparison is
independent of the assumed infusion duration.

```{r nca-compare}
published_nca <- auc_published |>
  transmute(arm, auclast = auc_mg_day_mL * 1000)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = nca_res,
  reference     = published_nca,
  by            = "arm",
  params        = "auclast",
  units         = c(auclast = "ug*day/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated steady-state AUCtau vs Zhu 2013 Results. * differs from reference by >20%.",
  align   = c("l", "l", "r", "r", "r")
)

nca_wide <- as.data.frame(nca_res) |>
  select(arm, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  left_join(published_nca, by = "arm", suffix = c("", "_published")) |>
  left_join(typ |> select(arm, cl), by = "arm") |>
  left_join(arms |> select(arm, mgkg), by = "arm") |>
  mutate(
    dose_mg        = mgkg * WT_REF,
    auc_dose_by_cl = dose_mg / cl,
    pct_vs_published = 100 * (auclast - auclast_published) / auclast_published,
    pct_vs_identity  = 100 * (auclast - auc_dose_by_cl) / auc_dose_by_cl
  )

stopifnot(
  # (a) Numerical: trapezoidal AUCtau must reproduce the closed form Dose/CL.
  #     Both sides use the same drawn parameters, so this is pure quadrature
  #     error and a tight bound is the right test.
  all(abs(nca_wide$pct_vs_identity) < 0.1),
  # (b) Structural: AUCtau must reproduce all four published values. The
  #     residual is the paper's own 2-decimal rounding (<= 0.5%).
  all(abs(nca_wide$pct_vs_published) < 1),
  # Guard: the comparison actually had all four arms to test.
  nrow(nca_wide) == 4L, !anyNA(nca_wide$pct_vs_published)
)
```

The pancreatic-versus-non-pancreatic exposure gap that motivates the paper's
dosing conclusion falls straight out:

```{r exposure-gap}
gap <- nca_wide |>
  select(arm, mgkg, auclast) |>
  mutate(group = ifelse(grepl("^Pancreatic", arm), "pancreatic", "non_pancreatic")) |>
  select(mgkg, group, auclast) |>
  tidyr::pivot_wider(names_from = group, values_from = auclast) |>
  mutate(ratio = non_pancreatic / pancreatic)

gap |>
  dplyr::rename(
    "Dose (mg/kg)"                    = mgkg,
    "Non-pancreatic AUCss (ug*day/mL)" = non_pancreatic,
    "Pancreatic AUCss (ug*day/mL)"     = pancreatic,
    "Exposure ratio"                   = ratio
  ) |>
  knitr::kable(digits = c(0, 0, 0, 2), caption = "Steady-state exposure gap by cancer type.")

stopifnot(
  # The exposure ratio is the reciprocal of the 1.7-fold CL difference.
  all(abs(gap$ratio - 1.154 / 0.679) < 0.02)
)
```

# Replicate published figures

## Figure 1 analogue -- projected profiles over the first four doses

Zhu 2013 Figure 1 overlays observed breast-cancer concentrations (Study 4,
12 mg/kg Q2W) on the median and 90% prediction interval projected from the final
model for the first four doses. Breast cancer is a non-pancreatic type, so
`TUMTP_PANC = 0`. The observed data are not available, so only the projection is
reproduced; the pancreatic arm is added for contrast.

```{r figure-1, fig.cap = "Replicates the projection in Figure 1 of Zhu 2013 (median and 90% prediction interval, 12 mg/kg Q2W, first four doses), with the pancreatic-cancer arm added for contrast."}
# rxode2 draws etas from its own RNG, which base `set.seed()` does not reach;
# `rxSetSeed()` is what makes the cohort below reproducible.
rxode2::rxSetSeed(20130926)
set.seed(20130926)
n_per_arm <- 200L
grid_vpc  <- sort(unique(c(
  seq(0, 56, by = 0.5),
  as.vector(outer(II * (0:3), c(0.04, 0.1, 0.25, 0.5, 1), "+"))
)))
grid_vpc <- grid_vpc[grid_vpc <= 56]

vpc_arms <- arms |> filter(mgkg == 12)

# One rxSolve per arm rather than one pooled call, for two reasons. (1) Cost:
# rxSolve on an rxUi scales super-linearly in subjects per call, so two calls of
# 200 are markedly cheaper than one of 400. (2) Reseeding inside the loop gives
# the two arms COMMON RANDOM NUMBERS -- subject i draws the same etas in both
# arms, so the pancreatic-vs-non-pancreatic contrast below is paired and
# reflects only the cancer-type effect, not sampling noise between two
# independent cohorts.
sim_vpc <- bind_rows(lapply(seq_len(nrow(vpc_arms)), function(i) {
  rxode2::rxSetSeed(20130926)
  ev <- make_arm(vpc_arms[i, ], n = n_per_arm, n_doses = 4L,
                 obs_times = grid_vpc, id_offset = 0L)
  stopifnot(!anyDuplicated(unique(ev[, c("id", "time", "evid")])))
  rxode2::rxSolve(mod, events = ev, keep = c("arm", "TUMTP_PANC")) |>
    as.data.frame()
}))

vpc_band <- sim_vpc |>
  # Drop t = 0: the first infusion has not yet delivered any drug there, so Cc
  # is exactly 0 and would be dropped by the log scale anyway.
  filter(!is.na(Cc), time > 0) |>
  group_by(arm, time) |>
  summarise(
    Q05 = quantile(Cc, 0.05), Q50 = median(Cc), Q95 = quantile(Cc, 0.95),
    .groups = "drop"
  )

ggplot(vpc_band, aes(time, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~arm) +
  scale_y_log10() +
  labs(
    x = "Time (days)", y = "Ganitumab serum concentration (ug/mL)",
    title = "Projected ganitumab profiles, 12 mg/kg Q2W, first four doses",
    caption = "Replicates the projection in Figure 1 of Zhu 2013."
  ) +
  theme_bw()

stopifnot(
  # Guard: the band was actually computed over both arms and the full window.
  nrow(vpc_band) > 0L, dplyr::n_distinct(vpc_band$arm) == 2L,
  # Every simulated concentration from the end of the first infusion onward is
  # above the assay LLOQ of 20.2 ng/mL (0.0202 ug/mL), consistent with the
  # paper excluding only 2% of samples as below quantification or in error.
  min(vpc_band$Q05[vpc_band$time >= 1]) > 0.0202,
  # Cancer-type separation, asserted PER SUBJECT rather than on cohort
  # quantiles. The two arms share eta draws (common random numbers), so subject
  # i differs between arms only by TUMTP_PANC -- every one of the 200 subjects
  # must therefore show a lower trough in the pancreatic arm. This is an exact
  # 200/200 test rather than a noise-limited comparison of two distributions.
  {
    tail_paired <- sim_vpc |>
      filter(!is.na(Cc), time == max(grid_vpc)) |>
      select(id, arm, Cc) |>
      mutate(arm = ifelse(grepl("^Pancreatic", arm), "panc", "nonpanc")) |>
      tidyr::pivot_wider(names_from = arm, values_from = Cc)
    nrow(tail_paired) == n_per_arm && all(tail_paired$panc < tail_paired$nonpanc)
  }
)
```

# Covariate sensitivity

The published exponents are the transferable part of the covariate model. This
panel sweeps each continuous covariate deterministically (random effects zeroed,
all other covariates at reference) and reports the resulting steady-state
exposure relative to the reference patient. `AUCss` is inversely proportional to
CL, so each curve is `(Cov / Cov_ref)^(-exponent)`.

```{r covariate-sensitivity, fig.cap = "Deterministic covariate sensitivity of steady-state exposure, non-pancreatic patient at 12 mg/kg Q2W."}
cov_grid <- bind_rows(
  tibble(covariate = "WT (kg)",    value = seq(45, 115, by = 5)) |>
    mutate(WT = value, ALB = ALB_REF, CREAT = CREAT_REF),
  tibble(covariate = "ALB (g/L)",  value = seq(25, 50, by = 2.5)) |>
    mutate(WT = WT_REF, ALB = value, CREAT = CREAT_REF),
  tibble(covariate = "CREAT (mg/dL)", value = seq(0.5, 1.6, by = 0.1)) |>
    mutate(WT = WT_REF, ALB = ALB_REF, CREAT = value)
) |>
  mutate(id = row_number(), TUMTP_PANC = 0L)

events_cov <- bind_rows(
  cov_grid |> mutate(time = 0, amt = 12 * WT_REF, evid = 1L, cmt = "central",
                     dur = INFUSION_DUR),
  cov_grid |> mutate(time = 1, amt = NA_real_, evid = 0L, cmt = "central",
                     dur = NA_real_)
) |>
  arrange(id, time, desc(evid))

# Each row of `cov_grid` is its own subject, so the covariate labels are
# recovered by joining on `id` rather than via `keep=`. rxSolve does not return
# a column named `value`, so keeping it is not an option here.
sim_cov <- rxode2::rxSolve(mod_typical, events = events_cov) |>
  as.data.frame() |>
  mutate(id = as.integer(as.character(id))) |>
  group_by(id) |>
  summarise(cl = first(cl), .groups = "drop") |>
  left_join(cov_grid |> select(id, covariate, value), by = "id") |>
  mutate(auc_relative = 0.679 / cl)

stopifnot(
  # Guard: the join must cover every grid point, or a silently-dropped covariate
  # arm would make the slope tests below vacuous.
  nrow(sim_cov) == nrow(cov_grid), !anyNA(sim_cov$covariate), !anyNA(sim_cov$cl)
)

ggplot(sim_cov, aes(value, auc_relative)) +
  geom_line() +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  facet_wrap(~covariate, scales = "free_x") +
  labs(x = "Covariate value", y = "Steady-state exposure relative to reference") +
  theme_bw()

stopifnot(
  # Each curve passes through 1 at its own reference value.
  abs(sim_cov$auc_relative[sim_cov$covariate == "ALB (g/L)" &
                             sim_cov$value == ALB_REF] - 1) < 1e-8,
  abs(sim_cov$auc_relative[sim_cov$covariate == "CREAT (mg/dL)" &
                             sim_cov$value == CREAT_REF] - 1) < 1e-8,
  # Directions match the published signs: exposure falls with weight, and
  # rises with albumin and with creatinine.
  {
    slope <- function(cv) {
      d <- sim_cov |> filter(covariate == cv) |> arrange(value)
      stats::coef(stats::lm(log(auc_relative) ~ log(value), data = d))[["log(value)"]]
    }
    all(abs(slope("WT (kg)")       - (-0.984)) < 1e-6,
        abs(slope("ALB (g/L)")     - ( 0.859)) < 1e-6,
        abs(slope("CREAT (mg/dL)") - ( 0.394)) < 1e-6)
  }
)
```

# Cross-check against the paper's non-compartmental clearances

Zhu 2013 also reports mean (SD) NCA clearances in mL/kg/day from individual
intensive profiles. These are shown for completeness but are **not** asserted
on: they are arithmetic means over small single-cycle subsets whose covariate
values are unpublished, obtained from sampling windows of 336-672 h against a
terminal half-life of 7-10 days.

```{r nca-cl-crosscheck}
nca_cl <- tibble::tribble(
  ~cohort,                              ~n,  ~mean_ml_kg_day, ~sd, ~panc,
  "Non-pancreatic, no gemcitabine (Study 2)", 33L, 10.1, 3.8, 0L,
  "Non-pancreatic, + gemcitabine (Study 3)",  11L, 11.1, 3.1, 0L,
  "Pancreatic, + gemcitabine (Study 1)",      13L, 21.1, 6.4, 1L
) |>
  mutate(
    model_ml_kg_day = ifelse(panc == 1L, 1.154, 0.679) / WT_REF * 1000,
    pct_diff        = 100 * (model_ml_kg_day - mean_ml_kg_day) / mean_ml_kg_day
  )

nca_cl |>
  select(cohort, n, mean_ml_kg_day, sd, model_ml_kg_day, pct_diff) |>
  dplyr::rename(
    "Cohort"                          = cohort,
    "N"                               = n,
    "Zhu 2013 NCA mean CL (mL/kg/day)" = mean_ml_kg_day,
    "SD"                              = sd,
    "Model typical CL (mL/kg/day)"     = model_ml_kg_day,
    "% difference"                     = pct_diff
  ) |>
  knitr::kable(digits = 1, caption = "Model typical CL vs the paper's non-compartmental clearances (context only, not asserted).")
```

The model's typical CL sits within one SD of the two non-pancreatic NCA means
but is 27% below the pancreatic NCA mean. That direction is expected: NCA
`Dose / AUCinf` overestimates CL whenever the terminal phase is under-sampled,
and Study 1's pancreatic cohort was sampled to 336 h against a ~7-day terminal
half-life. The population estimate, which borrows the terminal phase across all
99 subjects, is the more reliable of the two. **The model was not adjusted to
close this gap.**

# Assumptions and deviations

## Errata and unresolved items in the source

* **Negative signs on the albumin and creatinine exponents.** Most text and
  markdown extractions of this Wiley PDF render "ALB on CL" and "CR on CL" as
  bare magnitudes, which invites reading them as positive. They are negative.
  The journal encodes the minus sign as the C0 control byte `0x03` rather than
  an ASCII hyphen, so decoding the byte stream recovers the signs directly --
  `pdftotext -layout | cat -A` over Table 1 gives:

  ```
  WT on CL       0.984 (12.8)     <- no control byte: POSITIVE
  ALB on CL    ^C0.859 (22.2)     <- ^C = 0x03 = minus: NEGATIVE
  CR on CL     ^C0.394 (24.7)     <- ^C = 0x03 = minus: NEGATIVE
  ```

  The *within-table contrast* is what makes this conclusive rather than
  suggestive: weight carries no control byte while albumin and creatinine both
  do, so the absent signs are not a blanket artifact applied uniformly to every
  row. The same encoding shows up in the Results prose, where the dMOF values
  appear as `^C1255.55`, `^C55.971` and `^C34.848` -- all quantities that must
  be negative. Corroborated independently by the paper's prose in three places:
  Results ("higher baseline albumin and creatinine values were associated with
  decreased ganitumab CL"), and Discussion ("baseline albumin was negatively
  correlated with CL for all cancer types" and "lower muscle mass in patients
  with pancreatic cancer may lead to lower creatinine levels and higher CL of
  ganitumab"). Encoded as `-0.859` and `-0.394`.

* **Reference body weight 74.7 kg, not the 70 kg named in the Methods.**
  Zhu 2013 equation (3) defines the reference as the cohort median but reports
  no medians; the "(e.g., 70 kg for body weight)" parenthetical in that sentence
  illustrates the concept rather than stating the value. The reference weight
  was recovered by inverting the four published steady-state AUC values, which
  agree on 74.7 kg to three significant figures and rule out 70 kg outright (see
  "Recovering the reference body weight"). This is a derived, non-paper-stated
  value.

* **Reference albumin (40 g/L) and creatinine (1 mg/dL) are assumed.** These are
  rounded standard adult values. Zhu 2013 reports no baseline-demographics
  table, and no published quantity depends on albumin or creatinine separately,
  so no inversion analogous to the weight derivation is possible. The published
  *exponents* are unaffected and are the transferable part of the covariate
  model; a user applying this model to their own cohort should re-centre on
  their own medians, which rescales the typical CL but preserves every covariate
  effect. Zhu 2013's Discussion notes that cachexia lowers creatinine in
  pancreatic cancer, so the true cohort median may sit below 1 mg/dL.

* **Infusion duration is not stated** anywhere in Zhu 2013 or its supplement.
  The vignette uses a 1-hour intravenous infusion, consistent with the
  Supplemental Table 1 sampling schemes ("prior and end of infusion"; cycle-1
  samples at 0, 1 and 3 h). The choice affects Cmax and Tmax only:
  `AUCtau,ss = Dose / CL` is exactly independent of it, so every AUC assertion
  in this vignette is immune to the assumption. The model file itself carries no
  infusion duration -- it is supplied by the event table.

* **No demographic distributions are published**, so the virtual cohorts vary
  only the model's published random effects and hold covariates at the reference
  values. A covariate-varying virtual population would be invention rather than
  replication. Covariate effects are instead demonstrated deterministically.

* **IIV scale.** Zhu 2013 Table 1 reports each IIV as a percentage. Under the
  paper's own notation in equation (1) -- "estimated variance v^2 (hereafter ...
  v is referred to as OMEGA)" -- the tabulated percentage is OMEGA, the SD on
  the log scale, so the variance is `(percent / 100)^2`. This reading is
  corroborated by the residual error: Table 1's "Proportional error (%) 16.4"
  matches the Results text "RV decreased from 31.5% to 16.5%", which is an
  SD-scale percentage rather than a variance. The alternative back-transform
  `omega^2 = log(1 + CV^2)` would change the IIV on Vp (60.2%) by about 8% and
  the others by under 2%.

* **Residual error is proportional only.** Methods equation (2) writes a
  combined additive-plus-proportional model, but the final structure is
  proportional: the Results state the base model used "a proportional error
  model", and Table 1 reports no additive term.

* **Inter-occasion variability was evaluated and not included** ("Since CL and
  Vc were similar across occasions, and changes in either parameter over time
  were negligible, IOV was not incorporated in the base model"), so the model
  carries none.

## Covariates screened but not retained

Zhu 2013 screened a wide covariate set and retained only weight, albumin,
creatinine and cancer type. The screening ran in two stages, and the model file
records which stage rejected each covariate:

1. **Stage 1** (Methods) -- graphical exploration plus Pearson correlation
   against posthoc CL and Vc. Twenty-three items entered: age, WT, BMI, sex,
   race, CR, ALB, total bilirubin, AST, ALT, ALP, disease stage at enrollment,
   ECOG performance status, platelet count, neutrophils, fasting blood glucose,
   BUN, LDH, IGF-1, IGFBP3, gemcitabine coadministration, study, and
   MDRD-estimated GFR. Creatinine clearance derived from WT and CR was
   deliberately never screened, as collinear with its own inputs.
2. **Stage 2** (Results) -- formal stepwise forward addition at *P* < .01
   followed by backward elimination at *P* < .005, applied to the stage-1
   survivors: on CL, WT, sex, CR, ALB, AST, ALT, ALP, platelets, neutrophils,
   IGF-1, IGFBP3, ECOG, gemcitabine and GFR; on Vc, WT, sex, ALP and
   gemcitabine.

The rejected covariates are documented in the model file's
`covariatesDataExcluded` metadata rather than `covariateData`, since none
carries a published point estimate:

```{r excluded-covariates, echo = FALSE}
excluded <- as.list(ui$meta)$covariatesDataExcluded
tibble(
  Covariate = names(excluded),
  Description = vapply(excluded, function(x) x$description, character(1))
) |>
  knitr::kable(caption = "Covariates screened by Zhu 2013 but not retained in the final model.")
```

Six screened concepts are recorded as a comment in the model file rather than as
`covariatesDataExcluded` entries, because no entry can be written honestly for
them. Platelet count, IGF-1, IGFBP3 and disease stage have no canonical
covariate-register name, and registering one for a covariate that no model
references would leave an unused name in the register. Race is registered only
as per-group indicators (`RACE_WHITE`, `RACE_BLACK`, ...), and Zhu 2013 never
states which race levels its cohort contained, so no indicator can be named
without inventing the composition. Study was screened as a categorical, but the
paper pools Studies 1-3 and reports no per-study coefficient.

One elimination is worth singling out. MDRD-estimated GFR reached stage 2 on CL
and was dropped, while **serum creatinine -- the input to that very estimate --
was retained**. That is not a contradiction but a clue to the mechanism: a
150 kDa IgG1 is not renally cleared, so creatinine is not acting here as a
renal-function marker. It is acting as a marker of muscle mass, which is exactly
the reading the Discussion gives when it attributes the faster pancreatic
clearance to cachexia -- "lower muscle mass in patients with pancreatic cancer
may lead to lower creatinine levels and higher CL of ganitumab". Normalising
creatinine into a GFR estimate divides out the very signal that made it
predictive.

The most consequential of these is gemcitabine coadministration. Tested alone it
was significant (dMOF = -34.848, P < .001), but it was confounded with pancreatic
cancer type, which was the stronger effect (dMOF = -55.971); once cancer type
entered the model gemcitabine became non-significant (P > .2). The paper
corroborated this non-compartmentally, and the resulting conclusion -- that
disease state rather than the chemotherapy backbone drives ganitumab exposure --
is the paper's headline finding.
