---
title: "KN012 denosumab biosimilar vs reference denosumab (Zhu 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{KN012 denosumab biosimilar vs reference denosumab (Zhu 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhu_2026_denosumab_kn012"))$reference`
* Article: <https://doi.org/10.1016/j.bonr.2026.101916>
* Supplement (Fig. S1 model schema, Tables S3-S6 demographics, parameters and simulated exposures): <https://doi.org/10.1016/j.bonr.2026.101916>

Zhu 2026 is a 12-month, double-blind, randomised, active-controlled Phase III
study (CTR20201555) comparing KN012, a candidate denosumab biosimilar, against
US-sourced reference denosumab (Prolia) in Chinese postmenopausal women with
osteoporosis at high risk of fracture. Population PK was a secondary objective.

The authors fitted **two separate population PK models**, one per treatment arm,
rather than a single pooled model with a treatment covariate. This library
reproduces that structure faithfully as two model files:

* `Zhu_2026_denosumab_kn012` - the KN012 biosimilar arm (Supplementary Table 4)
* `Zhu_2026_denosumab_reference` - the US reference denosumab arm (Supplementary Table 5)

Both share the same structure: one compartment, first-order subcutaneous
absorption, and **parallel linear and Michaelis-Menten elimination** from the
central compartment, with body weight on linear clearance as a power term
normalised to 55.5 kg (Supplementary Fig. 1).

```{r descriptions, echo = FALSE, results = "asis"}
for (nm in c("Zhu_2026_denosumab_kn012", "Zhu_2026_denosumab_reference")) {
  cat("* **", nm, "** - ",
      rxode2::rxode(readModelDb(nm))$description, "\n\n", sep = "")
}
```

# Population

The population PK analysis set comprised 277 of the 280 randomised participants
(Zhu 2026 Results 3.5): 139 in the KN012 arm contributing 922 serum
concentrations, and 138 in the US-denosumab arm contributing 925. All were
Chinese postmenopausal women aged 50-80 years with a T-score between -2.5 and
-4.0 at the total hip or lumbar spine plus at least one additional fracture
risk factor. Baseline demographics were well balanced between arms
(Supplementary Table 3): mean age 64.8 years (range 51.0-78.0) and mean weight
55.5 kg (35.0-76.0) for KN012, versus 65.3 years (50.0-80.0) and 56.3 kg
(38.0-77.0) for US-denosumab. The cohort was 98.6% Han Chinese and renally
unremarkable for its age (mean creatinine clearance 73.3 and 75.1 mL/min
respectively, with no participant below 30 mL/min).

Each participant received 60 mg subcutaneously at month 0 and month 6, with
daily calcium carbonate D3 600 mg and vitamin D 800 IU throughout. Twelve
participants per arm were intensively sampled; the remainder were sparsely
sampled (Supplementary Table 1). Serum concentrations were measured by
electrochemiluminescence on the Meso Scale Discovery platform. Estimation used
FOCE with interaction in NONMEM 7.4.

The same information is available programmatically via each model's
`population` metadata, e.g.
`readModelDb("Zhu_2026_denosumab_kn012")()$population`.

Age, weight, BMI, BSA, baseline ALT, AST, albumin, total protein, total
bilirubin, creatinine clearance and anti-drug-antibody status were all screened
by stepwise forward addition and backward elimination; **only body weight on
clearance survived** in either model.

# Source trace

Every `ini()` entry carries an in-file comment naming its source location. The
table below collects them. "S4"/"S5" are Supplementary Tables 4 and 5; both
report the final-model estimate with its relative standard error and 95%
confidence interval, alongside a 1000-run bootstrap.

| Equation / parameter | KN012 | US-denosumab | Source location |
|----------------------|-------|--------------|-----------------|
| `lvc` (Vc, L) | 0.448 | 0.485 | S4 / S5 row `Vc, L` |
| `lcl` (CL at 55.5 kg, L/day) | 0.0983 | 0.116 | S4 / S5 row `CL, L/day` |
| `lka` (KA, 1/day) | 0.0136 | 0.0149 | S4 / S5 row `KA, 1/day` |
| `lvmax` (Vmax, ug/mL/day) | 0.321 | 0.304 | S4 / S5 row `Vmax, mg/day` (unit corrected - see Errata) |
| `lkm` (Km, ug/mL) | 0.0549 | 0.058 | S4 / S5 row `Km, ug/mL` |
| `e_wt_cl` (exponent on WT/55.5) | 1.34 | 1.16 | S4 / S5 row `Weight on CL` and table footnote |
| `etalcl` (omega^2) | 0.0573713 | 0.0343715 | S4 / S5 row `omega(CL),%` = 24.3% / 18.7% CV |
| `etalvmax` (omega^2) | 0.1567849 | 0.2003587 | S4 / S5 row `omega(Vmax), %` = 41.2% / 47.1% CV |
| `propSd` | 0.248 | 0.202 | S4 / S5 row `sigma (Prop), %` |
| `addSd` (ug/mL) | not in model | 0.014 | S5 row `sigma (Add), ug/mL` |
| `d/dt(depot)`, `d/dt(central)` | n/a | n/a | Supplementary Fig. 1 schema |
| `CL = CL_TV * (WT/55.5)^theta` | n/a | n/a | Footnote to S4 / S5 |

The IIV rows are printed as percentages under the symbol omega, i.e. as
coefficients of variation for log-normally distributed parameters; the packaged
variances are `omega^2 = log(1 + CV^2)`.

```{r params, echo = FALSE}
bind_rows(lapply(
  c(KN012 = "Zhu_2026_denosumab_kn012",
    `US-denosumab` = "Zhu_2026_denosumab_reference"),
  function(nm) {
    ui <- rxode2::rxode(readModelDb(nm))
    tibble(parameter = names(ui$theta), value = unname(ui$theta))
  }
), .id = "treatment") |>
  pivot_wider(names_from = treatment, values_from = value) |>
  knitr::kable(digits = 5, caption = "Packaged fixed-effect parameters (log scale where prefixed `l`).")
```

# Validation strategy

Zhu 2026 publishes a directly reproducible target. Supplementary Table 6 reports
the exposure of **a typical 60 kg subject after a single 60 mg dose**, simulated
from each final model. Because it is a typical-value prediction with no
between-subject variability, reproducing it is a deterministic check and is
gated tightly below. Main-text Table 4 additionally reports the observed,
NCA-derived bioequivalence comparison from the trial data itself.

Three checks follow:

1. **Mass balance** - the dose must be fully accounted for by the two
   elimination pathways. This is the gate that would catch rxode2 silently
   discarding the explicit ODE in favour of an analytic one-compartment
   solution, which would drop the Michaelis-Menten term entirely.
2. **Typical-value exposure** - reproduce Supplementary Table 6 via PKNCA.
3. **Virtual cohort** - a stochastic simulation of the trial regimen, checked
   against the observed Table 4 exposures on robust statistics only.

```{r typical-sim}
dose_mg <- 60
wt_ref <- 60          # Supplementary Table 6: "a typical 60kg subject"
six_months <- 182.5   # days; the paper's AUC0-6mon window and 2nd-dose time

models <- c(KN012 = "Zhu_2026_denosumab_kn012",
            `US-denosumab` = "Zhu_2026_denosumab_reference")

# Typical-value (zeroRe) single-dose profile per product. Observations are
# placed on the `central` ODE state, never on the algebraic observable `Cc`;
# rxode2 returns Cc as a column at those rows.
typ_ev <-
  rxode2::et(amt = dose_mg, cmt = "depot") |>
  rxode2::et(seq(0, 1500, by = 0.25), cmt = "central") |>
  as.data.frame()
typ_ev$WT <- wt_ref

sim_typ <- bind_rows(lapply(models, function(nm) {
  m <- rxode2::zeroRe(rxode2::rxode(readModelDb(nm)))
  rxode2::rxSolve(m, typ_ev, returnType = "data.frame")
}), .id = "treatment") |>
  filter(!is.na(Cc))

stopifnot(nrow(sim_typ) > 0, !anyNA(sim_typ$Cc), all(sim_typ$Cc >= 0))
```

## Check 1 - mass balance

With no bioavailability parameter in either model the whole 60 mg dose enters
the central compartment, so at the end of a long solve the dose must equal the
amount removed by linear clearance plus the amount removed by the saturable
pathway, plus whatever is left in the system.

```{r mass-balance}
trap <- function(t, y) sum(diff(t) * (head(y, -1) + tail(y, -1)) / 2)

mass <- sim_typ |>
  group_by(treatment) |>
  summarise(
    linear_mg = cl[1] * trap(time, Cc),
    # the saturable arm is vmax * vc * Cc / (km + Cc) on the amount scale
    mm_mg = trap(time, vmax[1] * vc[1] * Cc / (km[1] + Cc)),
    residual_mg = depot[n()] + central[n()],
    .groups = "drop"
  ) |>
  mutate(total_mg = linear_mg + mm_mg + residual_mg,
         pct_error = 100 * (total_mg - dose_mg) / dose_mg,
         mm_fraction_pct = 100 * mm_mg / dose_mg)

mass |>
  rename("Treatment" = treatment, "Linear (mg)" = linear_mg,
         "Michaelis-Menten (mg)" = mm_mg, "Remaining (mg)" = residual_mg,
         "Total (mg)" = total_mg, "Error (%)" = pct_error,
         "MM share of dose (%)" = mm_fraction_pct) |>
  knitr::kable(digits = 3, caption = "Mass balance of a single 60 mg dose in a typical 60 kg subject.")

# Deterministic check: both sides use the same solve, so the only difference is
# trapezoidal error on a 0.25-day grid. A tight bound is correct here.
stopifnot(all(abs(mass$pct_error) < 0.5))

# The saturable pathway must be materially active. If rxode2 were to auto-solve
# the model analytically and discard the explicit ODE, the Michaelis-Menten term
# would vanish and this share would collapse to zero.
stopifnot(all(mass$mm_fraction_pct > 30))
```

Roughly 45% of the dose is cleared by the saturable pathway at this dose level,
consistent with the paper's statement that denosumab elimination "approaches
linearity at doses >= 60 mg while exhibiting nonlinear target-mediated
disposition below 60 mg" (Discussion).

## Check 2 - typical-value exposure via PKNCA

```{r pknca-typical}
nca_conc <- sim_typ |>
  select(treatment, time, Cc) |>
  mutate(id = 1L)

# Time-zero anchor: pre-dose concentration is 0 for an extravascular dose.
nca_conc <- bind_rows(
  nca_conc,
  nca_conc |> distinct(treatment, id) |> mutate(time = 0, Cc = 0)
) |>
  distinct(treatment, id, time, .keep_all = TRUE) |>
  arrange(treatment, id, time)

conc_obj <- PKNCA::PKNCAconc(nca_conc, Cc ~ time | treatment + id)

dose_df <- nca_conc |>
  distinct(treatment, id) |>
  mutate(time = 0, amt = dose_mg)
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id)

intervals <- data.frame(
  start = c(0, 0),
  end = c(six_months, Inf),
  cmax = c(TRUE, FALSE),
  tmax = c(TRUE, FALSE),
  auclast = c(TRUE, FALSE),
  aucinf.obs = c(FALSE, TRUE),
  half.life = c(FALSE, TRUE)
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_raw <- as.data.frame(nca_res)

# PKNCA computes a requested parameter's dependencies in whichever interval
# needs them, so a name can appear in BOTH intervals: `tmax` is returned for
# 0-182.5 because we asked for it, and again for 0-Inf because `lambda.z` (and
# hence `aucinf.obs` / `half.life`) depends on it. Pivoting without pinning the
# interval turns those duplicates into a list-column, which is not numeric and
# breaks every downstream comparison. Select each parameter from the interval it
# was actually requested in.
nca_wide <-
  bind_rows(
    nca_raw |>
      filter(end == six_months, PPTESTCD %in% c("cmax", "tmax", "auclast")),
    nca_raw |>
      filter(is.infinite(end), PPTESTCD %in% c("aucinf.obs", "half.life"))
  ) |>
  select(treatment, PPTESTCD, PPORRES) |>
  pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

# Guard the list-column failure mode explicitly rather than letting it surface
# three chunks later as "non-numeric argument to binary operator".
stopifnot(
  nrow(nca_wide) == 2L,
  all(vapply(nca_wide[, c("cmax", "auclast", "aucinf.obs")], is.numeric, logical(1)))
)
```

Supplementary Table 6 reports Cmax, AUC0-6mon and AUCinf for each product. The
paper's `AUC0-6mon` is the AUC to the second dose, which for a single-dose
profile is `auclast` over 0-182.5 days.

```{r nca-compare}
# Zhu 2026 Supplementary Table 6, simulated typical 60 kg subject, 60 mg.
published <- tibble::tribble(
  ~treatment,      ~cmax, ~auclast, ~aucinf.obs,
  "KN012",          5.07,   297.05,      298.91,
  "US-denosumab",   4.84,   266.00,      267.49
)

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_wide |> select(treatment, cmax, auclast, aucinf.obs),
  reference = published,
  by = "treatment",
  units = c(cmax = "ug/mL", auclast = "day*ug/mL", aucinf.obs = "day*ug/mL"),
  tolerance_pct = 20
)

knitr::kable(
  cmp,
  caption = "Simulated vs Zhu 2026 Supplementary Table 6 (typical 60 kg subject, single 60 mg dose). * differs from reference by >20%.",
  align = c("l", "l", "r", "r", "r")
)
```

```{r nca-gate}
chk <- nca_wide |>
  left_join(published, by = "treatment", suffix = c("_sim", "_pub")) |>
  mutate(
    cmax_pct = 100 * (cmax_sim - cmax_pub) / cmax_pub,
    auc6_pct = 100 * (auclast_sim - auclast_pub) / auclast_pub,
    aucinf_pct = 100 * (aucinf.obs_sim - aucinf.obs_pub) / aucinf.obs_pub
  )

chk |>
  select(treatment, cmax_pct, auc6_pct, aucinf_pct) |>
  rename("Treatment" = treatment, "Cmax (% diff)" = cmax_pct,
         "AUC0-6mon (% diff)" = auc6_pct, "AUCinf (% diff)" = aucinf_pct) |>
  knitr::kable(digits = 2, caption = "Deviation from the published typical-value exposures.")

# These are typical-value (zeroRe) predictions compared against the paper's own
# typical-value simulation: no subject is drawn, so nothing here varies with the
# rxode2 RNG or the solver thread count, and a tight bound is appropriate.
# Observed magnitudes are all below 0.3%; the 2% bound leaves headroom for
# trapezoidal and extrapolation differences without being unfalsifiable.
stopifnot(
  all(abs(chk$cmax_pct) < 2),
  all(abs(chk$auc6_pct) < 2),
  all(abs(chk$aucinf_pct) < 2)
)
```

All six published values are reproduced to better than 0.3%. Note that this
agreement **requires** the Vmax unit correction documented in the Errata below;
with Vmax taken as the printed mg/day the same six values are missed by 23-50%.

```{r figure-typical}
sim_typ |>
  filter(time <= 365) |>
  ggplot(aes(time, Cc, colour = treatment)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = six_months, linetype = "dotted") +
  labs(x = "Time (days)", y = "Serum concentration (ug/mL)",
       colour = NULL,
       title = "Typical 60 kg subject, single 60 mg subcutaneous dose",
       caption = paste("Reproduces Zhu 2026 Supplementary Table 6.",
                       "Dotted line marks the 6-month second-dose time.")) +
  theme(legend.position = "top")
```

The profile is markedly flip-flop: the absorption half-life implied by
KA = 0.0136/day is about 51 days, far longer than the disposition half-life, so
absorption is rate-limiting and Tmax falls near day 10-13.

## Check 3 - virtual cohort under the trial regimen

```{r cohort}
# rxode2's RNG is partitioned per solver thread, so the drawn cohort differs
# between this machine and CI. Every assertion below is written on robust
# statistics that hold for any cohort the model can produce.
set.seed(20260913)
rxode2::rxSetSeed(20260913)

n_per_arm <- 200L

# Weight: mean +/- SD from Zhu 2026 Results 3.5, truncated to the observed
# ranges in Supplementary Table 3.
arm_wt <- list(
  KN012 = list(mean = 55.5, sd = 7.47, lo = 35, hi = 76, offset = 0L,
               model = "Zhu_2026_denosumab_kn012"),
  `US-denosumab` = list(mean = 56.3, sd = 7.89, lo = 38, hi = 77, offset = 1000L,
                        model = "Zhu_2026_denosumab_reference")
)

make_cohort <- function(spec, n) {
  wt <- pmin(pmax(rnorm(n, spec$mean, spec$sd), spec$lo), spec$hi)
  subj <- tibble(id = spec$offset + seq_len(n), WT = wt)
  obs_times <- sort(unique(c(seq(0, 365, by = 3.5), six_months)))
  bind_rows(
    subj |> tidyr::crossing(time = c(0, six_months)) |>
      mutate(evid = 1L, amt = dose_mg, cmt = "depot"),
    subj |> tidyr::crossing(time = obs_times) |>
      mutate(evid = 0L, amt = NA_real_, cmt = "central")
  ) |>
    arrange(id, time, desc(evid))
}

events <- bind_rows(lapply(names(arm_wt), function(a) {
  make_cohort(arm_wt[[a]], n_per_arm) |> mutate(treatment = a)
}))

stopifnot(
  !anyDuplicated(unique(events[, c("id", "time", "evid")])),
  length(unique(events$id)) == 2L * n_per_arm
)
```

```{r cohort-sim}
sim_cohort <- bind_rows(lapply(names(arm_wt), function(a) {
  m <- rxode2::rxode(readModelDb(arm_wt[[a]]$model))
  ev <- events |> filter(treatment == a) |> select(-treatment)
  rxode2::rxSolve(m, ev, keep = "WT", returnType = "data.frame") |>
    mutate(treatment = a)
})) |>
  filter(!is.na(Cc))

stopifnot(nrow(sim_cohort) > 0, !anyNA(sim_cohort$Cc))
```

```{r figure-vpc}
sim_cohort |>
  group_by(treatment, time) |>
  summarise(Q05 = quantile(Cc, 0.05), Q50 = median(Cc),
            Q95 = quantile(Cc, 0.95), .groups = "drop") |>
  ggplot(aes(time, Q50, colour = treatment, fill = treatment)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.8) +
  labs(x = "Time (days)", y = "Serum concentration (ug/mL)",
       colour = NULL, fill = NULL,
       title = "Simulated 12-month profile, 60 mg at months 0 and 6",
       caption = paste("Median with 5th-95th percentile band,", n_per_arm,
                       "subjects per arm. Compare Zhu 2026 Supplementary Fig. 4 (pcVPC).")) +
  theme(legend.position = "top")
```

```{r cohort-exposure}
cohort_nca <- sim_cohort |>
  filter(time <= six_months) |>
  group_by(treatment, id) |>
  summarise(cmax = max(Cc), auc6 = trap(time, Cc), .groups = "drop")

# Zhu 2026 main-text Table 4: observed, NCA-derived exposures from the trial.
observed_tbl4 <- tibble::tribble(
  ~treatment,     ~cmax_obs, ~auc6_obs,
  "KN012",             5.47,    319.89,
  "US-denosumab",      5.08,    274.76
)

cohort_summary <- cohort_nca |>
  group_by(treatment) |>
  summarise(cmax_med = median(cmax), auc6_med = median(auc6), .groups = "drop") |>
  left_join(observed_tbl4, by = "treatment") |>
  mutate(cmax_pct = 100 * (cmax_med - cmax_obs) / cmax_obs,
         auc6_pct = 100 * (auc6_med - auc6_obs) / auc6_obs)

cohort_summary |>
  rename("Treatment" = treatment, "Cmax median (ug/mL)" = cmax_med,
         "AUC0-6mon median (day*ug/mL)" = auc6_med,
         "Cmax observed" = cmax_obs, "AUC0-6mon observed" = auc6_obs,
         "Cmax (% diff)" = cmax_pct, "AUC0-6mon (% diff)" = auc6_pct) |>
  knitr::kable(digits = 2,
               caption = "Simulated cohort medians vs the observed NCA exposures of Zhu 2026 Table 4.")

# The two sides differ by a physical mechanism (a drawn weight and CL/Vmax per
# subject, versus the trial's own sparse-sampling NCA on real patients), so this
# is asserted on the median only and with a wide bound. It is a sanity check
# that the cohort sits in the right place, not a reproduction gate; Check 2 is
# the reproduction gate.
stopifnot(all(abs(cohort_summary$cmax_pct) < 35),
          all(abs(cohort_summary$auc6_pct) < 35))
```

The simulated cohort medians land within about 4% of the observed Table 4
exposures in both arms, which is closer than this check is designed to
guarantee. The residual sign is worth a note: the cohort medians sit slightly
*above* the model's own typical-value prediction of Supplementary Table 6
(5.65 against 5.07 for KN012 Cmax), because the cohort is drawn at the arm's
mean weight of 55.5 kg while Supplementary Table 6 fixes a 60 kg subject.
Weight acts on linear clearance with a positive exponent, so a lighter cohort
clears more slowly and is more exposed. The two sides also differ in kind -
Table 4 is the trial's own NCA on a sparse sampling grid, whereas this is a
model-based exposure on a virtual weight distribution - which is why the gate
above is deliberately wide and asserted on the median only.

```{r biosimilarity}
# Zhu 2026 main-text Table 4: GMR of KN012 to US-denosumab from the trial's own
# ANOVA on log-transformed NCA exposures.
published_gmr <- c(cmax_med = 107.68, auc6_med = 116.43)
metric_label <- c(cmax_med = "Cmax", auc6_med = "AUC0-6mon")

ratio <- cohort_summary |>
  select(treatment, cmax_med, auc6_med) |>
  pivot_longer(-treatment) |>
  pivot_wider(names_from = treatment, values_from = value) |>
  mutate(ratio = KN012 / `US-denosumab`,
         metric = unname(metric_label[name]),
         published_pct = unname(published_gmr[name]))

ratio |>
  select(metric, KN012, `US-denosumab`, ratio, published_pct) |>
  rename("Metric" = metric, "KN012" = KN012,
         "US-denosumab" = `US-denosumab`, "Simulated ratio" = ratio,
         "Zhu 2026 Table 4 GMR (%)" = published_pct) |>
  knitr::kable(digits = 3,
               caption = paste("Exposure ratio of KN012 to US-denosumab in the",
                               "simulated cohort, against the trial's published GMR."))

# Zhu 2026 concluded biosimilarity because every 90% CI fell within 0.80-1.25.
# The median ratio of a simulated cohort is not the paper's GMR of a fitted
# ANOVA, so this asserts only the qualitative conclusion the paper draws.
stopifnot(all(ratio$ratio > 0.80), all(ratio$ratio < 1.25))
```

# Assumptions and deviations

## Errata - the printed unit of Vmax is wrong

Supplementary Tables 4 and 5 print the saturable elimination parameter as
**`Vmax, mg/day`**. Read literally, the Michaelis-Menten term would be
`Vmax * Cc / (Km + Cc)` on the amount scale. Taken that way, neither model
reproduces the paper's own simulated exposures in Supplementary Table 6:

| | Cmax | AUC0-6mon | AUCinf |
|---|---|---|---|
| KN012, Vmax as printed (mg/day) | -29% | -50% | -50% |
| KN012, Vmax as ug/mL/day | +0.04% | +0.03% | +0.22% |
| US-denosumab, Vmax as printed (mg/day) | -23% | -43% | -43% |
| US-denosumab, Vmax as ug/mL/day | -0.04% | +0.03% | +0.16% |

Treating Vmax as a **concentration** rate in ug/mL/day, so that the amount-scale
rate is `Vmax * Vc * Cc / (Km + Cc)`, reproduces all six published values to
within 0.22%. Three points make this a unit-tag slip rather than a fitted
correction:

1. The scale factor is each model's **own** Vc - 0.448 L for KN012 and 0.485 L
   for US-denosumab. A single fudge constant would not reproduce both.
2. Six independent published targets are matched simultaneously by one
   structural change with no free parameter.
3. No value of Km can account for the gap: raising Km reduces the saturable
   loss, but no single Km reproduces Cmax and AUCinf jointly. Solving for the Km
   that makes KN012's AUCinf land exactly on 298.91 with Vmax kept as the
   printed mg/day requires Km = 1.64 ug/mL, thirty times the published 0.0549,
   and even then Cmax is 14.1% low (4.36 against 5.07).

The packaged models therefore keep the printed value 0.321 / 0.304 verbatim in
`ini()` and apply the `* vc` factor in `model()`, with the reasoning recorded
inline. Supplementary Fig. 1 labels the saturable arm `Vmax/(Km+Conc)` and the
figure caption describes Km as an "amount", while the table gives Km in ug/mL;
the schema and the reproduced simulation both support Km being a concentration,
which is how it is encoded.

## Other assumptions

- **No bioavailability parameter.** Neither model reports an F. Both products
  were given subcutaneously only, with no intravenous reference, so CL and Vc
  are apparent values (CL/F, Vc/F) even though the source tables print them
  without the `/F` qualifier. The packaged models take F = 1, which is what
  makes the mass-balance check above close on the full 60 mg.
- **IIV scale.** Supplementary Tables 4 and 5 report inter-individual
  variability as a percentage under the symbol omega, with no separate CV%
  column to settle the scale. This is read as a coefficient of variation for
  log-normally distributed parameters and converted with
  `omega^2 = log(1 + CV^2)`. The alternative reading, omega as a log-scale SD
  in percent, changes the variances by under 3% and would not be distinguishable
  from the published summaries.
- **Weight centering.** Both footnotes normalise to 55.5 kg, which is the KN012
  arm's mean weight. The US-denosumab arm's own mean is 56.3 kg, but its
  footnote still prints 55.5, so 55.5 is reproduced for both as a shared model
  constant rather than re-derived per arm.
- **Residual error differs between the two models.** KN012 has a proportional
  term only; US-denosumab has combined proportional and additive. This is as
  published, not an omission.
- **AUC0-6mon window.** Taken as 0-182.5 days. The paper says "6 months" without
  giving a day count; 182.5 is half of 365 and coincides with the second dose.
- **Virtual cohort weights** are drawn from a normal distribution using the
  mean and SD published in Results 3.5, truncated to the ranges in
  Supplementary Table 3. The paper does not publish the weight distribution's
  shape.
- **No PD model is packaged.** Zhu 2026 reports bone mineral density and bone
  turnover marker endpoints (Figs. 2 and 3) descriptively, via ANCOVA on
  percent change from baseline. No exposure-response or turnover model links
  them to the PK, so there is no PD model in the paper to extract. The sibling
  model `Choi_2025_denosumab` provides a denosumab PK/PD model with a
  lumbar-spine BMD indirect-response component.
- **No anti-drug-antibody effect.** ADA was screened as a covariate and not
  retained in either model; the paper reports no clinically meaningful PK
  difference between ADA-positive and ADA-negative participants.
