---
title: "Ropeginterferon alfa-2b (Zhu 2021)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ropeginterferon alfa-2b (Zhu 2021)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zhu_2021_ropeginterferon"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zhu_2021_ropeginterferon"))$description`
* Article (open access): <https://doi.org/10.3389/fphar.2021.673492>
* Supplementary Material (Supplementary Tables 1-3, one DOCX): available from
  the article landing page and from Europe PMC under PMC8193675.

Zhu 2021 is the first population PK model of ropeginterferon alfa-2b. It
pools two single-dose phase I studies in healthy volunteers to ask whether
the PK differs between Caucasian and Chinese subjects, and concludes it does
not once body weight is accounted for. A later analysis of the same drug in
Chinese and Japanese patients with polycythaemia vera, with the same
quasi-equilibrium TMDD structure, is packaged as `Qin_2025_ropeginterferon`.

# Population

Fifty-seven healthy adults contributed 894 serum concentrations (Zhu 2021
Table 1):

* **Study A09-102** (Montreal, Canada): 30 Caucasian men given a single
  subcutaneous dose of 24, 48, 90, 180, 225 or 270 ug; body weight
  79.4 +/- 9.41 kg; 456 observations.
* **Study A17-101** (Beijing, China): 27 Chinese adults (15 men, 12 women)
  given a single subcutaneous dose of 90, 180 or 270 ug; body weight
  68.1 +/- 10 kg; 438 observations.

Ages were 18-45 years by protocol (mean 32.3 years). Sampling ran from
pre-dose to 672 h. Nine subjects (six Caucasian, three Chinese) were
excluded for drop-out.

```{r population}
str(readModelDb("Zhu_2021_ropeginterferon")()$population)
```

# Source trace

Every `ini()` value carries an in-file comment pointing to its source. The
table collects them.

| Equation / parameter | Value | Source location |
|---|---|---|
| `lka` | log(0.14 / 24) 1/h | Table 2, ka 0.14 1/day |
| `ltlag` | log(0.426) h | Table 2, tlag 0.426 h |
| `lcl` | log(0.778 / 24) L/h | Table 2, CL/F 0.778 L/day (70 kg) |
| `lvc` | log(2.32) L | Table 2, V/F |
| `lrbase` | log(0.111) ng/mL | Table 2, R0 |
| `lkint` | fixed(log(0.0788)) 1/h | Table 2 (Fixed); Results, Base Model |
| `lkdeg` | log(0.544) 1/h | Table 2, kdeg |
| `lkd` | fixed(log(0.142)) ng/mL | Table 2 (Fixed); Results, Base Model |
| `e_wt_cl` | 0.927 | Table 2 'Impact of body weight'; Equation 15 |
| `etalcl`, `etalvc`, `etalka` | 35.7%, 90.8%, 63.5% CV | Table 2, Between subject variability |
| `propSd`, `addSd` | 0.187, 0.342 ng/mL | Table 2, Residual unexplained variability; Equation 12 |
| `d/dt(depot)` | | Equation 2 |
| `d/dt(central)` (total drug) | | Equation 3 |
| `d/dt(total_target)` | | Equation 4 |
| `total_target(0) = R0`, `ksyn = R0 * kdeg` | | Equation 5; Figure 1B |
| free drug `cfree` | | Equation 6 |
| `CL/F = 0.778 * (WT/70)^0.927` | | Equation 15 |

# Unit check: kint and kdeg are per hour

Table 2 prints CL/F (L/day) and ka (1/day) in days but tlag, kint and kdeg in
hours, and Supplementary Table 2 repeats exactly the same mix. The two readings
of the target-mediated rates give very different models, so the choice is
tested here against the paper's own noncompartmental analysis
(Supplementary Table 1, arithmetic means of the observed data) with a
deterministic typical-value solve at the Caucasian mean weight.

```{r unit-check, message = FALSE, warning = FALSE}
mod <- readModelDb("Zhu_2021_ropeginterferon")
mod_typ <- mod |> rxode2::zeroRe()

typical_auc <- function(m, doses, wt) {
  ev <- dplyr::bind_rows(
    tibble::tibble(id = seq_along(doses), time = 0, evid = 1L, amt = doses, cmt = "depot"),
    tidyr::expand_grid(id = seq_along(doses), time = seq(0, 2000, by = 1)) |>
      dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  ) |>
    dplyr::arrange(id, time, dplyr::desc(evid)) |>
    dplyr::mutate(WT = wt)
  s <- rxode2::rxSolve(m, ev, returnType = "data.frame")
  s |>
    dplyr::filter(!is.na(Cc)) |>
    dplyr::distinct(id, time, .keep_all = TRUE) |>
    dplyr::group_by(id) |>
    dplyr::summarise(
      auc = sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2),
      cmax = max(Cc),
      .groups = "drop"
    ) |>
    dplyr::mutate(dose = doses)
}

doses_cau <- c(24, 48, 90, 180, 225, 270)
nca_cau <- c(373.97, 620.61, 1243.54, 3891.48, 4352.23, 6184.44)

per_hour <- typical_auc(mod_typ, doses_cau, wt = 79.4)
# The alternative reading: kint and kdeg per day, i.e. 24-fold slower.
mod_day <- mod_typ |>
  rxode2::ini(lkint = log(0.0788 / 24), lkdeg = log(0.544 / 24))
per_day <- typical_auc(mod_day, doses_cau, wt = 79.4)

unit_tab <- tibble::tibble(
  dose = doses_cau,
  nca_mean = nca_cau,
  per_hour = per_hour$auc,
  per_day = per_day$auc
) |>
  dplyr::mutate(
    pct_hour = 100 * (per_hour - nca_mean) / nca_mean,
    pct_day = 100 * (per_day - nca_mean) / nca_mean
  )
unit_tab |>
  dplyr::rename(
    "Dose (ug)" = dose,
    "NCA mean AUC0-inf (ng*h/mL)" = nca_mean,
    "kint/kdeg per hour" = per_hour,
    "kint/kdeg per day" = per_day,
    "% diff, per hour" = pct_hour,
    "% diff, per day" = pct_day
  ) |>
  knitr::kable(digits = 0, caption = "Typical-value AUC0-inf at 79.4 kg against the Supplementary Table 1 Caucasian NCA means.")

# Dose-normalised exposure rises from 24 to 270 ug in the observed NCA
# (22.9 vs 15.6 ng*h/mL per ug, ratio 1.47) because the target-mediated
# arm saturates. Only the per-hour reading reproduces that (ratio ~1.9);
# read per day, the target arm is negligible and the ratio falls below 1.
nonlin <- function(x) (x$auc[6] / 270) / (x$auc[1] / 24)
c(per_hour = nonlin(per_hour), per_day = nonlin(per_day))
stopifnot(
  nonlin(per_hour) > 1.3,
  nonlin(per_day) < 1.1,
  # The per-hour reading stays within 30% of every observed mean; the
  # per-day reading overpredicts the three lowest doses more than 2-fold.
  all(abs(unit_tab$pct_hour) < 30),
  all(unit_tab$pct_day[1:3] > 100)
)
```

The per-hour reading reproduces the observed greater-than-proportional rise
in dose-normalised exposure (ratio about 1.9 against the observed 1.47) and
lands within 30% of every observed mean AUC, each of which comes from about
five subjects. The per-day reading overpredicts the three lowest doses
by more than 100%. The model is therefore run in hours with kint, kdeg
and tlag as printed and CL/F and ka divided by 24.

# Virtual cohort

Nine arms reproduce the study design: six Caucasian dose groups and three
Chinese dose groups, 100 subjects each. Body weight is drawn from a normal
distribution with the Table 1 mean and SD of each study, redrawn (not
clamped) outside 45-120 kg.

```{r cohort}
set.seed(2021)

draw_wt <- function(n, mean, sd) {
  wt <- rnorm(n, mean, sd)
  bad <- wt < 45 | wt > 120
  while (any(bad)) {
    wt[bad] <- rnorm(sum(bad), mean, sd)
    bad <- wt < 45 | wt > 120
  }
  wt
}

obs_times <- c(0, 1, 3, 6, 9, 12, 16, 24, 36, 48, 72, 96, 120, 144, 168,
               192, 240, 288, 336, 504, 672)

make_arm <- function(n, dose, pop, wt_mean, wt_sd, id_offset) {
  ids <- id_offset + seq_len(n)
  wt <- draw_wt(n, wt_mean, wt_sd)
  dose_rows <- tibble::tibble(id = ids, time = 0, evid = 1L, amt = dose, cmt = "depot")
  obs_rows <- tidyr::expand_grid(id = ids, time = obs_times) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::left_join(tibble::tibble(id = ids, WT = wt), by = "id") |>
    dplyr::mutate(
      population = pop,
      dose = dose,
      treatment = paste0(pop, " ", dose, " ug")
    ) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}

arms <- tibble::tribble(
  ~pop,        ~dose, ~wt_mean, ~wt_sd,
  "Caucasian",    24,     79.4,   9.41,
  "Caucasian",    48,     79.4,   9.41,
  "Caucasian",    90,     79.4,   9.41,
  "Caucasian",   180,     79.4,   9.41,
  "Caucasian",   225,     79.4,   9.41,
  "Caucasian",   270,     79.4,   9.41,
  "Chinese",      90,     68.1,  10,
  "Chinese",     180,     68.1,  10,
  "Chinese",     270,     68.1,  10
)
n_per_arm <- 100L
events <- dplyr::bind_rows(lapply(seq_len(nrow(arms)), function(i) {
  make_arm(n_per_arm, arms$dose[i], arms$pop[i], arms$wt_mean[i], arms$wt_sd[i],
           id_offset = (i - 1L) * n_per_arm)
}))
stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
sim <- rxode2::rxSolve(mod, events = events,
                       keep = c("population", "dose", "treatment", "WT")) |>
  as.data.frame()
```

# Replicate published figures

## Figure 2: concentration-time profiles by dose

```{r figure-2}
sim |>
  dplyr::filter(time > 0) |>
  dplyr::group_by(population, dose, time) |>
  dplyr::summarise(
    Q10 = quantile(Cc, 0.10), Q50 = median(Cc), Q90 = quantile(Cc, 0.90),
    .groups = "drop"
  ) |>
  dplyr::mutate(dose = factor(paste(dose, "ug"), levels = paste(doses_cau, "ug"))) |>
  ggplot(aes(time, Q50, colour = population, fill = population)) +
  geom_ribbon(aes(ymin = Q10, ymax = Q90), alpha = 0.2, colour = NA) +
  geom_line() +
  facet_wrap(~dose) +
  scale_y_log10() +
  labs(x = "Time (h)", y = "Total serum ropeginterferon alfa-2b (ng/mL)",
       colour = NULL, fill = NULL,
       caption = "Median and 10th-90th percentiles of the simulation. Compare Figure 2B of Zhu 2021.")
```

## Figure 3: apparent clearance falls with dose

Figure 3 of Zhu 2021 plots NCA CL/F (L/h) against dose; the fitted trend
falls from about 0.078 L/h at 24 ug to about 0.05 L/h at 270 ug in Caucasians
and from about 0.07 to 0.04 L/h in Chinese subjects.

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |> dplyr::mutate(time = 0, Cc = 0)
) |>
  dplyr::distinct(id, treatment, time, .keep_all = TRUE) |>
  dplyr::arrange(id, treatment, time)

dose_df <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::select(id, time, amt, treatment)

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "ng/mL", timeu = "h")
dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "ug")
intervals <- data.frame(start = 0, end = Inf, cmax = TRUE, tmax = TRUE,
                        auclast = TRUE, aucinf.obs = TRUE, half.life = TRUE)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

nca_ind <- as.data.frame(nca_res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "auclast", "aucinf.obs")) |>
  dplyr::select(treatment, id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES) |>
  dplyr::left_join(dplyr::distinct(dose_df, id, amt), by = "id") |>
  dplyr::left_join(dplyr::distinct(sim, id, population), by = "id") |>
  # ug / (ng*h/mL) = L/h
  dplyr::mutate(clf = amt / aucinf.obs)

nca_ind |>
  ggplot(aes(amt, clf, colour = population)) +
  geom_point(alpha = 0.15, position = position_jitter(width = 3)) +
  stat_summary(fun = median, geom = "line", linewidth = 1) +
  facet_wrap(~population) +
  scale_x_continuous(breaks = doses_cau) +
  coord_cartesian(ylim = c(0, 0.2)) +
  labs(x = "Dose (ug)", y = "CL/F = Dose / AUC0-inf (L/h)", colour = NULL,
       caption = "Simulated subjects; line = median by dose. Compare Figure 3 of Zhu 2021.")

clf_med <- nca_ind |>
  dplyr::group_by(population, amt) |>
  dplyr::summarise(clf = median(clf), .groups = "drop")
clf_med
stopifnot(
  # The saturable target arm makes CL/F fall with dose in both populations.
  with(clf_med, clf[population == "Caucasian" & amt == 24] /
         clf[population == "Caucasian" & amt == 270]) > 1.3,
  with(clf_med, clf[population == "Chinese" & amt == 90] /
         clf[population == "Chinese" & amt == 270]) > 1.1
)
```

## Comparison against the published NCA (Supplementary Table 1)

Supplementary Table 1 reports arithmetic means of the observed data, so the
simulated values below are arithmetic means across simulated subjects.

```{r nca-compare}
published <- tibble::tribble(
  ~treatment,           ~cmax, ~auclast, ~aucinf.obs,
  "Caucasian 24 ug",     1.82,   298.12,      373.97,
  "Caucasian 48 ug",     2.42,   391.68,      620.61,
  "Caucasian 90 ug",     5.27,  1147.50,     1243.54,
  "Caucasian 180 ug",   20.68,  3718.03,     3891.48,
  "Caucasian 225 ug",   21.26,  4210.42,     4352.23,
  "Caucasian 270 ug",   25.21,  5995.99,     6184.44,
  "Chinese 90 ug",       4.63,  1057.50,     1280.31,
  "Chinese 180 ug",     14.63,  3422.24,     3516.24,
  "Chinese 270 ug",     24.14,  6983.06,     7998.29
)

sim_mean <- nca_ind |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(
    cmax = mean(cmax), auclast = mean(auclast),
    aucinf.obs = mean(aucinf.obs, na.rm = TRUE),
    .groups = "drop"
  )

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = sim_mean,
  reference = published,
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(cmp, caption = "Simulated vs. published mean NCA. * differs from the reference by >20%.")

pct <- sim_mean |>
  dplyr::inner_join(published, by = "treatment", suffix = c("_sim", "_pub")) |>
  dplyr::mutate(
    d_cmax = 100 * (cmax_sim - cmax_pub) / cmax_pub,
    d_auc = 100 * (aucinf.obs_sim - aucinf.obs_pub) / aucinf.obs_pub
  )
stopifnot(
  # Centre: a mis-transcribed CL, V, ka or unit moves every arm together.
  abs(median(pct$d_auc)) < 15,
  abs(median(pct$d_cmax)) < 25,
  # Envelope across arms (each observed arm has only 5-10 subjects).
  quantile(abs(pct$d_auc), 0.9) < 45
)
```

Simulated mean AUC0-inf lies within 20% of the published mean in every arm.
The observed arms hold 5-10 subjects each, so Cmax, which also depends on the
large V/F and ka variability, scatters more widely about the simulation. The
observed 48 ug AUClast is well below its AUC0-inf (392 vs 621 ng\*h/mL) because
concentrations fell below the 50 pg/mL quantification limit early, which a
simulation without censoring does not reproduce.

## Figure 6: steady-state exposure at 100 and 200 ug every 2 weeks

```{r figure-6}
make_ss <- function(n, dose, pop, wt_mean, wt_sd, id_offset) {
  ids <- id_offset + seq_len(n)
  wt <- draw_wt(n, wt_mean, wt_sd)
  tau <- 336
  n_dose <- 10
  t_last <- (n_dose - 1) * tau
  dose_rows <- tidyr::expand_grid(id = ids, time = (0:(n_dose - 1)) * tau) |>
    dplyr::mutate(evid = 1L, amt = dose, cmt = "depot")
  obs_rows <- tidyr::expand_grid(id = ids, time = t_last + c(0, 1, 3, 6, 12, seq(24, tau, by = 12))) |>
    dplyr::mutate(evid = 0L, amt = 0, cmt = "central")
  dplyr::bind_rows(dose_rows, obs_rows) |>
    dplyr::left_join(tibble::tibble(id = ids, WT = wt), by = "id") |>
    dplyr::mutate(population = pop, regimen = paste(dose, "ug Q2W"),
                  treatment = paste(pop, regimen)) |>
    dplyr::arrange(id, time, dplyr::desc(evid))
}
ss_arms <- tidyr::expand_grid(dose = c(100, 200), pop = c("Caucasian", "Chinese")) |>
  dplyr::mutate(wt_mean = ifelse(pop == "Caucasian", 79.4, 68.1),
                wt_sd = ifelse(pop == "Caucasian", 9.41, 10))
ev_ss <- dplyr::bind_rows(lapply(seq_len(nrow(ss_arms)), function(i) {
  make_ss(200L, ss_arms$dose[i], ss_arms$pop[i], ss_arms$wt_mean[i], ss_arms$wt_sd[i],
          id_offset = (i - 1L) * 200L)
}))
stopifnot(!anyDuplicated(unique(ev_ss[, c("id", "time", "evid")])))

sim_ss <- rxode2::rxSolve(mod, events = ev_ss,
                          keep = c("population", "regimen", "treatment")) |>
  as.data.frame()

t_last <- 9 * 336
conc_ss <- PKNCA::PKNCAconc(
  sim_ss |> dplyr::filter(!is.na(Cc)) |> dplyr::select(id, time, Cc, treatment),
  Cc ~ time | treatment + id
)
dose_ss <- PKNCA::PKNCAdose(
  ev_ss |> dplyr::filter(evid == 1) |> dplyr::select(id, time, amt, treatment),
  amt ~ time | treatment + id
)
res_ss <- PKNCA::pk.nca(PKNCA::PKNCAdata(
  conc_ss, dose_ss,
  intervals = data.frame(start = t_last, end = t_last + 336, auclast = TRUE)
))
auc_ss <- as.data.frame(res_ss) |>
  dplyr::filter(PPTESTCD == "auclast") |>
  dplyr::left_join(dplyr::distinct(sim_ss, id, population, regimen), by = "id")

auc_ss |>
  ggplot(aes(PPORRES, fill = population, colour = population)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~regimen) +
  coord_cartesian(xlim = c(0, 30000)) +
  labs(x = "AUCss (ng*h/mL)", y = "Density", fill = NULL, colour = NULL,
       caption = "Simulated AUC over the 10th dosing interval. Compare Figure 6 of Zhu 2021.")

fig6 <- auc_ss |>
  dplyr::group_by(regimen, population) |>
  dplyr::summarise(
    sim_p10 = quantile(PPORRES, 0.1), sim_p50 = median(PPORRES),
    sim_p90 = quantile(PPORRES, 0.9), .groups = "drop"
  ) |>
  dplyr::left_join(
    tibble::tribble(
      ~regimen,       ~population,  ~pub_p10, ~pub_p90,
      "100 ug Q2W",   "Caucasian",      1867,    11595,
      "100 ug Q2W",   "Chinese",        2231,    13403,
      "200 ug Q2W",   "Caucasian",      3776,    14890,
      "200 ug Q2W",   "Chinese",        4384,    16922
    ),
    by = c("regimen", "population")
  )
fig6 |>
  dplyr::rename(
    "Regimen" = regimen, "Population" = population,
    "Simulated P10" = sim_p10, "Simulated median" = sim_p50, "Simulated P90" = sim_p90,
    "Published P10" = pub_p10, "Published P90" = pub_p90
  ) |>
  knitr::kable(digits = 0, caption = "AUCss 80% intervals (ng*h/mL): simulation vs. Zhu 2021 Results / Figure 6.")

stopifnot(
  # The paper's qualitative conclusion: Chinese exposure is similar to, and
  # slightly higher than, Caucasian exposure at the same dose (lower weight,
  # so lower CL/F). Compare medians, which are robust to the cohort draw.
  all(with(fig6, sim_p50[population == "Chinese"] / sim_p50[population == "Caucasian"]) > 0.8),
  all(with(fig6, sim_p50[population == "Chinese"] / sim_p50[population == "Caucasian"]) < 1.5)
)

# Upper bound from the linear arm alone: with no target-mediated
# elimination, AUC over one interval at steady state is exactly
# Dose / (CL/F), so its 90th percentile at the mean weight is
# Dose / CL_typ * exp(qnorm(0.9) * omega_CL). The target arm can only
# lower the AUC, so no reading of Table 2 reaches the published P90.
omega_cl <- sqrt(log(0.357^2 + 1))
linear_p90 <- fig6 |>
  dplyr::mutate(
    dose = ifelse(regimen == "100 ug Q2W", 100, 200),
    wt = ifelse(population == "Caucasian", 79.4, 68.1),
    lin_p90 = dose / (0.778 / 24 * (wt / 70)^0.927) * exp(qnorm(0.9) * omega_cl)
  )
linear_p90 |> dplyr::select(regimen, population, lin_p90, pub_p90)
stopifnot(all(linear_p90$lin_p90 < 0.7 * linear_p90$pub_p90))
```

The simulated 80% intervals are narrower and lower than those printed for
Figure 6. The linear part of the model alone bounds the published upper
limits: with no target-mediated elimination at all, the steady-state AUC over
one interval is exactly Dose / (CL/F), and the 90th percentile of that
quantity at 100 ug is about 4,300 ng\*h/mL for a 79.4 kg subject and about
4,900 ng\*h/mL for a 68.1 kg subject given the 35.7% BSV on CL/F, against the
published 11,595 and 13,403 ng\*h/mL. The target-mediated arm can only lower
the AUC further. The Figure 6 intervals therefore cannot be
reproduced from Table 2 by any reading of its units; the Simulx settings used
for Figure 6 are not reported, so the discrepancy is recorded rather than
resolved. The single-dose NCA comparison above, which is the paper's direct
data summary, is reproduced.

# Assumptions and deviations

* **Mixed printed time units.** Table 2 and Supplementary Table 2 give CL/F
  in L/day and ka in 1/day, but tlag, kint and kdeg in hours. The model runs
  in hours; CL/F and ka are divided by 24 in `ini()`. The per-hour reading of
  kint and kdeg is the one the paper's own single-dose NCA supports (see the
  unit check above). tlag is 0.426 h either way in effect: observed profiles
  are quantifiable at the 1 h sample.
* **Observed quantity.** The paper defines both Ctotal and Cfree (Equations
  3-6) but does not state which one the observations were fitted to. The
  model reports total drug, `Cc = central / vc`, the conventional choice for a
  quasi-equilibrium TMDD model fitted to a sandwich immunoassay; above about
  1 ng/mL the two differ by at most the receptor pool (well under 1 ng/mL).
* **Between-subject variability scale.** Table 2 reports BSV as CV% without
  stating the conversion. The variances are `omega^2 = log(CV^2 + 1)`, the
  exact log-normal relation; with the approximation `omega^2 = CV^2` the V/F
  variance would be 0.82 instead of 0.60.
* **Residual error.** Equation 12, `Y = IPRED * (1 + eps_prop) + eps_add`, with
  the two epsilons reported separately, is encoded as the nlmixr2 combined
  additive + proportional error; the 18.7% and 0.342 ng/mL are taken as
  standard deviations, as the Table 2 units indicate.
* **Weight reference.** Equation 15 divides weight by 70 kg, and the abstract
  and Discussion quote CL/F "in 70-kg subjects"; the Methods text calls the
  reference "the median value", which is not reported. 70 kg is used as
  printed.
* **Virtual cohort.** Weights are normal with the Table 1 mean and SD for each
  study; sex is not needed because it is not in the model. The simulation
  applies no quantification-limit censoring.
* **Figure 6.** The published steady-state 80% intervals are not reproduced
  (see above).
