---
title: "SPT-07A (D-borneol) whole-body PBPK in rats, dogs and humans (Zhu 2024)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SPT-07A (D-borneol) whole-body PBPK in rats, dogs and humans (Zhu 2024)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

Zhu 2024 builds a single whole-body PBPK structure and parameterises it three
times -- once per species -- so the paper contributes three model files that
share this vignette.

```{r load-models}
model_names <- c(
  rat   = "Zhu_2024_borneol_rat_pbpk",
  dog   = "Zhu_2024_borneol_dog_pbpk",
  human = "Zhu_2024_borneol_human_pbpk"
)
# readModelDb() returns the model FUNCTION; rxode2::rxode() resolves it to a ui.
uis <- lapply(model_names, function(nm) rxode2::rxode(readModelDb(nm)))
```

* Citation: `r uis$rat$reference`
* Article: <https://doi.org/10.3390/pharmaceutics16121596>
* PMC full text: <https://www.ncbi.nlm.nih.gov/pmc/articles/PMC11676658/>

There is no supplementary material for this article, and no erratum or
corrigendum was located.

SPT-07A is D-borneol, a bicyclic monoterpene in development in China for acute
ischaemic stroke. The model is built bottom-up from in vitro metabolism data:
SPT-07A is cleared almost entirely by UDP-glucuronosyltransferase (UGT)
glucuronidation, and unusually the kidney and intestine carry a substantial
share of it, which is what motivated the whole-body treatment. All tissues are
perfusion-rate limited except adipose, which is permeability-limited because
rat tissue-distribution work showed marked adipose accumulation.

```{r descriptions, results = "asis", echo = FALSE}
for (sp in names(uis)) {
  cat("\n**", sp, "** -- ", uis[[sp]]$description, "\n\n", sep = "")
}
```

# Population

The three species were studied as follows (Methods 2.3.1-2.3.4). Rats
(`r uis$rat$population$n_subjects` Sprague-Dawley animals across the kinetic and
tissue-distribution studies) received 0.5, 1 and 2 mg/kg as a single IV bolus
plus a 1 mg/kg qd multiple-dose arm, with plasma sampled to 240 min. Six beagle
dogs received 0.25, 0.5 and 1 mg/kg as single IV boluses in a three-period
cross-over, plus a multiple-dose arm, with plasma sampled to 300 min. The human
data are **not original to this paper** -- Methods 2.3.4 states they were cited
from the paper's reference [7]: 36 healthy volunteers in three cohorts of 12
received 10, 20 or 40 mg as a 1 h IV infusion, single dose on day 0 and then
q12h for 7 days beginning at 48 h.

Two sampling details matter for the validation below: rats and dogs received an
IV **bolus** with the first sample at 5 min, whereas humans received a 1 h
**infusion**. That difference is the whole explanation for the Table 3
discrepancy discussed in the Errata.

```{r population}
str(uis$human$population)
```

# Source trace

Every `ini()` entry carries an in-file comment naming its source location; the
table below collects the model-level provenance in one place.

| Equation / parameter block | Source location |
|----------------------------|-----------------|
| Perfusion-limited tissue ODE | Equation (7), page 7 |
| Arterial blood ODE | Equation (8), page 7 |
| Venous blood ODE | Equation (9), page 7 |
| Lung ODE | Equation (10), page 7 |
| Adipose vascular sub-compartment ODE (`vp_adipose`, paper's C1) | Equation (11), page 7 |
| Adipose extravascular ODE (`adipose`, paper's C2) | Equation (12), page 7 |
| Interspecies PS scaling `PS_i = PS_rat * (W_i / W_rat)` | Equation (13), page 7 |
| Liver ODE (eliminating, receives splanchnic inflow) | Equation (14), page 7 |
| Kidney ODE (eliminating) | Equation (15), page 8 |
| Intestine ODE (eliminating) | Equation (16), page 8 |
| Organ volumes, blood flows, `Kt:pl`, `Rb`, `fu,p` (all 3 species) | Table 1 |
| Adipose `V1` / `V2` split (all 3 species) | Paper text below Equation (12) (PK-Sim 11.2) |
| Unbound whole-organ intrinsic clearances (all 3 species) | Table 2, `CL int, in vivo,u` column |
| Well-stirred organ clearances used as a check below | Results 3.1.3 |
| Predicted vs observed Cmax / AUC / CL / t1/2 | Table 3 |
| `PS` (adipose permeability-surface product) | **Not reported anywhere** -- back-solved from Figure 4J; see Errata |

# Structural checks

Before comparing against published exposures, two checks confirm the physiology
and the elimination algebra were transcribed correctly. Both are exact
arithmetic identities, so they either hold or they do not.

## Blood-flow mass balance

Every tissue flow must sum to the cardiac output printed on the Table 1 lung
row, and the printed total hepatic inflow must equal the hepatic artery plus the
three splanchnic organs that drain into the liver.

```{r mass-balance}
flow_check <- lapply(names(uis), function(sp) {
  th <- uis[[sp]]$theta
  tissue_sum <- sum(th[c("Qheart", "Qbrain", "Qmuscle", "Qadipose", "Qskin",
                         "Qkidney", "Qother", "Qhepart", "Qspleen", "Qstomach",
                         "Qintestine")])
  liver_in <- sum(th[c("Qhepart", "Qspleen", "Qstomach", "Qintestine")])
  data.frame(
    Species             = sp,
    `Sum of tissue flows` = tissue_sum,
    `Cardiac output (Table 1)` = unname(th["Qtotal"]),
    `Difference (%)`    = 100 * (tissue_sum - th["Qtotal"]) / th["Qtotal"],
    `Total hepatic inflow` = liver_in,
    check.names = FALSE
  )
}) |> dplyr::bind_rows()

knitr::kable(
  flow_check, digits = 3, row.names = FALSE,
  caption = paste(
    "Blood-flow mass balance. The rat closes exactly; the dog and human tables",
    "carry small rounding differences (0.13% and 0.006%) which are reproduced",
    "as printed rather than silently reconciled. Total hepatic inflow matches",
    "the Table 1 'Liver' blood-flow row exactly in all three species",
    "(rat 12.30, dog 323.33, human 1518.33 mL/min)."
  )
)
```

## Well-stirred organ clearances

Equations (14)-(16) drive elimination with the plasma-equivalent concentration
`C_t / K_t:pl`, so each eliminating organ behaves as a well-stirred compartment
with blood clearance `Q * fu,b * CLint / (Q + fu,b * CLint)`, where
`fu,b = fu,p / Rb`. Reproducing the paper's own Results 3.1.3 numbers from the
packaged `ini()` values is a direct test that the elimination term was encoded
correctly.

```{r well-stirred}
well_stirred <- function(ui, organ) {
  th <- ui$theta
  clint <- unname(th[paste0("CLint", organ)]) * unname(th["BW"])
  fub   <- unname(th["fup"]) / unname(th["Rb"])
  q <- switch(organ,
    Liver     = sum(th[c("Qhepart", "Qspleen", "Qstomach", "Qintestine")]),
    Kidney    = unname(th["Qkidney"]),
    Intestine = unname(th["Qintestine"])
  )
  unname(q * fub * clint / (q + fub * clint) / th["BW"])
}

# Results 3.1.3, mL/min/kg. Dog intestine is absent (no glucuronidation in DIMs).
published_cl <- tibble::tribble(
  ~Species, ~Organ,      ~Published,
  "rat",    "Liver",     46.6,
  "rat",    "Kidney",    24.4,
  "rat",    "Intestine", 3.90,
  "dog",    "Liver",     37.8,
  "dog",    "Kidney",    5.50,
  "human",  "Liver",     20.4,
  "human",  "Kidney",    6.14,
  "human",  "Intestine", 0.109
)

cl_check <- published_cl |>
  dplyr::mutate(
    Recomputed = mapply(function(s, o) well_stirred(uis[[s]], o), Species, Organ),
    `% diff`   = 100 * (Recomputed - Published) / Published
  )

knitr::kable(
  cl_check, digits = c(0, 0, 3, 3, 2),
  caption = paste(
    "Well-stirred organ blood clearances (mL/min/kg) recomputed from the",
    "packaged ini() values versus the values the paper reports in Results",
    "3.1.3. All eight agree to the printed precision."
  )
)
```

```{r well-stirred-gate}
stopifnot(nrow(cl_check) == 8L, max(abs(cl_check$`% diff`)) < 1)
```

# Simulation

The models are deterministic typical-value simulators: the paper never reports
the magnitude of the parameter variability behind its 5th-95th percentile bands,
so no IIV and no residual error are encoded. Every simulation below is therefore
a single typical subject per dose arm (`omega = NA` is *not* passed -- these
models declare no etas).

```{r sim-helpers}
solve_arm <- function(ui, dose_mg, times, infusion_min = 0, dose_times = 0) {
  rate_val <- if (infusion_min > 0) dose_mg / infusion_min else 0
  dosing <- data.frame(
    time = dose_times, amt = dose_mg, evid = 1L,
    cmt = "venous", rate = rate_val
  )
  obs <- data.frame(
    time = times, amt = NA_real_, evid = 0L,
    cmt = "venous", rate = 0            # ODE state name, never the observable "Cc"
  )
  ev <- dplyr::arrange(dplyr::bind_rows(dosing, obs), time, dplyr::desc(evid))
  rxode2::rxSolve(ui, ev, atol = 1e-10, rtol = 1e-10, returnType = "data.frame")
}

# A grid that resolves the post-bolus distribution phase finely enough for NCA.
grid_for <- function(tmax) {
  g <- c(
    seq(0, 5, by = 0.05), seq(5, 60, by = 0.25),
    seq(60, 600, by = 2), seq(600, max(tmax, 600), by = 10)
  )
  sort(unique(g[g <= tmax]))
}
```

```{r run-single-dose}
arms <- tibble::tribble(
  ~species, ~dose_label,  ~dose_mg,        ~infusion_min,
  "rat",    "0.5 mg/kg",  0.5   * 0.25,    0,
  "rat",    "1 mg/kg",    1     * 0.25,    0,
  "rat",    "2 mg/kg",    2     * 0.25,    0,
  "dog",    "0.25 mg/kg", 0.25  * 8.5,     0,
  "dog",    "0.5 mg/kg",  0.5   * 8.5,     0,
  "dog",    "1 mg/kg",    1     * 8.5,     0,
  "human",  "10 mg",      10,              60,
  "human",  "20 mg",      20,              60,
  "human",  "40 mg",      40,              60
)

TMAX <- 6000  # min; >= 15 terminal half-lives in every species
sims <- arms |>
  dplyr::rowwise() |>
  dplyr::group_map(function(a, ...) {
    solve_arm(uis[[a$species]], a$dose_mg, grid_for(TMAX), a$infusion_min) |>
      dplyr::mutate(species = a$species, dose_label = a$dose_label,
                    dose_mg = a$dose_mg)
  }) |>
  dplyr::bind_rows() |>
  dplyr::mutate(arm = paste(species, dose_label))
```

## Plasma concentration-time profiles (Figures 3, 5 and 6)

```{r fig-plasma, fig.height = 4, fig.cap = "Replicates Figure 3 (rats), Figure 5 (dogs) and Figure 6 (humans) of Zhu 2024: simulated typical-value plasma SPT-07A after single intravenous doses."}
sims |>
  dplyr::filter(!dplyr::near(time, 0), time <= 300, Cc > 1e-12) |>
  dplyr::mutate(species = factor(species, c("rat", "dog", "human"))) |>
  ggplot(aes(time, Cc * 1000, colour = dose_label)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~species, scales = "free_y") +
  scale_y_log10() +
  labs(x = "Time (min)", y = "Plasma SPT-07A (ng/mL)", colour = "Dose") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Rat tissue distribution (Figure 4)

Figure 4 of the paper shows plasma plus ten tissues in rats after 2 mg/kg. The
adipose panel (4J) is the one that carries information about `PS`: it is the
only tissue whose simulated profile *rises* to a peak rather than declining
monotonically, which is the signature of permeability-limited uptake.

```{r fig-tissues, fig.height = 6, fig.cap = "Replicates Figure 4B-K of Zhu 2024: simulated typical-value tissue concentrations in the rat after a 2 mg/kg IV bolus."}
rat2 <- solve_arm(uis$rat, 2 * 0.25, grid_for(300))
th_rat <- uis$rat$theta

tissue_conc <- rat2 |>
  dplyr::transmute(
    time,
    plasma    = Cc,
    heart     = heart     / (th_rat[["Vheart"]]     * 1e-3),
    liver     = liver     / (th_rat[["Vliver"]]     * 1e-3),
    spleen    = spleen    / (th_rat[["Vspleen"]]    * 1e-3),
    stomach   = stomach   / (th_rat[["Vstomach"]]   * 1e-3),
    brain     = brain     / (th_rat[["Vbrain"]]     * 1e-3),
    intestine = intestine / (th_rat[["Vintestine"]] * 1e-3),
    muscle    = muscle    / (th_rat[["Vmuscle"]]    * 1e-3),
    lung      = lung      / (th_rat[["Vlung"]]      * 1e-3),
    adipose   = adipose   / (th_rat[["Vadipose2"]]  * 1e-3),
    kidney    = kidney    / (th_rat[["Vkidney"]]    * 1e-3)
  ) |>
  tidyr::pivot_longer(-time, names_to = "tissue", values_to = "conc") |>
  dplyr::filter(!dplyr::near(time, 0), conc > 1e-12)

ggplot(tissue_conc, aes(time, conc * 1000)) +
  geom_line(linewidth = 0.6) +
  facet_wrap(~tissue, scales = "free_y", ncol = 4) +
  scale_y_log10() +
  labs(x = "Time (min)", y = "Concentration (ng/mL or ng/g)") +
  theme_bw()
```

### Recovering `PS` from Figure 4J

`PS` is the single parameter in the whole model that was fitted to in vivo data,
and its value is not reported. It was recovered by digitising the median line of
Figure 4J and solving for the `PS` that reproduces it; the optimiser returns
0.499 mL/min, i.e. the authors' round 0.5. The digitised points are reproduced
below with their provenance so a reviewer can audit the back-solve.

```{r ps-check, fig.height = 3.5, fig.cap = "Simulated rat adipose (extravascular) concentration at the back-solved PS = 0.5 mL/min versus the digitised median line of Figure 4J."}
# Digitised from Figure 4J (rat adipose, 2 mg/kg IV bolus), median line.
fig4j <- tibble::tibble(
  time = c(5, 30, 60, 90),
  conc = c(430, 410, 278, 205)     # ng/g
)

adipose_pred <- tissue_conc |>
  dplyr::filter(tissue == "adipose") |>
  dplyr::mutate(conc_ng = conc * 1000)

ps_cmp <- fig4j |>
  dplyr::mutate(
    Simulated = approx(adipose_pred$time, adipose_pred$conc_ng, xout = time)$y,
    `% diff`  = 100 * (Simulated - conc) / conc
  ) |>
  dplyr::rename("Time (min)" = time, "Digitised Fig 4J (ng/g)" = conc)

knitr::kable(ps_cmp, digits = c(0, 0, 1, 1),
             caption = "Back-solved PS = 0.5 mL/min against the digitised Figure 4J median.")

ggplot(adipose_pred, aes(time, conc_ng)) +
  geom_line(linewidth = 0.7) +
  geom_point(data = fig4j, aes(time, conc), colour = "firebrick", size = 2.5) +
  labs(x = "Time (min)", y = "Rat adipose SPT-07A (ng/g)",
       subtitle = "Line: packaged model. Points: digitised Figure 4J median.") +
  theme_bw()
```

```{r ps-gate}
# The back-solve is only credible if it reproduces the digitised curve closely.
stopifnot(max(abs(ps_cmp$`% diff`)) < 15)
```

# PKNCA validation against Table 3

```{r pknca}
nca_input <- sims |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(id = arm)

dose_input <- arms |>
  dplyr::mutate(id = paste(species, dose_label), time = 0) |>
  dplyr::select(id, time, dose_mg, species, dose_label)

conc_obj <- PKNCA::PKNCAconc(nca_input, Cc ~ time | id)
dose_obj <- PKNCA::PKNCAdose(as.data.frame(dose_input), dose_mg ~ time | id)

intervals <- data.frame(
  start = 0, end = Inf,
  cmax = TRUE, aucinf.obs = TRUE
)

res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))

# PKNCA also returns the DEPENDENCIES of the requested parameters -- notably its
# own `half.life` / `lambda.z`, which aucinf.obs needs. Those extra rows must be
# dropped: a windowed half-life is appended below, and leaving PKNCA's
# asymptotic one in place would silently produce two half.life rows per arm,
# which ncaComparisonTable() would then average.
nca <- as.data.frame(res) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "aucinf.obs")) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  dplyr::left_join(dose_input, by = "id")
```

Terminal half-life is deliberately **not** taken from a 0-Inf PKNCA interval.
This model has a slowly-equilibrating adipose compartment, so its true
asymptotic terminal slope (rat 60, dog 139, human 217 min) is much shallower
than anything a finite study could observe; comparing that against a published
half-life would compare two different quantities.

Instead one uniform rule is applied to all three species -- fit the terminal
slope over the **final half of that species' sampling window** -- and the
resulting windows are stated explicitly. Rats were sampled to 240 min and dogs
to 300 min (Methods 2.3.1 and 2.3.3). The clinical sampling schedule is not
given in this paper (the human data are cited from reference [7]), so a 12 h
window is assumed, consistent with the q12h multiple-dose regimen. The rule is
fixed in advance and applied identically to every species rather than tuned per
arm.

```{r half-life}
# end of the sampling window (min); the slope is fitted over the final half
hl_end   <- c(rat = 240, dog = 300, human = 720)
hl_start <- hl_end / 2

half_life_windowed <- sims |>
  dplyr::group_by(species, dose_label) |>
  dplyr::group_modify(function(d, k) {
    sp <- k$species
    w <- d[d$time >= hl_start[[sp]] & d$time <= hl_end[[sp]] & d$Cc > 0, ]
    data.frame(PPTESTCD = "half.life",
               PPORRES  = unname(log(2) / -coef(lm(log(w$Cc) ~ w$time))[2]))
  }) |>
  dplyr::ungroup()

nca <- dplyr::bind_rows(nca, half_life_windowed)

# Guard against the duplicate-row trap above: ncaComparisonTable() takes the
# MEDIAN over repeated (group, parameter) rows, so a stray duplicate would be
# silently averaged rather than reported or flagged.
stopifnot(!any(duplicated(nca[, c("species", "dose_label", "PPTESTCD")])))
```

Table 3 reports predicted Cmax (humans only -- the animal arms were IV boluses,
for which the paper tabulates no Cmax), AUC and terminal half-life. The
comparison below uses the paper's **predicted** column, since reproducing the
authors' own simulation is what validates the packaged implementation.

```{r nca-comparison}
# Zhu 2024 Table 3, "Pre" (predicted) columns.
# AUC in ug*min/mL == mg*min/L, which is the unit of Cc * time here.
published <- tibble::tribble(
  ~species, ~dose_label,  ~cmax,     ~aucinf.obs, ~half.life,
  "rat",    "0.5 mg/kg",  NA_real_,  3.78,        NA_real_,
  "rat",    "1 mg/kg",    NA_real_,  7.56,        58.2,
  "rat",    "2 mg/kg",    NA_real_,  15.1,        58.2,
  "dog",    "0.25 mg/kg", NA_real_,  3.60,        NA_real_,
  "dog",    "0.5 mg/kg",  NA_real_,  7.19,        76.7,
  "dog",    "1 mg/kg",    NA_real_,  14.4,        NA_real_,
  "human",  "10 mg",      0.0701,    5.76,        179,
  "human",  "20 mg",      0.1400,    11.5,        179,
  "human",  "40 mg",      0.2810,    23.0,        179
) |>
  tidyr::pivot_longer(c(cmax, aucinf.obs, half.life),
                      names_to = "PPTESTCD", values_to = "PPORRES") |>
  dplyr::filter(!is.na(PPORRES))

cmp <- nlmixr2lib::ncaComparisonTable(
  simulated     = dplyr::select(nca, species, dose_label, PPTESTCD, PPORRES),
  reference     = published,
  by            = c("species", "dose_label"),
  units         = c(cmax = "mg/L", aucinf.obs = "mg*min/L", half.life = "min"),
  tolerance_pct = 20
)

cmp |>
  dplyr::rename("Species" = species, "Dose" = dose_label) |>
  knitr::kable(
    caption = paste(
      "Simulated NCA from the packaged models versus the PREDICTED column of",
      "Zhu 2024 Table 3. * marks a difference above 20%. The human arms -- the",
      "only ones dosed as an infusion -- agree throughout; every starred row is",
      "a rat or dog AUC, for the reason set out in the Errata."
    ),
    align = c("l", "l", "l", "r", "r", "r")
  )
```

The human arms are the decisive test, because they are a pure bottom-up
prediction: no human in vivo data entered the model at any point.

```{r human-gate}
human_rows <- cmp[cmp$species == "human", ]
pct <- suppressWarnings(as.numeric(gsub("[^0-9.+-]", "", human_rows$`% diff`)))
# Tightened to the accuracy actually achieved (worst human row ~5%), so this
# gate fails on a regression rather than merely on a gross error.
stopifnot(nrow(human_rows) == 9L, !anyNA(pct), max(abs(pct)) < 8)
```

## Dose proportionality

The model contains no saturable process, so exposure must be exactly
dose-proportional within each species. This is a structural self-check rather
than a comparison against the paper.

```{r dose-prop}
nca |>
  dplyr::filter(PPTESTCD == "aucinf.obs") |>
  dplyr::group_by(species) |>
  dplyr::summarise(
    `CL (mL/min/kg)` = paste(round(sort(unique(
      round(dose_mg / PPORRES / c(rat = 0.25, dog = 8.5, human = 70)[species[1]] * 1000, 2)
    ))), collapse = ", "),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Clearance is identical across dose levels within each species, as a linear model requires.")
```

## Multiple dosing (human, q12h)

For a linear system the steady-state AUC over a dosing interval equals the
single-dose AUC to infinity. Table 3's human multiple-dose predictions (5.87,
11.7 and 23.5) are indeed within ~2% of its single-dose predictions (5.76, 11.5
and 23.0), which is consistent with that identity; the simulation reproduces it.

```{r multiple-dose}
md <- solve_arm(
  uis$human, 20, times = sort(unique(c(seq(0, 9 * 1440, by = 5)))),
  infusion_min = 60,
  dose_times = c(0, 2880 + seq(0, 13) * 720)   # single dose, then q12h x 7 days from 48 h
)

tau_window <- md |> dplyr::filter(time >= 2880 + 12 * 720, time <= 2880 + 13 * 720)
auc_tau <- with(tau_window, sum(diff(time) * (head(Cc, -1) + tail(Cc, -1)) / 2))
auc_sd  <- nca$PPORRES[nca$id == "human 20 mg" & nca$PPTESTCD == "aucinf.obs"]

knitr::kable(
  data.frame(
    Quantity = c("Steady-state AUC over tau (12 h), 20 mg q12h",
                 "Single-dose AUC to infinity, 20 mg",
                 "Zhu 2024 Table 3 predicted, 20 mg multiple dose",
                 "Zhu 2024 Table 3 predicted, 20 mg single dose"),
    `Value (mg*min/L)` = round(c(auc_tau, auc_sd, 11.7, 11.5), 2),
    check.names = FALSE
  ),
  caption = "Linear-system identity: steady-state AUC over a dosing interval equals the single-dose AUC to infinity."
)
```

```{r md-plot, fig.height = 3.5, fig.cap = "Replicates the human multiple-dose panel of Figure 6: 20 mg 1 h infusion, single dose on day 0 then q12h from 48 h."}
md |>
  dplyr::filter(!dplyr::near(time, 0), Cc > 1e-12) |>
  ggplot(aes(time / 60, Cc * 1000)) +
  geom_line(linewidth = 0.5) +
  labs(x = "Time (h)", y = "Plasma SPT-07A (ng/mL)") +
  theme_bw()
```

```{r md-gate}
stopifnot(abs(auc_tau - auc_sd) / auc_sd < 0.02)
```

# Assumptions and deviations

## Errata and source inconsistencies

**1. `PS` is not reported anywhere and was back-solved from a figure.** Methods
2.5.1 states only that "PS is the permeability-surface product, which was
estimated by fitting this value to the adipose concentration data in rats." The
fitted number appears in no table, figure caption or text, and the article has
no supplementary material. It was recovered here by digitising the median line
of Figure 4J and solving for the `PS` that reproduces it; the optimiser returns
0.499 mL/min against the authors' evident round value of 0.5, and at `PS = 0.5`
the model matches the digitised curve to within about 5% (table above). Because
`PS` governs distribution and not elimination, **AUC and clearance are
mathematically independent of it** -- the back-solve affects the adipose
profile and the terminal phase only, and none of the AUC or clearance validation
above depends on it. The dog and human values (17.0 and 140 mL/min) follow from
Equation (13), which as printed carries **no exponent**: the scaling is linear
in body weight, not allometric.

**2. Table 3's rat and dog predicted AUC cannot be reconciled with the paper's
own equations, and the profiles show the equations are right.** The packaged
implementation reproduces the human Table 3 predictions to within about 2% on
Cmax, AUC and clearance at all three doses, and reproduces all eight of the
paper's Results 3.1.3 organ clearances exactly. For rats and dogs, however, the
simulated AUC to infinity is a constant 1.93-fold higher than the Table 3
predicted column at *every* dose level, even though the predicted rat terminal
half-life matches (58.7 min simulated versus 58.2 reported). Two further
observations locate the problem in the Table 3 summary rather than in the model:

* The paper's own stated systemic clearance for the rat -- 75.0 mL/min/kg,
  Results 3.1.3 -- agrees with the implemented equations, whereas Table 3's
  predicted rat clearance of 129-132 mL/min/kg does not. The two numbers appear
  in the same paper and are mutually inconsistent, since AUC = Dose / CL.
* Digitising the rat plasma profile of Figure 4A gives 107, 51.3 and 29.8 ng/mL
  at 30, 60 and 90 min, against 110, 49.6 and 30.0 ng/mL from the packaged
  model. The *figures* therefore agree with the printed equations; only the
  Table 3 AUC/CL column for the animal species does not.

The most likely mechanism is the sampling window. Rats and dogs received an IV
**bolus** with the first sample at 5 min, so an NCA that starts at the first
observed concentration misses the steep 0-5 min distribution phase entirely;
humans received a 1 h **infusion**, which has no such spike, and their numbers
reconcile. Consistent with this, the paper's rat and dog predicted AUC values
sit between the model's AUC to infinity and its AUC computed from 5 min onward.
Following the standing policy that a printed equation outranks a conflicting
summary, the equations as printed are what is implemented; the rat and dog AUC
rows of the comparison table above are starred for this reason and should not be
read as an implementation defect.

**3. Dog multiple-dose regimen is stated inconsistently.** Methods 2.3.3 says
the dogs received 1 mg/kg qd for 7 days; Table 3 labels the dog multiple-dose row
0.5 mg/kg and reports the same predicted AUC (7.19) as its 0.5 mg/kg single-dose
row. The Table 3 label is self-consistent with the reported exposure and is the
one recorded in the model's `population` metadata.

**4. Blood-flow rounding.** The dog tissue flows sum to 1121.46 mL/min against a
printed cardiac output of 1120 (0.13%), and the human flows to 5600.33 against
5600 (0.006%). Both are reproduced as printed rather than reconciled, so the
packaged values match Table 1 byte for byte. The rat closes exactly.

## Assumptions

* **No IIV and no residual error.** The paper's 5th-95th percentile bands come
  from 100 virtual subjects, but the magnitude and distribution of the
  underlying parameter variability are never reported. Rather than invent
  variances, both are omitted: `propSd` is `fixed(0)` and no etas are declared.
  The packaged models are therefore deterministic typical-value simulators and
  cannot reproduce the prediction intervals in Figures 3-6, only the median
  lines.
* **Terminal half-life is window-dependent, and the window is assumed.** The
  permeability-limited adipose compartment makes the model's asymptotic terminal
  slope (rat 60, dog 139, human 217 min) much shallower than any observable
  half-life, so the comparison uses the final half of each species' sampling
  window. Rat and dog windows come from Methods 2.3.1 and 2.3.3; the human
  window is **assumed** to be 12 h, since this paper does not report the
  clinical sampling schedule. The same rule is applied to all three species
  without per-arm tuning, and it recovers the paper's predicted half-lives to
  within 5% in every species -- but a reader who assumes a different clinical
  window will get a different human number, and none of the AUC, Cmax or
  clearance conclusions depends on this choice.
* **Observation compartment.** The paper does not state which blood pool its
  simulated plasma concentration is drawn from. Venous plasma is used
  (`Cc <- c_venous / Rb`), matching the sampling site for the dog (forearm vein)
  and human (venous) studies; the rat study sampled the femoral artery.
  Arterial and venous concentrations differ materially only during the first few
  minutes after a bolus.
* **Blood versus plasma.** Because `Kt:pl` is a tissue-to-*plasma* ratio while
  the perfusion terms carry `Kt:pl / Rb`, the arterial, venous and adipose
  vascular states hold *blood* concentrations; the reported plasma concentration
  is obtained by dividing by `Rb`. This follows directly from Equations (7)-(10)
  and is what makes the recomputed organ clearances match Results 3.1.3.
* **Body weight is not a covariate.** Organ volumes and blood flows are the
  absolute Table 1 values for the reference animal (0.25 kg rat, 8.5 kg beagle,
  70 kg adult) and are not weight-scaled. `BW` appears only in the per-kilogram
  intrinsic-clearance terms of Equations (14)-(16). Simulating a different body
  weight requires rescaling the volumes and flows as well, which the paper does
  not parameterise.
* **`MPPGL` / `MPPGK` / `MPPGI` are not model parameters.** These
  microsomal-protein scalars appear in Table 1 but act upstream, converting the
  measured in vitro `CLint,u` into the whole-organ `CLint,in vivo,u` values of
  Table 2 that the ODEs actually use. Only the Table 2 outputs are packaged.
* **Compartment naming.** The paper's adipose sub-compartments C1 and C2 are
  mapped to the registered canonicals `vp_adipose` (vascular) and `adipose`
  (the tissue). The registered `is_adipose` was deliberately not used for C2,
  because the `is_` prefix denotes interstitial *fluid*, whereas for a small
  lipophilic monoterpene with `Kadipose:plasma` of 1.72-2.84 the extravascular
  space is predominantly intracellular lipid. The paper's "rest of body" is the
  registered `other`.
