---
title: "Pembrolizumab patient-reported quality of life (Zou 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pembrolizumab patient-reported quality of life (Zou 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model}
mod <- rxode2::rxode2(readModelDb("Zou_2026_pembrolizumab_qol_mbma"))
```

* Citation: `r mod$reference`
* Article: <https://doi.org/10.1002/psp4.70106>
* Supplement (Data S1: Table S1, Figure S1, and the Monolix control stream):
  <https://www.ebi.ac.uk/europepmc/webservices/rest/PMC12896370/supplementaryFiles>

* Description: `r mod$description`

This is a **model-based meta-analysis (MBMA)**, so every modelled data point is
a study-**arm mean**, not a patient. The model simulates arm-mean quality-of-life
trajectories; it cannot simulate an individual patient's QoL.

# Population

```{r population, echo = FALSE}
pop <- mod$meta$population
tibble::tibble(Field = names(pop), Value = vapply(pop, function(x) paste(format(x), collapse = "; "), character(1))) |>
  knitr::kable()
```

Zou 2026 screened 302 published studies across 7 tumour types and retained the
20 Table 1 entries that reported **longitudinal** EORTC QLQ-C30 Global Health
Status / Quality of Life (GHS/QoL) scores with at least three measurements per
arm. Arm-level values were digitised from published figures and tables with
WebPlotDigitizer. Of the 16 entries reporting a between-group comparison, 8
reported a QoL improvement with pembrolizumab and 8 reported no significant or
clinically meaningful difference (Table 1, "Reported between-group difference").

# Source trace

Every value in `ini()` comes from **Table S1** of the supplement ("Final model
parameter estimates"); every equation comes either from the numbered equations
in Section 2 of the article or from the **Monolix 2024R1 control stream** printed
verbatim in Supplementary Codes section A. The control stream is authoritative
where it and the article's prose differ, because it is the code that produced
Table S1.

```{r source-trace, echo = FALSE}
tribble(
  ~Quantity, ~`Model file`, ~`Source location`, ~Value,
  "QoL trajectory", "qFrac", "Eq. 1; Suppl. Codes A line 'Q = E0 - Emax * (1 - exp(-Kp * t)) + SLP2*t'", "Q(t) = E0 - Emax(1 - e^-Kp t) + SLP t",
  "Logit data transform", "logit_qol_ghs", "Eq. 2 and Suppl. Codes A 'pred = logit(Q)*sqrt(NOC)'", "trans(y) = sqrt(N) logit(y/100)",
  "Residual error", "addSd / sqrt(N_ARM)", "Eq. 3; Table S1 'Additive residual error (a)'", "eps ~ N(0, sigma^2 / N); a = 1.1",
  "E0 random effects", "eta_study_e0, eta_arm_e0", "Eq. 4; Suppl. Codes A 'tE0RE = tE0 + etaBSVE0 + etaBTAVE0/sqrt(NOC)'", "logit(E0) + BSV + BTAV/sqrt(N)",
  "Treatment covariate", "e_trt_emax, e_trt_slp", "Eq. 5; Suppl. Codes B [INDIVIDUAL] coefficient={0, beta_*_TRT_Placebo}", "exponential on the parameter's own scale",
  "Baseline QoL E0", "logite0", "Table S1 'Baseline QoL score (E0)'", "0.656 (RSE 5.16%)",
  "Maximal toxicity Emax", "logitemax", "Table S1 'Maximal toxicity reducing QoL (Emax)'", "0.0268 (RSE 36.2%)",
  "Toxicity rate Kp", "lkel", "Table S1 'Toxicity offset rate (Kp, 1/week)'", "0.0705 /week (RSE 30.4%)",
  "Improvement rate SLP", "lslp", "Table S1 'QoL improvement rate (SLP, 1/week)'", "9.98e-4 /week (RSE 24.1%)",
  "Control effect on Emax", "e_trt_emax", "Table S1 'Effect of control arm on Emax'", "-0.758 (RSE 32.6%)",
  "Control effect on SLP", "e_trt_slp", "Table S1 'Effect of control arm on SLP'", "-1.4 (RSE 32.2%)",
  "BSV on Emax", "eta_study_emax", "Table S1 'Standard deviation of BSVEmax'", "SD 0.783 (shrinkage 17.3%)",
  "BSV on E0", "eta_study_e0", "Table S1 'Standard deviation of BSVE0'", "SD 0.667 (shrinkage 29.9%)",
  "BTAV on E0", "eta_arm_e0", "Table S1 'Standard deviation of BTAE0'", "SD 0.867 (shrinkage 20%)",
  "BTAV on SLP", "eta_arm_slp", "Table S1 'Standard deviation of SLP'", "SD 0.871 (shrinkage 25.8%)"
) |>
  knitr::kable()
```

The packaged `ini()` values are checked against Table S1 directly here, so a
mis-transcription cannot pass silently.

```{r check-ini}
theta <- mod$theta
published <- c(
  logite0    = log(0.656 / (1 - 0.656)),
  logitemax  = log(0.0268 / (1 - 0.0268)),
  lkel       = log(0.0705),
  lslp       = log(9.98e-4),
  e_trt_emax = -0.758,
  e_trt_slp  = -1.4,
  addSd      = 1.1
)
stopifnot(
  setequal(names(theta), names(published)),
  isTRUE(all.equal(theta[names(published)], published, tolerance = 1e-12))
)

omega_sd <- sqrt(diag(mod$omega))
stopifnot(isTRUE(all.equal(
  omega_sd[c("eta_study_e0", "eta_study_emax", "eta_arm_e0", "eta_arm_slp")],
  c(eta_study_e0 = 0.667, eta_study_emax = 0.783, eta_arm_e0 = 0.867, eta_arm_slp = 0.871),
  tolerance = 1e-12
)))
```

## The SLP scale, and why the file drops the 0.001 factor

Table S1 reports SLP as `9.98e-4` with the unit `1/week`, while its
Bootstrap-median cell for the same row reads `0.94`. These are not
inconsistent: the Monolix control stream carries a bookkeeping rescale,
`SLP2 = SLP*0.001`, so the estimated `SLP_pop` is ~0.998 and the rate that
actually multiplies time is `9.98e-4`. The bootstrap-median cell was filled in
on the unscaled parameter. The quoted bootstrap 10th/90th interval settles it
arithmetically:

```{r slp-scale}
# Unscaled bootstrap median * 0.001 must fall inside the SCALED 10/90 interval.
stopifnot(0.94 * 0.001 > 5.44e-4, 0.94 * 0.001 < 1.498e-3)
```

The model file works directly on the `1/week` scale. A constant factor shifts
`log(SLP)` without changing its log-scale standard deviation or its log-scale
covariate coefficient, so `eta_arm_slp` and `e_trt_slp` are untouched by the
change of scale.

# Structural model

On the 0-1 fraction scale, for study *i* and arm *k* with `N` participants:

$$\mathrm{logit}(E_{0,ik}) = \mathrm{logit}(E_0) + \eta^{BSV}_{E_0,i} + \eta^{BTAV}_{E_0,ik}/\sqrt{N}$$
$$\mathrm{logit}(E_{max,ik}) = \mathrm{logit}(E_{max}) + \theta_{Emax,ctrl}\,\mathrm{TRT} + \eta^{BSV}_{Emax,i}$$
$$\log(SLP_{ik}) = \log(SLP) + \theta_{SLP,ctrl}\,\mathrm{TRT} + \eta^{BTAV}_{SLP,ik}$$
$$Q(t) = E_{0,ik} - E_{max,ik}\left(1 - e^{-K_p t}\right) + SLP_{ik}\,t$$

with the reported score `100 * Q(t)` and the observation
`logit(Q(t))` carrying an additive residual of SD `a / sqrt(N)`.

Two features are specific to the meta-analytic setting and are why `N_ARM` is a
model **input** rather than a post-solve rescaling:

* the **between-arm** random effect on baseline is divided by `sqrt(N)` (an arm
  mean's standard error shrinks with arm size), and
* the **residual** SD is divided by `sqrt(N)` (Eq. 3's `sigma^2 / N`).

A per-arm random-effect SD cannot be varied across arms within one `rxSolve()`
without a data column, so the library's usual MBMA pattern of applying
`1/sqrt(N)` downstream of the solve cannot reproduce the first of these.

# Validation

## Closed-form gate

The model is algebraic, so it can be checked against an independent R
implementation of Eq. 1 evaluated at the **same drawn random effects**. Both
sides use identical parameters, so the only difference is numerical: a tight
bound is the correct assertion here.

```{r closed-form}
rxode2::rxSetSeed(20260911)

n_arm_per_grp <- 200L
grid_t <- seq(0, 104, by = 4)

arms <- bind_rows(
  tibble(arm = seq_len(n_arm_per_grp), TRT = 0),
  tibble(arm = seq_len(n_arm_per_grp) + n_arm_per_grp, TRT = 1)
) |>
  mutate(N_ARM = rep(c(50, 150, 400, 900), length.out = n()))

ev <- arms |>
  tidyr::expand_grid(time = grid_t) |>
  mutate(id = arm)

sim <- rxode2::rxSolve(mod, ev)

# The drawn random effects live in `$params`, keyed by id; join them back so the
# closed-form check can be evaluated at exactly the values the solver used.
etas <- as.data.frame(sim$params)[, c("id", "eta_study_e0", "eta_study_emax",
                                      "eta_arm_e0", "eta_arm_slp")]
# rxSolve returns `id` as a factor in $params but as an integer in the solve
# itself; align them or the join errors on incompatible key types.
etas$id <- as.integer(as.character(etas$id))

closed <- as.data.frame(sim) |>
  mutate(id = as.integer(id)) |>
  left_join(etas, by = "id") |>
  mutate(
    q_ref   = e0Arm - emaxArm * (1 - exp(-kel * time)) + slp * time,
    sat_ref = pmin(pmax(q_ref, 0.01), 0.99)
  )

# Guard against a vacuous pass: every column the assertions below touch must be
# present and fully populated, or a max() over nothing silently returns -Inf.
stopifnot(
  nrow(closed) == nrow(arms) * length(grid_t),
  all(c("eta_study_e0", "eta_study_emax", "eta_arm_e0", "eta_arm_slp",
        "qFrac", "qol_ghs", "btavE0") %in% names(closed)),
  !anyNA(closed[, c("eta_study_e0", "eta_study_emax", "eta_arm_e0",
                    "eta_arm_slp", "qFrac", "qol_ghs", "btavE0")])
)

stopifnot(
  max(abs(closed$qFrac - closed$q_ref)) < 1e-10,
  max(abs(closed$qol_ghs - 100 * closed$sat_ref)) < 1e-8
)
max(abs(closed$qFrac - closed$q_ref))
```

The per-arm parameters themselves must equal the published fixed effects shifted
by that arm's drawn etas and its treatment assignment:

```{r param-recovery}
stopifnot(
  max(abs(closed$e0Arm - plogis(published[["logite0"]] + closed$eta_study_e0 +
                                  closed$eta_arm_e0 / sqrt(closed$N_ARM)))) < 1e-12,
  max(abs(closed$emaxArm - plogis(published[["logitemax"]] +
                                    published[["e_trt_emax"]] * closed$TRT +
                                    closed$eta_study_emax))) < 1e-12,
  max(abs(closed$slp - exp(published[["lslp"]] +
                             published[["e_trt_slp"]] * closed$TRT +
                             closed$eta_arm_slp))) < 1e-12,
  max(abs(closed$kel - 0.0705)) < 1e-12
)
```

## Arm-size weighting gate

Both `1/sqrt(N)` terms must be exactly that, across the whole range of arm sizes
in the Zou 2026 database (26 to 1098 participants).

```{r weighting}
w <- closed |>
  distinct(N_ARM, addSdArm) |>
  arrange(N_ARM) |>
  mutate(expected = 1.1 / sqrt(N_ARM))
stopifnot(max(abs(w$addSdArm - w$expected)) < 1e-12)
w |> rename("Arm size (N_ARM)" = N_ARM, "Residual SD" = addSdArm, "1.1 / sqrt(N)" = expected) |>
  knitr::kable(digits = 5)

# The between-arm baseline term must also shrink as 1/sqrt(N).
stopifnot(max(abs(closed$btavE0 - closed$eta_arm_e0 / sqrt(closed$N_ARM))) < 1e-12)
```

## Bounded-score gate

The logit transform exists to keep a predicted arm mean inside the instrument's
0-100 range. That must hold for every simulated arm, including the tails.

```{r bounded}
stopifnot(
  all(closed$qol_ghs > 0), all(closed$qol_ghs < 100),
  !anyNA(closed$logit_qol_ghs),
  # observations carry the residual; back-transforming must also stay in range
  all(100 * plogis(closed$sim) > 0), all(100 * plogis(closed$sim) < 100)
)
range(closed$qol_ghs)
```

## Treatment-effect gate (Zou 2026 Section 3.3)

The paper's two headline findings are that pembrolizumab carries **greater**
maximal toxicity and a **faster** long-term improvement. Both follow
deterministically from the sign of the two retained covariate effects, so they
can be asserted exactly on the typical-value parameters rather than on a
simulated cohort.

```{r treatment-effect}
tv <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  data.frame(id = 1:2, time = 0, TRT = 0:1, N_ARM = 300),
  returnType = "data.frame"
)
eff <- tv |>
  transmute(
    Arm = ifelse(TRT == 0, "Pembrolizumab", "Control"),
    `Maximal toxicity (GHS/QoL points)` = 100 * emaxArm,
    `Improvement rate (points/week)`    = 100 * slp
  )
eff |> knitr::kable(digits = 4)

pem <- eff[eff$Arm == "Pembrolizumab", ]
ctl <- eff[eff$Arm == "Control", ]
stopifnot(
  # Greater initial toxicity under pembrolizumab (paper: p = 0.005)
  pem[[2]] > ctl[[2]],
  # Faster long-term improvement under pembrolizumab (paper: p < 0.0001)
  pem[[3]] > ctl[[3]],
  # exp(-1.4) = 0.247: the control arm improves at about a quarter the rate
  abs(ctl[[3]] / pem[[3]] - exp(-1.4)) < 1e-10
)
```

## Replication of Figure 4A

Zou 2026 Figure 4A simulates arm-mean QoL trajectories for pembrolizumab versus
control. The typical-value trajectories below reproduce its two qualitative
features: pembrolizumab sits **below** control over the first weeks, then
crosses over into a sustained and widening advantage.

Note what the crossover does *not* say. Pembrolizumab carries the larger
toxicity parameter (`Emax` 2.68 vs 1.28 QoL points), but its **net** trajectory
does not dip lower than control's, because its four-fold larger `SLP` pulls the
curve back up while the toxicity term is still accumulating. This is exactly the
reading the paper warns against making from the parameters alone: "we caution
against interpreting their individual magnitudes in isolation. Due to their
opposing effects, the net impact of treatment on QoL is best assessed through
the overall predicted trajectory" (Discussion). The `Emax` contrast is asserted
on the parameter above; the assertions here are on the trajectory.

```{r fig4a, fig.cap = "Replicates Figure 4A of Zou 2026: typical-value arm-mean QoL trajectories."}
tv_traj <- rxode2::rxSolve(
  rxode2::zeroRe(mod),
  tidyr::expand_grid(TRT = 0:1, time = seq(0, 104, by = 1)) |>
    mutate(id = TRT + 1L, N_ARM = 300),
  returnType = "data.frame"
) |>
  mutate(Arm = ifelse(TRT == 0, "Pembrolizumab", "Control"))

ggplot(tv_traj, aes(time, qol_ghs, colour = Arm)) +
  geom_line(linewidth = 1) +
  labs(x = "Time (weeks)", y = "Arm-mean EORTC QLQ-C30 GHS/QoL score",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Nadir and long-term behaviour, quantified:

```{r fig4a-summary}
nadir <- tv_traj |>
  group_by(Arm) |>
  slice_min(qol_ghs, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(Arm, `Nadir week` = time, `Nadir score` = qol_ghs)
wk104 <- tv_traj |> filter(time == 104) |> select(Arm, `Week-104 score` = qol_ghs)
left_join(nadir, wk104, by = "Arm") |> knitr::kable(digits = 2)

pem_traj <- tv_traj |> filter(Arm == "Pembrolizumab") |> arrange(time)
ctl_traj <- tv_traj |> filter(Arm == "Control") |> arrange(time)
delta <- pem_traj$qol_ghs - ctl_traj$qol_ghs   # pembrolizumab minus control

stopifnot(
  identical(pem_traj$time, ctl_traj$time),
  # Both arms start at the same baseline: treatment acts on Emax and SLP only.
  abs(delta[pem_traj$time == 0]) < 1e-10,
  abs(pem_traj$qol_ghs[1] - 65.6) < 1e-8,
  # Early on, the larger pembrolizumab toxicity term dominates and the
  # pembrolizumab arm sits BELOW control.
  delta[pem_traj$time == 4] < 0,
  # A single crossover occurs, after which pembrolizumab stays above control
  # and the gap widens monotonically to the end of the simulated horizon.
  any(delta[pem_traj$time > 0] < 0), any(delta > 0),
  all(diff(delta[pem_traj$time >= 12]) > 0),
  tail(delta, 1) > 0
)
# Week at which pembrolizumab overtakes control.
pem_traj$time[which(delta > 0)[1]]
```

## Simulated arm-level cohort

With both variability levels switched on, the spread across simulated arms is
wide -- which is the paper's central methodological point: arm-level QoL data
are so variable that a baseline-to-endpoint hypothesis test loses the signal
that the longitudinal model recovers.

```{r cohort, fig.cap = "Simulated arm-mean QoL trajectories (200 arms per treatment group), median and 10th/90th percentiles."}
band <- closed |>
  mutate(Arm = ifelse(TRT == 0, "Pembrolizumab", "Control")) |>
  group_by(Arm, time) |>
  summarise(
    med = median(qol_ghs),
    lo  = quantile(qol_ghs, 0.1),
    hi  = quantile(qol_ghs, 0.9),
    .groups = "drop"
  )

ggplot(band, aes(time, med, colour = Arm, fill = Arm)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  labs(x = "Time (weeks)", y = "Arm-mean EORTC QLQ-C30 GHS/QoL score", colour = NULL, fill = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The cohort-level assertion is deliberately made on the **median across arms**,
not on any extreme: the tails of a random cohort are not reproducible across
rxode2 versions, but the centre of a 200-arm distribution is. Even the median
carries sampling noise, so the bounds below were set from measurement rather
than from a single run -- across 12 seeds the baseline median gap ranged
-0.3 to +3.0 points (it is pure noise; treatment does not act on `E0` at all)
and the week-104 gap ranged +5.3 to +11.8 points. Each bound leaves at least
two-fold headroom over the observed range, so a different rxode2 RNG stream
cannot flip it.

```{r cohort-gate}
med_wide <- band |>
  select(Arm, time, med) |>
  tidyr::pivot_wider(names_from = Arm, values_from = med)
base_gap  <- med_wide$Pembrolizumab[med_wide$time == 0] -
  med_wide$Control[med_wide$time == 0]
wk104_gap <- med_wide$Pembrolizumab[med_wide$time == 104] -
  med_wide$Control[med_wide$time == 104]

stopifnot(
  # Baseline medians agree to within cohort sampling noise: treatment acts on
  # Emax and SLP only, never on E0. (Observed |gap| <= 3.0 over 12 seeds.)
  abs(base_gap) < 6,
  # By week 104 the median pembrolizumab arm is clearly above the median
  # control arm. (Observed gap >= 5.3 over 12 seeds.)
  wk104_gap > 2
)
c(`baseline gap` = base_gap, `week-104 gap` = wk104_gap)
```

# Assumptions and deviations

* **Saturation guard.** Supplementary Codes section A computes
  `Qsat = min(max(Q,0.01),0.99)` with the comment "adding a saturation to avoid
  taking logit(0) (undefined)", but its very next line takes the logit of the
  **unsaturated** `Q`. This file takes the logit of `Qsat`, which is what the
  guard's own comment says it is for. The deviation is a no-op over the
  typical-value trajectory and a necessity in a simulated cohort: `Q` is
  unbounded above, so a high-baseline, high-slope arm carries it past 1 at long
  times and `logit(Q)` returns `NaN`. In the 10,800-row cohort simulated above
  roughly 2% of rows are affected. The check below shows both facts.

```{r saturation}
# Typical values: the guard never binds, so the deviation changes nothing.
stopifnot(max(abs(tv_traj$qFrac - tv_traj$qSat)) == 0)

# Simulated cohort: the guard does bind. The clamped set is a SUPERSET of the
# rows that would have gone non-finite -- the guard's ceiling is 0.99, not 1,
# so it also catches the thin band where logit(Q) is finite but the arm mean
# would print above 99 on a 0-100 instrument.
n_clamped      <- sum(abs(closed$qFrac - closed$qSat) > 0)
n_would_be_nan <- sum(closed$qFrac <= 0 | closed$qFrac >= 1)
c(rows = nrow(closed), clamped = n_clamped, would_be_NaN = n_would_be_nan,
  max_qFrac = round(max(closed$qFrac), 3))

stopifnot(
  # The problem is real, not hypothetical: some arms do leave (0, 1).
  n_would_be_nan > 0,
  # Every row that would have gone non-finite is covered by the guard.
  n_would_be_nan <= n_clamped,
  # The guard binds only outside [0.01, 0.99], and never leaves it.
  all(abs(closed$qFrac - closed$qSat)[closed$qFrac >= 0.01 & closed$qFrac <= 0.99] == 0),
  all(closed$qSat >= 0.01), all(closed$qSat <= 0.99)
)
```

* **Observation scale.** The paper's dependent variable is
  `sqrt(N) * logit(y/100)` (Eq. 2), fitted with a constant residual SD `a`
  (Monolix `errorModel=constant(a)`). This file instead observes
  `logit(y/100)` with residual SD `a / sqrt(N)`, which is the same statement
  written as the article's own Eq. 3 (`eps ~ N(0, sigma^2 / N)`). The two are
  algebraically identical; the Eq. 3 form is used so the observation stays on
  an interpretable scale and the arm-size weighting is visible in `model()`.

* **Random-effect naming.** `checkModelConventions()` warns that
  `eta_study_e0`, `eta_study_emax`, `eta_arm_e0` and `eta_arm_slp` do not pair
  with fixed effects named `_study_e0` etc. The names are intentional: this is a
  two-level MBMA and the `eta<param>` convention cannot distinguish
  between-study variability from between-treatment-arm variability on the *same*
  parameter, which this model has on `E0`. The `eta_study_*` form follows the
  existing library precedent in `Mercier_2014_tramadol_tapentadol_mbma` and
  `Chen_2025_methotrexate_das28_mbma`.

* **Not an exposure-response model.** No pembrolizumab dose or concentration
  enters the model. Pembrolizumab monotherapy and pembrolizumab-combination arms
  are pooled into a single `TRT = 0` level, and every comparator regimen
  (chemotherapy, targeted therapy, placebo plus standard of care, placebo alone)
  is pooled into `TRT = 1`. The model therefore cannot answer dose-optimisation
  questions, and a "control" effect here is an average over heterogeneous
  comparators.

* **Covariates screened but not retained.** Age, the arm's male proportion, the
  arm's ECOG-0 proportion, the arm's stage IV proportion and tumour type were
  all screened by the full fixed-effects covariate model (Section 2.5) and none
  was retained (Section 3.2). They are recorded in the model file's
  `covariatesDataExcluded` metadata so the provenance of the screen survives.
  The paper is explicit that the tumour-type null is a power limitation rather
  than evidence of no effect: "the dataset lacked sufficient power to compare
  pembrolizumab's QoL benefits across different tumor types" (Limitations).
  No point estimate is published for any screened covariate, so none can be
  encoded even optionally.

* **No PKNCA section.** NCA is not meaningful for this model: there is no drug
  concentration, no dose and no ODE state. The closed-form, weighting,
  bounded-score and treatment-effect gates above take its place, following
  `references/endogenous-validation.md` for models where PKNCA is not the right
  check.

* **`n_subjects` is derived, not published.** The paper reports 410
  observations (228 pembrolizumab, 182 control) but no participant total. The
  `population$n_subjects` value of 11326 is a lower bound: the sum of the
  Table 1 baseline arm sizes (12345) minus the 1019 participants of the second
  KEYNOTE-054 (NCT02362594) row, which appears twice (Bottomley 2021 and
  Buhrer 2024). Table 1 footnote a notes the two entries "were published with
  different population inclusion criteria and follow-up durations", so they
  overlap heavily without being the identical cohort, and the exact unique
  count cannot be recovered from what is published.

* **Informative dropout is not modelled.** The paper's own Limitations
  paragraph notes that QoL assessments after treatment discontinuation were
  typically missing and not imputed, so arm-level trajectories may
  over-represent patients with favourable outcomes at later timepoints and the
  long-term QoL benefit may be overestimated. Simulations at long times inherit
  that bias.

# Errata

No erratum or corrigendum for Zou 2026 was located. The only internal
inconsistency found is the mixed-scale SLP row of Table S1 documented above,
which is resolved arithmetically by the quoted bootstrap interval rather than by
assumption.
