---
title: "PSA disease progression under leuprorelin (Zou 2020)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PSA disease progression under leuprorelin (Zou 2020)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# The model

Zou et al. (2020) built a population disease-progression model of serum
prostate-specific antigen (PSA) in 264 men with hormone-sensitive prostate
cancer who were treated with the LHRH agonist leuprorelin, using a US medical
claims database (Humana, 2007-2011). Two resistance mechanisms were compared;
the final model is the paper's "clonal selection" Model II (AIC 295.7 against
865.4 for the "adaptation" Model I), which splits the observed baseline PSA
(`PSA_BL`) into

* a **drug-resistant** fraction `R = exp(-RP)` that grows first-order at `GR`
  from the baseline draw onwards (state `growth`), and
* a **drug-sensitive** fraction `1 - R` that grows first-order at `GS` until the
  first leuprorelin dose and is killed first-order at `DS` from then on (state
  `shrink`).

`PSA = growth + shrink`, so `PSA(0) = PSA_BL` exactly. Hemoglobin enters as a
power covariate on `RP`, baseline PSA as a power covariate on `DS`, and
antiandrogen use within 30 days of leuprorelin initiation as an exponential
covariate on `DS`.

This is a disease-progression model with no pharmacokinetics: leuprorelin
enters only through the per-subject time of its first dose,
`T_SCAN_TO_DOSE`, measured from the baseline PSA draw. There is therefore
**no PKNCA section** -- there is no drug concentration for non-compartmental
analysis to act on. Validation is structural (closed-form and covariate
identities) plus a reproduction of the paper's own simulated nadir, time to
nadir and PSA-progression rates.

```{r model-load}
mod_fun <- readModelDb("Zou_2020_leuprorelin")
mod <- rxode2::rxode2(mod_fun)
mod
```

# Population

```{r population}
str(mod$meta$population)
```

The 264 subjects (Table 1) had a median age of 80 years (60-100): 189
Caucasian, 59 Black and 16 Hispanic/other, mostly from the US South (196).
Median baseline PSA was 8.50 ng/mL (0.200-782) and median hemoglobin 13.6 g/dL
(6.80-17.4); 33 subjects received an antiandrogen within 30 days of starting
leuprorelin. Patients with a falling pre-treatment PSA, an undetectable baseline
or on-treatment PSA, incomplete demographics or hemoglobin below 6 g/dL were
excluded, and PSA observed after the start of continuous antiandrogen therapy,
surgery, radiotherapy or chemotherapy was dropped. PSA below 0.1 ng/mL (the
LLOQ) was retained as censored.

# Source trace

```{r source-trace}
trace <- tibble::tribble(
  ~Item, ~Value, ~Source,
  "Model II resistant clone", "PSA_R = BAS * R * exp(GR * t)", "Methods, Eq 2",
  "Model II sensitive clone", "PSA_S = BAS * (1 - R) * exp(GS * t_s) * exp(-DS * t_k)", "Methods, Eq 3",
  "Time split", "t_s = min(t, t1); t_k = max(0, t - t1)", "Methods, text after Eq 3",
  "Resistant fraction", "R = exp(-RP)", "Methods, text after Eq 3",
  "IIV form", "theta_i = theta_typical * exp(eta_i), all parameters", "Methods, Eq 4",
  "Residual error", "additive on log PSA", "Methods, Eq 5; Results",
  "Covariate on RP", "RP = theta_RP * (HGB / 13.6)^theta_HGB_RP", "Results, Eq 16",
  "Covariates on DS", "DS = theta_DS * (BAS / 8.5)^theta_BAS_DS * exp(IND_AND * theta_AND_DS)", "Results, Eq 17",
  "lkse (DS)", "3.78e-2 1/day", "Table 2",
  "lkge_sens (GS)", "1.96e-3 1/day", "Table 2",
  "lrp (RP)", "3.94", "Table 2",
  "lkge (GR)", "6.54e-4 1/day", "Table 2",
  "etalkse", "omega^2 = 0.453", "Table 2",
  "etalkge_sens", "omega^2 = 2.59", "Table 2",
  "etalrp", "omega^2 = 0.944", "Table 2",
  "etalkge", "omega^2 = 3.76 (row labelled 'DR')", "Table 2",
  "e_hgb_rp", "2.30", "Table 2",
  "e_psa_bl_kse", "0.174", "Table 2",
  "e_antiandrogen_kse", "0.677", "Table 2",
  "expSd", "0.201", "Table 2"
)
knitr::kable(trace)
```

# Structural checks

## The ODE form reproduces the published closed form

The model writes Eqs 2-3 as two first-order ODEs whose sensitive-clone rate
switches from `+GS` to `-DS` at `T_SCAN_TO_DOSE`. Solving it for a stochastic
cohort with a spread of treatment-start delays and comparing each subject's
PSA against Eqs 2-3 evaluated at the same drawn parameters checks the encoding.
Both sides use identical parameters, so the difference is numerical error only
and a tight bound applies.

```{r closed-form-gate}
set.seed(2020)
n_cf <- 50
cf_cov <- tibble(
  id = seq_len(n_cf),
  PSA_BL = exp(rnorm(n_cf, log(8.5), 1)),
  HGB = rnorm(n_cf, 13.6, 1.5),
  CONMED_ANTIANDROGEN = rbinom(n_cf, 1, 0.125),
  T_SCAN_TO_DOSE = sample(c(0, 14, 45, 90), n_cf, replace = TRUE)
)
cf_times <- c(0, 7, 14, 30, 45, 60, 90, 120, 180, 270, 365, 540, 730)
cf_ev <- tidyr::crossing(cf_cov, time = cf_times) |>
  mutate(evid = 0, amt = 0) |>
  arrange(id, time)

rxode2::rxSetSeed(2020)
cf_sim <- rxode2::rxSolve(
  mod, cf_ev,
  keep = c("PSA_BL", "T_SCAN_TO_DOSE"),
  returnType = "data.frame",
  atol = 1e-10, rtol = 1e-10
)

cf_chk <- cf_sim |>
  mutate(
    t_s = pmin(time, T_SCAN_TO_DOSE),
    t_k = pmax(0, time - T_SCAN_TO_DOSE),
    psa_closed = PSA_BL * fres * exp(kge * time) +
      PSA_BL * (1 - fres) * exp(kge_sens * t_s) * exp(-kse * t_k),
    abs_err = abs(PSA - psa_closed),
    # Extreme etas start some resistant clones near 1e-6 ng/mL, comparable
    # to the solver's absolute tolerance, and the ODE then carries a relative
    # error of order 1e-5 as that clone grows. A structural error (a wrong
    # rate, a missed switch) would be off by tens of percent, so a 1e-4
    # relative bound, plus a small absolute term for near-zero values, still
    # isolates the encoding from numerical noise.
    within = abs_err <= 1e-4 * psa_closed + 1e-8
  )
summary(cf_chk$abs_err / cf_chk$psa_closed)
stopifnot(all(cf_chk$within))
```

## Typical covariate effects match the Results

The Results quote typical `RP` (and the implied `R`) at the cohort 5th
percentile, median and 95th percentile of hemoglobin, and typical `DS` at the
same percentiles of baseline PSA. These are deterministic functions of Table 2.

```{r covariate-gate}
th <- mod$theta
rp_typ <- function(hgb) exp(th[["lrp"]]) * (hgb / 13.6)^th[["e_hgb_rp"]]
ds_typ <- function(bas, and = 0) {
  exp(th[["lkse"]]) * (bas / 8.5)^th[["e_psa_bl_kse"]] *
    exp(th[["e_antiandrogen_kse"]] * and)
}
cov_tab <- tibble(
  Quantity = c(
    "RP, HGB 10.9", "RP, HGB 13.6", "RP, HGB 16.0",
    "R (%), HGB 10.9", "R (%), HGB 13.6", "R (%), HGB 16.0",
    "DS (1/day), PSA_BL 0.515", "DS (1/day), PSA_BL 8.5", "DS (1/day), PSA_BL 120",
    "DS increase with antiandrogen (%)",
    "GS doubling time (day)", "GR doubling time (day)"
  ),
  Model = c(
    rp_typ(c(10.9, 13.6, 16.0)),
    100 * exp(-rp_typ(c(10.9, 13.6, 16.0))),
    ds_typ(c(0.515, 8.5, 120)),
    100 * (ds_typ(8.5, 1) / ds_typ(8.5, 0) - 1),
    log(2) / exp(th[["lkge_sens"]]),
    log(2) / exp(th[["lkge"]])
  ),
  Paper = c(
    2.37, 3.94, 5.73, 9.36, 1.94, 0.326,
    2.32e-2, 3.78e-2, 5.99e-2, 96.8, 354, 1060
  )
) |>
  mutate(`Difference (%)` = 100 * (Model / Paper - 1))
knitr::kable(cov_tab, digits = 4)
stopifnot(all(abs(cov_tab$`Difference (%)`) < 1.5))
```

All agree to the three significant figures the paper prints.

# Virtual cohort and simulated PSA profiles

The cohort draws covariates to match Table 1: baseline PSA log-normal with
median 8.5 ng/mL and the 5th/95th percentiles (0.515 / 120 ng/mL) the Results
quote; hemoglobin normal with median 13.6 g/dL and the quoted 5th/95th
percentiles (10.9 / 16.0 g/dL), truncated to the observed range; antiandrogen
use with probability 33/264. The paper does not report the delay from the
baseline draw to the first dose, so leuprorelin starts at time 0.

```{r cohort}
set.seed(71)
n_sub <- 200
bl_sdlog <- (log(120) - log(0.515)) / (2 * qnorm(0.95))
hgb_sd <- (16.0 - 10.9) / (2 * qnorm(0.95))
cohort <- tibble(
  id = seq_len(n_sub),
  PSA_BL = pmin(pmax(exp(rnorm(n_sub, log(8.5), bl_sdlog)), 0.2), 782),
  HGB = pmin(pmax(rnorm(n_sub, 13.6, hgb_sd), 6.8), 17.4),
  CONMED_ANTIANDROGEN = rbinom(n_sub, 1, 33 / 264),
  T_SCAN_TO_DOSE = 0
)
obs_times <- c(0, seq(7, 1095, by = 14))
ev <- tidyr::crossing(cohort, time = obs_times) |>
  mutate(evid = 0, amt = 0) |>
  arrange(id, time)

rxode2::rxSetSeed(71)
sim <- rxode2::rxSolve(mod, ev, returnType = "data.frame")
```

```{r vpc, fig.cap = "Simulated PSA (individual predictions) in 200 virtual subjects starting leuprorelin at time 0: median with 5th-95th percentile band. The observed-data counterpart is the prediction-corrected VPC of S2 Fig of Zou 2020."}
sim |>
  group_by(time) |>
  summarise(
    q05 = quantile(PSA, 0.05), q50 = median(PSA), q95 = quantile(PSA, 0.95),
    .groups = "drop"
  ) |>
  ggplot(aes(time / 30.4375, q50)) +
  geom_ribbon(aes(ymin = q05, ymax = q95), alpha = 0.25) +
  geom_line() +
  geom_hline(yintercept = 0.1, linetype = "dotted", colour = "blue") +
  scale_y_log10() +
  labs(x = "Time since leuprorelin start (months)", y = "PSA (ng/mL)")
```

```{r archetypes, fig.cap = "Typical-subject PSA split into its drug-sensitive and drug-resistant parts at the 5th percentile, median and 95th percentile of hemoglobin, with leuprorelin starting 60 days after the baseline draw. Compare the decomposition shown for individual subjects in Figure 4 of Zou 2020."}
arch <- tibble(
  id = 1:3, HGB = c(10.9, 13.6, 16.0), PSA_BL = 8.5,
  CONMED_ANTIANDROGEN = 0, T_SCAN_TO_DOSE = 60
) |>
  tidyr::crossing(time = seq(0, 1095, by = 5)) |>
  mutate(evid = 0, amt = 0) |>
  arrange(id, time)
arch_sim <- rxode2::rxSolve(rxode2::zeroRe(mod), arch, keep = "HGB", returnType = "data.frame")
arch_sim |>
  select(time, HGB, Total = PSA, Sensitive = shrink, Resistant = growth) |>
  pivot_longer(c(Total, Sensitive, Resistant), names_to = "part", values_to = "psa") |>
  ggplot(aes(time / 30.4375, psa, colour = part)) +
  geom_line() +
  geom_vline(xintercept = 60 / 30.4375, linetype = "dashed", colour = "red") +
  facet_wrap(~ paste("HGB", HGB, "g/dL")) +
  scale_y_log10() +
  labs(x = "Time since baseline PSA (months)", y = "PSA (ng/mL)", colour = NULL)
```

# Reproducing the paper's simulations

The paper simulated 1000 subjects per scenario at the 5th percentile, median
and 95th percentile of one continuous covariate (the others at their medians,
no antiandrogen), or with and without antiandrogen, and reported

* the median PSA nadir and median time to nadir (Discussion), and
* the percentage of subjects with PSA progression by PCWG criteria -- a rise of
  at least 25% and at least 2 ng/mL above the nadir, confirmed three or more
  weeks later -- within one, two and three years of starting leuprorelin
  (Figure 3 and Results).

The simulated quantities below are individual predictions (between-subject
variability, no residual error) on a daily grid, with treatment starting at
time 0. After the nadir each individual prediction rises monotonically, so any
first crossing of the PCWG threshold is automatically confirmed three weeks
later. Because the closed-form check above shows the ODE and Eqs 2-3 agree,
the scenarios are evaluated with the closed form directly from the packaged
model's `theta` and `omega`, which allows 4000 draws per scenario at negligible
cost.

```{r pcwg-sim}
om <- mod$omega
eta_names <- c("etalkse", "etalkge_sens", "etalrp", "etalkge")
sim_scenario <- function(n, hgb = 13.6, bas = 8.5, and = 0, days = 0:1095) {
  # The four etas are uncorrelated (diagonal omega, Table 2).
  eta <- vapply(eta_names, function(nm) rnorm(n, 0, sqrt(om[nm, nm])), numeric(n))
  kse <- ds_typ(bas, and) * exp(eta[, "etalkse"])
  rp <- rp_typ(hgb) * exp(eta[, "etalrp"])
  kge <- exp(th[["lkge"]] + eta[, "etalkge"])
  fres <- exp(-rp)
  psa <- bas * (fres * exp(outer(kge, days)) + (1 - fres) * exp(-outer(kse, days)))
  nadir <- apply(psa, 1, min)
  t_nadir <- days[apply(psa, 1, which.min)]
  thr <- pmax(1.25 * nadir, nadir + 2)
  t_prog <- vapply(seq_len(n), function(i) {
    j <- which(psa[i, ] >= thr[i] & days > t_nadir[i])
    if (length(j) > 0) days[j[1]] else Inf
  }, numeric(1))
  tibble(
    nadir = median(nadir), t_nadir = median(t_nadir),
    p1 = 100 * mean(t_prog <= 365), p2 = 100 * mean(t_prog <= 730),
    p3 = 100 * mean(t_prog <= 1095)
  )
}
set.seed(354)
scen <- tibble::tribble(
  ~Scenario, ~hgb, ~bas, ~and,
  "HGB 10.9 g/dL", 10.9, 8.5, 0,
  "Median covariates", 13.6, 8.5, 0,
  "HGB 16.0 g/dL", 16.0, 8.5, 0,
  "PSA_BL 0.515 ng/mL", 13.6, 0.515, 0,
  "PSA_BL 120 ng/mL", 13.6, 120, 0,
  "Antiandrogen use", 13.6, 8.5, 1
)
scen_res <- scen |>
  rowwise() |>
  mutate(res = list(sim_scenario(4000, hgb, bas, and))) |>
  ungroup() |>
  tidyr::unnest(res)
```

```{r pcwg-table}
paper_prog <- tibble::tribble(
  ~Scenario, ~p1_paper, ~p3_paper,
  "HGB 10.9 g/dL", 19.8, 38.8,
  "Median covariates", 13.9, 28.2,
  "HGB 16.0 g/dL", 10.5, 22.1,
  "PSA_BL 0.515 ng/mL", 6.9, 15.8,
  "PSA_BL 120 ng/mL", 25.6, 47.8,
  "Antiandrogen use", NA, 28.5
)
prog_cmp <- scen_res |>
  left_join(paper_prog, by = "Scenario") |>
  select(Scenario, p1, p1_paper, p2, p3, p3_paper)
prog_cmp |>
  dplyr::rename(
    "Progression by 1 y (%), simulated" = p1,
    "Progression by 1 y (%), paper" = p1_paper,
    "Progression by 2 y (%), simulated" = p2,
    "Progression by 3 y (%), simulated" = p3,
    "Progression by 3 y (%), paper" = p3_paper
  ) |>
  knitr::kable(digits = 1)
```

Replicates Figure 3 of Zou 2020. The two-year percentages are shown in Figure 3
but not quoted numerically in the text, so they have no paper column.

```{r pcwg-gate}
d1 <- prog_cmp$p1 - prog_cmp$p1_paper
d3 <- prog_cmp$p3 - prog_cmp$p3_paper
stopifnot(
  # Every quoted rate within 5 percentage points (the paper's own Monte Carlo
  # standard error at n = 1000 is about 1.5 points at these rates).
  all(abs(d1) < 5, na.rm = TRUE),
  all(abs(d3) < 5, na.rm = TRUE),
  # The covariate orderings the paper reports.
  with(scen_res, p3[Scenario == "HGB 10.9 g/dL"] > p3[Scenario == "Median covariates"]),
  with(scen_res, p3[Scenario == "Median covariates"] > p3[Scenario == "HGB 16.0 g/dL"]),
  with(scen_res, p3[Scenario == "PSA_BL 120 ng/mL"] > p3[Scenario == "Median covariates"]),
  with(scen_res, p3[Scenario == "Median covariates"] > p3[Scenario == "PSA_BL 0.515 ng/mL"])
)
```

```{r nadir-table}
paper_nadir <- tibble::tribble(
  ~Scenario, ~nadir_paper, ~t_nadir_paper,
  "HGB 10.9 g/dL", 1.19, 155,
  "Median covariates", 0.290, 198,
  "HGB 16.0 g/dL", 0.0574, 240,
  "PSA_BL 0.515 ng/mL", 0.0220, 294,
  "PSA_BL 120 ng/mL", 3.47, 135
)
nadir_cmp <- scen_res |>
  inner_join(paper_nadir, by = "Scenario") |>
  select(Scenario, nadir, nadir_paper, t_nadir, t_nadir_paper) |>
  mutate(
    nadir_ratio = nadir / nadir_paper,
    t_nadir_ratio = t_nadir / t_nadir_paper
  )
nadir_cmp |>
  dplyr::rename(
    "Median nadir (ng/mL), simulated" = nadir,
    "Median nadir (ng/mL), paper" = nadir_paper,
    "Median time to nadir (day), simulated" = t_nadir,
    "Median time to nadir (day), paper" = t_nadir_paper,
    "Nadir ratio" = nadir_ratio,
    "Time-to-nadir ratio" = t_nadir_ratio
  ) |>
  knitr::kable(digits = 3)
stopifnot(
  # The median nadir spans two orders of magnitude across scenarios; its
  # sample median is noisy at the paper's n = 1000, so allow 30%.
  all(abs(log(nadir_cmp$nadir_ratio)) < log(1.3)),
  all(abs(nadir_cmp$t_nadir_ratio - 1) < 0.15)
)
```

The nadir values reproduce the two-orders-of-magnitude spread the Discussion
reports, and time to nadir agrees within about 7%. The simulated median nadir
at median covariates (about 0.24-0.25 ng/mL against the quoted 0.290) sits at
the low end of the agreement; the sample median of a quantity this skewed moves
by tens of percent between Monte Carlo cohorts of 1000.

# Assumptions and deviations

* **Row label 'DR' in Table 2.** The fourth IIV row is labelled `omega^2 DR`;
  Methods states IIV was estimated on all four structural parameters and there
  is no `DR` parameter, so it is taken as the IIV of `GR`.
* **Table 2 IIV values are variances.** The column is headed `omega^2`. The
  reproduction of the Figure 3 progression rates and the Discussion nadir values
  confirms the variance reading.
* **Residual error.** Eq 5 is additive on log PSA, encoded as `lnorm(expSd)`
  with `expSd = 0.201`. The paper retained PSA below the 0.1 ng/mL LLOQ as
  censored observations; censoring is a property of the data, not the model.
* **Treatment start.** Only the first leuprorelin dose enters the model. The
  analysis pooled continuous and intermittent regimens (which the claims data
  could not distinguish), and dose intensity was tested and not retained. The
  vignette simulations start treatment at time 0 because the paper does not
  report the baseline-to-first-dose interval.
* **Simulation settings for Figure 3.** The paper states only that 1000
  subjects per scenario were simulated in R with the final model. The
  reproduction assumes individual predictions without residual error, the
  non-varied covariates at their medians, no antiandrogen in the continuous
  covariate scenarios (the median rows of the hemoglobin and baseline-PSA
  panels are identical, 13.9% and 28.2%, and equal the no-antiandrogen
  three-year value), and progression timed from the start of leuprorelin. The
  close agreement supports these readings.
* **Covariate cohort.** Table 1 gives medians and ranges only; the log-normal
  baseline-PSA and normal hemoglobin distributions are built from the 5th and
  95th percentiles quoted in the Results.
* **Screened but not retained covariates** (age, AST, ALT, serum creatinine,
  alkaline phosphatase, albumin, race and region) are documented in the model
  file's `covariatesDataExcluded`, not used.
