---
title: "Pretomanid (Zou 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pretomanid (Zou 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r load-model}
ui <- rxode2::rxode(readModelDb("Zou_2022_pretomanid"))
```

# Model and source

* Citation: `r ui$reference`
* Description: `r ui$description`

Zou 2022 characterised a pediatric dispersible-tablet formulation (DTF) of
pretomanid against the marketed 200 mg tablet (MF) in study CL-011, a phase 1
single-dose, four-period crossover in healthy adults. Each participant received
10, 50 and 200 mg (4 x 50 mg) DTF and 200 mg MF, separated by 7-day washouts.
Only the panel dosed after an FDA standard high-fat breakfast was modelled,
because pretomanid is labelled to be taken with food.

The starting point was the Salinger 2019 pooled model (also in this package as
`Salinger_2019_pretomanid`). Zou 2022 kept its one-compartment disposition and
its transit-plus-absorption-compartment layout, but reduced the transit chain
from three compartments to one:

```
dose --> transit1 --> depot --> central --> eliminated
           ktr         ka         kel
```

with `ktr = 1 / MTT`, so the mean absorption time is `MAT = 1/KA + MTT`. As in
the Salinger 2019 model, `depot` is the source control stream's `ABS`
compartment and sits downstream of the transit compartment; **oral doses enter
`transit1`**, which is why the model file declares `dosing <- "transit1"`.

Covariates in the final model:

* body weight, allometric with fixed exponents 0.75 (CL) and 1 (V) around a
  55 kg reference participant;
* formulation, on KA only: the marketed tablet absorbs 1.65-fold faster than
  the dispersible tablet. Formulation was tested on bioavailability (ratio
  1.00, 90% CI 0.87-1.14) and not retained;
* dose, as a power function on relative bioavailability,
  `F = (DOSE/200)^0.0822`.

Interindividual variability is on CL and V (correlated). Inter-occasion
variability, one occasion per crossover period, is on F, KA and MTT. The
residual error is combined proportional and additive, and both components are
multiplied by 3.46 for samples taken less than 10 h after the dose.

```{r structure}
cat("ODE states:", paste(ui$state, collapse = " -> "), "\n")
cat("Observation:", paste(ui$predDf$var, collapse = ", "), "\n")
stopifnot(identical(ui$state, c("transit1", "depot", "central")))
```

# Population

The model was fitted to the fed panel of CL-011: 24 healthy adults (16 male,
8 female) contributing 1377 concentrations. Zou 2022 Table 1 ('Fed' column)
reports a median age of 39.0 years (range 23-50), median weight 76.2 kg
(64.4-117), median height 170 cm (155-193) and median BMI 27.0 kg/m^2
(22.6-31.5). The study was run at a single site in San Antonio, Texas. The same
information is stored in the model's `population` metadata:

```{r population}
str(readModelDb("Zou_2022_pretomanid")()$population, give.attr = FALSE)
```

# Source trace

Every value comes from Zou 2022 Table 2 and the NONMEM control stream printed in
Section C of the Electronic Supplementary Material (ESM). The control stream's
`$THETA`, `$OMEGA` and `$SIGMA` hold the final estimates: each one rounds to
the corresponding Table 2 entry.

| Quantity | Value | Source |
|---|---|---|
| `lka` (KA, DTF) | log(0.396) 1/h | Table 2 'Absorption rate of DTF'; ESM THETA(3) |
| `e_form_mf_ka` | 1.65 | Table 2 'Proportional effect (theta3) of MF on KA'; ESM THETA(5) |
| `lmtt` | log(1.13) h | Table 2 'Mean transit time'; ESM THETA(4) |
| `lcl` | log(2.81) L/h | Table 2 'Apparent clearance'; ESM THETA(1) |
| `lvc` | log(68.0) L | Table 2 'Apparent volume of distribution'; ESM THETA(2) |
| `e_wt_cl`, `e_wt_vc` | 0.75, 1 (fixed) | Table 2 weight-scaling rows; ESM `$PK` TVLCL / TVLV2 |
| Weight reference | 55 kg | Table 2 footnote a; ESM `LOG(WT/55)` |
| `lfdepot` | log(1) (fixed) | Table 2 'Bioavailability (F)' |
| `e_dose_fdepot` | 0.0822 | Table 2 'Dose effect (theta4) on F'; ESM THETA(6) |
| `etalcl`, `etalvc` block | 0.0622, 0.0111, 0.00751 | ESM `$OMEGA BLOCK(2)`; Table 2 CVs 24.9% and 8.67% |
| `etaiov_fdepot_*` | 0.00561 | ESM `$OMEGA BLOCK(3)` 'IOV.F1'; Table 2 7.49% |
| `etaiov_ka_*` | 0.284 | ESM 'IOV.KA'; Table 2 53.3% |
| `etaiov_mtt_*` | 0.77 | ESM 'IOV.MATT'; Table 2 87.8% |
| `propSd` | 0.0889 | Table 2 'Proportional error' 8.89%; ESM `$SIGMA` 0.0079 |
| `addSd` | 0.401 ng/mL | Table 2 'Additive error'; ESM `$SIGMA` 0.161 |
| `e_tad_early_ruv` | 3.46 | Table 2 'Time-varying error term (theta5)'; ESM THETA(7) |
| `ktr = 1/MTT`, dose into transit compartment | -- | ESM `$MODEL` and `$PK` (K13 = KTR, K32 = KA) |
| `F = (DOSE/200)^theta4` | -- | Table 2; ESM `TVLF1 = LOG(DOSE/200) * THETA(6)` |
| `ERRT` = 3.46 when TAD < 10 h | -- | Table 2 footnote d; ESM `$ERROR` |

Table 2 footnote b states that the IIV, IOV and proportional-error entries are
`sqrt(variance) x 100`, and the control-stream variances reproduce them:
`sqrt(0.0622) = 0.249`, `sqrt(0.00751) = 0.0867`, `sqrt(0.00561) = 0.0749`,
`sqrt(0.284) = 0.533`, `sqrt(0.77) = 0.878`, `sqrt(0.0079) = 0.0889` and
`sqrt(0.161) = 0.401`.

# Typical-value checks against the paper

The paper states three derived typical values that follow from the parameters
alone, so they are checked exactly.

```{r typical-checks}
p <- as.list(ui$theta)
mat_dtf <- 1 / exp(p$lka) + exp(p$lmtt)
mat_mf <- 1 / (exp(p$lka) * p$e_form_mf_ka) + exp(p$lmtt)
frel <- (c(10, 50) / 200)^p$e_dose_fdepot

typical <- data.frame(
  Quantity = c(
    "MAT, dispersible tablet (h)",
    "MAT, marketed tablet (h)",
    "MAT ratio DTF / MF (%)",
    "Relative F, 10 mg vs 200 mg (%)",
    "Relative F, 50 mg vs 200 mg (%)"
  ),
  Paper = c(3.7, 2.7, 137, 78, 89),
  Model = c(mat_dtf, mat_mf, 100 * mat_dtf / mat_mf, 100 * frel)
)
typical$Model <- signif(typical$Model, 3)
knitr::kable(
  typical,
  caption = "MAT from the Discussion and Results Section 3.2; relative bioavailability from Results Section 3.2."
)
# Each paper value is printed to 2-3 significant figures; the model must
# reproduce it to that rounding.
stopifnot(
  abs(mat_dtf - 3.7) < 0.05,
  abs(mat_mf - 2.7) < 0.05,
  abs(100 * mat_dtf / mat_mf - 137) < 1,
  abs(100 * frel - c(78, 89)) < 0.5
)
```

# Virtual cohort

A crossover cohort mirroring CL-011: every virtual participant receives all
four treatments, one per period, with a 168-h interval between doses and the
CL-011 sampling schedule (0-96 h after each dose). Treatments are assigned to
periods in a fixed order (MF 200 mg, DTF 200 mg, DTF 50 mg, DTF 10 mg); with
the same inter-occasion variance on every occasion and a half-life near 13 h,
period order and carry-over do not affect the comparison. Body weight is
log-normal around the fed-panel median of 76.2 kg, truncated to the observed
64.4-117 kg range.

```{r cohort}
n_sub <- 200
set.seed(20221163)
wt <- numeric(0)
while (length(wt) < n_sub) {
  draw <- rlnorm(2 * n_sub, log(76.2), 0.15)
  wt <- c(wt, draw[draw >= 64.4 & draw <= 117])
}
wt <- wt[seq_len(n_sub)]

treatments <- data.frame(
  OCC = 1:4,
  treatment = c("A: 200 mg MF", "B: 200 mg DTF", "C: 50 mg DTF", "D: 10 mg DTF"),
  DOSE_PRETOMANID_MG = c(200, 200, 50, 10),
  FORM_PRETOMANID_DT = c(0, 1, 1, 1)
)
period_h <- 168
sample_times <- c(0, 0.5, 1, 2, 3, 4, 6, 8, 12, 16, 24, 36, 48, 72, 96)

dose_rows <- tidyr::expand_grid(id = seq_len(n_sub), OCC = 1:4) |>
  dplyr::mutate(
    time = (OCC - 1) * period_h,
    evid = 1L,
    amt = treatments$DOSE_PRETOMANID_MG[OCC],
    cmt = "transit1"
  )
obs_rows <- tidyr::expand_grid(id = seq_len(n_sub), OCC = 1:4, tad = sample_times) |>
  dplyr::mutate(
    time = (OCC - 1) * period_h + tad,
    evid = 0L,
    amt = 0,
    cmt = "central"
  ) |>
  dplyr::select(-tad)

events <- dplyr::bind_rows(dose_rows, obs_rows) |>
  dplyr::left_join(treatments |> dplyr::select(-treatment), by = "OCC") |>
  dplyr::mutate(WT = wt[id]) |>
  dplyr::arrange(id, time, dplyr::desc(evid)) |>
  dplyr::select(id, time, evid, amt, cmt, dplyr::everything())

stopifnot(
  !anyNA(events),
  dplyr::n_distinct(events$id) == n_sub,
  sum(events$evid == 1) == 4 * n_sub
)
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20221163)
sim <- rxode2::rxSolve(ui, events, returnType = "data.frame")
# Recover the occasion and time after dose from the clock time: each dose
# opens a 168-h period, and every sample falls 0-96 h into its period.
sim <- sim |>
  dplyr::select(-dplyr::any_of(c("OCC", "DOSE_PRETOMANID_MG", "FORM_PRETOMANID_DT"))) |>
  dplyr::mutate(
    OCC = as.integer(floor(time / period_h)) + 1L,
    tad_h = time - (OCC - 1) * period_h
  ) |>
  dplyr::left_join(treatments, by = "OCC")
stopifnot(
  nrow(sim) == n_sub * 4 * length(sample_times),
  !anyNA(sim$Cc),
  all(sim$tad_h %in% sample_times)
)
```

## Visual predictive check (replicates Figure 2)

Figure 2 of Zou 2022 shows the 10th, 50th and 90th percentiles of the fed-panel
concentrations by treatment on a log scale. The simulated percentiles below
include residual error (`sim`) and are the model's counterpart of those bands.

```{r vpc, fig.width = 7, fig.height = 5}
vpc <- sim |>
  dplyr::filter(tad_h > 0) |>
  dplyr::group_by(treatment, tad_h) |>
  dplyr::summarise(
    p10 = quantile(sim, 0.10),
    p50 = quantile(sim, 0.50),
    p90 = quantile(sim, 0.90),
    .groups = "drop"
  )
ggplot(vpc, aes(tad_h)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = p50), colour = "steelblue4") +
  facet_wrap(~treatment) +
  scale_y_log10() +
  labs(
    x = "Time after dose (h)",
    y = "Pretomanid concentration (ng/mL)",
    caption = "Simulated median and 10th-90th percentiles; replicates Figure 2 of Zou 2022."
  ) +
  theme_bw()
```

## Absorption-phase residual error

The residual SD is 3.46-fold larger for samples taken less than 10 h after the
dose. At the 200 mg concentrations the proportional term dominates the
additive one, so the ratio of the relative-residual SDs before and after 10 h
recovers the multiplier.

```{r ruv-check}
ruv <- sim |>
  dplyr::filter(DOSE_PRETOMANID_MG == 200, tad_h > 0, Cc > 50) |>
  dplyr::mutate(rel_resid = (sim - Cc) / Cc, early = tad_h < 10) |>
  dplyr::group_by(early) |>
  dplyr::summarise(sd_rel = sd(rel_resid), n = dplyr::n(), .groups = "drop")
knitr::kable(ruv, digits = 4, caption = "Relative-residual SD before (TRUE) and after (FALSE) 10 h, 200 mg arms.")
ratio <- ruv$sd_rel[ruv$early] / ruv$sd_rel[!ruv$early]
cat("Early / late residual SD ratio:", signif(ratio, 3), "(model multiplier 3.46)\n")
# Thousands of residuals per group put the SD ratio within a few percent of
# 3.46 (the additive term lowers it slightly); omitting the multiplier gives
# 1, and applying it to every sample also gives 1.
stopifnot(ratio > 3.0, ratio < 3.9)
```

# Non-compartmental analysis

NCA uses the individual predictions (`Cc`, no residual error) on the CL-011
sampling schedule, per treatment. Reference values are the fed-panel geometric
least-squares means in the ESM: Table D1 for treatments A and B, and Table D5
(primary analysis, 'Test' = fed) for treatments C and D. The simulated values
are cohort medians, which for log-normally distributed exposure estimate the
same quantity as a geometric mean.

```{r pknca}
conc <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::mutate(conc = pmax(Cc, 0)) |>
  dplyr::select(id, treatment, time = tad_h, conc)
dose <- events |>
  dplyr::filter(evid == 1) |>
  dplyr::left_join(treatments |> dplyr::select(OCC, treatment), by = "OCC") |>
  dplyr::transmute(id, treatment, time = 0, amt)

conc_obj <- PKNCA::PKNCAconc(conc, conc ~ time | treatment + id)
dose_obj <- PKNCA::PKNCAdose(dose, amt ~ time | treatment + id)
intervals <- data.frame(
  start = c(0, 0),
  end = c(96, Inf),
  cmax = c(TRUE, FALSE),
  tmax = c(TRUE, FALSE),
  auclast = c(TRUE, FALSE),
  aucinf.obs = c(FALSE, TRUE),
  half.life = c(FALSE, TRUE)
)
nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_long <- as.data.frame(nca_res$result) |>
  dplyr::filter(PPTESTCD %in% c("cmax", "tmax", "auclast", "aucinf.obs", "half.life"))
```

```{r nca-compare}
reference <- data.frame(
  treatment = c("A: 200 mg MF", "B: 200 mg DTF", "C: 50 mg DTF", "D: 10 mg DTF"),
  cmax = c(2120, 1820, 384, 75.3),
  auclast = c(53900, 54200, 12000, 1970),
  aucinf.obs = c(56100, 56700, 12400, 2030)
)
cmp <- nlmixr2lib::ncaComparisonTable(
  simulated = nca_long,
  reference = reference,
  by = "treatment",
  units = c(cmax = "ng/mL", auclast = "ng*h/mL", aucinf.obs = "ng*h/mL"),
  tolerance_pct = 20
)
knitr::kable(
  cmp,
  caption = paste(
    "Zou 2022 ESM Tables D1 and D5 (fed-panel geometric LS means) vs simulated",
    "cohort medians. AUC0-t is AUC over 0-96 h. * differs by more than 20%."
  )
)
```

```{r nca-gate}
pct <- nca_long |>
  dplyr::group_by(treatment, PPTESTCD) |>
  dplyr::summarise(sim = median(PPORRES), .groups = "drop") |>
  dplyr::inner_join(
    tidyr::pivot_longer(reference, -treatment, names_to = "PPTESTCD", values_to = "pub"),
    by = c("treatment", "PPTESTCD")
  ) |>
  dplyr::mutate(pct_diff = 100 * (sim / pub - 1))
auc_pct <- pct$pct_diff[pct$PPTESTCD != "cmax"]
cmax_pct <- pct$pct_diff[pct$PPTESTCD == "cmax"]
# AUC is set by F * dose / CL, so a mis-transcribed clearance, weight
# reference, dose exponent or unit moves every arm by tens of percent.
# Cmax is underpredicted in every arm, by about 10-25%, which is the peak
# underprediction the paper reports for its own model (Discussion), so it
# is gated on its centre with an envelope that admits cohort noise.
stopifnot(
  abs(median(auc_pct)) < 8,
  max(abs(auc_pct)) < 15,
  abs(median(cmax_pct)) < 25,
  max(abs(cmax_pct)) < 35
)
```

Simulated and observed AUCs agree across the 20-fold dose range. The largest
AUC difference is at 10 mg, where the power model's relative bioavailability
of 78% sits slightly above the observed dose-normalised exposure (the observed
10 mg/200 mg AUC0-inf ratio is about 72%).

Cmax is underpredicted in every arm, most for the marketed 200 mg tablet and
the 10 mg dispersible tablet (about 20-25% below the observed geometric means)
and least at 50 mg. This reproduces a limitation of the published model rather
than an encoding error: Zou 2022 reports that its own peak concentrations were
underpredicted at 10 mg and for the 200 mg marketed tablet (Discussion and ESM
Section B VPCs). The model was built to predict AUC, the exposure metric used
for pediatric dose selection.

# Assumptions and deviations

* **Fed conditions only.** The model was fitted to the fed panel; it carries no
  food effect and should not be used to simulate fasted dosing. The fasted
  panel was analysed by NCA only, and the dose dependence of exposure reverses
  under fasting (NCA slope 0.824 versus 1.086 fed).
* **Weight reference.** Clearance and volume are referenced to 55 kg, inherited
  from Salinger 2019, even though the fed panel weighed 64.4-117 kg. The
  typical values `CL = 2.81 L/h` and `V = 68.0 L` therefore describe a lighter
  participant than any enrolled.
* **Formulation indicator polarity.** The source column `DTF` (1 = dispersible
  tablet) is carried as `FORM_PRETOMANID_DT`; the model forms the
  marketed-formulation indicator `1 - FORM_PRETOMANID_DT` internally, as the
  control stream does.
* **Occasions.** One occasion per crossover period (`OCC` 1-4). The ESM
  `$OMEGA BLOCK(3) SAME` blocks are encoded as independent per-occasion etas
  with occasions 2-4 fixed to the occasion-1 variances; the BLOCK(3)
  off-diagonals are zero in the source, so the etas are uncorrelated. Records
  with `OCC` outside 1-4 receive no inter-occasion variability.
* **Mu-referencing warning.** rxode2 has no occasion level for random effects,
  so the per-occasion etas are combined through occasion indicators. rxode2
  warns that these etas are not mu-referenced. The warning concerns estimation
  speed only; simulation is unaffected.
* **Time-varying residual error.** The control stream switches on a `TAD`
  data column; the model uses rxode2's `tad()`, the time since the most recent
  dose, which is the same quantity for single doses separated by washouts.
  NONMEM's `Y = IPRED + ERRT * (EPS(1) * IPRED + EPS(2))` with a diagonal
  `$SIGMA` is a combined error whose variances add (`combined2()`).
* **Unpublished covariance.** The CL-V covariance (0.0111, correlation 0.51) is
  printed only in the ESM control stream, not in Table 2; the control stream's
  other estimates all match Table 2, so it is taken as the final estimate.
* **NCA reference statistic.** The ESM reports geometric least-squares means;
  the comparison uses simulated cohort medians.
* **Peak concentrations.** Simulated Cmax is about 10-25% below the observed
  geometric means, as the source reports for its own model; AUC is reproduced
  within about 6%.
* **Virtual cohort.** Body weight is log-normal (median 76.2 kg, log-SD 0.15)
  truncated to the fed-panel range; sex, age and height do not enter the model.
