---
title: "Bedaquiline (Zou 2022)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bedaquiline (Zou 2022)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

* Citation: `r rxode2::rxode(readModelDb("Zou_2022_bedaquiline"))$reference`
* Description: `r rxode2::rxode(readModelDb("Zou_2022_bedaquiline"))$description`
* Article: <https://doi.org/10.1128/aac.00811-22> (open access, PMC9578397)

Zou 2022 developed a population PK model for oral bedaquiline (BDQ) from 246
plasma concentrations in 99 Chinese adults with multidrug-resistant pulmonary
tuberculosis, sampled during the 200 mg three-times-weekly continuation phase.
A one-compartment model with first-order absorption described the data. Two
covariates were retained on apparent clearance (CL/F): serum gamma-glutamyl
transferase (GGT) as a power term centred on the dataset median (28.9 U/L), and
the AGBL4 rs319952 GG genotype as an additive shift. The final covariate model
(Results Equations 1 and 2) is

$$
\mathrm{CL}/F =
\begin{cases}
4.54 \cdot \left(\dfrac{\mathrm{GGT}}{28.9}\right)^{-0.476}, & \text{rs319952 AA or AG} \\[2ex]
4.54 \cdot \left(\dfrac{\mathrm{GGT}}{28.9}\right)^{-0.476} - 1.4, & \text{rs319952 GG.}
\end{cases}
$$

The genotype effect is an **absolute** 1.4 L/h reduction, not a fractional one;
this is the one point where reproducing the published simulation table pins the
structural form unambiguously (see below).

# Population

```{r population}
str(readModelDb("Zou_2022_bedaquiline")()$population)
```

The analysis cohort was 99 patients (mean age 38.1 years, range 11-78; mean
weight 58 kg, range 17-90.7) recruited October 2020 to October 2021 at a single
centre in Shenzhen. GGT averaged 36.4 U/L (range 8-271.8). AGBL4 rs319952
genotypes were AA/AG/GG = 31/39/13, with 16 of 99 ungenotyped and imputed to the
most frequent class (AG). The sex split of the cohort is not reported, so it is
not simulated here.

# Source trace

Every value is also traced in a comment beside its `ini()` line in
`inst/modeldb/specificDrugs/Zou_2022_bedaquiline.R`.

| Quantity | Value | Source |
|---|---|---|
| Ka | 0.447 1/h | Table 3, Ka row |
| CL/F (typical, GGT = 28.9, AA/AG) | 4.54 L/h | Table 3, CL/F row; Results Eq. 1 |
| Vc/F | 227 L | Table 3, Vc/F row |
| GGT power exponent on CL/F | -0.476 | Table 3, "u GGT on CL/F"; Results Eq. 1 |
| rs319952 GG additive shift on CL/F | -1.4 L/h | Table 3, "u rs319952 on CL/F"; Results Eq. 2 |
| GGT centring value | 28.9 U/L | Results Eqs. 1-2 (dataset median) |
| IIV CL/F | 38.7% | Table 3, h(CL/F) |
| IIV Vc/F | 83.5% | Table 3, h(Vc/F) |
| Proportional residual | 32.2% | Table 3, eps prop |

The IIV percentages are read as coefficients of variation, so the internal
variance is `log(1 + CV^2)`; the table does not state the convention, but the
distinction is immaterial for the deterministic replication of Table 4 below
(which uses typical values only).

# Reproducing the published dosage simulation (Table 4)

Zou 2022 Table 4 reports the steady-state weekly AUC, Cmax, and Cmin for eight
typical-patient scenarios: two rs319952 strata x four GGT levels (10, 30, 50,
100 U/L), under 200 mg three times weekly. We reproduce it with the typical
(zero-eta) model. The three weekly doses are evenly spaced (every 56 h), which
is what matches the published numbers.

```{r sim_table4}
modTyp <- rxode2::zeroRe(rxode2::rxode(readModelDb("Zou_2022_bedaquiline")))

# 200 mg TIW, doses every 56 h, 30 weeks to steady state.
doseTimes <- sort(as.vector(outer(c(0, 56, 112), 168 * (0:29), "+")))

grid <- expand.grid(
  GGT = c(10, 30, 50, 100),
  GG = c(0L, 1L)
)

simOne <- function(GGT, GG) {
  ev <- rxode2::et(amt = 200, time = doseTimes, cmt = "depot")
  ev <- rxode2::et(ev, seq(168 * 29, 168 * 30, by = 0.1), cmt = "central")
  d <- as.data.frame(ev)
  d$GGT <- GGT
  d$SNP_AGBL4_RS319952_GG <- GG
  r <- rxode2::rxSolve(modTyp, d, returnType = "data.frame")
  r <- r[r$time >= 168 * 29, ]
  data.frame(
    genotype = ifelse(GG == 1L, "G", "A&AG"),
    GGT = GGT,
    aucweekly = sum(diff(r$time) * (head(r$Cc, -1) + tail(r$Cc, -1)) / 2),
    cmax = max(r$Cc),
    cmin = min(r$Cc)
  )
}

simTab <- do.call(rbind, Map(simOne, grid$GGT, grid$GG))

published <- tibble::tribble(
  ~genotype, ~GGT, ~aucweekly_pub, ~cmax_pub, ~cmin_pub,
  "A&AG", 10, 79.746, 0.859, 0.176,
  "A&AG", 30, 134.529, 1.165, 0.460,
  "A&AG", 50, 171.561, 1.377, 0.666,
  "A&AG", 100, 238.620, 1.767, 1.051,
  "G", 10, 97.977, 0.959, 0.266,
  "G", 30, 196.080, 1.519, 0.806,
  "G", 50, 286.080, 2.046, 1.328,
  "G", 100, 538.383, 3.538, 2.815
)

cmp <- simTab |>
  dplyr::left_join(published, by = c("genotype", "GGT")) |>
  dplyr::mutate(
    AUCweekly = sprintf("%.1f / %.1f", aucweekly, aucweekly_pub),
    Cmax = sprintf("%.3f / %.3f", cmax, cmax_pub),
    Cmin = sprintf("%.3f / %.3f", cmin, cmin_pub)
  ) |>
  dplyr::select(genotype, GGT, AUCweekly, Cmax, Cmin) |>
  dplyr::rename(
    "rs319952" = genotype,
    "GGT (U/L)" = GGT,
    "AUCweekly (sim / pub)" = AUCweekly,
    "Cmax (sim / pub)" = Cmax,
    "Cmin (sim / pub)" = Cmin
  )

knitr::kable(
  cmp,
  caption = "Simulated vs. Zou 2022 Table 4 (mg/L and mg/L*h).",
  align = c("l", "r", "r", "r", "r")
)

# Hard gate: the replication reproduces the published table to within rounding.
stopifnot(
  max(abs(simTab$aucweekly - published$aucweekly_pub)) < 0.5,
  max(abs(simTab$cmax - published$cmax_pub)) < 0.01,
  max(abs(simTab$cmin - published$cmin_pub)) < 0.01
)
```

Every cell agrees with the paper to three significant figures. Because the
weekly AUC is exactly `600 / (CL/F)` (three 200 mg doses per week divided by the
apparent clearance), the AUC column alone would already pin CL/F; the GG-stratum
AUCs only reconcile when the 1.4 L/h genotype effect is applied **additively
after** the GGT power term, confirming the structural form.

# Steady-state concentration-time profiles (Figure 3)

```{r fig3, fig.width = 7, fig.height = 4}
profile <- function(GGT, GG) {
  ev <- rxode2::et(amt = 200, time = doseTimes, cmt = "depot")
  ev <- rxode2::et(ev, seq(168 * 29, 168 * 30, by = 0.25), cmt = "central")
  d <- as.data.frame(ev)
  d$GGT <- GGT
  d$SNP_AGBL4_RS319952_GG <- GG
  r <- rxode2::rxSolve(modTyp, d, returnType = "data.frame")
  r <- r[r$time >= 168 * 29, ]
  data.frame(
    time = r$time - 168 * 29,
    Cc = r$Cc,
    GGT = factor(GGT),
    genotype = ifelse(GG == 1L, "rs319952 GG", "rs319952 AA & AG")
  )
}

prof <- do.call(rbind, Map(profile, grid$GGT, grid$GG))

ggplot(prof, aes(time, Cc, colour = GGT)) +
  geom_line() +
  facet_wrap(~genotype) +
  labs(
    x = "Time within a steady-state week (h)",
    y = "Bedaquiline (mg/L)",
    colour = "GGT (U/L)",
    title = "Replicates Figure 3 of Zou 2022"
  ) +
  theme_bw()
```

# PKNCA validation

We confirm the steady-state exposures with PKNCA for the reference scenario
(GGT = 30 U/L, each genotype), computing over one 168 h steady-state week.

```{r pknca}
ncaScenarios <- expand.grid(GGT = 30, GG = c(0L, 1L))

concList <- Map(function(GGT, GG) {
  ev <- rxode2::et(amt = 200, time = doseTimes, cmt = "depot")
  ev <- rxode2::et(ev, seq(168 * 29, 168 * 30, by = 0.1), cmt = "central")
  d <- as.data.frame(ev)
  d$GGT <- GGT
  d$SNP_AGBL4_RS319952_GG <- GG
  r <- rxode2::rxSolve(modTyp, d, returnType = "data.frame")
  r <- r[r$time >= 168 * 29, ]
  data.frame(
    id = 1L,
    treatment = ifelse(GG == 1L, "G", "A&AG"),
    time = r$time - 168 * 29,
    Cc = r$Cc
  )
}, ncaScenarios$GGT, ncaScenarios$GG)

conc <- dplyr::bind_rows(concList) |>
  dplyr::filter(!is.na(Cc))

dose <- conc |>
  dplyr::group_by(treatment) |>
  dplyr::summarise(time = 0, amt = 200, .groups = "drop") |>
  dplyr::mutate(id = 1L)

ncaData <- PKNCA::PKNCAdata(
  PKNCA::PKNCAconc(conc, Cc ~ time | treatment),
  PKNCA::PKNCAdose(dose, amt ~ time | treatment),
  intervals = data.frame(
    start = 0, end = 168,
    cmax = TRUE, cmin = TRUE, auclast = TRUE
  )
)
ncaRes <- PKNCA::pk.nca(ncaData)

ncaWide <- as.data.frame(ncaRes) |>
  dplyr::select(treatment, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

knitr::kable(
  ncaWide |>
    dplyr::rename(
      "rs319952" = treatment,
      "Cmax (mg/L)" = cmax,
      "Cmin (mg/L)" = cmin,
      "AUC0-168h (mg/L*h)" = auclast
    ),
  digits = 3,
  caption = "PKNCA steady-state NCA over one 168 h week at GGT = 30 U/L."
)

# The PKNCA weekly AUC must match the published Table 4 AUCweekly values
# (134.5 for A&AG, 196.1 for G at GGT = 30).
aucByTrt <- ncaWide$auclast[match(c("A&AG", "G"), ncaWide$treatment)]
stopifnot(
  abs(aucByTrt[1] - 134.529) < 1,
  abs(aucByTrt[2] - 196.080) < 1
)
```

The PKNCA weekly AUC reproduces the published Table 4 values (134.5 and
196.1 mg/L*h at GGT = 30 U/L) within rounding, independently confirming the
clearance parameterisation.

# Assumptions and deviations

- **rs319952 genotype effect is additive, not fractional.** Results Equation 2
  subtracts 1.4 L/h from CL/F for GG subjects. The exact reproduction of
  Table 4 confirms this reading; a multiplicative interpretation would miss the
  GG-stratum exposures badly.
- **GGT centring value.** The power term is centred on 28.9 U/L, the dataset
  median printed in Equations 1 and 2. The cohort mean (36.4 U/L) is not used.
- **Dosing schedule.** The paper's Table 4 simulation uses 200 mg three times
  weekly; the published Cmax/Cmin only reproduce when the three weekly doses are
  evenly spaced (every 56 h). The clinical regimen (Monday/Wednesday/Friday) is
  not evenly spaced, but the authors' simulation used even spacing, so the
  maintainers use it here to match the reference table.
- **Sex not simulated.** The cohort sex split is not reported; sex was screened
  on CL/F but not retained, so it does not enter the model.
- **Ungenotyped subjects.** 16 of 99 patients were not genotyped at rs319952 and
  were imputed to AG by the authors, entering the fit with the GG indicator = 0.
- **IIV convention.** The Table 3 IIV percentages are read as CVs
  (`omega^2 = log(1 + CV^2)`); the deterministic replications above are
  unaffected by the choice.
- No correction or erratum notice was found for this article as of 2026-10-09.
