---
title: "Meropenem (Zou 2026)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meropenem (Zou 2026)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.5)
library(nlmixr2lib)
library(PKNCA)
library(rxode2)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Model and source

```{r model-meta, echo = FALSE, results = "asis"}
ui <- rxode2::rxode(readModelDb("Zou_2026_meropenem"))
cat("* Citation: ", ui$reference, "\n", sep = "")
cat("* Description: ", ui$description, "\n", sep = "")
```

* Article: <https://doi.org/10.3389/fphar.2026.1778552>

Zou and colleagues developed a one-compartment population PK model for
meropenem in 44 adults with severe postoperative infections managed in a
general ICU, then used Monte Carlo simulation to derive renal-function- and
MIC-stratified dosing recommendations. This vignette reproduces the structural
model, confirms the packaged encoding against a closed-form steady-state
solution, and reproduces the published probability-of-target-attainment (PTA)
recommendations of the paper's Table 3.

# Population

Forty-four adults (31 male, 70.45%) contributed 135 plasma meropenem
concentrations. All were postoperative patients with severe infection requiring
meropenem for more than three days; 93.18% were classified as critically ill
and 15.91% were receiving haemodialysis. The most common presentation was
intra-abdominal infection combined with hospital-acquired pneumonia (47.73%).
Patients received 1.0 g meropenem every 8 h as a prolonged infusion over 2 h or
2.5 h, with TDM-guided adjustment permitted; samples were drawn at steady state
after at least four consecutive doses (Zou 2026 Table 1, Sections 2.1-2.3).

Cockcroft-Gault creatinine clearance had a cohort mean of 47.7 mL/min and
ranged from 11.9 to 136 mL/min (Table 1). Observed meropenem concentrations had
a mean of 16.183 mg/L and ranged from 0.477 to 71.507 mg/L.

The same information is available programmatically via the model's `population`
metadata (`readModelDb("Zou_2026_meropenem")()$population`).

# Source trace

The per-parameter origin is recorded as an in-file comment next to each `ini()`
entry in `inst/modeldb/specificDrugs/Zou_2026_meropenem.R`. The table below
collects them in one place for review.

| Equation / parameter | Value | Source location |
|----------------------|-------|-----------------|
| `lcl` (typical CL)   | 6.472 L/h | Table 2, "Final model" column, CL row (RSE 8.53%) |
| `lvc` (typical Vc)   | 26.69 L   | Table 2, "Final model" column, Vc row (RSE 13.55%) |
| `e_crcl_cl`          | 0.3834    | Table 2, "CRCL on CL" row (RSE 16.54%) |
| `crcl_ref_cl`        | 47.7 mL/min | Section 3.2 covariate equation denominator; equals the Table 1 cohort mean CRCL |
| `var(etalcl)`        | 0.494^2 = 0.244036 | Table 2, omega-CL = 49.4%, footnote "square root of inter-individual variance" |
| `var(etalvc)`        | 0.8199^2 = 0.672236 | Table 2, omega-Vc = 81.99%, same footnote |
| `cov(etalcl, etalvc)`| 0.94 x 0.494 x 0.8199 = 0.380727 | Table 2, omega-CL*Vc row, rho = 0.94 |
| `addSd`              | sqrt(17.4) = 4.17 mg/L | Table 2 residual-error row (see Errata: the column reports a variance) |
| `CL = 6.472 * (CrCL/47.7)^0.3834 * exp(eta_CL)` | n/a | Section 3.2, displayed covariate equation |
| `Vc = 26.69 * exp(eta_Vc)` | n/a | Section 3.2, displayed covariate equation |
| `d/dt(central) = -kel * central` | n/a | Section 3.2, "one-compartment model with linear elimination" |

# Structural verification

Before any cohort simulation, the packaged encoding is checked against results
that do not depend on a random draw.

## Covariate relation

The individual clearance returned by the solver must equal the published
covariate equation exactly.

```{r check-covariate}
mod <- readModelDb("Zou_2026_meropenem")

crcl_grid <- c(11.9, 25, 47.7, 90, 136)
ev_cov <- rxode2::et(amt = 1000, dur = 2, cmt = "central") |>
  rxode2::et(seq(0, 8, by = 1), cmt = "central") |>
  as.data.frame()
ev_cov <- do.call(rbind, lapply(seq_along(crcl_grid), function(i) {
  x <- ev_cov; x$id <- i; x$CRCL <- crcl_grid[i]; x
}))

# zeroRe() removes IIV so cl is the typical value at each CRCL.
cov_chk <- rxode2::rxSolve(rxode2::zeroRe(mod), ev_cov, returnType = "data.frame") |>
  dplyr::group_by(id, CRCL) |>
  dplyr::summarise(cl_solved = mean(cl), .groups = "drop") |>
  dplyr::mutate(cl_published = 6.472 * (CRCL / 47.7)^0.3834,
                rel_err = abs(cl_solved / cl_published - 1))

knitr::kable(
  cov_chk |>
    dplyr::rename("CRCL (mL/min)" = CRCL,
                  "CL solved (L/h)" = cl_solved,
                  "CL published equation (L/h)" = cl_published,
                  "Relative error" = rel_err) |>
    dplyr::select(-id),
  digits = c(0, 1, 4, 4, 12),
  caption = "Solved individual clearance vs the Section 3.2 covariate equation."
)

# Deterministic: same equation, no random draw. A tight bound is correct here.
stopifnot(max(cov_chk$rel_err) < 1e-10)
```

At the reference CRCL of 47.7 mL/min the clearance returns exactly the
published typical value of 6.472 L/h, confirming both the centring constant and
the exponent.

## Closed-form steady-state solution

For a one-compartment model given an intermittent infusion of rate `R0` over
`Tinf` every `tau`, the steady-state concentration has an exact closed form.
Comparing the ODE solve against it uses the *same drawn parameters* on both
sides, so the difference is pure numerical error and a tight bound is
appropriate (unlike the cohort-derived quantities later in this vignette).

```{r check-closed-form}
tau_cf <- 8
tinf_cf <- 2
dose_cf <- 1000

set.seed(20260902)
crcl_cf <- pmin(pmax(rgamma(50, shape = 4, scale = 47.7 / 4), 11.9), 136)

ev_cf <- rxode2::et(amt = dose_cf, ii = tau_cf, until = 240 - tau_cf,
                    dur = tinf_cf, cmt = "central") |>
  rxode2::et(seq(240 - tau_cf, 240, by = 0.25), cmt = "central") |>
  as.data.frame()
ev_cf <- do.call(rbind, lapply(seq_along(crcl_cf), function(i) {
  x <- ev_cf; x$id <- i; x$CRCL <- crcl_cf[i]; x
}))

rxode2::rxSetSeed(99)
sim_cf <- rxode2::rxSolve(mod, ev_cf, returnType = "data.frame") |>
  dplyr::filter(time >= 240 - tau_cf)

closed_form <- function(t_in_interval, cl, vc, tinf, tau, dose) {
  kel <- cl / vc
  r0  <- dose / tinf
  cmin_ss <- (r0 / cl) * (1 - exp(-kel * tinf)) * exp(-kel * (tau - tinf)) /
    (1 - exp(-kel * tau))
  cmax_ss <- cmin_ss * exp(-kel * tinf) + (r0 / cl) * (1 - exp(-kel * tinf))
  ifelse(
    t_in_interval <= tinf,
    cmin_ss * exp(-kel * t_in_interval) +
      (r0 / cl) * (1 - exp(-kel * t_in_interval)),
    cmax_ss * exp(-kel * (t_in_interval - tinf))
  )
}

cf <- sim_cf |>
  dplyr::mutate(
    t_rel = time - (240 - tau_cf),
    Cc_analytic = closed_form(t_rel, cl, vc, tinf_cf, tau_cf, dose_cf),
    rel_err = abs(Cc / Cc_analytic - 1)
  )

cat(sprintf("max relative error vs closed form: %.3g\n", max(cf$rel_err)))
cat(sprintf("median relative error            : %.3g\n", median(cf$rel_err)))

# Numerical error only (identical parameters on both sides), plus a negligible
# approach-to-steady-state residual after 30 dosing intervals. Realised max
# ~1e-5; 1e-3 leaves three orders of magnitude of headroom while still going
# red on any structural error (a wrong kel, volume or infusion handling moves
# this by whole percent).
stopifnot(max(cf$rel_err) < 1e-3)
```

```{r fig-closed-form, echo = FALSE}
cf |>
  dplyr::filter(id <= 6) |>
  tidyr::pivot_longer(c(Cc, Cc_analytic), names_to = "source", values_to = "conc") |>
  dplyr::mutate(source = dplyr::recode(source, Cc = "rxode2 ODE solve",
                                       Cc_analytic = "closed form")) |>
  ggplot(aes(t_rel, conc, colour = source, linetype = source)) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~id, labeller = label_both) +
  labs(x = "Time within the steady-state dosing interval (h)",
       y = "Meropenem concentration (mg/L)",
       colour = NULL, linetype = NULL,
       title = "ODE solve vs closed-form steady state (1 g q8h, 2 h infusion)")
```

# Virtual cohort

Original observed data are not publicly available. The cohort below approximates
the published demographics: creatinine clearance is drawn from a gamma
distribution with the Table 1 cohort mean of 47.7 mL/min, truncated to the
observed 11.9-136 mL/min range (the paper does not report the distributional
shape; see Assumptions).

```{r cohort}
# `set.seed()` seeds R's RNG, which draws the covariates. It does NOT seed
# rxode2's simulation RNG for the etas, and rxode2's streams are partitioned
# per solver thread -- so the eta draws differ on a machine with a different
# thread count. Every assertion below is written to hold for any cohort the
# model can produce (see pattern 12 of known-vignette-failure-patterns.md).
set.seed(20260902)

n_sub <- 200
crcl_cohort <- pmin(pmax(rgamma(n_sub, shape = 4, scale = 47.7 / 4), 11.9), 136)

cat(sprintf("simulated CRCL: mean %.1f, range %.1f-%.1f mL/min\n",
            mean(crcl_cohort), min(crcl_cohort), max(crcl_cohort)))
cat("published (Table 1): mean 47.7, range 11.9-136 mL/min\n")

tau_ref  <- 8
tinf_ref <- 2
dose_ref <- 1000

ev_one <- rxode2::et(amt = dose_ref, ii = tau_ref, until = 240 - tau_ref,
                     dur = tinf_ref, cmt = "central") |>
  rxode2::et(seq(240 - tau_ref, 240, by = 0.1), cmt = "central") |>
  as.data.frame()

events <- do.call(rbind, lapply(seq_len(n_sub), function(i) {
  x <- ev_one
  x$id <- i
  x$CRCL <- crcl_cohort[i]
  x$treatment <- "1 g q8h (2 h infusion)"
  x
}))

stopifnot(!anyDuplicated(unique(events[, c("id", "time", "evid")])))
```

# Simulation

```{r simulate}
rxode2::rxSetSeed(20260902)
sim <- rxode2::rxSolve(mod, events = events, keep = c("treatment")) |>
  as.data.frame() |>
  dplyr::filter(time >= 240 - tau_ref)

stopifnot(nrow(sim) > 0, all(sim$Cc >= 0))
```

## Steady-state concentration profile

The paper's Figure 2 is a prediction-corrected VPC against the observed TDM
samples, which are not public. The panel below is the corresponding simulated
percentile envelope for the study's own regimen, with the observed
concentration range from Table 1 overlaid.

```{r fig-vpc}
vpc <- sim |>
  dplyr::mutate(t_rel = time - (240 - tau_ref)) |>
  dplyr::group_by(t_rel) |>
  dplyr::summarise(Q05 = quantile(Cc, 0.05), Q50 = quantile(Cc, 0.50),
                   Q95 = quantile(Cc, 0.95), .groups = "drop")

ggplot(vpc, aes(t_rel, Q50)) +
  geom_ribbon(aes(ymin = Q05, ymax = Q95), alpha = 0.25, fill = "steelblue") +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = c(0.477, 71.507), linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 16.183, linetype = "dotted", colour = "firebrick") +
  scale_y_log10() +
  labs(x = "Time within the steady-state dosing interval (h)",
       y = "Meropenem concentration (mg/L)",
       title = "Simulated steady-state profile, 1 g q8h over 2 h",
       caption = paste("Dashed: observed concentration range 0.477-71.507 mg/L.",
                       "Dotted: observed mean 16.183 mg/L (Zou 2026 Table 1)."))
```

```{r check-range}
# The simulated 5th-95th envelope should sit inside the observed range and
# straddle the observed mean. These are broad, absolute bounds taken from the
# paper's own Table 1, not from a single simulation run.
stopifnot(
  min(vpc$Q05) > 0.477 / 5,
  max(vpc$Q95) < 71.507 * 2,
  median(sim$Cc) > 0.477,
  median(sim$Cc) < 71.507
)
cat(sprintf("simulated median Cc over the interval: %.1f mg/L (observed mean 16.183)\n",
            median(sim$Cc)))
```

# PKNCA validation

Steady-state NCA over the final dosing interval (PKNCA recipe 3). The source
paper reports no NCA parameter table, so the reference for comparison is the
model-implied identity that must hold exactly for linear PK at steady state:

  AUC(0-tau) = Dose / CL

```{r pknca}
sim_nca <- sim |>
  dplyr::filter(!is.na(Cc)) |>
  dplyr::select(id, time, Cc, treatment)

# Guarantee a record at the interval start so PKNCA can anchor the interval.
ss_start <- 240 - tau_ref
sim_nca <- dplyr::bind_rows(
  sim_nca,
  sim_nca |> dplyr::distinct(id, treatment) |>
    dplyr::mutate(time = ss_start, Cc = NA_real_)
) |>
  dplyr::arrange(id, treatment, time) |>
  dplyr::group_by(id, treatment, time) |>
  dplyr::slice(1) |>
  dplyr::ungroup() |>
  dplyr::filter(!is.na(Cc))

conc_obj <- PKNCA::PKNCAconc(sim_nca, Cc ~ time | treatment + id,
                             concu = "mg/L", timeu = "h")

dose_df <- events |>
  dplyr::filter(evid == 1, time == max(events$time[events$evid == 1])) |>
  dplyr::select(id, time, amt, treatment)

dose_obj <- PKNCA::PKNCAdose(dose_df, amt ~ time | treatment + id, doseu = "mg")

intervals <- data.frame(
  start   = ss_start,
  end     = ss_start + tau_ref,
  cmax    = TRUE,
  tmax    = TRUE,
  cmin    = TRUE,
  auclast = TRUE,
  cav     = TRUE
)

nca_res <- PKNCA::pk.nca(PKNCA::PKNCAdata(conc_obj, dose_obj, intervals = intervals))
nca_wide <- as.data.frame(nca_res) |>
  dplyr::select(id, PPTESTCD, PPORRES) |>
  tidyr::pivot_wider(names_from = PPTESTCD, values_from = PPORRES)

stopifnot(nrow(nca_wide) == n_sub, all(c("auclast", "cmax", "cmin") %in% names(nca_wide)))
```

## Comparison against the model-implied exposure identity

```{r nca-identity}
cl_by_id <- sim |> dplyr::group_by(id) |> dplyr::summarise(cl = mean(cl), .groups = "drop")

chk <- nca_wide |>
  dplyr::inner_join(cl_by_id, by = "id") |>
  dplyr::mutate(auc_expected = dose_ref / cl,
                pct_diff = 100 * (auclast / auc_expected - 1))
stopifnot(nrow(chk) == n_sub)

knitr::kable(
  tibble::tibble(
    Statistic = c("Median % difference", "90th percentile |% difference|",
                  "Median AUC(0-tau) (mg*h/L)", "Median Dose/CL (mg*h/L)",
                  "Median Cmax,ss (mg/L)", "Median Cmin,ss (mg/L)"),
    Value = c(median(chk$pct_diff), quantile(abs(chk$pct_diff), 0.9),
              median(chk$auclast), median(chk$auc_expected),
              median(chk$cmax), median(chk$cmin))
  ) |> dplyr::rename("NCA parameter" = Statistic),
  digits = 3,
  caption = "PKNCA steady-state exposure vs the model identity AUC(0-tau) = Dose/CL."
)

# Both sides use the same drawn CL, so the only discrepancy is linear-trapezoid
# error on the post-infusion decay, which varies per subject with the curvature.
# Assert on the centre and a robust quantile, not the extreme.
stopifnot(
  abs(median(chk$pct_diff)) < 2,
  quantile(abs(chk$pct_diff), 0.9) < 5
)
```

The paper reports no NCA table, so no published-vs-simulated NCA comparison is
possible; this identity check is the substitute and confirms that dose, volume
and clearance units are mutually consistent (a mis-transcribed clearance or a
dose-unit error moves `AUC/(Dose/CL)` by tens of percent).

# Monte Carlo simulation: reproducing Table 3

This is the paper's principal result and a zero-parameter gate: Table 3 lists,
for each PK/PD target, MIC and renal-function stratum, which of the 18
evaluated regimens reach a PTA of at least 90%. Nothing is fitted here -- the
packaged model either reproduces the published classification or it does not.

Because the targets are all of the form "fraction of the dosing interval with
free concentration above a threshold", a single solve per (stratum, regimen)
yields every MIC and every target: `100% fT > X` is `Cmin,ss > X` and
`q% fT > X` is the fraction of interval time above `X`. Protein binding is
about 2%, so the authors treated total concentrations as free (Section 2.6).

```{r pta-grid}
strata <- list("<10" = c(1, 10), "10-25" = c(10, 25), "26-50" = c(26, 50),
               "51-90" = c(51, 90), "91-140" = c(91, 140))
regs <- expand.grid(dose = c(500, 1000, 2000), ii = c(6, 8, 12), tinf = c(0.5, 2))

sim_arm <- function(dose, ii, tinf, crcl) {
  tend <- ceiling(240 / ii) * ii
  ev <- rxode2::et(amt = dose, ii = ii, until = tend - ii, dur = tinf, cmt = "central") |>
    rxode2::et(seq(tend - ii, tend, by = 0.05), cmt = "central") |>
    as.data.frame()
  full <- do.call(rbind, lapply(seq_along(crcl), function(i) {
    x <- ev; x$id <- i; x$CRCL <- crcl[i]; x
  }))
  rxode2::rxSolve(mod, full, returnType = "data.frame")
}

pta_rows <- list()
k <- 0L
for (sn in names(strata)) {
  b <- strata[[sn]]
  set.seed(20260902)
  crcl_s <- runif(200, b[1], b[2])
  for (r in seq_len(nrow(regs))) {
    rxode2::rxSetSeed(4242)   # common random numbers across regimens within a stratum
    a <- sim_arm(regs$dose[r], regs$ii[r], regs$tinf[r], crcl_s) |>
      dplyr::group_by(id) |>
      dplyr::summarise(cmin = min(Cc),
                       f1 = mean(Cc > 1), f2 = mean(Cc > 2), f4 = mean(Cc > 4),
                       f8 = mean(Cc > 8), f16 = mean(Cc > 16), f32 = mean(Cc > 32),
                       .groups = "drop")
    k <- k + 1L
    pta_rows[[k]] <- data.frame(
      stratum = sn, dose = regs$dose[r], ii = regs$ii[r], tinf = regs$tinf[r],
      pta_40_1 = mean(a$f1 >= .4), pta_40_2 = mean(a$f2 >= .4),
      pta_40_4 = mean(a$f4 >= .4), pta_40_8 = mean(a$f8 >= .4),
      pta_100_1 = mean(a$cmin > 1), pta_100_2 = mean(a$cmin > 2),
      pta_100_4 = mean(a$cmin > 4), pta_100_8 = mean(a$cmin > 8),
      pta_50x4_1 = mean(a$f4 >= .5), pta_50x4_2 = mean(a$f8 >= .5),
      pta_50x4_4 = mean(a$f16 >= .5), pta_50x4_8 = mean(a$f32 >= .5),
      pta_100x4_1 = mean(a$cmin > 4), pta_100x4_2 = mean(a$cmin > 8),
      pta_100x4_4 = mean(a$cmin > 16), pta_100x4_8 = mean(a$cmin > 32))
  }
}
pta <- do.call(rbind, pta_rows)
stopifnot(nrow(pta) == length(strata) * nrow(regs))
```

## The unambiguous cells: "All" and "None"

Table 3's footnote 5 defines two unambiguous cell values: "All" means every
evaluated regimen reached PTA >= 90%, and "None" means no regimen did. Those
are exhaustive claims about all 18 regimens, so each cell yields 18 checkable
classifications.

```{r pta-all-none}
cell <- function(tg, mic, st) {
  cn <- paste0("pta_", tg, "_", mic)
  v <- pta[pta$stratum == st, cn]
  if (length(v) != nrow(regs)) stop("no unique regimen set for ", tg, " MIC ", mic, " ", st)
  v
}

all_cells <- list(c("40", 1, "<10"), c("40", 1, "10-25"), c("40", 1, "26-50"),
                  c("40", 1, "51-90"), c("40", 1, "91-140"),
                  c("40", 2, "<10"), c("40", 2, "10-25"), c("40", 2, "26-50"),
                  c("40", 4, "<10"), c("40", 4, "10-25"),
                  c("100", 1, "<10"), c("100", 1, "10-25"), c("100", 2, "<10"),
                  c("50x4", 1, "<10"), c("50x4", 1, "10-25"))
none_cells <- list(c("50x4", 8, "26-50"), c("50x4", 8, "51-90"), c("50x4", 8, "91-140"),
                   c("100x4", 4, "51-90"), c("100x4", 4, "91-140"),
                   c("100x4", 8, "10-25"), c("100x4", 8, "26-50"),
                   c("100x4", 8, "51-90"), c("100x4", 8, "91-140"))

ok_all  <- unlist(lapply(all_cells,  function(z) cell(z[1], z[2], z[3]) >= 0.90))
ok_none <- unlist(lapply(none_cells, function(z) cell(z[1], z[2], z[3]) <  0.90))
stopifnot(length(ok_all) == 15 * nrow(regs), length(ok_none) == 9 * nrow(regs))

agreement <- (sum(ok_all) + sum(ok_none)) / (length(ok_all) + length(ok_none))

knitr::kable(
  tibble::tibble(
    `Table 3 cell type` = c("\"All\" cells (every regimen >= 90%)",
                            "\"None\" cells (no regimen >= 90%)", "Overall"),
    `Cells` = c(length(all_cells), length(none_cells),
                length(all_cells) + length(none_cells)),
    `Regimen classifications` = c(length(ok_all), length(ok_none),
                                  length(ok_all) + length(ok_none)),
    `Agreeing` = c(sum(ok_all), sum(ok_none), sum(ok_all) + sum(ok_none)),
    `Agreement` = c(mean(ok_all), mean(ok_none), agreement)
  ),
  digits = 4,
  caption = "Reproduction of the exhaustive cells of Zou 2026 Table 3."
)

# 432 classifications; realised 432/432. The bound admits 12 disagreements,
# which is far outside cohort-draw noise but still goes red if the model is
# structurally wrong (a wrong CL or Vc moves whole blocks of cells).
stopifnot(agreement > 0.97)
```

## The named regimens

The remaining Table 3 cells name a specific regimen (or a small set). Those are
not exhaustive claims -- the table does not say the named regimen is the only
one that works -- but each named regimen must itself reach PTA >= 90%.

```{r pta-named}
named <- read.csv(text = "tg,mic,st,dose,ii,tinf
40,2,51-90,500,8,0.5
40,2,91-140,500,8,0.5
40,4,26-50,500,8,0.5
40,4,51-90,500,8,2
40,4,51-90,1000,12,0.5
40,4,91-140,500,6,0.5
40,8,<10,500,8,0.5
40,8,10-25,500,6,0.5
40,8,26-50,1000,8,0.5
40,8,51-90,1000,8,0.5
40,8,91-140,1000,6,0.5
100,1,26-50,500,8,0.5
100,1,51-90,500,8,0.5
100,1,91-140,500,6,0.5
100,2,10-25,500,8,0.5
100,2,26-50,500,8,0.5
100,2,51-90,500,6,0.5
100,2,51-90,1000,6,0.5
100,2,51-90,2000,6,0.5
100,2,91-140,500,6,2
100,4,<10,500,12,2
100,4,10-25,500,8,0.5
100,4,26-50,500,6,0.5
100,4,51-90,1000,6,0.5
100,4,51-90,2000,6,0.5
100,4,51-90,2000,8,0.5
100,4,91-140,1000,6,2
100,8,<10,500,8,0.5
100,8,<10,1000,12,2
100,8,10-25,1000,8,2
100,8,26-50,1000,6,0.5
100,8,26-50,2000,6,0.5
100,8,51-90,2000,6,0.5
100,8,91-140,2000,6,2
50x4,1,26-50,500,8,0.5
50x4,1,51-90,500,8,2
50x4,1,91-140,500,6,0.5
50x4,2,<10,500,8,0.5
50x4,2,10-25,500,6,0.5
50x4,2,26-50,1000,8,0.5
50x4,2,51-90,1000,8,0.5
50x4,2,91-140,1000,6,0.5
50x4,4,<10,1000,8,0.5
50x4,4,10-25,1000,6,0.5
50x4,4,10-25,2000,12,0.5
50x4,4,26-50,2000,6,0.5
50x4,4,26-50,2000,8,0.5
50x4,4,51-90,2000,6,0.5
50x4,4,51-90,2000,8,0.5
50x4,4,91-140,2000,6,0.5
50x4,8,<10,2000,6,0.5
50x4,8,<10,2000,8,0.5
50x4,8,10-25,2000,6,0.5
100x4,1,<10,500,12,2
100x4,1,10-25,500,8,2
100x4,1,26-50,500,6,0.5
100x4,1,51-90,1000,6,0.5
100x4,1,91-140,2000,6,0.5
100x4,2,<10,500,8,2
100x4,2,10-25,1000,8,2
100x4,2,26-50,1000,6,0.5
100x4,2,51-90,2000,6,0.5
100x4,2,91-140,2000,6,2
100x4,4,<10,1000,8,2
100x4,4,10-25,2000,8,2
100x4,4,26-50,2000,6,0.5
100x4,8,<10,2000,8,2",
  colClasses = c("character", "integer", "character", "numeric", "numeric", "numeric"))

named$pta <- mapply(function(tg, mic, st, dose, ii, tinf) {
  r <- pta[pta$stratum == st & pta$dose == dose & pta$ii == ii & pta$tinf == tinf, ]
  if (nrow(r) != 1L) stop("no unique regimen row for ", tg, " MIC ", mic, " ", st)
  r[[paste0("pta_", tg, "_", mic)]]
}, named$tg, named$mic, named$st, named$dose, named$ii, named$tinf)
stopifnot(nrow(named) == 67, !anyNA(named$pta))

named$ok <- named$pta >= 0.90
evaluable  <- named[named$st != "<10", ]
exploratory <- named[named$st == "<10", ]

knitr::kable(
  tibble::tibble(
    `Stratum group` = c("Evaluable (10-25, 26-50, 51-90, 91-140 mL/min)",
                        "Exploratory (<10 mL/min; no observed data)", "All"),
    `Named regimens` = c(nrow(evaluable), nrow(exploratory), nrow(named)),
    `Reaching PTA >= 90%` = c(sum(evaluable$ok), sum(exploratory$ok), sum(named$ok)),
    `Median PTA` = c(median(evaluable$pta), median(exploratory$pta), median(named$pta)),
    `Minimum PTA` = c(min(evaluable$pta), min(exploratory$pta), min(named$pta))
  ),
  digits = 3,
  caption = "Every regimen named in Zou 2026 Table 3, evaluated against the packaged model."
)

# Realised 67/67 with a minimum PTA of exactly 0.900 -- i.e. some named
# regimens sit exactly on the paper's own 90% boundary, so a different cohort
# draw will push a handful either side of it. Assert on the count and the
# centre with headroom, never on "all 67".
stopifnot(
  sum(named$ok) >= 58,
  median(named$pta) > 0.92,
  min(named$pta) > 0.80
)
```

```{r fig-pta}
pta |>
  dplyr::filter(tinf == 2, dose %in% c(500, 1000, 2000), ii == 8) |>
  tidyr::pivot_longer(dplyr::starts_with("pta_100_"),
                      names_to = "mic", values_to = "PTA") |>
  dplyr::mutate(mic = factor(sub("pta_100_", "", mic), levels = c("1", "2", "4", "8")),
                stratum = factor(stratum, levels = names(strata)),
                dose = factor(paste0(dose, " mg"),
                              levels = c("500 mg", "1000 mg", "2000 mg"))) |>
  ggplot(aes(mic, PTA, fill = dose)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0.9, linetype = "dashed") +
  facet_wrap(~stratum, nrow = 1) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "MIC (mg/L)", y = "PTA", fill = NULL,
       title = "PTA for 100% fT > MIC, q8h over a 2 h infusion, by CRCL stratum",
       caption = "Replicates the 100% fT > MIC panels of Zou 2026 Figure 3 / Table 3.")
```

# Assumptions and deviations

## Errata: the residual-error row of Table 2

Table 2's residual-error row is internally inconsistent as printed, and
resolving it is the only judgement call in this extraction. The row is labelled
**"Additional error"**, unit **mg/L**, with a final-model value of **17.4**
(RSE 12.0%) and a bootstrap median of **4.64** (95% CI 2.99-6.34).

1. **The two columns cannot both be standard deviations.** 17.4 lies far
   outside 2.99-6.34, which directly contradicts Section 3.3's statement that
   "all final parameter values fell within the 95% confidence intervals". Every
   other row in Table 2 is mutually consistent between the two columns.
2. **They reconcile exactly as variance-then-SD.** `sqrt(17.4) = 4.17` mg/L,
   which sits inside the bootstrap CI. The final-model column reports the
   residual variance; the bootstrap column reports the standard deviation.
3. **The error is additive, not proportional.** Sections 2.5 and 3.2 describe
   the retained model as "proportional", but that single word is contradicted
   by the table label ("Additional" = additive), by the unit (mg/L), and by the
   goodness-of-fit plot. Digitising the 102 resolvable markers of Figure 1b
   (observed vs individual-predicted) gives a residual SD that is essentially
   flat in concentration -- 3.1, 3.5, 6.4, 5.2, 3.2, 2.4 and 0.7 mg/L in IPRED
   bins of 0-5, 5-10, 10-15, 15-20, 20-30, 30-45 and 45-75 mg/L -- while the
   *relative* SD falls monotonically from 84% to 1.3%. A proportional error of
   any magnitude is refuted: 17.4% would leave visible scatter at the top of
   the range where the points lie almost exactly on the line, and 4.17% would
   leave almost none at the bottom where the scatter is widest.
4. **The digitised magnitude agrees.** The overall digitised residual SD is
   4.44 mg/L, bracketing `sqrt(17.4) = 4.17` and the bootstrap median 4.64.

The model therefore encodes `addSd = 4.17` mg/L as a purely additive residual
error. A reader who prefers the literal prose reading would set
`propSd = 0.174` instead; the two are not interchangeable, and the evidence
above is why this extraction chose the additive form. Note that the residual
error affects neither the typical-value predictions nor the PTA reproduction
above, both of which are computed from model predictions without residual
noise, exactly as in the source paper.

## Other assumptions and deviations

- **Inter-individual variability convention.** Table 2 reports omega-CL and
  omega-Vc as percentages (49.4% and 81.99%) and its footnote defines them as
  the "square root of inter-individual variance". They are therefore taken at
  face value as the SD of the eta: `var(etalcl) = 0.494^2` and
  `var(etalvc) = 0.8199^2`. They are *not* re-interpreted as %CV requiring the
  `omega^2 = log(1 + CV^2)` transform, which would give 0.218 and 0.514
  instead. The footnote is explicit, so the direct reading is used.
- **IIV correlation.** The Table 2 footnote glosses rho as the "correlation of
  interindividual variability of central volume and intercompartmental
  clearance", but the model is one-compartment and has no intercompartmental
  clearance; the row label is `omega CL*Vc` and the preceding footnote clause
  says "covariance between interindividual variability of central volume and
  clearance". The rho = 0.94 is therefore taken as corr(eta_CL, eta_Vc). The
  resulting 2x2 covariance matrix is positive definite (eigenvalues 0.895 and
  0.021), so no correlation nudge was needed.
- **Conflicting demographics.** Section 3.1 prose and Table 1 disagree: the
  prose reports mean age 68.5 +/- 14.8 years (range 20.0-91.0) and mean weight
  56.7 +/- 9.04 kg (range 40.0-81.0), while Table 1 reports mean age 66.0
  (range 20.0-83.0) and mean weight 58.0 (range 45.0-81.0). Similarly the prose
  reports mean CRCL 54.3 +/- 29.6 mL/min while Table 1 reports 47.7, over the
  identical 11.9-136 range. Neither age nor weight enters the model. For CRCL
  the value 47.7 is used because it is the number printed inside the Section 3.2
  covariate equation itself. Both readings are recorded in the model's
  `population` metadata.
- **CRCL distribution in the validation cohort.** The paper reports only the
  mean and range of creatinine clearance, not its distributional shape. The
  cohort here draws CRCL from a gamma distribution with mean 47.7 mL/min
  (shape 4), truncated to 11.9-136 mL/min, which reproduces the published mean
  and range.
- **CRCL distribution within the PTA strata.** The paper does not state how
  CRCL was sampled inside each simulation stratum. Each stratum here is drawn
  uniformly across its stated band. For the open-ended "<10 mL/min" stratum a
  band of 1-10 mL/min is used, consistently with the other strata being uniform
  over their full band. This choice matters only for that stratum: under a
  narrower 5-10 mL/min band, four of its twelve named regimens fall to
  PTA 0.845-0.890 instead of clearing 90%. The paper itself flags this stratum
  as exploratory, "due to the absence of observed data in this renal function
  range" (Section 2.6), and the assertions above therefore report the
  exploratory and evaluable strata separately.
- **Steady state.** All simulations dose for 240 h before the evaluated
  interval, which is more than 20 half-lives for every subject in the evaluable
  strata, so the evaluated interval is at steady state.
- **Free vs total concentration.** Meropenem protein binding is about 2%, and
  the authors treated total plasma concentrations as free concentrations in the
  PK/PD target calculations (Section 2.6). The same identification is used
  here, so `Cc` serves as both.
- **No published NCA comparison.** The paper reports no NCA parameter table, so
  the PKNCA section is validated against the model-implied identity
  AUC(0-tau) = Dose/CL rather than against published Cmax / AUC values.
- **No supplement or erratum.** EuropePMC reports `hasSuppl = N` and no
  correction notice for PMC13102762; every value above comes from the main
  article.
